### 2011-08-23 APENDICE.tex
iniciado repositório. tentando com somente estes arquivos:
	new file:   APENDICE.tex
	new file:   agradecimentos.tex
	new file:   bibdatabase.bib
	new file:   conclusoes.tex
	new file:   dedicatoria.tex
	new file:   devEresultados.tex
	new file:   estiloifsc.sty
	new file:   fundamentos.tex
	new file:   introducao.tex
	new file:   mapaSitio.tex
	new file:   pretextual.tex
	new file:   resumoeabstract.tex

\section{Web}

\subsection{Protocolos e Tecnologias}

\subsubsection{Tecnologias padrão}
HTML, CSS, Javascript, PHP, Mysql.

\subsubsection{Frameworks}
Django, Ruby etc.

\subsection{Internet em Números}

Quantos usuários X, quantos sites/minuto cada usuáriosvisita em média.
Quais os computadores em uso hoje tanto no terceiro mundo quanto no primeiro,
processamento, memória RAM, resolução etc.

\subsection{e-democracy}

Focamos nossos desenvolvimentos em ferramentas de inclusão
e monitoramento político,i.e. diagnosticamos a necessidade
e investimos em \emph{e-democracy}, como são chamadas tais ferramentas atualmente.

### 2011-08-23 agradecimentos.tex
iniciado repositório. tentando com somente estes arquivos:
	new file:   APENDICE.tex
	new file:   agradecimentos.tex
	new file:   bibdatabase.bib
	new file:   conclusoes.tex
	new file:   dedicatoria.tex
	new file:   devEresultados.tex
	new file:   estiloifsc.sty
	new file:   fundamentos.tex
	new file:   introducao.tex
	new file:   mapaSitio.tex
	new file:   pretextual.tex
	new file:   resumoeabstract.tex

\begin{itemize}
	\item Aos meus pais ...
	\item Aos amigos ...
	\item Aos professores que ...
\end{itemize}

### 2011-08-23 conclusoes.tex
iniciado repositório. tentando com somente estes arquivos:
	new file:   APENDICE.tex
	new file:   agradecimentos.tex
	new file:   bibdatabase.bib
	new file:   conclusoes.tex
	new file:   dedicatoria.tex
	new file:   devEresultados.tex
	new file:   estiloifsc.sty
	new file:   fundamentos.tex
	new file:   introducao.tex
	new file:   mapaSitio.tex
	new file:   pretextual.tex
	new file:   resumoeabstract.tex

%% ------------------------------------------------------------------------- %%
\chapter{Conclusoes e trabalhos futuros} %Nome do capítulo.
\label{cap:intro}

Conclusões sobre o material apresentado nesta dissertação e trabalhos a serem realizados.

### 2011-08-23 devEresultados.tex
iniciado repositório. tentando com somente estes arquivos:
	new file:   APENDICE.tex
	new file:   agradecimentos.tex
	new file:   bibdatabase.bib
	new file:   conclusoes.tex
	new file:   dedicatoria.tex
	new file:   devEresultados.tex
	new file:   estiloifsc.sty
	new file:   fundamentos.tex
	new file:   introducao.tex
	new file:   mapaSitio.tex
	new file:   pretextual.tex
	new file:   resumoeabstract.tex

%% ------------------------------------------------------------------------- %%
\chapter{Desenvolvimentos e resultados} %Nome do capítulo.
\label{cap:intro} %Rótulo para futura referência ao capítulo. Em qualquer lugar da tese, você poderá citar este capítulo através de ~\ref{cap:introducao}. Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})

\section{Web}

Difusão de informação com ênfase na facilitação
da apropriação de tecnologias e de instancias políticas.

\subsection{Sítios}

FDDCA

Ferramenta de comunicação

Cadastro dos pontos

Meu site pessoal

\subsection{Conteúdos}

\subsubsection{textos}
Carta mídias livres

Artigos,

Tutoriais

\subsection{Articulação}

Em rede.

\section{Áudio}

\subsection{ABT}

Manual ABT

Linkar com ABD

\subsection{FDP}

Artigo FDP

\subsection{Experimentos abertos}

Audioexperiments. Estúdio livre.

\subsection{Tutoriais}

Screencasts. Textos.

### 2011-08-23 fundamentos.tex
iniciado repositório. tentando com somente estes arquivos:
	new file:   APENDICE.tex
	new file:   agradecimentos.tex
	new file:   bibdatabase.bib
	new file:   conclusoes.tex
	new file:   dedicatoria.tex
	new file:   devEresultados.tex
	new file:   estiloifsc.sty
	new file:   fundamentos.tex
	new file:   introducao.tex
	new file:   mapaSitio.tex
	new file:   pretextual.tex
	new file:   resumoeabstract.tex

%% ------------------------------------------------------------------------- %%
\chapter{Fundamentos teóricos e técnicos} %Nome do capítulo.
\label{cap:intro} %Rótulo para futura referência ao capítulo. Em qualquer lugar da tese, você poderá citar este capítulo através de ~\ref{cap:introducao}. Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})

\section{Áudio}

\subsection{Som como fenômeno físico}

Ver livros em casa.

\subsection{Som digitalizado}

Ver livros em casa.

Representação digital do som.

Padrões PCM. Padrões compactados.

\subsection{Processamento de áudio}

Ver livros em casa.

Procedimentos básicos. Processamento
espectral. Outros processamentos.

Vincular a um apêndice com excertos de código.

\section{Linguagens e modularidade}

Discursar brevemente sobre as diversas linguagens,
inclusive voltadas para áudio. E sobre a reutilização
de trechos de código.

\section{Sistema Operacional e Web}

\subsection{Breve Histórico}

O hardware estava ficando mais ambundante, sendo replicado em 
vários lugares. Modelos eram vendidos com diferenças de hardware 
de mínimas a discrepantes. Neste contexto, começou-se a
pensar em uma camada de código que abstraísse o hardware, assim
como a linguagem C se propôs.

Iniciou-se assim o desenvolvimento de Sistemas Operacionais (SO).
Um SO é um software que gerencia os recursos do computador e ajuda
também em abstrair o hardware. Orientamos a esta excelênte linha do
tempo da Wikipédia em língua inglesa \cite{solinhadotempo}:

\subsection{Unix e GNU/LINUX}

Resumo de utilização. No apêndice o resumo de:
The Unix programing envinoment
e
Running Linux

\subsection{Web e Browsers}

Dados do livrinho do google.

Possibilidades de uso.

Breve Histórico.

### 2011-08-23 introducao.tex
iniciado repositório. tentando com somente estes arquivos:
	new file:   APENDICE.tex
	new file:   agradecimentos.tex
	new file:   bibdatabase.bib
	new file:   conclusoes.tex
	new file:   dedicatoria.tex
	new file:   devEresultados.tex
	new file:   estiloifsc.sty
	new file:   fundamentos.tex
	new file:   introducao.tex
	new file:   mapaSitio.tex
	new file:   pretextual.tex
	new file:   resumoeabstract.tex

%% ------------------------------------------------------------------------- %%
\chapter{Introdução} %Nome do capítulo.
\label{cap:intro} %Rótulo para futura referência ao capítulo. Em qualquer lugar da tese, você poderá citar este capítulo através de ~\ref{cap:introducao}. Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})

O livre fluxo de informação

%% ------------------------------------------------------------------------- %%

\subsection{Código Aberto e Software Livre} %Nome da seção.

Qualquer licença de software livre é também uma licença de código aberto (Open Source), a diferença entre as duas nomenclaturas reside essencialmente na sua apresentação e na filosofia.

\subsection{Softwares para compartilhamento} %Nome da seção.
\label{sec:soft_compar} %Rótulo para futura referência à seção. Em qualquer lugar da tese, você poderá citar esta seção através de \label{sec:consideracoes_preliminares}. O mesmo vale para todas as próximas seções e capítulos.

Softwares para dar suporte a esta conectividade.

%% ------------------------------------------------------------------------- %%
\subsection{Licenças livres para conteúdo}
\label{sec:licencas_livres}

Principais tipos de licenças

Licenças voltadas para software.

Licenças voltadas para outras mídias.

\subsubsection{Principais licenças}
\label{sec:princ_licencas}

\section{Contextualização}
\label{sec:context}

Descentralização: Fator caminhão e espalhamento de informação. Descentralização
é uma constante nas estruturas e ferramentas mais relevantes como redes
sociais, comunicadores que agora funcionam por broadcast e protocolos de
compartilhamento de dados, em especial arquivos inteiros, unidades mais
bem organizadas.

Aumento da conectividade através de avanços em telecomunicações e transportes.

\subsubsection{Dinâmica de desenvolvimento}
\label{sec:din_dev}

\subsubsection{Compartilhamento de Tec}
\label{sec:comp_tec}

Terreno fértil -> semear

\subsection{Mutimídia} % Onde se detecta a ênfase em áudio aqui???
\label{sec:multimidia}

\subsubsection{Gargalo}
\label{sec:gargalo}

\subsubsection{Comparacao das soluções em software livre e proprietários}
\label{sec:sl_prop}

\subsection{Identificação de tarefas pouco suportadas}
\label{sec:tarefas_n_sup}

\section{Objetivos}
\label{sec:objetivos}

\subsection{Difusão de práticas}
\label{sec:difusao}

\subsubsection{Tutoriais}
\label{sec:tutoriais}

\begin{itemize}
    \item Tutoriais para programação em python.

Screen casts e texto.

    \item Tutoriais para fazer plugins.
\end{itemize}

\subsubsection{Uso de software livre}
\label{sec:uso_sl}

\subsubsection{Estudo de caso da difusão de conteúdos livres}
\label{sec:dif_sl}

Estamos desenvolvendo trabalhos em análise de comparação
entre sucessões de tendências filosóficas ~\cite{philome} e
musicais ~\cite{musime} e trabalhos em processamento de escrita
e fala ~\cite{rede-associacoes, complenet, enfmc, ifsc}.

Os estudos subsequentes a este trabalhos devem desenvolver
um estudo sobre a dinâmica de geração e propagação de conteúdo na rede.
Comparando textos e ustilizando medidas de similaridade separa-se
o que um texto não contém e o que é semelhante, i.e. servem de propagação
um do outro. A criatividade será enfocada também como estudos de dinâmicas
de atributos de redes complexas de palavras, como as redes de associação ~\cite{rede-associacoes}.

\subsection{Extensão de recursos em software livre - multimídia}
\label{sec:extensao}

\subsubsection{Manipulação e tratamento de áudio}
\label{sec:manip-audio}

\subsubsection{Plugins feitos}
\label{sec:tutoriais}

\subsubsection{Performance em tempo real}
\label{sec:perf}

\subsubsection{Síntese de sons e estruturas musicais}
\label{sec:sintese}

\subsubsection{Analise}
\label{sec:analise}

### 2011-08-23 pretextual.tex
iniciado repositório. tentando com somente estes arquivos:
	new file:   APENDICE.tex
	new file:   agradecimentos.tex
	new file:   bibdatabase.bib
	new file:   conclusoes.tex
	new file:   dedicatoria.tex
	new file:   devEresultados.tex
	new file:   estiloifsc.sty
	new file:   fundamentos.tex
	new file:   introducao.tex
	new file:   mapaSitio.tex
	new file:   pretextual.tex
	new file:   resumoeabstract.tex

\capa

% \pretextualchapter{}

\folhaderosto
% 
\pretextualchapter{}
	\thispagestyle{plain}
	\noindent \parbox{5.7in}{\centering AUTORIZO A REPRODUÇÃO E DIVULGAÇÃO TOTAL OU PARCIAL DESTE TRABALHO, POR QUALQUER MEIO CONVENCIONAL OU ELETRÔNICO, PARA FINS DE ESTUDO E PESQUISA, DESDE QUE CITADA A FONTE.}

\pretextualchapter{}
	Lado dedicado à folha de aprovação. Apagar isso antes de imprimir a versão oficial

\pretextualchapter{Dedicatória} %Título da página dedicatória
	\input{dedicatoria} %Para incluir o texto de dedicatória,use o arquivo dedicatória.tex

\pretextualchapter{Agradecimentos} %Título da página de agradecimentos
	\input{agradecimentos} %Para incluir o texto de agradecimentos,use o arquivo agradecimentos.tex

\pretextualchapter{}
	\begin{epigrafetop}
		{They must find it difficult ... \\ Those who have taken authority as the truth, \\ rather than truth as the authority.} %frase
		{Gerald Massey (1828 - 1907)} %referência
	\end{epigrafetop}

% \,      a small space
% \:      a medium space
% \;      a large space
% \quad   a really large space
% \qquad  a huge space
% \!      a negative space (moves things back to the left)

	\begin{epigrafemid}
		{from Pantheon import Obatalá\, \, \, \\
                from World import sharing, kitten \\
                \vspace{.1in}
                while True: \; \; \; \; \; \; \; \; \; \; \; \; \; \; \; \,\\
                    if sharing == 0: \; \; \; \; \;  \; \; \; \; \; \\
                        Obatalá.kill( kitten ) \; \; \;}
        {Autoria coletiva e anônima (2010)}
	\end{epigrafemid}

	\begin{epigrafebot}
		{Do not worry about your difficulties in Mathematics. \\ I can assure you mine are still greater.} %frase
		{Albert Einstein (1879 - 1955)} %referência
	\end{epigrafebot}

\resumoeabstract %Para incluir o texto do resumo e abstract,use o arquivo resumoeabstract.tex

\listadefiguras %Comando que gera a lista de figuras (AUTOMÁTICO)

\listadetabelas %Comando que gera a lista de tabelas (AUTOMÁTICO)

\pretextualchapter{Lista de Abreviaturas}
	\begin{listaespecial}[BIGNAMEWIDTH]
		\item[OSS] Programas de Código Aberto \emph{(Open Source Software)}
		\item[GNU] \emph{GNU is not UNIX}
		\item[GPL] \emph{General Public Licence}
	\end{listaespecial}

\pretextualchapter{Lista de Símbolos}
	\begin{listaespecial}[BIGNAMEWIDTH]
		\item[q] Fluxo de massa em um trecho do duto
		\item[q_i] Fluxo de massa entrando em um trecho do duto
	\end{listaespecial}

\sumario

### 2011-08-23 resumoeabstract.tex
iniciado repositório. tentando com somente estes arquivos:
	new file:   APENDICE.tex
	new file:   agradecimentos.tex
	new file:   bibdatabase.bib
	new file:   conclusoes.tex
	new file:   dedicatoria.tex
	new file:   devEresultados.tex
	new file:   estiloifsc.sty
	new file:   fundamentos.tex
	new file:   introducao.tex
	new file:   mapaSitio.tex
	new file:   pretextual.tex
	new file:   resumoeabstract.tex

\begin{resumo}

O código é repetidamente destacado como o bem mais valioso produzido em nosso tempo.
O código aberto - e sua abordagem politico-filosófica chamada de software livre -
se propõe a constituir um repertório tecnológico da humanidade, acessível por
qualquer ser humano. Esta proposta orienta o que é chamado de cultura digital,
um agregado de práticas e explorações recentes e advindas da popularização do digital.
É tido como o vetor de transformação de nossa época por excelência.

Após a contextualização, expomos os fundamentos teóricos de uma série de contribuições em áudio, música
e web, todas feitas em código aberto. Os resultados antingidos, assim como as repercussões
destes desenvolvimentos são então sistematizados por escrito. Por fim, como um produto adicional
deste trabalho, é delineado o mapa de um sítio no qual os desenvolvimentos desta tese
são agregados. Tal sítio é dedicado a manter as contribuições de forma
continuada e sistematizada.

$\phantom{linha em branco}$\\
Palavras-chave: código aberto, cultura digital, áudio, música, internet.

\end{resumo}

\begin{abstract}

tradução de:

O software livre se propõe a constituir um repertório tecnológico da humanidade. O código
é destacado como o bem mais valioso que é produzido em nosso tempo.

A cultura digital é o vetor de transformação de nossa época por excelência e seu representante
maior é o software livre.

Dada a contextualização, esmeramos as contribuições em código aberto de áudio e música
que foi desenvolvida, assim como alguns desenvolvimentos web e suas repercussões.

$\phantom{linha em branco}$\\
Keywords: open source, digital culture, audio, music, internet.

\end{abstract}

### 2011-08-23 agradecimentos.tex
dando um bom tapa inicial na dissertação. 	new file:   README.txt 	modified:   agradecimentos.tex 	modified:   bibdatabase.bib 	modified:   dedicatoria.tex 	modified:   devEresultados.tex 	modified:   dissertacao.tex 	modified:   fundamentos.tex 	modified:   introducao.tex 	modified:   resumoeabstract.tex

@@
-\begin{itemize}
-	\item Aos meus pais ...
-	\item Aos amigos ...
-	\item Aos professores que ...
-\end{itemize}
+    Agradeço aos meus pais e familiares por todo o suporte e companhia. Aos
+meus amigos pelo carinho. Aos professores por tanta atenção e compreensão.
+Aos colegas colecionadores de entendimentos por ter-nos construído até aqui.
+Agradeço a Felipe Machado pela visão, a Fabiana 'Goa' Sherine pelo entendimento,
+a Glerm Soares pela transcendência, a Fabianne Baveldi pelo modelo, a Daniel
+Marostegan e Carneiro pelas lições. Não posso deixar de agradecer nominalmente
+ao Prof. Luciano da Fontoura Costa, que esteve presente na quase totalidade deste
+trabalho e notavelmente elevou-o a outro patamar. Finalmente, agradeço ao meu 
+orientator Prof. Osvaldo Novais de Oliveira Júnior pelo inestimável acompanhamento
+e notável paciência e dedicação em todos os momentos.

### 2011-08-23 devEresultados.tex
dando um bom tapa inicial na dissertação. 	new file:   README.txt 	modified:   agradecimentos.tex 	modified:   bibdatabase.bib 	modified:   dedicatoria.tex 	modified:   devEresultados.tex 	modified:   dissertacao.tex 	modified:   fundamentos.tex 	modified:   introducao.tex 	modified:   resumoeabstract.tex

@@
 Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})
+
+\section{Materiais didáticos}
+
+\subsection{Tutoriais em texto e código}
+
+\begin{itemize}
+    \item Tutorial de python para áudio e som.
+
+    \item Tutoriais de filtros e amostragem via python.
+
+    \item Tutorial de plugins lv2
+
+    \item Microtutoriais Django ~\cite{dmicrotuts}.
+
+    \item Philosometrics
+
+    \item Carta mídias livres
+
+    \item Textos de cunho sociológico, transformador
+
+de      fabi borges catadores@gmail.com por  riseup.net 
+responder a     submidialogia@lists.riseup.net
+para    submidialogia@lists.riseup.net
+data    23 de agosto de 2011 12:26
+assunto Re: [submidialogia] livro sub- publicação
+lista de e-mails        <submidialogia.lists.riseup.net> Filtrar as mensagens dessa lista de e-mails
+enviado por     lists.riseup.net
+assinado por    riseup.net
+cancelar inscrição      Cancelar a inscrição para essa lista de e-mails
+        Importante principalmente porque você frequentemente lê mensagens com esse marcador.
+ocultar detalhes 12:26 (6 minutos atrás)
+entao, eu fui recebendo textos durante esse tempo,
+alguns tao atrazados como dos sem satelites, mas muita gente mandou;
+
+aqui os autores:
+
+os internacionais nao sao muitos, o joni kempf (o que bebe ouro do hardware), o barbrook (futuros imaginarios),
+Hamdy heda (da revolucao egipcia), o pedro soller (summerlab), maria llopis (pos porno),  talvez a bronac, nao entregou ainda.
+
+dos brasileiros, renato fabri, ruiz, pasteur, morgana e caio, ju dornelles, mari marcassa, coletivo errorista, adriana velozo-drica, tiago pimentel, felipe fonseca, thiago novaes, lelex, felipe ribeiro(?), maira, verenilde,  bartolina silva, poro, vitoria amaro, fabib (eu),
+
+entao, precisa publicar agora,
+uma equipe para publicacao e,,,
+
+bjs
+f
+
+\end{itemize}
+
+\subsection{Screencasts}
+
+\begin{itemize}
+    \item Python para áudio e música
+
+    \item Canal Macambira
+No coletivo Macambira estão sendo produzidos materiais em screencasts sobre
+diversas.
+\end{itemize}
+
+\subsection{Figusdevpack (FDP)}
+
+Um ambiente de interação da comunidade de Python e Música 
+para compartilhamento de códigos e excertos. Baseado principalmente
+em documentação organizada sobre as práticas e as bibliotecas
+existentes para python. Assim como scripts compartilhados que
+fazem uso de objetos e módulos específicos. Este trabalho foi aceito na
+maior conferência de áudio em linux, a Linux Audio Conference de 2008
+(LAC2008) e está sendo reativado por mim em conjunto com Vilson Vieira
+e outros desenvolvedores de áudio. Este projeto está em desenvolvimento
+do site do Estúdio Livre [estudio livre] com repositório no sourceforge [source force].
+
+As principais fontes sobre estre trabalho é a página de desenvolvimento da ideia
+que está em constante mudança ~\cite{http://estudiolivre.org/tiki-index.php?page=fdp&highlight=fdp fdpel}
+e, o artigo que foi aceito no LAC de 2008 ~\cite{http://www.estudiolivre.org/el-gallery_view.php?arquivoId=8221 fdplac2008}
+e o repositório ~\cite{http://sourceforge.net/projects/fdpack/develop fdpsf}.
+
+\section{Áudio}
+
+\subsection{ABeatTracker (ABT)}
+
+Manual ABT
+
+Linkar com ABD
+
+\subsection{Emotional Kernel Panic (EKP)}
+
+Artigo FDP
+
+\subsection{Finite Groups in Granular Synthesis (FIGGS)}
+
+\subsection{Plugins LADSPA e lv2}
+
+\subsection{Processamento de voz via redes complexas}
+
+\subsection{Experimentos abertos}
+
+Audioexperiments. Estúdio livre.
 
@@
 
-\section{Áudio}
+\section{Disponibilização e Desenvolvimento}
 
-\subsection{ABT}
+\subsection{Wiki}
 
-Manual ABT
+\subsection{Trac}
 
-Linkar com ABD
+\subsection{Screencasts - Vimeo}
 
-\subsection{FDP}
+\subsection{AA}
 
-Artigo FDP
-
-\subsection{Experimentos abertos}
-
-Audioexperiments. Estúdio livre.
-
-\subsection{Tutoriais}
-
-Screencasts. Textos.
+\subsection{Outras fontes}

### 2011-08-23 fundamentos.tex
Creio que seja a útilma modificada tão inicial. 	modified:   fundamentos.tex
dando um bom tapa inicial na dissertação. 	new file:   README.txt 	modified:   agradecimentos.tex 	modified:   bibdatabase.bib 	modified:   dedicatoria.tex 	modified:   devEresultados.tex 	modified:   dissertacao.tex 	modified:   fundamentos.tex 	modified:   introducao.tex 	modified:   resumoeabstract.tex

@@
 
-\subsection{Som como fenômeno físico}
+\subsection{O som como fenômeno físico}
 
-Ver livros em casa.
+O som é o fenômeno físico de propagação de ondas mecânicas. Ele pode
+somente ocorrer na presença da matéria, ou seja, o vácuo é um
+isolante acústico perfeito.
+
+\subsection{A audição e a psicofísica}
+
+A audição é definida com a percepção de vibrações mecânicas com
+ou sem a contribuição direta do tato. Em diversos
+seres vivos, encontramos um órgão dedicado a captar e transmitir
+estas vibrações mecânicas (som) para o sistema nervoso.
+
+O ouvido humano cumpre exatamente este papel. Um esquema básico
+do ouvido humano conta com 3 divisões. No ouvido externo tem a orelha.
+No médio, tem os martelinhos e o tímpano. No ouvido interno tem a cóclea
+e os cíclios. Uma visualização simplificada do ouvido humano está abaixo.
+
+FIGURA MOSTRANDO ENTRADA DE VIBRAÇÃO E TRANSMISSÃO AO CÉREBRO
+
+Vale notar que a audição conta em grande parte com o tato e potencialmente
+outros órgãos, como o estômago.
+
+Já a psicofísica busca relacionar características físicas do fenômeno sonoro
+com as percepções individuais causadas\footnote{PsycoPy para escrever experimentos psicofísicos em Python: http://www.psychopy.org/}.
+Importantes exemplos destas relações
+é a escala logarítimica que a percepção linear das alturas assumem no
+domínio da frequência e as curvas iso-audíveis (conhecidas como curvas de Fletcher-Münsen).
+
+Outro importante aspecto da psicofísica do som - e da música - é a resolução da percepção
+e a capacidade de notar características do espectro em contextos específicos. O mascaramento
+de frequências é a base na qual se faz compressores de tamanhos de arquivos de áudio. Sabe-se
+que, por exemplo, no extremo grave, não conseguimos distinguir duas senóides a quase um tom de
+diferença. Outros fenômenos psicoacústicos estudados com fins tecnológicos, científicos e artísticos são: espacialidade,
+relação da percepção de volume com espectro agudo e reverberação, compreensibilidade de sinais de fala, etc.
 
@@
 
+\subsection{Breve histórico}
+
 Discursar brevemente sobre as diversas linguagens,
@@
 
+\subsection{C/C++}
+
+\subsection{Python}
+
 \section{Sistema Operacional e Web}
 
-\subsection{Breve Histórico}
+\subsection{Breve histórico}
 

### 2011-08-23 introducao.tex
dando um bom tapa inicial na dissertação. 	new file:   README.txt 	modified:   agradecimentos.tex 	modified:   bibdatabase.bib 	modified:   dedicatoria.tex 	modified:   devEresultados.tex 	modified:   dissertacao.tex 	modified:   fundamentos.tex 	modified:   introducao.tex 	modified:   resumoeabstract.tex

@@
 
-\subsection{Código Aberto e Software Livre} %Nome da seção.
+\section{O som e o áudio}
 
+\subsection{Breve histórico e práticas atuais}
+
+\subsection{Campos de estudo e atuação}
+
+\section{Código aberto, software livre e cultura livre} %Nome da seção.
+
+O código é repetidamente destacado como o bem mais valioso produzido em nosso tempo.
+O código aberto - e sua abordagem politico-filosófica chamada de software livre -
+se propõe a constituir um repertório tecnológico da humanidade, acessível por
+qualquer ser humano. Esta proposta orienta o que é chamado de cultura digital,
+um agregado de práticas e explorações recentes e advindas da popularização do digital.
+É tido como o vetor de transformação de nossa época por excelência.
 Qualquer licença de software livre é também uma licença de código aberto (Open Source), a diferença entre as duas nomenclaturas reside essencialmente na sua apresentação e na filosofia.
 
-\subsection{Softwares para compartilhamento} %Nome da seção.
+\subsection{Cultura de compartilhamento} %Nome da seção.
 \label{sec:soft_compar} %Rótulo para futura referência à seção.
@@
 %% ------------------------------------------------------------------------- %%
-\subsection{Licenças livres para conteúdo}
+\subsection{Licenças livres para software e para outras mídias}
 \label{sec:licencas_livres}
@@
 
+Licenças fechadas.
+
 \subsubsection{Principais licenças}
@@
 
-\section{Contextualização}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%
+\section{Estado das Coisas}
 \label{sec:context}
 
-Descentralização: Fator caminhão e espalhamento de informação. Descentralização
-é uma constante nas estruturas e ferramentas mais relevantes como redes
-sociais, comunicadores que agora funcionam por broadcast e protocolos de
-compartilhamento de dados, em especial arquivos inteiros, unidades mais
-bem organizadas.
+\subsection{A convergência das mídias}
+\label{sec:midiamultimidia}
 
-Aumento da conectividade através de avanços em telecomunicações e transportes.
+HTML5 e Javascript deixando interativo. 3D, conexões rápidas que possuem banda para video HD e
+baixa latência. Popularização de hardware capaz destes processamentos.
+
+Tendência: 
+\begin{itemize}
+    \item intensa utilização de audiovisual em contraposição ao
+anterior predominio de textos simples.
+    \item a utilização de periféricos para resultar saída diferenciada - p.ex.
+impressora - ou entrada especializada - p.ex. touch pad.
+\end{itemize}
+
+\subsection{Multimídia}
+
+\subsubsection{Gargalos na interatividade e no estado do software}
+\label{sec:gargalos}
+
+Deficiencias em disponibilização da tecnologia atual: muita
+tecnologia fechada.
+
+Comparação do software proprietário com o livre para áudio:
+
+XXXXXXXXXXXXXXXXXXXXX
+
+Citação da parte de video da thread 'Abrindo o Códifo Fechado'.
+
+Para certas utilizações, a latência não permite a interatividade como 
+a obtida em presença do objeto examinado seja ele físico ou apenas um BD.
+A medida do 'agora' é de até 50 ms para humanos [Roads, Microsound] e a latência entre comunicadores é
+quase sempre maior que isso [tabela de latências, média, por distância, etc].
+
+Contribuimos na frente do software.
 
@@
 
-\subsubsection{Compartilhamento de Tec}
+A modularidade do sistema \emph{Unix type} e as compensações: jack,
+facilidades de instalação e segurança, tudo hackeável.
+
+\subsubsection{Compartilhamento}
 \label{sec:comp_tec}
@@
 
-\subsection{Mutimídia} % Onde se detecta a ênfase em áudio aqui???
-\label{sec:multimidia}
-
-\subsubsection{Gargalo}
-\label{sec:gargalo}
+%%%%%%%%
+\subsection{Identificação de tarefas pouco suportadas}
+\label{sec:tarefas_n_sup}
 
@@
 
-\subsection{Identificação de tarefas pouco suportadas}
-\label{sec:tarefas_n_sup}
-
 \section{Objetivos}
@@
 
-\subsection{Difusão de práticas}
-\label{sec:difusao}
-
-\subsubsection{Tutoriais}
+\subsection{Difusão de tecnologias através de tutoriais}
 \label{sec:tutoriais}
@@
 
-\subsubsection{Uso de software livre}
+\subsection{Difusão de tecnologias através do uso de software livre}
 \label{sec:uso_sl}
+
+Descentralização: Fator caminhão e espalhamento de informação. Descentralização
+é uma constante nas estruturas e ferramentas mais relevantes como redes
+sociais, comunicadores que agora funcionam por broadcast e protocolos de
+compartilhamento de dados, em especial arquivos inteiros, unidades mais
+bem organizadas.
+
+Aumento da conectividade através de avanços em telecomunicações e transportes.
 
@@
 
-\subsection{Extensão de recursos em software livre - multimídia}
+\subsection{Extensão de recursos em software livre para multimídia}
 \label{sec:extensao}

### 2011-08-23 resumoeabstract.tex
dando um bom tapa inicial na dissertação. 	new file:   README.txt 	modified:   agradecimentos.tex 	modified:   bibdatabase.bib 	modified:   dedicatoria.tex 	modified:   devEresultados.tex 	modified:   dissertacao.tex 	modified:   fundamentos.tex 	modified:   introducao.tex 	modified:   resumoeabstract.tex

@@
 
-O código é repetidamente destacado como o bem mais valioso produzido em nosso tempo.
-O código aberto - e sua abordagem politico-filosófica chamada de software livre -
-se propõe a constituir um repertório tecnológico da humanidade, acessível por
-qualquer ser humano. Esta proposta orienta o que é chamado de cultura digital,
-um agregado de práticas e explorações recentes e advindas da popularização do digital.
-É tido como o vetor de transformação de nossa época por excelência.
-
-Após a contextualização, expomos os fundamentos teóricos de uma série de contribuições em áudio, música
-e web, todas feitas em código aberto. Os resultados antingidos, assim como as repercussões
-destes desenvolvimentos são então sistematizados por escrito. Por fim, como um produto adicional
-deste trabalho, é delineado o mapa de um sítio no qual os desenvolvimentos desta tese
-são agregados. Tal sítio é dedicado a manter as contribuições de forma
-continuada e sistematizada.
+Após uma breve contextualização sobre o status que o código e o áudio digital possuem na cultura atual,
+expomos os fundamentos do manuseio do som digital através do software em código aberto: fenômeno
+físico (e psicofísico), representação digital e procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
+são então abordadas, já com vistas a esta tarefa lidar com o áudio digital. Ainda no campo do software,
+os sistemas operacionais são então perspassados e o ambiente de desenvolvimento Unix é descrito de um ponto
+de vista bem resumido e ferramental. Os resultados antingidos com os fundamentos aqui expostos,
+assim como as repercussões destes desenvolvimentos são então sistematizados por escrito. 
+Por fim, como um produto adicional relacionado a esta dissertação, é delineado o mapa de exposição e comunicação no qual
+os desenvolvimentos deste trabalho são agregados. Esta estrutura é dedicada a manter as contribuições
+de forma continuada e sistematizada através de webpages, wikis, repositórios, aplicativos de acompanhamento de projetos
+e uma tecnologia desenvolvida neste trabalho: A Autogestão Algorítmica.
 
 $\phantom{linha em branco}$\\
-Palavras-chave: código aberto, cultura digital, áudio, música, internet.
+Palavras-chave: áudio, programação, acústica, código aberto, cultura digital
 
@@
 
-O software livre se propõe a constituir um repertório tecnológico da humanidade. O código
-é destacado como o bem mais valioso que é produzido em nosso tempo.
-
-A cultura digital é o vetor de transformação de nossa época por excelência e seu representante
-maior é o software livre.
-
-Dada a contextualização, esmeramos as contribuições em código aberto de áudio e música
-que foi desenvolvida, assim como alguns desenvolvimentos web e suas repercussões.
+Após uma breve contextualização sobre o status que o código e o áudio digital possuem na cultura atual,
+expomos os fundamentos do manuseio do som digital através do software em código aberto: fenômeno
+físico (e psicofísico), representação digital e procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
+são então abordadas, já com vistas a esta tarefa lidar com o áudio digital. Ainda no campo do software,
+os sistemas operacionais são então perspassados e o ambiente de desenvolvimento Unix é descrito de um ponto
+de vista bem resumido e ferramental. Os resultados antingidos com os fundamentos aqui expostos,
+assim como as repercussões destes desenvolvimentos são então sistematizados por escrito. 
+Por fim, como um produto adicional relacionado a esta dissertação, é delineado o mapa de exposição e comunicação no qual
+os desenvolvimentos deste trabalho são agregados. Esta estrutura é dedicada a manter as contribuições
+de forma continuada e sistematizada através de webpages, wikis, repositórios, aplicativos de acompanhamento de projetos
+e uma tecnologia desenvolvida neste trabalho: A Autogestão Algorítmica.
 
 $\phantom{linha em branco}$\\
-Keywords: open source, digital culture, audio, music, internet.
+Keywords: audio, programming, acoustics, open source, digital culture
 

### 2011-08-24 introducao.tex
tomando forma real

@@
 
-O livre fluxo de informação
+O som é tipicamente utilizado para fala, sinais em geral\footnote{p.ex. buzina} e música.
+Caracteriza-se por ser um sentido   O livre fluxo de informação
 
@@
 \subsection{Breve histórico e práticas atuais}
+
+As explicações físicas do fenômeno do som no decorrer do tempo. Teorias
+corpusculares, escola de Tales. Helmholtz.
 
@@
 %%%%%%%%%%%%%%%%%%%
-\section{Estado das Coisas}
+\section{Estado das coisas}
 \label{sec:context}
@@
 
-\subsection{Difusão de tecnologias através de tutoriais}
+\subsection{Difusão de tecnologias e práticas}
 \label{sec:tutoriais}
+
+Descentralização: Fator caminhão e espalhamento de informação. Descentralização
+é uma constante nas estruturas e ferramentas mais relevantes como redes
+sociais, comunicadores que agora funcionam por broadcast e protocolos de
+compartilhamento de dados, em especial arquivos inteiros, unidades mais
+bem organizadas.
+
+Aumento da conectividade através de avanços em telecomunicações e transportes.
+
+\subsubsection{Tutoriais}
+\label{sec:uso_sl}
 
@@
 \end{itemize}
-
-\subsection{Difusão de tecnologias através do uso de software livre}
-\label{sec:uso_sl}
-
-Descentralização: Fator caminhão e espalhamento de informação. Descentralização
-é uma constante nas estruturas e ferramentas mais relevantes como redes
-sociais, comunicadores que agora funcionam por broadcast e protocolos de
-compartilhamento de dados, em especial arquivos inteiros, unidades mais
-bem organizadas.
-
-Aumento da conectividade através de avanços em telecomunicações e transportes.
 

### 2011-08-24 pretextual.tex
quotation no Newton modified:   pretextual.tex

@@
 	\begin{epigrafebot}
-		{Do not worry about your difficulties in Mathematics. \\ I can assure you mine are still greater.} %frase
-		{Albert Einstein (1879 - 1955)} %referência
+		{If I have seen further, it is by standing on the shoulders of giants.} %frase
+		{Sir Isaac Newton (1642 - 1727)} %referência
 	\end{epigrafebot}

### 2011-08-25 agradecimentos.tex
tomando forma real
trabalhando na dissertação como um todo ainda 	modified:   agradecimentos.tex 	modified:   bibdatabase.bib 	modified:   dissertacao.tex 	modified:   estiloifsc.sty 	modified:   pretextual.tex 	modified:   resumoeabstract.tex

--- 2011-08-23 agradecimentos.tex
@@
-\begin{itemize}
-	\item Aos meus pais ...
-	\item Aos amigos ...
-	\item Aos professores que ...
-\end{itemize}
+    Agradeço aos meus pais e familiares por todo o suporte e companhia. Aos
+meus amigos pelo carinho. Aos professores por tanta atenção e compreensão.
+Aos colegas colecionadores de entendimentos por ter-nos construído até aqui.
+Agradeço a Felipe Machado pela visão, a Fabiana 'Goa' Sherine pelo entendimento,
+a Glerm Soares pela transcendência, a Fabianne Baveldi pelo modelo, a Daniel
+Marostegan e Carneiro pelas lições, entendimentos e confianças. Agradeço ao profícuo Vilson Vieira, que tornou-se especialmente próximo
+das contribuições, construindo sobre elas e com elas pois trabalha com áudio e tecnologias livres também
+e está morando em São Carlos estudando estas mesmas questões, mas com outros focos. Ricardo Fabbri, meu irmão,
+também está trabalhando conosco e cuida de processamento de imagens e video. Fortíssima referência
+para mim, tanto para os assuntos tratados nesta dissertação, quanto para questões humanas. Agradecimento
+forte ao Lab Macambira, nominalmente estes assíduos: [Equipe Lab Macambira]. Um profundo agradecimento
+aos meus orientadores de IC: Adolfo Maia Junior e Jônatas Manzolli. Agradeço a José Augusto Mannis
+dos trabalhos de estudante, pelo acompanhamento e ensinos sobre manipulação de áudio e música como idioma.
+Agradeço fortemente ao Prof. Dr. Rafael Santos Mendes pela pesquisa em Wavelets de mestrado em engenharia elétrica
+praticamente completa na FEEC/Unicamp que não defendendi por razões circunstanciais. Agradeço aos professores da FEEC pelos ensinos de
+primeiríssima qualidade em diversas disciplinas como processamento de sinais, teoria da informação, 
+circuitos elétricos, processos estocásticos
+e programação bio-inspirada. Sem dúvida agradecimentos especiais para os assíduos mentores
+deste trabalho: ao orientador-colaborador Prof. Luciano da Fontoura Costa, que
+mistura criatividade com clareza de pensamento. Finalmente, agradeço ao meu 
+orientator Prof. Osvaldo Novais de Oliveira Júnior pelo inestimável acompanhamento
+e notável paciência, dedicação e precisão em todos os momentos.

### 2011-08-25 pretextual.tex
trabalhando na dissertação como um todo ainda 	modified:   agradecimentos.tex 	modified:   bibdatabase.bib 	modified:   dissertacao.tex 	modified:   estiloifsc.sty 	modified:   pretextual.tex 	modified:   resumoeabstract.tex

@@
 	\begin{epigrafebot}
-		{If I have seen further, it is by standing on the shoulders of giants.} %frase
+		{If I have seen further, \\it is by standing on the shoulders of giants.} %frase
 		{Sir Isaac Newton (1642 - 1727)} %referência
@@
 		\item[GPL] \emph{General Public Licence}
+                \item[AA] \emph{Algorithmic Autoregularion}
+                \item[AC] \emph{Autogestão Coletiva}
+                \item[CC] \emph{Creative Commons}
+                \item[ABT] \emph{ABeatTracker}
+                \item[ABD] \emph{ABeatDetector}
+                \item[PD] \emph{Pure Data}
+                \item[CMDCA] \emph{Conselho Municipal de defesa dos Direitos da Cirança e do Adolescente}
+                \item[SOS] \emph{Saúde Olha Sabedoria: uma proposta de coleta e difusão de conhecimentos relacionados à saúde}
+                \item[LADSPA] \emph{Linux Audio Developers Simple Plugin API}
+                \item[LV2] \emph{LADSPA Version 2}
+                \item[JACK] \emph{Jack Audio Connection Kit}
+                \item[SIP] \emph{Scilab Imaging Processing toolbox}
+                \item[MIT] \emph{Massachusetts Institute of Technology}
+                \item[NUMPY] \emph{Numerical Python}
+                \item[SCIPY] \emph{Scientific Python}
+                \item[ONU] \emph{Organização das Nações Unidas}
 	\end{listaespecial}

### 2011-08-25 resumoeabstract.tex
trabalhando na dissertação como um todo ainda 	modified:   agradecimentos.tex 	modified:   bibdatabase.bib 	modified:   dissertacao.tex 	modified:   estiloifsc.sty 	modified:   pretextual.tex 	modified:   resumoeabstract.tex

@@
 Após uma breve contextualização sobre o status que o código e o áudio digital possuem na cultura atual,
-expomos os fundamentos do manuseio do som digital através do software em código aberto: fenômeno
-físico (e psicofísico), representação digital e procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
+expomos os fundamentos do manuseio do som digitalizado através do software em código aberto. Procuramos abordar como um todo o fenômeno
+físico (e psicofísico) do som, a sua representação digital e os procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
 são então abordadas, já com vistas a esta tarefa lidar com o áudio digital. Ainda no campo do software,

### 2011-08-26 fundamentos.tex
modified:   fundamentos.tex
modified:   APENDICE.tex 	modified:   fundamentos.tex 	modified:   introducao.tex

@@
 O som é o fenômeno físico de propagação de ondas mecânicas. Ele pode
-somente ocorrer na presença da matéria, ou seja, o vácuo é um
-isolante acústico perfeito.
+somente ocorrer na presença de matéria, como ar, água, ferro ou madeira.
+São ondas de pressão que se deslocam com velocidade caracterizada principalmente
+pelo meio material e também por condições como a temperatura. O vácuo é um isolante acústico perfeito.
+
+Tipicamente a velocidade do som é de $~340 m/s $ e é
+relativamente estável\footnote{$343.2 m/s$ em ar seco a 20 °C (68 °F)}.
+As modificações de sonoridades com o clima é devido a alterações
+dos materiais que os geram e acústica do ambiente, não devido a alterações
+desta velocidade de propagação do som no meio atmosférico.
+
+Falar de fonons e coisas assim? Quais aspectos do som devemos cobrir?
+
+Pegar um livro bom do Prof. Luciano sobre o assunto pois não estou me animando a ficar
+escrevendo sobre ondas
 
@@
 
-\subsection{Som digitalizado}
+\subsection{Áudio digitalizado}
 
-Ver livros em casa.
+Enquanto estamos interessados, na variação de pressão do ar, podemos de fato utilizar
+um microfone para converter esta variação de pressão e transformar em sinal elétrico
+(geralmente voltagem).
+Convertemos assim o som (fenômeno físico) em áudio (sinal elétrico relacionado).
+É neste momento ponto em que o sinal é amostrado no tempo, geralmente através do uso de
+um conversor A/D\footnote{Conversor Analógico para Digital. Para as aplicações descritas aqui, é tipicamente uma placa som.}. Assim, um evento sônico
+torna-se um sinal digital, usufruindo das facilidades para armazenamento, transmissão e
+processamento da nossa tecnologia digital atual.
 
-Representação digital do som.
+Esta amostragem do sinal de áudio é bastante elaborada e dependente de manejo de hardware. Para
+aprofundar este tema, recomendamos [][][].
 
-Padrões PCM. Padrões compactados.
+A representação digital do som pode ser feita de diversas formas.
+A representação PCM (Pulse Code Modulation) canônica possui amostras
+igualmente espaçadas no tempo e paraa cada uma utiliza precisão com o mesmo
+número de bytes.
+
+[Figura com onda amostrada]
+
+Note que o áudio digital possue também uma velocidade no espaço de amostragem.
+Quantidades como $8kHz$, $44.1kHz$, $48kHz$
+descrevem quantidade de amostras a cada segundo com que o áudio examinado foi feito ou
+está sendo utilizado.
+
+O importante Teorema de Nyquist demonstra que em um sinal digital, só se sonsegue representar
+frequências até metade da taxa de amostragem. Ou seja, em um áudio com $48kHz$ de taxa de amostragem,
+conseguimos representar frequências ate $24kHz$. A frequência de Nyquist é exatamente a metade
+da frequência de amostragem e é a frequência mais aguda que aquele sinal pode representar.
+
+Além da representação PCM, pode-se utilizar outros formatos para o armazenamento do áudio. Alguns
+destes formatos apresentam perda de qualidade e são mais indicados para serem transmitidos via rede ou
+para armazenamento. Abaixo está uma tabela com alguns dos formatos mais importantes que possuímos:
+
+[Tabela de formatos]
 
@@
 
-Ver livros em casa.
+Síntese, processamento ou análise.
+
+\begin{itemize}
+    \item Geral
+    \item Voz
+    \item Música
+\end{itemize}
 
@@
 Breve Histórico.
+
+No apêndice resumo de:
+http://www.html5rocks.com/en/tutorials/internals/howbrowserswork/

### 2011-08-26 introducao.tex
modified:   APENDICE.tex 	modified:   fundamentos.tex 	modified:   introducao.tex

@@
 
-O som é tipicamente utilizado para fala, sinais em geral\footnote{p.ex. buzina} e música.
-Caracteriza-se por ser um sentido   O livre fluxo de informação
+É um trabalho criterioso definir quais pontos cabem serem abordados em um trabalho como este.
+Como abordar a atuação tecnológica e científica em uma área criativa? Como contemplar
+a cultura de compartilhamento e diversidade em que está inserida?
+
+Optamos neste trabalho pela síntese e pela simplicidade. Por outro lado, esta dissertação
+se propõe a ser um guia relativamente completo para uma atuação eficiênte em desenvolvimento
+de software em multimídia, mais especificamente voltados áudio e tecnologias livres.
+
+Estaremos abordando o som, o áudio, o áudio digital. Práticas de programação voltadas para
+áudio e também as plataformas para as quais desenvolvemos código: Sistemas Operacionais e Browsers.
+
+Com estas tecnologias em mãos, foram criados diversos aplicativos com fins artísticos,
+técnicos, científicos e sociais. Como consequência, foi criada uma equipe para tratar
+destas questões, o Lab Macambira, e um sistema rico de comunicação e exposição de todos
+os desenvolvimentos aqui expostos.
 
@@
 \section{O som e o áudio}
+
+O som é tipicamente utilizado para fala, sinais em geral\footnote{p.ex. buzina} e música.
+Caracteriza-se por ser um sentido   O livre fluxo de informação
 

### 2011-08-31 introducao.tex
caminhando com a escrita. 	modified:   dissertacao.tex 	modified:   fundamentos.tex 	modified:   introducao.tex

@@
 Como abordar a atuação tecnológica e científica em uma área criativa? Como contemplar
-a cultura de compartilhamento e diversidade em que está inserida?
-
-Optamos neste trabalho pela síntese e pela simplicidade. Por outro lado, esta dissertação
-se propõe a ser um guia relativamente completo para uma atuação eficiênte em desenvolvimento
+a cultura de compartilhamento e diversidade em que esta produção tecnológica está inserida?
+Com estas questões em mente, optamos neste trabalho pela síntese e pela simplicidade.
+Além disso, esta dissertação se propõe a ser um guia relativamente completo para uma atuação eficiênte em desenvolvimento
 de software em multimídia, mais especificamente voltados áudio e tecnologias livres.
@@
 \label{sec:midiamultimidia}
+
+Os anos de 1999 e 2000 compreendem o conhecido 'marco inicial da primeira década digital'.
+Embora seja principalmente uma forma de delimitar um processo-chave já em andamento, a este período é atribuído
+a delineação dos contornos da 'sociedade em rede'. Esta mudança paradigmática da nossa
+sociedade é sentida agora em maior intensidade nos negócios, na comunicação e na cultura. Estes dois
+últimos fatores são cruciais para o trabalho aqui apresentado dada a notoriedade do papel do
+fenômeno sonoro para a comunicação e expressões culturais.
+
+O termo 'convergência das mídias' não é utilizado para indicar algum dispositivo ou produto que
+lida com várias mídias. Para isso utiliza-se o termo multimídia, que é tema do próximo tópico.
+A convergência das mídias se refere à mundaneidade no lidar com suportes diferentes. Procura-se
+sobre um assunto e lidamos com imagens, textos, videos, áudios e até outras formas diferenciadas
+de mídia, como o código de liguagem de programação.
 

### 2011-09-01 devEresultados.tex
tomando forma real
ainda só adicionando. Modificado o título e iniciada a descrição do EKP.
	modified:   devEresultados.tex
	modified:   dissertacao.tex
	modified:   fundamentos.tex

@@
 
-\section{Áudio}
+\section{Áudio e Música}
 
@@
 
-Artigo FDP
+Em 2008, colaborando intensamente com o CDTL
+(Centro de Desenvolvimento de Tecnologias Livres)\footnote{foi uma associação civil formada e desmembrada em 2008 e sediada em Recife, PE}
+foi lançada a ideia de utilizar o estado do sistema operacional - especialmente o kernel linux - para
+geração de sons. Surge o Emotional Kernel Panic (EKP). Desde o inicio, foram definidos três finalidades
+para esta exploração do SO:
+
+\begin{itemize}
+    \item Didáticos
+    \item Artísticos
+    \item Monitoramento do SO
+\end{itemize}
+
+http://trac.assembla.com/audioexperiments/browser/ekp-base
 
@@
 \subsection{Plugins LADSPA e lv2}
+
+\subsection{Protocolo de compressão de áudio via wavelets, polinômios e permutações}
 
@@
 
+\subsection{Tecnologias sociais de alta demanda}
+
 \subsection{Sítios}
@@
 \subsection{Conteúdos}
-
-\subsubsection{textos}
-Carta mídias livres
-
-Artigos,
-
-Tutoriais
 
@@
 
+\subsection{Audio Experiments (Æ)}
+
 \subsection{Outras fontes}

### 2012-01-17 APENDICE.tex
modified:   APENDICE.tex 	modified:   fundamentos.tex 	modified:   introducao.tex
feita uma revisao quase que completa 	modified:   APENDICE.tex 	modified:   agradecimentos.tex 	modified:   dedicatoria.tex 	modified:   devEresultados.tex 	modified:   fundamentos.tex 	modified:   resumoeabstract.tex

@@
 \section{Web}
+
+\subsection{Números binários}
 
@@
 
-\subsubsection{Tecnologias padrão}
-HTML, CSS, Javascript, PHP, Mysql.
-
-\subsubsection{Frameworks}
-Django, Ruby etc.
+\subsubsection{Glossário de termos de tecnologias utilizadas nesta dissertação}
+HTML, CSS, Javascript, PHP, Mysql, Django, Ruby..
 
 \subsection{Internet em Números}
-
 Quantos usuários X, quantos sites/minuto cada usuáriosvisita em média.
@@
 processamento, memória RAM, resolução etc.
+Tablets e celulares
 
 \subsection{e-democracy}
-
 Focamos nossos desenvolvimentos em ferramentas de inclusão

### 2012-01-17 agradecimentos.tex
feita uma revisao quase que completa 	modified:   APENDICE.tex 	modified:   agradecimentos.tex 	modified:   dedicatoria.tex 	modified:   devEresultados.tex 	modified:   fundamentos.tex 	modified:   resumoeabstract.tex

@@
-    Agradeço aos meus pais e familiares por todo o suporte e companhia. Aos
-meus amigos pelo carinho. Aos professores por tanta atenção e compreensão.
+Em preimeiro lugar, agradeço aos meus pais e demais familiares. Agradeço aos
+meus amigos pelo carinho e companhia. Aos professores por tanta atenção e compreensão.
 Aos colegas colecionadores de entendimentos por ter-nos construído até aqui.
 Agradeço a Felipe Machado pela visão, a Fabiana 'Goa' Sherine pelo entendimento,
-a Glerm Soares pela transcendência, a Fabianne Baveldi pelo modelo, a Daniel
-Marostegan e Carneiro pelas lições, entendimentos e confianças. Agradeço ao profícuo Vilson Vieira, que tornou-se especialmente próximo
-das contribuições, construindo sobre elas e com elas pois trabalha com áudio e tecnologias livres também
-e está morando em São Carlos estudando estas mesmas questões, mas com outros focos. Ricardo Fabbri, meu irmão,
-também está trabalhando conosco e cuida de processamento de imagens e video. Fortíssima referência
+a Glerm Soares pela transcendência, a Fabianne Baveldi pela excelência, a Daniel
+Marostegan e Carneiro pelas lições e confianças. Agradeço ao profícuo Vilson Vieira, que tornou-se especialmente próximo
+das contribuições citadas neste trabalho, construindo sobre elas e com elas. Especial agradecimento
+ao meu irmao Ricardo Fabbri que também se aproximou desta contribuição cuidando de processamento de imagens e video. Fortíssima referência
 para mim, tanto para os assuntos tratados nesta dissertação, quanto para questões humanas. Agradecimento
@@
 aos meus orientadores de IC: Adolfo Maia Junior e Jônatas Manzolli. Agradeço a José Augusto Mannis
-dos trabalhos de estudante, pelo acompanhamento e ensinos sobre manipulação de áudio e música como idioma.
+pelos trabalhos de estudante, pelo acompanhamento e ensinos sobre manipulação de áudio e música como idioma.
 Agradeço fortemente ao Prof. Dr. Rafael Santos Mendes pela pesquisa em Wavelets de mestrado em engenharia elétrica
@@
 circuitos elétricos, processos estocásticos
-e programação bio-inspirada. Sem dúvida agradecimentos especiais para os assíduos mentores
-deste trabalho: ao orientador-colaborador Prof. Luciano da Fontoura Costa, que
+e programação bio-inspirada. Sem sombra de dúvida agradecimentos especiais para os assíduos mentores
+deste trabalho: ao orientador-colaborador Prof. Luciano da Fontoura Costa, que, dentre outras qualidades,
 mistura criatividade com clareza de pensamento. Finalmente, agradeço ao meu 

### 2012-01-17 fundamentos.tex
caminhando com a escrita. 	modified:   dissertacao.tex 	modified:   fundamentos.tex 	modified:   introducao.tex
ainda só adicionando. Modificado o título e iniciada a descrição do EKP.
	modified:   devEresultados.tex
	modified:   dissertacao.tex
	modified:   fundamentos.tex
feita uma revisao quase que completa 	modified:   APENDICE.tex 	modified:   agradecimentos.tex 	modified:   dedicatoria.tex 	modified:   devEresultados.tex 	modified:   fundamentos.tex 	modified:   resumoeabstract.tex

@@
 São ondas de pressão que se deslocam com velocidade caracterizada principalmente
-pelo meio material e também por condições como a temperatura. O vácuo é um isolante acústico perfeito.
+pelo meio material e também pelas condições como a temperatura e a pressão.
+O vácuo é um isolante acústico perfeito.
 
@@
 
-Pegar um livro bom do Prof. Luciano sobre o assunto pois não estou me animando a ficar
-escrevendo sobre ondas
-
 \subsection{A audição e a psicofísica}
@@
 O ouvido humano cumpre exatamente este papel. Um esquema básico
-do ouvido humano conta com 3 divisões. No ouvido externo tem a orelha.
-No médio, tem os martelinhos e o tímpano. No ouvido interno tem a cóclea
+do ouvido humano conta com 3 divisões. No ouvido externo está a orelha.
+No médio, os martelinhos e o tímpano. No ouvido interno tem a cóclea
 e os cíclios. Uma visualização simplificada do ouvido humano está abaixo.
@@
 Já a psicofísica busca relacionar características físicas do fenômeno sonoro
-com as percepções individuais causadas\footnote{PsycoPy para escrever experimentos psicofísicos em Python: http://www.psychopy.org/}.
+com as percepções individuais resultantes\footnote{PsycoPy para escrever experimentos psicofísicos em Python: http://www.psychopy.org/}.
 Importantes exemplos destas relações
-é a escala logarítimica que a percepção linear das alturas assumem no
+são a escala logarítimica que a percepção linear das alturas assume no
 domínio da frequência e as curvas iso-audíveis (conhecidas como curvas de Fletcher-Münsen).
+EXPLICAR MELHOR?
 
@@
 e a capacidade de notar características do espectro em contextos específicos. O mascaramento
-de frequências é a base na qual se faz compressores de tamanhos de arquivos de áudio. Sabe-se
+de frequências é a base na qual se faz compactadores de tamanhos de arquivos de áudio. Sabe-se
 que, por exemplo, no extremo grave, não conseguimos distinguir duas senóides a quase um tom de
-diferença. Outros fenômenos psicoacústicos estudados com fins tecnológicos, científicos e artísticos são: espacialidade,
-relação da percepção de volume com espectro agudo e reverberação, compreensibilidade de sinais de fala, etc.
+diferença. Alguns outros importantes fenômenos psicoacústicos estudados com fins tecnológicos, científicos e artísticos são: espacialidade,
+relação da percepção de volume com espectro agudo e reverberação e compreensibilidade de sinais de fala.
 
 \subsection{Áudio digitalizado}
-
-Enquanto estamos interessados, na variação de pressão do ar, podemos de fato utilizar
-um microfone para converter esta variação de pressão e transformar em sinal elétrico
-(geralmente voltagem).
+Enquanto estamos interessados na variação de pressão do ar, podemos de fato utilizar
+um microfone para converter esta variação de pressão em sinal elétrico (usualmente voltagem).
 Convertemos assim o som (fenômeno físico) em áudio (sinal elétrico relacionado).
-É neste momento ponto em que o sinal é amostrado no tempo, geralmente através do uso de
-um conversor A/D\footnote{Conversor Analógico para Digital. Para as aplicações descritas aqui, é tipicamente uma placa som.}. Assim, um evento sônico
+Neste processo, o sinal é amostrado no tempo geralmente através do uso de
+um conversor A/D\footnote{Conversor Analógico para Digital. Para as aplicações descritas aqui é tipicamente uma placa som.}. Assim, um evento sônico
 torna-se um sinal digital, usufruindo das facilidades para armazenamento, transmissão e
@@
 Esta amostragem do sinal de áudio é bastante elaborada e dependente de manejo de hardware. Para
-aprofundar este tema, recomendamos [][][].
+aprofundar este tema, recomendamos [][][] e uma representacao esquematica deste processo
+se encontra na figura \figure{fig:AD}.
+
+FIGURA MOSTRANDO UMA ONDA ANALOGICA E ELA REPRESENTADA POR PONTOS.
 
@@
 A representação PCM (Pulse Code Modulation) canônica possui amostras
-igualmente espaçadas no tempo e paraa cada uma utiliza precisão com o mesmo
+igualmente espaçadas no tempo e para cada uma utiliza precisão com o mesmo
 número de bytes.
 
-[Figura com onda amostrada]
+[Figura com onda amostrada com limites superior inferior e valores das amostras]
 
@@
 Quantidades como $8kHz$, $44.1kHz$, $48kHz$
-descrevem quantidade de amostras a cada segundo com que o áudio examinado foi feito ou
+descrevem quantidade de amostras em cada segundo com que o áudio examinado foi feito ou
 está sendo utilizado.
 
-O importante Teorema de Nyquist demonstra que em um sinal digital, só se sonsegue representar
+O importante Teorema de Nyquist demonstra que em um sinal digital só se consegue representar
 frequências até metade da taxa de amostragem. Ou seja, em um áudio com $48kHz$ de taxa de amostragem,
-conseguimos representar frequências ate $24kHz$. A frequência de Nyquist é exatamente a metade
+conseguimos representar frequências de até $24kHz$. A frequência de Nyquist é exatamente a metade
 da frequência de amostragem e é a frequência mais aguda que aquele sinal pode representar.
+(PROVA ELEGANTE DO TEOREMA DE Nyquist? TAMB REFERENCIAR OS SCRIPTS PARA ESTUDO DO FENOMENO)
 
@@
 destes formatos apresentam perda de qualidade e são mais indicados para serem transmitidos via rede ou
-para armazenamento. Abaixo está uma tabela com alguns dos formatos mais importantes que possuímos:
+para armazenamento\footnote{Os formatos que apresentam perda de qualidade com relação
+ao PCM são chamados formatos \emph{lossy}, os que não apresentam perda de qualidade
+são chamados \emph{loseless}}. Abaixo está uma tabela com alguns dos formatos mais importantes que possuímos:
 
@@
 
-Síntese, processamento ou análise.
+Síntese, alteração ou análise.
 
@@
 
-Procedimentos básicos. Processamento
+Procedimentos básicos: Lookuptable, sample. Processamento
 espectral. Outros processamentos.
@@
 
+Resumo de:
+The C programming Language
+
+C++?
+
+sndlib, SndObj, libsndfile
+
 \subsection{Python}
+
+Python in a Nutshell
+
+Tutorial padrão do van Rossum.
+
+Python the Hard Way?
+
+Pyo, interfaces para áudio e sinais Pywt, numpy, audiolab
+
+\subsection{Linguagens Dedicadas}
+
+Csound, ChucK, Supercollider, Puredata, Haskore,
 
@@
 
-O hardware estava ficando mais ambundante, sendo replicado em 
-vários lugares. Modelos eram vendidos com diferenças de hardware 
-de mínimas a discrepantes. Neste contexto, começou-se a
+O hardware estava sendo replicado mais facilmente e
+modelos eram vendidos com diferenças
+muitas vezes mínimas. Neste contexto, começou-se a
 pensar em uma camada de código que abstraísse o hardware, assim
@@
 
+No trabalho com tecnologias abertas, ainda é bastante cotidiana
+a utilização de sistemas Unix. O áudio e música não são excessões.
+Assim, faz-se cabivel uma abordagem breve dos sistemas GNU/Linux
+inclusive porque todos os desenvolvimentos desta dissertação se deram
+neste Sistema Operacional e algumas destas contribuições se revestiram
+poéticas específicas desta plataforma, como o ABT e o EKP.
+
 \subsection{Unix e GNU/LINUX}
-
 Resumo de utilização. No apêndice o resumo de:
@@
 Running Linux
+
+sox, dmesg > /dev/dsp, programas dedicados para áudio e para a música
 
@@
 http://www.html5rocks.com/en/tutorials/internals/howbrowserswork/
+
+Audiolet, que mais?
+
+\section{Sistemas de Controle de Versão}
+No desenvolvimento e compartilhamento efetivo dos códigos, faz-se
+necessária a utilização de ferramentas de registro dos históricos destes
+códigos. Isso é tão vital - ou mais vital - do que aprodução efetiva de código.
+
+Desta forma, abordaremos de forma bastante resumida o tema, de forma a deixar
+registrado o uso cotidiano que se faz de SCV para áudio e música, que não
+é bastante simples, ao contrário das usuais abordagens que se faz em manuais e livros
+sobre este assunto.
+
+\subsection{Breve histórico}
+Historico Cvs, Svn, Bitbucket, git, mercurial, baazar
+
+\subsection{Funcionamento básico}
+
+Usual: Após alterações finalizadas (que não quebrem a execução do código), registra-se conjuntos de alterações, e estas alterações são
+enviadas para um repositório central ou ao menos compartilhadas com outros.
+
+Variações: para manutenção pessoal de código. Conflitos de alterações.
+
+\subsection{Git}
+
+Roteiros básicos. Ver nas páginas do LM e do Automata.

### 2012-01-17 resumoeabstract.tex
feita uma revisao quase que completa 	modified:   APENDICE.tex 	modified:   agradecimentos.tex 	modified:   dedicatoria.tex 	modified:   devEresultados.tex 	modified:   fundamentos.tex 	modified:   resumoeabstract.tex

--- 2011-08-23 resumoeabstract.tex
@@
 
-Após uma breve contextualização sobre o status que o código e o áudio digital possuem na cultura atual,
-expomos os fundamentos do manuseio do som digital através do software em código aberto: fenômeno
-físico (e psicofísico), representação digital e procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
+Nesta dissertacao, expomos os fundamentos do manuseio do som digitalizado através do software em código aberto.
+Após uma breve contextualização sobre o status que o código e o áudio digital possuem na cultura atual, procuramos abordar como um todo o fenômeno
+físico (e psicofísico) do som, a sua representação digital e os procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
 são então abordadas, já com vistas a esta tarefa lidar com o áudio digital. Ainda no campo do software,
 os sistemas operacionais são então perspassados e o ambiente de desenvolvimento Unix é descrito de um ponto
-de vista bem resumido e ferramental. Os resultados antingidos com os fundamentos aqui expostos,
-assim como as repercussões destes desenvolvimentos são então sistematizados por escrito. 
+de vista bem resumido e ferramental. Um capítulo especial eh dedicado às linguagens dedicadas ao áudio e à música.
+Os resultados antingidos com os fundamentos aqui expostos são então sistematizados com ênfase nas ferramentas desenvolvidas.
+Apresentamos também uma sistematização das repercussões destes desenvolvimentos.
 Por fim, como um produto adicional relacionado a esta dissertação, é delineado o mapa de exposição e comunicação no qual
-os desenvolvimentos deste trabalho são agregados. Esta estrutura é dedicada a manter as contribuições
-de forma continuada e sistematizada através de webpages, wikis, repositórios, aplicativos de acompanhamento de projetos
-e uma tecnologia desenvolvida neste trabalho: A Autogestão Algorítmica.
+os desenvolvimentos deste trabalho estão agregados. Esta estrutura é dedicada a manter as contribuições
+de forma continuada e sistematizada através de webpages, wikis, repositórios, um canal IRC, aplicativos de acompanhamento de projetos
+e uma das tecnologias desenvolvidas neste trabalho: A Autogestão Algorítmica.
 
@@
 
-Após uma breve contextualização sobre o status que o código e o áudio digital possuem na cultura atual,
-expomos os fundamentos do manuseio do som digital através do software em código aberto: fenômeno
-físico (e psicofísico), representação digital e procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
+Nesta dissertacao, expomos os fundamentos do manuseio do som digitalizado através do software em código aberto.
+Após uma breve contextualização sobre o status que o código e o áudio digital possuem na cultura atual, procuramos abordar como um todo o fenômeno
+físico (e psicofísico) do som, a sua representação digital e os procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
 são então abordadas, já com vistas a esta tarefa lidar com o áudio digital. Ainda no campo do software,
 os sistemas operacionais são então perspassados e o ambiente de desenvolvimento Unix é descrito de um ponto
-de vista bem resumido e ferramental. Os resultados antingidos com os fundamentos aqui expostos,
-assim como as repercussões destes desenvolvimentos são então sistematizados por escrito. 
+de vista bem resumido e ferramental. Um capítulo especial eh dedicado às linguagens dedicadas ao áudio e à música.
+Os resultados antingidos com os fundamentos aqui expostos são então sistematizados com ênfase nas ferramentas desenvolvidas.
+Apresentamos também uma sistematização das repercussões destes desenvolvimentos.
 Por fim, como um produto adicional relacionado a esta dissertação, é delineado o mapa de exposição e comunicação no qual
-os desenvolvimentos deste trabalho são agregados. Esta estrutura é dedicada a manter as contribuições
-de forma continuada e sistematizada através de webpages, wikis, repositórios, aplicativos de acompanhamento de projetos
+os desenvolvimentos deste trabalho estão agregados. Esta estrutura é dedicada a manter as contribuições
+de forma continuada e sistematizada através de webpages, wikis, repositórios, um canal IRC, aplicativos de acompanhamento de projetos
 e uma tecnologia desenvolvida neste trabalho: A Autogestão Algorítmica.

### 2012-01-18 introducao.tex
revisada a apresentacao

@@
 Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})
+Partindo do som como um fenomeno fíisico, esta dissertação se propõe a ser um guia relativamente 
+completo para uma atuação eficiênte no desenvolvimento
+de software voltado ao áudio e tecnologias livres. Optamos
+realizar este trabalho através da síntese e da simplicidade.
 
-É um trabalho criterioso definir quais pontos cabem serem abordados em um trabalho como este.
-Como abordar a atuação tecnológica e científica em uma área criativa? Como contemplar
-a cultura de compartilhamento e diversidade em que esta produção tecnológica está inserida?
-Com estas questões em mente, optamos neste trabalho pela síntese e pela simplicidade.
-Além disso, esta dissertação se propõe a ser um guia relativamente completo para uma atuação eficiênte em desenvolvimento
-de software em multimídia, mais especificamente voltados áudio e tecnologias livres.
-
-Estaremos abordando o som, o áudio, o áudio digital. Práticas de programação voltadas para
-áudio e também as plataformas para as quais desenvolvemos código: Sistemas Operacionais e Browsers.
+Estaremos abordando o som, o áudio, o áudio digital e práticas de programação voltadas para
+áudio. Não menos importantes são as plataformas nas quais desenvolvemos e executamos código:
+Sistemas Operacionais e \emph{Browsers}.
 
@@
 
-%% ------------------------------------------------------------------------- %%
+Este trabalho tange algumas questões ainda mais gerais e difíceis.
+Como abordar a atuação tecnológica e científica em uma área criativa? Como contemplar
+a cultura de compartilhamento e diversidade em que esta produção tecnológica está inserida?
+Enfim, é um trabalho criterioso definir quais os pontos a serem abordados em um trabalho como este.
 
 \section{O som e o áudio}
-
 O som é tipicamente utilizado para fala, sinais em geral\footnote{p.ex. buzina} e música.
-Caracteriza-se por ser um sentido   O livre fluxo de informação
+Caracteriza-se por ser um sentido (estudar som na wikipedia e talvez helmholtz e zuckerkandl)
 
   \subsection{Breve histórico e práticas atuais}
-
 As explicações físicas do fenômeno do som no decorrer do tempo. Teorias
-corpusculares, escola de Tales. Helmholtz.
+corpusculares, escola de Tales. Helmholtz. Surgimento do áudio. Áudio digital.
 
   \subsection{Campos de estudo e atuação}
+som: Acúsitica, psicoacúsica, luthearia
+áudio: microfones, caixas de som
+áudio digital: conversores a/d d/a, programação voltada ao áudio e música
 
-\section{Código aberto, software livre e cultura livre} %Nome da seção.
-
+\section{Código aberto, software livre e cultura livre}
 O código é repetidamente destacado como o bem mais valioso produzido em nosso tempo.
@@
 É tido como o vetor de transformação de nossa época por excelência.
-Qualquer licença de software livre é também uma licença de código aberto (Open Source), a diferença entre as duas nomenclaturas reside essencialmente na sua apresentação e na filosofia.
+Qualquer licença de software livre é também uma licença de código aberto
+(Open Source), a diferença entre as duas nomenclaturas reside essencialmente
+na sua apresentação e na filosofia.
 
-\subsection{Cultura de compartilhamento} %Nome da seção.
-\label{sec:soft_compar} %Rótulo para futura referência à seção.
-Em qualquer lugar da tese, você poderá citar esta seção através de \label{sec:consideracoes_preliminares}.
-O mesmo vale para todas as próximas seções e capítulos.
+  \subsection{Cultura de compartilhamento}
+  \label{sec:soft_compar}
+Proposta de compartilhamento. Softwares para dar suporte a esta conectividade.
 
-Softwares para dar suporte a esta conectividade.
-
-%% ------------------------------------------------------------------------- %%
   \subsection{Licenças livres para software e para outras mídias}
   \label{sec:licencas_livres}
-
 Principais tipos de licenças
@@
   \label{sec:princ_licencas}
+Licenças uma a uma. GPL, LGPL, BSD, MIT, Beerware, etc
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%
 \section{Estado das coisas}
 \label{sec:context}
+Aqui no Brasil: as redes existentes, carta midias livres, fluxo de informação..
+No mundo: mudança de paradigma com relação a tudo que jah rolou e rolará:
+protestos, modismos, entidades, leis e tentativas de leis.
 
@@
 últimos fatores são cruciais para o trabalho aqui apresentado dada a notoriedade do papel do
-fenômeno sonoro para a comunicação e expressões culturais.
+fenômeno sonoro para a comunicação e expressões culturais. Mais recentemente, podemos falar claramente
+do papel do código nestas duas frentes.
 
@@
 
-HTML5 e Javascript deixando interativo. 3D, conexões rápidas que possuem banda para video HD e
+HTML5 CSS3 e Javascript deixando interativo. 3D, conexões rápidas que possuem banda para video HD e
 baixa latência. Popularização de hardware capaz destes processamentos.
@@
     \label{sec:gargalos}
-
 Deficiencias em disponibilização da tecnologia atual: muita
@@
 
-Citação da parte de video da thread 'Abrindo o Códifo Fechado'.
+Citação da parte de video da thread 'Abrindo o Código Fechado'.
 
@@
 
-Contribuimos na frente do software.
+Contribuimos na frente do software. Para isso veja [sessao dos resultados]
 
@@
     \label{sec:din_dev}
+A modularidade do sistema \emph{Unix type}. Codigo como substancia modular
+e reativa. Colaboratividade, listas de discussão, IRC.
 
-A modularidade do sistema \emph{Unix type} e as compensações: jack,
+No caso do áudio: as compensações: jack,
 facilidades de instalação e segurança, tudo hackeável.
@@
     \label{sec:comp_tec}
-
-Terreno fértil -> semear
+Terreno fértil -> semear que a primazia do compartilhamento se encarrega de
+desencadear os desenvolvimentos ligitimos.
 
@@
     \label{sec:tarefas_n_sup}
+[ESCREVER NO FINAL]
 
@@
     \label{sec:sl_prop}
+[ESCREVER NO FINAL E ABORDAR BREVEMENTE AS QUESTOES DE VIDEO]
 
@@
 \label{sec:objetivos}
+Aqui analisamos os objetivos deste trabalho. Eles refletem os objetivos
+dos empreendimentos de SL em geral e tambem das investidas no audiovisual e nas artes.
 
@@
   \label{sec:tutoriais}
-
 Descentralização: Fator caminhão e espalhamento de informação. Descentralização
@@
     \item Tutoriais para programação em python.
-
 Screen casts e texto.
@@
     \label{sec:dif_sl}
-
 Estamos desenvolvendo trabalhos em análise de comparação
@@
 
-Os estudos subsequentes a este trabalhos devem desenvolver
+Os estudos subsequentes a estes trabalhos devem desenvolver
 um estudo sobre a dinâmica de geração e propagação de conteúdo na rede.
-Comparando textos e ustilizando medidas de similaridade separa-se
+Comparando textos e utilizando medidas de similaridade separa-se
 o que um texto não contém e o que é semelhante, i.e. servem de propagação

### 2012-01-22 agradecimentos.tex
geralzona para fazer rodar o latex aqui. agora compila.
Revisao geral inteira feita

@@
-Em preimeiro lugar, agradeço aos meus pais e demais familiares. Agradeço aos
+Em primeiro lugar, agradeço aos meus pais e demais familiares. Agradeço aos
 meus amigos pelo carinho e companhia. Aos professores por tanta atenção e compreensão.
@@
 das contribuições citadas neste trabalho, construindo sobre elas e com elas. Especial agradecimento
-ao meu irmao Ricardo Fabbri que também se aproximou desta contribuição cuidando de processamento de imagens e video. Fortíssima referência
+ao meu irmão Ricardo Fabbri que também se aproximou destas contribuições cuidando de processamento de imagens e video. Fortíssima referência
 para mim, tanto para os assuntos tratados nesta dissertação, quanto para questões humanas. Agradecimento
@@
 Agradeço fortemente ao Prof. Dr. Rafael Santos Mendes pela pesquisa em Wavelets de mestrado em engenharia elétrica
-praticamente completa na FEEC/Unicamp que não defendendi por razões circunstanciais. Agradeço aos professores da FEEC pelos ensinos de
+praticamente completa na FEEC/Unicamp, titulo este que não defendendi por razões circunstanciais. Agradeço aos professores da FEEC pelos ensinos de
 primeiríssima qualidade em diversas disciplinas como processamento de sinais, teoria da informação, 
 circuitos elétricos, processos estocásticos
-e programação bio-inspirada. Sem sombra de dúvida agradecimentos especiais para os assíduos mentores
+e programação bio-inspirada. Sem sombra de dúvida maiores agradecimentos para os assíduos mentores
 deste trabalho: ao orientador-colaborador Prof. Luciano da Fontoura Costa, que, dentre outras qualidades,

### 2012-01-22 fundamentos.tex
geralzona para fazer rodar o latex aqui. agora compila.
Revisao geral inteira feita

@@
 \chapter{Fundamentos teóricos e técnicos} %Nome do capítulo.
-\label{cap:intro} %Rótulo para futura referência ao capítulo.
+% \label{cap:intro} %Rótulo para futura referência ao capítulo.
 Em qualquer lugar da tese, você poderá citar este capítulo através de ~\ref{cap:introducao}.
@@
 aprofundar este tema, recomendamos [][][] e uma representacao esquematica deste processo
-se encontra na figura \figure{fig:AD}.
+se encontra na figura.	 
+% \figure{fig:AD}.
 
@@
 
-[Tabela de formatos]
+TABELA DE FORMATOS
 
@@
 
-sox, dmesg > /dev/dsp, programas dedicados para áudio e para a música
+% sox, dmesg > /dev/dsp, programas dedicados para áudio e para a música
 

### 2012-01-22 pretextual.tex
revisada a apresentacao
geralzona para fazer rodar o latex aqui. agora compila.
Revisao geral inteira feita

@@
 		\item[OSS] Programas de Código Aberto \emph{(Open Source Software)}
-		\item[GNU] \emph{GNU is not UNIX}
+		\item[GNU] \emph{GNU is Not UNIX}
 		\item[GPL] \emph{General Public Licence}
-                \item[AA] \emph{Algorithmic Autoregularion}
-                \item[AC] \emph{Autogestão Coletiva}
+                \item[AA] \emph{Algorithmic Autoregularion ou Acrônimo Ambíguo}
+                \item[AC] \emph{Autogestão Coletiva ou Ágora Communs}
                 \item[CC] \emph{Creative Commons}
@@
                 \item[ONU] \emph{Organização das Nações Unidas}
+                \item[SL] \emph{Software Livre}
+                \item[EKP] \emph{Emotional Kernel Panic}
+                \item[SOS] \emph{Saúde Olha Sabedoria}
 	\end{listaespecial}
@@
 	\begin{listaespecial}[BIGNAMEWIDTH]
-		\item[q] Fluxo de massa em um trecho do duto
-		\item[q_i] Fluxo de massa entrando em um trecho do duto
+		\item[\$] Indica que o resto da linha é um comando bash
+		\item[\#] Resto da linha é comentario em código Bash ou Python
+		\item[\\\\] Resto da linha é comentario em código C/C++
 	\end{listaespecial}

### 2012-01-22 resumoeabstract.tex
geralzona para fazer rodar o latex aqui. agora compila.
Revisao geral inteira feita

@@
 Nesta dissertacao, expomos os fundamentos do manuseio do som digitalizado através do software em código aberto.
-Após uma breve contextualização sobre o status que o código e o áudio digital possuem na cultura atual, procuramos abordar como um todo o fenômeno
+Após uma brevíssima contextualização sobre o status que o código e o áudio digital possuem na cultura atual, procuramos abordar como um todo o fenômeno
 físico (e psicofísico) do som, a sua representação digital e os procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
@@
 Os resultados antingidos com os fundamentos aqui expostos são então sistematizados com ênfase nas ferramentas desenvolvidas.
-Apresentamos também uma sistematização das repercussões destes desenvolvimentos.
+Apresentamos também uma série de repercussões destes desenvolvimentos.
 Por fim, como um produto adicional relacionado a esta dissertação, é delineado o mapa de exposição e comunicação no qual
@@
 $\phantom{linha em branco}$\\
-Palavras-chave: áudio, programação, acústica, código aberto, cultura digital
+Palavras-chave: áudio, programação, acústica, código aberto, cultura digital, cultura hacker
 
@@
 de forma continuada e sistematizada através de webpages, wikis, repositórios, um canal IRC, aplicativos de acompanhamento de projetos
-e uma tecnologia desenvolvida neste trabalho: A Autogestão Algorítmica.
+e uma das tecnologias desenvolvidas neste trabalho: A Autogestão Algorítmica.
 
 $\phantom{linha em branco}$\\
-Keywords: audio, programming, acoustics, open source, digital culture
+Keywords: audio, programming, acoustics, open source, digital culture, hacker culture
 

### 2012-01-22 devEresultados.tex
feita uma revisao quase que completa 	modified:   APENDICE.tex 	modified:   agradecimentos.tex 	modified:   dedicatoria.tex 	modified:   devEresultados.tex 	modified:   fundamentos.tex 	modified:   resumoeabstract.tex
revisao geral dos materiais didaticos em textos.
	modified:   README.txt
	modified:   devEresultados.tex
	modified:   pretextual.tex

@@
 \chapter{Desenvolvimentos e resultados} %Nome do capítulo.
-\label{cap:intro} %Rótulo para futura referência ao capítulo.
+\label{cap:resultados} %Rótulo para futura referência ao capítulo.
 Em qualquer lugar da tese, você poderá citar este capítulo através de ~\ref{cap:introducao}.
@@
     \item Tutorial de python para áudio e som.
+Este tutorial foi levado para Berlim no LAC 2007 e sofreu melhoras desde entao. Esta
+primeira versao ficou resumida em forma de texto no EL\footnote{http://estudiolivre.org/python-e-som-tutorial}. Em 2010
+a Associacao Python Brasil escolheu este trabalho, então já mais amadurecido, para ser apresentado no
+FISL em Porto Alegre. Como consequencia, foi feita uma série de video-tutoriais bastante utilizados\footnote{http://estudiolivre.org/tiki-index.php?page=Video+Tutoriais}.
+Este tutorial foi comentado em listas em que o autor não participa (e outras em que o autor participa).
 
     \item Tutoriais de filtros e amostragem via python.
+Voltados para explicitar principios fundamentais de áudio, estes tutoriais
+são baseados código Python e o equivalente em C. Pequenas explicações são
+dadas com o intuito de orientar a exploração inteligente destes \emph{snippets}.
+
+{\bf Teorema de Amostragem}: estes scripts visam a experimentacao inteligente com
+o Teorema de Nyquist. (descricao)
+{\bf Filtros}: alem da explicitação sobre as diferenças entre filtros FIR e IIR,
+duas utilizações clássicas destes filtros estão implementadas: Wavelets (FIR) e Quad (IIR)
+
+Estes códigos podem ser baixados no repositório SVN do AudioExperiments. E os textos estão
+na wiki (nos digitais ou EL, recriar pois os CDTL foram apagados)
 
     \item Tutorial de plugins lv2
+Dadas as dificuldades que o desenvolvimento dos \emph{plugins} de áudio apresenta,
+desenvolvi um tutorial passo a passo com plugins que rodam em todas as etapas.
+Ele é baseado em uma interface C++ para este padrão de plugin que eh implementado
+em C. Os códigos e os textos estão todos em repositório.
 
     \item Microtutoriais Django ~\cite{dmicrotuts}.
+Estes 'microtutoriais' são baseados nos conceitos de \emph{scripts mínimos} e
+\emph{alterações puntuais}. O primeiro conjunto de microtutoriais é dedicado
+a reconstruir o tutorial oficial do django de forma condensada e não prolixa.
+O segundo destes conjuntos é dedicado a instrumentalizar de fato o leitor com
+o entendimento do funcionamento dos princípios fundamentais deste framework.
 
     \item Philosometrics
+Embora este não seja um trabalho didático propriamente dito, ele tem este intuito
+no cerne de sua concepção e surgimento. Em decorrência dele, surgiu o  Musimetrics,
+o Cinemetrics e o Literametrics. Além disso, ele é um belo exemplo da
+utilização das ciências duras para a análise de ciências humanas e foi acolhido
+como tal em alguns momentos.
 
     \item Carta mídias livres
+Texto criado em decorrência da participação da comissão de seleção no
+'Prêmio Mídias Livres', a convite do Ministério da Cultura por 'notório saber'.
+Esta carta é um documento único no seu conteúdo, deixando às claras
+o conceito de Mídias Livres como mídias não aprisionadas pelo conceito
+de propriedade, ou seja, que priorizam a sua livre circulação e a possibilidade
+de geração de materiais derivados, assim como sua geração aberta ao colaborativo e comunitário.
 
     \item Textos de cunho sociológico, transformador
+Produção mais numerosa que as anteriores, se caracteriza por métodos não convencionais
+de abordagem dos assuntos e de escrita. Em especial utiliza-se pseudônimos para
+auxiliar a despersonificação, gerando textos menos presos à satisfação da auto-imagem, dente
+outras qualidades. A utiliação de psudônimos é um costume muito apreciado em diversos meios,
+e as pesquisas tem confirmado as vantagens que a prática apresenta e confirma\footnote{http://disqus.com/research/pseudonyms/}.
+
+O autor destes textos se dá ao direito de não revelar seus psudônimos - embora muitos deles
+sejam publicamente conhecidos - para conservar as consequências desta prática na
+forma mais pura. Como comprovante desta produção, deixamos uma mensagem sobre a publicação
+de textos em mídia impressa com autores internacionais,
+confirmando a participação do autor desta dissertação, mas cujo
+nome não consta na publicação.
+
+!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
 
@@
 
+!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+
 \end{itemize}
@@
 diversas.
+    \begin{itemize}
+	\item Live-Coding
+	\item Raspagem de dados
+    \end{itemize}
 \end{itemize}
@@
 
+\subsection{AirHackTable}
+
 \subsection{Experimentos abertos}
@@
 
-Cadastro dos pontos
+(Cadastro dos pontos?)
+
+AA, SOS, Catalogo de Ideias, etc
 
@@
 
+Wiki?
+
 \subsection{Articulação}
 
-Em rede.
+IRC, Emails
 
@@
 
+\subsection{IRC}
+
+\subsection{Etherpads}
+
 \subsection{Outras fontes}

### 2012-02-03 tutoriais.tex
arrumacao principalmente quanto aos tutoriais, revisao

--- 2011-08-23 devEresultados.tex
@@
-%% ------------------------------------------------------------------------- %%
-\chapter{Desenvolvimentos e resultados} %Nome do capítulo.
-\label{cap:intro} %Rótulo para futura referência ao capítulo.
-Em qualquer lugar da tese, você poderá citar este capítulo através de ~\ref{cap:introducao}.
-Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})
-
-\section{Materiais didáticos}
+\chapter{Tutoriais e outros materiais didáticos} %Nome do capítulo.
+\label{cap:tutsDidads}
 
 \subsection{Tutoriais em texto e código}
-
 \begin{itemize}
@@
 
-de      fabi borges catadores@gmail.com por  riseup.net 
-responder a     submidialogia@lists.riseup.net
-para    submidialogia@lists.riseup.net
-data    23 de agosto de 2011 12:26
-assunto Re: [submidialogia] livro sub- publicação
-lista de e-mails        <submidialogia.lists.riseup.net> Filtrar as mensagens dessa lista de e-mails
-enviado por     lists.riseup.net
-assinado por    riseup.net
-cancelar inscrição      Cancelar a inscrição para essa lista de e-mails
-        Importante principalmente porque você frequentemente lê mensagens com esse marcador.
-ocultar detalhes 12:26 (6 minutos atrás)
-entao, eu fui recebendo textos durante esse tempo,
-alguns tao atrazados como dos sem satelites, mas muita gente mandou;
-
-aqui os autores:
-
-os internacionais nao sao muitos, o joni kempf (o que bebe ouro do hardware), o barbrook (futuros imaginarios),
-Hamdy heda (da revolucao egipcia), o pedro soller (summerlab), maria llopis (pos porno),  talvez a bronac, nao entregou ainda.
-
-dos brasileiros, renato fabri, ruiz, pasteur, morgana e caio, ju dornelles, mari marcassa, coletivo errorista, adriana velozo-drica, tiago pimentel, felipe fonseca, thiago novaes, lelex, felipe ribeiro(?), maira, verenilde,  bartolina silva, poro, vitoria amaro, fabib (eu),
-
-entao, precisa publicar agora,
-uma equipe para publicacao e,,,
-
-bjs
-f
-
 \end{itemize}
@@
     \item Canal Macambira
-No coletivo Macambira estão sendo produzidos materiais em screencasts sobre
-diversas.
+
+    \begin{itemize}
+	\item Live-Coding
+	\item Raspagem de dados
+    \end{itemize}
 \end{itemize}
@@
 \subsection{Figusdevpack (FDP)}
-
-Um ambiente de interação da comunidade de Python e Música 
-para compartilhamento de códigos e excertos. Baseado principalmente
-em documentação organizada sobre as práticas e as bibliotecas
-existentes para python. Assim como scripts compartilhados que
-fazem uso de objetos e módulos específicos. Este trabalho foi aceito na
-maior conferência de áudio em linux, a Linux Audio Conference de 2008
-(LAC2008) e está sendo reativado por mim em conjunto com Vilson Vieira
-e outros desenvolvedores de áudio. Este projeto está em desenvolvimento
-do site do Estúdio Livre [estudio livre] com repositório no sourceforge [source force].
-
-As principais fontes sobre estre trabalho é a página de desenvolvimento da ideia
-que está em constante mudança ~\cite{http://estudiolivre.org/tiki-index.php?page=fdp&highlight=fdp fdpel}
-e, o artigo que foi aceito no LAC de 2008 ~\cite{http://www.estudiolivre.org/el-gallery_view.php?arquivoId=8221 fdplac2008}
-e o repositório ~\cite{http://sourceforge.net/projects/fdpack/develop fdpsf}.
-
-\section{Áudio}
-
-\subsection{ABeatTracker (ABT)}
-
-Manual ABT
-
-Linkar com ABD
-
-\subsection{Emotional Kernel Panic (EKP)}
-
-Artigo FDP
-
-\subsection{Finite Groups in Granular Synthesis (FIGGS)}
-
-\subsection{Plugins LADSPA e lv2}
-
-\subsection{Processamento de voz via redes complexas}
-
-\subsection{Experimentos abertos}
-
-Audioexperiments. Estúdio livre.
-
-\section{Web}
-
-Difusão de informação com ênfase na facilitação
-da apropriação de tecnologias e de instancias políticas.
-
-\subsection{Sítios}
-
-FDDCA
-
-Ferramenta de comunicação
-
-Cadastro dos pontos
-
-Meu site pessoal
-
-\subsection{Conteúdos}
-
-\subsubsection{textos}
-Carta mídias livres
-
-Artigos,
-
-Tutoriais
-
-\subsection{Articulação}
-
-Em rede.
-
-\section{Disponibilização e Desenvolvimento}
-
-\subsection{Wiki}
-
-\subsection{Trac}
-
-\subsection{Screencasts - Vimeo}
-
-\subsection{AA}
-
-\subsection{Outras fontes}

### 2012-02-06 agradecimentos.tex
terminada a revisao geral da dissertacao

@@
 Marostegan e Carneiro pelas lições e confianças. Agradeço ao profícuo Vilson Vieira, que tornou-se especialmente próximo
-das contribuições citadas neste trabalho, construindo sobre elas e com elas. Especial agradecimento
-ao meu irmão Ricardo Fabbri que também se aproximou destas contribuições cuidando de processamento de imagens e video. Fortíssima referência
-para mim, tanto para os assuntos tratados nesta dissertação, quanto para questões humanas. Agradecimento
+das contribuições citadas neste trabalho, construindo sobre elas e com elas. Condizente com a fortíssima referência
+para que é para mim, tanto para os assuntos tratados nesta dissertação quanto para questões humanas, um especial agradecimento
+ao meu irmão Ricardo Fabbri que também se aproximou destas contribuições cuidando de processamento de imagens e video. Agradecimento
 forte ao Lab Macambira, nominalmente estes assíduos: [Equipe Lab Macambira]. Um profundo agradecimento

### 2012-02-06 devEresultados.tex
terminada a revisao geral da dissertacao

@@
 
+Os vários materiais didáticos produzidos constam no apêndice
+deste trabalho. Um único destes será exposto a seguir por sua
+capacidade de agregar os conteúdos dos capítulos anterioes:
+os tutoriais de filtros e amostragem.
+
 \begin{itemize}
-    \item Tutorial de python para áudio e som.
+    \item {\bf Tutorial de python para áudio e som}
+
 Este tutorial foi levado para Berlim no LAC 2007 e sofreu melhoras desde entao. Esta
@@
 
-    \item Tutoriais de filtros e amostragem via python.
+    \item {\bf Tutoriais de filtros e amostragem via python}
+
 Voltados para explicitar principios fundamentais de áudio, estes tutoriais
@@
 
-{\bf Teorema de Amostragem}: estes scripts visam a experimentacao inteligente com
+\emph{Teorema de Amostragem}: estes scripts visam a experimentacao inteligente com
 o Teorema de Nyquist. (descricao)
-{\bf Filtros}: alem da explicitação sobre as diferenças entre filtros FIR e IIR,
+
+\emph{Filtros}: alem da explicitação sobre as diferenças entre filtros FIR e IIR,
 duas utilizações clássicas destes filtros estão implementadas: Wavelets (FIR) e Quad (IIR)
@@
 
-    \item Tutorial de plugins lv2
+    \item {\bf Tutorial de plugins lv2}
+
 Dadas as dificuldades que o desenvolvimento dos \emph{plugins} de áudio apresenta,
@@
 
-    \item Microtutoriais Django ~\cite{dmicrotuts}.
+    \item {\bf Microtutoriais Django ~\cite{dmicrotuts}}
+
 Estes 'microtutoriais' são baseados nos conceitos de \emph{scripts mínimos} e
@@
 
-    \item Philosometrics
+    \item {\bf Philosometrics}
+
 Embora este não seja um trabalho didático propriamente dito, ele tem este intuito
@@
 
-    \item Carta mídias livres
+    \item {\bf Carta mídias livres}
+
 Texto criado em decorrência da participação da comissão de seleção no
@@
 
-    \item Textos de cunho sociológico, transformador
+    \item {\bf Textos de cunho sociológico, transformador}
+
 Produção mais numerosa que as anteriores, se caracteriza por métodos não convencionais
@@
 
-\subsection{Protocolo de compressão de áudio via wavelets, polinômios e permutações}
+\subsection{Protocolo de compactação de áudio via wavelets, polinômios e permutações}
 
@@
 \subsection{AirHackTable}
+
+\subsection{Livecoding}
 

### 2012-02-06 resumoeabstract.tex
terminada a revisao geral da dissertacao

@@
 
-Nesta dissertacao, expomos os fundamentos do manuseio do som digitalizado através do software em código aberto.
+Nesta dissertacao, expomos os fundamentos do manuseio do som digitalizado através de recursos em código aberto.
 Após uma brevíssima contextualização sobre o status que o código e o áudio digital possuem na cultura atual, procuramos abordar como um todo o fenômeno
 físico (e psicofísico) do som, a sua representação digital e os procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
-são então abordadas, já com vistas a esta tarefa lidar com o áudio digital. Ainda no campo do software,
-os sistemas operacionais são então perspassados e o ambiente de desenvolvimento Unix é descrito de um ponto
-de vista bem resumido e ferramental. Um capítulo especial eh dedicado às linguagens dedicadas ao áudio e à música.
-Os resultados antingidos com os fundamentos aqui expostos são então sistematizados com ênfase nas ferramentas desenvolvidas.
-Apresentamos também uma série de repercussões destes desenvolvimentos.
+são então abordadas já com vistas ã tarefa de lidar com o áudio digital. Um capítulo especial é dedicado às linguagens dedicadas ao áudio e à música. Ainda no campo do software,
+o ambiente de desenvolvimento Unix é descrito de um ponto
+de vista bem resumido e ferramental, pois é a base para todos os desenvolvimentos expostos nesta dissertação.
+Os resultados antingidos com os fundamentos expostos são então sistematizados com ênfase nas ferramentas desenvolvidas.
+Apresentamos também uma série de repercussões destes desenvolvimentos, em especial uma frente de ação criada: o Lab Macambira.
 Por fim, como um produto adicional relacionado a esta dissertação, é delineado o mapa de exposição e comunicação no qual
@@
 de forma continuada e sistematizada através de webpages, wikis, repositórios, um canal IRC, aplicativos de acompanhamento de projetos
-e uma das tecnologias desenvolvidas neste trabalho: A Autogestão Algorítmica.
+e uma das tecnologias desenvolvidas neste trabalho: a Autogestão Algorítmica.
 
 $\phantom{linha em branco}$\\
-Palavras-chave: áudio, programação, acústica, código aberto, cultura digital, cultura hacker
+Palavras-chave: acústica, áudio, código aberto, cultura digital, cultura hacker, programação, som
 
@@
 de forma continuada e sistematizada através de webpages, wikis, repositórios, um canal IRC, aplicativos de acompanhamento de projetos
-e uma das tecnologias desenvolvidas neste trabalho: A Autogestão Algorítmica.
+e uma das tecnologias desenvolvidas neste trabalho: a Autogestão Algorítmica.
 

### 2012-02-06 introducao.tex
ajeitado primeiro bloco da introducao

@@
 
-Estaremos abordando o som, o áudio, o áudio digital e práticas de programação voltadas para
-áudio. Não menos importantes são as plataformas nas quais desenvolvemos e executamos código:
-Sistemas Operacionais e \emph{Browsers}.
+Abordamos o som, o áudio, o áudio digital e práticas de programação voltadas para
+áudio. Não menos importantes são as plataformas nas quais desenvolvemos e executamos código
+- usualmente GNU/Linux ou BSD e navegadores de páginas de internet (\emph{browsers}) - e
+os sistemas de controle de versão dos códigos.
 
@@
 a cultura de compartilhamento e diversidade em que esta produção tecnológica está inserida?
-Enfim, é um trabalho criterioso definir quais os pontos a serem abordados em um trabalho como este.
+Enfim, é um trabalho criterioso definir quais os pontos a serem abordados em
+um trabalho como este.
 

### 2012-02-18 devEresultados.tex
escrito sobre o FIGGUS, partindo para minimum-fi

@@
 
-\subsection{Finite Groups in Granular Synthesis (FIGGS)}
+\subsection{Finite Groups in Granular and Unit Synthesis (FIGGUS)}
+
+Desenvolvimento iniciado em 2006 com Prof. Adolfo Maia Junior para
+tratar de simetrias na música, com vistas à composição musical atraves
+de métodos matemáticos. Mais especificamente, a proposta resultou em
+um programa voltado para o uso de Grupos Algébricos para síntese
+granular e síntese de estruturas musicais. O nome dado
+foi FIGGUS, sigla de FInite Groups in Granular and Unit Synthesis.
+
+A Síntese Granular é uma área bem estabelecida tanto na acústica quanto
+na Computação Musical. O Físico Gabor, quantum sonoro, phonon, etc..
+[achar artigos feitos com adolfo]
+
+Na atual reescrita, embora ainda bastante atras do FIGGUS original,
+mesmo sem contar a interface grafica, o codigo roda inteiro em python
+com as biblitecas imbutidas por padrão. Isso permite com que o FIGGUS
+sintetize todo um EP usando somente os comandos:
+
+[[ARRUMAR A PARADA PARA O CODIGO]]
+git clone
+python setup.py install
+python RUNME
 

### 2012-02-20 devEresultados.tex
resstruturado a parte de audio e musica para metade dos itens

@@
 
-\subsection{ABeatTracker (ABT)}
+exp abertos
+
+Do produção de códigos em tempo real para música, chamado livecoding,
+vamos para ferramentas computacionais para áudio e música, com o minimum-fi
+e FIGGUS. Já fazendo ponte do código com outras interfaces, adentramos AHT e EKP,
+que interfaceiam com o mundo físico e material (AHT) e com o hardware, onde
+nosso entendimento em codigo é efetuado em procedimentos.
+
+  \subsection{Experimentos abertos}
+
+Todos os desenvolvimentos desta dissertação estão em repositórios abertos\cite{repositorios-tese-dev},
+e alguns foram especialmente importantes como percurso para o que é apresentado neste
+escrito. Ou seja,
+
+Focaremos no código para a arte sonora,
+incluindo a musical, e portanto fazendo parte da cultura e fazendo cultura.
+\footnote{As chamadas culturas biopunk, ciberpunk, cipherpunk, hacker, digital e outras mais,
+possibilitadas aos recentes desenvolvimentos em telecomunicação, dizem respeito em menor
+ou maior grau ao código.}
+
+A seguir apresentamos alguns bons exemplos dos desenvolvimentos
+desta dissertação que não são de cunho artístico.
+
+      \subsubsection{Plugins LADSPA e lv2}
+      [repos AE, LM, wiki EL, historico CDTL]
+
+      \subsubsection{Protocolo de compactação de áudio via wavelets, polinômios e permutações}
+      [achar artigo feec]
+
+      \subsubsection{Processamento de voz via redes complexas}
+      [repo do audio experiments. passar para o LM e chamar o TT]
+
+  \subsection{Livecoding e ABeatTracker (ABT)}
+
+  [Apresentar Livecoding.]
+
+  Apenas parte dos conhecedores do assunto reconhecem
+  o ABT como livecoding. Alguns
+  consideram a linguagem por macros mais especializadas e poderosas como
+  sendo um tipo de interface escrita, nao uma linguagem de programação
+  propriamente dita.
+
+  Argumentar em favor do ABT como livecoding, apresentá-lo e depois
+  faze-lo
 
@@
 
-\subsection{Emotional Kernel Panic (EKP)}
+      \subsubsection{Audioexperiments, Estudiolivre e Foobarbaz}
 
-Em 2008, colaborando intensamente com o CDTL
-(Centro de Desenvolvimento de Tecnologias Livres)\footnote{foi uma associação civil formada e desmembrada em 2008 e sediada em Recife, PE}
-foi lançada a ideia de utilizar o estado do sistema operacional - especialmente o kernel linux - para
-geração de sons. Surge o Emotional Kernel Panic (EKP). Desde o inicio, foram definidos três finalidades
-para esta exploração do SO:
+      Audioexperiments. Estúdio livre. LM
 
+  \subsection{Musica em Tempo Diferido}
+
+      \subsubsection{Minimum-fi}
+
+Em Abril de 2011 foi trabalhado um codigo dedicado a materializar o minimo
+para se conseguir expressar ideias musicais em geral. Esta empreitada
+resultou no minimum-fi, codigo em um unico arquivo, que sintetiza
+musicas inteiras. O os algoritmos propriamente ditos somam menos de
+50 linhas e nao mais que 5 funções.
+
+Os princípios são:
 \begin{itemize}
-    \item Didáticos
-    \item Artísticos
-    \item Monitoramento do SO
+  \item Deve-se ter um mecanismo de sintese simples mas que
+permita varios timbres.
+  \item Deve-se ser capaz de construir series de notas e acordes.
 \end{itemize}
 
-http://trac.assembla.com/audioexperiments/browser/ekp-base
+Para o primeiro item, se prestam as funções lookup e lookupcruz. A função
+lookup já foi apresentada na sessão anterior, quando tratado de Python já visando o áudio.
+O lookupcruz ex explicado a seguir.
 
-\subsection{Finite Groups in Granular and Unit Synthesis (FIGGUS)}
+[codigo do lookupcruz e explicacao]
 
-Desenvolvimento iniciado em 2006 com Prof. Adolfo Maia Junior para
+As sequencias de notas e os acordes foram resolvidos com 2 outras funcoes simples,
+que utilizam lookup e lookupcruz.
+
+[codigo do fazSequencia e explicacao]
+
+[codigo do fazAcorde e explicacao]
+
+A função somador eh para automatizar a soma amostra a amostra e completar com zeros a
+sequencia menor.
+
+[codigo do somador, explicacao e figura de uma sequencia maior/menor que a outra]
+
+Depois disso é usufruir com estruturas criadas. Por exemplo: depois de construidas
+algumas tabelas para serem usadas como diferentes timbres, pode-se criar as
+escalas completamente simetricas assim:
+
+escala\_1=range(12) # cromática ascendente sem inclusão da oitava
+escala\_2=range(0,12,2) # tons inteiros
+escala\_3=range(0,12,3) # diminutão
+escala\_4=range(0,12,4) # terças maiores
+escala\_6=range(0,12,6) # trítonos
+
+Sendo cada unidade um semitom. Assim as triades maiores e menores
+são especificadas com simplicidade:
+
+maior=[0,4,7]
+menor=[0,3,7]
+
+E series podem ser anotadas junto a variacoes:
+
+b1=[0]*4
+b2=[0]*7+[7]
+b3=  [0,7,0,1]
+b3\_2=[0,7,8,7]
+b4=[0,7,7,7]
+b5=[0,7]
+b6=[0,19,7,12]
+b7=[0,-1,0,-1]
+b8=[0,-1,7,-1]
+b9=[0,1,0,-1]
+b10=[0,7,6,7]
+
+Agora eh sintetizar e mixar. A sintese de sequencias e acordes sao feitos tipicamente
+atraves de comandos desta forma:
+
+fazSequencia(nt, f, fator, d, s, senoide, dente)
+fazAcorde([0,7,14,21,28,35,42,49,56,63,70,77,84]
+
+Aqui uns procedimentos basicos utilizados nas montagens:
+
+seq=[1.5*i for i in fazSequencia(nt, f, fator, d, s, senoide, dente)] # amplificacao
+
+seq3=[(i+j) for i,j in zip(seq1,seq2)] # mixagem
+
+Gerando uma serie de acordes:
+
+ac=[]
+for i in xrange(8): # a cada 2 tempos
+    ac+=fazAcorde([0,7,14,21,28,35,42,49,56,63,70,77,84], f, fator, d, s, senoide, dente)
+    ac+=[0]*int((d+s)*44100)
+
+E por fim, mixando quando os comprimentos nao sao os mesmos:
+
+seq2=somador(ac,seq2)
+
+\subsubsection{Finite Groups in Granular and Unit Synthesis (FIGGUS)}
+
+Desenvolvimento iniciado em 2006 com o físico-matemático Prof. Adolfo Maia Junior para
 tratar de simetrias na música, com vistas à composição musical atraves
@@
 
-\subsection{Plugins LADSPA e lv2}
+  \subsection{Música na matéria}
+      \subsubsection{Emotional Kernel Panic (EKP)}
 
-\subsection{Protocolo de compactação de áudio via wavelets, polinômios e permutações}
+Em 2008, colaborando intensamente com o CDTL
+(Centro de Desenvolvimento de Tecnologias Livres)\footnote{foi uma associação civil formada e desmembrada em 2008 e sediada em Recife, PE}
+foi lançada a ideia de utilizar o estado do sistema operacional - especialmente o kernel linux - para
+geração de sons. Surge o Emotional Kernel Panic (EKP). Desde o inicio, foram definidos três finalidades
+para esta exploração do SO:
 
-\subsection{Processamento de voz via redes complexas}
+\begin{itemize}
+    \item Didáticos
+    \item Artísticos
+    \item Monitoramento do SO
+\end{itemize}
 
-\subsection{AirHackTable}
+http://trac.assembla.com/audioexperiments/browser/ekp-base
 
-\subsection{Livecoding}
-
-\subsection{Experimentos abertos}
-
-Audioexperiments. Estúdio livre.
+      \subsubsection{AirHackTable}
 

### 2012-02-20 devEresultados.tex
feitas alteracoes estruturais no msc todo

@@
 
-  \subsection{Experimentos abertos}
+  \subsection{Experimentos abertos em áudio: LADSPAs, Wavelets e Redes Complexas}
 
@@
 A seguir apresentamos alguns bons exemplos dos desenvolvimentos
-desta dissertação que não são de cunho artístico.
+desta dissertação especificamente em áudio, sem o envolvimento direto da música.
 
@@
 
-  \subsection{Livecoding e ABeatTracker (ABT)}
+  \subsection{Música em Tempo Real: Livecoding e ABeatTracker (ABT)}
 
@@
 
-  Argumentar em favor do ABT como livecoding, apresentá-lo e depois
-  faze-lo
+      \subsubsection{ABeatTracker}
 
@@
 
-  \subsection{Musica em Tempo Diferido}
+  \subsection{Musica em Tempo Diferido: Minimum-fi e FIGGUS}
 
@@
 
-  \subsection{Música na matéria}
+  \subsection{Música na Matéria: EKP e AHT}
       \subsubsection{Emotional Kernel Panic (EKP)}

### 2012-02-20 fundamentos.tex
terminada a revisao geral da dissertacao
feitas alteracoes estruturais no msc todo

@@
 
-\subsection{O som como fenômeno físico}
+  \subsection{O som e a audição}
 
@@
 
-\subsection{A audição e a psicofísica}
+      \subsubsection{A audição e a psicofísica}
 
@@
 
-\subsection{C/C++}
+  \subsection{Som em linguagens de uso geral: C/C++ e Python}
+
+      \subsubsection{C/C++}
 
@@
 
-\subsection{Python}
+      \subsubsection{Python}
 
@@
 
-\subsection{Linguagens Dedicadas}
+  \subsection{Linguagens Dedicadas ao áudio e à música}
+Csound, ChucK, Supercollider, Puredata, Haskore
 
-Csound, ChucK, Supercollider, Puredata, Haskore,
+  \subsection{Do meio e do sistema: Unix, Browsers e Git}
 
-\section{Sistema Operacional e Web}
-
-\subsection{Breve histórico}
-
-O hardware estava sendo replicado mais facilmente e
-modelos eram vendidos com diferenças
-muitas vezes mínimas. Neste contexto, começou-se a
-pensar em uma camada de código que abstraísse o hardware, assim
-como a linguagem C se propôs.
-
-Iniciou-se assim o desenvolvimento de Sistemas Operacionais (SO).
-Um SO é um software que gerencia os recursos do computador e ajuda
-também em abstrair o hardware. Orientamos a esta excelênte linha do
-tempo da Wikipédia em língua inglesa \cite{solinhadotempo}:
-
-No trabalho com tecnologias abertas, ainda é bastante cotidiana
-a utilização de sistemas Unix. O áudio e música não são excessões.
-Assim, faz-se cabivel uma abordagem breve dos sistemas GNU/Linux
-inclusive porque todos os desenvolvimentos desta dissertação se deram
-neste Sistema Operacional e algumas destas contribuições se revestiram
-poéticas específicas desta plataforma, como o ABT e o EKP.
-
-\subsection{Unix e GNU/LINUX}
-Resumo de utilização. No apêndice o resumo de:
-The Unix programing envinoment
-e
-Running Linux
-
-% sox, dmesg > /dev/dsp, programas dedicados para áudio e para a música
-
-\subsection{Web e Browsers}
-
-Dados do livrinho do google.
-
-Possibilidades de uso.
-
-Breve Histórico.
-
-No apêndice resumo de:
-http://www.html5rocks.com/en/tutorials/internals/howbrowserswork/
-
-Audiolet, que mais?
-
-\section{Sistemas de Controle de Versão}
-No desenvolvimento e compartilhamento efetivo dos códigos, faz-se
-necessária a utilização de ferramentas de registro dos históricos destes
-códigos. Isso é tão vital - ou mais vital - do que aprodução efetiva de código.
-
-Desta forma, abordaremos de forma bastante resumida o tema, de forma a deixar
-registrado o uso cotidiano que se faz de SCV para áudio e música, que não
-é bastante simples, ao contrário das usuais abordagens que se faz em manuais e livros
-sobre este assunto.
-
-\subsection{Breve histórico}
-Historico Cvs, Svn, Bitbucket, git, mercurial, baazar
-
-\subsection{Funcionamento básico}
-
-Usual: Após alterações finalizadas (que não quebrem a execução do código), registra-se conjuntos de alterações, e estas alterações são
-enviadas para um repositório central ou ao menos compartilhadas com outros.
-
-Variações: para manutenção pessoal de código. Conflitos de alterações.
-
-\subsection{Git}
-
-Roteiros básicos. Ver nas páginas do LM e do Automata.
+% \section{Plataformas: Unix e Navegadores da Web}
+% 
+% \subsection{Breve histórico}
+% 
+% O hardware estava sendo replicado mais facilmente e
+% modelos eram vendidos com diferenças
+% muitas vezes mínimas. Neste contexto, começou-se a
+% pensar em uma camada de código que abstraísse o hardware, assim
+% como a linguagem C se propôs.
+% 
+% Iniciou-se assim o desenvolvimento de Sistemas Operacionais (SO).
+% Um SO é um software que gerencia os recursos do computador e ajuda
+% também em abstrair o hardware. Orientamos a esta excelênte linha do
+% tempo da Wikipédia em língua inglesa \cite{solinhadotempo}:
+% 
+% No trabalho com tecnologias abertas, ainda é bastante cotidiana
+% a utilização de sistemas Unix. O áudio e música não são excessões.
+% Assim, faz-se cabivel uma abordagem breve dos sistemas GNU/Linux
+% inclusive porque todos os desenvolvimentos desta dissertação se deram
+% neste Sistema Operacional e algumas destas contribuições se revestiram
+% poéticas específicas desta plataforma, como o ABT e o EKP.
+% 
+% \subsection{GNU/Linux}
+% Resumo de utilização. No apêndice o resumo de:
+% The Unix programing envinoment
+% e
+% Running Linux
+% 
+% % sox, dmesg > /dev/dsp, programas dedicados para áudio e para a música
+% 
+% \subsection{Browsers}
+% 
+% Dados do livrinho do google.
+% 
+% Possibilidades de uso.
+% 
+% Breve Histórico.
+% 
+% No apêndice resumo de:
+% http://www.html5rocks.com/en/tutorials/internals/howbrowserswork/
+% 
+% Audiolet, que mais?
+% 
+% \section{Sistemas de Controle de Versão}
+% No desenvolvimento e compartilhamento efetivo dos códigos, faz-se
+% necessária a utilização de ferramentas de registro dos históricos destes
+% códigos. Isso é tão vital - ou mais vital - do que aprodução efetiva de código.
+% 
+% Desta forma, abordaremos de forma bastante resumida o tema, de forma a deixar
+% registrado o uso cotidiano que se faz de SCV para áudio e música, que não
+% é bastante simples, ao contrário das usuais abordagens que se faz em manuais e livros
+% sobre este assunto.
+% 
+% \subsection{Breve histórico}
+% Historico Cvs, Svn, Bitbucket, git, mercurial, baazar
+% 
+% \subsection{Funcionamento básico}
+% 
+% Usual: Após alterações finalizadas (que não quebrem a execução do código), registra-se conjuntos de alterações, e estas alterações são
+% enviadas para um repositório central ou ao menos compartilhadas com outros.
+% 
+% Variações: para manutenção pessoal de código. Conflitos de alterações.
+% 
+% \subsection{Git}
+% 
+% Roteiros básicos. Ver nas páginas do LM e do Automata.

### 2012-02-20 introducao.tex
arrumada a introducao

@@
 
-  \subsection{Breve histórico e práticas atuais}
+	  \subsubsection{Breve histórico e práticas atuais}
 	As explicações físicas do fenômeno do som no decorrer do tempo. Teorias
@@
 
-  \subsection{Campos de estudo e atuação}
+	  \subsubsection{Campos de conhecimento e atuação}
 	som: Acúsitica, psicoacúsica, luthearia

### 2012-02-20 devEresultados.tex
terminada reestruturacao da dissertacao

@@
 Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})
-
-\section{Materiais didáticos}
-
-\subsection{Tutoriais em texto e código}
-
-Os vários materiais didáticos produzidos constam no apêndice
-deste trabalho. Um único destes será exposto a seguir por sua
-capacidade de agregar os conteúdos dos capítulos anterioes:
-os tutoriais de filtros e amostragem.
-
-\begin{itemize}
-    \item {\bf Tutorial de python para áudio e som}
-
-Este tutorial foi levado para Berlim no LAC 2007 e sofreu melhoras desde entao. Esta
-primeira versao ficou resumida em forma de texto no EL\footnote{http://estudiolivre.org/python-e-som-tutorial}. Em 2010
-a Associacao Python Brasil escolheu este trabalho, então já mais amadurecido, para ser apresentado no
-FISL em Porto Alegre. Como consequencia, foi feita uma série de video-tutoriais bastante utilizados\footnote{http://estudiolivre.org/tiki-index.php?page=Video+Tutoriais}.
-Este tutorial foi comentado em listas em que o autor não participa (e outras em que o autor participa).
-
-    \item {\bf Tutoriais de filtros e amostragem via python}
-
-Voltados para explicitar principios fundamentais de áudio, estes tutoriais
-são baseados código Python e o equivalente em C. Pequenas explicações são
-dadas com o intuito de orientar a exploração inteligente destes \emph{snippets}.
-
-\emph{Teorema de Amostragem}: estes scripts visam a experimentacao inteligente com
-o Teorema de Nyquist. (descricao)
-
-\emph{Filtros}: alem da explicitação sobre as diferenças entre filtros FIR e IIR,
-duas utilizações clássicas destes filtros estão implementadas: Wavelets (FIR) e Quad (IIR)
-
-Estes códigos podem ser baixados no repositório SVN do AudioExperiments. E os textos estão
-na wiki (nos digitais ou EL, recriar pois os CDTL foram apagados)
-
-    \item {\bf Tutorial de plugins lv2}
-
-Dadas as dificuldades que o desenvolvimento dos \emph{plugins} de áudio apresenta,
-desenvolvi um tutorial passo a passo com plugins que rodam em todas as etapas.
-Ele é baseado em uma interface C++ para este padrão de plugin que eh implementado
-em C. Os códigos e os textos estão todos em repositório.
-
-    \item {\bf Microtutoriais Django ~\cite{dmicrotuts}}
-
-Estes 'microtutoriais' são baseados nos conceitos de \emph{scripts mínimos} e
-\emph{alterações puntuais}. O primeiro conjunto de microtutoriais é dedicado
-a reconstruir o tutorial oficial do django de forma condensada e não prolixa.
-O segundo destes conjuntos é dedicado a instrumentalizar de fato o leitor com
-o entendimento do funcionamento dos princípios fundamentais deste framework.
-
-    \item {\bf Philosometrics}
-
-Embora este não seja um trabalho didático propriamente dito, ele tem este intuito
-no cerne de sua concepção e surgimento. Em decorrência dele, surgiu o  Musimetrics,
-o Cinemetrics e o Literametrics. Além disso, ele é um belo exemplo da
-utilização das ciências duras para a análise de ciências humanas e foi acolhido
-como tal em alguns momentos.
-
-    \item {\bf Carta mídias livres}
-
-Texto criado em decorrência da participação da comissão de seleção no
-'Prêmio Mídias Livres', a convite do Ministério da Cultura por 'notório saber'.
-Esta carta é um documento único no seu conteúdo, deixando às claras
-o conceito de Mídias Livres como mídias não aprisionadas pelo conceito
-de propriedade, ou seja, que priorizam a sua livre circulação e a possibilidade
-de geração de materiais derivados, assim como sua geração aberta ao colaborativo e comunitário.
-
-    \item {\bf Textos de cunho sociológico, transformador}
-
-Produção mais numerosa que as anteriores, se caracteriza por métodos não convencionais
-de abordagem dos assuntos e de escrita. Em especial utiliza-se pseudônimos para
-auxiliar a despersonificação, gerando textos menos presos à satisfação da auto-imagem, dente
-outras qualidades. A utiliação de psudônimos é um costume muito apreciado em diversos meios,
-e as pesquisas tem confirmado as vantagens que a prática apresenta e confirma\footnote{http://disqus.com/research/pseudonyms/}.
-
-O autor destes textos se dá ao direito de não revelar seus psudônimos - embora muitos deles
-sejam publicamente conhecidos - para conservar as consequências desta prática na
-forma mais pura. Como comprovante desta produção, deixamos uma mensagem sobre a publicação
-de textos em mídia impressa com autores internacionais,
-confirmando a participação do autor desta dissertação, mas cujo
-nome não consta na publicação.
-
-!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-
-de      fabi borges catadores@gmail.com por  riseup.net 
-responder a     submidialogia@lists.riseup.net
-para    submidialogia@lists.riseup.net
-data    23 de agosto de 2011 12:26
-assunto Re: [submidialogia] livro sub- publicação
-lista de e-mails        <submidialogia.lists.riseup.net> Filtrar as mensagens dessa lista de e-mails
-enviado por     lists.riseup.net
-assinado por    riseup.net
-cancelar inscrição      Cancelar a inscrição para essa lista de e-mails
-        Importante principalmente porque você frequentemente lê mensagens com esse marcador.
-ocultar detalhes 12:26 (6 minutos atrás)
-entao, eu fui recebendo textos durante esse tempo,
-alguns tao atrazados como dos sem satelites, mas muita gente mandou;
-
-aqui os autores:
-
-os internacionais nao sao muitos, o joni kempf (o que bebe ouro do hardware), o barbrook (futuros imaginarios),
-Hamdy heda (da revolucao egipcia), o pedro soller (summerlab), maria llopis (pos porno),  talvez a bronac, nao entregou ainda.
-
-dos brasileiros, renato fabri, ruiz, pasteur, morgana e caio, ju dornelles, mari marcassa, coletivo errorista, adriana velozo-drica, tiago pimentel, felipe fonseca, thiago novaes, lelex, felipe ribeiro(?), maira, verenilde,  bartolina silva, poro, vitoria amaro, fabib (eu),
-
-entao, precisa publicar agora,
-uma equipe para publicacao e,,,
-
-bjs
-f
-
-!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-
-\end{itemize}
-
-\subsection{Screencasts}
-
-\begin{itemize}
-    \item Python para áudio e música
-
-    \item Canal Macambira
-No coletivo Macambira estão sendo produzidos materiais em screencasts sobre
-diversas.
-    \begin{itemize}
-	\item Live-Coding
-	\item Raspagem de dados
-    \end{itemize}
-\end{itemize}
-
-\subsection{Figusdevpack (FDP)}
-
-Um ambiente de interação da comunidade de Python e Música 
-para compartilhamento de códigos e excertos. Baseado principalmente
-em documentação organizada sobre as práticas e as bibliotecas
-existentes para python. Assim como scripts compartilhados que
-fazem uso de objetos e módulos específicos. Este trabalho foi aceito na
-maior conferência de áudio em linux, a Linux Audio Conference de 2008
-(LAC2008) e está sendo reativado por mim em conjunto com Vilson Vieira
-e outros desenvolvedores de áudio. Este projeto está em desenvolvimento
-do site do Estúdio Livre [estudio livre] com repositório no sourceforge [source force].
-
-As principais fontes sobre estre trabalho é a página de desenvolvimento da ideia
-que está em constante mudança ~\cite{http://estudiolivre.org/tiki-index.php?page=fdp&highlight=fdp fdpel}
-e, o artigo que foi aceito no LAC de 2008 ~\cite{http://www.estudiolivre.org/el-gallery_view.php?arquivoId=8221 fdplac2008}
-e o repositório ~\cite{http://sourceforge.net/projects/fdpack/develop fdpsf}.
 
@@
 
-\subsection{Tecnologias sociais de alta demanda}
+      \subsection{Tecnologias sociais de alta demanda: Sitios, Conteúdos e Articulação}
 
-\subsection{Sítios}
+      \subsubsection{Sítios}
 
@@
 
-\subsection{Conteúdos}
+      \subsubsection{Conteúdos}
 
@@
 
-\subsection{Articulação}
+      \subsubsection{Articulação}
 
@@
 
-\section{Disponibilização e Desenvolvimento}
+\subsection{Disponibilização e desenvolvimento conjunto: wikis, etherpads, AA, Trac, IRC ..}
 
-\subsection{Wiki}
+\subsubsection{Wiki}
 
-\subsection{Trac}
+\subsubsection{Trac}
 
-\subsection{Screencasts - Vimeo}
+\subsubsection{Screencasts - Vimeo}
 
-\subsection{AA}
+\subsubsection{AA}
 
-\subsection{Audio Experiments (Æ)}
+\subsubsection{Audio Experiments (Æ)}
 
-\subsection{IRC}
+\subsubsection{IRC}
 
-\subsection{Etherpads}
+\subsubsection{Etherpads}
 
-\subsection{Outras fontes}
+\subsubsection{Outras fontes}
+
+\section{Materiais didáticos}
+
+  \subsection{Tutoriais em texto e código: Filtros, Nyquist e plugins LADSPA}
+
+Os vários materiais didáticos produzidos constam no apêndice
+deste trabalho. Um único destes será exposto a seguir por sua
+capacidade de agregar os conteúdos dos capítulos anterioes:
+os tutoriais de filtros e amostragem.
+
+\begin{itemize}
+    \item {\bf Tutorial de python para áudio e som}
+
+Este tutorial foi levado para Berlim no LAC 2007 e sofreu melhoras desde entao. Esta
+primeira versao ficou resumida em forma de texto no EL\footnote{http://estudiolivre.org/python-e-som-tutorial}. Em 2010
+a Associacao Python Brasil escolheu este trabalho, então já mais amadurecido, para ser apresentado no
+FISL em Porto Alegre. Como consequencia, foi feita uma série de video-tutoriais bastante utilizados\footnote{http://estudiolivre.org/tiki-index.php?page=Video+Tutoriais}.
+Este tutorial foi comentado em listas em que o autor não participa (e outras em que o autor participa).
+
+    \item {\bf Tutoriais de filtros e amostragem via python}
+
+Voltados para explicitar principios fundamentais de áudio, estes tutoriais
+são baseados código Python e o equivalente em C. Pequenas explicações são
+dadas com o intuito de orientar a exploração inteligente destes \emph{snippets}.
+
+\emph{Teorema de Amostragem}: estes scripts visam a experimentacao inteligente com
+o Teorema de Nyquist. (descricao)
+
+\emph{Filtros}: alem da explicitação sobre as diferenças entre filtros FIR e IIR,
+duas utilizações clássicas destes filtros estão implementadas: Wavelets (FIR) e Quad (IIR)
+
+Estes códigos podem ser baixados no repositório SVN do AudioExperiments. E os textos estão
+na wiki (nos digitais ou EL, recriar pois os CDTL foram apagados)
+
+    \item {\bf Tutorial de plugins lv2}
+
+Dadas as dificuldades que o desenvolvimento dos \emph{plugins} de áudio apresenta,
+desenvolvi um tutorial passo a passo com plugins que rodam em todas as etapas.
+Ele é baseado em uma interface C++ para este padrão de plugin que eh implementado
+em C. Os códigos e os textos estão todos em repositório.
+
+    \item {\bf Microtutoriais Django ~\cite{dmicrotuts}}
+
+Estes 'microtutoriais' são baseados nos conceitos de \emph{scripts mínimos} e
+\emph{alterações puntuais}. O primeiro conjunto de microtutoriais é dedicado
+a reconstruir o tutorial oficial do django de forma condensada e não prolixa.
+O segundo destes conjuntos é dedicado a instrumentalizar de fato o leitor com
+o entendimento do funcionamento dos princípios fundamentais deste framework.
+
+    \item {\bf Philosometrics}
+
+Embora este não seja um trabalho didático propriamente dito, ele tem este intuito
+no cerne de sua concepção e surgimento. Em decorrência dele, surgiu o  Musimetrics,
+o Cinemetrics e o Literametrics. Além disso, ele é um belo exemplo da
+utilização das ciências duras para a análise de ciências humanas e foi acolhido
+como tal em alguns momentos.
+
+    \item {\bf Carta mídias livres}
+
+Texto criado em decorrência da participação da comissão de seleção no
+'Prêmio Mídias Livres', a convite do Ministério da Cultura por 'notório saber'.
+Esta carta é um documento único no seu conteúdo, deixando às claras
+o conceito de Mídias Livres como mídias não aprisionadas pelo conceito
+de propriedade, ou seja, que priorizam a sua livre circulação e a possibilidade
+de geração de materiais derivados, assim como sua geração aberta ao colaborativo e comunitário.
+
+    \item {\bf Textos de cunho sociológico, transformador}
+
+Produção mais numerosa que as anteriores, se caracteriza por métodos não convencionais
+de abordagem dos assuntos e de escrita. Em especial utiliza-se pseudônimos para
+auxiliar a despersonificação, gerando textos menos presos à satisfação da auto-imagem, dente
+outras qualidades. A utiliação de psudônimos é um costume muito apreciado em diversos meios,
+e as pesquisas tem confirmado as vantagens que a prática apresenta e confirma\footnote{http://disqus.com/research/pseudonyms/}.
+
+O autor destes textos se dá ao direito de não revelar seus psudônimos - embora muitos deles
+sejam publicamente conhecidos - para conservar as consequências desta prática na
+forma mais pura. Como comprovante desta produção, deixamos uma mensagem sobre a publicação
+de textos em mídia impressa com autores internacionais,
+confirmando a participação do autor desta dissertação, mas cujo
+nome não consta na publicação.
+
+!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+
+de      fabi borges catadores@gmail.com por  riseup.net 
+responder a     submidialogia@lists.riseup.net
+para    submidialogia@lists.riseup.net
+data    23 de agosto de 2011 12:26
+assunto Re: [submidialogia] livro sub- publicação
+lista de e-mails        <submidialogia.lists.riseup.net> Filtrar as mensagens dessa lista de e-mails
+enviado por     lists.riseup.net
+assinado por    riseup.net
+cancelar inscrição      Cancelar a inscrição para essa lista de e-mails
+        Importante principalmente porque você frequentemente lê mensagens com esse marcador.
+ocultar detalhes 12:26 (6 minutos atrás)
+entao, eu fui recebendo textos durante esse tempo,
+alguns tao atrazados como dos sem satelites, mas muita gente mandou;
+
+aqui os autores:
+
+os internacionais nao sao muitos, o joni kempf (o que bebe ouro do hardware), o barbrook (futuros imaginarios),
+Hamdy heda (da revolucao egipcia), o pedro soller (summerlab), maria llopis (pos porno),  talvez a bronac, nao entregou ainda.
+
+dos brasileiros, renato fabri, ruiz, pasteur, morgana e caio, ju dornelles, mari marcassa, coletivo errorista, adriana velozo-drica, tiago pimentel, felipe fonseca, thiago novaes, lelex, felipe ribeiro(?), maira, verenilde,  bartolina silva, poro, vitoria amaro, fabib (eu),
+
+entao, precisa publicar agora,
+uma equipe para publicacao e,,,
+
+bjs
+f
+
+!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+
+\end{itemize}
+
+\subsection{Screencasts e outros materiais em video e em texto}
+
+\begin{itemize}
+    \item Python para áudio e música
+	  Texto - palestras - videos
+
+    \item Canal Macambira
+No Macambira estão sendo produzidos materiais em screencasts sobre
+diversas cenas de hackeamento.
+
+    \begin{itemize}
+	\item Live-Coding
+	\item Raspagem de dados
+    \end{itemize}
+\end{itemize}

### 2012-02-20 tutoriais.tex
terminada reestruturacao da dissertacao

@@
-\chapter{Tutoriais e outros materiais didáticos} %Nome do capítulo.
+\chapter{Mais tutoriais e outros materiais didáticos} %Nome do capítulo.
 \label{cap:tutsDidads}
@@
 \subsection{Figusdevpack (FDP)}
+
+Um ambiente de interação da comunidade de Python e Música 
+para compartilhamento de códigos e excertos. Baseado principalmente
+em documentação organizada sobre as práticas e as bibliotecas
+existentes para python. Assim como scripts compartilhados que
+fazem uso de objetos e módulos específicos. Este trabalho foi aceito na
+maior conferência de áudio em linux, a Linux Audio Conference de 2008
+(LAC2008) e está sendo reativado por mim em conjunto com Vilson Vieira
+e outros desenvolvedores de áudio. Este projeto está em desenvolvimento
+do site do Estúdio Livre [estudio livre] com repositório no sourceforge [source force].
+
+As principais fontes sobre estre trabalho é a página de desenvolvimento da ideia
+que está em constante mudança ~\cite{http://estudiolivre.org/tiki-index.php?page=fdp&highlight=fdp fdpel}
+e, o artigo que foi aceito no LAC de 2008 ~\cite{http://www.estudiolivre.org/el-gallery_view.php?arquivoId=8221 fdplac2008}
+e o repositório ~\cite{http://sourceforge.net/projects/fdpack/develop fdpsf}.

### 2012-02-24 devEresultados.tex
resolvida a questao do codigo python na dissertacao

@@
 escalas completamente simetricas assim:
+% \begin{python}
 
-escala\_1=range(12) # cromática ascendente sem inclusão da oitava
+\code{}{ex.py}
+
+asd}{escala\_1=range(12) # cromática ascendente sem inclusão da oitava
 escala\_2=range(0,12,2) # tons inteiros
@@
 escala\_4=range(0,12,4) # terças maiores
-escala\_6=range(0,12,6) # trítonos
+escala\_6=range(0,12,6) # trítonos}
+% \end{python}
+% \input{py}
 

### 2012-02-24 devEresultados.tex
questao do codigo python jah encaminhado na dissertacao

--- 2012-02-20 devEresultados.tex
@@
 
-escala\_1=range(12) # cromática ascendente sem inclusão da oitava
-escala\_2=range(0,12,2) # tons inteiros
-escala\_3=range(0,12,3) # diminutão
-escala\_4=range(0,12,4) # terças maiores
-escala\_6=range(0,12,6) # trítonos
+\code{Escalas Simétricas na grade dos 12 semintos e no âmbito da oitava}{python_snippets/escalas_simetricas.py}
+
+% escala\_1=range(12) # cromática ascendente sem inclusão da oitava
+% escala\_2=range(0,12,2) # tons inteiros
+% escala\_3=range(0,12,3) # diminutão
+% escala\_4=range(0,12,4) # terças maiores
+% escala\_6=range(0,12,6) # trítonos
+% \end{python}
+% \input{py}
 
 Sendo cada unidade um semitom. Assim as triades maiores e menores
-são especificadas com simplicidade:
+são especificadas com simplicidade. Note que aos acordes \emph{diminutão} e
+\emph{aumentado} correspodem as mesmas notas das escalas simetricas de terças menores
+e maiores respectivamente:
 
-maior=[0,4,7]
-menor=[0,3,7]
+\code{Acordes anotados como listas em python}{python_snippets/acordes.py}
 
@@
 
-b1=[0]*4
-b2=[0]*7+[7]
-b3=  [0,7,0,1]
-b3\_2=[0,7,8,7]
-b4=[0,7,7,7]
-b5=[0,7]
-b6=[0,19,7,12]
-b7=[0,-1,0,-1]
-b8=[0,-1,7,-1]
-b9=[0,1,0,-1]
-b10=[0,7,6,7]
+\code{Series diversas}{python_snippets/series.py}
 
@@
 
-fazSequencia(nt, f, fator, d, s, senoide, dente)
-fazAcorde([0,7,14,21,28,35,42,49,56,63,70,77,84]
+\code{Sintetizando sequências e acordes}{python_snippets/seqs_acordes.py}
 
@@
 
-seq=[1.5*i for i in fazSequencia(nt, f, fator, d, s, senoide, dente)] # amplificacao
-
-seq3=[(i+j) for i,j in zip(seq1,seq2)] # mixagem
+\code{Amplificação e mixagem}{python_snippets/amp_mix.py}
 
@@
 
-ac=[]
-for i in xrange(8): # a cada 2 tempos
-    ac+=fazAcorde([0,7,14,21,28,35,42,49,56,63,70,77,84], f, fator, d, s, senoide, dente)
-    ac+=[0]*int((d+s)*44100)
-
-E por fim, mixando quando os comprimentos nao sao os mesmos:
-
-seq2=somador(ac,seq2)
+\code{Acordes periódicos}{python_snippets/acordes_periodicos.py}
 

### 2012-02-24 python_snippets/escalas_simetricas.py

@@ -1,0 +1,5 @@
+escala_1=range(12) # cromatica ascendente sem inclusao da oitava
+escala_2=range(0,12,2) # tons inteiros
+escala_3=range(0,12,3) # tercas menores
+escala_4=range(0,12,4) # tercas maiores
+escala_6=range(0,12,6) # tritonos

### 2012-02-24 python_snippets/seqs_acordes.py

@@ -1,0 +1,2 @@
+fazSequencia(nt, f, fator, d, s, senoide, dente)
+fazAcorde([0,7,14,21,28,35,42,49,56,63,70,77,84],f, fator, d, s, senoide, dente)

### 2012-02-24 python_snippets/series.py

@@ -1,0 +1,11 @@
+b1=[0]*4
+b2=[0]*7+[7]
+b3=  [0,7,0,1]
+b3_2=[0,7,8,7]
+b4=[0,7,7,7]
+b5=[0,7]
+b6=[0,19,7,12]
+b7=[0,-1,0,-1]
+b8=[0,-1,7,-1]
+b9=[0,1,0,-1]
+b10=[0,7,6,7]

### 2012-02-28 devEresultados.tex
feito do minimum-fi, explicitando ele como base

@@
 
-  \subsection{Música em Tempo Real: Livecoding e ABeatTracker (ABT)}
-
-  [Apresentar Livecoding.]
-
-  Apenas parte dos conhecedores do assunto reconhecem
-  o ABT como livecoding. Alguns
-  consideram a linguagem por macros mais especializadas e poderosas como
-  sendo um tipo de interface escrita, nao uma linguagem de programação
-  propriamente dita.
-
-      \subsubsection{ABeatTracker}
-
-  Manual ABT
-
-  Linkar com ABD
-
-      \subsubsection{Audioexperiments, Estudiolivre e Foobarbaz}
-
-      Audioexperiments. Estúdio livre. LM
-
   \subsection{Musica em Tempo Diferido: Minimum-fi e FIGGUS}
@@
 para se conseguir expressar ideias musicais em geral. Esta empreitada
-resultou no minimum-fi, codigo em um unico arquivo, que sintetiza
-musicas inteiras. O os algoritmos propriamente ditos somam menos de
+resultou no minimum-fi, código em um único arquivo, que sintetiza
+músicas inteiras. O os algoritmos propriamente ditos somam menos de
 50 linhas e nao mais que 5 funções.
@@
 permita varios timbres.
-  \item Deve-se ser capaz de construir series de notas e acordes.
+  \item Deve-se ser capaz de construir unidades e series de unidades (notas e acordes, por exemplo).
 \end{itemize}
 
-Para o primeiro item, se prestam as funções lookup e lookupcruz. A função
+Para o primeiro item, se prestam comumente os procedimentos de busca em tabelas com formas
+de ondas em alta resolução. O procedimento é barato e, como vimos nos fundamentos,
+não acrescentam ruídos relevantes ao sinal.
+
+O segundo item presta-se à discretização do espaço musical. Unidades como batidas e notas
+tornam mais eficiente a comunicação pois a quantidade
+de estruturas sugeridas é maior~\cite{Roederer} no discreto doq eu no contínuo. Roederer
+aponta que as próprias notas dos instrumentos musicais são um reflexo de que é mais eficiente
+o uso do discreto do que do contínuo.
+
+Desta forma, se mostram úteis para fazer sequencias de unidades, concatená-las. Quando as unidades
+são notas, as sequencias sustapostas no tempo de unidades são melodias ou linhas melódicas. As
+sequencias sobrepostas no tempo são comumente pensados como acordes, mas podem ser tidos simplesmente
+como sobreposições circunstanciais de duas linhas melódicas.
+
+As duas construções básicas escolhidas, baseadas na dicotomia melodia/harmonia, são
+as funções fazSequencia e fazAcorde. Vale notar elas são absolutamente equivalentes e dispensáveis
+em uma análise puramente conceitual. Os procedimentos de mixagem e concatenação são
+plenamente capazes de realizar o que estas funções realizam. Aliás, as funções nada
+mais são do que usos típicos e quase caricatos destes procedimentos: no fazAcorde a mixagem
+sobrepõe no tempo todas as unidades, no fazSequencia as unidades são todas juntapostas no tempo.
+
+ A função
 lookup já foi apresentada na sessão anterior, quando tratado de Python já visando o áudio.
-O lookupcruz ex explicado a seguir.
+O lookupcruz é explicado a seguir.
 
@@
 python RUNME
+
+  \subsection{Música em Tempo Real: Livecoding e ABeatTracker (ABT)}
+
+  [Apresentar Livecoding.]
+
+  Apenas parte dos conhecedores do assunto reconhecem
+  o ABT como livecoding. Alguns
+  consideram a linguagem por macros mais especializadas e poderosas como
+  sendo um tipo de interface escrita, nao uma linguagem de programação
+  propriamente dita.
+
+      \subsubsection{ABeatTracker}
+
+  Manual ABT
+
+  Linkar com ABD
+
+      \subsubsection{Audioexperiments, Estudiolivre e Foobarbaz}
+
+      Audioexperiments. Estúdio livre. LM
 

### 2012-03-04 devEresultados.tex
adicionado sobre tempo diferido e intro ao minimum-fi e FIGGUS

@@
   \subsection{Musica em Tempo Diferido: Minimum-fi e FIGGUS}
+
+\begin{quotation}
+\small
+'The increasing dominance of graphic interfaces for music software obscured 
+the continuing presence of the command-line tradition, 
+the code writer, the hacker. The code writing of deferred time 
+computer programming may be assembled out of time order, debugged and optimized.'
+
+\emph{Simon Emmerson, Living electronic music, 2007}
+\end{quotation}
+
+A realização musical em tempo diferido é o paradigma inicial da música computacional.
+Iniciando com o Music V, a proposta foi depois desenvolvido com o CSound. Pode-se dizer
+que até hoje é a forma como compositores usualmente pensam a música: pensando
+e escrevendo as estruturas, que depois são executadas por instrumentistas.
+
+Assim como a composição instrumental permite um trabalho mais minucioso do que
+a improvisação instrumental, a realização musical em tempo diferido permite um
+detalhamento maior dos procedimentos do que a realização em tempo real. Por este
+mesmo motivo, trataremos inicialmente de dois trabalhos em tempo diferido,
+aliás paradigmaticos.
+
+O primeiro, chamado de \emph{minimum-fi}\footnote{hi-fi, low-fi, minimum-fi, ou seja,
+o mínimo de qualidade para assegurar existência e consistência
+da criação} sintetiza uma música inteira, amostra por amostra, através de princípios
+claros de síntese sonora e organização musical. O segundo, chamado de \emph{FIGGUS} (FInite
+Groups in Granular Synthesis), utiliza os princípios do minimum-fi e constitui
+um módulo Python completo, cuja proposta é a utilização de simetrias, através de
+permutações e de Teoria de Grupos, para a composição de músicas. Como demonstração
+das capacidades do FIGGUS, ele gera um EP inteiro com um único comando. Este é o
+\emph{PPEPPS}\footnote{Pure Python EP: Projeto Solvente}, como veremos a seguir.
 

### 2012-03-05 devEresultados.tex
mudancas relacionadas ao minimum-fi

@@
 Assim como a composição instrumental permite um trabalho mais minucioso do que
-a improvisação instrumental, a realização musical em tempo diferido permite um
+a improvisação instrumental, a realização musical em tempo diferido usualmente permite um
 detalhamento maior dos procedimentos do que a realização em tempo real. Por este
@@
 claros de síntese sonora e organização musical. O segundo, chamado de \emph{FIGGUS} (FInite
-Groups in Granular Synthesis), utiliza os princípios do minimum-fi e constitui
+Groups in Granular and Unit Synthesis), utiliza os princípios do minimum-fi e constitui
 um módulo Python completo, cuja proposta é a utilização de simetrias, através de
@@
 
+Desta forma, buscaremos expor nesta sessão, atraves de dois exemplos reais,
+os princípios do uso de código para a síntese \emph{musical}, desde as amostras
+relativas a uma nota com dada frequencia, amplitude e timbre, até a confecção
+de uma ferramenta derivada, já incorporando propostas musicais e estruturas
+mais elaboradas.
+
       \subsubsection{Minimum-fi}
 
-Em Abril de 2011 foi trabalhado um codigo dedicado a materializar o minimo
+Em Abril de 2011 surgiu a necessidade estética e também tecnológica
+de realizar um código dedicado a materializar \emph{o mínimo}
 para se conseguir expressar ideias musicais em geral. Esta empreitada
 resultou no minimum-fi, código em um único arquivo, que sintetiza
-músicas inteiras. O os algoritmos propriamente ditos somam menos de
-50 linhas e nao mais que 5 funções.
+músicas inteiras. Os algorítmos propriamente ditos somam 
+53 linhas em 5 funções. O resto são estruturas musicais utilizadas neles.
 
@@
 \begin{itemize}
-  \item Deve-se ter um mecanismo de sintese simples mas que
-permita varios timbres.
-  \item Deve-se ser capaz de construir unidades e series de unidades (notas e acordes, por exemplo).
+  \item Deve-se ter um mecanismo de síntese que
+permita varios timbres e frequência fundamental.
+  \item Deve-se ser capaz de construir unidades e séries de unidades (notas e acordes, por exemplo).
 \end{itemize}
@@
 Para o primeiro item, se prestam comumente os procedimentos de busca em tabelas com formas
-de ondas em alta resolução. O procedimento é barato e, como vimos nos fundamentos,
-não acrescentam ruídos relevantes ao sinal.
+de ondas em alta resolução (chamado \emph{lookup table}). O procedimento é barato e de qualidade alta
+(não acrescentam ruídos relevantes ao sinal).
+
+\code{Procedimento de Lookup Table}{python_snippets/lookuptable.py}
 

### 2012-03-05 devEresultados.tex
andado nos procedimentos de lookup table e as tabelas mais usuais

@@
 que até hoje é a forma como compositores usualmente pensam a música: pensando
-e escrevendo as estruturas, que depois são executadas por instrumentistas.
+e escrevendo as estruturas, que depois são executadas por instrumentistas ou aparelhos eletronicos.
 
@@
 
-Desta forma, buscaremos expor nesta sessão, atraves de dois exemplos reais,
-os princípios do uso de código para a síntese \emph{musical}, desde as amostras
+Desta forma, esta sessão exemplifica e explicita - através de dois exemplos reais - os princípios do uso de código para a síntese \emph{musical}, desde as amostras
 relativas a uma nota com dada frequencia, amplitude e timbre, até a confecção
@@
 músicas inteiras. Os algorítmos propriamente ditos somam 
-53 linhas em 5 funções. O resto são estruturas musicais utilizadas neles.
+53 linhas em 5 funções. Com estas funções, estruturas musicais podem ser criadas amostra por amostra, resultando em notas, blocos e estruturas hierarquicamente superiores.
 
@@
   \item Deve-se ter um mecanismo de síntese que
-permita varios timbres e frequência fundamental.
+permita vários timbres e controle sobre a frequência fundamental.
   \item Deve-se ser capaz de construir unidades e séries de unidades (notas e acordes, por exemplo).
@@
 de ondas em alta resolução (chamado \emph{lookup table}). O procedimento é barato e de qualidade alta
-(não acrescentam ruídos relevantes ao sinal).
+(não acrescentam ruídos relevantes ao sinal). A seguir está o procedimento da lookuptable:
 
 \code{Procedimento de Lookup Table}{python_snippets/lookuptable.py}
+
+O primeiro ponto importante é a tabela em si. É de comum conhecimento que
+tabelas com 1024 são mais que suficientes para os usos musicais e a diferença
+de qualidade (relação sinal/ruido) não muda consideravelmente com a utilização
+de tabelas maiores. A seguir dispomos algumas destas tabelas, mais especificamente, dispomos as tabelas para os formatos de onda mais usuais.
+
+\code{Formas de Onda Tradicionais}{python_snippets/formasDeOnda.py}
 

### 2012-03-05 python_snippets/formasDeOnda.py

@@ -1,0 +1,5 @@
+T=2048
+senoide=np.sin(np.linspace(0,2*np.pi,T,endpoint=False))
+dente=np.linspace(-1,1,T,endpoint=False)
+triangular=np.hstack((dente[::2],dente[-1:0:-2]))
+quadrada=np.hstack((np.ones(T/2),np.ones(T/2)*-1))

### 2012-03-05 python_snippets/lookuptable.py

@@ -1,7 +1,13 @@
-T=len(table)
-SI= freq * T / 44100
+# fixados a frequencia (Hz), a duracao (segundos) e a tabela (1 ciclo da onda)
+# e a taxa de amostragem (44,1kHz, 48kHz, 8kHz, etc)
+# exemplo:
+# freq = 200
+# dur = .2 
+# tabela = '1 ciclo de senoide ou outra forma de onda em 1024 amostras' 
+T=len(tabela)
+incremento=freq*T/taxa_amostragem
 ap=0
-samples=[]
-for i in xrange(int(dur*44100)):
-    samples.append(table[int(ap)])
-    ap = (SI + ap)
+amostras=[]
+for i in xrange(int(dur*taxa_amostragem)):
+    amostras.append(tabela[int(ap)])
+    ap=incremento+ap

### 2012-06-28 devEresultados.tex
retomada e repassada ateh o ABT nao incluso

@@
 
-exp abertos
+Os resultados explicitados nesta dissertação estão focados no processamento de 
+áudio voltado para a música e na síntese de estruturas musicais.
 
-Do produção de códigos em tempo real para música, chamado livecoding,
-vamos para ferramentas computacionais para áudio e música, com o minimum-fi
-e FIGGUS. Já fazendo ponte do código com outras interfaces, adentramos AHT e EKP,
-que interfaceiam com o mundo físico e material (AHT) e com o hardware, onde
-nosso entendimento em codigo é efetuado em procedimentos.
+Por razões didáticas, dividimos este capítulo em 4 partes, a saber: experimentos
+abertos em áudio e música (incluindo plugins LADSPA, Wavelets, Redes Complexas e outros
+scripts experimentais), música em tempo diferido (ou seja, que não é feito em tempo real),
+música em tempo real, e música na matéria (suporte físico em hardware).
+
+Com isso, desejamos percorrer minimamente a prática musical através do código.
+com exemplos reais de aplicação e em uso pelo autor, por membros do LabMacambira.sf.net,
+por parceiros, colaboradores e por usuários eventuais das naturezas mais diversas.
 
@@
 que até hoje é a forma como compositores usualmente pensam a música: pensando
-e escrevendo as estruturas, que depois são executadas por instrumentistas ou aparelhos eletronicos.
+e escrevendo as estruturas, que depois são executadas por instrumentistas ou aparelhos eletrônicos.
 
@@
 detalhamento maior dos procedimentos do que a realização em tempo real. Por este
-mesmo motivo, trataremos inicialmente de dois trabalhos em tempo diferido,
-aliás paradigmaticos.
+mesmo motivo, trataremos inicialmente de dois trabalhos em tempo diferido.
+Aliás, como veremos a seguir, as abordagens são paradigmáticas para a música
+computacional praticada até os dias de hoje.
 
-O primeiro, chamado de \emph{minimum-fi}\footnote{hi-fi, low-fi, minimum-fi, ou seja,
+O primeiro destes feitos, chamado de \emph{minimum-fi}\footnote{hi-fi, low-fi, minimum-fi, ou seja,
 o mínimo de qualidade para assegurar existência e consistência
@@
 permutações e de Teoria de Grupos, para a composição de músicas. Como demonstração
-das capacidades do FIGGUS, ele gera um EP inteiro com um único comando. Este é o
+das capacidades do FIGGUS, ele gera um EP (E... Play, pegar significado dele) inteiro com um único comando. Este é o
 \emph{PPEPPS}\footnote{Pure Python EP: Projeto Solvente}, como veremos a seguir.
 
-Desta forma, esta sessão exemplifica e explicita - através de dois exemplos reais - os princípios do uso de código para a síntese \emph{musical}, desde as amostras
+Desta forma, esta sessão exemplifica e explicita - através de dois exemplos reais - os
+princípios do uso de código para a síntese \emph{musical}, desde as amostras
 relativas a uma nota com dada frequencia, amplitude e timbre, até a confecção
@@
 de realizar um código dedicado a materializar \emph{o mínimo}
-para se conseguir expressar ideias musicais em geral. Esta empreitada
+para expressar ideias musicais em geral. Esta empreitada
 resultou no minimum-fi, código em um único arquivo, que sintetiza
-músicas inteiras. Os algorítmos propriamente ditos somam 
-53 linhas em 5 funções. Com estas funções, estruturas musicais podem ser criadas amostra por amostra, resultando em notas, blocos e estruturas hierarquicamente superiores.
+músicas inteiras. Os algorítmos propriamente ditos, escritos em Python, somam 
+53 linhas e inclui 5 funções. Com estas funções, estruturas musicais podem 
+ser criadas amostra por amostra, resultando em notas, blocos e estruturas hierarquicamente superiores.
 
-Os princípios são:
+Os princípios, bastante simples, são:
 \begin{itemize}
-  \item Deve-se ter um mecanismo de síntese que
+  \item Deve-se ter um mecanismo de síntese sonora que
 permita vários timbres e controle sobre a frequência fundamental.
@@
 
-Para o primeiro item, se prestam comumente os procedimentos de busca em tabelas com formas
+Para o primeiro item, se prestam comumente os procedimentos de busca em tabelas/vetores com formas
 de ondas em alta resolução (chamado \emph{lookup table}). O procedimento é barato e de qualidade alta
@@
 
-O primeiro ponto importante é a tabela em si. É de comum conhecimento que
+O primeiro ponto importante é a tabela em si, usualmente unidimensional. É de comum conhecimento que
 tabelas com 1024 são mais que suficientes para os usos musicais e a diferença
 de qualidade (relação sinal/ruido) não muda consideravelmente com a utilização
-de tabelas maiores. A seguir dispomos algumas destas tabelas, mais especificamente, dispomos as tabelas para os formatos de onda mais usuais.
+de tabelas maiores. A seguir dispomos algumas destas tabelas, mais 
+especificamente, dispomos as tabelas para os formatos de onda mais usuais.
 
@@
 
+Através da utilização do lookup sucessivo (procura-se um valor na primeira tabela
+e este valor indica o valor na segunda tabela a ser utilizado), executamos
+um waveshaping. Este procedimento é bastante apreciado pela simplicidade e eficácia
+na síntese de tímbres diversos e ricos em harmônicos e evolução temporal.
+
+[codigo do lookupcruz e explicacao]
+
 O segundo item presta-se à discretização do espaço musical. Unidades como batidas e notas
 tornam mais eficiente a comunicação pois a quantidade
-de estruturas sugeridas é maior~\cite{Roederer} no discreto doq eu no contínuo. Roederer
+de estruturas sugeridas é maior e as estruturas são mais claras~\cite{Roederer} no discreto do que no contínuo. Roederer
 aponta que as próprias notas dos instrumentos musicais são um reflexo de que é mais eficiente
@@
 
-Desta forma, se mostram úteis para fazer sequencias de unidades, concatená-las. Quando as unidades
-são notas, as sequencias sustapostas no tempo de unidades são melodias ou linhas melódicas. As
+Desta forma, unidades bem definidas se mostram úteis na prática musical 
+para fazer sequencias de unidades, concatená-las. Quando as unidades
+são notas, as sequencias de unidades justapostas no tempo são melodias ou linhas melódicas. As
 sequencias sobrepostas no tempo são comumente pensados como acordes, mas podem ser tidos simplesmente
-como sobreposições circunstanciais de duas linhas melódicas.
+como sobreposições circunstanciais de duas linhas melódicas. Isso, claro, segundo
+a sistematização clássica e usual da música~\cite{Lacerda}.
 
-As duas construções básicas escolhidas, baseadas na dicotomia melodia/harmonia, são
-as funções fazSequencia e fazAcorde. Vale notar elas são absolutamente equivalentes e dispensáveis
-em uma análise puramente conceitual. Os procedimentos de mixagem e concatenação são
+As duas construções básicas explicitadas, baseadas na dicotomia melodia/harmonia
+(horizontalidade/verticalidade, justaposição/sobreposição), são
+as funções \emph{fazSequencia} e \emph{fazAcorde}. Vale notar elas são absolutamente 
+equivalentes em uma análise puramente conceitual, i.e. uma delas pode ser omitida segundo
+a teoria. Isso fica particularmente óbvio quando se nota que os procedimentos de mixagem e concatenação são
 plenamente capazes de realizar o que estas funções realizam. Aliás, as funções nada
@@
 
- A função
-lookup já foi apresentada na sessão anterior, quando tratado de Python já visando o áudio.
-O lookupcruz é explicado a seguir.
-
-[codigo do lookupcruz e explicacao]
-
-As sequencias de notas e os acordes foram resolvidos com 2 outras funcoes simples,
-que utilizam lookup e lookupcruz.
+Como pode notar a seguir, as sequencias de notas e os acordes, em última instância, são utilizações específicas
+das 2 funcoes de síntese sonora explicadas anteriormente: lookup e lookupcruz.
 
@@
 
-A função somador eh para automatizar a soma amostra a amostra e completar com zeros a
-sequencia menor.
+A última das cinco funções utilizadas é uma soma amostra a amostra de dois sons. Para isso,
+é necessário completar com zeros a sequencia com o menor número de amostras:
 
@@
 
-Sendo cada unidade um semitom. Assim as triades maiores e menores
+Sendo cada unidade um semitom. Claro que podemos também utilizar as usuais
+escalas tonais maior e menor natural, harmônica e melódica:
+
+[codigo das escalas maior e menores]
+
+A utilização de intervalos menores que o semitom\footnote{Prática também conhecida
+como \emph{microtonalidade}} mostra-se trivial em
+nossa implementação:
+
+[escalas com intervalos microtonais]
+
+Da mesma forma, as triades maiores e menores
 são especificadas com simplicidade. Note que aos acordes \emph{diminutão} e
@@
 
-E series podem ser anotadas junto a variacoes:
+E séries podem ser anotadas junto a variacoes:
 
-\code{Series diversas}{python_snippets/series.py}
+\code{Séries diversas}{python_snippets/series.py}
 
-Agora eh sintetizar e mixar. A sintese de sequencias e acordes sao feitos tipicamente
-atraves de comandos desta forma:
+Com este arcabouço, o passo seguinte é sintetizar e mixar, resultando
+em sequências musicais. A síntese de sequências e acordes são feitos tipicamente
+através de comandos desta forma:
 
 \code{Sintetizando sequências e acordes}{python_snippets/seqs_acordes.py}
+\code{Acordes periódicos}{python_snippets/acordes_periodicos.py}
 
-Aqui uns procedimentos basicos utilizados nas montagens:
+Em posse destas sequências e acordes, formamos estruturas hierarquicamente superiores
+através da concatenação de estruturas, da mixagem de estruturas, e da amplificação
+(ou atenuação) seletiva das mesmas:
 
 \code{Amplificação e mixagem}{python_snippets/amp_mix.py}
-
-Gerando uma serie de acordes:
-
-\code{Acordes periódicos}{python_snippets/acordes_periodicos.py}
 
@@
 Desenvolvimento iniciado em 2006 com o físico-matemático Prof. Adolfo Maia Junior para
-tratar de simetrias na música, com vistas à composição musical atraves
+tratar de simetrias na música, com vistas à composição musical através
 de métodos matemáticos. Mais especificamente, a proposta resultou em
@@
 granular e síntese de estruturas musicais. O nome dado
-foi FIGGUS, sigla de FInite Groups in Granular and Unit Synthesis.
+foi FIGGUS, sigla de FInite Groups in Granular and Unit Synthesis\footnote{Também utilizamos
+o nome FIGGS (FInite Group in Granular Synthesis) dado que o termo \emph{unit synthesis} não
+é usual na literatura. Posteriormente o principal autor deste trabalho recorreu novamente
+ao FIGGUS pois o maior uso da abordagem é para síntese de estruturas musicais, não para
+síntese de amálgamas sonoros tipicamente resultantes da síntese granular}.
 
@@
 
-Na atual reescrita, embora ainda bastante atras do FIGGUS original,
-mesmo sem contar a interface grafica, o codigo roda inteiro em python
+Na atual reescrita, embora ainda bastante atrás do FIGGUS original,
+e mesmo sem a interface gráfica, o código roda inteiro em python
 com as biblitecas imbutidas por padrão. Isso permite com que o FIGGUS
@@
 
+[Explicar o conceito de simetria e permutações]
+[Expor sobre uma ou duas músicas feitas com o FIGGUS]
+
   \subsection{Música em Tempo Real: Livecoding e ABeatTracker (ABT)}
 
-  [Apresentar Livecoding.]
+Com os avanços computacionais recentes, tornou-se usual a síntese sonora
+em tempo real. Com isso, surgiram liguagens dedicadas para o áudio e a música
+em sua maioria dedicadas - ou ao menos capacitadas - para o uso 
+em tempo de execussão [citar PD, SC, ChucK]. Em outras palavras, estas linguagens
+possibilitam que o usuário ouça o resultante sonoro do código utilizado e altere
+o código com resultado imediato no processamento e resultado que escuta.
+
+Nesta linha de exploração musical do código, iremos expor a seguir sobre
+nossas investidas em \emph{Livecoding} (escrita de código em tempo real
+com vistas à performance pública) e o ABeatTracker (uma linguagem por macros
+para execussão sonora rítmica em conjunto com instrumentos tradicionais e outras
+fontes sonoras/musicais externas).
+
+        \subsubsection{Livecoding}
+
+Recentemente, grupos de ponta em música experimental no mundo todo estão
+desenvolvendo apresentações musicais públicas baseadas na escrita
+de código ao vivo. Usualmente, se projeta o código para que a audiência possa
+ver o que está sendo escrito, no rítmo em que se escreve, e se projeta também
+o resultante sonoro por autofalantes.
+
+As motivações para isso são variadas. Expomos a seguir de forma topificada
+um condensado do que nos orientou a tal prática. Vale ressaltar que são
+motivações igualmente presentes em outros grupos, embora não necessariamente
+em todos ou da mesmícima forma.
+
+\begin{itemize}
+    \item A performance musical por computador carece de recursos performáticos
+    à altura das execussões com instrumentos tradicionais. Os gestos são por demais
+    discretos e a concentração do performer é bastante focada na tela do computador.
+    \item O feedback auditivo do código projetado permite que o espectador infira
+    significados dos códigos. Mesmo que de forma superficial, este recurso do livecoding
+    é em muito capaz de desmistificar a programação de computadores, por muito considerada
+    completamente intangível.;
+    \item O código em si é um recurso poderosíssimo, que permite ao usuário controlar
+    os sons produzidos amostra por amostra ou em escalas maiores, como notas, compassos, fraseados
+    inteiros ou mesmo em escalas maiores de tempo, como minutos, horas, dias e semanas.
+    \item O compartilhamento do código é usual, leve e eficiente como entrega completamente
+    aberta de tecnologia e proposta estética.
+\end{itemize}
+
+Desta forma, iniciamos em 2011 uma linha de atuação com Livecoding que resultou em uma performance
+no \emph{V Festival Contato}. Utilizamos a linguagem ChucK por apresentar os recursos que
+consideramos mais apropriados, embora de forma alguma isso seja consensual na prática atual
+de livecoding.
+
+A apresentação contou com recursos adicionais para agregar interesse, como a utilização
+do \emph{cowsay} (para enviar mensagens enquanto se desenrolava a música) e trajetórias
+de um ponto no fundo do código para sugerir a vertigem do sono REM. Estes recursos
+podem ser vistos em uso claramente nos videos demonstrativos [citar videos do vimeo].
+As linhas do cowsay e o script em processing relacionados estão no Apêndice XX...
+
+Duas pessoas executaram livecoding simultaneamente. A saber, Vilson Vieira executava
+rítmos, batidas bastante marcadas que serviam como base. Renato Fabbri, autor do presente
+trabalho, executava linhas fluídas quasi melódicas que formavam arcos maiores. Nos intervalos
+das investidas no formato citado, havia interlúdios em que ambos se revesavam com músicas
+curtas e inusitadas, como em um duelo[disponibilizar músicas em links e colocar de referencia].
+
+[Colocar os códigos meus e do Vilson e explicar o funcionamento e uso]
 

### 2012-06-28 resumoeabstract.tex
questao do codigo python jah encaminhado na dissertacao
retomada e repassada ateh o ABT nao incluso

@@
 
-Nesta dissertacao, expomos os fundamentos do manuseio do som digitalizado através de recursos em código aberto.
+Nesta dissertação, expomos os fundamentos do manuseio do som digitalizado através de recursos em código aberto como
+uma forma de manifestação cultural.
 Após uma brevíssima contextualização sobre o status que o código e o áudio digital possuem na cultura atual, procuramos abordar como um todo o fenômeno
@@
 Os resultados antingidos com os fundamentos expostos são então sistematizados com ênfase nas ferramentas desenvolvidas.
-Apresentamos também uma série de repercussões destes desenvolvimentos, em especial uma frente de ação criada: o Lab Macambira.
+Apresentamos também uma série de repercussões destes desenvolvimentos, em especial uma frente de ação criada: o LabMacambira.sf.net.
 Por fim, como um produto adicional relacionado a esta dissertação, é delineado o mapa de exposição e comunicação no qual
@@
 de forma continuada e sistematizada através de webpages, wikis, repositórios, um canal IRC, aplicativos de acompanhamento de projetos
-e uma das tecnologias desenvolvidas neste trabalho: a Autogestão Algorítmica.
+e uma das tecnologias desenvolvidas neste trabalho: a Autoregulação Algorítmica.
 
 $\phantom{linha em branco}$\\
-Palavras-chave: acústica, áudio, código aberto, cultura digital, cultura hacker, programação, som
+Palavras-chave: som, acústica, áudio, código aberto, cultura digital, cultura hacker, programação
 
@@
 de forma continuada e sistematizada através de webpages, wikis, repositórios, um canal IRC, aplicativos de acompanhamento de projetos
-e uma das tecnologias desenvolvidas neste trabalho: a Autogestão Algorítmica.
+e uma das tecnologias desenvolvidas neste trabalho: a Autoregulação Algorítmica.
 

### 2012-06-28 devEresultados.tex
terminada a repassada na parte musical, agora eh preencher melhor as partes

@@
 
-  Apenas parte dos conhecedores do assunto reconhecem
-  o ABT como livecoding. Alguns
-  consideram a linguagem por macros mais especializadas e poderosas como
-  sendo um tipo de interface escrita, nao uma linguagem de programação
-  propriamente dita.
+      \subsubsection{ABeatTracker (ABT)}
+O ABT é uma linguagem/ferramenta que dispara linhas rítmicas através de macros que especificam
+as células rítmicas, amostras sonoras que são utilizadas como conteúdo sonoro destas
+linhas, modos de leitura destas amostras sonoras, e variáveis randômicas utilizadas
+para execussão da linha. Além disso, o ABT dispõe de variáveis globais que podem ser alteradas
+a qualquer momento pelo usuário, como BPM, velocidade de leitura das amostras e variáveis
+randômicas globais (que se somam às individuais).
 
-      \subsubsection{ABeatTracker}
+Em pouco explicitaremos a forma de utilização do ABT. De antemão, é pertinente apresentar
+uma pequena discussão a respeito do que o ABT é considerado e sobre os propósitos desta
+ferramenta. Em primeiro lugar, vale manter em mente que o ABT tem um funcionamento específico
+que pode transcender o que se conhece por livecoding. As macros pré-estabelecidas engendram um
+conjunto de recursos pré-estabelecidos bem definido, o que contrasta com a ideia de uma 'linguagem
+de programação' que tenha capacidades mais amplas. De qualquer forma, linguagens com domínios
+específicos não são raras e por vezes o ABT foi descrito como uma linguagem.
 
-  Manual ABT
+Sobre os propósitos do ABT, em primeiro lugar ele se dispõe a ser um instrumento computacional
+essencialmente rítmico. Junto a esta proposta, vem a necessidade da utilização em conjunto com
+outros instrumentos, externos ao ABT, ao computador em que estiver rodando e possivelmente externo
+com relação a qualquer computador. Para isso foi elaborado o ABD (ABeatDetector), no qual o usuário
+tamborila os rítmos que estiver ouvindo ou imaginando para que o ABT sincronize o pulso e utilize
+células rítmicas relacionadas. A análise feita pelo ABD resulta em uma série de rítmos explicitados
+por sequências de compassos que encapsulem durações regulares do rítmo tamborilado. Estes rítmos
+relacionados ao tamborilar do usuário são chamados de harmônicos e podem ser selecionados
+prontamente para o disparo de linhas melódicas.
 
-  Linkar com ABD
+No Apêndice XX está o manual de utilização do ABT e todo o código do ABT (e do ABD) está
+disponivel online (citar repos). Exploramos abaixo detalhes relevantes da implementação,
+em especial a forma de funcionamento do ABD.
 
-      \subsubsection{Audioexperiments, Estudiolivre e Foobarbaz}
-
-      Audioexperiments. Estúdio livre. LM
+[códigos-chave e explicações sobre o ABT e o ABD]
 
   \subsection{Música na Matéria: EKP e AHT}
+
+Embora o foco deste trabalho seja na exploração musical através de códigos (abertos!),
+nesta sessão iniciamos uma explicação simples, clara e factual de como estas
+investidas transcendem o código e até mesmo a música em si.
+
+Mais relevantes que os desenvolvimentos em si, são as mobilizações criadas nos
+entornos e os engajamentos. Isso ficará evidente nas próximas - curtas - sessões
+destes desenvolvimentos e resultados.
+
+Aqui apresentamos dois trabalhos que geraram alguma movimentação de pessoas,
+resultando em reuniões, desenvolvimentos, pesquisas e apresentações propriamente
+ditas. O primeiro utiliza o estado do hardware como entrada, o segundo se trata
+de uma mesa escultural para flutuação de origamis que resulta em um instrumento musical
+bastante lúdico.
+
       \subsubsection{Emotional Kernel Panic (EKP)}
@@
 foi lançada a ideia de utilizar o estado do sistema operacional - especialmente o kernel linux - para
-geração de sons. Surge o Emotional Kernel Panic (EKP). Desde o inicio, foram definidos três finalidades
+geração de sons. Surge o Emotional Kernel Panic (EKP) na colaboração excepcional de
+Felipe Machado\footnote{Importantíssimo para o desenvolvimento da Cultura Digital no Brasil [citar fontes e programas governamentais]},
+Ricardo Brazileiro\footnote{Artivista Digital bastante ativo [citar fontes]} e o primeiro autor do presente trabalho.
+
+Desde o inicio, foram definidos três finalidades
 para esta exploração do SO:
+
+[visitar o README do EKP e achar os scripts do Brazileiro sobre o EKP]
 
@@
 
-http://trac.assembla.com/audioexperiments/browser/ekp-base
+Toda a base do EKP está aqui: http://trac.assembla.com/audioexperiments/browser/ekp-base
+
+[Patches desenvolvidos por Brazileiro/Machado e rascunho do EKP-Monitor]
 
       \subsubsection{AirHackTable}
+
+A AirHackTable é um instrumento musical eletrônico controlado por origamis (dobraduras de papel), construída na forma de uma mesa.
+Nela, uma rede de coolers reciclados faz flutuar origamis de geometria e cores variadas.
+Os movimentos dos origamis são captados por webcam e interpretados em tempo real por software de processamento de imagens, gerando padrões que controlam a transformação sonora da música.
+Dessa forma, pode-se dizer que os sons gerados refletem o voo dos origamis de acordo com suas geometrias (que geram trajetórias de voo caracteristicas).
+
+Na versão atual da AirHacktable, cada cor (vermelho, amarelo, azul, verde, preto, ou branco) controla uma voz, e a posição do origami na mesa modula aspectos do som.
+Em outras palavras, se o origami está flutuando mais à esquerda, o som sai no canal à esquerda da mesa, e se o origami está mais próximo da câmera, o volume aumenta, e se está mais afastado do operador, o som se torna mais agudo.
+
+[imagens da AHT e dos patches em operação]
+
+   \subsection{Música no Espírito: Sabrina Kawahara, Audioexperiments, EstudioLivre.org, CDTL, JuntaDados.org, Devolts.org, MSST, LabMacambira.sf.net}
+
+Esta é uma sessão menor, dedicada a apontar repercussões emergentes destas empreitadas em comunidades diferentes
+e então tornar compreensível o desdobramento que estes códigos dedicados à música tiveram em processos sociais
+e mobilizações civis.
+
+Grupos foram montados em torno destes desenvolvimentos. Como os propósitos de compartilhamento
+e apropriação tecnológica estavam no cerne dos grupos relacionados, as investidas
+naturalmente tomaram teores engajados socialmente. A questão do empoderamento das pontas
+e da criação de um patrimônio tecnológico da humanidade é consequência quase imediata
+das posturas de compartilhamento e apropriação citados.
+
+Discorremos brevemente sobre cada uma das iniciativas citadas:
+
+\begin{itemize}
+    \item Sabrina Kawahara
+    \item Audioexperiments
+    \item Estudiolivre.org
+    \item CDTL
+    \item JuntaDados.org
+    \item Devolts.org
+    \item MSST
+    \item LabMacambira.sf.net
+    \item Outros relacionados: MuSa, Metareciclagem, Submidialogia, Tainã
+\end{itemize}
 

### 2012-06-30 devEresultados.tex
repassado especialmente a parte de experimentos abertos em audio

@@
 Os resultados explicitados nesta dissertação estão focados no processamento de 
-áudio voltado para a música e na síntese de estruturas musicais.
+áudio através de códigos em liguagens de programação de computadores e
+voltado para a música e na síntese de estruturas musicais. O tema é amplo
+e pertinente o suficiente para ter gerado ampla bibliografia e um histórico
+que remonta desde a própria história da computação e das linguagens de programação.
+Desta forma, entendemos razoável abordar as empreitadas realizadas pelos autores
+da presente dissertação, com foco nas realizações do primeiro autor.
 
-Por razões didáticas, dividimos este capítulo em 4 partes, a saber: experimentos
+De forma satisfatória, tais realizações possuem um espectro amplo
+dentro do escopo proposto e constitui um percurso relativamente completo
+pelas possibilidades e decorrências da abordagem da música pelo código computacional\footnote{ao menos dentro do se denomina 'Cultura Livre', veja o primeiro capítulo do presente escrito para maiores informações}.
+Por razões didáticas, dividimos este capítulo em 5 partes, a saber: experimentos
 abertos em áudio e música (incluindo plugins LADSPA, Wavelets, Redes Complexas e outros
 scripts experimentais), música em tempo diferido (ou seja, que não é feito em tempo real),
-música em tempo real, e música na matéria (suporte físico em hardware).
+música em tempo real,  música na matéria (suporte físico em hardware) e música no espírito
+(considerando especialmente as mobilizações humanas relacionadas).
 
-Com isso, desejamos percorrer minimamente a prática musical através do código.
+Com isso desejamos expor a prática musical através do código
 com exemplos reais de aplicação e em uso pelo autor, por membros do LabMacambira.sf.net,
@@
 
-Todos os desenvolvimentos desta dissertação estão em repositórios abertos\cite{repositorios-tese-dev},
-e alguns foram especialmente importantes como percurso para o que é apresentado neste
-escrito. Ou seja,
+Todos os desenvolvimentos desta dissertação estão em repositórios abertos\cite{repositorios-tese-dev}.
+Alguns foram especialmente importantes como percurso para o que é apresentado neste
+escrito. Ou seja, são explorações que manifestaram diferentes aspectos
+do trabalho do código voltado para o áudio e a música.
 
-Focaremos no código para a arte sonora,
-incluindo a musical, e portanto fazendo parte da cultura e fazendo cultura.
+O código para a arte sonora, incluindo a musical, manifesta-se como cultura pois é fruto de práticas
+espontâneas, diárias e coletivas.
 \footnote{As chamadas culturas biopunk, ciberpunk, cipherpunk, hacker, digital e outras mais,
 possibilitadas aos recentes desenvolvimentos em telecomunicação, dizem respeito em menor
-ou maior grau ao código.}
+ou maior grau à produção de código como cultura.}
 
 A seguir apresentamos alguns bons exemplos dos desenvolvimentos
-desta dissertação especificamente em áudio, sem o envolvimento direto da música.
+desta dissertação especificamente em áudio, sem o envolvimento direto da música. São
+códigos de necessidade para música, mas de uso mais geral e cuja implementação
+requer muito mais de processamento de sinais e computação do que de música propriamente dito.
 
@@
       [repos AE, LM, wiki EL, historico CDTL]
+LADSPA (Linux Audio Developers Simple Plugin API) é a API livre\footnote{Veja o Capítulo 1 para a definição de livre neste contexto.} estabelecida de plugins de áudio até a presente data.
+A última versão liberada desta API é a 1.1, em uso corrente até hoje e é de 2002 segundo os arquivos de cabeçalho relacionados.
+
+A exploração de APIs específicas foge ao escopo do presente texto e os códigos completos podem ser encontrados pela rede, junto a tutoriais e plugins de exemplo.
+O código destes desenvolvimentos e pertinente ao áudio são relativos à síntese e remoção de ruídos.
+Da síntese, exitem abordagens mais puristas, que no caso sintetiza ruído branco e filtra para resultar nos espectros desejados (veja logo abaixo para maiores especificações).
+Uma abordagem menos purista porém mais eficiente é a de usar uma amostra curta do ruído e reproduzí-la indefinidamente através de \emph{cross-fades}.
+
+Já na remoção de ruídos as abordagens são as mais variadas e extremamente dependentes da aplicação.
+A complexidade dos algorítmos pode atingir níveis de especialidade em processamento de sinais que merecem um trabalho dedicado.
+Aqui iremos expor somente sobre remoção de ruído \emph{'Hum'} causado usualmente pela corrente alternada que alimenta os equipamentos utilizados.
+
+Abaixo dispomos explicações breves sobre os diferentes ruídos e os códigos relacionados. Preferimos expor em Python os códigos de síntese
+por questão de clareza e coerência com o resto do texto\footnote{Os códigos em C/C++ e a implementação como plugin lv2 está no repositório git LV2 como pode ser visto em: http://labmacambira.git.sourceforge.net/git/gitweb-index.cgi}.
+
+O procedimento de reprodução do ruído através de uma amostra repetida e em crossfade é a seguinte:
+
+[código da reprodução do ruído em loop e croos-fade ~10 linhas]
+
+Sintetizando diferentes ruídos\footnote{Dado o propósito do texto e da implementação, utilizamos nomenclaturas utilizadas na música para os ruídos, como se pode notar.}:
+\begin{itemize}
+    \item Ruído Branco: 
+    \item Ruído Rosa:
+    \item Ruído Marrom:
+    \item Ruído Violeta:
+    \item Ruído Azul:
+    \item Ruído Violeta:
+    \item Ruído Cinza:
+    (nao oficiais)
+    \item Ruído Vermelho:
+    \item Ruído Laranja:
+    \item Ruído Preto:
+    \item Ruído Branco Ruidoso:
+    \item Ruído Preto Ruidoso:
+    \item Clicks:
+\end{itemize}
+
+A remoção de ruído Hum é baseada em uma sequência de filtros notch (ver fundamentos) idealmente dispostos nos harmônicos da frequência de oscilação da corrente elétrica.
+Cada um destes filtros deve permitir ajustes finos no fator de qualidade e também na frequência central pois trata-se de um sistema real passivel dos mais diversos efeitos de distorção do modelo ideal.
+Abaixo expomos o código em Python para tal \emph{removedor de ruído} como é chamado.
+O código C/C++ e a implementação como plugin está no mesmo repositório disponibilizado para a síntese de ruídos.
+
+[código e explicação do remodedor de ruído Hum]
 
       \subsubsection{Protocolo de compactação de áudio via wavelets, polinômios e permutações}
-      [achar artigo feec]
+Criado por Rafael Santos Mendes em conjunto com o primeiro autor deste trabalho e em decorrência do mestrado de André Luvizotto, este método
+consiste em aproximar cada folha de uma decomposição WaveletPacket por um polinômio e uma permutação.
+A permutação facilita a aproximação polinomial, pois leva folhas WP de coeficientes ordenados para sucessões melhor comportadas de seus valores\footnote{O método é descrito em um artigo de 2009: Compressão de Áudio via Wavelets, Aproximações Polinomiais e Permutações [Fabbri, Mendes]}.
+Representamos cada permutação por um conjunto ordenado de números inteiros.
+Já o polinômimo possui poucos coeficientes (usualmente 6 ou 8) e os representamos como um conjunto ordenado de variáveis de tipo ponto flutuante.
 
       \subsubsection{Processamento de voz via redes complexas}
-      [repo do audio experiments. passar para o LM e chamar o TT]
+Conversar com Chu e com Luciano e decidir se retiro esta sessão ou mantenho.
 
-  \subsection{Musica em Tempo Diferido: Minimum-fi e FIGGUS}
+      [codigos bem comentados no audio experiments. Apresentação do trabalho "Speech Polarity Detection using Complex Networks Measures: First Explorations no Complenet de 2010"]
+
+  \subsection{Música em Tempo Diferido: Minimum-fi e FIGGUS}
 
@@
 
-   \subsection{Música no Espírito: Sabrina Kawahara, Audioexperiments, EstudioLivre.org, CDTL, JuntaDados.org, Devolts.org, MSST, LabMacambira.sf.net}
+   \subsection{Música no Espírito: Sabrina Kawahara, Audioexperiments, EstudioLivre.org, CDTL, juntaDados.org, Devolts.org, MSST, LabMacambira.sf.net}
 

### 2012-06-30 pretextual.tex
revisao geral dos materiais didaticos em textos.
	modified:   README.txt
	modified:   devEresultados.tex
	modified:   pretextual.tex
terminada a revisao geral da dissertacao
terminada a repassada na parte musical, agora eh preencher melhor as partes
repassado especialmente a parte de experimentos abertos em audio

@@
                 \item[SOS] \emph{Saúde Olha Sabedoria}
+                \item[FISL] \emph{Festival Internacional de Software Livre}
+                \item[BPM] \emph{Batidas Por Minuto (medida de andamento musical)}
+                \item[API] \emph{Application Programming Interface ou Interface de Programação de Aplicativos}
+                \item[WP] \emph{Wavelet Packet}
+
+		\item[EL] \emph{Estúdio Livre}
+		\item[AE] \emph{AudioExperiments}
 	\end{listaespecial}
@@
 		\item[\#] Resto da linha é comentario em código Bash ou Python
-		\item[\\\\] Resto da linha é comentario em código C/C++
+		\item[] Resto da linha é comentario em código C/C++
 	\end{listaespecial}

### 2012-07-05 devEresultados.tex
andamentos feitos nos resultados, com dicas de vilson vieira e outros

@@
 Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})
-
-\section{Áudio e Música}
-
 Os resultados explicitados nesta dissertação estão focados no processamento de 
@@
 voltado para a música e na síntese de estruturas musicais. O tema é amplo
-e pertinente o suficiente para ter gerado ampla bibliografia e um histórico
+e pertinente o suficiente para contar atualmente com ampla bibliografia e um histórico
 que remonta desde a própria história da computação e das linguagens de programação.
-Desta forma, entendemos razoável abordar as empreitadas realizadas pelos autores
-da presente dissertação, com foco nas realizações do primeiro autor.
+Desta forma, entendemos cabível focar este texto nas empreitadas realizadas pelos autores
+da presente dissertação, com foco nas realizações do primeiro autor. São usos
+reais destas tecnologias, manifestadas de forma cultural e com repercussões
+sociais, como poderá é relatado neste mesmo capítulo.
 
-De forma satisfatória, tais realizações possuem um espectro amplo
+Tais realizações possuem um espectro amplo
 dentro do escopo proposto e constitui um percurso relativamente completo
 pelas possibilidades e decorrências da abordagem da música pelo código computacional\footnote{ao menos dentro do se denomina 'Cultura Livre', veja o primeiro capítulo do presente escrito para maiores informações}.
-Por razões didáticas, dividimos este capítulo em 5 partes, a saber: experimentos
-abertos em áudio e música (incluindo plugins LADSPA, Wavelets, Redes Complexas e outros
-scripts experimentais), música em tempo diferido (ou seja, que não é feito em tempo real),
-música em tempo real,  música na matéria (suporte físico em hardware) e música no espírito
-(considerando especialmente as mobilizações humanas relacionadas).
 
-Com isso desejamos expor a prática musical através do código
-com exemplos reais de aplicação e em uso pelo autor, por membros do LabMacambira.sf.net,
-por parceiros, colaboradores e por usuários eventuais das naturezas mais diversas.
+Estaremos abordando experimentos abertos em áudio, abordagens musicais pelo código, incluindo música em tempo real, diferido, na matéria e repercussões no tecido social.
+De forma a relatar os resultados destas investidas, segue uma sessão sobre ações coletivas realizadas através das mobilizações que estes códigos criaram, e as estruturas virtuais mantidas na intenet relacionadas a estas mobilizações e ações.
+Por fim, alguns materiais didáticos criados pelo autor - que possuem algum papel emblemático - são também expostos por comporem o arsenal de repercussões desta investida no em som e código.
 
-  \subsection{Experimentos abertos em áudio: LADSPAs, Wavelets e Redes Complexas}
+  \section{Experimentos abertos em áudio: LADSPAs, Wavelets e Redes Complexas}
 
@@
       [codigos bem comentados no audio experiments. Apresentação do trabalho "Speech Polarity Detection using Complex Networks Measures: First Explorations no Complenet de 2010"]
+
+\section{Áudio e Música}
+
+Por razões didáticas, dividimos este capítulo em 4 partes, a saber: música em tempo diferido (ou seja, que não é feito em tempo real),
+música em tempo real,  música na matéria (suporte físico em hardware) e música no tecido social
+(considerando especialmente as mobilizações humanas relacionadas).
+
+Com isso desejamos expor a prática musical através do código
+com exemplos reais de aplicação e em uso pelo autor, por membros do LabMacambira.sf.net,
+por parceiros, colaboradores e por usuários eventuais das naturezas mais diversas.
 
@@
     \item O compartilhamento do código é usual, leve e eficiente como entrega completamente
-    aberta de tecnologia e proposta estética.
+    aberta da tecnologia relacionada à proposta estética.
 \end{itemize}
@@
 
-   \subsection{Música no Espírito: Sabrina Kawahara, Audioexperiments, EstudioLivre.org, CDTL, juntaDados.org, Devolts.org, MSST, LabMacambira.sf.net}
+   \subsection{Música no Tecido Social: Sabrina Kawahara, Audioexperiments, EstudioLivre.org, CDTL, juntaDados.org, Devolts.org, MSST, LabMacambira.sf.net}
 

### 2012-07-11 devEresultados.tex
tapa grande no minimum-fi, quase terminado

@@
 
-Em Abril de 2011 surgiu a necessidade estética e também tecnológica
-de realizar um código dedicado a materializar \emph{o mínimo}
-para expressar ideias musicais em geral. Esta empreitada
-resultou no minimum-fi, código em um único arquivo, que sintetiza
-músicas inteiras. Os algorítmos propriamente ditos, escritos em Python, somam 
+Existe uma perene matiz estética e também tecnológica
+de realizar uma dada tarefa com \emph{o mínimo} necessário.
+Esta matiz é para a música igualmente
+fundamental e reconhecida como
+princípio de unidade e coerência\footnote{Este princípio tanto é fundamental
+que as escolas musicais possuem técnicas específicas, músicas possuem suas
+próprias convenções mantidas por toda a sua duração, os arcos mantém características,
+enfim, podemos até mesmo concluir que a simetria define o escopo.}. Em código computacional
+a empreitada para manifestar este princípio ele próprio de forma mínima
+resultou no \emph{minimum-fi.py}, código Python em um único arquivo curto que sintetiza
+músicas com segundo as estruturas especificadas em linha\footnote{Em música se fala da \emph{bula} (que equivaleria
+às funções em si e das especificações da linguagem) e a partitura ou 'música' propriamente ditas (que equivale à .}.
+Na versão atual, de 2012 mas adiantada em 2011, os algorítmos 
+em Python propriamente ditos somam 
 53 linhas e inclui 5 funções. Com estas funções, estruturas musicais podem 
-ser criadas amostra por amostra, resultando em notas, blocos e estruturas hierarquicamente superiores.
+ser criadas padrão a padrão, nota a nota, amostra por amostra. Na prática e posto em
+linguagem cotidiana, resultam em notas que formam
+blocos e estruturas hierarquicamente superiores.
 
-Os princípios, bastante simples, são:
+Os princípios, bastante simples\footnote{Osvaldo Lacerda, em seu livro \emph{Compendio de Teoria Elementar da Musica} fala das propriedades do som musical de forma equivalente.}, são:
 \begin{itemize}
   \item Deve-se ter um mecanismo de síntese sonora que
-permita vários timbres e controle sobre a frequência fundamental.
-  \item Deve-se ser capaz de construir unidades e séries de unidades (notas e acordes, por exemplo).
+possibilite a geração de unidades sonoras com diferentes timbres, controle sobre a frequência fundamental e duração específica\footnote{Também entra aqui o volume, mas como o que controlamos de fato é a intensidade relativa entre as notas, resolvemos por bem omitir
+esta parte até que fiquem claros os códigos relativos a isso, logo abaixo.}.
+  \item Deve-se ser capaz de construir não só unidades sonoras, mas também séries de unidades, sejam sobrepostas (acordes, por exemplo)
+  ou justapostas (melodias, por exemplo).
 \end{itemize}
@@
 um waveshaping. Este procedimento é bastante apreciado pela simplicidade e eficácia
-na síntese de tímbres diversos e ricos em harmônicos e evolução temporal.
+na síntese de tímbres diversos e ricos em harmônicos e evolução temporal. Embora
+uma explicação exaustiva do waveshaping fuja ao escopo deste trabalho, este
+método se caracteriza pela aplicação de uma função não linear ao sinal de entrada.
+Neste caso o sinal de entrada é gerado pelo primeiro lookup, a função não linear aplicada
+é representada pela segunda tabela e aplicada pelo segundo lookup.
 
-[codigo do lookupcruz e explicacao]
+Existem muitas formas de se executar waveshaping, a seguir segue o que usamos no minimum-fi, que se sustenta principalmente por ser leve e simples:
 
-O segundo item presta-se à discretização do espaço musical. Unidades como batidas e notas
+\code{Waveshaping com Lookups sucessivas}{python_snippets/lookup_cruz.py}
+
+O segundo item - dos dois princípios expostos sobre o minimum-fi - presta-se à discretização do espaço musical. Unidades como batidas e notas
 tornam mais eficiente a comunicação pois a quantidade
-de estruturas sugeridas é maior e as estruturas são mais claras~\cite{Roederer} no discreto do que no contínuo. Roederer
-aponta que as próprias notas dos instrumentos musicais são um reflexo de que é mais eficiente
-o uso do discreto do que do contínuo.
+de estruturas sugeridas é maior e as estruturas são mais claras no discreto do que no contínuo~\cite{Roederer}. Roederer chega a
+apontar que as próprias notas dos instrumentos musicais são um reflexo de que é mais eficiente
+o uso do discreto do que do contínuo para a geração de estruturas musicais.
 
-Desta forma, unidades bem definidas se mostram úteis na prática musical 
-para fazer sequencias de unidades, concatená-las. Quando as unidades
-são notas, as sequencias de unidades justapostas no tempo são melodias ou linhas melódicas. As
-sequencias sobrepostas no tempo são comumente pensados como acordes, mas podem ser tidos simplesmente
+De fato, unidades bem definidas se mostram úteis na prática musical 
+para fazer sequências de unidades, concatená-las. Quando as unidades
+são notas, as sequências de unidades justapostas no tempo são melodias ou linhas melódicas. As
+sequências sobrepostas no tempo são comumente pensadas como acordes, mas podem ser tidas simplesmente
 como sobreposições circunstanciais de duas linhas melódicas. Isso, claro, segundo
-a sistematização clássica e usual da música~\cite{Lacerda}.
+a sistematização clássica e usual da música~\cite{Lacerda}\footnote{Vole assinalar aqui que a música do século XX apresentou diversos modelos teóricos que quebram com este entendimento simplificado sobre a música, suas unidades básicas e estruturas relacionadas}.
 
-As duas construções básicas explicitadas, baseadas na dicotomia melodia/harmonia
+As duas construções básicas explicitadas a seguir, baseadas na dicotomia melodia/harmonia
 (horizontalidade/verticalidade, justaposição/sobreposição), são
-as funções \emph{fazSequencia} e \emph{fazAcorde}. Vale notar elas são absolutamente 
-equivalentes em uma análise puramente conceitual, i.e. uma delas pode ser omitida segundo
-a teoria. Isso fica particularmente óbvio quando se nota que os procedimentos de mixagem e concatenação são
+as funções \emph{fazSequencia} e \emph{fazAcorde} no minimum-fi. Vale notar elas são absolutamente 
+equivalentes em uma análise puramente conceitual, i.e.
+uma delas pode ser omitida segundo algumas teorizações.
+Isso fica particularmente óbvio quando se nota que os procedimentos de mixagem e concatenação são
 plenamente capazes de realizar o que estas funções realizam. Aliás, as funções nada
@@
 
-Como pode notar a seguir, as sequencias de notas e os acordes, em última instância, são utilizações específicas
-das 2 funcoes de síntese sonora explicadas anteriormente: lookup e lookupcruz.
+Como pode-se notar a seguir, as sequências de notas e os acordes, em última instância, são utilizações específicas das 2 funções de síntese sonora explicadas anteriormente: lookup e lookupcruz.
 
-[codigo do fazSequencia e explicacao]
+\code{Realização de Sequências de Notas}{python_snippets/fazSequencia.py}
 
-[codigo do fazAcorde e explicacao]
+\code{Realização de Acordes de Notas}{python_snippets/fazAcorde.py}
 
 A última das cinco funções utilizadas é uma soma amostra a amostra de dois sons. Para isso,
-é necessário completar com zeros a sequencia com o menor número de amostras:
+é necessário completar com zeros a sequência com o menor número de amostras para somar rapidamente:
 
-[codigo do somador, explicacao e figura de uma sequencia maior/menor que a outra]
+\code{Somador (função auxiliar)}{python_snippets/somador.py}
 
@@
 algumas tabelas para serem usadas como diferentes timbres, pode-se criar as
-escalas completamente simetricas assim:
+escalas completamente simétricas na oitava cromática assim:
 
-\code{Escalas Simétricas na grade dos 12 semintos e no âmbito da oitava}{python_snippets/escalas_simetricas.py}
+\code{Escalas Completamente Simétricas (na grade dos 12 semitons e no âmbito da oitava)}{python_snippets/escalas_simetricas.py}
 
@@
 
-Sendo cada unidade um semitom. Claro que podemos também utilizar as usuais
+Sendo cada unidade um semitom. Ou seja, cada unidade é o fato que fica presente na conta $f_0 . 2^(1/12) . fator $ e resulta na frequência exata da nota em um sistema temperado ideal.
+
+Claro que podemos também utilizar as usuais
 escalas tonais maior e menor natural, harmônica e melódica:
 
-[codigo das escalas maior e menores]
+\code{Escalas diatônicas tonais (maiores e menores)}{python_snippets/escalas_Mm.py}
 
@@
 como \emph{microtonalidade}} mostra-se trivial em
-nossa implementação:
+nossa implementação. Basta representar sequências de fracionários
+que são partes proporcionais dos semitons:
 
-[escalas com intervalos microtonais]
+\code{Escalas Microtonais (com quartos e oitavos de tom e intervalos menores)}{python_snippets/escalas_microtonais.py}
 

### 2012-07-11 python_snippets/escalas_Mm.py

@@ -1,0 +1,9 @@
+# A escala maior é compreendida como tendo forma única:
+escala_maior=[0,2,4,5,7,9,11]
+# Já a escala menor tem 3 formas:
+escala_menor_natural=[0,2,3,5,7,8,10]
+escala_menor_hamonica=[0,2,3,5,7,8,11]
+# além disso, a escala melódica tem 
+# a forma ascendente distinta da descendente:
+escala_menor_melodica_ascendente=[0,2,3,5,7,9,11]
+escala_menor_melodica_descendente=[10,8,7,5,3,2,0]

### 2012-07-11 python_snippets/fazAcorde.py

@@ -1,0 +1,16 @@
+def fazAcorde(notas, f0, fator, dur_nota, dur_silencio, funcao1, funcao2=0):
+    seq=[]
+    for nota in notas:
+        freq=f0*fator**nota
+        if funcao2:
+            amostras=lookupcruz(funcao1,funcao2,dur_nota,freq)
+        else:
+            amostras=lookup(funcao1,dur_nota,freq)
+        amostras=amostras+[0]*int(dur_silencio*taxa_amostragem)
+        # para sobrepor unidades de tamanhos diferentes
+        # é conveniente uma função auxiliar
+        seq=somador(seq,amostras) 
+    # normalizaçào em duas linhas:
+    smin=min(seq); smax=max(seq)
+    seq= [(-.5+(i-smin)/(smax-smin)) for i in seq]
+    return seq

### 2012-07-11 python_snippets/fazSequencia.py

@@ -1,0 +1,16 @@
+# Note as variáveis que são fixadas de antemão
+def fazSequencia(notas, f0, fator, dur_nota, dur_silencio, funcao1, funcao2=0):
+    seq=[] # a sequência é iniciada vazia
+    for nota in notas: # para cada nota
+        # a frequência é a fundamental vezes 2 elevado ao número de semitons:
+        freq=f0*fator**nota 
+        if funcao2: # caso a funcao2 seja especificada, fazer o waveshaping
+            amostras=lookupcruz(funcao1,funcao2,dur_nota,freq)
+        else:
+            amostras=lookup(funcao1,dur_nota,freq)
+        amostras += [0]*int(taxa_amostragem*dur_silencio)
+        seq += amostras
+    # normalização em 2 passos:
+    smin=min(seq); smax=max(seq)
+    seq= [(-.5+(i-smin)/(smax-smin)) for i in seq]
+    return seq # retorna a sequencia pronta

### 2012-07-11 python_snippets/somador.py

@@ -1,0 +1,10 @@
+def somador(seq1,seq2):
+    # Basicamente adiciona zeros na sequencia menor para somar rapidamente
+    diff=len(seq1)-len(seq2)
+    if diff > 0:
+        seq2+=[0]*diff
+    elif diff< 0:
+        seq1+=[0]*-diff
+        
+    seq=[(i+j) for i,j in zip(seq1,seq2)]
+    return seq

### 2012-07-11 devEresultados.tex
terminada a parte do minimum-fi

@@
 
-Sendo cada unidade um semitom. Ou seja, cada unidade é o fato que fica presente na conta $f_0 . 2^(1/12) . fator $ e resulta na frequência exata da nota em um sistema temperado ideal.
+Sendo cada unidade um semitom.
+Ou seja, cada unidade é o fator que fica presente na conta $f_0 .
+2^{(1/12)} .
+fator $ e resulta na frequência exata da nota relativa a $f_0$ em um sistema temperado ideal.
 
-Claro que podemos também utilizar as usuais
+Podemos também utilizar as usuais
 escalas tonais maior e menor natural, harmônica e melódica:
@@
 A utilização de intervalos menores que o semitom\footnote{Prática também conhecida
-como \emph{microtonalidade}} mostra-se trivial em
+como \emph{microtonalismo}, que gera/utiliza \emph{microtonalidade}.} mostra-se trivial em
 nossa implementação. Basta representar sequências de fracionários
-que são partes proporcionais dos semitons:
+que são partes linearmente proporcionais dos semitons:
 
@@
 são especificadas com simplicidade. Note que aos acordes \emph{diminutão} e
-\emph{aumentado} correspodem as mesmas notas das escalas simetricas de terças menores
+\emph{aumentado} correspodem as mesmas notas das escalas simétricas de terças menores
 e maiores respectivamente:
@@
 
-E séries podem ser anotadas junto a variacoes:
+E séries/sequências podem ser anotadas junto a variacoes:
 
@@
 \code{Sintetizando sequências e acordes}{python_snippets/seqs_acordes.py}
+
+Já a síntese de estruturas compostas (p.ex. sequências de acordes e sobreposição
+de linhas melódicas), são feitas com os recursos usuais da liguagem. Neste caso
+usamos listas em Python e implementações em Numpy (mais eficientes em tempo de
+execussão e também na simplicidade do código) estão no apêndice. Estes mesmos
+procedimentos são praticamente os mesmos em Scilab, C/C++, Javascript, PHP, etc.
+
+A seguir demonstramos para fins didáticos, a construção de acordes periódicos em
+python puro\footnote{A implementação não didática destas três linhas de código
+podem ser agrupadas em uma só linha.}:
+
 \code{Acordes periódicos}{python_snippets/acordes_periodicos.py}
 
-Em posse destas sequências e acordes, formamos estruturas hierarquicamente superiores
+Em posse destas sequências, acordes e recursos da linguagem,
+formamos estruturas hierarquicamente superiores
 através da concatenação de estruturas, da mixagem de estruturas, e da amplificação

### 2012-07-11 python_snippets/series.py

@@ -1,6 +1,7 @@
+# padrão rítmico e variações
 b1=[0]*4
 b2=[0]*7+[7]
-b3=  [0,7,0,1]
+b3=[0,7,0,1]
 b3_2=[0,7,8,7]
 b4=[0,7,7,7]
 b5=[0,7]
@@ -9,3 +10,11 @@ b7=[0,-1,0,-1]
 b8=[0,-1,7,-1]
 b9=[0,1,0,-1]
 b10=[0,7,6,7]
+
+# Série utilizada por Anton Webern
+# em suas Variações Op.30
+c=[0,1,4,3,2,5,6,9,8,7,10,11]
+
+# Sequência em mixolídio comum, por exemplo,
+# Em algumas manifestações do folclore nordestino:
+d=[0,4,7,10,9,7,10]

### 2012-07-11 devEresultados.tex
atualizando, finalizacoes minimas minimum-fi e figgus iniciando

@@
 
-        \subsubsection{Finite Groups in Granular and Unit Synthesis (FIGGUS)}
+Neste ponto, basta criarmos músicas e sequências de interesse estético ou para pesquisa.
+Dado o ferramental, os encadeamentos dependem de intensões estéticas, entendimentos musicais e estruturas abstratas que mantém a coerência e interesse em uma peça musical.
+Para o leitor mais interessado, deixamos para o Apêndice X um exemplo de música feita com o minimum-fi.
 
-Desenvolvimento iniciado em 2006 com o físico-matemático Prof. Adolfo Maia Junior para
+\vspace{10 mm}
+
+        \subsubsection{FIGGUS: FInite Groups in Granular and Unit Synthesis}
+
+Neste ponto - por razões didáticas e também de coerência com os conhecimentos
+e tecnologias acumuladas nesta exposição - é que introduzimos um foco especial nas estruturas musicais
+desenvolvidas. São elas que constituem uma linguagem musical e músicas propriamente
+ditas\footnote{E especificamos o arcabouço para geração de sons musicais em unidades
+e estruturas compostas logo acima na subsessão sobre o minimum-fi.}.
+
+O ponto alto nesse desenvolvimento foi iniciado em 2006, bem anterior
+ao nascimento do minimum-fi\footnote{Embora este já estivesse sendo concebido.} 
+com o físico-matemático Prof. Adolfo Maia Junior para
 tratar de simetrias na música, com vistas à composição musical através
 de métodos matemáticos. Mais especificamente, a proposta resultou em
-um programa voltado para o uso de Grupos Algébricos para síntese
-granular e síntese de estruturas musicais. O nome dado
+um programa voltado para a síntese
+granular e síntese de estruturas musicais através de Grupos Algébricos. O nome dado
 foi FIGGUS, sigla de FInite Groups in Granular and Unit Synthesis\footnote{Também utilizamos
 o nome FIGGS (FInite Group in Granular Synthesis) dado que o termo \emph{unit synthesis} não
-é usual na literatura. Posteriormente o principal autor deste trabalho recorreu novamente
-ao FIGGUS pois o maior uso da abordagem é para síntese de estruturas musicais, não para
-síntese de amálgamas sonoros tipicamente resultantes da síntese granular}.
+é usual na literatura. Posteriormente o primeiro autor deste trabalho recorreu novamente
+ao uso do nome FIGGUS. Isso foi motivado principalmente pelo fato de que
+o maior uso da técnica é para síntese de estruturas musicais, não para
+síntese de amálgamas sonoros (ou timbres mesmo) tipicamente resultantes da síntese granular}.
 
@@
 
-Na atual reescrita, embora ainda bastante atrás do FIGGUS original,
-e mesmo sem a interface gráfica, o código roda inteiro em python
+Na atual reescrita, embora ainda bastante atrás do FIGGUS original quanto
+à interface gráfica, a ferramenta opera diretamente em Python puro,
 com as biblitecas imbutidas por padrão. Isso permite com que o FIGGUS
@@
 
-[[ARRUMAR A PARADA PARA O CODIGO]]
-git clone
-python setup.py install
-python RUNME
+\code{Utilizando o FIGGUS para Sintetizar um EP}{python_snippets/synth_ep.py}
 

### 2012-07-11 python_snippets/amp_mix.py

@@ -1,11 +1,11 @@
-## De posse das notas, fundamental, fator, durações e ondas...
+#-# De posse das notas, fundamental, fator, durações e ondas #-#
 # sequência do restrogrado das notas com waveshapping:
 sequencia1 = fazSequencia(nt[::-1], f, fator, d, s, senoide, dente)
 # sequência original das notas uma quinta justa acima,
 # com dente de serra:
 sequencia2 = fazSequencia(nt, f*(3/2.), fator, d, s, dente)
 
-## Procedimentos básicos:
+#-# Procedimentos básicos #-#
 # concatenação
 seq = sequencia1+sequencia2 
 # amplificação e concatenação

### 2012-07-11 devEresultados.tex
revisada geral e tb andada c o principal

@@
 nossa implementação. Basta representar sequências de fracionários
-que são partes linearmente proporcionais dos semitons:
+que são partes linearmente proporcionais dos semitons ou modificar o fator $f$:
 
@@
 
-E séries/sequências podem ser anotadas junto a variacoes:
+E séries/sequências podem ser anotadas junto a variações:
 
@@
 Para o leitor mais interessado, deixamos para o Apêndice X um exemplo de música feita com o minimum-fi.
+A seguir utilizamos esta base apresentada para sintetizar estruturas musicais e então um EP.
 
@@
 e tecnologias acumuladas nesta exposição - é que introduzimos um foco especial nas estruturas musicais
-desenvolvidas. São elas que constituem uma linguagem musical e músicas propriamente
+desenvolvidas. São elas - inseridas em um momento histórico e executadas por instrumentos específicos com as técnicas de época - que constituem uma linguagem musical e músicas propriamente
 ditas\footnote{E especificamos o arcabouço para geração de sons musicais em unidades
@@
 O ponto alto nesse desenvolvimento foi iniciado em 2006, bem anterior
-ao nascimento do minimum-fi\footnote{Embora este já estivesse sendo concebido.} 
+ao nascimento do minimum-fi, 
 com o físico-matemático Prof. Adolfo Maia Junior para
-tratar de simetrias na música, com vistas à composição musical através
+tratar de simetrias na música com vistas à composição musical através
 de métodos matemáticos. Mais especificamente, a proposta resultou em

### 2012-07-11 introducao.tex
terminada reestruturacao da dissertacao
arrumarinha geral
arrumado provisoriamente o template do codigo e tamb arrumada na intro mudei o Estado das coisas...
revisada geral e tb andada c o principal

@@
 Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})
-Partindo do som como um fenomeno fíisico, esta dissertação se propõe a ser um guia relativamente 
-completo para uma atuação eficiênte no desenvolvimento
-de software voltado ao áudio e tecnologias livres. Optamos
+Partindo do som como um fenomeno fíisico, esta dissertação 
+cobre casos de uso musicais de software voltado ao áudio e tecnologias livres. Optamos
 realizar este trabalho através da síntese e da simplicidade.
@@
 
-  \section{Estado das coisas}
+	  \subsubsection{Domínio público e outras categorias}
+
+  \section{Notas sobre o estado do software}
   \label{sec:context}
@@
 
-      \subsection{A convergência das mídias}
+      \subsection{A convergência das mídias e o multimídia}
       \label{sec:midiamultimidia}
@@
 
-      \subsection{Multimídia}
-
 	  \subsubsection{Gargalos na interatividade e no estado do software}
@@
 
+%%%%%%%%
+      \subsection{Identificação de tarefas pouco suportadas}
+      \label{sec:tarefas_n_sup}
+  [ESCREVER NO FINAL]
+
 	  \subsubsection{Compartilhamento}
@@
 	  desencadear os desenvolvimentos ligitimos.
-
-%%%%%%%%
-      \subsection{Identificação de tarefas pouco suportadas}
-      \label{sec:tarefas_n_sup}
-  [ESCREVER NO FINAL]
 

### 2012-07-11 python_snippets/fazSequencia.py

@@ -2,15 +2,19 @@
 def fazSequencia(notas, f0, fator, dur_nota, dur_silencio, funcao1, funcao2=0):
     seq=[] # a sequência é iniciada vazia
     for nota in notas: # para cada nota
-        # a frequência é a fundamental vezes 2 elevado ao número de semitons:
+        # a frequência é a fundamental vezes 2
+        # elevado ao número de semitons:
         freq=f0*fator**nota 
-        if funcao2: # caso a funcao2 seja especificada, fazer o waveshaping
+        if funcao2: # caso a funcao2 seja especificada
+            # fazer o waveshaping:
             amostras=lookupcruz(funcao1,funcao2,dur_nota,freq)
         else:
+            # ou faz com a forma de onda simples:
             amostras=lookup(funcao1,dur_nota,freq)
+        # soma o silêncio
         amostras += [0]*int(taxa_amostragem*dur_silencio)
-        seq += amostras
+        seq += amostras # concatena na sequência principal
     # normalização em 2 passos:
     smin=min(seq); smax=max(seq)
     seq= [(-.5+(i-smin)/(smax-smin)) for i in seq]
-    return seq # retorna a sequencia pronta
+    return seq # retorna a sequência pronta

### 2012-07-15 devEresultados.tex
mudancas de antes dessa sessao

@@
 
-Neste ponto - por razões didáticas e também de coerência com os conhecimentos
-e tecnologias acumuladas nesta exposição - é que introduzimos um foco especial nas estruturas musicais
-desenvolvidas. São elas - inseridas em um momento histórico e executadas por instrumentos específicos com as técnicas de época - que constituem uma linguagem musical e músicas propriamente
-ditas\footnote{E especificamos o arcabouço para geração de sons musicais em unidades
-e estruturas compostas logo acima na subsessão sobre o minimum-fi.}.
+Em coerência com os conhecimentos
+e tecnologias já apresentados nesta exposição\footnote{Especificamos o arcabouço para geração de sons musicais em unidades
+e estruturas compostas logo acima na subsessão sobre o minimum-fi.}, introduzimos agora
+um foco especial nas estruturas musicais
+desenvolvidas. São elas, inseridas em um momento histórico e
+executadas por instrumentos específicos com as técnicas de época,
+que constituem uma linguagem musical e músicas propriamente
+ditas[ref]. O FIGGUS constitui uma técnica composicional
+manifestada em software como ferramenta de síntese de
+estruturas musicais por estruturas matemáticas específicas
+para representação de simetrias.
 
-O ponto alto nesse desenvolvimento foi iniciado em 2006, bem anterior
-ao nascimento do minimum-fi, 
-com o físico-matemático Prof. Adolfo Maia Junior para
+Este desenvolvimento foi iniciado em 2006 com o físico-matemático Prof. Adolfo Maia Junior - bem anterior
+ao nascimento do minimum-fi - para
 tratar de simetrias na música com vistas à composição musical através
-de métodos matemáticos. Mais especificamente, a proposta resultou em
+de métodos matemáticos\footnote{Duas iniciações científicas trataram do assunto, convenios XXX e YYYY.}. Mais especificamente, a proposta resultou em
 um programa voltado para a síntese
@@
 
-A Síntese Granular é uma área bem estabelecida tanto na acústica quanto
-na Computação Musical. O Físico Gabor, quantum sonoro, phonon, etc..
-[achar artigos feitos com adolfo]
-
 Na atual reescrita, embora ainda bastante atrás do FIGGUS original quanto
@@
 \code{Utilizando o FIGGUS para Sintetizar um EP}{python_snippets/synth_ep.py}
+
+A Síntese Granular é uma área bem estabelecida tanto na acústica quanto
+na Computação Musical e se caracteriza pela geração de sons bastante curtos
+e em quantidade massiva. Tipicamente os sons possuem entre 5 e 40 milissegundos
+e a quantidade destes \emph{microsons}\footnote{Grãos sonoros e microsons são jargões
+típicos da síntese granular usados para indicar sons com as durações especificadas
+no texto.} pode chegar a milhares por segundo. O tratamento específico da
+síntese granular foge ao escopo deste trabalho e indicamos ao leitor interessado
+os artigos produzidos sobre Síntese Granular e Teoria de Grupos[refs AES e SBCM].
+Assim, focamos o texto a seguir em Grupos Finitos e Síntese de Estruturas.
+
+Matematicamente, as \emph{simetrias} são descritas por Grupos. Nas artes é de comum conhecimento o papel absolutamente central
+que as simetrias possuem. Na música, para citar somente alguns exemplos simples,
+temos os numerosos estudos de simetrias na música de J. S. Bach, os jogos
+de dados de Mozart e os usos recorrentes da proporção áurea na música
+de Béla Bártok.
+
+Embora as aplicações de simetrias neste caso possam se desenvolver
+de diversas formas com interesse estético no plano sonoro, em
+nossas pesquisas, uma delas se mostrou particularmente coerente
+e pertinente: as permutações de unidades sonoras. Primeiro,
+através do Teorema de Cayley, todo grupo é isomorfo
+a um grupo de permutação\footnote{Mais especificamente,
+todo grupo é isomorfo a um subgrupo do grupo simétrico agindo
+em G.}, i.e. seus elementos são diretamente relacionados
+a permutações. Além disso, na música as permutações estão
+no núcleo de diversos procedimentos canônicos
 

### 2012-07-15 python_snippets/escalas_microtonais.py

@@ -9,10 +9,10 @@ bar = [0,2,3,4.75,5,5.25,7,8,11]
 # Escala não ordenada de valores randômicos:
 baz = [1.197,9.580,8.309,8.726,2.355]
 
-# # Ou modifique o fator f
-# # com que calcula as frequências:
+#-# Ou modifique o fator f
+# # com que calcula as frequências: #-#
 f = 2**(1/19.) # 19 graus na escala
-# melodia em sistema de 19 graus
+# melodia em sistema de 19 graus:
 omega = [1,5,9,9,15,2,3] 
 # padrão rítmico em 19 graus:
 theta = [0,0,0,7,7,7,7]

### 2012-07-17 devEresultados.tex
modificacoes pequenas que fiz ontem ou +- por aih

@@
 
+Desta forma, a ferramenta fica muito mais simples para experimentações
+e implementações adicionais pois sua utilização se dá através
+do próprio código no qual tudo é especificado. Esta abordagem se mostra particularmente
+pertinente pois a estagnação do mecanismo de síntese de estruturas musicais
+beira a inviabilização do uso artístico ou torna necessário o uso
+de recursos externos.
+Outro uso pretendido para o FIGGUS
+é a síntese de tímbres e amálgamas sonoros através da
+Síntese Granular. Embora nosso foco seja outro, cabe algumas breves
+palavras sobre o assunto.
+
 A Síntese Granular é uma área bem estabelecida tanto na acústica quanto
@@
 os artigos produzidos sobre Síntese Granular e Teoria de Grupos[refs AES e SBCM].
-Assim, focamos o texto a seguir em Grupos Finitos e Síntese de Estruturas.
+Assim, focamos o texto a seguir em Grupos Finitos e Síntese de Estruturas Musicais.
 
-Matematicamente, as \emph{simetrias} são descritas por Grupos. Nas artes é de comum conhecimento o papel absolutamente central
+Nas artes é de comum conhecimento o papel absolutamente central
 que as simetrias possuem. Na música, para citar somente alguns exemplos simples,
@@
 de dados de Mozart e os usos recorrentes da proporção áurea na música
-de Béla Bártok.
+de Béla Bártok. Matematicamente, as \emph{simetrias} são descritas por Grupos,
+e estes são definidos da seguinte forma:
+
+[desc dos grupos]
 

### 2012-07-17 devEresultados.tex
adicionado sobre grupos

@@
 de Béla Bártok. Matematicamente, as \emph{simetrias} são descritas por Grupos,
-e estes são definidos da seguinte forma:
+e estes são definidos como sendo um conjunto (chamemos de $G$)
+munido de uma operação (seja $\bullet$), formando um grupo $(G,\bullet)$
+satisfazendo as seguintes propriedades:
 
-[desc dos grupos]
+\begin{enumerate}
+    \item Fechamento: $\forall g_1, g_2  \in G \Rightarrow g_1 \bullet g_2 \in G$
+    \item Associatividade: $(g_1 \bullet g_2) \bullet g_3 = g_1 \bullet (g_2 \bullet g_3), \forall g_1, g_2, g_3 \in G$
+    \item Existência do elemento neutro: $\exists e \in G : g \bullet e = e \bullet g = g$
+    \item Existência dos inversos: $\forall g \in G \exists g^{-1} : g \bullet g^{-1} = g^{-1} \bullet g = e$
+\end{enumerate}
+
+Adicionalmente, vale complementar que se um grupo tiver um número
+finito de elementos, ele é dito finito, caso contrário é chamado
+infinito. Todos os elementos de um grupo finito são cíclicos pois
+operado com si próprio um número suficiente de vezes resulta em si
+mesmo. Um grupo é dito comutativo ou abeliano caso a propriedade
+comutativa seja satisfeita para todos os elementos: $\forall g_1, g_2 \in G \Rightarrow g_1 \bullet g_2 = g_2 \bullet g_1$
 

### 2012-07-18 devEresultados.tex
andamentos com o FIGGUS, falando de grupos e do codigo em si, falta labutar +++

@@
 \begin{enumerate}
-    \item Fechamento: $\forall g_1, g_2  \in G \Rightarrow g_1 \bullet g_2 \in G$
-    \item Associatividade: $(g_1 \bullet g_2) \bullet g_3 = g_1 \bullet (g_2 \bullet g_3), \forall g_1, g_2, g_3 \in G$
-    \item Existência do elemento neutro: $\exists e \in G : g \bullet e = e \bullet g = g$
-    \item Existência dos inversos: $\forall g \in G \exists g^{-1} : g \bullet g^{-1} = g^{-1} \bullet g = e$
+    \item Fechamento: $\forall \ g_1, g_2  \in G \Rightarrow g_1 \bullet g_2 \in G$
+    \item Associatividade: $(g_1 \bullet g_2) \bullet g_3 = g_1 \bullet (g_2 \bullet g_3), \forall \ g_1, g_2, g_3 \in G$
+    \item Existência do elemento neutro: $\exists \ e \in G : g \bullet e = e \bullet g = g$
+    \item Existência dos inversos: $\forall \ g \in G \ \exists \ g^{-1} : g \bullet g^{-1} = g^{-1} \bullet g = e$
 \end{enumerate}
@@
 mesmo. Um grupo é dito comutativo ou abeliano caso a propriedade
-comutativa seja satisfeita para todos os elementos: $\forall g_1, g_2 \in G \Rightarrow g_1 \bullet g_2 = g_2 \bullet g_1$
+comutativa seja satisfeita para todos os elementos:
+$\forall \ g_1, g_2 \in G \Rightarrow g_1 \bullet g_2 = g_2 \bullet g_1$.
 
-Embora as aplicações de simetrias neste caso possam se desenvolver
-de diversas formas com interesse estético no plano sonoro, em
-nossas pesquisas, uma delas se mostrou particularmente coerente
+São numerosas as aplicações de grupos e estes
+usos para as artes possam se desenvolver
+de diversas formas focadas em simetrias
+e com interesse estético, em
+nossas pesquisas uma delas se mostrou particularmente coerente
 e pertinente: as permutações de unidades sonoras. Primeiro,
-através do Teorema de Cayley, todo grupo é isomorfo
+através do Teorema de Cayley, todo grupo finito é isomorfo
 a um grupo de permutação\footnote{Mais especificamente,
@@
 a permutações. Além disso, na música as permutações estão
-no núcleo de diversos procedimentos canônicos
+no núcleo de diversos procedimentos canônicos como a
+retrogradação e outras técnicas de variação.
 
-[Explicar o conceito de simetria e permutações]
 [Expor sobre uma ou duas músicas feitas com o FIGGUS]
+
+No FIGGUS, implementamos o grão sonoro (no caso mais usado
+como unidade) como uma classe que possui apenas
+os seguintes atributos (e não possui funcões): duração (segundos),
+frequência (Hz), timbre (identificador para usar mediante implementações convenientes), intensidade (pico $\in \ [0,1]$), e duração dos fades (in e out em segundos). Aqui a classe:
+
+\code{Grão Sonoro Básico}{python_snippets/fgrao.py}
+
+O FIGGUS funciona com base em uma sequência de grãos posta de antemão
+na qual operam as permutações. Além da sequência de grãos,
+é conveniente também manter o número de grãos da sequência e uma
+lista com os índices dos grãos para executar as operações necessárias para gerarmos as músicas (ver abaixo). A classe utilizada para representar as sequências é assim:
+
+\code{Sequência de Grãos}{python_snippets/fsequencia.py}
+
+Aos grãos em sequência neste ponto são aplicadas permutações.
+Para isso é bastante conveniente representar as permutações em classesseparadas e os padrões de aplicação destas permutações para facilitar
+desenvolvimentos com clareza.
+No caso a classe de permutações só possui os parâmetros da permutação em si e do tamando dela (número de elementos que ela utiliza).
+É conveniente implementar uma tradução da notação cíclica para a notação direta da permutação e é uma das facilidades da implementação por classe que fizemos.
+A classe do padrão de permutação possui também um período de aplicação da permutação, ou seja, de quantas em quantas leituras da sequência que aplicamos a permutação.
+As classes ficaram assim:
+
+\code{Permutações e Padrão de Permutações}{python_snippets/fpermutaçao.py}
 

### 2012-07-18 python_snippets/fgrao.py

@@ -1,0 +1,11 @@
+class UnitGrain:
+	"""Sonic minimum unit, be it a grain or a note
+	
+	Write something about phonons and granular synthesis.
+	(Gabor and Roads)"""
+	def __init__(self,duration=1,freq=220,timbre=0,intensidade=1,fades=0):
+        	self.duration=duration
+	        self.freq=freq
+	        self.timbre=timbre
+	        self.intensidade=intensidade
+	        self.fades=fades

### 2012-07-18 python_snippets/fpermutacoes.py

@@ -1,0 +1,28 @@
+class Permutation:
+    """Simple permutation.
+    
+    see http://en.wikipedia.org/wiki/Permutation for nomenclature
+    
+    If one-line notation, it runs like:
+    permuted=[original[i] for i in Permutation.one_line]
+    
+    If cycle notation:
+    unfold to one-line notation
+    
+    note
+    one_line = (2,3,1) # is equivalent to
+    cycle=(1,2,3)
+    
+    one_line=(2,5,4,3,1) # is equivalent to
+    cycle=(1,2,5)(3,4)"""
+    def __init__(self, one_line=(2,3,1)):
+        self.one_line = one_line
+        self.size=len(one_line)
+        
+
+class PermutationPattern(Permutation):
+    """A Permutation with periodicity/incidence_index information"""
+    period=1
+    def changePeriodicity(self, new_val=1):
+        self.period=new_val
+

### 2012-07-18 python_snippets/fsequencia.py

@@ -1,0 +1,6 @@
+class Sequence:
+    """Sequence of sonic atoms, our 'UnitGrain' classes"""
+    def __init__(self, ordered_unit_grains=[UnitGrain()]*3):
+        self.ordered_unit_grains=ordered_unit_grains
+        self.unit_count=len(ordered_unit_grains)
+        self.active_grains=range(self.unit_count)

### 2012-07-19 devEresultados.tex
pequena alteracao que nao pude continuar por falta de rede
feito com codigo ateh o figgus

@@
 É conveniente implementar uma tradução da notação cíclica para a notação direta da permutação e é uma das facilidades da implementação por classe que fizemos.
-A classe do padrão de permutação possui também um período de aplicação da permutação, ou seja, de quantas em quantas leituras da sequência que aplicamos a permutação.
+A classe do padrão de permutação possui também um período de aplicação da permutação, ou seja, de quantas em quantas leituras da sequência aplicamos a permutação.
 As classes ficaram assim:
 
-\code{Permutações e Padrão de Permutações}{python_snippets/fpermutaçao.py}
+\code{Permutações e Padrões de Permutações}{python_snippets/fpermutacoes.py}
+
+De posse dos grãos, da sequência, das permutações e do padrão de permutações estamos possibilitados de realizar a estrutura musical em si.
+Bastam adicionalmente, especificar o número desejado de iterações da sequência.
+Com isso, a sequência de grãos é lida um número de vezes, aplicando as permutações na sequência segundo o padrão de aplicação das pemutações, de forma a resultar em uma sequência sonora musical.
+Note que se a permutação usar menos elementos que a sequência possui, alguns destes elementos ficarão estáticos nas iterações da sequência no padrão sonoro.
+
+Na síntese do vetor sonoro relacionado, é também crucial estabelecer a taxa de amostragem preterida, como podemos perceber na função de síntese dos vetores, da classe Pattern que cuida desta realização do padrão em si.
+Deixamos a seguir esta função, de forma que faltará apenas a escrita destes vetores sonoros em arquivo de áudio, como veremos a seguir.
+
+\code{Realização do Padrão Musical}{python_snippets/fpadraosonoro.py}
+
+Em posse não apenas de representações abstratas do padrão musical a ser realizado, mas dos próprios vetores sonoros relacionados à representação digital da música a ser realizada, podemos escrever este vetor como um arquivo de áudio propriamente dito.
+O mais conveniente neste caso é escrever um arquivo PCM (Pulse Code Modulation) em padrão amplamente utilizado e reconhecido.
+Ambos os padrões WAV e AIFF satisfazem estes requisitos e escolhemos o WAV por ser o padrão de CDs (Compact Discs) e também o mais utilizado por programas de áudio.
+Mais especificamente, o padrão de CDs é WAV com 44100 amostras por segundo e 16 bits por amostra\footnote{No formato WAV de CD, cada amostra é representada por uma variável inteira 'signed' de 16 bits.}.
+As amostras dos vetores sonoros são calculados, por conveniência e por convenção, no âmbito $[-1,1]$ e precisam ser normalizados para o âmbito $[-32767,32768]$ e truncados em números inteiros.
+Depois disso precisam ser escritos em um arquivo com os bits como na convenção da linguagem C/C++.
+A bibioteca \emph{struct} cuida dessa escrita do inteiro no formato correto, e a biblioteca \emph{wave} escreve o cabeçalho no formato WAV adequado.
+Assim, a clase de escrita do vetor sonoro em arquivo comum fica assim:
+
+\code{Escrita do Vetor Sonoro em Arquivo WAV}{python_snippets/fio.py}
 

### 2012-07-19 python_snippets/fio.py

@@ -1,0 +1,59 @@
+import wave, struct
+class IOUtils:
+    def recordPattern(self,filename,pattern):
+    	self.recordFile(pattern.sonic_vector,[],"sound.wav",pattern.SR)
+	
+    def recordFile(self, sonic_vector=[], sonic_vector2=[],
+                         filename="sound.wav", samplerate=44100):
+    	sound = wave.open(filename,'w')
+    	sound.setframerate(samplerate)
+
+    	sound.setsampwidth(2) # Always 16bit/sample (2 bytes)
+    	if not sonic_vector2:
+    	    sound.setnchannels(1) # Mono
+	        sonic_vector=self.boundVector(sonic_vector)
+
+	        sonic_vector=[i*(2**15-1) for i in sonic_vector]
+	        sound.writeframes(struct.pack('h'*len(sonic_vector),
+                                    *[int(i) for i in sonic_vector]))
+    	else:
+	        sound.setnchannels(2) # stereo
+    	    sonic_vector=self.boundVector(sonic_vector)
+    	    sonic_vector2=self.boundVector(sonic_vector2)
+
+    	    sonic_vector=[i*(2**15-1) for i in sonic_vector]
+    	    sonic_vector2=[i*(2**15-1) for i in sonic_vector2]
+    	    SV=[]
+    	    for i,j in zip(sonic_vector,sonic_vector2):
+        		SV.extend((i,j))
+	        sound.writeframes(struct.pack('h'*len(SV),*SV))
+    	sound.close()
+
+    def boundVector(self,vector):
+ 	 	"""Bound vector in the [-1,1] interval"""
+ 		svmin=min(vector)
+ 		svmax=max(vector)
+ 		ambit=svmax-svmin
+ 		if svmax>1 or svmin<-1:
+ 		    if svmax-svmin > 2:
+  				i=0
+  				for sample in vector:
+  				    new_sample=(sample-svmin)/ambit # results in [0,1]
+  				    new_sample=new_sample*2-1 # results in [-1,1]
+  				    vector[i]=new_sample
+  				    i+=1
+ 		    elif svmax > 1:
+  				offset=svmax-ambit/2
+  				i=0
+  				for sample in vector:
+  				    new_sample=sample - offset
+  				    vector[i]=new_sample
+  				    i+=1
+ 		    elif svmin < -1:
+  				offset=-svmin -ambit/2
+  				i=0
+  				for sample in vector:
[…]

### 2012-07-23 devEresultados.tex
primeiras adequacoes do texto para uma passada menos em codigo computacional

@@
 permutações e de Teoria de Grupos, para a composição de músicas. Como demonstração
-das capacidades do FIGGUS, ele gera um EP (E... Play, pegar significado dele) inteiro com um único comando. Este é o
+das capacidades do FIGGUS, ele gera um EP\footnote{Extended Play, uma album musical maior que um single mas menor que um LP (Long Play) inteiro} com um único comando. Este é o
 \emph{PPEPPS}\footnote{Pure Python EP: Projeto Solvente}, como veremos a seguir.
@@
 de ondas em alta resolução (chamado \emph{lookup table}). O procedimento é barato e de qualidade alta
-(não acrescentam ruídos relevantes ao sinal). A seguir está o procedimento da lookuptable:
+(não acrescentam ruídos relevantes ao sinal).
+
+Para realizarmos a busca de forma eficiente, precisamos de tabelas que contenham cada uma
+uma forma de onda. As tradicionais são a senóide e as ondas quadrada, triangular e dente de serra.
+Mostramos a seguir estas formas de onda e também uma forma de onda retirada de um som real.
+
+De posse destas formas de onda, A seguir está o procedimento da lookuptable:
 

### 2012-08-16 devEresultados.tex
pequena revisao dos escritos e adicionados scripts, figura e anotacao sobre oscilacoes polinomiais

@@
 
-Por razões didáticas, dividimos este capítulo em 4 partes, a saber: música em tempo diferido (ou seja, que não é feito em tempo real),
+Por razões didáticas, dividimos este capítulo em 4 partes: música em tempo diferido (ou seja, que não é feito em tempo real),
 música em tempo real,  música na matéria (suporte físico em hardware) e música no tecido social
@@
 mesmo motivo, trataremos inicialmente de dois trabalhos em tempo diferido.
-Aliás, como veremos a seguir, as abordagens são paradigmáticas para a música
+Aliás, como veremos a seguir, as abordagens completam uma base para a música
 computacional praticada até os dias de hoje.
 
-O primeiro destes feitos, chamado de \emph{minimum-fi}\footnote{hi-fi, low-fi, minimum-fi, ou seja,
+Tratando da síntese sonora musical como fenômeno físico,  sintetiza uma música amostra por amostra, através de princípios
+claros de síntese sonora e organização musical\footnote{Veja no APÊNDICE XX: o \emph{minimum-fi} hi-fi, low-fi, minimum-fi, ou seja,
 o mínimo de qualidade para assegurar existência e consistência
-da criação} sintetiza uma música inteira, amostra por amostra, através de princípios
-claros de síntese sonora e organização musical. O segundo, chamado de \emph{FIGGUS} (FInite
+da criação}. O segundo, chamado de \emph{FIGGUS} (FInite
 Groups in Granular and Unit Synthesis), utiliza os princípios do minimum-fi e constitui

### 2012-08-17 devEresultados.tex
inicio da transicao de linguagem e ideia

@@
 
-Tratando da síntese sonora musical como fenômeno físico,  sintetiza uma música amostra por amostra, através de princípios
+O som musical pode ser caracterizado físicamente, e sua síntese digital com base em nossa banda de audição e no teorema de Nyquist, sintetiza uma música amostra por amostra, através de princípios
 claros de síntese sonora e organização musical\footnote{Veja no APÊNDICE XX: o \emph{minimum-fi} hi-fi, low-fi, minimum-fi, ou seja,

### 2012-08-17 devEresultados.tex
iniciado novo texto sobre musica no audio digital, psicofisica e sequencias matematicas

@@
 
-\section{Áudio e Música}
+\section{Música no Áudio Digital}
+
+A música é tradicionalmente pensada através de átomos chamados notas. Em sua concepção
+simples, estas notas possuem ao menos duração, volume, altura e timbre. Embora estas
+sejam características psicofísicas, podemos lidar com elas através de suas características
+fisicamente mensuráveis.
+
+Em primeiro lugar, estamos tratando diretamente da representação digital do som. Isso constitui em uma
+sequência de amostras igualmente espaçadas no tempo. Desta forma, definimos a
+taxa de amostragem $t_a$ como o número de amostras por segundo do áudio digital.
+A sequência $T_i={t_i}$ possui as amostras separadas por $\delta_a=1/t_a$ segundos.
+
+Desta forma, visando uma nota musical de duração $\Delta$,
+teremos uma sequência de $ \Delta . t_a $ amostras:
+
+\begin{equation}
+T_{i}^{\Delta}={\{t_i\}}_{0}^{\Delta . t_a - 1}
+\end{equation}
+
+O volume depende da reverberação e distribuição dos harmônicos - dentre outras medidas psicofísicas~\cite{chowningVolume} - mas num primeiro momento podemos lidar com ele de forma prática e geral através da potência da onda:
+
+\begin{equation}
+pot(T_i^{\Delta})=\frac{\sum_{i=0}^{\Delta t_a -1} t_i^2}{\Delta t_a}
+\end{equation}
+
+Como o volume final dependerá da amplificação do sinal nos falantes de quem ouvir, o que conta mais é a potência relativa de uma nota em relação às outras ou de um trecho da música em relação ao resto.
+Lembrando que, devido à nossa percepção logarítmica, a redução para uma mesma fração é sentido como a mesma diminuição do volume.
+
+\section{Áudio e Música (deprecated)}
 

### 2012-08-18 devEresultados.tex
descrito matematicamente as propriedades fisicas da nota musical do som digitalizado, jah adentrando timbre

--- 2012-08-17 devEresultados.tex
@@
 
-\section{Áudio e Música}
+\section{Música no Som Digitalizado}
+
+A música é tradicionalmente pensada através de átomos chamados notas. Em sua concepção
+simples, estas notas possuem ao menos duração, volume, altura e timbre. Embora estas
+sejam características psicofísicas, podemos lidar com elas através de suas características
+fisicamente mensuráveis.
+
+Em primeiro lugar, estamos tratando diretamente da representação digital do som. Isso em uma
+sequência de amostras igualmente espaçadas no tempo. Desta forma, definimos a
+frequência (ou taxa) de amostragem $f_a$ como o número de amostras por segundo do áudio digital.
+A sequência $T_i={t_i}$ possui as amostras separadas por $\delta_a=1/f_a$ segundos.
+
+Desta forma, visando uma nota musical de duração $\Delta$,
+teremos uma sequência de $ \Delta . f_a $ amostras\footnote{Observe o abuso de notação:
+o limite superior de uma sequência pertence aos naturais $ \mathbb{N} $, mas $ \Delta . f_a $
+só satisfaz esta condição em casos muito excepcionais. Procederemos assim por simplicidade. 
+Explicitaremos em texto quando for crucial o procedimento relativo à parte fracionária dos índices
+ou de seus limites.}:
+
+\begin{equation}
+T_{i}^{\Delta}={\{t_i\}}_{i=0}^{\Delta . f_a - 1}
+\end{equation}
+
+O volume depende da reverberação e distribuição dos harmônicos - dentre outras medidas psicofísicas~\cite{chowningVolume} - mas num primeiro momento podemos lidar com ele de forma prática e geral através da potência da onda:
+
+\begin{equation}
+pot(T_i^{\Delta})=\frac{\sum_{i=0}^{\Delta . f_a -1} t_i^2}{\Delta . f_a}
+\end{equation}
+
+Como o volume final dependerá da amplificação do sinal nos falantes, o mais importante é a potência relativa de uma nota em relação às outras ou de um trecho da música em relação ao resto.
+As diferenças de volume são medidas em decibéis, que são
+calculadas diretamente com as amplitudes, energias ou potências\footnote{Lembrando que, devido à nossa percepção logarítmica,
+em um som de volume $v$ a redução da potência $p$ para uma mesma fração $\nu . p $ 
+com $\nu \in [0,1]$ é sentido como a mesma diminuição do volume $v-\kappa$ com $\kappa \geq 0$.}:
+
+\begin{equation}\label{decibeis}
+V_{dB}=10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)}
+\end{equation}
+
+A quantidade adimensional é denotada pela unidade decibéis ($dB$). A cada 10 $dB$s atribuímos canonicamente
+a sensação de "volume dobrado". Importantes resultados de referência são os equivalente em $dB$s de se dobrar
+a amplitude ou a potência (este último equivale a dobrar a energia do sinal):
+
+\begin{equation*}
+se \quad  t_i^{'}=2 . t_i \quad \Rightarrow \quad pot(T^{'}_i)=4 . pot(T_i) \quad \Rightarrow \quad V^{'}_{dB}=10log_{10} 4 \quad  \approx \quad 6 dB
+\end{equation*}
+\begin{equation*}
+se \quad pot(T^{'}_i)=2 pot(T_i) \quad \Rightarrow \quad V^{'}_{dB}=10log_{10} 2 \quad \approx \quad 3 dB
+\end{equation*}
+
+E o quanto se deve amplificar uma sequência para que tenha o volume dobrado ($10dB$s a mais):
+
+\begin{gather}
+10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)} = 10 \quad \Rightarrow \quad \sum_{i=0}^{\Delta.f_a-1}t^{'2}_i=10\sum_{i=0}^{\Delta.f_a-1}t_i^2=\sum_{i=0}^{\Delta.f_a-1}(\sqrt{10}.t_i)^2 \\
+\therefore \quad t^{'}_i=\sqrt{10}t_i \quad \Rightarrow \quad t^{'}_i \approx 3,16t_i
+\end{gather}
+
+Estes valores servem de referência para os aumentos e dominuições dos valores absolutos que compõem as
+sequências de amostras sonoras para atingir efeitos musicais.
+
+Neste ponto, correspondendo a uma partícula musical (uma nota), estabelecemos uma sequência cuja
+duração e intensidade conseguimos controlar\footnote{Basta alterarmos o tamanho da sequência para alterarmos
+a duração ou multiplicarmos as os elementos da sequência por algum valor normalizador}.
+
+A altura é especificada pela frequência fundamental. A frequência fundamental $f$ especifica uma periodicidade temporal
+com ciclo de duração $\delta_f = 1/f$. Esta, multiplicada pela taxa de amostragem $f_a$ resulta no número de amostras
+do ciclo $\lambda_f=\delta_f . f_a$. Resulta evidente que, sendo $T_i^f$ a sequência sonora de frequência fundamental $f$:
+
+\begin{equation}\label{periodicidade}
+     T^f_i=\{ t_i^f \}=\{ t^f_{i+\lambda_f}  \}= \{ t^f_{i+\frac{f_a}{f}} \}
+\end{equation}
+
+Enquanto o período da onda corresponde a uma frequência fundamental, o percurso
+da onda sonora dentro do período define um espectro harmônico e portanto
+um timbre\footnote{O timbre é uma característica subjetiva e complexa. Fisicamente,
+o timbre é multidimensional e dado pelo comportamento temporalmente dinâmico
+de energias em componentes harmônicas e ruidosas. Usamos aqui o resultado de que espectros
+sonoros com diferenças mínimas resultam em timbres diferentes. Vale salientar que nem tudo
+o que se atribui ao timbre se acha manifesto em diferenças espectrais. Uma mesma nota
+possui diferentes matizes espectrais, um mesmo instrumento possui diferentes timbres. Mesmo
+assim se fala que uma nota e um instrumento possuem um timbre. Por último, até
+aspectos culturais ou circunstanciais alteram nossa percepção timbrística.
+}.
+
+O caso mais simples é do espectro que consiste somente
+de sua própria fundamental $f$. Este é o caso da senóide, frequência em movimento oscilatório puro chamado
+movimento harmônico simples. Seja $S_i^f$ uma sequência cujas amostras
+$s_i^f$ descrevem uma senóide de frequência $f$:
+
+\begin{equation}\label{senoide}
+     S^f_i=\{ s^f_i \}=\Bigl\{ \sin\bigl(\frac{i}{\lambda_f} . 2\pi\bigr)  \Bigr\}
+\end{equation}
+
+Onde $\lambda_f=f_a/f$ é o número de amostras do período.
+
+\section{Áudio e Música (deprecated)}
 

### 2012-08-18 devEresultados.tex
feito ateh frequencia fundamental e formas de onda adiantado

@@
 \begin{equation}\label{senoide}
-     S^f_i=\{ s^f_i \}=\Bigl\{ \sin\bigl(\frac{i}{\lambda_f} . 2\pi\bigr)  \Bigr\}
+     S^f_i=\{ s^f_i \}=\Bigl\{ \sin\bigl(2\pi \frac{i}{\lambda_f} \bigr)  \Bigr\} = \Bigl\{ \sin\bigl(2\pi f \frac{i}{f_a}\bigr)  \Bigr\} 
 \end{equation}
 
-Onde $\lambda_f=f_a/f$ é o número de amostras do período.
+Onde $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\delta_a} \;$ é o número de amostras do período.
+
+De forma semelhante, outras formas de onda são utilizadas na música por suas qualidades
+espectrais e simplicidade.
+
+\begin{equation}\label{denteDeSerra}
+     D^f_i=\{ d^f_i \}=\Bigl\{ 2\frac{i\,\%\lambda_f}{\lambda_f} -1 \Bigr\}
+\end{equation}
+
+\begin{equation}\label{triangular}
+     T^f_i=\{ t^f_i \}=\Bigl\{1- \Bigl| 2 - 4\frac{i\,\%\lambda_f}{\lambda_f} \Bigr| \Bigr\}
+\end{equation}
+
+\begin{equation}\label{quadrada}
+     Q^f_i=\{ q^f_i \}= \left\{
+         \begin{array}{l l}
+              1 & \quad \text{para} \; \; (i\,\%\lambda_f)   <  \lambda_f /2  \\
+              -1 & \quad \text{caso contrário}\\
+         \end{array} \right.
+\end{equation}
+
+Qualquer outra forma de onda é composta de proporções únicas dentre as componentes harmônicas.
+As mais utilizadas estão abaixo na figura:
+
+\begin{figure}[h!]
+    \centering
+    \caption{Formas de onda musicais básicas}
+        \includegraphics[width=\textwidth]{figuras/formasDeOnda3b}
+\end{figure}
 

### 2012-08-18 devEresultados.tex
adiantado ateh som real. No ponto das visualizacao dos espectros, sequencias de notas e usos musicais

@@
 De forma semelhante, outras formas de onda são utilizadas na música por suas qualidades
-espectrais e simplicidade.
+espectrais e simplicidade. A senóide é um ponto isolado no espectro, estas formas
+de onda apresentam cadeias de componentes harmônicos e usos específicos.
 
+A dente de serra apresenta todos os componentes da série
+harmônica com energia decrescente de $-6dB/oitava$:
 \begin{equation}\label{denteDeSerra}
@@
 
+A forma de onda triangular apresenta somente os harmonicos ímpares caindo a $-12dB/oitava$:
 \begin{equation}\label{triangular}
@@
 
+A onda quadrada apresenta os harmônicos ímpares caindo a $-6dB/oitava$:
 \begin{equation}\label{quadrada}
@@
 
-Qualquer outra forma de onda é composta de proporções únicas dentre as componentes harmônicas.
-As mais utilizadas estão abaixo na figura:
+A dente de serra é um ponto de partida comum para a síntese subtrativa pois possui
+ambos os harmônicos pares e ímpares e em grande quantidade (alguma filtragem atenuante
+é útil nos médios e agudos para que o som ganhe naturalidade e fique mais agradável).
+Por este mesmo motivo, os harmônicos atenuados da onda triangular
+a faz a mais funcional - dentre as citadas - para ser usada sem nenhum tratamento.
+
+Para citar também um uso possível da onda quadrada, ela pode ser usada em uma síntese
+subtrativa que vise imitar um clarinete. Este instrumento também só apresenta os
+componentes ímpares e o espectro da onda quadrada pode convir com sua energia abundante
+nas altas energias.
 
@@
         \includegraphics[width=\textwidth]{figuras/formasDeOnda3b}
+        \label{fig:formasDeOnda}
 \end{figure}
+
+Qualquer outra forma de onda é composta de proporções específicas de componentes harmônicas.
+Quanto maior a curvatura do trecho na forma de onda, maior a contribuição para a
+concentração de energia nos harmônicos agudos.
+
+A figura ~\ref{fig:formasDeOnda} apresenta
+as formas de onda descritas nas equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} para $\lambda_f=100$ (período
+de $100$ amostras).
+ Se $t_a=44,1 kHz$, como em padrões PCM e de Compact Disks, a onda possui frequência fundamental $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz $.
+Um lá\footnote{Um lá 4, logo acima do dó central, no segundo espaço do pentagrama na clave de sol comum.}, seja qual for a forma de onda.
+
+A onda 'som real' é um período extraído de um som real relativamente comportado.
+Ele possuí $\lambda_f=114$ amostras\footnote{Caso também utilizado diretamente em $f_a=44,1kHz$, é um pouco mais grave que os $441 Hz$ das formas de onda artificiais.}.
+Note que, definindo $ R_i=\{ r_i \}_0^{\lambda_f-1}$ a sequência de amostras do som real da figura ~\ref{fig:formasDeOnda}.
+$R_i$ pode ser tomado como base para um som $T_i^f$ da seguinte forma:
+
+\begin{equation}\label{sampleandoFormaDeOnda}
+     T^f_i=\{ t_i^f \}=\Bigl\{ r_{i\%\lambda_{f}} \Bigr\}
+\end{equation}
+
+O som resultante possui o espectro momentâneo do som original. Por ser repetido de forma identica,
+seu espectro é perfeitamente harmônico, sem os ruídos e variações típicas do fenômeno natural.
+
+[plot dos espectros]
+
+[resumo das sequencias com duracao, frequencia, forma de onda e intensidade]
+
+[usos musicais]
 

### 2012-08-19 scripts/espectros_oboe.py
adicionadas figuras comparativas de espectro, ainda ajeitando

@@ -1,0 +1,90 @@
+#-*- coding: utf8 -*-
+# http://matplotlib.sourceforge.net/examples/api/legend_demo.html
+# 
+import pylab as p, numpy as n, scikits.audiolab as a
+
+# uma nota inteira de oboe; 150529 amostras
+oboe=a.wavread("22686__acclivity__oboe-a-440.wav")[0]
+oboe=oboe[:,0]+oboe[:,1] # estereo para mono
+
+# um periodo da nota (recortado no audacity buscando zeros)
+# nao se repete o zero, por isso o [:-1], resulta 98 amostras
+poboe=a.wavread("22686__acclivity__oboe-a-440_periodo.wav")[0][:-1]
+# fazendo um sinal do mesmo tamanho:
+poboe2=n.array(list(poboe)*1510) # 147980 
+#poboe2=n.array(list(poboe)*1536) # len(oboe)/len(poboe) nao bate tanto pois pegamos uma frequencia relativamente baixa == periodo grande 
+#poboe2=poboe[n.arange(oboe.shape[0])%poboe.shape[0]] # Descobrir pq nao funciona
+
+
+o_s=n.fft.fft(oboe)
+o_a=n.abs(o_s)
+
+p_s=n.fft.fft(poboe2)
+p_a=n.abs(p_s)
+
+
+
+
+i=n.arange(poboe2.shape[0])
+foo=(p_a>50).nonzero()
+p.plot(i[foo],p_a[foo],"o",label=u"período amostrado")
+ii=list(i[foo])
+
+i=n.arange(oboe.shape[0])
+p.plot(i,o_a,label=u"nota de oboé natural")
+p.legend(loc="upper right")
+
+ticks=[]
+ticks=[r"$f%i$" % (i,) for i in xrange(len(ii))]
+
+p.yticks((0,11000),(0,'11k'))
+p.xticks([0] + ii + [22000],[0] + ticks + ["22k"])
+
+p.xlim(0,22000,)
+p.ylim(-300,11000)
+p.ylabel(r'valor absoluto $\sqrt{a^2+b^2}$ da componente complexa $\rightarrow$')
+p.xlabel(u'componente do espectro em frequência' + r'$\rightarrow$')
+p.show()
+
+
+#s_=senoide[indices%T]
+#s_s=n.fft.fft(s_)
[…]

### 2012-08-20 devEresultados.tex
adicionadas figuras comparativas de espectro, ainda ajeitando
melhoradas figuras e adentrado espectro sonoro

@@
     \caption{Formas de onda musicais básicas}
-        \includegraphics[width=\textwidth]{figuras/formasDeOnda3b}
+        \includegraphics[width=\textwidth]{figuras/formasDeOnda5}
         \label{fig:formasDeOnda}
@@
 Se $t_a=44,1 kHz$, como em padrões PCM e de Compact Disks, a onda possui frequência fundamental $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz $.
-Um lá\footnote{Um lá 4, logo acima do dó central, no segundo espaço do pentagrama na clave de sol comum.}, seja qual for a forma de onda.
+Um lá\footnote{Um lá 4, logo acima do dó central, no segundo espaço do pentagrama na clave de sol comum.}, seja qual for a forma de onda dentre as artificiais apresentadas.
 
-A onda 'som real' é um período extraído de um som real relativamente comportado.
-Ele possuí $\lambda_f=114$ amostras\footnote{Caso também utilizado diretamente em $f_a=44,1kHz$, é um pouco mais grave que os $441 Hz$ das formas de onda artificiais.}.
-Note que, definindo $ R_i=\{ r_i \}_0^{\lambda_f-1}$ a sequência de amostras do som real da figura ~\ref{fig:formasDeOnda}.
+Estas formas de onda possuem usos especiais na música e seus espectros estão dispostos na figura ~\ref{fig:espectroDeOndas}.
+É importante notar as componentes isoladas e exatamente harmônicas dos espectros.
+A senóide consiste de um nódulo único no espectro, frequência pura.
+A dente de serra é a únida com a série harmônica completa (pares e ímpares).
+Já as ondas triangular e quadrada possuem as mesmas componentes espectrais, mas com decaimentos de $-12dB/oitava$ e $-6dB/oitava$.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Espectros das ondas sonoras musicais artificiais básicas}
+        \includegraphics[width=\textwidth]{figuras/espectroDeOndas5}
+        \label{fig:espectroDeOndas}
+\end{figure}
+
+O espectro harmônico é formado pelas frequências múltiplas da frequência fundamental $f_n=n.f_0$.
+Como nossa percepção segue a escala logarítmica, o espectro possui notas diferentes da frequência fundamental.
+Além disso, o número de harmônicos será limitado pela frequência máxima $t_a/2$.
+Musicalmente crucial aqui é internalizar que se um sinal real foi decomposto por fourier e possui uma componente em uma frequência $fn$ com alguma energia
+   \footnote{A energia total equivale à soma dos quadrados das amplitudes como as de pressao/voltagem pelo tempo,e.g. ~\ref{fig:formasDeOnda}.
+As componentes espectrais - e.g. ~\ref{fig:espectroDeOndas} e ~\ref{fig:espectroOboe} -
+também podem ser elevadas ao quadrado para resultar em quantificação de energia e somam-se na energia total. As energias se equivalem pelo teorema de Parseval.},
+então existe uma contribuição senoidal, puramente harmônica no som.
+Esta energia concentrada especificamente na frequência $f_n$ é separado pelo ouvido\footnote{A cognição cerebral se dá basicamente com estas componentes.
+Muita informação é obtida pelo processamento destes sinais acústicos por diversas diversas formas de vida e se baseia em modulações de frequência e amplitude.}
+
+A onda rotulada como 'som real amostrado' na figura ~\ref{fig:formasDeOnda} é um período extraído de um som real relativamente comportado.
+Ele possuí $\lambda_f=114$ amostras\footnote{Caso também utilizado diretamente em $f_a=44,1kHz$, é um pouco mais grave que os $441 Hz$ das formas de onda artificiais $\frac{44100}{114}=385,84Hz$.}.
+A onda de oboé foi amostrada de um lá 4 também em $44,1kHz$.
+O período escolhido para a amostragem é relativamente curto, com 98 amostras corresponde a $\frac{44100}{98}=450 Hz$.
+Compararemos abaixo o conteúdo espectral da nota original com o resultante da utilização destas amostras.
+
+Note que, definindo $ R_i=\{ r_i \}_0^{\lambda_f-1}$ a sequência de amostras do som real da figura ~\ref{fig:formasDeOnda},
 $R_i$ pode ser tomado como base para um som $T_i^f$ da seguinte forma:
@@
 \begin{equation}\label{sampleandoFormaDeOnda}
-     T^f_i=\{ t_i^f \}=\Bigl\{ r_{i\%\lambda_{f}} \Bigr\}
+     T^f_i=\{ t_i^f \}=\Bigl\{ r_{(i\,\%\lambda_{f})} \Bigr\}
 \end{equation}
@@
 O som resultante possui o espectro momentâneo do som original. Por ser repetido de forma identica,
-seu espectro é perfeitamente harmônico, sem os ruídos e variações típicas do fenômeno natural.
+seu espectro é perfeitamente harmônico, sem os ruídos e variações típicas do fenômeno natural. Isso pode ser visto claramente na figura ~\ref{fig:espectroOboe} onde a nota original do oboé e uma nota 
+artificial - de mesma duração e cujas amostras consistem no mesmo período da figura ~\ref{fig:formasDeOnda} - estão dispostas juntamente.
+O espectro natural possui variações na frequência dos harmônicos, nas suas intensidades e uma quantidade de ruído.
+Já a nota cujo período foi amostrado possui espectro perfeitamente harmônico.
 
-[plot dos espectros]
+\begin{figure}[h!]
+    \centering
+    \caption{Espectros das ondas sonoras do oboé natural e de período amostrado}
+        \includegraphics[width=\textwidth]{figuras/espectroOboeAmostradoNatural2}
+        \label{fig:espectroOboe}
+\end{figure}
 

### 2012-08-20 scripts/formas_de_onda.py

@@ -1,7 +1,7 @@
 #-*- coding: utf8 -*-
 # http://matplotlib.sourceforge.net/examples/api/legend_demo.html
 # 
-import pylab as p, numpy as n, scikits.audiolab as a
+import pylab as p, numpy as n, scikits.audiolab as a, matplotlib as m
 
 T=100
 #T=1024
@@ -12,7 +12,10 @@ triangular=n.hstack ((n.linspace(-1,1,T/2,endpoint=False),n.linspace(1,-1,T/2,en
 quadrada=n.hstack ((n.ones(T/2),n.ones(T/2)* -1))
 
 onda=a.wavread("ondaReal.wav")[0] # onda real
-onda=((onda-onda.min())/(onda.max()-onda.min()))*2-1
+onda=((onda-onda.min())/(onda.max()-onda.min()))*2-1 # normalizando
+onda2=a.wavread("22686__acclivity__oboe-a-440_periodo.wav")[0] # periodo do oboe
+onda2=((onda2-onda2.min())/(onda2.max()-onda2.min()))*2-1 # normalizando
+
 
 p.subplot(211)
 #p.title("Formas de onda")
@@ -20,7 +23,6 @@ p.plot(senoide,'o',linewidth=3, label=u"senóide")
 p.plot(dente,'o',linewidth=3, label="dente de serra")
 p.plot(triangular,'o',linewidth=3, label="triangular")
 p.plot(quadrada,'o',linewidth=3, label="quadrada")
-p.plot([0],"yo",label="som real (abaixo)")
 p.xlim(0,T)
 p.ylim(-1.1,1.1)
 
@@ -39,15 +41,39 @@ frame.set_facecolor('0.80')
 for ll in l.get_lines():
         ll.set_linewidth(1.5)  # the legend line width
 #p.legend()
+p.ylabel(r'amplitude sintetizada')
 
-p.subplot(212)
-p.plot(onda,"yo",linewidth=3)
-p.plot(onda,"y",linewidth=3)
+fig=p.subplot(212)
+i=n.arange(len(onda))
+p.plot(i,onda,"yo",linewidth=3, label="som real amostrado")
+p.plot(i,onda,"y",linewidth=3)
+i=n.linspace(0,len(onda),len(onda2),endpoint=False)
+p.plot(i, onda2,"ko", linewidth=3, label=u"período de um oboé em 440Hz")
+p.plot(i, onda2,"k", linewidth=3)
 p.xlim(0,onda.shape[0])
[…]

### 2012-08-20 devEresultados.tex
andado com a discussao das frequencias nas amostras

@@
 Além disso, o número de harmônicos será limitado pela frequência máxima $t_a/2$.
-Musicalmente crucial aqui é internalizar que se um sinal real foi decomposto por fourier e possui uma componente em uma frequência $fn$ com alguma energia
+Musicalmente crucial aqui é internalizar que se um sinal real foi decomposto por fourier e possui uma componente em uma frequência $f_n$ com alguma energia
    \footnote{A energia total equivale à soma dos quadrados das amplitudes como as de pressao/voltagem pelo tempo,e.g. ~\ref{fig:formasDeOnda}.
@@
 também podem ser elevadas ao quadrado para resultar em quantificação de energia e somam-se na energia total. As energias se equivalem pelo teorema de Parseval.},
-então existe uma contribuição senoidal, puramente harmônica no som.
-Esta energia concentrada especificamente na frequência $f_n$ é separado pelo ouvido\footnote{A cognição cerebral se dá basicamente com estas componentes.
-Muita informação é obtida pelo processamento destes sinais acústicos por diversas diversas formas de vida e se baseia em modulações de frequência e amplitude.}
+então existe uma contribuição senoidal, puramente harmônica no som. Esta energia concentrada especificamente na frequência $f_n$ é separada
+ pelo ouvido para adentrar em um nível cognitivo de processamento\footnote{Esta separação em frequência é realizada por diversas espécies através de mecanismos similares à cóclea humana.}.
+Estas componentes são geralmente as principais responsáveis pela qualidade chamada timbre e, caso não apresentem proporções harmônicas (relações de pequenos números), são ruídos, i.e.
+não notas com frequência fundamental como o lá 4 ($440 Hz$) de nossos exemplos.
+Além disso, nossa noção de altura em um complexo sonoro é baseada na semelhança do espectro com a série harmônica.
+
+Podemos ver abaixo como se dá isso na representação discreta.
+O mínimo necessário para representar uma oscilação são 2 amostras, resultando na frequência de Nyquist $f_{Ny}=\frac{t_a}{2}$.
+Todas as sequências fixas de apenas 3 amostras também representam
+somente 1 frequência, pois sua primeira harmônica usaria 1,5 amostras e ultrapassa o limite inferior de 2 amostras mínimas.
+Com 4 amostras, podemos representar 1 ou 2 frequências em proporções diferentes.
+A figura ~\ref{fig:amostras4} mostra estas frequências sendo realizadas compondo as formas de onda básicas e se remontando.
+
+A figura ~\ref{fig:amostras8} mostra as decomposições senoidais para o caso de 8 amostras.
+\begin{figure}[h!]
+    \centering
+    \caption{Frequência máxima em qualquer $t_a$}
+        \includegraphics[width=\textwidth]{figuras/amostras2}
+        \label{fig:amostras2}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{Frequência única em quaisquer 3 amostras}
+        \includegraphics[width=\textwidth]{figuras/amostras3}
+        \label{fig:amostras2}
+\end{figure}
 

### 2012-08-20 scripts/amostras3.py

@@ -1,0 +1,29 @@
+# abs: 0.889857529063, a1: -0.878235820903, b1: -0.143346659928, fas: 0.163221149167A
+import pylab as p, numpy as n
+f=n.fft.fft
+n3=n.array([-0.66522536,  0.29577169,  0.13024923])
+
+ff=f(n3)
+
+a0=n.real(ff[0])
+b0=n.imag(ff[0])
+
+ab1=n.abs(ff[1])
+a1=n.real(ff[1])
+b1=n.imag(ff[1])
+#fas=b1/a1
+fas=n.arctan(a1/b1) # fas/2 = 0.70450095568094961
+print("abs: %s, a1: %s, b1: %s, fas: %s" % (ab1,a1,b1,fas))
+
+ii=n.linspace(-0.5*n.pi,1.5*n.pi,200)
+iii=n.linspace(-0.5,2.5,200)
+#s=(2/3.)*-ab1*n.cos(ii-fas+.5+.3+.05)+a0/3; p.plot(iii,s,"m--"); p.plot(n3,"bo");p.show()
+
+p.ylabel(r"amplitude $\rightarrow$")
+p.xlabel(r"tempo $\rightarrow$")
+s=(2/3.)*-ab1*n.cos(ii+fas/2)+a0/3; p.plot(iii,s,"m--"); p.plot(n3,"bo");p.show()
+
+#s=(2/3.)*-ab1*n.cos(ii+0.68677885598324417)+a0/3; p.plot(iii,s,"m--"); p.plot(n3,"bo");p.show()
+#s=.67*-ab1*n.cos(ii-fas+.5+.3+.05)-.1+.02; p.plot(iii,s,"m--"); p.plot(n3,"bo");p.show()
+#p.plot(iii,s,"m--")
+#p.plot(n3,"bo")

### 2012-08-21 devEresultados.tex
arrumada a sessao da musica no audio digital

@@
 
+\subsection{Caracterização da nota musical}
 A música é tradicionalmente pensada através de átomos chamados notas. Em sua concepção
@@
 fisicamente mensuráveis.
-
 Em primeiro lugar, estamos tratando diretamente da representação digital do som. Isso em uma
 sequência de amostras igualmente espaçadas no tempo. Desta forma, definimos a
-frequência (ou taxa) de amostragem $f_a$ como o número de amostras por segundo do áudio digital.
-A sequência $T_i={t_i}$ possui as amostras separadas por $\delta_a=1/f_a$ segundos.
-
+frequência (ou taxa) de amostragem $f_a$ como o número de amostras por segundo do áudio digital e
+a sequência $T_i=\{t_i\}$ como um conjunto ordenado de amostras reais separadas por $\delta_a=1/f_a$ segundos.
+
+\subsubsection{Duração}
 Desta forma, visando uma nota musical de duração $\Delta$,
 teremos uma sequência de $ \Delta . f_a $ amostras\footnote{Observe o abuso de notação:
-o limite superior de uma sequência pertence aos naturais $ \mathbb{N} $, mas $ \Delta . f_a $
+o limite superior de uma sequência é um número natural, mas $ \Delta . f_a $
 só satisfaz esta condição em casos muito excepcionais. Procederemos assim por simplicidade. 
@@
 
+\subsubsection{Volume}
 O volume depende da reverberação e distribuição dos harmônicos - dentre outras medidas psicofísicas~\cite{chowningVolume} - mas num primeiro momento podemos lidar com ele de forma prática e geral através da potência da onda:
@@
 Como o volume final dependerá da amplificação do sinal nos falantes, o mais importante é a potência relativa de uma nota em relação às outras ou de um trecho da música em relação ao resto.
-As diferenças de volume são medidas em decibéis, que são
+As diferenças de volume são medidas em decibeis, que são
 calculadas diretamente com as amplitudes, energias ou potências\footnote{Lembrando que, devido à nossa percepção logarítmica,
@@
 
-A quantidade adimensional é denotada pela unidade decibéis ($dB$). A cada 10 $dB$s atribuímos canonicamente
+A quantidade adimensional é denotada pela unidade decibeis ($dB$). A cada 10 $dB$s atribuímos canonicamente
 a sensação de "volume dobrado". Importantes resultados de referência são os equivalente em $dB$s de se dobrar
-a amplitude ou a potência (este último equivale a dobrar a energia do sinal):
+a amplitude ou a potência:
 
@@
 
-E o quanto se deve amplificar uma sequência para que tenha o volume dobrado ($10dB$s a mais):
+Outro valor de referência é a amplificação necessária para que uma sequência tenha o volume dobrado ($10dB$s a mais):
 
@@
 
-Estes valores servem de referência para os aumentos e dominuições dos valores absolutos que compõem as
-sequências de amostras sonoras para atingir efeitos musicais.
+Estes valores servem de guia para os aumentos e diminuições dos valores absolutos que compõem as
+sequências de amostras sonoras que visam atingir efeitos musicais.
 
@@
 
-A altura é especificada pela frequência fundamental. A frequência fundamental $f$ especifica uma periodicidade temporal
-com ciclo de duração $\delta_f = 1/f$. Esta, multiplicada pela taxa de amostragem $f_a$ resulta no número de amostras
-do ciclo $\lambda_f=\delta_f . f_a$. Resulta evidente que, sendo $T_i^f$ a sequência sonora de frequência fundamental $f$:
+\subsubsection{Altura\footnote{Com ganho de simplicidade e sem perda de generalidade, considere o índice $i=a/b$ como o inteiro mais próximo da divisão $a/b$.}}
+
+A altura é especificada pela frequência fundamental. A frequência fundamental $f_0$ especifica uma periodicidade temporal
+com ciclo de duração $\delta_{f_0} = 1/f_0$. Esta duração multiplicada pela taxa de amostragem $f_a$ resulta no número de amostras
+do ciclo $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$. Resulta evidente que, sendo $T_i^f$ a sequência sonora de frequência fundamental $f=f_0$:
 
 \begin{equation}\label{periodicidade}
-     T^f_i=\{ t_i^f \}=\{ t^f_{i+\lambda_f}  \}= \{ t^f_{i+\frac{f_a}{f}} \}
-\end{equation}
-
+     T^f_i=\{ t_i^f \}=\{ t^f_{i+\lambda_{f}}  \}= \{ t^f_{i+\frac{f_a}{f}} \}
+\end{equation}
+
+\subsubsection{Timbre}
 Enquanto o período da onda corresponde a uma frequência fundamental, o percurso
@@
 o timbre é multidimensional e dado pelo comportamento temporalmente dinâmico
-de energias em componentes harmônicas e ruidosas. Usamos aqui o resultado de que espectros
-sonoros com diferenças mínimas resultam em timbres diferentes. Vale salientar que nem tudo
+de energias em componentes espectrais tanto harmônicas quanto ruidosas.  Vale salientar que nem tudo
 o que se atribui ao timbre se acha manifesto em diferenças espectrais. Uma mesma nota
@@
 aspectos culturais ou circunstanciais alteram nossa percepção timbrística.
+Na música, espectros
+sonoros com diferenças mínimas resultam em
+ timbres com diferenças expressivamente cruciais.
 }.
 
-O caso mais simples é do espectro que consiste somente
+O caso mais simples (e mais importante, como veremos a diante) é o do espectro que consiste somente
 de sua própria fundamental $f$. Este é o caso da senóide, frequência em movimento oscilatório puro chamado
@@
 
-Onde $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\delta_a} \;$ é o número de amostras do período.
+Onde $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\delta_a} \;$ é o número de amostras do período\footnote{Neste ponto já se tem toda a base para música "reduced-fi", ver APÊNDICE ~\ref{APENDICE XX}}.
 
 De forma semelhante, outras formas de onda são utilizadas na música por suas qualidades
-espectrais e simplicidade. A senóide é um ponto isolado no espectro, estas formas
+espectrais e simplicidade. Enquanto a senóide é um ponto isolado no espectro, estas formas
 de onda apresentam cadeias de componentes harmônicos e usos específicos.
+As formas de onda especificadas em ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} são dispostas graficamente em ~\ref{fig:formasDeOnda}. 
+São as formas de onda artificiais tradicionalmente utilizadas na música para síntese e controle oscilatório de variáveis e apresentam diversos usos fora da música.
 
@@
 
-A forma de onda triangular apresenta somente os harmonicos ímpares caindo a $-12dB/oitava$:
+A forma de onda triangular apresenta somente os harmônicos ímpares caindo a $-12dB/oitava$:
 \begin{equation}\label{triangular}
@@
 
-A onda quadrada apresenta os harmônicos ímpares caindo a $-6dB/oitava$:
+A onda quadrada apresenta somente os harmônicos ímpares caindo a $-6dB/oitava$:
 \begin{equation}\label{quadrada}
@@
 subtrativa que vise imitar um clarinete. Este instrumento também só apresenta os
-componentes ímpares e o espectro da onda quadrada pode convir com sua energia abundante
-nas altas energias.
+componentes ímpares do espectro harmônico e a onda quadrada pode convir com sua energia abundante nas altas frequências.
 
@@
 
-Qualquer outra forma de onda é composta de proporções específicas de componentes harmônicas.
-Quanto maior a curvatura do trecho na forma de onda, maior a contribuição para a
-concentração de energia nos harmônicos agudos.
-
 A figura ~\ref{fig:formasDeOnda} apresenta
@@
 Além disso, o número de harmônicos será limitado pela frequência máxima $t_a/2$.
-Musicalmente crucial aqui é internalizar que se um sinal real foi decomposto por fourier e possui uma componente em uma frequência $f_n$ com alguma energia
+
+Musicalmente crucial aqui é internalizar que a presença de uma componente em uma frequência $f_n$ na decomposição por fourier possui alguma energia
    \footnote{A energia total equivale à soma dos quadrados das amplitudes como as de pressao/voltagem pelo tempo,e.g. ~\ref{fig:formasDeOnda}.
@@
 também podem ser elevadas ao quadrado para resultar em quantificação de energia e somam-se na energia total. As energias se equivalem pelo teorema de Parseval.},
-então existe uma contribuição senoidal, puramente harmônica no som. Esta energia concentrada especificamente na frequência $f_n$ é separada
+se e somente se existe uma contribuição senoidal, puramente harmônica no som, naquela frequência $f_n$. Esta energia concentrada especificamente na frequência $f_n$ é separada
  pelo ouvido para adentrar em um nível cognitivo de processamento\footnote{Esta separação em frequência é realizada por diversas espécies através de mecanismos similares à cóclea humana.}.
-Estas componentes são geralmente as principais responsáveis pela qualidade chamada timbre e, caso não apresentem proporções harmônicas (relações de pequenos números), são ruídos, i.e.
+  As componentes senoidáis são geralmente as principais responsáveis pela qualidade chamada timbre e, caso não apresentem proporções harmônicas (relações de pequenos números), são ruídos, i.e.
 não notas com frequência fundamental como o lá 4 ($440 Hz$) de nossos exemplos.
@@
 
-Podemos ver abaixo como se dá isso na representação discreta.
-O mínimo necessário para representar uma oscilação são 2 amostras, resultando na frequência de Nyquist $f_{Ny}=\frac{t_a}{2}$.
-Todas as sequências fixas de apenas 3 amostras também representam
-somente 1 frequência, pois sua primeira harmônica usaria 1,5 amostras e ultrapassa o limite inferior de 2 amostras mínimas.
-Com 4 amostras, podemos representar 1 ou 2 frequências em proporções diferentes.
-A figura ~\ref{fig:amostras4} mostra estas frequências sendo realizadas compondo as formas de onda básicas e se remontando.
-
-A figura ~\ref{fig:amostras8} mostra as decomposições senoidais para o caso de 8 amostras.
-\begin{figure}[h!]
-    \centering
-    \caption{Frequência máxima em qualquer $t_a$}
-        \includegraphics[width=\textwidth]{figuras/amostras2}
-        \label{fig:amostras2}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \caption{Frequência única em quaisquer 3 amostras}
-        \includegraphics[width=\textwidth]{figuras/amostras3}
-        \label{fig:amostras2}
-\end{figure}
-
+No caso de uma forma de onda fixa (e de tamanho fixo), o espectro é sempre harmônico e estático.
+Fixada a frequência (inverso do comprimento da onda), cada forma de onda é composta de proporções específicas das componentes harmônicas e
+quanto maior a curvatura do trecho na forma de onda, maior a contribuição do trecho para a
+concentração de energia nos harmônicos agudos.
+
+Podemos ver isso claramente em sons reais.
 A onda rotulada como 'som real amostrado' na figura ~\ref{fig:formasDeOnda} é um período extraído de um som real relativamente comportado.
@@
 O período escolhido para a amostragem é relativamente curto, com 98 amostras corresponde a $\frac{44100}{98}=450 Hz$.
+Pode-se perceber o espectro rico em frequências do oboé e o espectro do mais grave do som real.
 Compararemos abaixo o conteúdo espectral da nota original com o resultante da utilização destas amostras.
@@
         \label{fig:espectroOboe}
+\end{figure}
+
+\subsubsection{O espectro no som amostrado}
+Além deste papel-chave no timbre e cognição, a presença e comportamento destas componentes é peculiar no som discretizado.
+Considere um sinal $T_i$ e sua decomposição de Fourier $\mathcal{F}\{T_i\}=C_i=\{c_i\}_0^{\Lambda-1}$.
+Sabemos que a recomposição segue a conversão\footnote{Lembrando que o fator $\frac{1}{\Lambda}$ pode ser distribuído dentre a transformada e a reconstrução como preferir}:
+
+\begin{equation}\label{recomposicaoFourier}
+t_i = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}c_ke^{j . \frac{2\pi k}{\Lambda} i } = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}(a_k+ j . b_k)\left[cos(w_k i) +j . sen(w_k i)\right]
+\end{equation}
+
+Onde $c_k = a_k + j .
+b_k$ e $w_k=\frac{2\pi k}{\Lambda}$.
+Nosso som possui amostras $t_i$ reais, e sabemos a contribuição de cada frequênciindexada com $k$ e de valor $c_k$ pelos termos reais da equação ~\ref{recomposicaoFourier} que resultam em senóides simples, cada uma com uma amplitude e fase:
+
+\begin{equation}\label{moduloEfase}
+\begin{split}
+t_i& = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}\left[a_k cos(w_k i) -b_k sen(w_k i)\right] \\
+   & = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]
+\end{split}
+\end{equation}
+
+Podemos notar claramente pela equação ~\ref{moduloEfase} que o termo imaginário apenas acrescenta uma fase à senóide real.
+Ao mesmo tempo também altera a amplitude mas a informação de amplitude já está em $a_k$ somente, i.e.
+os termos imaginários da decomposição espectral por Fourier proporcionam a varredura de fase
+ $[-\frac{\pi}{2},+\frac{\pi}{2}]$.
+O sinal de $a_k$ especifica se estamos do lado direito ou esquerdo do circulo trigonométrico, completando a varredura completa de fase (os intervalos $[-\frac{\pi}{2},+\frac{\pi}{2}]$ e $[\frac{\pi}{2},\frac{3\pi}{2}]$ completam $2\pi$).
+
+\begin{figure}[h!]
+    \centering
+    \caption{Oscilação de 2 amostras (frequência máxima em qualquer $t_a$)}
+        \includegraphics[width=\textwidth]{figuras/amostras2c}
+        \label{fig:amostras2}
+\end{figure}
+
+A figura ~\ref{fig:amostras2} exibe duas amostras e a componente espectral que contém.
+A decomposição de Fourier resulta neste caso em somente dois coeficientes $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$.
+O papel das amplitudes $a_k$ fica bem claro:
+ $a_0$ é controla o deslocamento de fixo\footnote{chamado de bias ou offset} e $a_1$ especifica a amplitude da oscilação em si.
+A frequência que possui a energia\footnote{Lembrando que as energias garantidamente se equivalem pela $\frac{1}{\Lambda} .
+\sum_{k=0}^{\Lambda -1}c_k^2 = \sum_{i=0}^{\Lambda-1}t_i^2$ (Teorema de Parseval)}
+ $e_k=\frac{(c_k)^2}{\Lambda=2}$ é dada pela relação $f_k=k .
+\frac{t_a}{\Lambda=2} $.
+Ou seja, no caso de duas amostras só temos energia nas frequências $f_0=0$ e $f_1=\frac{t_a}{\Lambda=2}=t_{\text{máx}}$.
+
+Este caso é de especial importância pois o mínimo necessário para representar uma oscilação são 2 amostras, e disso resulta a frequência de Nyquist $f_{\text{máx}}=\frac{t_a}{2}$.
+Esta é a frequência máxima presente em um som amostrado com $t_a$ amostras por segundo\footnote{Qualquer sinal amostrado possui esta característica, não somente o som digitalizado.}.
+
+Vejamos o que acontece no caso de 3 amostras. Todas as sequências fixas $T_i$ de apenas 3 amostras também apresentam
+somente 1 frequência, pois sua primeira harmônica usaria 1,5 amostras e ultrapassa o limite inferior de 2 amostras mínimas (a frequência da harmônica ultrapassaria a de Nyquist pois:  $\; \frac{t_a}{3}.2 > \frac{t_a}{2} $).
+Se $\Lambda=3$,
+os coeficiêntes $\{c_k\}_0^{\Lambda-1=2}$ da decomposição por Fourier apresentam-se em 
+3 componentes frequenciais.
+
+Partimos de $\Lambda$ amostras reais $t_i$ e temos $\Lambda$ coeficientes complexos $c_k=a_k+j.b_k$.
+Os coeficientes $c_k$ se equivalem dois a dois\footnote{Módulos iguais e fases com sinais opostos; $a_{k1}=a_{k2}$ e $b_{k1}=-b_{k2}$.} pois correspondem às frequências $f_k = k.\frac{t_a}{\Lambda}, \; k \in \{0...\Lambda-1\} $
+ e quando $ f_k > f_{\text{máx}} = \frac{t_a}{2} $ (o que já exclui $f_0=0$ e $f_k=\frac{t_a}{2}$), a frequência $f_k$ é espelhada para $f_l=\frac{t_a}{2} - (f_k-\frac{t_a}{2})=t_a-f_k=t_a - k\frac{t_a}{\Lambda}=(\Lambda-k)\frac{t_a}{\Lambda} \Rightarrow f_k=f_{\Lambda-k}$.
+O mesmo pode ser observado com $w_k=f_k.\frac{2\pi}{t_a}$ e $w_{\text{máx}}=\pi$ resultando em $w_k=w_{\Lambda-k}$.
+
+  Dito de outra forma, em uma decomposição de $\Lambda$ amostras, as $\Lambda$ componentes frequenciais $\{c_i\}_0^{\Lambda-1}$ resultantes
+   são equivalentes em pares.
+   Excepcional excessão para $f_0$ e, no caso de $\Lambda$ ser par, de $f_{\Lambda/2}=f_{\text{máx}}=\frac{t_a}{2}$.
+Além disso, estas duas frequências (a frequência zero e a frequência máxima) não são representadas com variação de fase e, portanto, são estritamente reais.
+Assim, podemos
+   concluir que o número $\tau$ de pares de coeficientes equivalentes é:
+
+\begin{equation}\label{coefsPareados}
+\tau = \frac{\Lambda - \Lambda \% 2}{2} +\Lambda \% 2 -1
+\end{equation}
+
+e resultam evidentes as equivalências ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} e ~\ref{equivalenciasFases}:
+
+\begin{equation}\label{equivalenciasFreqs}
+f_{k}=f_{\Lambda-k}\;, \;\; w_{k}=w_{\Lambda-k}\;\;\;, \quad \;\; \forall \quad 1 \leq k \leq \tau  
+\end{equation}
+
+Como $a_k = a_{\Lambda -k}\;\;$ e $\;\;b_k = - b_{\Lambda -k}$:
+
+\begin{equation}\label{equivalenciasModulos}
+\sqrt{a_k^2 + b_k^2} = \sqrt{a_{\Lambda - k}^2 + b_{\Lambda -k}^2} \;\;, \quad \;\; \forall \quad 1 \leq k \leq \tau  \\
+\end{equation}
+
+\begin{equation}\label{equivalenciasFases}
+tg^{-1}\left(\frac{b_k}{a_k}\right)=-tg^{-1}\left(\frac{b_{\Lambda -k}}{a_{\Lambda - k}}\right)\;\;,\quad \;\; \forall \quad 1 \leq k \leq \tau
+\end{equation}
+
+Com $k \in \mathbb{N}$.
+A observação da equação de reconstrução para o sinal real ~\ref{moduloEfase} em conjunto com as equivalências dos módulos e fases ~\ref{equivalenciasModulos} e ~\ref{equivalenciasFases}
+mostra a combinação em fase das componentes em $w_k$ e $w_{\Lambda-k}$:
+
+\begin{multline}
+w_k=-w_{\Lambda-k}, a_k=a_{\Lambda-k},  b_k=-b_{\Lambda-k} \Rightarrow   \\
+\Rightarrow \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]
+\end{multline}
+
+Com 4 amostras, podemos representar 1 ou 2 frequências em proporções diferentes.
+A figura ~\ref{fig:amostras4} mostra estas frequências sendo realizadas compondo as formas de onda básicas e se remontando.
+
+A figura ~\ref{fig:amostras8} mostra as decomposições senoidais para o caso de 8 amostras.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Frequência única em quaisquer 3 amostras}
+        \includegraphics[width=\textwidth]{figuras/amostras3}
+        \label{fig:amostras3}
 \end{figure}

### 2012-08-22 devEresultados.tex
encaminhadas as figuras de 2,3 e 4 amostras

@@
 
+%\subsection{Caracterização da nota musical discretizada}
 \subsection{Caracterização da nota musical}
-A música é tradicionalmente pensada através de átomos chamados notas. Em sua concepção
-simples, estas notas possuem ao menos duração, volume, altura e timbre. Embora estas
+%\subsection{Caracterização da nota musical digital}
+A música é tradicionalmente pensada através de unidades chamadas notas. Em sua concepção
+simples, as notas possuem ao menos duração, volume, altura e timbre. Embora estas
 sejam características psicofísicas, podemos lidar com elas através de suas características
@@
 com ciclo de duração $\delta_{f_0} = 1/f_0$. Esta duração multiplicada pela taxa de amostragem $f_a$ resulta no número de amostras
-do ciclo $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$. Resulta evidente que, sendo $T_i^f$ a sequência sonora de frequência fundamental $f=f_0$:
+do ciclo $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$. Evidente que, sendo $T_i^f$ a sequência sonora de frequência fundamental $f=f_0$:
 
@@
 de onda apresentam cadeias de componentes harmônicos e usos específicos.
-As formas de onda especificadas em ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} são dispostas graficamente em ~\ref{fig:formasDeOnda}. 
-São as formas de onda artificiais tradicionalmente utilizadas na música para síntese e controle oscilatório de variáveis e apresentam diversos usos fora da música.
+As formas de onda especificadas em ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} são dispostas graficamente na figura ~\ref{fig:formasDeOnda}. 
+São as formas de onda artificiais tradicionalmente usadas na música para síntese e controle oscilatório de variáveis, apresentam diversos usos também fora da música.
 
 A dente de serra apresenta todos os componentes da série
-harmônica com energia decrescente de $-6dB/oitava$:
+harmônica com energia decrescente de $-6dB/oitava$. A sequência de amostras temporais pode ser descrita da seguinte forma:
 \begin{equation}\label{denteDeSerra}
@@
 A senóide consiste de um nódulo único no espectro, frequência pura.
-A dente de serra é a únida com a série harmônica completa (pares e ímpares).
+A dente de serra é a única com a série harmônica completa (pares e ímpares).
 Já as ondas triangular e quadrada possuem as mesmas componentes espectrais, mas com decaimentos de $-12dB/oitava$ e $-6dB/oitava$.
@@
 
-O espectro harmônico é formado pelas frequências múltiplas da frequência fundamental $f_n=n.f_0$.
+O espectro harmônico é formado pelas frequências múltiplas da frequência fundamental $f_n=(n+1).f_0$.
 Como nossa percepção segue a escala logarítmica, o espectro possui notas diferentes da frequência fundamental.
-Além disso, o número de harmônicos será limitado pela frequência máxima $t_a/2$.
-
-Musicalmente crucial aqui é internalizar que a presença de uma componente em uma frequência $f_n$ na decomposição por fourier possui alguma energia
-   \footnote{A energia total equivale à soma dos quadrados das amplitudes como as de pressao/voltagem pelo tempo,e.g. ~\ref{fig:formasDeOnda}.
+Além disso, o número de harmônicos será limitado pela frequência máxima $f_a/2$.
+
+Musicalmente crucial aqui é internalizar que a presença de
+energia\footnote{A energia total equivale à soma dos quadrados das amplitudes como as de pressao/voltagem pelo tempo,e.g. ~\ref{fig:formasDeOnda}.
 As componentes espectrais - e.g. ~\ref{fig:espectroDeOndas} e ~\ref{fig:espectroOboe} -
-também podem ser elevadas ao quadrado para resultar em quantificação de energia e somam-se na energia total. As energias se equivalem pelo teorema de Parseval.},
-se e somente se existe uma contribuição senoidal, puramente harmônica no som, naquela frequência $f_n$. Esta energia concentrada especificamente na frequência $f_n$ é separada
+também podem ser elevadas ao quadrado para resultar em quantificação de energia e somam-se na energia total. As energias se equivalem pelo teorema de Parseval.}
+em uma componente de frequência $f_n$ na decomposição por fourier 
+implica na presença de uma oscilação senoidal na constituição do som, puramente harmônica no som e naquela frequência $f_n$. Esta energia concentrada especificamente na frequência $f_n$ é separada
  pelo ouvido para adentrar em um nível cognitivo de processamento\footnote{Esta separação em frequência é realizada por diversas espécies através de mecanismos similares à cóclea humana.}.
   As componentes senoidáis são geralmente as principais responsáveis pela qualidade chamada timbre e, caso não apresentem proporções harmônicas (relações de pequenos números), são ruídos, i.e.
-não notas com frequência fundamental como o lá 4 ($440 Hz$) de nossos exemplos.
-Além disso, nossa noção de altura em um complexo sonoro é baseada na semelhança do espectro com a série harmônica.
+não são notas com frequência fundamental como o lá 4 ($440 Hz$) de nossos exemplos.
+Além disso, nossa noção de altura absoluta em um complexo sonoro é baseada na semelhança do espectro com a série harmônica.
 
 No caso de uma forma de onda fixa (e de tamanho fixo), o espectro é sempre harmônico e estático.
-Fixada a frequência (inverso do comprimento da onda), cada forma de onda é composta de proporções específicas das componentes harmônicas e
+Fixada a frequência fundamental (inverso do comprimento da onda), cada forma de onda é composta de proporções específicas das componentes harmônicas e
 quanto maior a curvatura do trecho na forma de onda, maior a contribuição do trecho para a
@@
 Pode-se perceber o espectro rico em frequências do oboé e o espectro do mais grave do som real.
-Compararemos abaixo o conteúdo espectral da nota original com o resultante da utilização destas amostras.
+A figura ~\ref{fig:espectroOboe}
+expõe em conjunto o conteúdo espectral da nota original e o resultante da utilização destas amostras.
 
@@
 
-O som resultante possui o espectro momentâneo do som original. Por ser repetido de forma identica,
-seu espectro é perfeitamente harmônico, sem os ruídos e variações típicas do fenômeno natural. Isso pode ser visto claramente na figura ~\ref{fig:espectroOboe} onde a nota original do oboé e uma nota 
+O som resultante possui o espectro momentâneo do som original. Por ser repetido de forma idêntica,
+seu espectro é perfeitamente harmônico, sem os ruídos e variações típicas do fenômeno natural.
+Isso pode ser visto claramente na figura ~\ref{fig:espectroOboe} onde os espectros da nota original do oboé e de uma nota
 artificial - de mesma duração e cujas amostras consistem no mesmo período da figura ~\ref{fig:formasDeOnda} - estão dispostas juntamente.
@@
 \subsubsection{O espectro no som amostrado}
-Além deste papel-chave no timbre e cognição, a presença e comportamento destas componentes é peculiar no som discretizado.
+Além deste papel-chave no timbre e na cognição como um todo, a presença e comportamento destas componentes é peculiar no som discretizado.
 Considere um sinal $T_i$ e sua decomposição de Fourier $\mathcal{F}\{T_i\}=C_i=\{c_i\}_0^{\Lambda-1}$.
-Sabemos que a recomposição segue a conversão\footnote{Lembrando que o fator $\frac{1}{\Lambda}$ pode ser distribuído dentre a transformada e a reconstrução como preferir}:
+Sabemos que a recomposição segue a conversão das componentes frequenciais em amostras temporais\footnote{Lembrando que o fator $\frac{1}{\Lambda}$ pode ser distribuído dentre a transformada e a reconstrução como preferir}:
 
@@
 b_k$ e $w_k=\frac{2\pi k}{\Lambda}$.
-Nosso som possui amostras $t_i$ reais, e sabemos a contribuição de cada frequênciindexada com $k$ e de valor $c_k$ pelos termos reais da equação ~\ref{recomposicaoFourier} que resultam em senóides simples, cada uma com uma amplitude e fase:
+Nosso som possui amostras $t_i$ reais que resultam das contribuições de cada componente de frequência $w_k$ e cujo $c_k$ regula o módulo e a fase.
+A parte real da equação ~\ref{recomposicaoFourier} nos fornece a componente de forma mais clara:
 
@@
 Podemos notar claramente pela equação ~\ref{moduloEfase} que o termo imaginário apenas acrescenta uma fase à senóide real.
-Ao mesmo tempo também altera a amplitude mas a informação de amplitude já está em $a_k$ somente, i.e.
+Ao mesmo tempo $b_k$ também altera a amplitude mas a informação de amplitude já está em $a_k$, i.e.
 os termos imaginários da decomposição espectral por Fourier proporcionam a varredura de fase
@@
     \caption{Oscilação de 2 amostras (frequência máxima em qualquer $t_a$)}
-        \includegraphics[width=\textwidth]{figuras/amostras2c}
+        \includegraphics[width=\textwidth]{figuras/amostras2c_}
         \label{fig:amostras2}
@@
 O papel das amplitudes $a_k$ fica bem claro:
- $a_0$ é controla o deslocamento de fixo\footnote{chamado de bias ou offset} e $a_1$ especifica a amplitude da oscilação em si.
+ $a_0$ é controla o deslocamento fixo\footnote{chamado de bias ou offset} e $a_1$ especifica a amplitude da oscilação em si.
 A frequência que possui a energia\footnote{Lembrando que as energias garantidamente se equivalem pela $\frac{1}{\Lambda} .
@@
  $e_k=\frac{(c_k)^2}{\Lambda=2}$ é dada pela relação $f_k=k .
-\frac{t_a}{\Lambda=2} $.
-Ou seja, no caso de duas amostras só temos energia nas frequências $f_0=0$ e $f_1=\frac{t_a}{\Lambda=2}=t_{\text{máx}}$.
-
-Este caso é de especial importância pois o mínimo necessário para representar uma oscilação são 2 amostras, e disso resulta a frequência de Nyquist $f_{\text{máx}}=\frac{t_a}{2}$.
-Esta é a frequência máxima presente em um som amostrado com $t_a$ amostras por segundo\footnote{Qualquer sinal amostrado possui esta característica, não somente o som digitalizado.}.
+\frac{f_a}{\Lambda=2} $.
+Ou seja, no caso de duas amostras só temos energia nas frequências $f_0=0$ e $f_1=\frac{f_a}{\Lambda=2}=f_{\text{máx}}$.
+
+Este caso é de especial importância pois o mínimo necessário para representar uma oscilação são 2 amostras e disso resulta a frequência de Nyquist $f_{\text{máx}}=\frac{f_a}{2}$.
+Esta é a frequência máxima presente em um som amostrado com $f_a$ amostras por segundo\footnote{Qualquer sinal amostrado possui esta característica, não somente o som digitalizado.}.
 
 Vejamos o que acontece no caso de 3 amostras. Todas as sequências fixas $T_i$ de apenas 3 amostras também apresentam
-somente 1 frequência, pois sua primeira harmônica usaria 1,5 amostras e ultrapassa o limite inferior de 2 amostras mínimas (a frequência da harmônica ultrapassaria a de Nyquist pois:  $\; \frac{t_a}{3}.2 > \frac{t_a}{2} $).
+somente 1 frequência, pois sua primeira harmônica usaria 1,5 amostras e ultrapassa o limite inferior de 2 amostras mínimas (a frequência da harmônica ultrapassaria a de Nyquist pois:  $\; \frac{f_a}{3}.2 > \frac{f_a}{2} $).
 Se $\Lambda=3$,
 os coeficiêntes $\{c_k\}_0^{\Lambda-1=2}$ da decomposição por Fourier apresentam-se em 
-3 componentes frequenciais.
+3 componentes frequenciais. Uma delas é a frequência zero, as outras duas contribuem de forma igual na reconstrução da senóide com $f=f_a/3$.
+
+\begin{figure}[h!]
+    \centering
+    \caption{3 amostras apresentam uma única frequência}
+        \includegraphics[width=\textwidth]{figuras/amostras3b}
+        \label{fig:amostras3}
+\end{figure}
 
 Partimos de $\Lambda$ amostras reais $t_i$ e temos $\Lambda$ coeficientes complexos $c_k=a_k+j.b_k$.
-Os coeficientes $c_k$ se equivalem dois a dois\footnote{Módulos iguais e fases com sinais opostos; $a_{k1}=a_{k2}$ e $b_{k1}=-b_{k2}$.} pois correspondem às frequências $f_k = k.\frac{t_a}{\Lambda}, \; k \in \{0...\Lambda-1\} $
- e quando $ f_k > f_{\text{máx}} = \frac{t_a}{2} $ (o que já exclui $f_0=0$ e $f_k=\frac{t_a}{2}$), a frequência $f_k$ é espelhada para $f_l=\frac{t_a}{2} - (f_k-\frac{t_a}{2})=t_a-f_k=t_a - k\frac{t_a}{\Lambda}=(\Lambda-k)\frac{t_a}{\Lambda} \Rightarrow f_k=f_{\Lambda-k}$.
-O mesmo pode ser observado com $w_k=f_k.\frac{2\pi}{t_a}$ e $w_{\text{máx}}=\pi$ resultando em $w_k=w_{\Lambda-k}$.
+Os coeficientes $c_k$ se equivalem dois a dois\footnote{Módulos iguais e fases com sinais opostos; $a_{k1}=a_{k2}$ e $b_{k1}=-b_{k2}$.} pois correspondem às frequências $f_k = k.\frac{f_a}{\Lambda}, \; k \in \{0...\Lambda-1\} $
+ e quando $ f_k > f_{\text{máx}} = \frac{f_a}{2} $ (o que já exclui $f_0=0$ e $f_k=\frac{f_a}{2}$), a frequência $f_k$ é espelhada para $f_l=\frac{f_a}{2} - (f_k-\frac{f_a}{2})=f_a-f_k=f_a - k\frac{f_a}{\Lambda}=(\Lambda-k)\frac{f_a}{\Lambda} \Rightarrow f_k\equiv f_{\Lambda-k}$.
+
+ O mesmo pode ser observado com 
+ $w_k=f_k.\frac{2\pi}{f_a}$ e lembrando da periodicidade $2\pi$, que resulta em $w_k=-w_{\Lambda-k}$.
+Como o coseno é uma função par e a tangente inversa é impar, as componentes em $w_k$ e $w_{\Lambda-k}$ se somam na equação de reconstrução das amostras reais mostrada em ~\ref{recomposicaoFourier}.
 
@@
    são equivalentes em pares.
-   Excepcional excessão para $f_0$ e, no caso de $\Lambda$ ser par, de $f_{\Lambda/2}=f_{\text{máx}}=\frac{t_a}{2}$.
+   Excepcional excessão para $f_0$ e, no caso de $\Lambda$ ser par, de $f_{\Lambda/2}=f_{\text{máx}}=\frac{f_a}{2}$.
 Além disso, estas duas frequências (a frequência zero e a frequência máxima) não são representadas com variação de fase e, portanto, são estritamente reais.
@@
 \begin{equation}\label{equivalenciasFreqs}
-f_{k}=f_{\Lambda-k}\;, \;\; w_{k}=w_{\Lambda-k}\;\;\;, \quad \;\; \forall \quad 1 \leq k \leq \tau  
+f_{k}\equiv f_{\Lambda-k}\;, \;\; w_{k}=-w_{\Lambda-k}\;\;\;, \quad \;\; \forall \quad 1 \leq k \leq \tau  
 \end{equation}
@@
 
-\begin{multline}
-w_k=-w_{\Lambda-k}, a_k=a_{\Lambda-k},  b_k=-b_{\Lambda-k} \Rightarrow   \\
-\Rightarrow \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]
-\end{multline}
+\begin{equation}
+t_i = \frac{a_0}{\Lambda} + \frac{2}{\Lambda}\sum_{k=1}^{\tau}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]+ \frac{ a_{\Lambda/2+1}}{\Lambda}.(1-\Lambda\% 2)
+\end{equation}
+
+Assim, a exemplo da figura ~\ref{fig:amostras3}, a transformada de Fourier de 3 amostras possui 2 coeficientes imaginários que possuem quantidades iguais de energia na mesma frequência.
 
 Com 4 amostras, podemos representar 1 ou 2 frequências em proporções diferentes.
-A figura ~\ref{fig:amostras4} mostra estas frequências sendo realizadas compondo as formas de onda básicas e se remontando.
-
-A figura ~\ref{fig:amostras8} mostra as decomposições senoidais para o caso de 8 amostras.
+A figura ~\ref{fig:amostras4} mostra estas frequências duas componentes se somando em uma forma de onda.
+A figura ~\ref{fig:formas4} realiza explicita os harmônicos para as formas de onda básicas das equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} e figura ~\ref{fig:formasDeOnda}.
 
@@
     \centering
-    \caption{Frequência única em quaisquer 3 amostras}
-        \includegraphics[width=\textwidth]{figuras/amostras3}
-        \label{fig:amostras3}
+    \caption{Componentes frequenciais em 4 amostras}
+        \includegraphics[width=\textwidth]{figuras/amostras4}
+        \label{fig:amostras4}
 \end{figure}
+
+A figura ~\ref{fig:amostras8} mostra as decomposições senoidais para o caso de 8 amostras.
 

### 2012-08-22 scripts/amostras2b.py

@@ -31,7 +31,7 @@ iii=n.linspace(-0.5,1.5,200) # domínio temporal das amostras
 s=(ab1/2)*n.cos(ii+fas)+a0/2 # Reconstruindo senóide para o domínio dado
 p.plot(iii,s,"m--") # plotada a senóide passando pelas amostras
 
-
+p.xticks((-1,0,1,2),(r"$-\delta_a$",r"$0$",r"$\delta_a$",r"$2\delta_a$"))
 # valores cruciais em y
 p.yticks((0,a0/2-ab1/2,a0/2,a0/2+ab1/2), ('0',r'$\frac{a_0-|a_1|}{2}$',r'$\frac{a_0}{2}$',r'$\frac{a_0+|a_1|}{2}$'), color = 'k', size = 20)
 p.plot((-10,10),(a0/2,)*2,"k--")

### 2012-08-22 scripts/amostras3b.py

@@ -4,7 +4,13 @@
 import pylab as p, numpy as n, scikits.audiolab as a
 f=n.fft.fft
 
-n3=n.random.rand(3)*2-1
+#n3=n.random.rand(3)*2-1
+n3=n.array([-0.20356754,  0.91942605,  0.44866268])
+#n3=n.array([-0.40+.3,  0.+.3,  0.4+.3])
+#n3=n.array([-0.40+.3,  0.4+.3,  0.4+.3])
+#n3=n.array([-0.40+.3, - 0.4+.3,  0.4+.3])
+#n3=n.array([+0.40+.3, - 0+.3, - 0.4+.3])
+#n3=n.array([-0.40+.3, + 0.4+.3, - 0.4+.3])
 p.plot(n3,"bo")
 
 ff=f(n3)
@@ -26,8 +32,14 @@ iii=n.linspace(0-fr2,3-fr2,200) # [0,3] == 3 \delta_a, período
 s=(2/3.)*ab1*n.cos(ii+fas)+a0/3
 p.plot(iii,s,"m--")
 
+p.plot((-10,10),(a0/3,)*2,"k--")
+p.plot((-10,10),(a0/3-2*ab1/3,)*2,"k--")
+p.plot((-10,10),(a0/3+2*ab1/3,)*2,"k--")
+p.xticks((-1,0,1,2,3),(r"$-\delta_a$",r"$0$",r"$\delta_a$",r"$2\delta_a$",r"$3\delta_a$"))
+p.yticks((0,a0/3-2*ab1/3,a0/3,a0/3+2*ab1/3), ('0',r'$\frac{a_0-2|c_1|}{3}$',r'$\frac{a_0}{3}$',r'$\frac{a_0+2|c_1|}{3}$'), color = 'k', size = 20)
+
 p.xlim(-1.2,3.2)
-p.ylim(-1.1,1.1)
+p.ylim(-1.1+.5,1.1)
 
 p.ylabel(r"amplitude $\rightarrow$")
 p.xlabel(r"tempo $\rightarrow$")

### 2012-08-22 scripts/amostras4.py

@@ -1,0 +1,59 @@
+#-*- coding: utf8 -*-
+# http://matplotlib.sourceforge.net/examples/api/legend_demo.html
+# 
+import pylab as p, numpy as n, scikits.audiolab as a
+f=n.fft.fft
+
+#n4=n.random.rand(4)*2-1
+n4=n.array([ 0.58003705, -0.30828309, -0.29797696, -0.99219078])
+p.plot(n4,"bo")
+
+ff=f(n4)
+
+# Primeiro componente, 0Hz
+a0=n.real(ff[0]) 
+b0=n.imag(ff[0]) # sempre zero
+
+# Segundo componente, t_a/N Hz
+ab1=n.abs(ff[1]) # (a**2+b**2)**0.5
+a1=n.real(ff[1])
+b1=n.imag(ff[1])
+fas=n.arctan(b1/a1) # fase fas=n.angle(f[1])
+if a1<0: fas+=n.pi # segundo e terceiro quadrantes somam pi
+print("abs: %s, a1: %s, b1: %s, fas: %s" % (ab1,a1,b1,fas))
+
+# Segundo componente, t_a/N Hz
+ab2=n.abs(ff[2]) # (a**2+b**2)**0.5
+a2=n.real(ff[2])
+b2=n.imag(ff[2])
+fas2=n.arctan(b2/a2) # fase fas=n.angle(f[1])
+if a2<0: fas2+=n.pi # segundo e terceiro quadrantes somam pi
+print("abs: %s, a2: %s, b2: %s, fas2: %s" % (ab2,a2,b2,fas2))
+
+
+
+
+fr1=(2*n.pi/4)/2 #ciclo em 4 * \delta_a, metade do intervalo angular 
+fr2=1/2. # metade de \delta_a
+#fr1=fr2=0
+ii=n.linspace(0-fr1,2*n.pi-fr1,200) # [0,2*pi] ciclo completo
+iii=n.linspace(0-fr2,4-fr2,200) # [0,3] == 3 \delta_a, período
+
+s=(1/4.)*ab2*n.cos(2*ii+fas2)+(2/4.)*ab1*n.cos(ii+fas)+a0/4
+p.plot(iii,s,"m", linewidth=3, label=r"$\oplus$")
+ss=(1/4.)*ab2*n.cos(2*ii+fas2)
+sss=(2/4.)*ab1*n.cos(ii+fas)
[…]

### 2012-08-22 devEresultados.tex
feito do espectro do som amostrado

@@
 Com 4 amostras, podemos representar 1 ou 2 frequências em proporções diferentes.
-A figura ~\ref{fig:amostras4} mostra estas frequências duas componentes se somando em uma forma de onda.
-A figura ~\ref{fig:formas4} realiza explicita os harmônicos para as formas de onda básicas das equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} e figura ~\ref{fig:formasDeOnda}.
+A figura ~\ref{fig:amostras4} mostra uma forma de onda de 4 amostras e suas duas componentes.
+Note que as contribuições individuais se somam de fato na forma de onda original, e que as curvaturas maiores são fruto da frequência mais aguda.
+
+A figura ~\ref{fig:formas4} explicita os harmônicos em 4 amostras nas formas de onda básicas das equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} e figura ~\ref{fig:formasDeOnda}.
+Todas consistem em apenas 1 senóide, com excessão da dente de serra que possui os harmônicos pares.
 
@@
     \caption{Componentes frequenciais em 4 amostras}
-        \includegraphics[width=\textwidth]{figuras/amostras4}
+        \includegraphics[width=\textwidth]{figuras/amostras4_}
         \label{fig:amostras4}
@@
 
-A figura ~\ref{fig:amostras8} mostra as decomposições senoidais para o caso de 8 amostras.
-
-[resumo das sequencias com duracao, frequencia, forma de onda e intensidade]
-
-[usos musicais]
+\begin{figure}[h!]
+    \centering
+    \caption{Formas de onda básicas em 4 amostras}
+        \includegraphics[width=\textwidth]{figuras/amostras4formas}
+        \label{fig:formas4}
+\end{figure}
+
+A figura ~\ref{fig:amostras6} mostra as decomposições senoidais para o caso de 6 amostras.
+A figura ~\ref{fig:formas6} decompõe as formas de onda básicas.
+Note que as ondas se diferenciam espectralmente.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Componentes frequenciais em 6 amostras}
+        \includegraphics[width=\textwidth]{figuras/amostras6}
+        \label{fig:amostras6}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \caption{Formas de onda básicas em 6 amostras}
+        \includegraphics[width=\textwidth]{figuras/amostras6formas_}
+        \label{fig:formas6}
+\end{figure}
+
+\subsubsection{A nota caracterizada}
+
+\subsubsection{Usos musicais}
 

### 2012-08-22 scripts/amostras4formasDeOnda.py

@@ -1,0 +1,75 @@
+#-*- coding: utf8 -*-
+# http://matplotlib.sourceforge.net/examples/api/legend_demo.html
+# 
+import pylab as p, numpy as n, scikits.audiolab as a
+f=n.fft.fft
+
+#n4=n.random.rand(4)*2-1
+n4=n.array([ 0.58003705, -0.30828309, -0.29797696, -0.99219078])
+T=4
+#T=1024
+l=n.linspace(0,2*n.pi,T,endpoint=False)
+senoide=n.sin(l)
+dente=n.linspace (-1,1,T) # dente de serra
+triangular=n.hstack ((n.linspace(-1,1,T/2,endpoint=False),n.linspace(1,-1,T/2,endpoint=False)))
+quadrada=n.hstack ((n.ones(T/2),n.ones(T/2)* -1))
+
+ondas=[senoide,dente,triangular,quadrada]
+ondas_n=[u"senóide","dente de serra","triangular","quadrada"]
+
+for i in xrange(4):
+    p.subplot(int("22"+str(i+1)))
+    p.title(ondas_n[i])
+    n4=ondas[i]
+    p.plot(n4,"bo")
+
+    ff=f(n4)
+
+    # Primeiro componente, 0Hz
+    a0=n.real(ff[0]) 
+    b0=n.imag(ff[0]) # sempre zero
+
+    # Segundo componente, t_a/N Hz
+    ab1=n.abs(ff[1]) # (a**2+b**2)**0.5
+    a1=n.real(ff[1])
+    b1=n.imag(ff[1])
+    fas=n.arctan(b1/a1) # fase fas=n.angle(f[1])
+    if a1<0: fas+=n.pi # segundo e terceiro quadrantes somam pi
+    print("abs: %s, a1: %s, b1: %s, fas: %s" % (ab1,a1,b1,fas))
+
+    # Segundo componente, t_a/N Hz
+    ab2=n.abs(ff[2]) # (a**2+b**2)**0.5
+    a2=n.real(ff[2])
+    b2=n.imag(ff[2])
+    if a2: fas2=n.arctan(b2/a2) # fase fas=n.angle(f[1])
+    else: fas2=0
+    if a2<0: fas2+=n.pi # segundo e terceiro quadrantes somam pi
[…]

### 2012-08-22 devEresultados.tex
resolvido texto sobre a nota basica

@@
 
-\subsubsection{A nota caracterizada}
+\subsubsection{A nota básica}
+
+Escolhamos $f$ tal que $f$ divide $f_a$   \footnote{Esta escolha facilita as demonstrações pois os períodos são quantidades inteiras de amostras.
+Transporemos esta limitação na prática com o uso de uma tabela de amostragem ou com o uso direto de funções periódicas,e.g.
+$t_i=sen(x.i)$}.
+Uma sequência $T_i$ de amostras sonoras periódicas separadas por $\delta_a=1/f_a$ descreve uma nota musical de frequência $f$ Herz e duração $\Delta$ segundos se possuir a periodicidade $\lambda_f=f_a/f$ e tamanho $f_a .
+\Delta$:
+
+\begin{equation}
+T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{f_a . \Delta}=\{t_{i \; \% \left( \frac{f_a}{f} \right) } \}_0^{f_a . \Delta}
+\end{equation}
+
+Note que a nota por si só não especifica um tímbre.
+Mesmo assim, faz-se necessária a escolha de uma forma de onda para que as amostras $t_i$ tenham um valor estabelecido individualmente.
+Um único período dentre as ondas básicas pode ser utilizado para a especificação da nota da seguinte forma.
+
+Seja $f$ a frequência da nota, $\delta_f=1/f$ o seu comprimento de onda e $L_i^{f,\, \delta_f} $ a sequência que descreve um período da onda $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ (ver equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada}; $R_i^f$ é uma onda real amostrada),
+
+\begin{equation}
+L_i^{f , \delta_f } = \{ l_i^f \}_0^{\delta_f . f_a -1}=\{ l_i^f \}_0^{\lambda_f-1}
+\end{equation}
+
+Então a sequência $T_i$ consistirá em uma nota de duração $\Delta$ e frequência $f$ se:
+
+\begin{equation}
+T_i^{f,\; \Delta}=\{t_i^f\}_0^{f_a . \Delta}=\{l^f_{i\,\%\left(\frac{f_a}{f}\right)}\}_0^{f_a . \Delta}
+\end{equation}
 

### 2012-08-23 devEresultados.tex
terminada primeira versao das notas, revisando ultimas partes

--- 2012-08-22 devEresultados.tex
@@
 com ciclo de duração $\delta_{f_0} = 1/f_0$. Esta duração multiplicada pela taxa de amostragem $f_a$ resulta no número de amostras
-do ciclo $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$. Evidente que, sendo $T_i^f$ a sequência sonora de frequência fundamental $f=f_0$:
+do ciclo $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$. Evidente que, sendo $T_i^f$ a sequência sonora de frequência fundamental $f$:
 
@@
 
-O caso mais simples (e mais importante, como veremos a diante) é o do espectro que consiste somente
+O caso mais simples (e mais importante, como veremos adiante) é o do espectro que consiste somente
 de sua própria fundamental $f$. Este é o caso da senóide, frequência em movimento oscilatório puro chamado
@@
     \caption{Espectros das ondas sonoras musicais artificiais básicas}
-        \includegraphics[width=\textwidth]{figuras/espectroDeOndas5}
+        \includegraphics[width=\textwidth]{figuras/espectroDeOndas5b}
         \label{fig:espectroDeOndas}
@@
 Musicalmente crucial aqui é internalizar que a presença de
-energia\footnote{A energia total equivale à soma dos quadrados das amplitudes como as de pressao/voltagem pelo tempo,e.g. ~\ref{fig:formasDeOnda}.
+energia\footnote{A energia total equivale à soma dos quadrados das amplitudes como as de pressao/voltagem pelo tempo, e.g. figura ~\ref{fig:formasDeOnda}.
 As componentes espectrais - e.g. ~\ref{fig:espectroDeOndas} e ~\ref{fig:espectroOboe} -
@@
  pelo ouvido para adentrar em um nível cognitivo de processamento\footnote{Esta separação em frequência é realizada por diversas espécies através de mecanismos similares à cóclea humana.}.
-  As componentes senoidáis são geralmente as principais responsáveis pela qualidade chamada timbre e, caso não apresentem proporções harmônicas (relações de pequenos números), são ruídos, i.e.
+  As componentes senoidais são geralmente as principais responsáveis pela qualidade chamada timbre e, caso não apresentem proporções harmônicas (relações de pequenos números), são ruídos, i.e.
 não são notas com frequência fundamental como o lá 4 ($440 Hz$) de nossos exemplos.
@@
 O período escolhido para a amostragem é relativamente curto, com 98 amostras corresponde a $\frac{44100}{98}=450 Hz$.
-Pode-se perceber o espectro rico em frequências do oboé e o espectro do mais grave do som real.
+Pode-se perceber o espectro rico em frequências do oboé e o espectro mais grave do som real.
 A figura ~\ref{fig:espectroOboe}
@@
 Onde $c_k = a_k + j .
-b_k$ e $w_k=\frac{2\pi k}{\Lambda}$.
+b_k$ pondera a amplitude e fase de cada frequência $w_k=\frac{2\pi k}{\Lambda}$.
 Nosso som possui amostras $t_i$ reais que resultam das contribuições de cada componente de frequência $w_k$ e cujo $c_k$ regula o módulo e a fase.
@@
 O papel das amplitudes $a_k$ fica bem claro:
- $a_0$ é controla o deslocamento fixo\footnote{chamado de bias ou offset} e $a_1$ especifica a amplitude da oscilação em si.
+ $a_0$ é o deslocamento fixo\footnote{chamado de bias ou offset} e $a_1$ especifica a amplitude da oscilação em si.
 A frequência que possui a energia\footnote{Lembrando que as energias garantidamente se equivalem pela $\frac{1}{\Lambda} .
@@
     \caption{Componentes frequenciais em 4 amostras}
-        \includegraphics[width=\textwidth]{figuras/amostras4_}
+        \includegraphics[width=\textwidth]{figuras/amostras4__}
         \label{fig:amostras4}
@@
 
-\subsubsection{A nota caracterizada}
+\subsubsection{A nota básica}
+
+Escolhamos $f$ tal que $f$ divida $f_a$   \footnote{Esta escolha facilita as demonstrações pois os períodos são quantidades inteiras de amostras.
+Transporemos esta limitação na prática com o uso de uma tabela de amostragem ou com o uso direto de funções periódicas,e.g.
+$t_i=sen(x.i)$}.
+Uma sequência $T_i$ de amostras sonoras separadas por $\delta_a=1/f_a$ descreve uma nota musical de frequência $f$ Herz e duração $\Delta$ segundos se, e somente se, possuir a periodicidade $\lambda_f=f_a/f$ e tamanho $f_a .
+\Delta$:
+
+\begin{equation}
+T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{f_a . \Delta}=\{t_{i \; \% \left( \frac{f_a}{f} \right) } \}_0^{f_a . \Delta}
+\end{equation}
+
+Note que a nota por si só não especifica um timbre.
+Mesmo assim, faz-se necessária a escolha de uma forma de onda para que as amostras $t_i$ tenham um valor estabelecido individualmente.
+Um único período dentre as ondas básicas pode ser utilizado para a especificação da nota da seguinte forma:
+
+seja $f$ a frequência da nota, $\delta_f=1/f$ o seu comprimento de onda e $L_i^{f,\, \delta_f} $ a sequência que descreve um período da onda $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ (ver equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada}; $R_i^f$ é uma onda real amostrada):
+
+\begin{equation}
+L_i^{f , \delta_f } = \{ l_i^f \}_0^{\delta_f . f_a -1}=\{ l_i^f \}_0^{\lambda_f-1}
+\end{equation}
+
+Então a sequência $T_i$ consistirá em uma nota de duração $\Delta$ e frequência $f$ se:
+
+\begin{equation}
+T_i^{f,\; \Delta}=\{t_i^f\}_0^{f_a . \Delta}=\{l^f_{i\,\%\left(\frac{f_a}{f}\right)}\}_0^{f_a . \Delta}
+\end{equation}
 
 \subsubsection{Usos musicais}
+
+Em posse da nota básica, podemos montar estruturas musicais com
+sequências destas partículas. Caso somemos $N$ sequências ($T_{k,i}=\{t_{k,i}\}, \;\; k \in 0...N-1$) de mesmo tamanho, seus conteúdos espectrais serão sobrepostos e a isso damos o nome de mixagem:
+
+\begin{equation}
+\{t_i\}=\left \{ \sum_{k=0}^{N-1}t_{k,i} \right \}
+\end{equation}
+
+Pode-se completar com zeros para somar sequências de tamanhos diferentes.
+O resultado da mixagem de notas musicais é a harmonia musical, cujos intervalos entre as frequências e os acordes de notas simultâneas regem aspectos subjetivos e abstratos da música e sua apreciação.
+
+As sequências podem também ser concatenadas no tempo.
+Caso as sequências $\{t_{k,i}\}_0^{\Lambda_k-1}$ de tamanhos $\Lambda_k$  representem notas musicais, sua concatenação em $t_i$ resultará em melodias ou sequências musicais mais simples:
+
+\begin{equation}
+\{t_i\}=\{t_{k,i}\}, \;\; k\text{ menor inteiro } : \quad \Lambda_k > i -\sum_{j=0}^{k-1}\Lambda_j
+\end{equation}
+
+As montagens musicais \emph{reduced-fi} demonstra este uso de justaposição temporal das notas, resultando em uma peça homofônica.
+O princípio vertical está demonstrado nos \emph{quadros sonoros}.
+Ambas as peças estão em código Python no APÊNDICE XX e podem ser escutadas e vizualizadas nos links [][][].
+
+A seguir desenvolveremos as notas musicais com parâmetros que evoluem no tempo, como glissandos e envoltórias.
+A filtragem de componentes e os ruídos finalizam sobre as notas musicais.
+A estruturação musical destas notas é então abordada com vistas ás trajetórias cíclicas ou que convirjam ou diverjam.
 

### 2012-08-23 devEresultados.tex
realizada primeira parte da musica no som digitalizado

@@
 
-Como $a_k = a_{\Lambda -k}\;\;$ e $\;\;b_k = - b_{\Lambda -k}$:
-
-\begin{equation}\label{equivalenciasModulos}
-\sqrt{a_k^2 + b_k^2} = \sqrt{a_{\Lambda - k}^2 + b_{\Lambda -k}^2} \;\;, \quad \;\; \forall \quad 1 \leq k \leq \tau  \\
-\end{equation}
-
-\begin{equation}\label{equivalenciasFases}
-tg^{-1}\left(\frac{b_k}{a_k}\right)=-tg^{-1}\left(\frac{b_{\Lambda -k}}{a_{\Lambda - k}}\right)\;\;,\quad \;\; \forall \quad 1 \leq k \leq \tau
-\end{equation}
-
-Com $k \in \mathbb{N}$.
-A observação da equação de reconstrução para o sinal real ~\ref{moduloEfase} em conjunto com as equivalências dos módulos e fases ~\ref{equivalenciasModulos} e ~\ref{equivalenciasFases}
-mostra a combinação em fase das componentes em $w_k$ e $w_{\Lambda-k}$:
-
-\begin{equation}
-t_i = \frac{a_0}{\Lambda} + \frac{2}{\Lambda}\sum_{k=1}^{\tau}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]+ \frac{ a_{\Lambda/2+1}}{\Lambda}.(1-\Lambda\% 2)
-\end{equation}
-
-Assim, a exemplo da figura ~\ref{fig:amostras3}, a transformada de Fourier de 3 amostras possui 2 coeficientes imaginários que possuem quantidades iguais de energia na mesma frequência.
-
-Com 4 amostras, podemos representar 1 ou 2 frequências em proporções diferentes.
-A figura ~\ref{fig:amostras4} mostra uma forma de onda de 4 amostras e suas duas componentes.
-Note que as contribuições individuais se somam de fato na forma de onda original, e que as curvaturas maiores são fruto da frequência mais aguda.
-
-A figura ~\ref{fig:formas4} explicita os harmônicos em 4 amostras nas formas de onda básicas das equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} e figura ~\ref{fig:formasDeOnda}.
-Todas consistem em apenas 1 senóide, com excessão da dente de serra que possui os harmônicos pares.
-
 \begin{figure}[h!]
@@
 
+Como $a_k = a_{\Lambda -k}\;\;$ e $\;\;b_k = - b_{\Lambda -k}$:
+
+\begin{equation}\label{equivalenciasModulos}
+\sqrt{a_k^2 + b_k^2} = \sqrt{a_{\Lambda - k}^2 + b_{\Lambda -k}^2} \;\;, \quad \;\; \forall \quad 1 \leq k \leq \tau  \\
+\end{equation}
+
+\begin{equation}\label{equivalenciasFases}
+tg^{-1}\left(\frac{b_k}{a_k}\right)=-tg^{-1}\left(\frac{b_{\Lambda -k}}{a_{\Lambda - k}}\right)\;\;,\quad \;\; \forall \quad 1 \leq k \leq \tau
+\end{equation}
+
+Com $k \in \mathbb{N}$.
+A observação da equação de reconstrução para o sinal real ~\ref{moduloEfase} em conjunto com as equivalências dos módulos e fases ~\ref{equivalenciasModulos} e ~\ref{equivalenciasFases}
+mostra a combinação em fase das componentes em $w_k$ e $w_{\Lambda-k}$:
+
+\begin{equation}
+t_i = \frac{a_0}{\Lambda} + \frac{2}{\Lambda}\sum_{k=1}^{\tau}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]+ \frac{ a_{\Lambda/2+1}}{\Lambda}.(1-\Lambda\% 2)
+\end{equation}
+
 \begin{figure}[h!]
@@
 
-A figura ~\ref{fig:amostras6} mostra as decomposições senoidais para o caso de 6 amostras.
-A figura ~\ref{fig:formas6} decompõe as formas de onda básicas.
-Note que as ondas se diferenciam espectralmente.
+Assim, a exemplo da figura ~\ref{fig:amostras3}, a transformada de Fourier de 3 amostras possui 2 coeficientes imaginários que possuem quantidades iguais de energia na mesma frequência.
+
+Com 4 amostras, podemos representar 1 ou 2 frequências em proporções diferentes.
+A figura ~\ref{fig:amostras4} mostra uma forma de onda de 4 amostras e suas duas componentes.
+Note que as contribuições individuais se somam de fato na forma de onda original, e que as curvaturas maiores são fruto da frequência mais aguda.
+
+A figura ~\ref{fig:formas4} explicita os harmônicos em 4 amostras nas formas de onda básicas das equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} e figura ~\ref{fig:formasDeOnda}.
+Todas consistem em apenas 1 senóide, com excessão da dente de serra que possui os harmônicos pares.
+
+A figura ~\ref{fig:amostras6} mostra as decomposições senoidais para o caso de 6 amostras. A figura ~\ref{fig:formas6} decompõe as formas de onda básicas para o caso de 6 amostras.
+ Note que neste caso de 6 amostras as ondas se diferenciam espectralmente: as quadrada e triangular possuem as mesmas componentes, mas em proporções diferentes, já a dente de serra possui uma componente a mais.
 
@@
 
-Escolhamos $f$ tal que $f$ divida $f_a$   \footnote{Esta escolha facilita as demonstrações pois os períodos são quantidades inteiras de amostras.
+Escolhamos $f$ tal que $f$ divida $f_a$  
+ \footnote{Esta escolha facilita as demonstrações pois os períodos são quantidades inteiras de amostras.
 Transporemos esta limitação na prática com o uso de uma tabela de amostragem ou com o uso direto de funções periódicas,e.g.
-$t_i=sen(x.i)$}.
+$t_i=sen(x.i)$.
+Veja próxima sessão.}.
 Uma sequência $T_i$ de amostras sonoras separadas por $\delta_a=1/f_a$ descreve uma nota musical de frequência $f$ Herz e duração $\Delta$ segundos se, e somente se, possuir a periodicidade $\lambda_f=f_a/f$ e tamanho $f_a .
@@
 As sequências podem também ser concatenadas no tempo.
-Caso as sequências $\{t_{k,i}\}_0^{\Lambda_k-1}$ de tamanhos $\Lambda_k$  representem notas musicais, sua concatenação em $t_i$ resultará em melodias ou sequências musicais mais simples:
+Caso as sequências $\{t_{k,i}\}_0^{\Lambda_k-1}$ de tamanhos $\Lambda_k$  representem notas musicais, sua concatenação em $t_i$ resultará em sequências musicais simples ou melodias:
 
 \begin{equation}
-\{t_i\}=\{t_{k,i}\}, \;\; k\text{ menor inteiro } : \quad \Lambda_k > i -\sum_{j=0}^{k-1}\Lambda_j
-\end{equation}
-
-As montagens musicais \emph{reduced-fi} demonstra este uso de justaposição temporal das notas, resultando em uma peça homofônica.
-O princípio vertical está demonstrado nos \emph{quadros sonoros}.
+\{t_i\}_0^{\sum\Delta}=\{t_{k,i}\}_0^{\sum\Delta}, \;\; k\text{ menor inteiro } : \quad \Lambda_k > i -\sum_{j=0}^{k-1}\Lambda_j
+\end{equation}
+
+A montagem musical \emph{reduced-fi} demonstra de forma isolada este uso de justaposição temporal das notas, resultando em uma peça homofônica.
+O princípio vertical está demonstrado nos \emph{quadros sonoros}, sons estáticos com espectros peculiares.
 Ambas as peças estão em código Python no APÊNDICE XX e podem ser escutadas e vizualizadas nos links [][][].
 
-A seguir desenvolveremos as notas musicais com parâmetros que evoluem no tempo, como glissandos e envoltórias.
-A filtragem de componentes e os ruídos finalizam sobre as notas musicais.
-A estruturação musical destas notas é então abordada com vistas ás trajetórias cíclicas ou que convirjam ou diverjam.
+Com isso estabelecemos nossa nota musical básica no som digital.
+Trabalharemos a seguir estas unidades, a evolução temporal de seus conteúdos e desenvolveremos as notas musicais com parâmetros que evoluem no tempo, como glissandos e envoltórias.
+A filtragem de componentes e os ruídos finalizam sobre a constituição da nota musical como unidade isolada.
+A estruturação musical destas notas é então abordada do ponto de vista das estruturas fora do tempo (como as escalas), trajetórias cíclicas ou que convirjam ou divirjam.
+
+\clearpage
 

### 2012-08-27 scripts/amostras6formas.py
atualizadas figuras de decomposicoes espectrais de 4 e 6 amostras

@@ -87,7 +87,7 @@ for i in xrange(4):
     #p.legend(loc="upper right")
 
     p.xlim(-1.2,6.2)
-    p.ylim(-1.1,1.1)
+    p.ylim(-1.3,1.3)
 
     p.xticks((-1,0,1,2,3,4,5,6),(r"$-\delta_a$",r"$0$",r"$\delta_a$",r"$2\delta_a$",r"$3\delta_a$",r"$4\delta_a$",r"$5\delta_a$",r"$6\delta_a$"))
 

### 2012-08-29 devEresultados.tex
atualizadas figuras de decomposicoes espectrais de 4 e 6 amostras
pequena correcao do ~lu e apontadas 2 sessoes
iniciada parte 2 com intro e lookuptable

@@
 %\subsection{Caracterização da nota musical discretizada}
-\subsection{Caracterização da nota musical}
+\subsection{Caracterização da nota musical digital}
 %\subsection{Caracterização da nota musical digital}
@@
     \caption{Formas de onda básicas em 4 amostras}
-        \includegraphics[width=\textwidth]{figuras/amostras4formas}
+        \includegraphics[width=\textwidth]{figuras/amostras4formas__}
         \label{fig:formas4}
@@
     \caption{Formas de onda básicas em 6 amostras}
-        \includegraphics[width=\textwidth]{figuras/amostras6formas_}
+        \includegraphics[width=\textwidth]{figuras/amostras6formas___}
         \label{fig:formas6}
@@
 
-\begin{equation}
+\begin{equation}\label{periodoUnico}
 L_i^{f , \delta_f } = \{ l_i^f \}_0^{\delta_f . f_a -1}=\{ l_i^f \}_0^{\lambda_f-1}
@@
 \clearpage
+
+\section{Variações na nota musical básica}
+
+Nossa nota musical digital básica está bem definida com os parâmetros:
+duração, altura, intensidade (volume) e timbre. Esta é uma definição
+útil e paradigmática, mas de forma alguma esgota o que entendemos por
+uma nota musical.
+
+Em primeiro lugar, as características da nota se modificam no decorrer
+da própria nota, i.e. em uma nota real os parâmetros
+não são absolutamente fixos. Por exemplo, em uma nota de piano
+de 3 segundos, a intensidade tem início abrupto e decaimento progressivo
+além de variações do espectro instantâneo, com harmônicos que
+decaem antes dos outros e alguns que aparecem com o tempo.
+Estas variações não necessariamente são levadas
+em conta na síntese sonora, mas constituem-se em guias para a
+síntese de sons com interesses musicais pois é como eles
+se apresentam na natureza.
+
+Explorar todas as formas pelas quais estas variações ocorrem está fora
+do escopo de qualquer trabalho dada a considerável sensibilidade do ouvido humano
+e a complexidade da nossa cognição. Mesmo assim, podemos apontar
+os recursos primários para estas variações dos parâmetros da nota
+básica.
+
+\subsection{Tabela de Busca}
+
+Mais conhecido pelo termo em inglês, a \emph(Lookup table) (ou simplesmente
+LUT), é um procedimento simples que consiste em criar uma tabela
+de referência e coletar as amostras destas tabelas segundo a 
+necessidade\footnote{Em contraposição ao cálculo das amostras individuais
+para cada caso.}. Em nosso caso, usaremos tabelas de busca para os
+sons criados, as notas. Assim, estas tabelas são unidimensionais e com
+valores reais.
+
+O procedimento de busca em tabelas é visto geralmente como um
+artifício de otimização. Na música seu uso transcende este
+primeiro, facilitando as operações e permitindo que um único
+período de onda possa ser usado para sintetizar sons em toda a gama
+audível de frequências.
+
+Descrevemos esta busca. Seja $\Lambda_{LUT}$ o tamanho 
+da tabela e $L_i^{LUT}$ a tabela de elementos $l_i$ de um
+período de onda qualquer (veja equação ~\ref{periodoUnico}). Um som
+de duração $T_i^{f,\Delta}$ de frequência $f$ e duração $\Delta$
+pode ser feito a partir de $L_i^{LUT}$ da seguinte forma:
+
+\begin{equation}
+T_i^{f,\Delta}=\{t_i^f\}_0^{f_a . \Delta} = \{ l_{\lfloor i . \gamma \rfloor  \% \Lambda} \}_0^{f_a . \Delta} , \gamma = \frac{f . \Lambda}{f_a}
+\end{equation}
+
+\section{Organização de notas musicais em música}
 

### 2012-08-30 devEresultados.tex
discorrido sobre lookup tables e rascunhado roteiro sobre variacoes na nota digital

--- 2012-08-23 devEresultados.tex
@@
 \clearpage
+
+\subsection{Variações na nota musical básica}
+
+Nossa nota musical digital básica está bem definida com os parâmetros:
+duração, altura, intensidade (volume) e timbre. Esta é uma modelagem
+útil e paradigmática, mas de forma alguma esgota o que entendemos por
+uma nota musical.
+
+Em primeiro lugar, as características da nota se modificam no decorrer
+da própria nota, i.e. em uma nota real os parâmetros
+não são absolutamente fixos. Por exemplo, em uma nota de piano
+de 3 segundos, a intensidade tem início abrupto e decaimento progressivo
+além de variações do espectro instantâneo, com harmônicos que
+decaem antes dos outros e alguns que aparecem com o tempo.
+Estas variações não necessariamente são levadas
+em conta na síntese sonora, mas constituem-se em guias para a
+síntese de sons para usos musicais pois é como eles
+se apresentam na natureza\footnote{A regra de ouro
+aqui é: para que um som isolado disperte interesse
+por si só, faça como que tenha variações internas.}.
+
+Explorar todas as formas pelas quais estas variações ocorrem está fora
+do escopo de qualquer trabalho dada a considerável sensibilidade do ouvido humano
+e a complexidade da nossa cognição sonora. Mesmo assim, podemos apontar
+os recursos primários para estas variações dos parâmetros da nota
+básica.
+
+\subsubsection{Tabela de Busca}
+
+Mais conhecido pelo termo em inglês, a \emph{Lookup table} (ou simplesmente
+LUT), é um procedimento simples que consiste em criar uma tabela
+de referência e coletar as amostras destas tabelas segundo a 
+necessidade\footnote{Em contraposição ao cálculo das amostras individuais
+para cada caso.}. Em nosso caso, usaremos tabelas de busca para os
+sons criados, as notas. Assim, estas tabelas são unidimensionais e com
+valores reais. Além disso, a tabela consiste nas amostras
+de um período de onda, que usamos para sintetizar períodos
+de tamanhos diferentes, resultando em frequências diferentes.
+
+O procedimento de busca em tabelas é visto geralmente como um
+artifício de otimização. Na música seu uso transcende este
+primeiro, facilitando as operações e permitindo que um único
+período de onda possa ser usado para sintetizar sons em toda a banda
+de frequências audíveis.
+
+Descrevemos esta busca. Seja $\widetilde{\Lambda}$ o tamanho 
+da tabela e $\widetilde{L_i} = \{ \widetilde{l}_i \}_0^{\widetilde{\Lambda} -1}$ a tabela de elementos $\widetilde{l_i}$ de um
+período de onda qualquer (veja equação ~\ref{periodoUnico}). Uma sequência
+$T_i^{f,\Delta}$ com amostras de um som de frequência $f$ e duração $\Delta$
+pode ser feita a partir de $\widetilde{L_i}$ da seguinte forma:
+
+\begin{equation}
+T_i^{f,\Delta}=\{t_i^f\}_0^{f_a .
+\Delta-1} = \{ \widetilde{l}_{\lfloor i .
+\gamma \rfloor  \% \widetilde{\Lambda} } \}_0^{f_a .
+\Delta-1}\; , \quad \text{onde} \;\; \gamma = \frac{f .
+\widetilde{\Lambda}}{f_a}
+\end{equation}
+
+Ou seja, dada a tabela de busca, basta termos a sequência de índices corretos
+para sintetizar o som. No termo $i . \gamma$ não importa se tomamos o inteiro
+como a parte inteira da multiplicação, como o maior inteiro acima da multiplicação
+ou se arredondamos. Além disso, para fins de síntese em $44,1 kHz$,
+ o padrão é usar $\widetilde{\Lambda} = 1024$ amostras, pois já não gera ruído
+ relevante no espectro audível.
+
+ A expressão que define a variável $\gamma$ pode ser facilmente compreendida da
+ seguinte forma: a variável $i$ é acrescida de $f_a$ em $1$ segundo (pela
+ definição de $f_a$ e $i$, pois temos $f_a$ amostras $i$ em $1$ segundo). Caso a dividamos por $f_a$, teremos $\frac{i}{f_a}$,
+fração esta que é acrescida de $1$ a cada $1$ segundo. Multiplicado pelo comprimento da
+ tabela $\widetilde{\Lambda}$, teremos $\frac{i . \widetilde{\Lambda}}{f_a}$
+ que resulta na varredura completa da tabela $\widetilde{L_i}$ em 
+ 1 segundo com o acrescimo de $\widetilde{\Lambda}$ em $1$ segundo. Por fim,
+ se multiplicarmos pela frequência $f$ que queremos, teremos $\frac{i . f . \widetilde{\Lambda}}{f_a}$
+ que resulta em $f$ varreduras completas da tabela $\widetilde{L_i}$ em $1$ segundo, i.e. a sequência
+ resultante apresenta frequência $f$.
+
+Importantes considerações: $f$ é qualquer, só há limitantes nas frequências
+graves quando $\widetilde{\Lambda}$ não é suficientemente grande para a taxa de amostragem
+$f_a$\footnote{$\widetilde{\Lambda} = 1024$ é suficiente para sintetizar
+sequências $T_i^f$ com qualquer $f$ no espectro.}. O procedimento de busca em tabela
+é bastante barato, substituindo cálculos por buscas simples (por isso geralmente
+é entendido como um processo de otimização). Salvo quando assinalado,
+usaremos este procedimento para todos os casos cabíveis a seguir pois
+simplifica diversas rotinas e é a forma usual de realizar estas sínteses.
+
+\subsubsection{Variações incrementais de frequência e intensidade}
+
+\subsubsection{Filtros}
+
+\subsubsection{Ruídos}
+
+\subsubsection{Usos musicais}
+
+\subsection{Organização de notas musicais em música}
 

### 2012-08-30 devEresultados.tex
feitas melhoras no lookuptable, finalizada

@@
 do escopo de qualquer trabalho dada a considerável sensibilidade do ouvido humano
-e a complexidade da nossa cognição sonora. Mesmo assim, podemos apontar
-os recursos primários para estas variações dos parâmetros da nota
+e a complexidade da nossa cognição sonora. Desta forma, apontaremos
+recursos primários para estas variações dos parâmetros na nossa nota
 básica.
+
+Iniciaremos com uma técnica que não resulta propriamente
+em variações internas da nota, mas simplificará esta exposição
+sem restringir o escopo, além de ser um notável recurso de otimização
+computacional.
 
@@
 T_i^{f,\Delta}=\{t_i^f\}_0^{f_a .
-\Delta-1} = \{ \widetilde{l}_{\lfloor i .
-\gamma \rfloor  \% \widetilde{\Lambda} } \}_0^{f_a .
-\Delta-1}\; , \quad \text{onde} \;\; \gamma = \frac{f .
-\widetilde{\Lambda}}{f_a}
+\Delta-1} = \{ \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} } \}_{0}^{f_a .
+\Delta-1}\; , \quad \text{onde} \;\; \gamma_i = \left \lfloor i .
+f \frac{ \widetilde{\Lambda}}{f_a} \right \rfloor
 \end{equation}
@@
 Ou seja, dada a tabela de busca, basta termos a sequência de índices corretos
-para sintetizar o som. No termo $i . \gamma$ não importa se tomamos o inteiro
-como a parte inteira da multiplicação, como o maior inteiro acima da multiplicação
+para sintetizar o som. Para o cálculo do inteiro $\gamma_i$, convencionamos
+atribuir a ele o maior inteiro menor que $i.f\frac{\widetilde{\Lambda}}{f_a}$,
+mas  não importa se o tomamos como a parte inteira da multiplicação, como o maior inteiro acima da multiplicação
 ou se arredondamos. Além disso, para fins de síntese em $44,1 kHz$,
@@
 
- A expressão que define a variável $\gamma$ pode ser facilmente compreendida da
+ A expressão que define a variável $\gamma_i$ pode ser facilmente compreendida da
  seguinte forma: a variável $i$ é acrescida de $f_a$ em $1$ segundo (pela
@@
  que resulta em $f$ varreduras completas da tabela $\widetilde{L_i}$ em $1$ segundo, i.e. a sequência
- resultante apresenta frequência $f$.
+ resultante apresenta a frequência fundamental $f$.
 
@@
 sequências $T_i^f$ com qualquer $f$ no espectro.}. O procedimento de busca em tabela
-é bastante barato, substituindo cálculos por buscas simples (por isso geralmente
+é computacionalmente bastante barato, substituindo cálculos por buscas simples (por isso geralmente
 é entendido como um processo de otimização). Salvo quando assinalado,
-usaremos este procedimento para todos os casos cabíveis a seguir pois
+a seguir usaremos este procedimento para todos os casos cabíveis pois
 simplifica diversas rotinas e é a forma usual de realizar estas sínteses.

### 2012-08-31 devEresultados.tex
arrumacoes da variacao na nota

@@
 para sintetizar o som. Para o cálculo do inteiro $\gamma_i$, convencionamos
-atribuir a ele o maior inteiro menor que $i.f\frac{\widetilde{\Lambda}}{f_a}$,
-mas  não importa se o tomamos como a parte inteira da multiplicação, como o maior inteiro acima da multiplicação
-ou se arredondamos. Além disso, para fins de síntese em $44,1 kHz$,
+atribuir a ele o maior inteiro menor que $i.f\frac{\widetilde{\Lambda}}{f_a}$.
+Esta aproximação introduz um ruído, mas conseguimos que este ruído seja desprezível
+mesmo se o tomarmos como o maior inteiro acima da multiplicação
+ou se arredondarmos por outros métodos. Além disso, para fins de síntese, em $f_a=44,1 kHz$
  o padrão é usar $\widetilde{\Lambda} = 1024$ amostras, pois já não gera ruído
@@
  seguinte forma: a variável $i$ é acrescida de $f_a$ em $1$ segundo (pela
- definição de $f_a$ e $i$, pois temos $f_a$ amostras $i$ em $1$ segundo). Caso a dividamos por $f_a$, teremos $\frac{i}{f_a}$,
+ definição de $f_a$ e $i$, pois temos $f_a$ amostras $t_i$ em $1$ segundo). Caso a dividamos por $f_a$, teremos $\frac{i}{f_a}$,
 fração esta que é acrescida de $1$ a cada $1$ segundo. Multiplicado pelo comprimento da
- tabela $\widetilde{\Lambda}$, teremos $\frac{i . \widetilde{\Lambda}}{f_a}$
+ tabela $\widetilde{\Lambda}$, teremos $i \frac{\widetilde{\Lambda}}{f_a}$
  que resulta na varredura completa da tabela $\widetilde{L_i}$ em 
- 1 segundo com o acrescimo de $\widetilde{\Lambda}$ em $1$ segundo. Por fim,
- se multiplicarmos pela frequência $f$ que queremos, teremos $\frac{i . f . \widetilde{\Lambda}}{f_a}$
+ 1 segundo com o acréscimo de $\widetilde{\Lambda}$ em $1$ segundo. Por fim,
+ se multiplicarmos pela frequência $f$ que queremos, teremos $i . f \frac{\widetilde{\Lambda}}{f_a}$
  que resulta em $f$ varreduras completas da tabela $\widetilde{L_i}$ em $1$ segundo, i.e. a sequência
@@
 Importantes considerações: $f$ é qualquer, só há limitantes nas frequências
-graves quando $\widetilde{\Lambda}$ não é suficientemente grande para a taxa de amostragem
-$f_a$\footnote{$\widetilde{\Lambda} = 1024$ é suficiente para sintetizar
+graves quando o tamanho da tabela $\widetilde{\Lambda}$ não é suficientemente grande para a taxa de amostragem
+$f_a$\footnote{Como citado acima, $\widetilde{\Lambda} = 1024$ é suficiente para sintetizar
 sequências $T_i^f$ com qualquer $f$ no espectro.}. O procedimento de busca em tabela
@@
 a seguir usaremos este procedimento para todos os casos cabíveis pois
-simplifica diversas rotinas e é a forma usual de realizar estas sínteses.
+simplifica diversas rotinas e é computacionalmente coerente.
 
 \subsubsection{Variações incrementais de frequência e intensidade}
+
+Nossa percepção segue a lei logarítmica...
+
+Em uma nota de duração $\Delta = \frac{\Lambda}{f_a}$, a frequência $f=f_i$ varia de $f_0$ até $f_{\Lambda -1}$
+linearmente. Podemos então escrever que:
+
+\begin{equation}\label{freqLinear}
+f_i=f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \quad ,\quad \quad i \;\in\; \mathbb{N}, \quad i \;\in\; [0,\Lambda-1]
+\end{equation}
+
+\begin{equation}\label{indiceLinear}
+\gamma_i=\left \lfloor i .
+f_i\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor i  \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right ] \right \rfloor   \quad ,\quad \quad i \;\in\; \mathbb{N}, \quad i \;\in\; [0,\Lambda-1]
+\end{equation}
+
+\begin{equation}\label{serieLinear}
+\{t_i^{\overline{f_0,f_{\Lambda-1}}}\}_0^{\Lambda-1}=\{\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\}_0^{\Lambda-1}
+\end{equation}
+
+Desta forma, podemos calcular os elementos $t_i^{\overline{f_0,f_{\Lambda-1}}}$ com base no período $\{\widetilde{l}_i\}_0^{\Lambda-1}$.
+
+As equações ~\ref{freqLinear}, ~\ref{indiceLinear} e ~\ref{serieLinear} são relativas à progressão linear
+da frequência. Nossa percepção segue a progressão logarítmica, então uma progressão de frequência
+\emph{percebida} como linear segue a progressão logarítmica:
+
+\begin{equation}
+f_i=f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}
+\end{equation}
+
+\begin{equation}
+f_i=f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}
+\end{equation}
 

### 2012-09-01 devEresultados.tex
deduzida e explicitada a frequência de cada amostra e os resultantes da lookup table para variacoes tidas como lineares pelo ouvido mas que sao logaritmicas

@@
 
-Desta forma, podemos calcular os elementos $t_i^{\overline{f_0,f_{\Lambda-1}}}$ com base no período $\{\widetilde{l}_i\}_0^{\Lambda-1}$.
+Desta forma, podemos calcular os elementos $t_i^{\overline{f_0,f_{\Lambda-1}}}$
+com base no período $\{\widetilde{l}_i\}_0^{\Lambda-1}$. Note que na equação
+~\ref{freqLinear}, o termo $f_{\Lambda -1} - f_0$ é o âmbito do contra-domínio de $f_i$
+e que $\Lambda -1$ é âmbito do domínio. Como $i$ varre o domínio $[0,\Lambda-1]$ e
+temos o deslocamento $f_0$ na imagem, $f_i$ varre o contra-domínio enquanto $i$
+na medida em que $i$ varre o domínio\footnote{Estas digressões são importantes
+para entendermos o significado musical do comportamento destas sequências de amostras}.
 
@@
 da frequência. Nossa percepção segue a progressão logarítmica, então uma progressão de frequência
-\emph{percebida} como linear segue a progressão logarítmica:
-
-\begin{equation}
-f_i=f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}
-\end{equation}
+\emph{percebida} como linear segue a progressão logarítmica. Fazendo a varredura com $i$
+de $0$ até $\Lambda-1$, podemos dizer que: $f_i=f_0 . 2^{\frac{i}{\Lambda-1} n_8}$ onde 
+$n_8=\log_2\frac{f_{\Lambda-q}}{f_0}$ é o número de oitavas entre $f_0$ e $f_{\Lambda-1}$.
+De forma que $f_i=f_0 . 2^{\frac{i}{\Lambda-1}\log_2\frac{f_{\Lambda-1}}{f_0}}=
+f_0 . 2^{\log_2\left ( \frac{f_{\Lambda-1}}{f_0} \right )^{\frac{i}{\Lambda-1}}}=
+f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}$. Portanto,
+nossas equações de referência para a síntese de transições de frequência tidas como
+lineares pelo ouvido são:
 

### 2012-09-01 devEresultados.tex
descrita a variacao de volume pelas amostras de um som digitalizado

@@
 
-\begin{equation}
+\begin{equation}\label{potencia}
 pot(T_i^{\Delta})=\frac{\sum_{i=0}^{\Delta . f_a -1} t_i^2}{\Delta . f_a}
@@
 
+Ou seja, é necessário pouco mais que triplicar a amplitude para que tenhamos volume drobrado.
+
 Estes valores servem de guia para os aumentos e diminuições dos valores absolutos que compõem as
-sequências de amostras sonoras que visam atingir efeitos musicais.
+sequências de amostras sonoras que visam atingir efeitos musicais. A conversão direta de decibeis
+em amplitude se dá da seguinte forma:
+
+\begin{equation}\label{ampDec}
+A = 10^{\frac{V_{dB}}{20}}
+\end{equation}
+
+Onde $A$ é o fator multiplicativo entre as médias dos valores absolutos das amplitudes dos sinais comparados.
 
@@
 
-\begin{equation}\label{serieLinear}
-\{t_i^{\overline{f_0,f_{\Lambda-1}}}\}_0^{\Lambda-1}=\{\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\}_0^{\Lambda-1}
-\end{equation}
-
-Desta forma, podemos calcular os elementos $t_i^{\overline{f_0,f_{\Lambda-1}}}$
+\begin{equation}\label{serieAmostralLin}
+\{t_i^{\;\overline{f_0,f_{\Lambda-1}}}\}_0^{\Lambda-1}=\{\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\}_0^{\Lambda-1}
+\end{equation}
+
+Desta forma, podemos calcular os elementos $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$
 com base no período $\{\widetilde{l}_i\}_0^{\Lambda-1}$. Note que na equação
@@
 
-As equações ~\ref{freqLinear}, ~\ref{indiceLinear} e ~\ref{serieLinear} são relativas à progressão linear
+As equações ~\ref{freqLinear}, ~\ref{indiceLinear} e ~\ref{serieAmostral} são relativas à progressão linear
 da frequência. Nossa percepção segue a progressão logarítmica, então uma progressão de frequência
@@
 f_i=f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}
+\end{equation}
+
+\begin{equation}\label{indiceExponencial}
+\gamma_i=\left \lfloor i .
+f_i\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor i f_0 \frac{\widetilde{\Lambda}}{f_a} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}} \right \rfloor   \;\; ,\quad \quad i \;\in\; \mathbb{N}, \quad i \;\in\; [0,\Lambda-1]
+\end{equation}
+
+\begin{equation}\label{serieAmostralLog}
+\{t_i^{\;\overline{f_0,f_{\Lambda-1}}}\}_0^{\Lambda-1}=\{\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\}_0^{\Lambda-1}
+\end{equation}
+
+Onde usamos a equação ~\ref{indiceExponencial} para calcular a sequência $\{\gamma_i\}_0^{\Lambda-1}$ de índices para para
+serem buscados na nossa tabela que contém um único período da forma de onda.
+Isso é explicitado na equação ~\ref{serieAmostralLog}, onde estabelecemos a relação
+ $t_i^{\;\overline{f_0,f_{\Lambda-1}}}=\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}$.
+Note que a equação ~\ref{serieAmostralLog} se mantém (permanece idêntica à equação ~\ref{serieAmostralLin}).
+
+O termo $\frac{i}{\Lambda-1}$ varre o intervalo $[0,1]$ e podemos elevá-lo a uma potência
+para que o início da transição seja mais suave ou mais abrupto (compensado pelo final
+da transição). Este procedimento é mais comum para o caso de variações de volume através da energia
+da onda vibratória\footnote{O aumento do volume pode ocorrer por diversas características, como
+a reverberação e a concentração de harmônicos agudos e a energia. A mais facilmente controlada
+é a energia da onda (veja equação ~\ref{potencia}) e a energia também pode variar de diferentes formas. A forma mais
+simples e conveniente geralmente é a variação de amplitude. A variação de energia sem variação de
+amplitude é comumente chamado de \emph{compressão sonora} e é muito útil e corriqueiro na prática
+de produção musical atual}. Nestes casos, o procedimento comum é multiplicar a sequência original
+(seja ela gerada ou pré-estabelecida) pela sequência $\left ( \frac{i}{\Lambda-1} \right )^\alpha$
+onde alpha é o coeficiente citado.
+
+Assim, para variações de amplitude, aplicamos as seguintes fórmulas:
+
+\begin{equation}
+\{a_i\}_0^{\Lambda-1}=\left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}=\left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1} \text{ com } a_0=1
+\end{equation}
+
+\begin{equation}
+T_i^{'}=T_i \odot A_i = \{t_i . a_i\}_0^{\Lambda-1}=\left \{ t_i . (a_{\Lambda-1} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
+\end{equation}
+
+Nosso intuito aqui é modificar a amplitude de um sinal original, seja qual for ele. Desta forma, tomamos
+$a_0=1$ de forma a iniciar a nova sequência com a amplitude original e então ir modificando com o tempo.
+Esta restrição faz com que nosso termo $a_{\Lambda-1}$ seja efetivamente a variação da amplitude.
+Aqui pode muito nos auxiliar a conversão de decibeis para amplitude apresentada na equação ~\ref{ampDec},
+resultando em:
+
+\begin{equation}
+T_i^{'}=\left \{ t_i . (10^{\frac{V_{dB}}{20}} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
 \end{equation}

### 2012-09-02 devEresultados.tex
escrito sobre variacoes incrementais de frequencia e amplitude, fechado com vibrato e tremolo

@@
 
-Nossa percepção segue a lei logarítmica...
+Segundo a lei de Weber e Fechner, nossa percepção tem uma relação logarítmica com
+o estímulo que a causa. Mesmo assim, dado o uso, explicitaremos também a variação
+linear. Começaremos por ela por razões didáticas.
 
@@
 
-As equações ~\ref{freqLinear}, ~\ref{indiceLinear} e ~\ref{serieAmostral} são relativas à progressão linear
+As equações ~\ref{freqLinear}, ~\ref{indiceLinear} e ~\ref{serieAmostralLin} são relativas à progressão linear
 da frequência. Nossa percepção segue a progressão logarítmica, então uma progressão de frequência
@@
 de $0$ até $\Lambda-1$, podemos dizer que: $f_i=f_0 . 2^{\frac{i}{\Lambda-1} n_8}$ onde 
-$n_8=\log_2\frac{f_{\Lambda-q}}{f_0}$ é o número de oitavas entre $f_0$ e $f_{\Lambda-1}$.
+$n_8=\log_2\frac{f_{\Lambda-1}}{f_0}$ é o número de oitavas entre $f_0$ e $f_{\Lambda-1}$.
 De forma que $f_i=f_0 . 2^{\frac{i}{\Lambda-1}\log_2\frac{f_{\Lambda-1}}{f_0}}=
@@
 Nosso intuito aqui é modificar a amplitude de um sinal original, seja qual for ele. Desta forma, tomamos
-$a_0=1$ de forma a iniciar a nova sequência com a amplitude original e então ir modificando com o tempo.
+$a_0=1$ para iniciar a nova sequência com a amplitude original e então ir modificando com o decorrer das amostras.
 Esta restrição faz com que nosso termo $a_{\Lambda-1}$ seja efetivamente a variação da amplitude.
+Caso $\alpha=1$, a variação de amplitude segue exatamente a progressão geométrica que caracteriza
+a percepção linear.
+
+Para resultar em uma amplificação linear, basta usarmos umas série $\{a_i\}$ diferente:
+
+\begin{equation}
+a_i=a_0 + (a_{\Lambda-1}-a_0)\frac{i}{\Lambda-1}
+\end{equation}
+
+Com $a_0=1$ para preservar a amplitude inicial original.
+
 Aqui pode muito nos auxiliar a conversão de decibeis para amplitude apresentada na equação ~\ref{ampDec},
@@
 \end{equation}
+
+para o caso geral de variações de amplitude segundo a progressão geométrica.
+
+\subsubsection{Tremolo e Vibrato}
+
+Enquanto o vibrato é uma variação periódica de altura (frequência),
+o tremolo é uma variação periódica de volume (intensidade).
+
+Para realizarmos o caso mais geral, façamos uma sequência $t_i'$
+de frequência $f'$ com o auxílio
+de uma segunda tabela $\widetilde{M}_i$ de tamanho $\widetilde{\Lambda}_M$ que apresente também
+um período de onda que oscila entre $[-1,1]$. Descrevemos um vibrato da seguinte forma:
+
+\begin{equation}
+\gamma_i'=\left \lfloor i f' \frac{\widetilde{\Lambda}_M}{f_a} \right \rfloor
+\end{equation}
+
+\begin{equation}
+t_i'=\widetilde{m}_{\gamma_i' \;\% \widetilde{\Lambda}_M}
+\end{equation}
+
+\begin{equation}
+f_i=f \left ( \frac{f + \mu }{f} \right )^{t_i'}
+\end{equation}
+
+\begin{equation}
+\gamma_i = \left \lfloor i . f_i \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor i \frac{\widetilde{\Lambda}}{f_a}f \left ( \frac{f + \mu }{f} \right )^{t_i'}  \right \rfloor
+\end{equation}
+
+\begin{equation}
+T_i^{f, vbr(f')}=\{ t_i^{f,vbr(f')} \}_0^{\Lambda-1}=\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \}_0^{\Lambda-1}
+\end{equation}
+
+Para a correta realização do vibrato, é importante atenção para as duas tabelas e sequências.
+A tabela $\widetilde{M}_i$ de tamanho $\widetilde{\Lambda}_M$ e a sequência de índices $\gamma_i'$ formam a sequência $t_i'$
+ que é o padrão da oscilação da frequência enquanto
+a tabela $\widetilde{L}_i$ de tamanho $\widetilde{\Lambda}$ e a sequência de índices $\gamma_i$ formam $t_i$ que é o som em si.
+
+O caso do tremolo é semelhante. Dado $f'$, $\gamma_i'$ e $t_i'$ permanecem os mesmos. Calculamos
+a sequência de amplitudes a serem multiplicadas pela sequência original $t_i$ da
+sequinte forma:
+
+\begin{equation}
+a_i=10^{t_i' \frac{V_{dB}}{20}} = a_{\text{máx}}^{t_i'}
+\end{equation}
+
+\begin{equation}
+T_i^{f, tr(f')}=\{ t_i^{f,tr(f')} \}_0^{\Lambda-1}=\{ t_i^{f} . a_i \}_0^{\Lambda-1}=\{t_i^f . a_{\text{máx}}^{t_i'}\}_0^{\Lambda-1}
+\end{equation}
+
+Onde $V_{dB}$ são os decibeis do tremolo e $a_{\text{máx}}$ é o ganho máximo de amplitude envolvido.
+A medição em decibeis é bastante conveniente neste caso pois o aumento máximo de amplitude
+não equivale à diminuição máxima relacionada, enquanto a diferença em decibeis se mantém.
 

### 2012-09-02 scripts/fourier.py

@@ -1,72 +1,0 @@
-#-*- coding: utf8 -*-
-# http://matplotlib.sourceforge.net/examples/api/legend_demo.html
-# 
-import pylab as p, numpy as n, scikits.audiolab as a
-f=n.fft.fft
-fi=n.fft.ifft
-a=[-1,1]
-#-1 * cos(n.pi*l)
-#p.plot([0,1],a,"bo") ##########3
-ii=n.linspace(-0.5*n.pi,1.5*n.pi,200)
-iii=n.linspace(-0.5,1.5,200)
-s=-n.cos(ii)
-#p.plot(iii,s,"m--") ###########3
-
-
-p.ylabel(r"amplitude $\rightarrow$")
-p.xlabel(r"tempo $\rightarrow$")
-#p.show()
-
-
-
-
-n3=n.random.rand(3)*2-1
-
-ff=f(n3)
-a0=n.real(ff[0])
-b0=n.imag(ff[0])
-
-ab1=n.abs(ff[1])
-a1=n.real(ff[1])
-b1=n.imag(ff[1])
-#fas=b1/a1
-fas=n.arctan(b1/a1) # fase
-fas=n.arctan(b1/a1)*3
-print("abs: %s, a1: %s, b1: %s, fas: %s" % (ab1,a1,b1,fas))
-
-#vfas=(n.sin(fas)-sin(fas))
-
-ii=n.linspace(-0.5*n.pi,1.5*n.pi,200)
-iii=n.linspace(-0.5,2.5,200)
-#s=-ab1*n.cos(ii-fas)
-#p.plot(iii,s,"m--")
-#p.plot(n3,"bo")
-
-s=(2/3.)*ab1*n.cos(ii)+a0/3; p.plot(iii,s,"m--"); p.plot(n3,"bo");p.show()
-
-
-p.xlim(-1.2,3.2)
-#p.xlim(0,T2*.56)
-p.ylim(-1.1,1.1)
-#p.show()
-
-
-
-d=[-.6,-.2,.2,.6]
-t=[-1,0,1,0]
-q=[-1,-1,1,1]
-
-da1=n.pi
-
-da2=n.pi/2
-
-das=n.linspace(-n.pi/2,-5*n.pi/2,4,endpoint=False)
-s1=n.sin(das)
-s2=n.array(a*2)
-
-#p.plot(s1)
-#p.plot(s2)
-#p.plot(s1+s2)
-#p.plot(d)
-#p.show()
-

### 2012-09-02 devEresultados.tex
terminadas variacoes incrementais, tremolo e vibrato, am e fm

@@
 
-Segundo a lei de Weber e Fechner, nossa percepção tem uma relação logarítmica com
+Segundo revela a lei de Weber e Fechner, nossa percepção tem uma relação logarítmica com
 o estímulo que a causa. Mesmo assim, dado o uso, explicitaremos também a variação
-linear. Começaremos por ela por razões didáticas.
+linear e começaremos descrevendo esta por razões didáticas.
 
@@
 e que $\Lambda -1$ é âmbito do domínio. Como $i$ varre o domínio $[0,\Lambda-1]$ e
-temos o deslocamento $f_0$ na imagem, $f_i$ varre o contra-domínio enquanto $i$
+temos o deslocamento $f_0$ na imagem, $f_i$ varre o contra-domínio 
 na medida em que $i$ varre o domínio\footnote{Estas digressões são importantes
@@
 As equações ~\ref{freqLinear}, ~\ref{indiceLinear} e ~\ref{serieAmostralLin} são relativas à progressão linear
-da frequência. Nossa percepção segue a progressão logarítmica, então uma progressão de frequência
-\emph{percebida} como linear segue a progressão logarítmica. Fazendo a varredura com $i$
+da frequência. Como assinalado, nossa percepção segue a progressão logarítmica:
+uma progressão de frequência
+\emph{percebida} como linear segue uma progressão exponencial\footnote{Ou,
+dito ainda de outra forma, uma progressão geométrica da frequência
+é percebido como uma progressão aritmética de alturas}.
+
+Fazendo a varredura com $i$
 de $0$ até $\Lambda-1$, podemos dizer que: $f_i=f_0 . 2^{\frac{i}{\Lambda-1} n_8}$ onde 
@@
 
-Para realizarmos o caso mais geral, façamos uma sequência $t_i'$
+Iniciemos pelo vibrato. Para realizarmos o caso mais geral, façamos uma sequência $t_i'$
 de frequência $f'$ com o auxílio
 de uma segunda tabela $\widetilde{M}_i$ de tamanho $\widetilde{\Lambda}_M$ que apresente também
-um período de onda que oscila entre $[-1,1]$. Descrevemos um vibrato da seguinte forma:
-
-\begin{equation}
+um período de onda que oscila entre $[-1,1]$. Descrevemos uma sequência $t_i^{vbr(f')}$ com
+um vibrato de frequência $f'$ e profundidade  $\mu$ (quantos herz é a oscilação no pico superior)
+ou $\nu$ (quantos semitons) da seguinte forma:
+
+\begin{equation}\label{vbrGamma}
 \gamma_i'=\left \lfloor i f' \frac{\widetilde{\Lambda}_M}{f_a} \right \rfloor
@@
 
-\begin{equation}
+\begin{equation}\label{vbrAux}
 t_i'=\widetilde{m}_{\gamma_i' \;\% \widetilde{\Lambda}_M}
@@
 
-\begin{equation}
-f_i=f \left ( \frac{f + \mu }{f} \right )^{t_i'}
-\end{equation}
-
-\begin{equation}
-\gamma_i = \left \lfloor i . f_i \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor i \frac{\widetilde{\Lambda}}{f_a}f \left ( \frac{f + \mu }{f} \right )^{t_i'}  \right \rfloor
-\end{equation}
-
-\begin{equation}
+\begin{equation}\label{vbrF}
+f_i=f \left ( \frac{f + \mu }{f} \right )^{t_i'}=f . 2^{t_i'\frac{\nu}{12}}
+\end{equation}
+
+\begin{equation}\label{vbrGamma}
+\gamma_i = \left \lfloor i .
+f_i \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor i \frac{\widetilde{\Lambda}}{f_a}f \left ( \frac{f + \mu }{f} \right )^{t_i'}  \right \rfloor= \left \lfloor i \frac{\widetilde{\Lambda}}{f_a}f .
+2^{t_i'\frac{\nu}{12}}  \right \rfloor
+\end{equation}
+
+\begin{equation}\label{vbrT}
 T_i^{f, vbr(f')}=\{ t_i^{f,vbr(f')} \}_0^{\Lambda-1}=\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \}_0^{\Lambda-1}
@@
 a tabela $\widetilde{L}_i$ de tamanho $\widetilde{\Lambda}$ e a sequência de índices $\gamma_i$ formam $t_i$ que é o som em si.
-
-O caso do tremolo é semelhante. Dado $f'$, $\gamma_i'$ e $t_i'$ permanecem os mesmos. Calculamos
+As variáveis $\mu$ e $\nu$ quantificam a intensidade do vibrato: $\mu$ é uma medida direta da quantidade
+de Herz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação.
+Note que $\nu=\log_{2}\frac{f+\mu}{f} $ é conveniente neste caso pois o aumento máximo de frequência
+não equivale à diminuição máxima, mas a variação de semitons se mantém.
+
+O caso do tremolo é semelhante e $f'$, $\gamma_i'$ e $t_i'$ permanecem os mesmos. Calculamos
 a sequência de amplitudes a serem multiplicadas pela sequência original $t_i$ da
@@
 
-\begin{equation}
+\begin{equation}\label{trA}
 a_i=10^{t_i' \frac{V_{dB}}{20}} = a_{\text{máx}}^{t_i'}
@@
 
-\begin{equation}
-T_i^{f, tr(f')}=\{ t_i^{f,tr(f')} \}_0^{\Lambda-1}=\{ t_i^{f} . a_i \}_0^{\Lambda-1}=\{t_i^f . a_{\text{máx}}^{t_i'}\}_0^{\Lambda-1}
-\end{equation}
-
-Onde $V_{dB}$ são os decibeis do tremolo e $a_{\text{máx}}$ é o ganho máximo de amplitude envolvido.
+\begin{equation}\label{trT}
+T_i^{tr(f')}=\{ t_i^{tr(f')} \}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}=\{t_i .10^{t_i' \frac{V_{dB}}{20}}    \}_0^{\Lambda-1}=\{t_i . a_{\text{máx}}^{t_i'}\}_0^{\Lambda-1}
+\end{equation}
+
+Onde $V_{dB}$ é profundidade da oscilação em decibeis do tremolo e $a_{\text{máx}}$ é o ganho máximo de amplitude envolvido.
 A medição em decibeis é bastante conveniente neste caso pois o aumento máximo de amplitude
 não equivale à diminuição máxima relacionada, enquanto a diferença em decibeis se mantém.
+
+Com as equações ~\ref{vbrT}, ~\ref{vbrGamma}, ~\ref{vbrF}, ~\ref{vbrAux} e  ~\ref{vbrGamma}
+são utilizadas para síntese de frequência modulada (FM) nos casos em que $f'$ é
+maior que $20Hz$ onde as oscilações de frequência resultam em padrões complexos
+de componentes espectrais harmônicas e não harmônicas. As equações ~\ref{trA}
+e ~\ref{trT} (junto às ~\ref{vbrT} e ~\ref{vbrGamma} para a realização da sequência
+auxiliar $t_i'$) resultam na síntese de amplitude modulada (AM)\footnote{As sínteses
+FM e AM são técnicas com diversas amplicações na música e também em outras áreas,
+principalmente nas telecomunicações para transferência de informações
+via ondas eletromagnéticas. O leitor interessado pode visitar os itens relacionados
+da nossa bibliografia para uma abordagem musical (veja especialmente ~\ref{} e ~\ref{}}.
 

### 2012-09-04 devEresultados.tex
encaminhada discussao sobre filtros, falta fechar melhor

@@
 As sequências podem também ser concatenadas no tempo.
-Caso as sequências $\{t_{k,i}\}_0^{\Lambda_k-1}$ de tamanhos $\Lambda_k$  representem notas musicais, sua concatenação em $t_i$ resultará em sequências musicais simples ou melodias:
+Caso as sequências $\{t_{k,i}\}_0^{\Lambda_k-1}$ de tamanhos $\Lambda_k$  representem $k$ notas musicais, sua concatenação em $t_i$ resultará em sequências musicais simples ou melodias:
 
@@
 FM e AM são técnicas com diversas amplicações na música e também em outras áreas,
-principalmente nas telecomunicações para transferência de informações
+especialmente conhecidos são os usos telecomunicações para transferência de informações
 via ondas eletromagnéticas. O leitor interessado pode visitar os itens relacionados
-da nossa bibliografia para uma abordagem musical (veja especialmente ~\ref{} e ~\ref{}}.
+da nossa bibliografia para uma abordagem musical (veja especialmente ~\ref{} e ~\ref{})}.
 
 \subsubsection{Filtros}
+O recurso básico para modificar a contribuição das componentees espectrais
+na composição de um som é a filtragem. Os filtros proporcionam
+ganhos e atenuações diferentes para porções diferentes do espectro,
+possibilitando que haja diminuição ou remoção de ruídos e modificações
+de timbres quando desejável. A aplicação de filtros pode
+ser parte constituinte da síntese ou poder ser feita posteriormente
+como parte dos processos tipicamente chamados de tratamento sonoro.
+No estudo de filtros, a concepção usual é a de multiplicação de espectros
+(espectro do som multiplicado pelo espectro do filtro).
+
+É importante frizar que a elaboração de filtros
+constitui uma área reconhecidamente complexa, com literatura
+e pacotes de software dedicados. Desta maneira, o texto a seguir
+é um resumo do que é utilizado na prática e de forma alguma cobre
+todas as possibilidades. Apontaremos utilizações destes filtros junto
+às especificações.
+
+\begin{itemize}
+
+\item Filtros de resposta ao impulso finita (FIR)
+
+Esta classe de filtros é conhecida
+pela sua sigla FIR (do inglês Finite Impulse Response)
+e é caracterizada por possuir uma representação amostral
+finita. Esta representação amostral é chamada
+de 'resposta ao impulso'. Os filtros FIR são aplicados ao som
+digitalizado pela convolucão do som com sua representação
+amostral:
+
+\begin{equation}
+convolucao\;de\;filtro\;fir
+\end{equation}
+
+Importante aqui é notar a vantegem e desvantagem do filtro FIR.
+Podemos especificar a filtragem intensionada através do perfil
+espectral a ser multiplicado pelo espectro que do som. Através da transformada
+de Fourier inversa, obtemos a resposta ao impulso. Assim,
+filtros FIR permitem que apliquemos os filtros exatos
+que desejarmos.
+
+\item Filtros de resposta ao impulso infinita (IIR)
+
+A exemplo dos filtros FIR, esta classe de filtros é
+conhecida pela sigla IIR (do inglês Infinite Impulse Response)
+e é caracterizada por possuir uma representação amostral (resposta
+ao impulso) infinita. Sua aplicação é usualmente feita pela equação
+a diferenças:
+
+\begin{equation}
+equacao\;a\;diferencas\;filtro\;iir
+\end{equation}
+
+Como pode ser notado, os filtros IIR são muito mais requerem
+menos processamento computacional para serem aplicados. Além disso,
+podem ser aplicados com maior localidade. Por este motivo, são os mais usuais
+para aplicações em geral, mas a obtenção de filtros IIR com características
+desejadas muitas vezes revela-se uma tarefa nada trivial\footnote{Encorajamos
+o leitor interessado a visitar nossa bibliografia, em especial os livros
+~\ref{} e ~\ref{}.}.
+
+\end{itemize}
+
+Os fitros possuem ainda uma segunda classificação,
+conforme o perfil espectral que aplicam. A seguir
+apontamos os tipos mais cruciais para a prática musical.
+
+\begin{enumerate}
+\item Passa-baixas
+
+\item Passa-altas
+\item Passa-banda
+\item Rejeita-banda
+\item Nó
+\end{enumerate}
 
@@
 \subsection{Organização de notas musicais em música}
+
+\clearpage
 

### 2012-09-05 devEresultados.tex
save geral para dar um tapa melhor nos filtros, mudando abordagem

@@
 na medida em que $i$ varre o domínio\footnote{Estas digressões são importantes
-para entendermos o significado musical do comportamento destas sequências de amostras}.
+para entendermos o significado musical do comportamento destas sequências de amostras.}.
 

### 2012-09-07 devEresultados.tex
feita a sessao sobre o filtro fir, faltam as figuras

@@
 
-\subsubsection{Filtros}
-O recurso básico para modificar a contribuição das componentees espectrais
-na composição de um som é a filtragem. Os filtros proporcionam
+\subsubsection{Aplicação de Filtros Digitais}
+Explicitaremos a aplicação de filtros por convolução
+e por equação a diferenças. As possibilidades
+aqui são variadas e a complexidade facilmente
+foge ao escopo deste trabalho\footnote{A elaboração de filtros
+constitui uma área reconhecidamente complexa, com literatura
+e pacotes de software dedicados. 
+Recomendamos ao leitor
+interessado nossa bibliografia, em especial ~\ref{} e ~\ref{}.},
+de forma que focamos em explicitar o que ocorre com as sequências
+de amostras nestes dois casos.
+
+Os equalizadores proporcionam
 ganhos e atenuações diferentes para porções diferentes do espectro,
 possibilitando que haja diminuição ou remoção de ruídos e modificações
-de timbres quando desejável. A aplicação de filtros pode
+de timbre quando desejável. A aplicação de filtros pode
 ser parte constituinte da síntese ou poder ser feita posteriormente
 como parte dos processos tipicamente chamados de tratamento sonoro.
-No estudo de filtros, a concepção usual é a de multiplicação de espectros
-(espectro do som multiplicado pelo espectro do filtro).
-
-É importante frizar que a elaboração de filtros
-constitui uma área reconhecidamente complexa, com literatura
-e pacotes de software dedicados. Desta maneira, o texto a seguir
-é um resumo do que é utilizado na prática e de forma alguma cobre
-todas as possibilidades. Apontaremos utilizações destes filtros junto
-às especificações.
 
@@
 
-\item Filtros de resposta ao impulso finita (FIR)
-
-Esta classe de filtros é conhecida
+\item  Convolução e filtros de resposta ao impulso finita (FIR)
+
+Os filtros aplicados por convolução são conhecidos
 pela sua sigla FIR (do inglês Finite Impulse Response)
-e é caracterizada por possuir uma representação amostral
-finita. Esta representação amostral é chamada
-de 'resposta ao impulso'. Os filtros FIR são aplicados ao som
-digitalizado pela convolucão do som com sua representação
-amostral:
+e são caracterizados por possuirem uma representação amostral
+finita no tempo. Esta representação amostral é chamada
+de 'resposta ao impulso' $h_i$. Os filtros FIR são aplicados ao som
+digitalizado pela convolucão do som com sua 
+resposta ao impulso. Para nossos fins, esta
+convolução fica definida como:
 
 \begin{equation}
-convolucao\;de\;filtro\;fir
-\end{equation}
-
-Importante aqui é notar a vantegem e desvantagem do filtro FIR.
-Podemos especificar a filtragem intensionada através do perfil
-espectral a ser multiplicado pelo espectro que do som. Através da transformada
-de Fourier inversa, obtemos a resposta ao impulso. Assim,
-filtros FIR permitem que apliquemos os filtros exatos
-que desejarmos.
+t_i'=(t*h)_i=\sum_{j=0}^{min(\Lambda_h,i)}h_{j} . t_{i-j}=\sum_{j=max(i-\Lambda_h,0)}^{i}t_j . h_{i-j}
+\end{equation}
+
+Ou seja, o som $\{t_i'\}$ resultante da convolução de $\{t_i\}$ com a resposta ao impulso $\{h_i\}$
+tem cada i-ésima amostra $t_i$ substituída pela soma de suas últimas amostras $t_{i-j}$
+multiplicadas pela resposta ao impulso em si ($h_i$). Isso é facilmente
+percebido pela visualização da convolução de sequências arbitrárias e dispomos
+para este fim a figura \ref{}.
+
+Com este procedimento podemos aplicar reverberadores, equalizadores, delays
+e vários outros tipos de filtros para fins de tratamento sonoro ou
+efeitos musicais/artísticos.
+
+Para a obtenção da resposta ao impulso, pode-se recorrer a medições
+físicas ou à síntese direta do filtro. Para citar um exemplo
+de medições diretas, uma resposta ao impulso para a aplicação
+de reverberação pode resultar da gravação acústica do ambiente ao disparar
+um estalo (cuja gravação se aproxima da resposta ao impulso diretamente) ou ao disparar uma
+varredura de frequência (cuja gravação se aproxima da resposta em frequência).
+Ambas resultam em respostas ao impulso
+que, convoluídas com a sequência sonora, resultam nela própria
+com uma reverberação que se assemelha ao ambiente de medição.
+
+Como exemplo de respostas ao impulso sintetizadas, pode-se
+descrever um perfil espectral, realizar a transformada inversa
+de fourier e convoluir o resultante com o som, realizando
+uma filtragem cujo perfil é próximo ao especificado inicialmente\footnote{Especialmente
+se for grande o número de amostras utilizado para especificar o perfil espectral
+do filtro.}. Outro exemplo simples é a aplicação de delays através
+da convolução do som com uma resposta ao impulso que possui um impulso
+em cada reincidência do som na linha de delays intensionada. Estes dois exemplos
+estão mostrados na figura ~\ref{}.
 

### 2012-09-07 devEresultados.tex
feita parte da nota musical inteira, mas o final soh topifiquei

@@
 
-Os equalizadores proporcionam
-ganhos e atenuações diferentes para porções diferentes do espectro,
-possibilitando que haja diminuição ou remoção de ruídos e modificações
-de timbre quando desejável. A aplicação de filtros pode
+A aplicação de filtros pode
 ser parte constituinte da síntese ou poder ser feita posteriormente
@@
 em cada reincidência do som na linha de delays intensionada. Estes dois exemplos
-estão mostrados na figura ~\ref{}.
+estão mostrados na figura ~\ref{}. Note que, pelo método citado,
+filtros dos tipos citados abaixo em ~\ref{passaBaixas},~\ref{passaAltas} podem ser sintetizados.
 
@@
 
-A exemplo dos filtros FIR, esta classe de filtros é
+Esta classe de filtros é
 conhecida pela sigla IIR (do inglês Infinite Impulse Response)
@@
 \begin{equation}
-equacao\;a\;diferencas\;filtro\;iir
-\end{equation}
-
-Como pode ser notado, os filtros IIR são muito mais requerem
-menos processamento computacional para serem aplicados. Além disso,
-podem ser aplicados com maior localidade. Por este motivo, são os mais usuais
-para aplicações em geral, mas a obtenção de filtros IIR com características
-desejadas muitas vezes revela-se uma tarefa nada trivial\footnote{Encorajamos
-o leitor interessado a visitar nossa bibliografia, em especial os livros
-~\ref{} e ~\ref{}.}.
+t_i' = \frac{1}{a_0}\left ( \sum_{j=0}^Jb_j . t_{i-j} \sum_{k=1}^Ka_k . t_{i-k}' \right )
+\end{equation}
+
+Apontamos filtros IIR paradigmáticos e bastante usuais abaixo. Existem
+diversos métodos e ferraemtas para a elaboração de filtros IIR,
+listamos abaixo uma seleção com fins didáticos e para consulta futura\footnote{Vale também
+apontar que todos os filtros abaixo são utilizados para equalização.}.
+São filtros de primeira ordem e Butterworth.
 
 \end{itemize}
-
-Os fitros possuem ainda uma segunda classificação,
-conforme o perfil espectral que aplicam. A seguir
-apontamos os tipos mais cruciais para a prática musical.
 
@@
 \item Passa-baixas
-
 \item Passa-altas
@@
 \item Nó
+\item Biquad
 \end{enumerate}
@@
 \subsubsection{Ruídos}
+Geração estatística de ruídos, distribuições, nomes, equações de geração.
 
 \subsubsection{Usos musicais}
+Oscilacoes, tremolos e vibratos. Variacoes
+progressivas das oscilacoes dos tremolos e vibratos
+como AM e FM.
+ADSR
+Linhas de delays
+Filtragem dependente do tempo
+Músicas: monolitos, minimum-fi
 

### 2012-09-10 devEresultados.tex
encaminhada parte 3 com escalas e acordes

@@
 
-Apontamos filtros IIR paradigmáticos e bastante usuais abaixo. Existem
-diversos métodos e ferraemtas para a elaboração de filtros IIR,
+Apontamos filtros IIR paradigmáticos e bastante úteis e usuais abaixo. Existem
+diversos métodos e ferrametas para a elaboração de filtros IIR,
 listamos abaixo uma seleção com fins didáticos e para consulta futura\footnote{Vale também
 apontar que todos os filtros abaixo são utilizados para equalização.}.
-São filtros de primeira ordem e Butterworth.
+São filtros de primeira ordem. Rotinas excelentes para cálculo de coeficientes
+de filtros Chebichev\footnote{Filtros Butterworth e Elípticos podem
+ser considerados como casos específicos dos Filtros do tipo Chebichev.} podem
+ser encontradas nas nossas referências, em especial em ~\ref{JOS} e ~\ref{dspguide}.
+Para uma exposição aprofundada do assunto, recomendamos ~\ref{Openheim}.
 
@@
 \begin{enumerate}
-\item Passa-baixas
-\item Passa-altas
-\item Passa-banda
-\item Rejeita-banda
-\item Nó
-\item Biquad
+\item Passa-baixas: polo simples com $a_0=0,15$ e $b_1=0,85$, fórmula geral para $x=e^{-2\pi f_c},\; x \in [0,1]$:
+
+\begin{equation}
+\begin{split}
+a_0 & =  1-x \\
+b_1 & =  x
+\end{split}
+\end{equation}
+
+\item Passa-altas: polo simples com $a_0=0,93$, $a_1=-0.93$ e $b_1=0.86$ para $x=e^{-2\pi f_c},\; x \in [0,1]$.
+
+\begin{equation}
+\begin{split}
+a_0 & =  \frac{1+x}{2} \\
+a_1 & =  -\frac{1+x}{2} \\
+b_1 & =  x
+\end{split}
+\end{equation}
+
+%\item Passa-banda
+%\item Rejeita-banda
+\item Nó: dadas a frequência $f$ e a largura de banda $BW$ (medida em 0,707 da amplitude),
+ambos dados como fracoes de $f_a$, portanto $f,BW \in [0,0,5]$,
+e as variáveis auxiliares $K$ e $R$ definidas como\footnote{A menor largura
+de banda que pode ser utilizada com precisão simples é por volta de $BW=0,0003$}:
+
+\begin{equation}
+\begin{split}
+R & = 1 - 3BW \\
+K & = \frac{1-2R\cos(2\pi f) + R^2}{2 - 2 \cos (2 \pi f)}
+\end{split}
+\end{equation}
+
+Temos o filtro nó passa banda:
+
+\begin{equation}
+\begin{split}
+a_0 & =  1 - K \\
+a_1 & =  2(K-R)\cos (2\pi f) \\
+a_2 & =  R^2-K \\
+b_1 & =  2R \cos (2\pi f) \\
+b_2 & =  -R^2
+\end{split}
+\end{equation}
+
+e o nó rejeita banda:
+
+\begin{equation}
+\begin{split}
+a_0 & =  K \\
+a_1 & =  -2K\cos (2\pi f) \\
+a_2 & =  K \\
+b_1 & =  2R \cos (2\pi f) \\
+b_2 & =  -R^2
+\end{split}
+\end{equation}
+
+\item Biquad: pela especificação de uma frequência central, da qualidade
+e da intensidade do filtro, este filtro é simples e usual para áudio,
+permitindo ajustes mais finos. Diversas receitas podem ser encontradas
+na literatura, recomendamos especialmente as diferentes especificações
+em ~\ref{musicDSP} e ~\ref{dspguide}.
+
 \end{enumerate}
@@
 Filtragem dependente do tempo
+cascatas de filtros
 Músicas: monolitos, minimum-fi
 
+\clearpage
 \subsection{Organização de notas musicais em música}
+Seja $ S=\{T_{i,1},T_{i,2}...T_{i,N}\} =\bigcup_{j \in [1,N]} T_{i,j} $ uma sequência de $N$ eventos
+musicais. Para simplificar, sejam notas musicais estes eventos musicais.
+Considerando $S$ como uma estrutura musical, existem diversas técnicas
+para tornar estas estruturas interessantes e agradáveis.
+
+\subsubsection{Afinação, escalas e acordes}
+Quando dobramos a frequência, temos uma oitava ascendente $f_2=2f_1$.
+Doze semitons equidistantes para o ouvido formam uma oitava,
+portanto $f_2=2^{\frac{1}{12}}f_1$ forma um semitom ascendente.
+O fator $\varepsilon=2^{\frac{1}{12}}$ forma uma grade de notas
+no espectro audível, em que as frequências fundamentais possíveis
+estão separadas por intervalos múltiplos de $\varepsilon$\footnote{Este
+é um modelo idealizado, os intrumentos reais possuem desvios das frequências
+fundamentais como citadas}.
+
+Esta divisão da oitava em doze notas é o cânone da música ocidental clássica,
+além de usos cerimoniais/religiosos e étnicos. Nesta afinação, chamada cromática
+de temperamento igual, temos 5 divisões perfeitamente simétricas, consideradas
+escalas pela simetria interna e usos fáceis que disso resultam. Descrevamos
+as escalas como sequências de fatores aos quais $\varepsilon$ é elevado para resultar
+no fator a ser multiplicado por $f_0$:
+
+\begin{equation}
+\begin{split}
+\text{cromática} = E_i^c = \{e_i^c\}_0^{11} =  \{0,1,2,3,4,5,6,7,8,9,10,11\} = \{i\}_0^{11}\\
+\text{tons inteiros} = E_i^t = \{e_i^t\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
+\text{terças menores} = E_i^{tm} = \{e_i^{tm}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
+\text{terças maiores} = E_i^{tM} = \{e_i^{tM}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
+\text{trítonos} = E_i^{tt} = \{e_i^{tt}\}_0^{1} = \{ 0, 6 \} = \{6.i\}_0^1
+\end{split}
+\end{equation}
+
+Assim, a terceira nota da escala em tons inteiros com $f_0=200Hz$
+é $f_3=\varepsilon^{e_3^t} . f_0 = 2^{\frac{6}{12}} . 200 \approxeq 282.843 Hz$. Estas
+'escalas' ou padrões, geram estruturas estáveis pelas simetrias internas, e podem ser
+repetidas de forma eficiênte e sustentada. Discutiremos mais sobre simetrias e voltaremos
+a estas divisões da oitava por igual.
+
+As escalas diatonicas também são facilmente especificadas, assim como
+as escalas tonais maior e menor:
+
+\begin{equation}
+\begin{split}
+\text{menor natural} = \text{modo eólico} = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
+\text{menor harmônica} = E_i^{mh} = \{e_i^{mh}\}_0^6 = \{0,2,3,5,7,8,11\} \\
+\text{menor melódica} = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
+\text{modo lócrio} = E_i^{mlo} = \{e_i^{mlo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
+\text{maior} = \text{modo jônico} = E_i^M = \{e_i^M\}_0^6 = \{0,2,4,5,7,9,11\} \\
+\text{modo dórico} = E_i^{md} = \{e_i^{md}\}_0^6 = \{0,2,3,5,7,9,10\} \\
+\text{modo frígio} = E_i^{mf} = \{e_i^{mf}\}_0^6 = \{0,1,3,5,7,8,10\} \\
+\text{modo lídio} = E_i^{ml}=\{e_i^{ml}\}_0^6 = \{0,2,4,6,7,9,11\} \\
+\text{modo mixolídio} = E_i^{mmi} = \{e_i^{mmi}\}_0^6 = \{0,2,4,5,7,9,10\}
+\end{split}
+\end{equation}
+
+Todas as escalas diatônicas (salvo as menores harmônica e melódica)
+seguem o padrão de intervalos sucessivos
+tom, tom, semitom, tom, tom, tom, semitom. De forma
+que podemos escrever:
+\begin{equation}
+\begin{split}
+d_i & =\{2,2,1,2,2,2,1\} \\
+e_0 & =0 \\
+e_i & =d_{(i+\kappa)\%7}+e_{i-1} \quad para \;\;  i > 0
+\end{split}
+\end{equation}
+
+Com $\kappa \in \mathbb{N}$. Por exemplo, consideremos o modo lídio. Uma breve
+inspeção revela que $e_i^{ml}=d_{(i+2)\%7}+e_{i-1}^{ml}$ com $\kappa=2$. Para cada
+modo, temos um único valor de $\kappa$ em qualquer intervalo $\in [x,x+6], \forall x \in \mathbb{N}$.
+
+Várias outras escalas e sequências são facilmente representadas desta forma. Em especial,
+lembramos aqui das pentatônicas e os modos de transposição limitados de messiaen ~\ref{}.
+
+A ocorrência simultânea de notas é observada musicalmente em grande medida
+através dos acordes. Listamos os acordes triádicos sem inversão e na forma fechada,
+com a fundamental em $0$:
+
+\begin{equation}
+\begin{split}
+\text{tríade maior} = A_i^M= \{a_i^M\}_0^2=\{0,4,7\} \\ 
+\text{tríade menor} = A_i^m = \{a_i^m\}_0^2=\{0,3,7\} \\
+\text{tríade diminuta} = A_i^d = \{a_i^d\}_0^2=\{0,3,6\} \\
+\text{tríade aumentada} = A_i^a = \{a_i^a\}_0^2=\{0,4,8\}
+\end{split}
+\end{equation}
+
+Nestes casos, seguindo as mesmas restrições, basta acrescentar $10$ para ter o acorde
+com sétima menor ou $11$ para resultar em sétima maior. Para realizar inversões e posições abertas,
+basta contar o número de semitons a partir da nota mais grave.
+
+Para a realização de microtonalidade, podemos usar reais não inteiros
+para descrever a sequência de alturas ou modificar o fator $\varepsilon$
+e usar inteiros ainda. Por exemplo, uma afinação muito
+interessante por contemplar a série harmônica quase perfeitamente,
+é proposta na forma da divisão da oitava em $53$ notas, de forma
+que $\varepsilon=2^{\frac{1}{53}}$. As sequências de alturas
+continuam sendo compostas por inteiros com o novo $\varepsilon$.
+Note que se $S_i$ é uma sequência de alturas $s_i$ e mudamos de $\varepsilon_1$ para
+$\varepsilon_2$,
+resultamos em uma nova sequência $S_i'$ com $s_i'=s_i \frac{\varepsilon_1}{\varepsilon_2}$.
+A música tradicional tailandesa utiliza uma divisão da oitava em sete notas $\varepsilon=2^{\frac{1}{7}}$,
+resultando intervalos que pouco se assemelham às possibilidades da
+divisão de doze notas.
+
+\subsubsection{Rítmo}
+
+\subsubsection{Estruturas direcionais}
+Arcos, sequências convergentes ou divergentes.
+
+\subsubsection{Estruturas cíclicas}
+Arcos. permutacoes
+
+\subsubsection{Usos Musicais}
+Algumas restrições funcionais possíveis: manter-se em harmonias triádicas e encadear os acordes
+que possuam relações de quintas justas dentre as fundamentais ou dentre uma fundamental
+e a terça do acorde seguinte.
+
+Explorações microtonais, oitava em 5 notas etc
+FIGGUS
 

### 2012-09-15 devEresultados.tex
iniciada exposicao sistematica sobre ritmo

@@
 e usar inteiros ainda. Por exemplo, uma afinação muito
-interessante por contemplar a série harmônica quase perfeitamente,
+interessante por contemplar a série harmônica consideravelmente precisa,
 é proposta na forma da divisão da oitava em $53$ notas, de forma
@@
 \subsubsection{Rítmo}
+A noção rítmica é dependente de eventos separados por durações. Estes eventos são discerníveis se separados
+por ao menos $50-63ms$. Para que a separação entre eles possa ser apreciada, ela deve ser um pouco maior,
+por volta de $100ms$ ($10Hz$)(\ref{microsound}). Podemos sumarizar a transição do domínio da apreciação
+de durações como alturas para a apreciação em rítmo da seguinte forma(~\ref{alfaix, microsound}):
+
+\begin{center}
+{\tiny 
+\begin{tabular}{ l | r r r r   r r r    r r r || r r || r r r r r r }
+\hline
+-           & \multicolumn{10}{c}{$\underleftarrow{\text{\bf zona de percepção temporal em rítmo}}$} & \multicolumn{2}{c}{transição} & \multicolumn{3}{c}{-} \\
+duração (s) & {\bf ...}     & {\bf 32,}     & {\bf 16,}   & {\bf 8,}  & {\bf 4,}   & {\bf 2,}   & {\bf 1,}   & {\bf 1/2,} & {\bf 1/4,} & {\bf 1/8,} & $\frac{1}{16}=62,5ms$ , & $\frac{1}{20}=50ms$ & {\color{Gray} 1/40} & {\color{Gray} 1/80  } & {\color{Gray} 1/160 } & {\color{Gray} 1/320 } & {\color{Gray} 1/640 } & {\color{Gray} ...
+} \\
+frequência (Hz) & {\color{Gray} ...} & {\color{Gray} 1/32,}   & {\color{Gray} 1/16,} & {\color{Gray} 1/8,} & {\color{Gray} 1/4,} & {\color{Gray} 1/2,} &  {\color{Gray} 1,}  & {\color{Gray} 2,}   & {\color{Gray} 4,}   & {\color{Gray} 8,}    & 16,  & 20   & {\bf 40}   & {\bf 80}   & {\bf 160}   & {\bf 320}   & {\bf 640}   & {\bf ...} \\
+-           & \multicolumn{10}{c}{ - } & \multicolumn{2}{c}{transição} & \multicolumn{6}{c}{$\overrightarrow{\text{\bf zona de percepção temporal em altura}}$} \\
+\hline
+\label{altRit}
+%\ca
+
+\end{tabular}
+}
+\end{center}
+
+A banda de durações marcada como transição está minimizada, pois os limites não são bem definidos
+e são dependentes da pessoa e de características do som como o conteúdo espectral e intensidade.
+
+A métrica rítmica costuma se basear em uma duração básica chamada pulso. O pulso tipicamente
+compreende durações entre $0,25-1,5s$ ($40-240BPM$) que se costuma associar às durações
+entre batidas do coração ou entre os passos ao caminhar.
+
+O pulso é subdividido em partes iguais e também é 
+repetido sequencialmente. Estas relações (de divisão e de concatenação) costumam
+seguir relações de números inteiros de baixa ordem
+\footnote{Em ordem crescente de ocorrência na música
+escrita e étnica, podemos
+dizer que as divisões do pulso musical e seus agrupamentos
+sequenciais no tempo ocorrem com fatores 2, 3, depois seus múltiplos
+4 (2 grupos de 2) e 6 (dois grupos de 3 ou 3 grupos de 2)
+e então 8 (2 grupos de 4), 9 (3 grupos de 3) e por último os primos 5 e 7, completando 1-9 (~\ref{jocelyn}, ~\ref{favre}).}.
+
+As relações duais costumam a ocorrer em rítmos de dança
+e ocasiões festivas e são chamadas imperfeitas. As relações triádicas ocorrem
+mais na música ritualística e relacionada ao sagrado
+e são ditas perfeitas.
+
+Seja $P_i^j$ a $i$-ésimo unidade de pulsos no nivel $j$ de agrupamento. Assim,
+$P^1_{10}$ é o décimo pulso, $P^{2}_3$ é a terceira unidade de agrupamento de pulsos,
+$P^{-1}_2$ é a segunda unidade da subdivisão do pulso. Especial atenção para
+os limites de $j$ pois as divisões do pulso devem resultar em durações apreciáveis
+como rítmo e também as junções do pulso deve resultar algo apreciável como, no máximo
+do nosso escopo música. Dito de outra forma: a duração de $P^{min(j)}_i$, para $\forall \; i$
+é maior que $50ms$ e as durações de $P^{\text{máx}(j)}_i$ \emph{somadas}  em $\forall i$
+devem resultar em uma duração menor do que alguns minutos ou, no máximo poucas horas.
+Sempre que $i$ soma três índices temos uma relação \emph{perfecta}, quando soma dois
+ou quatro índices temos uma relação \emph{imperfecta}.
+
+Podemos convencionar $P^{-1,1,3}_{3,2,2}$ como a terceira subdivisão $P^{-1}_3$ do segundo
+pulso $P^1_2$ do terceiro aglomerado $P^3_2$ de aglomerado de pulsos ($P^2_1$ omitido por simplicidade).
+Tipicamente, para um dado $j$, $i$ tende a se manter.
+
+A cada unidade ou conjunto de unidades $P_i^j$ pode ser associada uma sequência de amostras temporais $T_i$
+que formam uma nota musical. A associação de $P^{-1,1}_{ i_{\text{fraco}},x }$ e $P^{-1,1}_{0,x+1}$ é o que
+se chama de síncopa.
+
+Um compasso 4 por 4 possui $P^1_i$ com $i$ tipicamente contando 2, 4 ou 8 subdivisões. E
+em $P^2_i$, $i$ conta 4 unidades.
 

### 2012-09-16 devEresultados.tex
contemplando discussao ritmica com figura e discussao

@@
 A noção rítmica é dependente de eventos separados por durações. Estes eventos são discerníveis se separados
-por ao menos $50-63ms$. Para que a separação entre eles possa ser apreciada, ela deve ser um pouco maior,
-por volta de $100ms$ ($10Hz$)(\ref{microsound}). Podemos sumarizar a transição do domínio da apreciação
-de durações como alturas para a apreciação em rítmo da seguinte forma(~\ref{alfaix, microsound}):
+por ao menos $50-63ms$. Para que a separação entre eles possa ser apreciada, ela deve ser ainda um pouco maior,
+por volta de $100ms$ ($10Hz$)(tem um resumo interessante na página 42 do \ref{microsound}). Podemos sumarizar 
+a transição do domínio da apreciação
+de durações como alturas para a apreciação em rítmo da seguinte forma (~\ref{alfaix, microsound}):
 
@@
 A métrica rítmica costuma se basear em uma duração básica chamada pulso. O pulso tipicamente
-compreende durações entre $0,25-1,5s$ ($40-240BPM$) que se costuma associar às durações
-entre batidas do coração ou entre os passos ao caminhar.
+compreende durações entre $0,25-1,5s$ ($40-240BPM$). Na educação musical e estudos cognitivistas,
+costuma-se associar esta gama de velocidade de pulsação às durações entre batidas 
+do coração ou entre os passos ao caminhar.
 
 O pulso é subdividido em partes iguais e também é 
-repetido sequencialmente. Estas relações (de divisão e de concatenação) costumam
+repetido sequencialmente (veja a figura ~\ref{fig:pulsoSubAgl}. Estas relações (de divisão e de concatenação) costumam
 seguir relações de números inteiros de baixa ordem
@@
 4 (2 grupos de 2) e 6 (dois grupos de 3 ou 3 grupos de 2)
-e então 8 (2 grupos de 4), 9 (3 grupos de 3) e por último os primos 5 e 7, completando 1-9 (~\ref{jocelyn}, ~\ref{favre}).}.
+e então 8 (2 grupos de 4), 9 (3 grupos de 3) e por 
+último os primos 5 e 7, completando 1-9 (~\ref{jocelyn}, ~\ref{favre}).}.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Formas de onda musicais básicas}
+        \includegraphics[width=\textwidth]{figuras/formasDeOnda5}
+        \label{fig:formasDeOnda}
+\end{figure}
 
@@
 em $P^2_i$, $i$ conta 4 unidades.
+
+Nos tempos binários, os tempos fortes são
 

### 2012-09-16 scripts/metricaMusical.py

@@ -1,0 +1,14 @@
+#-*- coding: utf8 -*-
+import numpy as n, pylab as p
+
+resolucao=9*8 # 72
+
+p.plot((0,34),(1,1)); p.plot((36,72),(1,1))
+
+p.plot((0,16),(2,2)); p.plot((36,52),(2,2))
+p.plot((18,34),(2,2)); p.plot((54,72),(2,2))
+
+#p.plot(xs2,s2)
+p.ylim(-1,4)
+p.show()
+

### 2012-09-17 devEresultados.tex
feita figura para descricao da metrica intra e extra pulso

@@
     \centering
-    \caption{Formas de onda musicais básicas}
-        \includegraphics[width=\textwidth]{figuras/formasDeOnda5}
-        \label{fig:formasDeOnda}
+    \caption{Divisões e aglomerações do pulso musical para estabelecimento de métrica\footnote{Ao lado esquerdo temos as divisões da semínima estabelecida como pulso.
+Ao lado direito temos exemplos de fórmulas de compasso que especificam as mesmas métricas, mas na escala produzida pela algomeração e ocorrências do pulso musical.}}
+        \includegraphics[width=\textwidth]{figuras/metricaMusical}
+        \label{fig:pulsoSubAgl}
 \end{figure}

### 2012-09-17 scripts/metricaMusical.py

@@ -2,13 +2,52 @@
 import numpy as n, pylab as p
 
 resolucao=9*8 # 72
+#axL = p.subplot(2,1,1)
+fig=p.figure()
+ax1=fig.add_subplot(111)
+ax2=ax1.twinx()
 
-p.plot((0,34),(1,1)); p.plot((36,72),(1,1))
+#p.plot((0,34),(1,1)); p.plot((36,72),(1,1))
 
-p.plot((0,16),(2,2)); p.plot((36,52),(2,2))
-p.plot((18,34),(2,2)); p.plot((54,72),(2,2))
+#p.plot((0,16),(2,2)); p.plot((36,52),(2,2))
+#p.plot((18,34),(2,2)); p.plot((54,72),(2,2))
 
-#p.plot(xs2,s2)
-p.ylim(-1,4)
+ordem=[0,1,4,2,-1,5,-2,3,6]
+colors=['g','b','r','b','k','r','k','b','r']
+
+for i in xrange(9):
+    for j in xrange(i+1):
+        p.plot((j*resolucao/(i+1), (j+1)*resolucao/(i+1)-2),(ordem[i],ordem[i]),c=colors[i%len(colors)], linewidth=3)
+
+p.plot((-1000,-1000),'r',linewidth=3,label=u'compassos compostos, divisões ternárias, modus perfectus')
+p.plot((-1000,-1000),'b',linewidth=3,label=u'compassos simples, divisões binárias, modus imperfectus')
+p.plot((-1000,-1000),'k',linewidth=3,label=u'compassos em 5 e 7, quiálteras de 5 e 7')
+p.legend(loc="upper right")
+
+
+ax1.set_ylim(-3,8)
+ax2.set_ylim(-3,8)
+p.xlim(-4,74)
+
+ax1.set_xticks(())
+ax1.set_ylabel(r"$\leftarrow$"+u"divisões do pulso musical"+r"$\rightarrow$")
+ax2.set_ylabel(r"$\leftarrow$"+u"fórmulas de compasso"+r"$\rightarrow$")
+
+ax1.set_yticks((-2,-1,0,1,2,3,4,5,6))
[…]

### 2012-09-18 devEresultados.tex
terminada primeira versao do ritmo musical

@@
 \hline
--           & \multicolumn{10}{c}{$\underleftarrow{\text{\bf zona de percepção temporal em rítmo}}$} & \multicolumn{2}{c}{transição} & \multicolumn{3}{c}{-} \\
+           & \multicolumn{10}{c}{$\underleftarrow{\text{\bf zona de percepção de durações em rítmo}}$} & \multicolumn{2}{c}{transição} & \multicolumn{3}{c}{-} \\
 duração (s) & {\bf ...}     & {\bf 32,}     & {\bf 16,}   & {\bf 8,}  & {\bf 4,}   & {\bf 2,}   & {\bf 1,}   & {\bf 1/2,} & {\bf 1/4,} & {\bf 1/8,} & $\frac{1}{16}=62,5ms$ , & $\frac{1}{20}=50ms$ & {\color{Gray} 1/40} & {\color{Gray} 1/80  } & {\color{Gray} 1/160 } & {\color{Gray} 1/320 } & {\color{Gray} 1/640 } & {\color{Gray} ...
@@
 frequência (Hz) & {\color{Gray} ...} & {\color{Gray} 1/32,}   & {\color{Gray} 1/16,} & {\color{Gray} 1/8,} & {\color{Gray} 1/4,} & {\color{Gray} 1/2,} &  {\color{Gray} 1,}  & {\color{Gray} 2,}   & {\color{Gray} 4,}   & {\color{Gray} 8,}    & 16,  & 20   & {\bf 40}   & {\bf 80}   & {\bf 160}   & {\bf 320}   & {\bf 640}   & {\bf ...} \\
--           & \multicolumn{10}{c}{ - } & \multicolumn{2}{c}{transição} & \multicolumn{6}{c}{$\overrightarrow{\text{\bf zona de percepção temporal em altura}}$} \\
+           & \multicolumn{10}{c}{ - } & \multicolumn{2}{c}{transição} & \multicolumn{6}{c}{$\overrightarrow{\text{\bf zona de percepção de durações em altura}}$} \\
 \hline
-\label{altRit}
 %\ca
-
 \end{tabular}
@@
 
-A banda de durações marcada como transição está minimizada, pois os limites não são bem definidos
-e são dependentes da pessoa e de características do som como o conteúdo espectral e intensidade.
+A banda de durações marcada como transição está minimizada pois os limites não são bem definidos.
+A duração específica em que começamos a perceber nitidademente uma frequência fundamental ou uma separação entre as ocorrências é dependente da pessoa e de características do som como conteúdo espectral e intensidade.
 
 A métrica rítmica costuma se basear em uma duração básica chamada pulso. O pulso tipicamente
-compreende durações entre $0,25-1,5s$ ($40-240BPM$). Na educação musical e estudos cognitivistas,
-costuma-se associar esta gama de velocidade de pulsação às durações entre batidas 
+compreende durações entre $0,25-1,5s$ (respectivamente $240$ e $40BPM$). Na educação musical e estudos cognitivistas,
+costuma-se associar esta gama de frequências de pulsação às durações entre batidas 
 do coração ou entre os passos ao caminhar.
@@
 O pulso é subdividido em partes iguais e também é 
-repetido sequencialmente (veja a figura ~\ref{fig:pulsoSubAgl}. Estas relações (de divisão e de concatenação) costumam
+repetido sequencialmente. Estas relações (de divisão e de concatenação) costumam
 seguir relações de números inteiros de baixa ordem
@@
 dizer que as divisões do pulso musical e seus agrupamentos
-sequenciais no tempo ocorrem com fatores 2, 3, depois seus múltiplos
-4 (2 grupos de 2) e 6 (dois grupos de 3 ou 3 grupos de 2)
-e então 8 (2 grupos de 4), 9 (3 grupos de 3) e por 
-último os primos 5 e 7, completando 1-9 (~\ref{jocelyn}, ~\ref{favre}).}.
+sequenciais no tempo ocorrem com fatores 2, 4 e 8, depois 3, 6 (dois grupos de 3 ou 3 grupos de 2) e 9 (3 grupos de 3). Por 
+último os primos 5 e 7, completando 1-9 (~\ref{jocelyn}, ~\ref{favre}).
+Outras métricas são incidentes (divisões ou agrupamentos em 13, 17, etc), mas não são tão usuais.
+As métricas, por mais complexas que pareçam, costumam ser decompostas em divisões como estas (inclusive as de fator 5 e 7).}.
+Uma exposição esquemática está na figura ~\ref{fig:pulsoSubAgl}.
 
@@
 
-As relações duais costumam a ocorrer em rítmos de dança
+As relações duais (compassos simples e divisões binárias) costumam a ocorrer em rítmos de dança
 e ocasiões festivas e são chamadas imperfeitas. As relações triádicas ocorrem
-mais na música ritualística e relacionada ao sagrado
+incidem mais na música ritualística e relacionada ao sagrado
 e são ditas perfeitas.
 
-Seja $P_i^j$ a $i$-ésimo unidade de pulsos no nivel $j$ de agrupamento. Assim,
-$P^1_{10}$ é o décimo pulso, $P^{2}_3$ é a terceira unidade de agrupamento de pulsos,
-$P^{-1}_2$ é a segunda unidade da subdivisão do pulso. Especial atenção para
+As divisões mais fortes (acentuadas) são as que consistem nas cabeças dos tempos. Nas divisões binárias (2, 4 e 8 dos casos considerados aqui),
+as unidades consideradas fortes se revesam com as fracas
+(e.g. a divisão em 4 resulta em forte, fraco, (meio) forte, fraco).
+Nas divisões ternárias (3, 6 e 9 dos casos considerados neste escrito) a
+à unidade forte (primeira) se sucedem 2 unidades fracas (e.g. a divisão em 3 resulta em forte, fraco fraco)\footnote{A divisão em 6 é considerada composta
+ mas pode ocorrer também como uma divisão binária.
+ Uma divisão binária que sofre então uma divisão ternária
+ resulta em duas unidades divididas em três unidades cada: forte (subdividifo em forte fraco, fraco) e fraco (subdividido em forte, fraco, fraco).
+A outra forma de ocorrer a divisão em 6 é através de 
+uma divisão ternária que sofre então uma divisão binária, resultando em:
+uma unidade forte (subdividido em forte e fraco) e duas unidades fracas (subdivididas em forte e fraco).}.
+
+A acentuação em tempos fracos é chamada contratempo, as notas que se iniciam em tempos fracos e cuja duração percorre tempos fortes são as chamadas síncopas.
+
+As notas podem ocorrer dentro e fora destas divisões (da \emph{'métrica musical'}).
+Nos casos mais comportados, as notas ocorrem exatamente nestas divisões, com maior incidência para ataques nos tempos fortes.
+
+Com o fim de sistematizar e colocar em termos formais, podemos convencionar
+o pulso como nível $j=0$ de agrupamento, o nível $j=-1$ 
+é a primeira subdivisão do pulso, o nível $j=1$ como a primeira algomeração dos pulsos e assim por diante. 
+Assim, seja $P_i^j$ a $i$-ésima unidade de 
+pulsos no nível $j$ de agrupamento\footnote{Assim,
+$P^0_{10}$ é o décimo pulso, $P^{1}_3$ é a terceira unidade de agrupamento de pulsos,
+$P^{-1}_2$ é a segunda unidade da subdivisão do pulso.}.
+
+Especial atenção para
 os limites de $j$ pois as divisões do pulso devem resultar em durações apreciáveis
 como rítmo e também as junções do pulso deve resultar algo apreciável como, no máximo
-do nosso escopo música. Dito de outra forma: a duração de $P^{min(j)}_i$, para $\forall \; i$
+do nosso escopo, música. Dito de outra forma: a duração de $P^{min(j)}_i$, para $\forall \; i$
 é maior que $50ms$ e as durações de $P^{\text{máx}(j)}_i$ \emph{somadas}  em $\forall i$
 devem resultar em uma duração menor do que alguns minutos ou, no máximo poucas horas.
-Sempre que $i$ soma três índices temos uma relação \emph{perfecta}, quando soma dois
-ou quatro índices temos uma relação \emph{imperfecta}.
-
-Podemos convencionar $P^{-1,1,3}_{3,2,2}$ como a terceira subdivisão $P^{-1}_3$ do segundo
-pulso $P^1_2$ do terceiro aglomerado $P^3_2$ de aglomerado de pulsos ($P^2_1$ omitido por simplicidade).
-Tipicamente, para um dado $j$, $i$ tende a se manter.
+
+Para cada nível $j$ temos alguns índices $i$. Sempre que $i$ soma três 
+(ou múltiplo de três) índices temos uma relação \emph{perfecta}, 
+quando soma dois, quatro ou oito índices temos uma relação \emph{imperfecta}.
+
+Podemos utilizar a notação
+
+\begin{equation}
+P^{ \{ j_k \} }_{\left \{ \{ i_{l} \}_0^{\alpha_{k}-1} \right \} } \quad \text{onde}\; \alpha_k \; \text{é o número de partições do nível} j_k
+\end{equation}
+
+para especificar qualquer unidade da subdivisão da métrica, com níveis $j_k$ e unidades $i_l$\footnote{Como um exemplo, podemos escrever $P^{-1,0,2}_{3,2,1}$  como a terceira subdivisão $P^{-1}_3$ do segundo
+pulso $P^0_2$ do terceiro aglomerado $P^2_2$ de aglomerado de pulsos (perceba o termo $P^1_1$ omitido por simplicidade).}.
 
 A cada unidade ou conjunto de unidades $P_i^j$ pode ser associada uma sequência de amostras temporais $T_i$
-que formam uma nota musical. A associação de $P^{-1,1}_{ i_{\text{fraco}},x }$ e $P^{-1,1}_{0,x+1}$ é o que
-se chama de síncopa.
-
-Um compasso 4 por 4 possui $P^1_i$ com $i$ tipicamente contando 2, 4 ou 8 subdivisões. E
-em $P^2_i$, $i$ conta 4 unidades.
-
-Nos tempos binários, os tempos fortes são
+que formam uma nota musical.
+
+Usualmente lidamos com a distribuição rítmica dos eventos musicais com este arcabouço em mente. Assim como acontece com outras características
+musicais, lidamos com a estrutura rítmica sem fórmulas pre-estabelecidas, deixando procedimentos sistemáticos para quando é pertinente.
 

### 2012-09-18 devEresultados.tex
terminado rascunho das estruturas direcionais

@@
 \subsubsection{Estruturas direcionais}
-Arcos, sequências convergentes ou divergentes.
+
+Dadas as estruturas básicas musicais frequenciais (acordes e escalas) e rítmicas (divisões e aglomerações simples, compostas e complexas) é natural pensar em como manifestar estas estruturas de forma musical.
+
+Neste ponto, encontramos alguns artifícios consagrados pelo tempo e pelo uso.
+Apresentaremos aqui uma possibilidade musical dada por estruturas direcionais.
+Na parte seguinte exploraremos as estruturas recorrentes e cíclicas.
+
+As estruturas direcionais nos possibiitam os recursos mais usuais: arcos melódicos e sequências que convergem ou divergem.
+
+Os arcos melódicos são compostos por duas sequências convergentes: 
+uma que atinge um ápice (chamado literalmente de clímax) e 
+outra que volta do ápice à nota em que se estava antes.
+
+Seja $S_i=\{s_i\}_0^{\Lambda_S-1}$ uma sequência convergente. A sequência 
+$R_i=\{r_i\}_0^{2\Lambda_S -2}=\{s_{(\Lambda_S-1-|\Lambda_S-1-i|)}\}_0^{2\Lambda_S-2}$ é uma sequência perfeitamente simétrica.
+Pensada musicalmente, o clímax está exatamente no meio da sequência.
+Canonicamente, distinguimos também entre arcos cujos clímax estão deslocados do meio para o final e para o começo (e casos extremos disso, em que os climax estão no extato começo ou final).
 
@@
 e a terça do acorde seguinte.
+
+Melodias e harmonias.
 

### 2012-09-19 devEresultados.tex
terminada versao preliminar sobre as estruturas direcionais

--- 2012-09-18 devEresultados.tex
@@
 \subsubsection{Estruturas direcionais}
-Arcos, sequências convergentes ou divergentes.
+
+Dadas as estruturas musicais básicas tanto frequenciais (acordes e escalas) quanto 
+rítmicas (divisões e aglomerações simples, compostas e complexas) é 
+natural pensar em como realizar estas estruturas de forma musical.
+
+Neste ponto, encontramos alguns artifícios consagrados pelo tempo e pelo uso. 
+Apresentamos aqui, de forma bastante breve, o pensamento musical dado por estruturas direcionais.
+Na parte seguinte exploramos as estruturas recorrentes e cíclicas.
+
+As estruturas direcionais nos possibiitam os recursos mais usuais: arcos melódicos e sequências que convergem ou divergem.
+
+Os arcos melódicos são compostos por duas sequências convergentes: 
+uma que atinge um ápice (chamado literalmente de clímax pela teoria musical canônica) e 
+outra que, de forma paradigmática, volta do ápice à região da nota de partida.
+
+Seja $S_i=\{s_i\}_0^{\Lambda_S-1}$ uma sequência crescente. A sequência 
+$R_i=\{r_i\}_0^{2\Lambda_S -2}=\{s_{(\Lambda_S-1-|\Lambda_S-1-i|)}\}_0^{2\Lambda_S-2}$ 
+é uma sequência perfeitamente simétrica\footnote{Apresenta simetria especular perfeita, i.e.
+a segunda metade é uma versão espelhada da primeira.}.
+Pensada musicalmente, o clímax está exatamente no meio da sequência.
+Canonicamente, distinguimos também entre arcos cujos clímax estão deslocados do meio para o final e para o começo (e casos extremos disso, em que os climax estão no extato começo ou final).
 

### 2012-09-19 devEresultados.tex
feita base da discussao das estruturas ciclicas

@@
 a segunda metade é uma versão espelhada da primeira.}.
-Pensada musicalmente, o clímax está exatamente no meio da sequência.
-Canonicamente, distinguimos também entre arcos cujos clímax estão deslocados do meio para o final e para o começo (e casos extremos disso, em que os climax estão no extato começo ou final).
+Pensada segundo conceitos musicais, o clímax está exatamente no meio da sequência.
+Canonicamente, distinguimos também entre arcos cujos clímax estão deslocados do meio para o final 
+e para o começo (e casos extremos disso, em que os climax estão no extato começo ou final)~\ref{schoenbergComp}.
+Por razões ilustrativas, apresentamos no APÊNDICE XX alguns exemplos 
+destas estruturas em casos da história da música.
 
 \subsubsection{Estruturas cíclicas}
-Arcos. permutacoes
+Os arcos musicais nem sempre são fruto de estruturas direcionais como as citadas
+acima. Um caso crítico é o de estruturas que se modificam gradualmente
+até atingirem o estado original, formando um arco cíclico.
+
+Neste caso, faremos uma exposição específica sobre permutações dada
+a solidez desta abordagem segundo princípios musicais, matemáticos e
+filosóficos\footnote{De forma bastante resumida, apontamos o
+entendimento filosófico de que o pensamento humano se baseia
+em reconhecimentos de semelhanças e diferenças dentre os estímulos
+e objetos~\ref{deleuze}. Portanto, segundo este entendimento,
+as simetrias se encontram no cerne do processo cognitivo. Matematicamente,
+as simetrias são representadas por grupos algébricos e os grupos finitos, por sua vez,
+são sempre isomorfos a um grupo de permutações. Desta forma, podemos dizer que
+as permutações são capazes de representar quaisquer simetrias em um sistema finito.
+Diretamente na música encontramos permutações de forma bastante explícita, inclusive
+na análise de música étnica~\ref{permMusic}.
+Na tradição musical a permutação está presente em diversas técnicas~\ref{zamacois} e
+apontamos em especial a técnica \emph{Change Ringing} que consiste essencialmente
+em padrões de permutações dentre um fixado número de notas~\ref{change}.}.
+
+Toda permutação pode ser operada com outra permutação, resultando na permutação
+que representa a modificação de elementos fixados se aplicada uma permutação
+depois a outra:
+
+\begin{equation}
+p_1 \bullet p_2 = p_3, \quad \forall \; p_1, p_2 \; \text{permutações finitas}
+\end{equation}
+
+Uma permutação finita aplicada sucessivamente um número suficiente de vezes
+resulta na ordenação inicial dos elementos:
+
+\begin{equation}
+(p)^n = e \quad : \quad \text{se} \; p \; \text{finito} \; \Rightarrow \; \exists \;\; n  \in \mathbb{N} \; \text{e finito}
+\end{equation}
+
+O menor valor de $n$ é chamado de \emph{ordem} da permutação $p$. O elemento $e$ é chamado de elemento neutro e é
+a permutação que leva uma ordenação para ela mesma.
 

### 2012-09-21 devEresultados.tex
pequenas revisoes e encaminhamento da discussao das estruturas ciclicas

@@
 
-\subsubsection{A nota básica}
+\subsubsection{A nota básica}\label{notaBasica}
 
@@
 
-Com $\kappa \in \mathbb{N}$. Por exemplo, consideremos o modo lídio. Uma breve
-inspeção revela que $e_i^{ml}=d_{(i+2)\%7}+e_{i-1}^{ml}$ com $\kappa=2$. Para cada
+Com $\kappa \in \mathbb{N}$. Para cada
 modo, temos um único valor de $\kappa$ em qualquer intervalo $\in [x,x+6], \forall x \in \mathbb{N}$.
 
+Por exemplo, consideremos o modo lídio. Uma breve
+inspeção revela que $e_i^{ml}=d_{(i+2)\%7}+e_{i-1}^{ml}$. Temos, portanto, $\kappa=2$
+para o modo lídio. 
 Várias outras escalas e sequências são facilmente representadas desta forma. Em especial,
@@
 Nestes casos, seguindo as mesmas restrições, basta acrescentar $10$ para ter o acorde
-com sétima menor ou $11$ para resultar em sétima maior. Para realizar inversões e posições abertas,
-basta contar o número de semitons a partir da nota mais grave.
-
-Para a realização de microtonalidade, podemos usar reais não inteiros
+com sétima menor ou $11$ para resultar em sétima maior\footnote{As sétimas
+menor e maior são intervalos de $10$ e $11$ semitons respectivamente.}. Para realizar inversões e posições abertas,
+basta contar o número de semitons de cada nota a partir da nota mais grave.
+
+Para a realização de microtonalidade\footnote{O uso explícito de
+intervalos menores que o semitom é chamado microtonalismo. Isso pode ter usos
+ornamentais ou estruturantes da música em si. A
+divisão da oitava em $12$ notas igualmente espaçadas possui fundamentos físicos mas
+não deixa de ser uma \emph{convenção}
+(reconhecidamente adotada pela música erudita clássica de origem européia). Outras afinações
+são também incidentes na manifestação musical humana, para citar somente um exemplo, a
+música tradicional tailandesa utiliza uma divisão da oitava em sete notas igualmente
+espaçadas ($\varepsilon=2^{\frac{1}{7}}$),
+resultando em intervalos que pouco se assemelham aos intervalos presentes na
+divisão de doze notas.}, podemos usar reais não inteiros
 para descrever a sequência de alturas ou modificar o fator $\varepsilon$
 e usar inteiros ainda. Por exemplo, uma afinação muito
-interessante por contemplar a série harmônica consideravelmente precisa,
+interessante por contemplar a série harmônica de forma consideravelmente precisa,
 é proposta na forma da divisão da oitava em $53$ notas, de forma
@@
 resultamos em uma nova sequência $S_i'$ com $s_i'=s_i \frac{\varepsilon_1}{\varepsilon_2}$.
-A música tradicional tailandesa utiliza uma divisão da oitava em sete notas $\varepsilon=2^{\frac{1}{7}}$,
-resultando intervalos que pouco se assemelham às possibilidades da
-divisão de doze notas.
 
@@
 a permutação que leva uma ordenação para ela mesma.
+
+A utilização de permutações na música pode ser resumida da seguinte forma:
+seja $L_i$ uma lista de eventos musicais (para simplificar, considere
+estes eventos musicais como sendo notas musicais) e $p$ uma permutação.
+$L_i'=p(L_i)$ consiste nos mesmos elementos de $L_i$ com a ordem entre eles
+modificada.
+
+Importante notar que não é necessário permutar os elementos de $L_i$, mas somente
+alguma ou algumas de suas características. Assim, seja $p^f$ uma permutação agindo
+nas frequências e $L_i$ uma sequência de notas básicas como expostas
+ao final de ~\ref{notaBasica}. A nova sequência $L_i'=p^f(L_i)$ consiste nas mesmas
+notas musicais, na mesma ordem, com as frequências fundamentais permutadas segundo
+o padrão que $p^f$ apresenta.
+
+Por último, duas sutilezas deste procedimento.
+Primeiro, a permutação $p$ não precisa envolver todos os elementos de $L_i$, i.e. ela
+pode operar em subconjuntos de $L_i$. Segundo, nem todos os eventos $L_i$ precisam ser executados a
+cada consulta de estado realizada.
+
+Para simplificar, considere novamente os eventos de $L_i$ 
+como notas musicais. Assim, se $i$ é vai de $0$ a $n>3$, a cada compasso
+de $4$ notas podemos executar somente as primeiras $4$ notas. As outras notas de
+$L_i$ podem ser incidentes nos compassos em que as permutações aloque
+estas notas para as primeiras quatro notas de $L_i$.
+
+Nada impede que tenhamos permutações operando no mesmo conjunto, cada uma destas permutações
+(chamemos de $p_i$)
+pode possui: domínio das notas em que opera (frequência, duração, fades, intensidade, etc),
+modo de incidência (opera em $L_i$ a cada uma ou duas utilizações de $L_i$) e
 

### 2012-09-22 devEresultados.tex
finalizada ateh parte 3, menos usos musicais das partes 2 e 3 e coisas minimas

@@
 Nada impede que tenhamos permutações operando no mesmo conjunto, cada uma destas permutações
-(chamemos de $p_i$)
-pode possui: domínio das notas em que opera (frequência, duração, fades, intensidade, etc),
-modo de incidência (opera em $L_i$ a cada uma ou duas utilizações de $L_i$) e
+$p_i$
+possui: dimensões das notas em que opera (frequência, duração, fades, intensidade, etc),
+modo de incidência (opera em $L_i$ a cada uma ou duas utilizações de $L_i$).
+Na realização das notas de $L_i$, uma solução fácil e coerente é executar as primeiras $n$ notas\footnote{A execução de notas disjuntas de $L_i$ equivale a modificar a permutação e executar as primeiras notas.}.
+
+Esta abordagem resultou em alguns trabalhos ~\ref{figgusOriginal} e implementações computacionais.
+Descrevemos um uso coerente desta técnica abaixo e apresentamos no APÊNCIDE XX a implementação computacional disponibilizada em ~\ref{repoFiggus}
 

### 2012-09-30 devEresultados.tex
revisado ateh metade da segunda parte - variacoes na nota musical digital

@@
 
-%\subsection{Caracterização da nota musical discretizada}
-\subsection{Caracterização da nota musical digital}
 %\subsection{Caracterização da nota musical digital}
-A música é tradicionalmente pensada através de unidades chamadas notas. Em sua concepção
+\subsection{Caracterização da nota musical discretizada}
+Em diversos contextos, como na teoria musical tradicional européia,
+a música é pensada através de 
+unidades chamadas notas e 
+estas unidades pensadas como "átomos" constituintes da música.
+Atualmente, estas notas
+são tidas como discretizações
+que facilitam e enriquecem o fluxo de informação através da música 
+e como um paradigma de proposta musical (\ref{Roederer, Lacerda}).
+
+Em sua concepção
 simples, as notas possuem ao menos duração, volume, altura e timbre. Embora estas
@@
 Em primeiro lugar, estamos tratando diretamente da representação digital do som. Isso em uma
-sequência de amostras igualmente espaçadas no tempo. Desta forma, definimos a
+sequência de amostras igualmente espaçadas no tempo. Definimos a
 frequência (ou taxa) de amostragem $f_a$ como o número de amostras por segundo do áudio digital e
@@
 Desta forma, visando uma nota musical de duração $\Delta$,
-teremos uma sequência de $ \Delta . f_a $ amostras\footnote{Observe o abuso de notação:
-o limite superior de uma sequência é um número natural, mas $ \Delta . f_a $
-só satisfaz esta condição em casos muito excepcionais. Procederemos assim por simplicidade. 
-Explicitaremos em texto quando for crucial o procedimento relativo à parte fracionária dos índices
-ou de seus limites.}:
-
-\begin{equation}
-T_{i}^{\Delta}={\{t_i\}}_{i=0}^{\Delta . f_a - 1}
-\end{equation}
+teremos uma sequência de $ \lfloor \Delta . f_a \rfloor $ amostras\footnote{O
+limite superior de uma sequência é um número natural mas $ \Delta . f_a $
+só satisfaz esta condição em casos muito excepcionais. É necessário
+escolher um inteiro próximo de $\Delta . f_a$. Por simplicidade,
+escolheremos sempre a parte inteira da fração, descrita por $\lfloor \Delta . f_a \rfloor$.}:
+
+\begin{equation}
+T_{i}^{\Delta}={\{t_i\}}_{i=0}^{\lfloor \Delta . f_a \rfloor -1}
+\end{equation}
+
+Denotaremos $\Lambda = \lfloor \Delta . f_a \rfloor$ como o número de amostras da sequência, de forma que $T_i=\{t_i\}_0^{\Lambda-1}$.
 
@@
 A quantidade adimensional é denotada pela unidade decibeis ($dB$). A cada 10 $dB$s atribuímos canonicamente
-a sensação de "volume dobrado". Importantes resultados de referência são os equivalente em $dB$s de se dobrar
+a sensação de "volume dobrado". Importantes resultados de referência são os equivalentes em $dB$s de se dobrar
 a amplitude ou a potência:
@@
 
-Outro valor de referência é a amplificação necessária para que uma sequência tenha o volume dobrado ($10dB$s a mais):
+Outro valor de referência é a amplificação de amplitude
+necessária para que uma sequência tenha o volume dobrado ($10dB$s a mais):
 
@@
 Estes valores servem de guia para os aumentos e diminuições dos valores absolutos que compõem as
-sequências de amostras sonoras que visam atingir efeitos musicais. A conversão direta de decibeis
-em amplitude se dá da seguinte forma:
+sequências de amostras sonoras com propósitos musicais. A conversão direta de decibeis
+em ganho ou atenuação de amplitude se dá da seguinte forma:
 
@@
 
-Onde $A$ é o fator multiplicativo entre as médias dos valores absolutos das amplitudes dos sinais comparados.
-
-Neste ponto, correspondendo a uma partícula musical (uma nota), estabelecemos uma sequência cuja
+Onde $A$ é o fator multiplicativo que relaciona as amplitudes do sinal antes e depois da amplificação.
+
+Neste ponto, correspondendo a uma partícula musical (uma nota), estabelecemos uma sequência $T_i$ cuja
 duração e intensidade conseguimos controlar\footnote{Basta alterarmos o tamanho da sequência para alterarmos
-a duração ou multiplicarmos as os elementos da sequência por algum valor normalizador}.
-
-\subsubsection{Altura\footnote{Com ganho de simplicidade e sem perda de generalidade, considere o índice $i=a/b$ como o inteiro mais próximo da divisão $a/b$.}}
+a duração ou multiplicarmos as os elementos da sequência por algum valor para resultar em um ganho de intensidade e, portanto, de volume.}.
+
+\subsubsection{Altura}
 
@@
 com ciclo de duração $\delta_{f_0} = 1/f_0$. Esta duração multiplicada pela taxa de amostragem $f_a$ resulta no número de amostras
-do ciclo $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$. Evidente que, sendo $T_i^f$ a sequência sonora de frequência fundamental $f$:
+do ciclo $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$. É evidente que, sendo $T_i^f$ a sequência sonora de frequência fundamental $f$:
 
@@
 
+Considerando $\frac{f_a}{f}$ inteiro. Na sessão seguinte estaremos sintetizando sons cuja frequência $f$ não divide $f_a$, por hora podemos aceitar esta restrição sem perda de generalidade.
+
 \subsubsection{Timbre}
 Enquanto o período da onda corresponde a uma frequência fundamental, o percurso
-da onda sonora dentro do período define um espectro harmônico e portanto
+da onda sonora dentro do período - chamado de forma de onda - define um espectro harmônico e portanto
 um timbre\footnote{O timbre é uma característica subjetiva e complexa. Fisicamente,
@@
 espectrais e simplicidade. Enquanto a senóide é um ponto isolado no espectro, estas formas
-de onda apresentam cadeias de componentes harmônicos e usos específicos.
+de onda apresentam cadeias de componentes harmônicas e usos específicos.
 As formas de onda especificadas em ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} são dispostas graficamente na figura ~\ref{fig:formasDeOnda}. 
-São as formas de onda artificiais tradicionalmente usadas na música para síntese e controle oscilatório de variáveis, apresentam diversos usos também fora da música.
+São as formas de onda artificiais tradicionalmente usadas na música para síntese e controle oscilatório de variáveis, apresentam diversos usos também fora da música(\ref{Openheim}).
 
@@
 é útil nos médios e agudos para que o som ganhe naturalidade e fique mais agradável).
-Por este mesmo motivo, os harmônicos atenuados da onda triangular
-a faz a mais funcional - dentre as citadas - para ser usada sem nenhum tratamento.
+Os harmônicos atenuados da onda triangular
+a faz a mais funcional - dentre as citadas - para ser usada sem nenhum tratamento para síntese de notas musicais.
 
@@
 Estas formas de onda possuem usos especiais na música e seus espectros estão dispostos na figura ~\ref{fig:espectroDeOndas}.
-É importante notar as componentes isoladas e exatamente harmônicas dos espectros.
+É importante notar as componentes isoladas e exatamente harmônicas dos espectros,
+resultado de um período mantido fixo com exatidão.
 A senóide consiste de um nódulo único no espectro, frequência pura.
@@
 Como nossa percepção segue a escala logarítmica, o espectro possui notas diferentes da frequência fundamental.
-Além disso, o número de harmônicos será limitado pela frequência máxima $f_a/2$.
+Além disso, o número de harmônicos será limitado pela frequência máxima $f_a/2$ (pelo Teorema de Nyquist).
 
 Musicalmente crucial aqui é internalizar que a presença de
-energia\footnote{A energia total equivale à soma dos quadrados das amplitudes como as de pressao/voltagem pelo tempo, e.g. figura ~\ref{fig:formasDeOnda}.
+energia\footnote{A energia total equivale à soma dos quadrados das amplitudes
+(como as de pressao/voltagem no tempo), e.g. figura ~\ref{fig:formasDeOnda}.
 As componentes espectrais - e.g. ~\ref{fig:espectroDeOndas} e ~\ref{fig:espectroOboe} -
-também podem ser elevadas ao quadrado para resultar em quantificação de energia e somam-se na energia total. As energias se equivalem pelo teorema de Parseval.}
+também podem ser elevadas ao quadrado para resultar em quantificação de energia e somam-se na energia total.
+As energias se equivalem pelo teorema de Parseval: $\frac{1}{\Lambda} .
+\sum_{k=0}^{\Lambda -1}c_k^2 = \sum_{i=0}^{\Lambda-1}t_i^2$.}
 em uma componente de frequência $f_n$ na decomposição por fourier 
 implica na presença de uma oscilação senoidal na constituição do som, puramente harmônica no som e naquela frequência $f_n$. Esta energia concentrada especificamente na frequência $f_n$ é separada
- pelo ouvido para adentrar em um nível cognitivo de processamento\footnote{Esta separação em frequência é realizada por diversas espécies através de mecanismos similares à cóclea humana.}.
+ pelo ouvido para adentrar em um nível cognitivo de processamento\footnote{Esta separação em frequência é realizada por diversas espécies através de mecanismos similares à cóclea humana(\ref{musicaEsom}).}.
   As componentes senoidais são geralmente as principais responsáveis pela qualidade chamada timbre e, caso não apresentem proporções harmônicas (relações de pequenos números), são ruídos, i.e.
@@
 O período escolhido para a amostragem é relativamente curto, com 98 amostras corresponde a $\frac{44100}{98}=450 Hz$.
-Pode-se perceber o espectro rico em frequências do oboé e o espectro mais grave do som real.
-A figura ~\ref{fig:espectroOboe}
-expõe em conjunto o conteúdo espectral da nota original e o resultante da utilização destas amostras.
+Pode-se perceber, através das curvaturas, o espectro rico em frequências do oboé e o espectro mais grave do som real.
 
@@
 O som resultante possui o espectro momentâneo do som original. Por ser repetido de forma idêntica,
-seu espectro é perfeitamente harmônico, sem os ruídos e variações típicas do fenômeno natural.
-Isso pode ser visto claramente na figura ~\ref{fig:espectroOboe} onde os espectros da nota original do oboé e de uma nota
+seu espectro é perfeitamente harmônico, sem os ruídos e variações típicas do fenômeno natural. Isso pode ser 
+visto claramente na figura \ref{fig:espectroOboe} onde os espectros da nota original do oboé e de uma nota 
 artificial - de mesma duração e cujas amostras consistem no mesmo período da figura ~\ref{fig:formasDeOnda} - estão dispostas juntamente.
@@
 \subsubsection{O espectro no som amostrado}
-Além deste papel-chave no timbre e na cognição como um todo, a presença e comportamento destas componentes é peculiar no som discretizado.
+Além deste papel-chave no timbre e na cognição como um todo, a presença e comportamento destas componentes senoidais é peculiar no som discretizado.
 Considere um sinal $T_i$ e sua decomposição de Fourier $\mathcal{F}\{T_i\}=C_i=\{c_i\}_0^{\Lambda-1}$.
@@
 Onde $c_k = a_k + j .
-b_k$ pondera a amplitude e fase de cada frequência $w_k=\frac{2\pi k}{\Lambda}$.
-Nosso som possui amostras $t_i$ reais que resultam das contribuições de cada componente de frequência $w_k$ e cujo $c_k$ regula o módulo e a fase.
+b_k$ pondera a amplitude e fase de cada frequência: $w_k=\frac{2\pi k}{\Lambda}$ em radianos ou $f_k=w_k\frac{f_a}{2\pi}=\frac{k.f_a}{\Lambda}$ em Herz.
+Nosso som possui amostras $t_i$ reais que resultam das contribuições de cada componente de frequência $w_k=\frac{2\pi k}{\Lambda}$ e cujo $c_k$ regula o módulo e a fase.
 A parte real da equação ~\ref{recomposicaoFourier} nos fornece a componente de forma mais clara:
@@
 Podemos notar claramente pela equação ~\ref{moduloEfase} que o termo imaginário apenas acrescenta uma fase à senóide real.
-Ao mesmo tempo $b_k$ também altera a amplitude mas a informação de amplitude já está em $a_k$, i.e.
-os termos imaginários da decomposição espectral por Fourier proporcionam a varredura de fase
- $[-\frac{\pi}{2},+\frac{\pi}{2}]$.
+Ao mesmo tempo $b_k$ também altera a amplitude mas a informação de amplitude já está em $a_k$,
+i.e. os termos imaginários $b_k$ da decomposição espectral por Fourier proporcionam a varredura de fase
+ $[-\frac{\pi}{2},+\frac{\pi}{2}]$, basta observar o termo $tg^{-1}\left (\frac{b_k}{a_k}\right )$ que possui este contra-domínio.
 O sinal de $a_k$ especifica se estamos do lado direito ou esquerdo do circulo trigonométrico, completando a varredura completa de fase (os intervalos $[-\frac{\pi}{2},+\frac{\pi}{2}]$ e $[\frac{\pi}{2},\frac{3\pi}{2}]$ completam $2\pi$).
@@
     \centering
-    \caption{Oscilação de 2 amostras (frequência máxima em qualquer $t_a$)}
+    \caption{Oscilação de 2 amostras (frequência máxima em qualquer $f_a$)}
         \includegraphics[width=\textwidth]{figuras/amostras2c_}
@@
 A figura ~\ref{fig:amostras2} exibe duas amostras e a componente espectral que contém.
-A decomposição de Fourier resulta neste caso em somente dois coeficientes $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$.
-O papel das amplitudes $a_k$ fica bem claro:
- $a_0$ é o deslocamento fixo\footnote{chamado de bias ou offset} e $a_1$ especifica a amplitude da oscilação em si.
-A frequência que possui a energia\footnote{Lembrando que as energias garantidamente se equivalem pela $\frac{1}{\Lambda} .
-\sum_{k=0}^{\Lambda -1}c_k^2 = \sum_{i=0}^{\Lambda-1}t_i^2$ (Teorema de Parseval)}
- $e_k=\frac{(c_k)^2}{\Lambda=2}$ é dada pela relação $f_k=k .
-\frac{f_a}{\Lambda=2} $.
-Ou seja, no caso de duas amostras só temos energia nas frequências $f_0=0$ e $f_1=\frac{f_a}{\Lambda=2}=f_{\text{máx}}$.
+A decomposição de Fourier resulta neste caso em somente dois coeficientes $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relativos à frequências $\{f_k\}_0^1=\{w_k\frac{f_a}{2\pi}\}_0^1=\{k\frac{f_a}{\Lambda=2}\}_0^1=\{0,\frac{f_a}{2}=f_{\text{máx}}\}$
+com energias $e_k=\frac{(c_k)^2}{\Lambda=2}$. O papel das amplitudes $a_k$ fica bem claro:
+ $\frac{a_0}{2}$ é o deslocamento fixo\footnote{Chamado em vários contextos de bias ou offset.} e $\frac{a_1}{2}$ especifica a amplitude da oscilação em si.
+  é dada pela relação $f_k=k . \frac{f_a}{\Lambda=2}$.
 
@@
 Vejamos o que acontece no caso de 3 amostras. Todas as sequências fixas $T_i$ de apenas 3 amostras também apresentam
-somente 1 frequência, pois sua primeira harmônica usaria 1,5 amostras e ultrapassa o limite inferior de 2 amostras mínimas (a frequência da harmônica ultrapassaria a de Nyquist pois:  $\; \frac{f_a}{3}.2 > \frac{f_a}{2} $).
+somente 1 frequência, pois sua primeira harmônica usaria 1,5 amostras e ultrapassa o limite inferior de 2 amostras mínimas (a frequência da harmônica ultrapassaria a de Nyquist pois:  $\; \frac{2.
+f_a}{3} > \frac{f_a}{2} $).
 Se $\Lambda=3$,
 os coeficiêntes $\{c_k\}_0^{\Lambda-1=2}$ da decomposição por Fourier apresentam-se em 
-3 componentes frequenciais. Uma delas é a frequência zero, as outras duas contribuem de forma igual na reconstrução da senóide com $f=f_a/3$.
+3 componentes frequenciais. Uma delas é relativa à frequência zero ($c_0$), as outras duas ($c_1$ e $c_2$) contribuem de forma igual na reconstrução da senóide com $f=f_a/3$.
 
@@
 
-Partimos de $\Lambda$ amostras reais $t_i$ e temos $\Lambda$ coeficientes complexos $c_k=a_k+j.b_k$.
-Os coeficientes $c_k$ se equivalem dois a dois\footnote{Módulos iguais e fases com sinais opostos; $a_{k1}=a_{k2}$ e $b_{k1}=-b_{k2}$.} pois correspondem às frequências $f_k = k.\frac{f_a}{\Lambda}, \; k \in \{0...\Lambda-1\} $
- e quando $ f_k > f_{\text{máx}} = \frac{f_a}{2} $ (o que já exclui $f_0=0$ e $f_k=\frac{f_a}{2}$), a frequência $f_k$ é espelhada para $f_l=\frac{f_a}{2} - (f_k-\frac{f_a}{2})=f_a-f_k=f_a - k\frac{f_a}{\Lambda}=(\Lambda-k)\frac{f_a}{\Lambda} \Rightarrow f_k\equiv f_{\Lambda-k}$.
+Sempre que partimos de $\Lambda$ amostras reais $t_i$ resultamos em $\Lambda$ coeficientes complexos $c_k=a_k+j.b_k$.
+Os coeficientes $c_k$ se equivalem dois a dois\footnote{Parte real igual e imaginária com sinal trocado: $a_{k1}=a_{k2}$ e $b_{k1}=-b_{k2}$.
+Como consequência temos módulos iguais e fases com sinais opostos} correspondendo às frequências $f_k = k\frac{f_a}{\Lambda}, \; k \in \{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \} $.
+Quando $k> \frac{\Lambda}{2} $
+a frequência $f_k$ é espelhada em $\frac{f_a}{2}$ da seguinte forma $f_k=\frac{f_a}{2} - (f_k-\frac{f_a}{2})=f_a-f_k=f_a - k\frac{f_a}{\Lambda}=(\Lambda-k)\frac{f_a}{\Lambda} \;\;\;\; \Rightarrow \;\;\;\; f_k\equiv f_{\Lambda-k} \; , \forall \;\; k>\frac{\Lambda}{2}$.
 
@@
    são equivalentes em pares.
-   Excepcional excessão para $f_0$ e, no caso de $\Lambda$ ser par, de $f_{\Lambda/2}=f_{\text{máx}}=\frac{f_a}{2}$.
+   Excepcional excessão para $f_0$ e, no caso de $\Lambda$ ser par, de $f_{\Lambda/2}=f_{\text{máx}}=\frac{f_a}{2}$ , ambas as componentes são isoladas, i.e.
+não existe outra componente na frequência além dela mesma.
 Além disso, estas duas frequências (a frequência zero e a frequência máxima) não são representadas com variação de fase e, portanto, são estritamente reais.
@@
 \begin{equation}
-t_i = \frac{a_0}{\Lambda} + \frac{2}{\Lambda}\sum_{k=1}^{\tau}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]+ \frac{ a_{\Lambda/2+1}}{\Lambda}.(1-\Lambda\% 2)
+t_i = \frac{a_0}{\Lambda} + \frac{2}{\Lambda}\sum_{k=1}^{\tau}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]+ \frac{ a_{\Lambda/2}}{\Lambda}.(1-\Lambda\% 2)
 \end{equation}
@@
 
-Assim, a exemplo da figura ~\ref{fig:amostras3}, a transformada de Fourier de 3 amostras possui 2 coeficientes imaginários que possuem quantidades iguais de energia na mesma frequência.
-
-Com 4 amostras, podemos representar 1 ou 2 frequências em proporções diferentes.
-A figura ~\ref{fig:amostras4} mostra uma forma de onda de 4 amostras e suas duas componentes.
-Note que as contribuições individuais se somam de fato na forma de onda original, e que as curvaturas maiores são fruto da frequência mais aguda.
+Assim, a exemplo da figura ~\ref{fig:amostras3}, a transformada de Fourier de 3 amostras possui 2 coeficientes frequenciais que possuem quantidades iguais de energia na mesma frequência.
+
+Com 4 amostras, podemos representar 1 ou 2 frequências em proporções diferentes. A figura ~\ref{fig:amostras4} mostra uma 
+forma de onda de 4 amostras e suas duas componentes. 
+Note que as contribuições individuais se somam de fato na forma de onda 
+original, e que as curvaturas maiores são fruto da frequência mais aguda
+enquanto um deslocamento fixo da somatória das componentes é fruto
+da componente na frequência zero.
 
@@
 Escolhamos $f$ tal que $f$ divida $f_a$  
- \footnote{Esta escolha facilita as demonstrações pois os períodos são quantidades inteiras de amostras.
-Transporemos esta limitação na prática com o uso de uma tabela de amostragem ou com o uso direto de funções periódicas,e.g.
+ \footnote{Esta escolha facilita as demonstrações pois os períodos resultam em quantidades inteiras de amostras.
+Transporemos esta limitação na prática com o uso de uma tabela de amostragem ou com o uso direto de funções periódicas, e.g.
 $t_i=sen(x.i)$.
@@
 \begin{equation}
-T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{f_a . \Delta}=\{t_{i \; \% \left( \frac{f_a}{f} \right) } \}_0^{f_a . \Delta}
+T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{f_a . \Delta-1}=\{t_{i \; \% \left( \frac{f_a}{f} \right) } \}_0^{f_a . \Delta-1}
 \end{equation}
@@
 \begin{equation}
-T_i^{f,\; \Delta}=\{t_i^f\}_0^{f_a . \Delta}=\{l^f_{i\,\%\left(\frac{f_a}{f}\right)}\}_0^{f_a . \Delta}
+T_i^{f,\; \Delta}=\{t_i^f\}_0^{f_a . \Delta-1}=\{l^f_{i\,\%\left(\frac{f_a}{f}\right)}\}_0^{f_a . \Delta-1}
 \end{equation}
@@
 As sequências podem também ser concatenadas no tempo.
-Caso as sequências $\{t_{k,i}\}_0^{\Lambda_k-1}$ de tamanhos $\Lambda_k$  representem $k$ notas musicais, sua concatenação em $t_i$ resultará em sequências musicais simples ou melodias:
-
-\begin{equation}
-\{t_i\}_0^{\sum\Delta}=\{t_{k,i}\}_0^{\sum\Delta}, \;\; k\text{ menor inteiro } : \quad \Lambda_k > i -\sum_{j=0}^{k-1}\Lambda_j
+Caso as sequências $\{t_{k,i}\}_0^{\Lambda_k-1}$ de tamanhos $\Lambda_k$  representem $k$ notas musicais, sua concatenação em uma única sequência $T_i$ resultará em uma sequência musical simples ou melodia:
+
+\begin{equation}
+\{t_i\}_0^{\sum\Delta_k-1}=\{t_{l,i}\}_0^{\sum\Delta_k-1}, \;\; l\text{ menor inteiro } : \quad \Lambda_l > i -\sum_{j=0}^{l-1}\Lambda_j
 \end{equation}
@@
 O princípio vertical está demonstrado nos \emph{quadros sonoros}, sons estáticos com espectros peculiares.
-Ambas as peças estão em código Python no APÊNDICE XX e podem ser escutadas e vizualizadas nos links [][][].
+Ambas as peças estão em código Python no APÊNDICE XX e podem ser escutadas e vizualizadas nos links \ref{reducedLink} e \ref{quadrosLink}.
 
@@
 LUT), é um procedimento simples que consiste em criar uma tabela
-de referência e coletar as amostras destas tabelas segundo a 
+de referência e coletar as amostras desta tabela segundo a 
 necessidade\footnote{Em contraposição ao cálculo das amostras individuais
-para cada caso.}. Em nosso caso, usaremos tabelas de busca para os
+de forma analítica.}. Em nosso caso, usaremos tabelas de busca para os
 sons criados, as notas. Assim, estas tabelas são unidimensionais e com
@@
 Ou seja, dada a tabela de busca, basta termos a sequência de índices corretos
-para sintetizar o som. Para o cálculo do inteiro $\gamma_i$, convencionamos
+para sintetizar o som em qualquer frequência. Para o cálculo do inteiro $\gamma_i$, convencionamos
 atribuir a ele o maior inteiro menor que $i.f\frac{\widetilde{\Lambda}}{f_a}$.
@@
 $f_a$\footnote{Como citado acima, $\widetilde{\Lambda} = 1024$ é suficiente para sintetizar
-sequências $T_i^f$ com qualquer $f$ no espectro.}. O procedimento de busca em tabela
+sequências $T_i^f$ com qualquer $f$ no espectro audível.}. O procedimento de busca em tabela
 é computacionalmente bastante barato, substituindo cálculos por buscas simples (por isso geralmente
 é entendido como um processo de otimização). Salvo quando assinalado,
-a seguir usaremos este procedimento para todos os casos cabíveis pois
+no texto que segue usaremos este procedimento para todos os casos cabíveis pois
 simplifica diversas rotinas e é computacionalmente coerente.
@@
 com base no período $\{\widetilde{l}_i\}_0^{\Lambda-1}$. Note que na equação
-~\ref{freqLinear}, o termo $f_{\Lambda -1} - f_0$ é o âmbito do contra-domínio de $f_i$
+\ref{freqLinear}, o termo $f_{\Lambda -1} - f_0$ é o âmbito do contra-domínio de $f_i$
 e que $\Lambda -1$ é âmbito do domínio. Como $i$ varre o domínio $[0,\Lambda-1]$ e
@@
 
-As equações ~\ref{freqLinear}, ~\ref{indiceLinear} e ~\ref{serieAmostralLin} são relativas à progressão linear
+As equações \ref{freqLinear}, \ref{indiceLinear} e \ref{serieAmostralLin} são relativas à progressão linear
 da frequência. Como assinalado, nossa percepção segue a progressão logarítmica:
@@
 
-Onde usamos a equação ~\ref{indiceExponencial} para calcular a sequência $\{\gamma_i\}_0^{\Lambda-1}$ de índices para para
+Onde usamos a equação \ref{indiceExponencial} para calcular a sequência $\{\gamma_i\}_0^{\Lambda-1}$ de índices para para
 serem buscados na nossa tabela que contém um único período da forma de onda.
@@
  $t_i^{\;\overline{f_0,f_{\Lambda-1}}}=\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}$.
-Note que a equação ~\ref{serieAmostralLog} se mantém (permanece idêntica à equação ~\ref{serieAmostralLin}).
+Note que a equação \ref{serieAmostralLog} se mantém (permanece idêntica à equação \ref{serieAmostralLin}).
 
@@
 
-\begin{equation}
+\begin{equation}\label{transAmp}
 T_i^{'}=T_i \odot A_i = \{t_i . a_i\}_0^{\Lambda-1}=\left \{ t_i . (a_{\Lambda-1} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
@@
 
-Para resultar em uma amplificação linear, basta usarmos umas série $\{a_i\}$ diferente:
+Para realizar um \emph{fade out} (diminuição progressiva do volume até o volume zero),
+podemos tomar $a_0=1$ e $a_{\Lambda-1}=0$. Para um \emph{fade in} (acréscimo progressivo de volume a partir do volume zero), podemos
+tomar $a_0=1$, $a_{\Lambda-1}=2$ e subtrair $1$ de toda sequência: $\{a_i'\}_0^{\Lambda-1}=\{a_i-1\}_0^{\Lambda-1}$.
+
+Para resultar em uma amplificação linear, basta usarmos uma sequência $\{a_i\}$ adequada:
 
@@
 
-Com $a_0=1$ para preservar a amplitude inicial original.
-
-Aqui pode muito nos auxiliar a conversão de decibeis para amplitude apresentada na equação ~\ref{ampDec},
-resultando em:
-
-\begin{equation}
-T_i^{'}=\left \{ t_i . (10^{\frac{V_{dB}}{20}} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
+Aqui pode muito nos auxiliar a conversão de decibeis para amplitude apresentada na equação ~\ref{ampDec}.
+Junto à equação \ref{transAmp}, resulta na sequinte equação para uma transição de $V_{dB}$:
+
+\begin{equation}
+T_i^{'}=\left \{ t_i (10^{\frac{V_{dB}}{20}} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
 \end{equation}
@@
 As variáveis $\mu$ e $\nu$ quantificam a intensidade do vibrato: $\mu$ é uma medida direta da quantidade
-de Herz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação.
+de Herz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação ($2\nu$ é o número de semitons entre os picos superioes e inferiores da oscilação de frequência do vibrato).
 Note que $\nu=\log_{2}\frac{f+\mu}{f} $ é conveniente neste caso pois o aumento máximo de frequência

### 2012-09-30 scripts/amostras4.py

@@ -6,7 +6,7 @@ f=n.fft.fft
 
 #n4=n.random.rand(4)*2-1
 n4=n.array([ 0.58003705, -0.30828309, -0.29797696, -0.99219078])
-p.plot(n4,"bo")
+#p.plot(n4,"bo")
 
 ff=f(n4)
 

### 2012-10-01 devEresultados.tex
adicionada script e figura da convolucao

@@
 
-Onde $V_{dB}$ é profundidade da oscilação em decibeis do tremolo e $a_{\text{máx}}$ é o ganho máximo de amplitude envolvido.
-A medição em decibeis é bastante conveniente neste caso pois o aumento máximo de amplitude
+Onde $V_{dB}$ é profundidade da oscilação em decibeis do tremolo e $a_{\text{máx}}=10^{\frac{V_{dB}}{20}}$
+ é o ganho máximo de amplitude envolvido.
+A medição em decibeis é bastante pertinente pois o aumento máximo de amplitude
 não equivale à diminuição máxima relacionada, enquanto a diferença em decibeis se mantém.
 
-Com as equações ~\ref{vbrT}, ~\ref{vbrGamma}, ~\ref{vbrF}, ~\ref{vbrAux} e  ~\ref{vbrGamma}
-são utilizadas para síntese de frequência modulada (FM) nos casos em que $f'$ é
-maior que $20Hz$ onde as oscilações de frequência resultam em padrões complexos
-de componentes espectrais harmônicas e não harmônicas. As equações ~\ref{trA}
-e ~\ref{trT} (junto às ~\ref{vbrT} e ~\ref{vbrGamma} para a realização da sequência
-auxiliar $t_i'$) resultam na síntese de amplitude modulada (AM)\footnote{As sínteses
+As equações ~\ref{vbrT}, ~\ref{vbrGamma}, ~\ref{vbrF} e ~\ref{vbrAux}
+são utilizadas para síntese nos casos em que $f'$ é
+maior que $20Hz$ e as oscilações de frequência resultam em padrões complexos
+de componentes espectrais harmônicas e não harmônicas. É a chamada síntese
+por frequência modulada (FM).
+
+As equações ~\ref{trA} e ~\ref{trT}
+(junto às ~\ref{vbrT} e ~\ref{vbrGamma} para a realização da sequência auxiliar $t_i'$) 
+resultam na síntese chamada de amplitude modulada (AM)
+quando $f'$ é maior que $20Hz$\footnote{As sínteses
 FM e AM são técnicas com diversas amplicações na música e também em outras áreas,
@@
 via ondas eletromagnéticas. O leitor interessado pode visitar os itens relacionados
-da nossa bibliografia para uma abordagem musical (veja especialmente ~\ref{} e ~\ref{})}.
+da nossa bibliografia para uma abordagem musical (veja especialmente ~\ref{} e ~\ref{})
+ou aproveitar nossa abordagem breve sobre o assunto nos usos musicais desta sessão, logo abaixo.}.
 
 \subsubsection{Aplicação de Filtros Digitais}
-Explicitaremos a aplicação de filtros por convolução
-e por equação a diferenças. As possibilidades
+ As possibilidades
 aqui são variadas e a complexidade facilmente
@@
 Recomendamos ao leitor
-interessado nossa bibliografia, em especial ~\ref{} e ~\ref{}.},
+interessado uma visita à nossa bibliografia, em especial ~\ref{} e ~\ref{}.},
 de forma que focamos em explicitar o que ocorre com as sequências
-de amostras nestes dois casos.
-
+de amostras e em descrever os procedimentos usuais.
 A aplicação de filtros pode
@@
 como parte dos processos tipicamente chamados de tratamento sonoro.
+
+Explicitaremos a aplicação de filtros por convolução
+e por equação a diferenças.
 
@@
 percebido pela visualização da convolução de sequências arbitrárias e dispomos
-para este fim a figura \ref{}.
+para este fim a figura \ref{fig:conv}.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Convolução (interpretação gráfica)}
+        \includegraphics[width=\textwidth]{figuras/convolucao_}
+        \label{fig:conv}
+\end{figure}
 

### 2012-10-02 devEresultados.tex
adicionadas figuras e melhoradas discussoes ateh filtros iir

@@
 \begin{equation}\label{indiceLinear}
-\gamma_i=\left \lfloor i .
-f_i\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor i  \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right ] \right \rfloor   \quad ,\quad \quad i \;\in\; \mathbb{N}, \quad i \;\in\; [0,\Lambda-1]
+\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i=\left \lfloor \sum_{j=0}^{i} f_i\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor \sum_{j=0}^{i}  \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right ] \right \rfloor
 \end{equation}
@@
 \begin{equation}\label{indiceExponencial}
-\gamma_i=\left \lfloor i .
-f_i\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor i f_0 \frac{\widetilde{\Lambda}}{f_a} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}} \right \rfloor   \;\; ,\quad \quad i \;\in\; \mathbb{N}, \quad i \;\in\; [0,\Lambda-1]
+\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i=\left \lfloor \sum_{j=0}^{i} f_i\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor \sum_{j=0}^{i} f_0 \frac{\widetilde{\Lambda}}{f_a} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}} \right \rfloor
 \end{equation}
@@
 \begin{equation}\label{vbrGamma}
-\gamma_i = \left \lfloor i .
-f_i \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor i \frac{\widetilde{\Lambda}}{f_a}f \left ( \frac{f + \mu }{f} \right )^{t_i'}  \right \rfloor= \left \lfloor i \frac{\widetilde{\Lambda}}{f_a}f .
+\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_i \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f \left ( \frac{f + \mu }{f} \right )^{t_i'}  \right \rfloor= \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f .
 2^{t_i'\frac{\nu}{12}}  \right \rfloor
@@
 \end{equation}
+
+\begin{figure}[h!]
+    \centering
+    \caption{Espectrograma de um 'vibrato' senoidal de $3Hz$ e profundidade de uma oitava em uma dente de serra de $1000Hz$}
+        \includegraphics[width=\textwidth]{figuras/vibrato___}
+        \label{fig:vibrato}
+\end{figure}
 
@@
 
+A Figura \ref{fig:vibrato} é o espectrograma de um vibrato artificial de uma nota em
+$1000Hz$ (entre um si e um dó) e cujo desvio da frequência atinge uma oitava
+para cima e para baixo. Note o padrão senoidal do vibrato e as compontentes harmônicas
+da dente de serra. Estas qualidades não são praticáveis em instrumentos
+usuais, introduzindo novidade nas possibilidades musicais. Qualquer forma de onda pode
+ser utilizada para gerar o som e o padrão de oscilação do vibrato. Também é crucial a
+frequência de oscilação e o desvido de altura envolvido\footnote{O desvio de altura
+é chamado profundidade e é dado por conveniência em semitons ou cents.}.
+
 O caso do tremolo é semelhante e $f'$, $\gamma_i'$ e $t_i'$ permanecem os mesmos. Calculamos
@@
 não equivale à diminuição máxima relacionada, enquanto a diferença em decibeis se mantém.
+
+A figura~\ref{fig:tremolo} mostra a amplitude das sequências $\{a_i\}_0^{\Lambda-1}$ e $\{t_i'\}_0^{\Lambda-1}$
+para três oscilações de um trêmolo com forma da dente de serra. A curvatura é devido à progressão logarítmica de
+intensidade.
+A frequência do trêmolo é de $1,5Hz$ pois $f_a=44,1kHz \; \Rightarrow \; \text{duração} = \frac{i_{\text{máx}}=82000}{f_a}= 2s \; \Rightarrow \; \frac{3\text{oscilações}}{2s}=1,5$ oscilações por segundo ($Hz$).
+
+\begin{figure}[h!]
+    \centering
+    \caption{Tremolo de profundidade $12dB$ com padrão oscilatório de uma dente de serra $1,5Hz$ em uma senóide de $40Hz$}
+        \includegraphics[width=\textwidth]{figuras/tremolo}
+        \label{fig:tremolo}
+\end{figure}
 
@@
 resposta ao impulso. Para nossos fins, esta
-convolução fica definida como:
+convolução fica definida como\footnote{Convencionamos $t_i=0$ para as
+amostras não definidas de antemão.}:
 
@@
 percebido pela visualização da convolução de sequências arbitrárias e dispomos
-para este fim a figura \ref{fig:conv}.
+para este fim a figura \ref{fig:conv}. Observe que a resposta ao impulso $\{h_i\}$
+é disposta na forma retrograda e que mostramos na figura os cálculos explícitos
+das amostras $t_{12}'$ e $t_{32}'$ do sinal convoluído. O sinal resultante possui
+sempre o tamanho $\Lambda_t+\Lambda_h -1=\Lambda_{t'}$.
 
@@
     \centering
-    \caption{Convolução (interpretação gráfica)}
-        \includegraphics[width=\textwidth]{figuras/convolucao_}
+    \caption{Interpretação gráfica da convolução}
+        \includegraphics[width=\textwidth]{figuras/convolucao______}
         \label{fig:conv}
@@
 que, convoluídas com a sequência sonora, resultam nela própria
-com uma reverberação que se assemelha ao ambiente de medição.
+com uma reverberação que se assemelha à reverberação do ambiente 
+em que ocorreu a medição.
 
@@
 se for grande o número de amostras utilizado para especificar o perfil espectral
-do filtro.}. Outro exemplo simples é a aplicação de delays através
+do filtro. Este procedimento deve ser testado caso a caso pois estes procedimentos
+espectrais não possuem comportamento previsível.}.
+
+Outro exemplo simples é a aplicação de delays através
 da convolução do som com uma resposta ao impulso que possui um impulso
-em cada reincidência do som na linha de delays intensionada. Estes dois exemplos
-estão mostrados na figura ~\ref{}. Note que, pelo método citado,
-filtros dos tipos citados abaixo em ~\ref{passaBaixas},~\ref{passaAltas} podem ser sintetizados.
+em cada reincidência do som na linha de delays intensionada.
+Para ilustrar essa aplicação, dispomos a figura~\ref{fig:delays}
+em que pode-se observar o deslocamento causado pela convolução
+com o impulso. Dependendo da densidade dos impulsos, o resultado
+é de caráter rítmico (20 impulsos por segundo ou menos) ou de amálgama
+sonoro (20-40 impulsos por segundo ou mais). Neste último caso,
+ocorrem processos tipicamente vinculados à síntese granular e às
+equalizações.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Convolução com o impulso e linhas de delays}
+        \includegraphics[width=\textwidth]{figuras/delays_}
+        \label{fig:delays}
+\end{figure}
 
@@
 Oscilacoes, tremolos e vibratos. Variacoes
+~15Hz tremolo rugoso
+varias possibilidades de combinacao de intensidades, formas de onda e frequencias
+
 progressivas das oscilacoes dos tremolos e vibratos

### 2012-10-02 scripts/delays.py

@@ -1,0 +1,94 @@
+#-*- coding: utf8 -*-
+import numpy as n, pylab as p
+
+p.subplot(311)
+fa=4410
+
+# Som curto
+sdur=0.005 # segundos
+f=200. #hz
+ii=fa*sdur
+s=n.sin(n.linspace(0,2*200*2*n.pi*sdur,ii, endpoint=False))
+
+# resposta ao impulso com um pulso
+ddur=0.02
+iii=fa*ddur
+xi=2*iii/3
+hi=n.zeros(iii)
+hi[xi]=1
+som = n.convolve(s,hi)
+
+p.plot(s+6,'bo', label=r"$\{t_i\}_0^{\Lambda-1}$")
+p.plot(hi+3,'ro', label=r"$\{h_i=\delta_{\xi-i}\}_0^{\Lambda_h-1}$")
+p.plot(som,'go', label=r"$\{(t*h)_i\}_0^{\Lambda+\Lambda_h-2}$")
+p.legend(loc="upper right")
+p.xticks((0,ii-1,xi-1,xi+ii-1,ii+iii-2),(r"0",r"$\Lambda-1$",r"$\xi$",r"$\xi + \Lambda-1$", r"$\Lambda+\Lambda_h-2$"),fontsize='16')
+p.yticks((),())
+p.ylim(-1.3,7.3)
+p.xlim(-5,ii+iii-2+5)
+p.plot([ii-1,ii-1],[-2,9],"y", linewidth=5,alpha=.4)
+p.plot([xi -1,xi -1],[-2,9],"y", linewidth=5,alpha=.4)
+p.plot([xi +ii-1,xi +ii-1],[-2,9],"y", linewidth=5,alpha=.4)
+
+p.ylabel(r"deslocamento", fontsize=16, fontweight='bold')
+
+
+
+p.subplot(312)
+
+
+ddur=0.2
+iii=fa*ddur
+hi=n.zeros(iii)
+pulsos=iii*2/100 # percentagem de incidencias no delay
+xis=n.random.randint(0,iii-iii*3/100,pulsos)
+xis=n.array([330, 173, 629, 387,  40, 617, 373, 522, 416, 326, 290, 264,  16, 831, 442, 123, 677])
+hi[xis]=n.ones(pulsos)
+som = n.convolve(hi,s)
+
[…]

### 2012-10-02 scripts/tremolo.py

@@ -1,0 +1,45 @@
+#-*- coding: utf8 -*-
+import numpy as n, pylab as p, scikits.audiolab as a
+
+fa=44100
+Dv=2048 # tamanho da tabela do tremolo
+fv=1.5 # frequência do tremolo
+mu=12. # desvio maximo em dB do tremolo (profundidade)
+f=40. # freq do som em si
+D=1024 # tamanho da tabela do som
+dur=2 # duração
+
+x=n.linspace(0,2*n.pi,Dv,endpoint=False)
+tabv=n.sin(x) # tabela senoidal para o tremolo
+
+# Padrao do vibrato
+ii=n.arange(fa * dur) # amostras em dur segundos
+gv=n.array(ii*fv*float(D)/fa, n.int) # indices para pegar na tabela
+
+### Som em si
+tab=n.linspace(-1,1,D) # dente de serra
+tv=10**(tab[gv%D]*mu/20) # desvio instantaneo de amplitude para cada amostra
+
+gi=n.array(ii*f*(Dv/float(fa)), n.int) # a movimentacao na tabela total, jah inteiro
+ti=tabv[gi%Dv]*tv
+p.plot(ti,label=r"$T_i^{tr(f'=1,5Hz)}=\{t_i.a_i\}_0^{\Lambda-1}$", linewidth=2)
+ti=((ti-ti.min())/(ti.max()-ti.min()))*2-1 # normalizando
+a.wavwrite(ti,"tremolo.wav",fa)
+
+gi=n.array(  ii * (D/float(fa)) * f  , n.int ) % D
+t=tab[ gi ]
+a.wavwrite(t,"original.wav",fa)
+
+p.ylabel(r"amplitude $\quad \rightarrow $", fontsize=16)
+p.xlabel(r"$i\quad \rightarrow$",fontsize=26)
+
+p.xlim(-2000,ii[-1]+2000)
+p.ylim(-4.3,6)
+p.xticks((0,20000,40000,60000,80000,88200),(r"0",20000,40000,60000,80000,88200))
+
+p.plot(tv, label =r"$a_i=10^{t_i'\,\frac{V_dB=12\,dB}{20}}$", linewidth=4 )
+p.legend(loc="upper left")
+ltext = p.gca().get_legend().get_texts()
+p.setp(ltext[0], fontsize = 20)
+p.setp(ltext[1], fontsize = 20)
+p.show()

### 2012-10-02 scripts/vibrato.py

@@ -3,46 +3,40 @@ import numpy as n, pylab as p, scikits.audiolab as a
 
 fa=44100
 Dv=2048 # tamanho da tabela do vibrato
-fv=6.
-nu=12. # desvio maximo em semitons do vibrato
-f=2440. # freq do som em si
+fv=1.5 # frequência do vibrato
+nu=12. # desvio maximo em semitons do vibrato (profundidade)
+f=1040. # freq do som em si
 D=1024 # tamanho da tabela do som
 dur=2
 
 x=n.linspace(0,2*n.pi,Dv,endpoint=False)
 tabv=n.sin(x) # tabela senoidal para o vibrato
 
+# Padrao do vibrato
 ii=n.arange(fa * dur) # 2 segundos
-
 gv=n.array(ii*fv*float(Dv)/fa, n.int) # indices
-tv=tabv[gv%Dv]*nu # amostras do vibrato em desvios de semitons
+tv=tabv[gv%Dv]*nu # desvio instantaneo de semitons para cada amostra
+fi=f*(  2.**(  tv/12.  )   ) # frequência em Hz em cada amostra
 
-fi=f*(  2.**(  tv/12.  )   ) # amostras do vibrato em Hz
+### Som em si
+tab=n.linspace(-1,1,D) # dente de serra
 
-dD=fi*(Dv/float(fa))
-gi=n.cumsum(dD,0,n.int)
-#gi=n.array(gi,n.int)
-
-#for i in xrange(1,len(tv)):
-#    gi[i]=gi[i-1]+fi[i-1]*(Dv/float(fa))
-#gi=n.array(gi,n.int)
-#gi=n.array( fi * (Dv/float(fa)  ) +  ii * f  * (Dv/float(fa)  ) , n.int ) % Dv
-
-#gi=  ii * (Dv/float(fa)) * f  
-#gi+=(fi-f)*(Dv/float(fa))
-#gi=n.array(gi,n.int)
-
-ti=tabv[gi%Dv]
+dD=fi*(D/float(fa)) # a movimentação na tabela por amostra
+gi=n.cumsum(dD,0,n.int) # a movimentacao na tabela total, jah inteiro
+ti=tabv[gi%D]
 a.wavwrite(ti,"vibrato.wav",fa)
 
-ff=n.ones(ii.shape[0])*f
-gi=n.array(  ii * (Dv/float(fa)) * ff  , n.int ) % Dv
[…]

### 2012-10-07 devEresultados.tex
commit geralzao, mta coisa sendo feita

@@
     \caption{Convolução com o impulso e linhas de delays}
-        \includegraphics[width=\textwidth]{figuras/delays_}
+        \includegraphics[width=\textwidth]{figuras/delays__}
         \label{fig:delays}
@@
 conhecida pela sigla IIR (do inglês Infinite Impulse Response)
-e é caracterizada por possuir uma representação amostral (resposta
-ao impulso) infinita. Sua aplicação é usualmente feita pela equação
+e é caracterizada por possuir uma representação temporal
+infinita, i.e. a resposta ao impulso não converge para zero. Sua aplicação é usualmente feita pela equação
 a diferenças:
@@
 \begin{equation}
-t_i' = \frac{1}{a_0}\left ( \sum_{j=0}^Jb_j . t_{i-j} \sum_{k=1}^Ka_k . t_{i-k}' \right )
-\end{equation}
-
-Apontamos filtros IIR paradigmáticos e bastante úteis e usuais abaixo. Existem
-diversos métodos e ferrametas para a elaboração de filtros IIR,
-listamos abaixo uma seleção com fins didáticos e para consulta futura\footnote{Vale também
-apontar que todos os filtros abaixo são utilizados para equalização.}.
-São filtros de primeira ordem. Rotinas excelentes para cálculo de coeficientes
+t_i' = \frac{1}{b_0}\left ( \sum_{j=0}^Ja_j . t_{i-j} - \sum_{k=1}^Kb_k . t_{i-k}' \right )
+\end{equation}
+
+com $b_0=1$ na grande maioria dos casos pois basta normalizarmos as variáveis
+resultando em filtro idêntico: $a_j'=\frac{a_j}{b_0}$ e $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
+
+Apontamos filtros IIR bastante simples, úteis e usuais abaixo. Existem
+diversos métodos e ferrametas para a elaboração de filtros IIR
+e esta é uma seleção com fins didáticos e para consulta futura por
+utilidade.
+São filtros de primeira e segunda ordem bem comportados cujas
+filtragens realizadas dispomos na figura \ref{fig:iir}.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda}}
+        \includegraphics[width=\textwidth]{figuras/iir_}
+        \label{fig:iir}
+\end{figure}
+
+No caso dos filtros simples, a frequência de corte $f_c$ é onde 
+o filtro realiza uma atenuação de $-3dB \approx 0.707 $ da energia original.
+No caso dos filtros passa e rejeita banda, esta mesma atenuação é
+resultado de duas especificações: $f_c$ e a largura de banda $bw$
+(em ambas as frequências $f \pm bw$ há uma atenuação de 0.707 da energia original).
+Para concordar com a nossa audição, as energias em cada frequência estão dispostas em escala
+logarítmica.
+
+Para conseguir filtros cujas respostas em frequência possuem formatos específicos,
+o recurso mais em mãos aqui é fazer cascata destes filtros (aplicá-los sucessivamente) ou usar um biquad\footnote{Abreviação
+de 'biquadrado' pois sua função de transferência possue dois polos e dois zeros, i.e. sua
+forma normal consiste em dois polinômios quadráticos formando uma fração com um no numerador e o outro no denominador
+$\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1+ b_1.z^{-1} +b_2 . z^{-2}}$.}.
+
+Rotinas para cálculo de coeficientes
 de filtros Chebichev\footnote{Filtros Butterworth e Elípticos podem
@@
 \begin{enumerate}
-\item Passa-baixas: polo simples com $a_0=0,15$ e $b_1=0,85$, fórmula geral para $x=e^{-2\pi f_c},\; x \in [0,1]$:
-
-\begin{equation}
+\item Passa-baixas de polo simples. Fórmula geral:
+com a referência da frequência de corte $f_c \in (0,\frac{1}{2})$,
+fração da frequência de amostragem $f_a$
+em que se conseque aproximadamente uma atenuação de $-3dB$,
+calculamos os coeficientes do filtro IIR
+$a_0$ e $b_1$ 
+através da variável intermediária: $x \in [0,1]$:
+
+\begin{equation}\label{eq:passa-baixas}
 \begin{split}
+x & =e^{-2\pi f_c} \\
 a_0 & =  1-x \\
@@
 
-\item Passa-altas: polo simples com $a_0=0,93$, $a_1=-0.93$ e $b_1=0.86$ para $x=e^{-2\pi f_c},\; x \in [0,1]$.
-
-\begin{equation}
+\item Passa-altas de polo simples. A fórmula geral,
+com frequência de corte $f_c \in (0,\frac{1}{2})$, fração
+de $f_a$ em que conseguimos aproximadamente uma atenuação de $3dB$, calculada através da variável
+intermediária $x \in [0,1]$.
+
+\begin{equation}\label{eq:passa-altas}
 \begin{split}
-a_0 & =  \frac{1+x}{2} \\
-a_1 & =  -\frac{1+x}{2} \\
+x & =e^{-2\pi f_c} \\
+a_0 & =  \frac{x+1}{2} \\
+a_1 & =  -\frac{x+1}{2} \\
 b_1 & =  x
@@
 e as variáveis auxiliares $K$ e $R$ definidas como\footnote{A menor largura
-de banda que pode ser utilizada com precisão simples é por volta de $BW=0,0003$}:
+de banda que pode ser utilizada com precisão simples é por volta de $BW=0,0003$.}:
 
@@
 
-\begin{equation}
+\begin{equation}\label{eq:passa-banda}
 \begin{split}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:rejeita-banda}
 \begin{split}

### 2012-10-07 scripts/LUT.py

@@ -1,0 +1,13 @@
+#-*- coding: utf8 -*-
+import pylab as p, numpy as n
+
+#p.xlabel(r"$\gamma_i=i.f\frac{\widetilde{\Lambda}}{f_a}$", fontsize=26)
+p.xlabel(r"amostras $\rightarrow$", fontsize=26)
+p.plot(n.sin(n.linspace(0,2*n.pi,1024, endpoint=False)),"ro",ms=8)
+int(155*f*T/fa)%T
+
+p.xlim(745,804)
+p.ylim(-1.005,-0.97)
+
+p.show()
+

### 2012-10-07 scripts/delays.py

@@ -22,13 +22,14 @@ p.plot(s+6,'bo', label=r"$\{t_i\}_0^{\Lambda-1}$")
 p.plot(hi+3,'ro', label=r"$\{h_i=\delta_{\xi-i}\}_0^{\Lambda_h-1}$")
 p.plot(som,'go', label=r"$\{(t*h)_i\}_0^{\Lambda+\Lambda_h-2}$")
 p.legend(loc="upper right")
-p.xticks((0,ii-1,xi-1,xi+ii-1,ii+iii-2),(r"0",r"$\Lambda-1$",r"$\xi$",r"$\xi + \Lambda-1$", r"$\Lambda+\Lambda_h-2$"),fontsize='16')
+p.xticks((0,ii-1,xi-1,xi+ii-1,ii+iii-2, iii-1),(r"0", r"$\Lambda-1$",r"$\xi$", r"$\xi + \Lambda-1$", r"$\Lambda+\Lambda_h-2$", r"$\Lambda_h$"),fontsize='16')
 p.yticks((),())
 p.ylim(-1.3,7.3)
 p.xlim(-5,ii+iii-2+5)
 p.plot([ii-1,ii-1],[-2,9],"y", linewidth=5,alpha=.4)
+p.plot([len(hi)-1,len(hi)-1],[-2,9],"y", linewidth=5,alpha=.4)
 p.plot([xi -1,xi -1],[-2,9],"y", linewidth=5,alpha=.4)
-p.plot([xi +ii-1,xi +ii-1],[-2,9],"y", linewidth=5,alpha=.4)
+p.plot([xi +ii-1,xi +ii-1],[-2,9],"y-.", linewidth=5,alpha=.4)
 
 p.ylabel(r"deslocamento", fontsize=16, fontweight='bold')
 
@@ -46,8 +47,8 @@ xis=n.array([330, 173, 629, 387,  40, 617, 373, 522, 416, 326, 290, 264,  16, 83
 hi[xis]=n.ones(pulsos)
 som = n.convolve(hi,s)
 
-p.plot(hi+3,'ro', label=r"$\{h_i=\sum_{j=0}^{\Lambda_j-1}\delta_{\xi_j-i}\}_0^{\Lambda_h-1}$")
-p.plot(som,'go', label=r"$\{(t*h)_i\}_0^{\Lambda+\Lambda_h-2}$")
[…]

### 2012-10-07 scripts/filters/espectrosJanela.py

@@ -1,29 +1,26 @@
 #-*- coding: utf8 -*-
 import numpy as n, pylab as p
-ifft=n.fft.ifft
-ones=n.ones
-subplot=p.subplot
-plot=p.plot
-show=p.show
 
+N=60000 # N par
 
-N=101
+coeffs=n.random.random((N,2))*200-100
 
-e=ones(N)
-e[0]=0
-sa=[]
-for i in xrange(1,(N-1)/2+1):
-    e[i]=0
-    e[-i]=0
-    s=ifft(e)
-    sa.append(s)
+# Fazer complexos com os primeiros N elementos
+# depois normalizar para norma ==1
 
+j=n.complex(0,1)
 
+coeffs=coeffs[:,0]+coeffs[:,1]*j
+coeffs=coeffs/n.abs(coeffs)
 
+# real par, imaginaria impar
+coeffs[N/2+1:]=n.real(coeffs[1:N/2])[::-1] - j*n.imag(coeffs[1:N/2])[::-1]
 
-#i=0
-#for s in ss:
-    #i+=1
-    #subplot("%i1%i" % (len(ss),i))
-    #plot(s)
-#show()
+coeffs[0]=0.
+coeffs[N/2]=0.9
+
+noise=n.fft.ifft(coeffs)
+
+p.plot(n.real(noise))
+p.show()
+

### 2012-10-07 scripts/filters/iir.py

@@ -1,0 +1,191 @@
+#-*- coding: utf8 -*-
+import numpy as n, pylab as p
+
+def lp(fc, npoints=1000):
+    x=n.e**(-2*n.pi*fc) # fc=> freq de corte em 3dB
+    # coeficientes
+    a0=1-x
+    b1=x
+    # aplicação da equação a diferenças
+    sinal=[a0]
+    for i in xrange(npoints):
+        sinal.append(sinal[-1]*b1)
+    # cálculo do espectro
+    fft=n.fft.fft(sinal)
+    m=n.abs(fft)
+    return m
+
+def hp(fc, npoints=1000):
+    x=n.e**(-2*n.pi*fc) # fc=> freq de corte em 3dB
+    a0=(1+x)/2
+    a1=-(1+x)/2
+    b1=x
+    sinal=[a0]
+    sinal+=[a1+sinal[-1]*b1]
+    # aplicação da equação a diferenças
+    for i in xrange(npoints):
+        sinal.append(sinal[-1]*b1)
+    # cálculo do espectro
+    fft=n.fft.fft(sinal)
+    m=n.abs(fft)
+    return m
+
+def notch(f,bw,ftype='bs'):
+    """Use ftype=='bp' para filtro passa banda """
+    if ftype=='bs':
+        r=1-3*bw
+        k=(1-2*r*n.cos(2*n.pi*f)+r**2)/(2-2*n.cos(2*n.pi*f))
+        
+        # coefs rejeita banda
+        a0=k
+        a1=-2*k*n.cos(2*n.pi*f)
+        a2=k
+        b1=2*r*n.cos(2*n.pi*f)
+        b2=-r**2
+        
+        sinal=[a0]
+        sinal+=[a1+sinal[-1]*b1]
+        sinal+=[a2+sinal[-1]*b1+sinal[-2]*b2]
+        for i in xrange(44100):
+            sinal.append(sinal[-1]*b1+sinal[-2]*b2)
+        fft=n.fft.fft(sinal)
+        m=n.abs(fft)
+        return m
+    else:
+        r=1-3*bw
[…]

### 2012-10-08 devEresultados.tex
terminando discussao dos filtros iir com o biquad

@@
     \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda}}
-        \includegraphics[width=\textwidth]{figuras/iir_}
+        \includegraphics[width=\textwidth]{figuras/iir___}
         \label{fig:iir}
@@
 
-No caso dos filtros simples, a frequência de corte $f_c$ é onde 
+No caso dos filtros de ordem simples, a frequência de corte $f_c$ é onde 
 o filtro realiza uma atenuação de $-3dB \approx 0.707 $ da energia original.
@@
 resultado de duas especificações: $f_c$ e a largura de banda $bw$
-(em ambas as frequências $f \pm bw$ há uma atenuação de 0.707 da energia original).
-Para concordar com a nossa audição, as energias em cada frequência estão dispostas em escala
-logarítmica.
+pois em ambas as frequências $f \pm bw$ há uma atenuação de 0.707 da energia original.
+
+Existe amplificação do som no caso dos filtros passa e rejeita banda, quando a frequência
+de corte é baixa e a largura de banda é larga o suficiente. Nos agudos, quando a largura
+de banda é grande o suficiente, estes filtros apresentam
+somente um desvio do perfil esperado, expandindo a envoltória um pouco mais do lado grave da banda em
+evidência.
 
 Para conseguir filtros cujas respostas em frequência possuem formatos específicos,
-o recurso mais em mãos aqui é fazer cascata destes filtros (aplicá-los sucessivamente) ou usar um biquad\footnote{Abreviação
+o recurso mais simples a esta altura
+é fazer cascata destes filtros (aplicá-los sucessivamente).
+
+Outra possibilidade é usar o último filtro que desta
+coleção: o biquad\footnote{Abreviação
 de 'biquadrado' pois sua função de transferência possue dois polos e dois zeros, i.e. sua
@@
 $\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1+ b_1.z^{-1} +b_2 . z^{-2}}$.}.
+Existem diversas receitas para esta classe de filtro e
+recomendamos ~\ref{music-dsp} por possuir as 4 funcionalidades da figura \ref{fig:iir}.
 

### 2012-10-08 scripts/filters/biquad.py

@@ -1,0 +1,18 @@
+#-*- coding: utf8 -*-
+import numpy as n, pylab as p
+
+fa=44100.
+f0=100
+
+dBgain=-20. # para frequência central
+
+
+
+
+
+
+
+
+
+
+

### 2012-10-08 scripts/filters/iirlog.py

@@ -74,26 +74,43 @@ fa=44100.
 
 ax=p.subplot(221)
 #ax.set_xscale('log')
-fcs=[0.005,0.05,0.1,0.2,0.3,0.4]
-N=10000
-poss=[(.1*(N/2),0.15),(0.25*(N/2.),0.40),(.37*(N/2.),0.53),(.5*(N/2),0.7),(0.6*(N/2.),0.82),(0.7*(N/2.),0.90)]
+fcs=[0.0000001,0.000001,0.002,0.01,0.3]
+poss=[(2,-30),(3,-22),(4,-18),(5,-10),(5.5,-4.55)]
+N=1000
 ds=fa/N
 fs=[i*ds for i in xrange(1,N/2+1)]
 fs=n.log2(fs)
 fs=n.hstack(([2*fs[0]-fs[1]],fs))
-#p.plot((fs[0],fs[0]),(-100,100),'c',linewidth=3)
+p.plot((fs[0],fs[0]),(-100,100),'c',linewidth=3)
 for fc,pos in zip(fcs,poss):
     m=lp(fc,N)
-    p.plot(m[:len(m)/2+1])
+    #fs=n.logspace(n.log2(fa/N),n.log2(fa/2),len(m)/2-1,base=2)
+    #fs=n.hstack(([0],fs))
+    #fs[0]=0
+    #p.plot(20*n.log10(m[:len(m)/2]))
+    p.plot(fs,20*n.log10(m[:len(m)/2+1]),'ro')
+    p.plot(fs,20*n.log10(m[:len(m)/2+1]))
     p.text(pos[0],pos[1],r"$f_c=%s$" % (fc,))
 #    p.plot([fc*len(m),fc*len(m)],[-1000,1000])
 ii=range(1,11)
-p.ylabel(u"amplitude"+r"$\; \rightarrow$")
-p.xlabel(u"frequência "+r"$  \; \rightarrow$")
-p.xticks((0,len(m)/8,len(m)/4,3*len(m)/8,len(m)/2),(r"0",r"$\frac{f_a}{8}$", r"$\frac{f_a}{4}$", r"$\frac{3 . f_a}{8}$", r"$\frac{f_a}{2}$"),fontsize='16')
+p.text(0.2,-43,r"$ f_c \sim $ 44, 176, 882, 4410 e 13230 $Hz$", fontsize=16)
+p.text(0.2,-48,r"em $44,1kHz$", fontsize=16)
[…]

### 2012-10-09 devEresultados.tex
adentrada descricao e sintese de ruidos

@@
 
-Outra possibilidade é usar o último filtro que desta
-coleção: o biquad\footnote{Abreviação
+Outra possibilidade é usar alguma receita de filtro
+biquad\footnote{Abreviação
 de 'biquadrado' pois sua função de transferência possue dois polos e dois zeros, i.e. sua
 forma normal consiste em dois polinômios quadráticos formando uma fração com um no numerador e o outro no denominador
-$\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1+ b_1.z^{-1} +b_2 . z^{-2}}$.}.
-Existem diversas receitas para esta classe de filtro e
-recomendamos ~\ref{music-dsp} por possuir as 4 funcionalidades da figura \ref{fig:iir}.
-
-Rotinas para cálculo de coeficientes
+$\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1+ b_1.z^{-1} +b_2 . z^{-2}}$.}
+ou rotinas para cálculo de coeficientes
 de filtros Chebichev\footnote{Filtros Butterworth e Elípticos podem
-ser considerados como casos específicos dos Filtros do tipo Chebichev.} podem
-ser encontradas nas nossas referências, em especial em ~\ref{JOS} e ~\ref{dspguide}.
+ser considerados como casos específicos dos Filtros do tipo Chebichev.}.
+Ambas as possibilidades são exploradas
+por títulos em nossas referências, em especial em ~\ref{JOS} e ~\ref{dspguide} e na coleção de filtros da comunidade \emph{Music DSP}, da Universiade de Columbia \ref{music-dsp}.
 Para uma exposição aprofundada do assunto, recomendamos ~\ref{Openheim}.
@@
 \begin{enumerate}
-\item Passa-baixas de polo simples. Fórmula geral:
-com a referência da frequência de corte $f_c \in (0,\frac{1}{2})$,
+\item Passa-baixas de polo simples. A fórmula geral tem
+por referência da frequência de corte $f_c \in (0,\frac{1}{2})$,
 fração da frequência de amostragem $f_a$
-em que se conseque aproximadamente uma atenuação de $-3dB$,
+em que há aproximadamente uma atenuação de $3dB$,
 calculamos os coeficientes do filtro IIR
@@
 %\item Rejeita-banda
-\item Nó: dadas a frequência $f$ e a largura de banda $BW$ (medida em 0,707 da amplitude),
-ambos dados como fracoes de $f_a$, portanto $f,BW \in [0,0,5]$,
-e as variáveis auxiliares $K$ e $R$ definidas como\footnote{A menor largura
-de banda que pode ser utilizada com precisão simples é por volta de $BW=0,0003$.}:
+\item Nó (\emph{notch filter}). Este filtro é parametrizado
+pela frequência central $f$ e a largura de banda $bw$
+(medida em 0,707 da amplitude, i.e. atenuação de $3dB$),
+ambos dados como fracoes de $f_a$, portanto $f,\; bw \in (0,0,5)$.
+Por conveniência definimos as variáveis auxiliares $K$ e $R$ como:
 
@@
 
-Temos o filtro nó passa banda:
+A partir das quais temos dois filtros. O filtro passa banda:
 
@@
 
-e o nó rejeita banda:
+e o filtro rejeita banda:
 
@@
 
-\item Biquad: pela especificação de uma frequência central, da qualidade
-e da intensidade do filtro, este filtro é simples e usual para áudio,
-permitindo ajustes mais finos. Diversas receitas podem ser encontradas
-na literatura, recomendamos especialmente as diferentes especificações
-em ~\ref{musicDSP} e ~\ref{dspguide}.
+%\item Biquad: pela especificação de uma frequência central, da qualidade
+%e da intensidade do filtro, este filtro é simples e usual para áudio,
+%permitindo ajustes mais finos. Diversas receitas podem ser encontradas
+%na literatura, recomendamos especialmente as diferentes especificações
+%em ~\ref{musicDSP} e ~\ref{dspguide}.
 
@@
 \subsubsection{Ruídos}
-Geração estatística de ruídos, distribuições, nomes, equações de geração.
+De forma geral, os sons sem altura definida 
+são chamados ruídos.
+Estes sons estão presentes como constituintes importantes dos sons musicais de altura definida,
+como os ruídos presentes nas notas do piano, do violino, etc. Além disso, os instrumentos
+de percussão, em grande parte, não possuem altura definida e seus sons
+são muitas vezes compreendidos como ruídos. Na música eletrônica (incluindo
+a eletroacústica), os ruídos possuem usos amplos, diversificados e comumente
+idiomáticos.
+
+A ausência de uma altura definida é fruto da ausência de características
+harmônicas perceptíveis nas componentes do som. Assim, pode-se inferir
+as incontáveis possibilidades de gerar ruídos. A mais imediata é a utilização
+de valores aleatórios para a geração da sequência sonora $T_i$. Este
+é um método atraente, visto a disponibilidade
+das consagradas distribuições de probabilidade em diversas linguagens de programação,
+mas os resultados não são tão uteis, tendendo geralmente ao ruído branco
+e com difíceis previsões do espectro resultante.
+
+A outra possibilidade é a geração de ruído através do espectro desejado. A partir
+dele, executamos a transformada inversa de fourier. Neste caso, é importante
+que realizemos a distribuição espectral com cuidado pois queremos um perfil
+do módulo do espectro, e a fase é aleatória\footnote{Caso utilizemos a mesma fase
+(ou fases com forte correlação) para
+as componentes envolvidas, o som sintetizado possuirá energia bastante concentrada
+em alguns trechos.}.
+
+Abaixo elencamos alguns ruídos chamados \emph{coloridos} por terem
+sido associados a cores. Lidaremos somente com ruídos de espectro estático.
+
+\begin{itemize}
+
+\item O ruído branco possui este nome por possuir energia distribuida
+igualmente por todas as frequências. Podemos obter a realização
+do ruído branco com as transformada inversa dos seguintes coeficientes
+para $\Lambda$:
+
+\begin{equation}\label{eq:ruido-branco}
+\begin{split}
+c_0 & =0 \quad \text{pois não queremos bias} \\
+c_i & =e^{j.x}\;,\;\; x \; \text{randômico} \; \in \; [0,2\pi]\;,\;\; i \; \in \; [1, \, \Lambda/2-1] \\
+c_{\Lambda/2} & = 10^{-\frac{1}{20}} \approx 0.89 \quad \text{se $\Lambda$ par}\\ 
+c_i & = c_{\Lambda - i}^*\;,\;\; \text{para}\;  i \; > \;  \Lambda/2
+\end{split}
+\end{equation}
+
+O valor de $c_i$ calculado pela exponencial é apenas um artifício para resultar em módulo unitário e fase aleatória.
+Já $c_{\Lambda/2}$ é sempre puramente real (como vimos na sessão anterior) e possui a queda de $1dB$ pois não é resultande de duas componentes como as outras frequências, que por isso recebem $1dB$ adicional ($1dB$ equivale à intensidade adicionada com uma fonte a mais).
+
+\item O ruído rosa possui uma queda de $3dB$ por oitava. Este ruído é muito usual no teste de equipamentos e montagens de aparelhos além de presença destacada na natureza.
+
+\begin{equation}
+\begin{split}
+f_i & = i \frac{f_a}{\Lambda} \\
+\alpha_i & = (10^{-\frac{3}{20}})^{log_2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \;\;, \quad f_{\text{min}} \approx 15 Hz \\
+c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \\
+c_i & =e^{j.x} . \alpha_i\;,\;\; x \; \text{randômico} \; \in \; [0,2\pi]\;,\;\; i \; \in \; [i_0...\Lambda/2-1] \\
+c_{\Lambda/2} & = 10^{-\frac{1}{20}} . \alpha_{\Lambda/2} \quad \text{se $\Lambda$ par}\\ 
+c_i & = c_{\Lambda - i}^*\;,\;\; \text{para}\;  i \; > \;  \Lambda/2
+\end{split}
+\end{equation}
+
+\end{itemize}
+
+%Geração estatística de ruídos, distribuições, nomes, equações de geração.
 

### 2012-10-10 devEresultados.tex
terminada a discussao dos ruidos

@@
 das consagradas distribuições de probabilidade em diversas linguagens de programação,
-mas os resultados não são tão uteis, tendendo geralmente ao ruído branco
+mas os resultados não são tão úteis, tendendo geralmente ao ruído branco
 e com difíceis previsões do espectro resultante.
@@
 igualmente por todas as frequências. Podemos obter a realização
-do ruído branco com as transformada inversa dos seguintes coeficientes
-para $\Lambda$:
+do ruído branco com a transformada inversa dos seguintes coeficientes:
 
@@
 c_0 & =0 \quad \text{pois não queremos bias} \\
-c_i & =e^{j.x}\;,\;\; x \; \text{randômico} \; \in \; [0,2\pi]\;,\;\; i \; \in \; [1, \, \Lambda/2-1] \\
+c_i & =e^{j.x}\;,\;\; j^2=-1 \;, \;\; x \; \text{randômico} \; \in \; [0,2\pi]\;,\;\; i \; \in \; [1, \, \frac{\Lambda}{2}-1] \\
 c_{\Lambda/2} & = 10^{-\frac{1}{20}} \approx 0.89 \quad \text{se $\Lambda$ par}\\ 
@@
 
-\begin{equation}
+\begin{equation}\label{eq:rosa}
 \begin{split}
-f_i & = i \frac{f_a}{\Lambda} \\
-\alpha_i & = (10^{-\frac{3}{20}})^{log_2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \;\;, \quad f_{\text{min}} \approx 15 Hz \\
+f_i & = i \frac{f_a}{\Lambda} \;, \;\; \quad i \leq \frac{\Lambda}{2} \\
+\alpha_i & = (10^{-\frac{3}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \;\;, \quad f_{\text{min}} \approx 15 Hz \\
 c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \\
-c_i & =e^{j.x} . \alpha_i\;,\;\; x \; \text{randômico} \; \in \; [0,2\pi]\;,\;\; i \; \in \; [i_0...\Lambda/2-1] \\
+c_i & =e^{j.x} . \alpha_i\;,\;\; j^2=-1 \;, \;\;\  x \; \text{randômico} \; \in \; [0,2\pi]\;,\;\; i \; \in \; [i_0, \frac{\Lambda}{2}-1] \\
 c_{\Lambda/2} & = 10^{-\frac{1}{20}} . \alpha_{\Lambda/2} \quad \text{se $\Lambda$ par}\\ 
@@
 
+A frequência mínima $f_{\text{min}}$ pode ser escolhida com base no limite da audição, pois não escutamos frequência absoluta abaixo de $\approx\; 20Hz$ como altura.
+
+O resto dos ruídos podem ser feitos com base neste procedimento descrito para 
+o ruído rosa. Basta que modifiquemos alguns detalhes. Em especial a equação que define $\alpha_i$.
+
+\item O ruído marrom deve seu nome ao Robert Brown, que descobriu o movimento browniano.
+Embora esta origem seja um tanto díspar do que poderiamos considerar uma associação ligítima com a cor marrom, o ruído ficou consagrado com este nome.
+De qualquer forma, é bastante comum concordarmos que, dados os ruídos branco e rosa, ao ruído marrom, menos estridente,
+é satisfatória a associação com o ruído marrom.
+
+O que caracteriza este ruído é a queda de $6dB$ por oitava. Desta forma, basta modificarmos a equação de $\alpha_i$ 
+no conjunto \ref{eq:rosa} para:
+
+\begin{equation}
+\alpha_i=(10^{-\frac{6}{20}})^{\log _2 \left( \frac{f_i}{f_{\text{min}}} \right )}
+\end{equation}
+
+\item Ruído azul é aquele em que há um ganho de $3dB$ por oitava em uma banda limitada. Para
+a síntese deste ruído, basta estipularmos a frequência mínima $f_{\text{min}}$ e a frequência
+máxima $f_{\text{máx}}$. Assim, também com base no conjunto de equações \ref{eq:rosa}, 
+basta atentarmos para as sequintes identidades:
+
+\begin{equation}\label{eq:azul}
+\begin{split}
+\alpha_i & = (10^{\frac{3}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \\
+c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \;\; \text{ou} \;\; f_i>f_{\text{máx}} \\
+\end{split}
+\end{equation}
+
+\item O ruído violeta é similar ao ruído azul, mas o ganho é de $6dB$ por oitava:
+
+\begin{equation}\label{eq:violeta}
+\alpha_i = (10^{\frac{6}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \;\;, \quad f_{\text{min}} \approx 15 Hz \\
+\end{equation}
+
+\item O ruído preto possui perdas de mais de $6dB$ por oitava, assim:
+
+\begin{equation}
+\alpha_i=(10^{-\frac{\beta}{20}})^{\log _2 \left( \frac{f_i}{f_{\text{min}}} \right )}\;\;, \quad \beta > 6
+\end{equation}
+
+\item Um pouco diferente dos exemplos acima é o ruído cinza, definido como
+um ruído branco sujeito a uma das curvas iso-audíveis. Como estas curvas são resultados
+experimentais, precisamos neste caso da curva para a realização dos fatores multiplicativos
+$\alpha_i$.
+
 \end{itemize}
+
+Com base nestes ruídos expostos podemos prever a infinidade de ruídos utilizados. Importante
+neste momento é entender que foram expostos somente ruídos com espectro estático. Existem classificações
+de ruídos com variações do espectro no decorrer do tempo. Existem também ruídos que
+são fundamentalmente transientes, como os clicks e os chirps. O primeiro é modelado
+facilmente por um impulso relativamente isolado, enquanto o segundo é uma varredura rápida de 
+alguma banda de frequência.
 

### 2012-10-11 devEresultados.tex
feita figura e breve discussao dos ruidos

@@
 
-\begin{equation}\label{eq:ruido-branco}
+\begin{equation}\label{eq:branco}
 \begin{split}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:marrom}
 \alpha_i=(10^{-\frac{6}{20}})^{\log _2 \left( \frac{f_i}{f_{\text{min}}} \right )}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:preto}
 \alpha_i=(10^{-\frac{\beta}{20}})^{\log _2 \left( \frac{f_i}{f_{\text{min}}} \right )}\;\;, \quad \beta > 6
@@
 alguma banda de frequência.
+
+Os ruídos das equações \ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom},
+\ref{eq:azul}, \ref{eq:violeta} estão na figura \ref{fig:ruidos}. Vale notar
+que os espectros foram feitos com a mesma fase nos coeficientes, de forma que
+pode-se observar a contribuição dos harmônicos agudos no traçado geral dado
+pelas frequências graves e muito explícito no ruído marrom.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda}}
+        \includegraphics[width=\textwidth]{figuras/ruidos___}
+        \label{fig:ruidos}
+\end{figure}
 

### 2012-10-11 scripts/filters/ruidos.py

@@ -3,8 +3,9 @@ import numpy as n, pylab as p, scikits.audiolab as a
 
 N = 100000 # N sempre par
 fa=44100.
-################
-# geracao do ruido branco
+
+# diferença em das frequências entre coeficiêntes vizinhos:
+df=fa/N
 
 # geracao de espectro com modulo 1 uniforme
 # e fase aleatoria
@@ -16,31 +17,104 @@ coefs[N/2+1:]=n.real(coefs[1:N/2])[::-1] - 1j*n.imag(coefs[1:N/2])[::-1]
 coefs[0]=0. # sem bias
 coefs[N/2]=0.9 # freq max eh real e 0.9
 
-ruido=n.fft.ifft(coefs)
-r=n.real(ruido)
-r=((r-r.min())/(r.max()-r.min()))*2-1
-a.wavwrite(r,'branco.wav',44100)
 
-p.plot(n.real(ruido))
-p.show()
-
-
-################3
-# Ruido rosa
-df=fa/N
-fi=n.arange(coefs.shape[0])*df
+# as frequências relativas a cada coeficiente
+fi=n.arange(coefs.shape[0])*df # acima de N/2 nao vale)
 f0=15. # iniciamos o ruido em 15 Hz
 i0=n.floor(f0/df) # primeiro coeff a valer
 coefs[:i0]=n.zeros(i0)
 f0=fi[i0]
 
+
+# para plotar:
+ii=1000
+ie=1200
+
+
+###########
+# Ruido Violeta:
+# a cada oitava, ganhamos 6dB
+fator=10.**(6/20.)
+
+alphai=fator**(n.log2(fi[i0:]/f0))
+c=n.copy(coefs)
+c[i0:]=c[i0:]*alphai
+# real par, imaginaria impar
+
+
+c[N/2+1:]=n.real(c[1:N/2])[::-1] - 1j*n.imag(c[1:N/2])[::-1]
+
+ruido=n.fft.ifft(c)
+r=n.real(ruido)
+r=((r-r.min())/(r.max()-r.min()))*2-1
+a.wavwrite(r,'violeta.wav',44100)
+
+p.subplot(521)
+p.title(u'ruído violeta')
+p.ylim(-10,220)
[…]

### 2012-10-12 devEresultados.tex
arrumando fm

@@
 
-\item O ruído preto possui perdas de mais de $6dB$ por oitava, assim:
+\item O ruído preto possui perdas maiores que $6dB$ por oitava, assim:
 
@@
 \subsubsection{Usos musicais}
+A este ponto as possibilidades musicais explodiram. Possuímos unidades
+musicais com características próprias de altura (dada pela frequência),
+tímbre (dada pela forma de onda),
+volume (dada pela intensidade) e duração (dada pelo número de amostras).
+Em uma nota, estas características podem ser tratadas
+de forma absoluta ou ao longo de sua duração,
+com a única excessão da duração em si.
+
+Desta forma, os usos musicais que apresentamos é uma coleção de possibilidades
+com o objetivo de exemplificar manipulações sonoras que resultem algo
+musical, por razões variadas e abordadas mais a fundo na próxima sessão.
+
+Uma primeira possibilidade interessante é o estabelecimento de vínculos
+entre os parâmetros do trêmolo e do vibrato e algum parâmetro da nota básica,
+digamos a frequência. Assim, podemos estabelecer que quanto mais aguda é a nota,
+mais alta é a frequência do vibrato e do trêmolo mas menor a sua profundidade.
+Desta forma, tomando por base as equações \ref{vbrGamma}, \ref{vbrF} e \ref{trA}
+podemos escrever:
+
+\begin{equation}
+\begin{split}
+f^{vbr} = f^{tr} & = func_a(f) \\
+\nu & = func_b(f) \\
+V_{dB} & = func_c(f)
+\end{split}
+\end{equation}
+
+Com $f^{vbr}$ e $f^{tr}$ sendo $f'$ nas equações de referência, ou seja, a frequência
+de oscilação do vibrato e do tremolo. Já $\nu$ e $V_{dB}$ são as profundidades
+do vibrato e do tremolo, respectivamente. A montagem musical 
+\emph{tremolos, vibratos e a frequência} explora
+recursos como este e variações da forma de onda da oscilação. As possibilidades
+são numerosas e ricas e, aliás, os tremolos e os vibratos naturais ocorrem
+juntos, i.e. quase todo tremolo natural também apresenta oscilação de frequência
+e vice versa.
+
+Estas oscilações são ornamentos muito apreciados e possuem nomes
+específicos. Em linguagem pianística,
+os tremolos são ocorrências em que se alternam rapidamente 2 notas (com
+e casos de expansão), ou seja, seriam mais facilmente associados aos vibratos
+segundo a nomenclatura que estamos usando\footnote{Nossa nomenclatura tem por base
+a teoria musical tradicional, que abarca nomenclaturas de diversos intrumentos e
+tradições, de forma que não conflui com algumas definições de domínios mais específicos.}.
+
+Algo interessante ocorre se aumentarmos progressivamente $f'$.
+Primeiramente o limiar de frequência
+para audição do fenômeno sonoro como
+altura ($\approx 20Hz$) gera rugosidades para
+ambos tremolos e vibratos. A sequência \emph{Rugosidades}
+explora este limiar com concomitâncias de tremolos e vibratos na mesma
+voz, com intensidades e formas de onda diferentes.
+
+Se continuarmos aumentando a frequência, deixamos de ouvir estas oscilações
+como eventos identificáveis. Neste caso, as oscilações se apresentam como espectro
+audível como altura, $f'$ e a forma de onda realizam alterações espectrais no som
+original $T_i$ de formas diferentes para os tremolos e para os vibratos, são as 
+chamadas sínteses AM (Amplitude Modulation) e FM (Frequency Modulation).
+Estas são técnicas muito poderosas e conhecidas, com aplicações em sintetizadores
+históricos como o Yamaha XXX e com aplicações fora da música, como em telecomunicações
+(ex. radios AM e FM).
+
+Para nosso caso, em resumo, podemos entender a síntese FM através do caso entre senóides
+e decompor os sinais em seus espectros de Fourier (i.e. senoidais) para casos mais complexos.
+Assim, a síntese FM realizada com um vibrato de frequência $f'$ em um som $T_i$ de frequência $f$
+gera bandas centradas em $f$ e distantes $f'$ entre si:
+
+\begin{equation}
+\begin{split}
+\{t_i'\}_0^{\Lambda -1} & = \left \{ \cos(f . 2 \pi \frac{i}{\Lambda-1} + \nu . sen( f' . 2 \pi \frac{i}{\Lambda-1} )  ) \right \}_0^{\Lambda-1}  = \\
+ & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\nu) \cos \left ( f . 2 \pi \frac{i}{\Lambda-1} + k . f' . 2 \pi \frac{i}{\Lambda-1} \right )  \right \}_0^{\Lambda-1} = \\
+ & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\nu) \cos \left ( (f+k.f') . 2 \pi \frac{i}{\Lambda-1} \right )  \right \}_0^{\Lambda-1}
+\end{split}
+\end{equation}
+
+onde:
+
+\begin{equation}
+J_k(v) = 
+\end{equation}
+
+filtro parametrizado auto com fmin fmax min e erro admitido (gera fir)
+som de radio com filtragem q tira o que nao estiver entre 700Hz e 3kHz
+chorus c a soma de varios do mesmo com pequenas variacoes de freq, amp
+e flitragens.
+
 Oscilacoes, tremolos e vibratos. Variacoes

### 2012-10-12 devEresultados.tex
ajeitado FM, exmplorando

@@
 Assim, a síntese FM realizada com um vibrato de frequência $f'$ em um som $T_i$ de frequência $f$
+e duração $\Delta$ segundos (assumimos $\Lambda = \Delta . f_a $ amostras)
 gera bandas centradas em $f$ e distantes $f'$ entre si:
@@
 \begin{split}
-\{t_i'\}_0^{\Lambda -1} & = \left \{ \cos(f . 2 \pi \frac{i}{\Lambda-1} + \nu . sen( f' . 2 \pi \frac{i}{\Lambda-1} )  ) \right \}_0^{\Lambda-1}  = \\
- & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\nu) \cos \left ( f . 2 \pi \frac{i}{\Lambda-1} + k . f' . 2 \pi \frac{i}{\Lambda-1} \right )  \right \}_0^{\Lambda-1} = \\
- & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\nu) \cos \left ( (f+k.f') . 2 \pi \frac{i}{\Lambda-1} \right )  \right \}_0^{\Lambda-1}
+\{t_i'\}_0^{\Lambda -1} & = \left \{ \cos \left [f . 2 \pi \frac{i}{f_a-1} + \mu . sen \left ( f' . 2 \pi \frac{i}{ f_a -1 } \right ) \right ] \right \}_0^{\Lambda-1}  = \\
+ & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ f . 2 \pi \frac{i}{f_a-1} + k . f' . 2 \pi \frac{i}{f_a-1} \right ]  \right \}_0^{\Lambda-1} = \\
+ & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ (f+k.f') . 2 \pi \frac{i}{f_a-1} \right ]  \right \}_0^{\Lambda-1}
 \end{split}
@@
 
-onde:
-
-\begin{equation}
-J_k(v) = 
+onde 
+\begin{equation}
+J_k(\mu) = \frac{2}{\pi} \int_0^{\frac{\pi}{2}}\left [ cos \left (\overline{k}\;\frac{\pi}{2} + \mu .
+\sin w \right ) .
+cos \left ( \overline{k}\;\frac{\pi}{2} + k .
+w \right ) \right ] dw \quad , \quad \overline{k} = k \% 2 \;\;,\;\; k \in \mathbb{N}
+\end{equation}
+é a função de Bessel \cite{BesselCCRMA,JOSFM}.
+Note que, nestas equações, a variação de frequência não respeita a progressão geométrica de frequência que acompanha a percepçao de altura.
+Caso utilizemos as equações \ref{vbrF} teríamos uma distribuição um pouco diferente dos harmônicos:
+
+\begin{equation}
+\begin{split}
+\{t_i'\}_0^{\Lambda -1} & = \left \{ \cos \left [f_i .
+2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1}  = \left \{ \cos \left [f\left ( \frac{f+\mu}{f} \right )^{sen \left ( f' .
+2 \pi \frac{i}{ f_a -1 } \right )} .
+2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1}  =  \\
+ & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ f . 2 \pi \frac{i}{f_a-1} + k . f' . 2 \pi \frac{i}{f_a-1} \right ]  \right \}_0^{\Lambda-1} = \\
+ & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ (f+k.f') . 2 \pi \frac{i}{f_a-1} \right ]  \right \}_0^{\Lambda-1}
+\end{split}
 \end{equation}

### 2012-10-13 devEresultados.tex
terminada parte 2, falta tratar do som stereo

@@
 
-\begin{equation}
+\begin{equation}\label{ex:mixagem}
 \{t_i\}=\left \{ \sum_{k=0}^{N-1}t_{k,i} \right \}
@@
 
-Estas oscilações são ornamentos muito apreciados e possuem nomes
+Estas oscilações - aplicadas à amplitude, frequência, posição - são ornamentos muito apreciados e possuem nomes
 específicos. Em linguagem pianística,
 os tremolos são ocorrências em que se alternam rapidamente 2 notas (com
-e casos de expansão), ou seja, seriam mais facilmente associados aos vibratos
-segundo a nomenclatura que estamos usando\footnote{Nossa nomenclatura tem por base
-a teoria musical tradicional, que abarca nomenclaturas de diversos intrumentos e
-tradições, de forma que não conflui com algumas definições de domínios mais específicos.}.
+casos de expansão do conceito), ou seja, seriam mais facilmente associados aos vibratos
+segundo a nomenclatura aqui utilizada\footnote{Veja nosso glossário.}.
 
@@
 Se continuarmos aumentando a frequência, deixamos de ouvir estas oscilações
-como eventos identificáveis. Neste caso, as oscilações se apresentam como espectro
+como eventos identificáveis. Neste caso, as oscilações se apresentam na forma de espectro
 audível como altura, $f'$ e a forma de onda realizam alterações espectrais no som
@@
 Assim, a síntese FM realizada com um vibrato de frequência $f'$ em um som $T_i$ de frequência $f$
-e duração $\Delta$ segundos (assumimos $\Lambda = \Delta . f_a $ amostras)
+e duração $\Delta$ segundos (assumimos $\Lambda = \lfloor \Delta . f_a \rfloor $ amostras)
 gera bandas centradas em $f$ e distantes $f'$ entre si:
@@
 2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1}  =  \\
- & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ f . 2 \pi \frac{i}{f_a-1} + k . f' . 2 \pi \frac{i}{f_a-1} \right ]  \right \}_0^{\Lambda-1} = \\
- & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ (f+k.f') . 2 \pi \frac{i}{f_a-1} \right ]  \right \}_0^{\Lambda-1}
+ & = \left \{ \cos \left [f\left ( 2^{\frac{\nu}{12}} \right )^{sen \left ( f' .
+2 \pi \frac{i}{ f_a -1 } \right )} .
+2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1} =\left \{ \cos \left [f\left ( 2^{\frac{\nu}{12} \sin \left ( f' .
+2 \pi \frac{i}{ f_a -1 } \right )  } \right ) .
+2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1} \Rightarrow \\
+& \Rightarrow \text{usando: } \left ( 2^x=e^{x \ln 2}, \;\; e^x=\sum_0^\infty \frac{x^n}{n!},\;\; \sin^nx=y_1(n)\sum_k y_2(n,k).\sin[y_3(n,k,x)]   \right ) \Rightarrow \\
+\Rightarrow \{t_i'\}_0^{\Lambda -1} & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu,f) \cos \left [ y(f.f', f_a) \right ]  \right \}_0^{\Lambda-1}
 \end{split}
@@
 
-filtro parametrizado auto com fmin fmax min e erro admitido (gera fir)
-som de radio com filtragem q tira o que nao estiver entre 700Hz e 3kHz
-chorus c a soma de varios do mesmo com pequenas variacoes de freq, amp
-e flitragens.
-
-Oscilacoes, tremolos e vibratos. Variacoes
-~15Hz tremolo rugoso
-varias possibilidades de combinacao de intensidades, formas de onda e frequencias
-
-progressivas das oscilacoes dos tremolos e vibratos
-como AM e FM.
-ADSR
-Linhas de delays
-Filtragem dependente do tempo
-cascatas de filtros
-Músicas: monolitos, minimum-fi
+Onde, por brevidade, citamos as identidades úteis para resultar na
+forma final do som resultante, que explicita o conteúdo em frequências puras.
+Podemos observar que, para a FM, se utilizada a progressão exponencial de frequência, o resultado
+é menos comportado. A função de Bessel dependerá não somente da profundidade de oscilação,
+mas também da frequência original do som. Além disso, as frequências dependem do produto das frequências
+do som e da oscilação ($f.f'$), o que também - a princípio - não convém por introduzir complexidade.
+
+O caso da modulação de amplitude (AM) é mais simples:
+
+\begin{equation}\label{eq:am}
+\begin{split}
+\{t_i'\}_0^{\Lambda-1} & =\{(1+a_i) .
+t_i\}_0^{\Lambda-1}= \left \{ \left [ 1+M.\sin \left ( f'.2\pi\frac{i}{f_a -1} \right ) \right] .
+P .\sin \left ( f.2\pi\frac{i}{f_a -1} \right ) \right \}_0^{\Lambda-1} = \\
+& = P.\sin \left( f.2\pi\frac{i}{f_a -1}  \right ) + \frac{P.M}{2} \left [ \sin \left( (f-f').2\pi\frac{i}{f_a -1}  \right ) + \sin \left( (f+f').2\pi\frac{i}{f_a -1}  \right ) \right ]
+\end{split}
+\end{equation}
+
+Ou seja, o sinal resultante é o sinal original (chamada portadora, com conteúdos harmônicos como a senóide de frequência $f$)
+e a reprodução de seu conteúdo espectral acima e abaixo da frequência
+original, distantes na proporção de cada harmônico ($f'$) da moduladora.
+
+No caso de usarmos a modulação na escala logarítmica como nosso tremolo, temos novamente
+um resultado menos comportado:
+\begin{equation}\label{eq:am}
+\begin{split}
+\{t_i'\}_0^{\Lambda-1} & =\{(a_{\text{máx}})^{ti'} .
+t_i\}_0^{\Lambda-1}= \left \{ a_{\text{máx}}^{\sin \left ( f'.2\pi\frac{i}{f_a -1} \right )} .
+P .\sin \left ( f.2\pi\frac{i}{f_a -1} \right ) \right \}_0^{\Lambda-1}
+\end{split}
+\end{equation}
+
+Que nos leva pelos mesmos recursos a um espectro com diversas bandas de frequência distante
+em proporções múltiplas da frequência da moduladora.
+
+Com relação à convolução, podemos estabelecer um pulso musical (a exemplo de um pulso BPM)
+e distribuir impulsos no decorrer deste pulso, de forma a estabelecer métricas e rítmos\footnote{Lembrando
+que a convolução com o impulso resulta no som deslocado ao instande de ocorrência do impulso}.
+Por exemplo, se escolhemos 2 impulsos igualmente espaçados, estaremos fazendo uma
+divisão binária básica do pulso. Se escolhemos dois sinais, um com 2 pulsos e outro com 3 pulsos,
+ambos com os impulsos igualmente espaçados na duração do pulso, resultamos na manutenção
+do pulso, com uma marcação rítmica usada tanto em divisões binárias quanto ternárias em diversos
+estilos de música étnica e tradicional. 
+Os próprios valores absolutos destes impulsos resultam em proporções entre as amplitudes dos sinais
+convoluídos.
+Exploramos este recurso da métrica
+estabelecida pela convolução com impulsos na peça musical \emph{trem de impulsos caipiras}.
+
+Já com os filtros as possibilidades explodem. Pode-se convoluir um sinal para reverberá-lo, para
+remover algum ruído, para gerar distorções ou tratamento com intuito estético mesmo. Por exemplo,
+aplicando um filtro bassa banda em que deixamos passar somente entre $1kHz$ e $3kHz$, geramos um som
+que parece de telefone ou de televisão antiga. Ao remover com alguma precisão somente
+a frequência de oscilação da rede elétrica (usualmente $50Hz$ ou $60Hz$) e harmônicas, podemos remover
+ruídos introduzidos pelos equipamentos de áudio utilizados. Um uso mais incrementado
+e propriamente musical seria realizar filtragens em bandas específicas e usar estas bandas
+pré estabelecida como um parâmetro adicional das notas.
+
+Um recurso musicalmente forte é a filtragem dependente do tempo, no qual convém os filtros IIR, em que
+os valores do filtro podem ser modificados - com algum cuidado - ao longo de sua aplicação. Cascatas
+destes filtros podem realizar filtragens complexas e mais precisas. A montagem \emph{ruídos e faixas} explora
+este recursos, realizando filtragens em ruídos diversos, com sequênciamento de disposição de bandas
+e ruídos específicos, e filtragens que variam no tempo.
+
+Um uso incrementado destes recursos todos seria a realização de um efeito chamado \emph{chorus}. Ao
+exemplo do que ocorre com um coro de cantores, neste efeito o som é realizado com diversas pequenas modificações,
+potencialmente aleatórias, em paramêtros como frequência central, presença (ou ausência) de vibrato
+e trêmolo e suas características, equalizações, volume etc. Para o resultado final, estas versões do som
+inicial é então mixado como em \ref{eq:mixagem}. A peça \emph{coro infantil} realiza chorus de formas
+diferentes em sons diferentes.
 

### 2012-10-18 devEresultados.tex
adicionado glossario de termos
feita parte de espacialização na nota digital basica

@@
 
-%\subsection{Caracterização da nota musical digital}
 \subsection{Caracterização da nota musical discretizada}
-Em diversos contextos, como na teoria musical tradicional européia,
+Em diversos contextos artísticos e teóricos, 
 a música é pensada através de 
 unidades chamadas notas e 
-estas unidades pensadas como "átomos" constituintes da música.
+estas unidades compreendidas como "átomos" constituintes da música~\cite{Wisnick, Lovelock, Webern}.
 Atualmente, estas notas
-são tidas como discretizações
+são tidas como um paradigma de proposta musical
+e, de um ponto de vista cognitivo, como discretizações
 que facilitam e enriquecem o fluxo de informação através da música
-e como um paradigma de proposta musical (\ref{Roederer, Lacerda}).
+~\cite{Roederer, Lacerda}.
 
 Em sua concepção
-simples, as notas possuem ao menos duração, volume, altura e timbre. Embora estas
-sejam características psicofísicas, podemos lidar com elas através de suas características
-fisicamente mensuráveis.
-Em primeiro lugar, estamos tratando diretamente da representação digital do som. Isso em uma
-sequência de amostras igualmente espaçadas no tempo. Definimos a
+simples e canônica, as notas possuem ao menos duração, volume, altura e timbre~\cite{Lacerda}. Embora estas
+sejam características apreciadas por leis psicofísicas, podemos lidar com elas através de suas características
+fisicamente determináveis~\cite{Roederer}.
+
+Com isso em mente, definiremos as características básicas da nota musical digital através da lida direta com a representação digital do som.
+Para nós, este consiste  em amostras sequenciais
+igualmente espaçadas no tempo. As amostras assim dispostas,
+relativas à onda mecânica do som que representam,
+em conjunto formam a chamada \emph{modulação por código de pulsos}\footnote{\emph{Pulse Code Modulation},
+de onde vem a sigla PCM, é o formato de codificação de som
+padrão em Compact Discs e outras mídias de alta fidelidade e sem compactação com perdas.}.
+
+ Definimos a
 frequência (ou taxa) de amostragem $f_a$ como o número de amostras por segundo do áudio digital e
@@
 só satisfaz esta condição em casos muito excepcionais. É necessário
-escolher um inteiro próximo de $\Delta . f_a$. Por simplicidade,
+escolher um inteiro próximo de $\Delta . f_a$ e admitir algum erro. Por simplicidade,
 escolheremos sempre a parte inteira da fração, descrita por $\lfloor \Delta . f_a \rfloor$.}:
@@
 \subsubsection{Volume}
-O volume depende da reverberação e distribuição dos harmônicos - dentre outras medidas psicofísicas~\cite{chowningVolume} - mas num primeiro momento podemos lidar com ele de forma prática e geral através da potência da onda:
+O volume depende da reverberação e distribuição dos harmônicos - dentre outras medidas psicofísicas~\cite{Chowning} - mas num primeiro momento podemos obter resultados satisfatórios e de forma prática através da potência da onda:
 
 \begin{equation}\label{potencia}
-pot(T_i^{\Delta})=\frac{\sum_{i=0}^{\Delta . f_a -1} t_i^2}{\Delta . f_a}
-\end{equation}
-
-Como o volume final dependerá da amplificação do sinal nos falantes, o mais importante é a potência relativa de uma nota em relação às outras ou de um trecho da música em relação ao resto.
-As diferenças de volume são medidas em decibeis, que são
-calculadas diretamente com as amplitudes, energias ou potências\footnote{Lembrando que, devido à nossa percepção logarítmica,
+pot(T_i)=\frac{\sum_{i=0}^{\Lambda -1} t_i^2}{\Lambda}
+\end{equation}
+
+Como o volume final dependerá da amplificação do sinal nos auto-falantes, o mais importante é a potência relativa de uma nota em relação às outras ou de um trecho da música em relação ao resto.
+As diferenças de volume são medidas em decibels, que são
+calculadas diretamente com as amplitudes através das energias ou potências\footnote{Lembrando que, devido à nossa percepção logarítmica,
 em um som de volume $v$ a redução da potência $p$ para uma mesma fração $\nu . p $ 
@@
 
-\begin{equation}\label{decibeis}
+\begin{equation}\label{decibels}
 V_{dB}=10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)}
@@
 
-A quantidade adimensional é denotada pela unidade decibeis ($dB$). A cada 10 $dB$s atribuímos canonicamente
+A quantidade adimensional $V_{dB}$ é medida pela unidade decibel ($dB$). A cada 10 $dB$ atribuímos canonicamente
 a sensação de "volume dobrado". Importantes resultados de referência são os equivalentes em $dB$s de se dobrar
@@
 Estes valores servem de guia para os aumentos e diminuições dos valores absolutos que compõem as
-sequências de amostras sonoras com propósitos musicais. A conversão direta de decibeis
+sequências de amostras sonoras com propósitos musicais. A conversão direta de decibels
 em ganho ou atenuação de amplitude se dá da seguinte forma:
@@
 
+\subsubsection{Altura}
+
 Neste ponto, correspondendo a uma partícula musical (uma nota), estabelecemos uma sequência $T_i$ cuja
 duração e intensidade conseguimos controlar\footnote{Basta alterarmos o tamanho da sequência para alterarmos
-a duração ou multiplicarmos as os elementos da sequência por algum valor para resultar em um ganho de intensidade e, portanto, de volume.}.
-
-\subsubsection{Altura}
+a duração e multiplicarmos os elementos da sequência por algum valor para resultar em um ganho de intensidade e, portanto, de volume.}.
 
 A altura é especificada pela frequência fundamental. A frequência fundamental $f_0$ especifica uma periodicidade temporal
-com ciclo de duração $\delta_{f_0} = 1/f_0$. Esta duração multiplicada pela taxa de amostragem $f_a$ resulta no número de amostras
-do ciclo $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$. É evidente que, sendo $T_i^f$ a sequência sonora de frequência fundamental $f$:
+com ciclo de duração $\delta_{f_0} = 1/f_0$. Esta duração multiplicada pela frequência de amostragem $f_a$ resulta no número de amostras
+do ciclo $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$.
+
+Por motivos didáticos, escolhamos por equanto $f$ que divida $f_a$ de forma que $\lambda_f$ resulte inteiro.
+É evidente que sendo $T_i^f$ a sequência sonora de frequência fundamental $f$:
 
@@
 
-Considerando $\frac{f_a}{f}$ inteiro. Na sessão seguinte estaremos sintetizando sons cuja frequência $f$ não divide $f_a$, por hora podemos aceitar esta restrição sem perda de generalidade.
+Na sessão seguinte estaremos sintetizando sons cuja frequência $f$ não divide $f_a$, por hora podemos aceitar esta restrição sem perda de generalidade do conteúdo.
 
@@
 de energias em componentes espectrais tanto harmônicas quanto ruidosas.  Vale salientar que nem tudo
-o que se atribui ao timbre se acha manifesto em diferenças espectrais. Uma mesma nota
-possui diferentes matizes espectrais, um mesmo instrumento possui diferentes timbres. Mesmo
-assim se fala que uma nota e um instrumento possuem um timbre. Por último, até
-aspectos culturais ou circunstanciais alteram nossa percepção timbrística.
-Na música, espectros
-sonoros com diferenças mínimas resultam em
- timbres com diferenças expressivamente cruciais.
+o que se atribui ao timbre se acha manifesto em diferenças espectrais. Mesmo
+assim se fala que uma nota e um instrumento possuem um timbre.
+Além disso, a palavra \emph{timbre} é utilizada para designar coisas diferentes: uma mesma nota
+possui diferentes timbres, um mesmo instrumento possui diferentes timbres, dois instrumentos da mesma família possuem o mesmo timbre que o caracterizam mas possuem timbres diferentes porque são instrumentos diferentes.
+Por último, até
+aspectos culturais ou circunstanciais alteram nossa percepção do timbre.
 }.
-
-O caso mais simples (e mais importante, como veremos adiante) é o do espectro que consiste somente
-de sua própria fundamental $f$. Este é o caso da senóide, frequência em movimento oscilatório puro chamado
+Musicalmente, importa manter em mente que espectros sonoros com diferenças mínimas resultam em timbres com diferenças expressivamente cruciais e que, portanto, podemos produzir timbres diferentes através de espectros diferentes~\cite{Roederer}.
+
+O caso mais simples (e mais importante, como veremos no texto que segue) é o do espectro que consiste somente
+de sua própria fundamental $f$. Este é o caso da senoide, frequência em movimento oscilatório puro chamado
 movimento harmônico simples. Seja $S_i^f$ uma sequência cujas amostras
-$s_i^f$ descrevem uma senóide de frequência $f$:
+$s_i^f$ descrevem uma senoide de frequência $f$:
 
@@
 
-Onde $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\delta_a} \;$ é o número de amostras do período\footnote{Neste ponto já se tem toda a base para música "reduced-fi", ver APÊNDICE ~\ref{APENDICE XX}}.
+Onde $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\delta_a} \;$ é o número de amostras do período\footnote{Neste ponto já se tem toda a base para música \emph{reduced-fi}.}.
 
 De forma semelhante, outras formas de onda são utilizadas na música por suas qualidades
-espectrais e simplicidade. Enquanto a senóide é um ponto isolado no espectro, estas formas
+espectrais e simplicidade. Enquanto a senoide é um ponto isolado no espectro, estas formas
 de onda apresentam cadeias de componentes harmônicas e usos específicos.
-As formas de onda especificadas em ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} são dispostas graficamente na figura ~\ref{fig:formasDeOnda}. 
-São as formas de onda artificiais tradicionalmente usadas na música para síntese e controle oscilatório de variáveis, apresentam diversos usos também fora da música(\ref{Openheim}).
-
-A dente de serra apresenta todos os componentes da série
+Além da senóide (eq.~\ref{senoide}), as formas de onda especificadas em~\ref{denteDeSerra},~\ref{triangular} e~\ref{quadrada} são dispostas graficamente na figura~\ref{fig:formasDeOnda}. 
+São as formas de onda artificiais tradicionalmente usadas na música para síntese e controle oscilatório de variáveis apresentam diversos usos também fora da música~\cite{Openheim}.
+
+A dente de serra dispõe todos os componentes da série
 harmônica com energia decrescente de $-6dB/oitava$. A sequência de amostras temporais pode ser descrita da seguinte forma:
@@
 A dente de serra é um ponto de partida comum para a síntese subtrativa pois possui
-ambos os harmônicos pares e ímpares e em grande quantidade (alguma filtragem atenuante
-é útil nos médios e agudos para que o som ganhe naturalidade e fique mais agradável).
-Os harmônicos atenuados da onda triangular
+ambos os harmônicos pares e ímpares e em grande quantidade. Para fins musicais, estas formas de onda são bastante ricas em harmônicos agudos.
+Assim, para a maior parte dos casos, alguma filtragem atenuante
+é útil nos médios e agudos para que o som ganhe naturalidade e fique mais agradável.
+Os harmônicos relativamente atenuados da onda triangular
 a faz a mais funcional - dentre as citadas - para ser usada sem nenhum tratamento para síntese de notas musicais.
@@
     \caption{Formas de onda musicais básicas}
-        \includegraphics[width=\textwidth]{figuras/formasDeOnda5}
+        \includegraphics[width=\textwidth]{figuras/formasDeOnda6}
         \label{fig:formasDeOnda}
@@
 de $100$ amostras).
-Se $t_a=44,1 kHz$, como em padrões PCM e de Compact Disks, a onda possui frequência fundamental $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz $.
+Se $t_a=44,1 kHz$, como no padrão PCM de Compact Disks, a onda possui frequência fundamental $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz $.
 Um lá\footnote{Um lá 4, logo acima do dó central, no segundo espaço do pentagrama na clave de sol comum.}, seja qual for a forma de onda dentre as artificiais apresentadas.
@@
 resultado de um período mantido fixo com exatidão.
-A senóide consiste de um nódulo único no espectro, frequência pura.
+A senoide consiste de um nódulo único no espectro, frequência pura.
 A dente de serra é a única com a série harmônica completa (pares e ímpares).
@@
     \caption{Espectros das ondas sonoras musicais artificiais básicas}
-        \includegraphics[width=\textwidth]{figuras/espectroDeOndas5b}
+        \includegraphics[width=\textwidth]{figuras/espectroDeOndas6}
         \label{fig:espectroDeOndas}
@@
 O espectro harmônico é formado pelas frequências múltiplas da frequência fundamental $f_n=(n+1).f_0$.
-Como nossa percepção segue a escala logarítmica, o espectro possui notas diferentes da frequência fundamental.
+Como nossa percepção segue uma progressão geométrica de frequências, o espectro possui notas diferentes da frequência fundamental.
 Além disso, o número de harmônicos será limitado pela frequência máxima $f_a/2$ (pelo Teorema de Nyquist).
@@
 energia\footnote{A energia total equivale à soma dos quadrados das amplitudes
-(como as de pressao/voltagem no tempo), e.g. figura ~\ref{fig:formasDeOnda}.
-As componentes espectrais - e.g. ~\ref{fig:espectroDeOndas} e ~\ref{fig:espectroOboe} -
+(como as de pressao/voltagem no tempo), e.g. figura~\ref{fig:formasDeOnda}.
+As componentes espectrais - e.g. figuras~\ref{fig:espectroDeOndas} e ~\ref{fig:espectroOboe} -
 também podem ser elevadas ao quadrado para resultar em quantificação de energia e somam-se na energia total.
@@
 implica na presença de uma oscilação senoidal na constituição do som, puramente harmônica no som e naquela frequência $f_n$. Esta energia concentrada especificamente na frequência $f_n$ é separada
- pelo ouvido para adentrar em um nível cognitivo de processamento\footnote{Esta separação em frequência é realizada por diversas espécies através de mecanismos similares à cóclea humana(\ref{musicaEsom}).}.
-  As componentes senoidais são geralmente as principais responsáveis pela qualidade chamada timbre e, caso não apresentem proporções harmônicas (relações de pequenos números), são ruídos, i.e.
-não são notas com frequência fundamental como o lá 4 ($440 Hz$) de nossos exemplos.
-Além disso, nossa noção de altura absoluta em um complexo sonoro é baseada na semelhança do espectro com a série harmônica.
+ pelo ouvido para adentrar em um nível cognitivo de processamento\footnote{Esta separação em frequência é realizada por diversas espécies através de mecanismos similares à cóclea humana~\cite{Roederer}.}.
+  As componentes senoidais são geralmente as principais responsáveis pela qualidade chamada timbre e, caso não apresentem proporções harmônicas (relações de pequenos números), o som é percebido ruidoso, i.e.
+não são notas com frequência fundamental estabelecida unívocamente como um lá 4 ($440 Hz$) de nossos exemplos.
+Além disso, nossa noção de altura absoluta em um complexo sonoro é baseada na semelhança do espectro com a série harmônica~\cite{Roederer}.
 
@@
 A onda de oboé foi amostrada de um lá 4 também em $44,1kHz$.
-O período escolhido para a amostragem é relativamente curto, com 98 amostras corresponde a $\frac{44100}{98}=450 Hz$.
-Pode-se perceber, através das curvaturas, o espectro rico em frequências do oboé e o espectro mais grave do som real.
+O período escolhido para a amostragem é relativamente curto, com 98 amostras corresponde a 
+uma frequência de $\frac{44100}{98}=450 Hz$. Pode-se perceber, através das curvaturas, o espectro rico em 
+frequências agudas do oboé e o espectro mais grave do som real.
 
@@
 \subsubsection{O espectro no som amostrado}
-Além deste papel-chave no timbre e na cognição como um todo, a presença e comportamento destas componentes senoidais é peculiar no som discretizado.
-Considere um sinal $T_i$ e sua decomposição de Fourier $\mathcal{F}\{T_i\}=C_i=\{c_i\}_0^{\Lambda-1}$.
+Além deste papel-chave na cognição (especialmente no timbre), a presença e comportamento destas componentes senoidais 
+no som discretizado possui particularidades.
+Considere um sinal $T_i$ e sua decomposição de Fourier $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$.
 Sabemos que a recomposição segue a conversão das componentes frequenciais em amostras temporais\footnote{Lembrando que o fator $\frac{1}{\Lambda}$ pode ser distribuído dentre a transformada e a reconstrução como preferir}:
@@
 \begin{equation}\label{recomposicaoFourier}
-t_i = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}c_ke^{j . \frac{2\pi k}{\Lambda} i } = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}(a_k+ j . b_k)\left[cos(w_k i) +j . sen(w_k i)\right]
+t_i = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}c_ke^{j \frac{2\pi k}{\Lambda} i } = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}(a_k+ j . b_k)\left[cos(w_k i) +j . sen(w_k i)\right]
 \end{equation}
@@
 Onde $c_k = a_k + j .
-b_k$ pondera a amplitude e fase de cada frequência: $w_k=\frac{2\pi k}{\Lambda}$ em radianos ou $f_k=w_k\frac{f_a}{2\pi}=\frac{k.f_a}{\Lambda}$ em Herz.
+b_k$ pondera a amplitude e fase de cada frequência: $w_k=\frac{2\pi k}{\Lambda}$ em radianos ou $f_k=w_k\frac{f_a}{2\pi}=\frac{k.f_a}{\Lambda}$ em Herz (com atenção os respectivos limites em $\pi$ e em $\frac{f_a}{2}$).
 Nosso som possui amostras $t_i$ reais que resultam das contribuições de cada componente de frequência $w_k=\frac{2\pi k}{\Lambda}$ e cujo $c_k$ regula o módulo e a fase.
@@
 
-Podemos notar claramente pela equação ~\ref{moduloEfase} que o termo imaginário apenas acrescenta uma fase à senóide real.
-Ao mesmo tempo $b_k$ também altera a amplitude mas a informação de amplitude já está em $a_k$,
-i.e. os termos imaginários $b_k$ da decomposição espectral por Fourier proporcionam a varredura de fase
+Podemos notar claramente pela equação ~\ref{moduloEfase} que o termo imaginário acrescenta uma fase à senoide real.
+Os termos imaginários $b_k$ da decomposição espectral por Fourier proporcionam a varredura de fase
  $[-\frac{\pi}{2},+\frac{\pi}{2}]$, basta observar o termo $tg^{-1}\left (\frac{b_k}{a_k}\right )$ que possui este contra-domínio.
-O sinal de $a_k$ especifica se estamos do lado direito ou esquerdo do circulo trigonométrico, completando a varredura completa de fase (os intervalos $[-\frac{\pi}{2},+\frac{\pi}{2}]$ e $[\frac{\pi}{2},\frac{3\pi}{2}]$ completam $2\pi$).
+O sinal de $a_k$ especifica se estamos do lado direito ou esquerdo do circulo trigonométrico, completando a varredura completa de fase (os intervalos $[-\frac{\pi}{2},+\frac{\pi}{2}]$ e $[\frac{\pi}{2},\frac{3\pi}{2}]$ se justapõem em $2\pi$).
 
@@
     \caption{Oscilação de 2 amostras (frequência máxima em qualquer $f_a$)}
-        \includegraphics[width=\textwidth]{figuras/amostras2c_}
+        \includegraphics[width=\textwidth]{figuras/amostras2c__}
         \label{fig:amostras2}
@@
 com energias $e_k=\frac{(c_k)^2}{\Lambda=2}$. O papel das amplitudes $a_k$ fica bem claro:
- $\frac{a_0}{2}$ é o deslocamento fixo\footnote{Chamado em vários contextos de bias ou offset.} e $\frac{a_1}{2}$ especifica a amplitude da oscilação em si.
-  é dada pela relação $f_k=k . \frac{f_a}{\Lambda=2}$.
+ $\frac{a_0}{2}$ é o deslocamento fixo\footnote{Chamado em vários contextos de \emph{bias} ou \emph{offset}.} e $\frac{a_1}{2}$ especifica a amplitude da oscilação em si, dada pela relação $f_k=k .
+\frac{f_a}{\Lambda=2}$.
 
@@
 os coeficiêntes $\{c_k\}_0^{\Lambda-1=2}$ da decomposição por Fourier apresentam-se em 
-3 componentes frequenciais. Uma delas é relativa à frequência zero ($c_0$), as outras duas ($c_1$ e $c_2$) contribuem de forma igual na reconstrução da senóide com $f=f_a/3$.
+3 componentes frequenciais. Uma delas é relativa à frequência zero ($c_0$), as outras duas ($c_1$ e $c_2$) contribuem de forma igual na reconstrução da senoide com $f=f_a/3$.
 
@@
 Quando $k> \frac{\Lambda}{2} $
-a frequência $f_k$ é espelhada em $\frac{f_a}{2}$ da seguinte forma $f_k=\frac{f_a}{2} - (f_k-\frac{f_a}{2})=f_a-f_k=f_a - k\frac{f_a}{\Lambda}=(\Lambda-k)\frac{f_a}{\Lambda} \;\;\;\; \Rightarrow \;\;\;\; f_k\equiv f_{\Lambda-k} \; , \forall \;\; k>\frac{\Lambda}{2}$.
+a frequência $f_k$ é espelhada em $\frac{f_a}{2}$ da seguinte forma $f_k=\frac{f_a}{2} - (f_k-\frac{f_a}{2})=f_a-f_k=f_a - k\frac{f_a}{\Lambda}=(\Lambda-k)\frac{f_a}{\Lambda} \;\;\;\; \Rightarrow \;\;\;\; f_k\equiv f_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
 
@@
    Excepcional excessão para $f_0$ e, no caso de $\Lambda$ ser par, de $f_{\Lambda/2}=f_{\text{máx}}=\frac{f_a}{2}$ , ambas as componentes são isoladas, i.e.
-não existe outra componente na frequência além dela mesma.
-Além disso, estas duas frequências (a frequência zero e a frequência máxima) não são representadas com variação de fase e, portanto, são estritamente reais.
-Assim, podemos
+não existe outra componente na frequência $f_0$ ou $f_{\Lambda/2}$ (se $\Lambda$ par) além dela mesma.
+Pois $f_{\Lambda/2}=f_{(\Lambda-\Lambda/2) = \Lambda/2}$ e $f_0=f_{(\Lambda-0)=\Lambda}=f_0$.
+
+Além disso, estas duas frequências (a frequência zero e a frequência máxima) não são representadas com variação de fase e, portanto, são estritamente reais. Assim, podemos 
    concluir que o número $\tau$ de pares de coeficientes equivalentes é:
@@
 Com $k \in \mathbb{N}$.
-A observação da equação de reconstrução para o sinal real ~\ref{moduloEfase} em conjunto com as equivalências dos módulos e fases ~\ref{equivalenciasModulos} e ~\ref{equivalenciasFases}
-mostra a combinação em fase das componentes em $w_k$ e $w_{\Lambda-k}$:
+A observação da equação de reconstrução para o sinal real ~\ref{moduloEfase} em conjunto com as equivalências dos módulos e fases ~\ref{equivalenciasModulos} e ~\ref{equivalenciasFases}, o número de coeficientes pareados \ref{coefsPareados} e equivalência de pares de frequências \ref{equivalenciasFreqs}
+expõe o caso geral da combinação em fase das componentes em cada amostra $t_i$:
 
@@
 A figura ~\ref{fig:formas4} explicita os harmônicos em 4 amostras nas formas de onda básicas das equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} e figura ~\ref{fig:formasDeOnda}.
-Todas consistem em apenas 1 senóide, com excessão da dente de serra que possui os harmônicos pares.
-
-A figura ~\ref{fig:amostras6} mostra as decomposições senoidais para o caso de 6 amostras. A figura ~\ref{fig:formas6} decompõe as formas de onda básicas para o caso de 6 amostras.
+Todas consistem em apenas 1 senoide, com excessão da dente de serra que possui os harmônicos pares.
+
+A figura ~\ref{fig:amostras6} mostra uma decomposição senoidal para o caso de 6 amostras. A figura ~\ref{fig:formas6} decompõe as formas de onda básicas para o caso de 6 amostras.
  Note que neste caso de 6 amostras as ondas se diferenciam espectralmente: as quadrada e triangular possuem as mesmas componentes, mas em proporções diferentes, já a dente de serra possui uma componente a mais.
@@
 
+\subsubsection{Localização espacial}
+Embora não seja uma das quatro qualidades básicas tradicionalmente elegidas para caracterizar uma nota musical,
+a localização espacial é também importante para um som
+pois este comumente é emitido por uma fonte localizada, como ocorre com um instrumento musical ou uma pessoa. Além disso, a espacialização é atributo bastante valorizado
+ especialmente por audiófilos e pela indústria fonográfica~\cite{floEsp}.
+Segundo a compreensão atual do fenômeno de percepção da localização espacial do som, nosso sistema nervoso utiliza
+três informações: o atraso de chegada do som entre um ouvido e o outro, a diferença de intensidade do som em cada ouvido e, por último mas também crucial, a 
+filtragem realizada pelo corpo, incluindo toráx, cabeça e orelhas~\cite{Roederer, hrtf, Heeger}.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Detecção de localização espacial de fonte sonora: DTI e DII}
+        \includegraphics[width=.5\textwidth]{figuras/espacializacao___}
+        \label{fig:spac}
+\end{figure}
+
+	Se consideradas somente as incidências diretas em cada ouvido, podemos dispor as diferenças de tempo e intensidade em equações simples. Dada a distância entre os ouvidos $\zeta\approx 21,5cm$,
+um objeto localizado em $(x,y)$ conforme a figura~\ref{fig:spac}
+está distante de cada ouvido:
+
+\begin{equation}
+\begin{split}
+d & =\sqrt{\left (x-\frac{\zeta}{2} \right )^2+y^2} \\
+d' & =\sqrt{\left (x+\frac{\zeta}{2} \right )^2 + y^2}
+\end{split}
+\end{equation}
+
+e cálculos imediatos nos levam à DTI (Diferença de Tempo Interaural)\footnote{Constata-se que $\zeta \approx 21,5cm$ para um humano adulto.}:
+
+\begin{equation}
+DTI=\frac{d'-d}{v_{som\;no\;ar}\approx 343,2 }\quad \text{segundos}
+\end{equation}
+
+e à DII (Diferença de Intensidade Interaural):
+\begin{equation}
+DII=20\log_{10}\left (\frac{d'}{d}\right) \quad dBs
+\end{equation}
+
+Convertendo para amplitude temos $DII_a=\frac{d'}{d}$. A $DDI_a$ pode
+ser utilizada para multipicação da sequência e realizado um sinal estéreo junto à sequência original.
+
+Embora simples até aqui, a localização espacial depende drasticamente de outras pistas. De forma canonica, resolvemos pelas
+DTI e DII somente o ângulo horizontal (azimutal) $\theta$ dado por:
+
+\begin{equation}
+\theta=\tan^{-1}\left ( \frac{y}{ x }  \right )
+\end{equation}
+
+Mesmo assim, há dificuldades quando $\theta$ incinde sobre o chamado "cone de confusão" em que um mesmo par de especificações DTI, DII resultam de vários dos pontos 
+do cone.
+Nestes pontos, a inferência do ângulo azimutal depende especialmente da filtragem atenuante nos agudos, pois a cabeça interfere um tanto mais nas ondas mecânicas agudas do que das graves~\cite{Heeger,hrtf}.
+
+A figura~\ref{fig:spac} mostra o cone de confusão e esta sombra do crânio, além da convenção usual da especificação do ângulo azimutal positivo para a direita.
+
+Já a localização completa, incluindo distância e elevação da fonte sonora, é dada pela função de transferência de cabeça (HRTF - do inglês \emph{Head Related Transfer Function}, bastante pessoal\ref{hrtf}.
+Exitem bases abertas e conhecidas de HRTF como \emph{CIPIC}~\cite{CIPIC} e podemos aplicar estas funções de transferência em um som por convolução~\ref{eq:conv}.
+
 \subsubsection{A nota básica}\label{notaBasica}
@@
 Escolhamos $f$ tal que $f$ divida $f_a$  
- \footnote{Esta escolha facilita as demonstrações pois os períodos resultam em quantidades inteiras de amostras.
-Transporemos esta limitação na prática com o uso de uma tabela de amostragem ou com o uso direto de funções periódicas, e.g.
-$t_i=sen(x.i)$.
-Veja próxima sessão.}.
-Uma sequência $T_i$ de amostras sonoras separadas por $\delta_a=1/f_a$ descreve uma nota musical de frequência $f$ Herz e duração $\Delta$ segundos se, e somente se, possuir a periodicidade $\lambda_f=f_a/f$ e tamanho $f_a .
-\Delta$:
-
-\begin{equation}
-T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{f_a . \Delta-1}=\{t_{i \; \% \left( \frac{f_a}{f} \right) } \}_0^{f_a . \Delta-1}
+ \footnote{Como apontado anteriormente, esta escolha facilita as demonstrações.
+Transporemos esta limitação na próxima sessão.}. 
+Uma sequência $T_i$ de amostras sonoras separadas por $\delta_a=1/f_a$ descreve uma nota musical de frequência $f$ Herz e duração $\Delta$ segundos se, e somente se, possuir a periodicidade $\lambda_f=f_a/f$
+ e tamanho $\Lambda=\lfloor f_a . \Delta \rfloor $:
+
+\begin{equation}
+T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{\Lambda-1}= \left \{t_{i \; \% \left( \frac{f_a}{f} \right) } \right \}_0^{\Lambda-1}
 \end{equation}
@@
 
-seja $f$ a frequência da nota, $\delta_f=1/f$ o seu comprimento de onda e $L_i^{f,\, \delta_f} $ a sequência que descreve um período da onda $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ (ver equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada}; $R_i^f$ é uma onda real amostrada):
+seja $f$ a frequência da nota e escolhamos $f$ tal que $f$ divida $f_a$ (transporemos esta limitação na sessão seguinte).
+Assim, o inteiro $\lambda_f=\frac{f_a}{f}$ é o número de amostras do período.
+Temos que $L_i^{f,\, \delta_f} $ é
+a sequência que descreve um período da onda $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ de duração 
+$\delta_f=1/f$ (ver equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada}; $R_i^f$ 
+é uma onda real amostrada):
 
@@
 \begin{equation}
-T_i^{f,\; \Delta}=\{t_i^f\}_0^{f_a . \Delta-1}=\{l^f_{i\,\%\left(\frac{f_a}{f}\right)}\}_0^{f_a . \Delta-1}
+T_i^{f,\; \Delta}=\{t_i^f\}_0^{\lfloor f_a . \Delta \rfloor -1}=\left \{ l^f_{i\,\%\left(\frac{f_a}{f}\right)} \right \}_0^{\Lambda-1}
 \end{equation}
@@
 
-Aqui pode muito nos auxiliar a conversão de decibeis para amplitude apresentada na equação ~\ref{ampDec}.
+Aqui pode muito nos auxiliar a conversão de decibels para amplitude apresentada na equação ~\ref{ampDec}.
 Junto à equação \ref{transAmp}, resulta na sequinte equação para uma transição de $V_{dB}$:
@@
 
-Onde $V_{dB}$ é profundidade da oscilação em decibeis do tremolo e $a_{\text{máx}}=10^{\frac{V_{dB}}{20}}$
+Onde $V_{dB}$ é profundidade da oscilação em decibels do tremolo e $a_{\text{máx}}=10^{\frac{V_{dB}}{20}}$
  é o ganho máximo de amplitude envolvido.
-A medição em decibeis é bastante pertinente pois o aumento máximo de amplitude
-não equivale à diminuição máxima relacionada, enquanto a diferença em decibeis se mantém.
+A medição em decibels é bastante pertinente pois o aumento máximo de amplitude
+não equivale à diminuição máxima relacionada, enquanto a diferença em decibels se mantém.
 
@@
     \centering
-    \caption{Tremolo de profundidade $12dB$ com padrão oscilatório de uma dente de serra $1,5Hz$ em uma senóide de $40Hz$}
+    \caption{Tremolo de profundidade $12dB$ com padrão oscilatório de uma dente de serra $1,5Hz$ em uma senoide de $40Hz$}
         \includegraphics[width=\textwidth]{figuras/tremolo}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:conv}
 t_i'=(t*h)_i=\sum_{j=0}^{min(\Lambda_h,i)}h_{j} . t_{i-j}=\sum_{j=max(i-\Lambda_h,0)}^{i}t_j . h_{i-j}
@@
 
-Para nosso caso, em resumo, podemos entender a síntese FM através do caso entre senóides
+Para nosso caso, em resumo, podemos entender a síntese FM através do caso entre senoides
 e decompor os sinais em seus espectros de Fourier (i.e. senoidais) para casos mais complexos.
@@
 
-Ou seja, o sinal resultante é o sinal original (chamada portadora, com conteúdos harmônicos como a senóide de frequência $f$)
+Ou seja, o sinal resultante é o sinal original (chamada portadora, com conteúdos harmônicos como a senoide de frequência $f$)
 e a reprodução de seu conteúdo espectral acima e abaixo da frequência
@@
 Exploramos este recurso da métrica
-estabelecida pela convolução com impulsos na peça musical \emph{trem de impulsos caipiras}.
+estabelecida pela convolução com impulsos na sequência musical \emph{trezinho de caipiras impulsivos}.
 
@@
 Para realizarmos a busca de forma eficiente, precisamos de tabelas que contenham cada uma
-uma forma de onda. As tradicionais são a senóide e as ondas quadrada, triangular e dente de serra.
+uma forma de onda. As tradicionais são a senoide e as ondas quadrada, triangular e dente de serra.
 Mostramos a seguir estas formas de onda e também uma forma de onda retirada de um som real.

### 2012-10-18 pretextual.tex
adicionado glossario de termos
feita parte de espacialização na nota digital basica

@@
 		\item[AE] \emph{AudioExperiments}
+		\item[PCM] \emph{Pulse Code Modulation (modulação por código de pulsos)}
+		\item[DTI] \emph{Diferença de Tempo Interaural (\emph{Interaural Time Difference} - ITD)}
+		\item[DII] \emph{Diferença de Intensidade Interaural (\emph{Interaural Intensity Difference)} - IID ou \emph{Interaural Level Difference} - ILD}
 	\end{listaespecial}
@@
 
+\pretextualchapter{Glossário de Termos}
+	\begin{listaespecial2}[BIGNAMEWIDTH]
+		\item[Montagem musical :] algumas ideias musicais rascunhadas, sem a pretensão de ser uma peça ou uma obra musical. Sinônimo de Sequência musical.
+		\item[Sequência musical :] sinônimo de montagem musical. 
+		\item[Tessitura :] âmbito de altura ($\log(freq)$) do objeto em discussão (instrumento, passagem musical, etc).
+	\end{listaespecial2}
+
 \sumario

### 2012-10-18 scripts/espacializacao.py

@@ -1,0 +1,50 @@
+#-*- coding: utf8 -*-
+import numpy as n, pylab as p, Image
+
+im = Image.open('../figuras/cabeca.png')
+
+height = im.size[1]
+
+# We need a float array between 0-1, rather than
+# a uint8 array between 0-255
+im2 = n.array(im).astype(n.float) / 255
+
+fig = p.figure()
+
+ax=p.subplot(111)
+
+from matplotlib._png import read_png
+import matplotlib.pyplot as plt
+from matplotlib.offsetbox import TextArea, DrawingArea, OffsetImage, AnnotationBbox
+from matplotlib.cbook import get_sample_data
+
+#fn = get_sample_data("../figuras/cabeca.png")
+arr_lena = read_png("../figuras/cabeca.png")
+
+imagebox = OffsetImage(arr_lena, zoom=0.2)
+
+xy = (0.5, 0.7)
+
+ab = AnnotationBbox(imagebox, xy,
+		xybox=(120., -80.),
+		xycoords='data',
+		boxcoords="offset points",
+		pad=0.5,
+		arrowprops=dict(arrowstyle="->",connectionstyle="angle,angleA=0,angleB=90,rad=3")
+		   )
+
+
+ax.add_artist(ab)
+
+# With newer (1.0) versions of matplotlib, you can 
+# use the "zorder" kwarg to make the image overlay
+# the plot, rather than hide behind it... (e.g. zorder=10)
+fig.figimage(im2, 100, fig.bbox.ymax - height-100, zorder=10)
+
+
+p.xlim(0,200)
+p.ylim(-100,100)
+
+p.show()
+
+

### 2012-10-18 devEresultados.tex
terminada a parte ateh a espacializacao

@@
 
-A figura ~\ref{fig:amostras6} mostra uma decomposição senoidal para o caso de 6 amostras. A figura ~\ref{fig:formas6} decompõe as formas de onda básicas para o caso de 6 amostras.
- Note que neste caso de 6 amostras as ondas se diferenciam espectralmente: as quadrada e triangular possuem as mesmas componentes, mas em proporções diferentes, já a dente de serra possui uma componente a mais.
+A figura ~\ref{fig:amostras6} mostra uma decomposição senoidal para o caso de 6 amostras e a figura ~\ref{fig:formas6} decompõe as formas de onda básicas para o caso de 6 amostras.
+ Note que neste caso de 6 amostras todas as ondas se diferenciam no espectro: as quadrada e triangular possuem as mesmas componentes, mas em proporções diferentes, já a dente de serra possui uma componente a mais.
 
@@
 pois este comumente é emitido por uma fonte localizada, como ocorre com um instrumento musical ou uma pessoa. Além disso, a espacialização é atributo bastante valorizado
- especialmente por audiófilos e pela indústria fonográfica~\cite{floEsp}.
+ por audiófilos e pela indústria fonográfica~\cite{floEsp}.
 Segundo a compreensão atual do fenômeno de percepção da localização espacial do som, nosso sistema nervoso utiliza
-três informações: o atraso de chegada do som entre um ouvido e o outro, a diferença de intensidade do som em cada ouvido e, por último mas também crucial, a 
+três informações: o atraso de chegada do som entre um ouvido e o outro, a diferença de intensidade do som direto em cada ouvido e a 
 filtragem realizada pelo corpo, incluindo toráx, cabeça e orelhas~\cite{Roederer, hrtf, Heeger}.
@@
 
-Convertendo para amplitude temos $DII_a=\frac{d'}{d}$. A $DDI_a$ pode
-ser utilizada para multipicação da sequência e realizado um sinal estéreo junto à sequência original.
-
-Embora simples até aqui, a localização espacial depende drasticamente de outras pistas. De forma canonica, resolvemos pelas
+Convertendo para amplitude temos $DII_a=\frac{d'}{d}$. A $DII_a$ pode
+ser utilizada como constante multiplicativa do canal direito de um sinal sonoro estéreo ($\{t_i\}_0^{\Lambda -1}=\{DII_a .
+t_i'\}_0^{\Lambda -1}$).
+Pode-se utilizar a DII junto à DTI como adiantamento no tempo do canal direito com relação ao esquerdo, vínculo crucial para a percepção de localidade em sons graves e em sonoridades percussivas~\cite{Heeger}.
+Considerando $\Lambda_{DTI}=\lfloor DTI . f_a \rfloor $, podemos escrever:
+
+\begin{equation}
+\begin{split}
+\Lambda_{DTI} & = \left \lfloor \frac{d'-d}{343,2}  f_a \right \rfloor \\
+DII_a & = \frac{d'}{d} \\
+\{t_{(i+\Lambda_{DTI})}\}_{\Lambda_{DTI}}^{\Lambda+\Lambda_{DTI}-1} & =\{DII_a . t_i'\}_0^{\Lambda-1} \\
+\{t_i\}_0^{\Lambda_{DTI}-1} & = 0
+\end{split}
+\end{equation}
+
+Com $t_i$ o canal direito e $t_i'$ o canal esquerdo. Caso $\Lambda_{DTI} < 0 $, basta trocar $t_i$ por $t_i'$  e utilizar $\Lambda_{DTI}'= | \Lambda_{DTI} | $.
+
+Embora consideravelmente simples até aqui, a localização espacial depende drasticamente de outras pistas. De forma canonica, resolvemos pelas
 DTI e DII somente o ângulo horizontal (azimutal) $\theta$ dado por:
@@
 
-A figura~\ref{fig:spac} mostra o cone de confusão e esta sombra do crânio, além da convenção usual da especificação do ângulo azimutal positivo para a direita.
-
-Já a localização completa, incluindo distância e elevação da fonte sonora, é dada pela função de transferência de cabeça (HRTF - do inglês \emph{Head Related Transfer Function}, bastante pessoal\ref{hrtf}.
-Exitem bases abertas e conhecidas de HRTF como \emph{CIPIC}~\cite{CIPIC} e podemos aplicar estas funções de transferência em um som por convolução~\ref{eq:conv}.
+A figura~\ref{fig:spac} mostra esta sombra acústica do crânio, importante para a percepção do ângulo azimutal da fonte no cone de confusão.
+O cone em si não foi disposto na figura pois não é exatamente um cone e suas dimensões precisas não são especificadas na literatura~\cite{confCone}, mas pode ser entendido como um cone com o ápice no meio da cabeca e saindo por cada uma das orelhas.
+
+Já a localização completa, incluindo distância e elevação da fonte sonora, é dada pela função de transferência de cabeça (HRTF - do inglês \emph{Head Related Transfer Function}~\cite{hrtf}.
+Exitem bases abertas e conhecidas de HRTF como \emph{CIPIC}~\cite{CIPIC} e podemos aplicar estas funções de transferência em um som por convolução (veja equação~\ref{eq:conv}).
+Estas funções de transferência mudam de acordo com o corpo do indivíduo e existem diversas técnicas para resultar em funções de transferência que sejam utilizáveis de forma mais universal~\cite{lazaSPA}.
 

### 2012-10-19 devEresultados.tex
adentrada variacoes na nota musical

@@
 do cone.
-Nestes pontos, a inferência do ângulo azimutal depende especialmente da filtragem atenuante nos agudos, pois a cabeça interfere um tanto mais nas ondas mecânicas agudas do que das graves~\cite{Heeger,hrtf}.
+Nestes pontos, a inferência do ângulo azimutal depende especialmente da filtragem atenuante nos agudos, pois a cabeça interfere um tanto mais nas ondas mecânicas agudas do que nas graves~\cite{Heeger,hrtf}.
 
@@
 Pode-se completar com zeros para somar sequências de tamanhos diferentes.
-O resultado da mixagem de notas musicais é a harmonia musical, cujos intervalos entre as frequências e os acordes de notas simultâneas regem aspectos subjetivos e abstratos da música e sua apreciação.
+O resultado da mixagem de notas musicais é a harmonia musical, cujos intervalos entre as frequências e os acordes de notas simultâneas regem aspectos subjetivos e abstratos da música e sua apreciação~\cite{Harmonia}.
 
@@
 Estas variações não necessariamente são levadas
-em conta na síntese sonora, mas constituem-se em guias para a
-síntese de sons para usos musicais pois é como eles
+em conta na síntese sonora, mas constituem-se em guias de 
+síntese sonora para usos musicais pois é como os sons
 se apresentam na natureza\footnote{A regra de ouro

### 2012-10-20 devEresultados.tex
feitas figuras para mixagem e concatenacao e arrumadas outras coisinhas

@@
 
+\begin{figure}[h!]
+    \centering
+    \caption{Mixagem de três sequências sonoras}
+        \includegraphics[width=\textwidth]{figuras/mixagem}
+        \label{fig:mixagem}
+\end{figure}
+
+Ilustramos na figura~\ref{fig:mixagem} este processo de superposição de ondas sonoras discretizadas\footnote{A figura dispõe 100 amostras, de onde podemos concluir que, se $f_a=44,1kHz$, as frequências da dente de serra, da onda quadrada e da senoide são,
+respectivamente, $\frac{f_a}{50}=882Hz$, $\frac{f_a}{25}=1764Hz$ e $\frac{f_a}{20}=2205Hz$. A duração do trecho é bastante curto $\frac{f_a=44,1kHz}{100} \approx 2 \text{ milisegundos}$.}.
+ As notas mixadas são em grande parte separadas pelo ouvido por leis físicas de ressonância e pelo sistema nervoso~\cite{Roederer}.
+
 Pode-se completar com zeros para somar sequências de tamanhos diferentes.
@@
 
+Este mecanismo é demonstrado de forma ilustrativa na figura~\ref{fig:concatenacao} com as mesmas sequências utilizadas na figura ~\ref{fig:mixagem}.
+ As sequências são curtas para as taxas de amostragem usuais (cerca de $7$ milisegundos ao total se $f_a=44,1kHz$), mas pode-se observar de forma explícita a concatenação de sequências sonoras e o exemplo é real (cada nota tem a duração maior que $100ms$ se $f_a<1kHz$).
+
+\begin{figure}[h!]
+    \centering
+    \caption{Concatenação de três sequências sonoras}
+        \includegraphics[width=\textwidth]{figuras/concatenacao}
+        \label{fig:concatenacao}
+\end{figure}
+
 A montagem musical \emph{reduced-fi} demonstra de forma isolada este uso de justaposição temporal das notas, resultando em uma peça homofônica.
 O princípio vertical está demonstrado nos \emph{quadros sonoros}, sons estáticos com espectros peculiares.
-Ambas as peças estão em código Python no APÊNDICE XX e podem ser escutadas e vizualizadas nos links \ref{reducedLink} e \ref{quadrosLink}.
+Ambas as peças estão em código Python no APÊNDICE~\ref{cap:codigoPecas} e podem ser escutadas e vizualizadas nos links~\cite{reduced-fi} e~\cite{quadros}.
 

### 2012-10-20 scripts/mixa_concatena.py

@@ -1,0 +1,40 @@
+#-*- coding: utf8 -*-
+import numpy as n, pylab as p
+
+c=2
+
+# dois períodos de uma dente de serra:
+d=n.linspace(-1,1,50)
+d=n.hstack((d,d))
+d=n.hstack((d,d))
+d=d[:len(d)/c]
+
+# quatro periodos de uma onda quadrada
+q=n.hstack((  n.ones(13)*-1,n.ones(12)  ))
+q=n.hstack((q,q))
+q=n.hstack((q,q))
+q=n.hstack((q,q))
+q=q[:len(q)/c]
+
+# senoide ou ruido
+s=n.linspace(0,10*2*n.pi,200,endpoint=False)
+s=n.sin(s)[:len(s)/c]
+e=5.0
+ee=4
+p.plot(s+ee+3*e,'o')
+p.plot(q+ee+2*e,'o')
+p.plot(d+ee+e,'o')
+p.plot(s+q+d,'o')
+
+p.ylim(-3.5,21)
+p.xlim(-5,105)
+
+p.text(47,11.4-.5,"+",fontsize=48)
+p.text(47,11.4+e-.5,"+",fontsize=48)
+p.text(46,11.4-e-2,"=",fontsize=78)
+
+p.xticks((),())
+p.yticks((),())
+
+
+p.show()

### 2012-10-26 devEresultados.tex
modificacoes menores

@@
 duração, altura, intensidade (volume) e timbre. Esta é uma modelagem
-útil e paradigmática, mas de forma alguma esgota o que entendemos por
+útil e paradigmática, mas não esgota o que entendemos por
 uma nota musical.
@@
 da própria nota, i.e. em uma nota real os parâmetros
-não são absolutamente fixos. Por exemplo, em uma nota de piano
+não são absolutamente fixos~\cite{Chowning}. Por exemplo, em uma nota de piano
 de 3 segundos, a intensidade tem início abrupto e decaimento progressivo
@@
 decaem antes dos outros e alguns que aparecem com o tempo.
-Estas variações não necessariamente são levadas
-em conta na síntese sonora, mas constituem-se em guias de 
+Estas variações não são obrigatórias e sim orientações da
 síntese sonora para usos musicais pois é como os sons
@@
 aqui é: para que um som isolado disperte interesse
-por si só, faça como que tenha variações internas.}.
+por si só, faça como que tenha variações internas~\cite{Roederer}.}.
 
@@
 e a complexidade da nossa cognição sonora. Desta forma, apontaremos
-recursos primários para estas variações dos parâmetros na nossa nota
+recursos primários para estas variações das características na nossa nota
 básica.
@@
 
-Mais conhecido pelo termo em inglês, a \emph{Lookup table} (ou simplesmente
+Mais conhecido pelo termo em inglês, a \emph{Lookup Table} (ou simplesmente
 LUT), é um procedimento simples que consiste em criar uma tabela
@@
 período de onda possa ser usado para sintetizar sons em toda a banda
-de frequências audíveis.
+de frequências audíveis, qualquer que seja a forma de onda amostrada.
 
@@
 período de onda qualquer (veja equação ~\ref{periodoUnico}). Uma sequência
-$T_i^{f,\Delta}$ com amostras de um som de frequência $f$ e duração $\Delta$
+$T_i^{f,\,\Delta}$ com amostras de um som de frequência $f$ e duração $\Delta$
 pode ser feita a partir de $\widetilde{L_i}$ da seguinte forma:
@@
 \begin{equation}
-T_i^{f,\Delta}=\{t_i^f\}_0^{f_a .
-\Delta-1} = \{ \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} } \}_{0}^{f_a .
-\Delta-1}\; , \quad \text{onde} \;\; \gamma_i = \left \lfloor i .
+T_i^{f,\,\Delta}=\{t_i^f\}_0^{\lfloor f_a .
+\Delta \rfloor -1} = \{ \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} } \}_{0}^{\Lambda-1}\; , \quad \text{onde} \;\; \gamma_i = \left \lfloor i .
 f \frac{ \widetilde{\Lambda}}{f_a} \right \rfloor
@@
  o padrão é usar $\widetilde{\Lambda} = 1024$ amostras, pois já não gera ruído
- relevante no espectro audível.
+ relevante no espectro audível~\cite{Geiger}.
 
@@
 graves quando o tamanho da tabela $\widetilde{\Lambda}$ não é suficientemente grande para a taxa de amostragem
-$f_a$\footnote{Como citado acima, $\widetilde{\Lambda} = 1024$ é suficiente para sintetizar
-sequências $T_i^f$ com qualquer $f$ no espectro audível.}. O procedimento de busca em tabela
+$f_a$. O procedimento de busca em tabela
 é computacionalmente bastante barato, substituindo cálculos por buscas simples (por isso geralmente
@@
 simplifica diversas rotinas e é computacionalmente coerente.
+Para um uso clássico das LUTs na síntese sonora musical, recomendamos procurar pela chamada \emph{Wavetable Synthesis} que consiste em várias LUTs utilizadas em conjunto através da mixagem para gerar uma nota musical quasi-periódica~\cite{Cook,Wavetable}.
 
@@
 Segundo revela a lei de Weber e Fechner, nossa percepção tem uma relação logarítmica com
-o estímulo que a causa. Mesmo assim, dado o uso, explicitaremos também a variação
+o estímulo que a causa~\cite{Weber-Fechner}. Mesmo assim, dado o uso, explicitaremos também a variação
 linear e começaremos descrevendo esta por razões didáticas.
@@
 
+Onde $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a}$ é o incremento da LUT entre duas amostras dada a frequência do som na primeira amostra.
+
 Desta forma, podemos calcular os elementos $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$
-com base no período $\{\widetilde{l}_i\}_0^{\Lambda-1}$. Note que na equação
-\ref{freqLinear}, o termo $f_{\Lambda -1} - f_0$ é o âmbito do contra-domínio de $f_i$
-e que $\Lambda -1$ é âmbito do domínio. Como $i$ varre o domínio $[0,\Lambda-1]$ e
-temos o deslocamento $f_0$ na imagem, $f_i$ varre o contra-domínio 
-na medida em que $i$ varre o domínio\footnote{Estas digressões são importantes
-para entendermos o significado musical do comportamento destas sequências de amostras.}.
+com base no período $\{\widetilde{l}_i\}_0^{\Lambda-1}$.
 
@@
 dito ainda de outra forma, uma progressão geométrica da frequência
-é percebido como uma progressão aritmética de alturas}.
+é percebido como uma progressão aritmética de alturas.}.
 

### 2012-10-27 devEresultados.tex
feito tudo ate tremolo e vibrato que precisam ser migrados

@@
 \begin{equation}\label{serieAmostralLin}
-\{t_i^{\;\overline{f_0,f_{\Lambda-1}}}\}_0^{\Lambda-1}=\{\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\}_0^{\Lambda-1}
+\{t_i^{\;\overline{f_0,\, f_{\Lambda-1}}}\}_0^{\Lambda-1}=\{\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\}_0^{\Lambda-1}
 \end{equation}
@@
 \begin{equation}\label{serieAmostralLog}
-\{t_i^{\;\overline{f_0,f_{\Lambda-1}}}\}_0^{\Lambda-1}=\{\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\}_0^{\Lambda-1}
+\{t_i^{\;\overline{f_0,\,f_{\Lambda-1}}}\}_0^{\Lambda-1}=\{\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\}_0^{\Lambda-1}
 \end{equation}
@@
 Onde usamos a equação \ref{indiceExponencial} para calcular a sequência $\{\gamma_i\}_0^{\Lambda-1}$ de índices para para
-serem buscados na nossa tabela que contém um único período da forma de onda.
+serem buscados na nossa tabela que contém um único período da forma de onda representado em uma sequência de $\widetilde{\Lambda}$
+pontos igualmente espaçados (amostras).
 Isso é explicitado na equação ~\ref{serieAmostralLog}, onde estabelecemos a relação
@@
 Aqui pode muito nos auxiliar a conversão de decibels para amplitude apresentada na equação ~\ref{ampDec}.
-Junto à equação \ref{transAmp}, resulta na sequinte equação para uma transição de $V_{dB}$:
+Junto à equação \ref{transAmp}, resulta na sequinte equação para uma transição de $V_{dB}$ decibels:
 

### 2012-10-28 codigoPecas.tex
lipada geral e arrumando AM e FM

\chapter{Código Computacional das Peças Musicais}
\label{cap:codigoPecas}

\subsection{Reduced-fi}

\subsection{Quadros Sonoros}

### 2012-10-28 devEresultados.tex
lipada geral e arrumando AM e FM

@@
 
-\begin{equation}\label{ex:mixagem}
+\begin{equation}\label{eq:mixagem}
 \{t_i\}=\left \{ \sum_{k=0}^{N-1}t_{k,i} \right \}
@@
 
-\subsubsection{Tremolo e Vibrato}
+Veremos adiante a aplicação das transições lineares para as oscilações 
+de amplitude e frequência nas chamadas sínteses AM e FM e a aplicação das transições
+logarítmicas para os trêmolos e vibratos.
+
+\subsubsection{Aplicação de Filtros Digitais}
+ As possibilidades
+aqui são variadas e a complexidade facilmente
+foge ao escopo deste trabalho\footnote{A elaboração de filtros
+constitui uma área reconhecidamente complexa, com literatura
+e pacotes de software dedicados. 
+Recomendamos ao leitor
+interessado uma visita à nossa bibliografia, em especial~\cite{Openheim,smith}.},
+de forma que focamos em explicitar o que ocorre com as sequências
+de amostras e em descrever os procedimentos usuais.
+A aplicação de filtros pode
+ser parte constituinte da síntese ou feita posteriormente
+como parte dos processos tipicamente chamados de tratamento sonoro.
+
+Explicitaremos a aplicação de filtros por convolução
+e por equação a diferenças.
+
+\begin{itemize}
+
+\item  Convolução e filtros de resposta ao impulso finita (FIR)
+
+Os filtros aplicados por convolução são conhecidos
+pela sua sigla FIR (do inglês Finite Impulse Response)
+e são caracterizados por possuirem uma representação amostral
+finita no tempo. Esta representação amostral é chamada
+de 'resposta ao impulso' $h_i$. Os filtros FIR são aplicados ao som
+digitalizado pela convolucão do som com sua 
+resposta ao impulso. Para nossos fins, esta
+convolução fica definida como\footnote{Convencionamos $t_i=0$ para as
+amostras não definidas de antemão.}:
+
+\begin{equation}\label{eq:conv}
+\begin{split}
+\{t_i'\}_0^{\Lambda_t+\Lambda_h-2\; = \;\Lambda_{t\, '}-1} =\{(t*h)_i\}_0^{\Lambda_{t \, '}-1} & =\left \{ \sum_{j=0}^{min(\Lambda_h,i)}h_{j} . t_{i-j} \right \}_0^{\Lambda_{t\, '}-1} \\
+    & =\left \{ \sum_{j=max(i-\Lambda_h,0)}^{i}t_j . h_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
+\end{split}
+\end{equation}
+
+Ou seja, o som $\{t_i'\}$ resultante da convolução de $\{t_i\}$ com a resposta ao impulso $\{h_i\}$
+tem cada i-ésima amostra $t_i$ substituída pela soma de suas últimas $\Lambda_h$ amostras $\{t_{i-j}\}_{j=0}^{\Lambda_h-1}$
+multiplicadas pela resposta ao impulso $\{h_i\}_0^{\Lambda_h-1}$ em si. Isso é facilmente
+percebido pela visualização gráfica da convolução de sequências arbitrárias e dispomos
+para este fim a figura \ref{fig:conv}. Observe que a resposta ao impulso $\{h_i\}$
+é disposta na forma retrograda e que mostramos na figura os cálculos explícitos
+das amostras $t_{12}'$ e $t_{32}'$ do sinal convoluído. O sinal resultante possui
+sempre o tamanho $\Lambda_t+\Lambda_h -1=\Lambda_{t'}$.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Interpretação gráfica da convolução}
+        \includegraphics[width=\textwidth]{figuras/convolucao______}
+        \label{fig:conv}
+\end{figure}
+
+Com este procedimento podemos aplicar reverberadores, equalizadores, delays
+e vários outros tipos de filtros para fins de tratamento sonoro ou
+efeitos musicais/artísticos.
+
+Para a obtenção da resposta ao impulso, pode-se recorrer a medições
+físicas ou à síntese direta do filtro. Para citar um exemplo
+de medições diretas, uma resposta ao impulso para a aplicação
+de reverberação pode resultar da gravação acústica do ambiente ao disparar
+um estalo (cuja gravação se aproxima da resposta ao impulso diretamente) ou ao disparar uma
+varredura de frequência (cuja gravação se aproxima da resposta em frequência).
+Ambas resultam em respostas ao impulso
+que, convoluídas com a sequência sonora, resultam na própria sequência
+com uma reverberação que se assemelha à reverberação do ambiente 
+em que ocorreu a medição.
+
+Como exemplo de resposta ao impulso sintetizadas, pode-se
+descrever um perfil espectral, realizar a transformada inversa
+de fourier e convoluir o resultante com o som, realizando
+uma filtragem cujo perfil é próximo ao especificado inicialmente.
+Note que este procedimento funciona particularmente bem se for grande o número de amostras utilizadas para especificar o perfil espectral do filtro\footnote{O que resulta em mais processamento computacional.}.
+
+Outro exemplo simples e útil vem do deslocamento temporal causado pela convolução com o impulso deslocado. Assim, podemos aplicar de linhas de \emph{delays} através
+da convolução do som com uma resposta ao impulso que possui um impulso
+em cada reincidência do som na linha de delays intensionada.
+Para ilustrar essa aplicação, dispomos a figura~\ref{fig:delays}
+em que pode-se observar o deslocamento causado pela convolução
+com o impulso. Dependendo da densidade dos impulsos, o resultado
+é de caráter rítmico (20 impulsos por segundo ou menos) ou de amálgama
+sonoro (20-40 impulsos por segundo ou mais). Neste último caso,
+ocorrem processos tipicamente vinculados à síntese granular e às
+equalizações.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Convolução com o impulso e linhas de delays}
+        \includegraphics[width=\textwidth]{figuras/delays__}
+        \label{fig:delays}
+\end{figure}
+
+\item Filtros de resposta ao impulso infinita (IIR)
+
+Esta classe de filtros é
+conhecida pela sigla IIR (do inglês Infinite Impulse Response)
+e é caracterizada por possuir uma representação temporal
+infinita, i.e. a resposta ao impulso não converge para zero. Sua aplicação é usualmente feita pela equação
+a diferenças:
+
+\begin{equation}
+t_i' = \frac{1}{b_0}\left ( \sum_{j=0}^Ja_j . t_{i-j} - \sum_{k=1}^Kb_k . t_{i-k}' \right )
+\end{equation}
+
+com $b_0=1$ na grande maioria dos casos pois basta normalizarmos as variáveis
+resultando em filtro idêntico: $a_j'=\frac{a_j}{b_0}$ e $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
+
+Apontamos filtros IIR bastante simples, úteis e usuais abaixo. Existem
+diversos métodos e ferrametas para a elaboração de filtros IIR
+e esta é uma seleção com fins didáticos e para consulta futura por
+utilidade.
+São filtros de primeira e segunda ordem bem comportados cujas
+filtragens realizadas dispomos na figura~\ref{fig:iir}.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda}}
+        \includegraphics[width=\textwidth]{figuras/iir___}
+        \label{fig:iir}
+\end{figure}
+
+No caso dos filtros de ordem simples, a frequência de corte $f_c$ é onde 
+o filtro realiza uma atenuação de $-3dB \approx 0.707 $ da amplitude original.
+No caso dos filtros passa e rejeita banda, esta mesma atenuação é
+resultado de duas especificações: $f_c$ (neste caso melhor compreendida como 'frequência central') e a largura de banda $bw$
+pois em ambas as frequências $f_c \pm bw$ há uma atenuação de $\approx 0.707$ da amplitude original.
+
+Observe que existe amplificação do som no caso dos filtros passa e rejeita banda quando a frequência
+de corte é baixa e a largura de banda é grande o suficiente. Nos agudos, estes filtros apresentam
+somente um desvio do perfil esperado, expandindo a envoltória um pouco mais do lado grave da banda em
+evidência.
+
+Para conseguir filtros cujas respostas em frequência possuem outras envoltórias (para o módulo),
+o recurso mais simples a esta altura
+é fazer cascata destes filtros aplicando-os sucessivamente.
+
+Outra possibilidade é usar alguma receita de filtro
+biquad\footnote{Abreviação
+de 'biquadrado' pois sua função de transferência possue dois polos e dois zeros, i.e. sua
+forma normal consiste em dois polinômios quadráticos formando uma fração
+$\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1+ b_1.z^{-1} +b_2 . z^{-2}}$.}
+ou rotinas para cálculo de coeficientes
+de filtros Chebichev\footnote{Filtros Butterworth e Elípticos podem
+ser considerados como casos específicos dos Filtros do tipo Chebichev~\cite{Openheim,smith}.}.
+Ambas as possibilidades são exploradas
+por títulos em nossas referências, em especial recomendamos~\cite{JOSFM,smith} e na coleção de filtros da comunidade \emph{Music-DSP}, da Universiade de Columbia~\cite{music-dsp}.
+Para uma exposição aprofundada do assunto, recomendamos~\cite{Openheim}.
+
+\end{itemize}
+
+\begin{enumerate}
+\item Passa-baixas de polo simples com gráfico do canto superior esquerdo da figura~\ref{fig:iir}. A fórmula geral tem
+por referência da frequência de corte $f_c \in (0,\frac{1}{2})$,
+fração da frequência de amostragem $f_a$
+em que há aproximadamente uma atenuação de $3dB$.
+Calculamos os coeficientes do filtro IIR
+$a_0$ e $b_1$ 
+através da variável intermediária: $x \in [0,1]$:
+
+\begin{equation}\label{eq:passa-baixas}
+\begin{split}
+x & =e^{-2\pi f_c} \\
+a_0 & =  1-x \\
+b_1 & =  x
+\end{split}
+\end{equation}
+
+\item Passa-altas de polo simples com gráfico do canto superior direito da figura~\ref{fig:iir}. A fórmula geral,
+com frequência de corte $f_c \in (0,\frac{1}{2})$, é calculada através da variável
+intermediária $x \in [0,1]$:
+
+\begin{equation}\label{eq:passa-altas}
+\begin{split}
+x & =e^{-2\pi f_c} \\
+a_0 & =  \frac{x+1}{2} \\
+a_1 & =  -\frac{x+1}{2} \\
+b_1 & =  x
+\end{split}
+\end{equation}
+
+%\item Passa-banda
+%\item Rejeita-banda
+\item Nó (\emph{notch filter}). Este filtro é parametrizado
+pela frequência central\footnote{Não
+confundir com a frequência de corte também notada aqui por $f_c$ nos filtros passa baixas e passa altas.} $f_c$
+e a largura de banda $bw$
+($f_c \pm bw$ resulta em 0.707 da amplitude, i.e. atenuação de $3dB$),
+ambos dados como fracoes de $f_a$, portanto $f,\; bw \in (0,0.5)$.
+Por conveniência definimos as variáveis auxiliares $K$ e $R$ como:
+
+\begin{equation}
+\begin{split}
+R & = 1 - 3BW \\
+K & = \frac{1-2R\cos(2\pi f_c) + R^2}{2 - 2 \cos (2 \pi f_c)}
+\end{split}
+\end{equation}
+
+A partir das quais temos dois filtros. O filtro passa banda que dispomos no canto inferior esquerdo da figura~\ref{fig:iir}:
+
+\begin{equation}\label{eq:passa-banda}
+\begin{split}
+a_0 & =  1 - K \\
+a_1 & =  2(K-R)\cos (2\pi f) \\
+a_2 & =  R^2-K \\
+b_1 & =  2R \cos (2\pi f) \\
+b_2 & =  -R^2
+\end{split}
+\end{equation}
+
+e o filtro rejeita banda:
+
+\begin{equation}\label{eq:rejeita-banda}
+\begin{split}
+a_0 & =  K \\
+a_1 & =  -2K\cos (2\pi f) \\
+a_2 & =  K \\
+b_1 & =  2R \cos (2\pi f) \\
+b_2 & =  -R^2
+\end{split}
+\end{equation}
+
+disposto na parte inferior esquerda da figura~\ref{fig:iir}.
+
+%\item Biquad: pela especificação de uma frequência central, da qualidade
+%e da intensidade do filtro, este filtro é simples e usual para áudio,
+%permitindo ajustes mais finos. Diversas receitas podem ser encontradas
+%na literatura, recomendamos especialmente as diferentes especificações
+%em ~\ref{musicDSP} e ~\ref{dspguide}.
+
+\end{enumerate}
+
+\subsubsection{Ruídos}
+De forma geral, os sons sem altura definida 
+são chamados ruídos~\cite{Lacerda}.
+Estes sons estão presentes como constituintes importantes dos sons musicais de altura definida,
+como os ruídos presentes nas notas do piano, do violino, etc. Além disso, os instrumentos
+de percussão, em grande parte, não possuem altura definida e seus sons
+são muitas vezes compreendidos como ruídos~\cite{Roederer}. Na música eletrônica (incluindo
+a eletroacústica), os ruídos possuem usos amplos, diversificados e comumente
+idiomáticos~\cite{Cook}.
+
+A ausência de uma altura definida é fruto da ausência de uma organização harmônica perceptível nas componentes senoidais que formam o som. Assim, pode-se inferir
+as incontáveis possibilidades de gerar ruídos. A mais imediata é a utilização
+de valores aleatórios para a geração da sequência sonora $T_i$. Este
+é um método atraente, visto a disponibilidade
+das consagradas distribuições de probabilidade em diversas linguagens de programação,
+mas os resultados não são tão úteis, tendendo geralmente ao ruído branco
+e com difíceis previsões do espectro resultante~\cite{Cook}.
+
+Outra possibilidade é a geração de ruído através do espectro desejado. A partir
+dele, executamos a transformada inversa de Fourier. Neste caso, é importante
+que realizemos a distribuição espectral com cuidado pois queremos um perfil
+do módulo do espectro e a fase é aleatória\footnote{Caso utilizemos a mesma fase
+(ou fases com forte correlação) para
+as componentes envolvidas, o som sintetizado possuirá energia bastante concentrada
+em alguns trechos.}. Veja a figura~\ref{fig:ruidos} e o código Python que a gerou para uma implementação.
+
+\begin{figure}[htpq!]
+    \centering
+    \caption{Ruídos coloridos realizados através das equações~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul}, \ref{eq:violeta}: espectros e ondas sonoras resultantes}
+        \includegraphics[width=\textwidth]{figuras/ruidos___}
+        \label{fig:ruidos}
+\end{figure}
+
+Abaixo elencamos alguns ruídos de espectros estáticos. São chamados \emph{coloridos} por terem sido associados a cores.
+
+\begin{itemize}
+
+\item O ruído branco deve seu nome por possuir energia distribuida
+igualmente por todas as frequências. Podemos obter a realização
+do ruído branco com a transformada inversa dos seguintes coeficientes:
+
+\begin{equation}\label{eq:branco}
+\begin{split}
+c_0 & =0 \quad \text{pois não queremos bias} \\
+c_i & =e^{j.x}\;,\;\; j^2=-1 \;, \;\; x \; \text{randômico} \; \in \; [0,2\pi]\;,\;\; i \; \in \; \left[1, \, \frac{\Lambda}{2}-1\right] \\
+c_{\Lambda/2} & = 1 \quad\quad \text{(se $\Lambda$ par)}\\ 
+c_i & = c_{\Lambda - i}^*\;,\;\; \text{para}\;  i \; > \;  \frac{\Lambda}{2}
+\end{split}
+\end{equation}
+
+O valor de $c_i$ calculado pela exponencial é apenas um artifício para resultar em módulo unitário e fase aleatória.
+Já $c_{\Lambda/2}$ é sempre puramente real (como vimos na sessão anterior).
+
+\item O ruído rosa possui uma queda de $3dB$ por oitava. Este ruído é muito usual no teste de equipamentos e montagens de aparelhos além de presença destacada na natureza~\cite{Roederer}.
+
+\begin{equation}\label{eq:rosa}
+\begin{split}
+f_{\text{min}} & \approx 15 Hz \\
+f_i & = i \frac{f_a}{\Lambda} \;, \;\; \quad i \;\leq\; \frac{\Lambda}{2},\;\; i\;\in\;\mathbb{N}  \\
+\alpha_i & = (10^{-\frac{3}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )}  \\
+c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \\
+c_i & =e^{j.x} . \alpha_i\;,\;\; j^2=-1 \;, \;\;\  x \;\; \text{randômico} \; \in \; [0,2\pi]\;,\;\; \forall \; i \; : f_{\text{min}} \le f_i < f_{\lceil \Lambda/2-1 \rceil}  \\
+c_{\Lambda/2} & = \alpha_{\Lambda/2} \quad\quad \text{(se $\Lambda$ par)}\\ 
+c_i & = c_{\Lambda - i}^*\;,\;\; \text{para}\;  i \; > \;  \Lambda/2
+\end{split}
+\end{equation}
+
+A frequência mínima $f_{\text{min}}$ pode ser escolhida com base no limite da audição, pois não escutamos como altura de um som uma frequência absoluta abaixo de $\approx\; 20Hz$.
+
+O resto dos ruídos podem ser feitos com base neste procedimento descrito para 
+o ruído rosa. Basta que modifiquemos alguns detalhes. Em especial a equação que define $\alpha_i$.
+
+\item O ruído marrom deve seu nome ao Robert Brown, que descreveu o movimento browniano.
+Embora esta origem seja um tanto díspar do que poderiamos considerar motivo para uma associaçãocom a cor marrom, o ruído sonoro ficou consagrado com este nome.
+De qualquer forma, é bastante comum declarar satisfatória a associação do ruído com a cor marrom, dados os ruídos branco e rosa mais estridentes e relacionados a cores mais intensas~\cite{marrom}.
+
+O que caracteriza este ruído é a queda de $6dB$ por oitava. Desta forma, basta modificarmos a equação de $\alpha_i$ 
+no conjunto \ref{eq:rosa} para:
+
+\begin{equation}\label{eq:marrom}
+\alpha_i=(10^{-\frac{6}{20}})^{\log _2 \left( \frac{f_i}{f_{\text{min}}} \right )}
+\end{equation}
+
+\item Ruído azul é aquele em que há um ganho de $3dB$ por oitava em uma banda limitada. Para
+a síntese deste ruído, basta estipularmos a frequência mínima $f_{\text{min}}$ e a frequência
+máxima $f_{\text{máx}}$. Assim, também com base no conjunto de equações \ref{eq:rosa}, 
+basta atentarmos para as sequintes identidades:
+
+\begin{equation}\label{eq:azul}
+\begin{split}
+\alpha_i & = (10^{\frac{3}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \\
+c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \;\; \text{ou} \;\; f_i>f_{\text{máx}} \\
+\end{split}
+\end{equation}
+
+\item O ruído violeta é similar ao ruído azul, mas o ganho é de $6dB$ por oitava:
+
+\begin{equation}\label{eq:violeta}
+\alpha_i = (10^{\frac{6}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \;\;, \quad f_{\text{min}} \approx 15 Hz \\
+\end{equation}
+
+\item O ruído preto possui perdas maiores que $6dB$ por oitava, assim:
+
+\begin{equation}\label{eq:preto}
+\alpha_i=(10^{-\frac{\beta}{20}})^{\log _2 \left( \frac{f_i}{f_{\text{min}}} \right )}\;\;, \quad \beta > 6
+\end{equation}
+
+\item Um pouco diferente dos exemplos acima é o ruído cinza, definido como
+um ruído branco sujeito a uma das curvas iso-audíveis. Como estas curvas são resultados
+experimentais, precisamos neste caso da curva para a realização dos fatores multiplicativos
+$\alpha_i$.
+
+\end{itemize}
+
+Com base nestes ruídos expostos podemos prever a infinidade de ruídos utilizados. Importante
+neste momento é entender que foram expostos somente ruídos com espectro estático. Existem classificações
+de ruídos com variações do espectro no decorrer do tempo. Existem também ruídos que
+são fundamentalmente transientes, como os clicks e os chirps. O primeiro é modelado
+facilmente por um impulso relativamente isolado, enquanto o segundo é uma varredura rápida de 
+alguma banda de frequência~\cite{Cook}.
+
+Os ruídos das equações \ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom},
+\ref{eq:azul}, \ref{eq:violeta} estão na figura \ref{fig:ruidos}. Vale notar
+que os espectros foram feitos com a mesma fase nos coeficientes, de forma que
+pode-se observar a contribuição dos harmônicos agudos no traçado geral dado
+pelas frequências graves e muito explícito no ruído marrom.
+
+\subsubsection{Usos musicais parte 1: trêmolo e vibrato, AM e FM}
 
@@
 de uma segunda tabela $\widetilde{M}_i$ de tamanho $\widetilde{\Lambda}_M$ que apresente também
-um período de onda que oscila entre $[-1,1]$. Descrevemos uma sequência $t_i^{vbr(f')}$ com
+um período de onda que oscila entre $[-1,1]$. Descrevemos uma sequência sonora $t_i^{vbr(f',\,\nu)}$ com
 um vibrato de frequência $f'$ e profundidade  $\mu$ (quantos herz é a oscilação no pico superior)
-ou $\nu$ (quantos semitons) da seguinte forma:
+ou $\nu$ (amplitude do vibrato em semitons) da seguinte forma:
 
@@
     \centering
-    \caption{Espectrograma de um 'vibrato' senoidal de $3Hz$ e profundidade de uma oitava em uma dente de serra de $1000Hz$}
+    \caption{Espectrograma de um som com vibrato senoidal de $3Hz$ e profundidade de uma oitava em uma dente de serra de $1000Hz$}
         \includegraphics[width=\textwidth]{figuras/vibrato___}
@@
 As variáveis $\mu$ e $\nu$ quantificam a intensidade do vibrato: $\mu$ é uma medida direta da quantidade
-de Herz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação ($2\nu$ é o número de semitons entre os picos superioes e inferiores da oscilação de frequência do vibrato).
+de Herz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação ($2\nu$ é o número de semitons entre os picos superioes e inferiores de oscilação da frequência do som $t_i$ causada pelo vibrato).
 Note que $\nu=\log_{2}\frac{f+\mu}{f} $ é conveniente neste caso pois o aumento máximo de frequência
@@
 para cima e para baixo. Note o padrão senoidal do vibrato e as compontentes harmônicas
-da dente de serra. Estas qualidades não são praticáveis em instrumentos
-usuais, introduzindo novidade nas possibilidades musicais. Qualquer forma de onda pode
-ser utilizada para gerar o som e o padrão de oscilação do vibrato. Também é crucial a
+da dente de serra. Qualquer forma de onda pode
+ser utilizada para gerar o som e o padrão de oscilação do vibrato, o que é tão crucial para os usos musicais quanto a
 frequência de oscilação e o desvido de altura envolvido\footnote{O desvio de altura
-é chamado profundidade e é dado por conveniência em semitons ou cents.}.
+é chamado profundidade do vibrato e é dado por conveniência em semitons ou cents.}.
+Estas qualidades não são praticáveis em instrumentos musicais tradicionais, introduzindo novidade nas possibilidades musicais.
 
@@
 
-Onde $V_{dB}$ é profundidade da oscilação em decibels do tremolo e $a_{\text{máx}}=10^{\frac{V_{dB}}{20}}$
+Onde $V_{dB}$ é a profundidade da oscilação em decibels do trêmolo e $a_{\text{máx}}=10^{\frac{V_{dB}}{20}}$
  é o ganho máximo de amplitude envolvido.
@@
 
+A peça \emph{vibra e treme} explora estes recursos dos tremolos e vibratos
+com frequências $f'$
+e profundidades ($\nu$ e $V_{dB}$) diferentes.
+Usadas em associação e isoladamente e com variações progressivas dos parâmetros.
+A peça desenvolve também uma comparação entre os vibratos e tremolos que se desenvolvem em escala logarítmica e em escala linear.
+
 \begin{figure}[h!]
     \centering
-    \caption{Tremolo de profundidade $12dB$ com padrão oscilatório de uma dente de serra $1,5Hz$ em uma senoide de $40Hz$}
+    \caption{Tremolo de profundidade $12dB$ com padrão oscilatório de uma dente de serra em $1,5Hz$ em uma senoide de $40Hz$ (considerada taxa de amostragem $f_a=44,1kHz$)}
         \includegraphics[width=\textwidth]{figuras/tremolo}
@@
 
-\subsubsection{Aplicação de Filtros Digitais}
- As possibilidades
-aqui são variadas e a complexidade facilmente
-foge ao escopo deste trabalho\footnote{A elaboração de filtros
-constitui uma área reconhecidamente complexa, com literatura
-e pacotes de software dedicados. 
-Recomendamos ao leitor
-interessado uma visita à nossa bibliografia, em especial ~\ref{} e ~\ref{}.},
-de forma que focamos em explicitar o que ocorre com as sequências
-de amostras e em descrever os procedimentos usuais.
-A aplicação de filtros pode
-ser parte constituinte da síntese ou poder ser feita posteriormente
-como parte dos processos tipicamente chamados de tratamento sonoro.
-
-Explicitaremos a aplicação de filtros por convolução
-e por equação a diferenças.
-
-\begin{itemize}
-
-\item  Convolução e filtros de resposta ao impulso finita (FIR)
-
-Os filtros aplicados por convolução são conhecidos
-pela sua sigla FIR (do inglês Finite Impulse Response)
-e são caracterizados por possuirem uma representação amostral
-finita no tempo. Esta representação amostral é chamada
-de 'resposta ao impulso' $h_i$. Os filtros FIR são aplicados ao som
-digitalizado pela convolucão do som com sua 
-resposta ao impulso. Para nossos fins, esta
-convolução fica definida como\footnote{Convencionamos $t_i=0$ para as
-amostras não definidas de antemão.}:
-
-\begin{equation}\label{eq:conv}
-t_i'=(t*h)_i=\sum_{j=0}^{min(\Lambda_h,i)}h_{j} . t_{i-j}=\sum_{j=max(i-\Lambda_h,0)}^{i}t_j . h_{i-j}
-\end{equation}
-
-Ou seja, o som $\{t_i'\}$ resultante da convolução de $\{t_i\}$ com a resposta ao impulso $\{h_i\}$
-tem cada i-ésima amostra $t_i$ substituída pela soma de suas últimas amostras $t_{i-j}$
-multiplicadas pela resposta ao impulso em si ($h_i$). Isso é facilmente
-percebido pela visualização da convolução de sequências arbitrárias e dispomos
-para este fim a figura \ref{fig:conv}. Observe que a resposta ao impulso $\{h_i\}$
-é disposta na forma retrograda e que mostramos na figura os cálculos explícitos
-das amostras $t_{12}'$ e $t_{32}'$ do sinal convoluído. O sinal resultante possui
-sempre o tamanho $\Lambda_t+\Lambda_h -1=\Lambda_{t'}$.
-
-\begin{figure}[h!]
-    \centering
-    \caption{Interpretação gráfica da convolução}
-        \includegraphics[width=\textwidth]{figuras/convolucao______}
-        \label{fig:conv}
-\end{figure}
-
-Com este procedimento podemos aplicar reverberadores, equalizadores, delays
-e vários outros tipos de filtros para fins de tratamento sonoro ou
-efeitos musicais/artísticos.
-
-Para a obtenção da resposta ao impulso, pode-se recorrer a medições
-físicas ou à síntese direta do filtro. Para citar um exemplo
-de medições diretas, uma resposta ao impulso para a aplicação
-de reverberação pode resultar da gravação acústica do ambiente ao disparar
-um estalo (cuja gravação se aproxima da resposta ao impulso diretamente) ou ao disparar uma
-varredura de frequência (cuja gravação se aproxima da resposta em frequência).
-Ambas resultam em respostas ao impulso
-que, convoluídas com a sequência sonora, resultam nela própria
-com uma reverberação que se assemelha à reverberação do ambiente 
-em que ocorreu a medição.
-
-Como exemplo de respostas ao impulso sintetizadas, pode-se
-descrever um perfil espectral, realizar a transformada inversa
-de fourier e convoluir o resultante com o som, realizando
-uma filtragem cujo perfil é próximo ao especificado inicialmente\footnote{Especialmente
-se for grande o número de amostras utilizado para especificar o perfil espectral
-do filtro. Este procedimento deve ser testado caso a caso pois estes procedimentos
-espectrais não possuem comportamento previsível.}.
-
-Outro exemplo simples é a aplicação de delays através
-da convolução do som com uma resposta ao impulso que possui um impulso
-em cada reincidência do som na linha de delays intensionada.
-Para ilustrar essa aplicação, dispomos a figura~\ref{fig:delays}
-em que pode-se observar o deslocamento causado pela convolução
-com o impulso. Dependendo da densidade dos impulsos, o resultado
-é de caráter rítmico (20 impulsos por segundo ou menos) ou de amálgama
-sonoro (20-40 impulsos por segundo ou mais). Neste último caso,
-ocorrem processos tipicamente vinculados à síntese granular e às
-equalizações.
-
-\begin{figure}[h!]
-    \centering
-    \caption{Convolução com o impulso e linhas de delays}
-        \includegraphics[width=\textwidth]{figuras/delays__}
-        \label{fig:delays}
-\end{figure}
-
-\item Filtros de resposta ao impulso infinita (IIR)
-
-Esta classe de filtros é
-conhecida pela sigla IIR (do inglês Infinite Impulse Response)
-e é caracterizada por possuir uma representação temporal
-infinita, i.e. a resposta ao impulso não converge para zero. Sua aplicação é usualmente feita pela equação
-a diferenças:
-
-\begin{equation}
-t_i' = \frac{1}{b_0}\left ( \sum_{j=0}^Ja_j . t_{i-j} - \sum_{k=1}^Kb_k . t_{i-k}' \right )
-\end{equation}
-
-com $b_0=1$ na grande maioria dos casos pois basta normalizarmos as variáveis
-resultando em filtro idêntico: $a_j'=\frac{a_j}{b_0}$ e $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
-
-Apontamos filtros IIR bastante simples, úteis e usuais abaixo. Existem
-diversos métodos e ferrametas para a elaboração de filtros IIR
-e esta é uma seleção com fins didáticos e para consulta futura por
-utilidade.
-São filtros de primeira e segunda ordem bem comportados cujas
-filtragens realizadas dispomos na figura \ref{fig:iir}.
-
-\begin{figure}[h!]
-    \centering
-    \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda}}
-        \includegraphics[width=\textwidth]{figuras/iir___}
-        \label{fig:iir}
-\end{figure}
-
-No caso dos filtros de ordem simples, a frequência de corte $f_c$ é onde 
-o filtro realiza uma atenuação de $-3dB \approx 0.707 $ da energia original.
-No caso dos filtros passa e rejeita banda, esta mesma atenuação é
-resultado de duas especificações: $f_c$ e a largura de banda $bw$
-pois em ambas as frequências $f \pm bw$ há uma atenuação de 0.707 da energia original.
-
-Existe amplificação do som no caso dos filtros passa e rejeita banda, quando a frequência
-de corte é baixa e a largura de banda é larga o suficiente. Nos agudos, quando a largura
-de banda é grande o suficiente, estes filtros apresentam
-somente um desvio do perfil esperado, expandindo a envoltória um pouco mais do lado grave da banda em
-evidência.
-
-Para conseguir filtros cujas respostas em frequência possuem formatos específicos,
-o recurso mais simples a esta altura
-é fazer cascata destes filtros (aplicá-los sucessivamente).
-
-Outra possibilidade é usar alguma receita de filtro
-biquad\footnote{Abreviação
-de 'biquadrado' pois sua função de transferência possue dois polos e dois zeros, i.e. sua
-forma normal consiste em dois polinômios quadráticos formando uma fração com um no numerador e o outro no denominador
-$\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1+ b_1.z^{-1} +b_2 . z^{-2}}$.}
-ou rotinas para cálculo de coeficientes
-de filtros Chebichev\footnote{Filtros Butterworth e Elípticos podem
-ser considerados como casos específicos dos Filtros do tipo Chebichev.}.
-Ambas as possibilidades são exploradas
-por títulos em nossas referências, em especial em ~\ref{JOS} e ~\ref{dspguide} e na coleção de filtros da comunidade \emph{Music DSP}, da Universiade de Columbia \ref{music-dsp}.
-Para uma exposição aprofundada do assunto, recomendamos ~\ref{Openheim}.
-
-\end{itemize}
-
-\begin{enumerate}
-\item Passa-baixas de polo simples. A fórmula geral tem
-por referência da frequência de corte $f_c \in (0,\frac{1}{2})$,
-fração da frequência de amostragem $f_a$
-em que há aproximadamente uma atenuação de $3dB$,
-calculamos os coeficientes do filtro IIR
-$a_0$ e $b_1$ 
-através da variável intermediária: $x \in [0,1]$:
-
-\begin{equation}\label{eq:passa-baixas}
-\begin{split}
-x & =e^{-2\pi f_c} \\
-a_0 & =  1-x \\
-b_1 & =  x
-\end{split}
-\end{equation}
-
-\item Passa-altas de polo simples. A fórmula geral,
-com frequência de corte $f_c \in (0,\frac{1}{2})$, fração
-de $f_a$ em que conseguimos aproximadamente uma atenuação de $3dB$, calculada através da variável
-intermediária $x \in [0,1]$.
-
-\begin{equation}\label{eq:passa-altas}
-\begin{split}
-x & =e^{-2\pi f_c} \\
-a_0 & =  \frac{x+1}{2} \\
-a_1 & =  -\frac{x+1}{2} \\
-b_1 & =  x
-\end{split}
-\end{equation}
-
-%\item Passa-banda
-%\item Rejeita-banda
-\item Nó (\emph{notch filter}). Este filtro é parametrizado
-pela frequência central $f$ e a largura de banda $bw$
-(medida em 0,707 da amplitude, i.e. atenuação de $3dB$),
-ambos dados como fracoes de $f_a$, portanto $f,\; bw \in (0,0,5)$.
-Por conveniência definimos as variáveis auxiliares $K$ e $R$ como:
-
-\begin{equation}
-\begin{split}
-R & = 1 - 3BW \\
-K & = \frac{1-2R\cos(2\pi f) + R^2}{2 - 2 \cos (2 \pi f)}
-\end{split}
-\end{equation}
-
-A partir das quais temos dois filtros. O filtro passa banda:
-
-\begin{equation}\label{eq:passa-banda}
-\begin{split}
-a_0 & =  1 - K \\
-a_1 & =  2(K-R)\cos (2\pi f) \\
-a_2 & =  R^2-K \\
-b_1 & =  2R \cos (2\pi f) \\
-b_2 & =  -R^2
-\end{split}
-\end{equation}
-
-e o filtro rejeita banda:
-
-\begin{equation}\label{eq:rejeita-banda}
-\begin{split}
-a_0 & =  K \\
-a_1 & =  -2K\cos (2\pi f) \\
-a_2 & =  K \\
-b_1 & =  2R \cos (2\pi f) \\
-b_2 & =  -R^2
-\end{split}
-\end{equation}
-
-%\item Biquad: pela especificação de uma frequência central, da qualidade
-%e da intensidade do filtro, este filtro é simples e usual para áudio,
-%permitindo ajustes mais finos. Diversas receitas podem ser encontradas
-%na literatura, recomendamos especialmente as diferentes especificações
-%em ~\ref{musicDSP} e ~\ref{dspguide}.
-
-\end{enumerate}
-
-\subsubsection{Ruídos}
-De forma geral, os sons sem altura definida 
-são chamados ruídos.
-Estes sons estão presentes como constituintes importantes dos sons musicais de altura definida,
-como os ruídos presentes nas notas do piano, do violino, etc. Além disso, os instrumentos
-de percussão, em grande parte, não possuem altura definida e seus sons
-são muitas vezes compreendidos como ruídos. Na música eletrônica (incluindo
-a eletroacústica), os ruídos possuem usos amplos, diversificados e comumente
-idiomáticos.
-
-A ausência de uma altura definida é fruto da ausência de características
-harmônicas perceptíveis nas componentes do som. Assim, pode-se inferir
-as incontáveis possibilidades de gerar ruídos. A mais imediata é a utilização
-de valores aleatórios para a geração da sequência sonora $T_i$. Este
-é um método atraente, visto a disponibilidade
-das consagradas distribuições de probabilidade em diversas linguagens de programação,
-mas os resultados não são tão úteis, tendendo geralmente ao ruído branco
-e com difíceis previsões do espectro resultante.
-
-A outra possibilidade é a geração de ruído através do espectro desejado. A partir
-dele, executamos a transformada inversa de fourier. Neste caso, é importante
-que realizemos a distribuição espectral com cuidado pois queremos um perfil
-do módulo do espectro, e a fase é aleatória\footnote{Caso utilizemos a mesma fase
-(ou fases com forte correlação) para
-as componentes envolvidas, o som sintetizado possuirá energia bastante concentrada
-em alguns trechos.}.
-
-Abaixo elencamos alguns ruídos chamados \emph{coloridos} por terem
-sido associados a cores. Lidaremos somente com ruídos de espectro estático.
-
-\begin{itemize}
-
-\item O ruído branco possui este nome por possuir energia distribuida
-igualmente por todas as frequências. Podemos obter a realização
-do ruído branco com a transformada inversa dos seguintes coeficientes:
-
-\begin{equation}\label{eq:branco}
-\begin{split}
-c_0 & =0 \quad \text{pois não queremos bias} \\
-c_i & =e^{j.x}\;,\;\; j^2=-1 \;, \;\; x \; \text{randômico} \; \in \; [0,2\pi]\;,\;\; i \; \in \; [1, \, \frac{\Lambda}{2}-1] \\
-c_{\Lambda/2} & = 10^{-\frac{1}{20}} \approx 0.89 \quad \text{se $\Lambda$ par}\\ 
-c_i & = c_{\Lambda - i}^*\;,\;\; \text{para}\;  i \; > \;  \Lambda/2
-\end{split}
-\end{equation}
-
-O valor de $c_i$ calculado pela exponencial é apenas um artifício para resultar em módulo unitário e fase aleatória.
-Já $c_{\Lambda/2}$ é sempre puramente real (como vimos na sessão anterior) e possui a queda de $1dB$ pois não é resultande de duas componentes como as outras frequências, que por isso recebem $1dB$ adicional ($1dB$ equivale à intensidade adicionada com uma fonte a mais).
-
-\item O ruído rosa possui uma queda de $3dB$ por oitava. Este ruído é muito usual no teste de equipamentos e montagens de aparelhos além de presença destacada na natureza.
-
-\begin{equation}\label{eq:rosa}
-\begin{split}
-f_i & = i \frac{f_a}{\Lambda} \;, \;\; \quad i \leq \frac{\Lambda}{2} \\
-\alpha_i & = (10^{-\frac{3}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \;\;, \quad f_{\text{min}} \approx 15 Hz \\
-c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \\
-c_i & =e^{j.x} . \alpha_i\;,\;\; j^2=-1 \;, \;\;\  x \; \text{randômico} \; \in \; [0,2\pi]\;,\;\; i \; \in \; [i_0, \frac{\Lambda}{2}-1] \\
-c_{\Lambda/2} & = 10^{-\frac{1}{20}} . \alpha_{\Lambda/2} \quad \text{se $\Lambda$ par}\\ 
-c_i & = c_{\Lambda - i}^*\;,\;\; \text{para}\;  i \; > \;  \Lambda/2
-\end{split}
-\end{equation}
-
-A frequência mínima $f_{\text{min}}$ pode ser escolhida com base no limite da audição, pois não escutamos frequência absoluta abaixo de $\approx\; 20Hz$ como altura.
-
-O resto dos ruídos podem ser feitos com base neste procedimento descrito para 
-o ruído rosa. Basta que modifiquemos alguns detalhes. Em especial a equação que define $\alpha_i$.
-
-\item O ruído marrom deve seu nome ao Robert Brown, que descobriu o movimento browniano.
-Embora esta origem seja um tanto díspar do que poderiamos considerar uma associação ligítima com a cor marrom, o ruído ficou consagrado com este nome.
-De qualquer forma, é bastante comum concordarmos que, dados os ruídos branco e rosa, ao ruído marrom, menos estridente,
-é satisfatória a associação com o ruído marrom.
-
-O que caracteriza este ruído é a queda de $6dB$ por oitava. Desta forma, basta modificarmos a equação de $\alpha_i$ 
-no conjunto \ref{eq:rosa} para:
-
-\begin{equation}\label{eq:marrom}
-\alpha_i=(10^{-\frac{6}{20}})^{\log _2 \left( \frac{f_i}{f_{\text{min}}} \right )}
-\end{equation}
-
-\item Ruído azul é aquele em que há um ganho de $3dB$ por oitava em uma banda limitada. Para
-a síntese deste ruído, basta estipularmos a frequência mínima $f_{\text{min}}$ e a frequência
-máxima $f_{\text{máx}}$. Assim, também com base no conjunto de equações \ref{eq:rosa}, 
-basta atentarmos para as sequintes identidades:
-
-\begin{equation}\label{eq:azul}
-\begin{split}
-\alpha_i & = (10^{\frac{3}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \\
-c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \;\; \text{ou} \;\; f_i>f_{\text{máx}} \\
-\end{split}
-\end{equation}
-
-\item O ruído violeta é similar ao ruído azul, mas o ganho é de $6dB$ por oitava:
-
-\begin{equation}\label{eq:violeta}
-\alpha_i = (10^{\frac{6}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \;\;, \quad f_{\text{min}} \approx 15 Hz \\
-\end{equation}
-
-\item O ruído preto possui perdas maiores que $6dB$ por oitava, assim:
-
-\begin{equation}\label{eq:preto}
-\alpha_i=(10^{-\frac{\beta}{20}})^{\log _2 \left( \frac{f_i}{f_{\text{min}}} \right )}\;\;, \quad \beta > 6
-\end{equation}
-
-\item Um pouco diferente dos exemplos acima é o ruído cinza, definido como
-um ruído branco sujeito a uma das curvas iso-audíveis. Como estas curvas são resultados
-experimentais, precisamos neste caso da curva para a realização dos fatores multiplicativos
-$\alpha_i$.
-
-\end{itemize}
-
-Com base nestes ruídos expostos podemos prever a infinidade de ruídos utilizados. Importante
-neste momento é entender que foram expostos somente ruídos com espectro estático. Existem classificações
-de ruídos com variações do espectro no decorrer do tempo. Existem também ruídos que
-são fundamentalmente transientes, como os clicks e os chirps. O primeiro é modelado
-facilmente por um impulso relativamente isolado, enquanto o segundo é uma varredura rápida de 
-alguma banda de frequência.
-
-Os ruídos das equações \ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom},
-\ref{eq:azul}, \ref{eq:violeta} estão na figura \ref{fig:ruidos}. Vale notar
-que os espectros foram feitos com a mesma fase nos coeficientes, de forma que
-pode-se observar a contribuição dos harmônicos agudos no traçado geral dado
-pelas frequências graves e muito explícito no ruído marrom.
-
-\begin{figure}[h!]
-    \centering
-    \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda}}
-        \includegraphics[width=\textwidth]{figuras/ruidos___}
-        \label{fig:ruidos}
-\end{figure}
-
-%Geração estatística de ruídos, distribuições, nomes, equações de geração.
-
-\subsubsection{Usos musicais}
-A este ponto as possibilidades musicais explodiram. Possuímos unidades
-musicais com características próprias de altura (dada pela frequência),
-tímbre (dada pela forma de onda),
-volume (dada pela intensidade) e duração (dada pelo número de amostras).
-Em uma nota, estas características podem ser tratadas
-de forma absoluta ou ao longo de sua duração,
-com a única excessão da duração em si.
-
-Desta forma, os usos musicais que apresentamos é uma coleção de possibilidades
-com o objetivo de exemplificar manipulações sonoras que resultem algo
-musical, por razões variadas e abordadas mais a fundo na próxima sessão.
-
-Uma primeira possibilidade interessante é o estabelecimento de vínculos
-entre os parâmetros do trêmolo e do vibrato e algum parâmetro da nota básica,
-digamos a frequência. Assim, podemos estabelecer que quanto mais aguda é a nota,
-mais alta é a frequência do vibrato e do trêmolo mas menor a sua profundidade.
-Desta forma, tomando por base as equações \ref{vbrGamma}, \ref{vbrF} e \ref{trA}
-podemos escrever:
-
-\begin{equation}
-\begin{split}
-f^{vbr} = f^{tr} & = func_a(f) \\
-\nu & = func_b(f) \\
-V_{dB} & = func_c(f)
-\end{split}
-\end{equation}
-
-Com $f^{vbr}$ e $f^{tr}$ sendo $f'$ nas equações de referência, ou seja, a frequência
-de oscilação do vibrato e do tremolo. Já $\nu$ e $V_{dB}$ são as profundidades
-do vibrato e do tremolo, respectivamente. A montagem musical 
-\emph{tremolos, vibratos e a frequência} explora
-recursos como este e variações da forma de onda da oscilação. As possibilidades
-são numerosas e ricas e, aliás, os tremolos e os vibratos naturais ocorrem
-juntos, i.e. quase todo tremolo natural também apresenta oscilação de frequência
-e vice versa.
-
-Estas oscilações - aplicadas à amplitude, frequência, posição - são ornamentos muito apreciados e possuem nomes
-específicos. Em linguagem pianística,
-os tremolos são ocorrências em que se alternam rapidamente 2 notas (com
-casos de expansão do conceito), ou seja, seriam mais facilmente associados aos vibratos
-segundo a nomenclatura aqui utilizada\footnote{Veja nosso glossário.}.
+----
 
@@
 
-\begin{equation}
+\begin{equation}\label{eq:fmEsp}
 \begin{split}
@@
 \end{equation}
-é a função de Bessel \cite{BesselCCRMA,JOSFM}.
+é a função de Bessel~\cite{BesselCCRMA,JOSFM}.
 Note que, nestas equações, a variação de frequência não respeita a progressão geométrica de frequência que acompanha a percepçao de altura.
-Caso utilizemos as equações \ref{vbrF} teríamos uma distribuição um pouco diferente dos harmônicos:
-
-\begin{equation}
-\begin{split}
-\{t_i'\}_0^{\Lambda -1} &  = \left \{ \cos \left [f_i .
-2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1}  = \left \{ \cos \left [f\left ( \frac{f+\mu}{f} \right )^{sen \left ( f' .
-2 \pi \frac{i}{ f_a -1 } \right )} .
-2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1}  =  \\
- & = \left \{ \cos \left [f\left ( 2^{\frac{\nu}{12}} \right )^{sen \left ( f' .
-2 \pi \frac{i}{ f_a -1 } \right )} .
-2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1} =\left \{ \cos \left [f\left ( 2^{\frac{\nu}{12} \sin \left ( f' .
-2 \pi \frac{i}{ f_a -1 } \right )  } \right ) .
-2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1} \Rightarrow \\
-& \Rightarrow \text{usando: } \left ( 2^x=e^{x \ln 2}, \;\; e^x=\sum_0^\infty \frac{x^n}{n!},\;\; \sin^nx=y_1(n)\sum_k y_2(n,k).\sin[y_3(n,k,x)]   \right ) \Rightarrow \\
-\Rightarrow \{t_i'\}_0^{\Lambda -1} & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu,f) \cos \left [ y(f.f', f_a) \right ]  \right \}_0^{\Lambda-1}
-\end{split}
-\end{equation}
-
-Onde, por brevidade, citamos as identidades úteis para resultar na
-forma final do som resultante, que explicita o conteúdo em frequências puras.
-Podemos observar que, para a FM, se utilizada a progressão exponencial de frequência, o resultado
-é menos comportado. A função de Bessel dependerá não somente da profundidade de oscilação,
-mas também da frequência original do som. Além disso, as frequências dependem do produto das frequências
-do som e da oscilação ($f.f'$), o que também - a princípio - não convém por introduzir complexidade.
+Caso utilizemos as equações~\ref{vbrF} teríamos uma distribuição diferente dos harmônicos.
+Para este fim, dispomos o APÊNDICE~\ref{cap:fmam} onde calculamos o conteúdo espectral da síntese FM feita com oscilações na escala logarítmica.
+De fato, como pode-se notar nos cálculos, o comportamento simples que torna a FM é privilégio das variações lineares utilizadas em~\ref{eq:fmEsp}.
 
@@
 
-No caso de usarmos a modulação na escala logarítmica como nosso tremolo, temos novamente
-um resultado menos comportado:
-\begin{equation}\label{eq:am}
+ Além do comparativo qualitativo que é explorado entre as oscilações nas escalas diferentes, é uma ponte conveniente com as sínteses FM e AM.
+
+\subsubsection{Usos musicais livres}
+A este ponto as possibilidades musicais explodiram. Possuímos unidades
+musicais com características próprias de altura (dada pela frequência),
+tímbre (dada pela forma de onda),
+volume (dada pela intensidade) e duração (dada pelo número de amostras).
+Em uma nota, estas características podem ser consideradas
+de forma absoluta ou tratadas ao longo de sua duração,
+com a única excessão da duração em si.
+
+Desta forma, os usos musicais que apresentamos é uma coleção de possibilidades
+com o objetivo de exemplificar manipulações sonoras que resultem algo
+musical, por razões variadas e aprofundadas na próxima sessão.
+
+Uma primeira possibilidade interessante é o estabelecimento de vínculos
+entre os parâmetros do trêmolo e do vibrato e algum parâmetro da nota básica,
+digamos a frequência. Assim, podemos estabelecer que quanto mais aguda é a nota,
+mais alta é a frequência do vibrato e do trêmolo mas menor a sua profundidade.
+Desta forma, tomando por base as equações \ref{vbrGamma}, \ref{vbrF} e \ref{trA}
+podemos escrever:
+
+\begin{equation}
 \begin{split}
-\{t_i'\}_0^{\Lambda-1} & =\{(a_{\text{máx}})^{ti'} .
-t_i\}_0^{\Lambda-1}= \left \{ a_{\text{máx}}^{\sin \left ( f'.2\pi\frac{i}{f_a -1} \right )} .
-P .\sin \left ( f.2\pi\frac{i}{f_a -1} \right ) \right \}_0^{\Lambda-1}
+f^{vbr} = f^{tr} & = func_a(f) \\
+\nu & = func_b(f) \\
+V_{dB} & = func_c(f)
 \end{split}
@@
 
-Que nos leva pelos mesmos recursos a um espectro com diversas bandas de frequência distante
-em proporções múltiplas da frequência da moduladora.
+Com $f^{vbr}$ e $f^{tr}$ sendo $f'$ nas equações de referência, ou seja, a frequência
+de oscilação do vibrato e do tremolo utilizada na equação~\ref{vbrGamma}. Já $\nu$ e $V_{dB}$ são as profundidades
+do vibrato e do tremolo, respectivamente. A montagem musical 
+\emph{tremolos, vibratos e a frequência} explora
+recursos como este e variações da forma de onda da oscilação. As possibilidades
+são numerosas e ricas e, aliás, os tremolos e os vibratos naturais ocorrem
+juntos, i.e. quase todo tremolo natural também apresenta oscilação de frequência
+e vice versa.
+
+Estas oscilações - aplicadas à amplitude, frequência, posição espacial - são ornamentos muito apreciados e possuem nomes
+específicos. Em linguagem pianística,
+os tremolos são ocorrências em que se alternam rapidamente 2 notas (com
+casos de expansão do conceito), ou seja, seriam mais facilmente associados aos vibratos
+segundo a nomenclatura aqui utilizada\footnote{Veja nosso glossário.}.
+
+---------
 
@@
 Exploramos este recurso da métrica
-estabelecida pela convolução com impulsos na sequência musical \emph{trezinho de caipiras impulsivos}.
-
-Já com os filtros as possibilidades explodem. Pode-se convoluir um sinal para reverberá-lo, para
+estabelecida pela convolução com impulsos na sequência musical \emph{trenzinho de caipiras impulsivos}.
+
+Já com os filtros as possibilidades explodem ainda mais vertiginosamente. Pode-se convoluir um sinal para reverberá-lo, para
 remover algum ruído, para gerar distorções ou tratamento com intuito estético mesmo. Por exemplo,
@@
 e trêmolo e suas características, equalizações, volume etc. Para o resultado final, estas versões do som
-inicial é então mixado como em \ref{eq:mixagem}. A peça \emph{coro infantil} realiza chorus de formas
+inicial é então mixado (ver equação \ref{eq:mixagem}). A peça \emph{chorus infantil} realiza chorus de formas
 diferentes em sons diferentes.

### 2012-10-28 fmam.tex
lipada geral e arrumando AM e FM

\chapter{Síntese FM e AM em escala logarítmica}
\label{cap:fmam}

\subsection{FM}
\begin{equation}
\begin{split}
\{t_i'\}_0^{\Lambda -1} &  = \left \{ \cos \left [f_i . 2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1}  = \left \{ \cos \left [f\left ( \frac{f+\mu}{f} \right )^{sen \left ( f' . 2 \pi \frac{i}{ f_a -1 } \right )} . 2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1}  =  \\
 & = \left \{ \cos \left [f\left ( 2^{\frac{\nu}{12}} \right )^{sen \left ( f' . 2 \pi \frac{i}{ f_a -1 } \right )} . 2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1} =\left \{ \cos \left [f\left ( 2^{\frac{\nu}{12} \sin \left ( f' . 2 \pi \frac{i}{ f_a -1 } \right )  } \right ) . 2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1} \Rightarrow \\
& \Rightarrow \text{usando: } \left ( 2^x=e^{x \ln 2}, \;\; e^x=\sum_0^\infty \frac{x^n}{n!},\;\; \sin^nx=y_1(n)\sum_k y_2(n,k).\sin[y_3(n,k,x)]   \right ) \Rightarrow \\
\Rightarrow \{t_i'\}_0^{\Lambda -1} & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu,f) \cos \left [ y(f.f', f_a) \right ]  \right \}_0^{\Lambda-1}
\end{split}
\end{equation}

Onde, por brevidade, citamos as identidades úteis para resultar na
forma final do som resultante, que explicita o conteúdo em frequências puras.
Podemos observar que, para a FM, se utilizada a progressão exponencial de frequência, o resultado
é menos comportado. A função de Bessel dependerá não somente da profundidade de oscilação,
mas também da frequência original do som. Além disso, as frequências dependem do produto das frequências
do som e da oscilação ($f.f'$), o que também - a princípio - não convém por introduzir complexidade.

\subsection{AM}
No caso de usarmos a modulação na escala logarítmica como nosso tremolo, temos novamente
um resultado menos comportado:
\begin{equation}\label{eq:am}
\begin{split}
\{t_i'\}_0^{\Lambda-1} & =\{(a_{\text{máx}})^{ti'} . t_i\}_0^{\Lambda-1}= \left \{ a_{\text{máx}}^{\sin \left ( f'.2\pi\frac{i}{f_a -1} \right )} . P .\sin \left ( f.2\pi\frac{i}{f_a -1} \right ) \right \}_0^{\Lambda-1} 
\end{split}
\end{equation}

Que nos leva pelos mesmos recursos a um espectro com diversas bandas de frequência distante
em proporções múltiplas da frequência da moduladora.

### 2012-10-28 scripts/convolucao.py

@@ -3,15 +3,15 @@ import numpy as n, pylab as p
 
 ax=p.subplot(111)
 ax.annotate(r"$t_{12}'=\sum_{j=0}^9 \, h_j.t_{12-j}$", xy=(12.2, -2.65),  xycoords='data',
-                xytext=(21, -3.95), textcoords='data',
+                xytext=(23, -3.75), textcoords='data',
                 arrowprops=dict(facecolor='black', shrink=0.05),
-                horizontalalignment='right', verticalalignment='top',
+                horizontalalignment='right', verticalalignment='top', fontsize=20
                 )
 
-ax.annotate(r"$t_{32}'=\sum_{j=0}^9 \, h_j.t_{32-j}$", xy=(31.8, -1.75),  xycoords='data',
-                xytext=(30, -3.95), textcoords='data',
+ax.annotate(r"$t_{32}'=\sum_{j=0}^9 \, h_j.t_{32-j}$", xy=(32.2, -1.75),  xycoords='data',
+                xytext=(43, -3.75), textcoords='data',
                 arrowprops=dict(facecolor='black', shrink=0.05),
-                horizontalalignment='right', verticalalignment='top',
+                horizontalalignment='right', verticalalignment='top', fontsize=20
                 )
 
 
@@ -28,7 +28,7 @@ p.plot(t,'bo', label=r'trecho de sinal sonoro $\{t_i\}_0^{47}$')
 #h=1./x[1:11]**.5
 h=n.random.random(10)
 h=n.array([ 0.5591728 ,  0.59152829,  0.43285462,  0.8870076 ,  0.44892785,0.33476906,  0.8808893 ,  0.39040725,  0.56887214,  0.54278373])
-p.plot(range(3,13),h[::-1]+1,'ro',label=r'retrogrado da resposta ao impulso $\{h_{i-j}\}_{j=max(i-\Lambda_h,0)}^i$')
+p.plot(range(3,13),h[::-1]+1,'ro',label=r'retrogrado da resposta ao impulso $\{h_{i-j}\}_{j\; =\; max(i-\Lambda_h,0)}^i$')
 for i in xrange(3,12):
     p.plot([i,i],[-1,2],'y-.')
 p.plot([12,12],[-5,2], 'y-.',linewidth=3)
[…]

### 2012-10-28 devEresultados.tex
repassada geral na parte 2 terminada

@@
 
-Para realizar um \emph{fade out} (diminuição progressiva do volume até o volume zero),
-podemos tomar $a_0=1$ e $a_{\Lambda-1}=0$. Para um \emph{fade in} (acréscimo progressivo de volume a partir do volume zero), podemos
-tomar $a_0=1$, $a_{\Lambda-1}=2$ e subtrair $1$ de toda sequência: $\{a_i'\}_0^{\Lambda-1}=\{a_i-1\}_0^{\Lambda-1}$.
+Algum cuidado é necessário para lidar com $a=0$. Para realizar um \emph{fade out} (diminuição progressiva do volume até o volume zero),
+podemos tomar $a_0=2$ e $a_{\Lambda-1}=1$ e subtrair $1$ de toda a sequência $\{a_i'\}=\left\{2\left(\frac{1}{2}\right)^{\frac{i}{\Lambda-1}}-1\right\}$.
+Para um \emph{fade in} (acréscimo progressivo de volume a partir do volume zero), podemos
+tomar $a_0'=1$, $a_{\Lambda-1}'=2$ e subtrair $1$ de toda sequência: $\{a_i\}_0^{\Lambda-1}=\left\{2^{\frac{i}{\Lambda-1}}-1\right\}_0^{\Lambda-1}$.
 
@@
 Enquanto o vibrato é uma variação periódica de altura (frequência),
-o tremolo é uma variação periódica de volume (intensidade).
+o tremolo é uma variação periódica de volume (intensidade).\footnote{Alguns instrumentos e contextos musicais usam nomenclaturas diferentes.
+Por exemplo, no piano, o chamado trêmolo é um vibrato e um tremolo segundo a classificação aqui utilizada.
+As definições presentes neste trabalho tem por base uma literatura mais abrangente do que a utilizada para um único instrumento, prática ou tradição musical e comum em contextos de teoria musical e musica eletrônica~\cite{Lacerda,Harmonia}.}
 
@@
 \begin{equation}\label{vbrT}
-T_i^{f, vbr(f')}=\{ t_i^{f,vbr(f')} \}_0^{\Lambda-1}=\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \}_0^{\Lambda-1}
+T_i^{f, vbr(f',\,\nu)}=\{ t_i^{f,vbr(f',\,\nu)} \}_0^{\Lambda-1}=\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \}_0^{\Lambda-1}
 \end{equation}
@@
 
-A peça \emph{vibra e treme} explora estes recursos dos tremolos e vibratos
+A montagem musical \emph{Vibra e treme} explora estes recursos dos tremolos e vibratos
 com frequências $f'$
 e profundidades ($\nu$ e $V_{dB}$) diferentes.
-Usadas em associação e isoladamente e com variações progressivas dos parâmetros.
-A peça desenvolve também uma comparação entre os vibratos e tremolos que se desenvolvem em escala logarítmica e em escala linear.
+Usadas em associação e isoladamente os tremolos e vibratos e com variações progressivas dos parâmetros\footnote{Vale lembrar que os tremolos e vibratos ocorrem muitas vezes juntos em instrumentos tradicionais e na voz.}.
+A peça desenvolve também uma comparação entre os vibratos e tremolos em escala logarítmica e em escala linear para uma apreciação qualitativa.
 
@@
     \centering
-    \caption{Tremolo de profundidade $12dB$ com padrão oscilatório de uma dente de serra em $1,5Hz$ em uma senoide de $40Hz$ (considerada taxa de amostragem $f_a=44,1kHz$)}
+    \caption{Tremolo de profundidade $V_{dB}=12dB$ com padrão oscilatório de uma dente de serra em $f'=1.5Hz$ em uma senoide de $f=40Hz$ (considerada taxa de amostragem $f_a=44,1kHz$)}
         \includegraphics[width=\textwidth]{figuras/tremolo}
@@
 \end{figure}
-
-As equações ~\ref{vbrT}, ~\ref{vbrGamma}, ~\ref{vbrF} e ~\ref{vbrAux}
-são utilizadas para síntese nos casos em que $f'$ é
-maior que $20Hz$ e as oscilações de frequência resultam em padrões complexos
-de componentes espectrais harmônicas e não harmônicas. É a chamada síntese
-por frequência modulada (FM).
-
-As equações ~\ref{trA} e ~\ref{trT}
-(junto às ~\ref{vbrT} e ~\ref{vbrGamma} para a realização da sequência auxiliar $t_i'$) 
-resultam na síntese chamada de amplitude modulada (AM)
-quando $f'$ é maior que $20Hz$\footnote{As sínteses
-FM e AM são técnicas com diversas amplicações na música e também em outras áreas,
-especialmente conhecidos são os usos telecomunicações para transferência de informações
-via ondas eletromagnéticas. O leitor interessado pode visitar os itens relacionados
-da nossa bibliografia para uma abordagem musical (veja especialmente ~\ref{} e ~\ref{})
-ou aproveitar nossa abordagem breve sobre o assunto nos usos musicais desta sessão, logo abaixo.}.
-
-----
 
@@
 altura ($\approx 20Hz$) gera rugosidades para
-ambos tremolos e vibratos. A sequência \emph{Rugosidades}
+ambos tremolos e vibratos. Estas rugosidades são muito apreciadas
+tanto na tradição erudita quanto na música eletrônica atual, especialmente no \emph{Dubstep}~\cite{DubRugoso}.
+A rugosidade também é gerada através de conteúdos espectrais que geram batimentos~\cite{Porres}.  A sequência \emph{Rugosidades}
 explora este limiar com concomitâncias de tremolos e vibratos na mesma
@@
 como eventos identificáveis. Neste caso, as oscilações se apresentam na forma de espectro
-audível como altura, $f'$ e a forma de onda realizam alterações espectrais no som
+audível como altura, $f'$, $\mu$ e a forma de onda realizam alterações espectrais no som
 original $T_i$ de formas diferentes para os tremolos e para os vibratos, são as 
-chamadas sínteses AM (Amplitude Modulation) e FM (Frequency Modulation).
-Estas são técnicas muito poderosas e conhecidas, com aplicações em sintetizadores
-históricos como o Yamaha XXX e com aplicações fora da música, como em telecomunicações
-(ex. radios AM e FM).
-
-Para nosso caso, em resumo, podemos entender a síntese FM através do caso entre senoides
+chamadas sínteses AM (\emph{Amplitude Modulation}) e FM (\emph{Frequency Modulation}).
+Estas são técnicas muito poderosas e conhecidas, com clássicas aplicações em sintetizadores
+históricos como o Yamaha XXX e com aplicações fora da música, como em telecomunicações para transmissão de informação via ondas eletromagnéticas (ex. radios AM e FM).
+
+Para fins musicais e em resumo, podemos entender a síntese FM através do caso entre senoides
 e decompor os sinais em seus espectros de Fourier (i.e. senoidais) para casos mais complexos.
-Assim, a síntese FM realizada com um vibrato de frequência $f'$ em um som $T_i$ de frequência $f$
+Assim, a síntese FM realizada com um 'vibrato' senoidal de frequência $f'$ e profundidade $\mu$ em um som também senoidal $T_i$ de frequência $f$
 e duração $\Delta$ segundos (assumimos $\Lambda = \lfloor \Delta . f_a \rfloor $ amostras)
@@
 \end{equation}
-é a função de Bessel~\cite{BesselCCRMA,JOSFM}.
-Note que, nestas equações, a variação de frequência não respeita a progressão geométrica de frequência que acompanha a percepçao de altura.
+é a função de Bessel~\cite{BesselCCRMA,JOSFM} que, na FM, especifica as amplitudes de cada componente.
+Note que, nestas equações, a variação de frequência não respeita a progressão geométrica de frequência que acompanha a percepção de altura.
 Caso utilizemos as equações~\ref{vbrF} teríamos uma distribuição diferente dos harmônicos.
 Para este fim, dispomos o APÊNDICE~\ref{cap:fmam} onde calculamos o conteúdo espectral da síntese FM feita com oscilações na escala logarítmica.
-De fato, como pode-se notar nos cálculos, o comportamento simples que torna a FM é privilégio das variações lineares utilizadas em~\ref{eq:fmEsp}.
+De fato, como pode-se notar nos cálculos, o comportamento simples que torna a FM atraente é obtido somente com as variações lineares utilizadas em~\ref{eq:fmEsp}.
 
@@
 
-Ou seja, o sinal resultante é o sinal original (chamada portadora, com conteúdos harmônicos como a senoide de frequência $f$)
+Ou seja, o som resultante é o original
 e a reprodução de seu conteúdo espectral acima e abaixo da frequência
-original, distantes na proporção de cada harmônico ($f'$) da moduladora.
-
- Além do comparativo qualitativo que é explorado entre as oscilações nas escalas diferentes, é uma ponte conveniente com as sínteses FM e AM.
-
-\subsubsection{Usos musicais livres}
+original, distantes $f'$ de $f$.
+Observe que utilizamos novamente a variação na escala linear de amplitude.
+Dispomos no APÊNDICE~\ref{cap:fmam} uma exposição do espectro da AM se realizada na escala logarítmica de amplitude, que também perde o comportamento simples.
+
+A realização da FM e AM através de LUTs é consideravelmente simples.
+A frequência $f$ a ser modulada, chamada portadora, é modulada pela $f'$, chamada moduladora.
+No jargão de FM e AM, $\mu$ e $\alpha=10^{\frac{V_{dB}}{20}}$ são chamados índices de modulação.
+Assim, temos
+para o padrão de vibração da moduladora $\{t_i'\}$ as mesmas equações utilizadas para os tremolos e vibratos:
+
+\begin{equation}\label{fmGamma}
+\gamma_i'=\left \lfloor i f' \frac{\widetilde{\Lambda}_M}{f_a} \right \rfloor
+\end{equation}
+
+\begin{equation}\label{fmAux}
+t_i'=\widetilde{m}_{\gamma_i' \;\% \widetilde{\Lambda}_M}
+\end{equation}
+
+Para aplicação da moduladora $\{t_i'\}$ na portadora $\{t_i\}$,
+temos, para a FM:
+
+\begin{equation}\label{fmF}
+f_i=f + \mu . t_i'
+\end{equation}
+
+\begin{equation}\label{fmGamma}
+\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_i \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}(f+\mu .
+t_i') \right\rfloor
+\end{equation}
+
+\begin{equation}\label{fmT}
+T_i^{f,\, FM(f',\,\mu)}=\{ t_i^{f,\,FM(f',\,\mu)} \}_0^{\Lambda-1}=\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \}_0^{\Lambda-1}
+\end{equation}
+
+Onde $\widetilde{l}$ é um período da forma de onda de comprimento $\widetilde{\Lambda}$ da portadora.
+
+Para realizar a AM, basta modular $\{t_i\}$ com $\{t_i'\}$ com as equações (lembrando que $\alpha=10^{\frac{V_{dB}}{20}}$):
+
+\begin{equation}\label{amA}
+a_i=1 + \alpha . t_i'
+\end{equation}
+
+\begin{equation}\label{trT}
+T_i^{f,\,AM(f',\,\alpha)}=\{ t_i^{f,\,AM(f',\,\alpha)} \}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}=\{t_i . (1 + \alpha . t_i')    \}_0^{\Lambda-1}
+\end{equation}
+
+\subsubsection{Usos musicais parte 2}
 A este ponto as possibilidades musicais explodiram. Possuímos unidades
@@
 \emph{tremolos, vibratos e a frequência} explora
-recursos como este e variações da forma de onda da oscilação. As possibilidades
-são numerosas e ricas e, aliás, os tremolos e os vibratos naturais ocorrem
-juntos, i.e. quase todo tremolo natural também apresenta oscilação de frequência
-e vice versa.
-
-Estas oscilações - aplicadas à amplitude, frequência, posição espacial - são ornamentos muito apreciados e possuem nomes
-específicos. Em linguagem pianística,
-os tremolos são ocorrências em que se alternam rapidamente 2 notas (com
-casos de expansão do conceito), ou seja, seriam mais facilmente associados aos vibratos
-segundo a nomenclatura aqui utilizada\footnote{Veja nosso glossário.}.
-
----------
-
-Com relação à convolução, podemos estabelecer um pulso musical (a exemplo de um pulso BPM)
+recursos como este e variações da forma de onda da oscilação com vínculos, de modo a formar um \emph{idioma musical}\footnote{Veja na próxima sessão.}.
+
+Com relação à convolução, podemos estabelecer uma duração como pulso musical - a exemplo de um pulso BPM - 
 e distribuir impulsos no decorrer deste pulso, de forma a estabelecer métricas e rítmos\footnote{Lembrando
-que a convolução com o impulso resulta no som deslocado ao instande de ocorrência do impulso}.
+que a convolução com o impulso resulta no som deslocado ao instante de ocorrência do impulso.}.
 Por exemplo, se escolhemos 2 impulsos igualmente espaçados, estaremos fazendo uma
@@
 do pulso, com uma marcação rítmica usada tanto em divisões binárias quanto ternárias em diversos
-estilos de música étnica e tradicional. 
+estilos de música étnica e tradicional~\cite{Gramani}. 
 Os próprios valores absolutos destes impulsos resultam em proporções entre as amplitudes dos sinais
@@
 Já com os filtros as possibilidades explodem ainda mais vertiginosamente. Pode-se convoluir um sinal para reverberá-lo, para
-remover algum ruído, para gerar distorções ou tratamento com intuito estético mesmo. Por exemplo,
-aplicando um filtro bassa banda em que deixamos passar somente entre $1kHz$ e $3kHz$, geramos um som
+remover algum ruído, para gerar distorções ou para tratamento com intuito estético mesmo (soa melhor por algum motivo). Por exemplo,
+com a aplicação de um filtro passa banda em que deixamos passar somente entre $1kHz$ e $3kHz$, geramos um som
 que parece de telefone ou de televisão antiga. Ao remover com alguma precisão somente
@@
 e propriamente musical seria realizar filtragens em bandas específicas e usar estas bandas
-pré estabelecida como um parâmetro adicional das notas.
-
-Um recurso musicalmente forte é a filtragem dependente do tempo, no qual convém os filtros IIR, em que
+pré-estabelecidas como um parâmetro adicional das notas.
+
+Um recurso considerado musicalmente forte pelo resultado que se aproxima dos instrumentos tradicionais é a filtragem dependente do tempo~\cite{Roederer}, no qual convém os filtros IIR, em que
 os valores do filtro podem ser modificados - com algum cuidado - ao longo de sua aplicação. Cascatas
 destes filtros podem realizar filtragens complexas e mais precisas. A montagem \emph{ruídos e faixas} explora
-este recursos, realizando filtragens em ruídos diversos, com sequênciamento de disposição de bandas
-e ruídos específicos, e filtragens que variam no tempo.
-
-Um uso incrementado destes recursos todos seria a realização de um efeito chamado \emph{chorus}. Ao
+este recursos, realizando filtragens em ruídos diversos e síntese de ruídos diversos.
+
+Estes recursos todos utilizados em conjunto podem incindir na realização de um efeito chamado \emph{chorus}. A
 exemplo do que ocorre com um coro de cantores, neste efeito o som é realizado com diversas pequenas modificações,
@@
 e trêmolo e suas características, equalizações, volume etc. Para o resultado final, estas versões do som
-inicial é então mixado (ver equação \ref{eq:mixagem}). A peça \emph{chorus infantil} realiza chorus de formas
+inicial são então mixadas (ver equação \ref{eq:mixagem}). A peça \emph{chorus infantil} realiza chorus de formas
 diferentes em sons diferentes.
+
+Para terminar esta sessão com algo simples, vamos descrever a envoltória de volume chamada ADSR (sigla de \emph{Atack-Decay-Sustain-Release}).
+A envoltória ADSR possui numerosas implementações em sintetizadores de diversas procedências.
+Uma implementação pioneira pode ser encontrada no Hammond Novachord de 1938~\cite{ADSR} e algumas variantes são citadas logo abaixo.
+
+A envoltória ADSR canônica é caracterizada por 4 parâmetros: duração do ataque (que o som demora para atingir seu volume máximo), duração do decaimento (segue ao ataque imediatamente), nível de volume de sustentação (em que o volume fica estável após o decaimento) e duração de soltura (após a sustentação, nesta duração o volume decai a zero).
+Observer que o tempo de sustentação não está especificado, isso depende da nota em si.
+Note que o tempo de sustentação não é especificado, pois é resultante da duração em si menos os tempos de ataque, decaimento e soltura.
+
+A aplicação da envoltória ADSR com durações $\Delta_A$, $\Delta_D$ e $\Delta_R$, duração total $\Delta$ e nível de sustentação $a_S$, dado
+como fração da amplitude máxima, em uma sequência sonora $t_i$ qualquer pode ser feita da seguinte forma:
+
+\begin{equation}
+\begin{split}
+\{a_i\}_0^{\Lambda_A-1} & = \left\{2^{\frac{i}{\Lambda_A-1}}-1\right\}_0^{\Lambda_A-1} \\
+\{a_i\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} & =\left\{a_S^{\frac{i-\Lambda_A}{\Lambda_D-1}}  \right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} \\
+\{ a_i \}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} & =\left\{ a_S \right\}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} \\
+\{ a_i \}_{\Lambda-\Lambda_R}^{\Lambda-1} & =\left\{ (a_S+1)\left(\frac{1}{a_S+1} \right)^{\frac{i+\Lambda_R-\Lambda}{\Lambda-\Lambda_R}} -1 \right\}_{\Lambda-\Lambda_R}^{\Lambda-1}
+\end{split}
+\end{equation}
+
+Com $\Lambda_X=\lfloor \Delta . f_a \rfloor\;\;\forall\;\; X$.
+
+\begin{equation}
+\left\{t_i^{ADSR}\right\}_0^{\Lambda-1} =\{t_i . a_i\}_0^{\Lambda-1}
+\end{equation}
+
+\begin{figure}[h!]
+    \centering
+    \caption{Envelope ADSR canônico e uma sequência sonora arbitrária sendo submetida ao envelope}
+        \includegraphics[width=\textwidth]{figuras/adsr}
+        \label{fig:adsr}
+\end{figure}
+
+Podemos ver esquematicamente esta envoltória na figura~\ref{fig:adsr}. O envelope ADSR descrito é
+uma implementação clássica e paradigmática que comporta variações diversas. Por exemplo,
+entre o ataque e o decaimento, pode-se adicionar uma partição adicional em que a amplitude
+máxima perdura. Outro exemplo comum é a elaboração de traçados mais elaborados para o
+ataque ou para o decaimento.
 

### 2012-10-28 scripts/adsr.py

@@ -1,0 +1,59 @@
+#-*- coding: utf8 -*-
+import numpy as n, pylab as p
+
+ax=p.subplot(111)
+
+De=2*100. #duracao total (\Delta)
+DA=2*20. # duracao do ataque \Delta_A
+DD=2*20. # duracao do decay \Delta_D
+DR=2*20. # duracao do release \Delta_R
+SS=.4 # fração da amplitude em que ocorre o sustain
+
+A=2**(n.arange(DA)/(DA-1))-1 # amostras do ataque
+s=n.copy(A) 
+D=SS**(n.arange(DA,DA+DD)/DD-1) # amostras do decay
+s=n.hstack((  s,  D  ))
+S=SS*n.ones(De-DR-(DA+DD)) # amostras do sustain
+s=n.hstack((  s, S  ))
+R=(SS+1)*(1/(SS+1))**((n.arange(De-DR,De)+DR-De)/(DR))-1 # amostras do release
+s=n.hstack((  s,  R  ))
+
+p.plot(A,                      'bo', markersize=4)
+p.plot(n.arange(DA,DA+DD),D,   'go', markersize=4)
+p.plot(n.arange(DA+DD,De-DR),S,'ko', markersize=4)
+p.plot(n.arange(De-DR,De),R,   'ro', markersize=4)
+
+p.text(2*6 ,.7,"A",fontsize=48,color='blue')
+p.text(2*32,.7,"D",fontsize=48,color='green')
+p.text(2*56,.5,"S",fontsize=48,color='black')
+p.text(2*91,.3,"R",fontsize=48,color='red')
+
+som=n.random.random(De)*2-1
+som=n.sin(n.linspace(0,45*2*n.pi,De,endpoint=False))
+som=som*s
+p.plot(som,'c*',markersize=9,label=u"Som amostrado submetido ao envelope ADSR")
+p.plot(som,'c')
+
+p.legend(loc="upper right",prop={'size':16})
+p.xlim(-10,De+10)
+p.ylim(-1.2,1.2)
+
+p.yticks((-1,0,SS,1),(-1,0,r"$a_S$",1),fontsize=26)
+p.xticks(())
+
+
+p.arrow(0,-1.1,DA,0,length_includes_head=True,shape='full',head_width=.1,color='cyan'     ,width=0.009)
+p.arrow(DA,-1.1,-DA,0,length_includes_head=True,shape='full',head_width=.1,color='blue'   ,width=0.009)  
[…]

### 2012-10-28 devEresultados.tex
adicionadas figuras para lookuptable e para transicoes

@@
 
+\begin{figure}[h!]
+    \centering
+    \caption{Lookup Table}
+        \includegraphics[width=\textwidth]{figuras/lut}
+        \label{fig:lut}
+\end{figure}
+
 O procedimento de busca em tabelas é visto geralmente como um
@@
 \begin{equation}
-T_i^{f,\,\Delta}=\{t_i^f\}_0^{\lfloor f_a .
-\Delta \rfloor -1} = \{ \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} } \}_{0}^{\Lambda-1}\; , \quad \text{onde} \;\; \gamma_i = \left \lfloor i .
+T_i^{f,\,\Delta}=\left\{t_i^f\right\}_0^{\lfloor \, f_a .
+\Delta \, \rfloor -1} = \left\{ \, \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} }\, \right\}_{0}^{\Lambda-1}\; , \quad \text{onde} \;\; \gamma_i = \left \lfloor i .
 f \frac{ \widetilde{\Lambda}}{f_a} \right \rfloor
@@
 Ou seja, dada a tabela de busca, basta termos a sequência de índices corretos
-para sintetizar o som em qualquer frequência. Para o cálculo do inteiro $\gamma_i$, convencionamos
+para sintetizar o som em qualquer frequência. Dispomos graficamente
+o procedimento de LUT na figura~\ref{fig:lut} onde calculamos explicitamente
+também uma amostra de $\{t_i\}$ a partir de $\{\widetilde{l}_i\}$ para uma frequência
+de $f=200Hz$ e $\widetilde{\Lambda}=128$ considerada a taxa de amostragem em $f_a=44.1kHz$.
+
+Para o cálculo do inteiro $\gamma_i$, convencionamos
 atribuir a ele o maior inteiro menor que $i.f\frac{\widetilde{\Lambda}}{f_a}$.
@@
 mesmo se o tomarmos como o maior inteiro acima da multiplicação
-ou se arredondarmos por outros métodos. Além disso, para fins de síntese, em $f_a=44,1 kHz$
+ou se arredondarmos por outros métodos. Além disso, para fins de síntese, em $f_a=44.1 kHz$
  o padrão é usar $\widetilde{\Lambda} = 1024$ amostras, pois já não gera ruído
@@
 Segundo revela a lei de Weber e Fechner, nossa percepção tem uma relação logarítmica com
-o estímulo que a causa~\cite{Weber-Fechner}. Mesmo assim, dado o uso, explicitaremos também a variação
+o estímulo que a causa~\cite{Weber-Fechner}, em outras palavras, um estímulo em progressão exponencial
+é percebido como linear. Mesmo assim, dado o uso, explicitaremos também a variação
 linear e começaremos descrevendo esta por razões didáticas.
@@
 \begin{equation}\label{indiceLinear}
-\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i=\left \lfloor \sum_{j=0}^{i} f_i\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor \sum_{j=0}^{i}  \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right ] \right \rfloor
+\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i=\left \lfloor \sum_{j=0}^{i} f_i\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor i \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right ] \right \rfloor
 \end{equation}
@@
 Note que a equação \ref{serieAmostralLog} se mantém (permanece idêntica à equação \ref{serieAmostralLin}).
+
+\begin{figure}[h!]
+    \centering
+    \caption{Transições de intensidade para diferentes valores de $\alpha$ (veja equação~\ref{transAmp})}
+        \includegraphics[width=\textwidth]{figuras/transicao}
+        \label{fig:transicao}
+\end{figure}
 
@@
 Caso $\alpha=1$, a variação de amplitude segue exatamente a progressão geométrica que caracteriza
-a percepção linear.
+a percepção linear. A figura~\ref{fig:transicao} exibe as transições para diferentes valores de $\alpha$.
 
@@
 
+\begin{figure}[h!]
+    \centering
+    \caption{Interpretação gráfica da convolução}
+        \includegraphics[width=\textwidth]{figuras/convolucao______}
+        \label{fig:conv}
+\end{figure}
+
 Os filtros aplicados por convolução são conhecidos
@@
 sempre o tamanho $\Lambda_t+\Lambda_h -1=\Lambda_{t'}$.
-
-\begin{figure}[h!]
-    \centering
-    \caption{Interpretação gráfica da convolução}
-        \includegraphics[width=\textwidth]{figuras/convolucao______}
-        \label{fig:conv}
-\end{figure}
 
@@
 filtragens realizadas dispomos na figura~\ref{fig:iir}.
-
-\begin{figure}[h!]
-    \centering
-    \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda}}
-        \includegraphics[width=\textwidth]{figuras/iir___}
-        \label{fig:iir}
-\end{figure}
 
@@
 através da variável intermediária: $x \in [0,1]$:
+
+\begin{figure}[h!]
+    \centering
+    \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda}}
+        \includegraphics[width=\textwidth]{figuras/iir___}
+        \label{fig:iir}
+\end{figure}
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[htpq!]
     \centering

### 2012-10-28 scripts/progressao.py

@@ -1,20 +1,0 @@
-# para testar progressoes logaritmicas e outras
-import numpy as n, pylab as p
-
-f0=200.
-ff=600.
-ii=1000 # tamanho da sequencia
-iis=n.arange(ii)
-
-fi=f0*(ff/f0)**(      (iis/float(ii-1))    )
-
-for i in xrange(10):
-    fi=f0*(ff/f0)**(      (iis/float(ii-1))**i    )
-    p.plot(fi)
-
-fi=f0+(ff-f0)*(i/(ii-1))
-
-
-p.plot(fi)
-p.show()
-

### 2012-10-29 devEresultados.tex
revisao completa da parte 2 terminada

@@
 necessidade\footnote{Em contraposição ao cálculo das amostras individuais
-de forma analítica.}. Em nosso caso, usaremos tabelas de busca para os
+de forma analítica e permitindo o uso de funções sem possibilidade de cálculo direto
+(por serem amostras recolhidas diretamente de algum fenômeno natural, por exemplo).}. Em nosso caso, usaremos tabelas de busca para os
 sons criados, as notas. Assim, estas tabelas são unidimensionais e com
@@
 também uma amostra de $\{t_i\}$ a partir de $\{\widetilde{l}_i\}$ para uma frequência
-de $f=200Hz$ e $\widetilde{\Lambda}=128$ considerada a taxa de amostragem em $f_a=44.1kHz$.
+de $f=200Hz$, $\widetilde{\Lambda}=128$ e considerada a taxa de amostragem em $f_a=44.1kHz$.
 
@@
 simplifica diversas rotinas e é computacionalmente coerente.
-Para um uso clássico das LUTs na síntese sonora musical, recomendamos procurar pela chamada \emph{Wavetable Synthesis} que consiste em várias LUTs utilizadas em conjunto através da mixagem para gerar uma nota musical quasi-periódica~\cite{Cook,Wavetable}.
+
+O uso de LUTs é bastante difundido nas implementações computacionais
+voltadas para música e para um uso clássico que explora com ênfase
+as LUTs na síntese sonora musical, recomendamos procurar pela chamada 
+\emph{Wavetable Synthesis} que consiste em várias LUTs utilizadas em 
+conjunto através da mixagem para gerar uma nota musical quasi-periódica~\cite{Cook,Wavetable}.
 
@@
 Segundo revela a lei de Weber e Fechner, nossa percepção tem uma relação logarítmica com
-o estímulo que a causa~\cite{Weber-Fechner}, em outras palavras, um estímulo em progressão exponencial
+o estímulo que a causa~\cite{Weber-Fechner}. Em outras palavras, um estímulo em progressão exponencial
 é percebido como linear. Mesmo assim, dado o uso, explicitaremos também a variação
@@
 \begin{equation}\label{indiceLinear}
-\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i=\left \lfloor \sum_{j=0}^{i} f_i\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor i \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right ] \right \rfloor
+\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i=\left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{j}{\Lambda-1} \right ] \right \rfloor
 \end{equation}
@@
 \begin{equation}\label{serieAmostralLin}
-\{t_i^{\;\overline{f_0,\, f_{\Lambda-1}}}\}_0^{\Lambda-1}=\{\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\}_0^{\Lambda-1}
+\left\{t_i^{\;\overline{f_0,\, f_{\Lambda-1}}}\right\}_0^{\Lambda-1}=\left\{\,\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\,\right\}_0^{\Lambda-1}
 \end{equation}
@@
 As equações \ref{freqLinear}, \ref{indiceLinear} e \ref{serieAmostralLin} são relativas à progressão linear
-da frequência. Como assinalado, nossa percepção segue a progressão logarítmica:
+da frequência. Como assinalado para o caso geral, também aqui
 uma progressão de frequência
@@
 \begin{equation}\label{indiceExponencial}
-\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i=\left \lfloor \sum_{j=0}^{i} f_i\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor \sum_{j=0}^{i} f_0 \frac{\widetilde{\Lambda}}{f_a} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}} \right \rfloor
+\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i=\left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor \sum_{j=0}^{i} f_0 \frac{\widetilde{\Lambda}}{f_a} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{j}{\Lambda -1}} \right \rfloor
 \end{equation}
@@
 \begin{equation}\label{serieAmostralLog}
-\{t_i^{\;\overline{f_0,\,f_{\Lambda-1}}}\}_0^{\Lambda-1}=\{\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\}_0^{\Lambda-1}
+\left\{t_i^{\;\overline{f_0,\,f_{\Lambda-1}}}\right\}_0^{\Lambda-1}=\left\{\,\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\,\right\}_0^{\Lambda-1}
 \end{equation}
@@
  $t_i^{\;\overline{f_0,f_{\Lambda-1}}}=\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}$.
-Note que a equação \ref{serieAmostralLog} se mantém (permanece idêntica à equação \ref{serieAmostralLin}).
+Note que a equação \ref{serieAmostralLog} se mantém (é idêntica à equação \ref{serieAmostralLin}).
 
@@
     \centering
-    \caption{Transições de intensidade para diferentes valores de $\alpha$ (veja equação~\ref{transAmp})}
+    \caption{Transições de intensidade para diferentes valores de $\alpha$ (veja equações~\ref{seqAmp} e ~\ref{transAmp})}
         \includegraphics[width=\textwidth]{figuras/transicao}
@@
 da transição). Este procedimento é mais comum para o caso de variações de volume através da energia
-da onda vibratória\footnote{O aumento do volume pode ocorrer por diversas características, como
-a reverberação e a concentração de harmônicos agudos e a energia. A mais facilmente controlada
-é a energia da onda (veja equação ~\ref{potencia}) e a energia também pode variar de diferentes formas. A forma mais
-simples e conveniente geralmente é a variação de amplitude. A variação de energia sem variação de
-amplitude é comumente chamado de \emph{compressão sonora} e é muito útil e corriqueiro na prática
-de produção musical atual}. Nestes casos, o procedimento comum é multiplicar a sequência original
-(seja ela gerada ou pré-estabelecida) pela sequência $\left ( \frac{i}{\Lambda-1} \right )^\alpha$
-onde alpha é o coeficiente citado.
+da onda vibratória\footnote{O alteração do volume pode ocorrer por diversas características como
+a reverberação, a concentração de harmônicos agudos e a energia. Para nós, neste ponto da exposição,
+a mais facilmente controlada é a energia da onda (veja equação ~\ref{potencia}) e a energia também pode variar de diferentes formas.
+Uma forma mais simples e conveniente é a variar a amplitude pela multiplicação da sequência toda
+por algum número real. A variação de energia sem variação de
+amplitude é comumente chamada de \emph{compressão sonora} e é muito útil e corriqueira na prática
+de produção musical atual~\cite{compressao}.}. Podemos obter este resultado multiplicando a sequência original
+(seja ela gerada ou pré-estabelecida) pela sequência $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$
+onde $\alpha$ é o coeficiente citado e $a_{\Lambda-1}$ é fração da amplitude original que se visa atingir ao final da transição.
 
@@
 
-\begin{equation}
+\begin{equation}\label{seqAmp}
 \{a_i\}_0^{\Lambda-1}=\left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}=\left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1} \text{ com } a_0=1
@@
 Caso $\alpha=1$, a variação de amplitude segue exatamente a progressão geométrica que caracteriza
-a percepção linear. A figura~\ref{fig:transicao} exibe as transições para diferentes valores de $\alpha$.
-
-Algum cuidado é necessário para lidar com $a=0$. Para realizar um \emph{fade out} (diminuição progressiva do volume até o volume zero),
-podemos tomar $a_0=2$ e $a_{\Lambda-1}=1$ e subtrair $1$ de toda a sequência $\{a_i'\}=\left\{2\left(\frac{1}{2}\right)^{\frac{i}{\Lambda-1}}-1\right\}$.
-Para um \emph{fade in} (acréscimo progressivo de volume a partir do volume zero), podemos
-tomar $a_0'=1$, $a_{\Lambda-1}'=2$ e subtrair $1$ de toda sequência: $\{a_i\}_0^{\Lambda-1}=\left\{2^{\frac{i}{\Lambda-1}}-1\right\}_0^{\Lambda-1}$.
-
-Para resultar em uma amplificação linear, basta usarmos uma sequência $\{a_i\}$ adequada:
+a percepção linear.
+A figura~\ref{fig:transicao} exibe as transições para diferentes valores de $\alpha$ e para a transição entre os valores $1$ e $2$ (se $1$ e $2$ relacionam amplitudes, há um ganho de $\approx 6dB$).
+
+Algum cuidado é necessário para lidar com $a=0$. Observe que
+na equação~\ref{seqAmp}, se $a_0=0$ haverá divisão por zero e
+se $a_{\Lambda-1}=0$, há uma multiplicação por zero. Ambos os casos
+tornam o procedimento inútil e provém do fato de que nenhum número diferente de zero pode ser representado como uma proporção com relação ao zero.
+Podemos resolver isso escolhendo um número suficientemente pequeno.
+Podemos aceitar $-120dB\;\Rightarrow a=10^{\frac{-120}{20}}=10^{-6}$ ou valores ainda menores como nosso volume mínimo no caso de um
+\emph{fade in} ($a_0=10^{-6}$) ou de um \emph{fade out} ($a_{\Lambda-1}=10^{-6}$).
+
+Para resultar em uma amplificação linear, mas não linear para a percepção, basta usarmos uma sequência $\{a_i\}$ adequada:
 
@@
 Aqui pode muito nos auxiliar a conversão de decibels para amplitude apresentada na equação ~\ref{ampDec}.
-Junto à equação \ref{transAmp}, resulta na sequinte equação para uma transição de $V_{dB}$ decibels:
+Junto à equação \ref{transAmp}, resulta na seguinte equação para uma transição de $V_{dB}$ decibels:
 
@@
 para o caso geral de variações de amplitude segundo a progressão geométrica.
+Notamos que a transição mais natural e percebida como linear acorre com $\alpha=1$ e que quanto maior o valor de $\alpha$ mais suave é a introdução do som.
+Vale assinalar que $\alpha>1$ resulta em transições de volume muitas vezes chamadas de \emph{slow fade} enquanto $\alpha<1$ resulta em \emph{fast fade}~\cite{fades}.
 
@@
 finita no tempo. Esta representação amostral é chamada
-de 'resposta ao impulso' $h_i$. Os filtros FIR são aplicados ao som
-digitalizado pela convolucão do som com sua 
-resposta ao impulso. Para nossos fins, esta
-convolução fica definida como\footnote{Convencionamos $t_i=0$ para as
-amostras não definidas de antemão.}:
+de 'resposta ao impulso' $h_i$. Os filtros FIR são aplicados temporalmente ao som
+digitalizado pela convolucão do som com a 
+resposta ao impulso do filtro\footnote{Pode-se aplicar o filtro do domínio espectral através da multiplicação das transformadas de Fourier de ambos o som e a resposta ao impulso, e então realizada a transformada inversa de Fourier do espectro resultante~\cite{Openheim}.}.
+Para nossos fins, esta
+convolução fica definida como:
 
@@
 
+Onde consideramos $t_i=0$ para as
+amostras não definidas de antemão.
 Ou seja, o som $\{t_i'\}$ resultante da convolução de $\{t_i\}$ com a resposta ao impulso $\{h_i\}$
 tem cada i-ésima amostra $t_i$ substituída pela soma de suas últimas $\Lambda_h$ amostras $\{t_{i-j}\}_{j=0}^{\Lambda_h-1}$
-multiplicadas pela resposta ao impulso $\{h_i\}_0^{\Lambda_h-1}$ em si. Isso é facilmente
+multiplicadas uma a uma pela resposta ao impulso $\{h_i\}_0^{\Lambda_h-1}$ em si. Isso é facilmente
 percebido pela visualização gráfica da convolução de sequências arbitrárias e dispomos
@@
 com uma reverberação que se assemelha à reverberação do ambiente 
-em que ocorreu a medição.
-
-Como exemplo de resposta ao impulso sintetizadas, pode-se
+em que ocorreu a medição~\cite{Cook}.
+
+Como exemplo de resposta ao impulso sintetizada, pode-se
 descrever um perfil espectral, realizar a transformada inversa
@@
 Ambas as possibilidades são exploradas
-por títulos em nossas referências, em especial recomendamos~\cite{JOSFM,smith} e na coleção de filtros da comunidade \emph{Music-DSP}, da Universiade de Columbia~\cite{music-dsp}.
+por títulos em nossas referências, em especial recomendamos~\cite{JOSFM,smith} e a coleção de filtros da comunidade \emph{Music-DSP}, da Universiade de Columbia~\cite{music-dsp}.
 Para uma exposição aprofundada do assunto, recomendamos~\cite{Openheim}.
@@
 $a_0$ e $b_1$ 
-através da variável intermediária: $x \in [0,1]$:
+através da variável intermediária: $x \in [e^{-\pi},1]$:
 
@@
 com frequência de corte $f_c \in (0,\frac{1}{2})$, é calculada através da variável
-intermediária $x \in [0,1]$:
+intermediária $x \in [e^{-\pi},1]$:
 
@@
 \begin{split}
-R & = 1 - 3BW \\
+R & = 1 - 3bw \\
 K & = \frac{1-2R\cos(2\pi f_c) + R^2}{2 - 2 \cos (2 \pi f_c)}
@@
 a_0 & =  1 - K \\
-a_1 & =  2(K-R)\cos (2\pi f) \\
+a_1 & =  2(K-R)\cos (2\pi f_c) \\
 a_2 & =  R^2-K \\
-b_1 & =  2R \cos (2\pi f) \\
+b_1 & =  2R \cos (2\pi f_c) \\
 b_2 & =  -R^2
@@
 a_0 & =  K \\
-a_1 & =  -2K\cos (2\pi f) \\
+a_1 & =  -2K\cos (2\pi f_c) \\
 a_2 & =  K \\
-b_1 & =  2R \cos (2\pi f) \\
+b_1 & =  2R \cos (2\pi f_c) \\
 b_2 & =  -R^2
@@
 são muitas vezes compreendidos como ruídos~\cite{Roederer}. Na música eletrônica (incluindo
-a eletroacústica), os ruídos possuem usos amplos, diversificados e comumente
+a eletroacústica e gêneros de pista de dança), os ruídos possuem usos amplos, diversificados e comumente
 idiomáticos~\cite{Cook}.
@@
 
-A frequência mínima $f_{\text{min}}$ pode ser escolhida com base no limite da audição, pois não escutamos como altura de um som uma frequência absoluta abaixo de $\approx\; 20Hz$.
+A frequência mínima $f_{\text{min}}$ pode ser escolhida com base no limite da audição, pois não escutamos como altura uma componente sonora cuja frequência esteja abaixo de $\approx\; 20Hz$.
 
@@
 \item O ruído marrom deve seu nome ao Robert Brown, que descreveu o movimento browniano.
-Embora esta origem seja um tanto díspar do que poderiamos considerar motivo para uma associaçãocom a cor marrom, o ruído sonoro ficou consagrado com este nome.
+Embora esta origem seja um tanto díspar do que poderíamos considerar motivo para uma associação com a cor marrom, o ruído sonoro ficou consagrado com este nome.
 De qualquer forma, é bastante comum declarar satisfatória a associação do ruído com a cor marrom, dados os ruídos branco e rosa mais estridentes e relacionados a cores mais intensas~\cite{marrom}.
@@
 \begin{equation}\label{vbrGamma}
-\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_i \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f \left ( \frac{f + \mu }{f} \right )^{t_i'}  \right \rfloor= \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f .
-2^{t_i'\frac{\nu}{12}}  \right \rfloor
+\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f \left ( \frac{f + \mu }{f} \right )^{t_j'}  \right \rfloor= \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f .
+2^{t_j'\frac{\nu}{12}}  \right \rfloor
 \end{equation}
@@
     \centering
-    \caption{Espectrograma de um som com vibrato senoidal de $3Hz$ e profundidade de uma oitava em uma dente de serra de $1000Hz$}
+    \caption{Espectrograma de um som com vibrato senoidal de $3Hz$ e profundidade de uma oitava em uma dente de serra de $1000Hz$ (considerada $f_a=44.1kHz$)}
         \includegraphics[width=\textwidth]{figuras/vibrato___}
@@
 As variáveis $\mu$ e $\nu$ quantificam a intensidade do vibrato: $\mu$ é uma medida direta da quantidade
-de Herz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação ($2\nu$ é o número de semitons entre os picos superioes e inferiores de oscilação da frequência do som $t_i$ causada pelo vibrato).
+de Herz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação ($2\nu$ é o número de semitons entre os picos superioes e inferiores de oscilação da frequência do som $\{t_i\}$ causada pelo vibrato).
 Note que $\nu=\log_{2}\frac{f+\mu}{f} $ é conveniente neste caso pois o aumento máximo de frequência
@@
 ser utilizada para gerar o som e o padrão de oscilação do vibrato, o que é tão crucial para os usos musicais quanto a
-frequência de oscilação e o desvido de altura envolvido\footnote{O desvio de altura
-é chamado profundidade do vibrato e é dado por conveniência em semitons ou cents.}.
-Estas qualidades não são praticáveis em instrumentos musicais tradicionais, introduzindo novidade nas possibilidades musicais.
+frequência de oscilação e o desvio de altura envolvido\footnote{O desvio de altura
+é chamado profundidade do vibrato e é geralmetne dado por conveniência em semitons ou cents.}.
+Estas qualidades não são praticáveis em instrumentos musicais tradicionais, introduzindo novidade nas possibilidades artísticas.
 
@@
 \begin{equation}\label{trA}
-a_i=10^{t_i' \frac{V_{dB}}{20}} = a_{\text{máx}}^{t_i'}
+a_i=10^{\frac{V_{dB}}{20}t_i' } = a_{\text{máx}}^{t_i'}
 \end{equation}
@@
 e profundidades ($\nu$ e $V_{dB}$) diferentes.
-Usadas em associação e isoladamente os tremolos e vibratos e com variações progressivas dos parâmetros\footnote{Vale lembrar que os tremolos e vibratos ocorrem muitas vezes juntos em instrumentos tradicionais e na voz.}.
+Usados em associação e isoladamente os tremolos e vibratos e com variações progressivas dos parâmetros\footnote{Vale lembrar que os tremolos e vibratos ocorrem muitas vezes juntos em instrumentos tradicionais e na voz.}.
 A peça desenvolve também uma comparação entre os vibratos e tremolos em escala logarítmica e em escala linear para uma apreciação qualitativa.
@@
 Algo interessante ocorre se aumentarmos progressivamente $f'$.
-Primeiramente o limiar de frequência
+Primeiramente a aproximação do limiar de frequência
 para audição do fenômeno sonoro como
@@
 original $T_i$ de formas diferentes para os tremolos e para os vibratos, são as 
-chamadas sínteses AM (\emph{Amplitude Modulation}) e FM (\emph{Frequency Modulation}).
+chamadas sínteses AM (\emph{Amplitude Modulation}) e FM (\emph{Frequency Modulation}), respectivamente.
 Estas são técnicas muito poderosas e conhecidas, com clássicas aplicações em sintetizadores
-históricos como o Yamaha XXX e com aplicações fora da música, como em telecomunicações para transmissão de informação via ondas eletromagnéticas (ex. radios AM e FM).
+históricos como o \emph{Yamaha DX7} e com aplicações fora da música, como em telecomunicações para transmissão de informação via ondas eletromagnéticas (ex. radios AM e FM).
 
@@
 \end{equation}
-é a função de Bessel~\cite{BesselCCRMA,JOSFM} que, na FM, especifica as amplitudes de cada componente.
-Note que, nestas equações, a variação de frequência não respeita a progressão geométrica de frequência que acompanha a percepção de altura.
+é a função de Bessel~\cite{BesselCCRMA,JOSFM} que, na FM, como podemos notar, especifica a amplitude de cada componente.
+Note que, nestas equações, a variação de frequência introduzida por $\{t_i'\}$ não respeita a progressão geométrica de frequência que acompanha a percepção de altura.
 Caso utilizemos as equações~\ref{vbrF} teríamos uma distribuição diferente dos harmônicos.
@@
 
-Ou seja, o som resultante é o original
+Ou seja, o som resultante é o som original
 e a reprodução de seu conteúdo espectral acima e abaixo da frequência
@@
 Observe que utilizamos novamente a variação na escala linear de amplitude.
-Dispomos no APÊNDICE~\ref{cap:fmam} uma exposição do espectro da AM se realizada na escala logarítmica de amplitude, que também perde o comportamento simples.
+Dispomos no APÊNDICE~\ref{cap:fmam} uma exposição do espectro da AM se realizada a oscilação na escala logarítmica de amplitude, que também perde o comportamento simples.
 
@@
 \begin{equation}\label{fmGamma}
-\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_i \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}(f+\mu .
-t_i') \right\rfloor
+\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}(f+\mu .
+t_j') \right\rfloor
 \end{equation}
@@
 \begin{equation}\label{fmT}
-T_i^{f,\, FM(f',\,\mu)}=\{ t_i^{f,\,FM(f',\,\mu)} \}_0^{\Lambda-1}=\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \}_0^{\Lambda-1}
+T_i^{f,\, FM(f',\,\mu)}=\left\{ t_i^{f,\,FM(f',\,\mu)} \right\}_0^{\Lambda-1}=\left\{\,\widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \,\right\}_0^{\Lambda-1}
 \end{equation}
@@
 
-Para realizar a AM, basta modular $\{t_i\}$ com $\{t_i'\}$ com as equações (lembrando que $\alpha=10^{\frac{V_{dB}}{20}}$):
+Para realizar a AM, basta modular $\{t_i\}$ com $\{t_i'\}$ através das equações (lembrando que $\alpha=10^{\frac{V_{dB}}{20}}$):
 
@@
 \begin{equation}\label{trT}
-T_i^{f,\,AM(f',\,\alpha)}=\{ t_i^{f,\,AM(f',\,\alpha)} \}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}=\{t_i . (1 + \alpha . t_i')    \}_0^{\Lambda-1}
+T_i^{f,\,AM(f',\,\alpha)}=\left\{ t_i^{f,\,AM(f',\,\alpha)} \right\}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}=\{t_i . (1 + \alpha . t_i')    \}_0^{\Lambda-1}
 \end{equation}
@@
 de oscilação do vibrato e do tremolo utilizada na equação~\ref{vbrGamma}. Já $\nu$ e $V_{dB}$ são as profundidades
-do vibrato e do tremolo, respectivamente. A montagem musical 
-\emph{tremolos, vibratos e a frequência} explora
+do vibrato e do tremolo, respectivamente. As funções $func_a$,
+$func_b$ e $func_c$ são arbitrárias e dependentes das intensões musicais. A montagem musical 
+\emph{Tremolos, vibratos e a frequência} explora
 recursos como este e variações da forma de onda da oscilação com vínculos, de modo a formar um \emph{idioma musical}\footnote{Veja na próxima sessão.}.
@@
 estabelecida pela convolução com impulsos na sequência musical \emph{trenzinho de caipiras impulsivos}.
+Os recursos explorados incluem a criação de amálgamas sonoros provenientes da síntese granular e esta montagem é já uma ponte para a próxima sessão.
+Veja especialmente a figura~\ref{fig:pulsoSubAgl}.
 
@@
 
-Um recurso considerado musicalmente forte pelo resultado que se aproxima dos instrumentos tradicionais é a filtragem dependente do tempo~\cite{Roederer}, no qual convém os filtros IIR, em que
-os valores do filtro podem ser modificados - com algum cuidado - ao longo de sua aplicação. Cascatas
+Um recurso considerado musicalmente forte pelo resultado que se aproxima dos instrumentos tradicionais é a filtragem dependente do tempo~\cite{Roederer}. Cascatas
 destes filtros podem realizar filtragens complexas e mais precisas. A montagem \emph{ruídos e faixas} explora
@@
 e trêmolo e suas características, equalizações, volume etc. Para o resultado final, estas versões do som
-inicial são então mixadas (ver equação \ref{eq:mixagem}). A peça \emph{chorus infantil} realiza chorus de formas
+inicial são então mixadas (ver equação \ref{eq:mixagem}). A peça \emph{Chorus infantil} realiza chorus de formas
 diferentes em sons diferentes.
@@
 
-A envoltória ADSR canônica é caracterizada por 4 parâmetros: duração do ataque (que o som demora para atingir seu volume máximo), duração do decaimento (segue ao ataque imediatamente), nível de volume de sustentação (em que o volume fica estável após o decaimento) e duração de soltura (após a sustentação, nesta duração o volume decai a zero).
-Observer que o tempo de sustentação não está especificado, isso depende da nota em si.
+A envoltória ADSR canônica é caracterizada por 4 parâmetros: duração do ataque (tempo que o som demora para atingir seu volume máximo), duração do decaimento (segue ao ataque imediatamente), nível de volume de sustentação (em que o volume fica estável após o decaimento) e duração de soltura (após a sustentação, nesta duração o volume decai a zero).
 Note que o tempo de sustentação não é especificado, pois é resultante da duração em si menos os tempos de ataque, decaimento e soltura.

### 2012-10-29 devEresultados.tex
feita geral ateh o ritmo

@@
 
-\begin{equation}
+\begin{equation}\label{eq:concatenacao}
 \{t_i\}_0^{\sum\Delta_k-1}=\{t_{l,i}\}_0^{\sum\Delta_k-1}, \;\; l\text{ menor inteiro } : \quad \Lambda_l > i -\sum_{j=0}^{l-1}\Lambda_j
@@
 \subsection{Organização de notas musicais em música}
-Seja $ S=\{T_{i,1},T_{i,2}...T_{i,N}\} =\bigcup_{j \in [1,N]} T_{i,j} $ uma sequência de $N$ eventos
+Seja $ S_j=\{T_{i,1},T_{i,2}...T_{i,N}\} =\bigcup_{j \in [1,N]} T_{i,j} $ uma sequência de $N$ eventos
 musicais. Para simplificar, sejam notas musicais estes eventos musicais.
@@
 para tornar estas estruturas interessantes e agradáveis.
+
+Os elementos de $S_j$ podem ser mixados (veja equação~\ref{eq:mixagem} e a figura~\ref{fig:mixagem}) formando acordes ou concatenados (veja equação~\ref{eq:concatenacao} e a figura~\ref{fig:concatenacao}) formando sequências melódicas.
 
@@
 no espectro audível, em que as frequências fundamentais possíveis
-estão separadas por intervalos múltiplos de $\varepsilon$\footnote{Este
-é um modelo idealizado, os intrumentos reais possuem desvios das frequências
-fundamentais como citadas}.
+estão separadas por intervalos múltiplos de $\varepsilon$.
+Esta precisão absoluta é possível somente em implementações
+computacionais, pois os intrumentos reais possuem desvios destas notas para melhor harmonia entre os harmônicos~\cite{Roederer}.
 
@@
 de temperamento igual, temos 5 divisões perfeitamente simétricas, consideradas
-escalas pela simetria interna e usos fáceis que disso resultam. Descrevamos
-as escalas como sequências de fatores aos quais $\varepsilon$ é elevado para resultar
-no fator a ser multiplicado por $f_0$:
+escalas pelas simetrias internas e usos fáceis que disso resultam. Descrevamos
+as escalas como sequências de fatores aos quais $\varepsilon$ é elevado, resultando
+nos fatores a serem multiplicados por $f_0$ para obtenção dos graus de cada escala:
 
@@
 \begin{split}
-\text{cromática} = E_i^c = \{e_i^c\}_0^{11} =  \{0,1,2,3,4,5,6,7,8,9,10,11\} = \{i\}_0^{11}\\
-\text{tons inteiros} = E_i^t = \{e_i^t\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
-\text{terças menores} = E_i^{tm} = \{e_i^{tm}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
-\text{terças maiores} = E_i^{tM} = \{e_i^{tM}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
-\text{trítonos} = E_i^{tt} = \{e_i^{tt}\}_0^{1} = \{ 0, 6 \} = \{6.i\}_0^1
+\text{cromática} & = E_i^c = \{e_i^c\}_0^{11} =  \{0,1,2,3,4,5,6,7,8,9,10,11\} = \{i\}_0^{11}\\
+\text{tons inteiros} & = E_i^t = \{e_i^t\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
+\text{terças menores} & = E_i^{tm} = \{e_i^{tm}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
+\text{terças maiores} & = E_i^{tM} = \{e_i^{tM}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
+\text{trítonos} & = E_i^{tt} = \{e_i^{tt}\}_0^{1} = \{ 0, 6 \} = \{6.i\}_0^1
 \end{split}
@@
 \begin{split}
-\text{menor natural} = \text{modo eólico} = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
-\text{menor harmônica} = E_i^{mh} = \{e_i^{mh}\}_0^6 = \{0,2,3,5,7,8,11\} \\
-\text{menor melódica} = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
-\text{modo lócrio} = E_i^{mlo} = \{e_i^{mlo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
-\text{maior} = \text{modo jônico} = E_i^M = \{e_i^M\}_0^6 = \{0,2,4,5,7,9,11\} \\
-\text{modo dórico} = E_i^{md} = \{e_i^{md}\}_0^6 = \{0,2,3,5,7,9,10\} \\
-\text{modo frígio} = E_i^{mf} = \{e_i^{mf}\}_0^6 = \{0,1,3,5,7,8,10\} \\
-\text{modo lídio} = E_i^{ml}=\{e_i^{ml}\}_0^6 = \{0,2,4,6,7,9,11\} \\
-\text{modo mixolídio} = E_i^{mmi} = \{e_i^{mmi}\}_0^6 = \{0,2,4,5,7,9,10\}
+\text{menor natural} = \text{modo eólico} & = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
+\text{menor harmônica} & = E_i^{mh} = \{e_i^{mh}\}_0^6 = \{0,2,3,5,7,8,11\} \\
+\text{menor melódica} & = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
+\text{modo lócrio} & = E_i^{mlo} = \{e_i^{mlo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
+\text{maior}  = \text{modo jônico} & = E_i^M = \{e_i^M\}_0^6 = \{0,2,4,5,7,9,11\} \\
+\text{modo dórico} & = E_i^{md} = \{e_i^{md}\}_0^6 = \{0,2,3,5,7,9,10\} \\
+\text{modo frígio} & = E_i^{mf} = \{e_i^{mf}\}_0^6 = \{0,1,3,5,7,8,10\} \\
+\text{modo lídio} & = E_i^{ml}=\{e_i^{ml}\}_0^6 = \{0,2,4,6,7,9,11\} \\
+\text{modo mixolídio} & = E_i^{mmi} = \{e_i^{mmi}\}_0^6 = \{0,2,4,5,7,9,10\}
 \end{split}
@@
 Várias outras escalas e sequências são facilmente representadas desta forma. Em especial,
-lembramos aqui das pentatônicas e os modos de transposição limitados de messiaen ~\ref{}.
+lembramos aqui das pentatônicas e os modos de transposição limitados de messiaen~\cite{Messiaen}.
 
@@
 resultando em intervalos que pouco se assemelham aos intervalos presentes na
-divisão de doze notas.}, podemos usar reais não inteiros
+divisão de doze notas~\cite{Wisnick}.},
+podemos usar reais não inteiros
 para descrever a sequência de alturas ou modificar o fator $\varepsilon$
@@
 por ao menos $50-63ms$. Para que a separação entre eles possa ser apreciada, ela deve ser ainda um pouco maior,
-por volta de $100ms$ ($10Hz$)(tem um resumo interessante na página 42 do \ref{microsound}). Podemos sumarizar 
+por volta de $100ms$ ($10Hz$) (tem um resumo interessante na página 42 do \ref{microsound}). Podemos sumarizar 
 a transição do domínio da apreciação

### 2012-10-30 devEresultados.tex
jah revisando ultima parte e dando as devidas incrementadas

@@
 musicais. Para simplificar, sejam notas musicais estes eventos musicais.
-Considerando $S$ como uma estrutura musical, existem diversas técnicas
+Considerando $S$ uma estrutura musical, existem diversas técnicas
 para tornar estas estruturas interessantes e agradáveis.
 
-Os elementos de $S_j$ podem ser mixados (veja equação~\ref{eq:mixagem} e a figura~\ref{fig:mixagem}) formando acordes ou concatenados (veja equação~\ref{eq:concatenacao} e a figura~\ref{fig:concatenacao}) formando sequências melódicas.
-
-\subsubsection{Afinação, escalas e acordes}
-Quando dobramos a frequência, temos uma oitava ascendente $f_2=2f_1$.
+Os elementos de $S_j$ podem ser mixados (veja equação~\ref{eq:mixagem} e a figura~\ref{fig:mixagem}) formando intervalos e acordes ou concatenados (veja equação~\ref{eq:concatenacao} e a figura~\ref{fig:concatenacao}) formando sequências melódicas.
+
+\subsubsection{Afinação, escalas, acordes e intervalos}
+Quando dobramos a frequência, temos uma oitava ascendente ($f=2f_0$).
 Doze semitons equidistantes para o ouvido formam uma oitava,
-portanto $f_2=2^{\frac{1}{12}}f_1$ forma um semitom ascendente.
-O fator $\varepsilon=2^{\frac{1}{12}}$ forma uma grade de notas
+portanto, se $f=2^{\frac{1}{12}}f_0$, temos um semitom entre $f_0$
+e $f$.
+O fator $\varepsilon=2^{\frac{1}{12}}$, equivalente a um semitom, forma uma grade de notas
 no espectro audível, em que as frequências fundamentais possíveis
 estão separadas por intervalos múltiplos de $\varepsilon$.
-Esta precisão absoluta é possível somente em implementações
-computacionais, pois os intrumentos reais possuem desvios destas notas para melhor harmonia entre os harmônicos~\cite{Roederer}.
+Esta precisão absoluta é característica de implementações
+computacionais, pois os intrumentos reais possuem desvios destas frequências para melhor compatibilizar os harmônicos
+de suas notas~\cite{Roederer}.
 
 Esta divisão da oitava em doze notas é o cânone da música ocidental clássica,
-além de usos cerimoniais/religiosos e étnicos. Nesta afinação, chamada cromática
+além de usos cerimoniais/religiosos e étnicos~\cite{Wisnick}.
+Os intervalos entre quaisquer duas notas da escala de 12 notas em que se subdivide a oitava possuem nomes específicos.
+Notados em proporções de $\varepsilon$ entre as notas (i.e.
+um semitom), estes intervalos são:
+
+\begin{equation}
+\begin{split}
+\text{uníssono} & = 0 \\
+\text{semitom ou segunda menor} & = 1 \\
+\text{tom, tom inteiro ou segunda maior} & = 2 \\
+\text{terça menor} & = 3 \\
+\text{terça maior} & = 4 \\
+\text{quarta justa} & = 5 \\
+\text{quarta aumentada, quinta diminuta ou trítono} & = 6 \\
+\text{quinta justa} & = 7 \\
+\text{sexta menor} & = 8 \\
+\text{sexta maior} & = 9 \\
+\text{sétima menor} & = 10 \\
+\text{sétima maior} & = 11 \\
+\text{oitava} & = 12
+\end{split}
+\end{equation}
+
+Nesta nomenclatura (padrão no Brasil há algumas décadas),
+um intervalo justo (uníssono, quarta, quinta e oitava), ou um intervalo maior (segunda maior, terça maior, sexta maior ou sétima maior), acrescido de um semitom, resulta em um intervalo aumentado.
+Um intervalo justo, ou um intervalo menor (segunda menor, terça menor, sexta menor ou sétima menor), decrescido de um semitom, resulta em um intervalo diminuto.
+Um intervalo maior, descrescido de um semitom,
+resulta em um intervalo menor.
+Um intervalo menor, acrescido de um semitom, resulta em um intervalo maior.
+Um intervalo aumentado, acrescido de um semitom, resulta em um intervalo 'mais que aumentado' e um intervalo diminuto, descrescido de um semitom, resulta em um intervalo 'mais que diminuto'\footnote{A
+ necessidade de da nomenclatura, notação e uso de intervalos aumentados/diminutos e mais que aumentados/diminutos é consequência do sistema tonal, em que os graus da escala (veja abaixo) são realmente notas diferentes, com funções e usos específicos.
+Assim, em uma escala de dó bemol maior, a tônica - primeiro grau - é dó bemol, não si, e a sensível - sétimo grau - é si bemol, não lá sustenido ou dó duplo bemol.}.
+
+Caso as notas sejam dispostas sequencialmente, a ordem das notas, primeiro a nota mais grave ou a mais aguda, resulta em um intervalo ascendente ou descendente, respectivamente. 
+Passando a nota mais grave para a oitava acima, ou a nota mais aguda uma oitava para baixo, invertemos o intervalo.
+Um pouco de exploração do sistema revela que um intervalo, somado à sua inversão, resulta 9.
+Um intervalo maior invertido resulta em um intervalo menor e vice-versa.
+Um intervalo aumentado invertido resulta diminuto e vice-versa, assim como o mais que aumentado resulta em mais que diminuto e vice-versa.
+Um intervalo justo invertido resulta igualmente justo.
+
+Um intervalo maior que a oitava é classificado como intervalo composto e leva o nome do intervalo entre as mesmas notas, mas na mesma oitava.
+
+Este arcabouço resume com exatidão a teoria dos intervalos musicais~\cite{Lacerda} e nos será útil adiante.
+Dispomos esta nomenclatura também por completude desta exposição, não há necessidade alguma de decorar estes nomes ou apreender estes detalhes todos enquanto não há uso efetivo.
+
+Na escala cromática
 de temperamento igual, temos 5 divisões perfeitamente simétricas, consideradas
 escalas pelas simetrias internas e usos fáceis que disso resultam. Descrevamos
-as escalas como sequências de fatores aos quais $\varepsilon$ é elevado, resultando
+as escalas como sequências de inteiros aos quais $\varepsilon$ é elevado, resultando
 nos fatores a serem multiplicados por $f_0$ para obtenção dos graus de cada escala:
@@
 
-As escalas diatonicas também são facilmente especificadas, assim como
+As escalas diatônicas também são facilmente especificadas, assim como
 as escalas tonais maior e menor:
@@
 \text{menor natural} = \text{modo eólico} & = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
-\text{menor harmônica} & = E_i^{mh} = \{e_i^{mh}\}_0^6 = \{0,2,3,5,7,8,11\} \\
-\text{menor melódica} & = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
 \text{modo lócrio} & = E_i^{mlo} = \{e_i^{mlo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
@@
 
+Observe que nas escalas diatônicas só estão presentes intervalos maiores, menores e justos dentre quaisquer duas notas escohidas.
+Única excessão para o trítono, que se apresenta como quarta aumenta ou quinta diminuta, dependendo da nota mais grave que se leva em consideração.
+
 Todas as escalas diatônicas (salvo as menores harmônica e melódica)
@@
 lembramos aqui das pentatônicas e os modos de transposição limitados de messiaen~\cite{Messiaen}.
+
+A escala menor se apresenta em duas formas adicionais, dependentes do uso, melódico ou harmônico:
+
+\begin{equation}
+\begin{split}
+\text{menor natural (igual acima)} & = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
+\text{menor harmônica} & = E_i^{mh} = \{e_i^{mh}\}_0^6 = \{0,2,3,5,7,8,11\} \\
+\text{menor melódica} & = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
+\end{split}
+\end{equation}
+
+e podem ser utilizadas livremente, embora tradicionamente se associem aos citados usos melódicos ou harmônicos.
+Observe o traçado ascendente e descendente da escala menor melódica.
+Isso é necessário para que se apresente a sensível (sétimo e último grau separado por um semitom da oitava) no trajeto ascendente, o que não é necessário quando descende, retomando a forma natural.
+Já a harmônica apresenta a sensível, mas não evita o intervalo de segunda aumentada entre o sexto e sétimo graus, pois não precisa contemplar trajetória melódica, apenas apresentar a sensível tão cara ao sistema tonal~\cite{Harmonia}.
 
@@
 
-Nestes casos, seguindo as mesmas restrições, basta acrescentar $10$ para ter o acorde
+Basta acrescentar $10$ para ter o acorde
 com sétima menor ou $11$ para resultar em sétima maior\footnote{As sétimas
-menor e maior são intervalos de $10$ e $11$ semitons respectivamente.}. Para realizar inversões e posições abertas,
-basta contar o número de semitons de cada nota a partir da nota mais grave.
+menor e maior são intervalos de $10$ e $11$ semitons respectivamente.}. As inversões e posições abertas
+podem ser obtidas contando o número de semitons de cada nota a partir da nota mais grave.
 
@@
 é proposta na forma da divisão da oitava em $53$ notas, de forma
-que $\varepsilon=2^{\frac{1}{53}}$. As sequências de alturas
+que $\varepsilon=2^{\frac{1}{53}}$~\cite{microtonalidade}. As sequências de alturas
 continuam sendo compostas por inteiros com o novo $\varepsilon$.
@@
 \subsubsection{Rítmo}
-A noção rítmica é dependente de eventos separados por durações. Estes eventos são discerníveis se separados
+A noção rítmica é dependente de eventos separados por durações~\cite{rythm}. Estes eventos são discerníveis individualmente se separados
 por ao menos $50-63ms$. Para que a separação entre eles possa ser apreciada, ela deve ser ainda um pouco maior,
-por volta de $100ms$ ($10Hz$) (tem um resumo interessante na página 42 do \ref{microsound}). Podemos sumarizar 
+por volta de $100ms$~\cite{microsound}. Podemos sumarizar 
 a transição do domínio da apreciação
-de durações como alturas para a apreciação em rítmo da seguinte forma (~\ref{alfaix, microsound}):
+de durações como alturas para a apreciação em rítmo da seguinte forma ~\cite{Alfaix, microsound}:
 
@@
 
-A banda de durações marcada como transição está minimizada pois os limites não são bem definidos.
-A duração específica em que começamos a perceber nitidademente uma frequência fundamental ou uma separação entre as ocorrências é dependente da pessoa e de características do som como conteúdo espectral e intensidade.
+A banda de durações marcada como transição está minimizada pois os limites não são bem definidos: a duração específica em que começamos a perceber nitidademente uma frequência fundamental ou uma separação entre as ocorrências é dependente da pessoa e de características do som como conteúdo espectral e intensidade~\cite{microsound,Roederer}.
 
 A métrica rítmica costuma se basear em uma duração básica chamada pulso. O pulso tipicamente
-compreende durações entre $0,25-1,5s$ (respectivamente $240$ e $40BPM$). Na educação musical e estudos cognitivistas,
+compreende durações entre $0.25-1.5s$ (respectivamente $240$ e $40BPM$). Na educação musical e estudos cognitivistas,
 costuma-se associar esta gama de frequências de pulsação às durações entre batidas 
-do coração ou entre os passos ao caminhar.
+do coração ou entre os passos ao caminhar~\cite{Lacerda,Roederer}.
 
@@
 sequenciais no tempo ocorrem com fatores 2, 4 e 8, depois 3, 6 (dois grupos de 3 ou 3 grupos de 2) e 9 (3 grupos de 3). Por 
-último os primos 5 e 7, completando 1-9 (~\ref{jocelyn}, ~\ref{favre}).
-Outras métricas são incidentes (divisões ou agrupamentos em 13, 17, etc), mas não são tão usuais.
-As métricas, por mais complexas que pareçam, costumam ser decompostas em divisões como estas (inclusive as de fator 5 e 7).}.
+último os primos 5 e 7, completando 1-9.
+Outras métricas são incidentes (divisões ou agrupamentos em 13, 17, etc), mas não são tão usuais e ocorrem principalmente em música experimental e erudita do século XX para frente.
+As métricas, por mais complexas que pareçam, costumam ser decompostas em divisões como estas~\cite{Gramani,Roederer}.}.
 Uma exposição esquemática está na figura ~\ref{fig:pulsoSubAgl}.
@@
     \centering
-    \caption{Divisões e aglomerações do pulso musical para estabelecimento de métrica\footnote{Ao lado esquerdo temos as divisões da semínima estabelecida como pulso.
-Ao lado direito temos exemplos de fórmulas de compasso que especificam as mesmas métricas, mas na escala produzida pela algomeração e ocorrências do pulso musical.}}
+    \caption{Divisões e aglomerações do pulso musical para estabelecimento de métrica.
+Ao lado esquerdo temos as divisões da semínima estabelecida como pulso.
+Ao lado direito temos exemplos de fórmulas de compasso que especificam as mesmas métricas, mas na escala produzida pelas ocorrências e aglomerações do pulso musical.}
         \includegraphics[width=\textwidth]{figuras/metricaMusical}
@@
 As relações duais (compassos simples e divisões binárias) costumam a ocorrer em rítmos de dança
-e ocasiões festivas e são chamadas imperfeitas. As relações triádicas ocorrem
+e ocasiões festivas e são chamadas imperfeitas. As relações triádicas
 incidem mais na música ritualística e relacionada ao sagrado
@@
 
-As divisões mais fortes (acentuadas) são as que consistem nas cabeças dos tempos. Nas divisões binárias (2, 4 e 8 dos casos considerados aqui),
+As unidades mais fortes (acentuadas) são as que consistem nas cabeças dos tempos. Nas divisões binárias (2, 4 e 8 dos casos considerados aqui),
 as unidades consideradas fortes se revesam com as fracas
 (e.g. a divisão em 4 resulta em forte, fraco, (meio) forte, fraco).
-Nas divisões ternárias (3, 6 e 9 dos casos considerados neste escrito) a
+Nas divisões ternárias (3, 6 e 9 dos casos considerados neste escrito)
 à unidade forte (primeira) se sucedem 2 unidades fracas (e.g. a divisão em 3 resulta em forte, fraco fraco)\footnote{A divisão em 6 é considerada composta
@@
  Uma divisão binária que sofre então uma divisão ternária
- resulta em duas unidades divididas em três unidades cada: forte (subdividifo em forte fraco, fraco) e fraco (subdividido em forte, fraco, fraco).
+ resulta em duas unidades divididas em três unidades cada: forte (subdividido em forte fraco, fraco) e fraco (subdividido em forte, fraco, fraco).
 A outra forma de ocorrer a divisão em 6 é através de 
 uma divisão ternária que sofre então uma divisão binária, resultando em:
-uma unidade forte (subdividido em forte e fraco) e duas unidades fracas (subdivididas em forte e fraco).}.
-
-A acentuação em tempos fracos é chamada contratempo, as notas que se iniciam em tempos fracos e cuja duração percorre tempos fortes são as chamadas síncopas.
+uma unidade forte (subdividido em forte e fraco) e duas unidades fracas (subdivididas em forte e fraco cada).}.
+
+A acentuação em tempos fracos é chamada contratempo, as notas que se iniciam em tempos fracos e cuja duração percorrem tempos fortes são as chamadas síncopas.
 
@@
 Nos casos mais comportados, as notas ocorrem exatamente nestas divisões, com maior incidência para ataques nos tempos fortes.
+Em casos extremos, não se pode perceber a métrica~\cite{Roederer}.
+Variações (usualmente) bastante pequenas na grade compõem o que apreciamos como interpretação musical e diferenças entre estilos, junto a variações de intensidades, sobreposições de notas e timbre~\cite{Cook}.
 
@@
 é a primeira subdivisão do pulso, o nível $j=1$ como a primeira algomeração dos pulsos e assim por diante.
-Assim, seja $P_i^j$ a $i$-ésima unidade de 
-pulsos no nível $j$ de agrupamento\footnote{Assim,
-$P^0_{10}$ é o décimo pulso, $P^{1}_3$ é a terceira unidade de agrupamento de pulsos,
-$P^{-1}_2$ é a segunda unidade da subdivisão do pulso.}.
+
+Assim, $P_i^j$ é a $i$-ésima unidade de 
+pulsos no nível $j$ de agrupamento:
+$P^0_{10}$ é o décimo pulso, $P^{1}_3$ é a terceira unidade de agrupamento de pulsos (possivelmente terceiro compasso),
+$P^{-1}_2$ é a segunda unidade da subdivisão do pulso.
 
@@
 os limites de $j$ pois as divisões do pulso devem resultar em durações apreciáveis
-como rítmo e também as junções do pulso deve resultar algo apreciável como, no máximo
-do nosso escopo, música. Dito de outra forma: a duração de $P^{min(j)}_i$, para $\forall \; i$
+como rítmo e também as junções do pulso devem resultar algo apreciável como, no máximo
+do nosso escopo, uma música ou um conjunto coeso de músicas. Dito de outra forma: a duração de $P^{min(j)}_i$, para $\forall \; i$
 é maior que $50ms$ e as durações de $P^{\text{máx}(j)}_i$ \emph{somadas}  em $\forall i$
-devem resultar em uma duração menor do que alguns minutos ou, no máximo poucas horas.
+devem resultar em uma duração menor do que alguns minutos ou, no máximo, poucas horas.
 
@@
 
-Podemos utilizar a notação
-
-\begin{equation}
-P^{ \{ j_k \} }_{\left \{ \{ i_{l} \}_0^{\alpha_{k}-1} \right \} } \quad \text{onde}\; \alpha_k \; \text{é o número de partições do nível} j_k
-\end{equation}
-
-para especificar qualquer unidade da subdivisão da métrica, com níveis $j_k$ e unidades $i_l$\footnote{Como um exemplo, podemos escrever $P^{-1,0,2}_{3,2,1}$  como a terceira subdivisão $P^{-1}_3$ do segundo
-pulso $P^0_2$ do terceiro aglomerado $P^2_2$ de aglomerado de pulsos (perceba o termo $P^1_1$ omitido por simplicidade).}.
-
-A cada unidade ou conjunto de unidades $P_i^j$ pode ser associada uma sequência de amostras temporais $T_i$
-que formam uma nota musical.
-
-Usualmente lidamos com a distribuição rítmica dos eventos musicais com este arcabouço em mente. Assim como acontece com outras características
+Para especificar qualquer unidade (nota), de uma
+dada sequência musical que tenha métrica,
+podemos utilizar a notação:
+
+\begin{equation}
+P^{ \{ j_k \} }_{ \{ i_{k} \}}
+\end{equation}
+
+em que $j_k$ é o nível de aglomeração e $i_k$ é a ordem
+da unidade em si.
+
+Como um exemplo, podemos escrever $P^{-1,0,2}_{3,2,1}$  como a terceira subdivisão $P^{-1}_3$ do segundo
+pulso $P^0_2$ do terceiro aglomerado de aglomerado de pulsos $P^2_2$ (perceba o termo $P^1_1$ omitido por simplicidade).
+
+Cada unidade ou conjunto de unidades $P_i^j$ pode ser associada a uma sequência de amostras temporais $T_i$ que formam uma nota musical.
+
+Assim como acontece com outras características
 musicais, lidamos com a estrutura rítmica sem fórmulas pre-estabelecidas, deixando procedimentos sistemáticos para quando é pertinente.
@@
 rítmicas (divisões e aglomerações simples, compostas e complexas) é 
-natural pensar em como realizar estas estruturas de forma musical.
+natural pensar em como realizar estas estruturas de forma que tenha coesão e sentido~\cite{Boulez}.
 
@@
 Os arcos melódicos são compostos por duas sequências convergentes: 
-uma que atinge um ápice (chamado literalmente de clímax pela teoria musical canônica) e 
+uma que atinge um ápice (chamado literalmente de clímax pela teoria musical tradicional) e 
 outra que, de forma paradigmática, volta do ápice à região da nota de partida.
+Canonicamente, distinguimos entre arcos cujos clímax estão: no começo, no meio, no final, na primeira metade e na segunda metade~\cite{Schoenberg}.
+Dispomos esquematicamente a disposição destes clímax na figura~\ref{fig:climax}.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Distinções canônicas do clímax musical em uma melodia ou arcos em outras escalas}
+        \includegraphics[width=\textwidth]{figuras/climax}
+        \label{fig:climax}
+\end{figure}
 
@@
 Pensada segundo conceitos musicais, o clímax está exatamente no meio da sequência.
-Canonicamente, distinguimos também entre arcos cujos clímax estão deslocados do meio para o final 
-e para o começo (e casos extremos disso, em que os climax estão no extato começo ou final)~\ref{schoenbergComp}.
-Por razões ilustrativas, apresentamos no APÊNDICE XX alguns exemplos 
-destas estruturas em casos da história da música.
+Podemos modificar isso fazendo com que alguma das metades seja expandida ou diminuída ou substituindo alguma das metades por outra sequência de tamanho diferente.
+É atraente este recurso pois contamos com toda a teoria de sequências, já bastante desenvolvida e estabelecida.
+Comparadas as sequências estudadas na teoria das sequências e as utilizadas na música, podemos concluir que existe um interesse especial por sequências relativamente curtas comparadas com as encontradas na teoria matemática~\cite{Guidorizzo,Schoenberg}.
 
@@
 Neste caso, faremos uma exposição específica sobre permutações dada
-a solidez desta abordagem segundo princípios musicais, matemáticos e
-filosóficos\footnote{De forma bastante resumida, apontamos o
+a completude desta abordagem segundo princípios musicais, matemáticos e
+filosóficos\footnote{De
+forma bastante resumida, apontamos o
 entendimento filosófico de que o pensamento humano se baseia
-em reconhecimentos de semelhanças e diferenças dentre os estímulos
-e objetos~\ref{deleuze}. Portanto, segundo este entendimento,
-as simetrias se encontram no cerne do processo cognitivo. Matematicamente,
+na percepção de semelhanças e diferenças dentre os estímulos
+e objetos~\cite{Deleuze}. Portanto, segundo este entendimento,
+as simetrias se encontram no cerne do processo cognitivo pois conceitualmente são as equivalências em si. Matematicamente,
 as simetrias são representadas por grupos algébricos e os grupos finitos, por sua vez,
@@
 Diretamente na música encontramos permutações de forma bastante explícita, inclusive
-na análise de música étnica~\ref{permMusic}.
-Na tradição musical a permutação está presente em diversas técnicas~\ref{zamacois} e
+na análise de música étnica~\cite{permMusic}.
+Na tradição musical a permutação está presente em diversas técnicas~\cite{Zamacois} e
 apontamos em especial a técnica \emph{Change Ringing} que consiste essencialmente
-em padrões de permutações dentre um fixado número de notas~\ref{change}.}.
+em padrões de permutações dentre um fixado número de notas~\cite{change}.} e a consequente realização de dois trabalhos~\cite{figgusOriginal, figgusEspacializacao}.
 
 Toda permutação pode ser operada com outra permutação, resultando na permutação
-que representa a modificação de elementos fixados se aplicada uma permutação
+que representa a ordenação que resulta se aplicada uma permutação
 depois a outra:
@@
 O menor valor de $n$ é chamado de \emph{ordem} da permutação $p$. O elemento $e$ é chamado de elemento neutro e é
-a permutação que leva uma ordenação para ela mesma.
+a permutação que leva uma ordenação para ela mesma. Aqui já se pode notar o comportamento cíclico causado pela aplicação sucessiva da mesma permutação.
 
 A utilização de permutações na música pode ser resumida da seguinte forma:
-seja $L_i$ uma lista de eventos musicais (para simplificar, considere
-estes eventos musicais como sendo notas musicais) e $p$ uma permutação.
-$L_i'=p(L_i)$ consiste nos mesmos elementos de $L_i$ com a ordem entre eles
+seja $L_i$ uma sequência de eventos musicais (para simplificar, considere
+estes eventos como sendo notas musicais) e $p$ uma permutação.
+$L_i'=p(L_i)$ consiste nos mesmos elementos de $L_i$ com a ordenação
 modificada.
@@
 ao final de ~\ref{notaBasica}. A nova sequência $L_i'=p^f(L_i)$ consiste nas mesmas
-notas musicais, na mesma ordem, com as frequências fundamentais permutadas segundo
+notas musicais, na mesma ordem e com as mesmas características, com as frequências fundamentais permutadas segundo
 o padrão que $p^f$ apresenta.
@@
 Para simplificar, considere novamente os eventos de $L_i$ 
-como notas musicais. Assim, se $i$ é vai de $0$ a $n>3$, a cada compasso
+como notas musicais. Assim, se $i$ é vai de $0$ a $n$, e $n>4$, a cada compasso
 de $4$ notas podemos executar somente as primeiras $4$ notas. As outras notas de
-$L_i$ podem ser incidentes nos compassos em que as permutações aloque
+$L_i$ podem ser incidentes nos compassos em que as permutações aloquem
 estas notas para as primeiras quatro notas de $L_i$.
 
-Nada impede que tenhamos permutações operando no mesmo conjunto, cada uma destas permutações
-$p_i$
-possui: dimensões das notas em que opera (frequência, duração, fades, intensidade, etc),
-modo de incidência (opera em $L_i$ a cada uma ou duas utilizações de $L_i$).
+Podemos dispor permutações que operem no mesmo conjunto, cada uma destas permutações
+$p_i$ possui: dimensões das notas em que opera (frequência, duração, fades, intensidade, etc), modo de incidência (opera em $L_i$ a cada uma ou duas utilizações de $L_i$), período de aplicação (a cada quantas consultas é aplicada a permutação).
 Na realização das notas de $L_i$, uma solução fácil e coerente é executar as primeiras $n$ notas\footnote{A execução de notas disjuntas de $L_i$ equivale a modificar a permutação e executar as primeiras notas.}.
 
-Esta abordagem resultou em alguns trabalhos ~\ref{figgusOriginal} e implementações computacionais.
-Descrevemos um uso coerente desta técnica abaixo e apresentamos no APÊNCIDE XX a implementação computacional disponibilizada em ~\ref{repoFiggus}
+Esta abordagem resultou em alguns trabalhos~\cite{figgusOriginal,figgusEspacializacao} e implementações computacionais~\cite{repoFIGGUS,repoFIGGUSold}.
+Descrevemos um uso coerente desta técnica abaixo e apresentamos no APÊNCIDE~\ref{cap:FIGGUScode} a implementação computacional disponibilizada em~\cite{repoFIGGUS}
 

### 2012-10-30 scripts/climaxMel.py

@@ -1,0 +1,26 @@
+#-*- coding: utf8 -*-
+import numpy as n, pylab as p
+
+
+c1=n.hstack(( n.linspace(0,1,1000,endpoint=False),n.linspace(1,0,1000)    )) # simetrico
+c2=n.hstack(( n.linspace(0,1,500,endpoint=False), n.linspace(1,0,1500)    ))
+c3=n.hstack(( n.linspace(0,1,1500,endpoint=False),n.linspace(1,0,500)    ))
+c4=n.linspace(1,0,2000)
+c5=n.linspace(0,1,2000)
+
+p.plot(c1,label=u"clímax no meio",lw=5)
+p.plot(c2,label=u"clímax na primeira metade",lw=5)
+p.plot(c3,label=u"clímax na segunda metade", lw=5)
+p.plot(c4,label=u"clímax no começo",         lw=5)
+p.plot(c5,label=u"clímax no fim",            lw=5)
+p.legend(loc="lower center",prop={'size':22})
+
+p.xlim(-50,2050)
+p.ylim(-.18,1.02)
+
+p.xticks((),())
+p.yticks((),())
+
+
+p.show()
+

### 2012-10-30 devEresultados.tex
andando

@@
 
-Toda permutação pode ser operada com outra permutação, resultando na permutação
-que representa a ordenação que resulta se aplicada uma permutação
-depois a outra:
-
-\begin{equation}
-p_1 \bullet p_2 = p_3, \quad \forall \; p_1, p_2 \; \text{permutações finitas}
-\end{equation}
-
-Uma permutação finita aplicada sucessivamente um número suficiente de vezes
+Qualquer conjunto de permutações podem ser consideradas geradoras de grupos algébricos. Assim, vale citarmos as propriedades que definem um grupo $G$:
+
+\begin{equation}
+\begin{split}
+\forall p_1,p_2 \in G \Rightarrow p_1 \bullet p_2 & = p_3 \in G\quad\text{(propriedade de fechamento)} \\
+\forall p_1,p_2,p_3 \in G \Rightarrow (p_1\bullet p_2)\bullet p_3 & = p_1\bullet (p_2\bullet p_3)\quad\text{(propriedade da associatividade)} \\
+\exists e \in G : p \bullet e & = e \bullet p \forall p \in G\quad\text{(existência do elemento neutro)} \\
+\forall p \in G \exists p^{-1} : p\bullet p^{-1} & =p^{-1}\bullet p = e\quad\text{(existência do inverso)}
+\end{split}
+\end{equation}
+
+A primeira destas propriedades nos é de particular importância, pois concluímos que toda permutação pode ser operada com outra permutação.
+De fato, podemos aplicar uma permutação, depois outra e, se compararmos com a ordenação inicial, temos uma permutação resultante.
+As outras propriedades resultam triviais para nosso caso.
+
+Alguns teoremas nos são caros.
+Por exemplo, segue imediatamente que, em um grupo finito, todo elemento operado consigo mesmo um número de vezes resulta no elemento neutro.
+O menor número de vezes necessário para atingir o elemento neutro é chamado de ordem do elemento.
+Assim, uma permutação finita aplicada sucessivamente um número suficiente de vezes
 resulta na ordenação inicial dos elementos:
@@
 
-O menor valor de $n$ é chamado de \emph{ordem} da permutação $p$. O elemento $e$ é chamado de elemento neutro e é
+O menor valor de $n$ é chamado de \emph{ordem} da permutação $p$. O elemento $e$ é o elemento neutro e é
 a permutação que leva uma ordenação para ela mesma. Aqui já se pode notar o comportamento cíclico causado pela aplicação sucessiva da mesma permutação.

### 2012-10-30 devEresultados.tex
andando

--- 2012-10-30 devEresultados.tex
@@
 
-\subsubsection{Afinação, escalas, acordes e intervalos}
+\subsubsection{Afinação, intervalos, escalas e acordes}
 Quando dobramos a frequência, temos uma oitava ascendente ($f=2f_0$).
@@
 Notados em proporções de $\varepsilon$ entre as notas (i.e.
-um semitom), estes intervalos são:
+um semitom), estes intervalos são, junto a suas notações abreviadas usuais:
 
@@
 \begin{split}
-\text{uníssono} & = 0 \\
-\text{semitom ou segunda menor} & = 1 \\
-\text{tom, tom inteiro ou segunda maior} & = 2 \\
-\text{terça menor} & = 3 \\
-\text{terça maior} & = 4 \\
-\text{quarta justa} & = 5 \\
-\text{quarta aumentada, quinta diminuta ou trítono} & = 6 \\
-\text{quinta justa} & = 7 \\
-\text{sexta menor} & = 8 \\
-\text{sexta maior} & = 9 \\
-\text{sétima menor} & = 10 \\
-\text{sétima maior} & = 11 \\
-\text{oitava} & = 12
+\text{uníssono} = 1J & = 0 \\
+\text{semitom ou segunda menor} =2m & = 1 \\
+\text{tom, tom inteiro ou segunda maior} =2M & = 2 \\
+\text{terça menor} = 3m & = 3 \\
+\text{terça maior} = 3M & = 4 \\
+\text{quarta justa} = 4J & = 5 \\
+\text{quarta aumentada, quinta diminuta ou trítono} = 4aum = 5dim = Tri & = 6 \\
+\text{quinta justa} = 5J & = 7 \\
+\text{sexta menor} = 6m & = 8 \\
+\text{sexta maior} = 6M & = 9 \\
+\text{sétima menor} = 7m & = 10 \\
+\text{sétima maior} = 7M & = 11 \\
+\text{oitava} = 8J & = 12
 \end{split}
@@
 Um intervalo aumentado, acrescido de um semitom, resulta em um intervalo 'mais que aumentado' e um intervalo diminuto, descrescido de um semitom, resulta em um intervalo 'mais que diminuto'\footnote{A
- necessidade de da nomenclatura, notação e uso de intervalos aumentados/diminutos e mais que aumentados/diminutos é consequência do sistema tonal, em que os graus da escala (veja abaixo) são realmente notas diferentes, com funções e usos específicos.
+ necessidade da nomenclatura, notação e uso de intervalos aumentados/diminutos e mais que aumentados/diminutos é consequência do sistema tonal, em que os graus da escala (veja abaixo) são realmente notas diferentes, com funções e usos específicos.
 Assim, em uma escala de dó bemol maior, a tônica - primeiro grau - é dó bemol, não si, e a sensível - sétimo grau - é si bemol, não lá sustenido ou dó duplo bemol.}.
 
-Caso as notas sejam dispostas sequencialmente, a ordem das notas, primeiro a nota mais grave ou a mais aguda, resulta em um intervalo ascendente ou descendente, respectivamente. 
+Caso as notas sejam dispostas sequencialmente, o intervalo é dito melódico.
+A ordem das notas, primeiro a nota mais grave ou a mais aguda, resulta em um intervalo ascendente ou descendente, respectivamente.
 Passando a nota mais grave para a oitava acima, ou a nota mais aguda uma oitava para baixo, invertemos o intervalo.
@@
 
+Caso as notas soem simultaneamente, o intervalo é dito harmônico.
+Os intervalos de uníssono, oitava, quinta justa, quarta justa, terças e sextas são considerados {\bf consonantes}.
+As segundas, sétimas e trítonos são considerados {\bf dissonantes}.
+Esta classificação não é absoluta.
+A sétima menor por vezes é considerada consonante, assim como algumas culturas entoam o trítono como intervalo consonante~\cite{Roederer,Wisnick}.
+
 Um intervalo maior que a oitava é classificado como intervalo composto e leva o nome do intervalo entre as mesmas notas, mas na mesma oitava.
@@
 Dispomos esta nomenclatura também por completude desta exposição, não há necessidade alguma de decorar estes nomes ou apreender estes detalhes todos enquanto não há uso efetivo.
+A montagem \emph{Intervalos entre alturas} explora estes intervalos de formas isoladas e de formas diversas.
 
@@
 repetidas de forma eficiênte e sustentada. Discutiremos mais sobre simetrias e voltaremos
-a estas divisões da oitava por igual.
+a estas divisões da oitava por igual. A montagem \emph{Cristais} expõe cada uma destas escalas tanto melódica quanto harmonicamente.
 
@@
 
-e podem ser utilizadas livremente, embora tradicionamente se associem aos citados usos melódicos ou harmônicos.
+e podem ser utilizadas livremente, embora tradicionamente se associem aos usos melódicos ou harmônicos sugeridos pelos seus nomes.
 Observe o traçado ascendente e descendente da escala menor melódica.
@@
 A ocorrência simultânea de notas é observada musicalmente em grande medida
-através dos acordes. Listamos os acordes triádicos sem inversão e na forma fechada,
+através dos acordes.
+Destes, a base na musica tonal são as tríades, que consistem em duas terças sucessivas.
+Listamos os acordes triádicos sem inversão e na forma fechada\footnote{Um acorde invertido é aquele que apresenta no grave outra nota que não a fundamental.
+A posição fechada é aquela em que não cabe nota alguma do acorde entre quaisquer duas notas consecutivas~\cite{Lacerda}.},
 com a fundamental em $0$:
@@
 
-Basta acrescentar $10$ para ter o acorde
-com sétima menor ou $11$ para resultar em sétima maior\footnote{As sétimas
-menor e maior são intervalos de $10$ e $11$ semitons respectivamente.}. As inversões e posições abertas
+Basta acrescentar $10$ ao final para ter o acorde (uma tétrade)
+com sétima menor ou $11$ para resultar em sétima maior. As inversões e posições abertas
 podem ser obtidas contando o número de semitons de cada nota a partir da nota mais grave.
@@
 
+\subsubsection{Rudimentos de harmonia}
+
+Abordaremos de forma bastante resumida e voltada para nossas aplicações
+computacionais a harmonia tonal. Para harmonia modal e atonal recomendamos~\cite{harmEXT}.
+
+Na chamada harmonia tradicional, as escalas maior e menor são usadas
+de base para tríades feitas em cada grau da escala com duas terças consecutivas.
+É conveniente, assim, notar os graus da escala pelos números da seguinte forma: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$.
+
+Assim, as tríades formadas em cada grau é de fácil notação $\hat{1},\hat{3},\hat{5}$ é o acorde de primeiro grau, formado pelo primeiro grau da escala e central para uma música tonal.
+Em segundo lugar temos o acorde do quinto grau $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sustenido no caso da escala menor) e em terceiro o do quarto grau $\hat{4},\hat{6},\hat{1}$.
+Depois consideramos as outras tríades todas.
+Ao longo da história, formas diferentes de lidar com estas tríades foram desenvolvidas.
+Estas convenções e técnicas apresentadas ao longo da história é o objeto de estudo da harmonia tradicional~\cite{Harmonia}.
+
+A chamada harmonia funcional atribui funções a estes três acordes centrais e busca compreender seus usos através destas funções.
+O acorde formado sobre o primeiro grau é chamado de acorde de tônica e tem a função de manter um centro e um "chão" na música.
+O formado sobre o quinto grau é chamado de acorde de dominante e tem a função de tender à tônica, direcionar a música para ela.
+A tríade formada sobre o quarto grau é chamada de subdominante e tem a função de se distanciar da tônica.
+
+A estes três acordes, são associadas as outras tríades.
+Na escala maior, a associada relativa (tônica relativa, subdominante relativa e dominante relativa) é a tríade formada uma terça abaixo e a associada anti-relativa (tônica anti-relativa, subdominante anti-relativa e a dominante anti-relativa) é a tríade formada na terça acima.
+Na escala menor ocorre o mesmo, mas a tríade a uma terça abaixo é chamada anti-relativa e a tríade a uma terça acima é chamada de relativa.
+As exatas funções e efeitos musicais destes acordes é motivo de bastante controvérsia.
+Dispomos da tabela~\ref{tab:harmonia} esta relação entre as tríades formadas em cada grau da escala.
+
+\begin{table}[htpq!]
+\centering
+\begin{tabular}{l | c | r}
+relativa & acorde principal da função & antirelativa \\\hline\hline
+$\hat{6},\hat{1},\hat{2}$ & tônica:       $\hat{1},\hat{3},\hat{5}$ & $\hat{3}, \hat{5},      \hat{7}$ \\
+$\hat{3},\hat{5},\hat{7}$ & dominante:    $\hat{5},\hat{7},\hat{2}$ & [$\hat{7},\hat{2},\hat{4}$] \\
+$\hat{2},\hat{4},\hat{6}$ & subdominante: $\hat{4},\hat{6},\hat{1}$ & $\hat{6},\hat{1},       \hat{3}$
+\end{tabular}
+\caption{Resumo das funções harmônicas tonais para escala maior}
+\label{tab:harmonia}
+\end{table}
+
+Na escala menor, a relativa é a tríade iniciada uma terça acima e a antirelativa uma terça abaixo,
+ou seja, se invertem com relação à tabela~\ref{tab:harmonia}. Outro detalhe destas tríades em modo
+menor é a alteração do $\hat{7}$ um semitom acima para que fique somente a um semitom da tônica,
+permitindo a dominante de fato (que deve tender à tônica). Assim, a dominante é sempre maior, tanto
+em escalas
+maior ou menor.
+
+A alteração do sétimo grau
+$\hat{7}$ para atender a esta necessidade tonal dá origem a diferentes versões sobre
+a definição da dominante relativa ou antirelativa numa escala menor e recomendamos,
+para maiores detalhes, alguns títulos de nossa bibliografia~\cite{Koellheuteur,Harmonia,Schachter,harmFunc}.
+
+Voltando à escala maior, a dominante anti-relativa forma um acorde diminuto, de forma que geralmente se considera esta tríade uma 'dominante com sétima e sem fundamental'~\cite{Koellheuteur} e não se considera nenhuma dominante anti-relativa.
+Outra possibilidade encontrada na literatura é a elevação do $\hat{4}$ em um semitom para compor uma dominante anti-relativa que resulte em uma tríade consonante estável.
+
+Cada um destes acordes pode ser confirmado e se desenvolver com uma dominante e uma subdominante individuais (acorde com base na tríade formada a uma quinta acima e uma quinta abaixo respectivamente).
+
+Por último, vale apontar as medianas\footnote{Também chamadas de medianas cromáticas.}, que são acordes formados também a uma terça, mas com alterações com relação ao acorde de origem.
+Caso haja duas alterações cromáticas, i.e.
+duas notas alteradas por um semitom cada com relação ao acorde original, a mediana é chamada 'duplamente cromática'.
+Esta relação entre acordes é considerada de tonalismo avançado, por vezes até de dissolução do tonalismo, e tem efeitos fortes e marcantes.
+Foi muito utilizada do final do romantismo em diante por Wagner, Lizt, Richard Strauss dentre outros e é bastante simples de ser realizada~\cite{Harmonia,Schachter}.
+
+As modulações e sutilezas de encaminhamento de vozes podem ser encontradas em títulos diversos como os citados acima e constituem uma discussão longa e ainda presente.
+Apontamos somente que, para a condução de diferentes vozes, os movimentos que levam a notas vizinhas costumam soar mais coerentes e que as modulações se dão principalmente através de uma dominante individual que afirma a mudança da tônica e através de alterações cromáticas e enharmonias (mudança de função de um acorde).
+A importância da dominante torna ela o alvo primário e natural das modulações, resultando no circulo das quintas~\cite{Harmonia,Schachter,Koellheuteur,Harmony}.
+
+A montagem musical \emph{Vagas matizes} explora estas relações entre acordes.
+
+\subsubsection{Rudimentos de Contraponto}
+
+A condução de linhas melódicas simultâneas, chamadas vozes,
+é matéria do campo de estudo chamado contraponto.
+A nossa bibliografia aponta títulos para estudos específicos, geralmente percorrendo gêneros de condução de vozes e desembocando em gêneros escolásticos como cânones, invenções e fugas.
+Beethoven resumiu de forma satisfatória para uma breve exposição como esta da seguinte forma: todo intervalo dissonante é resolvido por grau conjunto descendente da voz superior, só é permitido movimento paralelo em terças ou sextas e não mais do que três movimentos consecutivos e é proibido atingir quinta ou oitava por movimento direto, pois é considerada uma 'quinta oculta' ou 'oitava oculta'~\cite{Fux,Tragtenberg,SchoenbergContra}.
+
 \subsubsection{Rítmo}
@@
 
-As unidades mais fortes (acentuadas) são as que consistem nas cabeças dos tempos. Nas divisões binárias (2, 4 e 8 dos casos considerados aqui),
+As unidades mais fortes (acentuadas) são as que consistem nas 'cabeças das divisões'.
+A cabeça de um tempo ou uma unidade dividida é a primeira parte da subdivisão.
+Nas divisões binárias (2, 4 e 8 dos casos considerados aqui),
 as unidades consideradas fortes se revesam com as fracas
@@
 
-Assim, $P_i^j$ é a $i$-ésima unidade de 
+Desta forma, $P_i^j$ é a $i$-ésima unidade de 
 pulsos no nível $j$ de agrupamento:
@@
 do nosso escopo, uma música ou um conjunto coeso de músicas. Dito de outra forma: a duração de $P^{min(j)}_i$, para $\forall \; i$
-é maior que $50ms$ e as durações de $P^{\text{máx}(j)}_i$ \emph{somadas}  em $\forall i$
+é maior que $50ms$ e as durações de $P^{\text{máx}(j)}_i$ \emph{somadas}  em $\forall \; i$
 devem resultar em uma duração menor do que alguns minutos ou, no máximo, poucas horas.
@@
 (ou múltiplo de três) índices temos uma relação \emph{perfecta}, 
-quando soma dois, quatro ou oito índices temos uma relação \emph{imperfecta}.
+quando soma dois, quatro ou oito índices temos uma relação \emph{imperfecta} (veja figura~\ref{fig:pulsoSubAgl}.
 
@@
 
-Seja $S_i=\{s_i\}_0^{\Lambda_S-1}$ uma sequência crescente. A sequência 
-$R_i=\{r_i\}_0^{2\Lambda_S -2}=\{s_{(\Lambda_S-1-|\Lambda_S-1-i|)}\}_0^{2\Lambda_S-2}$ 
+Seja $U_i=\{u_i\}_0^{\Lambda_U-1}$ uma sequência crescente. A sequência 
+$R_i=\{r_i\}_0^{2\Lambda_U -2}=\{u_{(\Lambda_U-1-|\Lambda_U-1-i|)}\}_0^{2\Lambda_U-2}$ 
 é uma sequência perfeitamente simétrica\footnote{Apresenta simetria especular perfeita, i.e.
@@
 Comparadas as sequências estudadas na teoria das sequências e as utilizadas na música, podemos concluir que existe um interesse especial por sequências relativamente curtas comparadas com as encontradas na teoria matemática~\cite{Guidorizzo,Schoenberg}.
+Teoricamente, estas sequências aplicadas a qualquer característica dos eventos musicais resulta em arcos.
+Assim, podemos conceber para a mesma sequência de eventos um certo número de arcos distintos, com tamanhos diferentes e climax diferentes para as características escolhidas.
+Embora este seja um recurso
+bastante interessante e útil, vale manter em mente que a correlação dos arcos resulta em coerência da escuta~\cite{Schachter}.
 
@@
 a completude desta abordagem segundo princípios musicais, matemáticos e
-filosóficos\footnote{De
+filosóficos além da incidência musical histórica\footnote{De
 forma bastante resumida, apontamos o
@@
 apontamos em especial a técnica \emph{Change Ringing} que consiste essencialmente
-em padrões de permutações dentre um fixado número de notas~\cite{change}.} e a consequente realização de dois trabalhos~\cite{figgusOriginal, figgusEspacializacao}.
-
-Toda permutação pode ser operada com outra permutação, resultando na permutação
-que representa a ordenação que resulta se aplicada uma permutação
-depois a outra:
-
-\begin{equation}
-p_1 \bullet p_2 = p_3, \quad \forall \; p_1, p_2 \; \text{permutações finitas}
-\end{equation}
-
-Uma permutação finita aplicada sucessivamente um número suficiente de vezes
-resulta na ordenação inicial dos elementos:
-
-\begin{equation}
-(p)^n = e \quad : \quad \text{se} \; p \; \text{finito} \; \Rightarrow \; \exists \;\; n  \in \mathbb{N} \; \text{e finito}
-\end{equation}
-
-O menor valor de $n$ é chamado de \emph{ordem} da permutação $p$. O elemento $e$ é chamado de elemento neutro e é
-a permutação que leva uma ordenação para ela mesma. Aqui já se pode notar o comportamento cíclico causado pela aplicação sucessiva da mesma permutação.
+em padrões de permutações dentre um fixado número de notas~\cite{change}.} e a consequente realização de dois trabalhos acadêmicos~\cite{figgusOriginal, figgusEspacializacao}.
+
+Qualquer conjunto de permutações pode ser utilizado como gerador de grupos algébricos~\cite{groups}. Assim, vale citarmos as propriedades que definem um grupo $G$:
+
+\begin{align}
+\forall \;\; p_1,p_2 \in G \Rightarrow\quad\quad\quad\;\; p_1 \bullet p_2 & = p_3 \in G  \quad\quad\quad\;\;\;\text{(propriedade de fechamento)} \\
+\forall \;\; p_1,p_2,p_3 \in G \Rightarrow\quad (p_1\bullet p_2)\bullet p_3 & = p_1\bullet (p_2\bullet p_3)\quad\;  \text{(propriedade da associatividade)} \\
+\exists \;\; e \in G :\quad\quad\quad\quad\; p \bullet e & = e \bullet p \;\;\;\; \forall p \in G  \quad \text{(existência do elemento neutro)} \\
+\forall \;\; p \in G, \;\exists\; p^{-1} :\quad\quad\quad\;  p\bullet p^{-1} & =p^{-1}\bullet p = e  \quad\quad\;\text{(existência do inverso)}
+\end{align}
+
+A primeira destas propriedades nos é de particular importância, pois concluímos que toda permutação pode ser operada com outra permutação.
+De fato, podemos aplicar uma permutação, depois outra e, se compararmos com a ordenação inicial, temos uma permutação resultante.
+As outras propriedades resultam triviais para nosso caso.
+
+Alguns teoremas nos são caros.
+Por exemplo, segue imediatamente que, em um grupo finito, todo elemento $p$ operado consigo mesmo um número suficiente de vezes resulta no elemento neutro $p^n=e$, pois caso contrário a propriedade do fechamento resultaria em um grupo infinito gerado por $p$.
+O menor $n\;:\;p^n=e$ é chamado de ordem do elemento.
+Assim, uma permutação finita $p$ aplicada sucessivamente um número suficiente de vezes resulta na ordenação inicial dos elementos, formando um ciclo, i.e.
+um arco musical se os elementos ordenados são elementos musicais e se consultados em cada estágio.
+
+Estes arcos, que levam uma ordenação de elementos na mesma ordenação após um certo número de aplicações de permutações, podem ser efetuados pelo uso conjunto de diferentes permutações. Como exemplo
+histórico, a tradição chamada \emph{change ringing} concebe música através de sinos tocados um após o outro e então tocados novamente, mas em uma ordem diferente.
+Este processo se repete até que se atinja a ordenação inicial e o conjunto de ordenações diferentes percorridas consiste no que é chamado de um \emph{peal}.
+Dispomos na tabela~\ref{tab:change} um \emph{peal} tradicional de 3 sinos (1, 2 e 3), que explora todas as suas ordenações.
+Cada linha apresenta uma ordenação dos sinos a ser tocada.
+As permutações agem aqui entre as linhas.
+% {\color{blue} AA}
+
+\begin{table}[htpq!]
+\centering
+\begin{tabular}{l c r}
+\textcolor{red}{1} & \textcolor{blue}{2} & \textcolor{green}{3} \\
+\textcolor{blue}{2} & \textcolor{red}{1} & \textcolor{green}{3} \\
+\textcolor{blue}{2} & \textcolor{green}{3} & \textcolor{red}{1} \\
+\textcolor{green}{3} & \textcolor{blue}{2} & \textcolor{red}{1} \\
+\textcolor{green}{3} & \textcolor{red}{1} & \textcolor{blue}{2} \\
+\textcolor{red}{1} & \textcolor{green}{3} & \textcolor{blue}{2} \\
+\textcolor{red}{1} & \textcolor{blue}{2} & \textcolor{green}{3}
+\end{tabular}
+\caption{Change Ringing: \emph{Peal} (padrão) com 3 sinos}
+\label{tab:change}
+\end{table}
+
+Para maiores detalhes desta tradição, recomendamos~\cite{change}.
+Importante para nós aqui é notar a existência de tradições musicais consciêntes da importância das permutações na música e observar que, neste caso, a estrutura musical consiste em permutações propriamente ditas e algumas permutações diferentes estão operando.
+Assim, fica claro o comportamento cíclico causado pela aplicação sucessiva da mesma permutação
+e a consistência musical deste procedimento.
 
@@
 Primeiro, a permutação $p$ não precisa envolver todos os elementos de $L_i$, i.e. ela
-pode operar em subconjuntos de $L_i$. Segundo, nem todos os eventos $L_i$ precisam ser executados a
+pode operar em subconjuntos de $L_i$. Segundo, nem todos os elementos $l_i$ precisam ser executados a
 cada consulta de estado realizada.
@@
 Podemos dispor permutações que operem no mesmo conjunto, cada uma destas permutações
-$p_i$ possui: dimensões das notas em que opera (frequência, duração, fades, intensidade, etc), modo de incidência (opera em $L_i$ a cada uma ou duas utilizações de $L_i$), período de aplicação (a cada quantas consultas é aplicada a permutação).
+$p_i$ possui ao menos, segundo nossa exposição acima, vinculadas à permutação em si: dimensões das notas em que opera (frequência, duração, fades, intensidade, etc) e período de incidência (a cada quantas consultas é aplicada a permutação).
 Na realização das notas de $L_i$, uma solução fácil e coerente é executar as primeiras $n$ notas\footnote{A execução de notas disjuntas de $L_i$ equivale a modificar a permutação e executar as primeiras notas.}.

### 2012-10-30 scripts/metricaMusical.py

@@ -21,7 +21,7 @@ for i in xrange(9):
 
 p.plot((-1000,-1000),'r',linewidth=3,label=u'compassos compostos, divisões ternárias, modus perfectus')
 p.plot((-1000,-1000),'b',linewidth=3,label=u'compassos simples, divisões binárias, modus imperfectus')
-p.plot((-1000,-1000),'k',linewidth=3,label=u'compassos em 5 e 7, quiálteras de 5 e 7')
+p.plot((-1000,-1000),'k',linewidth=3,label=u'compassos complexos em 5 e 7, quiálteras de 5 e 7')
 p.legend(loc="upper right")
 
 

### 2012-10-31 FIGGUScode.tex
adicionado arquivo de apendice

\chapter{Código do FIGGUS (Finite Groups in Granular and Unit Synthesis}
\label{cap:FIGGUScode}

### 2012-10-31 devEresultados.tex
andando

@@
     \caption{Espectros das ondas sonoras do oboé natural e de período amostrado}
-        \includegraphics[width=\textwidth]{figuras/espectroOboeAmostradoNatural2}
+        \includegraphics[width=\textwidth]{figuras/espectroOboeAmostradoNatural3}
         \label{fig:espectroOboe}
@@
 musicais com características próprias de altura (dada pela frequência),
-tímbre (dada pela forma de onda),
-volume (dada pela intensidade) e duração (dada pelo número de amostras).
+tímbre (dado pela forma de onda),
+volume (dado pela intensidade) e duração (dada pelo número de amostras).
 Em uma nota, estas características podem ser consideradas
@@
 
-Os elementos de $S_j$ podem ser mixados (veja equação~\ref{eq:mixagem} e a figura~\ref{fig:mixagem}) formando intervalos e acordes ou concatenados (veja equação~\ref{eq:concatenacao} e a figura~\ref{fig:concatenacao}) formando sequências melódicas.
+Os elementos de $S_j$ podem ser sobrepostos por mixagem (veja equação~\ref{eq:mixagem} e a figura~\ref{fig:mixagem}) formando intervalos e acordes ou concatenados (veja equação~\ref{eq:concatenacao} e a figura~\ref{fig:concatenacao}) formando sequências melódicas e rítmos.
 
@@
 
-Nesta nomenclatura (padrão no Brasil há algumas décadas),
+Nesta nomenclatura (um dos padrões presentes no Brasil~\cite{Lacerda}),
 um intervalo justo (uníssono, quarta, quinta e oitava), ou um intervalo maior (segunda maior, terça maior, sexta maior ou sétima maior), acrescido de um semitom, resulta em um intervalo aumentado.
@@
 
-Este arcabouço resume com exatidão a teoria dos intervalos musicais~\cite{Lacerda} e nos será útil adiante.
+Este arcabouço resume com exatidão a teoria tradicional dos intervalos musicais~\cite{Lacerda} e nos será útil adiante.
 Dispomos esta nomenclatura também por completude desta exposição, não há necessidade alguma de decorar estes nomes ou apreender estes detalhes todos enquanto não há uso efetivo.
-A montagem \emph{Intervalos entre alturas} explora estes intervalos de formas isoladas e de formas diversas.
+A montagem \emph{Intervalos entre alturas} explora estes intervalos de formas isoladas e diversas.
 
@@
 relativa & acorde principal da função & antirelativa \\\hline\hline
-$\hat{6},\hat{1},\hat{2}$ & tônica:       $\hat{1},\hat{3},\hat{5}$ & $\hat{3}, \hat{5},      \hat{7}$ \\
+$\hat{6},\hat{1},\hat{3}$ & tônica:       $\hat{1},\hat{3},\hat{5}$ & $\hat{3}, \hat{5},      \hat{7}$ \\
 $\hat{3},\hat{5},\hat{7}$ & dominante:    $\hat{5},\hat{7},\hat{2}$ & [$\hat{7},\hat{2},\hat{4}$] \\
@@
 
-A montagem musical \emph{Vagas matizes} explora estas relações entre acordes.
+A montagem musical \emph{Acorde cedo} explora estas relações entre acordes.
 

### 2012-10-31 scripts/adsr.py

@@ -33,13 +33,14 @@ som=n.sin(n.linspace(0,45*2*n.pi,De,endpoint=False))
 som=som*s
 p.plot(som,'c*',markersize=9,label=u"Som amostrado submetido ao envelope ADSR")
 p.plot(som,'c')
+p.plot((len(som),len(som)),(-.8,0),"k--")
 
 p.legend(loc="upper right",prop={'size':16})
 p.xlim(-10,De+10)
 p.ylim(-1.2,1.2)
 
 p.yticks((-1,0,SS,1),(-1,0,r"$a_S$",1),fontsize=26)
-p.xticks(())
+p.xticks((len(som),),(r"$\Lambda$",),fontsize=26)
 
 
 p.arrow(0,-1.1,DA,0,length_includes_head=True,shape='full',head_width=.1,color='cyan'     ,width=0.009)

### 2012-11-01 devEresultados.tex
finalizada versao .9 do escrito

@@
 
-\subsection{Variações na nota musical básica}
+\subsection{Variações na nota musical básica}\label{varInternas}
 
@@
 
-\begin{equation}\label{vbrGamma}
+\begin{equation}\label{vbrGamma2}
 \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f \left ( \frac{f + \mu }{f} \right )^{t_j'}  \right \rfloor= \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f .
@@
 Caso as notas soem simultaneamente, o intervalo é dito harmônico.
-Os intervalos de uníssono, oitava, quinta justa, quarta justa, terças e sextas são considerados {\bf consonantes}.
-As segundas, sétimas e trítonos são considerados {\bf dissonantes}.
+Os intervalos de uníssono, oitava e quinta justa são consonâncias perfeitas.
+As terças e sextas são consideradas consonancias imperfeitas.
+As segundas, sétimas e trítonos são considerados dissonantes.
 Esta classificação não é absoluta.
+A discrepância mais comum é com relação à quarta justa, que é tida como consonante ou dissonante de acordo com o contexto e arcabouço teórico.
+Podemos considerá-la consonante sem muitos riscos.
 A sétima menor por vezes é considerada consonante, assim como algumas culturas entoam o trítono como intervalo consonante~\cite{Roederer,Wisnick}.
@@
 
-\begin{equation}
+\begin{equation}\label{escSim}
 \begin{split}
@@
 
-\begin{equation}
+\begin{equation}\label{triades}
 \begin{split}
@@
 $\hat{6},\hat{1},\hat{3}$ & tônica:       $\hat{1},\hat{3},\hat{5}$ & $\hat{3}, \hat{5},      \hat{7}$ \\
-$\hat{3},\hat{5},\hat{7}$ & dominante:    $\hat{5},\hat{7},\hat{2}$ & [$\hat{7},\hat{2},\hat{4}$] \\
+$\hat{3},\hat{5},\hat{7}$ & dominante:    $\hat{5},\hat{7},\hat{2}$ & [ $\hat{7},\hat{2},\hat{4}\#$ ] \\
 $\hat{2},\hat{4},\hat{6}$ & subdominante: $\hat{4},\hat{6},\hat{1}$ & $\hat{6},\hat{1},       \hat{3}$
@@
 menor é a alteração do $\hat{7}$ um semitom acima para que fique somente a um semitom da tônica,
-permitindo a dominante de fato (que deve tender à tônica). Assim, a dominante é sempre maior, tanto
-em escalas
-maior ou menor.
-
-A alteração do sétimo grau
-$\hat{7}$ para atender a esta necessidade tonal dá origem a diferentes versões sobre
-a definição da dominante relativa ou antirelativa numa escala menor e recomendamos,
-para maiores detalhes, alguns títulos de nossa bibliografia~\cite{Koellheuteur,Harmonia,Schachter,harmFunc}.
-
-Voltando à escala maior, a dominante anti-relativa forma um acorde diminuto, de forma que geralmente se considera esta tríade uma 'dominante com sétima e sem fundamental'~\cite{Koellheuteur} e não se considera nenhuma dominante anti-relativa.
-Outra possibilidade encontrada na literatura é a elevação do $\hat{4}$ em um semitom para compor uma dominante anti-relativa que resulte em uma tríade consonante estável.
+permitindo a dominante de fato (que deve ser maior e tender à tônica).
+Assim, a dominante é sempre maior, tanto em escalas maior ou menor.
+Por este motivo, a dominante relativa permanece a uma terça abaixo e a antirelativa a uma terça acima na escala menor.
+
+A dominante anti-relativa forma um acorde menor, exigindo a alteração do quarto grau um semitom para cima $\hat{7}\#$.
+O acorde diminuto $\hat{7},\hat{2},\hat{4}$, geralmente é considerado uma 'dominante com sétima e sem fundamental'~\cite{Koellheuteur}.
 
@@
 As modulações e sutilezas de encaminhamento de vozes podem ser encontradas em títulos diversos como os citados acima e constituem uma discussão longa e ainda presente.
-Apontamos somente que, para a condução de diferentes vozes, os movimentos que levam a notas vizinhas costumam soar mais coerentes e que as modulações se dão principalmente através de uma dominante individual que afirma a mudança da tônica e através de alterações cromáticas e enharmonias (mudança de função de um acorde).
+Apontamos somente que, para a condução de diferentes vozes, os movimentos que levam a notas vizinhas costumam soar mais coerentes, a importância maior da voz mais aguda e mais grave (e menor das vozes internas) e que as modulações se dão principalmente através de uma dominante individual que afirma a mudança da tônica e através de alterações cromáticas e enharmonias (mudança de função de um acorde).
 A importância da dominante torna ela o alvo primário e natural das modulações, resultando no circulo das quintas~\cite{Harmonia,Schachter,Koellheuteur,Harmony}.
@@
 é matéria do campo de estudo chamado contraponto.
-A nossa bibliografia aponta títulos para estudos específicos, geralmente percorrendo gêneros de condução de vozes e desembocando em gêneros escolásticos como cânones, invenções e fugas.
-Beethoven resumiu de forma satisfatória para uma breve exposição como esta da seguinte forma: todo intervalo dissonante é resolvido por grau conjunto descendente da voz superior, só é permitido movimento paralelo em terças ou sextas e não mais do que três movimentos consecutivos e é proibido atingir quinta ou oitava por movimento direto, pois é considerada uma 'quinta oculta' ou 'oitava oculta'~\cite{Fux,Tragtenberg,SchoenbergContra}.
+A nossa bibliografia aponta títulos para estudos específicos, geralmente percorrendo formas sistemáticas de condução de vozes e desembocando em gêneros escolásticos como cânones, invenções e fugas.
+É possível resumir
+as regras principais do contraponto e é dito que o próprio Beethoven - dentre outros - esboçou uma síntese.
+
+Como
+resumo de nossa bibliografia, apontamos que as dissonâncias só podem ser atingidas por movimento de
+grau conjunto (notas vizinhas) ou resolvidas
+imediatamente por movimento descendente de grau conjunto da voz superior. Além disso,
+as consonâncias perfeitas só devem ser atacadas por movimento contrário ou oblíquo entre
+as vozes. O movimento paralelo só é permitido entre consonâncias imperfeitas e não mais
+do que 3 vezes consecutivas~\cite{Fux,Tragtenberg,SchoenbergContra}.
+Estas regras tem como um dos objetivos principais, preservar a independência entre as vozes.
+Dispomos os movimentos tidos como base para o contraponto na figura~\ref{fig:movContraponto}.
+Observe o movimento paralelo como um caso específico do movimento oblíquo.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Movimentos diferenciados pelo contraponto}
+        \includegraphics[width=\textwidth]{figuras/movContraponto}
+        \label{fig:movContraponto}
+\end{figure}
 
@@
 
-\subsubsection{Estruturas cíclicas}
+\subsubsection{Estruturas cíclicas}\label{estCic}
 Os arcos musicais nem sempre são fruto de estruturas direcionais como as citadas
@@
 
-\begin{align}
+\begin{equation}\label{eq:groups}
+\begin{split}
 \forall \;\; p_1,p_2 \in G \Rightarrow\quad\quad\quad\;\; p_1 \bullet p_2 & = p_3 \in G  \quad\quad\quad\;\;\;\text{(propriedade de fechamento)} \\
@@
 \forall \;\; p \in G, \;\exists\; p^{-1} :\quad\quad\quad\;  p\bullet p^{-1} & =p^{-1}\bullet p = e  \quad\quad\;\text{(existência do inverso)}
-\end{align}
+\end{split}
+\end{equation}
 
@@
 
+As permutações são usualmente escritas em notação cíclica ou natural.
+Por brevidade, vamos abordar somente a notação natural.
+Esta notação consiste na ordem dos índices originais considerados resultante da aplicação da permutação.
+Assim,
+convencionada a ordenação original dada pela sequência de seus índices $[0\;1\;2\;3\;4\;5\;...]$ a permutação é notada pela sequência que produz (ex. $[1\;3\;7\;0\;...]$).
+
 Importante notar que não é necessário permutar os elementos de $L_i$, mas somente
@@
 
+\subsubsection{Idioma musical?}
+
+De certa forma central à experiência músical é a concepção de 'idioma musical'.
+Existem diversas empreitadas para modelar e explorar diferentes entendimentos
+sobre a linguagem musical e linguística aplicada à música e para discernimento entre 
+o que seriam diferentes
+'idiomas musicais'\cite{Lerdahl, Harmonia, Schachter,Alfaix}.
+
+Apontamos aqui que, de forma simplificada e resumida, um idioma musical é fruto da
+repetição de elementos e da repetição de relações entre elementos presentes no decorrer da música\footnote{A
+música é tradicionalmente pensada através de dicotomias bastante relacionadas umas com as outras,
+como: repetição e variação, relaxamento e tensão, equilíbrio e desequilíbrio, consonância e dissonância, etc}.
+
 \subsubsection{Usos Musicais}
-Algumas restrições funcionais possíveis: manter-se em harmonias triádicas e encadear os acordes
-que possuam relações de quintas justas dentre as fundamentais ou dentre uma fundamental
-e a terça do acorde seguinte.
-
-Melodias e harmonias.
-
-Explorações microtonais, oitava em 5 notas etc
-FIGGUS
+
+Primeiro, a nota básica foi definida e suas características postas em termos
+claros e quantitativos (sessão~\ref{notaBasica}). Em seguida, adentramos a nota musical e suas transições internas,
+além de tratamentos imediatos (sessão~\ref{varInternas}). Por fim, nesta última sessão, apresentamos
+recursos consagrados para organizar estas notas musicais em música.
+
+Assim, atingimos uma gama de recursos com uma infinidade de possibilidades de resultados,
+situação típica e tão cara às artes~\cite{Harmonia,Webern}.
+
+Existem explorações e estudos individuais para cada recurso apresentado.
+Por exemplo, pode-se obter as harmonias triádicas 'sujas' (com notas não pertencentes à tríade) através de sobreposições de quartas justas.
+Outro exemplo interessante é a presença simultânea de rítmos em diferentes métricas, constituindo o que chamamos de \emph{polirritmia}.
+A montagem musical \emph{Poli-hit mia} explora estas métricas simultâneas através de trem de impulsos convoluídos com as notas que compõem cada linha.
+
+As escalas microtonais são bastante importantes na música do século XX~\cite{microtonalidade} e possuem vários usos com resultados muito marcantes, como os usos de quartos de tom na música indiana e possibilidades de batimentos.
+A sequência musical \emph{MicroTom} explora estes recursos de forma explícita, incluindo melodias microtonais e harmonias microtonais com várias notas em um âmbito de alturas bastante reduzido.
+
+Novamente, vínculos entre parâmetros são formas poderosas de resultar músicas ou ao menos montagens musicais.
+O número de notas permutadas podem variar no decorrer da música, revelando vínculo com a duração da peça.
+As harmonias podem consituir-se triádicas (eqs.~\ref{triades}) com notas replicadas em várias oitavas e mais numerosas quanto menor a profundidade e frequência de vibratos (eqs.~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}), dentre outras incontáveis possibilidades.
+
+Um uso particularmente coerente, do ponto de vista das simetrias apresentadas nas divisões da oitava (eqs.~\ref{escSim}) e das simetrias apresentadas através das permutações (tabela~\ref{tab:change} e eqs.~\ref{eq:groups}) é o uso conjunto de ambos.
+Nas peças \emph{3 trios} esta associação é feita de forma sistemática para possibilitar uma audição dedicada a esta associação.
+
+O \emph{PPEPPS} (Pure Python EP: Projeto Solvente) é um EP inteiro sintetizado com os recursos apresentados neste trabalho.
+Com pouca parametrização, o programa resulta músicas inteiras, permitindo a composição de músicas e conjuntos de músicas com facilidade.
+As possibilidades de compartilhamento envolvem obter este album (ou outro feito de forma semelhante) baixando algumas linhas de código e rodando para obter uma pasta com as músicas.
 

### 2012-11-01 scripts/climaxMel.py

@@ -21,6 +21,7 @@ p.ylim(-.18,1.02)
 p.xticks((),())
 p.yticks((),())
 
+p.xlabel(r"tempo $\rightarrow$", fontsize=19)
 
 p.show()
 

### 2012-11-01 resumoeabstract.tex
adianto de figura e resumo

@@
 \begin{resumo}
+
+Este texto expõe de forma sistemática os elementos básicos da música em termos do som em sua representação
+discreta. Esta abordagem é de utilidade
+já comprovada pela ampla utilização feita em diversos softwares
+para criação musical e tratamento sonoro. Há interesse especial
+de músicos com aproximação das áreas tecnológicas
+de por engenheiros
+ou programadores com interesses em música ou mesmo áudio em geral.
+Mesmo assim, a literatura relacionada
+não expõe a teoria e realização de forma concisa, resultando em conteúdos
+espalhados em diversos títulos de diferentes autores. Além disso, pouco se expõe sobre a
+efetiva implementação destes recursos. Em um primeiro momento, caracterizamos
+a nota musical no som digitalizado e relacionamos de forma quantitativa
+as suas características musicais básicas aos aspectos do sinal digital. Em seguida,
+adentramos a nota básica e expomos formas de realizar as variações internas - como trêmolos, vibratos e
+variações espectrais - tão
+caras às empreitadas artísticas e típicas de notas realizadas por instrumentos reais.
+Por se encontrar em um nível sintético, as possibilidades são expandidas e, com elas,
+os recursos artísticos disponíveis para criação. Por fim, saímos da canônica
+unidade musical mínima que é a nota e lidamos com a organização de notas musicais
+em música. Neste momento são apresentados recursos musicais consagrados para geração
+de estruturas musicais através das notas, como a métrica rítmica, os intervalos de altura e os
+ciclos.
+Como resultados, além desta exposição sistemática com equações de aplicação explícita,
+está disponibilizado um conjunto de scripts
+com as implementações computacionais destes procedimentos e exemplos sonoros audíveis em equipamentos
+comuns. Dentre estes exemplos sonoros, encontramos sons com ornamentos e outros exemplos simples. Também
+dispomos montagens musicais com fins de explorar os recursos apresentados. Ao final, apresentamos
+uma implementação capaz de sintetizar discos inteiros através de alguma parametrização.
+Este arcabouço dispertou interesses e foi responsável por empreitadas coletivas diversas,
+que fogem ao escopo deste trabalho e estão dispostas em apêndices sumplementares.
 

### 2012-11-01 scripts/concatena.py

@@ -23,8 +23,12 @@ e=5.0
 ee=4
 p.xticks((),())
 p.yticks((),())
-p.plot(n.hstack((d,q,s)),'o')
+#p.plot(n.hstack((d,q,s)),'o')
+p.plot(d,'o')
+p.plot(range(len(d),len(d)+len(q)),q,'o')
+p.plot(range(len(d)+len(q),len(d)+len(q)+len(s)),s,'o')
 p.plot(n.hstack((d,q,s)))
+p.xlabel(r"tempo $\rightarrow$",fontsize=22)
 p.ylim(-1.5,1.5)
 p.xlim(-5,305)
 p.show()

### 2012-11-05 introducao.tex
primeira limpada braba para versao final do msc

@@
 \chapter{Introdução} %Nome do capítulo.
-\label{cap:intro} %Rótulo para futura referência ao capítulo.
-Em qualquer lugar da tese, você poderá citar este capítulo através de ~\ref{cap:introducao}.
-Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})
-Partindo do som como um fenomeno fíisico, esta dissertação 
-cobre casos de uso musicais de software voltado ao áudio e tecnologias livres. Optamos
-realizar este trabalho através da síntese e da simplicidade.
-
-Abordamos o som, o áudio, o áudio digital e práticas de programação voltadas para
-áudio. Não menos importantes são as plataformas nas quais desenvolvemos e executamos código
-- usualmente GNU/Linux ou BSD e navegadores de páginas de internet (\emph{browsers}) - e
-os sistemas de controle de versão dos códigos.
-
-Com estas tecnologias em mãos, foram criados diversos aplicativos com fins artísticos,
-técnicos, científicos e sociais. Como consequência, foi criada uma equipe para tratar
-destas questões, o Lab Macambira, e um sistema rico de comunicação e exposição de todos
-os desenvolvimentos aqui expostos.
-
-Este trabalho tange algumas questões ainda mais gerais e difíceis.
-Como abordar a atuação tecnológica e científica em uma área criativa? Como contemplar
-a cultura de compartilhamento e diversidade em que esta produção tecnológica está inserida?
-Enfim, é um trabalho criterioso definir quais os pontos a serem abordados em
-um trabalho como este.
+\label{cap:intro}
 
   \section{O som e o áudio}
-  O som é tipicamente utilizado para fala, sinais em geral\footnote{p.ex. buzina} e música.
-  Caracteriza-se por ser um sentido (estudar som na wikipedia e talvez helmholtz e zuckerkandl)
-
-	  \subsubsection{Breve histórico e práticas atuais}
-	As explicações físicas do fenômeno do som no decorrer do tempo. Teorias
-	corpusculares, escola de Tales. Helmholtz. Surgimento do áudio. Áudio digital.
-
-	  \subsubsection{Campos de conhecimento e atuação}
-	som: Acúsitica, psicoacúsica, luthearia
-	áudio: microfones, caixas de som
-	áudio digital: conversores a/d d/a, programação voltada ao áudio e música
 
   \section{Código aberto, software livre e cultura livre}
-  O código é repetidamente destacado como o bem mais valioso produzido em nosso tempo.
-  O código aberto - e sua abordagem politico-filosófica chamada de software livre -
-  se propõe a constituir um repertório tecnológico da humanidade, acessível por
-  qualquer ser humano. Esta proposta orienta o que é chamado de cultura digital,
-  um agregado de práticas e explorações recentes e advindas da popularização do digital.
-  É tido como o vetor de transformação de nossa época por excelência.
-  Qualquer licença de software livre é também uma licença de código aberto
-  (Open Source), a diferença entre as duas nomenclaturas reside essencialmente
-  na sua apresentação e na filosofia.
 
       \subsection{Cultura de compartilhamento}
-      \label{sec:soft_compar}
-    Proposta de compartilhamento. Softwares para dar suporte a esta conectividade.
 
       \subsection{Licenças livres para software e para outras mídias}
-      \label{sec:licencas_livres}
-    Principais tipos de licenças
-
-    Licenças voltadas para software.
-
-    Licenças voltadas para outras mídias.
-
-    Licenças fechadas.
-
 	  \subsubsection{Principais licenças}
 	  \label{sec:princ_licencas}
-	Licenças uma a uma. GPL, LGPL, BSD, MIT, Beerware, etc
 
 	  \subsubsection{Domínio público e outras categorias}
-
-  \section{Notas sobre o estado do software}
-  \label{sec:context}
-  Aqui no Brasil: as redes existentes, carta midias livres, fluxo de informação..
-  No mundo: mudança de paradigma com relação a tudo que jah rolou e rolará:
-  protestos, modismos, entidades, leis e tentativas de leis.
 
@@
 	  quase sempre maior que isso [tabela de latências, média, por distância, etc].
-
 	  Contribuimos na frente do software. Para isso veja [sessao dos resultados]
-
-	  \subsubsection{Dinâmica de desenvolvimento}
-	  \label{sec:din_dev}
-	  A modularidade do sistema \emph{Unix type}. Codigo como substancia modular
-	  e reativa. Colaboratividade, listas de discussão, IRC.
-
-	  No caso do áudio: as compensações: jack,
-	  facilidades de instalação e segurança, tudo hackeável.
-
-%%%%%%%%
-      \subsection{Identificação de tarefas pouco suportadas}
-      \label{sec:tarefas_n_sup}
-  [ESCREVER NO FINAL]
 
@@
   \label{sec:tutoriais}
-Descentralização: Fator caminhão e espalhamento de informação. Descentralização
-é uma constante nas estruturas e ferramentas mais relevantes como redes
-sociais, comunicadores que agora funcionam por broadcast e protocolos de
-compartilhamento de dados, em especial arquivos inteiros, unidades mais
-bem organizadas.
-
-Aumento da conectividade através de avanços em telecomunicações e transportes.
-
     \subsubsection{Tutoriais}
@@
 
-    \subsubsection{Estudo de caso da difusão de conteúdos livres}
-    \label{sec:dif_sl}
-Estamos desenvolvendo trabalhos em análise de comparação
-entre sucessões de tendências filosóficas ~\cite{philome} e
-musicais ~\cite{musime} e trabalhos em processamento de escrita
-e fala ~\cite{rede-associacoes, complenet, enfmc, ifsc}.
-
-Os estudos subsequentes a estes trabalhos devem desenvolver
-um estudo sobre a dinâmica de geração e propagação de conteúdo na rede.
-Comparando textos e utilizando medidas de similaridade separa-se
-o que um texto não contém e o que é semelhante, i.e. servem de propagação
-um do outro. A criatividade será enfocada também como estudos de dinâmicas
-de atributos de redes complexas de palavras, como as redes de associação ~\cite{rede-associacoes}.
-
   \subsection{Extensão de recursos em software livre para multimídia}
@@
 
-  \subsubsection{Plugins feitos}
-  \label{sec:tutoriais}
-
-  \subsubsection{Performance em tempo real}
-  \label{sec:perf}
-
   \subsubsection{Síntese de sons e estruturas musicais}
@@
 
-  \subsubsection{Analise}
+  \subsubsection{Capacitação para performance em tempo real e análise}
+  \label{sec:perf}
   \label{sec:analise}

### 2012-11-05 pretextual.tex
adentrada variacoes na nota musical
lipada geral e arrumando AM e FM
primeira limpada braba para versao final do msc

@@
 	\begin{epigrafetop}
-		{They must find it difficult ... \\ Those who have taken authority as the truth, \\ rather than truth as the authority.} %frase
-		{Gerald Massey (1828 - 1907)} %referência
-	\end{epigrafetop}
-
-% \,      a small space
-% \:      a medium space
-% \;      a large space
-% \quad   a really large space
-% \qquad  a huge space
-% \!      a negative space (moves things back to the left)
-
-	\begin{epigrafemid}
 		{from Pantheon import Obatalá\, \, \, \\
@@
         {Autoria coletiva e anônima (2010)}
-	\end{epigrafemid}
-
-	\begin{epigrafebot}
+	\end{epigrafetop}
+% \,      a small space
+% \:      a medium space
+% \;      a large space
+% \quad   a really large space
+% \qquad  a huge space
+% \!      a negative space (moves things back to the left) 
+%	\begin{epigrafemid}
+%		{They must find it difficult ... \\ Those who have taken authority as the truth, \\ rather than truth as the authority.} %frase
+%		{Gerald Massey (1828 - 1907)} %referência
+%	\end{epigrafemid}
+	\begin{epigrafemid}
 		{If I have seen further, \\it is by standing on the shoulders of giants.} %frase
 		{Sir Isaac Newton (1642 - 1727)} %referência
+	\end{epigrafemid}
+	\vspace{-1cm}
+	\begin{epigrafebot}
+		{Electronic music used pure sounds, completely calibrated. You had to think digitally, as it were, in a way that allowed you to extend serial ideas into other parameters through technology.} %frase
+		{Luc Ferrari (1929 - 2005)} %referência
 	\end{epigrafebot}
-
 \resumoeabstract %Para incluir o texto do resumo e abstract,use o arquivo resumoeabstract.tex
@@
 	\begin{listaespecial}[BIGNAMEWIDTH]
-		\item[\$] Indica que o resto da linha é um comando bash
+		\item[\$] Indica que o resto da linha é um comando bash$\Lambda_{DTI}=DTI . f_a$
 		\item[\#] Resto da linha é comentario em código Bash ou Python
@@
 		\item[Tessitura :] âmbito de altura ($\log(freq)$) do objeto em discussão (instrumento, passagem musical, etc).
+
+		\item[Delay :] embora a palavra signifique simplesmente atraso, no contexto de música a áudio o \emph{delay} comumente se refere a um efeito causado por diversas repetições de um som original, com atrasos diferentes.
 	\end{listaespecial2}

### 2012-11-05 agradecimentos.tex
primeira limpada braba para versao final do msc

--- 2012-01-22 agradecimentos.tex
@@
 Agradeço a Felipe Machado pela visão, a Fabiana 'Goa' Sherine pelo entendimento,
-a Glerm Soares pela transcendência, a Fabianne Baveldi pela excelência, a Daniel
-Marostegan e Carneiro pelas lições e confianças. Agradeço ao profícuo Vilson Vieira, que tornou-se especialmente próximo
-das contribuições citadas neste trabalho, construindo sobre elas e com elas. Especial agradecimento
-ao meu irmão Ricardo Fabbri que também se aproximou destas contribuições cuidando de processamento de imagens e video. Fortíssima referência
-para mim, tanto para os assuntos tratados nesta dissertação, quanto para questões humanas. Agradecimento
-forte ao Lab Macambira, nominalmente estes assíduos: [Equipe Lab Macambira]. Um profundo agradecimento
+a Glerm Soares pela transcendência, a Fabianne Baveldi pela excelência, dentre tantos outros que não há páginas para escrever aqui.
+Agradeço ao profícuo Vilson Vieira, que tornou-se especialmente próximo
+das contribuições citadas neste trabalho, construindo sobre elas e com elas. Condizente com a fortíssima referência
+para que é para mim, tanto para os assuntos tratados nesta dissertação quanto para questões humanas, um especial agradecimento
+ao meu irmão Ricardo Fabbri que também se aproximou destas contribuições tanto cuidando de processamento de imagens e video quanto articulando e amadurecendo as ideias. Agradecimento
+forte ao LabMacambira.sf.net. Um profundo agradecimento
 aos meus orientadores de IC: Adolfo Maia Junior e Jônatas Manzolli. Agradeço a José Augusto Mannis
@@
 circuitos elétricos, processos estocásticos
-e programação bio-inspirada. Sem sombra de dúvida maiores agradecimentos para os assíduos mentores
-deste trabalho: ao orientador-colaborador Prof. Luciano da Fontoura Costa, que, dentre outras qualidades,
-mistura criatividade com clareza de pensamento. Finalmente, agradeço ao meu 
-orientator Prof. Osvaldo Novais de Oliveira Júnior pelo inestimável acompanhamento
-e notável paciência, dedicação e precisão em todos os momentos.
+e programação bio-inspirada. Sem sombra de dúvida, os maiores agradecimentos para os assíduos mentores
+deste trabalho: ao orientador-colaborador Prof. Luciano da Fontoura Costa e ao meu 
+orientator Prof. Osvaldo Novais de Oliveira Júnior. Ambos proporcionaram inestimável acompanhamento e notável paciência, dedicação e precisão em todos os momentos.

### 2012-11-05 conclusoes.tex
terminada a revisao geral da dissertacao
primeira limpada braba para versao final do msc

@@
 Conclusões sobre o material apresentado nesta dissertação e trabalhos a serem realizados.
+
+Redes complexas para estudo do dev de SL (sf.net) e de listas de discussao (GMANE).
+    \subsubsection{Estudo de caso da difusão de conteúdos livres}
+    \label{sec:dif_sl}
+Estamos desenvolvendo trabalhos em análise de comparação
+entre sucessões de tendências filosóficas ~\cite{philome} e
+musicais ~\cite{musime} e trabalhos em processamento de escrita
+e fala ~\cite{rede-associacoes, complenet, enfmc, ifsc}.
+
+Os estudos subsequentes a estes trabalhos devem desenvolver
+um estudo sobre a dinâmica de geração e propagação de conteúdo na rede.
+Comparando textos e utilizando medidas de similaridade separa-se
+o que um texto não contém e o que é semelhante, i.e. servem de propagação
+um do outro. A criatividade será enfocada também como estudos de dinâmicas
+de atributos de redes complexas de palavras, como as redes de associação ~\cite{rede-associacoes}.

### 2012-11-05 devEresultados.tex
primeira limpada braba para versao final do msc

@@
 %% ------------------------------------------------------------------------- %%
-\chapter{Desenvolvimentos e resultados} %Nome do capítulo.
+\chapter{Desenvolvimentos e resultados: música no som digitalizado} %Nome do capítulo.
 \label{cap:resultados} %Rótulo para futura referência ao capítulo.
@@
 Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})
-Os resultados explicitados nesta dissertação estão focados no processamento de 
-áudio através de códigos em liguagens de programação de computadores e
-voltado para a música e na síntese de estruturas musicais. O tema é amplo
-e pertinente o suficiente para contar atualmente com ampla bibliografia e um histórico
-que remonta desde a própria história da computação e das linguagens de programação.
-Desta forma, entendemos cabível focar este texto nas empreitadas realizadas pelos autores
-da presente dissertação, com foco nas realizações do primeiro autor. São usos
-reais destas tecnologias, manifestadas de forma cultural e com repercussões
-sociais, como poderá é relatado neste mesmo capítulo.
-
-Tais realizações possuem um espectro amplo
-dentro do escopo proposto e constitui um percurso relativamente completo
-pelas possibilidades e decorrências da abordagem da música pelo código computacional\footnote{ao menos dentro do se denomina 'Cultura Livre', veja o primeiro capítulo do presente escrito para maiores informações}.
-
-Estaremos abordando experimentos abertos em áudio, abordagens musicais pelo código, incluindo música em tempo real, diferido, na matéria e repercussões no tecido social.
-De forma a relatar os resultados destas investidas, segue uma sessão sobre ações coletivas realizadas através das mobilizações que estes códigos criaram, e as estruturas virtuais mantidas na intenet relacionadas a estas mobilizações e ações.
-Por fim, alguns materiais didáticos criados pelo autor - que possuem algum papel emblemático - são também expostos por comporem o arsenal de repercussões desta investida no em som e código.
-
-  \section{Experimentos abertos em áudio: LADSPAs, Wavelets e Redes Complexas}
-
-Todos os desenvolvimentos desta dissertação estão em repositórios abertos\cite{repositorios-tese-dev}.
-Alguns foram especialmente importantes como percurso para o que é apresentado neste
-escrito. Ou seja, são explorações que manifestaram diferentes aspectos
-do trabalho do código voltado para o áudio e a música.
-
-O código para a arte sonora, incluindo a musical, manifesta-se como cultura pois é fruto de práticas
-espontâneas, diárias e coletivas.
-\footnote{As chamadas culturas biopunk, ciberpunk, cipherpunk, hacker, digital e outras mais,
-possibilitadas aos recentes desenvolvimentos em telecomunicação, dizem respeito em menor
-ou maior grau à produção de código como cultura.}
-
-A seguir apresentamos alguns bons exemplos dos desenvolvimentos
-desta dissertação especificamente em áudio, sem o envolvimento direto da música. São
-códigos de necessidade para música, mas de uso mais geral e cuja implementação
-requer muito mais de processamento de sinais e computação do que de música propriamente dito.
-
-      \subsubsection{Plugins LADSPA e lv2}
-      [repos AE, LM, wiki EL, historico CDTL]
-LADSPA (Linux Audio Developers Simple Plugin API) é a API livre\footnote{Veja o Capítulo 1 para a definição de livre neste contexto.} estabelecida de plugins de áudio até a presente data.
-A última versão liberada desta API é a 1.1, em uso corrente até hoje e é de 2002 segundo os arquivos de cabeçalho relacionados.
-
-A exploração de APIs específicas foge ao escopo do presente texto e os códigos completos podem ser encontrados pela rede, junto a tutoriais e plugins de exemplo.
-O código destes desenvolvimentos e pertinente ao áudio são relativos à síntese e remoção de ruídos.
-Da síntese, exitem abordagens mais puristas, que no caso sintetiza ruído branco e filtra para resultar nos espectros desejados (veja logo abaixo para maiores especificações).
-Uma abordagem menos purista porém mais eficiente é a de usar uma amostra curta do ruído e reproduzí-la indefinidamente através de \emph{cross-fades}.
-
-Já na remoção de ruídos as abordagens são as mais variadas e extremamente dependentes da aplicação.
-A complexidade dos algorítmos pode atingir níveis de especialidade em processamento de sinais que merecem um trabalho dedicado.
-Aqui iremos expor somente sobre remoção de ruído \emph{'Hum'} causado usualmente pela corrente alternada que alimenta os equipamentos utilizados.
-
-Abaixo dispomos explicações breves sobre os diferentes ruídos e os códigos relacionados. Preferimos expor em Python os códigos de síntese
-por questão de clareza e coerência com o resto do texto\footnote{Os códigos em C/C++ e a implementação como plugin lv2 está no repositório git LV2 como pode ser visto em: http://labmacambira.git.sourceforge.net/git/gitweb-index.cgi}.
-
-O procedimento de reprodução do ruído através de uma amostra repetida e em crossfade é a seguinte:
-
-[código da reprodução do ruído em loop e croos-fade ~10 linhas]
-
-Sintetizando diferentes ruídos\footnote{Dado o propósito do texto e da implementação, utilizamos nomenclaturas utilizadas na música para os ruídos, como se pode notar.}:
-\begin{itemize}
-    \item Ruído Branco: 
-    \item Ruído Rosa:
-    \item Ruído Marrom:
-    \item Ruído Violeta:
-    \item Ruído Azul:
-    \item Ruído Violeta:
-    \item Ruído Cinza:
-    (nao oficiais)
-    \item Ruído Vermelho:
-    \item Ruído Laranja:
-    \item Ruído Preto:
-    \item Ruído Branco Ruidoso:
-    \item Ruído Preto Ruidoso:
-    \item Clicks:
-\end{itemize}
-
-A remoção de ruído Hum é baseada em uma sequência de filtros notch (ver fundamentos) idealmente dispostos nos harmônicos da frequência de oscilação da corrente elétrica.
-Cada um destes filtros deve permitir ajustes finos no fator de qualidade e também na frequência central pois trata-se de um sistema real passivel dos mais diversos efeitos de distorção do modelo ideal.
-Abaixo expomos o código em Python para tal \emph{removedor de ruído} como é chamado.
-O código C/C++ e a implementação como plugin está no mesmo repositório disponibilizado para a síntese de ruídos.
-
-[código e explicação do remodedor de ruído Hum]
-
-      \subsubsection{Protocolo de compactação de áudio via wavelets, polinômios e permutações}
-Criado por Rafael Santos Mendes em conjunto com o primeiro autor deste trabalho e em decorrência do mestrado de André Luvizotto, este método
-consiste em aproximar cada folha de uma decomposição WaveletPacket por um polinômio e uma permutação.
-A permutação facilita a aproximação polinomial, pois leva folhas WP de coeficientes ordenados para sucessões melhor comportadas de seus valores\footnote{O método é descrito em um artigo de 2009: Compressão de Áudio via Wavelets, Aproximações Polinomiais e Permutações [Fabbri, Mendes]}.
-Representamos cada permutação por um conjunto ordenado de números inteiros.
-Já o polinômimo possui poucos coeficientes (usualmente 6 ou 8) e os representamos como um conjunto ordenado de variáveis de tipo ponto flutuante.
-
-      \subsubsection{Processamento de voz via redes complexas}
-Conversar com Chu e com Luciano e decidir se retiro esta sessão ou mantenho.
-
-      [codigos bem comentados no audio experiments. Apresentação do trabalho "Speech Polarity Detection using Complex Networks Measures: First Explorations no Complenet de 2010"]
-
-\section{Música no Som Digitalizado}
-
-\subsection{Caracterização da nota musical discretizada}
+
+\section{Caracterização da nota musical discretizada}
 Em diversos contextos artísticos e teóricos, 
@@
 
-\subsubsection{Duração}
+\subsection{Duração}
 Desta forma, visando uma nota musical de duração $\Delta$,
@@
 
-\subsubsection{Volume}
+\subsection{Volume}
 O volume depende da reverberação e distribuição dos harmônicos - dentre outras medidas psicofísicas~\cite{Chowning} - mas num primeiro momento podemos obter resultados satisfatórios e de forma prática através da potência da onda:
@@
 
-\subsubsection{Altura}
+\subsection{Altura}
 
@@
 
-\subsubsection{Timbre}
+\subsection{Timbre}
 Enquanto o período da onda corresponde a uma frequência fundamental, o percurso
@@
 
-\subsubsection{O espectro no som amostrado}
+\subsection{O espectro no som amostrado}
 Além deste papel-chave na cognição (especialmente no timbre), a presença e comportamento destas componentes senoidais 
@@
 
-\subsubsection{Localização espacial}
+\subsection{Localização espacial}
 Embora não seja uma das quatro qualidades básicas tradicionalmente elegidas para caracterizar uma nota musical,
@@
 
-\subsubsection{A nota básica}\label{notaBasica}
+\subsection{A nota básica}\label{notaBasica}
 
@@
 
-\subsubsection{Usos musicais}
+\subsection{Usos musicais}
 
@@
 
-\subsection{Variações na nota musical básica}\label{varInternas}
+\section{Variações na nota musical básica}\label{varInternas}
 
@@
 
-\subsubsection{Tabela de Busca}
+\subsection{Tabela de busca}
 
@@
 
-\subsubsection{Variações incrementais de frequência e intensidade}
+\subsection{Variações incrementais de frequência e intensidade}
 
@@
 
-\subsubsection{Aplicação de Filtros Digitais}
+\subsection{Aplicação de filtros digitais}
  As possibilidades
@@
 
-\subsubsection{Ruídos}
+\subsection{Ruídos}
 De forma geral, os sons sem altura definida 
@@
 
-\subsubsection{Usos musicais parte 1: trêmolo e vibrato, AM e FM}
+\subsection{Usos musicais parte 1: trêmolo e vibrato, AM e FM}
 
@@
 
-\subsubsection{Usos musicais parte 2}
+\subsection{Usos musicais parte 2}
 A este ponto as possibilidades musicais explodiram. Possuímos unidades
@@
 \clearpage
-\subsection{Organização de notas musicais em música}
+\section{Organização de notas musicais em música}
 Seja $ S_j=\{T_{i,1},T_{i,2}...T_{i,N}\} =\bigcup_{j \in [1,N]} T_{i,j} $ uma sequência de $N$ eventos
@@
 
-\subsubsection{Afinação, intervalos, escalas e acordes}
+\subsection{Afinação, intervalos, escalas e acordes}
 Quando dobramos a frequência, temos uma oitava ascendente ($f=2f_0$).
@@
 
-\subsubsection{Rudimentos de harmonia}
+\subsection{Rudimentos de harmonia}
 
@@
 
-\subsubsection{Rudimentos de Contraponto}
+\subsection{Rudimentos de contraponto}
 
@@
 
-\subsubsection{Rítmo}
+\subsection{Rítmo}
 A noção rítmica é dependente de eventos separados por durações~\cite{rythm}. Estes eventos são discerníveis individualmente se separados
@@
 
-\subsubsection{Estruturas direcionais}
+\subsection{Estruturas direcionais}
 
@@
 
-\subsubsection{Estruturas cíclicas}\label{estCic}
+\subsection{Estruturas cíclicas}\label{estCic}
 Os arcos musicais nem sempre são fruto de estruturas direcionais como as citadas
@@
 
-\subsubsection{Idioma musical?}
+\subsection{Idioma musical?}
 
@@
 
-\subsubsection{Usos Musicais}
+\subsection{Usos musicais}
 
@@
 As possibilidades de compartilhamento envolvem obter este album (ou outro feito de forma semelhante) baixando algumas linhas de código e rodando para obter uma pasta com as músicas.
-
-\clearpage
-
-\section{Áudio e Música (deprecated)}
-
-Por razões didáticas, dividimos este capítulo em 4 partes: música em tempo diferido (ou seja, que não é feito em tempo real),
-música em tempo real,  música na matéria (suporte físico em hardware) e música no tecido social
-(considerando especialmente as mobilizações humanas relacionadas).
-
-Com isso desejamos expor a prática musical através do código
-com exemplos reais de aplicação e em uso pelo autor, por membros do LabMacambira.sf.net,
-por parceiros, colaboradores e por usuários eventuais das naturezas mais diversas.
-
-  \subsection{Música em Tempo Diferido: Minimum-fi e FIGGUS}
-
-\begin{quotation}
-\small
-'The increasing dominance of graphic interfaces for music software obscured 
-the continuing presence of the command-line tradition, 
-the code writer, the hacker. The code writing of deferred time 
-computer programming may be assembled out of time order, debugged and optimized.'
-
-\emph{Simon Emmerson, Living electronic music, 2007}
-\end{quotation}
-
-A realização musical em tempo diferido é o paradigma inicial da música computacional.
-Iniciando com o Music V, a proposta foi depois desenvolvido com o CSound. Pode-se dizer
-que até hoje é a forma como compositores usualmente pensam a música: pensando
-e escrevendo as estruturas, que depois são executadas por instrumentistas ou aparelhos eletrônicos.
-
-Assim como a composição instrumental permite um trabalho mais minucioso do que
-a improvisação instrumental, a realização musical em tempo diferido usualmente permite um
-detalhamento maior dos procedimentos do que a realização em tempo real. Por este
-mesmo motivo, trataremos inicialmente de dois trabalhos em tempo diferido.
-Aliás, como veremos a seguir, as abordagens completam uma base para a música
-computacional praticada até os dias de hoje.
-
-O som musical pode ser caracterizado físicamente, e sua síntese digital com base em nossa banda de audição e no teorema de Nyquist, sintetiza uma música amostra por amostra, através de princípios
-claros de síntese sonora e organização musical\footnote{Veja no APÊNDICE XX: o \emph{minimum-fi} hi-fi, low-fi, minimum-fi, ou seja,
-o mínimo de qualidade para assegurar existência e consistência
-da criação}. O segundo, chamado de \emph{FIGGUS} (FInite
-Groups in Granular and Unit Synthesis), utiliza os princípios do minimum-fi e constitui
-um módulo Python completo, cuja proposta é a utilização de simetrias, através de
-permutações e de Teoria de Grupos, para a composição de músicas. Como demonstração
-das capacidades do FIGGUS, ele gera um EP\footnote{Extended Play, uma album musical maior que um single mas menor que um LP (Long Play) inteiro} com um único comando. Este é o
-\emph{PPEPPS}\footnote{Pure Python EP: Projeto Solvente}, como veremos a seguir.
-
-Desta forma, esta sessão exemplifica e explicita - através de dois exemplos reais - os
-princípios do uso de código para a síntese \emph{musical}, desde as amostras
-relativas a uma nota com dada frequencia, amplitude e timbre, até a confecção
-de uma ferramenta derivada, já incorporando propostas musicais e estruturas
-mais elaboradas.
-
-      \subsubsection{Minimum-fi}
-
-Existe uma perene matiz estética e também tecnológica
-de realizar uma dada tarefa com \emph{o mínimo} necessário.
-Esta matiz é para a música igualmente
-fundamental e reconhecida como
-princípio de unidade e coerência\footnote{Este princípio tanto é fundamental
-que as escolas musicais possuem técnicas específicas, músicas possuem suas
-próprias convenções mantidas por toda a sua duração, os arcos mantém características,
-enfim, podemos até mesmo concluir que a simetria define o escopo.}. Em código computacional
-a empreitada para manifestar este princípio ele próprio de forma mínima
-resultou no \emph{minimum-fi.py}, código Python em um único arquivo curto que sintetiza
-músicas com segundo as estruturas especificadas em linha\footnote{Em música se fala da \emph{bula} (que equivaleria
-às funções em si e das especificações da linguagem) e a partitura ou 'música' propriamente ditas (que equivale à .}.
-Na versão atual, de 2012 mas adiantada em 2011, os algorítmos 
-em Python propriamente ditos somam 
-53 linhas e inclui 5 funções. Com estas funções, estruturas musicais podem 
-ser criadas padrão a padrão, nota a nota, amostra por amostra. Na prática e posto em
-linguagem cotidiana, resultam em notas que formam
-blocos e estruturas hierarquicamente superiores.
-
-Os princípios, bastante simples\footnote{Osvaldo Lacerda, em seu livro \emph{Compendio de Teoria Elementar da Musica} fala das propriedades do som musical de forma equivalente.}, são:
-\begin{itemize}
-  \item Deve-se ter um mecanismo de síntese sonora que
-possibilite a geração de unidades sonoras com diferentes timbres, controle sobre a frequência fundamental e duração específica\footnote{Também entra aqui o volume, mas como o que controlamos de fato é a intensidade relativa entre as notas, resolvemos por bem omitir
-esta parte até que fiquem claros os códigos relativos a isso, logo abaixo.}.
-  \item Deve-se ser capaz de construir não só unidades sonoras, mas também séries de unidades, sejam sobrepostas (acordes, por exemplo)
-  ou justapostas (melodias, por exemplo).
-\end{itemize}
-
-Para o primeiro item, se prestam comumente os procedimentos de busca em tabelas/vetores com formas
-de ondas em alta resolução (chamado \emph{lookup table}). O procedimento é barato e de qualidade alta
-(não acrescentam ruídos relevantes ao sinal).
-
-Para realizarmos a busca de forma eficiente, precisamos de tabelas que contenham cada uma
-uma forma de onda. As tradicionais são a senoide e as ondas quadrada, triangular e dente de serra.
-Mostramos a seguir estas formas de onda e também uma forma de onda retirada de um som real.
-
-De posse destas formas de onda, A seguir está o procedimento da lookuptable:
-
-\code{Procedimento de Lookup Table}{python_snippets/lookuptable.py}
-
-O primeiro ponto importante é a tabela em si, usualmente unidimensional. É de comum conhecimento que
-tabelas com 1024 são mais que suficientes para os usos musicais e a diferença
-de qualidade (relação sinal/ruido) não muda consideravelmente com a utilização
-de tabelas maiores. A seguir dispomos algumas destas tabelas, mais 
-especificamente, dispomos as tabelas para os formatos de onda mais usuais.
-
-\code{Formas de Onda Tradicionais}{python_snippets/formasDeOnda.py}
-
-Através da utilização do lookup sucessivo (procura-se um valor na primeira tabela
-e este valor indica o valor na segunda tabela a ser utilizado), executamos
-um waveshaping. Este procedimento é bastante apreciado pela simplicidade e eficácia
-na síntese de tímbres diversos e ricos em harmônicos e evolução temporal. Embora
-uma explicação exaustiva do waveshaping fuja ao escopo deste trabalho, este
-método se caracteriza pela aplicação de uma função não linear ao sinal de entrada.
-Neste caso o sinal de entrada é gerado pelo primeiro lookup, a função não linear aplicada
-é representada pela segunda tabela e aplicada pelo segundo lookup.
-
-Existem muitas formas de se executar waveshaping, a seguir segue o que usamos no minimum-fi, que se sustenta principalmente por ser leve e simples:
-
-\code{Waveshaping com Lookups sucessivas}{python_snippets/lookup_cruz.py}
-
-O segundo item - dos dois princípios expostos sobre o minimum-fi - presta-se à discretização do espaço musical. Unidades como batidas e notas
-tornam mais eficiente a comunicação pois a quantidade
-de estruturas sugeridas é maior e as estruturas são mais claras no discreto do que no contínuo~\cite{Roederer}. Roederer chega a
-apontar que as próprias notas dos instrumentos musicais são um reflexo de que é mais eficiente
-o uso do discreto do que do contínuo para a geração de estruturas musicais.
-
-De fato, unidades bem definidas se mostram úteis na prática musical 
-para fazer sequências de unidades, concatená-las. Quando as unidades
-são notas, as sequências de unidades justapostas no tempo são melodias ou linhas melódicas. As
-sequências sobrepostas no tempo são comumente pensadas como acordes, mas podem ser tidas simplesmente
-como sobreposições circunstanciais de duas linhas melódicas. Isso, claro, segundo
-a sistematização clássica e usual da música~\cite{Lacerda}\footnote{Vole assinalar aqui que a música do século XX apresentou diversos modelos teóricos que quebram com este entendimento simplificado sobre a música, suas unidades básicas e estruturas relacionadas}.
-
-As duas construções básicas explicitadas a seguir, baseadas na dicotomia melodia/harmonia
-(horizontalidade/verticalidade, justaposição/sobreposição), são
-as funções \emph{fazSequencia} e \emph{fazAcorde} no minimum-fi. Vale notar elas são absolutamente 
-equivalentes em uma análise puramente conceitual, i.e.
-uma delas pode ser omitida segundo algumas teorizações.
-Isso fica particularmente óbvio quando se nota que os procedimentos de mixagem e concatenação são
-plenamente capazes de realizar o que estas funções realizam. Aliás, as funções nada
-mais são do que usos típicos e quase caricatos destes procedimentos: no fazAcorde a mixagem
-sobrepõe no tempo todas as unidades, no fazSequencia as unidades são todas juntapostas no tempo.
-
-Como pode-se notar a seguir, as sequências de notas e os acordes, em última instância, são utilizações específicas das 2 funções de síntese sonora explicadas anteriormente: lookup e lookupcruz.
-
-\code{Realização de Sequências de Notas}{python_snippets/fazSequencia.py}
-
-\code{Realização de Acordes de Notas}{python_snippets/fazAcorde.py}
-
-A última das cinco funções utilizadas é uma soma amostra a amostra de dois sons. Para isso,
-é necessário completar com zeros a sequência com o menor número de amostras para somar rapidamente:
-
-\code{Somador (função auxiliar)}{python_snippets/somador.py}
-
-Depois disso é usufruir com estruturas criadas. Por exemplo: depois de construidas
-algumas tabelas para serem usadas como diferentes timbres, pode-se criar as
-escalas completamente simétricas na oitava cromática assim:
-
-\code{Escalas Completamente Simétricas (na grade dos 12 semitons e no âmbito da oitava)}{python_snippets/escalas_simetricas.py}
-
-% escala\_1=range(12) # cromática ascendente sem inclusão da oitava
-% escala\_2=range(0,12,2) # tons inteiros
-% escala\_3=range(0,12,3) # diminutão
-% escala\_4=range(0,12,4) # terças maiores
-% escala\_6=range(0,12,6) # trítonos
-% \end{python}
-% \input{py}
-
-Sendo cada unidade um semitom.
-Ou seja, cada unidade é o fator que fica presente na conta $f_0 .
-2^{(1/12)} .
-fator $ e resulta na frequência exata da nota relativa a $f_0$ em um sistema temperado ideal.
-
-Podemos também utilizar as usuais
-escalas tonais maior e menor natural, harmônica e melódica:
-
-\code{Escalas diatônicas tonais (maiores e menores)}{python_snippets/escalas_Mm.py}
-
-A utilização de intervalos menores que o semitom\footnote{Prática também conhecida
-como \emph{microtonalismo}, que gera/utiliza \emph{microtonalidade}.} mostra-se trivial em
-nossa implementação. Basta representar sequências de fracionários
-que são partes linearmente proporcionais dos semitons ou modificar o fator $f$:
-
-\code{Escalas Microtonais (com quartos e oitavos de tom e intervalos menores)}{python_snippets/escalas_microtonais.py}
-
-Da mesma forma, as triades maiores e menores
-são especificadas com simplicidade. Note que aos acordes \emph{diminutão} e
-\emph{aumentado} correspodem as mesmas notas das escalas simétricas de terças menores
-e maiores respectivamente:
-
-\code{Acordes anotados como listas em python}{python_snippets/acordes.py}
-
-E séries/sequências podem ser anotadas junto a variações:
-
-\code{Séries diversas}{python_snippets/series.py}
-
-Com este arcabouço, o passo seguinte é sintetizar e mixar, resultando
-em sequências musicais. A síntese de sequências e acordes são feitos tipicamente
-através de comandos desta forma:
-
-\code{Sintetizando sequências e acordes}{python_snippets/seqs_acordes.py}
-
-Já a síntese de estruturas compostas (p.ex. sequências de acordes e sobreposição
-de linhas melódicas), são feitas com os recursos usuais da liguagem. Neste caso
-usamos listas em Python e implementações em Numpy (mais eficientes em tempo de
-execussão e também na simplicidade do código) estão no apêndice. Estes mesmos
-procedimentos são praticamente os mesmos em Scilab, C/C++, Javascript, PHP, etc.
-
-A seguir demonstramos para fins didáticos, a construção de acordes periódicos em
-python puro\footnote{A implementação não didática destas três linhas de código
-podem ser agrupadas em uma só linha.}:
-
-\code{Acordes periódicos}{python_snippets/acordes_periodicos.py}
-
-Em posse destas sequências, acordes e recursos da linguagem,
-formamos estruturas hierarquicamente superiores
-através da concatenação de estruturas, da mixagem de estruturas, e da amplificação
-(ou atenuação) seletiva das mesmas:
-
-\code{Amplificação e mixagem}{python_snippets/amp_mix.py}
-
-Neste ponto, basta criarmos músicas e sequências de interesse estético ou para pesquisa.
-Dado o ferramental, os encadeamentos dependem de intensões estéticas, entendimentos musicais e estruturas abstratas que mantém a coerência e interesse em uma peça musical.
-Para o leitor mais interessado, deixamos para o Apêndice X um exemplo de música feita com o minimum-fi.
-A seguir utilizamos esta base apresentada para sintetizar estruturas musicais e então um EP.
-
-\vspace{10 mm}
-
-        \subsubsection{FIGGUS: FInite Groups in Granular and Unit Synthesis}
-
-Em coerência com os conhecimentos
-e tecnologias já apresentados nesta exposição\footnote{Especificamos o arcabouço para geração de sons musicais em unidades
-e estruturas compostas logo acima na subsessão sobre o minimum-fi.}, introduzimos agora
-um foco especial nas estruturas musicais
-desenvolvidas. São elas, inseridas em um momento histórico e
-executadas por instrumentos específicos com as técnicas de época,
-que constituem uma linguagem musical e músicas propriamente
-ditas[ref]. O FIGGUS constitui uma técnica composicional
-manifestada em software como ferramenta de síntese de
-estruturas musicais por estruturas matemáticas específicas
-para representação de simetrias.
-
-Este desenvolvimento foi iniciado em 2006 com o físico-matemático Prof. Adolfo Maia Junior - bem anterior
-ao nascimento do minimum-fi - para
-tratar de simetrias na música com vistas à composição musical através
-de métodos matemáticos\footnote{Duas iniciações científicas trataram do assunto, convenios XXX e YYYY.}. Mais especificamente, a proposta resultou em
-um programa voltado para a síntese
-granular e síntese de estruturas musicais através de Grupos Algébricos. O nome dado
-foi FIGGUS, sigla de FInite Groups in Granular and Unit Synthesis\footnote{Também utilizamos
-o nome FIGGS (FInite Group in Granular Synthesis) dado que o termo \emph{unit synthesis} não
-é usual na literatura. Posteriormente o primeiro autor deste trabalho recorreu novamente
-ao uso do nome FIGGUS. Isso foi motivado principalmente pelo fato de que
-o maior uso da técnica é para síntese de estruturas musicais, não para
-síntese de amálgamas sonoros (ou timbres mesmo) tipicamente resultantes da síntese granular}.
-
-Na atual reescrita, embora ainda bastante atrás do FIGGUS original quanto
-à interface gráfica, a ferramenta opera diretamente em Python puro,
-com as biblitecas imbutidas por padrão. Isso permite com que o FIGGUS
-sintetize todo um EP usando somente os comandos:
-
-\code{Utilizando o FIGGUS para Sintetizar um EP}{python_snippets/synth_ep.py}
-
-Desta forma, a ferramenta fica muito mais simples para experimentações
-e implementações adicionais pois sua utilização se dá através
-do próprio código no qual tudo é especificado. Esta abordagem se mostra particularmente
-pertinente pois a estagnação do mecanismo de síntese de estruturas musicais
-beira a inviabilização do uso artístico ou torna necessário o uso
-de recursos externos.
-Outro uso pretendido para o FIGGUS
-é a síntese de tímbres e amálgamas sonoros através da
-Síntese Granular. Embora nosso foco seja outro, cabe algumas breves
-palavras sobre o assunto.
-
-A Síntese Granular é uma área bem estabelecida tanto na acústica quanto
-na Computação Musical e se caracteriza pela geração de sons bastante curtos
-e em quantidade massiva. Tipicamente os sons possuem entre 5 e 40 milissegundos
-e a quantidade destes \emph{microsons}\footnote{Grãos sonoros e microsons são jargões
-típicos da síntese granular usados para indicar sons com as durações especificadas
-no texto.} pode chegar a milhares por segundo. O tratamento específico da
-síntese granular foge ao escopo deste trabalho e indicamos ao leitor interessado
-os artigos produzidos sobre Síntese Granular e Teoria de Grupos[refs AES e SBCM].
-Assim, focamos o texto a seguir em Grupos Finitos e Síntese de Estruturas Musicais.
-
-Nas artes é de comum conhecimento o papel absolutamente central
-que as simetrias possuem. Na música, para citar somente alguns exemplos simples,
-temos os numerosos estudos de simetrias na música de J. S. Bach, os jogos
-de dados de Mozart e os usos recorrentes da proporção áurea na música
-de Béla Bártok. Matematicamente, as \emph{simetrias} são descritas por Grupos,
-e estes são definidos como sendo um conjunto (chamemos de $G$)
-munido de uma operação (seja $\bullet$), formando um grupo $(G,\bullet)$
-satisfazendo as seguintes propriedades:
-
-\begin{enumerate}
-    \item Fechamento: $\forall \ g_1, g_2  \in G \Rightarrow g_1 \bullet g_2 \in G$
-    \item Associatividade: $(g_1 \bullet g_2) \bullet g_3 = g_1 \bullet (g_2 \bullet g_3), \forall \ g_1, g_2, g_3 \in G$
-    \item Existência do elemento neutro: $\exists \ e \in G : g \bullet e = e \bullet g = g$
-    \item Existência dos inversos: $\forall \ g \in G \ \exists \ g^{-1} : g \bullet g^{-1} = g^{-1} \bullet g = e$
-\end{enumerate}
-
-Adicionalmente, vale complementar que se um grupo tiver um número
-finito de elementos, ele é dito finito, caso contrário é chamado
-infinito. Todos os elementos de um grupo finito são cíclicos pois
-operado com si próprio um número suficiente de vezes resulta em si
-mesmo. Um grupo é dito comutativo ou abeliano caso a propriedade
-comutativa seja satisfeita para todos os elementos:
-$\forall \ g_1, g_2 \in G \Rightarrow g_1 \bullet g_2 = g_2 \bullet g_1$.
-
-São numerosas as aplicações de grupos e estes
-usos para as artes possam se desenvolver
-de diversas formas focadas em simetrias
-e com interesse estético, em
-nossas pesquisas uma delas se mostrou particularmente coerente
-e pertinente: as permutações de unidades sonoras. Primeiro,
-através do Teorema de Cayley, todo grupo finito é isomorfo
-a um grupo de permutação\footnote{Mais especificamente,
-todo grupo é isomorfo a um subgrupo do grupo simétrico agindo
-em G.}, i.e. seus elementos são diretamente relacionados
-a permutações. Além disso, na música as permutações estão
-no núcleo de diversos procedimentos canônicos como a
-retrogradação e outras técnicas de variação.
-
-[Expor sobre uma ou duas músicas feitas com o FIGGUS]
-
-No FIGGUS, implementamos o grão sonoro (no caso mais usado
-como unidade) como uma classe que possui apenas
-os seguintes atributos (e não possui funcões): duração (segundos),
-frequência (Hz), timbre (identificador para usar mediante implementações convenientes), intensidade (pico $\in \ [0,1]$), e duração dos fades (in e out em segundos). Aqui a classe:
-
-\code{Grão Sonoro Básico}{python_snippets/fgrao.py}
-
-O FIGGUS funciona com base em uma sequência de grãos posta de antemão
-na qual operam as permutações. Além da sequência de grãos,
-é conveniente também manter o número de grãos da sequência e uma
-lista com os índices dos grãos para executar as operações necessárias para gerarmos as músicas (ver abaixo). A classe utilizada para representar as sequências é assim:
-
-\code{Sequência de Grãos}{python_snippets/fsequencia.py}
-
-Aos grãos em sequência neste ponto são aplicadas permutações.
-Para isso é bastante conveniente representar as permutações em classesseparadas e os padrões de aplicação destas permutações para facilitar
-desenvolvimentos com clareza.
-No caso a classe de permutações só possui os parâmetros da permutação em si e do tamando dela (número de elementos que ela utiliza).
-É conveniente implementar uma tradução da notação cíclica para a notação direta da permutação e é uma das facilidades da implementação por classe que fizemos.
-A classe do padrão de permutação possui também um período de aplicação da permutação, ou seja, de quantas em quantas leituras da sequência aplicamos a permutação.
-As classes ficaram assim:
-
-\code{Permutações e Padrões de Permutações}{python_snippets/fpermutacoes.py}
-
-De posse dos grãos, da sequência, das permutações e do padrão de permutações estamos possibilitados de realizar a estrutura musical em si.
-Bastam adicionalmente, especificar o número desejado de iterações da sequência.
-Com isso, a sequência de grãos é lida um número de vezes, aplicando as permutações na sequência segundo o padrão de aplicação das pemutações, de forma a resultar em uma sequência sonora musical.
-Note que se a permutação usar menos elementos que a sequência possui, alguns destes elementos ficarão estáticos nas iterações da sequência no padrão sonoro.
-
-Na síntese do vetor sonoro relacionado, é também crucial estabelecer a taxa de amostragem preterida, como podemos perceber na função de síntese dos vetores, da classe Pattern que cuida desta realização do padrão em si.
-Deixamos a seguir esta função, de forma que faltará apenas a escrita destes vetores sonoros em arquivo de áudio, como veremos a seguir.
-
-\code{Realização do Padrão Musical}{python_snippets/fpadraosonoro.py}
-
-Em posse não apenas de representações abstratas do padrão musical a ser realizado, mas dos próprios vetores sonoros relacionados à representação digital da música a ser realizada, podemos escrever este vetor como um arquivo de áudio propriamente dito.
-O mais conveniente neste caso é escrever um arquivo PCM (Pulse Code Modulation) em padrão amplamente utilizado e reconhecido.
-Ambos os padrões WAV e AIFF satisfazem estes requisitos e escolhemos o WAV por ser o padrão de CDs (Compact Discs) e também o mais utilizado por programas de áudio.
-Mais especificamente, o padrão de CDs é WAV com 44100 amostras por segundo e 16 bits por amostra\footnote{No formato WAV de CD, cada amostra é representada por uma variável inteira 'signed' de 16 bits.}.
-As amostras dos vetores sonoros são calculados, por conveniência e por convenção, no âmbito $[-1,1]$ e precisam ser normalizados para o âmbito $[-32767,32768]$ e truncados em números inteiros.
-Depois disso precisam ser escritos em um arquivo com os bits como na convenção da linguagem C/C++.
-A bibioteca \emph{struct} cuida dessa escrita do inteiro no formato correto, e a biblioteca \emph{wave} escreve o cabeçalho no formato WAV adequado.
-Assim, a clase de escrita do vetor sonoro em arquivo comum fica assim:
-
-\code{Escrita do Vetor Sonoro em Arquivo WAV}{python_snippets/fio.py}
-
-  \subsection{Música em Tempo Real: Livecoding e ABeatTracker (ABT)}
-
-Com os avanços computacionais recentes, tornou-se usual a síntese sonora
-em tempo real. Com isso, surgiram liguagens dedicadas para o áudio e a música
-em sua maioria dedicadas - ou ao menos capacitadas - para o uso 
-em tempo de execussão [citar PD, SC, ChucK]. Em outras palavras, estas linguagens
-possibilitam que o usuário ouça o resultante sonoro do código utilizado e altere
-o código com resultado imediato no processamento e resultado que escuta.
-
-Nesta linha de exploração musical do código, iremos expor a seguir sobre
-nossas investidas em \emph{Livecoding} (escrita de código em tempo real
-com vistas à performance pública) e o ABeatTracker (uma linguagem por macros
-para execussão sonora rítmica em conjunto com instrumentos tradicionais e outras
-fontes sonoras/musicais externas).
-
-        \subsubsection{Livecoding}
-
-Recentemente, grupos de ponta em música experimental no mundo todo estão
-desenvolvendo apresentações musicais públicas baseadas na escrita
-de código ao vivo. Usualmente, se projeta o código para que a audiência possa
-ver o que está sendo escrito, no rítmo em que se escreve, e se projeta também
-o resultante sonoro por autofalantes.
-
-As motivações para isso são variadas. Expomos a seguir de forma topificada
-um condensado do que nos orientou a tal prática. Vale ressaltar que são
-motivações igualmente presentes em outros grupos, embora não necessariamente
-em todos ou da mesmícima forma.
-
-\begin{itemize}
-    \item A performance musical por computador carece de recursos performáticos
-    à altura das execussões com instrumentos tradicionais. Os gestos são por demais
-    discretos e a concentração do performer é bastante focada na tela do computador.
-    \item O feedback auditivo do código projetado permite que o espectador infira
-    significados dos códigos. Mesmo que de forma superficial, este recurso do livecoding
-    é em muito capaz de desmistificar a programação de computadores, por muito considerada
-    completamente intangível.;
-    \item O código em si é um recurso poderosíssimo, que permite ao usuário controlar
-    os sons produzidos amostra por amostra ou em escalas maiores, como notas, compassos, fraseados
-    inteiros ou mesmo em escalas maiores de tempo, como minutos, horas, dias e semanas.
-    \item O compartilhamento do código é usual, leve e eficiente como entrega completamente
-    aberta da tecnologia relacionada à proposta estética.
-\end{itemize}
-
-Desta forma, iniciamos em 2011 uma linha de atuação com Livecoding que resultou em uma performance
-no \emph{V Festival Contato}. Utilizamos a linguagem ChucK por apresentar os recursos que
-consideramos mais apropriados, embora de forma alguma isso seja consensual na prática atual
-de livecoding.
-
-A apresentação contou com recursos adicionais para agregar interesse, como a utilização
-do \emph{cowsay} (para enviar mensagens enquanto se desenrolava a música) e trajetórias
-de um ponto no fundo do código para sugerir a vertigem do sono REM. Estes recursos
-podem ser vistos em uso claramente nos videos demonstrativos [citar videos do vimeo].
-As linhas do cowsay e o script em processing relacionados estão no Apêndice XX...
-
-Duas pessoas executaram livecoding simultaneamente. A saber, Vilson Vieira executava
-rítmos, batidas bastante marcadas que serviam como base. Renato Fabbri, autor do presente
-trabalho, executava linhas fluídas quasi melódicas que formavam arcos maiores. Nos intervalos
-das investidas no formato citado, havia interlúdios em que ambos se revesavam com músicas
-curtas e inusitadas, como em um duelo[disponibilizar músicas em links e colocar de referencia].
-
-[Colocar os códigos meus e do Vilson e explicar o funcionamento e uso]
-
-      \subsubsection{ABeatTracker (ABT)}
-O ABT é uma linguagem/ferramenta que dispara linhas rítmicas através de macros que especificam
-as células rítmicas, amostras sonoras que são utilizadas como conteúdo sonoro destas
-linhas, modos de leitura destas amostras sonoras, e variáveis randômicas utilizadas
-para execussão da linha. Além disso, o ABT dispõe de variáveis globais que podem ser alteradas
-a qualquer momento pelo usuário, como BPM, velocidade de leitura das amostras e variáveis
-randômicas globais (que se somam às individuais).
-
-Em pouco explicitaremos a forma de utilização do ABT. De antemão, é pertinente apresentar
-uma pequena discussão a respeito do que o ABT é considerado e sobre os propósitos desta
-ferramenta. Em primeiro lugar, vale manter em mente que o ABT tem um funcionamento específico
-que pode transcender o que se conhece por livecoding. As macros pré-estabelecidas engendram um
-conjunto de recursos pré-estabelecidos bem definido, o que contrasta com a ideia de uma 'linguagem
-de programação' que tenha capacidades mais amplas. De qualquer forma, linguagens com domínios
-específicos não são raras e por vezes o ABT foi descrito como uma linguagem.
-
-Sobre os propósitos do ABT, em primeiro lugar ele se dispõe a ser um instrumento computacional
-essencialmente rítmico. Junto a esta proposta, vem a necessidade da utilização em conjunto com
-outros instrumentos, externos ao ABT, ao computador em que estiver rodando e possivelmente externo
-com relação a qualquer computador. Para isso foi elaborado o ABD (ABeatDetector), no qual o usuário
-tamborila os rítmos que estiver ouvindo ou imaginando para que o ABT sincronize o pulso e utilize
-células rítmicas relacionadas. A análise feita pelo ABD resulta em uma série de rítmos explicitados
-por sequências de compassos que encapsulem durações regulares do rítmo tamborilado. Estes rítmos
-relacionados ao tamborilar do usuário são chamados de harmônicos e podem ser selecionados
-prontamente para o disparo de linhas melódicas.
-
-No Apêndice XX está o manual de utilização do ABT e todo o código do ABT (e do ABD) está
-disponivel online (citar repos). Exploramos abaixo detalhes relevantes da implementação,
-em especial a forma de funcionamento do ABD.
-
-[códigos-chave e explicações sobre o ABT e o ABD]
-
-  \subsection{Música na Matéria: EKP e AHT}
-
-Embora o foco deste trabalho seja na exploração musical através de códigos (abertos!),
-nesta sessão iniciamos uma explicação simples, clara e factual de como estas
-investidas transcendem o código e até mesmo a música em si.
-
-Mais relevantes que os desenvolvimentos em si, são as mobilizações criadas nos
-entornos e os engajamentos. Isso ficará evidente nas próximas - curtas - sessões
-destes desenvolvimentos e resultados.
-
-Aqui apresentamos dois trabalhos que geraram alguma movimentação de pessoas,
-resultando em reuniões, desenvolvimentos, pesquisas e apresentações propriamente
-ditas. O primeiro utiliza o estado do hardware como entrada, o segundo se trata
-de uma mesa escultural para flutuação de origamis que resulta em um instrumento musical
-bastante lúdico.
-
-      \subsubsection{Emotional Kernel Panic (EKP)}
-
-Em 2008, colaborando intensamente com o CDTL
-(Centro de Desenvolvimento de Tecnologias Livres)\footnote{foi uma associação civil formada e desmembrada em 2008 e sediada em Recife, PE}
-foi lançada a ideia de utilizar o estado do sistema operacional - especialmente o kernel linux - para
-geração de sons. Surge o Emotional Kernel Panic (EKP) na colaboração excepcional de
-Felipe Machado\footnote{Importantíssimo para o desenvolvimento da Cultura Digital no Brasil [citar fontes e programas governamentais]},
-Ricardo Brazileiro\footnote{Artivista Digital bastante ativo [citar fontes]} e o primeiro autor do presente trabalho.
-
-Desde o inicio, foram definidos três finalidades
-para esta exploração do SO:
-
-[visitar o README do EKP e achar os scripts do Brazileiro sobre o EKP]
-
-\begin{itemize}
-    \item Didáticos
-    \item Artísticos
-    \item Monitoramento do SO
-\end{itemize}
-
-Toda a base do EKP está aqui: http://trac.assembla.com/audioexperiments/browser/ekp-base
-
-[Patches desenvolvidos por Brazileiro/Machado e rascunho do EKP-Monitor]
-
-      \subsubsection{AirHackTable}
-
-A AirHackTable é um instrumento musical eletrônico controlado por origamis (dobraduras de papel), construída na forma de uma mesa.
-Nela, uma rede de coolers reciclados faz flutuar origamis de geometria e cores variadas.
-Os movimentos dos origamis são captados por webcam e interpretados em tempo real por software de processamento de imagens, gerando padrões que controlam a transformação sonora da música.
-Dessa forma, pode-se dizer que os sons gerados refletem o voo dos origamis de acordo com suas geometrias (que geram trajetórias de voo caracteristicas).
-
-Na versão atual da AirHacktable, cada cor (vermelho, amarelo, azul, verde, preto, ou branco) controla uma voz, e a posição do origami na mesa modula aspectos do som.
-Em outras palavras, se o origami está flutuando mais à esquerda, o som sai no canal à esquerda da mesa, e se o origami está mais próximo da câmera, o volume aumenta, e se está mais afastado do operador, o som se torna mais agudo.
-
-[imagens da AHT e dos patches em operação]
-
-   \subsection{Música no Tecido Social: Sabrina Kawahara, Audioexperiments, EstudioLivre.org, CDTL, juntaDados.org, Devolts.org, MSST, LabMacambira.sf.net}
-
-Esta é uma sessão menor, dedicada a apontar repercussões emergentes destas empreitadas em comunidades diferentes
-e então tornar compreensível o desdobramento que estes códigos dedicados à música tiveram em processos sociais
-e mobilizações civis.
-
-Grupos foram montados em torno destes desenvolvimentos. Como os propósitos de compartilhamento
-e apropriação tecnológica estavam no cerne dos grupos relacionados, as investidas
-naturalmente tomaram teores engajados socialmente. A questão do empoderamento das pontas
-e da criação de um patrimônio tecnológico da humanidade é consequência quase imediata
-das posturas de compartilhamento e apropriação citados.
-
-Discorremos brevemente sobre cada uma das iniciativas citadas:
-
-\begin{itemize}
-    \item Sabrina Kawahara
-    \item Audioexperiments
-    \item Estudiolivre.org
-    \item CDTL
-    \item JuntaDados.org
-    \item Devolts.org
-    \item MSST
-    \item LabMacambira.sf.net
-    \item Outros relacionados: MuSa, Metareciclagem, Submidialogia, Tainã
-\end{itemize}
-
-\section{Web}
-
-  Difusão de informação com ênfase na facilitação
-  da apropriação de tecnologias e de instancias políticas.
-
-   \subsection{Tecnologias sociais de alta demanda: Sitios, Conteúdos e Articulação}
-
-      \subsubsection{Sítios}
-
-      FDDCA
-
-      Ferramenta de comunicação
-
-      (Cadastro dos pontos?)
-
-      AA, SOS, Catalogo de Ideias, etc
-
-      Meu site pessoal
-
-      \subsubsection{Conteúdos}
-
-      Wiki?
-
-      \subsubsection{Articulação}
-
-      IRC, Emails
-
-\subsection{Disponibilização e desenvolvimento conjunto: wikis, etherpads, AA, Trac, IRC ..}
-
-\subsubsection{Wiki}
-
-\subsubsection{Trac}
-
-\subsubsection{Screencasts - Vimeo}
-
-\subsubsection{AA}
-
-\subsubsection{Audio Experiments (Æ)}
-
-\subsubsection{IRC}
-
-\subsubsection{Etherpads}
-
-\subsubsection{Outras fontes}
-
-\section{Materiais didáticos}
-
-  \subsection{Tutoriais em texto e código: Filtros, Nyquist e plugins LADSPA}
-
-Os vários materiais didáticos produzidos constam no apêndice
-deste trabalho. Um único destes será exposto a seguir por sua
-capacidade de agregar os conteúdos dos capítulos anterioes:
-os tutoriais de filtros e amostragem.
-
-\begin{itemize}
-    \item {\bf Tutorial de python para áudio e som}
-
-Este tutorial foi levado para Berlim no LAC 2007 e sofreu melhoras desde entao. Esta
-primeira versao ficou resumida em forma de texto no EL\footnote{http://estudiolivre.org/python-e-som-tutorial}. Em 2010
-a Associacao Python Brasil escolheu este trabalho, então já mais amadurecido, para ser apresentado no
-FISL em Porto Alegre. Como consequencia, foi feita uma série de video-tutoriais bastante utilizados\footnote{http://estudiolivre.org/tiki-index.php?page=Video+Tutoriais}.
-Este tutorial foi comentado em listas em que o autor não participa (e outras em que o autor participa).
-
-    \item {\bf Tutoriais de filtros e amostragem via python}
-
-Voltados para explicitar principios fundamentais de áudio, estes tutoriais
-são baseados código Python e o equivalente em C. Pequenas explicações são
-dadas com o intuito de orientar a exploração inteligente destes \emph{snippets}.
-
-\emph{Teorema de Amostragem}: estes scripts visam a experimentacao inteligente com
-o Teorema de Nyquist. (descricao)
-
-\emph{Filtros}: alem da explicitação sobre as diferenças entre filtros FIR e IIR,
-duas utilizações clássicas destes filtros estão implementadas: Wavelets (FIR) e Quad (IIR)
-
-Estes códigos podem ser baixados no repositório SVN do AudioExperiments. E os textos estão
-na wiki (nos digitais ou EL, recriar pois os CDTL foram apagados)
-
-    \item {\bf Tutorial de plugins lv2}
-
-Dadas as dificuldades que o desenvolvimento dos \emph{plugins} de áudio apresenta,
-desenvolvi um tutorial passo a passo com plugins que rodam em todas as etapas.
-Ele é baseado em uma interface C++ para este padrão de plugin que eh implementado
-em C. Os códigos e os textos estão todos em repositório.
-
-    \item {\bf Microtutoriais Django ~\cite{dmicrotuts}}
-
-Estes 'microtutoriais' são baseados nos conceitos de \emph{scripts mínimos} e
-\emph{alterações puntuais}. O primeiro conjunto de microtutoriais é dedicado
-a reconstruir o tutorial oficial do django de forma condensada e não prolixa.
-O segundo destes conjuntos é dedicado a instrumentalizar de fato o leitor com
-o entendimento do funcionamento dos princípios fundamentais deste framework.
-
-    \item {\bf Philosometrics}
-
-Embora este não seja um trabalho didático propriamente dito, ele tem este intuito
-no cerne de sua concepção e surgimento. Em decorrência dele, surgiu o  Musimetrics,
-o Cinemetrics e o Literametrics. Além disso, ele é um belo exemplo da
-utilização das ciências duras para a análise de ciências humanas e foi acolhido
-como tal em alguns momentos.
-
-    \item {\bf Carta mídias livres}
-
-Texto criado em decorrência da participação da comissão de seleção no
-'Prêmio Mídias Livres', a convite do Ministério da Cultura por 'notório saber'.
-Esta carta é um documento único no seu conteúdo, deixando às claras
-o conceito de Mídias Livres como mídias não aprisionadas pelo conceito
-de propriedade, ou seja, que priorizam a sua livre circulação e a possibilidade
-de geração de materiais derivados, assim como sua geração aberta ao colaborativo e comunitário.
-
-    \item {\bf Textos de cunho sociológico, transformador}
-
-Produção mais numerosa que as anteriores, se caracteriza por métodos não convencionais
-de abordagem dos assuntos e de escrita. Em especial utiliza-se pseudônimos para
-auxiliar a despersonificação, gerando textos menos presos à satisfação da auto-imagem, dente
-outras qualidades. A utiliação de psudônimos é um costume muito apreciado em diversos meios,
-e as pesquisas tem confirmado as vantagens que a prática apresenta e confirma\footnote{http://disqus.com/research/pseudonyms/}.
-
-O autor destes textos se dá ao direito de não revelar seus psudônimos - embora muitos deles
-sejam publicamente conhecidos - para conservar as consequências desta prática na
-forma mais pura. Como comprovante desta produção, deixamos uma mensagem sobre a publicação
-de textos em mídia impressa com autores internacionais,
-confirmando a participação do autor desta dissertação, mas cujo
-nome não consta na publicação.
-
-!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-
-de      fabi borges catadores@gmail.com por  riseup.net 
-responder a     submidialogia@lists.riseup.net
-para    submidialogia@lists.riseup.net
-data    23 de agosto de 2011 12:26
-assunto Re: [submidialogia] livro sub- publicação
-lista de e-mails        <submidialogia.lists.riseup.net> Filtrar as mensagens dessa lista de e-mails
-enviado por     lists.riseup.net
-assinado por    riseup.net
-cancelar inscrição      Cancelar a inscrição para essa lista de e-mails
-        Importante principalmente porque você frequentemente lê mensagens com esse marcador.
-ocultar detalhes 12:26 (6 minutos atrás)
-entao, eu fui recebendo textos durante esse tempo,
-alguns tao atrazados como dos sem satelites, mas muita gente mandou;
-
-aqui os autores:
-
-os internacionais nao sao muitos, o joni kempf (o que bebe ouro do hardware), o barbrook (futuros imaginarios),
-Hamdy heda (da revolucao egipcia), o pedro soller (summerlab), maria llopis (pos porno),  talvez a bronac, nao entregou ainda.
-
-dos brasileiros, renato fabri, ruiz, pasteur, morgana e caio, ju dornelles, mari marcassa, coletivo errorista, adriana velozo-drica, tiago pimentel, felipe fonseca, thiago novaes, lelex, felipe ribeiro(?), maira, verenilde,  bartolina silva, poro, vitoria amaro, fabib (eu),
-
-entao, precisa publicar agora,
-uma equipe para publicacao e,,,
-
-bjs
-f
-
-!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-
-\end{itemize}
-
-\subsection{Screencasts e outros materiais em video e em texto}
-
-\begin{itemize}
-    \item Python para áudio e música
-	  Texto - palestras - videos
-
-    \item Canal Macambira
-No Macambira estão sendo produzidos materiais em screencasts sobre
-diversas cenas de hackeamento.
-
-    \begin{itemize}
-	\item Live-Coding
-	\item Raspagem de dados
-    \end{itemize}
-\end{itemize}

### 2012-11-05 musicaExtra.tex
primeira limpada braba para versao final do msc

\chapter{Música digital em domínios não digitais}
\label{cap:musicaExtra}

Os resultados explicitados nesta dissertação estão focados no processamento de 
áudio através de códigos em liguagens de programação de computadores e
voltado para a música e na síntese de estruturas musicais. O tema é amplo
e pertinente o suficiente para contar atualmente com ampla bibliografia e um histórico
que remonta desde a própria história da computação e das linguagens de programação.
Desta forma, entendemos cabível focar este texto nas empreitadas realizadas pelos autores
da presente dissertação, com foco nas realizações do primeiro autor. São usos
reais destas tecnologias, manifestadas de forma cultural e com repercussões
sociais, como poderá é relatado neste mesmo capítulo.

Tais realizações possuem um espectro amplo
dentro do escopo proposto e constitui um percurso relativamente completo
pelas possibilidades e decorrências da abordagem da música pelo código computacional\footnote{ao menos dentro do se denomina 'Cultura Livre', veja o primeiro capítulo do presente escrito para maiores informações}.

Estaremos abordando experimentos abertos em áudio, abordagens musicais pelo código, incluindo música em tempo real, diferido, na matéria e repercussões no tecido social. De forma a relatar os resultados destas investidas, segue uma sessão sobre ações coletivas realizadas através das mobilizações que estes códigos criaram, e as estruturas virtuais mantidas na intenet relacionadas a estas mobilizações e ações. Por fim, alguns materiais didáticos criados pelo autor - que possuem algum papel emblemático - são também expostos por comporem o arsenal de repercussões desta investida no em som e código.

  \section{Experimentos abertos em áudio: LADSPAs, Wavelets e Redes Complexas}

Todos os desenvolvimentos desta dissertação estão em repositórios abertos\cite{repositorios-tese-dev}.
Alguns foram especialmente importantes como percurso para o que é apresentado neste
escrito. Ou seja, são explorações que manifestaram diferentes aspectos
do trabalho do código voltado para o áudio e a música.

O código para a arte sonora, incluindo a musical, manifesta-se como cultura pois é fruto de práticas
espontâneas, diárias e coletivas.
\footnote{As chamadas culturas biopunk, ciberpunk, cipherpunk, hacker, digital e outras mais,
possibilitadas aos recentes desenvolvimentos em telecomunicação, dizem respeito em menor
ou maior grau à produção de código como cultura.}

A seguir apresentamos alguns bons exemplos dos desenvolvimentos
desta dissertação especificamente em áudio, sem o envolvimento direto da música. São
códigos de necessidade para música, mas de uso mais geral e cuja implementação
requer muito mais de processamento de sinais e computação do que de música propriamente dito.

      \subsubsection{Plugins LADSPA e lv2}
      [repos AE, LM, wiki EL, historico CDTL]
LADSPA (Linux Audio Developers Simple Plugin API) é a API livre\footnote{Veja o Capítulo 1 para a definição de livre neste contexto.} estabelecida de plugins de áudio até a presente data. A última versão liberada desta API é a 1.1, em uso corrente até hoje e é de 2002 segundo os arquivos de cabeçalho relacionados.

A exploração de APIs específicas foge ao escopo do presente texto e os códigos completos podem ser encontrados pela rede, junto a tutoriais e plugins de exemplo. O código destes desenvolvimentos e pertinente ao áudio são relativos à síntese e remoção de ruídos. Da síntese, exitem abordagens mais puristas, que no caso sintetiza ruído branco e filtra para resultar nos espectros desejados (veja logo abaixo para maiores especificações). Uma abordagem menos purista porém mais eficiente é a de usar uma amostra curta do ruído e reproduzí-la indefinidamente através de \emph{cross-fades}.

Já na remoção de ruídos as abordagens são as mais variadas e extremamente dependentes da aplicação. A complexidade dos algorítmos pode atingir níveis de especialidade em processamento de sinais que merecem um trabalho dedicado. Aqui iremos expor somente sobre remoção de ruído \emph{'Hum'} causado usualmente pela corrente alternada que alimenta os equipamentos utilizados.

Abaixo dispomos explicações breves sobre os diferentes ruídos e os códigos relacionados. Preferimos expor em Python os códigos de síntese
por questão de clareza e coerência com o resto do texto\footnote{Os códigos em C/C++ e a implementação como plugin lv2 está no repositório git LV2 como pode ser visto em: http://labmacambira.git.sourceforge.net/git/gitweb-index.cgi}.

O procedimento de reprodução do ruído através de uma amostra repetida e em crossfade é a seguinte:

[código da reprodução do ruído em loop e croos-fade ~10 linhas]

Sintetizando diferentes ruídos\footnote{Dado o propósito do texto e da implementação, utilizamos nomenclaturas utilizadas na música para os ruídos, como se pode notar.}:
\begin{itemize}
    \item Ruído Branco: 
    \item Ruído Rosa:
    \item Ruído Marrom:
    \item Ruído Violeta:
    \item Ruído Azul:
    \item Ruído Violeta:
    \item Ruído Cinza:
    (nao oficiais)
    \item Ruído Vermelho:
    \item Ruído Laranja:
    \item Ruído Preto:
    \item Ruído Branco Ruidoso:
    \item Ruído Preto Ruidoso:
    \item Clicks:
\end{itemize}

A remoção de ruído Hum é baseada em uma sequência de filtros notch (ver fundamentos) idealmente dispostos nos harmônicos da frequência de oscilação da corrente elétrica. Cada um destes filtros deve permitir ajustes finos no fator de qualidade e também na frequência central pois trata-se de um sistema real passivel dos mais diversos efeitos de distorção do modelo ideal. Abaixo expomos o código em Python para tal \emph{removedor de ruído} como é chamado. O código C/C++ e a implementação como plugin está no mesmo repositório disponibilizado para a síntese de ruídos.

[código e explicação do remodedor de ruído Hum]

\section{Áudio e Música}

Por razões didáticas, dividimos este capítulo em 4 partes: música em tempo diferido (ou seja, que não é feito em tempo real),
música em tempo real,  música na matéria (suporte físico em hardware) e música no tecido social
(considerando especialmente as mobilizações humanas relacionadas).

Com isso desejamos expor a prática musical através do código
com exemplos reais de aplicação e em uso pelo autor, por membros do LabMacambira.sf.net,
por parceiros, colaboradores e por usuários eventuais das naturezas mais diversas.

  \subsection{Música em Tempo Diferido: Minimum-fi e FIGGUS}

\begin{quotation}
\small
'The increasing dominance of graphic interfaces for music software obscured 
the continuing presence of the command-line tradition, 
the code writer, the hacker. The code writing of deferred time 
computer programming may be assembled out of time order, debugged and optimized.'

\emph{Simon Emmerson, Living electronic music, 2007}
\end{quotation}

A realização musical em tempo diferido é o paradigma inicial da música computacional.
Iniciando com o Music V, a proposta foi depois desenvolvido com o CSound. Pode-se dizer
que até hoje é a forma como compositores usualmente pensam a música: pensando
e escrevendo as estruturas, que depois são executadas por instrumentistas ou aparelhos eletrônicos.

Assim como a composição instrumental permite um trabalho mais minucioso do que
a improvisação instrumental, a realização musical em tempo diferido usualmente permite um
detalhamento maior dos procedimentos do que a realização em tempo real. Por este
mesmo motivo, trataremos inicialmente de dois trabalhos em tempo diferido.
Aliás, como veremos a seguir, as abordagens completam uma base para a música
computacional praticada até os dias de hoje.

O som musical pode ser caracterizado físicamente, e sua síntese digital com base em nossa banda de audição e no teorema de Nyquist, sintetiza uma música amostra por amostra, através de princípios
claros de síntese sonora e organização musical\footnote{Veja no APÊNDICE XX: o \emph{minimum-fi} hi-fi, low-fi, minimum-fi, ou seja,
o mínimo de qualidade para assegurar existência e consistência
da criação}. O segundo, chamado de \emph{FIGGUS} (FInite
Groups in Granular and Unit Synthesis), utiliza os princípios do minimum-fi e constitui
um módulo Python completo, cuja proposta é a utilização de simetrias, através de
permutações e de Teoria de Grupos, para a composição de músicas. Como demonstração
das capacidades do FIGGUS, ele gera um EP\footnote{Extended Play, uma album musical maior que um single mas menor que um LP (Long Play) inteiro} com um único comando. Este é o
\emph{PPEPPS}\footnote{Pure Python EP: Projeto Solvente}, como veremos a seguir.

Desta forma, esta sessão exemplifica e explicita - através de dois exemplos reais - os
princípios do uso de código para a síntese \emph{musical}, desde as amostras
relativas a uma nota com dada frequencia, amplitude e timbre, até a confecção
de uma ferramenta derivada, já incorporando propostas musicais e estruturas
mais elaboradas.

      \subsubsection{Minimum-fi}

Existe uma perene matiz estética e também tecnológica
de realizar uma dada tarefa com \emph{o mínimo} necessário.
Esta matiz é para a música igualmente
fundamental e reconhecida como
princípio de unidade e coerência\footnote{Este princípio tanto é fundamental
que as escolas musicais possuem técnicas específicas, músicas possuem suas
próprias convenções mantidas por toda a sua duração, os arcos mantém características,
enfim, podemos até mesmo concluir que a simetria define o escopo.}. Em código computacional
a empreitada para manifestar este princípio ele próprio de forma mínima
resultou no \emph{minimum-fi.py}, código Python em um único arquivo curto que sintetiza
músicas com segundo as estruturas especificadas em linha\footnote{Em música se fala da \emph{bula} (que equivaleria
às funções em si e das especificações da linguagem) e a partitura ou 'música' propriamente ditas (que equivale à .}.
Na versão atual, de 2012 mas adiantada em 2011, os algorítmos 
em Python propriamente ditos somam 
53 linhas e inclui 5 funções. Com estas funções, estruturas musicais podem 
ser criadas padrão a padrão, nota a nota, amostra por amostra. Na prática e posto em
linguagem cotidiana, resultam em notas que formam
blocos e estruturas hierarquicamente superiores.

Os princípios, bastante simples\footnote{Osvaldo Lacerda, em seu livro \emph{Compendio de Teoria Elementar da Musica} fala das propriedades do som musical de forma equivalente.}, são:
\begin{itemize}
  \item Deve-se ter um mecanismo de síntese sonora que
possibilite a geração de unidades sonoras com diferentes timbres, controle sobre a frequência fundamental e duração específica\footnote{Também entra aqui o volume, mas como o que controlamos de fato é a intensidade relativa entre as notas, resolvemos por bem omitir
esta parte até que fiquem claros os códigos relativos a isso, logo abaixo.}.
  \item Deve-se ser capaz de construir não só unidades sonoras, mas também séries de unidades, sejam sobrepostas (acordes, por exemplo)
  ou justapostas (melodias, por exemplo).
\end{itemize}

Para o primeiro item, se prestam comumente os procedimentos de busca em tabelas/vetores com formas
de ondas em alta resolução (chamado \emph{lookup table}). O procedimento é barato e de qualidade alta
(não acrescentam ruídos relevantes ao sinal).

Para realizarmos a busca de forma eficiente, precisamos de tabelas que contenham cada uma
uma forma de onda. As tradicionais são a senoide e as ondas quadrada, triangular e dente de serra.
Mostramos a seguir estas formas de onda e também uma forma de onda retirada de um som real.

De posse destas formas de onda, A seguir está o procedimento da lookuptable:

\code{Procedimento de Lookup Table}{python_snippets/lookuptable.py}

O primeiro ponto importante é a tabela em si, usualmente unidimensional. É de comum conhecimento que
tabelas com 1024 são mais que suficientes para os usos musicais e a diferença
de qualidade (relação sinal/ruido) não muda consideravelmente com a utilização
de tabelas maiores. A seguir dispomos algumas destas tabelas, mais 
especificamente, dispomos as tabelas para os formatos de onda mais usuais.

\code{Formas de Onda Tradicionais}{python_snippets/formasDeOnda.py}

Através da utilização do lookup sucessivo (procura-se um valor na primeira tabela
e este valor indica o valor na segunda tabela a ser utilizado), executamos
um waveshaping. Este procedimento é bastante apreciado pela simplicidade e eficácia
na síntese de tímbres diversos e ricos em harmônicos e evolução temporal. Embora
uma explicação exaustiva do waveshaping fuja ao escopo deste trabalho, este
método se caracteriza pela aplicação de uma função não linear ao sinal de entrada.
Neste caso o sinal de entrada é gerado pelo primeiro lookup, a função não linear aplicada
é representada pela segunda tabela e aplicada pelo segundo lookup.

Existem muitas formas de se executar waveshaping, a seguir segue o que usamos no minimum-fi, que se sustenta principalmente por ser leve e simples:

\code{Waveshaping com Lookups sucessivas}{python_snippets/lookup_cruz.py}

O segundo item - dos dois princípios expostos sobre o minimum-fi - presta-se à discretização do espaço musical. Unidades como batidas e notas
tornam mais eficiente a comunicação pois a quantidade
de estruturas sugeridas é maior e as estruturas são mais claras no discreto do que no contínuo~\cite{Roederer}. Roederer chega a
apontar que as próprias notas dos instrumentos musicais são um reflexo de que é mais eficiente
o uso do discreto do que do contínuo para a geração de estruturas musicais.

De fato, unidades bem definidas se mostram úteis na prática musical 
para fazer sequências de unidades, concatená-las. Quando as unidades
são notas, as sequências de unidades justapostas no tempo são melodias ou linhas melódicas. As
sequências sobrepostas no tempo são comumente pensadas como acordes, mas podem ser tidas simplesmente
como sobreposições circunstanciais de duas linhas melódicas. Isso, claro, segundo
a sistematização clássica e usual da música~\cite{Lacerda}\footnote{Vole assinalar aqui que a música do século XX apresentou diversos modelos teóricos que quebram com este entendimento simplificado sobre a música, suas unidades básicas e estruturas relacionadas}.

As duas construções básicas explicitadas a seguir, baseadas na dicotomia melodia/harmonia
(horizontalidade/verticalidade, justaposição/sobreposição), são
as funções \emph{fazSequencia} e \emph{fazAcorde} no minimum-fi. Vale notar elas são absolutamente 
equivalentes em uma análise puramente conceitual, i.e. uma delas pode ser omitida segundo algumas teorizações. Isso fica particularmente óbvio quando se nota que os procedimentos de mixagem e concatenação são
plenamente capazes de realizar o que estas funções realizam. Aliás, as funções nada
mais são do que usos típicos e quase caricatos destes procedimentos: no fazAcorde a mixagem
sobrepõe no tempo todas as unidades, no fazSequencia as unidades são todas juntapostas no tempo.

Como pode-se notar a seguir, as sequências de notas e os acordes, em última instância, são utilizações específicas das 2 funções de síntese sonora explicadas anteriormente: lookup e lookupcruz.

\code{Realização de Sequências de Notas}{python_snippets/fazSequencia.py}

\code{Realização de Acordes de Notas}{python_snippets/fazAcorde.py}

A última das cinco funções utilizadas é uma soma amostra a amostra de dois sons. Para isso,
é necessário completar com zeros a sequência com o menor número de amostras para somar rapidamente:

\code{Somador (função auxiliar)}{python_snippets/somador.py}

Depois disso é usufruir com estruturas criadas. Por exemplo: depois de construidas
algumas tabelas para serem usadas como diferentes timbres, pode-se criar as
escalas completamente simétricas na oitava cromática assim:

\code{Escalas Completamente Simétricas (na grade dos 12 semitons e no âmbito da oitava)}{python_snippets/escalas_simetricas.py}

% escala\_1=range(12) # cromática ascendente sem inclusão da oitava
% escala\_2=range(0,12,2) # tons inteiros
% escala\_3=range(0,12,3) # diminutão
% escala\_4=range(0,12,4) # terças maiores
% escala\_6=range(0,12,6) # trítonos
% \end{python}
% \input{py}

Sendo cada unidade um semitom. Ou seja, cada unidade é o fator que fica presente na conta $f_0 . 2^{(1/12)} . fator $ e resulta na frequência exata da nota relativa a $f_0$ em um sistema temperado ideal.

Podemos também utilizar as usuais
escalas tonais maior e menor natural, harmônica e melódica:

\code{Escalas diatônicas tonais (maiores e menores)}{python_snippets/escalas_Mm.py}

A utilização de intervalos menores que o semitom\footnote{Prática também conhecida
como \emph{microtonalismo}, que gera/utiliza \emph{microtonalidade}.} mostra-se trivial em
nossa implementação. Basta representar sequências de fracionários
que são partes linearmente proporcionais dos semitons ou modificar o fator $f$:

\code{Escalas Microtonais (com quartos e oitavos de tom e intervalos menores)}{python_snippets/escalas_microtonais.py}

Da mesma forma, as triades maiores e menores
são especificadas com simplicidade. Note que aos acordes \emph{diminutão} e
\emph{aumentado} correspodem as mesmas notas das escalas simétricas de terças menores
e maiores respectivamente:

\code{Acordes anotados como listas em python}{python_snippets/acordes.py}

E séries/sequências podem ser anotadas junto a variações:

\code{Séries diversas}{python_snippets/series.py}

Com este arcabouço, o passo seguinte é sintetizar e mixar, resultando
em sequências musicais. A síntese de sequências e acordes são feitos tipicamente
através de comandos desta forma:

\code{Sintetizando sequências e acordes}{python_snippets/seqs_acordes.py}

Já a síntese de estruturas compostas (p.ex. sequências de acordes e sobreposição
de linhas melódicas), são feitas com os recursos usuais da liguagem. Neste caso
usamos listas em Python e implementações em Numpy (mais eficientes em tempo de
execussão e também na simplicidade do código) estão no apêndice. Estes mesmos
procedimentos são praticamente os mesmos em Scilab, C/C++, Javascript, PHP, etc.

A seguir demonstramos para fins didáticos, a construção de acordes periódicos em
python puro\footnote{A implementação não didática destas três linhas de código
podem ser agrupadas em uma só linha.}:

\code{Acordes periódicos}{python_snippets/acordes_periodicos.py}

Em posse destas sequências, acordes e recursos da linguagem,
formamos estruturas hierarquicamente superiores
através da concatenação de estruturas, da mixagem de estruturas, e da amplificação
(ou atenuação) seletiva das mesmas:

\code{Amplificação e mixagem}{python_snippets/amp_mix.py}

Neste ponto, basta criarmos músicas e sequências de interesse estético ou para pesquisa. Dado o ferramental, os encadeamentos dependem de intensões estéticas, entendimentos musicais e estruturas abstratas que mantém a coerência e interesse em uma peça musical. Para o leitor mais interessado, deixamos para o Apêndice X um exemplo de música feita com o minimum-fi. A seguir utilizamos esta base apresentada para sintetizar estruturas musicais e então um EP.

\vspace{10 mm}

        \subsubsection{FIGGUS: FInite Groups in Granular and Unit Synthesis}

Em coerência com os conhecimentos
e tecnologias já apresentados nesta exposição\footnote{Especificamos o arcabouço para geração de sons musicais em unidades
e estruturas compostas logo acima na subsessão sobre o minimum-fi.}, introduzimos agora
um foco especial nas estruturas musicais
desenvolvidas. São elas, inseridas em um momento histórico e
executadas por instrumentos específicos com as técnicas de época,
que constituem uma linguagem musical e músicas propriamente
ditas[ref]. O FIGGUS constitui uma técnica composicional
manifestada em software como ferramenta de síntese de
estruturas musicais por estruturas matemáticas específicas
para representação de simetrias.

Este desenvolvimento foi iniciado em 2006 com o físico-matemático Prof. Adolfo Maia Junior - bem anterior
ao nascimento do minimum-fi - para
tratar de simetrias na música com vistas à composição musical através
de métodos matemáticos\footnote{Duas iniciações científicas trataram do assunto, convenios XXX e YYYY.}. Mais especificamente, a proposta resultou em
um programa voltado para a síntese
granular e síntese de estruturas musicais através de Grupos Algébricos. O nome dado
foi FIGGUS, sigla de FInite Groups in Granular and Unit Synthesis\footnote{Também utilizamos
o nome FIGGS (FInite Group in Granular Synthesis) dado que o termo \emph{unit synthesis} não
é usual na literatura. Posteriormente o primeiro autor deste trabalho recorreu novamente
ao uso do nome FIGGUS. Isso foi motivado principalmente pelo fato de que
o maior uso da técnica é para síntese de estruturas musicais, não para
síntese de amálgamas sonoros (ou timbres mesmo) tipicamente resultantes da síntese granular}.

Na atual reescrita, embora ainda bastante atrás do FIGGUS original quanto
à interface gráfica, a ferramenta opera diretamente em Python puro,
com as biblitecas imbutidas por padrão. Isso permite com que o FIGGUS
sintetize todo um EP usando somente os comandos:

\code{Utilizando o FIGGUS para Sintetizar um EP}{python_snippets/synth_ep.py}

Desta forma, a ferramenta fica muito mais simples para experimentações
e implementações adicionais pois sua utilização se dá através
do próprio código no qual tudo é especificado. Esta abordagem se mostra particularmente
pertinente pois a estagnação do mecanismo de síntese de estruturas musicais
beira a inviabilização do uso artístico ou torna necessário o uso
de recursos externos.
Outro uso pretendido para o FIGGUS
é a síntese de tímbres e amálgamas sonoros através da
Síntese Granular. Embora nosso foco seja outro, cabe algumas breves
palavras sobre o assunto.

A Síntese Granular é uma área bem estabelecida tanto na acústica quanto
na Computação Musical e se caracteriza pela geração de sons bastante curtos
e em quantidade massiva. Tipicamente os sons possuem entre 5 e 40 milissegundos
e a quantidade destes \emph{microsons}\footnote{Grãos sonoros e microsons são jargões
típicos da síntese granular usados para indicar sons com as durações especificadas
no texto.} pode chegar a milhares por segundo. O tratamento específico da
síntese granular foge ao escopo deste trabalho e indicamos ao leitor interessado
os artigos produzidos sobre Síntese Granular e Teoria de Grupos[refs AES e SBCM].
Assim, focamos o texto a seguir em Grupos Finitos e Síntese de Estruturas Musicais.

Nas artes é de comum conhecimento o papel absolutamente central
que as simetrias possuem. Na música, para citar somente alguns exemplos simples,
temos os numerosos estudos de simetrias na música de J. S. Bach, os jogos
de dados de Mozart e os usos recorrentes da proporção áurea na música
de Béla Bártok. Matematicamente, as \emph{simetrias} são descritas por Grupos,
e estes são definidos como sendo um conjunto (chamemos de $G$)
munido de uma operação (seja $\bullet$), formando um grupo $(G,\bullet)$
satisfazendo as seguintes propriedades:

\begin{enumerate}
    \item Fechamento: $\forall \ g_1, g_2  \in G \Rightarrow g_1 \bullet g_2 \in G$
    \item Associatividade: $(g_1 \bullet g_2) \bullet g_3 = g_1 \bullet (g_2 \bullet g_3), \forall \ g_1, g_2, g_3 \in G$
    \item Existência do elemento neutro: $\exists \ e \in G : g \bullet e = e \bullet g = g$
    \item Existência dos inversos: $\forall \ g \in G \ \exists \ g^{-1} : g \bullet g^{-1} = g^{-1} \bullet g = e$
\end{enumerate}

Adicionalmente, vale complementar que se um grupo tiver um número
finito de elementos, ele é dito finito, caso contrário é chamado
infinito. Todos os elementos de um grupo finito são cíclicos pois
operado com si próprio um número suficiente de vezes resulta em si
mesmo. Um grupo é dito comutativo ou abeliano caso a propriedade
comutativa seja satisfeita para todos os elementos:
$\forall \ g_1, g_2 \in G \Rightarrow g_1 \bullet g_2 = g_2 \bullet g_1$.

São numerosas as aplicações de grupos e estes
usos para as artes possam se desenvolver
de diversas formas focadas em simetrias
e com interesse estético, em
nossas pesquisas uma delas se mostrou particularmente coerente
e pertinente: as permutações de unidades sonoras. Primeiro,
através do Teorema de Cayley, todo grupo finito é isomorfo
a um grupo de permutação\footnote{Mais especificamente,
todo grupo é isomorfo a um subgrupo do grupo simétrico agindo
em G.}, i.e. seus elementos são diretamente relacionados
a permutações. Além disso, na música as permutações estão
no núcleo de diversos procedimentos canônicos como a
retrogradação e outras técnicas de variação.

[Expor sobre uma ou duas músicas feitas com o FIGGUS]

No FIGGUS, implementamos o grão sonoro (no caso mais usado
como unidade) como uma classe que possui apenas
os seguintes atributos (e não possui funcões): duração (segundos),
frequência (Hz), timbre (identificador para usar mediante implementações convenientes), intensidade (pico $\in \ [0,1]$), e duração dos fades (in e out em segundos). Aqui a classe:

\code{Grão Sonoro Básico}{python_snippets/fgrao.py}

O FIGGUS funciona com base em uma sequência de grãos posta de antemão
na qual operam as permutações. Além da sequência de grãos,
é conveniente também manter o número de grãos da sequência e uma
lista com os índices dos grãos para executar as operações necessárias para gerarmos as músicas (ver abaixo). A classe utilizada para representar as sequências é assim:

\code{Sequência de Grãos}{python_snippets/fsequencia.py}

Aos grãos em sequência neste ponto são aplicadas permutações.
Para isso é bastante conveniente representar as permutações em classesseparadas e os padrões de aplicação destas permutações para facilitar
desenvolvimentos com clareza. No caso a classe de permutações só possui os parâmetros da permutação em si e do tamando dela (número de elementos que ela utiliza). É conveniente implementar uma tradução da notação cíclica para a notação direta da permutação e é uma das facilidades da implementação por classe que fizemos. A classe do padrão de permutação possui também um período de aplicação da permutação, ou seja, de quantas em quantas leituras da sequência aplicamos a permutação. As classes ficaram assim:

\code{Permutações e Padrões de Permutações}{python_snippets/fpermutacoes.py}

De posse dos grãos, da sequência, das permutações e do padrão de permutações estamos possibilitados de realizar a estrutura musical em si. Bastam adicionalmente, especificar o número desejado de iterações da sequência. Com isso, a sequência de grãos é lida um número de vezes, aplicando as permutações na sequência segundo o padrão de aplicação das pemutações, de forma a resultar em uma sequência sonora musical. Note que se a permutação usar menos elementos que a sequência possui, alguns destes elementos ficarão estáticos nas iterações da sequência no padrão sonoro.

Na síntese do vetor sonoro relacionado, é também crucial estabelecer a taxa de amostragem preterida, como podemos perceber na função de síntese dos vetores, da classe Pattern que cuida desta realização do padrão em si. Deixamos a seguir esta função, de forma que faltará apenas a escrita destes vetores sonoros em arquivo de áudio, como veremos a seguir.

\code{Realização do Padrão Musical}{python_snippets/fpadraosonoro.py}

Em posse não apenas de representações abstratas do padrão musical a ser realizado, mas dos próprios vetores sonoros relacionados à representação digital da música a ser realizada, podemos escrever este vetor como um arquivo de áudio propriamente dito. O mais conveniente neste caso é escrever um arquivo PCM (Pulse Code Modulation) em padrão amplamente utilizado e reconhecido. Ambos os padrões WAV e AIFF satisfazem estes requisitos e escolhemos o WAV por ser o padrão de CDs (Compact Discs) e também o mais utilizado por programas de áudio. Mais especificamente, o padrão de CDs é WAV com 44100 amostras por segundo e 16 bits por amostra\footnote{No formato WAV de CD, cada amostra é representada por uma variável inteira 'signed' de 16 bits.}. As amostras dos vetores sonoros são calculados, por conveniência e por convenção, no âmbito $[-1,1]$ e precisam ser normalizados para o âmbito $[-32767,32768]$ e truncados em números inteiros. Depois disso precisam ser escritos em um arquivo com os bits como na convenção da linguagem C/C++. A bibioteca \emph{struct} cuida dessa escrita do inteiro no formato correto, e a biblioteca \emph{wave} escreve o cabeçalho no formato WAV adequado. Assim, a clase de escrita do vetor sonoro em arquivo comum fica assim:

\code{Escrita do Vetor Sonoro em Arquivo WAV}{python_snippets/fio.py}

  \subsection{Música em Tempo Real: Livecoding e ABeatTracker (ABT)}

Com os avanços computacionais recentes, tornou-se usual a síntese sonora
em tempo real. Com isso, surgiram liguagens dedicadas para o áudio e a música
em sua maioria dedicadas - ou ao menos capacitadas - para o uso 
em tempo de execussão [citar PD, SC, ChucK]. Em outras palavras, estas linguagens
possibilitam que o usuário ouça o resultante sonoro do código utilizado e altere
o código com resultado imediato no processamento e resultado que escuta.

Nesta linha de exploração musical do código, iremos expor a seguir sobre
nossas investidas em \emph{Livecoding} (escrita de código em tempo real
com vistas à performance pública) e o ABeatTracker (uma linguagem por macros
para execussão sonora rítmica em conjunto com instrumentos tradicionais e outras
fontes sonoras/musicais externas).

        \subsubsection{Livecoding}

Recentemente, grupos de ponta em música experimental no mundo todo estão
desenvolvendo apresentações musicais públicas baseadas na escrita
de código ao vivo. Usualmente, se projeta o código para que a audiência possa
ver o que está sendo escrito, no rítmo em que se escreve, e se projeta também
o resultante sonoro por autofalantes.

As motivações para isso são variadas. Expomos a seguir de forma topificada
um condensado do que nos orientou a tal prática. Vale ressaltar que são
motivações igualmente presentes em outros grupos, embora não necessariamente
em todos ou da mesmícima forma.

\begin{itemize}
    \item A performance musical por computador carece de recursos performáticos
    à altura das execussões com instrumentos tradicionais. Os gestos são por demais
    discretos e a concentração do performer é bastante focada na tela do computador.
    \item O feedback auditivo do código projetado permite que o espectador infira
    significados dos códigos. Mesmo que de forma superficial, este recurso do livecoding
    é em muito capaz de desmistificar a programação de computadores, por muito considerada
    completamente intangível.;
    \item O código em si é um recurso poderosíssimo, que permite ao usuário controlar
    os sons produzidos amostra por amostra ou em escalas maiores, como notas, compassos, fraseados
    inteiros ou mesmo em escalas maiores de tempo, como minutos, horas, dias e semanas.
    \item O compartilhamento do código é usual, leve e eficiente como entrega completamente
    aberta da tecnologia relacionada à proposta estética.
\end{itemize}

Desta forma, iniciamos em 2011 uma linha de atuação com Livecoding que resultou em uma performance
no \emph{V Festival Contato}. Utilizamos a linguagem ChucK por apresentar os recursos que
consideramos mais apropriados, embora de forma alguma isso seja consensual na prática atual
de livecoding.

A apresentação contou com recursos adicionais para agregar interesse, como a utilização
do \emph{cowsay} (para enviar mensagens enquanto se desenrolava a música) e trajetórias
de um ponto no fundo do código para sugerir a vertigem do sono REM. Estes recursos
podem ser vistos em uso claramente nos videos demonstrativos [citar videos do vimeo].
As linhas do cowsay e o script em processing relacionados estão no Apêndice XX...

Duas pessoas executaram livecoding simultaneamente. A saber, Vilson Vieira executava
rítmos, batidas bastante marcadas que serviam como base. Renato Fabbri, autor do presente
trabalho, executava linhas fluídas quasi melódicas que formavam arcos maiores. Nos intervalos
das investidas no formato citado, havia interlúdios em que ambos se revesavam com músicas
curtas e inusitadas, como em um duelo[disponibilizar músicas em links e colocar de referencia].

[Colocar os códigos meus e do Vilson e explicar o funcionamento e uso]

      \subsubsection{ABeatTracker (ABT)}
O ABT é uma linguagem/ferramenta que dispara linhas rítmicas através de macros que especificam
as células rítmicas, amostras sonoras que são utilizadas como conteúdo sonoro destas
linhas, modos de leitura destas amostras sonoras, e variáveis randômicas utilizadas
para execussão da linha. Além disso, o ABT dispõe de variáveis globais que podem ser alteradas
a qualquer momento pelo usuário, como BPM, velocidade de leitura das amostras e variáveis
randômicas globais (que se somam às individuais).

Em pouco explicitaremos a forma de utilização do ABT. De antemão, é pertinente apresentar
uma pequena discussão a respeito do que o ABT é considerado e sobre os propósitos desta
ferramenta. Em primeiro lugar, vale manter em mente que o ABT tem um funcionamento específico
que pode transcender o que se conhece por livecoding. As macros pré-estabelecidas engendram um
conjunto de recursos pré-estabelecidos bem definido, o que contrasta com a ideia de uma 'linguagem
de programação' que tenha capacidades mais amplas. De qualquer forma, linguagens com domínios
específicos não são raras e por vezes o ABT foi descrito como uma linguagem.

Sobre os propósitos do ABT, em primeiro lugar ele se dispõe a ser um instrumento computacional
essencialmente rítmico. Junto a esta proposta, vem a necessidade da utilização em conjunto com
outros instrumentos, externos ao ABT, ao computador em que estiver rodando e possivelmente externo
com relação a qualquer computador. Para isso foi elaborado o ABD (ABeatDetector), no qual o usuário
tamborila os rítmos que estiver ouvindo ou imaginando para que o ABT sincronize o pulso e utilize
células rítmicas relacionadas. A análise feita pelo ABD resulta em uma série de rítmos explicitados
por sequências de compassos que encapsulem durações regulares do rítmo tamborilado. Estes rítmos
relacionados ao tamborilar do usuário são chamados de harmônicos e podem ser selecionados
prontamente para o disparo de linhas melódicas.

No Apêndice XX está o manual de utilização do ABT e todo o código do ABT (e do ABD) está
disponivel online (citar repos). Exploramos abaixo detalhes relevantes da implementação,
em especial a forma de funcionamento do ABD.

[códigos-chave e explicações sobre o ABT e o ABD]

  \subsection{Música na Matéria: EKP e AHT}

Embora o foco deste trabalho seja na exploração musical através de códigos (abertos!),
nesta sessão iniciamos uma explicação simples, clara e factual de como estas
investidas transcendem o código e até mesmo a música em si.

Mais relevantes que os desenvolvimentos em si, são as mobilizações criadas nos
entornos e os engajamentos. Isso ficará evidente nas próximas - curtas - sessões
destes desenvolvimentos e resultados.

Aqui apresentamos dois trabalhos que geraram alguma movimentação de pessoas,
resultando em reuniões, desenvolvimentos, pesquisas e apresentações propriamente
ditas. O primeiro utiliza o estado do hardware como entrada, o segundo se trata
de uma mesa escultural para flutuação de origamis que resulta em um instrumento musical
bastante lúdico.

      \subsubsection{Emotional Kernel Panic (EKP)}

Em 2008, colaborando intensamente com o CDTL
(Centro de Desenvolvimento de Tecnologias Livres)\footnote{foi uma associação civil formada e desmembrada em 2008 e sediada em Recife, PE}
foi lançada a ideia de utilizar o estado do sistema operacional - especialmente o kernel linux - para
geração de sons. Surge o Emotional Kernel Panic (EKP) na colaboração excepcional de
Felipe Machado\footnote{Importantíssimo para o desenvolvimento da Cultura Digital no Brasil [citar fontes e programas governamentais]},
Ricardo Brazileiro\footnote{Artivista Digital bastante ativo [citar fontes]} e o primeiro autor do presente trabalho.

Desde o inicio, foram definidos três finalidades
para esta exploração do SO:

[visitar o README do EKP e achar os scripts do Brazileiro sobre o EKP]

\begin{itemize}
    \item Didáticos
    \item Artísticos
    \item Monitoramento do SO
\end{itemize}

Toda a base do EKP está aqui: http://trac.assembla.com/audioexperiments/browser/ekp-base

[Patches desenvolvidos por Brazileiro/Machado e rascunho do EKP-Monitor]

      \subsubsection{AirHackTable}

A AirHackTable é um instrumento musical eletrônico controlado por origamis (dobraduras de papel), construída na forma de uma mesa. Nela, uma rede de coolers reciclados faz flutuar origamis de geometria e cores variadas. Os movimentos dos origamis são captados por webcam e interpretados em tempo real por software de processamento de imagens, gerando padrões que controlam a transformação sonora da música. Dessa forma, pode-se dizer que os sons gerados refletem o voo dos origamis de acordo com suas geometrias (que geram trajetórias de voo caracteristicas).

Na versão atual da AirHacktable, cada cor (vermelho, amarelo, azul, verde, preto, ou branco) controla uma voz, e a posição do origami na mesa modula aspectos do som. Em outras palavras, se o origami está flutuando mais à esquerda, o som sai no canal à esquerda da mesa, e se o origami está mais próximo da câmera, o volume aumenta, e se está mais afastado do operador, o som se torna mais agudo.

[imagens da AHT e dos patches em operação]

   \subsection{Música no Tecido Social: Sabrina Kawahara, Audioexperiments, EstudioLivre.org, CDTL, juntaDados.org, Devolts.org, MSST, LabMacambira.sf.net}

Esta é uma sessão menor, dedicada a apontar repercussões emergentes destas empreitadas em comunidades diferentes
e então tornar compreensível o desdobramento que estes códigos dedicados à música tiveram em processos sociais
e mobilizações civis.

Grupos foram montados em torno destes desenvolvimentos. Como os propósitos de compartilhamento
e apropriação tecnológica estavam no cerne dos grupos relacionados, as investidas
naturalmente tomaram teores engajados socialmente. A questão do empoderamento das pontas
e da criação de um patrimônio tecnológico da humanidade é consequência quase imediata
das posturas de compartilhamento e apropriação citados.

Discorremos brevemente sobre cada uma das iniciativas citadas:

\begin{itemize}
    \item Sabrina Kawahara
    \item Audioexperiments
    \item Estudiolivre.org
    \item CDTL
    \item JuntaDados.org
    \item Devolts.org
    \item MSST
    \item LabMacambira.sf.net
    \item Outros relacionados: MuSa, Metareciclagem, Submidialogia, Tainã
\end{itemize}

\section{Web}

  Difusão de informação com ênfase na facilitação
  da apropriação de tecnologias e de instancias políticas.

   \subsection{Tecnologias sociais de alta demanda: Sitios, Conteúdos e Articulação}

      \subsubsection{Sítios}

      FDDCA

      Ferramenta de comunicação

      (Cadastro dos pontos?)

      AA, SOS, Catalogo de Ideias, etc

      Meu site pessoal

      \subsubsection{Conteúdos}

      Wiki?

      \subsubsection{Articulação}

      IRC, Emails

\subsection{Disponibilização e desenvolvimento conjunto: wikis, etherpads, AA, Trac, IRC ..}

\subsubsection{Wiki}

\subsubsection{Trac}

\subsubsection{Screencasts - Vimeo}

\subsubsection{AA}

\subsubsection{Audio Experiments (Æ)}

\subsubsection{IRC}

\subsubsection{Etherpads}

\subsubsection{Outras fontes}

\section{Materiais didáticos}

  \subsection{Tutoriais em texto e código: Filtros, Nyquist e plugins LADSPA}

Os vários materiais didáticos produzidos constam no apêndice
deste trabalho. Um único destes será exposto a seguir por sua
capacidade de agregar os conteúdos dos capítulos anterioes:
os tutoriais de filtros e amostragem.

\begin{itemize}
    \item {\bf Tutorial de python para áudio e som}

Este tutorial foi levado para Berlim no LAC 2007 e sofreu melhoras desde entao. Esta
primeira versao ficou resumida em forma de texto no EL\footnote{http://estudiolivre.org/python-e-som-tutorial}. Em 2010
a Associacao Python Brasil escolheu este trabalho, então já mais amadurecido, para ser apresentado no
FISL em Porto Alegre. Como consequencia, foi feita uma série de video-tutoriais bastante utilizados\footnote{http://estudiolivre.org/tiki-index.php?page=Video+Tutoriais}.
Este tutorial foi comentado em listas em que o autor não participa (e outras em que o autor participa).

    \item {\bf Tutoriais de filtros e amostragem via python}

Voltados para explicitar principios fundamentais de áudio, estes tutoriais
são baseados código Python e o equivalente em C. Pequenas explicações são
dadas com o intuito de orientar a exploração inteligente destes \emph{snippets}.

\emph{Teorema de Amostragem}: estes scripts visam a experimentacao inteligente com
o Teorema de Nyquist. (descricao)

\emph{Filtros}: alem da explicitação sobre as diferenças entre filtros FIR e IIR,
duas utilizações clássicas destes filtros estão implementadas: Wavelets (FIR) e Quad (IIR)

Estes códigos podem ser baixados no repositório SVN do AudioExperiments. E os textos estão
na wiki (nos digitais ou EL, recriar pois os CDTL foram apagados)

    \item {\bf Tutorial de plugins lv2}

Dadas as dificuldades que o desenvolvimento dos \emph{plugins} de áudio apresenta,
desenvolvi um tutorial passo a passo com plugins que rodam em todas as etapas.
Ele é baseado em uma interface C++ para este padrão de plugin que eh implementado
em C. Os códigos e os textos estão todos em repositório.

    \item {\bf Microtutoriais Django ~\cite{dmicrotuts}}

Estes 'microtutoriais' são baseados nos conceitos de \emph{scripts mínimos} e
\emph{alterações puntuais}. O primeiro conjunto de microtutoriais é dedicado
a reconstruir o tutorial oficial do django de forma condensada e não prolixa.
O segundo destes conjuntos é dedicado a instrumentalizar de fato o leitor com
o entendimento do funcionamento dos princípios fundamentais deste framework.

    \item {\bf Philosometrics}

Embora este não seja um trabalho didático propriamente dito, ele tem este intuito
no cerne de sua concepção e surgimento. Em decorrência dele, surgiu o  Musimetrics,
o Cinemetrics e o Literametrics. Além disso, ele é um belo exemplo da
utilização das ciências duras para a análise de ciências humanas e foi acolhido
como tal em alguns momentos.

    \item {\bf Carta mídias livres}

Texto criado em decorrência da participação da comissão de seleção no
'Prêmio Mídias Livres', a convite do Ministério da Cultura por 'notório saber'.
Esta carta é um documento único no seu conteúdo, deixando às claras
o conceito de Mídias Livres como mídias não aprisionadas pelo conceito
de propriedade, ou seja, que priorizam a sua livre circulação e a possibilidade
de geração de materiais derivados, assim como sua geração aberta ao colaborativo e comunitário.

    \item {\bf Textos de cunho sociológico, transformador}

Produção mais numerosa que as anteriores, se caracteriza por métodos não convencionais
de abordagem dos assuntos e de escrita. Em especial utiliza-se pseudônimos para
auxiliar a despersonificação, gerando textos menos presos à satisfação da auto-imagem, dente
outras qualidades. A utiliação de psudônimos é um costume muito apreciado em diversos meios,
e as pesquisas tem confirmado as vantagens que a prática apresenta e confirma\footnote{http://disqus.com/research/pseudonyms/}.

O autor destes textos se dá ao direito de não revelar seus psudônimos - embora muitos deles
sejam publicamente conhecidos - para conservar as consequências desta prática na
forma mais pura. Como comprovante desta produção, deixamos uma mensagem sobre a publicação
de textos em mídia impressa com autores internacionais,
confirmando a participação do autor desta dissertação, mas cujo
nome não consta na publicação.

!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

de      fabi borges catadores@gmail.com por  riseup.net 
responder a     submidialogia@lists.riseup.net
para    submidialogia@lists.riseup.net
data    23 de agosto de 2011 12:26
assunto Re: [submidialogia] livro sub- publicação
lista de e-mails        <submidialogia.lists.riseup.net> Filtrar as mensagens dessa lista de e-mails
enviado por     lists.riseup.net
assinado por    riseup.net
cancelar inscrição      Cancelar a inscrição para essa lista de e-mails
        Importante principalmente porque você frequentemente lê mensagens com esse marcador.
ocultar detalhes 12:26 (6 minutos atrás)
entao, eu fui recebendo textos durante esse tempo,
alguns tao atrazados como dos sem satelites, mas muita gente mandou;

aqui os autores:

os internacionais nao sao muitos, o joni kempf (o que bebe ouro do hardware), o barbrook (futuros imaginarios),
Hamdy heda (da revolucao egipcia), o pedro soller (summerlab), maria llopis (pos porno),  talvez a bronac, nao entregou ainda.

dos brasileiros, renato fabri, ruiz, pasteur, morgana e caio, ju dornelles, mari marcassa, coletivo errorista, adriana velozo-drica, tiago pimentel, felipe fonseca, thiago novaes, lelex, felipe ribeiro(?), maira, verenilde,  bartolina silva, poro, vitoria amaro, fabib (eu),

entao, precisa publicar agora,
uma equipe para publicacao e,,,

bjs
f

!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

\end{itemize}

\subsection{Screencasts e outros materiais em video e em texto}

\begin{itemize}
    \item Python para áudio e música
	  Texto - palestras - videos

    \item Canal Macambira
No Macambira estão sendo produzidos materiais em screencasts sobre
diversas cenas de hackeamento.

    \begin{itemize}
	\item Live-Coding
	\item Raspagem de dados
    \end{itemize}
\end{itemize}

### 2012-11-05 resumoeabstract.tex
primeira limpada braba para versao final do msc

--- 2012-06-28 resumoeabstract.tex
@@
 
-Nesta dissertação, expomos os fundamentos do manuseio do som digitalizado através de recursos em código aberto como
-uma forma de manifestação cultural.
-Após uma brevíssima contextualização sobre o status que o código e o áudio digital possuem na cultura atual, procuramos abordar como um todo o fenômeno
-físico (e psicofísico) do som, a sua representação digital e os procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
-são então abordadas já com vistas ã tarefa de lidar com o áudio digital. Um capítulo especial é dedicado às linguagens dedicadas ao áudio e à música. Ainda no campo do software,
-o ambiente de desenvolvimento Unix é descrito de um ponto
-de vista bem resumido e ferramental, pois é a base para todos os desenvolvimentos expostos nesta dissertação.
-Os resultados antingidos com os fundamentos expostos são então sistematizados com ênfase nas ferramentas desenvolvidas.
-Apresentamos também uma série de repercussões destes desenvolvimentos, em especial uma frente de ação criada: o LabMacambira.sf.net.
-Por fim, como um produto adicional relacionado a esta dissertação, é delineado o mapa de exposição e comunicação no qual
-os desenvolvimentos deste trabalho estão agregados. Esta estrutura é dedicada a manter as contribuições
-de forma continuada e sistematizada através de webpages, wikis, repositórios, um canal IRC, aplicativos de acompanhamento de projetos
-e uma das tecnologias desenvolvidas neste trabalho: a Autoregulação Algorítmica.
+A representação dos elementos básicos da música, tais como notas musicais, ornamentos e estruturas intervalares, em termos do sinal digital do som é muito utilizada em softwares e rotinas para criação musical e tratamento sonoro.
+Não há, entretanto, um sistema unificado e conciso que relacione matematicamente esses elementos a sequências temporais derivadas do som digitalizado.
+Nesta dissertação, implementamos um sistema computacional, em que cada elemento musical é representado por equações que resultam claramente nas sequências temporais do som na sua representação discretizada.
+O primeiro elemento, a nota musical básica com duração, volume, altura e timbre, é relacionado quantitativamente às características do sinal digital.
+Dado este alicerce, apreendemos a realização de suas variações internas, como trêmolos, vibratos e flutuações espectrais.
+Isso permite sintetizar notas e sonoridades que dispertam interesse estético e com inspiração nos instrumentos musicais reais, mas cujas possibilidades são expandidas por se tratar de uma prática artificial com resultados sintéticos.
+A partir dessa representação das notas musicais, expande-se para sua organização em música, com o que é possível gerar estruturas musicais através de recursos consagrados, como a métrica rítmica, os intervalos de altura e os ciclos.
+Todas as equações estão implementadas na linguagem Python, com scripts que resultam em figuras e exemplos sonoros claros e simples.
+Com esse paradigma de implementação em código aberto, esta contribuição pode ser expandida em processos de co-autoria e utilizada livremente por músicos, engenheiros e outros interessados.
+A realidade e eficácia destas contribuições foi comprovada com a síntese de montagens musicais em cada uma das etapas assinaladas: nota básica, nota incrementada, notas em música.
+É possível, também, sintetizar albuns inteiros através de alguma parametrização dada por um usuário.
+Subprodutos deste arcabouço já foram disponibilizados na Internet na forma de músicas geradas por scripts, experimentos psicoacústicos e iniciações hackers pelo grupo labMacambira.sf.net / \#labmacambira @ Freenode.
 
@@
 tradução de:
-
 Nesta dissertação, expomos os fundamentos do manuseio do som digitalizado através de recursos em código aberto como

### 2012-11-05 scripts/adsr.py
atualizando figura e pequeno abuso de notacao

@@ -31,7 +31,7 @@ p.text(2*91,.3,"R",fontsize=48,color='red')
 som=n.random.random(De)*2-1
 som=n.sin(n.linspace(0,45*2*n.pi,De,endpoint=False))
 som=som*s
-p.plot(som,'c*',markersize=9,label=u"Som amostrado submetido ao envelope ADSR")
+p.plot(som,'c*',markersize=9,label=u"Som amostrado submetido à envoltória ADSR")
 p.plot(som,'c')
 p.plot((len(som),len(som)),(-.8,0),"k--")
 

### 2012-11-08 devEresultados.tex
atualizando figura e pequeno abuso de notacao
encaminhado resumo e rearranjada

@@
 %% ------------------------------------------------------------------------- %%
-\chapter{Desenvolvimentos e resultados: música no som digitalizado} %Nome do capítulo.
+\chapter{Música no som digitalizado} %Nome do capítulo.
 \label{cap:resultados} %Rótulo para futura referência ao capítulo.
@@
 \begin{gather}
-10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)} = 10 \quad \Rightarrow \quad \sum_{i=0}^{\Delta.f_a-1}t^{'2}_i=10\sum_{i=0}^{\Delta.f_a-1}t_i^2=\sum_{i=0}^{\Delta.f_a-1}(\sqrt{10}.t_i)^2 \\
+10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)} = 10 \quad \Rightarrow \quad \sum_{i=0}^{\lfloor \Delta.f_a \rfloor -1}t^{'2}_i=10\sum_{i=0}^{\Lambda-1}t_i^2=\sum_{i=0}^{\Lambda-1}(\sqrt{10}.t_i)^2 \\
 \therefore \quad t^{'}_i=\sqrt{10}t_i \quad \Rightarrow \quad t^{'}_i \approx 3,16t_i
@@
     \centering
-    \caption{Envelope ADSR canônico e uma sequência sonora arbitrária sendo submetida ao envelope}
+    \caption{Envoltória ADSR canônica e uma sequência sonora arbitrária sendo submetida à envoltória}
         \includegraphics[width=\textwidth]{figuras/adsr}

### 2012-11-08 resumoeabstract.tex
encaminhado resumo e rearranjada

--- 2012-06-28 resumoeabstract.tex
@@
 
-Nesta dissertação, expomos os fundamentos do manuseio do som digitalizado através de recursos em código aberto como
-uma forma de manifestação cultural.
-Após uma brevíssima contextualização sobre o status que o código e o áudio digital possuem na cultura atual, procuramos abordar como um todo o fenômeno
-físico (e psicofísico) do som, a sua representação digital e os procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
-são então abordadas já com vistas ã tarefa de lidar com o áudio digital. Um capítulo especial é dedicado às linguagens dedicadas ao áudio e à música. Ainda no campo do software,
-o ambiente de desenvolvimento Unix é descrito de um ponto
-de vista bem resumido e ferramental, pois é a base para todos os desenvolvimentos expostos nesta dissertação.
-Os resultados antingidos com os fundamentos expostos são então sistematizados com ênfase nas ferramentas desenvolvidas.
-Apresentamos também uma série de repercussões destes desenvolvimentos, em especial uma frente de ação criada: o LabMacambira.sf.net.
-Por fim, como um produto adicional relacionado a esta dissertação, é delineado o mapa de exposição e comunicação no qual
-os desenvolvimentos deste trabalho estão agregados. Esta estrutura é dedicada a manter as contribuições
-de forma continuada e sistematizada através de webpages, wikis, repositórios, um canal IRC, aplicativos de acompanhamento de projetos
-e uma das tecnologias desenvolvidas neste trabalho: a Autoregulação Algorítmica.
+A representação dos elementos básicos da música - tais como notas musicais, ornamentos e estruturas intervalares - em termos do som discretizado é bastante utilizada em softwares e rotinas para criação musical e tratamento sonoro.
+Não há, entretanto, um sistema unificado e conciso que relacione matematicamente esses elementos às sequências de amostras no tempo.
+Nesta dissertação, implementamos um sistema computacional MASSA (Música e Áudio em Sequências e Séries Amostrais) em que cada elemento musical é descrito por equações que resultam diretamente nas sequências temporais do som em sua representação discretizada.
+O elemento fundamental, a nota musical básica com duração, volume, altura e timbre, é relacionado quantitativamente às características do sinal digital.
+As variações internas das notas, como trêmolos, vibratos e flutuações espectrais, também são contempladas, o que permite sintetizar notas e sonoridades com inspiração nos instrumentos musicais reais além de sonoridades novas.
+A partir dessa representação das notas musicais, é possível gerar estruturas musicais através de recursos consagrados, como a métrica rítmica, os intervalos de altura e os ciclos.
+As equações deram origem a uma implementação na linguagem Python, com scripts que resultam em exemplos sonoros simples e imagens como espectrogramas, variação da amplitude e princípios organizacionais da música.
+A eficácia do sistema foi comprovada com a síntese de pequenas peças usando notas básicas, notas incrementadas e notas em música.
+É possível, também, sintetizar álbuns inteiros através de colagens de scripts e parametrização especificada pelo usuário.
+Com o paradigma de implementação em código aberto, o sistema pode ser expandido em processos de co-autoria e usado livremente por músicos e outros interessados.
+De fato, o sistema já vem sendo empregado por usuários externos para fins diversos, que incluem produção de músicas, experimentos psicoacústicos e difusão da linguagem computacional explorando o apelo lúdico do audiovisual.
 

### 2012-11-08 introducao.tex
garibadinha no esqueleto

@@
 
-  \section{O som e o áudio}
+    \section{Do som ao áudio digital}
 
-  \section{Código aberto, software livre e cultura livre}
+    \section{Arte sonora e rudimentos de teoria musical}
 
-      \subsection{Cultura de compartilhamento}
-
-      \subsection{Licenças livres para software e para outras mídias}
-	  \subsubsection{Principais licenças}
-	  \label{sec:princ_licencas}
-
-	  \subsubsection{Domínio público e outras categorias}
-
-      \subsection{A convergência das mídias e o multimídia}
-      \label{sec:midiamultimidia}
-
-      Os anos de 1999 e 2000 compreendem o conhecido 'marco inicial da primeira década digital'.
-      Embora seja principalmente uma forma de delimitar um processo-chave já em andamento, a este período é atribuído
-      a delineação dos contornos da 'sociedade em rede'. Esta mudança paradigmática da nossa
-      sociedade é sentida agora em maior intensidade nos negócios, na comunicação e na cultura. Estes dois
-      últimos fatores são cruciais para o trabalho aqui apresentado dada a notoriedade do papel do
-      fenômeno sonoro para a comunicação e expressões culturais. Mais recentemente, podemos falar claramente
-      do papel do código nestas duas frentes.
-
-      O termo 'convergência das mídias' não é utilizado para indicar algum dispositivo ou produto que
-      lida com várias mídias. Para isso utiliza-se o termo multimídia, que é tema do próximo tópico.
-      A convergência das mídias se refere à mundaneidade no lidar com suportes diferentes. Procura-se
-      sobre um assunto e lidamos com imagens, textos, videos, áudios e até outras formas diferenciadas
-      de mídia, como o código de liguagem de programação.
-
-      HTML5 CSS3 e Javascript deixando interativo. 3D, conexões rápidas que possuem banda para video HD e
-      baixa latência. Popularização de hardware capaz destes processamentos.
-
-      Tendência:
-      \begin{itemize}
-	  \item intensa utilização de audiovisual em contraposição ao
-      anterior predominio de textos simples.
-	  \item a utilização de periféricos para resultar saída diferenciada - p.ex.
-      impressora - ou entrada especializada - p.ex. touch pad.
-      \end{itemize}
-
-	  \subsubsection{Gargalos na interatividade e no estado do software}
-	  \label{sec:gargalos}
-	  Deficiencias em disponibilização da tecnologia atual: muita
-	  tecnologia fechada.
-
-	  Comparação do software proprietário com o livre para áudio:
-
-	  XXXXXXXXXXXXXXXXXXXXX
-
-	  Citação da parte de video da thread 'Abrindo o Código Fechado'.
-
-	  Para certas utilizações, a latência não permite a interatividade como 
-	  a obtida em presença do objeto examinado seja ele físico ou apenas um BD.
-	  A medida do 'agora' é de até 50 ms para humanos [Roads, Microsound] e a latência entre comunicadores é
-	  quase sempre maior que isso [tabela de latências, média, por distância, etc].
-	  Contribuimos na frente do software. Para isso veja [sessao dos resultados]
-
-	  \subsubsection{Compartilhamento}
-	  \label{sec:comp_tec}
-	  Terreno fértil -> semear que a primazia do compartilhamento se encarrega de
-	  desencadear os desenvolvimentos ligitimos.
-
-	  \subsubsection{Comparacao das soluções em software livre e proprietários}
-	  \label{sec:sl_prop}
-      [ESCREVER NO FINAL E ABORDAR BREVEMENTE AS QUESTOES DE VIDEO]
+    \section{Disponibilização computacional}
 
@@
     \label{sec:objetivos}
-Aqui analisamos os objetivos deste trabalho. Eles refletem os objetivos
-dos empreendimentos de SL em geral e tambem das investidas no audiovisual e nas artes.
-
-  \subsection{Difusão de tecnologias e práticas}
-  \label{sec:tutoriais}
-    \subsubsection{Tutoriais}
-    \label{sec:uso_sl}
-
-\begin{itemize}
-    \item Tutoriais para programação em python.
-Screen casts e texto.
-
-    \item Tutoriais para fazer plugins.
-\end{itemize}
-
-  \subsection{Extensão de recursos em software livre para multimídia}
-  \label{sec:extensao}
-
-  \subsubsection{Manipulação e tratamento de áudio}
-  \label{sec:manip-audio}
-
-  \subsubsection{Síntese de sons e estruturas musicais}
-  \label{sec:sintese}
-
-  \subsubsection{Capacitação para performance em tempo real e análise}
-  \label{sec:perf}
-  \label{sec:analise}

### 2012-11-08 pretextual.tex
garibadinha no esqueleto

@@
 \pretextualchapter{}
+%	\begin{epigrafetop}
+%		{from Pantheon import Obatalá\, \, \, \\
+%                from World import sharing, kitten \\
+%                \vspace{.1in}
+%                while True: \; \; \; \; \; \; \; \; \; \; \; \; \; \; \; \,\\
+%                    if sharing == 0: \; \; \; \; \;  \; \; \; \; \; \\
+%                        Obatalá.kill( kitten ) \; \; \;}
+%        {Autoria coletiva e anônima (2010)}
+%	\end{epigrafetop}
+
 	\begin{epigrafetop}
-		{from Pantheon import Obatalá\, \, \, \\
-                from World import sharing, kitten \\
-                \vspace{.1in}
-                while True: \; \; \; \; \; \; \; \; \; \; \; \; \; \; \; \,\\
-                    if sharing == 0: \; \; \; \; \;  \; \; \; \; \; \\
-                        Obatalá.kill( kitten ) \; \; \;}
-        {Autoria coletiva e anônima (2010)}
+		{Music is a hidden arithmetic exercise of the soul, which does not know that it is counting.}
+        {Gottfried Leibniz (1646-1716)}
 	\end{epigrafetop}
+
 % \,      a small space
@@
 	\begin{epigrafemid}
-		{If I have seen further, \\it is by standing on the shoulders of giants.} %frase
-		{Sir Isaac Newton (1642 - 1727)} %referência
+		{Music is a hidden metaphysical exercise of the soul, which does not know that it is philosophizing.} %frase
+		{Arthur Schopenhauer (1788-1860)} %referência
 	\end{epigrafemid}

### 2012-11-09 introducao.tex
fazendo a introducao

@@
 
+O som pode ser definido como a propagação ondulatória e mecânica em um meio elástico
+ou como a excitação dos mecanismos auditivos que resultam em sua percepção~\cite{Everest}. No primeiro
+caso, estamos tratando de leis puramente físicas. No segundo caso, estamos lidando com um
+fenômeno psicofísico. Nesta última abordagem encontramos lugar para a música, que escapa,
+portanto, às leis fatídicas dos fenômenos naturais~\cite{Roederer}. A delicada tarefa de capturar estruturas
+e artifícios musicais diretamente relacionados às oscilações do som
+é exatamente a tarefa proposta por este trabalho. Para isso, entendemos cabível
+abordar o som em sua representação digital e dispor os resultados em fórmulas matemáticas e implementações computacionais.
+
     \section{Do som ao áudio digital}
 
+Em sua conceituação puramente física, o som é uma onda mecânica logitudinal de pressão e pode se propagar por qualquer meio material.
+Embora o corpo humano possa captar sons cujas frequências estão fora da banda compreendida aproximadamente entre $20Hz$ e $20 kHz$, estas são as frequências chamadas audíveis e apreciadas pelo nosso aparelho auditivo.
+ Considerada a velocidade do som no ar de $\approx 343.2\,m/s$ em condições usuais,
+estes limites correspondem respectivamente aos comprimentos de onda $\frac{343.2}{20} = 17.16\,m$ e $\frac{343.2}{20000}=17.16\,mm$.
+
+A efetiva audição humana destas vibrações consiste em um fenômeno complexo ainda sob intensa pesquisa, involvendo captações pelos ossos, estômago e orelha, além de funções de transferência da cabeça e dorso e processamento pelo sistema nervoso.
+Existe, no entanto, um órgão dedicado à tarefa de captura destas ondas, ao qual chamamos ouvido e cujo funcionamento decompõe o som em seu espectro senoidal antes de passar o estímulo para o sistema nervoso~\cite{Roederer}.
+Este papel central das componentes senoidais do som é crucial para os fenomenos musicais, o que pode ser notado tanto na composição de sons de interesse para a música, quanto nas afinações e escalas~\cite{floEsp}.
+
+A representação do som é chamada áudio e usualmente usamos est termo para designar uma representação elétrica do som\footnote{Embora haja esta tendência e tradição de uso, Os termos
+som e áudio são, na verdade, usados de forma intercambeável.}.
+O áudio pode ser proveniente da captura do som por microfones ou da síntese direta.
+O áudio digital, também chamado som digital, consiste na representação sonora através de protocolos digitais.
+Alguns destes protocolos são bastante elaborados, em geral visando resultar em quantidades reduzidas de dados para facilitar armazenamento e transferência dos arquivos.
+Outras representações digitais do som são bastante diretas, consistindo em amostras igualmente espaçadas no tempo e cujas aplitudes individuais são registradas com um mesmo número de bits.
+Esta representação do som por amostras separadas por intervalos regulares $\lambda_a$ é a forma padrão de representação do com em tempo discreto e é chamada de modulação por código de pulsos (denotada pela sigla PCM do inglêm Pulse Code Modulation).
+Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$ (também chamada de taxa de amostragem) e profundidade de bit que é o número de bits utilizados para representar cada amostra.
+Dispomos na figura~\ref{fig:PCM} um esquema gráfico de um áudio PCM.
+
+\begin{figure}[h!]
+    \centering
+    \caption{Som digital em modulação por código de pulsos (PCM}
+        \includegraphics[width=\textwidth]{figuras/foo}
+        \label{fig:PCM}
+\end{figure}
+
+Nesta representação, é de conhecimento comum pelo teorema de Nyquist que não podemos representar frequências acima da metade da frequência de amostragem.
+Assim, se buscamos apreender as frequências audíveis, precisamos utilizar uma taxa de amostragem que seja ao menos o dobro da frequência mais alta capturada pelo sistema auditivo $f_a=2.20kHz=40kHz$.
+Este raciocínio está na base da utilização das frequências de amostragens $f_a=44.1kHz$ e $f_a=48kHz$, ambas utilizadas amplamente e são as taxas de amostragens padrão em \emph{Compact Disks} (CDs) e em sistemas de Rádio e TV, respectivamente.
+
+Do ponto de vista didático, é notável que praticamente todo processamento espectral sonoro seja praticável por somatórios quando tratamos do som discretizado.
+Estes mesmos procedimentos são obtidos por integrais quando estamos lidando com o som analógico, o que é consideravelmente mais complicado.
+
     \section{Arte sonora e rudimentos de teoria musical}
+
+A música é comumente definida como a arte manifesta pelos sons e silêncios.
+Embora esta concepção esteja canonicamente correta, para as concepções usuais, ela está mais associada a uma ideia de arte sonora do que à música propriamente dita.
+Para um ouvinte comum - e boa parte dos especialistas - a concepção de música exige também a presença de uma métrica rítmica além de organizações de alturas que formem melodias e harmonias (veja sessão~\ref{notasMusica}).
+De qualquer forma, é importante estabelecer este entendimento e reconhecer que a música do século XX ampliou esta concepção tradicional de música feita por notas que formam rítmos e melodias.
+Isso ocorreu primeiramente na música de concerto, especialmente nas chamadas correntes eletrônica, concreta e eletroacústica.
+Já na década de 90 ficou claro que a música popular, especialmente as músicas eletrônicas de dança, tinham por sua vez se desdobrado para abrangir não só sons de altura definida e organizações temporais bem estabelecidas dentro de métricas simples, mas também amálgamas sonoros ruidosos e disposições temporais fluídas e com sobreposições.
 

### 2012-11-09 devEresultados.tex
fazendo a introducao

@@
 \clearpage
-\section{Organização de notas musicais em música}
+\section{Organização de notas musicais em música}\label{notasMusica}
 Seja $ S_j=\{T_{i,1},T_{i,2}...T_{i,N}\} =\bigcup_{j \in [1,N]} T_{i,j} $ uma sequência de $N$ eventos

### 2012-11-09 introducao.tex
terminada primeira versao da introducao

@@
 
-A música é comumente definida como a arte manifesta pelos sons e silêncios.
-Embora esta concepção esteja canonicamente correta, para as concepções usuais, ela está mais associada a uma ideia de arte sonora do que à música propriamente dita.
-Para um ouvinte comum - e boa parte dos especialistas - a concepção de música exige também a presença de uma métrica rítmica além de organizações de alturas que formem melodias e harmonias (veja sessão~\ref{notasMusica}).
-De qualquer forma, é importante estabelecer este entendimento e reconhecer que a música do século XX ampliou esta concepção tradicional de música feita por notas que formam rítmos e melodias.
+A música é comumente descrita como a arte manifesta pelos sons e silêncios.
+Embora esta definição esteja canonicamente correta, para as concepções usuais, ela está mais associada a uma ideia de arte sonora do que à música propriamente dita pois para um ouvinte comum - e boa parte dos especialistas - uma 'música que seja música' pressupõe também a presença de uma métrica rítmica além de organizações de alturas que formem melodias e harmonias (veja sessão~\ref{notasMusica}).
+De qualquer forma, é importante estabelecer este entendimento e reconhecer que a música do século XX ampliou esta concepção tradicional de música feita por notas que formam estruturas baseadas em rítmos, melodias e harmonias bem identificáveis.
 Isso ocorreu primeiramente na música de concerto, especialmente nas chamadas correntes eletrônica, concreta e eletroacústica.
-Já na década de 90 ficou claro que a música popular, especialmente as músicas eletrônicas de dança, tinham por sua vez se desdobrado para abrangir não só sons de altura definida e organizações temporais bem estabelecidas dentro de métricas simples, mas também amálgamas sonoros ruidosos e disposições temporais fluídas e com sobreposições.
+Já na década de 90 ficou claro que também a música popular, especialmente as músicas eletrônicas de dança, tinham por sua vez se desdobrado para abarcar não só os sons de altura definida e os rganizações temporais bem estabelecidas dentro de métricas simples, mas também amálgamas sonoros ruidosos e disposições temporais fluídas e com sobreposições.
+Mesmo assim, a nota musical permanece paradigmática como 'unidade fundamental' das estruturas musicais, além disso, podem se desdobrar em estruturas não relevantes como notas em si, mas como sons que se desenvolvem no tempo, contemplando estes desenvolvimentos musicais recentes que foram supracitados (veja sessão~\ref{varInternas}).
+
+A teoria musical é uma ampla área de estudos com várias abordagens divergentes e que engloba assuntos são diversos quanto psicoacústica, convenções culturais de organização de alturas (e.g.
+escalas e seus usos) e formas musicais tradicionais, como a forma sonata e rondos, que consistem em estruturas paradigmáticas para uma música inteira.
+Neste trabalho, a abordagem deste assunto é bastante vinculada às aplicações dispostas no capítulo seguinte e estão apontadas no próprio texto mediante necessidade.
+Assim, não consideramos apropriado discorrer sobre estes assuntos aqui de forma independente, e apontamos para isso nossa bibliografia para títulos consagrados.
+Recomendamos também o acompanhamento de um especialista, pois, na música, é comum um mesmo termo ser utilizado para designar diferentes conceitos e um mesmo conceito ser vinculado a diferentes termos, mesmo em contexto escolásticos.
 
     \section{Disponibilização computacional}
+Os desenvolvimentos apresentados nesta dissertação incluem \emph{scripts} (i.e.
+pequenos programas) escritos em Python para disponibilização das tecnologias de forma explícita.
+Entendemos uma linguagem simples como o Python ou Scilab apropriada para esta tarefa
+por ser de compreensão bastante simples, aproximando-se do que chamamos de pseudo-código~\cite{tutPython}.
+Outra faceta importante desta disponibilização é a utilização de um sistema de controle de versão para os textos, códigos e figuras.
+Assim, pode-se observar o desenvolvimento do trabalho e executar modificações sem tantas preocupações, pois todas elas ficam registradas individualmente e são reversíveis.
+Utilizamos para isso o sistema Git, tida como a segunda grande contribuição de Linus Torvals, sendo o próprio kernel Linux a primeira.
+
+Os scripts estão dispostos em duas versões: em Python puro, i.e.
+sem bibliotecas externas, e em Python com Numpy e Audiolab.
+Esta segunda versão executa as rotinas através de chamadas para Fortran, o que torna o trabalho muito eficiente em termos de processamento computacional e, consequentemente, em termos da duração necessária para a execussão dos algorítmos.
+
+Algum código já foi transcrito para Javascript com muita facilidade, o que torna estas contribuições potencialmente utilizáveis em navegadores usuais, como o Firefox e o Chromium.
+
+Estas tecnologias são todas abertas, i.e.
+estão publicadas em licensas que permitem o uso, cópia, distribuição e utilização de quaisquer partes para geração de produtos derivados.
+Desta forma, os trabalhos aqui descritos estão disponíveis como parte de um repertório tecnológico livre que constitui um patromônio da humanidade e é vinculado usualmente às comunidades de Software Livre e de Código Aberto\footnote{A comunidade e movimento chamada \emph{'Open Source'} é uma corrente que entende a publicação em licensas abertas de código computacional (e outras tecnologias) como sendo uma vantagem pragmática, possibilitando facilidades de desenvolvimento e vantagens de aprendizado e mercadológicas.
+A comunidade e movimento chamada \emph{'Free Software'} engloba este entendimento, mas adiciona a abordagem filosófica da liberdade e compartilhamento, e dá ênfase a isso.
+Ambas as correntes reforçam o entendimento de que o código computacional é o 'bem mais  precioso produzido atualmente' pois consiste em tecnologia condensada, reativa (pois roda, processa ou gera resultados), modular (podendo ser copiada partes e coladas juntas de forma coerente, ainda rodam) e replicada sem custo adicional (pois a cópia de texto tem custo baixíssimo).
+Muito frequentemente isso tudo está disposto em poucas linhas escritas.}.
+Constatamos que, desta forma, facilitamos as colaborações em processos de co-autoria e geração de materiais subsequêntes.
 
@@
    \label{sec:objetivos}
+Nosso objetivo principal nesta dissertação é dispor um sistema unificado que relacione elementos básicos da música às sequências amostrais do som digital.
+Este conhecimento é existente, tanto que diversos softwares comerciais apresentam desenvolvimentos relacionados, mas não há uma apresentação unificada e concisa destas relações.
+Assim, dispomos um texto coerente em que os elementos musicais são apresentados junto às sequências temporais envolvidas.
+Além disso, como parte da efetiva contriuição, entregamos implementações em código computacional (Python) destas equações e scripts que resultam em pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais}.
+
+Dada esta direção bem estabelecida, existem alguns objetivos secundários que entendemos relevantes o suficientes para citarmos aqui.
+Em primeiro lugar, a difusão de conhecimentos em programação e compreensão do código computacional é bastante potencializada através de práticas lúdicas, no caso a música.
+Outro objetivo considerado é a apresentação de um arcabouço de síntese sonora e musical com controle amostral, para o qual encontramos potenciais usos em experimentos psico-acústicos e síntese em alta definição (\emph{hi-fi}).
+Talvez por último, buscamos apresentar estes conteúdos de forma didática, quase um verdaeiro tutorial, possibilitando uma compreensão e uso facilitados, pois os assuntos tratados são de reconhecida complexidade: processamento de sinais, música, psico-acústica, para citar somente alguns exemplos.

### 2012-11-09 introducao.tex
adicionada figura da modulacao por codigo de pulsos

@@
 Dispomos na figura~\ref{fig:PCM} um esquema gráfico de um áudio PCM.
+Observe que os $2**4=16$ grados disponíveis para a amplitude de cada amostra junto ao espaçamento regular $\lambda_a$ introduzem um erro de quatização.
+Estaremos apontando as formas de lidar com este ruído sempre que pertinente.
 
@@
     \centering
-    \caption{Som digital em modulação por código de pulsos (PCM}
-        \includegraphics[width=\textwidth]{figuras/foo}
+    \caption{Som digital em modulação por código de pulsos (PCM): 25 amostras representadas por 4 bits cada uma.}
+        \includegraphics[width=\textwidth]{figuras/pcm}
         \label{fig:PCM}
@@
 Talvez por último, buscamos apresentar estes conteúdos de forma didática, quase um verdaeiro tutorial, possibilitando uma compreensão e uso facilitados, pois os assuntos tratados são de reconhecida complexidade: processamento de sinais, música, psico-acústica, para citar somente alguns exemplos.
+Deste ponto de vista didático, também reconhecemos importante a apresentação deste material na forma de hipertexto, em que cada script e exemplo sonoro/musical seja acessível junto ao material teórico.

### 2012-11-09 conclusoes.tex
primeira versao completa da dissertacao, sem contar os apendices e faltando uma garibada

--- 2011-08-23 conclusoes.tex
@@
 %% ------------------------------------------------------------------------- %%
-\chapter{Conclusoes e trabalhos futuros} %Nome do capítulo.
-\label{cap:intro}
+\chapter{Conclusões e trabalhos futuros} %Nome do capítulo.
+\label{cap:conclusao}
 
-Conclusões sobre o material apresentado nesta dissertação e trabalhos a serem realizados.
+Os resultados apresentados no capítulo anterior estão bastante de acordo
+com os objetivos estabelecidos: conseguimos estabelecer um sistema conciso
+que relaciona elementos musicais ao som digital. Dispomos também um conjunto
+de scripts que implementam estas relações, ao qual chamamos MASSA (Música
+e Áudio em Sequências e Séries Amostrais). Talvez em último lugar, mas igualmente
+importante, realizamos uma exposição destes desenvovimentos de forma didática no
+capítulo anterior.
+
+Este trabalho também teve resultados não previstos, como a formação de grupos
+de interesse em torno da questão criativa impulsionada por ferramentas computacionais,
+em especial pelo código de programação. Neste contexto, se desta o grupo
+labMacambira.sf.net, que reúne colaboradores de todo o Brasil e alguns fora do país.
+Este grupo tem vida própria com relação ao papel do orientado
+deste trabalho, e já apresentou contribuições relevantes em diferentes áreas
+como Democracia Direta Digital, ferramentas digitais de georeferenciamento e
+atividades artísticas e educacional, como cursos, worshops e apresentações artísticas.
+Vários destes resultados estão trabalhados nos APÊNDICES~\ref{cap:mapa-sitio} e~\ref{cap:musicaExtra} além de todo acervo online criado, que ultrapassa 700 videos, diversos softwares originais e contribuições em softwares externos utilizados no mundo todo, como o Firefox, Evince, LibreOffice, GEM/Puredata, para citar somente alguns exemplos.
+
+As possibilidades previstos por estes resultados envolvem a criação de interfaces de geração de ruídos e outros sons em alta fidelidade (hi-fi), o auxílio na realização de experimentos psicoacústicos e, principalmente, a utilização destes resultados para fins artísticos e didáticos.
+A incorporação da linguagem de programação é bastante facilitada através de recursos audiovisuais, o que já realizamos por práticas de \emph{livecoding} e cursos focados em ferramentas especializadas, como o Puredata e o ChucK, e está em muito potencializada e confirmada através dos resultados descritos neste trabalho.
+
+A disposição online destes conteúdos na forma de hipertexto junto aos códigos e exemplos sonoros, todos em licencas livres, não só facilita colaborações e geração de subprodutos em co-autoria, mas também que nossas contribuições tenham usos e se desdobrem de formas que ainda não concebemos.

### 2012-11-10 introducao.tex
retoques finais

--- 2012-11-09 introducao.tex
@@
 portanto, às leis fatídicas dos fenômenos naturais~\cite{Roederer}. A delicada tarefa de capturar estruturas
-e artifícios musicais diretamente relacionados às oscilações do som
+e artifícios musicais diretamente relacionados às oscilações do som discretizado
 é exatamente a tarefa proposta por este trabalho. Para isso, entendemos cabível
@@
 
-Em sua conceituação puramente física, o som é uma onda mecânica logitudinal de pressão e pode se propagar por qualquer meio material.
+Em sua conceituação puramente física, o som é uma onda mecânica longitudinal de pressão e pode se propagar por qualquer meio material.
 Embora o corpo humano possa captar sons cujas frequências estão fora da banda compreendida aproximadamente entre $20Hz$ e $20 kHz$, estas são as frequências chamadas audíveis e apreciadas pelo nosso aparelho auditivo.
@@
 Existe, no entanto, um órgão dedicado à tarefa de captura destas ondas, ao qual chamamos ouvido e cujo funcionamento decompõe o som em seu espectro senoidal antes de passar o estímulo para o sistema nervoso~\cite{Roederer}.
-Este papel central das componentes senoidais do som é crucial para os fenomenos musicais, o que pode ser notado tanto na composição de sons de interesse para a música, quanto nas afinações e escalas~\cite{floEsp}.
+Este papel central das componentes senoidais do som é crucial para os fenômenos musicais, o que pode ser notado tanto na composição de sons de interesse para a música, quanto nas afinações e escalas~\cite{floEsp}.
 
-A representação do som é chamada áudio e usualmente usamos est termo para designar uma representação elétrica do som\footnote{Embora haja esta tendência e tradição de uso, Os termos
-som e áudio são, na verdade, usados de forma intercambeável.}.
-O áudio pode ser proveniente da captura do som por microfones ou da síntese direta.
-O áudio digital, também chamado som digital, consiste na representação sonora através de protocolos digitais.
+A representação do som é chamada áudio e usualmente usamos este termo para designar uma representação elétrica do som\footnote{Embora haja esta tendência e tradição de uso, os termos
+som e áudio são, na verdade, usados de forma intercambiável.}.
+O áudio pode ser proveniente da captura do som por microfones ou de síntese diretamente.
+O áudio digital consiste na sua representação sonora através de protocolos também digitais.
 Alguns destes protocolos são bastante elaborados, em geral visando resultar em quantidades reduzidas de dados para facilitar armazenamento e transferência dos arquivos.
-Outras representações digitais do som são bastante diretas, consistindo em amostras igualmente espaçadas no tempo e cujas aplitudes individuais são registradas com um mesmo número de bits.
-Esta representação do som por amostras separadas por intervalos regulares $\lambda_a$ é a forma padrão de representação do com em tempo discreto e é chamada de modulação por código de pulsos (denotada pela sigla PCM do inglêm Pulse Code Modulation).
-Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$ (também chamada de taxa de amostragem) e profundidade de bit que é o número de bits utilizados para representar cada amostra.
+Outras representações digitais do som são bastante diretas, consistindo em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
+Esta representação do som por amostras separadas por intervalos regulares $\lambda_a$ é a forma padrão de representação do som em tempo discreto e é chamada de modulação por código de pulsos (denotada pela sigla PCM do inglê Pulse Code Modulation).
+Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$ (também chamada de taxa de amostragem) e profundidade de bit, sendo este o número de bits utilizados para representar cada amostra.
 Dispomos na figura~\ref{fig:PCM} um esquema gráfico de um áudio PCM.
+Observe que os $2^4=16$ grados disponíveis para a amplitude de cada amostra junto ao espaçamento regular $\lambda_a$ introduzem um erro de quantização.
+Estaremos apontando as formas de lidar com o ruído causado por estes erros de quantização sempre que pertinente.
 
@@
 Nesta representação, é de conhecimento comum pelo teorema de Nyquist que não podemos representar frequências acima da metade da frequência de amostragem.
-Assim, se buscamos apreender as frequências audíveis, precisamos utilizar uma taxa de amostragem que seja ao menos o dobro da frequência mais alta capturada pelo sistema auditivo $f_a=2.20kHz=40kHz$.
+Assim, se buscamos apreender as frequências audíveis, precisamos utilizar uma taxa de amostragem que seja ao menos o dobro da frequência mais alta capturada pelo sistema auditivo $f_a=2\times 20kHz=40kHz$.
 Este raciocínio está na base da utilização das frequências de amostragens $f_a=44.1kHz$ e $f_a=48kHz$, ambas utilizadas amplamente e são as taxas de amostragens padrão em \emph{Compact Disks} (CDs) e em sistemas de Rádio e TV, respectivamente.
@@
 
-A música é comumente definida como a arte manifesta pelos sons e silêncios.
-Embora esta concepção esteja canonicamente correta, para as concepções usuais, ela está mais associada a uma ideia de arte sonora do que à música propriamente dita.
-Para um ouvinte comum - e boa parte dos especialistas - a concepção de música exige também a presença de uma métrica rítmica além de organizações de alturas que formem melodias e harmonias (veja sessão~\ref{notasMusica}).
-De qualquer forma, é importante estabelecer este entendimento e reconhecer que a música do século XX ampliou esta concepção tradicional de música feita por notas que formam rítmos e melodias.
+A música é comumente descrita como a arte manifesta pelos sons e silêncios.
+Embora esta definição esteja canonicamente correta, para as concepções usuais, ela está mais associada a uma ideia de arte sonora do que à música propriamente dita, pois para um ouvinte comum - e boa parte dos especialistas - uma 'música que seja música' pressupõe também a presença de uma métrica rítmica além de organizações de alturas que formem melodias e harmonias (veja sessão~\ref{notasMusica}).
+De qualquer forma, é importante reconhecer que a música do século XX ampliou esta concepção tradicional de música feita por notas que formam estruturas baseadas em ritmos, melodias e harmonias bem identificáveis.
 Isso ocorreu primeiramente na música de concerto, especialmente nas chamadas correntes eletrônica, concreta e eletroacústica.
-Já na década de 90 ficou claro que a música popular, especialmente as músicas eletrônicas de dança, tinham por sua vez se desdobrado para abrangir não só sons de altura definida e organizações temporais bem estabelecidas dentro de métricas simples, mas também amálgamas sonoros ruidosos e disposições temporais fluídas e com sobreposições.
+Já na década de 90 ficou claro que também a música popular, especialmente as músicas eletrônicas de dança, tinham por sua vez se desdobrado para abarcar não só os sons de altura definida e os organizações temporais bem estabelecidas dentro de métricas simples, mas também amálgamas sonoros ruidosos e disposições temporais fluídas e com sobreposições.
+Mesmo assim, a nota musical permanece paradigmática como 'unidade fundamental' das estruturas musicais.
+Além disso, as notas podem se desdobrar em sons que se desenvolvem no tempo, contemplando estes desenvolvimentos musicais recentes que foram supracitados (veja sessão~\ref{varInternas}).
+
+A teoria musical é uma ampla área de estudos com várias abordagens divergentes e que engloba assuntos tão diversos quanto psicoacústica, convenções culturais de organização de alturas (e.g.
+escalas e seus usos) e formas musicais tradicionais, como a forma sonata e rondós, que consistem em estruturas paradigmáticas para uma música inteira.
+Neste trabalho, a abordagem deste assunto é bastante vinculada às aplicações dispostas no capítulo seguinte e estão apontadas no próprio texto mediante necessidade.
+Assim, não consideramos apropriado discorrer sobre estes assuntos aqui de forma independente, e para isso dispomos os títulos referenciados em nossa bibliografia.
+Recomendamos também o acompanhamento de um especialista, pois, na música, é comum um mesmo termo ser utilizado para designar diferentes conceitos e um mesmo conceito ser vinculado a diferentes termos, mesmo em contextos escolásticos.
 
     \section{Disponibilização computacional}
+Os desenvolvimentos apresentados nesta dissertação incluem \emph{scripts} (i.e.
+pequenos programas) escritos em Python para disponibilização das tecnologias de forma explícita.
+Entendemos uma linguagem como o Python ou Scilab apropriada para esta tarefa
+por ser de compreensão bastante simples, aproximando-se do que chamamos de pseudo-código~\cite{tutPython}.
+Outra faceta importante desta disponibilização é a utilização de um sistema de controle de versão para os textos, códigos e figuras.
+Assim, pode-se observar o desenvolvimento do trabalho e executar modificações sem tantas preocupações, pois todas elas ficam registradas individualmente e são reversíveis.
+Utilizamos para isso o sistema Git, tida como a segunda grande contribuição de Linus Torvals, sendo o próprio kernel Linux a primeira.
+
+Os \emph{scripts} estão dispostos em duas versões: em Python puro, i.e.
+sem bibliotecas externas, e em Python com Numpy e Audiolab.
+Esta segunda versão executa as rotinas através de chamadas à linguagem Fortran, muito eficiente em termos de processamento computacional e, consequentemente, em termos da duração necessária para a execução dos algoritmos.
+
+Algum código já foi transcrito para JavaScript com muita facilidade, o que torna estas contribuições potencialmente utilizáveis em navegadores bastante difundidos, como o Firefox e o Chromium.
+
+Estas tecnologias são todas abertas, i.e.
+estão publicadas em licenças que permitem o uso, cópia, distribuição e utilização de quaisquer partes para geração de produtos derivados.
+Desta forma, os trabalhos aqui descritos estão disponíveis como parte de um repertório tecnológico livre que constitui um patrimônio da humanidade e é vinculado usualmente às comunidades de Software Livre e de Código Aberto\footnote{A comunidade e movimento chamada \emph{'Open Source'} é uma corrente que entende a publicação em licenças abertas de código computacional (e outras tecnologias) como sendo uma vantagem pragmática, possibilitando facilidades de desenvolvimento e vantagens de aprendizado e mercadológicas.
+A comunidade e movimento chamada \emph{'Free Software'} engloba este entendimento, mas adiciona a abordagem filosófica da liberdade e compartilhamento, e dá ênfase a isso.
+Ambas as correntes reforçam o entendimento de que o código computacional é o 'bem mais  precioso produzido atualmente' pois consiste em tecnologia condensada, reativa (pois executa, processa ou gera resultados), modular (pode ter partes copiadas reutilizadas eficientemente) e replicada sem custo adicional (pois a cópia de texto tem custo baixíssimo).
+Muito frequentemente isso tudo está disposto em poucas linhas escritas.}.
+Constatamos que, desta forma, facilitamos as colaborações em processos de co-autoria e geração de materiais subsequentes.
 
@@
    \label{sec:objetivos}
+Nosso objetivo principal nesta dissertação é dispor um sistema unificado que relacione elementos básicos da música às sequências amostrais do som digital.
+Este conhecimento é existente, tanto que diversos softwares comerciais apresentam desenvolvimentos relacionados, mas não há uma apresentação unificada e concisa destas relações.
+Assim, dispomos um texto coerente em que os elementos musicais são apresentados junto às sequências temporais envolvidas.
+Além disso, como parte da efetiva contribuição, entregamos implementações em código computacional (Python) destas equações e \emph{scripts} que resultam em pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais.}.
+
+Dada esta direção bem estabelecida, consideramos alguns objetivos secundários.
+Em primeiro lugar, a difusão de conhecimentos em programação e compreensão do código computacional é bastante potencializada através de práticas lúdicas, no caso a música.
+Outro objetivo considerado é a apresentação de um arcabouço de síntese sonora e musical com controle amostral, para o qual encontramos potenciais usos em experimentos psico-acústicos e síntese em alta definição (\emph{hi-fi}).
+Talvez por último, buscamos apresentar estes conteúdos de forma didática, quase um verdadeiro tutorial, possibilitando uma compreensão e uso facilitados, pois os assuntos tratados são de reconhecida complexidade: processamento de sinais, música, psico-acústica, para citar somente alguns exemplos.
+Deste ponto de vista didático, também reconhecemos importante a apresentação deste material na forma de hipertexto, em que cada \emph{script} e exemplo sonoro/musical seja acessível junto ao material teórico.

### 2012-11-10 introducao.tex
feitas pequenas alteracoes e arrumacao primeira no apendice

@@
 A comunidade e movimento chamada \emph{'Free Software'} engloba este entendimento, mas adiciona a abordagem filosófica da liberdade e compartilhamento, e dá ênfase a isso.
-Ambas as correntes reforçam o entendimento de que o código computacional é o 'bem mais  precioso produzido atualmente' pois consiste em tecnologia condensada, reativa (pois executa, processa ou gera resultados), modular (pode ter partes copiadas reutilizadas eficientemente) e replicada sem custo adicional (pois a cópia de texto tem custo baixíssimo).
+Ambas as correntes reforçam o entendimento de que o código computacional é o 'bem mais  precioso produzido atualmente' pois consiste em tecnologia condensada, reativa (pois executa, processa ou gera resultados), modular (pode ter partes copiadas e reutilizadas eficientemente) e replicada sem custo adicional (pois a cópia de texto tem custo baixíssimo).
 Muito frequentemente isso tudo está disposto em poucas linhas escritas.}.

### 2012-11-10 conclusoes.tex

--- 2011-08-23 conclusoes.tex
@@
 %% ------------------------------------------------------------------------- %%
-\chapter{Conclusoes e trabalhos futuros} %Nome do capítulo.
-\label{cap:intro}
+\chapter{Conclusões e trabalhos futuros} %Nome do capítulo.
+\label{cap:conclusao}
 
-Conclusões sobre o material apresentado nesta dissertação e trabalhos a serem realizados.
+Os resultados apresentados no capítulo anterior estão bastante de acordo
+com os objetivos: conseguimos estabelecer um sistema conciso
+que relaciona elementos musicais ao som digital. Dispomos também um conjunto
+de scripts que implementam estas relações, ao qual chamamos MASSA (Música
+e Áudio em Sequências e Séries Amostrais). Talvez em último lugar, mas igualmente
+importante, realizamos uma exposição destes desenvovimentos de forma didática no
+capítulo anterior.
+
+Este trabalho também teve resultados não previstos, como a formação de grupos
+de interesse em torno da questão criativa impulsionada por ferramentas computacionais,
+em especial pelo código de programação. Neste contexto, se destaca o grupo
+labMacambira.sf.net, que reúne colaboradores de todo o Brasil e alguns fora do país.
+Este grupo tem vida própria com relação ao papel do orientado
+deste trabalho, e já apresentou contribuições relevantes em diferentes áreas
+como Democracia Direta Digital, ferramentas digitais de georeferenciamento e
+atividades artísticas e educacionais, como cursos, workshops e apresentações artísticas.
+Vários destes resultados estão apresentados nos Apêndices~\ref{cap:mapa-sitio} e~\ref{cap:musicaExtra} além de todo acervo online criado, que ultrapassa 700 videos, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos.
+
+As possibilidades previstas por estes resultados envolvem a criação de interfaces de geração de ruídos e outros sons em alta fidelidade (hi-fi), o auxílio na realização de experimentos psicoacústicos e, principalmente, a utilização destes resultados para fins artísticos e didáticos.
+A incorporação da linguagem de programação é bastante facilitada através de recursos audiovisuais, o que já realizamos por práticas de \emph{livecoding} e cursos focados em ferramentas especializadas, como o Puredata e o ChucK, e está em muito potencializada e confirmada através dos resultados descritos neste trabalho.
+
+A disposição online destes conteúdos na forma de hipertexto junto aos códigos e exemplos sonoros, todos em licenças livres, não só facilita colaborações e geração de subprodutos em co-autoria, mas também que nossas contribuições tenham usos e se desdobrem de formas que ainda não concebemos.

### 2012-11-10 devEresultados.tex

@@
 Com relação à convolução, podemos estabelecer uma duração como pulso musical - a exemplo de um pulso BPM - 
-e distribuir impulsos no decorrer deste pulso, de forma a estabelecer métricas e rítmos\footnote{Lembrando
+e distribuir impulsos no decorrer deste pulso, de forma a estabelecer métricas e ritmos\footnote{Lembrando
 que a convolução com o impulso resulta no som deslocado ao instante de ocorrência do impulso.}.
@@
 
-Os elementos de $S_j$ podem ser sobrepostos por mixagem (veja equação~\ref{eq:mixagem} e a figura~\ref{fig:mixagem}) formando intervalos e acordes ou concatenados (veja equação~\ref{eq:concatenacao} e a figura~\ref{fig:concatenacao}) formando sequências melódicas e rítmos.
+Os elementos de $S_j$ podem ser sobrepostos por mixagem (veja equação~\ref{eq:mixagem} e a figura~\ref{fig:mixagem}) formando intervalos e acordes ou concatenados (veja equação~\ref{eq:concatenacao} e a figura~\ref{fig:concatenacao}) formando sequências melódicas e ritmos.
 
@@
 
-\subsection{Rítmo}
+\subsection{Ritmo}
 A noção rítmica é dependente de eventos separados por durações~\cite{rythm}. Estes eventos são discerníveis individualmente se separados
@@
 a transição do domínio da apreciação
-de durações como alturas para a apreciação em rítmo da seguinte forma ~\cite{Alfaix, microsound}:
+de durações como alturas para a apreciação em ritmo da seguinte forma ~\cite{Alfaix, microsound}:
 
@@
 \hline
-           & \multicolumn{10}{c}{$\underleftarrow{\text{\bf zona de percepção de durações em rítmo}}$} & \multicolumn{2}{c}{transição} & \multicolumn{3}{c}{-} \\
+           & \multicolumn{10}{c}{$\underleftarrow{\text{\bf zona de percepção de durações em ritmo}}$} & \multicolumn{2}{c}{transição} & \multicolumn{3}{c}{-} \\
 duração (s) & {\bf ...}     & {\bf 32,}     & {\bf 16,}   & {\bf 8,}  & {\bf 4,}   & {\bf 2,}   & {\bf 1,}   & {\bf 1/2,} & {\bf 1/4,} & {\bf 1/8,} & $\frac{1}{16}=62,5ms$ , & $\frac{1}{20}=50ms$ & {\color{Gray} 1/40} & {\color{Gray} 1/80  } & {\color{Gray} 1/160 } & {\color{Gray} 1/320 } & {\color{Gray} 1/640 } & {\color{Gray} ...
@@
 
-As relações duais (compassos simples e divisões binárias) costumam a ocorrer em rítmos de dança
+As relações duais (compassos simples e divisões binárias) costumam a ocorrer em ritmos de dança
 e ocasiões festivas e são chamadas imperfeitas. As relações triádicas
@@
 os limites de $j$ pois as divisões do pulso devem resultar em durações apreciáveis
-como rítmo e também as junções do pulso devem resultar algo apreciável como, no máximo
+como ritmo e também as junções do pulso devem resultar algo apreciável como, no máximo
 do nosso escopo, uma música ou um conjunto coeso de músicas. Dito de outra forma: a duração de $P^{min(j)}_i$, para $\forall \; i$
@@
 
-De certa forma central à experiência músical é a concepção de 'idioma musical'.
+De certa forma central à experiência musical é a concepção de 'idioma musical'.
 Existem diversas empreitadas para modelar e explorar diferentes entendimentos
@@
 Por exemplo, pode-se obter as harmonias triádicas 'sujas' (com notas não pertencentes à tríade) através de sobreposições de quartas justas.
-Outro exemplo interessante é a presença simultânea de rítmos em diferentes métricas, constituindo o que chamamos de \emph{polirritmia}.
+Outro exemplo interessante é a presença simultânea de ritmos em diferentes métricas, constituindo o que chamamos de \emph{polirritmia}.
 A montagem musical \emph{Poli-hit mia} explora estas métricas simultâneas através de trem de impulsos convoluídos com as notas que compõem cada linha.

### 2012-11-10 fundamentos.tex
terminada reestruturacao da dissertacao

--- 2012-01-22 fundamentos.tex
@@
 
-\section{Linguagens e modularidade}
-
-\subsection{Breve histórico}
-
-Discursar brevemente sobre as diversas linguagens,
-inclusive voltadas para áudio. E sobre a reutilização
-de trechos de código.
-
-\subsection{C/C++}
-
-Resumo de:
-The C programming Language
-
-C++?
-
-sndlib, SndObj, libsndfile
+\section{Disponibilização da tecnologia em código}
 
@@
 
-Python in a Nutshell
-
-Tutorial padrão do van Rossum.
-
-Python the Hard Way?
-
-Pyo, interfaces para áudio e sinais Pywt, numpy, audiolab
-
-\subsection{Linguagens Dedicadas}
-
-Csound, ChucK, Supercollider, Puredata, Haskore,
-
-\section{Sistema Operacional e Web}
-
-\subsection{Breve histórico}
-
-O hardware estava sendo replicado mais facilmente e
-modelos eram vendidos com diferenças
-muitas vezes mínimas. Neste contexto, começou-se a
-pensar em uma camada de código que abstraísse o hardware, assim
-como a linguagem C se propôs.
-
-Iniciou-se assim o desenvolvimento de Sistemas Operacionais (SO).
-Um SO é um software que gerencia os recursos do computador e ajuda
-também em abstrair o hardware. Orientamos a esta excelênte linha do
-tempo da Wikipédia em língua inglesa \cite{solinhadotempo}:
-
-No trabalho com tecnologias abertas, ainda é bastante cotidiana
-a utilização de sistemas Unix. O áudio e música não são excessões.
-Assim, faz-se cabivel uma abordagem breve dos sistemas GNU/Linux
-inclusive porque todos os desenvolvimentos desta dissertação se deram
-neste Sistema Operacional e algumas destas contribuições se revestiram
-poéticas específicas desta plataforma, como o ABT e o EKP.
-
-\subsection{Unix e GNU/LINUX}
-Resumo de utilização. No apêndice o resumo de:
-The Unix programing envinoment
-e
-Running Linux
-
-% sox, dmesg > /dev/dsp, programas dedicados para áudio e para a música
-
-\subsection{Web e Browsers}
-
-Dados do livrinho do google.
-
-Possibilidades de uso.
-
-Breve Histórico.
-
-No apêndice resumo de:
-http://www.html5rocks.com/en/tutorials/internals/howbrowserswork/
-
-Audiolet, que mais?
-
-\section{Sistemas de Controle de Versão}
-No desenvolvimento e compartilhamento efetivo dos códigos, faz-se
-necessária a utilização de ferramentas de registro dos históricos destes
-códigos. Isso é tão vital - ou mais vital - do que aprodução efetiva de código.
-
-Desta forma, abordaremos de forma bastante resumida o tema, de forma a deixar
-registrado o uso cotidiano que se faz de SCV para áudio e música, que não
-é bastante simples, ao contrário das usuais abordagens que se faz em manuais e livros
-sobre este assunto.
-
-\subsection{Breve histórico}
-Historico Cvs, Svn, Bitbucket, git, mercurial, baazar
-
-\subsection{Funcionamento básico}
-
-Usual: Após alterações finalizadas (que não quebrem a execução do código), registra-se conjuntos de alterações, e estas alterações são
-enviadas para um repositório central ou ao menos compartilhadas com outros.
-
-Variações: para manutenção pessoal de código. Conflitos de alterações.
-
   \subsection{Git}
-
-Roteiros básicos. Ver nas páginas do LM e do Automata.

### 2012-11-10 musicaExtra.tex

@@
 \end{itemize}
+
+\subsection{Figusdevpack (FDP)}
+
+Um ambiente de interação da comunidade de Python e Música 
+para compartilhamento de códigos e excertos. Baseado principalmente
+em documentação organizada sobre as práticas e as bibliotecas
+existentes para python. Assim como scripts compartilhados que
+fazem uso de objetos e módulos específicos. Este trabalho foi aceito na
+maior conferência de áudio em linux, a Linux Audio Conference de 2008
+(LAC2008) e está sendo reativado por mim em conjunto com Vilson Vieira
+e outros desenvolvedores de áudio. Este projeto está em desenvolvimento
+do site do Estúdio Livre [estudio livre] com repositório no sourceforge [source force].
+
+As principais fontes sobre estre trabalho é a página de desenvolvimento da ideia
+que está em constante mudança ~\cite{http://estudiolivre.org/tiki-index.php?page=fdp&highlight=fdp fdpel}
+e, o artigo que foi aceito no LAC de 2008 ~\cite{http://www.estudiolivre.org/el-gallery_view.php?arquivoId=8221 fdplac2008}
+e o repositório ~\cite{http://sourceforge.net/projects/fdpack/develop fdpsf}.

### 2012-11-10 resumoeabstract.tex

@@
 A representação dos elementos básicos da música - tais como notas musicais, ornamentos e estruturas intervalares - em termos do som discretizado é bastante utilizada em softwares e rotinas para criação musical e tratamento sonoro.
-Não há, entretanto, um sistema unificado e conciso que relacione matematicamente esses elementos às sequências de amostras no tempo.
-Nesta dissertação, implementamos um sistema computacional MASSA (Música e Áudio em Sequências e Séries Amostrais) em que cada elemento musical é descrito por equações que resultam diretamente nas sequências temporais do som em sua representação discretizada.
+Não há, entretanto, um sistema unificado e conciso que relacione matematicamente estes elementos às sequências de amostras no tempo.
+Nesta dissertação, implementamos um sistema computacional ao qual chamamos MASSA (Música e Áudio em Sequências e Séries Amostrais) em que cada elemento musical é descrito por equações que resultam diretamente nas sequências temporais do som em sua representação discretizada.
 O elemento fundamental, a nota musical básica com duração, volume, altura e timbre, é relacionado quantitativamente às características do sinal digital.
 As variações internas das notas, como trêmolos, vibratos e flutuações espectrais, também são contempladas, o que permite sintetizar notas e sonoridades com inspiração nos instrumentos musicais reais além de sonoridades novas.
-A partir dessa representação das notas musicais, é possível gerar estruturas musicais através de recursos consagrados, como a métrica rítmica, os intervalos de altura e os ciclos.
-As equações deram origem a uma implementação na linguagem Python, com scripts que resultam em exemplos sonoros simples e imagens como espectrogramas, variação da amplitude e princípios organizacionais da música.
+A partir dessa representação das notas, é possível gerar estruturas musicais através de recursos consagrados, como a métrica rítmica, os intervalos de altura e os ciclos.
+As equações deram origem a uma implementação na linguagem Python, com \emph{scripts} que resultam em exemplos sonoros simples e imagens como espectrogramas, variações de amplitude e princípios organizacionais da música.
 A eficácia do sistema foi comprovada com a síntese de pequenas peças usando notas básicas, notas incrementadas e notas em música.
@@
 Com o paradigma de implementação em código aberto, o sistema pode ser expandido em processos de co-autoria e usado livremente por músicos e outros interessados.
-De fato, o sistema já vem sendo empregado por usuários externos para fins diversos, que incluem produção de músicas, experimentos psicoacústicos e difusão da linguagem computacional explorando o apelo lúdico do audiovisual.
+De fato, o sistema já vem sendo empregado por usuários externos para fins diversos, que incluem a produção de músicas, apresentações artísticas, experimentos psicoacústicos e difusão da linguagem computacional explorando o apelo lúdico do audiovisual.
 

### 2012-11-10 conclusoes.tex
garibada na intro e conclusoes

@@
 atividades artísticas e educacionais, como cursos, workshops e apresentações artísticas.
-Vários destes resultados estão apresentados nos Apêndices~\ref{cap:mapa-sitio} e~\ref{cap:musicaExtra} além de todo acervo online criado, que ultrapassa 700 videos, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos.
+Vários destes resultados estão apresentados no Apêndice~\ref{cap:musicaExtra} e no acervo online criado, que ultrapassa 700 videos, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos~\cite{siteLM,wikiLM,vimeoLM}.
 
-As possibilidades previstas por estes resultados envolvem a criação de interfaces de geração de ruídos e outros sons em alta fidelidade (hi-fi), o auxílio na realização de experimentos psicoacústicos e, principalmente, a utilização destes resultados para fins artísticos e didáticos.
+As possibilidades de trabalhos futuros abertas por estes resultados envolvem a criação de interfaces de geração de ruídos e outros sons em alta fidelidade (hi-fi), o auxílio na realização de experimentos psicoacústicos e, principalmente, a utilização destes resultados para fins artísticos e didáticos.
 A incorporação da linguagem de programação é bastante facilitada através de recursos audiovisuais, o que já realizamos por práticas de \emph{livecoding} e cursos focados em ferramentas especializadas, como o Puredata e o ChucK, e está em muito potencializada e confirmada através dos resultados descritos neste trabalho.

### 2012-11-10 introducao.tex
garibada na intro e conclusoes

@@
 \label{cap:intro} 
+\epigraph{"Tradicionalmente a notação musical é vista como um código através do qual sons, ideias musicais ou indicações para execução musical são registrados sob forma escrita."}{Edson S.
+Zampronha, \\ Notação, Representação e Composição, 2000}
 
@@
 portanto, às leis fatídicas dos fenômenos naturais~\cite{Roederer}. A delicada tarefa de capturar estruturas
-e artifícios musicais diretamente relacionados às oscilações do som discretizado
+e artifícios musicais através das características do som discretizado
 é exatamente a tarefa proposta por este trabalho. Para isso, entendemos cabível
@@
 
-Em sua conceituação puramente física, o som é uma onda mecânica longitudinal de pressão e pode se propagar por qualquer meio material.
+Em sua conceituação puramente física, o som é uma onda mecânica longitudinal de pressão e pode se propagar por qualquer meio material~\cite{}.
 Embora o corpo humano possa captar sons cujas frequências estão fora da banda compreendida aproximadamente entre $20Hz$ e $20 kHz$, estas são as frequências chamadas audíveis e apreciadas pelo nosso aparelho auditivo.
@@
 A representação do som é chamada áudio e usualmente usamos este termo para designar uma representação elétrica do som\footnote{Embora haja esta tendência e tradição de uso, os termos
-som e áudio são, na verdade, usados de forma intercambiável.}.
+som e áudio são, na verdade, usados de forma intercambiável~\cite{Everest}.}.
 O áudio pode ser proveniente da captura do som por microfones ou de síntese diretamente.
 O áudio digital consiste na sua representação sonora através de protocolos também digitais.
-Alguns destes protocolos são bastante elaborados, em geral visando resultar em quantidades reduzidas de dados para facilitar armazenamento e transferência dos arquivos.
+Alguns destes protocolos são bastante elaborados, em geral visando resultar em quantidades reduzidas de dados para facilitar armazenamento e transferência dos arquivos~\cite{procFala}.
 Outras representações digitais do som são bastante diretas, consistindo em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
-Esta representação do som por amostras separadas por intervalos regulares $\lambda_a$ é a forma padrão de representação do som em tempo discreto e é chamada de modulação por código de pulsos (denotada pela sigla PCM do inglê Pulse Code Modulation).
-Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$ (também chamada de taxa de amostragem) e profundidade de bit, sendo este o número de bits utilizados para representar cada amostra.
+Esta representação do som por amostras separadas por intervalos regulares $\lambda_a$ é a forma padrão de representação do som em tempo discreto e é chamada de modulação por código de pulsos e é denotada pela sigla PCM do inglês \emph{Pulse Code Modulation}.
+Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$ (também chamada de taxa de amostragem) e profundidade de bit, sendo este o número de bits utilizados para representar cada amostra~\cite{protocolosAudio}.
 Dispomos na figura~\ref{fig:PCM} um esquema gráfico de um áudio PCM.
@@
 Assim, se buscamos apreender as frequências audíveis, precisamos utilizar uma taxa de amostragem que seja ao menos o dobro da frequência mais alta capturada pelo sistema auditivo $f_a=2\times 20kHz=40kHz$.
-Este raciocínio está na base da utilização das frequências de amostragens $f_a=44.1kHz$ e $f_a=48kHz$, ambas utilizadas amplamente e são as taxas de amostragens padrão em \emph{Compact Disks} (CDs) e em sistemas de Rádio e TV, respectivamente.
+Este raciocínio está na base da utilização das frequências de amostragens $f_a=44.1kHz$ e $f_a=48kHz$, ambas utilizadas amplamente e são as taxas de amostragens padrão em \emph{Compact Disks} (CDs) e em sistemas de Rádio e TV, respectivamente~\cite{protocolosAudio}.
 
 Do ponto de vista didático, é notável que praticamente todo processamento espectral sonoro seja praticável por somatórios quando tratamos do som discretizado.
-Estes mesmos procedimentos são obtidos por integrais quando estamos lidando com o som analógico, o que é consideravelmente mais complicado.
+Estes mesmos procedimentos são efetuados por integrais quando estamos lidando com o som analógico, o que é consideravelmente mais complicado.
 
-    \section{Arte sonora e rudimentos de teoria musical}
+    \section{Arte sonora e teoria musical}
 
@@
 Mesmo assim, a nota musical permanece paradigmática como 'unidade fundamental' das estruturas musicais.
-Além disso, as notas podem se desdobrar em sons que se desenvolvem no tempo, contemplando estes desenvolvimentos musicais recentes que foram supracitados (veja sessão~\ref{varInternas}).
+Além disso, as notas podem se desdobrar em sons que se desenvolvem no tempo, contemplando estes desenvolvimentos musicais recentes que foram supracitados (veja sessão~\ref{varInternas}).~\cite{Wisnick,Webern,Lerdahl,Cook}
 
 A teoria musical é uma ampla área de estudos com várias abordagens divergentes e que engloba assuntos tão diversos quanto psicoacústica, convenções culturais de organização de alturas (e.g.
-escalas e seus usos) e formas musicais tradicionais, como a forma sonata e rondós, que consistem em estruturas paradigmáticas para uma música inteira.
+uso de escalas) e formas musicais tradicionais, como a forma sonata e rondós, que consistem em estruturas paradigmáticas para uma música inteira~\cite{Zamacois,Schoenberg,microsound}.
 Neste trabalho, a abordagem deste assunto é bastante vinculada às aplicações dispostas no capítulo seguinte e estão apontadas no próprio texto mediante necessidade.
@@
 Assim, pode-se observar o desenvolvimento do trabalho e executar modificações sem tantas preocupações, pois todas elas ficam registradas individualmente e são reversíveis.
-Utilizamos para isso o sistema Git, tida como a segunda grande contribuição de Linus Torvals, sendo o próprio kernel Linux a primeira.
+Utilizamos para isso o sistema Git, tida como a segunda grande contribuição de Linus Torvals, sendo o próprio kernel Linux a primeira~\cite{gitBook}.
 
-Os \emph{scripts} estão dispostos em duas versões: em Python puro, i.e.
-sem bibliotecas externas, e em Python com Numpy e Audiolab.
+Os \emph{scripts} estão dispostos em duas versões: em Python puro~\cite{python}, i.e.
+sem bibliotecas externas, e em Python com Numpy e Audiolab~\cite{numpy,audiolab}.
 Esta segunda versão executa as rotinas através de chamadas à linguagem Fortran, muito eficiente em termos de processamento computacional e, consequentemente, em termos da duração necessária para a execução dos algoritmos.
 
-Algum código já foi transcrito para JavaScript com muita facilidade, o que torna estas contribuições potencialmente utilizáveis em navegadores bastante difundidos, como o Firefox e o Chromium.
+Algum código já foi transcrito por terceiros para JavaScript com muita facilidade, o que torna estas contribuições potencialmente utilizáveis em navegadores bastante difundidos, como o Firefox e o Chromium.
 
@@
 Ambas as correntes reforçam o entendimento de que o código computacional é o 'bem mais  precioso produzido atualmente' pois consiste em tecnologia condensada, reativa (pois executa, processa ou gera resultados), modular (pode ter partes copiadas e reutilizadas eficientemente) e replicada sem custo adicional (pois a cópia de texto tem custo baixíssimo).
-Muito frequentemente isso tudo está disposto em poucas linhas escritas.}.
-Constatamos que, desta forma, facilitamos as colaborações em processos de co-autoria e geração de materiais subsequentes.
+Muito frequentemente isso tudo está disposto em poucas linhas escritas.~\cite{Raymond,Lessig}}.
+Constatamos que, desta forma, ficam facilitadas as colaborações em processos de co-autoria e geração de materiais subsequentes.
 
@@
 Nosso objetivo principal nesta dissertação é dispor um sistema unificado que relacione elementos básicos da música às sequências amostrais do som digital.
-Este conhecimento é existente, tanto que diversos softwares comerciais apresentam desenvolvimentos relacionados, mas não há uma apresentação unificada e concisa destas relações.
-Assim, dispomos um texto coerente em que os elementos musicais são apresentados junto às sequências temporais envolvidas.
-Além disso, como parte da efetiva contribuição, entregamos implementações em código computacional (Python) destas equações e \emph{scripts} que resultam em pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais.}.
+Este conhecimento já é existente ao menos em parte, tanto que diversos softwares comerciais apresentam desenvolvimentos relacionados~\cite{Steinberg, Waves, Sony}, mas não há uma apresentação unificada e concisa destas relações.
+Assim, a seguir apresentamos um texto coerente em que os elementos musicais são apresentados junto às sequências temporais envolvidas.
+Além disso, como parte da efetiva contribuição, entregamos implementações em código computacional (Python) destas relações e \emph{scripts} que resultam em pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais.}.
 

### 2012-11-10 musicaExtra.tex
melhores no apendice sobre live coding

@@
 
-Os resultados explicitados nesta dissertação estão focados no processamento de 
-áudio através de códigos em liguagens de programação de computadores e
-voltado para a música e na síntese de estruturas musicais. O tema é amplo
-e pertinente o suficiente para contar atualmente com ampla bibliografia e um histórico
-que remonta desde a própria história da computação e das linguagens de programação.
-Desta forma, entendemos cabível focar este texto nas empreitadas realizadas pelos autores
-da presente dissertação, com foco nas realizações do primeiro autor. São usos
-reais destas tecnologias, manifestadas de forma cultural e com repercussões
-sociais, como poderá é relatado neste mesmo capítulo.
+Os resultados explicitados nesta dissertação estão focados no
+processamento de áudio através de códigos em liguagens de programação
+de computadores e voltado para a música e na síntese de estruturas
+musicais. O tema é amplo e pertinente o suficiente para contar
+atualmente com ampla bibliografia e um histórico que remonta desde a
+própria história da computação e das linguagens de programação.  Desta
+forma, entendemos cabível focar este texto nas empreitadas realizadas
+pelos autores da presente dissertação, com foco nas realizações do
+primeiro autor. São usos reais destas tecnologias, manifestadas de
+forma cultural e com repercussões sociais, como é relatado neste mesmo
+capítulo.
 
-Tais realizações possuem um espectro amplo
-dentro do escopo proposto e constitui um percurso relativamente completo
-pelas possibilidades e decorrências da abordagem da música pelo código computacional\footnote{ao menos dentro do se denomina 'Cultura Livre', veja o primeiro capítulo do presente escrito para maiores informações}.
+Tais realizações possuem um espectro amplo dentro do escopo proposto e
+constitui um percurso relativamente completo pelas possibilidades e
+decorrências da abordagem da música pelo código
+computacional\footnote{ao menos dentro do que se denomina 'Cultura Livre',
+  veja o primeiro capítulo do presente escrito para maiores
+  informações}.
 
-Estaremos abordando experimentos abertos em áudio, abordagens musicais pelo código, incluindo música em tempo real, diferido, na matéria e repercussões no tecido social.
-De forma a relatar os resultados destas investidas, segue uma sessão sobre ações coletivas realizadas através das mobilizações que estes códigos criaram, e as estruturas virtuais mantidas na intenet relacionadas a estas mobilizações e ações.
-Por fim, alguns materiais didáticos criados pelo autor - que possuem algum papel emblemático - são também expostos por comporem o arsenal de repercussões desta investida no em som e código.
+Estaremos relatando experimentos abertos em áudio, abordagens musicais
+pelo código, incluindo música em tempo real, diferido, na matéria e
+repercussões no tecido social. De forma a relatar os resultados destas
+investidas, segue uma sessão sobre ações coletivas realizadas através
+das mobilizações que estes códigos criaram, e as estruturas virtuais
+mantidas na Intenet e relacionadas a estas mobilizações e ações. Por
+fim, alguns materiais didáticos criados pelo autor -- que possuem
+algum papel emblemático -- são também expostos por comporem o arsenal
+de repercussões desta investida no em som e código.
 
@@
 
-Todos os desenvolvimentos desta dissertação estão em repositórios abertos\cite{repositorios-tese-dev}.
-Alguns foram especialmente importantes como percurso para o que é apresentado neste
-escrito. Ou seja, são explorações que manifestaram diferentes aspectos
-do trabalho do código voltado para o áudio e a música.
+Todos os desenvolvimentos desta dissertação estão em repositórios
+abertos\cite{repositorios-tese-dev}.  Alguns foram especialmente
+importantes como percurso para o que é apresentado neste escrito. Ou
+seja, são explorações que manifestaram diferentes aspectos do trabalho
+do código voltado para o áudio e a música.
 
@@
 espontâneas, diárias e coletivas.
-\footnote{As chamadas culturas biopunk, ciberpunk, cipherpunk, hacker, digital e outras mais,
-possibilitadas aos recentes desenvolvimentos em telecomunicação, dizem respeito em menor
-ou maior grau à produção de código como cultura.}
+\footnote{As chamadas culturas biopunk, ciberpunk, cipherpunk, hacker,
+  digital e outras mais, possibilitadas aos recentes desenvolvimentos
+  em telecomunicação, dizem respeito em menor ou maior grau à produção
+  de código como cultura.}
 
-A seguir apresentamos alguns bons exemplos dos desenvolvimentos
-desta dissertação especificamente em áudio, sem o envolvimento direto da música. São
-códigos de necessidade para música, mas de uso mais geral e cuja implementação
-requer muito mais de processamento de sinais e computação do que de música propriamente dito.
+A seguir apresentamos alguns bons exemplos dos desenvolvimentos desta
+dissertação especificamente em áudio, sem o envolvimento direto da
+música. São códigos de necessidade para música, mas de uso mais geral
+e cuja implementação requer muito mais de processamento de sinais e
+computação do que de música propriamente dito.
 
@@
       [repos AE, LM, wiki EL, historico CDTL]
-LADSPA (Linux Audio Developers Simple Plugin API) é a API livre\footnote{Veja o Capítulo 1 para a definição de livre neste contexto.} estabelecida de plugins de áudio até a presente data.
-A última versão liberada desta API é a 1.1, em uso corrente até hoje e é de 2002 segundo os arquivos de cabeçalho relacionados.
 
-A exploração de APIs específicas foge ao escopo do presente texto e os códigos completos podem ser encontrados pela rede, junto a tutoriais e plugins de exemplo.
-O código destes desenvolvimentos e pertinente ao áudio são relativos à síntese e remoção de ruídos.
-Da síntese, exitem abordagens mais puristas, que no caso sintetiza ruído branco e filtra para resultar nos espectros desejados (veja logo abaixo para maiores especificações).
-Uma abordagem menos purista porém mais eficiente é a de usar uma amostra curta do ruído e reproduzí-la indefinidamente através de \emph{cross-fades}.
+LADSPA (Linux Audio Developers Simple Plugin API) é a API
+livre\footnote{Veja o Capítulo 1 para a definição de livre neste
+  contexto.}  estabelecida de plugins de áudio até a presente data. A
+última versão liberada desta API é a 1.1, em uso corrente até hoje e é
+de 2002 segundo os arquivos de cabeçalho dos códigos relacionados.
 
-Já na remoção de ruídos as abordagens são as mais variadas e extremamente dependentes da aplicação.
-A complexidade dos algorítmos pode atingir níveis de especialidade em processamento de sinais que merecem um trabalho dedicado.
-Aqui iremos expor somente sobre remoção de ruído \emph{'Hum'} causado usualmente pela corrente alternada que alimenta os equipamentos utilizados.
+A exploração de APIs específicas foge ao escopo do presente texto e os
+códigos completos podem ser encontrados pela rede, junto a tutoriais e
+plugins de exemplo. O código destes desenvolvimentos e pertinente ao
+áudio são relativos à síntese e remoção de ruídos. Da síntese, existem
+abordagens mais puristas, que no caso sintetiza ruído branco e filtra
+para resultar nos espectros desejados (veja logo abaixo para maiores
+especificações). Uma abordagem menos purista porém mais eficiente é a
+de usar uma amostra curta do ruído e reproduzí-la indefinidamente
+através de \emph{cross-fades}.
 
-Abaixo dispomos explicações breves sobre os diferentes ruídos e os códigos relacionados. Preferimos expor em Python os códigos de síntese
-por questão de clareza e coerência com o resto do texto\footnote{Os códigos em C/C++ e a implementação como plugin lv2 está no repositório git LV2 como pode ser visto em: http://labmacambira.git.sourceforge.net/git/gitweb-index.cgi}.
+Já na remoção de ruídos as abordagens são as mais variadas e
+extremamente dependentes da aplicação. A complexidade dos algorítmos
+pode atingir níveis de especialidade em processamento de sinais que
+merecem um trabalho dedicado. Aqui iremos expor somente sobre remoção
+de ruído \emph{'Hum'} causado usualmente pela corrente alternada que
+alimenta os equipamentos utilizados.
 
-O procedimento de reprodução do ruído através de uma amostra repetida e em crossfade é a seguinte:
+Abaixo dispomos explicações breves sobre os diferentes ruídos e os
+códigos relacionados. Preferimos expor em Python os códigos de síntese
+por questão de clareza e coerência com o resto do texto\footnote{Os
+  códigos em C/C++ e a implementação como plugin lv2 está no
+  repositório git LV2 como pode ser visto em:
+  http://labmacambira.git.sourceforge.net/git/gitweb-index.cgi}.
+
+O procedimento de reprodução do ruído através de uma amostra repetida
+e em \textit{cross-fade} é a seguinte:
 
@@
 
-Sintetizando diferentes ruídos\footnote{Dado o propósito do texto e da implementação, utilizamos nomenclaturas utilizadas na música para os ruídos, como se pode notar.}:
+Sintetizando diferentes ruídos\footnote{Dado o propósito do texto e da
+  implementação, utilizamos nomenclaturas utilizadas na música para os
+  ruídos, como se pode notar.}:
+
 \begin{itemize}
@@
 
-A remoção de ruído Hum é baseada em uma sequência de filtros notch (ver fundamentos) idealmente dispostos nos harmônicos da frequência de oscilação da corrente elétrica.
-Cada um destes filtros deve permitir ajustes finos no fator de qualidade e também na frequência central pois trata-se de um sistema real passivel dos mais diversos efeitos de distorção do modelo ideal.
-Abaixo expomos o código em Python para tal \emph{removedor de ruído} como é chamado.
-O código C/C++ e a implementação como plugin está no mesmo repositório disponibilizado para a síntese de ruídos.
+A remoção de ruído Hum é baseada em uma sequência de filtros notch
+(ver fundamentos) idealmente dispostos nos harmônicos da frequência de
+oscilação da corrente elétrica. Cada um destes filtros deve permitir
+ajustes finos no fator de qualidade e também na frequência central
+pois trata-se de um sistema real passível dos mais diversos efeitos de
+distorção do modelo ideal. Abaixo expomos o código em Python para tal
+\emph{removedor de ruído} como é chamado. O código C/C++ e a
+implementação como plugin está no mesmo repositório disponibilizado
+para a síntese de ruídos.
 
-[código e explicação do remodedor de ruído Hum]
+[código e explicação do removedor de ruído Hum]
 
@@
 
-Por razões didáticas, dividimos este capítulo em 4 partes: música em tempo diferido (ou seja, que não é feito em tempo real),
-música em tempo real,  música na matéria (suporte físico em hardware) e música no tecido social
-(considerando especialmente as mobilizações humanas relacionadas).
+Por razões didáticas, dividimos este capítulo em 4 partes: música em
+tempo diferido (ou seja, que não é feito em tempo real), música em
+tempo real, música na matéria (suporte físico em hardware) e música no
+tecido social (considerando especialmente as mobilizações humanas
+relacionadas).
 
-Com isso desejamos expor a prática musical através do código
-com exemplos reais de aplicação e em uso pelo autor, por membros do LabMacambira.sf.net,
-por parceiros, colaboradores e por usuários eventuais das naturezas mais diversas.
+Com isso desejamos expor a prática musical através do código com
+exemplos reais de aplicação e em uso pelo autor, por membros do
+LabMacambira.sf.net, por parceiros, colaboradores e por usuários
+eventuais das naturezas mais diversas.
+
+%%%%%%%%%%%%%%5 PULANDO AQUI, QUEM SABE COLOCAR EM OUTRO TEX %%%%%%%
 
@@
 \begin{quotation}
-\small
-'The increasing dominance of graphic interfaces for music software obscured 
-the continuing presence of the command-line tradition, 
-the code writer, the hacker. The code writing of deferred time 
-computer programming may be assembled out of time order, debugged and optimized.'
+\small 'The increasing dominance of graphic interfaces for music
+software obscured the continuing presence of the command-line
+tradition, the code writer, the hacker. The code writing of deferred
+time computer programming may be assembled out of time order, debugged
+and optimized.'
 
@@
 
-A realização musical em tempo diferido é o paradigma inicial da música computacional.
-Iniciando com o Music V, a proposta foi depois desenvolvido com o CSound. Pode-se dizer
-que até hoje é a forma como compositores usualmente pensam a música: pensando
-e escrevendo as estruturas, que depois são executadas por instrumentistas ou aparelhos eletrônicos.
+A realização musical em tempo diferido é o paradigma inicial da música
+computacional. Iniciando com o Music V, a proposta foi depois
+desenvolvido com o CSound. Pode-se dizer que até hoje é a forma como
+compositores usualmente pensam a música: pensando e escrevendo as
+estruturas, que depois são executadas por instrumentistas ou aparelhos
+eletrônicos.
 
-Assim como a composição instrumental permite um trabalho mais minucioso do que
-a improvisação instrumental, a realização musical em tempo diferido usualmente permite um
-detalhamento maior dos procedimentos do que a realização em tempo real. Por este
-mesmo motivo, trataremos inicialmente de dois trabalhos em tempo diferido.
-Aliás, como veremos a seguir, as abordagens completam uma base para a música
-computacional praticada até os dias de hoje.
+Assim como a composição instrumental permite um trabalho mais
+minucioso do que a improvisação instrumental, a realização musical em
+tempo diferido usualmente permite um detalhamento maior dos
+procedimentos do que a realização em tempo real. Por este mesmo
+motivo, trataremos inicialmente de dois trabalhos em tempo diferido.
+Aliás, como veremos a seguir, as abordagens completam uma base para a
+música computacional praticada até os dias de hoje.
 
-O som musical pode ser caracterizado físicamente, e sua síntese digital com base em nossa banda de audição e no teorema de Nyquist, sintetiza uma música amostra por amostra, através de princípios
-claros de síntese sonora e organização musical\footnote{Veja no APÊNDICE XX: o \emph{minimum-fi} hi-fi, low-fi, minimum-fi, ou seja,
-o mínimo de qualidade para assegurar existência e consistência
-da criação}. O segundo, chamado de \emph{FIGGUS} (FInite
-Groups in Granular and Unit Synthesis), utiliza os princípios do minimum-fi e constitui
-um módulo Python completo, cuja proposta é a utilização de simetrias, através de
-permutações e de Teoria de Grupos, para a composição de músicas. Como demonstração
-das capacidades do FIGGUS, ele gera um EP\footnote{Extended Play, uma album musical maior que um single mas menor que um LP (Long Play) inteiro} com um único comando. Este é o
-\emph{PPEPPS}\footnote{Pure Python EP: Projeto Solvente}, como veremos a seguir.
+O som musical pode ser caracterizado fisicamente, e sua síntese
+digital com base em nossa banda de audição e no teorema de Nyquist,
+sintetiza uma música amostra por amostra, através de princípios claros
+de síntese sonora e organização musical\footnote{Veja em seguida o
+  \emph{minimum-fi} hi-fi, low-fi, minimum-fi, ou seja, o mínimo de
+  qualidade para assegurar existência e consistência da criação}. O
+segundo, chamado de \emph{FIGGUS} (FInite Groups in Granular and Unit
+Synthesis), utiliza os princípios do minimum-fi e constitui um módulo
+Python completo, cuja proposta é a utilização de simetrias, através de
+permutações e de Teoria de Grupos, para a composição de músicas. Como
+demonstração das capacidades do FIGGUS, ele gera um
+EP\footnote{Extended Play, uma álbum musical maior que um single mas
+  menor que um LP (Long Play) inteiro} com um único comando. Este é o
+\emph{PPEPPS}\footnote{Pure Python EP: Projeto Solvente}, como veremos
+a seguir.
 
-Desta forma, esta sessão exemplifica e explicita - através de dois exemplos reais - os
-princípios do uso de código para a síntese \emph{musical}, desde as amostras
-relativas a uma nota com dada frequencia, amplitude e timbre, até a confecção
-de uma ferramenta derivada, já incorporando propostas musicais e estruturas
-mais elaboradas.
+Desta forma, esta sessão exemplifica e explicita - através de dois
+exemplos reais - os princípios do uso de código para a síntese
+\emph{musical}, desde as amostras relativas a uma nota com dada
+frequencia, amplitude e timbre, até a confecção de uma ferramenta
+derivada, já incorporando propostas musicais e estruturas mais
+elaboradas.
 
@@
 
-Existe uma perene matiz estética e também tecnológica
-de realizar uma dada tarefa com \emph{o mínimo} necessário.
-Esta matiz é para a música igualmente
-fundamental e reconhecida como
-princípio de unidade e coerência\footnote{Este princípio tanto é fundamental
-que as escolas musicais possuem técnicas específicas, músicas possuem suas
-próprias convenções mantidas por toda a sua duração, os arcos mantém características,
-enfim, podemos até mesmo concluir que a simetria define o escopo.}. Em código computacional
-a empreitada para manifestar este princípio ele próprio de forma mínima
-resultou no \emph{minimum-fi.py}, código Python em um único arquivo curto que sintetiza
-músicas com segundo as estruturas especificadas em linha\footnote{Em música se fala da \emph{bula} (que equivaleria
-às funções em si e das especificações da linguagem) e a partitura ou 'música' propriamente ditas (que equivale à .}.
-Na versão atual, de 2012 mas adiantada em 2011, os algorítmos 
-em Python propriamente ditos somam 
-53 linhas e inclui 5 funções. Com estas funções, estruturas musicais podem 
-ser criadas padrão a padrão, nota a nota, amostra por amostra. Na prática e posto em
-linguagem cotidiana, resultam em notas que formam
-blocos e estruturas hierarquicamente superiores.
+Existe uma perene matiz estética e também tecnológica de realizar uma
+dada tarefa com \emph{o mínimo} necessário.  Esta matiz é para a
+música igualmente fundamental e reconhecida como princípio de unidade
+e coerência\footnote{Este princípio tanto é fundamental que as escolas
+  musicais possuem técnicas específicas, músicas possuem suas próprias
+  convenções mantidas por toda a sua duração, os arcos mantém
+  características, enfim, podemos até mesmo concluir que a simetria
+  define o escopo.}. Em código computacional a empreitada para
+manifestar este princípio ele próprio de forma mínima resultou no
+\emph{minimum-fi.py}, código Python em um único arquivo curto que
+sintetiza músicas segundo as estruturas especificadas em
+linha\footnote{Em música se fala da \emph{bula} (que equivaleria às
+  funções em si e das especificações da linguagem) e a partitura ou
+  'música' propriamente ditas (que equivale à .}.  Na versão atual, de
+2012 mas adiantada em 2011, os algorítmos em Python propriamente ditos
+somam 53 linhas e incluem apenas 5 funções. Com estas funções,
+estruturas musicais podem ser criadas padrão a padrão, nota a nota,
+amostra por amostra. Na prática e posto em linguagem cotidiana,
+resultam em notas que formam blocos e estruturas hierarquicamente
+superiores.
 
-Os princípios, bastante simples\footnote{Osvaldo Lacerda, em seu livro \emph{Compendio de Teoria Elementar da Musica} fala das propriedades do som musical de forma equivalente.}, são:
+Os princípios, bastante simples\footnote{Osvaldo Lacerda, em seu livro
+  \emph{Compêndio de Teoria Elementar da Música} fala das propriedades
+  do som musical de forma equivalente.}, são:
 \begin{itemize}
-  \item Deve-se ter um mecanismo de síntese sonora que
-possibilite a geração de unidades sonoras com diferentes timbres, controle sobre a frequência fundamental e duração específica\footnote{Também entra aqui o volume, mas como o que controlamos de fato é a intensidade relativa entre as notas, resolvemos por bem omitir
-esta parte até que fiquem claros os códigos relativos a isso, logo abaixo.}.
-  \item Deve-se ser capaz de construir não só unidades sonoras, mas também séries de unidades, sejam sobrepostas (acordes, por exemplo)
-  ou justapostas (melodias, por exemplo).
+  \item Deve-se ter um mecanismo de síntese sonora que possibilite a
+    geração de unidades sonoras com diferentes timbres, controle sobre
+    a frequência fundamental e duração específica\footnote{Também
+      entra aqui o volume, mas como o que controlamos de fato é a
+      intensidade relativa entre as notas, resolvemos por bem omitir
+      esta parte até que fiquem claros os códigos relativos a isso,
+      logo abaixo.}.
+  \item Deve-se ser capaz de construir não só unidades sonoras, mas
+    também séries de unidades, sejam sobrepostas (acordes, por
+    exemplo) ou justapostas (melodias, por exemplo).
 \end{itemize}
 
-Para o primeiro item, se prestam comumente os procedimentos de busca em tabelas/vetores com formas
-de ondas em alta resolução (chamado \emph{lookup table}). O procedimento é barato e de qualidade alta
-(não acrescentam ruídos relevantes ao sinal).
+Para o primeiro item, se prestam comumente os procedimentos de busca
+em tabelas/vetores com formas de ondas em alta resolução (chamado
+\emph{lookup table}). O procedimento é barato computacionalmente e de
+qualidade alta (não acrescentam ruídos relevantes ao sinal).
 
-Para realizarmos a busca de forma eficiente, precisamos de tabelas que contenham cada uma
-uma forma de onda. As tradicionais são a senoide e as ondas quadrada, triangular e dente de serra.
-Mostramos a seguir estas formas de onda e também uma forma de onda retirada de um som real.
+Para realizarmos a busca de forma eficiente, precisamos de tabelas que
+contenham cada forma de onda. As tradicionais são a senoide e
+as ondas quadrada, triangular e dente de serra.  Mostramos a seguir
+estas formas de onda e também uma forma de onda retirada de um som
+real.
 
-De posse destas formas de onda, A seguir está o procedimento da lookuptable:
+De posse destas formas de onda, A seguir está o procedimento da
+lookuptable:
 
@@
 
-O primeiro ponto importante é a tabela em si, usualmente unidimensional. É de comum conhecimento que
-tabelas com 1024 são mais que suficientes para os usos musicais e a diferença
-de qualidade (relação sinal/ruido) não muda consideravelmente com a utilização
-de tabelas maiores. A seguir dispomos algumas destas tabelas, mais 
-especificamente, dispomos as tabelas para os formatos de onda mais usuais.
+O primeiro ponto importante é a tabela em si, usualmente
+unidimensional. É de comum conhecimento que tabelas com 1024 são mais
+que suficientes para os usos musicais e a diferença de qualidade
+(relação sinal/ruido) não muda consideravelmente com a utilização de
+tabelas maiores. A seguir dispomos algumas destas tabelas, mais
+especificamente, dispomos as tabelas para os formatos de onda mais
+usuais.
 
@@
 
-  \subsection{Música em Tempo Real: Livecoding e ABeatTracker (ABT)}
+%%%%%%%%%%%%%%%%%% FIM DO PULO %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-Com os avanços computacionais recentes, tornou-se usual a síntese sonora
-em tempo real. Com isso, surgiram liguagens dedicadas para o áudio e a música
-em sua maioria dedicadas - ou ao menos capacitadas - para o uso 
-em tempo de execussão [citar PD, SC, ChucK]. Em outras palavras, estas linguagens
-possibilitam que o usuário ouça o resultante sonoro do código utilizado e altere
-o código com resultado imediato no processamento e resultado que escuta.
+\subsection{Música em Tempo Real: Live coding e ABeatTracker (ABT)}
 
-Nesta linha de exploração musical do código, iremos expor a seguir sobre
-nossas investidas em \emph{Livecoding} (escrita de código em tempo real
-com vistas à performance pública) e o ABeatTracker (uma linguagem por macros
-para execussão sonora rítmica em conjunto com instrumentos tradicionais e outras
-fontes sonoras/musicais externas).
+Com os avanços computacionais recentes, tornou-se usual a síntese
+sonora em tempo real. Com isso, surgiram linguagens chamadas de
+domínio específico, para o áudio e a música, em sua maioria dedicadas
+-- ou ao menos capacitadas -- para o uso em tempo de
+execução. Puredata, SuperCollider, ChucK e ixilang são exemplos de
+linguagens largamente utilizadas para a composição musical e síntese
+sonora em tempo real. Em outras palavras, estas linguagens
+possibilitam que o usuário ouça o resultante sonoro do código
+utilizado e altere o código com resultado imediato no processamento e
+resultado que escuta.
 
-        \subsubsection{Livecoding}
+Nesta linha de exploração musical do código, iremos expor a seguir
+sobre nossas investidas em \emph{Live coding} (escrita de código em
+tempo real com vistas à performance pública) através de linguagens em
+contínuo desenvolvimento como ChucK, e o ABeatTracker (uma linguagem
+por macros para execução sonora rítmica em conjunto com instrumentos
+tradicionais e outras fontes sonoras/musicais externas).
 
-Recentemente, grupos de ponta em música experimental no mundo todo estão
-desenvolvendo apresentações musicais públicas baseadas na escrita
-de código ao vivo. Usualmente, se projeta o código para que a audiência possa
-ver o que está sendo escrito, no rítmo em que se escreve, e se projeta também
-o resultante sonoro por autofalantes.
+\subsubsection{Live coding}
 
-As motivações para isso são variadas. Expomos a seguir de forma topificada
-um condensado do que nos orientou a tal prática. Vale ressaltar que são
-motivações igualmente presentes em outros grupos, embora não necessariamente
-em todos ou da mesmícima forma.
+Recentemente, grupos de ponta em música experimental no mundo todo
+estão desenvolvendo apresentações musicais públicas baseadas na
+escrita de código ao vivo. Destes vale ressaltar os grupos pioneiros
+slub e Benoît and the Mandelbrots e as ``orquestras de laptops''
+PLOrk, SLOrk e DubLOrk\footnote{Para maiores detalhes vide
+  \url{http://toplap.org}}. Usualmente, se projeta o código para que a
+audiência possa ver o que está sendo escrito, no rítmo em que se
+escreve, e se projeta também o resultante sonoro por autofalantes.
+
+As motivações para isso são variadas. Expomos a seguir de forma
+topificada um condensado do que nos orientou a tal prática. Vale
+ressaltar que são motivações igualmente presentes em outros
+grupos\footnote{Vide ``Manifesto Live coding'' em
+  \url{http://toplap.org/wiki/ManifestoDraft}}, embora não
+necessariamente em todos ou da mesmícima forma.
 
 \begin{itemize}
-    \item A performance musical por computador carece de recursos performáticos
-    à altura das execussões com instrumentos tradicionais. Os gestos são por demais
-    discretos e a concentração do performer é bastante focada na tela do computador.
-    \item O feedback auditivo do código projetado permite que o espectador infira
-    significados dos códigos. Mesmo que de forma superficial, este recurso do livecoding
-    é em muito capaz de desmistificar a programação de computadores, por muito considerada
-    completamente intangível.;
-    \item O código em si é um recurso poderosíssimo, que permite ao usuário controlar
-    os sons produzidos amostra por amostra ou em escalas maiores, como notas, compassos, fraseados
-    inteiros ou mesmo em escalas maiores de tempo, como minutos, horas, dias e semanas.
-    \item O compartilhamento do código é usual, leve e eficiente como entrega completamente
-    aberta da tecnologia relacionada à proposta estética.
+    \item A performance musical por computador carece de recursos
+      performáticos à altura das execuções com instrumentos
+      tradicionais. Os gestos são por demais discretos e a
+      concentração do performer é bastante focada na tela do
+      computador.
+    \item O feedback auditivo do código projetado permite que o
+      espectador infira significados dos códigos. Mesmo que de forma
+      superficial, este recurso do Live coding é em muito capaz de
+      desmistificar a programação de computadores e sua aplicação em
+      computação musical, por muito considerada completamente
+      intangível.
+    \item O código em si é um recurso poderosíssimo, que permite ao
+      usuário controlar os sons produzidos amostra por amostra ou em
+      escalas maiores, como notas, compassos, fraseados inteiros ou
+      mesmo em escalas maiores de tempo, como minutos, horas, dias e
+      semanas.
+    \item O compartilhamento do código é usual, leve e eficiente como
+      entrega completamente aberta da tecnologia relacionada à
+      proposta estética. Isso motiva não só o programador à aplicar
+      seus conhecimentos na música, mas também o músico a adentrar o
+      uso de linguagens de programação para expressão de suas ideias
+      musicais.
 \end{itemize}
 
-Desta forma, iniciamos em 2011 uma linha de atuação com Livecoding que resultou em uma performance
-no \emph{V Festival Contato}. Utilizamos a linguagem ChucK por apresentar os recursos que
-consideramos mais apropriados, embora de forma alguma isso seja consensual na prática atual
-de livecoding.
+Desta forma, iniciamos em 2011 uma linha de atuação em Live coding com
+a criação do duo \emph{FooBarBaz} (Variáveis Metasintáticas) composto
+pelo autor Renato Fabbri e Vilson Vieira. Essa atuação resultou em uma
+performance no \emph{V Festival Contato}. É interessante ressaltar que
+essa foi a primeira apresentação de Live coding no Brasil, e a maior
+apresentação internacional em número de ouvintes (por volta de 5 mil
+pessoas) que se tenha relato.
 
-A apresentação contou com recursos adicionais para agregar interesse, como a utilização
-do \emph{cowsay} (para enviar mensagens enquanto se desenrolava a música) e trajetórias
-de um ponto no fundo do código para sugerir a vertigem do sono REM. Estes recursos
-podem ser vistos em uso claramente nos videos demonstrativos [citar videos do vimeo].
-As linhas do cowsay e o script em processing relacionados estão no Apêndice XX...
+Utilizamos a linguagem ChucK por apresentar os recursos que
+consideramos mais apropriados, embora de forma alguma isso seja
+consensual na prática atual de Live coding. Ainda, a apresentação
+contou com recursos adicionais para agregar interesse, como a
+utilização do \emph{cowsay} (para enviar mensagens enquanto se
+desenrolava a música) e trajetórias de um ponto no fundo do código
+para sugerir a vertigem do sono REM. Estes recursos podem ser vistos
+em uso claramente nos videos demonstrativos\footnote{Disponíveis em
+  \url{http://vimeo.com/33012735}, \url{http://vimeo.com/33018740},
+  \url{http://vimeo.com/33019291}, \url{http://vimeo.com/33025717} e
+  \url{http://vimeo.com/33025913}.}  As linhas do \emph{cowsay} e o
+\emph{script} em linguagem Processing relacionados, assim como maiores
+detalhes sobre o duo e fotos da apresentação podem ser vistos em
+\url{http://wiki.nosdigitais.teia.org.br/FooBarBaz}.
 
-Duas pessoas executaram livecoding simultaneamente. A saber, Vilson Vieira executava
-rítmos, batidas bastante marcadas que serviam como base. Renato Fabbri, autor do presente
-trabalho, executava linhas fluídas quasi melódicas que formavam arcos maiores. Nos intervalos
-das investidas no formato citado, havia interlúdios em que ambos se revesavam com músicas
-curtas e inusitadas, como em um duelo[disponibilizar músicas em links e colocar de referencia].
+Duas pessoas executaram Live coding simultaneamente. A saber, Vilson
+Vieira executava ritmos, batidas bastante marcadas que serviam como
+base. Renato Fabbri, autor do presente trabalho, executava linhas
+fluídas quasi melódicas que formavam arcos maiores. Nos intervalos das
+investidas no formato citado, havia interlúdios em que ambos se
+revesavam com músicas curtas e inusitadas, como em um duelo. A mixagem
+e espacialização dos dois canais de áudio foram controladas por um
+patch em Puredata que permitia o \textit{cross-fading} entre os canais
+através da detecção do movimento das mãos. Isso foi possível através
+do objeto em Puredata/GEM chamado \emph{color classify}, criado por
+Ricardo Fabbri e posto em uso por Gilson Beck durante a
+apresentação. Esse mesmo objeto foi utilizado no instrumento
+AirHackTable, discutido na seção~\ref{aht}.
 
-[Colocar os códigos meus e do Vilson e explicar o funcionamento e uso]
+A prática de Live coding requer o uso de linguagens que permitam a
+execução rápida de instruções curtas, incentivando assim o improviso
+musical através do código. Com esse objetivo, ambos \emph{live coders}
+usaram editores de texto -- vim e Emacs -- que permitiam a comunicação
+imediata com a linguagem de síntese sonora ChucK.  Acreditando que a
+música, especialmente sua vertente experimental, se baseia fortemente
+na organização de trechos de arquivos de áudio ou notas dadas a
+instrumentos sintetizados, foram gerados códigos que permitissem tal
+manipulação. Renato utilizou de trechos (``esqueletos'') minimalistas
+de código em ChucK para reproduzir arcos longos e estruturas
+melódicas, como os que seguem:
+
+\code{Interface para controle de parâmetros em tempo
+  real}{chuck_snippets/bar.ck}
+
+Ao mesmo tempo, Vilson utilizou dos códigos seguintes, também em
+ChucK, que permitem facilmente especificar a posição do arquivo de
+áudio que deve ser executada, sua velocidade de execução, duração e
+amplitude. Uma interface minimalista foi utilizada para permitir a
+alteração rápida desses parâmetros em tempo de execução.
+
+\code{Interface para controle de parâmetros em tempo
+  real}{chuck_snippets/foo.ck}
+
+\code{Classe para \textit{sampling} de arquivos de
+  áudio}{chuck_snippets/foosp.ck}
+
+Esses desenvolvimentos todos inspiraram a criação de uma linguagem
+específica para Live coding, chamada Vivace, que está em constante
+desenvolvimento e tem seu código e maiores detalhes disponíveis em
+\url{http://automata.github.com/vivace}. Já foram realizadas novas
+apresentações utilizando essa linguagem, a notar: Hacklab do Velho, em
+São Carlos (SP), Semana da Comunicação FAAP (SP), Palco UFSCar (SP) e
+Semana Nacional da Ciência e Tecnologia (SP). Nelas houve a
+predominância do apelo às mídias populares, em uma reinauguração do
+gênero da \emph{pop art}: trechos de vídeos de uma novela brasileira
+eram rearanjadas em tempo real, contrastando com linhas rítmicas e
+melódicas executadas em sincronia com o material visual\footnote{Uma
+  amostra está disponível on-line para ser utilizada por qualquer
+  pessoa que possua um navegador Chrome:
+  \url{http://pulapirata.com/skills/vivace}}.
 
 \subsubsection{ABeatTracker (ABT)}
-O ABT é uma linguagem/ferramenta que dispara linhas rítmicas através de macros que especificam
-as células rítmicas, amostras sonoras que são utilizadas como conteúdo sonoro destas
-linhas, modos de leitura destas amostras sonoras, e variáveis randômicas utilizadas
-para execussão da linha. Além disso, o ABT dispõe de variáveis globais que podem ser alteradas
-a qualquer momento pelo usuário, como BPM, velocidade de leitura das amostras e variáveis
-randômicas globais (que se somam às individuais).
 
-Em pouco explicitaremos a forma de utilização do ABT. De antemão, é pertinente apresentar
-uma pequena discussão a respeito do que o ABT é considerado e sobre os propósitos desta
-ferramenta. Em primeiro lugar, vale manter em mente que o ABT tem um funcionamento específico
-que pode transcender o que se conhece por livecoding. As macros pré-estabelecidas engendram um
-conjunto de recursos pré-estabelecidos bem definido, o que contrasta com a ideia de uma 'linguagem
-de programação' que tenha capacidades mais amplas. De qualquer forma, linguagens com domínios
-específicos não são raras e por vezes o ABT foi descrito como uma linguagem.
+O ABT é uma linguagem/ferramenta que dispara linhas rítmicas através
+de macros que especificam as células rítmicas, amostras sonoras que
+são utilizadas como conteúdo sonoro destas linhas, modos de leitura
+destas amostras sonoras, e variáveis randômicas utilizadas para
+execussão da linha. Além disso, o ABT dispõe de variáveis globais que
+podem ser alteradas a qualquer momento pelo usuário, como BPM,
+velocidade de leitura das amostras e variáveis randômicas globais (que
+se somam às individuais).
 
-Sobre os propósitos do ABT, em primeiro lugar ele se dispõe a ser um instrumento computacional
-essencialmente rítmico. Junto a esta proposta, vem a necessidade da utilização em conjunto com
-outros instrumentos, externos ao ABT, ao computador em que estiver rodando e possivelmente externo
-com relação a qualquer computador. Para isso foi elaborado o ABD (ABeatDetector), no qual o usuário
-tamborila os rítmos que estiver ouvindo ou imaginando para que o ABT sincronize o pulso e utilize
-células rítmicas relacionadas. A análise feita pelo ABD resulta em uma série de rítmos explicitados
-por sequências de compassos que encapsulem durações regulares do rítmo tamborilado. Estes rítmos
-relacionados ao tamborilar do usuário são chamados de harmônicos e podem ser selecionados
-prontamente para o disparo de linhas melódicas.
+Em pouco explicitaremos a forma de utilização do ABT. De antemão, é
+pertinente apresentar uma pequena discussão a respeito do que o ABT é
+considerado e sobre os propósitos desta ferramenta. Em primeiro lugar,
+vale manter em mente que o ABT tem um funcionamento específico que
+pode transcender o que se conhece por Live coding. As macros
+pré-estabelecidas engendram um conjunto de recursos pré-estabelecidos
+bem definido, o que contrasta com a ideia de uma 'linguagem de
+programação' que tenha capacidades mais amplas. De qualquer forma,
+linguagens com domínios específicos não são raras e por vezes o ABT
+foi descrito como uma linguagem.
 
-No Apêndice XX está o manual de utilização do ABT e todo o código do ABT (e do ABD) está
-disponivel online (citar repos). Exploramos abaixo detalhes relevantes da implementação,
-em especial a forma de funcionamento do ABD.
+Sobre os propósitos do ABT, em primeiro lugar ele se dispõe a ser um
+instrumento computacional essencialmente rítmico. Junto a esta
+proposta, vem a necessidade da utilização em conjunto com outros
+instrumentos, externos ao ABT, ao computador em que estiver executando
+e possivelmente externo com relação a qualquer computador. Para isso
+foi elaborado o ABD (ABeatDetector), no qual o usuário tamborila no
+teclado do computador os rítmos que estiver ouvindo ou imaginando para
+que o ABT sincronize o pulso e utilize células rítmicas
+relacionadas. A análise feita pelo ABD resulta em uma série de ritmos
+explicitados por sequências de compassos que encapsulem durações
+regulares do ritmo tamborilado. Estes ritmos relacionados ao
+tamborilar do usuário são chamados de harmônicos e podem ser
+selecionados prontamente para o disparo de linhas melódicas.
+
+No Apêndice XX está o manual de utilização do ABT e todo o código do
+ABT (e do ABD) está disponivel online (citar repos). Exploramos abaixo
+detalhes relevantes da implementação, em especial a forma de
+funcionamento do ABD.
 
@@
 \subsection{Música na Matéria: EKP e AHT}
+\label{aht}
 
-Embora o foco deste trabalho seja na exploração musical através de códigos (abertos!),
-nesta sessão iniciamos uma explicação simples, clara e factual de como estas
-investidas transcendem o código e até mesmo a música em si.
+Embora o foco deste trabalho seja na exploração musical através de
+códigos (abertos!), nesta sessão iniciamos uma explicação simples,
+clara e factual de como estas investidas transcendem o código e até
+mesmo a música em si.
 
-Mais relevantes que os desenvolvimentos em si, são as mobilizações criadas nos
-entornos e os engajamentos. Isso ficará evidente nas próximas - curtas - sessões
-destes desenvolvimentos e resultados.
+Mais relevantes que os desenvolvimentos em si, são as mobilizações
+criadas nos entornos e os engajamentos. Isso ficará evidente nas
+próximas - curtas - sessões destes desenvolvimentos e resultados.
 
-Aqui apresentamos dois trabalhos que geraram alguma movimentação de pessoas,
-resultando em reuniões, desenvolvimentos, pesquisas e apresentações propriamente
-ditas. O primeiro utiliza o estado do hardware como entrada, o segundo se trata
-de uma mesa escultural para flutuação de origamis que resulta em um instrumento musical
-bastante lúdico.
+Aqui apresentamos dois trabalhos que geraram alguma movimentação de
+pessoas, resultando em reuniões, desenvolvimentos, pesquisas e
+apresentações propriamente ditas. O primeiro utiliza o estado do
+hardware como entrada, o segundo se trata de uma mesa escultural para
+flutuação de origamis que resulta em um instrumento musical bastante
+lúdico.
 
@@
 
-Em 2008, colaborando intensamente com o CDTL
-(Centro de Desenvolvimento de Tecnologias Livres)\footnote{foi uma associação civil formada e desmembrada em 2008 e sediada em Recife, PE}
-foi lançada a ideia de utilizar o estado do sistema operacional - especialmente o kernel linux - para
-geração de sons. Surge o Emotional Kernel Panic (EKP) na colaboração excepcional de
-Felipe Machado\footnote{Importantíssimo para o desenvolvimento da Cultura Digital no Brasil [citar fontes e programas governamentais]},
-Ricardo Brazileiro\footnote{Artivista Digital bastante ativo [citar fontes]} e o primeiro autor do presente trabalho.
+Em 2008, colaborando intensamente com o CDTL (Centro de
+Desenvolvimento de Tecnologias Livres)\footnote{foi uma associação
+  civil formada e desmembrada em 2008 e sediada em Recife, PE} foi
+lançada a ideia de utilizar o estado do sistema operacional -
+especialmente o kernel linux - para geração de sons. Surge o Emotional
+Kernel Panic (EKP) na colaboração excepcional de Felipe
+Machado\footnote{Importantíssimo para o desenvolvimento da Cultura
+  Digital no Brasil [citar fontes e programas governamentais]},
+Ricardo Brazileiro\footnote{Artivista Digital bastante ativo [citar
+    fontes]} e o primeiro autor do presente trabalho.
 
-Desde o inicio, foram definidos três finalidades
-para esta exploração do SO:
+Desde o inicio, foram definidos três finalidades para esta exploração
+do SO:
 
@@
 
-A AirHackTable é um instrumento musical eletrônico controlado por origamis (dobraduras de papel), construída na forma de uma mesa.
-Nela, uma rede de coolers reciclados faz flutuar origamis de geometria e cores variadas.
-Os movimentos dos origamis são captados por webcam e interpretados em tempo real por software de processamento de imagens, gerando padrões que controlam a transformação sonora da música.
-Dessa forma, pode-se dizer que os sons gerados refletem o voo dos origamis de acordo com suas geometrias (que geram trajetórias de voo caracteristicas).
+A AirHackTable (AHT) é um instrumento musical eletrônico controlado por
+origamis (dobraduras de papel), construída na forma de uma mesa. Nela,
+uma rede de coolers reciclados faz flutuar origamis de geometria e
+cores variadas. Os movimentos dos origamis são captados por webcam e
+interpretados em tempo real por software de processamento de imagens,
+gerando padrões que controlam a transformação sonora da música. Dessa
+forma, pode-se dizer que os sons gerados refletem o voo dos origamis
+de acordo com suas geometrias (que geram trajetórias de voo
+caracteristicas).
 
-Na versão atual da AirHacktable, cada cor (vermelho, amarelo, azul, verde, preto, ou branco) controla uma voz, e a posição do origami na mesa modula aspectos do som.
-Em outras palavras, se o origami está flutuando mais à esquerda, o som sai no canal à esquerda da mesa, e se o origami está mais próximo da câmera, o volume aumenta, e se está mais afastado do operador, o som se torna mais agudo.
+Na versão atual da AirHacktable, cada cor (vermelho, amarelo, azul,
+verde, preto, ou branco) controla uma voz, e a posição do origami na
+mesa modula aspectos do som. Em outras palavras, se o origami está
+flutuando mais à esquerda, o som sai no canal à esquerda da mesa, e se
+o origami está mais próximo da câmera, o volume aumenta, e se está
+mais afastado do operador, o som se torna mais agudo.
 
-[imagens da AHT e dos patches em operação]
+A AHT segue a filosofia de desenvolvimento contínuo, comum ao software
+livre, que atualmente foi incorporado ao movimento chamado hardware
+livre ou aberto.
+
+[citar apresentações no contato, avav, oficinas sesc, chico, oficina semana imagem e som]
+
+[imagens da AHT e dos patches em operação, ver com renato quais ele quer]
 
@@
 
-Esta é uma sessão menor, dedicada a apontar repercussões emergentes destas empreitadas em comunidades diferentes
-e então tornar compreensível o desdobramento que estes códigos dedicados à música tiveram em processos sociais
-e mobilizações civis.
+Esta é uma sessão menor, dedicada a apontar repercussões emergentes
+destas empreitadas em comunidades diferentes e então tornar
+compreensível o desdobramento que estes códigos dedicados à música
+tiveram em processos sociais e mobilizações civis.
 
-Grupos foram montados em torno destes desenvolvimentos. Como os propósitos de compartilhamento
-e apropriação tecnológica estavam no cerne dos grupos relacionados, as investidas
-naturalmente tomaram teores engajados socialmente. A questão do empoderamento das pontas
-e da criação de um patrimônio tecnológico da humanidade é consequência quase imediata
-das posturas de compartilhamento e apropriação citados.
+Grupos foram montados em torno destes desenvolvimentos. Como os
+propósitos de compartilhamento e apropriação tecnológica estavam no
+cerne dos grupos relacionados, as investidas naturalmente tomaram
+teores engajados socialmente. A questão do empoderamento das pontas e
+da criação de um patrimônio tecnológico da humanidade é consequência
+quase imediata das posturas de compartilhamento e apropriação citados.
 

### 2012-11-11 devEresultados.tex
garibada na intro e conclusoes
realocados pastas de musicas e arquivos

@@
 Você escolhe o argumento de \label e pode ser qualquer coisa (Ex: \label{Procedimento_Experimental})
+\begin{quotation}
+\small
+'The increasing dominance of graphic interfaces for music software obscured 
+the continuing presence of the command-line tradition, 
+the code writer, the hacker. The code writing of deferred time 
+computer programming may be assembled out of time order, debugged and optimized.'
+
+\emph{Simon Emmerson, Living electronic music, 2007}
+\end{quotation}
 
@@
 relativas à onda mecânica do som que representam,
-em conjunto formam a chamada \emph{modulação por código de pulsos}\footnote{\emph{Pulse Code Modulation},
-de onde vem a sigla PCM, é o formato de codificação de som
-padrão em Compact Discs e outras mídias de alta fidelidade e sem compactação com perdas.}.
+em conjunto formam o chamado áudio PCM (veja sessão~\ref{sec:audio}).
 
@@
 \begin{equation}
-DTI=\frac{d'-d}{v_{som\;no\;ar}\approx 343,2 }\quad \text{segundos}
+DTI=\frac{d'-d}{v_{som\;no\;ar}\approx 343.2 }\quad \text{segundos}
 \end{equation}
@@
 \begin{equation}
-DII=20\log_{10}\left (\frac{d'}{d}\right) \quad dBs
-\end{equation}
-
-Convertendo para amplitude temos $DII_a=\frac{d'}{d}$. A $DII_a$ pode
-ser utilizada como constante multiplicativa do canal direito de um sinal sonoro estéreo ($\{t_i\}_0^{\Lambda -1}=\{DII_a .
-t_i'\}_0^{\Lambda -1}$).
+DII=20\log_{10}\left (\frac{d}{d'}\right) \quad dBs
+\end{equation}
+
+Convertendo para amplitude temos $DII_a=\frac{d}{d'}$. A $DII_a$ pode
+ser utilizada como constante multiplicativa do canal direito de um sinal sonoro estéreo ($\{t_i'\}_0^{\Lambda -1}=\{DII_a .
+t_i\}_0^{\Lambda -1}$).
 Pode-se utilizar a DII junto à DTI como adiantamento no tempo do canal direito com relação ao esquerdo, vínculo crucial para a percepção de localidade em sons graves e em sonoridades percussivas~\cite{Heeger}.
@@
 \Lambda_{DTI} & = \left \lfloor \frac{d'-d}{343,2}  f_a \right \rfloor \\
-DII_a & = \frac{d'}{d} \\
-\{t_{(i+\Lambda_{DTI})}\}_{\Lambda_{DTI}}^{\Lambda+\Lambda_{DTI}-1} & =\{DII_a . t_i'\}_0^{\Lambda-1} \\
-\{t_i\}_0^{\Lambda_{DTI}-1} & = 0
+DII_a & = \frac{d}{d'} \\
+\{t_{(i+\Lambda_{DTI})}'\}_{\Lambda_{DTI}}^{\Lambda+\Lambda_{DTI}-1} & =\{DII_a . t_i\}_0^{\Lambda-1} \\
+\{t_i'\}_0^{\Lambda_{DTI}-1} & = 0
 \end{split}
@@
 O princípio vertical está demonstrado nos \emph{quadros sonoros}, sons estáticos com espectros peculiares.
-Ambas as peças estão em código Python no APÊNDICE~\ref{cap:codigoPecas} e podem ser escutadas e vizualizadas nos links~\cite{reduced-fi} e~\cite{quadros}.
+Ambas as peças estão em código Python no Apêndice~\ref{cap:codigoPecas} e podem ser escutadas e vizualizadas nos links~\cite{reduced-fi} e~\cite{quadros}.
 
@@
 Caso utilizemos as equações~\ref{vbrF} teríamos uma distribuição diferente dos harmônicos.
-Para este fim, dispomos o APÊNDICE~\ref{cap:fmam} onde calculamos o conteúdo espectral da síntese FM feita com oscilações na escala logarítmica.
+Para este fim, dispomos o Apêndice~\ref{cap:fmam} onde calculamos o conteúdo espectral da síntese FM feita com oscilações na escala logarítmica.
 De fato, como pode-se notar nos cálculos, o comportamento simples que torna a FM atraente é obtido somente com as variações lineares utilizadas em~\ref{eq:fmEsp}.
@@
 Observe que utilizamos novamente a variação na escala linear de amplitude.
-Dispomos no APÊNDICE~\ref{cap:fmam} uma exposição do espectro da AM se realizada a oscilação na escala logarítmica de amplitude, que também perde o comportamento simples.
+Dispomos no Apêndice~\ref{cap:fmam} uma exposição do espectro da AM se realizada a oscilação na escala logarítmica de amplitude, que também perde o comportamento simples.
 

### 2012-11-11 scripts/sessao2.1/eqs.py

@@ -16,7 +16,6 @@ a.wavwrite(T_i,"silencio.wav",f_a)
 
 ############## 2.1.2 Volume
 ### 2.2
-
 Lambda=100 # o som terá 100 amostras
 T_i=n.random.random(Lambda) # 100 amostras quaisquer
 pot=(T_i**2).sum()/Lambda #potência
@@ -146,17 +145,4 @@ a.wavwrite(T_i,"mixados.wav",f_a)
 T_i=n.hstack((T1_i,T2_i,T3_i))
 # escrevendo arquivo
 a.wavwrite(T_i,"concatenados.wav",f_a)
-
-
-
-
-
-
-
-
-
-
-
-
-
 

### 2012-11-14 devEresultados.tex
feitas eqs da sessao 2.2.1

@@
 necessidade\footnote{Em contraposição ao cálculo das amostras individuais
-de forma analítica e permitindo o uso de funções sem possibilidade de cálculo direto
-(por serem amostras recolhidas diretamente de algum fenômeno natural, por exemplo).}. Em nosso caso, usaremos tabelas de busca para os
+de forma analítica. A LUT permite o uso de funções sem possibilidade de cálculo direto como
+por exemplo, amostras recolhidas diretamente de algum fenômeno natural.}. Em nosso caso, usaremos tabelas de busca para os
 sons criados, as notas. Assim, estas tabelas são unidimensionais e com
@@
 Descrevemos esta busca. Seja $\widetilde{\Lambda}$ o tamanho 
-da tabela e $\widetilde{L_i} = \{ \widetilde{l}_i \}_0^{\widetilde{\Lambda} -1}$ a tabela de elementos $\widetilde{l_i}$ de um
+da tabela e $\widetilde{L_i} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ a tabela de elementos $\widetilde{l_i}$ de um
 período de onda qualquer (veja equação ~\ref{periodoUnico}). Uma sequência
@@
 o procedimento de LUT na figura~\ref{fig:lut} onde calculamos explicitamente
-também uma amostra de $\{t_i\}$ a partir de $\{\widetilde{l}_i\}$ para uma frequência
+também uma amostra de $\{t_i\}$ a partir de $\left\{\,\widetilde{l}_i\,\right\}$ para uma frequência
 de $f=200Hz$, $\widetilde{\Lambda}=128$ e considerada a taxa de amostragem em $f_a=44.1kHz$.
+Esta não é uma configuração praticável, como veremos abaixo, mas possibilita uma
+disposição gráfica bastante clara do procedimento.
 

### 2012-11-14 devEresultados.tex
finalizados todos os scritps das equacoes da sessao 2.2

@@
 da transição). Este procedimento é mais comum para o caso de variações de volume através da energia
-da onda vibratória\footnote{O alteração do volume pode ocorrer por diversas características como
-a reverberação, a concentração de harmônicos agudos e a energia. Para nós, neste ponto da exposição,
-a mais facilmente controlada é a energia da onda (veja equação ~\ref{potencia}) e a energia também pode variar de diferentes formas.
-Uma forma mais simples e conveniente é a variar a amplitude pela multiplicação da sequência toda
+da onda vibratória\footnote{A mudança do volume (qualidade psicofísica) ocorre por diversas características 
+do som como a reverberação e a concentração de harmônicos agudos, e não apenas pela energia da onda.
+Para nós, entretanto, a mais facilmente controlada é de fato a energia da onda (veja equação ~\ref{potencia}) e esta também pode variar de diferentes formas.
+Uma forma mais simples e conveniente é variar a amplitude pela multiplicação da sequência toda
 por algum número real. A variação de energia sem variação de
-amplitude é comumente chamada de \emph{compressão sonora} e é muito útil e corriqueira na prática
-de produção musical atual~\cite{compressao}.}. Podemos obter este resultado multiplicando a sequência original
+amplitude é comumente chamada de \emph{compressão sonora} e é muito útil e corriqueira na
+produção musical atual~\cite{compressao}.}. Podemos obter este resultado multiplicando a sequência original
 (seja ela gerada ou pré-estabelecida) pela sequência $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$
@@
 \begin{equation}
-T_i^{'}=\left \{ t_i (10^{\frac{V_{dB}}{20}} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
+T_i^{'}=\left \{ t_i \left(10^{\frac{V_{dB}}{20}} \right)^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
 \end{equation}
@@
 de amplitude e frequência nas chamadas sínteses AM e FM e a aplicação das transições
-logarítmicas para os trêmolos e vibratos.
+logarítmicas para os tremolos e vibratos.
 
@@
 finita no tempo. Esta representação amostral é chamada
-de 'resposta ao impulso' $h_i$. Os filtros FIR são aplicados temporalmente ao som
+de 'resposta ao impulso' $\{h_i\}$. Os filtros FIR são aplicados temporalmente ao som
 digitalizado pela convolucão do som com a 
@@
 \begin{equation}
-t_i' = \frac{1}{b_0}\left ( \sum_{j=0}^Ja_j . t_{i-j} - \sum_{k=1}^Kb_k . t_{i-k}' \right )
+t_i' = \frac{1}{b_0}\left ( \sum_{j=0}^Ja_j . t_{i-j} + \sum_{k=1}^Kb_k . t_{i-k}' \right )
 \end{equation}
@@
 de 'biquadrado' pois sua função de transferência possue dois polos e dois zeros, i.e. sua
-forma normal consiste em dois polinômios quadráticos formando uma fração
-$\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1+ b_1.z^{-1} +b_2 . z^{-2}}$.}
+forma normal consiste em dois polinômios quadráticos formando uma fração:
+$\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1- b_1.z^{-1} -b_2 . z^{-2}}$.}
 ou rotinas para cálculo de coeficientes
@@
 das consagradas distribuições de probabilidade em diversas linguagens de programação,
-mas os resultados não são tão úteis, tendendo geralmente ao ruído branco
-e com difíceis previsões do espectro resultante~\cite{Cook}.
+mas os resultados não são tão úteis, tendendo geralmente ao ruído branco.\cite{Cook}
 
@@
 
-Abaixo elencamos alguns ruídos de espectros estáticos. São chamados \emph{coloridos} por terem sido associados a cores.
+Abaixo elencamos alguns ruídos de espectro estático. São chamados \emph{coloridos} por terem sido associados a cores.
 
@@
 f_i & = i \frac{f_a}{\Lambda} \;, \;\; \quad i \;\leq\; \frac{\Lambda}{2},\;\; i\;\in\;\mathbb{N}  \\
-\alpha_i & = (10^{-\frac{3}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )}  \\
+\alpha_i & = \left(10^{-\frac{3}{20}}\right)^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )}  \\
 c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \\
@@
 
-\subsection{Usos musicais parte 1: trêmolo e vibrato, AM e FM}
+\subsection{Usos musicais parte 1: tremolo e vibrato, AM e FM}
 
@@
 o tremolo é uma variação periódica de volume (intensidade).\footnote{Alguns instrumentos e contextos musicais usam nomenclaturas diferentes.
-Por exemplo, no piano, o chamado trêmolo é um vibrato e um tremolo segundo a classificação aqui utilizada.
+Por exemplo, no piano, o chamado tremolo é um vibrato e um tremolo segundo a classificação aqui utilizada.
 As definições presentes neste trabalho tem por base uma literatura mais abrangente do que a utilizada para um único instrumento, prática ou tradição musical e comum em contextos de teoria musical e musica eletrônica~\cite{Lacerda,Harmonia}.}
@@
 
-Onde $V_{dB}$ é a profundidade da oscilação em decibels do trêmolo e $a_{\text{máx}}=10^{\frac{V_{dB}}{20}}$
+Onde $V_{dB}$ é a profundidade da oscilação em decibels do tremolo e $a_{\text{máx}}=10^{\frac{V_{dB}}{20}}$
  é o ganho máximo de amplitude envolvido.
@@
 A figura~\ref{fig:tremolo} mostra a amplitude das sequências $\{a_i\}_0^{\Lambda-1}$ e $\{t_i'\}_0^{\Lambda-1}$
-para três oscilações de um trêmolo com forma da dente de serra. A curvatura é devido à progressão logarítmica de
+para três oscilações de um tremolo com forma da dente de serra. A curvatura é devido à progressão logarítmica de
 intensidade.
-A frequência do trêmolo é de $1,5Hz$ pois $f_a=44,1kHz \; \Rightarrow \; \text{duração} = \frac{i_{\text{máx}}=82000}{f_a}= 2s \; \Rightarrow \; \frac{3\text{oscilações}}{2s}=1,5$ oscilações por segundo ($Hz$).
+A frequência do tremolo é de $1,5Hz$ pois $f_a=44,1kHz \; \Rightarrow \; \text{duração} = \frac{i_{\text{máx}}=82000}{f_a}= 2s \; \Rightarrow \; \frac{3\text{oscilações}}{2s}=1,5$ oscilações por segundo ($Hz$).
 
@@
 tanto na tradição erudita quanto na música eletrônica atual, especialmente no \emph{Dubstep}~\cite{DubRugoso}.
-A rugosidade também é gerada através de conteúdos espectrais que geram batimentos~\cite{Porres}.  A sequência \emph{Rugosidades}
+A rugosidade também é gerada através de conteúdos espectrais que geram batimentos~\cite{Porres}.  A sequência \emph{Bela Rugosi's Dade}
 explora este limiar com concomitâncias de tremolos e vibratos na mesma
@@
 como eventos identificáveis. Neste caso, as oscilações se apresentam na forma de espectro
-audível como altura, $f'$, $\mu$ e a forma de onda realizam alterações espectrais no som
+audível como altura. Assim, $f'$, $\mu$ e a forma de onda realizam alterações espectrais no som
 original $T_i$ de formas diferentes para os tremolos e para os vibratos, são as 
@@
 Uma primeira possibilidade interessante é o estabelecimento de vínculos
-entre os parâmetros do trêmolo e do vibrato e algum parâmetro da nota básica,
+entre os parâmetros do tremolo e do vibrato e algum parâmetro da nota básica,
 digamos a frequência. Assim, podemos estabelecer que quanto mais aguda é a nota,
-mais alta é a frequência do vibrato e do trêmolo mas menor a sua profundidade.
+mais alta é a frequência do vibrato e do tremolo mas menor a sua profundidade.
 Desta forma, tomando por base as equações \ref{vbrGamma}, \ref{vbrF} e \ref{trA}
@@
 potencialmente aleatórias, em paramêtros como frequência central, presença (ou ausência) de vibrato
-e trêmolo e suas características, equalizações, volume etc. Para o resultado final, estas versões do som
+e tremolo e suas características, equalizações, volume etc. Para o resultado final, estas versões do som
 inicial são então mixadas (ver equação \ref{eq:mixagem}). A peça \emph{Chorus infantil} realiza chorus de formas
@@
 \begin{split}
-\{a_i\}_0^{\Lambda_A-1} & = \left\{2^{\frac{i}{\Lambda_A-1}}-1\right\}_0^{\Lambda_A-1} \\
+\{a_i\}_0^{\Lambda_A-1} & = \left\{\xi\left(\frac{1}{\xi}\right)^{\frac{i}{\Lambda_A-1}}\right\}_0^{\Lambda_A-1} \\
 \{a_i\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} & =\left\{a_S^{\frac{i-\Lambda_A}{\Lambda_D-1}}  \right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} \\
 \{ a_i \}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} & =\left\{ a_S \right\}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} \\
-\{ a_i \}_{\Lambda-\Lambda_R}^{\Lambda-1} & =\left\{ (a_S+1)\left(\frac{1}{a_S+1} \right)^{\frac{i+\Lambda_R-\Lambda}{\Lambda-\Lambda_R}} -1 \right\}_{\Lambda-\Lambda_R}^{\Lambda-1}
-\end{split}
-\end{equation}
-
-Com $\Lambda_X=\lfloor \Delta . f_a \rfloor\;\;\forall\;\; X$.
+\{ a_i \}_{\Lambda-\Lambda_R}^{\Lambda-1} & =\left\{ a_S\left(\frac{\xi}{a_S} \right)^{\frac{i-(\Lambda-\Lambda_R)}{\Lambda_R-1}} \right\}_{\Lambda-\Lambda_R}^{\Lambda-1}
+\end{split}
+\end{equation}
+
+Com $\Lambda_X=\lfloor \Delta .
+f_a \rfloor\;\;\forall\;\; X \; \in (A,D,R,\;)$ e $\xi$ um valor bastante pequeno que torne o \emph{fade in} e o \emph{fade out} satisfatórios, e.g.
+$\xi=10^{\frac{-120}{20}}=10^{-6}\;$ ou $\;\xi=10^{\frac{-180}{20}}=10^{-9}$.
 
@@
 máxima perdura. Outro exemplo comum é a elaboração de traçados mais elaborados para o
-ataque ou para o decaimento.
+ataque ou para o decaimento. A montagem musical \emph{ADa e SaRa} explora diversas destas
+configurações da envoltória ADSR.
 

### 2012-11-14 scripts/adsr.py

@@ -8,14 +8,15 @@ DA=2*20. # duracao do ataque \Delta_A
 DD=2*20. # duracao do decay \Delta_D
 DR=2*20. # duracao do release \Delta_R
 SS=.4 # fração da amplitude em que ocorre o sustain
+xi=1e-2 # -180dB para iniciar o fade in e finalizar o fade out
 
-A=2**(n.arange(DA)/(DA-1))-1 # amostras do ataque
+A=xi*(1./xi)**(n.arange(DA)/(DA-1)) # amostras do ataque
 s=n.copy(A) 
-D=SS**(n.arange(DA,DA+DD)/DD-1) # amostras do decay
+D=SS**((n.arange(DA,DA+DD)-DA)/(DD-1)) # amostras do decay
 s=n.hstack((  s,  D  ))
 S=SS*n.ones(De-DR-(DA+DD)) # amostras do sustain
 s=n.hstack((  s, S  ))
-R=(SS+1)*(1/(SS+1))**((n.arange(De-DR,De)+DR-De)/(DR))-1 # amostras do release
+R=(SS)*(xi/SS)**(  (n.arange(De-DR,De)+DR-De)/(DR-1)  ) # amostras do release
 s=n.hstack((  s,  R  ))
 
 p.plot(A,                      'bo', markersize=4)

### 2012-11-15 scripts/pecas2.2/vibraEtreme.py
adiantadas pecas da sessao 2

@@ -1,5 +1,7 @@
 #-*- coding: utf8 -*-
 import numpy as n, scikits.audiolab as a
+H=n.hstack
+V=n.vstack
 
 # montagem dedicada a explorar tremolos e vibratos
 # independentemente e combinados
@@ -43,5 +45,226 @@ Tr_i=n.hstack(  ( foo , foo*-1 )   )
 # Dente de Serra:
 D_i=n.linspace(-1,1,Lt)
 
+
+def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
+    Lambda=n.floor(f_a*d)
+    ii=n.arange(Lambda)
+    Lv=float(len(S_i))
+
+    Gammav_i=n.floor(ii*fv*Lv/f_a) # índices para a LUT
+    Gammav_i=n.array(Gammav_i,n.int)
+    Tv_i=tabv[Gammav_i%int(Lv)] # padrão de variação do vibrato para cada amostra
+
+    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) # frequência em Hz em cada amostra
+
+    D_gamma_i=F_i*(Lt/float(f_a)) # a movimentação na tabela por amostra
+    Gamma_i=n.cumsum(D_gamma_i) # a movimentação na tabela total
+    Gamma_i=n.floor( Gamma_i) # já os índices
+    Gamma_i=n.array( Gamma_i, dtype=n.int) # já os índices
+    return Tr_i[Gamma_i%int(Lt)] # busca dos índices na tabela
+
+aa=v()
+bb=v(nu=4.)
+cc=v(f=300)
+dd=v()
+a.wavwrite(n.hstack((aa,bb,cc,dd)),"vbr.wav",f_a)
+
+aa=v(f=800)
+bb=v(nu=7.)
+cc=v(f=300,fv=3)
[…]

### 2012-11-15 devEresultados.tex
adiantadas pecas da sessao 2

@@
 de onda apresentam cadeias de componentes harmônicas e usos específicos.
-Além da senóide (eq.~\ref{senoide}), as formas de onda especificadas em~\ref{denteDeSerra},~\ref{triangular} e~\ref{quadrada} são dispostas graficamente na figura~\ref{fig:formasDeOnda}. 
+Além da senoide (eq.~\ref{senoide}), as formas de onda especificadas em~\ref{denteDeSerra},~\ref{triangular} e~\ref{quadrada} são dispostas graficamente na figura~\ref{fig:formasDeOnda}. 
 São as formas de onda artificiais tradicionalmente usadas na música para síntese e controle oscilatório de variáveis apresentam diversos usos também fora da música~\cite{Openheim}.

### 2012-11-18 pretextual.tex
arrumada formatacao de todo pretextual

@@
 \pretextualchapter{}
-	Lado dedicado à folha de aprovação. Apagar isso antes de imprimir a versão oficial
+[[	Lado dedicado à folha de aprovação. Apagar isso antes de imprimir a versão oficial ]]
 
-\pretextualchapter{Dedicatória} %Título da página dedicatória
+\afterpage{\blankpage}
+
+\pretextualchapter{DEDICATÓRIA} %Título da página dedicatória
 	\input{dedicatoria} %Para incluir o texto de dedicatória,use o arquivo dedicatória.tex
 
-\pretextualchapter{Agradecimentos} %Título da página de agradecimentos
+\afterpage{\blankpage}
+
+\pretextualchapter{AGRADECIMENTOS} %Título da página de agradecimentos
 	\input{agradecimentos} %Para incluir o texto de agradecimentos,use o arquivo agradecimentos.tex
+
+\afterpage{\blankpage}
 
@@
 	\end{epigrafebot}
+
+\afterpage{\blankpage}
+
 \resumoeabstract %Para incluir o texto do resumo e abstract,use o arquivo resumoeabstract.tex
+
+\afterpage{\blankpage}
 
@@
 
-\listadetabelas %Comando que gera a lista de tabelas (AUTOMÁTICO)
+%\listadetabelas %Comando que gera a lista de tabelas (AUTOMÁTICO)
 
-\pretextualchapter{Lista de Abreviaturas}
+\pretextualchapter{Lista de relações descritas e implementadas}
 	\begin{listaespecial}[BIGNAMEWIDTH]
-		\item[OSS] Programas de Código Aberto \emph{(Open Source Software)}
-		\item[GNU] \emph{GNU is Not UNIX}
-		\item[GPL] \emph{General Public Licence}
-                \item[AA] \emph{Algorithmic Autoregularion ou Acrônimo Ambíguo}
-                \item[AC] \emph{Autogestão Coletiva ou Ágora Communs}
-                \item[CC] \emph{Creative Commons}
-                \item[ABT] \emph{ABeatTracker}
-                \item[ABD] \emph{ABeatDetector}
-                \item[PD] \emph{Pure Data}
-                \item[CMDCA] \emph{Conselho Municipal de defesa dos Direitos da Cirança e do Adolescente}
-                \item[SOS] \emph{Saúde Olha Sabedoria: uma proposta de coleta e difusão de conhecimentos relacionados à saúde}
-                \item[LADSPA] \emph{Linux Audio Developers Simple Plugin API}
-                \item[LV2] \emph{LADSPA Version 2}
-                \item[JACK] \emph{Jack Audio Connection Kit}
-                \item[SIP] \emph{Scilab Imaging Processing toolbox}
-                \item[MIT] \emph{Massachusetts Institute of Technology}
-                \item[NUMPY] \emph{Numerical Python}
-                \item[SCIPY] \emph{Scientific Python}
-                \item[ONU] \emph{Organização das Nações Unidas}
-                \item[SL] \emph{Software Livre}
-                \item[EKP] \emph{Emotional Kernel Panic}
-                \item[SOS] \emph{Saúde Olha Sabedoria}
-                \item[FISL] \emph{Festival Internacional de Software Livre}
-                \item[BPM] \emph{Batidas Por Minuto (medida de andamento musical)}
-                \item[API] \emph{Application Programming Interface ou Interface de Programação de Aplicativos}
-                \item[WP] \emph{Wavelet Packet}
-
-		\item[EL] \emph{Estúdio Livre}
-		\item[AE] \emph{AudioExperiments}
-		\item[PCM] \emph{Pulse Code Modulation (modulação por código de pulsos)}
-		\item[DTI] \emph{Diferença de Tempo Interaural (\emph{Interaural Time Difference} - ITD)}
-		\item[DII] \emph{Diferença de Intensidade Interaural (\emph{Interaural Intensity Difference)} - IID ou \emph{Interaural Level Difference} - ILD}
+		\item[AM]: modulação de amplitude como descrita em~\ref{}. Implementação computacional em Python com Numpy e Scikits/Audiolab no apêndice~\ref{}.
 	\end{listaespecial}
 
-\pretextualchapter{Lista de Símbolos}
+\afterpage{\blankpage}
+
+\pretextualchapter{Lista de peças musicais}
 	\begin{listaespecial}[BIGNAMEWIDTH]
-		\item[\$] Indica que o resto da linha é um comando bash$\Lambda_{DTI}=DTI . f_a$
-		\item[\#] Resto da linha é comentario em código Bash ou Python
-		\item[] Resto da linha é comentario em código C/C++
+		\item[Chorus infantil]: relações harmônicas da seção~\ref{}. Música disponibilizada online~\cite{}. Arquivo Python no apêndice~\ref{}.
 	\end{listaespecial}
 
-\pretextualchapter{Glossário de Termos}
-	\begin{listaespecial2}[BIGNAMEWIDTH]
-		\item[Montagem musical :] algumas ideias musicais rascunhadas, sem a pretensão de ser uma peça ou uma obra musical. Sinônimo de Sequência musical.
-		\item[Sequência musical :] sinônimo de montagem musical. 
-		\item[Tessitura :] âmbito de altura ($\log(freq)$) do objeto em discussão (instrumento, passagem musical, etc).
-
-		\item[Delay :] embora a palavra signifique simplesmente atraso, no contexto de música a áudio o \emph{delay} comumente se refere a um efeito causado por diversas repetições de um som original, com atrasos diferentes.
-	\end{listaespecial2}
+\afterpage{\blankpage}
+%\pretextualchapter{Lista de Abreviaturas}
+%	\begin{listaespecial}[BIGNAMEWIDTH]
+%		\item[OSS] Programas de Código Aberto \emph{(Open Source Software)}
+%		\item[GNU] \emph{GNU is Not UNIX}
+%		\item[GPL] \emph{General Public Licence}
+%                \item[AA] \emph{Algorithmic Autoregularion ou Acrônimo Ambíguo}
+%                \item[AC] \emph{Autogestão Coletiva ou Ágora Communs}
+%                \item[CC] \emph{Creative Commons}
+%                \item[ABT] \emph{ABeatTracker}
+%                \item[ABD] \emph{ABeatDetector}
+%                \item[PD] \emph{Pure Data}
+%                \item[CMDCA] \emph{Conselho Municipal de defesa dos Direitos da Cirança e do Adolescente}
+%                \item[SOS] \emph{Saúde Olha Sabedoria: uma proposta de coleta e difusão de conhecimentos relacionados à saúde}
+%                \item[LADSPA] \emph{Linux Audio Developers Simple Plugin API}
+%                \item[LV2] \emph{LADSPA Version 2}
+%                \item[JACK] \emph{Jack Audio Connection Kit}
+%                \item[SIP] \emph{Scilab Imaging Processing toolbox}
+%                \item[MIT] \emph{Massachusetts Institute of Technology}
+%                \item[NUMPY] \emph{Numerical Python}
+%                \item[SCIPY] \emph{Scientific Python}
+%                \item[ONU] \emph{Organização das Nações Unidas}
+%                \item[SL] \emph{Software Livre}
+%                \item[EKP] \emph{Emotional Kernel Panic}
+%                \item[SOS] \emph{Saúde Olha Sabedoria}
+%                \item[FISL] \emph{Festival Internacional de Software Livre}
+%                \item[BPM] \emph{Batidas Por Minuto (medida de andamento musical)}
+%                \item[API] \emph{Application Programming Interface ou Interface de Programação de Aplicativos}
+%                \item[WP] \emph{Wavelet Packet}
+%
+%		\item[EL] \emph{Estúdio Livre}
+%		\item[AE] \emph{AudioExperiments}
+%		\item[PCM] \emph{Pulse Code Modulation (modulação por código de pulsos)}
+%		\item[DTI] \emph{Diferença de Tempo Interaural (\emph{Interaural Time Difference} - ITD)}
+%		\item[DII] \emph{Diferença de Intensidade Interaural (\emph{Interaural Intensity Difference)} - IID ou \emph{Interaural Level Difference} - ILD}
+%	\end{listaespecial} 
+%
 

### 2012-11-18 agradecimentos.tex
terminado pretextual dedicatoria e epigrafe

@@
-Em primeiro lugar, agradeço aos meus pais e demais familiares. Agradeço aos
-meus amigos pelo carinho e companhia. Aos professores por tanta atenção e compreensão.
-Aos colegas colecionadores de entendimentos por ter-nos construído até aqui.
-Agradeço a Felipe Machado pela visão, a Fabiana 'Goa' Sherine pelo entendimento,
-a Glerm Soares pela transcendência, a Fabianne Baveldi pela excelência, dentre tantos outros que não há páginas para escrever aqui.
+Agradeço ao Prof. Luciano da Fontoura Costa e ao Prof. Osvaldo Novais de Oliveira Junior pela oportunidade e pela orientação deste trabalho.
+
+\vspace{4 mm}
+
+Agradeço ao corpo de funcionários do IFSC, pela prestatividade e eficiência em todos os momentos que precisei.
+
+\vspace{4 mm}
+
+Agradeço à minha família, em especial à minha esposa Thaís Teixeira Fabbri e ao meu filho Antônio Anzoategui Fabbri.
+
+\vspace{4 mm}
+
+Agradeço ao meu irmão Ricardo Fabbri e ao amigo Vilson Vieira pelas profícuas colaborações acadêmicas, artísticas e socialmente engajadas.
+
+\vspace{4 mm}
+
+Agradeço aos participantes do labMacambira.sf.net pelas colaborações diretas e indiretas nos software, apresentações e articulações desde junho de 2011.
+Em especial agradeco ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela e Patrícia Ferraz, Glerm Soares, Vanessa Ferreira e Sérgio Teixeira de Carvalho.
+Agradeço aos demais colaboradores das listas de email, IRC, AA e outros canais.
+
+\vspace{4 mm}
+
+Agradeço ao Cultura Viva e aos pontos de cultura pelo apoio fundamental a este e outros trabalho.
+Em especial agradeço ao CDTL (PE), JuntaDados.org (BA), Nós Digitais (SP), Casa dos Meninos (SP), Nina Griô (SP), Pontão da Eco (RJ), Casa de Cultura Tainã (SP) e à Comissão Nacional dos Pontos de Cultura - CNPdC (GO).
+
+\vspace{4 mm}
+
+Agradeço às comunidades de cultura e software livre por todos os conhecimentos e tecnologias repassados e que compõem este trabalho.
+
+%
+%
+%
+%Em primeiro lugar, agradeço aos meus pais e demais familiares. Agradeço aos
+%meus amigos pelo carinho e companhia. Aos professores por tanta atenção e compreensão.
+%Aos colegas colecionadores de entendimentos por ter-nos construído até aqui.
+%Agradeço a Felipe Machado pela visão, a Fabiana 'Goa' Sherine pelo entendimento,
+%a Glerm Soares pela transcendência, a Fabianne Baveldi pela excelência, dentre tantos outros que não há páginas para escrever aqui.
 Agradeço ao profícuo Vilson Vieira, que tornou-se especialmente próximo
-das contribuições citadas neste trabalho, construindo sobre elas e com elas. Condizente com a fortíssima referência
-para que é para mim, tanto para os assuntos tratados nesta dissertação quanto para questões humanas, um especial agradecimento
-ao meu irmão Ricardo Fabbri que também se aproximou destas contribuições tanto cuidando de processamento de imagens e video quanto articulando e amadurecendo as ideias. Agradecimento
-forte ao LabMacambira.sf.net. Um profundo agradecimento
-aos meus orientadores de IC: Adolfo Maia Junior e Jônatas Manzolli. Agradeço a José Augusto Mannis
-pelos trabalhos de estudante, pelo acompanhamento e ensinos sobre manipulação de áudio e música como idioma.
-Agradeço fortemente ao Prof. Dr. Rafael Santos Mendes pela pesquisa em Wavelets de mestrado em engenharia elétrica
-praticamente completa na FEEC/Unicamp, titulo este que não defendendi por razões circunstanciais. Agradeço aos professores da FEEC pelos ensinos de
-primeiríssima qualidade em diversas disciplinas como processamento de sinais, teoria da informação, 
-circuitos elétricos, processos estocásticos
-e programação bio-inspirada. Sem sombra de dúvida, os maiores agradecimentos para os assíduos mentores
-deste trabalho: ao orientador-colaborador Prof. Luciano da Fontoura Costa e ao meu 
-orientator Prof. Osvaldo Novais de Oliveira Júnior. Ambos proporcionaram inestimável acompanhamento e notável paciência, dedicação e precisão em todos os momentos.
+%das contribuições citadas neste trabalho, construindo sobre elas e com elas. Condizente com a fortíssima referência
+%para que é para mim, tanto para os assuntos tratados nesta dissertação quanto para questões humanas, um especial agradecimento
+%ao meu irmão Ricardo Fabbri que também se aproximou destas contribuições tanto cuidando de processamento de imagens e video quanto articulando e amadurecendo as ideias. Agradecimento
+%forte ao LabMacambira.sf.net. Um profundo agradecimento
+%aos meus orientadores de IC: Adolfo Maia Junior e Jônatas Manzolli. Agradeço a José Augusto Mannis
+%pelos trabalhos de estudante, pelo acompanhamento e ensinos sobre manipulação de áudio e música como idioma.
+%Agradeço fortemente ao Prof. Dr. Rafael Santos Mendes pela pesquisa em Wavelets de mestrado em engenharia elétrica
+%praticamente completa na FEEC/Unicamp, titulo este que não defendendi por razões circunstanciais. Agradeço aos professores da FEEC pelos ensinos de
+%primeiríssima qualidade em diversas disciplinas como processamento de sinais, teoria da informação, 
+%circuitos elétricos, processos estocásticos
+%e programação bio-inspirada. Sem sombra de dúvida, os maiores agradecimentos para os assíduos mentores
+%deste trabalho: ao orientador-colaborador Prof. Luciano da Fontoura Costa e ao meu 
+%orientator Prof. Osvaldo Novais de Oliveira Júnior. Ambos proporcionaram inestimável acompanhamento e notável paciência, dedicação e precisão em todos os momentos.

### 2012-11-18 pretextual.tex
terminado pretextual dedicatoria e epigrafe

@@
 
-\afterpage{\blankpage}
-
 \pretextualchapter{}
-%	\begin{epigrafetop}
-%		{from Pantheon import Obatalá\, \, \, \\
-%                from World import sharing, kitten \\
-%                \vspace{.1in}
-%                while True: \; \; \; \; \; \; \; \; \; \; \; \; \; \; \; \,\\
-%                    if sharing == 0: \; \; \; \; \;  \; \; \; \; \; \\
-%                        Obatalá.kill( kitten ) \; \; \;}
-%        {Autoria coletiva e anônima (2010)}
-%	\end{epigrafetop}
-
 	\begin{epigrafetop}
-		{Music is a hidden arithmetic exercise of the soul, which does not know that it is counting.}
+		{Music is a hidden arithmetic exercise of the soul, \\ which does not know that it is counting.}
         {Gottfried Leibniz (1646-1716)}
@@
 	\begin{epigrafemid}
-		{Music is a hidden metaphysical exercise of the soul, which does not know that it is philosophizing.} %frase
+		{Music is a hidden metaphysical exercise of the soul, \\ which does not know that it is philosophizing.} %frase
 		{Arthur Schopenhauer (1788-1860)} %referência
@@
 	\vspace{-1cm}
+%	\begin{epigrafebot}
+%		{Electronic music used pure sounds, completely calibrated. You had to think digitally, as it were, in a way that allowed you to extend serial ideas into other parameters through technology.} %frase
+%		{Luc Ferrari (1929 - 2005)} %referência
+%	\end{epigrafebot}
+
 	\begin{epigrafebot}
-		{Electronic music used pure sounds, completely calibrated. You had to think digitally, as it were, in a way that allowed you to extend serial ideas into other parameters through technology.} %frase
-		{Luc Ferrari (1929 - 2005)} %referência
+		{from Pantheon import Obatalá\, \, \, \\
+                from World import sharing, kitten \\
+                \vspace{.1in}
+                while True: \; \; \; \; \; \; \; \; \; \; \; \; \; \; \; \,\\
+                    if sharing == 0: \; \; \; \; \;  \; \; \; \; \; \\
+                        Obatalá.kill( kitten ) \; \; \;}
+        {Autoria coletiva e anônima (2010)}
 	\end{epigrafebot}

### 2012-11-18 agradecimentos.tex
arrumada geral, adentrando texto

@@
 
-Agradeço ao meu irmão Ricardo Fabbri e ao amigo Vilson Vieira pelas profícuas colaborações acadêmicas, artísticas e socialmente engajadas.
+Agradeço ao meu irmão Ricardo Fabbri e ao amigo Vilson Vieira da Silva Junior pelas profícuas colaborações acadêmicas, artísticas e socialmente engajadas.
 
@@
 Agradeço aos participantes do labMacambira.sf.net pelas colaborações diretas e indiretas nos software, apresentações e articulações desde junho de 2011.
-Em especial agradeco ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela e Patrícia Ferraz, Glerm Soares, Vanessa Ferreira e Sérgio Teixeira de Carvalho.
+Em especial agradeco ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela e Patrícia Ferraz, Glerm Soares, Vanessa Ferreira, Sérgio Teixeira de Carvalho.
 Agradeço aos demais colaboradores das listas de email, IRC, AA e outros canais.
@@
 
-Agradeço ao Cultura Viva e aos pontos de cultura pelo apoio fundamental a este e outros trabalho.
-Em especial agradeço ao CDTL (PE), JuntaDados.org (BA), Nós Digitais (SP), Casa dos Meninos (SP), Nina Griô (SP), Pontão da Eco (RJ), Casa de Cultura Tainã (SP) e à Comissão Nacional dos Pontos de Cultura - CNPdC (GO).
+Agradeço ao Cultura Viva e aos pontos de cultura pelo apoio fundamental a este e outros trabalhos.
+Agradeço em especial a estes Pontões e Cultura Digital e Ação Griô pelo suporte a estes desenvolvimentos em ações regionais e nacionais: CDTL (PE), JuntaDados.org (BA), Nós Digitais (SP), Casa dos Meninos (SP), Nina Griô (SP), Pontão da Eco (RJ), Casa de Cultura Tainã (SP) e à Comissão Nacional dos Pontos de Cultura - CNPdC (GO).
 

### 2012-11-18 resumoeabstract.tex
arrumada formatacao de todo pretextual
arrumada geral, adentrando texto

--- 2012-11-08 resumoeabstract.tex
@@
-\begin{resumo}
+\begin{resumo2}
+\vspace{-10mm}
+FABBRI, R. \textbf{\ABNTtitulodata}. 2012. 151p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, 2012.
+\vspace{15mm}
 
-A representação dos elementos básicos da música - tais como notas musicais, ornamentos e estruturas intervalares - em termos do som discretizado é bastante utilizada em softwares e rotinas para criação musical e tratamento sonoro.
-Não há, entretanto, um sistema unificado e conciso que relacione matematicamente esses elementos às sequências de amostras no tempo.
-Nesta dissertação, implementamos um sistema computacional MASSA (Música e Áudio em Sequências e Séries Amostrais) em que cada elemento musical é descrito por equações que resultam diretamente nas sequências temporais do som em sua representação discretizada.
+A representação dos elementos básicos da música - tais como notas, ornamentos e estruturas intervalares - em termos do som discretizado é bastante utilizada em software e rotinas para criação musical e tratamento sonoro.
+Não há, entretanto, uma abordagem concisa que relacione estes elementos às amostras sonoras.
+Nesta dissertação, cada elemento musical é descrito por equações que resultam diretamente nas sequências temporais do som em sua representação discretizada.
 O elemento fundamental, a nota musical básica com duração, volume, altura e timbre, é relacionado quantitativamente às características do sinal digital.
-As variações internas das notas, como trêmolos, vibratos e flutuações espectrais, também são contempladas, o que permite sintetizar notas e sonoridades com inspiração nos instrumentos musicais reais além de sonoridades novas.
-A partir dessa representação das notas musicais, é possível gerar estruturas musicais através de recursos consagrados, como a métrica rítmica, os intervalos de altura e os ciclos.
-As equações deram origem a uma implementação na linguagem Python, com scripts que resultam em exemplos sonoros simples e imagens como espectrogramas, variação da amplitude e princípios organizacionais da música.
-A eficácia do sistema foi comprovada com a síntese de pequenas peças usando notas básicas, notas incrementadas e notas em música.
-É possível, também, sintetizar álbuns inteiros através de colagens de scripts e parametrização especificada pelo usuário.
-Com o paradigma de implementação em código aberto, o sistema pode ser expandido em processos de co-autoria e usado livremente por músicos e outros interessados.
-De fato, o sistema já vem sendo empregado por usuários externos para fins diversos, que incluem produção de músicas, experimentos psicoacústicos e difusão da linguagem computacional explorando o apelo lúdico do audiovisual.
+As variações internas, como tremolos, vibratos e flutuações espectrais, também são contempladas, o que permite sintetizar notas com inspiração nos instrumentos musicais reais além de sonoridades novas.
+A partir desta representação das notas, dispomos de recursos para a geração de estruturas musicais, como a métrica rítmica, os intervalos de altura e os ciclos.
+As equações deram origem a uma caixa de ferramentas computacionais: \emph{scripts} que realizam cada equação e geram exemplos sonoros simples.
+Nomeamos \massa, Música e Áudio em Sequências e Séries Amostrais, este ferramental e sua eficácia foi comprovada com a síntese de pequenas peças usando notas básicas, notas incrementadas e notas em música.
+É possível, também, sintetizar álbuns inteiros através de colagens dos scripts e parametrização especificada pelo usuário.
+Com o paradigma de implementação em código aberto, a (caixa de ferramentas) \massa\  pode ser expandida em processos de co-autoria e usada livremente por músicos, engenheiros e outros interessados.
+De fato, o sistema já foi empregado por usuários externos para a produção de músicas, apresentações artísticas, experimentos psico-acústicos e a difusão da linguagem computacional através do apelo lúdico dos artefatos audiovisuais.
 
 $\phantom{linha em branco}$\\
-Palavras-chave: som, acústica, áudio, código aberto, cultura digital, cultura hacker, programação
+Palavras-chave: Som.
+Áudio.
+Áudio digital.
+Acústica.
+Música.
+Música digital.
+Psicofísica.
+Alta fidelidade.
+Hi-fi.
+Hifi.
+Síntese sonora.
+Síntese musical.
+Python.
+Livecoding.
+Arte e tecnologia.
+Código aberto.
+Cultura hacker.
+Cultura digital.
+Cultura livre.
 
-\end{resumo}
+\end{resumo2}
 
-\begin{abstract}
+\afterpage{\blankpage}
 
-tradução de:
-Nesta dissertação, expomos os fundamentos do manuseio do som digitalizado através de recursos em código aberto como
-uma forma de manifestação cultural.
-Após uma brevíssima contextualização sobre o status que o código e o áudio digital possuem na cultura atual, procuramos abordar como um todo o fenômeno
-físico (e psicofísico) do som, a sua representação digital e os procedimentos clássicos de manuseio. As linguagens de programação C/C++ e Python
-são então abordadas já com vistas ã tarefa de lidar com o áudio digital. Um capítulo especial é dedicado às linguagens dedicadas ao áudio e à música. Ainda no campo do software,
-o ambiente de desenvolvimento Unix é descrito de um ponto
-de vista bem resumido e ferramental, pois é a base para todos os desenvolvimentos expostos nesta dissertação.
-Os resultados antingidos com os fundamentos expostos são então sistematizados com ênfase nas ferramentas desenvolvidas.
-Apresentamos também uma série de repercussões destes desenvolvimentos, em especial uma frente de ação criada: o LabMacambira.sf.net.
-Por fim, como um produto adicional relacionado a esta dissertação, é delineado o mapa de exposição e comunicação no qual
-os desenvolvimentos deste trabalho estão agregados. Esta estrutura é dedicada a manter as contribuições
-de forma continuada e sistematizada através de webpages, wikis, repositórios, um canal IRC, aplicativos de acompanhamento de projetos
-e uma das tecnologias desenvolvidas neste trabalho: a Autoregulação Algorítmica.
+\begin{abstract2}
+\vspace{-10mm}
+FABBRI, R. \textbf{Music on digital audio: psychophysical approach and toolbox}. 2012. 151p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, 2012.
+\vspace{15mm}
+
+The representation of music's basic elements - such as notes, ornaments and intervalar structures - in terms of dicrete-time audio is used in software and routines for musical creation and sound design.
+Even so, there is no unified approach that relates directly this elements to the sound discrete-time samples.
+In this dissertation, each musical element is described by equations that result directly on the sonic time samples.
+The fundamental element, the musical note with duration, volume, pitch and timbre, is related quantitatively to the characteristics of the digital signal.
+Internal variations, such as tremolos, vibratos and spectral fluctuations are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities.
+With this representation of notes, we predispose canonical resources for the generation of musical structures, such as rhythmic meter, pitch intervals and cycles.
+The equations gave rise to a computer toolbox: scripts that accomplish each equation and result simple sonic examples.
+We name \mass, Music and Audio in Sequences and Samples, this resources and its efficacy was confirmed by the synthesis of small musical pieces using basic notes, incremented notes and notes in music.
+It is possible, as well, to synthesize whole albums through collage of the scripts and parametrization specified by the user.
+With the paradigm of open source implementation, the \mass\ toolbox can be expanded in processes of co-authorship and used freely by musicians, engineers and other interested parties.
+In fact, the system was already employed by external users for diverse purposes which include music production, artistic presentations, psicoacoustic experiments and computer language diffusion by means of the playful appeal of audiovisual artifacts.
 
 $\phantom{linha em branco}$\\
-Keywords: audio, programming, acoustics, open source, digital culture, hacker culture
+Keywords: Sound.
+Audio.
+Digital audio.
+Acoustics.
+Music.
+Digital music.
+Psicophysics.
+High fidelity.
+Hi-fi.
+Hifi.
+Sound synthesis.
+Musical synthesis.
+Python.
+Livecoding.
+Art and technology.
+Open source software.
+Hacker culture.
+Digital culture.
+Free culture.
 
-\end{abstract}
+\end{abstract2}

### 2012-11-18 pretextual.tex
arrumada geral, adentrando texto

@@
                     if sharing == 0: \; \; \; \; \;  \; \; \; \; \; \\
-                        Obatalá.kill( kitten ) \; \; \;}
+                        Obatalá.kill( kitten )}
         {Autoria coletiva e anônima (2010)}

### 2012-11-18 devEresultados.tex
finalizada toda intro

@@
 
-\section{Caracterização da nota musical discretizada}
+\section{Caracterização da nota musical discretizada}\label{sec:notaDisc}
 Em diversos contextos artísticos e teóricos, 

### 2012-11-18 introducao.tex
realocados pastas de musicas e arquivos
arrumada geral, adentrando texto
finalizada toda intro

--- 2012-11-10 introducao.tex
@@
 \label{cap:intro} 
+\epigraph{"Tradicionalmente a notação musical é vista como um código através do qual sons, ideias musicais ou indicações para execução musical são registrados sob forma escrita."}{Edson S.
+Zampronha.\cite{Zampronha} \\}
 
-O som pode ser definido como a propagação ondulatória e mecânica em um meio elástico
-ou como a excitação dos mecanismos auditivos que resultam em sua percepção~\cite{Everest}. No primeiro
-caso, estamos tratando de leis puramente físicas. No segundo caso, estamos lidando com um
-fenômeno psicofísico. Nesta última abordagem encontramos lugar para a música, que escapa,
-portanto, às leis fatídicas dos fenômenos naturais~\cite{Roederer}. A delicada tarefa de capturar estruturas
-e artifícios musicais diretamente relacionados às oscilações do som discretizado
-é exatamente a tarefa proposta por este trabalho. Para isso, entendemos cabível
-abordar o som em sua representação digital e dispor os resultados em fórmulas matemáticas e implementações computacionais.
+Representar estruturas e artifícios musicais através das características do som discretizado
+é exatamente a tarefa proposta por este trabalho.
+Os resultados estão em relações matemáticas e implementações computacionais.
+Uma descrição está no capítulo~\ref{cap:resultados} e o conjunto de \emph{scripts} disponibilizados no Apêndice~\ref{cap:codigoProc} e online.\cite{MASSA} O \emph{toolbox} recebeu o nome \massa\ (música e áudio em sequências e séries amostrais) e foi utilizado para fazer pequenas peças e montagens focadas nos princípios expostos.
+O Apêndice~\ref{cap:codigoPecas} possui uma relação destas montagens assim como o diretório \emph{exemplos\_de\_uso} da \massa.
 
-    \section{Do som ao áudio digital}
+    \section{Do som ao áudio digital}\label{sec:audio}
 
-Em sua conceituação puramente física, o som é uma onda mecânica longitudinal de pressão e pode se propagar por qualquer meio material.
-Embora o corpo humano possa captar sons cujas frequências estão fora da banda compreendida aproximadamente entre $20Hz$ e $20 kHz$, estas são as frequências chamadas audíveis e apreciadas pelo nosso aparelho auditivo.
- Considerada a velocidade do som no ar de $\approx 343.2\,m/s$ em condições usuais,
-estes limites correspondem respectivamente aos comprimentos de onda $\frac{343.2}{20} = 17.16\,m$ e $\frac{343.2}{20000}=17.16\,mm$.
+O som é uma onda mecânica longitudinal de pressão.
+A banda de frequências compreendida entre $20Hz$ e $20 kHz$ é apreciada pelo nosso aparelho auditivo e possui variações dependentes da pessoa, das condições climáticas e do som em si.
+ Considerada a velocidade do som no ar de $\approx 343.2\,m/s$,
+estes limites correspondem respectivamente aos comprimentos de onda $\frac{343.2}{20} = 17.16\,m$ e $\frac{343.2}{20000}=17.16\,mm$.\cite{Roederer}
 
-A efetiva audição humana destas vibrações consiste em um fenômeno complexo ainda sob intensa pesquisa, involvendo captações pelos ossos, estômago e orelha, além de funções de transferência da cabeça e dorso e processamento pelo sistema nervoso.
-Existe, no entanto, um órgão dedicado à tarefa de captura destas ondas, ao qual chamamos ouvido e cujo funcionamento decompõe o som em seu espectro senoidal antes de passar o estímulo para o sistema nervoso~\cite{Roederer}.
-Este papel central das componentes senoidais do som é crucial para os fenômenos musicais, o que pode ser notado tanto na composição de sons de interesse para a música, quanto nas afinações e escalas~\cite{floEsp}.
+A percepção humana do som envolve captações pelos ossos, estômago e orelha, funções de transferência da cabeça e dorso e processamento pelo sistema nervoso.
+Existe, além disso, um órgão dedicado à tarefa de captura destas ondas, ao qual chamamos ouvido e cujo funcionamento decompõe o som em seu espectro senoidal e passa para o sistema nervoso.\cite{Roederer} Estas componentes senoidais são cruciais para os fenômenos musicais, como pode-se observar na composição dos sons de interesse para a música quanto nas afinações e escalas.\cite{floEsp} A subseção~\ref{sec:notaDisc} expõe a presença de senóides no som discretizado e caracteriza a nota musical básica.
 
-A representação do som é chamada áudio e usualmente usamos este termo para designar uma representação elétrica do som\footnote{Embora haja esta tendência e tradição de uso, os termos
-som e áudio são, na verdade, usados de forma intercambiável.}.
-O áudio pode ser proveniente da captura do som por microfones ou de síntese diretamente.
-O áudio digital consiste na sua representação sonora através de protocolos também digitais.
-Alguns destes protocolos são bastante elaborados, em geral visando resultar em quantidades reduzidas de dados para facilitar armazenamento e transferência dos arquivos.
-Outras representações digitais do som são bastante diretas, consistindo em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
-Esta representação do som por amostras separadas por intervalos regulares $\lambda_a$ é a forma padrão de representação do som em tempo discreto e é chamada de modulação por código de pulsos (denotada pela sigla PCM do inglê Pulse Code Modulation).
-Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$ (também chamada de taxa de amostragem) e profundidade de bit, sendo este o número de bits utilizados para representar cada amostra.
-Dispomos na figura~\ref{fig:PCM} um esquema gráfico de um áudio PCM.
-Observe que os $2^4=16$ grados disponíveis para a amplitude de cada amostra junto ao espaçamento regular $\lambda_a$ introduzem um erro de quantização.
-Estaremos apontando as formas de lidar com o ruído causado por estes erros de quantização sempre que pertinente.
+A representação do som é chamada áudio\footnote{Os termos
+som e áudio são muitas vezes usados de forma intercambiável.~\cite{Everest}} e este pode ser proveniente da captura do som por microfones ou da síntese.
+Muitas vezes, o áudio digital é obtido através protocolos que facilitam o armazenamento e transferência dos arquivos.~\cite{procFala} Em outros momentos, a representação digital do som consiste em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
+Estas amostras separadas por intervalos regulares $\lambda_a$ constituem a forma padrão de representação do som em tempo discreto chamada de modulação por código de pulsos (PCM do inglês \emph{Pulse Code Modulation}).
+Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$, também chamada de taxa de amostragem, e a profundidade de bit que é o número de bits utilizados para representar a amplitude de cada amostra.\cite{protocolosAudio} A figura~\ref{fig:PCM} exibe $25$ amostras de um áudio PCM com $4$ bits cada.
+Os $2^4=16$ grados para a amplitude de cada amostra junto ao espaçamento regular $\lambda_a$ introduzem um erro de quantização.
+O ruído causado por estes erros diminuem junto à diminuição estes espaçamentos.
 
@@
 
-Nesta representação, é de conhecimento comum pelo teorema de Nyquist que não podemos representar frequências acima da metade da frequência de amostragem.
-Assim, se buscamos apreender as frequências audíveis, precisamos utilizar uma taxa de amostragem que seja ao menos o dobro da frequência mais alta capturada pelo sistema auditivo $f_a=2\times 20kHz=40kHz$.
-Este raciocínio está na base da utilização das frequências de amostragens $f_a=44.1kHz$ e $f_a=48kHz$, ambas utilizadas amplamente e são as taxas de amostragens padrão em \emph{Compact Disks} (CDs) e em sistemas de Rádio e TV, respectivamente.
+Pelo teorema de Nyquist, constata-se que a metade da frequência de amostragem é a frequência máxima do sinal.
+Assim, para apreender as frequências audíveis, é necessária uma taxa de amostragem que seja ao menos o dobro da frequência mais aguda $f_a \geq 2\times 20kHz=40kHz$.
+Este raciocínio está na base da utilização das frequências de amostragens $f_a=44.1kHz$ e $f_a=48kHz$, ambas padrão em \emph{Compact Disks} (CDs) e em sistemas de Rádio e TV, respectivamente.\cite{protocolosAudio}
 
-Do ponto de vista didático, é notável que praticamente todo processamento espectral sonoro seja praticável por somatórios quando tratamos do som discretizado.
-Estes mesmos procedimentos são obtidos por integrais quando estamos lidando com o som analógico, o que é consideravelmente mais complicado.
+    \section{Arte sonora e teoria musical}
 
-    \section{Arte sonora e rudimentos de teoria musical}
+A música é definida canonicamente como a arte manifesta pelos sons e silêncios.
+Para um ouvinte comum - e boa parte dos especialistas - uma 'música que seja música' pressupõe também a presença de uma métrica rítmica e organizações de alturas que formem melodias e harmonias como explicadas na sessão~\ref{notasMusica}.
+A música do século XX ampliou esta concepção tradicional de música.
+Isso ocorreu na música de concerto, especialmente nas correntes concreta, eletrônica e eletroacústica.
+Já na década de 90 era evidente que também a música popular, especialmente as músicas eletrônicas de dança, tinham se desdobrado para abarcar sons sem altura definida e organizações temporais fora de métricas simples.
+Mesmo assim, a nota permanece paradigmática como 'unidade fundamental' das estruturas musicais e, na prática, podem se desdobrar em sons que contemplam estes desenvolvimentos recentes.
+Adentramos a nota musical e sua expansão na sessão~\ref{varInternas}.\cite{Wisnick,Webern,Lerdahl,Cook}
 
-A música é comumente descrita como a arte manifesta pelos sons e silêncios.
-Embora esta definição esteja canonicamente correta, para as concepções usuais, ela está mais associada a uma ideia de arte sonora do que à música propriamente dita, pois para um ouvinte comum - e boa parte dos especialistas - uma 'música que seja música' pressupõe também a presença de uma métrica rítmica além de organizações de alturas que formem melodias e harmonias (veja sessão~\ref{notasMusica}).
-De qualquer forma, é importante reconhecer que a música do século XX ampliou esta concepção tradicional de música feita por notas que formam estruturas baseadas em ritmos, melodias e harmonias bem identificáveis.
-Isso ocorreu primeiramente na música de concerto, especialmente nas chamadas correntes eletrônica, concreta e eletroacústica.
-Já na década de 90 ficou claro que também a música popular, especialmente as músicas eletrônicas de dança, tinham por sua vez se desdobrado para abarcar não só os sons de altura definida e os organizações temporais bem estabelecidas dentro de métricas simples, mas também amálgamas sonoros ruidosos e disposições temporais fluídas e com sobreposições.
-Mesmo assim, a nota musical permanece paradigmática como 'unidade fundamental' das estruturas musicais.
-Além disso, as notas podem se desdobrar em sons que se desenvolvem no tempo, contemplando estes desenvolvimentos musicais recentes que foram supracitados (veja sessão~\ref{varInternas}).
+A teoria musical engloba assuntos tão diversos quanto psico-acústica, manifestações culturais e formalismos.
+O texto do capítulo~\ref{cap:resultados} aborda estes assuntos mediante necessidade e assinala complementos externos.\cite{Zamacois,Schoenberg,microsound}
 
-A teoria musical é uma ampla área de estudos com várias abordagens divergentes e que engloba assuntos tão diversos quanto psicoacústica, convenções culturais de organização de alturas (e.g.
-escalas e seus usos) e formas musicais tradicionais, como a forma sonata e rondós, que consistem em estruturas paradigmáticas para uma música inteira.
-Neste trabalho, a abordagem deste assunto é bastante vinculada às aplicações dispostas no capítulo seguinte e estão apontadas no próprio texto mediante necessidade.
-Assim, não consideramos apropriado discorrer sobre estes assuntos aqui de forma independente, e para isso dispomos os títulos referenciados em nossa bibliografia.
-Recomendamos também o acompanhamento de um especialista, pois, na música, é comum um mesmo termo ser utilizado para designar diferentes conceitos e um mesmo conceito ser vinculado a diferentes termos, mesmo em contextos escolásticos.
+    \section{Implementação computacional}
+Os resultados apresentados desta dissertação incluem \emph{scripts}, i.e.
+pequenos programas para melhor disponibilidade e validação das tecnologias, que constituem a caixa de ferramentas \massa.
+Liberamos este conjunto de códigos em domínio público através de repositórios Git públicos.\cite{gitBook}
 
-    \section{Disponibilização computacional}
-Os desenvolvimentos apresentados nesta dissertação incluem \emph{scripts} (i.e.
-pequenos programas) escritos em Python para disponibilização das tecnologias de forma explícita.
-Entendemos uma linguagem como o Python ou Scilab apropriada para esta tarefa
-por ser de compreensão bastante simples, aproximando-se do que chamamos de pseudo-código~\cite{tutPython}.
-Outra faceta importante desta disponibilização é a utilização de um sistema de controle de versão para os textos, códigos e figuras.
-Assim, pode-se observar o desenvolvimento do trabalho e executar modificações sem tantas preocupações, pois todas elas ficam registradas individualmente e são reversíveis.
-Utilizamos para isso o sistema Git, tida como a segunda grande contribuição de Linus Torvals, sendo o próprio kernel Linux a primeira.
-
-Os \emph{scripts} estão dispostos em duas versões: em Python puro, i.e.
-sem bibliotecas externas, e em Python com Numpy e Audiolab.
-Esta segunda versão executa as rotinas através de chamadas à linguagem Fortran, muito eficiente em termos de processamento computacional e, consequentemente, em termos da duração necessária para a execução dos algoritmos.
-
-Algum código já foi transcrito para JavaScript com muita facilidade, o que torna estas contribuições potencialmente utilizáveis em navegadores bastante difundidos, como o Firefox e o Chromium.
+Os \emph{scripts} estão em Python e fazem uso das bibliotecas externas Numpy e Scikits/Audiolab que realizam chamadas à linguagem Fortran para maior eficiência computacional.
+Parte deste código foi transcrito para JavaScript e Python nativos com facilidade, o que aponta para um uso destas contribuições em navegadores como o Firefox e o Chromium.\cite{numpy,audiolab,tutPython,python}
 
 Estas tecnologias são todas abertas, i.e.
-estão publicadas em licenças que permitem o uso, cópia, distribuição e utilização de quaisquer partes para geração de produtos derivados.
-Desta forma, os trabalhos aqui descritos estão disponíveis como parte de um repertório tecnológico livre que constitui um patrimônio da humanidade e é vinculado usualmente às comunidades de Software Livre e de Código Aberto\footnote{A comunidade e movimento chamada \emph{'Open Source'} é uma corrente que entende a publicação em licenças abertas de código computacional (e outras tecnologias) como sendo uma vantagem pragmática, possibilitando facilidades de desenvolvimento e vantagens de aprendizado e mercadológicas.
-A comunidade e movimento chamada \emph{'Free Software'} engloba este entendimento, mas adiciona a abordagem filosófica da liberdade e compartilhamento, e dá ênfase a isso.
-Ambas as correntes reforçam o entendimento de que o código computacional é o 'bem mais  precioso produzido atualmente' pois consiste em tecnologia condensada, reativa (pois executa, processa ou gera resultados), modular (pode ter partes copiadas reutilizadas eficientemente) e replicada sem custo adicional (pois a cópia de texto tem custo baixíssimo).
-Muito frequentemente isso tudo está disposto em poucas linhas escritas.}.
-Constatamos que, desta forma, facilitamos as colaborações em processos de co-autoria e geração de materiais subsequentes.
+estão publicadas em licenças que permitem o uso, cópia, distribuição e utilização de quaisquer partes para estudo e geração de produtos derivados.
+Desta forma, o trabalho aqui descrito está disponível como parte de um patrimônio tecnológico da humanidade\footnote{A comunidade e movimento chamada \emph{'Open Source'} entende a publicação de código computacional (e outras tecnologias) em licenças abertas como uma vantagem pragmática que facilita o desenvolvimento de \emph{software} e apresenta vantagens de pedagógicas e mercadológicas.
+A comunidade e movimento chamada \emph{'Free Software'} engloba este entendimento, mas adiciona a abordagem filosófica da liberdade e compartilhamento, dando ênfase a isso.
+Ambas as correntes reforçam o entendimento de que o código computacional é o 'bem mais precioso produzido atualmente' pois consiste em tecnologia condensada, reativa (pois executa, processa ou gera resultados), modular (partes são copiadas e reutilizadas eficientemente) e replicada sem custo adicional (pois a cópia de texto tem custo baixíssimo).\cite{Raymond,Lessig}} e facilita os processos de co-autoria.
 
@@
    \label{sec:objetivos}
-Nosso objetivo principal nesta dissertação é dispor um sistema unificado que relacione elementos básicos da música às sequências amostrais do som digital.
-Este conhecimento é existente, tanto que diversos softwares comerciais apresentam desenvolvimentos relacionados, mas não há uma apresentação unificada e concisa destas relações.
-Assim, dispomos um texto coerente em que os elementos musicais são apresentados junto às sequências temporais envolvidas.
-Além disso, como parte da efetiva contribuição, entregamos implementações em código computacional (Python) destas equações e \emph{scripts} que resultam em pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais.}.
+O objetivo principal desta dissertação é apresentar um sistema unificado que relacione elementos básicos da música às sequências amostrais do áudio PCM.
+Assim, o capítulo seguinte é um texto conciso em que os elementos musicais são apresentados junto às amostras temporais resultantes.
+Para validação e compartilhamento, as implementações em código computacional destas relações e de pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais.} foram reunidas em uma \emph{toolbox} chamada \massa\ e disponibilizadas online e nos Apêndices~\ref{cap:codigoProc} e~\ref{cap:codigoPecas}.
 
 Dada esta direção bem estabelecida, consideramos alguns objetivos secundários.
-Em primeiro lugar, a difusão de conhecimentos em programação e compreensão do código computacional é bastante potencializada através de práticas lúdicas, no caso a música.
+Em primeiro lugar, visamos a difusão da compreensão do código computacional através de práticas lúdicas, no caso a música.
 Outro objetivo considerado é a apresentação de um arcabouço de síntese sonora e musical com controle amostral, para o qual encontramos potenciais usos em experimentos psico-acústicos e síntese em alta definição (\emph{hi-fi}).
 Talvez por último, buscamos apresentar estes conteúdos de forma didática, quase um verdadeiro tutorial, possibilitando uma compreensão e uso facilitados, pois os assuntos tratados são de reconhecida complexidade: processamento de sinais, música, psico-acústica, para citar somente alguns exemplos.
-Deste ponto de vista didático, também reconhecemos importante a apresentação deste material na forma de hipertexto, em que cada \emph{script} e exemplo sonoro/musical seja acessível junto ao material teórico.
+Deste ponto de vista pedagógico, também reconhecemos importante a apresentação destes resultados na forma de hipertexto, em que cada \emph{script} e exemplo sonoro/musical seja acessível junto ao material teórico.

### 2012-11-19 devEresultados.tex
feita ateh localizacao espacial

--- 2012-11-15 devEresultados.tex
@@
 
-\emph{Simon Emmerson, Living electronic music, 2007}
+\emph{Simon Emmerson, Living electronic music.\cite{Emmerson}}
 \end{quotation}
 
-\section{Caracterização da nota musical discretizada}
+\section{Caracterização da nota musical em tempo discreto}\label{sec:notaDisc}
 Em diversos contextos artísticos e teóricos, 
@@
 unidades chamadas notas e 
-estas unidades compreendidas como "átomos" constituintes da música~\cite{Wisnick, Lovelock, Webern}.
+estas unidades compreendidas como "átomos" constituintes da música.\cite{Wisnick, Lovelock, Webern}
 Atualmente, estas notas
@@
 e, de um ponto de vista cognitivo, como discretizações
-que facilitam e enriquecem o fluxo de informação através da música
-~\cite{Roederer, Lacerda}.
-
-Em sua concepção
-simples e canônica, as notas possuem ao menos duração, volume, altura e timbre~\cite{Lacerda}. Embora estas
-sejam características apreciadas por leis psicofísicas, podemos lidar com elas através de suas características
-fisicamente determináveis~\cite{Roederer}.
-
-Com isso em mente, definiremos as características básicas da nota musical digital através da lida direta com a representação digital do som.
-Para nós, este consiste  em amostras sequenciais
-igualmente espaçadas no tempo. As amostras assim dispostas,
-relativas à onda mecânica do som que representam,
-em conjunto formam o chamado áudio PCM (veja sessão~\ref{sec:audio}).
-
+que facilitam e enriquecem o fluxo de informação através da música.\cite{Roederer, Lacerda}
+
+Canonicamente, as notas possuem ao menos duração, volume, altura e timbre.\cite{Lacerda} Estas
+são qualidades apreciadas por leis psicofísicas e pode-se lidar com elas através de caracterizações físicas.\cite{Roederer}
+Assim, as características da nota musical digital básica é dada pelas amostras igualmente espaçadas no tempo
+da onda mecânica que representam (veja sessão~\ref{sec:audio} sobre áudio PCM).
+
+Todas as relações desta seção estão no Apêndice~\ref{sec:cod1}, as montagens musicais \emph{Quadros sonoros} e \emph{Reduced-fi} estão nos Apêndices~\ref{ap:quadros} e~\ref{ap:reduced}.
+Estas implementações estão também disponíveis online como parte do \emph{toolbox} \massa.\cite{MASSA}
+
+\subsection{Duração}
  Definimos a
-frequência (ou taxa) de amostragem $f_a$ como o número de amostras por segundo do áudio digital e
+frequência (ou taxa) de amostragem $f_a$ como o número de amostras por segundo e
 a sequência $T_i=\{t_i\}$ como um conjunto ordenado de amostras reais separadas por $\delta_a=1/f_a$ segundos.
 
-\subsection{Duração}
-Desta forma, visando uma nota musical de duração $\Delta$,
-teremos uma sequência de $ \lfloor \Delta . f_a \rfloor $ amostras\footnote{O
+Assim, uma nota musical de duração $\Delta$
+se apresenta como uma sequência de $ \lfloor \Delta . f_a \rfloor $ amostras\footnote{O
 limite superior de uma sequência é um número natural mas $ \Delta . f_a $
 só satisfaz esta condição em casos muito excepcionais. É necessário
-escolher um inteiro próximo de $\Delta . f_a$ e admitir algum erro. Por simplicidade,
-escolheremos sempre a parte inteira da fração, descrita por $\lfloor \Delta . f_a \rfloor$.}:
+escolher um inteiro próximo de $\Delta . f_a$ e admitir algum erro. Por simplicidade, será considerada sempre a parte inteira da multiplicação, descrita por $\lfloor \Delta . f_a \rfloor$.}:
 
@@
 \subsection{Volume}
-O volume depende da reverberação e distribuição dos harmônicos - dentre outras medidas psicofísicas~\cite{Chowning} - mas num primeiro momento podemos obter resultados satisfatórios e de forma prática através da potência da onda:
+A sensação de volume sonoro depende da reverberação e distribuição dos harmônicos, dentre outras características trabalhamos na seção~\ref{varInternas}, e podemos obter variações do volume através da potência da onda~\cite{Chowning}:
 
@@
 
-Como o volume final dependerá da amplificação do sinal nos auto-falantes, o mais importante é a potência relativa de uma nota em relação às outras ou de um trecho da música em relação ao resto.
-As diferenças de volume são medidas em decibels, que são
-calculadas diretamente com as amplitudes através das energias ou potências\footnote{Lembrando que, devido à nossa percepção logarítmica,
+Como o volume final dependerá da amplificação do sinal nos auto-falantes, o crucial é a potência relativa de uma nota em relação às outras ou de um trecho da música em relação ao resto.
+As diferenças de volume são medidas em decibels, e estes são
+calculados diretamente com as amplitudes através das energias ou potências\footnote{Lembrando que, devido à nossa percepção logarítmica,
 em um som de volume $v$ a redução da potência $p$ para uma mesma fração $\nu . p $ 
@@
 
-A quantidade adimensional $V_{dB}$ é medida pela unidade decibel ($dB$). A cada 10 $dB$ atribuímos canonicamente
-a sensação de "volume dobrado". Importantes resultados de referência são os equivalentes em $dB$s de se dobrar
+A quantidade $V_{dB}$ possui a unidade decibel ($dB$). A cada 10 $dB$ se atribui canonicamente
+a sensação de "volume dobrado". Importantes resultados de referência são os equivalentes em $dB$ de se dobrar
 a amplitude ou a potência:
@@
 
-Outro valor de referência é a amplificação de amplitude
-necessária para que uma sequência tenha o volume dobrado ($10dB$s a mais):
+Outro valor de referência é o ganho de amplitude
+necessário para que uma sequência tenha o volume dobrado ($10dB$ a mais):
 
@@
 
-Ou seja, é necessário pouco mais que triplicar a amplitude para que tenhamos volume drobrado.
+Ou seja, é necessário pouco mais que triplicar a amplitude para um volume drobrado.
 
@@
 
-Neste ponto, correspondendo a uma partícula musical (uma nota), estabelecemos uma sequência $T_i$ cuja
-duração e intensidade conseguimos controlar\footnote{Basta alterarmos o tamanho da sequência para alterarmos
-a duração e multiplicarmos os elementos da sequência por algum valor para resultar em um ganho de intensidade e, portanto, de volume.}.
-
-A altura é especificada pela frequência fundamental. A frequência fundamental $f_0$ especifica uma periodicidade temporal
-com ciclo de duração $\delta_{f_0} = 1/f_0$. Esta duração multiplicada pela frequência de amostragem $f_a$ resulta no número de amostras
+Neste ponto, a partícula musical (nota) é uma sequência $T_i$ cuja duração e volume correspondem ao tamanho da sequência e amplitude de suas amostras.
+A altura é especificada pela frequência fundamental $f_0$ com ciclo de duração $\delta_{f_0} = 1/f_0$.
+Esta duração multiplicada pela frequência de amostragem $f_a$ resulta no número de amostras
 do ciclo $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$.
 
-Por motivos didáticos, escolhamos por equanto $f$ que divida $f_a$ de forma que $\lambda_f$ resulte inteiro.
-É evidente que sendo $T_i^f$ a sequência sonora de frequência fundamental $f$:
+Por motivos didáticos, seja $f_0$ tal que divida $f_a$ e $\lambda_{f_0}$ resulte inteiro.
+Se $T_i^f$ é uma sequência sonora de frequência fundamental $f$, resulta evidente que:
 
 \begin{equation}\label{periodicidade}
-     T^f_i=\{ t_i^f \}=\{ t^f_{i+\lambda_{f}}  \}= \{ t^f_{i+\frac{f_a}{f}} \}
-\end{equation}
-
-Na sessão seguinte estaremos sintetizando sons cuja frequência $f$ não divide $f_a$, por hora podemos aceitar esta restrição sem perda de generalidade do conteúdo.
+     T^f_i=\left\{ t_i^f \right\}=\left\{ t^f_{i+\lambda_{f}}  \right\}= \left\{ t^f_{i+\frac{f_a}{f}} \right\}
+\end{equation}
+
+Na seção seguinte serão contempladas frequências $f$ que não dividem $f_a$ e esta restrição não implica na perda de generalidade do conteúdo desta seção.
 
@@
 o timbre é multidimensional e dado pelo comportamento temporalmente dinâmico
-de energias em componentes espectrais tanto harmônicas quanto ruidosas.  Vale salientar que nem tudo
-o que se atribui ao timbre se acha manifesto em diferenças espectrais. Mesmo
-assim se fala que uma nota e um instrumento possuem um timbre.
+de energias em componentes espectrais tanto harmônicas quanto ruidosas.
 Além disso, a palavra \emph{timbre} é utilizada para designar coisas diferentes: uma mesma nota
-possui diferentes timbres, um mesmo instrumento possui diferentes timbres, dois instrumentos da mesma família possuem o mesmo timbre que o caracterizam mas possuem timbres diferentes porque são instrumentos diferentes.
-Por último, até
+possui diferentes timbres, um mesmo instrumento possui diferentes timbres, dois instrumentos da mesma família possuem o mesmo timbre que a caracteriza mas possuem timbres diferentes porque são instrumentos diferentes.
+  Vale salientar que nem tudo
+o que se atribui ao timbre se acha manifesto em diferenças espectrais e que até
 aspectos culturais ou circunstanciais alteram nossa percepção do timbre.
 }.
-Musicalmente, importa manter em mente que espectros sonoros com diferenças mínimas resultam em timbres com diferenças expressivamente cruciais e que, portanto, podemos produzir timbres diferentes através de espectros diferentes~\cite{Roederer}.
-
-O caso mais simples (e mais importante, como veremos no texto que segue) é o do espectro que consiste somente
-de sua própria fundamental $f$. Este é o caso da senoide, frequência em movimento oscilatório puro chamado
+Musicalmente, importa que espectros sonoros com diferenças mínimas resultam em timbres com diferenças expressivamente cruciais e que, portanto, pode-se produzir timbres diferentes através de espectros diferentes.\cite{Roederer}
+
+O caso mais simples (e mais importante, como mostra o texto que segue) é o do espectro que consiste somente
+em sua própria fundamental $f$. Este é o caso da senoide, frequência em movimento oscilatório puro chamado
 movimento harmônico simples. Seja $S_i^f$ uma sequência cujas amostras
@@
 
-Onde $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\delta_a} \;$ é o número de amostras do período\footnote{Neste ponto já se tem toda a base para música \emph{reduced-fi}.}.
+Onde $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\delta_a} \;$ é o número de amostras do período\footnote{Neste ponto já se tem toda a base para música \emph{Reduced-fi} do Apêndice~\ref{ap:reduced}.}.
 
 De forma semelhante, outras formas de onda são utilizadas na música por suas qualidades
-espectrais e simplicidade. Enquanto a senoide é um ponto isolado no espectro, estas formas
-de onda apresentam cadeias de componentes harmônicas e usos específicos.
-Além da senoide (eq.~\ref{senoide}), as formas de onda especificadas em~\ref{denteDeSerra},~\ref{triangular} e~\ref{quadrada} são dispostas graficamente na figura~\ref{fig:formasDeOnda}. 
-São as formas de onda artificiais tradicionalmente usadas na música para síntese e controle oscilatório de variáveis apresentam diversos usos também fora da música~\cite{Openheim}.
-
-A dente de serra dispõe todos os componentes da série
+espectrais e simplicidade. Enquanto a senoide é um ponto isolado no espectro, estas 
+ondas apresentam cadeias de componentes harmônicas.
+As formas de onda especificadas nas equações~\ref{senoide},~\ref{denteDeSerra},~\ref{triangular} e~\ref{quadrada} estão na figura~\ref{fig:formasDeOnda}.
+São as formas de onda artificiais tradicionalmente usadas na música para síntese e controle oscilatório de variáveis e apresentam diversos usos também fora da música.\cite{Openheim}
+
+A dente de serra apresenta todas as componentes da série
 harmônica com energia decrescente de $-6dB/oitava$. A sequência de amostras temporais pode ser descrita da seguinte forma:
 \begin{equation}\label{denteDeSerra}
-     D^f_i=\{ d^f_i \}=\Bigl\{ 2\frac{i\,\%\lambda_f}{\lambda_f} -1 \Bigr\}
+     D^f_i=\left\{ d^f_i \right\}=\left\{ 2\frac{i\,\%\lambda_f}{\lambda_f} -1 \right\}
 \end{equation}
@@
 \begin{equation}\label{triangular}
-     T^f_i=\{ t^f_i \}=\Bigl\{1- \Bigl| 2 - 4\frac{i\,\%\lambda_f}{\lambda_f} \Bigr| \Bigr\}
+     T^f_i=\left\{ t^f_i \right\}=\left\{1- \left| 2 - 4\frac{i\,\%\lambda_f}{\lambda_f} \right| \right\}
 \end{equation}
@@
 \begin{equation}\label{quadrada}
-     Q^f_i=\{ q^f_i \}= \left\{
+     Q^f_i=\left\{ q^f_i \right\}= \left\{
          \begin{array}{l l}
@@
 A dente de serra é um ponto de partida comum para a síntese subtrativa pois possui
-ambos os harmônicos pares e ímpares e em grande quantidade. Para fins musicais, estas formas de onda são bastante ricas em harmônicos agudos.
-Assim, para a maior parte dos casos, alguma filtragem atenuante
-é útil nos médios e agudos para que o som ganhe naturalidade e fique mais agradável.
+ambos os harmônicos pares e ímpares e em grande quantidade.
+Para fins musicais, estas formas de onda são excessivamente ricas em harmônicos agudos e uma filtragem atenuante nos médios e agudos é útil para que o som ganhe naturalidade e fique mais agradável.
 Os harmônicos relativamente atenuados da onda triangular
-a faz a mais funcional - dentre as citadas - para ser usada sem nenhum tratamento para síntese de notas musicais.
+a faz a mais funcional - dentre as citadas - para ser usada sem nenhum tratamento na síntese de notas musicais.
 
@@
 subtrativa que vise imitar um clarinete. Este instrumento também só apresenta os
-componentes ímpares do espectro harmônico e a onda quadrada pode convir com sua energia abundante nas altas frequências.
-
-\begin{figure}[h!]
-    \centering
+componentes ímpares do espectro harmônico e a onda quadrada convém com sua energia abundante nas altas frequências.
+
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=\textwidth]{figuras/formasDeOnda6}
     \caption{Formas de onda musicais básicas}
-        \includegraphics[width=\textwidth]{figuras/formasDeOnda6}
         \label{fig:formasDeOnda}
@@
 Se $t_a=44,1 kHz$, como no padrão PCM de Compact Disks, a onda possui frequência fundamental $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz $.
-Um lá\footnote{Um lá 4, logo acima do dó central, no segundo espaço do pentagrama na clave de sol comum.}, seja qual for a forma de onda dentre as artificiais apresentadas.
-
-Estas formas de onda possuem usos especiais na música e seus espectros estão dispostos na figura ~\ref{fig:espectroDeOndas}.
-É importante notar as componentes isoladas e exatamente harmônicas dos espectros,
-resultado de um período mantido fixo com exatidão.
+Um lá\footnote{Um lá 4, logo acima do dó central, no segundo espaço do pentagrama na clave de sol comum.}, independente da forma de onda.
+
+O espectro de cada forma de onda básica está na figura ~\ref{fig:espectroDeOndas}.
+As componentes isoladas e exatamente harmônicas dos espectros correspondem a um período rigorosamente fixo.
 A senoide consiste de um nódulo único no espectro, frequência pura.
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/espectroDeOndas6}
     \caption{Espectros das ondas sonoras musicais artificiais básicas}
-        \includegraphics[width=\textwidth]{figuras/espectroDeOndas6}
         \label{fig:espectroDeOndas}
@@
 O espectro harmônico é formado pelas frequências múltiplas da frequência fundamental $f_n=(n+1).f_0$.
-Como nossa percepção segue uma progressão geométrica de frequências, o espectro possui notas diferentes da frequência fundamental.
+Como a percepção humana de altura segue uma progressão geométrica de frequências, o espectro possui notas diferentes da frequência fundamental.
 Além disso, o número de harmônicos será limitado pela frequência máxima $f_a/2$ (pelo Teorema de Nyquist).
@@
 Musicalmente crucial aqui é internalizar que a presença de
-energia\footnote{A energia total equivale à soma dos quadrados das amplitudes
-(como as de pressao/voltagem no tempo), e.g. figura~\ref{fig:formasDeOnda}.
-As componentes espectrais - e.g. figuras~\ref{fig:espectroDeOndas} e ~\ref{fig:espectroOboe} -
-também podem ser elevadas ao quadrado para resultar em quantificação de energia e somam-se na energia total.
-As energias se equivalem pelo teorema de Parseval: $\frac{1}{\Lambda} .
-\sum_{k=0}^{\Lambda -1}c_k^2 = \sum_{i=0}^{\Lambda-1}t_i^2$.}
-em uma componente de frequência $f_n$ na decomposição por fourier 
-implica na presença de uma oscilação senoidal na constituição do som, puramente harmônica no som e naquela frequência $f_n$. Esta energia concentrada especificamente na frequência $f_n$ é separada
- pelo ouvido para adentrar em um nível cognitivo de processamento\footnote{Esta separação em frequência é realizada por diversas espécies através de mecanismos similares à cóclea humana~\cite{Roederer}.}.
-  As componentes senoidais são geralmente as principais responsáveis pela qualidade chamada timbre e, caso não apresentem proporções harmônicas (relações de pequenos números), o som é percebido ruidoso, i.e.
-não são notas com frequência fundamental estabelecida unívocamente como um lá 4 ($440 Hz$) de nossos exemplos.
-Além disso, nossa noção de altura absoluta em um complexo sonoro é baseada na semelhança do espectro com a série harmônica~\cite{Roederer}.
-
-No caso de uma forma de onda fixa (e de tamanho fixo), o espectro é sempre harmônico e estático.
-Fixada a frequência fundamental (inverso do comprimento da onda), cada forma de onda é composta de proporções específicas das componentes harmônicas e
+energia
+em uma componente de frequência $f_n$ significa
+ uma oscilação na constituição do som, puramente harmônica e naquela frequência $f_n$. Esta energia concentrada especificamente na frequência $f_n$ é separada
+ pelo ouvido para adentrar em um nível cognitivo de processamento\footnote{Esta separação em frequência é realizada por diversas espécies através de mecanismos similares à cóclea humana.\cite{Roederer}}.
+  As componentes senoidais são geralmente as principais responsáveis pela qualidade chamada timbre.
+Caso não se apresentem em proporções harmônicas (relações de pequenos números), o som é percebido como ruidoso ou dissonante e não com uma sonoridade de frequência fundamental estabelecida unívocamente.
+Além disso, a noção de altura absoluta em um complexo sonoro é baseada na semelhança do espectro com a série harmônica.\cite{Roederer}
+
+No caso de uma forma de onda fixa e de tamanho fixo, o espectro é sempre harmônico e estático. Cada forma de onda é composta de proporções específicas das componentes harmônicas e 
 quanto maior a curvatura do trecho na forma de onda, maior a contribuição do trecho para a
 concentração de energia nos harmônicos agudos.
-
-Podemos ver isso claramente em sons reais.
-A onda rotulada como 'som real amostrado' na figura ~\ref{fig:formasDeOnda} é um período extraído de um som real relativamente comportado.
-Ele possuí $\lambda_f=114$ amostras\footnote{Caso também utilizado diretamente em $f_a=44,1kHz$, é um pouco mais grave que os $441 Hz$ das formas de onda artificiais $\frac{44100}{114}=385,84Hz$.}.
+Pode-se constatar isso em sons reais.
+A onda rotulada como 'som real amostrado' na figura ~\ref{fig:formasDeOnda} é um período de $\lambda_f=114$ amostras extraído de um som real relativamente comportado.
 A onda de oboé foi amostrada de um lá 4 também em $44,1kHz$.
@@
 
-Note que, definindo $ R_i=\{ r_i \}_0^{\lambda_f-1}$ a sequência de amostras do som real da figura ~\ref{fig:formasDeOnda},
+Definindo $ R_i=\{ r_i \}_0^{\lambda_f-1}$ a sequência de amostras do som real da figura ~\ref{fig:formasDeOnda},
 $R_i$ pode ser tomado como base para um som $T_i^f$ da seguinte forma:
@@
 seu espectro é perfeitamente harmônico, sem os ruídos e variações típicas do fenômeno natural. Isso pode ser 
-visto claramente na figura \ref{fig:espectroOboe} onde os espectros da nota original do oboé e de uma nota 
-artificial - de mesma duração e cujas amostras consistem no mesmo período da figura ~\ref{fig:formasDeOnda} - estão dispostas juntamente.
-O espectro natural possui variações na frequência dos harmônicos, nas suas intensidades e uma quantidade de ruído.
+visto na figura~\ref{fig:espectroOboe} onde dispomos os espectros da nota original do oboé e de uma nota 
+artificial de mesma duração e cujas amostras consistem no mesmo período da figura~\ref{fig:formasDeOnda}.
+O espectro natural possui variações nas frequências dos harmônicos, nas suas intensidades e uma quantidade de ruído.
 Já a nota cujo período foi amostrado possui espectro perfeitamente harmônico.
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/espectroOboeAmostradoNatural3}
     \caption{Espectros das ondas sonoras do oboé natural e de período amostrado}
-        \includegraphics[width=\textwidth]{figuras/espectroOboeAmostradoNatural3}
         \label{fig:espectroOboe}
@@
 \subsection{O espectro no som amostrado}
-Além deste papel-chave na cognição (especialmente no timbre), a presença e comportamento destas componentes senoidais 
+A presença e comportamento destas componentes senoidais 
 no som discretizado possui particularidades.
 Considere um sinal $T_i$ e sua decomposição de Fourier $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$.
-Sabemos que a recomposição segue a conversão das componentes frequenciais em amostras temporais\footnote{Lembrando que o fator $\frac{1}{\Lambda}$ pode ser distribuído dentre a transformada e a reconstrução como preferir}:
+A recomposição é a soma das componentes frequenciais em amostras temporais\footnote{Lembrando que o fator $\frac{1}{\Lambda}$ pode ser distribuído dentre a transformada e a reconstrução como preferir.}:
 
@@
 Onde $c_k = a_k + j .
-b_k$ pondera a amplitude e fase de cada frequência: $w_k=\frac{2\pi k}{\Lambda}$ em radianos ou $f_k=w_k\frac{f_a}{2\pi}=\frac{k.f_a}{\Lambda}$ em Herz (com atenção os respectivos limites em $\pi$ e em $\frac{f_a}{2}$).
-Nosso som possui amostras $t_i$ reais que resultam das contribuições de cada componente de frequência $w_k=\frac{2\pi k}{\Lambda}$ e cujo $c_k$ regula o módulo e a fase.
-A parte real da equação ~\ref{recomposicaoFourier} nos fornece a componente de forma mais clara:
+b_k$ dita a amplitude e fase de cada frequência: $w_k=\frac{2\pi}{\Lambda}k$ em radianos ou $f_k=w_k\frac{f_a}{2\pi}=\frac{f_a}{\Lambda}k$ em Herz,com atenção para os respectivos limites em $\pi$ e em $\frac{f_a}{2}$ dados pelo Teorema de Nyquist.
+
+No caso específico de um sinal sonoro, as amostras $t_i$ são reais e dadas pela parte real da equação ~\ref{recomposicaoFourier}:
 
@@
 
-Podemos notar claramente pela equação ~\ref{moduloEfase} que o termo imaginário acrescenta uma fase à senoide real.
-Os termos imaginários $b_k$ da decomposição espectral por Fourier proporcionam a varredura de fase
- $[-\frac{\pi}{2},+\frac{\pi}{2}]$, basta observar o termo $tg^{-1}\left (\frac{b_k}{a_k}\right )$ que possui este contra-domínio.
-O sinal de $a_k$ especifica se estamos do lado direito ou esquerdo do circulo trigonométrico, completando a varredura completa de fase (os intervalos $[-\frac{\pi}{2},+\frac{\pi}{2}]$ e $[\frac{\pi}{2},\frac{3\pi}{2}]$ se justapõem em $2\pi$).
-
-\begin{figure}[h!]
-    \centering
+A equação ~\ref{moduloEfase} mostra que o termo imaginário de $c_k$ acrescenta uma fase à senoide real, i.e.
+os termos imaginários $b_k$ da decomposição espectral por Fourier proporcionam a varredura de fase
+ $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right]$ dada pelo termo $tg^{-1}\left(\frac{b_k}{a_k}\right)$ que possui esta imagem.
+O sinal de $a_k$ especifica se o lado direito ou esquerdo do circulo trigonométrico,  o que completa a varredura de fase: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.
+
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=\textwidth]{figuras/amostras2c__}
     \caption{Oscilação de 2 amostras (frequência máxima em qualquer $f_a$)}
-        \includegraphics[width=\textwidth]{figuras/amostras2c__}
         \label{fig:amostras2}
@@
 
-A figura ~\ref{fig:amostras2} exibe duas amostras e a componente espectral que contém.
-A decomposição de Fourier resulta neste caso em somente dois coeficientes $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relativos à frequências $\{f_k\}_0^1=\{w_k\frac{f_a}{2\pi}\}_0^1=\{k\frac{f_a}{\Lambda=2}\}_0^1=\{0,\frac{f_a}{2}=f_{\text{máx}}\}$
-com energias $e_k=\frac{(c_k)^2}{\Lambda=2}$. O papel das amplitudes $a_k$ fica bem claro:
- $\frac{a_0}{2}$ é o deslocamento fixo\footnote{Chamado em vários contextos de \emph{bias} ou \emph{offset}.} e $\frac{a_1}{2}$ especifica a amplitude da oscilação em si, dada pela relação $f_k=k .
-\frac{f_a}{\Lambda=2}$.
+A figura ~\ref{fig:amostras2} exibe duas amostras e as componentes espectrais que contém.
+A decomposição de Fourier resulta neste caso em um único par de coeficientes $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relativos às frequências $\{f_k\}_0^1=\left\{w_k\frac{f_a}{2\pi}\right\}_0^1=\left\{k\frac{f_a}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_a}{2}=f_{\text{máx}}\right\}$
+com energias $e_k=\frac{(c_k)^2}{\Lambda=2}$. O papel das amplitudes $a_k$ é nítido com
+ $\frac{a_0}{2}$ o deslocamento fixo\footnote{Chamado de \emph{bias} ou \emph{offset}.} e $\frac{a_1}{2}$ a amplitude da oscilação em si, dada pela relação $f_k=k \frac{f_a}{\Lambda=2}$.
 
@@
 
-Vejamos o que acontece no caso de 3 amostras. Todas as sequências fixas $T_i$ de apenas 3 amostras também apresentam
-somente 1 frequência, pois sua primeira harmônica usaria 1,5 amostras e ultrapassa o limite inferior de 2 amostras mínimas (a frequência da harmônica ultrapassaria a de Nyquist pois:  $\; \frac{2.
-f_a}{3} > \frac{f_a}{2} $).
-Se $\Lambda=3$,
-os coeficiêntes $\{c_k\}_0^{\Lambda-1=2}$ da decomposição por Fourier apresentam-se em 
+Todas as sequências fixas $T_i$ de apenas $3$ amostras também apresentam
+somente $1$ frequência, pois sua primeira harmônica usaria $1.5$ amostras e ultrapassa o limite inferior de 2 amostras mínimas, i.e.
+a frequência da harmônica ultrapassaria a de Nyquist pois:  $\; \frac{2.
+f_a}{3} > \frac{f_a}{2} $.
+Os coeficiêntes $\{c_k\}_0^{\Lambda-1=2}$ apresentam-se em 
 3 componentes frequenciais. Uma delas é relativa à frequência zero ($c_0$), as outras duas ($c_1$ e $c_2$) contribuem de forma igual na reconstrução da senoide com $f=f_a/3$.
@@
     \centering
-    \caption{3 amostras apresentam uma única frequência}
         \includegraphics[width=\textwidth]{figuras/amostras3b}
+    \caption{3 amostras apresentam fixas apresentam uma só frequência não nula. $c_1=c_2^*$ e $w_1 \equiv w_2$}
         \label{fig:amostras3}
@@
 
-Sempre que partimos de $\Lambda$ amostras reais $t_i$ resultamos em $\Lambda$ coeficientes complexos $c_k=a_k+j.b_k$.
-Os coeficientes $c_k$ se equivalem dois a dois\footnote{Parte real igual e imaginária com sinal trocado: $a_{k1}=a_{k2}$ e $b_{k1}=-b_{k2}$.
-Como consequência temos módulos iguais e fases com sinais opostos} correspondendo às frequências $f_k = k\frac{f_a}{\Lambda}, \; k \in \{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \} $.
-Quando $k> \frac{\Lambda}{2} $
+Sempre que partimos de $\Lambda$ amostras reais $t_i$ resultamos em $\Lambda$ coeficientes complexos $c_k=a_k+j.b_k$, os coeficientes $c_k$ se equivalem dois a dois correspondendo às mesmas frequências e com contribuições idênticas\footnote{Parte real igual e imaginária com sinal trocado: $a_{k1}=a_{k2}$ e $b_{k1}=-b_{k2}$.
+Como consequência temos módulos iguais e fases com sinais opostos}.
+Lembrando que $f_k = k\frac{f_a}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $, pode-se anotar que, quando $k> \frac{\Lambda}{2} $,
 a frequência $f_k$ é espelhada em $\frac{f_a}{2}$ da seguinte forma $f_k=\frac{f_a}{2} - (f_k-\frac{f_a}{2})=f_a-f_k=f_a - k\frac{f_a}{\Lambda}=(\Lambda-k)\frac{f_a}{\Lambda} \;\;\;\; \Rightarrow \;\;\;\; f_k\equiv f_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
@@
    são equivalentes em pares.
-   Excepcional excessão para $f_0$ e, no caso de $\Lambda$ ser par, de $f_{\Lambda/2}=f_{\text{máx}}=\frac{f_a}{2}$ , ambas as componentes são isoladas, i.e.
+   Excepcional exceção para $f_0$ e, no caso de $\Lambda$ ser par, de $f_{\Lambda/2}=f_{\text{máx}}=\frac{f_a}{2}$ , ambas as componentes são isoladas, i.e.
 não existe outra componente na frequência $f_0$ ou $f_{\Lambda/2}$ (se $\Lambda$ par) além dela mesma.
@@
 
-Além disso, estas duas frequências (a frequência zero e a frequência máxima) não são representadas com variação de fase e, portanto, são estritamente reais. Assim, podemos 
+Além disso, estas duas frequências (a frequência zero e a frequência máxima) não são representadas com variação de fase e, portanto, são estritamente reais. Assim, pode-se
    concluir que o número $\tau$ de pares de coeficientes equivalentes é:
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/amostras4__}
     \caption{Componentes frequenciais em 4 amostras}
-        \includegraphics[width=\textwidth]{figuras/amostras4__}
         \label{fig:amostras4}
@@
 Com $k \in \mathbb{N}$.
+
  A observação da equação de reconstrução para o sinal real ~\ref{moduloEfase} em conjunto com as equivalências dos módulos e fases ~\ref{equivalenciasModulos} e ~\ref{equivalenciasFases}, o número de coeficientes pareados \ref{coefsPareados} e equivalência de pares de frequências \ref{equivalenciasFreqs}
-expõe o caso geral da combinação em fase das componentes em cada amostra $t_i$:
+expõe o caso geral da combinação das componentes em cada amostra $t_i$:
 
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/amostras4formas__}
     \caption{Formas de onda básicas em 4 amostras}
-        \includegraphics[width=\textwidth]{figuras/amostras4formas__}
         \label{fig:formas4}
@@
 
-Assim, a exemplo da figura ~\ref{fig:amostras3}, a transformada de Fourier de 3 amostras possui 2 coeficientes frequenciais que possuem quantidades iguais de energia na mesma frequência.
-
-Com 4 amostras, podemos representar 1 ou 2 frequências em proporções diferentes. A figura ~\ref{fig:amostras4} mostra uma 
+Assim, a exemplo da figura ~\ref{fig:amostras3}, a transformada de Fourier de 3 amostras possui 2 coeficientes frequenciais com quantidades iguais de energia na mesma frequência.
+
+Com 4 amostras, pode-se representar 1 ou 2 frequências em proporções quaisquer. A figura ~\ref{fig:amostras4} mostra uma 
 forma de onda de 4 amostras e suas duas componentes. 
-Note que as contribuições individuais se somam de fato na forma de onda 
-original, e que as curvaturas maiores são fruto da frequência mais aguda
+As contribuições individuais se somam na forma de onda 
+original, e uma breve inspeção revela que as curvaturas maiores são fruto da frequência mais aguda
 enquanto um deslocamento fixo da somatória das componentes é fruto
 da componente na frequência zero.
-
-A figura ~\ref{fig:formas4} explicita os harmônicos em 4 amostras nas formas de onda básicas das equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} e figura ~\ref{fig:formasDeOnda}.
-Todas consistem em apenas 1 senoide, com excessão da dente de serra que possui os harmônicos pares.
-
-A figura ~\ref{fig:amostras6} mostra uma decomposição senoidal para o caso de 6 amostras e a figura ~\ref{fig:formas6} decompõe as formas de onda básicas para o caso de 6 amostras.
- Note que neste caso de 6 amostras todas as ondas se diferenciam no espectro: as quadrada e triangular possuem as mesmas componentes, mas em proporções diferentes, já a dente de serra possui uma componente a mais.
-
-\begin{figure}[h!]
-    \centering
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=\textwidth]{figuras/amostras6}
     \caption{Componentes frequenciais em 6 amostras}
-        \includegraphics[width=\textwidth]{figuras/amostras6}
         \label{fig:amostras6}
@@
 
-\begin{figure}[h!]
-    \centering
+A figura ~\ref{fig:formas4} explicita os harmônicos em 4 amostras nas formas de onda básicas das equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada}.
+Todas consistem em apenas 1 senoide, com exceção da dente de serra que possui os harmônicos pares.
+
+A figura ~\ref{fig:amostras6} mostra uma decomposição senoidal para o caso de 6 amostras e a figura ~\ref{fig:formas6} decompõe as formas de onda básicas.
+ Neste caso todas as ondas se diferenciam no espectro: as quadrada e triangular possuem as mesmas componentes, mas em proporções diferentes, já a dente de serra possui uma componente a mais.
+
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=\textwidth]{figuras/amostras6formas___}
     \caption{Formas de onda básicas em 6 amostras}
-        \includegraphics[width=\textwidth]{figuras/amostras6formas___}
         \label{fig:formas6}
@@
 a localização espacial é também importante para um som
-pois este comumente é emitido por uma fonte localizada, como ocorre com um instrumento musical ou uma pessoa. Além disso, a espacialização é atributo bastante valorizado
+pois este comumente é emitido por uma fonte localizada, como um instrumento musical ou uma pessoa. Além disso, a espacialização é atributo bastante valorizado
  por audiófilos e pela indústria fonográfica~\cite{floEsp}.
@@
     \centering
+        \includegraphics[width=.5\textwidth]{figuras/espacializacao___}
     \caption{Detecção de localização espacial de fonte sonora: DTI e DII}
-        \includegraphics[width=.5\textwidth]{figuras/espacializacao___}
         \label{fig:spac}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/mixagem}
     \caption{Mixagem de três sequências sonoras}
-        \includegraphics[width=\textwidth]{figuras/mixagem}
         \label{fig:mixagem}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/concatenacao}
     \caption{Concatenação de três sequências sonoras}
-        \includegraphics[width=\textwidth]{figuras/concatenacao}
         \label{fig:concatenacao}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/lut}
     \caption{Lookup Table}
-        \includegraphics[width=\textwidth]{figuras/lut}
         \label{fig:lut}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/transicao}
     \caption{Transições de intensidade para diferentes valores de $\alpha$ (veja equações~\ref{seqAmp} e ~\ref{transAmp})}
-        \includegraphics[width=\textwidth]{figuras/transicao}
         \label{fig:transicao}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/convolucao______}
     \caption{Interpretação gráfica da convolução}
-        \includegraphics[width=\textwidth]{figuras/convolucao______}
         \label{fig:conv}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/delays__}
     \caption{Convolução com o impulso e linhas de delays}
-        \includegraphics[width=\textwidth]{figuras/delays__}
         \label{fig:delays}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/iir___}
     \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda}}
-        \includegraphics[width=\textwidth]{figuras/iir___}
         \label{fig:iir}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/ruidos___}
     \caption{Ruídos coloridos realizados através das equações~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul}, \ref{eq:violeta}: espectros e ondas sonoras resultantes}
-        \includegraphics[width=\textwidth]{figuras/ruidos___}
         \label{fig:ruidos}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/vibrato___}
     \caption{Espectrograma de um som com vibrato senoidal de $3Hz$ e profundidade de uma oitava em uma dente de serra de $1000Hz$ (considerada $f_a=44.1kHz$)}
-        \includegraphics[width=\textwidth]{figuras/vibrato___}
         \label{fig:vibrato}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/tremolo}
     \caption{Tremolo de profundidade $V_{dB}=12dB$ com padrão oscilatório de uma dente de serra em $f'=1.5Hz$ em uma senoide de $f=40Hz$ (considerada taxa de amostragem $f_a=44,1kHz$)}
-        \includegraphics[width=\textwidth]{figuras/tremolo}
         \label{fig:tremolo}
@@
 de forma absoluta ou tratadas ao longo de sua duração,
-com a única excessão da duração em si.
+com a única exceção da duração em si.
 
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/adsr}
     \caption{Envoltória ADSR canônica e uma sequência sonora arbitrária sendo submetida à envoltória}
-        \includegraphics[width=\textwidth]{figuras/adsr}
+\end{figure}
         \label{fig:adsr}
-\end{figure}
 
@@
 Observe que nas escalas diatônicas só estão presentes intervalos maiores, menores e justos dentre quaisquer duas notas escohidas.
-Única excessão para o trítono, que se apresenta como quarta aumenta ou quinta diminuta, dependendo da nota mais grave que se leva em consideração.
+Única exceção para o trítono, que se apresenta como quarta aumenta ou quinta diminuta, dependendo da nota mais grave que se leva em consideração.
 
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/movContraponto}
     \caption{Movimentos diferenciados pelo contraponto}
-        \includegraphics[width=\textwidth]{figuras/movContraponto}
         \label{fig:movContraponto}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/metricaMusical}
     \caption{Divisões e aglomerações do pulso musical para estabelecimento de métrica.
@@
 Ao lado direito temos exemplos de fórmulas de compasso que especificam as mesmas métricas, mas na escala produzida pelas ocorrências e aglomerações do pulso musical.}
-        \includegraphics[width=\textwidth]{figuras/metricaMusical}
         \label{fig:pulsoSubAgl}
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/climax}
     \caption{Distinções canônicas do clímax musical em uma melodia ou arcos em outras escalas}
-        \includegraphics[width=\textwidth]{figuras/climax}
         \label{fig:climax}

### 2012-11-19 devEresultados.tex
terminada revisao completa da parte 2.1, adicionando scripts e dados nas tabelas

@@
         \includegraphics[width=\textwidth]{figuras/amostras3b}
-    \caption{3 amostras apresentam fixas apresentam uma só frequência não nula. $c_1=c_2^*$ e $w_1 \equiv w_2$}
+    \caption{3 amostras fixas apresentam uma só frequência não nula. $c_1=c_2^*$ e $w_1 \equiv w_2$}
         \label{fig:amostras3}
@@
 
+\subsection{A nota básica}\label{notaBasica}
+
+Seja $f$ tal que $f$ divida $f_a$\footnote{Como apontado anteriormente, esta limitação facilita a exposição sem perda de generalidade.
+Esta limitação será superada no início da próxima seção.}.
+Uma sequência $T_i$ de amostras sonoras separadas por $\delta_a=1/f_a$ descreve uma nota musical de frequência $f$ Herz e duração $\Delta$ segundos se, e somente se, possuir a periodicidade $\lambda_f=f_a/f$
+ e tamanho $\Lambda=\lfloor f_a . \Delta \rfloor $:
+
+\begin{equation}
+T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{\Lambda-1}= \left \{t_{i \; \% \left( \frac{f_a}{f} \right) } \right \}_0^{\Lambda-1}
+\end{equation}
+
+Note que a nota por si só não especifica um timbre.
+Mesmo assim, faz-se necessária a escolha de uma forma de onda para que as amostras $t_i$ tenham um valor estabelecido individualmente.
+Um único período dentre as ondas básicas pode ser utilizado para a especificação da nota da seguinte forma:
+
+$\lambda_f=\frac{f_a}{f}$ é o número de amostras do período. Seja $L_i^{f,\, \delta_f} $
+a sequência que descreve um período da onda $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ de duração 
+$\delta_f=1/f$, dadas pelas equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} e onde $R_i^f$ 
+uma onda real amostrada:
+
+\begin{equation}\label{periodoUnico}
+L_i^{f , \delta_f } = \left\{ l_i^f \right\}_0^{\delta_f . f_a -1}=\left\{ l_i^f \right\}_0^{\lambda_f-1}
+\end{equation}
+
+Então a sequência $T_i$ consistirá em uma nota de duração $\Delta$ e frequência $f$ se:
+
+\begin{equation}
+T_i^{f,\; \Delta}=\left\{t_i^f\right\}_0^{\lfloor f_a . \Delta \rfloor -1}=\left \{ l^f_{i\,\%\left(\frac{f_a}{f}\right)} \right \}_0^{\Lambda-1}
+\end{equation}
+
 \subsection{Localização espacial}
-Embora não seja uma das quatro qualidades básicas tradicionalmente elegidas para caracterizar uma nota musical,
-a localização espacial é também importante para um som
-pois este comumente é emitido por uma fonte localizada, como um instrumento musical ou uma pessoa. Além disso, a espacialização é atributo bastante valorizado
- por audiófilos e pela indústria fonográfica~\cite{floEsp}.
-Segundo a compreensão atual do fenômeno de percepção da localização espacial do som, nosso sistema nervoso utiliza
+Embora não seja uma das quatro qualidades básicas tradicionais de uma nota musical, esta possui sempre uma localização espacial. Além disso, a espacialização é atributo bastante valorizado
+ por audiófilos e pela indústria fonográfica.\cite{floEsp}
+Acredita-se que a percepção da localização espacial do som se dê em nosso sistema nervoso através destas
 três informações: o atraso de chegada do som entre um ouvido e o outro, a diferença de intensidade do som direto em cada ouvido e a 
-filtragem realizada pelo corpo, incluindo toráx, cabeça e orelhas~\cite{Roederer, hrtf, Heeger}.
+filtragem realizada pelo corpo, incluindo toráx, cabeça e orelhas.\cite{Roederer, hrtf, Heeger}
 
@@
 
-	Se consideradas somente as incidências diretas em cada ouvido, podemos dispor as diferenças de tempo e intensidade em equações simples. Dada a distância entre os ouvidos $\zeta\approx 21,5cm$,
+	Se consideradas somente as incidências diretas em cada ouvido, as equações resultam simples.
+Dada a separação entre os ouvidos\footnote{Constata-se que $\zeta \approx 21,5cm$ para um humano adulto.} $\zeta$,
 um objeto localizado em $(x,y)$ conforme a figura~\ref{fig:spac}
@@
 
-e cálculos imediatos nos levam à DTI (Diferença de Tempo Interaural)\footnote{Constata-se que $\zeta \approx 21,5cm$ para um humano adulto.}:
+e cálculos imediatos resultam na Diferença de Tempo Interaural:
 
@@
 
-e à DII (Diferença de Intensidade Interaural):
+e na Diferença de Intensidade Interaural:
 \begin{equation}
@@
 Convertendo para amplitude temos $DII_a=\frac{d}{d'}$. A $DII_a$ pode
-ser utilizada como constante multiplicativa do canal direito de um sinal sonoro estéreo ($\{t_i'\}_0^{\Lambda -1}=\{DII_a .
-t_i\}_0^{\Lambda -1}$).
-Pode-se utilizar a DII junto à DTI como adiantamento no tempo do canal direito com relação ao esquerdo, vínculo crucial para a percepção de localidade em sons graves e em sonoridades percussivas~\cite{Heeger}.
-Considerando $\Lambda_{DTI}=\lfloor DTI . f_a \rfloor $, podemos escrever:
+ser utilizada como constante multiplicativa do canal direito de um sinal sonoro estéreo: $\{t_i'\}_0^{\Lambda -1}=\{DII_a .
+t_i\}_0^{\Lambda -1}$.
+Pode-se utilizar a DII junto à DTI como adiantamento no tempo do canal direito com relação ao esquerdo, vínculo crucial para a localização em sons graves e em sonoridades percussivas.\cite{Heeger}
+Considerando $\Lambda_{DTI}=\lfloor DTI . f_a \rfloor$, pode-se escrever:
 
@@
 DII_a & = \frac{d}{d'} \\
-\{t_{(i+\Lambda_{DTI})}'\}_{\Lambda_{DTI}}^{\Lambda+\Lambda_{DTI}-1} & =\{DII_a . t_i\}_0^{\Lambda-1} \\
-\{t_i'\}_0^{\Lambda_{DTI}-1} & = 0
+\left\{t_{(i+\Lambda_{DTI})}'\right\}_{\Lambda_{DTI}}^{\Lambda+\Lambda_{DTI}-1} & =\left\{DII_a . t_i\right\}_0^{\Lambda-1} \\
+\left\{t_i'\right\}_0^{\Lambda_{DTI}-1} & = 0
 \end{split}
@@
 
-Embora consideravelmente simples até aqui, a localização espacial depende drasticamente de outras pistas. De forma canonica, resolvemos pelas
-DTI e DII somente o ângulo horizontal (azimutal) $\theta$ dado por:
+Embora consideravelmente simples até aqui, a localização espacial depende drasticamente de outras pistas. De forma canonica, pela
+DTI e DII especifica-se somente o ângulo horizontal (azimutal) $\theta$ dado por:
 
@@
 do cone.
-Nestes pontos, a inferência do ângulo azimutal depende especialmente da filtragem atenuante nos agudos, pois a cabeça interfere um tanto mais nas ondas mecânicas agudas do que nas graves~\cite{Heeger,hrtf}.
+Nestes pontos, a inferência do ângulo azimutal depende especialmente da filtragem atenuante nos agudos, pois a cabeça interfere um tanto mais nas ondas mecânicas agudas do que nas graves.\cite{Heeger,hrtf}
 
 A figura~\ref{fig:spac} mostra esta sombra acústica do crânio, importante para a percepção do ângulo azimutal da fonte no cone de confusão.
-O cone em si não foi disposto na figura pois não é exatamente um cone e suas dimensões precisas não são especificadas na literatura~\cite{confCone}, mas pode ser entendido como um cone com o ápice no meio da cabeca e saindo por cada uma das orelhas.
-
-Já a localização completa, incluindo distância e elevação da fonte sonora, é dada pela função de transferência de cabeça (HRTF - do inglês \emph{Head Related Transfer Function}~\cite{hrtf}.
-Exitem bases abertas e conhecidas de HRTF como \emph{CIPIC}~\cite{CIPIC} e podemos aplicar estas funções de transferência em um som por convolução (veja equação~\ref{eq:conv}).
-Estas funções de transferência mudam de acordo com o corpo do indivíduo e existem diversas técnicas para resultar em funções de transferência que sejam utilizáveis de forma mais universal~\cite{lazaSPA}.
-
-\subsection{A nota básica}\label{notaBasica}
-
-Escolhamos $f$ tal que $f$ divida $f_a$  
- \footnote{Como apontado anteriormente, esta escolha facilita as demonstrações.
-Transporemos esta limitação na próxima sessão.}. 
-Uma sequência $T_i$ de amostras sonoras separadas por $\delta_a=1/f_a$ descreve uma nota musical de frequência $f$ Herz e duração $\Delta$ segundos se, e somente se, possuir a periodicidade $\lambda_f=f_a/f$
- e tamanho $\Lambda=\lfloor f_a . \Delta \rfloor $:
-
-\begin{equation}
-T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{\Lambda-1}= \left \{t_{i \; \% \left( \frac{f_a}{f} \right) } \right \}_0^{\Lambda-1}
-\end{equation}
-
-Note que a nota por si só não especifica um timbre.
-Mesmo assim, faz-se necessária a escolha de uma forma de onda para que as amostras $t_i$ tenham um valor estabelecido individualmente.
-Um único período dentre as ondas básicas pode ser utilizado para a especificação da nota da seguinte forma:
-
-seja $f$ a frequência da nota e escolhamos $f$ tal que $f$ divida $f_a$ (transporemos esta limitação na sessão seguinte).
-Assim, o inteiro $\lambda_f=\frac{f_a}{f}$ é o número de amostras do período.
-Temos que $L_i^{f,\, \delta_f} $ é
-a sequência que descreve um período da onda $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ de duração 
-$\delta_f=1/f$ (ver equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada}; $R_i^f$ 
-é uma onda real amostrada):
-
-\begin{equation}\label{periodoUnico}
-L_i^{f , \delta_f } = \{ l_i^f \}_0^{\delta_f . f_a -1}=\{ l_i^f \}_0^{\lambda_f-1}
-\end{equation}
-
-Então a sequência $T_i$ consistirá em uma nota de duração $\Delta$ e frequência $f$ se:
-
-\begin{equation}
-T_i^{f,\; \Delta}=\{t_i^f\}_0^{\lfloor f_a . \Delta \rfloor -1}=\left \{ l^f_{i\,\%\left(\frac{f_a}{f}\right)} \right \}_0^{\Lambda-1}
-\end{equation}
+O cone em si não foi disposto na figura pois não é exatamente um cone e suas dimensões precisas não são especificadas na literatura, mas pode ser entendido como um cone com o ápice no meio da cabeca e saindo por cada uma das orelhas.\cite{confCone}
+
+Já a localização completa, incluindo distância e elevação da fonte sonora, é dada pela função de transferência de cabeça (HRTF - do inglês \emph{Head Related Transfer Function}).\cite{hrtf} Exitem bases abertas e conhecidas de HRTF como a CIPIC e pode-se aplicar estas funções de transferência em um som por convolução (veja equação~\ref{eq:conv}).\cite{CIPIC} O corpo do indivíduo altera bastante as filtragens realizadas e existem técnicas para resultar em HRTFs que sejam - como proposta - utilizáveis de forma universal.\cite{lazaSPA}
 
@@
 
-Em posse da nota básica, podemos montar estruturas musicais com
-sequências destas partículas. Caso somemos $N$ sequências ($T_{k,i}=\{t_{k,i}\}, \;\; k \in 0...N-1$) de mesmo tamanho, seus conteúdos espectrais serão sobrepostos e a isso damos o nome de mixagem:
+Em posse da nota básica, pode-se realizar estruturas musicais com
+sequências destas partículas. A soma de $N$ sequências $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ de mesmo tamanho $\Lambda$, resulta em seus conteúdos espectrais sobrepostos em um processo de mixagem sonora:
 
 \begin{equation}\label{eq:mixagem}
-\{t_i\}=\left \{ \sum_{k=0}^{N-1}t_{k,i} \right \}
-\end{equation}
-
-\begin{figure}[h!]
-    \centering
-        \includegraphics[width=\textwidth]{figuras/mixagem}
-    \caption{Mixagem de três sequências sonoras}
+\{t_i\}_0^{\Lambda-1}=\left \{ \sum_{k=0}^{N-1}t_{k,i} \right \}_0^{\Lambda-1}
+\end{equation}
+
+\begin{figure}[h!]
+    {\centering
+        \includegraphics[width=\textwidth]{figuras/mixagem}}
+    \caption{Mixagem de três sequências sonoras. }
+
         \label{fig:mixagem}
@@
 
-Ilustramos na figura~\ref{fig:mixagem} este processo de superposição de ondas sonoras discretizadas\footnote{A figura dispõe 100 amostras, de onde podemos concluir que, se $f_a=44,1kHz$, as frequências da dente de serra, da onda quadrada e da senoide são,
-respectivamente, $\frac{f_a}{50}=882Hz$, $\frac{f_a}{25}=1764Hz$ e $\frac{f_a}{20}=2205Hz$. A duração do trecho é bastante curto $\frac{f_a=44,1kHz}{100} \approx 2 \text{ milisegundos}$.}.
- As notas mixadas são em grande parte separadas pelo ouvido por leis físicas de ressonância e pelo sistema nervoso~\cite{Roederer}.
-
+A figura~\ref{fig:mixagem} ilustra este processo de superposição de ondas sonoras discretizadas.
+A figura dispõe 100 amostras, de onde pode-se concluir que, se $f_a=44.1kHz$, as frequências da dente de serra, da onda quadrada e da senoide são,
+respectivamente, $\frac{f_a}{100/2}=882Hz$, $\frac{f_a}{100/4}=1764Hz$ e $\frac{f_a}{100/5}=2205Hz$.
+A duração do trecho é bastante curto $\frac{f_a=44.1kHz}{100} \approx 2 \text{ milisegundos}$.
 Pode-se completar com zeros para somar sequências de tamanhos diferentes.
-O resultado da mixagem de notas musicais é a harmonia musical, cujos intervalos entre as frequências e os acordes de notas simultâneas regem aspectos subjetivos e abstratos da música e sua apreciação~\cite{Harmonia}.
+
+As notas mixadas são em grande parte separadas pelo ouvido por leis físicas de ressonância e pelo sistema nervoso.\cite{Roederer} O resultado da mixagem de notas musicais é a harmonia musical, cujos intervalos entre as frequências e os acordes de notas simultâneas regem aspectos subjetivos e abstratos da música e sua apreciação.\cite{Harmonia}
 
@@
 Este mecanismo é demonstrado de forma ilustrativa na figura~\ref{fig:concatenacao} com as mesmas sequências utilizadas na figura ~\ref{fig:mixagem}.
- As sequências são curtas para as taxas de amostragem usuais (cerca de $7$ milisegundos ao total se $f_a=44,1kHz$), mas pode-se observar de forma explícita a concatenação de sequências sonoras e o exemplo é real (cada nota tem a duração maior que $100ms$ se $f_a<1kHz$).
-
-\begin{figure}[h!]
-    \centering
-        \includegraphics[width=\textwidth]{figuras/concatenacao}
+ As sequências são curtas para as taxas de amostragem usuais, mas pode-se observar a concatenação de sequências sonoras. Além disso, cada nota tem a duração maior que $100ms$ se $f_a<1kHz$.
+
+\begin{figure}[h!]
+{    \centering
+        \includegraphics[width=\textwidth]{figuras/concatenacao}}
     \caption{Concatenação de três sequências sonoras}
@@
 O princípio vertical está demonstrado nos \emph{quadros sonoros}, sons estáticos com espectros peculiares.
-Ambas as peças estão em código Python no Apêndice~\ref{cap:codigoPecas} e podem ser escutadas e vizualizadas nos links~\cite{reduced-fi} e~\cite{quadros}.
-
-Com isso estabelecemos nossa nota musical básica no som digital.
-Trabalharemos a seguir estas unidades, a evolução temporal de seus conteúdos e desenvolveremos as notas musicais com parâmetros que evoluem no tempo, como glissandos e envoltórias.
-A filtragem de componentes e os ruídos finalizam sobre a constituição da nota musical como unidade isolada.
-A estruturação musical destas notas é então abordada do ponto de vista das estruturas fora do tempo (como as escalas), trajetórias cíclicas ou que convirjam ou divirjam.
+Ambas as peças estão em código Python nos Apêndices~\ref{ap:quadros} e~\ref{ap:reduced} e estão disponíveis como parte da \emph{toolbox} \massa.\cite{MASSA}
+
+Está descrita a nota musical digital básica e a seção a seguite desenvolve a evolução temporal de seus conteúdos, como nos glissandos e nas envoltórias.
+A filtragem de componentes espectrais e a geração dos ruídos completam a constituição da nota musical como unidade isolada.
+A seção~\ref{notasMusica} é dedicada à estruturação destas notas em música através de métricas e trajetórias.
 

### 2012-11-19 agradecimentos.tex
iniciada secao 2.2

@@
 Agradeço aos participantes do labMacambira.sf.net pelas colaborações diretas e indiretas nos software, apresentações e articulações desde junho de 2011.
-Em especial agradeco ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela e Patrícia Ferraz, Glerm Soares, Vanessa Ferreira, Sérgio Teixeira de Carvalho.
+Em especial agradeco ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela, Patrícia Ferraz, Glerm Soares, Vanessa Ferreira, Sérgio Teixeira de Carvalho.
 Agradeço aos demais colaboradores das listas de email, IRC, AA e outros canais.
@@
 Agradeço ao Cultura Viva e aos pontos de cultura pelo apoio fundamental a este e outros trabalhos.
-Agradeço em especial a estes Pontões e Cultura Digital e Ação Griô pelo suporte a estes desenvolvimentos em ações regionais e nacionais: CDTL (PE), JuntaDados.org (BA), Nós Digitais (SP), Casa dos Meninos (SP), Nina Griô (SP), Pontão da Eco (RJ), Casa de Cultura Tainã (SP) e à Comissão Nacional dos Pontos de Cultura - CNPdC (GO).
+Agradeço em especial a estes Pontões e Cultura Digital e Ação Griô pelo suporte a este trabalho em ações regionais e nacionais: CDTL (PE), JuntaDados.org (BA), Nós Digitais (SP), Casa dos Meninos (SP), Nina Griô (SP), Pontão da Eco (RJ), Casa de Cultura Tainã (SP) e à Comissão Nacional dos Pontos de Cultura - CNPdC (GO).
 
@@
 
-Agradeço às comunidades de cultura e software livre por todos os conhecimentos e tecnologias repassados e que compõem este trabalho.
+Agradeço às comunidades de cultura e software livre por todos os conhecimentos e tecnologias repassados e que compõem esta contribuição.
 

### 2012-11-19 codigoPecas.tex
feita ateh localizacao espacial
iniciada secao 2.2

@@
 \label{cap:codigoPecas}
+Todas as peças a seguir se propõem a exemplificar as relações apresentadas no capítulo~\ref{cap:resultados} e são disponibilizadas online junto ao toolbox \massa.\cite{MASSA}
 
-\subsection{Reduced-fi}
+\subsection{Quadros sonoros}\label{ap:quadros}
+Montagem musical 'quadros sonoros' para demonstração da mixagem pela soma direta de sequências amostrais.
+Resultam em 5 pequenas peças de sonoridades estáticas.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:notaDisc}.
 
-\subsection{Quadros Sonoros}
+\code{Quadros sonoros 1-5}{scripts/pecas2.1/quadrosSonoros.py}
+
+\subsection{Reduced-fi}\label{ap:reduced}
+Pequena peça musical para demonstração da concatenação de sequências amostrais como notas musicais.
+Resulta em uma pequena peça de 25 segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:notaDisc} e disponibilizada online junto ao toolbox \massa.
+
+\code{reduced-fi}{scripts/pecas2.1/reduced-fi-limpo.py}

### 2012-11-19 codigoProcedimentos.tex
primeira limpada braba para versao final do msc
finalizada toda intro
feita ateh localizacao espacial
iniciada secao 2.2

--- 2012-10-28 codigoPecas.tex
@@
-\chapter{Código Computacional das Peças Musicais}
-\label{cap:codigoPecas}
+\chapter{Código computacional dos procedimentos expostos no capítulo~\ref{cap:resultados}}
+\label{cap:codigoProc}
 
-\subsection{Reduced-fi}
+\section{Códigos das relações descritas na seção~\ref{sec:notaDisc}}\label{sec:cod1}
+Todas as equações da seção~\ref{sec:notaDisc} estão descritas abaixo em Python e disponibizadas como parte da toolbox \massa.
 
-\subsection{Quadros Sonoros}
+\code{equações da seção~\ref{sec:notaDisc} em código python}{scripts/secao2.1/eqs2.1.py}
+
+\section{Códigos das relações descritas na seção~\ref{varInternas}}\label{sec:cod2}
+Todas as equações da seção~\ref{varInternas} estão descritas abaixo em Python e disponibizadas como parte da toolbox \massa.
+
+\code{equações da seção~\ref{varInternas} em código python}{scripts/secao2.2/eqs2.2.py}

### 2012-11-19 devEresultados.tex
iniciada secao 2.2

@@
 
-\begin{equation}
+\begin{equation}\label{eq:dur}
 T_{i}^{\Delta}={\{t_i\}}_{i=0}^{\lfloor \Delta . f_a \rfloor -1}
@@
 
-\begin{equation}\label{potencia}
+\begin{equation}\label{eq:potencia}
 pot(T_i)=\frac{\sum_{i=0}^{\Lambda -1} t_i^2}{\Lambda}
@@
 Está descrita a nota musical digital básica e a seção a seguite desenvolve a evolução temporal de seus conteúdos, como nos glissandos e nas envoltórias.
-A filtragem de componentes espectrais e a geração dos ruídos completam a constituição da nota musical como unidade isolada.
-A seção~\ref{notasMusica} é dedicada à estruturação destas notas em música através de métricas e trajetórias.
+A filtragem de componentes espectrais e a geração dos ruídos completam a constituição da nota musical como unidade isolada e se desdobra na seção~\ref{notasMusica}, dedicada à estruturação destas notas em música através de métricas e trajetórias.
 
@@
 
-Nossa nota musical digital básica está bem definida com os parâmetros:
+A nota musical digital básica está bem definida na seção~\ref{sec:notaDisc} com os parâmetros:
 duração, altura, intensidade (volume) e timbre. Esta é uma modelagem
-útil e paradigmática, mas não esgota o que entendemos por
+útil e paradigmática, mas não esgota o que se entedente por
 uma nota musical.
@@
 da própria nota, i.e. em uma nota real os parâmetros
-não são absolutamente fixos~\cite{Chowning}. Por exemplo, em uma nota de piano
-de 3 segundos, a intensidade tem início abrupto e decaimento progressivo
-além de variações do espectro instantâneo, com harmônicos que
+não são absolutamente fixos.\cite{Chowning} Por exemplo, uma nota de piano
+de 3 segundos tem a intensidade com início abrupto e decaimento progressivo,
+além de variações do espectro, com harmônicos que
 decaem antes dos outros e alguns que aparecem com o tempo.
@@
 aqui é: para que um som isolado disperte interesse
-por si só, faça como que tenha variações internas~\cite{Roederer}.}.
+por si só, faça como que tenha variações internas.\cite{Roederer}}.
 
@@
 do escopo de qualquer trabalho dada a considerável sensibilidade do ouvido humano
-e a complexidade da nossa cognição sonora. Desta forma, apontaremos
-recursos primários para estas variações das características na nossa nota
+e a complexidade da nossa cognição sonora. Desta forma, serão apontados
+recursos primários para estas variações das características na nota
 básica.
+
+Todas as relações descritas nesta seção estão no Apêndice~\ref{sec:cod2}
 
@@
 do som como a reverberação e a concentração de harmônicos agudos, e não apenas pela energia da onda.
-Para nós, entretanto, a mais facilmente controlada é de fato a energia da onda (veja equação ~\ref{potencia}) e esta também pode variar de diferentes formas.
+Para nós, entretanto, a mais facilmente controlada é de fato a energia da onda (veja equação~\ref{eq:potencia}) e esta também pode variar de diferentes formas.
 Uma forma mais simples e conveniente é variar a amplitude pela multiplicação da sequência toda

### 2012-11-19 introducao.tex
iniciada secao 2.2

@@
 A representação do som é chamada áudio\footnote{Os termos
-som e áudio são muitas vezes usados de forma intercambiável.~\cite{Everest}} e este pode ser proveniente da captura do som por microfones ou da síntese.
-Muitas vezes, o áudio digital é obtido através protocolos que facilitam o armazenamento e transferência dos arquivos.~\cite{procFala} Em outros momentos, a representação digital do som consiste em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
+som e áudio são muitas vezes usados de forma intercambiável.\cite{Everest}} e este pode ser proveniente da captura do som por microfones ou da síntese.
+Muitas vezes, o áudio digital é obtido através protocolos que facilitam o armazenamento e transferência dos arquivos.\cite{procFala} Em outros momentos, a representação digital do som consiste em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
 Estas amostras separadas por intervalos regulares $\lambda_a$ constituem a forma padrão de representação do som em tempo discreto chamada de modulação por código de pulsos (PCM do inglês \emph{Pulse Code Modulation}).
@@
     \centering
+        \includegraphics[width=\textwidth]{figuras/pcm}
     \caption{Som digital em modulação por código de pulsos (PCM): 25 amostras representadas por 4 bits cada uma.}
-        \includegraphics[width=\textwidth]{figuras/pcm}
         \label{fig:PCM}
@@
 A comunidade e movimento chamada \emph{'Free Software'} engloba este entendimento, mas adiciona a abordagem filosófica da liberdade e compartilhamento, dando ênfase a isso.
-Ambas as correntes reforçam o entendimento de que o código computacional é o 'bem mais precioso produzido atualmente' pois consiste em tecnologia condensada, reativa (pois executa, processa ou gera resultados), modular (partes são copiadas e reutilizadas eficientemente) e replicada sem custo adicional (pois a cópia de texto tem custo baixíssimo).\cite{Raymond,Lessig}} e facilita os processos de co-autoria.
+Ambas as correntes reforçam o entendimento de que o código computacional é o 'bem mais precioso produzido atualmente' pois consiste em tecnologia condensada, reativa (executa, processa ou gera resultados), modular (partes são copiadas e reutilizadas eficientemente) e replicada sem custo adicional (a cópia de texto tem custo baixíssimo).\cite{Raymond,Lessig}} e facilita os processos de co-autoria.
 

### 2012-11-19 scripts/pecas2.1/reduced-fi-limpo.py

@@ -69,9 +69,10 @@ smin=min(seq); smax=max(seq)
 seq= [(-.5+(i-smin)/(smax-smin)) for i in seq] # normalizando
 sound = wave.open('musica.wav','w')
 sound.setframerate(44100)
-sound.setsampwidth(2) # Always 16bit/sample (2 bytes)
-sound.setnchannels(1) # Mono
-sonic_vector=[i*(2**15-1) for i in seq] #signed 16 bit
-sound.writeframes(struct.pack('h'*len(sonic_vector),*[int(i) for i in sonic_vector]))
+sound.setsampwidth(2) # sempre 16bit/sample (2 bytes)
+sound.setnchannels(1) # mono
+sonic_vector=[i*(2**15-1) for i in seq] # 16 bit com sinal
+sound.writeframes(struct.pack('h'*len(sonic_vector),\
+                      *[int(i) for i in sonic_vector]))
 sound.close()
 print "arquivo musica.wav gravado"

### 2012-11-19 devEresultados.tex
adicionados codigos e pecas da secao 2.2

@@
 
-Todas as relações descritas nesta seção estão no Apêndice~\ref{sec:cod2}
+Todas as relações descritas nesta seção estão no Apêndice~\ref{sec:cod2}, as montagens musicais \emph{Tansita para metro}, \emph{Vibra e treme}, \emph{Tremolos, vibratos e a frequência}, \emph{Trenzinho de caipiras impulsivos}, \emph{Ruidosa faixa}, \emph{Bela rugosi}, \emph{Chorus infantil}, \emph{ADa e SaRa} estão nos Apêndices~\ref{ap:transita}, \ref{ap:vibra}, \ref{ap:tremolos}, \ref{ap:trenzinho}, \ref{ap:ruidosa}, \ref{ap:bela}, \ref{ap:chorus} e \ref{ap:ada}.
+Estes códigos são parte da caixa de ferramentas \massa, disponível online.\cite{MASSA}
 

### 2012-11-19 codigoPecas.tex
adicionados codigos e pecas da secao 2.2

@@
 Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:notaDisc}.
-
 \code{Quadros sonoros 1-5}{scripts/pecas2.1/quadrosSonoros.py}
@@
 Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:notaDisc} e disponibilizada online junto ao toolbox \massa.
+\code{reduced-fi}{scripts/pecas2.1/reduced-fi-limpo.py}
 
-\code{reduced-fi}{scripts/pecas2.1/reduced-fi-limpo.py}
+\subsection{Transita para metro}\label{ap:transita}
+Montagem musical 'transita para metro' para . Resultam em 1 pequenas peças com chirps. Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:varInternas}.
+\code{Transita para metro}{scripts/pecas2.2/transitaParaMetro.py}
+
+\subsection{Vibra e treme}\label{ap:vibra}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
+\code{Tremolos, vibratos e a frequência}{scripts/pecas2.2/vibraEtreme.py}
+
+\subsection{Tremolos, vibratos e a frequência}\label{ap:tremolos}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
+\code{Tremolos, vibratos e a frequência}{scripts/pecas2.2/TremolosVibratosEaFrequencia.py}
+
+\subsection{Trenzinho de caipiras impulsivos}\label{ap:trenzinho}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
+\code{Trenzinho de caipiras impulsivos}{scripts/pecas2.2/trenzinhoCaipiraImpulsivos.py}
+
+\subsection{Ruidosa faixa}\label{ap:ruidosa}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
+\code{Ruidosa faixa}{scripts/pecas2.2/ruidosaFaixa4.py}
+
+\subsection{Bela Rugosi}\label{ap:bela}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
+\code{Bela Rugosi}{scripts/pecas2.2/bellaRugosi.py}
+
+\subsection{Chorus infantil}\label{ap:chorus}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
+\code{Bela Rugosi}{scripts/pecas2.2/chorusInfantilb.py}
+
+\subsection{ADa e SaRa}\label{ap:ada}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
+\code{ADa e SaRa}{scripts/pecas2.2/ADa_e_SaRa.py}

### 2012-11-19 devEresultados.tex
ateh ruidos

--- 2012-11-19 devEresultados.tex
@@
 
-\begin{equation*}
+\begin{equation}\label{eq:ampVol}
 se \quad  t_i^{'}=2 . t_i \quad \Rightarrow \quad pot(T^{'}_i)=4 . pot(T_i) \quad \Rightarrow \quad V^{'}_{dB}=10log_{10} 4 \quad  \approx \quad 6 dB
-\end{equation*}
-\begin{equation*}
+\end{equation}
+\begin{equation}
 se \quad pot(T^{'}_i)=2 pot(T_i) \quad \Rightarrow \quad V^{'}_{dB}=10log_{10} 2 \quad \approx \quad 3 dB
-\end{equation*}
+\end{equation}
 
@@
 Em primeiro lugar, as características da nota se modificam no decorrer
-da própria nota, i.e. em uma nota real os parâmetros
-não são absolutamente fixos.\cite{Chowning} Por exemplo, uma nota de piano
+da própria nota.\cite{Chowning} Por exemplo, uma nota de piano
 de 3 segundos tem a intensidade com início abrupto e decaimento progressivo,
@@
 se apresentam na natureza\footnote{A regra de ouro
-aqui é: para que um som isolado disperte interesse
+aqui é: para que um som isolado desperte interesse
 por si só, faça como que tenha variações internas.\cite{Roederer}}.
@@
 do escopo de qualquer trabalho dada a considerável sensibilidade do ouvido humano
-e a complexidade da nossa cognição sonora. Desta forma, serão apontados
+e a complexidade da nossa cognição sonora e, a seguir, serão apontados
 recursos primários para estas variações das características na nota
@@
 
-Todas as relações descritas nesta seção estão no Apêndice~\ref{sec:cod2}
-
-Iniciaremos com uma técnica que não resulta propriamente
-em variações internas da nota, mas simplificará esta exposição
-sem restringir o escopo, além de ser um notável recurso de otimização
-computacional.
+Todas as relações descritas nesta seção estão no Apêndice~\ref{sec:cod2}.
+As montagens musicais \emph{Tansita para metro}, \emph{Vibra e treme}, \emph{Tremolos, vibratos e a frequência}, \emph{Trenzinho de caipiras impulsivos}, \emph{Ruidosa faixa}, \emph{Bela rugosi}, \emph{Chorus infantil}, \emph{ADa e SaRa} estão nos Apêndices~\ref{ap:transita}, \ref{ap:vibra}, \ref{ap:tremolos}, \ref{ap:trenzinho}, \ref{ap:ruidosa}, \ref{ap:bela}, \ref{ap:chorus} e \ref{ap:ada}.
+Estes códigos são parte da caixa de ferramentas \massa, disponível online.\cite{MASSA}
 
@@
 
-Mais conhecido pelo termo em inglês, a \emph{Lookup Table} (ou simplesmente
-LUT), é um procedimento simples que consiste em criar uma tabela
-de referência e coletar as amostras desta tabela segundo a 
-necessidade\footnote{Em contraposição ao cálculo das amostras individuais
-de forma analítica. A LUT permite o uso de funções sem possibilidade de cálculo direto como
-por exemplo, amostras recolhidas diretamente de algum fenômeno natural.}. Em nosso caso, usaremos tabelas de busca para os
-sons criados, as notas. Assim, estas tabelas são unidimensionais e com
-valores reais. Além disso, a tabela consiste nas amostras
-de um período de onda, que usamos para sintetizar períodos
-de tamanhos diferentes, resultando em frequências diferentes.
+Mais conhecida pelo termo em inglês, a \emph{Lookup Table} (ou simplesmente
+LUT), é uma estrutura de dados para
+consultas indexadas usada
+frequentemente para reduzir a complexidade computacional
+e por
+permitir o uso de funções sem possibilidade de cálculo direto, como
+amostras recolhidas da natureza.
 
@@
 
-O procedimento de busca em tabelas é visto geralmente como um
-artifício de otimização. Na música seu uso transcende este
-primeiro, facilitando as operações e permitindo que um único
+Na música seu uso transcende estes
+primeiros, facilitando as operações e permitindo que um único
 período de onda possa ser usado para sintetizar sons em toda a banda
@@
 
-Descrevemos esta busca. Seja $\widetilde{\Lambda}$ o tamanho 
-da tabela e $\widetilde{L_i} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ a tabela de elementos $\widetilde{l_i}$ de um
+Seja $\widetilde{\Lambda}$ o tamanho 
+do período e $\widetilde{L_i} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ os elementos $\widetilde{l_i}$ de um
 período de onda qualquer (veja equação ~\ref{periodoUnico}). Uma sequência
@@
 
-Ou seja, dada a tabela de busca, basta termos a sequência de índices corretos
-para sintetizar o som em qualquer frequência. Dispomos graficamente
-o procedimento de LUT na figura~\ref{fig:lut} onde calculamos explicitamente
-também uma amostra de $\{t_i\}$ a partir de $\left\{\,\widetilde{l}_i\,\right\}$ para uma frequência
+Ou seja, com os índices corretos ($\gamma_i\%\widetilde{\Lambda}$)
+da LUT, pode-se sintetizar o som em qualquer frequência. 
+A figura~\ref{fig:lut} ilustra
+o cálculo de uma amostra de $\{t_i\}$ 
+a partir de $\left\{\,\widetilde{l}_i\,\right\}$ para uma frequência
 de $f=200Hz$, $\widetilde{\Lambda}=128$ e considerada a taxa de amostragem em $f_a=44.1kHz$.
-Esta não é uma configuração praticável, como veremos abaixo, mas possibilita uma
-disposição gráfica bastante clara do procedimento.
-
-Para o cálculo do inteiro $\gamma_i$, convencionamos
-atribuir a ele o maior inteiro menor que $i.f\frac{\widetilde{\Lambda}}{f_a}$.
-Esta aproximação introduz um ruído, mas conseguimos que este ruído seja desprezível
-mesmo se o tomarmos como o maior inteiro acima da multiplicação
-ou se arredondarmos por outros métodos. Além disso, para fins de síntese, em $f_a=44.1 kHz$
+Esta não é uma configuração praticável, como assinalado abaixo, mas possibilita uma
+disposição gráfica do procedimento.
+
+O cálculo do inteiro $\gamma_i$ introduz um ruído, 
+e este diminui com o aumento de $\widetilde{\Lambda}$.
+Para fins de síntese, em $f_a=44.1 kHz$
  o padrão é usar $\widetilde{\Lambda} = 1024$ amostras, pois já não gera ruído
- relevante no espectro audível~\cite{Geiger}.
-
- A expressão que define a variável $\gamma_i$ pode ser facilmente compreendida da
- seguinte forma: a variável $i$ é acrescida de $f_a$ em $1$ segundo (pela
- definição de $f_a$ e $i$, pois temos $f_a$ amostras $t_i$ em $1$ segundo). Caso a dividamos por $f_a$, teremos $\frac{i}{f_a}$,
-fração esta que é acrescida de $1$ a cada $1$ segundo. Multiplicado pelo comprimento da
- tabela $\widetilde{\Lambda}$, teremos $i \frac{\widetilde{\Lambda}}{f_a}$
- que resulta na varredura completa da tabela $\widetilde{L_i}$ em 
- 1 segundo com o acréscimo de $\widetilde{\Lambda}$ em $1$ segundo. Por fim,
- se multiplicarmos pela frequência $f$ que queremos, teremos $i . f \frac{\widetilde{\Lambda}}{f_a}$
- que resulta em $f$ varreduras completas da tabela $\widetilde{L_i}$ em $1$ segundo, i.e. a sequência
+ relevante no espectro audível. O método de arredondamento ou interpolação não é decisivo.\cite{Geiger}
+
+ A expressão que define a variável $\gamma_i$ pode ser compreendida da
+ seguinte forma: $i$ é acrescida de $f_a$ em $1$ segundo.
+ Caso seja dividida pela frequência de amostragem, resulta $\frac{i}{f_a}$,
+que é acrescida de $1$ a cada $1$ segundo. Multiplicada pelo comprimento do período, resulta $i \frac{\widetilde{\Lambda}}{f_a}$
+ que varre o 
+  período em $1$ segundo. Por fim,
+ com a frequência $f$, resulta $i . f \frac{\widetilde{\Lambda}}{f_a}$
+ que completa $f$ varreduras do período  $\widetilde{\Lambda}$ em $1$ segundo, i.e. a sequência
  resultante apresenta a frequência fundamental $f$.
@@
 é entendido como um processo de otimização). Salvo quando assinalado,
-no texto que segue usaremos este procedimento para todos os casos cabíveis pois
-simplifica diversas rotinas e é computacionalmente coerente.
+no texto que usará este procedimento para todos os casos cabíveis pois
+simplifica as rotinas e é computacionalmente coerente.
 
 O uso de LUTs é bastante difundido nas implementações computacionais
-voltadas para música e para um uso clássico que explora com ênfase
-as LUTs na síntese sonora musical, recomendamos procurar pela chamada 
+voltadas para música e um uso clássico que explora com ênfase
+as LUTs na síntese sonora musical, é a  chamada 
 \emph{Wavetable Synthesis} que consiste em várias LUTs utilizadas em 
-conjunto através da mixagem para gerar uma nota musical quasi-periódica~\cite{Cook,Wavetable}.
+conjunto através da mixagem para gerar uma nota musical quasi-periódica.~\cite{Cook,Wavetable}.
 
@@
 
-Segundo revela a lei de Weber e Fechner, nossa percepção tem uma relação logarítmica com
-o estímulo que a causa~\cite{Weber-Fechner}. Em outras palavras, um estímulo em progressão exponencial
-é percebido como linear. Mesmo assim, dado o uso, explicitaremos também a variação
-linear e começaremos descrevendo esta por razões didáticas.
+Segundo a lei de Weber e Fechner, a percepção humana tem uma relação logarítmica com
+o estímulo que a causa.\cite{Weber-Fechner} Em outras palavras, um estímulo em progressão exponencial
+é percebido como linear.
+Por razões didáticas e dado o uso nas AM e FM (veja subseção~\ref{subsec:tvaf}), a variação linear será abordada primeiro.
 
 Em uma nota de duração $\Delta = \frac{\Lambda}{f_a}$, a frequência $f=f_i$ varia de $f_0$ até $f_{\Lambda -1}$
-linearmente. Podemos então escrever que:
+linearmente. Pode-se escrever:
 
 \begin{equation}\label{freqLinear}
-f_i=f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \quad ,\quad \quad i \;\in\; \mathbb{N}, \quad i \;\in\; [0,\Lambda-1]
+F_i=\{f_i\}_0^{\Lambda-1}=\left\{f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right\}_0^{\Lambda-1}
 \end{equation}
@@
 
-Desta forma, podemos calcular os elementos $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$
-com base no período $\{\widetilde{l}_i\}_0^{\Lambda-1}$.
+Desta forma, pode-se calcular os elementos $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$
+com base no período $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$.
 
@@
 dito ainda de outra forma, uma progressão geométrica da frequência
-é percebido como uma progressão aritmética de alturas.}.
-
-Fazendo a varredura com $i$
-de $0$ até $\Lambda-1$, podemos dizer que: $f_i=f_0 . 2^{\frac{i}{\Lambda-1} n_8}$ onde 
+é percebida como uma progressão aritmética de alturas.}.
+Pode-se escrever que: $f_i=f_0 . 2^{\frac{i}{\Lambda-1} n_8}$ onde 
 $n_8=\log_2\frac{f_{\Lambda-1}}{f_0}$ é o número de oitavas entre $f_0$ e $f_{\Lambda-1}$.
@@
 f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}$. Portanto,
-nossas equações de referência para a síntese de transições de frequência tidas como
-lineares pelo ouvido são:
-
-\begin{equation}
-f_i=f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}
+ as equações de transições de frequência
+lineares para o ouvido são:
+
+\begin{equation}
+F_i=\{f_i\}_0^{\Lambda-1}=\left\{f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}} \right\}_0^{\Lambda-1}
 \end{equation}
@@
 \end{equation}
-
-Onde usamos a equação \ref{indiceExponencial} para calcular a sequência $\{\gamma_i\}_0^{\Lambda-1}$ de índices para para
-serem buscados na nossa tabela que contém um único período da forma de onda representado em uma sequência de $\widetilde{\Lambda}$
-pontos igualmente espaçados (amostras).
-Isso é explicitado na equação ~\ref{serieAmostralLog}, onde estabelecemos a relação
- $t_i^{\;\overline{f_0,f_{\Lambda-1}}}=\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}$.
-Note que a equação \ref{serieAmostralLog} se mantém (é idêntica à equação \ref{serieAmostralLin}).
 
@@
 
-O termo $\frac{i}{\Lambda-1}$ varre o intervalo $[0,1]$ e podemos elevá-lo a uma potência
-para que o início da transição seja mais suave ou mais abrupto (compensado pelo final
-da transição). Este procedimento é mais comum para o caso de variações de volume através da energia
-da onda vibratória\footnote{A mudança do volume (qualidade psicofísica) ocorre por diversas características 
-do som como a reverberação e a concentração de harmônicos agudos, e não apenas pela energia da onda.
-Para nós, entretanto, a mais facilmente controlada é de fato a energia da onda (veja equação~\ref{eq:potencia}) e esta também pode variar de diferentes formas.
-Uma forma mais simples e conveniente é variar a amplitude pela multiplicação da sequência toda
-por algum número real. A variação de energia sem variação de
-amplitude é comumente chamada de \emph{compressão sonora} e é muito útil e corriqueira na
-produção musical atual~\cite{compressao}.}. Podemos obter este resultado multiplicando a sequência original
+O termo $\frac{i}{\Lambda-1}$ varre o intervalo $[0,1]$ e pode-se elevá-lo a uma potência
+para que o início da transição seja mais suave ou abrupta.
+Este procedimento é útil para variações de energia
+da onda vibratória para alteração do volume\footnote{A mudança do volume (qualidade psicofísica) ocorre através de diferentes características 
+do som, como a reverberação e a concentração de harmônicos agudos, dentre as quais está energia da onda.
+A manipulada com mais facilidade é a energia da onda (veja equação~\ref{eq:potencia}) e esta também pode variar de diferentes formas.
+Uma forma mais simples é variar a amplitude através da multiplicação da sequência toda
+por um número real. A variação de energia sem variação de
+amplitude é a \emph{compressão sonora}, útil na
+produção musical atual.\cite{compressao}}. Basta multiplicar a sequência original
 (seja ela gerada ou pré-estabelecida) pela sequência $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$
@@
 
-Assim, para variações de amplitude, aplicamos as seguintes fórmulas:
+Assim, para variações de amplitude:
 
@@
 
-Nosso intuito aqui é modificar a amplitude de um sinal original, seja qual for ele. Desta forma, tomamos
-$a_0=1$ para iniciar a nova sequência com a amplitude original e então ir modificando com o decorrer das amostras.
-Esta restrição faz com que nosso termo $a_{\Lambda-1}$ seja efetivamente a variação da amplitude.
+Pode-se tomar $a_0=1$ para iniciar a nova sequência com a amplitude original e então ir modificando com o decorrer das amostras.
+Esta restrição faz com que o termo $a_{\Lambda-1}$ seja a variação da amplitude.
 Caso $\alpha=1$, a variação de amplitude segue exatamente a progressão geométrica que caracteriza
 a percepção linear.
-A figura~\ref{fig:transicao} exibe as transições para diferentes valores de $\alpha$ e para a transição entre os valores $1$ e $2$ (se $1$ e $2$ relacionam amplitudes, há um ganho de $\approx 6dB$).
-
-Algum cuidado é necessário para lidar com $a=0$. Observe que
-na equação~\ref{seqAmp}, se $a_0=0$ haverá divisão por zero e
+A figura~\ref{fig:transicao} exibe as transições para diferentes valores de $\alpha$ e para a transição entre os valores $1$ e $2$, um ganho de $\approx 6dB$ segundo a equação~\ref{eq:ampVol}.
+
+Algum cuidado é necessário para lidar com $a=0$.
+Na equação~\ref{seqAmp}, se $a_0=0$ há divisão por zero e
 se $a_{\Lambda-1}=0$, há uma multiplicação por zero. Ambos os casos
-tornam o procedimento inútil e provém do fato de que nenhum número diferente de zero pode ser representado como uma proporção com relação ao zero.
-Podemos resolver isso escolhendo um número suficientemente pequeno.
-Podemos aceitar $-120dB\;\Rightarrow a=10^{\frac{-120}{20}}=10^{-6}$ ou valores ainda menores como nosso volume mínimo no caso de um
-\emph{fade in} ($a_0=10^{-6}$) ou de um \emph{fade out} ($a_{\Lambda-1}=10^{-6}$).
-
-Para resultar em uma amplificação linear, mas não linear para a percepção, basta usarmos uma sequência $\{a_i\}$ adequada:
+tornam o procedimento inútil pois nenhum número diferente de zero pode ser representado como uma proporção com relação ao zero.
+Pode-se resolver isso escolhendo um número suficientemente pequeno como $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ como o volume mínimo no caso de um
+\emph{fade in} ($a_0=10^{-4}$) ou de um \emph{fade out} ($a_{\Lambda-1}=10^{-4}$).
+
+Para resultar em uma amplificação linear, mas não linear para a percepção, basta uma sequência $\{a_i\}$ adequada:
 
@@
 
-Aqui pode muito nos auxiliar a conversão de decibels para amplitude apresentada na equação ~\ref{ampDec}.
-Junto à equação \ref{transAmp}, resulta na seguinte equação para uma transição de $V_{dB}$ decibels:
-
-\begin{equation}
-T_i^{'}=\left \{ t_i \left(10^{\frac{V_{dB}}{20}} \right)^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
+Aqui convém a conversão de decibels para amplitude apresentada na equação ~\ref{ampDec}.
+Junto à equação \ref{transAmp}, resulta na transição de $V_{dB}$ decibels:
+
+\begin{equation}
+T_i^{'}=\left\{ t_i 10^{\frac{V_{dB}}{20}\left( \frac{i}{\Lambda-1} \right)^\alpha} \right\}_0^{\Lambda-1}
 \end{equation}
@@
 para o caso geral de variações de amplitude segundo a progressão geométrica.
-Notamos que a transição mais natural e percebida como linear acorre com $\alpha=1$ e que quanto maior o valor de $\alpha$ mais suave é a introdução do som.
-Vale assinalar que $\alpha>1$ resulta em transições de volume muitas vezes chamadas de \emph{slow fade} enquanto $\alpha<1$ resulta em \emph{fast fade}~\cite{fades}.
-
-Veremos adiante a aplicação das transições lineares para as oscilações 
-de amplitude e frequência nas chamadas sínteses AM e FM e a aplicação das transições
+Quanto maior o valor de $\alpha$, mais suave é a introdução do som e mais intenso o final do fade.
+$\alpha>1$ resulta em transições de volume muitas vezes chamadas de \emph{slow fade} enquanto $\alpha<1$ resulta em \emph{fast fade}.\cite{fades}
+
+As transições lineares serão utilizadas para
+de as chamadas sínteses AM e FM e a aplicação das transições
 logarítmicas para os tremolos e vibratos.
@@
 \subsection{Aplicação de filtros digitais}
- As possibilidades
-aqui são variadas e a complexidade facilmente
-foge ao escopo deste trabalho\footnote{A elaboração de filtros
+Esta subseção limita-se a uma descrição
+do processamento das sequências, por convolução
+e equação a diferenças, e em aplicações
+imediatas, pois a complexidade facilmente
+foge ao escopo\footnote{A elaboração de filtros
 constitui uma área reconhecidamente complexa, com literatura
@@
 Recomendamos ao leitor
-interessado uma visita à nossa bibliografia, em especial~\cite{Openheim,smith}.},
-de forma que focamos em explicitar o que ocorre com as sequências
-de amostras e em descrever os procedimentos usuais.
-A aplicação de filtros pode
+interessado uma visita à nossa bibliografia.\cite{Openheim,smith}}. A aplicação de filtros pode
 ser parte constituinte da síntese ou feita posteriormente
 como parte dos processos tipicamente chamados de tratamento sonoro.
-
-Explicitaremos a aplicação de filtros por convolução
-e por equação a diferenças.
 
@@
 finita no tempo. Esta representação amostral é chamada
-de 'resposta ao impulso' $\{h_i\}$. Os filtros FIR são aplicados temporalmente ao som
+de 'resposta ao impulso' $\{h_i\}$. Os filtros FIR são aplicados 
+no domínio temporal ao som
 digitalizado pela convolucão do som com a 
-resposta ao impulso do filtro\footnote{Pode-se aplicar o filtro do domínio espectral através da multiplicação das transformadas de Fourier de ambos o som e a resposta ao impulso, e então realizada a transformada inversa de Fourier do espectro resultante~\cite{Openheim}.}.
-Para nossos fins, esta
+resposta ao impulso do filtro\footnote{Pode-se aplicar o filtro do domínio espectral através da multiplicação das transformadas de Fourier de ambos o som e a resposta ao impulso, e então realizada a transformada inversa de Fourier do espectro resultante.\cite{Openheim}}.
+Para os fins deste trabalho, a
 convolução fica definida como:
@@
 \begin{split}
-\{t_i'\}_0^{\Lambda_t+\Lambda_h-2\; = \;\Lambda_{t\, '}-1} =\{(t*h)_i\}_0^{\Lambda_{t \, '}-1} & =\left \{ \sum_{j=0}^{min(\Lambda_h,i)}h_{j} . t_{i-j} \right \}_0^{\Lambda_{t\, '}-1} \\
-    & =\left \{ \sum_{j=max(i-\Lambda_h,0)}^{i}t_j . h_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
-\end{split}
-\end{equation}
-
-Onde consideramos $t_i=0$ para as
+\left\{t_i'\right\}_0^{\Lambda_t+\Lambda_h-2\; = \;\Lambda_{t\, '}-1} =\{(T_j*H_j)_i\}_0^{\Lambda_{t \, '}-1} & =\left \{ \sum_{j=0}^{min(\Lambda_h-1,i)}h_{j} .
+t_{i-j} \right \}_0^{\Lambda_{t\, '}-1} \\
+    & =\left \{ \sum_{j=max(i+1-\Lambda_h,0)}^{i}t_j . h_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
+\end{split}
+\end{equation}
+
+Onde $t_i=0$ para as
 amostras não definidas de antemão.
@@
 tem cada i-ésima amostra $t_i$ substituída pela soma de suas últimas $\Lambda_h$ amostras $\{t_{i-j}\}_{j=0}^{\Lambda_h-1}$
-multiplicadas uma a uma pela resposta ao impulso $\{h_i\}_0^{\Lambda_h-1}$ em si. Isso é facilmente
-percebido pela visualização gráfica da convolução de sequências arbitrárias e dispomos
-para este fim a figura \ref{fig:conv}. Observe que a resposta ao impulso $\{h_i\}$
-é disposta na forma retrograda e que mostramos na figura os cálculos explícitos
-das amostras $t_{12}'$ e $t_{32}'$ do sinal convoluído. O sinal resultante possui
+multiplicadas uma a uma pela resposta ao impulso $\{h_i\}_0^{\Lambda_h-1}$ em si como na figura~\ref{fig:conv}. A resposta ao impulso $\{h_i\}$
+é percorrida na forma retrograda e os cálculos explícitos
+das amostras $t_{12}'$ e $t_{32}'$ são duas amostras calculadas
+pela convolução $(T_j*H_j)_i=t_i'$. O sinal resultante possui
 sempre o tamanho $\Lambda_t+\Lambda_h -1=\Lambda_{t'}$.
 
-Com este procedimento podemos aplicar reverberadores, equalizadores, delays
+Com este procedimento pode-se aplicar reverberadores, equalizadores, delays
 e vários outros tipos de filtros para fins de tratamento sonoro ou
@@
 
-Para a obtenção da resposta ao impulso, pode-se recorrer a medições
-físicas ou à síntese direta do filtro. Para citar um exemplo
-de medições diretas, uma resposta ao impulso para a aplicação
-de reverberação pode resultar da gravação acústica do ambiente ao disparar
-um estalo (cuja gravação se aproxima da resposta ao impulso diretamente) ou ao disparar uma
-varredura de frequência (cuja gravação se aproxima da resposta em frequência).
-Ambas resultam em respostas ao impulso
+A resposta ao impulso pode provir de medições
+físicas ou à síntese direta do filtro. Uma resposta 
+ao impulso para a aplicação
+de reverberação pode resultar da gravação do ao disparar
+um estalo que se assemelhe a um impulso ou 
+de uma varredura em senoide, que transformada se aproxima
+da resposta em frequência.
+
+Ambas são respostas ao impulso
 que, convoluídas com a sequência sonora, resultam na própria sequência
 com uma reverberação que se assemelha à reverberação do ambiente 
-em que ocorreu a medição~\cite{Cook}.
-
-Como exemplo de resposta ao impulso sintetizada, pode-se
-descrever um perfil espectral, realizar a transformada inversa
-de fourier e convoluir o resultante com o som, realizando
-uma filtragem cujo perfil é próximo ao especificado inicialmente.
-Note que este procedimento funciona particularmente bem se for grande o número de amostras utilizadas para especificar o perfil espectral do filtro\footnote{O que resulta em mais processamento computacional.}.
-
-Outro exemplo simples e útil vem do deslocamento temporal causado pela convolução com o impulso deslocado. Assim, podemos aplicar de linhas de \emph{delays} através
-da convolução do som com uma resposta ao impulso que possui um impulso
-em cada reincidência do som na linha de delays intensionada.
-Para ilustrar essa aplicação, dispomos a figura~\ref{fig:delays}
-em que pode-se observar o deslocamento causado pela convolução
+em que ocorreu a medição.\cite{Cook}
+
+A transformada inversa
+de fourier de uma envoltória par e real é uma
+resposta ao impulso de um FIR. Este realiza
+uma filtragem em frequência com a envoltória.
+Quanto maior o número de amostras maior
+a resolução a envoltória e também 
+o processamento computacional pois a convolução é cara.
+
+Uma propriedade importante é o deslocamento temporal causado pela convolução com o impulso deslocado. Embora caro computacionalmente, 
+pode-se criar linhas de \emph{delays} através da convolução do som com uma resposta ao impulso que possui um impulso
+para cada reincidência do som.
+Na figura~\ref{fig:delays}
+pode-se observar o deslocamento causado pela convolução
 com o impulso. Dependendo da densidade dos impulsos, o resultado
@@
 sonoro (20-40 impulsos por segundo ou mais). Neste último caso,
-ocorrem processos tipicamente vinculados à síntese granular e às
-equalizações.
+ocorrem processos tipicamente vinculados à síntese granular, delays, reverbs e equalizações.
 
@@
         \includegraphics[width=\textwidth]{figuras/delays__}
-    \caption{Convolução com o impulso e linhas de delays}
+    \caption{Convolução com o impulsoi, linhas de delays e síntese granular}
         \label{fig:delays}
@@
 
-com $b_0=1$ na grande maioria dos casos pois basta normalizarmos as variáveis
-resultando em filtro idêntico: $a_j'=\frac{a_j}{b_0}$ e $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
-
-Apontamos filtros IIR bastante simples, úteis e usuais abaixo. Existem
+com $b_0=1$ na grande maioria dos casos pois basta normalizar as variáveis para
+resultar em filtro idêntico: $a_j'=\frac{a_j}{b_0}$ e $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
+
+Existem
 diversos métodos e ferrametas para a elaboração de filtros IIR
-e esta é uma seleção com fins didáticos e para consulta futura por
+e segue abaixo uma seleção com fins didáticos e para consulta futura por
 utilidade.
-São filtros de primeira e segunda ordem bem comportados cujas
-filtragens realizadas dispomos na figura~\ref{fig:iir}.
+São filtros bem comportados e cujas
+filtragens estão na figura~\ref{fig:iir}.
 
@@
 No caso dos filtros passa e rejeita banda, esta mesma atenuação é
-resultado de duas especificações: $f_c$ (neste caso melhor compreendida como 'frequência central') e a largura de banda $bw$
-pois em ambas as frequências $f_c \pm bw$ há uma atenuação de $\approx 0.707$ da amplitude original.
-
-Observe que existe amplificação do som no caso dos filtros passa e rejeita banda quando a frequência
+resultado de duas especificações: $f_c$ (neste caso melhor compreendida como 'frequência central') e a largura de banda $bw$,
+em ambas as frequências $f_c \pm bw$ há uma atenuação de $\approx 0.707$ da amplitude original.
+
+Existe amplificação do som no caso dos filtros passa e rejeita banda quando a frequência
 de corte é baixa e a largura de banda é grande o suficiente. Nos agudos, estes filtros apresentam
-somente um desvio do perfil esperado, expandindo a envoltória um pouco mais do lado grave da banda em
+somente um desvio do perfil esperado, expandindo a envoltória para o lado grave da banda em
 evidência.
 
-Para conseguir filtros cujas respostas em frequência possuem outras envoltórias (para o módulo),
-o recurso mais simples a esta altura
-é fazer cascata destes filtros aplicando-os sucessivamente.
-
-Outra possibilidade é usar alguma receita de filtro
+Para filtros cujas respostas em frequência possuem outras envoltórias (para o módulo),
+pode-se realizar cascatas destes filtros aplicando-os sucessivamente.
+Outra possibilidade é utilizar alguma receita de filtro
 biquad\footnote{Abreviação
@@
 de filtros Chebichev\footnote{Filtros Butterworth e Elípticos podem
-ser considerados como casos específicos dos Filtros do tipo Chebichev~\cite{Openheim,smith}.}.
+ser considerados como casos específicos dos Filtros do tipo Chebichev.\cite{Openheim,smith}}.
 Ambas as possibilidades são exploradas
-por títulos em nossas referências, em especial recomendamos~\cite{JOSFM,smith} e a coleção de filtros da comunidade \emph{Music-DSP}, da Universiade de Columbia~\cite{music-dsp}.
-Para uma exposição aprofundada do assunto, recomendamos~\cite{Openheim}.
+por títulos em nossas referências, em especial~\cite{JOSFM,smith} e a coleção de filtros da comunidade \emph{Music-DSP}, da Universiade de Columbia.\cite{music-dsp,Openheim}
 
@@
 \begin{enumerate}
-\item Passa-baixas de polo simples com gráfico do canto superior esquerdo da figura~\ref{fig:iir}. A fórmula geral tem
+\item Passa-baixas de polo simples com módulo da resposta em frequência no canto superior esquerdo da figura~\ref{fig:iir}. A fórmula geral tem
 por referência da frequência de corte $f_c \in (0,\frac{1}{2})$,
@@
 em que há aproximadamente uma atenuação de $3dB$.
-Calculamos os coeficientes do filtro IIR
+Os coeficientes do filtro IIR
 $a_0$ e $b_1$ 
-através da variável intermediária: $x \in [e^{-\pi},1]$:
+são dados através da variável intermediária: $x \in [e^{-\pi},1]$:
 
@@
 
-\item Passa-altas de polo simples com gráfico do canto superior direito da figura~\ref{fig:iir}. A fórmula geral,
+\item Passa-altas de polo simples o módulo da resposta em frequência no canto superior direito da figura~\ref{fig:iir}. A fórmula geral,
 com frequência de corte $f_c \in (0,\frac{1}{2})$, é calculada através da variável
@@
 \item Nó (\emph{notch filter}). Este filtro é parametrizado
-pela frequência central\footnote{Não
-confundir com a frequência de corte também notada aqui por $f_c$ nos filtros passa baixas e passa altas.} $f_c$
+pela frequência central\footnote{ Atenção com a frequência de corte também $f_c$ nos filtros passa baixas e passa altas.} $f_c$
 e a largura de banda $bw$
-($f_c \pm bw$ resulta em 0.707 da amplitude, i.e. atenuação de $3dB$),
+- $f_c \pm bw$ resultam em 0.707 da amplitude, i.e. atenuação de $3dB$ -
 ambos dados como fracoes de $f_a$, portanto $f,\; bw \in (0,0.5)$.
-Por conveniência definimos as variáveis auxiliares $K$ e $R$ como:
+
+Por facilidade, sejam as variáveis auxiliares $K$ e $R$:
 
@@
 
-A partir das quais temos dois filtros. O filtro passa banda que dispomos no canto inferior esquerdo da figura~\ref{fig:iir}:
+O filtro passa banda do canto inferior esquerdo da figura~\ref{fig:iir}:
 
@@
 
+com o módulo de sua resposta em frequência 
 disposto na parte inferior esquerda da figura~\ref{fig:iir}.
@@
 
-\subsection{Usos musicais parte 1: tremolo e vibrato, AM e FM}
+\subsection{Usos musicais parte 1: tremolo e vibrato, AM e FM}\label{subsec:tvaf}
 

### 2012-11-19 scripts/convolucao.py
ateh ruidos

@@ -2,13 +2,13 @@
 import numpy as n, pylab as p
 
 ax=p.subplot(111)
-ax.annotate(r"$t_{12}'=\sum_{j=0}^9 \, h_j.t_{12-j}$", xy=(12.2, -2.65),  xycoords='data',
+ax.annotate(r"$t_{12}'=\sum_{j=3}^{12} \, t_j.h_{12-j}$", xy=(12.2, -2.65),  xycoords='data',
                 xytext=(23, -3.75), textcoords='data',
                 arrowprops=dict(facecolor='black', shrink=0.05),
                 horizontalalignment='right', verticalalignment='top', fontsize=20
                 )
 
-ax.annotate(r"$t_{32}'=\sum_{j=0}^9 \, h_j.t_{32-j}$", xy=(32.2, -1.75),  xycoords='data',
+ax.annotate(r"$t_{32}'=\sum_{j=24}^{32} \, t_j.h_{32-j}$", xy=(32.2, -1.75),  xycoords='data',
                 xytext=(43, -3.75), textcoords='data',
                 arrowprops=dict(facecolor='black', shrink=0.05),
                 horizontalalignment='right', verticalalignment='top', fontsize=20
@@ -28,7 +28,7 @@ p.plot(t,'bo', label=r'trecho de sinal sonoro $\{t_i\}_0^{47}$')
 #h=1./x[1:11]**.5
 h=n.random.random(10)
 h=n.array([ 0.5591728 ,  0.59152829,  0.43285462,  0.8870076 ,  0.44892785,0.33476906,  0.8808893 ,  0.39040725,  0.56887214,  0.54278373])
-p.plot(range(3,13),h[::-1]+1,'ro',label=r'retrogrado da resposta ao impulso $\{h_{i-j}\}_{j\; =\; max(i-\Lambda_h,0)}^i$')
+p.plot(range(3,13),h[::-1]+1,'ro',label=r'retrogrado da resposta ao impulso $\{h_{i-j}\}_{j\; =\; max(i+1-\Lambda_h,0)}^i$')
 for i in xrange(3,12):
     p.plot([i,i],[-1,2],'y-.')
 p.plot([12,12],[-5,2], 'y-.',linewidth=3)
@@ -40,7 +40,7 @@ for i in xrange(23,32):
[…]

### 2012-11-20 devEresultados.tex
adentrando 2.3

@@
 De forma geral, os sons sem altura definida 
-são chamados ruídos~\cite{Lacerda}.
-Estes sons estão presentes como constituintes importantes dos sons musicais de altura definida,
+são chamados ruídos.\cite{Lacerda}
+Estes sons são constituintes importantes dos sons musicais de altura definida,
 como os ruídos presentes nas notas do piano, do violino, etc. Além disso, os instrumentos
 de percussão, em grande parte, não possuem altura definida e seus sons
-são muitas vezes compreendidos como ruídos~\cite{Roederer}. Na música eletrônica (incluindo
-a eletroacústica e gêneros de pista de dança), os ruídos possuem usos amplos, diversificados e comumente
-idiomáticos~\cite{Cook}.
-
-A ausência de uma altura definida é fruto da ausência de uma organização harmônica perceptível nas componentes senoidais que formam o som. Assim, pode-se inferir
-as incontáveis possibilidades de gerar ruídos. A mais imediata é a utilização
-de valores aleatórios para a geração da sequência sonora $T_i$. Este
-é um método atraente, visto a disponibilidade
-das consagradas distribuições de probabilidade em diversas linguagens de programação,
+são em geral compreendidos como ruídos.\cite{Roederer} Na música eletrônica, incluindo
+a eletroacústica e gêneros de pista de dança, os ruídos possuem usos diversificados e comumente
+idiomáticos.\cite{Cook}
+
+A ausência de uma altura definida é fruto da ausência de uma organização harmônica perceptível nas componentes senoidais que formam o som. Assim,
+são incontáveis as possibilidades de gerar ruídos.
+A utilização
+de valores aleatórios para a geração da sequência sonora $T_i$
+é um método atraente,
 mas os resultados não são tão úteis, tendendo geralmente ao ruído branco.\cite{Cook}
 
-Outra possibilidade é a geração de ruído através do espectro desejado. A partir
-dele, executamos a transformada inversa de Fourier. Neste caso, é importante
-que realizemos a distribuição espectral com cuidado pois queremos um perfil
-do módulo do espectro e a fase é aleatória\footnote{Caso utilizemos a mesma fase
-(ou fases com forte correlação) para
-as componentes envolvidas, o som sintetizado possuirá energia bastante concentrada
-em alguns trechos.}. Veja a figura~\ref{fig:ruidos} e o código Python que a gerou para uma implementação.
+Outra possibilidade é a geração de ruído através do espectro desejadoi, a partir
+do qual executamos a transformada inversa de Fourier.
+A distribuição espectral deve ser feita com cuidado 
+pois caso utilizemos a mesma fase,
+ou fases com forte correlação, 
+o som sintetizado possuirá energia bastante concentrada
+em alguns trechos de sua duração.
 
@@
 Abaixo elencamos alguns ruídos de espectro estático. São chamados \emph{coloridos} por terem sido associados a cores. 
+A figura~\ref{fig:ruidos} dipoe lado a lado o perfil espectral e
+a sequência sonora. Os ruídos foram gerados com a mesma fase, então
+pode-se observar o resultado das contribuições em diferentes regiões do espectro.
 
@@
 \item O ruído branco deve seu nome por possuir energia distribuida
-igualmente por todas as frequências. Podemos obter a realização
-do ruído branco com a transformada inversa dos seguintes coeficientes:
+igualmente por todas as frequências. Podse realizar
+o ruído branco com a transformada inversa dos seguintes coeficientes:
 
@@
 
-A frequência mínima $f_{\text{min}}$ pode ser escolhida com base no limite da audição, pois não escutamos como altura uma componente sonora cuja frequência esteja abaixo de $\approx\; 20Hz$.
+A frequência mínima $f_{\text{min}}$ pode ser escolhida com base no limite da audição, pois não se escuta como altura uma componente sonora cuja frequência esteja abaixo de $\approx\; 20Hz$.
 
 O resto dos ruídos podem ser feitos com base neste procedimento descrito para 
-o ruído rosa. Basta que modifiquemos alguns detalhes. Em especial a equação que define $\alpha_i$.
+o ruído rosa, basta que modificar detalhes, em especial a equação que define $\alpha_i$.
 
 \item O ruído marrom deve seu nome ao Robert Brown, que descreveu o movimento browniano.
-Embora esta origem seja um tanto díspar do que poderíamos considerar motivo para uma associação com a cor marrom, o ruído sonoro ficou consagrado com este nome.
+Embora esta origem seja um tanto díspar do que pode-se considerar motivo para uma associação com a cor marrom, o ruído sonoro ficou consagrado com este nome.
 De qualquer forma, é bastante comum declarar satisfatória a associação do ruído com a cor marrom, dados os ruídos branco e rosa mais estridentes e relacionados a cores mais intensas~\cite{marrom}.
 
-O que caracteriza este ruído é a queda de $6dB$ por oitava. Desta forma, basta modificarmos a equação de $\alpha_i$ 
-no conjunto \ref{eq:rosa} para:
+O que caracteriza este ruído é a queda de $6dB$ por oitava. Desta forma, $\alpha_i$ 
+no conjunto \ref{eq:rosa} fica:
 
@@
 
-\item Ruído azul é aquele em que há um ganho de $3dB$ por oitava em uma banda limitada. Para
-a síntese deste ruído, basta estipularmos a frequência mínima $f_{\text{min}}$ e a frequência
-máxima $f_{\text{máx}}$. Assim, também com base no conjunto de equações \ref{eq:rosa}, 
-basta atentarmos para as sequintes identidades:
+\item No ruído azul há ganho de $3dB$ por oitava em uma banda limitada
+pela frequência mínima $f_{\text{min}}$ e a frequência
+máxima $f_{\text{máx}}$. Assim, também com base no conjunto de equações \ref{eq:rosa}:
 
@@
 
-\item Um pouco diferente dos exemplos acima é o ruído cinza, definido como
-um ruído branco sujeito a uma das curvas iso-audíveis. Como estas curvas são resultados
-experimentais, precisamos neste caso da curva para a realização dos fatores multiplicativos
-$\alpha_i$.
+\item O ruído cinza e definido como
+um ruído branco sujeito a uma das curvas iso-audíveis. Estas curvas são resultados
+experimentais e necessárias para a obtenção de $\alpha_i$. Uma implementação da ISO 226, a última revisão destas curvas, está na toolbox \massa.\cite{MASSA}
 
@@
 
-Com base nestes ruídos expostos podemos prever a infinidade de ruídos utilizados. Importante
-neste momento é entender que foram expostos somente ruídos com espectro estático. Existem classificações
+Foram expostos somente ruídos com espectro estático. Existem classificações
 de ruídos com variações do espectro no decorrer do tempo. Existem também ruídos que
 são fundamentalmente transientes, como os clicks e os chirps. O primeiro é modelado
-facilmente por um impulso relativamente isolado, enquanto o segundo é uma varredura rápida de 
-alguma banda de frequência~\cite{Cook}.
+facilmente por um impulso relativamente isolado, enquanto o segundo não é um ruído, mas uma varredura rápida de 
+alguma banda de frequência.\cite{Cook}
 
 Os ruídos das equações \ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom},
-\ref{eq:azul}, \ref{eq:violeta} estão na figura \ref{fig:ruidos}. Vale notar
-que os espectros foram feitos com a mesma fase nos coeficientes, de forma que
-pode-se observar a contribuição dos harmônicos agudos no traçado geral dado
-pelas frequências graves e muito explícito no ruído marrom.
+\ref{eq:azul}, \ref{eq:violeta} estão na figura \ref{fig:ruidos}. Os espectros foram feitos com a mesma fase em cada coeficiente de mesma frequência, de forma que
+pode-se observar a contribuição dos harmônicos agudos e
+das frequências graves.
 
@@
 Por exemplo, no piano, o chamado tremolo é um vibrato e um tremolo segundo a classificação aqui utilizada.
-As definições presentes neste trabalho tem por base uma literatura mais abrangente do que a utilizada para um único instrumento, prática ou tradição musical e comum em contextos de teoria musical e musica eletrônica~\cite{Lacerda,Harmonia}.}
-
-Iniciemos pelo vibrato. Para realizarmos o caso mais geral, façamos uma sequência $t_i'$
-de frequência $f'$ com o auxílio
-de uma segunda tabela $\widetilde{M}_i$ de tamanho $\widetilde{\Lambda}_M$ que apresente também
-um período de onda que oscila entre $[-1,1]$. Descrevemos uma sequência sonora $t_i^{vbr(f',\,\nu)}$ com
-um vibrato de frequência $f'$ e profundidade  $\mu$ (quantos herz é a oscilação no pico superior)
-ou $\nu$ (amplitude do vibrato em semitons) da seguinte forma:
+As definições presentes neste trabalho tem por base uma literatura mais abrangente do que a utilizada para um único instrumento, prática ou tradição musical e comum em contextos de teoria musical e musica eletrônica.\cite{Lacerda,Harmonia}}
+
+Para o caso geral, o vibrato é descrito da seguinte forma:
 
@@
 \begin{equation}\label{vbrT}
-T_i^{f, vbr(f',\,\nu)}=\{ t_i^{f,vbr(f',\,\nu)} \}_0^{\Lambda-1}=\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \}_0^{\Lambda-1}
+T_i^{f, vbr(f',\,\nu)}=\left\{ t_i^{f,vbr(f',\,\nu)} \right\}_0^{\Lambda-1}=\left\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \right\}_0^{\Lambda-1}
 \end{equation}
@@
 de Herz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação ($2\nu$ é o número de semitons entre os picos superioes e inferiores de oscilação da frequência do som $\{t_i\}$ causada pelo vibrato).
-Note que $\nu=\log_{2}\frac{f+\mu}{f} $ é conveniente neste caso pois o aumento máximo de frequência
+É conveniente $\nu=\log_{2}\frac{f+\mu}{f} $ neste caso pois o aumento máximo de frequência
 não equivale à diminuição máxima, mas a variação de semitons se mantém.
@@
 $1000Hz$ (entre um si e um dó) e cujo desvio da frequência atinge uma oitava
-para cima e para baixo. Note o padrão senoidal do vibrato e as compontentes harmônicas
-da dente de serra. Qualquer forma de onda pode
-ser utilizada para gerar o som e o padrão de oscilação do vibrato, o que é tão crucial para os usos musicais quanto a
-frequência de oscilação e o desvio de altura envolvido\footnote{O desvio de altura
+para cima e para baixo. Qualquer forma de onda pode
+ser utilizada para gerar o som e o padrão de oscilação do vibrato
+em quaisquer frequência de oscilação e desvio de altura envolvidos\footnote{O desvio de altura
 é chamado profundidade do vibrato e é geralmetne dado por conveniência em semitons ou cents.}.
@@
 
-O caso do tremolo é semelhante e $f'$, $\gamma_i'$ e $t_i'$ permanecem os mesmos. Calculamos
-a sequência de amplitudes a serem multiplicadas pela sequência original $t_i$ da
-sequinte forma:
+O caso do tremolo é semelhante: $f'$, $\gamma_i'$ e $t_i'$ permanecem os mesmos.
+A sequência de amplitudes a serem multiplicadas pela sequência original $t_i$ fica:
 
@@
  é o ganho máximo de amplitude envolvido.
-A medição em decibels é bastante pertinente pois o aumento máximo de amplitude
+A medição em decibels é pertinente pois o aumento máximo de amplitude
 não equivale à diminuição máxima relacionada, enquanto a diferença em decibels se mantém.
@@
 
-A montagem musical \emph{Vibra e treme} explora estes recursos dos tremolos e vibratos
+A montagem musical \emph{Vibra e treme} explora estes recursos dos tremolos e vibratos em associação e isoladamente
 com frequências $f'$
-e profundidades ($\nu$ e $V_{dB}$) diferentes.
-Usados em associação e isoladamente os tremolos e vibratos e com variações progressivas dos parâmetros\footnote{Vale lembrar que os tremolos e vibratos ocorrem muitas vezes juntos em instrumentos tradicionais e na voz.}.
+e profundidades ($\nu$ e $V_{dB}$) diferentes, variações progressivas dos parâmetros\footnote{Os tremolos e vibratos ocorrem muitas vezes juntos em instrumentos tradicionais e na voz.}.
 A peça desenvolve também uma comparação entre os vibratos e tremolos em escala logarítmica e em escala linear para uma apreciação qualitativa.
+Seu código está no Apêndice~\ref{ap:vibra} e disponível online como parte da \emph{toolbox} \massa.
 
@@
 
-Algo interessante ocorre se aumentarmos progressivamente $f'$.
-Primeiramente a aproximação do limiar de frequência
+No aumento progressico de $f'$,
+a aproximação do limiar de frequência
 para audição do fenômeno sonoro como
@@
 ambos tremolos e vibratos. Estas rugosidades são muito apreciadas
-tanto na tradição erudita quanto na música eletrônica atual, especialmente no \emph{Dubstep}~\cite{DubRugoso}.
-A rugosidade também é gerada através de conteúdos espectrais que geram batimentos~\cite{Porres}.  A sequência \emph{Bela Rugosi's Dade}
+tanto na tradição erudita quanto na música eletrônica atual, especialmente no \emph{Dubstep}.\cite{DubRugoso}
+A rugosidade também é gerada através de conteúdos espectrais que geram batimentos.\cite{Porres}  A sequência \emph{Bela Rugosi}
 explora este limiar com concomitâncias de tremolos e vibratos na mesma
-voz, com intensidades e formas de onda diferentes.
-
-Se continuarmos aumentando a frequência, deixamos de ouvir estas oscilações
-como eventos identificáveis. Neste caso, as oscilações se apresentam na forma de espectro
+voz, com intensidades e formas de onda diferentes. Seu código está o Apêndice~\ref{ap:rugosi} e disponível online como parte da \massa.
+
+Aumentando ainda mais a frequência, estas oscilações
+deixam de ser eventos identificáveis. 
+Neste caso, as oscilações são
 audível como altura. Assim, $f'$, $\mu$ e a forma de onda realizam alterações espectrais no som
@@
 
-Para fins musicais e em resumo, podemos entender a síntese FM através do caso entre senoides
+Para fins musicais e em resumo, pode-se entender a síntese FM através do caso entre senoides
 e decompor os sinais em seus espectros de Fourier (i.e. senoidais) para casos mais complexos.
 Assim, a síntese FM realizada com um 'vibrato' senoidal de frequência $f'$ e profundidade $\mu$ em um som também senoidal $T_i$ de frequência $f$
-e duração $\Delta$ segundos (assumimos $\Lambda = \lfloor \Delta . f_a \rfloor $ amostras)
 gera bandas centradas em $f$ e distantes $f'$ entre si:
@@
 \begin{split}
-\{t_i'\}_0^{\Lambda -1} & = \left \{ \cos \left [f . 2 \pi \frac{i}{f_a-1} + \mu . sen \left ( f' . 2 \pi \frac{i}{ f_a -1 } \right ) \right ] \right \}_0^{\Lambda-1}  = \\
- & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ f . 2 \pi \frac{i}{f_a-1} + k . f' . 2 \pi \frac{i}{f_a-1} \right ]  \right \}_0^{\Lambda-1} = \\
- & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ (f+k.f') . 2 \pi \frac{i}{f_a-1} \right ]  \right \}_0^{\Lambda-1}
+\{t_i'\} & = \left \{ \cos \left [f . 2 \pi \frac{i}{f_a-1} + \mu . sen \left ( f' . 2 \pi \frac{i}{ f_a -1 } \right ) \right ] \right \} = \\
+ & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ f . 2 \pi \frac{i}{f_a-1} + k . f' . 2 \pi \frac{i}{f_a-1} \right ]  \right \} = \\
+ & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ (f+k.f') . 2 \pi \frac{i}{f_a-1} \right ]  \right \}
 \end{split}
@@
 \end{equation}
-é a função de Bessel~\cite{BesselCCRMA,JOSFM} que, na FM, como podemos notar, especifica a amplitude de cada componente.
-Note que, nestas equações, a variação de frequência introduzida por $\{t_i'\}$ não respeita a progressão geométrica de frequência que acompanha a percepção de altura.
-Caso utilizemos as equações~\ref{vbrF} teríamos uma distribuição diferente dos harmônicos.
-Para este fim, dispomos o Apêndice~\ref{cap:fmam} onde calculamos o conteúdo espectral da síntese FM feita com oscilações na escala logarítmica.
-De fato, como pode-se notar nos cálculos, o comportamento simples que torna a FM atraente é obtido somente com as variações lineares utilizadas em~\ref{eq:fmEsp}.
+é a função de Bessel~\cite{BesselCCRMA,JOSFM} que na FM especifica a amplitude de cada componente.
+
+Nestas equações, a variação de frequência introduzida por $\{t_i'\}$ não respeita a progressão geométrica de frequência que acompanha a percepção de altura, mas sim a equação~\ref{freqLinear}.
+A utilização das equações~\ref{vbrF} para a FM está no Apêndice~\ref{cap:fmam} onde é calculado o conteúdo espectral da síntese FM feita com oscilações na escala logarítmica.
+De fato, o comportamento simples que torna a FM atraente é obtido somente com as variações lineares utilizadas em~\ref{eq:fmEsp}.
 
@@
 original, distantes $f'$ de $f$.
-Observe que utilizamos novamente a variação na escala linear de amplitude.
-Dispomos no Apêndice~\ref{cap:fmam} uma exposição do espectro da AM se realizada a oscilação na escala logarítmica de amplitude, que também perde o comportamento simples.
-
-A realização da FM e AM através de LUTs é consideravelmente simples.
-A frequência $f$ a ser modulada, chamada portadora, é modulada pela $f'$, chamada moduladora.
+Novamente, isso é obtido com a variação na escala linear de amplitude.
+No Apêndice~\ref{cap:fmam} está uma exposição do espectro da AM realizada com a oscilação na escala logarítmica de amplitude.
+Esta também perde o comportamento simples.
+
+A sequência $T_i$ de frequência $f$, chamada portadora, é modulada pela $f'$, chamada moduladora.
 No jargão de FM e AM, $\mu$ e $\alpha=10^{\frac{V_{dB}}{20}}$ são chamados índices de modulação.
-Assim, temos
-para o padrão de vibração da moduladora $\{t_i'\}$ as mesmas equações utilizadas para os tremolos e vibratos:
+Assim,
+para o padrão de vibração da moduladora $\{t_i'\}$ as equações:
 
@@
 
-Para aplicação da moduladora $\{t_i'\}$ na portadora $\{t_i\}$,
-temos, para a FM:
+Para aplicação da moduladora $\{t_i'\}$ na portadora $\{t_i\}$
+por FM:
 
@@
 
-Para realizar a AM, basta modular $\{t_i\}$ com $\{t_i'\}$ através das equações (lembrando que $\alpha=10^{\frac{V_{dB}}{20}}$):
+Para realizar a AM, basta modular $\{t_i\}$ com $\{t_i'\}$ através das equações:
 
@@
 \subsection{Usos musicais parte 2}
-A este ponto as possibilidades musicais explodiram. Possuímos unidades
-musicais com características próprias de altura (dada pela frequência),
-tímbre (dado pela forma de onda),
-volume (dado pela intensidade) e duração (dada pelo número de amostras).
-Em uma nota, estas características podem ser consideradas
+A este ponto as possibilidades musicais explodiram. As
+características de altura (dada pela frequência),
+tímbre (dado pela forma de onda e filtragens),
+volume (dado pela intensidade) e duração (dada pelo número de amostras)
+podem ser consideradas
 de forma absoluta ou tratadas ao longo de sua duração,
@@
 
-Desta forma, os usos musicais que apresentamos é uma coleção de possibilidades
+Desta forma, os usos musicais aqui dispostos são uma coleção de possibilidades
 com o objetivo de exemplificar manipulações sonoras que resultem algo
@@
 
-Uma primeira possibilidade interessante é o estabelecimento de vínculos
+Uma primeira possibilidade interessante é a utilização de vínculos
 entre os parâmetros do tremolo e do vibrato e algum parâmetro da nota básica,
-digamos a frequência. Assim, podemos estabelecer que quanto mais aguda é a nota,
-mais alta é a frequência do vibrato e do tremolo mas menor a sua profundidade.
-Desta forma, tomando por base as equações \ref{vbrGamma}, \ref{vbrF} e \ref{trA}
-podemos escrever:
+como a frequência. Assim, pode-se estabelecer
+a frequência do vibrato  é diretamente proporcional à altura à qual a profundidade do tremolo é inversamente proporcional.
+Desta forma, com as equações \ref{vbrGamma}, \ref{vbrF} e \ref{trA}
+pode-se escrever:
 
@@
 
-Com $f^{vbr}$ e $f^{tr}$ sendo $f'$ nas equações de referência, ou seja, a frequência
-de oscilação do vibrato e do tremolo utilizada na equação~\ref{vbrGamma}. Já $\nu$ e $V_{dB}$ são as profundidades
+Com $f^{vbr}$ e $f^{tr}$ como $f'$ nas equações de referência, ou seja, a frequência
+de oscilação do vibrato e do tremolo utilizadas na equação~\ref{vbrGamma}. Já $\nu$ e $V_{dB}$ são as profundidades
 do vibrato e do tremolo, respectivamente. As funções $func_a$,
-$func_b$ e $func_c$ são arbitrárias e dependentes das intensões musicais. A montagem musical 
+$func_b$ e $func_c$ são arbitrárias e dependentes das intensões musicais. A montagem 
 \emph{Tremolos, vibratos e a frequência} explora
 recursos como este e variações da forma de onda da oscilação com vínculos, de modo a formar um \emph{idioma musical}\footnote{Veja na próxima sessão.}.
-
-Com relação à convolução, podemos estabelecer uma duração como pulso musical - a exemplo de um pulso BPM - 
+Seu código está no Apêndice~\ref{ap:tremolos} e também disponivel online como parte da caixa de ferramentas \massa.
+
+Com relação à convolução, pode-se estabelecer uma duração como pulso musical - a exemplo de um pulso BPM - 
 e distribuir impulsos no decorrer deste pulso, de forma a estabelecer métricas e ritmos\footnote{Lembrando
 que a convolução com o impulso resulta no som deslocado ao instante de ocorrência do impulso.}.
-Por exemplo, se escolhemos 2 impulsos igualmente espaçados, estaremos fazendo uma
-divisão binária básica do pulso. Se escolhemos dois sinais, um com 2 pulsos e outro com 3 pulsos,
-ambos com os impulsos igualmente espaçados na duração do pulso, resultamos na manutenção
+Por exemplo, 2 impulsos igualmente espaçados fazem uma
+divisão binária básica do pulso. Dois sinais, um com 2 pulsos e outro com 3 pulsos,
+ambos com os impulsos igualmente espaçados na duração do pulso, resultam na manutenção
 do pulso, com uma marcação rítmica usada tanto em divisões binárias quanto ternárias em diversos
-estilos de música étnica e tradicional~\cite{Gramani}. 
+estilos de música étnica e tradicional.\cite{Gramani} 
 Os próprios valores absolutos destes impulsos resultam em proporções entre as amplitudes dos sinais
 convoluídos.
-Exploramos este recurso da métrica
-estabelecida pela convolução com impulsos na sequência musical \emph{trenzinho de caipiras impulsivos}.
+Este recurso da métrica
+estabelecida pela convolução com impulsos é explorado na montagem \emph{Trenzinho de caipiras impulsivos}.
 Os recursos explorados incluem a criação de amálgamas sonoros provenientes da síntese granular e esta montagem é já uma ponte para a próxima sessão.
 Veja especialmente a figura~\ref{fig:pulsoSubAgl}.
-
-Já com os filtros as possibilidades explodem ainda mais vertiginosamente. Pode-se convoluir um sinal para reverberá-lo, para
-remover algum ruído, para gerar distorções ou para tratamento com intuito estético mesmo (soa melhor por algum motivo). Por exemplo,
-com a aplicação de um filtro passa banda em que deixamos passar somente entre $1kHz$ e $3kHz$, geramos um som
-que parece de telefone ou de televisão antiga. Ao remover com alguma precisão somente
-a frequência de oscilação da rede elétrica (usualmente $50Hz$ ou $60Hz$) e harmônicas, podemos remover
-ruídos introduzidos pelos equipamentos de áudio utilizados. Um uso mais incrementado
-e propriamente musical seria realizar filtragens em bandas específicas e usar estas bandas
+O código da peça está na seção~\ref{ap:trenzinho} e disponível como parte da toolbox \massa.
+
+Com os filtros as possibilidades explodem ainda mais vertiginosamente. Pode-se convoluir um sinal para reverberá-lo, para
+remover algum ruído, para gerar distorções ou para tratamento com intuito estético mesmo. Por exemplo,
+a aplicação de um filtro passa banda em que deixa passar somente entre $1kHz$ e $3kHz$, resulta em sons
+que parecem de telefone ou de televisão antiga. Ao remover com alguma precisão somente
+a frequência de oscilação da rede elétrica (usualmente $50Hz$ ou $60Hz$) e harmônicas, pode-se remover
+ruídos introduzidos por equipamentos de áudio.
+Um uso mais incrementado
+e propriamente musical é realizar filtragens em bandas específicas e usar estas bandas
 pré-estabelecidas como um parâmetro adicional das notas.
 
-Um recurso considerado musicalmente forte pelo resultado que se aproxima dos instrumentos tradicionais é a filtragem dependente do tempo~\cite{Roederer}. Cascatas
-destes filtros podem realizar filtragens complexas e mais precisas. A montagem \emph{ruídos e faixas} explora
-este recursos, realizando filtragens em ruídos diversos e síntese de ruídos diversos.
+Inspirado em nos instrumentos tradicionais, pode-se aplicar uma a filtragem dependente do tempo~\cite{Roederer}.
+Cascatas
+destes filtros podem realizar filtragens complexas e mais precisas. A montagem \emph{Ruidosa faixa} explora
+este recursos, realizando filtragens em ruídos diversos e síntese de ruídos diversos. O código da peça está no Apêndice~\ref{ap:ruidosa} e disponível online como parte da \massa.
 
@@
 inicial são então mixadas (ver equação \ref{eq:mixagem}). A peça \emph{Chorus infantil} realiza chorus de formas
-diferentes em sons diferentes.
-
-Para terminar esta sessão com algo simples, vamos descrever a envoltória de volume chamada ADSR (sigla de \emph{Atack-Decay-Sustain-Release}).
-A envoltória ADSR possui numerosas implementações em sintetizadores de diversas procedências.
-Uma implementação pioneira pode ser encontrada no Hammond Novachord de 1938~\cite{ADSR} e algumas variantes são citadas logo abaixo.
+diferentes em sons diferentes e seu código está no Apêndice~\ref{ap:chorus}. A \emph{toolbox} \massa\ disponibiliza online este \emph{script}.
+
+A variação de volume no decorrer de um som é crucial para nossa percepção de timbre.
+A envoltória de volume chamada ADSR (sigla de \emph{Atack-Decay-Sustain-Release}) possui numerosas implementações em sintetizadores em hardware e software.
+Uma implementação pioneira pode ser encontrada no Hammond Novachord de 1938 e algumas variantes são citadas logo abaixo.\cite{ADSR}
 
@@
 Com $\Lambda_X=\lfloor \Delta .
-f_a \rfloor\;\;\forall\;\; X \; \in (A,D,R,\;)$ e $\xi$ um valor bastante pequeno que torne o \emph{fade in} e o \emph{fade out} satisfatórios, e.g.
-$\xi=10^{\frac{-120}{20}}=10^{-6}\;$ ou $\;\xi=10^{\frac{-180}{20}}=10^{-9}$.
+f_a \rfloor\;\;\forall\;\; X \; \in (A,D,R,\;)$ e $\xi$ um valor pequeno que torne o \emph{fade in} e o \emph{fade out} satisfatórios, e.g.
+$\xi=10^{\frac{-80}{20}}=10^{-4}\;$ ou $\;\xi=10^{\frac{-40}{20}}=10^{-2}$.
+Quanto menor for $\xi$ mais lento é o face, a exemplo de $\alpha$ da figura~\ref{fig:transicao} .
 
@@
 
-Podemos ver esquematicamente esta envoltória na figura~\ref{fig:adsr}. O envelope ADSR descrito é
-uma implementação clássica e paradigmática que comporta variações diversas. Por exemplo,
+Esquematicamente, a figura~\ref{fig:adsr} mostra esta envoltória ADSR, 
+uma implementação clássica que comporta variações diversas. Por exemplo,
 entre o ataque e o decaimento, pode-se adicionar uma partição adicional em que a amplitude
-máxima perdura. Outro exemplo comum é a elaboração de traçados mais elaborados para o
+máxima perdura. Outro exemplo comum é o uso de traçados mais elaborados para o
 ataque ou para o decaimento. A montagem musical \emph{ADa e SaRa} explora diversas destas

### 2012-11-20 codigoPecas.tex
feita toda correcao ateh conclusai

@@
 \code{ADa e SaRa}{scripts/pecas2.2/ADa_e_SaRa.py}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%% SECAO 2.3
+
+\subsection{Intervalos entre alturas}\label{ap:intervalos}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
+\code{Intervalos entre alturas}{scripts/pecas2.3/intervalosEntreAlturas.py}
+
+\subsection{Acorde cedo}\label{ap:acorde}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
+\code{Acorde cedo}{scripts/pecas2.3/acordeCedo.py}
+
+\subsection{Conta ponto}\label{ap:conta}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
+\code{Cristais}{scripts/pecas2.3/contaPonto.py}
+
+\subsection{Cristais}\label{ap:cristais}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
+\code{Cristais}{scripts/pecas2.3/cristais.py}
+
+\subsection{Micro tom}\label{ap:micro}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
+\code{Cristais}{scripts/pecas2.3/microTom.py}
+
+\subsection{Poli Hit Mia}\label{ap:poli}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
+\code{Cristais}{scripts/pecas2.3/poliHitMia.py}
+
+\subsection{Dirracional}\label{ap:dirracional}
+Pequena peça musical para demonstração da .
+Resulta em uma pequena peça de XX segundos.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
+\code{Cristais}{scripts/pecas2.3/dirracional.py}

### 2012-11-20 devEresultados.tex
feita toda correcao ateh conclusai

@@
 Seja $ S_j=\{T_{i,1},T_{i,2}...T_{i,N}\} =\bigcup_{j \in [1,N]} T_{i,j} $ uma sequência de $N$ eventos
-musicais. Para simplificar, sejam notas musicais estes eventos musicais.
-Considerando $S$ uma estrutura musical, existem diversas técnicas
-para tornar estas estruturas interessantes e agradáveis.
-
-Os elementos de $S_j$ podem ser sobrepostos por mixagem (veja equação~\ref{eq:mixagem} e a figura~\ref{fig:mixagem}) formando intervalos e acordes ou concatenados (veja equação~\ref{eq:concatenacao} e a figura~\ref{fig:concatenacao}) formando sequências melódicas e ritmos.
+musicais. Para simplificar, sejam notas estes eventos musicais
+e $S_j$ uma estrutura musical. Esta seção é didicada às técnicas
+que torna $S_j$ interessante e agradável na audição.
+
+Os elementos de $S_j$ podem ser sobrepostos por mixagem, como na equação~\ref{eq:mixagem} e figura~\ref{fig:mixagem}, formando intervalos e acordes.
+Resultam no 'pensamento vertical' da música.
+A concatenação de $S_j$, como na equação~\ref{eq:concatenacao} e na figura~\ref{fig:concatenacao}, forma sequências melódicas e ritmos, associados ao 'pensamento horizontal' na música.
+A frequência fundamental $f$ e o momento de início (ataque) são as características mais importantes dos elementos de $S_j$, possibilitando, grosso modo, a preseça de músicas de alturas (harmonia e melodia) e a preseça da métrica temporal e ritmos, respectivamente.
 
 \subsection{Afinação, intervalos, escalas e acordes}
-Quando dobramos a frequência, temos uma oitava ascendente ($f=2f_0$).
+O dobro da frequência é uma oitava ascendente ($f=2f_0$).
 Doze semitons equidistantes para o ouvido formam uma oitava,
-portanto, se $f=2^{\frac{1}{12}}f_0$, temos um semitom entre $f_0$
+portanto, se $f=2^{\frac{1}{12}}f_0$, há um semitom entre $f_0$
 e $f$.
-O fator $\varepsilon=2^{\frac{1}{12}}$, equivalente a um semitom, forma uma grade de notas
-no espectro audível, em que as frequências fundamentais possíveis
+O fator $\varepsilon=2^{\frac{1}{12}}$, de um semitom, forma uma grade de notas
+no espectro audível. Fixada uma frequência $f$, as frequências fundamentais possíveis
 estão separadas por intervalos múltiplos de $\varepsilon$.
@@
 computacionais, pois os intrumentos reais possuem desvios destas frequências para melhor compatibilizar os harmônicos
-de suas notas~\cite{Roederer}.
+de suas notas.\cite{Roederer}
 
 Esta divisão da oitava em doze notas é o cânone da música ocidental clássica,
-além de usos cerimoniais/religiosos e étnicos~\cite{Wisnick}.
-Os intervalos entre quaisquer duas notas da escala de 12 notas em que se subdivide a oitava possuem nomes específicos.
+além de incindir em usos cerimoniais/religiosos e étnicos.\cite{Wisnick} Os intervalos entre quaisquer dois graus da escala de 12 notas.
 Notados em proporções de $\varepsilon$ entre as notas (i.e.
@@
 
-Nesta nomenclatura (um dos padrões presentes no Brasil~\cite{Lacerda}),
+Nesta nomenclatura, um dos padrões presentes no Brasil,
 um intervalo justo (uníssono, quarta, quinta e oitava), ou um intervalo maior (segunda maior, terça maior, sexta maior ou sétima maior), acrescido de um semitom, resulta em um intervalo aumentado.
@@
  necessidade da nomenclatura, notação e uso de intervalos aumentados/diminutos e mais que aumentados/diminutos é consequência do sistema tonal, em que os graus da escala (veja abaixo) são realmente notas diferentes, com funções e usos específicos.
-Assim, em uma escala de dó bemol maior, a tônica - primeiro grau - é dó bemol, não si, e a sensível - sétimo grau - é si bemol, não lá sustenido ou dó duplo bemol.}.
-
-Caso as notas sejam dispostas sequencialmente, o intervalo é dito melódico.
+Assim, em uma escala de dó bemol maior, a tônica - primeiro grau - é dó bemol, não si, e a sensível - sétimo grau - é si bemol, não lá sustenido ou dó duplo bemol.}.\cite{Lacerda}
+
+Caso as notas sejam dispostas sequencialmente, o intervalo é melódico.
 A ordem das notas, primeiro a nota mais grave ou a mais aguda, resulta em um intervalo ascendente ou descendente, respectivamente.
-Passando a nota mais grave para a oitava acima, ou a nota mais aguda uma oitava para baixo, invertemos o intervalo.
-Um pouco de exploração do sistema revela que um intervalo, somado à sua inversão, resulta 9.
+Passando a nota mais grave para a oitava acima, ou a nota mais aguda uma oitava para baixo, o intervalo é invertido.
+Um intervalo, somado à sua inversão, resulta 9.
 Um intervalo maior invertido resulta em um intervalo menor e vice-versa.
@@
 
-Caso as notas soem simultaneamente, o intervalo é dito harmônico.
+Caso as notas soem simultaneamente, o intervalo é harmônico.
 Os intervalos de uníssono, oitava e quinta justa são consonâncias perfeitas.
-As terças e sextas são consideradas consonancias imperfeitas.
+As terças e sextas são consonancias imperfeitas.
 As segundas, sétimas e trítonos são considerados dissonantes.
 Esta classificação não é absoluta.
-A discrepância mais comum é com relação à quarta justa, que é tida como consonante ou dissonante de acordo com o contexto e arcabouço teórico.
-Podemos considerá-la consonante sem muitos riscos.
-A sétima menor por vezes é considerada consonante, assim como algumas culturas entoam o trítono como intervalo consonante~\cite{Roederer,Wisnick}.
+A discrepância mais comum é com relação à quarta justa, que é tida como consonante perfeita ou dissonante de acordo com o contexto e arcabouço teórico.
+Pode-se considerá-la consonante como regra geral.
+A sétima menor por vezes é considerada consonante, assim como algumas culturas entoam o trítono como intervalo consonante.\cite{Roederer,Wisnick}
 
@@
 
-Este arcabouço resume com exatidão a teoria tradicional dos intervalos musicais~\cite{Lacerda} e nos será útil adiante.
-Dispomos esta nomenclatura também por completude desta exposição, não há necessidade alguma de decorar estes nomes ou apreender estes detalhes todos enquanto não há uso efetivo.
-A montagem \emph{Intervalos entre alturas} explora estes intervalos de formas isoladas e diversas.
+Este arcabouço resume com exatidão a teoria tradicional dos intervalos musicais.\cite{Lacerda} A montagem \emph{Intervalos entre alturas} explora estes intervalos de formas isoladas e diversas.
+O código está no Apêndice~\ref{ap:intervalos} e disponível online com a \emph{toolbox} \massa.
 
 Na escala cromática
-de temperamento igual, temos 5 divisões perfeitamente simétricas, consideradas
-escalas pelas simetrias internas e usos fáceis que disso resultam. Descrevamos
-as escalas como sequências de inteiros aos quais $\varepsilon$ é elevado, resultando
-nos fatores a serem multiplicados por $f_0$ para obtenção dos graus de cada escala:
+de temperamento igual, existem 5 divisões perfeitamente simétricas, consideradas
+escalas pelas simetrias internas e usos fáceis que disso resultam. Como sequências de inteiros aos quais $\varepsilon$ é elevado
+para multiplicar $f_0$, as escalas são:
 
@@
 repetidas de forma eficiênte e sustentada. Discutiremos mais sobre simetrias e voltaremos
-a estas divisões da oitava por igual. A montagem \emph{Cristais} expõe cada uma destas escalas tanto melódica quanto harmonicamente.
-
-As escalas diatônicas também são facilmente especificadas, assim como
-as escalas tonais maior e menor:
-
-\begin{equation}
+a estas divisões da oitava por igual.
+A montagem \emph{Cristais} expõe cada uma destas escalas tanto melódica quanto harmonicamente e seu código está no Apêndice~\ref{ap:cristais}.
+Como parte da \massa, este código é disponibilizado online.
+
+As escalas diatônicas:
+
+\begin{equation}\label{eq:escalas}
 \begin{split}
@@
 
-Observe que nas escalas diatônicas só estão presentes intervalos maiores, menores e justos dentre quaisquer duas notas escohidas.
-Única exceção para o trítono, que se apresenta como quarta aumenta ou quinta diminuta, dependendo da nota mais grave que se leva em consideração.
-
-Todas as escalas diatônicas (salvo as menores harmônica e melódica)
+só possuem intervalos maiores, menores e justos. Única exceção para o trítono, que se apresenta como quarta aumenta ou quinta diminuta.
+
+Todas as escalas diatônicas
 seguem o padrão de intervalos sucessivos
@@
 Com $\kappa \in \mathbb{N}$. Para cada
-modo, temos um único valor de $\kappa$ em qualquer intervalo $\in [x,x+6], \forall x \in \mathbb{N}$.
-
-Por exemplo, consideremos o modo lídio. Uma breve
-inspeção revela que $e_i^{ml}=d_{(i+2)\%7}+e_{i-1}^{ml}$. Temos, portanto, $\kappa=2$
+modo, há um único valor de $\kappa$ em qualquer intervalo $\in [x,x+6], \forall x \in \mathbb{N}$.
+
+Por exemplo, uma breve
+inspeção revela que $e_i^{ml}=d_{(i+2)\%7}+e_{i-1}^{ml}$. Portanto, $\kappa=2$
 para o modo lídio.
-Várias outras escalas e sequências são facilmente representadas desta forma. Em especial,
-lembramos aqui das pentatônicas e os modos de transposição limitados de messiaen~\cite{Messiaen}.
-
-A escala menor se apresenta em duas formas adicionais, dependentes do uso, melódico ou harmônico:
+
+A escala menor possui duas formas adicionais, melódica e harmônica:
 
@@
 
-e podem ser utilizadas livremente, embora tradicionamente se associem aos usos melódicos ou harmônicos sugeridos pelos seus nomes.
-Observe o traçado ascendente e descendente da escala menor melódica.
-Isso é necessário para que se apresente a sensível (sétimo e último grau separado por um semitom da oitava) no trajeto ascendente, o que não é necessário quando descende, retomando a forma natural.
-Já a harmônica apresenta a sensível, mas não evita o intervalo de segunda aumentada entre o sexto e sétimo graus, pois não precisa contemplar trajetória melódica, apenas apresentar a sensível tão cara ao sistema tonal~\cite{Harmonia}.
-
-A ocorrência simultânea de notas é observada musicalmente em grande medida
-através dos acordes.
-Destes, a base na musica tonal são as tríades, que consistem em duas terças sucessivas.
-Listamos os acordes triádicos sem inversão e na forma fechada\footnote{Um acorde invertido é aquele que apresenta no grave outra nota que não a fundamental.
-A posição fechada é aquela em que não cabe nota alguma do acorde entre quaisquer duas notas consecutivas~\cite{Lacerda}.},
-com a fundamental em $0$:
+O traçado ascendente e descendente da escala menor melódica é necessário para a presença da sensível (sétimo e último grau separado por um semitom da oitava) no trajeto ascendente, o que não é necessário quando descende, retomando a forma natural.
+Já a escala harmônica apresenta a sensível, mas não evita o intervalo de segunda aumentada entre o sexto e sétimo graus, pois não precisa contemplar trajetória melódica, apenas apresentar a sensível tão cara ao sistema tonal (a sensível tende à tônica, afirmando-a).\cite{Harmonia}
+
+Outras escalas podem ser representadas da mesma forma, como as pentatônicas e os modos de transposição limitados de messiaen.\cite{Messiaen}
+
+A ocorrência simultânea de notas é observada através dos acordes.
+Destes, a base na musica tonal são as tríades.
+Estas constituem-se de duas terças sucessivas, em 3 notas: fundamental, terça e quinta.
+Os acordes triádicos sem inversão e na forma fechada\footnote{Um acorde invertido é aquele que apresenta no grave outra nota que não a fundamental.
+A posição fechada é aquela em que não cabe nota alguma do acorde entre quaisquer duas notas consecutivas.\cite{Lacerda}},
+com a fundamental em $0$ são escritos;
 
@@
 com sétima menor ou $11$ para resultar em sétima maior. As inversões e posições abertas
-podem ser obtidas contando o número de semitons de cada nota a partir da nota mais grave.
-
-Para a realização de microtonalidade\footnote{O uso explícito de
-intervalos menores que o semitom é chamado microtonalismo. Isso pode ter usos
-ornamentais ou estruturantes da música em si. A
-divisão da oitava em $12$ notas igualmente espaçadas possui fundamentos físicos mas
+podem ser obtidas contando o número de semitons de cada nota a partir da nota mais grave ou somando $12$ à fundamental, terça ou quinta.
+
+Para a realização de microtonalidade\footnote{O uso de
+intervalos menores que o semitom é chamado microtonalidade e tem usos
+ornamentais e estruturantes da música. A
+divisão da oitava em $12$ notas possui fundamentos físicos mas
 não deixa de ser uma \emph{convenção}
-(reconhecidamente adotada pela música erudita clássica de origem européia). Outras afinações
-são também incidentes na manifestação musical humana, para citar somente um exemplo, a
+adotada inclusive pela música erudita clássica de origem européia. Outras afinações
+são incidentes. Para citar somente um exemplo, a
 música tradicional tailandesa utiliza uma divisão da oitava em sete notas igualmente
@@
 resultando em intervalos que pouco se assemelham aos intervalos presentes na
-divisão de doze notas~\cite{Wisnick}.},
-podemos usar reais não inteiros
+divisão de doze notas.\cite{Wisnick}},
+pode-se usar reais não inteiros
 para descrever a sequência de alturas ou modificar o fator $\varepsilon$
-e usar inteiros ainda. Por exemplo, uma afinação muito
-interessante por contemplar a série harmônica de forma consideravelmente precisa,
-é proposta na forma da divisão da oitava em $53$ notas, de forma
-que $\varepsilon=2^{\frac{1}{53}}$~\cite{microtonalidade}. As sequências de alturas
-continuam sendo compostas por inteiros com o novo $\varepsilon$.
-Note que se $S_i$ é uma sequência de alturas $s_i$ e mudamos de $\varepsilon_1$ para
+e continuar usando inteiros. Por exemplo, uma afinação
+bastante próxima da série harmônica em si,
+é proposta na forma da divisão da oitava em $53$ notas:
+$\varepsilon=2^{\frac{1}{53}}$.\cite{microtonalidade}
+As notas
+continuam inteiras com o novo $\varepsilon$.
+Note que se $S_i$ é uma sequência de alturas $s_i$ e mapeamos uma sequência em $\varepsilon_1$ para
 $\varepsilon_2$,
 resultamos em uma nova sequência $S_i'$ com $s_i'=s_i \frac{\varepsilon_1}{\varepsilon_2}$.
+A montagem musical \emph{Micro tom} explora recursos microtonais e seu código está no Apêndice~\ref{ap:micro}, assim como parte da \massa\ online.
 
@@
 
-Abordaremos de forma bastante resumida e voltada para nossas aplicações
-computacionais a harmonia tonal. Para harmonia modal e atonal recomendamos~\cite{harmEXT}.
-
-Na chamada harmonia tradicional, as escalas maior e menor são usadas
-de base para tríades feitas em cada grau da escala com duas terças consecutivas.
-É conveniente, assim, notar os graus da escala pelos números da seguinte forma: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$.
-
-Assim, as tríades formadas em cada grau é de fácil notação $\hat{1},\hat{3},\hat{5}$ é o acorde de primeiro grau, formado pelo primeiro grau da escala e central para uma música tonal.
-Em segundo lugar temos o acorde do quinto grau $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sustenido no caso da escala menor) e em terceiro o do quarto grau $\hat{4},\hat{6},\hat{1}$.
-Depois consideramos as outras tríades todas.
-Ao longo da história, formas diferentes de lidar com estas tríades foram desenvolvidas.
-Estas convenções e técnicas apresentadas ao longo da história é o objeto de estudo da harmonia tradicional~\cite{Harmonia}.
+Acordes como os listados em~\ref{triades} formados em cada
+grau das escalas em~\ref{eq:escalas} formam campos harmônicos
+das escalas. As progressões de acordes utilizadas e regras para seus encadeamentos é o objeto de estudo da harmonia musical.
+Mesmo uma melodia monofônica gera campos harmônicos.
+
+Não respeitar 
+os encadeamentos básicos do sistema tonal resulta
+em harmonias modal e atonal. Caso
+as notas coincidam com algum modo diatônico~\ref{eq:escalas}
+ou sejam em número pequeno, podemos dizer que a
+harmonia é modal. Caso contrário, atonal.\cite{harmEXT}.
+
+Na harmonia tonal tradicional,
+as escalas maior e menor são
+base para tríades em cada grau 
+da escala: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$.
+Pode-se anotar $\hat{1},\hat{3},\hat{5}$ para o acorde do primeiro grau, formado pelo primeiro grau da escala e central para uma música tonal.
+Secundários, o acorde do quinto grau $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sustenido no caso da escala menor) e o acorde do quarto grau $\hat{4},\hat{6},\hat{1}$.
+Depois são considerados os outros graus.
+As convenções e técnicas apresentadas ao longo da história é o objeto de estudo da harmonia tradicional.\cite{Harmonia}
 
 A chamada harmonia funcional atribui funções a estes três acordes centrais e busca compreender seus usos através destas funções.
-O acorde formado sobre o primeiro grau é chamado de acorde de tônica e tem a função de manter um centro e um "chão" na música.
-O formado sobre o quinto grau é chamado de acorde de dominante e tem a função de tender à tônica, direcionar a música para ela.
-A tríade formada sobre o quarto grau é chamada de subdominante e tem a função de se distanciar da tônica.
+O acorde formado sobre o primeiro grau é o acorde de tônica e tem a função de manter um centro, um "chão" na música.
+O acorde formado sobre o quinto grau é o acorde de dominante e tem a função de tender à tônica, direcionar a música para ela.
+A tríade formada sobre o quarto grau é a subdominante e tem a função de se distanciar da tônica.
+O sistema se baseia em afirmar a tônica através de encadeamentos tônica-dominante-tônica expandidos com outros acordes das formas mais diversas.
 
@@
 As exatas funções e efeitos musicais destes acordes é motivo de bastante controvérsia.
-Dispomos da tabela~\ref{tab:harmonia} esta relação entre as tríades formadas em cada grau da escala.
+A tabela~\ref{tab:harmonia} mostra a relação entre as tríades formadas em cada grau da escala.
 
@@
 
+A dominante anti-relativa forma um acorde menor, exigindo a alteração do quarto grau um semitom para cima $\hat{7}\#$.
+O acorde diminuto $\hat{7},\hat{2},\hat{4}$, geralmente é considerado uma 'dominante com sétima e sem fundamental'.\cite{Koellheuteur}
+
 Na escala menor, a relativa é a tríade iniciada uma terça acima e a antirelativa uma terça abaixo,
@@
 menor é a alteração do $\hat{7}$ um semitom acima para que fique somente a um semitom da tônica,
-permitindo a dominante de fato (que deve ser maior e tender à tônica).
+permitindo a dominante (que deve ser maior e tender à tônica).
 Assim, a dominante é sempre maior, tanto em escalas maior ou menor.
@@
 
-A dominante anti-relativa forma um acorde menor, exigindo a alteração do quarto grau um semitom para cima $\hat{7}\#$.
-O acorde diminuto $\hat{7},\hat{2},\hat{4}$, geralmente é considerado uma 'dominante com sétima e sem fundamental'~\cite{Koellheuteur}.
-
 Cada um destes acordes pode ser confirmado e se desenvolver com uma dominante e uma subdominante individuais (acorde com base na tríade formada a uma quinta acima e uma quinta abaixo respectivamente).
 
-Por último, vale apontar as medianas\footnote{Também chamadas de medianas cromáticas.}, que são acordes formados também a uma terça, mas com alterações com relação ao acorde de origem.
+As medianas\footnote{Também chamadas de medianas cromáticas.}, são acordes formados também a uma terça, mas com alterações com relação ao acorde de origem.
 Caso haja duas alterações cromáticas, i.e.
 duas notas alteradas por um semitom cada com relação ao acorde original, a mediana é chamada 'duplamente cromática'.
-Esta relação entre acordes é considerada de tonalismo avançado, por vezes até de dissolução do tonalismo, e tem efeitos fortes e marcantes.
-Foi muito utilizada do final do romantismo em diante por Wagner, Lizt, Richard Strauss dentre outros e é bastante simples de ser realizada~\cite{Harmonia,Schachter}.
-
-As modulações e sutilezas de encaminhamento de vozes podem ser encontradas em títulos diversos como os citados acima e constituem uma discussão longa e ainda presente.
-Apontamos somente que, para a condução de diferentes vozes, os movimentos que levam a notas vizinhas costumam soar mais coerentes, a importância maior da voz mais aguda e mais grave (e menor das vozes internas) e que as modulações se dão principalmente através de uma dominante individual que afirma a mudança da tônica e através de alterações cromáticas e enharmonias (mudança de função de um acorde).
-A importância da dominante torna ela o alvo primário e natural das modulações, resultando no circulo das quintas~\cite{Harmonia,Schachter,Koellheuteur,Harmony}.
-
-A montagem musical \emph{Acorde cedo} explora estas relações entre acordes.
-
-\subsection{Rudimentos de contraponto}
+Esta relação entre acordes é considerada de tonalismo avançado, por vezes até de dissolução do tonalismo, e tem efeitos fortes e marcantes embora perfeitamente consonantes.
+As medianas foram utilizadas a partir do final do romantismo por Wagner, Lizt, Richard Strauss dentre outros e é bastante simples de ser realizada.\cite{Harmonia,Schachter}
+
+A modulação é a troca de tonalidade e se dão principalmente através de: uma dominante individual que afirma a mudança da tônica; alterações cromáticas para um acorde da nova tonalidade; e enharmonia que é a mudança de função de um acorde com as mesmas notas.
+A importância da dominante torna ela o pivô natural das modulações, resultando no circulo das quintas.\cite{Harmonia,Schachter,Koellheuteur,Harmony}
+
+A montagem musical \emph{Acorde cedo} explora estas relações entre acordes. Seu código está disponível do Apêndice~\ref{ap:acorde} e online como parte da \massa.\cite{MASSA}
+
+\subsection{Contraponto}
+Na condução de diferentes vozes, os movimentos a notas vizinhas soam mais coerentes. A importância melódica recai sobre as vozes mais aguda e mais grave.
 
 A condução de linhas melódicas simultâneas, chamadas vozes,
-é matéria do campo de estudo chamado contraponto.
-A nossa bibliografia aponta títulos para estudos específicos, geralmente percorrendo formas sistemáticas de condução de vozes e desembocando em gêneros escolásticos como cânones, invenções e fugas.
-É possível resumir
-as regras principais do contraponto e é dito que o próprio Beethoven - dentre outros - esboçou uma síntese.
-
-Como
-resumo de nossa bibliografia, apontamos que as dissonâncias só podem ser atingidas por movimento de
-grau conjunto (notas vizinhas) ou resolvidas
-imediatamente por movimento descendente de grau conjunto da voz superior. Além disso,
-as consonâncias perfeitas só devem ser atacadas por movimento contrário ou oblíquo entre
-as vozes. O movimento paralelo só é permitido entre consonâncias imperfeitas e não mais
-do que 3 vezes consecutivas~\cite{Fux,Tragtenberg,SchoenbergContra}.
-Estas regras tem como um dos objetivos principais, preservar a independência entre as vozes.
-Dispomos os movimentos tidos como base para o contraponto na figura~\ref{fig:movContraponto}.
-Observe o movimento paralelo como um caso específico do movimento oblíquo.
+é matéria do campo de estudo chamado contraponto. A bibliografia
+geralmente percorre formas sistemáticas de condução de vozes e desemboca em gêneros escolásticos como cânones, invenções e fugas.
+É possível resumir regras principais do contraponto e é dito que o próprio Beethoven - dentre outros - esboçou uma síntese deste tipo.
 
@@
 
+O propósito do contraponto é conduzir as vozes 
+de forma que soem independentes. Crucial para isso
+são as movimentações relativas,
+categorizadas em: movimento direto, oblíquo e contrário
+conforme a figura~\ref{fig:movContraponto}. O movimento paralelo
+é um tipo de movimento oblíquo.
+A regra de ouro é cuidar que os movimentos diretos
+não resulte em consonância perfeita. O movimento paralelo
+deve só ocorrer entre consonâncias imperfeitas e não mais
+do que três vezes consecutivas. As dissonâncias podem
+ser não admitidas ou usadas seguidas 
+e precedidas de consonâncias em graus conjuntos.\cite{Fux,Tragtenberg,SchoenbergContra}
+
+Estas regras foram usadas na montagem \emph{Conta ponto}, o código está no Apêndice~\ref{ap:conta} e disponível online junto à \massa.
+
 \subsection{Ritmo}
-A noção rítmica é dependente de eventos separados por durações~\cite{rythm}. Estes eventos são discerníveis individualmente se separados
-por ao menos $50-63ms$. Para que a separação entre eles possa ser apreciada, ela deve ser ainda um pouco maior,
-por volta de $100ms$~\cite{microsound}. Podemos sumarizar 
-a transição do domínio da apreciação
-de durações como alturas para a apreciação em ritmo da seguinte forma ~\cite{Alfaix, microsound}:
+A noção rítmica é depende de eventos separados por durações.\cite{rythm} Estes eventos podem ser ouvidos individualmente se separados
+por ao menos $50-63ms$. Para que a separação entre eles possa ser apreciada como duração, ela deve ser maior,
+por volta de $100ms$.\cite{microsound} Pode-se sumarizar 
+a transição da
+de durações ouvidas alturas para a apreciação em ritmo da seguinte forma:\cite{Alfaix, microsound}
 
@@
 
-A banda de durações marcada como transição está minimizada pois os limites não são bem definidos: a duração específica em que começamos a perceber nitidademente uma frequência fundamental ou uma separação entre as ocorrências é dependente da pessoa e de características do som como conteúdo espectral e intensidade~\cite{microsound,Roederer}.
+A banda de durações marcada como transição está minimizada pois os limites não são bem definidos: a duração em que começa-se a perceber uma frequência fundamental ou uma separação entre as ocorrências é dependente da pessoa e de características do som.\cite{microsound,Roederer}
 
@@
 costuma-se associar esta gama de frequências de pulsação às durações entre batidas 
-do coração ou entre os passos ao caminhar~\cite{Lacerda,Roederer}.
+do coração ou entre os passos ao caminhar.\cite{Lacerda,Roederer}
 
@@
 \footnote{Em ordem crescente de ocorrência na música
-escrita e étnica, podemos
-dizer que as divisões do pulso musical e seus agrupamentos
-sequenciais no tempo ocorrem com fatores 2, 4 e 8, depois 3, 6 (dois grupos de 3 ou 3 grupos de 2) e 9 (3 grupos de 3). Por 
-último os primos 5 e 7, completando 1-9.
-Outras métricas são incidentes (divisões ou agrupamentos em 13, 17, etc), mas não são tão usuais e ocorrem principalmente em música experimental e erudita do século XX para frente.
-As métricas, por mais complexas que pareçam, costumam ser decompostas em divisões como estas~\cite{Gramani,Roederer}.}.
-Uma exposição esquemática está na figura ~\ref{fig:pulsoSubAgl}.
+escrita e étnica,
+as divisões do pulso musical e seus agrupamentos
+sequenciais no tempo ocorrem com fatores 2, 4 e 8, depois 3, 6 (dois grupos de 3 ou 3 grupos de 2) e 9 e 12 (3 e 4 grupos de 3). Por 
+último os primos 5 e 7, completando 1-9 e 12.
+Outras métricas são incidentes (divisões ou agrupamentos em 13, 17, etc), embora menos usuais e principalmente em contextos de música experimental e erudita do século XX e XXI.
+Por mais complexas que pareçam, as métricas costumam ser composições e decomposições de 1-9 partes iguais.\cite{Gramani,Roederer}.}.
+Uma exposição esquemática está na figura~\ref{fig:pulsoSubAgl}.
 
@@
     \caption{Divisões e aglomerações do pulso musical para estabelecimento de métrica.
-Ao lado esquerdo temos as divisões da semínima estabelecida como pulso.
-Ao lado direito temos exemplos de fórmulas de compasso que especificam as mesmas métricas, mas na escala produzida pelas ocorrências e aglomerações do pulso musical.}
+Ao lado esquerdo estão as divisões da semínima estabelecida como pulso.
+Ao lado direito, fórmulas de compasso que especificam as mesmas métricas, mas na escala das aglomerações do pulso musical.}
         \label{fig:pulsoSubAgl}
@@
 
-As unidades mais fortes (acentuadas) são as que consistem nas 'cabeças das divisões'.
-A cabeça de um tempo ou uma unidade dividida é a primeira parte da subdivisão.
-Nas divisões binárias (2, 4 e 8 dos casos considerados aqui),
+As unidades mais fortes (acentuadas) são as que consistem nas 'cabeças das divisões'. A cabeça de uma unidade é a primeira parte da subdivisão. Nas divisões binárias (2, 4 e 8 dos casos considerados),
 as unidades consideradas fortes se revesam com as fracas
-(e.g. a divisão em 4 resulta em forte, fraco, (meio) forte, fraco).
-Nas divisões ternárias (3, 6 e 9 dos casos considerados neste escrito)
+(e.g. a divisão em 4 resulta em forte, fraco, meio-forte, fraco).
+Nas divisões ternárias (3, 6 e 9)
 à unidade forte (primeira) se sucedem 2 unidades fracas (e.g. a divisão em 3 resulta em forte, fraco fraco)\footnote{A divisão em 6 é considerada composta
@@
 
-A acentuação em tempos fracos é chamada contratempo, as notas que se iniciam em tempos fracos e cuja duração percorrem tempos fortes são as chamadas síncopas.
-
-As notas podem ocorrer dentro e fora destas divisões (da \emph{'métrica musical'}).
-Nos casos mais comportados, as notas ocorrem exatamente nestas divisões, com maior incidência para ataques nos tempos fortes.
-Em casos extremos, não se pode perceber a métrica~\cite{Roederer}.
-Variações (usualmente) bastante pequenas na grade compõem o que apreciamos como interpretação musical e diferenças entre estilos, junto a variações de intensidades, sobreposições de notas e timbre~\cite{Cook}.
-
-Com o fim de sistematizar e colocar em termos formais, podemos convencionar
+A acentuação em tempo fraco é o contratempo, as notas iniciadas em tempo fraco e cuja duração se prolonga sobre tempos fortes são as síncopas.
+
+As notas podem ocorrer dentro e fora destas divisões da \emph{'métrica musical'}.
+Nos casos mais comportados, as notas ocorrem exatamente nestas divisões, com maior incidência em ataques nos tempos fortes.
+Em casos extremos, não se pode perceber a métrica.\cite{Roederer} Variações pequenas na grade ajudam a compor a interpretação musical e diferenças entre estilos.\cite{Cook}
+
+Pode-se convencionar
 o pulso como nível $j=0$ de agrupamento, o nível $j=-1$ 
-é a primeira subdivisão do pulso, o nível $j=1$ como a primeira algomeração dos pulsos e assim por diante.
+a primeira subdivisão do pulso, o nível $j=1$ a primeira algomeração dos pulsos e assim por diante.
 
@@
 Especial atenção para
-os limites de $j$ pois as divisões do pulso devem resultar em durações apreciáveis
-como ritmo e também as junções do pulso devem resultar algo apreciável como, no máximo
-do nosso escopo, uma música ou um conjunto coeso de músicas. Dito de outra forma: a duração de $P^{min(j)}_i$, para $\forall \; i$
+os limites de $j$: as divisões do pulso são durações apreciáveis
+como ritmo; as junções do pulso somam, no máximo
+do escopo, uma música ou um conjunto coeso de músicas. Dito de outra forma: a duração de $P^{min(j)}_i$, $\forall \; i$
 é maior que $50ms$ e as durações de $P^{\text{máx}(j)}_i$ \emph{somadas}  em $\forall \; i$
@@
 
-Para cada nível $j$ temos alguns índices $i$. Sempre que $i$ soma três 
-(ou múltiplo de três) índices temos uma relação \emph{perfecta}, 
-quando soma dois, quatro ou oito índices temos uma relação \emph{imperfecta} (veja figura~\ref{fig:pulsoSubAgl}.
-
-Para especificar qualquer unidade (nota), de uma
-dada sequência musical que tenha métrica,
-podemos utilizar a notação:
+Cada nível $j$ possui alguns índices $i$. Sempre que $i$ soma três 
+(ou múltiplo de três) índices há uma relação \emph{perfecta}, 
+quando soma dois, quatro ou oito índices há uma relação \emph{imperfecta}, como na figura~\ref{fig:pulsoSubAgl}.
+
+Qualquer unidade (nota), de uma
+dada sequência musical que tenha métrica resulta
+assim especificada:
 
@@
 
-Como um exemplo, podemos escrever $P^{-1,0,2}_{3,2,1}$  como a terceira subdivisão $P^{-1}_3$ do segundo
-pulso $P^0_2$ do terceiro aglomerado de aglomerado de pulsos $P^2_2$ (perceba o termo $P^1_1$ omitido por simplicidade).
-
-Cada unidade ou conjunto de unidades $P_i^j$ pode ser associada a uma sequência de amostras temporais $T_i$ que formam uma nota musical.
-
-Assim como acontece com outras características
-musicais, lidamos com a estrutura rítmica sem fórmulas pre-estabelecidas, deixando procedimentos sistemáticos para quando é pertinente.
+Como um exemplo, $P^{-1,0,1}_{3,2,2}$  é a terceira subdivisão $P^{-1}_3$ do segundo
+pulso $P^0_2$ do segundo aglomerado de pulsos $P^1_2$.
+
+Cada unidade ou conjunto de unidades $P_i^j$ pode ser associada a uma sequência de amostras temporais $T_i$ que forma uma nota musical.
+
+A montagem \emph{Poli Hit Mia} utiliza as diferentes métricas e está no Apêndice~\ref{ap:poli} e disponível online como parte da \massa.
 
@@
 rítmicas (divisões e aglomerações simples, compostas e complexas) é 
-natural pensar em como realizar estas estruturas de forma que tenha coesão e sentido~\cite{Boulez}.
-
-Neste ponto, encontramos alguns artifícios consagrados pelo tempo e pelo uso. 
-Apresentamos aqui, de forma bastante breve, o pensamento musical dado por estruturas direcionais.
-Na parte seguinte exploramos as estruturas recorrentes e cíclicas.
-
-As estruturas direcionais nos possibiitam os recursos mais usuais: arcos melódicos e sequências que convergem ou divergem.
-
-Os arcos melódicos são compostos por duas sequências convergentes: 
-uma que atinge um ápice (chamado literalmente de clímax pela teoria musical tradicional) e 
-outra que, de forma paradigmática, volta do ápice à região da nota de partida.
-Canonicamente, distinguimos entre arcos cujos clímax estão: no começo, no meio, no final, na primeira metade e na segunda metade~\cite{Schoenberg}.
-Dispomos esquematicamente a disposição destes clímax na figura~\ref{fig:climax}.
+natural pensar em como realizar estas estruturas de forma que tenha coesão e sentido.\cite{Boulez}
+
+Para tal, os arcos são a base: partindo de algum lugar e voltando, forma-se um arco. Uma música sem nenhum arco pode ser compreendida como uma música sem coesão.
+Na subseção seguinte exploramos arcos cíclicos e neste
+os provenientes de estruturas direcionais.
+
+Os arcos podem ser decompostos em duas sequências convergentes: 
+uma que atinge o ápice (chamado literalmente de clímax pela teoria musical tradicional) e 
+outra que, de forma paradigmática, volta do ápice à região de partida.
+Canonicamente, distingue-se entre arcos cujos clímax estão: no começo, no meio, no final, na primeira metade e na segunda metade como esquematicamente a dispostos na figura~\ref{fig:climax}.\cite{Schoenberg}
 
@@
         \includegraphics[width=\textwidth]{figuras/climax}
-    \caption{Distinções canônicas do clímax musical em uma melodia ou arcos em outras escalas}
+    \caption{Distinções canônicas do clímax musical em uma melodia e outros domínios}
         \label{fig:climax}
@@
 $R_i=\{r_i\}_0^{2\Lambda_U -2}=\{u_{(\Lambda_U-1-|\Lambda_U-1-i|)}\}_0^{2\Lambda_U-2}$ 
-é uma sequência perfeitamente simétrica\footnote{Apresenta simetria especular perfeita, i.e.
-a segunda metade é uma versão espelhada da primeira.}.
+é uma sequência apresenta simetria especular perfeita, i.e.
+a segunda metade é uma versão espelhada da primeira.
 Pensada segundo conceitos musicais, o clímax está exatamente no meio da sequência.
-Podemos modificar isso fazendo com que alguma das metades seja expandida ou diminuída ou substituindo alguma das metades por outra sequência de tamanho diferente.
-É atraente este recurso pois contamos com toda a teoria de sequências, já bastante desenvolvida e estabelecida.
-Comparadas as sequências estudadas na teoria das sequências e as utilizadas na música, podemos concluir que existe um interesse especial por sequências relativamente curtas comparadas com as encontradas na teoria matemática~\cite{Guidorizzo,Schoenberg}.
-Teoricamente, estas sequências aplicadas a qualquer característica dos eventos musicais resulta em arcos.
-Assim, podemos conceber para a mesma sequência de eventos um certo número de arcos distintos, com tamanhos diferentes e climax diferentes para as características escolhidas.
-Embora este seja um recurso
-bastante interessante e útil, vale manter em mente que a correlação dos arcos resulta em coerência da escuta~\cite{Schachter}.
+Pode-se modificar isso com a substituição por sequências de tamanhos diferentes.
+Toda a teoria matemática de sequências, já estabelecida, pode ser utilizada para geração destes arcos.\cite{Guidorizzo,Schoenberg}
+Teoricamente, estas sequências aplicadas desta forma
+a qualquer característica dos eventos musicais resulta em arcos,
+pois resultam no afastamento e aproximação de uma parametrização inicial. 
+Assim, é possível para uma mesma sequência de eventos possuir um número de arcos distintos, com tamanhos e climax diferentes. Este é um recurso
+interessante e útil e a correlação dos arcos resulta na coerência da escuta.\cite{Schachter}
 
 \subsection{Estruturas cíclicas}\label{estCic}
-Os arcos musicais nem sempre são fruto de estruturas direcionais como as citadas
-acima. Um caso crítico é o de estruturas que se modificam gradualmente
-até atingirem o estado original, formando um arco cíclico.
-
-Neste caso, faremos uma exposição específica sobre permutações dada
-a completude desta abordagem segundo princípios musicais, matemáticos e
-filosóficos além da incidência musical histórica\footnote{De
-forma bastante resumida, apontamos o
-entendimento filosófico de que o pensamento humano se baseia
+
+O entendimento filosófico de que o pensamento humano se baseia
 na percepção de semelhanças e diferenças dentre os estímulos
-e objetos~\cite{Deleuze}. Portanto, segundo este entendimento,
-as simetrias se encontram no cerne do processo cognitivo pois conceitualmente são as equivalências em si. Matematicamente,
-as simetrias são representadas por grupos algébricos e os grupos finitos, por sua vez,
-são sempre isomorfos a um grupo de permutações. Desta forma, podemos dizer que
-as permutações são capazes de representar quaisquer simetrias em um sistema finito.
-Diretamente na música encontramos permutações de forma bastante explícita, inclusive
-na análise de música étnica~\cite{permMusic}.
-Na tradição musical a permutação está presente em diversas técnicas~\cite{Zamacois} e
-apontamos em especial a técnica \emph{Change Ringing} que consiste essencialmente
-em padrões de permutações dentre um fixado número de notas~\cite{change}.} e a consequente realização de dois trabalhos acadêmicos~\cite{figgusOriginal, figgusEspacializacao}.
-
-Qualquer conjunto de permutações pode ser utilizado como gerador de grupos algébricos~\cite{groups}. Assim, vale citarmos as propriedades que definem um grupo $G$:
+e objetos coloca
+as simetrias no cerne do processo cognitivo.\cite{Deleuze}
+Matematicamente,
+as simetrias são grupos algébricos e os grupos finitos
+são sempre isomorfos a um grupo de permutações. 
+Pode-se dizer que
+as permutações representam quaisquer simetrias em um sistema finito.
+Na música, as permutações estão são ubíquas
+e estão presentes em técnicas.
+A aplicação sucessivas das permutações gera arcos cíclicos.\cite{change,Zamacois,permMusic}
+Esta abordagem resultou em dois trabalhos acadêmicos para geração de estruturas musicais.\cite{figgusOriginal, figgusEspacializacao}
+
+Qualquer conjunto de permutações pode ser utilizado como gerador de grupos algébricos.\cite{groups} As propriedades que definem um grupo $G$ são:
 
@@
 
-A primeira destas propriedades nos é de particular importância, pois concluímos que toda permutação pode ser operada com outra permutação.
-De fato, podemos aplicar uma permutação, depois outra e, se compararmos com a ordenação inicial, temos uma permutação resultante.
-As outras propriedades resultam triviais para nosso caso.
-
-Alguns teoremas nos são caros.
-Por exemplo, segue imediatamente que, em um grupo finito, todo elemento $p$ operado consigo mesmo um número suficiente de vezes resulta no elemento neutro $p^n=e$, pois caso contrário a propriedade do fechamento resultaria em um grupo infinito gerado por $p$.
+Da primeira propriedade conclui-se que toda permutação pode ser operada com outra permutação.
+De fato, pode-se aplicar uma permutação, depois outra e, se comparadas as ordenações inicial e final, temos uma permutação resultante.
+
+Todo elemento $p$ operado consigo mesmo um número suficiente de vezes resulta no elemento neutro $p^n=e$, pois caso contrário a propriedade do fechamento resultaria em um grupo infinito gerado por $p$.
 O menor $n\;:\;p^n=e$ é chamado de ordem do elemento.
-Assim, uma permutação finita $p$ aplicada sucessivamente um número suficiente de vezes resulta na ordenação inicial dos elementos, formando um ciclo, i.e.
-um arco musical se os elementos ordenados são elementos musicais e se consultados em cada estágio.
-
-Estes arcos, que levam uma ordenação de elementos na mesma ordenação após um certo número de aplicações de permutações, podem ser efetuados pelo uso conjunto de diferentes permutações. Como exemplo
+Assim, uma permutação finita $p$ aplicada sucessivamente resulta na ordenação inicial dos elementos, formando um ciclo.
+Um arco musical se consultados os elementos em cada estágio.
+
+Estes arcos podem ser efetuados pelo uso conjunto de diferentes permutações. Como exemplo
 histórico, a tradição chamada \emph{change ringing} concebe música através de sinos tocados um após o outro e então tocados novamente, mas em uma ordem diferente.
-Este processo se repete até que se atinja a ordenação inicial e o conjunto de ordenações diferentes percorridas consiste no que é chamado de um \emph{peal}.
-Dispomos na tabela~\ref{tab:change} um \emph{peal} tradicional de 3 sinos (1, 2 e 3), que explora todas as suas ordenações.
+Este processo se repete até que se atinja a ordenação inicial.
+O conjunto de ordenações diferentes percorridas é um \emph{peal}.
+A tabela~\ref{tab:change} dispoe um \emph{peal} tradicional de 3 sinos (1, 2 e 3) que explora todas as suas ordenações.
 Cada linha apresenta uma ordenação dos sinos a ser tocada.
-As permutações agem aqui entre as linhas.
-% {\color{blue} AA}
+As permutações estão entre as linhas.
 
@@
 
-Para maiores detalhes desta tradição, recomendamos~\cite{change}.
-Importante para nós aqui é notar a existência de tradições musicais consciêntes da importância das permutações na música e observar que, neste caso, a estrutura musical consiste em permutações propriamente ditas e algumas permutações diferentes estão operando.
-Assim, fica claro o comportamento cíclico causado pela aplicação sucessiva da mesma permutação
-e a consistência musical deste procedimento.
+Neste caso, a estrutura musical consiste em permutações propriamente ditas e algumas permutações diferentes estão operando para resultar em um comportamento cíclico.
 
 A utilização de permutações na música pode ser resumida da seguinte forma:
-seja $L_i$ uma sequência de eventos musicais (para simplificar, considere
-estes eventos como sendo notas musicais) e $p$ uma permutação.
+seja $L_i$ uma sequência de eventos musicais (e.g. notas) e $p$ uma permutação.
 $L_i'=p(L_i)$ consiste nos mesmos elementos de $L_i$ com a ordenação
 modificada.
-
 As permutações são usualmente escritas em notação cíclica ou natural. 
-Por brevidade, vamos abordar somente a notação natural.
-Esta notação consiste na ordem dos índices originais considerados resultante da aplicação da permutação.
-Assim,
+Esta notação consiste na ordem dos índices 
+resultante da permutação. Assim,
 convencionada a ordenação original dada pela sequência de seus índices $[0\;1\;2\;3\;4\;5\;...]$ a permutação é notada pela sequência que produz (ex. $[1\;3\;7\;0\;...]$).
 
-Importante notar que não é necessário permutar os elementos de $L_i$, mas somente
-alguma ou algumas de suas características. Assim, seja $p^f$ uma permutação agindo
+Não é necessário permutar os elementos de $L_i$, mas somente
+alguma ou algumas de suas características. Assim, seja $p^f$ uma permutação 
 nas frequências e $L_i$ uma sequência de notas básicas como expostas
-ao final de ~\ref{notaBasica}. A nova sequência $L_i'=p^f(L_i)$ consiste nas mesmas
-notas musicais, na mesma ordem e com as mesmas características, com as frequências fundamentais permutadas segundo
-o padrão que $p^f$ apresenta.
+ao final de~\ref{notaBasica}. A nova sequência $L_i'=p^f(L_i)$ consiste nas mesmas
+notas musicais, na mesma ordem e com as mesmas características, com as frequências fundamentais permutadas segundo o padrão que $p^f$ apresenta.
 
@@
 Primeiro, a permutação $p$ não precisa envolver todos os elementos de $L_i$, i.e. ela
-pode operar em subconjuntos de $L_i$. Segundo, nem todos os elementos $l_i$ precisam ser executados a
-cada consulta de estado realizada.
-
-Para simplificar, considere novamente os eventos de $L_i$ 
-como notas musicais. Assim, se $i$ é vai de $0$ a $n$, e $n>4$, a cada compasso
-de $4$ notas podemos executar somente as primeiras $4$ notas. As outras notas de
-$L_i$ podem ser incidentes nos compassos em que as permutações aloquem
+pode operar em subconjuntos de $L_i$. Segundo, nem todos os elementos $l_i$ precisam ser executados a cada consulta de estado realizada.
+
+Para exemplificar, seja $L_i$ 
+uma sequência de notas musicais $l_i$. 
+Se $i$ é vai de $0$ a $n$, e $n>4$, a cada compasso
+de $4$ notas pode-se executar as primeiras $4$ notas. As outras notas de
+$L_i$ podem incidir nos compassos em que as permutações aloquem
 estas notas para as primeiras quatro notas de $L_i$.
 
-Podemos dispor permutações que operem no mesmo conjunto, cada uma destas permutações
-$p_i$ possui ao menos, segundo nossa exposição acima, vinculadas à permutação em si: dimensões das notas em que opera (frequência, duração, fades, intensidade, etc) e período de incidência (a cada quantas consultas é aplicada a permutação).
-Na realização das notas de $L_i$, uma solução fácil e coerente é executar as primeiras $n$ notas\footnote{A execução de notas disjuntas de $L_i$ equivale a modificar a permutação e executar as primeiras notas.}.
-
-Esta abordagem resultou em alguns trabalhos~\cite{figgusOriginal,figgusEspacializacao} e implementações computacionais~\cite{repoFIGGUS,repoFIGGUSold}.
-Descrevemos um uso coerente desta técnica abaixo e apresentamos no APÊNCIDE~\ref{cap:FIGGUScode} a implementação computacional disponibilizada em~\cite{repoFIGGUS}
+Cada uma destas permutações
+$p_i$ possui, ao menos, segundo a exposição acima: dimensões das notas em que opera (frequência, duração, fades, intensidade, etc) e período de incidência (a cada quantas consultas é aplicada a permutação).
+Na realização das notas de $L_i$, uma forma fácil e coerente é executar as primeiras $n$ notas\footnote{A execução de notas disjuntas de $L_i$ equivale a modificar a permutação e executar as primeiras notas.}.
+
+Esta abordagem resultou em alguns trabalhos e implementações computacionais.\cite{repoFIGGUS,repoFIGGUSold,figgusOriginal,figgusEspacializacao} Descrevemos um uso coerente desta técnica abaixo e apresentamos no Apêndice~\ref{cap:FIGGUScode} a implementação computacional disponibilizada em~\cite{repoFIGGUS}.
 
@@
 
-De certa forma central à experiência musical é a concepção de 'idioma musical'.
-Existem diversas empreitadas para modelar e explorar diferentes entendimentos
-sobre a linguagem musical e linguística aplicada à música e para discernimento entre 
+Existem diversas empreitadas que se propõem a modelar e explorar entendimentos
+sobre a linguagem musical e da linguística aplicada à música ou ainda
+para discernimento entre 
 o que seriam diferentes
-'idiomas musicais'\cite{Lerdahl, Harmonia, Schachter,Alfaix}.
-
-Apontamos aqui que, de forma simplificada e resumida, um idioma musical é fruto da
-repetição de elementos e da repetição de relações entre elementos presentes no decorrer da música\footnote{A
-música é tradicionalmente pensada através de dicotomias bastante relacionadas umas com as outras,
-como: repetição e variação, relaxamento e tensão, equilíbrio e desequilíbrio, consonância e dissonância, etc}.
+'idiomas musicais'.\cite{Lerdahl, Harmonia, Schachter,Alfaix}
+
+De forma simplificada, um idioma musical é fruto da
+repetição de elementos e da repetição de relações entre elementos presentes no decorrer da música. Bastante relacionado é o uso de dicotomias 
+como: repetição e variação, relaxamento e tensão, equilíbrio e desequilíbrio, consonância e dissonância, etc.
 
@@
 Primeiro, a nota básica foi definida e suas características postas em termos
-claros e quantitativos (sessão~\ref{notaBasica}). Em seguida, adentramos a nota musical e suas transições internas,
-além de tratamentos imediatos (sessão~\ref{varInternas}). Por fim, nesta última sessão, apresentamos
-recursos consagrados para organizar estas notas musicais em música.
-
-Assim, atingimos uma gama de recursos com uma infinidade de possibilidades de resultados,
-situação típica e tão cara às artes~\cite{Harmonia,Webern}.
-
-Existem explorações e estudos individuais para cada recurso apresentado.
+claros e quantitativos (sessão~\ref{notaBasica}).
+Em seguida, a nota foi adentrada e foram compreendidas suas transições internas e de tratamentos imediatos (sessão~\ref{varInternas}).
+Por fim, esta última sessão dedica-se a organizar estas notas musicais em música.
+A gama de recursos com uma infinidade de possibilidades de resultados,
+é situação típica e cara às artes.\cite{Harmonia,Webern}
+
+Existem estudos para cada recurso apresentado.
 Por exemplo, pode-se obter as harmonias triádicas 'sujas' (com notas não pertencentes à tríade) através de sobreposições de quartas justas.
@@
 A montagem musical \emph{Poli-hit mia} explora estas métricas simultâneas através de trem de impulsos convoluídos com as notas que compõem cada linha.
-
-As escalas microtonais são bastante importantes na música do século XX~\cite{microtonalidade} e possuem vários usos com resultados muito marcantes, como os usos de quartos de tom na música indiana e possibilidades de batimentos.
-A sequência musical \emph{MicroTom} explora estes recursos de forma explícita, incluindo melodias microtonais e harmonias microtonais com várias notas em um âmbito de alturas bastante reduzido.
-
-Novamente, vínculos entre parâmetros são formas poderosas de resultar músicas ou ao menos montagens musicais.
-O número de notas permutadas podem variar no decorrer da música, revelando vínculo com a duração da peça.
+Seu código está no Apêndice~\ref{ap:poli} e disponível online como parte da \massa.
+
+As escalas microtonais são importantes na música do século XX~\cite{microtonalidade} e possuem usos com resultados muito marcantes, como os quartos de tom na música indiana.
+A sequência musical \emph{MicroTom} explora estes recursos de forma dedicada, incluindo melodias microtonais e harmonias microtonais com várias notas em um âmbito de alturas bastante reduzido.
+Seu código está no Apêndice~\ref{ap:micro} e disponível online como parte da \massa.
+
+Os vínculos entre parâmetros são formas poderosas de resultar peças e montagens musicais.
+O número de notas permutadas pode variar no decorrer da música, revelando vínculo com a duração da peça.
 As harmonias podem consituir-se triádicas (eqs.~\ref{triades}) com notas replicadas em várias oitavas e mais numerosas quanto menor a profundidade e frequência de vibratos (eqs.~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}), dentre outras incontáveis possibilidades.
@@
 Nas peças \emph{3 trios} esta associação é feita de forma sistemática para possibilitar uma audição dedicada a esta associação.
-
-O \emph{PPEPPS} (Pure Python EP: Projeto Solvente) é um EP inteiro sintetizado com os recursos apresentados neste trabalho.
+Esta é uma peça instrumental e não consta dentre os códigos dos apêndices e da \massa.
+
+O \emph{PPEPPS} (Pure Python EP: Projeto Solvente) é um EP sintetizado com os recursos apresentados neste trabalho.
 Com pouca parametrização, o programa resulta músicas inteiras, permitindo a composição de músicas e conjuntos de músicas com facilidade.
-As possibilidades de compartilhamento envolvem obter este album (ou outro feito de forma semelhante) baixando algumas linhas de código e rodando para obter uma pasta com as músicas.
+As possibilidades de compartilhamento envolvem obter este album (ou outro feito de forma semelhante) através depoucas linhas de código e, pela execução, obter uma pasta com as músicas.

### 2012-11-20 conclusoes.tex
terminada revisao completa

@@
 
-Os resultados apresentados no capítulo anterior estão bastante de acordo
-com os objetivos: conseguimos estabelecer um sistema conciso
-que relaciona elementos musicais ao som digital. Dispomos também um conjunto
-de scripts que implementam estas relações, ao qual chamamos MASSA (Música
-e Áudio em Sequências e Séries Amostrais). Talvez em último lugar, mas igualmente
-importante, realizamos uma exposição destes desenvovimentos de forma didática no
-capítulo anterior.
+No capítulo anterior está um sistema conciso
+que relaciona elementos musicais ao som digital. Um conjunto
+de scripts implementam estas relações, em conjunto nomeadas \massa (Música
+e Áudio em Sequências e Séries Amostrais). 
+A exposição didática destes desenvovimentos no
+capítulo anterior destina-se a facilitar a utilização
+do arcabouço.
+
+As possibilidades abertas por estes resultados envolvem a criação de interfaces de geração de ruídos e outros sons em alta fidelidade (hi-fi), experimentos psicoacústicos e a utilização destes resultados para fins artísticos e didáticos.
+A incorporação de conhecimentos
+em programação é bastante facilitada através de recursos audiovisuais, o que já realizamos por práticas de \emph{livecoding} e cursos focados em ferramentas especializadas, como o Puredata e o ChucK.
+
+Está prevista a utilização destes resultados com
+métodos generativos para geração de materiais artísticos.
 
 Este trabalho também teve resultados não previstos, como a formação de grupos
-de interesse em torno da questão criativa impulsionada por ferramentas computacionais,
-em especial pelo código de programação. Neste contexto, se destaca o grupo
+de interesse em torno da questão criativa aliada à computação.
+Neste contexto, se destaca o grupo
 labMacambira.sf.net, que reúne colaboradores de todo o Brasil e alguns fora do país.
@@
 
-As possibilidades de trabalhos futuros abertas por estes resultados envolvem a criação de interfaces de geração de ruídos e outros sons em alta fidelidade (hi-fi), o auxílio na realização de experimentos psicoacústicos e, principalmente, a utilização destes resultados para fins artísticos e didáticos.
-A incorporação da linguagem de programação é bastante facilitada através de recursos audiovisuais, o que já realizamos por práticas de \emph{livecoding} e cursos focados em ferramentas especializadas, como o Puredata e o ChucK, e está em muito potencializada e confirmada através dos resultados descritos neste trabalho.
-
-A disposição online destes conteúdos na forma de hipertexto junto aos códigos e exemplos sonoros, todos em licenças livres, não só facilita colaborações e geração de subprodutos em co-autoria, mas também que nossas contribuições tenham usos e se desdobrem de formas que ainda não concebemos.
+A disposição online destes conteúdos na forma de hipertexto junto aos códigos e exemplos sonoros, todos em licenças livres, facilita colaborações e geração de subprodutos em co-autoria, e também permite que esta contribuição tenha uso e se desdobrem.

### 2012-11-25 introducao.tex
commit geral antes de correcoes do prof. chu

@@
 é exatamente a tarefa proposta por este trabalho.
-Os resultados estão em relações matemáticas e implementações computacionais.
-Uma descrição está no capítulo~\ref{cap:resultados} e o conjunto de \emph{scripts} disponibilizados no Apêndice~\ref{cap:codigoProc} e online.\cite{MASSA} O \emph{toolbox} recebeu o nome \massa\ (música e áudio em sequências e séries amostrais) e foi utilizado para fazer pequenas peças e montagens focadas nos princípios expostos.
+Os resultados são relações matemáticas e suas implementações computacionais.
+Uma descrição está no capítulo~\ref{cap:resultados} e o conjunto de \emph{scripts} disponibilizados no Apêndice~\ref{cap:codigoProc} e online.\cite{MASSA} A \emph{toolbox} resultante recebeu o nome \massa\ (música e áudio em sequências e séries amostrais) e foi utilizado para fazer pequenas peças e montagens focadas nos princípios expostos.
 O Apêndice~\ref{cap:codigoPecas} possui uma relação destas montagens assim como o diretório \emph{exemplos\_de\_uso} da \massa.
@@
 O som é uma onda mecânica longitudinal de pressão.
-A banda de frequências compreendida entre $20Hz$ e $20 kHz$ é apreciada pelo nosso aparelho auditivo e possui variações dependentes da pessoa, das condições climáticas e do som em si.
- Considerada a velocidade do som no ar de $\approx 343.2\,m/s$,
+A banda de frequências compreendida entre $20Hz$ e $20 kHz$ é apreciada pelo aparelho auditivo humano com variações dependentes da pessoa, das condições climáticas e do som em si.
+ Considerada a velocidade do som no ar $\approx 343.2\,m/s$,
 estes limites correspondem respectivamente aos comprimentos de onda $\frac{343.2}{20} = 17.16\,m$ e $\frac{343.2}{20000}=17.16\,mm$.\cite{Roederer}
@@
 A percepção humana do som envolve captações pelos ossos, estômago e orelha, funções de transferência da cabeça e dorso e processamento pelo sistema nervoso.
-Existe, além disso, um órgão dedicado à tarefa de captura destas ondas, ao qual chamamos ouvido e cujo funcionamento decompõe o som em seu espectro senoidal e passa para o sistema nervoso.\cite{Roederer} Estas componentes senoidais são cruciais para os fenômenos musicais, como pode-se observar na composição dos sons de interesse para a música quanto nas afinações e escalas.\cite{floEsp} A subseção~\ref{sec:notaDisc} expõe a presença de senóides no som discretizado e caracteriza a nota musical básica.
+Além disso, o ouvido é um órgão dedicado à tarefa de captura destas ondas.
+Seu funcionamento decompõe o som em seu espectro senoidal e passa para o sistema nervoso.\cite{Roederer} Estas componentes senoidais são cruciais para os fenômenos musicais, como pode-se observar na composição dos sons de interesse para a música quanto nas afinações e escalas.\cite{floEsp} A subseção~\ref{sec:notaDisc} expõe a presença de senóides no som discretizado e caracteriza a nota musical básica.
 
-A representação do som é chamada áudio\footnote{Os termos
-som e áudio são muitas vezes usados de forma intercambiável.\cite{Everest}} e este pode ser proveniente da captura do som por microfones ou da síntese.
-Muitas vezes, o áudio digital é obtido através protocolos que facilitam o armazenamento e transferência dos arquivos.\cite{procFala} Em outros momentos, a representação digital do som consiste em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
+A representação do som é o áudio\footnote{Os termos
+som e áudio são muitas vezes usados de forma intercambiável.\cite{Everest}} e este pode provir da captura do som por microfones ou da síntese.
+Muitas vezes, o áudio digital é obtido através de protocolos que facilitam o armazenamento e transferência dos arquivos.\cite{procFala} Em outros momentos, a representação digital do som consiste em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
 Estas amostras separadas por intervalos regulares $\lambda_a$ constituem a forma padrão de representação do som em tempo discreto chamada de modulação por código de pulsos (PCM do inglês \emph{Pulse Code Modulation}).
@@
 
-A música é definida canonicamente como a arte manifesta pelos sons e silêncios.
+A música é definida como a arte manifesta pelos sons e silêncios.
 Para um ouvinte comum - e boa parte dos especialistas - uma 'música que seja música' pressupõe também a presença de uma métrica rítmica e organizações de alturas que formem melodias e harmonias como explicadas na sessão~\ref{notasMusica}.
@@
    \label{sec:objetivos}
-O objetivo principal desta dissertação é apresentar um sistema unificado que relacione elementos básicos da música às sequências amostrais do áudio PCM.
-Assim, o capítulo seguinte é um texto conciso em que os elementos musicais são apresentados junto às amostras temporais resultantes.
+O objetivo principal desta dissertação é apresentar de forma unificada relações entre elementos básicos da música e as sequências amostrais do áudio PCM.
+O capítulo seguinte é um texto conciso em que os elementos musicais são apresentados junto às amostras temporais resultantes.
 Para validação e compartilhamento, as implementações em código computacional destas relações e de pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais.} foram reunidas em uma \emph{toolbox} chamada \massa\ e disponibilizadas online e nos Apêndices~\ref{cap:codigoProc} e~\ref{cap:codigoPecas}.
 
-Dada esta direção bem estabelecida, consideramos alguns objetivos secundários.
-Em primeiro lugar, visamos a difusão da compreensão do código computacional através de práticas lúdicas, no caso a música.
-Outro objetivo considerado é a apresentação de um arcabouço de síntese sonora e musical com controle amostral, para o qual encontramos potenciais usos em experimentos psico-acústicos e síntese em alta definição (\emph{hi-fi}).
-Talvez por último, buscamos apresentar estes conteúdos de forma didática, quase um verdadeiro tutorial, possibilitando uma compreensão e uso facilitados, pois os assuntos tratados são de reconhecida complexidade: processamento de sinais, música, psico-acústica, para citar somente alguns exemplos.
-Deste ponto de vista pedagógico, também reconhecemos importante a apresentação destes resultados na forma de hipertexto, em que cada \emph{script} e exemplo sonoro/musical seja acessível junto ao material teórico.
+Dada esta direção central, consideramos alguns objetivos secundários.
+Destes, destaca-se a difusão da compreensão do código computacional através de práticas lúdicas, no caso a música.
+Outro objetivo considerado é a apresentação de um arcabouço de síntese sonora e musical com controle amostral, para o qual há potenciais usos em experimentos psico-acústicos e síntese em alta definição (\emph{hi-fi}).
+Também é considerada a apresentação destes conteúdos de forma didática, quase um tutorial, possibilitando uma compreensão e uso facilitados, pois os assuntos tratados são de reconhecida complexidade: processamento de sinais, música, psico-acústica, para citar somente alguns exemplos.
+Deste ponto de vista pedagógico, também se presta a apresentação destes resultados na forma de hipertexto, em que cada \emph{script} e exemplo sonoro/musical seja acessível junto ao material teórico.

### 2012-11-25 conclusoes.tex
commit geral antes de correcoes do prof. chu

@@
 No capítulo anterior está um sistema conciso
-que relaciona elementos musicais ao som digital. Um conjunto
-de scripts implementam estas relações, em conjunto nomeadas \massa (Música
+que relaciona elementos musicais ao som digital. Scripts 
+implementam estas relações, e em conjunto foram nomeados \massa\ (Música
 e Áudio em Sequências e Séries Amostrais). 
@@
 em programação é bastante facilitada através de recursos audiovisuais, o que já realizamos por práticas de \emph{livecoding} e cursos focados em ferramentas especializadas, como o Puredata e o ChucK.
-
 Está prevista a utilização destes resultados com
 métodos generativos para geração de materiais artísticos.
+
+A disposição online destes conteúdos na forma de hipertexto junto aos códigos e exemplos sonoros, todos em licenças livres, facilita colaborações e geração de subprodutos em co-autoria, e com isso a expansão da \massa com novas implementações e desenvolvimentos das montagens musicais.
+
+Explorações sistemáticas de parametrizações (dos tremolos, da ADSR, etc) em alta fidelidade tem utilidade artística e é possibilidata por este trabalho com controle amostral.
 
@@
 deste trabalho, e já apresentou contribuições relevantes em diferentes áreas
-como Democracia Direta Digital, ferramentas digitais de georeferenciamento e
+como Democracia Direta Digital, ferramentas de georeferenciamento e
 atividades artísticas e educacionais, como cursos, workshops e apresentações artísticas.
 Vários destes resultados estão apresentados no Apêndice~\ref{cap:musicaExtra} e no acervo online criado, que ultrapassa 700 videos, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos~\cite{siteLM,wikiLM,vimeoLM}.
-
-A disposição online destes conteúdos na forma de hipertexto junto aos códigos e exemplos sonoros, todos em licenças livres, facilita colaborações e geração de subprodutos em co-autoria, e também permite que esta contribuição tenha uso e se desdobrem.

### 2012-11-26 introducao.tex
adiantada introducao, indo para cap 2 jah

@@
 Representar estruturas e artifícios musicais através das características do som discretizado
-é exatamente a tarefa proposta por este trabalho.
+é a tarefa proposta por este trabalho.
 Os resultados são relações matemáticas e suas implementações computacionais.
-Uma descrição está no capítulo~\ref{cap:resultados} e o conjunto de \emph{scripts} disponibilizados no Apêndice~\ref{cap:codigoProc} e online.\cite{MASSA} A \emph{toolbox} resultante recebeu o nome \massa\ (música e áudio em sequências e séries amostrais) e foi utilizado para fazer pequenas peças e montagens focadas nos princípios expostos.
+Uma descrição teórica está no capítulo~\ref{cap:resultados} e o conjunto de \emph{scripts} disponibilizados no Apêndice~\ref{cap:codigoProc} e online.\cite{MASSA} A \emph{toolbox} resultante recebeu o nome \massa\ (música e áudio em sequências e séries amostrais) e foi utilizada para fazer pequenas peças e montagens focadas nos princípios expostos.
 O Apêndice~\ref{cap:codigoPecas} possui uma relação destas montagens assim como o diretório \emph{exemplos\_de\_uso} da \massa.
@@
 Além disso, o ouvido é um órgão dedicado à tarefa de captura destas ondas.
-Seu funcionamento decompõe o som em seu espectro senoidal e passa para o sistema nervoso.\cite{Roederer} Estas componentes senoidais são cruciais para os fenômenos musicais, como pode-se observar na composição dos sons de interesse para a música quanto nas afinações e escalas.\cite{floEsp} A subseção~\ref{sec:notaDisc} expõe a presença de senóides no som discretizado e caracteriza a nota musical básica.
+Seu funcionamento decompõe o som em seu espectro senoidal e passa para o sistema nervoso.\cite{Roederer} Estas componentes senoidais são cruciais para os fenômenos musicais, como se pode observar na composição dos sons de interesse para a música quanto nas afinações e escalas.\cite{floEsp} A subseção~\ref{sec:notaDisc} expõe a presença de senóides no som discretizado e caracteriza a nota musical básica.
 
@@
 som e áudio são muitas vezes usados de forma intercambiável.\cite{Everest}} e este pode provir da captura do som por microfones ou da síntese.
-Muitas vezes, o áudio digital é obtido através de protocolos que facilitam o armazenamento e transferência dos arquivos.\cite{procFala} Em outros momentos, a representação digital do som consiste em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
-Estas amostras separadas por intervalos regulares $\lambda_a$ constituem a forma padrão de representação do som em tempo discreto chamada de modulação por código de pulsos (PCM do inglês \emph{Pulse Code Modulation}).
+Muitas vezes, o áudio digital é obtido através de protocolos que facilitam o armazenamento e transferência dos arquivos.\cite{procFala} A representação digital do som pode consistir em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
+Estas amostras separadas por intervalos regulares $\lambda_a$ constituem a forma padrão de representação do som em tempo discreto, chamada de modulação por código de pulsos (PCM do inglês \emph{Pulse Code Modulation}).
 Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$, também chamada de taxa de amostragem, e a profundidade de bit que é o número de bits utilizados para representar a amplitude de cada amostra.\cite{protocolosAudio} A figura~\ref{fig:PCM} exibe $25$ amostras de um áudio PCM com $4$ bits cada.
 Os $2^4=16$ grados para a amplitude de cada amostra junto ao espaçamento regular $\lambda_a$ introduzem um erro de quantização.
-O ruído causado por estes erros diminuem junto à diminuição estes espaçamentos.
+O ruído causado por estes erros diminuem com a diminuição destes espaçamentos.
 
@@
 A música é definida como a arte manifesta pelos sons e silêncios.
-Para um ouvinte comum - e boa parte dos especialistas - uma 'música que seja música' pressupõe também a presença de uma métrica rítmica e organizações de alturas que formem melodias e harmonias como explicadas na sessão~\ref{notasMusica}.
+Para um ouvinte comum - e boa parte dos especialistas - uma 'música que seja música' pressupõe também uma métrica rítmica e organizações de alturas que formem melodias e harmonias como explicadas na seção~\ref{notasMusica}.
 A música do século XX ampliou esta concepção tradicional de música.
@@
 Mesmo assim, a nota permanece paradigmática como 'unidade fundamental' das estruturas musicais e, na prática, podem se desdobrar em sons que contemplam estes desenvolvimentos recentes.
-Adentramos a nota musical e sua expansão na sessão~\ref{varInternas}.\cite{Wisnick,Webern,Lerdahl,Cook}
+A nota musical e sua expansão são abordadas na seção~\ref{varInternas}.\cite{Wisnick,Webern,Lerdahl,Cook}
 
@@
 Liberamos este conjunto de códigos em domínio público através de repositórios Git públicos.\cite{gitBook}
-
 Os \emph{scripts} estão em Python e fazem uso das bibliotecas externas Numpy e Scikits/Audiolab que realizam chamadas à linguagem Fortran para maior eficiência computacional.
@@
 estão publicadas em licenças que permitem o uso, cópia, distribuição e utilização de quaisquer partes para estudo e geração de produtos derivados.
-Desta forma, o trabalho aqui descrito está disponível como parte de um patrimônio tecnológico da humanidade\footnote{A comunidade e movimento chamada \emph{'Open Source'} entende a publicação de código computacional (e outras tecnologias) em licenças abertas como uma vantagem pragmática que facilita o desenvolvimento de \emph{software} e apresenta vantagens de pedagógicas e mercadológicas.
+Desta forma, o trabalho aqui descrito está disponível e facilita os processos de co-autoria\footnote{A comunidade e movimento chamada \emph{'Open Source'} entende a publicação de código computacional (e outras tecnologias) em licenças abertas como uma vantagem pragmática que facilita o desenvolvimento de \emph{software} e apresenta vantagens de pedagógicas e mercadológicas.
 A comunidade e movimento chamada \emph{'Free Software'} engloba este entendimento, mas adiciona a abordagem filosófica da liberdade e compartilhamento, dando ênfase a isso.
-Ambas as correntes reforçam o entendimento de que o código computacional é o 'bem mais precioso produzido atualmente' pois consiste em tecnologia condensada, reativa (executa, processa ou gera resultados), modular (partes são copiadas e reutilizadas eficientemente) e replicada sem custo adicional (a cópia de texto tem custo baixíssimo).\cite{Raymond,Lessig}} e facilita os processos de co-autoria.
+Ambas as correntes reforçam o entendimento de que o código computacional é o 'bem mais precioso produzido atualmente' pois consiste em tecnologia condensada, reativa (executa, processa ou gera resultados), modular (partes são copiadas e reutilizadas eficientemente) e replicada sem custo adicional (a cópia de texto tem custo baixíssimo).\cite{Raymond,Lessig}}.
 

### 2012-11-26 scripts/pcm.py
adiantada introducao, indo para cap 2 jah

@@ -1,5 +1,5 @@
 #-*- coding: utf8 -*-
-import numpy as n, pylab as p
+import numpy as n, pylab as p, string
 
 # plotando um som em PCM de 4 bits de profundidade
 
@@ -12,12 +12,18 @@ b=n.sin(aa)+n.sin(4*aa)
 b*=.5
 b*=2**3-1 # 4 bits
 
-p.plot(aa,b,"g",lw=6)
+p.plot(aa,b,"g",lw=6,label=u"sinal analógico")
 
 c=n.round(b)
-p.plot(aa[::10],c[::10],'ro',ms=9)
+p.plot(aa[::10],c[::10],'ro',ms=9,label="amostras digitais")
 
-p.yticks(range(-7,8), fontsize=18)
+def bits(i,n):
+    return tuple(("0","1")[i>>j & 1] for j in xrange(n-1,-1,-1)) 
+
+foo=[bits(i,4) for i in xrange(2**4)]
+bar=[string.join(ii,"") for ii in foo]
+p.yticks(range(-7,9),bar, fontsize=18)
+
 xlabs=[r"$\lambda_a$"]*len(aa[::10])
 xlabs=[str(i)+xlabs[i] for i in xrange(len(xlabs))]
 xlabs[0]=0
@@ -25,6 +31,8 @@ xlabs[1]=r"$\lambda_a$"
 p.xticks((aa[::10]),xlabs)
 
 p.xlim(-.3,aa[::10][-1]+0.5)
+
+p.legend(prop={'size':26})
 
 p.xlabel(r"tempo $\rightarrow$",fontsize=20)
 p.ylabel(r"amplitude $\rightarrow$",fontsize=24)

### 2012-11-26 introducao.tex
final da parte 2.2

@@
 estão publicadas em licenças que permitem o uso, cópia, distribuição e utilização de quaisquer partes para estudo e geração de produtos derivados.
-Desta forma, o trabalho aqui descrito está disponível e facilita os processos de co-autoria\footnote{A comunidade e movimento chamada \emph{'Open Source'} entende a publicação de código computacional (e outras tecnologias) em licenças abertas como uma vantagem pragmática que facilita o desenvolvimento de \emph{software} e apresenta vantagens de pedagógicas e mercadológicas.
+Desta forma, o trabalho aqui descrito está disponível e facilita os processos de co-autoria\footnote{A comunidade e movimento chamada \emph{'Open Source'} entende a publicação de código computacional (e outras tecnologias) em licenças abertas como uma vantagem pragmática que facilita o desenvolvimento de \emph{software} e apresenta vantagens pedagógicas e mercadológicas.
 A comunidade e movimento chamada \emph{'Free Software'} engloba este entendimento, mas adiciona a abordagem filosófica da liberdade e compartilhamento, dando ênfase a isso.

### 2012-11-26 devEresultados.tex
final da parte 2.2

@@
 que facilitam e enriquecem o fluxo de informação através da música.\cite{Roederer, Lacerda}
-
 Canonicamente, as notas possuem ao menos duração, volume, altura e timbre.\cite{Lacerda} Estas
-são qualidades apreciadas por leis psicofísicas e pode-se lidar com elas através de caracterizações físicas.\cite{Roederer}
-Assim, as características da nota musical digital básica é dada pelas amostras igualmente espaçadas no tempo
-da onda mecânica que representam (veja sessão~\ref{sec:audio} sobre áudio PCM).
+são qualidades que podem ser tratadas quantitativamente.\cite{Roederer}
+Assim, as características da nota musical digital básica são dadas pelas amostras igualmente espaçadas no tempo
+da onda mecânica que representam (veja seção~\ref{sec:audio} sobre áudio PCM).
 
@@
 se apresenta como uma sequência de $ \lfloor \Delta . f_a \rfloor $ amostras\footnote{O
-limite superior de uma sequência é um número natural mas $ \Delta . f_a $
+limite superior de uma sequência é um número natural, mas $ \Delta . f_a $
 só satisfaz esta condição em casos muito excepcionais. É necessário
@@
 \subsection{Volume}
-A sensação de volume sonoro depende da reverberação e distribuição dos harmônicos, dentre outras características trabalhamos na seção~\ref{varInternas}, e podemos obter variações do volume através da potência da onda~\cite{Chowning}:
+A sensação de volume sonoro depende da reverberação e distribuição dos harmônicos, dentre outras características trabalhadas na seção~\ref{varInternas}, e podemos obter variações do volume através da potência da onda~\cite{Chowning}:
 
@@
 
-Como o volume final dependerá da amplificação do sinal nos auto-falantes, o crucial é a potência relativa de uma nota em relação às outras ou de um trecho da música em relação ao resto.
+Como o volume final dependerá da amplificação do sinal nos alto-falantes, o crucial é a potência relativa de uma nota em relação às outras ou de um trecho da música em relação ao resto.
 As diferenças de volume são medidas em decibels, e estes são
@@
 
-A quantidade $V_{dB}$ possui a unidade decibel ($dB$). A cada 10 $dB$ se atribui canonicamente
+A quantidade $V_{dB}$ possui a unidade decibel ($dB$). A cada 10 $dB$ se atribui
 a sensação de "volume dobrado". Importantes resultados de referência são os equivalentes em $dB$ de se dobrar
@@
 Ou seja, é necessário pouco mais que triplicar a amplitude para um volume drobrado.
-
 Estes valores servem de guia para os aumentos e diminuições dos valores absolutos que compõem as
@@
 
-Neste ponto, a partícula musical (nota) é uma sequência $T_i$ cuja duração e volume correspondem ao tamanho da sequência e amplitude de suas amostras.
+Recapitulando, a partícula musical (nota) é uma sequência $T_i$ cuja duração e volume correspondem ao tamanho da sequência e amplitude de suas amostras.
 A altura é especificada pela frequência fundamental $f_0$ com ciclo de duração $\delta_{f_0} = 1/f_0$.
@@
 Por motivos didáticos, seja $f_0$ tal que divida $f_a$ e $\lambda_{f_0}$ resulte inteiro.
-Se $T_i^f$ é uma sequência sonora de frequência fundamental $f$, resulta evidente que:
+Se $T_i^f$ é uma sequência sonora de frequência fundamental $f$, então:
 
@@
 Para citar também um uso possível da onda quadrada, ela pode ser usada em uma síntese
-subtrativa que vise imitar um clarinete. Este instrumento também só apresenta os
+subtrativa que vise a imitar um clarinete. Este instrumento também só apresenta os
 componentes ímpares do espectro harmônico e a onda quadrada convém com sua energia abundante nas altas frequências.
@@
 seu espectro é perfeitamente harmônico, sem os ruídos e variações típicas do fenômeno natural. Isso pode ser 
-visto na figura~\ref{fig:espectroOboe} onde dispomos os espectros da nota original do oboé e de uma nota 
+visto na figura~\ref{fig:espectroOboe}, que mostra os espectros da nota original do oboé e de uma nota 
 artificial de mesma duração e cujas amostras consistem no mesmo período da figura~\ref{fig:formasDeOnda}.
@@
 
-A figura ~\ref{fig:amostras2} exibe duas amostras e as componentes espectrais que contém.
+A figura ~\ref{fig:amostras2} exibe duas amostras e as componentes espectrais que contêm.
 A decomposição de Fourier resulta neste caso em um único par de coeficientes $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relativos às frequências $\{f_k\}_0^1=\left\{w_k\frac{f_a}{2\pi}\right\}_0^1=\left\{k\frac{f_a}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_a}{2}=f_{\text{máx}}\right\}$
@@
 
-Sempre que partimos de $\Lambda$ amostras reais $t_i$ resultamos em $\Lambda$ coeficientes complexos $c_k=a_k+j.b_k$, os coeficientes $c_k$ se equivalem dois a dois correspondendo às mesmas frequências e com contribuições idênticas\footnote{Parte real igual e imaginária com sinal trocado: $a_{k1}=a_{k2}$ e $b_{k1}=-b_{k2}$.
+$\Lambda$ amostras reais $t_i$ resultam em $\Lambda$ coeficientes complexos $c_k=a_k+j.b_k$.
+Os coeficientes $c_k$ se equivalem dois a dois correspondendo às mesmas frequências e com contribuições idênticas\footnote{Parte real igual e imaginária com sinal trocado: $a_{k1}=a_{k2}$ e $b_{k1}=-b_{k2}$.
 Como consequência temos módulos iguais e fases com sinais opostos}.
-Lembrando que $f_k = k\frac{f_a}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $, pode-se anotar que, quando $k> \frac{\Lambda}{2} $,
+Lembrando que $f_k = k\frac{f_a}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $, pode-se notar que, quando $k> \frac{\Lambda}{2} $,
 a frequência $f_k$ é espelhada em $\frac{f_a}{2}$ da seguinte forma $f_k=\frac{f_a}{2} - (f_k-\frac{f_a}{2})=f_a-f_k=f_a - k\frac{f_a}{\Lambda}=(\Lambda-k)\frac{f_a}{\Lambda} \;\;\;\; \Rightarrow \;\;\;\; f_k\equiv f_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
@@
  $w_k=f_k.\frac{2\pi}{f_a}$ e lembrando da periodicidade $2\pi$, que resulta em $w_k=-w_{\Lambda-k}$.
-Como o coseno é uma função par e a tangente inversa é impar, as componentes em $w_k$ e $w_{\Lambda-k}$ se somam na equação de reconstrução das amostras reais mostrada em ~\ref{recomposicaoFourier}.
-
-  Dito de outra forma, em uma decomposição de $\Lambda$ amostras, as $\Lambda$ componentes frequenciais $\{c_i\}_0^{\Lambda-1}$ resultantes
+Como o coseno é uma função par e a tangente inversa é impar, as componentes em $w_k$ e $w_{\Lambda-k}$ se somam na equação de reconstrução das amostras reais disposta na equação~\ref{recomposicaoFourier}.
+
+  Ou seja, em uma decomposição de $\Lambda$ amostras, as $\Lambda$ componentes frequenciais $\{c_i\}_0^{\Lambda-1}$ resultantes
    são equivalentes em pares.
-   Excepcional exceção para $f_0$ e, no caso de $\Lambda$ ser par, de $f_{\Lambda/2}=f_{\text{máx}}=\frac{f_a}{2}$ , ambas as componentes são isoladas, i.e.
+   Exceção para $f_0$ e, no caso de $\Lambda$ ser par, de $f_{\Lambda/2}=f_{\text{máx}}=\frac{f_a}{2}$ , ambas as componentes são isoladas, i.e.
 não existe outra componente na frequência $f_0$ ou $f_{\Lambda/2}$ (se $\Lambda$ par) além dela mesma.
 Pois $f_{\Lambda/2}=f_{(\Lambda-\Lambda/2) = \Lambda/2}$ e $f_0=f_{(\Lambda-0)=\Lambda}=f_0$.
-
 Além disso, estas duas frequências (a frequência zero e a frequência máxima) não são representadas com variação de fase e, portanto, são estritamente reais. Assim, pode-se
@@
 As contribuições individuais se somam na forma de onda 
-original, e uma breve inspeção revela que as curvaturas maiores são fruto da frequência mais aguda
-enquanto um deslocamento fixo da somatória das componentes é fruto
+original, e uma breve inspeção revela que as curvaturas maiores são fruto da frequência mais aguda,
+enquanto um deslocamento fixo da somatória das componentes advém
 da componente na frequência zero.
@@
         \includegraphics[width=\textwidth]{figuras/amostras6}
-    \caption{Componentes frequenciais em 6 amostras}
+    \caption{Componentes frequenciais em 6 amostras: 3 senoides se somam ao \emph{bias}}
         \label{fig:amostras6}
@@
         \includegraphics[width=\textwidth]{figuras/amostras6formas___}
-    \caption{Formas de onda básicas em 6 amostras}
+    \caption{Formas de onda básicas em 6 amostras: as ondas triangular e quadrada possuem os harmônicos ímpares, mas em proporções e fases diferentes; a dente de serra possui também o harmônico par.}
         \label{fig:formas6}
@@
 
-Note que a nota por si só não especifica um timbre.
+A nota por si só não especifica um timbre.
 Mesmo assim, faz-se necessária a escolha de uma forma de onda para que as amostras $t_i$ tenham um valor estabelecido individualmente.
@@
 a sequência que descreve um período da onda $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ de duração 
-$\delta_f=1/f$, dadas pelas equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} e onde $R_i^f$ 
+$\delta_f=1/f$, dadas pelas equações ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} e ~\ref{quadrada} e onde $R_i^f$ é
 uma onda real amostrada:
@@
         \includegraphics[width=.5\textwidth]{figuras/espacializacao___}
-    \caption{Detecção de localização espacial de fonte sonora: DTI e DII}
+    \caption{Detecção de localização espacial de fonte sonora: esquema utilizado para cálculo da diferença de tempo interaural (DTI) e da diferença de intensidade interaural (DII).}
         \label{fig:spac}
@@
 
-Embora consideravelmente simples até aqui, a localização espacial depende drasticamente de outras pistas. De forma canonica, pela
+Embora consideravelmente simples até aqui, a localização espacial depende drasticamente de outras pistas. Pela
 DTI e DII especifica-se somente o ângulo horizontal (azimutal) $\theta$ dado por:
@@
 
-Mesmo assim, há dificuldades quando $\theta$ incinde sobre o chamado "cone de confusão" em que um mesmo par de especificações DTI, DII resultam de vários dos pontos 
+Mesmo assim, há dificuldades quando $\theta$ incide sobre o chamado "cone de confusão" em que um mesmo par de especificações DTI, DII resultam de vários dos pontos 
 do cone.
@@
 
-Já a localização completa, incluindo distância e elevação da fonte sonora, é dada pela função de transferência de cabeça (HRTF - do inglês \emph{Head Related Transfer Function}).\cite{hrtf} Exitem bases abertas e conhecidas de HRTF como a CIPIC e pode-se aplicar estas funções de transferência em um som por convolução (veja equação~\ref{eq:conv}).\cite{CIPIC} O corpo do indivíduo altera bastante as filtragens realizadas e existem técnicas para resultar em HRTFs que sejam - como proposta - utilizáveis de forma universal.\cite{lazaSPA}
+Já a localização completa, incluindo distância e elevação da fonte sonora, é dada pela função de transferência de cabeça (HRTF - do inglês \emph{Head Related Transfer Function}).\cite{hrtf} Exitem bases abertas e conhecidas de HRTF como a CIPIC e pode-se aplicar estas funções de transferência em um som por convolução (veja equação~\ref{eq:conv}).\cite{CIPIC} O corpo do indivíduo altera bastante as filtragens realizadas e existem técnicas para gerar HRTFs que sejam - como proposta - utilizáveis de forma universal.\cite{lazaSPA}
 
@@
 
-Em posse da nota básica, pode-se realizar estruturas musicais com
-sequências destas partículas. A soma de $N$ sequências $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ de mesmo tamanho $\Lambda$, resulta em seus conteúdos espectrais sobrepostos em um processo de mixagem sonora:
+A partir da nota básica, pode-se realizar estruturas musicais com
+sequências destas partículas. A soma de $N$ sequências $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ de mesmo tamanho $\Lambda$ resulta em seus conteúdos espectrais sobrepostos em um processo de mixagem sonora:
 
@@
 As sequências podem também ser concatenadas no tempo.
-Caso as sequências $\{t_{k,i}\}_0^{\Lambda_k-1}$ de tamanhos $\Lambda_k$  representem $k$ notas musicais, sua concatenação em uma única sequência $T_i$ resultará em uma sequência musical simples ou melodia:
+Caso as sequências $\{t_{k,i}\}_0^{\Lambda_k-1}$ de tamanhos $\Lambda_k$  representem $k$ notas musicais, sua concatenação em uma única sequência $T_i$ resulta em uma sequência musical simples ou melodia:
 
@@
 
-Este mecanismo é demonstrado de forma ilustrativa na figura~\ref{fig:concatenacao} com as mesmas sequências utilizadas na figura ~\ref{fig:mixagem}.
+Este mecanismo é demonstrado de forma ilustrativa na figura~\ref{fig:concatenacao} com as mesmas sequências da figura ~\ref{fig:mixagem}.
  As sequências são curtas para as taxas de amostragem usuais, mas pode-se observar a concatenação de sequências sonoras. Além disso, cada nota tem a duração maior que $100ms$ se $f_a<1kHz$.
@@
 
-Está descrita a nota musical digital básica e a seção a seguite desenvolve a evolução temporal de seus conteúdos, como nos glissandos e nas envoltórias.
+Está descrita a nota musical digital básica e a seção a seguinte desenvolve a evolução temporal de seus conteúdos, como nos glissandos e nas envoltórias.
 A filtragem de componentes espectrais e a geração dos ruídos completam a constituição da nota musical como unidade isolada e se desdobra na seção~\ref{notasMusica}, dedicada à estruturação destas notas em música através de métricas e trajetórias.
@@
 
-A nota musical digital básica está bem definida na seção~\ref{sec:notaDisc} com os parâmetros:
+A nota musical digital básica foi definida na seção~\ref{sec:notaDisc} com os parâmetros:
 duração, altura, intensidade (volume) e timbre. Esta é uma modelagem
-útil e paradigmática, mas não esgota o que se entedente por
+útil e paradigmática, mas não esgota o que se entende por
 uma nota musical.
@@
 Estas variações não são obrigatórias e sim orientações da
-síntese sonora para usos musicais pois é como os sons
+síntese sonora para usos musicais, pois é como os sons
 se apresentam na natureza\footnote{A regra de ouro
@@
 por si só, faça como que tenha variações internas.\cite{Roederer}}. 
-
 Explorar todas as formas pelas quais estas variações ocorrem está fora
 do escopo de qualquer trabalho dada a considerável sensibilidade do ouvido humano
-e a complexidade da nossa cognição sonora e, a seguir, serão apontados
+e a complexidade da nossa cognição sonora. A seguir, serão apontados
 recursos primários para estas variações das características na nota
 básica.
-
 Todas as relações descritas nesta seção estão no Apêndice~\ref{sec:cod2}.
@@
 amostras recolhidas da natureza. 
-
-\begin{figure}[h!]
-    \centering
-        \includegraphics[width=\textwidth]{figuras/lut}
-    \caption{Lookup Table}
-        \label{fig:lut}
-\end{figure}
-
 Na música seu uso transcende estes
@@
 de frequências audíveis, qualquer que seja a forma de onda amostrada.
+
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=\textwidth]{figuras/lut}
+    \caption{Lookup Table}
+        \label{fig:lut}
+\end{figure}
 
@@
 da onda vibratória para alteração do volume\footnote{A mudança do volume (qualidade psicofísica) ocorre através de diferentes características 
-do som, como a reverberação e a concentração de harmônicos agudos, dentre as quais está energia da onda.
+do som, como a reverberação e a concentração de harmônicos agudos, dentre as quais está a energia da onda.
 A manipulada com mais facilidade é a energia da onda (veja equação~\ref{eq:potencia}) e esta também pode variar de diferentes formas.
@@
 
-Aqui convém a conversão de decibels para amplitude apresentada na equação ~\ref{ampDec}.
-Junto à equação \ref{transAmp}, resulta na transição de $V_{dB}$ decibels:
+Aqui convém a conversão de decibels para amplitude. Assim, as equações ~\ref{ampDec} e \ref{transAmp}
+resultam na transição de $V_{dB}$ decibels:
 
@@
 para o caso geral de variações de amplitude segundo a progressão geométrica.
-Quanto maior o valor de $\alpha$, mais suave é a introdução do som e mais intenso o final do fade.
+Quanto maior o valor de $\alpha$, mais suave é a introdução do som e mais intenso o final da transição.
 $\alpha>1$ resulta em transições de volume muitas vezes chamadas de \emph{slow fade} enquanto $\alpha<1$ resulta em \emph{fast fade}.\cite{fades}
 
-As transições lineares serão utilizadas para
-de as chamadas sínteses AM e FM e a aplicação das transições
+As transições lineares serão usadas para
+as sínteses AM e FM e a aplicação das transições
 logarítmicas para os tremolos e vibratos.
@@
 no domínio temporal ao som
-digitalizado pela convolucão do som com a 
+digitalizado pela convolução do som com a 
 resposta ao impulso do filtro\footnote{Pode-se aplicar o filtro do domínio espectral através da multiplicação das transformadas de Fourier de ambos o som e a resposta ao impulso, e então realizada a transformada inversa de Fourier do espectro resultante.\cite{Openheim}}.
@@
 tem cada i-ésima amostra $t_i$ substituída pela soma de suas últimas $\Lambda_h$ amostras $\{t_{i-j}\}_{j=0}^{\Lambda_h-1}$
-multiplicadas uma a uma pela resposta ao impulso $\{h_i\}_0^{\Lambda_h-1}$ em si como na figura~\ref{fig:conv}. A resposta ao impulso $\{h_i\}$
-é percorrida na forma retrograda e os cálculos explícitos
-das amostras $t_{12}'$ e $t_{32}'$ são duas amostras calculadas
+multiplicadas uma a uma pelas amostras da resposta ao impulso $\{h_i\}_0^{\Lambda_h-1}$. Este
+procedimento está ilustrado na figura~\ref{fig:conv}, onde a resposta ao impulso $\{h_i\}$
+é percorrida na forma retrógrada e
+$t_{12}'$ e $t_{32}'$ são duas amostras calculadas
 pela convolução $(T_j*H_j)_i=t_i'$. O sinal resultante possui
@@
 A resposta ao impulso pode provir de medições
-físicas ou à síntese direta do filtro. Uma resposta 
+físicas ou da síntese. Uma resposta 
 ao impulso para a aplicação
-de reverberação pode resultar da gravação do ao disparar
+de reverberação pode resultar da gravação sonora em um ambiente ao disparar
 um estalo que se assemelhe a um impulso ou 
@@
 da resposta em frequência.
-
 Ambas são respostas ao impulso
 que, convoluídas com a sequência sonora, resultam na própria sequência
-com uma reverberação que se assemelha à reverberação do ambiente 
+com uma reverberação que se assemelha àquela do ambiente 
 em que ocorreu a medição.\cite{Cook}
@@
 A transformada inversa
-de fourier de uma envoltória par e real é uma
+de Fourier de uma envoltória par e real é uma
 resposta ao impulso de um FIR. Este realiza
@@
 Quanto maior o número de amostras maior
-a resolução a envoltória e também 
-o processamento computacional pois a convolução é cara.
+a resolução da envoltória e também 
+o processamento computacional, pois a convolução é cara.
 
@@
         \includegraphics[width=\textwidth]{figuras/delays__}
-    \caption{Convolução com o impulsoi, linhas de delays e síntese granular}
+    \caption{Convolução com o impulso: deslocamento, linhas de delays e síntese granular}
         \label{fig:delays}
@@
 e é caracterizada por possuir uma representação temporal
-infinita, i.e. a resposta ao impulso não converge para zero. Sua aplicação é usualmente feita pela equação
-a diferenças:
-
-\begin{equation}
+infinita, i.e. a resposta ao impulso não converge para zero.
+Sua aplicação é usualmente feita pela equação:
+
+\begin{equation}\label{eq:diferencas}
 t_i' = \frac{1}{b_0}\left ( \sum_{j=0}^Ja_j . t_{i-j} + \sum_{k=1}^Kb_k . t_{i-k}' \right )
@@
 
-com $b_0=1$ na grande maioria dos casos pois basta normalizar as variáveis para
-resultar em filtro idêntico: $a_j'=\frac{a_j}{b_0}$ e $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
+com $b_0=1$ na grande maioria dos casos pois pode-se normalizar as variáveis:
+$a_j'=\frac{a_j}{b_0}$ e $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
+A equação~\ref{eq:diferencas} é chamada 'equação a diferenças' por exibir as amostras resultantes $\left\{t_i'\right\}$
+através das diferenças entre as amostras originais $\{t_i\}$ e as amostras resultantes anteriores $\left\{t_{i-k}'\right\}$.
 
@@
 No caso dos filtros passa e rejeita banda, esta mesma atenuação é
-resultado de duas especificações: $f_c$ (neste caso melhor compreendida como 'frequência central') e a largura de banda $bw$,
+resultado de duas especificações: $f_c$ (neste caso mais bem compreendida como 'frequência central') e a largura de banda $bw$,
 em ambas as frequências $f_c \pm bw$ há uma atenuação de $\approx 0.707$ da amplitude original.
-
 Existe amplificação do som no caso dos filtros passa e rejeita banda quando a frequência
@@
 biquad\footnote{Abreviação
-de 'biquadrado' pois sua função de transferência possue dois polos e dois zeros, i.e. sua
+de 'biquadrado' pois sua função de transferência possui dois polos e dois zeros, i.e. sua
 forma normal consiste em dois polinômios quadráticos formando uma fração:
@@
         \includegraphics[width=\textwidth]{figuras/iir___}
-    \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda}}
+    \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda} para diferentes frequências de corte, frequências centrais e larguras de banda.}
         \label{fig:iir}
@@
 
-\item Passa-altas de polo simples o módulo da resposta em frequência no canto superior direito da figura~\ref{fig:iir}. A fórmula geral,
+\item Passa-altas de polo simples com o módulo da resposta em frequência no canto superior direito da figura~\ref{fig:iir}. A fórmula geral,
 com frequência de corte $f_c \in (0,\frac{1}{2})$, é calculada através da variável
@@
 e a largura de banda $bw$
-- $f_c \pm bw$ resultam em 0.707 da amplitude, i.e. atenuação de $3dB$ -
-ambos dados como fracoes de $f_a$, portanto $f,\; bw \in (0,0.5)$.
+- $f_c \pm bw$, que resultam em $0.707$ da amplitude, i.e. atenuação de $3dB$ -
+ambos dados como frações de $f_a$, portanto $f,\; bw \in (0,0.5)$.
 
@@
 
-O filtro passa banda do canto inferior esquerdo da figura~\ref{fig:iir}:
+O filtro passa banda do canto inferior esquerdo da figura~\ref{fig:iir}
+possui os seguintes coeficientes para a equação~\ref{eq:diferencas}:
 
@@
 
-e o filtro rejeita banda:
+Os coeficientes do filtro rejeita banda são:
 
@@
 são chamados ruídos.\cite{Lacerda}
-Estes sons são constituintes importantes dos sons musicais de altura definida,
+Estes são constituintes importantes dos sons musicais de altura definida,
 como os ruídos presentes nas notas do piano, do violino, etc. Além disso, os instrumentos
@@
 a eletroacústica e gêneros de pista de dança, os ruídos possuem usos diversificados e comumente
-idiomáticos.\cite{Cook}
+característicos do estilo musical.\cite{Cook}
 
@@
 
-Outra possibilidade é a geração de ruído através do espectro desejadoi, a partir
+Outra possibilidade é a geração de ruído através do espectro desejado, a partir
 do qual executamos a transformada inversa de Fourier.
@@
 Abaixo elencamos alguns ruídos de espectro estático. São chamados \emph{coloridos} por terem sido associados a cores. 
-A figura~\ref{fig:ruidos} dipoe lado a lado o perfil espectral e
+A figura~\ref{fig:ruidos} mostra lado a lado o perfil espectral e
 a sequência sonora. Os ruídos foram gerados com a mesma fase, então
@@
 
-\item O ruído branco deve seu nome por possuir energia distribuida
-igualmente por todas as frequências. Podse realizar
+\item O ruído branco deve seu nome por possuir energia distribuída
+igualmente por todas as frequências. Pode-se realizar
 o ruído branco com a transformada inversa dos seguintes coeficientes:
@@
 O valor de $c_i$ calculado pela exponencial é apenas um artifício para resultar em módulo unitário e fase aleatória.
-Já $c_{\Lambda/2}$ é sempre puramente real (como vimos na sessão anterior).
+Já $c_{\Lambda/2}$ é sempre puramente real (como vimos na seção anterior).
 
@@
 
-O resto dos ruídos podem ser feitos com base neste procedimento descrito para 
-o ruído rosa, basta que modificar detalhes, em especial a equação que define $\alpha_i$.
-
-\item O ruído marrom deve seu nome ao Robert Brown, que descreveu o movimento browniano.
+Os ruídos restantes podem ser feitos com base no procedimento descrito para 
+o ruído rosa, bastando que modificar detalhes, em especial a equação que define $\alpha_i$.
+
+\item O ruído marrom deve seu nome a Robert Brown, que descreveu o movimento browniano.
 Embora esta origem seja um tanto díspar do que pode-se considerar motivo para uma associação com a cor marrom, o ruído sonoro ficou consagrado com este nome.
-De qualquer forma, é bastante comum declarar satisfatória a associação do ruído com a cor marrom, dados os ruídos branco e rosa mais estridentes e relacionados a cores mais intensas~\cite{marrom}.
+De qualquer forma, é bastante comum declarar satisfatória a associação do ruído com a cor marrom, uma vez que os ruídos branco e rosa são mais estridentes e relacionados a cores mais intensas~\cite{marrom}.
 
@@
 
-\item O ruído cinza e definido como
+\item O ruído cinza é definido como
 um ruído branco sujeito a uma das curvas iso-audíveis. Estas curvas são resultados
@@
 \ref{eq:azul}, \ref{eq:violeta} estão na figura \ref{fig:ruidos}. Os espectros foram feitos com a mesma fase em cada coeficiente de mesma frequência, de forma que
-pode-se observar a contribuição dos harmônicos agudos e
+se pode observar a contribuição dos harmônicos agudos e
 das frequências graves.
 
-\subsection{Usos musicais parte 1: tremolo e vibrato, AM e FM}\label{subsec:tvaf}
-
-Enquanto o vibrato é uma variação periódica de altura (frequência),
+\subsection{Usos musicais - Parte 1: tremolo e vibrato, AM e FM}\label{subsec:tvaf}
+
+O vibrato é uma variação periódica de altura (frequência) e 
 o tremolo é uma variação periódica de volume (intensidade).\footnote{Alguns instrumentos e contextos musicais usam nomenclaturas diferentes.
@@
 As definições presentes neste trabalho tem por base uma literatura mais abrangente do que a utilizada para um único instrumento, prática ou tradição musical e comum em contextos de teoria musical e musica eletrônica.\cite{Lacerda,Harmonia}}
-
 Para o caso geral, o vibrato é descrito da seguinte forma:
@@
 As variáveis $\mu$ e $\nu$ quantificam a intensidade do vibrato: $\mu$ é uma medida direta da quantidade
-de Herz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação ($2\nu$ é o número de semitons entre os picos superioes e inferiores de oscilação da frequência do som $\{t_i\}$ causada pelo vibrato).
+de Herz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação ($2\nu$ é o número de semitons entre os picos superiores e inferiores de oscilação da frequência do som $\{t_i\}$ causada pelo vibrato).
 É conveniente $\nu=\log_{2}\frac{f+\mu}{f} $ neste caso pois o aumento máximo de frequência
@@
 A figura~\ref{fig:tremolo} mostra a amplitude das sequências $\{a_i\}_0^{\Lambda-1}$ e $\{t_i'\}_0^{\Lambda-1}$
-para três oscilações de um tremolo com forma da dente de serra. A curvatura é devido à progressão logarítmica de
+para três oscilações de um tremolo com forma da dente de serra. A curvatura é devida à progressão logarítmica de
 intensidade.
@@
 
-No aumento progressico de $f'$,
+No aumento progressivo de $f'$,
 a aproximação do limiar de frequência
@@
 explora este limiar com concomitâncias de tremolos e vibratos na mesma
-voz, com intensidades e formas de onda diferentes. Seu código está o Apêndice~\ref{ap:rugosi} e disponível online como parte da \massa.
+voz, com intensidades e formas de onda diferentes. Seu código está o Apêndice~\ref{ap:bela} e disponível online como parte da \massa.
 
@@
 Neste caso, as oscilações são
-audível como altura. Assim, $f'$, $\mu$ e a forma de onda realizam alterações espectrais no som
-original $T_i$ de formas diferentes para os tremolos e para os vibratos, são as 
+audíveis como altura. Assim, $f'$, $\mu$ e a forma de onda realizam alterações espectrais no som
+original $T_i$ de formas diferentes para os tremolos e para os vibratos. São as 
 chamadas sínteses AM (\emph{Amplitude Modulation}) e FM (\emph{Frequency Modulation}), respectivamente.
-Estas são técnicas muito poderosas e conhecidas, com clássicas aplicações em sintetizadores
-históricos como o \emph{Yamaha DX7} e com aplicações fora da música, como em telecomunicações para transmissão de informação via ondas eletromagnéticas (ex. radios AM e FM).
+Estas são técnicas conhecidas, com aplicações em sintetizadores
+como o \emph{Yamaha DX7}, e com aplicações fora da música, como em telecomunicações para transmissão de informação via ondas eletromagnéticas (ex. rádios AM e FM).
 
@@
 Nestas equações, a variação de frequência introduzida por $\{t_i'\}$ não respeita a progressão geométrica de frequência que acompanha a percepção de altura, mas sim a equação~\ref{freqLinear}.
-A utilização das equações~\ref{vbrF} para a FM está no Apêndice~\ref{cap:fmam} onde é calculado o conteúdo espectral da síntese FM feita com oscilações na escala logarítmica.
-De fato, o comportamento simples que torna a FM atraente é obtido somente com as variações lineares utilizadas em~\ref{eq:fmEsp}.
+A utilização das equações~\ref{vbrF} para a FM está no Apêndice~\ref{cap:fmam}, onde é calculado o conteúdo espectral da síntese FM obtida com oscilações na escala logarítmica.
+De fato, o comportamento simples que torna a FM atraente é obtido somente com as variações lineares em~\ref{eq:fmEsp}.
 
@@
 
-Ou seja, o som resultante é o som original
+Ou seja, o som resultante é o original
 e a reprodução de seu conteúdo espectral acima e abaixo da frequência
@@
 
-\subsection{Usos musicais parte 2}
+\subsection{Usos musicais - Parte 2}
 A este ponto as possibilidades musicais explodiram. As
 características de altura (dada pela frequência),
-tímbre (dado pela forma de onda e filtragens),
+timbre (dado pela forma de onda e filtragens),
 volume (dado pela intensidade) e duração (dada pelo número de amostras)
@@
 com o objetivo de exemplificar manipulações sonoras que resultem algo
-musical, por razões variadas e aprofundadas na próxima sessão.
+musical, por razões variadas e aprofundadas na próxima seção.
 
@@
 \emph{Tremolos, vibratos e a frequência} explora
-recursos como este e variações da forma de onda da oscilação com vínculos, de modo a formar um \emph{idioma musical}\footnote{Veja na próxima sessão.}.
+recursos como este e variações da forma de onda da oscilação com vínculos, de modo a formar um \emph{idioma musical}\footnote{Veja na próxima seção.}.
 Seu código está no Apêndice~\ref{ap:tremolos} e também disponivel online como parte da caixa de ferramentas \massa.
@@
 estabelecida pela convolução com impulsos é explorado na montagem \emph{Trenzinho de caipiras impulsivos}.
-Os recursos explorados incluem a criação de amálgamas sonoros provenientes da síntese granular e esta montagem é já uma ponte para a próxima sessão.
+Os recursos explorados incluem a criação de amálgamas sonoros provenientes da síntese granular e esta montagem é já uma ponte para a próxima seção.
 Veja especialmente a figura~\ref{fig:pulsoSubAgl}.
@@
 Primeiro, a nota básica foi definida e suas características postas em termos
-claros e quantitativos (sessão~\ref{notaBasica}).
-Em seguida, a nota foi adentrada e foram compreendidas suas transições internas e de tratamentos imediatos (sessão~\ref{varInternas}).
-Por fim, esta última sessão dedica-se a organizar estas notas musicais em música.
+claros e quantitativos (seção~\ref{notaBasica}).
+Em seguida, a nota foi adentrada e foram compreendidas suas transições internas e de tratamentos imediatos (seção~\ref{varInternas}).
+Por fim, esta última seção dedica-se a organizar estas notas musicais em música.
 A gama de recursos com uma infinidade de possibilidades de resultados,

### 2012-11-26 devEresultados.tex
tudo ateh secao 2.3

@@
 
-Uma primeira possibilidade interessante é a utilização de vínculos
+Uma primeira possibilidade interessante é usar vínculos
 entre os parâmetros do tremolo e do vibrato e algum parâmetro da nota básica,
-como a frequência. Assim, pode-se estabelecer
-a frequência do vibrato  é diretamente proporcional à altura à qual a profundidade do tremolo é inversamente proporcional.
+como a frequência. Assim, pode-se estabelecer que
+a frequência do vibrato é diretamente proporcional à altura, e a profundidade do tremolo é inversamente proporcional
+à altura.
 Desta forma, com as equações \ref{vbrGamma}, \ref{vbrF} e \ref{trA}
@@
 Com $f^{vbr}$ e $f^{tr}$ como $f'$ nas equações de referência, ou seja, a frequência
-de oscilação do vibrato e do tremolo utilizadas na equação~\ref{vbrGamma}. Já $\nu$ e $V_{dB}$ são as profundidades
+de oscilação do vibrato e do tremolo da equação~\ref{vbrGamma}. Já $\nu$ e $V_{dB}$ são as profundidades
 do vibrato e do tremolo, respectivamente. As funções $func_a$,
-$func_b$ e $func_c$ são arbitrárias e dependentes das intensões musicais. A montagem 
+$func_b$ e $func_c$ são arbitrárias e dependentes das intenções musicais. A montagem 
 \emph{Tremolos, vibratos e a frequência} explora
 recursos como este e variações da forma de onda da oscilação com vínculos, de modo a formar um \emph{idioma musical}\footnote{Veja na próxima seção.}.
-Seu código está no Apêndice~\ref{ap:tremolos} e também disponivel online como parte da caixa de ferramentas \massa.
+Seu código está no Apêndice~\ref{ap:tremolos} e também disponível online como parte da caixa de ferramentas \massa.
 
@@
 
-A envoltória ADSR canônica é caracterizada por 4 parâmetros: duração do ataque (tempo que o som demora para atingir seu volume máximo), duração do decaimento (segue ao ataque imediatamente), nível de volume de sustentação (em que o volume fica estável após o decaimento) e duração de soltura (após a sustentação, nesta duração o volume decai a zero).
+A envoltória ADSR escolástica é caracterizada por 4 parâmetros: duração do ataque (tempo que o som demora para atingir seu volume máximo), duração do decaimento (segue ao ataque imediatamente), nível de volume de sustentação (em que o volume fica estável após o decaimento) e duração de soltura (após a sustentação, nesta duração o volume decai a zero).
 Note que o tempo de sustentação não é especificado, pois é resultante da duração em si menos os tempos de ataque, decaimento e soltura.
@@
 A aplicação da envoltória ADSR com durações $\Delta_A$, $\Delta_D$ e $\Delta_R$, duração total $\Delta$ e nível de sustentação $a_S$, dado
-como fração da amplitude máxima, em uma sequência sonora $t_i$ qualquer pode ser feita da seguinte forma:
+como fração da amplitude máxima, em uma sequência sonora $t_i$ qualquer, pode ser feita da seguinte forma:
 
@@
 $\xi=10^{\frac{-80}{20}}=10^{-4}\;$ ou $\;\xi=10^{\frac{-40}{20}}=10^{-2}$.
-Quanto menor for $\xi$ mais lento é o face, a exemplo de $\alpha$ da figura~\ref{fig:transicao} .
+Quanto menor for $\xi$ mais lento é o \emph{fade}, a exemplo de $\alpha$ da figura~\ref{fig:transicao} .
 
@@
 máxima perdura. Outro exemplo comum é o uso de traçados mais elaborados para o
-ataque ou para o decaimento. A montagem musical \emph{ADa e SaRa} explora diversas destas
-configurações da envoltória ADSR.
+ataque ou para o decaimento. A montagem musical \emph{ADa e SaRa}, disponível no Apêndice~\ref{ap:ada}
+e na \massa, explora diversas destas configurações da envoltória ADSR
 

### 2012-11-26 conclusoes.tex
terminada ateh conclusoes (incluso)

@@
 
-A disposição online destes conteúdos na forma de hipertexto junto aos códigos e exemplos sonoros, todos em licenças livres, facilita colaborações e geração de subprodutos em co-autoria, e com isso a expansão da \massa com novas implementações e desenvolvimentos das montagens musicais.
-
+A disposição online destes conteúdos na forma de hipertexto junto aos códigos e exemplos sonoros, todos em licenças livres, facilita colaborações e geração de subprodutos em co-autoria, e com isso a expansão da \massa\ com novas implementações e desenvolvimentos das montagens musicais.
 Explorações sistemáticas de parametrizações (dos tremolos, da ADSR, etc) em alta fidelidade tem utilidade artística e é possibilidata por este trabalho com controle amostral.
@@
 de interesse em torno da questão criativa aliada à computação.
-Neste contexto, se destaca o grupo
+Neste contexto, destaca-se o grupo
 labMacambira.sf.net, que reúne colaboradores de todo o Brasil e alguns fora do país.
-Este grupo tem vida própria com relação ao papel do orientado
-deste trabalho, e já apresentou contribuições relevantes em diferentes áreas
+Este grupo
+já apresentou contribuições relevantes em diferentes áreas
 como Democracia Direta Digital, ferramentas de georeferenciamento e
 atividades artísticas e educacionais, como cursos, workshops e apresentações artísticas.
-Vários destes resultados estão apresentados no Apêndice~\ref{cap:musicaExtra} e no acervo online criado, que ultrapassa 700 videos, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos~\cite{siteLM,wikiLM,vimeoLM}.
+Vários destes resultados estão apresentados no Apêndice~\ref{cap:musicaExtra} e no acervo online criado, que ultrapassa 700 vídeos, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos~\cite{siteLM,wikiLM,vimeoLM}.

### 2012-11-26 devEresultados.tex
terminada ateh conclusoes (incluso)

@@
 
-Estes recursos todos utilizados em conjunto podem incindir na realização de um efeito chamado \emph{chorus}. A
+Estes recursos todos utilizados em conjunto podem incidir na realização de um efeito chamado \emph{chorus}. A
 exemplo do que ocorre com um coro de cantores, neste efeito o som é realizado com diversas pequenas modificações,
@@
 musicais. Para simplificar, sejam notas estes eventos musicais
-e $S_j$ uma estrutura musical. Esta seção é didicada às técnicas
-que torna $S_j$ interessante e agradável na audição.
+e $S_j$ uma estrutura musical. Esta seção é dedicada às técnicas
+que tornam $S_j$ interessante e agradável na audição.
 
@@
 A concatenação de $S_j$, como na equação~\ref{eq:concatenacao} e na figura~\ref{fig:concatenacao}, forma sequências melódicas e ritmos, associados ao 'pensamento horizontal' na música.
-A frequência fundamental $f$ e o momento de início (ataque) são as características mais importantes dos elementos de $S_j$, possibilitando, grosso modo, a preseça de músicas de alturas (harmonia e melodia) e a preseça da métrica temporal e ritmos, respectivamente.
+A frequência fundamental $f$ e o momento de início (ataque) são as características mais importantes dos elementos de $S_j$, possibilitando, grosso modo, a presença de músicas de alturas (harmonia e melodia) e a presença da métrica temporal e ritmos, respectivamente.
 
@@
 Esta precisão absoluta é característica de implementações
-computacionais, pois os intrumentos reais possuem desvios destas frequências para melhor compatibilizar os harmônicos
+computacionais, pois os intrumentos reais possuem desvios destas frequências para mais bem compatibilizar os harmônicos
 de suas notas.\cite{Roederer}
@@
 Esta divisão da oitava em doze notas é o cânone da música ocidental clássica,
-além de incindir em usos cerimoniais/religiosos e étnicos.\cite{Wisnick} Os intervalos entre quaisquer dois graus da escala de 12 notas.
-Notados em proporções de $\varepsilon$ entre as notas (i.e.
+além de incidir em usos cerimoniais/religiosos e étnicos.\cite{Wisnick} Notados em proporções de $\varepsilon$ entre as notas (i.e.
 um semitom), estes intervalos são, junto a suas notações abreviadas usuais:
@@
 Passando a nota mais grave para a oitava acima, ou a nota mais aguda uma oitava para baixo, o intervalo é invertido.
-Um intervalo, somado à sua inversão, resulta 9.
+Um intervalo, somado à sua inversão, resulta 9 (7m inverte para 2M: $7+2=9$).
 Um intervalo maior invertido resulta em um intervalo menor e vice-versa.
@@
 Um intervalo justo invertido resulta igualmente justo.
+Um intervalo maior que a oitava é classificado como intervalo composto e leva o nome do intervalo entre as mesmas notas, mas na mesma oitava.
 
@@
 Os intervalos de uníssono, oitava e quinta justa são consonâncias perfeitas.
-As terças e sextas são consonancias imperfeitas.
+As terças e sextas são consonâncias imperfeitas.
 As segundas, sétimas e trítonos são considerados dissonantes.
@@
 
-Um intervalo maior que a oitava é classificado como intervalo composto e leva o nome do intervalo entre as mesmas notas, mas na mesma oitava.
-
 Este arcabouço resume com exatidão a teoria tradicional dos intervalos musicais.\cite{Lacerda} A montagem \emph{Intervalos entre alturas} explora estes intervalos de formas isoladas e diversas.
@@
 'escalas' ou padrões, geram estruturas estáveis pelas simetrias internas, e podem ser
-repetidas de forma eficiênte e sustentada. Discutiremos mais sobre simetrias e voltaremos
+repetidas de forma eficiente e sustentada. Discutiremos mais sobre simetrias e voltaremos
 a estas divisões da oitava por igual.
@@
 Já a escala harmônica apresenta a sensível, mas não evita o intervalo de segunda aumentada entre o sexto e sétimo graus, pois não precisa contemplar trajetória melódica, apenas apresentar a sensível tão cara ao sistema tonal (a sensível tende à tônica, afirmando-a).\cite{Harmonia}
-
-Outras escalas podem ser representadas da mesma forma, como as pentatônicas e os modos de transposição limitados de messiaen.\cite{Messiaen}
+Outras escalas podem ser representadas da mesma forma, como as pentatônicas e os modos de transposição limitados de Messiaen.\cite{Messiaen}
 
 A ocorrência simultânea de notas é observada através dos acordes.
-Destes, a base na musica tonal são as tríades.
+Destes, a base na música tonal são as tríades.
 Estas constituem-se de duas terças sucessivas, em 3 notas: fundamental, terça e quinta.
@@
 espaçadas ($\varepsilon=2^{\frac{1}{7}}$),
-resultando em intervalos que pouco se assemelham aos intervalos presentes na
+resultando em intervalos que pouco se assemelham aos intervalos na
 divisão de doze notas.\cite{Wisnick}},
 pode-se usar reais não inteiros
-para descrever a sequência de alturas ou modificar o fator $\varepsilon$
+para descrever a sequência de alturas, ou modificar o fator $\varepsilon$
 e continuar usando inteiros. Por exemplo, uma afinação
-bastante próxima da série harmônica em si,
+bastante próxima da série harmônica em si
 é proposta na forma da divisão da oitava em $53$ notas:
@@
 continuam inteiras com o novo $\varepsilon$.
-Note que se $S_i$ é uma sequência de alturas $s_i$ e mapeamos uma sequência em $\varepsilon_1$ para
-$\varepsilon_2$,
-resultamos em uma nova sequência $S_i'$ com $s_i'=s_i \frac{\varepsilon_1}{\varepsilon_2}$.
-A montagem musical \emph{Micro tom} explora recursos microtonais e seu código está no Apêndice~\ref{ap:micro}, assim como parte da \massa\ online.
+Note que se $S_i$ é uma sequência de alturas relacionadas por $\varepsilon_1$,
+um mapeamento para notas em $\varepsilon_2$
+resulta em uma nova sequência $S_i'$ com $s_i'=s_i \frac{\varepsilon_1}{\varepsilon_2}$.
+A montagem musical \emph{Micro tom} explora recursos microtonais e seu código está no Apêndice~\ref{ap:micro}, assim como na \massa\ online.
 
@@
 Depois são considerados os outros graus.
-As convenções e técnicas apresentadas ao longo da história é o objeto de estudo da harmonia tradicional.\cite{Harmonia}
+As convenções e técnicas apresentadas ao longo da história
+são objeto de estudo da harmonia tradicional.\cite{Harmonia}
 
@@
 O acorde formado sobre o primeiro grau é o acorde de tônica e tem a função de manter um centro, um "chão" na música.
-O acorde formado sobre o quinto grau é o acorde de dominante e tem a função de tender à tônica, direcionar a música para ela.
+O acorde formado sobre o quinto grau é a dominante e tem a função de tender à tônica, direcionar a música para ela.
 A tríade formada sobre o quarto grau é a subdominante e tem a função de se distanciar da tônica.
@@
 
-As medianas\footnote{Também chamadas de medianas cromáticas.}, são acordes formados também a uma terça, mas com alterações com relação ao acorde de origem.
+As medianas\footnote{Também chamadas de medianas cromáticas.} são acordes formados também a uma terça, mas com alterações com relação ao acorde de origem.
 Caso haja duas alterações cromáticas, i.e.
@@
 
-A modulação é a troca de tonalidade e se dão principalmente através de: uma dominante individual que afirma a mudança da tônica; alterações cromáticas para um acorde da nova tonalidade; e enharmonia que é a mudança de função de um acorde com as mesmas notas.
-A importância da dominante torna ela o pivô natural das modulações, resultando no circulo das quintas.\cite{Harmonia,Schachter,Koellheuteur,Harmony}
+A modulação é a troca de tonalidade e se dão principalmente através de: uma dominante individual que afirma a mudança da tônica; alterações cromáticas para um acorde da nova tonalidade; e 'enarmonia' que é a mudança de função de um acorde com as mesmas notas.
+A importância da dominante torna ela o pivô natural das modulações, resultando no círculo das quintas.\cite{Harmonia,Schachter,Koellheuteur,Harmony}
 
@@
 \subsection{Contraponto}
-Na condução de diferentes vozes, os movimentos a notas vizinhas soam mais coerentes. A importância melódica recai sobre as vozes mais aguda e mais grave.
 
@@
 O propósito do contraponto é conduzir as vozes 
-de forma que soem independentes. Crucial para isso
+de forma que soem independentes. Cruciais para isso
 são as movimentações relativas,
@@
 conforme a figura~\ref{fig:movContraponto}. O movimento paralelo
-é um tipo de movimento oblíquo.
+é um movimento oblíquo.
 A regra de ouro é cuidar que os movimentos diretos
-não resulte em consonância perfeita. O movimento paralelo
+não terminem em consonância perfeita. O movimento paralelo
 deve só ocorrer entre consonâncias imperfeitas e não mais
@@
 ser não admitidas ou usadas seguidas 
-e precedidas de consonâncias em graus conjuntos.\cite{Fux,Tragtenberg,SchoenbergContra}
+e precedidas de consonâncias em graus conjuntos.
+Os movimentos que levam a nota a uma vizinha soam mais coerentes.
+Na presença de 3 ou mais vozes,
+a importância melódica recai sobre as vozes mais aguda e mais grave.\cite{Fux,Tragtenberg,SchoenbergContra}
 
@@
 por volta de $100ms$.\cite{microsound} Pode-se sumarizar 
-a transição da
-de durações ouvidas alturas para a apreciação em ritmo da seguinte forma:\cite{Alfaix, microsound}
+a transição
+de durações ouvidas como alturas para a apreciação em ritmo da seguinte forma:\cite{Alfaix, microsound}
 
@@
 
-A banda de durações marcada como transição está minimizada pois os limites não são bem definidos: a duração em que começa-se a perceber uma frequência fundamental ou uma separação entre as ocorrências é dependente da pessoa e de características do som.\cite{microsound,Roederer}
+A banda de durações marcada como transição está minimizada pois os limites não são bem definidos: a duração em que se começa a perceber uma frequência fundamental ou uma separação entre as ocorrências é dependente da pessoa e de características do som.\cite{microsound,Roederer}
 
@@
 
-As relações duais (compassos simples e divisões binárias) costumam a ocorrer em ritmos de dança
-e ocasiões festivas e são chamadas imperfeitas. As relações triádicas
+As relações duais (compassos simples e divisões binárias) costumam ocorrer em ritmos de dança
+e ocasiões festivas, e são chamadas imperfeitas. As relações triádicas
 incidem mais na música ritualística e relacionada ao sagrado
@@
 As unidades mais fortes (acentuadas) são as que consistem nas 'cabeças das divisões'. A cabeça de uma unidade é a primeira parte da subdivisão. Nas divisões binárias (2, 4 e 8 dos casos considerados),
-as unidades consideradas fortes se revesam com as fracas
-(e.g. a divisão em 4 resulta em forte, fraco, meio-forte, fraco).
+as unidades consideradas fortes se revezam com as fracas
+(e.g. a divisão em 4 é forte, fraco, meio-forte, fraco).
 Nas divisões ternárias (3, 6 e 9)
-à unidade forte (primeira) se sucedem 2 unidades fracas (e.g. a divisão em 3 resulta em forte, fraco fraco)\footnote{A divisão em 6 é considerada composta
+à unidade forte (primeira) se sucedem 2 unidades fracas (e.g. a divisão em 3 é forte, fraco fraco)\footnote{A divisão em 6 é considerada composta
  mas pode ocorrer também como uma divisão binária.
@@
 Qualquer unidade (nota), de uma
-dada sequência musical que tenha métrica resulta
-assim especificada:
+dada sequência musical que tenha métrica pode ser
+univocamente assim especificada:
 
@@
 pulso $P^0_2$ do segundo aglomerado de pulsos $P^1_2$.
-
 Cada unidade ou conjunto de unidades $P_i^j$ pode ser associada a uma sequência de amostras temporais $T_i$ que forma uma nota musical.
 
-A montagem \emph{Poli Hit Mia} utiliza as diferentes métricas e está no Apêndice~\ref{ap:poli} e disponível online como parte da \massa.
+A montagem \emph{Poli Hit Mia} utiliza as diferentes métricas e está no Apêndice~\ref{ap:poli}, disponível online como parte da \massa.
 
@@
 Dadas as estruturas musicais básicas tanto frequenciais (acordes e escalas) quanto 
-rítmicas (divisões e aglomerações simples, compostas e complexas) é 
+rítmicas (divisões e aglomerações simples, compostas e complexas), é 
 natural pensar em como realizar estas estruturas de forma que tenha coesão e sentido.\cite{Boulez}
-
 Para tal, os arcos são a base: partindo de algum lugar e voltando, forma-se um arco. Uma música sem nenhum arco pode ser compreendida como uma música sem coesão.
-Na subseção seguinte exploramos arcos cíclicos e neste
+Na subseção seguinte exploramos arcos cíclicos e nesta
 os provenientes de estruturas direcionais.
@@
 outra que, de forma paradigmática, volta do ápice à região de partida.
-Canonicamente, distingue-se entre arcos cujos clímax estão: no começo, no meio, no final, na primeira metade e na segunda metade como esquematicamente a dispostos na figura~\ref{fig:climax}.\cite{Schoenberg}
+Distingue-se entre arcos cujos clímax estão: no começo, no meio, no final, na primeira metade e na segunda metade como esquematicamente a dispostos na figura~\ref{fig:climax}.\cite{Schoenberg}
 
@@
 Seja $U_i=\{u_i\}_0^{\Lambda_U-1}$ uma sequência crescente. A sequência 
-$R_i=\{r_i\}_0^{2\Lambda_U -2}=\{u_{(\Lambda_U-1-|\Lambda_U-1-i|)}\}_0^{2\Lambda_U-2}$ 
-é uma sequência apresenta simetria especular perfeita, i.e.
+$R_i=\{r_i\}_0^{2\Lambda_U -2}=\left\{u_{(\Lambda_U-1-|\Lambda_U-1-i|)}\right\}_0^{2\Lambda_U-2}$ 
+é uma sequência que apresenta simetria especular perfeita, i.e.
 a segunda metade é uma versão espelhada da primeira.
-Pensada segundo conceitos musicais, o clímax está exatamente no meio da sequência.
+Segundo conceitos musicais, o clímax está exatamente no meio da sequência.
 Pode-se modificar isso com a substituição por sequências de tamanhos diferentes.
@@
 Teoricamente, estas sequências aplicadas desta forma
-a qualquer característica dos eventos musicais resulta em arcos,
+a qualquer característica dos eventos musicais produz arcos,
 pois resultam no afastamento e aproximação de uma parametrização inicial. 
-Assim, é possível para uma mesma sequência de eventos possuir um número de arcos distintos, com tamanhos e climax diferentes. Este é um recurso
+Assim, é possível para uma mesma sequência de eventos possuir um número de arcos distintos, com tamanhos e clímax diferentes. Este é um recurso
 interessante e útil e a correlação dos arcos resulta na coerência da escuta.\cite{Schachter}
@@
 as permutações representam quaisquer simetrias em um sistema finito.
-Na música, as permutações estão são ubíquas
-e estão presentes em técnicas.
-A aplicação sucessivas das permutações gera arcos cíclicos.\cite{change,Zamacois,permMusic}
-Esta abordagem resultou em dois trabalhos acadêmicos para geração de estruturas musicais.\cite{figgusOriginal, figgusEspacializacao}
+Na música, as permutações são ubíquas
+e estão presentes em técnicas, o que confirma seu papel central.
+A aplicação sucessiva das permutações gera arcos cíclicos.\cite{change,Zamacois,permMusic}
+A esta abordagem foram dedicados dois trabalhos acadêmicos para geração de estruturas musicais.\cite{figgusOriginal, figgusEspacializacao}
 
@@
 
-Todo elemento $p$ operado consigo mesmo um número suficiente de vezes resulta no elemento neutro $p^n=e$, pois caso contrário a propriedade do fechamento resultaria em um grupo infinito gerado por $p$.
+Todo elemento $p$ operado consigo mesmo um número suficiente de vezes $n$ atinge o elemento neutro $p^n=e$, pois caso contrário com a propriedade do fechamento se prova que o grupo é infinito (gerado $p$).
 O menor $n\;:\;p^n=e$ é chamado de ordem do elemento.
-Assim, uma permutação finita $p$ aplicada sucessivamente resulta na ordenação inicial dos elementos, formando um ciclo.
-Um arco musical se consultados os elementos em cada estágio.
+Assim, uma permutação finita $p$ aplicada sucessivamente atinge a ordenação inicial dos elementos, formando um ciclo.
+Este ciclo, se utilizado para parametros de notas musicais, resulta em um arco musical.
 
@@
 O conjunto de ordenações diferentes percorridas é um \emph{peal}.
-A tabela~\ref{tab:change} dispoe um \emph{peal} tradicional de 3 sinos (1, 2 e 3) que explora todas as suas ordenações.
+A tabela~\ref{tab:change} representa um \emph{peal} tradicional de 3 sinos (1, 2 e 3) que explora todas as suas ordenações.
 Cada linha apresenta uma ordenação dos sinos a ser tocada.
 As permutações estão entre as linhas.
+Neste caso, a estrutura musical consiste em permutações propriamente ditas e algumas permutações diferentes estão operando para resultar em um comportamento cíclico.
 
@@
 
-Neste caso, a estrutura musical consiste em permutações propriamente ditas e algumas permutações diferentes estão operando para resultar em um comportamento cíclico.
-
 A utilização de permutações na música pode ser resumida da seguinte forma:
@@
 pode operar em subconjuntos de $L_i$. Segundo, nem todos os elementos $l_i$ precisam ser executados a cada consulta de estado realizada.
-
 Para exemplificar, seja $L_i$ 
 uma sequência de notas musicais $l_i$. 
-Se $i$ é vai de $0$ a $n$, e $n>4$, a cada compasso
+Se $i$ vai de $0$ a $n$, e $n>4$, a cada compasso
 de $4$ notas pode-se executar as primeiras $4$ notas. As outras notas de
@@
 Cada uma destas permutações
-$p_i$ possui, ao menos, segundo a exposição acima: dimensões das notas em que opera (frequência, duração, fades, intensidade, etc) e período de incidência (a cada quantas consultas é aplicada a permutação).
+$p_i$ possui, ao menos, segundo a exposição acima: dimensões das notas em que opera (frequência, duração, \emph{fades}, intensidade, etc) e período de incidência (a cada quantas consultas é aplicada a permutação).
 Na realização das notas de $L_i$, uma forma fácil e coerente é executar as primeiras $n$ notas\footnote{A execução de notas disjuntas de $L_i$ equivale a modificar a permutação e executar as primeiras notas.}.
@@
 'idiomas musicais'.\cite{Lerdahl, Harmonia, Schachter,Alfaix}
-
 De forma simplificada, um idioma musical é fruto da
@@
 claros e quantitativos (seção~\ref{notaBasica}).
-Em seguida, a nota foi adentrada e foram compreendidas suas transições internas e de tratamentos imediatos (seção~\ref{varInternas}).
-Por fim, esta última seção dedica-se a organizar estas notas musicais em música.
-A gama de recursos com uma infinidade de possibilidades de resultados,
+Em seguida, a composição interna da nota foi abordada, e compreendidas suas transições internas e tratamentos imediatos (seção~\ref{varInternas}).
+Por fim, esta presente seção dedica-se a organizar estas notas musicais em música.
+A gama de recursos com uma infinidade de possibilidades de resultados
 é situação típica e cara às artes.\cite{Harmonia,Webern}
@@
 As escalas microtonais são importantes na música do século XX~\cite{microtonalidade} e possuem usos com resultados muito marcantes, como os quartos de tom na música indiana.
-A sequência musical \emph{MicroTom} explora estes recursos de forma dedicada, incluindo melodias microtonais e harmonias microtonais com várias notas em um âmbito de alturas bastante reduzido.
+A sequência musical \emph{MicroTom} explora estes recursos, incluindo melodias microtonais e harmonias microtonais com várias notas em um âmbito de alturas bastante reduzido.
 Seu código está no Apêndice~\ref{ap:micro} e disponível online como parte da \massa.
 
-Os vínculos entre parâmetros são formas poderosas de resultar peças e montagens musicais.
+Os vínculos entre parâmetros são formas poderosas de obter peças e montagens musicais.
 O número de notas permutadas pode variar no decorrer da música, revelando vínculo com a duração da peça.
@@
 
-Um uso particularmente coerente, do ponto de vista das simetrias apresentadas nas divisões da oitava (eqs.~\ref{escSim}) e das simetrias apresentadas através das permutações (tabela~\ref{tab:change} e eqs.~\ref{eq:groups}) é o uso conjunto de ambos.
-Nas peças \emph{3 trios} esta associação é feita de forma sistemática para possibilitar uma audição dedicada a esta associação.
-Esta é uma peça instrumental e não consta dentre os códigos dos apêndices e da \massa.
+As simetrias apresentadas nas divisões da oitava (eqs.~\ref{escSim}) e das simetrias apresentadas através das permutações (tabela~\ref{tab:change} e eqs.~\ref{eq:groups}) podem ser usadas em conjunto.
+Nas peças \emph{3 trios} esta associação é feita de forma sistemática para possibilitar uma audição a ela dedicada.
+Esta é uma peça instrumental e não consta dentre os códigos dos Apêndices e da \massa.
 
 O \emph{PPEPPS} (Pure Python EP: Projeto Solvente) é um EP sintetizado com os recursos apresentados neste trabalho.
-Com pouca parametrização, o programa resulta músicas inteiras, permitindo a composição de músicas e conjuntos de músicas com facilidade.
-As possibilidades de compartilhamento envolvem obter este album (ou outro feito de forma semelhante) através depoucas linhas de código e, pela execução, obter uma pasta com as músicas.
+Com pouca parametrização, o programa gera músicas inteiras, permitindo a composição de músicas e conjuntos de músicas com facilidade.
+As possibilidades de compartilhamento permitem obter este album (ou outro feito de forma semelhante) através de poucas linhas de código e, pela execução, obter uma pasta com as músicas.

### 2012-11-27 codigoProcedimentos.tex
feita toda correcao ateh conclusai
arrumada lista de figuras, indo para arrumação de código

@@
 \section{Códigos das relações descritas na seção~\ref{sec:notaDisc}}\label{sec:cod1}
-Todas as equações da seção~\ref{sec:notaDisc} estão descritas abaixo em Python e disponibizadas como parte da toolbox \massa.
+Todas as equações da seção~\ref{sec:notaDisc} estão descritas abaixo em Python e disponibizadas como parte da toolbox \massa.\cite{MASSA}
 
@@
 \code{equações da seção~\ref{varInternas} em código python}{scripts/secao2.2/eqs2.2.py}
+
+\section{Códigos das relações descritas na seção~\ref{notasMusica}}\label{sec:cod3}
+Todas as equações da seção~\ref{notasMusica} estão descritas abaixo em Python e disponibizadas como parte da toolbox \massa.
+
+\code{equações da seção~\ref{notasMusica} em código python}{scripts/secao2.3/eqs2.3.py}

### 2012-11-27 devEresultados.tex
arrumada lista de figuras, indo para arrumação de código

@@
 
-\begin{equation}
-\begin{split}
-\{a_i\}_0^{\Lambda_A-1} & = \left\{\xi\left(\frac{1}{\xi}\right)^{\frac{i}{\Lambda_A-1}}\right\}_0^{\Lambda_A-1} \\
-\{a_i\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} & =\left\{a_S^{\frac{i-\Lambda_A}{\Lambda_D-1}}  \right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} \\
+\begin{equation}\label{eq:adsr}
+\begin{split}
+\{a_i\}_0^{\Lambda_A-1} & = \left\{\xi\left(\frac{1}{\xi}\right)^{\frac{i}{\Lambda_A-1}}\right\}_0^{\Lambda_A-1} \;\;\quad\quad\quad \text{ou} \;\;\quad \left\{\frac{i}{\Lambda_A-1}\right\}_0^{\Lambda_A}\\
+\{a_i\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} & =\left\{a_S^{\frac{i-\Lambda_A}{\Lambda_D-1}}  \right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} \;\;\quad \quad\quad \text{ou} \quad\;\; \left\{1-(1-a_S)\frac{i-\Lambda_A}{\Lambda_D-1}\right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1}\\
 \{ a_i \}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} & =\left\{ a_S \right\}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} \\
-\{ a_i \}_{\Lambda-\Lambda_R}^{\Lambda-1} & =\left\{ a_S\left(\frac{\xi}{a_S} \right)^{\frac{i-(\Lambda-\Lambda_R)}{\Lambda_R-1}} \right\}_{\Lambda-\Lambda_R}^{\Lambda-1}
+\{ a_i \}_{\Lambda-\Lambda_R}^{\Lambda-1} & =\left\{ a_S\left(\frac{\xi}{a_S} \right)^{\frac{i-(\Lambda-\Lambda_R)}{\Lambda_R-1}} \right\}_{\Lambda-\Lambda_R}^{\Lambda-1} \quad\;\; \text{ou} \quad\;\; \left\{ a_S - a_S\frac{i+\Lambda_R-\Lambda}{\Lambda_R-1}\right\}_{\Lambda-\Lambda_R}^{\Lambda-1}
 \end{split}
@@
 $\xi=10^{\frac{-80}{20}}=10^{-4}\;$ ou $\;\xi=10^{\frac{-40}{20}}=10^{-2}$.
-Quanto menor for $\xi$ mais lento é o \emph{fade}, a exemplo de $\alpha$ da figura~\ref{fig:transicao} .
-
-\begin{equation}
-\left\{t_i^{ADSR}\right\}_0^{\Lambda-1} =\{t_i . a_i\}_0^{\Lambda-1}
-\end{equation}
-
-\begin{figure}[htpq!]
-    \centering
-        \includegraphics[width=\textwidth]{figuras/adsr}
-    \caption{Envoltória ADSR canônica e uma sequência sonora arbitrária sendo submetida à envoltória}
-\end{figure}
-        \label{fig:adsr}
-
+Quanto menor for $\xi$ mais lento é o \emph{fade}, a exemplo de $\alpha$ da figura~\ref{fig:transicao}.
+Já os termos do lado direito de~\ref{eq:adsr} podem realizar as entrada e saída do som a partir da intensidade zero, por serem lineares.
 Esquematicamente, a figura~\ref{fig:adsr} mostra esta envoltória ADSR, 
@@
 e na \massa, explora diversas destas configurações da envoltória ADSR
+
+\begin{equation}
+\left\{t_i^{ADSR}\right\}_0^{\Lambda-1} =\{t_i . a_i\}_0^{\Lambda-1}
+\end{equation}
+
+\begin{figure}[htpq!]
+    \centering
+        \includegraphics[width=\textwidth]{figuras/adsr}
+    \caption{Envoltória ADSR e uma sequência sonora arbitrária sendo submetida à envoltória. A variação linear de amplitude está acima. Abaixo a variação de amplitude é exponencial.}
+\end{figure}
+        \label{fig:adsr}
 
@@
 Cada nível $j$ possui alguns índices $i$. Sempre que $i$ soma três 
-(ou múltiplo de três) índices há uma relação \emph{perfecta}, 
-quando soma dois, quatro ou oito índices há uma relação \emph{imperfecta}, como na figura~\ref{fig:pulsoSubAgl}.
+(ou múltiplo de três) índices há uma relação perfeita, 
+quando soma dois, quatro ou oito índices há uma relação imperfeita, como na figura~\ref{fig:pulsoSubAgl}.
 

### 2012-11-27 scripts/adsr.py

@@ -43,7 +43,7 @@ p.text(2*91,.3,"R",fontsize=48,color='red')
 som=n.random.random(De)*2-1
 som=n.sin(n.linspace(0,45*2*n.pi,De,endpoint=False))
 som=som*s
-p.plot(som,'c*',markersize=9,label=u"Som amostrado submetido à envoltória ADSR")
+p.plot(som,'c*',markersize=9,label=u"Som amostrado submetido à envoltória ADSR com variação linear")
 p.plot(som,'c')
 p.plot((len(som),len(som)),(-.8,0),"k--")
 
@@ -54,19 +54,102 @@ p.ylim(-1.2,1.2)
 p.yticks((-1,0,SS,1),(-1,0,r"$a_S$",1),fontsize=26)
 p.xticks((len(som),),(r"$\Lambda$",),fontsize=26)
 
-
-p.arrow(0,-1.1,DA,0,length_includes_head=True,shape='full',head_width=.1,color='cyan'     ,width=0.009)
-p.arrow(DA,-1.1,-DA,0,length_includes_head=True,shape='full',head_width=.1,color='blue'   ,width=0.009)  
-p.arrow(DA+DD,-1.1,-DD,0,length_includes_head=True,shape='full',head_width=.1,color='cyan',width=0.009)
-p.arrow(DA,-1.1,DD,0,length_includes_head=True,shape='full',head_width=.1,color='green'   ,width=0.009)
-p.arrow(De,-1.1,-DR,0,length_includes_head=True,shape='full',head_width=.1,color='red'    ,width=0.009)
-p.arrow(De-DR,-1.1,DR,0,length_includes_head=True,shape='full',head_width=.1,color='red'  ,width=0.009)
-
-p.text(2*6+3 ,-1,r"$\Lambda_A$",fontsize=38,color='blue')
-p.text(2*32-9,-1,r"$\Lambda_D$",fontsize=38,color='green')
-p.text(2*91-7,-1,r"$\Lambda_R$",fontsize=38,color='red')
-
 p.plot((-20,7),(SS,SS),'k--')
+###############################
+######################
+
+De=2*100. #duracao total (\Delta)
+DA=2*20. # duracao do ataque \Delta_A
+DD=2*20. # duracao do decay \Delta_D
[…]

### 2012-11-27 resumoeabstract.tex
arrumada lista de figuras, indo para arrumação de código

@@
 
-The representation of music's basic elements - such as notes, ornaments and intervalar structures - in terms of dicrete-time audio is used in software and routines for musical creation and sound design.
-Even so, there is no unified approach that relates directly this elements to the sound discrete-time samples.
-In this dissertation, each musical element is described by equations that result directly on the sonic time samples.
+The representation of the basic elements of music - such as notes, ornaments and intervalar structures - in terms of discrete audio signal is often used in software for music creation and design.
+Nevertheless, there is no unified approach that relates these elements to the sound discrete samples.
+In this dissertation, each musical element is described by equations that represent the sonic time samples, which are then implemented in scripts within a software toolbox, referred to as \massa\ (Music and Audio in Sequences and Samples).
 The fundamental element, the musical note with duration, volume, pitch and timbre, is related quantitatively to the characteristics of the digital signal.
-Internal variations, such as tremolos, vibratos and spectral fluctuations are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities.
-With this representation of notes, we predispose canonical resources for the generation of musical structures, such as rhythmic meter, pitch intervals and cycles.
-The equations gave rise to a computer toolbox: scripts that accomplish each equation and result simple sonic examples.
-We name \mass, Music and Audio in Sequences and Samples, this resources and its efficacy was confirmed by the synthesis of small musical pieces using basic notes, incremented notes and notes in music.
-It is possible, as well, to synthesize whole albums through collage of the scripts and parametrization specified by the user.
-With the paradigm of open source implementation, the \mass\ toolbox can be expanded in processes of co-authorship and used freely by musicians, engineers and other interested parties.
-In fact, the system was already employed by external users for diverse purposes which include music production, artistic presentations, psicoacoustic experiments and computer language diffusion by means of the playful appeal of audiovisual artifacts.
+Internal variations, such as tremolos, vibratos and spectral fluctuations, are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities.
+With this representation of notes, resources are provided for the generation of musical structures, such as rhythmic meter, pitch intervals and cycles.
+The efficacy of \massa\ was confirmed by the synthesis of small musical pieces using basic notes, incremented notes and notes in music.
+Also possible is to synthesize whole albums through collage of the scripts and parameterization specified by the user.
+With the paradigm of open source implementation, \massa toolbox can be expanded in co-authorship processes and used freely by musicians, engineers and other interested parties.
+In fact, \massa has already been employed by external users for diverse purposes which include music production, artistic presentations, psychoacoustic experiments and computer language diffusion where the appeal of audiovisual artifacts is exploited.
 

### 2012-11-27 codigoProcedimentos.tex
inserido todo codigo da secao 2.1

@@
 
-\section{Códigos das relações descritas na seção~\ref{sec:notaDisc}}\label{sec:cod1}
-Todas as equações da seção~\ref{sec:notaDisc} estão descritas abaixo em Python e disponibizadas como parte da toolbox \massa.\cite{MASSA}
+\section{Código Python das relações descritas na seção~\ref{sec:notaDisc}}\label{sec:cod1}
+Todas as equações da seção~\ref{sec:notaDisc} estão implementadas abaixo em Python e disponibizadas como parte da toolbox \massa.\cite{MASSA}
 
-\code{equações da seção~\ref{sec:notaDisc} em código python}{scripts/secao2.1/eqs2.1.py}
+\code{equações da seção~\ref{sec:notaDisc} em Python}{scripts/secao2.1/eqs2.1.py}
 

### 2012-11-27 codigoProcedimentos.tex
revisto codigos todos 2.1 e 2.2

@@
 
-\section{Códigos das relações descritas na seção~\ref{varInternas}}\label{sec:cod2}
-Todas as equações da seção~\ref{varInternas} estão descritas abaixo em Python e disponibizadas como parte da toolbox \massa.
+\clearpage
 
-\code{equações da seção~\ref{varInternas} em código python}{scripts/secao2.2/eqs2.2.py}
+\section{Código Python das relações descritas na seção~\ref{varInternas}}\label{sec:cod2}
+Todas as equações da seção~\ref{varInternas} estão implementadas abaixo em Python e disponibizadas como parte da toolbox \massa.\cite{MASSA}
 
-\section{Códigos das relações descritas na seção~\ref{notasMusica}}\label{sec:cod3}
-Todas as equações da seção~\ref{notasMusica} estão descritas abaixo em Python e disponibizadas como parte da toolbox \massa.
+\code{equações da seção~\ref{varInternas} em Python}{scripts/secao2.2/eqs2.2.py}
 
-\code{equações da seção~\ref{notasMusica} em código python}{scripts/secao2.3/eqs2.3.py}
+\clearpage
+
+\section{Código Python das relações descritas na seção~\ref{notasMusica}}\label{sec:cod3}
+Todas as equações da seção~\ref{notasMusica} estão implementadas abaixo em Python e disponibizadas como parte da toolbox \massa.\cite{MASSA}
+
+\code{equações da seção~\ref{notasMusica} em Python}{scripts/secao2.3/eqs2.3.py}

### 2012-11-27 scripts/secao2.1/eqs2.1.py

@@ -73,9 +73,9 @@ Tf_i=1.-n.abs(2.-(4./lambda_f)*(ii%lambda_f))
 ### 2.13 Onda quadrada
 Qf_i=((ii%lambda_f)<(lambda_f/2))*2-1
 
-r=a.wavread("22686__acclivity__oboe-a-440_periodo.wav")[0]
+Rf_i=a.wavread("22686__acclivity__oboe-a-440_periodo.wav")[0]
 ### 2.14 Período amostrado
-Tf_i=r[n.int64(ii)%len(r)]
+Tf_i=Rf_i[n.int64(ii)%len(Rf_i)]
 
 
 ############## 2.1.5 O espectro no som amostrado

### 2012-11-27 devEresultados.tex
inserido todo codigo da secao 2.1
revisto codigos todos 2.1 e 2.2

@@
 \begin{equation}\label{equivalenciasFreqs}
-f_{k}\equiv f_{\Lambda-k}\;, \;\; w_{k}=-w_{\Lambda-k}\;\;\;, \quad \;\; \forall \quad 1 \leq k \leq \tau  
+f_{k}\equiv f_{\Lambda-k}\;, \;\; w_{k}\equiv-w_{\Lambda-k}\;\;\;, \quad \;\; \forall \quad 1 \leq k \leq \tau  
 \end{equation}
@@
 
-Para resultar em uma amplificação linear, mas não linear para a percepção, basta uma sequência $\{a_i\}$ adequada:
+Para uma amplificação linear, mas não linear para a percepção, basta usar uma sequência $\{a_i\}$ adequada:
 
@@
 Aqui convém a conversão de decibels para amplitude. Assim, as equações ~\ref{ampDec} e \ref{transAmp}
-resultam na transição de $V_{dB}$ decibels:
+especificam a transição de $V_{dB}$ decibels:
 

### 2012-11-27 scripts/secao2.2/eqs2.2.py

@@ -24,6 +24,7 @@ D_i=n.linspace(-1,1,Lt)
 
 # som real, importar período e
 # usar T correto: o número de amostras do período
+Rf_i=a.wavread("22686__acclivity__oboe-a-440_periodo.wav")[0]
 
 f=110. # Hz
 Delta=3.4 # segundos
@@ -32,41 +33,44 @@ Lambda=int(Delta*f_a)
 # Amostras:
 ii=n.arange(Lambda)
 
-### 2.30
+### 2.32 LUT
 Gamma_i=n.array(ii*f*Lt/f_a,dtype=n.int)
-L_i=Tr_i # podemos usar igualmente S_i, Q_i, D_i ou qualquer período de som real suficientemente grande
+# Pode-se usar S_i, Q_i, D_i ou qualquer período de som real
+# suficientemente grande
+L_i=Tr_i 
 TfD_i=L_i[Gamma_i%Lt]
 
 
 ############## 2.2.2 Variações incrementais de frequência e amplitude
 # VARIAÇÕES DE FREQUÊNCIA
 f_0=100. # freq inicial em Hz
-f_f=300. # frq final em Hz
+f_f=300. # freq final em Hz
 Delta=2.4 # duração
 
 Lambda=int(f_a*Delta)
 ii=n.arange(Lambda)
-### 2.31 - variação linear
+### 2.33 - variação linear
 f_i=f_0+(f_f-f_0)*ii/(float(Lambda)-1)
-### 2.32
+### 2.34 coeficientes para a LUT
 D_gamma_i=f_i*Lt/f_a
 Gamma_i=n.cumsum(D_gamma_i)
 Gamma_i=n.array(Gamma_i,dtype=n.int)
-### 2.33
+### 2.35 som resultante
 Tf0ff_i=L_i[Gamma_i%Lt]
 
-### 2.34 - variação exponencial
+#### 2.36 - variação exponencial
 f_i=f_0*(f_f/f_0)**(ii/(float(Lambda)-1))
-### 2.35
+### 2.37 coeficientes para a LUT
 D_gamma_i=f_i*Lt/f_a
 Gamma_i=n.cumsum(D_gamma_i)
 Gamma_i=n.array(Gamma_i,dtype=n.int)
-### 2.36
+### 2.38 som resultante
 Tf0ff_i=L_i[Gamma_i%Lt]
 
 
 # VARIAÇÕES DE AMPLITUDE
[…]

### 2012-11-27 codigoPecas.tex
vista todas as pecas da 2.1

@@
 \label{cap:codigoPecas}
-Todas as peças a seguir se propõem a exemplificar as relações apresentadas no capítulo~\ref{cap:resultados} e são disponibilizadas online junto ao toolbox \massa.\cite{MASSA}
+Todas as peças a seguir foram feitas para exemplificar as relações apresentadas no capítulo~\ref{cap:resultados} e são disponibilizadas online junto ao toolbox \massa.\cite{MASSA}
 
@@
 Montagem musical 'quadros sonoros' para demonstração da mixagem pela soma direta de sequências amostrais.
-Resultam em 5 pequenas peças de sonoridades estáticas.
+Esta rotina sintetiza 5 pequenas peças de sonoridades estáticas.
 Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:notaDisc}.
@@
 Resulta em uma pequena peça de 25 segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:notaDisc} e disponibilizada online junto ao toolbox \massa.
-\code{reduced-fi}{scripts/pecas2.1/reduced-fi-limpo.py}
+Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:notaDisc} e disponibilizada online junto ao toolbox \massa.\cite{MASSA}
+\code{reduced-fi}{scripts/pecas2.1/reduced-fi.py}
 

### 2012-11-27 scripts/pecas2.1/quadrosSonoros.py

@@ -2,23 +2,22 @@
 import numpy as n, scikits.audiolab as a
 
 # Peça musical baseada em sons estáticos
-# são mixagens de sons básicos apresentados na sessão 2.1
+# são mixagens de sons básicos apresentados na seção 2.1
 
 f_a=44100 # 44.1kHz, frequência de amostragem de CDs
-Delta=30. # cada quadro terá exatos 180 segundos == 3 minutos
-Lambda=Delta*f_a # número de amostras
+Delta=180. # cada quadro terá Delta segundos
+Lambda=int(Delta*f_a) # número de amostras
 
 ii=n.linspace(0,Delta*2*n.pi,Lambda,endpoint=False)
 
-# frequencias que dividem f_a
+# frequências que dividem f_a
 fs=[]
 for i in xrange(1,f_a/2+1):
     if f_a/float(i)==int(f_a/i):
         fs.append(i)
 
-### Quadro 1: senoides no grave em batimento e no agudo uma
-# dente de serra bem suave
-
+### Quadro 1: senoides no grave em batimento e
+### no agudo uma dente de serra bem suave
 f1=100
 f2=100.5
 
@@ -26,7 +25,7 @@ som=n.sin(ii*f1)+n.sin(ii*f2)
 dente_aguda=(n.arange(Lambda)%4-2)
 som+=dente_aguda/80
 
-# normalizando no intervalo [-1,1]
+# normalição no intervalo [-1,1]
 som=((som -som.min())/(som.max()-som.min()))*2-1
 
 a.wavwrite(som,"quadro1.wav",f_a)
@@ -54,7 +53,7 @@ em uma frequência de amostragem de %iHz" % (len(som),f_a))
 
 ### Quadro 3: estereofonia alternada no espectro harmônico
 f=50.
-fs3=[f*i for i in xrange(1,7)] # 6 harmonicos
+fs3=[f*i for i in xrange(1,7)] # 6 harmônicos
 som_d=n.zeros(Lambda)
 som_e=n.zeros(Lambda)
 ii=n.linspace(0,Delta*2*n.pi,Lambda,endpoint=False)
@@ -110,7 +109,7 @@ print("quadro 4 feito em quadro4.wav (estéreo), são %i amostras \
 em uma frequência de amostragem de %iHz" % (len(som),f_a))
 
 
-### Quadro 5: Dente de serra grave bate com harmonico em cada lado
[…]

### 2012-11-27 codigoPecas.tex
feito tudo ateh tremolos, vibratos e a freq

@@
 Pequena peça musical para demonstração da concatenação de sequências amostrais como notas musicais.
-Resulta em uma pequena peça de 25 segundos.
+Sintetiza uma pequena peça de 25 segundos em Python puro, i.e.
+sem utilizar bibliotecas externas como Numpy ou Scikits/Audiolab.
 Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:notaDisc} e disponibilizada online junto ao toolbox \massa.\cite{MASSA}
@@
 \subsection{Transita para metro}\label{ap:transita}
-Montagem musical 'transita para metro' para . Resultam em 1 pequenas peças com chirps. Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:varInternas}.
+Sintetiza em 1 pequena peça de 49 segundos com varreduras de frequência, chamados chirps.
+Também é proposta para explorações de transições de intensidade.
+Peça demonstrativa dos conceitos apresentados na subseção~\ref{subsec:vars} e disponibilizada junto à \massa.\cite{MASSA}
 \code{Transita para metro}{scripts/pecas2.2/transitaParaMetro.py}
@@
 \subsection{Vibra e treme}\label{ap:vibra}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
-\code{Tremolos, vibratos e a frequência}{scripts/pecas2.2/vibraEtreme.py}
+Para explorações de tremolos e vibratos como expostos na subseção~\ref{subsec:tvaf}, sintetiza 17 pequenas montagens de 8 a 24 segundos cada.
+Script disponibilizado online junto ao toolbox \massa.\cite{MASSA}
+\code{Vibra e treme}{scripts/pecas2.2/vibraEtreme.py}
 

### 2012-11-27 scripts/pecas2.2/vibraEtreme.py

@@ -53,11 +53,13 @@ def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
 
     Gammav_i=n.floor(ii*fv*Lv/f_a) # índices para a LUT
     Gammav_i=n.array(Gammav_i,n.int)
-    Tv_i=tabv[Gammav_i%int(Lv)] # padrão de variação do vibrato para cada amostra
+    # padrão de variação do vibrato para cada amostra
+    Tv_i=tabv[Gammav_i%int(Lv)] 
 
-    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) # frequência em Hz em cada amostra
-
-    D_gamma_i=F_i*(Lt/float(f_a)) # a movimentação na tabela por amostra
+    # frequência em Hz em cada amostra
+    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) 
+    # a movimentação na tabela por amostra
+    D_gamma_i=F_i*(Lt/float(f_a)) 
     Gamma_i=n.cumsum(D_gamma_i) # a movimentação na tabela total
     Gamma_i=n.floor( Gamma_i) # já os índices
     Gamma_i=n.array( Gamma_i, dtype=n.int) # já os índices
@@ -121,7 +123,8 @@ cc=v(d=7,f=300,fv=3,nu=4)
 cc2=v(d=7,f=300,fv=3,nu=4,tabv=Tr_i)
 dd=v(tabv=Tr_i,d=7)
 dd2=v(d=7)
-a.wavwrite(V((H((aa2,aa4,bb,cc2,dd2)), H((aa,aa3,bb,cc,dd)))).T,"vbr8.wav",f_a)
+a.wavwrite(V((H((aa2,aa4,bb,cc2,dd2)), H((aa,aa3,bb,cc,dd)))).T,\
+                                                 "vbr8.wav",f_a)
 
 
 aa=v(f=100, fv=3,nu=5,tabv=D_i)
@@ -133,7 +136,8 @@ cc=v(d=7,f=300,fv=3,nu=14)
 cc2=v(d=7,f=300,fv=3,nu=4,tabv=Tr_i)
 dd=v(tabv=Tr_i,d=7)
 dd2=v(tabv=D_i,d=7)
[…]

### 2012-11-27 codigoPecas.tex
terminadas todas as pecas da 2.2

@@
 
+\clearpage
+
 \subsection{Reduced-fi}\label{ap:reduced}
@@
 
+\clearpage
+
 \subsection{Transita para metro}\label{ap:transita}
@@
 
+\clearpage
 \subsection{Vibra e treme}\label{ap:vibra}
@@
 
+\clearpage
+
 \subsection{Tremolos, vibratos e a frequência}\label{ap:tremolos}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
+Pequena montagem musical para demonstração de vínculo entre os parâmetros dos tremolos e vibratos com a frequência fundamental central da nota.
+Resulta em 19 pequenas montagens de 4-32 segundos.
+Peça demonstrativa dos conceitos apresentados na subseção~\ref{subsec:mus2} e disponibilizada online junto ao toolbox \massa.\cite{MASSA}
 \code{Tremolos, vibratos e a frequência}{scripts/pecas2.2/TremolosVibratosEaFrequencia.py}
 
+\clearpage
+
 \subsection{Trenzinho de caipiras impulsivos}\label{ap:trenzinho}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
-\code{Trenzinho de caipiras impulsivos}{scripts/pecas2.2/trenzinhoCaipiraImpulsivos.py}
+Pequena peça musical para demonstração do deslocamento causado pela convolução com o impulso.
+Sintetiza 11 pequenas montages cada uma com duração de 4-240 segundos.
+Peça demonstrativa dos conceitos apresentados na subseção~\ref{subsec:filtros} e disponibilizada online junto ao toolbox \massa.
+\code{trenzinho de caipiras impulsivos}{scripts/pecas2.2/trenzinhoCaipiraImpulsivos.py}
+
+\clearpage
 
 \subsection{Ruidosa faixa}\label{ap:ruidosa}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
-\code{Ruidosa faixa}{scripts/pecas2.2/ruidosaFaixa4.py}
+Pequena peça musical de 240 segundo spara demonstração de filtragens diversas em ruídos e para reverberação, conceitos apresentados na subseção~\ref{subsec:filtros} e disponibilizada online junto ao toolbox \massa.\cite{MASSA}
+\code{ruidosa faixa}{scripts/pecas2.2/ruidosaFaixa4.py}
+
+\clearpage
 
 \subsection{Bela Rugosi}\label{ap:bela}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
-\code{Bela Rugosi}{scripts/pecas2.2/bellaRugosi.py}
+Pequena peça musical de 96 segundos para demonstração da rugosidade causada por oscilações com frequencias $\approx 13-30 Hz$, conceitos apresentados na subseção~\ref{subsec:tvaf}.
+A peça é também disponibilizada online junto à toolbox \massa.\cite{MASSA}
+\code{bela rugosi}{scripts/pecas2.2/bellaRugosi.py}
+
+\clearpage
 
 \subsection{Chorus infantil}\label{ap:chorus}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
-\code{Bela Rugosi}{scripts/pecas2.2/chorusInfantilb.py}
+Script para demonstração do efeito \emph{chorus} (inspirado em coro de cantores) .
+Sintetiza 4 pequenas montagens de 8-32 segundos com os conceitos apresentados na subseção~\ref{subsec:mus2} e disponibilizada online junto ao toolbox \massa.\cite{MASSA}
+\code{chorus infantil}{scripts/pecas2.2/chorusInfantil.py}
+
+\clearpage
 
@@
 
+\clearpage
 \subsection{Intervalos entre alturas}\label{ap:intervalos}
@@
 
+\clearpage
 \subsection{Acorde cedo}\label{ap:acorde}
@@
 \code{Acorde cedo}{scripts/pecas2.3/acordeCedo.py}
+
+\clearpage
 
@@
 
+\clearpage
+
 \subsection{Cristais}\label{ap:cristais}
@@
 \code{Cristais}{scripts/pecas2.3/cristais.py}
+
+\clearpage
 
@@
 
+\clearpage
+
 \subsection{Poli Hit Mia}\label{ap:poli}
@@
 
+\clearpage
+
 \subsection{Dirracional}\label{ap:dirracional}

### 2012-11-27 scripts/pecas2.2/chorusInfantil.py

@@ -6,7 +6,7 @@ V=n.vstack
 f_a = 44100. # Hz, frequência de amostragem
 
 ############## 2.2.1 Tabela de busca (LUT)
-Lambda_tilde=Lt=1024.
+Lambda_tilde=Lt=1024.*16
 
 # Senoide
 foo=n.linspace(0,2*n.pi,Lt,endpoint=False)
@@ -26,15 +26,17 @@ D_i=n.linspace(-1,1,Lt)
 def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
     Lambda=n.floor(f_a*d)
     ii=n.arange(Lambda)
-    Lv=float(len(S_i))
+    Lv=float(len(tabv))
 
     Gammav_i=n.floor(ii*fv*Lv/f_a) # índices para a LUT
     Gammav_i=n.array(Gammav_i,n.int)
-    Tv_i=tabv[Gammav_i%int(Lv)] # padrão de variação do vibrato para cada amostra
+    # padrão de variação do vibrato para cada amostra
+    Tv_i=tabv[Gammav_i%int(Lv)] 
 
-    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) # frequência em Hz em cada amostra
-
-    D_gamma_i=F_i*(Lt/float(f_a)) # a movimentação na tabela por amostra
+    # frequência em Hz em cada amostra
+    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) 
+    # a movimentação na tabela por amostra
+    D_gamma_i=F_i*(Lt/float(f_a))
     Gamma_i=n.cumsum(D_gamma_i) # a movimentação na tabela total
     Gamma_i=n.floor( Gamma_i) # já os índices
     Gamma_i=n.array( Gamma_i, dtype=n.int) # já os índices
@@ -46,11 +48,10 @@ def A(fa=2.,V_dB=10.,d=2.,taba=S_i):
     Lt=float(len(taba))
     Gammaa_i=n.floor(ii*fa*Lt/f_a) # índices para a LUT
     Gammaa_i=n.array(Gammaa_i,n.int)
-### 2.55 padrão de oscilação do vibrato
[…]

### 2012-11-27 devEresultados.tex
feito todo cógido das sessões, indo para as peças
feito tudo ateh tremolos, vibratos e a freq
terminadas todas as pecas da 2.2

@@
 
-\subsection{Variações incrementais de frequência e intensidade}
+\subsection{Variações incrementais de frequência e intensidade}\label{subsec:vars}
 
@@
 
-\subsection{Aplicação de filtros digitais}
+\subsection{Aplicação de filtros digitais}\label{subsec:filtros}
 Esta subseção limita-se a uma descrição
@@
 
-\subsection{Usos musicais - Parte 2}
+\subsection{Usos musicais - Parte 2}\label{subsec:mus2}
 A este ponto as possibilidades musicais explodiram. As
@@
 
-\begin{center}
-{\tiny 
+\begin{table}[htpq!]
+\tiny
+\centering
 \begin{tabular}{  l | r r r r   r r r    r r r || r r || r r r r r r }
@@
 \hline
-%\ca
 \end{tabular}
-}
-\end{center}
+\caption{Transição das durações ouvidas individualmente para alturas.}
+\label{tab:duracoes}
+\end{table}
 

### 2012-11-27 scripts/pecas2.2/TremolosVibratosEaFrequencia.py

@@ -26,15 +26,17 @@ D_i=n.linspace(-1,1,Lt)
 def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
     Lambda=n.floor(f_a*d)
     ii=n.arange(Lambda)
-    Lv=float(len(S_i))
+    Lv=float(len(tabv))
 
     Gammav_i=n.floor(ii*fv*Lv/f_a) # índices para a LUT
     Gammav_i=n.array(Gammav_i,n.int)
-    Tv_i=tabv[Gammav_i%int(Lv)] # padrão de variação do vibrato para cada amostra
+    # padrão de variação do vibrato para cada amostra
+    Tv_i=tabv[Gammav_i%int(Lv)] 
 
-    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) # frequência em Hz em cada amostra
-
-    D_gamma_i=F_i*(Lt/float(f_a)) # a movimentação na tabela por amostra
+    # frequência em Hz em cada amostra
+    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) 
+    # a movimentação na tabela por amostra
+    D_gamma_i=F_i*(Lt/float(f_a))
     Gamma_i=n.cumsum(D_gamma_i) # a movimentação na tabela total
     Gamma_i=n.floor( Gamma_i) # já os índices
     Gamma_i=n.array( Gamma_i, dtype=n.int) # já os índices
@@ -46,8 +48,8 @@ def A(fa=2.,V_dB=10.,d=2.,taba=S_i):
     Lt=float(len(taba))
     Gammaa_i=n.floor(ii*fa*Lt/f_a) # índices para a LUT
     Gammaa_i=n.array(Gammaa_i,n.int)
-### 2.55 padrão de oscilação do vibrato
-    A_i=taba[Gammaa_i%int(Lt)] # padrão de variação da amplitude do tremolo para cada amostra
+    # variação da amplitude em cada amostra
+    A_i=taba[Gammaa_i%int(Lt)] 
[…]

### 2012-11-27 scripts/pecas2.2/bellaRugosi.py

@@ -5,9 +5,9 @@ V=n.vstack
 
 f_a = 44100. # Hz, frequência de amostragem
 
-############## 2.2.1 Tabela de busca (LUT)
-Lambda_tilde=Lt=1024.
-
+# Tamanho da LUT > 2**10 para usar também em oscilacoes
+# de comprimento de onda grandes (LFO)
+Lambda_tilde=Lt=(2.**5)*1024.
 # Senoide
 foo=n.linspace(0,2*n.pi,Lt,endpoint=False)
 S_i=n.sin(foo) # um período da senóide com T amostras
@@ -26,19 +26,24 @@ D_i=n.linspace(-1,1,Lt)
 def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
     Lambda=n.floor(f_a*d)
     ii=n.arange(Lambda)
-    Lv=float(tabv)
+    Lv=float(len(tabv))
 
     Gammav_i=n.floor(ii*fv*Lv/f_a) # índices para a LUT
     Gammav_i=n.array(Gammav_i,n.int)
-    Tv_i=tabv[Gammav_i%int(Lv)] # padrão de variação do vibrato para cada amostra
+    # padrão de variação do vibrato para cada amostra
+    Tv_i=tabv[Gammav_i%int(Lv)] 
 
-    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) # frequência em Hz em cada amostra
-
-    D_gamma_i=F_i*(Lt/float(f_a)) # a movimentação na tabela por amostra
+    # frequência em Hz em cada amostra
+    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) 
+    # a movimentação na tabela por amostra
+    D_gamma_i=F_i*(Lt/float(f_a))
     Gamma_i=n.cumsum(D_gamma_i) # a movimentação na tabela total
     Gamma_i=n.floor( Gamma_i) # já os índices
     Gamma_i=n.array( Gamma_i, dtype=n.int) # já os índices
     return tab[Gamma_i%int(Lt)] # busca dos índices na tabela
+
+
+
 
 
 
@@ -122,5 +127,4 @@ aa=n.vstack(( aa, (zz+aa1).T*.5, (zz+aa2).T*.5,(zz+aa3).T*.5))
 
 print("BellaRugosiSdadE.wav escrito")
[…]

### 2012-11-27 scripts/pecas2.2/ruidosaFaixa4.py

@@ -22,7 +22,6 @@ Tr_i=n.hstack(  ( foo , foo*-1 )   )
 # Dente de Serra:
 D_i=n.linspace(-1,1,Lt)
 
-
 def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
     Lambda=n.floor(f_a*d)
     ii=n.arange(Lambda)
@@ -30,11 +29,13 @@ def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
 
     Gammav_i=n.floor(ii*fv*Lv/f_a) # índices para a LUT
     Gammav_i=n.array(Gammav_i,n.int)
-    Tv_i=tabv[Gammav_i%int(Lv)] # padrão de variação do vibrato para cada amostra
+    # padrão de variação do vibrato para cada amostra
+    Tv_i=tabv[Gammav_i%int(Lv)] 
 
-    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) # frequência em Hz em cada amostra
-
-    D_gamma_i=F_i*(Lt/float(f_a)) # a movimentação na tabela por amostra
+    # frequência em Hz em cada amostra
+    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) 
+    # a movimentação na tabela por amostra
+    D_gamma_i=F_i*(Lt/float(f_a))
     Gamma_i=n.cumsum(D_gamma_i) # a movimentação na tabela total
     Gamma_i=n.floor( Gamma_i) # já os índices
     Gamma_i=n.array( Gamma_i, dtype=n.int) # já os índices
@@ -46,8 +47,8 @@ def A(fa=2.,V_dB=10.,d=2.,taba=S_i):
     Lt=float(len(taba))
     Gammaa_i=n.floor(ii*fa*Lt/f_a) # índices para a LUT
     Gammaa_i=n.array(Gammaa_i,n.int)
-### 2.55 padrão de oscilação do vibrato
-    A_i=taba[Gammaa_i%int(Lt)] # padrão de variação da amplitude do tremolo para cada amostra
+    # variação da amplitude em cada amostra
+    A_i=taba[Gammaa_i%int(Lt)] 
     A_i=A_i*10.**(V_dB/20.)
[…]

### 2012-11-27 scripts/pecas2.2/trenzinhoCaipiraImpulsivos.py

@@ -30,15 +30,18 @@ def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
 
     Gammav_i=n.floor(ii*fv*Lv/f_a) # índices para a LUT
     Gammav_i=n.array(Gammav_i,n.int)
-    Tv_i=tabv[Gammav_i%int(Lv)] # padrão de variação do vibrato para cada amostra
+    # padrão de variação do vibrato para cada amostra
+    Tv_i=tabv[Gammav_i%int(Lv)] 
 
-    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) # frequência em Hz em cada amostra
-
-    D_gamma_i=F_i*(Lt/float(f_a)) # a movimentação na tabela por amostra
+    # frequência em Hz em cada amostra
+    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) 
+    # a movimentação na tabela por amostra
+    D_gamma_i=F_i*(Lt/float(f_a))
     Gamma_i=n.cumsum(D_gamma_i) # a movimentação na tabela total
     Gamma_i=n.floor( Gamma_i) # já os índices
     Gamma_i=n.array( Gamma_i, dtype=n.int) # já os índices
     return tab[Gamma_i%int(Lt)] # busca dos índices na tabela
+
 
 def A(fa=2.,V_dB=10.,d=2.,taba=S_i):
     Lambda=n.floor(f_a*d)
@@ -46,8 +49,8 @@ def A(fa=2.,V_dB=10.,d=2.,taba=S_i):
     Lt=float(len(taba))
     Gammaa_i=n.floor(ii*fa*Lt/f_a) # índices para a LUT
     Gammaa_i=n.array(Gammaa_i,n.int)
-### 2.55 padrão de oscilação do vibrato
-    A_i=taba[Gammaa_i%int(Lt)] # padrão de variação da amplitude do tremolo para cada amostra
+    # variação da amplitude em cada amostra
+    A_i=taba[Gammaa_i%int(Lt)] 
     A_i=A_i*10.**(V_dB/20.)
     return A_i
 
[…]

### 2012-11-27 codigoPecas.tex
feitas todas as pecas menos FIGGUS

@@
 
+\section{Peças referentes à seção~\ref{sec:notaDisc}}
+
 \subsection{Quadros sonoros}\label{ap:quadros}
-Montagem musical 'quadros sonoros' para demonstração da mixagem pela soma direta de sequências amostrais.
+Montagem musical 'quadros sonoros' para demonstração da mixagem pela soma direta de sequências amostrais (subseção~\ref{subsec:basMus}).
 Esta rotina sintetiza 5 pequenas peças de sonoridades estáticas.
@@
 \subsection{Reduced-fi}\label{ap:reduced}
-Pequena peça musical para demonstração da concatenação de sequências amostrais como notas musicais.
+Pequena peça musical para a concatenação de sequências amostrais como notas musicais (subseção~\ref{subsec:masMus}).
 Sintetiza uma pequena peça de 25 segundos em Python puro, i.e.
@@
 \clearpage
+\section{Peças referentes à seção~\ref{varInternas}}
 
@@
 \subsection{ADa e SaRa}\label{ap:ada}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{varInternas} e disponibilizada online junto ao toolbox \massa.
+Pequenas peças musicais para demonstração da envoltória ADSR.
+Sintetiza 7 peças de 5-17 segundos, cetradas nos conceitos apresentados na subseção~\ref{subsec:mus2} e disponibilizada online junto à \massa.\cite{MASSA}
 \code{ADa e SaRa}{scripts/pecas2.2/ADa_e_SaRa.py}
 
+\clearpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\section{Peças referentes à seção~\ref{notasMusica}}
+\subsection{Intervalos entre alturas}\label{ap:intervalos}
+Pequena peça musical de 45 segundos para exploração do sistema de intervalos, demonstrativa dos conceitos apresentados na subseção~\ref{subsec:intervalos}, e disponibilizada online junto ao toolbox \massa.
+\code{intervalos entre alturas}{scripts/pecas2.3/intervalosEntreAlturas.py}
+
 \clearpage
-\subsection{Intervalos entre alturas}\label{ap:intervalos}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
-\code{Intervalos entre alturas}{scripts/pecas2.3/intervalosEntreAlturas.py}
+
+\subsection{Cristais}\label{ap:cristais}
+Pequena peça musical de 64 segundos para demonstração das escalas simétricas, conceitos apresentados na subseção~\ref{subsec:intervalos}, e disponibilizada online junto à \massa.\cite{MASSA}
+\code{cristais}{scripts/pecas2.3/cristais.py}
+
+\clearpage
+
+\subsection{Micro tom}\label{ap:micro}
+Pequena peça musical para demonstração da microtonalidade de quartos de tom e de sétimos de oitava, conceitos apresentados na subseção~\ref{subsec:intervalos}, e disponibilizada online junto à \massa.\cite{MASSA}
+\code{micro tom}{scripts/pecas2.3/microTom.py}
 
@@
 \subsection{Acorde cedo}\label{ap:acorde}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
-\code{Acorde cedo}{scripts/pecas2.3/acordeCedo.py}
+Pequena peça musical de 40 segundos para demonstração dos encadeamentos tonais básicos, conceitos apresentados na subseção~\ref{subsec:harmonia}, e disponibilizada online como parte da \massa.\cite{MASSA}
+\code{acorde cedo}{scripts/pecas2.3/acordeCedo.py}
 
@@
 \subsection{Conta ponto}\label{ap:conta}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
-\code{Cristais}{scripts/pecas2.3/contaPonto.py}
-
-\clearpage
-
-\subsection{Cristais}\label{ap:cristais}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
-\code{Cristais}{scripts/pecas2.3/cristais.py}
-
-\clearpage
-
-\subsection{Micro tom}\label{ap:micro}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
-\code{Cristais}{scripts/pecas2.3/microTom.py}
+Pequena montagem musical de 17 segundos para demonstração de vozes conduzidas por regras simples para manterem independência, conceitos apresentados na subseção~\ref{subsec:contraponto}, e disponibilizada online junto à \massa.\cite{MASSA}
+\code{conta ponto}{scripts/pecas2.3/contaPonto.py}
 
@@
 \subsection{Poli Hit Mia}\label{ap:poli}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
-\code{Cristais}{scripts/pecas2.3/poliHitMia.py}
+Pequena montagem musical de 99 segundos para demonstração da métrica musical e das diferentes contagens, conceitos apresentados na seção~\ref{subsec:ritmo}, e disponibilizada online junto à \massa.\cite{MASSA}
+\code{poli hit mia}{scripts/pecas2.3/poliHitMia.py}
 
@@
 \subsection{Dirracional}\label{ap:dirracional}
-Pequena peça musical para demonstração da .
-Resulta em uma pequena peça de XX segundos.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{notasMusica} e disponibilizada online junto ao toolbox \massa.
-\code{Cristais}{scripts/pecas2.3/dirracional.py}
+Pequena peça musical de 22 segundos para demonstração das estruturas direcionais e posicionalmentos dos clímax, conceitos apresentados na seção~\ref{notasMusica}, e disponibilizada online junto à \massa.\cite{MASSA}
+\code{dirracional}{scripts/pecas2.3/dirracional.py}

### 2012-11-27 scripts/pecas2.3/acordeCedo.py

@@ -22,19 +22,20 @@ Tr_i=n.hstack(  ( foo , foo*-1 )   )
 # Dente de Serra:
 D_i=n.linspace(-1,1,Lt)
 
-
 def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
     Lambda=n.floor(f_a*d)
     ii=n.arange(Lambda)
-    Lv=float(len(S_i))
+    Lv=float(len(tabv))
 
     Gammav_i=n.floor(ii*fv*Lv/f_a) # índices para a LUT
     Gammav_i=n.array(Gammav_i,n.int)
-    Tv_i=tabv[Gammav_i%int(Lv)] # padrão de variação do vibrato para cada amostra
+    # padrão de variação do vibrato para cada amostra
+    Tv_i=tabv[Gammav_i%int(Lv)] 
 
-    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) # frequência em Hz em cada amostra
-
-    D_gamma_i=F_i*(Lt/float(f_a)) # a movimentação na tabela por amostra
+    # frequência em Hz em cada amostra
+    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) 
+    # a movimentação na tabela por amostra
+    D_gamma_i=F_i*(Lt/float(f_a))
     Gamma_i=n.cumsum(D_gamma_i) # a movimentação na tabela total
     Gamma_i=n.floor( Gamma_i) # já os índices
     Gamma_i=n.array( Gamma_i, dtype=n.int) # já os índices
@@ -46,10 +47,11 @@ def A(fa=2.,V_dB=10.,d=2.,taba=S_i):
     Lt=float(len(taba))
     Gammaa_i=n.floor(ii*fa*Lt/f_a) # índices para a LUT
     Gammaa_i=n.array(Gammaa_i,n.int)
-### 2.55 padrão de oscilação do vibrato
-    A_i=taba[Gammaa_i%int(Lt)] # padrão de variação da amplitude do tremolo para cada amostra
+    # variação da amplitude em cada amostra
+    A_i=taba[Gammaa_i%int(Lt)] 
     A_i=A_i*10.**(V_dB/20.)
     return A_i
+
 
 
 def adsr(som,A=10.,D=20.,S=-20.,R=100.,xi=1e-2):
[…]

### 2012-11-27 scripts/pecas2.3/contaPonto.py

@@ -5,7 +5,6 @@ V=n.vstack
 
 f_a = 44100. # Hz, frequência de amostragem
 
-############## 2.2.1 Tabela de busca (LUT)
 Lambda_tilde=Lt=1024.*16
 
 # Senoide
@@ -26,15 +25,17 @@ D_i=n.linspace(-1,1,Lt)
 def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
     Lambda=n.floor(f_a*d)
     ii=n.arange(Lambda)
-    Lv=float(len(S_i))
+    Lv=float(len(tabv))
 
     Gammav_i=n.floor(ii*fv*Lv/f_a) # índices para a LUT
     Gammav_i=n.array(Gammav_i,n.int)
-    Tv_i=tabv[Gammav_i%int(Lv)] # padrão de variação do vibrato para cada amostra
+    # padrão de variação do vibrato para cada amostra
+    Tv_i=tabv[Gammav_i%int(Lv)] 
 
-    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) # frequência em Hz em cada amostra
-
-    D_gamma_i=F_i*(Lt/float(f_a)) # a movimentação na tabela por amostra
+    # frequência em Hz em cada amostra
+    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) 
+    # a movimentação na tabela por amostra
+    D_gamma_i=F_i*(Lt/float(f_a))
     Gamma_i=n.cumsum(D_gamma_i) # a movimentação na tabela total
     Gamma_i=n.floor( Gamma_i) # já os índices
     Gamma_i=n.array( Gamma_i, dtype=n.int) # já os índices
@@ -46,11 +47,10 @@ def A(fa=2.,V_dB=10.,d=2.,taba=S_i):
     Lt=float(len(taba))
     Gammaa_i=n.floor(ii*fa*Lt/f_a) # índices para a LUT
     Gammaa_i=n.array(Gammaa_i,n.int)
-### 2.55 padrão de oscilação do vibrato
-    A_i=taba[Gammaa_i%int(Lt)] # padrão de variação da amplitude do tremolo para cada amostra
+    # variação da amplitude em cada amostra
+    A_i=taba[Gammaa_i%int(Lt)] 
[…]

### 2012-11-27 scripts/pecas2.3/dirracional.py

@@ -27,15 +27,17 @@ D_i=n.linspace(-1,1,Lt)
 def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
     Lambda=n.floor(f_a*d)
     ii=n.arange(Lambda)
-    Lv=float(len(S_i))
+    Lv=float(len(tabv))
 
     Gammav_i=n.floor(ii*fv*Lv/f_a) # índices para a LUT
     Gammav_i=n.array(Gammav_i,n.int)
-    Tv_i=tabv[Gammav_i%int(Lv)] # padrão de variação do vibrato para cada amostra
+    # padrão de variação do vibrato para cada amostra
+    Tv_i=tabv[Gammav_i%int(Lv)] 
 
-    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) # frequência em Hz em cada amostra
-
-    D_gamma_i=F_i*(Lt/float(f_a)) # a movimentação na tabela por amostra
+    # frequência em Hz em cada amostra
+    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) 
+    # a movimentação na tabela por amostra
+    D_gamma_i=F_i*(Lt/float(f_a))
     Gamma_i=n.cumsum(D_gamma_i) # a movimentação na tabela total
     Gamma_i=n.floor( Gamma_i) # já os índices
     Gamma_i=n.array( Gamma_i, dtype=n.int) # já os índices
@@ -47,10 +49,11 @@ def A(fa=2.,V_dB=10.,d=2.,taba=S_i):
     Lt=float(len(taba))
     Gammaa_i=n.floor(ii*fa*Lt/f_a) # índices para a LUT
     Gammaa_i=n.array(Gammaa_i,n.int)
-### 2.55 padrão de oscilação do vibrato
-    A_i=taba[Gammaa_i%int(Lt)] # padrão de variação da amplitude do tremolo para cada amostra
+    # variação da amplitude em cada amostra
+    A_i=taba[Gammaa_i%int(Lt)] 
     A_i=A_i*10.**(V_dB/20.)
     return A_i
+
 
 
 def adsr(som,A=10.,D=20.,S=-20.,R=100.,xi=1e-2):
@@ -75,7 +78,8 @@ def adsr(som,A=10.,D=20.,S=-20.,R=100.,xi=1e-2):
     return som*A_i
 
[…]

### 2012-11-27 devEresultados.tex
feitas todas as pecas menos FIGGUS

@@
 
-\subsection{Usos musicais}
+\subsection{Usos musicais}\label{subsec:basMus}
 
@@
 logarítmicas para os tremolos e vibratos.
+Uma exploração não oscilatória destas variações
+está na montagem musical \emph{Transita para metro},
+cujo código está no Apêndice~\ref{ap:transita} e online
+na \massa.\cite{MASSA}
 
@@
 
-\subsection{Afinação, intervalos, escalas e acordes}
+\subsection{Afinação, intervalos, escalas e acordes}\label{subsec:intervalos}
 O dobro da frequência é uma oitava ascendente ($f=2f_0$).
@@
 
-\subsection{Rudimentos de harmonia}
+\subsection{Rudimentos de harmonia}\label{subsec:harmonia}
 
@@
 
-\subsection{Contraponto}
+\subsection{Contraponto}\label{subsec:contraponto}
 
@@
 
-\subsection{Ritmo}
+\subsection{Ritmo}\label{subsec:ritmo}
 A noção rítmica é depende de eventos separados por durações.\cite{rythm} Estes eventos podem ser ouvidos individualmente se separados
@@
 interessante e útil e a correlação dos arcos resulta na coerência da escuta.\cite{Schachter}
+
+A montagem sonora \emph{Dirracional} expõe estes arcos em estruturas direcionais. Seu código está no Apêndice~\ref{ap:dirracional} e disponível online como parte da \massa.\cite{MASSA}
 

### 2012-11-27 scripts/pecas2.2/ADa_e_SaRa.py

@@ -5,7 +5,6 @@ V=n.vstack
 
 f_a = 44100. # Hz, frequência de amostragem
 
-############## 2.2.1 Tabela de busca (LUT)
 Lambda_tilde=Lt=1024.
 
 # Senoide

### 2012-11-27 scripts/pecas2.3/intervalosEntreAlturas.py

@@ -22,19 +22,20 @@ Tr_i=n.hstack(  ( foo , foo*-1 )   )
 # Dente de Serra:
 D_i=n.linspace(-1,1,Lt)
 
-
-def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
+def v(f=200.,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
     Lambda=n.floor(f_a*d)
     ii=n.arange(Lambda)
-    Lv=float(len(S_i))
+    Lv=float(len(tabv))
 
     Gammav_i=n.floor(ii*fv*Lv/f_a) # índices para a LUT
     Gammav_i=n.array(Gammav_i,n.int)
-    Tv_i=tabv[Gammav_i%int(Lv)] # padrão de variação do vibrato para cada amostra
+    # padrão de variação do vibrato para cada amostra
+    Tv_i=tabv[Gammav_i%int(Lv)] 
 
-    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) # frequência em Hz em cada amostra
-
-    D_gamma_i=F_i*(Lt/float(f_a)) # a movimentação na tabela por amostra
+    # frequência em Hz em cada amostra
+    F_i=f*(   2.**(  Tv_i*nu/12.  )   ) 
+    # a movimentação na tabela por amostra
+    D_gamma_i=F_i*(Lt/float(f_a))
     Gamma_i=n.cumsum(D_gamma_i) # a movimentação na tabela total
     Gamma_i=n.floor( Gamma_i) # já os índices
     Gamma_i=n.array( Gamma_i, dtype=n.int) # já os índices
@@ -46,10 +47,11 @@ def A(fa=2.,V_dB=10.,d=2.,taba=S_i):
     Lt=float(len(taba))
     Gammaa_i=n.floor(ii*fa*Lt/f_a) # índices para a LUT
     Gammaa_i=n.array(Gammaa_i,n.int)
-### 2.55 padrão de oscilação do vibrato
-    A_i=taba[Gammaa_i%int(Lt)] # padrão de variação da amplitude do tremolo para cada amostra
+    # variação da amplitude em cada amostra
+    A_i=taba[Gammaa_i%int(Lt)] 
     A_i=A_i*10.**(V_dB/20.)
     return A_i
+
 
 
 def adsr(som,A=10.,D=20.,S=-20.,R=100.,xi=1e-2):

### 2012-11-27 scripts/secao2.3/eqs2.3.py

@@ -153,10 +153,10 @@ Vm=[2.,7.,10.] # quinto grau menor nao eh dominante, é quinto grau
 
 ############## 2.3.3 Contraponto
 
-def contraNotaNotaSup(alturas=[0,2,4,5,5,0,2,0,2,2,2,0\
-                               ,7,5,4,4,4,0,2,4,5,5,5]):
+def contraNotaNotaSup(alturas=[0,2,4,5,5,0,2,0,2,2,2,0,7,\
+                                     5,4,4,4,0,2,4,5,5,5]):
     """Realiza rotina de independência das vozes
-    
+
     Limitado em 1 oitava acima da nota"""
     primeiraNota=alturas[0]+(7,12)[n.random.randint(2)]
     contra=[primeiraNota]
@@ -164,22 +164,22 @@ def contraNotaNotaSup(alturas=[0,2,4,5,5,0,2,0,2,2,2,0\
     i=0
     cont=0 # contador de paralelas
     reg=0 # registrador de intervalo em que se fez a paralela
-    for al in alturas[-1]:
+    for al in alturas[:-1]:
         mov_cf=alturas[i:i+2]
-        atual_cf,seguinte_cf=mov_cA
+        atual_cf,seguinte_cf=mov_cf
         if seguinte_cf-atual_cf>0:
-            direct="asc"
+            mov="asc"
         elif seguinte_cf-atual_cf<0:
-            direct="asc"
+            mov="asc"
         else:
-            direct="obl"
+            mov="obl"
 
         # possibilidades por consonancia
         possiveis=[seguinte_cf+interval for interval in\
                                     [0,3,4,5,7,8,9,12]]
         movs=[]
         for pos in possiveis:
-            if pos -cotra[i] < 0:
+            if pos -contra[i] < 0:
                 movs.append("desc")
             if pos - contra[i] > 0:
                 movs.append("asc")
@@ -200,14 +200,14 @@ def contraNotaNotaSup(alturas=[0,2,4,5,5,0,2,0,2,2,2,0\
             if mt == "direto": # mov direto
                 # n aceita intervalo perfeito
                 if nota-seguinte_cf in (0,7,8,12):
-                    possiveis.remove(inter)
+                    possiveis.remove(nota)
         ok=0
         while not ok:
-            nnota=possiveis[n.random.randint(len(possiveis)]
-            if nnota-seguinte_fc==contra[i]-atual_cf: # paralelo
[…]

### 2012-11-28 FIGGUScode.tex
arrumados todos os codigos dos apendices

@@
-\chapter{Código do FIGGUS (Finite Groups in Granular and Unit Synthesis}
+\chapter{\emph{Finite Groups in Granular and Unit Synthesis} e a síntese de um EP}
 \label{cap:FIGGUScode}
+\section{\emph{FiGGUS}}
+
+Escrito como um módulo python, o \figgus\ sintetiza estruturas
+musicais através de permutações, como apresentadas na subseção~\ref{estCic}. Utiliza os princípios da \massa\
+e os códigos foram reescritos para python nativo, i.e. sem a 
+utilização de bibliotecas externas como o Numpy e o Audiolab, para facilitar
+o uso de terceiros. O \figgus\ é parte da toolbox \massa\ como exemplo de implementação dos princípios da \emph{toolbox} com as biblitecas padrão da linguagem Python.\cite{MASSA}
+
+\subsection{FIGGUS.py}\label{ap:figgus.py}
+Arquivo principal, possui todas as rotinas.
+\code{figgus/FIGGUS.py}{FIGGUS/figgus/FIGGUS.py}
+
+\subsection{tables.py}\label{ap:tables.py}
+Arquivo auxiliar para tratar as tabelas separadamente.
+\code{figgus/tables.py}{FIGGUS/figgus/tables.py}
+
+\subsection{\_\_init\_\_.py}\label{ap:init.py}
+Inicialização do módulo.
+\code{figgus/\_\_init\_\_.py}{FIGGUS/figgus/__init__.py}
+
+%%%%%%%%%%%%%%%%%%%%
+\section{PPEPPS: músicas de um EP solvente}
+O PPEPPS (Pure Python EP: Projeto Solvente) usa
+o \figgus\ para sintetizar um EP inteiro. As músicas
+estão abaixo, junto com o arquivo que executa cada uma.
+
+\subsection{RUNME make EP MUSIC.py}\label{ap:RUNME.py}
+Arquivo que executa os outros um por um para sintetizar as músicas do EP.
+\code{RUNME\_make\_now\_an\_EP\_MUSIC.py}{FIGGUS/RUNME_make_now_an_EP_MUSIC.py}
+
+\subsection{Éter}\label{ap:triangulo3b.py}
+\code{examples/triangulo3\_B.py}{FIGGUS/examples/triangulo3_B.py}
+\clearpage
+\subsection{Butano}\label{ap:triangulo4b}
+\code{examples/try4\_B.py}{FIGGUS/examples/try4_B.py}
+\clearpage
+
+\subsection{Thinner}\label{ap:triangulo5b}
+\code{examples/try5\_B.py}{FIGGUS/examples/try5_B.py}
+\clearpage
+
+\subsection{Tolueno}\label{ap:triangulo6b}
+\code{examples/try6\_B.py}{FIGGUS/examples/try6_B.py}
+\clearpage
+
+\subsection{Benzina}\label{ap:try7}
+\code{examples/try7.py}{FIGGUS/examples/try7.py}
+\clearpage
+
+\subsection{LSA}\label{ap:try3b}
+\code{examples/try3\_B.py}{FIGGUS/examples/try3_B.py}
+\clearpage
+
+\subsection{Clorofórmio}\label{ap:try2}
+\code{examples/try2.py}{FIGGUS/examples/try2.py}
+\clearpage
+
+\subsection{Água}\label{ap:try5}
+\code{examples/try5.py}{FIGGUS/examples/try5.py}
+\clearpage

### 2012-11-28 fmam.tex
arrumados todos os codigos dos apendices

@@
 
-\subsection{FM}
 \begin{equation}
@@
 
-Onde, por brevidade, citamos as identidades úteis para resultar na
-forma final do som resultante, que explicita o conteúdo em frequências puras.
-Podemos observar que, para a FM, se utilizada a progressão exponencial de frequência, o resultado
-é menos comportado. A função de Bessel dependerá não somente da profundidade de oscilação,
+Por brevidade, as identidades úteis para a abtenção
+da equação final, que explicita o conteúdo em frequências puras.
+Portanto, se utilizada a progressão exponencial de frequência,
+o espectro da FM perde a simplicidade.
+A função de Bessel dependerá não somente da profundidade de oscilação,
 mas também da frequência original do som. Além disso, as frequências dependem do produto das frequências
-do som e da oscilação ($f.f'$), o que também - a princípio - não convém por introduzir complexidade.
+da portadora e da moduladora ($f.f'$), o que também introduz complexidade.
 
-\subsection{AM}
-No caso de usarmos a modulação na escala logarítmica como nosso tremolo, temos novamente
-um resultado menos comportado:
+No caso da modulação de amplitude na escala logarítmica:
 \begin{equation}\label{eq:am}
@@
 
-Que nos leva pelos mesmos recursos a um espectro com diversas bandas de frequência distante
-em proporções múltiplas da frequência da moduladora.
+através das mesmas três identidades, observa-se um espectro com complexidade similar.

### 2012-11-28 musicaExtra.tex
arrumados todos os codigos dos apendices

@@
 
-Os resultados explicitados nesta dissertação estão focados no
-processamento de áudio através de códigos em liguagens de programação
-de computadores e voltado para a música e na síntese de estruturas
-musicais. O tema é amplo e pertinente o suficiente para contar
-atualmente com ampla bibliografia e um histórico que remonta desde a
-própria história da computação e das linguagens de programação.  Desta
-forma, entendemos cabível focar este texto nas empreitadas realizadas
-pelos autores da presente dissertação, com foco nas realizações do
-primeiro autor. São usos reais destas tecnologias, manifestadas de
+As tecnologias expostas neste trabalho tiveram
+usos reais, foram utilizadas de
 forma cultural e com repercussões sociais, como é relatado neste mesmo
 capítulo.
+Pode-se dividir estes resultados assim: experimentos abertos em áudio, música em tempo real, em tempo diferido, música na matéria e repercussões no tecido social.
 
-Tais realizações possuem um espectro amplo dentro do escopo proposto e
-constitui um percurso relativamente completo pelas possibilidades e
-decorrências da abordagem da música pelo código
-computacional\footnote{ao menos dentro do que se denomina 'Cultura Livre',
-  veja o primeiro capítulo do presente escrito para maiores
-  informações}.
-
-Estaremos relatando experimentos abertos em áudio, abordagens musicais
-pelo código, incluindo música em tempo real, diferido, na matéria e
-repercussões no tecido social. De forma a relatar os resultados destas
-investidas, segue uma sessão sobre ações coletivas realizadas através
-das mobilizações que estes códigos criaram, e as estruturas virtuais
-mantidas na Intenet e relacionadas a estas mobilizações e ações. Por
-fim, alguns materiais didáticos criados pelo autor -- que possuem
-algum papel emblemático -- são também expostos por comporem o arsenal
-de repercussões desta investida no em som e código.
+Há uma listagem abaixo das ações coletivas realizadas através
+das mobilizações que estes códigos criaram e as estruturas virtuais
+mantidas \emph{online}.
+ Por fim, alguns materiais didáticos são exibidos como consequência da abordagem da programação através do audiovisual.
 
@@
 Todos os desenvolvimentos desta dissertação estão em repositórios
-abertos\cite{repositorios-tese-dev}.  Alguns foram especialmente
-importantes como percurso para o que é apresentado neste escrito. Ou
-seja, são explorações que manifestaram diferentes aspectos do trabalho
-do código voltado para o áudio e a música.
+abertos.\cite{repoDissertacao} Alguns foram especialmente
+importantes como percurso para o que é apresentado neste escrito.
 
@@
 
-\code{Procedimento de Lookup Table}{python_snippets/lookuptable.py}
+%\code{Procedimento de Lookup Table}{python_snippets/lookuptable.py}
 
@@
 
-\code{Formas de Onda Tradicionais}{python_snippets/formasDeOnda.py}
+%\code{Formas de Onda Tradicionais}{python_snippets/formasDeOnda.py}
 
@@
 
-\code{Waveshaping com Lookups sucessivas}{python_snippets/lookup_cruz.py}
+%\code{Waveshaping com Lookups sucessivas}{python_snippets/lookup_cruz.py}
 
@@
 
-\code{Realização de Sequências de Notas}{python_snippets/fazSequencia.py}
+%\code{Realização de Sequências de Notas}{python_snippets/fazSequencia.py}
 
-\code{Realização de Acordes de Notas}{python_snippets/fazAcorde.py}
+%\code{Realização de Acordes de Notas}{python_snippets/fazAcorde.py}
 
@@
 
-\code{Somador (função auxiliar)}{python_snippets/somador.py}
+%\code{Somador (função auxiliar)}{python_snippets/somador.py}
 
@@
 
-\code{Escalas Completamente Simétricas (na grade dos 12 semitons e no âmbito da oitava)}{python_snippets/escalas_simetricas.py}
+%\code{Escalas Completamente Simétricas (na grade dos 12 semitons e no âmbito da oitava)}{python_snippets/escalas_simetricas.py}
 
@@
 
-\code{Escalas diatônicas tonais (maiores e menores)}{python_snippets/escalas_Mm.py}
+%\code{Escalas diatônicas tonais (maiores e menores)}{python_snippets/escalas_Mm.py}
 
@@
 
-\code{Escalas Microtonais (com quartos e oitavos de tom e intervalos menores)}{python_snippets/escalas_microtonais.py}
+%\code{Escalas Microtonais (com quartos e oitavos de tom e intervalos menores)}{python_snippets/escalas_microtonais.py}
 
@@
 
-\code{Acordes anotados como listas em python}{python_snippets/acordes.py}
+%\code{Acordes anotados como listas em python}{python_snippets/acordes.py}
 
@@
 
-\code{Séries diversas}{python_snippets/series.py}
+%\code{Séries diversas}{python_snippets/series.py}
 
@@
 
-\code{Sintetizando sequências e acordes}{python_snippets/seqs_acordes.py}
+%\code{Sintetizando sequências e acordes}{python_snippets/seqs_acordes.py}
 
@@
 
-\code{Acordes periódicos}{python_snippets/acordes_periodicos.py}
+%\code{Acordes periódicos}{python_snippets/acordes_periodicos.py}
 
@@
 
-\code{Amplificação e mixagem}{python_snippets/amp_mix.py}
+%\code{Amplificação e mixagem}{python_snippets/amp_mix.py}
 
@@
 
-\code{Utilizando o FIGGUS para Sintetizar um EP}{python_snippets/synth_ep.py}
+%\code{Utilizando o FIGGUS para Sintetizar um EP}{python_snippets/synth_ep.py}
 
@@
 
-\code{Grão Sonoro Básico}{python_snippets/fgrao.py}
+%\code{Grão Sonoro Básico}{python_snippets/fgrao.py}
 
@@
 
-\code{Sequência de Grãos}{python_snippets/fsequencia.py}
+%\code{Sequência de Grãos}{python_snippets/fsequencia.py}
 
@@
 
-\code{Permutações e Padrões de Permutações}{python_snippets/fpermutacoes.py}
+%\code{Permutações e Padrões de Permutações}{python_snippets/fpermutacoes.py}
 
@@
 
-\code{Realização do Padrão Musical}{python_snippets/fpadraosonoro.py}
+%\code{Realização do Padrão Musical}{python_snippets/fpadraosonoro.py}
 
@@
 
-\code{Escrita do Vetor Sonoro em Arquivo WAV}{python_snippets/fio.py}
+%\code{Escrita do Vetor Sonoro em Arquivo WAV}{python_snippets/fio.py}
 

### 2012-12-27 introducao.tex
arrumando numeracao
atualizando

@@
 %% ------------------------------------------------------------------------- %%
+\setcounter{page}{23}
 \chapter{Introdução} %Nome do capítulo.
@@
 
-    \section{Do som ao áudio digital}\label{sec:audio}
+    \section{Som em áudio digital}\label{sec:audio}
 
@@
 Assim, para apreender as frequências audíveis, é necessária uma taxa de amostragem que seja ao menos o dobro da frequência mais aguda $f_a \geq 2\times 20kHz=40kHz$.
-Este raciocínio está na base da utilização das frequências de amostragens $f_a=44.1kHz$ e $f_a=48kHz$, ambas padrão em \emph{Compact Disks} (CDs) e em sistemas de Rádio e TV, respectivamente.\cite{protocolosAudio}
+Este raciocínio está na base da utilização das frequências de amostragem $f_a=44.1kHz$ e $f_a=48kHz$, ambas padrão em \emph{Compact Disks} (CDs) e em sistemas de Rádio e TV, respectivamente.\cite{protocolosAudio}
 
@@
 Isso ocorreu na música de concerto, especialmente nas correntes concreta, eletrônica e eletroacústica.
-Já na década de 90 era evidente que também a música popular, especialmente as músicas eletrônicas de dança, tinham se desdobrado para abarcar sons sem altura definida e organizações temporais fora de métricas simples.
+Já na década de 90 era evidente que também a música popular, especialmente as músicas eletrônicas de dança, tinham incorporado sons sem altura definida e organizações temporais fora de métricas simples.
 Mesmo assim, a nota permanece paradigmática como 'unidade fundamental' das estruturas musicais e, na prática, podem se desdobrar em sons que contemplam estes desenvolvimentos recentes.

### 2012-12-27 devEresultados.tex
atualizando

@@
 
-Todo elemento $p$ operado consigo mesmo um número suficiente de vezes $n$ atinge o elemento neutro $p^n=e$, pois caso contrário com a propriedade do fechamento se prova que o grupo é infinito (gerado $p$).
+Todo elemento $p$ operado consigo mesmo um número suficiente de vezes $n$ atinge o elemento neutro $p^n=e$, pois caso contrário com a propriedade do fechamento se prova que o grupo é infinito (gerado por $p$).
 O menor $n\;:\;p^n=e$ é chamado de ordem do elemento.

### 2013-01-12 devEresultados.tex
feito apendice E ateh musica em tmepo real

@@
 
-\subsection{Tabela de busca}
+\subsection{Tabela de busca}\label{subsec:lookup}
 
@@
 
-\subsection{Ruídos}
+\subsection{Ruídos}\label{subsec:ruidos}
 De forma geral, os sons sem altura definida 

### 2013-01-12 musicaExtra.tex
feito apendice E ateh musica em tmepo real

@@
 
-As tecnologias expostas neste trabalho tiveram
-usos reais, foram utilizadas de
-forma cultural e com repercussões sociais, como é relatado neste mesmo
+As tecnologias expostas neste trabalho tiveram origem em
+usos reais, advindas
+de práticas culturais e voltadas para repercussões sociais.
+Este material humano e artístico é o assunto deste
 capítulo.
-Pode-se dividir estes resultados assim: experimentos abertos em áudio, música em tempo real, em tempo diferido, música na matéria e repercussões no tecido social.
-
-Há uma listagem abaixo das ações coletivas realizadas através
-das mobilizações que estes códigos criaram e as estruturas virtuais
+Pode-se dividir estes resultados em:
+experimentos abertos em áudio,
+música em tempo real,
+música em tempo diferido,
+música na matéria e
+repercussões no tecido social.
+Estes tópicos são tratados separadamente
+abaixo em termos do que é
 mantidas \emph{online}.
- Por fim, alguns materiais didáticos são exibidos como consequência da abordagem da programação através do audiovisual.
+Estas atividades extrapolam os usos musicais, especialmente devido às
+finalidades pedagódicas e sociais dos materiais audiovisuais.
+Assim, por completude, alguns materiais didáticos
+e ferramentas web são descritas.
 
@@
 Todos os desenvolvimentos desta dissertação estão em repositórios
-abertos.\cite{repoDissertacao} Alguns foram especialmente
-importantes como percurso para o que é apresentado neste escrito.
+abertos.\cite{repoDissertacao}
 
-O código para a arte sonora, incluindo a musical, manifesta-se como cultura pois é fruto de práticas
+O código computacional feito para a arte sonora, incluindo a musical,
+ manifesta-se como cultura pois é fruto de práticas
 espontâneas, diárias e coletivas.
 \footnote{As chamadas culturas biopunk, ciberpunk, cipherpunk, hacker,
-  digital e outras mais, possibilitadas aos recentes desenvolvimentos
+  digital e outras mais, possibilitadas pelos desenvolvimentos
   em telecomunicação, dizem respeito em menor ou maior grau à produção
-  de código como cultura.}
-
-A seguir apresentamos alguns bons exemplos dos desenvolvimentos desta
-dissertação especificamente em áudio, sem o envolvimento direto da
-música. São códigos de necessidade para música, mas de uso mais geral
-e cuja implementação requer muito mais de processamento de sinais e
-computação do que de música propriamente dito.
+  de código como cultura. As elaborações de conceitos e ferramentas
+  voltados para o compartilhamento nestas culturas estão na gênese do que
+  se entende como 'Cultura Livre'.}
+A seguir estão alguns exemplos de resultados
+destas incidências culturais,
+especificamente em áudio
+ e música.
 
 \subsubsection{Plugins LADSPA e lv2}
-      [repos AE, LM, wiki EL, historico CDTL]
 
 LADSPA (Linux Audio Developers Simple Plugin API) é a API
-livre\footnote{Veja o Capítulo 1 para a definição de livre neste
-  contexto.}  estabelecida de plugins de áudio até a presente data. A
-última versão liberada desta API é a 1.1, em uso corrente até hoje e é
+livre de plugins de áudio até a presente data. A
+última versão é a 1.1, em uso corrente até hoje e é
 de 2002 segundo os arquivos de cabeçalho dos códigos relacionados.
+O LADSPA \emph{version} 2, abreviado lv2, é uma segunda versão do
+protocolo também estabelecida e estável, mas extensível, motivo pelo
+qual está sempre em desenvolvimento.
 
-A exploração de APIs específicas foge ao escopo do presente texto e os
-códigos completos podem ser encontrados pela rede, junto a tutoriais e
-plugins de exemplo. O código destes desenvolvimentos e pertinente ao
-áudio são relativos à síntese e remoção de ruídos. Da síntese, existem
-abordagens mais puristas, que no caso sintetiza ruído branco e filtra
-para resultar nos espectros desejados (veja logo abaixo para maiores
-especificações). Uma abordagem menos purista porém mais eficiente é a
-de usar uma amostra curta do ruído e reproduzí-la indefinidamente
-através de \emph{cross-fades}.
+Na síntese, para maior qualidade, 
+pode-se fazer a recomposição a partir do espectro desejado, como na seção~\ref{subsec:ruidos}.
+Menos purista porém mais eficiente é
+usar uma amostra curta do ruído e reproduzi-la indefinidamente
+através de \emph{cross-fades}. Esta eficiência é desejada em um plugin,
+pois seus usos comuns estão aplicações em tempo real.
+
+Através deste trabalho, foi disponibilizado um pacote de plugins lv2
+que sintetizam os ruídos branco, azul, violeta, rosa e marrom. Os
+  códigos em C e a implementação como plugin lv2 estão no
+  repositório git como pode ser visto em \emph{online}\footnote{
+      este link é o README do repositódio git com todos os códigos dos
+      plugins:
+  http://labmacambira.git.sourceforge.net/git/gitweb-index.cgi.}.
 
@@
 extremamente dependentes da aplicação. A complexidade dos algorítmos
-pode atingir níveis de especialidade em processamento de sinais que
-merecem um trabalho dedicado. Aqui iremos expor somente sobre remoção
-de ruído \emph{'Hum'} causado usualmente pela corrente alternada que
+atinge níveis de especialidade em processamento de sinais que já
+mereceram trabalhos dedicados. Adiante está a implementação de um
+removedor de ruído \emph{'Hum'}. Este ruído é
+geralemte causado pela corrente alternada que
 alimenta os equipamentos utilizados.
 
-Abaixo dispomos explicações breves sobre os diferentes ruídos e os
-códigos relacionados. Preferimos expor em Python os códigos de síntese
-por questão de clareza e coerência com o resto do texto\footnote{Os
-  códigos em C/C++ e a implementação como plugin lv2 está no
-  repositório git LV2 como pode ser visto em:
-  http://labmacambira.git.sourceforge.net/git/gitweb-index.cgi}.
-
-O procedimento de reprodução do ruído através de uma amostra repetida
-e em \textit{cross-fade} é a seguinte:
-
-[código da reprodução do ruído em loop e croos-fade ~10 linhas]
-
-Sintetizando diferentes ruídos\footnote{Dado o propósito do texto e da
-  implementação, utilizamos nomenclaturas utilizadas na música para os
-  ruídos, como se pode notar.}:
-
-\begin{itemize}
-    \item Ruído Branco: 
-    \item Ruído Rosa:
-    \item Ruído Marrom:
-    \item Ruído Violeta:
-    \item Ruído Azul:
-    \item Ruído Violeta:
-    \item Ruído Cinza:
-    (nao oficiais)
-    \item Ruído Vermelho:
-    \item Ruído Laranja:
-    \item Ruído Preto:
-    \item Ruído Branco Ruidoso:
-    \item Ruído Preto Ruidoso:
-    \item Clicks:
-\end{itemize}
-
-A remoção de ruído Hum é baseada em uma sequência de filtros notch
-(ver fundamentos) idealmente dispostos nos harmônicos da frequência de
+A remoção de ruído \emph{Hum} é baseada em uma sequência de filtros nó rejeita
+banda (veja subseção~\ref{subsec:filtros}
+idealmente dispostos nos harmônicos da frequência de
 oscilação da corrente elétrica. Cada um destes filtros deve permitir
@@
 pois trata-se de um sistema real passível dos mais diversos efeitos de
-distorção do modelo ideal. Abaixo expomos o código em Python para tal
-\emph{removedor de ruído} como é chamado. O código C/C++ e a
+distorção do modelo ideal. O código C/C++ e a
 implementação como plugin está no mesmo repositório disponibilizado
@@
 
-[código e explicação do removedor de ruído Hum]
+\subsubsection{\emph{Wavelets}}
+Alguns dos experimentos em \emph{wavelets} praticados estão em
+Audioexperiments.\cite{Audioexperiments} Uma destas investidas
+superou em muito um experimento e constitui um protocolo de compactação de áudio desenvolvido com o Prof.
+Rafael Santos Mendes na FEEC/UNICAMP como consequência do mestrado realizado por André Luvizzoto sob orientação do mesmo professor.
+O método está bem descrito no artigo desenvolvido.
+Em resumo, o método consiste particionar o áudio com sobreposições para contemplar aplicações reais sem as distorções de borda da decomposição e recomposição.
+Em seguida, o áudio é decomposto em árvore \emph{Wavelet Packet}.
+Cada folha é então ordenada, os coeficientes com baixa energia eliminadose um polinômio encontrado pelo método dos mínimos quadrados é usado para representar os coeficientes restantes.
+Detalhes de escrita e leitura do polinômio completam o protocolo.
+Aplicações para stream de voz em tempo real foram vislumbradas e testes adicionais propostos.\cite{FabbriWavelets}
+
+\subsubsection{Redes Complexas}
+Com o fim de detectar nuances emocionais no uso da linguagem natural, foram desenvidos alguns trabalhos em análise de fala e de texto, ambos focados no uso de redes complexas.
+Três trabalhos: dois de processamento de texto e um de fala.
+O primeiro trabalho de processamento de texto foi apresentado na ACL de 2010, em Upsala, na Suécia, pelo Prof.
+Thiago Pardo, atual presidente da ACL.\cite{FabbriACL,FabbriComplenetVoz,FabbriComplenetTexto}
+
+Os trabalhos focaram em distinguir polaridade em excertos de textos e de fala.
+Nos textos, a rede é formada a partir da sequência de palavras, em que cada palavra é um vértice e a sucessão imediata cria uma aresta.
+O pré-processamento dos textos consiste na lematização e na retiradas das \emph{stop-words} e da pontuação.
+Nas falas, as bandas de frequência utilizadas são vértices e a transição entre uma banda e outra é uma aresta.
+Ambas redes com peso e podem ser utilizadas como direcionais ou não.
+Os trabalhos consistiam em formar estas redes com os bancos de dados correlatos, extrair medidas e aplicar reconhecimento de padrão.
+Ambos os estudos não fazem uso do conteúdo semântico, mas sim topológico dos sinais recebidos, e chegam a mais 80\% de cobertura ou precisão em alguns casos, podendo contribuir com métodos tradicionais que fazem uso de redes semânticas, por exemplo.
 
@@
 
-Por razões didáticas, dividimos este capítulo em 4 partes: música em
+Este capítulo está dividido em 4 partes: música em
 tempo diferido (ou seja, que não é feito em tempo real), música em
-tempo real, música na matéria (suporte físico em hardware) e música no
+tempo real, música na matéria (suporte físico em \emph{hardware}) e música no
 tecido social (considerando especialmente as mobilizações humanas
@@
 
-Com isso desejamos expor a prática musical através do código com
-exemplos reais de aplicação e em uso pelo autor, por membros do
-LabMacambira.sf.net, por parceiros, colaboradores e por usuários
+Com isso desejamos expor uma incidência da prática musical através do código com
+exemplos reais de aplicação e em utilização pelo autor, membros do
+LabMacambira.sf.net, parceiros, colaboradores e por usuários
 eventuais das naturezas mais diversas.
-
-%%%%%%%%%%%%%%5 PULANDO AQUI, QUEM SABE COLOCAR EM OUTRO TEX %%%%%%%
 
@@
 \begin{quotation}
-\small 'The increasing dominance of graphic interfaces for music
-software obscured the continuing presence of the command-line
-tradition, the code writer, the hacker. The code writing of deferred
-time computer programming may be assembled out of time order, debugged
-and optimized.'
+\small 'I also find that intelligent people always respect the intelligence needed to construct a simple structure in a clear way that really works.'
 
-\emph{Simon Emmerson, Living electronic music, 2007}
+\emph{Tom Johnson}
 \end{quotation}
@@
 A realização musical em tempo diferido é o paradigma inicial da música
-computacional. Iniciando com o Music V, a proposta foi depois
-desenvolvido com o CSound. Pode-se dizer que até hoje é a forma como
-compositores usualmente pensam a música: pensando e escrevendo as
-estruturas, que depois são executadas por instrumentistas ou aparelhos
+computacional, iniciando com o Music V, depois
+com o CSound. Pode-se dizer que até hoje é a forma como
+compositores usualmente pensam a música: concebendo e escrevendo as
+estruturas que depois são executadas por instrumentistas ou aparelhos
 eletrônicos.
 
-Assim como a composição instrumental permite um trabalho mais
-minucioso do que a improvisação instrumental, a realização musical em
-tempo diferido usualmente permite um detalhamento maior dos
+Assim como a composição instrumental permite alguns refinamentos estruturais não presentes na improvisação instrumental,
+a realização musical em
+tempo diferido permite um detalhamento maior dos
 procedimentos do que a realização em tempo real. Por este mesmo
-motivo, trataremos inicialmente de dois trabalhos em tempo diferido.
-Aliás, como veremos a seguir, as abordagens completam uma base para a
-música computacional praticada até os dias de hoje.
+motivo.
 
-O som musical pode ser caracterizado fisicamente, e sua síntese
-digital com base em nossa banda de audição e no teorema de Nyquist,
-sintetiza uma música amostra por amostra, através de princípios claros
-de síntese sonora e organização musical\footnote{Veja em seguida o
-  \emph{minimum-fi} hi-fi, low-fi, minimum-fi, ou seja, o mínimo de
-  qualidade para assegurar existência e consistência da criação}. O
-segundo, chamado de \emph{FIGGUS} (FInite Groups in Granular and Unit
-Synthesis), utiliza os princípios do minimum-fi e constitui um módulo
-Python completo, cuja proposta é a utilização de simetrias, através de
-permutações e de Teoria de Grupos, para a composição de músicas. Como
-demonstração das capacidades do FIGGUS, ele gera um
-EP\footnote{Extended Play, uma álbum musical maior que um single mas
-  menor que um LP (Long Play) inteiro} com um único comando. Este é o
+O som musical pode ser caracterizado fisicamente, tornando possível a  sua síntese
+digital. Estes sons e a organização deles são objeto do Capítulo 2 da presente dissertação.
+A utilização destes sons diretamente, utilizando somente os recursos básicos de justaposição e sobreposição, está na pequena peça \emph{minimum-fi}. Já o \figgus\ (FInite Groups in Granular and Unit
+Synthesis), utiliza os princípios do minimum-fi e utiliza simetrias, através de
+permutações e de Teoria de Grupos, para a composição de músicas.
+O \figgus\ gera um
+EP\footnote{\emph{Extended Play}, um álbum musical maior que um single mas
+  menor que um LP (Long Play) inteiro.} com um único comando. Este é o
 \emph{PPEPPS}\footnote{Pure Python EP: Projeto Solvente}, como veremos
@@
 Desta forma, esta sessão exemplifica e explicita - através de dois
-exemplos reais - os princípios do uso de código para a síntese
+exemplos reais - o uso de código para a síntese
 \emph{musical}, desde as amostras relativas a uma nota com dada
@@
 
-\subsubsection{Minimum-fi}
+\subsubsection{\emph{Minimum-fi}}
 
-Existe uma perene matiz estética e também tecnológica de realizar uma
-dada tarefa com \emph{o mínimo} necessário.  Esta matiz é para a
-música igualmente fundamental e reconhecida como princípio de unidade
+Existe uma perene matiz estética e também tecnológica dedicada a realizar uma
+dada tarefa com \emph{o mínimo} necessário.  Na música, esta matiz é
+igualmente presente e decorre em grande parte do princípio de unidade
 e coerência\footnote{Este princípio tanto é fundamental que as escolas
@@
   convenções mantidas por toda a sua duração, os arcos mantém
-  características, enfim, podemos até mesmo concluir que a simetria
-  define o escopo.}. Em código computacional a empreitada para
+  características, enfim, podemos até mesmo concluir que as simetrias
+  definem o escopo musical.}. 
+  Em código computacional, a empreitada para
 manifestar este princípio ele próprio de forma mínima resultou no
 \emph{minimum-fi.py}, código Python em um único arquivo curto que
-sintetiza músicas segundo as estruturas especificadas em
-linha\footnote{Em música se fala da \emph{bula} (que equivaleria às
-  funções em si e das especificações da linguagem) e a partitura ou
-  'música' propriamente ditas (que equivale à .}.  Na versão atual, de
+sintetiza a música segundo as estruturas especificadas em
+linha. Na versão atual, de
 2012 mas adiantada em 2011, os algorítmos em Python propriamente ditos
 somam 53 linhas e incluem apenas 5 funções. Com estas funções,
-estruturas musicais podem ser criadas padrão a padrão, nota a nota,
-amostra por amostra. Na prática e posto em linguagem cotidiana,
-resultam em notas que formam blocos e estruturas hierarquicamente
+estruturas musicais são criadas padrão a padrão, nota a nota,
+amostra por amostra. Na prática, e em linguagem cotidiana,
+as notas formam blocos e estruturas hierarquicamente
 superiores.
@@
   \emph{Compêndio de Teoria Elementar da Música} fala das propriedades
-  do som musical de forma equivalente.}, são:
+  do som musical e sua organização de forma condizente.}, são:
 \begin{itemize}
@@
     geração de unidades sonoras com diferentes timbres, controle sobre
-    a frequência fundamental e duração específica\footnote{Também
-      entra aqui o volume, mas como o que controlamos de fato é a
-      intensidade relativa entre as notas, resolvemos por bem omitir
-      esta parte até que fiquem claros os códigos relativos a isso,
-      logo abaixo.}.
-  \item Deve-se ser capaz de construir não só unidades sonoras, mas
-    também séries de unidades, sejam sobrepostas (acordes, por
-    exemplo) ou justapostas (melodias, por exemplo).
+    a frequência fundamental, duração e volume, como especificado na seção~\ref{sec:notasDisc}.
+  \item Deve-se ser capaz de construir
+    séries de unidades, sejam sobrepostas (e.g. acordes)
+    ou justapostas (e.g. melodias).
 \end{itemize}
 
-Para o primeiro item, se prestam comumente os procedimentos de busca
-em tabelas/vetores com formas de ondas em alta resolução (chamado
-\emph{lookup table}). O procedimento é barato computacionalmente e de
-qualidade alta (não acrescentam ruídos relevantes ao sinal).
-
-Para realizarmos a busca de forma eficiente, precisamos de tabelas que
-contenham cada forma de onda. As tradicionais são a senoide e
-as ondas quadrada, triangular e dente de serra.  Mostramos a seguir
-estas formas de onda e também uma forma de onda retirada de um som
-real.
-
-De posse destas formas de onda, A seguir está o procedimento da
-lookuptable:
-
-%\code{Procedimento de Lookup Table}{python_snippets/lookuptable.py}
-
-O primeiro ponto importante é a tabela em si, usualmente
-unidimensional. É de comum conhecimento que tabelas com 1024 são mais
-que suficientes para os usos musicais e a diferença de qualidade
-(relação sinal/ruido) não muda consideravelmente com a utilização de
-tabelas maiores. A seguir dispomos algumas destas tabelas, mais
-especificamente, dispomos as tabelas para os formatos de onda mais
-usuais.
-
-%\code{Formas de Onda Tradicionais}{python_snippets/formasDeOnda.py}
+Para o primeiro item, se prestam os procedimentos de busca
+em tabelas/vetores com formas de ondas em alta resolução, chamado
+\emph{lookup table}. O procedimento é barato computacionalmente,
+com resultados diversificados e tidos como de alta qualidade
+pois não acrescentam ruídos relevantes ao sinal. Uma descrição
+do procedimento está na subseção~\ref{subsec:lookup}.
 
 Através da utilização do lookup sucessivo (procura-se um valor na primeira tabela
-e este valor indica o valor na segunda tabela a ser utilizado), executamos
-um waveshaping. Este procedimento é bastante apreciado pela simplicidade e eficácia
+e este valor indica o valor na segunda tabela a ser utilizado), executa-se
+um \emph{waveshaping}. Este procedimento é bastante apreciado pela simplicidade e eficácia
 na síntese de tímbres diversos e ricos em harmônicos e evolução temporal. Embora
@@
 método se caracteriza pela aplicação de uma função não linear ao sinal de entrada.
-Neste caso o sinal de entrada é gerado pelo primeiro lookup, a função não linear aplicada
-é representada pela segunda tabela e aplicada pelo segundo lookup.
+Neste caso o sinal de entrada é gerado pela primeira busca, a função não linear aplicada
+é a segunda busca, na outra tabela.
 
@@
 
-%\code{Waveshaping com Lookups sucessivas}{python_snippets/lookup_cruz.py}
+\code{Waveshaping com consultas sucessivas a tabelas}{python_snippets/lookup_cruz.py}
 
-O segundo item - dos dois princípios expostos sobre o minimum-fi - presta-se à discretização do espaço musical. Unidades como batidas e notas
+O segundo item - dos dois princípios expostos sobre o minimum-fi - presta-se à discretização do espaço musical.
+Unidades como batidas e notas
 tornam mais eficiente a comunicação pois a quantidade
-de estruturas sugeridas é maior e as estruturas são mais claras no discreto do que no contínuo~\cite{Roederer}. Roederer chega a
+de estruturas sugeridas é maior e as estruturas são mais claras no discreto do que no contínuo.\cite{Roederer} Roederer chega a
 apontar que as próprias notas dos instrumentos musicais são um reflexo de que é mais eficiente
@@
 De fato, unidades bem definidas se mostram úteis na prática musical 
-para fazer sequências de unidades, concatená-las. Quando as unidades
-são notas, as sequências de unidades justapostas no tempo são melodias ou linhas melódicas. As
-sequências sobrepostas no tempo são comumente pensadas como acordes, mas podem ser tidas simplesmente
-como sobreposições circunstanciais de duas linhas melódicas. Isso, claro, segundo
-a sistematização clássica e usual da música~\cite{Lacerda}\footnote{Vole assinalar aqui que a música do século XX apresentou diversos modelos teóricos que quebram com este entendimento simplificado sobre a música, suas unidades básicas e estruturas relacionadas}.
+para fazer sequências de unidades, concatená-las no tempo. Quando as unidades
+são notas com frequência definida, as sequências de unidades justapostas no tempo tendem a ser compreendidas como melodias ou linhas melódicas. As
+sequências sobrepostas no tempo são comumente compreendidas como acordes, mas podem ser tidas simplesmente
+como sobreposições circunstanciais de duas ou mais linhas melódicas\footnote{A música do século XX apresentou diversos modelos teóricos que quebram com este entendimento simplificado sobre a música, suas unidades básicas e estruturas relacionadas}.\cite{Lacerda}
 
-As duas construções básicas explicitadas a seguir, baseadas na dicotomia melodia/harmonia
-(horizontalidade/verticalidade, justaposição/sobreposição), são
+As duas construções básicas explicitadas, baseadas na dicotomia melodia/harmonia - relacionada à horizontalidade/verticalidade, justaposição/sobreposição - são
 as funções \emph{fazSequencia} e \emph{fazAcorde} no minimum-fi. Vale notar elas são absolutamente 
 equivalentes em uma análise puramente conceitual, i.e.
-uma delas pode ser omitida segundo algumas teorizações.
+uma delas pode ser omitida sem perda das possibilidades musicais.
 Isso fica particularmente óbvio quando se nota que os procedimentos de mixagem e concatenação são
 plenamente capazes de realizar o que estas funções realizam. Aliás, as funções nada
-mais são do que usos típicos e quase caricatos destes procedimentos: no fazAcorde a mixagem
-sobrepõe no tempo todas as unidades, no fazSequencia as unidades são todas juntapostas no tempo.
+mais são do que usos típicos e quase caricatos destes procedimentos: no \emph{fazAcorde} a mixagem
+sobrepõe no tempo todas as unidades, no \emph{fazSequencia} as unidades são todas juntapostas no tempo.
 
@@
 
-%\code{Realização de Sequências de Notas}{python_snippets/fazSequencia.py}
+\code{Realização de Sequências de Notas}{python_snippets/fazSequencia.py}
 
-%\code{Realização de Acordes de Notas}{python_snippets/fazAcorde.py}
+\code{Realização de Acordes de Notas}{python_snippets/fazAcorde.py}
 
@@
 
-%\code{Somador (função auxiliar)}{python_snippets/somador.py}
+\code{Somador (função auxiliar)}{python_snippets/somador.py}
 
-Depois disso é usufruir com estruturas criadas. Por exemplo: depois de construidas
-algumas tabelas para serem usadas como diferentes timbres, pode-se criar as
-escalas completamente simétricas na oitava cromática assim:
+Com isso o básico está coberto e o foco vai para a criação de estruturas. Por exemplo: 
+pode-se criar as
+escalas completamente simétricas na oitava cromática, escalas diatônicas ou até microtonais, como as descritas na subseção~\ref{subsec:intervalos}.
 
-%\code{Escalas Completamente Simétricas (na grade dos 12 semitons e no âmbito da oitava)}{python_snippets/escalas_simetricas.py}
+O estabelecimento de pequenas sequências é bastante útil para
+reutilizações, variações e geração de materiais derivados, por exemplo: 
+\code{Sequências diversas}{python_snippets/series.py}
 
-% escala\_1=range(12) # cromática ascendente sem inclusão da oitava
-% escala\_2=range(0,12,2) # tons inteiros
-% escala\_3=range(0,12,3) # diminutão
-% escala\_4=range(0,12,4) # terças maiores
-% escala\_6=range(0,12,6) # trítonos
-% \end{python}
-% \input{py}
+O passo seguinte é sintetizar e mixar para a obtenção 
+de sequências musicais. A síntese de sequências e acordes
+podem ser feitos desta forma:
 
-Sendo cada unidade um semitom.
-Ou seja, cada unidade é o fator que fica presente na conta $f_0 .
-2^{(1/12)} .
-fator $ e resulta na frequência exata da nota relativa a $f_0$ em um sistema temperado ideal.
+\code{Sintetizando sequências e acordes}{python_snippets/seqs_acordes.py}
 
-Podemos também utilizar as usuais
-escalas tonais maior e menor natural, harmônica e melódica:
-
-%\code{Escalas diatônicas tonais (maiores e menores)}{python_snippets/escalas_Mm.py}
-
-A utilização de intervalos menores que o semitom\footnote{Prática também conhecida
-como \emph{microtonalismo}, que gera/utiliza \emph{microtonalidade}.} mostra-se trivial em
-nossa implementação. Basta representar sequências de fracionários
-que são partes linearmente proporcionais dos semitons ou modificar o fator $f$:
-
-%\code{Escalas Microtonais (com quartos e oitavos de tom e intervalos menores)}{python_snippets/escalas_microtonais.py}
-
-Da mesma forma, as triades maiores e menores
-são especificadas com simplicidade. Note que aos acordes \emph{diminutão} e
-\emph{aumentado} correspodem as mesmas notas das escalas simétricas de terças menores
-e maiores respectivamente:
-
-%\code{Acordes anotados como listas em python}{python_snippets/acordes.py}
-
-E séries/sequências podem ser anotadas junto a variações:
-
-%\code{Séries diversas}{python_snippets/series.py}
-
-Com este arcabouço, o passo seguinte é sintetizar e mixar, resultando
-em sequências musicais. A síntese de sequências e acordes são feitos tipicamente
-através de comandos desta forma:
-
-%\code{Sintetizando sequências e acordes}{python_snippets/seqs_acordes.py}
-
-Já a síntese de estruturas compostas (p.ex. sequências de acordes e sobreposição
-de linhas melódicas), são feitas com os recursos usuais da liguagem. Neste caso
-usamos listas em Python e implementações em Numpy (mais eficientes em tempo de
-execussão e também na simplicidade do código) estão no apêndice. Estes mesmos
+Já a síntese de estruturas compostas (e.g. sequências de acordes e sobreposição
+de linhas melódicas), é feita com os recursos usuais da liguagem:
+listas em Python e vetores Numpy (mais eficientes em tempo de
+execussão e também na simplicidade do código). Estes mesmos
 procedimentos são praticamente os mesmos em Scilab, C/C++, Javascript, PHP, etc.
 
-A seguir demonstramos para fins didáticos, a construção de acordes periódicos em
-python puro\footnote{A implementação não didática destas três linhas de código
-podem ser agrupadas em uma só linha.}:
+A seguir, para fins didáticos, está a construção de acordes periódicos em
+python puro\footnote{Uma implementação não didática destas três linhas de código
+pode ser feita em uma só linha.}:
 
-%\code{Acordes periódicos}{python_snippets/acordes_periodicos.py}
+\code{Acordes periódicos}{python_snippets/acordes_periodicos.py}
 
 Em posse destas sequências, acordes e recursos da linguagem,
-formamos estruturas hierarquicamente superiores
+são formadas estruturas hierarquicamente superiores
 através da concatenação de estruturas, da mixagem de estruturas, e da amplificação
@@
 
-%\code{Amplificação e mixagem}{python_snippets/amp_mix.py}
+\code{Amplificação e mixagem}{python_snippets/amp_mix.py}
 
-Neste ponto, basta criarmos músicas e sequências de interesse estético ou para pesquisa.
-Dado o ferramental, os encadeamentos dependem de intensões estéticas, entendimentos musicais e estruturas abstratas que mantém a coerência e interesse em uma peça musical.
-Para o leitor mais interessado, deixamos para o Apêndice X um exemplo de música feita com o minimum-fi.
+Neste ponto, basta criar músicas e sequências de interesse estético ou para pesquisa.
+Os encadeamentos dependem de intensões estéticas, entendimentos musicais e estruturas abstratas que mantém a coerência e o interesse em uma peça musical.
+Para o leitor mais interessado, recomendamos uma visita ao capítulo~\ref{cap:resultados} e à \emph{toolbox} \massa, onde pode-se encontrar o próprio script \emph{minimum-fi.py} e outros experimentos.
 A seguir utilizamos esta base apresentada para sintetizar estruturas musicais e então um EP.
@@
 
-Em coerência com os conhecimentos
-e tecnologias já apresentados nesta exposição\footnote{Especificamos o arcabouço para geração de sons musicais em unidades
-e estruturas compostas logo acima na subsessão sobre o minimum-fi.}, introduzimos agora
-um foco especial nas estruturas musicais
-desenvolvidas. São elas, inseridas em um momento histórico e
+Nesta subseção há um foco especial nas estruturas musicais.
+São elas, inseridas em um momento histórico e
 executadas por instrumentos específicos com as técnicas de época,
-que constituem uma linguagem musical e músicas propriamente
-ditas[ref]. O FIGGUS constitui uma técnica composicional
-manifestada em software como ferramenta de síntese de
-estruturas musicais por estruturas matemáticas específicas
-para representação de simetrias.
+que constituem em grande parte uma linguagem musical e músicas propriamente
+ditas. O \figgus\ constitui uma técnica composicional
+manifesta em software como ferramenta de síntese de
+estruturas musicais. Esta técnica
+consiste na utilização de estruturas matemáticas
+para a representação de simetrias.
 
-Este desenvolvimento foi iniciado em 2006 com o físico-matemático Prof. Adolfo Maia Junior - bem anterior
-ao nascimento do minimum-fi - para
+O \figgus\ foi iniciado em 2006 com o físico-matemático Prof. Adolfo Maia Junior (do IMECC e NICS, ambos da UNICAMP) - bem anterior
+ao nascimento do \emph{minimum-fi} - para
 tratar de simetrias na música com vistas à composição musical através
-de métodos matemáticos\footnote{Duas iniciações científicas trataram do assunto, convenios XXX e YYYY.}. Mais especificamente, a proposta resultou em
+de métodos matemáticos\footnote{Duas iniciações científicas trataram do assunto.}. Mais especificamente, a proposta resultou em
 um programa voltado para a síntese
 granular e síntese de estruturas musicais através de Grupos Algébricos. O nome dado
-foi FIGGUS, sigla de FInite Groups in Granular and Unit Synthesis\footnote{Também utilizamos
+foi \figgus, sigla de FInite Groups in Granular and Unit Synthesis\footnote{Também foi usado
 o nome FIGGS (FInite Group in Granular Synthesis) dado que o termo \emph{unit synthesis} não
 é usual na literatura. Posteriormente o primeiro autor deste trabalho recorreu novamente
-ao uso do nome FIGGUS. Isso foi motivado principalmente pelo fato de que
-o maior uso da técnica é para síntese de estruturas musicais, não para
+ao uso do nome \figgus. Isso foi motivado pelo
+uso da técnica para síntese de estruturas musicais, não para
 síntese de amálgamas sonoros (ou timbres mesmo) tipicamente resultantes da síntese granular}.
 
-Na atual reescrita, embora ainda bastante atrás do FIGGUS original quanto
+Na atual reescrita, embora ainda bastante atrás do \figgus\ original quanto
 à interface gráfica, a ferramenta opera diretamente em Python puro,
@@
 
-%\code{Utilizando o FIGGUS para Sintetizar um EP}{python_snippets/synth_ep.py}
+\code{Utilizando o FIGGUS para Sintetizar um EP}{python_snippets/synth_ep.py}
 
-Desta forma, a ferramenta fica muito mais simples para experimentações
-e implementações adicionais pois sua utilização se dá através
-do próprio código no qual tudo é especificado. Esta abordagem se mostra particularmente
-pertinente pois a estagnação do mecanismo de síntese de estruturas musicais
-beira a inviabilização do uso artístico ou torna necessário o uso
-de recursos externos.
-Outro uso pretendido para o FIGGUS
+Desta forma, a ferramenta 
+é simples de ser usada para experimentações
+e implementações adicionais.
+Outro uso planejado para o \figgus\
 é a síntese de tímbres e amálgamas sonoros através da
-Síntese Granular. Embora nosso foco seja outro, cabe algumas breves
+Síntese Granular. Embora o foco atual seja outro, cabe algumas breves
 palavras sobre o assunto.
@@
 na Computação Musical e se caracteriza pela geração de sons bastante curtos
-e em quantidade massiva. Tipicamente os sons possuem entre 5 e 40 milissegundos
-e a quantidade destes \emph{microsons}\footnote{Grãos sonoros e microsons são jargões
-típicos da síntese granular usados para indicar sons com as durações especificadas
-no texto.} pode chegar a milhares por segundo. O tratamento específico da
-síntese granular foge ao escopo deste trabalho e indicamos ao leitor interessado
-os artigos produzidos sobre Síntese Granular e Teoria de Grupos[refs AES e SBCM].
-Assim, focamos o texto a seguir em Grupos Finitos e Síntese de Estruturas Musicais.
+e em quantidade massiva. Tipicamente, os sons possuem entre 5 e 40 milissegundos
+e a quantidade destes \emph{microsons}\footnote{Grãos sonoros e microsons são jargões equivalentes
+típicos da síntese granular. São usados para indicar sons com durações
+bastante curtas, como assinalado no texto.} pode chegar a milhares por segundo. O tratamento específico da
+síntese granular foge ao escopo deste trabalho. O leitor interessado pode consultar os artigos produzidos sobre Síntese Granular e Teoria de Grupos.\cite{figgusOriginal,figgusEspacializacao}
+Desta forma, o texto a seguir concentra-se em Grupos Finitos para a síntese de estruturas musicais.
 
@@
 que as simetrias possuem. Na música, para citar somente alguns exemplos simples,
-temos os numerosos estudos de simetrias na música de J. S. Bach, os jogos
+há os numerosos estudos de simetrias na música de J. S. Bach, os jogos
 de dados de Mozart e os usos recorrentes da proporção áurea na música
 de Béla Bártok. Matematicamente, as \emph{simetrias} são descritas por Grupos,
-e estes são definidos como sendo um conjunto (chamemos de $G$)
+e estes são definidos como um conjunto (seja $G$)
 munido de uma operação (seja $\bullet$), formando um grupo $(G,\bullet)$
-satisfazendo as seguintes propriedades:
+satisfazendo as propriedades descritas na subseção~\ref{estCic}.
 
-\begin{enumerate}
-    \item Fechamento: $\forall \ g_1, g_2  \in G \Rightarrow g_1 \bullet g_2 \in G$
-    \item Associatividade: $(g_1 \bullet g_2) \bullet g_3 = g_1 \bullet (g_2 \bullet g_3), \forall \ g_1, g_2, g_3 \in G$
-    \item Existência do elemento neutro: $\exists \ e \in G : g \bullet e = e \bullet g = g$
-    \item Existência dos inversos: $\forall \ g \in G \ \exists \ g^{-1} : g \bullet g^{-1} = g^{-1} \bullet g = e$
-\end{enumerate}
-
-Adicionalmente, vale complementar que se um grupo tiver um número
-finito de elementos, ele é dito finito, caso contrário é chamado
-infinito. Todos os elementos de um grupo finito são cíclicos pois
-operado com si próprio um número suficiente de vezes resulta em si
-mesmo. Um grupo é dito comutativo ou abeliano caso a propriedade
-comutativa seja satisfeita para todos os elementos:
-$\forall \ g_1, g_2 \in G \Rightarrow g_1 \bullet g_2 = g_2 \bullet g_1$.
-
-São numerosas as aplicações de grupos e estes
-usos para as artes possam se desenvolver
-de diversas formas focadas em simetrias
-e com interesse estético, em
-nossas pesquisas uma delas se mostrou particularmente coerente
-e pertinente: as permutações de unidades sonoras. Primeiro,
-através do Teorema de Cayley, todo grupo finito é isomorfo
-a um grupo de permutação\footnote{Mais especificamente,
-todo grupo é isomorfo a um subgrupo do grupo simétrico agindo
-em G.}, i.e. seus elementos são diretamente relacionados
-a permutações. Além disso, na música as permutações estão
-no núcleo de diversos procedimentos canônicos como a
-retrogradação e outras técnicas de variação.
-
-[Expor sobre uma ou duas músicas feitas com o FIGGUS]
-
-No FIGGUS, implementamos o grão sonoro (no caso mais usado
-como unidade) como uma classe que possui apenas
-os seguintes atributos (e não possui funcões): duração (segundos),
-frequência (Hz), timbre (identificador para usar mediante implementações convenientes), intensidade (pico $\in \ [0,1]$), e duração dos fades (in e out em segundos). Aqui a classe:
+No \figgus, o grão ou unidade sonora
+é uma classe que possui apenas
+os seguintes atributos: duração (segundos),
+frequência (Hz), timbre (identificador para usar mediante implementações convenientes), intensidade (pico $\in \ [0,1]$), e duração dos fades (in e out em segundos).
 
@@
 
-O FIGGUS funciona com base em uma sequência de grãos posta de antemão
-na qual operam as permutações. Além da sequência de grãos,
-é conveniente também manter o número de grãos da sequência e uma
-lista com os índices dos grãos para executar as operações necessárias para gerarmos as músicas (ver abaixo). A classe utilizada para representar as sequências é assim:
+O \figgus\ funciona com base em uma sequência de grãos especificada inicialmente e na qual operam as permutações.
 
@@
 
-Aos grãos em sequência neste ponto são aplicadas permutações.
-Para isso é bastante conveniente representar as permutações em classesseparadas e os padrões de aplicação destas permutações para facilitar
-desenvolvimentos com clareza.
-No caso a classe de permutações só possui os parâmetros da permutação em si e do tamando dela (número de elementos que ela utiliza).
-É conveniente implementar uma tradução da notação cíclica para a notação direta da permutação e é uma das facilidades da implementação por classe que fizemos.
-A classe do padrão de permutação possui também um período de aplicação da permutação, ou seja, de quantas em quantas leituras da sequência aplicamos a permutação.
-As classes ficaram assim:
+Aos grãos em sequência são aplicadas permutações.
+Para isso é bastante conveniente representar as permutações em classes próprias.
+A classe do padrão de permutação possui também um período de aplicação da permutação, ou seja, de quantas em quantas leituras da sequência apermutação é aplicada.
 
@@
 
-De posse dos grãos, da sequência, das permutações e do padrão de permutações estamos possibilitados de realizar a estrutura musical em si.
-Bastam adicionalmente, especificar o número desejado de iterações da sequência.
-Com isso, a sequência de grãos é lida um número de vezes, aplicando as permutações na sequência segundo o padrão de aplicação das pemutações, de forma a resultar em uma sequência sonora musical.
+De posse dos grãos, da sequência, das permutações e do padrão de permutações, pode-se realizar a estrutura musical em si.
+Basta adicionalmente especificar o número desejado de iterações da sequência.
+Com isso, a sequência de grãos é lida um número de vezes, aplicando as permutações na sequência de grãos segundo o padrão especificado, de forma a resultar em uma sequência musical.
 Note que se a permutação usar menos elementos que a sequência possui, alguns destes elementos ficarão estáticos nas iterações da sequência no padrão sonoro.
-
-Na síntese do vetor sonoro relacionado, é também crucial estabelecer a taxa de amostragem preterida, como podemos perceber na função de síntese dos vetores, da classe Pattern que cuida desta realização do padrão em si.
-Deixamos a seguir esta função, de forma que faltará apenas a escrita destes vetores sonoros em arquivo de áudio, como veremos a seguir.
 
@@
 
-Em posse não apenas de representações abstratas do padrão musical a ser realizado, mas dos próprios vetores sonoros relacionados à representação digital da música a ser realizada, podemos escrever este vetor como um arquivo de áudio propriamente dito.
-O mais conveniente neste caso é escrever um arquivo PCM (Pulse Code Modulation) em padrão amplamente utilizado e reconhecido.
-Ambos os padrões WAV e AIFF satisfazem estes requisitos e escolhemos o WAV por ser o padrão de CDs (Compact Discs) e também o mais utilizado por programas de áudio.
-Mais especificamente, o padrão de CDs é WAV com 44100 amostras por segundo e 16 bits por amostra\footnote{No formato WAV de CD, cada amostra é representada por uma variável inteira 'signed' de 16 bits.}.
-As amostras dos vetores sonoros são calculados, por conveniência e por convenção, no âmbito $[-1,1]$ e precisam ser normalizados para o âmbito $[-32767,32768]$ e truncados em números inteiros.
-Depois disso precisam ser escritos em um arquivo com os bits como na convenção da linguagem C/C++.
+Em posse não apenas de representações abstratas do padrão musical a ser realizado, mas dos próprios vetores sonoros relacionados à representação digital da música a ser realizada, pode-se escrever um arquivo de áudio propriamente dito.
+O mais conveniente neste caso é escrever um arquivo PCM (Pulse Code Modulation) em algum padrão amplamente utilizado e reconhecido.
+Ambos WAV e AIFF satisfazem estes requisitos.
+Mais especificamente, o padrão de CDs é WAV com 44100 amostras por segundo e 16 bits por amostra.
+As amostras dos vetores sonoros são calculados, por conveniência e convenção, no âmbito $[-1,1]$ e precisam ser normalizados para o âmbito $[-32767,32768]$ e truncados em números inteiros.
+Depois disso devem ser escritos em um arquivo com os \emph{bits} como na convenção da linguagem C/C++.
 A bibioteca \emph{struct} cuida dessa escrita do inteiro no formato correto, e a biblioteca \emph{wave} escreve o cabeçalho no formato WAV adequado.
@@
 
-%\code{Escrita do Vetor Sonoro em Arquivo WAV}{python_snippets/fio.py}
+\code{Escrita do Vetor Sonoro em Arquivo WAV}{python_snippets/fio.py}
 

### 2013-01-12 python_snippets/fazAcorde.py

@@ -1,4 +1,5 @@
-def fazAcorde(notas, f0, fator, dur_nota, dur_silencio, funcao1, funcao2=0):
+def fazAcorde(notas, f0, fator, dur_nota, dur_silencio,\
+                                     funcao1, funcao2=0):
     seq=[]
     for nota in notas:
         freq=f0*fator**nota

### 2013-01-12 python_snippets/fazSequencia.py

@@ -1,5 +1,5 @@
 # Note as variáveis que são fixadas de antemão
-def fazSequencia(notas, f0, fator, dur_nota, dur_silencio,
+def fazSequencia(notas, f0, fator, dur_nota, dur_silencio,\
                                         funcao1, funcao2=0):
     seq=[] # a sequência é iniciada vazia
     for nota in notas: # para cada nota

### 2013-01-12 python_snippets/seqs_acordes.py

@@ -6,16 +6,14 @@ dur_nota=0.5
 dur_silencio=0.1
 onda1=senoide
 onda2=dente_de_serra
-
-# fazendo a sequência com senóides:
+# realização da sequência com senóides:
 sequencia = fazSequencia( notas,
                           fundamental, 
                           fator, 
                           dur_nota, 
                           dur_silencio, 
                           onda1 ) 
-
-# fazendo o acorde com waveshaping
+# realização do acorde com waveshaping
 acorde = fazAcorde( notas
                     fundamental,
                     fator,

### 2013-01-12 musicaExtra.tex
terminada musica em tempo real

@@
 
-%%%%%%%%%%%%%%%%%% FIM DO PULO %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \subsection{Música em Tempo Real: Live coding e ABeatTracker (ABT)}
 
-Com os avanços computacionais recentes, tornou-se usual a síntese
-sonora em tempo real. Com isso, surgiram linguagens chamadas de
+Tornou-se usual a síntese
+sonora em tempo real mesmo em \emph{laptops} populares.
+Assim, surgiram linguagens de
 domínio específico, para o áudio e a música, em sua maioria dedicadas
--- ou ao menos capacitadas -- para o uso em tempo de
-execução. Puredata, SuperCollider, ChucK e ixilang são exemplos de
+- ou ao menos capacitadas - para interação em tempo de
+execução como \emph{Puredata}, \emph{SuperCollider}, \emph{ChucK}
+ e \emph{ixilang}. Todos estes são exemplos de
 linguagens largamente utilizadas para a composição musical e síntese
@@
 possibilitam que o usuário ouça o resultante sonoro do código
-utilizado e altere o código com resultado imediato no processamento e
-resultado que escuta.
+utilizado e altere o código com resultados imediatos no processamento e
+sonoridades produzidas.
 
-Nesta linha de exploração musical do código, iremos expor a seguir
-sobre nossas investidas em \emph{Live coding} (escrita de código em
-tempo real com vistas à performance pública) através de linguagens em
-contínuo desenvolvimento como ChucK, e o ABeatTracker (uma linguagem
-por macros para execução sonora rítmica em conjunto com instrumentos
-tradicionais e outras fontes sonoras/musicais externas).
+A exploração estética destas ferramentas 
+em performances musicais é a proposta do \emph{Live coding}.
+Esta prática de \emph{performance} se dedica
+especialmente à escrita de código em
+tempo real com vistas à projeção visual dos códigos enquanto
+são escritos, junto à projeção sonora que produzem.
+O ABeatTracker ainda pode ser considerado
+livecoding, mas já é um intermediário, com traços
+de um aplicativo ou ferramenta, embora os comandos sejam
+pela escrita. É uma linguagem
+com macros para execução de rítmos através de \emph{samples}
+em conjunto com instrumentos
+tradicionais e outras fontes sonoras/musicais externas.
 
@@
 
-Recentemente, grupos de ponta em música experimental no mundo todo
+Recentemente, grupos de ponta em música experimental
 estão desenvolvendo apresentações musicais públicas baseadas na
-escrita de código ao vivo. Destes vale ressaltar os grupos pioneiros
-slub e Benoît and the Mandelbrots e as ``orquestras de laptops''
-PLOrk, SLOrk e DubLOrk\footnote{Para maiores detalhes vide
-  \url{http://toplap.org}}. Usualmente, se projeta o código para que a
+escrita de código ao vivo. Este é um fenômeno cultural e
+estético, do qual vale ressaltar os grupos pioneiros
+\emph{Slub} e \emph{Benoît and the Mandelbrots} assim como
+as 'orquestras de \emph{laptops}'
+\emph{PLOrk}, \emph{SLOrk} e \emph{DubLOrk}\footnote{Para maiores detalhes, veja:
+  \url{http://toplap.org}.}. Usualmente, o código é projetado para que a
 audiência possa ver o que está sendo escrito, no rítmo em que se
@@
 
-As motivações para isso são variadas. Expomos a seguir de forma
-topificada um condensado do que nos orientou a tal prática. Vale
-ressaltar que são motivações igualmente presentes em outros
-grupos\footnote{Vide ``Manifesto Live coding'' em
-  \url{http://toplap.org/wiki/ManifestoDraft}}, embora não
-necessariamente em todos ou da mesmícima forma.
+Estas são motivações presentes em diferentes
+grupos\footnote{Vide 'Manifesto Live coding' em
+  \url{http://toplap.org/wiki/ManifestoDraft}.}, embora não
+necessariamente ou da mesmícima forma:
 
 \begin{itemize}
-    \item A performance musical por computador carece de recursos
-      performáticos à altura das execuções com instrumentos
-      tradicionais. Os gestos são por demais discretos e a
-      concentração do performer é bastante focada na tela do
+    \item A apresentação musical, com o uso do computador como instrumento,
+    carece de recursos
+      performáticos visuais dos instrumentos tradicionais.
+       Os gestos são por demais discretos e a
+      concentração do \emph{performer}/instrumentista é bastante focada na tela do
       computador.
-    \item O feedback auditivo do código projetado permite que o
-      espectador infira significados dos códigos. Mesmo que de forma
-      superficial, este recurso do Live coding é em muito capaz de
+    \item O \emph{feedback} auditivo do código projetado permite que o
+      espectador infira significados dos códigos. 
+      Este recurso do \emph{livecoding} é usado para
       desmistificar a programação de computadores e sua aplicação em
-      computação musical, por muito considerada completamente
+      computação musical, comumente considerada
       intangível.
-    \item O código em si é um recurso poderosíssimo, que permite ao
+    \item O código em si é um recurso poderosíssimo que permite ao
       usuário controlar os sons produzidos amostra por amostra ou em
-      escalas maiores, como notas, compassos, fraseados inteiros ou
+      escalas maiores de tempo, como notas, compassos, fraseados inteiros ou
       mesmo em escalas maiores de tempo, como minutos, horas, dias e
@@
     \item O compartilhamento do código é usual, leve e eficiente como
-      entrega completamente aberta da tecnologia relacionada à
-      proposta estética. Isso motiva não só o programador à aplicar
+      entrega de tecnologia valiosa para a
+      proposta estética. Isso motiva não só o programador a aplicar
       seus conhecimentos na música, mas também o músico a adentrar o
@@
 
-Desta forma, iniciamos em 2011 uma linha de atuação em Live coding com
-a criação do duo \emph{FooBarBaz} (Variáveis Metasintáticas) composto
-pelo autor Renato Fabbri e Vilson Vieira. Essa atuação resultou em uma
-performance no \emph{V Festival Contato}. É interessante ressaltar que
-essa foi a primeira apresentação de Live coding no Brasil, e a maior
-apresentação internacional em número de ouvintes (por volta de 5 mil
-pessoas) que se tenha relato.
+Desta forma, foi iniciada em 2011 uma linha de atuação em \emph{livecoding} com
+a criação do duo \emph{FooBar} composto por Vilson Vieira e o autor
+deste escrito. Este duo se desenvolveu no
+trio \emph{FooBarBaz}/Variáveis Metasintáticas composto
+pelo duo e por Gilson Beck na mesa de som, usando detecção de cor no laptop
+para incrementar a apresentação. Este trio
+realizou uma \emph{performance} no \emph{V Festival Contato}.
+ É interessante ressaltar que, até onde se sabe,
+essa foi a primeira apresentação de \emph{livecoding} no Brasil, e a maior
+apresentação internacional em tamanho de platéia, estima-se que entre 3 e 5 mil
+pessoas estavam presentes.
 
-Utilizamos a linguagem ChucK por apresentar os recursos que
-consideramos mais apropriados, embora de forma alguma isso seja
-consensual na prática atual de Live coding. Ainda, a apresentação
+Foi usada a linguagem emph{ChucK} por apresentar os recursos que
+o duo inicial considerou mais apropriados,
+ embora de forma alguma isso seja
+consensual na prática atual de \emph{livecoding}. Além disso, a apresentação
 contou com recursos adicionais para agregar interesse, como a
-utilização do \emph{cowsay} (para enviar mensagens enquanto se
-desenrolava a música) e trajetórias de um ponto no fundo do código
-para sugerir a vertigem do sono REM. Estes recursos podem ser vistos
-em uso claramente nos videos demonstrativos\footnote{Disponíveis em
+utilização do \emph{cowsay} para enviar mensagens enquanto se
+desenrolava a música. Em especial, as trajetórias de um ponto branco
+no fundo do código sugeria o sono REM como uma experiência coletiva de alteração
+do estado.
+Estes recursos podem ser vistos
+em uso nos videos demonstrativos\footnote{Disponíveis em
   \url{http://vimeo.com/33012735}, \url{http://vimeo.com/33018740},
   \url{http://vimeo.com/33019291}, \url{http://vimeo.com/33025717} e
-  \url{http://vimeo.com/33025913}.}  As linhas do \emph{cowsay} e o
+  \url{http://vimeo.com/33025913}.}
+  As linhas do \emph{cowsay} e o
 \emph{script} em linguagem Processing relacionados, assim como maiores
 detalhes sobre o duo e fotos da apresentação podem ser vistos em
-\url{http://wiki.nosdigitais.teia.org.br/FooBarBaz}.
+\url{http://wiki.nosdigitais.teia.org.br/FooBarBaz}. Estavam presentes
+no palco outros membros do labMacambira.sourceforge.net: Ricardo Fabbri,
+Alexandre 'Bzum', Larissa.
 
-Duas pessoas executaram Live coding simultaneamente. A saber, Vilson
-Vieira executava ritmos, batidas bastante marcadas que serviam como
-base. Renato Fabbri, autor do presente trabalho, executava linhas
-fluídas quasi melódicas que formavam arcos maiores. Nos intervalos das
-investidas no formato citado, havia interlúdios em que ambos se
+Especificamente sobre a prática do \emph{livecoding},
+duas pessoas executaram \emph{scripts} em \emph{Chuck} simultaneamente.
+A saber, Vilson
+Vieira executou ritmos, batidas bastante marcadas que serviam como
+base. Renato Fabbri, autor deste texto, executava linhas
+fluídas quasi-melódicas que formavam arcos maiores.
+Havia também interlúdios em que ambos se
 revesavam com músicas curtas e inusitadas, como em um duelo. A mixagem
@@
 
-A prática de Live coding requer o uso de linguagens que permitam a
-execução rápida de instruções curtas, incentivando assim o improviso
-musical através do código. Com esse objetivo, ambos \emph{live coders}
-usaram editores de texto -- vim e Emacs -- que permitiam a comunicação
-imediata com a linguagem de síntese sonora ChucK.  Acreditando que a
-música, especialmente sua vertente experimental, se baseia fortemente
-na organização de trechos de arquivos de áudio ou notas dadas a
-instrumentos sintetizados, foram gerados códigos que permitissem tal
-manipulação. Renato utilizou de trechos (``esqueletos'') minimalistas
-de código em ChucK para reproduzir arcos longos e estruturas
-melódicas, como os que seguem:
+A prática de \emph{livecoding} requer o uso de 
+instruções curtas, incentivando o improviso
+musical através do código. Ambos \emph{livecoders}
+usaram bons editores de texto
+para a escrita dos \emph{scripts} em \emph{ChucK}: Vim e Emacs.
+Renato utilizou aspectos estruturais de alto nivel, 
+arcos longos e estruturas
+melódicas e minimalistas, como os que seguem:
 
@@
 
-Ao mesmo tempo, Vilson utilizou dos códigos seguintes, também em
-ChucK, que permitem facilmente especificar a posição do arquivo de
-áudio que deve ser executada, sua velocidade de execução, duração e
-amplitude. Uma interface minimalista foi utilizada para permitir a
+Ao mesmo tempo, Vilson fez uso de scripts para
+facilmente especificar a posição do arquivo de
+áudio que deve ser iniciada a execução, assim como a velocidade
+reprodução do áudio, duração e
+amplitude. Uma interface simples foi permitiu a
 alteração rápida desses parâmetros em tempo de execução.
@@
 Esses desenvolvimentos todos inspiraram a criação de uma linguagem
-específica para Live coding, chamada Vivace, que está em constante
-desenvolvimento e tem seu código e maiores detalhes disponíveis em
-\url{http://automata.github.com/vivace}. Já foram realizadas novas
+específica para \emph{livecoding}, chamada \emph{Vivace}, que está em constante
+desenvolvimento e tem seu código e maiores detalhes disponíveis online
+de forma livre\footnote{Veja: 
+\url{http://automata.github.com/vivace}.}. Já foram realizadas novas
 apresentações utilizando essa linguagem, a notar: Hacklab do Velho, em
@@
 predominância do apelo às mídias populares, em uma reinauguração do
-gênero da \emph{pop art}: trechos de vídeos de uma novela brasileira
-eram rearanjadas em tempo real, contrastando com linhas rítmicas e
-melódicas executadas em sincronia com o material visual\footnote{Uma
-  amostra está disponível on-line para ser utilizada por qualquer
-  pessoa que possua um navegador Chrome:
-  \url{http://pulapirata.com/skills/vivace}}.
+gênero da \emph{pop art}: trechos de vídeos de uma novela brasileira: 'Avenida Brasil'
+eram rearanjados em tempo real em sincronia com linhas rítmicas e
+melódicas\footnote{Uma
+  amostra está disponível \emph{online} para ser utilizada por qualquer
+  em qualquer computador com o navegador \emph{Google-Chrome}:
+  \url{http://pulapirata.com/skills/vivace}.}.
+
+  Catalizadas pela aproximação de "Gera" Magela e Caleb Mascarenhas Luporini, as
+   apresentações e amadurecimeentos de 2012 desembocaram na
+  na escrita coletiva de códigos (acesso e controle da platéia 
+  no código que está sendo escrito), utilização de monstros e ascii art e
+  e sonoridades bizarras. Com isso, está sendo proposto
+  um gênero de \emph{livecoding} batizado de '\emph{Freakcoding}', possivelmente o primeiro subgênero
+  de \emph{livecoding}.
 
@@
 
-O ABT é uma linguagem/ferramenta que dispara linhas rítmicas através
-de macros que especificam as células rítmicas, amostras sonoras que
+O ABT é uma linguagem que dispara linhas rítmicas através
+de macros. Nestas, especifica-se as células rítmicas, amostras sonoras que
 são utilizadas como conteúdo sonoro destas linhas, modos de leitura
-destas amostras sonoras, e variáveis randômicas utilizadas para
+destas amostras sonoras e variáveis randômicas utilizadas para
 execussão da linha. Além disso, o ABT dispõe de variáveis globais que
-podem ser alteradas a qualquer momento pelo usuário, como BPM,
+podem ser alteradas a qualquer momento pelo usuário também com macros, como BPM,
 velocidade de leitura das amostras e variáveis randômicas globais (que
 se somam às individuais).
-
-Em pouco explicitaremos a forma de utilização do ABT. De antemão, é
-pertinente apresentar uma pequena discussão a respeito do que o ABT é
-considerado e sobre os propósitos desta ferramenta. Em primeiro lugar,
-vale manter em mente que o ABT tem um funcionamento específico que
-pode transcender o que se conhece por Live coding. As macros
-pré-estabelecidas engendram um conjunto de recursos pré-estabelecidos
+As macros
+pré-estabelecidas constituem um conjunto de recursos pré-estabelecidos
 bem definido, o que contrasta com a ideia de uma 'linguagem de
@@
 
-Sobre os propósitos do ABT, em primeiro lugar ele se dispõe a ser um
-instrumento computacional essencialmente rítmico. Junto a esta
-proposta, vem a necessidade da utilização em conjunto com outros
-instrumentos, externos ao ABT, ao computador em que estiver executando
+O ABT é um instrumento computacional essencialmente rítmico. Junto a esta
+proposta, está a necessidade da utilização em conjunto com outros
+instrumentos, externos ao ABT, ao computador em que o ABT está sendo executado
 e possivelmente externo com relação a qualquer computador. Para isso
@@
 que o ABT sincronize o pulso e utilize células rítmicas
-relacionadas. A análise feita pelo ABD resulta em uma série de ritmos
+relacionadas. A análise feita pelo ABD resulta em uma série de rítmos
 explicitados por sequências de compassos que encapsulem durações
-regulares do ritmo tamborilado. Estes ritmos relacionados ao
+regulares do rítmo tamborilado. Estes ritmos relacionados ao
 tamborilar do usuário são chamados de harmônicos e podem ser
-selecionados prontamente para o disparo de linhas melódicas.
+selecionados prontamente para o disparo de linhas melódicas no ABT.
+De fato, atualmente o ABD é parte do ABT.
 
-No Apêndice XX está o manual de utilização do ABT e todo o código do
-ABT (e do ABD) está disponivel online (citar repos). Exploramos abaixo
-detalhes relevantes da implementação, em especial a forma de
-funcionamento do ABD.
-
-[códigos-chave e explicações sobre o ABT e o ABD]
+Todo o código do
+ABT está disponivel online junto à documentação para uso\footnote{Veja o
+repositório: git://labmacambira.git.sourceforge.net/labmacambira/audioArt/ABT}.
 

### 2013-01-13 musicaExtra.tex
terminada musica em tempo real

@@
 Embora o foco deste trabalho seja na exploração musical através de
-códigos (abertos!), nesta sessão iniciamos uma explicação simples,
+códigos computacionais, nesta sessão está explicação simples,
 clara e factual de como estas investidas transcendem o código e até
@@
 
-Mais relevantes que os desenvolvimentos em si, são as mobilizações
-criadas nos entornos e os engajamentos. Isso ficará evidente nas
-próximas - curtas - sessões destes desenvolvimentos e resultados.
+Por vezes, mais relevantes que os próprios desenvolvimentos são as mobilizações
+criadas nos entornos e os engajamentos.
 
-Aqui apresentamos dois trabalhos que geraram alguma movimentação de
-pessoas, resultando em reuniões, desenvolvimentos, pesquisas e
+A seguir estão dois trabalhos que geraram alguma mobilização,
+resultando em trocas, reuniões, desenvolvimentos, pesquisas e
 apresentações propriamente ditas. O primeiro utiliza o estado do
-hardware como entrada, o segundo se trata de uma mesa escultural para
-flutuação de origamis que resulta em um instrumento musical bastante
+hardware como entrada, o segundo consiste na
+flutuação de origamis para a obtenção de um instrumento musical
 lúdico.
@@
 
-Em 2008, colaborando intensamente com o CDTL (Centro de
-Desenvolvimento de Tecnologias Livres)\footnote{foi uma associação
-  civil formada e desmembrada em 2008 e sediada em Recife, PE} foi
+Em 2008, em colaboração intensa com o CDTL (Centro de
+Desenvolvimento de Tecnologias Livres)\footnote{Uma associação
+  civil formada e desmembrada em 2008, sediada em Recife, PE.} foi
 lançada a ideia de utilizar o estado do sistema operacional -
-especialmente o kernel linux - para geração de sons. Surge o Emotional
-Kernel Panic (EKP) na colaboração excepcional de Felipe
-Machado\footnote{Importantíssimo para o desenvolvimento da Cultura
-  Digital no Brasil [citar fontes e programas governamentais]},
-Ricardo Brazileiro\footnote{Artivista Digital bastante ativo [citar
-    fontes]} e o primeiro autor do presente trabalho.
+especialmente o kernel linux - para gerar de sons. Surge então o 
+'\emph{Emotional Kernel Panic}' (EKP) na parceria de Felipe
+Machado (então coordenador do CDTL),
+Ricardo Brazileiro (artista conhecedor de tecnologias livres)
+e o primeiro autor do presente trabalho.
 
-Desde o inicio, foram definidos três finalidades para esta exploração
-do SO:
-
-[visitar o README do EKP e achar os scripts do Brazileiro sobre o EKP]
+Foram definidos três finalidades para esta exploração
+do Sistema Operacional:
 
 \begin{itemize}
-\item Didáticos
-\item Artísticos
-\item Monitoramento do SO
+\item Pedagógicos, através da utilização de um sentido que não o visual para a tarefa de analisar o Sistema.
+\item Artísticos, para apresentações musicais/audiovisuais.
+\item Monitoramento do SO, pela emissão de sons periódicos relativos à carga de processamento, memória, uso de rede, etc.
 \end{itemize}
 
-Toda a base do EKP está aqui: http://trac.assembla.com/audioexperiments/browser/ekp-base
-
-[Patches desenvolvidos por Brazileiro/Machado e rascunho do EKP-Monitor]
+Esta empreitada se desdobrou em apresentações na Conferência Internacional de PureData (2009), SESC (2009),
+e no III Festival Contato (2009). No Festival Internacional de Software Livre de 2010 foi apresentada
+uma pesquisa sobre o EKP com diversos patches feitos por Ricardo Brazileiro e Felipe Machado, assim
+como amadurecimentos da proposta.
+Uma implementação conceitual do EKP está aqui: \url{http://trac.assembla.com/audioexperiments/browser/ekp-base}.
 
@@
 uma rede de coolers reciclados faz flutuar origamis de geometria e
-cores variadas. Os movimentos dos origamis são captados por webcam e
+cores variadas. Os movimentos dos origamis são captados por \emph{webcam} e
 interpretados em tempo real por software de processamento de imagens,
 gerando padrões que controlam a transformação sonora da música. Dessa
-forma, pode-se dizer que os sons gerados refletem o voo dos origamis
-de acordo com suas geometrias (que geram trajetórias de voo
-caracteristicas).
+forma, pode-se dizer que os sons gerados refletem o vôo dos origamis
+de acordo com suas geometrias (que geram trajetórias de vôo
+caracteristicas). Dito de maneira mais teórica, as estruturas fora
+do tempo (geometrias dos origamis) estão sendo mapeadas em estruturas
+temporais (trajetórias dos origamis em seus vôos), o que é um macete
+da música erudita desde tempos antigos.
 
 Na versão atual da AirHacktable, cada cor (vermelho, amarelo, azul,
-verde, preto, ou branco) controla uma voz, e a posição do origami na
-mesa modula aspectos do som. Em outras palavras, se o origami está
-flutuando mais à esquerda, o som sai no canal à esquerda da mesa, e se
-o origami está mais próximo da câmera, o volume aumenta, e se está
-mais afastado do operador, o som se torna mais agudo.
+verde, preto, ou branco) controla uma voz, já a posição do origami na
+mesa controla aspectos do som. Como um exeplo, pode-se
+configurar a AHT de forma que, se o origami está
+flutuando para a esquerda, o som também se move para esquerda, se
+o origami está mais próximo da câmera, o volume é maior, e se está
+mais afastando do operador, o som se torna mais agudo.
 
@@
 livre, que atualmente foi incorporado ao movimento chamado hardware
-livre ou aberto.
+livre ou aberto. A concepção inicial foi de Chico Simões (mestre de maracatu
+e conhecedor de tecnologias livres) e o autor
+deste texto, através do amadurecimento de possiveis instrumentos musicais de papel. 
+Logo neste estágio inicial, foram aproximados Vilson Vieira, Ricardo Fabbri, Fábio Simões
+e Daniel Penalva em reuniões presenciais. Destas reuniões, saíram ideias
+diversas, escritas em um \emph{pad} aberto\footnote{Veja: \url{http://pontaopad.me/origami-sensores}.},
+e, por fim, a AHT em si.
 
-[citar apresentações no contato, avav, oficinas sesc, chico, oficina semana imagem e som]
+A primeira realização da mesa se deu no V Festival Contato, no \emph{Espaço Macambira},
+com apresentação na Teia Casa de Criação, durante o mesmo festival. A estrutura
+da mesa é também de papel (papelão, e foi feita
+por Francisco Simões por pesquisa própria. O uso do reticulado de ventoinhas foi fruto
+de discussões coletivas e de experimentos do coletivo labMacambira.sourceforge.net,
+com especial empenho do Francisco e uso de sua luthearia de percussão: \emph{Tora Tambores}.
 
-[imagens da AHT e dos patches em operação, ver com renato quais ele quer]
+Em 2012 houve a aproximação de Caleb Mascarenhas Luporini e "Gera" Magela.
+Foram feitas oficinas no SESC Pinheiros e SESC Belenzinho e apresentação
+no AVAV\footnote{Evento mensal que ocorre na cidade São Paulo: AudioVisual Ao Vivo.},
+todas com o foco na construção e uso da AHT.
 
-\subsection{Música no Tecido Social: Sabrina Kawahara, Audioexperiments, EstudioLivre.org, CDTL, juntaDados.org, Devolts.org, MSST, LabMacambira.sf.net}
+\subsection{Música no tecido social: Sabrina Kawahara, Audioexperiments, EstudioLivre.org, CDTL, juntaDados.org, Devolts.org, MSST, LabMacambira.sf.net}
 
-Esta é uma sessão menor, dedicada a apontar repercussões emergentes
-destas empreitadas em comunidades diferentes e então tornar
-compreensível o desdobramento que estes códigos dedicados à música
-tiveram em processos sociais e mobilizações civis.
-
-Grupos foram montados em torno destes desenvolvimentos. Como os
-propósitos de compartilhamento e apropriação tecnológica estavam no
-cerne dos grupos relacionados, as investidas naturalmente tomaram
-teores engajados socialmente. A questão do empoderamento das pontas e
-da criação de um patrimônio tecnológico da humanidade é consequência
-quase imediata das posturas de compartilhamento e apropriação citados.
-
-Discorremos brevemente sobre cada uma das iniciativas citadas:
+A singularidade do cruzamento artístico, cultural e tecnológico
+dos trabalhos expostos também é desenvolvida por diferentes grupos.
+Os propósitos de compartilhamento e apropriação tecnológica estão no
+cerne destas comunidades, de forma que as investidas naturalmente tomam
+rumos engajados socialmente. Em especial, o empoderamento civil e
+a criação de um patrimônio tecnológico da humanidade são consequências
+imediatas das posturas de compartilhamento e apropriação citados.
 
 \begin{itemize}
-    \item Sabrina Kawahara
-    \item Audioexperiments
-    \item Estudiolivre.org
-    \item CDTL
-    \item JuntaDados.org
-    \item Devolts.org
-    \item MSST
-    \item LabMacambira.sf.net
-    \item Outros relacionados: MuSa, Metareciclagem, Submidialogia, Tainã
+    \item Sabrina Kawahara: formado pelo autor deste texto, Guilherme Lunhani,
+    Guilherme Rebecchi, Israel Laurindo, Rodrigo Felício e Otávio Martigli,
+    este grupo dedicou-se à realização
+    de apresentações conjuntas, bem como a escrita de peças e textos através
+    de dinâmicas coletivas.
+    \item Audioexperiments (Æ): iniciado por Tiago Tavares, o autor do texto e aproximações de
+    Daniel Pastore. O grupo se dedicou à publicação de textos didáticos e produção
+    de código em repositórios públicos. Em destaque, o repositório no \emph{Assembla}
+    possui códigos em Python e C/C++ que realizam diversos experimentos em áudio, música
+    e inteligência artificial\footnote{Veja a página principal do espaço: \url{https://www.assembla.com/spaces/audioexperiments}.}.
+    \item Estudios Livres: são ambientes de estudo, produção e distribuição de mídias livres.
+    Formam, sem conjunto, um movimento e uma rede social brasileira, tida como única no mundo, embora
+    o conceito seja conhecido internacionalmente\footnote{Há uma plataforma aberta com capacidades
+    de \emph{wiki}, blog e midiateca em \url{http://estudiolivre.org}.}. Tanto a comunidade quanto
+    a lista operacional e plataforma da comunidade foram essenciais nos amadurecimentos de questões
+    relacionadas às mídias livres e ao uso de Python para áudio e música, como pode ser visto
+    na Carta Mídias Livres e no Tutorial de Python para Áudio e Música, abordados abaixo
+    em~\ref{subsec:tuts} deste mesmo Apêndice.
+    \item CDTL (Centro de Desenvolvimento de Tecnologias Livres): foi uma associação civil com base em Recife/PE, dedicada ao desenvolvimento e difusão de tecnologias livres.
+Esta associação proporcionou suporte para os primeiros materiais didáticos e investidas em plugins de áudio através do convênio estabelecido com o Ministério da Cultura de Pontão de Cultura Digital.
+    \item JuntaDados.org: é um grupo com membros em todo o território brasileiro e ainda ativo.
+Embora tenha havido um momento de maior institucionalidade, durante o convênio de Pontão de Cultura Digital com o Ministério da Cultura e parceria com a Universidade Estadual da Bahia (UNEB), o grupo é marcado pela descentralização e autonomia dos envolvidos.
+O autor é membro fundador e ativo do juntaDados, em conjunto, especialmente, com Fabiana "Goa" Sherine e Marcelo Soares Souza.
+Neste grupo foram desenvolvidas ideias como o EKP, o ABT, os tutoriais de Python para Áudio e escritos socialmente engajados.
+    \item Devolts.org: responsável por Esporos de Pesquisa e Experimentação do programa Cultura Viva, os "Jardins Devolts" cultivam listas nacionais de trocas de conhecimentos livres e cultivo de informações voltadas para música com usos especiais de \emph{PureData}, \emph{Arduinos} e \emph{Python}.
+Este é um meio de troca rico e alternativo a alguns meios mais politizados, como a lista do \emph{estudiolivre.org}.
+    \item MSST (Movimento dos Sem Satélite): com um manifesto e propostas coerentes, mas nada óbvias, como a demanda específica por satélites civís, abordagens tecnomágicas, o grupo é uma instâncias iniciáticas em termos tecnológicos e conceituais.
+    \item LabMacambira.sf.net: espaço virtual de programadores dedicados ao audiovisual e propósitos socialmente engajados.
+Fundado em parceria do autor deste trabalho e de Vilson Vieira, Daniel Marostegan e Carneiro e Ricardo Fabbri e de parte dos investimentos da segunda etapa do convênio de Pontão de Cultura Digital Nós Digitais, da Teia Casa de Criação.
+Em um primeiro momento, membros foram remunerados como mentores e como aprendizes\footnote{Renato Fabbri, Ricardo Fabbri, Vilson Vieira, Marcos Mendonça e Gilson Beck mentoraram; Alexandre "Bzum" Negrão, Lucas Zambianchi, Larissa Arruda, Nivaldo Henrique Bondança, Fernando Gorodscy, Andres Martano desenvolveram como bolsistas; Danilo Shiga e Marcos Murad foram parceiros colaboradores; Francisco Simões, Fábio Simões, Caleb Mascarenhas Luporini, "Gera" Magela, Edson "Presto" Correia, Daniel Penalva se aproximaram para atividades específicas, cruciais em todo ano já sem os investimentos do convênio.}.
+Em um segundo momento, aproximações e parcerias diversas ocorreram.
+Detalhes estão mas subseções seguintes de \emph{Web} e materiais didáticos.
+    \item Submidialogia: grupo bastante relacionado aos encontros de '\emph{submidialogias}, ao Descentro e às ideias de submídia.
+A lista de emails do grupo possui membros de diversos dos outros grupos citados.
+    \item Outros relacionados: o MuSa foi um grupo de JoinVille importante neste trabalho por ter sido fundado por Vilson Vieira, que possui presença forte no LabMacambira.sf.net e nos trabalhos aqui descritos.
+O  Metareciclagem é uma rede engajada em apropriação tecnológica e empoderamento civil com bastante presença dos outros grupos aqui citados e muito movimentada.
+A Casa de Cultura Tainã é uma entidade cultural de Campinas/SP, referência no uso de tecnologias livres e atividades socialmente engajadas, e onde muito do que está aqui descrito foi inspirado.
 \end{itemize}
@@
 \section{Web}
-
+Houve tentativas, como do grupo APISTO, 
   Difusão de informação com ênfase na facilitação
@@
 
-  \subsection{Tutoriais em texto e código: Filtros, Nyquist e plugins LADSPA}
+  \subsection{Tutoriais em texto e código: Filtros, Nyquist e plugins LADSPA}\label{subsec:tuts}
 

### 2013-01-13 musicaExtra.tex
terminada musica em tempo real

@@
 
-\section{Web}
-Houve tentativas, como do grupo APISTO, 
-  Difusão de informação com ênfase na facilitação
-  da apropriação de tecnologias e de instancias políticas.
-
-   \subsection{Tecnologias sociais de alta demanda: Sitios, Conteúdos e Articulação}
-
-      \subsubsection{Sítios}
-
-      FDDCA
-
-      Ferramenta de comunicação
-
-      (Cadastro dos pontos?)
-
-      AA, SOS, Catalogo de Ideias, etc
-
-      Meu site pessoal
-
-      \subsubsection{Conteúdos}
-
-      Wiki?
-
-      \subsubsection{Articulação}
-
-      IRC, Emails
-
-\subsection{Disponibilização e desenvolvimento conjunto: wikis, etherpads, AA, Trac, IRC ..}
-
-\subsubsection{Wiki}
-
-\subsubsection{Trac}
-
-\subsubsection{Screencasts - Vimeo}
-
-\subsubsection{AA}
-
-\subsubsection{Audio Experiments (Æ)}
-
-\subsubsection{IRC}
-
-\subsubsection{Etherpads}
-
-\subsubsection{Outras fontes}
-
 \section{Materiais didáticos}
 
-  \subsection{Tutoriais em texto e código: Filtros, Nyquist e plugins LADSPA}\label{subsec:tuts}
-
-Os vários materiais didáticos produzidos constam no apêndice
-deste trabalho. Um único destes será exposto a seguir por sua
-capacidade de agregar os conteúdos dos capítulos anterioes:
-os tutoriais de filtros e amostragem.
+  \subsection{ Tutoriais em texto e código: python, filtros, nyquist e plugins LADSPA, metrics, carta mídias livres, contra-cultura digital }\label{subsec:tuts}
 
@@
 e o repositório ~\cite{http://sourceforge.net/projects/fdpack/develop fdpsf}.
+
+\section{Web}
+  Difusão de informação com ênfase na facilitação
+  da apropriação de tecnologias e de instancias políticas.
+
+   \subsection{Tecnologias sociais de alta demanda: Sitios, Conteúdos e Articulação}
+
+      \subsubsection{Sítios}
+
+      FDDCA
+
+      Ferramenta de comunicação
+
+      (Cadastro dos pontos?)
+
+      AA, SOS, Catalogo de Ideias, etc
+
+      Meu site pessoal
+
+      \subsubsection{Conteúdos}
+
+      Wiki?
+
+      \subsubsection{Articulação}
+
+      IRC, Emails
+
+\subsection{Disponibilização e desenvolvimento conjunto: wikis, etherpads, AA, Trac, IRC ..}
+
+\subsubsection{Wiki}
+
+\subsubsection{Trac}
+
+\subsubsection{Screencasts - Vimeo}
+
+\subsubsection{AA}
+
+\subsubsection{Audio Experiments (Æ)}
+
+\subsubsection{IRC}
+
+\subsubsection{Etherpads}
+
+\subsubsection{Outras fontes}

### 2013-01-14 musicaExtra.tex
atualizando dissertacao, apendice extra de mus

@@
 Este tutorial foi levado para Berlim no LAC 2007 e sofreu melhoras desde entao. Esta
-primeira versao ficou resumida em forma de texto no EL\footnote{http://estudiolivre.org/python-e-som-tutorial}. Em 2010
-a Associacao Python Brasil escolheu este trabalho, então já mais amadurecido, para ser apresentado no
-FISL em Porto Alegre. Como consequencia, foi feita uma série de video-tutoriais bastante utilizados\footnote{http://estudiolivre.org/tiki-index.php?page=Video+Tutoriais}.
-Este tutorial foi comentado em listas em que o autor não participa (e outras em que o autor participa).
+primeira versao ficou resumida em forma de texto no EL\footnote{\url{http://estudiolivre.org/python-e-som-tutorial}.}. Em 2010
+a Associacao Python Brasil escolheu este trabalho, então já mais maduro, para ser apresentado no
+Festival Internacional de \emph{Software} Livre, em Porto Alegre.
+Como consequência, foi feita uma série de video-tutoriais bastante utilizados\footnote{http://estudiolivre.org/tiki-index.php?page=Video+Tutoriais}.
+Este tutorial foi comentado em listas de \emph{emails} e mensagens pessoais. Os vídeos são tidos como iniciáticos em Python por diferentes pessoas ligadas ao movimento de \emph{software} livre.
 
-    \item {\bf Tutoriais de filtros e amostragem via python}
+    \item {\bf Tutoriais de filtros e amostragem via python }
 
-Voltados para explicitar principios fundamentais de áudio, estes tutoriais
+Voltados para explicitar fundamentos de áudio digital, estes tutoriais
 são baseados código Python e o equivalente em C. Pequenas explicações são

### 2013-01-14 musicaExtra.tex
descritos os textos + pedagogicos, parei nos videos

@@
 A lista de emails do grupo possui membros de diversos dos outros grupos citados.
-    \item Outros relacionados: o MuSa foi um grupo de JoinVille importante neste trabalho por ter sido fundado por Vilson Vieira, que possui presença forte no LabMacambira.sf.net e nos trabalhos aqui descritos.
+    \item Outros grupos relacionados: o MuSa foi um grupo de JoinVille importante neste trabalho por ter sido fundado por Vilson Vieira, que possui presença forte no LabMacambira.sf.net e nos trabalhos aqui descritos.
 O  Metareciclagem é uma rede engajada em apropriação tecnológica e empoderamento civil com bastante presença dos outros grupos aqui citados e muito movimentada.
@@
 
-Este tutorial foi levado para Berlim no LAC 2007 e sofreu melhoras desde entao. Esta
-primeira versao ficou resumida em forma de texto no EL\footnote{\url{http://estudiolivre.org/python-e-som-tutorial}.}. Em 2010
+Este tutorial foi apresentado parcialmente em Berlim no LAC 2007 e, desde então, foi melhorado algumas vezes. Esta
+primeira versao ficou resumida em forma de texto na plataforma
+Estúdio Livre\footnote{\url{http://estudiolivre.org/python-e-som-tutorial}.}.
+Um agradecimento especial para Fábio Furlanete e Marília Chiozo pelas contribuições.
+Em 2010
 a Associacao Python Brasil escolheu este trabalho, então já mais maduro, para ser apresentado no
-Festival Internacional de \emph{Software} Livre, em Porto Alegre.
-Como consequência, foi feita uma série de video-tutoriais bastante utilizados\footnote{http://estudiolivre.org/tiki-index.php?page=Video+Tutoriais}.
-Este tutorial foi comentado em listas de \emph{emails} e mensagens pessoais. Os vídeos são tidos como iniciáticos em Python por diferentes pessoas ligadas ao movimento de \emph{software} livre.
+Festival Internacional de \emph{Software} Livre, em Porto Alegre. Como consequência, 
+foi feita uma série de video-tutoriais\footnote{http://estudiolivre.org/tiki-index.php?page=Video+Tutoriais}.
+Os vídeos são tidos como iniciáticos em Python por diferentes pessoas ligadas ao movimento de \emph{software} livre.
 
@@
 dadas com o intuito de orientar a exploração inteligente destes \emph{snippets}.
+\emph{Teorema de Amostragem}: estes \emph{scripts} executam experimentações ilustrativas com
+o Teorema de Nyquist e figuras.
+\emph{Filtros}: além de filtros FIR e IIR simples,
+duas utilizações clássicas destes filtros estão implementadas
+de forma didática: Wavelets (FIR) e Quad (IIR).
 
-\emph{Teorema de Amostragem}: estes scripts visam a experimentacao inteligente com
-o Teorema de Nyquist. (descricao)
-
-\emph{Filtros}: alem da explicitação sobre as diferenças entre filtros FIR e IIR,
-duas utilizações clássicas destes filtros estão implementadas: Wavelets (FIR) e Quad (IIR)
-
-Estes códigos podem ser baixados no repositório SVN do AudioExperiments. E os textos estão
-na wiki (nos digitais ou EL, recriar pois os CDTL foram apagados)
+Estes códigos didáticos podem ser baixados do repositório \emph{AudioExperiments}.
 
@@
 Dadas as dificuldades que o desenvolvimento dos \emph{plugins} de áudio apresenta,
-desenvolvi um tutorial passo a passo com plugins que rodam em todas as etapas.
+foi feito um tutorial passo a passo com plugins que rodam em todas as etapas.
 Ele é baseado em uma interface C++ para este padrão de plugin que eh implementado
@@
 o entendimento do funcionamento dos princípios fundamentais deste framework.
+Uma das primeiras incidências dos \emph{scripts} mínimos, muito usados no
+labMacambira.sourceforge.net para passagens de tecnologias de forma precisa.
 
-    \item {\bf Philosometrics}
+     \item {\bf Figusdevpack (FDP)}
 
-Embora este não seja um trabalho didático propriamente dito, ele tem este intuito
-no cerne de sua concepção e surgimento. Em decorrência dele, surgiu o  Musimetrics,
-o Cinemetrics e o Literametrics. Além disso, ele é um belo exemplo da
-utilização das ciências duras para a análise de ciências humanas e foi acolhido
-como tal em alguns momentos.
+Idealizado como um meio de interação da comunidade de Python e Música 
+para compartilhamento de códigos e excertos, o FDP foi Baseado principalmente
+em documentação organizada sobre as práticas e as bibliotecas
+existentes para Python. Parte desta documentação
+é proposta como \emph{scripts} que
+fazem uso de objetos e módulos de forma isolada, puntual.
+ Este trabalho foi aceito na
+maior conferência de áudio em linux, a Linux Audio Conference de 2008
+(LAC2008) e foi reativado algumas vezes pelo autor deste texto
+em conjunto com Vilson Vieira, Ivan Marin
+e outros desenvolvedores. Este projeto está
+documentado na plataforma do Estúdio Livre\footnote{
+Veja a página do PDF \url{http://estudiolivre.org/tiki-index.php?page=fdp&highlight=fdp}, o artigo aceito no LAC2008 \url{http://www.estudiolivre.org/el-gallery_view.php?arquivoId=8221}
+e o repositório do FDP \url{http://sourceforge.net/projects/fdpack/develop fdpsf}.}.
 
@@
 'Prêmio Mídias Livres', a convite do Ministério da Cultura por 'notório saber'.
-Esta carta é um documento único no seu conteúdo, deixando às claras
-o conceito de Mídias Livres como mídias não aprisionadas pelo conceito
+A participação consistia em analisar os inscritos no Prêmio Mídias Livres e distribuir 4 milhões de reais dentre categorias regionais e nacionais.
+Esta carta é um documento único, deixando às claras
+o conceito de Mídias Livres como não aprisionadas pelo conceito
 de propriedade, ou seja, que priorizam a sua livre circulação e a possibilidade
-de geração de materiais derivados, assim como sua geração aberta ao colaborativo e comunitário.
+de geração de materiais derivados.
+Há o viés de priorizar processos colaborativos, comunitários e setores da sociedade menos contemplados na geração e circulação midiática\footnote{Veja: \url{http://www.estudiolivre.org/carta-pts-midias-livres}}.
 
-    \item {\bf Textos de cunho sociológico, transformador}
+    \item {\bf Philosometrics}
 
-Produção mais numerosa que as anteriores, se caracteriza por métodos não convencionais
-de abordagem dos assuntos e de escrita. Em especial utiliza-se pseudônimos para
-auxiliar a despersonificação, gerando textos menos presos à satisfação da auto-imagem, dente
-outras qualidades. A utiliação de psudônimos é um costume muito apreciado em diversos meios,
-e as pesquisas tem confirmado as vantagens que a prática apresenta e confirma\footnote{http://disqus.com/research/pseudonyms/}.
+Em decorrência deste trabalho, surgiu o  Musimetrics,
+o Cinemetrics e o Literametrics. Uma publicação no
+JSTAT condensa e compara as análises de filosofia
+e música.\cite{FabbriJSTAT}
+Além disso, ele é uma
+utilização das ciências duras para 
+assunto mais incidente em ciências humanas.
 
-O autor destes textos se dá ao direito de não revelar seus psudônimos - embora muitos deles
-sejam publicamente conhecidos - para conservar as consequências desta prática na
-forma mais pura. Como comprovante desta produção, deixamos uma mensagem sobre a publicação
-de textos em mídia impressa com autores internacionais,
-confirmando a participação do autor desta dissertação, mas cujo
-nome não consta na publicação.
+    \item {\bf Textos socialmente engajados }
 
-!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-
-de      fabi borges catadores@gmail.com por  riseup.net 
-responder a     submidialogia@lists.riseup.net
-para    submidialogia@lists.riseup.net
-data    23 de agosto de 2011 12:26
-assunto Re: [submidialogia] livro sub- publicação
-lista de e-mails        <submidialogia.lists.riseup.net> Filtrar as mensagens dessa lista de e-mails
-enviado por     lists.riseup.net
-assinado por    riseup.net
-cancelar inscrição      Cancelar a inscrição para essa lista de e-mails
-        Importante principalmente porque você frequentemente lê mensagens com esse marcador.
-ocultar detalhes 12:26 (6 minutos atrás)
-entao, eu fui recebendo textos durante esse tempo,
-alguns tao atrazados como dos sem satelites, mas muita gente mandou;
-
-aqui os autores:
-
-os internacionais nao sao muitos, o joni kempf (o que bebe ouro do hardware), o barbrook (futuros imaginarios),
-Hamdy heda (da revolucao egipcia), o pedro soller (summerlab), maria llopis (pos porno),  talvez a bronac, nao entregou ainda.
-
-dos brasileiros, renato fabri, ruiz, pasteur, morgana e caio, ju dornelles, mari marcassa, coletivo errorista, adriana velozo-drica, tiago pimentel, felipe fonseca, thiago novaes, lelex, felipe ribeiro(?), maira, verenilde,  bartolina silva, poro, vitoria amaro, fabib (eu),
-
-entao, precisa publicar agora,
-uma equipe para publicacao e,,,
-
-bjs
-f
-
-!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+O uso de pseudônimos é um costume apreciado em diversos meios. As pesquisas informais confirmam vantagens desta prática\footnote{Veja: \url{http://disqus.com/research/pseudonyms/}}.
+Em especial utiliza-se pseudônimos para
+auxiliar a despersonificação, gerando textos menos presos à satisfação da auto-imagem.
+Como resultado, além de aumento de produtividade, constuma-se conseguir também compreensões diferentes.
+Destes textos, dois foram publicados em duas publicações relevantes: O Contracultura Digital e o Peixe Morto, este último relacionado aos submidialogias\footnote{Veja: \url{http://mutgamb.org/blog/Submidialogias-Peixe-Morto-para-Baixar} e \url{http://culturadigital.br/contraculturadigital/2012/02/01/publicacao-contraculturadigital/}.}.
 
@@
 
-\subsection{Screencasts e outros materiais em video e em texto}
+\subsection{Screencasts e outros materiais em video}
 
@@
     \item Python para áudio e música
-	  Texto - palestras - videos
+    Relacionado ao tutorial em texto citado acima, foram feitos videos sobre
 
@@
 \end{itemize}
-
-\subsection{Figusdevpack (FDP)}
-
-Um ambiente de interação da comunidade de Python e Música 
-para compartilhamento de códigos e excertos. Baseado principalmente
-em documentação organizada sobre as práticas e as bibliotecas
-existentes para python. Assim como scripts compartilhados que
-fazem uso de objetos e módulos específicos. Este trabalho foi aceito na
-maior conferência de áudio em linux, a Linux Audio Conference de 2008
-(LAC2008) e está sendo reativado por mim em conjunto com Vilson Vieira
-e outros desenvolvedores de áudio. Este projeto está em desenvolvimento
-do site do Estúdio Livre [estudio livre] com repositório no sourceforge [source force].
-
-As principais fontes sobre estre trabalho é a página de desenvolvimento da ideia
-que está em constante mudança ~\cite{http://estudiolivre.org/tiki-index.php?page=fdp&highlight=fdp fdpel}
-e, o artigo que foi aceito no LAC de 2008 ~\cite{http://www.estudiolivre.org/el-gallery_view.php?arquivoId=8221 fdplac2008}
-e o repositório ~\cite{http://sourceforge.net/projects/fdpack/develop fdpsf}.
 

### 2013-01-17 musicaExtra.tex
finalizada primeira passada no Apendice todo

@@
 \begin{itemize}
-    \item Python para áudio e música
-    Relacionado ao tutorial em texto citado acima, foram feitos videos sobre
+    \item {\bf Python para áudio e música}
+    Relacionado ao tutorial em texto citado acima, foram feitos videos sobre a utilização da linguagem Python,
+    em grande parte através de usos em \emph{IPython}, para leitura e escrita de arquivos de áudio com bibliotecas
+    padrão e externas, assim como uso básico de wavelets,OSC e outros recursos.
 
-    \item Canal Macambira
-No Macambira estão sendo produzidos materiais em screencasts sobre
-diversas cenas de hackeamento.
+    \item {\bf Canal Macambira no Vimeo}
+    Atualmente, são mais de 700 videos sobre sessões de \emph{hacking} e experiências de arte
+    e tecnologia. Feitos pelo LabMacambira.sourceforge.net, o canal tem videos do auto
+    deste escrito, e de: Larissa Arruda, Alexandre Negrão, Lucas Zambianchi, Andres Martano,
+    Fernando Gorodscki, Chico Simões, Daniel Marcicano, Daniel Pizetta, Daniel Penalva, Nivaldo Bondança,
+    Danilo Shiga, Marcos Murad, Ricardo Fabbri e Vilson Vieira.\footnote{O canal pode ser acessado em:
+    \url{http://vimeo.com/channels/labmacambira}.}
+    \begin{itemize}
+	    \item \emph{Livecoding}: como exemplo dos videos do canal, foram expostos os processos utilizados
+        na apresentação de \emph{livecoding} com ChucK, vacas e sono REM.\footnote{Veja: \url{https://vimeo.com/33012735},
+        \url{https://vimeo.com/33018740}, \url{https://vimeo.com/33019291}, \url{https://vimeo.com/33025717},
+        \url{https://vimeo.com/33025913}.}
 
-    \begin{itemize}
-	\item Live-Coding
-	\item Raspagem de dados
+	    \item Raspagem de dados: outro exemplo, menos artístico e mais engajado, é uma breve explicação
+        sobre o processo de raspagem de dados em páginas HTML.\footnote{Veja: \url{vimeo.com/channels/labmacambira/2681879}.}
+
+        \item Outros videos incluem gambiarras artísticas, uso de scripts/ferramentas para audiovisual, etc.
     \end{itemize}
@@
 \section{Web}
-  Difusão de informação com ênfase na facilitação
-  da apropriação de tecnologias e de instancias políticas.
+Esta dissertação é focada em uma descrição física do áudio digital enquanto música e
+usos em código desta abordagem.
+Como pode-se observar acima, neste mesmo Apêndice, este trabalho, disponibilizado e conebido como tecnologias livres, tem dimensões bastante consideráveis em aspectos socialmente engajados e artísticos.
+Esta seção exibe os feitos que extrapolam a música por completo.
 
@@
 
-      FDDCA
+        Alguns \emph{sites} foram criados, como parte de trabalhos engajados e tentativa de manutenção das estruturas do labMacambira.sourceforge.net.
+O primeiro site com alguma relevância seja talvez o site para a Casa dos Meninos / LIDAS, em que foi feita uma estrutura para acolher as conferências setoriais e municipais dos direitos da criança e do adolescente da cidade de São Paulo.
+Outros exemplos, mais recentes, são as páginas para a Rede Nacional de Cultura Ambiental Afrobrasileira e os blogs para a Rio+20 das secretarias do Ministério da Cultura.
 
-      Ferramenta de comunicação
-
-      (Cadastro dos pontos?)
-
-      AA, SOS, Catalogo de Ideias, etc
-
-      Meu site pessoal
+Nest emeio tempo também foram criadas várias ferramentas \emph{web}, que rodam em navegadores comuns, como \emph{Firefox}. Destes, vale citar:
+\begin{itemize}
+    \item SOS: sistema para coleta e difusão de conhecimentos populares e étnicos ligados à saúde.
+    \item Ágora Communs, Novo Ágora: sistema de deliberações \emph{online}.
+    \item Economia criativa: rascunho de sistema para facilitar circulação de bens me 4 eixos: venda de produtos, aluguél de equipamentos, aluguél de lugares e prestação de serviços.
+    \item Catálogo de Ideias: ferramenta simples para catálogo de ideias sobre temas diferentes, junto a palestras, etc.
+    \item Mapeamento dos 301 pontos de SP: mapeamento dos 301 pontos de cultura estaduais.
+    \item Mapas Coletivos: ferramenta para criação e publicação de mapas com possibilidades colaborativas.
+    \item Leitura de \emph{shapefiles} no Mapas de Vista: implementação de leitura de arquivos do tipo \emph{shapefile} com dados geográficos em um plugin e tema de \emph{Wordpress}.
+    \item Maper: atualmente, um rascunho de uma ferramenta de publicação de mapas com vistas a publicações no facebook e outros formatos.
+\end{itemize}
 
@@
 
-      Wiki?
+      Os conteúdos estão em diversas páginas da \emph{wiki} do Nós Digitais e do Estúdio Livre.\footnote{Veja: \url{http://wiki.nosdigitais.teia.org.br/Lab_Macambira}, \url{http://www.estudiolivre.org/el-user.php?view_user=gk}.} Tratam principamente de documentações de \emph{software} e receitas para acesso a repositórios ou instalações.
+Nestas documentações também estão apontamentos de recursos artísticos e apresentações realizadas com a AHT, \emph{livecoding} e outras propostas.
 
-      \subsubsection{Articulação}
+      \subsubsection{Articulação, disponibilização e desenvolvimento conjunto}
 
-      IRC, Emails
-
-\subsection{Disponibilização e desenvolvimento conjunto: wikis, etherpads, AA, Trac, IRC ..}
-
-\subsubsection{Wiki}
-
-\subsubsection{Trac}
-
-\subsubsection{Screencasts - Vimeo}
+As articulações se dão principalmente através do IRC, do AA e de emails, além de conversas presenciais.
+O IRC é uma forma de comunicação leve e focada na comunicação coletiva.
+O AA está explicado abaixo.
+Não menos importante são os \emph{Etherpads} e as páginas \emph{wiki} criadas, permitindo proposições e escritas rápidas e em várias mãos.
 
@@
 
-\subsubsection{Audio Experiments (Æ)}
-
-\subsubsection{IRC}
-
-\subsubsection{Etherpads}
-
-\subsubsection{Outras fontes}
+Uma das tecnologias desenvolvidas nesta empreitada se propõe a estabelecer formas de compartilhar andamentos e processos.
+As motivações incluem transparência civil e organização assíncrona de grupos descentralizados.
+O sistema se baseia no envio de mensagens periódicas sobre as atividades que estão sendo desempenhadas, evitando foco em produtos e propaganda.

### 2013-01-18 pretextual.tex
iniciada secao 2.2
terminada lista de relacoes

@@
 
-\afterpage{\blankpage}
-
-\pretextualchapter{DEDICATÓRIA} %Título da página dedicatória
-	\input{dedicatoria} %Para incluir o texto de dedicatória,use o arquivo dedicatória.tex
+%\afterpage{\blankpage}
+%
+%\pretextualchapter{DEDICATÓRIA} %Título da página dedicatória
+%	\input{dedicatoria} %Para incluir o texto de dedicatória,use o arquivo dedicatória.tex
 
@@
 %\listadetabelas %Comando que gera a lista de tabelas (AUTOMÁTICO)
+\afterpage{\blankpage}
 
-\pretextualchapter{Lista de relações descritas e implementadas}
+\pretextualchapter{Lista de relações analíticas descritas e implementadas computacionalmente no Apêndice~\ref{cap:codigoProc}}
 	\begin{listaespecial}[BIGNAMEWIDTH]
-		\item[AM]: modulação de amplitude como descrita em~\ref{}. Implementação computacional em Python com Numpy e Scikits/Audiolab no apêndice~\ref{}.
+        \item[Equação~\ref{eq:dur}] Duração em áudio PCM.
+        \item[Equação~\ref{eq:potencia}] Potência.
+        \item[Equação~\ref{decibels}] Decibels entre dois áudios.
+        \item[Equação~\ref{eq:ampVol}]  Amplitude dobrada em decibels.
+        \item[Equação~\ref{eq:potVol}] Potência dobrada em decibels.
+        \item[Equação~\ref{eq:dobraVol}] Variação de amplitude no volume dobrado (10 dBs).
+        \item[Equação~\ref{ampDec}] Conversão de decibels em amplitude.
+        \item[Equação~\ref{periodicidade}] Representação básica de uma sequência periódica.
+        \item[Equação~\ref{senoide}] Sequência infinita senoidal.
+        \item[Equação~\ref{denteDeSerra}] Sequência infinita de uma onda 'dente de serra'.
+        \item[Equação~\ref{triangular}] Sequência infinita de uma onda 'triangular'.
+        \item[Equação~\ref{quadrada}] Sequência infinita de uma onda 'quadrada'.
+        \item[Equação~\ref{sampleandoFormaDeOnda}] Sequência infinita de uma forma de onda \emph{sampleada}.
+        \item[Equação~\ref{recomposicaoFourier}] Recomposição das amostras temporais com base nos coeficientes espectrais.
+        \item[Equação~\ref{moduloEfase}] Recomposição das amostras temporais reais em termos do módulo e fase.
+        \item[Equação~\ref{coefsPareados}] Número de pares de coeficientes relativos à mesma frequência.
+        \item[Equação~\ref{equivalenciasFreqs}] Coeficientes equivalentes em termos das frequêcias que representam.
+        \item[Equação~\ref{equivalenciasModulos}] Equivalências dos módulos dos coeficientes espectrais.
+        \item[Equação~\ref{equivalenciasFases}] Equivalências das fases dos coeficientes espectrais.
+        \item[Equação~\ref{eq:reconsCompleta}] Reconstrução das amostras temporais em termos dos coeficientes pareados e independentes.
+        \item[Equação~\ref{eq:notaBasica}] Nota básica com duração, frequência e altura.
+        \item[Equação~\ref{periodoUnico}] Forma de onda sintética ou amostrada.
+        \item[Equação~\ref{eq:notaBasicaTimbre}] Nota básica com forma de onda especificada.
+        \item[Equação~\ref{eq:distOuvidos}] Distância de uma fonte sonora a cada ouvido.
+        \item[Equação~\ref{eq:dti}] Diferença de Tempo Interaural (DTI).
+        \item[Equação~\ref{eq:dii}] Diferença de Intensidade Interaural (DII).
+        \item[Equação~\ref{eq:locImpl}] Sequência binaural PCM com  DTI e DII para localização espacial.
+        \item[Equação~\ref{eq:angulo}] Ângulo resolvido pela implementação da DTI e DII.
+        \item[Equação~\ref{eq:mixagem}] Mixagem de N sequências.
+        \item[Equação~\ref{eq:concatenacao}] Concatenação de N sequências.
+        \item[Equação~\ref{eq:lut}] Procedimento de busca em tabelas (Lookup Table).
+        \item[Equação~\ref{freqLinear}] Frequências em cada amostra em uma variação linear.
+        \item[Equação~\ref{indiceLinear}] Índices para busca na tabela em uma variação linear de frequência.
+        \item[Equação~\ref{serieAmostralLin}] Sequência amostral em uma variação linear de frequências.
+        \item[Equação~\ref{freqExponencial}] Frequências em cada amostra em uma variação exponencial.
+        \item[Equação~\ref{indiceExponencial}] Índices para busca na tabela em uma variação exponencial de frequência.
+        \item[Equação~\ref{serieAmostralLog}] Sequência amostral em uma variação exponencial de frequências.
+        \item[Equação~\ref{seqAmp}] Sequência de amplitudes em uma variação exponencial.
+        \item[Equação~\ref{transAmp}] Sequência amostral em uma variação exponencial de amplitude.
+        \item[Equação~\ref{seqAmpLin}] Sequência de amplitudes em uma variação linear de amplitude.
+        \item[Equação~\ref{seqAmpDB}] Sequência amostral em uma variação exponencial de amplitude dada em decibels.
+        \item[Equação~\ref{eq:conv}] Convolução de sequências reais e finitas.
+        \item[Equação~\ref{eq:diferencas}] Equação a diferenças para aplicação temporal de filtros IIR.
+        \item[Equação~\ref{eq:passa-baixas}] Coeficientes de um filtro passa-baixas bem comportado de primeira ordem com frequência de corte variável.
+        \item[Equação~\ref{eq:passa-altas}] Coeficientes de um filtro passa-altas bem comportado de primeira ordem com frequência de corte variável.
+        \item[Equação~\ref{eq:varAux}] Variáveis auxiliares de um filtro nó de segunda ordem.
+        \item[Equação~\ref{eq:passa-banda}] Coeficientes de um filtro passa-banda de segunda ordem com frequência central e largura de banda variáveis.
+        \item[Equação~\ref{eq:rejeita-banda}] Coeficientes de um filtro rejeita-banda de segunda ordem com frequência central e largura de banda variáveis.
+        \item[Equação~\ref{eq:branco}] Coeficientes espectrais para síntese de ruído branco.
+        \item[Equação~\ref{eq:rosa}] Coeficientes espectrais para síntese de ruído rosa.
+        \item[Equação~\ref{eq:marrom}] Coeficientes espectrais para síntese de ruído marrom.
+        \item[Equação~\ref{eq:azul}] Coeficientes espectrais para síntese de ruído azul.
+        \item[Equação~\ref{eq:violeta}] Coeficientes espectrais para síntese de ruído violeta.
+        \item[Equação~\ref{eq:preto}] Coeficientes espectrais para síntese de ruído preto.
+        \item[Equação~\ref{vbrGamma}] Índices auxiliares para um vibrato de frequência variável.
+        \item[Equação~\ref{vbrAux}] Expoentes auxiliares para um vibrato de frequência variável.
+        \item[Equação~\ref{vbrF}] Frequências por amostra de um vibrato de frequência e profundidade variáveis.
+        \item[Equação~\ref{vbrGamma2}] Índices um vibrato de frequência e profundidade variáveis.
+        \item[Equação~\ref{vbrT}] Amostras de um áudio com vibrato de frequência e profundidade variáveis.
+        \item[Equação~\ref{trA}] Amplitudes por amostra de um tremolo de frequência e profundidade variáveis.
+        \item[Equação~\ref{trT}] Amostras de um áudio com tremolo de frequência e profundidade variáveis.
+        \item[Equação~\ref{eq:fmEsp}] Espectro da síntese FM.
+        \item[Equação~\ref{eq:amEsp}] Espectro da síntese AM.
+        \item[Equação~\ref{fmGammaAux}] Índices auxiliares para síntese FM.
+        \item[Equação~\ref{fmAux}] Sequência auxiliar para síntese FM.
+        \item[Equação~\ref{fmF}] Sequência de frequências para cada amostra para síntese FM.
+        \item[Equação~\ref{fmGamma}] Índices para síntese FM.
+        \item[Equação~\ref{fmT}] Sequência amostral resultante da síntese FM.
+        \item[Equação~\ref{amA}] Sequência de amplitudes por amostra na síntese AM.
+        \item[Equação~\ref{amT}] Sequência amostral resultante da síntese AM.
+        \item[Equação~\ref{eq:vinculos}] Exemplo de vínculo entre a frequência e parâmetros do tremolo e do vibrato em um som.
+        \item[Equação~\ref{eq:adsr}] Envoltória ADSR com variações lineares e logarítmicas.
+        \item[Equação~\ref{eq:adsrApl}] Aplicação da envoltória ADSR em uma sequência arbitrária.
+        \item[Equação~\ref{eq:intervalos}] Intervalos musicais em número de semitons.
+        \item[Equação~\ref{escSim}] Escalas simétricas na oitava divida em 12 notas.
+        \item[Equação~\ref{eq:escalas}] Escalas diatônicas.
+        \item[Equação~\ref{eq:relacaoDia}] Sucessão de intervalos em uma escala diatônica.
+        \item[Equação~\ref{eq:escalasMenores}] Escalas menores natural, harmônica e melódica.
+        \item[Equação~\ref{triades}] Tríades maior, menor, diminuta e aumentada.
+        \item[Subseção~\ref{subsec:intervalos}] Relações microtonais através de frações de semitons ou através de quantidades inteiras de divisões arbitrárias da oitava.
+        \item[Subseção~\ref{subsec:harmonia}] Relações básicas de harmonia tonal.
+        \item[Subseção~\ref{subsec:contraponto}] Regras básicas de condução de vozes com independência.
+        \item[Subseção~\ref{subsec:ritmo}] Relações básicas de métrica e rítmica.
+        \item[Subseção~\ref{subsec:dir}] Formação de arcos musicais através de estruturas direcionais.
+        \item[Subseção~\ref{estCic}] Estruturas cíclicas para a síntese musical.
+        \item[Equação~\ref{eq:groups}] Propriedades básicas de um grupo algébrico.
+
 	\end{listaespecial}
@@
 \afterpage{\blankpage}
-
-\pretextualchapter{Lista de peças musicais}
-	\begin{listaespecial}[BIGNAMEWIDTH]
-		\item[Chorus infantil]: relações harmônicas da seção~\ref{}. Música disponibilizada online~\cite{}. Arquivo Python no apêndice~\ref{}.
-	\end{listaespecial}
-
-\afterpage{\blankpage}
+%
+%\afterpage{\blankpage}
+%
+%\pretextualchapter{Lista de peças musicais}
+%	\begin{listaespecial}[BIGNAMEWIDTH]
+%		\item[Chorus infantil]: relações harmônicas da seção~\ref{}. Música disponibilizada online~\cite{}. Arquivo Python no apêndice~\ref{}.
+%	\end{listaespecial} 
+%
+%\afterpage{\blankpage}
 %\pretextualchapter{Lista de Abreviaturas}

### 2013-01-18 devEresultados.tex
terminada lista de relacoes

@@
 \end{equation}
-\begin{equation}
+\begin{equation}\label{eq:potVol}
 se \quad pot(T^{'}_i)=2 pot(T_i) \quad \Rightarrow \quad V^{'}_{dB}=10log_{10} 2 \quad \approx \quad 3 dB
@@
 
-\begin{gather}
-10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)} = 10 \quad \Rightarrow \quad \sum_{i=0}^{\lfloor \Delta.f_a \rfloor -1}t^{'2}_i=10\sum_{i=0}^{\Lambda-1}t_i^2=\sum_{i=0}^{\Lambda-1}(\sqrt{10}.t_i)^2 \\
-\therefore \quad t^{'}_i=\sqrt{10}t_i \quad \Rightarrow \quad t^{'}_i \approx 3,16t_i
-\end{gather}
+\begin{align}
+10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)} = 10 \quad \Rightarrow \quad \sum_{i=0}^{\lfloor \Delta.f_a \rfloor -1}t^{'2}_i=10\sum_{i=0}^{\Lambda-1}t_i^2=\sum_{i=0}^{\Lambda-1}(\sqrt{10}.t_i)^2 \label{eq:dobraAmp}\\
+\therefore \quad t^{'}_i=\sqrt{10}t_i \quad \Rightarrow \quad t^{'}_i \approx 3,16t_i\label{eq:dobraVol}
+\end{align}
 
@@
 
-\begin{equation}
+\begin{equation}\label{eq:reconsCompleta}
 t_i = \frac{a_0}{\Lambda} + \frac{2}{\Lambda}\sum_{k=1}^{\tau}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]+ \frac{ a_{\Lambda/2}}{\Lambda}.(1-\Lambda\% 2)
@@
 
-\begin{equation}
-T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{\Lambda-1}= \left \{t_{i \; \% \left( \frac{f_a}{f} \right) } \right \}_0^{\Lambda-1}
+\begin{equation}\label{eq:notaBasica}
+T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{\Lambda-1}= \left \{t^f_{i \; \% \left( \frac{f_a}{f} \right) } \right \}_0^{\Lambda-1}
 \end{equation}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:notaBasicaTimbre}
 T_i^{f,\; \Delta}=\left\{t_i^f\right\}_0^{\lfloor f_a . \Delta \rfloor -1}=\left \{ l^f_{i\,\%\left(\frac{f_a}{f}\right)} \right \}_0^{\Lambda-1}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:distOuvidos}
 \begin{split}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:dti}
 DTI=\frac{d'-d}{v_{som\;no\;ar}\approx 343.2 }\quad \text{segundos}
@@
 e na Diferença de Intensidade Interaural:
-\begin{equation}
+\begin{equation}\label{eq:dii}
 DII=20\log_{10}\left (\frac{d}{d'}\right) \quad dBs
@@
 
-\begin{equation}
+\begin{equation}\label{eq:locImpl}
 \begin{split}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:angulo}
 \theta=\tan^{-1}\left ( \frac{y}{ x }  \right )
@@
 
-\begin{equation}
+\begin{equation}\label{eq:lut}
 T_i^{f,\,\Delta}=\left\{t_i^f\right\}_0^{\lfloor \, f_a .
@@
 
-\begin{equation}
+\begin{equation}\label{freqExponencial}
 F_i=\{f_i\}_0^{\Lambda-1}=\left\{f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}} \right\}_0^{\Lambda-1}
@@
 
-\begin{equation}
+\begin{equation}\label{seqAmpLin}
 a_i=a_0 + (a_{\Lambda-1}-a_0)\frac{i}{\Lambda-1}
@@
 
-\begin{equation}
+\begin{equation}\label{seqAmpDB}
 T_i^{'}=\left\{ t_i 10^{\frac{V_{dB}}{20}\left( \frac{i}{\Lambda-1} \right)^\alpha} \right\}_0^{\Lambda-1}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:varAux}
 \begin{split}
@@
 onde 
-\begin{equation}
+\begin{equation}\label{eq:Bessel}
 J_k(\mu) = \frac{2}{\pi} \int_0^{\frac{\pi}{2}}\left [ cos \left (\overline{k}\;\frac{\pi}{2} + \mu .
@@
 
-\begin{equation}\label{eq:am}
+\begin{equation}\label{eq:amEsp}
 \begin{split}
@@
 
-\begin{equation}\label{fmGamma}
+\begin{equation}\label{fmGammaAux}
 \gamma_i'=\left \lfloor i f' \frac{\widetilde{\Lambda}_M}{f_a} \right \rfloor
@@
 
-\begin{equation}\label{trT}
+\begin{equation}\label{amT}
 T_i^{f,\,AM(f',\,\alpha)}=\left\{ t_i^{f,\,AM(f',\,\alpha)} \right\}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}=\{t_i . (1 + \alpha . t_i')    \}_0^{\Lambda-1}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:vinculos}
 \begin{split}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:adsrApl}
 \left\{t_i^{ADSR}\right\}_0^{\Lambda-1} =\{t_i . a_i\}_0^{\Lambda-1}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:intervalos}
 \begin{split}
@@
 que podemos escrever:
-\begin{equation}
+\begin{equation}\label{eq:relacaoDia}
 \begin{split}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:escalasMenores}
 \begin{split}
@@
 
-\subsection{Estruturas direcionais}
+\subsection{Estruturas direcionais}\label{subsec:dir}
 

### 2013-01-18 conclusoes.tex
feitas alteracoes propostas pelo prof. Luciano

@@
 No capítulo anterior está um sistema conciso
-que relaciona elementos musicais ao som digital. Scripts 
+que relaciona elementos musicais ao som digital. \emph{Scripts} 
 implementam estas relações, e em conjunto foram nomeados \massa\ (Música
@@
 
-As possibilidades abertas por estes resultados envolvem a criação de interfaces de geração de ruídos e outros sons em alta fidelidade (hi-fi), experimentos psicoacústicos e a utilização destes resultados para fins artísticos e didáticos.
+As possibilidades abertas por estes resultados envolvem a criação de interfaces de geração de ruídos e outros sons em alta fidelidade (\emph{hi-fi}), experimentos psicoacústicos e a utilização destes resultados para fins artísticos e didáticos.
 A incorporação de conhecimentos

### 2013-01-18 devEresultados.tex
feitas alteracoes propostas pelo prof. Luciano

@@
         \includegraphics[width=\textwidth]{figuras/formasDeOnda6}
-    \caption{Formas de onda musicais básicas}
+    \caption{Formas de onda musicais básicas. As formas de onda sintéticas estão em (a) e as formas de onda reais estão em (b).}
         \label{fig:formasDeOnda}
@@
         \includegraphics[width=\textwidth]{figuras/espectroDeOndas6}
-    \caption{Espectros das ondas sonoras musicais artificiais básicas}
+    \caption{Espectros das ondas sonoras musicais artificiais básicas.
+As senoide tem o especro puntual, a triangular apresenta somente os harmônicos ímpares, caindo a 6dB por oitava; a onda quadrada tem somente os harmônicos ímpares, caindo a 12dB por oitava; a onda dente de serra apresenta todos os harmônicos, caindo a 6dB por oitava.}
         \label{fig:espectroDeOndas}
@@
         \includegraphics[width=\textwidth]{figuras/espectroOboeAmostradoNatural3}
-    \caption{Espectros das ondas sonoras do oboé natural e de período amostrado}
+    \caption{Espectros das ondas sonoras de uma nota de oboé natural e de período amostrado.
+O som natural possui flutuações nos harmonicos e ruídos, já o som de período amostrado possui espectro perfeitamente harmônico.}
         \label{fig:espectroOboe}
@@
         \includegraphics[width=\textwidth]{figuras/amostras2c__}
-    \caption{Oscilação de 2 amostras (frequência máxima em qualquer $f_a$)}
+    \caption{Oscilação de 2 amostras (frequência máxima em qualquer $f_a$).
+O primeiro coeficiente reflete o deslocamento (\emph{offset} ou \emph{bias}) e o segundo coeficiente especifica a amplitude da oscilação.}
         \label{fig:amostras2}
@@
         \includegraphics[width=\textwidth]{figuras/amostras3b}
-    \caption{3 amostras fixas apresentam uma só frequência não nula. $c_1=c_2^*$ e $w_1 \equiv w_2$}
+    \caption{3 amostras fixas apresentam uma só frequência não nula. $c_1=c_2^*$ e $w_1 \equiv w_2$.}
         \label{fig:amostras3}
@@
         \includegraphics[width=\textwidth]{figuras/amostras4__}
-    \caption{Componentes frequenciais em 4 amostras}
+    \caption{Componentes frequenciais em 4 amostras.}
         \label{fig:amostras4}
@@
         \includegraphics[width=\textwidth]{figuras/amostras4formas__}
-    \caption{Formas de onda básicas em 4 amostras}
+    \caption{Formas de onda básicas em 4 amostras.}
         \label{fig:formas4}
@@
         \includegraphics[width=\textwidth]{figuras/amostras6}
-    \caption{Componentes frequenciais em 6 amostras: 3 senoides se somam ao \emph{bias}}
+    \caption{Componentes frequenciais em 6 amostras: 3 senoides se somam ao \emph{bias}.}
         \label{fig:amostras6}
@@
         \includegraphics[width=\textwidth]{figuras/mixagem}}
-    \caption{Mixagem de três sequências sonoras. }
+    \caption{Mixagem de três sequências sonoras. As amplitudes são sobrepostas diretamente.}
 
@@
         \includegraphics[width=\textwidth]{figuras/concatenacao}}
-    \caption{Concatenação de três sequências sonoras}
+    \caption{Concatenação de três sequências sonoras através da justaposição temporal de suas amostras.}
         \label{fig:concatenacao}
@@
         \includegraphics[width=\textwidth]{figuras/lut}
-    \caption{Lookup Table}
+    \caption{Procedimento de busca em tabela (conhecido como \emph{Lookup Table}) para síntese de sons em frequências diferentes a partir de uma única forma de onda em alta resolução.}
         \label{fig:lut}
@@
         \includegraphics[width=\textwidth]{figuras/transicao}
-    \caption{Transições de intensidade para diferentes valores de $\alpha$ (veja equações~\ref{seqAmp} e ~\ref{transAmp})}
+    \caption{Transições de intensidade para diferentes valores de $\alpha$ (veja equações~\ref{seqAmp} e ~\ref{transAmp}).}
         \label{fig:transicao}
@@
         \includegraphics[width=\textwidth]{figuras/convolucao______}
-    \caption{Interpretação gráfica da convolução}
+    \caption{Interpretação gráfica da convolução. Cada amostra resultante é a soma das amostras anteriores de um sinal uma a uma multiplicadas pelas amostras retrógradas do outro sinal.}
         \label{fig:conv}
@@
         \includegraphics[width=\textwidth]{figuras/delays__}
-    \caption{Convolução com o impulso: deslocamento, linhas de delays e síntese granular}
+    \caption{Convolução com o impulso: deslocamento (a), linhas de delays (b) e síntese granular~(c). Dispostos em ordem crescente de densidade de pulsos.}
         \label{fig:delays}
@@
         \includegraphics[width=\textwidth]{figuras/iir___}
-    \caption{Filtragens realizadas pelos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda} para diferentes frequências de corte, frequências centrais e larguras de banda.}
+    \caption{Módulos da resposta em frequência (a), (b), (c) e (d) respectivamente dos filtros IIR das equações \ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} e \ref{eq:rejeita-banda} para diferentes frequências de corte, frequências centrais e larguras de banda.}
         \label{fig:iir}
@@
         \includegraphics[width=\textwidth]{figuras/ruidos___}
-    \caption{Ruídos coloridos realizados através das equações~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul}, \ref{eq:violeta}: espectros e ondas sonoras resultantes}
+    \caption{Ruídos coloridos realizados através das equações~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul}, \ref{eq:violeta}: espectros e ondas sonoras resultantes.}
         \label{fig:ruidos}
@@
 
-\subsection{Usos musicais - Parte 1: tremolo e vibrato, AM e FM}\label{subsec:tvaf}
+\subsection{Tremolo e vibrato, AM e FM}\label{subsec:tvaf}
 
@@
         \includegraphics[width=\textwidth]{figuras/vibrato___}
-    \caption{Espectrograma de um som com vibrato senoidal de $3Hz$ e profundidade de uma oitava em uma dente de serra de $1000Hz$ (considerada $f_a=44.1kHz$)}
+    \caption{Espectrograma de um som com vibrato senoidal de $3Hz$ e profundidade de uma oitava em uma dente de serra de $1000Hz$ (considerada $f_a=44.1kHz$).}
         \label{fig:vibrato}
@@
         \includegraphics[width=\textwidth]{figuras/tremolo}
-    \caption{Tremolo de profundidade $V_{dB}=12dB$ com padrão oscilatório de uma dente de serra em $f'=1.5Hz$ em uma senoide de $f=40Hz$ (considerada taxa de amostragem $f_a=44,1kHz$)}
+    \caption{Tremolo de profundidade $V_{dB}=12dB$ com padrão oscilatório de uma dente de serra em $f'=1.5Hz$ em uma senoide de $f=40Hz$ (considerada taxa de amostragem $f_a=44,1kHz$).}
         \label{fig:tremolo}
@@
         \includegraphics[width=\textwidth]{figuras/adsr}
-    \caption{Envoltória ADSR e uma sequência sonora arbitrária sendo submetida à envoltória. A variação linear de amplitude está acima. Abaixo a variação de amplitude é exponencial.}
+    \caption{Envoltória ADSR (\emph{Attack, Decay, Sustain, Release}) e uma sequência sonora arbitrária submetida à envoltória.
+A variação linear de amplitude está acima.
+Abaixo a variação de amplitude é exponencial.}
 \end{figure}
@@
 \clearpage
-\section{Organização de notas musicais em música}\label{notasMusica}
+\section{Organização de notas em música}\label{notasMusica}
 Seja $ S_j=\{T_{i,1},T_{i,2}...T_{i,N}\} =\bigcup_{j \in [1,N]} T_{i,j} $ uma sequência de $N$ eventos
@@
 \end{tabular}
-\caption{Resumo das funções harmônicas tonais para escala maior}
+\caption{Resumo das funções harmônicas tonais para a escala maior.
+A tônica é o centro da música, a dominante tende à tônica e a subdominante se distancia da tônica.
+Os três acordes podem, a princípio, serem substituídos livremente pela relativa ou antirelativa.}
 \label{tab:harmonia}
@@
         \includegraphics[width=\textwidth]{figuras/movContraponto}
-    \caption{Movimentos diferenciados pelo contraponto}
+    \caption{Movimentos diferenciados pelo contraponto. Visando preservar a independência entre as vozes, são distinguidos 3 tipos de movimentos.}
         \label{fig:movContraponto}
@@
         \includegraphics[width=\textwidth]{figuras/climax}
-    \caption{Distinções canônicas do clímax musical em uma melodia e outros domínios}
+    \caption{Distinções canônicas do clímax musical em uma melodia e outros domínios.}
         \label{fig:climax}
@@
 o que seriam diferentes
-'idiomas musicais'.\cite{Lerdahl, Harmonia, Schachter,Alfaix}
+`idiomas musicais'.\cite{Lerdahl, Harmonia, Schachter,Alfaix}
 De forma simplificada, um idioma musical é fruto da

### 2013-01-18 musicaExtra.tex
feitas alteracoes propostas pelo prof. Luciano

@@
 abaixo em termos do que é
-mantidas \emph{online}.
+mantido \emph{online}.
 Estas atividades extrapolam os usos musicais, especialmente devido às
@@
 Assim, por completude, alguns materiais didáticos
-e ferramentas web são descritas.
+e ferramentas \emph{web} são descritas.
 

### 2013-01-18 resumoeabstract.tex
feito apendice E ateh musica em tmepo real
feitas alteracoes propostas pelo prof. Luciano

@@
 \vspace{-10mm}
-FABBRI, R. \textbf{Music on digital audio: psychophysical approach and toolbox}. 2012. 151p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, 2012.
+FABBRI, R. \textbf{Music on digital audio: psychophysical description and toolbox}. 2012. 151p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, 2012.
 \vspace{15mm}
@@
 The efficacy of \massa\ was confirmed by the synthesis of small musical pieces using basic notes, incremented notes and notes in music.
-Also possible is to synthesize whole albums through collage of the scripts and parameterization specified by the user.
+It is possible to synthesize whole albums through collage of the scripts and parameterization specified by the user.
 With the paradigm of open source implementation, \massa toolbox can be expanded in co-authorship processes and used freely by musicians, engineers and other interested parties.

### 2013-01-18 scripts/adsr.py

@@ -2,6 +2,8 @@
 import numpy as n, pylab as p
 
 ax=p.subplot(111)
+p.xlabel(r"tempo $\rightarrow$",fontsize=19)
+p.ylabel(r"amplitude $\rightarrow$", fontsize=19)
 
 De=2*100. #duracao total (\Delta)
 DA=2*20. # duracao do ataque \Delta_A
@@ -128,7 +130,6 @@ p.ylim(-3.8,1.8)
 
 
 p.yticks((-1-2.5,0-2.5,SS-2.5,1-2.5),(-1,0,r"$a_S$",1),fontsize=26)
-
 
 
 

### 2013-01-18 scripts/filters/iir.py

@@ -91,7 +91,7 @@ ii=range(1,11)
 p.ylabel(u"amplitude"+r"$\; \rightarrow$")
 p.xlabel(u"frequência "+r"$  \; \rightarrow$")
 p.xticks((0,len(m)/8,len(m)/4,3*len(m)/8,len(m)/2),(r"0",r"$\frac{f_a}{8}$", r"$\frac{f_a}{4}$", r"$\frac{3 . f_a}{8}$", r"$\frac{f_a}{2}$"),fontsize='16')
-p.title("Passa baixas ordem simples")
+p.title("(a) Passa baixas ordem simples")
 #p.xlim(xvals[0]-1,n.log2(fa/2))
 #p.ylim(-123,3)
 
@@ -108,7 +108,7 @@ for fc,pos in zip(fcs,poss):
 p.ylabel(u"amplifitude "+r"$\; \rightarrow$")
 p.xlabel(u"frequência "+r"$  \; \rightarrow$")
 p.xticks((0,len(m)/8,len(m)/4,3*len(m)/8,len(m)/2),(r"0",r"$\frac{f_a}{8}$", r"$\frac{f_a}{4}$", r"$\frac{3 . f_a}{8}$", r"$\frac{f_a}{2}$"),fontsize='16')
-p.title("Passa altas ordem simples")
+p.title("(b) Passa altas ordem simples")
 
 
 ##########
@@ -133,7 +133,7 @@ p.xticks((0,int(0.05*len(m)),int(0.25*len(m)),int(0.45*len(m))),(r"0",r"$f_c=\fr
 p.yticks((0,0.5,1,1.5),(0,0.5,1,1.5))
 p.xlim(0,len(m)/2)
 p.ylim(0,2.5)
-p.title("Rejeita banda de polo duplo")
+p.title("(c) Rejeita banda de polo duplo")
 p.legend(loc="upper left", labelspacing=0,prop={'size':16})
 
 p.ylabel(u"amplifitude "+r"$\; \rightarrow$")
@@ -163,7 +163,7 @@ p.xticks((0,int(0.05*len(m)),int(0.25*len(m)),int(0.45*len(m))),(r"0",r"$f_c=\fr
 p.yticks((0,0.5,1,1.5,2),(0,0.5,1,1.5,2))
 p.xlim(0,len(m)/2)
 #p.ylim(0,2.5)
-p.title("Passa banda de polo duplo")
+p.title("(d) Passa banda de polo duplo")
[…]

### 2013-01-18 devEresultados.tex
dada a primeira mao na bibliografia

@@
 A figura~\ref{fig:spac} mostra esta sombra acústica do crânio, importante para a percepção do ângulo azimutal da fonte no cone de confusão.
-O cone em si não foi disposto na figura pois não é exatamente um cone e suas dimensões precisas não são especificadas na literatura, mas pode ser entendido como um cone com o ápice no meio da cabeca e saindo por cada uma das orelhas.\cite{confCone}
+O cone em si não foi disposto na figura pois não é exatamente um cone e suas dimensões precisas não são especificadas na literatura, mas pode ser entendido como um cone com o ápice no meio da cabeca e saindo por cada uma das orelhas.\cite{hrtf}
 
@@
 Uma forma mais simples é variar a amplitude através da multiplicação da sequência toda
-por um número real. A variação de energia sem variação de
+por um número real. O aumento de energia sem variação de
 amplitude é a \emph{compressão sonora}, útil na
-produção musical atual.\cite{compressao}}. Basta multiplicar a sequência original
+produção musical atual.\cite{guillaume}}. Basta multiplicar a sequência original
 (seja ela gerada ou pré-estabelecida) pela sequência $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$
@@
 Quanto maior o valor de $\alpha$, mais suave é a introdução do som e mais intenso o final da transição.
-$\alpha>1$ resulta em transições de volume muitas vezes chamadas de \emph{slow fade} enquanto $\alpha<1$ resulta em \emph{fast fade}.\cite{fades}
+$\alpha>1$ resulta em transições de volume muitas vezes chamadas de \emph{slow fade} enquanto $\alpha<1$ resulta em \emph{fast fade}.\cite{guillaume}
 
@@
 $a_0$ e $b_1$ 
-são dados através da variável intermediária: $x \in [e^{-\pi},1]$:
+são dados através da variável intermediária $x \in [e^{-\pi},1]$:
 
@@
 Embora esta origem seja um tanto díspar do que pode-se considerar motivo para uma associação com a cor marrom, o ruído sonoro ficou consagrado com este nome.
-De qualquer forma, é bastante comum declarar satisfatória a associação do ruído com a cor marrom, uma vez que os ruídos branco e rosa são mais estridentes e relacionados a cores mais intensas~\cite{marrom}.
+De qualquer forma, é bastante comum declarar satisfatória a associação do ruído com a cor marrom, uma vez que os ruídos branco e rosa são mais estridentes e relacionados a cores mais intensas~\cite{Cook,guillaume}.
 
@@
 ambos tremolos e vibratos. Estas rugosidades são muito apreciadas
-tanto na tradição erudita quanto na música eletrônica atual, especialmente no \emph{Dubstep}.\cite{DubRugoso}
-A rugosidade também é gerada através de conteúdos espectrais que geram batimentos.\cite{Porres}  A sequência \emph{Bela Rugosi}
+tanto na tradição erudita quanto na música eletrônica atual, especialmente no \emph{Dubstep}.
+A rugosidade também é gerada através de conteúdos espectrais que geram batimentos.\cite{Porres,porres2009} A sequência \emph{Bela Rugosi}
 explora este limiar com concomitâncias de tremolos e vibratos na mesma
@@
 
-\subsection{Usos musicais - Parte 2}\label{subsec:mus2}
+\subsection{Usos musicais}\label{subsec:mus2}
 A este ponto as possibilidades musicais explodiram. As
@@
 Esta relação entre acordes é considerada de tonalismo avançado, por vezes até de dissolução do tonalismo, e tem efeitos fortes e marcantes embora perfeitamente consonantes.
-As medianas foram utilizadas a partir do final do romantismo por Wagner, Lizt, Richard Strauss dentre outros e é bastante simples de ser realizada.\cite{Harmonia,Schachter}
+As medianas foram utilizadas a partir do final do romantismo por Wagner, Lizt, Richard Strauss dentre outros e é bastante simples de ser realizada.\cite{Harmonia,Salzer}
 
 A modulação é a troca de tonalidade e se dão principalmente através de: uma dominante individual que afirma a mudança da tônica; alterações cromáticas para um acorde da nova tonalidade; e 'enarmonia' que é a mudança de função de um acorde com as mesmas notas.
-A importância da dominante torna ela o pivô natural das modulações, resultando no círculo das quintas.\cite{Harmonia,Schachter,Koellheuteur,Harmony}
+A importância da dominante torna ela o pivô natural das modulações, resultando no círculo das quintas.\cite{Harmonia,Salzer,Koellheuteur,Harmony}
 
@@
 \subsection{Ritmo}\label{subsec:ritmo}
-A noção rítmica é depende de eventos separados por durações.\cite{rythm} Estes eventos podem ser ouvidos individualmente se separados
+A noção rítmica é depende de eventos separados por durações.\cite{Lacerda} Estes eventos podem ser ouvidos individualmente se separados
 por ao menos $50-63ms$. Para que a separação entre eles possa ser apreciada como duração, ela deve ser maior,
@@
 Assim, é possível para uma mesma sequência de eventos possuir um número de arcos distintos, com tamanhos e clímax diferentes. Este é um recurso
-interessante e útil e a correlação dos arcos resulta na coerência da escuta.\cite{Schachter}
+interessante e útil e a correlação dos arcos resulta na coerência da escuta.\cite{Salzer}
 
@@
 
-Qualquer conjunto de permutações pode ser utilizado como gerador de grupos algébricos.\cite{groups} As propriedades que definem um grupo $G$ são:
+Qualquer conjunto de permutações pode ser utilizado como gerador de grupos algébricos.\cite{permMusic} As propriedades que definem um grupo $G$ são:
 
@@
 
-Esta abordagem resultou em alguns trabalhos e implementações computacionais.\cite{repoFIGGUS,repoFIGGUSold,figgusOriginal,figgusEspacializacao} Descrevemos um uso coerente desta técnica abaixo e apresentamos no Apêndice~\ref{cap:FIGGUScode} a implementação computacional disponibilizada em~\cite{repoFIGGUS}.
+Esta abordagem resultou em alguns trabalhos e implementações computacionais.\cite{figgusOriginal,figgusEspacializacao} Abaixo está um uso coerente desta técnica abaixo e no Apêndice~\ref{cap:FIGGUScode} apontada a implementação computacional disponibilizada em.\cite{MASSA}
 
@@
 o que seriam diferentes
-`idiomas musicais'.\cite{Lerdahl, Harmonia, Schachter,Alfaix}
+`idiomas musicais'.\cite{Lerdahl, Harmonia, Salzer,Alfaix}
 De forma simplificada, um idioma musical é fruto da

### 2013-01-18 introducao.tex
dada a primeira mao na bibliografia

@@
 som e áudio são muitas vezes usados de forma intercambiável.\cite{Everest}} e este pode provir da captura do som por microfones ou da síntese.
-Muitas vezes, o áudio digital é obtido através de protocolos que facilitam o armazenamento e transferência dos arquivos.\cite{procFala} A representação digital do som pode consistir em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
+Muitas vezes, o áudio digital é obtido através de protocolos que facilitam o armazenamento e transferência dos arquivos.
+A representação digital do som pode consistir em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
 Estas amostras separadas por intervalos regulares $\lambda_a$ constituem a forma padrão de representação do som em tempo discreto, chamada de modulação por código de pulsos (PCM do inglês \emph{Pulse Code Modulation}).
-Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$, também chamada de taxa de amostragem, e a profundidade de bit que é o número de bits utilizados para representar a amplitude de cada amostra.\cite{protocolosAudio} A figura~\ref{fig:PCM} exibe $25$ amostras de um áudio PCM com $4$ bits cada.
+Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$, também chamada de taxa de amostragem, e a profundidade de bit que é o número de bits utilizados para representar a amplitude de cada amostra.
+A figura~\ref{fig:PCM} exibe $25$ amostras de um áudio PCM com $4$ bits cada.
 Os $2^4=16$ grados para a amplitude de cada amostra junto ao espaçamento regular $\lambda_a$ introduzem um erro de quantização.
-O ruído causado por estes erros diminuem com a diminuição destes espaçamentos.
+O ruído causado por estes erros diminuem com a diminuição destes espaçamentos.\cite{audioMedia}
 
@@
 Assim, para apreender as frequências audíveis, é necessária uma taxa de amostragem que seja ao menos o dobro da frequência mais aguda $f_a \geq 2\times 20kHz=40kHz$.
-Este raciocínio está na base da utilização das frequências de amostragem $f_a=44.1kHz$ e $f_a=48kHz$, ambas padrão em \emph{Compact Disks} (CDs) e em sistemas de Rádio e TV, respectivamente.\cite{protocolosAudio}
+Este raciocínio está na base da utilização das frequências de amostragem $f_a=44.1kHz$ e $f_a=48kHz$, ambas padrão em \emph{Compact Disks} (CDs) e em sistemas de Rádio e TV, respectivamente.\cite{audioMedia}
 

### 2013-01-19 scripts/secao2.2/eqs2.2.py
atualizados codigos das 3 secoes com base no pep8

@@ -1,436 +1,428 @@
 #-*- coding: utf8 -*-
-import numpy as n, scikits.audiolab as a
+import numpy as n
+import scikits.audiolab as a
 
-f_a = 44100 # Hz, frequência de amostragem
+f_a = 44100  # Hz, frequência de amostragem
 
 ############## 2.2.1 Tabela de busca (LUT)
 # tamanho da tabela: use par para não conflitar abaixo
 # e ao menos 1024
-Lambda_tilde=Lt=1024
+Lambda_tilde = Lt = 1024
 
 # Senoide
-foo=n.linspace(0,2*n.pi,Lt,endpoint=False)
-S_i=n.sin(foo) # um período da senoide com T amostras
+foo = n.linspace(0, 2*n.pi, Lt, endpoint=False)
+S_i = n.sin(foo)  # um período da senoide com T amostras
 
 # Quadrada:
-Q_i=n.hstack(  ( n.ones(Lt/2)*-1 , n.ones(Lt/2) )  )
+Q_i = n.hstack((n.ones(Lt/2)*-1, n.ones(Lt/2)))
 
 # Triangular:
-foo=n.linspace(-1,1,Lt/2,endpoint=False)
-Tr_i=n.hstack(  ( foo , foo*-1 )   )
+foo = n.linspace(-1, 1, Lt/2, endpoint=False)
+Tr_i = n.hstack((foo, foo*-1))
 
 # Dente de Serra:
-D_i=n.linspace(-1,1,Lt)
+D_i = n.linspace(-1, 1, Lt)
 
 # som real, importar período e
 # usar T correto: o número de amostras do período
-Rf_i=a.wavread("22686__acclivity__oboe-a-440_periodo.wav")[0]
+Rf_i = a.wavread("22686__acclivity__oboe-a-440_periodo.wav")[0]
 
-f=110. # Hz
-Delta=3.4 # segundos
-Lambda=int(Delta*f_a)
+f = 110.  # Hz
+Delta = 3.4  # segundos
+Lambda = int(Delta*f_a)
 
 # Amostras:
-ii=n.arange(Lambda)
+ii = n.arange(Lambda)
 
 ### 2.32 LUT
[…]

### 2013-01-19 FIGGUScode.tex
modificacoes pequenas para arrumar lista de figuras e tabelas
feita atualizacao dos apendices de codigos

@@
 \label{cap:FIGGUScode}
-\section{\emph{FiGGUS}}
+\section{FIGGUS}
 
@@
 Arquivo principal, possui todas as rotinas.
-\code{figgus/FIGGUS.py}{FIGGUS/figgus/FIGGUS.py}
+%\code{figgus/FIGGUS.py}{FIGGUS/figgus/FIGGUS.py}
 
@@
 Arquivo auxiliar para tratar as tabelas separadamente.
-\code{figgus/tables.py}{FIGGUS/figgus/tables.py}
+%\code{figgus/tables.py}{FIGGUS/figgus/tables.py}
 
@@
 Inicialização do módulo.
-\code{figgus/\_\_init\_\_.py}{FIGGUS/figgus/__init__.py}
+%\code{figgus/\_\_init\_\_.py}{FIGGUS/figgus/__init__.py}
 
@@
 \subsection{Éter}\label{ap:triangulo3b.py}
-\code{examples/triangulo3\_B.py}{FIGGUS/examples/triangulo3_B.py}
+\code{}{FIGGUS/examples/triangulo3_B.py}
 \clearpage
-\subsection{Butano}\label{ap:triangulo4b}
-\code{examples/try4\_B.py}{FIGGUS/examples/try4_B.py}
+%\subsection{Butano}\label{ap:triangulo4b}
+%\code{examples/try4\_B.py}{FIGGUS/examples/try4_B.py}
+%\clearpage
+
+%\subsection{Thinner}\label{ap:triangulo5b}
+%\code{examples/try5\_B.py}{FIGGUS/examples/try5_B.py}
+%\clearpage
+%
+%\subsection{Tolueno}\label{ap:triangulo6b}
+%\code{examples/try6\_B.py}{FIGGUS/examples/try6_B.py}
+%\clearpage
+
+\subsection{Benzina}\label{ap:try7}
+\code{}{FIGGUS/examples/try7.py}
+\clearpage
+%
+%\subsection{LSA}\label{ap:try3b}
+%\code{examples/try3\_B.py}{FIGGUS/examples/try3_B.py}
+%\clearpage
+
+\subsection{Clorofórmio}\label{ap:try2}
+\code{}{FIGGUS/examples/try2.py}
 \clearpage
 
-\subsection{Thinner}\label{ap:triangulo5b}
-\code{examples/try5\_B.py}{FIGGUS/examples/try5_B.py}
-\clearpage
-
-\subsection{Tolueno}\label{ap:triangulo6b}
-\code{examples/try6\_B.py}{FIGGUS/examples/try6_B.py}
-\clearpage
-
-\subsection{Benzina}\label{ap:try7}
-\code{examples/try7.py}{FIGGUS/examples/try7.py}
-\clearpage
-
-\subsection{LSA}\label{ap:try3b}
-\code{examples/try3\_B.py}{FIGGUS/examples/try3_B.py}
-\clearpage
-
-\subsection{Clorofórmio}\label{ap:try2}
-\code{examples/try2.py}{FIGGUS/examples/try2.py}
-\clearpage
-
-\subsection{Água}\label{ap:try5}
-\code{examples/try5.py}{FIGGUS/examples/try5.py}
-\clearpage
+%\subsection{Água}\label{ap:try5}
+%\code{examples/try5.py}{FIGGUS/examples/try5.py}
+%\clearpage

### 2013-01-19 scripts/pecas2.1/quadrosSonoros.py

@@ -1,132 +1,121 @@
 #-*- coding: utf8 -*-
-import numpy as n, scikits.audiolab as a
+import numpy as n
+import scikits.audiolab as a
 
 # Peça musical baseada em sons estáticos
 # são mixagens de sons básicos apresentados na seção 2.1
 
-f_a=44100 # 44.1kHz, frequência de amostragem de CDs
-Delta=180. # cada quadro terá Delta segundos
-Lambda=int(Delta*f_a) # número de amostras
+f_a = 44100  # 44.1kHz, frequência de amostragem de CDs
+Delta = 180.  # cada quadro terá Delta segundos
+Lambda = int(Delta*f_a)  # número de amostras
 
-ii=n.linspace(0,Delta*2*n.pi,Lambda,endpoint=False)
+ii = n.linspace(0, Delta*2*n.pi, Lambda, endpoint=False)
 
 # frequências que dividem f_a
-fs=[]
-for i in xrange(1,f_a/2+1):
-    if f_a/float(i)==int(f_a/i):
+fs = []
+for i in xrange(1, f_a/2+1):
+    if f_a/float(i) == int(f_a/i):
         fs.append(i)
 
 ### Quadro 1: senoides no grave em batimento e
 ### no agudo uma dente de serra bem suave
-f1=100
-f2=100.5
+f1 = 100
+f2 = 100.5
 
-som=n.sin(ii*f1)+n.sin(ii*f2)
-dente_aguda=(n.arange(Lambda)%4-2)
-som+=dente_aguda/80
+som = n.sin(ii*f1)+n.sin(ii*f2)
+dente_aguda = (n.arange(Lambda) % 4-2)
+som += dente_aguda/80
 
 # normalição no intervalo [-1,1]
-som=((som -som.min())/(som.max()-som.min()))*2-1
+som = ((som - som.min())/(som.max()-som.min()))*2-1
 
-a.wavwrite(som,"quadro1.wav",f_a)
+a.wavwrite(som, "quadro1.wav", f_a)
 print(u"quadro 1 feito em quadro1.wav (mono), são %i amostras \
-em uma frequência de amostragem de %iHz" % (len(som),f_a) )
+em uma frequência de amostragem de %iHz" % (len(som), f_a))
 
 
[…]

### 2013-01-19 codigoPecas.tex
feita atualizacao dos apendices de codigos

@@
 
+\clearpage
 \section{Peças referentes à seção~\ref{sec:notaDisc}}
@@
 \subsection{Quadros sonoros}\label{ap:quadros}
-Montagem musical 'quadros sonoros' para demonstração da mixagem pela soma direta de sequências amostrais (subseção~\ref{subsec:basMus}).
+A montagem musical 'quadros sonoros' é decidada para demonstração da mixagem pela soma direta de sequências amostrais (subseção~\ref{subsec:basMus}).
 Esta rotina sintetiza 5 pequenas peças de sonoridades estáticas.
 Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:notaDisc}.
+Houve relatos em listas de \emph{emails} de que estas sonoridades causaram alteração do estado de consciência.
+Em especial, o quadro 5 causou em diferentes pessoas o mesmo efeito: a sensação de escutar sons da mandimbula.
+
 \code{Quadros sonoros 1-5}{scripts/pecas2.1/quadrosSonoros.py}
-
 \clearpage
@@
 \subsection{Reduced-fi}\label{ap:reduced}
-Pequena peça musical para a concatenação de sequências amostrais como notas musicais (subseção~\ref{subsec:masMus}).
+Pequena peça musical para a concatenação de sequências amostrais como notas musicais (subseção~\ref{subsec:basMus}).
 Sintetiza uma pequena peça de 25 segundos em Python puro, i.e.
-sem utilizar bibliotecas externas como Numpy ou Scikits/Audiolab.
-Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:notaDisc} e disponibilizada online junto ao toolbox \massa.\cite{MASSA}
+sem utilizar bibliotecas externas como \emph{Numpy} ou \emph{Scikits/Audiolab}.
+Peça demonstrativa dos conceitos apresentados na seção~\ref{sec:notaDisc} e disponibilizada online junto ao \emph{toolbox} \massa.\cite{MASSA}
 \code{reduced-fi}{scripts/pecas2.1/reduced-fi.py}
@@
 \subsection{Transita para metro}\label{ap:transita}
-Sintetiza em 1 pequena peça de 49 segundos com varreduras de frequência, chamados chirps.
+Sintetiza em 1 pequena peça de 49 segundos com varreduras de frequência, chamados \emph{chirps}.
 Também é proposta para explorações de transições de intensidade.
@@
 Script disponibilizado online junto ao toolbox \massa.\cite{MASSA}
-\code{Vibra e treme}{scripts/pecas2.2/vibraEtreme.py}
-
-\clearpage
+%\code{Vibra e treme}{scripts/pecas2.2/vibraEtreme.py}
 
@@
 Peça demonstrativa dos conceitos apresentados na subseção~\ref{subsec:mus2} e disponibilizada online junto ao toolbox \massa.\cite{MASSA}
-\code{Tremolos, vibratos e a frequência}{scripts/pecas2.2/TremolosVibratosEaFrequencia.py}
-
-\clearpage
+%\code{Tremolos, vibratos e a frequência}{scripts/pecas2.2/TremolosVibratosEaFrequencia.py}
 
@@
 Peça demonstrativa dos conceitos apresentados na subseção~\ref{subsec:filtros} e disponibilizada online junto ao toolbox \massa.
-\code{trenzinho de caipiras impulsivos}{scripts/pecas2.2/trenzinhoCaipiraImpulsivos.py}
-
-\clearpage
+%\code{trenzinho de caipiras impulsivos}{scripts/pecas2.2/trenzinhoCaipiraImpulsivos.py}
 
@@
 Sintetiza 4 pequenas montagens de 8-32 segundos com os conceitos apresentados na subseção~\ref{subsec:mus2} e disponibilizada online junto ao toolbox \massa.\cite{MASSA}
-\code{chorus infantil}{scripts/pecas2.2/chorusInfantil.py}
-
-\clearpage
+%\code{chorus infantil}{scripts/pecas2.2/chorusInfantil.py}
 
@@
 Sintetiza 7 peças de 5-17 segundos, cetradas nos conceitos apresentados na subseção~\ref{subsec:mus2} e disponibilizada online junto à \massa.\cite{MASSA}
-\code{ADa e SaRa}{scripts/pecas2.2/ADa_e_SaRa.py}
+%\code{ADa e SaRa}{scripts/pecas2.2/ADa_e_SaRa.py}
 
@@
 Pequena peça musical de 45 segundos para exploração do sistema de intervalos, demonstrativa dos conceitos apresentados na subseção~\ref{subsec:intervalos}, e disponibilizada online junto ao toolbox \massa.
-\code{intervalos entre alturas}{scripts/pecas2.3/intervalosEntreAlturas.py}
-
-\clearpage
+%\code{intervalos entre alturas}{scripts/pecas2.3/intervalosEntreAlturas.py}
 
@@
 \clearpage
+
 \subsection{Acorde cedo}\label{ap:acorde}
 Pequena peça musical de 40 segundos para demonstração dos encadeamentos tonais básicos, conceitos apresentados na subseção~\ref{subsec:harmonia}, e disponibilizada online como parte da \massa.\cite{MASSA}
-\code{acorde cedo}{scripts/pecas2.3/acordeCedo.py}
-
-\clearpage
+%\code{acorde cedo}{scripts/pecas2.3/acordeCedo.py}
 
@@
 Pequena montagem musical de 17 segundos para demonstração de vozes conduzidas por regras simples para manterem independência, conceitos apresentados na subseção~\ref{subsec:contraponto}, e disponibilizada online junto à \massa.\cite{MASSA}
-\code{conta ponto}{scripts/pecas2.3/contaPonto.py}
-
-\clearpage
+%\code{conta ponto}{scripts/pecas2.3/contaPonto.py}
 
@@
 Pequena montagem musical de 99 segundos para demonstração da métrica musical e das diferentes contagens, conceitos apresentados na seção~\ref{subsec:ritmo}, e disponibilizada online junto à \massa.\cite{MASSA}
-\code{poli hit mia}{scripts/pecas2.3/poliHitMia.py}
-
-\clearpage
+%\code{poli hit mia}{scripts/pecas2.3/poliHitMia.py}
 

### 2013-01-19 scripts/pecas2.3/dirracional.py

@@ -54,8 +54,6 @@ def A(fa=2.,V_dB=10.,d=2.,taba=S_i):
     A_i=A_i*10.**(V_dB/20.)
     return A_i
 
-
-
 def adsr(som,A=10.,D=20.,S=-20.,R=100.,xi=1e-2):
     a_S=10**(S/20.)
     Lambda=len(som)

### 2013-01-21 agradecimentos.tex
commit geral antes de correcoes do prof. chu
atualizando dissertacao, varias coisinhas

@@
 Agradeço ao corpo de funcionários do IFSC, pela prestatividade e eficiência em todos os momentos que precisei.
+
+\vspace{4 mm}
+
+Agradeço aos Prof. Rafael Santos Mendes e Prof. Adolfo Maia Junior pelas orientações na gênese deste trabalho.
 
@@
 Agradeço aos participantes do labMacambira.sf.net pelas colaborações diretas e indiretas nos software, apresentações e articulações desde junho de 2011.
-Em especial agradeco ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela, Patrícia Ferraz, Glerm Soares, Vanessa Ferreira, Sérgio Teixeira de Carvalho.
+Em especial agradeco ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela, Guilherme Lunhani, Patrícia Ferraz, Glerm Soares, Vanessa Ferreira, Sérgio Teixeira de Carvalho.
 Agradeço aos demais colaboradores das listas de email, IRC, AA e outros canais.

### 2013-01-21 musicaExtra.tex
atualizados codigos das 3 secoes com base no pep8
atualizando dissertacao, varias coisinhas

@@
 Estas atividades extrapolam os usos musicais, especialmente devido às
-finalidades pedagódicas e sociais dos materiais audiovisuais.
+finalidades pedagódicas e sociais das práticas audiovisuais.
 Assim, por completude, alguns materiais didáticos
 e ferramentas \emph{web} são descritas.
+Todos os desenvolvimentos desta dissertação estão em repositórios
+abertos.\cite{repoDissertacao}
 
 \section{Experimentos abertos em áudio: LADSPAs, Wavelets e Redes Complexas}
-
-Todos os desenvolvimentos desta dissertação estão em repositórios
-abertos.\cite{repoDissertacao}
 
@@
   em telecomunicação, dizem respeito em menor ou maior grau à produção
-  de código como cultura. As elaborações de conceitos e ferramentas
+  de código computacional como uma manifestação cultural. As elaborações de conceitos e ferramentas
   voltados para o compartilhamento nestas culturas estão na gênese do que
@@
 LADSPA (Linux Audio Developers Simple Plugin API) é a API
-livre de plugins de áudio até a presente data. A
+livre de \emph{plugins} de áudio até a presente data. A
 última versão é a 1.1, em uso corrente até hoje e é
@@
 usar uma amostra curta do ruído e reproduzi-la indefinidamente
-através de \emph{cross-fades}. Esta eficiência é desejada em um plugin,
-pois seus usos comuns estão aplicações em tempo real.
+através de \emph{cross-fades}. Esta eficiência é desejada em um \emph{plugin},
+pois seus usos comuns incluem aplicações em tempo real.
 
@@
   códigos em C e a implementação como plugin lv2 estão no
-  repositório git como pode ser visto em \emph{online}\footnote{
-      este link é o README do repositódio git com todos os códigos dos
+  repositório git como pode ser visto em \emph{online}\footnote{Este link é o \emph{README} do repositódio \emph{git} com os códigos dos
       plugins:
-  http://labmacambira.git.sourceforge.net/git/gitweb-index.cgi.}.
+  \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/LV2;a=summary}.}.
 
@@
 A remoção de ruído \emph{Hum} é baseada em uma sequência de filtros nó rejeita
-banda (veja subseção~\ref{subsec:filtros}
+banda (veja subseção~\ref{subsec:filtros})
 idealmente dispostos nos harmônicos da frequência de
@@
 distorção do modelo ideal. O código C/C++ e a
-implementação como plugin está no mesmo repositório disponibilizado
+implementação como \emph{plugin} está no mesmo repositório disponibilizado
 para a síntese de ruídos.
@@
 O método está bem descrito no artigo desenvolvido.
-Em resumo, o método consiste particionar o áudio com sobreposições para contemplar aplicações reais sem as distorções de borda da decomposição e recomposição.
+Em resumo, o método consiste em particionar o áudio com sobreposições para contemplar aplicações reais sem as distorções de borda da decomposição e recomposição.
 Em seguida, o áudio é decomposto em árvore \emph{Wavelet Packet}.
-Cada folha é então ordenada, os coeficientes com baixa energia eliminadose um polinômio encontrado pelo método dos mínimos quadrados é usado para representar os coeficientes restantes.
-Detalhes de escrita e leitura do polinômio completam o protocolo.
-Aplicações para stream de voz em tempo real foram vislumbradas e testes adicionais propostos.\cite{FabbriWavelets}
+Cada folha é então ordenada, os coeficientes com baixa energia eliminados e um polinômio encontrado pelo método dos mínimos quadrados é usado para representar os coeficientes restantes.
+Detalhes de escrita e leitura dos polinômios e das permutações completam o protocolo.
+Aplicações para \emph{stream} de voz em tempo real foram vislumbradas e testes adicionais propostos.\cite{FabbriWavelets}
 
 \subsubsection{Redes Complexas}
-Com o fim de detectar nuances emocionais no uso da linguagem natural, foram desenvidos alguns trabalhos em análise de fala e de texto, ambos focados no uso de redes complexas.
-Três trabalhos: dois de processamento de texto e um de fala.
+Com o fim de detectar nuanças emocionais no uso da linguagem natural, foram feitos alguns trabalhos em análise de fala e de texto, ambos focados no uso de redes complexas.
 O primeiro trabalho de processamento de texto foi apresentado na ACL de 2010, em Upsala, na Suécia, pelo Prof.
@@
 Nas falas, as bandas de frequência utilizadas são vértices e a transição entre uma banda e outra é uma aresta.
-Ambas redes com peso e podem ser utilizadas como direcionais ou não.
+Ambas as redes possuem arestas com peso e podem ser utilizadas como direcionais ou não.
 Os trabalhos consistiam em formar estas redes com os bancos de dados correlatos, extrair medidas e aplicar reconhecimento de padrão.
@@
 
-\section{Áudio e Música}
+\section{Áudio e música}
 
 Este capítulo está dividido em 4 partes: música em
-tempo diferido (ou seja, que não é feito em tempo real), música em
+tempo diferido (ou seja, que não é realizada em tempo real), música em
 tempo real, música na matéria (suporte físico em \emph{hardware}) e música no
-tecido social (considerando especialmente as mobilizações humanas
-relacionadas).
-
-Com isso desejamos expor uma incidência da prática musical através do código com
-exemplos reais de aplicação e em utilização pelo autor, membros do
+tecido social (mobilizações humanas).
+O texto se baseia em exemplos reais da prática musical através do código, com aplicações e em utilização pelo autor, membros do
 LabMacambira.sf.net, parceiros, colaboradores e por usuários
@@
 
-\subsection{Música em Tempo Diferido: Minimum-fi e FIGGUS}
+\subsection{Música em tempo diferido: \emph{minimum-fi} e FIGGUS}
 
@@
 A realização musical em tempo diferido é o paradigma inicial da música
-computacional, iniciando com o Music V, depois
-com o CSound. Pode-se dizer que até hoje é a forma como
+computacional, iniciada com o \emph{Music V}, depois
+com o \emph{CSound}. Pode-se dizer que até hoje é a forma como
 compositores usualmente pensam a música: concebendo e escrevendo as
@@
 eletrônicos.
-
 Assim como a composição instrumental permite alguns refinamentos estruturais não presentes na improvisação instrumental,
@@
 tempo diferido permite um detalhamento maior dos
-procedimentos do que a realização em tempo real. Por este mesmo
-motivo.
+procedimentos do que a realização em tempo real.
 
@@
 digital. Estes sons e a organização deles são objeto do Capítulo 2 da presente dissertação.
-A utilização destes sons diretamente, utilizando somente os recursos básicos de justaposição e sobreposição, está na pequena peça \emph{minimum-fi}. Já o \figgus\ (FInite Groups in Granular and Unit
-Synthesis), utiliza os princípios do minimum-fi e utiliza simetrias, através de
+A utilização destes sons de forma direta, utilizando somente os recursos básicos de justaposição e sobreposição, está na pequena peça \emph{minimum-fi}.
+Já o \figgus\ (FInite Groups in Granular and Unit
+Synthesis), utiliza os princípios do \emph{minimum-fi} e utiliza simetrias através de
 permutações e de Teoria de Grupos, para a composição de músicas.
@@
 EP\footnote{\emph{Extended Play}, um álbum musical maior que um single mas
-  menor que um LP (Long Play) inteiro.} com um único comando. Este é o
-\emph{PPEPPS}\footnote{Pure Python EP: Projeto Solvente}, como veremos
+menor que um LP (\emph{Long Play}) inteiro.} com um único comando. Este é o
+\emph{PPEPPS}\footnote{Pure Python EP: Projeto Solvente.}, como veremos
 a seguir.
 
-Desta forma, esta sessão exemplifica e explicita - através de dois
+Desta forma, esta seção exemplifica e explicita - através de dois
 exemplos reais - o uso de código para a síntese
@@
 frequencia, amplitude e timbre, até a confecção de uma ferramenta
-derivada, já incorporando propostas musicais e estruturas mais
+derivada, que já incorpora propostas musicais e estruturas mais
 elaboradas.
@@
 
-Os princípios, bastante simples\footnote{Osvaldo Lacerda, em seu livro
+Os princípios, bastante simples, são\footnote{Osvaldo Lacerda, em seu livro
   \emph{Compêndio de Teoria Elementar da Música} fala das propriedades
-  do som musical e sua organização de forma condizente.}, são:
+  do som musical e sua organização de forma condizente.}:
 \begin{itemize}
@@
     geração de unidades sonoras com diferentes timbres, controle sobre
-    a frequência fundamental, duração e volume, como especificado na seção~\ref{sec:notasDisc}.
+    a frequência fundamental, duração e volume, como especificado na seção~\ref{sec:notaDisc}.
   \item Deve-se ser capaz de construir
@@
 
-Para o primeiro item, se prestam os procedimentos de busca
+Para o primeiro item, presta-se o procedimento de busca
 em tabelas/vetores com formas de ondas em alta resolução, chamado
@@
 com resultados diversificados e tidos como de alta qualidade
-pois não acrescentam ruídos relevantes ao sinal. Uma descrição
+pois não acrescenta ruídos relevantes ao sinal. Uma descrição
 do procedimento está na subseção~\ref{subsec:lookup}.
 
-Através da utilização do lookup sucessivo (procura-se um valor na primeira tabela
+Através da utilização do \emph{lookup} sucessivo (procura-se um valor na primeira tabela
 e este valor indica o valor na segunda tabela a ser utilizado), executa-se
@@
 é a segunda busca, na outra tabela.
-
-Existem muitas formas de se executar waveshaping, a seguir segue o que usamos no minimum-fi, que se sustenta principalmente por ser leve e simples:
+Existem muitas formas de se executar um \emph{waveshaping}, a seguir segue o que usamos no \emph{minimum-fi}, que se sustenta principalmente por ser leve e simples:
 
@@
 
-O segundo item - dos dois princípios expostos sobre o minimum-fi - presta-se à discretização do espaço musical.
+O segundo item - dos dois princípios expostos sobre o \emph{minimum-fi} - presta-se à discretização do espaço musical.
 Unidades como batidas e notas
 tornam mais eficiente a comunicação pois a quantidade
-de estruturas sugeridas é maior e as estruturas são mais claras no discreto do que no contínuo.\cite{Roederer} Roederer chega a
-apontar que as próprias notas dos instrumentos musicais são um reflexo de que é mais eficiente
-o uso do discreto do que do contínuo para a geração de estruturas musicais.
+de estruturas sugeridas é maior e as estruturas são mais explícitas no discreto do que no contínuo. Roederer
+aponta que as próprias notas dos instrumentos musicais são um reflexo de que é mais eficiente
+o uso do discreto do que do contínuo para a geração de estruturas musicais.\cite{Roederer}
 
 De fato, unidades bem definidas se mostram úteis na prática musical 
-para fazer sequências de unidades, concatená-las no tempo. Quando as unidades
+para fazer sequências de unidades. Quando as unidades
 são notas com frequência definida, as sequências de unidades justapostas no tempo tendem a ser compreendidas como melodias ou linhas melódicas. As
@@
 
-As duas construções básicas explicitadas, baseadas na dicotomia melodia/harmonia - relacionada à horizontalidade/verticalidade, justaposição/sobreposição - são
-as funções \emph{fazSequencia} e \emph{fazAcorde} no minimum-fi. Vale notar elas são absolutamente 
+As duas construções básicas explicitadas - baseadas na dicotomia melodia/harmonia, horizontalidade/verticalidade, justaposição/sobreposição - são realizadas através
+das funções \emph{fazSequencia} e \emph{fazAcorde} no \emph{minimum-fi}. Vale notar elas são absolutamente 
 equivalentes em uma análise puramente conceitual, i.e.
@@
 
-Como pode-se notar a seguir, as sequências de notas e os acordes, em última instância, são utilizações específicas das 2 funções de síntese sonora explicadas anteriormente: lookup e lookupcruz.
+Como pode-se notar a seguir, as sequências de notas e os acordes, em última instância, são usos específicos das 2 funções de síntese sonora implementadas: \emph{lookup} e \emph{lookupcruz}.
 
@@
 
-O passo seguinte é sintetizar e mixar para a obtenção 
+O passo seguinte é sintetizar, mixar e concatenar para a obtenção 
 de sequências musicais. A síntese de sequências e acordes
-podem ser feitos desta forma:
+pode ser feita desta forma:
 
@@
 listas em Python e vetores Numpy (mais eficientes em tempo de
-execussão e também na simplicidade do código). Estes mesmos
-procedimentos são praticamente os mesmos em Scilab, C/C++, Javascript, PHP, etc.
+execução e também na simplicidade do código). Estes 
+procedimentos são praticamente os mesmos se implementados em Scilab, C/C++, Javascript, PHP, etc.
 
@@
 
-Nesta subseção há um foco especial nas estruturas musicais.
+Nesta subseção há um foco nas estruturas musicais.
 São elas, inseridas em um momento histórico e
@@
 consiste na utilização de estruturas matemáticas
-para a representação de simetrias.
+que representam simetrias para a organização de materiais musicais.
 
@@
 tratar de simetrias na música com vistas à composição musical através
-de métodos matemáticos\footnote{Duas iniciações científicas trataram do assunto.}. Mais especificamente, a proposta resultou em
-um programa voltado para a síntese
-granular e síntese de estruturas musicais através de Grupos Algébricos. O nome dado
+de métodos matemáticos\footnote{Duas iniciações científicas trataram do assunto.}. Mais especificamente, a proposta gerou
+um programa de síntese
+granular e de estruturas musicais através de Grupos Algébricos. O nome dado
 foi \figgus, sigla de FInite Groups in Granular and Unit Synthesis\footnote{Também foi usado
@@
 é usual na literatura. Posteriormente o primeiro autor deste trabalho recorreu novamente
-ao uso do nome \figgus. Isso foi motivado pelo
-uso da técnica para síntese de estruturas musicais, não para
-síntese de amálgamas sonoros (ou timbres mesmo) tipicamente resultantes da síntese granular}.
+ao uso do nome \figgus. Isso foi motivado pela utilidade
+da técnica para síntese de estruturas musicais, que, nos usos que o \figgus\ teve, foi maior do que a utilidade para síntese de amálgamas sonoros tipicamente resultantes da síntese granular}.
 
@@
 Outro uso planejado para o \figgus\
-é a síntese de tímbres e amálgamas sonoros através da
+é a síntese de timbres e amálgamas sonoros através da
 Síntese Granular. Embora o foco atual seja outro, cabe algumas breves
@@
 satisfazendo as propriedades descritas na subseção~\ref{estCic}.
-
 No \figgus, o grão ou unidade sonora
@@
 frequência (Hz), timbre (identificador para usar mediante implementações convenientes), intensidade (pico $\in \ [0,1]$), e duração dos fades (in e out em segundos).
-
-%\code{Grão Sonoro Básico}{python_snippets/fgrao.py}
-
 O \figgus\ funciona com base em uma sequência de grãos especificada inicialmente e na qual operam as permutações.
-
-%\code{Sequência de Grãos}{python_snippets/fsequencia.py}
-
 Aos grãos em sequência são aplicadas permutações.
@@
 
-%\code{Permutações e Padrões de Permutações}{python_snippets/fpermutacoes.py}
-
-De posse dos grãos, da sequência, das permutações e do padrão de permutações, pode-se realizar a estrutura musical em si.
+Em posse dos grãos, da sequência, das permutações e do padrão de permutações, pode-se realizar a estrutura musical em si.
 Basta adicionalmente especificar o número desejado de iterações da sequência.
@@
 
-%\code{Realização do Padrão Musical}{python_snippets/fpadraosonoro.py}
-
-Em posse não apenas de representações abstratas do padrão musical a ser realizado, mas dos próprios vetores sonoros relacionados à representação digital da música a ser realizada, pode-se escrever um arquivo de áudio propriamente dito.
-O mais conveniente neste caso é escrever um arquivo PCM (Pulse Code Modulation) em algum padrão amplamente utilizado e reconhecido.
+A partir das representações abstratas do padrão musical, são feitos os vetores sonoros da representação digital da música a ser realizada e pode-se escrever um arquivo de áudio propriamente dito.
+O mais conveniente, neste caso, é escrever um arquivo PCM (Pulse Code Modulation) em algum padrão amplamente utilizado e reconhecido.
 Ambos WAV e AIFF satisfazem estes requisitos.
@@
 
-\subsection{Música em Tempo Real: Live coding e ABeatTracker (ABT)}
+\vspace{10mm}
+\subsection{Música em tempo real: \emph{Livecoding} e ABeatTracker (ABT)}
 
@@
 A exploração estética destas ferramentas 
-em performances musicais é a proposta do \emph{Live coding}.
+em performances musicais é a proposta do \emph{Livecoding}.
 Esta prática de \emph{performance} se dedica
@@
 livecoding, mas já é um intermediário, com traços
-de um aplicativo ou ferramenta, embora os comandos sejam
+de um aplicativo ou ferramenta, embora os comandos sejam acessados
 pela escrita. É uma linguagem
@@
 
-\subsubsection{Live coding}
+\subsubsection{\emph{Livecoding}}
 
@@
       concentração do \emph{performer}/instrumentista é bastante focada na tela do
-      computador.
+      computador. Recursos performáticos são, portanto, preciosíssimos.
     \item O \emph{feedback} auditivo do código projetado permite que o
@@
 trio \emph{FooBarBaz}/Variáveis Metasintáticas composto
-pelo duo e por Gilson Beck na mesa de som, usando detecção de cor no laptop
+pelo duo e por Gilson Beck, este atuante na mesa de som e usando detecção de cor no \emph{laptop}
 para incrementar a apresentação. Este trio
@@
 
-Foi usada a linguagem emph{ChucK} por apresentar os recursos que
+Foi usada a linguagem \emph{ChucK} por apresentar os recursos que
 o duo inicial considerou mais apropriados,
@@
 no fundo do código sugeria o sono REM como uma experiência coletiva de alteração
-do estado.
+do estado de consciência.
 Estes recursos podem ser vistos
@@
 \url{http://wiki.nosdigitais.teia.org.br/FooBarBaz}. Estavam presentes
-no palco outros membros do labMacambira.sourceforge.net: Ricardo Fabbri,
-Alexandre 'Bzum', Larissa.
+no palco também outros membros do labMacambira.sourceforge.net, davam suporte em gravações e outros auxílios: Ricardo Fabbri,
+Alexandre 'Bzum', Daniel Penalva e Ivan Marin.
 
@@
 e espacialização dos dois canais de áudio foram controladas por um
-patch em Puredata que permitia o \textit{cross-fading} entre os canais
+\emph{patch} em Puredata que permitia o \textit{cross-fading} entre os canais
 através da detecção do movimento das mãos. Isso foi possível através
@@
 A prática de \emph{livecoding} requer o uso de 
-instruções curtas, incentivando o improviso
-musical através do código. Ambos \emph{livecoders}
+instruções curtas, que incentivem o improviso
+musical através do código. Ambos os \emph{livecoders}
 usaram bons editores de texto
@@
 facilmente especificar a posição do arquivo de
-áudio que deve ser iniciada a execução, assim como a velocidade
+áudio em que deve ser iniciada a execução, assim como a velocidade de
 reprodução do áudio, duração e
-amplitude. Uma interface simples foi permitiu a
+amplitude. Uma interface simples permitiu a
 alteração rápida desses parâmetros em tempo de execução.
@@
 
-Esses desenvolvimentos todos inspiraram a criação de uma linguagem
+Estes desenvolvimentos inspiraram a criação de uma linguagem
 específica para \emph{livecoding}, chamada \emph{Vivace}, que está em constante
@@
 gênero da \emph{pop art}: trechos de vídeos de uma novela brasileira: 'Avenida Brasil'
-eram rearanjados em tempo real em sincronia com linhas rítmicas e
+eram rearranjados em tempo real em sincronia com linhas rítmicas e
 melódicas\footnote{Uma
@@
 
-  Catalizadas pela aproximação de "Gera" Magela e Caleb Mascarenhas Luporini, as
+  Catalizadas pela aproximação de "Gera" Magela, Caleb Mascarenhas Luporini e Guilherme Lunhani as
    apresentações e amadurecimeentos de 2012 desembocaram na
-  na escrita coletiva de códigos (acesso e controle da platéia 
-  no código que está sendo escrito), utilização de monstros e ascii art e
-  e sonoridades bizarras. Com isso, está sendo proposto
+  escrita coletiva de códigos (acesso e controle da platéia 
+  no código que está sendo escrito), utilização de monstros e \emph{ascii art} e
+  e sonoridades bizarras. Com isso, está sendo observado e proposto
   um gênero de \emph{livecoding} batizado de '\emph{Freakcoding}', possivelmente o primeiro subgênero
   de \emph{livecoding}.
-
+Foi escrito um \emph{manifesto} sobre o \emph{freakcoding} e um artigo acadêmico.
+Estão sendo planejados mais apresentações e desenvolvimentos tecnológicos, todos motivados pelo \emph{livecoding} e este subgênero brasileiro que está despontando.
+  \vspace{10mm}
 \subsubsection{ABeatTracker (ABT)}
@@
 destas amostras sonoras e variáveis randômicas utilizadas para
-execussão da linha. Além disso, o ABT dispõe de variáveis globais que
-podem ser alteradas a qualquer momento pelo usuário também com macros, como BPM,
+execução da linha. Além disso, o ABT dispõe de variáveis globais que
+podem ser alteradas a qualquer momento pelo usuário também com macros, como BPM, volume,
 velocidade de leitura das amostras e variáveis randômicas globais (que
@@
 linguagens com domínios específicos não são raras e por vezes o ABT
-foi descrito como uma linguagem.
+foi descrito como uma linguagem por usuários e interessados.
 
@@
 explicitados por sequências de compassos que encapsulem durações
-regulares do rítmo tamborilado. Estes ritmos relacionados ao
+regulares do rítmo tamborilado. Estes rítmos relacionados ao
 tamborilar do usuário são chamados de harmônicos e podem ser
@@
 De fato, atualmente o ABD é parte do ABT.
-
 Todo o código do
@@
 
-\subsection{Música na Matéria: EKP e AHT}
+\subsection{Música na matéria: EKP e AHT}
 \label{aht}
 
-Embora o foco deste trabalho seja na exploração musical através de
-códigos computacionais, nesta sessão está explicação simples,
+Embora o foco deste trabalho seja a exploração musical através de
+códigos computacionais, nesta seção está uma explicação simples,
 clara e factual de como estas investidas transcendem o código e até
 mesmo a música em si.
-
 Por vezes, mais relevantes que os próprios desenvolvimentos são as mobilizações
 criadas nos entornos e os engajamentos.
-
 A seguir estão dois trabalhos que geraram alguma mobilização,
@@
 
-Foram definidos três finalidades para esta exploração
-do Sistema Operacional:
+Foram reconhecidas três finalidades para esta exploração
+do sistema operacional:
 
 \begin{itemize}
-\item Pedagógicos, através da utilização de um sentido que não o visual para a tarefa de analisar o Sistema.
-\item Artísticos, para apresentações musicais/audiovisuais.
-\item Monitoramento do SO, pela emissão de sons periódicos relativos à carga de processamento, memória, uso de rede, etc.
+\item Pedagógica, através da utilização de um sentido que não o visual para a tarefa de analisar o sistema.
+\item Artística, para apresentações musicais/audiovisuais.
+\item Monitoramento do sistema operacional, pela emissão de sons periódicos relativos à carga de processamento, memória, uso de rede, etc.
 \end{itemize}
@@
 origamis (dobraduras de papel), construída na forma de uma mesa. Nela,
-uma rede de coolers reciclados faz flutuar origamis de geometria e
+uma rede de \emph{coolers} reciclados faz flutuar origamis de geometria e
 cores variadas. Os movimentos dos origamis são captados por \emph{webcam} e
 interpretados em tempo real por software de processamento de imagens,
-gerando padrões que controlam a transformação sonora da música. Dessa
+gerando padrões que controlam a transformação sonora da música. Desta
 forma, pode-se dizer que os sons gerados refletem o vôo dos origamis
@@
 Na versão atual da AirHacktable, cada cor (vermelho, amarelo, azul,
-verde, preto, ou branco) controla uma voz, já a posição do origami na
-mesa controla aspectos do som. Como um exeplo, pode-se
+verde, preto, ou branco) controla uma voz. Já a posição do origami na
+mesa controla aspectos do som. Como um exemplo, pode-se
 configurar a AHT de forma que, se o origami está
@@
 
-A AHT segue a filosofia de desenvolvimento contínuo, comum ao software
-livre, que atualmente foi incorporado ao movimento chamado hardware
+A AHT segue a filosofia de desenvolvimento contínuo, comum ao \emph{software}
+livre, que atualmente foi incorporado ao movimento chamado \emph{hardware}
 livre ou aberto. A concepção inicial foi de Chico Simões (mestre de maracatu
@@
 Logo neste estágio inicial, foram aproximados Vilson Vieira, Ricardo Fabbri, Fábio Simões
-e Daniel Penalva em reuniões presenciais. Destas reuniões, saíram ideias
+e Daniel Penalva em reuniões presenciais. Destas reuniões saíram ideias
 diversas, escritas em um \emph{pad} aberto\footnote{Veja: \url{http://pontaopad.me/origami-sensores}.},
@@
 com apresentação na Teia Casa de Criação, durante o mesmo festival. A estrutura
-da mesa é também de papel (papelão, e foi feita
+da mesa é também de papel (papelão), e foi feita
 por Francisco Simões por pesquisa própria. O uso do reticulado de ventoinhas foi fruto
 de discussões coletivas e de experimentos do coletivo labMacambira.sourceforge.net,
-com especial empenho do Francisco e uso de sua luthearia de percussão: \emph{Tora Tambores}.
+com especial empenho do Francisco e uso de sua luthieria de percussão: \emph{Tora Tambores}.
 
-Em 2012 houve a aproximação de Caleb Mascarenhas Luporini e "Gera" Magela.
+Em 2012 houve a aproximação de Caleb Mascarenhas Luporini e "Gera" Magela nesta frente de atuação.
 Foram feitas oficinas no SESC Pinheiros e SESC Belenzinho e apresentação
@@
 imediatas das posturas de compartilhamento e apropriação citados.
+Segue uma lista parcial dos grupos mais relevantes para o presente trabalho, tanto pela influência que estes grupos/redes tiveram na formação e atuação do primeiro autor quanto pela ressonância que os trabalhos aqui expostos encontram.
 
@@
     este grupo dedicou-se à realização
-    de apresentações conjuntas, bem como a escrita de peças e textos através
+    de apresentações públicas com peças musicais, bem como à escrita de peças e textos através
     de dinâmicas coletivas.
@@
     \item Estudios Livres: são ambientes de estudo, produção e distribuição de mídias livres.
-    Formam, sem conjunto, um movimento e uma rede social brasileira, tida como única no mundo, embora
+    Formam, em conjunto, um movimento e uma rede social brasileira, tida como única no mundo, embora
     o conceito seja conhecido internacionalmente\footnote{Há uma plataforma aberta com capacidades
@@
 Este é um meio de troca rico e alternativo a alguns meios mais politizados, como a lista do \emph{estudiolivre.org}.
-    \item MSST (Movimento dos Sem Satélite): com um manifesto e propostas coerentes, mas nada óbvias, como a demanda específica por satélites civís, abordagens tecnomágicas, o grupo é uma instâncias iniciáticas em termos tecnológicos e conceituais.
+    \item MSST (Movimento dos Sem Satélite): com um manifesto e propostas coerentes, mas nada óbvias, como a demanda específica por satélites civis, abordagens tecnomágicas, o grupo é uma instância iniciática em termos tecnológicos e conceituais.
     \item LabMacambira.sf.net: espaço virtual de programadores dedicados ao audiovisual e propósitos socialmente engajados.
 Fundado em parceria do autor deste trabalho e de Vilson Vieira, Daniel Marostegan e Carneiro e Ricardo Fabbri e de parte dos investimentos da segunda etapa do convênio de Pontão de Cultura Digital Nós Digitais, da Teia Casa de Criação.
-Em um primeiro momento, membros foram remunerados como mentores e como aprendizes\footnote{Renato Fabbri, Ricardo Fabbri, Vilson Vieira, Marcos Mendonça e Gilson Beck mentoraram; Alexandre "Bzum" Negrão, Lucas Zambianchi, Larissa Arruda, Nivaldo Henrique Bondança, Fernando Gorodscy, Andres Martano desenvolveram como bolsistas; Danilo Shiga e Marcos Murad foram parceiros colaboradores; Francisco Simões, Fábio Simões, Caleb Mascarenhas Luporini, "Gera" Magela, Edson "Presto" Correia, Daniel Penalva se aproximaram para atividades específicas, cruciais em todo ano já sem os investimentos do convênio.}.
+Em um primeiro momento, membros foram remunerados como mentores e como aprendizes\footnote{Renato Fabbri, Ricardo Fabbri, Vilson Vieira, Marcos Mendonça e Gilson Beck mentoraram; Alexandre "Bzum" Negrão, Lucas Zambianchi, Larissa Arruda, Nivaldo Henrique Bondança, Fernando Gorodscy, Andres Martano desenvolveram como bolsistas; Danilo Shiga e Marcos Murad foram parceiros colaboradores; Francisco Simões, Fábio Simões, Caleb Mascarenhas Luporini, "Gera" Magela, Edson "Presto" Correia, Daniel Penalva se aproximaram para atividades específicas, cruciais em todo ano de 2012 já sem os investimentos do convênio.}.
 Em um segundo momento, aproximações e parcerias diversas ocorreram.
@@
     \item Outros grupos relacionados: o MuSa foi um grupo de JoinVille importante neste trabalho por ter sido fundado por Vilson Vieira, que possui presença forte no LabMacambira.sf.net e nos trabalhos aqui descritos.
-O  Metareciclagem é uma rede engajada em apropriação tecnológica e empoderamento civil com bastante presença dos outros grupos aqui citados e muito movimentada.
+A Metareciclagem é uma rede engajada em apropriação tecnológica e empoderamento civil com bastante presença dos outros grupos aqui citados e muito movimentada.
 A Casa de Cultura Tainã é uma entidade cultural de Campinas/SP, referência no uso de tecnologias livres e atividades socialmente engajadas, e onde muito do que está aqui descrito foi inspirado.
@@
 
-    \item {\bf Microtutoriais Django ~\cite{dmicrotuts}}
+    \item {\bf Microtutoriais Django\footnote{Veja: \url{https://vimeo.com/channels/labmacambira.}} }
 

### 2013-01-21 pretextual.tex
feitas alteracoes propostas pelo prof. Luciano
atualizando dissertacao, varias coisinhas

@@
 \pretextualchapter{}
-[[	Lado dedicado à folha de aprovação. Apagar isso antes de imprimir a versão oficial ]]
 
@@
 %\listadetabelas %Comando que gera a lista de tabelas (AUTOMÁTICO)
-\afterpage{\blankpage}
 
-\pretextualchapter{Lista de relações analíticas descritas e implementadas computacionalmente no Apêndice~\ref{cap:codigoProc}}
+\pretextualchapter{LISTA DE RELAÇÕES ANALÍTICAS DESCRITAS E IMPLEMENTADAS COMPUTACIONALMENTE NO APÊNDICE~\ref{cap:codigoProc}}
 	\begin{listaespecial}[BIGNAMEWIDTH]
@@
 
-\afterpage{\blankpage}
 %

### 2013-01-21 resumoeabstract.tex
atualizando dissertacao, varias coisinhas

@@
 \vspace{-10mm}
-FABBRI, R. \textbf{\ABNTtitulodata}. 2012. 151p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, 2012.
+FABBRI, R. \textbf{\ABNTtitulodata}. 2013. 197p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, São Carlos, São Carlos, 2013.
 \vspace{15mm}
@@
 $\phantom{linha em branco}$\\
-Palavras-chave: Som.
-Áudio.
-Áudio digital.
-Acústica.
-Música.
-Música digital.
-Psicofísica.
-Alta fidelidade.
-Hi-fi.
-Hifi.
-Síntese sonora.
-Síntese musical.
-Python.
-Livecoding.
-Arte e tecnologia.
-Código aberto.
-Cultura hacker.
-Cultura digital.
-Cultura livre.
+Palavras-chave: Som. Áudio digital. Música. Psicofísica. Arte e tecnologia.
 
@@
 \vspace{-10mm}
-FABBRI, R. \textbf{Music on digital audio: psychophysical description and toolbox}. 2012. 151p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, 2012.
+FABBRI, R. \textbf{Music on digital audio: psychophysical description and toolbox}. 2013. 197p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, 2013.
 \vspace{15mm}
@@
 $\phantom{linha em branco}$\\
-Keywords: Sound.
-Audio.
-Digital audio.
-Acoustics.
-Music.
-Digital music.
-Psicophysics.
-High fidelity.
-Hi-fi.
-Hifi.
-Sound synthesis.
-Musical synthesis.
-Python.
-Livecoding.
-Art and technology.
-Open source software.
-Hacker culture.
-Digital culture.
-Free culture.
+Keywords: Sound. Digital audio. Music. Psicophysics. Art and technology.
 

### 2013-01-22 musicaExtra.tex
terminada revisao do apendice e

@@
 A lista de emails do grupo possui membros de diversos dos outros grupos citados.
-    \item Outros grupos relacionados: o MuSa foi um grupo de JoinVille importante neste trabalho por ter sido fundado por Vilson Vieira, que possui presença forte no LabMacambira.sf.net e nos trabalhos aqui descritos.
+    \item Outros grupos relacionados: o MuSa foi um grupo de Joinville importante neste trabalho por ter sido fundado por Vilson Vieira, que possui presença forte no LabMacambira.sf.net e nos trabalhos aqui descritos.
 A Metareciclagem é uma rede engajada em apropriação tecnológica e empoderamento civil com bastante presença dos outros grupos aqui citados e muito movimentada.
+A lista de emails ligada ao movimento Transparência Hacker foi também de importância central como espaço focado em questões de transparência e dados abertos.
 A Casa de Cultura Tainã é uma entidade cultural de Campinas/SP, referência no uso de tecnologias livres e atividades socialmente engajadas, e onde muito do que está aqui descrito foi inspirado.
+O Hacklab do Velho foi fundado por pessoas ligadas ao labMacambira.sf.net e por moradores do alojamento velho da USP, campus de São Carlos, e acolheu diversas experimentações e articulações.
+Outros grupos com alguma importância foram a Casa Fora do Eixo de São Carlos, a Teia Casa de Criação, a Nuvem estação rural de experimentação tecnológica e o grupo que organiza o AVAV Áudiovisual Ao Vivo, ligado ao Epicentro Cultural.
 \end{itemize}
@@
 
-  \subsection{ Tutoriais em texto e código: python, filtros, nyquist e plugins LADSPA, metrics, carta mídias livres, contra-cultura digital }\label{subsec:tuts}
+  \subsection{ Tutoriais em texto e código: python, filtros e nyquist, plugins lv2, metrics, carta mídias livres, contra-cultura digital }\label{subsec:tuts}
 
@@
 Este tutorial foi apresentado parcialmente em Berlim no LAC 2007 e, desde então, foi melhorado algumas vezes. Esta
-primeira versao ficou resumida em forma de texto na plataforma
+primeira versão ficou resumida em forma de texto na plataforma
 Estúdio Livre\footnote{\url{http://estudiolivre.org/python-e-som-tutorial}.}.
@@
 Festival Internacional de \emph{Software} Livre, em Porto Alegre. Como consequência, 
-foi feita uma série de video-tutoriais\footnote{http://estudiolivre.org/tiki-index.php?page=Video+Tutoriais}.
+foi feita uma série de video-tutoriais\footnote{\url{http://estudiolivre.org/tiki-index.php?page=Video+Tutoriais}.}.
 Os vídeos são tidos como iniciáticos em Python por diferentes pessoas ligadas ao movimento de \emph{software} livre.
@@
 Voltados para explicitar fundamentos de áudio digital, estes tutoriais
-são baseados código Python e o equivalente em C. Pequenas explicações são
+são baseados em pequenos \emph{scripts} escritos em Python que exploram conceitos puntuais. Pequenas explicações são
 dadas com o intuito de orientar a exploração inteligente destes \emph{snippets}.
@@
 de forma didática: Wavelets (FIR) e Quad (IIR).
-
-Estes códigos didáticos podem ser baixados do repositório \emph{AudioExperiments}.
+Estes códigos didáticos podem ser baixados do repositório \emph{audioArt}\footnote{Veja \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/audioArt;}.}.
 
@@
 Dadas as dificuldades que o desenvolvimento dos \emph{plugins} de áudio apresenta,
-foi feito um tutorial passo a passo com plugins que rodam em todas as etapas.
-Ele é baseado em uma interface C++ para este padrão de plugin que eh implementado
-em C. Os códigos e os textos estão todos em repositório.
+foi feito um tutorial passo a passo com \emph{plugins} operantes em todas as etapas.
+Os códigos e os textos estão todos em repositório\footnote{Veja \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/lv2Tut;}.}.
 
@@
 O segundo destes conjuntos é dedicado a instrumentalizar de fato o leitor com
-o entendimento do funcionamento dos princípios fundamentais deste framework.
-Uma das primeiras incidências dos \emph{scripts} mínimos, muito usados no
+o entendimento do funcionamento dos princípios fundamentais deste \emph{framework}.
+A importância deste material reside no fato de ser uma das primeiras incidências dos \emph{scripts} mínimos, muito usados no
 labMacambira.sourceforge.net para passagens de tecnologias de forma precisa.
@@
 Idealizado como um meio de interação da comunidade de Python e Música 
-para compartilhamento de códigos e excertos, o FDP foi Baseado principalmente
+para compartilhamento de \emph{scripts} inteiros e excertos, o FDP foi Baseado principalmente
 em documentação organizada sobre as práticas e as bibliotecas
@@
 
-Texto criado em decorrência da participação da comissão de seleção no
-'Prêmio Mídias Livres', a convite do Ministério da Cultura por 'notório saber'.
-A participação consistia em analisar os inscritos no Prêmio Mídias Livres e distribuir 4 milhões de reais dentre categorias regionais e nacionais.
+Texto criado em decorrência da participação do autor do presente escrito na comissão de seleção no
+``Prêmio Mídias Livres'', a convite do Ministério da Cultura por ``notório saber''.
+A participação consistiu em avaliar os inscritos no Prêmio Mídias Livres e distribuir 4 milhões de reais dentre categorias regionais e nacionais.
 Esta carta é um documento único, deixando às claras
@@
 o Cinemetrics e o Literametrics. Uma publicação no
-JSTAT condensa e compara as análises de filosofia
-e música.\cite{FabbriJSTAT}
-Além disso, ele é uma
+\emph{Journal of Statistical Mechanics: Theory and Experiment}
+condensa e compara as análises de filosofia
+e música.\cite{FabbriSTAT}
+Além disso, é uma
 utilização das ciências duras para 
@@
 
-\subsection{Screencasts e outros materiais em video}
+ \subsection{\emph{Screencasts} e outros materiais em video}
 
 \begin{itemize}
-    \item {\bf Python para áudio e música}
+    \item {\bf Python para áudio e música:}
     Relacionado ao tutorial em texto citado acima, foram feitos videos sobre a utilização da linguagem Python,
@@
     Atualmente, são mais de 700 videos sobre sessões de \emph{hacking} e experiências de arte
-    e tecnologia. Feitos pelo LabMacambira.sourceforge.net, o canal tem videos do auto
+    e tecnologia. Feitos pelo LabMacambira.sourceforge.net, o canal tem videos do autor
     deste escrito, e de: Larissa Arruda, Alexandre Negrão, Lucas Zambianchi, Andres Martano,
     Fernando Gorodscki, Chico Simões, Daniel Marcicano, Daniel Pizetta, Daniel Penalva, Nivaldo Bondança,
-    Danilo Shiga, Marcos Murad, Ricardo Fabbri e Vilson Vieira.\footnote{O canal pode ser acessado em:
-    \url{http://vimeo.com/channels/labmacambira}.}
+    Danilo Shiga, Marcos Murad, Ricardo Fabbri e Vilson Vieira\footnote{O canal pode ser acessado em:
+    \url{http://vimeo.com/channels/labmacambira}.}. Exemplos especiais dos videos disponibilizados no canal são:
     \begin{itemize}
-	    \item \emph{Livecoding}: como exemplo dos videos do canal, foram expostos os processos utilizados
-        na apresentação de \emph{livecoding} com ChucK, vacas e sono REM.\footnote{Veja: \url{https://vimeo.com/33012735},
+	    \item \emph{Livecoding}: foram expostos os processos utilizados
+        na apresentação de \emph{livecoding} com ChucK, vacas e sono REM\footnote{Veja: \url{https://vimeo.com/33012735},
         \url{https://vimeo.com/33018740}, \url{https://vimeo.com/33019291}, \url{https://vimeo.com/33025717},
-        \url{https://vimeo.com/33025913}.}
+        \url{https://vimeo.com/33025913}.}.
 
-	    \item Raspagem de dados: outro exemplo, menos artístico e mais engajado, é uma breve explicação
-        sobre o processo de raspagem de dados em páginas HTML.\footnote{Veja: \url{vimeo.com/channels/labmacambira/2681879}.}
+	    \item Raspagem de dados: menos artístico e mais engajado, é uma breve explicação
+        sobre o processo de raspagem de dados em páginas HTML\footnote{Veja: \url{vimeo.com/channels/labmacambira/2681879}.}.
 
@@
 usos em código desta abordagem.
-Como pode-se observar acima, neste mesmo Apêndice, este trabalho, disponibilizado e conebido como tecnologias livres, tem dimensões bastante consideráveis em aspectos socialmente engajados e artísticos.
+Como pode-se observar acima, neste mesmo Apêndice, este trabalho, disponibilizado e concebido como tecnologias livres, tem dimensões bastante consideráveis em aspectos socialmente engajados e artísticos.
 Esta seção exibe os feitos que extrapolam a música por completo.
 
-   \subsection{Tecnologias sociais de alta demanda: Sitios, Conteúdos e Articulação}
+   \subsection{Tecnologias sociais de alta demanda: Sítios, Conteúdos e Articulação}
 
@@
 
-Nest emeio tempo também foram criadas várias ferramentas \emph{web}, que rodam em navegadores comuns, como \emph{Firefox}. Destes, vale citar:
+        Nest meio tempo também foram criadas várias ferramentas \emph{web}, que rodam em navegadores comuns, como o \emph{Firefox}.
+Destes, vale citar\footnote{Os arquivos fonte estão nos repositórios em \url{http://labmacambira.git.sourceforge.net/git/gitweb-index.cgi}.}:
 \begin{itemize}
@@
     \item Ágora Communs, Novo Ágora: sistema de deliberações \emph{online}.
-    \item Economia criativa: rascunho de sistema para facilitar circulação de bens me 4 eixos: venda de produtos, aluguél de equipamentos, aluguél de lugares e prestação de serviços.
+    \item Economia Criativa: rascunho de sistema para facilitar circulação de bens me 4 eixos: venda de produtos, aluguél de equipamentos, aluguél de lugares e prestação de serviços.
     \item Catálogo de Ideias: ferramenta simples para catálogo de ideias sobre temas diferentes, junto a palestras, etc.
     \item Mapeamento dos 301 pontos de SP: mapeamento dos 301 pontos de cultura estaduais.
-    \item Mapas Coletivos: ferramenta para criação e publicação de mapas com possibilidades colaborativas.
+    \item Mapas Coletivos: ferramenta para criação e publicação de mapas com possibilidades colaborativas. Bastante usada por pessoas de meios diferentes pela usabilidade e proposta pertinente.
     \item Leitura de \emph{shapefiles} no Mapas de Vista: implementação de leitura de arquivos do tipo \emph{shapefile} com dados geográficos em um plugin e tema de \emph{Wordpress}.
-    \item Maper: atualmente, um rascunho de uma ferramenta de publicação de mapas com vistas a publicações no facebook e outros formatos.
+    \item Maper: atualmente, um rascunho de uma ferramenta de publicação de Mapas com Vistas a publicações no facebook e outros formatos.
 \end{itemize}
@@
 Nestas documentações também estão apontamentos de recursos artísticos e apresentações realizadas com a AHT, \emph{livecoding} e outras propostas.
+Chama a atenção a quantidade de instâncias de \emph{etherpads} usadas, o labMacambira.sf.net chega a criar vários destes epads por dia, para escrita simultânea a várias mãos\footnote{Veja: \url{http://pontaopad.me/epads}}.
 
@@
 O AA está explicado abaixo.
-Não menos importante são os \emph{Etherpads} e as páginas \emph{wiki} criadas, permitindo proposições e escritas rápidas e em várias mãos.
+Não menos importante são os \emph{Etherpads} e as páginas \emph{wiki} criadas, permitindo proposições e escritas rápidas por diferentes pessoas.
 
@@
 O sistema se baseia no envio de mensagens periódicas sobre as atividades que estão sendo desempenhadas, evitando foco em produtos e propaganda.
+A ferramenta pode ser usada no IRC ou em linha de comando.
+Existem algumas interfaces \emph{web} para exibir as mensagens e dados do AA\footnote{Veja: \url{http://wiki.nosdigitais.teia.org.br/AA}.}.
+Estão planejados desenvolvimentos do AA para que seja utilizável através do \emph{chat} em diferentes redes sociais.
+Também está sendo considerada a utilização do AA para distribuição de verba de acordo com as dedicações comprovadas pelos mecanismos do AA, tornando possível que se trabalhe remuneradamente pelo bem comum.

### 2013-01-23 agradecimentos.tex
feita revisao ateh as simetrias dos coeficientes

@@
 
-Agradeço aos participantes do labMacambira.sf.net pelas colaborações diretas e indiretas nos software, apresentações e articulações desde junho de 2011.
+Agradeço aos participantes do labMacambira.sf.net pelas colaborações diretas e indiretas nos \emph{software}, apresentações e articulações desde junho de 2011.
 Em especial agradeco ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela, Guilherme Lunhani, Patrícia Ferraz, Glerm Soares, Vanessa Ferreira, Sérgio Teixeira de Carvalho.

### 2013-01-23 introducao.tex
atualizando dissertacao, varias coisinhas
feita revisao ateh as simetrias dos coeficientes

@@
 %% ------------------------------------------------------------------------- %%
-\setcounter{page}{23}
 \chapter{Introdução} %Nome do capítulo.
+\setcounter{page}{29}
 \label{cap:intro} 
@@
 Representar estruturas e artifícios musicais através das características do som discretizado
-é a tarefa proposta por este trabalho.
+é a proposta deste trabalho.
 Os resultados são relações matemáticas e suas implementações computacionais.
@@
 A percepção humana do som envolve captações pelos ossos, estômago e orelha, funções de transferência da cabeça e dorso e processamento pelo sistema nervoso.
-Além disso, o ouvido é um órgão dedicado à tarefa de captura destas ondas.
-Seu funcionamento decompõe o som em seu espectro senoidal e passa para o sistema nervoso.\cite{Roederer} Estas componentes senoidais são cruciais para os fenômenos musicais, como se pode observar na composição dos sons de interesse para a música quanto nas afinações e escalas.\cite{floEsp} A subseção~\ref{sec:notaDisc} expõe a presença de senóides no som discretizado e caracteriza a nota musical básica.
+Além disso, o ouvido é um órgão dedicado à captura destas ondas.
+Seu funcionamento decompõe o som em seu espectro senoidal e passa para o sistema nervoso.\cite{Roederer} Estas componentes senoidais são cruciais para os fenômenos musicais, como se pode observar tanto na composição dos sons de interesse para a música quanto nas afinações e escalas.\cite{floEsp} A subseção~\ref{sec:notaDisc} expõe a presença de senoides no som discretizado e caracteriza a nota musical básica.
 
@@
 som e áudio são muitas vezes usados de forma intercambiável.\cite{Everest}} e este pode provir da captura do som por microfones ou da síntese.
-Muitas vezes, o áudio digital é obtido através de protocolos que facilitam o armazenamento e transferência dos arquivos.
-A representação digital do som pode consistir em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de bits.
+Muitas vezes, o áudio digital é especificado através de protocolos que facilitam o armazenamento e transferência dos arquivos.
+A representação digital do som pode consistir em amostras igualmente espaçadas no tempo e cujas amplitudes individuais são registradas com um mesmo número de \emph{bits}.
 Estas amostras separadas por intervalos regulares $\lambda_a$ constituem a forma padrão de representação do som em tempo discreto, chamada de modulação por código de pulsos (PCM do inglês \emph{Pulse Code Modulation}).
-Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$, também chamada de taxa de amostragem, e a profundidade de bit que é o número de bits utilizados para representar a amplitude de cada amostra.
-A figura~\ref{fig:PCM} exibe $25$ amostras de um áudio PCM com $4$ bits cada.
+Um som digital PCM é caracterizado pela frequência de amostragem $f_a=\frac{1}{\lambda_a}$, também chamada de taxa de amostragem, e a profundidade de \emph{bit} que é o número de \emph{bits} utilizados para representar a amplitude de cada amostra.
+A figura~\ref{fig:PCM} exibe $25$ amostras de um áudio PCM com $4$ \emph{bits} cada.
 Os $2^4=16$ grados para a amplitude de cada amostra junto ao espaçamento regular $\lambda_a$ introduzem um erro de quantização.
@@
 Já na década de 90 era evidente que também a música popular, especialmente as músicas eletrônicas de dança, tinham incorporado sons sem altura definida e organizações temporais fora de métricas simples.
-Mesmo assim, a nota permanece paradigmática como 'unidade fundamental' das estruturas musicais e, na prática, podem se desdobrar em sons que contemplam estes desenvolvimentos recentes.
+Mesmo assim, a nota permanece paradigmática como 'unidade fundamental' das estruturas musicais e, na prática, pode se desdobrar em sons que contemplam estes desenvolvimentos recentes.
 A nota musical e sua expansão são abordadas na seção~\ref{varInternas}.\cite{Wisnick,Webern,Lerdahl,Cook}
@@
 pequenos programas para melhor disponibilidade e validação das tecnologias, que constituem a caixa de ferramentas \massa.
-Liberamos este conjunto de códigos em domínio público através de repositórios Git públicos.\cite{gitBook}
+Este conjunto de códigos está disponibilizado em domínio público através de repositórios Git públicos.\cite{gitBook}
 Os \emph{scripts} estão em Python e fazem uso das bibliotecas externas Numpy e Scikits/Audiolab que realizam chamadas à linguagem Fortran para maior eficiência computacional.
@@
 O capítulo seguinte é um texto conciso em que os elementos musicais são apresentados junto às amostras temporais resultantes.
-Para validação e compartilhamento, as implementações em código computacional destas relações e de pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais.} foram reunidas em uma \emph{toolbox} chamada \massa\ e disponibilizadas online e nos Apêndices~\ref{cap:codigoProc} e~\ref{cap:codigoPecas}.
+Para validação e compartilhamento, as implementações em código computacional destas relações e de pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais.} foram reunidas em uma \emph{toolbox} chamada \massa\ e disponibilizadas online e parcialmente nos Apêndices~\ref{cap:codigoProc} e~\ref{cap:codigoPecas}.
 
-Dada esta direção central, consideramos alguns objetivos secundários.
-Destes, destaca-se a difusão da compreensão do código computacional através de práticas lúdicas, no caso a música.
+Dos objetivos secundários, destaca-se a difusão da compreensão do código computacional através de práticas lúdicas, no caso a música.
 Outro objetivo considerado é a apresentação de um arcabouço de síntese sonora e musical com controle amostral, para o qual há potenciais usos em experimentos psico-acústicos e síntese em alta definição (\emph{hi-fi}).
-Também é considerada a apresentação destes conteúdos de forma didática, quase um tutorial, possibilitando uma compreensão e uso facilitados, pois os assuntos tratados são de reconhecida complexidade: processamento de sinais, música, psico-acústica, para citar somente alguns exemplos.
+Também é considerada a apresentação destes conteúdos de forma didática, quase um tutorial, o que possibilita compreensão e uso facilitados.
+Esta exposição amistosa faz-se significativa pois os assuntos tratados são de reconhecida complexidade: processamento de sinais, música, psico-acústica, para citar somente alguns exemplos.
 Deste ponto de vista pedagógico, também se presta a apresentação destes resultados na forma de hipertexto, em que cada \emph{script} e exemplo sonoro/musical seja acessível junto ao material teórico.

### 2013-01-23 scripts/pecas2.3/intervalosEntreAlturas.py

@@ -97,7 +97,7 @@ def inv(I):
     """retorna intervalo inverso de I: 0<= I <=12"""
     return 12-I
 
-# intervalo harmonico
+# intervalo harmônico
 def intervaloHarmonico(f,I,d=.3):
     return (  v(f,d=d)+v(f*2.**(I/12.),d=d)  )*0.5
 # intervalo melódico

### 2013-01-23 scripts/secao2.3/eqs2.3.py

@@ -53,7 +53,7 @@ def inv(I):
     return 12-I
 
 
-# intervalo harmonico
+# intervalo harmônico
 def intervaloHarmonico(f, I):
     return (v(f)+v(f*2.**(I/12.)))*0.5
 

### 2013-01-23 devEresultados.tex
atualizando dissertacao, varias coisinhas
feita revisao ateh as simetrias dos coeficientes

@@
 Canonicamente, as notas possuem ao menos duração, volume, altura e timbre.\cite{Lacerda} Estas
-são qualidades que podem ser tratadas quantitativamente.\cite{Roederer}
+são qualidades tratáveis quantitativamente.\cite{Roederer}
 Assim, as características da nota musical digital básica são dadas pelas amostras igualmente espaçadas no tempo
@@
 \subsection{Duração}
- Definimos a
-frequência (ou taxa) de amostragem $f_a$ como o número de amostras por segundo e
-a sequência $T_i=\{t_i\}$ como um conjunto ordenado de amostras reais separadas por $\delta_a=1/f_a$ segundos.
-
-Assim, uma nota musical de duração $\Delta$
+A frequência (ou taxa) de amostragem $f_a$ 
+é definida como o número de amostras por segundo. Seja
+a sequência $T_i=\{t_i\}$ um conjunto ordenado de amostras reais separadas por $\delta_a=1/f_a$ segundos.
+Uma nota musical de duração $\Delta$
 se apresenta como uma sequência de $ \lfloor \Delta . f_a \rfloor $ amostras\footnote{O
@@
 só satisfaz esta condição em casos muito excepcionais. É necessário
-escolher um inteiro próximo de $\Delta . f_a$ e admitir algum erro. Por simplicidade, será considerada sempre a parte inteira da multiplicação, descrita por $\lfloor \Delta . f_a \rfloor$.}:
+escolher um inteiro próximo de $\Delta .
+f_a$ e admitir algum erro.
+Por simplicidade, será considerada sempre a parte inteira da multiplicação, descrita por $\lfloor \Delta .
+f_a \rfloor$ e aceito o erro de até $-\delta_a$ segundos.
+Por exemplo, $\delta_a=1/44100 \approx 2,3.10^{-5}$, por volta de 23 microssegundos, o que é razoável para usos musicais.}:
 
@@
 
-Denotaremos $\Lambda = \lfloor \Delta . f_a \rfloor$ como o número de amostras da sequência, de forma que $T_i=\{t_i\}_0^{\Lambda-1}$.
+Seja $\Lambda = \lfloor \Delta . f_a \rfloor$ o número de amostras da sequência, de forma que $T_i=\{t_i\}_0^{\Lambda-1}$.
 
 \subsection{Volume}
-A sensação de volume sonoro depende da reverberação e distribuição dos harmônicos, dentre outras características trabalhadas na seção~\ref{varInternas}, e podemos obter variações do volume através da potência da onda~\cite{Chowning}:
+A sensação de volume sonoro depende da reverberação e distribuição dos harmônicos, dentre outras características trabalhadas na seção~\ref{varInternas}.
+Pode-se obter variações do volume através da potência da onda~\cite{Chowning}:
 
@@
 
-Como o volume final dependerá da amplificação do sinal nos alto-falantes, o crucial é a potência relativa de uma nota em relação às outras ou de um trecho da música em relação ao resto.
+O volume final dependerá sempre da amplificação do sinal nos alto-falantes, assim o crucial é a potência relativa de uma nota em relação às outras ou de um trecho da música em relação ao resto.
 As diferenças de volume são medidas em decibels, e estes são
-calculados diretamente com as amplitudes através das energias ou potências\footnote{Lembrando que, devido à nossa percepção logarítmica,
+calculados diretamente com as amplitudes através das energias ou potências\footnote{Lembrando que, devido à percepção logarítmica,
 em um som de volume $v$ a redução da potência $p$ para uma mesma fração $\nu . p $ 
-com $\nu \in [0,1]$ é sentido como a mesma diminuição do volume $v-\kappa$ com $\kappa \geq 0$.}:
+com $\nu \in [0,1]$ é sentido como a mesma diminuição $\kappa$ do volume $v-\kappa$ com $\kappa \geq 0$.}:
 
@@
 
-Para citar também um uso possível da onda quadrada, ela pode ser usada em uma síntese
+Para citar também uma utilidade da onda quadrada, ela pode ser usada em uma síntese
 subtrativa que vise a imitar um clarinete. Este instrumento também só apresenta os
@@
 de $100$ amostras).
-Se $t_a=44,1 kHz$, como no padrão PCM de Compact Disks, a onda possui frequência fundamental $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz $.
+Se $t_a=44,1 kHz$, como no padrão PCM de \emph{Compact Disks}, a onda possui frequência fundamental $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz $.
 Um lá\footnote{Um lá 4, logo acima do dó central, no segundo espaço do pentagrama na clave de sol comum.}, independente da forma de onda.
@@
 A dente de serra é a única com a série harmônica completa (pares e ímpares).
-Já as ondas triangular e quadrada possuem as mesmas componentes espectrais, mas com decaimentos de $-12dB/oitava$ e $-6dB/oitava$.
+Já as ondas triangular e quadrada possuem as mesmas componentes espectrais, mas com decaimentos de $-12dB/oitava$ e $-6dB/oitava$ respectivamente.
 
@@
   As componentes senoidais são geralmente as principais responsáveis pela qualidade chamada timbre.
-Caso não se apresentem em proporções harmônicas (relações de pequenos números), o som é percebido como ruidoso ou dissonante e não com uma sonoridade de frequência fundamental estabelecida unívocamente.
+Caso não se apresentem em proporções harmônicas (relações de pequenos números), o som é percebido como ruidoso ou dissonante e não com uma sonoridade de frequência fundamental estabelecida de forma unívoca.
 Além disso, a noção de altura absoluta em um complexo sonoro é baseada na semelhança do espectro com a série harmônica.\cite{Roederer}
@@
 Pode-se constatar isso em sons reais.
-A onda rotulada como 'som real amostrado' na figura ~\ref{fig:formasDeOnda} é um período de $\lambda_f=114$ amostras extraído de um som real relativamente comportado.
+A onda rotulada como ``som real amostrado'' na figura ~\ref{fig:formasDeOnda} é um período de $\lambda_f=114$ amostras extraído de um som real relativamente comportado.
 A onda de oboé foi amostrada de um lá 4 também em $44,1kHz$.
@@
 
-Definindo $ R_i=\{ r_i \}_0^{\lambda_f-1}$ a sequência de amostras do som real da figura ~\ref{fig:formasDeOnda},
-$R_i$ pode ser tomado como base para um som $T_i^f$ da seguinte forma:
+A sequência 
+$ R_i=\{ r_i \}_0^{\lambda_f-1}$ de amostras do som real da figura ~\ref{fig:formasDeOnda}
+ pode ser tomada como base para um som $T_i^f$ da seguinte forma:
 
@@
     \caption{Espectros das ondas sonoras de uma nota de oboé natural e de período amostrado.
-O som natural possui flutuações nos harmonicos e ruídos, já o som de período amostrado possui espectro perfeitamente harmônico.}
+O som natural possui flutuações nos harmônicos e ruídos, já o som de período amostrado possui espectro perfeitamente harmônico.}
         \label{fig:espectroOboe}
@@
  $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right]$ dada pelo termo $tg^{-1}\left(\frac{b_k}{a_k}\right)$ que possui esta imagem.
-O sinal de $a_k$ especifica se o lado direito ou esquerdo do circulo trigonométrico,  o que completa a varredura de fase: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.
+O sinal de $a_k$ especifica o lado direito ou esquerdo do circulo trigonométrico,  o que completa a varredura de fase: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.
 
@@
 A figura ~\ref{fig:amostras2} exibe duas amostras e as componentes espectrais que contêm.
-A decomposição de Fourier resulta neste caso em um único par de coeficientes $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relativos às frequências $\{f_k\}_0^1=\left\{w_k\frac{f_a}{2\pi}\right\}_0^1=\left\{k\frac{f_a}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_a}{2}=f_{\text{máx}}\right\}$
+A decomposição de Fourier possui neste caso um único par de coeficientes $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relativos às frequências $\{f_k\}_0^1=\left\{w_k\frac{f_a}{2\pi}\right\}_0^1=\left\{k\frac{f_a}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_a}{2}=f_{\text{máx}}\right\}$
 com energias $e_k=\frac{(c_k)^2}{\Lambda=2}$. O papel das amplitudes $a_k$ é nítido com
  $\frac{a_0}{2}$ o deslocamento fixo\footnote{Chamado de \emph{bias} ou \emph{offset}.} e $\frac{a_1}{2}$ a amplitude da oscilação em si, dada pela relação $f_k=k \frac{f_a}{\Lambda=2}$.
-
 Este caso é de especial importância pois o mínimo necessário para representar uma oscilação são 2 amostras e disso resulta a frequência de Nyquist $f_{\text{máx}}=\frac{f_a}{2}$.
@@
 Como consequência temos módulos iguais e fases com sinais opostos}.
-Lembrando que $f_k = k\frac{f_a}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $, pode-se notar que, quando $k> \frac{\Lambda}{2} $,
+Lembrando que $f_k = k\frac{f_a}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $.
+Quando $k> \frac{\Lambda}{2} $,
 a frequência $f_k$ é espelhada em $\frac{f_a}{2}$ da seguinte forma $f_k=\frac{f_a}{2} - (f_k-\frac{f_a}{2})=f_a-f_k=f_a - k\frac{f_a}{\Lambda}=(\Lambda-k)\frac{f_a}{\Lambda} \;\;\;\; \Rightarrow \;\;\;\; f_k\equiv f_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
@@
 
-e resultam evidentes as equivalências ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} e ~\ref{equivalenciasFases}:
+e ficam evidentes as equivalências ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} e ~\ref{equivalenciasFases}:
 
@@
 \centering
+\caption{Resumo das funções harmônicas tonais para a escala maior.
+A tônica é o centro da música, a dominante tende à tônica e a subdominante se distancia da tônica.
+Os três acordes podem, a princípio, serem substituídos livremente pela relativa ou antirelativa.}
 \begin{tabular}{l | c | r}
@@
 \end{tabular}
-\caption{Resumo das funções harmônicas tonais para a escala maior.
-A tônica é o centro da música, a dominante tende à tônica e a subdominante se distancia da tônica.
-Os três acordes podem, a princípio, serem substituídos livremente pela relativa ou antirelativa.}
 \label{tab:harmonia}
@@
 \centering
+\caption{Transição das durações ouvidas individualmente para alturas.}
 \begin{tabular}{  l | r r r r   r r r    r r r || r r || r r r r r r }
@@
 \end{tabular}
-\caption{Transição das durações ouvidas individualmente para alturas.}
 \label{tab:duracoes}
@@
 \centering
+\caption{Change Ringing: \emph{Peal} (padrão) com 3 sinos}
 \begin{tabular}{l c r}
@@
 \end{tabular}
-\caption{Change Ringing: \emph{Peal} (padrão) com 3 sinos}
 \label{tab:change}

### 2013-01-23 pretextual.tex
feita revisao ateh as simetrias dos coeficientes

@@
 	\begin{listaespecial}[BIGNAMEWIDTH]
-        \item[Equação~\ref{eq:dur}] Duração em áudio PCM.
+        \item[Equação~\ref{eq:dur}] Sequência de amostras de um áudio PCM.
         \item[Equação~\ref{eq:potencia}] Potência.
@@
         \item[Equação~\ref{eq:dobraVol}] Variação de amplitude no volume dobrado (10 dBs).
-        \item[Equação~\ref{ampDec}] Conversão de decibels em amplitude.
+        \item[Equação~\ref{ampDec}] Conversão de decibels em variação de amplitude.
         \item[Equação~\ref{periodicidade}] Representação básica de uma sequência periódica.
@@
         \item[Equação~\ref{eq:concatenacao}] Concatenação de N sequências.
-        \item[Equação~\ref{eq:lut}] Procedimento de busca em tabelas (Lookup Table).
+        \item[Equação~\ref{eq:lut}] Procedimento de busca em tabelas (\emph{Lookup Table}).
         \item[Equação~\ref{freqLinear}] Frequências em cada amostra em uma variação linear.
@@
         \item[Subseção~\ref{estCic}] Estruturas cíclicas para a síntese musical.
-        \item[Equação~\ref{eq:groups}] Propriedades básicas de um grupo algébrico.
+%        \item[Equação~\ref{eq:groups}] Propriedades básicas de um grupo algébrico.
 

### 2013-01-23 resumoeabstract.tex
feita revisao ateh as simetrias dos coeficientes

@@
 \vspace{-10mm}
-FABBRI, R. \textbf{\ABNTtitulodata}. 2013. 197p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, São Carlos, São Carlos, 2013.
+FABBRI, R. \textbf{\ABNTtitulodata}. 2013. 197p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, São Carlos, 2013.
 \vspace{15mm}
@@
 \vspace{-10mm}
-FABBRI, R. \textbf{Music on digital audio: psychophysical description and toolbox}. 2013. 197p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, 2013.
+FABBRI, R. \textbf{Music on digital audio: psychophysical description and toolbox}. 2013. 197p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, São Carlos, 2013.
 \vspace{15mm}

### 2013-01-23 devEresultados.tex
feita revisao completa, repassada e bib, jah era

@@
 Os coeficientes $c_k$ se equivalem dois a dois correspondendo às mesmas frequências e com contribuições idênticas\footnote{Parte real igual e imaginária com sinal trocado: $a_{k1}=a_{k2}$ e $b_{k1}=-b_{k2}$.
-Como consequência temos módulos iguais e fases com sinais opostos}.
+Como consequência os módulos são iguais e as fases possuem sinais opostos}.
 Lembrando que $f_k = k\frac{f_a}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $.
@@
 Seja $f$ tal que $f$ divida $f_a$\footnote{Como apontado anteriormente, esta limitação facilita a exposição sem perda de generalidade.
-Esta limitação será superada no início da próxima seção.}.
+A limitação será superada no início da próxima seção.}.
 Uma sequência $T_i$ de amostras sonoras separadas por $\delta_a=1/f_a$ descreve uma nota musical de frequência $f$ Herz e duração $\Delta$ segundos se, e somente se, possuir a periodicidade $\lambda_f=f_a/f$
@@
 
-	Se consideradas somente as incidências diretas em cada ouvido, as equações resultam simples.
-Dada a separação entre os ouvidos\footnote{Constata-se que $\zeta \approx 21,5cm$ para um humano adulto.} $\zeta$,
+	Se consideradas somente as incidências diretas em cada ouvido, as equações são simples.
+Dada a separação $\zeta$ entre os ouvidos\footnote{Constata-se que $\zeta \approx 21,5cm$ para um humano adulto.},
 um objeto localizado em $(x,y)$ conforme a figura~\ref{fig:spac}
@@
 \begin{equation}\label{eq:dii}
-DII=20\log_{10}\left (\frac{d}{d'}\right) \quad dBs
-\end{equation}
-
-Convertendo para amplitude temos $DII_a=\frac{d}{d'}$. A $DII_a$ pode
+DII=20\log_{10}\left (\frac{d}{d'}\right) \quad decibels
+\end{equation}
+
+Convertendo para amplitude, obtem-se $DII_a=\frac{d}{d'}$. A $DII_a$ pode
 ser utilizada como constante multiplicativa do canal direito de um sinal sonoro estéreo: $\{t_i'\}_0^{\Lambda -1}=\{DII_a .
@@
 Pode-se utilizar a DII junto à DTI como adiantamento no tempo do canal direito com relação ao esquerdo, vínculo crucial para a localização em sons graves e em sonoridades percussivas.\cite{Heeger}
-Considerando $\Lambda_{DTI}=\lfloor DTI . f_a \rfloor$, pode-se escrever:
+Considerando $\Lambda_{DTI}=\lfloor DTI . f_a \rfloor$:
 
@@
 
+com $x,y$ tais como representados na figura~\ref{fig:spac}.
 Mesmo assim, há dificuldades quando $\theta$ incide sobre o chamado "cone de confusão" em que um mesmo par de especificações DTI, DII resultam de vários dos pontos
@@
 
-A figura~\ref{fig:spac} mostra esta sombra acústica do crânio, importante para a percepção do ângulo azimutal da fonte no cone de confusão.
-O cone em si não foi disposto na figura pois não é exatamente um cone e suas dimensões precisas não são especificadas na literatura, mas pode ser entendido como um cone com o ápice no meio da cabeca e saindo por cada uma das orelhas.\cite{hrtf}
+A figura~\ref{fig:spac} mostra também esta sombra acústica do crânio, importante para a percepção do ângulo azimutal da fonte no cone de confusão.
+O cone em si não foi disposto na figura pois não é exatamente um cone e suas dimensões precisas não foram encontradas na literatura visitada, mas pode ser entendido como um cone com o ápice no meio da cabeca e saindo por cada uma das orelhas.\cite{hrtf}
 
@@
 
-A partir da nota básica, pode-se realizar estruturas musicais com
-sequências destas partículas. A soma de $N$ sequências $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ de mesmo tamanho $\Lambda$ resulta em seus conteúdos espectrais sobrepostos em um processo de mixagem sonora:
+A partir da nota básica, cabe realizar estruturas musicais com
+sequências destas partículas.
+A soma dos elementos de mesmo índice de $N$ sequências $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ de mesmo tamanho $\Lambda$ resulta em seus conteúdos espectrais sobrepostos em um processo de mixagem sonora:
 
@@
 A duração do trecho é bastante curto $\frac{f_a=44.1kHz}{100} \approx 2 \text{ milisegundos}$.
-Pode-se completar com zeros para somar sequências de tamanhos diferentes.
+Basta completar com zeros para somar sequências de tamanhos diferentes.
 
@@
 As sequências podem também ser concatenadas no tempo.
-Caso as sequências $\{t_{k,i}\}_0^{\Lambda_k-1}$ de tamanhos $\Lambda_k$  representem $k$ notas musicais, sua concatenação em uma única sequência $T_i$ resulta em uma sequência musical simples ou melodia:
+Caso as sequências $\{t_{k,i}\}_0^{\Lambda_k-1}$ de tamanhos $\Lambda_k$  representem $k$ notas musicais, sua concatenação em uma única sequência $T_i$ é em uma sequência musical simples ou melodia:
 
@@
 
-A montagem musical \emph{reduced-fi} demonstra de forma isolada este uso de justaposição temporal das notas, resultando em uma peça homofônica.
+A montagem musical \emph{reduced-fi} explora de forma isolada este uso de justaposição temporal das notas, resultando em uma peça homofônica.
 O princípio vertical está demonstrado nos \emph{quadros sonoros}, sons estáticos com espectros peculiares.
@@
 
-Está descrita a nota musical digital básica e a seção a seguinte desenvolve a evolução temporal de seus conteúdos, como nos glissandos e nas envoltórias.
+Está descrita a nota musical digital básica e a seção a seguinte desenvolve a evolução temporal de seus conteúdos, como nos glissandos e nas envoltórias de volume.
 A filtragem de componentes espectrais e a geração dos ruídos completam a constituição da nota musical como unidade isolada e se desdobra na seção~\ref{notasMusica}, dedicada à estruturação destas notas em música através de métricas e trajetórias.
@@
 básica.
-Todas as relações descritas nesta seção estão no Apêndice~\ref{sec:cod2}.
+Todas as relações descritas nesta seção estão implementadas em Python no Apêndice~\ref{sec:cod2}.
 As montagens musicais \emph{Tansita para metro}, \emph{Vibra e treme}, \emph{Tremolos, vibratos e a frequência}, \emph{Trenzinho de caipiras impulsivos}, \emph{Ruidosa faixa}, \emph{Bela rugosi}, \emph{Chorus infantil}, \emph{ADa e SaRa} estão nos Apêndices~\ref{ap:transita}, \ref{ap:vibra}, \ref{ap:tremolos}, \ref{ap:trenzinho}, \ref{ap:ruidosa}, \ref{ap:bela}, \ref{ap:chorus} e \ref{ap:ada}.
@@
 $T_i^{f,\,\Delta}$ com amostras de um som de frequência $f$ e duração $\Delta$
-pode ser feita a partir de $\widetilde{L_i}$ da seguinte forma:
+pode ser obtida a partir de $\widetilde{L_i}$ da seguinte forma:
 
@@
  A expressão que define a variável $\gamma_i$ pode ser compreendida da
- seguinte forma: $i$ é acrescida de $f_a$ em $1$ segundo.
+ seguinte forma: $i$ é acrescida de $f_a$ a cada $1$ segundo.
  Caso seja dividida pela frequência de amostragem, resulta $\frac{i}{f_a}$,
@@
 O termo $\frac{i}{\Lambda-1}$ varre o intervalo $[0,1]$ e pode-se elevá-lo a uma potência
-para que o início da transição seja mais suave ou abrupta.
+para que o início da transição seja mais suave ou abrupto.
 Este procedimento é útil para variações de energia
@@
 A distribuição espectral deve ser feita com cuidado 
-pois caso utilizemos a mesma fase,
+pois caso se utilize a mesma fase,
 ou fases com forte correlação, 
@@
 \begin{split}
-c_0 & =0 \quad \text{pois não queremos bias} \\
+c_0 & =0 \quad \text{pois evita-se bias} \\
 c_i & =e^{j.x}\;,\;\; j^2=-1 \;, \;\; x \; \text{randômico} \; \in \; [0,2\pi]\;,\;\; i \; \in \; \left[1, \, \frac{\Lambda}{2}-1\right] \\
@@
 Pode-se considerá-la consonante como regra geral.
-A sétima menor por vezes é considerada consonante, assim como algumas culturas entoam o trítono como intervalo consonante.\cite{Roederer,Wisnick}
-
-Este arcabouço resume com exatidão a teoria tradicional dos intervalos musicais.\cite{Lacerda} A montagem \emph{Intervalos entre alturas} explora estes intervalos de formas isoladas e diversas.
-O código está no Apêndice~\ref{ap:intervalos} e disponível online com a \emph{toolbox} \massa.
+A sétima menor por vezes é considerada consonante, assim como algumas culturas entoam o trítono como intervalo consonante.
+Por vezes, as segundas menores, sétimas maiores e trítonos são consideradas dissonãncias forte, as segundas maiores, sétimas menores consideradas dissonâncias fracas.
+As oitavas, quintas e uníssonos consonâncias fortes, terças e sextas consonâncias fracas.\cite{Roederer,Wisnick}
+
+Esta descrição resume a teoria tradicional dos intervalos musicais.\cite{Lacerda} A montagem \emph{Intervalos entre alturas} explora estes intervalos de formas isoladas e diversas.
+O código está no Apêndice~\ref{ap:intervalos} e disponível online com a \emph{toolbox} \massa.\cite{massa}
 
@@
 'escalas' ou padrões, geram estruturas estáveis pelas simetrias internas, e podem ser
-repetidas de forma eficiente e sustentada. Discutiremos mais sobre simetrias e voltaremos
-a estas divisões da oitava por igual.
+repetidas de forma eficiente e sustentada. Abaixo há mais sobre simetrias e
+sobre estas divisões da oitava por igual.
 A montagem \emph{Cristais} expõe cada uma destas escalas tanto melódica quanto harmonicamente e seu código está no Apêndice~\ref{ap:cristais}.
@@
 tom, tom, semitom, tom, tom, tom, semitom. De forma
-que podemos escrever:
+que pode-se escrever:
 \begin{equation}\label{eq:relacaoDia}
@@
 as notas coincidam com algum modo diatônico~\ref{eq:escalas}
-ou sejam em número pequeno, podemos dizer que a
+ou sejam em número pequeno, pode-se dizer que a
 harmonia é modal. Caso contrário, atonal.\cite{harmEXT}.
@@
 Da primeira propriedade conclui-se que toda permutação pode ser operada com outra permutação.
-De fato, pode-se aplicar uma permutação, depois outra e, se comparadas as ordenações inicial e final, temos uma permutação resultante.
+De fato, pode-se aplicar uma permutação, depois outra e, se comparadas as ordenações inicial e final, há uma permutação resultante.
 

### 2013-01-23 scripts/pecas2.2/bellaRugosi.py

@@ -43,10 +43,6 @@ def v(f=200,d=2.,tab=S_i,fv=2.,nu=2.,tabv=S_i):
     return tab[Gamma_i%int(Lt)] # busca dos índices na tabela
 
 
-
-
-
-
 dd=v(tabv=Tr_i ,d=2,fv=35.,nu=7.0)
 dd2=v(tabv=D_i ,d=2,fv=35.,nu=7.0)
 
@@ -66,7 +62,6 @@ aa3=n.array(v(fv=.25/2.,f=200,nu=9,d=8.))
 
 aa=n.vstack(( (zz+aa1).T*.5, (zz+aa2).T*.5,(zz+aa3).T*.5))
 
-
 dd= v(tab=Tr_i,tabv=Tr_i ,d=2,fv=35.,nu=7.0)
 dd2=v(tab=Tr_i,tabv=D_i ,d=2,fv=35.,nu=7.0)
 
@@ -85,7 +80,6 @@ aa2=n.array(v(tabv=Tr_i,fv=.25/2.,f=200,nu=9,d=8.))
 aa3=n.array(v(fv=.25/2.,f=200,nu=9,d=8.))
 
 aa=n.vstack(( aa, (zz+aa1).T*.5, (zz+aa2).T*.5,(zz+aa3).T*.5))
-
 
 dd= v(tab=Q_i,tabv=Tr_i ,d=2,fv=35.,nu=7.0)
 dd2=v(tab=Q_i,tabv=D_i ,d=2,fv=35.,nu=7.0)
@@ -127,4 +121,3 @@ aa=n.vstack(( aa, (zz+aa1).T*.5, (zz+aa2).T*.5,(zz+aa3).T*.5))
 
 print("BellaRugosiSdadE.wav escrito")
 a.wavwrite(aa,"BellaRugosiSdadE.wav",f_a)
-

### 2013-01-23 codigoPecas.tex
feita revisao completa, repassada e bib, jah era

@@
 \label{cap:codigoPecas}
-Todas as peças a seguir foram feitas para exemplificar as relações apresentadas no capítulo~\ref{cap:resultados} e são disponibilizadas online junto ao toolbox \massa.\cite{MASSA}
+Todas as peças a seguir foram feitas para exemplificar as relações apresentadas no capítulo~\ref{cap:resultados} e são disponibilizadas online junto ao \emph{toolbox} \massa.\cite{MASSA}
 
@@
 \subsection{Quadros sonoros}\label{ap:quadros}
-A montagem musical 'quadros sonoros' é decidada para demonstração da mixagem pela soma direta de sequências amostrais (subseção~\ref{subsec:basMus}).
+A montagem musical `quadros sonoros' é dedicada para demonstração da mixagem pela soma direta de sequências amostrais (subseção~\ref{subsec:basMus}).
 Esta rotina sintetiza 5 pequenas peças de sonoridades estáticas.
@@
 Houve relatos em listas de \emph{emails} de que estas sonoridades causaram alteração do estado de consciência.
-Em especial, o quadro 5 causou em diferentes pessoas o mesmo efeito: a sensação de escutar sons da mandimbula.
+Em especial, o quadro 5 causou em diferentes pessoas o mesmo efeito: a sensação de amplificação e distorção dos sons da mandimbula.
 
@@
 \subsection{Micro tom}\label{ap:micro}
-Pequena peça musical para demonstração da microtonalidade de quartos de tom e de sétimos de oitava, conceitos apresentados na subseção~\ref{subsec:intervalos}, e disponibilizada online junto à \massa.\cite{MASSA}
+Pequena peça musical para demonstração da microtonalidade.
+Usados de quartos de tom e de sétimos de oitava, conceitos apresentados na subseção~\ref{subsec:intervalos}.
+Disponibilizada \emph{online} junto à \massa.\cite{MASSA}
 \code{micro tom}{scripts/pecas2.3/microTom.py}
@@
 \subsection{Dirracional}\label{ap:dirracional}
-Pequena peça musical de 22 segundos para demonstração das estruturas direcionais e posicionalmentos dos clímax, conceitos apresentados na seção~\ref{notasMusica}, e disponibilizada online junto à \massa.\cite{MASSA}
+Pequena peça musical de 22 segundos para demonstração das estruturas direcionais e posicionamentos dos clímax, conceitos apresentados na seção~\ref{notasMusica}, e disponibilizada online junto à \massa.\cite{MASSA}
 \code{dirracional}{scripts/pecas2.3/dirracional.py}

### 2013-01-23 conclusoes.tex
feita revisao completa, repassada e bib, jah era

@@
 atividades artísticas e educacionais, como cursos, workshops e apresentações artísticas.
-Vários destes resultados estão apresentados no Apêndice~\ref{cap:musicaExtra} e no acervo online criado, que ultrapassa 700 vídeos, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos~\cite{siteLM,wikiLM,vimeoLM}.
+Vários destes resultados estão apresentados no Apêndice~\ref{cap:musicaExtra} e no acervo online criado, que ultrapassa 700 vídeos, documentações escritas, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos~\cite{siteLM,wikiLM,vimeoLM}.
+
+Está havendo um aumento no número de pesquisas em música em andamento no campus de São Carlos da USP, o que sugere facilidade para estabelecer parcerias.
+As publicações acadêmicas efetivadas durante este mestrado também apontam para uma multidisciplinaridade,
+tratando diretamente de questões humanas como artes, filosofia, humor e linguagem falada e escrita, através de artifícios lineares e estatísticos.\cite{FabbriSTAT,FabbriACL,FabbriComplenetVoz,FabbriComplenetTexto} Os desdobramentos estão alcançando redes sociais e teorias epistemológicas com base em pesquisas prévias, dos orientadores deste trabalho, em redes complexas e processamento de linguagem natural\footnote{Os títulos, por serem numerosos, não constram na bibliografia.}.

### 2013-01-23 musicaExtra.tex
feita revisao ateh as simetrias dos coeficientes
feita revisao completa, repassada e bib, jah era

@@
 
-   \subsection{Tecnologias sociais de alta demanda: Sítios, Conteúdos e Articulação}
+   \subsection{Tecnologias sociais: Sítios, Conteúdos e Articulação}
 
@@
 Estão planejados desenvolvimentos do AA para que seja utilizável através do \emph{chat} em diferentes redes sociais.
-Também está sendo considerada a utilização do AA para distribuição de verba de acordo com as dedicações comprovadas pelos mecanismos do AA, tornando possível que se trabalhe remuneradamente pelo bem comum.
+Também está sendo considerada a utilização do AA para distribuição de verba de acordo com as dedicações comprovadas pelos mecanismos do AA, tornando possível que se trabalhe de forma remunerada pelo bem comum.
+
+\section{Momento atual e previsões}
+Está sendo escrito sobre \emph{livecoding}, especialmente
+ligado ao Vivace e ao subgênero \emph{Freakcoding}, que recebeu um manifesto recentemente.
+Um artigo explicando os vínculos entre a proposta estética e o programa Vivace está sendo escrito por várias mãos.
+O georeferenciamento, perene na atuação do labMacambira.sf.net, tomou agora a direção de arrumar animações de fluxos em arestas de grafos dispostos em mapas.
+Apresentações artísticas estão previstas para uso da AHT e experimentações de \emph{livecoding}.
+Um \emph{crowdfunding} foi feito para experimentação de difusão na rede e respostas, com estudos em andamento sobre comportamentos no \emph{Facebook} e listas de \emph{emails}.
+O Cultura Viva, programa do governo federal que apoiou parte destas iniciativas, indica fortalecimento, tornando-se política de estado, não somente do governo que a implantou.
+As aproximações federais são observadas também por parte do MEC (Ministério da Educação), com a prestação de uma acessoria em 2012, e do MinC, devido à Lei Cultura Viva.
+Comunicações mais moderadas continuam com o MCTi, MC e MMA.
+Os canais de IRC e emails do labMacambira.sf.net estão bastante ativos e focados em execução de tarefas, e, dentre os desenvolvimentos em código previstos, está o aproveitamento desta dissertação para implementações em JavaScript, experimentações, apresentações, publicações, usos didáticos e estudo propriamente dito.
+O alto rendimento do labmacambira.sourceforge.net no GSoC de 2012, e o forte empenho de alunos do IPRJ/UERJ, de Nova Friburgo, através do suporte oferecido pelo prof.
+Ricardo Fabbri, da mesma instituição, apontam para uma vida própria que estes andamentos tomaram.
+Em São Paulo, a atuação de Geraldo Magela e Caleb Luporini se propõe a difundir estas práticas em circuitos artísticos e de ensino.
+Por IRC, emails privados e listas, os participantes do grupo estabelecem trocas com pessoas em diversas localidades no Brasil e no mundo para fins de desenvolvimento tecnológico, trabalhos, elaborações didáticas e convivências, mesmo que brandas, por protocolos de comunicação, preferencialmente o IRC.

### 2013-01-23 scripts/pecas2.1/reduced-fi.py

@@ -5,7 +5,8 @@
 # ou
 # python reduced-fi.py
 
-import math as m, random as r
+import math as m
+import random as r
 
 def p(freq):
     return 2*m.pi*freq/44100

### 2013-01-23 scripts/pecas2.3/cristais.py

@@ -71,6 +71,8 @@ def adsr(som,A=10.,D=20.,S=-20.,R=100.,xi=1e-2):
     A_i=n.hstack(  (A_i,R)  )
 
     return som*A_i
+
+
 ### 2..7.5. Escalas simétricas
 Ec_i=[0.,1.,2.,3.,4.,5.,6.,7.,8.,9.,10.,11.]
 Et_i=[0.,2.,4.,6.,8.,10.]
@@ -84,19 +86,15 @@ notast_D=[adsr( v(400*2**(-i/12.),d=.2 ) ) for i in Et_i]
 s=H( notast_D +notast_A+notast_D)
 s1=n.copy(s)
 
-
-
 notastm_A=[adsr( v(200*2**(i/12.),d=.2 ) ) for i in Etm_i]
 
 notastm_D=[adsr( v(400*2**(-i/12.),d=.2 ) ) for i in Etm_i]
 s=H(( notastm_D +notastm_A+notastm_D))
 
-
 notastM_A=[adsr( v(200*2**(i/12.),d=.2 ) ) for i in  EtM_i]
 
 notastM_D=[adsr( v(400*2**(-i/12.),d=.2 ) ) for i in EtM_i]
 s=H(( [list(s)]+ notastM_D +notastM_A+notastM_D))
-
 
 notastt_A=[adsr( v(200*2**(i/12.),d=.2 ) ) for i in  Ett_i]
 
@@ -107,10 +105,6 @@ s=H((s,s[::-1]))
 s=H((s,s[::2],s[::-2]))
 s=H((  s,s+H((s[::2],s[::-2])),
        s+H((s[::4],s[::-4],s[::4],s[::-8],s[::-8])) ))
-
-
-
-
 
 s=((s-s.min())/(s.max()-s.min()))*2.-1.
 a.wavwrite(s,"cristais.wav",f_a) # escrita do som

### 2013-01-27 conclusoes.tex
pequeno erro de digitacao nos agradecimentos

--- 2013-01-18 conclusoes.tex
@@
 atividades artísticas e educacionais, como cursos, workshops e apresentações artísticas.
-Vários destes resultados estão apresentados no Apêndice~\ref{cap:musicaExtra} e no acervo online criado, que ultrapassa 700 vídeos, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos~\cite{siteLM,wikiLM,vimeoLM}.
+Vários destes resultados estão apresentados no Apêndice~\ref{cap:musicaExtra} e no acervo online criado, que ultrapassa 700 vídeos, documentações escritas, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos.\cite{siteLM,wikiLM,vimeoLM}
+
+Há um aumento no número de pesquisas em música em andamento no campus de São Carlos da USP, o que sugere facilidade para estabelecer parcerias.
+As publicações acadêmicas efetivadas durante este mestrado também apontam para uma multidisciplinaridade,
+tratando diretamente de questões humanas como artes, filosofia, humor e linguagem falada e escrita, através de artifícios lineares e estatísticos.\cite{FabbriSTAT,FabbriACL,FabbriComplenetVoz,FabbriComplenetTexto} Os desdobramentos estão alcançando redes sociais e teorias epistemológicas com base em pesquisas prévias, dos orientadores deste trabalho, com forte presença de redes complexas e processamento de linguagem natural.

### 2013-01-27 agradecimentos.tex
pequeno erro de digitacao nos agradecimentos

@@
 Agradeço aos participantes do labMacambira.sf.net pelas colaborações diretas e indiretas nos \emph{software}, apresentações e articulações desde junho de 2011.
-Em especial agradeco ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela, Guilherme Lunhani, Patrícia Ferraz, Glerm Soares, Vanessa Ferreira, Sérgio Teixeira de Carvalho.
+Em especial agradeço ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela, Guilherme Lunhani, Patrícia Ferraz, Glerm Soares, Vanessa Ferreira, Sérgio Teixeira de Carvalho.
 Agradeço aos demais colaboradores das listas de email, IRC, AA e outros canais.

### 2013-01-27 musicaExtra.tex
pequeno erro de digitacao nos agradecimentos

@@
 Em São Paulo, a atuação de Geraldo Magela e Caleb Luporini se propõe a difundir estas práticas em circuitos artísticos e de ensino.
-Por IRC, emails privados e listas, os participantes do grupo estabelecem trocas com pessoas em diversas localidades no Brasil e no mundo para fins de desenvolvimento tecnológico, trabalhos, elaborações didáticas e convivências, mesmo que brandas, por protocolos de comunicação, preferencialmente o IRC.
+Por emails privados, listas ou IRC, os participantes do grupo estabelecem trocas com pessoas em diversas localidades no Brasil e no mundo para fins de desenvolvimento tecnológico, trabalhos, elaborações didáticas e convivências, mesmo que brandas, por protocolos de comunicação.
+A proposta no momento é focar em algo que contemple estas coisas todas da melhor forma, aliando cobertura dos interesses e precisão no aprofundamento, o que está despontado nos estudos de \emph{logs} de IRC e listas de \emph{emails}, com bancos de dados massivos e públicos.
+As possibilidades artísticas, engajadas, científicas e de instrução e interesse pessoal são confluentes através da modelagem por redes complexas, o que sugere a unificação do conhecimento, traço marcante da área e observado com frequência.

### 2013-01-27 resumoeabstract.tex
pequeno erro de digitacao nos agradecimentos

@@
 \vspace{-10mm}
-FABBRI, R. \textbf{\ABNTtitulodata}. 2013. 197p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, São Carlos, 2013.
+FABBRI, R. \textbf{\ABNTtitulodata}. 2013. 213p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, São Carlos, 2013.
 \vspace{15mm}
@@
 \vspace{-10mm}
-FABBRI, R. \textbf{Music on digital audio: psychophysical description and toolbox}. 2013. 197p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, São Carlos, 2013.
+FABBRI, R. \textbf{Music on digital audio: psychophysical description and toolbox}. 2013. 213p. Dissertação (Mestrado) - Instituto de Física de São Carlos, Universidade de São Paulo, São Carlos, 2013.
 \vspace{15mm}
@@
 It is possible to synthesize whole albums through collage of the scripts and parameterization specified by the user.
-With the paradigm of open source implementation, \massa toolbox can be expanded in co-authorship processes and used freely by musicians, engineers and other interested parties.
-In fact, \massa has already been employed by external users for diverse purposes which include music production, artistic presentations, psychoacoustic experiments and computer language diffusion where the appeal of audiovisual artifacts is exploited.
+With the paradigm of open source implementation, \massa\ toolbox can be expanded in co-authorship processes and used freely by musicians, engineers and other interested parties.
+In fact, \massa\ has already been employed by external users for diverse purposes which include music production, artistic presentations, psychoacoustic experiments and computer language diffusion where the appeal of audiovisual artifacts is exploited.
 
 $\phantom{linha em branco}$\\
-Keywords: Sound. Digital audio. Music. Psicophysics. Art and technology.
+Keywords: Sound. Digital audio. Music. Psychophysics. Art and technology.
 

### 2013-02-13 agradecimentos.tex
atualizacao de coisas menores

@@
 Agradeço aos participantes do labMacambira.sf.net pelas colaborações diretas e indiretas nos \emph{software}, apresentações e articulações desde junho de 2011.
-Em especial agradeço ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela, Guilherme Lunhani, Patrícia Ferraz, Glerm Soares, Vanessa Ferreira, Sérgio Teixeira de Carvalho.
+Em especial agradeço ao Daniel Penalva, Caleb Macarenhas, Chico Simões, Fábio Simões, Daniel Marostegan, Marcos Mendonça, Geraldo Magela Rocha, Guilherme Lunhani, Patrícia Ferraz, Glerm Soares, Danilo Shiga, Edson "Presto"\, Corrêa, Vanessa Ferreira e Sérgio Teixeira de Carvalho.
 Agradeço aos demais colaboradores das listas de email, IRC, AA e outros canais.

### 2013-02-18 resumoeabstract.tex
colocado italico em 3 palavras de origem estrangeira no resumo

@@
 
-A representação dos elementos básicos da música - tais como notas, ornamentos e estruturas intervalares - em termos do som discretizado é bastante utilizada em software e rotinas para criação musical e tratamento sonoro.
+A representação dos elementos básicos da música - tais como notas, ornamentos e estruturas intervalares - em termos do som discretizado é bastante utilizada em \emph{software} e rotinas para criação musical e tratamento sonoro.
 Não há, entretanto, uma abordagem concisa que relacione estes elementos às amostras sonoras.
@@
 O elemento fundamental, a nota musical básica com duração, volume, altura e timbre, é relacionado quantitativamente às características do sinal digital.
-As variações internas, como tremolos, vibratos e flutuações espectrais, também são contempladas, o que permite sintetizar notas com inspiração nos instrumentos musicais reais além de sonoridades novas.
+As variações internas, como \emph{tremolos}, vibratos e flutuações espectrais, também são contempladas, o que permite sintetizar notas com inspiração nos instrumentos musicais reais além de sonoridades novas.
 A partir desta representação das notas, dispomos de recursos para a geração de estruturas musicais, como a métrica rítmica, os intervalos de altura e os ciclos.
@@
 Nomeamos \massa, Música e Áudio em Sequências e Séries Amostrais, este ferramental e sua eficácia foi comprovada com a síntese de pequenas peças usando notas básicas, notas incrementadas e notas em música.
-É possível, também, sintetizar álbuns inteiros através de colagens dos scripts e parametrização especificada pelo usuário.
+É possível, também, sintetizar álbuns inteiros através de colagens dos \emph{scripts} e parametrização especificada pelo usuário.
 Com o paradigma de implementação em código aberto, a (caixa de ferramentas) \massa\  pode ser expandida em processos de co-autoria e usada livremente por músicos, engenheiros e outros interessados.

### 2013-02-18 introducao.tex
adicionados italicos em online

@@
 Os resultados são relações matemáticas e suas implementações computacionais.
-Uma descrição teórica está no capítulo~\ref{cap:resultados} e o conjunto de \emph{scripts} disponibilizados no Apêndice~\ref{cap:codigoProc} e online.\cite{MASSA} A \emph{toolbox} resultante recebeu o nome \massa\ (música e áudio em sequências e séries amostrais) e foi utilizada para fazer pequenas peças e montagens focadas nos princípios expostos.
+Uma descrição teórica está no capítulo~\ref{cap:resultados} e o conjunto de \emph{scripts} disponibilizados no Apêndice~\ref{cap:codigoProc} e \emph{online}.\cite{MASSA} A caixa de ferramentas (\emph{toolbox}) resultante recebeu o nome \massa\ (música e áudio em sequências e séries amostrais) e foi utilizada para fazer pequenas peças e montagens focadas nos princípios expostos.
 O Apêndice~\ref{cap:codigoPecas} possui uma relação destas montagens assim como o diretório \emph{exemplos\_de\_uso} da \massa.
@@
 O capítulo seguinte é um texto conciso em que os elementos musicais são apresentados junto às amostras temporais resultantes.
-Para validação e compartilhamento, as implementações em código computacional destas relações e de pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais.} foram reunidas em uma \emph{toolbox} chamada \massa\ e disponibilizadas online e parcialmente nos Apêndices~\ref{cap:codigoProc} e~\ref{cap:codigoPecas}.
+Para validação e compartilhamento, as implementações em código computacional destas relações e de pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais.} foram reunidas em uma \emph{toolbox} chamada \massa\ e disponibilizadas \emph{online} e parcialmente nos Apêndices~\ref{cap:codigoProc} e~\ref{cap:codigoPecas}.
 

### 2013-03-20 devEresultados.tex
pequeno erro de digitacao nos agradecimentos
atualizacao de coisas menores
arrumadas chaves do vibrato em altura
terminada pequenas correcoes antes de adentrar 2.3

@@
 A quantidade $V_{dB}$ possui a unidade decibel ($dB$). A cada 10 $dB$ se atribui
-a sensação de "volume dobrado". Importantes resultados de referência são os equivalentes em $dB$ de se dobrar
+a sensação de "volume dobrado". Referências úteis são os $10dB$ por grado na escala
+de intensidades: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} e \emph{fortissimo}. Valores ainda
+mais cruciais são equivalentes em $dB$ de se dobrar
 a amplitude ou a potência:
@@
 
-Outro valor de referência é o ganho de amplitude
+e o ganho de amplitude
 necessário para que uma sequência tenha o volume dobrado ($10dB$ a mais):
@@
 
+\afterpage{\blankpage}
 \clearpage
@@
 é chamado profundidade do vibrato e é geralmetne dado por conveniência em semitons ou cents.}.
-Estas qualidades não são praticáveis em instrumentos musicais tradicionais, introduzindo novidade nas possibilidades artísticas.
+Estas oscilações com formas precisas e amplitudes arbitrárias não são praticáveis em instrumentos musicais tradicionais, introduzindo novidade nas possibilidades artísticas.
 
@@
 \begin{equation}\label{trT}
-T_i^{tr(f')}=\{ t_i^{tr(f')} \}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}=\{t_i .10^{t_i' \frac{V_{dB}}{20}}    \}_0^{\Lambda-1}=\{t_i . a_{\text{máx}}^{t_i'}\}_0^{\Lambda-1}
+T_i^{tr(f')}=\left \{ t_i^{tr(f')} \right \}_0^{\Lambda-1}=\{ t_i .
+a_i \}_0^{\Lambda-1}=\left \{t_i .10^{t_i' \frac{V_{dB}}{20}}    \right \}_0^{\Lambda-1}=\left\{t_i .
+a_{\text{máx}}^{t_i'} \right\}_0^{\Lambda-1}
 \end{equation}
@@
 P .\sin \left ( f.2\pi\frac{i}{f_a -1} \right ) \right \}_0^{\Lambda-1} = \\
-& = P.\sin \left( f.2\pi\frac{i}{f_a -1}  \right ) + \frac{P.M}{2} \left [ \sin \left( (f-f').2\pi\frac{i}{f_a -1}  \right ) + \sin \left( (f+f').2\pi\frac{i}{f_a -1}  \right ) \right ]
+                       & =  \left\{P.\sin \left( f.2\pi\frac{i}{f_a -1}  \right ) + \frac{P.M}{2} \left [ \sin \left( (f-f').2\pi\frac{i}{f_a -1}  \right ) + \sin \left( (f+f').2\pi\frac{i}{f_a -1}  \right ) \right ] \right \}_0^{\Lambda-1}
 \end{split}
@@
 Esta descrição resume a teoria tradicional dos intervalos musicais.\cite{Lacerda} A montagem \emph{Intervalos entre alturas} explora estes intervalos de formas isoladas e diversas.
-O código está no Apêndice~\ref{ap:intervalos} e disponível online com a \emph{toolbox} \massa.\cite{massa}
+O código está no Apêndice~\ref{ap:intervalos} e disponível online com a \emph{toolbox} \massa.\cite{MASSA}
 
@@
 \begin{split}
-d_i & =\{2,2,1,2,2,2,1\} \\
+\{d_i\} & =\{2,2,1,2,2,2,1\} \\
 e_0 & =0 \\

### 2013-03-21 devEresultados.tex
terminada parte de intervalos, indo para escalas

@@
 Seja $ S_j=\{T_{i,1},T_{i,2}...T_{i,N}\} =\bigcup_{j \in [1,N]} T_{i,j} $ uma sequência de $N$ eventos
-musicais. Para simplificar, sejam notas estes eventos musicais
-e $S_j$ uma estrutura musical. Esta seção é dedicada às técnicas
+musicais. Por simplicidade, sejam notas estes eventos musicais
+e seja $S_j$ chamada uma 'estrutura musical'. Esta seção é dedicada às técnicas
 que tornam $S_j$ interessante e agradável na audição.
@@
 Os elementos de $S_j$ podem ser sobrepostos por mixagem, como na equação~\ref{eq:mixagem} e figura~\ref{fig:mixagem}, formando intervalos e acordes.
-Resultam no 'pensamento vertical' da música.
+Este é o 'pensamento vertical' da música.
 A concatenação de $S_j$, como na equação~\ref{eq:concatenacao} e na figura~\ref{fig:concatenacao}, forma sequências melódicas e ritmos, associados ao 'pensamento horizontal' na música.
-A frequência fundamental $f$ e o momento de início (ataque) são as características mais importantes dos elementos de $S_j$, possibilitando, grosso modo, a presença de músicas de alturas (harmonia e melodia) e a presença da métrica temporal e ritmos, respectivamente.
-
-\subsection{Afinação, intervalos, escalas e acordes}\label{subsec:intervalos}
+A frequência fundamental $f$ e o momento de início (ataque) são, em geral, as características mais importantes dos elementos de $S_j$.
+Estas viabilizam músicas de alturas (harmonia e melodia) e a presença da métrica temporal e ritmos, respectivamente.
+
+\subsection{Afinação}
 O dobro da frequência é uma oitava ascendente ($f=2f_0$).
+A divisão da oitava em doze notas é o cânone da música ocidental clássica,
+além de usos cerimoniais/religiosos e étnicos observados fora da tradição ocidental.\cite{Wisnick}
 Doze semitons equidistantes para o ouvido formam uma oitava,
@@
 e $f$.
-O fator $\varepsilon=2^{\frac{1}{12}}$, de um semitom, forma uma grade de notas
+O fator $\varepsilon=2^{\frac{1}{12}}$, um semitom, forma uma grade de notas
 no espectro audível. Fixada uma frequência $f$, as frequências fundamentais possíveis
@@
 Esta precisão absoluta é característica de implementações
-computacionais, pois os intrumentos reais possuem desvios destas frequências para mais bem compatibilizar os harmônicos
-de suas notas.\cite{Roederer}
-
-Esta divisão da oitava em doze notas é o cânone da música ocidental clássica,
-além de incidir em usos cerimoniais/religiosos e étnicos.\cite{Wisnick} Notados em proporções de $\varepsilon$ entre as notas (i.e.
-um semitom), estes intervalos são, junto a suas notações abreviadas usuais:
-
-\begin{equation}\label{eq:intervalos}
-\begin{split}
-\text{uníssono} = 1J & = 0 \\
-\text{semitom ou segunda menor} =2m & = 1 \\
-\text{tom, tom inteiro ou segunda maior} =2M & = 2 \\
-\text{terça menor} = 3m & = 3 \\
-\text{terça maior} = 3M & = 4 \\
-\text{quarta justa} = 4J & = 5 \\
-\text{quarta aumentada, quinta diminuta ou trítono} = 4aum = 5dim = Tri & = 6 \\
-\text{quinta justa} = 5J & = 7 \\
-\text{sexta menor} = 6m & = 8 \\
-\text{sexta maior} = 6M & = 9 \\
-\text{sétima menor} = 7m & = 10 \\
-\text{sétima maior} = 7M & = 11 \\
-\text{oitava} = 8J & = 12
-\end{split}
-\end{equation}
-
-Nesta nomenclatura, um dos padrões presentes no Brasil,
-um intervalo justo (uníssono, quarta, quinta e oitava), ou um intervalo maior (segunda maior, terça maior, sexta maior ou sétima maior), acrescido de um semitom, resulta em um intervalo aumentado.
-Um intervalo justo, ou um intervalo menor (segunda menor, terça menor, sexta menor ou sétima menor), decrescido de um semitom, resulta em um intervalo diminuto.
-Um intervalo maior, descrescido de um semitom,
-resulta em um intervalo menor.
-Um intervalo menor, acrescido de um semitom, resulta em um intervalo maior.
-Um intervalo aumentado, acrescido de um semitom, resulta em um intervalo 'mais que aumentado' e um intervalo diminuto, descrescido de um semitom, resulta em um intervalo 'mais que diminuto'\footnote{A
- necessidade da nomenclatura, notação e uso de intervalos aumentados/diminutos e mais que aumentados/diminutos é consequência do sistema tonal, em que os graus da escala (veja abaixo) são realmente notas diferentes, com funções e usos específicos.
-Assim, em uma escala de dó bemol maior, a tônica - primeiro grau - é dó bemol, não si, e a sensível - sétimo grau - é si bemol, não lá sustenido ou dó duplo bemol.}.\cite{Lacerda}
-
-Caso as notas sejam dispostas sequencialmente, o intervalo é melódico.
-A ordem das notas, primeiro a nota mais grave ou a mais aguda, resulta em um intervalo ascendente ou descendente, respectivamente.
-Passando a nota mais grave para a oitava acima, ou a nota mais aguda uma oitava para baixo, o intervalo é invertido.
-Um intervalo, somado à sua inversão, resulta 9 (7m inverte para 2M: $7+2=9$).
-Um intervalo maior invertido resulta em um intervalo menor e vice-versa.
-Um intervalo aumentado invertido resulta diminuto e vice-versa, assim como o mais que aumentado resulta em mais que diminuto e vice-versa.
-Um intervalo justo invertido resulta igualmente justo.
-Um intervalo maior que a oitava é classificado como intervalo composto e leva o nome do intervalo entre as mesmas notas, mas na mesma oitava.
-
-Caso as notas soem simultaneamente, o intervalo é harmônico.
-Os intervalos de uníssono, oitava e quinta justa são consonâncias perfeitas.
-As terças e sextas são consonâncias imperfeitas.
-As segundas, sétimas e trítonos são considerados dissonantes.
-Esta classificação não é absoluta.
-A discrepância mais comum é com relação à quarta justa, que é tida como consonante perfeita ou dissonante de acordo com o contexto e arcabouço teórico.
-Pode-se considerá-la consonante como regra geral.
-A sétima menor por vezes é considerada consonante, assim como algumas culturas entoam o trítono como intervalo consonante.
-Por vezes, as segundas menores, sétimas maiores e trítonos são consideradas dissonãncias forte, as segundas maiores, sétimas menores consideradas dissonâncias fracas.
-As oitavas, quintas e uníssonos consonâncias fortes, terças e sextas consonâncias fracas.\cite{Roederer,Wisnick}
-
-Esta descrição resume a teoria tradicional dos intervalos musicais.\cite{Lacerda} A montagem \emph{Intervalos entre alturas} explora estes intervalos de formas isoladas e diversas.
-O código está no Apêndice~\ref{ap:intervalos} e disponível online com a \emph{toolbox} \massa.\cite{MASSA}
-
-Na escala cromática
-de temperamento igual, existem 5 divisões perfeitamente simétricas, consideradas
-escalas pelas simetrias internas e usos fáceis que disso resultam. Como sequências de inteiros aos quais $\varepsilon$ é elevado
-para multiplicar $f_0$, as escalas são:
-
-\begin{equation}\label{escSim}
-\begin{split}
-\text{cromática} & = E_i^c = \{e_i^c\}_0^{11} =  \{0,1,2,3,4,5,6,7,8,9,10,11\} = \{i\}_0^{11}\\
-\text{tons inteiros} & = E_i^t = \{e_i^t\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
-\text{terças menores} & = E_i^{tm} = \{e_i^{tm}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
-\text{terças maiores} & = E_i^{tM} = \{e_i^{tM}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
-\text{trítonos} & = E_i^{tt} = \{e_i^{tt}\}_0^{1} = \{ 0, 6 \} = \{6.i\}_0^1
-\end{split}
-\end{equation}
-
-Assim, a terceira nota da escala em tons inteiros com $f_0=200Hz$
-é $f_3=\varepsilon^{e_3^t} . f_0 = 2^{\frac{6}{12}} . 200 \approxeq 282.843 Hz$. Estas
-'escalas' ou padrões, geram estruturas estáveis pelas simetrias internas, e podem ser
-repetidas de forma eficiente e sustentada. Abaixo há mais sobre simetrias e
-sobre estas divisões da oitava por igual.
-A montagem \emph{Cristais} expõe cada uma destas escalas tanto melódica quanto harmonicamente e seu código está no Apêndice~\ref{ap:cristais}.
-Como parte da \massa, este código é disponibilizado online.
-
-As escalas diatônicas:
-
-\begin{equation}\label{eq:escalas}
-\begin{split}
-\text{menor natural} = \text{modo eólico} & = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
-\text{modo lócrio} & = E_i^{mlo} = \{e_i^{mlo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
-\text{maior}  = \text{modo jônico} & = E_i^M = \{e_i^M\}_0^6 = \{0,2,4,5,7,9,11\} \\
-\text{modo dórico} & = E_i^{md} = \{e_i^{md}\}_0^6 = \{0,2,3,5,7,9,10\} \\
-\text{modo frígio} & = E_i^{mf} = \{e_i^{mf}\}_0^6 = \{0,1,3,5,7,8,10\} \\
-\text{modo lídio} & = E_i^{ml}=\{e_i^{ml}\}_0^6 = \{0,2,4,6,7,9,11\} \\
-\text{modo mixolídio} & = E_i^{mmi} = \{e_i^{mmi}\}_0^6 = \{0,2,4,5,7,9,10\}
-\end{split}
-\end{equation}
-
-só possuem intervalos maiores, menores e justos. Única exceção para o trítono, que se apresenta como quarta aumenta ou quinta diminuta.
-
-Todas as escalas diatônicas
-seguem o padrão de intervalos sucessivos
-tom, tom, semitom, tom, tom, tom, semitom. De forma
-que pode-se escrever:
-\begin{equation}\label{eq:relacaoDia}
-\begin{split}
-\{d_i\} & =\{2,2,1,2,2,2,1\} \\
-e_0 & =0 \\
-e_i & =d_{(i+\kappa)\%7}+e_{i-1} \quad para \;\;  i > 0
-\end{split}
-\end{equation}
-
-Com $\kappa \in \mathbb{N}$. Para cada
-modo, há um único valor de $\kappa$ em qualquer intervalo $\in [x,x+6], \forall x \in \mathbb{N}$.
-
-Por exemplo, uma breve
-inspeção revela que $e_i^{ml}=d_{(i+2)\%7}+e_{i-1}^{ml}$. Portanto, $\kappa=2$
-para o modo lídio.
-
-A escala menor possui duas formas adicionais, melódica e harmônica:
-
-\begin{equation}\label{eq:escalasMenores}
-\begin{split}
-\text{menor natural (igual acima)} & = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
-\text{menor harmônica} & = E_i^{mh} = \{e_i^{mh}\}_0^6 = \{0,2,3,5,7,8,11\} \\
-\text{menor melódica} & = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
-\end{split}
-\end{equation}
-
-O traçado ascendente e descendente da escala menor melódica é necessário para a presença da sensível (sétimo e último grau separado por um semitom da oitava) no trajeto ascendente, o que não é necessário quando descende, retomando a forma natural.
-Já a escala harmônica apresenta a sensível, mas não evita o intervalo de segunda aumentada entre o sexto e sétimo graus, pois não precisa contemplar trajetória melódica, apenas apresentar a sensível tão cara ao sistema tonal (a sensível tende à tônica, afirmando-a).\cite{Harmonia}
-Outras escalas podem ser representadas da mesma forma, como as pentatônicas e os modos de transposição limitados de Messiaen.\cite{Messiaen}
-
-A ocorrência simultânea de notas é observada através dos acordes.
-Destes, a base na música tonal são as tríades.
-Estas constituem-se de duas terças sucessivas, em 3 notas: fundamental, terça e quinta.
-Os acordes triádicos sem inversão e na forma fechada\footnote{Um acorde invertido é aquele que apresenta no grave outra nota que não a fundamental.
-A posição fechada é aquela em que não cabe nota alguma do acorde entre quaisquer duas notas consecutivas.\cite{Lacerda}},
-com a fundamental em $0$ são escritos;
-
-\begin{equation}\label{triades}
-\begin{split}
-\text{tríade maior} = A_i^M= \{a_i^M\}_0^2=\{0,4,7\} \\ 
-\text{tríade menor} = A_i^m = \{a_i^m\}_0^2=\{0,3,7\} \\
-\text{tríade diminuta} = A_i^d = \{a_i^d\}_0^2=\{0,3,6\} \\
-\text{tríade aumentada} = A_i^a = \{a_i^a\}_0^2=\{0,4,8\}
-\end{split}
-\end{equation}
-
-Basta acrescentar $10$ ao final para ter o acorde (uma tétrade)
-com sétima menor ou $11$ para resultar em sétima maior. As inversões e posições abertas
-podem ser obtidas contando o número de semitons de cada nota a partir da nota mais grave ou somando $12$ à fundamental, terça ou quinta.
+computacionais, pois os intrumentos reais possuem desvios destas frequências para melhor compatibilizar os harmônicos
+de suas notas. Além disso, o fator fixo $\varepsilon=2^{\frac{1}{12}}$ caracteriza
+a afinação de temperamento por igual. Há afinações com intervalos propostos como razões de inteiros de baixa ordem, com
+fundamentação em leis físicas, e foram propostas antes do advento do temperamento por igual.\cite{Roederer}
+
+Duas afinações emblemáticas são:
+\begin{itemize}
+    \item A {\bf intonação justa} consiste em notas da escala diatônica descritas como razão de inteiros de pequena ordem como apontados pela série harmônica.
+As razões básicas estão contidas no modo jônico (dó a dó nas teclas brancas do piano, veja abaixo na subseção~\ref{subsec:escalas}): 1, 9/8, 5/4, 4/3, 3/2, 5/3, 15/8, 2/1.
+Os intervalos são considerados com relação às notas das escala e são usados o semitom 16/15, o 'tom menor' 10/9 e o 'tom maior' 9/8.
+Há diferentes formas de realizar uma divisão de 12 notas.
+    \item A {\bf afinação pitagórica} baseia-se no intervalo 3/2 (quinta justa).
+O modo jônico fica: 1, 9/8, 81/64, 4/3, 3/2, 27/16, 243/128, 2/1.
+Os intervalos são também considerados com relação às notas da escala.
+Além dos intervalos do modo, são usados a segunda menor 256/243, a terça menor 32/27, a quarta aumentada 729/512, a quinta diminuta 1024/729, a sexta menor 128/81 e a sétima menor 16/9.
+\end{itemize}
 
@@
 pode-se usar reais não inteiros
-para descrever a sequência de alturas, ou modificar o fator $\varepsilon$
+para a sequência de alturas, ou modificar o fator $\varepsilon=2^{\frac{1}{12}}$
 e continuar usando inteiros. Por exemplo, uma afinação
@@
 é proposta na forma da divisão da oitava em $53$ notas:
-$\varepsilon=2^{\frac{1}{53}}$.\cite{microtonalidade}
-As notas
-continuam inteiras com o novo $\varepsilon$.
+$\varepsilon_2=2^{\frac{1}{53}}$.\cite{microtonalidade}
+As notas nesta divisão da oitava em $53$ notas se relacionam por inteiros
+com $\varepsilon_2$.
 Note que se $S_i$ é uma sequência de alturas relacionadas por $\varepsilon_1$,
 um mapeamento para notas em $\varepsilon_2$
-resulta em uma nova sequência $S_i'$ com $s_i'=s_i \frac{\varepsilon_1}{\varepsilon_2}$.
+constitui uma nova sequência $S_i'$ com $s_i'=s_i \frac{\varepsilon_1}{\varepsilon_2}$.
 A montagem musical \emph{Micro tom} explora recursos microtonais e seu código está no Apêndice~\ref{ap:micro}, assim como na \massa\ online.
+
+\subsection{Intervalos}\label{subsec:intervalos}
+Em proporções de $\varepsilon=2^{\frac{1}{12}}$ entre as frequências das notas (i.e.
+um semitom), os intervalos do sistema de 12 notas são representados por inteiros.
+A tabela~\ref{eq:intervalos} resume as características de cada intervalo: sua nomenclatura tradicional, características de consonância e dissonância e número de semitons de cada um.
+
+\begin{table}[htpq!]
+\centering
+\caption{Intervalos musicais, suas notações tradicionais, classificações básicas de dissonância e número de semitons.
+As consonâncias perfeitas são os uníssonos, as quintas e as oitavas justas (J). As consonâncias imperfeitas são
+as terças e as sextas maiores (M) e menores (m). As dissonâncias fortes são as segundas menores e sétimas maiores. As dissonâncias
+brandas são as segundas maiores e as sétimas menores. O primeiro caso especial consite na quarta justa, que é consonante perfeita
+se considerada uma inversão da quinta justa, caso contrário pode ser considerada uma dissonância ou uma consonância imperfeita.
+O segundo caso especial é o trítono (4aum, 5dim, tri).
+Este é consonante em algumas culturas, já para a música tonal, o trítono indica dominante e busca fundamentalmente sua resolução em uma terça ou sexta e, por esta instabilidade, é considerado intervalo dissonante.}
+\begin{tabular}{| c | c | c | }\hline
+    \multicolumn{3}{|c|}{\bf consonâncias}  \\\hline
+   & notação tradicional & número de semitons \\
+   perfeitas: & 1J, 5J, 8J & 0, 7, 12 \\
+ imperfeitas: & 3m, 3M, 6m, 6M & 3, 4, 8, 9 \\\hline\hline
+    \multicolumn{3}{|c|}{\bf dissonâncias} \\\hline
+ & notação tradicional & número de semitons \\
+ fortes: & 2m, 7M & 1, 11 \\
+ brandas: & 2M, 7m & 2, 10 \\\hline\hline
+    \multicolumn{3}{|c|}{\bf casos especiais} \\\hline
+ & notação tradicional & número de semitons \\
+ consonante ou dissonante: & 4J & 5 \\
+ dissonante na tradição ocidental: & trítono, 4aum, 5dim & 6 \\\hline
+\end{tabular}\label{eq:intervalos}
+\end{table}
+
+A nomenclatura, com base em imposições e conveniências do sistema tonal, e de aspectos práticos da manipulação de notas, pode ser especificada assim:\cite{Roederer,Wisnick}
+\begin{itemize}
+        \item Intervalos por número de grados entre as notas: primeira (uníssono), segunda, terça, quarta, quinta, sexta, sétima, oitava.
+Nona, décima, décima primeira, etc, são os intervalos compostos, de uma ou mais oitavas + um intervalo dentro da oitava, que caracteriza o intervalo composto.
+São representados pelos dígitos numéricos: 1, 3, 5 é um uníssono, uma terça, e uma quinta.
+        \item Qualidades de cada intervalo: as consonâncias perfeitas - i.e.
+uníssono, quarta, quinta e oitava - são 'justas'.
+As consonâncias imperfeitas - i.e.
+terças e sextas - e as dissonâncias - i.e.
+segundas e sétimas - podem ser maiores ou menores.
+Excessão para o trítono.
+        \item A quarta justa é tida como consonante perfeita ou dissonante de acordo com o contexto e arcabouço teórico.
+Como regra geral, pode ser considerada consonante salvo casos em que é usada de passagem para uma quinta ou terça como resolução.
+        \item O trítono é dissonante na música ocidental por caracterizar a "dominante" no sistema tonal (veja abaixo na subseção~\ref{subsec:harmonia}) e representar instabilidade.
+Algumas culturas entoam o intervalo como consonante.
+        \item Um intervalo maior, descrescido de um semitom, resulta em um intervalo menor. Um intervalo menor, acrescido de um semitom, resulta em um intervalo maior.
+        \item Um intervalo justo (uníssono, quarta justa, quinta justa, oitava justa), ou um intervalo maior (segunda maior 2M, terça maior 3M, sexta maior 6M ou sétima maior 7M), acrescido de um semitom, resulta em um intervalo aumentado (p.ex.
+terça aumentada 3aum com cinco semitons).
+A quarta aumentada é também chamada de trítono (4aum ~ tri).
+        \item Um intervalo justo, ou um intervalo menor (segunda menor 2m, terça menor 3m, sexta menor 6m ou sétima menor 7m), decrescido de um semitom, resulta em um intervalo diminuto.
+A quinta diminuta é também chamda de trítono (5dim ~ tri).
+        \item Um intervalo aumentado, acrescido de um semitom, resulta em um intervalo 'mais que aumentado' e um intervalo diminuto, descrescido de um semitom, resulta em um intervalo 'mais que diminuto'.
+        \item Caso as notas soem simultaneamente, o intervalo é harmônico.
+        \item Caso as notas soem em sequência no tempo, o intervalo é melódico.
+A ordem das notas, primeiro a nota mais grave ou a mais aguda, resulta em um intervalo ascendente ou descendente, respectivamente.
+        \item Passando a nota mais grave para a oitava acima, ou a nota mais aguda uma oitava para baixo, o intervalo é invertido.
+Um intervalo, somado à sua inversão, resulta 9 (7m inverte para 2M: $7m+2M=9-$).
+Um intervalo maior invertido resulta em um intervalo menor e vice-versa.
+Um intervalo aumentado invertido resulta diminuto e vice-versa, assim como o mais que aumentado resulta em mais que diminuto e vice-versa.
+Um intervalo justo invertido resulta igualmente justo.
+        \item Um intervalo maior que a oitava é dito 'intervalo composto' e é classificado como o intervalo entre as mesmas notas, mas na mesma oitava.
+São também especificados por 7 acrescido deste intervalo: 11J é uma oitava mais uma quarta (7+4J==11J), 9M é uma oitava mais uma segunda (7+2M=9M).
+\end{itemize}
+
+Os intervalos aumentados/diminutos e mais que aumentados/diminutos são consequências do sistema tonal.
+Os graus da escala (veja abaixo na subseção~\ref{subsec:escalas}) são realmente notas diferentes, com funções e usos específicos.
+Assim, em uma escala de dó bemol maior, a tônica - primeiro grau - é dó bemol, não si, e a sensível - sétimo grau - é si bemol, não lá sustenido ou dó duplo bemol.
+De forma semelhante, o segundo grau de uma escala pode estar a um semitom do primeiro grau, assim como a sensível (sétimo grau a um semitom ascendente do primeiro), momento no qual há uma terça diminuta entre os dois semitons do sétimo e segundo grau da escala pois o primeiro grau está entre os dois graus próximos a ele: segundo e sensível.\cite{Lacerda}
+
+Esta descrição resume a teoria tradicional dos intervalos musicais.\cite{Lacerda} A montagem \emph{Intervalos entre alturas} explora estes intervalos de formas isoladas e diversas.
+O código está no Apêndice~\ref{ap:intervalos} e disponível online com a \emph{toolbox} \massa.\cite{MASSA}
+
+\subsection{Escalas}\label{subsec:escalas}
+Na escala cromática
+de temperamento igual, existem 5 divisões perfeitamente simétricas, consideradas
+escalas pelas simetrias internas e usos fáceis que disso resultam. Como sequências de inteiros aos quais $\varepsilon$ é elevado
+para multiplicar $f_0$, as escalas são:
+
+\begin{equation}\label{escSim}
+\begin{split}
+\text{cromática} & = E_i^c = \{e_i^c\}_0^{11} =  \{0,1,2,3,4,5,6,7,8,9,10,11\} = \{i\}_0^{11}\\
+\text{tons inteiros} & = E_i^t = \{e_i^t\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
+\text{terças menores} & = E_i^{tm} = \{e_i^{tm}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
+\text{terças maiores} & = E_i^{tM} = \{e_i^{tM}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
+\text{trítonos} & = E_i^{tt} = \{e_i^{tt}\}_0^{1} = \{ 0, 6 \} = \{6.i\}_0^1
+\end{split}
+\end{equation}
+
+Assim, a terceira nota da escala em tons inteiros com $f_0=200Hz$
+é $f_3=\varepsilon^{e_3^t} . f_0 = 2^{\frac{6}{12}} . 200 \approxeq 282.843 Hz$. Estas
+'escalas' ou padrões, geram estruturas estáveis pelas simetrias internas, e podem ser
+repetidas de forma eficiente e sustentada. Abaixo há mais sobre simetrias e
+sobre estas divisões da oitava por igual.
+A montagem \emph{Cristais} expõe cada uma destas escalas tanto melódica quanto harmonicamente e seu código está no Apêndice~\ref{ap:cristais}.
+Como parte da \massa, este código é disponibilizado online.
+
+As escalas diatônicas:
+
+\begin{equation}\label{eq:escalas}
+\begin{split}
+\text{menor natural} = \text{modo eólico} & = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
+\text{modo lócrio} & = E_i^{mlo} = \{e_i^{mlo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
+\text{maior}  = \text{modo jônico} & = E_i^M = \{e_i^M\}_0^6 = \{0,2,4,5,7,9,11\} \\
+\text{modo dórico} & = E_i^{md} = \{e_i^{md}\}_0^6 = \{0,2,3,5,7,9,10\} \\
+\text{modo frígio} & = E_i^{mf} = \{e_i^{mf}\}_0^6 = \{0,1,3,5,7,8,10\} \\
+\text{modo lídio} & = E_i^{ml}=\{e_i^{ml}\}_0^6 = \{0,2,4,6,7,9,11\} \\
+\text{modo mixolídio} & = E_i^{mmi} = \{e_i^{mmi}\}_0^6 = \{0,2,4,5,7,9,10\}
+\end{split}
+\end{equation}
+
+só possuem intervalos maiores, menores e justos. Única exceção para o trítono, que se apresenta como quarta aumenta ou quinta diminuta.
+
+Todas as escalas diatônicas
+seguem o padrão de intervalos sucessivos
+tom, tom, semitom, tom, tom, tom, semitom. De forma
+que pode-se escrever:
+\begin{equation}\label{eq:relacaoDia}
+\begin{split}
+\{d_i\} & =\{2,2,1,2,2,2,1\} \\
+e_0 & =0 \\
+e_i & =d_{(i+\kappa)\%7}+e_{i-1} \quad para \;\;  i > 0
+\end{split}
+\end{equation}
+
+Com $\kappa \in \mathbb{N}$. Para cada
+modo, há um único valor de $\kappa$ em qualquer intervalo $\in [x,x+6], \forall x \in \mathbb{N}$.
+
+Por exemplo, uma breve
+inspeção revela que $e_i^{ml}=d_{(i+2)\%7}+e_{i-1}^{ml}$. Portanto, $\kappa=2$
+para o modo lídio.
+
+A escala menor possui duas formas adicionais, melódica e harmônica:
+
+\begin{equation}\label{eq:escalasMenores}
+\begin{split}
+\text{menor natural (igual acima)} & = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
+\text{menor harmônica} & = E_i^{mh} = \{e_i^{mh}\}_0^6 = \{0,2,3,5,7,8,11\} \\
+\text{menor melódica} & = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
+\end{split}
+\end{equation}
+
+O traçado ascendente e descendente da escala menor melódica é necessário para a presença da sensível (sétimo e último grau separado por um semitom da oitava) no trajeto ascendente, o que não é necessário quando descende, retomando a forma natural.
+Já a escala harmônica apresenta a sensível, mas não evita o intervalo de segunda aumentada entre o sexto e sétimo graus, pois não precisa contemplar trajetória melódica, apenas apresentar a sensível tão cara ao sistema tonal (a sensível tende à tônica, afirmando-a).\cite{Harmonia}
+Outras escalas podem ser representadas da mesma forma, como as pentatônicas e os modos de transposição limitados de Messiaen.\cite{Messiaen}
+
+\subsection{Acordes}\label{subsec:acordes}
+A ocorrência simultânea de notas é observada através dos acordes.
+Destes, a base na música tonal são as tríades.
+Estas constituem-se de duas terças sucessivas, em 3 notas: fundamental, terça e quinta.
+Os acordes triádicos sem inversão e na forma fechada\footnote{Um acorde invertido é aquele que apresenta no grave outra nota que não a fundamental.
+A posição fechada é aquela em que não cabe nota alguma do acorde entre quaisquer duas notas consecutivas.\cite{Lacerda}},
+com a fundamental em $0$ são escritos;
+
+\begin{equation}\label{triades}
+\begin{split}
+\text{tríade maior} = A_i^M= \{a_i^M\}_0^2=\{0,4,7\} \\ 
+\text{tríade menor} = A_i^m = \{a_i^m\}_0^2=\{0,3,7\} \\
+\text{tríade diminuta} = A_i^d = \{a_i^d\}_0^2=\{0,3,6\} \\
+\text{tríade aumentada} = A_i^a = \{a_i^a\}_0^2=\{0,4,8\}
+\end{split}
+\end{equation}
+
+Basta acrescentar $10$ ao final para ter o acorde (uma tétrade)
+com sétima menor ou $11$ para resultar em sétima maior. As inversões e posições abertas
+podem ser obtidas contando o número de semitons de cada nota a partir da nota mais grave ou somando $12$ à fundamental, terça ou quinta.
 

### 2013-03-21 devEresultados.tex
adiantado bastante a parte de harmonia

@@
 \subsection{Escalas}\label{subsec:escalas}
-Na escala cromática
-de temperamento igual, existem 5 divisões perfeitamente simétricas, consideradas
-escalas pelas simetrias internas e usos fáceis que disso resultam. Como sequências de inteiros aos quais $\varepsilon$ é elevado
+Uma escala é um conjunto ordenado de alturas. Usualmente, as escalas se repetem
+a cada oitava. A sequências ascendente com todas as notas da divisão da oitava em 12 intervalos iguais
+separados pela razão $\varepsilon=2^{\frac{1}{12}}$ é a escala cromática de temperamento igual.
+Há 5 divisões perfeitamente simétricas da oitava dentro da escala cromática.
+Estas divisões são consideradas
+escalas pelos usos fáceis e peculiares que disso provém. Como inteiros aos quais $\varepsilon=2^{\frac{1}{12}}$ é elevado
 para multiplicar $f_0$, as escalas são:
@@
 'escalas' ou padrões, geram estruturas estáveis pelas simetrias internas, e podem ser
-repetidas de forma eficiente e sustentada. Abaixo há mais sobre simetrias e
-sobre estas divisões da oitava por igual.
+repetidas de forma eficiente e sustentada.
+Abaixo, na seção~\ref{estCic}.
 A montagem \emph{Cristais} expõe cada uma destas escalas tanto melódica quanto harmonicamente e seu código está no Apêndice~\ref{ap:cristais}.
@@
 
-só possuem intervalos maiores, menores e justos. Única exceção para o trítono, que se apresenta como quarta aumenta ou quinta diminuta.
+possuem apenas intervalos maiores, menores e justos. Única exceção para o trítono, que se apresenta como quarta aumenta ou quinta diminuta.
 
@@
 seguem o padrão de intervalos sucessivos
-tom, tom, semitom, tom, tom, tom, semitom. De forma
-que pode-se escrever:
+tom, tom, semitom, tom, tom, tom, semitom, e
+pode-se escrever:
 \begin{equation}\label{eq:relacaoDia}
@@
 Com $\kappa \in \mathbb{N}$. Para cada
-modo, há um único valor de $\kappa$ em qualquer intervalo $\in [x,x+6], \forall x \in \mathbb{N}$.
-
+modo, há um único valor de $\kappa \in [0,6]$.
 Por exemplo, uma breve
@@
 
-O traçado ascendente e descendente da escala menor melódica é necessário para a presença da sensível (sétimo e último grau separado por um semitom da oitava) no trajeto ascendente, o que não é necessário quando descende, retomando a forma natural.
+O traçado ascendente e descendente da escala menor melódica é necessário para a presença da sensível (sétimo e último grau separado por um semitom da oitava e realça a polarização na tônica) no trajeto ascendente, o que não é necessário quando descende, retomando a forma natural.
 Já a escala harmônica apresenta a sensível, mas não evita o intervalo de segunda aumentada entre o sexto e sétimo graus, pois não precisa contemplar trajetória melódica, apenas apresentar a sensível tão cara ao sistema tonal (a sensível tende à tônica, afirmando-a).\cite{Harmonia}
@@
 Estas constituem-se de duas terças sucessivas, em 3 notas: fundamental, terça e quinta.
-Os acordes triádicos sem inversão e na forma fechada\footnote{Um acorde invertido é aquele que apresenta no grave outra nota que não a fundamental.
-A posição fechada é aquela em que não cabe nota alguma do acorde entre quaisquer duas notas consecutivas.\cite{Lacerda}},
-com a fundamental em $0$ são escritos;
+Um acorde invertido é aquele que apresenta no grave outra nota que não a fundamental.
+A posição fechada é aquela em que não cabe nota alguma do acorde entre quaisquer duas notas consecutivas.\cite{Lacerda}
+As tríades, sem inversão, na forma fechada e com a fundamental em $0$, são:
 
@@
 
-Basta acrescentar $10$ ao final para ter o acorde (uma tétrade)
-com sétima menor ou $11$ para resultar em sétima maior. As inversões e posições abertas
-podem ser obtidas contando o número de semitons de cada nota a partir da nota mais grave ou somando $12$ à fundamental, terça ou quinta.
-
-\subsection{Rudimentos de harmonia}\label{subsec:harmonia}
-
-Acordes como os listados em~\ref{triades} formados em cada
-grau das escalas em~\ref{eq:escalas} formam campos harmônicos
-das escalas. As progressões de acordes utilizadas e regras para seus encadeamentos é o objeto de estudo da harmonia musical.
-Mesmo uma melodia monofônica gera campos harmônicos.
-
-Não respeitar 
-os encadeamentos básicos do sistema tonal resulta
-em harmonias modal e atonal. Caso
-as notas coincidam com algum modo diatônico~\ref{eq:escalas}
-ou sejam em número pequeno, pode-se dizer que a
-harmonia é modal. Caso contrário, atonal.\cite{harmEXT}.
-
-Na harmonia tonal tradicional,
-as escalas maior e menor são
-base para tríades em cada grau 
+Para considerar outra terça sobreposta à quinta, basta acrescentar $10$ ao final para a tétrade
+com sétima menor ou $11$ para a tétrade com sétima maior. As inversões e posições abertas
+podem ser obtidas com a adição seletiva de $12$ às componentes.
+
+Acordes triádicos incompletos, com notas adicionais (acordes 'sujos'), e não triádicos são comuns.
+Orientações gerais são:
+\begin{itemize}
+    \item Uma quinta constitui uma fundamental confirmada pelo intervalo.
+    \item A terça maior ou menor aponta a qualidade maior ou menor do acorde.
+    \item Todo trítono, especialmente se formado entre uma terça maior e uma sétima menor, tende a resolver em uma terça ou uma sexta.
+    \item Na repetição, a ordem de preferência é: a fundamental, a quinta, a terça e a sétima.
+    \item Pode-se formar acordes com notas diferentes das triádicas, especialmente se dentro de alguma lógica recorrente ou encadeamento musical que justifique as notas diferentes.
+    \item Acordes formados por sucessões de intervalos diferentes da terças, como as quartas ou as segundas, são recorrentes dentre composições de tonalismo avançado ou experimentais.
+    \item A repetição de sucessões de acordes (ou suas características) fixa uma trajetória pela recorrência e permite a introdução de formações exóticas sem que haja incoerência.
+\end{itemize}
+
+\subsection{Harmonia}\label{subsec:harmonia}
+
+Omitir os encadeamentos básicos do sistema tonal é a chave para a obtenção de 
+harmonias modal e atonal. No caso desta ausência de estruturas tonais mínimas,
+se as notas coincidirem com alguma escala diatônica (veja as equações ~\ref{eq:escalas})
+ou forem em número pequeno, pode-se dizer que a
+harmonia é modal. Caso encadeamentos tonais básicos estejam ausentes, a notas não coincidam
+com alguma das escalas diatônicas e forem diversos e dissonantes o suficiente para evitar redução por
+polarizações, a harmonia é atonal.
+Nesta classificação, a harmonia modal não é tonal e não é atonal.
+A harmonia modal está reduzida à incidência de notas dentro de escalas diatônicas.
+Pode-se perceber, pelo conceito, que a harmonia atonal é difícil de ser realizada.\cite{harmEXT}
+
+De fato, as técnicas de música atonal
+visam estruturas que evitam o vínculo da audição a modos e relações tonais. A
+dificuldade é tamanha que o dodecafonismo surgiu. A proposta do dodecafonismo é
+usar um conjunto de notas, idealmente usa-se as 12 notas, e executar uma a uma destas notas
+na mesma ordem. Neste contexto, a tônica fica difícil de se estabelecer. Mesmo assim, a audição ocidental
+procura traços tonais nas música e facilmente os encontra por caminhos inesperados e por vezes tortuosos.
+A utilização de
+intervalos dissonantes, especialmente trítonos sem resoluções, nonas, segundas e sétimas, reforça
+a ausência de tonalidade. Neste contexto, para criação da peça, pode-se:
+\begin{itemize}
+    \item Repetir notas, como artifício de sua incidência anterior, não adiciona informação relevante.
+    \item Tocar notas adjacentes ao mesmo tempo, em bloco.
+    \item As durações podem ser livres, desde que respeitada ordem de aparição das notas.
+    \item Para variação, além dos recursos de ampliação, transposição e translação, são usados o inverso, o retrógrado e o retrógrado do inverso.
+Veja nas subseções~\ref{subsec:motivos} e~\ref{subsec:usosmusicais3} para maiores detalhes.
+    \item Variações de intrumentação, articulação, espacialização e outras possibilidades de apresentação da estrutura de notas.
+\end{itemize}
+
+A harmonia atonal pode ser observada, de forma paradigmática, dentro destas condições.
+Vale apontar que boa parte do que escreveram os compositores
+emblemáticos para o dodecafonismo, como Alban Berg e o próprio Schoenberg, consiste
+aplicações parciais e livres destas técnicas. Várias peças fazem até mesmo pontes
+propositais entre técnicas tonais e atonais.
+
+No século XX, músicas rítmicas
+e com ênfase em sonoridades/timbres ampliaram 
+as concepções de tonalidade e harmonia. Ainda assim, a harmonia tonal tem forte presença
+nas vertentes artísticas e comerciais. O próprio dodecafonismo considera-se de natureza
+tonal pois consiste na negação de suas características de polarização.
+
+Na música tonal ou modal, acordes, como os listados nas equações~\ref{triades}, constituídos com a fundamental em cada
+grau de escalas especificadas nas equações~\ref{eq:escalas}, formam os pilares do campo harmônico.
+A observação de progressões incidentes de acordes e regras para encadeamentos é o objeto de estudo da harmonia musical.
+Mesmo uma melodia monofônica gera campos harmônicos e pode-se observar acordes sugeridos por cada passagem.
+
+Na 'harmonia tonal tradicional',
+a escala pode ter como tônica (primeiro grau da escala) qualquer nota e pode ser maior (com as mesmas notas do modo jônico) ou menor (as notas do eólico constituem a escala "menor natural" e esta possui versões harmônica e melódica, veja a equação~\ref{eq:escalasMenores}.
+A escala escolhida é
+base para tríades, cada uma com a fundamental em um grau 
 da escala: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$.
+Para a formação do acorde, são consideradas a terceira nota e a quinta nota acima da fundamental, considerando a própria nota como a primeira e utilizando as notas da escala.
 Pode-se anotar $\hat{1},\hat{3},\hat{5}$ para o acorde do primeiro grau, formado pelo primeiro grau da escala e central para uma música tonal.
-Secundários, o acorde do quinto grau $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sustenido no caso da escala menor) e o acorde do quarto grau $\hat{4},\hat{6},\hat{1}$.
+Secundários são os acordes do quinto grau $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sustenido no caso da escala menor) e do quarto grau $\hat{4},\hat{6},\hat{1}$.
 Depois são considerados os outros graus.
-As convenções e técnicas apresentadas ao longo da história
-são objeto de estudo da harmonia tradicional.\cite{Harmonia}
-
-A chamada harmonia funcional atribui funções a estes três acordes centrais e busca compreender seus usos através destas funções.
-O acorde formado sobre o primeiro grau é o acorde de tônica e tem a função de manter um centro, um "chão" na música.
-O acorde formado sobre o quinto grau é a dominante e tem a função de tender à tônica, direcionar a música para ela.
-A tríade formada sobre o quarto grau é a subdominante e tem a função de se distanciar da tônica.
+A harmonia tradicional consiste em convenções e técnicas estilísticas de encadeamento destes acordes, formados em cada grau da escala.\cite{Harmonia}
+
+A 'harmonia funcional' atribui funções a estes três acordes centrais e busca compreender seus usos através destas funções.
+O acorde formado sobre o primeiro grau é o acorde de tônica (T ou t se tônica maior ou menor) e tem a função de manter um centro, um "chão" na música.
+O acorde formado sobre o quinto grau é a dominante (D, a dominante é sempre maior) e tem a função de tender à tônica, direcionar a música para ela.
+A tríade formada sobre o quarto grau é a subdominante (S ou s se subdominante maior ou menor) e tem a função de distanciar a música da tônica.
 O sistema se baseia em afirmar a tônica através de encadeamentos tônica-dominante-tônica expandidos com outros acordes das formas mais diversas.
@@
 A estes três acordes, são associadas as outras tríades.
-Na escala maior, a associada relativa (tônica relativa, subdominante relativa e dominante relativa) é a tríade formada uma terça abaixo e a associada anti-relativa (tônica anti-relativa, subdominante anti-relativa e a dominante anti-relativa) é a tríade formada na terça acima.
-Na escala menor ocorre o mesmo, mas a tríade a uma terça abaixo é chamada anti-relativa e a tríade a uma terça acima é chamada de relativa.
+Na escala maior, a associada relativa (tônica relativa Tr, subdominante relativa Sr e dominante relativa Dr) é a tríade formada uma terça abaixo e a associada anti-relativa (tônica anti-relativa Ta, subdominante anti-relativa Sa e a dominante anti-relativa Da) é a tríade formada na terça acima.
+Na escala menor ocorre o mesmo, mas a tríade a uma terça abaixo é chamada anti-relativa (tA, sA) e a tríade a uma terça acima é chamada de relativa (tR, sR).
 As exatas funções e efeitos musicais destes acordes é motivo de bastante controvérsia.
@@
 A tônica é o centro da música, a dominante tende à tônica e a subdominante se distancia da tônica.
-Os três acordes podem, a princípio, serem substituídos livremente pela relativa ou antirelativa.}
+Os três acordes podem, a princípio, serem substituídos livremente pelas respectivas relativas ou anti-relativas.}
 \begin{tabular}{l | c | r}
-relativa & acorde principal da função & antirelativa \\\hline\hline
+relativa & acorde principal da função & anti-relativa \\\hline\hline
 $\hat{6},\hat{1},\hat{3}$ & tônica:       $\hat{1},\hat{3},\hat{5}$ & $\hat{3}, \hat{5},      \hat{7}$ \\
@@
 
-Na escala menor, a relativa é a tríade iniciada uma terça acima e a antirelativa uma terça abaixo,
+Na escala menor, a relativa é a tríade iniciada uma terça acima e a anti-relativa uma terça abaixo,
 ou seja, se invertem com relação à tabela~\ref{tab:harmonia}. Outro detalhe destas tríades em modo
@@
 Assim, a dominante é sempre maior, tanto em escalas maior ou menor.
-Por este motivo, a dominante relativa permanece a uma terça abaixo e a antirelativa a uma terça acima na escala menor.
-
-Cada um destes acordes pode ser confirmado e se desenvolver com uma dominante e uma subdominante individuais (acorde com base na tríade formada a uma quinta acima e uma quinta abaixo respectivamente).
+Por este motivo, mesmo em tom menor, a dominante relativa permanece a uma terça abaixo e a antirelativa a uma terça acima.
+
+Cada um destes acordes pode ser confirmado e se desenvolver com uma execução de sua dominante ou subdominante individual (acorde com base na tríade formada a uma quinta acima e uma quinta abaixo respectivamente).
+Estas dominantes/subdominantes indiduais, por sua vez, possuem também subdominantes e dominantes individuais passíveis de uso.
+Assim, em uma dada tonalidade, pode ocorrer qualquer acorde, por mais distante que seja do campo harmônico e das notas da escala, desde que a ocorrência apresente um percurso coerente de dominantes e subdominantes até a tonalidade de origem.
 
@@
 
+\subsection{Motivos e unidades maiores}\label{subsec:motivos}
+
 \subsection{Estruturas direcionais}\label{subsec:dir}
@@
 
-\subsection{Usos musicais}
+\subsection{Usos musicais}\label{subsec:usosmusicais3}
 

### 2013-03-21 devEresultados.tex
terminada parte de contraponto e ritmo

@@
 
-As medianas\footnote{Também chamadas de medianas cromáticas.} são acordes formados também a uma terça, mas com alterações com relação ao acorde de origem.
+As medianas, ou 'medianas cromáticas', são duas para cada acorde: a mediana cromática superior, formada com a fundamental na terça do acorde original, e a inferior, formada com a quinta na terça do acorde original.
+São acordes formados também a uma terça, mas com uma alteração cromática com relação ao acorde de origem.
 Caso haja duas alterações cromáticas, i.e.
 duas notas alteradas por um semitom cada com relação ao acorde original, a mediana é chamada 'duplamente cromática'.
-Esta relação entre acordes é considerada de tonalismo avançado, por vezes até de dissolução do tonalismo, e tem efeitos fortes e marcantes embora perfeitamente consonantes.
-As medianas foram utilizadas a partir do final do romantismo por Wagner, Lizt, Richard Strauss dentre outros e é bastante simples de ser realizada.\cite{Harmonia,Salzer}
-
-A modulação é a troca de tonalidade e se dão principalmente através de: uma dominante individual que afirma a mudança da tônica; alterações cromáticas para um acorde da nova tonalidade; e 'enarmonia' que é a mudança de função de um acorde com as mesmas notas.
-A importância da dominante torna ela o pivô natural das modulações, resultando no círculo das quintas.\cite{Harmonia,Salzer,Koellheuteur,Harmony}
-
-A montagem musical \emph{Acorde cedo} explora estas relações entre acordes. Seu código está disponível do Apêndice~\ref{ap:acorde} e online como parte da \massa.\cite{MASSA}
+Também são duas para cada acorde: a superior, com terça na quinta do acorde original, e a inferior, com terça na fundamental da tríade original.
+Observe que um acorde maior possui medianas maiores e medianas duplamente cromáticas maiores.
+Um acorde menor possui medianas menores e medianas duplamente cromáticas menores.
+Esta relação entre acordes é considerada de tonalismo avançado, por vezes até de expansão e dissolução do tonalismo, e tem efeitos fortes e marcantes embora perfeitamente consonantes.
+As medianas foram utilizadas a partir do final do romantismo por Wagner, Lizt, Richard Strauss dentre outros e são bastante simples de serem realizadas.\cite{Harmonia,Salzer}
+
+A modulação é a mudança da tonalidade em que se encontra a musica. Caracteriza-se a modulação através da observação
+das tonalidades de partida e chegada e da forma de transição. As tonalidades são sempre relacionadas por quintas e suas relativas e antirelativas. São formas de efetuar a modulação:
+\begin{itemize}
+    \item A transposição do discurso para a nova tonalidade, sem preparação alguma.
+É procedimento típico do barroco embora seja incidente em outros períodos.
+Por vezes chamada de modulação frasal ou modulação sem preparação.
+    \item O uso cuidadoso de uma dominante individual, e possivelmente também a subdominante individual, para afirmar a mudança da tônica e campo harmônica.
+    \item Uso de alterações cromáticas para um acorde da nova tonalidade. Chamada de modulação cromática.
+    \item O destaque para uma única nota, possivelmente repetida ou suspensa sem acompanhamento, comum às tonalidades de saída e chegada, constitui uma forma peculiar de introduzir o novo campo harmônico.
+    \item Mudança da função, sem a modificação das notas em si, de um acorde para contemplar nova tonalidade. Procedimento chamado de enarmonia.
+    \item A manutenção do centro tonal e mudança da qualidade maior para menor (ou vice-versa) da tonalidade é a modulação paralela. A tonalidade de mesma tônica e outra qualidade é chamada homônima.
+\end{itemize}
+
+A importância da dominante torna ela pivô natural das modulações, o que desemboca no círculo das quintas.\cite{Harmonia,Salzer,Koellheuteur,Harmony} A montagem musical \emph{Acorde cedo} explora estas relações entre acordes.
+Seu código está disponível do Apêndice~\ref{ap:acorde} e online como parte da \massa.\cite{MASSA}
 
@@
 é matéria do campo de estudo chamado contraponto. A bibliografia
-geralmente percorre formas sistemáticas de condução de vozes e desemboca em gêneros escolásticos como cânones, invenções e fugas.
+percorre formas sistemáticas de condução de vozes e desemboca em gêneros escolásticos como cânones, invenções e fugas.
 É possível resumir regras principais do contraponto e é dito que o próprio Beethoven - dentre outros - esboçou uma síntese deste tipo.
@@
         \includegraphics[width=\textwidth]{figuras/movContraponto}
-    \caption{Movimentos diferenciados pelo contraponto. Visando preservar a independência entre as vozes, são distinguidos 3 tipos de movimentos.}
+    \caption{Movimentos diferenciados pelo contraponto com vistas a preservar a independência entre as vozes.
+3 tipos de movimentos: direto, contrário e oblíquo, categorizam as possibilidades.
+O movimento paralelo é um tipo de movimento direto.}
         \label{fig:movContraponto}
@@
 de forma que soem independentes. Cruciais para isso
-são as movimentações relativas,
+são as movimentações relativas, das vozes duas a duas,
 categorizadas em: movimento direto, oblíquo e contrário
@@
 ser não admitidas ou usadas seguidas 
-e precedidas de consonâncias em graus conjuntos.
+e precedidas de consonâncias em graus conjuntos, i.e. notas vizinhas na escala.
 Os movimentos que levam a nota a uma vizinha soam mais coerentes.
 Na presença de 3 ou mais vozes,
-a importância melódica recai sobre as vozes mais aguda e mais grave.\cite{Fux,Tragtenberg,SchoenbergContra}
+a importância melódica recai sobre as vozes mais aguda e mais grave, neta ordem.\cite{Fux,Tragtenberg,SchoenbergContra}
 
@@
 \subsection{Ritmo}\label{subsec:ritmo}
-A noção rítmica é depende de eventos separados por durações.\cite{Lacerda} Estes eventos podem ser ouvidos individualmente se separados
+A noção rítmica é dependente de eventos separados por durações.\cite{Lacerda} Estes eventos podem ser ouvidos individualmente se separados
 por ao menos $50-63ms$. Para que a separação entre eles possa ser apreciada como duração, ela deve ser maior,
@@
 costuma-se associar esta gama de frequências de pulsação às durações entre batidas 
-do coração ou entre os passos ao caminhar.\cite{Lacerda,Roederer}
+do coração, da inspiração/expiração ou entre os passos ao caminhar.\cite{Lacerda,Roederer}
 
@@
 as divisões do pulso musical e seus agrupamentos
-sequenciais no tempo ocorrem com fatores 2, 4 e 8, depois 3, 6 (dois grupos de 3 ou 3 grupos de 2) e 9 e 12 (3 e 4 grupos de 3). Por 
+sequenciais no tempo são: 2, 4 e 8, depois 3, 6 (dois grupos de 3 ou 3 grupos de 2) e 9 e 12 (3 e 4 grupos de 3). Por 
 último os primos 5 e 7, completando 1-9 e 12.
-Outras métricas são incidentes (divisões ou agrupamentos em 13, 17, etc), embora menos usuais e principalmente em contextos de música experimental e erudita do século XX e XXI.
+Outras métricas são menos usuais, como divisões ou agrupamentos em 13, 17, etc, e são incidentes principalmente em contextos de música experimental e erudita do século XX e XXI.
 Por mais complexas que pareçam, as métricas costumam ser composições e decomposições de 1-9 partes iguais.\cite{Gramani,Roederer}.}.
@@
 As relações duais (compassos simples e divisões binárias) costumam ocorrer em ritmos de dança
-e ocasiões festivas, e são chamadas imperfeitas. As relações triádicas
+e ocasiões festivas, e são chamadas imperfeitas. As relações ternárias
 incidem mais na música ritualística e relacionada ao sagrado
@@
 pulsos no nível $j$ de agrupamento:
-$P^0_{10}$ é o décimo pulso, $P^{1}_3$ é a terceira unidade de agrupamento de pulsos (possivelmente terceiro compasso),
+$P^0_{10}$ é o décimo pulso, $P^{1}_3$ é a terceira unidade de agrupamento de pulsos (é possivel que seja o terceiro compasso),
 $P^{-1}_2$ é a segunda unidade da subdivisão do pulso.
@@
 os limites de $j$: as divisões do pulso são durações apreciáveis
-como ritmo; as junções do pulso somam, no máximo
-do escopo, uma música ou um conjunto coeso de músicas. Dito de outra forma: a duração de $P^{min(j)}_i$, $\forall \; i$
-é maior que $50ms$ e as durações de $P^{\text{máx}(j)}_i$ \emph{somadas}  em $\forall \; i$
-devem resultar em uma duração menor do que alguns minutos ou, no máximo, poucas horas.
-
-Cada nível $j$ possui alguns índices $i$. Sempre que $i$ soma três 
+como ritmo; além disso, as junções do pulso somam, no máximo
+do escopo, uma música ou um conjunto coeso de músicas. Dito de outra forma: a duração de $P^{min(j)}_i$, $\forall \; i$,
+é maior que $50ms$ e as durações somadas $\sum_{\forall i}P^{\text{máx}(j)}_i$
+é menor do que alguns minutos ou, no máximo, poucas horas.
+
+Cada nível $j$ possui alguns índices $i$. Sempre que $i$ possuir três valores diferentes
 (ou múltiplo de três) índices há uma relação perfeita, 
-quando soma dois, quatro ou oito índices há uma relação imperfeita, como na figura~\ref{fig:pulsoSubAgl}.
+quando é somente múltiplo dois, quatro ou oito índices há uma relação imperfeita, como na figura~\ref{fig:pulsoSubAgl}.
 

### 2013-03-22 devEresultados.tex
terminada revisao da parte 2.3

@@
 \section{Organização de notas em música}\label{notasMusica}
-Seja $ S_j=\{T_{i,1},T_{i,2}...T_{i,N}\} =\bigcup_{j \in [1,N]} T_{i,j} $ uma sequência de $N$ eventos
-musicais. Por simplicidade, sejam notas estes eventos musicais
-e seja $S_j$ chamada uma 'estrutura musical'. Esta seção é dedicada às técnicas
-que tornam $S_j$ interessante e agradável na audição.
+Seja $ S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1} $ uma sequência $S_j$ de $H$ eventos
+musicais $s_j$. Seja $S_j$ chamada uma 'estrutura musical' composta de eventos $s_j$ que são também estruturas musicais,
+p.ex. notas.
+Esta seção é dedicada às técnicas que tornam $S_j$ interessante e agradável na audição.
 
@@
 
-Pode-se convencionar
-o pulso como nível $j=0$ de agrupamento, o nível $j=-1$ 
+Seja o pulso o nível $j=0$ de agrupamento, o nível $j=-1$ 
 a primeira subdivisão do pulso, o nível $j=1$ a primeira algomeração dos pulsos e assim por diante.
@@
 
-\subsection{Motivos e unidades maiores}\label{subsec:motivos}
-
-\subsection{Estruturas direcionais}\label{subsec:dir}
-
+\subsection{Repetição e variação: motivos e unidades maiores}\label{subsec:motivos}
 Dadas as estruturas musicais básicas tanto frequenciais (acordes e escalas) quanto 
 rítmicas (divisões e aglomerações simples, compostas e complexas), é 
-natural pensar em como realizar estas estruturas de forma que tenha coesão e sentido.\cite{Boulez}
-Para tal, os arcos são a base: partindo de algum lugar e voltando, forma-se um arco. Uma música sem nenhum arco pode ser compreendida como uma música sem coesão.
-Na subseção seguinte exploramos arcos cíclicos e nesta
-os provenientes de estruturas direcionais.
+natural apresentar estas estruturas de forma que tenham coesão e sentido.\cite{Boulez}
+Para tal, é fundamental o conceito de arcos: partindo de algum lugar e voltando, forma-se um arco. 
+A audição de linhas melódicas e harmônicas é permeada de arcos musicais pela natureza cognitiva
+da escuta musical. Pode-se considerar a nota o menor arco, cada motivo e melodia também um arco.
+Cada tempo e cada subdivisão, cada compasso e secção da música, constitui um arco próprio.
+Uma música, cujos arcos não apresentam consistência entre si, pode ser compreendida como uma música sem coesão.
+A sensação de coerência provém, em grande parte, do tratamento hábil dos arcos de uma peça.
+
+Os arcos músicais são estruturas abstratas e passíveis de operações básicas. Um arco espectral, como um
+acorde, pode ser invertido, ampliado, permutado, por exemplo. Os arcos temporais, como uma melodia, um motivo, um compasso ou
+uma nota, são igualmente passíveis de variações. Lembrando que $S_j=\left\{s_j=T_i^j=\{t_i^{j}\}_0^{\Lambda_j-1}\right\}_0^{H-1}$ é uma
+sequência de $H$ eventos musicais $s_j$, cada evento com suas $\Lambda_j$ amostras $t_i^j$ (veja no início desta seção~\ref{notasMusica}), as técnicas básicas podem ser descritas assim:
+
+\begin{itemize}
+    \item A translação temporal é o deslocamento $\delta$ do material para um outro instante $\Gamma'=\Gamma + \delta$ da música.
+É uma variação do material com modificação na localização no decorrer da música: $\left\{s_j'\right\}=\left\{s_j^{\Gamma'}\right\}=\left\{s_j^{\Gamma+\delta}\right\}$ onde $\Gamma$ é a duração entre o começo da peça (ou trecho considerado) e o primeiro evento $s_0$ da estrutura $S_j$ original.
+Observe que $\delta$ é o deslocamento no tempo.
+    \item A dilatação ou contração temporal é a alteração da duração de cada arco por um fator $\mu\,:\; s_j'^{\Delta}=s_j^{\mu_j . \Delta}$. Possivelmente, $\mu_j=\mu$ constante.
+    \item Reversão temporal consiste em gerar uma sequência com os elementos em ordem invertida da sequência original $S_j$, assim: $S_j'=\left\{s_j'\right\}_0^{H-1}=\left\{s_{(H-j-1)}\right\}_0^{H-1}$.
+    \item A inversão intervalar é a inversão do sentido dos intervalos percorridos pelo material.
+A inversão é estrita se o número de semitons esteja sendo usado como referência para a operação: $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{-\varepsilon_j .
+f_0}\right\}$, onde $\varepsilon_j$ é o fator entre a frequência do evento $s_j$ e a frequência de $s_0$.
+A inversão é tonal caso as distâncias sejam consideradas em termos de números de graus da escala $E_k$: $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{-\varepsilon^{\left(e_{\left(j_e\right)}\right)} .
+f_0}\right\}_0^{H-1}$ onde $j_e=e_{j}^{inv}$ é o índice $k=j_e$ em $E_k$ da nota do evento $s_j$.
+    \item A inserção e remoção de materiais na estrutura $S_j$ pode ser ornamental ou estruturante: $S_j'=\{s_j'\}=\{s_j \text{ se condição A, caso contrário } r_j\}$, para qualquer material musical $r_j$, incluindo o instante vazio.
+Elementos de $R_j$ podem ser inseridos no começo, como uma prefixo de $S_j$, ao final, como um sufixo, ou no meio, dividindo $S_j$ ou fazendo dele o prefixo e o sufixo.
+Os dois materiais podem se misturar das formas mais diversas.
+    \item Modificações de articulação, instrumentação e espacialização $s_j'=s_j^{*_j}$, com $*_j$ a nova característica incorporada pelo elemento $s_j'$.
+    \item Acompanhamento, tanto a instrumentação quanto as linhas melódicas que presentes na ocorrência de $S_j$ podem sofrer modificações e ser considerada uma variação de $S_j$ em si.
+\end{itemize}
+
+Com estes procedimentos, outros são derivados: o retrógrado do inverso, uma contração temporal com um sufixo externo, etc.
+A estruturas musicais ressoam mentalmente devido à própria natureza do pensamento.
+As ideias dizem respeito a objetos/contextos. Em suas várias facetas,
+uma ideia lida com o mesmo número de elementos e aspectos conectivos entre eles.
+A música, através de sintonia com estas estruturas mentais, suscita impressões. É, assim, desencadeado
+todo um processo de ressonância mental e cortical responsável pelos sentimentos, lembranças e imaginações
+típicas de uma audição musical atenta. Esta atividade neurológica contribui para a terapia
+músical, conhecida pela utilidade em casos de depressão e dano neurológico.
+Considera-se que as regiões do cérebro responsáveis pelo processamento auditivo são também usadas para outras atividades, incluindo linguísticas e matemáticas.\cite{Sacks,Roederer}
+
+Como princípios gerais que orientam a criação de novos materiais musicais, utiliza-se estruturas paradigmáticas.
+Uma delas, central, é o dipolo tensão/relaxamento.
+Relaciona-se com este dipolo, cada outro dipolo tradicional: tônica/dominante, repetição/variação, consonância/dissonância, coerência/rompimento, simetria/assimetria, chegada/saída, perto/longe, parado/em movimento, etc.
+Já as relações ternárias tendem a se relacionar com o circular e a unificação.
+A subseção seguinte é dedicada aos arcos direcionais e, 
+no seguinte, os arcos cíclicos.
+
+\subsection{Estruturas direcionais}\label{subsec:dir}
 
 Os arcos podem ser decompostos em duas sequências convergentes: 
-uma que atinge o ápice (chamado literalmente de clímax pela teoria musical tradicional) e 
+uma que atinge o ápice, e 
 outra que, de forma paradigmática, volta do ápice à região de partida.
-Distingue-se entre arcos cujos clímax estão: no começo, no meio, no final, na primeira metade e na segunda metade como esquematicamente a dispostos na figura~\ref{fig:climax}.\cite{Schoenberg}
+Este ápice é chamado de clímax pela teoria musical tradicional.
+Distingue-se entre arcos cujos clímax estão: no começo, no meio, no final, na primeira metade e na segunda metade da duração considerada.
+Estas estruturas estão na figura~\ref{fig:climax}.
+O parâmetro que varia pode não existir, no caso o arco consiste somente em uma estrutura de referência.\cite{Schoenberg}
 
@@
         \includegraphics[width=\textwidth]{figuras/climax}
-    \caption{Distinções canônicas do clímax musical em uma melodia e outros domínios.}
+    \caption{Distinções canônicas do clímax musical em uma melodia e outros domínios.
+As possibilidades diferenciadas são: clímax no começo, clímax na primeira metade, clímax no meio, clímax na segunda metade, clímax no fim.
+Não está especificado o eixo das ordenadas pois pode não haver variação paramétrica real, neste caso a estrutura é uma referência.}
         \label{fig:climax}
@@
 
-Seja $U_i=\{u_i\}_0^{\Lambda_U-1}$ uma sequência crescente. A sequência 
-$R_i=\{r_i\}_0^{2\Lambda_U -2}=\left\{u_{(\Lambda_U-1-|\Lambda_U-1-i|)}\right\}_0^{2\Lambda_U-2}$ 
+Seja $S_i=\{s_i\}_0^{H-1}$ uma sequência crescente. A sequência 
+$R_i=\{r_i\}_0^{2H -2}=\left\{s_{(H-1-|H-1-i|)}\right\}_0^{2H-2}$ 
 é uma sequência que apresenta simetria especular perfeita, i.e.
@@
 Pode-se modificar isso com a substituição por sequências de tamanhos diferentes.
-Toda a teoria matemática de sequências, já estabelecida, pode ser utilizada para geração destes arcos.\cite{Guidorizzo,Schoenberg}
+Toda a teoria matemática de sequências, já estabelecida e ensinada corriqueiramente em cursos de cálculo III, pode ser utilizada para geração destes arcos.\cite{Guidorizzo,Schoenberg}
 Teoricamente, estas sequências aplicadas desta forma
 a qualquer característica dos eventos musicais produz arcos,
-pois resultam no afastamento e aproximação de uma parametrização inicial. 
+pois implicam no afastamento e retorno de uma parametrização inicial. 
 Assim, é possível para uma mesma sequência de eventos possuir um número de arcos distintos, com tamanhos e clímax diferentes. Este é um recurso
@@
 
-O entendimento filosófico de que o pensamento humano se baseia
+O entendimento filosófico de que o pensamento humano é fundamentado
 na percepção de semelhanças e diferenças dentre os estímulos
@@
 Matematicamente,
-as simetrias são grupos algébricos e os grupos finitos
-são sempre isomorfos a um grupo de permutações. 
+as simetrias são grupos algébricos e um grupo finito
+é sempre isomorfo a um grupo de permutações. 
 Pode-se dizer que
@@
 Da primeira propriedade conclui-se que toda permutação pode ser operada com outra permutação.
-De fato, pode-se aplicar uma permutação, depois outra e, se comparadas as ordenações inicial e final, há uma permutação resultante.
-
-Todo elemento $p$ operado consigo mesmo um número suficiente de vezes $n$ atinge o elemento neutro $p^n=e$, pois caso contrário com a propriedade do fechamento se prova que o grupo é infinito (gerado por $p$).
+De fato, pode-se aplicar uma permutação $p_1$, depois outra $p_2$ e, se comparadas as ordenações inicial e final, há uma permutação $p_3$.
+
+Todo elemento $p$ operado consigo mesmo um número suficiente de vezes $n$ atinge o elemento neutro $p^n=e$, caso contrário o grupo seria infinito (gerado por $p$).
 O menor $n\;:\;p^n=e$ é chamado de ordem do elemento.
 Assim, uma permutação finita $p$ aplicada sucessivamente atinge a ordenação inicial dos elementos, formando um ciclo.
-Este ciclo, se utilizado para parametros de notas musicais, resulta em um arco musical.
+Este ciclo, se utilizado para parametros de notas musicais, implica em um arco musical cíclico.
 
@@
 As permutações estão entre as linhas.
-Neste caso, a estrutura musical consiste em permutações propriamente ditas e algumas permutações diferentes estão operando para resultar em um comportamento cíclico.
+Neste caso, a estrutura musical consiste em permutações propriamente ditas e algumas permutações diferentes operam para o comportamento cíclico.
 
@@
 \centering
-\caption{Change Ringing: \emph{Peal} (padrão) com 3 sinos}
+\caption{Change Ringing: \emph{Peal} (padrão) com 3 sinos. As permutações estão entre as ordenações. Cada linha é uma ordednação dos sinos, cada ordenaçao é tocada, uma linha por vez.}
 \begin{tabular}{l c r}
@@
 seja $L_i$ uma sequência de eventos musicais (e.g. notas) e $p$ uma permutação.
-$L_i'=p(L_i)$ consiste nos mesmos elementos de $L_i$ com a ordenação
-modificada.
-As permutações são usualmente escritas em notação cíclica ou natural. 
-Esta notação consiste na ordem dos índices 
+$L_i'=p(L_i)$ consiste nos mesmos elementos de $L_i$ mas em ordem diferente.
+As permutações podem ser escritas em duas notações: cíclica ou natural. 
+A notação natural consiste na ordem dos índices 
 resultante da permutação. Assim,
-convencionada a ordenação original dada pela sequência de seus índices $[0\;1\;2\;3\;4\;5\;...]$ a permutação é notada pela sequência que produz (ex. $[1\;3\;7\;0\;...]$).
+convencionada a ordenação original dada pela sequência de seus índices $[0\;1\;2\;3\;4\;5\;...]$ a permutação é notada pela sequência que produz (ex.
+$[1\;3\;7\;0\;...]$).
+Na notação cíclica, a permutação é a troca de um elemento pelo
+da frente, e o último pelo primeiro.
 
@@
 
-Por último, duas sutilezas deste procedimento.
-Primeiro, a permutação $p$ não precisa envolver todos os elementos de $L_i$, i.e. ela
-pode operar em subconjuntos de $L_i$. Segundo, nem todos os elementos $l_i$ precisam ser executados a cada consulta de estado realizada.
+Duas sutilezas deste procedimento.
+1) A permutação $p$ não precisa envolver todos os elementos de $L_i$, i.e. ela
+pode operar em subconjuntos de $L_i$. 2) Nem todos os elementos $l_i$ precisam ser executados a cada consulta de estado realizada.
 Para exemplificar, seja $L_i$ 
@@
 
-Esta abordagem resultou em alguns trabalhos e implementações computacionais.\cite{figgusOriginal,figgusEspacializacao} Abaixo está um uso coerente desta técnica abaixo e no Apêndice~\ref{cap:FIGGUScode} apontada a implementação computacional disponibilizada em.\cite{MASSA}
+No Apêndice~\ref{cap:FIGGUScode} apontada a implementação computacional disponibilizada em.\cite{MASSA,figgusOriginal,figgusEspacializacao}
 
@@
 Existem diversas empreitadas que se propõem a modelar e explorar entendimentos
-sobre a linguagem musical e da linguística aplicada à música ou ainda
+sobre a 'linguagem musical', a 'linguística aplicada à música' ou ainda
 para discernimento entre 
@@
 `idiomas musicais'.\cite{Lerdahl, Harmonia, Salzer,Alfaix}
-De forma simplificada, um idioma musical é fruto da
-repetição de elementos e da repetição de relações entre elementos presentes no decorrer da música. Bastante relacionado é o uso de dicotomias 
-como: repetição e variação, relaxamento e tensão, equilíbrio e desequilíbrio, consonância e dissonância, etc.
+De forma simples, um idioma musical é fruto da escolha de materiais básicos e
+repetição de elementos e da repetição de relações entre elementos presentes no decorrer da música. Nestas questões, as dicotomias são salientes,
+como explicadas na subseção~\ref{subsec:motivos}: repetição e variação, relaxamento e tensão, equilíbrio e desequilíbrio, consonância e dissonância, etc.
 
@@
 
-Primeiro, a nota básica foi definida e suas características postas em termos
-claros e quantitativos (seção~\ref{notaBasica}).
-Em seguida, a composição interna da nota foi abordada, e compreendidas suas transições internas e tratamentos imediatos (seção~\ref{varInternas}).
-Por fim, esta presente seção dedica-se a organizar estas notas musicais em música.
-A gama de recursos com uma infinidade de possibilidades de resultados
+Primeiro, a nota básica foi definida e caracterizada em termos
+claros e quantitativos (seção~\ref{sec:notaDisc}).
+Em seguida, a composição interna da nota foi abordada, e compreendidas as transições internas e tratamentos imediatos (seção~\ref{varInternas}).
+Por fim, esta presente seção dedica-se a organizar estas notas em música.
+A gama de recursos e a consequênte infinidade de possibilidades de resultados
 é situação típica e cara às artes.\cite{Harmonia,Webern}
@@
 Por exemplo, pode-se obter as harmonias triádicas 'sujas' (com notas não pertencentes à tríade) através de sobreposições de quartas justas.
-Outro exemplo interessante é a presença simultânea de ritmos em diferentes métricas, constituindo o que chamamos de \emph{polirritmia}.
+Outro exemplo interessante é a presença simultânea de ritmos em diferentes métricas, constituindo o que chama-se de \emph{polirritmia}.
 A montagem musical \emph{Poli-hit mia} explora estas métricas simultâneas através de trem de impulsos convoluídos com as notas que compõem cada linha.
@@
 
-As escalas microtonais são importantes na música do século XX~\cite{microtonalidade} e possuem usos com resultados muito marcantes, como os quartos de tom na música indiana.
+As escalas microtonais são importantes na música do século XX~\cite{microtonalidade} e possuem resultados muito marcantes, como os quartos de tom na música indiana.
 A sequência musical \emph{MicroTom} explora estes recursos, incluindo melodias microtonais e harmonias microtonais com várias notas em um âmbito de alturas bastante reduzido.
@@
 
-Os vínculos entre parâmetros são formas poderosas de obter peças e montagens musicais.
+Como também apontado na subseção~\ref{subsec:mus2}, os vínculos entre parâmetros são formas poderosas de se obter peças e montagens musicais.
 O número de notas permutadas pode variar no decorrer da música, revelando vínculo com a duração da peça.
-As harmonias podem consituir-se triádicas (eqs.~\ref{triades}) com notas replicadas em várias oitavas e mais numerosas quanto menor a profundidade e frequência de vibratos (eqs.~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}), dentre outras incontáveis possibilidades.
-
-As simetrias apresentadas nas divisões da oitava (eqs.~\ref{escSim}) e das simetrias apresentadas através das permutações (tabela~\ref{tab:change} e eqs.~\ref{eq:groups}) podem ser usadas em conjunto.
+As harmonias podem constituir-se triádicas (eqs.~\ref{triades}) com notas replicadas em várias oitavas e mais numerosas quanto menor a profundidade e frequência de vibratos (eqs.~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}), dentre outras incontáveis possibilidades.
+
+As simetrias apresentadas nas divisões da oitava (eqs.~\ref{escSim}) e as simetrias apresentadas através das permutações (tabela~\ref{tab:change} e eqs.~\ref{eq:groups}) podem ser usadas em conjunto.
 Nas peças \emph{3 trios} esta associação é feita de forma sistemática para possibilitar uma audição a ela dedicada.
-Esta é uma peça instrumental e não consta dentre os códigos dos Apêndices e da \massa.
+Esta é uma peça instrumental e não consta dentre os códigos dos Apêndices e da \massa.\cite{3Trios}
 
@@
 Com pouca parametrização, o programa gera músicas inteiras, permitindo a composição de músicas e conjuntos de músicas com facilidade.
-As possibilidades de compartilhamento permitem obter este album (ou outro feito de forma semelhante) através de poucas linhas de código e, pela execução, obter uma pasta com as músicas.
+Através de poucas linhas de código e, pela execução, pode-se obter uma pasta com as músicas.
+Esta facilidade e entrega tecnológica abre possibilidades estéticas, de compartilhamento e educacionais.

### 2013-03-22 trabalhosRelacionados.tex
finalizado 2.3 e conclusoes

--- 2012-11-28 fmam.tex
@@
-\chapter{Síntese FM e AM em escala logarítmica}
-\label{cap:fmam}
+\chapter{Trabalhos relacionados e caracterização das contribuições destre trabalho}
+\label{cap:trabalhosRelacionados}
 
@@
 
-Por brevidade, as identidades úteis para a abtenção
+Por brevidade, as identidades úteis para a obtenção
 da equação final, que explicita o conteúdo em frequências puras.

### 2013-03-22 conclusoes.tex
finalizado 2.3 e conclusoes

@@
 Está prevista a utilização destes resultados com
-métodos generativos para geração de materiais artísticos.
+métodos de inteligência artificial para geração de materiais artísticos.
 
@@
 Explorações sistemáticas de parametrizações (dos tremolos, da ADSR, etc) em alta fidelidade tem utilidade artística e é possibilidata por este trabalho com controle amostral.
+Tal descrição analítica precisa, junto às implementações computacionais, não foi atingida anteriormente, como mostra o Apêndice~\ref{cap:trabalhosRelacionados} com uma visita aos trabalhos relacionados.
 
@@
 atividades artísticas e educacionais, como cursos, workshops e apresentações artísticas.
-Vários destes resultados estão apresentados no Apêndice~\ref{cap:musicaExtra} e no acervo online criado, que ultrapassa 700 vídeos, documentações escritas, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos.\cite{siteLM,wikiLM,vimeoLM}
+Vários destes resultados estão no Apêndice~\ref{cap:musicaExtra} e no acervo online, que ultrapassa 700 vídeos, documentações escritas, diversos software originais e contribuições em software externos utilizados no mundo todo, como o Firefox, Scilab, LibreOffice, GEM/Puredata, para citar somente alguns exemplos.\cite{siteLM,wikiLM,vimeoLM}
 
-Há um aumento no número de pesquisas em música em andamento no campus de São Carlos da USP, o que sugere facilidade para estabelecer parcerias.
+Há um aumento no número de pesquisas relacionadas à música em andamento no campus de São Carlos da USP, o que sugere facilidade para estabelecer parcerias.
 As publicações acadêmicas efetivadas durante este mestrado também apontam para uma multidisciplinaridade,

### 2013-03-23 devEresultados.tex
arrumacoes gerais

@@
     \item Repetir notas, como artifício de sua incidência anterior, não adiciona informação relevante.
-    \item Tocar notas adjacentes ao mesmo tempo, em bloco.
-    \item As durações podem ser livres, desde que respeitada ordem de aparição das notas.
+    \item Soar notas adjacentes ao mesmo tempo, em bloco.
+    \item Apresentar durações livremente, desde que respeitada ordem de aparição das notas.
     \item Para variação, além dos recursos de ampliação, transposição e translação, são usados o inverso, o retrógrado e o retrógrado do inverso.
@@
     \item Reversão temporal consiste em gerar uma sequência com os elementos em ordem invertida da sequência original $S_j$, assim: $S_j'=\left\{s_j'\right\}_0^{H-1}=\left\{s_{(H-j-1)}\right\}_0^{H-1}$.
+    \item A translação em altura é o deslocamento $\tau$ do material para uma frequência/altura diferente $\Xi'=\Xi + \tau$ da música.
+É uma variação com modificação na localização em altura do material: $\left\{s_j'\right\}=\left\{s_j^{\Xi'}\right\}=\left\{s_j^{\Xi+\tau}\right\}$ onde $\Xi$ é a altura do trecho $S_j$ ou do primeiro evento $s_0$ da estrutura $S_j$ original.
+Observe que $\tau$ é o deslocamento em altura.
+Caso $\tau$ seja dado em semitons, o deslocamento em frequência é $\tau_f=f_0.2^{\frac{\tau}{12}}$ onde $f_0$ é resultado da adoção de alguma referência: $f_0=\Xi_{f_0}Hz\;\sim \Xi_0$ valor absoluto de altura.
+Para a frequência para qualquer valor de altura: $\Xi_f=\Xi_{f_0}.2^{\frac{\Xi-\Xi_0}{12}}$.
+        Para altura de qualquer valor de frequência: $\Xi=\Xi_0 +12 .
+\log_2\left(\frac{\Xi_f}{\Xi_{f_0}}\right)$.
+No protocolo MIDI, $\Xi_{f_0}=55Hz$ enquanto $\Xi_0=33$ de altura marca o lá 1, outro ponto de referência é o $\Xi_{f_0}=440Hz$ que corresponde a $\Xi_0=69$.
+A diferença da unidade equivale ao semitom.
+Não se pode dizer que a unidade é o semitom pois $\Xi=1$ não é um semitom, é uma nota de frequência audível como rítmo, com menos de 9 ocorrências por segundo (veja a tabela~\ref{tab:duracoes}).
     \item A inversão intervalar é a inversão do sentido dos intervalos percorridos pelo material.
@@
     \item Modificações de articulação, instrumentação e espacialização $s_j'=s_j^{*_j}$, com $*_j$ a nova característica incorporada pelo elemento $s_j'$.
-    \item Acompanhamento, tanto a instrumentação quanto as linhas melódicas que presentes na ocorrência de $S_j$ podem sofrer modificações e ser considerada uma variação de $S_j$ em si.
+    \item Acompanhamento. Tanto a instrumentação quanto as linhas melódicas que presentes na ocorrência de $S_j$ podem sofrer modificações e ser considerada uma variação de $S_j$ em si.
 \end{itemize}

### 2013-03-23 trabalhosRelacionados.tex
arrumacoes gerais

@@
 
-\begin{equation}
-\begin{split}
-\{t_i'\}_0^{\Lambda -1} &  = \left \{ \cos \left [f_i .
-2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1}  = \left \{ \cos \left [f\left ( \frac{f+\mu}{f} \right )^{sen \left ( f' .
-2 \pi \frac{i}{ f_a -1 } \right )} .
-2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1}  =  \\
- & = \left \{ \cos \left [f\left ( 2^{\frac{\nu}{12}} \right )^{sen \left ( f' .
-2 \pi \frac{i}{ f_a -1 } \right )} .
-2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1} =\left \{ \cos \left [f\left ( 2^{\frac{\nu}{12} \sin \left ( f' .
-2 \pi \frac{i}{ f_a -1 } \right )  } \right ) .
-2 \pi \frac{i}{f_a-1} \right ] \right \}_0^{\Lambda-1} \Rightarrow \\
-& \Rightarrow \text{usando: } \left ( 2^x=e^{x \ln 2}, \;\; e^x=\sum_0^\infty \frac{x^n}{n!},\;\; \sin^nx=y_1(n)\sum_k y_2(n,k).\sin[y_3(n,k,x)]   \right ) \Rightarrow \\
-\Rightarrow \{t_i'\}_0^{\Lambda -1} & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu,f) \cos \left [ y(f.f', f_a) \right ]  \right \}_0^{\Lambda-1}
-\end{split}
-\end{equation}
+Os trabalhos relacionados a esta dissertação são numerosos. Dentre as causas disso, pode-se apontar:
 
-Por brevidade, as identidades úteis para a obtenção
-da equação final, que explicita o conteúdo em frequências puras.
-Portanto, se utilizada a progressão exponencial de frequência,
-o espectro da FM perde a simplicidade.
-A função de Bessel dependerá não somente da profundidade de oscilação,
-mas também da frequência original do som. Além disso, as frequências dependem do produto das frequências
-da portadora e da moduladora ($f.f'$), o que também introduz complexidade.
+\begin{itemize}
+    \item Natureza interdisciplinar entre música, computação e física.
+    \item Há um interesse generalizado em música por parte das pessoas que compõem a sociedade.
+    \item A programação de computadores está se difundindo notavelmente.
+    \item As rotinas descritas neste trabalho são essenciais para boa parte dos \emph{software} voltados para áudio e música.
+\end{itemize}
 
-No caso da modulação de amplitude na escala logarítmica:
-\begin{equation}\label{eq:am}
-\begin{split}
-\{t_i'\}_0^{\Lambda-1} & =\{(a_{\text{máx}})^{ti'} .
-t_i\}_0^{\Lambda-1}= \left \{ a_{\text{máx}}^{\sin \left ( f'.2\pi\frac{i}{f_a -1} \right )} .
-P .\sin \left ( f.2\pi\frac{i}{f_a -1} \right ) \right \}_0^{\Lambda-1}
-\end{split}
-\end{equation}
+Vale observar que, embora estas rotinas estejam presentes em diversas implementações livres e proprietárias, suas descrições precisas se encontram somente em código computacional.
+A maior contribuição desta dissertação é exatamente a descrição analítica das qualidades básicas que compõem elementos musicais no áudio digital.
+A apresentação didática dos fenômenos envolvidos também não foi encontrada na literatura visitada, o que, junto com as implementações em código Python destas relações e de peças musicais que as exemplifiquem, forma uma contribuição simples e convidativa embora inédita e multidisciplinar.
 
-através das mesmas três identidades, observa-se um espectro com complexidade similar.
+Este capítulo é dedicado aos trabalhos similares ou relacionados.
+Os livros mais próximos são descritos.
+Na sequência, são apontadas as implementações computacionais proprietárias e livres.
+Por fim, esta dissertação é posicionada com relação aos trabalhos relacionados.
+
+\section{Trabalhos relacionados}
+
+\subsection{Livros}
+
+\subsection{Bibliotecas, linguagens e conjuntos de implementações computacionais voltados para música}
+
+\subsection{Aprofundamento sobre esta dissertação com base nos trabalhos visitados}

### 2013-03-23 devEresultados.tex
caminhado mais com 2.3 e apendice F

@@
 
-\subsection{Afinação}
+\subsection{Afinação, intervalos, escalas e acordes}
+\subsubsection{Afinação}
 O dobro da frequência é uma oitava ascendente ($f=2f_0$).
@@
 
-\subsection{Intervalos}\label{subsec:intervalos}
+\subsubsection{Intervalos}\label{subsec:intervalos}
 Em proporções de $\varepsilon=2^{\frac{1}{12}}$ entre as frequências das notas (i.e.
@@
 
-\subsection{Escalas}\label{subsec:escalas}
+\subsubsection{Escalas}\label{subsec:escalas}
 Uma escala é um conjunto ordenado de alturas. Usualmente, as escalas se repetem
@@
 
-\subsection{Acordes}\label{subsec:acordes}
+\subsubsection{Acordes}\label{subsec:acordes}
 A ocorrência simultânea de notas é observada através dos acordes.
@@
 
-\subsection{Harmonia}\label{subsec:harmonia}
-
+\subsection{Harmonias atonal, tonal, expansão e modulação}\label{subsec:harmonia}
+\subsubsection{Harmonia atonal}
 Omitir os encadeamentos básicos do sistema tonal é a chave para a obtenção de 
@@
 ou forem em número pequeno, pode-se dizer que a
-harmonia é modal. Caso encadeamentos tonais básicos estejam ausentes, a notas não coincidam
-com alguma das escalas diatônicas e forem diversos e dissonantes o suficiente para evitar redução por
+harmonia é modal. Caso encadeamentos tonais básicos estejam ausentes, as notas não coincidam
+com alguma das escalas diatônicas e forem diversas e dissonantes (com relação às outras notas) o suficiente para evitar redução por
 polarizações, a harmonia é atonal.
@@
 
+\subsubsection{Harmonia tonal}
 No século XX, músicas rítmicas
@@
 
+\subsubsection{Expansão tonal: funções individuais e medianas cromáticas}
 Cada um destes acordes pode ser confirmado e se desenvolver com uma execução de sua dominante ou subdominante individual (acorde com base na tríade formada a uma quinta acima e uma quinta abaixo respectivamente).
@@
 
+\subsubsection{Modulação}
 A modulação é a mudança da tonalidade em que se encontra a musica. Caracteriza-se a modulação através da observação
@@
 A diferença da unidade equivale ao semitom.
-Não se pode dizer que a unidade é o semitom pois $\Xi=1$ não é um semitom, é uma nota de frequência audível como rítmo, com menos de 9 ocorrências por segundo (veja a tabela~\ref{tab:duracoes}).
+Não se pode dizer que a unidade é o semitom pois $\Xi=1$ não é um semitom, é uma nota de frequência audível como ritmo, com menos de 9 ocorrências por segundo (veja a tabela~\ref{tab:duracoes}).
     \item A inversão intervalar é a inversão do sentido dos intervalos percorridos pelo material.
@@
 f_0}\right\}_0^{H-1}$ onde $j_e=e_{j}^{inv}$ é o índice $k=j_e$ em $E_k$ da nota do evento $s_j$.
+    \item Rotação de elementos musicais é a transferência do último elemento para a posição do primeiro e o deslocamento deste ao penúltimo uma posição para frente.
+Pode-se definir a rotação de $\tilde{n}$ posições por $S'_n=S_{(n+\tilde{n})\%H}$.
+Caso $\tilde{n}<0$, basta usar $\tilde{n}'=H-\tilde{n}$.
+É razoável associar $\tilde{n}>0$ com a rotação horária e $\tilde{n}<0$ com a rotação anti-horária.
+Mais sobre rotações na subseção~\ref{estCic}.
     \item A inserção e remoção de materiais na estrutura $S_j$ pode ser ornamental ou estruturante: $S_j'=\{s_j'\}=\{s_j \text{ se condição A, caso contrário } r_j\}$, para qualquer material musical $r_j$, incluindo o instante vazio.
@@
 Com estes procedimentos, outros são derivados: o retrógrado do inverso, uma contração temporal com um sufixo externo, etc.
-A estruturas musicais ressoam mentalmente devido à própria natureza do pensamento.
-As ideias dizem respeito a objetos/contextos. Em suas várias facetas,
+A estruturas musicais ressoam no sistema cognitivo devido à própria natureza do pensamento. Em suas várias facetas,
 uma ideia lida com o mesmo número de elementos e aspectos conectivos entre eles.
 A música, através de sintonia com estas estruturas mentais, suscita impressões. É, assim, desencadeado
-todo um processo de ressonância mental e cortical responsável pelos sentimentos, lembranças e imaginações
-típicas de uma audição musical atenta. Esta atividade neurológica contribui para a terapia
+todo um processo de ressonância mental e neurológica responsável pelos sentimentos, lembranças e imaginações
+típicas de uma audição musical atenta. Esta atividade cortical contribui para a terapia
 músical, conhecida pela utilidade em casos de depressão e dano neurológico.
@@
 interessante e útil e a correlação dos arcos resulta na coerência da escuta.\cite{Salzer}
+
+Na prática, e historicamente, tem especial importância a razão áurea e, por isso, à série de Fibonacci.
+A série de lucas é uma generalização da série de Fibonacci e pode ser compreendida facilmente.
+Dados dois números quaisquer $x_0$ e $x_1$, obtem-se a série de Lucas: $x_n=x_{n-1}+x_{n-2}$.
+Quanto maior for $n$, mais a razão $\frac{x_{n}}{x_{n_1}}$ se aproxima da razão áurea: $1.61803398875...$.
+A série converge rápido mesmo para valores iniciais bem discrepantes.
+Seja $x_0=1$ e $x_1=100$, e $y_n=\frac{x_n}{x_{n+1}}$ uma sequência auxiliar.
+O erro em percentagem dos primeiros valores desta sequência com relação à proporção áurea é, aproximadamente, $\{ e_n \} =\left\{100\frac{y_n}{1.61803398875}-100 \right\}_1^{10}=\{6080.33, -37.57, 23, -7.14, 2.937, -1.09, 0.42, -0.1601, 0.06125, -0.02338\}$.
+A sequência de Fibonacci apresenta aproximadamente a mesma progressão de erros, mas começa já no segundo passo pois $\frac{1}{1}\approx\frac{101}{100}$.
 

### 2013-03-23 trabalhosRelacionados.tex
caminhado mais com 2.3 e apendice F

@@
 A apresentação didática dos fenômenos envolvidos também não foi encontrada na literatura visitada, o que, junto com as implementações em código Python destas relações e de peças musicais que as exemplifiquem, forma uma contribuição simples e convidativa embora inédita e multidisciplinar.
+Vale apontar que, no início da escrita desta dissertação, a caixa de ferramentas não estava prevista, ela foi fruto das equações e descrições precisas, o que tornou imediata a escrita dos scripts que compõem a \emph{toolbox} \massa.
 
@@
 
+\begin{enumerate}
+    \item \emph{Music For Geeks And Nerds: learn more about music with Python and a little bit of math}
+        \begin{itemize}
+            \item {\bf Descrição:} com exemplos em código computacional e sonoros, este excelênte livro de Pedro Kroeger aborda conceitos de notas, afinações, especificação da nota Midi e conversão entre nomenclaturas latina (dó-ré-mi) e anglo-saxã (C-D-E).
+Lida também com operações musicais fundamentais como transposição, inversão e afins, combinações randômicas, por Fibonacci.
+Explora estas organizações tanto para acordes quanto para combinações horizontais (melódicas).
+Apresenta o básico sobre a constituição dos sons, batimentos, série harmônica, e um aprofundamento sobre as afinações.
+Por fim, aponta os recursos de ampliação temporal e de tessitura (alturas) de um dado conjunto de notas.
+Com isso, aponta o dipolo repetição/variação.
+Faz vínculos de apreensão de estruturas com peças de Josquin des Prez, Bach, Rachmaninoff e Steve Reich.
+            \item {\bf Aspecto complementar: as formalizações de operações dentro da notação tradicional são preciosos adendos às questões naturais abordadas na presente dissertação.
+A notação em si capta aspectos estruturais do sistema tonal e de 12 notas.
+Além disso, com a notação abre-se uma ponte com as tradições musicais eruditas.}
+            \item {\bf Aspecto diferencial: o livro não desenvolve descrição precisa de aspectos musicais do som em si e não há foco em relacionar qualidades psicofísicas aos elementos musicais.}
+            \item {\bf Contribuições diretas:}
+        \end{itemize}
+    \item  \emph{The Theory and Technique of Electronic Music}
+        \begin{itemize}
+            \item {\bf Descrição:} um livro de Miller Puckette, de reconhecida complexidade, se define, nas palavras de Max Matheus (Prefácio) "The Theory and Technique of Electronic Music is a uniquely complete source of information for the computer synthesis of rich and interesting musical timbres".
+O livro começa com medições do som e controle de parâmetros, cai em síntese, modulações, métodos espectrais, atrasos e reverberações e termina com filtros.
+            \item {\bf Aspecto complementar:} o livro apresenta diversos procedimentos valioso para síntese, tratamento e análise.
+O texto busca ser de computação musical em geral e todo o texto é acompanhado de exemplos em Puredata, que é uma excelênte, linguagem de programação por patches voltada para audiovisual.
+Puredata é a linguagem de programação mais difundida na música acadêmica e tecnológica em geral.
+            \item {\bf Aspecto diferencial:} salvo raras excessões, o livro não apresenta uma descrição analítica das amostras sonoras com relação aos procedimentos, assim, não relaciona de forma precisa as qualidades físicas do som.
+Tampouco se aprofunda em aspectos formais da teoria musical tradicional.
+            \item {\bf Contribuições diretas:} na página 92, há uma solução para a o \emph{fade-in} e o \emph{fade-out} que, se feitos em progressão geométrica, demora a cair ao inaudível.
+A curva "quártica" atinge o zero e se é bastante próxima da progressão exponencial, especialmente nas intensidades maiores: $a_n = \left\{\left(\frac{n}{\Lambda-1}\right)^4\right\}_0^{\Lambda-1}$.
+Outra contribuição é a descrição prática do uso ideal de 1000 ou mais linhas de atrasos por segundo para simular a reverberação.
+Também deixa claro que há uma equalização na atenuação do som refletido, e que esta equalização tende ser mais atenuante nos agudos.
+        \end{itemize}
+    \item \emph{}
+        \begin{itemize}
+            \item {\bf Descrição:}
+            \item {\bf Aspecto complementar:}
+            \item {\bf Aspecto diferencial:}
+            \item {\bf Contribuições diretas:}
+        \end{itemize}
+\end{enumerate}
+
 \subsection{Bibliotecas, linguagens e conjuntos de implementações computacionais voltados para música}

### 2013-03-23 devEresultados.tex
caminhada a revisao de trabalhos relacionados e os acrescimos no capitulo 2

@@
 
-\subsection{Localização espacial}
-Embora não seja uma das quatro qualidades básicas tradicionais de uma nota musical, esta possui sempre uma localização espacial. Além disso, a espacialização é atributo bastante valorizado
+\subsection{Localização espacial e espacialização}\label{subsec:spac}
+Embora não seja uma das quatro qualidades básicas tradicionais de uma nota musical, esta possui sempre uma localização espacial, que é a posição da fonte que a emitiu, no espaço físico tridimensional ordinário.
+Além disso, há um ambiente que reverbera a nota emitida, assunto ao qual a 'espacialização' é dedicada.
+Ambas, a espacialização e a localização espacial, são bastante valorizadas
  por audiófilos e pela indústria fonográfica.\cite{floEsp}
+
+ \subsubsection{Localização espacial}
 Acredita-se que a percepção da localização espacial do som se dê em nosso sistema nervoso através destas
@@
 Já a localização completa, incluindo distância e elevação da fonte sonora, é dada pela função de transferência de cabeça (HRTF - do inglês \emph{Head Related Transfer Function}).\cite{hrtf} Exitem bases abertas e conhecidas de HRTF como a CIPIC e pode-se aplicar estas funções de transferência em um som por convolução (veja equação~\ref{eq:conv}).\cite{CIPIC} O corpo do indivíduo altera bastante as filtragens realizadas e existem técnicas para gerar HRTFs que sejam - como proposta - utilizáveis de forma universal.\cite{lazaSPA}
+
+\subsubsection{Espacialização}
+Já a espacialização é, fundamentalmente, o resultado das reflexões e absorções da nota emitida nas superfícies do recinto/paisagem no qual a nota foi emitida.
+Pode-se simular reverberações reais não possíveis em sistemas reais.
+O som viaja a $343,2m/s$, e pode ser emitido da fonte com qualquer padrão de direcionalidade.
+Quando uma frente sonora encontra uma superfície do recinto, há uma reflexão.
+Nesta reflexão ocorrem tanto 1) a inversão da componente da velocidade de propagação que é perpendicular à superfície, quanto 2) a absorção de energia, especialmente nos agudos.
+As ondas se propagam até atingirem niveis considerados inaudíveis.
+Quando alguma frente de onda atingir o ouvido, é registrada e pode ser inteiramente descrita com: momento de chegada ao ouvido e o número de reflexões junto aos filtros de absorção de cada superfície que atingiu.
+Para experimentações, pode-se usar reflexões assimétricas com relação ao eixo perpendicular à superfície, ou ainda ganhos em determinadas bandas de frequência (tidos como 'ressonâncias'), ambas as características não são encontradas em sistemas reais.
+
+Existem algumas modelagens de reverberação menos atreladas ao cálculo de cada reflexão, exploram informações valiosas do ponto de vista auditivo.
+De fato, a reverberação pode ser modelada com um conjunto de 2 características temporais e no espectro:
+\begin{itemize}
+    \item Primeiro periodo: as 'primeiras reflexões' são mais intensas e esparças.
+    \item Segundo período: a 'reverberação tardia' é uma continuação de atrasos com um decaimento exponencial e ocorrências estatísticas.
+    \item Primeira banda: o grave possui algumas frequências de ressonância relativamente espaçadas.
+    \item Segunda banda: o médio e agudo possuem um decaimento progressivo e suave com flutuações estatísticas.
+\end{itemize}
+
+Smith III aponta que boas salas de concerto possuem um tempo total de reverberação de aproximadamente $1,9$ segundos.
+Aponta também que o período das primeiras reflexões deve ser de $0,1$ segundos.
+Estas quantidades apontam que, nas condições contempladas, há frentes de onda perceptíveis que se propagam até $652,08$ metros ($83.79k$ amostras em $f_a=44.1kHz$) antes de atingirem o ouvido.
+Além disso, as reflexões do som formam, para a audição, após a propagação por $34,32$ metros ($4.41k$ amostras em $f_a=44.1kHz$ ), um emaranhado cujas incidências são pouco distintas.
+Estas primeiras reflexões são particularmente importantes para a sensação de espaço.
+A primeira incidência é o som direto, descrito por DTI e DII das equações~\ref{eq:dti} e~\ref{eq:dii}.
+Admitindo-se que cada uma das primeiras reflexões se propagará, ao menos, $3-30m$ dependentes das dimensões da sala, a separação entre as primeiras reflexões é de, ao menos, $8-90$ milissegundos, correspondendo a uma separação de aproximadamente $350-4000$ amostras.
+Verifica-se experimentalmente que o número de reflexões por aumenta em proporção quadrática  $ \approx k.n^2$.
+Apontamentos do uso de convoluções e filtragens para facilitar estas implementações estão na subseção~\ref{subsec:mus2}, especialmente nos parágrafos sobre reverberação.
 
@@
 
-Uma primeira possibilidade interessante é usar vínculos
+\subsubsection{Vínculos entre características}
+
+Uma outra possibilidade interessante é usar vínculos
 entre os parâmetros do tremolo e do vibrato e algum parâmetro da nota básica,
@@
 
+\subsubsection{Fonte em movimento, receptor em movimento, efeito Doppler}
+
+Retomada a exposição da subseção~\ref{subsec:spac}, quando uma fonte ou um receptor está em movimento, suas características idealmente são calculadas amostra a amostra do sinal digital.
+As velocidades são decompostas com relação à direção de cada ouvido.
+Assim, dada a velocidade $v_s$ da fonte ($s$ de \emph{source}), positiva se a fonte se no sentido de se afastar do receptor e a velocidade $v_r$ do receptor, positiva no sentido de se aproximar da fonte, a frequência é dada pela conhecida fórmula do efeito Doppler:
+
+\begin{equation}
+    f=\left(\frac{c+v_r}{c+v_s}\right)f_0
+\end{equation}
+
+Com esta frequência e com as relações dadas pela nova DII da nova localização da fonte, pode-se realizar o efeito Doppler.
+Há um acréscimo para maior fidelidade ao fenômeno físico: um aumento da potência recebida $\Delta P=\left(\frac{c+v_r}{c+v_s}\right)P_0$, com $P_0$ a potência do sinal.
+
+\subsubsection{Ruidos e filtragens}
 Com os filtros as possibilidades explodem ainda mais vertiginosamente. Pode-se convoluir um sinal para reverberá-lo, para
@@
 
+\subsubsection{Reverberação}
+Retomando também a exposição da subseção~\ref{subsec:spac}, a reverberação tardia pode ser modelada como a convolução com um trecho de ruído colorido rosa, marrom ou preto, em decaimento exponencial de amplitude com relação também ao tempo.
+A atenuação nos agudos e irregularidade suave na resposta em frequência são contempladas de forma bastante satisfatória.\cite{JOSPhy,guillaume} Linhas de atrasos podem ser adicionadas como prefixo à ao ruído com o decaimento, assim comtemplando as duas partes temporais da reverberação: as primeiras reflexões e a reverberação tardia.
+Pode-se, para melhora qualidade, calcular geometricamente a localização da última superfície em que cada frente de onda refletiu antes de chegar ao ouvido nos primeiros $100-200$ milissegundos e aplicar um filtro passa-baixas como em~\ref{subsec:filtros}.
+O ruído colorido pode ser introduzido gradualmente, desde o instante inicial dado pela incidência direta (i.e.
+sem reflexão alguma e dada pela DTI e DII), em \emph{fade-in}, atingindo o máximo no começo da 'reveberação tardia', quando as incidências geométricas perdem sua importância para as estatísticas do ruído em decaimento.
+
+A reverberação é conhecida por causar bastante interesse no ouvinte e tornar as sonoridades mais agradáveis.
+Além disso, a modificação do espaço em que são reverberadas a música, ou as sonoridades individualmente, constitui um massete (quase um clichê) para causar supresa e interesse no ouvinte.
+
+\subsubsection{Envoltória ADSR}
 A variação de volume no decorrer de um som é crucial para nossa percepção de timbre.

### 2013-03-23 trabalhosRelacionados.tex
caminhada a revisao de trabalhos relacionados e os acrescimos no capitulo 2

@@
             \item {\bf Aspecto diferencial: o livro não desenvolve descrição precisa de aspectos musicais do som em si e não há foco em relacionar qualidades psicofísicas aos elementos musicais.}
-            \item {\bf Contribuições diretas:}
         \end{itemize}
@@
         \end{itemize}
+    \item \emph{Real Sound Synthesis for Interactive Applications}
+        \begin{itemize}
+            \item {\bf Descrição:} livro do Perry Cook, de 2002, discorre sobre fundamentos de áudio digital e modelagem física.
+Possui preciosas dicas de modelagens de sons reais com características de intrumentos tradicionais e efeitos com origem nos instrumentos analógicos.
+            \item {\bf Aspecto complementar:} implementa diversos instrumentos musicais e efeitos sonoros.
+Trata de uma biblioteca em C para síntese sonora que contempla boa parte das funcionalidades da \massa.
+            \item {\bf Aspecto diferencial:} O trabalho escapa à descrição das amostras sonoras em si e não possui uma sistematização de elementos musicais básicos em termos das características sonoras.
+            \item {\bf Contribuições diretas:}  a parte "1.3 Quantização" descreve de forma breve e suficiente o ruído de quantização e pode complementar à dissertação em "1.1 Som em áudio digital".
+O escrito sugere uma melhora na relação sinal/ruído de 6 decibels por bit utilizado na quantização de cada amostra do áudio PCM.
+O que indica diretamente uma relação sinal/ruído de 96dB para 16 bits/amostra (padrão de CD) e 48dB para 8 bits/amostra (comum em sistemas de voz).
+        \end{itemize}
+    \item \emph{Interação Tímbrica na Música Eletroacústica Mista}
+        \begin{itemize}
+            \item {\bf Descrição:} dissertação de Ignacio de Campos, possui diversas discussões cognitivas e musicais.
+            \item {\bf Aspecto complementar:} desenvolvem os apontamentos da dissertação e do Apêndice~\ref{cap:musicaExtra} com relação à \emph{performance} musical.
+            \item {\bf Aspecto diferencial:} o trabalho não lida de forma sistemática com as características do sinal digital de elementos musicais.
+            \item {\bf Contribuições diretas:} a parte "2.3.2 A Sensação de Identidade Tímbrica" pode completar a exposição sobre timbre em "2.1.4 Timbre" com os apontamentos de envoltória, inarmonicidade, jitter e shimmer.
+        \end{itemize}
+    \item \emph{Music, Cognition, and Computerized Sound: An Introduction to Psychoacoustics}
+        \begin{itemize}
+            \item {\bf Descrição:} outro livro do Perry Cook, desta vez são artigos de vários autores, ele cuidou da edição.
+Um valioso apanhado pertinente para a psicofísica de elementos musicais no áudio digital, com textos focados em aspectos cognitivos e físicos do som além de eventuais descrições de procedimentos elétricos e digitais envolvidos.
+O foco é qualitativo embora com bastante embasamento quantitativo.
+            \item {\bf Aspecto complementar:} o livro traz artigos de vários autores. As temáticas são pertinentes para a dissertação como cultura a respeito dos assuntos.
+            \item {\bf Aspecto diferencial:} o trabalho não é uma descrição de elementos musicais com relação às características do som digital.
+            \item {\bf Contribuições diretas:} talvez as partes mais interessantes para a dissertação sejam a subseção "23.10 Special Considerations in Psychoacoustic Research" e o capítulo "Appendix A: Suggested Lab Exercises".
+        \end{itemize}
+    \item \emph{Modelos Psicoacústicos de Dissonância para Eletrônica ao Vivo}
+        \begin{itemize}
+            \item {\bf Descrição:} tese de Alexandre Porres com considerações pertinentes sobre modelos psicoacústicos, dissonância e aspereza/rugosidade.
+Outros trabalhos, do mesmo autor, incluem um livro sobre computação musical em PD e apresentam, de forma acessível, procedimentos-chave de computação musical junto às implementações em Puredata.
+            \item {\bf Aspecto complementar:} a tese foca em teorias psicoacústicas e em descrições minuciosas de rugosidade e dissonâncias que podem acrescentar bastante ao que foi apresentado na dissertação.
+Além disso, há um viés prático com a utilização do Puredata.
+            \item {\bf Aspecto diferencial:} a trabalho não é uma descrição psicofísica de sequências amostrais relacionadas a elementos musicais.
+        \end{itemize}
+    \item \emph{5 livros do Julious O. Smith III}\cite{JOSFaust,JOSSpec,JOSFilt,JOSPhy,JOSFM}
+        \begin{itemize}
+            \item {\bf Descrição:} este autor possui diversos escritos, mais de 200 artigos publicados.
+Ao menos 5 livros de interesse para a dissertação: \emph{Mathematics of the Discrete Fourier Transform (DFT)} descreve a DFT minuciosamente, influindo FFT.
+O livro \emph{Introduction to Digital Filters (with audio applications)} trata de filtros de diversos tipos e descreve transfomadas, técnicas de design de filtros e análises de frequência e por polos e zeros.
+Além de abordar filtros não convencionais, o livro termina com implementações em matlab, Faust e PD e possibilidades de confeção de plugins.
+O livro \emph{Physical Audio Signal Processing (for virtual musical instruments and audio effects)} trata de modelagem física.
+O livro \emph{Spectral Audio Signal Processing} adentra STFT em detalhes das janelas utilizadas e aplicações da STFT.
+Também aborda Wavelets de forma superficial porém consistente.
+O livro \emph{Audio Signal Processing in Faust} trata da liguagem Faust e suas facilidades para o processamento de áudio e música.
+            \item {\bf Aspecto complementar:} a seção "Perceptual Aspects of Reverberation", do livro de filtros, pode fechar o assunto da espacialização junto à subseção "2.1.7 Localização espacial".
+As partes de modelagem de voz e de fundamentos físicos podem também servir de bons  complementos à dissertação.
+o livro sobre Faust complementa a dissertação por explicitar uma linguagem de domínio específico para áudio e música.
+O livro também contempla usos integrados com outros programas e linguagens via, por exemplo, o padrão de plugin LADSPA ou o protocolo OSC de comunicação entre programas para manipulação multimídia.
+            \item {\bf Aspecto diferencial:} estes trabalhos apresentam teorias fundamentais para o que se faz hoje em termos de programação para música, e não uma descrição de elementos musicais em termos das amostras digitais.
+            \item {\bf Contribuições diretas:} a parte de espacialização tem como referencia forte o livro de modelagem física.
+A clareza sobre a natureza da reverberação e sobre suas características foi toda disparada por este livro.
+        \end{itemize}
+    \item \emph{}
+        \begin{itemize}
+            \item {\bf Descrição:}
+            \item {\bf Aspecto complementar:}
+            \item {\bf Aspecto diferencial:}
+            \item {\bf Contribuições diretas:}
+        \end{itemize}
+    \item \emph{}
+        \begin{itemize}
+            \item {\bf Descrição:}
+            \item {\bf Aspecto complementar:}
+            \item {\bf Aspecto diferencial:}
+            \item {\bf Contribuições diretas:}
+        \end{itemize}
+    \item \emph{}
+        \begin{itemize}
+            \item {\bf Descrição:}
+            \item {\bf Aspecto complementar:}
+            \item {\bf Aspecto diferencial:}
+            \item {\bf Contribuições diretas:}
+        \end{itemize}
+    \item \emph{}
+        \begin{itemize}
+            \item {\bf Descrição:}
+            \item {\bf Aspecto complementar:}
+            \item {\bf Aspecto diferencial:}
+            \item {\bf Contribuições diretas:}
+        \end{itemize}
+    \item \emph{}
+        \begin{itemize}
+            \item {\bf Descrição:}
+            \item {\bf Aspecto complementar:}
+            \item {\bf Aspecto diferencial:}
+            \item {\bf Contribuições diretas:}
+        \end{itemize}
     \item \emph{}

### 2013-03-24 trabalhosRelacionados.tex
terminadas partes sobre espacializacao e sobre os livros relacionados

@@
         \end{itemize}
-    \item \emph{}
+    \item \emph{Music: A Mathematical Offering}
         \begin{itemize}
-            \item {\bf Descrição:}
-            \item {\bf Aspecto complementar:}
-            \item {\bf Aspecto diferencial:}
-            \item {\bf Contribuições diretas:}
+            \item {\bf Descrição:} este livro, escrito por Dave Benson, se ocupa dos fundamentos matemáticos do fenômeno e tradições musicais.
+            \item {\bf Aspecto complementar:} algumas afinações, escalas e explicações dos capítulos 5 e 6 (sobre afinações e escalas) podem facilitar algumas compreensões apontadas em subseção~\ref{subsec:afinacao}.
+O capítulo "9- Symmetry in Music" pode acrescentar nas estruturas cíclicas do capítulo 2.3 da dissertação, especialmente a parte de grupos, Cayley e órbitas.
+            \item {\bf Aspecto diferencial:} o livro não trata de áudio digital. Além disso, o trabalho é uma coleção de modelagens matemáticas, não uma descrição do fenômeno sonoro-musical.
         \end{itemize}
-    \item \emph{}
+    \item \emph{The Topos of Music: Geometric Logic of Concepts, Theory, and Performance}
         \begin{itemize}
-            \item {\bf Descrição:}
-            \item {\bf Aspecto complementar:}
-            \item {\bf Aspecto diferencial:}
-            \item {\bf Contribuições diretas:}
+            \item {\bf Descrição:} este impressionante livro de Guerino Mazzola possui quase 1500 páginas e explora diversos formalismos matemáticos para estruturas musicais.
+É um livro fascinante que atinge questões filosóficas.
+            \item {\bf Aspecto complementar:} podem servir de ilustração ou aprofundamento: a subseções "6.5 Regular and Circular Forms", "22.4 Paradigmatic Groups" e "22.5 Pseudo-metrics on Orbits" e as seções "8 Symmetries and Morphisms" e "11 Orbits" fundamentam e ampliam a parte de estruturas cíclicas de 2.3.
+A subsubseção "7.2.2 Local Meters and Local Rhythms" e a seção "21 Metrics and Rythmics" podem aprofundar na parte de "2.3.4 Ritmo".
+A Subsubseção 9.4.2.
+pode dar uma contribuição na abordagem artística do código computacional.
+A subseções "9.4 Undestanding Fine and Other Arts" e "10.1 Paradigmata in Musicology, Linguistics, and Mathematics" e as seções "47 Unfolding Geometry and Logic in Time" e "48 Local and Global Strategies in Composition" podem fundamentar melhor a parte 2.3.7.
+Idioma Musical.
+Os tomos "VI Harmony" e "VII Counterpoint" podem aprofundar as partes "2.3.2 Rudimentos de harmonia" e "2.3.3 Contraponto".
+O tomo "XV Appendix: Sound" é bastante interessate, com algumas partes que lembram bastante a dissertação: "A Common Parameter Spaces", "A.1 Physical Spaces", "A.2.1 Onset and Duration", "A.2.2 Amplitude and Crescendo", "A.2.3 Frequency and Glussando" e "B Auditory Physiology and Psychology".
+No tomo "XVI Appendix: Mathematical Basics", interessa a parte "C.3 Groups" em que trata de homomorfismos, produtos direto, semidireto e 'Wreath', teoremas de Sylow, classificação, grupons affinos gerais e grupos de permutação, a subseção "E.3 Categories of Modules and Affine Transformations", "F.2 Spectra of Commutative Rings".
+O tomo "XVII Appendix: Tables" pode ter algumas afinações para constar em "2.3.1 Afinação, intervalos, escalas e acordes".
+            \item {\bf Aspecto diferencial:} o livro não se ocupa da descrição de elementos musicais em termos do áudio digital.
         \end{itemize}
-    \item \emph{}
+    \item \emph{Symmetry as a Compositional Determinant}
         \begin{itemize}
-            \item {\bf Descrição:}
-            \item {\bf Aspecto complementar:}
-            \item {\bf Aspecto diferencial:}
-            \item {\bf Contribuições diretas:}
+            \item {\bf Descrição:} livro Larry J. Salomon, de 1973 e revisado em 2002, é uma leitura de procedimentos composicionais tradicionais tendo como base comum a simetria.
+            \item {\bf Aspecto complementar:} é um escrito bastante pertinete para toda a seção "2.3 Organização de notas em música", principalmente para a parte "2.3.6 Estruturas cíclicas" com as rotações, reflexões, translados e incidências tradicionais.
+            \item {\bf Aspecto diferencial:} o trabalho não relaciona as características musicais com o som como fenômeno físico.
+            \item {\bf Contribuições diretas:} foi um livro importante no amadurecimento das questões tratadas nesta dissertação por ser um livro focado em simetrias na música.
         \end{itemize}
-    \item \emph{}
+    \item \emph{The CSound Book}
         \begin{itemize}
-            \item {\bf Descrição:}
-            \item {\bf Aspecto complementar:}
-            \item {\bf Aspecto diferencial:}
-            \item {\bf Contribuições diretas:}
+            \item {\bf Descrição:} editado por Richard Boulanger, 2000, possui contribuições de diversos autores.
+Além de abordar pincípios da linguagem, seus funcionamentos internos e paradigmas musicais, é um apanhado de técnicas de síntese musical do som e estruturas de mais alto nivel.
+            \item {\bf Aspecto complementar:} apresenta a paradigmática linguagem de programação voltada para música: CSound.
+Junto a esta valiosa empreitada, consagrados recursos musicais eletrônicos são expostos.
+            \item {\bf Aspecto diferencial:} é focado no CSound e não se propõe a ser um livro de fundamentos com descrição formal das características do som relacionado às estruturas musicais. 
         \end{itemize}
-    \item \emph{}
+    \item \emph{The Audio Programming Book}
         \begin{itemize}
-            \item {\bf Descrição:}
-            \item {\bf Aspecto complementar:}
-            \item {\bf Aspecto diferencial:}
-            \item {\bf Contribuições diretas:}
+            \item {\bf Descrição:} de Richard Boulanger e Victor Lazzarini, 2011, um livro interessantíssimo que trata de programação em C/C++ voltado para áudio.
+Possui várias receitas tradicionais e fundamentais, todas implementadas em código C.
+Aborda CSound ao final.
+            \item {\bf Aspecto complementar:} aborda programação em C/C++, apresenta aspectos cruciais das implementações.
+            \item {\bf Aspecto diferencial:} não é uma descrição do comportamento amostral do sinal sonoro com relação aos elementos musicais.
         \end{itemize}
-    \item \emph{}
+    \item \emph{Computer Music Tutorial}
         \begin{itemize}
-            \item {\bf Descrição:}
-            \item {\bf Aspecto complementar:}
-            \item {\bf Aspecto diferencial:}
-            \item {\bf Contribuições diretas:}
+            \item {\bf Descrição:} de Curtis Roads, 1996, é já um clássico com algumas partes escritas por outros autores, como colaborações. 
+            \item {\bf Aspecto complementar:} o capítulo "VII - Psychoacoustics" pode ser complementar à dissertação em algum detalhes e para aprofundamento.
+            \item {\bf Aspecto diferencial:} Os textos do livro são de alto nivel, sem preocupação central com o rigor, mas sim com apontamentos qualitativos e historicos das subáreas da música computacional.
+        \end{itemize}
+    \item \emph{Music and Acoustics: From Instrument to Computer}
+        \begin{itemize}
+            \item {\bf Descrição:} um curto e preciso livro com os fundamentos acústicos, de modelagem de instrumentos musicais, de síntese e tratamento digitais.
+            \item {\bf Aspecto complementar:} apresenta técnicas de modelagem física, além de aspectos musicais e computacionais.
+            \item {\bf Aspecto diferencial:} não é uma descrição das amostras do sinal sonoro com relação às características musicais.
+            \item {\bf Contribuição direta:} a proposta de modelar a reverberação como um ruido com decaimento exponencial foi importante para a elaboração da exposição sobre reverberação na subsubseção~\ref{subsubsec:reverb}
         \end{itemize}

### 2013-03-24 devEresultados.tex
terminadas partes sobre espacializacao e sobre os livros relacionados

@@
 Com esta frequência e com as relações dadas pela nova DII da nova localização da fonte, pode-se realizar o efeito Doppler.
-Há um acréscimo para maior fidelidade ao fenômeno físico: um aumento da potência recebida $\Delta P=\left(\frac{c+v_r}{c+v_s}\right)P_0$, com $P_0$ a potência do sinal.
+Há um acréscimo para maior fidelidade ao fenômeno físico: um aumento da potência recebida.
+Pode-se assumir que este aumento em potência é proporcional à velocidade relativa, que, a cada segundo, acrescenta aquele trecho percorrido, pela forma de onda, de potência: $\Delta P=P_0\left(\frac{v_r-v_s}{343,2}\right)$, com $P_0$ a potência do sinal.
 
@@
 
-\subsubsection{Reverberação}
+\subsubsection{Reverberação}\label{subsubsec:reverb}
 Retomando também a exposição da subseção~\ref{subsec:spac}, a reverberação tardia pode ser modelada como a convolução com um trecho de ruído colorido rosa, marrom ou preto, em decaimento exponencial de amplitude com relação também ao tempo.
@@
 
-\subsection{Afinação, intervalos, escalas e acordes}
+\subsection{Afinação, intervalos, escalas e acordes}\label{subsec:afinacao}
 \subsubsection{Afinação}

### 2013-03-24 pretextual.tex
feita revisao completa, repassada e bib, jah era
feita revisao completa da dissertacao, arrumando pormenores

@@
 
+\afterpage{\blankpage}
 \pretextualchapter{LISTA DE RELAÇÕES ANALÍTICAS DESCRITAS E IMPLEMENTADAS COMPUTACIONALMENTE NO APÊNDICE~\ref{cap:codigoProc}}
@@
 
+\pretextualchapter{Elementos de notação utilizados na dissertação}
+	\begin{listaespecial2}[BIGNAMEWIDTH]
+        \item[$Hz$] abreviação de Herz, medida de frequência, número de ocorrências por segundo.
+        \item[$kHz$] abreviação de kilo Hertz, i.e. mil Herz.
+        \item[$m$] metro.
+        \item[$mm$] milimetro.
+        \item[$s$] segundo.
+        \item[$m/s$] metros por segundo.
+        \item[$f_a$] frequência de amostragem, taxa de amostragem.
+        \item[$\lambda_a$] duração da separação temporal entre um par de amostras consecutivas.
+        \item[PCM] sigla de \emph{Pulse Code Modulation}, veja seção~\ref{sec:audio}
+        \item[$S_i$] sequência $S$ indexada em $i$.
+        \item[$S_i=\{s_i\}_x^{y}$] sequência $S_i$ com elementos $s_i$com índices de $x$ a $y$ incluso.
+        \item[$\lfloor x \rfloor$] parte inteira de $x$.
+        \item[$V_{dB}$] volume em decibels.
+        \item[$\Rightarrow$] então.
+        \item[$\therefore$] portanto.
+        \item[$\approx$] aproximadamente.
+        \item[$\equiv$] equivalente.
+        \item[$\%$] operação módulo, resto da divisão.
+        \item[$\Lambda$] comprimento em amostras.
+        \item[$\widetilde{\Lambda}$] comprimento em amostras de período de onda.
+        \item[$\Delta$] duração temporal.
+        \item[$a_i$] fator multiplicativo de amplitude da $i$-ésima amostra.
+	\end{listaespecial2}
+
 %

### 2013-03-24 devEresultados.tex
feita revisao completa da dissertacao, arrumando pormenores

@@
 
-Onde $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\delta_a} \;$ é o número de amostras do período\footnote{Neste ponto já se tem toda a base para música \emph{Reduced-fi} do Apêndice~\ref{ap:reduced}.}.
+Onde $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\lambda_a} \;$ é o número de amostras do período\footnote{Neste ponto já se tem toda a base para música \emph{Reduced-fi} do Apêndice~\ref{ap:reduced}.}.
 

### 2013-03-24 introducao.tex
feita revisao completa da dissertacao, arrumando pormenores

@@
 Deste ponto de vista pedagógico, também se presta a apresentação destes resultados na forma de hipertexto, em que cada \emph{script} e exemplo sonoro/musical seja acessível junto ao material teórico.
+
+\section{Trabalhos relacionados}
+
+Dado o interesse humano pela música e a multidisciplinaridade inerente à esta dissertação, os trabalhos relacionados são numerosos.
+Assim, o Apêndice~\ref{cap:trabalhosRelacionados} é dedicado aos livros e implementações computacionais de interesse ou que apresentem similaridades com a descrição psicofísica de elementos musicais em termos do áudio digital.
+A visita indica aspectos inéditos deste trabalho, em especial a descrição analítica de elementos musicais básicos em termos das amostras sonoras e a descrição natural, formal e concisa de técnicas tradicionais da música.

### 2013-03-24 trabalhosRelacionados.tex
feita revisao completa da dissertacao, arrumando pormenores

@@
 
-\subsection{Livros}
+\subsection{Livros}\label{subsec:livros}
 
@@
 
-\subsection{Bibliotecas, linguagens e conjuntos de implementações computacionais voltados para música}
+\subsection{Bibliotecas, linguagens e conjuntos de implementações computacionais voltados para música}\label{subsec:bibs}
+{\bf Implementações fechadas.} As seguintes implementações fogem ao escopo do trabalho dada a inexistência de descrição amostral ou código que assegure uma implementação precisa e limpa.
+Como estas implementações são comerciais e o compromisso é com a satisfação do usuário, podem possuir pequenas adições de reverbs, delays e outros artifícios que, conhecidamente, fazem o resultado final soar mais agradável (chamado de \emph{'catchup'} no som ou na música).
+Observe que os plugins com funções adicionais às propostas pela classificação podem ser considerados adulterados.
+Exemplos desta adulteração são, em funções díspares do efeito especificado: aplicar um compressor de dinâmica para fazer o som ser ouvido como 'mais volumoso', aplicar reverb, dar ganho de equalização nos agudos ou nos graves, etc.
 
-\subsection{Aprofundamento sobre esta dissertação com base nos trabalhos visitados}
+\begin{enumerate}
+    \item \emph{Plugins da Waves}: vários plugins que simulam equipamentos consagrados de estúdio analógico, além de montagem e configurações típicas de músicos e produtores famosos.
+Há plugins concebidos/desenhados por artistas/produtores com notória competência.
+Há uma série de plugins de um só parâmetro.
+Há uma linha de compressores, \emph{gates} e \emph{de-essers}.
+Uma linha especial de \emph{plugins} de \emph{limiters}.
+Uma extensa linha de excelêntes \emph{plugins} de EQs.
+Uma linha de plugins de \emph{reverbs} e \emph{delays}.
+Outra de efeitos como Doppler, \emph{flanger}, \emph{morpher}, etc.
+Uma linha focada em \emph{plugins} de \emph{surround}.
+Uma linha bastante diferenciada de \emph{plugins} de remoção de ruído.
+Diversos \emph{plugins} para vocais.
+Alguns \emph{plugins} para alteração e modificação de altura.
+Outros analisadores.
+Por fim, alguns \emph{plugins} para criação de imagem estéreo e maximização da sensação do grave.
+    \item \emph{GRM Tools} é uma coleção de plugins para transformação sonora que atinge, na prática cotidiana, resultados únicos e poderosos.
+É um conjunto de efeitos bastante considerado na música eletroacústica.
+Pode interessar à dissertação como exemplo de implementações modelo.
+    \item A \emph{Steinberg} possui uma série de programas considerados da melhor qualidade, incluindo os sequenciadores Cubase/Nuendo e o editor de áudio Wavelab.
+Há também diversos plugins VST (que é o padrão de plugin de áudio mais difundido e é da Steinberg) para síntese e tratamento sonoro.
+A Steinberg está também fornecendo bancos de sons.
+    \item \emph{Synths} (i.e.
+sintetizadores) mais conhecidos: Reason e ReCycle são produtos da Propellerheads com bastante ênfase em música de pista e simulação computacional de equipamentos e configurações de canais de áudio em hardware.
+O Tassman e outros intrumentos da Applied Acoustics Systems proporcionam sons de qualidade impressionante.
+Os instrumentos da Native Instrumens são consagrados, em especial o sintetizador ABSynth, o Traktor para VJing e o Maschine para fazer batidas; além destes, o sintetizador modular Reaktor e o pacote Komplete são também bastante conhecidos.
+\end{enumerate}
+
+{\bf Implementações livres}. Podem ser consideradas documentações pois os códigos computacionais estão disponíveis e podem ser considerados especificações dos procedimentos envolvidos.
+
+\begin{enumerate}
+    \item \emph{SndObj} é uma biblioteca em C usada também como um módulo Python.
+O manual está claro: "o SndObj é uma unidade de programação que pode gerar sinais com especificações de áudio e parâmetros de controle".
+A biblioteca tem anos de existência, foi escrita e é mantida por Victor Lazzarini com contribuições da FFTW e do Frank Barknecht.
+A documentação, incluindo sítio, manual e códigos, lida com especificações técnicas da biblioteca como um todo e com as entradas e saídas de cada objeto.
+É complementar com relação às implementações dos Apêndices.
+Não possui descrição analítica dos procedimentos.
+    \item \emph{The MusicKit} é uma biblioteca em ObjC com mais de 10 anos de existência. Descreve-se como um sistema de \emph{software} orientado a objetos para criação musical. 
+    \item \emph{Abjad} é uma biblioteca Python que se define como um "sistema de \emph{software} interativo para ajudar compositores a escreverem complexas peças de notação musical de forma incremental e iterativa".
+O Abjad utiliza o Lilypond para renderizar as estruturas em notação musical.
+    \item \emph{music21}: esta biblioteca do MIT se descreve como um conjunto de ferramentas para musicologia auxiliada por computadores.
+Em grande parte, lida com a entrada e saída de elementos musicais em bancos de dados com obras dos compositores mais relevantes da música erudita.
+Trata principalmente de notação musical e análise de dados.
+Complementar à dissertação é esta própria ênfase em análise e notação musical.
+Diferencial que não trata de aspectos físicos do som ou sua representação amostrada no tempo, mas sim de aspectos que estão descritos pelo viés natural em "2.3 Notas em música".
+    \item Embora ainda possa ser usado como uma biblioteca, o \emph{AthenaCL} tem já características de uso que fogem à linguagem Python, com terminal próprio e outras interfaces.
+O autor, Christopher Arizza, também co-responsável pelo Music21, o define com um "sistema que é uma ferramenta computacional orientada a objetos e em código aberto".
+O manual possui diversas das ferramentas, os detalhes de uso e implementações.
+É complementar quanto ao trabalho o aspecto de ferramenta computacional com linguagem própria e ferramentas encapsuladas à disposição.
+Diferencial é que não tem um tratamento analítico e não se atém às amostras em si.
+    \item \emph{PD e Max/MSP}: são linguagens de domínio específico (para música) que aproveitam a interface das linguagem de patches.
+As caixinhas cujo texto especifica o objeto e cujas entradas e saídas são ligadas por cordinhas são tanto uma interface amigável quanto uma interface gráfica para utilização do programa.
+Ambas as linguagens foram feitas em grande parte por Miller Puckette, que cuida atualmente somente da versão livre, que é o PD.
+O Max/MSP é mantido pela empresa Cycling 74.
+Os manuais do Max/MSP são ótimas e até históricas introduções qualitativas e práticas.
+O livro do Puckette, citado em na lista de livros acima, é um dos materiais excelêntes que utilizam o PD.
+Complementar à dissertação há esta aproximação de um programa/sistema, com interface, e mais engessado do que uma linguagem de programação.
+Diferencial é que não há descrição analítica do que ocorre com as amostras.
+O código computacional, neste caso, é a documentação e Puckette explicita que o código do PD deva sempre ser completo se impresso em uma folha de papel.
+    \item \emph{Faust, Chuck, Supercollider, Nyquist, Impromptu}: linguagens de domínio específico como p PD e o Max/MSP, voltadas para a síntese de música em tempo real, mas escritas normalmente em linhas, não em patches.
+Complementar ao trabalho é que são implementações computacionais excelêntes para uso em tempo real.
+Diferencial é que há perda da desenvoltura para manipulações de áudio com controle amostral de precisão arbitrária.
+A descrição analítica também é inexistente.
+    \item \emph{CSound}: é uma a linguagem de programação voltada para música mais antiga que está em uso.
+As constantes produções acadêmicas em torno do CSound podem servir de descrições mais rigorosas do que as encontradas nas soluções anterires.
+De qualquer forma, não há uma documentação clara e concisa sobre os procedimentos básicos como o capítulo 2 da dissertação.
+    \item \emph{tuneR}: um pacote em R para análise de música. A documentação é um bom manual dos objetos e as características que extraem. 
+    \item \emph{Gwibber} e \emph{Vivace}: linguagens de programação voltadas para música (e video no caso do Vivace) utilizáveis via \emph{browser}.
+\end{enumerate}
+
+\subsection{Aprofundamento sobre esta dissertação com base nos trabalhos visitados nas subseções~\ref{subsec:livros} e~\ref{subsec:bibs}}
+
+Há livros de técnicas de música eletrônica, como o do Puckette.
+Já materiais como o livro do Pedro Koeger e a biblioteca music21 não tratam das amostras sonoras em momento algum, mas sim da representação e manipulação de estruturas musicais, assunto abordado na seção~\ref{notasMusica} do ponto de vista natural.
+Há também trabalhos focados em modelagens matemáticas de técnicas e princípios musicais, cuja consideração permeia toda a dissertação e aprofundamentos são bem vindos.
+Estes trabalhos, quando muito, apresentam princípios de sinais discretos, mas não se ocupam da descrição do comportamento amostral.
+Os trabalhos de Cook, Smith III e algumas implementações possuem descrições de fenômenos vibratórios em corpos rígidos e outras modelagens físicas.
+Pode-se propor expansões para usos conjuntos com alguns dos trabalhos, visitados ou não.
+Vale assinalar: com base na experiência prévia dos autores e nos trabalhos visitados em~\ref{subsec:livros} e~\ref{subsec:bibs}, a descrição analítica dos elementos musicais básicos em termos das amostras do áudio digital parece ser inédita.
+     Como há também a implementação em código livre das relações descritas nesta dissertação, abre-se um leque para experimentações científicas rigorosas assim como para usos artísticos com alta fidelidade e com precisão arbitrária.

### 2013-03-24 devEresultados.tex
feita revisao completa da dissertacao ateh 2.1

@@
 estas unidades compreendidas como "átomos" constituintes da música.\cite{Wisnick, Lovelock, Webern}
-Atualmente, estas notas
+Hoje, estas notas
 são tidas como um paradigma de proposta musical
@@
 Canonicamente, as notas possuem ao menos duração, volume, altura e timbre.\cite{Lacerda} Estas
-são qualidades tratáveis quantitativamente.\cite{Roederer}
-Assim, as características da nota musical digital básica são dadas pelas amostras igualmente espaçadas no tempo
-da onda mecânica que representam (veja seção~\ref{sec:audio} sobre áudio PCM).
+são qualidades tratáveis quantitativamente
+e ditadas pelas amostras igualmente espaçadas no tempo
+da onda mecânica que representam (veja seção~\ref{sec:audio} sobre áudio PCM).\cite{Roederer}
 

### 2013-03-24 introducao.tex
feita revisao completa da dissertacao ateh 2.1

@@
 Os resultados são relações matemáticas e suas implementações computacionais.
-Uma descrição teórica está no capítulo~\ref{cap:resultados} e o conjunto de \emph{scripts} disponibilizados no Apêndice~\ref{cap:codigoProc} e \emph{online}.\cite{MASSA} A caixa de ferramentas (\emph{toolbox}) resultante recebeu o nome \massa\ (música e áudio em sequências e séries amostrais) e foi utilizada para fazer pequenas peças e montagens focadas nos princípios expostos.
-O Apêndice~\ref{cap:codigoPecas} possui uma relação destas montagens assim como o diretório \emph{exemplos\_de\_uso} da \massa.
+Uma descrição teórica está no capítulo~\ref{cap:resultados} e o conjunto de \emph{scripts} disponibilizados no Apêndice~\ref{cap:codigoProc} e \emph{online}.
+A caixa de ferramentas (\emph{toolbox}) recebeu o nome \massa\ (música e áudio em sequências e séries amostrais) e foi utilizada para fazer pequenas peças e montagens focadas nos princípios expostos.
+O Apêndice~\ref{cap:codigoPecas} possui uma relação destas montagens assim como o diretório \emph{exemplos\_de\_uso} da \massa.\cite{MASSA}
 
@@
 Mesmo assim, a nota permanece paradigmática como 'unidade fundamental' das estruturas musicais e, na prática, pode se desdobrar em sons que contemplam estes desenvolvimentos recentes.
-A nota musical e sua expansão são abordadas na seção~\ref{varInternas}.\cite{Wisnick,Webern,Lerdahl,Cook}
+A definição e expansão da nota como unidade fundamental da música são abordados nas seções~\ref{varInternas} e~\ref{sec:notaDisc}, respectivamente.
+A seção~\ref{notasMusica} trata da organização das notas em estruras de alto nível.\cite{Wisnick,Webern,Lerdahl,Cook,Lacerda}
 
@@
 Os resultados apresentados desta dissertação incluem \emph{scripts}, i.e.
-pequenos programas para melhor disponibilidade e validação das tecnologias, que constituem a caixa de ferramentas \massa.
-Este conjunto de códigos está disponibilizado em domínio público através de repositórios Git públicos.\cite{gitBook}
+pequenos programas para melhor disponibilidade e validação das tecnologias.
+Estes constituem a caixa de ferramentas \massa, disponibilizada em domínio público através de repositórios Git abertos.\cite{gitBook}
 Os \emph{scripts} estão em Python e fazem uso das bibliotecas externas Numpy e Scikits/Audiolab que realizam chamadas à linguagem Fortran para maior eficiência computacional.
@@
 Dado o interesse humano pela música e a multidisciplinaridade inerente à esta dissertação, os trabalhos relacionados são numerosos.
-Assim, o Apêndice~\ref{cap:trabalhosRelacionados} é dedicado aos livros e implementações computacionais de interesse ou que apresentem similaridades com a descrição psicofísica de elementos musicais em termos do áudio digital.
+Assim, o Apêndice~\ref{cap:trabalhosRelacionados} é dedicado aos livros e implementações computacionais de interesse ou que apresentem similaridades com a descrição de elementos musicais em termos do áudio digital.
 A visita indica aspectos inéditos deste trabalho, em especial a descrição analítica de elementos musicais básicos em termos das amostras sonoras e a descrição natural, formal e concisa de técnicas tradicionais da música.

### 2013-03-25 devEresultados.tex
feito toda a revisao ateh a parte de intervalos

@@
 são qualidades tratáveis quantitativamente
-e ditadas pelas amostras igualmente espaçadas no tempo
-da onda mecânica que representam (veja seção~\ref{sec:audio} sobre áudio PCM).\cite{Roederer}
+e ditadas pelas amostras, igualmente espaçadas no tempo,
+da onda sonora (veja seção~\ref{sec:audio} sobre áudio PCM).\cite{Roederer}
 
@@
 do cone.
-Nestes pontos, a inferência do ângulo azimutal depende especialmente da filtragem atenuante nos agudos, pois a cabeça interfere um tanto mais nas ondas mecânicas agudas do que nas graves.\cite{Heeger,hrtf}
+Nestes pontos, a inferência do ângulo azimutal depende especialmente da filtragem atenuante nos agudos, pois a cabeça interfere um tanto mais nas ondas mecânicas agudas do que nas graves.\cite{Heeger,hrtf}  Também pertinente à audição de fonte lateral, quando o som é graves o suficiente, há uma difração e a onda chega ao ouvido $\approx 0,7ms$ depois.\cite{floEsp}
 
@@
 \subsubsection{Espacialização}
-Já a espacialização é, fundamentalmente, o resultado das reflexões e absorções da nota emitida nas superfícies do recinto/paisagem no qual a nota foi emitida.
-Pode-se simular reverberações reais não possíveis em sistemas reais.
-O som viaja a $343,2m/s$, e pode ser emitido da fonte com qualquer padrão de direcionalidade.
-Quando uma frente sonora encontra uma superfície do recinto, há uma reflexão.
+Já a espacialização é o resultado das reflexões e absorções do som nas superfícies do recinto/paisagem no qual a nota foi emitida.
+O som se propaga no ar a $\approx 343,2m/s$, e pode ser emitido da fonte com qualquer padrão de direcionalidade.
+Quando uma frente sonora encontra uma superfície, há uma reflexão.
 Nesta reflexão ocorrem tanto 1) a inversão da componente da velocidade de propagação que é perpendicular à superfície, quanto 2) a absorção de energia, especialmente nos agudos.
-As ondas se propagam até atingirem niveis considerados inaudíveis.
-Quando alguma frente de onda atingir o ouvido, é registrada e pode ser inteiramente descrita com: momento de chegada ao ouvido e o número de reflexões junto aos filtros de absorção de cada superfície que atingiu.
+As ondas se propagam até atingirem níveis inaudíveis.
+Quando alguma frente de onda atingir o ouvido, pode ser descrita com o momento de chegada ao ouvido e os filtros de absorção de cada superfície que atingiu.
+Pode-se simular reverberações não possíveis em sistemas reais.
 Para experimentações, pode-se usar reflexões assimétricas com relação ao eixo perpendicular à superfície, ou ainda ganhos em determinadas bandas de frequência (tidos como 'ressonâncias'), ambas as características não são encontradas em sistemas reais.
@@
 \begin{itemize}
-    \item Primeiro periodo: as 'primeiras reflexões' são mais intensas e esparças.
-    \item Segundo período: a 'reverberação tardia' é uma continuação de atrasos com um decaimento exponencial e ocorrências estatísticas.
+    \item Primeiro periodo: as 'primeiras reflexões' são mais intensas e esparsas.
+    \item Segundo período: a 'reverberação tardia' é praticamente uma sucessão densa de atrasos indistintos com um decaimento exponencial e ocorrências estatísticas.
     \item Primeira banda: o grave possui algumas frequências de ressonância relativamente espaçadas.
@@
 Smith III aponta que boas salas de concerto possuem um tempo total de reverberação de aproximadamente $1,9$ segundos.
-Aponta também que o período das primeiras reflexões deve ser de $0,1$ segundos.
-Estas quantidades apontam que, nas condições contempladas, há frentes de onda perceptíveis que se propagam até $652,08$ metros ($83.79k$ amostras em $f_a=44.1kHz$) antes de atingirem o ouvido.
-Além disso, as reflexões do som formam, para a audição, após a propagação por $34,32$ metros ($4.41k$ amostras em $f_a=44.1kHz$ ), um emaranhado cujas incidências são pouco distintas.
+Aponta também o período das primeiras reflexões de $0,1$ segundos.
+Estas quantidades sugerem que, nas condições contempladas, há frentes de onda perceptiveis que se propagam até $652.08$ metros ($83.79k$ amostras em $f_a=44.1kHz$) antes de atingirem o ouvido.
+Além disso, as reflexões do som formam, após a propagação por $34.32$ metros ($4.41k$ amostras em $f_a=44.1kHz$ ), um emaranhado cujas incidências são pouco distintas na audição.
 Estas primeiras reflexões são particularmente importantes para a sensação de espaço.
 A primeira incidência é o som direto, descrito por DTI e DII das equações~\ref{eq:dti} e~\ref{eq:dii}.
-Admitindo-se que cada uma das primeiras reflexões se propagará, ao menos, $3-30m$ dependentes das dimensões da sala, a separação entre as primeiras reflexões é de, ao menos, $8-90$ milissegundos, correspondendo a uma separação de aproximadamente $350-4000$ amostras.
-Verifica-se experimentalmente que o número de reflexões por aumenta em proporção quadrática  $ \approx k.n^2$.
+Admitindo que cada uma das primeiras reflexões, antes de chegar ao ouvido.
+se propagará, ao menos, $3-30m$ dependentes das dimensões da sala, a separação entre as primeiras reflexões é de, ao menos, $8-90$ milissegundos ($\approx 350-4000$ amostras em $f_a=44.1kHz$).
+Verifica-se experimentalmente que o número de reflexões aumenta em proporção quadrática  $ \approx k.n^2$.
 Apontamentos do uso de convoluções e filtragens para facilitar estas implementações estão na subseção~\ref{subsec:mus2}, especialmente nos parágrafos sobre reverberação.
@@
 \subsubsection{Reverberação}\label{subsubsec:reverb}
-Retomando também a exposição da subseção~\ref{subsec:spac}, a reverberação tardia pode ser modelada como a convolução com um trecho de ruído colorido rosa, marrom ou preto, em decaimento exponencial de amplitude com relação também ao tempo.
-A atenuação nos agudos e irregularidade suave na resposta em frequência são contempladas de forma bastante satisfatória.\cite{JOSPhy,guillaume} Linhas de atrasos podem ser adicionadas como prefixo à ao ruído com o decaimento, assim comtemplando as duas partes temporais da reverberação: as primeiras reflexões e a reverberação tardia.
-Pode-se, para melhora qualidade, calcular geometricamente a localização da última superfície em que cada frente de onda refletiu antes de chegar ao ouvido nos primeiros $100-200$ milissegundos e aplicar um filtro passa-baixas como em~\ref{subsec:filtros}.
+Com a mesma nomenclatura da subseção~\ref{subsec:spac} sobre espacialização, a reverberação tardia pode ser modelada como a convolução com um trecho de ruído colorido rosa, marrom ou preto, em decaimento exponencial de amplitude com relação também ao tempo.
+Assim, a atenuação nos agudos e irregularidade suave na resposta em frequência são contempladas de forma bastante satisfatória.\cite{JOSPhy,guillaume} Linhas de atrasos podem ser adicionadas como prefixo à ao ruído com o decaimento, assim comtemplando as duas partes temporais da reverberação: as primeiras reflexões e a reverberação tardia.
+Pode-se, para melhora de qualidade, calcular geometricamente a localização da última superfície em que cada frente de onda refletiu antes de chegar ao ouvido nos primeiros $100-200$ milissegundos e aplicar um filtro passa-baixas como em~\ref{subsec:filtros}.
 O ruído colorido pode ser introduzido gradualmente, desde o instante inicial dado pela incidência direta (i.e.
@@
 Esta precisão absoluta é característica de implementações
-computacionais, pois os intrumentos reais possuem desvios destas frequências para melhor compatibilizar os harmônicos
-de suas notas. Além disso, o fator fixo $\varepsilon=2^{\frac{1}{12}}$ caracteriza
+computacionais, os intrumentos reais possuem desvios destas frequências para melhor compatibilizar os harmônicos
+de suas notas. Além disso, a referência fixa $\varepsilon=2^{\frac{1}{12}}$ caracteriza
 a afinação de temperamento por igual. Há afinações com intervalos propostos como razões de inteiros de baixa ordem, com
-fundamentação em leis físicas, e foram propostas antes do advento do temperamento por igual.\cite{Roederer}
+fruto da observação de comportamentos físicos.  As primeiras destas afinações foram formalizadas por volta de 2 mil anos antes do advento do temperamento por igual.\cite{Roederer}
 
@@
 \begin{itemize}
-    \item A {\bf intonação justa} consiste em notas da escala diatônica descritas como razão de inteiros de pequena ordem como apontados pela série harmônica.
+    \item A {\bf intonação justa} consiste em notas da escala diatônica associadas a razões de inteiros de pequena ordem como apontados pela série harmônica.
 As razões básicas estão contidas no modo jônico (dó a dó nas teclas brancas do piano, veja abaixo na subseção~\ref{subsec:escalas}): 1, 9/8, 5/4, 4/3, 3/2, 5/3, 15/8, 2/1.
@@
 Note que se $S_i$ é uma sequência de alturas relacionadas por $\varepsilon_1$,
-um mapeamento para notas em $\varepsilon_2$
-constitui uma nova sequência $S_i'$ com $s_i'=s_i \frac{\varepsilon_1}{\varepsilon_2}$.
+um mapeamento para notas relacionadas por $\varepsilon_2$
+constitui uma nova sequência $S_i'=\left\{s_i'\right\}=\left\{ s_i \frac{\varepsilon_1}{\varepsilon_2}\right\}$.
 A montagem musical \emph{Micro tom} explora recursos microtonais e seu código está no Apêndice~\ref{ap:micro}, assim como na \massa\ online.
@@
 O segundo caso especial é o trítono (4aum, 5dim, tri).
-Este é consonante em algumas culturas, já para a música tonal, o trítono indica dominante e busca fundamentalmente sua resolução em uma terça ou sexta e, por esta instabilidade, é considerado intervalo dissonante.}
+Este é consonante em algumas culturas, já para a música tonal, o trítono indica dominante e busca sua resolução em uma terça ou sexta e, por esta instabilidade, é considerado intervalo dissonante.}
 \begin{tabular}{| c | c | c | }\hline

### 2013-03-25 devEresultados.tex
adiantada dissertacao ateh estruts ciclicas

@@
 O segundo caso especial é o trítono (4aum, 5dim, tri).
-Este é consonante em algumas culturas, já para a música tonal, o trítono indica dominante e busca sua resolução em uma terça ou sexta e, por esta instabilidade, é considerado intervalo dissonante.}
+Este é consonante em algumas culturas.
+Já para a música tonal, o trítono indica dominante e busca sua resolução em uma terça ou sexta e, por esta instabilidade, é considerado intervalo dissonante.}
 \begin{tabular}{| c | c | c | }\hline
@@
 Nona, décima, décima primeira, etc, são os intervalos compostos, de uma ou mais oitavas + um intervalo dentro da oitava, que caracteriza o intervalo composto.
-São representados pelos dígitos numéricos: 1, 3, 5 é um uníssono, uma terça, e uma quinta.
+Os intervalos são representados pelos dígitos numéricos, e.g.
+1, 3, 5 é um uníssono, uma terça e uma quinta.
         \item Qualidades de cada intervalo: as consonâncias perfeitas - i.e.
@@
 A ordem das notas, primeiro a nota mais grave ou a mais aguda, resulta em um intervalo ascendente ou descendente, respectivamente.
-        \item Passando a nota mais grave para a oitava acima, ou a nota mais aguda uma oitava para baixo, o intervalo é invertido.
+        \item Passada a nota mais grave para a oitava acima, ou a nota mais aguda uma oitava para baixo, o intervalo é invertido.
 Um intervalo, somado à sua inversão, resulta 9 (7m inverte para 2M: $7m+2M=9-$).
@@
         \item Um intervalo maior que a oitava é dito 'intervalo composto' e é classificado como o intervalo entre as mesmas notas, mas na mesma oitava.
-São também especificados por 7 acrescido deste intervalo: 11J é uma oitava mais uma quarta (7+4J==11J), 9M é uma oitava mais uma segunda (7+2M=9M).
+São também especificados por 7 acrescido deste intervalo: 11J é uma oitava mais uma quarta (7+4J==11J), 9M é uma oitava mais uma segunda maior (7+2M=9M).
 \end{itemize}
@@
 Assim, em uma escala de dó bemol maior, a tônica - primeiro grau - é dó bemol, não si, e a sensível - sétimo grau - é si bemol, não lá sustenido ou dó duplo bemol.
-De forma semelhante, o segundo grau de uma escala pode estar a um semitom do primeiro grau, assim como a sensível (sétimo grau a um semitom ascendente do primeiro), momento no qual há uma terça diminuta entre os dois semitons do sétimo e segundo grau da escala pois o primeiro grau está entre os dois graus próximos a ele: segundo e sensível.\cite{Lacerda}
+De forma semelhante, o segundo grau de uma escala pode estar a um semitom do primeiro grau, assim como a sensível (sétimo grau a um semitom ascendente do primeiro grau), momento no qual há uma terça diminuta entre os dois semitons do sétimo e segundo grau da escala pois o primeiro grau está entre os dois graus próximos a ele: segundo e sensível.\cite{Lacerda}
 
@@
 Uma escala é um conjunto ordenado de alturas. Usualmente, as escalas se repetem
-a cada oitava. A sequências ascendente com todas as notas da divisão da oitava em 12 intervalos iguais
-separados pela razão $\varepsilon=2^{\frac{1}{12}}$ é a escala cromática de temperamento igual.
+a cada oitava. A sequência ascendente com todas as notas da divisão da oitava em 12 intervalos iguais,
+separados pela razão $\varepsilon=2^{\frac{1}{12}}$, é a escala cromática de temperamento por igual.
 Há 5 divisões perfeitamente simétricas da oitava dentro da escala cromática.
@@
 repetidas de forma eficiente e sustentada.
-Abaixo, na seção~\ref{estCic}.
+Abaixo, na seção~\ref{estCic}, será retomado o assunto das simetrias.
 A montagem \emph{Cristais} expõe cada uma destas escalas tanto melódica quanto harmonicamente e seu código está no Apêndice~\ref{ap:cristais}.
@@
 
-O traçado ascendente e descendente da escala menor melódica é necessário para a presença da sensível (sétimo e último grau separado por um semitom da oitava e realça a polarização na tônica) no trajeto ascendente, o que não é necessário quando descende, retomando a forma natural.
+O traçado ascendente e descendente da escala menor melódica é necessário para a presença da sensível (sétimo e último grau, separado por um semitom da oitava, realça a polarização na tônica) no trajeto ascendente, o que não é necessário quando descende, retomando a forma natural.
 Já a escala harmônica apresenta a sensível, mas não evita o intervalo de segunda aumentada entre o sexto e sétimo graus, pois não precisa contemplar trajetória melódica, apenas apresentar a sensível tão cara ao sistema tonal (a sensível tende à tônica, afirmando-a).\cite{Harmonia}
@@
     \item Todo trítono, especialmente se formado entre uma terça maior e uma sétima menor, tende a resolver em uma terça ou uma sexta.
-    \item Na repetição, a ordem de preferência é: a fundamental, a quinta, a terça e a sétima.
+    \item Evita-se a duplicação de nota. Na necessidade de duplicação, a ordem de preferência é: a fundamental, a quinta, a terça e a sétima.
     \item Pode-se formar acordes com notas diferentes das triádicas, especialmente se dentro de alguma lógica recorrente ou encadeamento musical que justifique as notas diferentes.
@@
 \subsubsection{Harmonia atonal}
-Omitir os encadeamentos básicos do sistema tonal é a chave para a obtenção de 
+A omissão dos encadeamentos básicos do sistema tonal é chave para a obtenção de 
 harmonias modal e atonal. No caso desta ausência de estruturas tonais mínimas,
@@
 Nesta classificação, a harmonia modal não é tonal e não é atonal.
-A harmonia modal está reduzida à incidência de notas dentro de escalas diatônicas.
+A harmonia modal está reduzida à incidência de notas dentro de escalas diatônicas (ou simplificações) e à ausência de estruturas tonais.
 Pode-se perceber, pelo conceito, que a harmonia atonal é difícil de ser realizada.\cite{harmEXT}
@@
 \begin{itemize}
-    \item Repetir notas, como artifício de sua incidência anterior, não adiciona informação relevante.
-    \item Soar notas adjacentes ao mesmo tempo, em bloco.
-    \item Apresentar durações livremente, desde que respeitada ordem de aparição das notas.
+    \item Repetir notas. Ao considerar a repetição imediata um prolongamento da incidência anterior, notas iguais, sem outras notas entre elas, não adicionam informação relevante.
+    \item Soar notas adjacentes ao mesmo tempo, formando intervalos harmônicos e acordes.
+    \item Apresentar durações livremente, desde que respeitada a ordem de aparição das notas.
     \item Para variação, além dos recursos de ampliação, transposição e translação, são usados o inverso, o retrógrado e o retrógrado do inverso.
@@
 as concepções de tonalidade e harmonia. Ainda assim, a harmonia tonal tem forte presença
-nas vertentes artísticas e comerciais. O próprio dodecafonismo considera-se de natureza
-tonal pois consiste na negação de suas características de polarização.
+nas vertentes artísticas e comerciais. Considera-se o próprio dodecafonismo de natureza
+tonal pois consiste na negação das características tonais de polarização.
 
 Na música tonal ou modal, acordes, como os listados nas equações~\ref{triades}, constituídos com a fundamental em cada
-grau de escalas especificadas nas equações~\ref{eq:escalas}, formam os pilares do campo harmônico.
+grau de uma escala, como das equações~\ref{eq:escalas}, formam os pilares do campo harmônico.
 A observação de progressões incidentes de acordes e regras para encadeamentos é o objeto de estudo da harmonia musical.
@@
 Na 'harmonia tonal tradicional',
-a escala pode ter como tônica (primeiro grau da escala) qualquer nota e pode ser maior (com as mesmas notas do modo jônico) ou menor (as notas do eólico constituem a escala "menor natural" e esta possui versões harmônica e melódica, veja a equação~\ref{eq:escalasMenores}.
+a escala pode ter como tônica (primeiro grau da escala) qualquer nota e pode ser maior (com as mesmas notas do modo jônico) ou menor (as notas do eólico constituem a escala "menor natural" e esta possui versões harmônica e melódica, veja a equação~\ref{eq:escalasMenores}).
 A escala escolhida é
@@
 As exatas funções e efeitos musicais destes acordes é motivo de bastante controvérsia.
-A tabela~\ref{tab:harmonia} mostra a relação entre as tríades formadas em cada grau da escala.
+A tabela~\ref{tab:harmonia} mostra a relação entre as tríades formadas em cada grau da escala maior.
 
@@
 
-A dominante anti-relativa forma um acorde menor, exigindo a alteração do quarto grau um semitom para cima $\hat{7}\#$.
+É necessário que a dominante anti-relativa forme um acorde menor, por isso a alteração do quarto grau um semitom para cima $\hat{7}\#$.
 O acorde diminuto $\hat{7},\hat{2},\hat{4}$, geralmente é considerado uma 'dominante com sétima e sem fundamental'.\cite{Koellheuteur}
-
-Na escala menor, a relativa é a tríade iniciada uma terça acima e a anti-relativa uma terça abaixo,
-ou seja, se invertem com relação à tabela~\ref{tab:harmonia}. Outro detalhe destas tríades em modo
-menor é a alteração do $\hat{7}$ um semitom acima para que fique somente a um semitom da tônica,
+Em modo
+menor, há a alteração do $\hat{7}$ por um semitom ascendente para que haja somente um semitom de separação com a tônica,
 permitindo a dominante (que deve ser maior e tender à tônica).
-Assim, a dominante é sempre maior, tanto em escalas maior ou menor.
-Por este motivo, mesmo em tom menor, a dominante relativa permanece a uma terça abaixo e a antirelativa a uma terça acima.
+Assim, a dominante é sempre maior, tanto em escalas maior ou menor e, por este motivo, mesmo em tom menor, a dominante relativa permanece a uma terça abaixo e a antirelativa a uma terça acima.
 
 \subsubsection{Expansão tonal: funções individuais e medianas cromáticas}
-Cada um destes acordes pode ser confirmado e se desenvolver com uma execução de sua dominante ou subdominante individual (acorde com base na tríade formada a uma quinta acima e uma quinta abaixo respectivamente).
-Estas dominantes/subdominantes indiduais, por sua vez, possuem também subdominantes e dominantes individuais passíveis de uso.
+Cada um destes acordes pode ser confirmado e se desenvolver com uma execução de sua dominante ou subdominante individual, que é o acorde com base na tríade formada a uma quinta acima ou uma quinta abaixo respectivamente.
+Estas dominantes/subdominantes individuais, por sua vez, possuem também subdominantes e dominantes individuais passíveis de uso.
 Assim, em uma dada tonalidade, pode ocorrer qualquer acorde, por mais distante que seja do campo harmônico e das notas da escala, desde que a ocorrência apresente um percurso coerente de dominantes e subdominantes até a tonalidade de origem.
@@
 A modulação é a mudança da tonalidade em que se encontra a musica. Caracteriza-se a modulação através da observação
-das tonalidades de partida e chegada e da forma de transição. As tonalidades são sempre relacionadas por quintas e suas relativas e antirelativas. São formas de efetuar a modulação:
+das tonalidades de partida e chegada e da forma de transição. As tonalidades são sempre tidas como relacionadas por quintas e suas relativas e antirelativas. São formas de efetuar a modulação:
 \begin{itemize}
@@
 Por vezes chamada de modulação frasal ou modulação sem preparação.
-    \item O uso cuidadoso de uma dominante individual, e possivelmente também a subdominante individual, para afirmar a mudança da tônica e campo harmônica.
-    \item Uso de alterações cromáticas para um acorde da nova tonalidade. Chamada de modulação cromática.
+    \item O uso cuidadoso de uma dominante individual, e possivelmente também a subdominante individual, para afirmar a mudança da tônica e campo harmônico.
+    \item Uso de alterações cromáticas para atingir um acorde da nova tonalidade a partir de algum acorde da tonalidade anterior. Chamada de modulação cromática.
     \item O destaque para uma única nota, possivelmente repetida ou suspensa sem acompanhamento, comum às tonalidades de saída e chegada, constitui uma forma peculiar de introduzir o novo campo harmônico.
@@
 A condução de linhas melódicas simultâneas, chamadas vozes,
-é matéria do campo de estudo chamado contraponto. A bibliografia
+é o contraponto. A bibliografia
 percorre formas sistemáticas de condução de vozes e desemboca em gêneros escolásticos como cânones, invenções e fugas.
@@
 Na presença de 3 ou mais vozes,
-a importância melódica recai sobre as vozes mais aguda e mais grave, neta ordem.\cite{Fux,Tragtenberg,SchoenbergContra}
+a importância melódica recai sobre as vozes mais aguda e mais grave, nesta ordem.\cite{Fux,Tragtenberg,SchoenbergContra}
 
@@
 \subsection{Ritmo}\label{subsec:ritmo}
-A noção rítmica é dependente de eventos separados por durações.\cite{Lacerda} Estes eventos podem ser ouvidos individualmente se separados
-por ao menos $50-63ms$. Para que a separação entre eles possa ser apreciada como duração, ela deve ser maior,
+A noção rítmica é dependente de eventos separados por durações.\cite{Lacerda} Estes eventos podem ser ouvidos individualmente se espaçados
+por ao menos $50-63ms$. Para que a separação temporal entre eles possa ser apreciada como duração, ela deve ser maior,
 por volta de $100ms$.\cite{microsound} Pode-se sumarizar 
@@
 
-A acentuação em tempo fraco é o contratempo, as notas iniciadas em tempo fraco e cuja duração se prolonga sobre tempos fortes são as síncopas.
+A acentuação em tempo fraco é o contratempo, uma nota iniciada em tempo fraco e cuja duração se prolonga sobre um tempo forte é uma síncopa.
 
@@
 
-Cada nível $j$ possui alguns índices $i$. Sempre que $i$ possuir três valores diferentes
-(ou múltiplo de três) índices há uma relação perfeita, 
-quando é somente múltiplo dois, quatro ou oito índices há uma relação imperfeita, como na figura~\ref{fig:pulsoSubAgl}.
+Cada nível $j$ possui alguns índices $i$. Sempre que o índice $i$ possuir três valores diferentes
+(ou múltiplo de três), há uma relação perfeita. 
+Quando $i$ possuir somente múltiplo dois, quatro ou oito valores diferentes, há uma relação imperfeita, como na figura~\ref{fig:pulsoSubAgl}.
 
@@
 Os arcos músicais são estruturas abstratas e passíveis de operações básicas. Um arco espectral, como um
-acorde, pode ser invertido, ampliado, permutado, por exemplo. Os arcos temporais, como uma melodia, um motivo, um compasso ou
+acorde, pode ser invertido, ampliado e permutado, por exemplo. Os arcos temporais, como uma melodia, um motivo, um compasso ou
 uma nota, são igualmente passíveis de variações. Lembrando que $S_j=\left\{s_j=T_i^j=\{t_i^{j}\}_0^{\Lambda_j-1}\right\}_0^{H-1}$ é uma
@@
     \item Reversão temporal consiste em gerar uma sequência com os elementos em ordem invertida da sequência original $S_j$, assim: $S_j'=\left\{s_j'\right\}_0^{H-1}=\left\{s_{(H-j-1)}\right\}_0^{H-1}$.
-    \item A translação em altura é o deslocamento $\tau$ do material para uma frequência/altura diferente $\Xi'=\Xi + \tau$ da música.
+    \item A translação em altura é o deslocamento $\tau$ do material para uma altura diferente $\Xi'=\Xi + \tau$ da música.
 É uma variação com modificação na localização em altura do material: $\left\{s_j'\right\}=\left\{s_j^{\Xi'}\right\}=\left\{s_j^{\Xi+\tau}\right\}$ onde $\Xi$ é a altura do trecho $S_j$ ou do primeiro evento $s_0$ da estrutura $S_j$ original.
@@
     \item A inserção e remoção de materiais na estrutura $S_j$ pode ser ornamental ou estruturante: $S_j'=\{s_j'\}=\{s_j \text{ se condição A, caso contrário } r_j\}$, para qualquer material musical $r_j$, incluindo o instante vazio.
-Elementos de $R_j$ podem ser inseridos no começo, como uma prefixo de $S_j$, ao final, como um sufixo, ou no meio, dividindo $S_j$ ou fazendo dele o prefixo e o sufixo.
+Elementos de $R_j$ podem ser inseridos no começo, como um prefixo de $S_j$, ao final, como um sufixo, ou no meio, dividindo $S_j$ ou fazendo dele o prefixo e o sufixo.
 Os dois materiais podem se misturar das formas mais diversas.
     \item Modificações de articulação, instrumentação e espacialização $s_j'=s_j^{*_j}$, com $*_j$ a nova característica incorporada pelo elemento $s_j'$.
-    \item Acompanhamento. Tanto a instrumentação quanto as linhas melódicas que presentes na ocorrência de $S_j$ podem sofrer modificações e ser considerada uma variação de $S_j$ em si.
+    \item Acompanhamento. Tanto a instrumentação quanto as linhas melódicas presentes na ocorrência de $S_j$ podem sofrer modificações e ser considerada uma variação de $S_j$ em si.
 \end{itemize}
@@
 Com estes procedimentos, outros são derivados: o retrógrado do inverso, uma contração temporal com um sufixo externo, etc.
-A estruturas musicais ressoam no sistema cognitivo devido à própria natureza do pensamento. Em suas várias facetas,
+As estruturas musicais ressoam no sistema cognitivo devido à própria natureza do pensamento. Em suas várias facetas,
 uma ideia lida com o mesmo número de elementos e aspectos conectivos entre eles.
@@
 
-Como princípios gerais que orientam a criação de novos materiais musicais, utiliza-se estruturas paradigmáticas.
+As estruturas paradigmáticas orientam a criação de novos materiais musicais.
 Uma delas, central, é o dipolo tensão/relaxamento.
-Relaciona-se com este dipolo, cada outro dipolo tradicional: tônica/dominante, repetição/variação, consonância/dissonância, coerência/rompimento, simetria/assimetria, chegada/saída, perto/longe, parado/em movimento, etc.
+Relaciona-se com este dipolo, cada outro dipolo tradicional: tônica/dominante, repetição/variação, consonância/dissonância, coerência/rompimento, simetria/assimetria, igualdade/diferença, chegada/saída, perto/longe, parado/em movimento, etc.
 Já as relações ternárias tendem a se relacionar com o circular e a unificação.
-A subseção seguinte é dedicada aos arcos direcionais e, 
-no seguinte, os arcos cíclicos.
+A lúcida comunhão ternária, 'modus perfectus', se opõe ao dicotômico passional, 'modus imperfectus'.
+A seguir está uma exposição  dedicada aos arcos direcionais 
+e cíclicos.
 
@@
 Segundo conceitos musicais, o clímax está exatamente no meio da sequência.
-Pode-se modificar isso com a substituição por sequências de tamanhos diferentes.
+Pode-se modificar isso com o uso de sequências de tamanhos diferentes.
 Toda a teoria matemática de sequências, já estabelecida e ensinada corriqueiramente em cursos de cálculo III, pode ser utilizada para geração destes arcos.\cite{Guidorizzo,Schoenberg}
-Teoricamente, estas sequências aplicadas desta forma
-a qualquer característica dos eventos musicais produz arcos,
+Teoricamente, estas sequências, aplicadas desta forma
+a qualquer característica dos eventos musicais, produzem arcos,
 pois implicam no afastamento e retorno de uma parametrização inicial. 
@@
 
-Na prática, e historicamente, tem especial importância a razão áurea e, por isso, à série de Fibonacci.
-A série de lucas é uma generalização da série de Fibonacci e pode ser compreendida facilmente.
-Dados dois números quaisquer $x_0$ e $x_1$, obtem-se a série de Lucas: $x_n=x_{n-1}+x_{n-2}$.
+Na prática, e historicamente, tem especial importância a razão áurea e, por isso, a sequência de Fibonacci.
+A sequência de Lucas permite uma generalização da sequência de Fibonacci que pode ser compreendida facilmente.
+Dados dois números quaisquer $x_0$ e $x_1$, obtem-se a sequência de Lucas: $x_n=x_{n-1}+x_{n-2}$.
 Quanto maior for $n$, mais a razão $\frac{x_{n}}{x_{n_1}}$ se aproxima da razão áurea: $1.61803398875...$.
-A série converge rápido mesmo para valores iniciais bem discrepantes.
+A sequência converge rápido mesmo para valores iniciais bem discrepantes.
 Seja $x_0=1$ e $x_1=100$, e $y_n=\frac{x_n}{x_{n+1}}$ uma sequência auxiliar.
@@
 O entendimento filosófico de que o pensamento humano é fundamentado
-na percepção de semelhanças e diferenças dentre os estímulos
-e objetos coloca
+na percepção de semelhanças e diferenças, dentre os estímulos
+e objetos, coloca
 as simetrias no cerne do processo cognitivo.\cite{Deleuze}
@@
 Assim, uma permutação finita $p$ aplicada sucessivamente atinge a ordenação inicial dos elementos, formando um ciclo.
-Este ciclo, se utilizado para parametros de notas musicais, implica em um arco musical cíclico.
+Este ciclo, se utilizado para parâmetros de notas musicais, implica em um arco musical cíclico.
 

### 2013-03-25 devEresultados.tex
terminada correcao primeira

@@
 A utilização de permutações na música pode ser resumida da seguinte forma:
-seja $L_i$ uma sequência de eventos musicais (e.g. notas) e $p$ uma permutação.
-$L_i'=p(L_i)$ consiste nos mesmos elementos de $L_i$ mas em ordem diferente.
+seja $S_i=\{s_i\}$ uma sequência de eventos musicais $s_i$ (e.g. notas) e $p$ uma permutação.
+$S_i'=p(S_i)$ consiste nos mesmos elementos de $S_i$ mas em ordem diferente.
 As permutações podem ser escritas em duas notações: cíclica ou natural. 
@@
 
-Não é necessário permutar os elementos de $L_i$, mas somente
+Não é necessário permutar os elementos de $S_i$, mas somente
 alguma ou algumas de suas características. Assim, seja $p^f$ uma permutação 
-nas frequências e $L_i$ uma sequência de notas básicas como expostas
-ao final de~\ref{notaBasica}. A nova sequência $L_i'=p^f(L_i)$ consiste nas mesmas
+nas frequências e $S_i$ uma sequência de notas básicas como expostas
+ao final de~\ref{notaBasica}. A nova sequência $S_i'=p^f(S_i)=\left\{s_i^{p(f)}\right\}$ consiste nas mesmas
 notas musicais, na mesma ordem e com as mesmas características, com as frequências fundamentais permutadas segundo o padrão que $p^f$ apresenta.
@@
 Duas sutilezas deste procedimento.
-1) A permutação $p$ não precisa envolver todos os elementos de $L_i$, i.e. ela
-pode operar em subconjuntos de $L_i$. 2) Nem todos os elementos $l_i$ precisam ser executados a cada consulta de estado realizada.
-Para exemplificar, seja $L_i$ 
-uma sequência de notas musicais $l_i$. 
+1) A permutação $p$ não precisa envolver todos os elementos de $S_i$, i.e. ela
+pode operar em subconjuntos de $S_i$. 2) Nem todos os elementos $s_i$ precisam ser executados a cada consulta de estado realizada.
+Para exemplificar, seja $S_i$ 
+uma sequência de notas musicais $s_i$. 
 Se $i$ vai de $0$ a $n$, e $n>4$, a cada compasso
 de $4$ notas pode-se executar as primeiras $4$ notas. As outras notas de
-$L_i$ podem incidir nos compassos em que as permutações aloquem
-estas notas para as primeiras quatro notas de $L_i$.
-
-Cada uma destas permutações
-$p_i$ possui, ao menos, segundo a exposição acima: dimensões das notas em que opera (frequência, duração, \emph{fades}, intensidade, etc) e período de incidência (a cada quantas consultas é aplicada a permutação).
-Na realização das notas de $L_i$, uma forma fácil e coerente é executar as primeiras $n$ notas\footnote{A execução de notas disjuntas de $L_i$ equivale a modificar a permutação e executar as primeiras notas.}.
+$S_i$ podem incidir nos compassos em que as permutações aloquem
+estas notas para as primeiras quatro notas de $S_i$.
+
+A cada uma destas permutações
+$p_i$, segundo a exposição acima, relaciona-se: dimensões das notas em que opera (frequência, duração, \emph{fades}, intensidade, etc) e período de incidência (a cada quantas consultas é aplicada a permutação).
+Na realização das notas de $S_i$, uma forma fácil e coerente é executar as primeiras $n$ notas\footnote{A execução de notas disjuntas de $S_i$ equivale a modificar a permutação e executar as primeiras notas.}.
 

### 2013-03-26 trabalhosRelacionados.tex
pequena arrumada no apendice G

@@
-\chapter{Trabalhos relacionados e caracterização das contribuições destre trabalho}
+\chapter{Trabalhos relacionados e caracterização das contribuições deste trabalho}
 \label{cap:trabalhosRelacionados}
@@
 
-Vale observar que, embora estas rotinas estejam presentes em diversas implementações livres e proprietárias, suas descrições precisas se encontram somente em código computacional.
+Embora as rotinas, apresentadas nesta dissertação, estejam presentes em diversas implementações livres e proprietárias, suas descrições precisas se encontram somente em código computacional.
 A maior contribuição desta dissertação é exatamente a descrição analítica das qualidades básicas que compõem elementos musicais no áudio digital.
 A apresentação didática dos fenômenos envolvidos também não foi encontrada na literatura visitada, o que, junto com as implementações em código Python destas relações e de peças musicais que as exemplifiquem, forma uma contribuição simples e convidativa embora inédita e multidisciplinar.
-Vale apontar que, no início da escrita desta dissertação, a caixa de ferramentas não estava prevista, ela foi fruto das equações e descrições precisas, o que tornou imediata a escrita dos scripts que compõem a \emph{toolbox} \massa.
+No início da escrita desta dissertação, a caixa de ferramentas não estava prevista, ela foi fruto das equações e descrições precisas, o que tornou imediata a escrita dos scripts que compõem a \emph{toolbox} \massa.
 
@@
 Faz vínculos de apreensão de estruturas com peças de Josquin des Prez, Bach, Rachmaninoff e Steve Reich.
-            \item {\bf Aspecto complementar: as formalizações de operações dentro da notação tradicional são preciosos adendos às questões naturais abordadas na presente dissertação.
+            \item {\bf Aspecto complementar:} as formalizações de operações dentro da notação tradicional são preciosos adendos às questões naturais abordadas na presente dissertação.
 A notação em si capta aspectos estruturais do sistema tonal e de 12 notas.
-Além disso, com a notação abre-se uma ponte com as tradições musicais eruditas.}
-            \item {\bf Aspecto diferencial: o livro não desenvolve descrição precisa de aspectos musicais do som em si e não há foco em relacionar qualidades psicofísicas aos elementos musicais.}
+Além disso, com a notação abre-se uma ponte com as tradições musicais eruditas.
+            \item {\bf Aspecto diferencial:} o livro não desenvolve descrição precisa de aspectos musicais do som em si e não há foco em relacionar qualidades psicofísicas aos elementos musicais.
         \end{itemize}

### 2013-03-26 devEresultados.tex
liberado primerio paragrafo de harmonia para a definicao dos temas

@@
 \subsection{Harmonias atonal, tonal, expansão e modulação}\label{subsec:harmonia}
-\subsubsection{Harmonia atonal}
 A omissão dos encadeamentos básicos do sistema tonal é chave para a obtenção de 
@@
 
+\subsubsection{Harmonia atonal}
 De fato, as técnicas de música atonal

### 2013-04-02 introducao.tex
arrumacoes do prof. Chu na intro

@@
 Os \emph{scripts} estão em Python e fazem uso das bibliotecas externas Numpy e Scikits/Audiolab que realizam chamadas à linguagem Fortran para maior eficiência computacional.
-Parte deste código foi transcrito para JavaScript e Python nativos com facilidade, o que aponta para um uso destas contribuições em navegadores como o Firefox e o Chromium.\cite{numpy,audiolab,tutPython,python}
+Parte deste código foi transcrita para JavaScript e Python nativos com facilidade, o que aponta para um uso destas contribuições em navegadores como o Firefox e o Chromium.\cite{numpy,audiolab,tutPython,python}
 
@@
 
-Dado o interesse humano pela música e a multidisciplinaridade inerente à esta dissertação, os trabalhos relacionados são numerosos.
+Dado o interesse humano pela música e a multidisciplinaridade inerente a esta dissertação, os trabalhos relacionados são numerosos.
 Assim, o Apêndice~\ref{cap:trabalhosRelacionados} é dedicado aos livros e implementações computacionais de interesse ou que apresentem similaridades com a descrição de elementos musicais em termos do áudio digital.

### 2013-04-02 devEresultados.tex
feitas correcoes do prof. Chu no cap 2 de dev

@@
 
-Ou seja, é necessário pouco mais que triplicar a amplitude para um volume drobrado.
+Ou seja, é necessário pouco mais que triplicar a amplitude para um volume dobrado.
 Estes valores servem de guia para os aumentos e diminuições dos valores absolutos que compõem as
@@
 Os harmônicos relativamente atenuados da onda triangular
-a faz a mais funcional - dentre as citadas - para ser usada sem nenhum tratamento na síntese de notas musicais.
-
-Para citar também uma utilidade da onda quadrada, ela pode ser usada em uma síntese
+a fazem a mais funcional - dentre as citadas - para ser usada sem nenhum tratamento na síntese de notas musicais.
+
+Já a onda quadrada, pode ser usada na síntese
 subtrativa que vise a imitar um clarinete. Este instrumento também só apresenta os
@@
 Se $t_a=44,1 kHz$, como no padrão PCM de \emph{Compact Disks}, a onda possui frequência fundamental $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz $.
-Um lá\footnote{Um lá 4, logo acima do dó central, no segundo espaço do pentagrama na clave de sol comum.}, independente da forma de onda.
+Um lá\footnote{Um lá 4, logo acima do dó central, no segundo espaço do pentagrama na clave de sol comum.} seja qual for a forma de onda.
 
@@
     \caption{Espectros das ondas sonoras musicais artificiais básicas.
-As senoide tem o especro puntual, a triangular apresenta somente os harmônicos ímpares, caindo a 6dB por oitava; a onda quadrada tem somente os harmônicos ímpares, caindo a 12dB por oitava; a onda dente de serra apresenta todos os harmônicos, caindo a 6dB por oitava.}
+A senoide tem o especro puntual, a triangular apresenta somente os harmônicos ímpares, caindo a 6dB por oitava; a onda quadrada tem somente os harmônicos ímpares, caindo a 12dB por oitava; a onda dente de serra apresenta todos os harmônicos, caindo a 6dB por oitava.}
         \label{fig:espectroDeOndas}
@@
 Todas as sequências fixas $T_i$ de apenas $3$ amostras também apresentam
-somente $1$ frequência, pois sua primeira harmônica usaria $1.5$ amostras e ultrapassa o limite inferior de 2 amostras mínimas, i.e.
+somente $1$ frequência, pois sua primeira harmônica usaria $1,5$ amostras e ultrapassa o limite inferior de 2 amostras mínimas, i.e.
 a frequência da harmônica ultrapassaria a de Nyquist pois:  $\; \frac{2.
 f_a}{3} > \frac{f_a}{2} $.
-Os coeficiêntes $\{c_k\}_0^{\Lambda-1=2}$ apresentam-se em 
+Os coeficientes $\{c_k\}_0^{\Lambda-1=2}$ apresentam-se em 
 3 componentes frequenciais. Uma delas é relativa à frequência zero ($c_0$), as outras duas ($c_1$ e $c_2$) contribuem de forma igual na reconstrução da senoide com $f=f_a/3$.
@@
 
-Convertendo para amplitude, obtem-se $DII_a=\frac{d}{d'}$. A $DII_a$ pode
+Convertendo para amplitude, obtém-se $DII_a=\frac{d}{d'}$. A $DII_a$ pode
 ser utilizada como constante multiplicativa do canal direito de um sinal sonoro estéreo: $\{t_i'\}_0^{\Lambda -1}=\{DII_a .
@@
 do cone.
-Nestes pontos, a inferência do ângulo azimutal depende especialmente da filtragem atenuante nos agudos, pois a cabeça interfere um tanto mais nas ondas mecânicas agudas do que nas graves.\cite{Heeger,hrtf}  Também pertinente à audição de fonte lateral, quando o som é graves o suficiente, há uma difração e a onda chega ao ouvido $\approx 0,7ms$ depois.\cite{floEsp}
+Nestes pontos, a inferência do ângulo azimutal depende especialmente da filtragem atenuante nos agudos, pois a cabeça interfere um tanto mais nas ondas mecânicas agudas do que nas graves.\cite{Heeger,hrtf}  Também pertinente à audição de fonte lateral, quando o som é grave o suficiente, há uma difração e a onda chega ao ouvido $\approx 0,7ms$ depois.\cite{floEsp}
 
@@
 
-Já a localização completa, incluindo distância e elevação da fonte sonora, é dada pela função de transferência de cabeça (HRTF - do inglês \emph{Head Related Transfer Function}).\cite{hrtf} Exitem bases abertas e conhecidas de HRTF como a CIPIC e pode-se aplicar estas funções de transferência em um som por convolução (veja equação~\ref{eq:conv}).\cite{CIPIC} O corpo do indivíduo altera bastante as filtragens realizadas e existem técnicas para gerar HRTFs que sejam - como proposta - utilizáveis de forma universal.\cite{lazaSPA}
+Já a localização completa, incluindo distância e elevação da fonte sonora, é dada pela função de transferência de cabeça (HRTF - do inglês \emph{Head Related Transfer Function}).\cite{hrtf} Existem bases abertas e conhecidas de HRTF como a CIPIC e pode-se aplicar estas funções de transferência em um som por convolução (veja equação~\ref{eq:conv}).\cite{CIPIC} O corpo do indivíduo altera bastante as filtragens realizadas e existem técnicas para gerar HRTFs que sejam - como proposta - utilizáveis de forma universal.\cite{lazaSPA}
 
@@
 \begin{itemize}
-    \item Primeiro periodo: as 'primeiras reflexões' são mais intensas e esparsas.
+    \item Primeiro período: as 'primeiras reflexões' são mais intensas e esparsas.
     \item Segundo período: a 'reverberação tardia' é praticamente uma sucessão densa de atrasos indistintos com um decaimento exponencial e ocorrências estatísticas.
@@
 Aponta também o período das primeiras reflexões de $0,1$ segundos.
-Estas quantidades sugerem que, nas condições contempladas, há frentes de onda perceptiveis que se propagam até $652.08$ metros ($83.79k$ amostras em $f_a=44.1kHz$) antes de atingirem o ouvido.
+Estas quantidades sugerem que, nas condições contempladas, há frentes de onda perceptíveis que se propagam até $652.08$ metros ($83.79k$ amostras em $f_a=44.1kHz$) antes de atingirem o ouvido.
 Além disso, as reflexões do som formam, após a propagação por $34.32$ metros ($4.41k$ amostras em $f_a=44.1kHz$ ), um emaranhado cujas incidências são pouco distintas na audição.
@@
 A primeira incidência é o som direto, descrito por DTI e DII das equações~\ref{eq:dti} e~\ref{eq:dii}.
-Admitindo que cada uma das primeiras reflexões, antes de chegar ao ouvido.
-se propagará, ao menos, $3-30m$ dependentes das dimensões da sala, a separação entre as primeiras reflexões é de, ao menos, $8-90$ milissegundos ($\approx 350-4000$ amostras em $f_a=44.1kHz$).
+Admitindo que cada uma das primeiras reflexões, antes de chegar ao ouvido, se propagará, ao menos, $3-30m$ dependentes das dimensões da sala, a separação entre as primeiras reflexões é de, ao menos, $8-90$ milissegundos ($\approx 350-4000$ amostras em $f_a=44.1kHz$).
 Verifica-se experimentalmente que o número de reflexões aumenta em proporção quadrática  $ \approx k.n^2$.
@@
 
-Está descrita a nota musical digital básica e a seção a seguinte desenvolve a evolução temporal de seus conteúdos, como nos glissandos e nas envoltórias de volume.
+Está descrita a nota musical digital básica e a seção seguinte desenvolve a evolução temporal de seus conteúdos, como nos glissandos e nas envoltórias de volume.
 A filtragem de componentes espectrais e a geração dos ruídos completam a constituição da nota musical como unidade isolada e se desdobra na seção~\ref{notasMusica}, dedicada à estruturação destas notas em música através de métricas e trajetórias.
@@
 aqui é: para que um som isolado desperte interesse
-por si só, faça como que tenha variações internas.\cite{Roederer}}. 
+por si só, faça com que tenha variações internas.\cite{Roederer}}. 
 Explorar todas as formas pelas quais estas variações ocorrem está fora
@@
 Existem
-diversos métodos e ferrametas para a elaboração de filtros IIR
+diversos métodos e ferramentas para a elaboração de filtros IIR
 e segue abaixo uma seleção com fins didáticos e para consulta futura por
@@
 As velocidades são decompostas com relação à direção de cada ouvido.
-Assim, dada a velocidade $v_s$ da fonte ($s$ de \emph{source}), positiva se a fonte se no sentido de se afastar do receptor e a velocidade $v_r$ do receptor, positiva no sentido de se aproximar da fonte, a frequência é dada pela conhecida fórmula do efeito Doppler:
+Assim, dada a velocidade $v_s$ da fonte ($s$ de \emph{source}), positiva se a fonte se move no sentido de se afastar do receptor e a velocidade $v_r$ do receptor, positiva no sentido de se aproximar da fonte, a frequência é dada pela conhecida fórmula do efeito Doppler:
 
@@
 Há um acréscimo para maior fidelidade ao fenômeno físico: um aumento da potência recebida.
-Pode-se assumir que este aumento em potência é proporcional à velocidade relativa, que, a cada segundo, acrescenta aquele trecho percorrido, pela forma de onda, de potência: $\Delta P=P_0\left(\frac{v_r-v_s}{343,2}\right)$, com $P_0$ a potência do sinal.
+Pode-se pressupor que este aumento em potência é proporcional à velocidade relativa, que, a cada segundo, acrescenta aquele trecho percorrido, pela forma de onda, de potência: $\Delta P=P_0\left(\frac{v_r-v_s}{343,2}\right)$, com $P_0$ a potência do sinal.
 
@@
 
-Inspirado em nos instrumentos tradicionais, pode-se aplicar uma a filtragem dependente do tempo~\cite{Roederer}.
+Inspirado nos instrumentos tradicionais, pode-se aplicar uma a filtragem dependente do tempo~\cite{Roederer}.
 Cascatas
@@
 
-Estes recursos todos utilizados em conjunto podem incidir na realização de um efeito chamado \emph{chorus}. A
+Estes recursos utilizados em conjunto podem incidir na realização de um efeito chamado \emph{chorus}. A
 exemplo do que ocorre com um coro de cantores, neste efeito o som é realizado com diversas pequenas modificações,
@@
 Com a mesma nomenclatura da subseção~\ref{subsec:spac} sobre espacialização, a reverberação tardia pode ser modelada como a convolução com um trecho de ruído colorido rosa, marrom ou preto, em decaimento exponencial de amplitude com relação também ao tempo.
-Assim, a atenuação nos agudos e irregularidade suave na resposta em frequência são contempladas de forma bastante satisfatória.\cite{JOSPhy,guillaume} Linhas de atrasos podem ser adicionadas como prefixo à ao ruído com o decaimento, assim comtemplando as duas partes temporais da reverberação: as primeiras reflexões e a reverberação tardia.
+Assim, a atenuação nos agudos e irregularidade suave na resposta em frequência são contempladas de forma bastante satisfatória.\cite{JOSPhy,guillaume} Linhas de atrasos podem ser adicionadas como prefixo ao ruído com o decaimento, assim comtemplando as duas partes temporais da reverberação: as primeiras reflexões e a reverberação tardia.
 Pode-se, para melhora de qualidade, calcular geometricamente a localização da última superfície em que cada frente de onda refletiu antes de chegar ao ouvido nos primeiros $100-200$ milissegundos e aplicar um filtro passa-baixas como em~\ref{subsec:filtros}.
@@
 A reverberação é conhecida por causar bastante interesse no ouvinte e tornar as sonoridades mais agradáveis.
-Além disso, a modificação do espaço em que são reverberadas a música, ou as sonoridades individualmente, constitui um massete (quase um clichê) para causar supresa e interesse no ouvinte.
+Além disso, a modificação do espaço em que são reverberadas a música, ou as sonoridades individualmente, constitui um macete (quase um clichê) para causar surpresa e interesse no ouvinte.
 
@@
 Abaixo a variação de amplitude é exponencial.}
-\end{figure}
         \label{fig:adsr}
+\end{figure}
 
@@
 segundas e sétimas - podem ser maiores ou menores.
-Excessão para o trítono.
+Exceção para o trítono.
         \item A quarta justa é tida como consonante perfeita ou dissonante de acordo com o contexto e arcabouço teórico.
@@
 é $f_3=\varepsilon^{e_3^t} . f_0 = 2^{\frac{6}{12}} . 200 \approxeq 282.843 Hz$. Estas
-'escalas' ou padrões, geram estruturas estáveis pelas simetrias internas, e podem ser
+'escalas' ou padrões geram estruturas estáveis pelas simetrias internas, e podem ser
 repetidas de forma eficiente e sustentada.
@@
 Vale apontar que boa parte do que escreveram os compositores
-emblemáticos para o dodecafonismo, como Alban Berg e o próprio Schoenberg, consiste
+emblemáticos para o dodecafonismo, como Alban Berg e o próprio Schoenberg, consiste de
 aplicações parciais e livres destas técnicas. Várias peças fazem até mesmo pontes
@@
 grau de uma escala, como das equações~\ref{eq:escalas}, formam os pilares do campo harmônico.
-A observação de progressões incidentes de acordes e regras para encadeamentos é o objeto de estudo da harmonia musical.
+A observação de progressões incidentes de acordes e regras para encadeamentos é o objeto de estudo da harmonia músical.
 Mesmo uma melodia monofônica gera campos harmônicos e pode-se observar acordes sugeridos por cada passagem.
@@
 permitindo a dominante (que deve ser maior e tender à tônica).
-Assim, a dominante é sempre maior, tanto em escalas maior ou menor e, por este motivo, mesmo em tom menor, a dominante relativa permanece a uma terça abaixo e a antirelativa a uma terça acima.
+Assim, a dominante é sempre maior, tanto em escalas maior ou menor e, por este motivo, mesmo em tom menor, a dominante relativa permanece a uma terça abaixo e a antirrelativa a uma terça acima.
 
@@
 
-A importância da dominante torna ela pivô natural das modulações, o que desemboca no círculo das quintas.\cite{Harmonia,Salzer,Koellheuteur,Harmony} A montagem musical \emph{Acorde cedo} explora estas relações entre acordes.
+A importância da dominante a torna pivô natural das modulações, o que desemboca no círculo das quintas.\cite{Harmonia,Salzer,Koellheuteur,Harmony} A montagem musical \emph{Acorde cedo} explora estas relações entre acordes.
 Seu código está disponível do Apêndice~\ref{ap:acorde} e online como parte da \massa.\cite{MASSA}
@@
 a primeira subdivisão do pulso, o nível $j=1$ a primeira algomeração dos pulsos e assim por diante. 
-
 Desta forma, $P_i^j$ é a $i$-ésima unidade de 
@@
 A sequência de Lucas permite uma generalização da sequência de Fibonacci que pode ser compreendida facilmente.
-Dados dois números quaisquer $x_0$ e $x_1$, obtem-se a sequência de Lucas: $x_n=x_{n-1}+x_{n-2}$.
+Dados dois números quaisquer $x_0$ e $x_1$, obtém-se a sequência de Lucas: $x_n=x_{n-1}+x_{n-2}$.
 Quanto maior for $n$, mais a razão $\frac{x_{n}}{x_{n_1}}$ se aproxima da razão áurea: $1.61803398875...$.
@@
 
-No Apêndice~\ref{cap:FIGGUScode} apontada a implementação computacional disponibilizada em.\cite{MASSA,figgusOriginal,figgusEspacializacao}
+No Apêndice~\ref{cap:FIGGUScode} está a implementação computacional disponibilizada em.\cite{MASSA,figgusOriginal,figgusEspacializacao}
 
@@
 Por fim, esta presente seção dedica-se a organizar estas notas em música.
-A gama de recursos e a consequênte infinidade de possibilidades de resultados
+A gama de recursos e consequente infinidade de possibilidades de resultados
 é situação típica e cara às artes.\cite{Harmonia,Webern}

### 2013-04-02 trabalhosRelacionados.tex
adicionadas correcoes do prof. chu sobre os trabalhos relacionados

@@
 
-\section{Trabalhos relacionados}
-
-\subsection{Livros}\label{subsec:livros}
+\section{Livros}\label{subsec:livros}
 
@@
         \begin{itemize}
-            \item {\bf Descrição:} com exemplos em código computacional e sonoros, este excelênte livro de Pedro Kroeger aborda conceitos de notas, afinações, especificação da nota Midi e conversão entre nomenclaturas latina (dó-ré-mi) e anglo-saxã (C-D-E).
+            \item {\bf Descrição:} com exemplos em código computacional e sonoros, este excelente livro de Pedro Kroeger aborda conceitos de notas, afinações, especificação da nota Midi e conversão entre nomenclaturas latina (dó-ré-mi) e anglo-saxã (C-D-E).
 Lida também com operações musicais fundamentais como transposição, inversão e afins, combinações randômicas, por Fibonacci.
@@
 O livro começa com medições do som e controle de parâmetros, cai em síntese, modulações, métodos espectrais, atrasos e reverberações e termina com filtros.
-            \item {\bf Aspecto complementar:} o livro apresenta diversos procedimentos valioso para síntese, tratamento e análise.
-O texto busca ser de computação musical em geral e todo o texto é acompanhado de exemplos em Puredata, que é uma excelênte, linguagem de programação por patches voltada para audiovisual.
+            \item {\bf Aspecto complementar:} o livro apresenta diversos procedimentos valiosos para síntese, tratamento e análise.
+O texto busca ser de computação musical em geral e todo o texto é acompanhado de exemplos em Puredata, que é uma excelente linguagem de programação por patches voltada para audiovisual.
 Puredata é a linguagem de programação mais difundida na música acadêmica e tecnológica em geral.
-            \item {\bf Aspecto diferencial:} salvo raras excessões, o livro não apresenta uma descrição analítica das amostras sonoras com relação aos procedimentos, assim, não relaciona de forma precisa as qualidades físicas do som.
+            \item {\bf Aspecto diferencial:} salvo raras exceções, o livro não apresenta uma descrição analítica das amostras sonoras com relação aos procedimentos, assim, não relaciona de forma precisa as qualidades físicas do som.
 Tampouco se aprofunda em aspectos formais da teoria musical tradicional.
@@
 
-\subsection{Bibliotecas, linguagens e conjuntos de implementações computacionais voltados para música}\label{subsec:bibs}
+\section{Bibliotecas, linguagens e conjuntos de implementações computacionais voltados para música}\label{subsec:bibs}
 {\bf Implementações fechadas.} As seguintes implementações fogem ao escopo do trabalho dada a inexistência de descrição amostral ou código que assegure uma implementação precisa e limpa.
@@
 Uma linha especial de \emph{plugins} de \emph{limiters}.
-Uma extensa linha de excelêntes \emph{plugins} de EQs.
+Uma extensa linha de excelentes \emph{plugins} de EQs.
 Uma linha de plugins de \emph{reverbs} e \emph{delays}.
@@
 Os manuais do Max/MSP são ótimas e até históricas introduções qualitativas e práticas.
-O livro do Puckette, citado em na lista de livros acima, é um dos materiais excelêntes que utilizam o PD.
+O livro do Puckette, citado em na lista de livros acima, é um dos materiais excelentes que utilizam o PD.
 Complementar à dissertação há esta aproximação de um programa/sistema, com interface, e mais engessado do que uma linguagem de programação.
@@
     \item \emph{Faust, Chuck, Supercollider, Nyquist, Impromptu}: linguagens de domínio específico como p PD e o Max/MSP, voltadas para a síntese de música em tempo real, mas escritas normalmente em linhas, não em patches.
-Complementar ao trabalho é que são implementações computacionais excelêntes para uso em tempo real.
+Complementar ao trabalho é que são implementações computacionais excelentes para uso em tempo real.
 Diferencial é que há perda da desenvoltura para manipulações de áudio com controle amostral de precisão arbitrária.
 A descrição analítica também é inexistente.
-    \item \emph{CSound}: é uma a linguagem de programação voltada para música mais antiga que está em uso.
-As constantes produções acadêmicas em torno do CSound podem servir de descrições mais rigorosas do que as encontradas nas soluções anterires.
+    \item \emph{CSound}: linguagem de programação voltada para música mais antiga que está em uso.
+As constantes produções acadêmicas em torno do CSound podem servir de descrições mais rigorosas do que as encontradas nas soluções anteriores.
 De qualquer forma, não há uma documentação clara e concisa sobre os procedimentos básicos como o capítulo 2 da dissertação.
@@
 
-\subsection{Aprofundamento sobre esta dissertação com base nos trabalhos visitados nas subseções~\ref{subsec:livros} e~\ref{subsec:bibs}}
+\section{Aprofundamento sobre esta dissertação com base nos trabalhos visitados nas subseções~\ref{subsec:livros} e~\ref{subsec:bibs}}
 

### 2013-04-16 devEresultados.tex
feita implementação de efeito Doppler e descrição na secao 2.2

@@
 
-\subsection{Volume}
+\subsection{Volume}\label{subsec:volume}
 A sensação de volume sonoro depende da reverberação e distribuição dos harmônicos, dentre outras características trabalhadas na seção~\ref{varInternas}.
@@
 
-Já a onda quadrada, pode ser usada na síntese
+Já a onda quadrada pode ser usada na síntese
 subtrativa que vise a imitar um clarinete. Este instrumento também só apresenta os
@@
 A figura~\ref{fig:spac} mostra também esta sombra acústica do crânio, importante para a percepção do ângulo azimutal da fonte no cone de confusão.
-O cone em si não foi disposto na figura pois não é exatamente um cone e suas dimensões precisas não foram encontradas na literatura visitada, mas pode ser entendido como um cone com o ápice no meio da cabeca e saindo por cada uma das orelhas.\cite{hrtf}
+O cone em si não foi disposto na figura pois não é exatamente um cone e suas dimensões precisas não foram encontradas na literatura visitada e não são facilmente concebíveis, dadas as filtragens e a difração dependente do espectro do som em si.
+De toda forma, o cone de confusão pode ser entendido como um cone com o ápice no meio da cabeca e saindo por cada uma das orelhas.\cite{hrtf}
 
@@
 Aponta também o período das primeiras reflexões de $0,1$ segundos.
-Estas quantidades sugerem que, nas condições contempladas, há frentes de onda perceptíveis que se propagam até $652.08$ metros ($83.79k$ amostras em $f_a=44.1kHz$) antes de atingirem o ouvido.
-Além disso, as reflexões do som formam, após a propagação por $34.32$ metros ($4.41k$ amostras em $f_a=44.1kHz$ ), um emaranhado cujas incidências são pouco distintas na audição.
+Estas quantidades sugerem que, nas condições contempladas, há frentes de onda perceptíveis que se propagam até $652,08$ metros ($83,79k$ amostras em $f_a=44,1kHz$) antes de atingirem o ouvido.
+Além disso, as reflexões do som formam, após a propagação por $34,32$ metros ($4,41k$ amostras em $f_a=44,1kHz$ ), um emaranhado cujas incidências são pouco distintas na audição.
 Estas primeiras reflexões são particularmente importantes para a sensação de espaço.
@@
 
-Retomada a exposição da subseção~\ref{subsec:spac}, quando uma fonte ou um receptor está em movimento, suas características idealmente são calculadas amostra a amostra do sinal digital.
+Retomada a exposição da subseção~\ref{subsec:spac}, quando uma fonte ou um receptor está em movimento, suas características idealmente são atualizadas a cada amostra do sinal digital.
 As velocidades são decompostas com relação à direção de cada ouvido.
@@
 \begin{equation}
-    f=\left(\frac{c+v_r}{c+v_s}\right)f_0
+    f=\left(\frac{v_{som}+v_r}{v_{som}+v_s}\right)f_0
 \end{equation}
@@
 Há um acréscimo para maior fidelidade ao fenômeno físico: um aumento da potência recebida.
-Pode-se pressupor que este aumento em potência é proporcional à velocidade relativa, que, a cada segundo, acrescenta aquele trecho percorrido, pela forma de onda, de potência: $\Delta P=P_0\left(\frac{v_r-v_s}{343,2}\right)$, com $P_0$ a potência do sinal.
-
-\subsubsection{Ruidos e filtragens}
+Pode-se entender que este aumento em potência é proporcional à velocidade relativa, que, a cada segundo, acrescenta aquele trecho percorrido, pela forma de onda, de potência: $\Delta P=P_0\left(\frac{v_r-v_s}{343,2}\right)$, com $P_0$ a potência do sinal.
+
+Assim, pode-se obter a amplitude e frequência de uma fonte em movimento.
+Esteja esta fonte à frente do receptor
+a uma distância horizontal de $y_0$ metros e altura $z_0$ metros. A distância é dada por
+$D_i=\left\{ d_i=\sqrt{ y_{i}^{2}+z_{0}^{2} } \right\}_0^{\Lambda-1}$
+onde $y_i=y_0+v_s-v_r$ se consideradas $v_s$ e $v_r$ horizontais.
+A amplitude muda com a distância e com o fator de potência citado acima (veja a subseção~\ref{subsec:volume} para a conversão de potência para amplitude):
+
+\begin{equation}
+    A_i=\left\{ \frac{z_0}{d_i}A_{\Delta P}\right\}_0^{\Lambda-1} = \left\{ \frac{z_0}{\sqrt{y_i^2+z_0^2}} \sqrt{\frac{v_r-v_s}{343,2}+1}  \,\right\}_0^{\Lambda-1}
+\end{equation}
+
+Observe que o fator alterador de amplitude devido à distância é par enquanto a o fator devido à variação de potência é antissimétrico com relação ao cruzamento da fonte e do receptor.
+Já a frequência tem uma progressão simétrica com relação à altura, ou seja, os mesmos semitons (ou frações) acrescentados na aproximação são decrescidos no afastamento.
+Além disso, a transição é abrupta caso a fonte e o receptor se cruzem com exatidão, caso contrário, há uma progressão monotônica.
+No caso considerado, em que há uma altura fixa $z_0$, é necessário observar a componente de velocidade na direção entre o observador e a fonte:
+
+\begin{equation}
+    F_i=\{f_i\}_0^{\Lambda-1}=\left\{\frac{v_{som} + v_r\frac{y_i}{\sqrt{z_0^2+y_i^2}}}{v_{som}+v_s\frac{y_i}{\sqrt{z_0^2+y_i^2}}}f_0\right\}_0^{\Lambda-1}
+\end{equation}
+
+No Apêndice~\ref{sec:cod2} está uma implementação em Python do efeito Doppler como descrito acima, com as alterações de frequência e amplitude, e com o cruzamento da fonte com o receptor.
+
+\subsubsection{Filtragens e ruídos (subseções~\ref{subsec:ruidos} e~\ref{subsec:filtros})}
 Com os filtros as possibilidades explodem ainda mais vertiginosamente. Pode-se convoluir um sinal para reverberá-lo, para

### 2013-04-16 devEresultados.tex
descrito analisticamente e implementado reverberacao artificial segundo apontamentos de JO Smith III

@@
 sem reflexão alguma e dada pela DTI e DII), em \emph{fade-in}, atingindo o máximo no começo da 'reveberação tardia', quando as incidências geométricas perdem sua importância para as estatísticas do ruído em decaimento.
+
+Como um exemplo de implementação, seja $\Delta_1$ a duração do primeiro período e $\Delta_R$ a duração total da reverberação ($\Lambda_1=\Delta_1 f_a$, $\Lambda_R=\Delta_R f_a$). Pode-se
+associar uma probabilidade $p_i$ de uma reincidência do som na $i$-ésima amostra com amplitude em decaimento exponencial. Seguindo
+as indicações na subseção~\ref{subsec:spac}, a reverberação $R_i^1$ do primeiro período pode ser descrita desta forma:
+
+\begin{equation}
+    R_i^1=\left\{r_i^1\right\}_0^{\Lambda_1-1}\;:\;r_i^1=\left\{
+        \begin{array}{l l}
+            10^{\frac{V_{dB}}{20}\frac{i}{\Lambda_R-1}}\;  & \text{com probabilidade}\quad p_i=\left(\frac{i}{\Lambda_1}\right)^2 \\
+                                     0 \; & \text{com probabilidade}\quad 1-p_i \\
+        \end{array} \right.
+\end{equation}
+
+Onde $V_{dB}$ é o decaimento total considerado em decibels, tipicamente $-80dB$ ou $-120dB$.
+Já a reverberação $R_i^2$, do segundo período, pode ser emulada por um ruído marrom $R_i^m$ (ou rosa $R_i^r$) em decaimento exponencial:
+
+\begin{equation}
+    R_i^2=\left\{r_i^2\right\}_{\Lambda_1}^{\Lambda_R-1}=\left\{10^{\frac{V_{dB}}{20}\frac{i}{\Lambda_R-1}}\,.\,r_i^m\right\}_{\Lambda_1}^{\Lambda_R-1}
+\end{equation}
+
+Como:
+\begin{equation}
+    R_i=\left\{r_i\right\}_0^{\Lambda_R-1}\;:\;r_i=\left\{
+        \begin{array}{l l}
+            r_i^1\;  & \text{se }\quad i<\Lambda_1-1 \\
+                                     r_i^2 \; & \text{se}\quad \Lambda_1 \leq i < \Lambda_R-1 \\
+        \end{array} \right.
+\end{equation}
+
+a aplicação da reverberação representada por $R_i$ é a simples convolução de $R_i$ (chamada 'resposta ao impulso' da reverberação) com a sequência sonora $T_i$, tal como descrito na subseção~\ref{subsec:filtros}.
 

### 2013-04-16 trabalhosRelacionados.tex
feito apontamento para a ausencia de artigos sobre o tema

@@
 Pode-se propor expansões para usos conjuntos com alguns dos trabalhos, visitados ou não.
-Vale assinalar: com base na experiência prévia dos autores e nos trabalhos visitados em~\ref{subsec:livros} e~\ref{subsec:bibs}, a descrição analítica dos elementos musicais básicos em termos das amostras do áudio digital parece ser inédita.
-     Como há também a implementação em código livre das relações descritas nesta dissertação, abre-se um leque para experimentações científicas rigorosas assim como para usos artísticos com alta fidelidade e com precisão arbitrária.
+Vale assinalar: com base na experiência prévia dos autores, nos escassos artigos encontrados e nos trabalhos visitados em~\ref{subsec:livros} e~\ref{subsec:bibs}, a descrição analítica dos elementos musicais básicos em termos das amostras do áudio digital parece ser inédita.
+Como há também a implementação em código livre das relações descritas nesta dissertação na \emph{toolbox} \massa, abre-se um leque para experimentações científicas rigorosas assim como para usos artísticos com alta fidelidade e precisão arbitrária.\cite{MASSA}

### 2013-04-16 analiseMusical.tex
feito pequeno capitulo de analise musical

\chapter{Análise sonora e musical}
Este capítulo foi escrito para complementar a didática da dissertação pois há forte interesse em análise musical assistida por computador, tanto para a indústria, como para pesquisas acadêmicas e musicológicas. Assim, o texto a seguir é uma introdução rápida e deve ser apreciado como um conjunto despretencioso de apontamentos informais a conceitos básicos de análise de música em evidência com os recursos computacionais.

\section{Espectro em frequência}\label{subsec:especFreq}
Muitas vezes, se está preocupado com as componentes em frequência de um dado som ou estilo. Para tal, a transformada de Fourier é a solução de uso mais comum. A localização de energia no tempo e em frequência, por exemplo para observar as componentes do espectro no decorrer de uma sonoridade ou uma música, é geralmente feita com a transformada de Fourier janelada ou com a transformada \emph{wavelet} (transformada de ondeleta). Uma figura bidimensional da evolução do espectro no decorrer do tempo é o espectrograma (veja figura~\ref{fig:vibrato}). A transformada wavelet é bastante simples e computacionalmente barata, consistindo em um banco de filtros FIR. Por outro lado, o espectro apresentado não é senoidal, tão caro à cognição e à música. Além disso, a transformada wavelet é mais rígida com relação à resolução temporal e frequencial do que a transformada de Fourier janelada.

Com a observação espectral, verifica-se a presença de formantes, detecta-se frequências fundamentais, calcula-se centróides, observa-se decaimento e aparecimento de harmônicas e ruído, para citar somente alguns exemplos. O vínculo direto do timbre com o espectro de um som, mesmo não se sabendo os mecanismos cognitivos envolvidos, é compreensão básica.

\section{Eventos no tempo}\label{subsec:eventosTempo}
Com o quadrado das amostras temporais é obtida uma medida da energia no trecho considerado. Variações abruptas de energia indicam ataques, o que pode ser informativo do ponto de vista rítmico e até como indicação de estilos ou técnicas. Os decaimentos são importantes musicalmente, mas suas detecções são secundárias para a análise. Dados os ataques ao longo da duração, é capital e capcioso (difícil) adequá-los a alguma métrica. Para detecção de ataques de instrumentos diferentes no mesmo áudio, pode-se, por exemplo, realizar a separação do áudio por faixas do espectro ou procurar um espectro já conhecido.

\section{Análise de partituras}
As partituras são registros musicais que usam símbolos consagrados. A extração de características musicais a partir das partituras, é de uso musicológico e classificatório. Estão em uso diferentes formatos para a utilização computacional destas informações (como o Kern) e, como exemplo de uso, pode-se obter a sequência de intervalos de altura de linhas melódicas que percorrem toda a música ou as durações relativas de suas notas.

### 2013-04-16 conclusoes.tex
passado corretor ortografico

@@
 e Áudio em Sequências e Séries Amostrais). 
-A exposição didática destes desenvovimentos no
+A exposição didática destes desenvolvimentos no
 capítulo anterior destina-se a facilitar a utilização
@@
 A disposição online destes conteúdos na forma de hipertexto junto aos códigos e exemplos sonoros, todos em licenças livres, facilita colaborações e geração de subprodutos em co-autoria, e com isso a expansão da \massa\ com novas implementações e desenvolvimentos das montagens musicais.
-Explorações sistemáticas de parametrizações (dos tremolos, da ADSR, etc) em alta fidelidade tem utilidade artística e é possibilidata por este trabalho com controle amostral.
+Explorações sistemáticas de parametrizações (dos tremolos, da ADSR, etc) em alta fidelidade tem utilidade artística e é possibilitada por este trabalho com controle amostral.
 Tal descrição analítica precisa, junto às implementações computacionais, não foi atingida anteriormente, como mostra o Apêndice~\ref{cap:trabalhosRelacionados} com uma visita aos trabalhos relacionados.
@@
 já apresentou contribuições relevantes em diferentes áreas
-como Democracia Direta Digital, ferramentas de georeferenciamento e
+como Democracia Direta Digital, ferramentas de georreferenciamento e
 atividades artísticas e educacionais, como cursos, workshops e apresentações artísticas.

### 2013-04-16 devEresultados.tex
arrumadas numeracoes ateh 2.2
passado corretor ortografico

@@
     \caption{Espectros das ondas sonoras musicais artificiais básicas.
-A senoide tem o especro puntual, a triangular apresenta somente os harmônicos ímpares, caindo a 6dB por oitava; a onda quadrada tem somente os harmônicos ímpares, caindo a 12dB por oitava; a onda dente de serra apresenta todos os harmônicos, caindo a 6dB por oitava.}
+A senoide tem o espectro puntual, a triangular apresenta somente os harmônicos ímpares, caindo a 6dB por oitava; a onda quadrada tem somente os harmônicos ímpares, caindo a 12dB por oitava; a onda dente de serra apresenta todos os harmônicos, caindo a 6dB por oitava.}
         \label{fig:espectroDeOndas}
@@
 Onde $c_k = a_k + j .
-b_k$ dita a amplitude e fase de cada frequência: $w_k=\frac{2\pi}{\Lambda}k$ em radianos ou $f_k=w_k\frac{f_a}{2\pi}=\frac{f_a}{\Lambda}k$ em Herz,com atenção para os respectivos limites em $\pi$ e em $\frac{f_a}{2}$ dados pelo Teorema de Nyquist.
+b_k$ dita a amplitude e fase de cada frequência: $w_k=\frac{2\pi}{\Lambda}k$ em radianos ou $f_k=w_k\frac{f_a}{2\pi}=\frac{f_a}{\Lambda}k$ em Hertz,com atenção para os respectivos limites em $\pi$ e em $\frac{f_a}{2}$ dados pelo Teorema de Nyquist.
 
@@
 A limitação será superada no início da próxima seção.}.
-Uma sequência $T_i$ de amostras sonoras separadas por $\delta_a=1/f_a$ descreve uma nota musical de frequência $f$ Herz e duração $\Delta$ segundos se, e somente se, possuir a periodicidade $\lambda_f=f_a/f$
+Uma sequência $T_i$ de amostras sonoras separadas por $\delta_a=1/f_a$ descreve uma nota musical de frequência $f$ Hertz e duração $\Delta$ segundos se, e somente se, possuir a periodicidade $\lambda_f=f_a/f$
  e tamanho $\Lambda=\lfloor f_a . \Delta \rfloor $:
@@
 três informações: o atraso de chegada do som entre um ouvido e o outro, a diferença de intensidade do som direto em cada ouvido e a 
-filtragem realizada pelo corpo, incluindo toráx, cabeça e orelhas.\cite{Roederer, hrtf, Heeger}
+filtragem realizada pelo corpo, incluindo tórax, cabeça e orelhas.\cite{Roederer, hrtf, Heeger}
 
@@
 O cone em si não foi disposto na figura pois não é exatamente um cone e suas dimensões precisas não foram encontradas na literatura visitada e não são facilmente concebíveis, dadas as filtragens e a difração dependente do espectro do som em si.
-De toda forma, o cone de confusão pode ser entendido como um cone com o ápice no meio da cabeca e saindo por cada uma das orelhas.\cite{hrtf}
+De toda forma, o cone de confusão pode ser entendido como um cone com o ápice no meio da cabeça e saindo por cada uma das orelhas.\cite{hrtf}
 
@@
 respectivamente, $\frac{f_a}{100/2}=882Hz$, $\frac{f_a}{100/4}=1764Hz$ e $\frac{f_a}{100/5}=2205Hz$.
-A duração do trecho é bastante curto $\frac{f_a=44.1kHz}{100} \approx 2 \text{ milisegundos}$.
+A duração do trecho é bastante curto $\frac{f_a=44.1kHz}{100} \approx 2 \text{ milissegundos}$.
 Basta completar com zeros para somar sequências de tamanhos diferentes.
@@
 
-Está descrita a nota musical digital básica e a seção seguinte desenvolve a evolução temporal de seus conteúdos, como nos glissandos e nas envoltórias de volume.
+Está descrita a nota musical digital básica e a seção seguinte desenvolve a evolução temporal de seus conteúdos, como nos \emph{glissandi} e nas envoltórias de volume.
 A filtragem de componentes espectrais e a geração dos ruídos completam a constituição da nota musical como unidade isolada e se desdobra na seção~\ref{notasMusica}, dedicada à estruturação destas notas em música através de métricas e trajetórias.
@@
 Todas as relações descritas nesta seção estão implementadas em Python no Apêndice~\ref{sec:cod2}.
-As montagens musicais \emph{Tansita para metro}, \emph{Vibra e treme}, \emph{Tremolos, vibratos e a frequência}, \emph{Trenzinho de caipiras impulsivos}, \emph{Ruidosa faixa}, \emph{Bela rugosi}, \emph{Chorus infantil}, \emph{ADa e SaRa} estão nos Apêndices~\ref{ap:transita}, \ref{ap:vibra}, \ref{ap:tremolos}, \ref{ap:trenzinho}, \ref{ap:ruidosa}, \ref{ap:bela}, \ref{ap:chorus} e \ref{ap:ada}.
+As montagens musicais \emph{Transita para metro}, \emph{Vibra e treme}, \emph{Tremolos, vibratos e a frequência}, \emph{Trenzinho de caipiras impulsivos}, \emph{Ruidosa faixa}, \emph{Bela rugosi}, \emph{Chorus infantil}, \emph{ADa e SaRa} estão nos Apêndices~\ref{ap:transita}, \ref{ap:vibra}, \ref{ap:tremolos}, \ref{ap:trenzinho}, \ref{ap:ruidosa}, \ref{ap:bela}, \ref{ap:chorus} e \ref{ap:ada}.
 Estes códigos são parte da caixa de ferramentas \massa, disponível online.\cite{MASSA}
@@
 pela sua sigla FIR (do inglês Finite Impulse Response)
-e são caracterizados por possuirem uma representação amostral
+e são caracterizados por possuírem uma representação amostral
 finita no tempo. Esta representação amostral é chamada
@@
 
-Com este procedimento pode-se aplicar reverberadores, equalizadores, delays
+Com este procedimento pode-se aplicar reverberadores, equalizadores, \emph{delays}
 e vários outros tipos de filtros para fins de tratamento sonoro ou
@@
 Ambas são respostas ao impulso
-que, convoluídas com a sequência sonora, resultam na própria sequência
+que, convoluidas com a sequência sonora, resultam na própria sequência
 com uma reverberação que se assemelha àquela do ambiente 
@@
 Ambas as possibilidades são exploradas
-por títulos em nossas referências, em especial~\cite{JOSFM,smith} e a coleção de filtros da comunidade \emph{Music-DSP}, da Universiade de Columbia.\cite{music-dsp,Openheim}
+por títulos em nossas referências, em especial~\cite{JOSFM,smith} e a coleção de filtros da comunidade \emph{Music-DSP}, da Universidade de Columbia.\cite{music-dsp,Openheim}
 
@@
 As variáveis $\mu$ e $\nu$ quantificam a intensidade do vibrato: $\mu$ é uma medida direta da quantidade
-de Herz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação ($2\nu$ é o número de semitons entre os picos superiores e inferiores de oscilação da frequência do som $\{t_i\}$ causada pelo vibrato).
+de Hertz envolvidos no limite superior da oscilação e $\nu$ é a medida direta de semitons envolvidos na oscilação ($2\nu$ é o número de semitons entre os picos superiores e inferiores de oscilação da frequência do som $\{t_i\}$ causada pelo vibrato).
 É conveniente $\nu=\log_{2}\frac{f+\mu}{f} $ neste caso pois o aumento máximo de frequência
@@
 em quaisquer frequência de oscilação e desvio de altura envolvidos\footnote{O desvio de altura
-é chamado profundidade do vibrato e é geralmetne dado por conveniência em semitons ou cents.}.
+é chamado profundidade do vibrato e é geralmente dado por conveniência em semitons ou cents.}.
 Estas oscilações com formas precisas e amplitudes arbitrárias não são praticáveis em instrumentos musicais tradicionais, introduzindo novidade nas possibilidades artísticas.
@@
 Os próprios valores absolutos destes impulsos resultam em proporções entre as amplitudes dos sinais
-convoluídos.
+convoluidos.
 Este recurso da métrica
@@
 
-\begin{equation}
+\begin{equation}\label{eq:fDoppler}
     f=\left(\frac{v_{som}+v_r}{v_{som}+v_s}\right)f_0
@@
 
-\begin{equation}
+\begin{equation}\label{eq:aDoppler}
     A_i=\left\{ \frac{z_0}{d_i}A_{\Delta P}\right\}_0^{\Lambda-1} = \left\{ \frac{z_0}{\sqrt{y_i^2+z_0^2}} \sqrt{\frac{v_r-v_s}{343,2}+1}  \,\right\}_0^{\Lambda-1}
@@
 
-\begin{equation}
+\begin{equation}\label{eq:ffDoppler}
     F_i=\{f_i\}_0^{\Lambda-1}=\left\{\frac{v_{som} + v_r\frac{y_i}{\sqrt{z_0^2+y_i^2}}}{v_{som}+v_s\frac{y_i}{\sqrt{z_0^2+y_i^2}}}f_0\right\}_0^{\Lambda-1}
@@
 exemplo do que ocorre com um coro de cantores, neste efeito o som é realizado com diversas pequenas modificações,
-potencialmente aleatórias, em paramêtros como frequência central, presença (ou ausência) de vibrato
+potencialmente aleatórias, em parâmetros como frequência central, presença (ou ausência) de vibrato
 e tremolo e suas características, equalizações, volume etc. Para o resultado final, estas versões do som
@@
 Com a mesma nomenclatura da subseção~\ref{subsec:spac} sobre espacialização, a reverberação tardia pode ser modelada como a convolução com um trecho de ruído colorido rosa, marrom ou preto, em decaimento exponencial de amplitude com relação também ao tempo.
-Assim, a atenuação nos agudos e irregularidade suave na resposta em frequência são contempladas de forma bastante satisfatória.\cite{JOSPhy,guillaume} Linhas de atrasos podem ser adicionadas como prefixo ao ruído com o decaimento, assim comtemplando as duas partes temporais da reverberação: as primeiras reflexões e a reverberação tardia.
+Assim, a atenuação nos agudos e irregularidade suave na resposta em frequência são contempladas de forma bastante satisfatória.\cite{JOSPhy,guillaume} Linhas de atrasos podem ser adicionadas como prefixo ao ruído com o decaimento, assim contemplado as duas partes temporais da reverberação: as primeiras reflexões e a reverberação tardia.
 Pode-se, para melhora de qualidade, calcular geometricamente a localização da última superfície em que cada frente de onda refletiu antes de chegar ao ouvido nos primeiros $100-200$ milissegundos e aplicar um filtro passa-baixas como em~\ref{subsec:filtros}.
 O ruído colorido pode ser introduzido gradualmente, desde o instante inicial dado pela incidência direta (i.e.
-sem reflexão alguma e dada pela DTI e DII), em \emph{fade-in}, atingindo o máximo no começo da 'reveberação tardia', quando as incidências geométricas perdem sua importância para as estatísticas do ruído em decaimento.
+sem reflexão alguma e dada pela DTI e DII), em \emph{fade-in}, atingindo o máximo no começo da 'reverberação tardia', quando as incidências geométricas perdem sua importância para as estatísticas do ruído em decaimento.
 
@@
 
-\begin{equation}
+\begin{equation}\label{eq:p1rev}
     R_i^1=\left\{r_i^1\right\}_0^{\Lambda_1-1}\;:\;r_i^1=\left\{
@@
 
-\begin{equation}
+\begin{equation}\label{eq:p2rev}
     R_i^2=\left\{r_i^2\right\}_{\Lambda_1}^{\Lambda_R-1}=\left\{10^{\frac{V_{dB}}{20}\frac{i}{\Lambda_R-1}}\,.\,r_i^m\right\}_{\Lambda_1}^{\Lambda_R-1}
@@
 Como:
-\begin{equation}
+\begin{equation}\label{eq:rev}
     R_i=\left\{r_i\right\}_0^{\Lambda_R-1}\;:\;r_i=\left\{
         \begin{array}{l l}
-            r_i^1\;  & \text{se }\quad i<\Lambda_1-1 \\
+            1\; & \text{se }\quad i=0 \\
+            r_i^1\;  & \text{se }\quad 1\leq i<\Lambda_1-1 \\
                                      r_i^2 \; & \text{se}\quad \Lambda_1 \leq i < \Lambda_R-1 \\
@@
 \end{figure}
+
+\afterpage{\blankpage}
 
@@
 Esta precisão absoluta é característica de implementações
-computacionais, os intrumentos reais possuem desvios destas frequências para melhor compatibilizar os harmônicos
+computacionais, os instrumentos reais possuem desvios destas frequências para melhor compatibilizar os harmônicos
 de suas notas. Além disso, a referência fixa $\varepsilon=2^{\frac{1}{12}}$ caracteriza
@@
 não deixa de ser uma \emph{convenção}
-adotada inclusive pela música erudita clássica de origem européia. Outras afinações
+adotada inclusive pela música erudita clássica de origem europeia. Outras afinações
 são incidentes. Para citar somente um exemplo, a
@@
 as terças e as sextas maiores (M) e menores (m). As dissonâncias fortes são as segundas menores e sétimas maiores. As dissonâncias
-brandas são as segundas maiores e as sétimas menores. O primeiro caso especial consite na quarta justa, que é consonante perfeita
+brandas são as segundas maiores e as sétimas menores. O primeiro caso especial consiste na quarta justa, que é consonante perfeita
 se considerada uma inversão da quinta justa, caso contrário pode ser considerada uma dissonância ou uma consonância imperfeita.
@@
 Algumas culturas entoam o intervalo como consonante.
-        \item Um intervalo maior, descrescido de um semitom, resulta em um intervalo menor. Um intervalo menor, acrescido de um semitom, resulta em um intervalo maior.
+        \item Um intervalo maior, decrescido de um semitom, resulta em um intervalo menor. Um intervalo menor, acrescido de um semitom, resulta em um intervalo maior.
         \item Um intervalo justo (uníssono, quarta justa, quinta justa, oitava justa), ou um intervalo maior (segunda maior 2M, terça maior 3M, sexta maior 6M ou sétima maior 7M), acrescido de um semitom, resulta em um intervalo aumentado (p.ex.
@@
         \item Um intervalo justo, ou um intervalo menor (segunda menor 2m, terça menor 3m, sexta menor 6m ou sétima menor 7m), decrescido de um semitom, resulta em um intervalo diminuto.
-A quinta diminuta é também chamda de trítono (5dim ~ tri).
-        \item Um intervalo aumentado, acrescido de um semitom, resulta em um intervalo 'mais que aumentado' e um intervalo diminuto, descrescido de um semitom, resulta em um intervalo 'mais que diminuto'.
+A quinta diminuta é também chamada de trítono (5dim ~ tri).
+        \item Um intervalo aumentado, acrescido de um semitom, resulta em um intervalo 'mais que aumentado' e um intervalo diminuto, decrescido de um semitom, resulta em um intervalo 'mais que diminuto'.
         \item Caso as notas soem simultaneamente, o intervalo é harmônico.
@@
 Veja nas subseções~\ref{subsec:motivos} e~\ref{subsec:usosmusicais3} para maiores detalhes.
-    \item Variações de intrumentação, articulação, espacialização e outras possibilidades de apresentação da estrutura de notas.
+    \item Variações de instrumentação, articulação, espacialização e outras possibilidades de apresentação da estrutura de notas.
 \end{itemize}
@@
 grau de uma escala, como das equações~\ref{eq:escalas}, formam os pilares do campo harmônico.
-A observação de progressões incidentes de acordes e regras para encadeamentos é o objeto de estudo da harmonia músical.
+A observação de progressões incidentes de acordes e regras para encadeamentos é o objeto de estudo da harmonia musical.
 Mesmo uma melodia monofônica gera campos harmônicos e pode-se observar acordes sugeridos por cada passagem.
@@
 A modulação é a mudança da tonalidade em que se encontra a musica. Caracteriza-se a modulação através da observação
-das tonalidades de partida e chegada e da forma de transição. As tonalidades são sempre tidas como relacionadas por quintas e suas relativas e antirelativas. São formas de efetuar a modulação:
+das tonalidades de partida e chegada e da forma de transição. As tonalidades são sempre tidas como relacionadas por quintas e suas relativas e antirrelativas. São formas de efetuar a modulação:
 \begin{itemize}
@@
 Seja o pulso o nível $j=0$ de agrupamento, o nível $j=-1$ 
-a primeira subdivisão do pulso, o nível $j=1$ a primeira algomeração dos pulsos e assim por diante. 
+a primeira subdivisão do pulso, o nível $j=1$ a primeira aglomeração dos pulsos e assim por diante. 
 Desta forma, $P_i^j$ é a $i$-ésima unidade de 
 pulsos no nível $j$ de agrupamento:
-$P^0_{10}$ é o décimo pulso, $P^{1}_3$ é a terceira unidade de agrupamento de pulsos (é possivel que seja o terceiro compasso),
+$P^0_{10}$ é o décimo pulso, $P^{1}_3$ é a terceira unidade de agrupamento de pulsos (é possível que seja o terceiro compasso),
 $P^{-1}_2$ é a segunda unidade da subdivisão do pulso.
@@
 
-Os arcos músicais são estruturas abstratas e passíveis de operações básicas. Um arco espectral, como um
+Os arcos musicais são estruturas abstratas e passíveis de operações básicas. Um arco espectral, como um
 acorde, pode ser invertido, ampliado e permutado, por exemplo. Os arcos temporais, como uma melodia, um motivo, um compasso ou
@@
 típicas de uma audição musical atenta. Esta atividade cortical contribui para a terapia
-músical, conhecida pela utilidade em casos de depressão e dano neurológico.
+musical, conhecida pela utilidade em casos de depressão e dano neurológico.
 Considera-se que as regiões do cérebro responsáveis pelo processamento auditivo são também usadas para outras atividades, incluindo linguísticas e matemáticas.\cite{Sacks,Roederer}
@@
 \centering
-\caption{Change Ringing: \emph{Peal} (padrão) com 3 sinos. As permutações estão entre as ordenações. Cada linha é uma ordednação dos sinos, cada ordenaçao é tocada, uma linha por vez.}
+\caption{Change Ringing: \emph{Peal} (padrão) com 3 sinos. As permutações estão entre as ordenações. Cada linha é uma ordenação dos sinos, cada ordenação é tocada, uma linha por vez.}
 \begin{tabular}{l c r}
@@
 Outro exemplo interessante é a presença simultânea de ritmos em diferentes métricas, constituindo o que chama-se de \emph{polirritmia}.
-A montagem musical \emph{Poli-hit mia} explora estas métricas simultâneas através de trem de impulsos convoluídos com as notas que compõem cada linha.
+A montagem musical \emph{Poli-hit mia} explora estas métricas simultâneas através de trem de impulsos convoluidos com as notas que compõem cada linha.
 Seu código está no Apêndice~\ref{ap:poli} e disponível online como parte da \massa.

### 2013-04-16 introducao.tex
feito apontamento para a ausencia de artigos sobre o tema
passado corretor ortografico

@@
 \chapter{Introdução} %Nome do capítulo.
-\setcounter{page}{29}
+\setcounter{page}{33}
 \label{cap:intro} 
@@
 A definição e expansão da nota como unidade fundamental da música são abordados nas seções~\ref{varInternas} e~\ref{sec:notaDisc}, respectivamente.
-A seção~\ref{notasMusica} trata da organização das notas em estruras de alto nível.\cite{Wisnick,Webern,Lerdahl,Cook,Lacerda}
+A seção~\ref{notasMusica} trata da organização das notas em estruturas de alto nível.\cite{Wisnick,Webern,Lerdahl,Cook,Lacerda}
 
@@
 O capítulo seguinte é um texto conciso em que os elementos musicais são apresentados junto às amostras temporais resultantes.
-Para validação e compartilhamento, as implementações em código computacional destas relações e de pequenas peças musicais\footnote{Ou, de forma menos pretenciosa, montagens musicais, sequências musicais.} foram reunidas em uma \emph{toolbox} chamada \massa\ e disponibilizadas \emph{online} e parcialmente nos Apêndices~\ref{cap:codigoProc} e~\ref{cap:codigoPecas}.
+Para validação e compartilhamento, as implementações em código computacional destas relações e de pequenas peças musicais\footnote{Ou, de forma menos pretensiosa, montagens musicais, sequências musicais.} foram reunidas em uma \emph{toolbox} chamada \massa\ e disponibilizadas \emph{online} e parcialmente nos Apêndices~\ref{cap:codigoProc} e~\ref{cap:codigoPecas}.
 
@@
 Assim, o Apêndice~\ref{cap:trabalhosRelacionados} é dedicado aos livros e implementações computacionais de interesse ou que apresentem similaridades com a descrição de elementos musicais em termos do áudio digital.
+Não há ênfase em artigos pois foram poucos os encontrados.
 A visita indica aspectos inéditos deste trabalho, em especial a descrição analítica de elementos musicais básicos em termos das amostras sonoras e a descrição natural, formal e concisa de técnicas tradicionais da música.

### 2013-04-16 musicaExtra.tex
passado corretor ortografico

@@
 Estas atividades extrapolam os usos musicais, especialmente devido às
-finalidades pedagódicas e sociais das práticas audiovisuais.
+finalidades pedagógicas e sociais das práticas audiovisuais.
 Assim, por completude, alguns materiais didáticos
@@
   códigos em C e a implementação como plugin lv2 estão no
-  repositório git como pode ser visto em \emph{online}\footnote{Este link é o \emph{README} do repositódio \emph{git} com os códigos dos
+  repositório git como pode ser visto em \emph{online}\footnote{Este link é o \emph{README} do repositório \emph{git} com os códigos dos
       plugins:
@@
 Já na remoção de ruídos as abordagens são as mais variadas e
-extremamente dependentes da aplicação. A complexidade dos algorítmos
+extremamente dependentes da aplicação. A complexidade dos algoritmos
 atinge níveis de especialidade em processamento de sinais que já
@@
 removedor de ruído \emph{'Hum'}. Este ruído é
-geralemte causado pela corrente alternada que
+geralmente causado pela corrente alternada que
 alimenta os equipamentos utilizados.
@@
 \emph{musical}, desde as amostras relativas a uma nota com dada
-frequencia, amplitude e timbre, até a confecção de uma ferramenta
+frequência, amplitude e timbre, até a confecção de uma ferramenta
 derivada, que já incorpora propostas musicais e estruturas mais
@@
 linha. Na versão atual, de
-2012 mas adiantada em 2011, os algorítmos em Python propriamente ditos
+2012 mas adiantada em 2011, os algoritmos em Python propriamente ditos
 somam 53 linhas e incluem apenas 5 funções. Com estas funções,
@@
 um \emph{waveshaping}. Este procedimento é bastante apreciado pela simplicidade e eficácia
-na síntese de tímbres diversos e ricos em harmônicos e evolução temporal. Embora
+na síntese de timbres diversos e ricos em harmônicos e evolução temporal. Embora
 uma explicação exaustiva do waveshaping fuja ao escopo deste trabalho, este
@@
 mais são do que usos típicos e quase caricatos destes procedimentos: no \emph{fazAcorde} a mixagem
-sobrepõe no tempo todas as unidades, no \emph{fazSequencia} as unidades são todas juntapostas no tempo.
+sobrepõe no tempo todas as unidades, no \emph{fazSequencia} as unidades são todas justapostas no tempo.
 
@@
 Já a síntese de estruturas compostas (e.g. sequências de acordes e sobreposição
-de linhas melódicas), é feita com os recursos usuais da liguagem:
+de linhas melódicas), é feita com os recursos usuais da linguagem:
 listas em Python e vetores Numpy (mais eficientes em tempo de
@@
 Para isso é bastante conveniente representar as permutações em classes próprias.
-A classe do padrão de permutação possui também um período de aplicação da permutação, ou seja, de quantas em quantas leituras da sequência apermutação é aplicada.
+A classe do padrão de permutação possui também um período de aplicação da permutação, ou seja, de quantas em quantas leituras da sequência a permutação é aplicada.
 
@@
 Depois disso devem ser escritos em um arquivo com os \emph{bits} como na convenção da linguagem C/C++.
-A bibioteca \emph{struct} cuida dessa escrita do inteiro no formato correto, e a biblioteca \emph{wave} escreve o cabeçalho no formato WAV adequado.
+A biblioteca \emph{struct} cuida dessa escrita do inteiro no formato correto, e a biblioteca \emph{wave} escreve o cabeçalho no formato WAV adequado.
 Assim, a clase de escrita do vetor sonoro em arquivo comum fica assim:
@@
 pela escrita. É uma linguagem
-com macros para execução de rítmos através de \emph{samples}
+com macros para execução de ritmos através de \emph{samples}
 em conjunto com instrumentos
@@
   \url{http://toplap.org}.}. Usualmente, o código é projetado para que a
-audiência possa ver o que está sendo escrito, no rítmo em que se
+audiência possa ver o que está sendo escrito, no ritmo em que se
 escreve, e se projeta também o resultante sonoro por autofalantes.

### 2013-04-16 pretextual.tex
feita revisao completa da dissertacao ateh 2.1
adiantada dissertacao ateh estruts ciclicas
passado corretor ortografico

@@
         \item[Equação~\ref{eq:vinculos}] Exemplo de vínculo entre a frequência e parâmetros do tremolo e do vibrato em um som.
+        \item[Equação~\ref{eq:fDoppler}] Equação básica da frequência observada no efeito Doppler.
+        \item[Equação~\ref{eq:aDoppler}] Amplitude observada no efeito Doppler.
+        \item[Equação~\ref{eq:ffDoppler}] Frequência efetiva observada no efeito Doppler.
+        \item[Equação~\ref{eq:p1rev}] Primeiro período de uma reverberação.
+        \item[Equação~\ref{eq:p2rev}] Segundo período de uma reverberação.
+        \item[Equação~\ref{eq:rev}] Resposta ao impulso de uma reverberação.
         \item[Equação~\ref{eq:adsr}] Envoltória ADSR com variações lineares e logarítmicas.
@@
 
-\pretextualchapter{Elementos de notação utilizados na dissertação}
+\pretextualchapter{Elementos de notação}
 	\begin{listaespecial2}[BIGNAMEWIDTH]
@@
         \item[$m$] metro.
-        \item[$mm$] milimetro.
+        \item[$mm$] milímetro.
         \item[$s$] segundo.
@@
         \item[$S_i$] sequência $S$ indexada em $i$.
-        \item[$S_i=\{s_i\}_x^{y}$] sequência $S_i$ com elementos $s_i$com índices de $x$ a $y$ incluso.
+        \item[$\{s_i\}_x^{y}$] sequência com elementos $s_i$ com índices de $x$ a $y$ incluso.
         \item[$\lfloor x \rfloor$] parte inteira de $x$.
@@
         \item[$\therefore$] portanto.
+        \item[$:$] tal que.
+        \item[$\forall$] para todo.
         \item[$\approx$] aproximadamente.

### 2013-04-16 trabalhosRelacionados.tex
passado corretor ortografico

@@
             \item {\bf Descrição:} livro do Perry Cook, de 2002, discorre sobre fundamentos de áudio digital e modelagem física.
-Possui preciosas dicas de modelagens de sons reais com características de intrumentos tradicionais e efeitos com origem nos instrumentos analógicos.
+Possui preciosas dicas de modelagens de sons reais com características de instrumentos tradicionais e efeitos com origem nos instrumentos analógicos.
             \item {\bf Aspecto complementar:} implementa diversos instrumentos musicais e efeitos sonoros.
@@
 Ao menos 5 livros de interesse para a dissertação: \emph{Mathematics of the Discrete Fourier Transform (DFT)} descreve a DFT minuciosamente, influindo FFT.
-O livro \emph{Introduction to Digital Filters (with audio applications)} trata de filtros de diversos tipos e descreve transfomadas, técnicas de design de filtros e análises de frequência e por polos e zeros.
-Além de abordar filtros não convencionais, o livro termina com implementações em matlab, Faust e PD e possibilidades de confeção de plugins.
+O livro \emph{Introduction to Digital Filters (with audio applications)} trata de filtros de diversos tipos e descreve transformadas, técnicas de design de filtros e análises de frequência e por polos e zeros.
+Além de abordar filtros não convencionais, o livro termina com implementações em matlab, Faust e PD e possibilidades de confecção de plugins.
 O livro \emph{Physical Audio Signal Processing (for virtual musical instruments and audio effects)} trata de modelagem física.
@@
 Também aborda Wavelets de forma superficial porém consistente.
-O livro \emph{Audio Signal Processing in Faust} trata da liguagem Faust e suas facilidades para o processamento de áudio e música.
+O livro \emph{Audio Signal Processing in Faust} trata da linguagem Faust e suas facilidades para o processamento de áudio e música.
             \item {\bf Aspecto complementar:} a seção "Perceptual Aspects of Reverberation", do livro de filtros, pode fechar o assunto da espacialização junto à subseção "2.1.7 Localização espacial".
@@
             \item {\bf Descrição:} livro Larry J. Salomon, de 1973 e revisado em 2002, é uma leitura de procedimentos composicionais tradicionais tendo como base comum a simetria.
-            \item {\bf Aspecto complementar:} é um escrito bastante pertinete para toda a seção "2.3 Organização de notas em música", principalmente para a parte "2.3.6 Estruturas cíclicas" com as rotações, reflexões, translados e incidências tradicionais.
+            \item {\bf Aspecto complementar:} é um escrito bastante pertinente para toda a seção "2.3 Organização de notas em música", principalmente para a parte "2.3.6 Estruturas cíclicas" com as rotações, reflexões, translados e incidências tradicionais.
             \item {\bf Aspecto diferencial:} o trabalho não relaciona as características musicais com o som como fenômeno físico.
@@
             \item {\bf Descrição:} editado por Richard Boulanger, 2000, possui contribuições de diversos autores.
-Além de abordar pincípios da linguagem, seus funcionamentos internos e paradigmas musicais, é um apanhado de técnicas de síntese musical do som e estruturas de mais alto nivel.
+Além de abordar princípios da linguagem, seus funcionamentos internos e paradigmas musicais, é um apanhado de técnicas de síntese musical do som e estruturas de mais alto nível.
             \item {\bf Aspecto complementar:} apresenta a paradigmática linguagem de programação voltada para música: CSound.
@@
             \item {\bf Aspecto complementar:} o capítulo "VII - Psychoacoustics" pode ser complementar à dissertação em algum detalhes e para aprofundamento.
-            \item {\bf Aspecto diferencial:} Os textos do livro são de alto nivel, sem preocupação central com o rigor, mas sim com apontamentos qualitativos e historicos das subáreas da música computacional.
+            \item {\bf Aspecto diferencial:} Os textos do livro são de alto nível, sem preocupação central com o rigor, mas sim com apontamentos qualitativos e históricos das subáreas da música computacional.
         \end{itemize}

### 2013-04-20 tex/phi+mus.tex
adicionado template para artigo

\documentclass[
 aip,
 jmp,
 amsmath,amssymb,
 reprint,
]{revtex4-1}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{multirow}
\usepackage{color}
\begin{document}
\preprint{AIP/123-QED}

\title{A quantitative approach to evolution of music
  and philosophy}

\author{Vilson Vieira}
 \homepage{http://automata.cc}
 \email{vilson@void.cc}
  \affiliation{ 
Instituto de F\'isica de S\~ao Carlos, Universidade de S\~ao Paulo (IFSC/USP)
}

\author{Renato Fabbri}
 \homepage{http://www.estudiolivre.org/el-user.php?view\_user=gk}
 \email{renato.fabbri@gmail.com}
 \altaffiliation[Also at ]{IFSC-USP}

\author{Gonzalo Travieso}
  \email{gonzalo@ifsc.usp.br}
 \altaffiliation[Also at ]{IFSC-USP}

\author{Osvaldo N. Oliveira Jr.}
  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
  \email{chu@ifsc.usp.br}
 \altaffiliation[Also at ]{IFSC-USP}

\author{Luciano da Fontoura Costa}
  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
  \email{ldfcosta@gmail.com}
 \altaffiliation[Also at ]{IFSC-USP}

\date{\today}
\begin{abstract}
The development of new statistical and computational methods is
increasingly making it possible to bridge the gap between hard
sciences and humanities. In this study, we propose an approach based
on a quantitative evaluation of attributes of objects in fields of
humanities, from which concepts such as dialectics and opposition are
formally defined mathematically. As case studies, we analyzed the
temporal evolution of classical music and philosophy by obtaining data
for 8 features characterizing the corresponding fields for 7
well-known composers and philosophers, which were treated with
multivariate statistics and pattern recognition methods. A bootstrap
method was applied to avoid statistical bias caused by the small
sample data set, with which hundreds of artificial composers and
philosophers were generated, influenced by the 7 names originally
chosen.  Upon defining indices for opposition, skewness and
counter-dialectics, we confirmed the intuitive analysis of historians
in that classical music evolved according to a master-apprentice
tradition, while in philosophy changes were driven by
opposition. Though these case studies were meant only to show the
possibility of treating phenomena in humanities quantitatively,
including a quantitative measure of concepts such as dialectics and
opposition the results are encouraging for further application of the
approach presented here to many other areas, since it is entirely
generic.
\end{abstract}
\pacs{89.75.Fb,05.65.+b} % PACS, the Physics and Astronomy
\keywords{arts, music, musicology, philosophy, pattern recognition, statistics}
\maketitle

\section{\label{sec:level1}Introduction}

Philosophy and natural sciences were born together in the 6th century
B.C. within Greek civilization~\cite{Russel}. Logic started with
Aristotle, geometry with Thales and Euclid, arithmetic with
Diophantus, while others created algebra, astronomy and politics, with
all fields being part of a common knowledge
space~\cite{Papineau}. What we know today as modern philosophy and
science were not originally separate, and arts were also present
almost universally in this space, not only as cultural expression. The
harmonic properties of music were of interest to philosophers such as
Pythagoras. Aristotle with his Poetics discussed a dramatic theory for
theatre. In this multidisciplinary space, music was addressed with
mathematics and science with philosophy. Quantitative methods were
applied to explain humanities, while the scientific method had its
origin in philosophy. The segregation of this space into philosophy,
science and arts was inevitable in the scientific revolution, thus
leading to its division into individual, independent areas. Although
specific, these areas are growing in complexity and their domains
overlap. Methods from specific areas are not sufficient to deal with
their complexity.

In this study, we cross the interdisciplinary borders and apply a
quantitative method to philosophy and music, in an attempt to
understand how humanities evolve. While philosophy and music have
their history constantly analyzed and discussed by critics, the nature
of their works --- text documents and music scores --- are difficult
to analyze and interpretation is always subjective. Here we propose a
generic approach to analyze features from any field in a quantitative
manner. The approach, described in the next Section, was then applied
to characterize composers of classical music and philosophers. The
data collected from scores assigned to these composers and
philosophers were treated with statistical and pattern recognition
methods, with the conclusions drawn being then compared to the
literature based on critics of music and philosophy.

\subsection{The Approach}

The approach devised to study the evolution of music and philosophy is
completely generic, and can be applied to any subject. It consists of
the following steps:

\begin{enumerate}
  \item{Given a subject (or area) to be studied, objects belonging to
    this area are identified. Here, philosophers and composers were
    chosen arbitrarily for the fields of philosophy and classical
    music, respectively. In other applications, the choice could be
    objective; for example, in a study of metropolitan areas the
    choice of the objects (cities) would obey a very precise criterion
    (population).}

  \item{A set of attributes is established, which are used to
    characterize the objects. In our case, a few attributes were
    chosen arbitrarily based on well-known characteristics in the
    fields of philosophy and music. In order to make the analysis by
    humans feasible, the number of attributes had to be low. But in
    other types of work, the set could be chosen based on objective
    criteria and the number of attributes could be large.}

  \item{For a quantitative analysis, scores are assigned for each of
    the attributes to each of the objects. In this study, three of the
    authors assigned scores based on their knowledge of the fields
    under study. This can be generalized, with scores assigned with
    objective criteria. Moreover, schemes may be applied to check the
    quality of the assignment, for example using the Kappa
    index~\cite{Cohen1960} to verify agreement among the people who
    assigned the scores.}

  \item{The data generated from the steps above may be sparse and in
    small amount, therefore unsuitable for the application of
    statistical methods. In order to overcome this limitation, we
    introduced a step to verify the robustness of the analysis, which
    consists in generating ``artificial data'' via a bootstrap
    method~\cite{Boos2003}. It is worth noting that in other
    applications, this creation of artificial data may not be
    needed. For instance, we could have taken data for a much larger
    number of philosophers and composers. However, in this paper this
    would hamper the task of manual score assignment, and
    interpretation of the final results would be much more difficult
    (with so many names to be analyzed).}

  \item{Because the study is comparative, metrics must be found to
    identify similarities (or dissimilarities) among the objects, and
    projection methods are applied to visualize the data. There are
    several possibilities for this task, and here we used Pearson
    correlation and principal component analysis~\cite{Costa} to
    analyze the distances among objects. It served to establish a time
    line, with which the evolution of philosophy and music could be
    studied. Furthermore, with this time line it was possible to
    analyze the data in terms of concepts such as dialectics and
    opposition, which were obtained in a quantitative manner.}
\end{enumerate}

With regard to the specific methodology in the present work, we first
identified prominent music composers and philosophers along history,
and established a set of main musical and philosophical
features. Grades were then assigned to each of the composers and
philosophers for all features. The assignment of scores was not
arbitrary, for they were based on research about techniques and styles
used by composers and philosophers. The grades reveal a \emph{tendency
  index} on characteristics of composers or philosophers. For example,
to say that Bach is more contrapuntist than the other composers
corresponds to assigning a higher grade --- e.g.\ 8.0 or 9.0 --- to
the Barroque master and smaller grades to others. We chose a reduced
set of philosophers and composers for the sake of simplicity and
clarity. Then, to avoid statistical bias owing to the small number of
samples, we developed a bootstrap method~\cite{Varian} with which a
larger data set of 1000 new \emph{artificial composers and
  philosophers} were generated, being directly influenced by the
original sample and representing the contemporaries of the
philosophers and composers chosen.

The scores assigned to the characteristics of each composer --- and
philosopher --- define a state vector in its feature space. This
quantification of composers and philosophers allowed the application
of sound quantitative concepts and methods from multivariate
statistics~\cite{Papoulis, Wichern, Therrien} and pattern
recognition~\cite{Duda, Costa}. Correlations between these
characteristic vectors were identified and principal component
analysis (PCA)~\cite{Costa} was applied to represent the music --- and
philosophy --- history as a planar space where development may be
followed as vectorial movements. On this planar space, concepts like
dialectics, innovation and opposition, originally non-quantitative,
can be modeled as mathematical relations between individual states.

It is important to note that application of statistical analysis to
music is not recent. In musicology, statistical methods have been used
to identify many musical
characteristics. Simonton~\cite{Simonton1991829, Simonton1977791} used
time-series analysis to measure the creative productivity of composers
based on their music and popularity. Kozbelt~\cite{Kozbelt01012009,
  Kozbelt01012007} also analyzed productivity, but based on the
measure of performance time of the compositions and investigated the
relation between productivity and versatility. More recent
works~\cite{Kranenburg2004, Kranenburg2007} used machine-learning
algorithms to recognize musical styles of selected compositions.

In contrast to the studies above, we are not interested in applying
statistical analysis to music but on characterizing composers by
identification of scores based on their styles. On the other hand,
automatic information retrieval --- that has been applied to music ---
is not common in philosophy. The method proposed here is a way to
analyze both fields independently of the nature of their works ---
i.e.\ music pieces and textual documents --- but based on a
well-formed opinion of reviewers or critics on these fields and their
analysis of temporal evolution.

\section{Mathematical Description}

Each composer or philosopher and their characteristics
(i.e.\ opposition and skewness) are defined for each pair of
subsequent composers or philosophers along time. Therefore, the choice
of composers and philosophers is crucial for the time-evolution
analysis.  A sequence $S$ of $C$ music composers and $P$ philosophers
was chosen based on their relevance in each period of classical music
and western philosophy history, respectively. The set of $C$
measurements defined a $C$-dimensional space, henceforth referred to
as the \emph{musical space}. Likewise, we defined a $P$-dimensional
\emph{philosophical space} based on $P$ measurements.

The vector $\vec{v}_i$ for each composer or philosopher $i$ defines a
corresponding \emph{composer state} in the music space or
\emph{philosopher state} in the philosophy space. For the set of $C$
composers and $P$ philosophers, we defined the same relations
summarized in Table \ref{tab:tableRelations}. Some details about these
relations are worth noting from Figure \ref{fig.1} for composers, and
similar remarks apply to philosophers. Given a set of $C$ composers as
a time-sequence $S$, the \emph{average state} at time $i$ is defined
as $\vec{a}_i$. The \emph{opposite state} is defined as the
``counterpoint'' of a music state $\vec{v}_i$, considering its average
state: everything running along the opposite direction of $\vec{v}_i$
is understood as opposition. In other words, any displacement from
$\vec{v}_i$ along the direction $\vec{r}_i$ is a \emph{contrary move},
and any displacement from $\vec{v}_i$ along the direction $-\vec{r}_i$
is an \emph{emphasis move}. Given a composer state $\vec{v}_i$ and its
opposite state $\vec{r}_i$, the \emph{opposition vector} $\vec{D}_i$
is defined.

\begin{table}
  \caption{\label{tab:tableRelations}Description of mathematical
    relations for each composer or philosopher $i$, $j$ and $k$, given
    a set of $C$ composers or $P$ philosophers as a time-sequence $S$.
    The \emph{opposition index} $W_{i,j}$ quantifies how much a
    composer or philosopher state $i$ is opposed to $j$. The
    \emph{skewness index} $_{i,j}$ quantifies the extent to which the
    new composer or philosopher state $j$ departs from the
    corresponding opposition state. And the \emph{counter-dialectics}
    index quantifies how much a state $k$ could be considered a
    \emph{synthesis} between the \emph{thesis} state $i$ and the
    \emph{antithesis} state $j$.}

  \begin{ruledtabular}
    \begin{tabular}{ll}
      \\ 
      Average state      & $\vec{a}_i = \frac{1}{i}\sum_{k=1}^i\vec{v}_k.$ 
      \\ \\
      Opposite state     & $\vec{r}_i = \vec{v}_i + 2(\vec{a}_i - \vec{v}_i)$ 
      \\ \\
      Opposition vector  & $\vec{D}_i=\vec{r}_i - \vec{v}_i$ 
      \\ \\
      Composer or \\ 
      philosopher state \\move & $\vec{M}_{i,j} = \vec{v}_j - \vec{v}_i$ 
      \\ \\
      Opposition index  & $W_{i,j} = \frac{\left< \vec{M}_{i,j}, \vec{D}_i\right>}
      {||\vec{D}_i||^2}$ 
      \\ \\
      Skewness index     & $s_{i,j} = \sqrt{\frac{|\vec{v}_i-\vec{v}_j|^2
          |\vec{a}_i-\vec{v}_i|^2 - 
          [(\vec{v}_i-\vec{v}_j) . 
            (\vec{a}_i-\vec{v}_i)]^2}
        {|\vec{a}_i-\vec{v}_i|^2}}$ 
      \\ \\
      Counter-dialectics \\ 
      index              & $d_{i \rightarrow k} = 
      \frac{|\left< \vec{v}_j-\vec{v}_i,\vec{v}_k \right> + 
        \frac{1}{2}\left<\vec{v}_i-\vec{v}_j, \vec{v}_i+\vec{v}_j\right>|}
           {|\vec{v}_j-\vec{v}_i|}$
           \\
    \end{tabular}
  \end{ruledtabular}
\end{table}

\begin{figure}
  \begin{center}
    \includegraphics[width=0.35\textwidth]{phil-space-desc_}
  \end{center}
  \caption{\it Graphical representation of the measures derived from a 
    \emph{composer state move}.}
  \label{fig.1}
\end{figure}

For the time-sequence $S$ relations between pairs of composers can be
defined, as follows. The move between two successive \emph{composer
  states} at time $i$ and $j$ corresponds to the $\vec{M}_{i,j}$
vector extending from $\vec{v}_i$ to $\vec{v}_j$. Given the composer
state we can quantify the intensity of opposition by the projection of
$\vec{M}_{i,j}$ along the opposition vector $\vec{D}_i$, normalized,
yielding the \emph{opposition index} $W_{i,j}$. With the same composer
state, the \emph{skewness index} $s_{i,j}$ is the distance between
$\vec{v}_j$ and the line $L_i$ defined by the vector $\vec{D}_i$, thus
quantifying the extent into which the new composer state departs from
the corresponding opposition state. A relationship between a triple of
successive composers can also be defined. Taking $i$, $j$ and $k$ as
the \emph{thesis}, \emph{antithesis} and \emph{synthesis} states, the
\emph{counter-dialectics index} $d_{i \rightarrow k}$ was defined by
the distance between the composer state $\vec{v}_k$ and the middle
line $ML_{i,j}$ defined by the thesis and antithesis, as shown in
Figure \ref{fig.2}. In higher dimensional musical or philosophical
spaces, the middle-hyperplane defined by the points which are at equal
distances to both $\vec{v}_i$ and $\vec{v}_j$ should be used instead
of the middle line $ML_{i,j}$. The proposed equation for
counter-dialectics scales to hyperplanes.  The counter-dialectics
index is used here, instead of a dialectics index, to maintain
compatibility with the use of a distance from point to line as adopted
for the definition of skewness.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.35\textwidth]{dialetica_.eps}
  \end{center}
  \caption{\it Graphical representation of the quantification of
    dialectics.}
  \label{fig.2}
\end{figure}

\section{Characteristics}

To create the music and philosophy spaces we derived eight variables
corresponding to distinct characteristics commonly found in music
compositions and works by philosophers. While the selected
characteristics cannot be considered a summary of all the relevant
features in music and philosophy history, they are initial indicators
that reveal differences in style of composers and philosophers. We
emphasize that the focus of this work is not on the specific
characteristics used or the scores assigned, which can be disputed,
but on the techniques for a quantitative analysis.

\subsection{Musical Characteristics}

These characteristics are related to basic elements of music ---
melody, harmony, rhythm, timbre, form and
tessitura~\cite{BennettHistory} --- in addition to non-musical issues
like historical events that influenced compositions, such as the
importance of the Church. The eight characteristics are listed
below:\\

{\bf \em{ Sacred - Secular}} ($S$-$S_c$): the sacred or religious
music is composed through religious influence or used for its
purposes. \textit{Masses}, \textit{motets} and hymns, dedicated to the
Christian liturgy, are well-known examples~\cite{Lovelock}. Secular
music has no or minimal relation with religion and includes popular
songs like Italian madrigals and German
\textit{lieds}~\cite{BennettHistory}.

{\bf \em{ Short duration - Long duration}} ($D_s$-$D_l$): compositions
quantified as short duration have a few minutes of execution. Long
duration compositions have at least 20 minutes of execution. The same
criterion was adopted by Kozbelt~\cite{Kozbelt01012009,
  Kozbelt01012007} in his analysis of time execution.

{\bf \em{ Harmony - Counterpoint}} ($H$-$C$): harmony regards the
vertical combination of notes, while counterpoint focuses on
horizontal combinations~\cite{BennettHistory}.

{\bf \em{ Vocal - Instrumental}} ($V$-$I$): compositions using just
vocals (e.g.\ \emph{cantata}) or exclusively instruments
(e.g.\ \emph{sonata}). Note the use of vocals over instruments on
Sacred compositions~\cite{Lovelock}.

{\bf \em{ Non-discursive - Discursive}} ($D_n$-$D$): compositions
based or not on verbal discourse, like programmatic music or Baroque
rhetoric, where the composer wants to ``tell a story'' invoking images
to the listeners mind~\cite{BennettHistory}. Its contrary part is
known as \textit{absolute music} where music was aimed to be
appreciated simply by what it is.

{\bf \em{ Motivic Stability - Motivic Variety}} ($M_s$-$M_v$): motivic
pieces present equilibrium between repetition, reuse and variation of
melodic motives. Bach is noticeable for his \textit{variation} of
motives, contrasting with the constantly inventive use of new
materials by Mozart~\cite{Webern}.

{\bf \em{ Rhythmic Simplicity - Rhythmic Complexity}} ($R_s$-$R_c$):
presence or not of polyrhythms, the use of independent rhythms at the
same time --- also known as \textit{rhythmic
  counterpoint}\cite{BennettHistory} --- a characteristic frequently
found in Romanticism and the works of 20th-century composers like
Stravinsky.

{\bf \em{ Harmonic Stability - Harmonic Variety}} ($H_s$-$H_v$): rate
of tonality change along a piece or its stability. After the highly
polyphonic development in Renaissance, Webern regarded Beethoven as
the composer who returned to the maximum exploration of harmonic
variety~\cite{Webern}.

\subsection{Characteristics in Philosophy}

We derived eight variables corresponding to some of the most recurrent
philosophical issues~\cite{Russel,Papineau,Deleuze2}. Each of these
variables, which defined an axis in the philosophy space, are briefly
described in the following. \\

{\bf \em{ Rationalism - Empiricism}} (\emph{R-E}): the rationalists
claim that the human acquaintance of knowledge/concepts is
significantly independent of sense experience. Empiricists understand
sensory experience as the main way to gain knowledge. Empiricism is in
the origin of the scientific method where knowledge must be based on
sensible observation of the world instead of faith or
intuition. Frequently, rationalists understand the world as affected
by intrinsic properties of the human brain, in contrast to the
empiricist approach where the world would imprint itself onto our
minds (the principle of \emph{``tabula rasa''}). For rationalists,
deduction is the superior method for investigation and privileges the
reason instead of experience as the source of true
knowledge. Historically, Descartes, Espinoza and Leibniz introduced
the rationalism in modern philosophy.

{\bf \em{ Essence - Existence}} (\emph{E-E}): An existence-based
understanding of the world has its basis on the fact that things are
as an existent unit. In Existentialism, the life of an individual is
determined by its self, that constitutes its essence, and not a
predefined essence that defines what is to be human. Essence focuses
on a substance (e.g.\ intellectual) that precedes existence
itself. For essentialists, any specific entity has essential
characteristics necessary for its function and identity. For example,
to have two legs and the ability to run is not an essential
characteristic defining humans, because other animals have the same
characteristics. Plato was one of the first essentialists while
Nietzsche and Kierkegaard were fundamental to Existentialism in the
19th century.

{\bf \em{ Monism - Dualism}} (\emph{M-D}): Dualism requires the
division of the human person into two or more domains, such as matter
and soul. Monism is based on a unique ``category of being''. Plato is
a recognized dualist while his disciple, Aristotle, was a notable
monist.

{\bf \em{ Theocentrism - Anthropocentrism}} (\emph{T-A}): In
theocentrism, God is the most important thing in the universe. The
anthropocentric view has man as prevalent, with Nietzsche being its
main representative. It is important to distinguish theocentrism from
deism, where the figure of God is considered as an abstract entity, as
for Espinoza, but does not play a central role in the universe.

{\bf \em{ Holism - Reductionism}} (\emph{H-R}): Reductionism attempts
to explain the world in terms of simple components and their emerging
properties. Holists focus on the fact that the whole is more than its
constitutive parts.

{\bf \em{ Deductionism - Phenomenology}} (\emph{D-P}): Phenomenology
relies on systematic reflection of consciousness and what happens in
conscious acts. Deductionism is based on deriving conclusions from
axiomatic systems.

{\bf \em{ Determinism - Free Will}} (\emph{D-F}): Free will assumes
that humans make choices, which are not predetermined. Determinism
understands that every event is fatidic, e.g.\, perfectly determined
by prior states.

{\bf \em{ Naturalism - Mechanism}} (\emph{N-M}): Methodological
naturalism is the thinking basis of modern science, i.e. hypotheses
must be argued and tested in terms of natural laws. Mechanism attempts
to build explanation using logic-mathematical processes.

\subsection{Bootstrap method for sampling}
\label{bootstrap}

To eliminate the bias intrinsic in a small sample group, we used a
bootstrap method for generating \emph{artificial composers and
  philosophers} contemporaries of those seven chosen. The bootstrap
routine generated new scores $\vec{r}$, which are not totally random
because they follow a probability distribution that models the
original $n = 7$ scores, given by $p(\vec{r}) = \sum^n_{i=1}
e^{\frac{d_i}{2\sigma^2}}$ where $d_i$ is the distance between a
random score $\vec{r}$ and the original score chart. For each step a
value $p(\vec{r})$ is generated and compared with a random normalized
value, as in the Monte Carlo~\cite{Robert2011} method to choose a set
of samples. These samples simulate new randomized composers and
philosophers score charts --- while preserving the historical
influence of the main 7 original names in each field. Higher values of
$p(\vec{r})$ imply a stronger influence of the original scores over
$\vec{r}$. For the analysis we used 1000 bootstrap samples obtained by
the bootstrap process together with the original scores, taking
$\sigma = 1.1$. Other values for $\sigma$ were used yielding
distributions with bootstrap samples that did not affect the music or
philosophy space substantially.

\section{Results and Discussion}

Memorable composers were chosen as key representatives of classical
music, which we believe had impact on their contemporaries, thus
creating a concise parallel with music history. The chronological
sequence is presented in Table \ref{tab:table0} with each composer
related to his historical period. The same was done for philosophy
where a set of seven philosophers were chosen spanning the period from
Classical Greece until contemporary times, and ordered chronologically
as: Plato, Aristotle, Descartes, Espinoza, Kant, Nietzsche and
Deleuze, as shown in Table \ref{tab:table0b}.

 \begin{table}[ht]
   \caption{\label{tab:table0} Sequence of music composers ordered
     chronologically with the period each represents.}
   \begin{tabular}{|l||l|}
     \hline
     Composer        & Movement \\ \hline
     Monteverdi      & Renaissance \\
     Bach            & Baroque \\
     Mozart          & Classical \\
     Beethoven       & Classical $\to$ Romantic \\
     Brahms          & Romantic \\
     Stravinsky      & 20th-century \\
     Stockhausen     & Contemporary\\
     \hline
   \end{tabular}
 \end{table}

\begin{table}[ht]
  \caption{\label{tab:table0b} Sequence of philosophers ordered
    chronologically with the period each represents.}
  \begin{tabular}{|l||l|}
    \hline
    Philosopher     & Era \\ \hline
    Plato           & Ancient \\
    Aristotle       & Ancient  \\
    Descartes       & 17th-century  \\
    Espinoza        & 17th-century  \\
    Kant            & 18th-century  \\
    Nietzsche       & 19th century  \\
    Deleuze         & 20th-century  \\
    \hline
  \end{tabular}
\end{table}

The quantification of the eight characteristics for music and
philosophy was performed jointly by three of the authors of this
article, based on research of history of music and western
philosophy. The scores shown in Tables \ref{tab:tableAmus} and
\ref{tab:tableAphi} for philosophers and composers, respectively, were
numerical values between 1 and 9. Values closer to 1 reveal the
composer or philosopher tended to the first element of each
characteristic pair, and vice-versa.

\begin{table}[ht]
  \caption{\label{tab:tableAmus} Quantification of eight music
    characteristics for each of the seven composers.}
  \begin{ruledtabular}
    \begin{tabular}{|l|c|c|c|c|c|c|c|c|}
      \footnotesize Composers    & \tiny  $S$-$S_c$ 
      & \tiny  $D_s$-$D_l$ & \tiny  $H$-$C$ & \tiny  $V$-$I$ 
      & \tiny  $D_n$-$D$ & \tiny  $M_s$-$M_v$ & \tiny  $R_s$-$R_c$
      & \tiny  $H_s$-$H_v$  \\
      \hline
   \footnotesize Monteverdi  & 3.0 & 8.0 & 5.0 & 3.0 & 7.0 & 5.0 & 3.0 & 7.0 
   \\
   \footnotesize Bach        & 2.0 & 6.0 & 9.0 & 2.0 & 8.0 & 2.0 & 1.0 & 5.0 
   \\
   \footnotesize Mozart      & 6.0 & 4.0 & 1.0 & 6.0 & 6.0 & 7.0 & 2.0 & 2.0 
   \\
   \footnotesize Beethoven   & 7.0 & 8.0 & 2.5 & 8.0 & 5.0 & 4.0 & 4.0 & 7.0 
   \\
   % Chopin & 9.0 & 3.0 & 3.0 & 9.0 & 5.5 & 8.0 & 7.0 & 8.0 \\
   \footnotesize Brahms      & 6.0 & 6.0 & 4.0 & 7.0 & 4.5 & 6.5 & 5.0 & 7.0 
   \\
   \footnotesize Stravinsky  & 8.0 & 7.0 & 6.0 & 7.0 & 8.0 & 5.0 & 8.0 & 5.0 
   \\
   \footnotesize Stockhausen & 7.0 & 4.0 & 8.0 & 7.0 & 5.0 & 8.0 & 9.0 & 6.0 
   \\
    \end{tabular}
  \end{ruledtabular}
\end{table}

\begin{table}%\footnotesize%\scriptsize%\tiny
  \caption{\label{tab:tableAphi} Quantification of eight philosophy
    characteristics for each of the seven philosophers.}

  \begin{ruledtabular}
    \begin{tabular}{|l||c|c|c|c|c|c|c|c|}
      Philosophers & R-E & E-E & M-D & T-A & H-R & D-P & D-F & N-M \\ \hline
      Plato        & 3.0 & 3.5 & 9.0 & 5.0 & 4.5 & 3.5 & 5.0 & 4.5 \\
      Aristotle    & 8.0 & 7.5 & 7.0 & 5.5 & 7.5 & 8.0 & 2.5 & 2.5 \\
      Descartes    & 1.5 & 2.5 & 9.0 & 6.5 & 7.0 & 2.5 & 7.5 & 7.5 \\
      Espinoza     & 8.0 & 2.0 & 1.0 & 5.0 & 2.0 & 3.0 & 1.0 & 1.0 \\
      Kant         & 7.0 & 2.5 & 8.5 & 6.5 & 4.5 & 3.5 & 7.5 & 5.0 \\
      Nietzsche    & 7.5 & 9.0 & 1.0 & 9.0 & 5.0 & 8.0 & 1.0 & 1.5 \\
      Deleuze      & 5.5 & 7.5 & 1.0 & 8.0 & 2.5 & 5.5 & 5.0 & 6.0 \\
    \end{tabular}
  \end{ruledtabular}
\end{table}

This data set defines an 8-dimensional space for music or philosophy
where each dimension corresponds to a characteristic that applies to
all 7 composers or philosophers. Such small data sets are not adequate
for statistical analysis, which could be biased. This is the reason
why we used the bootstrap method for sampling described in
section~\ref{bootstrap}.

For the extended data set after applying the bootstrap method, Pearson
correlation coefficients between the eight characteristics chosen are
given in Table \ref{tab:tableBmus} for composers and in Table
\ref{tab:tableBphi} for philosophers. The coefficient was 0.69 for the
pairs $S$-$S_c$ (Sacred or Secular) and $V$-$I$ (Vocal or
Instrumental), which indicates that sacred music tends to be more
vocal than instrumental. The coefficient 0.56 for the pairs $S$-$S_c$
and $R_s$-$R_c$ (Rhythmic Simplicity or Complexity) also shows that
this genre does not commonly use polyrhythms. A negative coefficient
of -0.33 for the pair $V$-$I$ and $D_n$-$D$ (Non-discursive or
Discursive) indicated that composers who used just voices on their
compositions also preferred programmatic music techniques such as
baroque rhetoric. Strong correlations were also observed for
philosophers. For instance, the Pearson coefficient of $-0.46$ for
\emph{R-E} and \emph{N-M} suggests that rationalists tend to be also
mechanists. An even stronger correlation of $0.74$, now positive, is
observed between \emph{E-E} and \emph{D-P}, i.e. existentialists also
tend to be phenomenologists, as could be expected. Other important
correlations appeared between \emph{D-F} and \emph{N-M} (coefficient =
$0.61$) and between \emph{M-D} and \emph{D-F}, which seems to be
directly implied by religious background.

\begin{table}[ht]
  \caption{\label{tab:tableBmus}Pearson correlation coefficients
    between the eight musical characteristics. The entries with
    absolute values $\geq$ 0.30 have been highlighted.}
  \begin{ruledtabular}
    \begin{tabular}{|c||c|c|c|c|c|c|c|c|}
      -    & \tiny  $S$-$S_c$ & \tiny  $D_s$-$D_l$ & \tiny  $H$-$C$ 
      & \tiny  $V$-$I$ & \tiny  $D_n$-$D$ & \tiny  $M_s$-$M_v$ 
      & \tiny  $R_s$-$R_c$ & \tiny  $H_s$-$H_v$  \\ \hline

      \tiny$S$-$S_c$ 
      & - & -0.2 & -0.06 & \bf{0.69} & -0.18 & 0.19  & \bf{0.56} & -0.16 
      \\
      \tiny$D_s$-$D_l$ 
      & - & - & -0.14 & -0.13 & 0.2 & \bf{-0.48} & -0.2 & \bf{0.37}
      \\
      \tiny$H$-$C$ 
      & - & - & - & -0.23 & 0.26 & 0.05 & \bf{0.46} & 0.03 
      \\
      \tiny$V$-$I$ 
      & - & - & - & - & \bf{-0.33} & 0.17 & \bf{0.42} & -0.06 
      \\
      \tiny$D_n$-$D$ 
      & - & - & - & - & - & -0.3 & 0.02 & -0.22 
      \\
      \tiny$M_s$-$M_v$ 
      & - & - & - & - & - & - & 0.26 & -0.15
      \\
      \tiny$R_s$-$R_c$ 
      & - & - & - & - & - & - & - & -0.02 
      \\
      \tiny$H_s$-$H_v$ 
      & - & - & - & - & - & - & - & - 
      \\
    \end{tabular}
  \end{ruledtabular}
\end{table}

\begin{table}\footnotesize
  \caption{\label{tab:tableBphi}Pearson correlation coefficients
    between the eight philosophical characteristics.  The entries with
    absolute values $\geq$ 0.35 have been highlighted.}
  \begin{ruledtabular}
    \begin{tabular}{|c||c|c|c|c|c|c|c|c|}
      - & R-E & E-E & M-D & T-A & H-R & D-P & D-F & N-M \\ \hline
      R-E & - & \bf{0.37} & -0.23 & 0.15 & 0.1 & \bf{0.46} & -0.27 & \bf{-0.46} 
      \\
      E-E & - & - & \bf{-0.53} & 0.19 & 0.15 & \bf{0.74} & \bf{-0.61} & -0.3 
      \\
      M-D & - & - & - & \bf{-0.43} & \bf{0.41} & -0.3 & \bf{0.35} & 0.01 
      \\
      T-A & - & - & - & - & -0.21 & 0.06  & 0.19 & 0.26 
      \\
      H-R & - & - & - & - & - & 0.32 & -0.22 & -0.25 
      \\ 
      D-P & - & - & - & - & - & - & \bf{-0.63} & \bf{-0.47} 
      \\
      D-F & - & - & - & - & - & - & - & \bf{0.61}
      \\
      N-M & - & - & - & - & - & - & - & - 
      \\
    \end{tabular}
  \end{ruledtabular}
\end{table}

The data were analyzed with principal component analysis (PCA), from
which it was found that several of the characteristics contributed to
the variability of the data, as shown in Tables~\ref{tab:Deviatesmus}
and \ref{tab:Deviatesphi} in the Supporting Information. Figures
\ref{fig:pcamus} and \ref{fig:pcaphi} display a 2-dimensional space
with the first two main axes. The arrows follow the time sequence
along with the seven composers and philosophers. Each of these arrows
corresponds to a vectorial move from one composer or philosopher state
to another. For clarity, just the lines of the arrows are
preserved. The bootstrap samples define clusters around the original
composers and philosophers. In subsidiary experiments, we verified
that the results from the bootstrap method were robust. This was
performed by applying 1000 perturbations of the original scores by
adding to each score the values -2, -1, 0, 1 or 2 with uniform
probability. In other words, we tested if scoring errors could be
sufficient to cause relevant effects on the PCA
projections. Interestingly, the values of average and standard
deviation for both original and perturbed positions listed in Tables
\ref{tab:tableDmus} and \ref{tab:tableDphi} in the Supporting
Information show relatively small changes. It is therefore reasonable
to assume that small errors in the scores assigned had no significant
effect on the overall analysis.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{g1}
  \end{center}
  \caption{\it 2-dimensional projected music space.}
  \label{fig:pcamus}
\end{figure}

Bach is found far from the rest of the composers, which suggests his
key role acknowledged by other great composers like Beethoven and
Webern~\cite{Webern}: ``In fact Bach composed everything, concerned
himself with everything that gives food for thought!''. The greatest
subsequent change takes place from Bach to Mozart, reflecting a
substantial difference in style. There is a strong relationship
between Beethoven and Brahms, supporting the belief by the
\textit{virtuosi} Hans von B\"{u}low~\cite{Bulow} when he stated that
the $1^{st}$ Symphony of Brahms was, in reality, the \textit{$10^{th}$
  Symphony of Beethoven}, appointing Brahms as the true successor of
Beethoven. Stravinsky is near Beethoven and Brahms, presumably due to
his heterogeneity~\cite{BennettHistory, Lovelock}. Beethoven is also
near Mozart, who deeply influenced Beethoven, mainly in his early
works. For Webern, Beethoven was the unique classicist who really came
close to the coherence found in the pieces of the Burgundian School:
``Not even in Haydn and Mozart do we see these two forms as clearly as
in Beethoven. The period and the eight-bar sentence are at their
purest in Beethoven; in his predecessors we find only traces of
them''~\cite{Webern}. It could explain the move of Beethoven in
direction of the Renaissance Monteverdi. Stockhausen is a deviating
point when compared with the others, which could be more even so had
we considered vanguard characteristics --- e.g.\ timbre exploration by
using electronic devices~\cite{Lovelock} --- not shared by his
precursors.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.45\textwidth]{pca_filosofos_novo}
  \end{center}
  \caption{\it 2-dimensional projected philosophy space.}
  \label{fig:pcaphi}
\end{figure}

The opposition and skewness indices for each of the six moves among
composers states in Table \ref{tab:tableOImus} indicate that the
movements were driven by small opposition and strong skewness. In
other words, most movements seem to seek innovation rather than
opposition. Furthermore, the counter-dialectics indices in Table
\ref{tab:tableEmus} are smaller than for the philosophers, as will be
discussed later on.

\begin{table}[ht]
  \caption{\label{tab:tableOImus}Opposition and skewness indices for
    each of the six composers states moves}
  \begin{tabular}{|c||c|c|}
    \hline
    Musical Move & $W_{i,j}$ & $s_{i,j}$ \\
    \hline \hline
    Monteverdi $\to$ Bach             &   1.0     &  0.      \\
    Bach $\to$ Mozart                 &   1.0196  &  1.9042  \\
    Mozart $\to$ Beethoven            &   0.4991  &  2.8665  \\
    Beethoven $\to$ Brahms            &   0.2669  &  1.7495  \\
    Brahms $\to$ Stravinsky           &   0.4582  &  2.6844  \\
    Stravinsky $\to$ Stockhausen      &   0.2516  &  3.1348  \\
    \hline
  \end{tabular}
\end{table}

\begin{table}[ht]
  \caption{\label{tab:tableEmus} Counter-dialectics index for each of
    the five subsequent pairs of moves among composers states for the
    8 components.}
  \begin{tabular}{|c||c|}
    \hline
    Musical Triple & $d_{i \rightarrow k}$ \\
    \hline \hline
    Monteverdi $\to$ Bach $\to$ Mozart          & 2.0586 \\
    Bach $\to$ Mozart $\to$ Beethoven           & 1.2020 \\
    Mozart $\to$ Beethoven $\to$ Brahms         & 1.0769 \\
    Beethoven $\to$ Brahms $\to$ Stravinsky     & 0.2518 \\
    Brahms $\to$ Stravinsky $\to$ Stockhausen   & 0.2549 \\
    \hline
  \end{tabular}
\end{table}

As for philosophy, Figure \ref{fig:pcaphi} shows an opposite movement
from Plato to Aristotle, which confirmed the antagonistic view of
Aristotle when compared with Plato, even though Aristotle was his
disciple~\cite{Russel}. Opposition is present along all the moves
among philosophers states. This oscillatory pattern makes it possible
to identify two well defined groups. The first contained Aristotle,
Espinoza, Nietzsche and Deleuze, in the left side of the graph. The
other contained Plato, Descartes and Kant,in the right side.  This
division is consistent with the points of view shared by each group
member. Another two groups are identified in the y-axis, separating
all the philosophers from Nietzsche and Deleuze, who are represented
by the most distant points.

When opposition and skewness indices in Table \ref{tab:tableOIphi} are
considered, all the moves among philosophers states tend to take place
according to a well-defined, intense opposition from the average
state. This was already noticeable in the PCA analysis. An interesting
relationship is the minor opposition and strong skewness between
Nietzsche and Deleuze, suggesting the return to Nietzsche as noted in
the works of Deleuze while considering the vanguard characteristics of
the 20th century philosopher~\cite{Deleuze2}. Espinoza tended toward
Nietzsche, and their similarity was admitted by Nietzsche, by naming
Espinoza his direct precursor~\cite{Russel}. While similar to
Espinoza, Nietzsche presented strong opposition to Kant, consistent
with Nietzsche being the strongest objector to Kant
ideas~\cite{Russel}. Also surprising was the rather small skewness
among most of the moves among philosophers states, which would be
driven almost exclusively by opposition to the current philosopher
state.  The results for dialectics in Table \ref{tab:tableEphi} show a
progressively stronger dialectics among subsequent pairs of moves in
philosophers states.

\begin{table}
  \caption{\label{tab:tableOIphi}Opposition and skewness indices for
    each of the six philosophers states moves.}
  \begin{tabular}{|c||c|c|}
    \hline
    Philosophical Move & $W_{i,j}$ & $s_{i,j}$ \\
    \hline \hline
    Plato $\rightarrow$ Aristotle     & 1.0    & 0 \\
    Aristotle $\rightarrow$ Descartes & 0.8740 & 1.1205 \\
    Descartes $\rightarrow$ Espinoza  & 0.9137 & 2.3856 \\
    Espinoza $\rightarrow$ Kant       & 0.6014 & 1.6842 \\
    Kant $\rightarrow$ Nietzsche      & 1.1102 & 2.9716 \\
    Nietzsche $\rightarrow$ Deleuze   & 0.3584 & 2.4890 \\
    \hline
  \end{tabular}
\end{table}

\begin{table}
  \caption{\label{tab:tableEphi}Counter-dialectics index for each of
    the five subsequent pairs of philosophers states moves.}
  \begin{tabular}{|c||c|}
    \hline
    Philosophical Triple & $d_{i \rightarrow k}$ \\
    \hline \hline
    Plato $\rightarrow$ Aristotle $\rightarrow$ Descartes    & 3.0198 \\
    Aristotle $\rightarrow$ Descartes $\rightarrow$ Espinoza & 1.8916 \\
    Descartes $\rightarrow$ Espinoza $\rightarrow$ Kant      & 1.1536 \\
    Espinoza $\rightarrow$ Kant $\rightarrow$ Nietzsche      & 1.1530 \\
    Kant $\rightarrow$ Nietzsche $\rightarrow$ Deleuze       & 0.2705 \\
    \hline
  \end{tabular}
\end{table}

The analysis above indicates distinct characteristics for the
evolution of classical music and philosophy. Philosophers seem to have
developed their ideas driven by opposition ($W_{i,j}$), as shown in
Table \ref{tab:tableOIphi}, while composers tend to be more influenced
by their predecessors according to the dialectics measurements
($1/d_{i \rightarrow k}$). In general, the movements among composers
had minor opposition, thus reflecting the master-apprentice
tradition. There is then a crucial difference in the \textit{memory
  treatment} along the development of philosophy and music: using the
same techniques, we verified that a philosopher was influenced by
opposition of ideas from his direct predecessor, while composers were
influenced by their two predecessors. We can argue that philosophy
exhibits a \textit{memory-1} state, while music presents
\textit{memory-2}, with \textit{memory-N} having $N$ of past
generations that influenced a philosopher or composer. Furthermore,
Figure \ref{fig:comparingdialectics} shows a constant decrease in the
counter-dialectics index, which means a constant return to the origins
for the development of music based on the search for unity. Using the
words of Webern, the search for the ``comprehensibility'' but always
influenced by their old masters.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{compara_dialeticas2}
  \end{center}
  \caption{\it Comparison between composers and philosophers
    counter-dialectics indices}
  \label{fig:comparingdialectics}
\end{figure}

The comparison between composers and philosophers is complemented with
the Wards hierarchical clustering~\cite{Ward}, which clusters the
original scores taking into account their distance. The generated
dendrogram in Figure \ref{fig:dendrogrammus} shows composers according
to their similarity, while the corresponding dendrogram for
philosophers is shown in Figure \ref{fig:dendrogramphi}. These
dendrograms are consistent with the previous observations. For
example, Beethoven and Brahms are close, reflecting their
heritage. Stravinsky and Stockhausen form another cluster, while
Mozart appears on its own, like Bach and Monteverdi. Both relations
were also present in the planar space in Figure \ref{fig:pcamus}.

\begin{figure}[ht]
        \begin{center}
          \includegraphics[width=0.45\textwidth]{Clust_Compositores.eps}
        \end{center}
        \caption{\it Wards hierarchical clustering of the seven composers.}
        \label{fig:dendrogrammus}
\end{figure}

\begin{figure}
        \begin{center}
                \includegraphics[width=0.5\textwidth]{Dendrogram_.eps}
        \end{center}
        \caption{\it Wards hierarchical clustering of the seven
          philosophers considering all the eight features.}
        \label{fig:dendrogramphi}
\end{figure}

\section{Concluding Remarks}

An approach has been proposed which allows for a quantitative
assessment of features from any given subject or area. For classical
music and philosophy investigated here, a time line could be
established based on the characteristics of composers and
philosophers. The quantitative analysis involved the collection of
data for these characteristics, extended with a bootstrap method to
yield robustness to the statistical analysis, and the use of
multivariate statistics. Also important was the establishment of
indices for opposition, skewness and counter-dialectics. Though the
emphasis of our work has been on the method of analysis, the
interpretation of the data was already sufficient to draw conclusions
that confirm intuition and subjective analyses of classical music and
philosophy.

For instance, the results pointed to the development in music
following a dichotomy: while composers aim at innovation, creating
their own styles, their technique is based on the work of their
predecessors, in a master-apprentice tradition. Indeed, in the history
of music, composers developed their own styles with a continuous
search for coherence or unity. In the words of Anton
Webern~\cite{Webern}, ``[...] ever since music has been written most
great artists have striven to make this unity ever clearer. Everything
that has happened aims at this [...]''. A frequent inheritance of
style can be identified from one composer to another as a gradual
development from its predecessor, contrasting with the necessity for
innovation. Quoting Lovelock: ``[...] by experiment that progress is
possible; it is the man with the forward-looking type of mind [...]
who forces man out of the rut of `what was good enough for my father
is good enough for me'.''~\cite{Lovelock}. In contrast, our
quantitative analysis confirmed that philosophy appears to exhibit a
well-defined trend in innovation: unlike music, the quest for
difference seems to drive philosophical changes as expressed by Gilles
Deleuze~\cite{Deleuze}. According to Ferdinand de Saussurre's
principle~\cite{Saussure}, concepts (words) tend to be different in
the sense of meaning distinct things.  The paradigm of difference is
particularly important because it is related to the own dynamics of
philosophical evolution.

Again emphasizing that our aim was to propose a generic, quantitative
method, we highlight some limitations of the specific analysis made
here for music and philosophy. For the scores and choice of main
characteristics in music and philosophy were largely arbitrary and
could be disputed. Nevertheless, the perturbation analysis performed
in this work suggests that the effect of non-systematic errors in
assigning the scores does not seem to be critical and has little
overall impact on the conclusions drawn. Most importantly, the formal
quantitative methodology described here may be combined with other
methods, including information retrieval and natural language
processing, to investigate other issues in humanities. For example,
still connected with the present work, it can be adapted to the
investigation of musical and philosophical schools, individual pieces
(e.g.\ music suites or books), or even the contributions from the same
composers or philosophers along distinct periods of time. Obviously,
this methodology can also be applied to other areas such as poetry,
cinema and science.

\begin{acknowledgments}
Luciano da F. Costa thanks CNPq (308231/03-1) and FAPESP (05/00587-5)
for sponsorship. Gonzalo Travieso thanks CNPq (308118/2010-3) for
sponsorship. Vilson Vieira is grateful to CAPES.
\end{acknowledgments}

\appendix

\section{A Brief Explanation of Principal Component Analysis (PCA)}

PCA is a dimensionality reduction procedure performed through rotation
of axes.  It operates by concentrating dispersion/variance along the
first new axes, which are referred to as the principal components. The
technique consists in finding the eigenvectors and eigenvalues of the
covariance matrix of the corresponding random vectors (i.e.\ the
vectors associated with each philosophical state). The eigenvalues
correspond to the variances of the new variables.  When multiplied by
the original feature matrix, the eigenvectors yield the new random
variables which are fully uncorrelated. For a more extensive
explanation of PCA, please refer to~\cite{Costa} and references
therein.

\section{Supporting Information}

Tables \ref{tab:Deviatesmus} and \ref{tab:Deviatesphi} show the
normalized weights of the contributions of each original property on
the eight axes for composers and philosophers. Most of the
characteristics contribute almost equally in defining the axes.

\begin{table}[ht]
  \caption{\label{tab:Deviatesmus}Percentages of
    the contributions from each musical characteristic on the eight
    new main axes.}
  \begin{tabular}{|c||c|c|c|c|c|c|c|c|}
    \hline
    Musical         & \multirow{2}{*}{$C_1$} & \multirow{2}{*}{$C_2$} 
    & \multirow{2}{*}{$C_3$} & \multirow{2}{*}{$C_4$} & \multirow{2}{*}{$C_5$}
    & \multirow{2}{*}{$C_6$} & \multirow{2}{*}{$C_7$} & \multirow{2}{*}{$C_8$}
    \\
    Charac. & & & & & & & & \\
    \hline
    $S$-$S_c$   
    &  19.78  &   4.04  & 10.38 & 10.60 &  17.55  &  36.60  &  4.41 &  0.63 \\
    $D_s$-$D_l$            
    &  13.63  &   9.21  & 19.17 &  3.55 &   3.13  &   1.65  & 25.55 & 24.05 \\
    $H$-$C$                
    &   1.44  &  26.62  & 8.26 & 13.97 &  21.71  &   7.76  & 13.98 & 12.20 \\
    $V$-$I$                
    &  18.35  &  12.82  & 9.29 &  8.02 &   9.37  &  40.95  &  2.12 &  2.03 \\
    $D_n$-$D$              
    &   6.31  &  10.73  & 15.48 & 26.29 &  4.04  &   1.86  & 25.29 &  2.35 \\
    $M_s$-$M_v$            
    &  16.94  &  13.28  & 15.03 &  4.84 &  32.25  &  1.70  &  2.62 &  4.37 \\
    $R_s$-$R_c$            
    &  14.13  &   3.26  & 15.58 & 13.80 &   7.48  &  1.88  &  1.36 & 35.99 \\
    $H_s$-$H_v$            
    &   9.38  &  20.00  &  6.75 & 18.88 &   4.45  &  7.56  & 24.62 & 18.36 \\
    \hline
  \end{tabular}
\end{table}

\begin{table}
  \caption{\label{tab:Deviatesphi}Percentages of
    the contributions from each philosophical characteristic to the four 
    new main axes.}
  \begin{tabular}{|c||c|c|c|c|c|c|c|c|}
    \hline
    Philos.      & \multirow{2}{*}{$C_1$} & \multirow{2}{*}{$C_2$} 
    & \multirow{2}{*}{$C_3$} & \multirow{2}{*}{$C_4$} & \multirow{2}{*}{$C_5$} 
    & \multirow{2}{*}{$C_6$} & \multirow{2}{*}{$C_7$} & \multirow{2}{*}{$C_8$}
    \\
    Charac. & & & & & & & & \\
    \hline
    R-E & 13.35   &  1.81 & 13.11 & 31.12 & 15.06 & 10.99 &  8.20 &  3.56 \\
    E-E & 18.13   &  7.88 &  2.10 & 14.68 & 11.41 &  5.31 &  7.98 & 33.50 \\
    M-D & 10.07   & 24.59 & 10.71 &  2.91 &  0.69 & 19.39 & 20.95 &  2.91 \\
    T-A &  1.08   & 24.18 & 21.51 &  0.68 & 23.72 &  6.31 &  7,27 &  2.06 \\
    H-R &  5.90   & 21.49 & 22.29 & 14.90 &  6.35 & 18.81 &  8.69 &  2.05 \\
    D-P & 18.99   &  1.29 &  6.85 &  8.41 &  9.67 & 20.50 & 11.47 & 24.93 \\
    D-F & 14.57   & 13.67 &  8.37 & 18.93 & 21.12 &  9.06 & 12.77 & 14.98 \\
    N-M & 17.86   &  5.06 & 15.03 &  8.33 & 11.96 &  9.59 & 22.62 & 15.97 \\
    \hline
  \end{tabular}
\end{table}

\begin{table}
  \caption{\label{tab:tableDmus}Averages and standard deviations
    for each composer and for the 8 eigenvalues.}

  \begin{tabular}{|c||c|c|}
    \hline
    Composers & $\mu_{\Delta}$ & $\sigma_{\Delta}$ \\
    \hline
    Monteverdi     & 3.7347 & 0.8503 \\
    Bach           & 5.3561 & 0.9379 \\
    Mozart         & 4.4319 & 0.8911 \\
    Beethoven      & 3.4987 & 0.7851 \\
    Brahms         & 3.0449 & 0.6996 \\
    Stravinsky     & 3.6339 & 0.7960 \\
    Stockhausen    & 4.2143 & 0.9029 \\
    \hline \hline
    Eigenvalues & $\mu_{\Delta}$ & $\sigma_{\Delta}$ \\
    \hline
    $\lambda_1$ &  -0.1759 & 0.0045 \\
    $\lambda_2$ &  -0.0638 & 0.0026 \\
    $\lambda_3$ &  -0.0411 & 0.0021 \\
    $\lambda_4$ &  -0.0144 & 0.0019 \\
    $\lambda_5$ &   0.0578 & 0.0021 \\
    $\lambda_6$ &   0.0736 & 0.0023 \\
    $\lambda_7$ &   0.0080 & 0.0027 \\
    $\lambda_8$ &   0.0835 & 0.0030 \\
    \hline
  \end{tabular}
\end{table}

\begin{table}
  \caption{\label{tab:tableDphi}Average and standard deviation for
    each philosopher and for the 8 eigenvalues.  }
  \begin{tabular}{|c||c|c|}
    \hline
    Philosophers & $\mu_{\Delta}$ & $\sigma_{\Delta}$ \\
    \hline
    Plato          & 3.3263 & 0.7673 \\
    Aristotle      & 4.0896 & 0.8930 \\
    Descartes      & 4.3081 & 0.9225 \\
    Espinoza       & 4.9709 & 0.9131 \\
    Kant           & 3.2845 & 0.7749 \\
    Nietzsche      & 5.3195 & 0.9797 \\
    Deleuze        & 4.0990 & 0.8970 \\
    \hline \hline
    Eigenvalues & $\mu_{\Delta}$ & $\sigma_{\Delta}$ \\
    \hline
    $\lambda_1$ & -0.2618 & 0.0068 \\
    $\lambda_2$ & -0.0976 & 0.0035 \\
    $\lambda_3$ &  0.0154 & 0.0025 \\
    $\lambda_4$ &  0.0212 & 0.0024 \\
    $\lambda_5$ &  0.0697 & 0.0026 \\
    $\lambda_6$ &  0.0807 & 0.0030 \\
    $\lambda_7$ &  0.0877 & 0.0032 \\
    $\lambda_8$ &  0.0846 & 0.0036 \\
    \hline
  \end{tabular}
\end{table}

\nocite{*}
\bibliography{phi+mus}
\end{document}

### 2013-04-20 tex/phi+mus.tex
feito abstract

@@
 \usepackage{color}
+
+\newcommand{\massa}{{\large \textsc{massa}}}
+\newcommand{\mass}{{\large \textsc{mass}}}
+\newcommand{\figgus}{{\large \textsc{figgus}}}
+
 \begin{document}
@@
 
-\title{A quantitative approach to evolution of music
-  and philosophy}
-
-\author{Vilson Vieira}
- \homepage{http://automata.cc}
- \email{vilson@void.cc}
+\title{Psychophysical description of musical elements in digital audio}
+
+\author{Renato Fabbri}
+ \homepage{http://www.estudiolivre.org/el-user.php?view\_user=gk}
+ \email{renato.fabbri@gmail.com}
   \affiliation{ 
@@
 
-\author{Renato Fabbri}
- \homepage{http://www.estudiolivre.org/el-user.php?view\_user=gk}
- \email{renato.fabbri@gmail.com}
- \altaffiliation[Also at ]{IFSC-USP}
-
-\author{Gonzalo Travieso}
-  \email{gonzalo@ifsc.usp.br}
+\author{Luciano da Fontoura Costa}
+  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
+  \email{ldfcosta@gmail.com}
  \altaffiliation[Also at ]{IFSC-USP}
@@
 
-\author{Luciano da Fontoura Costa}
-  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
-  \email{ldfcosta@gmail.com}
- \altaffiliation[Also at ]{IFSC-USP}
-
 \date{\today}
 \begin{abstract}
-The development of new statistical and computational methods is
-increasingly making it possible to bridge the gap between hard
-sciences and humanities. In this study, we propose an approach based
-on a quantitative evaluation of attributes of objects in fields of
-humanities, from which concepts such as dialectics and opposition are
-formally defined mathematically. As case studies, we analyzed the
-temporal evolution of classical music and philosophy by obtaining data
-for 8 features characterizing the corresponding fields for 7
-well-known composers and philosophers, which were treated with
-multivariate statistics and pattern recognition methods. A bootstrap
-method was applied to avoid statistical bias caused by the small
-sample data set, with which hundreds of artificial composers and
-philosophers were generated, influenced by the 7 names originally
-chosen.  Upon defining indices for opposition, skewness and
-counter-dialectics, we confirmed the intuitive analysis of historians
-in that classical music evolved according to a master-apprentice
-tradition, while in philosophy changes were driven by
-opposition. Though these case studies were meant only to show the
-possibility of treating phenomena in humanities quantitatively,
-including a quantitative measure of concepts such as dialectics and
-opposition the results are encouraging for further application of the
-approach presented here to many other areas, since it is entirely
-generic.
+
+    The representation of the basic elements of music - such as notes, ornaments and intervalar structures - in terms of discrete audio signal is often used in software for music creation and design.
+Nevertheless, there is no unified approach that relates these elements to the sound discrete samples.
+In this article, each musical element is described by equations that represent the sonic time samples, which are all implemented in scripts within an open source software toolbox, referred to as \massa\ (Music and Audio in Sequences and Samples).
+The fundamental element, the musical note with duration, volume, pitch and timbre, is related quantitatively to the characteristics of the discrete-time signal.
+Internal variations, such as tremolos, vibratos and spectral fluctuations, are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities.
+With this representation of notes, resources are provided for the generation of musical structures, such as rhythmic meter, pitch intervals and cycles.
+The efficacy of the natural representations was confirmed by the synthesis of small musical pieces using basic notes, incremented notes and notes in music.
+It is possible to synthesize whole albums through collage of the scripts and parameterization.
+    The sample-based analitical description, and the paradigm of open source implementation, enables cientific experiments in precise and trustful ways.
+In fact, \massa\ has already been employed by external users for diverse purposes.
+From these, it is mentioned a not identified in literature acoustic effect recognized by diverse individuals in mailing lists and uses related do art and education.
+
 \end{abstract}
-\pacs{89.75.Fb,05.65.+b} % PACS, the Physics and Astronomy
-\keywords{arts, music, musicology, philosophy, pattern recognition, statistics}
+\pacs{*43.66.-x,43.66.+y,05.10.-a} % PACS, the Physics and Astronomy
+\keywords{music, digital audio, statistics, psychophysics}
 \maketitle
@@
 \section{\label{sec:level1}Introduction}
-
-Philosophy and natural sciences were born together in the 6th century
-B.C. within Greek civilization~\cite{Russel}. Logic started with
-Aristotle, geometry with Thales and Euclid, arithmetic with
-Diophantus, while others created algebra, astronomy and politics, with
-all fields being part of a common knowledge
-space~\cite{Papineau}. What we know today as modern philosophy and
-science were not originally separate, and arts were also present
-almost universally in this space, not only as cultural expression. The
-harmonic properties of music were of interest to philosophers such as
-Pythagoras. Aristotle with his Poetics discussed a dramatic theory for
-theatre. In this multidisciplinary space, music was addressed with
-mathematics and science with philosophy. Quantitative methods were
-applied to explain humanities, while the scientific method had its
-origin in philosophy. The segregation of this space into philosophy,
-science and arts was inevitable in the scientific revolution, thus
-leading to its division into individual, independent areas. Although
-specific, these areas are growing in complexity and their domains
-overlap. Methods from specific areas are not sufficient to deal with
-their complexity.
-
-In this study, we cross the interdisciplinary borders and apply a
-quantitative method to philosophy and music, in an attempt to
-understand how humanities evolve. While philosophy and music have
-their history constantly analyzed and discussed by critics, the nature
-of their works --- text documents and music scores --- are difficult
-to analyze and interpretation is always subjective. Here we propose a
-generic approach to analyze features from any field in a quantitative
-manner. The approach, described in the next Section, was then applied
-to characterize composers of classical music and philosophers. The
-data collected from scores assigned to these composers and
-philosophers were treated with statistical and pattern recognition
-methods, with the conclusions drawn being then compared to the
-literature based on critics of music and philosophy.
 

### 2013-04-20 tex/phi+mus.tex
feita metade da introducao

@@
 
-\title{Psychophysical description of musical elements in digital audio}
+\title{Psychophysics of musical elements in discrete-time representation of sound}
 
@@
 Nevertheless, there is no unified approach that relates these elements to the sound discrete samples.
-In this article, each musical element is described by equations that represent the sonic time samples, which are all implemented in scripts within an open source software toolbox, referred to as \massa\ (Music and Audio in Sequences and Samples).
+In this article, each musical element is described by equations in sonic time samples, which are all implemented in scripts within an open source software toolbox, referred to as \massa\ (Music and Audio in Sequences and Samples).
 The fundamental element, the musical note with duration, volume, pitch and timbre, is related quantitatively to the characteristics of the discrete-time signal.
@@
 With this representation of notes, resources are provided for the generation of musical structures, such as rhythmic meter, pitch intervals and cycles.
-The efficacy of the natural representations was confirmed by the synthesis of small musical pieces using basic notes, incremented notes and notes in music.
+The efficacy of these physical descriptions of basic musical elements was confirmed by the synthesis of small musical pieces within each frame: basic notes, incremented notes and notes in music.
 It is possible to synthesize whole albums through collage of the scripts and parameterization.
@@
 In fact, \massa\ has already been employed by external users for diverse purposes.
-From these, it is mentioned a not identified in literature acoustic effect recognized by diverse individuals in mailing lists and uses related do art and education.
+Among these, it is mentioned an acoustic effect recognized by diverse individuals in mailing lists but not found in literature, and uses related to art and education.
 
@@
 \pacs{*43.66.-x,43.66.+y,05.10.-a} % PACS, the Physics and Astronomy
-\keywords{music, digital audio, statistics, psychophysics}
+\keywords{psychophysics, acoustics, statistics, signal processing, digital audio, music}
 \maketitle
@@
 
-\subsection{The Approach}
-
-The approach devised to study the evolution of music and philosophy is
-completely generic, and can be applied to any subject. It consists of
-the following steps:
+Representing musical structures and artifices by it's discrete sound characteristics is the purpuse of this work.
+The results include mathematical relations and it's computer program implementations.
+Next section exposes the theoretical description, which is implemented as scripts in a one-to-one relation to the equations.
+
+\subsection{Sound and digital audio}
+
+Sound is a longitudinal wave of mechanical pressure.
+The frequency bandwidth between $20Hz$ and $20kHz$ is appreciated by human hearing system with boudaries dependent on the person, climate conditions and sonic charachteristics itself~\cite{Roederer}.
+If considereded the speed of sound of $\approx 343.2 m/s$, this limits corresponds to $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.
+
+Human perception of sound envolves captivation by bones, stomach, ears, tranfer functions of head and processing dorso and nervous system.
+Besides that, the ear is a dedicated organ to the capture of this waves.
+Its mechanism decomposes sound into its sinusoidal spectrum and delivers them to the nervous system.
+This sinusoidal components are crucial to musical fenomena, as one can observe in the composition of sounds with musical interest and in tunnings and scales.
+Subsection~\ref{subsec:dicNote} exposes the presence of sinusoids in discrete-time audio and charachterizes a basic musical note.
+
+The representation of sound is called audio (although these terms are often used without distinction), and this can be provenient from caption by microphones or from sythesis.
+Often enough, digital audio is specified by protocols that eases file storage and transfering, in cost of a direct representation or even some loss in quality.
+Standard representation of digital audio, on the other hand, consists of samples equally spaced by $\lambda_s$ durations in time, with each sample spacified by a sample number of bits.
+This is called the Pulse Code Modulation representation of sound (PCM).
+A PCM digital sound is charachterized by it's sampling frequency $f_s=\frac{1}{\lambda_s}$, also called sampling rate, and bit depth, which is the number of bits used of representing the amplitude of each sample.
+Figure~\ref{fig:PCM} shows $25$ samples of a PCM áudio with $4$ bits each.
+The $2^4=16$ steps for each sample, together with the regular spacing $\lambda_s$ between them, introduces a quantization error.
+This noise, caused by this errors, diminishes as these spacing diminishes.
+
+\begin{figure*}
+    \centering
+        \includegraphics[width=\textwidth]{figures/pcm}
+        \caption{Pulse Code Modulation (PCM) audio: an analogic signal is represented by 25 samples with 4 bits each.}
+        \label{fig:PCM}
+\end{figure*}
+
+By the Nyquist theorem, it is known that half the sampling frequency is the maximum frequency of the signal.
+Thus, it is necessary to have a sampling frequency at least twice the highest frequency heard by humans $f_s \geq 2\times 20kHz = 40kHz$.
+This is the basis for the use of the sampling frequencies $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and Broadcast systems (Radio and TV), respectively.
+
+\subsection{Sonic art and musical theory}
+
+A common definition for music is art in sound and silence.
+For the average listener - and a reasonable part of specialists -the notion of music presupposes rythmic and pitch organization such as explained in subsection~\ref{subsec:notesMusic}.
+Music from the twentieth century enlarged this traditional comprehension of music.
+This occured in concert music, specially in the concrete, electronic and electroacoustic styles.
+On the last decade of the century, it was evident that popular music has also incorporated sounds without defined pitch and temporal organization out of simple metrics.
+Even though, the note stands paradigmatic as a 'fundamental unit' of musical structures and, in practice, it can unfold in sounds that observe this recent developments.
+The definition and expansion of the musical note as the fundamental unit of music is approached in subsections~\ref{subsec:discNote} and~\ref{subsec:internalVar}, respectively.
+Subsection~\ref{subsec:notesMusic} tackles the organization of musical notes in a higher level~\cite{Wisnick,Webern,Lerdhal,Cook,Lacerda}.
+
+Musical theory embody topics as diverse as psicho-acoustics, cultural manifestations and formalisms.
+The section~\ref{sec:results} point this topics as needed and designate external complements~\cite{Zamacois,Schoenberg,microsound}.
 

### 2013-04-21 article/phi+mus.tex
feito ateh comeco das notas no artigo, arrumado coisa pouca na dissertacao

--- 2013-04-20 tex/phi+mus.tex
@@
 The section~\ref{sec:results} point this topics as needed and designate external complements~\cite{Zamacois,Schoenberg,microsound}.
+
+\subsection{Computational implementation}
+The results presented in this article are implemented inf the form of scripts, i.e.
+small programs for better availability and validation of the models as acessible techology.
+This constitute the \massa\ toolbox, available in public domain in an open Git repository~\cite{gitBook}.
+This scripts are written in Python and make use of external libraries Numpy and Scikits/Audiolab that performs calls to Fortran routines for better computational efficiency.
+Part of this code has been trascribed to JavaScript and native Python with readiness, what points to uses of this contribution in web browsers such as Firefox and Chromium~\cite{numpy, audiolab, tutpython, python}.
+
+This are all open technologies, that is, published in licenses that allows copy, distribution and use of any part for research and derivatives.
+This way, the work here presented is available and eases co-authorship processes~\cite{Raymond,Lessig}.
+
+\subsection{Objectives}
+\label{subsec:objectives}
+The mais goal of this article is to present a concise set of relations among musical basic elements and sequences of PCM audio samples.
+The next section is a minimum text in which music elements are presented side-by-side with the discrete-time samples they result.
+As validation and sharing, implementations on computer code of these relations and little musical pieces where gathered the \massa\ toolbox, available online.
+
+Secondary objectives include presenting a framework of sound synthesis with control at sample level, with potential uses in psichoacoustical experiments and high-fidelity synthesis.
+The didatic presentation of the content favours use and apprehension on a problem which calls diverse topics to be tackled: signal processing, music and psycho-acoustics, to name just a few.
+
+\subsection{Related work}
+Due to the general interest, and number of knowledge areas involved, there is a number of books and computer implementations that are of interest or present similarities to what is presented in this work.
+A more detailed comparrison of them is pointed out in the bibliography~\cite{dissertacao}.
+There is almost no articles which could be found on the topic.
+In summary, there are computer implementations that use this analitycal descriptions implicitly, but there is no such a concise and mathemathical description of the processes implemented, as they aim to be libraries for sound and music.
+There are books on the topic that cover various aspects of effects and physical modeling, but none of them carry a concise description of musical elements and structures, but focus on aspects of musical sounds and ways to mimic traditional instruments.
+
+\section{Characterization of the discrete-time musical note}
+In diverse artistic and theoretical contexts, music is though of as being constituted by units called notes and this units taken as "atoms" that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
+Today, this notes are undestood as a central element of certain musical paradigms.
+In a cognitive sense, the notes are seen as discretizations that eases and enrich the flow of information through music~\cite{Roederer, Lacerda}
+Canonically, a musical note posses at least duration, colume, pitch and tibre~\cite{Lacerda}. These are qualities which can be maneged quantitatively dictated by the evenly time spaced sound samples.
+
+All the relations on this section are implemented at the file \emph{eqs2.1.py} of the \massa\ toolbox.
+Musical pieces \emph{5 sonic portraits} and \emph{reduced-fi} are available online as proof of concept.
+
+\subsection{Duration}
+The sample frequency $f_s$ is defined as the number os samples in each second of the discrete-time signal.
+Be $T_i=\{t_i\}$ an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
+A Musical note od duration $\Delta$ seconds is presented as a sequence of $\lfloor \Delta .
+f_s \rfloor $ samples.
+That is, taken the integer part of the multiplication, it is admited an error of at most $-\lambda_a$ seconds, which, for musical porpuses, are usually fine, for $f_s=44.1kHz \;\;\Rightarrow\;\;\lambda_s=\frac{1}{44100}\approx 23$ microseconds.
+It is reasonable to state:
+
+\begin{equation}\label{eq:dur}
+T_{i}^{\Delta}={\{t_i\}}_{i=0}^{\lfloor \Delta . f_a \rfloor -1}
+\end{equation}
+
+Be $\Lambda=\lfloor \Delta . f_a \rfloor$ the number of samples in the sequence, so that $T_i=\{t_i\}_0^{\Lambda-1}$.
+
+\subsection{Volume}\label{subsec:volume}
+The sensation of sound volume depends on reverberation and harmonic distribution, among other characteristics worked on section~\ref{sec:varInternas}.
+One can get variations on volume through the potency of the wave~\cite{Chowning}:
+
+\begin{equation}\label{eq:potencia}
+pot(T_i)=\frac{\sum_{i=0}^{\Lambda -1} t_i^2}{\Lambda}
+\end{equation}
 

### 2013-04-21 introducao.tex
feita metade da introducao
feito ateh comeco das notas no artigo, arrumado coisa pouca na dissertacao

@@
 Mesmo assim, a nota permanece paradigmática como 'unidade fundamental' das estruturas musicais e, na prática, pode se desdobrar em sons que contemplam estes desenvolvimentos recentes.
-A definição e expansão da nota como unidade fundamental da música são abordados nas seções~\ref{varInternas} e~\ref{sec:notaDisc}, respectivamente.
+A definição e expansão da nota como unidade fundamental da música são abordados nas seções~\ref{sec:notaDisc} e~\ref{varInternas}, respectivamente.
 A seção~\ref{notasMusica} trata da organização das notas em estruturas de alto nível.\cite{Wisnick,Webern,Lerdahl,Cook,Lacerda}
@@
 Estas tecnologias são todas abertas, i.e.
-estão publicadas em licenças que permitem o uso, cópia, distribuição e utilização de quaisquer partes para estudo e geração de produtos derivados.
+estão publicadas em licenças que permitem cópia, distribuição e utilização de quaisquer partes para estudo e geração de produtos derivados.
 Desta forma, o trabalho aqui descrito está disponível e facilita os processos de co-autoria\footnote{A comunidade e movimento chamada \emph{'Open Source'} entende a publicação de código computacional (e outras tecnologias) em licenças abertas como uma vantagem pragmática que facilita o desenvolvimento de \emph{software} e apresenta vantagens pedagógicas e mercadológicas.

### 2013-04-22 article/phi+mus.tex
artigo escrito ateh parte dos timbres

@@
 In diverse artistic and theoretical contexts, music is though of as being constituted by units called notes and this units taken as "atoms" that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
-Today, this notes are undestood as a central element of certain musical paradigms.
+Nowadays, this notes are undestood as a central element of certain musical paradigms.
 In a cognitive sense, the notes are seen as discretizations that eases and enrich the flow of information through music~\cite{Roederer, Lacerda}
@@
 The sample frequency $f_s$ is defined as the number os samples in each second of the discrete-time signal.
-Be $T_i=\{t_i\}$ an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
+Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
 A Musical note od duration $\Delta$ seconds is presented as a sequence of $\lfloor \Delta .
@@
 The sensation of sound volume depends on reverberation and harmonic distribution, among other characteristics worked on section~\ref{sec:varInternas}.
-One can get variations on volume through the potency of the wave~\cite{Chowning}:
+One can get variations of volume through the potency of the wave~\cite{Chowning}:
 
@@
 
-\begin{enumerate}
-  \item{Given a subject (or area) to be studied, objects belonging to
-    this area are identified. Here, philosophers and composers were
-    chosen arbitrarily for the fields of philosophy and classical
-    music, respectively. In other applications, the choice could be
-    objective; for example, in a study of metropolitan areas the
-    choice of the objects (cities) would obey a very precise criterion
-    (population).}
+The final volume is dependent on the speakers amplification of the signal.
+Thus, what matters is the relative potency of a note in relation to the ones around it or the potency of a music section in relation to the rest.
+Differences in volume are measured in decibels, and these are calculated directly form the amplitudes through energy or potency:
 
-  \item{A set of attributes is established, which are used to
-    characterize the objects. In our case, a few attributes were
-    chosen arbitrarily based on well-known characteristics in the
-    fields of philosophy and music. In order to make the analysis by
-    humans feasible, the number of attributes had to be low. But in
-    other types of work, the set could be chosen based on objective
-    criteria and the number of attributes could be large.}
+\begin{equation}\label{decibels}
+V_{dB}=10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)}
+\end{equation}
 
-  \item{For a quantitative analysis, scores are assigned for each of
-    the attributes to each of the objects. In this study, three of the
-    authors assigned scores based on their knowledge of the fields
-    under study. This can be generalized, with scores assigned with
-    objective criteria. Moreover, schemes may be applied to check the
-    quality of the assignment, for example using the Kappa
-    index~\cite{Cohen1960} to verify agreement among the people who
-    assigned the scores.}
+The quantity $V_{dB}$ has the decibel unit ($dB$). 
+To each $10dB$ it is associated a "doubled volume".
+Handy references are the $10dB$ for each step in the intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} e \emph{fortissimo}.
+Other useful references are que equivalent in $dB$ of doubling amplitude or potency:
 
-  \item{The data generated from the steps above may be sparse and in
-    small amount, therefore unsuitable for the application of
-    statistical methods. In order to overcome this limitation, we
-    introduced a step to verify the robustness of the analysis, which
-    consists in generating ``artificial data'' via a bootstrap
-    method~\cite{Boos2003}. It is worth noting that in other
-    applications, this creation of artificial data may not be
-    needed. For instance, we could have taken data for a much larger
-    number of philosophers and composers. However, in this paper this
-    would hamper the task of manual score assignment, and
-    interpretation of the final results would be much more difficult
-    (with so many names to be analyzed).}
+\begin{equation}\label{eq:ampVol}
+\begin{split}
+t_i^{'}=2 . t_i \Rightarrow pot(T^{'}_i)=4 . pot(T_i) \Rightarrow \\ \Rightarrow V^{'}_{dB}=10log_{10} 4 \approx 6 dB
+\end{split}
+\end{equation}
 
-  \item{Because the study is comparative, metrics must be found to
-    identify similarities (or dissimilarities) among the objects, and
-    projection methods are applied to visualize the data. There are
-    several possibilities for this task, and here we used Pearson
-    correlation and principal component analysis~\cite{Costa} to
-    analyze the distances among objects. It served to establish a time
-    line, with which the evolution of philosophy and music could be
-    studied. Furthermore, with this time line it was possible to
-    analyze the data in terms of concepts such as dialectics and
-    opposition, which were obtained in a quantitative manner.}
-\end{enumerate}
+\begin{equation}\label{eq:potVol}
+\begin{split}
+pot(T^{'}_i)=2 pot(T_i) \Rightarrow \\ \Rightarrow V^{'}_{dB}=10log_{10} 2 \approx 3 dB
+\end{split}
+\end{equation}
 
-With regard to the specific methodology in the present work, we first
-identified prominent music composers and philosophers along history,
-and established a set of main musical and philosophical
-features. Grades were then assigned to each of the composers and
-philosophers for all features. The assignment of scores was not
-arbitrary, for they were based on research about techniques and styles
-used by composers and philosophers. The grades reveal a \emph{tendency
-  index} on characteristics of composers or philosophers. For example,
-to say that Bach is more contrapuntist than the other composers
-corresponds to assigning a higher grade --- e.g.\ 8.0 or 9.0 --- to
-the Barroque master and smaller grades to others. We chose a reduced
-set of philosophers and composers for the sake of simplicity and
-clarity. Then, to avoid statistical bias owing to the small number of
-samples, we developed a bootstrap method~\cite{Varian} with which a
-larger data set of 1000 new \emph{artificial composers and
-  philosophers} were generated, being directly influenced by the
-original sample and representing the contemporaries of the
-philosophers and composers chosen.
+and the amplitude gain for a sequence whose volume has been doubled ($10dB$):
 
-The scores assigned to the characteristics of each composer --- and
-philosopher --- define a state vector in its feature space. This
-quantification of composers and philosophers allowed the application
-of sound quantitative concepts and methods from multivariate
-statistics~\cite{Papoulis, Wichern, Therrien} and pattern
-recognition~\cite{Duda, Costa}. Correlations between these
-characteristic vectors were identified and principal component
-analysis (PCA)~\cite{Costa} was applied to represent the music --- and
-philosophy --- history as a planar space where development may be
-followed as vectorial movements. On this planar space, concepts like
-dialectics, innovation and opposition, originally non-quantitative,
-can be modeled as mathematical relations between individual states.
+\begin{equation}\label{eq:dobraVol}
+\begin{split}
+10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)} = 10 \quad \Rightarrow \\ \Rightarrow \quad \sum_{i=0}^{\lfloor \Delta.f_a \rfloor -1}t^{'2}_i=10\sum_{i=0}^{\Lambda-1}t_i^2=\sum_{i=0}^{\Lambda-1}(\sqrt{10}.t_i)^2 \\
+\therefore \quad t^{'}_i=\sqrt{10}t_i \quad \Rightarrow \quad t^{'}_i \approx 3,16t_i
+\end{split}
+\end{equation}
 
-It is important to note that application of statistical analysis to
-music is not recent. In musicology, statistical methods have been used
-to identify many musical
-characteristics. Simonton~\cite{Simonton1991829, Simonton1977791} used
-time-series analysis to measure the creative productivity of composers
-based on their music and popularity. Kozbelt~\cite{Kozbelt01012009,
-  Kozbelt01012007} also analyzed productivity, but based on the
-measure of performance time of the compositions and investigated the
-relation between productivity and versatility. More recent
-works~\cite{Kranenburg2004, Kranenburg2007} used machine-learning
-algorithms to recognize musical styles of selected compositions.
+Thus, it is necessary a little bit more than to triplicate the amplitude for a doubled volume.
+This values are guides for the increases and decreases on the absolute values on the sample sequences with musical purposes.
+The direct conversion from decibels to amplitude gain (or attenuation) is:
 
-In contrast to the studies above, we are not interested in applying
-statistical analysis to music but on characterizing composers by
-identification of scores based on their styles. On the other hand,
-automatic information retrieval --- that has been applied to music ---
-is not common in philosophy. The method proposed here is a way to
-analyze both fields independently of the nature of their works ---
-i.e.\ music pieces and textual documents --- but based on a
-well-formed opinion of reviewers or critics on these fields and their
-analysis of temporal evolution.
+\begin{equation}\label{ampDec}
+A = 10^{\frac{V_{dB}}{20}}
+\end{equation}
 
-\section{Mathematical Description}
+Where $A$ is the mutiplicative factor that relates the amplitudes before and after the amplification.
 
-Each composer or philosopher and their characteristics
-(i.e.\ opposition and skewness) are defined for each pair of
-subsequent composers or philosophers along time. Therefore, the choice
-of composers and philosophers is crucial for the time-evolution
-analysis.  A sequence $S$ of $C$ music composers and $P$ philosophers
-was chosen based on their relevance in each period of classical music
-and western philosophy history, respectively. The set of $C$
-measurements defined a $C$-dimensional space, henceforth referred to
-as the \emph{musical space}. Likewise, we defined a $P$-dimensional
-\emph{philosophical space} based on $P$ measurements.
+\subsection{Pitch}
+Recapitulating, the musical particle (note) is a sequence $T_i$ which duration and volume corresponds to the size of sequence and the amplitude of its samples.
+The pitch is specified by the fundamental frequency $f_0$ whose cycle has duration $\delta_{f_0}=1/f_0$.
+This duration, multiplied by the sampling frequency $f_s$ results on the number of samples of the cycle $\lambda_{f_0}=f_a .
+\delta_{f_0} =f_a/f_0$.
 
-The vector $\vec{v}_i$ for each composer or philosopher $i$ defines a
-corresponding \emph{composer state} in the music space or
-\emph{philosopher state} in the philosophy space. For the set of $C$
-composers and $P$ philosophers, we defined the same relations
-summarized in Table \ref{tab:tableRelations}. Some details about these
-relations are worth noting from Figure \ref{fig.1} for composers, and
-similar remarks apply to philosophers. Given a set of $C$ composers as
-a time-sequence $S$, the \emph{average state} at time $i$ is defined
-as $\vec{a}_i$. The \emph{opposite state} is defined as the
-``counterpoint'' of a music state $\vec{v}_i$, considering its average
-state: everything running along the opposite direction of $\vec{v}_i$
-is understood as opposition. In other words, any displacement from
-$\vec{v}_i$ along the direction $\vec{r}_i$ is a \emph{contrary move},
-and any displacement from $\vec{v}_i$ along the direction $-\vec{r}_i$
-is an \emph{emphasis move}. Given a composer state $\vec{v}_i$ and its
-opposite state $\vec{r}_i$, the \emph{opposition vector} $\vec{D}_i$
-is defined.
+For didatic reasons, be $f_0$ such that it divides $f_s$ and $\lambda_{f_0}$ results an integer. If $T_i^f$ is a sonic sequence of fundamental frequency $f$, then:
 
-\begin{table}
-  \caption{\label{tab:tableRelations}Description of mathematical
-    relations for each composer or philosopher $i$, $j$ and $k$, given
-    a set of $C$ composers or $P$ philosophers as a time-sequence $S$.
-    The \emph{opposition index} $W_{i,j}$ quantifies how much a
-    composer or philosopher state $i$ is opposed to $j$. The
-    \emph{skewness index} $_{i,j}$ quantifies the extent to which the
-    new composer or philosopher state $j$ departs from the
-    corresponding opposition state. And the \emph{counter-dialectics}
-    index quantifies how much a state $k$ could be considered a
-    \emph{synthesis} between the \emph{thesis} state $i$ and the
-    \emph{antithesis} state $j$.}
+\begin{equation}\label{periodicidade}
+     T^f_i=\left\{ t_i^f \right\}=\left\{ t^f_{i+\lambda_{f}}  \right\}= \left\{ t^f_{i+\frac{f_a}{f}} \right\}
+\end{equation}
 
-  \begin{ruledtabular}
-    \begin{tabular}{ll}
-      \\ 
-      Average state      & $\vec{a}_i = \frac{1}{i}\sum_{k=1}^i\vec{v}_k.$ 
-      \\ \\
-      Opposite state     & $\vec{r}_i = \vec{v}_i + 2(\vec{a}_i - \vec{v}_i)$ 
-      \\ \\
-      Opposition vector  & $\vec{D}_i=\vec{r}_i - \vec{v}_i$ 
-      \\ \\
-      Composer or \\ 
-      philosopher state \\move & $\vec{M}_{i,j} = \vec{v}_j - \vec{v}_i$ 
-      \\ \\
-      Opposition index  & $W_{i,j} = \frac{\left< \vec{M}_{i,j}, \vec{D}_i\right>}
-      {||\vec{D}_i||^2}$ 
-      \\ \\
-      Skewness index     & $s_{i,j} = \sqrt{\frac{|\vec{v}_i-\vec{v}_j|^2
-          |\vec{a}_i-\vec{v}_i|^2 - 
-          [(\vec{v}_i-\vec{v}_j) . 
-            (\vec{a}_i-\vec{v}_i)]^2}
-        {|\vec{a}_i-\vec{v}_i|^2}}$ 
-      \\ \\
-      Counter-dialectics \\ 
-      index              & $d_{i \rightarrow k} = 
-      \frac{|\left< \vec{v}_j-\vec{v}_i,\vec{v}_k \right> + 
-        \frac{1}{2}\left<\vec{v}_i-\vec{v}_j, \vec{v}_i+\vec{v}_j\right>|}
-           {|\vec{v}_j-\vec{v}_i|}$
-           \\
-    \end{tabular}
-  \end{ruledtabular}
-\end{table}
+In the next secion, frequencies $f$ that does not divide $f_s$ will be considered. This restriction does not imply in lost of generality of this section's content.
 
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=0.35\textwidth]{phil-space-desc_}
-  \end{center}
-  \caption{\it Graphical representation of the measures derived from a 
-    \emph{composer state move}.}
-  \label{fig.1}
-\end{figure}
+\subsection{Timbre}
+While the period of the wave corresponds to a fundamental frequency, the trajectory of the wave inside the period - called the waveform - defines a harmonic spectrum and, thus, a timbre\footnote{The timbre is a subjective and complex characteristic.
+Physicaly, the timbre is multidimentionsl and given by the temporal dinamics behavior of enegy in spectral components that are harmonic or noisy.
+Beyond that, the word timbre is used to designate different things: one same note has different timbres, a same instrument has different timbres, two intruments of the same family possesses the same timbre that blends them in the same family and different timbres as they are different instruments.
+It is worth to mention that not all that is associated to timbre is manifested in spectral traces, even cultural or circunstancial aspects alter our perception of timbre}.
+Musically, it matters that sonic spectra with minimum differences result in timbres with crucial expressive differences and that, hence, diffetent timbres can be produced by using different spectra\cite{Roederer}.
 
-For the time-sequence $S$ relations between pairs of composers can be
-defined, as follows. The move between two successive \emph{composer
-  states} at time $i$ and $j$ corresponds to the $\vec{M}_{i,j}$
-vector extending from $\vec{v}_i$ to $\vec{v}_j$. Given the composer
-state we can quantify the intensity of opposition by the projection of
-$\vec{M}_{i,j}$ along the opposition vector $\vec{D}_i$, normalized,
-yielding the \emph{opposition index} $W_{i,j}$. With the same composer
-state, the \emph{skewness index} $s_{i,j}$ is the distance between
-$\vec{v}_j$ and the line $L_i$ defined by the vector $\vec{D}_i$, thus
-quantifying the extent into which the new composer state departs from
-the corresponding opposition state. A relationship between a triple of
-successive composers can also be defined. Taking $i$, $j$ and $k$ as
-the \emph{thesis}, \emph{antithesis} and \emph{synthesis} states, the
-\emph{counter-dialectics index} $d_{i \rightarrow k}$ was defined by
-the distance between the composer state $\vec{v}_k$ and the middle
-line $ML_{i,j}$ defined by the thesis and antithesis, as shown in
-Figure \ref{fig.2}. In higher dimensional musical or philosophical
-spaces, the middle-hyperplane defined by the points which are at equal
-distances to both $\vec{v}_i$ and $\vec{v}_j$ should be used instead
-of the middle line $ML_{i,j}$. The proposed equation for
-counter-dialectics scales to hyperplanes.  The counter-dialectics
-index is used here, instead of a dialectics index, to maintain
-compatibility with the use of a distance from point to line as adopted
-for the definition of skewness.
+The simplest case (and most important, as the following texts shows) in that of the spectrum that consists only of the fundamental frequency $f$ itself.
+This is the sinusoid, frequency in pure oscilatory movement called 'simple harmonic movement'.
+Be $S_i^f$ a sequence whose samples $s_i^f$ describes a sibusoid of frequency $f$:
 
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=0.35\textwidth]{dialetica_.eps}
-  \end{center}
-  \caption{\it Graphical representation of the quantification of
-    dialectics.}
-  \label{fig.2}
-\end{figure}
-
-\section{Characteristics}
-
-To create the music and philosophy spaces we derived eight variables
-corresponding to distinct characteristics commonly found in music
-compositions and works by philosophers. While the selected
-characteristics cannot be considered a summary of all the relevant
-features in music and philosophy history, they are initial indicators
-that reveal differences in style of composers and philosophers. We
-emphasize that the focus of this work is not on the specific
-characteristics used or the scores assigned, which can be disputed,
-but on the techniques for a quantitative analysis.
-
-\subsection{Musical Characteristics}
-
-These characteristics are related to basic elements of music ---
-melody, harmony, rhythm, timbre, form and
-tessitura~\cite{BennettHistory} --- in addition to non-musical issues
-like historical events that influenced compositions, such as the
-importance of the Church. The eight characteristics are listed
-below:\\
-
-{\bf \em{ Sacred - Secular}} ($S$-$S_c$): the sacred or religious
-music is composed through religious influence or used for its
-purposes. \textit{Masses}, \textit{motets} and hymns, dedicated to the
-Christian liturgy, are well-known examples~\cite{Lovelock}. Secular
-music has no or minimal relation with religion and includes popular
-songs like Italian madrigals and German
-\textit{lieds}~\cite{BennettHistory}.
-
-{\bf \em{ Short duration - Long duration}} ($D_s$-$D_l$): compositions
-quantified as short duration have a few minutes of execution. Long
-duration compositions have at least 20 minutes of execution. The same
-criterion was adopted by Kozbelt~\cite{Kozbelt01012009,
-  Kozbelt01012007} in his analysis of time execution.
-
-{\bf \em{ Harmony - Counterpoint}} ($H$-$C$): harmony regards the
-vertical combination of notes, while counterpoint focuses on
-horizontal combinations~\cite{BennettHistory}.
-
-{\bf \em{ Vocal - Instrumental}} ($V$-$I$): compositions using just
-vocals (e.g.\ \emph{cantata}) or exclusively instruments
-(e.g.\ \emph{sonata}). Note the use of vocals over instruments on
-Sacred compositions~\cite{Lovelock}.
-
-{\bf \em{ Non-discursive - Discursive}} ($D_n$-$D$): compositions
-based or not on verbal discourse, like programmatic music or Baroque
-rhetoric, where the composer wants to ``tell a story'' invoking images
-to the listeners mind~\cite{BennettHistory}. Its contrary part is
-known as \textit{absolute music} where music was aimed to be
-appreciated simply by what it is.
-
-{\bf \em{ Motivic Stability - Motivic Variety}} ($M_s$-$M_v$): motivic
-pieces present equilibrium between repetition, reuse and variation of
-melodic motives. Bach is noticeable for his \textit{variation} of
-motives, contrasting with the constantly inventive use of new
-materials by Mozart~\cite{Webern}.
-
-{\bf \em{ Rhythmic Simplicity - Rhythmic Complexity}} ($R_s$-$R_c$):
-presence or not of polyrhythms, the use of independent rhythms at the
-same time --- also known as \textit{rhythmic
-  counterpoint}\cite{BennettHistory} --- a characteristic frequently
-found in Romanticism and the works of 20th-century composers like
-Stravinsky.
-
-{\bf \em{ Harmonic Stability - Harmonic Variety}} ($H_s$-$H_v$): rate
-of tonality change along a piece or its stability. After the highly
-polyphonic development in Renaissance, Webern regarded Beethoven as
-the composer who returned to the maximum exploration of harmonic
-variety~\cite{Webern}.
-
-\subsection{Characteristics in Philosophy}
-
-We derived eight variables corresponding to some of the most recurrent
-philosophical issues~\cite{Russel,Papineau,Deleuze2}. Each of these
-variables, which defined an axis in the philosophy space, are briefly
-described in the following. \\
-
-{\bf \em{ Rationalism - Empiricism}} (\emph{R-E}): the rationalists
-claim that the human acquaintance of knowledge/concepts is
-significantly independent of sense experience. Empiricists understand
-sensory experience as the main way to gain knowledge. Empiricism is in
-the origin of the scientific method where knowledge must be based on
-sensible observation of the world instead of faith or
-intuition. Frequently, rationalists understand the world as affected
-by intrinsic properties of the human brain, in contrast to the
-empiricist approach where the world would imprint itself onto our
-minds (the principle of \emph{``tabula rasa''}). For rationalists,
-deduction is the superior method for investigation and privileges the
-reason instead of experience as the source of true
-knowledge. Historically, Descartes, Espinoza and Leibniz introduced
-the rationalism in modern philosophy.
-
-{\bf \em{ Essence - Existence}} (\emph{E-E}): An existence-based
-understanding of the world has its basis on the fact that things are
-as an existent unit. In Existentialism, the life of an individual is
-determined by its self, that constitutes its essence, and not a
-predefined essence that defines what is to be human. Essence focuses
-on a substance (e.g.\ intellectual) that precedes existence
-itself. For essentialists, any specific entity has essential
-characteristics necessary for its function and identity. For example,
-to have two legs and the ability to run is not an essential
-characteristic defining humans, because other animals have the same
-characteristics. Plato was one of the first essentialists while
-Nietzsche and Kierkegaard were fundamental to Existentialism in the
-19th century.
-
-{\bf \em{ Monism - Dualism}} (\emph{M-D}): Dualism requires the
-division of the human person into two or more domains, such as matter
-and soul. Monism is based on a unique ``category of being''. Plato is
-a recognized dualist while his disciple, Aristotle, was a notable
-monist.
-
-{\bf \em{ Theocentrism - Anthropocentrism}} (\emph{T-A}): In
-theocentrism, God is the most important thing in the universe. The
-anthropocentric view has man as prevalent, with Nietzsche being its
-main representative. It is important to distinguish theocentrism from
-deism, where the figure of God is considered as an abstract entity, as
-for Espinoza, but does not play a central role in the universe.
-
-{\bf \em{ Holism - Reductionism}} (\emph{H-R}): Reductionism attempts
-to explain the world in terms of simple components and their emerging
-properties. Holists focus on the fact that the whole is more than its
-constitutive parts.
-
-{\bf \em{ Deductionism - Phenomenology}} (\emph{D-P}): Phenomenology
-relies on systematic reflection of consciousness and what happens in
-conscious acts. Deductionism is based on deriving conclusions from
-axiomatic systems.
-
-{\bf \em{ Determinism - Free Will}} (\emph{D-F}): Free will assumes
-that humans make choices, which are not predetermined. Determinism
-understands that every event is fatidic, e.g.\, perfectly determined
-by prior states.
-
-{\bf \em{ Naturalism - Mechanism}} (\emph{N-M}): Methodological
-naturalism is the thinking basis of modern science, i.e. hypotheses
-must be argued and tested in terms of natural laws. Mechanism attempts
-to build explanation using logic-mathematical processes.
-
-\subsection{Bootstrap method for sampling}
-\label{bootstrap}
-
-To eliminate the bias intrinsic in a small sample group, we used a
-bootstrap method for generating \emph{artificial composers and
-  philosophers} contemporaries of those seven chosen. The bootstrap
-routine generated new scores $\vec{r}$, which are not totally random
-because they follow a probability distribution that models the
-original $n = 7$ scores, given by $p(\vec{r}) = \sum^n_{i=1}
-e^{\frac{d_i}{2\sigma^2}}$ where $d_i$ is the distance between a
-random score $\vec{r}$ and the original score chart. For each step a
-value $p(\vec{r})$ is generated and compared with a random normalized
-value, as in the Monte Carlo~\cite{Robert2011} method to choose a set
-of samples. These samples simulate new randomized composers and
-philosophers score charts --- while preserving the historical
-influence of the main 7 original names in each field. Higher values of
-$p(\vec{r})$ imply a stronger influence of the original scores over
-$\vec{r}$. For the analysis we used 1000 bootstrap samples obtained by
-the bootstrap process together with the original scores, taking
-$\sigma = 1.1$. Other values for $\sigma$ were used yielding
-distributions with bootstrap samples that did not affect the music or
-philosophy space substantially.
-
-\section{Results and Discussion}
-
-Memorable composers were chosen as key representatives of classical
-music, which we believe had impact on their contemporaries, thus
-creating a concise parallel with music history. The chronological
-sequence is presented in Table \ref{tab:table0} with each composer
-related to his historical period. The same was done for philosophy
-where a set of seven philosophers were chosen spanning the period from
-Classical Greece until contemporary times, and ordered chronologically
-as: Plato, Aristotle, Descartes, Espinoza, Kant, Nietzsche and
-Deleuze, as shown in Table \ref{tab:table0b}.
-
- \begin{table}[ht]
-   \caption{\label{tab:table0} Sequence of music composers ordered
-     chronologically with the period each represents.}
-   \begin{tabular}{|l||l|}
-     \hline
-     Composer        & Movement \\ \hline
-     Monteverdi      & Renaissance \\
-     Bach            & Baroque \\
-     Mozart          & Classical \\
-     Beethoven       & Classical $\to$ Romantic \\
-     Brahms          & Romantic \\
-     Stravinsky      & 20th-century \\
-     Stockhausen     & Contemporary\\
-     \hline
-   \end{tabular}
- \end{table}
-
-\begin{table}[ht]
-  \caption{\label{tab:table0b} Sequence of philosophers ordered
-    chronologically with the period each represents.}
-  \begin{tabular}{|l||l|}
-    \hline
-    Philosopher     & Era \\ \hline
-    Plato           & Ancient \\
-    Aristotle       & Ancient  \\
-    Descartes       & 17th-century  \\
-    Espinoza        & 17th-century  \\
-    Kant            & 18th-century  \\
-    Nietzsche       & 19th century  \\
-    Deleuze         & 20th-century  \\
-    \hline
-  \end{tabular}
-\end{table}
-
-The quantification of the eight characteristics for music and
-philosophy was performed jointly by three of the authors of this
-article, based on research of history of music and western
-philosophy. The scores shown in Tables \ref{tab:tableAmus} and
-\ref{tab:tableAphi} for philosophers and composers, respectively, were
-numerical values between 1 and 9. Values closer to 1 reveal the
-composer or philosopher tended to the first element of each
-characteristic pair, and vice-versa.
-
-\begin{table}[ht]
-  \caption{\label{tab:tableAmus} Quantification of eight music
-    characteristics for each of the seven composers.}
-  \begin{ruledtabular}
-    \begin{tabular}{|l|c|c|c|c|c|c|c|c|}
-      \footnotesize Composers    & \tiny  $S$-$S_c$ 
-      & \tiny  $D_s$-$D_l$ & \tiny  $H$-$C$ & \tiny  $V$-$I$ 
-      & \tiny  $D_n$-$D$ & \tiny  $M_s$-$M_v$ & \tiny  $R_s$-$R_c$
-      & \tiny  $H_s$-$H_v$  \\
-      \hline
-   \footnotesize Monteverdi  & 3.0 & 8.0 & 5.0 & 3.0 & 7.0 & 5.0 & 3.0 & 7.0 
-   \\
-   \footnotesize Bach        & 2.0 & 6.0 & 9.0 & 2.0 & 8.0 & 2.0 & 1.0 & 5.0 
-   \\
-   \footnotesize Mozart      & 6.0 & 4.0 & 1.0 & 6.0 & 6.0 & 7.0 & 2.0 & 2.0 
-   \\
-   \footnotesize Beethoven   & 7.0 & 8.0 & 2.5 & 8.0 & 5.0 & 4.0 & 4.0 & 7.0 
-   \\
-   % Chopin & 9.0 & 3.0 & 3.0 & 9.0 & 5.5 & 8.0 & 7.0 & 8.0 \\
-   \footnotesize Brahms      & 6.0 & 6.0 & 4.0 & 7.0 & 4.5 & 6.5 & 5.0 & 7.0 
-   \\
-   \footnotesize Stravinsky  & 8.0 & 7.0 & 6.0 & 7.0 & 8.0 & 5.0 & 8.0 & 5.0 
-   \\
-   \footnotesize Stockhausen & 7.0 & 4.0 & 8.0 & 7.0 & 5.0 & 8.0 & 9.0 & 6.0 
-   \\
-    \end{tabular}
-  \end{ruledtabular}
-\end{table}
-
-\begin{table}%\footnotesize%\scriptsize%\tiny
-  \caption{\label{tab:tableAphi} Quantification of eight philosophy
-    characteristics for each of the seven philosophers.}
-
-  \begin{ruledtabular}
-    \begin{tabular}{|l||c|c|c|c|c|c|c|c|}
-      Philosophers & R-E & E-E & M-D & T-A & H-R & D-P & D-F & N-M \\ \hline
-      Plato        & 3.0 & 3.5 & 9.0 & 5.0 & 4.5 & 3.5 & 5.0 & 4.5 \\
-      Aristotle    & 8.0 & 7.5 & 7.0 & 5.5 & 7.5 & 8.0 & 2.5 & 2.5 \\
-      Descartes    & 1.5 & 2.5 & 9.0 & 6.5 & 7.0 & 2.5 & 7.5 & 7.5 \\
-      Espinoza     & 8.0 & 2.0 & 1.0 & 5.0 & 2.0 & 3.0 & 1.0 & 1.0 \\
-      Kant         & 7.0 & 2.5 & 8.5 & 6.5 & 4.5 & 3.5 & 7.5 & 5.0 \\
-      Nietzsche    & 7.5 & 9.0 & 1.0 & 9.0 & 5.0 & 8.0 & 1.0 & 1.5 \\
-      Deleuze      & 5.5 & 7.5 & 1.0 & 8.0 & 2.5 & 5.5 & 5.0 & 6.0 \\
-    \end{tabular}
-  \end{ruledtabular}
-\end{table}
-
-This data set defines an 8-dimensional space for music or philosophy
-where each dimension corresponds to a characteristic that applies to
-all 7 composers or philosophers. Such small data sets are not adequate
-for statistical analysis, which could be biased. This is the reason
-why we used the bootstrap method for sampling described in
-section~\ref{bootstrap}.
-
-For the extended data set after applying the bootstrap method, Pearson
-correlation coefficients between the eight characteristics chosen are
-given in Table \ref{tab:tableBmus} for composers and in Table
-\ref{tab:tableBphi} for philosophers. The coefficient was 0.69 for the
-pairs $S$-$S_c$ (Sacred or Secular) and $V$-$I$ (Vocal or
-Instrumental), which indicates that sacred music tends to be more
-vocal than instrumental. The coefficient 0.56 for the pairs $S$-$S_c$
-and $R_s$-$R_c$ (Rhythmic Simplicity or Complexity) also shows that
-this genre does not commonly use polyrhythms. A negative coefficient
-of -0.33 for the pair $V$-$I$ and $D_n$-$D$ (Non-discursive or
-Discursive) indicated that composers who used just voices on their
-compositions also preferred programmatic music techniques such as
-baroque rhetoric. Strong correlations were also observed for
-philosophers. For instance, the Pearson coefficient of $-0.46$ for
-\emph{R-E} and \emph{N-M} suggests that rationalists tend to be also
-mechanists. An even stronger correlation of $0.74$, now positive, is
-observed between \emph{E-E} and \emph{D-P}, i.e. existentialists also
-tend to be phenomenologists, as could be expected. Other important
-correlations appeared between \emph{D-F} and \emph{N-M} (coefficient =
-$0.61$) and between \emph{M-D} and \emph{D-F}, which seems to be
-directly implied by religious background.
-
-\begin{table}[ht]
-  \caption{\label{tab:tableBmus}Pearson correlation coefficients
-    between the eight musical characteristics. The entries with
-    absolute values $\geq$ 0.30 have been highlighted.}
-  \begin{ruledtabular}
-    \begin{tabular}{|c||c|c|c|c|c|c|c|c|}
-      -    & \tiny  $S$-$S_c$ & \tiny  $D_s$-$D_l$ & \tiny  $H$-$C$ 
-      & \tiny  $V$-$I$ & \tiny  $D_n$-$D$ & \tiny  $M_s$-$M_v$ 
-      & \tiny  $R_s$-$R_c$ & \tiny  $H_s$-$H_v$  \\ \hline
-
-      \tiny$S$-$S_c$ 
-      & - & -0.2 & -0.06 & \bf{0.69} & -0.18 & 0.19  & \bf{0.56} & -0.16 
-      \\
-      \tiny$D_s$-$D_l$ 
-      & - & - & -0.14 & -0.13 & 0.2 & \bf{-0.48} & -0.2 & \bf{0.37}
-      \\
-      \tiny$H$-$C$ 
-      & - & - & - & -0.23 & 0.26 & 0.05 & \bf{0.46} & 0.03 
-      \\
-      \tiny$V$-$I$ 
-      & - & - & - & - & \bf{-0.33} & 0.17 & \bf{0.42} & -0.06 
-      \\
-      \tiny$D_n$-$D$ 
-      & - & - & - & - & - & -0.3 & 0.02 & -0.22 
-      \\
-      \tiny$M_s$-$M_v$ 
-      & - & - & - & - & - & - & 0.26 & -0.15
-      \\
-      \tiny$R_s$-$R_c$ 
-      & - & - & - & - & - & - & - & -0.02 
-      \\
-      \tiny$H_s$-$H_v$ 
-      & - & - & - & - & - & - & - & - 
-      \\
-    \end{tabular}
-  \end{ruledtabular}
-\end{table}
-
-\begin{table}\footnotesize
-  \caption{\label{tab:tableBphi}Pearson correlation coefficients
-    between the eight philosophical characteristics.  The entries with
-    absolute values $\geq$ 0.35 have been highlighted.}
-  \begin{ruledtabular}
-    \begin{tabular}{|c||c|c|c|c|c|c|c|c|}
-      - & R-E & E-E & M-D & T-A & H-R & D-P & D-F & N-M \\ \hline
-      R-E & - & \bf{0.37} & -0.23 & 0.15 & 0.1 & \bf{0.46} & -0.27 & \bf{-0.46} 
-      \\
-      E-E & - & - & \bf{-0.53} & 0.19 & 0.15 & \bf{0.74} & \bf{-0.61} & -0.3 
-      \\
-      M-D & - & - & - & \bf{-0.43} & \bf{0.41} & -0.3 & \bf{0.35} & 0.01 
-      \\
-      T-A & - & - & - & - & -0.21 & 0.06  & 0.19 & 0.26 
-      \\
-      H-R & - & - & - & - & - & 0.32 & -0.22 & -0.25 
-      \\ 
-      D-P & - & - & - & - & - & - & \bf{-0.63} & \bf{-0.47} 
-      \\
-      D-F & - & - & - & - & - & - & - & \bf{0.61}
-      \\
-      N-M & - & - & - & - & - & - & - & - 
-      \\
-    \end{tabular}
-  \end{ruledtabular}
-\end{table}
-
-The data were analyzed with principal component analysis (PCA), from
-which it was found that several of the characteristics contributed to
-the variability of the data, as shown in Tables~\ref{tab:Deviatesmus}
-and \ref{tab:Deviatesphi} in the Supporting Information. Figures
-\ref{fig:pcamus} and \ref{fig:pcaphi} display a 2-dimensional space
-with the first two main axes. The arrows follow the time sequence
-along with the seven composers and philosophers. Each of these arrows
-corresponds to a vectorial move from one composer or philosopher state
-to another. For clarity, just the lines of the arrows are
-preserved. The bootstrap samples define clusters around the original
-composers and philosophers. In subsidiary experiments, we verified
-that the results from the bootstrap method were robust. This was
-performed by applying 1000 perturbations of the original scores by
-adding to each score the values -2, -1, 0, 1 or 2 with uniform
-probability. In other words, we tested if scoring errors could be
-sufficient to cause relevant effects on the PCA
-projections. Interestingly, the values of average and standard
-deviation for both original and perturbed positions listed in Tables
-\ref{tab:tableDmus} and \ref{tab:tableDphi} in the Supporting
-Information show relatively small changes. It is therefore reasonable
-to assume that small errors in the scores assigned had no significant
-effect on the overall analysis.
-
-\begin{figure}[htbp]
-  \begin{center}
-    \includegraphics[width=0.45\textwidth]{g1}
-  \end{center}
-  \caption{\it 2-dimensional projected music space.}
-  \label{fig:pcamus}
-\end{figure}
-
-Bach is found far from the rest of the composers, which suggests his
-key role acknowledged by other great composers like Beethoven and
-Webern~\cite{Webern}: ``In fact Bach composed everything, concerned
-himself with everything that gives food for thought!''. The greatest
-subsequent change takes place from Bach to Mozart, reflecting a
-substantial difference in style. There is a strong relationship
-between Beethoven and Brahms, supporting the belief by the
-\textit{virtuosi} Hans von B\"{u}low~\cite{Bulow} when he stated that
-the $1^{st}$ Symphony of Brahms was, in reality, the \textit{$10^{th}$
-  Symphony of Beethoven}, appointing Brahms as the true successor of
-Beethoven. Stravinsky is near Beethoven and Brahms, presumably due to
-his heterogeneity~\cite{BennettHistory, Lovelock}. Beethoven is also
-near Mozart, who deeply influenced Beethoven, mainly in his early
-works. For Webern, Beethoven was the unique classicist who really came
-close to the coherence found in the pieces of the Burgundian School:
-``Not even in Haydn and Mozart do we see these two forms as clearly as
-in Beethoven. The period and the eight-bar sentence are at their
-purest in Beethoven; in his predecessors we find only traces of
-them''~\cite{Webern}. It could explain the move of Beethoven in
-direction of the Renaissance Monteverdi. Stockhausen is a deviating
-point when compared with the others, which could be more even so had
-we considered vanguard characteristics --- e.g.\ timbre exploration by
-using electronic devices~\cite{Lovelock} --- not shared by his
-precursors.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=0.45\textwidth]{pca_filosofos_novo}
-  \end{center}
-  \caption{\it 2-dimensional projected philosophy space.}
-  \label{fig:pcaphi}
-\end{figure}
-
-The opposition and skewness indices for each of the six moves among
-composers states in Table \ref{tab:tableOImus} indicate that the
-movements were driven by small opposition and strong skewness. In
-other words, most movements seem to seek innovation rather than
-opposition. Furthermore, the counter-dialectics indices in Table
-\ref{tab:tableEmus} are smaller than for the philosophers, as will be
-discussed later on.
-
-\begin{table}[ht]
-  \caption{\label{tab:tableOImus}Opposition and skewness indices for
-    each of the six composers states moves}
-  \begin{tabular}{|c||c|c|}
-    \hline
-    Musical Move & $W_{i,j}$ & $s_{i,j}$ \\
-    \hline \hline
-    Monteverdi $\to$ Bach             &   1.0     &  0.      \\
-    Bach $\to$ Mozart                 &   1.0196  &  1.9042  \\
-    Mozart $\to$ Beethoven            &   0.4991  &  2.8665  \\
-    Beethoven $\to$ Brahms            &   0.2669  &  1.7495  \\
-    Brahms $\to$ Stravinsky           &   0.4582  &  2.6844  \\
-    Stravinsky $\to$ Stockhausen      &   0.2516  &  3.1348  \\
-    \hline
-  \end{tabular}
-\end{table}
-
-\begin{table}[ht]
-  \caption{\label{tab:tableEmus} Counter-dialectics index for each of
-    the five subsequent pairs of moves among composers states for the
-    8 components.}
-  \begin{tabular}{|c||c|}
-    \hline
-    Musical Triple & $d_{i \rightarrow k}$ \\
-    \hline \hline
-    Monteverdi $\to$ Bach $\to$ Mozart          & 2.0586 \\
-    Bach $\to$ Mozart $\to$ Beethoven           & 1.2020 \\
-    Mozart $\to$ Beethoven $\to$ Brahms         & 1.0769 \\
-    Beethoven $\to$ Brahms $\to$ Stravinsky     & 0.2518 \\
-    Brahms $\to$ Stravinsky $\to$ Stockhausen   & 0.2549 \\
-    \hline
-  \end{tabular}
-\end{table}
-
-As for philosophy, Figure \ref{fig:pcaphi} shows an opposite movement
-from Plato to Aristotle, which confirmed the antagonistic view of
-Aristotle when compared with Plato, even though Aristotle was his
-disciple~\cite{Russel}. Opposition is present along all the moves
-among philosophers states. This oscillatory pattern makes it possible
-to identify two well defined groups. The first contained Aristotle,
-Espinoza, Nietzsche and Deleuze, in the left side of the graph. The
-other contained Plato, Descartes and Kant,in the right side.  This
-division is consistent with the points of view shared by each group
-member. Another two groups are identified in the y-axis, separating
-all the philosophers from Nietzsche and Deleuze, who are represented
-by the most distant points.
-
-When opposition and skewness indices in Table \ref{tab:tableOIphi} are
-considered, all the moves among philosophers states tend to take place
-according to a well-defined, intense opposition from the average
-state. This was already noticeable in the PCA analysis. An interesting
-relationship is the minor opposition and strong skewness between
-Nietzsche and Deleuze, suggesting the return to Nietzsche as noted in
-the works of Deleuze while considering the vanguard characteristics of
-the 20th century philosopher~\cite{Deleuze2}. Espinoza tended toward
-Nietzsche, and their similarity was admitted by Nietzsche, by naming
-Espinoza his direct precursor~\cite{Russel}. While similar to
-Espinoza, Nietzsche presented strong opposition to Kant, consistent
-with Nietzsche being the strongest objector to Kant
-ideas~\cite{Russel}. Also surprising was the rather small skewness
-among most of the moves among philosophers states, which would be
-driven almost exclusively by opposition to the current philosopher
-state.  The results for dialectics in Table \ref{tab:tableEphi} show a
-progressively stronger dialectics among subsequent pairs of moves in
-philosophers states.
-
-\begin{table}
-  \caption{\label{tab:tableOIphi}Opposition and skewness indices for
-    each of the six philosophers states moves.}
-  \begin{tabular}{|c||c|c|}
-    \hline
-    Philosophical Move & $W_{i,j}$ & $s_{i,j}$ \\
-    \hline \hline
-    Plato $\rightarrow$ Aristotle     & 1.0    & 0 \\
-    Aristotle $\rightarrow$ Descartes & 0.8740 & 1.1205 \\
-    Descartes $\rightarrow$ Espinoza  & 0.9137 & 2.3856 \\
-    Espinoza $\rightarrow$ Kant       & 0.6014 & 1.6842 \\
-    Kant $\rightarrow$ Nietzsche      & 1.1102 & 2.9716 \\
-    Nietzsche $\rightarrow$ Deleuze   & 0.3584 & 2.4890 \\
-    \hline
-  \end{tabular}
-\end{table}
-
-\begin{table}
-  \caption{\label{tab:tableEphi}Counter-dialectics index for each of
-    the five subsequent pairs of philosophers states moves.}
-  \begin{tabular}{|c||c|}
-    \hline
-    Philosophical Triple & $d_{i \rightarrow k}$ \\
-    \hline \hline
-    Plato $\rightarrow$ Aristotle $\rightarrow$ Descartes    & 3.0198 \\
-    Aristotle $\rightarrow$ Descartes $\rightarrow$ Espinoza & 1.8916 \\
-    Descartes $\rightarrow$ Espinoza $\rightarrow$ Kant      & 1.1536 \\
-    Espinoza $\rightarrow$ Kant $\rightarrow$ Nietzsche      & 1.1530 \\
-    Kant $\rightarrow$ Nietzsche $\rightarrow$ Deleuze       & 0.2705 \\
-    \hline
-  \end{tabular}
-\end{table}
-
-The analysis above indicates distinct characteristics for the
-evolution of classical music and philosophy. Philosophers seem to have
-developed their ideas driven by opposition ($W_{i,j}$), as shown in
-Table \ref{tab:tableOIphi}, while composers tend to be more influenced
-by their predecessors according to the dialectics measurements
-($1/d_{i \rightarrow k}$). In general, the movements among composers
-had minor opposition, thus reflecting the master-apprentice
-tradition. There is then a crucial difference in the \textit{memory
-  treatment} along the development of philosophy and music: using the
-same techniques, we verified that a philosopher was influenced by
-opposition of ideas from his direct predecessor, while composers were
-influenced by their two predecessors. We can argue that philosophy
-exhibits a \textit{memory-1} state, while music presents
-\textit{memory-2}, with \textit{memory-N} having $N$ of past
-generations that influenced a philosopher or composer. Furthermore,
-Figure \ref{fig:comparingdialectics} shows a constant decrease in the
-counter-dialectics index, which means a constant return to the origins
-for the development of music based on the search for unity. Using the
-words of Webern, the search for the ``comprehensibility'' but always
-influenced by their old masters.
-
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics[width=0.45\textwidth]{compara_dialeticas2}
-  \end{center}
-  \caption{\it Comparison between composers and philosophers
-    counter-dialectics indices}
-  \label{fig:comparingdialectics}
-\end{figure}
-
-The comparison between composers and philosophers is complemented with
-the Wards hierarchical clustering~\cite{Ward}, which clusters the
-original scores taking into account their distance. The generated
-dendrogram in Figure \ref{fig:dendrogrammus} shows composers according
-to their similarity, while the corresponding dendrogram for
-philosophers is shown in Figure \ref{fig:dendrogramphi}. These
-dendrograms are consistent with the previous observations. For
-example, Beethoven and Brahms are close, reflecting their
-heritage. Stravinsky and Stockhausen form another cluster, while
-Mozart appears on its own, like Bach and Monteverdi. Both relations
-were also present in the planar space in Figure \ref{fig:pcamus}.
-
-\begin{figure}[ht]
-        \begin{center}
-          \includegraphics[width=0.45\textwidth]{Clust_Compositores.eps}
-        \end{center}
-        \caption{\it Wards hierarchical clustering of the seven composers.}
-        \label{fig:dendrogrammus}
-\end{figure}
-
-\begin{figure}
-        \begin{center}
-                \includegraphics[width=0.5\textwidth]{Dendrogram_.eps}
-        \end{center}
-        \caption{\it Wards hierarchical clustering of the seven
-          philosophers considering all the eight features.}
-        \label{fig:dendrogramphi}
-\end{figure}
-
-\section{Concluding Remarks}
-
-An approach has been proposed which allows for a quantitative
-assessment of features from any given subject or area. For classical
-music and philosophy investigated here, a time line could be
-established based on the characteristics of composers and
-philosophers. The quantitative analysis involved the collection of
-data for these characteristics, extended with a bootstrap method to
-yield robustness to the statistical analysis, and the use of
-multivariate statistics. Also important was the establishment of
-indices for opposition, skewness and counter-dialectics. Though the
-emphasis of our work has been on the method of analysis, the
-interpretation of the data was already sufficient to draw conclusions
-that confirm intuition and subjective analyses of classical music and
-philosophy.
-
-For instance, the results pointed to the development in music
-following a dichotomy: while composers aim at innovation, creating
-their own styles, their technique is based on the work of their
-predecessors, in a master-apprentice tradition. Indeed, in the history
-of music, composers developed their own styles with a continuous
-search for coherence or unity. In the words of Anton
-Webern~\cite{Webern}, ``[...] ever since music has been written most
-great artists have striven to make this unity ever clearer. Everything
-that has happened aims at this [...]''. A frequent inheritance of
-style can be identified from one composer to another as a gradual
-development from its predecessor, contrasting with the necessity for
-innovation. Quoting Lovelock: ``[...] by experiment that progress is
-possible; it is the man with the forward-looking type of mind [...]
-who forces man out of the rut of `what was good enough for my father
-is good enough for me'.''~\cite{Lovelock}. In contrast, our
-quantitative analysis confirmed that philosophy appears to exhibit a
-well-defined trend in innovation: unlike music, the quest for
-difference seems to drive philosophical changes as expressed by Gilles
-Deleuze~\cite{Deleuze}. According to Ferdinand de Saussurre's
-principle~\cite{Saussure}, concepts (words) tend to be different in
-the sense of meaning distinct things.  The paradigm of difference is
-particularly important because it is related to the own dynamics of
-philosophical evolution.
-
-Again emphasizing that our aim was to propose a generic, quantitative
-method, we highlight some limitations of the specific analysis made
-here for music and philosophy. For the scores and choice of main
-characteristics in music and philosophy were largely arbitrary and
-could be disputed. Nevertheless, the perturbation analysis performed
-in this work suggests that the effect of non-systematic errors in
-assigning the scores does not seem to be critical and has little
-overall impact on the conclusions drawn. Most importantly, the formal
-quantitative methodology described here may be combined with other
-methods, including information retrieval and natural language
-processing, to investigate other issues in humanities. For example,
-still connected with the present work, it can be adapted to the
-investigation of musical and philosophical schools, individual pieces
-(e.g.\ music suites or books), or even the contributions from the same
-composers or philosophers along distinct periods of time. Obviously,
-this methodology can also be applied to other areas such as poetry,
-cinema and science.
-
-\begin{acknowledgments}
-Luciano da F. Costa thanks CNPq (308231/03-1) and FAPESP (05/00587-5)
-for sponsorship. Gonzalo Travieso thanks CNPq (308118/2010-3) for
-sponsorship. Vilson Vieira is grateful to CAPES.
-\end{acknowledgments}
-
-\appendix
-
-\section{A Brief Explanation of Principal Component Analysis (PCA)}
-
-PCA is a dimensionality reduction procedure performed through rotation
-of axes.  It operates by concentrating dispersion/variance along the
-first new axes, which are referred to as the principal components. The
-technique consists in finding the eigenvectors and eigenvalues of the
-covariance matrix of the corresponding random vectors (i.e.\ the
-vectors associated with each philosophical state). The eigenvalues
-correspond to the variances of the new variables.  When multiplied by
-the original feature matrix, the eigenvectors yield the new random
-variables which are fully uncorrelated. For a more extensive
-explanation of PCA, please refer to~\cite{Costa} and references
-therein.
-
-\section{Supporting Information}
-
-Tables \ref{tab:Deviatesmus} and \ref{tab:Deviatesphi} show the
-normalized weights of the contributions of each original property on
-the eight axes for composers and philosophers. Most of the
-characteristics contribute almost equally in defining the axes.
-
-\begin{table}[ht]
-  \caption{\label{tab:Deviatesmus}Percentages of
-    the contributions from each musical characteristic on the eight
-    new main axes.}
-  \begin{tabular}{|c||c|c|c|c|c|c|c|c|}
-    \hline
-    Musical         & \multirow{2}{*}{$C_1$} & \multirow{2}{*}{$C_2$} 
-    & \multirow{2}{*}{$C_3$} & \multirow{2}{*}{$C_4$} & \multirow{2}{*}{$C_5$}
-    & \multirow{2}{*}{$C_6$} & \multirow{2}{*}{$C_7$} & \multirow{2}{*}{$C_8$}
-    \\
-    Charac. & & & & & & & & \\
-    \hline
-    $S$-$S_c$   
-    &  19.78  &   4.04  & 10.38 & 10.60 &  17.55  &  36.60  &  4.41 &  0.63 \\
-    $D_s$-$D_l$            
-    &  13.63  &   9.21  & 19.17 &  3.55 &   3.13  &   1.65  & 25.55 & 24.05 \\
-    $H$-$C$                
-    &   1.44  &  26.62  & 8.26 & 13.97 &  21.71  &   7.76  & 13.98 & 12.20 \\
-    $V$-$I$                
-    &  18.35  &  12.82  & 9.29 &  8.02 &   9.37  &  40.95  &  2.12 &  2.03 \\
-    $D_n$-$D$              
-    &   6.31  &  10.73  & 15.48 & 26.29 &  4.04  &   1.86  & 25.29 &  2.35 \\
-    $M_s$-$M_v$            
-    &  16.94  &  13.28  & 15.03 &  4.84 &  32.25  &  1.70  &  2.62 &  4.37 \\
-    $R_s$-$R_c$            
-    &  14.13  &   3.26  & 15.58 & 13.80 &   7.48  &  1.88  &  1.36 & 35.99 \\
-    $H_s$-$H_v$            
-    &   9.38  &  20.00  &  6.75 & 18.88 &   4.45  &  7.56  & 24.62 & 18.36 \\
-    \hline
-  \end{tabular}
-\end{table}
-
-\begin{table}
-  \caption{\label{tab:Deviatesphi}Percentages of
-    the contributions from each philosophical characteristic to the four 
-    new main axes.}
-  \begin{tabular}{|c||c|c|c|c|c|c|c|c|}
-    \hline
-    Philos.      & \multirow{2}{*}{$C_1$} & \multirow{2}{*}{$C_2$} 
-    & \multirow{2}{*}{$C_3$} & \multirow{2}{*}{$C_4$} & \multirow{2}{*}{$C_5$} 
-    & \multirow{2}{*}{$C_6$} & \multirow{2}{*}{$C_7$} & \multirow{2}{*}{$C_8$}
-    \\
-    Charac. & & & & & & & & \\
-    \hline
-    R-E & 13.35   &  1.81 & 13.11 & 31.12 & 15.06 & 10.99 &  8.20 &  3.56 \\
-    E-E & 18.13   &  7.88 &  2.10 & 14.68 & 11.41 &  5.31 &  7.98 & 33.50 \\
-    M-D & 10.07   & 24.59 & 10.71 &  2.91 &  0.69 & 19.39 & 20.95 &  2.91 \\
-    T-A &  1.08   & 24.18 & 21.51 &  0.68 & 23.72 &  6.31 &  7,27 &  2.06 \\
-    H-R &  5.90   & 21.49 & 22.29 & 14.90 &  6.35 & 18.81 &  8.69 &  2.05 \\
-    D-P & 18.99   &  1.29 &  6.85 &  8.41 &  9.67 & 20.50 & 11.47 & 24.93 \\
-    D-F & 14.57   & 13.67 &  8.37 & 18.93 & 21.12 &  9.06 & 12.77 & 14.98 \\
-    N-M & 17.86   &  5.06 & 15.03 &  8.33 & 11.96 &  9.59 & 22.62 & 15.97 \\
-    \hline
-  \end{tabular}
-\end{table}
-
-\begin{table}
-  \caption{\label{tab:tableDmus}Averages and standard deviations
-    for each composer and for the 8 eigenvalues.}
-
-  \begin{tabular}{|c||c|c|}
-    \hline
-    Composers & $\mu_{\Delta}$ & $\sigma_{\Delta}$ \\
-    \hline
-    Monteverdi     & 3.7347 & 0.8503 \\
-    Bach           & 5.3561 & 0.9379 \\
-    Mozart         & 4.4319 & 0.8911 \\
-    Beethoven      & 3.4987 & 0.7851 \\
-    Brahms         & 3.0449 & 0.6996 \\
-    Stravinsky     & 3.6339 & 0.7960 \\
-    Stockhausen    & 4.2143 & 0.9029 \\
-    \hline \hline
-    Eigenvalues & $\mu_{\Delta}$ & $\sigma_{\Delta}$ \\
-    \hline
-    $\lambda_1$ &  -0.1759 & 0.0045 \\
-    $\lambda_2$ &  -0.0638 & 0.0026 \\
-    $\lambda_3$ &  -0.0411 & 0.0021 \\
-    $\lambda_4$ &  -0.0144 & 0.0019 \\
-    $\lambda_5$ &   0.0578 & 0.0021 \\
-    $\lambda_6$ &   0.0736 & 0.0023 \\
-    $\lambda_7$ &   0.0080 & 0.0027 \\
-    $\lambda_8$ &   0.0835 & 0.0030 \\
-    \hline
-  \end{tabular}
-\end{table}
-
-\begin{table}
-  \caption{\label{tab:tableDphi}Average and standard deviation for
-    each philosopher and for the 8 eigenvalues.  }
-  \begin{tabular}{|c||c|c|}
-    \hline
-    Philosophers & $\mu_{\Delta}$ & $\sigma_{\Delta}$ \\
-    \hline
-    Plato          & 3.3263 & 0.7673 \\
-    Aristotle      & 4.0896 & 0.8930 \\
-    Descartes      & 4.3081 & 0.9225 \\
-    Espinoza       & 4.9709 & 0.9131 \\
-    Kant           & 3.2845 & 0.7749 \\
-    Nietzsche      & 5.3195 & 0.9797 \\
-    Deleuze        & 4.0990 & 0.8970 \\
-    \hline \hline
-    Eigenvalues & $\mu_{\Delta}$ & $\sigma_{\Delta}$ \\
-    \hline
-    $\lambda_1$ & -0.2618 & 0.0068 \\
-    $\lambda_2$ & -0.0976 & 0.0035 \\
-    $\lambda_3$ &  0.0154 & 0.0025 \\
-    $\lambda_4$ &  0.0212 & 0.0024 \\
-    $\lambda_5$ &  0.0697 & 0.0026 \\
-    $\lambda_6$ &  0.0807 & 0.0030 \\
-    $\lambda_7$ &  0.0877 & 0.0032 \\
-    $\lambda_8$ &  0.0846 & 0.0036 \\
-    \hline
-  \end{tabular}
-\end{table}
+\begin{equation}\label{senoide}
+     S^f_i=\{ s^f_i \}=\Bigl\{ \sin\bigl(2\pi \frac{i}{\lambda_f} \bigr)  \Bigr\} = \Bigl\{ \sin\bigl(2\pi f \frac{i}{f_a}\bigr)  \Bigr\} 
+\end{equation}
 

### 2013-04-22 devEresultados.tex
artigo escrito ateh parte dos timbres

@@
   Vale salientar que nem tudo
-o que se atribui ao timbre se acha manifesto em diferenças espectrais e que até
+o que se atribui ao timbre se acha manifesto em traços espectrais e que até
 aspectos culturais ou circunstanciais alteram nossa percepção do timbre.

### 2013-04-23 article/phi+mus.tex
feita passagem para artigo ateh o espectro no som amostrado

@@
 
+Where $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\lambda_a}$  is the number of samples in the period.
+
+In a similar fashion, other waveforms are used in music because of its spectral qualities and simplicity.
+While the sinusoid is an isolated point in the spectrum, these waves present a succession of harmonic components.
+These waveforms are specified in equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} are in figure~\ref{fig:formasDeOnda}.
+These artificial waveforms are traditionally used in music for synthesis and oscilatory control of variables, they also present diverse used outside music\cite{Openheim}.
+
+The sawtooth present all components of the harmonic series, with decreasing energy of $-6dB/octave$. The sequence of temporal samples can be described like this:
+
+\begin{equation}\label{sawTooth}
+     D^f_i=\left\{ d^f_i \right\}=\left\{ 2\frac{i\,\%\lambda_f}{\lambda_f} -1 \right\}
+\end{equation}
+
+The triangular waveform present only odd harminics falling with $-12dB/octave$:
+\begin{equation}\label{triangular}
+     T^f_i=\left\{ t^f_i \right\}=\left\{1- \left| 2 - 4\frac{i\,\%\lambda_f}{\lambda_f} \right| \right\}
+\end{equation}
+
+The square wave presets only odd harmonics falling at $-6dB/octave$:
+
+\begin{equation}\label{quadrada}
+     Q^f_i=\left\{ q^f_i \right\}= \left\{
+         \begin{array}{l l}
+              1 & \quad \text{for } \; \; (i\,\%\lambda_f)   <  \lambda_f /2  \\
+              -1 & \quad \text{otherwise}\\
+         \end{array} \right.
+\end{equation}
+
+The sawtooth is a common starting point for a subtractive synthesis, beacuse it has both odd and even harmonics with high energy.
+For musical intention, these waveforms are excessively rich in sharp harmonics and atenuant filtering on treble and midle parts of the spectrum is useful for reaching a more natural and pleasant sound.
+The relativelly attenuated harmonics of the triangle wave makes it the more functional - among the listed ones - to be used in the synthesis of musical notes without any treatment.
+
+The square wave can be used in a subtractive synthesis the aims to mimic a clarinet.
+This instrument has only the odd harmonic components and the square wave is convenient with its abundant energy in high frequencies.
+
+\begin{figure*}
+    \centering
+        \includegraphics[width=\textwidth]{figures/waveForms}
+    \caption{Basic musical waveforms. The synthetic  waveforms are in (a) and the real waveforms are in (b).}
+        \label{fig:formasDeOnda}
+\end{figure*}
+
+Figure~\ref{fig:formasDeOnda} presents the waveforms described in equations  ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} and ~\ref{quadrada} for $\lambda_f=100$ (period of $100$ samples).
+If $f_s=44.1kHz$, as PCM standard in Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz$, an A4, just above the central "C", whatever the waveform is.
+
+The spectrum of each basic waveform is in figure~\ref{fig:espectroDeondas}.
+The isolated and exaclty harmonic components of the spectrum is a consequence of the fixed period usage.
+The sinusoid consists of a one and only node in spectrum, pure frequency.
+The sawtooh is the only with a complete harmonic series (odd and even components).
+Triangular and square waves has the same components (odd harmonics), decaing at $-12dB/octave$ and $-6dB/octave$ respectivelly.
+
+\begin{figure*}
+    \centering
+        \includegraphics[width=\textwidth]{figures/waveSpectrum}
+    \caption{Spectrum of basic artificial musical waveforms.}
+        \label{fig:espectroDeOndas}
+\end{figure*}
+
+The harmonic spectrum is formed by frequencies multiple from the fundamental frequency $f_n=(n+1)f_0$.
+As the human linear perception of pitch follows the a geometric progression of frequencies, the spectrum has notes different from the fundamental frequency.
+Additionally, the number of harmonics will be limited to the maximum frequency $f_s/2$ (by Nyquist's theorem).
+
+Musically crucial here is to internalize that energy in an component of frequency $f_n$ means an oscilation in the constitution of the sound, purely harmonic and in that frequency $f_n$.
+This energy, specificaly concentrated on the $f_n$ is separated by the ear to enter a cognitive level of processing (this separation is done in many species is mechanisms similar to the human cochlea)~\cite{Roederer}.
+The sinusoidal components are usually the main responsibles for the quality we call timbre.
+If they are not presented in harmonic proportions (small number relations), the sound is perceived as noisy or dissonant and not as a sonority with a univocally stablished fundamental frequency.
+Furthermore, the notion of absolute pitch in a complex sonority is based relies on the similarity of the spectrum to the harmonic series~\cite{Roederer}.
+
+In the case of a fixed length period and waveform, the spectrum is perfectly harmonic and static.
+Each waveform is compound of specific proportions of harmonic components and the greater the curvature of a the part, the greater the contribution of the part to the energy on the high harmonics.
+That can be seen on real sounds.
+The wave rotulated as ``sampled real sound'' in figure~\ref{fig:formasDeOnda} is a period of $\Lambda_f=114$ samples, extracted from a relatively well behaved real sound.
+The oboe wave was sampled of a A4 also in $44.1kHz$.
+The chose period for sampling was a relatively short, with $98$ samples, corresonds to the frequency $\frac{44100}{98}=450Hz$.
+It can be noticed, from the curvatures, the oboe's rich spectrum in high frequencies and the lower spectrum of the real sound.
+
+The sequence 
+$ R_i=\{ r_i \}_0^{\lambda_f-1}$ of samples in the real sound of figure~\ref{fig:formasDeOnda} can be taken as a basis for a sound $T_i^f$ din the following way:
+
+\begin{equation}\label{sampleandoFormaDeOnda}
+     T^f_i=\{ t_i^f \}=\Bigl\{ r_{(i\,\%\lambda_{f})} \Bigr\}
+\end{equation}
+
+The resulting sound has a the momentary spectrum of the original sound.
+Because it is repeated in an identical form, the spectrum is perfectly harmonic, withoud noise and variations tipical of the natural phenomenon.
+This can be seen in figure~\ref{fig:espectroOboe}, that show the spectrum of the original oboe note and a note with same duration and whose samples consists of the repetition of cycle of figure~\ref{fig:formasDeOnda}.
+The natural spectrum exibit variations in the frequencies of the harmonics, its intensities and some noise.
+The note made from the sampled period has a perfecly harmonic spectrum
+
+\begin{figure*}
+    \centering
+        \includegraphics[width=\textwidth]{figures/oboeNaturalSampledSpectrum}
+    \caption{Spectrum of the sonic waves of a natural oboe note and one made from a sampled period.
+The natural sound has fluctuations in the harmonics and in noise, while the sampled period note has a perfectly harmonic spectrum.}
+        \label{fig:espectroOboe}
+\end{figure*}
+
 \nocite{*}

### 2013-04-23 devEresultados.tex
feita passagem para artigo ateh o espectro no som amostrado

@@
 de $100$ amostras).
-Se $t_a=44,1 kHz$, como no padrão PCM de \emph{Compact Disks}, a onda possui frequência fundamental $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz $.
+Se $f_a=44,1 kHz$, como no padrão PCM de \emph{Compact Disks}, a onda possui frequência fundamental $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz $.
 Um lá\footnote{Um lá 4, logo acima do dó central, no segundo espaço do pentagrama na clave de sol comum.} seja qual for a forma de onda.

### 2013-09-19 article/article.tex
fazendo o artigo, ainda na intro

% notas:
% campo harmônico => harmonic field?
% anti-relative => counter-relative? (discussão sobre harmonia funcional)
%    associada relative => associated relative? (ver a discussão para mais...)
% medianas cromáticas => cromatic mediant
% tonalidade => key
% homônica => homonymous

\documentclass[
 aip,
 jmp,
 amsmath,amssymb,
 reprint,
]{revtex4-1}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{multirow}
\usepackage{color}
\usepackage{xcolor}
\usepackage{enumerate}
\usepackage{float}

\newcommand{\massa}{{\large \textsc{massa}}}
\newcommand{\mass}{{\large \textsc{mass}}}
\newcommand{\figgus}{{\large \textsc{figgus}}}

\floatstyle{plain}
\newfloat{twocolequfloat}{t}{zzz}
\floatname{twocolequfloat}{Equation}

\begin{document}
\preprint{AIP/123-QED}

\title{Psychophysics of musical elements in the discrete-time representation of sound}

\author{Renato Fabbri}
 \homepage{http://www.ifsc.usp.br/~fabbri}
 \email{renato.fabbri@gmail.com}
  \affiliation{ 
Instituto de F\'isica de S\~ao Carlos, Universidade de S\~ao Paulo (IFSC/USP)
}

\author{Vilson Vieira da Silva Jr.}
  \homepage{http://automata.cc/}
  \email{vilson@void.cc}
 \altaffiliation[Also at ]{IFSC-USP}

\author{D\'ebora Corr\^ea}
  \homepage{}
  \email{debcris.cor@gmail.com}
 \altaffiliation[Instituto de Ci\^encias Matem\'aticas e de Computa\c{c}\~ao ]{ICMC-USP}

\author{Luciano da Fontoura Costa}
  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
  \email{ldfcosta@gmail.com}
 \altaffiliation[Also at ]{IFSC-USP}

 \author{Osvaldo N. Oliveira Jr.}
  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
  \email{chu@ifsc.usp.br}
 \altaffiliation[Also at ]{IFSC-USP}

\date{\today}
\begin{abstract}

    %The representation of the basic elements of music - such as notes, ornaments and interval structures - in terms of discrete audio signal is often used in software for music creation and design. Nevertheless, there is no unified approach that relates these elements to the sound discrete samples. In this article, each musical element is described by equations in sonic time samples, which are all implemented in scripts within an open source software toolbox, referred to as \massa\ (Music and Audio in Sequences and Samples). The fundamental element, the musical note with duration, volume, pitch and timbre, is related quantitatively to the characteristics of the discrete-time signal. Internal variations, such as tremolos, vibratos and spectral fluctuations, are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities. With this representation of notes, resources are provided for the generation of musical structures, such as rhythmic meter, pitch intervals and cycles. The efficacy of these physical descriptions of basic musical elements was confirmed by the synthesis of small musical pieces within each frame: basic notes, incremented notes and notes in music. It is possible to synthesize whole albums through collage of the scripts and parametrization. 
    %The sample-based analytical description, and the paradigm of open source implementation, enables scientific experiments in precise and trustful ways. In fact, \massa\ has already been employed by external users for diverse purposes. Among these, it is mentioned an acoustic effect recognized by diverse individuals in mailing lists but not found in literature, and uses related to art and education.

Notes, ornaments and intervals are examples of basic elements of music,  their representation as discrete-time digital audio signal plays a central role in software for music creation and design. Nevertheless, there is no systematic relation, in analytical terms, of these musical elements to the sound samples. Such a compendium is important since it enables scientific experiments in precise and trustful ways. In this paper, a comprehensive description of the musical elements within a unified approach is presented. Musical elements, like pitch, duration and timbre are expressed by equations on sample level. This quantitatively relates characteristics of the discrete-time signal to musical qualities. Internal variations, e.g., tremolos, vibratos and spectral fluctuations, are also considered as means to achieve variation within a note. Moreover, the generation of musical structures such as rhythmic meter, pitch intervals and cycles, are reached canonically with notes. The availability of these resources in scripts  (\massa\ - Music and Audio in Sequences and Samples) is also provided in public domain. Authors observe that the implementation of sample-domain analytical results as open source can encourage concise research.
As further illustrated in the paper, \massa\ has already been employed by users for diverse purposes, including acoustics experimentation, art and education. The efficacy of these physical descriptions of basic musical elements was confirmed by the synthesis of small musical pieces.
As shown, it is possible to synthesize whole albums through collage of scripts and parametrization.

\end{abstract}
\pacs{*43.66.-x,43.66.+y,05.10.-a} % PACS, the Physics and Astronomy
\keywords{psychophysics, acoustics, statistics, signal processing, digital audio, music}
\maketitle

\section{\label{sec:level1}Introduction}

Music is commonly defined as the art made by sounds and silences. Sound is the longitudinal wave of mechanical pressure. The frequency bandwidth between $20Hz$ and $20kHz$ is appreciated by the human hearing system with boundaries dependent on the person, climate conditions and sonic characteristics themselves~\cite{Roederer}. Considered the speed of sound as $\approx 343.2 m/s$, these limits corresponds to wavelengths of $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.

Human perception of sound involves captivation by bones, stomach, ears, transfer functions of head and processing torso and nervous system~\cite{}. Besides that, the ear is a dedicated organ responsible for the capture of these waves. Its mechanism decomposes sound into its sinusoidal spectrum and delivers them to the nervous system. The sinusoidal components are crucial to musical phenomena, as one can perceive the composition of sounds according to a specific musical interest, e.g., tunings and scales.

The representation of sound is commonly called audio (although these terms are often used without distinction). Audio expresses the waveform prominent from the capture of microphones or from synthesis. In order to account for the corresponding digital representation, digital audio is specified by protocols that eases file storage and transferring, usually with low loss of quality. Standard representation of digital audio consists of equally spacing samples by $\lambda_s$ durations in time, where each sample is specified by a sample number of bits. This process is called the Pulse Code Modulation (PCM) representation of sound. A PCM digital sound is characterized by it's sampling frequency $f_s=\frac{1}{\lambda_s}$ (also called the sampling rate), and by a bit depth, which is the number of bits used for representing the amplitude of each sample. With the purposes of illustrating this idea, Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with $4$ bits each. The $2^4=16$ possible steps for each sample, together with the regular spacing $\lambda_s$ between them, establish what is named as quantization error. The noise prominent of these errors diminishes conforming the spacing also does.

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/pcm}
        \caption{Pulse Code Modulation (PCM) audio: an analogical signal is represented by 25 samples with 4 bits each.}
        \label{fig:PCM}
\end{figure*}

The Nyquist theorem~\cite{} says that half of the sampling frequency is the maximum frequency of the signal. Thus, it is necessary to have a sampling frequency at least twice the highest frequency heard by humans, that is, $f_s \geq 2\times 20kHz = 40kHz$. This configures the basis for the use of sampling frequencies like $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and Broadcast systems (Radio and TV), respectively.

For the average listener -- and a reasonable part of specialists -- the notion of music presupposes rhythmic and pitch organization. I a simple  way, rhythm can be described as recurring pattern produced by notes differing in duration, pause (silence) and intensity. Pitch configures the perceptual attribute that allows sounds to be organized in a frequency-related scale~\cite{}. Music from the twentieth century enlarged this traditional comprehension of music. This occurred in the concert music, specially in the concrete, and in both electronic and electroacoustic styles. On the last decade of the century, it became evident that popular music had also incorporated sounds without defined pitch and temporal organization, to name a few of simple metrics. Even though, the note still stands paradigmatic as a 'fundamental unit' of musical structures and, in practice, it can unfold in sounds that observe this recent developments (**not clear here**).

\subsection{Contributions and paper organization}

This paper purposes to represent musical structures and artifices by their discrete sound characteristics. The results include mathematical relations and their computer program implementations. Despite the general interest and number of knowledge areas involved, there are only a few number of books and computer implementations that tackle the subjects presented in this work. However, they mainly focus on aspects of musical sounds and ways to mimic traditional instruments. A detailed compilation of them is pointed out in the bibliography~\cite{dissertacao}. To the best of the author's knowledge, there is a lack of articles on the topic. Moreover, although current computer implementations use the analytical descriptions presented in this study in a implicitly manner, it seems that there is not a concise and mathematical description of the processes implemented, as they aim here to be further used as libraries for sound and music.

In order to address the concise description of musical elements and structures, the objectives of this paper are:

\begin{enumerate}

\item Present a concise set of relations among musical basic elements and sequences of PCM audio samples.

\item Introduce a framework of sound synthesis with control at sample level and with potential uses in psycho-acoustical experiments and high-fidelity synthesis. The didactic presentation of the content favors its apprehension and usage. It is worthwhile to mention that this subject comprises diverse topics on signal processing, music and psycho-acoustics.

\item Provide the accessibility of the developed framework. The topics presented in this article are implemented as scripts, i.e. small computer programs using accessible technologies for better distribution and validation. This constitute the \massa\ toolbox, available in public domain in an open Git repository~\cite{gitBook}. The scripts are written in Python and make use of external libraries such as Numpy and Scikits/Audiolab, that performs calls to Fortran routines for better computational efficiency. Part of the scripts have been transcribed to JavaScript and native Python with readiness, what favors their use in Web browsers such as Firefox and Chromium~\cite{numpy, audiolab, tutpython, python}. Furthermore, these are all open technologies, that is, published using licenses that allows copy, distribution and use of any part for research and derivatives. This way, the work presented here is available and eases co-authorship processes~\cite{Raymond,Lessig}.

\end{enumerate}

The remaining parts of this work are organized as follows: section~\ref{sec:dicNote} exposes the presence of sinusoid in discrete-time audio and characterizes a basic musical note; the definition and expansion of the musical note as the essential unit of music is further approached in sections~\ref{sec:discNote} and~\ref{sec:internalVar}, respectively; section~\ref{sec:notesMusic} tackles the organization of musical notes in a higher level~\cite{Wisnick,Webern,Lerdhal,Cook,Lacerda}. Musical theory embody diverse topics as psycho-acoustics, cultural manifestations and formalisms. The section~\ref{sec:results} points out these topics as needed and designate external complements~\cite{Zamacois,Schoenberg,microsound}.

The next section is a minimum text in which music elements are presented side-by-side with the discrete-time samples they result. In order to account for validation and sharing, implementations on computer code of these relations and little musical pieces resulting from them are gathered in the \massa\ toolbox, online available.

%Representing musical structures and artifices by it's discrete sound characteristics is the purpose of this work. The results include mathematical relations and it's computer program implementations. Next section exposes the theoretical description, which is implemented as scripts in a one-to-one relation to the equations.

%\subsection{Sound and digital audio}

%Sound is a longitudinal wave of mechanical pressure. The frequency bandwidth between $20Hz$ and $20kHz$ is appreciated by human hearing system with boundaries dependent on the person, climate conditions and sonic characteristics itself~\cite{Roederer}. If considered the speed of sound of $\approx 343.2 m/s$, this limits corresponds to $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.

%Human perception of sound involves captivation by bones, stomach, ears, transfer functions of head and processing torso and nervous system. Besides that, the ear is a dedicated organ to the capture of this waves. Its mechanism decomposes sound into its sinusoidal spectrum and delivers them to the nervous system. This sinusoidal components are crucial to musical phenomena, as one can observe in the composition of sounds with musical interest and in tunings and scales. Subsection~\ref{subsec:dicNote} exposes the presence of sinusoid in discrete-time audio and characterizes a basic musical note.

%The representation of sound is called audio (although these terms are often used without distinction), and this can be provenient from caption by microphones or from synthesis. Often enough, digital audio is specified by protocols that eases file storage and transferring, in cost of a direct representation or even some loss in quality. Standard representation of digital audio, on the other hand, consists of samples equally spaced by $\lambda_s$ durations in time, with each sample specified by a sample number of bits. This is called the Pulse Code Modulation representation of sound (PCM). A PCM digital sound is characterized by it's sampling frequency $f_s=\frac{1}{\lambda_s}$, also called sampling rate, and bit depth, which is the number of bits used of representing the amplitude of each sample. Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with $4$ bits each. The $2^4=16$ possible steps for each sample, together with the regular spacing $\lambda_s$ between them, introduces a quantization error. This noise, caused by this errors, diminishes as these spacing diminishes.

%\begin{figure*}
%    \centering
%        \includegraphics[width=\textwidth]{pcm}
%        \caption{Pulse Code Modulation (PCM) audio: an analogical signal is %represented by 25 samples with 4 bits each.}
%        \label{fig:PCM}
%\end{figure*}

%By the Nyquist theorem, it is known that half the sampling frequency is the maximum frequency of the signal. Thus, it is necessary to have a sampling frequency at least twice the highest frequency heard by humans $f_s \geq 2\times 20kHz = 40kHz$. This is the basis for the use of the sampling frequencies $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and Broadcast systems (Radio and TV), respectively.

%\subsection{Sonic art and musical theory}

%A common definition for music is the art made by sounds and silences. For the average listener -- and a reasonable part of specialists -- the notion of music presupposes rhythmic and pitch organization such as explained in subsection~\ref{subsec:notesMusic}. Music from the twentieth century enlarged this traditional comprehension of music. This occurred in concert music, specially in the concrete, electronic and electroacoustic styles. On the last decade of the century, it was evident that popular music has also incorporated sounds without defined pitch and temporal organization out of simple metrics. Even though, the note stands paradigmatic as a 'fundamental unit' of musical structures and, in practice, it can unfold in sounds that observe this recent developments. The definition and expansion of the musical note as the fundamental unit of music is approached in subsections~\ref{subsec:discNote} and~\ref{subsec:internalVar}, respectively. Subsection~\ref{subsec:notesMusic} tackles the organization of musical notes in a higher level~\cite{Wisnick,Webern,Lerdhal,Cook,Lacerda}.

%Musical theory embody topics as diverse as psycho-acoustics, cultural manifestations and formalisms. The section~\ref{sec:results} point this topics as needed and designate external complements~\cite{Zamacois,Schoenberg,microsound}.

%\subsection{Computational implementation}

%The results presented in this article are implemented as scripts, i.e.\ small computer programs implemented using accessible technologies for better distribution and validation. This constitute the \massa\ toolbox, available in public domain in an open Git repository~\cite{gitBook}. This scripts are written in Python and make use of external libraries Numpy and Scikits/Audiolab that performs calls to Fortran routines for better computational efficiency. Part of this code has been transcribed to JavaScript and native Python with readiness, what points to uses of this contribution in Web browsers such as Firefox and Chromium~\cite{numpy, audiolab, tutpython, python}.

%This are all open technologies, that is, published using licenses that allows copy, distribution and use of any part for research and derivatives. This way, the work here presented is available and eases co-authorship processes~\cite{Raymond,Lessig}.

%\subsection{Objectives}
%\label{subsec:objectives}
%The main goal of this article is to present a concise set of relations among musical basic elements and sequences of PCM audio samples. The next section is a minimum text in which music elements are presented side-by-side with the discrete-time samples they result. As validation and sharing, implementations on computer code of these relations and little musical pieces where gathered the \massa\ toolbox, available online.

%Secondary objectives include presenting a framework of sound synthesis with control at sample level, with potential uses in psychoacoustical experiments and high-fidelity synthesis. The didactic presentation of the content favors use and apprehension on a problem which calls diverse topics to be tackled: signal processing, music and psycho-acoustics, to name just a few.

%\subsection{Related work}
%Due to the general interest, and number of knowledge areas involved, there is a number of books and computer implementations that are of interest or present similarities to what is presented in this work. A more detailed comparison of them is pointed out in the bibliography~\cite{dissertacao}. There is almost no articles which could be found on the topic. In summary, there are computer implementations that use this analytical descriptions implicitly, but there is no such a concise and mathematical description of the processes implemented, as they aim to be libraries for sound and music. There are books on the topic that cover various aspects of effects and physical modeling, but none of them carry a concise description of musical elements and structures, but focus on aspects of musical sounds and ways to mimic traditional instruments.

\section{Characterization of the discrete-time musical note} \label{sec:discNote}

In diverse artistic and theoretical contexts, music is though as being constituted by units called notes, and these units taken as "atoms" that constitute music itself~\cite{Wisnick, Lovelock, Webern}. (****nao entendi****, texto ruim)
Nowadays, these notes are understood as central elements of certain musical paradigms. In a cognitive perspective, the notes are seen as discretized elements that easy and enrich the flow of information through music~\cite{Roederer, Lacerda}.
Canonically, the principal properties of a musical note are duration, volume, pitch and timbre~\cite{Lacerda}. These qualities can be managed quantitatively, dictated by the evenly time spaced sound samples.

All the relations described in this section are implemented at the file \emph{eqs2.1.py} of the \massa\ toolbox. Musical pieces \emph{5 sonic portraits} and \emph{reduced-fi} are also available online in order to corroborate the concepts.

\subsection{Duration}

The sample frequency $f_s$ is defined as the number of samples in each second of the discrete-time signal. Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds. A musical note of duration $\Delta$ seconds is presented as a sequence of $\lfloor \Delta . f_s \rfloor $ samples. That is, taken the integer part of the multiplication, it is admitted an error of at most $-\lambda_a$ seconds, which are usually fine for musical purposes. Considering $f_s=44.1kHz \;\;\Rightarrow\;\;\lambda_s=\frac{1}{44100}\approx 23$ microseconds. It is reasonable to state that:

\begin{equation}\label{eq:dur}
T_{i}^{\Delta}={\{t_i\}}_{i=0}^{\lfloor \Delta . f_a \rfloor -1}
\end{equation}

Being $\Lambda=\lfloor \Delta . f_a \rfloor$ the number of samples in the sequence, thus $T_i=\{t_i\}_0^{\Lambda-1}$.

\subsection{Volume}\label{subsec:volume}

The sensation of sound volume depends on reverberation and harmonic distribution, among other characteristics described on section~\ref{sec:varInternas}. One can get volume variations through the potency of the wave~\cite{Chowning}:

\begin{equation}\label{eq:potencia}
pot(T_i)=\frac{\sum_{i=0}^{\Lambda -1} t_i^2}{\Lambda}
\end{equation}

The final volume is dependent on the speakers amplification of the signal. Thus, what matters is the relative potency of a note in relation to the other ones around it, or the potency of a music section in relation to the rest. Differences in volume are measured in decibels, calculated directly from the amplitudes through energy or potency:

\begin{equation}\label{decibels}
V_{dB}=10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)}
\end{equation}

The quantity $V_{dB}$ has the decibel unit ($dB$). 
For each $10dB$ it is associated a "doubled volume".
Handy references are the $10dB$ for each step in the intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} e \emph{fortissimo}. Other useful references are that equivalent in $dB$ of doubling amplitude or potency:

\begin{equation}\label{eq:ampVol}
\begin{split}
t_i^{'}=2 . t_i \Rightarrow pot(T^{'}_i)=4 . pot(T_i) \Rightarrow \\ \Rightarrow V^{'}_{dB}=10log_{10} 4 \approx 6 dB
\end{split}
\end{equation}

\begin{equation}\label{eq:potVol}
\begin{split}
pot(T^{'}_i)=2 pot(T_i) \Rightarrow \\ \Rightarrow V^{'}_{dB}=10log_{10} 2 \approx 3 dB
\end{split}
\end{equation}

and the amplitude gain for a sequence whose volume has been doubled ($10dB$) is:

\begin{equation}\label{eq:dobraVol}
\begin{split}
10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)} = 10 \quad \Rightarrow \\ \Rightarrow \quad \sum_{i=0}^{\lfloor \Delta.f_a \rfloor -1}t^{'2}_i=10\sum_{i=0}^{\Lambda-1}t_i^2=\sum_{i=0}^{\Lambda-1}(\sqrt{10}.t_i)^2 \\
\therefore \quad t^{'}_i=\sqrt{10}t_i \quad \Rightarrow \quad t^{'}_i \approx 3,16t_i
\end{split}
\end{equation}

Thus, it is necessary a little bit more than triplicate the amplitude for a doubled volume. These values are guides for increasing or decreasing the absolute values on the sample sequences with musical purposes. The conversion from decibels to amplitude gain (or attenuation) is straightforward:

\begin{equation}\label{ampDec}
A = 10^{\frac{V_{dB}}{20}}
\end{equation}

where $A$ is the multiplicative factor that relates the amplitudes before and after the amplification.

\subsection{Pitch}

As explained in the previous subsections, the musical particle (note) is a sequence $T_i$ in which duration and volume corresponds to the size of sequence and the amplitude of its samples. The pitch is specified by the fundamental frequency $f_0$ whose cycle has duration $\delta_{f_0}=1/f_0$. This duration, multiplied by the sampling frequency $f_s$ results on the number of samples per cycle: $\lambda_{f_0}=f_a . \delta_{f_0} =f_a/f_0$.

For didactic reasons, be $f_0$ such that it divides $f_s$ and $\lambda_{f_0}$ results an integer. If $T_i^f$ is a sonic sequence of fundamental frequency $f$, then:

\begin{equation}\label{periodicidade}
     T^f_i=\left\{ t_i^f \right\}=\left\{ t^f_{i+\lambda_{f}}  \right\}= \left\{ t^f_{i+\frac{f_a}{f}} \right\}
\end{equation}

In the next section, frequencies $f$ that does not divide $f_s$ will be considered. We will see that this restriction does not imply in lost of generality of this section's content.

\subsection{Timbre}

While the wave period corresponds to the fundamental frequency, the trajectory of the wave inside the period - called the waveform - defines a harmonic spectrum and, thus, a timbre\footnote{Timbre is a subjective and complex characteristic. Physically, the timbre is multidimensional and given by the temporal dynamics behavior of energy in the spectral components that are harmonic or noisy. Beyond that, the word timbre is used to designate different things: one same note has different timbres, a same instrument has different timbres, two instruments of the same family have, at the same time, the same timbre that blends them in the same family, and different timbres as they are different instruments. It is worth to mention that timbre is not always manifested in spectral traces, since cultural or circumstantial aspects alter our perception of timbre}. Musically, it matters that sonic spectra with minimum differences can result in timbres with crucial expressive differences and, consequently, different timbres can be produced by using different spectra\cite{Roederer}.

The simplest (and most important case, as we will see in the following) is the spectrum that consists of only of the fundamental frequency $f$. This defines the sinusoid: a frequency in pure oscillatory movement called 'simple harmonic movement'. Let $S_i^f$ be a sequence whose samples $s_i^f$ describes a sinusoid of frequency $f$:

\begin{equation}\label{senoide}
     S^f_i=\{ s^f_i \}=\Bigl\{ \sin\bigl(2\pi \frac{i}{\lambda_f} \bigr)  \Bigr\} = \Bigl\{ \sin\bigl(2\pi f \frac{i}{f_a}\bigr)  \Bigr\} 
\end{equation}

where $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\lambda_a}$  is the number of samples in the period.

In a similar fashion, other waveforms are applied in music considering their spectral qualities and simplicity. While the sinusoid is an isolated point in the spectrum, these additional waves present a succession of harmonic components of the sinusoid. These waveforms are specified in equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} and illustrated in Figure~\ref{fig:formasDeOnda}.
Although the artificial waveforms are traditionally used in music for synthesis and oscillatory control of variables, they are also useful outside music context\cite{Openheim}.

The sawtooth presents all components of the harmonic series with decreasing energy of $-6dB/octave$. The sequence of temporal samples can be described as following:

\begin{equation}\label{sawTooth}
     D^f_i=\left\{ d^f_i \right\}=\left\{ 2\frac{i\,\%\lambda_f}{\lambda_f} -1 \right\}
\end{equation}

The triangular waveform present only odd harmonics falling with $-12dB/octave$:
\begin{equation}\label{triangular}
     T^f_i=\left\{ t^f_i \right\}=\left\{1- \left| 2 - 4\frac{i\,\%\lambda_f}{\lambda_f} \right| \right\}
\end{equation}

The sawtooth is a common starting point for a subtractive synthesis, because it has both odd and even harmonics with high energy. For musical intention, these waveforms are excessively rich in sharp harmonics, and attenuator filtering on treble and middle parts of the spectrum is specially useful for reaching a more natural and pleasant sound. 
The relatively attenuated harmonics of the triangle wave makes it the more functional - among the listed ones - to be used in the synthesis of musical notes without any treatment.

The square wave preservers only odd harmonics falling at $-6dB/octave$:

\begin{equation}\label{quadrada}
     Q^f_i=\left\{ q^f_i \right\}= \left\{
         \begin{array}{l l}
              1 & \quad \text{for } \; \; (i\,\%\lambda_f)   <  \lambda_f /2  \\
              -1 & \quad \text{otherwise}\\
         \end{array} \right.
\end{equation}

The square wave can be used in a subtractive synthesis with the purposes of mimicking a clarinet. This instrument has only the odd harmonic components and the square wave is convenient with its abundant energy in high frequencies.

\begin{figure*}
    \centering
        \includegraphics[width=\columnwidth]{figures/waveForms}
    \caption{Basic musical waveforms: (a) the synthetic  waveforms; (b) the real waveforms.}
        \label{fig:formasDeOnda}
\end{figure*}

Figure~\ref{fig:formasDeOnda} presents the waveforms described in equations  ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} and ~\ref{quadrada} for $\lambda_f=100$ (period of $100$ samples). If $f_s=44.1kHz$, a PCM standard in Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz$, reflecting the A4, just above the central "C", whatever the waveform is.

The spectrum of each basic waveform is in Figure~\ref{fig:espectroDeondas}. The isolated and exactly harmonic components of the spectrum is a consequence of the fixed period usage. The sinusoid consists of a one and only node in spectrum (**esquisito o texto**), pure frequency. As we said before, the sawtooth is the only waveform with a complete harmonic series (odd and even components). Triangular and square waves has the same components (odd harmonics), decaying at $-12dB/octave$ and $-6dB/octave$, respectively.

\begin{figure*}
    \centering
        \includegraphics[width=\columnwidth]{figures/waveSpectrum}
    \caption{Spectrum of basic artificial musical waveforms.}
        \label{fig:espectroDeOndas}
\end{figure*}

The harmonic spectrum is formed by frequencies that are multiple from the fundamental frequency $f_n=(n+1)f_0$. As the human linear perception of pitch follows  a geometric progression of frequencies, the spectrum has notes different from the fundamental frequency. Additionally, the number of harmonics will be limited to the maximum frequency $f_s/2$ (by Nyquist's theorem).

From a musical perspective, to internalize that energy in a component of frequency $f_n$ means an oscillation in the constitution of the sound, purely harmonic and in that frequency $f_n$. This energy, specifically concentrated on the $f_n$, is separated by the ear to enter in a cognitive level of processing (this separation is done in many species with mechanisms similar to the human cochlea~\cite{Roederer}).

The sinusoidal components are usually the main responsible for the timbre quality. If they are not presented in harmonic proportions (small number relations), the sound is perceived as noisy or dissonant, in opposite of a sonority with a unequivocally established fundamental frequency when they are present. Furthermore, the notion of absolute pitch in a complex sonority is based relies on the similarity of the spectrum to the harmonic series~\cite{Roederer} (**nao entendi o q vcs querem dizer**).

In the case of a fixed length period and waveform, the spectrum is perfectly harmonic and static. Each waveform is compound of specific proportions of harmonic components and the greater the curvature of the part, the greater the contribution of the part to the energy on the high harmonics. This aspect can be seen on real sounds. The wave labeled as ``sampled real sound'' in Figure~\ref{fig:formasDeOnda} has a period of $\Lambda_f=114$ samples, extracted from a relatively well behaved real sound. The oboe wave was sampled for A4 also in $44.1kHz$. The chose period for sampling was a relatively short one, with $98$ samples, and corresponds to the frequency $\frac{44100}{98}=450Hz$. It can be noticed from the curvatures, the oboe's rich spectrum in high frequencies and the lower spectrum of the real sound.

The sequence 
$ R_i=\{ r_i \}_0^{\lambda_f-1}$ of samples in the real sound of Figure~\ref{fig:formasDeOnda} can be taken as basis for a sound $T_i^f$ in the following way:

\begin{equation}\label{sampleandoFormaDeOnda}
     T^f_i=\{ t_i^f \}=\Bigl\{ r_{(i\,\%\lambda_{f})} \Bigr\}
\end{equation}

The resulting sound has the momentary spectrum of the original sound. Consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise and with variations typical of the natural phenomenon. This can be observed in Figure~\ref{fig:espectroOboe}, that shows the spectrum of the original oboe note and a note with same duration and whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}. The natural spectrum exhibits variations in the frequencies of the harmonics, in their intensities and some noise. The note made from the sampled period has a perfectly harmonic spectrum.

\begin{figure*}
    \centering
        \includegraphics[width=\columnwidth]{figures/oboeNaturalSampledSpectrum}
    \caption{Spectrum of the sonic waves of a natural oboe note and one made from a sampled period. The natural sound has fluctuations in the harmonics and in its noise, while the sampled period note has a perfectly harmonic spectrum.}
        \label{fig:espectroOboe}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Spectrum at sampled sound}

The presence and behavior of these sinusoidal components in the discretized sound have some particularities. Considering a signal $T_i$ and its corresponding Fourier decomposition $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$, the recomposition is the sum of the frequencies components as time samples\footnote{It is important to note that the factor $\frac{1}{\Lambda}$ could be distributed among the Fourier transform and its reconstruction, as preferred.}:

\begin{multline}\label{recomposicaoFourier}
t_i = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}c_ke^{j \frac{2\pi k}{\Lambda} i } \\ = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}(a_k+ j . b_k)\left[cos(w_k i)   +j . sen(w_k i)\right]
\end{multline}

where $c_k = a_k + j . b_k$ defines the amplitude and phase of each frequency: $w_k=\frac{2\pi}{\Lambda}k$ in radians or $f_k=w_k\frac{f_a}{2\pi}=\frac{f_a}{\Lambda}k$ in Hertz, taking into account the respective limits in $\pi$ and in $\frac{f_a}{2}$ given by the Nyquist Theorem.

Equation~\ref{recomposicaoFourier} shows how the imaginary term $c_k$ adds a phase to the real sinusoid. For example, the imaginary terms $b_k$ from the Fourier spectral decomposition make possible the phase sweep $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right]$ given by $tg^{-1}\left(\frac{b_k}{a_k}\right)$ which has this image. The signal from $a_k$ specifies the right and left side of the trigonometric circle, complementing the phase sweeping: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.

For a sound signal, samples $t_i$ are real and are given by the real part of equation~\ref{recomposicaoFourier}:

\begin{equation}\label{moduloEfase}
\begin{split}
t_i& = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}\left[a_k cos(w_k i) -b_k sen(w_k i)\right] \\
   & = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]
\end{split}
\end{equation}

 \begin{figure}[h!]
     \centering
         \includegraphics[width=\columnwidth]{figures/amostras2c__}
     \caption{Oscillation for 2 samples (maximum frequency for any $f_a$). The first coefficient reflects a detachment (\emph{offset} or \emph{bias}) and the second coefficient specifies the oscillation amplitude.}
         \label{fig:amostras2}
 \end{figure}

Figure~\ref{fig:amostras2} shows two samples and its spectral components. In this case, the Fourier decomposition has one unique pair of coefficients $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relatives to frequencies $\{f_k\}_0^1=\left\{w_k\frac{f_a}{2\pi}\right\}_0^1=\left\{k\frac{f_a}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_a}{2}=f_{\text{max}}\right\}$
with energies $e_k=\frac{(c_k)^2}{\Lambda=2}$. The role of amplitudes $a_k$ is clearly observed with $\frac{a_0}{2}$, the fixed detachment\footnote{Also called \emph{bias} or \emph{offset}.} and $\frac{a_1}{2}$, oscillation own amplitude given by the relation $f_k=k \frac{f_a}{\Lambda=2}$.
This case have special relevance. It is necessary the of 2 samples to represent an oscillation and it yields the Nyquist frequency $f_{\text{max}}=\frac{f_a}{2}$. That is the maximum frequency in a sound sampled with $f_a$ samples per second\footnote{Any sampled signal has this property, not only the digitalized sound.}.

All fixed sequences $T_i$ of only $3$ samples also have just $1$ frequency, since their first harmonic has $1,5$ samples and exceeds the bottom limit of 2 samples, e.g.\ the frequency of the harmonic would exceed the Nyquist frequency:  $\; \frac{2. f_a}{3} > \frac{f_a}{2} $. 
The coefficients $\{c_k\}_0^{\Lambda-1=2}$ are present in 3 frequency components. One is relative to zero frequency ($c_0$), and the other two ($c_1$ and $c_2$) have the same role for the reconstruction of sinusoid with $f=f_a/3$.

 \begin{figure}[h!]
     \centering
         \includegraphics[width=\columnwidth]{figures/amostras3b}
     \caption{Three fixed samples presents only one non-null frequency. $c_1=c_2^*$ and $w_1 \equiv w_2$.}
         \label{fig:amostras3}
 \end{figure}

$\Lambda$ real samples $t_i$ result in $\Lambda$ complex coefficients $c_k=a_k+j.b_k$. The coefficients $c_k$ are equivalent two by two, corresponding to the same frequencies and with same relevance\footnote{Equal real part and imaginary with inverse order: $a_{k1}=a_{k2}$ and $b_{k1}=-b_{k2}$. As consequence the modules are equal and phases have inverse order.}. Remembering that $f_k = k\frac{f_a}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $. When $k > \frac{\Lambda}{2}$, the frequency $f_k$ is mirrored by $\frac{f_a}{2}$ in this way: $f_k=\frac{f_a}{2} - (f_k-\frac{f_a}{2})=f_a-f_k=f_a - k\frac{f_a}{\Lambda}=(\Lambda-k)\frac{f_a}{\Lambda} \;\;\;\; \Rightarrow \;\;\;\; f_k\equiv f_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.

The same could be observed in $w_k=f_k.\frac{2\pi}{f_a}$ and considering the periodicity $2\pi$, it results in $w_k=-w_{\Lambda-k}$. Given the cosine as an even function and the inverse tangent as an odd function, the components in $w_k$ and $w_{\Lambda-k}$ sums up in the equation that reconstructs the real samples, yielding equation~\ref{recomposicaoFourier}.

In other words, in a decomposition of $\Lambda$ samples, the $\Lambda$ given frequencies components $\{c_i\}_0^{\Lambda-1}$ are equivalents in pairs.
Except for $f_0$, and when having an even $\Lambda$, from $f_{\Lambda/2}=f_{\text{max}}=\frac{f_a}{2}$ both components are isolated, e.g.\ there is any other component in frequency $f_0$ or $f_{\Lambda/2}$ (if even $\Lambda$) regarding itself. 
It is true because $f_{\Lambda/2}=f_{(\Lambda-\Lambda/2) = \Lambda/2}$ and $f_0=f_{(\Lambda-0)=\Lambda}=f_0$.
Furthermore, these two frequencies (zero and maximum frequency) are not represented having phase variation, being strictly real. In this way, it is possible to conclude the number $\tau$ of equivalent coefficient pairs:

\begin{equation}\label{coefsPareados}
\tau = \frac{\Lambda - \Lambda \% 2}{2} +\Lambda \% 2 -1
\end{equation}

This make clear visible the equivalences ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} and ~\ref{equivalenciasFases}:

\begin{equation}\label{equivalenciasFreqs}
f_{k}\equiv f_{\Lambda-k}\;, \;\; w_{k}\equiv-w_{\Lambda-k}\;\;\;, \quad \;\; \forall \quad 1 \leq k \leq \tau  
\end{equation}

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{figures/amostras4__}
    \caption{Frequential components for 4 samples.}
        \label{fig:amostras4}
\end{figure}

Having $a_k = a_{\Lambda -k}\;\;$ and $\;\;b_k = - b_{\Lambda -k}$:

\begin{equation}\label{equivalenciasModulos}
\sqrt{a_k^2 + b_k^2} = \sqrt{a_{\Lambda - k}^2 + b_{\Lambda -k}^2} \;\;, \quad \;\; \forall \quad 1 \leq k \leq \tau  \\
\end{equation}

\begin{equation}\label{equivalenciasFases}
tg^{-1}\left(\frac{b_k}{a_k}\right)=-tg^{-1}\left(\frac{b_{\Lambda -k}}{a_{\Lambda - k}}\right)\;\;,\quad \;\; \forall \quad 1 \leq k \leq \tau
\end{equation}

with $k \in \mathbb{N}$.

The equation~\ref{moduloEfase} for the real signal reconstruction, the modules and phases equivalences~\ref{equivalenciasModulos} and~\ref{equivalenciasFases}, the number of paired coefficients~\ref{coefsPareados}, and equivalence of paired frequencies~\ref{equivalenciasFreqs} expose a general case for components combination in each sample $t_i$:

\begin{multline}\label{eq:reconsCompleta}
t_i = \frac{a_0}{\Lambda} + \frac{2}{\Lambda}\sum_{k=1}^{\tau}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]+ \\ \frac{ a_{\Lambda/2}}{\Lambda}.(1-\Lambda\% 2)
\end{multline}

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{figures/amostras4formas__}
    \caption{Basic wave forms for 4 samples.}
        \label{fig:formas4}
\end{figure}

Therefore, as in Figure~\ref{fig:amostras3}, the Fourier transform of 3 samples have 2 frequencies coefficients with same amount of energy in the same frequency.

With 4 samples it is possible to represent 1 or 2 frequencies with any proportions. Figure~\ref{fig:amostras4} depicts wave forms for 4 samples and its respective two components. The individual contributions sums together yielding the original wave form and a brief inspection reveals the major curvatures resulting from the higher frequency, while a fixed detachment of component summation results from the zero frequency component.

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{figures/amostras6}
    \caption{Frequencies components for 6 samples: 3 sinusoid sums with \emph{bias}.}
        \label{fig:amostras6}
\end{figure}

Figure~\ref{fig:formas4} shows the harmonics for 4 samples in basic wave forms of equations ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} and ~\ref{quadrada}. Their conjunction results in only 1 sinusoid, with exception to sawtooth wave, which have even harmonics.

Figure~\ref{fig:amostras6} presents a sinusoidal decomposition for the 6 samples case and Figure~\ref{fig:formas6} presents the decomposition of the basic wave forms.
In this case all wave forms are different in spectrum: square and triangular ones have the same components but with different proportions, while the sawtooth have an extra component.

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{figures/amostras6formas__}
    \caption{Basic wave forms for 6 samples: triangular and square wave forms have odd harmonics, with different proportions and phases; the sawtooth wave also has even harmonics.}
        \label{fig:formas6}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The basic note}\label{notaBasica}

Consider $f$ presented in a such way that divides $f_a$\footnote{As pointed before, this limitation simplify the explanation without losing generality, and will be overcome in the next section.}. A sequence $T_i$ of sound samples are separated by $\delta_a=1/f_a$ and describes a musical note with a frequency of $f$ Hertz and $\Delta$ seconds of duration if, and only if, it has the periodicity equal to $\lambda_f=f_a/f$ and size $\Lambda=\lfloor f_a . \Delta \rfloor $:

\begin{equation}\label{eq:notaBasica}
T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{\Lambda-1}= \left \{t^f_{i \; \% \left( \frac{f_a}{f} \right) } \right \}_0^{\Lambda-1}
\end{equation}

The note by itself does not specifies a timbre. Nevertheless, it is necessary to choose the right waveform so that the samples $t_i$ have a value established individually. A unique period from the basic waveforms can be used to specify the note, where $\lambda_f=\frac{f_a}{f}$ is the number of samples at the period. Here, $L_i^{f,\, \delta_f} $ is the sequence that describes a period of the waveform $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ with duration $\delta_f=1/f$ (as given by equations~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} and ~\ref{quadrada}) and $R_i^f$ is a sampled real waveform:

\begin{equation}\label{periodoUnico}
L_i^{f , \delta_f } = \left\{ l_i^f \right\}_0^{\delta_f . f_a -1}=\left\{ l_i^f \right\}_0^{\lambda_f-1}
\end{equation}

Therefore, the sequence $T_i$ will consist in a note of duration $\Delta$ and frequency $f$ if:

\begin{equation}\label{eq:notaBasicaTimbre}
T_i^{f,\; \Delta}=\left\{t_i^f\right\}_0^{\lfloor f_a . \Delta \rfloor -1}=\left \{ l^f_{i\,\%\left(\frac{f_a}{f}\right)} \right \}_0^{\Lambda-1}
\end{equation}

\subsection{Spatial localization and spatialization}\label{subsec:spac}

Although it is not one of its four basic properties, a musical note always has a spatial localization: the note source position at the ordinary physical space. In this context, the existence of an environment that reverbs the played note it is the matter of 'spatialization'. Both fields, spatialization and spatial localization, are widely valued by audiophiles and music industry~\cite{floEsp}.

\subsubsection{Spatial localization}

It is believed that the perception of spatial localization of sound occurs in our nervous system by three information: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering realized by the human body, including its chest, head and ears~\cite{Roederer, hrtf, Heeger}.

\begin{figure}[h!]
    \centering
        \includegraphics[width=.5\textwidth]{figures/espacializacao___}
    \caption{Detection of sound source spatial localization: schema used to calculate Interaural Time Difference (ITD) and Interaural Intensity Difference (IID).}
    \label{fig:spac}
\end{figure}

Considering only the direct incidences in each ear the equations are quite simple. Given the distance between ears $\zeta$\footnote{It is known that $\zeta \approx 21,5cm$ to an adult human}, an object placed at $(x,y)$ as in Figure~\ref{fig:spac} is distant of each ear by:

\begin{equation}\label{eq:distOuvidos}
\begin{split}
d & =\sqrt{\left (x-\frac{\zeta}{2} \right )^2+y^2} \\
d' & =\sqrt{\left (x+\frac{\zeta}{2} \right )^2 + y^2}
\end{split}
\end{equation}

\noindent and immediate calculations results in the Interaural Time Difference:

\begin{equation}\label{eq:dti}
ITD=\frac{d'-d}{v_{sound\;at\;air}\approx 343.2 }\quad \text{seconds}
\end{equation}

\noindent and in the Interaural Intensity Difference:

\begin{equation}\label{eq:dii}
IID=20\log_{10}\left (\frac{d}{d'}\right) \quad decibels
\end{equation}

Converting to amplitude it gives $IID_a=\frac{d}{d'}$. The $IID_a$ can be used as a multiplicative constant to the right channel of a stereo sound signal: $\{t_i'\}_0^{\Lambda -1}=\{IID_a . t_i\}_0^{\Lambda -1}$. It is possible to use IID together with ITD as a time advance for the right channel in relation with the left one. It is a crucial entail to localization in bass sounds and percussive sonorities~\cite{Heeger}. 
Considering $\Lambda_{ITD}=\lfloor ITD . f_a \rfloor$:

\begin{equation}\label{eq:locImpl}
\begin{split}
\Lambda_{ITD} & = \left \lfloor \frac{d'-d}{343,2}  f_a \right \rfloor \\
IID_a & = \frac{d}{d'} \\
\left\{t_{(i+\Lambda_{ITD})}'\right\}_{\Lambda_{ITD}}^{\Lambda+\Lambda_{ITD}-1} & =\left\{IID_a . t_i\right\}_0^{\Lambda-1} \\
\left\{t_i'\right\}_0^{\Lambda_{ITD}-1} & = 0
\end{split}
\end{equation}

\noindent with $t_i$ as the right channel and $t_i'$ the left channel. If $\Lambda_{ITD} < 0 $, it is only needed to change $t_i$ by $t_i'$ and to use $\Lambda_{ITD}'= | \Lambda_{ITD} | $.

Although simple until here, spatial localization depends considerably of other steps. Using ITD and IID it is possible to specify only the horizontal angle (azimuthal) $\theta$ given by:

\begin{equation}\label{eq:angulo}
\theta=\tan^{-1}\left ( \frac{y}{ x }  \right )
\end{equation}

\noindent with $x,y$ as presented in Figure~\ref{fig:spac}. Henceforth, there are problems when $\theta$ focuses in the so called "cone of confusion": where the same pair of ITD and IID results in a large number of points inside the code. On those points the inference of azimuthal angle depends specially of the attenuate filtering for high frequencies, since the head interferes much more in the treble mechanical waves than in the bass waves~\cite{Heeger,hrtf}. It is also relevant to the hearing of source side, where the sound has enough low frequencies there is a diffraction and the wave comes to the ear with a delay of $\approx 0,7ms$.\cite{floEsp}

Figure~\ref{fig:spac} also shows an acoustic shadow from cranium, an important phenomenon to perception of source azimuthal angle in the cone of confusion. The cone itself is not shown in Figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not encountered in the literature. This makes it hard to conceive given the filtering and diffraction dependent of the sound spectrum. In this way, the cone of confusion can be understood as a cone with its top side placed in the middle of head and growing out in the direction of each ear~\cite{hrtf}.

On the other hand, the complete localization, including height and distance of sound source, is given by the Head Related Transfer Function (HRTF)~\cite{hrtf}. There are well known open databases of HRTF as CIPIC and it is possible to apply those transfer functions in a sound signal by convolution (see equation~\ref{eq:conv})~\cite{CIPIC}. The human body considerably changes the filtering and there are techniques to generate HRTFs to be universally used, as expected~\cite{lazaSPA}.

\subsubsection{Spatialization}

Spatialization results from sound reflections and absorptions by room/environment surface where the note is played. The sound propagates throw the air with a speed of $\approx 343,2m/s$ and can be sent from a source without any directionality pattern. When a sound pulse encounters a surface there is reflection. In this reflection occurs: 1) the component inversion of propagation speed whose is normal to the surface, as 2) the energy absorption, specially in trebles. The sound waves propagate until they reach inaudible levels. The moment when some sound pulse reach the human ear can be described as the moment it reaches the ear and the absorption filters of each surface it has reached. It is possible to simulate reverberations that are impossible in real systems. To experimentation, it is possible to use asymmetric reflections with relation to the axis perpendicular to the surface, or to increase specific frequency bands (known as 'resonances'), both characteristics are now found in real systems.

There are some reverberation models that are less related to the calculation of each reflection, exploring valuable information of the auditory system. In fact, reverberation can be modeled with a set of 2 temporal and spectral characteristics:

\begin{itemize}
   \item First period: 'first reflections' are more intense and scattered.
   \item Second period: 'late reverberation' is practically a dense succession of indistinct delays with exponential decay and statistical occurrences.
   \item First band: bass has some resonance frequencies relatively spaced.
   \item Second band: mid and treble have progressive decay and smooth statistical fluctuations.
\end{itemize}

Smith III points out that reasonable concert rooms have total reverberation time of approx. $1,9$ seconds, and that the period of first reflections is around $0,1$ seconds. These values suggests that, in the given conditions, there are perceived wave pulses which propagates for $652,08$ meters long ($83,79k$ samples in $f_a=44,1kHz$) before reaching the ear. In addition, sound reflections made after propagation for $34,32$ meters long ($4,41k$ samples in $f_a=44,1kHz$) have a tangled form whose incidences are less distinct by hearing. These first reflections are particularly important to space sensation. The first incidence is the direct sound, described by ITD and IID of equations~\ref{eq:dti} and~\ref{eq:dii}. Admitting that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$ dependents of room dimension, the separation between the first reflection is at  $8-90ms$ ($\approx 350-4000$ samples in $f_a=44.1kHz$). It is possible to experimentally verify that the number of reflections increases with squared proportion $\approx k.n^2$. A discussion about the use of convolutions and filtering to favor these implementations is provided in subsection~\ref{subsec:mus2}, specially at paragraphs about reverberation.

\subsection{Musical use}\label{subsec:basMus}

Once we have well defined the basic note, it is time to build musical structures with sequences based on those particles. The sum of elements with same index of $N$ sequences $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ with same size $\Lambda$ results in overlapped spectral contents in a process called sound mixing:

\begin{equation}\label{eq:mixagem}
\{t_i\}_0^{\Lambda-1}=\left \{ \sum_{k=0}^{N-1}t_{k,i} \right \}_0^{\Lambda-1}
\end{equation}

\begin{figure}[h!]
    {\centering
        \includegraphics[width=\columnwidth]{figures/mixagem}}
    \caption{Mixing of three sound sequences. The amplitudes are directly overlapped.}
        \label{fig:mixagem}
\end{figure}

Figure~\ref{fig:mixagem} illustrates this overlapping process of discretized sound waves, disposing 100 samples. It is possible to conclude that, if $f_a=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_a}{100/2}=882Hz$, $\frac{f_a}{100/4}=1764Hz$ and $\frac{f_a}{100/5}=2205Hz$. The samples duration are very short $\frac{f_a=44.1kHz}{100} \approx 2 \text{ milliseconds}$. Someone just need to complete the sequence with zeros to sum up sequences with different sizes.

The mixed notes are generally separated by the ear according to the physic laws of resonance and by the nervous system~\cite{Roederer}.  This musical notes mixing process results in musical harmony whose intervals between frequencies and chords of simultaneous notes guide subjective and abstract aspects of music and its appreciation~\cite{Harmonia}.

The sequences can be concatenated in time. If the sequences $\{t_{k,i}\}_0^{\Lambda_k-1}$ of size $\Lambda_k$ represent $k$ musical notes, their concatenation in a unique sequence $T_i$ is a simple musical sequence or melody:

\begin{multline}\label{eq:concatenacao}
\{t_i\}_0^{\sum\Delta_k-1}=\{t_{l,i}\}_0^{\sum\Delta_k-1}, \;\; \\ l\text{ smallest integer } : \quad \Lambda_l > i -\sum_{j=0}^{l-1}\Lambda_j
\end{multline}

This mechanism is demonstrated in Figure~\ref{fig:concatenacao} with same sequences of Figure~\ref{fig:mixagem}. Although the sequences are short for the used sample rates, it is possible to observe the concatenation of sound sequences. Beside that, each note has it duration larger than $100ms$ if $f_a<1kHz$.

\begin{figure}[h!]
{    \centering
        \includegraphics[width=\columnwidth]{figures/concatenacao}}
    \caption{Concatenation of three sound sequences by temporal overlap of its samples.}
        \label{fig:concatenacao}
\end{figure}

The musical piece \emph{reduced-fi} explores the temporal overlapping of notes, resulting in a homophobic piece. The vertical principle is demonstrated at the \emph{sound boards}, static sounds with peculiar spectrum. Both pieces were written in Python at Appendices~\ref{ap:quadros} and~\ref{ap:reduced}, and are available as part of the \massa \emph{toolbox}.\cite{MASSA}

Now that the basic digital music note is described, the next section develops the temporal evolution of its contents as in \emph{glissandi} and volume envelopes. The filtering of spectral components and noise generation complements the definition of the musical note as a self contained unity. Section~\ref{notasMusica} is dedicated to the organization of these notes as music by using metrics and trajectories.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Variation in the basic musical note}\label{sec:internalVar}

The basic digital music note defined in section~\ref{sec:notaDisc} has following parameters: duration, pitch, intensity (volume) and timbre. This is a useful and paradigmatic modeling but does not covers all the aspects regarding a musical note.

At first place, the characteristics of the note are modified along the duration of the note itself~\cite{Chowning}. For example, a note played in a piano with duration of 3 seconds has abrupt intensity at its beginning time of execution. Then, it behaves with progressively decays, spectrum variations, presenting harmonics decaying before others and presenting harmonics showing up along the time.
Those variations are not mandatory but they are guidance from sound synthesis to musical uses, because it suitable reflects how sounds are represented in nature\footnote{The gold tip here is: to make a sound that incites interest by itself, do internal variations on it~\cite{Roederer}.}.
To explore all the possible ways which those variations occur is out of the scope of this study, given the considerably sensibility of the human ear and the complexity of our sound cognition. In the following we point out primary resources to produce those characteristic variations in the basic note. It is worthwhile to mention that all the described relations in the following section are currently implemented in Python at Appendix~\ref{sec:cod2}. The music pieces \emph{Transita para metro}, \emph{Vibra e treme}, \emph{Tremolos, vibratos e a frequência}, \emph{Trenzinho de caipiras impulsivos}, \emph{Ruidosa faixa}, \emph{Bela rugosi}, \emph{Chorus infantil}, \emph{ADa e SaRa} are listed in Appendices~\ref{ap:transita}, \ref{ap:vibra}, \ref{ap:tremolos}, \ref{ap:trenzinho}, \ref{ap:ruidosa}, \ref{ap:bela}, \ref{ap:chorus} and \ref{ap:ada}. The code is also part of the toolbox \massa, available online.\cite{MASSA}

\subsection{Lookup Table}\label{subsec:lookup}

The \emph{Lookup Table} (or simply LUT), is a data structure array which substitutes continuous and repetitive calculation by indexed access. It is frequently used to reduce computational complexity and to make possible the use of functions without the possibility of direct calculation, as the data sampled from nature.
In music its usage transcends the former, simplifying operations and making possible to use a single wave period to synthesize sounds in the whole audible spectrum, no matter what kind of wave was originally sampled.

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{figures/lut}
    \caption{Search procedure in an indexed array (known as \emph{Lookup Table}) to synthesize sounds in different frequency using an unique waveform with high resolution.}
        \label{fig:lut}                                                                                                            
\end{figure}

Being $\widetilde{\Lambda}$ the wave period size and $\widetilde{L_i} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ the elements $\widetilde{l_i}$ of any wave period (refer to Equation~\ref{periodoUnico}), a sequence $T_i^{f,\,\Delta}$ with samples of a sound with frequency $f$ and duration $\Delta$ can be obtained by means of $\widetilde{L_i}$ as:

\begin{multline}\label{eq:lut}
T_i^{f,\,\Delta}=\left\{t_i^f\right\}_0^{\lfloor \, f_a . \Delta \, \rfloor -1} = \left\{ \, \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} }\, \right\}_{0}^{\Lambda-1}\; , \quad \\ \text{where} \;\; \gamma_i = \left \lfloor i . f \frac{ \widetilde{\Lambda}}{f_a} \right \rfloor  
\end{multline}

In other words, with the right LUT indexes ($\gamma_i\%\widetilde{\Lambda}$) it is possible to synthesize a sound with any frequency. Figure~\ref{fig:lut} illustrates the calculation of a $\{t_i\}$ sample from $\left\{\,\widetilde{l}_i\,\right\}$ for a frequency of $f=200Hz$, $\widetilde{\Lambda}=128$ and considering the sample rate as $f_a=44.1kHz$.
Though this is not a practical configuration (as assigned below), it makes possible a graphical visualization of the procedure.

The calculation of the integer $\gamma_i$ introduces noise which decreases as $\widetilde{\Lambda}$ increases.
In order to use this calculation in sound synthesis, with $f_a=44.1 kHz$ the standard guidelines suggest the use of $\widetilde{\Lambda} = 1024$ samples, since it does not produce relevant noise considering the audible spectrum. The round or interpolation method are not decisive in this process~\cite{Geiger}.

The expression defining the variable $\gamma_i$ can be understood as $f_a$ being summed to $i$ at each $1$ second.
If $i$ is divided by the sample frequency, it results in $\frac{i}{f_a}$,
which is incremented in $1$ at each $1$ second. Multiplied by the period length it results in $i \frac{\widetilde{\Lambda}}{f_a}$ that look up the period in $1$ second. Finally, with frequency $f$ it results in $i . f \frac{\widetilde{\Lambda}}{f_a}$ which complements $f$ lookups of the period $\widetilde{\Lambda}$ in $1$ second, i.e., the resulting sequence presents the fundamental frequency $f$.

There are important considerations here: it is possible to use any value for $f$, the limits exist only in the loud frequencies when the size of table  $\widetilde{\Lambda}$ is not sufficient enough considering the sample rate $f_a$. The lookup table procedure is computationally costless and replaces the calculations by simple indexed searches (what is generally understood as an optimization process). Unless otherwise stated, this procedure will be used along all the text for every applicable case.

LUTs are broadly used in computational implementations for music. A classical usage of LUTs is known as \emph{Wavetable Synthesis}, which consists of many LUTs used together to generate a music note quasi-periodic~\cite{Cook,Wavetable}.

\subsection{Incremental Variations of Frequency and Intensity}\label{subsec:vars}

Following the Weber and Fechner~\cite{Weber-Fechner} law, the human perception has a logarithmic relation with the causing stimulus. In other words, a stimulus following a exponential progression is perceived as linear.
For didactic reasons and given its use in AM and FM synthesis (refer to subsection~\ref{subsec:tvaf}), the linear variation is discussed first.

Considering the note with duration $\Delta = \frac{\Lambda}{f_a}$, the frequency $f=f_i$ varies linearly from $f_0$ to $f_{\Lambda -1}$. In this way, it is possible to define the following:

\begin{equation}\label{freqLinear}
 F_i=\{f_i\}_0^{\Lambda-1}=\left\{f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right\}_0^{\Lambda-1}
\end{equation}

\begin{multline}\label{indiceLinear}
 \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i= \\ \left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{j}{\Lambda-1} \right ] \right \rfloor 
\end{multline}

\begin{equation}\label{serieAmostralLin}
 \left\{t_i^{\;\overline{f_0,\, f_{\Lambda-1}}}\right\}_0^{\Lambda-1}=\left\{\,\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\,\right\}_0^{\Lambda-1}
\end{equation}

where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a}$ is the LUT increment between two samples given the sound frequency in the first sample.

Thus, it is possible to calculate the elements $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ having the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$.

The equations~\ref{freqLinear}, \ref{indiceLinear} and \ref{serieAmostralLin} are related with the linear progression of the frequency. We recall that, for the general case, the frequency progression \emph{perceived} as linear follows an exponential progression\footnote{In other words, a geometric progression of frequency is perceived as an arithmetic progression of pitch.}.
It is possible to write this such that: $f_i=f_0 . 2^{\frac{i}{\Lambda-1} n_8}$ where  $n_8=\log_2\frac{f_{\Lambda-1}}{f_0}$ is the number of octaves between $f_0$ and $f_{\Lambda-1}$.
Therefore, $f_i=f_0 . 2^{\frac{i}{\Lambda-1}\log_2\frac{f_{\Lambda-1}}{f_0}}=
 f_0 . 2^{\log_2\left ( \frac{f_{\Lambda-1}}{f_0} \right )^{\frac{i}{\Lambda-1}}}=
 f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}$. Accordingly, the transition equations of linear frequency for the human ear are given by:

\begin{multline}\label{freqExponencial}
 F_i=\{f_i\}_0^{\Lambda-1}= \\ \left\{f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}} \right\}_0^{\Lambda-1}
\end{multline}

\begin{multline}\label{indiceExponencial}
 \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \\ \quad \gamma_i=\left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor \sum_{j=0}^{i} f_0 \frac{\widetilde{\Lambda}}{f_a} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{j}{\Lambda -1}} \right \rfloor
\end{multline}

\begin{equation}\label{serieAmostralLog}
 \left\{t_i^{\;\overline{f_0,\,f_{\Lambda-1}}}\right\}_0^{\Lambda-1}=\left\{\,\widetilde{l}_{\gamma_i \% \widetilde{\Lambda}}\,\right\}_0^{\Lambda-1}
\end{equation}

\begin{figure}[h!]
     \centering
         \includegraphics[width=\columnwidth]{figures/transicao}
     \caption{Intensity transitions to different values of $\alpha$ (see equations~\ref{seqAmp} and ~\ref{transAmp}).}
         \label{fig:transicao}
\end{figure}

The term $\frac{i}{\Lambda-1}$ looks up the interval $[0,1]$ and it is possible to power it to a value in such a way that the beginning of the transition will be smoother or steeper.
This procedure is useful for energy variations of a vibratory wave to change the volume intensity\footnote{The volume change (psychophisical quality) is consequence of different sound characteristics like reverberation and concentration of treble harmonics, among which is wave energy. Wave energy is the easiest one to modify (see Equation~\ref{eq:potencia}) and it can vary in many ways. The most simple way consists of modifying the amplitude by multiplying the whole sequence by a real number. The increased energy without amplitude variation is the \emph{sound compression}, useful nowadays for music production~\cite{guillaume}}. It is sufficient to multiply the original sequence (being it generated or predetermined) by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is the given coefficient and $a_{\Lambda-1}$ is a fraction of the original amplitude, which is the aimed value to change until the end of transition.

Thus, for amplitude variations:

\begin{multline}\label{seqAmp}
 \{a_i\}_0^{\Lambda-1}=\left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}= \\ \left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1} \text{ com } a_0=1
\end{multline}

\begin{equation}\label{transAmp}
 T_i^{'}=T_i \odot A_i = \{t_i . a_i\}_0^{\Lambda-1}=\left \{ t_i . (a_{\Lambda-1} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
\end{equation}

It is possible to take $a_0=1$ in order to start a new sequence with the original amplitude and then keep changing it with the following samples.
This restriction makes $a_{\Lambda-1}$ the amplitude variation term.
If $\alpha=1$, the amplitude variation follows the geometric progression whose defines the linear progression. Figure~\ref{fig:transicao} shows the transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by equation~\ref{eq:ampVol}.

An special attention should be driven while considering $a=0$.
In equation~\ref{seqAmp}, $a_0=0$ results in a division by zero and if $a_{\Lambda-1}=0$, there will be multiplication by zero. Both cases make the procedure useless, once no number different of zero can be represented as a ratio in relation of zero. It is possible to solve this dilemma choosing a number that is small enough like $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum volume for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$).

For linear amplification -- but not linear to perception -- it is sufficient to use an appropriate sequence $\{a_i\}$:

\begin{equation}\label{seqAmpLin}
a_i=a_0 + (a_{\Lambda-1}-a_0)\frac{i}{\Lambda-1}
\end{equation}

Here the conversation between decibels and amplitude is convenient. In this way, the equations~\ref{ampDec} and \ref{transAmp} specify a transition of $V_{dB}$ decibels:

\begin{equation}\label{seqAmpDB}
T_i^{'}=\left\{ t_i 10^{\frac{V_{dB}}{20}\left( \frac{i}{\Lambda-1} \right)^\alpha} \right\}_0^{\Lambda-1}
\end{equation}

\noindent for the general case of amplitude variations following a geometric progression. The greater the value of $\alpha$, the smoother the sound introduction and more intense its end. $\alpha>1$ results in volume transitions commonly called \emph{slow fade}, while $\alpha<1$ results in \emph{fast fade}.\cite{guillaume}

The linear transitions will be used for AM and FM synthesis and for applications of logarithmic transitions for tremolos and vibratos. A non-oscillatory exploration of those variations is in the music piece \emph{Transita para metro}, which code is in Appendix~\ref{ap:transita} and online at \massa.\cite{MASSA}

\subsection{Application of Digital Filters}\label{subsec:filtros}

This subsection is limited to a description of the procedure in sequences by convolution and differential equations, and the immediate applications. Its complexity exceeds to the scope of this study\footnote{The implementation of filters comprehends an area of recognized complexity with dedicated literature and software implementations. We recommended the reader to consult the bibliography~\cite{Openheim,smith}.}. The filters application can be part of the synthesis process or made individually as part of processes commonly called sound treatment.

\begin{itemize}
  \item  Convolution and finite impulse response (FIR) filters

\begin{figure}[h!]
     \centering
         \includegraphics[width=\columnwidth]{figures/convolucao}
     \caption{Graphical interpretation of convolution. Each resulting sample is the sum of the previous samples of a signal, with each one multiplied by the retrograde sequence of samples from the other signal.}
         \label{fig:conv}
\end{figure}

Filters applied by means of convolution are known by acronym FIR (Finite Impulse Response) and are characterized by having a time finite sample representation. This sample representation is called 'impulse response' $\{h_i\}$. The FIR filters are applied to the time domain of the digital sound by means of convolution with the respective impulse response of the filter\footnote{It is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both sound and the impulse response, and then performing the inverse Fourier transform in the resulting spectrum.\cite{Openheim}}. For this study, the convolution is defined as:

\begin{multline}\label{eq:conv}
 \begin{split}
 \left\{t_i'\right\}_0^{\Lambda_t+\Lambda_h-2\; = \;\Lambda_{t\, '}-1} =\{(T_j*H_j)_i\}_0^{\Lambda_{t \, '}-1} & = \\ \left \{ \sum_{j=0}^{min(\Lambda_h-1,i)}h_{j} . t_{i-j} \right \}_0^{\Lambda_{t\, '}-1} 
     & = \\ \left \{ \sum_{j=max(i+1-\Lambda_h,0)}^{i}t_j . h_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
 \end{split}
\end{multline}

\noindent where $t_i=0$ for the samples not given beforehand.
In other words, the sound $\{t_i'\}$ resulting from the convolution of $\{t_i\}$ with the given impulse response $\{h_i\}$ has each $i$-th sample $t_i$ overwritten by the sum of its last $\Lambda_h$ samples $\{t_{i-j}\}_{j=0}^{\Lambda_h-1}$ multiplied one-by-one by samples of the another impulse response $\{h_i\}_0^{\Lambda_h-1}$. This procedure is illustrated in Figure~\ref{fig:conv}, where the impulse response $\{h_i\}$ is traveled by its retrograde form, and $t_{12}'$ and $t_{32}'$ are two calculated samples using the convolution given by $(T_j*H_j)_i=t_i'$. The resulting signal always has the length of $\Lambda_t+\Lambda_h -1=\Lambda_{t'}$.

With this procedure it is possible to apply reverbs, equalizers, \emph{delays} to name a few of a variety of other filters for sound processing, as well as musical/artistic effects.

The impulse response can be provided by physical measures or by pure synthesis. An impulse response for reverb application can be obtained by sound recording of the environment when someone triggers a snap which resembles an impulse or a sinusoidal lookup whose Fourier transform approximates its frequency response. Both are impulse responses which, properly convoluted with the sound sequence, result in the own sequence with a reverberation that resembles the physical one in the original environment where the measure happened~\cite{Cook}.

The inverse Fourier transform of an envelope even and real is equivalent to a impulse response of a FIR filter. It performs a frequency filtering with the envelope. The greater the number of samples, the higher the envelope resolution and the computational complexity as well. The convolution is recognized as computationally expensive.

An important property is the time shift caused by convolution with a shifted impulse. Despite computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse to each relapse of sound.
In Figure~\ref{fig:delays} it is possible to observe the shifting caused by convolution with each impulse. Depending on the intensity of the impulses, the result is perceived in rhythm (around 20 impulses by second) or in pitch (around 20-40 impulses by second). In the last case, the processes resemble granular synthesis, delays, reverbs and equalization.

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{figures/delays}
    \caption{Convolution with impulse: shifting (a), delay lines (b) and granular synthesis~(c). Disposed in increasing order of its pulse density.}
        \label{fig:delays}
\end{figure}

\item Infinite impulse response (IIR) filters

This class of filters is known by the acronym IIR. They are characterized by having an infinite time representation, i.e.\ the impulse response does not converges to zero. Its application is usually made by the following equation:

\begin{equation}\label{eq:diferencas}
 t_i' = \frac{1}{b_0}\left ( \sum_{j=0}^Ja_j . t_{i-j} + \sum_{k=1}^Kb_k . t_{i-k}' \right )
\end{equation}

\noindent with $b_0=1$ in the most cases it is possible to normalize the variables: $a_j'=\frac{a_j}{b_0}$ and $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
Equation~\ref{eq:diferencas} is called 'difference equation' because the resulting samples $\left\{t_i'\right\}$ are given by differences between the original samples $\{t_i\}$ and the previous resulting ones $\left\{t_{i-k}'\right\}$.

There are many methods and tools to prepare IIR filters. The following texts list a selection of them for didactic purpose and to make easy future consulting. They are well behaved filters whose aspects are described in Figure~\ref{fig:iir}.

Considering filters with simple order, the cutoff frequency $f_c$ is the position where the filter performs an attenuation of $-3dB \approx 0.707 $ in the original amplitude.
For band-pass and band-reject (or 'notch') filters, this attenuation results on two specifications: $f_c$ (in this case, referred as 'center frequency') band width $bw$,
in both frequencies $f_c \pm bw$ there is an attenuation of $\approx 0.707$ in the original amplitude.
There is sound amplification in band-pass and band-reject filters when the cutoff frequency is low and the band width is large enough. In trebles those filters present only a shift in the expected waveform, expanding the envelope to the bass side of the band.

For filters whose frequency responses have other envelopes (for the module), it is possible to create chains and apply them successively. Another possibility is to use some biquad 'filter receipt'\footnote{Short for 'biquadratic': its transfer function has two poles and two zeros, i.e. its first direct form consists of two quadratic polynomials forming the fraction: $\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1- b_1.z^{-1} -b_2 . z^{-2}}$.} or procedures for the calculation of coefficients of Chebichev filters\footnote{Butterworth and Elliptical filters can be considered as specific cases of Chebichev filters.\cite{Openheim,smith}}.
Both possibilities are explored by the referenced studies, in special~\cite{JOSFM,smith}, and the collection of filters maintained by the \emph{Music-DSP} community from University of Columbia.\cite{music-dsp,Openheim}

\end{itemize}

\begin{enumerate}
  \item Low-pass with simple pole and module of the frequency response in the upper left corner of Figure~\ref{fig:iir}. The general equation has as reference by the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_a$ that there is approximately an attenuation of $3dB$.
  The coefficients $a_0$ and $b_1$ of the IIR filter are given by means of the intermediate variable $x \in [e^{-\pi},1]$:

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{figures/iir}
    \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of equations~\ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} and \ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
        \label{fig:iir}
\end{figure}

\begin{equation}\label{eq:passa-baixas}
 \begin{split}
 x & =e^{-2\pi f_c} \\
 a_0 & =  1-x \\
 b_1 & =  x
 \end{split}
\end{equation}

  \item High-pass filter with simple pole and modules of its frequency responses at the upper right corner of Figure~\ref{fig:iir}. The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of the intermediate variable $x \in [e^{-\pi},1]$:

\begin{equation}\label{eq:passa-altas}
 \begin{split}
 x & =e^{-2\pi f_c} \\
 a_0 & =  \frac{x+1}{2} \\
 a_1 & =  -\frac{x+1}{2} \\
 b_1 & =  x
 \end{split}
\end{equation}

%\item Passa-banda
%\item Rejeita-banda

\item Notch filter. This filter is parametrized by a center frequency\footnote{Attention with the cutoff frequency $f_c$ in low-pass and high-pass filters.} $f_c$ and band width $bw$ -- $f_c \pm bw$ that results in $0.707$ of the amplitude, i.e. attenuation of $3dB$ -- both given as fractions of $f_a$, therefore $f,\; bw \in (0,0.5)$.

In order to simplify, consider the auxiliary variables $K$ and $R$ defined as:

\begin{equation}\label{eq:varAux}
 \begin{split}
  R & = 1 - 3bw \\
  K & = \frac{1-2R\cos(2\pi f_c) + R^2}{2 - 2 \cos (2 \pi f_c)}
 \end{split}
\end{equation}

The band-pass filter in the lower left corner of Figure~\ref{fig:iir} has the following coefficients in equation~\ref{eq:diferencas}:

\begin{equation}\label{eq:passa-banda}
 \begin{split}
 a_0 & =  1 - K \\
 a_1 & =  2(K-R)\cos (2\pi f_c) \\
 a_2 & =  R^2-K \\
 b_1 & =  2R \cos (2\pi f_c) \\
 b_2 & =  -R^2
 \end{split}
\end{equation}

The coefficients of band-reject filter are:

\begin{equation}\label{eq:rejeita-banda}
 \begin{split}
 a_0 & =  K \\
 a_1 & =  -2K\cos (2\pi f_c) \\
 a_2 & =  K \\
 b_1 & =  2R \cos (2\pi f_c) \\
 b_2 & =  -R^2
\end{split}
\end{equation}

\noindent with the module of its frequency response represented in the lower left corner of the Figure~\ref{fig:iir}.

%\item Biquad: pela especificação de uma frequência central, da qualidade
%e da intensidade do filtro, este filtro é simples e usual para áudio,
%permitindo ajustes mais finos. Diversas receitas podem ser encontradas
%na literatura, recomendamos especialmente as diferentes especificações
%em ~\ref{musicDSP} e ~\ref{dspguide}.

\end{enumerate}

\subsection{Noise}\label{subsec:ruidos}

Generally, sounds without a defined pitch are called noise~\cite{Lacerda}.
They are important parts of the musical sounds with undefined pitch like the noise present in piano notes, violin, etc. Besides that, the majority of percussion instruments do not have defined pitch and their sounds are generally comprehended as noise~\cite{Roederer}. In electronic music, including electro-acoustic and dance genres, noise has diversified use, and commonly characterizes the music style~\cite{Cook}.

The absence of a defined pitch is due to the absence of a harmonic organization in the sinusoidal components that compose the sound.
In this way, there are many ways to generate noise.
The use of random values to generate the sound sequence $T_i$ is an attractive method but the result is generally not useful because it tends to the white noise~\cite{Cook}.

Another way is to generate noise using the desired spectrum, from which it is possible to perform the inverse Fourier transform.
The spectral distribution should be done with care because, if it uses the same phase or phases with strong correlation, the synthesized sound will include considerably amount of concentrated energy in some periods of its duration.

\begin{figure}[htpq!]
     \centering
         \includegraphics[width=\columnwidth]{figures/ruidos}
     \caption{Colors of noise generated by equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul} and \ref{eq:violeta}: resulting spectrum and waveforms.}
         \label{fig:ruidos}
\end{figure}

Some noise with static spectrum are listed below. They are called \emph{colors of noise} since they are associated with colors.
Figure~\ref{fig:ruidos} shows the spectrum profile and the corresponding sound sequence side-by-side. All the noise were generated with a same phase, making it possible to observe the contributions of different regions of the spectrum.

\begin{itemize}

 \item The white note has its name because its energy are distributed equally between all the frequencies. It is possible to realize the white noise with the inverse transform of the following coefficients:

\begin{equation}\label{eq:branco}
 \begin{split}
 c_0 & =0 \quad \text{avoiding bias} \\
 c_i & =e^{j.x}\;,\;\; j^2=-1 \;, \;\; x \; \text{random} \; \in \; [0,2\pi]\;,\;\; i \; \in \; \left[1, \, \frac{\Lambda}{2}-1\right] \\
 c_{\Lambda/2} & = 1 \quad\quad \text{(if even $\Lambda$)}\\ 
 c_i & = c_{\Lambda - i}^*\;,\;\; \text{for}\;  i \; > \;  \frac{\Lambda}{2}
 \end{split}
\end{equation}

The value $c_i$, calculated by the exponential, is only an artifice to obtain unitary module and random phase. Besides that, $c_{\Lambda/2}$ is always real (as discussed in the previous section).

 \item The pink noise has a decrease of $3dB$ for octave. This noise is useful for testing electronic devices, and it has prominent presence in nature~\cite{Roederer}.

\begin{multline}\label{eq:rosa}
 %\begin{split}
 \qquad \qquad \qquad f_{\text{min}}  \approx 15 Hz \\
 f_i  = i \frac{f_a}{\Lambda} \;, \;\; \quad i \;\leq\; \frac{\Lambda}{2},\;\; i\;\in\;\mathbb{N}  \\
 \alpha_i  = \left(10^{-\frac{3}{20}}\right)^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )}  \\
 c_i  =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \\
 c_i  =e^{j.x} . \alpha_i\;,\;\; j^2=-1 \;, \;\;\ \\ x \;\; \text{random} \; \in \; [0,2\pi]\;,\;\; \forall \; i \; : f_{\text{min}} \le f_i < f_{\lceil \Lambda/2-1 \rceil}  \\
 c_{\Lambda/2}  = \alpha_{\Lambda/2} \quad \text{(se $\Lambda$ par)} \\ 
 c_i  = c_{\Lambda - i}^*\;,\;\; \text{for}\;  i \; > \;  \Lambda/2 \qquad \qquad
 %\end{split}
\end{multline}

The minimum frequency $f_{\text{min}}$ can be chosen based on the human hearing limit, since no one listens to a pitch with a sound component which frequency is bellow $\approx\; 20Hz$.

Other noises can be made from the pink noise procedure by simply modifying some details, specially the equation that defines $\alpha_i$.

  \item The brown noise received this name after Robert Brown, who described the brownian movement\footnote{Although its origin is disparate given its association with the brown color, this noise become established with this specific name.
  Anyway, this association is satisfactory once the white and pink noises are more strident and associated with intense colors~\cite{Cook,guillaume}.}

What characterizes this noise is the decreasing of $6dB$ by octave. In this way, $\alpha_i$ in the set \ref{eq:rosa} is defined as:

\begin{equation}\label{eq:marrom}
 \alpha_i=(10^{-\frac{6}{20}})^{\log _2 \left( \frac{f_i}{f_{\text{min}}} \right )}
\end{equation}

 \item In the blue noise there is a gain of $3dB$ by octave in a band limited by the minimum frequency $f_{\text{min}}$ and the maximum frequency $f_{\text{max}}$. Therefore, the corresponding equation is also based on the equations set~\ref{eq:rosa}:

\begin{equation}\label{eq:azul}
 \begin{split}
 \alpha_i & = (10^{\frac{3}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \\
 c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \;\; \text{ou} \;\; f_i>f_{\text{max}} \\
 \end{split}
\end{equation}

 \item The violet noise is similar to the blue noise, but its gain is $6dB$ by octave:

\begin{equation}\label{eq:violeta}
 \alpha_i = (10^{\frac{6}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \;\;, \quad f_{\text{min}} \approx 15 Hz \\
\end{equation}

 \item The black noise has higher losses than $6dB$ for octave:

\begin{equation}\label{eq:preto}
 \alpha_i=(10^{-\frac{\beta}{20}})^{\log _2 \left( \frac{f_i}{f_{\text{min}}} \right )}\;\;, \quad \beta > 6
\end{equation}

 \item The gray noise is defined as a white noise subject to one of the ISO-audible curves. Those curves are obtained by experiments and imperative to obtain $\alpha_i$. An implementation of ISO 226, that is the last revision of those curves, is in the toolbox \massa.\cite{MASSA}

\end{itemize}

This subsection exposed only noises with static spectrum. There are also characterizations for noises with dynamic spectrum during the time, or noises which are fundamentally transient, like clicks and chirps. The former are easily modeled by an impulse relatively isolated, while chirps are not in fact a noise, but a fast scan of some given frequency band~\cite{Cook}.

The noise from equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom},
\ref{eq:azul} and \ref{eq:violeta} are presented in Figure~\ref{fig:ruidos}. The spectrum were build with the same phase and frequency for each coefficient, making it straightforward to observe the contribution of treble harmonics and bass frequencies.

\subsection{Tremolo and vibrato, AM and FM}\label{subsec:tvaf}

Vibrato is a period variation in pitch (frequency) and tremolo is a period variation in volume (intensity)\footnote{Some musical instruments and contexts use different terms. For example, in piano, the called tremolo is a vibrato and a tremolo is the classification used here. The presented definitions are common in contexts regarding music theory and electronic music. In addition, they are based on a broader literature than the one used for an specific instrument, practice or musical tradition~\cite{Lacerda,Harmonia}.}
For the general case, the vibrato is described as follows:

\begin{equation}\label{vbrGamma}
 \gamma_i'=\left \lfloor i f' \frac{\widetilde{\Lambda}_M}{f_a} \right \rfloor
\end{equation}

\begin{equation}\label{vbrAux}
 t_i'=\widetilde{m}_{\gamma_i' \;\% \widetilde{\Lambda}_M}
\end{equation}

\begin{equation}\label{vbrF}
 f_i=f \left ( \frac{f + \mu }{f} \right )^{t_i'}=f . 2^{t_i'\frac{\nu}{12}}
\end{equation}

\begin{multline}\label{vbrGamma2}
 \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \\ = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f \left ( \frac{f + \mu }{f} \right )^{t_j'}  \right \rfloor= \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f . 2^{t_j'\frac{\nu}{12}}  \right \rfloor
\end{multline}

\begin{equation}\label{vbrT}
 T_i^{f, vbr(f',\,\nu)}=\left\{ t_i^{f,vbr(f',\,\nu)} \right\}_0^{\Lambda-1}=\left\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \right\}_0^{\Lambda-1}
\end{equation}

\begin{figure}[h!]
     \centering
         \includegraphics[width=\columnwidth]{figures/vibrato}
     \caption{Spectrogram of a sound with sinusoidal vibrato of $3Hz$ and one octave depth in a $1000Hz$ sawtooth wave (considering $f_a=44.1kHz$).}
         \label{fig:vibrato}
\end{figure}

For the properly realization of vibrato, it is important to pay attention on the following tables (quais tabelas? seriam termos?) and sequences.
\textcolor{red}{Table $\widetilde{M}_i$ with length $\widetilde{\Lambda}_M$ and the sequence with indices $\gamma_i'$ make the sequence $t_i'$ which is the oscillation pattern in the frequency while table $\widetilde{L}_i$ with length $\widetilde{\Lambda}$ and the sequence with indices $\gamma_i$ make $t_i$ which is the sound itself.
Variables $\mu$ and $\nu$ quantify vibrato intensity: $\mu$ is a direct measure of how many Hertz are involved in the upper limit of the oscillation while $\nu$ is the direct measure of how many semitones (or half steps) are involved in the oscillation ($2\nu$ is the number of semitones between the upper and lower pikes of frequency oscillations of the sound $\{t_i\}$ caused by the vibrato).
It is convenient to use $\nu=\log_{2}\frac{f+\mu}{f} $ in this case because the maximum frequency increase is not equivalent to the maximum diminish, but the semitones variations remains.}

Figure~\ref{fig:vibrato} illustrates the spectrogram of an artificial vibrato for a note with $1000Hz$ (between a \emph{B} and a \emph{C}), and which frequency deviation reaches one octave above and one below. Any waveform can be used to generate a sound and a vibrato oscillation pattern given any oscillation frequency and pitch deviation\footnote{The pitch deviation is called 'vibrato depth' and is generally given as semitones or cents, as convenience}.
Those oscillations with precise forms and arbitrary amplitudes are not possible in traditional music instruments and, in this way, it introduces novelty in the artistic possibilities.

Tremolo is similar: $f'$, $\gamma_i'$ and $t_i'$ remains the same.
The amplitude sequence to be multiplied by the original sequence $t_i$ turns in:

\begin{equation}\label{trA}
 a_i=10^{\frac{V_{dB}}{20}t_i' } = a_{\text{max}}^{t_i'}
\end{equation}

\begin{multline}\label{trT}
 T_i^{tr(f')}=\left \{ t_i^{tr(f')} \right \}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}= \\ =\left \{t_i .10^{t_i' \frac{V_{dB}}{20}}    \right \}_0^{\Lambda-1}=\left\{t_i . a_{\text{max}}^{t_i'} \right\}_0^{\Lambda-1}
\end{multline}

\noindent where $V_{dB}$ is the oscillation depth in decibels of tremolo and $a_{\text{max}}=10^{\frac{V_{dB}}{20}}$ is the maximum amplitude gain.
The measurement in decibels is pertinent because the maximum increase in amplitude is not equivalent to the related maximum decrease, while the difference in decibels remains.

Figure~\ref{fig:tremolo} shows the amplitude of sequences $\{a_i\}_0^{\Lambda-1}$ and $\{t_i'\}_0^{\Lambda-1}$ for three oscillations of a tremolo with a sawtooth waveform. The curvature is due to the logarithmic progression of the intensity. The tremolo frequency is $1,5Hz$ because $f_a=44,1kHz \; \Rightarrow \; \text{duration} = \frac{i_{\text{max}}=82000}{f_a}= 2s \; \Rightarrow \; \frac{3\text{oscillations}}{2s}=1,5$ oscillations for second ($Hz$).

The music piece \emph{Vibra e treme} explores those possibilities given by tremolos and vibratos, both used in conjunction and independently, with frequencies $f'$, different depths ($\nu$ and $V_{dB}$), and progressive parameters variations \footnote{Tremolos and vibratos occur many times together in a traditional music instrument and voices.}. Aiming a qualitative appreciation, the piece also develops a comparison between vibratos and tremolos in logarithmic and linear scales. Its source code is in the Appendix~\ref{ap:vibra} and available online as part of the \massa \emph{toolbox}.

\begin{figure}[h!]
     \centering
         \includegraphics[width=\columnwidth]{figures/tremolo}
     \caption{Tremolo with a depth of $V_{dB}=12dB$ with a sawtooth waveform as its oscillatory pattern with $f'=1.5Hz$ in a sine of $f=40Hz$ (considering a sample frequency of $f_a=44,1kHz$).}
         \label{fig:tremolo}
\end{figure}

With progressive increasing of $f'$ for a linear proximity of frequency aiming
to hear the phenomena as pitch ($\approx 20Hz$) it (o que?) generates roughness to both
tremolos and vibratos. Those roughness are largely appreciated both in
traditional classical music and current electronic music, specially in the
\emph{Dubstep} genre.  Roughness is also generated by spectral content that
produce beating~\cite{Porres,porres2009}. The sequence \emph{Bela Rugosi}
explores the threshold with concomitants of tremolos and vibratos at the same
voice, with different intensity and waveforms. The respective code is presented in
Appendix~\ref{ap:bela} and available online in \massa.

Increasing even more the frequency makes those oscillations no longer remains
noticeable. In this case, the oscillations are audible as pitch. Then, $f'$,
$\mu$ and the waveform changes together the frequency of original sound $T_i$
in different ways for both tremolos and vibratos.  They are called AM
(\emph{Amplitude Modulation}) and FM (\emph{Frequency Modulation}) synthesis,
respectively.  These techniques are well known, with applications in
synthesizers like \emph{Yamaha DX7}, and even with applications outside music, like
in telecommunications to data transferring by means of electromagnetic waves
(e.g.\ AM and FM radios).

For musical ends, it is possible to understand FM based on the
case of sines and to decompose the signals into their respective Fourier
components (i.e.\ sinusoidal) to more complex cases. In this way, the FM
synthesis performed with a sinusoidal vibrato with frequency $f'$ and depth
$\mu$ in a sinusoidal sound $T_i$ with frequency $f$ generates bands
centered in $f$ and far from each other with a distance of $f'$:

\begin{equation}\label{eq:fmEsp}
\begin{split}
\{t_i'\} & = \left \{ \cos \left [f . 2 \pi \frac{i}{f_a-1} + \mu . sen \left ( f' . 2 \pi \frac{i}{ f_a -1 } \right ) \right ] \right \} = \\
 & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ f . 2 \pi \frac{i}{f_a-1} + k . f' . 2 \pi \frac{i}{f_a-1} \right ]  \right \} = \\
 & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ (f+k.f') . 2 \pi \frac{i}{f_a-1} \right ]  \right \}
\end{split}
\end{equation}

\noindent where

\begin{multline}\label{eq:Bessel}
J_k(\mu) = \\ = \frac{2}{\pi} \int_0^{\frac{\pi}{2}}\left [ cos \left (\overline{k}\;\frac{\pi}{2} + \mu . \sin w \right ) . cos \left ( \overline{k}\;\frac{\pi}{2} + k . w \right ) \right ] dw \\  \overline{k} = k \% 2 \;\;,\;\; k \in \mathbb{N}
\end{multline}

\noindent is the Bessel function~\cite{BesselCCRMA,JOSFM} which specifies the
amplitude of each component in FM synthesis.

On those equations, the frequency variation introduced by $\{t_i'\}$ does not
respect the geometric progression that follows the pitch perfection but reflects the
equation~\ref{freqLinear}. Equations~\ref{vbrF} are used for FM and are described
in Appendix~\ref{cap:fmam}, where the spectral content of the FM synthesis is
calculated and obtained with oscillations in the logarithmic scale. In
fact, what is attractive in FM is its simple behavior (only with linear
variations in~\ref{eq:fmEsp}).

For the amplitude modulation (AM):

\newcommand{\OneColEqu}[1]{%
\end{multicols}%
\begin{twocolequfloat}%
\begin{equation}
\{t_i'\}_0^{\Lambda-1} =\{(1+a_i) . t_i\}_0^{\Lambda-1} = \left \{ \left [ 1+M.\sin \left ( f'.2\pi\frac{i}{f_a -1} \right ) \right] .P .\sin \left ( f.2\pi\frac{i}{f_a -1} \right ) \right \}_0^{\Lambda-1} = \\ 
                        =  \left\{P.\sin \left( f.2\pi\frac{i}{f_a -1}  \right ) +  \frac{P.M}{2} \left [ \sin \left( (f-f').2\pi\frac{i}{f_a -1}  \right )  + \sin \left( (f+f').2\pi\frac{i}{f_a -1}  \right ) \right ] \right \}_0^{\Lambda-1}
\end{equation}
\end{twocolequfloat}%
\begin{multicols}{2}%
}

%\begin{widetext}
%\begin{multline*}\label{eq:amEsp}
%\begin{split}
%\{t_i'\}_0^{\Lambda-1} =\{(1+a_i) . t_i\}_0^{\Lambda-1} = \left \{ \left [ 1+M.\sin \left ( f'.2\pi\frac{i}{f_a -1} \right ) \right] .P .\sin \left ( f.2\pi\frac{i}{f_a -1} \right ) \right \}_0^{\Lambda-1} = \\ 
%                        =  \left\{P.\sin \left( f.2\pi\frac{i}{f_a -1}  \right %) +  \frac{P.M}{2} \left [ \sin \left( (f-f').2\pi\frac{i}{f_a -1}  \right )  + \sin \left( (f+f').2\pi\frac{i}{f_a -1}  \right ) \right ] \right \}_0^{\Lambda-1}
%\end{split}
%\end{multline*}
%\end{widetext}

The resulting sound is the original one together with the
reproduction of its spectral content below and above the original frequency,
with a distance of $f'$ from $f$. Again, this is obtained by variations in the
linear scale of the amplitude. Appendix~\ref{cap:fmam} has an exposition of the
spectrum considering an AM performed with an oscillation in the amplitude logarithmic scale. It also loses the simple behavior.

The sequence $T_i$ with frequency $f$, called 'carrier signal', is modulated by
$f'$, called 'modulation signal'. Assuming the FM and AM jargon, $\mu$ and
$\alpha=10^{\frac{V_{dB}}{20}}$ are called 'modulation indexes'. The following
equations are defined for the vibration pattern of the modulation signal
$\{t_i'\}$:

\begin{equation}\label{fmGammaAux}
\gamma_i'=\left \lfloor i f' \frac{\widetilde{\Lambda}_M}{f_a} \right \rfloor
\end{equation}

\begin{equation}\label{fmAux}
t_i'=\widetilde{m}_{\gamma_i' \;\% \widetilde{\Lambda}_M}
\end{equation}

The modulation signal $\{t_i'\}$ into the carrier signal $\{t_i\}$ for FM is applied as:

\begin{equation}\label{fmF}
f_i=f + \mu . t_i'
\end{equation}

\begin{equation}\label{fmGamma}
\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}(f+\mu . t_j') \right\rfloor
\end{equation}

\begin{equation}\label{fmT}
T_i^{f,\, FM(f',\,\mu)}=\left\{ t_i^{f,\,FM(f',\,\mu)} \right\}_0^{\Lambda-1}=\left\{\,\widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \,\right\}_0^{\Lambda-1}
\end{equation}

\noindent where $\widetilde{l}$ is the waveform period with a length
$\widetilde{\Lambda}$ for the carrier signal.

To perform AM someone just needs to modulate the signal $\{t_i\}$ with $\{t_i'\}$ using
the following equations:

\begin{equation}\label{amA}
a_i=1 + \alpha . t_i'
\end{equation}

\begin{multline}\label{amT}
T_i^{f,\,AM(f',\,\alpha)}=\left\{ t_i^{f,\,AM(f',\,\alpha)} \right\}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}= \\ \{t_i . (1 + \alpha . t_i')    \}_0^{\Lambda-1}
\end{multline}

\subsection{Musical usage}\label{subsec:mus2}

At this point, the musical possibilities have been exploded. All characteristics like pitch (given by frequency), timbre (given by the waveform and filters), volume (given by intensity) and duration (given by the number of samples) can be considered by themselves or treated during their duration (with the exception of the duration itself).

The following musical usages comprehend a collection of possibilities
with the purpose of exemplifying types of sound manipulation that results in
musical material. Some of them are deeply discussed in the next section.

\subsubsection{Relations between characteristics}

Another interesting possibility is to use the relations between parameters of
tremolo and vibrato, and some parameters of the basic note like frequency. In
this way, it is possible to say that vibrato frequency is proportionally
different from pitch, or the tremolo depth is inversely proportional to
pitch. Therefore, with equations \ref{vbrGamma}, \ref{vbrF} and \ref{trA}, it is
possible to define:

\begin{equation}\label{eq:vinculos}
\begin{split}
f^{vbr} = f^{tr} & = func_a(f) \\
\nu & = func_b(f) \\
V_{dB} & = func_c(f)
\end{split}
\end{equation}

\noindent with $f^{vbr}$ and $f^{tr}$ as $f'$ in the referenced equations. They can also be associated with the vibrato and tremolo oscillation frequency in
equation~\ref{vbrGamma}.

Besides that, $\nu$ and $V_{dB}$ are the respective depth values of vibrato and
tremolo. Functions $func_a$, $func_b$ and $func_c$ are arbitrary and dependents
on the musical intentions. The music piece \emph{Tremolos, vibratos e a
frequência} explores such characteristics and shows variations in the oscillation
waveform with specific relations with the purpose of building a \emph{musical
language}\footnote{Details in the next section.}. The corresponding code is on
Appendix~\ref{ap:tremolos} and it is also available online as part of the \massa
toolbox.

About convolution it is possible to use a musical pulse as duration -- like a BPM
pulse -- and to distribute impulses during this original pulse, aiming at
establishing metrics and rhythms\footnote{It is important to remember that the
convolution with a impulse results in a sound shifted to the moment that the
impulse occurred}.
For example, two impulses equally spaced builds a binary division into the
pulse. Two signals, one with 2 pulses and another with 3 pulses, both with
equally spaced impulses in the pulse duration, results in the pulse
maintenance with a rhythmic mark that is possible to use as binary or ternary
divisions in many ethnic or traditions music styles~\cite{Gramani}. 
The absolute values of the impulses results in proportions among the
amplitudes of the convoluted signals.
Using convolution with impulses as a metric is explored in the
music piece \emph{Trenzinho de caipiras impulsivos}. The features
embrace the creation of 'sound amalgam' based on granular synthesis and this
piece provides a link to the next section. Refer specially to the
Figure~\ref{fig:pulsoSubAgl}. The source code of the music piece is included in the section~\ref{ap:trenzinho} and online as well (\massa toolbox).

\subsubsection{Moving audio source and receptor, Doppler effect}

Resuming the exposition in subsection~\ref{subsec:spac}: when an audio source
and a receptor are moving, their characteristics are ideally updated at each
sample of the digital signal. The speeds are decomposed in relation of each ear
direction. In this way, given the audio source speed (or velocity) $v_s$, that
is positive if the source moves away from receptor, and receptor speed $v_r$,
that is positive when it gets closer of the audio source, the frequency is given
by the well known Doppler effect:

\begin{equation}\label{eq:fDoppler}
    f=\left(\frac{v_{sound}+v_r}{v_{sound}+v_s}\right)f_0
\end{equation}

With this frequency and relations given by the new IID from the new source
position, it is possible to create the Doppler effect. There is an addendum to
improve the fidelity of the physical phenomena: to increase the received
potency. It is possible to understand this potency gain as being proportional to the relative speed that, in each second, adds the traversed period, by means of a waveform with potency: $\Delta P=P_0\left(\frac{v_r-v_s}{343,2}\right)$, where $P_0$ represents the signal potency.

In this way it is possible to obtain both amplitude and frequency of a moving
audio source. Being this audio source in front of receptor with $y_0$ meters of horizontal distance and $z_0$ meters of height, the distance is given by
$D_i=\left\{ d_i=\sqrt{ y_{i}^{2}+z_{0}^{2} } \right\}_0^{\Lambda-1}$,
where $y_i=y_0+v_s-v_r$ if considering $v_s$ and $v_r$ both in
horizon. Amplitude changes with the distance and with the potency factor cited
above (see subsection~\ref{subsec:volume} for potency to amplitude conversion).

\begin{equation}\label{eq:aDoppler}
    A_i=\left\{ \frac{z_0}{d_i}A_{\Delta P}\right\}_0^{\Lambda-1} = \left\{ \frac{z_0}{\sqrt{y_i^2+z_0^2}} \sqrt{\frac{v_r-v_s}{343,2}+1}  \,\right\}_0^{\Lambda-1}
\end{equation}

Observe that the factor which changes the amplitude (caused by the distance) is even, while the factor caused by the potency variation is antisymmetric in
relation to the crossing of source with receptor. The frequency has a symmetric progression in relation to pitch. In other words, the same semitones (or fractions) added during the approach are decreased during the departure. Besides that, the transition becomes steep if both source and receptor intersects each other exactly at the same point, otherwise, there is a monotonic progression. In the given case, where there is a static height $z_0$, it is necessary to observe the speed component in the direction between the observer and audio source:

\begin{equation}\label{eq:ffDoppler}
    F_i=\{f_i\}_0^{\Lambda-1}=\left\{\frac{v_{sound} + v_r\frac{y_i}{\sqrt{z_0^2+y_i^2}}}{v_{sound}+v_s\frac{y_i}{\sqrt{z_0^2+y_i^2}}}f_0\right\}_0^{\Lambda-1}
\end{equation}

In the Appendix~\ref{sec:cod2} there is a Python implementation of the Doppler
effect as describe above, also considering the intersection between audio source and receptor.

\subsubsection{Filter and noises (subsections~\ref{subsec:ruidos} and~\ref{subsec:filtros})}

Using filters the possibilities are many. It is possible to
convolve a signal to reverberate it, to remove its noise, to generate
distortions or to treat the audio with aesthetically manipulation in mind. For
example, it is possible to simulate sounds from and old television if a
band-pass filter is applied to accept just sound between $1kHz$ and $3kHz$. Or
if someone removes (with some precision) just the frequency of electric
oscillation (usually $50Hz$ or $60Hz$) and the harmonics, it will remove noises
caused by audio devices. A more musical application is to perform filtering in
specific bands and to use those bands as an additional parameter to the notes.

Inspired by traditional music instruments, it is possible to apply a
time-dependent filter~\cite{Roederer}. Chains of those filters can perform
complex and more accurate filtering routines. The music piece \emph{Ruidosa
faixa} explores those features by using filters and many kinds of noise
synthesis. The source code is in Appendix~\ref{ap:ruidosa} and is available
online as part of \massa.

When used together the features can create an effect known
as \emph{chorus}. Based on what happens in a singers choir, in this
effect the sound is performed using small and potentially arbitrary
modifications of parameters like center frequency, presence (or absence) of
vibrato or tremolo and its characteristics, equalization, volume, etc. As a
final result, those versions of the original sound are mixed together (see
equation~\ref{eq:mixagem}). The music piece \emph{Chorus infantil} implements a
chorus in many ways with different sounds and its source code is listed in
Appendix~\ref{ap:chorus}. The script is also available in the \massa toolbox.

\subsubsection{Reverberation}\label{subsubsec:reverb}

Using the same spatialization terms of subsection~\ref{subsec:spac}, the
late reverberation can be modeled as a convolution with a period of pink, brown
or black noise, with exponential decay of amplitude and time related. In
this way, the treble attenuation and irregular smooth in the frequency response
are contemplated with great success. Delay lines can be added as prefix to noise
with the decay, and this contemplates both time parts of the reverberation: the
first reflections and the late reverberation. It is possible to improve the ??
quality by calculating the geometric localization of the last surface where each
wave front reflected before reach the ear in the fairest $100-200$ milliseconds. Then, one could apply a low-pass filter as described in subsection~\ref{subsec:filtros}. The colored
noise can be gradually introduced since the initial moment gives direct
incidence (i.e.\ without any reflection and given by ITD and IID),
with \emph{fade-in}, reaching its maximum at the beginning of the 'late
reverberation', when the geometric incidences loses their importance to the
statistics properties of the noise decay.

As an example, consider $\Delta_1$ as the duration of the first period and $\Delta_R$ as the duration of total reverberation ($\Lambda_1=\Delta_1 f_a$, $\Lambda_R=\Delta_R
f_a$). It is possible to add a probability $p_i$ of a sound reincidencing in the
$i$-th sample with amplitude with exponential decay. Following
subsection~\ref{subsec:spac}, the reverberation $R_i^1$ of the first period can
be described as:

\begin{multline}\label{eq:p1rev}
    R_i^1=\left\{r_i^1\right\}_0^{\Lambda_1-1}\;: \\ \;r_i^1=\left\{
        \begin{array}{l l}
            10^{\frac{V_{dB}}{20}\frac{i}{\Lambda_R-1}}\;  & \text{with probability}\quad p_i=\left(\frac{i}{\Lambda_1}\right)^2 \\
                                     0 \; & \text{with probability}\quad 1-p_i \\
        \end{array} \right.
\end{multline}

\noindent where $V_{dB}$ is the total decay in decibels, typically $-80dB$ or
        $-120dB$. Reverberation $R_i^2$ of the second period can be emulated by a
        brown noise $R_i^m$ (or by a pink noise $R_i^r$) with exponential decay:

\begin{equation}\label{eq:p2rev}
    R_i^2=\left\{r_i^2\right\}_{\Lambda_1}^{\Lambda_R-1}=\left\{10^{\frac{V_{dB}}{20}\frac{i}{\Lambda_R-1}}\,.\,r_i^m\right\}_{\Lambda_1}^{\Lambda_R-1}
\end{equation}

\noindent like:

\begin{equation}\label{eq:rev}
    R_i=\left\{r_i\right\}_0^{\Lambda_R-1}\;:\;r_i=\left\{
        \begin{array}{l l}
            1\; & \text{if }\quad i=0 \\
            r_i^1\;  & \text{if }\quad 1\leq i<\Lambda_1-1 \\
                                     r_i^2 \; & \text{se}\quad \Lambda_1 \leq i < \Lambda_R-1 \\
        \end{array} \right.
\end{equation}

\noindent the reverberation represented by $R_i$ can be applied as simple
        convolution of $R_i$ (called reverberation 'impulse response') with the sound sequence $T_i$ as described in subsection~\ref{subsec:filtros}.

Reverberation is well known by causing great interest in the listeners and to provide more enjoyable sonorities. Besides that, modifications in the reverb
space consists of a tricking (almost a \textit{clich\'{e}}) to induce surprise and interest in the listener.

\subsubsection{ADSR Envelopes}

The volume variation along the duration of sound is crucial to our timbre
perception. The volume envelope, known as ADSR (\emph{Attack-Decay-Sustain-Release}), has many implementations in both hardware and software synthesizers. A pioneer implementation can be found in the Hammond
Novachord synthesizer of 1938 and some variants are cited below~\cite{ADSR}.

The stochastic ADSR envelope is characterized by 4 parameters: attack duration
(time to sound reaches its maximum volume), decay duration (follows the attack
immediately), level of sustain volume (in which the volume remains stable after the decay) and release duration (after sustain, this is the duration until the volume decays to zero). Note that the sustain duration is not specified because it is the difference between the duration itself and the durations of attack, decay and sustain.

It is possible to apply the ADSR envelope with durations $\Delta_A$, $\Delta_D$ and $\Delta_R$, with total duration $\Delta$ and sustain level $a_S$, given by the fraction of the maximum amplitude, as a sound sequence $t_i$ defined as:

\begin{multline}\label{eq:adsr}
%\begin{split}
\{a_i\}_0^{\Lambda_A-1}  = \left\{\xi\left(\frac{1}{\xi}\right)^{\frac{i}{\Lambda_A-1}}\right\}_0^{\Lambda_A-1} \;\; \text{or} \;\;\quad \left\{\frac{i}{\Lambda_A-1}\right\}_0^{\Lambda_A}\\
\{a_i\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} =\left\{a_S^{\frac{i-\Lambda_A}{\Lambda_D-1}}  \right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} \;\;\quad \quad\quad \\ \text{or} \quad\;\; \left\{1-(1-a_S)\frac{i-\Lambda_A}{\Lambda_D-1}\right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1}\\
\{ a_i \}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} =\left\{ a_S \right\}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} \\
\{ a_i \}_{\Lambda-\Lambda_R}^{\Lambda-1}  =\left\{ a_S\left(\frac{\xi}{a_S} \right)^{\frac{i-(\Lambda-\Lambda_R)}{\Lambda_R-1}} \right\}_{\Lambda-\Lambda_R}^{\Lambda-1} \quad\;\; \\ \text{or} \quad\;\; \left\{ a_S - a_S\frac{i+\Lambda_R-\Lambda}{\Lambda_R-1}\right\}_{\Lambda-\Lambda_R}^{\Lambda-1} 
%\end{split}
\end{multline}

\noindent with $\Lambda_X=\lfloor \Delta . f_a \rfloor\;\;\forall\;\; X \; \in
(A,D,R,\;)$ and being $\xi$ a small value that provides a satisfactory \emph{fade in} and \emph{fade out}, e.g.\ $\xi=10^{\frac{-80}{20}}=10^{-4}\;$ or $\;\xi=10^{\frac{-40}{20}}=10^{-2}$. The lower the $\xi$, the slower the \emph{fade}, like the $\alpha$ illustrated in Figure~\ref{fig:transicao}. The terms in the right side of the equation~\ref{eq:adsr} attend for both introduction and ending of sound from the zero intensity because they are linear. Schematically, Figure~\ref{fig:adsr} shows the ADSR envelope, a classical implementation that supports many variations. For example, between attack and decay it is possible to add an extra partition where the maximum amplitude stays. Another common example is the use of more elaborated tracings to attack or decay. The music piece \emph{ADa e SaRa}, available both in
Appendix~\ref{ap:ada} and in \massa, explores many configurations of the ADSL envelope.

\begin{equation}\label{eq:adsrApl}
\left\{t_i^{ADSR}\right\}_0^{\Lambda-1} =\{t_i . a_i\}_0^{\Lambda-1}
\end{equation}

\begin{figure}[htpq!]
    \centering
        \includegraphics[width=\columnwidth]{figures/adsr}
    \caption{An ADSR envelope (\emph{Attack, Decay, Sustain, Release}) applied
        to an arbitrary sound sequence. The linear variation of the amplitude is
        above. Below the amplitude variation is exponential.}
        \label{fig:adsr}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\afterpage{\blankpage}

%\clearpage

\section{Notes organization in music}\label{notasMusica} \label{sec:notesMusic}

Consider $ S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1} $ as the sequence $S_j$ of $H$ musical events $s_j$. In addition, $S_j$, called a 'musical structure', as composed by events $s_j$ which are musical structure themselves, e.g.\ notes. This section is dedicated to techniques that make $S_j$ interesting and enjoyable for audition.

The elements of $S_j$ can be overlapped by mixing them together as in
equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, building intervals
and chords. This scenery reflects the 'vertical thought' in music. On the other hand, the concatenation of $S_j$
as in equation~\ref{eq:concatenacao} and in Figure ~\ref{fig:concatenacao},
builds melodic sequences and rhythms, which are associated with the musical
'horizontal thought' in music. The fundamental frequency $f$ and the beginning moment
(attack) are generally the most important characteristics of the elements in
$S_j$. Such characteristics make possible the creation of music made by pitches (both
harmony and melody) and by the presence of temporal metrics and rhythms.

\subsection{Tuning, intervals, scales and chords}\label{subsec:afinacao}

Here we describe characteristics in music related to harmony and melody.

\subsubsection{Tuning}

Doubling the frequency is equivalent to ascending one octave ($f=2f_0$).
The octave division in twelve notes is the canon of classical western music, although its usage has also been observed
outside western tradition, like ceremonial/religious and ethnic context~\cite{Wisnick}.

The factor given by $\varepsilon=2^{\frac{1}{12}}$ defines a semitone. It builds a note
grid along the hearing spectrum. Fixing a frequency $f$, all the possible
fundamental frequencies stay separated by intervals that are multiples of
$\varepsilon$. Twelve semitones (or half steps), equidistant to the human ear, make an
octave. Therefore, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
$f_0$ and $f$.

This absolute accuracy is common in computational implementations. For the real
music instruments, however, there are deviations of those frequencies in order to make
the harmonics of corresponding notes more compatible. Besides that, the fixed reference
$\varepsilon=2^{\frac{1}{12}}$ characterizes an equally tempered tuning. There
are tunings which respective intervals are ratios of low-order integers, originally based
on observations of physical behaviors. First of these tunings were formalized
around two thousand years before the existence of an equal temperament~\cite{Roederer}.

Two iconic tunings:

\begin{itemize}
    \item The {\bf just intonation} is defined as notes of the diatonic scale
    that are associated with ratios of low-order integers that defines the harmonic
    series. The basic ratios are the same in the Ionian mode (white piano keys
    from \do to \do (see below in subsection~\ref{subsec:escalas}): 1, 9/8, 5/4,
    4/3, 3/2, 5/3, 15/8, 2/1). 
    The intervals are considered as notes in a scale together with the following: the
    semitone 16/15, the 'minor tone' 10/9, and the 'major tone' 9/8. There are
    many ways to perform the division of the 12 notes.

    \item The {\bf Pythagorean tuning} is based on the interval 3/2 (perfect
    fifth). The Ionian mode becomes: 1, 9/8, 81/64, 4/3, 3/2, 27/16, 243/128,
    2/1. The intervals are also considered as notes in a scale. Besides the
    intervals in the mode, it is also used the minor second 256/243, the minor
    third 32/27, the augmented fourth 729/512, the diminished fifth 1024/729,
    the minor sixth 128/81 and the minor seventh 16/9. 
\end{itemize}

In order to attends for micro-tonality\footnote{Micro-tonality is the use of intervals less
than one semitone and have ornamental and structuring functionalities in music. The
division of the octave in $12$ notes is motivated by physical fundamentals but it is also
a \emph{convention} adopted even by western classical music. Other tunings are
present. To cite one example, the traditional Thai music uses a octave
division in seven notes equally spaced ($\varepsilon=2^{\frac{1}{7}}$),
resulting in intervals that have little in common with the intervals in the
diatonic scale~\cite{Wisnick}}. It is possible to use non-integer real values as
pitch sequences or even to maintain the integer values and change the factor $\varepsilon=2^{\frac{1}{12}}$. For example, 
a tuning really near of the harmonic series
itself is proposed with the division of the octave in $53$ notes:
$\varepsilon_2=2^{\frac{1}{53}}$~\cite{microtonalidade}.
In that division the notes are related by means of integers with
$\varepsilon_2$. Note that if $S_i$ is a pitch sequence related by means of
$\varepsilon_1$, a mapping for all the notes related by $\varepsilon_2$ defines
a new sequence $S_i'=\left\{s_i'\right\}=\left\{
s_i \frac{\varepsilon_1}{\varepsilon_2}\right\}$. The music piece \emph{Micro
tom} explores this micro-tonal features and its code is in
Appendix~\ref{ap:micro} and is part of \massa.

\subsubsection{Intervals}\label{subsec:intervalos}

Using the ratio $\varepsilon=2^{\frac{1}{12}}$ between the note frequencies
(i.e.\ one semitone) the intervals in the twelve note system can be represented by integers. Table~\ref{eq:intervalos} summarizes the characteristics of each interval: its traditional nomenclature, consonance and
dissonance properties, and number of semitones per interval.

\begin{table*}[htpq!]
\centering
\caption{The music intervals together with their traditional notation, basic classification for dissonance and consonance, and number of semitones. Perfect (P) unison, fifths and octaves are the perfect consonances. Major (M) and minor (m) thirds and sixths are the imperfect consonances. Minor seconds and major sevenths are the harsh dissonances. Major seconds and minor sevenths are the mild dissonances. Perfect fourth is the first special case, being a perfect consonance when considering it as the inversion of the perfect fifth and a dissonance or a imperfect consonance otherwise. The second special case is the tritone (A4 or aug4, d5 or dim5, tri,TT). This interval is consonant in some cultures. For tonal music, the tritone indicates dominant and searches for its resolution in a third or sixth. Due to this instability it is considered a dissonance interval.}
\begin{tabular}{| c | c | c | }\hline
    \multicolumn{3}{|c|}{\bf consonances}  \\\hline
   & traditional notation & number of semitones \\
   perfects: & P1, P5, P8 & 0, 7, 12 \\
 imperfects: & m3, M3, m6, M6 & 3, 4, 8, 9 \\\hline\hline
    \multicolumn{3}{|c|}{\bf dissonances} \\\hline
 & traditional notation & number of semitones \\
 fortes: & m2, M7 & 1, 11 \\
 brandas: & M2, m7 & 2, 10 \\\hline\hline
    \multicolumn{3}{|c|}{\bf special cases} \\\hline
 & traditional notation & number of semitones \\
 consonance or dissonance: & P4 & P5 \\
 dissonance in Western tradition: & tritone, aug4, dim5 & 6 \\\hline
\end{tabular}\label{eq:intervalos}
\end{table*}

The nomenclature is based both in impositions and for convenience of the tonal system and practical aspects of note manipulation and can be specified
as~\cite{Roederer,Wisnick}:

\begin{itemize}
                \item Intervals as number of ratios between notes: first
                (unison), second, third, forth, fifth, sixth, seventh and
                eighth. The ninth, tenth, eleventh, etc, are compound                       intervals made by one or more octaves + a interval inside the                 octave, which characterizes the compound interval. The                        intervals are represented by the numeric digits, e.g. 1, 3, 5                 are a unison, a third and a fifth, respectively.

                \item Qualities of each interval: the perfect consonances --
                i.e.\ unison, forth, fifth and octave -- are 'perfect'. The
                imperfect consonances -- i.e.\ thirds and sixths -- and
                dissonances -- i.e.\ seconds and sevenths -- can be major and
                minor. Exception for the tritone.

                \item The perfect fourth is both perfect consonance or
                dissonance according to the context and theoretical
                background. As general rule, it can be considered as                        consonance except when it is used as passage to resolve a                   fifth or a third.

                \item Tritone is dissonance in Western music because
                it characterizes the ``dominant'' in the tonal system (see
                subsection~\ref{subsec:harmonia}) and represents the
                instability. Some cultures chant the interval as consonance.

                \item A major interval decreased by one semitone results in a
                minor interval. A minor interval increased by one semitone
                results in a major interval.

                \item A perfect interval (unison, perfect forth, perfect                    fifth, perfect octave), or a major interval (major second M2,                 major third M3, major sixth M6 or major seventh M7), increased                 by one semitone results in an augmented interval (e.g.\                    augmented third aug3 with five semitones). The augmented forth                 is also called tritone (aug4 ~ tri ~ TT).

                \item A perfect interval or a minor interval (minor second m2,
                minor third m3, minor sixth m6 or minor seventh m7), decreased
                by one semitone results in a diminished interval. The                       diminished fifth is also called tritone (dim5 ~ tri ~ TT).

                \item An augmented interval increased by one semitone results                 in a 'doubly augmented' interval and a diminished interval
                decreased by one semitone results in a 'doubly diminished'
                interval.

                \item Notes played simultaneously configure a harmonic
                interval.

                \item Notes played as a sequence in time configure a
                melodic interval. The order of the notes: considering first                 the lowest note or the highest note, results in an ascending                 or descending interval, respectively.

                \item If the lower pitch is raised one octave, or if
                the upper pitch is lowered one octave below, the interval is
                inverted. When summed to its inversion an interval results in                 9 (e.g.\ m7 is inverted to M2: $m7+M2=9$). An inverted major
                interval results in a minor interval and vice-versa. An                     inverted augmented interval results in a diminished interval                 and vice-versa (like a doubly augmented results in a doubly
                diminished and vice-verse). An inverted perfect interval                    results in a perfect interval as well.

                \item An interval higher than an octave is called a 'compound
                interval' and is classified like the interval between the same
                notes but in the same octave. They are also specified adding a                 7 to the interval: P11 is an octave plus a forth ($7 + P4 =                  P11$), M9 is an octave plus a major second ($7 + M2 = M9$).
\end{itemize}

The augmented/diminished intervals and the doubly augmented/doubly diminished intervals are consequences of the tonal system. The scale degrees (subsection~\ref{subsec:escalas}) generally determine different pitches, with
specific uses and functions. Henceforth, in a \textit{C flat} major scale, the tonic -- first degree -- is \textit{C flat}, not \textit{B}, and the leading tone -- seventh degree -- is \textit{B flat}, not \textit{A sharp} or \textit{C double flat}. In a similar way, the second degree of a scale can be one semitone far from the first degree like the leading tone (seventh degree at one ascending semitone from the first degree), where there is a diminished third between the two semitones of the seventh and second scale degrees, consequence of the first degree be between the near two degrees: the second and the leading note~\cite{Lacerda}.

The descriptions presented here summarizes the traditional theory of the music
intervals~\cite{Lacerda}. The music piece \emph{Intervalos entre alturas}
explores these intervals in a independently and conjunction manner. The reader is invited to see the source code in Appendix~\ref{ap:intervalos} and available online in the \massa toolbox~\cite{MASSA}.

\subsubsection{Scales}\label{subsec:escalas}

Basically speaking, scale is a set of ordered pitches. Usually, scales repeat at each octave. The ascending sequence with all notes from the octave division in 12 equal intervals, separated by ratio $\varepsilon=2^{\frac{1}{12}}$, is known as the chromatic scale with equal temperament. There are 5 perfect symmetric divisions of the octave within the chromatic scale. These divisions are considered as scales by the easily and peculiar use they provide. Considering the integers that the factor $\varepsilon=2^{\frac{1}{12}}$ is powered to multiply $f_0$, the scales are the following:

\begin{multline}\label{escSim}
%\begin{split}
\text{chromatic}  = E_i^c = \\ = \{e_i^c\}_0^{11} =  \{0,1,2,3,4,5,6,7,8,9,10,11\} = \{i\}_0^{11}\\
\text{whole tones}  = E_i^t = \{e_i^t\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
\text{minor thirds}  = E_i^{tm} = \{e_i^{tm}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
\text{major thirds}  = E_i^{tM} = \{e_i^{tM}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
\text{tritones}  = E_i^{tt} = \{e_i^{tt}\}_0^{1} = \{ 0, 6 \} = \{6.i\}_0^1
%\end{split}
\end{multline}

Therefore, the third note of the whole tone scale with $f_0=200Hz$
is $f_3=\varepsilon^{e_3^t} . f_0 = 2^{\frac{6}{12}} . 200 \approxeq 282.843
Hz$. These 'scales' or patterns generate stable structures by their intern
symmetries and can be repeated in a efficient and sustained way. Section~\ref{estCic} specially discusses the symmetries. 
The music piece \emph{Cristais} exposes each one of these scales, in both melodic and
harmonic counterpart and the corresponding source code is presented in Appendix~\ref{ap:cristais} and as
part of \massa.

The diatonic scales are:

\begin{multline}\label{eq:escalas}
%\begin{split}
\text{natural minor} = \text{aeolian mode}  = \\ = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
\text{locrian mode}  = E_i^{mlo} = \{e_i^{mlo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
\text{major}  = \text{ionian mode}  =  E_i^M = \{e_i^M\}_0^6 = \{0,2,4,5,7,9,11\} \\
\text{dorian mode}  = E_i^{md} = \{e_i^{md}\}_0^6 = \{0,2,3,5,7,9,10\} \\
\text{phrygian mode}  = E_i^{mf} = \{e_i^{mf}\}_0^6 = \{0,1,3,5,7,8,10\} \\
\text{lydian mode}  = E_i^{ml}=\{e_i^{ml}\}_0^6 = \{0,2,4,6,7,9,11\} \\
\text{mixolydian mode} = E_i^{mmi} = \{e_i^{mmi}\}_0^6 = \{0,2,4,5,7,9,10\}
%\end{split}
\end{multline}

\noindent They have only the major, minor and perfect intervals. The unique exception is the tritone that is presented as a augmented forth or diminished fifth.

All the diatonic scales follow a pattern of successive intervals \textit{tone,
tone, semitone, tone, tone, tone, semitone}. Thus, it is possible to write:

\begin{equation}\label{eq:relacaoDia}
\begin{split}
\{d_i\} & =\{2,2,1,2,2,2,1\} \\
e_0 & =0 \\
e_i & =d_{(i+\kappa)\%7}+e_{i-1} \quad for \;\;  i > 0
\end{split}
\end{equation}

\noindent with $\kappa \in \mathbb{N}$. For each mode there is only one value
for $\kappa \in [0,6]$. For example, a brief inspection reveals that
$e_i^{ml}=d_{(i+2)\%7}+e_{i-1}^{ml}$. Then, $\kappa=2$ for the lydian mode.

The minor scales has two additional forms, named melodic and harmonic:

\begin{multline}\label{eq:escalasMenores}
%\begin{split}
\text{natural minor (same as above)}  = \\ = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
\text{harmonic minor}  = E_i^{mh} = \{e_i^{mh}\}_0^6 = \{0,2,3,5,7,8,11\} \\
\text{melodic minor}  = \\ = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
%\end{split}
\end{multline}

The ascending and descending contour of the melodic minor scale is necessary for the existence of the leading tone (seventh and last degree, separated by one semitone of the octave, enhances the tonic polarization) in the ascending trajectory, which is not necessary in the descending mode, since it recovers the normal form. The harmonic scale presents the leading tone but does not avoid the augmented second interval between the sixth and seventh degrees. In addition, it does not need to cover a melodic trajectory, just having to present the leading tone, crucial to the tonal system (the leading tone tends to the tonic, asserting it)~\cite{Harmonia}. 
Other scales can be represented in a similar form, like the pentatonic and the
modes of limited transposition by Messiaen~\cite{Messiaen}.

\subsubsection{Chords}\label{subsec:acordes}

The simultaneous occurrence of three or more notes is observed by means of chords. Their
base of tonal music are the triads. Triads are built by two successive thirds
within 3 notes: root, third and fifth. If the lower note of a chord is different from the root, this chord is referred
as inverted. A close position is where any note of chord fits between two consecutive
notes~\cite{Lacerda}. All the non-inverted triads in the close position form and with root in
$0$ are described as following:

\begin{equation}\label{triades}
\begin{split}
\text{major triad} = A_i^M= \{a_i^M\}_0^2=\{0,4,7\} \\ 
\text{minor triad} = A_i^m = \{a_i^m\}_0^2=\{0,3,7\} \\
\text{diminished triad} = A_i^d = \{a_i^d\}_0^2=\{0,3,6\} \\
\text{augmented triad} = A_i^a = \{a_i^a\}_0^2=\{0,4,8\}
\end{split}
\end{equation}

To consider another third superimposed to the fifth, is sufficient to add $10$
as the highest note in order to form a tetrad with minor seventh, or add $11$ in order to form a tetrad with major
seventh. The inversions and open positions can be obtained with a selective
addition of $12$ to the components.

Incomplete triadic chords, with extra notes ('dirty' chords), and non-triadic
are also common.

In the following we present general recommendations:

\begin{itemize}
        \item A fifth is a root (fundamental) confirmed by the interval.
        \item The major or minor third points to the major or minor quality
        of the chord.
        \item Every tritone, specially if built between a major third and a
        minor seventh, tends to resolve in a third or sixth.
        \item Note duplication is avoided. If duplication is needed, the
        preferred order is: the root, fifth, third and seventh.
        \item It is possible to build chords with notes different from triads,
        specially if they respect to a recurrent logic or musical chain that
        justifies these different notes.
        \item Chords built by successive intervals different of thirds -- like
        fourths and seconds -- are recurrent in compositions with advanced
        tonalism or experimental character.
        \item The repetition of chords successions (or their characteristics)
        fixes a trajectory by means of the recurrence and make possible the
        introduction of exotic formations without incoherence.
\end{itemize}

\subsection{Atonal and tonal harmonies, harmonic expansion and modulation}\label{subsec:harmonia}

Omission of the basic chaining is the key to obtain modal and atonal harmonies. In
case of absence of these minimal tonal structures, we say that the
harmony is modal if the notes match with some diatonic scale (see
equations~\ref{eq:escalas}) or if the notes are presented in a small number.
If basic tonal chaining are absent, the notes do not match any diatonic scale
and are diverse and dissonant (by relation with each other) enough to avoid
reduction by polarization, the harmony is atonal. In this classification, the
modal harmony are not tonal neither atonal. The modal harmony is reduced to the 
incidence of notes within the diatonic scale (or simplifications) and to the
absence of tonal structures. It is possible to note, following this concept, that
atonal harmony is hard to be realized~\cite{harmEXT}.

\subsubsection{Atonal harmony}

In fact, the techniques around atonal music comprehend structures for avoiding
the hearing relation with modes and tonality. This difficulty imposes the use of dodecafonism. 
The purpose of dedocafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
order (não sei se ficou mt claro). In this context, the tonic becomes difficult to settle. Nevertheless, the
Western hearing searches for tonal traces in musics and easily finds them by
unexpected paths and even tortuous.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The use of dissonant intervals (specially tritones without resolution), together
with ninths, seconds and sevenths, reinforces the absence of tonality. Considering
this context, it is possible to account for the following considerations while creating a music
piece:

\begin{itemize}
     \item Repeating notes. Considering the immediate repetition as an extension
     of the previous incidence, the use of same notes in sequence does not add relevant information.
     \item It is interesting to play adjacent notes at the same time, making harmonic intervals
     and chords.
     \item Present durations with liberty, respecting the notes order.
     \item In order to gain variation, one can use extension, transposition, translation, inversion, retrograde and retrograde inversion. See subsections~\ref{subsec:motivos}
     and~\ref{subsec:usosmusicais3} for more details.
     \item Account for the presence of note structures, it is possible to use variations in orchestration, articulation, spatialization, among others.
\end{itemize}

The atonal harmony can be observed, paradigmatically, within these presented
conditions. Most of them were written by great dodecafonic composers like Alban Berg and even Schoenberg. Many pieces create purposeful links between tonal and atonal techniques.

\subsubsection{Tonal harmony}

In the XX century, rhythmic music with emphasis on sonorities/timbres, extended
the concepts about tonality and harmony. Nevertheless, the
tonal harmony has strong influence in the artistic movements and commercial
venues. In addition, the dodecafonism is considered having tonal nature because it
was created to deny the tonal characteristics of polarization.

In tonal or modal music, the base of the harmonic filed are chords -- like the ones listed in
equations~\ref{triades}-- built with a root note for each degree of a scale -- like
the ones in equation~\ref{eq:escalas}. In this context, music harmony aims at studying the observation of incident chord progressions and chaining rules. Even a monophonic melody generates harmonic fields, making it possible to observe the suggested chords at each passage.

In the 'traditional tonal harmony', a scale has its corresponding tonic (first scale
degree) and can be major (with the same notes of the Ionian mode) or
minor (notes from Eolian mode form the 'natural minor', which has both a
harmonic and a melodic version, see equation~\ref{eq:escalasMenores}). The
scale is base on triads, each one with its root and representing a degree into
the scale: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$. 
To build triads, the third and the fifth note above the root is considered together with the root (or fundamental). It is possible to write $\hat{1},\hat{3},\hat{5}$ as the first degree chord, built by the first degree of scale and the center for a tonal music. The chords
of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when a minor
scale) and of the forth degree $\hat{4},\hat{6},\hat{1}$ are secondary. After
that, other degrees are considered. The traditional harmony comprises conventions and stylistic techniques to create chord chains, made in each scale degree~\cite{Harmonia}.

The 'functional harmony' ascribes functions to these three central chords and
tries to understand their use by means of these functions. The chord built under
the first degree is named the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its (role) function consisting of preserving a center, usually referred as a ``ground'' for the music. The chord built under the fifth degree is the \textbf{dominant} (\textit{D}, the dominant is always major) and its function is call for the tonic (we say that the dominant chord asks for a conclusion and this conclusion is usually the tonic chord). Thus, the dominant chord guides the music to the tonic. The triad built under the forth degree is the \textbf{subdominant} (\textit{S} or \textit{s} for a major or minor
subdominant, respectively) and its basically function is to distance the music from the tonic. The process aims at confirming the tonic using tonic-subdominant-tonic chains which are
expanded by using other chords in different ways.

The remaining triads are associated to these three most important chords. In the major scale, the associated relative (relative tonic \textit{Tr}, relative
subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built with a
third below, and the associated counter-relative (counter-relative
tonic \textit{Ta}, counter-relative subdominant \textit{Sa} and the
counter-relative dominant \textit{Da}) is the triad built with a third above. In the
minor scale the same happens, but the triad with a third below is called
counter-relative (tA, sA) and the triad with one third above is called relative (tR,
sR). The precise functions and musical effects of these chords are
controversial. Table~\ref{tab:harmonia} shows the relation between the triads
built in each degree of the major scale.

\begin{table}[htpq!]
\centering
\caption{Summary of tonal harmonic functions of the major scale. Tonic is the
music center, the dominant goes to the tonic and the subdominant moves the music away from
the tonic. The three chords can, at first, be freely replaced by their
respective relative and counter-relatives.}
\begin{tabular}{l | c | r}
relative & main chord of the function & counter-relative \\\hline\hline
$\hat{6},\hat{1},\hat{3}$ & tonic:       $\hat{1},\hat{3},\hat{5}$ & $\hat{3}, \hat{5},      \hat{7}$ \\
$\hat{3},\hat{5},\hat{7}$ & dominant:    $\hat{5},\hat{7},\hat{2}$ & [ $\hat{7},\hat{2},\hat{4}\#$ ] \\
$\hat{2},\hat{4},\hat{6}$ & subdominant: $\hat{4},\hat{6},\hat{1}$ & $\hat{6},\hat{1},       \hat{3}$
\end{tabular}
\label{tab:harmonia}
\end{table}

The dominant counter-relative should form a minor chord. It explains the change
in the forth degree by a semitone above $\hat{7}\#$. The diminished chord
$\hat{7},\hat{2},\hat{4}$, is generally considered a 'dominant seventh chord
with no root'~\cite{Koellheuteur}.
In the minor mode, there is the change in $\hat{7}$ by an ascending semitone to
make possible the existence of an unique semitone separating the tonic, and
making also possible the dominant (that should be major and goes to the tonic). In
this way, the dominant is always major, for both major and minor scale and,
because of that, even as a minor tone, the relative dominant remains a third
below and in the counter-relative, a third above.

\subsubsection{Tonal expansion: individual functions and chromatic mediants}

%% dominante e subdominante individualmente, ou, dominante e subdominante individual?

Each one of these chords can be confirmed and developed by performing their
dominant or individual subdominant, which is the chord based in the third formed
by a fifth above or by a fifth below, respectively. These dominants and individual
subdominants, in the same way, have also subdominants and individual dominants that
can be used. Given a tonality, any chord can occur, no matter
how distant it is from the harmonic field and from the notes within the
scale. The unique on condition is that the occurrence presents a coherent trajectory of
dominants and subdominants to the original tonality.

Mediants, or 'chromatic mediants', are present in each chord in two ways: the upper chromatic
mediant, formed by the root at the third of original chord; and the lower
chromatic mediant, formed by the fifth at the third of original chord. Both chords also are formed by a third, but with a chromatic alteration regarding
the original chord. If two chromatic alterations exist, i.e.\ two notes altered
by one semitone each regarding the original chord, the mediant is called
'doubly chromatic mediant'. Again, there are two forms for each chord: the upper form, with
a third in the fifth of the original chord; and the lower form, with a third in the root
of the original triad. It is interesting to observe that a major chord has upper chromatic mediants and upper doubly chromatic mediants. A minor chord has lower chromatic mediants and lower doubly chromatic mediants. This relation between chords is considered as
advanced tonalism, sometimes even considered as expansion and dissolution of tonalism, and, as consequence, it has strong and outstanding effects although perfectly consonant. The chromatic
mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss,
among others, and are quite simple to be realized~\cite{Harmonia,Salzer}.

\subsubsection{Modulation}

Modulation is the change of the key (tonic, or tonal center) in a music. It can be characterized by observing the key at the start and at end of the chord transitions. Keys are always conceived as related by fifths and their relative and counter-relatives. Forms to perform modulation include:

\begin{itemize}
    \item Transposing the discourse to a new key, without any preparation. It is
    a common Baroque procedure although incident at other periods as well. Sometimes it is 
    called phrasal modulation or unprepared modulation.
    \item  Using carefully an individual dominant and perhaps the individual
    subdominant, to confirm the change in key and harmonic field.
    \item Using chromatic alterations to reach a chord in the new key starting from a
    chord in the previous key. It is called chromatic modulation.
    \item Featuring a unique note, possibly repeated or suspended with no
    accompaniment, common to start and end a key, constitute a peculiar way
    to introduce the new harmonic field.
    \item Changing the function, without changing the notes, of a chord to
    contemplate a new key. This procedure is called enharmony.
    \item Maintaining the tonal center and changing the key quality from major to minor
    (or vice-verse) comprehends a parallel modulation. Key with same tonic and
    different quality is known as homonyms.
\end{itemize}

The dominant has great importance and is a natural pivot into modulations,
which leads to the circle of fifths~\cite{Harmonia,Salzer,Koellheuteur,Harmony}. 
The music piece \emph{Acorde cedo} explores these chord relations. Its code is
available both in Appendix~\ref{ap:acorde} and online as part of \massa~\cite{MASSA}.

\subsection{Counterpoint}\label{subsec:contraponto}

Counterpoint is defined as the conduction of simultaneous melodic lines, called voices. The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues. It is possible to
summarize the main counterpoint rules, and it is known that Beethoven --
among others -- outlined such similar synthesis.

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{figures/movContraponto}
    \caption{Different motions of counterpoint aiming to preserve independence
        between voices. There are 3 types of motions: direct, contrary and
        oblique. The parallel motion is a type of direct motion.}
        \label{fig:movContraponto}
\end{figure}

The purpose of counterpoint is to conduct voices in a way they sound
independent. In order to do that, the relative motion of voices (in pairs), is crucial and
categorized as: direct, oblique and contrary motion (see Figure~\ref{fig:movContraponto}). The parallel motion is an oblique motion.
The gold rule here is to take care with the direct motions, avoiding making them
lead to a perfect consonance. The parallel motion should occur only between
imperfect consonances and no more than three consecutive times. The dissonances
can be unadmitted or used when followed and preceded by consonances of joint
degrees, i.e.\ neighbor notes in a scale. The motions that lead a note to a
neighbor sound more coherent. When having 3 or more voices, the melodic
importance lies on the higher and lower voices, in this particular
order~\cite{Fux,Tragtenberg,SchoenbergContra}.

These rules were used in the music piece \emph{Conta ponto}, the source code is
in Appendix~\ref{ap:conta} and available online in \massa.

\subsection{Rhythm}\label{subsec:ritmo}

Basically speaking, the rhythm notion is dependent on events separated by durations~\cite{Lacerda}. These events can be heard individually if spaced by at least $50-63ms$. To make the temporal separation between them be perceived as duration, the period should be large enough, around $100ms$~\cite{microsound}.  
It is possible to summarize the duration transitions heard as rhythm and pitch,
in the following way~\cite{Alfaix, microsound}.

\begin{table*}[htpq!]
\tiny
\centering
\caption{Transition of durations heard individually until it turns into pitch.}
\begin{tabular}{  l | r r r r   r r r    r r r || r r || r r r r r r }
\hline
           & \multicolumn{10}{c}{$\underleftarrow{\text{\bf perception zone of
           duration as rhythm}}$} & \multicolumn{2}{c}{transition} & \multicolumn{3}{c}{-} \\
duration (s) & {\bf ...}     & {\bf 32,}     & {\bf 16,}   & {\bf 8,}  & {\bf 4,}   & {\bf 2,}   & {\bf 1,}   & {\bf 1/2,} & {\bf 1/4,} & {\bf 1/8,} & $\frac{1}{16}=62,5ms$ , & $\frac{1}{20}=50ms$ & {\color{gray} 1/40} & {\color{gray} 1/80  } & {\color{gray} 1/160 } & {\color{gray} 1/320 } & {\color{gray} 1/640 } & {\color{gray} ... } \\
frequency (Hz) & {\color{gray} ...} & {\color{gray} 1/32,}   & {\color{gray} 1/16,} & {\color{gray} 1/8,} & {\color{gray} 1/4,} & {\color{gray} 1/2,} &  {\color{gray} 1,}  & {\color{gray} 2,}   & {\color{gray} 4,}   & {\color{gray} 8,}    & 16,  & 20   & {\bf 40}   & {\bf 80}   & {\bf 160}   & {\bf 320}   & {\bf 640}   & {\bf ...} \\
           & \multicolumn{10}{c}{ - } & \multicolumn{2}{c}{transition}
           & \multicolumn{6}{c}{$\overrightarrow{\text{\bf perception zone of
           duration as pitch}}$} \\
\hline
\end{tabular}
\label{tab:duracoes}
\end{table*}

The duration region marked as transition is presented minimized because the limits
are not well defined. In fact, the duration where someone begins to perceive a
fundamental frequency or a separation between occurrences, depends on the
listener and sound characteristics~\cite{microsound,Roederer}.

The rhythmic metric is commonly based on a basic duration called pulse. The
pulse typically comprehends durations between $0.25-1.5s$ ($240$
and $40 BPM$, respectively). In music education and cognitively studies, it is common to
associate these range of pulse frequencies to the durations of the heart beat,
movements of inspiration and expiration and walk steps~\cite{Lacerda,Roederer}.

The pulse is subdivided into equal parts and repeated in sequence. These
relations (division and concatenation) usually follow relations of small
order integers\footnote{In ascending order of occurrence in written and ethnic
music, these indicate the musical pulse divisions and their sequential grouping at
time: 2, 4 and 8, after that 3, 6 (two groups of 3 or 3 groups of 2) and 9 and
12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, complementing
1-9 and 12. Other metrics are less common, like divisions and grouping in 13,
17, etc, and are mainly used in contexts of experimental music and classical
music of XX and XXI. No matter how complex they seem, metrics are commonly
compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.}.
A schematic exposition is shown in Figure~\ref{fig:pulsoSubAgl}.

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{figures/metricaMusical}
    \caption{Divisions and grouping in bars for the music pulse to establish a rhythmic
        metric. Divisions of the quarter note, established as the
        pulse, is presented in the left. The time signature whose specifies the same
        metrics but having as scale the groupings of the music pulse is presented in the right.}
        \label{fig:pulsoSubAgl}
\end{figure}

Dual relations (common time and binary divisions) are frequent in dance rhythms
and celebrations. Ternary relations are common in
ritualistic music and is related to the sacred. Dual relations are called imperfect, while and the ternary relations are called perfect.

The strong unities (accents) are the ones that fall in the 'head' of bars, the downbeats. The
head of a unity is the first part of the subdivision. In binary divisions (2, 4
and 8, for example), the unities originally strong turn into weak (não sei se isso ficou claro)
(e.g.\ division in 4 is: strong, weak, strong medium, weak). In ternary divisions
(3, 6 and 9) two weak unities succeeds the downbeat (e.g.\ division in 3 is:
strong, weak, weak)\footnote{Division in 6 is considered compound but can also
occur as a binary division. A binary division which suffers a ternary division
results in two units divided in tree units each: strong (subdivided in strong,
weak, weak) and weak (subdivided in strong, weak, weak). Another way to perform
the division in 6 is applying ternary division that suffers a binary division,
resulting in: a strong unit (subdivided in strong and weak) and two weak units
(subdivided in strong and weak for each).}

The accent in the weak beat is known as the backbeat, whereas a note starting in a weak
beat which duration coming across the strong beat is known as syncope.

Notes can occur inside and outside of these \emph{'music metric'} divisions. In
most behaved cases, notes occur exactly on these divisions, with broader
incidence on attacks of strong beats.
In extreme cases, the time metric cannot be perceived~\cite{Roederer}. 
Small grid variations help to build musical interpretation and differences
between styles~\cite{Cook}.

% ficou um tanto confuso:

Being pulse the grouping level $j=0$, level $j=-1$ the first pulse subdivision,
level $j=1$ the first pulse agglomeration and so on. In this way, $P_i^j$ is the
$i$-th unit of pulses at grouping level $j$: $P^0_{10}$ is the tenth pulse,
$P^{1}_3$ is the third pulse grouping unit (and, possibly, the third measure),
$P^{-1}_2$ is the second unit of pulse subdivision.

Special attention should be given to the limits of $j$: pulse divisions are
durations sensible as rhythm. Furthermore, the pulse joints sums, at its
maximum, a music or a cohesive set of musics. In other words, a duration given
by $P^{min(j)}_i$, $\forall \; i$, is greater than $50 ms$ and the durations
summed together $\sum_{\forall i}P^{\text{max}(j)}_i$ are less than a few
minutes or at most, a few hours.

Each level $j$ has some indexes $i$. A index $i$ always has three different
values (or multiple of three) that occur in a perfect relation. When $i$ has only
multiple of two, four or eight different values, than an imperfect relation occurs,
as shown in Figure~\ref{fig:pulsoSubAgl}.

Any unit (note) of a given musical sequence which has a time metric can be unequivocally
specified as:

\begin{equation}
P^{ \{ j_k \} }_{ \{ i_{k} \}}
\end{equation}

\noindent where $j_k$ is the grouping level and$i_k$ is the unit order itself.

As example, consider $P^{-1,0,1}_{3,2,2}$ as the third subdivision $P^{-1}_3$ of the
second pulse $P^0_2$ and of the second pulse group $P^1_2$. Each unit or unit set
$P_i^j$ can be associated with a sequence of temporal samples $T_i$ that forms a
music note.

The music piece \emph{Poli Hit Mia} explorers different metrics (avaleble in Appendix~\ref{ap:poli} and as part of \massa).

\subsection{Repetition and variation: motifs and large units}\label{subsec:motivos}

Given the basic music structures, either frequential (chords and scales) or
rhythmic (beat divisions and simple, compound and complex grouping), it is
natural to organize these structures in a cohesive way~\cite{Boulez}. To this end,
the concept of arcs are fundamental: we make an arc by departing from a place (music unit) and going to another one. The audition of melodic and harmonic lines is pervaded as
musical arcs thanks to the cognitive nature of the musical hearing. It is
possible to consider the note as the smaller arc, and each motif (definir motif aqui? nao foi falado ainda) and melody as another arc. Each time, beat subdivision, measure and music
section constitutes an arc. A music in which the arcs do not present consistency by
themselves, can be understood as a music with no cohesion. The coherence feeling
comes, mostly, from the skilled manipulation of arcs in a music piece.

Musical arcs are abstract structures and liable of basic operations. An spectral
arc, like a chord, can be inverted, enlarged and permuted, to cite some
examples. Temporal arcs, like a melody, a motif, a measure or a note are also
capable of variations. Remembering that
$S_j=\left\{s_j=T_i^j=\{t_i^{j}\}_0^{\Lambda_j-1}\right\}_0^{H-1}$ is a sequence
of $H$ musical events $s_j$, each event with its $\Lambda_j$ samples $t_i^j$
(refer to the beginning of this section~\ref{notasMusica}), the basic techniques
can be described as:

\begin{itemize}
        \item Temporal translation is a displacement
    $\delta$ of a specific material to another instant $\Gamma'=\Gamma + \delta$
    of the music. It is a variation that changes temporal localization in
    a music:
    $\left\{s_j'\right\}=\left\{s_j^{\Gamma'}\right\}=\left\{s_j^{\Gamma+\delta}\right\}$
    where $\Gamma$ is the duration between the beginning of the piece (or considered
    period) and the first event $s_0$ of original structure $S_j$. Observe that
    $\delta$ is the time offset of the displacement.

    \item Temporal expansion or contraction is a change in duration of each
    arc by a factor $\mu\,:\; s_j'^{\Delta}=s_j^{\mu_j . \Delta}$. Possibly,
    $\mu_j=\mu$ is constant.

    \item Temporal reversion consists of generating a sequence with elements
    in reverse order of the original sequence $S_j$, thus: $S_j'=\left\{s_j'\right\}_0^{H-1}=\left\{s_{(H-j-1)}\right\}_0^{H-1}$.

    \item Pitch translation is a displacement $\tau$ of a material for a
        different pitch $\Xi'=\Xi + \tau$. It is a variation that changes pitch
        localization of material:
        $\left\{s_j'\right\}=\left\{s_j^{\Xi'}\right\}=\left\{s_j^{\Xi+\tau}\right\}$
        where $\Xi$ is the pitch of a period $S_j$ or of an event $s_0$ of the
        original structure $S_j$. Observe that $\tau$ is the pitch offset of the
        displacement. If $\tau$ is given in semitones, the displacement in
        frequency is $\tau_f=f_0.2^{\frac{\tau}{12}}$ where $f_0$ results from
        the adoption of some reference: $f_0=\Xi_{f_0}Hz\;\sim \Xi_0$ absolute
        value of pitch. For the frequency of any pitch value: $\Xi_f=\Xi_{f_0}.2^{\frac{\Xi-\Xi_0}{12}}$. 
        For the pitch of any frequency value: $\Xi=\Xi_0 +12
        . \log_2\left(\frac{\Xi_f}{\Xi_{f_0}}\right)$.  
        In the MIDI protocol, $\Xi_{f_0}=55Hz$ while pitch $\Xi_0=33$ marks
        a \textit{A 1}, another reference point is $\Xi_{f_0}=440Hz$ which
        corresponds to $\Xi_0=69$. The unit difference is equal for semitones.
        It is not possible to say that the unit is the semitone because $\Xi=1$
        is not a semitone, but a note with an audible frequency as rhythm, with
        less than 9 occurrences each second (see Table~\ref{tab:duracoes}).

        \item Interval inversion is the inversion traversed by
        the material. The inversion is strict if the number of semitones is
        being used as a reference for the operation:
        $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{-\varepsilon_j . f_0}\right\}$, where
        $\varepsilon_j$ is the factor between the event frequency $s_j$ and the
        frequency of $s_0$. The inversion is said tonal if the distances are
        considered in terms of degree number of the scale $E_k$:
        $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{-\varepsilon^{\left(e_{\left(j_e\right)}\right)}
        . f_0}\right\}_0^{H-1}$ where $j_e=e_{j}^{inv}$ is the index $k=j_e$ in
        $E_k$ of the note of event $s_j$.

        \item Rotation of musical elements is the transference of the last
        element to the position of the first element, and the displacement of
        that to the penult, one position ahead. It is possible to define
        rotation of $\tilde{n}$ positions by $S'_n=S_{(n+\tilde{n})\%H}$. If
        $\tilde{n}<0$, it is enough to use $\tilde{n}'=H-\tilde{n}$. It is
        reasonable to associate $\tilde{n}>0$ with the clockwise rotation and
        $\tilde{n}<0$ with tne anti clockwise rotation. More information about rotations in presented in subsection~\ref{estCic}.

        \item The insertion and removal of materials into structure $S_j$ can be
    ornamental or structural: $S_j'=\{s_j'\}=\{s_j \text{ if condition A,
    otherwise } r_j\}$, for any music material $r_j$, including the empty
    instant. Elements of $R_j$ can be inserted at the beginning, like a prefix
    in $S_j$; at the end, as a suffix; or at the middle, dividing $S_j$ or making
    it the prefix or suffix. Both materials can be mixed in a variety of ways.

    \item Changes in articulation, orchestration and spatialization, 
    $s_j'=s_j^{*_j}$, where $*_j$ is the new characteristic incorporated by the
    element $s_j'$.

    \item Accompaniment. Both orchestration and melodic lines presented when
    $S_j$ occurs can suffer modifications and be considered as a variation of $S_j$ itself.
\end{itemize}

Other proceesings can be derived together with the presented ones, as, for example, the retrograde
inversion, temporal contraction with an external suffix, etc. The music
structures resound in the human cognitive system due to the own nature of
thought. In its many veneers, an idea read with the same number of elements and
connective aspects between them. The music, when tunned with those mental structures,
raises impressions. In this way, a whole process of mental and neurological
resonance is unleashed, responsible by the feelings, memories and imaginations,
typical to a mindful music listening. This cortical activity helps the
musical therapy, known by its utility in cases of depression and neurological
injury. It is known that regions of the human brain responsible by
hearing processing, are also used for other activities, even linguistic and mathematical~\cite{Sacks,Roederer}.

The paradigmatic structures guide the creation of new music material. One of the
most central of them is the dipole tension/relax. Traditional dipoles
are relate to tonic/dominant, repetition/variation,
consonance/dissonance, coherence/rupture, symmetry/asymmetry,
equality/difference, arrival/departure, near/far, stationary/moving,
etc. The ternary relations tend to relate to the circular and unification. The
lucid ternary communion, 'modus perfectus', opposes to the passionate
dichotomic, 'modus imperfectus'. Hereafter, there is an exposition dedicated to
directional and cyclic arcs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Directional structures}\label{subsec:dir}

The arcs can be decomposed in two convergent sequences: the first arc reaches the
apex and the second arc returns from the apex to the start region. This apex is called climax by traditional music theory. It is
possible to differ between arcs whose climax are placed at the beginning, middle, end,
and at the first and second half of the considered duration. These structures are
shown in Figure~\ref{fig:climax}. The parameter of variation does not exist, thus
the arc consists only of the reference structure~\cite{Schoenberg}.

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{figures/climax}
        \caption{Canonical distinctions of musical climax in a given melody and
        other domains. The considered possibilities are: climax at beginning,
        climax at first half, climax at middle, climax at second half and climax
        at the end. The y-axis is not properly specified since there exist no
        real parametric variation and, therefore, the structure is a reference.}
        \label{fig:climax}
\end{figure}

Consider $S_i=\{s_i\}_0^{H-1}$ as an increasing sequence. The sequence
$R_i=\{r_i\}_0^{2H -2}=\left\{s_{(H-1-|H-1-i|)}\right\}_0^{2H-2}$ is a sequence
that presents a perfect profiteer symmetry, i.e.\ the second half is the
mirrored version of the former. Following the musical concepts, the climax is
localized in the middle of the sequence. It is possible to modify this
fact using sequences with different sizes. All the mathematical theory of
sequences, already well established and taught routinely in calculus courses, can be
used to generate those arcs~\cite{Guidorizzo,Schoenberg}. 
Theoretically, when applied to any characteristic of music events, these sequences produce arcs, since they imply the deviation and return of an initial
parametrization. Henceforth, it is possible to have a
number of distinct arcs, with different sizes and climax for a same sequence of events. This is an interesting and useful resource, and the correlation of arcs results in coherence~\cite{Salzer}.

Historically (and practically nowadays), the golden ratio and the Fibonacci sequence have special importance. The Lucas sequence allows the generalization of
Fibonacci sequence, making it easy to understand. Given any two numbers $x_0$
and $x_1$, the Lucas sequence can be obtained as: $x_n=x_{n-1}+x_{n-2}$. The greater $n$ is, the more $\frac{x_{n}}{x_{n_1}}$ approaches the golden ratio:
$1.61803398875...$. The sequence converges fast even with high discrepant
initial values. Being $x_0=1$ and $x_1=100$, and $y_n=\frac{x_n}{x_{n+1}}$ an
auxiliary sequence, the error occurring in the first values with
relation to the golden ratio is, approximately, $\{ e_n \}
=\left\{100\frac{y_n}{1.61803398875}-100 \right\}_1^{10}=\{6080.33, -37.57, 23,
-7.14, 2.937, -1.09, 0.42, -0.1601, \\ 0.06125, -0.02338\}$. The Fibonacci sequence
presents about the same error progression but starts right at the second step since $\frac{1}{1}\approx\frac{101}{100}$.

The music piece \emph{Dirracional} exposes the use of arcs into directional structures.
Its source code is included in Appendix~\ref{ap:dirracional} and available online as part
of \massa~\cite{MASSA}.

\subsection{Cyclic structures}\label{estCic}

The philosophical understanding that the human thought is based on the perception of
similarities, differences, stimuli and objects, places the symmetries
at the core of the cognitive process~\cite{Deleuze}. Mathematically, symmetries
are algebraic finite groups, and a finite group is always isomorphic to a permutation
group. It is possible to say that permutations represent any symmetry in a
finite system (precisa de uma referência aqui). In music, permutations are ubiquitous and considerably present in techniques, which confirms their central role. The successive application of permutations generates cyclic arcs~\cite{change,Zamacois,permMusic}. Two academic studies were dedicated to this end, aiming at generating music structures~\cite{figgusOriginal, figgusEspacializacao}.

Any permutation set can be used as a generator of algebraic groups~\cite{permMusic}.
The properties defining a group $G$ are:

\begin{multline}\label{eq:groups}
%\begin{split}
\forall \;\; p_1,p_2 \in G \Rightarrow\quad\;\; p_1 \bullet p_2  =
p_3 \in G \\ \;\;\;\text{(closure property)} \\
\forall \;\; p_1,p_2,p_3 \in G \Rightarrow\quad (p_1\bullet p_2)\bullet p_3  =
p_1\bullet (p_2\bullet p_3)\quad\;  \\ \text{(associativity property)} \\
\exists \;\; e \in G :\; p \bullet e  = e \bullet
p \;\;\;\; \forall p \in G  \quad \\ \text{(identity element property)} \\
\forall \;\; p \in G, \;\exists\; p^{-1} :\quad\;  p\bullet p^{-1} 
=p^{-1}\bullet p = e \\ \; \text{(inverse element property)}
%\end{split}
\end{multline}

From the first property it is possible to conclude that any permutation can be
operated with another permutation. In fact, it is possible to apply a
permutation $p_1$ and another permutation $p_2$, and, comparing both initial and final
orderings, results in another permutation $p_3$.

Every element $p$ operated with itself a sufficient number of times $n$ reaches
the identity element $p^n=e$ (otherwise the group would be infinite, generated
by $p$). The lower $n\;:\;p^n=e$ is called the element order. Thus, a finite
permutation $p$, successively applied, reaches the initial ordering of its
elements, making a cycle. This cycle, if used as parameter of music notes,
implies in a cyclic musical arc.

These arcs can be established by using a set of different permutations. As a historic
example, the tradition called \emph{change ringing} conceives the music through
bells played one after other and then played again, but in a different
order. This process repeats until it reaches the initial ordering. The set of
different traversed orderings is a \emph{peal}. Table~\ref{tab:change}
represents one traditional \emph{peal} for 3 bells (1, 2 and 3) which explores
all possible orderings. Each line indicates one bell ordering to be
played. Permutations occur between each line. In this case, the music structure
consists of permutations by itself and some different permutations operate due to
the cyclic behavior.

\begin{table}[htpq!]
\centering
\caption{Change Ringing: Standard \emph{peal} for 3 bells. Permutations
occur between each line. Each line is a bell ordering and each ordering is played
        a line a time.} 
\begin{tabular}{l c r}
\textcolor{red}{1} & \textcolor{blue}{2} & \textcolor{green}{3} \\
\textcolor{blue}{2} & \textcolor{red}{1} & \textcolor{green}{3} \\
\textcolor{blue}{2} & \textcolor{green}{3} & \textcolor{red}{1} \\
\textcolor{green}{3} & \textcolor{blue}{2} & \textcolor{red}{1} \\
\textcolor{green}{3} & \textcolor{red}{1} & \textcolor{blue}{2} \\
\textcolor{red}{1} & \textcolor{green}{3} & \textcolor{blue}{2} \\
\textcolor{red}{1} & \textcolor{blue}{2} & \textcolor{green}{3}
\end{tabular}
\label{tab:change}
\end{table}

The use of permutations in music can be summarized in the following way:
consider $S_i=\{s_i\}$ as a sequence of musical events $s_i$ (e.g.\ notes), and a
permutation $p$. $S_i'=p(S_i)$ comprises the same elements of $S_i$ but in a
different order. The permutations can be written based on two notations: cyclic or
natural. The natural notation basically comprehends the order of the resulting indexes from
the permutation. Thus, agreed the original ordering given the sequence of its
indexes $[0\;1\;2\;3\;4\;5\;...]$, the permutation is noted by the sequence it
produces (ex. $[1\;3\;7\;0\;...]$). In the cyclic notation, permutation is the
changing of one element by the posterior one, and the last element by the first one.

It is not necessary to permute elements of $S_i$, but only some
characteristics. In this way, being $p^f$ a permutation in frequency and $S_i$ a
sequence of basic notes as exposed in the end of section~\ref{notaBasica}, the
new sequence $S_i'=p^f(S_i)=\left\{s_i^{p(f)}\right\}$ consists of the same
music notes, respecting same order and maintaining the same characteristics, but with the
fundamental frequencies permuted following the pattern that $p^f$ presents.

It is worthwhile to mention two subtleties of this procedure. First, the permutation $p$ does not have to involve all elements of $S_i$, i.e.\ it can operate in subsets of $S_i$. Second,  the elements $s_i$ do not have to be entirely executed at each state access. To exemplify,
consider $S_i$ as a sequence of music notes $s_i$. If $i$ goes from $0$ to $n$, and
$n>4$, at each measure of $4$ notes it is possible to execute the first $4$
notes. The other notes of $S_i$ can occur in measures where permutations set
aside those notes to the first four notes of $S_i$.

Each one of the permutations $p_i$ described above relates the note dimensions where it operates (frequency, duration, \emph{fades},
intensity, etc) and the period of incidence (at how much access a permutation is
applied). During realization of notes in $S_i$, an easy and coherent form is to
execute the first $n$ notes\footnote{The execution of disjoint notes of $S_i$
equals to modify the permutation and to execute the first notes.}.

Appendix~\ref{cap:FIGGUScode} and \massa present the computational implementation about permutations~\cite{MASSA,figgusOriginal,figgusEspacializacao}.

\subsection{Musical idiom?}

There are many studies intending to model and explore a 'musical language',
or 'linguistics applied to music' or even to discern between different 'musical
idioms'~\cite{Lerdahl, Harmonia, Salzer,Alfaix}. In a simple way, a musical idiom
is the result of a chosen material together with repetition of elements and
repetition of relations between elements present along the music. The related dichotomies are outstanding, as explained at subsection~\ref{subsec:motivos}: repetition and variation, relax and tension, equilibrium and instability, consonance and dissonance, etc.

\subsection{Musical usages}\label{subsec:usosmusicais3}

The basic note was defined and characterized in quantitative terms (section~\ref{sec:notaDisc}). Next, the internal note
composition was addressed and both internal transitions and immediate treatment
were understood (section~\ref{varInternas}). Finally, this section aims at
organizing these notes in music. The gamma of resources and consequent infinitude
of possibilities is a typical situation and highly relevant for art contexts~\cite{Harmonia,Webern}.

There are studies dedicated to each one of the presented resource. For example, it is possible to obtain the 'dirty' triadic harmonies (with notes without the triad) by
superpositions of dirty fourths. Another interesting example is the simultaneous
presence of rhythms in different metrics, constituting what is
called \emph{polyrhythms}. The music piece \emph{Poli-hit mia} explores these
simultaneous metrics by impulse train (??) convolved with notes that compound the
line. Its source code is included in Appendix~\ref{ap:poli} and available online as part
of \massa.

The microtonal scales are important for the music of the 20th
century~\cite{microtonalidade} and present diverse outstanding results, like the
fourths of tone in the Indian music. The musical sequence \emph{MicroTom}
explores these resources, including microtonal melodies and microtonal harmonies
with many notes in a very reduced note scope. Its code is presented in
Appendix~\ref{ap:micro} and available online in \massa.

As pointed in subsection~\ref{subsec:mus2}, the relations between
parameters are powerful ways to acquire music pieces. The number of permuted
notes can vary during the music, revealing relationship with piece
duration. Harmonies can be made of triads (eqs.~\ref{triades}) with replicated
notes at each octaves and more numerous as minor the depth and frequency of
vibratos (eqs.~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}),
among other uncountable possibilities.

The presented symmetries at octave divisions (eqs.~\ref{escSim}) and the
symmetries presented as permutations (Table~\ref{tab:change} and
eqs.~\ref{eq:groups}) can be used together. In the music piece \emph{3 trios}
this association is done in a systematic way in order to enable the required
listening. This is a instrumental piece, not included as a source
code~\cite{3Trios}.

The \emph{PPEPPS} (Pure Python EP: Projeto Solvente) is an EP synthesized using
the resources presented in this study. With minimal parametrization, the program
generates complete musics, allowing the easy composition of musics and sets of
musics. Using few lines of code it is possible to obtain
a directory with the generated musics. This facility and technological
demystification create aesthetically possibilities for both sharing and education.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusions and future works} %Nome do capítulo.
\label{cap:conclusao}

We have described a concise system that explores the musical elements and relate them to the digital signal. The reader is invited to access \emph{Scripts} and \massa, where these relations are computationally implemented. We hope the didactic report along the paper and the supplied scripts facilitates and encourages the use of the proposed framework.

The open possibilities provided by the techniques and results discussed in this paper involve psychoacoustic experiments and the creation of interfaces for the generation of music, noise and other music applications in a high fidelity (\emph{hi-fi}). It is worthwhile to mention the benefit of these results for artistic and didactic purposes. The incorporation of programming skills is facilitated by the provided visual aids. Initial \emph{live-coding} practices and guided courses based on this work have already been realized with successful. Examples are Puredata and Chuck.

This work systematically investigates the parameterization issues (like the tremolo, ADSR, etc.) in a high fidelity, which has significant artistic utility. Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature (as shown in Appendix~\ref{cap: trabalhosRelacionados}).

Besides, the free software license and online availability of the exposed content as hypertext, in conjunction with the respective codes and sound examples, strongly facilitates future collaborations and the generation of sub-products in a co-authorship fashion. As consequence, the expansion of \massa is feasible and straightforward, making it possible new developments and implementations of musical context.

In addition, this work permitted the formation of groups with common interests, like music creativity and computer music. Specially, the project \url{labMacambira.sf.net} groups Brazilian and foreign co-workers in order to offer relevant contributions in diverse areas like Digital Direct Democracy (não devo ter traduzido certo), georeferencing techniques, artistic and educational activities. Some of these reports are available in Appendix~\ref{cap:musicaExtra} and made online. There are more than 700 videos, written documentations, original softwares and contributions in well-known external softwares such as Firefox, Scilab, LibreOffice, GEM/Puredata, to name a few~\cite{siteLM,wikiLM,vimeoLM}.

Future works include the application of the reported results in machine learning and artificial intelligence methods for the generation of appealing artistic materials.

%% Há um aumento no número de pesquisas relacionadas à música em andamento no campus de São Carlos da USP, o que sugere facilidade para estabelecer parcerias.
%% As publicações acadêmicas efetivadas durante este mestrado também apontam para uma multidisciplinaridade,
%% tratando diretamente de questões humanas como artes, filosofia, humor e linguagem falada e escrita, através de artifícios lineares e estatísticos.\cite{FabbriSTAT,FabbriACL,FabbriComplenetVoz,FabbriComplenetTexto} Os desdobramentos estão alcançando redes sociais e teorias epistemológicas com base em pesquisas prévias, dos orientadores deste trabalho, com forte presença de redes complexas e processamento de linguagem natural.

\nocite{*}
\bibliography{phi+mus}
\end{document}

### 2013-09-19 article/article.tex
feito ateh I. A

@@
 The frequency bandwidth between $20Hz$ and $20kHz$ is appreciated by the human hearing system with boundaries dependent on the person, climate conditions and sonic characteristics themselves~\cite{Roederer}.
-Considered the speed of sound as $\approx 343.2 m/s$, these limits corresponds to wavelengths of $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.
-
-Human perception of sound involves captivation by bones, stomach, ears, transfer functions of head and processing torso and nervous system~\cite{}.
-Besides that, the ear is a dedicated organ responsible for the capture of these waves.
-Its mechanism decomposes sound into its sinusoidal spectrum and delivers them to the nervous system.
-The sinusoidal components are crucial to musical phenomena, as one can perceive the composition of sounds according to a specific musical interest, e.g., tunings and scales.
+Considered the speed of sound as $\approx 343.2 m/s$, these limits imply wavelengths of $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.
+
+Human perception of sound involves captivation by bones, stomach, ears, transfer functions of head and torso, and processing by the nervous system~\cite{}.
+Besides that, the ear is a dedicated organ for the appreciation of these waves.
+Its mechanism decomposes sound into its sinusoidal spectrum and delivers to the nervous system.
+The sinusoidal components are crucial to musical phenomena, as one can perceive in the composition of sounds of musical interest (such as harmonic sounds and noises, seen in sections~\ref{sec:discNote} and~\ref{sec:internalVar}), and sonic structures of musical interest (such as tunings and scales, in section~\ref{sec:notesMusic}).
 
 The representation of sound is commonly called audio (although these terms are often used without distinction).
-Audio expresses the waveform prominent from the capture of microphones or from synthesis.
-In order to account for the corresponding digital representation, digital audio is specified by protocols that eases file storage and transferring, usually with low loss of quality.
-Standard representation of digital audio consists of equally spacing samples by $\lambda_s$ durations in time, where each sample is specified by a sample number of bits.
-This process is called the Pulse Code Modulation (PCM) representation of sound.
-A PCM digital sound is characterized by it's sampling frequency $f_s=\frac{1}{\lambda_s}$ (also called the sampling rate), and by a bit depth, which is the number of bits used for representing the amplitude of each sample.
-With the purposes of illustrating this idea, Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with $4$ bits each.
-The $2^4=16$ possible steps for each sample, together with the regular spacing $\lambda_s$ between them, establish what is named as quantization error.
-The noise prominent of these errors diminishes conforming the spacing also does.
+Audio expresses waves prominent from the capture of microphones or from direct synthesis, although these sources entangle as captured sounds are processed to generate new sonorities.
+Digital audio specified by protocols that eases file transfering and storage often imply a quality loss.
+Standard representation of digital audio, on the other hand, assures perfect reconstruction of the analog wave, within any convenient precision.
+This paradigm consists of equally spaced samples, by $\lambda_s$ durations, each specified by a fixed number of bits.
+This is the Pulse Code Modulation (PCM) representation of sound.
+A sound in PCM audio is characterized by a sampling frequency $f_s=\frac{1}{\lambda_s}$ (also called the sampling rate), which is the number of samples used for representing a second of sound; and by a bit depth, which is the number of bits used for representing the amplitude of each sample.
+Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with $4$ bits each.
+The fixed $2^4=16$ values for the amplitude of each sample, with the regular spacing $\lambda_s$, yields the quantization error or noise.
+This noise diminishes as bit depth increases.
 
@@
 
-The Nyquist theorem~\cite{} says that half of the sampling frequency is the maximum frequency of the signal.
-Thus, it is necessary to have a sampling frequency at least twice the highest frequency heard by humans, that is, $f_s \geq 2\times 20kHz = 40kHz$.
-This configures the basis for the use of sampling frequencies like $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and Broadcast systems (Radio and TV), respectively.
-
-For the average listener -- and a reasonable part of specialists -- the notion of music presupposes rhythmic and pitch organization.
-I a simple  way, rhythm can be described as recurring pattern produced by notes differing in duration, pause (silence) and intensity.
-Pitch configures the perceptual attribute that allows sounds to be organized in a frequency-related scale~\cite{}.
-Music from the twentieth century enlarged this traditional comprehension of music.
-This occurred in the concert music, specially in the concrete, and in both electronic and electroacoustic styles.
-On the last decade of the century, it became evident that popular music had also incorporated sounds without defined pitch and temporal organization, to name a few of simple metrics.
-Even though, the note still stands paradigmatic as a 'fundamental unit' of musical structures and, in practice, it can unfold in sounds that observe this recent developments (**not clear here**).
+The Nyquist theorem~\cite{} states that the sampling frequency is twice the maximum frequency of the represented signal.
+Thus, for general musical purposes, it is necessary to have samples in a rate at least twice the highest frequency heard by humans, that is, $f_s \geq 2\times 20kHz = 40kHz$.
+This is the fundamental reason for the adoption of sampling frequencies like $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and broadcast systems (radio and television), respectively.
+
+Within this framework, musical notes can be characterized.
+The note still stands paradigmatic as the 'fundamental unit' of musical structures and, in practice, it can unfold into sounds that uphold other approaches\footnote{Music from the twentieth century enlarged this traditional comprehension of music.
+This occurred in the concert music, specially in the concrete, electronic and electroacoustic styles.
+On the last decade of the century, it became evident that popular music had also incorporated sounds without defined pitch and temporal organization, to name a few of simple characteristics.}.
+Notes are also convenient for another reason: the average listener -- and considerable part of the specialists -- presupposes rhythmic and pitch organization (made explicit in section~\ref{sec:notasMusic}) as fundamental musical properties, and these are developed in traditional musical theory in terms of notes.
 
@@
 \nocite{*}
-\bibliography{phi+mus}
+\bibliography{article}
 \end{document}

### 2013-09-19 article/article.tex
feita intro do artigo

@@
 
-This paper purposes to represent musical structures and artifices by their discrete sound characteristics.
-The results include mathematical relations and their computer program implementations.
-Despite the general interest and number of knowledge areas involved, there are only a few number of books and computer implementations that tackle the subjects presented in this work.
-However, they mainly focus on aspects of musical sounds and ways to mimic traditional instruments.
-A detailed compilation of them is pointed out in the bibliography~\cite{dissertacao}.
+This article aims at representing musical structures and artifices by their discrete-time sonic characteristics.
+Results include mathematical relations, usually in terms of samples, and their direct computer program implementations.
+Despite the general interests involved, there are few books and computer implementations that tackle the subject.
+These mainly focus on computer implementations and ways to mimic traditional instruments, with scattered mathematical formalisms.
+A compilation of the works and their contributions is in the bibliography~\cite{dissertacao}.
 To the best of the author's knowledge, there is a lack of articles on the topic.
-Moreover, although current computer implementations use the analytical descriptions presented in this study in a implicitly manner, it seems that there is not a concise and mathematical description of the processes implemented, as they aim here to be further used as libraries for sound and music.
-
-In order to address the concise description of musical elements and structures, the objectives of this paper are:
+Moreover, although current computer implementations use the analytical descriptions presented in this study in a implicit manner, it seems that there has been no concise and mathematical description of the processes implemented.
+
+In order to address this concise description of musical elements and structures, in terms of the digitilized sound, the objectives of this paper are:
 
@@
 \item Introduce a framework of sound synthesis with control at sample level and with potential uses in psycho-acoustical experiments and high-fidelity synthesis.
-The didactic presentation of the content favors its apprehension and usage.
-It is worthwhile to mention that this subject comprises diverse topics on signal processing, music and psycho-acoustics.
 
 \item Provide the accessibility of the developed framework.
-The topics presented in this article are implemented as scripts, i.e.
+The analytic relations presented in this article are implemented as public domain scripts, i.e.
 small computer programs using accessible technologies for better distribution and validation.
-This constitute the \massa\ toolbox, available in public domain in an open Git repository~\cite{gitBook}.
-The scripts are written in Python and make use of external libraries such as Numpy and Scikits/Audiolab, that performs calls to Fortran routines for better computational efficiency.
+This constitute the \massa\ toolbox, available in an open Git repository~\cite{gitBook}.
+These scripts are written in Python and make use of external libraries, such as Numpy that performs calls to Fortran routines for better computational efficiency.
 Part of the scripts have been transcribed to JavaScript and native Python with readiness, what favors their use in Web browsers such as Firefox and Chromium~\cite{numpy, audiolab, tutpython, python}.
-Furthermore, these are all open technologies, that is, published using licenses that allows copy, distribution and use of any part for research and derivatives.
-This way, the work presented here is available and eases co-authorship processes~\cite{Raymond,Lessig}.
+Furthermore, these are all open technologies, that is, published using licenses that allow copy, distribution and use of any part for research and derivatives.
+This way, the work presented here is embedded in recommended practices for availability and eases co-authorship processes~\cite{Raymond,Lessig}.
+
+\item To provide a didactic presentation of the content to favor its apprehension and usage.
+It is worthwhile to mention that this subject comprises diverse topics on signal processing, music, psycho-acoustics and programming.
 
@@
 
-The remaining parts of this work are organized as follows: section~\ref{sec:dicNote} exposes the presence of sinusoid in discrete-time audio and characterizes a basic musical note; the definition and expansion of the musical note as the essential unit of music is further approached in sections~\ref{sec:discNote} and~\ref{sec:internalVar}, respectively; section~\ref{sec:notesMusic} tackles the organization of musical notes in a higher level~\cite{Wisnick,Webern,Lerdhal,Cook,Lacerda}.
-Musical theory embody diverse topics as psycho-acoustics, cultural manifestations and formalisms.
-The section~\ref{sec:results} points out these topics as needed and designate external complements~\cite{Zamacois,Schoenberg,microsound}.
-
-The next section is a minimum text in which music elements are presented side-by-side with the discrete-time samples they result.
-In order to account for validation and sharing, implementations on computer code of these relations and little musical pieces resulting from them are gathered in the \massa\ toolbox, online available.
+The remaining parts of this work are organized as follows: section~\ref{sec:discNote} characterizes the basic musical note; internals of the musical note is further developed in section~\ref{sec:internalVar}; section~\ref{sec:notesMusic} tackles the organization of musical notes in higher levels of musical structure~\cite{Wisnick,Webern,Lerdhal,Cook,Lacerda}.
+As these descriptions embody topics such as psycho-acoustics, cultural traditions, formalisms and protocols, the text points to external complements as needed~\cite{Zamacois,Schoenberg,microsound}.
+
+The next section is a minimum text in which musical elements are presented side-by-side with the discrete-time samples they result.
+In order to account for validation and sharing, implementations on computer code of each one of these relations are gathered in the \massa\ toolbox together with little musical montage resulting from them.
 

### 2013-09-20 article/article.tex
feito ateh metade do timbre

@@
 
-In diverse artistic and theoretical contexts, music is though as being constituted by units called notes, and these units taken as "atoms" that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
-(****nao entendi****, texto ruim)
+In diverse artistic and theoretical contexts, music is conceived as constituted by fundamental units called notes, ``atoms'' that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
 Nowadays, these notes are understood as central elements of certain musical paradigms.
@@
 Canonically, the principal properties of a musical note are duration, volume, pitch and timbre~\cite{Lacerda}.
-These qualities can be managed quantitatively, dictated by the evenly time spaced sound samples.
+These qualities can be managed quantitatively, dictated by the evenly time spaced sonic samples.
 
@@
 Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
-A musical note of duration $\Delta$ seconds is presented as a sequence of $\lfloor \Delta .
+A musical note of duration $\Delta$ seconds is presented as a sequence $T_i^{\Delta}$ of $\lfloor \Delta .
 f_s \rfloor $ samples.
-That is, taken the integer part of the multiplication, it is admitted an error of at most $-\lambda_a$ seconds, which are usually fine for musical purposes.
-Considering $f_s=44.1kHz \;\;\Rightarrow\;\;\lambda_s=\frac{1}{44100}\approx 23$ microseconds.
-It is reasonable to state that:
+That is, the integer part of the multiplication is considered, and an error of at most $\delta_s$ missing seconds is admitted, which is usually fine for musical purposes.
+As an example, if $f_s=44.1kHz \;\;\Rightarrow\;\;\lambda_s=\frac{1}{44100}\approx 23$ microseconds.
+Thus:
 
@@
 
-Being $\Lambda=\lfloor \Delta . f_a \rfloor$ the number of samples in the sequence, thus $T_i=\{t_i\}_0^{\Lambda-1}$.
+With $\Lambda=\lfloor \Delta . f_a \rfloor$, the number of samples in the sequence, the more condensed notation is $T_i^\Delta=\{t_i\}_0^{\Lambda-1}$.
 
@@
 The quantity $V_{dB}$ has the decibel unit ($dB$). 
-For each $10dB$ it is associated a "doubled volume".
-Handy references are the $10dB$ for each step in the intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} e \emph{fortissimo}.
-Other useful references are that equivalent in $dB$ of doubling amplitude or potency:
+For each $10dB$ it is associated a ``doubled volume''.
+A handy reference is $10dB$ for each step in the intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} e \emph{fortissimo}.
+Other useful references are $dB$ values related to double amplitude or potency:
 
@@
 
-and the amplitude gain for a sequence whose volume has been doubled ($10dB$) is:
+and the amplitude gain for a sequence whose volume has been doubled ($10dB$):
 
@@
 10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)} = 10 \quad \Rightarrow \\ \Rightarrow \quad \sum_{i=0}^{\lfloor \Delta.f_a \rfloor -1}t^{'2}_i=10\sum_{i=0}^{\Lambda-1}t_i^2=\sum_{i=0}^{\Lambda-1}(\sqrt{10}.t_i)^2 \\
-\therefore \quad t^{'}_i=\sqrt{10}t_i \quad \Rightarrow \quad t^{'}_i \approx 3,16t_i
+\therefore \quad t^{'}_i=\sqrt{10}t_i \quad \Rightarrow \quad t^{'}_i \approx 3.16t_i
 \end{split}
@@
 
-Thus, it is necessary a little bit more than triplicate the amplitude for a doubled volume.
+As shown, it is necessary an amplitude increase by a bit more than a factor of 3 for a doubled volume.
 These values are guides for increasing or decreasing the absolute values on the sample sequences with musical purposes.
@@
 
-As explained in the previous subsections, the musical particle (note) is a sequence $T_i$ in which duration and volume corresponds to the size of sequence and the amplitude of its samples.
+As exposed in the previous subsections, to a note corresponds a sequence $T_i$ in which duration and volume are directly related to the size of the sequence and the amplitude of its samples, respectively.
 The pitch is specified by the fundamental frequency $f_0$ whose cycle has duration $\delta_{f_0}=1/f_0$.
-This duration, multiplied by the sampling frequency $f_s$ results on the number of samples per cycle: $\lambda_{f_0}=f_a .
+This duration, multiplied by the sampling frequency $f_s$ results on the number of samples per cycle: $\lambda_{f_0}=f_s .
 \delta_{f_0} =f_a/f_0$.
 
-For didactic reasons, be $f_0$ such that it divides $f_s$ and $\lambda_{f_0}$ results an integer. If $T_i^f$ is a sonic sequence of fundamental frequency $f$, then:
+For didactic reasons, be $f_0$ such that it divides $f_s$ and $\lambda_{f_0}$ results integer. If $T_i^f$ is a sonic sequence of fundamental frequency $f$, then:
 
@@
 
-In the next section, frequencies $f$ that does not divide $f_s$ will be considered. We will see that this restriction does not imply in lost of generality of this section's content.
+In the next section, frequencies $f$ that does not divide $f_s$ will be considered. This restriction does not imply in lost of generality of this section's content.
 
@@
 
-While the wave period corresponds to the fundamental frequency, the trajectory of the wave inside the period - called the waveform - defines a harmonic spectrum and, thus, a timbre\footnote{Timbre is a subjective and complex characteristic.
-Physically, the timbre is multidimensional and given by the temporal dynamics behavior of energy in the spectral components that are harmonic or noisy.
+In a sound with harmonic spectrum, the (wave) period corresponds to the duration of the cycle, given by the inverse of the fundamental frequency.
+The trajectory of the wave inside the period - called the waveform - defines a harmonic spectrum and, thus, a timbre\footnote{Timbre is a subjective and complex characteristic.
+Physically, the timbre is multidimensional and given by the temporal dynamics of energy in the spectral components that are harmonic or noisy.
 Beyond that, the word timbre is used to designate different things: one same note has different timbres, a same instrument has different timbres, two instruments of the same family have, at the same time, the same timbre that blends them in the same family, and different timbres as they are different instruments.
@@
 
-The simplest (and most important case, as we will see in the following) is the spectrum that consists of only of the fundamental frequency $f$.
-This defines the sinusoid: a frequency in pure oscillatory movement called 'simple harmonic movement'.
-Let $S_i^f$ be a sequence whose samples $s_i^f$ describes a sinusoid of frequency $f$:
-
-\begin{equation}\label{senoide}
-     S^f_i=\{ s^f_i \}=\Bigl\{ \sin\bigl(2\pi \frac{i}{\lambda_f} \bigr)  \Bigr\} = \Bigl\{ \sin\bigl(2\pi f \frac{i}{f_a}\bigr)  \Bigr\} 
-\end{equation}
-
-where $\lambda_f=\frac{f_a}{f}=\frac{\delta_f}{\lambda_a}$  is the number of samples in the period.
-
-In a similar fashion, other waveforms are applied in music considering their spectral qualities and simplicity.
-While the sinusoid is an isolated point in the spectrum, these additional waves present a succession of harmonic components of the sinusoid.
+The simplest case is the spectrum that consists of only of the fundamental frequency $f$.
+This defines the sinusoid: a frequency in pure oscillatory movement called `simple harmonic movement'.
+Be $S_i^f$ a sequence whose samples $s_i^f$ describes a sinusoid of frequency $f$:
+
+\begin{equation}\label{sinusoid}
+     S^f_i=\{ s^f_i \}=\Bigl\{ \sin\bigl(2\pi \frac{i}{\lambda_f} \bigr)  \Bigr\} = \Bigl\{ \sin\bigl(2\pi f \frac{i}{f_s}\bigr)  \Bigr\} 
+\end{equation}
+
+where $\lambda_f=\frac{f_s}{f}=\frac{\delta_f}{\lambda_s}$  is the number of samples in the period.
+
+In a similar fashion, other waveforms are applied in music for their spectral qualities and simplicity.
+While the sinusoid is an isolated point in the spectrum, these waves present a succession of harmonic components.
 These waveforms are specified in equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} and illustrated in Figure~\ref{fig:formasDeOnda}.
-Although the artificial waveforms are traditionally used in music for synthesis and oscillatory control of variables, they are also useful outside music context\cite{Openheim}.
-
-The sawtooth presents all components of the harmonic series with decreasing energy of $-6dB/octave$. The sequence of temporal samples can be described as following:
+These artificial waveforms are traditionally used in music for synthesis and oscillatory control of variables. They are also useful outside musical contexts\cite{Openheim}.
+
+The sawtooth presents all components of the harmonic series with decreasing energy of $-6dB/octave$. The sequence of temporal samples can be described as:
 
@@
 
-The sawtooth is a common starting point for a subtractive synthesis, because it has both odd and even harmonics with high energy.
-For musical intention, these waveforms are excessively rich in sharp harmonics, and attenuator filtering on treble and middle parts of the spectrum is specially useful for reaching a more natural and pleasant sound.
-The relatively attenuated harmonics of the triangle wave makes it the more functional - among the listed ones - to be used in the synthesis of musical notes without any treatment.
-
 The square wave preservers only odd harmonics falling at $-6dB/octave$:
 
-\begin{equation}\label{quadrada}
+\begin{equation}\label{square}
      Q^f_i=\left\{ q^f_i \right\}= \left\{
@@
 This instrument has only the odd harmonic components and the square wave is convenient with its abundant energy in high frequencies.
+The sawtooth is a common starting point for a subtractive synthesis, because it has both odd and even harmonics with high energy.
+In general, these waveforms are appreciated as excessively rich in sharp harmonics, and attenuator filtering on treble and middle parts of the spectrum is specially useful for reaching a more natural and pleasant sound.
+The relatively attenuated harmonics of the triangle wave makes it the more functional - among the listed ones - to be used in the synthesis of musical notes without any treatment.
+The sinusoid is often a nice choice, but a problematic one.
+While pleasant if not loud in a very high pitch (above 500Hz, it requires careful dosage), the pitch is not accuratly detected, specially in low frequencies.
+It requires a big amplitude to give sinusoid volume, if compared to other waveforms.
+Both particularities are seen as a consequence of the non-existence of pure sinusoidal sounds in nature~\ref{roederer}.
 
@@
     \centering
-        \includegraphics[width=\columnwidth]{figures/waveForms}
+        \includegraphics[width=\textwidth]{figures/waveForms}
     \caption{Basic musical waveforms: (a) the synthetic  waveforms; (b) the real waveforms.}
@@
 Figure~\ref{fig:formasDeOnda} presents the waveforms described in equations  ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} and ~\ref{quadrada} for $\lambda_f=100$ (period of $100$ samples).
-If $f_s=44.1kHz$, a PCM standard in Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz$, reflecting the A4, just above the central "C", whatever the waveform is.
-
-The spectrum of each basic waveform is in Figure~\ref{fig:espectroDeondas}.
+If $f_s=44.1kHz$, the PCM standard in Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz$, around A4, just above the central "C", whatever the waveform is.
+
+The spectrum of each basic waveform is in Figure~\ref{fig:espectroDeOndas}.
 The isolated and exactly harmonic components of the spectrum is a consequence of the fixed period usage.
-The sinusoid consists of a one and only node in spectrum (**esquisito o texto**), pure frequency.
-As we said before, the sawtooth is the only waveform with a complete harmonic series (odd and even components).
-Triangular and square waves has the same components (odd harmonics), decaying at $-12dB/octave$ and $-6dB/octave$, respectively.
+The sinusoid consists of an one and only node in the spectrum, pure frequency.
+The figure exibit the spectra described: the sawtooth is the only waveform with a complete harmonic series (odd and even components); triangular and square waves has the same components (odd harmonics), decaying at $-12dB/octave$ and $-6dB/octave$, respectively.
 
@@
 
-The harmonic spectrum is formed by frequencies that are multiple from the fundamental frequency $f_n=(n+1)f_0$.
-As the human linear perception of pitch follows  a geometric progression of frequencies, the spectrum has notes different from the fundamental frequency.
-Additionally, the number of harmonics will be limited to the maximum frequency $f_s/2$ (by Nyquist's theorem).
-
-From a musical perspective, to internalize that energy in a component of frequency $f_n$ means an oscillation in the constitution of the sound, purely harmonic and in that frequency $f_n$.
-This energy, specifically concentrated on the $f_n$, is separated by the ear to enter in a cognitive level of processing (this separation is done in many species with mechanisms similar to the human cochlea~\cite{Roederer}).
+The harmonic spectrum is formed by frequencies that are multiple of the fundamental frequency $f_n=(n+1)f_0$.
+As the human linear perception of pitch follows  a geometric progression of frequencies, the spectrum has notes different from the fundamental frequency (see equation~\ref{eq:serieHarmonica}).
+Additionally, the number of harmonics will be limited by the Nyquist frequency $f_s/2$.
+
+From a musical perspective, it is critical to internalize that energy in a component of frequency $f_n$ means an oscillation in the constitution of the sound, purely harmonic and in that frequency $f_n$.
+This energy, specifically concentrated on the frequency $f_n$, is separated by the ear for further cognitive processes (this separation is done in many species with mechanisms similar to the human cochlea~\cite{Roederer}).
 
@@
 
+Although not a scale, the harmonic series is often used as such. Besides, this point is a convenient moment to put it plain simple:
+
+\begin{equation}\label{eq:serieHarmonica}
+\begin{split}
+H_i = \{h_i\}_0^{19}=\{ & 0,12,19+0.02,  24,28-0.14, 31+0.2, 34-0.31, \\
+                     & 36, 38+0.04,40-0.14, 42-0.49, 43+0.02, \\
+                     & 44+0.41, 46-0.31, 47-0.12, \\
+                     & 48, 49+0.05, 50+0.04, 51-0.02, 52-0.14   \}
+\end{split}
+\end{equation}
+
+This scale is composed of the frequencies on harmonic spectra. Nature exhibits this frequencies within sounds and with all kinds of distortions.
+
 \subsubsection{Chords}\label{subsec:acordes}

### 2013-09-28 article/article.tex
atualizado artigo, voltando aa traducao

@@
 
-The sinusoidal components are usually the main responsible for the timbre quality.
-If they are not presented in harmonic proportions (small number relations), the sound is perceived as noisy or dissonant, in opposite of a sonority with a unequivocally established fundamental frequency when they are present.
-Furthermore, the notion of absolute pitch in a complex sonority is based relies on the similarity of the spectrum to the harmonic series~\cite{Roederer} (**nao entendi o q vcs querem dizer**).
-
-In the case of a fixed length period and waveform, the spectrum is perfectly harmonic and static.
-Each waveform is compound of specific proportions of harmonic components and the greater the curvature of the part, the greater the contribution of the part to the energy on the high harmonics.
-This aspect can be seen on real sounds.
-The wave labeled as ``sampled real sound'' in Figure~\ref{fig:formasDeOnda} has a period of $\Lambda_f=114$ samples, extracted from a relatively well behaved real sound.
+The sinusoidal components are usually the main responsibles for timbre qualities.
+If they are not presented in harmonic proportions (small number relations), the sound is perceived as noisy or dissonant, in opposition to sonorities with an unequivocally established fundamental.
+Furthermore, the notion of absolute pitch relies on the similarity of the spectrum to the harmonic series~\cite{Roederer}.
+
+In the case of a fixed length period and waveform, the spectrum is perfectly harmonic and static and  each waveform is compound of specific proportions of harmonic components.
+High  curvatures are sign of high harmonics in the wave.
+Figure~\ref{fig:formasDeOnda} depicts a wave, labeled as ``sampled real sound'', with a period of $\Lambda_f=114$ samples, extracted from a relatively well behaved recorded sound.
 The oboe wave was sampled for A4 also in $44.1kHz$.

### 2013-09-28 article/article.tex
revisada parte 2.1 completa

@@
     \centering
-        \includegraphics[width=\textwidth]{figures/waveForms}
+        \includegraphics[width=\textwidth]{figures/waveForms_}
     \caption{Basic musical waveforms: (a) the synthetic  waveforms; (b) the real waveforms.}
@@
 
-The harmonic spectrum is formed by frequencies that are multiple of the fundamental frequency $f_n=(n+1)f_0$.
+The harmonic spectrum is composed of frequencies $f_n$ that are multiples of the fundamental frequency: $f_n=(n+1)f_0$.
 As the human linear perception of pitch follows  a geometric progression of frequencies, the spectrum has notes different from the fundamental frequency (see equation~\ref{eq:serieHarmonica}).
@@
 Figure~\ref{fig:formasDeOnda} depicts a wave, labeled as ``sampled real sound'', with a period of $\Lambda_f=114$ samples, extracted from a relatively well behaved recorded sound.
-The oboe wave was sampled for A4 also in $44.1kHz$.
-The chose period for sampling was a relatively short one, with $98$ samples, and corresponds to the frequency $\frac{44100}{98}=450Hz$.
-It can be noticed from the curvatures, the oboe's rich spectrum in high frequencies and the lower spectrum of the real sound.
+The oboe wave was also sampled in $44.1kHz$.
+The chosen period for sampling was relatively short, with $98$ samples, and corresponds to the frequency $\frac{44100}{98}=450Hz$, which corresponds to a slightly out-of-tune A4 pitch.
+One can notice from the curvatures: the oboe's rich spectrum in high frequencies and the lower spectrum of the real sound.
 
@@
 
-The resulting sound has the momentary spectrum of the original sound.
-Consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise and with variations typical of the natural phenomenon.
+The resulting sound has the momentary spectrum of the original waveform.
+As a consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise and with variations typical of the natural phenomenon.
 This can be observed in Figure~\ref{fig:espectroOboe}, that shows the spectrum of the original oboe note and a note with same duration and whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}.
-The natural spectrum exhibits variations in the frequencies of the harmonics, in their intensities and some noise.
-The note made from the sampled period has a perfectly harmonic spectrum.
+Summing up, the natural spectrum exhibits variations in the frequencies of the harmonics, in their intensities and some noise, while the note made from the sampled period has a perfectly harmonic spectrum.
 
@@
         \includegraphics[width=\columnwidth]{figures/oboeNaturalSampledSpectrum}
-    \caption{Spectrum of the sonic waves of a natural oboe note and one made from a sampled period.
+    \caption{Spectrum of the sonic waves of a natural oboe note and from a sampled period.
 The natural sound has fluctuations in the harmonics and in its noise, while the sampled period note has a perfectly harmonic spectrum.}
@@
 The presence and behavior of these sinusoidal components in the discretized sound have some particularities.
-Considering a signal $T_i$ and its corresponding Fourier decomposition $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$, the recomposition is the sum of the frequencies components as time samples\footnote{It is important to note that the factor $\frac{1}{\Lambda}$ could be distributed among the Fourier transform and its reconstruction, as preferred.}:
-
-\begin{multline}\label{recomposicaoFourier}
-t_i = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}c_ke^{j \frac{2\pi k}{\Lambda} i } \\ = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}(a_k+ j . b_k)\left[cos(w_k i)   +j . sen(w_k i)\right]
-\end{multline}
+Considering a signal $T_i$ and its corresponding Fourier decomposition $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$, the recomposition is the sum of the frequency components as time samples\footnote{It is important to note that the factor $\frac{1}{\Lambda}$ could be distributed among the Fourier transform and its reconstruction, as preferred.}:
+
+\begin{equation}\label{recomposicaoFourier}
+\begin{split}
+t_i = & \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}c_ke^{j \frac{2\pi k}{\Lambda} i } \\ 
+    = & \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}(a_k+ j . b_k)\left[cos(w_k i)   +j . sen(w_k i)\right]
+\end{split}
+\end{equation}
 
@@
 b_k$ defines the amplitude and phase of each frequency: $w_k=\frac{2\pi}{\Lambda}k$ in radians or $f_k=w_k\frac{f_a}{2\pi}=\frac{f_a}{\Lambda}k$ in Hertz, taking into account the respective limits in $\pi$ and in $\frac{f_a}{2}$ given by the Nyquist Theorem.
-
-Equation~\ref{recomposicaoFourier} shows how the imaginary term $c_k$ adds a phase to the real sinusoid.
-For example, the imaginary terms $b_k$ from the Fourier spectral decomposition make possible the phase sweep $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right]$ given by $tg^{-1}\left(\frac{b_k}{a_k}\right)$ which has this image.
-The signal from $a_k$ specifies the right and left side of the trigonometric circle, complementing the phase sweeping: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.
+As can be inferred, $j$ is the complex number, with $j^2=-1$.
 
@@
 
+Equation~\ref{moduloEfase} shows how the imaginary term of $c_k$ adds a phase to the real sinusoid: the terms $b_k$ enables the phase sweep $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right]$ given by $tg^{-1}\left(\frac{b_k}{a_k}\right)$ which has this image.
+The terms $a_k$ specifies the right or left side of the trigonometric circle, which completes the phase domain: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.
+
  \begin{figure}[h!]
@@
          \includegraphics[width=\columnwidth]{figures/amostras2c__}
-     \caption{Oscillation for 2 samples (maximum frequency for any $f_a$).
+     \caption{Oscillation of 2 samples (maximum frequency for any $f_a$).
 The first coefficient reflects a detachment (\emph{offset} or \emph{bias}) and the second coefficient specifies the oscillation amplitude.}
@@
 Figure~\ref{fig:amostras2} shows two samples and its spectral components.
-In this case, the Fourier decomposition has one unique pair of coefficients $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relatives to frequencies $\{f_k\}_0^1=\left\{w_k\frac{f_a}{2\pi}\right\}_0^1=\left\{k\frac{f_a}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_a}{2}=f_{\text{max}}\right\}$
+In this case, the Fourier decomposition has one unique pair of coefficients $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relative to frequencies $\{f_k\}_0^1=\left\{w_k\frac{f_a}{2\pi}\right\}_0^1=\left\{k\frac{f_a}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_a}{2}=f_{\text{max}}\right\}$
 with energies $e_k=\frac{(c_k)^2}{\Lambda=2}$.
-The role of amplitudes $a_k$ is clearly observed with $\frac{a_0}{2}$, the fixed detachment\footnote{Also called \emph{bias} or \emph{offset}.} and $\frac{a_1}{2}$, oscillation own amplitude given by the relation $f_k=k \frac{f_a}{\Lambda=2}$.
+The role of amplitudes $a_k$ is clearly observed with $\frac{a_0}{2}$, the fixed offset\footnote{Also called \emph{bias}.} and $\frac{a_1}{2}$, oscillation amplitude with frequency given by the relation $f_k=k \frac{f_a}{\Lambda=2}$.
 This case have special relevance.
-It is necessary the of 2 samples to represent an oscillation and it yields the Nyquist frequency $f_{\text{max}}=\frac{f_a}{2}$.
-That is the maximum frequency in a sound sampled with $f_a$ samples per second\footnote{Any sampled signal has this property, not only the digitalized sound.}.
-
-All fixed sequences $T_i$ of only $3$ samples also have just $1$ frequency, since their first harmonic has $1,5$ samples and exceeds the bottom limit of 2 samples, e.g.\ the frequency of the harmonic would exceed the Nyquist frequency:  $\; \frac{2.
+It is necessary at least 2 samples to represent an oscillation and it yields the Nyquist frequency $f_{\text{max}}=\frac{f_a}{2}$, which is the maximum frequency in a sound sampled with $f_a$ samples per second\footnote{Any sampled signal has this property, not only the digitalized sound.}.
+
+All fixed sequences $T_i$ of only $3$ samples also have just $1$ frequency, since their first harmonic has $1,5$ samples and exceeds the bottom limit of 2 samples, i.e.\ the frequency of the harmonic would exceed the Nyquist frequency:  $\; \frac{2.
 f_a}{3} > \frac{f_a}{2} $.
@@
 $\Lambda$ real samples $t_i$ result in $\Lambda$ complex coefficients $c_k=a_k+j.b_k$.
-The coefficients $c_k$ are equivalent two by two, corresponding to the same frequencies and with same relevance\footnote{Equal real part and imaginary with inverse order: $a_{k1}=a_{k2}$ and $b_{k1}=-b_{k2}$.
-As consequence the modules are equal and phases have inverse order.}.
+The coefficients $c_k$ are equivalent two by two, corresponding to the same frequencies and with same contribution to its reconstruction.
+They are complex conjugates: $a_{k1}=a_{k2}$ and $b_{k1}=-b_{k2}$ and, as a consequence, the modules are equal and phases have opposite signs.
 Remembering that $f_k = k\frac{f_a}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $.
@@
 
-The same could be observed in $w_k=f_k.\frac{2\pi}{f_a}$ and considering the periodicity $2\pi$, it results in $w_k=-w_{\Lambda-k}$.
-Given the cosine as an even function and the inverse tangent as an odd function, the components in $w_k$ and $w_{\Lambda-k}$ sums up in the equation that reconstructs the real samples, yielding equation~\ref{recomposicaoFourier}.
-
-In other words, in a decomposition of $\Lambda$ samples, the $\Lambda$ given frequencies components $\{c_i\}_0^{\Lambda-1}$ are equivalents in pairs.
-Except for $f_0$, and when having an even $\Lambda$, from $f_{\Lambda/2}=f_{\text{max}}=\frac{f_a}{2}$ both components are isolated, e.g.\ there is any other component in frequency $f_0$ or $f_{\Lambda/2}$ (if even $\Lambda$) regarding itself.
-It is true because $f_{\Lambda/2}=f_{(\Lambda-\Lambda/2) = \Lambda/2}$ and $f_0=f_{(\Lambda-0)=\Lambda}=f_0$.
-Furthermore, these two frequencies (zero and maximum frequency) are not represented having phase variation, being strictly real.
+The same can be observed with $w_k=f_k.\frac{2\pi}{f_a}$ and the periodicity $2\pi$: it follows that $w_k=-w_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
+Given the cosine (an even function) and the inverse tangent (an odd function), the components in $w_k$ and $w_{\Lambda-k}$ contribute with coefficients $c_k$ and $c_{\Lambda-k}$ in the reconstruction of the real samples.
+
+In other words, in a decomposition of $\Lambda$ samples, the $\Lambda$ frequency components $\{c_i\}_0^{\Lambda-1}$ are equivalents in pairs,
+except for $f_0$, and, when $\Lambda$ is even, for $f_{\Lambda/2}=f_{\text{max}}=\frac{f_a}{2}$.
+Both components are isolated, i.e.\ there is one and only component in frequency $f_0$ or $f_{\Lambda/2}$ (if even $\Lambda$).
+One can verify this assertion considering $k=0$ and $k=\Lambda/2$. Uncoiled: $f_{\Lambda/2}=f_{(\Lambda-\Lambda/2) = \Lambda/2}$ and $f_0=f_{(\Lambda-0)=\Lambda}=f_0$.
+Furthermore, these two frequencies (zero and Nyquist frequency) does not have phase variation, being their coefficients strictly real.
 In this way, it is possible to conclude the number $\tau$ of equivalent coefficient pairs:
@@
 
-This make clear visible the equivalences ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} and ~\ref{equivalenciasFases}:
+This discussion makes clear the equivalences ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} and ~\ref{equivalenciasFases}:
 
@@
 
-\begin{figure}[h!]
+\begin{figure*}
     \centering
-        \includegraphics[width=\columnwidth]{figures/amostras4__}
-    \caption{Frequential components for 4 samples.}
+        \includegraphics[width=\textwidth]{figures/amostras4___}
+    \caption{Frequency components for 4 samples.}
         \label{fig:amostras4}
-\end{figure}
-
-Having $a_k = a_{\Lambda -k}\;\;$ and $\;\;b_k = - b_{\Lambda -k}$:
+\end{figure*}
+
+$T_i \; \Rightarrow \; a_k = a_{\Lambda -k}\;\;$ and $\;\;b_k = - b_{\Lambda -k}$, and thus:
 
@@
 
-The equation~\ref{moduloEfase} for the real signal reconstruction, the modules and phases equivalences~\ref{equivalenciasModulos} and~\ref{equivalenciasFases}, the number of paired coefficients~\ref{coefsPareados}, and equivalence of paired frequencies~\ref{equivalenciasFreqs} expose a general case for components combination in each sample $t_i$:
+To expose the general case for components combination in each sample $t_i$, one can gather relations in equation~\ref{moduloEfase} for the real signal reconstruction, relations of modules~\ref{equivalenciasModulos} and phases equivalences~\ref{equivalenciasFases}, the number of paired coefficients~\ref{coefsPareados}, and equivalence of paired frequencies~\ref{equivalenciasFreqs}:
 
@@
 
-\begin{figure}[h!]
+with $a_{\Lambda/2}=0$ if $\Lambda$ odd.
+
+\begin{figure*}
     \centering
-        \includegraphics[width=\columnwidth]{figures/amostras4formas__}
-    \caption{Basic wave forms for 4 samples.}
+        \includegraphics[width=\textwidth]{figures/amostras4formas__}
+    \caption{Basic wave forms with 4 samples.}
         \label{fig:formas4}
-\end{figure}
-
-Therefore, as in Figure~\ref{fig:amostras3}, the Fourier transform of 3 samples have 2 frequencies coefficients with same amount of energy in the same frequency.
-
-With 4 samples it is possible to represent 1 or 2 frequencies with any proportions.
-Figure~\ref{fig:amostras4} depicts wave forms for 4 samples and its respective two components.
-The individual contributions sums together yielding the original wave form and a brief inspection reveals the major curvatures resulting from the higher frequency, while a fixed detachment of component summation results from the zero frequency component.
-
-\begin{figure}[h!]
+\end{figure*}
+
+With 4 samples it is possible to represent 1 or 2 frequencies in any proportions (i.e.
+with independence).
+Figure~\ref{fig:amostras4} depicts the basic waveforms with 4 samples and its two (possible) components.
+The individual contributions sum to the original waveform and a brief inspection reveals the major curvatures resulting from the higher frequency, while the fixed offset is captured in the zero frequency component.
+
+\begin{figure*}
     \centering
-        \includegraphics[width=\columnwidth]{figures/amostras6}
-    \caption{Frequencies components for 6 samples: 3 sinusoid sums with \emph{bias}.}
+        \includegraphics[width=\textwidth]{figures/amostras6_}
+    \caption{Frequency components for 6 samples: 4 sinusoids, one of them is the \emph{bias} with zero frequency.}
         \label{fig:amostras6}
-\end{figure}
-
-Figure~\ref{fig:formas4} shows the harmonics for 4 samples in basic wave forms of equations ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} and ~\ref{quadrada}.
-Their conjunction results in only 1 sinusoid, with exception to sawtooth wave, which have even harmonics.
-
-Figure~\ref{fig:amostras6} presents a sinusoidal decomposition for the 6 samples case and Figure~\ref{fig:formas6} presents the decomposition of the basic wave forms.
-In this case all wave forms are different in spectrum: square and triangular ones have the same components but with different proportions, while the sawtooth have an extra component.
+\end{figure*}
+
+Figure~\ref{fig:formas4} shows the harmonics for the basic waveforms of equations~\ref{senoide},~\ref{denteDeSerra},~\ref{triangular} and~\ref{quadrada} for the case of 4 samples.
+There is only 1 sinusoid for each waveform, with the exception of the sawtooth, which has even harmonics.
+
+Figure~\ref{fig:amostras6} presents the sinusoidal decomposition for 6 samples and figure~\ref{fig:formas6} presents the decomposition of the basic wave forms.
+In this case, the waveforms have spectra with fundamental differences: square and triangular have the same components but with different proportions, while the sawtooth have an extra component.
 
@@
         \includegraphics[width=\columnwidth]{figures/amostras6formas__}
-    \caption{Basic wave forms for 6 samples: triangular and square wave forms have odd harmonics, with different proportions and phases; the sawtooth wave also has even harmonics.}
+    \caption{Basic waveforms with 6 samples: triangular and square waveforms have odd harmonics, with different proportions and phases; the sawtooth has even harmonics.}
         \label{fig:formas6}
@@
 
-Consider $f$ presented in a such way that divides $f_a$\footnote{As pointed before, this limitation simplify the explanation without losing generality, and will be overcome in the next section.}.
-A sequence $T_i$ of sound samples are separated by $\delta_a=1/f_a$ and describes a musical note with a frequency of $f$ Hertz and $\Delta$ seconds of duration if, and only if, it has the periodicity equal to $\lambda_f=f_a/f$ and size $\Lambda=\lfloor f_a .
+Be $f$ such that it divides $f_a$\footnote{As pointed before, this limitation simplify the explanation without losing generality, and will be overcome in the next section.}.
+A sequence $T_i$ of sonic samples separated by $\delta_a=1/f_a$ describes a musical note with a frequency of $f$ Hertz and $\Delta$ seconds of duration if, and only if, it has the periodicity $\lambda_f=f_a/f$ and size $\Lambda=\lfloor f_a .
 \Delta \rfloor $:
@@
 
-The note by itself does not specifies a timbre.
-Nevertheless, it is necessary to choose the right waveform so that the samples $t_i$ have a value established individually.
-A unique period from the basic waveforms can be used to specify the note, where $\lambda_f=\frac{f_a}{f}$ is the number of samples at the period.
+The note by itself does not specify a timbre.
+Nevertheless, it is necessary to choose a waveform for the samples $t_i$ to have a value.
+An unique period from the basic waveforms can be used to specify the note, where $\lambda_f=\frac{f_a}{f}$ is the number of samples at the period.
 Here, $L_i^{f,\, \delta_f} $ is the sequence that describes a period of the waveform $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ with duration $\delta_f=1/f$ (as given by equations~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} and ~\ref{quadrada}) and $R_i^f$ is a sampled real waveform:
@@
 Although it is not one of its four basic properties, a musical note always has a spatial localization: the note source position at the ordinary physical space.
-In this context, the existence of an environment that reverbs the played note it is the matter of 'spatialization'.
+The existence of an environment that reverberates the note is the matter of 'spatialization'.
 Both fields, spatialization and spatial localization, are widely valued by audiophiles and music industry~\cite{floEsp}.
@@
 
-It is believed that the perception of spatial localization of sound occurs in our nervous system by three information: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering realized by the human body, including its chest, head and ears~\cite{Roederer, hrtf, Heeger}.
+It is understood that the perception of sound localization occurs in our nervous system by three informations: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering realized by the human body, including its chest, head and ears~\cite{Roederer, hrtf, Heeger}.
 
@@
         \includegraphics[width=.5\textwidth]{figures/espacializacao___}
-    \caption{Detection of sound source spatial localization: schema used to calculate Interaural Time Difference (ITD) and Interaural Intensity Difference (IID).}
+    \caption{Detection of sound source localization: schema used to calculate Interaural Time Difference (ITD) and Interaural Intensity Difference (IID).}
     \label{fig:spac}
@@
 
-Considering only the direct incidences in each ear the equations are quite simple.
-Given the distance between ears $\zeta$\footnote{It is known that $\zeta \approx 21,5cm$ to an adult human}, an object placed at $(x,y)$ as in Figure~\ref{fig:spac} is distant of each ear by:
+Considering only the direct incidences in each ear, the equations are quite simple. An object placed at $(x,y)$, as in figure~\ref{fig:spac}, is distant of each ear by:
 
@@
 
-\noindent and immediate calculations results in the Interaural Time Difference:
+Where $\zeta$ is the distance between ears $\zeta$,  known to be $\zeta \approx 21,5cm$ in an adult human.
+Immediate calculations result in the ITD:
 
@@
 
-Converting to amplitude it gives $IID_a=\frac{d}{d'}$.
+\noindent which, converted to amplitude, yields $IID_a=\frac{d}{d'}$.
 The $IID_a$ can be used as a multiplicative constant to the right channel of a stereo sound signal: $\{t_i'\}_0^{\Lambda -1}=\{IID_a .
-t_i\}_0^{\Lambda -1}$.
-It is possible to use IID together with ITD as a time advance for the right channel in relation with the left one.
-It is a crucial entail to localization in bass sounds and percussive sonorities~\cite{Heeger}.
-Considering $\Lambda_{ITD}=\lfloor ITD . f_a \rfloor$:
+t_i\}_0^{\Lambda -1}$, where $\{t_i'\}$ are samples of the wave incident in the left ear.
+It is possible to use IID together with ITD as a time advance for the right channel.
+It is a crucial vestige to localization perception of bass sounds and percussive sonorities~\cite{Heeger}.
+With $\Lambda_{ITD}=\lfloor ITD . f_a \rfloor$:
 
@@
 
-\noindent with $t_i$ as the right channel and $t_i'$ the left channel. If $\Lambda_{ITD} < 0 $, it is only needed to change $t_i$ by $t_i'$ and to use $\Lambda_{ITD}'= | \Lambda_{ITD} | $.
-
-Although simple until here, spatial localization depends considerably of other steps. Using ITD and IID it is possible to specify only the horizontal angle (azimuthal) $\theta$ given by:
+\noindent with $t_i$ as the right channel and $t_i'$ the left channel.
+If $\Lambda_{ITD} < 0 $, it is only needed to change $t_i$ by $t_i'$ and to use $\Lambda_{ITD}'= | \Lambda_{ITD} | $ and $IID_a'=1 / IID_a$.
+
+Although simple until here, spatial localization depends considerably of other cues. By using only ITD and IID it is possible to specify solely the horizontal angle (azimuthal) $\theta$ given by:
 
@@
 \noindent with $x,y$ as presented in Figure~\ref{fig:spac}.
-Henceforth, there are problems when $\theta$ focuses in the so called "cone of confusion": where the same pair of ITD and IID results in a large number of points inside the code.
-On those points the inference of azimuthal angle depends specially of the attenuate filtering for high frequencies, since the head interferes much more in the treble mechanical waves than in the bass waves~\cite{Heeger,hrtf}.
-It is also relevant to the hearing of source side, where the sound has enough low frequencies there is a diffraction and the wave comes to the ear with a delay of $\approx 0,7ms$.\cite{floEsp}
-
-Figure~\ref{fig:spac} also shows an acoustic shadow from cranium, an important phenomenon to perception of source azimuthal angle in the cone of confusion.
-The cone itself is not shown in Figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not encountered in the literature.
-This makes it hard to conceive given the filtering and diffraction dependent of the sound spectrum.
-In this way, the cone of confusion can be understood as a cone with its top side placed in the middle of head and growing out in the direction of each ear~\cite{hrtf}.
+Henceforth, there are problems when $\theta$ fall upon the so called "cone of confusion": the same pair of ITD and IID results in a large number of points inside the cone.
+On those points the inference of the azimuthal angle depends specially of the attenuative filtering for high frequencies, since the head interferes much more in the treble than in bass waves~\cite{Heeger,hrtf}.
+Also relevant to the hearing of lateral sources is that low frequencies diffracts and the wave arrive to the opposite ear with a delay of $\approx 0,7ms$.\cite{floEsp}
+
+Figure~\ref{fig:spac} depicts the acoustic shadow of the cranium, an important phenomenon to perception of source azimuthal angle in the cone of confusion.
+The cone itself is not shown in figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not encountered in the literature.
+Given the filtering and diffraction dependent of the sound spectrum, it is hard, if not impossible, to correctly draw the confusion cone.
+Even so, the cone of confusion can be understood as a cone with its top placed in the middle of head and growing out in the direction of each ear~\cite{hrtf}.
 
 On the other hand, the complete localization, including height and distance of sound source, is given by the Head Related Transfer Function (HRTF)~\cite{hrtf}.
-There are well known open databases of HRTF as CIPIC and it is possible to apply those transfer functions in a sound signal by convolution (see equation~\ref{eq:conv})~\cite{CIPIC}.
-The human body considerably changes the filtering and there are techniques to generate HRTFs to be universally used, as expected~\cite{lazaSPA}.
+There are well known open databases of HRTF, such as CIPIC, and it is possible to apply those transfer functions in a sonic signal by convolution (see equation~\ref{eq:conv})~\cite{CIPIC}.
+Each human body has its filtering and there are techniques to generate HRTFs to be universally used~\cite{lazaSPA}.
 
@@
 Spatialization results from sound reflections and absorptions by room/environment surface where the note is played.
-The sound propagates throw the air with a speed of $\approx 343,2m/s$ and can be sent from a source without any directionality pattern.
+The sound propagates through the air with a speed of $\approx 343,2m/s$ and can be emitted from a source with any directionality pattern.
 When a sound pulse encounters a surface there is reflection.
-In this reflection occurs: 1) the component inversion of propagation speed whose is normal to the surface, as 2) the energy absorption, specially in trebles.
-The sound waves propagate until they reach inaudible levels.
-The moment when some sound pulse reach the human ear can be described as the moment it reaches the ear and the absorption filters of each surface it has reached.
+In this reflection occurs: 1) the inversion of propagation speed component that is normal to the surface;  2) energy absorption, specially in trebles.
+The sonic waves propagate until they reaches inaudible levels (and even further).
+As a sonic front reaches the human ear, it  can be described as the original sound, with the last reflection point as the source, and the absorption filters of each surface it has reached.
 It is possible to simulate reverberations that are impossible in real systems.
-To experimentation, it is possible to use asymmetric reflections with relation to the axis perpendicular to the surface, or to increase specific frequency bands (known as 'resonances'), both characteristics are now found in real systems.
-
-There are some reverberation models that are less related to the calculation of each reflection, exploring valuable information of the auditory system.
-In fact, reverberation can be modeled with a set of 2 temporal and spectral characteristics:
+For example, it is possible to use asymmetric reflections with relation to the axis perpendicular to the surface, or to increase specific frequency bands (known as 'resonances'), both characteristics are not found in real systems.
+
+There are some reverberation models that are less related to each independent reflection, exploring valuable information to the auditory system.
+In fact, reverberation can be modeled with a set of 2 temporal and spectral sections:
 
@@
    \item Second period: 'late reverberation' is practically a dense succession of indistinct delays with exponential decay and statistical occurrences.
-   \item First band: bass has some resonance frequencies relatively spaced.
+   \item First band: the bass has some resonance bandwidths relatively spaced.
    \item Second band: mid and treble have progressive decay and smooth statistical fluctuations.
@@
 
-Smith III points out that reasonable concert rooms have total reverberation time of approx.
-$1,9$ seconds, and that the period of first reflections is around $0,1$ seconds.
-These values suggests that, in the given conditions, there are perceived wave pulses which propagates for $652,08$ meters long ($83,79k$ samples in $f_a=44,1kHz$) before reaching the ear.
-In addition, sound reflections made after propagation for $34,32$ meters long ($4,41k$ samples in $f_a=44,1kHz$) have a tangled form whose incidences are less distinct by hearing.
-These first reflections are particularly important to space sensation.
+Smith III points that reasonable concert rooms have total reverberation time of  $\approx 1.9$ seconds, and that the period of first reflections is around $0.1$ seconds.
+These values suggest that, in the given conditions, there are perceived wave pulses which propagates for $652.08$ meters long ($83.79k$ samples in $f_a=44.1kHz$) before reaching the ear.
+In addition, sound reflections made after propagation for $34.32$ meters long ($4.41k$ samples in $f_a=44.1kHz$) have incidences less distinct by hearing.
+These first reflections are particularly important to spacial sensation.
 The first incidence is the direct sound, described by ITD and IID of equations~\ref{eq:dti} and~\ref{eq:dii}.
-Admitting that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$ dependents of room dimension, the separation between the first reflection is at  $8-90ms$ ($\approx 350-4000$ samples in $f_a=44.1kHz$).
+Admitting that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, dependent of room dimension, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_a=44.1kHz$).
 It is possible to experimentally verify that the number of reflections increases with squared proportion $\approx k.n^2$.
-A discussion about the use of convolutions and filtering to favor these implementations is provided in subsection~\ref{subsec:mus2}, specially at paragraphs about reverberation.
+A discussion about the use of convolutions and filtering to favor implementation of these phenomena is provided in subsection~\ref{subsec:mus2}, specially at the paragraphs about reverberation.
 
@@
 
-Once we have well defined the basic note, it is time to build musical structures with sequences based on those particles.
-The sum of elements with same index of $N$ sequences $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ with same size $\Lambda$ results in overlapped spectral contents in a process called sound mixing:
+Once defined the basic note, it is convenient to build musical structures with sequences based on these particles.
+The sum of elements with same index of $N$ sequences $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ with same size $\Lambda$ results in the overlapped spectral contents of each sequence, in a process called mixing:
 
@@
 
-Figure~\ref{fig:mixagem} illustrates this overlapping process of discretized sound waves, disposing 100 samples.
+Figure~\ref{fig:mixagem} illustrates this overlapping process of discretized sound waves, each with 100 samples.
 It is possible to conclude that, if $f_a=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_a}{100/2}=882Hz$, $\frac{f_a}{100/4}=1764Hz$ and $\frac{f_a}{100/5}=2205Hz$.
 The samples duration are very short $\frac{f_a=44.1kHz}{100} \approx 2 \text{ milliseconds}$.
-Someone just need to complete the sequence with zeros to sum up sequences with different sizes.
-
-The mixed notes are generally separated by the ear according to the physic laws of resonance and by the nervous system~\cite{Roederer}.
-This musical notes mixing process results in musical harmony whose intervals between frequencies and chords of simultaneous notes guide subjective and abstract aspects of music and its appreciation~\cite{Harmonia}.
-
-The sequences can be concatenated in time.
-If the sequences $\{t_{k,i}\}_0^{\Lambda_k-1}$ of size $\Lambda_k$ represent $k$ musical notes, their concatenation in a unique sequence $T_i$ is a simple musical sequence or melody:
-
-\begin{multline}\label{eq:concatenacao}
-\{t_i\}_0^{\sum\Delta_k-1}=\{t_{l,i}\}_0^{\sum\Delta_k-1}, \;\; \\ l\text{ smallest integer } : \quad \Lambda_l > i -\sum_{j=0}^{l-1}\Lambda_j
-\end{multline}
+One can complete the sequence with zeros to sum (mix) sequences with different sizes.
+
+The mixed notes are generally separated by the ear according to the physical laws of resonance and by the nervous system~\cite{Roederer}.
+This musical notes mixing process results in musical harmony whose intervals between frequencies and chords of simultaneous notes guide subjective and abstract aspects of music and its appreciation~\cite{Harmonia}, which is tackled in section~\ref{sec:notesMusic}.
+
+Sequences can be concatenated in time.
+If the sequences $\{t_{k,i}\}_0^{\Lambda_k-1}$ of size $\Lambda_k$ represent $k$ musical notes, their concatenation in a unique sequence $T_i$ is a simple melodic sequence, or melody of its own:
+
+\begin{equation}\label{eq:concatenacao}
+\begin{split}
+\{t_i\}_0^{\sum\Delta_k-1}= & \{t_{l,i}\}_0^{\sum\Delta_k-1}, \;\; \\ l\text{ smallest integer } : & \quad \Lambda_l > i -\sum_{j=0}^{l-1}\Lambda_j
+\end{split}
+\end{equation}
 
 This mechanism is demonstrated in Figure~\ref{fig:concatenacao} with same sequences of Figure~\ref{fig:mixagem}.
-Although the sequences are short for the used sample rates, it is possible to observe the concatenation of sound sequences.
-Beside that, each note has it duration larger than $100ms$ if $f_a<1kHz$.
+Although the sequences are short for the usual sample rates, it is easy to observe the concatenation of sound sequences.
+Besides that, each note has a duration larger than $100ms$ if $f_a<1kHz$.
 
@@
 
-The musical piece \emph{reduced-fi} explores the temporal overlapping of notes, resulting in a homophobic piece.
-The vertical principle is demonstrated at the \emph{sound boards}, static sounds with peculiar spectrum.
-Both pieces were written in Python at Appendices~\ref{ap:quadros} and~\ref{ap:reduced}, and are available as part of the \massa \emph{toolbox}.\cite{MASSA}
-
-Now that the basic digital music note is described, the next section develops the temporal evolution of its contents as in \emph{glissandi} and volume envelopes.
-The filtering of spectral components and noise generation complements the definition of the musical note as a self contained unity.
-Section~\ref{notasMusica} is dedicated to the organization of these notes as music by using metrics and trajectories.
+The musical piece \emph{reduced-fi} explores the temporal juxtaposition of notes, resulting in a homophobic piece.
+The vertical principle is demonstrated at the \emph{sounic pictures}, static sounds with peculiar spectrum.
+Both pieces were written in Python and are available as part of the \massa\ \emph{toolbox}.\cite{MASSA}
+
+With the basic musical digital note carefully described, the next section develops the temporal evolution of its contents as in \emph{glissandi} and volume envelopes.
+Filtering of spectral components and noise generation complements the musical note as a self contained unity.
+Section~\ref{notasMusica} is dedicated to the organization of these notes as music by using metrics and trajectories, with regards to traditional music theory.
 
@@
 \begin{split}
-H_i = \{h_i\}_0^{19}=\{ & 0,12,19+0.02,  24,28-0.14, 31+0.2, 34-0.31, \\
+H_i = & \{h_i\}_0^{19}= \\
+      & \{ 0,12,19+0.02,  24,28-0.14, 31+0.2, 34-0.31, \\
                      & 36, 38+0.04,40-0.14, 42-0.49, 43+0.02, \\

### 2013-10-02 article/article.tex
adicionado artigo com tudo corrigido ateh metade da discussao das LUTs

@@
 The basic digital music note defined in section~\ref{sec:notaDisc} has following parameters: duration, pitch, intensity (volume) and timbre.
-This is a useful and paradigmatic modeling but does not covers all the aspects regarding a musical note.
-
-At first place, the characteristics of the note are modified along the duration of the note itself~\cite{Chowning}.
-For example, a note played in a piano with duration of 3 seconds has abrupt intensity at its beginning time of execution.
-Then, it behaves with progressively decays, spectrum variations, presenting harmonics decaying before others and presenting harmonics showing up along the time.
-Those variations are not mandatory but they are guidance from sound synthesis to musical uses, because it suitable reflects how sounds are represented in nature\footnote{The gold tip here is: to make a sound that incites interest by itself, do internal variations on it~\cite{Roederer}.}.
-To explore all the possible ways which those variations occur is out of the scope of this study, given the considerably sensibility of the human ear and the complexity of our sound cognition.
-In the following we point out primary resources to produce those characteristic variations in the basic note.
-It is worthwhile to mention that all the described relations in the following section are currently implemented in Python at Appendix~\ref{sec:cod2}.
-The music pieces \emph{Transita para metro}, \emph{Vibra e treme}, \emph{Tremolos, vibratos e a frequência}, \emph{Trenzinho de caipiras impulsivos}, \emph{Ruidosa faixa}, \emph{Bela rugosi}, \emph{Chorus infantil}, \emph{ADa e SaRa} are listed in Appendices~\ref{ap:transita}, \ref{ap:vibra}, \ref{ap:tremolos}, \ref{ap:trenzinho}, \ref{ap:ruidosa}, \ref{ap:bela}, \ref{ap:chorus} and \ref{ap:ada}.
-The code is also part of the toolbox \massa, available online.\cite{MASSA}
+This is a useful and paradigmatic model, but does not exhaust all the aspects of a musical note.
+
+First of all, characteristics of the note modifie along the note itself~\cite{Chowning}.
+For example, a 3 seconds piano note has intensity with abrupt rise at beginning and progressive decay, has spectrum variations with harmonics decaying some others that emerge along time.
+These variations are not mandatory but sound synthesis guidance for musical uses, as it reflects how sounds appear in nature.
+This is so considered to be true, that there is a rule of thumb: to make a sound that incites interest by itself, do internal variations on it~\cite{Roederer}.
+To explore all the ways in which variations occur within a note is out of the scope of any work, given the sensibility of the human ear and the complexity of human sound cognition.
+In the following, primary resources are presented to produce variations in the basic note.
+It is worthwhile to recall that all the relations in this and other sections are implemented in Python and published as the public domain \massa\ toolbox.
+The musical pieces \emph{Transita para metro}, \emph{Vibra e treme}, \emph{Tremolos, vibratos e a frequência}, \emph{Trenzinho de caipiras impulsivos}, \emph{Ruidosa faixa}, \emph{Bela rugosi}, \emph{Chorus infantil}, \emph{ADa e SaRa} were made to validate and illustrate concepts of this section.
+The code that synthesizes these pieces are also part of the toolbox\cite{MASSA}.
 
@@
 
-The \emph{Lookup Table} (or simply LUT), is a data structure array which substitutes continuous and repetitive calculation by indexed access.
-It is frequently used to reduce computational complexity and to make possible the use of functions without the possibility of direct calculation, as the data sampled from nature.
-In music its usage transcends the former, simplifying operations and making possible to use a single wave period to synthesize sounds in the whole audible spectrum, no matter what kind of wave was originally sampled.
+The \emph{Lookup Table} (or simply LUT), is an array for indexed operations which substitutes continuous and repetitive calculation.
+It is used to reduce computational complexity and to use functions without direct calculation, as data sampled from nature.
+In music its usage transcends these applications, as it simplifies many operations and makes possible the use a single wave period to synthesize sounds in the whole audible spectrum, with any waveform.
 
@@
         \includegraphics[width=\columnwidth]{figures/lut}
-    \caption{Search procedure in an indexed array (known as \emph{Lookup Table}) to synthesize sounds in different frequency using an unique waveform with high resolution.}
+    \caption{Search in the \emph{lookup table} to synthesize sounds in different frequencies using an unique waveform with high resolution.}
         \label{fig:lut}                                                                                                            
@@
 
-Being $\widetilde{\Lambda}$ the wave period size and $\widetilde{L_i} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ the elements $\widetilde{l_i}$ of any wave period (refer to Equation~\ref{periodoUnico}), a sequence $T_i^{f,\,\Delta}$ with samples of a sound with frequency $f$ and duration $\Delta$ can be obtained by means of $\widetilde{L_i}$ as:
-
-\begin{multline}\label{eq:lut}
+Be $\widetilde{\Lambda}$ the wave period in samples and $\widetilde{L_i} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ the samples $\widetilde{l_i}$ (refer to Equation~\ref{periodoUnico}), a sequence $T_i^{f,\,\Delta}$ with samples of a sound with frequency $f$ and duration $\Delta$ can be obtained by means of $\widetilde{L_i}$:
+
+\begin{equation}\label{eq:lut}
+\begin{split}
 T_i^{f,\,\Delta}=\left\{t_i^f\right\}_0^{\lfloor \, f_a .
-\Delta \, \rfloor -1} = \left\{ \, \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} }\, \right\}_{0}^{\Lambda-1}\; , \quad \\ \text{where} \;\; \gamma_i = \left \lfloor i .
+\Delta \, \rfloor -1} = & \left\{ \, \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} }\, \right\}_{0}^{\Lambda-1}\; , \quad \\ \text{where} \;\; \gamma_i = & \left \lfloor i .
 f \frac{ \widetilde{\Lambda}}{f_a} \right \rfloor
-\end{multline}
-
-In other words, with the right LUT indexes ($\gamma_i\%\widetilde{\Lambda}$) it is possible to synthesize a sound with any frequency.
-Figure~\ref{fig:lut} illustrates the calculation of a $\{t_i\}$ sample from $\left\{\,\widetilde{l}_i\,\right\}$ for a frequency of $f=200Hz$, $\widetilde{\Lambda}=128$ and considering the sample rate as $f_a=44.1kHz$.
+\end{split}
+\end{equation}
+
+In other words, with the right LUT indexes ($\gamma_i\%\widetilde{\Lambda}$) it is possible to synthesize sounds at any frequency.
+Figure~\ref{fig:lut} illustrates the calculation of $\{t_i\}$ sample from $\left\{\,\widetilde{l}_i\,\right\}$ for $f=200Hz$, $\widetilde{\Lambda}=128$ and adopting the sample rate of $f_s=44.1kHz$.
 Though this is not a practical configuration (as assigned below), it makes possible a graphical visualization of the procedure.
@@
 The calculation of the integer $\gamma_i$ introduces noise which decreases as $\widetilde{\Lambda}$ increases.
-In order to use this calculation in sound synthesis, with $f_a=44.1 kHz$ the standard guidelines suggest the use of $\widetilde{\Lambda} = 1024$ samples, since it does not produce relevant noise considering the audible spectrum.
-The round or interpolation method are not decisive in this process~\cite{Geiger}.
-
-The expression defining the variable $\gamma_i$ can be understood as $f_a$ being summed to $i$ at each $1$ second.
-If $i$ is divided by the sample frequency, it results in $\frac{i}{f_a}$,
-which is incremented in $1$ at each $1$ second.
-Multiplied by the period length it results in $i \frac{\widetilde{\Lambda}}{f_a}$ that look up the period in $1$ second.
+In order to use this calculation in sound synthesis, with $f_s=44.1 kHz$, the standard guidelines suggests the use of $\widetilde{\Lambda} = 1024$ samples, since it does not produce relevant noise on the audible spectrum.
+The rounding or interpolation method are not decisive in this process~\cite{Geiger}.
+
+The expression defining the variable $\gamma_i$ can be understood as $f_s$ being added to $i$ at each second.
+If $i$ is divided by the sample frequency, $\frac{i}{f_a}$
+is incremented in $1$ at each second.
+Multiplied by the period, it results in $i \frac{\widetilde{\Lambda}}{f_a}$, which covers the period in each second.
 Finally, with frequency $f$ it results in $i .
-f \frac{\widetilde{\Lambda}}{f_a}$ which complements $f$ lookups of the period $\widetilde{\Lambda}$ in $1$ second, i.e., the resulting sequence presents the fundamental frequency $f$.
-
-There are important considerations here: it is possible to use any value for $f$, the limits exist only in the loud frequencies when the size of table  $\widetilde{\Lambda}$ is not sufficient enough considering the sample rate $f_a$.
-The lookup table procedure is computationally costless and replaces the calculations by simple indexed searches (what is generally understood as an optimization process).
+f \frac{\widetilde{\Lambda}}{f_a}$ which completes $f$ periods $\widetilde{\Lambda}$ in $1$ second, i.e.
+the resulting sequence presents the fundamental frequency $f$.
+
+There are important considerations here: it is possible to use practically any frequency $f$.
+Limits exist only in low frequencies when the size of the table  $\widetilde{\Lambda}$ is not sufficient for the sample rate $f_a$.
+The lookup procedure is virtually costless and replaces calculations by simple indexed searches (what is generally understood as an optimization process).
 Unless otherwise stated, this procedure will be used along all the text for every applicable case.
-
 LUTs are broadly used in computational implementations for music.
-A classical usage of LUTs is known as \emph{Wavetable Synthesis}, which consists of many LUTs used together to generate a music note quasi-periodic~\cite{Cook,Wavetable}.
+A classical usage of LUTs is known as \emph{Wavetable Synthesis}, which consists of many LUTs used together to generate a quasi-periodic music note~\cite{Cook,Wavetable}.
 
@@
 
-Following the Weber and Fechner~\cite{Weber-Fechner} law, the human perception has a logarithmic relation with the causing stimulus.
-In other words, a stimulus following a exponential progression is perceived as linear.
-For didactic reasons and given its use in AM and FM synthesis (refer to subsection~\ref{subsec:tvaf}), the linear variation is discussed first.
-
-Considering the note with duration $\Delta = \frac{\Lambda}{f_a}$, the frequency $f=f_i$ varies linearly from $f_0$ to $f_{\Lambda -1}$. In this way, it is possible to define the following:
+As stated by the Weber and Fechner~\cite{Weber-Fechner} law, the human perception has a logarithmic relation with the stimulus.
+In other words, the exponential progression of a stimulues is perceived as linear.
+For didactic reasons, and given its use in AM and FM synthesis (subsection~\ref{subsec:tvaf}), linear variation is discussed first.
+
+Consider a note with duration $\Delta = \frac{\Lambda}{f_a}$, in which the frequency $f=f_i$ varies linearly from $f_0$ to $f_{\Lambda -1}$. Thus:
 
@@
 
-\begin{multline}\label{indiceLinear}
- \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i= \\ \left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{j}{\Lambda-1} \right ] \right \rfloor
-\end{multline}
+\begin{equation}\label{indiceLinear}
+\begin{split}
+ \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i= \left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   = \\
+=  \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{j}{\Lambda-1} \right ] \right \rfloor 
+\end{split}
+\end{equation}
 
@@
 
-where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a}$ is the LUT increment between two samples given the sound frequency in the first sample.
-
-Thus, it is possible to calculate the elements $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ having the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$.
+where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a}$ is the LUT increment between two samples given the sound frequency of the first sample.
+Therefore, it is handy to calculate the elements $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ by means of the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$.
 

### 2013-10-02 article/article.tex
feito tudo ateh os filtros digitais

@@
 \begin{split}
- \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i= \left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   = \\
-=  \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{j}{\Lambda-1} \right ] \right \rfloor 
+ \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i= \left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor \\
+\gamma_i=  \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{j}{\Lambda-1} \right ] \right \rfloor 
 \end{split}
@@
 
-The equations~\ref{freqLinear}, \ref{indiceLinear} and \ref{serieAmostralLin} are related with the linear progression of the frequency.
-We recall that, for the general case, the frequency progression \emph{perceived} as linear follows an exponential progression\footnote{In other words, a geometric progression of frequency is perceived as an arithmetic progression of pitch.}.
-It is possible to write this such that: $f_i=f_0 . 2^{\frac{i}{\Lambda-1} n_8}$ where  $n_8=\log_2\frac{f_{\Lambda-1}}{f_0}$ is the number of octaves between $f_0$ and $f_{\Lambda-1}$.
+The equations~\ref{freqLinear},~\ref{indiceLinear} and~\ref{serieAmostralLin} are related with the linear progression of the frequency.
+As stated above, the frequency progression \emph{perceived} as linear follows an exponential progression, i.e.
+a geometric progression of frequency is perceived as an arithmetic progression of pitch.
+It is possible to write: $f_i=f_0 . 2^{\frac{i}{\Lambda-1} n_8}$ where  $n_8=\log_2\frac{f_{\Lambda-1}}{f_0}$ is the number of octaves between $f_0$ and $f_{\Lambda-1}$.
 Therefore, $f_i=f_0 . 2^{\frac{i}{\Lambda-1}\log_2\frac{f_{\Lambda-1}}{f_0}}=
  f_0 . 2^{\log_2\left ( \frac{f_{\Lambda-1}}{f_0} \right )^{\frac{i}{\Lambda-1}}}=
- f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}$. Accordingly, the transition equations of linear frequency for the human ear are given by:
-
-\begin{multline}\label{freqExponencial}
- F_i=\{f_i\}_0^{\Lambda-1}= \\ \left\{f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}} \right\}_0^{\Lambda-1}
-\end{multline}
-
-\begin{multline}\label{indiceExponencial}
- \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \\ \quad \gamma_i=\left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor   =\left \lfloor \sum_{j=0}^{i} f_0 \frac{\widetilde{\Lambda}}{f_a} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{j}{\Lambda -1}} \right \rfloor
-\end{multline}
+ f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}}$. Accordingly, the equations for linear pitch transition are:
+
+\begin{equation}\label{freqExponencial}
+ F_i=\{f_i\}_0^{\Lambda-1}=  \left\{f_0 \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{i}{\Lambda -1}} \right\}_0^{\Lambda-1}
+\end{equation}
+
+\begin{equation}\label{indiceExponencial}
+\begin{split}
+ \Delta_{\gamma_i}= f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow  \quad \gamma_i= & \left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor \\ \gamma_i = & \left \lfloor \sum_{j=0}^{i} f_0 \frac{\widetilde{\Lambda}}{f_a} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{j}{\Lambda -1}} \right \rfloor
+\end{split}
+\end{equation}
 
@@
 
-The term $\frac{i}{\Lambda-1}$ looks up the interval $[0,1]$ and it is possible to power it to a value in such a way that the beginning of the transition will be smoother or steeper.
-This procedure is useful for energy variations of a vibratory wave to change the volume intensity\footnote{The volume change (psychophisical quality) is consequence of different sound characteristics like reverberation and concentration of treble harmonics, among which is wave energy.
-Wave energy is the easiest one to modify (see Equation~\ref{eq:potencia}) and it can vary in many ways.
-The most simple way consists of modifying the amplitude by multiplying the whole sequence by a real number.
-The increased energy without amplitude variation is the \emph{sound compression}, useful nowadays for music production~\cite{guillaume}}.
-It is sufficient to multiply the original sequence (being it generated or predetermined) by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is the given coefficient and $a_{\Lambda-1}$ is a fraction of the original amplitude, which is the aimed value to change until the end of transition.
+The term $\frac{i}{\Lambda-1}$ convers the interval $[0,1]$ and it is possible to power it to a value in such a way that the beginning of the transition will be smoother or steeper.
+This procedure is useful for energy variations with the purpose of changing the volume\footnote{The volume (psychophysical quality) is consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is wave total energy.
+Wave energy is the easy to modify (see Equation~\ref{eq:potencia}) and it can vary in many ways.
+The simplest way consists of modifying the amplitude by multiplying the whole sequence by a real number.
+The increase of energy without amplitude variation is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}}.
+It is sufficient to multiply the original sequence by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is the given coefficient and $a_{\Lambda-1}$ is a fraction of the original amplitude, which is the value to be reached at the end of transition.
 
@@
 
-\begin{multline}\label{seqAmp}
- \{a_i\}_0^{\Lambda-1}=\left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}= \\ \left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1} \text{ com } a_0=1
-\end{multline}
-
+\begin{equation}\label{seqAmp}
+\begin{split}
+ \{a_i\}_0^{\Lambda-1}= & \left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}= \\ = & \left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1} \text{ (with } a_0=1 \text{)}
+\end{split}
+\end{equation}
+and
 \begin{equation}\label{transAmp}
- T_i^{'}=T_i \odot A_i = \{t_i . a_i\}_0^{\Lambda-1}=\left \{ t_i . (a_{\Lambda-1} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
-\end{equation}
-
-It is possible to take $a_0=1$ in order to start a new sequence with the original amplitude and then keep changing it with the following samples.
-This restriction makes $a_{\Lambda-1}$ the amplitude variation term.
-If $\alpha=1$, the amplitude variation follows the geometric progression whose defines the linear progression.
-Figure~\ref{fig:transicao} shows the transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by equation~\ref{eq:ampVol}.
-
-An special attention should be driven while considering $a=0$.
+\begin{split}
+ T_i^{'}=T_i \odot A_i = & \{t_i . a_i\}_0^{\Lambda-1} \\ = & \left \{ t_i . (a_{\Lambda-1} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
+\end{split}
+\end{equation}
+
+It is often convenient to have $a_0=1$ in order to start a new sequence with the original amplitude and then progressively change it.
+If $\alpha=1$, the amplitude variation follows the geometric progression whose defines the linear variation of volume.
+Figure~\ref{fig:transicao} depicts transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by equation~\ref{eq:ampVol}.
+
+Special attention should be dedicated while considering $a=0$.
 In equation~\ref{seqAmp}, $a_0=0$ results in a division by zero and if $a_{\Lambda-1}=0$, there will be multiplication by zero.
-Both cases make the procedure useless, once no number different of zero can be represented as a ratio in relation of zero.
+Both cases make the procedure useless, once no number different from zero can be represented as a ratio in relation of zero.
 It is possible to solve this dilemma choosing a number that is small enough like $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum volume for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$).
-
-For linear amplification -- but not linear to perception -- it is sufficient to use an appropriate sequence $\{a_i\}$:
+A linear fade can be used to reach zero amplitude.
+Another common solution is the use of the quartic polynomial term $x^4$, as it reaches zero without these difficulties and gets reasonably close to the curve with $\alpha=1$ as it withdraws from zero~\cite{Cook}.
+
+For linear amplification -- but not linear perception -- it is sufficient to use an appropriate sequence $\{a_i\}$:
 
@@
 
-Here the conversation between decibels and amplitude is convenient. In this way, the equations~\ref{ampDec} and \ref{transAmp} specify a transition of $V_{dB}$ decibels:
+Here the convertion between decibels and amplitude is convenient. In this way, the equations~\ref{ampDec} and \ref{transAmp} specify a transition of $V_{dB}$ decibels:
 
@@
 The greater the value of $\alpha$, the smoother the sound introduction and more intense its end.
-$\alpha>1$ results in volume transitions commonly called \emph{slow fade}, while $\alpha<1$ results in \emph{fast fade}.\cite{guillaume}
-
-The linear transitions will be used for AM and FM synthesis and for applications of logarithmic transitions for tremolos and vibratos.
-A non-oscillatory exploration of those variations is in the music piece \emph{Transita para metro}, which code is in Appendix~\ref{ap:transita} and online at \massa.\cite{MASSA}
+$\alpha>1$ results in volume transitions commonly called \emph{slow fade}, while $\alpha<1$ results in \emph{fast fade}\cite{guillaume}.
+
+The linear transitions will be used for AM and FM synthesis, while logarithmic transitions are proper tremolos and vibratos, as developed in subsection~\ref{subsec:tvaf}.
+A non-oscillatory exploration of these variations is in the music piece \emph{Transita para metro}, which code is online as part of the \massa\ toolbox\cite{MASSA}.
 

### 2013-10-02 article/article.tex
feito ateh filtros IIR

@@
 
-This subsection is limited to a description of the procedure in sequences by convolution and differential equations, and the immediate applications.
-Its complexity exceeds to the scope of this study\footnote{The implementation of filters comprehends an area of recognized complexity with dedicated literature and software implementations.
-We recommended the reader to consult the bibliography~\cite{Openheim,smith}.}.
-The filters application can be part of the synthesis process or made individually as part of processes commonly called sound treatment.
-
-\begin{itemize}
-  \item  Convolution and finite impulse response (FIR) filters
+This subsection is limited to a description of sequences processing, by convolution and differential equations, and immediate applications, as its complexity escapes the scope of this study\footnote{The implementation of filters comprehends an area of recognized complexity, with dedicated literature and software implementations.
+The reader is welcome to visit the bibliography for further discussion~\cite{Openheim,smith}.}.
+Filter applications can be part of the synthesis process or made subsequently as part of processes commonly called ``sound treatment''.
+\subsubsection{Convolution and finite impulse response (FIR) filters}
 
@@
          \includegraphics[width=\columnwidth]{figures/convolucao}
-     \caption{Graphical interpretation of convolution.
-Each resulting sample is the sum of the previous samples of a signal, with each one multiplied by the retrograde sequence of samples from the other signal.}
+     \caption{Graphical interpretation of convolution. Each resulting sample is the sum of the previous samples of a signal, with each one multiplied by the retrograde of the other sequence.}
          \label{fig:conv}
@@
 
-Filters applied by means of convolution are known by acronym FIR (Finite Impulse Response) and are characterized by having a time finite sample representation.
-This sample representation is called 'impulse response' $\{h_i\}$.
-The FIR filters are applied to the time domain of the digital sound by means of convolution with the respective impulse response of the filter\footnote{It is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both sound and the impulse response, and then performing the inverse Fourier transform in the resulting spectrum.\cite{Openheim}}.
-For this study, the convolution is defined as:
-
-\begin{multline}\label{eq:conv}
+Filters applied by means of convolution are known by the acronym FIR (Finite Impulse Response) and are characterized by having a finite sample representation.
+This sample representation is called `impulse response' $\{h_i\}$.
+FIR filters are applied in the time domain of digital sound by means of convolution with the respective impulse response of the filter\footnote{It is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both sound and the impulse response, and then performing the inverse Fourier transform in the resulting spectrum.\cite{Openheim}}.
+For the purposes of this work, the convolution is defined as:
+
+\begin{equation}\label{eq:conv}
  \begin{split}
- \left\{t_i'\right\}_0^{\Lambda_t+\Lambda_h-2\; = \;\Lambda_{t\, '}-1} =\{(T_j*H_j)_i\}_0^{\Lambda_{t \, '}-1} & = \\ \left \{ \sum_{j=0}^{min(\Lambda_h-1,i)}h_{j} .
-t_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
-     & = \\ \left \{ \sum_{j=max(i+1-\Lambda_h,0)}^{i}t_j . h_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
+ \left\{t_i'\right\}_0^{\Lambda_t+\Lambda_h-2\; = \;\Lambda_{t\, '}-1} = & \{(T_j*H_j)_i\}_0^{\Lambda_{t \, '}-1} \\ = & \left \{ \sum_{j=0}^{min(\Lambda_h-1,i)}h_{j} t_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
+     \\ = & \left \{ \sum_{j=max(i+1-\Lambda_h,0)}^{i}t_j h_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
  \end{split}
-\end{multline}
-
-\noindent where $t_i=0$ for the samples not given beforehand.
-In other words, the sound $\{t_i'\}$ resulting from the convolution of $\{t_i\}$ with the given impulse response $\{h_i\}$ has each $i$-th sample $t_i$ overwritten by the sum of its last $\Lambda_h$ samples $\{t_{i-j}\}_{j=0}^{\Lambda_h-1}$ multiplied one-by-one by samples of the another impulse response $\{h_i\}_0^{\Lambda_h-1}$.
-This procedure is illustrated in Figure~\ref{fig:conv}, where the impulse response $\{h_i\}$ is traveled by its retrograde form, and $t_{12}'$ and $t_{32}'$ are two calculated samples using the convolution given by $(T_j*H_j)_i=t_i'$.
-The resulting signal always has the length of $\Lambda_t+\Lambda_h -1=\Lambda_{t'}$.
-
-With this procedure it is possible to apply reverbs, equalizers, \emph{delays} to name a few of a variety of other filters for sound processing, as well as musical/artistic effects.
+\end{equation}
+
+\noindent where $t_i=0$ for the samples not given.
+In other words, the sound $\{t_i'\}$, resulting from the convolution of $\{t_i\}$, with the impulse response $\{h_i\}$, has each $i$-th sample $t_i$ overwritten by the sum of its last $\Lambda_h$ samples $\{t_{i-j}\}_{j=0}^{\Lambda_h-1}$ multiplied one-by-one by samples of the impulse response $\{h_i\}_0^{\Lambda_h-1}$.
+This procedure is illustrated in Figure~\ref{fig:conv}, where the impulse response $\{h_i\}$ is in its retrograde form, and $t_{12}'$ and $t_{32}'$ are two calculated samples using the convolution given by $(T_j*H_j)_i=t_i'$.
+The final signal always has the length of $\Lambda_t+\Lambda_h -1=\Lambda_{t'}$.
+
+With this procedure it is possible to apply reverberators, equalizers, \emph{delays}, to name a few of a variety of other filters for sound processing and to obtain musical/artistic effects.
 
 The impulse response can be provided by physical measures or by pure synthesis.
-An impulse response for reverb application can be obtained by sound recording of the environment when someone triggers a snap which resembles an impulse or a sinusoidal lookup whose Fourier transform approximates its frequency response.
-Both are impulse responses which, properly convoluted with the sound sequence, result in the own sequence with a reverberation that resembles the physical one in the original environment where the measure happened~\cite{Cook}.
-
-The inverse Fourier transform of an envelope even and real is equivalent to a impulse response of a FIR filter.
-It performs a frequency filtering with the envelope.
-The greater the number of samples, the higher the envelope resolution and the computational complexity as well.
-The convolution is recognized as computationally expensive.
+An impulse response for a reverberation application, for example, can be obtained by sound recording of the environment when someone triggers a snap which resembles an impulse, or obtained by a sinusoidal sweep whose Fourier transform approximates its frequency response.
+Both are impulse responses which, properly convoluted with the sound sequence, result in the own sound with a reverberation that resembles the original environment where the measure was made~\cite{Cook}.
+
+The inverse Fourier transform of an even and real envelope is an impulse response of a FIR filter.
+Convoluted with a sound, it performs the frequency filtering specified by the envelope.
+The greater the number of samples, the higher the envelope resolution and the computational complexity, which should often be weighted, for the convolution is expensive.
 
 An important property is the time shift caused by convolution with a shifted impulse.
-Despite computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse to each relapse of sound.
-In Figure~\ref{fig:delays} it is possible to observe the shifting caused by convolution with each impulse.
-Depending on the intensity of the impulses, the result is perceived in rhythm (around 20 impulses by second) or in pitch (around 20-40 impulses by second).
-In the last case, the processes resemble granular synthesis, delays, reverbs and equalization.
+Despite computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse for each reincidence of the sound.
+In figure~\ref{fig:delays} it is possible to observe the shift caused by convolution with an impulse.
+Depending on the intensity of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as pitch (from about 20 impulses per second and higher frequencies).
+In the latter case, the process resembles granular synthesis, reverbs and equalization.
 
@@
         \includegraphics[width=\columnwidth]{figures/delays}
-    \caption{Convolution with impulse: shifting (a), delay lines (b) and granular synthesis~(c). Disposed in increasing order of its pulse density.}
+    \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c). Disposed in increasing order of its pulse density.}
         \label{fig:delays}
@@
 
-\item Infinite impulse response (IIR) filters
+\subsubsection{Infinite impulse response (IIR) filters}
 
@@
 Both possibilities are explored by the referenced studies, in special~\cite{JOSFM,smith}, and the collection of filters maintained by the \emph{Music-DSP} community from University of Columbia.\cite{music-dsp,Openheim}
-
-\end{itemize}
 

### 2013-10-02 article/article.tex
feito ateh as receitas IIR

@@
 
-This class of filters is known by the acronym IIR.
-They are characterized by having an infinite time representation, i.e.\ the impulse response does not converges to zero.
+This class of filters, known by the acronym IIR, are characterized by having an infinite time representation, i.e.\ the impulse response does not converge to zero.
 Its application is usually made by the following equation:
@@
 
-\noindent with $b_0=1$ in the most cases it is possible to normalize the variables: $a_j'=\frac{a_j}{b_0}$ and $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
-Equation~\ref{eq:diferencas} is called 'difference equation' because the resulting samples $\left\{t_i'\right\}$ are given by differences between the original samples $\{t_i\}$ and the previous resulting ones $\left\{t_{i-k}'\right\}$.
-
-There are many methods and tools to prepare IIR filters.
-The following texts list a selection of them for didactic purpose and to make easy future consulting.
-They are well behaved filters whose aspects are described in Figure~\ref{fig:iir}.
-
-Considering filters with simple order, the cutoff frequency $f_c$ is the position where the filter performs an attenuation of $-3dB \approx 0.707 $ in the original amplitude.
-For band-pass and band-reject (or 'notch') filters, this attenuation results on two specifications: $f_c$ (in this case, referred as 'center frequency') band width $bw$,
-in both frequencies $f_c \pm bw$ there is an attenuation of $\approx 0.707$ in the original amplitude.
+In the most cases it is possible to normalize the variables: $a_j'=\frac{a_j}{b_0}$ and $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
+Equation~\ref{eq:diferencas} is called `difference equation' because the resulting samples $\left\{t_i'\right\}$ are given by differences between original samples $\{t_i\}$ and previous resulting ones $\left\{t_{i-k}'\right\}$.
+
+There are many methods and tools to obtain IIR filters.
+The text bellow lists a selection of them for didactic purposes and as a reference.
+They are well behaved filters whose aspects are described in figure~\ref{fig:iir}.
+
+For filters of simple order, the cutoff frequency $f_c$ is where the filter performs an attenuation of $-3dB \approx 0.707 $ of the original amplitude.
+For band-pass and band-reject (or 'notch') filters, this attenuation has two specifications: $f_c$ (in this case, the `center frequency') and bandwidth $bw$.
+In both frequencies $f_c \pm bw$ there is an attenuation of $\approx 0.707$ of the original amplitude.
 There is sound amplification in band-pass and band-reject filters when the cutoff frequency is low and the band width is large enough.
-In trebles those filters present only a shift in the expected waveform, expanding the envelope to the bass side of the band.
-
-For filters whose frequency responses have other envelopes (for the module), it is possible to create chains and apply them successively.
-Another possibility is to use some biquad 'filter receipt'\footnote{Short for 'biquadratic': its transfer function has two poles and two zeros, i.e.
+In trebles, those filters present only a deviation of the expected profile, expanding the envelope to the bass.
+
+For filters with other frequency responses, it is possible to apply them successively.
+Another possibility is to use a biquad 'filter receipt'\footnote{Short for 'biquadratic': its transfer function has two poles and two zeros, i.e.
 its first direct form consists of two quadratic polynomials forming the fraction: $\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1- b_1.z^{-1} -b_2 .
-z^{-2}}$.} or procedures for the calculation of coefficients of Chebichev filters\footnote{Butterworth and Elliptical filters can be considered as specific cases of Chebichev filters.\cite{Openheim,smith}}.
-Both possibilities are explored by the referenced studies, in special~\cite{JOSFM,smith}, and the collection of filters maintained by the \emph{Music-DSP} community from University of Columbia.\cite{music-dsp,Openheim}
+z^{-2}}$.} or the calculation of Chebichev filter coefficients\footnote{Butterworth and Elliptical filters can be considered as special cases of Chebichev filters~\cite{Openheim,smith}.}.
+Both alternatives are explored by referenced studies, in special~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University .\cite{music-dsp,Openheim}
 

### 2013-10-02 article/article.tex
ateh os ruidos

@@
 z^{-2}}$.} or the calculation of Chebichev filter coefficients\footnote{Butterworth and Elliptical filters can be considered as special cases of Chebichev filters~\cite{Openheim,smith}.}.
-Both alternatives are explored by referenced studies, in special~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University .\cite{music-dsp,Openheim}
-
-\begin{enumerate}
-  \item Low-pass with simple pole and module of the frequency response in the upper left corner of Figure~\ref{fig:iir}.
-The general equation has as reference by the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_a$ that there is approximately an attenuation of $3dB$.
-  The coefficients $a_0$ and $b_1$ of the IIR filter are given by means of the intermediate variable $x \in [e^{-\pi},1]$:
+Both alternatives are explored by referenced studies, in special~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University~\cite{music-dsp,Openheim}.
 
@@
 
+\begin{enumerate}
+  \item Low-pass with a simple pole, module of the frequency response in the upper left corner of figure~\ref{fig:iir}.
+The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_s$ in which occurs an attenuation of $3dB$.
+  The coefficients $a_0$ and $b_1$ of the IIR filter are given by means of the intermediate variable $x \in [e^{-\pi},1]$:
+
 \begin{equation}\label{eq:passa-baixas}
@@
 
-  \item High-pass filter with simple pole and modules of its frequency responses at the upper right corner of Figure~\ref{fig:iir}.
+  \item High-pass filter with a simple pole,  module of its frequency responses at the upper right corner of figure~\ref{fig:iir}.
 The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of the intermediate variable $x \in [e^{-\pi},1]$:
@@
 \item Notch filter.
-This filter is parametrized by a center frequency\footnote{Attention with the cutoff frequency $f_c$ in low-pass and high-pass filters.} $f_c$ and band width $bw$ -- $f_c \pm bw$ that results in $0.707$ of the amplitude, i.e.
-attenuation of $3dB$ -- both given as fractions of $f_a$, therefore $f,\; bw \in (0,0.5)$.
-
-In order to simplify, consider the auxiliary variables $K$ and $R$ defined as:
+This filter is parametrized by a center frequency $f_c$ and bandwidth $bw$, both given as fraction of $f_s$, therefore $f,\; bw \in (0,0.5)$.
+Both frequencies $f_c \pm bw$ has $\approx 0.707$ of the amplitude, i.e.
+attenuation of $3dB$.
+Auxiliary variables $K$ and $R$ are defined as:
 
@@
 
-The band-pass filter in the lower left corner of Figure~\ref{fig:iir} has the following coefficients in equation~\ref{eq:diferencas}:
+The band-pass filter in the lower left corner of figure~\ref{fig:iir} has the following coefficients:
 
@@
 
-The coefficients of band-reject filter are:
+The coefficients of band-reject filter, depicted in the lower right of figure~\ref{fig:iir}, are:
 
@@
 
-\noindent with the module of its frequency response represented in the lower left corner of the Figure~\ref{fig:iir}.
-
 %\item Biquad: pela especificação de uma frequência central, da qualidade
@@
 Generally, sounds without a defined pitch are called noise~\cite{Lacerda}.
-They are important parts of the musical sounds with undefined pitch like the noise present in piano notes, violin, etc.
-Besides that, the majority of percussion instruments do not have defined pitch and their sounds are generally comprehended as noise~\cite{Roederer}.
-In electronic music, including electro-acoustic and dance genres, noise has diversified use, and commonly characterizes the music style~\cite{Cook}.
+They are important musical sounds, as noise present in piano notes, violin, etc.
+Besides that, the majority of percussion instruments do not have precise pitch and their sounds are generally comprehended as noise~\cite{Roederer}.
+In electronic music, including electro-acoustic and dance genres, noise has diverse uses, and frequently characterizes the music style~\cite{Cook}.
 

### 2013-10-02 article/article.tex
arrumadas figuras anteriores que estavam pequenas e arrumado texto ateh os ruidos individuais

@@
 
-\begin{figure}[h!]
+\begin{figure*}
     \centering
-        \includegraphics[width=\columnwidth]{figures/lut}
+        \includegraphics[width=\textwidth]{figures/lut}
     \caption{Search in the \emph{lookup table} to synthesize sounds in different frequencies using an unique waveform with high resolution.}
         \label{fig:lut}                                                                                                            
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[h!]
+\begin{figure*}
      \centering
-         \includegraphics[width=\columnwidth]{figures/convolucao}
+         \includegraphics[width=\textwidth]{figures/convolucao}
      \caption{Graphical interpretation of convolution. Each resulting sample is the sum of the previous samples of a signal, with each one multiplied by the retrograde of the other sequence.}
          \label{fig:conv}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[h!]
+\begin{figure*}
     \centering
-        \includegraphics[width=\columnwidth]{figures/delays}
+        \includegraphics[width=\textwidth]{figures/delays}
     \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c). Disposed in increasing order of its pulse density.}
         \label{fig:delays}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[h!]
+\begin{figure*}
     \centering
-        \includegraphics[width=\columnwidth]{figures/iir}
+        \includegraphics[width=\textwidth]{figures/iir}
     \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of equations~\ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} and \ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
         \label{fig:iir}
-\end{figure}
+\end{figure*}
 
@@
 
-Generally, sounds without a defined pitch are called noise~\cite{Lacerda}.
+Generally, sounds without an explicit pitch are called noise~\cite{Lacerda}.
 They are important musical sounds, as noise present in piano notes, violin, etc.
-Besides that, the majority of percussion instruments do not have precise pitch and their sounds are generally comprehended as noise~\cite{Roederer}.
-In electronic music, including electro-acoustic and dance genres, noise has diverse uses, and frequently characterizes the music style~\cite{Cook}.
-
-The absence of a defined pitch is due to the absence of a harmonic organization in the sinusoidal components that compose the sound.
-In this way, there are many ways to generate noise.
-The use of random values to generate the sound sequence $T_i$ is an attractive method but the result is generally not useful because it tends to the white noise~\cite{Cook}.
-
-Another way is to generate noise using the desired spectrum, from which it is possible to perform the inverse Fourier transform.
-The spectral distribution should be done with care because, if it uses the same phase or phases with strong correlation, the synthesized sound will include considerably amount of concentrated energy in some periods of its duration.
-
-\begin{figure}[htpq!]
+Besides that, the majority of percussion instruments does not exhibit an unequivocal pitch and their sounds are generally regarded as noise~\cite{Roederer}.
+In electronic music, including electro-acoustic and dance genres, noise has diverse uses and frequently characterizes the music style~\cite{Cook}.
+
+The absence of a definite pitch is due to the lack of a perceptible harmonic organization in the sinusoidal components of the sound.
+Hence, there are many ways to generate noise.
+The use of random values to generate the sound sequence $T_i$ is an attractive method but not outstandingly useful because it tends to produce white noise with little or no variations~\cite{Cook}.
+
+Another way is to generate noise is by using the desired spectrum, from which it is possible to perform the inverse Fourier transform.
+The spectral distribution should be done with care: if phases of components present proeminent correlation, the synthesized sound will concentrate energy in some periods of its duration.
+
+\begin{figure*}
      \centering
-         \includegraphics[width=\columnwidth]{figures/ruidos}
-     \caption{Colors of noise generated by equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul} and \ref{eq:violeta}: resulting spectrum and waveforms.}
+         \includegraphics[width=\textwidth]{figures/ruidos}
+     \caption{Colors of noise generated by equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul} and \ref{eq:violeta}: spectrum and waveforms.}
          \label{fig:ruidos}
-\end{figure}
-
-Some noise with static spectrum are listed below. They are called \emph{colors of noise} since they are associated with colors.
-Figure~\ref{fig:ruidos} shows the spectrum profile and the corresponding sound sequence side-by-side.
-All the noise were generated with a same phase, making it possible to observe the contributions of different regions of the spectrum.
+\end{figure*}
+
+Some noise with static spectrum are listed below. They are called \emph{colored noise} since they are associated with colors.
+Figure~\ref{fig:ruidos} shows the spectrum profile and the corresponding sonic sequence side-by-side.
+All five noises were generated with the same phase for each component, making it possible to observe the contributions of different parts of the spectrum.
 

### 2013-10-02 article/article.tex
feita revisao ateh tremolo, vibr, am e fm

@@
 
- \item The white note has its name because its energy are distributed equally between all the frequencies.
-It is possible to realize the white noise with the inverse transform of the following coefficients:
+ \item The white note has its name because its energy is distributed equally among all frequencies. It is possible to obtain white noise with the inverse transform of the following coefficients:
 
@@
  \begin{split}
- c_0 & =0 \quad \text{avoiding bias} \\
- c_i & =e^{j.x}\;,\;\; j^2=-1 \;, \;\; x \; \text{random} \; \in \; [0,2\pi]\;,\;\; i \; \in \; \left[1, \, \frac{\Lambda}{2}-1\right] \\
- c_{\Lambda/2} & = 1 \quad\quad \text{(if even $\Lambda$)}\\ 
+ c_0 & =0 \;\text{,\quad to avoid bias} \\
+ c_i & =e^{j.x}\;,\; x \; \text{random} \; \in \; [0,2\pi]\;,\; i \; \in \; \left[1, \, \frac{\Lambda}{2}-1\right] \\
+ c_{\Lambda/2} & = 1 \; \text{, \; if $\Lambda$ even}\\ 
  c_i & = c_{\Lambda - i}^*\;,\;\; \text{for}\;  i \; > \;  \frac{\Lambda}{2}
@@
 
-The value $c_i$, calculated by the exponential, is only an artifice to obtain unitary module and random phase. Besides that, $c_{\Lambda/2}$ is always real (as discussed in the previous section).
-
- \item The pink noise has a decrease of $3dB$ for octave. This noise is useful for testing electronic devices, and it has prominent presence in nature~\cite{Roederer}.
+The exponential $e^{j.x}$, is an artifice to obtain unitary module and random phase for the value of $c_i$. Besides that, $c_{\Lambda/2}$ is always real (as discussed in the previous section).
+
+ \item The pink noise is charachterized by a decrease of $3dB$ per octave. This noise is useful for testing electronic devices, and it has prominent presence in nature~\cite{Roederer}.
 
@@
  c_i  =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \\
- c_i  =e^{j.x} . \alpha_i\;,\;\; j^2=-1 \;, \;\;\ \\ x \;\; \text{random} \; \in \; [0,2\pi]\;,\;\; \forall \; i \; : f_{\text{min}} \le f_i < f_{\lceil \Lambda/2-1 \rceil}  \\
- c_{\Lambda/2}  = \alpha_{\Lambda/2} \quad \text{(se $\Lambda$ par)} \\ 
+ c_i  =e^{j.x} . \alpha_i\;, \; x \; \text{random} \; \in \; [0,2\pi]\;,\;\; \forall \; i \; : f_{\text{min}} \le f_i < f_{\lceil \Lambda/2-1 \rceil}  \\
+ c_{\Lambda/2}  = \alpha_{\Lambda/2}\;, \; \text{if $\Lambda$ even} \\ 
  c_i  = c_{\Lambda - i}^*\;,\;\; \text{for}\;  i \; > \;  \Lambda/2 \qquad \qquad
@@
 
-The minimum frequency $f_{\text{min}}$ can be chosen based on the human hearing limit, since no one listens to a pitch with a sound component which frequency is bellow $\approx\; 20Hz$.
-
-Other noises can be made from the pink noise procedure by simply modifying some details, specially the equation that defines $\alpha_i$.
-
-  \item The brown noise received this name after Robert Brown, who described the brownian movement\footnote{Although its origin is disparate given its association with the brown color, this noise become established with this specific name.
-  Anyway, this association is satisfactory once the white and pink noises are more strident and associated with intense colors~\cite{Cook,guillaume}.}
-
-What characterizes this noise is the decreasing of $6dB$ by octave. In this way, $\alpha_i$ in the set \ref{eq:rosa} is defined as:
+The minimum frequency $f_{\text{min}}$ is chosen with regard to the human hearing, since a sound component with frequency bellow $\approx\; 20Hz$ is virtually inaudible.
+
+Other noises can be made by similar procedures. Simple modifications are needed, specially in the equation that defines $\alpha_i$.
+
+  \item The brown noise received this name after Robert Brown, who described the Brownian movement\footnote{Although its origin is disparate with its color association, this noise became established with this specific name, in musical contexts.
+Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more intense colors~\cite{Cook,guillaume}.}
+
+What characterizes this noise is the decrease of $6dB$ per octave. In this way, $\alpha_i$ in the equation set~\ref{eq:rosa} is:
 
@@
 
- \item In the blue noise there is a gain of $3dB$ by octave in a band limited by the minimum frequency $f_{\text{min}}$ and the maximum frequency $f_{\text{max}}$.
+ \item In the blue noise there is a gain of $3dB$ per octave in a band limited by the minimum frequency $f_{\text{min}}$ and the maximum frequency $f_{\text{max}}$.
 Therefore, the corresponding equation is also based on the equations set~\ref{eq:rosa}:
@@
  \alpha_i & = (10^{\frac{3}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \\
- c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \;\; \text{ou} \;\; f_i>f_{\text{max}} \\
+ c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \;\; \text{or} \;\; f_i>f_{\text{max}} \\
  \end{split}
@@
 
- \item The violet noise is similar to the blue noise, but its gain is $6dB$ by octave:
+ \item The violet noise is similar to the blue noise, but its gain is $6dB$ per octave:
 
 \begin{equation}\label{eq:violeta}
- \alpha_i = (10^{\frac{6}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )} \;\;, \quad f_{\text{min}} \approx 15 Hz \\
+ \alpha_i = (10^{\frac{6}{20}})^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )}
 \end{equation}
@@
  \item The gray noise is defined as a white noise subject to one of the ISO-audible curves.
-Those curves are obtained by experiments and imperative to obtain $\alpha_i$.
-An implementation of ISO 226, that is the last revision of those curves, is in the toolbox \massa.\cite{MASSA}
+Those curves are obtained by experiments and are imperative to obtain $\alpha_i$.
+An implementation of ISO 226, which is the last revision of these curves, is in the \massa\ toolbox~\cite{MASSA}.
 
@@
 
-This subsection exposed only noises with static spectrum.
-There are also characterizations for noises with dynamic spectrum during the time, or noises which are fundamentally transient, like clicks and chirps.
-The former are easily modeled by an impulse relatively isolated, while chirps are not in fact a noise, but a fast scan of some given frequency band~\cite{Cook}.
+This subsection exposed only noise with static spectrum.
+There are also characterizations for noise with dynamic spectrum during the time, and noises which are fundamentally transient, like clicks and chirps.
+The former are easily modeled by an impulse relatively isolated, while chirps are not in fact a noises, but a fast scan of some given frequency band~\cite{Cook}.
 
 The noise from equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom},
-\ref{eq:azul} and \ref{eq:violeta} are presented in Figure~\ref{fig:ruidos}.
-The spectrum were build with the same phase and frequency for each coefficient, making it straightforward to observe the contribution of treble harmonics and bass frequencies.
+\ref{eq:azul} and \ref{eq:violeta} are presented in figure~\ref{fig:ruidos}.
+As stated above, the spectra were built with the same phase and frequency for each coefficient, making it straightforward to observe the contribution of treble harmonics and bass frequencies.
 

### 2013-10-02 article/article.tex
feito ateh usos musicais das variacoes na nota musical

@@
 \begin{equation}\label{eq:dur}
-T_{i}^{\Delta}={\{t_i\}}_{i=0}^{\lfloor \Delta . f_a \rfloor -1}
+T_{i}^{\Delta}={\{t_i\}}_{i=0}^{\lfloor \Delta . f_s \rfloor -1}
 \end{equation}
@@
 
-Vibrato is a period variation in pitch (frequency) and tremolo is a period variation in volume (intensity)\footnote{Some musical instruments and contexts use different terms.
-For example, in piano, the called tremolo is a vibrato and a tremolo is the classification used here.
+Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in volume (intensity)\footnote{Some musical instruments jargon and contexts use different terms.
+For example, in piano, the called tremolo is a vibrato in the classification used here.
 The presented definitions are common in contexts regarding music theory and electronic music.
-In addition, they are based on a broader literature than the one used for an specific instrument, practice or musical tradition~\cite{Lacerda,Harmonia}.}
+In addition, they are based on a broader literature than the one used for a specific instrument, practice or musical tradition~\cite{Lacerda,Harmonia}.}
 For the general case, the vibrato is described as follows:
@@
 \begin{equation}\label{vbrGamma}
- \gamma_i'=\left \lfloor i f' \frac{\widetilde{\Lambda}_M}{f_a} \right \rfloor
+ \gamma_i'=\left \lfloor i f' \frac{\widetilde{\Lambda}_M}{f_s} \right \rfloor
 \end{equation}
@@
 \begin{multline}\label{vbrGamma2}
- \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \\ = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f \left ( \frac{f + \mu }{f} \right )^{t_j'}  \right \rfloor= \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f .
+ \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_s} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_s} \right \rfloor = \\ = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}f \left ( \frac{f + \mu }{f} \right )^{t_j'}  \right \rfloor= \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}f .
 2^{t_j'\frac{\nu}{12}}  \right \rfloor
@@
          \includegraphics[width=\columnwidth]{figures/vibrato}
-     \caption{Spectrogram of a sound with sinusoidal vibrato of $3Hz$ and one octave depth in a $1000Hz$ sawtooth wave (considering $f_a=44.1kHz$).}
+     \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave,  considered $f_s=44.1kHz$.}
          \label{fig:vibrato}
@@
 
-For the properly realization of vibrato, it is important to pay attention on the following tables (quais tabelas? seriam termos?) and sequences.
-\textcolor{red}{Table $\widetilde{M}_i$ with length $\widetilde{\Lambda}_M$ and the sequence with indices $\gamma_i'$ make the sequence $t_i'$ which is the oscillation pattern in the frequency while table $\widetilde{L}_i$ with length $\widetilde{\Lambda}$ and the sequence with indices $\gamma_i$ make $t_i$ which is the sound itself.
-Variables $\mu$ and $\nu$ quantify vibrato intensity: $\mu$ is a direct measure of how many Hertz are involved in the upper limit of the oscillation while $\nu$ is the direct measure of how many semitones (or half steps) are involved in the oscillation ($2\nu$ is the number of semitones between the upper and lower pikes of frequency oscillations of the sound $\{t_i\}$ caused by the vibrato).
-It is convenient to use $\nu=\log_{2}\frac{f+\mu}{f} $ in this case because the maximum frequency increase is not equivalent to the maximum diminish, but the semitones variations remains.}
-
-Figure~\ref{fig:vibrato} illustrates the spectrogram of an artificial vibrato for a note with $1000Hz$ (between a \emph{B} and a \emph{C}), and which frequency deviation reaches one octave above and one below.
-Any waveform can be used to generate a sound and a vibrato oscillation pattern given any oscillation frequency and pitch deviation\footnote{The pitch deviation is called 'vibrato depth' and is generally given as semitones or cents, as convenience}.
-Those oscillations with precise forms and arbitrary amplitudes are not possible in traditional music instruments and, in this way, it introduces novelty in the artistic possibilities.
+For the proper realization of the vibrato, it is important to pay attention to both tables and sequences.
+Table $\widetilde{M}_i$ with length $\widetilde{\Lambda}_M$ and the sequence with indices $\gamma_i'$ make the sequence $t_i'$ which is the oscillation pattern in the frequency while table $\widetilde{L}_i$ with length $\widetilde{\Lambda}$ and the sequence with indices $\gamma_i$ make $t_i$ which is the sound itself.
+Variables $\mu$ and $\nu$ quantify the vibrato intensity: $\mu$ is a direct measure of how many Hertz are involved in the upper limit of the oscillation while $\nu$ is the direct measure of how many semitones (or half steps) are involved in the oscillation ($2\nu$ is the number of semitones between the upper and lower peaks of frequency oscillations of the sound $\{t_i\}$ caused by the vibrato).
+It is convenient to use $\nu=\log_{2}\frac{f+\mu}{f} $ in this case because the maximum frequency increase is not equivalent to the maximum frequency diminish, but the semitone variation remains.
+
+Figure~\ref{fig:vibrato} is the spectrogram of an artificial vibrato for a note with $1000Hz$ (between a \emph{B} and a \emph{C}), in which pitch deviation reaches one octave above and one below.
+Practically any waveform can be used to generate a sound and the vibrato oscillation pattern, with virtually any oscillation frequency and pitch deviation\footnote{The pitch deviation is called 'vibrato depth' and is generally given as semitones or cents, as convenient}.
+Those oscillations with precise waveforms and arbitrary amplitudes are not possible in traditional music instruments and, in this way, it introduces novelty in the artistic possibilities.
 
@@
 \end{equation}
-
+and finally: 
 \begin{multline}\label{trT}
@@
 The curvature is due to the logarithmic progression of the intensity.
-The tremolo frequency is $1,5Hz$ because $f_a=44,1kHz \; \Rightarrow \; \text{duration} = \frac{i_{\text{max}}=82000}{f_a}= 2s \; \Rightarrow \; \frac{3\text{oscillations}}{2s}=1,5$ oscillations for second ($Hz$).
-
-The music piece \emph{Vibra e treme} explores those possibilities given by tremolos and vibratos, both used in conjunction and independently, with frequencies $f'$, different depths ($\nu$ and $V_{dB}$), and progressive parameters variations \footnote{Tremolos and vibratos occur many times together in a traditional music instrument and voices.}.
+The tremolo frequency is $1,5Hz$ because $f_a=44,1kHz \; \Rightarrow \; \text{duration} = \frac{i_{\text{max}}=82000}{f_a}= 2s \; \Rightarrow \; \frac{3\text{oscillations}}{2s}=1,5$ oscillations per second ($Hz$).
+
+The music piece \emph{Vibra e treme} explores these possibilities given by tremolos and vibratos, both used in conjunction and independently, with frequencies $f'$, different depths ($\nu$ and $V_{dB}$), and progressive parameters variations (tremolos and vibratos occur many times together in a traditional music instrument and voices).
 Aiming a qualitative appreciation, the piece also develops a comparison between vibratos and tremolos in logarithmic and linear scales.
-Its source code is in the Appendix~\ref{ap:vibra} and available online as part of the \massa \emph{toolbox}.
-
-\begin{figure}[h!]
+Its source code is available online as part of the \massa\ toolbox.
+
+\begin{figure*}
      \centering
-         \includegraphics[width=\columnwidth]{figures/tremolo}
-     \caption{Tremolo with a depth of $V_{dB}=12dB$ with a sawtooth waveform as its oscillatory pattern with $f'=1.5Hz$ in a sine of $f=40Hz$ (considering a sample frequency of $f_a=44,1kHz$).}
+         \includegraphics[width=\textwidth]{figures/tremolo}
+     \caption{Tremolo with a depth of $V_{dB}=12dB$, with a sawtooth waveform as its oscillatory pattern, with $f'=1.5Hz$ in a sine of $f=40Hz$ (considered a sample frequency of $f_s=44.1kHz$).}
          \label{fig:tremolo}
-\end{figure}
-
-With progressive increasing of $f'$ for a linear proximity of frequency aiming
-to hear the phenomena as pitch ($\approx 20Hz$) it (o que?) generates roughness to both
-tremolos and vibratos. Those roughness are largely appreciated both in
-traditional classical music and current electronic music, specially in the
-\emph{Dubstep} genre.  Roughness is also generated by spectral content that
-produce beating~\cite{Porres,porres2009}. The sequence \emph{Bela Rugosi}
-explores the threshold with concomitants of tremolos and vibratos at the same
-voice, with different intensity and waveforms. The respective code is presented in
-Appendix~\ref{ap:bela} and available online in \massa.
-
-Increasing even more the frequency makes those oscillations no longer remains
-noticeable. In this case, the oscillations are audible as pitch. Then, $f'$,
-$\mu$ and the waveform changes together the frequency of original sound $T_i$
+\end{figure*}
+
+The proximity of $f'$ to $20Hz$ generates rugosity in both tremolos and vibratos. This rugosity is largely appreciated both in
+traditional classical music and current electronic music, specially in the \emph{Dubstep} genre.
+Rugosity is also generated by spectral content that produce beating~\cite{Porres,porres2009}.
+The sequence \emph{Bela Rugosi} explores this rugosity threshold with concomitant tremolos and vibratos at the same
+voice, with different intensity and waveforms. The respective code is
+available online in the \massa\ toolbox.
+
+As frequency increases further, these oscillations no longer remains
+noticeable individually. In this case, the oscillations are audible as pitch. Then, $f'$,
+$\mu$ and the waveform together, changes the spectrum of original sound $T_i$
 in different ways for both tremolos and vibratos.  They are called AM
@@
 synthesizers like \emph{Yamaha DX7}, and even with applications outside music, like
-in telecommunications to data transferring by means of electromagnetic waves
-(e.g.\ AM and FM radios).
-
-For musical ends, it is possible to understand FM based on the
-case of sines and to decompose the signals into their respective Fourier
-components (i.e.\ sinusoidal) to more complex cases. In this way, the FM
+in telecommunications, for data transfer by means of electromagnetic waves (e.g.\ AM and FM radios).
+
+For musical goals, it is possible to understand FM based on the
+case of sines and, in occurrences of greater complexity, to decompose the signals into their respective Fourier components (i.e.\ sines as well). In this way, the FM
 synthesis performed with a sinusoidal vibrato with frequency $f'$ and depth
@@
 \begin{split}
-\{t_i'\} & = \left \{ \cos \left [f . 2 \pi \frac{i}{f_a-1} + \mu . sen \left ( f' . 2 \pi \frac{i}{ f_a -1 } \right ) \right ] \right \} = \\
- & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ f . 2 \pi \frac{i}{f_a-1} + k . f' . 2 \pi \frac{i}{f_a-1} \right ]  \right \} = \\
- & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ (f+k.f') . 2 \pi \frac{i}{f_a-1} \right ]  \right \}
+\{t_i'\} & = \left \{ \cos \left [f . 2 \pi \frac{i}{f_s-1} + \mu . sen \left ( f' . 2 \pi \frac{i}{ f_s -1 } \right ) \right ] \right \} = \\
+ & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ f . 2 \pi \frac{i}{f_s-1} + k . f' . 2 \pi \frac{i}{f_s-1} \right ]  \right \} = \\
+ & = \left \{ \sum_{k=-\infty}^{+\infty} J_k(\mu) \cos \left [ (f+k.f') . 2 \pi \frac{i}{f_s-1} \right ]  \right \}
 \end{split}
@@
 
-On those equations, the frequency variation introduced by $\{t_i'\}$ does not
-respect the geometric progression that follows the pitch perfection but reflects the
-equation~\ref{freqLinear}. Equations~\ref{vbrF} are used for FM and are described
-in Appendix~\ref{cap:fmam}, where the spectral content of the FM synthesis is
-calculated and obtained with oscillations in the logarithmic scale. In
-fact, what is attractive in FM is its simple behavior (only with linear
-variations in~\ref{eq:fmEsp}).
+On these equations, the frequency variation introduced by $\{t_i'\}$ does not
+respect the geometric progression that yield linear pitch variation, but reflects
+equation~\ref{freqLinear}.
+The use of equations~\ref{vbrF} for FM are described in~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
+In
+fact, the simple and attractive FM behavior are observed only with linear variations, such as in~\ref{eq:fmEsp}).
 
@@
 \{t_i'\}_0^{\Lambda-1} =\{(1+a_i) .
-t_i\}_0^{\Lambda-1} = \left \{ \left [ 1+M.\sin \left ( f'.2\pi\frac{i}{f_a -1} \right ) \right] .P .\sin \left ( f.2\pi\frac{i}{f_a -1} \right ) \right \}_0^{\Lambda-1} = \\
-                        =  \left\{P.\sin \left( f.2\pi\frac{i}{f_a -1}  \right ) +  \frac{P.M}{2} \left [ \sin \left( (f-f').2\pi\frac{i}{f_a -1}  \right )  + \sin \left( (f+f').2\pi\frac{i}{f_a -1}  \right ) \right ] \right \}_0^{\Lambda-1}
+t_i\}_0^{\Lambda-1} = \left \{ \left [ 1+M.\sin \left ( f'.2\pi\frac{i}{f_s -1} \right ) \right] .P .\sin \left ( f.2\pi\frac{i}{f_s -1} \right ) \right \}_0^{\Lambda-1} = \\
+                        =  \left\{P.\sin \left( f.2\pi\frac{i}{f_s -1}  \right ) +  \frac{P.M}{2} \left [ \sin \left( (f-f').2\pi\frac{i}{f_s -1}  \right )  + \sin \left( (f+f').2\pi\frac{i}{f_s -1}  \right ) \right ] \right \}_0^{\Lambda-1}
 \end{equation}
@@
 reproduction of its spectral content below and above the original frequency,
-with a distance of $f'$ from $f$. Again, this is obtained by variations in the
-linear scale of the amplitude. Appendix~\ref{cap:fmam} has an exposition of the
-spectrum considering an AM performed with an oscillation in the amplitude logarithmic scale. It also loses the simple behavior.
-
-The sequence $T_i$ with frequency $f$, called 'carrier signal', is modulated by
-$f'$, called 'modulation signal'. Assuming the FM and AM jargon, $\mu$ and
-$\alpha=10^{\frac{V_{dB}}{20}}$ are called 'modulation indexes'. The following
-equations are defined for the vibration pattern of the modulation signal
+with the distance $f'$ from $f$. Again, this is obtained by variations in the
+linear scale of the amplitude. An exposition of the
+spectrum of an AM performed with oscillations in the logarithmic amplitude scale is again in~\cite{dissertacao}, and it again loses the simple and attractive behavior.
+
+The sequence $T_i$, with frequency $f$, called `carrier', is modulated by
+$f'$, called 'modulator'. In FM and AM jargon, $\mu$ and
+$\alpha=10^{\frac{V_{dB}}{20}}$ are `modulation indexes'. The following
+equations are defined for the vibration pattern of the modulator sequence
 $\{t_i'\}$:
@@
 \begin{equation}\label{fmGammaAux}
-\gamma_i'=\left \lfloor i f' \frac{\widetilde{\Lambda}_M}{f_a} \right \rfloor
+\gamma_i'=\left \lfloor i f' \frac{\widetilde{\Lambda}_M}{f_s} \right \rfloor
 \end{equation}
@@
 
-The modulation signal $\{t_i'\}$ into the carrier signal $\{t_i\}$ for FM is applied as:
+The modulator $\{t_i'\}$ into the carrier $\{t_i\}$, for FM, is applied as:
 
@@
 \begin{equation}\label{fmGamma}
-\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_a} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}(f+\mu .
+\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_s} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_s} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}(f+\mu .
 t_j') \right\rfloor
@@
 
-\noindent where $\widetilde{l}$ is the waveform period with a length
-$\widetilde{\Lambda}$ for the carrier signal.
-
-To perform AM someone just needs to modulate the signal $\{t_i\}$ with $\{t_i'\}$ using
+\noindent where $\widetilde{l}$ is the waveform period with a length of
+$\widetilde{\Lambda}$ samples, used for the carrier signal.
+
+To perform AM, it is needed to modulate the signal $\{t_i\}$ with $\{t_i'\}$ using
 the following equations:
@@
 
-At this point, the musical possibilities have been exploded.
-All characteristics like pitch (given by frequency), timbre (given by the waveform and filters), volume (given by intensity) and duration (given by the number of samples) can be considered by themselves or treated during their duration (with the exception of the duration itself).
+At this point, the musical possibilities has exploded.
+All characteristics, like pitch (given by frequency), timbre (given by the waveform, filters and noise), volume (manipulated by intensity) and duration (given by the number of samples), can be considered in an absolute form or treated during duration of sound (with the exception of the duration itself, clearly).
 
 The following musical usages comprehend a collection of possibilities
-with the purpose of exemplifying types of sound manipulation that results in
+with the purpose of exemplifying types of sound manipulation that result in
 musical material. Some of them are deeply discussed in the next section.
@@
 
-As an example, consider $\Delta_1$ as the duration of the first period and $\Delta_R$ as the duration of total reverberation ($\Lambda_1=\Delta_1 f_a$, $\Lambda_R=\Delta_R
-f_a$). It is possible to add a probability $p_i$ of a sound reincidencing in the
+As an example, consider $\Delta_1$ as the duration of the first period and $\Delta_R$ as the duration of total reverberation ($\Lambda_1=\Delta_1 f_s$, $\Lambda_R=\Delta_R
+f_s$). It is possible to add a probability $p_i$ of a sound reincidencing in the
 $i$-th sample with amplitude with exponential decay. Following
@@
 
-\noindent with $\Lambda_X=\lfloor \Delta . f_a \rfloor\;\;\forall\;\; X \; \in
+\noindent with $\Lambda_X=\lfloor \Delta . f_s \rfloor\;\;\forall\;\; X \; \in
 (A,D,R,\;)$ and being $\xi$ a small value that provides a satisfactory \emph{fade in} and \emph{fade out}, e.g.\ $\xi=10^{\frac{-80}{20}}=10^{-4}\;$ or $\;\xi=10^{\frac{-40}{20}}=10^{-2}$.

### 2013-10-02 article/article.tex
ateh fonte em movimento

@@
 
-\subsection{Musical usage}\label{subsec:mus2}
-
-At this point, the musical possibilities has exploded.
-All characteristics, like pitch (given by frequency), timbre (given by the waveform, filters and noise), volume (manipulated by intensity) and duration (given by the number of samples), can be considered in an absolute form or treated during duration of sound (with the exception of the duration itself, clearly).
+\subsection{Musical usages}\label{subsec:mus2}
+
+At this point, the musical possibilities exploded.
+All characteristics, like pitch (given by frequency), timbre (given by the waveform, filters and noise), volume (manipulated by intensity) and duration (given by the number of samples), can be considered in an absolute form or treated during duration of sound (clearly, with the exception of duration itself).
 
@@
 
-Another interesting possibility is to use the relations between parameters of
-tremolo and vibrato, and some parameters of the basic note like frequency. In
-this way, it is possible to say that vibrato frequency is proportionally
-different from pitch, or the tremolo depth is inversely proportional to
+An interesting possibility is to use relations between parameters of
+a tremolo or vibrato, and some parameters of the basic note like frequency. In
+this way, it is possible to have a vibrato frequency proportional
+to note pitch, or a tremolo depth inversely proportional to
 pitch. Therefore, with equations \ref{vbrGamma}, \ref{vbrF} and \ref{trA}, it is
-possible to define:
+perfectly possible to arbitrate:
 
@@
 
-\noindent with $f^{vbr}$ and $f^{tr}$ as $f'$ in the referenced equations. They can also be associated with the vibrato and tremolo oscillation frequency in
+\noindent with $f^{vbr}$ and $f^{tr}$ as $f'$ in the referenced equations. They can also be associated with vibrato and tremolo oscillation frequency of
 equation~\ref{vbrGamma}. 
-
-Besides that, $\nu$ and $V_{dB}$ are the respective depth values of vibrato and
-tremolo. Functions $func_a$, $func_b$ and $func_c$ are arbitrary and dependents
-on the musical intentions. The music piece \emph{Tremolos, vibratos e a
-frequência} explores such characteristics and shows variations in the oscillation
-waveform with specific relations with the purpose of building a \emph{musical
-language}\footnote{Details in the next section.}. The corresponding code is on
-Appendix~\ref{ap:tremolos} and it is also available online as part of the \massa
-toolbox.
-
-About convolution it is possible to use a musical pulse as duration -- like a BPM
-pulse -- and to distribute impulses during this original pulse, aiming at
-establishing metrics and rhythms\footnote{It is important to remember that the
-convolution with a impulse results in a sound shifted to the moment that the
-impulse occurred}.
+ $\nu$ and $V_{dB}$ are the respective depth values of vibrato and
+tremolo. Functions $func_a$, $func_b$ and $func_c$ are arbitrary and dependent
+on musical intentions. The music piece \emph{Tremolos, vibratos e a
+frequência} explores such bonds and exhibit variations in the oscillation
+waveform with the purpose of building a \emph{musical
+language} (details in the next section). The corresponding code is on
+is also available online as part of the \massa\ toolbox.
+
+\subsubsection{Convolution for rhythm and meter}
+It is possible to use a musical pulse -- such as specified by a BPM
+tempo mark -- and to place an impulse at the start of each beat, with the purpose of
+establishing metric and rhythms: the
+convolution with an impulse shifts the sound to where
+the impulse.
 For example, two impulses equally spaced builds a binary division into the
-pulse. Two signals, one with 2 pulses and another with 3 pulses, both with
-equally spaced impulses in the pulse duration, results in the pulse
-maintenance with a rhythmic mark that is possible to use as binary or ternary
-divisions in many ethnic or traditions music styles~\cite{Gramani}. 
-The absolute values of the impulses results in proportions among the
-amplitudes of the convoluted signals.
-Using convolution with impulses as a metric is explored in the
-music piece \emph{Trenzinho de caipiras impulsivos}. The features
-embrace the creation of 'sound amalgam' based on granular synthesis and this
-piece provides a link to the next section. Refer specially to the
-Figure~\ref{fig:pulsoSubAgl}. The source code of the music piece is included in the section~\ref{ap:trenzinho} and online as well (\massa toolbox).
+pulse. Two signals, one with 2 impulses and another with 3 impulses, both 
+equally spaced in the pulse duration, yields the pulse
+maintenance with a rhythmic trace in which is easy to use binary or ternary
+divisions of the pulse, as found in many ethnic or traditional music styles~\cite{Gramani}. 
+Absolute values of the impulses implyes in proportions among the
+amplitudes of the sonic incidences.
+The use of convolution with impulses in this context is explored in the
+music piece \emph{Trenzinho de caipiras impulsivos}. These features
+embrace the creation of `sound amalgams' based on granular synthesis, for now refer specially to the
+figure~\ref{fig:pulsoSubAgl}. This piece is a link to next section contents. 
+The source code of the music piece is online, as part of the \massa toolbox.
 

### 2013-10-04 article/article.tex
refeita parte de impulsos para ritmo, falta de fonte em movimento para frente

@@
 It is possible to use a musical pulse -- such as specified by a BPM
-tempo mark -- and to place an impulse at the start of each beat, with the purpose of
+tempo -- and to place an impulse at the start of each beat, with the purpose of
 establishing metric and rhythms: the
-convolution with an impulse shifts the sound to where
-the impulse.
+convolution with an impulse shifts the sound 
+to impulse position.
 For example, two impulses equally spaced builds a binary division into the
-pulse. Two signals, one with 2 impulses and another with 3 impulses, both 
-equally spaced in the pulse duration, yields the pulse
-maintenance with a rhythmic trace in which is easy to use binary or ternary
-divisions of the pulse, as found in many ethnic or traditional music styles~\cite{Gramani}. 
+pulse. Two signals, one with 2 impulses and the other with 3 impulses, both 
+equally spaced in the pulse duration, yields a pulse
+maintenance with a rhythmic in which eases both binary or ternary
+divisions of the pulse, and this is found in many ethnic and traditional musical styles~\cite{Gramani}. 
 Absolute values of the impulses implyes in proportions among the
-amplitudes of the sonic incidences.
+amplitudes of the sonic reincidences.
 The use of convolution with impulses in this context is explored in the
@@
 
-\subsubsection{Moving audio source and receptor, Doppler effect}
-
-Resuming the exposition in subsection~\ref{subsec:spac}: when an audio source
+\subsubsection{Moving source and receptor, Doppler effect}
+
+Summarizing the exposition in subsection~\ref{subsec:spac}: when an audio source
 and a receptor are moving, their characteristics are ideally updated at each

### 2013-10-05 article/article.tex
feita ateh 3ira parte

@@
 
-Summarizing the exposition in subsection~\ref{subsec:spac}: when an audio source
-and a receptor are moving, their characteristics are ideally updated at each
-sample of the digital signal. The speeds are decomposed in relation of each ear
-direction. In this way, given the audio source speed (or velocity) $v_s$, that
-is positive if the source moves away from receptor, and receptor speed $v_r$,
-that is positive when it gets closer of the audio source, the frequency is given
-by the well known Doppler effect:
+Retaken the exposition in subsection~\ref{subsec:spac}, when an audio source
+or receptor is moving, its characteristics are ideally updated at each
+sample of the digital signal. Speed components should be found for each ear.
+In this way, given the audio source speed (or velocity) $v_s$,
+with positive values if the source moves away from receptor, and receptor speed $v_r$,
+positive when it gets closer to audio source, the frequency is given
+by the well known Doppler shift:
 
@@
 
-With this frequency and relations given by the new IID from the new source
+With both frequencies $f$ and $f_0$, and the IID from the new source
 position, it is possible to create the Doppler effect. There is an addendum to
 improve the fidelity of the physical phenomena: to increase the received
-potency.
-It is possible to understand this potency gain as being proportional to the relative speed that, in each second, adds the traversed period, by means of a waveform with potency: $\Delta P=P_0\left(\frac{v_r-v_s}{343,2}\right)$, where $P_0$ represents the signal potency.
-
-In this way it is possible to obtain both amplitude and frequency of a moving
+potency. It is possible to understand this potency gain as being proportional 
+to the wave shrinking or expansion: $\Delta P=P_0\left(\frac{v_r-v_s}{343.2}\right)$,
+ where $P_0$ is signal potency and $P$ the potency at receptor.
+
+In this way, it is possible to obtain both amplitude and frequency of a moving
 audio source. Being this audio source in front of receptor with $y_0$ meters of horizontal distance and $z_0$ meters of height, the distance is given by
 $D_i=\left\{ d_i=\sqrt{ y_{i}^{2}+z_{0}^{2} } \right\}_0^{\Lambda-1}$,
-where $y_i=y_0+v_s-v_r$ if considering $v_s$ and $v_r$ both in
-horizon. Amplitude changes with the distance and with the potency factor cited
+where $y_i=y_0+(v_s-v_r)\frac{i}{f_s}$ with $v_s$ and $v_r$ both
+horizontal, having null non-$y$ components. Amplitude changes with the distance and with the potency factor cited
 above (see subsection~\ref{subsec:volume} for potency to amplitude conversion).
@@
 
-Observe that the factor which changes the amplitude (caused by the distance) is even, while the factor caused by the potency variation is antisymmetric in
+The amplitude change caused by the distance is even, while the change caused by the potency variation is antisymmetric in
 relation to the crossing of source with receptor.
@@
 In other words, the same semitones (or fractions) added during the approach are decreased during the departure.
-Besides that, the transition becomes steep if both source and receptor intersects each other exactly at the same point, otherwise, there is a monotonic progression.
-In the given case, where there is a static height $z_0$, it is necessary to observe the speed component in the direction between the observer and audio source:
+Besides that, the transition is abrupt if source and receptor intersect with zero distance, otherwise, there is a monotonic progression.
+In the given case, where there is a static height $z_0$, observed the speed component in the direction given by observer and source positions, the frequencies $F_i$ at the observer is given by:
 
@@
 
-In the Appendix~\ref{sec:cod2} there is a Python implementation of the Doppler
-effect as describe above, also considering the intersection between audio source and receptor.
-
-\subsubsection{Filter and noises (subsections~\ref{subsec:ruidos} and~\ref{subsec:filtros})}
-
-Using filters the possibilities are many. It is possible to
-convolve a signal to reverberate it, to remove its noise, to generate
-distortions or to treat the audio with aesthetically manipulation in mind. For
-example, it is possible to simulate sounds from and old television if a
-band-pass filter is applied to accept just sound between $1kHz$ and $3kHz$. Or
-if someone removes (with some precision) just the frequency of electric
-oscillation (usually $50Hz$ or $60Hz$) and the harmonics, it will remove noises
-caused by audio devices. A more musical application is to perform filtering in
+There is a Python implementation of the Doppler
+effect as describe above, and also considering the intersection between audio source and receptor,
+at the \massa\ toolbox~\cite{massa}.
+
+\subsubsection{Filters and noises (subsections~\ref{subsec:ruidos} and~\ref{subsec:filtros})}
+
+With the use of filters, the possibilities are even wider.
+Convolve a signal to have a reverberated version of it, to remove its noise, to
+distort or to treat the audio aesthetically in other ways. For
+example, it is possible to simulate sounds as if originated in an old television or telefone,
+by the use of a band-pass filter, allowing only frequencies between $1kHz$ and $3kHz$.
+By rejecting the frequency of electric oscillation (usually $50Hz$ or $60Hz$) and harmonics, one can remove noises
+caused by audio devices connected to a usual power suply. A more musical application is to perform filtering in
 specific bands and to use those bands as an additional parameter to the notes.
@@
 Inspired by traditional music instruments, it is possible to apply a
-time-dependent filter~\cite{Roederer}. Chains of those filters can perform
+time-dependent filter~\cite{Roederer}. Chains of these filters can perform
 complex and more accurate filtering routines. The music piece \emph{Ruidosa
-faixa} explores those features by using filters and many kinds of noise
-synthesis. The source code is in Appendix~\ref{ap:ruidosa} and is available
-online as part of \massa.
-
-When used together the features can create an effect known
-as \emph{chorus}. Based on what happens in a singers choir, in this
-effect the sound is performed using small and potentially arbitrary
+faixa} explores filters and many kinds and noise
+synthesis. The source code is available
+online as part of \massa\ toolbox.
+
+A sound can be altered through different filtering and then mixed to create an effect known
+as \emph{chorus}. Based on what happens in a choir of singers,
+the sound is performed using small and potentially arbitrary
 modifications of parameters like center frequency, presence (or absence) of
@@
 equation~\ref{eq:mixagem}). The music piece \emph{Chorus infantil} implements a
-chorus in many ways with different sounds and its source code is listed in
-Appendix~\ref{ap:chorus}. The script is also available in the \massa toolbox.
+chorus in many ways with different sounds and its source code is also available in the \massa\ toolbox.
 
@@
 
-Using the same spatialization terms of subsection~\ref{subsec:spac}, the
-late reverberation can be modeled as a convolution with a period of pink, brown
-or black noise, with exponential decay of amplitude and time related. In
-this way, the treble attenuation and irregular smooth in the frequency response
-are contemplated with great success. Delay lines can be added as prefix to noise
+Using the same terms of subsection~\ref{subsec:spac}, the
+late reverberation can be achieved by a convolution with a section of pink, brown
+or black noise, with exponential decay of amplitude along time.
+Delay lines can be added as a prefix to the noise
 with the decay, and this contemplates both time parts of the reverberation: the
-first reflections and the late reverberation. It is possible to improve the ??
-quality by calculating the geometric localization of the last surface where each
-wave front reflected before reach the ear in the fairest $100-200$ milliseconds. Then, one could apply a low-pass filter as described in subsection~\ref{subsec:filtros}. The colored
-noise can be gradually introduced since the initial moment gives direct
-incidence (i.e.\ without any reflection and given by ITD and IID),
-with \emph{fade-in}, reaching its maximum at the beginning of the 'late
-reverberation', when the geometric incidences loses their importance to the
-statistics properties of the noise decay.
-
-As an example, consider $\Delta_1$ as the duration of the first period and $\Delta_R$ as the duration of total reverberation ($\Lambda_1=\Delta_1 f_s$, $\Lambda_R=\Delta_R
-f_s$). It is possible to add a probability $p_i$ of a sound reincidencing in the
-$i$-th sample with amplitude with exponential decay. Following
+first reflections and the late reverberation. It is possible to improve the
+quality by calculating the geometric trajectory and (mainly low-pass, subsection~\ref{subsec:filtros}) filtering by each surface where
+wavefront reflected before reaching the ear in the first $100-200$ milliseconds.
+The colored noise can be gradually introduced with a \emph{fade-in}: the initial moment given by direct
+incidence of sound (i.e.\ without any reflection and given by ITD and IID),
+reaching its maximum at the beginning of the 'late
+reverberation', when the geometric incidences loses their relevance to the
+statistical properties of noise decay.
+
+As an example, consider $\Delta_1$ as the duration of the first section and $\Delta_R$ as the duration of total reverberation ($\Lambda_1=\Delta_1 f_s$, $\Lambda_R=\Delta_R
+f_s$). Be $p_i$ the probability of a sound to repeat in the
+$i$-th sample. Amplitude falls with exponential decay. Following
 subsection~\ref{subsec:spac}, the reverberation $R_i^1$ of the first period can
@@
 
-\begin{multline}\label{eq:p1rev}
-    R_i^1=\left\{r_i^1\right\}_0^{\Lambda_1-1}\;: \\ \;r_i^1=\left\{
+\begin{equation}\label{eq:p1rev}
+\begin{split}
+    R_i^1=\left\{r_i^1\right\}_0^{\Lambda_1-1}\;:\quad\quad\quad\quad\quad\quad\quad\quad \\ 
+\;r_i^1=\left\{
         \begin{array}{l l}
@@
         \end{array} \right.
-\end{multline}
+\end{split}
+\end{equation}
 
@@
         $-120dB$. Reverberation $R_i^2$ of the second period can be emulated by a
-        brown noise $R_i^m$ (or by a pink noise $R_i^r$) with exponential decay:
+        brown noise $R_i^m$ (or by a pink noise $R_i^r$) with exponential amplitude decay of the waveform:
 
@@
 
-\noindent like:
+Considered:
 
@@
 
-\noindent the reverberation represented by $R_i$ can be applied as simple
+\noindent, a reverberated auralization of a sound can be achieved by simple
         convolution of $R_i$ (called reverberation 'impulse response') with the sound sequence $T_i$ as described in subsection~\ref{subsec:filtros}.
 
-Reverberation is well known by causing great interest in the listeners and to provide more enjoyable sonorities. Besides that, modifications in the reverb
-space consists of a tricking (almost a \textit{clich\'{e}}) to induce surprise and interest in the listener.
+Reverberation is well known by causing great interest in the listeners and to provide more enjoyable sonorities.
+Besides that, modifications in the reverberation space consists on a common resource (almost a \textit{clich\'{e}}) to surprise and attract the listener.
+An implementation of the reverb recipe described is on the \massa\ toolbox~\cite{massa}
 
@@
 The volume variation along the duration of sound is crucial to our timbre
-perception.
-The volume envelope, known as ADSR (\emph{Attack-Decay-Sustain-Release}), has many implementations in both hardware and software synthesizers.
+perception. The volume envelope known as ADSR (\emph{Attack-Decay-Sustain-Release}), 
+has many implementations in both hardware and software synthesizers. 
 A pioneer implementation can be found in the Hammond
@@
 
-The stochastic ADSR envelope is characterized by 4 parameters: attack duration
-(time to sound reaches its maximum volume), decay duration (follows the attack
-immediately), level of sustain volume (in which the volume remains stable after the decay) and release duration (after sustain, this is the duration until the volume decays to zero).
+The scholastic ADSR envelope is characterized by 4 parameters: attack duration
+(time in which sound reaches its maximum amplitude), decay duration (follows the attack
+immediately), level of sustain volume (in which the volume remains stable after the decay) and release duration (after sustain section, this is the duration needed for amplitude to reach zero or final value).
 Note that the sustain duration is not specified because it is the difference between the duration itself and the durations of attack, decay and sustain.
 
-It is possible to apply the ADSR envelope with durations $\Delta_A$, $\Delta_D$ and $\Delta_R$, with total duration $\Delta$ and sustain level $a_S$, given by the fraction of the maximum amplitude, as a sound sequence $t_i$ defined as:
-
-\begin{multline}\label{eq:adsr}
-%\begin{split}
-\{a_i\}_0^{\Lambda_A-1}  = \left\{\xi\left(\frac{1}{\xi}\right)^{\frac{i}{\Lambda_A-1}}\right\}_0^{\Lambda_A-1} \;\; \text{or} \;\;\quad \left\{\frac{i}{\Lambda_A-1}\right\}_0^{\Lambda_A}\\
-\{a_i\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} =\left\{a_S^{\frac{i-\Lambda_A}{\Lambda_D-1}}  \right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} \;\;\quad \quad\quad \\ \text{or} \quad\;\; \left\{1-(1-a_S)\frac{i-\Lambda_A}{\Lambda_D-1}\right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1}\\
-\{ a_i \}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} =\left\{ a_S \right\}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} \\
-\{ a_i \}_{\Lambda-\Lambda_R}^{\Lambda-1}  =\left\{ a_S\left(\frac{\xi}{a_S} \right)^{\frac{i-(\Lambda-\Lambda_R)}{\Lambda_R-1}} \right\}_{\Lambda-\Lambda_R}^{\Lambda-1} \quad\;\; \\ \text{or} \quad\;\; \left\{ a_S - a_S\frac{i+\Lambda_R-\Lambda}{\Lambda_R-1}\right\}_{\Lambda-\Lambda_R}^{\Lambda-1}
-%\end{split}
-\end{multline}
+The ADSR envelope with durations $\Delta_A$, $\Delta_D$ and $\Delta_R$, with total duration $\Delta$ and sustain level $a_S$, given by the fraction of the maximum amplitude, to be aplyed to any sound sequence $T_i=\{t_i\}$, can be presented as:
+
+\begin{equation}\label{eq:adsr}
+\begin{split}
+\{a_i\}_0^{\Lambda_A-1}  = & \left\{\xi\left(\frac{1}{\xi}\right)^{\frac{i}{\Lambda_A-1}}\right\}_0^{\Lambda_A-1} \quad \text{ or }\\ = & \left\{\frac{i}{\Lambda_A-1}\right\}_0^{\Lambda_A}\\
+\{a_i\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} = & \left\{a_S^{\frac{i-\Lambda_A}{\Lambda_D-1}}  \right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} \quad \text{ or } \\ = &  \left\{1-(1-a_S)\frac{i-\Lambda_A}{\Lambda_D-1}\right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1}\\
+\{ a_i \}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} = & \left\{ a_S \right\}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} \\
+\{ a_i \}_{\Lambda-\Lambda_R}^{\Lambda-1}  = & \left\{ a_S\left(\frac{\xi}{a_S} \right)^{\frac{i-(\Lambda-\Lambda_R)}{\Lambda_R-1}} \right\}_{\Lambda-\Lambda_R}^{\Lambda-1} \quad \text{ or } \\ = &  \left\{ a_S - a_S\frac{i+\Lambda_R-\Lambda}{\Lambda_R-1}\right\}_{\Lambda-\Lambda_R}^{\Lambda-1}
+\end{split}
+\end{equation}
 
@@
 The lower the $\xi$, the slower the \emph{fade}, like the $\alpha$ illustrated in Figure~\ref{fig:transicao}.
-The terms in the right side of the equation~\ref{eq:adsr} attend for both introduction and ending of sound from the zero intensity because they are linear.
-Schematically, Figure~\ref{fig:adsr} shows the ADSR envelope, a classical implementation that supports many variations.
-For example, between attack and decay it is possible to add an extra partition where the maximum amplitude stays.
-Another common example is the use of more elaborated tracings to attack or decay.
-The music piece \emph{ADa e SaRa}, available both in
-Appendix~\ref{ap:ada} and in \massa, explores many configurations of the ADSL envelope.
+The stated second in equations~\ref{eq:adsr} attend both introduction and ending of sound from zero intensity because they are linear.
+Schematically, figure~\ref{fig:adsr} shows the ADSR envelope in a classical implementation that supports many variations.
+For example, between attack and decay it is possible to add an extra partition where the maximum amplitude remains for more than a peak.
+Another common example is the use of more elaborated tracings of attack or decay.
+The music piece \emph{ADa e SaRa}, available
+in \massa, explores many configurations of the ADSL envelope.
 
@@
         to an arbitrary sound sequence. The linear variation of the amplitude is
-        above. Below the amplitude variation is exponential.}
+        above, in blue. Below the amplitude variation is exponential.}
         \label{fig:adsr}

### 2013-10-06 article/article.tex
feito ateh intervalos

@@
 
-\section{Notes organization in music}\label{notasMusica} \label{sec:notesMusic}
-
-Consider $ S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1} $ as the sequence $S_j$ of $H$ musical events $s_j$.
-In addition, $S_j$, called a 'musical structure', as composed by events $s_j$ which are musical structure themselves, e.g.\ notes.
+\section{Organization of notes in music}\label{notasMusica} \label{sec:notesMusic}
+
+Consider $ S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1} $, the sequence $S_j$ of $H$ musical events $s_j$.
+Be $S_j$ a `musical structure', composed by events $s_j$ which are musical structures themselves, e.g.\ notes.
 This section is dedicated to techniques that make $S_j$ interesting and enjoyable for audition.
 
-The elements of $S_j$ can be overlapped by mixing them together as in
-equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, building intervals
-and chords. This scenery reflects the 'vertical thought' in music. On the other hand, the concatenation of $S_j$
-as in equation~\ref{eq:concatenacao} and in Figure ~\ref{fig:concatenacao},
+The elements of $S_j$ can be overlapped by mixing them together, as in
+equation~\ref{eq:mixagem} and figure~\ref{fig:mixagem}, for building intervals
+and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of the events
+in $S_j$, as in equation~\ref{eq:concatenacao} and in figure ~\ref{fig:concatenacao},
 builds melodic sequences and rhythms, which are associated with the musical
-'horizontal thought' in music. The fundamental frequency $f$ and the beginning moment
-(attack) are generally the most important characteristics of the elements in
-$S_j$. Such characteristics make possible the creation of music made by pitches (both
+`horizontal thought'. The fundamental frequency $f$ and the starting moment
+(attack) are generally the most important characteristics of the elements $s_j$ in
+$S_j$. This makes possible the creation of music made by pitches (both
 harmony and melody) and by the presence of temporal metrics and rhythms.
@@
 
-Here we describe characteristics in music related to harmony and melody.
-
 \subsubsection{Tuning}
@@
 Doubling the frequency is equivalent to ascending one octave ($f=2f_0$).
-The octave division in twelve notes is the canon of classical western music, although its usage has also been observed
+The octave division in twelve steps is the canon of classical western music.
+Its usage has also been observed
 outside western tradition, like ceremonial/religious and ethnic context~\cite{Wisnick}.
-
-The factor given by $\varepsilon=2^{\frac{1}{12}}$ defines a semitone. It builds a note
-grid along the hearing spectrum. Fixing a frequency $f$, all the possible
-fundamental frequencies stay separated by intervals that are multiples of
-$\varepsilon$. Twelve semitones (or half steps), equidistant to the human ear, make an
-octave. Therefore, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
+In this system, ideally, the factor given by $\varepsilon=2^{\frac{1}{12}}$ defines a semitone.
+A note grid along the hearing spectrum in which, given the frequency $f$, all the possible
+fundamental frequencies are separated by intervals, multiples of
+$\varepsilon$. Twelve semitones (or half steps), equidistant to the human ear, reaches an
+octave, equivalent to human year. Therefore, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
 $f_0$ and $f$.
 
-This absolute accuracy is common in computational implementations. For the real
-music instruments, however, there are deviations of those frequencies in order to make
-the harmonics of corresponding notes more compatible. Besides that, the fixed reference
+This absolute accuracy is common in computational implementations. Real
+music instruments, however, exhibit deviations in order to make compatible
+deviations from the harmonics of notes. Besides that, the fixed reference
 $\varepsilon=2^{\frac{1}{12}}$ characterizes an equally tempered tuning. There
-are tunings which respective intervals are ratios of low-order integers, originally based
-on observations of physical behaviors. First of these tunings were formalized
-around two thousand years before the existence of an equal temperament~\cite{Roederer}.
-
-Two iconic tunings:
+are tunings in which intervals are ratios of low-order integers, based
+on observations of physical phenomena. These tunings were first formalized
+around two thousand years before the arrival of the equal temperament~\cite{Roederer}.
+). Two emblematic tunings are:
 
 \begin{itemize}
-    \item The {\bf just intonation} is defined as notes of the diatonic scale
-    that are associated with ratios of low-order integers that defines the harmonic
-    series. The basic ratios are the same in the Ionian mode (white piano keys
-    from \do to \do (see below in subsection~\ref{subsec:escalas}): 1, 9/8, 5/4,
-    4/3, 3/2, 5/3, 15/8, 2/1). 
-    The intervals are considered as notes in a scale together with the following: the
-    semitone 16/15, the 'minor tone' 10/9, and the 'major tone' 9/8. There are
-    many ways to perform the division of the 12 notes.
-
-    \item The {\bf Pythagorean tuning} is based on the interval 3/2 (perfect
-    fifth). The Ionian mode becomes: 1, 9/8, 81/64, 4/3, 3/2, 27/16, 243/128,
-    2/1. The intervals are also considered as notes in a scale. Besides the
-    intervals in the mode, it is also used the minor second 256/243, the minor
+    \item The {\bf just intonation}, defined by association of diatonic scale notes
+    with ratios of low-order integers, as found in the harmonic
+    series. Considered the ionian mode, the white piano keys
+    from C to C, as seen below in subsection~\ref{subsec:escalas}, i.e. intervals 1, 2M, 3M, 4J, 5J, 6M, 7M, 8,
+    are considered ratios of frequency: 1, 9/8, 5/4
+    4/3, 3/2, 5/3, 15/8, 2/1. 
+    The following intervals are also considered: the
+    semitone 16/15, the 'minor tone' 10/9, and the 'major tone' 9/8.
+    There are many ways to perform the division of the 12 notes in the just intonation.
+
+    \item The {\bf Pythagorean tuning}, based on the interval 3/2 (perfect
+    fifth). The Ionian mode ratios becomes: 1, 9/8, 81/64, 4/3, 3/2, 27/16, 243/128,
+    2/1. These intervals are also considered to yield notes in a scale. Besides the
+    intervals in the mode, also used are the minor second 256/243, the minor
     third 32/27, the augmented fourth 729/512, the diminished fifth 1024/729,
@@
 
-In order to attends for micro-tonality\footnote{Micro-tonality is the use of intervals less
+In order to attend for micro-tonality\footnote{Micro-tonality is the use of intervals less
 than one semitone and have ornamental and structuring functionalities in music. The
-division of the octave in $12$ notes is motivated by physical fundamentals but it is also
-a \emph{convention} adopted even by western classical music. Other tunings are
-present. To cite one example, the traditional Thai music uses a octave
+division of the octave in $12$ notes is motivated by physical fundaments but also
+a \emph{convention} adopted too by western classical music. Other tunings are
+incident. To point only one example, the traditional Thai music uses an octave
 division in seven notes equally spaced ($\varepsilon=2^{\frac{1}{7}}$),
-resulting in intervals that have little in common with the intervals in the
+which allows intervals that have little in common with the intervals in the
 diatonic scale~\cite{Wisnick}}. It is possible to use non-integer real values as
-pitch sequences or even to maintain the integer values and change the factor $\varepsilon=2^{\frac{1}{12}}$. For example, 
+factors of $\varepsilon=2^{\frac{1}{12}}$ between frequencies, 
+or to maintain integer values and change $\varepsilon$. For example, 
 a tuning really near of the harmonic series
-itself is proposed with the division of the octave in $53$ notes:
-$\varepsilon_2=2^{\frac{1}{53}}$~\cite{microtonalidade}.
-In that division the notes are related by means of integers with
-$\varepsilon_2$. Note that if $S_i$ is a pitch sequence related by means of
-$\varepsilon_1$, a mapping for all the notes related by $\varepsilon_2$ defines
+itself is proposed with the equal division of the octave in $53$ notes:
+$\varepsilon_2=2^{\frac{1}{53}}$.~\cite{microtonalidade}
+In this division, notes are related by means of integers, factors of
+$\varepsilon_2$ relating frequencies by a valid interval.
+ Note that if $S_i$ is a pitch sequence related by means of
+$\varepsilon_1$, a direct mapping for notes related by $\varepsilon_2$ is 
 a new sequence $S_i'=\left\{s_i'\right\}=\left\{
 s_i \frac{\varepsilon_1}{\varepsilon_2}\right\}$. The music piece \emph{Micro
-tom} explores this micro-tonal features and its code is in
-Appendix~\ref{ap:micro} and is part of \massa.
+tom} explores microtonal features and its code
+is part of \massa\ toolbox.
 

### 2013-10-06 article/article.tex
entrado em acordes

@@
 
-Using the ratio $\varepsilon=2^{\frac{1}{12}}$ between the note frequencies
+Using the ratio $\varepsilon=2^{\frac{1}{12}}$ between note frequencies
 (i.e.\ one semitone) the intervals in the twelve note system can be represented by integers.
-Table~\ref{eq:intervalos} summarizes the characteristics of each interval: its traditional nomenclature, consonance and
-dissonance properties, and number of semitones per interval.
+Table~\ref{eq:intervalos} summarizes each interval:
+ its traditional nomenclature, consonance and
+dissonance properties, and number of semitones.
 
@@
 \centering
-\caption{The music intervals together with their traditional notation, basic classification for dissonance and consonance, and number of semitones.
-Perfect (P) unison, fifths and octaves are the perfect consonances.
+\caption{Music intervals together with their traditional notation, basic classification for dissonance and consonance, and number of semitones.
+Unison, fifth and octave are the perfect (P) consonances.
 Major (M) and minor (m) thirds and sixths are the imperfect consonances.
@@
 Major seconds and minor sevenths are the mild dissonances.
-Perfect fourth is the first special case, being a perfect consonance when considering it as the inversion of the perfect fifth and a dissonance or a imperfect consonance otherwise.
-The second special case is the tritone (A4 or aug4, d5 or dim5, tri,TT).
+Perfect fourth is a special case, being a perfect consonance when considered as an inversion of the perfect fifth and a dissonance or a imperfect consonance otherwise.
+The second special case is the tritone (A4 or aug4, d5 or dim5, tri, TT).
 This interval is consonant in some cultures.
-For tonal music, the tritone indicates dominant and searches for its resolution in a third or sixth.
-Due to this instability it is considered a dissonance interval.}
+For tonal music, the tritone indicates dominant and seeks urgent resolution in a third or sixth.
+Due to this instability it is considered a dissonant interval.}
 \begin{tabular}{| c | c | c | }\hline
@@
  & traditional notation & number of semitones \\
- consonance or dissonance: & P4 & P5 \\
+ consonance or dissonance: & P4 & 5 \\
  dissonance in Western tradition: & tritone, aug4, dim5 & 6 \\\hline
@@
 
-The nomenclature is based both in impositions and for convenience of the tonal system and practical aspects of note manipulation and can be specified
-as~\cite{Roederer,Wisnick}:
+The nomenclature, based on conveniences for tonal music and practical aspects of note manipulation, can be specified
+as follows~\cite{Roederer,Wisnick}:
 
 \begin{itemize}
-                \item Intervals as number of ratios between notes: first
+                \item Intervals are seen first by the as number of steps between notes: first
                 (unison), second, third, forth, fifth, sixth, seventh and
-                eighth.
-The ninth, tenth, eleventh, etc, are compound                       intervals made by one or more octaves + a interval inside the                 octave, which characterizes the compound interval.
-The                        intervals are represented by the numeric digits, e.g.
-1, 3, 5                 are a unison, a third and a fifth, respectively.
-
-                \item Qualities of each interval: the perfect consonances --
+                eighth (octave). The ninth, tenth, eleventh, etc, are compound  
+                intervals made by one or more octaves plus a ``simple'' interval inside the
+                octave, which characterizes the compound interval. The
+                intervals are represented by numeric digits, e.g. 1, 3, 5
+                are a unison, a third and a fifth, respectively.
+
+                \item Qualities of each interval: perfect consonances --
                 i.e.\ unison, forth, fifth and octave -- are 'perfect'. The
@@
 
-                \item The perfect fourth is both perfect consonance or
-                dissonance according to the context and theoretical
-                background. As general rule, it can be considered as                        consonance except when it is used as passage to resolve a                   fifth or a third.
-
-                \item Tritone is dissonance in Western music because
-                it characterizes the ``dominant'' in the tonal system (see
+                \item The perfect fourth can be a perfect consonance or
+                a dissonance according to the context and theoretical
+                background. As a general rule, it can be considered
+                a consonance except when it resolves into a
+                fifth or a third with movimentation of just one note .
+
+                \item Tritone is a dissonance in Western music because
+                it characterizes the ``dominant'' chord in the tonal system (see
                 subsection~\ref{subsec:harmonia}) and represents the
-                instability. Some cultures chant the interval as consonance.
+                instability. Some cultures have the interval as a consonance,
+                using it in melodies and chants as a stable interval.
 
@@
 
-                \item A perfect interval (unison, perfect forth, perfect                    fifth, perfect octave), or a major interval (major second M2,                 major third M3, major sixth M6 or major seventh M7), increased                 by one semitone results in an augmented interval (e.g.\                    augmented third aug3 with five semitones).
-The augmented forth                 is also called tritone (aug4 ~ tri ~ TT).
+                \item A perfect interval (unison, perfect forth, perfect
+                fifth, perfect octave), or a major interval (major second M2,
+                major third M3, major sixth M6 or major seventh M7), increased
+                by one semitone results in an augmented interval (e.g.\
+                augmented third aug3 with five semitones). The augmented forth
+                is also called tritone (aug4 ~ tri ~ TT).
 
@@
                 minor third m3, minor sixth m6 or minor seventh m7), decreased
-                by one semitone results in a diminished interval. The                       diminished fifth is also called tritone (dim5 ~ tri ~ TT).
-
-                \item An augmented interval increased by one semitone results                 in a 'doubly augmented' interval and a diminished interval
-                decreased by one semitone results in a 'doubly diminished'
+                by one semitone results in a diminished interval. The
+                diminished fifth is also called tritone (dim5 ~ tri ~ TT).
+
+                \item An augmented interval increased by one semitone results
+                 in a `doubly augmented' interval and a diminished interval
+                decreased by one semitone results in a `doubly diminished'
                 interval.
@@
                 \item Notes played as a sequence in time configure a
-                melodic interval.
-The order of the notes: considering first                 the lowest note or the highest note, results in an ascending                 or descending interval, respectively.
+                melodic interval. The order of the notes: first
+                the lowest note or the highest note, results in an ascending
+                 or descending interval, respectively.
 
                 \item If the lower pitch is raised one octave, or if
-                the upper pitch is lowered one octave below, the interval is
-                inverted. When summed to its inversion an interval results in                 9 (e.g.\ m7 is inverted to M2: $m7+M2=9$). An inverted major
-                interval results in a minor interval and vice-versa.
-An                     inverted augmented interval results in a diminished interval                 and vice-versa (like a doubly augmented results in a doubly
-                diminished and vice-verse). An inverted perfect interval                    results in a perfect interval as well.
+                the upper pitch is lowered one octave, the interval is
+                inverted. The sum of an interval and its inversion is
+                9 (e.g.\ m7 is inverted to M2: $m7+M2=9$). An inverted major
+                interval results in a minor interval and vice-versa. An
+                inverted augmented interval results in a diminished interval
+                and vice-versa (inverting a doublyi-augmented results in a
+                doubly-diminished and vice-verse, etc.).
+                An inverted perfect interval results in a perfect interval as well.
 
@@
                 interval' and is classified like the interval between the same
-                notes but in the same octave.
-They are also specified adding a                 7 to the interval: P11 is an octave plus a forth ($7 + P4 =                  P11$), M9 is an octave plus a major second ($7 + M2 = M9$).
+                notes but in the same octave. They are also reached by adding a
+                7 to the interval: P11 is an octave plus a forth ($7 + P4 = P11$),
+                M9 is an octave plus a major second ($7 + M2 = M9$).
 \end{itemize}
@@
 The augmented/diminished intervals and the doubly augmented/doubly diminished intervals are consequences of the tonal system.
-The scale degrees (subsection~\ref{subsec:escalas}) generally determine different pitches, with
+Scale degrees (subsection~\ref{subsec:escalas}) are in fact different pitches, with
 specific uses and functions.
 Henceforth, in a \textit{C flat} major scale, the tonic -- first degree -- is \textit{C flat}, not \textit{B}, and the leading tone -- seventh degree -- is \textit{B flat}, not \textit{A sharp} or \textit{C double flat}.
-In a similar way, the second degree of a scale can be one semitone far from the first degree like the leading tone (seventh degree at one ascending semitone from the first degree), where there is a diminished third between the two semitones of the seventh and second scale degrees, consequence of the first degree be between the near two degrees: the second and the leading note~\cite{Lacerda}.
-
-The descriptions presented here summarizes the traditional theory of the music
+In a similar way, the second degree of a scale can be one semitone from first degree, like the leading tone (seventh degree at one ascending semitone from the first degree), where there is a diminished third between the two semitones of the seventh and second scale degrees~\cite{Lacerda}.
+
+The descriptions presented summarizes the traditional theory of the music
 intervals~\cite{Lacerda}. The music piece \emph{Intervalos entre alturas}
-explores these intervals in a independently and conjunction manner.
-The reader is invited to see the source code in Appendix~\ref{ap:intervalos} and available online in the \massa toolbox~\cite{MASSA}.
+explores these intervals in both independent and related manners. 
+The source code is available online in the \massa\ toolbox~\cite{MASSA}.
 
@@
 
-Basically speaking, scale is a set of ordered pitches.
-Usually, scales repeat at each octave.
-The ascending sequence with all notes from the octave division in 12 equal intervals, separated by ratio $\varepsilon=2^{\frac{1}{12}}$, is known as the chromatic scale with equal temperament.
-There are 5 perfect symmetric divisions of the octave within the chromatic scale.
-These divisions are considered as scales by the easily and peculiar use they provide.
-Considering the integers that the factor $\varepsilon=2^{\frac{1}{12}}$ is powered to multiply $f_0$, the scales are the following:
-
-\begin{multline}\label{escSim}
-%\begin{split}
-\text{chromatic}  = E_i^c = \\ = \{e_i^c\}_0^{11} =  \{0,1,2,3,4,5,6,7,8,9,10,11\} = \{i\}_0^{11}\\
-\text{whole tones}  = E_i^t = \{e_i^t\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
-\text{minor thirds}  = E_i^{tm} = \{e_i^{tm}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
-\text{major thirds}  = E_i^{tM} = \{e_i^{tM}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
-\text{tritones}  = E_i^{tt} = \{e_i^{tt}\}_0^{1} = \{ 0, 6 \} = \{6.i\}_0^1
-%\end{split}
-\end{multline}
+A scale is an ordered set of pitches. Usually, scales repeat at each octave. 
+The ascending sequence with all notes from the octave division in 12 equal intervals, 
+separated by the ratio $\varepsilon=2^{\frac{1}{12}}$, is known as the chromatic scale with equal temperament.
+There are 5 perfectly symmetric divisions of the octave within the chromatic scale. 
+These divisions are considered scales by easy and peculiar uses they provide.
+Considering integers $e_i$ used as powers of the factor $\varepsilon=2^{\frac{1}{12}}$
+to multiply lower note frequency ($f=\varepsilon^{e_i} f_0$),
+the scales are as follows:
+
+\begin{equation}\label{escSim}
+\begin{split}
+\text{chromatic}    & = E_i^c = \{e_i^c\}_0^{11} = \\
+                    & =  \{0,1,2,3,4,5,6,7,8,9,10,11\} = \{i\}_0^{11}\\
+\text{whole tones}  & = E_i^t = \{e_i^t\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
+\text{minor thirds} & = E_i^{tm} = \{e_i^{tm}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
+\text{major thirds} & = E_i^{tM} = \{e_i^{tM}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
+\text{tritones}     & = E_i^{tt} = \{e_i^{tt}\}_0^{1} = \{ 0, 6 \} = \{6.i\}_0^1
+\end{split}
+\end{equation}
 
@@
 is $f_3=\varepsilon^{e_3^t} . f_0 = 2^{\frac{6}{12}} . 200 \approxeq 282.843
-Hz$. These 'scales' or patterns generate stable structures by their intern
-symmetries and can be repeated in a efficient and sustained way. Section~\ref{estCic} specially discusses the symmetries. 
+Hz$. These `scales', or patterns, generate stable structures by their intern
+symmetries and can be repeated in an efficient and sustained way.
+Section~\ref{estCic} specially discusses symmetries. 
 The music piece \emph{Cristais} exposes each one of these scales, in both melodic and
-harmonic counterpart and the corresponding source code is presented in Appendix~\ref{ap:cristais} and as
-part of \massa.
-
-The diatonic scales are:
-
-\begin{multline}\label{eq:escalas}
-%\begin{split}
-\text{natural minor} = \text{aeolian mode}  = \\ = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
-\text{locrian mode}  = E_i^{mlo} = \{e_i^{mlo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
-\text{major}  = \text{ionian mode}  =  E_i^M = \{e_i^M\}_0^6 = \{0,2,4,5,7,9,11\} \\
-\text{dorian mode}  = E_i^{md} = \{e_i^{md}\}_0^6 = \{0,2,3,5,7,9,10\} \\
-\text{phrygian mode}  = E_i^{mf} = \{e_i^{mf}\}_0^6 = \{0,1,3,5,7,8,10\} \\
-\text{lydian mode}  = E_i^{ml}=\{e_i^{ml}\}_0^6 = \{0,2,4,6,7,9,11\} \\
-\text{mixolydian mode} = E_i^{mmi} = \{e_i^{mmi}\}_0^6 = \{0,2,4,5,7,9,10\}
-%\end{split}
-\end{multline}
-
-\noindent They have only the major, minor and perfect intervals. The unique exception is the tritone that is presented as a augmented forth or diminished fifth.
-
-All the diatonic scales follow a pattern of successive intervals \textit{tone,
+harmonic counterpart and their source code is part of \massa\ toolbox.
+
+Diatonic modes are:
+
+\begin{equation}\label{eq:escalas}
+\begin{split}
+\text{aeolian}    & = \text{natural minor scale} = \\
+                  & = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
+\text{locrian}    & = E_i^{mlo} = \{e_i^{mlo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
+\text{ionian}     & = \text{major scale} =  \\
+                       & = E_i^M = \{e_i^M\}_0^6 = \{0,2,4,5,7,9,11\} \\
+\text{dorian}     & = E_i^{md} = \{e_i^{md}\}_0^6 = \{0,2,3,5,7,9,10\} \\
+\text{phrygian}   & = E_i^{mf} = \{e_i^{mf}\}_0^6 = \{0,1,3,5,7,8,10\} \\
+\text{lydian}     & = E_i^{ml}=\{e_i^{ml}\}_0^6 = \{0,2,4,6,7,9,11\} \\
+\text{mixolydian} & = E_i^{mmi} = \{e_i^{mmi}\}_0^6 = \{0,2,4,5,7,9,10\}
+\end{split}
+\end{equation}
+
+\noindent They have only major, minor and perfect intervals.
+The unique exception is the tritone that is presented as an augmented forth or a diminished fifth.
+
+Diatonic scales follow a circular pattern of successive intervals \textit{tone,
 tone, semitone, tone, tone, tone, semitone}. Thus, it is possible to write:
@@
 \noindent with $\kappa \in \mathbb{N}$. For each mode there is only one value
-for $\kappa \in [0,6]$. For example, a brief inspection reveals that
+for $\kappa \in [0,6]$ by which $\{e_i\}$ matches. 
+For example, a brief inspection reveals that
 $e_i^{ml}=d_{(i+2)\%7}+e_{i-1}^{ml}$. Then, $\kappa=2$ for the lydian mode.
@@
 
-\begin{multline}\label{eq:escalasMenores}
-%\begin{split}
-\text{natural minor (same as above)}  = \\ = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
-\text{harmonic minor}  = E_i^{mh} = \{e_i^{mh}\}_0^6 = \{0,2,3,5,7,8,11\} \\
-\text{melodic minor}  = \\ = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
-%\end{split}
-\end{multline}
-
-The ascending and descending contour of the melodic minor scale is necessary for the existence of the leading tone (seventh and last degree, separated by one semitone of the octave, enhances the tonic polarization) in the ascending trajectory, which is not necessary in the descending mode, since it recovers the normal form.
-The harmonic scale presents the leading tone but does not avoid the augmented second interval between the sixth and seventh degrees.
-In addition, it does not need to cover a melodic trajectory, just having to present the leading tone, crucial to the tonal system (the leading tone tends to the tonic, asserting it)~\cite{Harmonia}.
+\begin{equation}\label{eq:escalasMenores}
+\begin{split}
+\text{natural minor}&  = E_i^m = \{e_i^m\}_0^6 = \\
+                                           &  = \{0,2,3,5,7,8,10\} \\
+\text{harmonic minor}                      &  = E_i^{mh} = \{e_i^{mh}\}_0^6 = \\
+                                           &  = \{0,2,3,5,7,8,11\} \\
+\text{melodic minor}                       &  = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \\
+                                           &  = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
+\end{split}
+\end{equation}
+
+The different ascending and descending contour of the melodic minor scale is required
+by the leading tone (seventh and last degree, separated by 
+one semitone from the octave, enhances tonic polarization) in the ascending
+ trajectory, which is not necessary in the descending mode, and therefore it recovers the normal form.
+The harmonic scale presents the leading tone but does not avoid the augmented second
+between the sixth and seventh degrees, it does not consider the melodic trajectory~\cite{Harmonia}. 
 Other scales can be represented in a similar form, like the pentatonic and the
@@
 
-Although not a scale, the harmonic series is often used as such. Besides, this point is a convenient moment to put it plain simple:
+Although not a scale, the harmonic series is often used as such.
+Besides, this point is a convenient moment to put it plain simple:
 
@@
 
-This scale is composed of the frequencies on harmonic spectra. Nature exhibits this frequencies within sounds and with all kinds of distortions.
+In this scale, the frequency of $i-th$ note $h_i$ the frequency of $i-th$ harmonic $f=\varepsilon^{h_i} f_0$from the spectra generated by $f_0$.
+Nature exhibits this frequencies within sounds and with all kinds of distortions.
 

### 2013-10-06 article/article.tex
feito ateh modulacao

@@
 
-The simultaneous occurrence of three or more notes is observed by means of chords. Their
-base of tonal music are the triads. Triads are built by two successive thirds
-within 3 notes: root, third and fifth. If the lower note of a chord is different from the root, this chord is referred
-as inverted. A close position is where any note of chord fits between two consecutive
-notes~\cite{Lacerda}. All the non-inverted triads in the close position form and with root in
-$0$ are described as following:
+The simultaneous occurrence of three or more notes is observed by means of chords.
+Chords are based on triads, specially in tonal music. Triads are built by two successive thirds
+within 3 notes: root, third and fifth. If the lower note of a chord is different from the root, this chord is 
+an inverted chord as oposed to the chord in its root position. A closed position is any in which no chord note 
+fits between two consecutive
+notes~\cite{Lacerda}, any non-closed position is an open position. In closed and fundamental position,
+and with fundamental in $0$, triads are described as:
 
@@
 
-To consider another third superimposed to the fifth, is sufficient to add $10$
+To have another third superimposed to the fifth, is sufficient to add $10$
 as the highest note in order to form a tetrad with minor seventh, or add $11$ in order to form a tetrad with major
-seventh. The inversions and open positions can be obtained with a selective
-addition of $12$ to the components.
+seventh. Inversions and open positions can be obtained with the 
+addition of $\pm 12$ to selected component.
 
@@
 
-In the following we present general recommendations:
+For general guidance:
 
 \begin{itemize}
-        \item A fifth is a root (fundamental) confirmed by the interval.
-        \item The major or minor third points to the major or minor quality
-        of the chord.
+        \item A fifth confirms the root (fundamental).
+        \item Major or minor thirds, suggests major or minor chord qualities.
         \item Every tritone, specially if built between a major third and a
-        minor seventh, tends to resolve in a third or sixth.
+        minor seventh, tends to resolve in a third or a sixth.
         \item Note duplication is avoided. If duplication is needed, the
         preferred order is: the root, fifth, third and seventh.
+        \item Note omission is avoided in the triad. If needed, the fifth is
+        first considered for omission, than third and, if needed, the fundamental.
         \item It is possible to build chords with notes different from triads,
-        specially if they respect to a recurrent logic or musical chain that
+        specially if they respect a recurrent logic or musical concatenation that
         justifies these different notes.
         \item Chords built by successive intervals different of thirds -- like
-        fourths and seconds -- are recurrent in compositions with advanced
+        fourths and seconds -- are recurrent in compositions of advanced
         tonalism or experimental character.
-        \item The repetition of chords successions (or their characteristics)
-        fixes a trajectory by means of the recurrence and make possible the
+        \item The repetition of chord successions (or their characteristics)
+        fixes a trajectory and makes possible the
         introduction of exotic formations without incoherence.
@@
 
-Omission of the basic chaining is the key to obtain modal and atonal harmonies. In
-case of absence of these minimal tonal structures, we say that the
-harmony is modal if the notes match with some diatonic scale (see
-equations~\ref{eq:escalas}) or if the notes are presented in a small number.
-If basic tonal chaining are absent, the notes do not match any diatonic scale
+Omission of basic tonal chaining is the key to obtain modal and atonal harmonies. In
+absence of these minimal tonal structures, the
+harmony is considered modal if the notes match with some diatonic scale (see
+equations~\ref{eq:escalas}) or if notes are presented in a small number.
+If basic tonal chainings are absent and notes does not match any diatonic scale
 and are diverse and dissonant (by relation with each other) enough to avoid
 reduction by polarization, the harmony is atonal. In this classification, the
-modal harmony are not tonal neither atonal. The modal harmony is reduced to the 
+modal harmony is not tonal or atonal and is reduced to the 
 incidence of notes within the diatonic scale (or simplifications) and to the
 absence of tonal structures. It is possible to note, following this concept, that
-atonal harmony is hard to be realized~\cite{harmEXT}.
+atonal harmony is hard to be realized and, indeed, no matter how dissonant and diverse
+is a set of notes, tonal harmonies arise if not avoided~\cite{harmEXT}.
 
@@
 
-In fact, the techniques around atonal music comprehend structures for avoiding
-the hearing relation with modes and tonality. This difficulty imposes the use of dodecafonism. 
-The purpose of dedocafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
-order (não sei se ficou mt claro). In this context, the tonic becomes difficult to settle. Nevertheless, the
-Western hearing searches for tonal traces in musics and easily finds them by
-unexpected paths and even tortuous.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The use of dissonant intervals (specially tritones without resolution), together
-with ninths, seconds and sevenths, reinforces the absence of tonality. Considering
-this context, it is possible to account for the following considerations while creating a music
-piece:
+In fact, the techniques around atonal music aim at structures for avoiding
+direct relation with modes and tonality. Manifesting such structures
+are of such difficulty that the dodecafonism arouse. 
+The purpose of dedocafonism is to use a set of notes (ideally 12 notes),
+and to perform each note, one by one, in the same
+order. In this context, the tonic becomes difficult to be stablished. Nevertheless, the
+Western hearing searches for tonal traces in music and easily finds them by
+unexpected and tortuous paths.
+
+The use of dissonant intervals (specially tritones) without resolution,
+reinforces the absence of tonality. Considering
+this context, while creating a music
+piece, it is possible to:
 
 \begin{itemize}
-     \item Repeating notes. Considering the immediate repetition as an extension
-     of the previous incidence, the use of same notes in sequence does not add relevant information.
-     \item It is interesting to play adjacent notes at the same time, making harmonic intervals
+     \item Repeat notes. By considering immediate repetition as an extension
+     of previous incidence, the use of same note in sequence does not add relevant information.
+     \item To play adjacent notes at the same time, making harmonic intervals
      and chords.
-     \item Present durations with liberty, respecting the notes order.
-     \item In order to gain variation, one can use extension, transposition, translation, inversion, retrograde and retrograde inversion. See subsections~\ref{subsec:motivos}
+     \item Present durations and pauses with liberty, respecting notes order.
+     \item Vary by extension, transposition, translation, inversion, retrograde and retrograde inversion. See subsections~\ref{subsec:motivos}
      and~\ref{subsec:usosmusicais3} for more details.
@@
 The atonal harmony can be observed, paradigmatically, within these presented
-conditions. Most of them were written by great dodecafonic composers like Alban Berg and even Schoenberg. Many pieces create purposeful links between tonal and atonal techniques.
+conditions (a simple dodecaphonic model). Most of what was written by great dodecafonic composers, 
+like Alban Berg and even Schoenberg, had the purpose of mixing tonal and atonal techniques.
 
@@
 
-In the XX century, rhythmic music with emphasis on sonorities/timbres, extended
-the concepts about tonality and harmony. Nevertheless, the
-tonal harmony has strong influence in the artistic movements and commercial
-venues. In addition, the dodecafonism is considered having tonal nature because it
-was created to deny the tonal characteristics of polarization.
-
-In tonal or modal music, the base of the harmonic filed are chords -- like the ones listed in
-equations~\ref{triades}-- built with a root note for each degree of a scale -- like
-the ones in equation~\ref{eq:escalas}.
-In this context, music harmony aims at studying the observation of incident chord progressions and chaining rules.
+In the XX century, rhythmic music and with emphasis on sonorities/timbres, extended
+the concepts of tonality and harmony. Hence, 
+tonal harmony has strong influence in artistic movements and commercial
+venues. In addition, the dodecafonism itself is considered of tonal nature because it
+denies tonal characteristics of polarization.
+
+In tonal or modal music, chords -- like the ones listed in
+equations~\ref{triades}-- built on the root note of each degree of a scale  -- such 
+as listed in equations~\ref{eq:escalas} --  form the pilars of the harmonic field. 
+Music harmony aims at studying the incidence of chord progressions and chaining rules. 
 Even a monophonic melody generates harmonic fields, making it possible to observe the suggested chords at each passage.
 
-In the 'traditional tonal harmony', a scale has its corresponding tonic (first scale
-degree) and can be major (with the same notes of the Ionian mode) or
-minor (notes from Eolian mode form the 'natural minor', which has both a
-harmonic and a melodic version, see equation~\ref{eq:escalasMenores}). The
-scale is base on triads, each one with its root and representing a degree into
-the scale: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$. 
+In the traditional tonal music, a scale has its tonic (first
+degree) on any note, and can be major (with the same notes of Ionian mode) or
+minor (same notes as Eolian mode, 'natural minor', which has both
+harmonic and melodic versions, see equations~\ref{eq:escalasMenores}). The
+scale is the base for triads, each with its root in a degree: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$. 
 To build triads, the third and the fifth note above the root is considered together with the root (or fundamental).
-It is possible to write $\hat{1},\hat{3},\hat{5}$ as the first degree chord, built by the first degree of scale and the center for a tonal music.
-The chords
+$\hat{1},\hat{3},\hat{5}$ is the first degree chord,
+ built on top of scale's first degree and central for tonal music. The chords
 of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when a minor
 scale) and of the forth degree $\hat{4},\hat{6},\hat{1}$ are secondary. After
-that, other degrees are considered. The traditional harmony comprises conventions and stylistic techniques to create chord chains, made in each scale degree~\cite{Harmonia}.
-
-The 'functional harmony' ascribes functions to these three central chords and
-tries to understand their use by means of these functions. The chord built under
-the first degree is named the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its (role) function consisting of preserving a center, usually referred as a ``ground'' for the music.
-The chord built under the fifth degree is the \textbf{dominant} (\textit{D}, the dominant is always major) and its function is call for the tonic (we say that the dominant chord asks for a conclusion and this conclusion is usually the tonic chord).
+these, other degrees are considered. The `traditional harmony' comprises conventions
+ and stylistic techniques to create chainings with such chords~\cite{Harmonia}.
+
+The `functional harmony' ascribes functions to these three central chords and
+tries to understand their use by means of these functions. The chord built on top of
+first degree is the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its function (role) consists of maintaining a center, usually referred as a ``ground'' for the music.
+The chord built on the fifth degree is the \textbf{dominant} (\textit{D}, the dominant is always major) and its function is to lean for the tonic
+(the dominant chord asks for a conclusion and this conclusion is the tonic).
 Thus, the dominant chord guides the music to the tonic.
 The triad built under the forth degree is the \textbf{subdominant} (\textit{S} or \textit{s} for a major or minor
-subdominant, respectively) and its basically function is to distance the music from the tonic. The process aims at confirming the tonic using tonic-subdominant-tonic chains which are
+subdominant, respectively) and its function is to deviate the music from the tonic. 
+The process aims at confirming the tonic using tonic-dominant-tonic chains which are
 expanded by using other chords in different ways.
 
-The remaining triads are associated to these three most important chords. In the major scale, the associated relative (relative tonic \textit{Tr}, relative
-subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built with a
+The remaining triads are associated to these three most important chords.
+ In the major scale, the associated relative (relative tonic \textit{Tr}, relative
+subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built a
 third below, and the associated counter-relative (counter-relative
 tonic \textit{Ta}, counter-relative subdominant \textit{Sa} and the
-counter-relative dominant \textit{Da}) is the triad built with a third above. In the
-minor scale the same happens, but the triad with a third below is called
-counter-relative (tA, sA) and the triad with one third above is called relative (tR,
+counter-relative dominant \textit{Da}) is the triad built in a third above. 
+In the minor scale the same happens, but the triad a third below is called
+counter-relative (tA, sA) and the triad a third above is called relative (tR,
 sR). The precise functions and musical effects of these chords are
-controversial. Table~\ref{tab:harmonia} shows the relation between the triads
+controversial. Table~\ref{tab:harmonia} shows relations between the triads
 built in each degree of the major scale.
@@
 in the forth degree by a semitone above $\hat{7}\#$. The diminished chord
-$\hat{7},\hat{2},\hat{4}$, is generally considered a 'dominant seventh chord
-with no root'~\cite{Koellheuteur}.
-In the minor mode, there is the change in $\hat{7}$ by an ascending semitone to
-make possible the existence of an unique semitone separating the tonic, and
-making also possible the dominant (that should be major and goes to the tonic). In
+$\hat{7},\hat{2},\hat{4}$, is generally considered a `dominant seventh chord
+with omissed root'~\cite{Koellheuteur}.
+In minor mode, there is a change in $\hat{7}$ by an ascending semitone to
+make possible the existence of an unique semitone separating $\hat{7}$ and $\hat{1}$,
+and making the dominant possible (that should be major and goes to the tonic). In
 this way, the dominant is always major, for both major and minor scale and,
-because of that, even as a minor tone, the relative dominant remains a third
+because of that, even in a minor tone, the relative dominant remains a third
 below and in the counter-relative, a third above.
@@
 Each one of these chords can be confirmed and developed by performing their
-dominant or individual subdominant, which is the chord based in the third formed
-by a fifth above or by a fifth below, respectively. These dominants and individual
-subdominants, in the same way, have also subdominants and individual dominants that
-can be used. Given a tonality, any chord can occur, no matter
+individual dominant or subdominant, which are the triads based
+a fifth above or a fifth below, respectively. These individual dominants and subdominants,
+in the same way, have also subdominants and individual dominants of their own.
+Given a tonality, any chord can occur, no matter
 how distant it is from the harmonic field and from the notes within the
-scale. The unique on condition is that the occurrence presents a coherent trajectory of
+scale. The unique condition is that the occurrence presents a coherent trajectory of
 dominants and subdominants to the original tonality.
 
-Mediants, or 'chromatic mediants', are present in each chord in two ways: the upper chromatic
+Mediants, or 'chromatic mediants', are two for each chord: the upper
 mediant, formed by the root at the third of original chord; and the lower
-chromatic mediant, formed by the fifth at the third of original chord. Both chords also are formed by a third, but with a chromatic alteration regarding
+mediant, formed by the fifth at the third of original chord.
+Both chords also are formed by a third, but with a chromatic alteration regarding
 the original chord. If two chromatic alterations exist, i.e.\ two notes altered
-by one semitone each regarding the original chord, the mediant is called
+by one semitone each regarding the original chord, it is a
 'doubly chromatic mediant'. Again, there are two forms for each chord: the upper form, with
-a third in the fifth of the original chord; and the lower form, with a third in the root
-of the original triad.
-It is interesting to observe that a major chord has upper chromatic mediants and upper doubly chromatic mediants.
-A minor chord has lower chromatic mediants and lower doubly chromatic mediants.
-This relation between chords is considered as
-advanced tonalism, sometimes even considered as expansion and dissolution of tonalism, and, as consequence, it has strong and outstanding effects although perfectly consonant. The chromatic
+a third in the fifth of the original triad; and the lower form, with a third in the root
+of the original triad. It is interesting to observe that a major chord has major chromatic
+ and doubly chromatic mediants. A minor chord has minor chromatic and doubly chromatic mediants.
+(Recall that relatives and counter-relatives have oposite major-minor quality.)
+This relation between chords is considered
+advanced tonalism, sometimes even considered expansion and dissolution of tonalism,
+with strong and impressing effects although perfectly consonant triads. Chromatic
 mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss,

### 2013-10-06 article/article.tex
feito ateh comeco de ritmo

@@
 
-Modulation is the change of the key (tonic, or tonal center) in a music.
-It can be characterized by observing the key at the start and at end of the chord transitions.
+Modulation is the change of key (tonic, or tonal center) in a music. 
+It is characterized by observance of start and end keys, and transition artifacts.
 Keys are always conceived as related by fifths and their relative and counter-relatives.
-Forms to perform modulation include:
+Ways to perform modulation:
 
@@
     called phrasal modulation or unprepared modulation.
-    \item  Using carefully an individual dominant and perhaps the individual
-    subdominant, to confirm the change in key and harmonic field.
-    \item Using chromatic alterations to reach a chord in the new key starting from a
-    chord in the previous key. It is called chromatic modulation.
-    \item Featuring a unique note, possibly repeated or suspended with no
-    accompaniment, common to start and end a key, constitute a peculiar way
+    \item  Careful use of an individual dominant, and perhaps also the individual
+    subdominant, to confirm change in key and harmonic field.
+    \item Use of chromatic alterations to reach a chord in the new key starting from a
+    chord in the previous key. Called chromatic modulation.
+    \item Featuring an unique note, possibly repeated or suspended with no
+    accompaniment, common to start and end keys, constitute a peculiar way
     to introduce the new harmonic field.
@@
     \item Maintaining the tonal center and changing the key quality from major to minor
-    (or vice-verse) comprehends a parallel modulation. Key with same tonic and
+    (or vice-verse) comprehends a parallel modulation. Keys with same tonic and
     different quality is known as homonyms.
@@
 which leads to the circle of fifths~\cite{Harmonia,Salzer,Koellheuteur,Harmony}.
+Other inventive ways to modulate are possible, to point but one common example,
+the minor thirds tetrad ($E_i^{tm}$ in equations~\ref{escSim}) can be susteined to bridge to other tonalities, with the ease
+of its both tritones.
 The music piece \emph{Acorde cedo} explores these chord relations. Its code is
-available both in Appendix~\ref{ap:acorde} and online as part of \massa~\cite{MASSA}.
+online as part of \massa\ toolbox~\cite{MASSA}.
 
@@
 
-Counterpoint is defined as the conduction of simultaneous melodic lines, called voices.
+Counterpoint is defined as the conduction of simultaneous melodic lines, or ``voices''.
 The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues.
 It is possible to
-summarize the main counterpoint rules, and it is known that Beethoven --
-among others -- outlined such similar synthesis.
+summarize counterpoint rules, and it is known that Beethoven --
+among others -- also outlined such a briefing of counterpoint.
 
@@
     \caption{Different motions of counterpoint aiming to preserve independence
-        between voices. There are 3 types of motions: direct, contrary and
+        between voices. There are 3 types of motion: direct, contrary and
         oblique. The parallel motion is a type of direct motion.}
@@
 
-The purpose of counterpoint is to conduct voices in a way they sound
-independent. In order to do that, the relative motion of voices (in pairs), is crucial and
-categorized as: direct, oblique and contrary motion (see Figure~\ref{fig:movContraponto}). The parallel motion is an oblique motion.
-The gold rule here is to take care with the direct motions, avoiding making them
-lead to a perfect consonance. The parallel motion should occur only between
-imperfect consonances and no more than three consecutive times. The dissonances
+The purpose of counterpoint is to conduct voices in a way that they sound
+independent. In order to do that, the relative motion of voices (in pairs) is crucial and
+categorized as: direct, oblique and contrary (see Figure~\ref{fig:movContraponto}).
+The parallel motion is an direct motion.
+The gold rule here is to take care of the direct motions, avoiding them
+when leading to a perfect consonance. The parallel motion should occur only between
+imperfect consonances and no more than three consecutive times. Dissonances
 can be unadmitted or used when followed and preceded by consonances of joint
-degrees, i.e.\ neighbor notes in a scale. The motions that lead a note to a
-neighbor sound more coherent. When having 3 or more voices, the melodic
-importance lies on the higher and lower voices, in this particular
-order~\cite{Fux,Tragtenberg,SchoenbergContra}.
+degrees, i.e.\ neighbor notes in a scale. The motions that lead to a
+neighbor note in the scale sound coherent. When having 3 or more voices, the melodic
+importance lies on the higher and lower voices, in this order~\cite{Fux,Tragtenberg,SchoenbergContra}.
 
 These rules were used in the music piece \emph{Conta ponto}, the source code is
-in Appendix~\ref{ap:conta} and available online in \massa.
+available online in \massa\ toolbox.
 
@@
 
-Basically speaking, the rhythm notion is dependent on events separated by durations~\cite{Lacerda}.
+Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}.
 These events can be heard individually if spaced by at least $50-63ms$. 
-To make the temporal separation between them be perceived as duration, the period should be large enough, around $100ms$~\cite{microsound}.
-It is possible to summarize the duration transitions heard as rhythm and pitch,
-in the following way~\cite{Alfaix, microsound}.
+For the temporal separation between them to be perceived as duration,
+the period should even a bit larger, around $100ms$~\cite{microsound}.  
+It is possible to summarize duration heard as rhythm or pitch, and its transition,
+as in table~\ref{tab:duracoes}~\cite{Alfaix, microsound}.
 
@@
 
-The duration region marked as transition is presented minimized because the limits
+The duration transition span is minimized because the limits
 are not well defined. In fact, the duration where someone begins to perceive a
 fundamental frequency or a separation between occurrences, depends on the
-listener and sound characteristics~\cite{microsound,Roederer}.
+listener and sonic characteristics~\cite{microsound,Roederer}.
 

### 2013-10-06 article/article.tex
feito ateh metade do ritmo Ternary relations are common in rit-

@@
 pulse typically comprehends durations between $0.25-1.5s$ ($240$
-and $40 BPM$, respectively). In music education and cognitively studies, it is common to
-associate these range of pulse frequencies to the durations of the heart beat,
-movements of inspiration and expiration and walk steps~\cite{Lacerda,Roederer}.
-
-The pulse is subdivided into equal parts and repeated in sequence. These
-relations (division and concatenation) usually follow relations of small
-order integers\footnote{In ascending order of occurrence in written and ethnic
-music, these indicate the musical pulse divisions and their sequential grouping at
-time: 2, 4 and 8, after that 3, 6 (two groups of 3 or 3 groups of 2) and 9 and
+and $40 BPM$, respectively). In music education and cognitive studies, it is common to
+associate these range of frequencies to the durations of the heart beat,
+movements of inspiration and expiration and steps of a walking or running person~\cite{Lacerda,Roederer}.
+
+The pulse is subdivided into equal parts and is also repeated in sequence.
+These relations (division and concatenation) usually follow relations of small
+order integers. The occurrences, in written and ethnic
+music, of musical pulse division (and their sequential grouping at
+time),  are, in ascendent disposition: 2, 4 and 8, after that 3, 6 (two groups of 3 or 3 groups of 2) and 9 and
 12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, complementing
-1-9 and 12. Other metrics are less common, like divisions and grouping in 13,
+1-9 and 12. Other metrics are less common, like division and grouping in 13,
 17, etc, and are mainly used in contexts of experimental music and classical
-music of XX and XXI. No matter how complex they seem, metrics are commonly
-compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.}.
-A schematic exposition is shown in Figure~\ref{fig:pulsoSubAgl}.
-
-\begin{figure}[h!]
+music of XX and XXI. No matter how complex they seem, metrics are common
+compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
+A schematic exposition is shown in figure~\ref{fig:pulsoSubAgl}.
+
+\begin{figure*}
     \centering
-        \includegraphics[width=\columnwidth]{figures/metricaMusical}
-    \caption{Divisions and grouping in bars for the music pulse to establish a rhythmic
-        metric. Divisions of the quarter note, established as the
-        pulse, is presented in the left. The time signature whose specifies the same
-        metrics but having as scale the groupings of the music pulse is presented in the right.}
+        \includegraphics[width=\textwidth]{figures/metricaMusical}
+    \caption{Divisions and groupings of the musical pulse for establishing a
+        metric. Divisions of the quarter note, regarded as the
+        pulse, is presented in the left. The time signature yielded by
+        groupings of the music pulse is presented in the right.}
         \label{fig:pulsoSubAgl}
-\end{figure}
-
-Dual relations (common time and binary divisions) are frequent in dance rhythms
-and celebrations. Ternary relations are common in
-ritualistic music and is related to the sacred. Dual relations are called imperfect, while and the ternary relations are called perfect.
-
-The strong unities (accents) are the ones that fall in the 'head' of bars, the downbeats. The
+\end{figure*}
+
+Dual relations (simple measures and binary divisions) are frequent in dance rhythms
+and celebrations, and are called ``imperfect''. Ternary relations are common in
+ritualistic music and is related to the sacred, and are called ``perfect''.
+
+Strong units (accents) fall in the `head' of bars, the downbeats, of which the first in
+every bar is the greatest. The
 head of a unity is the first part of the subdivision. In binary divisions (2, 4

### 2013-10-07 article/article.tex
feito ateh repeticao e variacao

@@
 every bar is the greatest. The
-head of a unity is the first part of the subdivision. In binary divisions (2, 4
-and 8, for example), the unities originally strong turn into weak (não sei se isso ficou claro)
-(e.g.\ division in 4 is: strong, weak, strong medium, weak). In ternary divisions
-(3, 6 and 9) two weak unities succeeds the downbeat (e.g.\ division in 3 is:
-strong, weak, weak)\footnote{Division in 6 is considered compound but can also
-occur as a binary division. A binary division which suffers a ternary division
-results in two units divided in tree units each: strong (subdivided in strong,
+head of a unit is the first part of the subdivision. In binary divisions (2, 4
+and 8), strong units alternate with weak units
+(e.g.\ division in 4 is: strong, weak, average strong, weak). In ternary divisions
+(3, 6 and 9) two weak units succeeds the downbeat (e.g.\ division in 3 is:
+strong, weak, weak). Division in 6 is considered compound but can also
+occur as a binary division. Binary division units which suffers a ternary division
+yields two units divided in tree units each: strong (subdivided in strong,
 weak, weak) and weak (subdivided in strong, weak, weak). Another way to perform
-the division in 6 is applying ternary division that suffers a binary division,
+the division in 6 is ternary division units that divides as binary,
 resulting in: a strong unit (subdivided in strong and weak) and two weak units
-(subdivided in strong and weak for each).}
-
-The accent in the weak beat is known as the backbeat, whereas a note starting in a weak
-beat which duration coming across the strong beat is known as syncope.
-
-Notes can occur inside and outside of these \emph{'music metric'} divisions. In
-most behaved cases, notes occur exactly on these divisions, with broader
-incidence on attacks of strong beats.
-In extreme cases, the time metric cannot be perceived~\cite{Roederer}. 
-Small grid variations help to build musical interpretation and differences
-between styles~\cite{Cook}.
-
-% ficou um tanto confuso:
-
-Being pulse the grouping level $j=0$, level $j=-1$ the first pulse subdivision,
-level $j=1$ the first pulse agglomeration and so on. In this way, $P_i^j$ is the
-$i$-th unit of pulses at grouping level $j$: $P^0_{10}$ is the tenth pulse,
-$P^{1}_3$ is the third pulse grouping unit (and, possibly, the third measure),
-$P^{-1}_2$ is the second unit of pulse subdivision.
-
-Special attention should be given to the limits of $j$: pulse divisions are
-durations sensible as rhythm. Furthermore, the pulse joints sums, at its
+(subdivided in strong and weak for each).
+
+The accent in the weak beat is the backbeat, whereas a note starting in a weak
+beat persisting across the strong beat is the syncope.
+
+Notes can occur inside and outside of these \emph{'musical metric'} divisions. In
+most well behaved cases, notes occur exactly on these divisions, with 
+greater incidence strong beat attacks.
+In extreme cases, time metric cannot be perceived~\cite{Roederer}. 
+Small variations along the grid are crucial for musical interpretation and
+different styles~\cite{Cook}.
+
+Be the pulse the grouping level $j=0$, level $j=-1$ is the first pulse subdivision,
+level $j=1$ is the first pulse agglomeration and so on. In this way, $P_i^j$ is the
+$i$-th unit at grouping level $j$: $P^0_{10}$ is the tenth pulse,
+$P^{1}_3$ is the third grouped unit (and, possibly, the third measure),
+$P^{-1}_2$ is the second part of pulse subdivision.
+
+The limits of $j$ are of special interest: pulse divisions are
+durations perceivable as rhythm; furthermore, the pulses sum, at its
 maximum, a music or a cohesive set of musics. In other words, a duration given
@@
 summed together $\sum_{\forall i}P^{\text{max}(j)}_i$ are less than a few
-minutes or at most, a few hours.
-
-Each level $j$ has some indexes $i$. A index $i$ always has three different
-values (or multiple of three) that occur in a perfect relation. When $i$ has only
-multiple of two, four or eight different values, than an imperfect relation occurs,
-as shown in Figure~\ref{fig:pulsoSubAgl}.
-
-Any unit (note) of a given musical sequence which has a time metric can be unequivocally
+minutes or, at most, a few hours.
+
+Each level $j$ has some sections $i$. When $i$ has three different
+values (or multiple of three) there is a perfect relation. When $i$ has only
+two, four or eight different values, than there is an imperfect relation,
+as shown in figure~\ref{fig:pulsoSubAgl}.
+
+Any unit (note) of a given musical sequence with a time metric can be unequivocally
 specified as:
@@
 
-\noindent where $j_k$ is the grouping level and$i_k$ is the unit order itself.
-
-As example, consider $P^{-1,0,1}_{3,2,2}$ as the third subdivision $P^{-1}_3$ of the
-second pulse $P^0_2$ and of the second pulse group $P^1_2$. Each unit or unit set
-$P_i^j$ can be associated with a sequence of temporal samples $T_i$ that forms a
-music note.
-
-The music piece \emph{Poli Hit Mia} explorers different metrics (avaleble in Appendix~\ref{ap:poli} and as part of \massa).
-
-\subsection{Repetition and variation: motifs and large units}\label{subsec:motivos}
+\noindent where $j_k$ is the grouping level and$i_k$ is the unit itself.
+
+As an example, consider $P^{-1,0,1}_{3,2,2}$ as the third subdivision $P^{-1}_3$ of the
+second pulse $P^0_2$ and of the second pulse group $P^1_2$ (possibly second measure).
+Each unit $P_i^j$ can be associated with a sequence of temporal samples $T_i$ that comprehends a
+note.
+
+The music piece \emph{Poli Hit Mia} explorers different metrics (available as part of \massa\ toolbox).
+
+\subsection{Repetition and variation: motifs and larger units}\label{subsec:motivos}
 

### 2013-10-07 article/article.tex
arrumando rep e var

@@
 
-Given the basic music structures, either frequential (chords and scales) or
-rhythmic (beat divisions and simple, compound and complex grouping), it is
-natural to organize these structures in a cohesive way~\cite{Boulez}. To this end,
-the concept of arcs are fundamental: we make an arc by departing from a place (music unit) and going to another one. The audition of melodic and harmonic lines is pervaded as
-musical arcs thanks to the cognitive nature of the musical hearing. It is
-possible to consider the note as the smaller arc, and each motif (definir motif aqui? nao foi falado ainda) and melody as another arc. Each time, beat subdivision, measure and music
-section constitutes an arc. A music in which the arcs do not present consistency by
-themselves, can be understood as a music with no cohesion. The coherence feeling
-comes, mostly, from the skilled manipulation of arcs in a music piece.
-
-Musical arcs are abstract structures and liable of basic operations. An spectral
+Given the basic musical structures, both frequential (chords and scales) and
+rhythmic (simple, compound and complex beat divisions and agglomerations), it is
+natural to present these structures in a coherent and meaningful way~\cite{Boulez}.
+For such, the concept of arcs is fundamental: by departing from a place and returning,
+an arc is made. The audition of melodic and harmonic lines is permeated by
+musical arcs due to cognitive nature of the musical hearing. 
+The note can be considered the smaller arc, and each motif and melody as an arc as well.
+Each beat and subdivision, each measure and music
+section, constitutes an own arc. A music in which the arcs do not present consistency with one another,
+can be understood as a music with no coherence. Coherence impression
+comes, mostly, from the skilled handling of arcs in a music piece.
+
+Musical arcs are abstract structures and liable of basic operations. A spectral
 arc, like a chord, can be inverted, enlarged and permuted, to cite some
-examples. Temporal arcs, like a melody, a motif, a measure or a note are also
+examples. Temporal arcs, like a melody, a motif, a measure or a note, are also
 capable of variations. Remembering that
@@
     where $\Gamma$ is the duration between the beginning of the piece (or considered
-    period) and the first event $s_0$ of original structure $S_j$. Observe that
+    section) and the first event $s_0$ of original structure $S_j$. Observe that
     $\delta$ is the time offset of the displacement.
@@
 
-    \item Pitch translation is a displacement $\tau$ of a material for a
-        different pitch $\Xi'=\Xi + \tau$. It is a variation that changes pitch
+    \item Pitch translation, or transposition, is a displacement $\tau$ of the material to a
+        different pitch $\Xi=\Xi_0 + \tau$. It is a variation that changes pitch
         localization of material:
         $\left\{s_j'\right\}=\left\{s_j^{\Xi'}\right\}=\left\{s_j^{\Xi+\tau}\right\}$
-        where $\Xi$ is the pitch of a period $S_j$ or of an event $s_0$ of the
-        original structure $S_j$. Observe that $\tau$ is the pitch offset of the
-        displacement. If $\tau$ is given in semitones, the displacement in
-        frequency is $\tau_f=f_0.2^{\frac{\tau}{12}}$ where $f_0$ results from
-        the adoption of some reference: $f_0=\Xi_{f_0}Hz\;\sim \Xi_0$ absolute
-        value of pitch. For the frequency of any pitch value: $\Xi_f=\Xi_{f_0}.2^{\frac{\Xi-\Xi_0}{12}}$. 
-        For the pitch of any frequency value: $\Xi=\Xi_0 +12
-        . \log_2\left(\frac{\Xi_f}{\Xi_{f_0}}\right)$.  
+        where $\Xi_0$ is a reference value, such as the pitch of a section $S_j$ or of the first event $s_0$.
+        If $\tau$ is given in semitones, the displacement in
+        frequency is $\tau_f=f_0.2^{\frac{\tau}{12}}$ where $f_0$ is the
+        reference frequency value: $f_0=\Xi_{f_0}Hz\;\sim \Xi$ absolute
+        value of pitch. The frequency of any pitch value is $f=f_0.2^{\frac{\tau}{12}}$s; 
+        and the pitch of any frequency is: $\Xi=\Xi_0 +12
+        \log_2\left(\frac{f}{f_0}\right)$.  
         In the MIDI protocol, $\Xi_{f_0}=55Hz$ while pitch $\Xi_0=33$ marks
-        a \textit{A 1}, another reference point is $\Xi_{f_0}=440Hz$ which
-        corresponds to $\Xi_0=69$. The unit difference is equal for semitones.
-        It is not possible to say that the unit is the semitone because $\Xi=1$
-        is not a semitone, but a note with an audible frequency as rhythm, with
-        less than 9 occurrences each second (see Table~\ref{tab:duracoes}).
-
-        \item Interval inversion is the inversion traversed by
-        the material. The inversion is strict if the number of semitones is
-        being used as a reference for the operation:
+        a \textit{A 1}. Another good MIDI reference is $\Xi_{f_0}=440Hz$ and
+        $\Xi_0=69$. The difference ($\Xi_1 - \Xi_2$) is measured in semitones.
+        $\Xi$ is not a measure in semitones: $\Xi=1$ is not a semitone,
+        it is a note with an audible frequency as rhythm, with
+        less than 9 occurrences each second (see table~\ref{tab:duracoes}).
+
+        \item Interval inversion is: 1) the inversion of note pitch order, within octave equivalence, such as exposed in~\ref{subsec:intervalos}, and $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{\varepsilon.
+f_0}\right\}$ with selective $s_j=2,\;1/2,\;\text{or}\;1$; or 2) the inversion of interval orientation.
+        In the former case, the number of semitones
+        is preserved in the ``strict invertion'':
         $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{-\varepsilon_j . f_0}\right\}$, where
-        $\varepsilon_j$ is the factor between the event frequency $s_j$ and the
+        $\varepsilon_j$ is the factor between the frequency of event $s_j$ and the
         frequency of $s_0$. The inversion is said tonal if the distances are
-        considered in terms of degree number of the scale $E_k$:
-        $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{-\varepsilon^{\left(e_{\left(j_e\right)}\right)}
-        . f_0}\right\}_0^{H-1}$ where $j_e=e_{j}^{inv}$ is the index $k=j_e$ in
-        $E_k$ of the note of event $s_j$.
+        considered in terms of the diaonic scale $E_k$:
+        $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{2^{\left(\frac{12-e_{\left(7-j_e\right)}}{12}\right)}
+        . f_0}\right\}_0^{H-1}$ where $j_e$ is the index in
+        $E_k$ of the note $s_j$.
 

### 2013-10-07 article/article.tex
feito ateh arcos direcionais

@@
 
-        \item Rotation of musical elements is the transference of the last
-        element to the position of the first element, and the displacement of
-        that to the penult, one position ahead. It is possible to define
-        rotation of $\tilde{n}$ positions by $S'_n=S_{(n+\tilde{n})\%H}$. If
+        \item Rotation of musical elements is the translation of all elements
+        a number of positions ahead or behind, with the care to fill empty positions
+        with events out of convenient slots. Thus, a
+        rotation of $\tilde{n}$ positions is $S'_n=S_{(n+\tilde{n})\%H}$. If
         $\tilde{n}<0$, it is enough to use $\tilde{n}'=H-\tilde{n}$. It is
-        reasonable to associate $\tilde{n}>0$ with the clockwise rotation and
-        $\tilde{n}<0$ with tne anti clockwise rotation. More information about rotations in presented in subsection~\ref{estCic}.
-
-        \item The insertion and removal of materials into structure $S_j$ can be
+        reasonable to associate $\tilde{n}<0$ (events advance) with the clockwise rotation and
+        $\tilde{n}<0$ (elements delay) with the anti-clockwise rotation. 
+        More information about rotations in presented in subsection~\ref{estCic}.
+
+        \item The insertion and removal of material in $S_j$ can be
     ornamental or structural: $S_j'=\{s_j'\}=\{s_j \text{ if condition A,
-    otherwise } r_j\}$, for any music material $r_j$, including the empty
-    instant. Elements of $R_j$ can be inserted at the beginning, like a prefix
-    in $S_j$; at the end, as a suffix; or at the middle, dividing $S_j$ or making
-    it the prefix or suffix. Both materials can be mixed in a variety of ways.
+    otherwise } r_j\}$, for any music material $r_j$, including silence.
+    Elements of $R_j$ can be inserted at the beginning, like a prefix
+    in $S_j$; at the end, as a suffix; or at the middle, dividing $S_j$ and making
+    it the prefix and suffix. Both materials can be mixed in a variety of ways.
 
@@
 
-Other proceesings can be derived together with the presented ones, as, for example, the retrograde
-inversion, temporal contraction with an external suffix, etc. The music
-structures resound in the human cognitive system due to the own nature of
-thought. In its many veneers, an idea read with the same number of elements and
-connective aspects between them. The music, when tunned with those mental structures,
-raises impressions. In this way, a whole process of mental and neurological
-resonance is unleashed, responsible by the feelings, memories and imaginations,
-typical to a mindful music listening. This cortical activity helps the
+With these presented processes, others are derived, as, for example, the inverted retrograde,
+a temporal contraction with an external suffix, etc. Music
+structures resonate in the human cognitive system due to the own nature of
+thought. In its many facets, an idea is observed with the same number of elements and
+connective aspects between them. The music, when tunned with mental structures,
+raises related impressions. In this way, a whole process of mental and neurological
+activity is unleashed, responsible by the feelings, memories and imaginations,
+typical of a diligent music listening. This cortical activity is critical to
 musical therapy, known by its utility in cases of depression and neurological
 injury. It is known that regions of the human brain responsible by
-hearing processing, are also used for other activities, even linguistic and mathematical~\cite{Sacks,Roederer}.
-
-The paradigmatic structures guide the creation of new music material. One of the
-most central of them is the dipole tension/relax. Traditional dipoles
+sonic processing, are also used for other activities, such as linguistic and mathematical~\cite{Sacks,Roederer}.
+
+Paradigmatic structures guide the creation of new musical material.
+One of the most central of them is the tension/relaxation dipole. Traditional dipoles
 are relate to tonic/dominant, repetition/variation,
 consonance/dissonance, coherence/rupture, symmetry/asymmetry,
-equality/difference, arrival/departure, near/far, stationary/moving,
-etc. The ternary relations tend to relate to the circular and unification. The
-lucid ternary communion, 'modus perfectus', opposes to the passionate
-dichotomic, 'modus imperfectus'. Hereafter, there is an exposition dedicated to
+equivalence/difference, arrival/departure, near/far, stationary/moving,
+etc. Ternary relations tend to relate to the circle and to unification. The
+lucid ternary communion, `modus perfectus', opposes to the passionate
+dichotomic, `modus imperfectus'. Hereafter, there is an exposition dedicated to
 directional and cyclic arcs.
@@
 
-\begin{figure}[h!]
+\begin{figure}
     \centering

### 2013-10-07 article/article.tex
feito ateh arcos ciclicos

@@
 
-The arcs can be decomposed in two convergent sequences: the first arc reaches the
-apex and the second arc returns from the apex to the start region. This apex is called climax by traditional music theory. It is
+The arcs can be decomposed in two convergent sequences: the first reaches the
+apex and the second returns from apex to start region.
+This apex is called climax by traditional music theory. It is
 possible to differ between arcs whose climax are placed at the beginning, middle, end,
 and at the first and second half of the considered duration. These structures are
-shown in Figure~\ref{fig:climax}. The parameter of variation does not exist, thus
-the arc consists only of the reference structure~\cite{Schoenberg}.
+shown in figure~\ref{fig:climax}. The varying parameter can be unexistent, a case in which
+the arc consists only of a reference structure~\cite{Schoenberg}.
 
@@
         \caption{Canonical distinctions of musical climax in a given melody and
-        other domains. The considered possibilities are: climax at beginning,
+        other domains. Considered possibilities are: climax at beginning,
         climax at first half, climax at middle, climax at second half and climax
-        at the end. The y-axis is not properly specified since there exist no
-        real parametric variation and, therefore, the structure is a reference.}
+        at the end. The x and y-axis is not properly specified since the parameters
+        can be non-existent, as in a reference structure.}
         \label{fig:climax}
@@
 
-Consider $S_i=\{s_i\}_0^{H-1}$ as an increasing sequence. The sequence
-$R_i=\{r_i\}_0^{2H -2}=\left\{s_{(H-1-|H-1-i|)}\right\}_0^{2H-2}$ is a sequence
-that presents a perfect profiteer symmetry, i.e.\ the second half is the
-mirrored version of the former. Following the musical concepts, the climax is
-localized in the middle of the sequence. It is possible to modify this
-fact using sequences with different sizes. All the mathematical theory of
+Consider the $S_i=\{s_i\}_0^{H-1}$ increasing sequence. The sequence
+$R_i=\{r_i\}_0^{2H -2}=\left\{s_{(H-1-|H-1-i|)}\right\}_0^{2H-2}$
+presents perfect specular symmetry, i.e.\ the second half is the
+mirrored version of the first. In musical terms, the climax is
+in the middle of the sequence. It is possible to modify this
+by using sequences with different sizes. All the mathematics of
 sequences, already well established and taught routinely in calculus courses, can be
-used to generate those arcs~\cite{Guidorizzo,Schoenberg}. 
-Theoretically, when applied to any characteristic of music events, these sequences produce arcs, since they imply the deviation and return of an initial
-parametrization. Henceforth, it is possible to have a
-number of distinct arcs, with different sizes and climax for a same sequence of events. This is an interesting and useful resource, and the correlation of arcs results in coherence~\cite{Salzer}.
-
-Historically (and practically nowadays), the golden ratio and the Fibonacci sequence have special importance. The Lucas sequence allows the generalization of
-Fibonacci sequence, making it easy to understand. Given any two numbers $x_0$
+used to generate these arcs~\cite{Guidorizzo,Schoenberg}. 
+Theoretically, when applied to any characteristic of musical events, 
+these sequences produce arcs, since they imply a deviation and return of an initial
+parametrization.
+Henceforth, it is possible for an only sequence to have
+numberous distinct arcs, with different sizes and climax. 
+This is an interesting and useful resource, and the correlation of arcs yields coherence~\cite{Salzer}.
+
+In practice, and historically, there is special incidence and use of the golden ratio.
+The Lucas sequence allows the generalization of
+Fibonacci sequence, making it easy to follow. Given any two numbers $x_0$
 and $x_1$, the Lucas sequence can be obtained as: $x_n=x_{n-1}+x_{n-2}$. The greater $n$ is, the more $\frac{x_{n}}{x_{n_1}}$ approaches the golden ratio:
@@
 initial values. Being $x_0=1$ and $x_1=100$, and $y_n=\frac{x_n}{x_{n+1}}$ an
-auxiliary sequence, the error occurring in the first values with
-relation to the golden ratio is, approximately, $\{ e_n \}
+auxiliary sequence, the error for the fraction of first values with
+respect to the golden ratio is, approximately, $\{ e_n \}
 =\left\{100\frac{y_n}{1.61803398875}-100 \right\}_1^{10}=\{6080.33, -37.57, 23,
 -7.14, 2.937, -1.09, 0.42, -0.1601, \\ 0.06125, -0.02338\}$. The Fibonacci sequence
-presents about the same error progression but starts right at the second step since $\frac{1}{1}\approx\frac{101}{100}$.
+presents exactly the same error progression, but starts at the second step of the most discrepant
+case ($\frac{1}{1}\approx\frac{100+1 = 101}{100}$).
 
 The music piece \emph{Dirracional} exposes the use of arcs into directional structures.
-Its source code is included in Appendix~\ref{ap:dirracional} and available online as part
-of \massa~\cite{MASSA}.
+Its source code is available online as part
+of \massa\ toolbox~\cite{MASSA}.
 

### 2013-10-07 article/article.tex
feito ateh idioma musical

@@
 
-The philosophical understanding that the human thought is based on the perception of
-similarities, differences, stimuli and objects, places the symmetries
-at the core of the cognitive process~\cite{Deleuze}. Mathematically, symmetries
-are algebraic finite groups, and a finite group is always isomorphic to a permutation
-group. It is possible to say that permutations represent any symmetry in a
-finite system (precisa de uma referência aqui).
-In music, permutations are ubiquitous and considerably present in techniques, which confirms their central role.
+The philosophical understanding that human thought is founded on the concept of
+similarities and differences (e.g. perceived in stimuli and objects), places symmetries
+at the core of cognitive processes~\cite{Deleuze}. Mathematically, symmetries
+are algebraic groups, and a finite group is always isomorphic to a permutation
+group (by Cayley's theorem). 
+In a way, this states that permutations represent any symmetry in a
+finite system~\cite{grupos,gruposFascination}.
+In music, permutations are ubiquitous and present in scholastic techniques,
+ which confirms their central role. 
 The successive application of permutations generates cyclic arcs~\cite{change,Zamacois,permMusic}. 
-Two academic studies were dedicated to this end, aiming at generating music structures~\cite{figgusOriginal, figgusEspacializacao}.
+Two academic studies were dedicated to generating musical structures~\cite{figgusOriginal, figgusEspacializacao}.
 
@@
 
-\begin{multline}\label{eq:groups}
-%\begin{split}
-\forall \;\; p_1,p_2 \in G \Rightarrow\quad\;\; p_1 \bullet p_2  =
-p_3 \in G \\ \;\;\;\text{(closure property)} \\
-\forall \;\; p_1,p_2,p_3 \in G \Rightarrow\quad (p_1\bullet p_2)\bullet p_3  =
-p_1\bullet (p_2\bullet p_3)\quad\;  \\ \text{(associativity property)} \\
-\exists \;\; e \in G :\; p \bullet e  = e \bullet
-p \;\;\;\; \forall p \in G  \quad \\ \text{(identity element property)} \\
-\forall \;\; p \in G, \;\exists\; p^{-1} :\quad\;  p\bullet p^{-1} 
-=p^{-1}\bullet p = e \\ \; \text{(inverse element property)}
-%\end{split}
-\end{multline}
-
-From the first property it is possible to conclude that any permutation can be
-operated with another permutation. In fact, it is possible to apply a
+\begin{equation}\label{eq:groups}
+\begin{split}
+\forall \;\; p_1,p_2 \in G \Rightarrow  \quad   & p_1 \bullet p_2  = p_3 \in G \\ 
+     & \text{(closure property)} \\
+\forall \;\; p_1,p_2,p_3 \in G \Rightarrow \quad & (p_1\bullet p_2)\bullet p_3  = p_1\bullet (p_2\bullet p_3) \\
+     & \text{(associativity property)} \\
+\exists \;\; e \in G :                  \quad    & p \bullet e  = e \bullet p \;\;\;\; \forall p \in G  \\ 
+     &  \text{(identity element existence)} \\
+\forall \;\; p \in G, \;\exists\; p^{-1} :\quad  &  p\bullet p^{-1}i =p^{-1}\bullet p = e \\
+     &  \text{(inverse element existence)}
+\end{split}
+\end{equation}
+
+From the first property follows that two permutations act as one permutation.
+In fact, it is possible to apply a
 permutation $p_1$ and another permutation $p_2$, and, comparing both initial and final
-orderings, results in another permutation $p_3$.
+orderings, observe another permutation $p_3$.
 
@@
 the identity element $p^n=e$ (otherwise the group would be infinite, generated
-by $p$). The lower $n\;:\;p^n=e$ is called the element order. Thus, a finite
+by $p$). The lower $n\;:\;p^n=e$ is the element order. Thus, a finite
 permutation $p$, successively applied, reaches the initial ordering of its
-elements, making a cycle. This cycle, if used as parameter of music notes,
-implies in a cyclic musical arc.
+elements, and makes a cycle. This cycle, if used for parameters of musical notes or structures,
+yields a cyclic musical arc.
 
 These arcs can be established by using a set of different permutations. As a historic
-example, the tradition called \emph{change ringing} conceives the music through
+example, the \emph{change ringing} tradition conceives music through
 bells played one after other and then played again, but in a different
@@
 different traversed orderings is a \emph{peal}. Table~\ref{tab:change}
-represents one traditional \emph{peal} for 3 bells (1, 2 and 3) which explores
+presents a traditional \emph{peal}, named ``Hunt'', for 3 bells (\textcolor{red}{1}, \textcolor{blue}{2} and \textcolor{green}{3}), which explores
 all possible orderings. Each line indicates one bell ordering to be
-played. Permutations occur between each line. In this case, the music structure
-consists of permutations by itself and some different permutations operate due to
-the cyclic behavior.
+played. Permutations occur between each line. In this case, music structure
+consists of permutations itself and some different permutations operate in the
+cyclic behavior.
 
@@
 \centering
-\caption{Change Ringing: Standard \emph{peal} for 3 bells. Permutations
-occur between each line. Each line is a bell ordering and each ordering is played
-        a line a time.} 
+\caption{Change Ringing: Traditional \emph{peal} for 3 bells. Permutations
+occur between each line. Each line is a bell ordering and each ordering is played at a time.} 
 \begin{tabular}{l c r}
@@
 The use of permutations in music can be summarized in the following way:
-consider $S_i=\{s_i\}$ as a sequence of musical events $s_i$ (e.g.\ notes), and a
+with $S_i=\{s_i\}$ a sequence of musical events $s_i$ (e.g.\ notes), and a
 permutation $p$. $S_i'=p(S_i)$ comprises the same elements of $S_i$ but in a
-different order. The permutations can be written based on two notations: cyclic or
+different order. Permutations has two notations: cyclic and
 natural. The natural notation basically comprehends the order of the resulting indexes from
@@
 indexes $[0\;1\;2\;3\;4\;5\;...]$, the permutation is noted by the sequence it
-produces (ex. $[1\;3\;7\;0\;...]$). In the cyclic notation, permutation is the
-changing of one element by the posterior one, and the last element by the first one.
-
-It is not necessary to permute elements of $S_i$, but only some
-characteristics. In this way, being $p^f$ a permutation in frequency and $S_i$ a
+produces (ex. $[1\;3\;7\;0\;...]$). In the cyclic notation, a permutation consists
+of swaps by elements and its posteriors, and the last element by the first one.
+E.g. $(1,2,5)(3,4)$ in cyclic notation is equivalent to $[0,2,5,4,3,1]$ in natural notation.
+
+In the auralization of a permutation, it is not necessary to permute elements of $S_i$,
+but only some characteristic. In this way, being $p$ a permutation and $S_i$ a
 sequence of basic notes as exposed in the end of section~\ref{notaBasica}, the
-new sequence $S_i'=p^f(S_i)=\left\{s_i^{p(f)}\right\}$ consists of the same
+sequence $S_i'=p^f(S_i)=\left\{s_i^{p(f)}\right\}$ consists of the same
 music notes, respecting same order and maintaining the same characteristics, but with the
-fundamental frequencies permuted following the pattern that $p^f$ presents.
-
-It is worthwhile to mention two subtleties of this procedure.
-First, the permutation $p$ does not have to involve all elements of $S_i$, i.e.\ it can operate in subsets of $S_i$.
-Second,  the elements $s_i$ do not have to be entirely executed at each state access.
-To exemplify,
-consider $S_i$ as a sequence of music notes $s_i$. If $i$ goes from $0$ to $n$, and
+fundamental frequencies permuted following the pattern that $p$ presents.
+
+Two subtleties of this procedure. 
+First, a permutation $p$ is not restricted to involve all elements of $S_i$,
+i.e.\ it can operate in a subset of $S_i$.
+Second,  not all elements $s_i$ need to be executed at each access to $S_i$.
+To exemplify, be $S_i$ a sequence of music notes $s_i$. 
+If $i$ goes from $0$ to $n$, and
 $n>4$, at each measure of $4$ notes it is possible to execute the first $4$
-notes. The other notes of $S_i$ can occur in measures where permutations set
-aside those notes to the first four notes of $S_i$.
-
-Each one of the permutations $p_i$ described above relates the note dimensions where it operates (frequency, duration, \emph{fades},
-intensity, etc) and the period of incidence (at how much access a permutation is
+notes. The other notes of $S_i$ can occur in measures where permutations 
+allocate those notes to the first four events $s'_i$ of $S'_i$.
+
+It is inalienable to each permutations $p_i$ described above:
+1) note dimensions where it operates (frequency, duration, \emph{fades},
+intensity, etc); and
+2) the period of incidence (how much consults before a permutation is
 applied). During realization of notes in $S_i$, an easy and coherent form is to
-execute the first $n$ notes\footnote{The execution of disjoint notes of $S_i$
-equals to modify the permutation and to execute the first notes.}.
-
-Appendix~\ref{cap:FIGGUScode} and \massa present the computational implementation about permutations~\cite{MASSA,figgusOriginal,figgusEspacializacao}.
+execute the first $n$ notes, the execution of disjoint sets of $S_i$
+equals modifying the permutation and executing the first $n$ notes.
+
+The \massa\ toolbox presents a computational implementation that isolates
+permutation aplication to sonic characteristics, in order to deliver
+musical structures~\cite{MASSA,figgusOriginal,figgusEspacializacao}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2013-10-07 article/article.tex
feita ateh conclusao

@@
 
-There are many studies intending to model and explore a 'musical language',
-or 'linguistics applied to music' or even to discern between different 'musical
+In numerous studies, there are models and explorations of a `musical language',
+many can somewhat be considered `linguistics applied to music'
+and some even discern different `musical
 idioms'~\cite{Lerdahl, Harmonia, Salzer,Alfaix}. In a simple way, a musical idiom
-is the result of a chosen material together with repetition of elements and
-repetition of relations between elements present along the music.
-The related dichotomies are outstanding, as explained at subsection~\ref{subsec:motivos}: repetition and variation, relax and tension, equilibrium and instability, consonance and dissonance, etc.
+is the result of chosen material together with repetition of elements and
+of relations between elements, along a music piece. In these matters, 
+dichotomies are prominent, as explained in subsection~\ref{subsec:motivos}:
+repetition and variation, relaxation and tension, equilibrium and instability, consonance and dissonance, etc.
 
@@
 
-The basic note was defined and characterized in quantitative terms (section~\ref{sec:notaDisc}). Next, the internal note
-composition was addressed and both internal transitions and immediate treatment
+First, the basic note was defined and characterized in quantitative terms (section~\ref{sec:notaDisc}).
+Next, internal note
+composition was addressed and both internal transitions and immediate sonic treatment
 were understood (section~\ref{varInternas}). Finally, this section aims at
 organizing these notes in music. The gamma of resources and consequent infinitude
-of possibilities is a typical situation and highly relevant for art contexts~\cite{Harmonia,Webern}.
-
-There are studies dedicated to each one of the presented resource. For example, it is possible to obtain the 'dirty' triadic harmonies (with notes without the triad) by
-superpositions of dirty fourths. Another interesting example is the simultaneous
+of praxis possibilities is a typical and highly relevant situation for artistic contexts~\cite{Harmonia,Webern}.
+
+There are studies for each of the presented resources.
+For example, it is possible to obtain `dirty' triadic harmonies (with notes out of the triad) by
+superposition of perfect fourths. 
+Another interesting example is the simultaneous
 presence of rhythms in different metrics, constituting what is
-called \emph{polyrhythms}. The music piece \emph{Poli-hit mia} explores these
-simultaneous metrics by impulse train (??) convolved with notes that compound the
-line. Its source code is included in Appendix~\ref{ap:poli} and available online as part
-of \massa.
-
-The microtonal scales are important for the music of the 20th
-century~\cite{microtonalidade} and present diverse outstanding results, like the
+called \emph{polyrhythm}. The music piece \emph{Poli-hit mia} explores these
+simultaneous metrics by impulse trains convolved with notes.
+Its source code is online as part of \massa\ toolbox~\cite{massa}.
+
+Microtonal scales are important for 20th
+century music~\cite{microtonalidade} and present diverse remarkable results, e.g.
 fourths of tone in the Indian music. The musical sequence \emph{MicroTom}
 explores these resources, including microtonal melodies and microtonal harmonies
-with many notes in a very reduced note scope. Its code is presented in
-Appendix~\ref{ap:micro} and available online in \massa.
-
-As pointed in subsection~\ref{subsec:mus2}, the relations between
-parameters are powerful ways to acquire music pieces. The number of permuted
+with many notes in a very reduced note scope. Its code is 
+available online in \massa\ toolbox~\cite{massa}.
+
+As in subsection~\ref{subsec:mus2}, relations between
+parameters are powerful ways to acquire musical pieces.
+The number of permuted
 notes can vary during the music, revealing relationship with piece
-duration. Harmonies can be made of triads (eqs.~\ref{triades}) with replicated
+duration. Harmonies can be made from triads (eqs.~\ref{triades}) with replicated
 notes at each octaves and more numerous as minor the depth and frequency of
@@
 
-The presented symmetries at octave divisions (eqs.~\ref{escSim}) and the
-symmetries presented as permutations (Table~\ref{tab:change} and
+The symmetries at octave divisions (eqs.~\ref{escSim}) and the
+symmetries presented as permutations (table~\ref{tab:change} and
 eqs.~\ref{eq:groups}) can be used together. In the music piece \emph{3 trios}
-this association is done in a systematic way in order to enable the required
-listening. This is a instrumental piece, not included as a source
-code~\cite{3Trios}.
-
-The \emph{PPEPPS} (Pure Python EP: Projeto Solvente) is an EP synthesized using
-the resources presented in this study. With minimal parametrization, the program
-generates complete musics, allowing the easy composition of musics and sets of
-musics. Using few lines of code it is possible to obtain
-a directory with the generated musics. This facility and technological
-demystification create aesthetically possibilities for both sharing and education.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+this association is done in a systematic way in order to enable a dedicated
+listening. This is an instrumental piece, not included as a source
+code but available online~\cite{3Trios}.
+
+\emph{PPEPPS} (Pure Python EP: Projeto Solvente) is an EP synthesized using
+resources presented in this study. With minimal parametrization, the scripts
+generate complete musici pieces, allowing easy composition of sets of
+musics. A simple script of a few lines specifies music delivered as 16 bit
+44.1kHz PCM files (WAVE). This facility and technological
+demystification create aesthetical possibilities for both sharing and education.
 

### 2013-10-07 article/article.tex
terminada revisao do texto do artigo

@@
 
-\section{Conclusions and future works} %Nome do capítulo.
+\section{Conclusions and further developments}
 \label{cap:conclusao}
 
-We have described a concise system that explores the musical elements and relate them to the digital signal.
+This concise presentation relates basic musical elements to digital sound.
 The reader is invited to access \emph{Scripts} and \massa, where these relations are computationally implemented.
-We hope the didactic report along the paper and the supplied scripts facilitates and encourages the use of the proposed framework.
-
-The open possibilities provided by the techniques and results discussed in this paper involve psychoacoustic experiments and the creation of interfaces for the generation of music, noise and other music applications in a high fidelity (\emph{hi-fi}).
-It is worthwhile to mention the benefit of these results for artistic and didactic purposes.
-The incorporation of programming skills is facilitated by the provided visual aids.
-Initial \emph{live-coding} practices and guided courses based on this work have already been realized with successful.
-Examples are Puredata and Chuck.
+The didactic report along the paper and the supplied scripts eases and encourages the use of the proposed framework.
+
+The possibilities provided by these results include psychoacoustic experiments, 
+interfaces for the generation of noise and other high fidelity sounds and musical structures,
+and for artistic and didactic purposes. The incorporation of programming skills is facilitated by visual aids, which
+was explored by \emph{live-coding} practices and courses based on this framework, with success.
 
 This work systematically investigates the parameterization issues (like the tremolo, ADSR, etc.) in a high fidelity, which has significant artistic utility.
-Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature (as shown in Appendix~\ref{cap: trabalhosRelacionados}).
-
-Besides, the free software license and online availability of the exposed content as hypertext, in conjunction with the respective codes and sound examples, strongly facilitates future collaborations and the generation of sub-products in a co-authorship fashion.
-As consequence, the expansion of \massa is feasible and straightforward, making it possible new developments and implementations of musical context.
-
-In addition, this work permitted the formation of groups with common interests, like music creativity and computer music.
-Specially, the project \url{labMacambira.sf.net} groups Brazilian and foreign co-workers in order to offer relevant contributions in diverse areas like Digital Direct Democracy (não devo ter traduzido certo), georeferencing techniques, artistic and educational activities.
-Some of these reports are available in Appendix~\ref{cap:musicaExtra} and made online.
-There are more than 700 videos, written documentations, original softwares and contributions in well-known external softwares such as Firefox, Scilab, LibreOffice, GEM/Puredata, to name a few~\cite{siteLM,wikiLM,vimeoLM}.
-
-Future works include the application of the reported results in machine learning and artificial intelligence methods for the generation of appealing artistic materials.
+Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, such as reviewed in ~\ref{dissertacao}.
+
+The free software license and online availability of the exposed content as hypertext, 
+with the respective codes and sonic examples, strongly facilitates future collaborations and the generation of sub-products in a co-authorship fashion.
+As consequence, the expansion of \massa\ is favoured and straightforward, easing new implementarions and musical pieces development.
+
+In addition, this framework permitted the formation of interests groups, with topics such as music creativity and computer music. 
+Specially, the project \url{labMacambira.sf.net} groups Brazilian and foreign co-workers in order to offer relevant contributions 
+in diverse areas like Digital Direct Democracy, georeferencing techniques, artistic and educational activities. 
+Some of these reports are available online~\cite{dissertacao}. There are more than 700 videos, written documentats,
+ original software applications and contributions in well-known external softwares such as Firefox, Scilab, LibreOffice,
+ GEM/Puredata, to name a few~\cite{siteLM,wikiLM,vimeoLM}.
+
+Future work includes application of these results in machine learning and artificial intelligence 
+methods for the generation of appealing artistic materials. Some psychoacoustical effects were detected,
+which needs validation and should be reported, specially with~\cite{quadrosSonoros}. Other foreseen advances
+are: JavaScript version of \massa\ toolbox, better hypermedia deliverables of this framework, use guides
+for different ends (musical composition, psychophysical experiments, sound synthesis, education, etc), creation
+of musical pieces and open experiments to be studied with EEG recordings and further analitical specification of musical
+elements in the discrete-time representation of sound as community feedback corroborate. 
+.
 

### 2013-10-07 article/article.tex
revisitado abstract

@@
 
-Notes, ornaments and intervals are examples of basic elements of music,  their representation as discrete-time digital audio signal plays a central role in software for music creation and design.
-Nevertheless, there is no systematic relation, in analytical terms, of these musical elements to the sound samples.
-Such a compendium is important since it enables scientific experiments in precise and trustful ways.
-In this paper, a comprehensive description of the musical elements within a unified approach is presented.
+Notes, ornaments and intervals are examples of basic musical elements.
+Their representation as discrete-time digital audio plays a central role in software for music creation and design.
+Nevertheless, there is no systematic relation, in analytical terms, of these musical elements to the sonic samples.
+Such a compendium enables scientific experiments in precise and trustful ways, among educational and artistic uses.
+This paper presents a comprehensive description of musical elements, within an unified approach.
 Musical elements, like pitch, duration and timbre are expressed by equations on sample level.
 This quantitatively relates characteristics of the discrete-time signal to musical qualities.
-Internal variations, e.g., tremolos, vibratos and spectral fluctuations, are also considered as means to achieve variation within a note.
-Moreover, the generation of musical structures such as rhythmic meter, pitch intervals and cycles, are reached canonically with notes.
-The availability of these resources in scripts  (\massa\ - Music and Audio in Sequences and Samples) is also provided in public domain.
+Internal variations, e.g. tremolos, vibratos and spectral fluctuations, are also considered
+to operate within a note. Moreover, the generation of musical structures such as rhythmic meter,
+pitch intervals and cycles, are attained canonically with notes. 
+The availability of these resources in scripts is provided in public domain within the \massa\ toolbox - Music and Audio in Sequences and Samples.
 Authors observe that the implementation of sample-domain analytical results as open source can encourage concise research.
-As further illustrated in the paper, \massa\ has already been employed by users for diverse purposes, including acoustics experimentation, art and education.
-The efficacy of these physical descriptions of basic musical elements was confirmed by the synthesis of small musical pieces.
+As further illustrated in the paper, \massa\ has already been employed by users for diverse purposes, 
+including acoustics experimentation, art and education. The efficacy of these physical descriptions 
+was confirmed by the synthesis of small musical pieces.
 As shown, it is possible to synthesize whole albums through collage of scripts and parametrization.

### 2013-10-07 article/article.tex
integrado bibliografia da dissertacao

@@
 On the last decade of the century, it became evident that popular music had also incorporated sounds without defined pitch and temporal organization, to name a few of simple characteristics.}.
-Notes are also convenient for another reason: the average listener -- and considerable part of the specialists -- presupposes rhythmic and pitch organization (made explicit in section~\ref{sec:notasMusic}) as fundamental musical properties, and these are developed in traditional musical theory in terms of notes.
+    Notes are also convenient for another reason: the average listener -- and considerable part of the specialists -- presupposes rhythmic and pitch organization (made explicit in section~\ref{sec:notesMusic}) as fundamental musical properties, and these are developed in traditional musical theory in terms of notes.
 
@@
 \nocite{*}
-\bibliography{article}
+\bibliography{article2}
 \end{document}

### 2013-10-07 article/article.tex
terminada repassada das figuras e bibliografia

@@
 
-\section{Characterization of the discrete-time musical note} \label{sec:discNote}
+\section{Characterization of the discrete-time musical note} \label{sec:discNote}\label{sec:notaDisc}
 
@@
 
-and the amplitude gain for a sequence whose volume has been doubled ($10dB$):
+\noindent and the amplitude gain for a sequence whose volume has been doubled ($10dB$):
 
@@
 
-where $A$ is the multiplicative factor that relates the amplitudes before and after the amplification.
+\noindent where $A$ is the multiplicative factor that relates the amplitudes before and after the amplification.
 
@@
 It requires a big amplitude to give sinusoid volume, if compared to other waveforms.
-Both particularities are seen as a consequence of the non-existence of pure sinusoidal sounds in nature~\ref{roederer}.
+Both particularities are seen as a consequence of the non-existence of pure sinusoidal sounds in nature~\cite{Roederer}.
 
@@
 
-Figure~\ref{fig:formasDeOnda} presents the waveforms described in equations  ~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} and ~\ref{quadrada} for $\lambda_f=100$ (period of $100$ samples).
+Figure~\ref{fig:formasDeOnda} presents the waveforms described in equations  ~\ref{sinusoid}, ~\ref{sawTooth}, ~\ref{triangular} and ~\ref{square} for $\lambda_f=100$ (period of $100$ samples).
 If $f_s=44.1kHz$, the PCM standard in Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz$, around A4, just above the central "C", whatever the waveform is.
@@
     \centering
-        \includegraphics[width=\columnwidth]{figures/waveSpectrum}
+        \includegraphics[width=\textwidth]{figures/waveSpectrum}
     \caption{Spectrum of basic artificial musical waveforms.}
@@
     \centering
-        \includegraphics[width=\columnwidth]{figures/oboeNaturalSampledSpectrum}
+        \includegraphics[width=\textwidth]{figures/oboeNaturalSampledSpectrum}
     \caption{Spectrum of the sonic waves of a natural oboe note and from a sampled period.
@@
 
-Figure~\ref{fig:formas4} shows the harmonics for the basic waveforms of equations~\ref{senoide},~\ref{denteDeSerra},~\ref{triangular} and~\ref{quadrada} for the case of 4 samples.
+Figure~\ref{fig:formas4} shows the harmonics for the basic waveforms of equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} for the case of 4 samples.
 There is only 1 sinusoid for each waveform, with the exception of the sawtooth, which has even harmonics.
@@
 An unique period from the basic waveforms can be used to specify the note, where $\lambda_f=\frac{f_a}{f}$ is the number of samples at the period.
-Here, $L_i^{f,\, \delta_f} $ is the sequence that describes a period of the waveform $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ with duration $\delta_f=1/f$ (as given by equations~\ref{senoide}, ~\ref{denteDeSerra}, ~\ref{triangular} and ~\ref{quadrada}) and $R_i^f$ is a sampled real waveform:
+Here, $L_i^{f,\, \delta_f} $ is the sequence that describes a period of the waveform $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ with duration $\delta_f=1/f$ (as given by equations~\ref{sinusoid}, ~\ref{sawTooth}, ~\ref{triangular} and ~\ref{square}) and $R_i^f$ is a sampled real waveform:
 
@@
 
-\section{Variation in the basic musical note}\label{sec:internalVar}
+\section{Variation in the basic musical note}\label{sec:internalVar}\label{sec:varInternas}
 
@@
 figure~\ref{fig:pulsoSubAgl}. This piece is a link to next section contents. 
-The source code of the music piece is online, as part of the \massa toolbox.
+The source code of the music piece is online, as part of the \massa\ toolbox\cite{MASSA}.
 
@@
 effect as describe above, and also considering the intersection between audio source and receptor,
-at the \massa\ toolbox~\cite{massa}.
+at the \massa\ toolbox~\cite{MASSA}.
 
@@
 Besides that, modifications in the reverberation space consists on a common resource (almost a \textit{clich\'{e}}) to surprise and attract the listener.
-An implementation of the reverb recipe described is on the \massa\ toolbox~\cite{massa}
+An implementation of the reverb recipe described is on the \massa\ toolbox~\cite{MASSA}
 
@@
 simultaneous metrics by impulse trains convolved with notes.
-Its source code is online as part of \massa\ toolbox~\cite{massa}.
+Its source code is online as part of \massa\ toolbox~\cite{MASSA}.
 
@@
 with many notes in a very reduced note scope. Its code is 
-available online in \massa\ toolbox~\cite{massa}.
+available online in \massa\ toolbox~\cite{MASSA}.
 
@@
 This work systematically investigates the parameterization issues (like the tremolo, ADSR, etc.) in a high fidelity, which has significant artistic utility.
-Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, such as reviewed in ~\ref{dissertacao}.
+Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, such as reviewed in ~\cite{dissertacao}.
 

### 2014-12-21 article/article.tex
atualizando artigo, revisao do resumo
incorporadas correcoes do antonio

@@
 \usepackage{float}
+\usepackage{hyperref}
 
@@
 
-\author{Luciano da Fontoura Costa}
-  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
-  \email{ldfcosta@gmail.com}
- \altaffiliation[Also at ]{IFSC-USP}
-
- \author{Osvaldo N. Oliveira Jr.}
-  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
-  \email{chu@ifsc.usp.br}
- \altaffiliation[Also at ]{IFSC-USP}
+%\author{Luciano da Fontoura Costa}
+%  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
+%  \email{ldfcosta@gmail.com}
+% \altaffiliation[Also at ]{IFSC-USP}
+%
+% \author{Osvaldo N. Oliveira Jr.}
+%  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
+%  \email{chu@ifsc.usp.br}
+% \altaffiliation[Also at ]{IFSC-USP}
 
@@
 Such a compendium enables scientific experiments in precise and trustful ways, among educational and artistic uses.
-This paper presents a comprehensive description of musical elements, within an unified approach.
+This paper presents a comprehensive description of music in digital audio, within an unified approach.
 Musical elements, like pitch, duration and timbre are expressed by equations on sample level.
@@
 
-Human perception of sound involves captivation by bones, stomach, ears, transfer functions of head and torso, and processing by the nervous system~\cite{}.
+Human perception of sound involves captivation by bones, stomach, ears, transfer functions of head and torso, and processing by the nervous system~\cite{Roederer}.
 Besides that, the ear is a dedicated organ for the appreciation of these waves.
 Its mechanism decomposes sound into its sinusoidal spectrum and delivers to the nervous system.
-The sinusoidal components are crucial to musical phenomena, as one can perceive in the composition of sounds of musical interest (such as harmonic sounds and noises, seen in sections~\ref{sec:discNote} and~\ref{sec:internalVar}), and sonic structures of musical interest (such as tunings and scales, in section~\ref{sec:notesMusic}).
+The sinusoidal components are crucial to musical phenomena, as one can perceive in the composition of sounds of musical interest (such as harmonic sounds and noises, exposed in sections~\ref{sec:discNote} and~\ref{sec:internalVar}), and sonic structures of musical interest (such as tunings and scales, in section~\ref{sec:notesMusic}).
 
@@
 This occurred in the concert music, specially in the concrete, electronic and electroacoustic styles.
-On the last decade of the century, it became evident that popular music had also incorporated sounds without defined pitch and temporal organization, to name a few of simple characteristics.}.
+On the last decade of the twentieth century, it became evident that popular music had also incorporated sounds without defined pitch and temporal organization, to name a few of simple characteristics.}.
     Notes are also convenient for another reason: the average listener -- and considerable part of the specialists -- presupposes rhythmic and pitch organization (made explicit in section~\ref{sec:notesMusic}) as fundamental musical properties, and these are developed in traditional musical theory in terms of notes.
@@
 
-The remaining parts of this work are organized as follows: section~\ref{sec:discNote} characterizes the basic musical note; internals of the musical note is further developed in section~\ref{sec:internalVar}; section~\ref{sec:notesMusic} tackles the organization of musical notes in higher levels of musical structure~\cite{Wisnick,Webern,Lerdhal,Cook,Lacerda}.
+The remaining parts of this work are organized as follows: section~\ref{sec:discNote} characterizes the basic musical note; section~\ref{sec:internalVar} further develops internals of the musical note; section~\ref{sec:notesMusic} tackles the organization of musical notes in higher levels of musical structure~\cite{Wisnick,Webern,Lerdhal,Cook,Lacerda}.
 As these descriptions embody topics such as psycho-acoustics, cultural traditions, formalisms and protocols, the text points to external complements as needed~\cite{Zamacois,Schoenberg,microsound}.
@@
 Physically, the timbre is multidimensional and given by the temporal dynamics of energy in the spectral components that are harmonic or noisy.
-Beyond that, the word timbre is used to designate different things: one same note has different timbres, a same instrument has different timbres, two instruments of the same family have, at the same time, the same timbre that blends them in the same family, and different timbres as they are different instruments.
+Beyond that, the word timbre is used to designate different things: one same note can have (be produced with) different timbres, an instrument has different timbres, two instruments of the same family have, at the same time, the same timbre that blends them in the same family, and different timbres as they are different instruments.
 It is worth to mention that timbre is not always manifested in spectral traces, since cultural or circumstantial aspects alter our perception of timbre}.
-Musically, it matters that sonic spectra with minimum differences can result in timbres with crucial expressive differences and, consequently, different timbres can be produced by using different spectra\cite{Roederer}.
+Sonic spectra with minimum differences can result in timbres with crucial expressive differences and, consequently, different timbres can be produced using different spectra\cite{Roederer}.
 
@@
 While the sinusoid is an isolated point in the spectrum, these waves present a succession of harmonic components.
-These waveforms are specified in equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} and illustrated in Figure~\ref{fig:formasDeOnda}.
+These waveforms are specified in equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square}, illustrated in Figure~\ref{fig:formasDeOnda}.
 These artificial waveforms are traditionally used in music for synthesis and oscillatory control of variables. They are also useful outside musical contexts\cite{Openheim}.
@@
 The isolated and exactly harmonic components of the spectrum is a consequence of the fixed period usage.
-The sinusoid consists of an one and only node in the spectrum, pure frequency.
+The sinusoid consists of an unique node in the spectrum, pure frequency.
 The figure exibit the spectra described: the sawtooth is the only waveform with a complete harmonic series (odd and even components); triangular and square waves has the same components (odd harmonics), decaying at $-12dB/octave$ and $-6dB/octave$, respectively.
@@
 
-In the case of a fixed length period and waveform, the spectrum is perfectly harmonic and static and  each waveform is compound of specific proportions of harmonic components.
+In the case of a fixed length period and waveform, the spectrum is perfectly harmonic and static, and each waveform is compound of specific proportions of harmonic components.
 High  curvatures are sign of high harmonics in the wave.
@@
 As a consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise and with variations typical of the natural phenomenon.
-This can be observed in Figure~\ref{fig:espectroOboe}, that shows the spectrum of the original oboe note and a note with same duration and whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}.
+This can be observed in Figure~\ref{fig:espectroOboe}, that shows the spectrum of the original oboe note and a note with same duration, whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}.
 Summing up, the natural spectrum exhibits variations in the frequencies of the harmonics, in their intensities and some noise, while the note made from the sampled period has a perfectly harmonic spectrum.
@@
 
- \begin{figure}[h!]
+ \begin{figure*}
      \centering
-         \includegraphics[width=\columnwidth]{figures/amostras2c__}
+         \includegraphics[width=\textwidth]{figures/amostras2c__}
      \caption{Oscillation of 2 samples (maximum frequency for any $f_a$).
@@
          \label{fig:amostras2}
- \end{figure}
+ \end{figure*}
 
@@
 
- \begin{figure}[h!]
+ \begin{figure*}
      \centering
-         \includegraphics[width=\columnwidth]{figures/amostras3b}
+         \includegraphics[width=\textwidth]{figures/amostras3b}
      \caption{Three fixed samples presents only one non-null frequency. $c_1=c_2^*$ and $w_1 \equiv w_2$.}
          \label{fig:amostras3}
- \end{figure}
+ \end{figure*}
 
@@
 
-\begin{figure}[h!]
+\begin{figure*}
     \centering
-        \includegraphics[width=\columnwidth]{figures/amostras6formas__}
+        \includegraphics[width=\textwidth]{figures/amostras6formas__}
     \caption{Basic waveforms with 6 samples: triangular and square waveforms have odd harmonics, with different proportions and phases; the sawtooth has even harmonics.}
         \label{fig:formas6}
-\end{figure}
+\end{figure*}
 
@@
 Although it is not one of its four basic properties, a musical note always has a spatial localization: the note source position at the ordinary physical space.
-The existence of an environment that reverberates the note is the matter of 'spatialization'.
+The reverberation the environment in which sound occurs is the matter of \emph{spatialization}.
 Both fields, spatialization and spatial localization, are widely valued by audiophiles and music industry~\cite{floEsp}.
@@
 For example, a 3 seconds piano note has intensity with abrupt rise at beginning and progressive decay, has spectrum variations with harmonics decaying some others that emerge along time.
-These variations are not mandatory but sound synthesis guidance for musical uses, as it reflects how sounds appear in nature.
+These variations are not mandatory, but sound synthesis guidance for musical uses, as it reflects how sounds appear in nature.
 This is so considered to be true, that there is a rule of thumb: to make a sound that incites interest by itself, do internal variations on it~\cite{Roederer}.
@@
 Table $\widetilde{M}_i$ with length $\widetilde{\Lambda}_M$ and the sequence with indices $\gamma_i'$ make the sequence $t_i'$ which is the oscillation pattern in the frequency while table $\widetilde{L}_i$ with length $\widetilde{\Lambda}$ and the sequence with indices $\gamma_i$ make $t_i$ which is the sound itself.
-Variables $\mu$ and $\nu$ quantify the vibrato intensity: $\mu$ is a direct measure of how many Hertz are involved in the upper limit of the oscillation while $\nu$ is the direct measure of how many semitones (or half steps) are involved in the oscillation ($2\nu$ is the number of semitones between the upper and lower peaks of frequency oscillations of the sound $\{t_i\}$ caused by the vibrato).
+Variables $\mu$ and $\nu$ quantify the vibrato intensity:
+\begin{itemize}
+    \item $\mu$ is a direct measure of how many Hertz are involved in the upper limit of the oscillation, while
+    \item $\nu$ is the direct measure of how many semitones (or half steps) are involved in the oscillation ($2\nu$ is the number of semitones between the upper and lower peaks of frequency oscillations of the sound $\{t_i\}$ caused by the vibrato).
+\end{itemize}
 It is convenient to use $\nu=\log_{2}\frac{f+\mu}{f} $ in this case because the maximum frequency increase is not equivalent to the maximum frequency diminish, but the semitone variation remains.
@@
 are of such difficulty that the dodecafonism arouse. 
-The purpose of dedocafonism is to use a set of notes (ideally 12 notes),
+The purpose of dodecafonism is to use a set of notes (ideally 12 notes),
 and to perform each note, one by one, in the same
@@
 scale is the base for triads, each with its root in a degree: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$. 
-To build triads, the third and the fifth note above the root is considered together with the root (or fundamental).
+To build triads, the third and the fifth notes above the root is considered together with the root (or fundamental).
 $\hat{1},\hat{3},\hat{5}$ is the first degree chord,

### 2014-12-22 article/article.tex
revisada bibliografia do artigo

@@
 
+\author{Antônio Pessoti}
+  \homepage{}
+  \email{debcris.cor@gmail.com}
+ \altaffiliation[Instituto de Ci\^encias Matem\'aticas e de Computa\c{c}\~ao ]{ICMC-USP}
+
 \author{D\'ebora Corr\^ea}
@@
 
-The Nyquist theorem~\cite{} states that the sampling frequency is twice the maximum frequency of the represented signal.
+The Nyquist theorem~\cite{Openheim} states that the sampling frequency is twice the maximum frequency of the represented signal.
 Thus, for general musical purposes, it is necessary to have samples in a rate at least twice the highest frequency heard by humans, that is, $f_s \geq 2\times 20kHz = 40kHz$.
@@
 
-The remaining parts of this work are organized as follows: section~\ref{sec:discNote} characterizes the basic musical note; section~\ref{sec:internalVar} further develops internals of the musical note; section~\ref{sec:notesMusic} tackles the organization of musical notes in higher levels of musical structure~\cite{Wisnick,Webern,Lerdhal,Cook,Lacerda}.
+The remaining parts of this work are organized as follows: section~\ref{sec:discNote} characterizes the basic musical note; section~\ref{sec:internalVar} further develops internals of the musical note; section~\ref{sec:notesMusic} tackles the organization of musical notes in higher levels of musical structure~\cite{Wisnick,Webern,Lerdahl,Cook,Lacerda}.
 As these descriptions embody topics such as psycho-acoustics, cultural traditions, formalisms and protocols, the text points to external complements as needed~\cite{Zamacois,Schoenberg,microsound}.
@@
 In a way, this states that permutations represent any symmetry in a
-finite system~\cite{grupos,gruposFascination}.
+finite system~\cite{gruposFascination}.
 In music, permutations are ubiquitous and present in scholastic techniques,

### 2014-12-22 article/article.tex
arrumada autoria para primeiro upload
retirado o caminho para as figuras do artigo para compatibilizar com o esperado pelo arxiv

@@
 \author{Renato Fabbri}
- \homepage{http://www.ifsc.usp.br/~fabbri}
  \email{renato.fabbri@gmail.com}
@@
 
-\author{Vilson Vieira da Silva Jr.}
+\author{Vilson Vieira da Silva Junior}
   \homepage{http://automata.cc/}
@@
 
-\author{Antônio Pessoti}
-  \homepage{}
+\author{Antônio Carlos Silvano Pessotti}
+  \email{antoniopessotti@gmail.com}
+ \altaffiliation[Instituto de Estudos da Linguagem ]{IEL/UNICAMP}
+
+\author{D\'ebora Cristina Corr\^ea}
   \email{debcris.cor@gmail.com}
- \altaffiliation[Instituto de Ci\^encias Matem\'aticas e de Computa\c{c}\~ao ]{ICMC-USP}
-
-\author{D\'ebora Corr\^ea}
-  \homepage{}
-  \email{debcris.cor@gmail.com}
- \altaffiliation[Instituto de Ci\^encias Matem\'aticas e de Computa\c{c}\~ao ]{ICMC-USP}
+ \altaffiliation[Instituto de Ci\^encias Matem\'aticas e de Computa\c{c}\~ao ]{ICMC/USP}
 
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/pcm}
+        \includegraphics[width=\textwidth]{pcm}
         \caption{Pulse Code Modulation (PCM) audio: an analogical signal is represented by 25 samples with 4 bits each.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/waveForms_}
+        \includegraphics[width=\textwidth]{waveForms_}
     \caption{Basic musical waveforms: (a) the synthetic  waveforms; (b) the real waveforms.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/waveSpectrum}
+        \includegraphics[width=\textwidth]{waveSpectrum}
     \caption{Spectrum of basic artificial musical waveforms.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/oboeNaturalSampledSpectrum}
+        \includegraphics[width=\textwidth]{oboeNaturalSampledSpectrum}
     \caption{Spectrum of the sonic waves of a natural oboe note and from a sampled period.
@@
      \centering
-         \includegraphics[width=\textwidth]{figures/amostras2c__}
+         \includegraphics[width=\textwidth]{amostras2c__}
      \caption{Oscillation of 2 samples (maximum frequency for any $f_a$).
@@
      \centering
-         \includegraphics[width=\textwidth]{figures/amostras3b}
+         \includegraphics[width=\textwidth]{amostras3b}
      \caption{Three fixed samples presents only one non-null frequency. $c_1=c_2^*$ and $w_1 \equiv w_2$.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/amostras4___}
+        \includegraphics[width=\textwidth]{amostras4___}
     \caption{Frequency components for 4 samples.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/amostras4formas__}
+        \includegraphics[width=\textwidth]{amostras4formas__}
     \caption{Basic wave forms with 4 samples.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/amostras6_}
+        \includegraphics[width=\textwidth]{amostras6_}
     \caption{Frequency components for 6 samples: 4 sinusoids, one of them is the \emph{bias} with zero frequency.}
@@
 
-Figure~\ref{fig:amostras6} presents the sinusoidal decomposition for 6 samples and figure~\ref{fig:formas6} presents the decomposition of the basic wave forms.
+Figure~\ref{fig:amostras6} presents the sinusoidal decomposition for 6 samples and Figure~\ref{fig:formas6} presents the decomposition of the basic wave forms.
 In this case, the waveforms have spectra with fundamental differences: square and triangular have the same components but with different proportions, while the sawtooth have an extra component.
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/amostras6formas__}
+        \includegraphics[width=\textwidth]{amostras6formas__}
     \caption{Basic waveforms with 6 samples: triangular and square waveforms have odd harmonics, with different proportions and phases; the sawtooth has even harmonics.}
@@
     \centering
-        \includegraphics[width=.5\textwidth]{figures/espacializacao___}
+        \includegraphics[width=.5\textwidth]{espacializacao___}
     \caption{Detection of sound source localization: schema used to calculate Interaural Time Difference (ITD) and Interaural Intensity Difference (IID).}
@@
 
-Considering only the direct incidences in each ear, the equations are quite simple. An object placed at $(x,y)$, as in figure~\ref{fig:spac}, is distant of each ear by:
+Considering only the direct incidences in each ear, the equations are quite simple. An object placed at $(x,y)$, as in Figure~\ref{fig:spac}, is distant of each ear by:
 
@@
 Figure~\ref{fig:spac} depicts the acoustic shadow of the cranium, an important phenomenon to perception of source azimuthal angle in the cone of confusion.
-The cone itself is not shown in figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not encountered in the literature.
+The cone itself is not shown in Figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not encountered in the literature.
 Given the filtering and diffraction dependent of the sound spectrum, it is hard, if not impossible, to correctly draw the confusion cone.
@@
     {\centering
-        \includegraphics[width=\columnwidth]{figures/mixagem}}
+        \includegraphics[width=\columnwidth]{mixagem}}
     \caption{Mixing of three sound sequences. The amplitudes are directly overlapped.}
@@
 {    \centering
-        \includegraphics[width=\columnwidth]{figures/concatenacao}}
+        \includegraphics[width=\columnwidth]{concatenacao}}
     \caption{Concatenation of three sound sequences by temporal overlap of its samples.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/lut}
+        \includegraphics[width=\textwidth]{lut}
     \caption{Search in the \emph{lookup table} to synthesize sounds in different frequencies using an unique waveform with high resolution.}
@@
      \centering
-         \includegraphics[width=\columnwidth]{figures/transicao}
+         \includegraphics[width=\columnwidth]{transicao}
      \caption{Intensity transitions to different values of $\alpha$ (see equations~\ref{seqAmp} and ~\ref{transAmp}).}
@@
      \centering
-         \includegraphics[width=\textwidth]{figures/convolucao}
+         \includegraphics[width=\textwidth]{convolucao}
      \caption{Graphical interpretation of convolution. Each resulting sample is the sum of the previous samples of a signal, with each one multiplied by the retrograde of the other sequence.}
@@
 Despite computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse for each reincidence of the sound.
-In figure~\ref{fig:delays} it is possible to observe the shift caused by convolution with an impulse.
+In Figure~\ref{fig:delays} it is possible to observe the shift caused by convolution with an impulse.
 Depending on the intensity of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as pitch (from about 20 impulses per second and higher frequencies).
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/delays}
+        \includegraphics[width=\textwidth]{delays}
     \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c). Disposed in increasing order of its pulse density.}
@@
 The text bellow lists a selection of them for didactic purposes and as a reference.
-They are well behaved filters whose aspects are described in figure~\ref{fig:iir}.
+They are well behaved filters whose aspects are described in Figure~\ref{fig:iir}.
 
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/iir}
+        \includegraphics[width=\textwidth]{iir}
     \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of equations~\ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} and \ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
@@
 \begin{enumerate}
-  \item Low-pass with a simple pole, module of the frequency response in the upper left corner of figure~\ref{fig:iir}.
+  \item Low-pass with a simple pole, module of the frequency response in the upper left corner of Figure~\ref{fig:iir}.
 The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_s$ in which occurs an attenuation of $3dB$.
@@
 
-  \item High-pass filter with a simple pole,  module of its frequency responses at the upper right corner of figure~\ref{fig:iir}.
+  \item High-pass filter with a simple pole,  module of its frequency responses at the upper right corner of Figure~\ref{fig:iir}.
 The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of the intermediate variable $x \in [e^{-\pi},1]$:
@@
 
-The band-pass filter in the lower left corner of figure~\ref{fig:iir} has the following coefficients:
+The band-pass filter in the lower left corner of Figure~\ref{fig:iir} has the following coefficients:
 
@@
 
-The coefficients of band-reject filter, depicted in the lower right of figure~\ref{fig:iir}, are:
+The coefficients of band-reject filter, depicted in the lower right of Figure~\ref{fig:iir}, are:
 
@@
      \centering
-         \includegraphics[width=\textwidth]{figures/ruidos}
+         \includegraphics[width=\textwidth]{ruidos}
      \caption{Colors of noise generated by equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul} and \ref{eq:violeta}: spectrum and waveforms.}
@@
 The noise from equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom},
-\ref{eq:azul} and \ref{eq:violeta} are presented in figure~\ref{fig:ruidos}.
+\ref{eq:azul} and \ref{eq:violeta} are presented in Figure~\ref{fig:ruidos}.
 As stated above, the spectra were built with the same phase and frequency for each coefficient, making it straightforward to observe the contribution of treble harmonics and bass frequencies.
@@
      \centering
-         \includegraphics[width=\columnwidth]{figures/vibrato}
+         \includegraphics[width=\columnwidth]{vibrato}
      \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave,  considered $f_s=44.1kHz$.}
@@
      \centering
-         \includegraphics[width=\textwidth]{figures/tremolo}
+         \includegraphics[width=\textwidth]{tremolo}
      \caption{Tremolo with a depth of $V_{dB}=12dB$, with a sawtooth waveform as its oscillatory pattern, with $f'=1.5Hz$ in a sine of $f=40Hz$ (considered a sample frequency of $f_s=44.1kHz$).}
@@
 embrace the creation of `sound amalgams' based on granular synthesis, for now refer specially to the
-figure~\ref{fig:pulsoSubAgl}. This piece is a link to next section contents. 
+Figure~\ref{fig:pulsoSubAgl}. This piece is a link to next section contents. 
 The source code of the music piece is online, as part of the \massa\ toolbox\cite{MASSA}.
@@
 The stated second in equations~\ref{eq:adsr} attend both introduction and ending of sound from zero intensity because they are linear.
-Schematically, figure~\ref{fig:adsr} shows the ADSR envelope in a classical implementation that supports many variations.
+Schematically, Figure~\ref{fig:adsr} shows the ADSR envelope in a classical implementation that supports many variations.
 For example, between attack and decay it is possible to add an extra partition where the maximum amplitude remains for more than a peak.
@@
     \centering
-        \includegraphics[width=\columnwidth]{figures/adsr}
+        \includegraphics[width=\columnwidth]{adsr}
     \caption{An ADSR envelope (\emph{Attack, Decay, Sustain, Release}) applied
@@
 The elements of $S_j$ can be overlapped by mixing them together, as in
-equation~\ref{eq:mixagem} and figure~\ref{fig:mixagem}, for building intervals
+equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, for building intervals
 and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of the events
-in $S_j$, as in equation~\ref{eq:concatenacao} and in figure ~\ref{fig:concatenacao},
+in $S_j$, as in equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao},
 builds melodic sequences and rhythms, which are associated with the musical
@@
     \centering
-        \includegraphics[width=\columnwidth]{figures/movContraponto}
+        \includegraphics[width=\columnwidth]{movContraponto}
     \caption{Different motions of counterpoint aiming to preserve independence
@@
 compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
-A schematic exposition is shown in figure~\ref{fig:pulsoSubAgl}.
+A schematic exposition is shown in Figure~\ref{fig:pulsoSubAgl}.
 
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/metricaMusical}
+        \includegraphics[width=\textwidth]{metricaMusical}
     \caption{Divisions and groupings of the musical pulse for establishing a
@@
 two, four or eight different values, than there is an imperfect relation,
-as shown in figure~\ref{fig:pulsoSubAgl}.
+as shown in Figure~\ref{fig:pulsoSubAgl}.
 
@@
 and at the first and second half of the considered duration. These structures are
-shown in figure~\ref{fig:climax}. The varying parameter can be unexistent, a case in which
+shown in Figure~\ref{fig:climax}. The varying parameter can be unexistent, a case in which
 the arc consists only of a reference structure~\cite{Schoenberg}.
@@
     \centering
-        \includegraphics[width=\columnwidth]{figures/climax}
+        \includegraphics[width=\columnwidth]{climax}
         \caption{Canonical distinctions of musical climax in a given melody and

### 2017-04-24 article/article.tex
incorporada revisao do chu

@@
 
-\author{Antônio Carlos Silvano Pessotti}
+\author{Ant\^onio Carlos Silvano Pessotti}
   \email{antoniopessotti@gmail.com}
@@
 %
-% \author{Osvaldo N. Oliveira Jr.}
-%  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
-%  \email{chu@ifsc.usp.br}
-% \altaffiliation[Also at ]{IFSC-USP}
+\author{Osvaldo N. Oliveira Jr.}
+\homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
+\email{chu@ifsc.usp.br}
+\altaffiliation[Also at ]{IFSC-USP}
 
@@
 
-Music is commonly defined as the art made by sounds and silences.
-Sound is the longitudinal wave of mechanical pressure.
-The frequency bandwidth between $20Hz$ and $20kHz$ is appreciated by the human hearing system with boundaries dependent on the person, climate conditions and sonic characteristics themselves~\cite{Roederer}.
-Considered the speed of sound as $\approx 343.2 m/s$, these limits imply wavelengths of $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.
-
-Human perception of sound involves captivation by bones, stomach, ears, transfer functions of head and torso, and processing by the nervous system~\cite{Roederer}.
-Besides that, the ear is a dedicated organ for the appreciation of these waves.
-Its mechanism decomposes sound into its sinusoidal spectrum and delivers to the nervous system.
-The sinusoidal components are crucial to musical phenomena, as one can perceive in the composition of sounds of musical interest (such as harmonic sounds and noises, exposed in sections~\ref{sec:discNote} and~\ref{sec:internalVar}), and sonic structures of musical interest (such as tunings and scales, in section~\ref{sec:notesMusic}).
-
-The representation of sound is commonly called audio (although these terms are often used without distinction).
-Audio expresses waves prominent from the capture of microphones or from direct synthesis, although these sources entangle as captured sounds are processed to generate new sonorities.
-Digital audio specified by protocols that eases file transfering and storage often imply a quality loss.
+Music is commonly defined as the art made by sounds and silences, where sound corresponds to the longitudinal wave of mechanical pressure.
+The human hearing system perceives sounds within the frequency bandwidth between $20Hz$ and $20kHz$, with the actual limits depending on the person, climate conditions and the sonic characteristics themselves~\cite{Roederer}.
+Since the speed of sound is $\approx 343.2 m/s$, these limits imply wavelengths of $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.
+Such perception involves stimuli in bones, stomach, ears, transfer functions of head and torso, and processing by the nervous system~\cite{Roederer}.
+Obviously, the ear is a dedicated organ for the appreciation of these waves, which decomposes them into their sinusoidal spectra and delivers to the nervous system.
+The sinusoidal components are crucial to musical phenomena, as one can perceive in the composition of sounds of musical interest (such as harmonic sounds and noises, discussed in sections~\ref{sec:discNote} and~\ref{sec:internalVar}), and sonic structures of musical interest (such as tunings and scales, in section~\ref{sec:notesMusic}).
+
+The representation of sound is commonly referred to as audio, although these terms are often used without distinction.
+Audio expresses waves from the capture by microphones or from direct synthesis, although these sources are not neatly distinguishable as captured sounds are processed to generate new sonorities.
+Digital audio specified by protocols that facilitate file transferring and storage often implies a quality loss.
 Standard representation of digital audio, on the other hand, assures perfect reconstruction of the analog wave, within any convenient precision.
-This paradigm consists of equally spaced samples, by $\lambda_s$ durations, each specified by a fixed number of bits.
+This paradigm consists of representing the audio with equally spaced samples, of $\lambda_s$ durations, each specified by a fixed number of bits.
 This is the Pulse Code Modulation (PCM) representation of sound.
@@
 The fixed $2^4=16$ values for the amplitude of each sample, with the regular spacing $\lambda_s$, yields the quantization error or noise.
-This noise diminishes as bit depth increases.
+This noise diminishes as the bit depth increases.
 
@@
     \centering
-        \includegraphics[width=\textwidth]{pcm}
+        \includegraphics[width=\textwidth]{figures/pcm}
         \caption{Pulse Code Modulation (PCM) audio: an analogical signal is represented by 25 samples with 4 bits each.}
@@
 Thus, for general musical purposes, it is necessary to have samples in a rate at least twice the highest frequency heard by humans, that is, $f_s \geq 2\times 20kHz = 40kHz$.
-This is the fundamental reason for the adoption of sampling frequencies like $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and broadcast systems (radio and television), respectively.
+This is the fundamental reason for the adoption of sampling frequencies such as $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and broadcast systems (radio and television), respectively.
 
@@
 The note still stands paradigmatic as the 'fundamental unit' of musical structures and, in practice, it can unfold into sounds that uphold other approaches\footnote{Music from the twentieth century enlarged this traditional comprehension of music.
-This occurred in the concert music, specially in the concrete, electronic and electroacoustic styles.
-On the last decade of the twentieth century, it became evident that popular music had also incorporated sounds without defined pitch and temporal organization, to name a few of simple characteristics.}.
+This occurred in the concert music, especially in the concrete, electronic and electroacoustic styles.
+In the 1990s, it became evident that popular music had also incorporated sounds without defined pitch and temporal organization, to name a few of simple characteristics.}.
     Notes are also convenient for another reason: the average listener -- and considerable part of the specialists -- presupposes rhythmic and pitch organization (made explicit in section~\ref{sec:notesMusic}) as fundamental musical properties, and these are developed in traditional musical theory in terms of notes.
@@
 
-In order to address this concise description of musical elements and structures, in terms of the digitilized sound, the objectives of this paper are:
+In order to address this concise description of musical elements and structures, in terms of the digitalized sound, the objectives of this paper are:
 
@@
 
-In diverse artistic and theoretical contexts, music is conceived as constituted by fundamental units called notes, ``atoms'' that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
-Nowadays, these notes are understood as central elements of certain musical paradigms.
-In a cognitive perspective, the notes are seen as discretized elements that easy and enrich the flow of information through music~\cite{Roederer, Lacerda}.
-Canonically, the principal properties of a musical note are duration, volume, pitch and timbre~\cite{Lacerda}.
-These qualities can be managed quantitatively, dictated by the evenly time spaced sonic samples.
-
-All the relations described in this section are implemented at the file \emph{eqs2.1.py} of the \massa\ toolbox.
-Musical pieces \emph{5 sonic portraits} and \emph{reduced-fi} are also available online in order to corroborate the concepts.
+In diverse artistic and theoretical contexts, music is conceived as comprising fundamental units referred to as notes, ``atoms'' that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
+These notes are now understood as central elements of certain musical paradigms.
+In a cognitive perspective, the notes are seen as discretized elements that facilitate and enrich the flow of information through music~\cite{Roederer, Lacerda}.
+Canonically, the principal properties of a musical note are duration, volume, pitch and timbre~\cite{Lacerda}, which can be quantified using the evenly time spaced sonic samples.
+
+All the relations described in this section are implemented in the file \emph{eqs2.1.py} of the \massa\ toolbox.
+The musical pieces \emph{5 sonic portraits} and \emph{reduced-fi} are also available online to corroborate the concepts.
 
@@
 
-The sensation of sound volume depends on reverberation and harmonic distribution, among other characteristics described on section~\ref{sec:varInternas}.
-One can get volume variations through the potency of the wave~\cite{Chowning}:
+The sensation of sound volume depends on reverberation and harmonic distribution, among other characteristics described in section~\ref{sec:varInternas}.
+One can get volume variations through the power of the wave~\cite{Chowning}:
 
 \begin{equation}\label{eq:potencia}
-pot(T_i)=\frac{\sum_{i=0}^{\Lambda -1} t_i^2}{\Lambda}
+pow(T_i)=\frac{\sum_{i=0}^{\Lambda -1} t_i^2}{\Lambda}
 \end{equation}
@@
 The final volume is dependent on the speakers amplification of the signal.
-Thus, what matters is the relative potency of a note in relation to the other ones around it, or the potency of a music section in relation to the rest.
+Thus, what matters is the relative power of a note in relation to the other ones around it, or the power of a music section in relation to the rest.
 Differences in volume are measured in decibels, calculated directly from the amplitudes through energy or potency:
@@
 
-As shown, it is necessary an amplitude increase by a bit more than a factor of 3 for a doubled volume.
-These values are guides for increasing or decreasing the absolute values on the sample sequences with musical purposes.
+An amplitude increase by a factor slightly above of 3 is required for yielding a doubled volume.
+These values are guides for increasing or decreasing the absolute values in the sample sequences with musical purposes.
 The conversion from decibels to amplitude gain (or attenuation) is straightforward:
@@
 
-\noindent where $A$ is the multiplicative factor that relates the amplitudes before and after the amplification.
+\noindent where $A$ is the multiplicative factor that relates the amplitudes before and after amplification.
 
@@
 
-As exposed in the previous subsections, to a note corresponds a sequence $T_i$ in which duration and volume are directly related to the size of the sequence and the amplitude of its samples, respectively.
+To a note corresponds a sequence $T_i$ in which duration and volume are directly related to the size of the sequence and the amplitude of its samples, respectively.
 The pitch is specified by the fundamental frequency $f_0$ whose cycle has duration $\delta_{f_0}=1/f_0$.
-This duration, multiplied by the sampling frequency $f_s$ results on the number of samples per cycle: $\lambda_{f_0}=f_s .
+This duration multiplied by the sampling frequency $f_s$ results in the number of samples per cycle: $\lambda_{f_0}=f_s .
 \delta_{f_0} =f_a/f_0$.
 
-For didactic reasons, be $f_0$ such that it divides $f_s$ and $\lambda_{f_0}$ results integer. If $T_i^f$ is a sonic sequence of fundamental frequency $f$, then:
+For didactic reasons, let $f_0$ be such that it divides $f_s$ and $\lambda_{f_0}$ results integer. If $T_i^f$ is a sonic sequence of fundamental frequency $f$, then:
 
@@
 
-In the next section, frequencies $f$ that does not divide $f_s$ will be considered. This restriction does not imply in lost of generality of this section's content.
+In the next section, frequencies $f$ that do not divide $f_s$ will be considered. This restriction does not imply in loss of generality of this current section's content.
 
@@
 
-In a sound with harmonic spectrum, the (wave) period corresponds to the duration of the cycle, given by the inverse of the fundamental frequency.
+In a sound with harmonic spectrum, the (wave) period corresponds to the cycle duration, given by the inverse of the fundamental frequency.
 The trajectory of the wave inside the period - called the waveform - defines a harmonic spectrum and, thus, a timbre\footnote{Timbre is a subjective and complex characteristic.
 Physically, the timbre is multidimensional and given by the temporal dynamics of energy in the spectral components that are harmonic or noisy.
-Beyond that, the word timbre is used to designate different things: one same note can have (be produced with) different timbres, an instrument has different timbres, two instruments of the same family have, at the same time, the same timbre that blends them in the same family, and different timbres as they are different instruments.
-It is worth to mention that timbre is not always manifested in spectral traces, since cultural or circumstantial aspects alter our perception of timbre}.
-Sonic spectra with minimum differences can result in timbres with crucial expressive differences and, consequently, different timbres can be produced using different spectra\cite{Roederer}.
-
-The simplest case is the spectrum that consists of only of the fundamental frequency $f$.
-This defines the sinusoid: a frequency in pure oscillatory movement called `simple harmonic movement'.
-Be $S_i^f$ a sequence whose samples $s_i^f$ describes a sinusoid of frequency $f$:
+In addition, the word timbre is used to designate different things: one same note can have (be produced with) different timbres, an instrument has different timbres, two instruments of the same family have, at the same time, the same timbre that blends them in the same family, and different timbres as they are different instruments.
+It is worth mentioning that timbre is not always manifested in spectral traces, since cultural or circumstantial aspects alter our perception of timbre}.
+Sonic spectra with minimum differences can result in timbres with crucial differences and, consequently, distinct timbres can be produced using different spectra\cite{Roederer}.
+
+The simplest case is the spectrum with only the fundamental frequency $f$, which is a sinusoid as in "simple harmonic oscillation".
+Let $S_i^f$ be a sequence whose samples $s_i^f$ describe a sinusoid of frequency $f$:
 
@@
 
-The triangular waveform present only odd harmonics falling with $-12dB/octave$:
+The triangular waveform has only odd harmonics falling with $-12dB/octave$:
 \begin{equation}\label{triangular}
@@
 
-The square wave preservers only odd harmonics falling at $-6dB/octave$:
+The square wave preserves only odd harmonics falling at $-6dB/octave$:
 
@@
 The square wave can be used in a subtractive synthesis with the purposes of mimicking a clarinet.
-This instrument has only the odd harmonic components and the square wave is convenient with its abundant energy in high frequencies.
+This instrument has only the odd harmonic components and the square wave is convenient with its abundant energy at high frequencies.
 The sawtooth is a common starting point for a subtractive synthesis, because it has both odd and even harmonics with high energy.
-In general, these waveforms are appreciated as excessively rich in sharp harmonics, and attenuator filtering on treble and middle parts of the spectrum is specially useful for reaching a more natural and pleasant sound.
+In general, these waveforms are appreciated as excessively rich in sharp harmonics, and attenuator filtering on treble and middle parts of the spectrum is especially useful for reaching a more natural and pleasant sound.
 The relatively attenuated harmonics of the triangle wave makes it the more functional - among the listed ones - to be used in the synthesis of musical notes without any treatment.
 The sinusoid is often a nice choice, but a problematic one.
-While pleasant if not loud in a very high pitch (above 500Hz, it requires careful dosage), the pitch is not accuratly detected, specially in low frequencies.
-It requires a big amplitude to give sinusoid volume, if compared to other waveforms.
+While pleasant if not loud in a very high pitch (above 500Hz, it requires careful dosage), the pitch is not accurately detected, particularly at low frequencies.
+It requires a large amplitude to give sinusoid volume, if compared to other waveforms.
 Both particularities are seen as a consequence of the non-existence of pure sinusoidal sounds in nature~\cite{Roederer}.
@@
     \centering
-        \includegraphics[width=\textwidth]{waveForms_}
+        \includegraphics[width=\textwidth]{figures/waveForms_}
     \caption{Basic musical waveforms: (a) the synthetic  waveforms; (b) the real waveforms.}
@@
 The spectrum of each basic waveform is in Figure~\ref{fig:espectroDeOndas}.
-The isolated and exactly harmonic components of the spectrum is a consequence of the fixed period usage.
-The sinusoid consists of an unique node in the spectrum, pure frequency.
-The figure exibit the spectra described: the sawtooth is the only waveform with a complete harmonic series (odd and even components); triangular and square waves has the same components (odd harmonics), decaying at $-12dB/octave$ and $-6dB/octave$, respectively.
+The isolated and exactly harmonic components of the spectrum is a consequence of the use of a fixed period.
+The sinusoid consists of a unique node in the spectrum, pure frequency.
+The figure exhibits the spectra described: the sawtooth is the only waveform with a complete harmonic series (odd and even components); triangular and square waves have the same components (odd harmonics), decaying at $-12dB/octave$ and $-6dB/octave$, respectively.
 
@@
     \centering
-        \includegraphics[width=\textwidth]{waveSpectrum}
+        \includegraphics[width=\textwidth]{figures/waveSpectrum}
     \caption{Spectrum of basic artificial musical waveforms.}
@@
 Additionally, the number of harmonics will be limited by the Nyquist frequency $f_s/2$.
-
 From a musical perspective, it is critical to internalize that energy in a component of frequency $f_n$ means an oscillation in the constitution of the sound, purely harmonic and in that frequency $f_n$.
-This energy, specifically concentrated on the frequency $f_n$, is separated by the ear for further cognitive processes (this separation is done in many species with mechanisms similar to the human cochlea~\cite{Roederer}).
-
-The sinusoidal components are usually the main responsibles for timbre qualities.
+This energy, specifically concentrated on the frequency $f_n$, is separated by the ear for further cognitive processes (this separation is done in several mechanisms similar to the human cochlea~\cite{Roederer}).
+
+The sinusoidal components are usually responsible for timbre qualities.
 If they are not presented in harmonic proportions (small number relations), the sound is perceived as noisy or dissonant, in opposition to sonorities with an unequivocally established fundamental.
 Furthermore, the notion of absolute pitch relies on the similarity of the spectrum to the harmonic series~\cite{Roederer}.
-
-In the case of a fixed length period and waveform, the spectrum is perfectly harmonic and static, and each waveform is compound of specific proportions of harmonic components.
-High  curvatures are sign of high harmonics in the wave.
+For a fixed length period and waveform, the spectrum is perfectly harmonic and static, and each waveform is comprised of specific proportions of harmonic components.
+High  curvatures are a sign of high harmonics in the wave.
 Figure~\ref{fig:formasDeOnda} depicts a wave, labeled as ``sampled real sound'', with a period of $\Lambda_f=114$ samples, extracted from a relatively well behaved recorded sound.
-The oboe wave was also sampled in $44.1kHz$.
-The chosen period for sampling was relatively short, with $98$ samples, and corresponds to the frequency $\frac{44100}{98}=450Hz$, which corresponds to a slightly out-of-tune A4 pitch.
-One can notice from the curvatures: the oboe's rich spectrum in high frequencies and the lower spectrum of the real sound.
-
-The sequence 
-$ R_i=\{ r_i \}_0^{\lambda_f-1}$ of samples in the real sound of Figure~\ref{fig:formasDeOnda} can be taken as basis for a sound $T_i^f$ in the following way:
+The oboe wave was also sampled at $44.1kHz$.
+The chosen period for sampling was relatively short, with $98$ samples, and corresponds to the frequency $\frac{44100}{98}=450Hz$, which is associated with a slightly out-of-tune A4 pitch.
+One can notice from the curvatures: the oboe's rich spectrum at high frequencies and the lower spectrum of the real sound.
+
+The sequence $ R_i=\{ r_i \}_0^{\lambda_f-1}$ of samples in the real sound of Figure~\ref{fig:formasDeOnda} can be taken as basis for a sound $T_i^f$ in the following way:
 
@@
 
-The resulting sound has the momentary spectrum of the original waveform.
-As a consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise and with variations typical of the natural phenomenon.
+The resulting sound has the spectrum of the original waveform.
+As a consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise or variations typical of the natural phenomenon.
 This can be observed in Figure~\ref{fig:espectroOboe}, that shows the spectrum of the original oboe note and a note with same duration, whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}.
@@
     \centering
-        \includegraphics[width=\textwidth]{oboeNaturalSampledSpectrum}
+        \includegraphics[width=\textwidth]{figures/oboeNaturalSampledSpectrum}
     \caption{Spectrum of the sonic waves of a natural oboe note and from a sampled period.
@@
 
-The presence and behavior of these sinusoidal components in the discretized sound have some particularities.
-Considering a signal $T_i$ and its corresponding Fourier decomposition $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$, the recomposition is the sum of the frequency components as time samples\footnote{It is important to note that the factor $\frac{1}{\Lambda}$ could be distributed among the Fourier transform and its reconstruction, as preferred.}:
+These sinusoidal components in the discretized sound have some particularities.
+Considering a signal $T_i$ and its corresponding Fourier decomposition $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$, the recomposition is the sum of the frequency components as time samples\footnote{The factor $\frac{1}{\Lambda}$ could be distributed among the Fourier transform and its reconstruction, as preferred.}:
 
@@
 b_k$ defines the amplitude and phase of each frequency: $w_k=\frac{2\pi}{\Lambda}k$ in radians or $f_k=w_k\frac{f_a}{2\pi}=\frac{f_a}{\Lambda}k$ in Hertz, taking into account the respective limits in $\pi$ and in $\frac{f_a}{2}$ given by the Nyquist Theorem.
-As can be inferred, $j$ is the complex number, with $j^2=-1$.
+$j$ is the complex number, with $j^2=-1$.
 
@@
 
-Equation~\ref{moduloEfase} shows how the imaginary term of $c_k$ adds a phase to the real sinusoid: the terms $b_k$ enables the phase sweep $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right]$ given by $tg^{-1}\left(\frac{b_k}{a_k}\right)$ which has this image.
-The terms $a_k$ specifies the right or left side of the trigonometric circle, which completes the phase domain: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.
+Equation~\ref{moduloEfase} shows how the imaginary term of $c_k$ adds a phase to the real sinusoid: the terms $b_k$ enable the phase sweep $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right]$ given by $tg^{-1}\left(\frac{b_k}{a_k}\right)$ which has this image.
+The terms $a_k$ specify the right or left side of the trigonometric circle, which completes the phase domain: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.
 
@@
      \centering
-         \includegraphics[width=\textwidth]{amostras2c__}
+         \includegraphics[width=\textwidth]{figures/amostras2c__}
      \caption{Oscillation of 2 samples (maximum frequency for any $f_a$).
@@
 
-Figure~\ref{fig:amostras2} shows two samples and its spectral components.
-In this case, the Fourier decomposition has one unique pair of coefficients $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relative to frequencies $\{f_k\}_0^1=\left\{w_k\frac{f_a}{2\pi}\right\}_0^1=\left\{k\frac{f_a}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_a}{2}=f_{\text{max}}\right\}$
+Figure~\ref{fig:amostras2} shows two samples and their spectral components, where the Fourier decomposition has one unique pair of coefficients $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relative to frequencies $\{f_k\}_0^1=\left\{w_k\frac{f_a}{2\pi}\right\}_0^1=\left\{k\frac{f_a}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_a}{2}=f_{\text{max}}\right\}$
 with energies $e_k=\frac{(c_k)^2}{\Lambda=2}$.
-The role of amplitudes $a_k$ is clearly observed with $\frac{a_0}{2}$, the fixed offset\footnote{Also called \emph{bias}.} and $\frac{a_1}{2}$, oscillation amplitude with frequency given by the relation $f_k=k \frac{f_a}{\Lambda=2}$.
-This case have special relevance.
-It is necessary at least 2 samples to represent an oscillation and it yields the Nyquist frequency $f_{\text{max}}=\frac{f_a}{2}$, which is the maximum frequency in a sound sampled with $f_a$ samples per second\footnote{Any sampled signal has this property, not only the digitalized sound.}.
-
-All fixed sequences $T_i$ of only $3$ samples also have just $1$ frequency, since their first harmonic has $1,5$ samples and exceeds the bottom limit of 2 samples, i.e.\ the frequency of the harmonic would exceed the Nyquist frequency:  $\; \frac{2.
+The role of amplitudes $a_k$ is clearly observed with $\frac{a_0}{2}$, the fixed offset\footnote{Also called \emph{bias}.} and $\frac{a_1}{2}$, oscillation amplitude with frequency given by $f_k=k \frac{f_a}{\Lambda=2}$.
+This case has special relevance.
+At least 2 samples are necessary to represent an oscillation and it yields the Nyquist frequency $f_{\text{max}}=\frac{f_a}{2}$, which is the maximum frequency in a sound sampled with $f_a$ samples per second\footnote{Any sampled signal has this property, not only the digitalized sound.}.
+
+All fixed sequences $T_i$ of only $3$ samples also have just $1$ frequency, since the first harmonic would have $1.5$ samples and exceeds the bottom limit of 2 samples, i.e.\ the frequency of the harmonic would exceed the Nyquist frequency:  $\; \frac{2.
 f_a}{3} > \frac{f_a}{2} $.
 The coefficients $\{c_k\}_0^{\Lambda-1=2}$ are present in 3 frequency components.
-One is relative to zero frequency ($c_0$), and the other two ($c_1$ and $c_2$) have the same role for the reconstruction of sinusoid with $f=f_a/3$.
+One is relative to zero frequency ($c_0$), and the other two ($c_1$ and $c_2$) have the same role for reconstructing a sinusoid with $f=f_a/3$.
 
@@
      \centering
-         \includegraphics[width=\textwidth]{amostras3b}
-     \caption{Three fixed samples presents only one non-null frequency. $c_1=c_2^*$ and $w_1 \equiv w_2$.}
+         \includegraphics[width=\textwidth]{figures/amostras3b}
+     \caption{Three fixed samples present only one non-null frequency. $c_1=c_2^*$ and $w_1 \equiv w_2$.}
          \label{fig:amostras3}
@@
 $\Lambda$ real samples $t_i$ result in $\Lambda$ complex coefficients $c_k=a_k+j.b_k$.
-The coefficients $c_k$ are equivalent two by two, corresponding to the same frequencies and with same contribution to its reconstruction.
+The coefficients $c_k$ are equivalent two by two, corresponding to the same frequencies and with the same contribution to its reconstruction.
 They are complex conjugates: $a_{k1}=a_{k2}$ and $b_{k1}=-b_{k2}$ and, as a consequence, the modules are equal and phases have opposite signs.
-Remembering that $f_k = k\frac{f_a}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $.
+Recalling that $f_k = k\frac{f_a}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $.
 When $k > \frac{\Lambda}{2}$, the frequency $f_k$ is mirrored by $\frac{f_a}{2}$ in this way: $f_k=\frac{f_a}{2} - (f_k-\frac{f_a}{2})=f_a-f_k=f_a - k\frac{f_a}{\Lambda}=(\Lambda-k)\frac{f_a}{\Lambda} \;\;\;\; \Rightarrow \;\;\;\; f_k\equiv f_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
 
-The same can be observed with $w_k=f_k.\frac{2\pi}{f_a}$ and the periodicity $2\pi$: it follows that $w_k=-w_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
+The same applies to $w_k=f_k.\frac{2\pi}{f_a}$ and the periodicity $2\pi$: it follows that $w_k=-w_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
 Given the cosine (an even function) and the inverse tangent (an odd function), the components in $w_k$ and $w_{\Lambda-k}$ contribute with coefficients $c_k$ and $c_{\Lambda-k}$ in the reconstruction of the real samples.
-
 In other words, in a decomposition of $\Lambda$ samples, the $\Lambda$ frequency components $\{c_i\}_0^{\Lambda-1}$ are equivalents in pairs,
 except for $f_0$, and, when $\Lambda$ is even, for $f_{\Lambda/2}=f_{\text{max}}=\frac{f_a}{2}$.
-Both components are isolated, i.e.\ there is one and only component in frequency $f_0$ or $f_{\Lambda/2}$ (if even $\Lambda$).
-One can verify this assertion considering $k=0$ and $k=\Lambda/2$. Uncoiled: $f_{\Lambda/2}=f_{(\Lambda-\Lambda/2) = \Lambda/2}$ and $f_0=f_{(\Lambda-0)=\Lambda}=f_0$.
-Furthermore, these two frequencies (zero and Nyquist frequency) does not have phase variation, being their coefficients strictly real.
-In this way, it is possible to conclude the number $\tau$ of equivalent coefficient pairs:
+Both components are isolated, i.e.\ there is one and only one component at frequency $f_0$ or $f_{\Lambda/2}$ (if $\Lambda$ is even).
+This assertion can be verified with $k=0$ and $k=\Lambda/2$ like this: $f_{\Lambda/2}=f_{(\Lambda-\Lambda/2) = \Lambda/2}$ and $f_0=f_{(\Lambda-0)=\Lambda}=f_0$.
+Furthermore, these two frequencies (zero and Nyquist frequency) do not have phase variation, their coefficients being strictly real. Therefore, the number $\tau$ of equivalent coefficient pairs is:
 
@@
 
-This discussion makes clear the equivalences ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} and ~\ref{equivalenciasFases}:
+This discussion makes it clear the equivalence ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} and ~\ref{equivalenciasFases}:
 
@@
     \centering
-        \includegraphics[width=\textwidth]{amostras4___}
+        \includegraphics[width=\textwidth]{figures/amostras4___}
     \caption{Frequency components for 4 samples.}
@@
 
-To expose the general case for components combination in each sample $t_i$, one can gather relations in equation~\ref{moduloEfase} for the real signal reconstruction, relations of modules~\ref{equivalenciasModulos} and phases equivalences~\ref{equivalenciasFases}, the number of paired coefficients~\ref{coefsPareados}, and equivalence of paired frequencies~\ref{equivalenciasFreqs}:
+To discuss the general case for components combination in each sample $t_i$, one can gather relations in equation~\ref{moduloEfase} for the real signal reconstruction, relations of modules~\ref{equivalenciasModulos} and phase equivalences~\ref{equivalenciasFases}, the number of paired coefficients~\ref{coefsPareados}, and equivalence of paired frequencies~\ref{equivalenciasFreqs}:
 
@@
     \centering
-        \includegraphics[width=\textwidth]{amostras4formas__}
+        \includegraphics[width=\textwidth]{figures/amostras4formas__}
     \caption{Basic wave forms with 4 samples.}
@@
 with independence).
-Figure~\ref{fig:amostras4} depicts the basic waveforms with 4 samples and its two (possible) components.
+Figure~\ref{fig:amostras4} depicts the basic waveforms with 4 samples and their two (possible) components.
 The individual contributions sum to the original waveform and a brief inspection reveals the major curvatures resulting from the higher frequency, while the fixed offset is captured in the zero frequency component.
@@
     \centering
-        \includegraphics[width=\textwidth]{amostras6_}
+        \includegraphics[width=\textwidth]{figures/amostras6_}
     \caption{Frequency components for 6 samples: 4 sinusoids, one of them is the \emph{bias} with zero frequency.}
@@
 
-Figure~\ref{fig:amostras6} presents the sinusoidal decomposition for 6 samples and Figure~\ref{fig:formas6} presents the decomposition of the basic wave forms.
-In this case, the waveforms have spectra with fundamental differences: square and triangular have the same components but with different proportions, while the sawtooth have an extra component.
+Figure~\ref{fig:amostras6} shows the sinusoidal decomposition for 6 samples, while Figure~\ref{fig:formas6} presents the decomposition of the basic wave forms.
+In this case, the waveforms have spectra with fundamental differences: square and triangular have the same components but with different proportions, while the sawtooth has an extra component.
 
@@
     \centering
-        \includegraphics[width=\textwidth]{amostras6formas__}
+        \includegraphics[width=\textwidth]{figures/amostras6formas__}
     \caption{Basic waveforms with 6 samples: triangular and square waveforms have odd harmonics, with different proportions and phases; the sawtooth has even harmonics.}
@@
 
-Be $f$ such that it divides $f_a$\footnote{As pointed before, this limitation simplify the explanation without losing generality, and will be overcome in the next section.}.
+Let $f$ be such that it divides $f_a$\footnote{As pointed before, this limitation simplifies the explanation without losing generality, and will be overcome in the next section.}.
 A sequence $T_i$ of sonic samples separated by $\delta_a=1/f_a$ describes a musical note with a frequency of $f$ Hertz and $\Delta$ seconds of duration if, and only if, it has the periodicity $\lambda_f=f_a/f$ and size $\Lambda=\lfloor f_a .
@@
 Nevertheless, it is necessary to choose a waveform for the samples $t_i$ to have a value.
-An unique period from the basic waveforms can be used to specify the note, where $\lambda_f=\frac{f_a}{f}$ is the number of samples at the period.
+A unique period from the basic waveforms can be used to specify the note, where $\lambda_f=\frac{f_a}{f}$ is the number of samples at the period.
 Here, $L_i^{f,\, \delta_f} $ is the sequence that describes a period of the waveform $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ with duration $\delta_f=1/f$ (as given by equations~\ref{sinusoid}, ~\ref{sawTooth}, ~\ref{triangular} and ~\ref{square}) and $R_i^f$ is a sampled real waveform:
@@
 
-Although it is not one of its four basic properties, a musical note always has a spatial localization: the note source position at the ordinary physical space.
-The reverberation the environment in which sound occurs is the matter of \emph{spatialization}.
-Both fields, spatialization and spatial localization, are widely valued by audiophiles and music industry~\cite{floEsp}.
+A musical note always has a spatial localization, even though this is not one of its four basic properties: the note source position at the ordinary physical space.
+The reverberation in the environment in which sound occurs is the matter of \emph{spatialization}.
+Both fields, spatialization and spatial localization, are widely valued by audiophiles and the music industry~\cite{floEsp}.
 
@@
 
-It is understood that the perception of sound localization occurs in our nervous system by three informations: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering realized by the human body, including its chest, head and ears~\cite{Roederer, hrtf, Heeger}.
+It is understood that the perception of sound localization occurs in our nervous system by three pieces of information: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, including its chest, head and ears~\cite{Roederer, hrtf, Heeger}.
 
@@
     \centering
-        \includegraphics[width=.5\textwidth]{espacializacao___}
+        \includegraphics[width=.5\textwidth]{figures/espacializacao___}
     \caption{Detection of sound source localization: schema used to calculate Interaural Time Difference (ITD) and Interaural Intensity Difference (IID).}
@@
 
-Where $\zeta$ is the distance between ears $\zeta$,  known to be $\zeta \approx 21,5cm$ in an adult human.
-Immediate calculations result in the ITD:
+Where $\zeta$ is the distance between ears $\zeta$, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in ITD:
 
@@
 \noindent which, converted to amplitude, yields $IID_a=\frac{d}{d'}$.
-The $IID_a$ can be used as a multiplicative constant to the right channel of a stereo sound signal: $\{t_i'\}_0^{\Lambda -1}=\{IID_a .
+$IID_a$ can be used as a multiplicative constant to the right channel of a stereo sound signal: $\{t_i'\}_0^{\Lambda -1}=\{IID_a .
 t_i\}_0^{\Lambda -1}$, where $\{t_i'\}$ are samples of the wave incident in the left ear.
@@
 
-Although simple until here, spatial localization depends considerably of other cues. By using only ITD and IID it is possible to specify solely the horizontal angle (azimuthal) $\theta$ given by:
+Spatial localization depends considerably on other cues. By using only ITD and IID it is possible to specify solely the horizontal angle (azimuthal) $\theta$ given by:
 
@@
 \noindent with $x,y$ as presented in Figure~\ref{fig:spac}.
-Henceforth, there are problems when $\theta$ fall upon the so called "cone of confusion": the same pair of ITD and IID results in a large number of points inside the cone.
-On those points the inference of the azimuthal angle depends specially of the attenuative filtering for high frequencies, since the head interferes much more in the treble than in bass waves~\cite{Heeger,hrtf}.
-Also relevant to the hearing of lateral sources is that low frequencies diffracts and the wave arrive to the opposite ear with a delay of $\approx 0,7ms$.\cite{floEsp}
+Henceforth, there are problems when $\theta$ falls within the so-called "cone of confusion": the same pair of ITD and IID results in a large number of points inside the cone.
+On those points the inference of the azimuthal angle depends especially on the attenuation filtering for high frequencies, since the head interferes much more in the treble than in bass waves~\cite{Heeger,hrtf}.
+Also relevant to the hearing of lateral sources is that low frequencies diffract and the wave arrives to the opposite ear with a delay of $\approx 0.7ms$.\cite{floEsp}
 
@@
 The cone itself is not shown in Figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not encountered in the literature.
-Given the filtering and diffraction dependent of the sound spectrum, it is hard, if not impossible, to correctly draw the confusion cone.
-Even so, the cone of confusion can be understood as a cone with its top placed in the middle of head and growing out in the direction of each ear~\cite{hrtf}.
+Given the filtering and diffraction dependent on the sound spectrum, it is hard, if not impossible, to correctly draw the confusion cone.
+Even so, the cone of confusion can be understood as a cone with its top placed in the middle of the head and growing out in the direction of each ear~\cite{hrtf}.
 
@@
 Spatialization results from sound reflections and absorptions by room/environment surface where the note is played.
-The sound propagates through the air with a speed of $\approx 343,2m/s$ and can be emitted from a source with any directionality pattern.
-When a sound pulse encounters a surface there is reflection.
-In this reflection occurs: 1) the inversion of propagation speed component that is normal to the surface;  2) energy absorption, specially in trebles.
-The sonic waves propagate until they reaches inaudible levels (and even further).
+The sound propagates through the air with a speed of $\approx 343.2m/s$ and can be emitted from a source with any directionality pattern.
+When a sound pulse encounters a surface there is reflection, and there are: 1) inversion of the propagation speed component normal to the surface;  2) energy absorption, especially in trebles.
+The sonic waves propagate until they reach inaudible levels (and even further).
 As a sonic front reaches the human ear, it can be described as the original sound, with the last reflection point as the source, and the absorption filters of each surface it has reached.
 It is possible to simulate reverberations that are impossible in real systems.
-For example, it is possible to use asymmetric reflections with relation to the axis perpendicular to the surface, or to increase specific frequency bands (known as 'resonances'), both characteristics are not found in real systems.
-
-There are some reverberation models that are less related to each independent reflection, exploring valuable information to the auditory system.
+For example, it is possible to use asymmetric reflections with relation to the axis perpendicular to the surface, or to increase specific frequency bands (known as 'resonances'); neither of these characteristics are found in real systems.
+
+There are reverberation models less related to each independent reflection, exploring valuable information to the auditory system.
 In fact, reverberation can be modeled with a set of 2 temporal and spectral sections:
@@
 Smith III points that reasonable concert rooms have total reverberation time of $\approx 1.9$ seconds, and that the period of first reflections is around $0.1$ seconds.
-These values suggest that, in the given conditions, there are perceived wave pulses which propagates for $652.08$ meters long ($83.79k$ samples in $f_a=44.1kHz$) before reaching the ear.
-In addition, sound reflections made after propagation for $34.32$ meters long ($4.41k$ samples in $f_a=44.1kHz$) have incidences less distinct by hearing.
-These first reflections are particularly important to spacial sensation.
+With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08$ m ($83.79k$ samples in $f_a=44.1kHz$) before reaching the ear.
+In addition, sound reflections made after propagation for $34.32$ m ($4.41k$ samples in $f_a=44.1kHz$) have incidences less distinct by hearing.
+These first reflections are particularly important to spatial sensation.
 The first incidence is the direct sound, described by ITD and IID of equations~\ref{eq:dti} and~\ref{eq:dii}.
-Admitting that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, dependent of room dimension, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_a=44.1kHz$).
-It is possible to experimentally verify that the number of reflections increases with squared proportion $\approx k.n^2$.
-A discussion about the use of convolutions and filtering to favor implementation of these phenomena is provided in subsection~\ref{subsec:mus2}, specially at the paragraphs about reverberation.
+Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_a=44.1kHz$).
+It is experimentally verifiable that the number of reflections increases with the square of $\approx k.n^2$.
+A discussion about the use of convolutions and filtering to favor implementation of these phenomena is provided in subsection~\ref{subsec:mus2}, particularly in the paragraphs about reverberation.
 
@@
 
-Once defined the basic note, it is convenient to build musical structures with sequences based on these particles.
-The sum of elements with same index of $N$ sequences $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ with same size $\Lambda$ results in the overlapped spectral contents of each sequence, in a process called mixing:
+Once the basic note is defined, it is convenient to build musical structures with sequences based on these particles.
+The sum of elements with same index of $N$ sequences $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ with same size $\Lambda$ results in the overlapped spectral contents of each sequence, in a process referred to as mixing:
 
@@
     {\centering
-        \includegraphics[width=\columnwidth]{mixagem}}
+        \includegraphics[width=\columnwidth]{figures/mixagem}}
     \caption{Mixing of three sound sequences. The amplitudes are directly overlapped.}
@@
 Figure~\ref{fig:mixagem} illustrates this overlapping process of discretized sound waves, each with 100 samples.
-It is possible to conclude that, if $f_a=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_a}{100/2}=882Hz$, $\frac{f_a}{100/4}=1764Hz$ and $\frac{f_a}{100/5}=2205Hz$.
-The samples duration are very short $\frac{f_a=44.1kHz}{100} \approx 2 \text{ milliseconds}$.
-One can complete the sequence with zeros to sum (mix) sequences with different sizes.
+If $f_a=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_a}{100/2}=882Hz$, $\frac{f_a}{100/4}=1764Hz$ and $\frac{f_a}{100/5}=2205Hz$.
+The samples duration is very short $\frac{f_a=44.1kHz}{100} \approx 2 \text{ milliseconds}$.
+One can complete the sequence with zeroes to sum (mix) sequences with different sizes.
 
 The mixed notes are generally separated by the ear according to the physical laws of resonance and by the nervous system~\cite{Roederer}.
-This musical notes mixing process results in musical harmony whose intervals between frequencies and chords of simultaneous notes guide subjective and abstract aspects of music and its appreciation~\cite{Harmonia}, which is tackled in section~\ref{sec:notesMusic}.
+This process of mixing musical notes results in musical harmony whose intervals between frequencies and chords of simultaneous notes guide subjective and abstract aspects of music and its appreciation~\cite{Harmonia}, which is addressed in section~\ref{sec:notesMusic}.
 
@@
 
-This mechanism is demonstrated in Figure~\ref{fig:concatenacao} with same sequences of Figure~\ref{fig:mixagem}.
+This mechanism is demonstrated in Figure~\ref{fig:concatenacao} with the same sequences of Figure~\ref{fig:mixagem}.
 Although the sequences are short for the usual sample rates, it is easy to observe the concatenation of sound sequences.
-Besides that, each note has a duration larger than $100ms$ if $f_a<1kHz$.
+In addition, each note has a duration larger than $100ms$ if $f_a<1kHz$.
 
@@
 {    \centering
-        \includegraphics[width=\columnwidth]{concatenacao}}
-    \caption{Concatenation of three sound sequences by temporal overlap of its samples.}
+        \includegraphics[width=\columnwidth]{figures/concatenacao}}
+    \caption{Concatenation of three sound sequences by temporal overlap of their samples.}
         \label{fig:concatenacao}
@@
 
-The musical piece \emph{reduced-fi} explores the temporal juxtaposition of notes, resulting in a homophobic piece.
+The musical piece \emph{reduced-fi} explores the temporal juxtaposition of notes, resulting in a homophonic piece.
 The vertical principle is demonstrated at the \emph{sounic pictures}, static sounds with peculiar spectrum.
@@
 With the basic musical digital note carefully described, the next section develops the temporal evolution of its contents as in \emph{glissandi} and volume envelopes.
-Filtering of spectral components and noise generation complements the musical note as a self contained unity.
+Filtering of spectral components and noise generation complements the musical note as a self-contained unity.
 Section~\ref{notasMusica} is dedicated to the organization of these notes as music by using metrics and trajectories, with regards to traditional music theory.
@@
 
-The basic digital music note defined in section~\ref{sec:notaDisc} has following parameters: duration, pitch, intensity (volume) and timbre.
-This is a useful and paradigmatic model, but does not exhaust all the aspects of a musical note.
-
-First of all, characteristics of the note modifie along the note itself~\cite{Chowning}.
-For example, a 3 seconds piano note has intensity with abrupt rise at beginning and progressive decay, has spectrum variations with harmonics decaying some others that emerge along time.
-These variations are not mandatory, but sound synthesis guidance for musical uses, as it reflects how sounds appear in nature.
-This is so considered to be true, that there is a rule of thumb: to make a sound that incites interest by itself, do internal variations on it~\cite{Roederer}.
+The basic digital music note defined in section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (volume) and timbre.
+This is a useful and paradigmatic model, but it does not exhaust all the aspects of a musical note.
+First of all, characteristics of the note modify along the note itself~\cite{Chowning}.
+For example, a 3 s piano note has intensity with abrupt rise at the beginning and progressive decay, has spectrum variations with harmonics decaying and some others emerging along time.
+These variations are not mandatory, but they are used in sound synthesis for music because they reflect how sounds appear in nature.
+This is considered to be so true that there is a rule of thumb: to make a sound that incites interest by itself, do internal variations on it~\cite{Roederer}.
 To explore all the ways in which variations occur within a note is out of the scope of any work, given the sensibility of the human ear and the complexity of human sound cognition.
@@
 
-The \emph{Lookup Table} (or simply LUT), is an array for indexed operations which substitutes continuous and repetitive calculation.
-It is used to reduce computational complexity and to use functions without direct calculation, as data sampled from nature.
-In music its usage transcends these applications, as it simplifies many operations and makes possible the use a single wave period to synthesize sounds in the whole audible spectrum, with any waveform.
+The \emph{Lookup Table} (or simply LUT) is an array for indexed operations which substitutes continuous and repetitive calculation.
+It is used to reduce computational complexity and employ functions without direct calculation, as data sampled from nature.
+In music its usage transcends these applications, as it simplifies many operations and makes it possible to use a single wave period to synthesize sounds in the whole audible spectrum, with any waveform.
 
@@
     \centering
-        \includegraphics[width=\textwidth]{lut}
-    \caption{Search in the \emph{lookup table} to synthesize sounds in different frequencies using an unique waveform with high resolution.}
+        \includegraphics[width=\textwidth]{figures/lut}
+    \caption{Search in the \emph{lookup table} to synthesize sounds at different frequencies using a unique waveform with high resolution.}
         \label{fig:lut}                                                                                                            
@@
 
-Be $\widetilde{\Lambda}$ the wave period in samples and $\widetilde{L_i} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ the samples $\widetilde{l_i}$ (refer to Equation~\ref{periodoUnico}), a sequence $T_i^{f,\,\Delta}$ with samples of a sound with frequency $f$ and duration $\Delta$ can be obtained by means of $\widetilde{L_i}$:
+Let $\widetilde{\Lambda}$ be the wave period in samples and $\widetilde{L_i} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ the samples $\widetilde{l_i}$ (refer to Equation~\ref{periodoUnico}), a sequence $T_i^{f,\,\Delta}$ with samples of a sound with frequency $f$ and duration $\Delta$ can be obtained by means of $\widetilde{L_i}$:
 
@@
 Figure~\ref{fig:lut} illustrates the calculation of $\{t_i\}$ sample from $\left\{\,\widetilde{l}_i\,\right\}$ for $f=200Hz$, $\widetilde{\Lambda}=128$ and adopting the sample rate of $f_s=44.1kHz$.
-Though this is not a practical configuration (as assigned below), it makes possible a graphical visualization of the procedure.
+Though this is not a practical configuration (as discussed below), it allows for a graphical visualization of the procedure.
 
 The calculation of the integer $\gamma_i$ introduces noise which decreases as $\widetilde{\Lambda}$ increases.
-In order to use this calculation in sound synthesis, with $f_s=44.1 kHz$, the standard guidelines suggests the use of $\widetilde{\Lambda} = 1024$ samples, since it does not produce relevant noise on the audible spectrum.
-The rounding or interpolation method are not decisive in this process~\cite{Geiger}.
+In order to use this calculation in sound synthesis, with $f_s=44.1 kHz$, the standard guidelines suggest the use of $\widetilde{\Lambda} = 1024$ samples, since it does not produce relevant noise on the audible spectrum.
+The rounding or interpolation method is not decisive in this process~\cite{Geiger}.
 
@@
 There are important considerations here: it is possible to use practically any frequency $f$.
-Limits exist only in low frequencies when the size of the table  $\widetilde{\Lambda}$ is not sufficient for the sample rate $f_a$.
+Limits exist only at low frequencies when the size of table $\widetilde{\Lambda}$ is not sufficient for the sample rate $f_a$.
 The lookup procedure is virtually costless and replaces calculations by simple indexed searches (what is generally understood as an optimization process).
@@
 As stated by the Weber and Fechner~\cite{Weber-Fechner} law, the human perception has a logarithmic relation with the stimulus.
-In other words, the exponential progression of a stimulues is perceived as linear.
+That is to say, the exponential progression of a stimulus is perceived as linear.
 For didactic reasons, and given its use in AM and FM synthesis (subsection~\ref{subsec:tvaf}), linear variation is discussed first.
@@
 Therefore, it is handy to calculate the elements $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ by means of the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$.
-
-The equations~\ref{freqLinear},~\ref{indiceLinear} and~\ref{serieAmostralLin} are related with the linear progression of the frequency.
+Equations~\ref{freqLinear},~\ref{indiceLinear} and~\ref{serieAmostralLin} are related with the linear progression of the frequency.
 As stated above, the frequency progression \emph{perceived} as linear follows an exponential progression, i.e.
 a geometric progression of frequency is perceived as an arithmetic progression of pitch.
-It is possible to write: $f_i=f_0 . 2^{\frac{i}{\Lambda-1} n_8}$ where  $n_8=\log_2\frac{f_{\Lambda-1}}{f_0}$ is the number of octaves between $f_0$ and $f_{\Lambda-1}$.
+It is possible to write: $f_i=f_0 .
+2^{\frac{i}{\Lambda-1} n_8}$ where  $n_8=\log_2\frac{f_{\Lambda-1}}{f_0}$ is the number of octaves between $f_0$ and $f_{\Lambda-1}$.
 Therefore, $f_i=f_0 . 2^{\frac{i}{\Lambda-1}\log_2\frac{f_{\Lambda-1}}{f_0}}=
@@
      \centering
-         \includegraphics[width=\columnwidth]{transicao}
-     \caption{Intensity transitions to different values of $\alpha$ (see equations~\ref{seqAmp} and ~\ref{transAmp}).}
+         \includegraphics[width=\columnwidth]{figures/transicao}
+     \caption{Intensity transitions for different values of $\alpha$ (see equations~\ref{seqAmp} and ~\ref{transAmp}).}
          \label{fig:transicao}
@@
 
-The term $\frac{i}{\Lambda-1}$ convers the interval $[0,1]$ and it is possible to power it to a value in such a way that the beginning of the transition will be smoother or steeper.
-This procedure is useful for energy variations with the purpose of changing the volume\footnote{The volume (psychophysical quality) is consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is wave total energy.
-Wave energy is the easy to modify (see Equation~\ref{eq:potencia}) and it can vary in many ways.
+The term $\frac{i}{\Lambda-1}$ covers the interval $[0,1]$ and it is possible raise it to a power in such a way that the beginning of the transition will be smoother or steeper.
+This procedure is useful for energy variations with the purpose of changing the volume\footnote{The volume (psychophysical quality) is a consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is the wave total energy.
+Wave energy is easy to modify (see Equation~\ref{eq:potencia}) and can be varied in many ways.
 The simplest way consists of modifying the amplitude by multiplying the whole sequence by a real number.
 The increase of energy without amplitude variation is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}}.
-It is sufficient to multiply the original sequence by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is the given coefficient and $a_{\Lambda-1}$ is a fraction of the original amplitude, which is the value to be reached at the end of transition.
+It is sufficient to multiply the original sequence by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is the given coefficient and $a_{\Lambda-1}$ is a fraction of the original amplitude, which is the value to be reached at the end of the transition.
 
@@
 
-It is often convenient to have $a_0=1$ in order to start a new sequence with the original amplitude and then progressively change it.
-If $\alpha=1$, the amplitude variation follows the geometric progression whose defines the linear variation of volume.
+It is often convenient to have $a_0=1$ to start a new sequence with the original amplitude and then progressively change it.
+If $\alpha=1$, the amplitude variation follows the geometric progression that defines the linear variation of volume.
 Figure~\ref{fig:transicao} depicts transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by equation~\ref{eq:ampVol}.
@@
 In equation~\ref{seqAmp}, $a_0=0$ results in a division by zero and if $a_{\Lambda-1}=0$, there will be multiplication by zero.
-Both cases make the procedure useless, once no number different from zero can be represented as a ratio in relation of zero.
+Both cases make the procedure useless, once no number different from zero can be represented as a ratio in relation to zero.
 It is possible to solve this dilemma choosing a number that is small enough like $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum volume for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$).
@@
 
-Here the convertion between decibels and amplitude is convenient. In this way, the equations~\ref{ampDec} and \ref{transAmp} specify a transition of $V_{dB}$ decibels:
+Here the conversion between decibels and amplitude is convenient, with equations~\ref{ampDec} and \ref{transAmp} specifying a transition of $V_{dB}$ decibels:
 
@@
 The linear transitions will be used for AM and FM synthesis, while logarithmic transitions are proper tremolos and vibratos, as developed in subsection~\ref{subsec:tvaf}.
-A non-oscillatory exploration of these variations is in the music piece \emph{Transita para metro}, which code is online as part of the \massa\ toolbox\cite{MASSA}.
+A non-oscillatory exploration of these variations is in the music piece \emph{Transita para metro}, whose code is online as part of the \massa\ toolbox\cite{MASSA}.
 
@@
 
-This subsection is limited to a description of sequences processing, by convolution and differential equations, and immediate applications, as its complexity escapes the scope of this study\footnote{The implementation of filters comprehends an area of recognized complexity, with dedicated literature and software implementations.
+This subsection is limited to a description of sequences processing, by convolution and differential equations, and immediate applications, as its complexity escapes the scope of this study\footnote{The implementation of filters encompasses an area of recognized complexity, with dedicated literature and software implementations.
 The reader is welcome to visit the bibliography for further discussion~\cite{Openheim,smith}.}.
-Filter applications can be part of the synthesis process or made subsequently as part of processes commonly called ``sound treatment''.
+Filter applications can be part of the synthesis process or made subsequently as part of processes commonly referred to as ``sound treatment''.
+
 \subsubsection{Convolution and finite impulse response (FIR) filters}
@@
      \centering
-         \includegraphics[width=\textwidth]{convolucao}
+         \includegraphics[width=\textwidth]{figures/convolucao}
      \caption{Graphical interpretation of convolution. Each resulting sample is the sum of the previous samples of a signal, with each one multiplied by the retrograde of the other sequence.}
@@
 FIR filters are applied in the time domain of digital sound by means of convolution with the respective impulse response of the filter\footnote{It is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both sound and the impulse response, and then performing the inverse Fourier transform in the resulting spectrum.\cite{Openheim}}.
-For the purposes of this work, the convolution is defined as:
+For the purposes of this work, convolution is defined as:
 
@@
 Both are impulse responses which, properly convoluted with the sound sequence, result in the own sound with a reverberation that resembles the original environment where the measure was made~\cite{Cook}.
-
 The inverse Fourier transform of an even and real envelope is an impulse response of a FIR filter.
 Convoluted with a sound, it performs the frequency filtering specified by the envelope.
-The greater the number of samples, the higher the envelope resolution and the computational complexity, which should often be weighted, for the convolution is expensive.
+The greater the number of samples, the higher the envelope resolution and the computational complexity, which should often be weighted, for convolution is expensive.
 
 An important property is the time shift caused by convolution with a shifted impulse.
-Despite computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse for each reincidence of the sound.
-In Figure~\ref{fig:delays} it is possible to observe the shift caused by convolution with an impulse.
+Despite being computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse for each reincidence of the sound.
+Figure~\ref{fig:delays} shows the shift caused by convolution with an impulse.
 Depending on the intensity of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as pitch (from about 20 impulses per second and higher frequencies).
@@
     \centering
-        \includegraphics[width=\textwidth]{delays}
-    \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c). Disposed in increasing order of its pulse density.}
+        \includegraphics[width=\textwidth]{figures/delays}
+    \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c). Represented in increasing order of its pulse density.}
         \label{fig:delays}
@@
 
-This class of filters, known by the acronym IIR, are characterized by having an infinite time representation, i.e.\ the impulse response does not converge to zero.
+This class of filters, known by the acronym IIR, is characterized by having an infinite time representation, i.e.\ the impulse response does not converge to zero.
 Its application is usually made by the following equation:
@@
 
-In the most cases it is possible to normalize the variables: $a_j'=\frac{a_j}{b_0}$ and $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
+In most cases the variables may be normalized: $a_j'=\frac{a_j}{b_0}$ and $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
 Equation~\ref{eq:diferencas} is called `difference equation' because the resulting samples $\left\{t_i'\right\}$ are given by differences between original samples $\{t_i\}$ and previous resulting ones $\left\{t_{i-k}'\right\}$.
@@
 There are many methods and tools to obtain IIR filters.
-The text bellow lists a selection of them for didactic purposes and as a reference.
+The text below lists a selection for didactic purposes and as a reference.
 They are well behaved filters whose aspects are described in Figure~\ref{fig:iir}.
-
 For filters of simple order, the cutoff frequency $f_c$ is where the filter performs an attenuation of $-3dB \approx 0.707 $ of the original amplitude.
@@
 z^{-2}}$.} or the calculation of Chebichev filter coefficients\footnote{Butterworth and Elliptical filters can be considered as special cases of Chebichev filters~\cite{Openheim,smith}.}.
-Both alternatives are explored by referenced studies, in special~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University~\cite{music-dsp,Openheim}.
+Both alternatives are explored by ~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University~\cite{music-dsp,Openheim}.
 
@@
     \centering
-        \includegraphics[width=\textwidth]{iir}
+        \includegraphics[width=\textwidth]{figures/iir}
     \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of equations~\ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} and \ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
@@
   \item Low-pass with a simple pole, module of the frequency response in the upper left corner of Figure~\ref{fig:iir}.
-The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_s$ in which occurs an attenuation of $3dB$.
-  The coefficients $a_0$ and $b_1$ of the IIR filter are given by means of the intermediate variable $x \in [e^{-\pi},1]$:
+The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_s$ in which an attenuation of $3dB$ occurs.
+The coefficients $a_0$ and $b_1$ of the IIR filter are given by the intermediate variable $x \in [e^{-\pi},1]$:
 
@@
 This filter is parametrized by a center frequency $f_c$ and bandwidth $bw$, both given as fraction of $f_s$, therefore $f,\; bw \in (0,0.5)$.
-Both frequencies $f_c \pm bw$ has $\approx 0.707$ of the amplitude, i.e.
+Both frequencies $f_c \pm bw$ have $\approx 0.707$ of the amplitude, i.e.
 attenuation of $3dB$.
-Auxiliary variables $K$ and $R$ are defined as:
+The auxiliary variables $K$ and $R$ are defined as:
 
@@
 
-Generally, sounds without an explicit pitch are called noise~\cite{Lacerda}.
-They are important musical sounds, as noise present in piano notes, violin, etc.
-Besides that, the majority of percussion instruments does not exhibit an unequivocal pitch and their sounds are generally regarded as noise~\cite{Roederer}.
+Sounds without an explicit pitch are generally called noise~\cite{Lacerda}.
+They are important musical sounds, as noise is present in piano notes, violin, etc.
+Furthermore, the majority of percussion instruments does not exhibit an unequivocal pitch and their sounds are generally regarded as noise~\cite{Roederer}.
 In electronic music, including electro-acoustic and dance genres, noise has diverse uses and frequently characterizes the music style~\cite{Cook}.
@@
 The use of random values to generate the sound sequence $T_i$ is an attractive method but not outstandingly useful because it tends to produce white noise with little or no variations~\cite{Cook}.
-
-Another way is to generate noise is by using the desired spectrum, from which it is possible to perform the inverse Fourier transform.
-The spectral distribution should be done with care: if phases of components present proeminent correlation, the synthesized sound will concentrate energy in some periods of its duration.
+Another possibility to generate noise is by using the desired spectrum, from which it is possible to perform the inverse Fourier transform.
+The spectral distribution should be done with care: if phases of components exhibit prominent correlation, the synthesized sound will concentrate energy in some periods of its duration.
 
@@
      \centering
-         \includegraphics[width=\textwidth]{ruidos}
+         \includegraphics[width=\textwidth]{figures/ruidos}
      \caption{Colors of noise generated by equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul} and \ref{eq:violeta}: spectrum and waveforms.}
@@
 
-Some noise with static spectrum are listed below. They are called \emph{colored noise} since they are associated with colors.
+Some noises with static spectrum are listed below. They are called \emph{colored noise} since they are associated with colors.
 Figure~\ref{fig:ruidos} shows the spectrum profile and the corresponding sonic sequence side-by-side.
@@
 
-The exponential $e^{j.x}$, is an artifice to obtain unitary module and random phase for the value of $c_i$. Besides that, $c_{\Lambda/2}$ is always real (as discussed in the previous section).
-
- \item The pink noise is charachterized by a decrease of $3dB$ per octave. This noise is useful for testing electronic devices, and it has prominent presence in nature~\cite{Roederer}.
+The exponential $e^{j.x}$ is a way to obtain unitary module and random phase for the value of $c_i$. In addition, $c_{\Lambda/2}$ is always real (as discussed in the previous section).
+
+ \item The pink noise is characterized by a decrease of $3dB$ per octave. This noise is useful for testing electronic devices, being prominent in nature~\cite{Roederer}.
 
@@
 
-The minimum frequency $f_{\text{min}}$ is chosen with regard to the human hearing, since a sound component with frequency bellow $\approx\; 20Hz$ is virtually inaudible.
-
-Other noises can be made by similar procedures. Simple modifications are needed, specially in the equation that defines $\alpha_i$.
+The minimum frequency $f_{\text{min}}$ is chosen with regard to the human hearing, since a sound component with frequency below $\approx\; 20Hz$ is virtually inaudible.
+
+Other noises can be made by similar procedures. Simple modifications are needed, especially in the equation that defines $\alpha_i$.
 
   \item The brown noise received this name after Robert Brown, who described the Brownian movement\footnote{Although its origin is disparate with its color association, this noise became established with this specific name, in musical contexts.
-Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more intense colors~\cite{Cook,guillaume}.}
-
-What characterizes this noise is the decrease of $6dB$ per octave. In this way, $\alpha_i$ in the equation set~\ref{eq:rosa} is:
+Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more intense colors~\cite{Cook,guillaume}.} What characterizes brown noise is the decrease of $6dB$ per octave, with $\alpha_i$ in equation set~\ref{eq:rosa} being:
 
@@
 
-This subsection exposed only noise with static spectrum.
+This subsection discussed only noise with static spectrum.
 There are also characterizations for noise with dynamic spectrum during the time, and noises which are fundamentally transient, like clicks and chirps.
-The former are easily modeled by an impulse relatively isolated, while chirps are not in fact a noises, but a fast scan of some given frequency band~\cite{Cook}.
+The former are easily modeled by an impulse relatively isolated, while chirps are not in fact a noise, but a fast scan of some given frequency band~\cite{Cook}.
 
@@
 \ref{eq:azul} and \ref{eq:violeta} are presented in Figure~\ref{fig:ruidos}.
-As stated above, the spectra were built with the same phase and frequency for each coefficient, making it straightforward to observe the contribution of treble harmonics and bass frequencies.
+The spectra were built with the same phase and frequency for each coefficient, making it straightforward to observe the contribution of treble harmonics and bass frequencies.
 
@@
 
-Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in volume (intensity)\footnote{Some musical instruments jargon and contexts use different terms.
-For example, in piano, the called tremolo is a vibrato in the classification used here.
-The presented definitions are common in contexts regarding music theory and electronic music.
+Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in volume (intensity)\footnote{The jargon may be different in other contexts.
+For example, in piano, tremolo is a vibrato in the classification used here.
+The definitions given here are common in contexts regarding music theory and electronic music.
 In addition, they are based on a broader literature than the one used for a specific instrument, practice or musical tradition~\cite{Lacerda,Harmonia}.}
-For the general case, the vibrato is described as follows:
+For the general case, vibrato is described as:
 
@@
      \centering
-         \includegraphics[width=\columnwidth]{vibrato}
-     \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave,  considered $f_s=44.1kHz$.}
+         \includegraphics[width=\columnwidth]{figures/vibrato}
+     \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave, with $f_s=44.1kHz$.}
          \label{fig:vibrato}
@@
 \end{itemize}
-It is convenient to use $\nu=\log_{2}\frac{f+\mu}{f} $ in this case because the maximum frequency increase is not equivalent to the maximum frequency diminish, but the semitone variation remains.
+
+It is convenient to use $\nu=\log_{2}\frac{f+\mu}{f} $ in this case because the maximum frequency increase is not equivalent to the maximum frequency decrease, but the semitone variation remains.
 
@@
 Practically any waveform can be used to generate a sound and the vibrato oscillation pattern, with virtually any oscillation frequency and pitch deviation\footnote{The pitch deviation is called 'vibrato depth' and is generally given as semitones or cents, as convenient}.
-Those oscillations with precise waveforms and arbitrary amplitudes are not possible in traditional music instruments and, in this way, it introduces novelty in the artistic possibilities.
+Those oscillations with precise waveforms and arbitrary amplitudes are not possible in traditional music instruments, and thus it introduces novelty in the artistic possibilities.
 
 Tremolo is similar: $f'$, $\gamma_i'$ and $t_i'$ remains the same.
-The amplitude sequence to be multiplied by the original sequence $t_i$ turns in:
+The amplitude sequence to be multiplied by the original sequence $t_i$ is:
 
@@
 \noindent where $V_{dB}$ is the oscillation depth in decibels of tremolo and $a_{\text{max}}=10^{\frac{V_{dB}}{20}}$ is the maximum amplitude gain.
-The measurement in decibels is pertinent because the maximum increase in amplitude is not equivalent to the related maximum decrease, while the difference in decibels remains.
+The measurement in decibels is suitable because the maximum increase in amplitude is not equivalent to the related maximum decrease, while the difference in decibels remains.
 
@@
 The curvature is due to the logarithmic progression of the intensity.
-The tremolo frequency is $1,5Hz$ because $f_a=44,1kHz \; \Rightarrow \; \text{duration} = \frac{i_{\text{max}}=82000}{f_a}= 2s \; \Rightarrow \; \frac{3\text{oscillations}}{2s}=1,5$ oscillations per second ($Hz$).
+The tremolo frequency is $1.5Hz$ because $f_a=44.1kHz \; \Rightarrow \; \text{duration} = \frac{i_{\text{max}}=82000}{f_a}= 2s \; \Rightarrow \; \frac{3\text{oscillations}}{2s}=1.5$ oscillations per second ($Hz$).
 
 The music piece \emph{Vibra e treme} explores these possibilities given by tremolos and vibratos, both used in conjunction and independently, with frequencies $f'$, different depths ($\nu$ and $V_{dB}$), and progressive parameters variations (tremolos and vibratos occur many times together in a traditional music instrument and voices).
-Aiming a qualitative appreciation, the piece also develops a comparison between vibratos and tremolos in logarithmic and linear scales.
+Aiming at a qualitative appreciation, the piece also develops a comparison between vibratos and tremolos in logarithmic and linear scales.
 Its source code is available online as part of the \massa\ toolbox.
@@
      \centering
-         \includegraphics[width=\textwidth]{tremolo}
-     \caption{Tremolo with a depth of $V_{dB}=12dB$, with a sawtooth waveform as its oscillatory pattern, with $f'=1.5Hz$ in a sine of $f=40Hz$ (considered a sample frequency of $f_s=44.1kHz$).}
+         \includegraphics[width=\textwidth]{figures/tremolo}
+     \caption{Tremolo with a depth of $V_{dB}=12dB$, with a sawtooth waveform as its oscillatory pattern, with $f'=1.5Hz$ in a sine of $f=40Hz$ (sample frequency $f_s=44.1kHz$).}
          \label{fig:tremolo}
@@
 
-The proximity of $f'$ to $20Hz$ generates rugosity in both tremolos and vibratos. This rugosity is largely appreciated both in
-traditional classical music and current electronic music, specially in the \emph{Dubstep} genre.
-Rugosity is also generated by spectral content that produce beating~\cite{Porres,porres2009}.
-The sequence \emph{Bela Rugosi} explores this rugosity threshold with concomitant tremolos and vibratos at the same
-voice, with different intensity and waveforms. The respective code is
-available online in the \massa\ toolbox.
-
-As frequency increases further, these oscillations no longer remains
-noticeable individually. In this case, the oscillations are audible as pitch. Then, $f'$,
-$\mu$ and the waveform together, changes the spectrum of original sound $T_i$
-in different ways for both tremolos and vibratos.  They are called AM
-(\emph{Amplitude Modulation}) and FM (\emph{Frequency Modulation}) synthesis,
+The proximity of $f'$ to $20Hz$ generates roughness in both tremolos and vibratos.
+This roughness is largely appreciated both in traditional classical music and current electronic music, especially in the \emph{Dubstep} genre.
+Roughness is also generated by spectral content that produces beating~\cite{Porres,porres2009}.
+The sequence \emph{Bela Rugosi} explores this roughness threshold with concomitant tremolos and vibratos at the same voice, with different intensity and waveforms.
+The corresponding code is available online in the \massa\ toolbox.
+
+As the frequency increases further, these oscillations no longer remain noticeable individually.
+In this case, the oscillations are audible as pitch.
+Then, $f'$, $\mu$ and the waveform together change the spectrum of original sound $T_i$ in different ways for both tremolos and vibratos.
+They are called AM (\emph{Amplitude Modulation}) and FM (\emph{Frequency Modulation}) synthesis,
 respectively. These techniques are well known, with applications in
-synthesizers like \emph{Yamaha DX7}, and even with applications outside music, like
-in telecommunications, for data transfer by means of electromagnetic waves (e.g.\ AM and FM radios).
+synthesizers like \emph{Yamaha DX7}, and even with applications outside music, as in telecommunications for data transfer by means of electromagnetic waves (e.g.\ AM and FM radios).
 
 For musical goals, it is possible to understand FM based on the
-case of sines and, in occurrences of greater complexity, to decompose the signals into their respective Fourier components (i.e.\ sines as well). In this way, the FM
-synthesis performed with a sinusoidal vibrato with frequency $f'$ and depth
-$\mu$ in a sinusoidal sound $T_i$ with frequency $f$ generates bands
-centered in $f$ and far from each other with a distance of $f'$:
+case of sines and, in occurrences of greater complexity, to decompose the signals into their respective Fourier components (i.e.\ sines as well).
+The FM synthesis performed with a sinusoidal vibrato with frequency $f'$ and depth $\mu$ in a sinusoidal sound $T_i$ with frequency $f$ generates bands centered in $f$ and far from each other with a distance of $f'$:
 
@@
 
-\noindent is the Bessel function~\cite{BesselCCRMA,JOSFM} which specifies the
-amplitude of each component in FM synthesis.
-
-On these equations, the frequency variation introduced by $\{t_i'\}$ does not
-respect the geometric progression that yield linear pitch variation, but reflects
-equation~\ref{freqLinear}.
-The use of equations~\ref{vbrF} for FM are described in~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
-In
-fact, the simple and attractive FM behavior are observed only with linear variations, such as in~\ref{eq:fmEsp}).
+\noindent is the Bessel function~\cite{BesselCCRMA,JOSFM} which specifies the amplitude of each component in FM synthesis.
+
+In these equations, the frequency variation introduced by $\{t_i'\}$ does not follow the geometric progression that yields linear pitch variation, but reflects equation~\ref{freqLinear}.
+The use of equations~\ref{vbrF} for FM is described in~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
+In fact, the simple and attractive FM behavior are observed only with linear variations, such as in~\ref{eq:fmEsp}).
 
@@
 The resulting sound is the original one together with the
-reproduction of its spectral content below and above the original frequency,
-with the distance $f'$ from $f$. Again, this is obtained by variations in the
-linear scale of the amplitude. An exposition of the
-spectrum of an AM performed with oscillations in the logarithmic amplitude scale is again in~\cite{dissertacao}, and it again loses the simple and attractive behavior.
-
-The sequence $T_i$, with frequency $f$, called `carrier', is modulated by
+reproduction of its spectral content below and above the original frequency, with the distance $f'$ from $f$.
+Again, this is obtained by variations in the linear scale of the amplitude.
+A discussion of the
+spectrum of an AM performed with oscillations in the logarithmic amplitude scale is available in~\cite{dissertacao}. The sequence $T_i$, with frequency $f$, called `carrier', is modulated by
 $f'$, called 'modulator'. In FM and AM jargon, $\mu$ and
-$\alpha=10^{\frac{V_{dB}}{20}}$ are `modulation indexes'. The following
-equations are defined for the vibration pattern of the modulator sequence
-$\{t_i'\}$:
+$\alpha=10^{\frac{V_{dB}}{20}}$ are `modulation indexes'. The following equations are defined for the vibration pattern of the modulator sequence $\{t_i'\}$:
 
@@
 
-\noindent where $\widetilde{l}$ is the waveform period with a length of
-$\widetilde{\Lambda}$ samples, used for the carrier signal.
-
-To perform AM, it is needed to modulate the signal $\{t_i\}$ with $\{t_i'\}$ using
-the following equations:
+\noindent where $\widetilde{l}$ is the waveform period with a length of $\widetilde{\Lambda}$ samples, used for the carrier signal.
+
+To perform AM, the signal $\{t_i\}$ needs to be modulated with $\{t_i'\}$ using the following equations:
 
@@
 
-At this point, the musical possibilities exploded.
+At this point, the musical possibilities are very wide.
 All characteristics, like pitch (given by frequency), timbre (given by the waveform, filters and noise), volume (manipulated by intensity) and duration (given by the number of samples), can be considered in an absolute form or treated during duration of sound (clearly, with the exception of duration itself).
-
-The following musical usages comprehend a collection of possibilities
-with the purpose of exemplifying types of sound manipulation that result in
-musical material. Some of them are deeply discussed in the next section.
+The following musical usages comprehend a collection of possibilities with the purpose of exemplifying types of sound manipulation that result in musical material.
+Some of them are discussed in depth in the next section.
 
@@
 An interesting possibility is to use relations between parameters of
-a tremolo or vibrato, and some parameters of the basic note like frequency. In
-this way, it is possible to have a vibrato frequency proportional
+a tremolo or vibrato, and some parameters of the basic note like frequency. It is possible to have a vibrato frequency proportional
 to note pitch, or a tremolo depth inversely proportional to
-pitch. Therefore, with equations \ref{vbrGamma}, \ref{vbrF} and \ref{trA}, it is
-perfectly possible to arbitrate:
+pitch. Therefore, with equations \ref{vbrGamma}, \ref{vbrF} and \ref{trA}, it is possible to choose:
 
@@
 
-\noindent with $f^{vbr}$ and $f^{tr}$ as $f'$ in the referenced equations. They can also be associated with vibrato and tremolo oscillation frequency of
-equation~\ref{vbrGamma}. 
+\noindent with $f^{vbr}$ and $f^{tr}$ as $f'$ in the referenced equations. They can also be associated with vibrato and tremolo oscillation frequency of equation~\ref{vbrGamma}. 
  $\nu$ and $V_{dB}$ are the respective depth values of vibrato and
-tremolo. Functions $func_a$, $func_b$ and $func_c$ are arbitrary and dependent
-on musical intentions. The music piece \emph{Tremolos, vibratos e a
-frequência} explores such bonds and exhibit variations in the oscillation
-waveform with the purpose of building a \emph{musical
-language} (details in the next section). The corresponding code is on
-is also available online as part of the \massa\ toolbox.
+tremolo. Functions $func_a$, $func_b$ and $func_c$ are arbitrary and dependent on musical intentions. The music piece \emph{Tremolos, vibratos e a
+frequência} explores such bonds and exhibits variations in the oscillation waveform with the purpose of building a \emph{musical
+language} (details in the next section). The corresponding code is also available online as part of the \massa\ toolbox.
 
 \subsubsection{Convolution for rhythm and meter}
-It is possible to use a musical pulse -- such as specified by a BPM
-tempo -- and to place an impulse at the start of each beat, with the purpose of
-establishing metric and rhythms: the
-convolution with an impulse shifts the sound 
-to impulse position.
-For example, two impulses equally spaced builds a binary division into the
-pulse. Two signals, one with 2 impulses and the other with 3 impulses, both 
-equally spaced in the pulse duration, yields a pulse
-maintenance with a rhythmic in which eases both binary or ternary
-divisions of the pulse, and this is found in many ethnic and traditional musical styles~\cite{Gramani}. 
-Absolute values of the impulses implyes in proportions among the
-amplitudes of the sonic reincidences.
-The use of convolution with impulses in this context is explored in the
-music piece \emph{Trenzinho de caipiras impulsivos}. These features
-embrace the creation of `sound amalgams' based on granular synthesis, for now refer specially to the
-Figure~\ref{fig:pulsoSubAgl}. This piece is a link to next section contents. 
+A musical pulse - such as specified by a BPM
+tempo - can be used with an impulse at the start of each beat, with the purpose of establishing metric and rhythms: the convolution with an impulse shifts the sound to impulse position.
+For example, two impulses equally spaced build a binary division into the
+pulse. Two signals, one with 2 impulses and the other with 3 impulses, both equally spaced in the pulse duration, yields a pulse
+maintenance with a rhythm in which eases both binary or ternary
+divisions of the pulse.
+This is found in many ethnic and traditional musical styles~\cite{Gramani}.
+Absolute values of the impulses implies in proportions among the amplitudes of the sonic reincidences.
+The use of convolution with impulses in this context is explored in the music piece \emph{Trenzinho de caipiras impulsivos}.
+These features embrace the creation of `sound amalgams' based on granular synthesis, refer to Figure~\ref{fig:pulsoSubAgl}.
+This piece is a link to next section contents.
 The source code of the music piece is online, as part of the \massa\ toolbox\cite{MASSA}.
@@
 
-Retaken the exposition in subsection~\ref{subsec:spac}, when an audio source
-or receptor is moving, its characteristics are ideally updated at each
-sample of the digital signal. Speed components should be found for each ear.
-In this way, given the audio source speed (or velocity) $v_s$,
-with positive values if the source moves away from receptor, and receptor speed $v_r$,
-positive when it gets closer to audio source, the frequency is given
-by the well known Doppler shift:
+Following the exposition in Subsection~\ref{subsec:spac}, when an audio source or receptor is moving, its characteristics are ideally updated at each sample of the digital signal.
+Speed components should be found for each ear.
+In this way, given the audio source speed (or velocity) $v_s$, with positive values if the source moves away from receptor, and receptor speed $v_r$, positive when it gets closer to audio source, the frequency is given by the well-known Doppler shift:
 
@@
 With both frequencies $f$ and $f_0$, and the IID from the new source
-position, it is possible to create the Doppler effect. There is an addendum to
-improve the fidelity of the physical phenomena: to increase the received
-potency. It is possible to understand this potency gain as being proportional 
-to the wave shrinking or expansion: $\Delta P=P_0\left(\frac{v_r-v_s}{343.2}\right)$,
- where $P_0$ is signal potency and $P$ the potency at receptor.
-
-In this way, it is possible to obtain both amplitude and frequency of a moving
-audio source. Being this audio source in front of receptor with $y_0$ meters of horizontal distance and $z_0$ meters of height, the distance is given by
+position, it is possible to create the Doppler effect.
+There is an addendum to improve the fidelity of the physical phenomena: to increase the received power, which may be understood as being proportional to the wave shrinking or expansion: $\Delta P=P_0\left(\frac{v_r-v_s}{343.2}\right)$,
+ where $P_0$ is signal power and $P$ the potency at receptor.
+Both amplitude and frequency of a moving audio source can be obtained.
+If this audio source is in front of the receptor with $y_0$ m of horizontal distance and $z_0$ m of height, the distance is given by
 $D_i=\left\{ d_i=\sqrt{ y_{i}^{2}+z_{0}^{2} } \right\}_0^{\Lambda-1}$,
 where $y_i=y_0+(v_s-v_r)\frac{i}{f_s}$ with $v_s$ and $v_r$ both
-horizontal, having null non-$y$ components. Amplitude changes with the distance and with the potency factor cited
-above (see subsection~\ref{subsec:volume} for potency to amplitude conversion).
+horizontal, having null non-$y$ components. Amplitude changes with the distance and with the potency factor mentioned above (see subsection~\ref{subsec:volume} for potency to amplitude conversion).
 
@@
 In other words, the same semitones (or fractions) added during the approach are decreased during the departure.
-Besides that, the transition is abrupt if source and receptor intersect with zero distance, otherwise, there is a monotonic progression.
-In the given case, where there is a static height $z_0$, observed the speed component in the direction given by observer and source positions, the frequencies $F_i$ at the observer is given by:
+Moreover, the transition is abrupt if source and receptor intersect with zero distance, otherwise, there is a monotonic progression.
+In the given case, where there is a static height $z_0$, the speed component in the direction given by the observer and source positions, the frequencies $F_i$ at the observer is given by:
 
@@
 
-There is a Python implementation of the Doppler
-effect as describe above, and also considering the intersection between audio source and receptor,
-at the \massa\ toolbox~\cite{MASSA}.
+There is a Python implementation of the Doppler effect, also considering the intersection between audio source and receptor, in \massa\ toolbox~\cite{MASSA}.
 
@@
 With the use of filters, the possibilities are even wider.
-Convolve a signal to have a reverberated version of it, to remove its noise, to
-distort or to treat the audio aesthetically in other ways. For
-example, it is possible to simulate sounds as if originated in an old television or telefone,
-by the use of a band-pass filter, allowing only frequencies between $1kHz$ and $3kHz$.
-By rejecting the frequency of electric oscillation (usually $50Hz$ or $60Hz$) and harmonics, one can remove noises
-caused by audio devices connected to a usual power suply. A more musical application is to perform filtering in
-specific bands and to use those bands as an additional parameter to the notes.
+Convolve a signal to have a reverberated version of it, to remove its noise, to distort or to treat the audio aesthetically in other ways.
+For example, sounds originated from an old television or telephone can be simulated with a band-pass filter, allowing only frequencies between $1kHz$ and $3kHz$.
+By rejecting the frequency of electric oscillation (usually $50Hz$ or $60Hz$) and harmonics, one can remove noises caused by audio devices connected to a usual power suply.
+A more musical application is to perform filtering in specific bands and to use those bands as an additional parameter to the notes.
 
 Inspired by traditional music instruments, it is possible to apply a
-time-dependent filter~\cite{Roederer}. Chains of these filters can perform
-complex and more accurate filtering routines. The music piece \emph{Ruidosa
-faixa} explores filters and many kinds and noise
-synthesis. The source code is available
-online as part of \massa\ toolbox.
-
-A sound can be altered through different filtering and then mixed to create an effect known
-as \emph{chorus}. Based on what happens in a choir of singers,
-the sound is performed using small and potentially arbitrary
-modifications of parameters like center frequency, presence (or absence) of
-vibrato or tremolo and its characteristics, equalization, volume, etc. As a
-final result, those versions of the original sound are mixed together (see
-equation~\ref{eq:mixagem}). The music piece \emph{Chorus infantil} implements a
-chorus in many ways with different sounds and its source code is also available in the \massa\ toolbox.
+time-dependent filter~\cite{Roederer}.
+Chains of these filters can perform complex and more accurate filtering routines.
+The music piece \emph{Ruidosa faixa} explores filters and many kinds and noise synthesis.
+The source code is available online as part of \massa\ toolbox.
+
+A sound can be altered through different filtering and then mixed to create an effect known as \emph{chorus}.
+Based on what happens in a choir of singers, the sound is performed using small and potentially arbitrary modifications of parameters like center frequency, presence (or absence) of vibrato or tremolo and its characteristics, equalization, volume, etc.
+As a final result, those versions of the original sound are mixed together (see equation~\ref{eq:mixagem}).
+The music piece \emph{Chorus infantil} implements a chorus in many ways with different sounds.
+Its source code is available in \massa\ toolbox.
 
@@
 Using the same terms of subsection~\ref{subsec:spac}, the
-late reverberation can be achieved by a convolution with a section of pink, brown
-or black noise, with exponential decay of amplitude along time.
-Delay lines can be added as a prefix to the noise
-with the decay, and this contemplates both time parts of the reverberation: the
-first reflections and the late reverberation. It is possible to improve the
-quality by calculating the geometric trajectory and (mainly low-pass, subsection~\ref{subsec:filtros}) filtering by each surface where
-wavefront reflected before reaching the ear in the first $100-200$ milliseconds.
+late reverberation can be achieved by a convolution with a section of pink, brown or black noise, with exponential decay of amplitude along time. Delay lines can be added as a prefix to the noise
+with the decay, and this contemplates both time parts of the reverberation: the first reflections and the late reverberation.
+Quality can be improved by varying the geometric trajectory and (mainly low-pass, subsection~\ref{subsec:filtros}) filtering by each surface where wavefront reflected before reaching the ear in the first $100-200$ ms.
 The colored noise can be gradually introduced with a \emph{fade-in}: the initial moment given by direct
-incidence of sound (i.e.\ without any reflection and given by ITD and IID),
-reaching its maximum at the beginning of the 'late
-reverberation', when the geometric incidences loses their relevance to the
-statistical properties of noise decay.
-
+incidence of sound (i.e.\ without any reflection and given by ITD and IID), reaching its maximum at the beginning of the 'late
+reverberation', when the geometric incidences lose their relevance to the statistical properties of noise decay.
 As an example, consider $\Delta_1$ as the duration of the first section and $\Delta_R$ as the duration of total reverberation ($\Lambda_1=\Delta_1 f_s$, $\Lambda_R=\Delta_R
-f_s$). Be $p_i$ the probability of a sound to repeat in the
-$i$-th sample. Amplitude falls with exponential decay. Following
-subsection~\ref{subsec:spac}, the reverberation $R_i^1$ of the first period can
-be described as:
+f_s$). Let $p_i$ be the probability of a sound to repeat in the
+$i$-th sample. The amplitude decreases exponentially. Following
+subsection~\ref{subsec:spac}, reverberation $R_i^1$ of the first period can be described as:
 
@@
 
-\noindent where $V_{dB}$ is the total decay in decibels, typically $-80dB$ or
-        $-120dB$. Reverberation $R_i^2$ of the second period can be emulated by a
-        brown noise $R_i^m$ (or by a pink noise $R_i^r$) with exponential amplitude decay of the waveform:
+\noindent where $V_{dB}$ is the total decay in decibels, typically $-80dB$ or $-120dB$.
+Reverberation $R_i^2$ of the second period can be emulated by a brown noise $R_i^m$ (or by a pink noise $R_i^r$) with exponential amplitude decay of the waveform:
 
@@
 
-Considered:
+With
 
@@
 
-\noindent, a reverberated auralization of a sound can be achieved by simple
-        convolution of $R_i$ (called reverberation 'impulse response') with the sound sequence $T_i$ as described in subsection~\ref{subsec:filtros}.
-
-Reverberation is well known by causing great interest in the listeners and to provide more enjoyable sonorities.
-Besides that, modifications in the reverberation space consists on a common resource (almost a \textit{clich\'{e}}) to surprise and attract the listener.
-An implementation of the reverb recipe described is on the \massa\ toolbox~\cite{MASSA}
+\noindent, a reverberated auralization of a sound can be achieved by simple convolution of $R_i$ (called reverberation `impulse response') with the sound sequence $T_i$ as described in subsection~\ref{subsec:filtros}.
+Reverberation is well known for causing great interest in listeners and to provide more enjoyable sonorities.
+Furthermore, modifications in the reverberation space consists on a common resource (almost a \textit{clich\'{e}}) to surprise and attract the listener.
+An implementation of the reverb recipe described here is in \massa\ toolbox~\cite{MASSA}
 
@@
 
-The volume variation along the duration of sound is crucial to our timbre
-perception. The volume envelope known as ADSR (\emph{Attack-Decay-Sustain-Release}), 
-has many implementations in both hardware and software synthesizers. 
-A pioneer implementation can be found in the Hammond
-Novachord synthesizer of 1938 and some variants are cited below~\cite{ADSR}.
-
-The scholastic ADSR envelope is characterized by 4 parameters: attack duration
-(time in which sound reaches its maximum amplitude), decay duration (follows the attack
-immediately), level of sustain volume (in which the volume remains stable after the decay) and release duration (after sustain section, this is the duration needed for amplitude to reach zero or final value).
+The variation of volume along the duration of sound is crucial to our timbre perception.
+The volume envelope known as ADSR (\emph{Attack-Decay-Sustain-Release}), has many implementations in both hardware and software synthesizers.
+A pioneering implementation can be found in the Hammond Novachord synthesizer of 1938 and some variants are mentioned below~\cite{ADSR}.
+The scholastic ADSR envelope is characterized by 4 parameters: attack duration (time in which sound reaches its maximum amplitude), decay duration (follows the attack immediately), level of sustained volume (in which the volume remains stable after the decay) and release duration (after sustained section, this is the duration needed for amplitude to reach zero or final value).
 Note that the sustain duration is not specified because it is the difference between the duration itself and the durations of attack, decay and sustain.
 
-The ADSR envelope with durations $\Delta_A$, $\Delta_D$ and $\Delta_R$, with total duration $\Delta$ and sustain level $a_S$, given by the fraction of the maximum amplitude, to be aplyed to any sound sequence $T_i=\{t_i\}$, can be presented as:
+The ADSR envelope with durations $\Delta_A$, $\Delta_D$ and $\Delta_R$, with total duration $\Delta$ and sustain level $a_S$, given by the fraction of the maximum amplitude, to be applied to any sound sequence $T_i=\{t_i\}$, can be presented as:
 
@@
 
-\noindent with $\Lambda_X=\lfloor \Delta . f_s \rfloor\;\;\forall\;\; X \; \in
-(A,D,R,\;)$ and being $\xi$ a small value that provides a satisfactory \emph{fade in} and \emph{fade out}, e.g.\ $\xi=10^{\frac{-80}{20}}=10^{-4}\;$ or $\;\xi=10^{\frac{-40}{20}}=10^{-2}$.
+\noindent with $\Lambda_X=\lfloor \Delta_X . f_s \rfloor\;\;\forall\;\; X \; \in
+(A,D,R\;)$ with $\xi$ being a small value that provides a satisfactory \emph{fade in} and \emph{fade out}, e.g.\ $\xi=10^{\frac{-80}{20}}=10^{-4}\;$ or $\;\xi=10^{\frac{-40}{20}}=10^{-2}$.
 The lower the $\xi$, the slower the \emph{fade}, like the $\alpha$ illustrated in Figure~\ref{fig:transicao}.
-The stated second in equations~\ref{eq:adsr} attend both introduction and ending of sound from zero intensity because they are linear.
+The right side of Equations~\ref{eq:adsr} can attend both introduction and ending of sound from zero intensity because they are linear.
 Schematically, Figure~\ref{fig:adsr} shows the ADSR envelope in a classical implementation that supports many variations.
@@
 
-\begin{figure}[htpq!]
-    \centering
-        \includegraphics[width=\columnwidth]{adsr}
-    \caption{An ADSR envelope (\emph{Attack, Decay, Sustain, Release}) applied
-        to an arbitrary sound sequence. The linear variation of the amplitude is
-        above, in blue. Below the amplitude variation is exponential.}
+\begin{figure}[htp!]
+    \centering
+        \includegraphics[width=\columnwidth]{figures/adsr}
+    \caption{An ADSR envelope (\emph{Attack, Decay, Sustain, Release}) applied to an arbitrary sound sequence.
+The linear variation of the amplitude is above, in blue.
+Below the amplitude variation is exponential.}
         \label{fig:adsr}
@@
 Consider $ S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1} $, the sequence $S_j$ of $H$ musical events $s_j$.
-Be $S_j$ a `musical structure', composed by events $s_j$ which are musical structures themselves, e.g.\ notes.
+Let $S_j$ be a `musical structure', composed by events $s_j$ which are musical structures themselves, e.g.\ notes.
 This section is dedicated to techniques that make $S_j$ interesting and enjoyable for audition.
@@
 The elements of $S_j$ can be overlapped by mixing them together, as in
-equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, for building intervals
-and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of the events
-in $S_j$, as in equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao},
-builds melodic sequences and rhythms, which are associated with the musical
-`horizontal thought'. The fundamental frequency $f$ and the starting moment
-(attack) are generally the most important characteristics of the elements $s_j$ in
-$S_j$. This makes possible the creation of music made by pitches (both
-harmony and melody) and by the presence of temporal metrics and rhythms.
+equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, for building intervals and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of events
+in $S_j$, as in equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao}, builds melodic sequences and rhythms, which are associated with the musical `horizontal thought'.
+The fundamental frequency $f$ and the starting moment
+(attack) are generally the most important characteristics of the elements $s_j$ in $S_j$.
+This makes it possible to create music made by pitches (both harmony and melody) and by temporal metrics and rhythms.
 
@@
 
-Doubling the frequency is equivalent to ascending one octave ($f=2f_0$).
-The octave division in twelve steps is the canon of classical western music.
-Its usage has also been observed
-outside western tradition, like ceremonial/religious and ethnic context~\cite{Wisnick}.
-In this system, ideally, the factor given by $\varepsilon=2^{\frac{1}{12}}$ defines a semitone.
-A note grid along the hearing spectrum in which, given the frequency $f$, all the possible
-fundamental frequencies are separated by intervals, multiples of
-$\varepsilon$. Twelve semitones (or half steps), equidistant to the human ear, reaches an
-octave, equivalent to human year. Therefore, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
+Doubling the frequency is equivalent to ascending one octave ($f=2f_0$). The octave division in twelve steps is the canon of classical western music. Its usage has also been observed
+outside western tradition, as in ceremonial/religious and ethnic context~\cite{Wisnick}.
+In this system, ideally the factor given by $\varepsilon=2^{\frac{1}{12}}$ defines a semitone.
+This constitutes a note grid along the spectrum in which, given the frequency $f$, all the possible fundamental frequencies are separated by intervals which are multiples of $\varepsilon$.
+Twelve semitones (or half steps), equidistant to the human ear, reach an
+octave, equivalent to the human year. Therefore, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
 $f_0$ and $f$.
 
-This absolute accuracy is common in computational implementations. Real
-music instruments, however, exhibit deviations in order to make compatible
-deviations from the harmonics of notes. Besides that, the fixed reference
-$\varepsilon=2^{\frac{1}{12}}$ characterizes an equally tempered tuning. There
-are tunings in which intervals are ratios of low-order integers, based
-on observations of physical phenomena. These tunings were first formalized
-around two thousand years before the arrival of the equal temperament~\cite{Roederer}.
+This absolute accuracy is common in computational implementations.
+Real music instruments, however, exhibit deviations in order to make compatible deviations from the harmonics of notes.
+In addition, the fixed reference $\varepsilon=2^{\frac{1}{12}}$ characterizes an equally tempered tuning.
+There are tunings in which intervals are ratios of low-order integers, based on observations of physical phenomena.
+These tunings were first formalized around two thousand years before the arrival of the equal temperament~\cite{Roederer}.
 ). Two emblematic tunings are:
@@
 \begin{itemize}
-    \item The {\bf just intonation}, defined by association of diatonic scale notes
-    with ratios of low-order integers, as found in the harmonic
-    series. Considered the ionian mode, the white piano keys
-    from C to C, as seen below in subsection~\ref{subsec:escalas}, i.e. intervals 1, 2M, 3M, 4J, 5J, 6M, 7M, 8,
-    are considered ratios of frequency: 1, 9/8, 5/4
+    \item The {\bf just intonation}, defined by association of diatonic scale notes with ratios of low-order integers, as found in the harmonic series.
+Considered the ionian mode (see Section~\ref{subsec:escalas}), we achieve notes correspondent to the white piano keys from C to C, i.e.
+intervals 1, 2M, 3M, 4J, 5J, 6M, 7M, 8, are considered ratios of frequency: 1, 9/8, 5/4
     4/3, 3/2, 5/3, 15/8, 2/1.
-    The following intervals are also considered: the
-    semitone 16/15, the 'minor tone' 10/9, and the 'major tone' 9/8.
+The following intervals are also considered: the semitone 16/15, the 'minor tone' 10/9, and the 'major tone' 9/8.
 There are many ways to perform the division of the 12 notes in the just intonation.
 
-    \item The {\bf Pythagorean tuning}, based on the interval 3/2 (perfect
-    fifth). The Ionian mode ratios becomes: 1, 9/8, 81/64, 4/3, 3/2, 27/16, 243/128,
-    2/1. These intervals are also considered to yield notes in a scale. Besides the
-    intervals in the mode, also used are the minor second 256/243, the minor
-    third 32/27, the augmented fourth 729/512, the diminished fifth 1024/729,
-    the minor sixth 128/81 and the minor seventh 16/9. 
+    \item The {\bf Pythagorean tuning}, based on the interval 3/2 (perfect fifth).
+The Ionian mode ratios become: 1, 9/8, 81/64, 4/3, 3/2, 27/16, 243/128, 2/1.
+These intervals are also considered to yield notes in a scale.
+Besides the intervals in the mode, also used are the minor second 256/243, the minor third 32/27, the augmented fourth 729/512, the diminished fifth 1024/729, the minor sixth 128/81 and the minor seventh 16/9.
 \end{itemize}
 
-In order to attend for micro-tonality\footnote{Micro-tonality is the use of intervals less
-than one semitone and have ornamental and structuring functionalities in music. The
-division of the octave in $12$ notes is motivated by physical fundaments but also
-a \emph{convention} adopted too by western classical music. Other tunings are
-incident. To point only one example, the traditional Thai music uses an octave
-division in seven notes equally spaced ($\varepsilon=2^{\frac{1}{7}}$),
-which allows intervals that have little in common with the intervals in the
-diatonic scale~\cite{Wisnick}}. It is possible to use non-integer real values as
-factors of $\varepsilon=2^{\frac{1}{12}}$ between frequencies, 
-or to maintain integer values and change $\varepsilon$. For example, 
-a tuning really near of the harmonic series
+In order to account for micro-tonality\footnote{Micro-tonality is the use of intervals less
+than one semitone and has ornamental and structuring functionalities in music.
+The division of the octave in $12$ notes is motivated by physical fundaments but also a \emph{convention} adopted by western classical music.
+Other tunings are incident.
+By way of illustration, the traditional Thai music uses an octave division in seven notes equally spaced ($\varepsilon=2^{\frac{1}{7}}$),
+which allows intervals that have little in common with the intervals in the diatonic scale~\cite{Wisnick}}.
+It is possible to use non-integer real values as factors of $\varepsilon=2^{\frac{1}{12}}$ between frequencies, or to maintain integer values and change $\varepsilon$.
+For example, a tuning really near the harmonic series
 itself is proposed with the equal division of the octave in $53$ notes:
@@
 In this division, notes are related by means of integers, factors of
-$\varepsilon_2$ relating frequencies by a valid interval.
- Note that if $S_i$ is a pitch sequence related by means of
-$\varepsilon_1$, a direct mapping for notes related by $\varepsilon_2$ is 
+$\varepsilon_2$ relating frequencies by a valid interval. Note that if $S_i$ is a pitch sequence related by means of $\varepsilon_1$, a direct mapping for notes related by $\varepsilon_2$ is 
 a new sequence $S_i'=\left\{s_i'\right\}=\left\{
 s_i \frac{\varepsilon_1}{\varepsilon_2}\right\}$. The music piece \emph{Micro
-tom} explores microtonal features and its code
+tom} uses microtonal features and its code
 is part of \massa\ toolbox.
@@
 \subsubsection{Intervals}\label{subsec:intervalos}
-
-Using the ratio $\varepsilon=2^{\frac{1}{12}}$ between note frequencies
-(i.e.\ one semitone) the intervals in the twelve note system can be represented by integers.
-Table~\ref{eq:intervalos} summarizes each interval:
- its traditional nomenclature, consonance and
+Using the ratio $\varepsilon=2^{\frac{1}{12}}$ between note frequencies (i.e.\ one semitone) the intervals in the twelve note system can be represented by integers.
+Table~\ref{eq:intervalos} summarizes each interval: its traditional nomenclature, consonance and
 dissonance properties, and number of semitones.
 
-\begin{table*}[htpq!]
+\begin{table*}[htp!]
 \centering
@@
 Major seconds and minor sevenths are the mild dissonances.
-Perfect fourth is a special case, being a perfect consonance when considered as an inversion of the perfect fifth and a dissonance or a imperfect consonance otherwise.
+Perfect fourth is a special case, as it is a perfect consonance when considered as an inversion of the perfect fifth and a dissonance or an imperfect consonance otherwise.
 The second special case is the tritone (A4 or aug4, d5 or dim5, tri, TT).
 This interval is consonant in some cultures.
-For tonal music, the tritone indicates dominant and seeks urgent resolution in a third or sixth.
-Due to this instability it is considered a dissonant interval.}
+For tonal music, the tritone indicates dominant and seeks urgent resolution in a third or sixth. Due to this instability it is considered a dissonant interval.}
 \begin{tabular}{| c | c | c | }\hline
@@
 \begin{itemize}
-                \item Intervals are seen first by the as number of steps between notes: first
-                (unison), second, third, forth, fifth, sixth, seventh and
-                eighth (octave). The ninth, tenth, eleventh, etc, are compound  
-                intervals made by one or more octaves plus a ``simple'' interval inside the
-                octave, which characterizes the compound interval. The
-                intervals are represented by numeric digits, e.g. 1, 3, 5
-                are a unison, a third and a fifth, respectively.
+                \item Intervals are seen first by the number of steps between notes: first (unison), second, third, forth, fifth, sixth, seventh and eighth (octave).
+The ninth, tenth, eleventh, etc, are compound intervals made by one or more octaves plus a ``simple'' interval inside the octave, which characterizes the compound interval.
+The intervals are represented by numeric digits, e.g.
+1, 3, 5 are a unison, a third and a fifth, respectively.
 
                 \item Qualities of each interval: perfect consonances --
-                i.e.\ unison, forth, fifth and octave -- are 'perfect'. The
-                imperfect consonances -- i.e.\ thirds and sixths -- and
-                dissonances -- i.e.\ seconds and sevenths -- can be major and
-                minor. Exception for the tritone.
-
-                \item The perfect fourth can be a perfect consonance or
-                a dissonance according to the context and theoretical
-                background. As a general rule, it can be considered
-                a consonance except when it resolves into a
-                fifth or a third with movimentation of just one note .
+                i.e.\ unison, forth, fifth and octave -- are 'perfect'.
+The imperfect consonances -- i.e.\ thirds and sixths -- and dissonances -- i.e.\ seconds and sevenths -- can be major and                 minor.
+Except for the tritone.
+
+                \item The perfect fourth can be a perfect consonance or a dissonance according to the context and theoretical background.
+As a general rule, it can be considered a consonance except when it resolves into a fifth or a third with movimentation of just one note.
 
                 \item Tritone is a dissonance in Western music because
-                it characterizes the ``dominant'' chord in the tonal system (see
-                subsection~\ref{subsec:harmonia}) and represents the
-                instability. Some cultures have the interval as a consonance,
-                using it in melodies and chants as a stable interval.
-
-                \item A major interval decreased by one semitone results in a
-                minor interval. A minor interval increased by one semitone
-                results in a major interval.
-
-                \item A perfect interval (unison, perfect forth, perfect
-                fifth, perfect octave), or a major interval (major second M2,
+                it characterizes the ``dominant'' chord in the tonal system (see subsection~\ref{subsec:harmonia}) and represents the    instability.
+Some cultures have the interval as a consonance,                 using it in melodies and chants as a stable interval.
+
+                \item A major interval decreased by one semitone results in a minor interval. A minor interval increased by one semitone results in a major interval.
+
+                \item A perfect interval (unison, perfect forth, perfect fifth, perfect octave), or a major interval (major second M2,
 major third M3, major sixth M6 or major seventh M7), increased
@@
 
-                \item A perfect interval or a minor interval (minor second m2,
-                minor third m3, minor sixth m6 or minor seventh m7), decreased
-                by one semitone results in a diminished interval. The
+                \item A perfect interval or a minor interval (minor second m2, minor third m3, minor sixth m6 or minor seventh m7), decreased by one semitone results in a diminished interval. The
 diminished fifth is also called tritone (dim5 ~ tri ~ TT).
 
-                \item An augmented interval increased by one semitone results
-                 in a `doubly augmented' interval and a diminished interval
-                decreased by one semitone results in a `doubly diminished'
-                interval.
+                \item An augmented interval increased by one semitone results in a `doubly augmented' interval and a diminished interval decreased by one semitone results in a `doubly diminished' interval.
 
@@
                 melodic interval. The order of the notes: first
-                the lowest note or the highest note, results in an ascending
-                 or descending interval, respectively.
+                the lowest note or the highest note, results in an ascending or descending interval, respectively.
 
@@
 
-                \item An interval higher than an octave is called a 'compound
-                interval' and is classified like the interval between the same
-                notes but in the same octave. They are also reached by adding a
-                7 to the interval: P11 is an octave plus a forth ($7 + P4 = P11$),
-                M9 is an octave plus a major second ($7 + M2 = M9$).
+                \item An interval higher than an octave is called a 'compound interval' and is classified like the interval between the same notes but in the same octave.
+They are also reached by adding a
+                7 to the interval: P11 is an octave plus a forth ($7 + P4 = P11$), M9 is an octave plus a major second ($7 + M2 = M9$).
 \end{itemize}
@@
 The augmented/diminished intervals and the doubly augmented/doubly diminished intervals are consequences of the tonal system.
-Scale degrees (subsection~\ref{subsec:escalas}) are in fact different pitches, with
-specific uses and functions.
+Scale degrees (subsection~\ref{subsec:escalas}) are in fact different pitches, with specific uses and functions.
 Henceforth, in a \textit{C flat} major scale, the tonic -- first degree -- is \textit{C flat}, not \textit{B}, and the leading tone -- seventh degree -- is \textit{B flat}, not \textit{A sharp} or \textit{C double flat}.
-In a similar way, the second degree of a scale can be one semitone from first degree, like the leading tone (seventh degree at one ascending semitone from the first degree), where there is a diminished third between the two semitones of the seventh and second scale degrees~\cite{Lacerda}.
-
-The descriptions presented summarizes the traditional theory of the music
-intervals~\cite{Lacerda}. The music piece \emph{Intervalos entre alturas}
-explores these intervals in both independent and related manners. 
-The source code is available online in the \massa\ toolbox~\cite{MASSA}.
+In a similar fashion, the second degree of a scale can be one semitone from first degree, like the leading tone (seventh degree at one ascending semitone from the first degree), where there is a diminished third between the two semitones of the seventh and second scale degrees~\cite{Lacerda}.
+
+The descriptions presented summarize the traditional theory of music intervals~\cite{Lacerda}.
+The music piece \emph{Intervalos entre alturas} explores these intervals in both independent and related manners.
+The source code is available online in \massa\ toolbox~\cite{MASSA}.
 
@@
 
-A scale is an ordered set of pitches. Usually, scales repeat at each octave. 
-The ascending sequence with all notes from the octave division in 12 equal intervals, 
-separated by the ratio $\varepsilon=2^{\frac{1}{12}}$, is known as the chromatic scale with equal temperament.
+A scale is an ordered set of pitches.
+Usually, scales repeat at each octave.
+The ascending sequence with all notes from the octave division in 12 equal intervals, separated by the ratio $\varepsilon=2^{\frac{1}{12}}$, is known as the chromatic scale with equal temperament.
 There are 5 perfectly symmetric divisions of the octave within the chromatic scale.
-These divisions are considered scales by easy and peculiar uses they provide.
+These divisions are considered scales owing to the easy and peculiar uses they provide.
 Considering integers $e_i$ used as powers of the factor $\varepsilon=2^{\frac{1}{12}}$
-to multiply lower note frequency ($f=\varepsilon^{e_i} f_0$),
+to multiply the lower note frequency ($f=\varepsilon^{e_i} f_0$),
 the scales are as follows:
@@
 
-Therefore, the third note of the whole tone scale with $f_0=200Hz$
-is $f_3=\varepsilon^{e_3^t} . f_0 = 2^{\frac{6}{12}} . 200 \approxeq 282.843
-Hz$. These `scales', or patterns, generate stable structures by their intern
-symmetries and can be repeated in an efficient and sustained way.
-Section~\ref{estCic} specially discusses symmetries. 
-The music piece \emph{Cristais} exposes each one of these scales, in both melodic and
-harmonic counterpart and their source code is part of \massa\ toolbox.
+Therefore, the third note of the whole tone scale with $f_0=200Hz$ is $f_3=\varepsilon^{e_3^t}. f_0 = 2^{\frac{6}{12}} . 200 \approxeq 282.843
+Hz$. These `scales', or patterns, generate stable structures by their intern symmetries and can be repeated in an efficient and sustained way. Section~\ref{estCic} discusses symmetries. 
+The music piece \emph{Cristais} uses each one of these scales, in both melodic and harmonic counterpart and their source code is part of \massa\ toolbox.
 
@@
 \noindent They have only major, minor and perfect intervals.
-The unique exception is the tritone that is presented as an augmented forth or a diminished fifth.
-
-Diatonic scales follow a circular pattern of successive intervals \textit{tone,
-tone, semitone, tone, tone, tone, semitone}. Thus, it is possible to write:
+The unique exception is the tritone presented as an augmented forth or a diminished fifth.
+
+Diatonic scales follow a circular pattern of successive intervals \textit{tone, tone, semitone, tone, tone, tone, semitone}. Thus, it is possible to write:
 
@@
 
-\noindent with $\kappa \in \mathbb{N}$. For each mode there is only one value
-for $\kappa \in [0,6]$ by which $\{e_i\}$ matches. 
+\noindent with $\kappa \in \mathbb{N}$. For each mode there is only one value for $\kappa \in [0,6]$ by which $\{e_i\}$ matches. 
 For example, a brief inspection reveals that
@@
 
-The minor scales has two additional forms, named melodic and harmonic:
+The minor scales have two additional forms, named melodic and harmonic:
 
@@
 
-The different ascending and descending contour of the melodic minor scale is required
-by the leading tone (seventh and last degree, separated by 
-one semitone from the octave, enhances tonic polarization) in the ascending
+The different ascending and descending contour of the melodic minor scale is required by the leading tone (seventh and last degree, separated by one semitone from the octave, enhances tonic polarization) in the ascending
  trajectory, which is not necessary in the descending mode, and therefore it recovers the normal form.
-The harmonic scale presents the leading tone but does not avoid the augmented second
-between the sixth and seventh degrees, it does not consider the melodic trajectory~\cite{Harmonia}. 
-Other scales can be represented in a similar form, like the pentatonic and the
-modes of limited transposition by Messiaen~\cite{Messiaen}.
-
-Although not a scale, the harmonic series is often used as such.
-Besides, this point is a convenient moment to put it plain simple:
+The harmonic scale presents the leading tone but does not avoid the augmented second between the sixth and seventh degrees; it does not consider the melodic trajectory~\cite{Harmonia}. 
+Other scales can be represented in a similar form, like the pentatonic and the modes of limited transposition by Messiaen~\cite{Messiaen}.
+
+Although it is not a scale, the harmonic series is often used as such.
+It is a convenient to write the notes related to each harmonic:
 
@@
 
-In this scale, the frequency of $i-th$ note $h_i$ the frequency of $i-th$ harmonic $f=\varepsilon^{h_i} f_0$from the spectra generated by $f_0$.
-Nature exhibits this frequencies within sounds and with all kinds of distortions.
+In this scale, the frequency of the $i-th$ note $h_i$ is the frequency of $i-th$ harmonic $f=\varepsilon^{h_i} f_0$ from the spectra generated by $f_0$.
+Nature exhibits these frequencies within sounds and with all kinds of distortions.
 
@@
 
-The simultaneous occurrence of three or more notes is observed by means of chords.
-Chords are based on triads, specially in tonal music. Triads are built by two successive thirds
-within 3 notes: root, third and fifth. If the lower note of a chord is different from the root, this chord is 
-an inverted chord as oposed to the chord in its root position. A closed position is any in which no chord note 
-fits between two consecutive
+The simultaneous occurrence of three or more notes is observed by means of chords. Chords are based on triads, especially in tonal music. Triads are built by two successive thirds
+within 3 notes: root, third and fifth.
+If the lower note of a chord is different from the root, this chord is an inverted chord as opposed to the chord in its root position.
+A closed position is any in which no chord note fits between two consecutive
 notes~\cite{Lacerda}, any non-closed position is an open position. In closed and fundamental position,
@@
 
-To have another third superimposed to the fifth, is sufficient to add $10$
-as the highest note in order to form a tetrad with minor seventh, or add $11$ in order to form a tetrad with major
+To have another third superimposed to the fifth, it is sufficient to add $10$ as the highest note in order to form a tetrad with minor seventh, or add $11$ in order to form a tetrad with major
 seventh. Inversions and open positions can be obtained with the 
-addition of $\pm 12$ to selected component.
-
-Incomplete triadic chords, with extra notes ('dirty' chords), and non-triadic
+addition of $\pm 12$ to the selected component. Incomplete triadic chords, with extra notes ('dirty' chords), and non-triadic
 are also common.
@@
         \item A fifth confirms the root (fundamental).
-        \item Major or minor thirds, suggests major or minor chord qualities.
-        \item Every tritone, specially if built between a major third and a
-        minor seventh, tends to resolve in a third or a sixth.
-        \item Note duplication is avoided. If duplication is needed, the
-        preferred order is: the root, fifth, third and seventh.
-        \item Note omission is avoided in the triad. If needed, the fifth is
-        first considered for omission, than third and, if needed, the fundamental.
-        \item It is possible to build chords with notes different from triads,
-        specially if they respect a recurrent logic or musical concatenation that
-        justifies these different notes.
-        \item Chords built by successive intervals different of thirds -- like
-        fourths and seconds -- are recurrent in compositions of advanced
-        tonalism or experimental character.
-        \item The repetition of chord successions (or their characteristics)
-        fixes a trajectory and makes possible the
-        introduction of exotic formations without incoherence.
+        \item Major or minor thirds suggest major or minor chord qualities.
+        \item Every tritone, especially if built between a major third and a minor seventh, tends to resolve into a third or a sixth.
+        \item Note duplication is avoided. If duplication is needed, the preferred order is: the root, fifth, third and seventh.
+        \item Note omission is avoided in the triad. If needed, the fifth is first considered for omission, than third and, if needed, the fundamental.
+        \item It is possible to build chords with notes different from triads, particularly if they obey a recurrent logic or musical concatenation that justifies these different notes.
+        \item Chords built by successive intervals different from thirds – like fourths and seconds -- are recurrent in compositions of advanced tonalism or experimental character.
+        \item The repetition of chord successions (or their characteristics) fixes a trajectory and makes it possible to 
+Introduce exotic formations without incoherence.
 \end{itemize}
@@
 
-Omission of basic tonal chaining is the key to obtain modal and atonal harmonies. In
-absence of these minimal tonal structures, the
-harmony is considered modal if the notes match with some diatonic scale (see
-equations~\ref{eq:escalas}) or if notes are presented in a small number.
-If basic tonal chainings are absent and notes does not match any diatonic scale
-and are diverse and dissonant (by relation with each other) enough to avoid
-reduction by polarization, the harmony is atonal. In this classification, the
-modal harmony is not tonal or atonal and is reduced to the 
-incidence of notes within the diatonic scale (or simplifications) and to the
-absence of tonal structures. It is possible to note, following this concept, that
-atonal harmony is hard to be realized and, indeed, no matter how dissonant and diverse
-is a set of notes, tonal harmonies arise if not avoided~\cite{harmEXT}.
+Omission of basic tonal chaining is the key to obtain modal and atonal harmonies. In the absence of these minimal tonal structures, 
+harmony is considered modal if the notes match with some diatonic scale (see equations~\ref{eq:escalas}) or if notes are presented in a small number.
+If basic tonal chainings are absent and notes do not match any diatonic scale and are diverse and dissonant (by relation with each other) enough to avoid reduction by polarization, harmony is atonal.
+In this classification, the modal harmony is not tonal or atonal and is reduced to the incidence of notes within the diatonic scale (or simplifications) and to the absence of tonal structures.
+Following this concept, one observes that atonal harmony is hard to be realized and, indeed, no matter how dissonant and diverse
+a set of notes is, tonal harmonies arise if not avoided~\cite{harmEXT}.
 
@@
 
-In fact, the techniques around atonal music aim at structures for avoiding
-direct relation with modes and tonality. Manifesting such structures
-are of such difficulty that the dodecafonism arouse. 
-The purpose of dodecafonism is to use a set of notes (ideally 12 notes),
-and to perform each note, one by one, in the same
-order. In this context, the tonic becomes difficult to be stablished. Nevertheless, the
-Western hearing searches for tonal traces in music and easily finds them by
-unexpected and tortuous paths.
-
-The use of dissonant intervals (specially tritones) without resolution,
-reinforces the absence of tonality. Considering
-this context, while creating a music
+In fact, the techniques around atonal music aim at structures for avoiding direct relation with modes and tonality.
+Manifesting such structures is of such difficulty that the dodecafonism arouse.
+The purpose of dodecafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
+order.
+In this context, the tonic becomes difficult to be established.
+Nevertheless, the western listener searches for tonal traces in music and easily finds them by unexpected and tortuous paths.
+The use of dissonant intervals (especially tritones) without resolution, reinforces the absence of tonality.
+In this context, while creating a music
 piece, it is possible to:
@@
 \begin{itemize}
-     \item Repeat notes. By considering immediate repetition as an extension
-     of previous incidence, the use of same note in sequence does not add relevant information.
-     \item To play adjacent notes at the same time, making harmonic intervals
-     and chords.
+     \item Repeat notes. By considering immediate repetition as an extension of previous incidence, the use of the same note in sequence does not add relevant information.
+     \item To play adjacent notes at the same time, making harmonic intervals and chords.
      \item Present durations and pauses with liberty, respecting notes order.
@@
      and~\ref{subsec:usosmusicais3} for more details.
-     \item Account for the presence of note structures, it is possible to use variations in orchestration, articulation, spatialization, among others.
+     \item Accounting for the presence of note structures, it is possible to use variations in orchestration, articulation, spatialization, among others.
 \end{itemize}
 
-The atonal harmony can be observed, paradigmatically, within these presented
-conditions (a simple dodecaphonic model). Most of what was written by great dodecafonic composers, 
-like Alban Berg and even Schoenberg, had the purpose of mixing tonal and atonal techniques.
+The atonal harmony can be observed, paradigmatically, within these presented conditions (a simple dodecaphonic model). Most of what was written by great dodecafonic composers, 
+e.g. Alban Berg and even Schoenberg, had the purpose of mixing tonal and atonal techniques.
 
@@
 
-In the XX century, rhythmic music and with emphasis on sonorities/timbres, extended
-the concepts of tonality and harmony. Hence, 
-tonal harmony has strong influence in artistic movements and commercial
-venues. In addition, the dodecafonism itself is considered of tonal nature because it
-denies tonal characteristics of polarization.
-
+In the XX century, rhythmic music and with emphasis on sonorities/timbres, extended the concepts of tonality and harmony.
+Hence, tonal harmony has strong influence on artistic movements and commercial venues.
+In addition, dodecafonism itself is considered of tonal nature because it denies tonal characteristics of polarization.
 In tonal or modal music, chords -- like the ones listed in
-equations~\ref{triades}-- built on the root note of each degree of a scale  -- such 
-as listed in equations~\ref{eq:escalas} --  form the pilars of the harmonic field. 
+equations~\ref{triades}-- built on the root note of each degree of a scale  -- such as listed in equations~\ref{eq:escalas} --  form the pilars of the harmonic field.
 Music harmony aims at studying the incidence of chord progressions and chaining rules.
@@
 In the traditional tonal music, a scale has its tonic (first
-degree) on any note, and can be major (with the same notes of Ionian mode) or
-minor (same notes as Eolian mode, 'natural minor', which has both
-harmonic and melodic versions, see equations~\ref{eq:escalasMenores}). The
+degree) on any note, and can be major (with the same notes of Ionian mode) or minor (same notes as Eolian mode, 'natural minor', which has both harmonic and melodic versions, see equations~\ref{eq:escalasMenores}).
+The
 scale is the base for triads, each with its root in a degree: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$. 
-To build triads, the third and the fifth notes above the root is considered together with the root (or fundamental).
+To build triads, the third and the fifth notes above the root are taken together with the root (or fundamental).
 $\hat{1},\hat{3},\hat{5}$ is the first degree chord,
- built on top of scale's first degree and central for tonal music. The chords
-of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when a minor
-scale) and of the forth degree $\hat{4},\hat{6},\hat{1}$ are secondary. After
-these, other degrees are considered. The `traditional harmony' comprises conventions
- and stylistic techniques to create chainings with such chords~\cite{Harmonia}.
-
-The `functional harmony' ascribes functions to these three central chords and
-tries to understand their use by means of these functions. The chord built on top of
-first degree is the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its function (role) consists of maintaining a center, usually referred as a ``ground'' for the music.
+ built on top of scale's first degree and central for tonal music.
+The chords of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when a minor scale) and of the forth degree $\hat{4},\hat{6},\hat{1}$ are secondary.
+After these, other degrees are considered.
+The `traditional harmony' comprises conventions and stylistic techniques to create chainings with such chords~\cite{Harmonia}.
+
+The `functional harmony' ascribes functions to these three central chords and tries to understand their use by means of these functions.
+The chord built on top of the first degree is the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its function (role) consists of maintaining a center, usually referred to as a ``ground'' for the music.
 The chord built on the fifth degree is the \textbf{dominant} (\textit{D}, the dominant is always major) and its function is to lean for the tonic
@@
 subdominant, respectively) and its function is to deviate the music from the tonic.
-The process aims at confirming the tonic using tonic-dominant-tonic chains which are
-expanded by using other chords in different ways.
-
-The remaining triads are associated to these three most important chords.
- In the major scale, the associated relative (relative tonic \textit{Tr}, relative
-subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built a
-third below, and the associated counter-relative (counter-relative
-tonic \textit{Ta}, counter-relative subdominant \textit{Sa} and the
-counter-relative dominant \textit{Da}) is the triad built in a third above. 
-In the minor scale the same happens, but the triad a third below is called
-counter-relative (tA, sA) and the triad a third above is called relative (tR,
+The process aims at confirming the tonic using tonic-dominant-tonic chains which are expanded by using other chords in different ways.
+
+The remaining triads are associated to these three most important chords. In the major scale, the associated relative (relative tonic \textit{Tr}, relative
+subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built a third below, and the associated counter-relative (counter-relative tonic \textit{Ta}, counter-relative subdominant \textit{Sa} and the counter-relative dominant \textit{Da}) is the triad built in a third above.
+In the minor scale the same happens, but the triad a third below is called counter-relative (tA, sA) and the triad a third above is called relative (tR,
 sR). The precise functions and musical effects of these chords are
-controversial. Table~\ref{tab:harmonia} shows relations between the triads
-built in each degree of the major scale.
-
-\begin{table}[htpq!]
-\centering
-\caption{Summary of tonal harmonic functions of the major scale. Tonic is the
-music center, the dominant goes to the tonic and the subdominant moves the music away from
+controversial. Table~\ref{tab:harmonia} shows relations between the triads built in each degree of the major scale.
+
+\begin{table}[htp!]
+\centering
+\caption{Summary of tonal harmonic functions of the major scale. Tonic is the music center, the dominant goes to the tonic and the subdominant moves the music away from
 the tonic. The three chords can, at first, be freely replaced by their
@@
 
-The dominant counter-relative should form a minor chord. It explains the change
-in the forth degree by a semitone above $\hat{7}\#$. The diminished chord
-$\hat{7},\hat{2},\hat{4}$, is generally considered a `dominant seventh chord
-with omissed root'~\cite{Koellheuteur}.
-In minor mode, there is a change in $\hat{7}$ by an ascending semitone to
-make possible the existence of an unique semitone separating $\hat{7}$ and $\hat{1}$,
-and making the dominant possible (that should be major and goes to the tonic). In
-this way, the dominant is always major, for both major and minor scale and,
-because of that, even in a minor tone, the relative dominant remains a third
-below and in the counter-relative, a third above.
+The dominant counter-relative should form a minor chord. It explains the change in the forth degree by a semitone above $\hat{7}\#$. The diminished chord
+$\hat{7},\hat{2},\hat{4}$, is generally considered a `dominant seventh chord with omitted root'~\cite{Koellheuteur}.
+In the minor mode, there is a change in $\hat{7}$ by an ascending semitone to yield a unique semitone separating $\hat{7}$ and $\hat{1}$, and making the dominant possible (that should be major and goes to the tonic).
+In this way, the dominant is always major, for both major and minor scale and, therefore, even in a minor tone the relative dominant remains a third below, and in the counter-relative, a third above.
 
@@
 
-Each one of these chords can be confirmed and developed by performing their
-individual dominant or subdominant, which are the triads based
-a fifth above or a fifth below, respectively. These individual dominants and subdominants,
-in the same way, have also subdominants and individual dominants of their own.
-Given a tonality, any chord can occur, no matter
-how distant it is from the harmonic field and from the notes within the
-scale. The unique condition is that the occurrence presents a coherent trajectory of
-dominants and subdominants to the original tonality.
+Each one of these chords can be confirmed and developed by performing their individual dominant or subdominant, which are the triads based
+on a fifth above or a fifth below, respectively. These individual dominants and subdominants,
+in the same way, have also subdominants and individual dominants of their own. Given a tonality, any chord can occur, no matter
+how distant it is from the harmonic field and from the notes within the scale.
+The unique condition is that the occurrence presents a coherent trajectory of dominants and subdominants to the original tonality.
 
 Mediants, or 'chromatic mediants', are two for each chord: the upper
-mediant, formed by the root at the third of original chord; and the lower
-mediant, formed by the fifth at the third of original chord.
-Both chords also are formed by a third, but with a chromatic alteration regarding
-the original chord. If two chromatic alterations exist, i.e.\ two notes altered
-by one semitone each regarding the original chord, it is a
-'doubly chromatic mediant'. Again, there are two forms for each chord: the upper form, with
-a third in the fifth of the original triad; and the lower form, with a third in the root
-of the original triad. It is interesting to observe that a major chord has major chromatic
- and doubly chromatic mediants. A minor chord has minor chromatic and doubly chromatic mediants.
-(Recall that relatives and counter-relatives have oposite major-minor quality.)
-This relation between chords is considered
-advanced tonalism, sometimes even considered expansion and dissolution of tonalism,
-with strong and impressing effects although perfectly consonant triads. Chromatic
-mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss,
+mediant, formed by the root at the third of original chord; and the lower mediant, formed by the fifth at the third of original chord.
+Both chords also are formed by a third, but with a chromatic alteration regarding the original chord.
+If two chromatic alterations exist, i.e.\ two notes altered by one semitone each regarding the original chord, it is a 'doubly chromatic mediant'.
+Again, there are two forms for each chord: the upper form, with a third in the fifth of the original triad; and the lower form, with a third in the root
+of the original triad.
+It is worth observing that a major chord has major chromatic and doubly chromatic mediants.
+A minor chord has minor chromatic and doubly chromatic mediants.
+(Recall that relatives and counter-relatives have opposite major-minor quality.)
+This relation between chords is considered advanced tonalism, sometimes even considered expansion and dissolution of tonalism,
+with strong and impressing effects although they are perfectly consonant triads. Chromatic mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss,
 among others, and are quite simple to be realized~\cite{Harmonia,Salzer}.
@@
 
-Modulation is the change of key (tonic, or tonal center) in a music. 
-It is characterized by observance of start and end keys, and transition artifacts.
-Keys are always conceived as related by fifths and their relative and counter-relatives.
-Ways to perform modulation:
+Modulation is the change of key (tonic, or tonal center) in a music, being characterized by start and end keys, and transition artifacts.
+Keys are always conceived as related by fifths and their relative and counter-relatives. Some ways to perform modulation include:
 
 \begin{itemize}
-    \item Transposing the discourse to a new key, without any preparation. It is
-    a common Baroque procedure although incident at other periods as well. Sometimes it is 
+    \item Transposing the discourse to a new key, without any preparation. It is a common Baroque procedure although incident at other periods as well. Sometimes it is 
     called phrasal modulation or unprepared modulation.
@@
     subdominant, to confirm change in key and harmonic field.
-    \item Use of chromatic alterations to reach a chord in the new key starting from a
-    chord in the previous key. Called chromatic modulation.
-    \item Featuring an unique note, possibly repeated or suspended with no
-    accompaniment, common to start and end keys, constitute a peculiar way
+    \item Use of chromatic alterations to reach a chord in the new key starting from a chord in the previous key. Called chromatic modulation.
+    \item Featuring a unique note, possibly repeated or suspended with no accompaniment, common to start and end keys, it constitutes a peculiar way
     to introduce the new harmonic field.
@@
     (or vice-verse) comprehends a parallel modulation. Keys with same tonic and
-    different quality is known as homonyms.
+    different quality are known as homonyms.
 \end{itemize}
@@
 which leads to the circle of fifths~\cite{Harmonia,Salzer,Koellheuteur,Harmony}.
-Other inventive ways to modulate are possible, to point but one common example,
-the minor thirds tetrad ($E_i^{tm}$ in equations~\ref{escSim}) can be susteined to bridge to other tonalities, with the ease
-of its both tritones.
-The music piece \emph{Acorde cedo} explores these chord relations. Its code is
-online as part of \massa\ toolbox~\cite{MASSA}.
+Other inventive ways to modulate are possible, to point but one common example, the minor thirds tetrad ($E_i^{tm}$ in equations~\ref{escSim}) can be sustained to bridge to other tonalities, with the ease of its both tritones.
+The music piece \emph{Acorde cedo} explores these chord relations, and is implmemented online as part of \massa\ toolbox~\cite{MASSA}.
 
@@
     \centering
-        \includegraphics[width=\columnwidth]{movContraponto}
+        \includegraphics[width=\columnwidth]{figures/movContraponto}
     \caption{Different motions of counterpoint aiming to preserve independence
@@
 
-The purpose of counterpoint is to conduct voices in a way that they sound
-independent. In order to do that, the relative motion of voices (in pairs) is crucial and
+The purpose of counterpoint is to conduct voices in a way that they sound independent. In order to do that, the relative motion of voices (in pairs) is crucial and
 categorized as: direct, oblique and contrary (see Figure~\ref{fig:movContraponto}).
-The parallel motion is an direct motion.
+The parallel motion is a direct motion.
 The gold rule here is to take care of the direct motions, avoiding them
 when leading to a perfect consonance. The parallel motion should occur only between
-imperfect consonances and no more than three consecutive times. Dissonances
-can be unadmitted or used when followed and preceded by consonances of joint
+imperfect consonances and no more than three consecutive times. Disonances can be unadmitted or used when followed and preceded by consonances of joint
 degrees, i.e.\ neighbor notes in a scale. The motions that lead to a
 neighbor note in the scale sound coherent. When having 3 or more voices, the melodic
-importance lies on the higher and lower voices, in this order~\cite{Fux,Tragtenberg,SchoenbergContra}.
-
-These rules were used in the music piece \emph{Conta ponto}, the source code is
+importance lies in the higher and lower voices, in this order~\cite{Fux,Tragtenberg,SchoenbergContra}.
+
+These rules were used in the music piece \emph{Conta ponto}, whose source code is
 available online in \massa\ toolbox.
@@
 
-Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}.
-These events can be heard individually if spaced by at least $50-63ms$. 
-For the temporal separation between them to be perceived as duration,
-the period should even a bit larger, around $100ms$~\cite{microsound}.  
+Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}, which can be heard individually if spaced by at least $50-63ms$.
+For the temporal separation between them to be perceived as duration, the period should even a bit larger, around $100ms$~\cite{microsound}.
 It is possible to summarize duration heard as rhythm or pitch, and its transition,
@@
 
-\begin{table*}[htpq!]
+\begin{table*}[htp!]
 \tiny
@@
 
-The rhythmic metric is commonly based on a basic duration called pulse. The
-pulse typically comprehends durations between $0.25-1.5s$ ($240$
-and $40 BPM$, respectively). In music education and cognitive studies, it is common to
-associate these range of frequencies to the durations of the heart beat,
-movements of inspiration and expiration and steps of a walking or running person~\cite{Lacerda,Roederer}.
-
-The pulse is subdivided into equal parts and is also repeated in sequence.
-These relations (division and concatenation) usually follow relations of small
+The rhythmic metric is commonly based on a basic duration called pulse, whose durations range between $0.25 and 1.5s$ ($240$
+and $40 BPM$, respectively).
+In music education and cognitive studies, it is common to associate these range of frequencies to the durations of the heart beat, movements of inspiration and expiration and steps of a walking or running person~\cite{Lacerda,Roederer}.
+
+The pulse is subdivided into equal parts and is also repeated in sequence. These relations (division and concatenation) usually follow relations of small
 order integers. The occurrences, in written and ethnic
 music, of musical pulse division (and their sequential grouping at
-time),  are, in ascendent disposition: 2, 4 and 8, after that 3, 6 (two groups of 3 or 3 groups of 2) and 9 and
-12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, complementing
-1-9 and 12. Other metrics are less common, like division and grouping in 13,
-17, etc, and are mainly used in contexts of experimental music and classical
-music of XX and XXI. No matter how complex they seem, metrics are common
-compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
-A schematic exposition is shown in Figure~\ref{fig:pulsoSubAgl}.
+time), are, in ascending order: 2, 4 and 8, after that 3, 6 (two groups of 3 or 3 groups of 2) and 9 and 12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, complementing
+1-9 and 12.
+Other metrics are less common, like division and grouping in 13, 17, etc, and are mainly used in contexts of experimental music and classical music of XX and XXI.
+No matter how complex they seem, metrics are common compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
+A schematic illustration is shown in Figure~\ref{fig:pulsoSubAgl}.
 
@@
     \centering
-        \includegraphics[width=\textwidth]{metricaMusical}
-    \caption{Divisions and groupings of the musical pulse for establishing a
-        metric. Divisions of the quarter note, regarded as the
-        pulse, is presented in the left. The time signature yielded by
-        groupings of the music pulse is presented in the right.}
+        \includegraphics[width=\textwidth]{figures/metricaMusical}
+    \caption{Divisions and groupings of the musical pulse for establishing a metric. Divisions of the quarter note, regarded as the
+        pulse, is presented on the left. The time signature yielded by
+        groupings of the music pulse is presented on the right.}
         \label{fig:pulsoSubAgl}
@@
 
-Dual relations (simple measures and binary divisions) are frequent in dance rhythms
-and celebrations, and are called ``imperfect''. Ternary relations are common in
-ritualistic music and is related to the sacred, and are called ``perfect''.
-
-Strong units (accents) fall in the `head' of bars, the downbeats, of which the first in
-every bar is the greatest. The
-head of a unit is the first part of the subdivision. In binary divisions (2, 4
+Dual relations (simple measures and binary divisions) are frequent in dance rhythms and celebrations, and are called ``imperfect''. Ternary relations are common in
+ritualistic music and is related to the sacred, and are called ``perfect''. Strong units (accents) fall in the `head' of bars, the downbeats, of which the first in
+every bar is the greatest. The head of a unit is the first part of the subdivision. In binary divisions (2, 4
 and 8), strong units alternate with weak units
 (e.g.\ division in 4 is: strong, weak, average strong, weak). In ternary divisions
-(3, 6 and 9) two weak units succeeds the downbeat (e.g.\ division in 3 is:
+(3, 6 and 9) two weak units succeed the downbeat (e.g.\ division in 3 is:
 strong, weak, weak). Division in 6 is considered compound but can also
-occur as a binary division. Binary division units which suffers a ternary division
-yields two units divided in tree units each: strong (subdivided in strong,
+occur as a binary division. Binary division units which suffers a ternary division yields two units divided into tree units each: strong (subdivided in strong,
 weak, weak) and weak (subdivided in strong, weak, weak). Another way to perform
@@
 
-The accent in the weak beat is the backbeat, whereas a note starting in a weak
-beat persisting across the strong beat is the syncope.
-
-Notes can occur inside and outside of these \emph{'musical metric'} divisions. In
-most well behaved cases, notes occur exactly on these divisions, with 
-greater incidence strong beat attacks.
+The accent in the weak beat is the backbeat, whereas a note starting in a weak beat persisting across the strong beat is the syncope.
+
+Notes can occur inside and outside of these \emph{'musical metric'} divisions. In most well-behaved cases, notes occur exactly on these divisions, with greater incidence on strong beat attacks.
 In extreme cases, time metric cannot be perceived~\cite{Roederer}. 
@@
 
-Be the pulse the grouping level $j=0$, level $j=-1$ is the first pulse subdivision,
-level $j=1$ is the first pulse agglomeration and so on. In this way, $P_i^j$ is the
-$i$-th unit at grouping level $j$: $P^0_{10}$ is the tenth pulse,
-$P^{1}_3$ is the third grouped unit (and, possibly, the third measure),
+Let the pulse be the grouping level $j=0$, the first pulse subdivision  be level $j=-1$,
+the first pulse agglomeration be level $j=1$ and so on.
+Hence, $P_i^j$ is the $i$-th unit at grouping level $j$: $P^0_{10}$ is the tenth pulse, $P^{1}_3$ is the third grouped unit (and, possibly, the third measure),
 $P^{-1}_2$ is the second part of pulse subdivision.
-
-The limits of $j$ are of special interest: pulse divisions are
-durations perceivable as rhythm; furthermore, the pulses sum, at its
-maximum, a music or a cohesive set of musics. In other words, a duration given
+The limits of $j$ are of special interest: pulse divisions are durations perceivable as rhythm; furthermore, the pulses sum, at its maximum, a music or a cohesive set of musics.
+In other words, a duration given
 by $P^{min(j)}_i$, $\forall \; i$, is greater than $50 ms$ and the durations
@@
 two, four or eight different values, than there is an imperfect relation,
-as shown in Figure~\ref{fig:pulsoSubAgl}.
-
-Any unit (note) of a given musical sequence with a time metric can be unequivocally
+as shown in Figure~\ref{fig:pulsoSubAgl}. Any unit (note) of a given musical sequence with a time metric can be unequivocally
 specified as:
@@
 As an example, consider $P^{-1,0,1}_{3,2,2}$ as the third subdivision $P^{-1}_3$ of the
-second pulse $P^0_2$ and of the second pulse group $P^1_2$ (possibly second measure).
-Each unit $P_i^j$ can be associated with a sequence of temporal samples $T_i$ that comprehends a
-note.
-
-The music piece \emph{Poli Hit Mia} explorers different metrics (available as part of \massa\ toolbox).
+second pulse $P^0_2$ and of the second pulse group $P^1_2$ (possibly second measure). Each unit $P_i^j$ can be associated with a sequence of temporal samples $T_i$ that comprehends a
+note. The music piece \emph{Poli Hit Mia} uses different metrics (available as part of \massa\ toolbox).
 
@@
 
-Given the basic musical structures, both frequential (chords and scales) and
-rhythmic (simple, compound and complex beat divisions and agglomerations), it is
+Given the basic musical structures, both frequential (chords and scales) and rhythmic (simple, compound and complex beat divisions and agglomerations), it is
 natural to present these structures in a coherent and meaningful way~\cite{Boulez}.
-For such, the concept of arcs is fundamental: by departing from a place and returning,
-an arc is made. The audition of melodic and harmonic lines is permeated by
-musical arcs due to cognitive nature of the musical hearing. 
+The concept of arcs is essential in this context: by departing from a place and returning, an arc is made.
+The audition of melodic and harmonic lines is permeated by
+musical arcs due to the cognitive nature of the musical hearing. 
 The note can be considered the smaller arc, and each motif and melody as an arc as well.
 Each beat and subdivision, each measure and music
-section, constitutes an own arc. A music in which the arcs do not present consistency with one another,
+section, constitutes an arc. A music in which the arcs do not present consistency with one another,
 can be understood as a music with no coherence. Coherence impression
@@
 
-Musical arcs are abstract structures and liable of basic operations. A spectral
-arc, like a chord, can be inverted, enlarged and permuted, to cite some
-examples. Temporal arcs, like a melody, a motif, a measure or a note, are also
-capable of variations. Remembering that
+Musical arcs are abstract structures and amenable to basic operations.
+A spectral arc, like a chord, can be inverted, enlarged and permuted, to mention a few possibilities.
+Temporal arcs, like a melody, a motif, a measure or a note, are also
+capable of variations. Recalling that
 $S_j=\left\{s_j=T_i^j=\{t_i^{j}\}_0^{\Lambda_j-1}\right\}_0^{H-1}$ is a sequence
@@
     where $\Gamma$ is the duration between the beginning of the piece (or considered
-    section) and the first event $s_0$ of original structure $S_j$. Observe that
+    section) and the first event $s_0$ of original structure $S_j$. One should observe that
     $\delta$ is the time offset of the displacement.
@@
         frequency is $\tau_f=f_0.2^{\frac{\tau}{12}}$ where $f_0$ is the
-        reference frequency value: $f_0=\Xi_{f_0}Hz\;\sim \Xi$ absolute
-        value of pitch. The frequency of any pitch value is $f=f_0.2^{\frac{\tau}{12}}$s; 
+        reference frequency value: $f_0=\Xi_{f_0}Hz\;\sim \Xi$ absolute value of pitch. The frequency of any pitch value is $f=f_0.2^{\frac{\tau}{12}}$s; 
         and the pitch of any frequency is: $\Xi=\Xi_0 +12
@@
         $\Xi_0=69$. The difference ($\Xi_1 - \Xi_2$) is measured in semitones.
-        $\Xi$ is not a measure in semitones: $\Xi=1$ is not a semitone,
-        it is a note with an audible frequency as rhythm, with
+        $\Xi$ is not a measure in semitones: $\Xi=1$ is not a semitone, it is a note with an audible frequency as rhythm, with
         less than 9 occurrences each second (see table~\ref{tab:duracoes}).
 
-        \item Interval inversion is: 1) the inversion of note pitch order, within octave equivalence, such as exposed in~\ref{subsec:intervalos}, and $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{\varepsilon.
+        \item Interval inversion is: 1) the inversion of note pitch order, within octave equivalence, such as described in~\ref{subsec:intervalos}, and $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{\varepsilon.
 f_0}\right\}$ with selective $s_j=2,\;1/2,\;\text{or}\;1$; or 2) the inversion of interval orientation.
@@
         rotation of $\tilde{n}$ positions is $S'_n=S_{(n+\tilde{n})\%H}$. If
-        $\tilde{n}<0$, it is enough to use $\tilde{n}'=H-\tilde{n}$. It is
+        $\tilde{n}<0$, it is sufficient to use $\tilde{n}'=H-\tilde{n}$. It is
         reasonable to associate $\tilde{n}<0$ (events advance) with the clockwise rotation and
@@
     Elements of $R_j$ can be inserted at the beginning, like a prefix
-    in $S_j$; at the end, as a suffix; or at the middle, dividing $S_j$ and making
+    in $S_j$; at the end, as a suffix; or in the middle, dividing $S_j$ and making
     it the prefix and suffix. Both materials can be mixed in a variety of ways.
@@
     \item Changes in articulation, orchestration and spatialization, 
-    $s_j'=s_j^{*_j}$, where $*_j$ is the new characteristic incorporated by the
+    $s_j'=s_j^{*_j}$, where $*_j$ is the new characteristic incorporated by 
     element $s_j'$.
 
-    \item Accompaniment. Both orchestration and melodic lines presented when
-    $S_j$ occurs can suffer modifications and be considered as a variation of $S_j$ itself.
+    \item Accompaniment. Both orchestration and melodic lines presented when $S_j$ occurs can suffer modifications and be considered as a variation of $S_j$ itself.
 \end{itemize}
 
-With these presented processes, others are derived, as, for example, the inverted retrograde,
-a temporal contraction with an external suffix, etc. Music
-structures resonate in the human cognitive system due to the own nature of
-thought. In its many facets, an idea is observed with the same number of elements and
-connective aspects between them. The music, when tunned with mental structures,
-raises related impressions. In this way, a whole process of mental and neurological
-activity is unleashed, responsible by the feelings, memories and imaginations,
-typical of a diligent music listening. This cortical activity is critical to
-musical therapy, known by its utility in cases of depression and neurological
-injury. It is known that regions of the human brain responsible by
-sonic processing, are also used for other activities, such as linguistic and mathematical~\cite{Sacks,Roederer}.
+From these presented processes, others are derived, as the inverted retrograde, the temporal contraction with an external suffix, etc.
+A whole process of mental and neurological activity is unleashed, responsible for feelings, memories and imaginations, typical of a diligent music listening.
+This cortical activity is critical to
+musical therapy, known by its utility in cases of depression and neurological injury.
+It is known that regions of the human brain responsible by sonic processing are also used for other activities, such as language-related and mathematics~\cite{Sacks,Roederer}.
 
 Paradigmatic structures guide the creation of new musical material.
-One of the most central of them is the tension/relaxation dipole. Traditional dipoles
-are relate to tonic/dominant, repetition/variation,
+One of the most central is the tension/relaxation dipole. Traditional dipoles are relate to tonic/dominant, repetition/variation,
 consonance/dissonance, coherence/rupture, symmetry/asymmetry,
@@
 
-The arcs can be decomposed in two convergent sequences: the first reaches the
-apex and the second returns from apex to start region.
+The arcs can be decomposed in two convergent sequences: the first reaches the apex and the second returns from apex to start region.
 This apex is called climax by traditional music theory. It is
-possible to differ between arcs whose climax are placed at the beginning, middle, end,
-and at the first and second half of the considered duration. These structures are
-shown in Figure~\ref{fig:climax}. The varying parameter can be unexistent, a case in which
+possible to distinguish between arcs whose climax is placed at the beginning, middle, end,
+and at the first and second half of the duration. These structures are
+shown in Figure~\ref{fig:climax}. The varying parameter can be non-existent, a case in which
 the arc consists only of a reference structure~\cite{Schoenberg}.
@@
     \centering
-        \includegraphics[width=\columnwidth]{climax}
+        \includegraphics[width=\columnwidth]{figures/climax}
         \caption{Canonical distinctions of musical climax in a given melody and
-        other domains. Considered possibilities are: climax at beginning,
-        climax at first half, climax at middle, climax at second half and climax
-        at the end. The x and y-axis is not properly specified since the parameters
-        can be non-existent, as in a reference structure.}
+        other domains. The possibilities considered are: climax at the beginning, at the first half, in the middle, in the second half and 
+        at the end. The x and y-axis is not properly specified since the parameters can be non-existent, as in a reference structure.}
         \label{fig:climax}
@@
 by using sequences with different sizes. All the mathematics of
-sequences, already well established and taught routinely in calculus courses, can be
-used to generate these arcs~\cite{Guidorizzo,Schoenberg}. 
+sequences, already well established and taught routinely in calculus courses, can be used to generate these arcs~\cite{Guidorizzo,Schoenberg}. 
 Theoretically, when applied to any characteristic of musical events, 
-these sequences produce arcs, since they imply a deviation and return of an initial
-parametrization.
-Henceforth, it is possible for an only sequence to have
+these sequences produce arcs, since they imply a deviation and return of an initial parametrization.
+Henceforth, it is possible for a given sequence to have
 numberous distinct arcs, with different sizes and climax. 
@@
 
-In practice, and historically, there is special incidence and use of the golden ratio.
-The Lucas sequence allows the generalization of
+In practice, and historically, there is special incidence and use of the golden ratio. The Lucas sequence allows the generalization of
 Fibonacci sequence, making it easy to follow. Given any two numbers $x_0$
@@
 $1.61803398875...$. The sequence converges fast even with high discrepant
-initial values. Being $x_0=1$ and $x_1=100$, and $y_n=\frac{x_n}{x_{n+1}}$ an
+initial values. If $x_0=1$ and $x_1=100$, and $y_n=\frac{x_n}{x_{n+1}}$ an
 auxiliary sequence, the error for the fraction of first values with
@@
 
-The music piece \emph{Dirracional} exposes the use of arcs into directional structures.
-Its source code is available online as part
+The music piece \emph{Dirracional} uses arcs into directional structures. Its source code is available online as part
 of \massa\ toolbox~\cite{MASSA}.
@@
 
-The philosophical understanding that human thought is founded on the concept of
-similarities and differences (e.g. perceived in stimuli and objects), places symmetries
-at the core of cognitive processes~\cite{Deleuze}. Mathematically, symmetries
-are algebraic groups, and a finite group is always isomorphic to a permutation
+The philosophical understanding that human thought is founded on the concept of similarities and differences (e.g. perceived in stimuli and objects), places symmetries
+at the core of cognitive processes~\cite{Deleuze}. Mathematically, symmetries are algebraic groups, and a finite group is always isomorphic to a permutation
 group (by Cayley's theorem). 
@@
 finite system~\cite{gruposFascination}.
-In music, permutations are ubiquitous and present in scholastic techniques,
+In music, permutations are ubiquitous in scholastic techniques,
  which confirms their central role. 
@@
 Two academic studies were dedicated to generating musical structures~\cite{figgusOriginal, figgusEspacializacao}.
-
 Any permutation set can be used as a generator of algebraic groups~\cite{permMusic}.
@@
 
-From the first property follows that two permutations act as one permutation.
-In fact, it is possible to apply a
-permutation $p_1$ and another permutation $p_2$, and, comparing both initial and final
-orderings, observe another permutation $p_3$.
-
-Every element $p$ operated with itself a sufficient number of times $n$ reaches
-the identity element $p^n=e$ (otherwise the group would be infinite, generated
+From the first property follows that two permutations act as one permutation. In fact, it is possible to apply a
+permutation $p_1$ and another permutation $p_2$, and, comparing both initial and final orderings, observe another permutation $p_3$.
+Every element $p$ operated with itself a sufficient number of times $n$ reaches the identity element $p^n=e$ (otherwise the group would be infinite, generated
 by $p$). The lower $n\;:\;p^n=e$ is the element order. Thus, a finite
@@
 example, the \emph{change ringing} tradition conceives music through
-bells played one after other and then played again, but in a different
+bells played one after another and then played again, but in a different
 order. This process repeats until it reaches the initial ordering. The set of
@@
 
-\begin{table}[htpq!]
+\begin{table}[htp!]
 \centering
@@
 permutation $p$. $S_i'=p(S_i)$ comprises the same elements of $S_i$ but in a
-different order. Permutations has two notations: cyclic and
-natural. The natural notation basically comprehends the order of the resulting indexes from
-the permutation. Thus, agreed the original ordering given the sequence of its
-indexes $[0\;1\;2\;3\;4\;5\;...]$, the permutation is noted by the sequence it
+different order. Permutations have two notations: cyclic and
+natural. The natural notation basically indicates the order of the resulting indexes from
+the permutation. Thus, given the original ordering of the sequence by its indexes $[0\;1\;2\;3\;4\;5\;...]$, the permutation is noted by the sequence of indexes it
 produces (ex. $[1\;3\;7\;0\;...]$). In the cyclic notation, a permutation consists
-of swaps by elements and its posteriors, and the last element by the first one.
+of swaps by elements and its successors, and the last element by the first one.
 E.g. $(1,2,5)(3,4)$ in cyclic notation is equivalent to $[0,2,5,4,3,1]$ in natural notation.
@@
 In the auralization of a permutation, it is not necessary to permute elements of $S_i$,
-but only some characteristic. In this way, being $p$ a permutation and $S_i$ a
-sequence of basic notes as exposed in the end of section~\ref{notaBasica}, the
+but only some characteristic. Thus, if $p$ is a permutation and $S_i$ is a sequence of basic notes as in the end of section~\ref{notaBasica}, the
 sequence $S_i'=p^f(S_i)=\left\{s_i^{p(f)}\right\}$ consists of the same
-music notes, respecting same order and maintaining the same characteristics, but with the
-fundamental frequencies permuted following the pattern that $p$ presents.
-
-Two subtleties of this procedure. 
-First, a permutation $p$ is not restricted to involve all elements of $S_i$,
-i.e.\ it can operate in a subset of $S_i$.
-Second,  not all elements $s_i$ need to be executed at each access to $S_i$.
-To exemplify, be $S_i$ a sequence of music notes $s_i$. 
+music notes, following the same order and maintaining the same characteristics, but with the
+fundamental frequencies permuted according to the pattern of $p$.
+
+Two subtleties of this procedure should be commented upon. 
+First, a permutation $p$ is not restricted to involve all elements of $S_i$, i.e.\ it can operate in a subset of $S_i$.
+Second, not all elements $s_i$ need to be executed at each access to $S_i$. To exemplify, let $S_i$ be a sequence of music notes $s_i$. 
 If $i$ goes from $0$ to $n$, and
 $n>4$, at each measure of $4$ notes it is possible to execute the first $4$
-notes. The other notes of $S_i$ can occur in measures where permutations 
+notes.
+The other notes of $S_i$ can occur in other measures where permutations 
 allocate those notes to the first four events $s'_i$ of $S'_i$.
 
-It is inalienable to each permutations $p_i$ described above:
+To each permutation $p_i$ described above, we have to determine:
 1) note dimensions where it operates (frequency, duration, \emph{fades},
@@
 execute the first $n$ notes, the execution of disjoint sets of $S_i$
-equals modifying the permutation and executing the first $n$ notes.
+is the same as modifying the permutation and executing the first $n$ notes.
 
@@
 dichotomies are prominent, as explained in subsection~\ref{subsec:motivos}:
-repetition and variation, relaxation and tension, equilibrium and instability, consonance and dissonance, etc.
+repetition and variation, relaxation and tension, equilibrium and instability, consonance and disonance, etc.
 
@@
 First, the basic note was defined and characterized in quantitative terms (section~\ref{sec:notaDisc}).
-Next, internal note
+Next, the internal note
 composition was addressed and both internal transitions and immediate sonic treatment
-were understood (section~\ref{varInternas}). Finally, this section aims at
-organizing these notes in music. The gamma of resources and consequent infinitude
+were understood (section~\ref{varInternas}). Finally, this section aims at organizing these notes in music. The gamma of resources and consequent infinitude
 of praxis possibilities is a typical and highly relevant situation for artistic contexts~\cite{Harmonia,Webern}.
@@
 There are studies for each of the presented resources.
-For example, it is possible to obtain `dirty' triadic harmonies (with notes out of the triad) by
-superposition of perfect fourths. 
+For example, it is possible to obtain `dirty' triadic harmonies (with notes out of the triad) by superposition of perfect fourths. 
 Another interesting example is the simultaneous
 presence of rhythms in different metrics, constituting what is
-called \emph{polyrhythm}. The music piece \emph{Poli-hit mia} explores these
-simultaneous metrics by impulse trains convolved with notes.
-Its source code is online as part of \massa\ toolbox~\cite{MASSA}.
+called \emph{polyrhythm}. The music piece \emph{Poli-hit mia} in \massa\ toolbox~\cite{MASSA} explores these simultaneous metrics by impulse trains convolved with notes.
 
@@
 century music~\cite{microtonalidade} and present diverse remarkable results, e.g.
-fourths of tone in the Indian music. The musical sequence \emph{MicroTom}
-explores these resources, including microtonal melodies and microtonal harmonies
-with many notes in a very reduced note scope. Its code is 
-available online in \massa\ toolbox~\cite{MASSA}.
+fourths of tone in the Indian music. The musical sequence \emph{MicroTom} in \massa\ toolbox~\cite{MASSA} explores these resources, including microtonal melodies and microtonal harmonies
+with many notes in a very reduced note scope.
 

### 2017-06-06 article/acm/body.tex
andando com revisao do artigo p acm

--- 2017-04-24 article/article.tex
@@
-% notas:
-% campo harmônico => harmonic field?
-% anti-relative => counter-relative? (discussão sobre harmonia funcional)
-%    associada relative => associated relative? (ver a discussão para mais...)
-% medianas cromáticas => cromatic mediant
-% tonalidade => key
-% homônica => homonymous
-
-\documentclass[
- aip,
- jmp,
- amsmath,amssymb,
- reprint,
-]{revtex4-1}
-\usepackage{graphicx}
-\usepackage{grffile}
-\usepackage{dcolumn}
-\usepackage{bm}
-\usepackage{multirow}
-\usepackage{color}
-\usepackage{xcolor}
-\usepackage{enumerate}
-\usepackage{float}
-\usepackage{hyperref}
-
-\newcommand{\massa}{{\large \textsc{massa}}}
-\newcommand{\mass}{{\large \textsc{mass}}}
-\newcommand{\figgus}{{\large \textsc{figgus}}}
-
-\floatstyle{plain}
-\newfloat{twocolequfloat}{t}{zzz}
-\floatname{twocolequfloat}{Equation}
-
-\begin{document}
-\preprint{AIP/123-QED}
-
-\title{Psychophysics of musical elements in the discrete-time representation of sound}
-
-\author{Renato Fabbri}
- \email{renato.fabbri@gmail.com}
-  \affiliation{ 
-Instituto de F\'isica de S\~ao Carlos, Universidade de S\~ao Paulo (IFSC/USP)
-}
-
-\author{Vilson Vieira da Silva Junior}
-  \homepage{http://automata.cc/}
-  \email{vilson@void.cc}
- \altaffiliation[Also at ]{IFSC-USP}
-
-\author{Ant\^onio Carlos Silvano Pessotti}
-  \email{antoniopessotti@gmail.com}
- \altaffiliation[Instituto de Estudos da Linguagem ]{IEL/UNICAMP}
-
-\author{D\'ebora Cristina Corr\^ea}
-  \email{debcris.cor@gmail.com}
- \altaffiliation[Instituto de Ci\^encias Matem\'aticas e de Computa\c{c}\~ao ]{ICMC/USP}
-
-%\author{Luciano da Fontoura Costa}
-%  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
-%  \email{ldfcosta@gmail.com}
-% \altaffiliation[Also at ]{IFSC-USP}
-%
-\author{Osvaldo N. Oliveira Jr.}
-\homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
-\email{chu@ifsc.usp.br}
-\altaffiliation[Also at ]{IFSC-USP}
-
-\date{\today}
-\begin{abstract}
-
-    %The representation of the basic elements of music - such as notes, ornaments and interval structures - in terms of discrete audio signal is often used in software for music creation and design.
-Nevertheless, there is no unified approach that relates these elements to the sound discrete samples.
-In this article, each musical element is described by equations in sonic time samples, which are all implemented in scripts within an open source software toolbox, referred to as \massa\ (Music and Audio in Sequences and Samples).
-The fundamental element, the musical note with duration, volume, pitch and timbre, is related quantitatively to the characteristics of the discrete-time signal.
-Internal variations, such as tremolos, vibratos and spectral fluctuations, are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities.
-With this representation of notes, resources are provided for the generation of musical structures, such as rhythmic meter, pitch intervals and cycles.
-The efficacy of these physical descriptions of basic musical elements was confirmed by the synthesis of small musical pieces within each frame: basic notes, incremented notes and notes in music.
-It is possible to synthesize whole albums through collage of the scripts and parametrization.
-    %The sample-based analytical description, and the paradigm of open source implementation, enables scientific experiments in precise and trustful ways.
-In fact, \massa\ has already been employed by external users for diverse purposes.
-Among these, it is mentioned an acoustic effect recognized by diverse individuals in mailing lists but not found in literature, and uses related to art and education.
-
-Notes, ornaments and intervals are examples of basic musical elements.
-Their representation as discrete-time digital audio plays a central role in software for music creation and design.
-Nevertheless, there is no systematic relation, in analytical terms, of these musical elements to the sonic samples.
-Such a compendium enables scientific experiments in precise and trustful ways, among educational and artistic uses.
-This paper presents a comprehensive description of music in digital audio, within an unified approach.
-Musical elements, like pitch, duration and timbre are expressed by equations on sample level.
-This quantitatively relates characteristics of the discrete-time signal to musical qualities.
-Internal variations, e.g. tremolos, vibratos and spectral fluctuations, are also considered
-to operate within a note. Moreover, the generation of musical structures such as rhythmic meter,
-pitch intervals and cycles, are attained canonically with notes. 
-The availability of these resources in scripts is provided in public domain within the \massa\ toolbox - Music and Audio in Sequences and Samples.
-Authors observe that the implementation of sample-domain analytical results as open source can encourage concise research.
-As further illustrated in the paper, \massa\ has already been employed by users for diverse purposes, 
-including acoustics experimentation, art and education. The efficacy of these physical descriptions 
-was confirmed by the synthesis of small musical pieces.
-As shown, it is possible to synthesize whole albums through collage of scripts and parametrization.
-
-\end{abstract}
-\pacs{*43.66.-x,43.66.+y,05.10.-a} % PACS, the Physics and Astronomy
-\keywords{psychophysics, acoustics, statistics, signal processing, digital audio, music}
-\maketitle
-
-\section{\label{sec:level1}Introduction}
+\section{Introduction}\label{sec:level1}
 
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/pcm}
+        \includegraphics[width=.7\textwidth]{figures/pcm}
         \caption{Pulse Code Modulation (PCM) audio: an analogical signal is represented by 25 samples with 4 bits each.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/waveForms_}
+        \includegraphics[width=.7\textwidth]{figures/waveForms_}
     \caption{Basic musical waveforms: (a) the synthetic  waveforms; (b) the real waveforms.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/waveSpectrum}
+        \includegraphics[width=.7\textwidth]{figures/waveSpectrum}
     \caption{Spectrum of basic artificial musical waveforms.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/oboeNaturalSampledSpectrum}
+        \includegraphics[width=.7\textwidth]{figures/oboeNaturalSampledSpectrum}
     \caption{Spectrum of the sonic waves of a natural oboe note and from a sampled period.
@@
      \centering
-         \includegraphics[width=\textwidth]{figures/amostras2c__}
+         \includegraphics[width=.7\textwidth]{figures/amostras2c__}
      \caption{Oscillation of 2 samples (maximum frequency for any $f_a$).
@@
      \centering
-         \includegraphics[width=\textwidth]{figures/amostras3b}
+         \includegraphics[width=.7\textwidth]{figures/amostras3b}
      \caption{Three fixed samples present only one non-null frequency. $c_1=c_2^*$ and $w_1 \equiv w_2$.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/amostras4___}
+        \includegraphics[width=.7\textwidth]{figures/amostras4___}
     \caption{Frequency components for 4 samples.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/amostras4formas__}
+        \includegraphics[width=.9\textwidth]{figures/amostras4formas__}
     \caption{Basic wave forms with 4 samples.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/amostras6_}
+        \includegraphics[width=.7\textwidth]{figures/amostras6_}
     \caption{Frequency components for 6 samples: 4 sinusoids, one of them is the \emph{bias} with zero frequency.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/amostras6formas__}
+        \includegraphics[width=.9\textwidth]{figures/amostras6formas__}
     \caption{Basic waveforms with 6 samples: triangular and square waveforms have odd harmonics, with different proportions and phases; the sawtooth has even harmonics.}
@@
 
-Spatialization results from sound reflections and absorptions by room/environment surface where the note is played.
+Spatialization results from sound reflections and absorptions by environment (e.g.
+room) surface where the note is played.
 The sound propagates through the air with a speed of $\approx 343.2m/s$ and can be emitted from a source with any directionality pattern.
@@
     {\centering
-        \includegraphics[width=\columnwidth]{figures/mixagem}}
+        \includegraphics[width=.7\columnwidth]{figures/mixagem_}}
     \caption{Mixing of three sound sequences. The amplitudes are directly overlapped.}
@@
 {    \centering
-        \includegraphics[width=\columnwidth]{figures/concatenacao}}
+        \includegraphics[width=.5\columnwidth]{figures/concatenacao_}}
     \caption{Concatenation of three sound sequences by temporal overlap of their samples.}
@@
      \centering
-         \includegraphics[width=\textwidth]{figures/convolucao}
+         \includegraphics[width=\textwidth]{figures/convolucao_}
      \caption{Graphical interpretation of convolution. Each resulting sample is the sum of the previous samples of a signal, with each one multiplied by the retrograde of the other sequence.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/iir}
+        \includegraphics[width=\textwidth]{figures/iir_}
     \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of equations~\ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} and \ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
@@
 \begin{figure*}
-     \centering
-         \includegraphics[width=\textwidth]{figures/ruidos}
+	\hspace*{-.75cm}
+         \includegraphics[width=1.1\textwidth]{figures/ruidos}
      \caption{Colors of noise generated by equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul} and \ref{eq:violeta}: spectrum and waveforms.}
@@
 For tonal music, the tritone indicates dominant and seeks urgent resolution in a third or sixth. Due to this instability it is considered a dissonant interval.}
-\begin{tabular}{| c | c | c | }\hline
-    \multicolumn{3}{|c|}{\bf consonances}  \\\hline
+\begin{tabular}{ c | c | c }\hline
+    \multicolumn{3}{c}{\bf consonances}  \\\hline
    & traditional notation & number of semitones \\
-   perfects: & P1, P5, P8 & 0, 7, 12 \\
- imperfects: & m3, M3, m6, M6 & 3, 4, 8, 9 \\\hline\hline
-    \multicolumn{3}{|c|}{\bf dissonances} \\\hline
+   perfect: & P1, P5, P8 & 0, 7, 12 \\
+ imperfect: & m3, M3, m6, M6 & 3, 4, 8, 9 \\\hline\hline
+    \multicolumn{3}{c}{\bf dissonances} \\\hline
  & traditional notation & number of semitones \\
- fortes: & m2, M7 & 1, 11 \\
- brandas: & M2, m7 & 2, 10 \\\hline\hline
-    \multicolumn{3}{|c|}{\bf special cases} \\\hline
+ strong: & m2, M7 & 1, 11 \\
+ weak: & M2, m7 & 2, 10 \\\hline\hline
+    \multicolumn{3}{c}{\bf special cases} \\\hline
  & traditional notation & number of semitones \\
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/metricaMusical}
+        \includegraphics[width=.9\textwidth]{figures/metricaMusical_}
     \caption{Divisions and groupings of the musical pulse for establishing a metric. Divisions of the quarter note, regarded as the
@@
 elements in the discrete-time representation of sound as community feedback corroborate.
-.
-
-%% Há um aumento no número de pesquisas relacionadas à música em andamento no campus de São Carlos da USP, o que sugere facilidade para estabelecer parcerias.
-%% As publicações acadêmicas efetivadas durante este mestrado também apontam para uma multidisciplinaridade,
-%% tratando diretamente de questões humanas como artes, filosofia, humor e linguagem falada e escrita, através de artifícios lineares e estatísticos.\cite{FabbriSTAT,FabbriACL,FabbriComplenetVoz,FabbriComplenetTexto} Os desdobramentos estão alcançando redes sociais e teorias epistemológicas com base em pesquisas prévias, dos orientadores deste trabalho, com forte presença de redes complexas e processamento de linguagem natural.
-
-\nocite{*}
-\bibliography{article2}
-\end{document}
+
+\begin{acks}
+
+The authors would like to thank ???
+
+The work is supported by the São Paulo Research Foundation under Grant ???.
+% 	\grantsponsor{GS501100001809}{National
+%   Natural Science Foundation of
+%   China}{http://dx.doi.org/10.13039/501100001809} under Grant
+% No.:~\grantnum{GS501100001809}{61273304\_a}
+% and~\grantnum[http://www.nnsf.cn/youngscientsts]{GS501100001809}{Young
+%   Scientsts' Support Program}.
+\end{acks}
+
+% Bibliography
+\bibliographystyle{ACM-Reference-Format}
+% \bibliography{sample-bibliography}
+\bibliography{../article2}

### 2017-06-06 article/acm/sample-acmsmall.tex
andando com revisao do artigo p acm

\documentclass[format=acmsmall, review=false, screen=true]{acmart}

\usepackage{booktabs} % For formal tables

\usepackage[ruled]{algorithm2e} % For algorithms
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

% Metadata Information
\acmJournal{TWEB}
\acmVolume{9}
\acmNumber{4}
\acmArticle{39}
\acmYear{2010}
\acmMonth{3}
\copyrightyear{2009}
%\acmArticleSeq{9}

% Copyright
%\setcopyright{acmcopyright}
\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{0000001.0000001}

% Paper history
\received{February 2007}
\received[revised]{March 2009}
\received[accepted]{June 2009}

\newcommand{\massa}{{\large \textsc{massa}}}
\newcommand{\mass}{{\large \textsc{mass}}}
\newcommand{\figgus}{{\large \textsc{figgus}}}

% Document starts
\begin{document}
% Title portion. Note the short title for running heads 
\title[Psychophysics of musical elements in digital audio]{Psychophysics of musical elements in the discrete-time representation of sound}  
\author{Renato Fabbri}
% \orcid{1234-5678-9012-3456}
\affiliation{%
  \institution{University of São Paulo}
  \department{Institute of Mathematics and Computer Sciences}
  \streetaddress{Avenida Trabalhador São Carlense, 400 - Centro}
  \city{São Carlos}
  \state{SP}
  \postcode{13566-590}
  \country{Brazil}}
\author{Vilson Vieira da Silva Junior}
\affiliation{%
  \institution{The Grid}
%  \department{Research and Development}
  \city{Berlin}
  \state{BE}
  \postcode{???}
  \country{DE}
}
\author{Ant\^onio Carlos Silvano Pessotti}
\affiliation{%
  \institution{Universidade Metodista de Piracicaba}
  \department{??}
  \city{Piracicaba}
  \state{SP}
  \postcode{???}
  \country{Brazil}}
\author{D\'ebora Cristina Corr\^ea}
\affiliation{%
  \institution{??}
  \department{??}
  \city{Piracicaba}
  \state{SP}
  \postcode{???}
  \country{AU}
  }
\author{Osvaldo N. Oliveira Jr.}
\affiliation{%
  \institution{University of São Paulo}
  \department{São Carlos Institute of Physics}
  \streetaddress{Avenida Trabalhador São Carlense, 400 - Centro}
  \city{São Carlos}
  \state{SP}
  \postcode{13566-590}
  \country{Brazil}
  }

\begin{abstract}
Notes, ornaments and intervals are examples of basic musical elements.
Their representation as discrete-time digital audio plays a central role in software for music creation and design.
Nevertheless, there is no systematic relation, in analytical terms, of these musical elements to the sonic samples.
Such a compendium enables scientific experiments in precise and trustful ways, among educational and artistic uses.
This paper presents a comprehensive description of music in digital audio, within an unified approach.
Musical elements, like pitch, duration and timbre are expressed by equations on sample level.
This quantitatively relates characteristics of the discrete-time signal to musical qualities.
Internal variations, e.g. tremolos, vibratos and spectral fluctuations, are also considered
to operate within a note. Moreover, the generation of musical structures such as rhythmic meter,
pitch intervals and cycles, are attained canonically with notes. 
The availability of these resources in scripts is provided in public domain within the \massa\ toolbox - Music and Audio in Sequences and Samples. Authors observe that the implementation of sample-domain analytical results as open source can encourage concise research.
As further illustrated in the paper, \massa\ has already been employed by users for diverse purposes, 
including acoustics experimentation, art and education. The efficacy of these physical descriptions 
was confirmed by the synthesis of small musical pieces.
As shown, it is possible to synthesize whole albums through collage of scripts and parametrization.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
\begin{CCSXML}
  <ccs2012>
    <concept>
      <concept_id>10010405.10010469.10010475</concept_id>
      <concept_desc>Applied computing~Sound and music computing</concept_desc>
      <concept_significance>500</concept_significance>
    </concept>
    <concept>
      <concept_id>10010147.10010341.10010342.10010343</concept_id>
      <concept_desc>Computing methodologies~Modeling methodologies</concept_desc>
      <concept_significance>300</concept_significance>
    </concept>
    <concept>
      <concept_id>10002944.10011122.10002945</concept_id>
      <concept_desc>General and reference~Surveys and overviews</concept_desc>
      <concept_significance>300</concept_significance>
    </concept>
    <concept>
      <concept_id>10002944.10011122.10002946</concept_id>
      <concept_desc>General and reference~Reference works</concept_desc>
      <concept_significance>300</concept_significance>
    </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Applied computing~Sound and music computing}
\ccsdesc[300]{Computing methodologies~Modeling methodologies}
\ccsdesc[300]{General and reference~Surveys and overviews}
\ccsdesc[300]{General and reference~Reference works}

%
% End generated code
%

\keywords{digital audio, music, signal processing, acoustics, psychophysics}

\thanks{This work is supported by FAPESP (grant number etc ???).}

\maketitle

% The default list of authors is too long for headers}
\renewcommand{\shortauthors}{R. Fabbri et al.}

\input{body}

\end{document}

### 2017-06-06 article/scripts/concatena.py

@@ -29,6 +29,7 @@ p.plot(range(len(d),len(d)+len(q)),q,'o')
 p.plot(range(len(d)+len(q),len(d)+len(q)+len(s)),s,'o')
 p.plot(n.hstack((d,q,s)))
 p.xlabel(r"time $\rightarrow$",fontsize=22)
+p.ylabel(r"amplitude $\rightarrow$",fontsize=22)
 p.ylim(-1.5,1.5)
 p.xlim(-5,305)
 p.show()

### 2017-06-06 article/scripts/convolucao.py

@@ -23,33 +23,34 @@ t=n.sin(x)
 t=n.hstack((n.hstack((n.linspace(-1,1,16),n.linspace(-1,1,16))),n.linspace(-1,1,16)))
 t*=.5
 
-p.plot(t,'bo', label=r'piece of sound signal $\{t_i\}_0^{47}$')
+p.plot(t,'bo', label=r'sonic signal $\{t_i\}_0^{47}$')
 
 #h=1./x[1:11]**.5
 h=n.random.random(10)
 h=n.array([ 0.5591728 ,  0.59152829,  0.43285462,  0.8870076 ,  0.44892785,0.33476906,  0.8808893 ,  0.39040725,  0.56887214,  0.54278373])
 p.plot(range(3,13),h[::-1]+1,'ro',label=r'retrograde of impulse response $\{h_{i-j}\}_{j\; =\; max(i+1-\Lambda_h,0)}^i$')
-for i in xrange(3,12):
+for i in range(3,12):
     p.plot([i,i],[-1,2],'y-.')
 p.plot([12,12],[-5,2], 'y-.',linewidth=3)
 
 
 p.plot(range(23,33),h[::-1]+1,'ro')
-for i in xrange(23,32):
+for i in range(23,32):
     p.plot([i,i],[-1,2],'y-.')
 p.plot([32,32],[-5,2], 'y-.',linewidth=3)
 
 c=n.convolve(t,h)
-p.plot(c-2.5,'go', label=u'convoluted signal ' + r'$\{\, (t*h)_{i}=\sum_{j=max(i+1-\Lambda_h,0)}^{i} t_j . h_{i-j}\;\}_{0}^{48+10-2=56}$')
+p.plot(c-2.5,'go', label=u'convoluted signal ' + r'$\{\, (t*h)_{i}=\sum_{j=max(i+1-\Lambda_h,0)}^{i} t_j .  h_{i-j}\;\}_{0}^{48+10-2=56}$')
 
 
 #p.xlim(-1.2,4.2)
-p.ylim(-7.2,2.5)
+p.ylim(-5.6,6.5)
 
 p.yticks((),())
 p.xticks((),())
-p.legend(loc="upper right")
+p.legend(loc="upper right", fontsize=20)
 #p.legend(loc="upper right",prop={'size':18})
 p.xlabel(r"i $\rightarrow$", fontsize=26)
+p.ylabel(r"amplitude $\rightarrow$", fontsize=26)
 
 p.show()

### 2017-06-06 article/scripts/metricaMusical.py

@@ -15,8 +15,8 @@ ax2=ax1.twinx()
 ordem=[0,1,4,2,-1,5,-2,3,6]
 colors=['g','b','r','b','k','r','k','b','r']
 
-for i in xrange(9):
-    for j in xrange(i+1):
+for i in range(9):
+    for j in range(i+1):
         p.plot((j*resolucao/(i+1), (j+1)*resolucao/(i+1)-2),(ordem[i],ordem[i]),c=colors[i%len(colors)], linewidth=3)
 
 p.plot((-1000,-1000),'r',linewidth=3,label=u'composed measures, ternary divisions, perfect mode')

### 2017-06-06 article/scripts/mixa_concatena.py

@@ -36,5 +36,7 @@ p.text(46,11.4-e-2,"=",fontsize=78)
 p.xticks((),())
 p.yticks((),())
 
+p.xlabel(r"time $\rightarrow$",fontsize=22)
+p.ylabel(r"amplitude $\rightarrow$",fontsize=22)
 
 p.show()

### 2017-07-01 article/acm/body.tex
melhoradas figuras do artigo

@@
      \centering
-         \includegraphics[width=.7\textwidth]{figures/amostras2c__}
+         \includegraphics[width=.7\textwidth]{figures/amostras2c___}
      \caption{Oscillation of 2 samples (maximum frequency for any $f_a$).
@@
      \centering
-         \includegraphics[width=.7\textwidth]{figures/amostras3b}
+         \includegraphics[width=.7\textwidth]{figures/amostras3b_}
      \caption{Three fixed samples present only one non-null frequency. $c_1=c_2^*$ and $w_1 \equiv w_2$.}
@@
     \centering
-        \includegraphics[width=.7\textwidth]{figures/amostras4___}
+        \includegraphics[width=.7\textwidth]{figures/amostras4____}
     \caption{Frequency components for 4 samples.}
@@
     \centering
-        \includegraphics[width=.9\textwidth]{figures/amostras4formas__}
+        \includegraphics[width=.9\textwidth]{figures/amostras4formas___}
     \caption{Basic wave forms with 4 samples.}
@@
     \centering
-        \includegraphics[width=.7\textwidth]{figures/amostras6_}
+        \includegraphics[width=.7\textwidth]{figures/amostras6__}
     \caption{Frequency components for 6 samples: 4 sinusoids, one of them is the \emph{bias} with zero frequency.}
@@
     \centering
-        \includegraphics[width=.9\textwidth]{figures/amostras6formas__}
+        \includegraphics[width=.9\textwidth]{figures/amostras6formas___}
     \caption{Basic waveforms with 6 samples: triangular and square waveforms have odd harmonics, with different proportions and phases; the sawtooth has even harmonics.}
@@
     {\centering
-        \includegraphics[width=.7\columnwidth]{figures/mixagem_}}
+        \includegraphics[width=.5\columnwidth]{figures/mixagem__}}
     \caption{Mixing of three sound sequences. The amplitudes are directly overlapped.}
@@
 {    \centering
-        \includegraphics[width=.5\columnwidth]{figures/concatenacao_}}
+        \includegraphics[width=.5\columnwidth]{figures/concatenacao__}}
     \caption{Concatenation of three sound sequences by temporal overlap of their samples.}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/lut}
-    \caption{Search in the \emph{lookup table} to synthesize sounds at different frequencies using a unique waveform with high resolution.}
+        \includegraphics[width=.7\textwidth]{figures/lut_}
+    \caption{Consultation in the \emph{lookup table} to synthesize sounds at different frequencies using a unique waveform with high resolution.
+	Each \emph{i-th} sample $t_i^f$ of a sound with frequency $f$ is related to the samples in the table $\widetilde{l}$\, by the relation $t_i^{f}=\widetilde{l}_{\left\lfloor i.f\frac{\widetilde{\Lambda}}{f_a} \right\rfloor \%\,\widetilde{\Lambda}}$ where $\widetilde{\Lambda}$ is the number of sample in $\widetilde{l}$ and $f_a$ is the sampling rate.}
         \label{fig:lut}                                                                                                            
@@
      \centering
-         \includegraphics[width=\columnwidth]{figures/transicao}
+         \includegraphics[width=.7\columnwidth]{figures/transicao_}
      \caption{Intensity transitions for different values of $\alpha$ (see equations~\ref{seqAmp} and ~\ref{transAmp}).}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/delays}
+        \includegraphics[width=\textwidth]{figures/delays_}
     \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c). Represented in increasing order of its pulse density.}

### 2017-07-01 article/scripts/amostras3b.py

@@ -11,6 +11,8 @@ n3=n.array([-0.20356754,  0.91942605,  0.44866268])
 #n3=n.array([-0.40+.3, - 0.4+.3,  0.4+.3])
 #n3=n.array([+0.40+.3, - 0+.3, - 0.4+.3])
 #n3=n.array([-0.40+.3, + 0.4+.3, - 0.4+.3])
+p.figure(figsize=(10.,5.))
+p.subplots_adjust(left=0.18,bottom=0.15,right=0.97,top=0.94)
 p.plot(n3,"bo")
 
 ff=f(n3)
@@ -35,12 +37,16 @@ p.plot(iii,s,"m--")
 p.plot((-10,10),(a0/3,)*2,"k--")
 p.plot((-10,10),(a0/3-2*ab1/3,)*2,"k--")
 p.plot((-10,10),(a0/3+2*ab1/3,)*2,"k--")
-p.xticks((-1,0,1,2,3),(r"$-\delta_a$",r"$0$",r"$\delta_a$",r"$2\delta_a$",r"$3\delta_a$"))
-p.yticks((0,a0/3-2*ab1/3,a0/3,a0/3+2*ab1/3), ('0',r'$\frac{a_0-2|c_1|}{3}$',r'$\frac{a_0}{3}$',r'$\frac{a_0+2|c_1|}{3}$'), color = 'k', size = 20)
+p.xticks((-1,0,1,2,3),(r"$-\delta_a$",r"$0$",r"$\delta_a$",r"$2\delta_a$",r"$3\delta_a$"),
+        size=20)
+p.yticks((a0/3-2*ab1/3,a0/3,a0/3+2*ab1/3),
+        (r'$\frac{a_0-2|c_1|}{3}$',r'$\frac{a_0}{3}$',r'$\frac{a_0+2|c_1|}{3}$'),
+        color = 'k', size = 33)
 
 p.xlim(-1.2,3.2)
 p.ylim(-1.1+.5,1.1)
 
-p.ylabel(r"amplitude $\rightarrow$")
-p.xlabel(r"time $\rightarrow$")
+p.ylabel(r"amplitude $\rightarrow$", fontsize=19)
+p.xlabel(r"time $\rightarrow$", fontsize=19)
+p.savefig("../figures/amostras3b_.png")
 p.show()

### 2017-07-01 article/scripts/amostras4formasDeOnda.py

@@ -17,8 +17,11 @@ quadrada=n.hstack ((n.ones(T/2),n.ones(T/2)* -1))
 ondas=[senoide,dente,triangular,quadrada]
 ondas_n=[u"sinusoid","sawtooth","triangle","square"]
 
-for i in xrange(4):
-    p.subplot(int("22"+str(i+1)))
+fig=p.figure(figsize=(12.,6.))
+fig.subplots_adjust(left=0.04,bottom=0.13,right=0.99,top=0.95,wspace=0.14, hspace=0.37)
+subs=[]
+for i in range(4):
+    subs.append(fig.add_subplot(int("22"+str(i+1))))
     p.title(ondas_n[i])
     n4=ondas[i]
     p.plot(n4,"bo")
@@ -45,9 +48,6 @@ for i in xrange(4):
     else: fas2=0
     if a2<0: fas2+=n.pi # segundo e terceiro quadrantes somam pi
     print("abs: %s, a2: %s, b2: %s, fas2: %s" % (ab2,a2,b2,fas2))
-
-
-
 
     fr1=(2*n.pi/4)/2 #ciclo em 4 * \delta_a, metade do intervalo angular 
     fr2=1/2. # metade de \delta_a
@@ -76,8 +76,11 @@ for i in xrange(4):
 
     p.xlim(-1.2,4.2)
     p.ylim(-1.49,1.49)
-    p.xticks((-1,0,1,2,3,4),(r"$-\delta_a$",r"$0$",r"$\delta_a$",r"$2\delta_a$",r"$3\delta_a$",r"$4\delta_a$"))
-
-#p.ylabel(r"amplitude $\rightarrow$")
-#p.xlabel(r"tempo $\rightarrow$")
+    p.yticks((),())
+    p.xticks((-1,0,1,2,3,4),(r"$-\delta_a$",r"$0$",r"$\delta_a$",r"$2\delta_a$",r"$3\delta_a$",r"$4\delta_a$"),
+            size=17)
+subs[0].set_ylabel(r"amplitude $\rightarrow$",
+            {"y": -.1, "fontsize": 20})
+subs[2].set_xlabel(r"time $\rightarrow$", {"x": 1.11, "fontsize": 20})
+p.savefig("../figures/amostras4formas___.png")
 p.show()

### 2017-07-01 article/scripts/amostras6formas.py

@@ -17,8 +17,11 @@ quadrada=n.hstack ((n.ones(T/2),n.ones(T/2)* -1))
 ondas=[senoide,dente,triangular,quadrada]
 ondas_n=[u"sinusoid","sawtooth","triangle","square"]
 
-for i in xrange(4):
-    p.subplot(int("22"+str(i+1)))
+fig=p.figure(figsize=(12.,6.))
+fig.subplots_adjust(left=0.04,bottom=0.13,right=0.99,top=0.95,wspace=0.14, hspace=0.37)
+subs=[]
+for i in range(4):
+    subs.append(fig.add_subplot(int("22"+str(i+1))))
 #n4=n.array([ 0.58003705, -0.30828309, -0.29797696, -0.99219078])
     p.title(ondas_n[i])
     n6=ondas[i]
@@ -89,8 +92,14 @@ for i in xrange(4):
     p.xlim(-1.2,6.2)
     p.ylim(-1.45,1.45)
 
-    p.xticks((-1,0,1,2,3,4,5,6),(r"$-\delta_a$",r"$0$",r"$\delta_a$",r"$2\delta_a$",r"$3\delta_a$",r"$4\delta_a$",r"$5\delta_a$",r"$6\delta_a$"))
+    p.yticks((),())
+    p.xticks((-1,0,1,2,3,4,5,6),(r"$-\delta_a$",r"$0$",r"$\delta_a$",r"$2\delta_a$",r"$3\delta_a$",r"$4\delta_a$",r"$5\delta_a$",r"$6\delta_a$"),
+            size=15)
+subs[0].set_ylabel(r"amplitude $\rightarrow$",
+            {"y": -.1, "fontsize": 20})
+subs[2].set_xlabel(r"time $\rightarrow$", {"x": 1.11, "fontsize": 20})
 
     #p.ylabel(r"amplitude $\rightarrow$")
     #p.xlabel(r"tempo $\rightarrow$")
+p.savefig("../figures/amostras6formas___.png")
 p.show()

### 2017-07-01 article/scripts/delays.py

@@ -1,95 +1,115 @@
 #-*- coding: utf-8 -*-
 import numpy as n, pylab as p
 
-p.subplot(311)
+fig=p.figure(figsize=(10.,5.8))
+p.subplots_adjust(left=0.05,bottom=0.065,right=0.96,top=0.59, hspace=0.34)
+sub1=fig.add_subplot(311)
 fa=4410
 
 # Som curto
-sdur=0.005 # segundos
+sdur=0.003 # segundos
 f=200. #hz
-ii=fa*sdur
-s=n.sin(n.linspace(0,2*200*2*n.pi*sdur,ii, endpoint=False))
+ii=int(fa*sdur)
+s=n.sin(n.linspace(0,2*400*2*n.pi*sdur,ii, endpoint=False))
 
 # resposta ao impulso com um pulso
-ddur=0.02
-iii=fa*ddur
-xi=2*iii/3
+ddur=0.020
+iii=int(fa*ddur)
+xi=(1.5*iii)//3
 hi=n.zeros(iii)
 hi[xi]=1
 som = n.convolve(s,hi)
 
-p.plot(s+6,'bo', label=r"$\{t_i\}_0^{\Lambda-1}$")
-p.plot(hi+3,'ro', label=r"$\{h_i=\delta_{\xi-i}\}_0^{\Lambda_h-1}$")
-p.plot(som,'go', label=r"$\{(t*h)_i\}_0^{\Lambda+\Lambda_h-2}$")
-p.legend(loc="upper right")
-p.xticks((0,ii-1,xi-1,xi+ii-1,ii+iii-2, iii-1),(r"0", r"$\Lambda-1$",r"$\xi$", r"$\xi + \Lambda-1$", r"$\Lambda+\Lambda_h-2$", r"$\Lambda_h$"),fontsize='16')
+p1=sub1.plot(s+6,'bo', label=r"$\{t_i\}_0^{\Lambda-1}$", ms=3)
+p2=sub1.plot(hi+3,'ro', label=r"$\{h_i=\delta_{\xi-i}\}_0^{\Lambda_h-1}$",
+        ms=3)
+p3=sub1.plot(som,'go', label=r"$\{(t*h)_i\}_0^{\Lambda+\Lambda_h-2}$", ms=3)
+# p.legend(loc="upper right",numpoints=5)
+p.xticks((0,ii-1,xi,xi+ii-1,ii+iii-2, iii-1),(r"0",
[…]

### 2017-07-01 article/scripts/mixa_concatena.py

@@ -21,22 +21,24 @@ s=n.linspace(0,10*2*n.pi,200,endpoint=False)
 s=n.sin(s)[:len(s)/c]
 e=5.0
 ee=4
+p.figure(figsize=(10.,5.))
+p.xticks((),())
+p.yticks((),())
+p.subplots_adjust(left=0.05,bottom=0.08,right=0.97,top=0.97)
 p.plot(s+ee+3*e,'o')
-p.plot(q+ee+2*e,'o')
-p.plot(d+ee+e,'o')
+p.plot(q+ee+1.93*e,'o')
+p.plot(d+ee+e*.8,'o')
 p.plot(s+q+d,'o')
 
 p.ylim(-3.5,21)
 p.xlim(-5,105)
 
-p.text(47,11.4-.5,"+",fontsize=48)
-p.text(47,11.4+e-.5,"+",fontsize=48)
-p.text(46,11.4-e-2,"=",fontsize=78)
-
-p.xticks((),())
-p.yticks((),())
+p.text(47,11.4+e-.75,"+",fontsize=48)
+p.text(47,11.4-1.65,"+",fontsize=48)
+p.text(45.5,11.4-e-3,"=",fontsize=68)
 
 p.xlabel(r"time $\rightarrow$",fontsize=22)
 p.ylabel(r"amplitude $\rightarrow$",fontsize=22)
 
+p.savefig("../figures/mixagem__.png")
 p.show()

### 2017-07-02 article/acm/body.tex
melhoradas figuras do artigo

@@
         \includegraphics[width=\textwidth]{figures/delays_}
-    \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c). Represented in increasing order of its pulse density.}
+    \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c).
+Represented in increasing order of its pulse density.
+The vertical axis is related to amplitude although one should keep in mind that each plot has two or three displaced signals.}
         \label{fig:delays}
@@
     \centering
-        \includegraphics[width=\textwidth]{figures/iir_}
+        \includegraphics[width=\textwidth]{figures/iir__}
     \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of equations~\ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} and \ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
@@
 	\hspace*{-.75cm}
-         \includegraphics[width=1.1\textwidth]{figures/ruidos}
-     \caption{Colors of noise generated by equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul} and \ref{eq:violeta}: spectrum and waveforms.}
+         \includegraphics[width=1.\textwidth]{figures/ruidos_}
+     \caption{Colors of noise generated by equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul} and \ref{eq:violeta}: spectrum and example waveforms.}
          \label{fig:ruidos}
@@
      \centering
-         \includegraphics[width=\columnwidth]{figures/vibrato}
+         \includegraphics[width=\columnwidth]{figures/vibrato_}
      \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave, with $f_s=44.1kHz$.}
@@
      \centering
-         \includegraphics[width=\textwidth]{figures/tremolo}
+         \includegraphics[width=\textwidth]{figures/tremolo_}
      \caption{Tremolo with a depth of $V_{dB}=12dB$, with a sawtooth waveform as its oscillatory pattern, with $f'=1.5Hz$ in a sine of $f=40Hz$ (sample frequency $f_s=44.1kHz$).}
@@
     \centering
-        \includegraphics[width=\columnwidth]{figures/adsr}
+        \includegraphics[width=\columnwidth]{figures/adsr_}
     \caption{An ADSR envelope (\emph{Attack, Decay, Sustain, Release}) applied to an arbitrary sound sequence.

### 2017-07-02 article/scripts/vibrato.py

@@ -1,5 +1,5 @@
 #-*- coding: utf-8 -*-
-import numpy as n, pylab as p, scikits.audiolab as a
+import numpy as n, pylab as p#, scikits.audiolab as a
 
 fa=44100 # frequência de amostragem
 Dv=2048 # tamanho da tabela do vibrato
@@ -24,19 +24,24 @@ tab=n.linspace(-1,1,D) # dente de serra
 dD=fi*(D/float(fa)) # a movimentação na tabela por amostra
 gi=n.cumsum(dD,0,n.int) # a movimentação na tabela total, já os índices
 ti=tabv[gi%D] # busca dos índices na tabela
-a.wavwrite(ti,"vibrato.wav",fa) # escrita do som
+# a.wavwrite(ti,"vibrato.wav",fa) # escrita do som
 
 gi=n.array(  ii * (D/float(fa)) * f  , n.int ) % Dv
 t=tabv[ gi ]
-a.wavwrite(t,"original.wav",fa)
+# a.wavwrite(t,"original.wav",fa)
 
+p.figure(figsize=(10.,5.))
+p.subplots_adjust(left=0.17,bottom=0.15,right=0.97,top=0.97)
 p.specgram(ti-ti.mean())
 p.colorbar()
 p.xlim(-2000,46100)
-p.xticks((0,10000,20000,30000,44000),(r"0",10000,20000,30000,44100))
+p.xticks((0,10000,20000,30000,44000),(r"0",10000,20000,30000,44100),
+        fontsize=16)
+p.yticks(fontsize=16)
+p.ylabel(u"frequency "+r"$ \in \; [0,\,\frac{f_a=44100}{2}=22050] \, \rightarrow $", fontsize=19)
+p.xlabel(r"samples $\quad \rightarrow$",fontsize=26)
+p.savefig("../figures/vibrato_.png")
 p.show()
 
-p.ylabel(u"frequency "+r"$ \in \; [0,\,\frac{f_a=44100}{2}=22050] \quad \rightarrow $", fontsize=16)
-p.xlabel(r"$i\quad \rightarrow$",fontsize=26)
 
 

### 2017-07-02 article/acm/body.tex
terminada revisao inicial de todas as figuras e tabelas

@@
     \centering
-        \includegraphics[width=\columnwidth]{figures/movContraponto}
+        \includegraphics[width=.8\columnwidth]{figures/movContraponto_}
     \caption{Different motions of counterpoint aiming to preserve independence
         between voices. There are 3 types of motion: direct, contrary and
-        oblique. The parallel motion is a type of direct motion.}
+        oblique. The parallel motion is a type of direct motion.
+	A voice is often called a 'melodic line' or a melody.}
         \label{fig:movContraponto}
@@
 \begin{table*}[htp!]
-\tiny
-\centering
-\caption{Transition of durations heard individually until it turns into pitch.}
-\begin{tabular}{  l | r r r r   r r r    r r r || r r || r r r r r r }
-\hline
-           & \multicolumn{10}{c}{$\underleftarrow{\text{\bf perception zone of
-           duration as rhythm}}$} & \multicolumn{2}{c}{transition} & \multicolumn{3}{c}{-} \\
-duration (s) & {\bf ...}     & {\bf 32,}     & {\bf 16,}   & {\bf 8,}  & {\bf 4,}   & {\bf 2,}   & {\bf 1,}   & {\bf 1/2,} & {\bf 1/4,} & {\bf 1/8,} & $\frac{1}{16}=62,5ms$ , & $\frac{1}{20}=50ms$ & {\color{gray} 1/40} & {\color{gray} 1/80  } & {\color{gray} 1/160 } & {\color{gray} 1/320 } & {\color{gray} 1/640 } & {\color{gray} ...
-} \\
-frequency (Hz) & {\color{gray} ...} & {\color{gray} 1/32,}   & {\color{gray} 1/16,} & {\color{gray} 1/8,} & {\color{gray} 1/4,} & {\color{gray} 1/2,} &  {\color{gray} 1,}  & {\color{gray} 2,}   & {\color{gray} 4,}   & {\color{gray} 8,}    & 16,  & 20   & {\bf 40}   & {\bf 80}   & {\bf 160}   & {\bf 320}   & {\bf 640}   & {\bf ...} \\
-           & \multicolumn{10}{c}{ - } & \multicolumn{2}{c}{transition}
-           & \multicolumn{6}{c}{$\overrightarrow{\text{\bf perception zone of
-           duration as pitch}}$} \\
-\hline
+\caption{Transition of durations heard as rhythm until heard as pitch.}
+\begin{tabular}{  l | r r r r   r r r    r r r || r r  } \hline
+& \multicolumn{10}{c}{\bf perception of durations as rhythm} & \multicolumn{2}{c}{}  \\
+	duration (s) & \multirow{2}{*}{\bf ...}     & {\bf 32,}     & {\bf 16,}   & {\bf 8,}  & {\bf 4,}   & {\bf 2,}   & {\bf 1,}   & {\bf 1/2,} & {\bf 1/4,} & {\bf 1/8,} & \multirow{2}{*}{ ...
+transition } & \\
+frequency (Hz) & & {\color{gray} 1/32,}   & {\color{gray} 1/16,} & {\color{gray} 1/8,} & {\color{gray} 1/4,} & {\color{gray} 1/2,} &  {\color{gray} 1,}  & {\color{gray} 2,}   & {\color{gray} 4,}   & {\color{gray} 8,} & & \\
+& \multicolumn{10}{c}{ - } & \multicolumn{2}{c}{} \\ \hline
 \end{tabular}
+\vspace{.2cm}
+
+	\begin{tabular}{  l | p{1.7cm}  || r r || r } \hline
+& \multicolumn{1}{c}{} & \multicolumn{2}{c}{transition} &  \\
+duration (s) & \multirow{2}{*}{\bf rhythm ...} & $\frac{1}{16}=62.5ms$ , & $\frac{1}{20}=50ms$ & \multirow{2}{*}{\bf ... pitch} \\
+frequency (Hz) & & 16, & 20 &  \\
+& \multicolumn{1}{c}{} & \multicolumn{2}{c}{transition} & \\ \hline
+\end{tabular}
+\hspace{3.36cm}
+\vspace{.2cm}
+
+	\begin{tabular}{  l | p{1.7cm} || r r r r r r }\hline
+& \multicolumn{1}{c}{} &  \\
+	duration (s) & \multirow{2}{*}{transition ...} & {\color{gray} 1/40} & {\color{gray} 1/80  } & {\color{gray} 1/160 } & {\color{gray} 1/320 } & {\color{gray} 1/640 } & \multirow{2}{*}{\bf ... } \\
+frequency (Hz) & & {\bf 40}   & {\bf 80}   & {\bf 160}   & {\bf 320}   & {\bf 640}   & \\
+& \multicolumn{1}{c}{} & \multicolumn{6}{c}{\bf perception of durations as pitch} \\ \hline
+\end{tabular}
+\hspace{2.70cm}
 \label{tab:duracoes}
@@
     \centering
-        \includegraphics[width=.9\textwidth]{figures/metricaMusical_}
+        \includegraphics[width=.9\textwidth]{figures/metricaMusical__}
     \caption{Divisions and groupings of the musical pulse for establishing a metric. Divisions of the quarter note, regarded as the
@@
     \centering
-        \includegraphics[width=\columnwidth]{figures/climax}
+        \includegraphics[width=.8\columnwidth]{figures/climax_}
         \caption{Canonical distinctions of musical climax in a given melody and

### 2017-07-02 article/scripts/climaxMel.py

@@ -1,7 +1,8 @@
 #-*- coding: utf-8 -*-
 import numpy as n, pylab as p
 
-
+p.figure(figsize=(10.,5.))
+p.subplots_adjust(left=0.04,bottom=0.06,right=0.99,top=0.99)
 c1=n.hstack(( n.linspace(0,1,1000,endpoint=False),n.linspace(1,0,1000)    )) # simetrico
 c2=n.hstack(( n.linspace(0,1,500,endpoint=False), n.linspace(1,0,1500)    ))
 c3=n.hstack(( n.linspace(0,1,1500,endpoint=False),n.linspace(1,0,500)    ))
@@ -13,7 +14,7 @@ p.plot(c2,label=u"climax at first half",lw=5)
 p.plot(c3,label=u"climax at second half", lw=5)
 p.plot(c4,label=u"climax at beginning",         lw=5)
 p.plot(c5,label=u"climax at end",            lw=5)
-p.legend(loc="lower center",prop={'size':22})
+p.legend(loc="lower center",prop={'size':16})
 
 p.xlim(-50,2050)
 p.ylim(-.18,1.02)
@@ -22,7 +23,8 @@ p.xticks((),())
 p.yticks((),())
 
 p.xlabel(r"time $\rightarrow$", fontsize=19)
-p.ylabel(u"parâmetro"+r"$\rightarrow$", fontsize=19)
+p.ylabel(u"parameter"+r"$\rightarrow$", fontsize=19)
 
+p.savefig("../figures/climax_.png")
 p.show()
 

### 2017-07-02 article/scripts/movimentosVozes.py

@@ -2,7 +2,7 @@
 import numpy as n, pylab as p
 
 p.figure(figsize=(10.,5.))
-p.subplots_adjust(left=0.17,bottom=0.15,right=0.97,top=0.97)
+p.subplots_adjust(left=0.04,bottom=0.08,right=0.95,top=0.99)
 ax = p.subplot(111)
 
 ##### movimentos discernidos no contraponto
@@ -67,16 +67,16 @@ for n1,n2 in zip(voz1,voz2):
 
 
 
-p.plot((1000,1000),(n1,n1),'b',lw=6, label=u"melodic line 1")
-p.plot((1000,1000),(n2,n2),'g',lw=6, label=u"melodic line 2")
-p.legend(loc="upper right",prop={'size':22})
+p.plot((1000,1000),(n1,n1),'b',lw=6, label=u"voice 1")
+p.plot((1000,1000),(n2,n2),'g',lw=6, label=u"voice 2")
+p.legend(loc="upper right",prop={'size':14})
 
 p.ylim(-10,3.6)
 p.xlim(-0.2,3.8)
 
-p.yticks((-8,-4,1),(u"oblique",u"opposite","direct"),rotation="90",fontsize=22)
+p.yticks((-8,-4,1),(u"oblique",u"contrary","direct"),rotation="90",fontsize=16, fontweight='bold')
 
-p.text(2.8,-1.6,"parallel\nmovement",fontsize="17")
+p.text(2.8,-1.6,"parallel\nmovement",fontsize="14")
 
 for line in ax.get_xticklines() + ax.get_yticklines():
     line.set_markersize(0)
@@ -89,5 +89,6 @@ ax2.set_ylabel(r"$\log(freq)$ $\rightarrow$",fontsize=26)
 p.ylim(-10,3.6)
 p.yticks(())
 
+p.savefig("../figures/movContraponto_.png")
 p.show()
 

### 2017-07-02 article/scripts/vibrato.py

@@ -42,6 +42,3 @@ p.ylabel(u"frequency "+r"$ \in \; [0,\,\frac{f_a=44100}{2}=22050] \, \rightarrow
 p.xlabel(r"samples $\quad \rightarrow$",fontsize=26)
 p.savefig("../figures/vibrato_.png")
 p.show()
-
-
-

### 2017-07-02 article/acm/body.tex
melhoradas no artigo. Acabei de verificar que preciso incorporar as revisoes do Chu

@@
 \section{Introduction}\label{sec:level1}
-
-Music is commonly defined as the art made by sounds and silences, where sound corresponds to the longitudinal wave of mechanical pressure.
-The human hearing system perceives sounds within the frequency bandwidth between $20Hz$ and $20kHz$, with the actual limits depending on the person, climate conditions and the sonic characteristics themselves~\cite{Roederer}.
-Since the speed of sound is $\approx 343.2 m/s$, these limits imply wavelengths of $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.
-Such perception involves stimuli in bones, stomach, ears, transfer functions of head and torso, and processing by the nervous system~\cite{Roederer}.
-Obviously, the ear is a dedicated organ for the appreciation of these waves, which decomposes them into their sinusoidal spectra and delivers to the nervous system.
-The sinusoidal components are crucial to musical phenomena, as one can perceive in the composition of sounds of musical interest (such as harmonic sounds and noises, discussed in sections~\ref{sec:discNote} and~\ref{sec:internalVar}), and sonic structures of musical interest (such as tunings and scales, in section~\ref{sec:notesMusic}).
-
-The representation of sound is commonly referred to as audio, although these terms are often used without distinction.
-Audio expresses waves from the capture by microphones or from direct synthesis, although these sources are not neatly distinguishable as captured sounds are processed to generate new sonorities.
-Digital audio specified by protocols that facilitate file transferring and storage often implies a quality loss.
-Standard representation of digital audio, on the other hand, assures perfect reconstruction of the analog wave, within any convenient precision.
-This paradigm consists of representing the audio with equally spaced samples, of $\lambda_s$ durations, each specified by a fixed number of bits.
+Music is usually defined as the art whose medium is sound.
+The definition might also state that the medium includes silences
+and temporal organization of structures, or that music
+is also a cultural activity or product~\cite{}.
+Sound is a physical phenomenon which corresponds to the longitudinal wave of mechanical pressure.
+The human auditory system perceives sounds within the frequency bandwidth between $20Hz$ and $20kHz$,
+with the actual boundaries depending on the person,
+climate conditions and the sonic characteristics themselves~\cite{Roederer}.
+Since the speed of sound is $\approx 343.2 m/s$,
+such frequency limits imply wavelengths of $\frac{343.2}{20} \approx 17.16\,m$ and
+$\frac{343.2}{20000} \approx 17.16\,mm$.
+Hearing involves stimuli in bones, stomach, ears,
+transfer functions of head and torso,
+and processing by the nervous system~\cite{Roederer}.
+The ear is a dedicated organ for the appreciation of these waves,
+which decomposes them into their sinusoidal spectra and delivers to the nervous system.
+The sinusoidal components are crucial to musical phenomena,
+as one can perceive in the constitution of sounds of musical interest
+(such as harmonic sounds and noises, discussed in sections~\ref{sec:discNote} and~\ref{sec:internalVar}),
+and higher level musical structures (such as tunings, scales and chords, in section~\ref{sec:notesMusic}).
+
+The representation of sound can take many forms,
+from musical scores and phonetic texts to electric analog signals and binary data,
+which includes arbitrary sets of features and spectral decompositions such as obtained through wavelets.
+When the representation mimics the mechanical wave directly, it might be referred to as audio,
+although the terms audio and sound are often used without distinction
+and has many definitions that depends on the context and the author~\cite{audioSound}.
+In this definition, which contemplates most common uses of the word,
+audio expresses sonic waves from direct synthesis or input by microphones,
+although these sources are not always neatly distinguishable
+e.g. as captured sounds are processed to generate new sonorities.
+Digital audio protocols that facilitate file
+storage and transfer often implies quality loss and are called \emph{lossy}.
+This is the case e.g. of MP3 and Ogg Vorbis~\cite{loss}.
+Non-lossy representations of digital audio, called \emph{lossless}, on the other hand,
+assures perfect reconstruction of the analog wave within any convenient precision.
+The standard paradigm of lossless audio consists of
+representing the sound with samples equally spaced
+by a duration $\lambda_s$, and specifying each sample by a fixed number of bits.
 This is the Pulse Code Modulation (PCM) representation of sound.
-A sound in PCM audio is characterized by a sampling frequency $f_s=\frac{1}{\lambda_s}$ (also called the sampling rate), which is the number of samples used for representing a second of sound; and by a bit depth, which is the number of bits used for representing the amplitude of each sample.
-Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with $4$ bits each.
-The fixed $2^4=16$ values for the amplitude of each sample, with the regular spacing $\lambda_s$, yields the quantization error or noise.
-This noise diminishes as the bit depth increases.
+A PCM audio format has two essential attributes:
+a sampling frequency $f_s=\frac{1}{\lambda_s}$ (also called the sampling rate),
+which is the number of samples used for representing a second of sound;
+and a bit depth, which is the number of bits used for representing the amplitude of each sample.
+Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with a bit depth of $4$,
+which yields $2^4=16$ possible values for the amplitude of each sample
+and a total of $4 \times 25= 100$ bits for representing the whole sound.
 
@@
         \includegraphics[width=.7\textwidth]{figures/pcm}
-        \caption{Pulse Code Modulation (PCM) audio: an analogical signal is represented by 25 samples with 4 bits each.}
+	\caption{Example of PCM audio: a sound wave is represented by 25 samples where each sample has an amplitude specified with 4 bits.}
         \label{fig:PCM}
@@
 
-The Nyquist theorem~\cite{Openheim} states that the sampling frequency is twice the maximum frequency of the represented signal.
-Thus, for general musical purposes, it is necessary to have samples in a rate at least twice the highest frequency heard by humans, that is, $f_s \geq 2\times 20kHz = 40kHz$.
-This is the fundamental reason for the adoption of sampling frequencies such as $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and broadcast systems (radio and television), respectively.
-
-Within this framework, musical notes can be characterized.
-The note still stands paradigmatic as the 'fundamental unit' of musical structures and, in practice, it can unfold into sounds that uphold other approaches\footnote{Music from the twentieth century enlarged this traditional comprehension of music.
-This occurred in the concert music, especially in the concrete, electronic and electroacoustic styles.
-In the 1990s, it became evident that popular music had also incorporated sounds without defined pitch and temporal organization, to name a few of simple characteristics.}.
-    Notes are also convenient for another reason: the average listener -- and considerable part of the specialists -- presupposes rhythmic and pitch organization (made explicit in section~\ref{sec:notesMusic}) as fundamental musical properties, and these are developed in traditional musical theory in terms of notes.
+The fixed sampling frequency and bit depth
+yields the quantization error or quantization noise.
+This noise diminishes as the bit depth increases
+while greater sampling frequency allows higher frequencies to be represented.
+The Nyquist theorem~\cite{Openheim} asserts that the sampling frequency
+is twice the maximum frequency of the represented signal.
+Thus, for general musical purposes, it is necessary to have samples in a rate
+at least twice the highest frequency heard by humans,
+that is, $f_s \geq 2\times 20kHz = 40kHz$.
+This is the fundamental reason for the adoption of sampling frequencies
+such as $44.1kHz$ and $48kHz$,
+which are the standards in Compact Disks (CD)
+and broadcast systems (radio and television), respectively.
+
+Within this framework for representing sounds,
+musical notes can be characterized.
+The note often stands as the 'fundamental unit' of musical structures
+(such as atoms in matter or cells in macroscopic organisms) and,
+in practice, it can unfold into sounds that uphold other approaches.
+This is of capital importance because science and scholastic artists
+widened the traditional comprehension of music in the twentieth century
+to encompass concepts beyond e.g. rhythm, melody and harmony.
+This is very evident e.g. in the concrete, electronic, electroacoustic,
+and spectral musical styles.
+In the 1990s, it became evident that popular music had also incorporated
+sounds amalgams and abstract discursive arcs,
+to name just a couple of simple characteristics.
+Notes are also convenient for another reason:
+the average listener -- and a considerable part of the specialists --
+presupposes rhythmic and pitch organization 
+(made explicit in section~\ref{sec:notesMusic})
+as fundamental musical properties,
+and these are developed in traditional musical theory in terms of notes.
+Thereafter, in this article we specify musical notes in PCM audio through equations
+and then indicate mechanisms for deriving higher level musical structures.
+We understand that this is not the unique approach to mathematically express music in
+digital audio, but musical theory and practical issues suggest that
+this is a proper framework for understanding and making computer music,
+as should become evident in the reminder of this text and be verifiable
+by usage of the \massa\ toolbox.
 
 \subsection{Contributions and paper organization}
-
-This article aims at representing musical structures and artifices by their discrete-time sonic characteristics.
-Results include mathematical relations, usually in terms of samples, and their direct computer program implementations.
-Despite the general interests involved, there are few books and computer implementations that tackle the subject.
-These mainly focus on computer implementations and ways to mimic traditional instruments, with scattered mathematical formalisms.
+This article aims at representing musical structures by their discrete-time sonic characteristics.
+Results include mathematical relations, usually in terms of musical characteristics and samples,
+concise musical theory considerations,
+and their implementations as software routines both as
+very raw and direct algorithms and in the context of rendering musical pieces.
+Despite the general interests involved,
+there are some books and computer implementations that tackle the subject.
+These mainly focus on computer implementations and ways to mimic traditional instruments,
+with scattered mathematical formalisms.
 A compilation of the works and their contributions is in the bibliography~\cite{dissertacao}.
-To the best of the author's knowledge, there is a lack of articles on the topic.
-Moreover, although current computer implementations use the analytical descriptions presented in this study in a implicit manner, it seems that there has been no concise and mathematical description of the processes implemented.
-
-In order to address this concise description of musical elements and structures, in terms of the digitalized sound, the objectives of this paper are:
+To the best of the authors' knowledge,
+although current music software
+use the analytical descriptions presented in this study,
+it seems that there has been no concise and mathematical description of them.
+The objectives of this paper are:
 
 \begin{enumerate}
-
 	\item Present a concise set of relations among musical basic elements and sequences of PCM audio samples. 
-
 	\item Introduce a framework of sound synthesis with control at sample level and with potential uses in psycho-acoustical experiments and high-fidelity synthesis.
-
 	\item Provide the accessibility of the developed framework.
@@
 This way, the work presented here is embedded in recommended practices for availability and eases co-authorship processes~\cite{Raymond,Lessig}.
-
 	\item To provide a didactic presentation of the content to favor its apprehension and usage.
 It is worthwhile to mention that this subject comprises diverse topics on signal processing, music, psycho-acoustics and programming.
-
 \end{enumerate}

### 2017-07-02 article/acm/sample-acmsmall.tex
terminada revisao inicial de todas as figuras e tabelas
melhoradas no artigo. Acabei de verificar que preciso incorporar as revisoes do Chu

@@
 \usepackage{booktabs} % For formal tables
+\usepackage{multirow}
 
@@
 \acmJournal{TWEB}
-\acmVolume{9}
-\acmNumber{4}
-\acmArticle{39}
-\acmYear{2010}
-\acmMonth{3}
+\acmVolume{0}
+\acmNumber{0}
+\acmArticle{0}
+\acmYear{0}
+\acmMonth{0}
 \copyrightyear{2009}
@@
 \affiliation{%
-  \institution{University of São Paulo}
+  \institution{University of S\~ao Paulo}
   \department{Institute of Mathematics and Computer Sciences}
-  \streetaddress{Avenida Trabalhador São Carlense, 400 - Centro}
-  \city{São Carlos}
+  \streetaddress{Avenida Trabalhador S\~ao Carlense, 400 - Centro}
+  \city{S\~ao Carlos}
   \state{SP}
@@
 \affiliation{%
-  \institution{University of São Paulo}
-  \department{São Carlos Institute of Physics}
-  \streetaddress{Avenida Trabalhador São Carlense, 400 - Centro}
-  \city{São Carlos}
+  \institution{University of S\~ao Paulo}
+  \department{S\~ao Carlos Institute of Physics}
+  \streetaddress{Avenida Trabalhador S\~ao Carlense, 400 - Centro}
+  \city{S\~ao Carlos}
   \state{SP}
@@
 \begin{abstract}
-Notes, ornaments and intervals are examples of basic musical elements.
-Their representation as discrete-time digital audio plays a central role in software for music creation and design.
-Nevertheless, there is no systematic relation, in analytical terms, of these musical elements to the sonic samples.
-Such a compendium enables scientific experiments in precise and trustful ways, among educational and artistic uses.
-This paper presents a comprehensive description of music in digital audio, within an unified approach.
-Musical elements, like pitch, duration and timbre are expressed by equations on sample level.
-This quantitatively relates characteristics of the discrete-time signal to musical qualities.
-Internal variations, e.g. tremolos, vibratos and spectral fluctuations, are also considered
-to operate within a note. Moreover, the generation of musical structures such as rhythmic meter,
-pitch intervals and cycles, are attained canonically with notes. 
-The availability of these resources in scripts is provided in public domain within the \massa\ toolbox - Music and Audio in Sequences and Samples.
-Authors observe that the implementation of sample-domain analytical results as open source can encourage concise research.
-As further illustrated in the paper, \massa\ has already been employed by users for diverse purposes, 
-including acoustics experimentation, art and education. The efficacy of these physical descriptions 
-was confirmed by the synthesis of small musical pieces.
-As shown, it is possible to synthesize whole albums through collage of scripts and parametrization.
+The representation of the basic elements of music - such as notes,
+ornaments and intervalar structures - in terms of discrete audio
+signals is often used in software for music creation and design.
+Nevertheless, there is no unified approach that relates these elements to the sound discrete samples.
+In this article, each musical element is related by equations to the discrete-time samples of sounds,
+and each of these relations are implemented in scripts within a software toolbox,
+referred to as \massa\ (Music and Audio in Sequences and Samples).
+The toolbox also includes this article and every script necessary to render the images
+and musical examples, pieces and album, which adds to the educational and artistic uses.
+The fundamental element, the musical note with duration, volume, pitch and timbre,
+is related quantitatively to the characteristics of the digital signal.
+Internal variations, such as tremolos, vibratos and spectral fluctuations,
+are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities.
+With this representation of notes, resources are provided for the generation of further musical structures,
+such as rhythmic meter, pitch intervals and cycles.
+Such a framework enables precise and trustful scientific experiments and is useful for education and art.
+The efficacy of \massa\ is confirmed by the synthesis of small musical pieces using basic notes,
+incremented notes and notes in music, which reflects the organization of the toolbox and of this article.
+It is possible to synthesize whole albums through collage of the scripts and parameterization specified by the user.
+With the paradigm of open source implementation, \massa\ toolbox can be promptly scrutinized, expanded in co-authorship processes and used freely by musicians, engineers and other interested parties.
+In fact, \massa\ has already been employed by external users for diverse purposes which include music production,
+artistic presentations, psychoacoustic experiments and computer language diffusion where the appeal of audiovisual artifacts is exploited for education.
 \end{abstract}
@@
 
-\keywords{digital audio, music, signal processing, acoustics, psychophysics}
+\keywords{music, acoustics, psychophysics, digital audio, signal processing}
 

### 2017-07-02 article/acm/body.tex
finalizada revisao da intro

@@
 Sound is a physical phenomenon which corresponds to the longitudinal wave of mechanical pressure.
-The human auditory system perceives sounds within the frequency bandwidth between $20Hz$ and $20kHz$,
+The human auditory system perceives sounds in the frequency bandwidth between $20Hz$ and $20kHz$,
 with the actual boundaries depending on the person,
@@
 The sinusoidal components are crucial to musical phenomena,
-as one can perceive in the constitution of sounds of musical interest
-(such as harmonic sounds and noises, discussed in sections~\ref{sec:discNote} and~\ref{sec:internalVar}),
-and higher level musical structures (such as tunings, scales and chords, in section~\ref{sec:notesMusic}).
+as one can recognize in the constitution of sounds of musical interest
+(such as harmonic sounds and noises, discussed in Sections~\ref{sec:discNote} and~\ref{sec:internalVar}),
+and higher level musical structures (such as tunings, scales and chords, in Section~\ref{sec:notesMusic}).
 
 The representation of sound can take many forms,
-from musical scores and phonetic texts to electric analog signals and binary data,
-which includes arbitrary sets of features and spectral decompositions such as obtained through wavelets.
-When the representation mimics the mechanical wave directly, it might be referred to as audio,
-although the terms audio and sound are often used without distinction
-and has many definitions that depends on the context and the author~\cite{audioSound}.
-In this definition, which contemplates most common uses of the word,
+from musical scores and texts in a phonetic language to electric analog signals and binary data.
+It includes arbitrary sets of features and spectral decompositions such as obtained through wavelets.
+Although the terms 'audio' and 'sound' are often used without distinction
+and have many definitions that depend on the context and the author~\cite{audioSound},
+audio most often means a representation of sound that
+depicts the amplitude of the mechanical wave through time.
+In this sense,
 audio expresses sonic waves from direct synthesis or input by microphones,
@@
 e.g. as captured sounds are processed to generate new sonorities.
-Digital audio protocols that facilitate file
-storage and transfer often implies quality loss and are called \emph{lossy}.
+Digital audio protocols often implies quality loss (to yield smaller files, ease storage and transfer)
+and are called \emph{lossy}.
 This is the case e.g. of MP3 and Ogg Vorbis~\cite{loss}.
-Non-lossy representations of digital audio, called \emph{lossless}, on the other hand,
-assures perfect reconstruction of the analog wave within any convenient precision.
+Non-lossy representations of digital audio, called \emph{lossless} protocols or formats,
+on the other hand, assures perfect reconstruction of the analog wave within any convenient precision.
 The standard paradigm of lossless audio consists of
 representing the sound with samples equally spaced
-by a duration $\lambda_s$, and specifying each sample by a fixed number of bits.
+by a duration $\lambda_s$, and specifying the amplitude of each sample by a fixed number of bits.
 This is the Pulse Code Modulation (PCM) representation of sound.
 A PCM audio format has two essential attributes:
-a sampling frequency $f_s=\frac{1}{\lambda_s}$ (also called the sampling rate),
+a sampling frequency $f_s=\frac{1}{\lambda_s}$ (also called e.g. sampling rate or sample rate),
 which is the number of samples used for representing a second of sound;
-and a bit depth, which is the number of bits used for representing the amplitude of each sample.
+and a bit depth, which is the number of bits for each sample.
 Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with a bit depth of $4$,
@@
         \includegraphics[width=.7\textwidth]{figures/pcm}
-	\caption{Example of PCM audio: a sound wave is represented by 25 samples where each sample has an amplitude specified with 4 bits.}
+	\caption{Example of PCM audio:
+	a sound wave is represented by 25 samples where each sample has an amplitude specified with 4 bits.}
         \label{fig:PCM}
@@
 The Nyquist theorem~\cite{Openheim} asserts that the sampling frequency
-is twice the maximum frequency of the represented signal.
-Thus, for general musical purposes, it is necessary to have samples in a rate
+is twice the maximum frequency that the represented signal can contain.
+Thus, for general musical purposes, it is necessary to use a sample rate of
 at least twice the highest frequency heard by humans,
@@
 widened the traditional comprehension of music in the twentieth century
-to encompass concepts beyond e.g. rhythm, melody and harmony.
-This is very evident e.g. in the concrete, electronic, electroacoustic,
+to encompass discourse without explicit rhythm, melody or harmony.
+This is very evident in e.g. in the concrete, electronic, electroacoustic,
 and spectral musical styles.
-In the 1990s, it became evident that popular music had also incorporated
-sounds amalgams and abstract discursive arcs,
-to name just a couple of simple characteristics.
+In the 1990s, it became evident that popular
+(commercial) music had also incorporated
+sounds amalgams and abstract discursive arcs~\footnote{There
+are well known incidence of such musical techniques in ethnic music,
+such as in Pygmy music, but western music theory assimilated them
+only in the last century~\cite{Wisnik}.}.
 Notes are also convenient for another reason:
@@
 presupposes rhythmic and pitch organization 
-(made explicit in section~\ref{sec:notesMusic})
+(made explicit in Section~\ref{sec:notesMusic})
 as fundamental musical properties,
 and these are developed in traditional musical theory in terms of notes.
-Thereafter, in this article we specify musical notes in PCM audio through equations
+Thereafter, in this article we describe musical notes in PCM audio through equations
 and then indicate mechanisms for deriving higher level musical structures.
@@
 by usage of the \massa\ toolbox.
-
-\subsection{Contributions and paper organization}
-This article aims at representing musical structures by their discrete-time sonic characteristics.
-Results include mathematical relations, usually in terms of musical characteristics and samples,
+Hopefully, the interested reader or programmer will be able to use
+this framework to synthesize music beyond traditional conceptualizations when intended.
+
+% \subsection{Contributions and paper organization}
+This article aims at representing musical structures and artifices
+by their discrete-time sonic characteristics.
+The results include mathematical relations,
+usually in terms of musical characteristics and samples,
 concise musical theory considerations,
@@
 Despite the general interests involved,
-there are some books and computer implementations that tackle the subject.
+there are only a few books and computer implementations that tackle the subject.
 These mainly focus on computer implementations and ways to mimic traditional instruments,
@@
 A compilation of the works and their contributions is in the bibliography~\cite{dissertacao}.
-To the best of the authors' knowledge,
-although current music software
-use the analytical descriptions presented in this study,
-it seems that there has been no concise and mathematical description of them.
+Articles on the topic appear to be lacking, to the best of our knowledge.
+Although current music software use the analytical descriptions presented here,
+there is no concise mathematical description of them, and it is far from trivial
+to achieve the equations by analyzing the available software implementations.
+
 The objectives of this paper are:
@@
 \begin{enumerate}
-	\item Present a concise set of relations among musical basic elements and sequences of PCM audio samples. 
-	\item Introduce a framework of sound synthesis with control at sample level and with potential uses in psycho-acoustical experiments and high-fidelity synthesis.
-	\item Provide the accessibility of the developed framework.
-The analytic relations presented in this article are implemented as public domain scripts, i.e.
-small computer programs using accessible technologies for better distribution and validation.
-This constitute the \massa\ toolbox, available in an open Git repository~\cite{gitBook}.
-These scripts are written in Python and make use of external libraries, such as Numpy that performs calls to Fortran routines for better computational efficiency.
-Part of the scripts have been transcribed to JavaScript and native Python with readiness, what favors their use in Web browsers such as Firefox and Chromium~\cite{numpy, audiolab, tutpython, python}.
-Furthermore, these are all open technologies, that is, published using licenses that allow copy, distribution and use of any part for research and derivatives.
-This way, the work presented here is embedded in recommended practices for availability and eases co-authorship processes~\cite{Raymond,Lessig}.
-	\item To provide a didactic presentation of the content to favor its apprehension and usage.
-It is worthwhile to mention that this subject comprises diverse topics on signal processing, music, psycho-acoustics and programming.
+	\item Present a concise set of mathematical relations between musical basic elements and sequences of PCM audio samples.
+	\item Introduce a framework of sound and musical synthesis with control at sample level which entails potential uses in psychoacoustic experiments and synthesis with extreme precision (discussed in Section~\ref{highPrecision}).
+	\item Provide a powerful theoretical framework which can be used to synthesize musical pieces and albums.
+	\item Provide approachability to the developed framework~\footnote{All
+		the analytic relations presented in this article are implemented as public domain scripts,
+		i.e. small computer programs using accessible technologies for distribution and validation.
+		This constitutes the \massa\ toolbox, available in an open Git repository~\cite{gitBook}.
+		These scripts are written in Python and make use of Numpy,
+		which performs numerical routines efficiently (e.g. through LAPACK).
+		Part of the scripts have been transcribed to JavaScript and native Python,
+		which favors their use in Web browsers such as Firefox and
+		Chromium~\cite{numpy, audiolab, tutpython, python}.
+		These are all open technologies, published using licenses that allow for copy,
+		distribution and use for research and generating derivatives.
+		Hence, the work presented here aims at being compliant with recommended
+		practices for availability and validation
+		and should ease co-authorship processes~\cite{Raymond,Lessig}.}.
+	\item Provide a didactic presentation of the content, which is highly multidisciplinary,
+		involving signal processing, music, psycho-acoustics and programming.
 \end{enumerate}
 
-The remaining parts of this work are organized as follows: section~\ref{sec:discNote} characterizes the basic musical note; section~\ref{sec:internalVar} further develops internals of the musical note; section~\ref{sec:notesMusic} tackles the organization of musical notes in higher levels of musical structure~\cite{Wisnick,Webern,Lerdahl,Cook,Lacerda}.
-As these descriptions embody topics such as psycho-acoustics, cultural traditions, formalisms and protocols, the text points to external complements as needed~\cite{Zamacois,Schoenberg,microsound}.
-
-The next section is a minimum text in which musical elements are presented side-by-side with the discrete-time samples they result.
-In order to account for validation and sharing, implementations on computer code of each one of these relations are gathered in the \massa\ toolbox together with little musical montage resulting from them.
-
-%Representing musical structures and artifices by it's discrete sound characteristics is the purpose of this work.
-The results include mathematical relations and it's computer program implementations.
-Next section exposes the theoretical description, which is implemented as scripts in a one-to-one relation to the equations.
-
-%\subsection{Sound and digital audio}
-
-%Sound is a longitudinal wave of mechanical pressure.
-The frequency bandwidth between $20Hz$ and $20kHz$ is appreciated by human hearing system with boundaries dependent on the person, climate conditions and sonic characteristics itself~\cite{Roederer}.
-If considered the speed of sound of $\approx 343.2 m/s$, this limits corresponds to $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.
-
-%Human perception of sound involves captivation by bones, stomach, ears, transfer functions of head and processing torso and nervous system.
-Besides that, the ear is a dedicated organ to the capture of this waves.
-Its mechanism decomposes sound into its sinusoidal spectrum and delivers them to the nervous system.
-This sinusoidal components are crucial to musical phenomena, as one can observe in the composition of sounds with musical interest and in tunings and scales.
-Subsection~\ref{subsec:dicNote} exposes the presence of sinusoid in discrete-time audio and characterizes a basic musical note.
-
-%The representation of sound is called audio (although these terms are often used without distinction), and this can be provenient from caption by microphones or from synthesis.
-Often enough, digital audio is specified by protocols that eases file storage and transferring, in cost of a direct representation or even some loss in quality.
-Standard representation of digital audio, on the other hand, consists of samples equally spaced by $\lambda_s$ durations in time, with each sample specified by a sample number of bits.
-This is called the Pulse Code Modulation representation of sound (PCM).
-A PCM digital sound is characterized by it's sampling frequency $f_s=\frac{1}{\lambda_s}$, also called sampling rate, and bit depth, which is the number of bits used of representing the amplitude of each sample.
-Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with $4$ bits each.
-The $2^4=16$ possible steps for each sample, together with the regular spacing $\lambda_s$ between them, introduces a quantization error.
-This noise, caused by this errors, diminishes as these spacing diminishes.
-
-%\begin{figure*}
-%    \centering
-%        \includegraphics[width=\textwidth]{pcm}
-%        \caption{Pulse Code Modulation (PCM) audio: an analogical signal is %represented by 25 samples with 4 bits each.}
-%        \label{fig:PCM}
-%\end{figure*}
-
-%By the Nyquist theorem, it is known that half the sampling frequency is the maximum frequency of the signal.
-Thus, it is necessary to have a sampling frequency at least twice the highest frequency heard by humans $f_s \geq 2\times 20kHz = 40kHz$.
-This is the basis for the use of the sampling frequencies $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and Broadcast systems (Radio and TV), respectively.
-
-%\subsection{Sonic art and musical theory}
-
-%A common definition for music is the art made by sounds and silences.
-For the average listener -- and a reasonable part of specialists -- the notion of music presupposes rhythmic and pitch organization such as explained in subsection~\ref{subsec:notesMusic}.
-Music from the twentieth century enlarged this traditional comprehension of music.
-This occurred in concert music, specially in the concrete, electronic and electroacoustic styles.
-On the last decade of the century, it was evident that popular music has also incorporated sounds without defined pitch and temporal organization out of simple metrics.
-Even though, the note stands paradigmatic as a 'fundamental unit' of musical structures and, in practice, it can unfold in sounds that observe this recent developments.
-The definition and expansion of the musical note as the fundamental unit of music is approached in subsections~\ref{subsec:discNote} and~\ref{subsec:internalVar}, respectively.
-Subsection~\ref{subsec:notesMusic} tackles the organization of musical notes in a higher level~\cite{Wisnick,Webern,Lerdhal,Cook,Lacerda}.
-
-%Musical theory embody topics as diverse as psycho-acoustics, cultural manifestations and formalisms.
-The section~\ref{sec:results} point this topics as needed and designate external complements~\cite{Zamacois,Schoenberg,microsound}.
-
-%\subsection{Computational implementation}
-
-%The results presented in this article are implemented as scripts, i.e.\ small computer programs implemented using accessible technologies for better distribution and validation.
-This constitute the \massa\ toolbox, available in public domain in an open Git repository~\cite{gitBook}.
-This scripts are written in Python and make use of external libraries Numpy and Scikits/Audiolab that performs calls to Fortran routines for better computational efficiency.
-Part of this code has been transcribed to JavaScript and native Python with readiness, what points to uses of this contribution in Web browsers such as Firefox and Chromium~\cite{numpy, audiolab, tutpython, python}.
-
-%This are all open technologies, that is, published using licenses that allows copy, distribution and use of any part for research and derivatives.
-This way, the work here presented is available and eases co-authorship processes~\cite{Raymond,Lessig}.
-
-%\subsection{Objectives}
-%\label{subsec:objectives}
-%The main goal of this article is to present a concise set of relations among musical basic elements and sequences of PCM audio samples.
-The next section is a minimum text in which music elements are presented side-by-side with the discrete-time samples they result.
-As validation and sharing, implementations on computer code of these relations and little musical pieces where gathered the \massa\ toolbox, available online.
-
-%Secondary objectives include presenting a framework of sound synthesis with control at sample level, with potential uses in psychoacoustical experiments and high-fidelity synthesis.
-The didactic presentation of the content favors use and apprehension on a problem which calls diverse topics to be tackled: signal processing, music and psycho-acoustics, to name just a few.
-
-%\subsection{Related work}
-%Due to the general interest, and number of knowledge areas involved, there is a number of books and computer implementations that are of interest or present similarities to what is presented in this work.
-A more detailed comparison of them is pointed out in the bibliography~\cite{dissertacao}.
-There is almost no articles which could be found on the topic.
-In summary, there are computer implementations that use this analytical descriptions implicitly, but there is no such a concise and mathematical description of the processes implemented, as they aim to be libraries for sound and music.
-There are books on the topic that cover various aspects of effects and physical modeling, but none of them carry a concise description of musical elements and structures, but focus on aspects of musical sounds and ways to mimic traditional instruments.
+The reminder of this article is organized as follows:
+Section~\ref{sec:discNote} characterizes the basic musical note;
+Section~\ref{sec:internalVar} develops internal dynamics of musical notes;
+Section~\ref{sec:notesMusic} tackles the organization of musical notes in
+higher level musical structures~\cite{Wisnick,Webern,Lerdahl,Cook,Lacerda}.
+As these descriptions require topics such as psychoacoustics, cultural traditions,
+and formalisms, the text points to external complements as needed~\cite{Zamacois,Schoenberg,microsound}
+and presents methods, results and discussions altogether.
+Section~\ref{cap:conclusao} holds final considerations and further work.
 

### 2017-07-04 article/acm/body.tex
revisado texto até localizacao espacial

@@
 \section{Characterization of the discrete-time musical note} \label{sec:discNote}\label{sec:notaDisc}
-
 In diverse artistic and theoretical contexts, music is conceived as comprising fundamental units referred to as notes, ``atoms'' that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
-These notes are now understood as central elements of certain musical paradigms.
+These notes are understood as central elements of certain musical paradigms.
 In a cognitive perspective, the notes are seen as discretized elements that facilitate and enrich the flow of information through music~\cite{Roederer, Lacerda}.
-Canonically, the principal properties of a musical note are duration, volume, pitch and timbre~\cite{Lacerda}, which can be quantified using the evenly time spaced sonic samples.
+Canonically, the basic characteristics of a musical note are duration, loudness, pitch and timbre~\cite{Lacerda}, which can be quantified using the evenly time spaced sonic samples.
 
@@
 \subsection{Duration}
-
-The sample frequency $f_s$ is defined as the number of samples in each second of the discrete-time signal.
-Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
-A musical note of duration $\Delta$ seconds is presented as a sequence $T_i^{\Delta}$ of $\lfloor \Delta .
-f_s \rfloor $ samples.
+The sample frequency $f_s$ is defined as the number of samples in each second of the discrete-time signal. Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
+A musical note of duration $\Delta$ seconds can be expressed as a sequence $T_i^{\Delta}$ of $\lfloor \Delta . f_s \rfloor$ samples.
 That is, the integer part of the multiplication is considered, and an error of at most $\delta_s$ missing seconds is admitted, which is usually fine for musical purposes.
@@
 
-With $\Lambda=\lfloor \Delta . f_a \rfloor$, the number of samples in the sequence, the more condensed notation is $T_i^\Delta=\{t_i\}_0^{\Lambda-1}$.
-
-\subsection{Volume}\label{subsec:volume}
-
-The sensation of sound volume depends on reverberation and harmonic distribution, among other characteristics described in section~\ref{sec:varInternas}.
-One can get volume variations through the power of the wave~\cite{Chowning}:
+With $\Lambda=\lfloor \Delta . f_s \rfloor$, the number of samples in the sequence, a more condensed notation is $T_i^\Delta=\{t_i\}_0^{\Lambda-1}$.
+
+\subsection{Loudness}\label{subsec:volume}
+The sensation of loudness depends on reverberation and spectrum, among other characteristics described in section~\ref{sec:varInternas}.
+One can achieve loudness variations through the power of the wave~\cite{Chowning}:
 
@@
 
-The final volume is dependent on the speakers amplification of the signal.
-Thus, what matters is the relative power of a note in relation to the other ones around it, or the power of a music section in relation to the rest.
-Differences in volume are measured in decibels, calculated directly from the amplitudes through energy or potency:
+The final loudness is dependent on the amplification of the signal by the speakers.
+Thus, what matters is the relative power of a note in relation to the others around it, or the power of a music section in relation to the rest.
+Differences in loudness are the result of complex psychophysical phenomena but can often be reasoned about in terms of decibels, calculated directly from the amplitudes through energy or potency:
 
@@
 The quantity $V_{dB}$ has the decibel unit ($dB$). 
-For each $10dB$ it is associated a ``doubled volume''.
-A handy reference is $10dB$ for each step in the intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} e \emph{fortissimo}.
+For each $10dB$ a ``doubled volume'' is associated.
+A handy reference is $10dB$ for each step in the musical intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} and \emph{fortissimo}.
 Other useful references are $dB$ values related to double amplitude or potency:
@@
 
-An amplitude increase by a factor slightly above of 3 is required for yielding a doubled volume.
+Thus, an amplitude increase by a factor slightly above 3 is required for achieving a doubled loudness.
 These values are guides for increasing or decreasing the absolute values in the sample sequences with musical purposes.
@@
 \subsection{Pitch}
-
-To a note corresponds a sequence $T_i$ in which duration and volume are directly related to the size of the sequence and the amplitude of its samples, respectively.
+To a note corresponds a sequence $T_i$ in which duration and loudness are directly related to the size of the sequence and the amplitude of its samples, respectively.
 The pitch is specified by the fundamental frequency $f_0$ whose cycle has duration $\delta_{f_0}=1/f_0$.
-This duration multiplied by the sampling frequency $f_s$ results in the number of samples per cycle: $\lambda_{f_0}=f_s .
+This duration, multiplied by the sampling frequency $f_s$, yields in the number of samples per cycle: $\lambda_{f_0}=f_s .
 \delta_{f_0} =f_a/f_0$.
 
-For didactic reasons, let $f_0$ be such that it divides $f_s$ and $\lambda_{f_0}$ results integer. If $T_i^f$ is a sonic sequence of fundamental frequency $f$, then:
+For didactic reasons, let $f_0$ be such that it divides $f_s$ and $\lambda_{f_0}$ results integer. If $T_i^f$ is a sonic sequence with fundamental frequency $f$, then:
 
@@
 \subsection{Timbre}
-
-In a sound with harmonic spectrum, the (wave) period corresponds to the cycle duration, given by the inverse of the fundamental frequency.
-The trajectory of the wave inside the period - called the waveform - defines a harmonic spectrum and, thus, a timbre\footnote{Timbre is a subjective and complex characteristic.
+In a sound with harmonic spectrum, there is a (wave) period which corresponds to the (wave) cycle duration and is given by the inverse of the fundamental frequency.
+The trajectory of the wave inside the period - called the waveform - defines a harmonic spectrum and, thus, entails a timbre\footnote{The timbre of a sound is a subjective and complex characteristic.
 Physically, the timbre is multidimensional and given by the temporal dynamics of energy in the spectral components that are harmonic or noisy.
@@
 
-The simplest case is the spectrum with only the fundamental frequency $f$, which is a sinusoid as in "simple harmonic oscillation".
-Let $S_i^f$ be a sequence whose samples $s_i^f$ describe a sinusoid of frequency $f$:
+The simplest case is the spectrum with only one (fundamental) frequency $f$, which is a sinusoid, often regarded as a ``pure'' oscillation in terms of the \emph{simple harmonic motion}.
+Let $S_i^f$ be a sequence whose samples $s_i^f$ describe a sinusoid with frequency $f$:
 
@@
 
-where $\lambda_f=\frac{f_s}{f}=\frac{\delta_f}{\lambda_s}$ is the number of samples in the period.
-
-In a similar fashion, other waveforms are applied in music for their spectral qualities and simplicity.
-While the sinusoid is an isolated point in the spectrum, these waves present a succession of harmonic components.
-These waveforms are specified in equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square}, illustrated in Figure~\ref{fig:formasDeOnda}.
+\noindent where $\lambda_f=\frac{f_s}{f}=\frac{\delta_f}{\lambda_s}$ is the number of samples in the period.
+
+In a similar fashion, other waveforms are used in music for their spectral qualities and simplicity.
+While the sinusoid is an isolated node in the spectrum, any other wave present a succession of harmonic components.
+The most basic waveforms are specified in equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square}, and are illustrated in Figure~\ref{fig:formasDeOnda}.
 These artificial waveforms are traditionally used in music for synthesis and oscillatory control of variables. They are also useful outside musical contexts\cite{Openheim}.
 
-The sawtooth presents all components of the harmonic series with decreasing energy of $-6dB/octave$. The sequence of temporal samples can be described as:
+The sawtooth presents all components of the harmonic series with decreasing energy of $-6dB/octave$\footnote{In 
+musical jargon, an ``octave'' $\{f,2f\}$ means a frequency $f$ and twice such frequency $2f$.}. The sequence of temporal samples can be described as:
 
@@
 The square wave preserves only odd harmonics falling at $-6dB/octave$:
-
 \begin{equation}\label{square}
@@
 In general, these waveforms are appreciated as excessively rich in sharp harmonics, and attenuator filtering on treble and middle parts of the spectrum is especially useful for reaching a more natural and pleasant sound.
-The relatively attenuated harmonics of the triangle wave makes it the more functional - among the listed ones - to be used in the synthesis of musical notes without any treatment.
+The relatively attenuated harmonics of the triangle wave makes it the more functional - among the listed ones - to be used in the synthesis of musical notes without any further processing.
 The sinusoid is often a nice choice, but a problematic one.
-While pleasant if not loud in a very high pitch (above 500Hz, it requires careful dosage), the pitch is not accurately detected, particularly at low frequencies.
-It requires a large amplitude to give sinusoid volume, if compared to other waveforms.
-Both particularities are seen as a consequence of the non-existence of pure sinusoidal sounds in nature~\cite{Roederer}.
+While pleasant if not loud in a very high pitch (above $\approx 500Hz$ it requires careful dosage), the pitch of a pure sinusoid is not accurately detected by the human auditory system, particularly at low frequencies.
+Also, it requires a large amplitude to give a sinusoid a level of loudness if compared to other waveforms.
+Both particularities are understood in the scientific literature as a consequence of the nonexistence of pure sinusoidal sounds in nature~\cite{Roederer}.
 
@@
         \includegraphics[width=.7\textwidth]{figures/waveForms_}
-    \caption{Basic musical waveforms: (a) the synthetic  waveforms; (b) the real waveforms.}
+    \caption{Basic musical waveforms: (a) the basic synthetic  waveforms; (b) real waveforms.}
         \label{fig:formasDeOnda}
@@
 
-Figure~\ref{fig:formasDeOnda} presents the waveforms described in equations  ~\ref{sinusoid}, ~\ref{sawTooth}, ~\ref{triangular} and ~\ref{square} for $\lambda_f=100$ (period of $100$ samples).
-If $f_s=44.1kHz$, the PCM standard in Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz$, around A4, just above the central "C", whatever the waveform is.
+Figure~\ref{fig:formasDeOnda} presents the waveforms described in equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} for $\lambda_f=100$ (i.e.
+period of $100$ samples).
+If $f_s=44.1kHz$, the PCM standard for Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Hz$, around A4, just above the central "C", whatever the waveform is.
 
@@
 The isolated and exactly harmonic components of the spectrum is a consequence of the use of a fixed period.
-The sinusoid consists of a unique node in the spectrum, pure frequency.
-The figure exhibits the spectra described: the sawtooth is the only waveform with a complete harmonic series (odd and even components); triangular and square waves have the same components (odd harmonics), decaying at $-12dB/octave$ and $-6dB/octave$, respectively.
+The sinusoid consists of a unique node in the spectrum.
+The figure exhibits the spectra described in the last paragraphs: the sawtooth is the only waveform with a complete harmonic series (odd and even components); triangular and square waves have the same components (odd harmonics), decaying at $-12dB/octave$ and $-6dB/octave$, respectively.
 
@@
 The harmonic spectrum is composed of frequencies $f_n$ that are multiples of the fundamental frequency: $f_n=(n+1)f_0$.
-As the human linear perception of pitch follows  a geometric progression of frequencies, the spectrum has notes different from the fundamental frequency (see equation~\ref{eq:serieHarmonica}).
+As the human linear perception of pitch follows a geometric progression of frequencies, the harmonic spectrum has notes different from the fundamental frequency (see equation~\ref{eq:serieHarmonica}).
 Additionally, the number of harmonics will be limited by the Nyquist frequency $f_s/2$.
-From a musical perspective, it is critical to internalize that energy in a component of frequency $f_n$ means an oscillation in the constitution of the sound, purely harmonic and in that frequency $f_n$.
-This energy, specifically concentrated on the frequency $f_n$, is separated by the ear for further cognitive processes (this separation is done in several mechanisms similar to the human cochlea~\cite{Roederer}).
-
-The sinusoidal components are usually responsible for timbre qualities.
-If they are not presented in harmonic proportions (small number relations), the sound is perceived as noisy or dissonant, in opposition to sonorities with an unequivocally established fundamental.
+From a musical perspective, it is critical to internalize that energy in a component with frequency $f_n$ is a sinusoidal oscillation in the constitution of the sound in that frequency $f_n$.
+This energy, specifically concentrated on $f_n$, is separated from other frequencies by the ear for further cognitive processes (this separation is done by diverse living organisms by mechanisms similar to the human cochlea~\cite{Roederer}).
+
+The sinusoidal components are responsible for timbre qualities.
+If their frequencies do not relate by small integers, the sound is perceived as noisy or dissonant, in opposition to sonorities with an unequivocally established fundamental.
 Furthermore, the notion of absolute pitch relies on the similarity of the spectrum to the harmonic series~\cite{Roederer}.
-For a fixed length period and waveform, the spectrum is perfectly harmonic and static, and each waveform is comprised of specific proportions of harmonic components.
-High  curvatures are a sign of high harmonics in the wave.
+For a fixed length period and waveform, the spectrum is perfectly harmonic and static, each waveform entails different energy in the (same) harmonic components.
+High curvatures are a sign of energy in the high harmonics.
 Figure~\ref{fig:formasDeOnda} depicts a wave, labeled as ``sampled real sound'', with a period of $\Lambda_f=114$ samples, extracted from a relatively well behaved recorded sound.
-The oboe wave was also sampled at $44.1kHz$.
-The chosen period for sampling was relatively short, with $98$ samples, and corresponds to the frequency $\frac{44100}{98}=450Hz$, which is associated with a slightly out-of-tune A4 pitch.
+The oboe waveform was also sampled at $44.1kHz$.
+The period is relatively short, with $98$ samples, and corresponds to the frequency $\frac{44100}{98}=450Hz$, which is associated with a slightly out-of-tune A4 pitch.
 One can notice from the curvatures: the oboe's rich spectrum at high frequencies and the lower spectrum of the real sound.
 
-The sequence $ R_i=\{ r_i \}_0^{\lambda_f-1}$ of samples in the real sound of Figure~\ref{fig:formasDeOnda} can be taken as basis for a sound $T_i^f$ in the following way:
+The sequence $R_i=\{ r_i \}_0^{\lambda_f-1}$ of samples in the real sound of Figure~\ref{fig:formasDeOnda} can be taken as basis for a sound $T_i^f$ in the following way:
 
@@
 The resulting sound has the spectrum of the original waveform.
-As a consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise or variations typical of the natural phenomenon.
-This can be observed in Figure~\ref{fig:espectroOboe}, that shows the spectrum of the original oboe note and a note with same duration, whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}.
-Summing up, the natural spectrum exhibits variations in the frequencies of the harmonics, in their intensities and some noise, while the note made from the sampled period has a perfectly harmonic spectrum.
+As a consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise or variations typical of natural phenomena.
+This can be observed in Figure~\ref{fig:espectroOboe}, which shows the spectrum of the original oboe note and a note with the same duration, whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}.
+Summing up, the natural spectrum exhibits variations in the frequencies of the harmonics, in their intensities and some noise, while the note made from the sampled period has a perfectly harmonic and static spectrum.
 
@@
         \includegraphics[width=.7\textwidth]{figures/oboeNaturalSampledSpectrum}
-    \caption{Spectrum of the sonic waves of a natural oboe note and from a sampled period.
-The natural sound has fluctuations in the harmonics and in its noise, while the sampled period note has a perfectly harmonic spectrum.}
+	\caption{Spectrum of the sonic waves of a natural oboe note and obtained through a sampled period.
+The natural sound has fluctuations in the harmonics and in its noise, while the sampled period note has a perfectly harmonic (and static) spectrum.}
         \label{fig:espectroOboe}
@@
 \subsection{Spectrum at sampled sound}
-
 These sinusoidal components in the discretized sound have some particularities.
-Considering a signal $T_i$ and its corresponding Fourier decomposition $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$, the recomposition is the sum of the frequency components as time samples\footnote{The factor $\frac{1}{\Lambda}$ could be distributed among the Fourier transform and its reconstruction, as preferred.}:
+Considering a signal $T_i$ and its corresponding Fourier decomposition $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$, the recomposition is the sum of the frequency components as time samples\footnote{The factor $\frac{1}{\Lambda}$ can be distributed among the Fourier transform and its reconstruction, as preferred.}:
 
@@
 
-where $c_k = a_k + j .
+\noindent where $c_k = a_k + j .
 b_k$ defines the amplitude and phase of each frequency: $w_k=\frac{2\pi}{\Lambda}k$ in radians or $f_k=w_k\frac{f_a}{2\pi}=\frac{f_a}{\Lambda}k$ in Hertz, taking into account the respective limits in $\pi$ and in $\frac{f_a}{2}$ given by the Nyquist Theorem.
@@
      \caption{Oscillation of 2 samples (maximum frequency for any $f_a$).
-The first coefficient reflects a detachment (\emph{offset} or \emph{bias}) and the second coefficient specifies the oscillation amplitude.}
+The first coefficient reflects a constant detachment (called \emph{offset} or \emph{bias}) and the second coefficient specifies the oscillation amplitude.}
          \label{fig:amostras2}
@@
 with energies $e_k=\frac{(c_k)^2}{\Lambda=2}$.
-The role of amplitudes $a_k$ is clearly observed with $\frac{a_0}{2}$, the fixed offset\footnote{Also called \emph{bias}.} and $\frac{a_1}{2}$, oscillation amplitude with frequency given by $f_k=k \frac{f_a}{\Lambda=2}$.
+The role of amplitudes $a_k$ is clearly observed with $\frac{a_0}{2}$, the fixed offset (also called \emph{bias}), and $\frac{a_1}{2}$, oscillation with frequency given by $f_k=k \frac{f_a}{\Lambda=2}$.
 This case has special relevance.
-At least 2 samples are necessary to represent an oscillation and it yields the Nyquist frequency $f_{\text{max}}=\frac{f_a}{2}$, which is the maximum frequency in a sound sampled with $f_a$ samples per second\footnote{Any sampled signal has this property, not only the digitalized sound.}.
+At least 2 samples are necessary to represent an oscillation and it yields the Nyquist frequency $f_{\text{max}}=\frac{f_a}{2}$, which is the maximum frequency in a sound sampled with $f_a$ samples per second.
+In fact, any sampled signal has this property, not only the digitalized sound.
 
 All fixed sequences $T_i$ of only $3$ samples also have just $1$ frequency, since the first harmonic would have $1.5$ samples and exceeds the bottom limit of 2 samples, i.e.\ the frequency of the harmonic would exceed the Nyquist frequency:  $\; \frac{2.
-f_a}{3} > \frac{f_a}{2} $.
+f_a}{3} > \frac{f_a}{2}$.
 The coefficients $\{c_k\}_0^{\Lambda-1=2}$ are present in 3 frequency components.
-One is relative to zero frequency ($c_0$), and the other two ($c_1$ and $c_2$) have the same role for reconstructing a sinusoid with $f=f_a/3$.
+One is relative to frequency zero ($c_0$), and the other two ($c_1$ and $c_2$) have the same role for reconstructing a sinusoid with $f=f_a/3$.
 
@@
 Given the cosine (an even function) and the inverse tangent (an odd function), the components in $w_k$ and $w_{\Lambda-k}$ contribute with coefficients $c_k$ and $c_{\Lambda-k}$ in the reconstruction of the real samples.
-In other words, in a decomposition of $\Lambda$ samples, the $\Lambda$ frequency components $\{c_i\}_0^{\Lambda-1}$ are equivalents in pairs,
+In other words, in a decomposition of $\Lambda$ samples, the $\Lambda$ frequency components $\{c_i\}_0^{\Lambda-1}$ are equivalent in pairs,
 except for $f_0$, and, when $\Lambda$ is even, for $f_{\Lambda/2}=f_{\text{max}}=\frac{f_a}{2}$.
@@
 
-This discussion makes it clear the equivalence ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} and ~\ref{equivalenciasFases}:
+This discussion makes it clear the equivalence in~\ref{equivalenciasFreqs},~\ref{equivalenciasModulos} and~\ref{equivalenciasFases}:
 
@@
 
-with $k \in \mathbb{N}$.
+\noindent with $k \in \mathbb{N}$.
 
@@
 
-with $a_{\Lambda/2}=0$ if $\Lambda$ odd.
+\noindent with $a_{\Lambda/2}=0$ if $\Lambda$ odd.
 
@@
 Figure~\ref{fig:amostras4} depicts the basic waveforms with 4 samples and their two (possible) components.
-The individual contributions sum to the original waveform and a brief inspection reveals the major curvatures resulting from the higher frequency, while the fixed offset is captured in the zero frequency component.
+The individual components sum to the original waveform and a brief inspection reveals the major curvatures resulting from the higher frequency, while the fixed offset is captured in the zero frequency component.
 
@@
 \subsection{The basic note}\label{notaBasica}
-
-Let $f$ be such that it divides $f_a$\footnote{As pointed before, this limitation simplifies the explanation without losing generality, and will be overcome in the next section.}.
-A sequence $T_i$ of sonic samples separated by $\delta_a=1/f_a$ describes a musical note with a frequency of $f$ Hertz and $\Delta$ seconds of duration if, and only if, it has the periodicity $\lambda_f=f_a/f$ and size $\Lambda=\lfloor f_a .
-\Delta \rfloor $:
+In a nutshell, let $f$ be such that it divides $f_a$\footnote{As mentioned before, this limitation simplifies the explanation for now and will be overcome in the next section.}.
+A sequence $T_i$ of sonic samples separated by $\delta_a=1/f_a$ expresses a musical note with a frequency of $f$ Hertz and $\Delta$ seconds of duration if, and only if, it has the periodicity $\lambda_f=f_a/f$ and size $\Lambda=\lfloor f_a .
+\Delta \rfloor$:
 
@@
 A unique period from the basic waveforms can be used to specify the note, where $\lambda_f=\frac{f_a}{f}$ is the number of samples at the period.
-Here, $L_i^{f,\, \delta_f} $ is the sequence that describes a period of the waveform $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ with duration $\delta_f=1/f$ (as given by equations~\ref{sinusoid}, ~\ref{sawTooth}, ~\ref{triangular} and ~\ref{square}) and $R_i^f$ is a sampled real waveform:
+Here, $L_i^{f,\, \delta_f} $ is the sequence that describes a period of the waveform $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ with duration $\delta_f=1/f$ (as given by equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square}) and $R_i^f$ is a sampled real waveform:
 
@@
 \subsection{Spatial localization and spatialization}\label{subsec:spac}
-
-A musical note always has a spatial localization, even though this is not one of its four basic properties: the note source position at the ordinary physical space.
+A musical note always has a spatial localization, even though this is not one of its four basic properties in canonical musical theory.
+The note has a source which has a position at the ordinary physical three dimensional space.
 The reverberation in the environment in which sound occurs is the matter of \emph{spatialization}.
-Both fields, spatialization and spatial localization, are widely valued by audiophiles and the music industry~\cite{floEsp}.
+Both concepts, spatialization and spatial localization, are widely valued by audiophiles and the music industry~\cite{floEsp}.
 
 \subsubsection{Spatial localization}
-
-It is understood that the perception of sound localization occurs in our nervous system by three pieces of information: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, including its chest, head and ears~\cite{Roederer, hrtf, Heeger}.
+It is understood that the perception of sound localization occurs in our nervous system by three pieces of information: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, specially its chest, head and ears~\cite{Roederer, hrtf, Heeger}.
 
@@
 
-\section{Variation in the basic musical note}\label{sec:internalVar}\label{sec:varInternas}
+\section{Variations in the basic musical note}\label{sec:internalVar}\label{sec:varInternas}
 

### 2017-07-07 article/acm/foobody.tex
added file ffor comparison

--- 2017-04-24 article/article.tex
@@
-% notas:
-% campo harmônico => harmonic field?
-% anti-relative => counter-relative? (discussão sobre harmonia funcional)
-%    associada relative => associated relative? (ver a discussão para mais...)
-% medianas cromáticas => cromatic mediant
-% tonalidade => key
-% homônica => homonymous
-
-\documentclass[
- aip,
- jmp,
- amsmath,amssymb,
- reprint,
-]{revtex4-1}
-\usepackage{graphicx}
-\usepackage{grffile}
-\usepackage{dcolumn}
-\usepackage{bm}
-\usepackage{multirow}
-\usepackage{color}
-\usepackage{xcolor}
-\usepackage{enumerate}
-\usepackage{float}
-\usepackage{hyperref}
-
-\newcommand{\massa}{{\large \textsc{massa}}}
-\newcommand{\mass}{{\large \textsc{mass}}}
-\newcommand{\figgus}{{\large \textsc{figgus}}}
-
-\floatstyle{plain}
-\newfloat{twocolequfloat}{t}{zzz}
-\floatname{twocolequfloat}{Equation}
-
-\begin{document}
-\preprint{AIP/123-QED}
-
-\title{Psychophysics of musical elements in the discrete-time representation of sound}
-
-\author{Renato Fabbri}
- \email{renato.fabbri@gmail.com}
-  \affiliation{ 
-Instituto de F\'isica de S\~ao Carlos, Universidade de S\~ao Paulo (IFSC/USP)
-}
-
-\author{Vilson Vieira da Silva Junior}
-  \homepage{http://automata.cc/}
-  \email{vilson@void.cc}
- \altaffiliation[Also at ]{IFSC-USP}
-
-\author{Ant\^onio Carlos Silvano Pessotti}
-  \email{antoniopessotti@gmail.com}
- \altaffiliation[Instituto de Estudos da Linguagem ]{IEL/UNICAMP}
-
-\author{D\'ebora Cristina Corr\^ea}
-  \email{debcris.cor@gmail.com}
- \altaffiliation[Instituto de Ci\^encias Matem\'aticas e de Computa\c{c}\~ao ]{ICMC/USP}
-
-%\author{Luciano da Fontoura Costa}
-%  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
-%  \email{ldfcosta@gmail.com}
-% \altaffiliation[Also at ]{IFSC-USP}
-%
-\author{Osvaldo N. Oliveira Jr.}
-\homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
-\email{chu@ifsc.usp.br}
-\altaffiliation[Also at ]{IFSC-USP}
-
-\date{\today}
-\begin{abstract}
-
-    %The representation of the basic elements of music - such as notes, ornaments and interval structures - in terms of discrete audio signal is often used in software for music creation and design.
-Nevertheless, there is no unified approach that relates these elements to the sound discrete samples.
-In this article, each musical element is described by equations in sonic time samples, which are all implemented in scripts within an open source software toolbox, referred to as \massa\ (Music and Audio in Sequences and Samples).
-The fundamental element, the musical note with duration, volume, pitch and timbre, is related quantitatively to the characteristics of the discrete-time signal.
-Internal variations, such as tremolos, vibratos and spectral fluctuations, are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities.
-With this representation of notes, resources are provided for the generation of musical structures, such as rhythmic meter, pitch intervals and cycles.
-The efficacy of these physical descriptions of basic musical elements was confirmed by the synthesis of small musical pieces within each frame: basic notes, incremented notes and notes in music.
-It is possible to synthesize whole albums through collage of the scripts and parametrization.
-    %The sample-based analytical description, and the paradigm of open source implementation, enables scientific experiments in precise and trustful ways.
-In fact, \massa\ has already been employed by external users for diverse purposes.
-Among these, it is mentioned an acoustic effect recognized by diverse individuals in mailing lists but not found in literature, and uses related to art and education.
-
-Notes, ornaments and intervals are examples of basic musical elements.
-Their representation as discrete-time digital audio plays a central role in software for music creation and design.
-Nevertheless, there is no systematic relation, in analytical terms, of these musical elements to the sonic samples.
-Such a compendium enables scientific experiments in precise and trustful ways, among educational and artistic uses.
-This paper presents a comprehensive description of music in digital audio, within an unified approach.
-Musical elements, like pitch, duration and timbre are expressed by equations on sample level.
-This quantitatively relates characteristics of the discrete-time signal to musical qualities.
-Internal variations, e.g. tremolos, vibratos and spectral fluctuations, are also considered
-to operate within a note. Moreover, the generation of musical structures such as rhythmic meter,
-pitch intervals and cycles, are attained canonically with notes. 
-The availability of these resources in scripts is provided in public domain within the \massa\ toolbox - Music and Audio in Sequences and Samples.
-Authors observe that the implementation of sample-domain analytical results as open source can encourage concise research.
-As further illustrated in the paper, \massa\ has already been employed by users for diverse purposes, 
-including acoustics experimentation, art and education. The efficacy of these physical descriptions 
-was confirmed by the synthesis of small musical pieces.
-As shown, it is possible to synthesize whole albums through collage of scripts and parametrization.
-
-\end{abstract}
-\pacs{*43.66.-x,43.66.+y,05.10.-a} % PACS, the Physics and Astronomy
-\keywords{psychophysics, acoustics, statistics, signal processing, digital audio, music}
-\maketitle
-
-\section{\label{sec:level1}Introduction}
-
-Music is commonly defined as the art made by sounds and silences, where sound corresponds to the longitudinal wave of mechanical pressure.
-The human hearing system perceives sounds within the frequency bandwidth between $20Hz$ and $20kHz$, with the actual limits depending on the person, climate conditions and the sonic characteristics themselves~\cite{Roederer}.
-Since the speed of sound is $\approx 343.2 m/s$, these limits imply wavelengths of $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.
-Such perception involves stimuli in bones, stomach, ears, transfer functions of head and torso, and processing by the nervous system~\cite{Roederer}.
-Obviously, the ear is a dedicated organ for the appreciation of these waves, which decomposes them into their sinusoidal spectra and delivers to the nervous system.
-The sinusoidal components are crucial to musical phenomena, as one can perceive in the composition of sounds of musical interest (such as harmonic sounds and noises, discussed in sections~\ref{sec:discNote} and~\ref{sec:internalVar}), and sonic structures of musical interest (such as tunings and scales, in section~\ref{sec:notesMusic}).
-
-The representation of sound is commonly referred to as audio, although these terms are often used without distinction.
-Audio expresses waves from the capture by microphones or from direct synthesis, although these sources are not neatly distinguishable as captured sounds are processed to generate new sonorities.
-Digital audio specified by protocols that facilitate file transferring and storage often implies a quality loss.
-Standard representation of digital audio, on the other hand, assures perfect reconstruction of the analog wave, within any convenient precision.
-This paradigm consists of representing the audio with equally spaced samples, of $\lambda_s$ durations, each specified by a fixed number of bits.
-This is the Pulse Code Modulation (PCM) representation of sound.
-A sound in PCM audio is characterized by a sampling frequency $f_s=\frac{1}{\lambda_s}$ (also called the sampling rate), which is the number of samples used for representing a second of sound; and by a bit depth, which is the number of bits used for representing the amplitude of each sample.
-Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with $4$ bits each.
-The fixed $2^4=16$ values for the amplitude of each sample, with the regular spacing $\lambda_s$, yields the quantization error or noise.
-This noise diminishes as the bit depth increases.
-
-\begin{figure*}
-    \centering
-        \includegraphics[width=\textwidth]{figures/pcm}
-        \caption{Pulse Code Modulation (PCM) audio: an analogical signal is represented by 25 samples with 4 bits each.}
-        \label{fig:PCM}
-\end{figure*}
-
-The Nyquist theorem~\cite{Openheim} states that the sampling frequency is twice the maximum frequency of the represented signal.
-Thus, for general musical purposes, it is necessary to have samples in a rate at least twice the highest frequency heard by humans, that is, $f_s \geq 2\times 20kHz = 40kHz$.
-This is the fundamental reason for the adoption of sampling frequencies such as $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and broadcast systems (radio and television), respectively.
-
-Within this framework, musical notes can be characterized.
-The note still stands paradigmatic as the 'fundamental unit' of musical structures and, in practice, it can unfold into sounds that uphold other approaches\footnote{Music from the twentieth century enlarged this traditional comprehension of music.
-This occurred in the concert music, especially in the concrete, electronic and electroacoustic styles.
-In the 1990s, it became evident that popular music had also incorporated sounds without defined pitch and temporal organization, to name a few of simple characteristics.}.
-    Notes are also convenient for another reason: the average listener -- and considerable part of the specialists -- presupposes rhythmic and pitch organization (made explicit in section~\ref{sec:notesMusic}) as fundamental musical properties, and these are developed in traditional musical theory in terms of notes.
-
-\subsection{Contributions and paper organization}
-
-This article aims at representing musical structures and artifices by their discrete-time sonic characteristics.
-Results include mathematical relations, usually in terms of samples, and their direct computer program implementations.
-Despite the general interests involved, there are few books and computer implementations that tackle the subject.
-These mainly focus on computer implementations and ways to mimic traditional instruments, with scattered mathematical formalisms.
-A compilation of the works and their contributions is in the bibliography~\cite{dissertacao}.
-To the best of the author's knowledge, there is a lack of articles on the topic.
-Moreover, although current computer implementations use the analytical descriptions presented in this study in a implicit manner, it seems that there has been no concise and mathematical description of the processes implemented.
-
-In order to address this concise description of musical elements and structures, in terms of the digitalized sound, the objectives of this paper are:
-
-\begin{enumerate}
-
-\item Present a concise set of relations among musical basic elements and sequences of PCM audio samples.
-
-\item Introduce a framework of sound synthesis with control at sample level and with potential uses in psycho-acoustical experiments and high-fidelity synthesis.
-
-\item Provide the accessibility of the developed framework.
-The analytic relations presented in this article are implemented as public domain scripts, i.e.
-small computer programs using accessible technologies for better distribution and validation.
-This constitute the \massa\ toolbox, available in an open Git repository~\cite{gitBook}.
-These scripts are written in Python and make use of external libraries, such as Numpy that performs calls to Fortran routines for better computational efficiency.
-Part of the scripts have been transcribed to JavaScript and native Python with readiness, what favors their use in Web browsers such as Firefox and Chromium~\cite{numpy, audiolab, tutpython, python}.
-Furthermore, these are all open technologies, that is, published using licenses that allow copy, distribution and use of any part for research and derivatives.
-This way, the work presented here is embedded in recommended practices for availability and eases co-authorship processes~\cite{Raymond,Lessig}.
-
-\item To provide a didactic presentation of the content to favor its apprehension and usage.
-It is worthwhile to mention that this subject comprises diverse topics on signal processing, music, psycho-acoustics and programming.
-
-\end{enumerate}
-
-The remaining parts of this work are organized as follows: section~\ref{sec:discNote} characterizes the basic musical note; section~\ref{sec:internalVar} further develops internals of the musical note; section~\ref{sec:notesMusic} tackles the organization of musical notes in higher levels of musical structure~\cite{Wisnick,Webern,Lerdahl,Cook,Lacerda}.
-As these descriptions embody topics such as psycho-acoustics, cultural traditions, formalisms and protocols, the text points to external complements as needed~\cite{Zamacois,Schoenberg,microsound}.
-
-The next section is a minimum text in which musical elements are presented side-by-side with the discrete-time samples they result.
-In order to account for validation and sharing, implementations on computer code of each one of these relations are gathered in the \massa\ toolbox together with little musical montage resulting from them.
-
-%Representing musical structures and artifices by it's discrete sound characteristics is the purpose of this work.
-The results include mathematical relations and it's computer program implementations.
-Next section exposes the theoretical description, which is implemented as scripts in a one-to-one relation to the equations.
-
-%\subsection{Sound and digital audio}
-
-%Sound is a longitudinal wave of mechanical pressure.
-The frequency bandwidth between $20Hz$ and $20kHz$ is appreciated by human hearing system with boundaries dependent on the person, climate conditions and sonic characteristics itself~\cite{Roederer}.
-If considered the speed of sound of $\approx 343.2 m/s$, this limits corresponds to $\frac{343.2}{20} = 17.16\,m$ and $\frac{343.2}{20000}=17.16\,mm$.
-
-%Human perception of sound involves captivation by bones, stomach, ears, transfer functions of head and processing torso and nervous system.
-Besides that, the ear is a dedicated organ to the capture of this waves.
-Its mechanism decomposes sound into its sinusoidal spectrum and delivers them to the nervous system.
-This sinusoidal components are crucial to musical phenomena, as one can observe in the composition of sounds with musical interest and in tunings and scales.
-Subsection~\ref{subsec:dicNote} exposes the presence of sinusoid in discrete-time audio and characterizes a basic musical note.
-
-%The representation of sound is called audio (although these terms are often used without distinction), and this can be provenient from caption by microphones or from synthesis.
-Often enough, digital audio is specified by protocols that eases file storage and transferring, in cost of a direct representation or even some loss in quality.
-Standard representation of digital audio, on the other hand, consists of samples equally spaced by $\lambda_s$ durations in time, with each sample specified by a sample number of bits.
-This is called the Pulse Code Modulation representation of sound (PCM).
-A PCM digital sound is characterized by it's sampling frequency $f_s=\frac{1}{\lambda_s}$, also called sampling rate, and bit depth, which is the number of bits used of representing the amplitude of each sample.
-Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with $4$ bits each.
-The $2^4=16$ possible steps for each sample, together with the regular spacing $\lambda_s$ between them, introduces a quantization error.
-This noise, caused by this errors, diminishes as these spacing diminishes.
-
-%\begin{figure*}
-%    \centering
-%        \includegraphics[width=\textwidth]{pcm}
-%        \caption{Pulse Code Modulation (PCM) audio: an analogical signal is %represented by 25 samples with 4 bits each.}
-%        \label{fig:PCM}
-%\end{figure*}
-
-%By the Nyquist theorem, it is known that half the sampling frequency is the maximum frequency of the signal.
-Thus, it is necessary to have a sampling frequency at least twice the highest frequency heard by humans $f_s \geq 2\times 20kHz = 40kHz$.
-This is the basis for the use of the sampling frequencies $44.1kHz$ and $48kHz$, standards in Compact Disks (CD) and Broadcast systems (Radio and TV), respectively.
-
-%\subsection{Sonic art and musical theory}
-
-%A common definition for music is the art made by sounds and silences.
-For the average listener -- and a reasonable part of specialists -- the notion of music presupposes rhythmic and pitch organization such as explained in subsection~\ref{subsec:notesMusic}.
-Music from the twentieth century enlarged this traditional comprehension of music.
-This occurred in concert music, specially in the concrete, electronic and electroacoustic styles.
-On the last decade of the century, it was evident that popular music has also incorporated sounds without defined pitch and temporal organization out of simple metrics.
-Even though, the note stands paradigmatic as a 'fundamental unit' of musical structures and, in practice, it can unfold in sounds that observe this recent developments.
-The definition and expansion of the musical note as the fundamental unit of music is approached in subsections~\ref{subsec:discNote} and~\ref{subsec:internalVar}, respectively.
-Subsection~\ref{subsec:notesMusic} tackles the organization of musical notes in a higher level~\cite{Wisnick,Webern,Lerdhal,Cook,Lacerda}.
-
-%Musical theory embody topics as diverse as psycho-acoustics, cultural manifestations and formalisms.
-The section~\ref{sec:results} point this topics as needed and designate external complements~\cite{Zamacois,Schoenberg,microsound}.
-
-%\subsection{Computational implementation}
-
-%The results presented in this article are implemented as scripts, i.e.\ small computer programs implemented using accessible technologies for better distribution and validation.
-This constitute the \massa\ toolbox, available in public domain in an open Git repository~\cite{gitBook}.
-This scripts are written in Python and make use of external libraries Numpy and Scikits/Audiolab that performs calls to Fortran routines for better computational efficiency.
-Part of this code has been transcribed to JavaScript and native Python with readiness, what points to uses of this contribution in Web browsers such as Firefox and Chromium~\cite{numpy, audiolab, tutpython, python}.
-
-%This are all open technologies, that is, published using licenses that allows copy, distribution and use of any part for research and derivatives.
-This way, the work here presented is available and eases co-authorship processes~\cite{Raymond,Lessig}.
-
-%\subsection{Objectives}
-%\label{subsec:objectives}
-%The main goal of this article is to present a concise set of relations among musical basic elements and sequences of PCM audio samples.
-The next section is a minimum text in which music elements are presented side-by-side with the discrete-time samples they result.
-As validation and sharing, implementations on computer code of these relations and little musical pieces where gathered the \massa\ toolbox, available online.
-
-%Secondary objectives include presenting a framework of sound synthesis with control at sample level, with potential uses in psychoacoustical experiments and high-fidelity synthesis.
-The didactic presentation of the content favors use and apprehension on a problem which calls diverse topics to be tackled: signal processing, music and psycho-acoustics, to name just a few.
-
-%\subsection{Related work}
-%Due to the general interest, and number of knowledge areas involved, there is a number of books and computer implementations that are of interest or present similarities to what is presented in this work.
-A more detailed comparison of them is pointed out in the bibliography~\cite{dissertacao}.
-There is almost no articles which could be found on the topic.
-In summary, there are computer implementations that use this analytical descriptions implicitly, but there is no such a concise and mathematical description of the processes implemented, as they aim to be libraries for sound and music.
-There are books on the topic that cover various aspects of effects and physical modeling, but none of them carry a concise description of musical elements and structures, but focus on aspects of musical sounds and ways to mimic traditional instruments.
-
 \section{Characterization of the discrete-time musical note} \label{sec:discNote}\label{sec:notaDisc}
-
 In diverse artistic and theoretical contexts, music is conceived as comprising fundamental units referred to as notes, ``atoms'' that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
@@
 \subsection{Duration}
-
-The sample frequency $f_s$ is defined as the number of samples in each second of the discrete-time signal.
-Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
-A musical note of duration $\Delta$ seconds is presented as a sequence $T_i^{\Delta}$ of $\lfloor \Delta .
-f_s \rfloor $ samples.
+The sample frequency $f_s$ is defined as the number of samples in each second of the discrete-time signal. Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
+A musical note of duration $\Delta$ seconds is presented as a sequence $T_i^{\Delta}$ of $\lfloor \Delta . f_s \rfloor $ samples.
 That is, the integer part of the multiplication is considered, and an error of at most $\delta_s$ missing seconds is admitted, which is usually fine for musical purposes.
@@
 \subsection{Volume}\label{subsec:volume}
-
 The sensation of sound volume depends on reverberation and harmonic distribution, among other characteristics described in section~\ref{sec:varInternas}.
@@
 The quantity $V_{dB}$ has the decibel unit ($dB$). 
-For each $10dB$ it is associated a ``doubled volume''.
-A handy reference is $10dB$ for each step in the intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} e \emph{fortissimo}.
+For each $10dB$ a ``doubled volume'' is associated.
+A handy reference is $10dB$ for each step in the intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} and \emph{fortissimo}.
 Other useful references are $dB$ values related to double amplitude or potency:
@@
 
-An amplitude increase by a factor slightly above of 3 is required for yielding a doubled volume.
+An amplitude increase by a factor slightly above 3 is required for yielding a doubled volume.
 These values are guides for increasing or decreasing the absolute values in the sample sequences with musical purposes.
@@
 \subsection{Pitch}
-
 To a note corresponds a sequence $T_i$ in which duration and volume are directly related to the size of the sequence and the amplitude of its samples, respectively.
@@
 \subsection{Timbre}
-
 In a sound with harmonic spectrum, the (wave) period corresponds to the cycle duration, given by the inverse of the fundamental frequency.
@@
 Figure~\ref{fig:formasDeOnda} presents the waveforms described in equations  ~\ref{sinusoid}, ~\ref{sawTooth}, ~\ref{triangular} and ~\ref{square} for $\lambda_f=100$ (period of $100$ samples).
-If $f_s=44.1kHz$, the PCM standard in Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Herz$, around A4, just above the central "C", whatever the waveform is.
+If $f_s=44.1kHz$, the PCM standard in Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Hz$, around A4, just above the central "C", whatever the waveform is.
 
@@
 As a consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise or variations typical of the natural phenomenon.
-This can be observed in Figure~\ref{fig:espectroOboe}, that shows the spectrum of the original oboe note and a note with same duration, whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}.
+This can be observed in Figure~\ref{fig:espectroOboe}, which shows the spectrum of the original oboe note and a note with same duration, whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}.
 Summing up, the natural spectrum exhibits variations in the frequencies of the harmonics, in their intensities and some noise, while the note made from the sampled period has a perfectly harmonic spectrum.
@@
 \subsection{Spectrum at sampled sound}
-
 These sinusoidal components in the discretized sound have some particularities.
@@
 
-This discussion makes it clear the equivalence ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} and ~\ref{equivalenciasFases}:
+This discussion makes it clear the equivalence in ~\ref{equivalenciasFreqs}, ~\ref{equivalenciasModulos} and ~\ref{equivalenciasFases}:
 
@@
 \subsection{The basic note}\label{notaBasica}
-
 Let $f$ be such that it divides $f_a$\footnote{As pointed before, this limitation simplifies the explanation without losing generality, and will be overcome in the next section.}.
@@
 \subsection{Spatial localization and spatialization}\label{subsec:spac}
-
 A musical note always has a spatial localization, even though this is not one of its four basic properties: the note source position at the ordinary physical space.
@@
 \subsubsection{Spatial localization}
-
 It is understood that the perception of sound localization occurs in our nervous system by three pieces of information: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, including its chest, head and ears~\cite{Roederer, hrtf, Heeger}.
@@
 
-Where $\zeta$ is the distance between ears $\zeta$, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in ITD:
+where $\zeta$ is the distance between ears $\zeta$, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in ITD:
 
@@
 
-Smith III points that reasonable concert rooms have total reverberation time of $\approx 1.9$ seconds, and that the period of first reflections is around $0.1$ seconds.
+Smith III states that reasonable concert rooms have total reverberation time of $\approx 1.9$ s, and that the period of first reflections is around $0.1$ s.
 With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08$ m ($83.79k$ samples in $f_a=44.1kHz$) before reaching the ear.
@@
 If $f_a=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_a}{100/2}=882Hz$, $\frac{f_a}{100/4}=1764Hz$ and $\frac{f_a}{100/5}=2205Hz$.
-The samples duration is very short $\frac{f_a=44.1kHz}{100} \approx 2 \text{ milliseconds}$.
+The samples duration is very short $\frac{f_a=44.1kHz}{100} \approx 2 \text{ ms}$.
 One can complete the sequence with zeroes to sum (mix) sequences with different sizes.
@@
 The musical pieces \emph{Transita para metro}, \emph{Vibra e treme}, \emph{Tremolos, vibratos e a frequência}, \emph{Trenzinho de caipiras impulsivos}, \emph{Ruidosa faixa}, \emph{Bela rugosi}, \emph{Chorus infantil}, \emph{ADa e SaRa} were made to validate and illustrate concepts of this section.
-The code that synthesizes these pieces are also part of the toolbox\cite{MASSA}.
+The code that synthesizes these pieces is also part of the toolbox\cite{MASSA}.
 
@@
 Finally, with frequency $f$ it results in $i .
-f \frac{\widetilde{\Lambda}}{f_a}$ which completes $f$ periods $\widetilde{\Lambda}$ in $1$ second, i.e.
+f \frac{\widetilde{\Lambda}}{f_a}$ which completes $f$ periods $\widetilde{\Lambda}$ in $1$ s, i.e.
 the resulting sequence presents the fundamental frequency $f$.
@@
 
-The term $\frac{i}{\Lambda-1}$ covers the interval $[0,1]$ and it is possible raise it to a power in such a way that the beginning of the transition will be smoother or steeper.
+The term $\frac{i}{\Lambda-1}$ covers the interval $[0,1]$ and it is possible to raise it to a power in such a way that the beginning of the transition will be smoother or steeper.
 This procedure is useful for energy variations with the purpose of changing the volume\footnote{The volume (psychophysical quality) is a consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is the wave total energy.
@@
 A linear fade can be used to reach zero amplitude.
-Another common solution is the use of the quartic polynomial term $x^4$, as it reaches zero without these difficulties and gets reasonably close to the curve with $\alpha=1$ as it withdraws from zero~\cite{Cook}.
+Another common solution is the use of the quartic polynomial term $x^4$, as it reaches zero without these difficulties and gets reasonably close to the curve with $\alpha=1$ as it departs from zero~\cite{Cook}.
 
@@
 z^{-2}}$.} or the calculation of Chebichev filter coefficients\footnote{Butterworth and Elliptical filters can be considered as special cases of Chebichev filters~\cite{Openheim,smith}.}.
-Both alternatives are explored by ~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University~\cite{music-dsp,Openheim}.
+Both alternatives are explored by ~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of Columbia University~\cite{music-dsp,Openheim}.
 
@@
 
-Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in volume (intensity)\footnote{The jargon may be different in other contexts.
+Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in volume (intensity).\footnote{The jargon may be different in other contexts.
 For example, in piano, tremolo is a vibrato in the classification used here.
@@
 The use of equations~\ref{vbrF} for FM is described in~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
-In fact, the simple and attractive FM behavior are observed only with linear variations, such as in~\ref{eq:fmEsp}).
-
-For the amplitude modulation (AM):
+In fact, the simple and attractive FM behavior is observed only with linear variations, such as in~\ref{eq:fmEsp}.
+
+For the amplitude modulation (AM): (esta frase está isolada? Deveria ver algo logo em seguida?
 
@@
 This is found in many ethnic and traditional musical styles~\cite{Gramani}.
-Absolute values of the impulses implies in proportions among the amplitudes of the sonic reincidences.
+Absolute values of the impulses imply in proportions among the amplitudes of the sonic reincidences.
 The use of convolution with impulses in this context is explored in the music piece \emph{Trenzinho de caipiras impulsivos}.
-These features embrace the creation of `sound amalgams' based on granular synthesis, refer to Figure~\ref{fig:pulsoSubAgl}.
+These features embrace the creation of `sound amalgams' based on granular synthesis; see Figure~\ref{fig:pulsoSubAgl}.
 This piece is a link to next section contents.
@@
 
-Following the exposition in Subsection~\ref{subsec:spac}, when an audio source or receptor is moving, its characteristics are ideally updated at each sample of the digital signal.
+According to the description in Subsection~\ref{subsec:spac}, when an audio source or receptor is moving, its characteristics are ideally updated at each sample of the digital signal.
 Speed components should be found for each ear.
-In this way, given the audio source speed (or velocity) $v_s$, with positive values if the source moves away from receptor, and receptor speed $v_r$, positive when it gets closer to audio source, the frequency is given by the well-known Doppler shift:
+Hence, given the audio source speed (or velocity) $v_s$, with positive values if the source moves away from receptor, and receptor speed $v_r$, positive when it gets closer to audio source, the frequency is given by the well-known Doppler shift:
 
@@
 There is an addendum to improve the fidelity of the physical phenomena: to increase the received power, which may be understood as being proportional to the wave shrinking or expansion: $\Delta P=P_0\left(\frac{v_r-v_s}{343.2}\right)$,
- where $P_0$ is signal power and $P$ the potency at receptor.
+ where $P_0$ is signal power and $P$ the potency at the receptor.
 Both amplitude and frequency of a moving audio source can be obtained.
@@
 Moreover, the transition is abrupt if source and receptor intersect with zero distance, otherwise, there is a monotonic progression.
-In the given case, where there is a static height $z_0$, the speed component in the direction given by the observer and source positions, the frequencies $F_i$ at the observer is given by:
+In the given case, where there is a static height $z_0$, the speed component in the direction given by the observer and source positions, the frequencies $F_i$ at the observer are given by:
 
@@
 For example, sounds originated from an old television or telephone can be simulated with a band-pass filter, allowing only frequencies between $1kHz$ and $3kHz$.
-By rejecting the frequency of electric oscillation (usually $50Hz$ or $60Hz$) and harmonics, one can remove noises caused by audio devices connected to a usual power suply.
+By rejecting the frequency of electric oscillation (usually $50Hz$ or $60Hz$) and harmonics, one can remove noises caused by audio devices connected to a usual power supply.
 A more musical application is to perform filtering in specific bands and to use those bands as an additional parameter to the notes.
@@
 
-\noindent, a reverberated auralization of a sound can be achieved by simple convolution of $R_i$ (called reverberation `impulse response') with the sound sequence $T_i$ as described in subsection~\ref{subsec:filtros}.
-Reverberation is well known for causing great interest in listeners and to provide more enjoyable sonorities.
-Furthermore, modifications in the reverberation space consists on a common resource (almost a \textit{clich\'{e}}) to surprise and attract the listener.
+\noindent a reverberated auralization of a sound can be achieved by simple convolution of $R_i$ (called reverberation `impulse response') with the sound sequence $T_i$ as described in subsection~\ref{subsec:filtros}.
+Reverberation is well known for causing great interest in listeners and to provide sonorities that are more enjoyable.
+Furthermore, modifications in the reverberation space consist in a common resource (almost a \textit{clich\'{e}}) to surprise and attract the listener.
 An implementation of the reverb recipe described here is in \massa\ toolbox~\cite{MASSA}
@@
 The variation of volume along the duration of sound is crucial to our timbre perception.
-The volume envelope known as ADSR (\emph{Attack-Decay-Sustain-Release}), has many implementations in both hardware and software synthesizers.
+The volume envelope known as ADSR (\emph{Attack-Decay-Sustain-Release}) has many implementations in both hardware and software synthesizers.
 A pioneering implementation can be found in the Hammond Novachord synthesizer of 1938 and some variants are mentioned below~\cite{ADSR}.
@@
 The lower the $\xi$, the slower the \emph{fade}, like the $\alpha$ illustrated in Figure~\ref{fig:transicao}.
-The right side of Equations~\ref{eq:adsr} can attend both introduction and ending of sound from zero intensity because they are linear.
+The right side of Equations~\ref{eq:adsr} can cater for both introduction and ending of sound from zero intensity because they are linear.
 Schematically, Figure~\ref{fig:adsr} shows the ADSR envelope in a classical implementation that supports many variations.
@@
 Twelve semitones (or half steps), equidistant to the human ear, reach an
-octave, equivalent to the human year. Therefore, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
+octave, equivalent to the human ear. Therefore, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
 $f_0$ and $f$.
@@
 This absolute accuracy is common in computational implementations.
-Real music instruments, however, exhibit deviations in order to make compatible deviations from the harmonics of notes.
+Real music instruments, however, exhibit deviations in order to make compatible departures from the harmonics of notes.
 In addition, the fixed reference $\varepsilon=2^{\frac{1}{12}}$ characterizes an equally tempered tuning.
@@
 These tunings were first formalized around two thousand years before the arrival of the equal temperament~\cite{Roederer}.
-). Two emblematic tunings are:
+Two emblematic tunings are:
 
@@
     \item The {\bf just intonation}, defined by association of diatonic scale notes with ratios of low-order integers, as found in the harmonic series.
-Considered the ionian mode (see Section~\ref{subsec:escalas}), we achieve notes correspondent to the white piano keys from C to C, i.e.
+In the ionian mode (see Section~\ref{subsec:escalas}), we achieve notes corresponding to the white piano keys from C to C, i.e.
 intervals 1, 2M, 3M, 4J, 5J, 6M, 7M, 8, are considered ratios of frequency: 1, 9/8, 5/4
@@
 
-In order to account for micro-tonality\footnote{Micro-tonality is the use of intervals less
+In order to account for micro-tonality,\footnote{Micro-tonality is the use of intervals less
 than one semitone and has ornamental and structuring functionalities in music.
@@
 By way of illustration, the traditional Thai music uses an octave division in seven notes equally spaced ($\varepsilon=2^{\frac{1}{7}}$),
-which allows intervals that have little in common with the intervals in the diatonic scale~\cite{Wisnick}}.
-It is possible to use non-integer real values as factors of $\varepsilon=2^{\frac{1}{12}}$ between frequencies, or to maintain integer values and change $\varepsilon$.
+which allows intervals that have little in common with the intervals in the diatonic scale~\cite{Wisnick}} non-integer real values can be used as factors of $\varepsilon=2^{\frac{1}{12}}$ between frequencies, or to maintain integer values and change $\varepsilon$.
 For example, a tuning really near the harmonic series
@@
    & traditional notation & number of semitones \\
-   perfects: & P1, P5, P8 & 0, 7, 12 \\
- imperfects: & m3, M3, m6, M6 & 3, 4, 8, 9 \\\hline\hline
+   perfect: & P1, P5, P8 & 0, 7, 12 \\
+ imperfect: & m3, M3, m6, M6 & 3, 4, 8, 9 \\\hline\hline
     \multicolumn{3}{|c|}{\bf dissonances} \\\hline
@@
 \begin{itemize}
-                \item Intervals are seen first by the number of steps between notes: first (unison), second, third, forth, fifth, sixth, seventh and eighth (octave).
+                \item Intervals are seen first by the number of steps between notes: first (unison), second, third, fourth, fifth, sixth, seventh and eighth (octave).
 The ninth, tenth, eleventh, etc, are compound intervals made by one or more octaves plus a ``simple'' interval inside the octave, which characterizes the compound interval.
@@
                 \item Qualities of each interval: perfect consonances --
-                i.e.\ unison, forth, fifth and octave -- are 'perfect'.
+                i.e.\ unison, fourth, fifth and octave -- are 'perfect'.
 The imperfect consonances -- i.e.\ thirds and sixths -- and dissonances -- i.e.\ seconds and sevenths -- can be major and                 minor.
@@
 
-                \item A perfect interval (unison, perfect forth, perfect fifth, perfect octave), or a major interval (major second M2,
+                \item A perfect interval (unison, perfect fourth, perfect fifth, perfect octave), or a major interval (major second M2,
 major third M3, major sixth M6 or major seventh M7), increased
 by one semitone results in an augmented interval (e.g.\
-augmented third aug3 with five semitones). The augmented forth
+augmented third aug3 with five semitones). The augmented fourth
 is also called tritone (aug4 ~ tri ~ TT).
@@
                 \item Notes played as a sequence in time configure a
-                melodic interval. The order of the notes: first
-                the lowest note or the highest note, results in an ascending or descending interval, respectively.
+                melodic interval.
+With regard to the order of the notes, when the lowest note comes first there is an ascending interval, while a descending interval is observed with the highest note first.
 
@@
                 inverted augmented interval results in a diminished interval
-                and vice-versa (inverting a doublyi-augmented results in a
-                doubly-diminished and vice-verse, etc.).
+                and vice-versa (inverting a doubly-augmented results in a
+                doubly-diminished and vice-versa, etc.).
                 An inverted perfect interval results in a perfect interval as well.
@@
 They are also reached by adding a
-                7 to the interval: P11 is an octave plus a forth ($7 + P4 = P11$), M9 is an octave plus a major second ($7 + M2 = M9$).
+                7 to the interval: P11 is an octave plus a fourth ($7 + P4 = P11$), M9 is an octave plus a major second ($7 + M2 = M9$).
 \end{itemize}
@@
 \noindent They have only major, minor and perfect intervals.
-The unique exception is the tritone presented as an augmented forth or a diminished fifth.
+The unique exception is the tritone presented as an augmented fourth or a diminished fifth.
 
@@
 Although it is not a scale, the harmonic series is often used as such.
-It is a convenient to write the notes related to each harmonic:
+It is convenient to write the notes related to each harmonic:
 
@@
         \item The repetition of chord successions (or their characteristics) fixes a trajectory and makes it possible to 
-Introduce exotic formations without incoherence.
+introduce exotic formations without incoherence.
 \end{itemize}
@@
 In fact, the techniques around atonal music aim at structures for avoiding direct relation with modes and tonality.
-Manifesting such structures is of such difficulty that the dodecafonism arouse.
+Manifesting such structures is of such difficulty that dodecafonism arises.
 The purpose of dodecafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
@@
 Nevertheless, the western listener searches for tonal traces in music and easily finds them by unexpected and tortuous paths.
-The use of dissonant intervals (especially tritones) without resolution, reinforces the absence of tonality.
+The use of dissonant intervals (especially tritones) without resolution reinforces the absence of tonality.
 In this context, while creating a music
@@
  built on top of scale's first degree and central for tonal music.
-The chords of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when a minor scale) and of the forth degree $\hat{4},\hat{6},\hat{1}$ are secondary.
+The chords of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when a minor scale) and of the fourth degree $\hat{4},\hat{6},\hat{1}$ are secondary.
 After these, other degrees are considered.
@@
 Thus, the dominant chord guides the music to the tonic.
-The triad built under the forth degree is the \textbf{subdominant} (\textit{S} or \textit{s} for a major or minor
+The triad built under the fourth degree is the \textbf{subdominant} (\textit{S} or \textit{s} for a major or minor
 subdominant, respectively) and its function is to deviate the music from the tonic.
@@
 
-The dominant counter-relative should form a minor chord. It explains the change in the forth degree by a semitone above $\hat{7}\#$. The diminished chord
+The dominant counter-relative should form a minor chord. It explains the change in the fourth degree by a semitone above $\hat{7}\#$. The diminished chord
 $\hat{7},\hat{2},\hat{4}$, is generally considered a `dominant seventh chord with omitted root'~\cite{Koellheuteur}.
@@
 
-Modulation is the change of key (tonic, or tonal center) in a music, being characterized by start and end keys, and transition artifacts.
+Modulation is the change of key (tonic, or tonal center) in music, being characterized by start and end keys, and transition artifacts.
 Keys are always conceived as related by fifths and their relative and counter-relatives. Some ways to perform modulation include:
@@
 \begin{itemize}
-    \item Transposing the discourse to a new key, without any preparation. It is a common Baroque procedure although incident at other periods as well. Sometimes it is 
+    \item Transposing the discourse to a new key, without any preparation. It is a common Baroque procedure although incident in other periods as well. Sometimes it is 
     called phrasal modulation or unprepared modulation.
@@
     \item Maintaining the tonal center and changing the key quality from major to minor
-    (or vice-verse) comprehends a parallel modulation. Keys with same tonic and
+    (or vice-versa) comprehends a parallel modulation. Keys with same tonic and
     different quality are known as homonyms.
@@
 Other inventive ways to modulate are possible, to point but one common example, the minor thirds tetrad ($E_i^{tm}$ in equations~\ref{escSim}) can be sustained to bridge to other tonalities, with the ease of its both tritones.
-The music piece \emph{Acorde cedo} explores these chord relations, and is implmemented online as part of \massa\ toolbox~\cite{MASSA}.
+The music piece \emph{Acorde cedo} explores these chord relations, and is implemented online as part of \massa\ toolbox~\cite{MASSA}.
 
@@
 when leading to a perfect consonance. The parallel motion should occur only between
-imperfect consonances and no more than three consecutive times. Disonances can be unadmitted or used when followed and preceded by consonances of joint
+imperfect consonances and no more than three consecutive times. Dissonances can be forbidden or used when followed and preceded by consonances of joint
 degrees, i.e.\ neighbor notes in a scale. The motions that lead to a
@@
 Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}, which can be heard individually if spaced by at least $50-63ms$.
-For the temporal separation between them to be perceived as duration, the period should even a bit larger, around $100ms$~\cite{microsound}.
+For the temporal separation between them to be perceived as duration, the period should even be a somewhat larger, around $100ms$~\cite{microsound}.
 It is possible to summarize duration heard as rhythm or pitch, and its transition,
@@
 are not well defined. In fact, the duration where someone begins to perceive a
-fundamental frequency or a separation between occurrences, depends on the
-listener and sonic characteristics~\cite{microsound,Roederer}.
-
-The rhythmic metric is commonly based on a basic duration called pulse, whose durations range between $0.25 and 1.5s$ ($240$
+fundamental frequency, or a separation between occurrences, depends on the
+listener and sonic characteristics~\cite{microsound,Roederer}. The rhythmic metric is commonly based on a basic duration called pulse, whose durations range between $0.25 and 1.5s$ ($240$
 and $40 BPM$, respectively).
-In music education and cognitive studies, it is common to associate these range of frequencies to the durations of the heart beat, movements of inspiration and expiration and steps of a walking or running person~\cite{Lacerda,Roederer}.
+In music education and cognitive studies, it is common to associate these range of frequencies with the durations of the heart beat, movements of inspiration and expiration and steps of a walking or running person~\cite{Lacerda,Roederer}.
 
@@
 1-9 and 12.
-Other metrics are less common, like division and grouping in 13, 17, etc, and are mainly used in contexts of experimental music and classical music of XX and XXI.
+Other metrics are less common, like division and grouping in 13, 17, etc, and are mainly used in contexts of experimental music and classical music of XX and XXI centuries.
 No matter how complex they seem, metrics are common compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
@@
 strong, weak, weak). Division in 6 is considered compound but can also
-occur as a binary division. Binary division units which suffers a ternary division yields two units divided into tree units each: strong (subdivided in strong,
+occur as a binary division. Binary division units which suffers a ternary division yields two units divided into three units each: strong (subdivided in strong,
 weak, weak) and weak (subdivided in strong, weak, weak). Another way to perform
@@
 
-\noindent where $j_k$ is the grouping level and$i_k$ is the unit itself.
+\noindent where $j_k$ is the grouping level and $i_k$ is the unit itself.
 
@@
 Each beat and subdivision, each measure and music
-section, constitutes an arc. A music in which the arcs do not present consistency with one another,
-can be understood as a music with no coherence. Coherence impression
+section, constitutes an arc. Music in which the arcs do not present consistency with one another can be understood as music with no coherence. Coherence impression
 comes, mostly, from the skilled handling of arcs in a music piece.
@@
     $\delta$ of a specific material to another instant $\Gamma'=\Gamma + \delta$
-    of the music. It is a variation that changes temporal localization in
-    a music:
+    of the music. It is a variation that changes temporal localization in a music piece:
     $\left\{s_j'\right\}=\left\{s_j^{\Gamma'}\right\}=\left\{s_j^{\Gamma+\delta}\right\}$
@@
         $\tilde{n}<0$ (elements delay) with the anti-clockwise rotation. 
-        More information about rotations in presented in subsection~\ref{estCic}.
+        Additional information about rotations is given in subsection~\ref{estCic}.
 
@@
 musical therapy, known by its utility in cases of depression and neurological injury.
-It is known that regions of the human brain responsible by sonic processing are also used for other activities, such as language-related and mathematics~\cite{Sacks,Roederer}.
+It is known that regions of the human brain responsible for sonic processing are also used for other activities, such as language-related and mathematics~\cite{Sacks,Roederer}.
 
 Paradigmatic structures guide the creation of new musical material.
-One of the most central is the tension/relaxation dipole. Traditional dipoles are relate to tonic/dominant, repetition/variation,
+One of the most central is the tension/relaxation dipole. Traditional dipoles are related to tonic/dominant, repetition/variation,
 consonance/dissonance, coherence/rupture, symmetry/asymmetry,
@@
 lucid ternary communion, `modus perfectus', opposes to the passionate
-dichotomic, `modus imperfectus'. Hereafter, there is an exposition dedicated to
+dichotomic, `modus imperfectus'. Hereafter, there is a description dedicated to
 directional and cyclic arcs.
@@
 The arcs can be decomposed in two convergent sequences: the first reaches the apex and the second returns from apex to start region.
-This apex is called climax by traditional music theory. It is
+This apex is called climax in traditional music theory. It is
 possible to distinguish between arcs whose climax is placed at the beginning, middle, end,
@@
         other domains. The possibilities considered are: climax at the beginning, at the first half, in the middle, in the second half and 
-        at the end. The x and y-axis is not properly specified since the parameters can be non-existent, as in a reference structure.}
+        at the end. The x and y-axis are not properly specified since the parameters can be non-existent, as in a reference structure.}
         \label{fig:climax}
@@
 played. Permutations occur between each line. In this case, music structure
-consists of permutations itself and some different permutations operate in the
+consists of permutations themselves and some different permutations operate in the
 cyclic behavior.
@@
 produces (ex. $[1\;3\;7\;0\;...]$). In the cyclic notation, a permutation consists
-of swaps by elements and its successors, and the last element by the first one.
+of swaps by elements and their successors, and the last element by the first one.
 E.g. $(1,2,5)(3,4)$ in cyclic notation is equivalent to $[0,2,5,4,3,1]$ in natural notation.
@@
 In the auralization of a permutation, it is not necessary to permute elements of $S_i$,
-but only some characteristic. Thus, if $p$ is a permutation and $S_i$ is a sequence of basic notes as in the end of section~\ref{notaBasica}, the
+but only a given characteristic. Thus, if $p$ is a permutation and $S_i$ is a sequence of basic notes as in the end of section~\ref{notaBasica}, the
 sequence $S_i'=p^f(S_i)=\left\{s_i^{p(f)}\right\}$ consists of the same
@@
 The \massa\ toolbox presents a computational implementation that isolates
-permutation aplication to sonic characteristics, in order to deliver
+permutation application to sonic characteristics, in order to deliver
 musical structures~\cite{MASSA,figgusOriginal,figgusEspacializacao}.
@@
 dichotomies are prominent, as explained in subsection~\ref{subsec:motivos}:
-repetition and variation, relaxation and tension, equilibrium and instability, consonance and disonance, etc.
+repetition and variation, relaxation and tension, equilibrium and instability, consonance and dissonance, etc.
 
@@
 The number of permuted
-notes can vary during the music, revealing relationship with piece
+notes can vary during the music piece, revealing relationship with piece
 duration. Harmonies can be made from triads (eqs.~\ref{triades}) with replicated
@@
 resources presented in this study. With minimal parametrization, the scripts
-generate complete musici pieces, allowing easy composition of sets of
-musics. A simple script of a few lines specifies music delivered as 16 bit
+generate complete music pieces, allowing easy composition of sets of
+music. A simple script of a few lines specifies music delivered as 16 bit
 44.1kHz PCM files (WAVE). This facility and technological
@@
 and for artistic and didactic purposes. The incorporation of programming skills is facilitated by visual aids, which
-was explored by \emph{live-coding} practices and courses based on this framework, with success.
+was explored by \emph{live-coding} practices and courses based on this framework.
 
@@
 
-The free software license and online availability of the exposed content as hypertext, 
-with the respective codes and sonic examples, strongly facilitates future collaborations and the generation of sub-products in a co-authorship fashion.
-As consequence, the expansion of \massa\ is favoured and straightforward, easing new implementarions and musical pieces development.
+The free software license and online availability of the content as hypertext, 
+with the respective codes and sonic examples, facilitate future collaborations and the generation of sub-products in a co-authorship fashion.
+As a consequence, the expansion of \massa\ is favoured and straightforward, easing new implementations and development of musical pieces.
 
 In addition, this framework permitted the formation of interests groups, with topics such as music creativity and computer music. 
-Specially, the project \url{labMacambira.sf.net} groups Brazilian and foreign co-workers in order to offer relevant contributions 
+In particular, the project \url{labMacambira.sf.net} groups Brazilian and foreign co-workers in order to offer relevant contributions 
 in diverse areas like Digital Direct Democracy, georeferencing techniques, artistic and educational activities. 
-Some of these reports are available online~\cite{dissertacao}. There are more than 700 videos, written documentats,
- original software applications and contributions in well-known external softwares such as Firefox, Scilab, LibreOffice,
+Some of these reports are available online~\cite{dissertacao}. There are more than 700 videos, written documentations,
+ original software applications and contributions in well-known external software such as Firefox, Scilab, LibreOffice,
  GEM/Puredata, to name a few~\cite{siteLM,wikiLM,vimeoLM}.
@@
 methods for the generation of appealing artistic materials. Some psychoacoustical effects were detected,
-which needs validation and should be reported, specially with~\cite{quadrosSonoros}. Other foreseen advances
-are: JavaScript version of \massa\ toolbox, better hypermedia deliverables of this framework, use guides
+which needs validation and should be reported, especially with~\cite{quadrosSonoros}. Other foreseen advances
+are: JavaScript version of \massa\ toolbox, better hypermedia deliverables of this framework, user guides
 for different ends (musical composition, psychophysical experiments, sound synthesis, education, etc), creation
-of musical pieces and open experiments to be studied with EEG recordings and further analitical specification of musical
-elements in the discrete-time representation of sound as community feedback corroborate. 
-.
+of musical pieces and open experiments to be studied with EEG recordings and further analytical specification of musical
+elements in the discrete-time representation of sound as feedback is received from the community.
+
+Acknowledgments
+This work was supported by CNPq and FAPESP (Brazil).
+
+Conferir número de referências, pois a lista parece maior do que as citações no texto?
 

### 2017-07-08 article/acm/body.tex
finished review of section 2

@@
 The note has a source which has a position at the ordinary physical three dimensional space.
-The reverberation in the environment in which sound occurs is the matter of \emph{spatialization}.
+This position is the spatial localization of the sound and might not be (modeled as) a single point.
+The reverberation in the environment in which sound occurs is one of the main topics of \emph{spatialization}, i.e.
+the consideration of the influence the environment has on the sound that reaches a listener.
 Both concepts, spatialization and spatial localization, are widely valued by audiophiles and the music industry~\cite{floEsp}.
@@
 
-Where $\zeta$ is the distance between ears $\zeta$, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in ITD:
+\noindent where $\zeta$ is the distance between ears $\zeta$, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in ITD:
 
@@
 It is possible to use IID together with ITD as a time advance for the right channel.
-It is a crucial vestige to localization perception of bass sounds and percussive sonorities~\cite{Heeger}.
+It is a crucial clue to localization perception of bass sounds and percussive sonorities~\cite{Heeger}.
 With $\Lambda_{ITD}=\lfloor ITD . f_a \rfloor$:
@@
 Given the filtering and diffraction dependent on the sound spectrum, it is hard, if not impossible, to correctly draw the confusion cone.
-Even so, the cone of confusion can be understood as a cone with its top placed in the middle of the head and growing out in the direction of each ear~\cite{hrtf}.
+Even so, the cone of confusion can be understood as a cone with its apex placed in the middle of the head and growing out in the direction of each ear~\cite{hrtf}.
 
@@
 \subsubsection{Spatialization}
-
-Spatialization results from sound reflections and absorptions by environment (e.g.
-room) surface where the note is played.
+Spatialization results from sound reflections and absorption by room/environment surface where the note is played.
 The sound propagates through the air with a speed of $\approx 343.2m/s$ and can be emitted from a source with any directionality pattern.
-When a sound pulse encounters a surface there is reflection, and there are: 1) inversion of the propagation speed component normal to the surface;  2) energy absorption, especially in trebles.
-The sonic waves propagate until they reach inaudible levels (and even further).
+When a sound pulse encounters a surface there are: 1) inversion of the propagation speed component normal to the surface;  2) energy absorption, especially in trebles.
+The sonic waves propagate until they reach inaudible levels (and further but then can often be neglected).
 As a sonic front reaches the human ear, it can be described as the original sound, with the last reflection point as the source, and the absorption filters of each surface it has reached.
@@
 There are reverberation models less related to each independent reflection, exploring valuable information to the auditory system.
-In fact, reverberation can be modeled with a set of 2 temporal and spectral sections:
+In fact, reverberation can be modeled with a set of two temporal and two spectral sections~\cite{JOSPhy}:
 
@@
 
-Smith III points that reasonable concert rooms have total reverberation time of $\approx 1.9$ seconds, and that the period of first reflections is around $0.1$ seconds.
-With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08$ m ($83.79k$ samples in $f_a=44.1kHz$) before reaching the ear.
+Smith III states that usual concert rooms have total reverberation time of $\approx 1.9$ seconds, and that the period of first reflections is around $0.1s$.
+With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08m$ ($83.79k$ samples in $f_a=44.1kHz$) before reaching the ear.
 In addition, sound reflections made after propagation for $34.32$ m ($4.41k$ samples in $f_a=44.1kHz$) have incidences less distinct by hearing.
@@
 \subsection{Musical use}\label{subsec:basMus}
-
-Once the basic note is defined, it is convenient to build musical structures with sequences based on these particles.
-The sum of elements with same index of $N$ sequences $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ with same size $\Lambda$ results in the overlapped spectral contents of each sequence, in a process referred to as mixing:
+Once the basic note is defined, it is didactically convenient to build musical structures with sequences based on these particles.
+The sum of elements with same index of $N$ sequences $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ with same size $\Lambda$ results in the overlapped spectral contents of each sequence, in a process called mixing:
 
@@
         \includegraphics[width=.5\columnwidth]{figures/mixagem__}}
-    \caption{Mixing of three sound sequences. The amplitudes are directly overlapped.}
+    \caption{Mixing of three sound sequences. The amplitudes are directly summed.}
         \label{fig:mixagem}
@@
 If $f_a=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_a}{100/2}=882Hz$, $\frac{f_a}{100/4}=1764Hz$ and $\frac{f_a}{100/5}=2205Hz$.
-The samples duration is very short $\frac{f_a=44.1kHz}{100} \approx 2 \text{ milliseconds}$.
+The samples duration is very short $\frac{f_a=44.1kHz}{100} \approx 2ms$.
 One can complete the sequence with zeroes to sum (mix) sequences with different sizes.
@@
 This mechanism is demonstrated in Figure~\ref{fig:concatenacao} with the same sequences of Figure~\ref{fig:mixagem}.
-Although the sequences are short for the usual sample rates, it is easy to observe the concatenation of sound sequences.
+Although the sequences are short for the usual sample rates, it is easy to visually observe the concatenation of sound sequences.
 In addition, each note has  a duration larger than $100ms$ if $f_a<1kHz$.

### 2017-07-08 article/acm/foobody.tex
finished review of section 2

@@
 \subsubsection{Spatialization}
-
 Spatialization results from sound reflections and absorptions by room/environment surface where the note is played.
@@
 \subsection{Musical use}\label{subsec:basMus}
-
 Once the basic note is defined, it is convenient to build musical structures with sequences based on these particles.

### 2017-07-08 article/acm/body.tex
finished review of section 2

@@
 
-\section{Variations in the basic musical note}\label{sec:internalVar}\label{sec:varInternas}
-
+\section{Variation in the basic musical note}\label{sec:internalVar}\label{sec:varInternas}
 The basic digital music note defined in section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (volume) and timbre.
 This is a useful and paradigmatic model, but it does not exhaust all the aspects of a musical note.
-First of all, characteristics of the note modify along the note itself~\cite{Chowning}.
-For example, a 3 s piano note has intensity with abrupt rise at the beginning and progressive decay, has spectrum variations with harmonics decaying and some others emerging along time.
+First of all, characteristics of the note change along the note itself~\cite{Chowning}.
+For example, a $3s$ piano note has intensity with abrupt rise at the beginning and progressive decay, has spectrum variations with harmonics decaying and some others emerging along time.
 These variations are not mandatory, but they are used in sound synthesis for music because they reflect how sounds appear in nature.
-This is considered to be so true that there is a rule of thumb: to make a sound that incites interest by itself, do internal variations on it~\cite{Roederer}.
+This is considered so important that there is a rule of thumb: to make a sound that incites interest by itself, arrange internal variations on it~\cite{Roederer}.
 To explore all the ways in which variations occur within a note is out of the scope of any work, given the sensibility of the human ear and the complexity of human sound cognition.
 In the following, primary resources are presented to produce variations in the basic note.
-It is worthwhile to recall that all the relations in this and other sections are implemented in Python and published as the public domain \massa\ toolbox.
-The musical pieces \emph{Transita para metro}, \emph{Vibra e treme}, \emph{Tremolos, vibratos e a frequência}, \emph{Trenzinho de caipiras impulsivos}, \emph{Ruidosa faixa}, \emph{Bela rugosi}, \emph{Chorus infantil}, \emph{ADa e SaRa} were made to validate and illustrate concepts of this section.
-The code that synthesizes these pieces are also part of the toolbox\cite{MASSA}.
+It is worthwhile to recall that all the relations in this and other sections are implemented in Python and published in the public domain \massa\ toolbox.
+The musical pieces \emph{Transit para meter}, \emph{Vibrates and shakes}, \emph{Tremolos, vibratos and the frequency}, \emph{Little train of impulsive hillbillies}, \emph{Noisy band}, \emph{Bela rugosi}, \emph{Children's chorus}, \emph{ADa and SaRa} were made to validate and illustrate concepts of this section.
+The code that synthesizes these pieces is also part of the toolbox\cite{MASSA}.
 
 \subsection{Lookup Table}\label{subsec:lookup}
-
 The \emph{Lookup Table} (or simply LUT) is an array for indexed operations which substitutes continuous and repetitive calculation.
-It is used to reduce computational complexity and employ functions without direct calculation, as data sampled from nature.
-In music its usage transcends these applications, as it simplifies many operations and makes it possible to use a single wave period to synthesize sounds in the whole audible spectrum, with any waveform.
+It is used to reduce computational complexity and for employing functions without calculating them directly, e.g.
+from data sampled from nature.
+In music its usage transcends these applications: it simplifies many operations and enables the use a single wave period to synthesize sounds in the whole audible spectrum, with any waveform.
 
@@
         \includegraphics[width=.7\textwidth]{figures/lut_}
-    \caption{Consultation in the \emph{lookup table} to synthesize sounds at different frequencies using a unique waveform with high resolution.
+    \caption{Search in the \emph{lookup table} to synthesize sounds at different frequencies using a unique waveform with high resolution.
 	Each \emph{i-th} sample $t_i^f$ of a sound with frequency $f$ is related to the samples in the table $\widetilde{l}$\, by the relation $t_i^{f}=\widetilde{l}_{\left\lfloor i.f\frac{\widetilde{\Lambda}}{f_a} \right\rfloor \%\,\widetilde{\Lambda}}$ where $\widetilde{\Lambda}$ is the number of sample in $\widetilde{l}$ and $f_a$ is the sampling rate.}
@@
 If $i$ is divided by the sample frequency, $\frac{i}{f_a}$
-is incremented in $1$ at each second.
-Multiplied by the period, it results in $i \frac{\widetilde{\Lambda}}{f_a}$, which covers the period in each second.
+is incremented by $1$ at each second.
+Multiplied by the period, it results in $i \frac{\widetilde{\Lambda}}{f_a}$, which covers the period in one second.
 Finally, with frequency $f$ it results in $i .
@@
 \subsection{Incremental Variations of Frequency and Intensity}\label{subsec:vars}
-
 As stated by the Weber and Fechner~\cite{Weber-Fechner} law, the human perception has a logarithmic relation with the stimulus.
@@
 
-where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a}$ is the LUT increment between two samples given the sound frequency of the first sample.
+\noindent where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a}$ is the LUT increment between two samples given the sound frequency of the first sample.
 Therefore, it is handy to calculate the elements $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ by means of the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$.
@@
 The simplest way consists of modifying the amplitude by multiplying the whole sequence by a real number.
-The increase of energy without amplitude variation is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}}.
+The increase of energy without amplitude variation is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}.}.
 It is sufficient to multiply the original sequence by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is the given coefficient and $a_{\Lambda-1}$ is a fraction of the original amplitude, which is the value to be reached at the end of the transition.
@@
 Thus, for amplitude variations:
-
 \begin{equation}\label{seqAmp}
@@
 It is often convenient to have $a_0=1$ to start a new sequence with the original amplitude and then progressively change it.
-If $\alpha=1$, the amplitude variation follows the geometric progression that defines the linear variation of volume.
+If $\alpha=1$, the amplitude variation follows the geometric progression that implies the linear variation of volume.
 Figure~\ref{fig:transicao} depicts transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by equation~\ref{eq:ampVol}.
@@
 In equation~\ref{seqAmp}, $a_0=0$ results in a division by zero and if $a_{\Lambda-1}=0$, there will be multiplication by zero.
-Both cases make the procedure useless, once no number different from zero can be represented as a ratio in relation to zero.
+Both cases make the procedure useless, once a ratio of any number in relation to zero is not well defined for our purposes.
 It is possible to solve this dilemma choosing a number that is small enough like $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum volume for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$).
-A linear fade can be used to reach zero amplitude.
-Another common solution is the use of the quartic polynomial term $x^4$, as it reaches zero without these difficulties and gets reasonably close to the curve with $\alpha=1$ as it withdraws from zero~\cite{Cook}.
+A linear fade can be used then to reach zero amplitude, if needed.
+Another common solution is the use of the quartic polynomial term $x^4$, as it reaches zero without these difficulties and gets reasonably close to the curve with $\alpha=1$ as it departs from zero~\cite{Cook}.
 
@@
 The linear transitions will be used for AM and FM synthesis, while logarithmic transitions are proper tremolos and vibratos, as developed in subsection~\ref{subsec:tvaf}.
-A non-oscillatory exploration of these variations is in the music piece \emph{Transita para metro}, whose code is online as part of the \massa\ toolbox\cite{MASSA}.
+A non-oscillatory exploration of these variations is in the music piece \emph{Transit para meter}, whose code is online as part of the \massa\ toolbox\cite{MASSA}.
 
 \subsection{Application of Digital Filters}\label{subsec:filtros}
-
-This subsection is limited to a description of sequences processing, by convolution and differential equations, and immediate applications, as its complexity escapes the scope of this study\footnote{The implementation of filters encompasses an area of recognized complexity, with dedicated literature and software implementations.
-The reader is welcome to visit the bibliography for further discussion~\cite{Openheim,smith}.}.
-Filter applications can be part of the synthesis process or made subsequently as part of processes commonly referred to as ``sound treatment''.
+This subsection is limited to a description of sequences processing, by convolution and difference equations, and immediate applications, as a thorough discussion of filtering is beyond the scope of this study\footnote{The implementation of filters encompasses an area of recognized complexity, with dedicated literature and software implementations~\cite{Openheim,smith}.}.
+Filter applications can be part of the synthesis process or made subsequently as part of processes commonly referred to as ``acoustic/sound treatment''.
 
 \subsubsection{Convolution and finite impulse response (FIR) filters}
-
 \begin{figure*}
@@
 This sample representation is called `impulse response' $\{h_i\}$.
-FIR filters are applied in the time domain of digital sound by means of convolution with the respective impulse response of the filter\footnote{It is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both sound and the impulse response, and then performing the inverse Fourier transform in the resulting spectrum.\cite{Openheim}}.
+FIR filters are applied in the time domain of digital sound by means of convolution with the respective impulse response of the filter\footnote{It is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both the sound and the impulse response, and then performing the inverse Fourier transform in the resulting spectrum.\cite{Openheim}}.
 For the purposes of this work, convolution is defined as:
@@
 
-With this procedure it is possible to apply reverberators, equalizers, \emph{delays}, to name a few of a variety of other filters for sound processing and to obtain musical/artistic effects.
+With this procedure it is possible to achieve reverberators, equalizers, \emph{delays}, to name a few of a variety of other filters for sound processing used to obtain musical/artistic effects.
 
 The impulse response can be provided by physical measures or by pure synthesis.
-An impulse response for a reverberation application, for example, can be obtained by sound recording of the environment when someone triggers a snap which resembles an impulse, or obtained by a sinusoidal sweep whose Fourier transform approximates its frequency response.
+An impulse response for a reverberation application, for example, can be obtained by sound recording of the environment when someone triggers a click which resembles an impulse, or obtained by a sinusoidal sweep whose Fourier transform approximates its frequency response.
 Both are impulse responses which, properly convoluted with the sound sequence, result in the own sound with a reverberation that resembles the original environment where the measure was made~\cite{Cook}.
@@
 Figure~\ref{fig:delays} shows the shift caused by convolution with an impulse.
-Depending on the intensity of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as pitch (from about 20 impulses per second and higher frequencies).
-In the latter case, the process resembles granular synthesis, reverbs and equalization.
+Depending on the density of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as pitch (from about 20 impulses per second and higher frequencies).
+In the latter case, the process resembles granular synthesis, reverberation and equalization.
 
@@
 Represented in increasing order of its pulse density.
-The vertical axis is related to amplitude although one should keep in mind that each plot has two or three displaced signals.}
+The vertical axis is related to amplitude although one should keep in mind that each subplot has two or three displaced signals.}
         \label{fig:delays}
@@
 \subsubsection{Infinite impulse response (IIR) filters}
-
 This class of filters, known by the acronym IIR, is characterized by having an infinite time representation, i.e.\ the impulse response does not converge to zero.
@@
 In most cases the variables may be normalized: $a_j'=\frac{a_j}{b_0}$ and $b_k'=\frac{b_k}{b_0} \Rightarrow b_0' = 1$.
-Equation~\ref{eq:diferencas} is called `difference equation' because the resulting samples $\left\{t_i'\right\}$ are given by differences between original samples $\{t_i\}$ and previous resulting ones $\left\{t_{i-k}'\right\}$.
+Equation~\ref{eq:diferencas} is called `difference equation' because the resulting samples $\left\{t_i'\right\}$ are given by differences between original samples $\{t_i\}$ and previous ones in the resulting signal $\left\{t_{i-k}'\right\}$.
 
@@
 For band-pass and band-reject (or 'notch') filters, this attenuation has two specifications: $f_c$ (in this case, the `center frequency') and bandwidth $bw$.
-In both frequencies $f_c \pm bw$ there is an attenuation of $\approx 0.707$ of the original amplitude.
+In both frequencies $f_c \pm bw$ there is an attenuation of $-3dB \approx 0.707$ of the original amplitude.
 There is sound amplification in band-pass and band-reject filters when the cutoff frequency is low and the band width is large enough.
@@
 
-For filters with other frequency responses, it is possible to apply them successively.
-Another possibility is to use a biquad 'filter receipt'\footnote{Short for 'biquadratic': its transfer function has two poles and two zeros, i.e.
-its first direct form consists of two quadratic polynomials forming the fraction: $\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1- b_1.z^{-1} -b_2 .
+It is possible to apply these filters successively in order to obtain filters with other frequency responses.
+Another possibility is to use a biquad 'filter recipe'\footnote{Short for 'biquadratic': its transfer function has two poles and two zeros, i.e.
+its first direct form consists of two quadratic polynomials in the fraction: $\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1- b_1.z^{-1} -b_2 .
 z^{-2}}$.} or the calculation of Chebichev filter coefficients\footnote{Butterworth and Elliptical filters can be considered as special cases of Chebichev filters~\cite{Openheim,smith}.}.
-Both alternatives are explored by ~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University~\cite{music-dsp,Openheim}.
+Both alternatives are explored by~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University~\cite{music-dsp,Openheim}.
 
@@
 The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_s$ in which an attenuation of $3dB$ occurs.
-The coefficients $a_0$ and $b_1$ of the IIR filter are given by the intermediate variable $x \in [e^{-\pi},1]$:
-
+The coefficients $a_0$ and $b_1$ of the IIR filter are given by $x \in [e^{-\pi},1]$:
 \begin{equation}\label{eq:passa-baixas}
@@
 \end{equation}
-
   \item High-pass filter with a simple pole, module of its frequency responses at the upper right corner of Figure~\ref{fig:iir}.
-The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of the intermediate variable $x \in [e^{-\pi},1]$:
-
+The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of $x \in [e^{-\pi},1]$:
 \begin{equation}\label{eq:passa-altas}
@@
 \end{equation}
-
-%\item Passa-banda
-%\item Rejeita-banda
-
 \item Notch filter.
@@
 The auxiliary variables $K$ and $R$ are defined as:
-
 \begin{equation}\label{eq:varAux}
@@
 \end{equation}
-
 The band-pass filter in the lower left corner of Figure~\ref{fig:iir} has the following coefficients:
-
 \begin{equation}\label{eq:passa-banda}
@@
 \end{equation}
-
 The coefficients of band-reject filter, depicted in the lower right of Figure~\ref{fig:iir}, are:
-
 \begin{equation}\label{eq:rejeita-banda}
@@
 \end{equation}
-
-%\item Biquad: pela especificação de uma frequência central, da qualidade
-%e da intensidade do filtro, este filtro é simples e usual para áudio,
-%permitindo ajustes mais finos. Diversas receitas podem ser encontradas
-%na literatura, recomendamos especialmente as diferentes especificações
-%em ~\ref{musicDSP} e ~\ref{dspguide}.
-
 \end{enumerate}
@@
 \subsection{Noise}\label{subsec:ruidos}
-
-Sounds without an explicit pitch are generally called noise~\cite{Lacerda}.
-They are important musical sounds, as noise is present in piano notes, violin, etc.
-Furthermore, the majority of percussion instruments does not exhibit an unequivocal pitch and their sounds are generally regarded as noise~\cite{Roederer}.
+Sounds without an easily recognizable pitch are generally called noise~\cite{Lacerda}.
+They are important musical sounds, as noise is present in real notes, e.g.
+emitted by a violin or a piano.
+Furthermore, many percussion instruments does not exhibit an unequivocal pitch and their sounds are generally regarded as noise~\cite{Roederer}.
 In electronic music, including electro-acoustic and dance genres, noise has diverse uses and frequently characterizes the music style~\cite{Cook}.
@@
 
-Some noises with static spectrum are listed below. They are called \emph{colored noise} since they are associated with colors.
+Some noises with static spectrum are listed below. They are called \emph{colored noise} since they are associated with colors for many reasons.
 Figure~\ref{fig:ruidos} shows the spectrum profile and the corresponding sonic sequence side-by-side.
-All five noises were generated with the same phase for each component, making it possible to observe the contributions of different parts of the spectrum.
+All five noises were generated with the same phase for each component, making it straightforward to observe the contributions of different parts of the spectrum.
 
 \begin{itemize}
-
- \item The white note has its name because its energy is distributed equally among all frequencies. It is possible to obtain white noise with the inverse transform of the following coefficients:
-
+ \item The white note has its name because its energy is distributed equally among all frequencies, such as the white color.
+It is possible to obtain white noise with the inverse transform of the following coefficients:
 \begin{equation}\label{eq:branco}
@@
  \item The pink noise is characterized by a decrease of $3dB$ per octave. This noise is useful for testing electronic devices, being prominent in nature~\cite{Roederer}. 
-
 \begin{multline}\label{eq:rosa}
@@
 \end{multline}
-
 The minimum frequency $f_{\text{min}}$ is chosen with regard to the human hearing, since a sound component with frequency below $\approx\; 20Hz$ is virtually inaudible.
@@
   \item The brown noise received this name after Robert Brown, who described the Brownian movement\footnote{Although its origin is disparate with its color association, this noise became established with this specific name, in musical contexts.
-Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more intense colors~\cite{Cook,guillaume}.} What characterizes brown noise is the decrease of $6dB$ per octave, with $\alpha_i$ in equation set~\ref{eq:rosa} being:
+Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more vivid colors~\cite{Cook,guillaume}.} What characterizes brown noise is the decrease of $6dB$ per octave, with $\alpha_i$ in equation set~\ref{eq:rosa} being:
 
@@
 An implementation of ISO 226, which is the last revision of these curves, is in the \massa\ toolbox~\cite{MASSA}.
-
 \end{itemize}
@@
 
-The noise from equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom},
-\ref{eq:azul} and \ref{eq:violeta} are presented in Figure~\ref{fig:ruidos}.
-The spectra were built with the same phase and frequency for each coefficient, making it straightforward to observe the contribution of treble harmonics and bass frequencies.
-
 \subsection{Tremolo and vibrato, AM and FM}\label{subsec:tvaf}
-
 Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in volume (intensity)\footnote{The jargon may be different in other contexts.
@@
     \item $\mu$ is a direct measure of how many Hertz are involved in the upper limit of the oscillation, while
-    \item $\nu$ is the direct measure of how many semitones (or half steps) are involved in the oscillation ($2\nu$ is the number of semitones between the upper and lower peaks of frequency oscillations of the sound $\{t_i\}$ caused by the vibrato).
+    \item $\nu$ is the direct measure of how many semitones (or half steps) are involved in the oscillation ($2\nu$ is the number of semitones between the upper and lower peaks of frequency oscillations of the sound $\{t_i\}$ in the vibrato).
 \end{itemize}

### 2017-07-08 article/acm/foobody.tex
finished review of section 2

@@
 \section{Variation in the basic musical note}\label{sec:internalVar}\label{sec:varInternas}
-
 The basic digital music note defined in section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (volume) and timbre.
@@
 \subsection{Lookup Table}\label{subsec:lookup}
-
 The \emph{Lookup Table} (or simply LUT) is an array for indexed operations which substitutes continuous and repetitive calculation.
@@
 \subsection{Application of Digital Filters}\label{subsec:filtros}
-
 This subsection is limited to a description of sequences processing, by convolution and differential equations, and immediate applications, as its complexity escapes the scope of this study\footnote{The implementation of filters encompasses an area of recognized complexity, with dedicated literature and software implementations.
@@
 \subsubsection{Convolution and finite impulse response (FIR) filters}
-
 \begin{figure*}
@@
 The coefficients $a_0$ and $b_1$ of the IIR filter are given by the intermediate variable $x \in [e^{-\pi},1]$:
-
 \begin{equation}\label{eq:passa-baixas}
@@
 \end{equation}
-
   \item High-pass filter with a simple pole, module of its frequency responses at the upper right corner of Figure~\ref{fig:iir}.
 The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of the intermediate variable $x \in [e^{-\pi},1]$:
-
 \begin{equation}\label{eq:passa-altas}
@@
 \subsection{Noise}\label{subsec:ruidos}
-
 Sounds without an explicit pitch are generally called noise~\cite{Lacerda}.
@@
 \begin{itemize}
-
  \item The white note has its name because its energy is distributed equally among all frequencies. It is possible to obtain white noise with the inverse transform of the following coefficients:
-
 \begin{equation}\label{eq:branco}
@@
 \subsection{Tremolo and vibrato, AM and FM}\label{subsec:tvaf}
-
 Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in volume (intensity).\footnote{The jargon may be different in other contexts.

### 2017-07-08 article/acm/body.tex
finished review of section 3

@@
 
-\subsubsection{Convolution and finite impulse response (FIR) filters}
+\subsubsection{Convolution and finite impulse response (FIR) filters}\label{subsec:conv}
 \begin{figure*}
@@
 
-It is convenient to use $\nu=\log_{2}\frac{f+\mu}{f} $ in this case because the maximum frequency increase is not equivalent to the maximum frequency decrease, but the semitone variation remains.
+It is convenient to use $\nu=\log_{2}\frac{f+\mu}{f} $ in this case because the maximum frequency increase is not equivalent to the maximum frequency decrease.
+The semitone/pitch variation is the invariant quantity.
 
 Figure~\ref{fig:vibrato} is the spectrogram of an artificial vibrato for a note with $1000Hz$ (between a \emph{B} and a \emph{C}), in which pitch deviation reaches one octave above and one below.
-Practically any waveform can be used to generate a sound and the vibrato oscillation pattern, with virtually any oscillation frequency and pitch deviation\footnote{The pitch deviation is called 'vibrato depth' and is generally given as semitones or cents, as convenient}.
+Practically any waveform can be used to generate a sound and the vibrato oscillation pattern, with virtually any oscillation frequency and pitch deviation\footnote{The pitch deviation is called 'vibrato depth' and is generally given in semitones or cents (one cent $= \frac{1}{100}$ of a semitone).}.
 Those oscillations with precise waveforms and arbitrary amplitudes are not possible in traditional music instruments, and thus it introduces novelty in the artistic possibilities.
@@
 \end{equation}
-and finally: 
+\noindent and finally: 
 \begin{multline}\label{trT}
@@
 The curvature is due to the logarithmic progression of the intensity.
-The tremolo frequency is $1.5Hz$ because $f_a=44.1kHz \; \Rightarrow \; \text{duration} = \frac{i_{\text{max}}=82000}{f_a}= 2s \; \Rightarrow \; \frac{3\text{oscillations}}{2s}=1.5$ oscillations per second ($Hz$).
-
-The music piece \emph{Vibra e treme} explores these possibilities given by tremolos and vibratos, both used in conjunction and independently, with frequencies $f'$, different depths ($\nu$ and $V_{dB}$), and progressive parameters variations (tremolos and vibratos occur many times together in a traditional music instrument and voices).
+The tremolo frequency is $1.5Hz$ if $f_a=44.1kHz$ because $\text{duration} = \frac{i_{\text{max}}=82000}{f_a}= 2s \; \Rightarrow \; \frac{3\text{oscillations}}{2s}=1.5$ oscillations per second.
+
+The music piece \emph{Vibrates and shakes} explores these possibilities given by tremolos and vibratos, both used in conjunction and independently, with frequencies $f'$, different depths ($\nu$ and $V_{dB}$), and progressive parameters variations (tremolos and vibratos occur many times together in a traditional music instrument and voices).
 Aiming at a qualitative appreciation, the piece also develops a comparison between vibratos and tremolos in logarithmic and linear scales.
@@
 Roughness is also generated by spectral content that produces beating~\cite{Porres,porres2009}.
-The sequence \emph{Bela Rugosi} explores this roughness threshold with concomitant tremolos and vibratos at the same voice, with different intensity and waveforms.
+The sequence \emph{Bela Rugosi} explores this roughness threshold with concomitant tremolos and vibratos at the same voice, with different intensities and waveforms.
 The corresponding code is available online in the \massa\ toolbox.
@@
 In this case, the oscillations are audible as pitch.
-Then, $f'$, $\mu$ and the waveform together change the spectrum of original sound $T_i$ in different ways for both tremolos and vibratos.
+Then, $f'$, $\mu$ and the waveform together change the audible spectrum of original sound $T_i$ in different ways for both tremolos and vibratos.
 They are called AM (\emph{Amplitude Modulation}) and FM (\emph{Frequency Modulation}) synthesis,
@@
 For musical goals, it is possible to understand FM based on the
-case of sines and, in occurrences of greater complexity, to decompose the signals into their respective Fourier components (i.e.\ sines as well).
-The FM synthesis performed with a sinusoidal vibrato with frequency $f'$ and depth $\mu$ in a sinusoidal sound $T_i$ with frequency $f$ generates bands centered in $f$ and far from each other with a distance of $f'$:
+case of sines and, when other waveforms are employed, to decompose the signals into their respective Fourier components (i.e.\ sines as well).
+The FM synthesis performed with a sinusoidal vibrato of frequency $f'$ and depth $\mu$ in a sinusoidal sound $T_i$ with frequency $f$ generates bands centered around $f$ and far from each other by $f'$:
 
@@
 In these equations, the frequency variation introduced by $\{t_i'\}$ does not follow the geometric progression that yields linear pitch variation, but reflects equation~\ref{freqLinear}.
-The use of equations~\ref{vbrF} for FM is described in~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
-In fact, the simple and attractive FM behavior are observed only with linear variations, such as in~\ref{eq:fmEsp}).
+The result of using equations~\ref{vbrF} for FM is described in~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
+In fact, the simple and attractive FM behavior is usually observed with linear oscillations, such as in~\ref{eq:fmEsp}, which yield less strident noisy sounds.
 
@@
 
-%\begin{widetext}
-%\begin{multline*}\label{eq:amEsp}
-%\begin{split}
-%\{t_i'\}_0^{\Lambda-1} =\{(1+a_i) .
-t_i\}_0^{\Lambda-1} = \left \{ \left [ 1+M.\sin \left ( f'.2\pi\frac{i}{f_a -1} \right ) \right] .P .\sin \left ( f.2\pi\frac{i}{f_a -1} \right ) \right \}_0^{\Lambda-1} = \\
-%                        =  \left\{P.\sin \left( f.2\pi\frac{i}{f_a -1}  \right %) +  \frac{P.M}{2} \left [ \sin \left( (f-f').2\pi\frac{i}{f_a -1}  \right )  + \sin \left( (f+f').2\pi\frac{i}{f_a -1}  \right ) \right ] \right \}_0^{\Lambda-1}
-%\end{split}
-%\end{multline*}
-%\end{widetext}
-
 The resulting sound is the original one together with the
@@
 
-The modulator $\{t_i'\}$ into the carrier $\{t_i\}$, for FM, is applied as:
+The modulator $\{t_i'\}$ is applied to the carrier $\{t_i\}$:
 
@@
 \subsection{Musical usages}\label{subsec:mus2}
-
-At this point, the musical possibilities are very wide.
-All characteristics, like pitch (given by frequency), timbre (given by the waveform, filters and noise), volume (manipulated by intensity) and duration (given by the number of samples), can be considered in an absolute form or treated during duration of sound (clearly, with the exception of duration itself).
-The following musical usages comprehend a collection of possibilities with the purpose of exemplifying types of sound manipulation that result in musical material.
+At this point the musical possibilities are very wide.
+All characteristics, like pitch (given by frequency), timbre (achieved by waveforms, filters and noise), volume (manipulated by intensity) and duration (given by the number of samples), can be considered in an absolute form or varied during the duration of the sound (clearly, with the exception of duration itself).
+The following musical usages comprehend a collection of possibilities with the purpose of exemplifying types of sound manipulations that result in musical material.
 Some of them are discussed in depth in the next section.
@@
 \subsubsection{Relations between characteristics}
-
 An interesting possibility is to use relations between parameters of
@@
 to note pitch, or a tremolo depth inversely proportional to
-pitch. Therefore, with equations \ref{vbrGamma}, \ref{vbrF} and \ref{trA}, it is possible to choose:
+pitch. Therefore, with equations \ref{vbrGamma}, \ref{vbrF} and \ref{trA}, it is possible to set:
 
@@
  $\nu$ and $V_{dB}$ are the respective depth values of vibrato and
-tremolo. Functions $func_a$, $func_b$ and $func_c$ are arbitrary and dependent on musical intentions. The music piece \emph{Tremolos, vibratos e a
-frequência} explores such bonds and exhibits variations in the oscillation waveform with the purpose of building a \emph{musical
+tremolo. Functions $func_a$, $func_b$ and $func_c$ are arbitrary and dependent on musical intentions. The music piece \emph{Tremolos, vibratos and the
+frequency} explores such bonds and exhibits variations in the oscillation waveform with the purpose of building a \emph{musical
 language} (details in the next section). The corresponding code is also available online as part of the \massa\ toolbox.
@@
 A musical pulse - such as specified by a BPM
-tempo - can be used with an impulse at the start of each beat, with the purpose of establishing metric and rhythms: the convolution with an impulse shifts the sound to impulse position.
-For example, two impulses equally spaced build a binary division into the
+tempo - can be implied by an impulse at the start of each beat, with the purpose of establishing metric and rhythms: the convolution with an impulse shifts the sound to impulse position, as stated in section~\ref{subsec:conv}.
+For example, two impulses equally spaced build a binary division in the
 pulse. Two signals, one with 2 impulses and the other with 3 impulses, both equally spaced in the pulse duration, yields a pulse
-maintenance with a rhythm in which eases both binary or ternary
-divisions of the pulse.
-This is found in many ethnic and traditional musical styles~\cite{Gramani}.
-Absolute values of the impulses implies in proportions among the amplitudes of the sonic reincidences.
-The use of convolution with impulses in this context is explored in the music piece \emph{Trenzinho de caipiras impulsivos}.
-These features embrace the creation of `sound amalgams' based on granular synthesis, refer to Figure~\ref{fig:pulsoSubAgl}.
-This piece is a link to next section contents.
+maintenance with a rhythm which eases both binary or ternary
+divisions of the pulse. This is found in many ethnic and traditional musical styles~\cite{Gramani}. Absolute values of the impulses entail
+proportions among the amplitudes of the sonic reincidences. The use of convolution with impulses in this context is explored in the music piece \emph{Little train of impulsive hillbillies}.
+These features encompass
+the creation of `sound amalgams' based on granular synthesis; see Figure~\ref{fig:pulsoSubAgl}.
+This piece is a link to next section contents because of the explicit organization of notes in higher level structures.
 The source code of the music piece is online, as part of the \massa\ toolbox\cite{MASSA}.
@@
 \subsubsection{Moving source and receptor, Doppler effect}
-
-Following the exposition in Subsection~\ref{subsec:spac}, when an audio source or receptor is moving, its characteristics are ideally updated at each sample of the digital signal.
-Speed components should be found for each ear.
-In this way, given the audio source speed (or velocity) $v_s$, with positive values if the source moves away from receptor, and receptor speed $v_r$, positive when it gets closer to audio source, the frequency is given by the well-known Doppler shift:
+According to the description in Subsection~\ref{subsec:spac}, when an audio source or receptor is moving, its characteristics are ideally updated at each sample of the digital signal (if fast computational rendering is not at stake).
+Velocity components should be found for each ear.
+Hence, given the audio source velocity $v_s$, with positive values if the source moves away from receptor, and receptor velocity $v_r$, positive when it gets closer to audio source, the frequency is given by the well-known shift in the Doppler effect:
 
@@
 With both frequencies $f$ and $f_0$, and the IID from the new source
-position, it is possible to create the Doppler effect.
+position, it is possible to simulate the Doppler effect.
 There is an addendum to improve the fidelity of the physical phenomena: to increase the received power, which may be understood as being proportional to the wave shrinking or expansion: $\Delta P=P_0\left(\frac{v_r-v_s}{343.2}\right)$,
- where $P_0$ is signal power and $P$ the potency at receptor.
-Both amplitude and frequency of a moving audio source can be obtained.
-If this audio source is in front of the receptor with $y_0$ m of horizontal distance and $z_0$ m of height, the distance is given by
+ where $P_0$ is signal power and $P$ the potency at the receptor.
+Both amplitude and frequency of a moving audio source can be modeled.
+If the audio source is in front of the receptor with $y_0$ meters of horizontal distance and $z_0$ meters of height, the distance is given by
 $D_i=\left\{ d_i=\sqrt{ y_{i}^{2}+z_{0}^{2} } \right\}_0^{\Lambda-1}$,
@@
 In other words, the same semitones (or fractions) added during the approach are decreased during the departure.
-Moreover, the transition is abrupt if source and receptor intersect with zero distance, otherwise, there is a monotonic progression.
-In the given case, where there is a static height $z_0$, the speed component in the direction given by the observer and source positions, the frequencies $F_i$ at the observer is given by:
+Moreover, the transition is abrupt if source and receptor intersect with zero distance, otherwise, there is a smooth progression.
+In the given case, where there is a static height $z_0$, the frequencies $F_i$ at the observer are given by:
 
@@
 \subsubsection{Filters and noises (subsections~\ref{subsec:ruidos} and~\ref{subsec:filtros})}
-
 With the use of filters, the possibilities are even wider.
-Convolve a signal to have a reverberated version of it, to remove its noise, to distort or to treat the audio aesthetically in other ways.
+Convolve a signal to have a reverberated version of it, to remove its noise, to distort or to handle
+the audio aesthetically in other ways.
 For example, sounds originated from an old television or telephone can be simulated with a band-pass filter, allowing only frequencies between $1kHz$ and $3kHz$.
-By rejecting the frequency of electric oscillation (usually $50Hz$ or $60Hz$) and harmonics, one can remove noises caused by audio devices connected to a usual power suply.
+By rejecting the frequency of an electric oscillation (usually $50Hz$ or $60Hz$) and the harmonics, one can remove noises caused by audio devices connected to the power supply.
 A more musical application is to perform filtering in specific bands and to use those bands as an additional parameter to the notes.
@@
 Chains of these filters can perform complex and more accurate filtering routines.
-The music piece \emph{Ruidosa faixa} explores filters and many kinds and noise synthesis.
+The music piece \emph{Noisy band} explores filters and many kinds and noise synthesis.
 The source code is available online as part of \massa\ toolbox.
@@
 A sound can be altered through different filtering and then mixed to create an effect known as \emph{chorus}.
-Based on what happens in a choir of singers, the sound is performed using small and potentially arbitrary modifications of parameters like center frequency, presence (or absence) of vibrato or tremolo and its characteristics, equalization, volume, etc.
+Based on what happens in a choir of singers, the sound is synthesized using small and potentially arbitrary modifications of parameters like center frequency, presence (or absence) of vibrato or tremolo and its characteristics, equalization, volume, etc.
 As a final result, those versions of the original sound are mixed together (see equation~\ref{eq:mixagem}).
-The music piece \emph{Chorus infantil} implements a chorus in many ways with different sounds.
+The music piece \emph{Child chorus} implements a chorus in many ways with different sounds.
 Its source code is available in \massa\ toolbox.
@@
 \subsubsection{Reverberation}\label{subsubsec:reverb}
-
 Using the same terms of subsection~\ref{subsec:spac}, the
 late reverberation can be achieved by a convolution with a section of pink, brown or black noise, with exponential decay of amplitude along time. Delay lines can be added as a prefix to the noise
-with the decay, and this contemplates both time parts of the reverberation: the first reflections and the late reverberation.
-Quality can be improved by varying the geometric trajectory and (mainly low-pass, subsection~\ref{subsec:filtros}) filtering by each surface where wavefront reflected before reaching the ear in the first $100-200$ ms.
+with the decay, and this accounts for both time parts of the reverberation: the early reflections and the late reverberation.
+Quality can be improved by varying the geometric trajectory and filtering by each surface where wavefront reflected before reaching the ear in the first $100-200ms$ (mainly with a LP).
 The colored noise can be gradually introduced with a \emph{fade-in}: the initial moment given by direct
@@
 
-With
+\noindent with
 
@@
 
-\noindent, a reverberated auralization of a sound can be achieved by simple convolution of $R_i$ (called reverberation `impulse response') with the sound sequence $T_i$ as described in subsection~\ref{subsec:filtros}.
-Reverberation is well known for causing great interest in listeners and to provide more enjoyable sonorities.
-Furthermore, modifications in the reverberation space consists on a common resource (almost a \textit{clich\'{e}}) to surprise and attract the listener.
+A reverberated auralization of a sound can be achieved by simple convolution of $R_i$ (called reverberation impulse response) with the sound sequence $T_i$, as described in subsection~\ref{subsec:filtros}.
+Reverberation is well known for causing great interest in listeners and to provide sonorities that are more enjoyable.
+Furthermore, modifications in the reverberation space consist in a common resource (almost a \textit{clich\'{e}}) to surprise and attract the listener.
 An implementation of the reverb recipe described here is in \massa\ toolbox~\cite{MASSA}
@@
 \subsubsection{ADSR Envelopes}
-
 The variation of volume along the duration of sound is crucial to our timbre perception.
-The volume envelope known as ADSR (\emph{Attack-Decay-Sustain-Release}), has many implementations in both hardware and software synthesizers.
+The volume envelope known as ADSR (\emph{Attack-Decay-Sustain-Release}) has many implementations in both hardware and software synthesizers.
 A pioneering implementation can be found in the Hammond Novachord synthesizer of 1938 and some variants are mentioned below~\cite{ADSR}.
-The scholastic ADSR envelope is characterized by 4 parameters: attack duration (time in which sound reaches its maximum amplitude), decay duration (follows the attack immediately), level of sustained volume (in which the volume remains stable after the decay) and release duration (after sustained section, this is the duration needed for amplitude to reach zero or final value).
-Note that the sustain duration is not specified because it is the difference between the duration itself and the durations of attack, decay and sustain.
-
-The ADSR envelope with durations $\Delta_A$, $\Delta_D$ and $\Delta_R$, with total duration $\Delta$ and sustain level $a_S$, given by the fraction of the maximum amplitude, to be applied to any sound sequence $T_i=\{t_i\}$, can be presented as:
+The canonical ADSR envelope is characterized by 4 parameters: attack duration (time in which sound reaches its maximum amplitude), decay duration (follows the attack immediately), level of sustained volume (in which the volume remains stable after the decay) and release duration (after sustained section, this is the duration needed for amplitude to reach zero or final value).
+Note that the sustain duration is not specified because it is the difference between the duration itself and the sum of the attack, decay and sustain durations.
+
+The ADSR envelope with durations $\Delta_A$, $\Delta_D$ and $\Delta_R$, with total duration $\Delta$ and sustain level $a_S$, given by the fraction of the maximum amplitude, to be applied to any sound sequence $T_i=\{t_i\}$, can be expressed as:
 
@@
 The lower the $\xi$, the slower the \emph{fade}, like the $\alpha$ illustrated in Figure~\ref{fig:transicao}.
-The right side of Equations~\ref{eq:adsr} can attend both introduction and ending of sound from zero intensity because they are linear.
+The right side of Equations~\ref{eq:adsr} can cater for both introduction and ending of sound from zero intensity because they are linear.
 Schematically, Figure~\ref{fig:adsr} shows the ADSR envelope in a classical implementation that supports many variations.
 For example, between attack and decay it is possible to add an extra partition where the maximum amplitude remains for more than a peak.
-Another common example is the use of more elaborated tracings of attack or decay.
-The music piece \emph{ADa e SaRa}, available
+Another common example is the use of more elaborated outlines of attack or decay.
+The music piece \emph{ADa and SaRa}, available
 in \massa, explores many configurations of the ADSL envelope.

### 2017-07-08 article/acm/foobody.tex
finished review of section 3

@@
 \subsection{Musical usages}\label{subsec:mus2}
-
 At this point, the musical possibilities are very wide.
@@
 maintenance with a rhythm in which eases both binary or ternary
-divisions of the pulse.
-This is found in many ethnic and traditional musical styles~\cite{Gramani}.
-Absolute values of the impulses imply in proportions among the amplitudes of the sonic reincidences.
-The use of convolution with impulses in this context is explored in the music piece \emph{Trenzinho de caipiras impulsivos}.
-These features embrace the creation of `sound amalgams' based on granular synthesis; see Figure~\ref{fig:pulsoSubAgl}.
+divisions of the pulse. This is found in many ethnic and traditional musical styles~\cite{Gramani}. Absolute values of the impulses imply
+proportions among the amplitudes of the sonic reincidences. The use of convolution with impulses in this context is explored in the music piece \emph{Trenzinho de caipiras impulsivos}.
+These features embrace
+the creation of `sound amalgams' based on granular synthesis; see Figure~\ref{fig:pulsoSubAgl}.
 This piece is a link to next section contents.
@@
 \subsubsection{Moving source and receptor, Doppler effect}
-
 According to the description in Subsection~\ref{subsec:spac}, when an audio source or receptor is moving, its characteristics are ideally updated at each sample of the digital signal.
@@
 \subsubsection{Filters and noises (subsections~\ref{subsec:ruidos} and~\ref{subsec:filtros})}
-
 With the use of filters, the possibilities are even wider.
-Convolve a signal to have a reverberated version of it, to remove its noise, to distort or to treat the audio aesthetically in other ways.
+Convolve a signal to have a reverberated version of it, to remove its noise, to distort or to treat
+the audio aesthetically in other ways.
 For example, sounds originated from an old television or telephone can be simulated with a band-pass filter, allowing only frequencies between $1kHz$ and $3kHz$.
@@
 \subsubsection{Reverberation}\label{subsubsec:reverb}
-
 Using the same terms of subsection~\ref{subsec:spac}, the

### 2017-07-09 article/acm/body.tex
reviewed until the end of tunings

@@
 
-A reverberated auralization of a sound can be achieved by simple convolution of $R_i$ (called reverberation impulse response) with the sound sequence $T_i$, as described in subsection~\ref{subsec:filtros}.
+A sound with simulated reverberation can be achieved by a simple convolution of $R_i$ (called reverberation impulse response) with the sound sequence $T_i$, as described in subsection~\ref{subsec:filtros}.
 Reverberation is well known for causing great interest in listeners and to provide sonorities that are more enjoyable.
@@
 \section{Organization of notes in music}\label{notasMusica} \label{sec:notesMusic}
-
-Consider $ S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1} $, the sequence $S_j$ of $H$ musical events $s_j$.
+Consider $S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1}$, the sequence $S_j$ of $H$ musical events $s_j$.
 Let $S_j$ be a `musical structure', composed by events $s_j$ which are musical structures themselves, e.g.\ notes.
 This section is dedicated to techniques that make $S_j$ interesting and enjoyable for audition.
+More specifically, what follows is a summary of academic music composition theory and praxis.
 
@@
 equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, for building intervals and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of events
-in $S_j$, as in equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao}, builds melodic sequences and rhythms, which are associated with the musical `horizontal thought'.
+in $S_j$, as in equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao}, builds melodic sequences and rhythms, which are associated with the `horizontal thought'.
 The fundamental frequency $f$ and the starting moment
@@
 This makes it possible to create music made by pitches (both harmony and melody) and by temporal metrics and rhythms.
+We will start by considering these aspects of musical organization as they are more traditional in music theory and usually easier to understand.
 
 \subsection{Tuning, intervals, scales and chords}\label{subsec:afinacao}
-
 \subsubsection{Tuning}
-
 Doubling the frequency is equivalent to ascending one octave ($f=2f_0$). The octave division in twelve steps is the canon of classical western music. Its usage has also been observed
-outside western tradition, as in ceremonial/religious and ethnic context~\cite{Wisnick}.
-In this system, ideally the factor given by $\varepsilon=2^{\frac{1}{12}}$ defines a semitone.
-This constitutes a note grid along the spectrum in which, given the frequency $f$, all the possible fundamental frequencies are separated by intervals which are multiples of $\varepsilon$.
-Twelve semitones (or half steps), equidistant to the human ear, reach an
-octave, equivalent to the human year. Therefore, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
+outside western tradition, e.g.
+in ceremonial/religious and ethnic contexts~\cite{Wisnick}.
+The semitones need not to be equivalent, as will become clear in the next paragraphs, but, roughly, the factor given by $\varepsilon=2^{\frac{1}{12}}$ defines a semitone.
+This entails a note grid along the spectrum in which, given the frequency $f$, all the possible fundamental frequencies are separated by intervals which are multiples of $\varepsilon$.
+Twelve successive semitones (or half steps), equidistants to the human ear, yield an
+octave, equivalent to the human ear. In other words, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
 $f_0$ and $f$.
 
-This absolute accuracy is common in computational implementations.
-Real music instruments, however, exhibit deviations in order to make compatible deviations from the harmonics of notes.
-In addition, the fixed reference $\varepsilon=2^{\frac{1}{12}}$ characterizes an equally tempered tuning.
-There are tunings in which intervals are ratios of low-order integers, based on observations of physical phenomena.
-These tunings were first formalized around two thousand years before the arrival of the equal temperament~\cite{Roederer}.
+This absolute accuracy is usual in computational implementations.
+Real music instruments, however, often present semitones that are not exactly $2^{\frac{1}{12}}$ because the (real) harmonics of the notes are not perfect multiples of the fundamental frequency.
+The fixed interval $\varepsilon=2^{\frac{1}{12}}$ characterizes an equally tempered tuning but there are other tunings.
+The first formalizations of tunings (that the scientific tradition has repported) date from around two thousand years before the advent of the equal temperament~\cite{Roederer}.
 ). Two emblematic tunings are:
-
 \begin{itemize}
-    \item The {\bf just intonation}, defined by association of diatonic scale notes with ratios of low-order integers, as found in the harmonic series.
-Considered the ionian mode (see Section~\ref{subsec:escalas}), we achieve notes correspondent to the white piano keys from C to C, i.e.
-intervals 1, 2M, 3M, 4J, 5J, 6M, 7M, 8, are considered ratios of frequency: 1, 9/8, 5/4
+    \item The {\bf just intonation}, defined by association of intervals with ratios of low-order integers, as found in the harmonic series.
+E.g.
+the white piano keys from C to C are achieved by the ratios of frequency: 1, 9/8, 5/4
     4/3, 3/2, 5/3, 15/8, 2/1.
-The following intervals are also considered: the semitone 16/15, the 'minor tone' 10/9, and the 'major tone' 9/8.
+The following intervals are also often considered: the semitone 16/15, the 'minor tone' 10/9, and the 'major tone' 9/8.
 There are many ways to perform the division of the 12 notes in the just intonation.
-
     \item The {\bf Pythagorean tuning}, based on the interval 3/2 (perfect fifth).
-The Ionian mode ratios become: 1, 9/8, 81/64, 4/3, 3/2, 27/16, 243/128, 2/1.
-These intervals are also considered to yield notes in a scale.
-Besides the intervals in the mode, also used are the minor second 256/243, the minor third 32/27, the augmented fourth 729/512, the diminished fifth 1024/729, the minor sixth 128/81 and the minor seventh 16/9.
+The 'white piano keys' become: 1, 9/8, 81/64, 4/3, 3/2, 27/16, 243/128, 2/1.
+Also often used are the 'minor second´ 256/243, the 'minor third' 32/27, the 'augmented fourth' 729/512, the 'diminished fifth' 1024/729, the 'minor sixth' 128/81 and the 'minor seventh' 16/9.
 \end{itemize}
 
-In order to account for micro-tonality\footnote{Micro-tonality is the use of intervals less
+In order to account for micro-tonality\footnote{Micro-tonality is the use of intervals smaller
 than one semitone and has ornamental and structuring functionalities in music.
-The division of the octave in $12$ notes is motivated by physical fundaments but also a \emph{convention} adopted by western classical music.
-Other tunings are incident.
-By way of illustration, the traditional Thai music uses an octave division in seven notes equally spaced ($\varepsilon=2^{\frac{1}{7}}$),
-which allows intervals that have little in common with the intervals in the diatonic scale~\cite{Wisnick}}.
-It is possible to use non-integer real values as factors of $\varepsilon=2^{\frac{1}{12}}$ between frequencies, or to maintain integer values and change $\varepsilon$.
-For example, a tuning really near the harmonic series
-itself is proposed with the equal division of the octave in $53$ notes:
-$\varepsilon_2=2^{\frac{1}{53}}$.~\cite{microtonalidade}
-In this division, notes are related by means of integers, factors of
-$\varepsilon_2$ relating frequencies by a valid interval. Note that if $S_i$ is a pitch sequence related by means of $\varepsilon_1$, a direct mapping for notes related by $\varepsilon_2$ is 
+The division of the octave in $12$ notes has physical grounds but is still a \emph{convention} adopted by western classical music.
+Other tunings are incident, e.g.
+a traditional Thai music style uses an octave division in seven notes equally spaced ($\varepsilon=2^{\frac{1}{7}}$),
+which allows intervals quite different than those found when $\varepsilon=2^{\frac{1}{12}}$~\cite{Wisnick}.} non-integer real values can be used as factors of $\varepsilon=2^{\frac{1}{12}}$ between frequencies, or one can maintain integer values and change $\varepsilon$.
+For example, a tuning that approximates the harmonic series
+is proposed with the equal division of the octave in $53$ notes:
+$\varepsilon=2^{\frac{1}{53}}$.~\cite{microtonalidade}
+Note that if $S_i$ is a pitch sequence related by means of $\varepsilon_1$, a direct mapping for notes related by $\varepsilon_2$ is 
 a new sequence $S_i'=\left\{s_i'\right\}=\left\{
 s_i \frac{\varepsilon_1}{\varepsilon_2}\right\}$. The music piece \emph{Micro
-tom} uses microtonal features and its code
+tone} uses microtonal features and its code
 is part of \massa\ toolbox.

### 2017-07-09 article/acm/foobody.tex
reviewed until the end of tunings

@@
 \subsubsection{ADSR Envelopes}
-
 The variation of volume along the duration of sound is crucial to our timbre perception.
@@
 \section{Organization of notes in music}\label{notasMusica} \label{sec:notesMusic}
-
 Consider $ S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1} $, the sequence $S_j$ of $H$ musical events $s_j$.
@@
 \subsection{Tuning, intervals, scales and chords}\label{subsec:afinacao}
-
 \subsubsection{Tuning}
-
 Doubling the frequency is equivalent to ascending one octave ($f=2f_0$). The octave division in twelve steps is the canon of classical western music. Its usage has also been observed
@@
 Two emblematic tunings are:
-
 \begin{itemize}
@@
 There are many ways to perform the division of the 12 notes in the just intonation.
-
     \item The {\bf Pythagorean tuning}, based on the interval 3/2 (perfect fifth).

### 2017-07-15 article/acm/body.tex
enhanced the intervals section

@@
 $\varepsilon=2^{\frac{1}{53}}$.~\cite{microtonalidade}
-Note that if $S_i$ is a pitch sequence related by means of $\varepsilon_1$, a direct mapping for notes related by $\varepsilon_2$ is 
-a new sequence $S_i'=\left\{s_i'\right\}=\left\{
+Note that if $S_i$ is a pitch sequence related by means of $\varepsilon_1$, the sequence with the same notes, but related by $\varepsilon_2$, is 
+$S_i'=\left\{s_i'\right\}=\left\{
 s_i \frac{\varepsilon_1}{\varepsilon_2}\right\}$. The music piece \emph{Micro
@@
 Using the ratio $\varepsilon=2^{\frac{1}{12}}$ between note frequencies (i.e.\ one semitone) the intervals in the twelve note system can be represented by integers.
-Table~\ref{eq:intervalos} summarizes each interval: its traditional nomenclature, consonance and
-dissonance properties, and number of semitones.
+Table~\ref{eq:intervalos} summarizes the intervals: traditional notation, qualifications of consonance and
+dissonances, and number of semitones.
 
@@
 \centering
-\caption{Music intervals together with their traditional notation, basic classification for dissonance and consonance, and number of semitones.
+\caption{Music intervals together with their traditional notation, basic classification for dissonances and consonances, and number of semitones.
 Unison, fifth and octave are the perfect (P) consonances.
@@
 Perfect fourth is a special case, as it is a perfect consonance when considered as an inversion of the perfect fifth and a dissonance or an imperfect consonance otherwise.
-The second special case is the tritone (A4 or aug4, d5 or dim5, tri, TT).
+Another special case is the tritone (A4 or aug4, d5 or dim5, tri or TT).
 This interval is consonant in some cultures.
-For tonal music, the tritone indicates dominant and seeks urgent resolution in a third or sixth. Due to this instability it is considered a dissonant interval.}
+	For tonal music, the tritone indicates dominant (chord or harmonic field, see Section~\ref{subsec:harmonia}) and seeks urgent resolution into a third or sixth.
+Due to this instability it is considered a dissonant interval.}
 \begin{tabular}{ c | c | c }\hline
@@
 
-The nomenclature, based on conveniences for tonal music and practical aspects of note manipulation, can be specified
+The nomenclature, based on conveniences for tonal music and practical aspects of manipulating notes, can be specified
 as follows~\cite{Roederer,Wisnick}:
@@
 \begin{itemize}
-                \item Intervals are seen first by the number of steps between notes: first (unison), second, third, forth, fifth, sixth, seventh and eighth (octave).
-The ninth, tenth, eleventh, etc, are compound intervals made by one or more octaves plus a ``simple'' interval inside the octave, which characterizes the compound interval.
-The intervals are represented by numeric digits, e.g.
-1, 3, 5 are a unison, a third and a fifth, respectively.
-
+	\item Intervals are inspected first by the number of steps between notes. The simple intervals are: first (unison), second, third, fourth, fifth, sixth, seventh and eighth (octave).
+		Each of these intervals are related to one step less the their names suggest: a third is an interval with two steps.
+		As can be noticed in Table~\ref{eq:interfvalos}, one step is not one semitone.
+		A step, in this sense, is yield by two consecutive notes in a musical scale.
+		A scale for now can be regarded as any arbitrary monotonic sequence of pitches and will be discussed in the next section.
+	\item The intervals are represented by numeric digits, e.g. 1, 3, 5 are a unison, a third and a fifth, respectively.
 	\item Qualities of each interval: perfect consonances --
-                i.e.\ unison, forth, fifth and octave -- are 'perfect'.
+                i.e.\ unison, fourth, fifth and octave -- are 'perfect'.
 The imperfect consonances -- i.e.\ thirds and sixths -- and dissonances -- i.e.\ seconds and sevenths -- can be major and minor.
-Except for the tritone.
-
+The tritone is an exception to this rule because it is a dissonant interval and cannot be major or minor.
 	\item The perfect fourth can be a perfect consonance or a dissonance according to the context and theoretical background.
-As a general rule, it can be considered a consonance except when it resolves into a fifth or a third with movimentation of just one note.
-
+As a general rule, it can be considered a consonance except when it is followed by a third or a fifth by the movement one of the notes.
 	\item Tritone is a dissonance in Western music because
-                it characterizes the ``dominant'' chord in the tonal system (see subsection~\ref{subsec:harmonia}) and represents the    instability.
-Some cultures have the interval as a consonance,                 using it in melodies and chants as a stable interval.
-
+		it characterizes the ``dominant'' chord in the tonal system (see subsection~\ref{subsec:harmonia}) and represents (or yields) instability.
+Some cultures considers the interval a consonance, using it in melodies and chants as a stable interval.
 	\item A major interval decreased by one semitone results in a minor interval. A minor interval increased by one semitone results in a major interval.
-
-                \item A perfect interval (unison, perfect forth, perfect fifth, perfect octave), or a major interval (major second M2,
-major third M3, major sixth M6 or major seventh M7), increased
+	\item A perfect interval (P1, P4, P5, or P8), or a major interval (M2, M3, M6 or M7), increased
 by one semitone results in an augmented interval (e.g.\
-augmented third aug3 with five semitones). The augmented forth
+aug3 has five semitones). The augmented forth
 is also called tritone (aug4 ~ tri ~ TT).
-
-                \item A perfect interval or a minor interval (minor second m2, minor third m3, minor sixth m6 or minor seventh m7), decreased by one semitone results in a diminished interval. The
+	\item A perfect interval or a minor interval (m2, m3, m6 or m7), decreased by one semitone results in a diminished interval. The
 diminished fifth is also called tritone (dim5 ~ tri ~ TT).
@@
 	\item An augmented interval increased by one semitone results in a `doubly augmented' interval and a diminished interval decreased by one semitone results in a `doubly diminished' interval.
-
-                \item Notes played simultaneously configure a harmonic
-                interval.
-
-                \item Notes played as a sequence in time configure a
-                melodic interval. The order of the notes: first
-                the lowest note or the highest note, results in an ascending or descending interval, respectively.
-
-                \item If the lower pitch is raised one octave, or if
-                the upper pitch is lowered one octave, the interval is
-                inverted. The sum of an interval and its inversion is
+	\item Notes played simultaneously yields a harmonic interval.
+	\item Notes played as a sequence in time yields a
+                melodic interval.  When the lowest note comes first there is an ascending interval, while a descending interval is observed when the highest comes note first. 
+	\item A simple interval is inverted if the lower pitch is raised one octave, or if
+                the upper pitch is lowered one octave. The sum of an interval and its inversion is
                 9 (e.g.\ m7 is inverted to M2: $m7+M2=9$). An inverted major
@@
                 and vice-versa (inverting a doublyi-augmented results in a
-                doubly-diminished and vice-verse, etc.).
-                An inverted perfect interval results in a perfect interval as well.
-
-                \item An interval higher than an octave is called a 'compound interval' and is classified like the interval between the same notes but in the same octave.
-They are also reached by adding a
+                doubly-diminished and vice-versa, etc).
+                An inverted perfect interval is a perfect interval as well.
+	\item An interval wider than an octave (e.g.
+ninth, tenth, eleventh) is called a 'compound interval' and is classified in terms of the interval between the same notes but in the same octave.
+Their notation can be achieved by adding a
                 7 to the interval: P11 is an octave plus a forth ($7 + P4 = P11$), M9 is an octave plus a major second ($7 + M2 = M9$).
@@
 
-The augmented/diminished intervals and the doubly augmented/doubly diminished intervals are consequences of the tonal system.
-Scale degrees (subsection~\ref{subsec:escalas}) are in fact different pitches, with specific uses and functions.
+The augmented/diminished intervals and the doubly augmented/doubly diminished intervals have the same number of semitones of other intervals (e.g.
+minor, major or perfect) and are consequences of the tonal system.
+Scale notes are in fact different pitches, with specific uses and functions.
 Henceforth, in a \textit{C flat} major scale, the tonic -- first degree -- is \textit{C flat}, not \textit{B}, and the leading tone -- seventh degree -- is \textit{B flat}, not \textit{A sharp} or \textit{C double flat}.
-In a similar fashion, the second degree of a scale can be one semitone from first degree, like the leading tone (seventh degree at one ascending semitone from the first degree), where there is a diminished third between the two semitones of the seventh and second scale degrees~\cite{Lacerda}.
-
-The descriptions presented summarize the traditional theory of music intervals~\cite{Lacerda}.
+To grasp what this entails for intervals, let the second degree (second note) of a scale to be one semitone from the first degree.
+Let also the leading tone (i.e.
+the seventh degree at one ascending semitone from the first degree).
+There is a diminished third between the seventh and second scale degrees~\cite{Lacerda}.
+Notice that the dim3 is only two semitones wide, as is the major second (or e.g. an doubly-augmented unisson!).
+
+This description summarizes the traditional theory of musical intervals~\cite{Lacerda}.
 The music piece \emph{Intervalos entre alturas} explores these intervals in both independent and related manners.
-The source code is available online in \massa\ toolbox~\cite{MASSA}.
+The source code is available online in the \massa\ toolbox~\cite{MASSA}.
 

### 2017-07-15 article/acm/foobody.tex
enhanced the intervals section

@@
 1, 3, 5 are a unison, a third and a fifth, respectively.
-
 	\item Qualities of each interval: perfect consonances --
@@
 Except for the tritone.
-
 	\item The perfect fourth can be a perfect consonance or a dissonance according to the context and theoretical background.
 As a general rule, it can be considered a consonance except when it resolves into a fifth or a third with movimentation of just one note.
-
 	\item Tritone is a dissonance in Western music because
@@
 Some cultures have the interval as a consonance,                 using it in melodies and chants as a stable interval.
-
 	\item A major interval decreased by one semitone results in a minor interval. A minor interval increased by one semitone results in a major interval.
-
 	\item A perfect interval (unison, perfect fourth, perfect fifth, perfect octave), or a major interval (major second M2,
@@
                 and vice-versa (inverting a doubly-augmented results in a
-                doubly-diminished and vice-versa, etc.).
+                doubly-diminished and vice-versa, etc).
                 An inverted perfect interval results in a perfect interval as well.
@@
 \subsubsection{Scales}\label{subsec:escalas}
-
 A scale is an ordered set of pitches.

### 2017-07-15 article/acm/body.tex
feita revisao de escalas e acordes

@@
 \subsubsection{Scales}\label{subsec:escalas}
-
 A scale is an ordered set of pitches.
+Strictly speaking, any (ordered) set of pitches can be considered a scale.
+The complexity about musical scales lean on the tradition, i.e.
+on the scales and their uses which result from practice throughout history.
 Usually, scales repeat at each octave.
-The ascending sequence with all notes from the octave division in 12 equal intervals, separated by the ratio $\varepsilon=2^{\frac{1}{12}}$, is known as the chromatic scale with equal temperament.
+The ascending sequence with all notes from the octave division in 12 equal intervals ($\varepsilon=2^{\frac{1}{12}}$) is known as the chromatic scale within the equal temperament.
 There are 5 perfectly symmetric divisions of the octave within the chromatic scale.
-These divisions are considered scales owing to the easy and peculiar uses they provide.
-Considering integers $e_i$ used as powers of the factor $\varepsilon=2^{\frac{1}{12}}$
-to multiply the lower note frequency ($f=\varepsilon^{e_i} f_0$),
-the scales are as follows:
-
+These divisions are often regarded as scales themselves owing to the easy and peculiar uses they entail.
+
+Let $e_i$ be integers indexed by $i$ such that
+$f=\varepsilon^{e_i} f_0$,
+where $f_0$ is any fixed frequency.
+The symmetric scales mentioned above can be expressed as:
 \begin{equation}\label{escSim}
@@
                     & =  \{0,1,2,3,4,5,6,7,8,9,10,11\} = \{i\}_0^{11}\\
-\text{whole tones}  & = E_i^t = \{e_i^t\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
-\text{minor thirds} & = E_i^{tm} = \{e_i^{tm}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
-\text{major thirds} & = E_i^{tM} = \{e_i^{tM}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
+	\text{whole tones}  & = E_i^{wt} = \{e_i^{wt}\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
+\text{minor thirds} & = E_i^{mt} = \{e_i^{mt}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
+\text{major thirds} & = E_i^{Mt} = \{e_i^{Mt}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
 \text{tritones}     & = E_i^{tt} = \{e_i^{tt}\}_0^{1} = \{ 0, 6 \} = \{6.i\}_0^1
@@
 
-Therefore, the third note of the whole tone scale with $f_0=200Hz$ is $f_3=\varepsilon^{e_3^t}. f_0 = 2^{\frac{6}{12}} . 200 \approxeq 282.843
-Hz$. These `scales', or patterns, generate stable structures by their intern symmetries and can be repeated in an efficient and sustained way. Section~\ref{estCic} discusses symmetries. 
+For example, the third note of the whole tone scale with $f_0=200Hz$ is $f_3=\varepsilon^{e_2^{wt}}. f_0 = 2^{\frac{4}{12}} . 200 \approxeq 251.98
+Hz$.
+These `scales', or patterns, generate stable structures by their internal symmetries and can be repeated in a sustained way which are musically effective.
+Section~\ref{estCic} discusses symmetries.
 The music piece \emph{Cristais} uses each one of these scales, in both melodic and harmonic counterpart and their source code is part of \massa\ toolbox.
 
-Diatonic modes are:
+The \emph{diatonic scales} are scales with seven notes in which the sequence consists of five whole tones and two semitones (in each octave). There are seven of them:
 
@@
                   & = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
-\text{locrian}    & = E_i^{mlo} = \{e_i^{mlo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
+\text{locrian}    & = E_i^{lo} = \{e_i^{lo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
 \text{ionian}     & = \text{major scale} =  \\
                        & = E_i^M = \{e_i^M\}_0^6 = \{0,2,4,5,7,9,11\} \\
-\text{dorian}     & = E_i^{md} = \{e_i^{md}\}_0^6 = \{0,2,3,5,7,9,10\} \\
-\text{phrygian}   & = E_i^{mf} = \{e_i^{mf}\}_0^6 = \{0,1,3,5,7,8,10\} \\
-\text{lydian}     & = E_i^{ml}=\{e_i^{ml}\}_0^6 = \{0,2,4,6,7,9,11\} \\
-\text{mixolydian} & = E_i^{mmi} = \{e_i^{mmi}\}_0^6 = \{0,2,4,5,7,9,10\}
+\text{dorian}     & = E_i^{d} = \{e_i^{d}\}_0^6 = \{0,2,3,5,7,9,10\} \\
+\text{phrygian}   & = E_i^{p} = \{e_i^{p}\}_0^6 = \{0,1,3,5,7,8,10\} \\
+\text{lydian}     & = E_i^{l}=\{e_i^{l}\}_0^6 = \{0,2,4,6,7,9,11\} \\
+\text{mixolydian} & = E_i^{mi} = \{e_i^{mi}\}_0^6 = \{0,2,4,5,7,9,10\}
 \end{split}
@@
 \noindent They have only major, minor and perfect intervals.
-The unique exception is the tritone presented as an augmented forth or a diminished fifth.
+The unique exception is the tritone found as an augmented fourth or a diminished fifth.
+% It is useful to name the diatonic scale degrees as: tonic, supertonic, mediant, subdominant, dominant, submediant, leading note (and then the tonic again).
+% Accordingly, the first degree is the tonic, the fifth degree is the dominant.
+% These names are unanimously used for the major scale and within tonal harmony (as presented in Section~\ref{subsec:harmonia}.
+% In other scales these names might be used by extension.
+% One critical exception is the dominant in the locrian scale: it is not a perfect fifth above the tonic, so it does not play the same role as in the other scales.
+% This is the main reason why the locrian is sometimes not considered a diatonic scale.
 
@@
 For example, a brief inspection reveals that
-$e_i^{ml}=d_{(i+2)\%7}+e_{i-1}^{ml}$. Then, $\kappa=2$ for the lydian mode.
-
-The minor scales have two additional forms, named melodic and harmonic:
+$e_i^{l}=d_{(i+2)\%7}+e_{i-1}^{l}$. Then, $\kappa=2$ for the lydian scale.
+
+The minor scale have two additional forms, named harmonic and melodic:
 
@@
 
-The different ascending and descending contour of the melodic minor scale is required by the leading tone (seventh and last degree, separated by one semitone from the octave, enhances tonic polarization) in the ascending
- trajectory, which is not necessary in the descending mode, and therefore it recovers the normal form.
-The harmonic scale presents the leading tone but does not avoid the augmented second between the sixth and seventh degrees; it does not consider the melodic trajectory~\cite{Harmonia}. 
-Other scales can be represented in a similar form, like the pentatonic and the modes of limited transposition by Messiaen~\cite{Messiaen}.
-
-Although it is not a scale, the harmonic series is often used as such.
-It is a convenient to write the notes related to each harmonic:
-
+The different ascending and descending contour of the melodic minor scale is required in tonal harmony contexts.
+The minor scale has one whole tone between the seventh and eigth (or first) degrees but
+the separation by one semitone is critical to the polarization of the first degree.
+This is not necessary in the descending trajectory, and therefore the scale recovers the standard form.
+The harmonic scale presents the modified seventh degree but does not avoid the augmented second between the sixth and seventh degrees; it does not consider the melodic trajectory and thus does not need to avoid the aug2~\cite{Harmonia}.
+
+Although it is not a traditional scale, the harmonic series is often used as such::
 \begin{equation}\label{eq:serieHarmonica}
@@
 
-In this scale, the frequency of the $i-th$ note $h_i$ is the frequency of $i-th$ harmonic $f=\varepsilon^{h_i} f_0$ from the spectra generated by $f_0$.
-Nature exhibits these frequencies within sounds and with all kinds of distortions.
+In this scale, the frequency of the $i$th note $h_i$ is the frequency of $i$th harmonic $f=\varepsilon^{h_i} f_0$ from the spectrum generated by $f_0$.
+Natural sounds have such frequencies (as discussed in Section~\ref{sec:discNote}) usually with deviations from the expected values and with noise.
+
+Many other scales can be expressed using the framework exposed in this section, e.g. the pentatonic scales and the modes of limited transposition of Messiaen~\cite{Messiaen}.
+
+One last observation: the words \emph{scale} and \emph{mode} are often used as synonyms both in the literature and in colloquial discussions.
+The word \emph{mode} can also be used to mean two other things:
+\begin{itemize}
+	\item an unordered set of pitches (i.e. an unordered scale).
+	\item A scale used in the context of modal harmony, in the sense presented in Section~\ref{subsec:harmonia}.
+\end{itemize}
 
 \subsubsection{Chords}\label{subsec:acordes}
-
-The simultaneous occurrence of three or more notes is observed by means of chords. Chords are based on triads, especially in tonal music. Triads are built by two successive thirds
+A musical chord is implied by the simultaneous occurrence of three or more notes. Chords are often based on triads, especially in tonal music. Triads are built by two successive thirds
 within 3 notes: root, third and fifth.
-If the lower note of a chord is different from the root, this chord is an inverted chord as opposed to the chord in its root position.
+If the lower note of a chord is the root, the chord is in the root position, otherwise it is an inverted chord.
 A closed position is any in which no chord note fits between two consecutive
-notes~\cite{Lacerda}, any non-closed position is an open position. In closed and fundamental position,
-and with fundamental in $0$, triads are described as:
-
+notes~\cite{Lacerda}, any non-closed position is an open position. In closed and fundamental positions,
+and with the fundamental in $0$, triads can be expressed as:
 \begin{equation}\label{triades}
@@
 are also common.
+These are often interpreted as the result of further extending the succession of thirds.
+E.g. $\{0,2,4,7\}$ will often be understood as a major chord with a major ninth (a major ninth has 14 semitones and $14-12 = 2$).
 
 For general guidance:
-
 \begin{itemize}
         \item A fifth confirms the root (fundamental).
-        \item Major or minor thirds suggest major or minor chord qualities.
+		There are theoretical discussion about why this happens, and the most usual arguments are that the fifth is the first (non-octave) harmonic of a note and that the harmonics of the fifth are in the harmonics of the fundamental.
+		Important here is to grasp the fact that musical theory and practice assures that the fifth establishes the fundamental as the root of a chord.
+        \item Major or minor thirds from the root entails major or minor chord qualities.
         \item Every tritone, especially if built between a major third and a minor seventh, tends to resolve into a third or a sixth.
-        \item Note duplication is avoided. If duplication is needed, the preferred order is: the root, fifth, third and seventh.
-        \item Note omission is avoided in the triad. If needed, the fifth is first considered for omission, than third and, if needed, the fundamental.
+        \item Note duplication is avoided. If duplication is needed, the preference is, in descending order: the root, fifth, third and seventh.
+        \item Note omission is avoided in the triad. If needed, the fifth is first considered for omission, then third and then the fundamental.
         \item It is possible to build chords with notes different from triads, particularly if they obey a recurrent logic or musical concatenation that justifies these different notes.
@@
         \item The repetition of chord successions (or their characteristics) fixes a trajectory and makes it possible to 
-Introduce exotic formations without incoherence.
+introduce exotic formations without implying in musical incoherence.
 \end{itemize}

### 2017-07-29 article/acm/body.tex
revisada parte de harmonia

@@
         \item Chords built by successive intervals different from thirds – like fourths and seconds -- are recurrent in compositions of advanced tonalism or experimental character.
-        \item The repetition of chord successions (or their characteristics) fixes a trajectory and makes it possible to 
-introduce exotic formations without implying in musical incoherence.
+        \item The repetition of chord successions (or of characteristics they hold) fixes a trajectory and makes it possible to 
+introduce exotic arrangements without implying musical incoherence.
 \end{itemize}
@@
 \subsection{Atonal and tonal harmonies, harmonic expansion and modulation}\label{subsec:harmonia}
-
-Omission of basic tonal chaining is the key to obtain modal and atonal harmonies. In the absence of these minimal tonal structures, 
-harmony is considered modal if the notes match with some diatonic scale (see equations~\ref{eq:escalas}) or if notes are presented in a small number.
-If basic tonal chainings are absent and notes do not match any diatonic scale and are diverse and dissonant (by relation with each other) enough to avoid reduction by polarization, harmony is atonal.
-In this classification, the modal harmony is not tonal or atonal and is reduced to the incidence of notes within the diatonic scale (or simplifications) and to the absence of tonal structures.
+Omission of basic tonal structures is the key to achieving modal and atonal harmonies. In the absence of minimal tonal organization,
+harmony is (usually) considered modal if the notes match with some diatonic scale (see equations~\ref{eq:escalas}) or if there is only a small number of notes.
+If basic tonal progressions are absent and notes do not match any diatonic scale and are enough diverse and dissonant (between themselves) to avoid reduction of the notes by polarization\footnote{By polarization we mean having some notes that are way more important than others and to which the other notes are ornaments or subordinates.}, the harmony is atonal.
+In this classification, the modal harmony is not tonal or atonal and is reduced to the incidence of notes within a (most often diatonic) scale and to the absence of tonal structures.
 Following this concept, one observes that atonal harmony is hard to be realized and, indeed, no matter how dissonant and diverse
-a set of notes is, tonal harmonies arise if not avoided~\cite{harmEXT}.
+a set of notes is, tonal harmonies arise very easily if not avoided~\cite{harmEXT}.
 
 \subsubsection{Atonal harmony}
-
-In fact, the techniques around atonal music aim at structures for avoiding direct relation with modes and tonality.
-Manifesting such structures is of such difficulty that the dodecafonism arouse.
+In fact, atonal music techniques avoid that a direct relation of the notes with modes and tonality be established.
+Manifesting such atonal structures is of such difficulty that the dodecafonism emerged.
 The purpose of dodecafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
@@
 In this context, the tonic becomes difficult to be established.
-Nevertheless, the western listener searches for tonal traces in music and easily finds them by unexpected and tortuous paths.
-The use of dissonant intervals (especially tritones) without resolution, reinforces the absence of tonality.
-In this context, while creating a music
+Nevertheless, the western listener automatically searches for tonal elements in music and obstinately finds them by unexpected and tortuous paths.
+The use of dissonant intervals (especially tritones) without resolution reinforces the absence of tonality.
+In this context, while creating a musical
 piece, it is possible to:
-
 \begin{itemize}
-     \item Repeat notes. By considering immediate repetition as an extension of previous incidence, the use of the same note in sequence does not add relevant information.
-     \item To play adjacent notes at the same time, making harmonic intervals and chords.
-     \item Present durations and pauses with liberty, respecting notes order.
-     \item Vary by extension, transposition, translation, inversion, retrograde and retrograde inversion. See subsections~\ref{subsec:motivos}
-     and~\ref{subsec:usosmusicais3} for more details.
-     \item Accounting for the presence of note structures, it is possible to use variations in orchestration, articulation, spatialization, among others.
+     \item Repeat notes. By considering immediate repetition as an extension of the previous incidence, the use of the same note in sequence does not add relevant information.
+     \item To play adjacent notes (e.g. of a dodecafonic progression) at the same time, making harmonic intervals and chords.
+     \item Use durations and pauses with liberty, respecting notes order.
+     \item Vary note sequences by temporal magnification and translation or pitch transposition and by pitch sequence inversion, retrograde and retrograde inversion.
+See subsections~\ref{subsec:motivos}
+     and~\ref{subsec:usosmusicais3} for what these terms mean.
+     \item Make variations in orchestration, articulation, spatialization, among other possibilities in presenting the same notes.
 \end{itemize}
 
-The atonal harmony can be observed, paradigmatically, within these presented conditions (a simple dodecaphonic model). Most of what was written by great dodecafonic composers, 
+The atonal harmony can be observed, paradigmatically, within these presented conditions (which is a simple dodecaphonic model). Most of what was written by great dodecafonic composers,
 e.g. Alban Berg and even Schoenberg, had the purpose of mixing tonal and atonal techniques.
+Most frequently, atonal music is not strictly dodecafonic, but "serial", i.e. they use the same kind of techniques based in a sequence of notes (called the series or row).
 
 \subsubsection{Tonal harmony}
-
-In the XX century, rhythmic music and with emphasis on sonorities/timbres, extended the concepts of tonality and harmony.
-Hence, tonal harmony has strong influence on artistic movements and commercial venues.
-In addition, dodecafonism itself is considered of tonal nature because it denies tonal characteristics of polarization.
+In the XX century, music with emphasis on sonorities/timbres and rhythm, extended the concepts of tonality and harmony.
+Even so, tonal harmony is very often in artistic movements and commercial venues.
+In addition, dodecafonism itself is sometimes considered of tonal nature because it was conceived to deny tonal characteristics of polarization.
 In tonal or modal music, chords -- like the ones listed in
-equations~\ref{triades}-- built on the root note of each degree of a scale  -- such as listed in equations~\ref{eq:escalas} --  form the pilars of the harmonic field.
-Music harmony aims at studying the incidence of chord progressions and chaining rules.
-Even a monophonic melody generates harmonic fields, making it possible to observe the suggested chords at each passage.
+equations~\ref{triades}-- built on the root note of each degree of a scale  -- such as listed in equations~\ref{eq:escalas} --  form the pillars of harmony.
+Tonal (and modal) harmony aims at understanding the incidence of chord progressions and chaining rules.
+Even a monophonic melody entails harmonic fields, making it possible to perceive the chord progression even in unaccompanied melodies.
 
 In the traditional tonal music, a scale has its tonic (first
-degree) on any note, and can be major (with the same notes of Ionian mode) or minor (same notes as Eolian mode, 'natural minor', which has both harmonic and melodic versions, see equations~\ref{eq:escalasMenores}).
+degree) on any note, and can be major (with the same notes of Ionian mode) or minor (same notes as Eolian mode, 'natural minor', which has both harmonic and melodic versions, as in equations~\ref{eq:escalasMenores}).
 The
@@
 $\hat{1},\hat{3},\hat{5}$ is the first degree chord,
- built on top of scale's first degree and central for tonal music.
-The chords of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when a minor scale) and of the forth degree $\hat{4},\hat{6},\hat{1}$ are secondary.
-After these, other degrees are considered.
-The `traditional harmony' comprises conventions and stylistic techniques to create chainings with such chords~\cite{Harmonia}.
-
-The `functional harmony' ascribes functions to these three central chords and tries to understand their use by means of these functions.
-The chord built on top of the first degree is the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its function (role) consists of maintaining a center, usually referred to as a ``ground'' for the music.
+ built on top of the scale's first degree and central for tonal music.
+The chords of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when in a minor scale) and of the forth degree $\hat{4},\hat{6},\hat{1}$ are also important.
+The triads build on the other degrees less important then these and are usually understood in relation to them.
+The `traditional harmony' comprises conventions and stylistic techniques to create progressions with such chords~\cite{Harmonia}.
+
+The `functional harmony' ascribes functions to the three main chords and tries to understand their use by means of these functions.
+The chord built on top of the first degree is the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its function (role) consists on maintaining a center, usually referred to as a ``ground'' for the music.
 The chord built on the fifth degree is the \textbf{dominant} (\textit{D}, the dominant is always major) and its function is to lean for the tonic
@@
 Thus, the dominant chord guides the music to the tonic.
-The triad built under the forth degree is the \textbf{subdominant} (\textit{S} or \textit{s} for a major or minor
+The triad built under the fourth degree is the \textbf{subdominant} (\textit{S} or \textit{s} for a major or minor
 subdominant, respectively) and its function is to deviate the music from the tonic.
-The process aims at confirming the tonic using tonic-dominant-tonic chains which are expanded by using other chords in different ways.
+The tonal discourse aims at confirming the tonic using tonic-dominant-tonic chains which are expanded by using other chords in numerous ways.
 
@@
 sR). The precise functions and musical effects of these chords are
-controversial. Table~\ref{tab:harmonia} shows relations between the triads built in each degree of the major scale.
+controversial but are basically the same as the chords they are associated to. Table~\ref{tab:harmonia} shows relations between the triads built in each degree of the major scale.
 
@@
 \centering
-\caption{Summary of tonal harmonic functions of the major scale. Tonic is the music center, the dominant goes to the tonic and the subdominant moves the music away from
-the tonic. The three chords can, at first, be freely replaced by their
-respective relative and counter-relatives.}
+\caption{Summary of tonal harmonic functions on the major scale. Tonic is the musical center, the dominant goes to the tonic and the subdominant moves the music away from
+the tonic. The three chords can, in principle, be freely replaced by their
+respective relative or counter-relatives.}
 \begin{tabular}{l | c | r}
@@
 
-The dominant counter-relative should form a minor chord. It explains the change in the forth degree by a semitone above $\hat{7}\#$. The diminished chord
-$\hat{7},\hat{2},\hat{4}$, is generally considered a `dominant seventh chord with omitted root'~\cite{Koellheuteur}.
-In the minor mode, there is a change in $\hat{7}$ by an ascending semitone to yield a unique semitone separating $\hat{7}$ and $\hat{1}$, and making the dominant possible (that should be major and goes to the tonic).
-In this way, the dominant is always major, for both major and minor scale and, therefore, even in a minor tone the relative dominant remains a third below, and in the counter-relative, a third above.
+The dominant counter-relative should form a minor chord. It explains the change in the forth degree by a semitone above $\hat{4}\#$. The diminished chord
+$\hat{7},\hat{2},\hat{4}$, is generally considered a `dominant seventh chord with the root omitted'~\cite{Koellheuteur}.
+In the minor mode, there is a change in $\hat{7}$ by an ascending semitone to achieve a separation between $\hat{7}$ and $\hat{1}$ of a semitone.
+This is important for the dominant function (which should lean to the tonic).
+In this way, the dominant is always major, for both major and minor scale and, therefore, even in a minor tone the relative dominant remains a third below, and the counter-relative remains a third above.
 
 \subsubsection{Tonal expansion: individual functions and chromatic mediants}
-
-%% dominante e subdominante individualmente, ou, dominante e subdominante individual?
-
-Each one of these chords can be confirmed and developed by performing their individual dominant or subdominant, which are the triads based
+Each chord can be stressed and developed by performing their individual dominant or subdominant, which are the triads based
 on a fifth above or a fifth below, respectively. These individual dominants and subdominants,
 in the same way, have also subdominants and individual dominants of their own. Given a tonality, any chord can occur, no matter
-how distant it is from the harmonic field and from the notes within the scale.
-The unique condition is that the occurrence presents a coherent trajectory of dominants and subdominants to the original tonality.
-
-Mediants, or 'chromatic mediants', are two for each chord: the upper
-mediant, formed by the root at the third of original chord; and the lower mediant, formed by the fifth at the third of original chord.
+how distant it is from the most basic chords and from the notes of the scale.
+The unique condition is that the occurrence presents a coherent trajectory of dominants and subdominants (or their relatives and counter-relatives) to the original tonality.
+
+There are two mediants, or 'chromatic mediants', for each chord: the upper
+mediant, formed with the root at the third of original chord; and the lower mediant, formed by the fifth at the third of original chord.
 Both chords also are formed by a third, but with a chromatic alteration regarding the original chord.
@@
 of the original triad.
-It is worth observing that a major chord has major chromatic and doubly chromatic mediants.
+A major chord has major chromatic and doubly chromatic mediants.
 A minor chord has minor chromatic and doubly chromatic mediants.
 (Recall that relatives and counter-relatives have opposite major-minor quality.)
-This relation between chords is considered advanced tonalism, sometimes even considered expansion and dissolution of tonalism,
-with strong and impressing effects although they are perfectly consonant triads. Chromatic mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss,
-among others, and are quite simple to be realized~\cite{Harmonia,Salzer}.
+This relation between chords is considered advanced tonalism, sometimes even considered as an expansion and dissolution of tonalism,
+with strong and impressive effects although they are simple, consonant triads.
+Chromatic mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss, among others~\cite{Harmonia,Salzer}.
 
 \subsubsection{Modulation}
-
-Modulation is the change of key (tonic, or tonal center) in a music, being characterized by start and end keys, and transition artifacts.
-Keys are always conceived as related by fifths and their relative and counter-relatives. Some ways to perform modulation include:
+Modulation is the change of key (tonic, or tonal center) in music, being characterized by start and end keys, and transition artifacts.
+Keys are always (thought of as) related by fifths and their relatives and counter-relatives. Some ways to perform modulation include:
 
 \begin{itemize}
-    \item Transposing the discourse to a new key, without any preparation. It is a common Baroque procedure although incident at other periods as well. Sometimes it is 
+    \item Transposing the discourse to a new key, without any preparation. It is a common Baroque procedure although incident in other periods as well. Sometimes it is 
     called phrasal modulation or unprepared modulation.
@@
     subdominant, to confirm change in key and harmonic field.
-    \item Use of chromatic alterations to reach a chord in the new key starting from a chord in the previous key. Called chromatic modulation.
+    \item Use of chromatic alterations to reach a chord in the new key by starting from a chord in the previous key. Called chromatic modulation.
     \item Featuring a unique note, possibly repeated or suspended with no accompaniment, common to start and end keys, it constitutes a peculiar way
@@
     \item Maintaining the tonal center and changing the key quality from major to minor
-    (or vice-verse) comprehends a parallel modulation. Keys with same tonic and
-    different quality are known as homonyms.
+    (or vice-versa) is a `parallel modulation'. Keys with same tonic but
+		different (major/minor) qualities are known as homonyms.
 \end{itemize}
 
-The dominant has great importance and is a natural pivot into modulations,
-which leads to the circle of fifths~\cite{Harmonia,Salzer,Koellheuteur,Harmony}.
+The dominant has great importance and is a natural pivot in modulations,
+a fact that leads to the circle of fifths~\cite{Harmonia,Salzer,Koellheuteur,Harmony}.
 Other inventive ways to modulate are possible, to point but one common example, the minor thirds tetrad ($E_i^{tm}$ in equations~\ref{escSim}) can be sustained to bridge to other tonalities, with the ease of its both tritones.
-The music piece \emph{Acorde cedo} explores these chord relations, and is implmemented online as part of \massa\ toolbox~\cite{MASSA}.
+The music piece \emph{Acorde cedo} explores these chord relations, and is implemented online as part of \massa\ toolbox~\cite{MASSA}.
 

### 2017-07-29 article/acm/foobody.tex
feita revisao de escalas e acordes
revisada parte de harmonia

@@
 \subsubsection{Chords}\label{subsec:acordes}
-
 The simultaneous occurrence of three or more notes is observed by means of chords. Chords are based on triads, especially in tonal music. Triads are built by two successive thirds
@@
 and with fundamental in $0$, triads are described as:
-
 \begin{equation}\label{triades}
@@
 For general guidance:
-
 \begin{itemize}
@@
 \subsection{Atonal and tonal harmonies, harmonic expansion and modulation}\label{subsec:harmonia}
-
 Omission of basic tonal chaining is the key to obtain modal and atonal harmonies. In the absence of these minimal tonal structures, 
@@
 \subsubsection{Atonal harmony}
-
 In fact, the techniques around atonal music aim at structures for avoiding direct relation with modes and tonality.
@@
 piece, it is possible to:
-
 \begin{itemize}
@@
 \subsubsection{Tonal harmony}
-
 In the XX century, rhythmic music and with emphasis on sonorities/timbres, extended the concepts of tonality and harmony.
@@
 \subsubsection{Tonal expansion: individual functions and chromatic mediants}
-
-%% dominante e subdominante individualmente, ou, dominante e subdominante individual?
-
 Each one of these chords can be confirmed and developed by performing their individual dominant or subdominant, which are the triads based
@@
 \subsubsection{Modulation}
-
 Modulation is the change of key (tonic, or tonal center) in music, being characterized by start and end keys, and transition artifacts.

### 2017-07-29 article/acm/body.tex
revisada parte de ritmo

@@
 \subsection{Counterpoint}\label{subsec:contraponto}
-
-Counterpoint is defined as the conduction of simultaneous melodic lines, or ``voices''.
-The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues.
+Counterpoint is the conduction of simultaneous melodic lines, or ``voices''.
+The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues~\cite{count1, count2}.
 It is possible to
 summarize counterpoint rules, and it is known that Beethoven --
-among others -- also outlined such a briefing of counterpoint.
+among others -- also outlined such a digest of counterpoint.
 
@@
 
-The purpose of counterpoint is to conduct voices in a way that they sound independent. In order to do that, the relative motion of voices (in pairs) is crucial and
-categorized as: direct, oblique and contrary (see Figure~\ref{fig:movContraponto}).
-The parallel motion is a direct motion.
-The gold rule here is to take care of the direct motions, avoiding them
-when leading to a perfect consonance. The parallel motion should occur only between
-imperfect consonances and no more than three consecutive times. Disonances can be unadmitted or used when followed and preceded by consonances of joint
-degrees, i.e.\ neighbor notes in a scale. The motions that lead to a
-neighbor note in the scale sound coherent. When having 3 or more voices, the melodic
-importance lies in the higher and lower voices, in this order~\cite{Fux,Tragtenberg,SchoenbergContra}.
+The purpose of (scholastic) counterpoint is to conduct voices in a way that they sound independent. In order to do that, the relative motion of voices (in pairs) is crucial and
+categorized as: direct, oblique and contrary, as depicted in Figure~\ref{fig:movContraponto}.
+The parallel motion is a direct motion in which the starting and final intervals are the same.
+The golden rule here is to take care of the direct motions, avoiding them
+when ending in a perfect consonance. The parallel motion should occur only between
+imperfect consonances and no more than three consecutive times. Dissonances can be forbidden or used when followed and preceded by consonances of neighbor
+degrees, i.e.\ adjacent notes in a scale. The motions that lead to a
+neighbor note in the scale sound coherent and are prioritized. When having 3 or more voices, the melodic
+importance lies mainly in the highest and then in the lowest of the voices~\cite{Fux,Tragtenberg,SchoenbergContra}.
 
@@
 \subsection{Rhythm}\label{subsec:ritmo}
-
 Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}, which can be heard individually if spaced by at least $50-63ms$.
 For the temporal separation between them to be perceived as duration, the period should even a bit larger, around $100ms$~\cite{microsound}.
-It is possible to summarize duration heard as rhythm or pitch, and its transition,
+It is possible to summarize the durations heard as rhythm or pitch, and its transition,
 as in table~\ref{tab:duracoes}~\cite{Alfaix, microsound}.
@@
 \begin{table*}[htp!]
-\caption{Transition of durations heard as rhythm until heard as pitch.}
+\caption{Durations heard as rhythm, as pitch and transition.}
 \begin{tabular}{  l | r r r r   r r r    r r r || r r  } \hline
@@
 
-The duration transition span is minimized because the limits
+The transition span in Table~\ref{tab:duracoes} is minimized because the limits
 are not well defined. In fact, the duration where someone begins to perceive a
 fundamental frequency or a separation between occurrences, depends on the
-listener and sonic characteristics~\cite{microsound,Roederer}.
-
-The rhythmic metric is commonly based on a basic duration called pulse, whose durations range between $0.25 and 1.5s$ ($240$
-and $40 BPM$, respectively).
-In music education and cognitive studies, it is common to associate these range of frequencies to the durations of the heart beat, movements of inspiration and expiration and steps of a walking or running person~\cite{Lacerda,Roederer}.
+listener and sonic characteristics~\cite{microsound,Roederer}. The rhythmic metric is commonly based on a basic duration called pulse, whose typical durations range between $0.25 and 1.5s$ ($240$
+and $40 BPM$, respectively~\footnote{BPM stands for Beats Per Minute and is just a frequency measure like Herz, but is the number of incidences per minute intead of second.
+BPM is often used as a measure of musical tempo and of heart rate.}).
+In music education and cognitive studies, it is common to associate this range of frequencies with the durations of the heart beat, movements of respiration and steps of a walking or running person~\cite{Lacerda,Roederer}.
 
 The pulse is subdivided into equal parts and is also repeated in sequence. These relations (division and concatenation) usually follow relations of small
-order integers. The occurrences, in written and ethnic
-music, of musical pulse division (and their sequential grouping at
-time), are, in ascending order: 2, 4 and 8, after that 3, 6 (two groups of 3 or 3 groups of 2) and 9 and 12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, complementing
+integers. By far, the most often musical pulse divisions (and their sequential groupings), in written and ethnic
+music, are: 2, 4 and 8; 3, 6 (two groups of 3 or 3 groups of 2), 9 and 12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, completing
 1-9 and 12.
-Other metrics are less common, like division and grouping in 13, 17, etc, and are mainly used in contexts of experimental music and classical music of XX and XXI.
-No matter how complex they seem, metrics are common compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
+Other metrics are less common, like division or grouping in 13, 17, etc, and are mainly used in experimental music or classical music of the XX and XXI centuries.
+No matter how complex they seem, metrics are almost always compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
 A schematic illustration is shown in Figure~\ref{fig:pulsoSubAgl}.
@@
 
-Dual relations (simple measures and binary divisions) are frequent in dance rhythms and celebrations, and are called ``imperfect''. Ternary relations are common in
-ritualistic music and is related to the sacred, and are called ``perfect''. Strong units (accents) fall in the `head' of bars, the downbeats, of which the first in
-every bar is the greatest. The head of a unit is the first part of the subdivision. In binary divisions (2, 4
+Binary divisions are frequent in dance rhythms and celebrations, and are called ``imperfect''. Ternary relations are common in
+ritualistic music and is related to the sacred, and are called ``perfect''.
+Strong units (accents) fall in the `head' of the units (the first subdivision) and are called downbeats.
+In binary divisions (2, 4
 and 8), strong units alternate with weak units
@@
 strong, weak, weak). Division in 6 is considered compound but can also
-occur as a binary division. Binary division units which suffers a ternary division yields two units divided into tree units each: strong (subdivided in strong,
-weak, weak) and weak (subdivided in strong, weak, weak). Another way to perform
-the division in 6 is ternary division units that divides as binary,
+occur as a binary division. Binary division units which suffer a ternary division yields two units divided into three units each: strong (subdivided in strong,
+weak, weak) and weak (also subdivided in strong, weak, weak). Another way to perform
+the division in 6 is ternary division units that subdivides as binary,
 resulting in: a strong unit (subdivided in strong and weak) and two weak units
-(subdivided in strong and weak for each).
-
-The accent in the weak beat is the backbeat, whereas a note starting in a weak beat persisting across the strong beat is the syncope.
-
-Notes can occur inside and outside of these \emph{'musical metric'} divisions. In most well-behaved cases, notes occur exactly on these divisions, with greater incidence on strong beat attacks.
-In extreme cases, time metric cannot be perceived~\cite{Roederer}. 
-Small variations along the grid are crucial for musical interpretation and
-different styles~\cite{Cook}.
+(subdivided in strong and weak each).
+
+An accent in the weak beat is the `backbeat', whereas a note starting in a weak beat persisting across the strong beat is a `syncope'.
+These are often found in ethnic and popular music and was used with parsimony in classical music before the XX century.
+
+Notes can occur inside and outside of these \emph{'musical metric'} divisions. In most well-behaved cases, notes occur exactly on these divisions, with greater incidence on strong beats.
+In extreme cases, rythmic metric cannot be perceived~\cite{Roederer}. 
+Small variations along the temporal grid are crucial for musical interpretation and
+styles~\cite{Cook}.
 
@@
 the first pulse agglomeration be level $j=1$ and so on.
-Hence, $P_i^j$ is the $i$-th unit at grouping level $j$: $P^0_{10}$ is the tenth pulse, $P^{1}_3$ is the third grouped unit (and, possibly, the third measure),
+Accordingly, let $P_i^j$ be the $i$-th unit at grouping level $j$: $P^0_{10}$ is the tenth pulse, $P^{1}_3$ is the third grouped unit (possibly the third measure),
 $P^{-1}_2$ is the second part of pulse subdivision.
-The limits of $j$ are of special interest: pulse divisions are durations perceivable as rhythm; furthermore, the pulses sum, at its maximum, a music or a cohesive set of musics.
+The limits of $j$ are of special interest: pulse divisions are durations perceivable as rhythm; furthermore, the pulses sum, at its maximum, a music or a cohesive set of musical pieces.
 In other words, a duration given
-by $P^{min(j)}_i$, $\forall \; i$, is greater than $50 ms$ and the durations
-summed together $\sum_{\forall i}P^{\text{max}(j)}_i$ are less than a few
+by $P^{min(j)}_i$, $\forall \; i$, should be greater than $50 ms$ and the durations
+summed together $\sum_{\forall i}P^{\text{max}(j)}_i$ should be less than a few
 minutes or, at most, a few hours.
@@
 Each level $j$ has some sections $i$. When $i$ has three different
-values (or multiple of three) there is a perfect relation. When $i$ has only
-two, four or eight different values, than there is an imperfect relation,
+values (or multiple of three) there is a perfect (i.e. ternary) relation. When $i$ has only
+two, four or eight possible values, than there is an imperfect relation,
 as shown in Figure~\ref{fig:pulsoSubAgl}. Any unit (note) of a given musical sequence with a time metric can be unequivocally
@@
 
-\noindent where $j_k$ is the grouping level and$i_k$ is the unit itself.
+\noindent where $j_k$ is the grouping level and $i_k$ is the unit itself.
 
 As an example, consider $P^{-1,0,1}_{3,2,2}$ as the third subdivision $P^{-1}_3$ of the
-second pulse $P^0_2$ and of the second pulse group $P^1_2$ (possibly second measure). Each unit $P_i^j$ can be associated with a sequence of temporal samples $T_i$ that comprehends a
+second pulse $P^0_2$ and of the second pulse group $P^1_2$ (possibly second measure). Each unit $P_i^j$ can be associated with a sequence of temporal samples $T_i$ that constitutes a
 note. The music piece \emph{Poli Hit Mia} uses different metrics (available as part of \massa\ toolbox).
@@
 \subsection{Repetition and variation: motifs and larger units}\label{subsec:motivos}
-
 Given the basic musical structures, both frequential (chords and scales) and rhythmic (simple, compound and complex beat divisions and agglomerations), it is

### 2017-07-29 article/acm/foobody.tex
revisada parte de ritmo

@@
 \subsection{Counterpoint}\label{subsec:contraponto}
-
 Counterpoint is defined as the conduction of simultaneous melodic lines, or ``voices''.
@@
 \subsection{Rhythm}\label{subsec:ritmo}
-
 Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}, which can be heard individually if spaced by at least $50-63ms$.

### 2017-07-29 article/acm/body.tex
finisehd first review of everything

@@
 The concept of arcs is essential in this context: by departing from a place and returning, an arc is made.
+One important case is the arc from and to the absence of a unit: from the beginning to the end.
 The audition of melodic and harmonic lines is permeated by
-musical arcs due to the cognitive nature of the musical hearing. 
-The note can be considered the smaller arc, and each motif and melody as an arc as well.
+arcs due to the cognitive nature of the musical hearing: as the mind divides an excerpt, and groups excerpts, each of the units yields an arc.
+Accordingly, the note can be considered the smallest (relevant) arc, and each motif and melody as an arc as well.
 Each beat and subdivision, each measure and music
-section, constitutes an arc. A music in which the arcs do not present consistency with one another,
-can be understood as a music with no coherence. Coherence impression
+section, constitutes an arc. Music in which the arcs do not present consistency with one another can be understood as music with no coherence. Coherence impression
 comes, mostly, from the skilled handling of arcs in a music piece.
@@
 Musical arcs are abstract structures and amenable to basic operations.
-A spectral arc, like a chord, can be inverted, enlarged and permuted, to mention a few possibilities.
+A spectral arc, like a chord, can be inverted, magnified and permuted, to mention just a few possibilities.
 Temporal arcs, like a melody, a motif, a measure or a note, are also
-capable of variations. Recalling that
-$S_j=\left\{s_j=T_i^j=\{t_i^{j}\}_0^{\Lambda_j-1}\right\}_0^{H-1}$ is a sequence
+prone to variations. Let
+$S_j=\left\{s_j=T_i^j=\{t_i^{j}\}_0^{\Lambda_j-1}\right\}_0^{H-1}$ be a sequence
 of $H$ musical events $s_j$, each event with its $\Lambda_j$ samples $t_i^j$
-(refer to the beginning of this section~\ref{notasMusica}), the basic techniques
-can be described as:
-
+(refer to the beginning of this section~\ref{notasMusica} if needed). Bellow is a list of basic techniques
+for variation.
 \begin{itemize}
@@
     where $\Gamma$ is the duration between the beginning of the piece (or considered
-    section) and the first event $s_0$ of original structure $S_j$. One should observe that
+    section) and the first event $s_0$ of original structure $S_j$, and
     $\delta$ is the time offset of the displacement.
@@
 
-    \item Temporal reversion consists of generating a sequence with elements
-    in reverse order of the original sequence $S_j$, thus: $S_j'=\left\{s_j'\right\}_0^{H-1}=\left\{s_{(H-j-1)}\right\}_0^{H-1}$.
+    \item Temporal reversion consists on generating a sequence with elements
+    in the reverse order of the original sequence $S_j$, thus: $S_j'=\left\{s_j'\right\}_0^{H-1}=\left\{s_{(H-j-1)}\right\}_0^{H-1}$.
 
@@
         and the pitch of any frequency is: $\Xi=\Xi_0 +12
-        \log_2\left(\frac{f}{f_0}\right)$.  
-        In the MIDI protocol, $\Xi_{f_0}=55Hz$ while pitch $\Xi_0=33$ marks
-        a \textit{A 1}. Another good MIDI reference is $\Xi_{f_0}=440Hz$ and
-        $\Xi_0=69$. The difference ($\Xi_1 - \Xi_2$) is measured in semitones.
+		\log_2\left(\frac{f}{f_0}\right)$.\footnote{In the MIDI protocol, $\Xi_{f_0}=55Hz$ when pitch $\Xi_0=33$
+	(an \textit{A1} note). Another good MIDI reference is $\Xi_{f_0}=440Hz$ and
+		$\Xi_0=69$ (\textit{A4}). The difference ($\Xi_1 - \Xi_2$) is in semitones.
         $\Xi$ is not a measure in semitones: $\Xi=1$ is not a semitone, it is a note with an audible frequency as rhythm, with
-        less than 9 occurrences each second (see table~\ref{tab:duracoes}).
-
-        \item Interval inversion is: 1) the inversion of note pitch order, within octave equivalence, such as described in~\ref{subsec:intervalos}, and $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{\varepsilon.
+		less than 9 occurrences each second (see table~\ref{tab:duracoes}).}
+
+        \item Interval inversion is either: 1) the inversion of note pitch order, within octave equivalence, such as described in~\ref{subsec:intervalos}, and $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{\varepsilon.
 f_0}\right\}$ with selective $s_j=2,\;1/2,\;\text{or}\;1$; or 2) the inversion of interval orientation.
@@
         a number of positions ahead or behind, with the care to fill empty positions
-        with events out of convenient slots. Thus, a
+        with events which are out of the slots. Thus, a
         rotation of $\tilde{n}$ positions is $S'_n=S_{(n+\tilde{n})\%H}$. If
         $\tilde{n}<0$, it is sufficient to use $\tilde{n}'=H-\tilde{n}$. It is
-        reasonable to associate $\tilde{n}<0$ (events advance) with the clockwise rotation and
+        usual to associate $\tilde{n}<0$ (events advance) with the clockwise rotation and
         $\tilde{n}<0$ (elements delay) with the anti-clockwise rotation.
-        More information about rotations in presented in subsection~\ref{estCic}.
+        Additional information about rotations is given in subsection~\ref{estCic}.
 
@@
 
-    \item Changes in articulation, orchestration and spatialization, 
+    \item Changes in articulation, orchestration and spatialization, or
     $s_j'=s_j^{*_j}$, where $*_j$ is the new characteristic incorporated by 
@@
 
-    \item Accompaniment. Both orchestration and melodic lines presented when $S_j$ occurs can suffer modifications and be considered as a variation of $S_j$ itself.
+    \item Accompaniment. Both orchestration and melodic lines present when $S_j$ occurs can suffer modifications and be considered as a variation of $S_j$ itself.
 \end{itemize}
 
-From these presented processes, others are derived, as the inverted retrograde, the temporal contraction with an external suffix, etc.
-A whole process of mental and neurological activity is unleashed, responsible for feelings, memories and imaginations, typical of a diligent music listening.
+% scramble 
+From these processes, many others are derived, such as the inverted retrograde, the temporal contraction with an external suffix, etc.
+Variations are often thought about in the terms above but are also often very loose, such as an arbitrary shuffle of the notes in a melody which the composer finds interesting.
+As a result, a whole process of mental and neurological activity is unleashed for relating the arcs, responsible for feelings, memories and imaginations, typical of a diligent music listening.
 This cortical activity is critical to
 musical therapy, known by its utility in cases of depression and neurological injury.
-It is known that regions of the human brain responsible by sonic processing are also used for other activities, such as language-related and mathematics~\cite{Sacks,Roederer}.
+Also, tt is known that regions of the human brain responsible for sonic processing are also used for other activities, such as for performing verbal discourse and mathematics~\cite{Sacks,Roederer}.
 
 Paradigmatic structures guide the creation of new musical material.
-One of the most central is the tension/relaxation dipole. Traditional dipoles are relate to tonic/dominant, repetition/variation,
+One of the most central is the tension/relaxation dipole. Traditional dipoles are tonic/dominant, repetition/variation,
 consonance/dissonance, coherence/rupture, symmetry/asymmetry,
 equivalence/difference, arrival/departure, near/far, stationary/moving,
-etc. Ternary relations tend to relate to the circle and to unification. The
+etc. All these dipoles are often thought of as parallel or even as equivalent. Ternary relations tend to relate to the circle and to unification. The
 lucid ternary communion, `modus perfectus', opposes to the passionate
-dichotomic, `modus imperfectus'. Hereafter, there is an exposition dedicated to
+dichotomic, `modus imperfectus'. Hereafter, there is a description dedicated to
 directional and cyclic arcs.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \subsection{Directional structures}\label{subsec:dir}
-
 The arcs can be decomposed in two convergent sequences: the first reaches the apex and the second returns from apex to start region.
 This apex is called climax by traditional music theory. It is
-possible to distinguish between arcs whose climax is placed at the beginning, middle, end,
+usual to distinguish between arcs whose climax is at the beginning, middle, end,
 and at the first and second half of the duration. These structures are
 shown in Figure~\ref{fig:climax}. The varying parameter can be non-existent, a case in which
-the arc consists only of a reference structure~\cite{Schoenberg}.
+the arc consists only of a reference structure~\cite{Schoenberg}\footnote{A case which resembles a note without the fundamental frequency.
+Brass instruments can emit such notes and the listener identifies the fundamental frequency even though it is absent.}.
 
@@
         \caption{Canonical distinctions of musical climax in a given melody and
-        other domains. The possibilities considered are: climax at the beginning, at the first half, in the middle, in the second half and 
-        at the end. The x and y-axis is not properly specified since the parameters can be non-existent, as in a reference structure.}
+        other arcs. The possibilities considered are: climax at the beginning, at the first half, in the middle, in the second half and 
+        at the end. The x and y-axis are not properly specified since the parameters can be non-existent, as in a reference structure.}
         \label{fig:climax}
@@
 Henceforth, it is possible for a given sequence to have
-numberous distinct arcs, with different sizes and climax. 
+numerous distinct arcs, with different sizes and climax. 
 This is an interesting and useful resource, and the correlation of arcs yields coherence~\cite{Salzer}.
@@
 $1.61803398875...$. The sequence converges fast even with high discrepant
-initial values. If $x_0=1$ and $x_1=100$, and $y_n=\frac{x_n}{x_{n+1}}$ an
-auxiliary sequence, the error for the fraction of first values with
+initial values. If $x_0=1$ and $x_1=100$, and $y_n=\frac{x_n}{x_{n+1}}$, the error for the first values with
 respect to the golden ratio is, approximately, $\{ e_n \}
@@
 -7.14, 2.937, -1.09, 0.42, -0.1601, \\ 0.06125, -0.02338\}$. The Fibonacci sequence
-presents exactly the same error progression, but starts at the second step of the most discrepant
-case ($\frac{1}{1}\approx\frac{100+1 = 101}{100}$).
+presents the same error progression, but starts at the second step of a more discrepant initial setting
+($\frac{1}{1}\approx\frac{100+1 = 101}{100}$).
 
@@
 \subsection{Cyclic structures}\label{estCic}
-
-The philosophical understanding that human thought is founded on the concept of similarities and differences (e.g. perceived in stimuli and objects), places symmetries
-at the core of cognitive processes~\cite{Deleuze}. Mathematically, symmetries are algebraic groups, and a finite group is always isomorphic to a permutation
+The philosophical understanding that human thought is founded on the concept of similarities and differences (e.g. perceived in stimuli), places symmetries
+at the core of cognition~\cite{Deleuze}. Mathematically, symmetries are algebraic groups, and a finite group is always isomorphic to a permutation
 group (by Cayley's theorem).
-In a way, this states that permutations represent any symmetry in a
+In a way, this states that permutations can express any symmetry in a
 finite system~\cite{gruposFascination}.
@@
 The successive application of permutations generates cyclic arcs~\cite{change,Zamacois,permMusic}.
-Two academic studies were dedicated to generating musical structures~\cite{figgusOriginal, figgusEspacializacao}.
+Two academic studies were dedicated to generating musical structures using permutation groups~\cite{figgusOriginal, figgusEspacializacao}.
 Any permutation set can be used as a generator of algebraic groups~\cite{permMusic}.
@@
 permutation $p_1$ and another permutation $p_2$, and, comparing both initial and final orderings, observe another permutation $p_3$.
-Every element $p$ operated with itself a sufficient number of times $n$ reaches the identity element $p^n=e$ (otherwise the group would be infinite, generated
-by $p$). The lower $n\;:\;p^n=e$ is the element order. Thus, a finite
+Every element $p$ operated with itself a sufficient number of times $n$ reaches the identity element $p^n=e$ (otherwise the group generated by $p$ would be infinite).
+The lower $n\;:\;p^n=e$ is the element order.
+Thus, a finite
 permutation $p$, successively applied, reaches the initial ordering of its
-elements, and makes a cycle. This cycle, if used for parameters of musical notes or structures,
-yields a cyclic musical arc.
+elements, and makes a cycle. This cycle, if used for parameters of notes or other musical structures,
+yields a cyclic arc.
 
@@
 bells played one after another and then played again, but in a different
-order. This process repeats until it reaches the initial ordering. The set of
-different traversed orderings is a \emph{peal}. Table~\ref{tab:change}
+order. This process repeats until it reaches the initial ordering. The sequence of
+different orderings is a \emph{peal}. Table~\ref{tab:change}
 presents a traditional \emph{peal}, named ``Hunt'', for 3 bells (\textcolor{red}{1}, \textcolor{blue}{2} and \textcolor{green}{3}), which explores
 all possible orderings. Each line indicates one bell ordering to be
-played. Permutations occur between each line. In this case, music structure
-consists of permutations itself and some different permutations operate in the
+played. Permutations occur between each line. In this case, the musical structure
+consists of permutations and some different permutations add into a
 cyclic behavior.
@@
 \subsection{Musical idiom?}
-
-In numerous studies, there are models and explorations of a `musical language',
-many can somewhat be considered `linguistics applied to music'
-and some even discern different `musical
-idioms'~\cite{Lerdahl, Harmonia, Salzer,Alfaix}. In a simple way, a musical idiom
-is the result of chosen material together with repetition of elements and
-of relations between elements, along a music piece. In these matters, 
+In numerous studies, there are models, discussions and exploitation of a `musical language'.
+Some of them are linguistic theories applied to music
+and some discern different `musical
+idioms'~\cite{Lerdahl, Harmonia, Salzer,Alfaix}. Simply put, a musical idiom
+is the result of chosen materials together with variation techniques and
+relations established between elements along a music piece. In these matters,
 dichotomies are prominent, as explained in subsection~\ref{subsec:motivos}:
-repetition and variation, relaxation and tension, equilibrium and instability, consonance and disonance, etc.
+repetition and variation, relaxation and tension, stability and instability, consonance and dissonance, etc.
+A thorough discussion of what can be considered a musical language is out of the scope of this article, but this brief consideration of the subject is useful as a onvergence of all the previous content.
 
 \subsection{Musical usages}\label{subsec:usosmusicais3}
-
-First, the basic note was defined and characterized in quantitative terms (section~\ref{sec:notaDisc}).
+First, the basic note was defined and characterized in quantitative terms in section~\ref{sec:notaDisc}.
 Next, the internal note
-composition was addressed and both internal transitions and immediate sonic treatment
-were understood (section~\ref{varInternas}). Finally, this section aims at organizing these notes in music. The gamma of resources and consequent infinitude
-of praxis possibilities is a typical and highly relevant situation for artistic contexts~\cite{Harmonia,Webern}.
-
-There are studies for each of the presented resources.
+composition was addressed within both internal transitions and elementary sonic treatment
+(section~\ref{varInternas}). Finally, this section aims at organizing these notes in music. The numerous resources and consequent infinitude
+of praxis possibilities is typical and highly relevant for artistic contexts~\cite{Harmonia,Webern}.
+
+There are studies and further developments for each of the presented resources.
 For example, it is possible to obtain `dirty' triadic harmonies (with notes out of the triad) by superposition of perfect fourths.
-Another interesting example is the simultaneous
-presence of rhythms in different metrics, constituting what is
+Another interesting example is the superimposition of rhythms in different metrics, constituting what is
 called \emph{polyrhythm}. The music piece \emph{Poli-hit mia} in \massa\ toolbox~\cite{MASSA} explores these simultaneous metrics by impulse trains convolved with notes.
@@
 Microtonal scales are important for 20th
-century music~\cite{microtonalidade} and present diverse remarkable results, e.g.
-fourths of tone in the Indian music. The musical sequence \emph{MicroTom} in \massa\ toolbox~\cite{MASSA} explores these resources, including microtonal melodies and microtonal harmonies
+century music~\cite{microtonalidade} and yielded diverse remarkable results throughout history, e.g.
+fourths of a tone ($\epsilon=2^{\frac{1}{24}}$) is often used in some genres of Indian and Arabic music.
+The musical sequence \emph{MicroTom} in \massa\ toolbox~\cite{MASSA} explores these resources, including microtonal melodies and harmonies
 with many notes in a very reduced note scope.
@@
 As in subsection~\ref{subsec:mus2}, relations between
-parameters are powerful ways to acquire musical pieces.
+parameters are powerful ways to achieve musical pieces.
 The number of permuted
 notes can vary during the music, revealing relationship with piece
-duration. Harmonies can be made from triads (eqs.~\ref{triades}) with replicated
-notes at each octaves and more numerous as minor the depth and frequency of
-vibratos (eqs.~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}),
-among other uncountable possibilities.
-
-The symmetries at octave divisions (eqs.~\ref{escSim}) and the
+duration. Harmonies can be made from triads (equations~\ref{triades}) with duplicated
+notes at each octave and more numerous duplication when the depth and frequency of
+vibratos are lower (equations ~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}).
+Incontestably, the possibilities are very wide, which is evidenced in the wide range of musical pieces and styles.
+
+The symmetries at octave divisions (equation~\ref{escSim}) and the
 symmetries presented as permutations (table~\ref{tab:change} and
-eqs.~\ref{eq:groups}) can be used together. In the music piece \emph{3 trios}
-this association is done in a systematic way in order to enable a dedicated
-listening. This is an instrumental piece, not included as a source
+equations~\ref{eq:groups}) can be used together. In the music piece \emph{3 trios}
+this association is done in a systematic way in order to achieve a specific style. This is an instrumental piece, not included as a source
 code but available online~\cite{3Trios}.
 
-\emph{PPEPPS} (Pure Python EP: Projeto Solvente) is an EP synthesized using
+\emph{PPEPPS} (Pure Python EP: Project Solvent) is an EP (Extended Play) synthesized using
 resources presented in this study. With minimal parametrization, the scripts
-generate complete musici pieces, allowing easy composition of sets of
-musics. A simple script of a few lines specifies music delivered as 16 bit
+generate complete musical pieces, allowing easy composition of sets of
+music. A simple script of a few lines specifies music delivered as 16 bit
 44.1kHz PCM files (WAVE). This facility and technological
-demystification create aesthetical possibilities for both sharing and education.
+arrangement creates aesthetic possibilities for both sharing and education.
 
@@
 \label{cap:conclusao}
-
-This concise presentation relates basic musical elements to digital sound.
-The reader is invited to access \emph{Scripts} and \massa, where these relations are computationally implemented.
-The didactic report along the paper and the supplied scripts eases and encourages the use of the proposed framework.
-
-The possibilities provided by these results include psychoacoustic experiments, 
-interfaces for the generation of noise and other high fidelity sounds and musical structures,
-and for artistic and didactic purposes. The incorporation of programming skills is facilitated by visual aids, which
-was explored by \emph{live-coding} practices and courses based on this framework, with success.
-
-This work systematically investigates the parameterization issues (like the tremolo, ADSR, etc.) in a high fidelity, which has significant artistic utility.
-Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, such as reviewed in ~\cite{dissertacao}.
-
-The free software license and online availability of the exposed content as hypertext, 
-with the respective codes and sonic examples, strongly facilitates future collaborations and the generation of sub-products in a co-authorship fashion.
-As consequence, the expansion of \massa\ is favoured and straightforward, easing new implementarions and musical pieces development.
-
-In addition, this framework permitted the formation of interests groups, with topics such as music creativity and computer music. 
-Specially, the project \url{labMacambira.sf.net} groups Brazilian and foreign co-workers in order to offer relevant contributions 
-in diverse areas like Digital Direct Democracy, georeferencing techniques, artistic and educational activities. 
-Some of these reports are available online~\cite{dissertacao}. There are more than 700 videos, written documentats,
- original software applications and contributions in well-known external softwares such as Firefox, Scilab, LibreOffice,
- GEM/Puredata, to name a few~\cite{siteLM,wikiLM,vimeoLM}.
-
-Future work includes application of these results in machine learning and artificial intelligence 
-methods for the generation of appealing artistic materials. Some psychoacoustical effects were detected,
-which needs validation and should be reported, specially with~\cite{quadrosSonoros}. Other foreseen advances
-are: JavaScript version of \massa\ toolbox, better hypermedia deliverables of this framework, use guides
-for different ends (musical composition, psychophysical experiments, sound synthesis, education, etc), creation
-of musical pieces and open experiments to be studied with EEG recordings and further analitical specification of musical
-elements in the discrete-time representation of sound as community feedback corroborate.
+We believe that this article is effective in relating musical elements to digital audio.
+We aimed at achieving a concise presentation of the subject because it involves many knowledge fields, and therefore can very easily blast into thousands of pages.
+Some readers might benefit from the text alone, but the \emph{Scripts} in the \massa toolbox, where all the relations presented here are directly and simply implemented as software (in Python), are very helpful for one to achieve elaborated implementations and deeper understandings.
+The scripts include routines that render musical pieces that illustrate the concepts.
+This is valuable since art (e.g. music) can involve many non-trivial processes and is often deeply glamurized, which results in a nearly unmanagable terrain for a newcommer.
+Moreover, this didactic report and the supplied open source scripts should facilitate the use of the proposed framework.
+
+The possibilities provided by this exposition pour from both the organization of knowledge and the ability to achieve sounds which are extremely true to the models.
+For example, one can produce noises with arbitrary spectrum resolution and a musical note can be synthesized with the parameters (e.g. of a vibrato) updated sample-by-sample.
+These abilities can be crucial e.g. for psychoacoustic experiments or for the development of
+interfaces to generate high quality musical sounds or pieces.
+Simply put, it is compelling
+for many scientific, artistic and didactic purposes.
+A fact evidenced by a number of uses this framework had in courses, in software implementations and in making music~\ref{videos,vivace,massaRepo,dissertacao}.
+
+This work systematically investigates parameterization issues (like for the tremolo, ADSR, etc.) in a sample-by-sample synthesis, which entails high/extreme fidelity and if of significant artistic convenience.
+Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, as far as the authors know, such as testified in the literature review in~\cite{dissertacao}, where the books, articles and open software that deal with the issue are related to this framework.
+
+The free software license and online availability of the content,
+with the respective codes and sonic examples, facilitate future collaborations and the generation of sub-products in a co-authorship fashion.
+As consequence, the expansion of \massa\ is favoured and straightforward, easing new implementations and development of musical pieces.
+The scripts can be divided in three groups: implementation of all the equations and topics of music theory; routines for rendering musical pieces that illustrate the concepts; scripts that render the figures of this article and the article itself.
+
+In addition, this framework permitted the formation of interest groups in topics such as musical creativity and computer music.
+In particular, the project \url{labMacambira.sourceforge.net} groups Brazilian and foreign co-workers
+in diverse areas that range from digital direct democracy and georeferencing to art and education.
+This was only possible because of the audiovisual utility in many contexts, in particular because of the knowledge and mastery condensed in the \massa\ framework.\footnote{There are more than 700 videos, written documentats,
+ original software applications and contributions in well-known external softwares such as Firefox, Scilab, LibreOffice, GEM/Puredata, to name a few~\cite{siteLM,wikiLM,vimeoLM}.
+Some of these efforts are available online~\cite{dissertacao}.
+It is obvious that all these contributions are a consequence of more that just \massa, but it is evident to the authors that \massa\ had a primary role in converging interests and attracting collaborators.}
+
+Future work should include application of these results in artificial intelligence
+for the generation of appealing artistic materials. Some psychoacoustical effects were detected,
+which needs validation and should be reported, specially with~\cite{quadrosSonoros}.\footnote{The sonic pictures where sent to a public mailing list and the fifth sonic picture was reported by some individuals to induce a state in which noises from the own tongue and teeth of the indidual echoed for some seconds~\cite{quadrosMetarec}.} Other foreseen advances
+are: a JavaScript version of the toolbox, better hypermedia deliverables of this framework, user guides
+for different goals (e.g. musical composition, psychophysical experiments, sound synthesis, education), creation
+of musical pieces and open experiments to be studied with EEG recordings, and further analytical specification of musical
+elements in the discrete-time representation of sound as feedback is received from the community.
 
 \begin{acks}
-
-The authors would like to thank ???
-
-The work is supported by the São Paulo Research Foundation under Grant ???.
-% 	\grantsponsor{GS501100001809}{National
-%   Natural Science Foundation of
-%   China}{http://dx.doi.org/10.13039/501100001809} under Grant
-% No.:~\grantnum{GS501100001809}{61273304\_a}
-% and~\grantnum[http://www.nnsf.cn/youngscientsts]{GS501100001809}{Young
-%   Scientsts' Support Program}.
+This work was supported by CNPq and FAPESP (Brazil).
 \end{acks}

### 2017-07-29 article/acm/foobody.tex
finisehd first review of everything

@@
 can be described as:
-
 \begin{itemize}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \subsection{Directional structures}\label{subsec:dir}
-
 The arcs can be decomposed in two convergent sequences: the first reaches the apex and the second returns from apex to start region.
@@
 \subsection{Cyclic structures}\label{estCic}
-
 The philosophical understanding that human thought is founded on the concept of similarities and differences (e.g. perceived in stimuli and objects), places symmetries
@@
 \subsection{Musical idiom?}
-
 In numerous studies, there are models and explorations of a `musical language',
@@
 \subsection{Musical usages}\label{subsec:usosmusicais3}
-
 First, the basic note was defined and characterized in quantitative terms (section~\ref{sec:notaDisc}).
@@
 \label{cap:conclusao}
-
 This concise presentation relates basic musical elements to digital sound.
@@
 In addition, this framework permitted the formation of interests groups, with topics such as music creativity and computer music. 
-In particular, the project \url{labMacambira.sf.net} groups Brazilian and foreign co-workers in order to offer relevant contributions 
+In particular, the project \url{labMacambira.sf.net} groups Brazilian and foreign co-workers in order to offer contributions
 in diverse areas like Digital Direct Democracy, georeferencing techniques, artistic and educational activities. 

### 2017-07-29 article/scripts/espectros_formas_de_onda.py
finished enhancing figures

@@ -1,7 +1,10 @@
 #-*- coding: utf-8 -*-
 # http://matplotlib.sourceforge.net/examples/api/legend_demo.html
 # 
-import pylab as p, numpy as n, scikits.audiolab as a
+import pylab as p, numpy as n
+
+p.figure(figsize=(10.,5.))
+p.subplots_adjust(left=0.09,bottom=0.15,right=0.96,top=0.96, hspace=0.4)
 
 T=30
 #T=1024
@@ -49,7 +52,7 @@ l=p.legend(loc="upper right")
 for t in l.get_texts():
     t.set_fontsize('x-large')
 
-p.yticks((0,20000),(0,"20k"), fontsize=16)
+p.yticks((0,20000),(0,"20k"), fontsize=17)
 #p.xticks((0,15000),(0,"15k"))
 # pop untill they leave f15 out of the game
 ii.pop()
@@ -66,16 +69,17 @@ ii.pop()
 ii.pop()
 ii.pop()
 ii.pop()
-ticks=[r"$f%i$" % (i,) for i in xrange(len(ii))]
-p.xticks([0] + ii + [16500], [0] + ticks + ["16,5k"], fontsize=16)
+ticks=[r"$f%i$" % (3*i,) for i in range(len(ii[::3]))]
+ticks_=[]
+for i in ticks:
+    ticks_+=[i]+[""]*2
+p.xticks([0] + ii , [0] + ticks_ , fontsize=18)
 
 p.xlim(0,16500)
 #p.xlim(0,T2*.56)
 p.ylim(-300,20000)
 p.ylabel(r'absolute value $\rightarrow$', fontsize=19)
-p.xlabel(r'spectrum component $\rightarrow$', fontsize=19)
+p.xlabel(r'spectral component $\rightarrow$', fontsize=19)
 
-
-
-
+p.savefig("../figures/waveSpectrum_.png")
 p.show()

### 2017-07-30 article/acm/body.tex
finished enhancing figures
iniciado documento para as listagens dos elementos no artigo

@@
+% \tableofcontents % enable to generate .toc file used in listings.tex
 \section{Introduction}\label{sec:level1}
@@
 and temporal organization of structures, or that music
-is also a cultural activity or product~\cite{}.
+is also a cultural activity or product.
 Sound is a physical phenomenon which corresponds to the longitudinal wave of mechanical pressure.
@@
 Although the terms 'audio' and 'sound' are often used without distinction
-and have many definitions that depend on the context and the author~\cite{audioSound},
-audio most often means a representation of sound that
-depicts the amplitude of the mechanical wave through time.
+and 'audio' has many definitions which depend on the context and the author,
+audio most often means a representation of sound by means
+of the wave amplitude through time.
 In this sense,
@@
 Digital audio protocols often implies quality loss (to yield smaller files, ease storage and transfer)
-and are called \emph{lossy}.
-This is the case e.g. of MP3 and Ogg Vorbis~\cite{loss}.
+and are called \emph{lossy}~\cite{loss}.
+This is the case e.g. of MP3 and Ogg Vorbis.
 Non-lossy representations of digital audio, called \emph{lossless} protocols or formats,
@@
     \centering
-        \includegraphics[width=.7\textwidth]{figures/pcm}
+        \includegraphics[width=.7\textwidth]{figures/pcm_}
 	\caption{Example of PCM audio:
-	a sound wave is represented by 25 samples where each sample has an amplitude specified with 4 bits.}
+	a sound wave is represented by 25 samples equally spaced in time where each sample has an amplitude specified with 4 bits.}
         \label{fig:PCM}
@@
 such as in Pygmy music, but western music theory assimilated them
-only in the last century~\cite{Wisnik}.}.
+only in the last century~\cite{Wisnick}.}.
 Notes are also convenient for another reason:
@@
 	\item Present a concise set of mathematical relations between musical basic elements and sequences of PCM audio samples.
-	\item Introduce a framework of sound and musical synthesis with control at sample level which entails potential uses in psychoacoustic experiments and synthesis with extreme precision (discussed in Section~\ref{highPrecision}).
+	\item Introduce a framework of sound and musical synthesis with control at sample level which entails potential uses in psychoacoustic experiments and synthesis with extreme precision (recap in Section~\ref{cap:conclusao}).
 	\item Provide a powerful theoretical framework which can be used to synthesize musical pieces and albums.
@@
 Section~\ref{sec:notesMusic} tackles the organization of musical notes in
-higher level musical structures~\cite{Wisnick,Webern,Lerdahl,Cook,Lacerda}.
-As these descriptions require topics such as psychoacoustics, cultural traditions,
-and formalisms, the text points to external complements as needed~\cite{Zamacois,Schoenberg,microsound}
+higher level musical structures~\cite{Wisnick,Webern,Lerdahl,Cook,Lacerda,Zamacois,Schoenberg,microsound}.
+As these descriptions require knowledge on topics such as psychoacoustics, cultural traditions,
+and mathematical formalisms, the text points to external complements as needed
 and presents methods, results and discussions altogether.
 Section~\ref{cap:conclusao} holds final considerations and further work.
+The Supporting Information document~\cite{SI} holds an informative listing of all the equations, figures, tables and sections in this document
+and the scripts in the \massa\ toolbox\footnote{The toolbox contains a collection of Python scripts which:
+\begin{itemize}
+    \item implement each of the equations;
+    \item render music and illustrate the concepts;
+    \item render each of the figures used in this article.
+\end{itemize}
+Additionally, the documentation of the toolbox consists of this article and the scripts themselves.}.
 
@@
 \subsection{Loudness}\label{subsec:volume}
-The sensation of loudness depends on reverberation and spectrum, among other characteristics described in section~\ref{sec:varInternas}.
+The sensation of loudness depends on reverberation and spectrum, among other characteristics described in Section~\ref{sec:varInternas}.
 One can achieve loudness variations through the power of the wave~\cite{Chowning}:
@@
 While the sinusoid is an isolated node in the spectrum, any other wave present a succession of harmonic components.
-The most basic waveforms are specified in equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square}, and are illustrated in Figure~\ref{fig:formasDeOnda}.
+The most basic waveforms are specified in Equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square}, and are illustrated in Figure~\ref{fig:formasDeOnda}.
 These artificial waveforms are traditionally used in music for synthesis and oscillatory control of variables. They are also useful outside musical contexts\cite{Openheim}.
@@
     \centering
-        \includegraphics[width=.7\textwidth]{figures/waveForms_}
+        \includegraphics[width=.7\textwidth]{figures/waveForms__}
     \caption{Basic musical waveforms: (a) the basic synthetic  waveforms; (b) real waveforms.}
@@
 
-Figure~\ref{fig:formasDeOnda} presents the waveforms described in equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} for $\lambda_f=100$ (i.e.
+Figure~\ref{fig:formasDeOnda} presents the waveforms described in Equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} for $\lambda_f=100$ (i.e.
 period of $100$ samples).
@@
     \centering
-        \includegraphics[width=.7\textwidth]{figures/waveSpectrum}
+        \includegraphics[width=.7\textwidth]{figures/waveSpectrum_}
     \caption{Spectrum of basic artificial musical waveforms.}
@@
 The harmonic spectrum is composed of frequencies $f_n$ that are multiples of the fundamental frequency: $f_n=(n+1)f_0$.
-As the human linear perception of pitch follows a geometric progression of frequencies, the harmonic spectrum has notes different from the fundamental frequency (see equation~\ref{eq:serieHarmonica}).
+As the human linear perception of pitch follows a geometric progression of frequencies, the harmonic spectrum has notes different from the fundamental frequency (see Equation~\ref{eq:serieHarmonica}).
 Additionally, the number of harmonics will be limited by the Nyquist frequency $f_s/2$.
@@
     \centering
-        \includegraphics[width=.7\textwidth]{figures/oboeNaturalSampledSpectrum}
+        \includegraphics[width=.7\textwidth]{figures/oboeNaturalSampledSpectrum_}
 	\caption{Spectrum of the sonic waves of a natural oboe note and obtained through a sampled period.
@@
 
-\subsection{Spectrum at sampled sound}
+\subsection{Spectrum of sampled sound}
 These sinusoidal components in the discretized sound have some particularities.
@@
 
-For a sound signal, samples $t_i$ are real and are given by the real part of equation~\ref{recomposicaoFourier}:
+For a sound signal, samples $t_i$ are real and are given by the real part of Equation~\ref{recomposicaoFourier}:
 
@@
 
-This discussion makes it clear the equivalence in~\ref{equivalenciasFreqs},~\ref{equivalenciasModulos} and~\ref{equivalenciasFases}:
+This discussion makes it clear the equivalence in Equations~\ref{equivalenciasFreqs},~\ref{equivalenciasModulos} and~\ref{equivalenciasFases}:
 
@@
 
-To discuss the general case for components combination in each sample $t_i$, one can gather relations in equation~\ref{moduloEfase} for the real signal reconstruction, relations of modules~\ref{equivalenciasModulos} and phase equivalences~\ref{equivalenciasFases}, the number of paired coefficients~\ref{coefsPareados}, and equivalence of paired frequencies~\ref{equivalenciasFreqs}:
+To discuss the general case for components combination in each sample $t_i$, one can gather relations in Equation~\ref{moduloEfase} for the real signal reconstruction, relations of modules (Equation~\ref{equivalenciasModulos}) and phase equivalences (Equation~\ref{equivalenciasFases}), the number of paired coefficients (Equation~\ref{coefsPareados}), and equivalence of paired frequencies (Equation~\ref{equivalenciasFreqs}):
 
@@
 
-Figure~\ref{fig:formas4} shows the harmonics for the basic waveforms of equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} for the case of 4 samples.
+Figure~\ref{fig:formas4} shows the harmonics for the basic waveforms of Equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} for the case of 4 samples.
 There is only 1 sinusoid for each waveform, with the exception of the sawtooth, which has even harmonics.
@@
 A unique period from the basic waveforms can be used to specify the note, where $\lambda_f=\frac{f_a}{f}$ is the number of samples at the period.
-Here, $L_i^{f,\, \delta_f} $ is the sequence that describes a period of the waveform $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ with duration $\delta_f=1/f$ (as given by equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square}) and $R_i^f$ is a sampled real waveform:
+Here, $L_i^{f,\, \delta_f} $ is the sequence that describes a period of the waveform $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ with duration $\delta_f=1/f$ (as given by Equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square}) and $R_i^f$ is a sampled real waveform:
 
@@
 On the other hand, the complete localization, including height and distance of sound source, is given by the Head Related Transfer Function (HRTF)~\cite{hrtf}.
-There are well known open databases of HRTF, such as CIPIC, and it is possible to apply those transfer functions in a sonic signal by convolution (see equation~\ref{eq:conv})~\cite{CIPIC}.
+There are well known open databases of HRTF, such as CIPIC, and it is possible to apply those transfer functions in a sonic signal by convolution (see Equation~\ref{eq:conv})~\cite{CIPIC}.
 Each human body has its filtering and there are techniques to generate HRTFs to be universally used~\cite{lazaSPA}.
@@
 There are reverberation models less related to each independent reflection, exploring valuable information to the auditory system.
-In fact, reverberation can be modeled with a set of two temporal and two spectral sections~\cite{JOSPhy}:
-
+In fact, reverberation can be modeled with a set of two temporal and two spectral regions~\cite{JOSPhy}:
 \begin{itemize}
@@
 These first reflections are particularly important to spatial sensation.
-The first incidence is the direct sound, described by ITD and IID of equations~\ref{eq:dti} and~\ref{eq:dii}.
+The first incidence is the direct sound, described by ITD and IID of Equations~\ref{eq:dti} and~\ref{eq:dii}.
 Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_a=44.1kHz$).
 It is experimentally verifiable that the number of reflections increases with the square of $\approx k.n^2$.
-A discussion about the use of convolutions and filtering to favor implementation of these phenomena is provided in subsection~\ref{subsec:mus2}, particularly in the paragraphs about reverberation.
-
-\subsection{Musical use}\label{subsec:basMus}
+A discussion about the use of convolutions and filtering to favor implementation of these phenomena is provided in Section~\ref{subsec:mus2}, particularly in the paragraphs about reverberation.
+
+\subsection{Musical usages}\label{subsec:basMus}
 Once the basic note is defined, it is didactically convenient to build musical structures with sequences based on these particles.
@@
 The mixed notes are generally separated by the ear according to the physical laws of resonance and by the nervous system~\cite{Roederer}.
-This process of mixing musical notes results in musical harmony whose intervals between frequencies and chords of simultaneous notes guide subjective and abstract aspects of music and its appreciation~\cite{Harmonia}, which is addressed in section~\ref{sec:notesMusic}.
+This process of mixing musical notes results in musical harmony whose intervals between frequencies and chords of simultaneous notes guide subjective and abstract aspects of music and its appreciation~\cite{Harmonia}, which is addressed in Section~\ref{sec:notesMusic}.
 
@@
 \section{Variation in the basic musical note}\label{sec:internalVar}\label{sec:varInternas}
-The basic digital music note defined in section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (volume) and timbre.
+The basic digital music note defined in Section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (volume) and timbre.
 This is a useful and paradigmatic model, but it does not exhaust all the aspects of a musical note.
@@
 
-\subsection{Lookup Table}\label{subsec:lookup}
+\subsection{Lookup table}\label{subsec:lookup}
 The \emph{Lookup Table} (or simply LUT) is an array for indexed operations which substitutes continuous and repetitive calculation.
 It is used to reduce computational complexity and for employing functions without calculating them directly, e.g.
-from data sampled from nature.
+from sampled data or hand picked values.
 In music its usage transcends these applications: it simplifies many operations and enables the use a single wave period to synthesize sounds in the whole audible spectrum, with any waveform.
@@
 
-\subsection{Incremental Variations of Frequency and Intensity}\label{subsec:vars}
+\subsection{Incremental variations of frequency and intensity}\label{subsec:vars}
 As stated by the Weber and Fechner~\cite{Weber-Fechner} law, the human perception has a logarithmic relation with the stimulus.
 That is to say, the exponential progression of a stimulus is perceived as linear.
-For didactic reasons, and given its use in AM and FM synthesis (subsection~\ref{subsec:tvaf}), linear variation is discussed first.
+For didactic reasons, and given its use in AM and FM synthesis (Section~\ref{subsec:tvaf}), linear variation is discussed first.
 
@@
          \includegraphics[width=.7\columnwidth]{figures/transicao_}
-     \caption{Intensity transitions for different values of $\alpha$ (see equations~\ref{seqAmp} and ~\ref{transAmp}).}
+     \caption{Intensity transitions for different values of $\alpha$ (see Equations~\ref{seqAmp} and~\ref{transAmp}).}
          \label{fig:transicao}
@@
 If $\alpha=1$, the amplitude variation follows the geometric progression that implies the linear variation of volume.
-Figure~\ref{fig:transicao} depicts transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by equation~\ref{eq:ampVol}.
+Figure~\ref{fig:transicao} depicts transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by Equation~\ref{eq:ampVol}.
 
 Special attention should be dedicated while considering $a=0$.
-In equation~\ref{seqAmp}, $a_0=0$ results in a division by zero and if $a_{\Lambda-1}=0$, there will be multiplication by zero.
+In Equation~\ref{seqAmp}, $a_0=0$ results in a division by zero and if $a_{\Lambda-1}=0$, there will be multiplication by zero.
 Both cases make the procedure useless, once a ratio of any number in relation to zero is not well defined for our purposes.
@@
 
-Here the conversion between decibels and amplitude is convenient, with equations~\ref{ampDec} and \ref{transAmp} specifying a transition of $V_{dB}$ decibels:
+Here the conversion between decibels and amplitude is convenient, with Equations~\ref{ampDec} and~\ref{transAmp} specifying a transition of $V_{dB}$ decibels:
 
@@
 
-The linear transitions will be used for AM and FM synthesis, while logarithmic transitions are proper tremolos and vibratos, as developed in subsection~\ref{subsec:tvaf}.
+The linear transitions will be used for AM and FM synthesis, while logarithmic transitions are proper tremolos and vibratos, as developed in Section~\ref{subsec:tvaf}.
 A non-oscillatory exploration of these variations is in the music piece \emph{Transit para meter}, whose code is online as part of the \massa\ toolbox\cite{MASSA}.
 
-\subsection{Application of Digital Filters}\label{subsec:filtros}
+\subsection{Application of digital filters}\label{subsec:filtros}
 This subsection is limited to a description of sequences processing, by convolution and difference equations, and immediate applications, as a thorough discussion of filtering is beyond the scope of this study\footnote{The implementation of filters encompasses an area of recognized complexity, with dedicated literature and software implementations~\cite{Openheim,smith}.}.
@@
         \includegraphics[width=\textwidth]{figures/iir__}
-    \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of equations~\ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} and \ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
+    \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of Equations~\ref{eq:passa-baixas},~\ref{eq:passa-altas},~\ref{eq:passa-banda} and~\ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
         \label{fig:iir}
@@
          \includegraphics[width=1.\textwidth]{figures/ruidos_}
-     \caption{Colors of noise generated by equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul} and \ref{eq:violeta}: spectrum and example waveforms.}
+     \caption{Colors of noise generated by Equations~\ref{eq:branco},~\ref{eq:rosa},~\ref{eq:marrom},~\ref{eq:azul} and~\ref{eq:violeta}: spectrum and example waveforms.}
          \label{fig:ruidos}
@@
   \item The brown noise received this name after Robert Brown, who described the Brownian movement\footnote{Although its origin is disparate with its color association, this noise became established with this specific name, in musical contexts.
-Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more vivid colors~\cite{Cook,guillaume}.} What characterizes brown noise is the decrease of $6dB$ per octave, with $\alpha_i$ in equation set~\ref{eq:rosa} being:
+Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more vivid colors~\cite{Cook,guillaume}.} What characterizes brown noise is the decrease of $6dB$ per octave, with $\alpha_i$ in Equations~\ref{eq:rosa} being:
 
@@
  \item In the blue noise there is a gain of $3dB$ per octave in a band limited by the minimum frequency $f_{\text{min}}$ and the maximum frequency $f_{\text{max}}$.
-Therefore, the corresponding equation is also based on the equations set~\ref{eq:rosa}:
+Therefore, the corresponding equation is also based on the Equations~\ref{eq:rosa}:
 
@@
 
-In these equations, the frequency variation introduced by $\{t_i'\}$ does not follow the geometric progression that yields linear pitch variation, but reflects equation~\ref{freqLinear}.
-The result of using equations~\ref{vbrF} for FM is described in~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
-In fact, the simple and attractive FM behavior is usually observed with linear oscillations, such as in~\ref{eq:fmEsp}, which yield less strident noisy sounds.
+In these equations, the frequency variation introduced by $\{t_i'\}$ does not follow the geometric progression that yields linear pitch variation, but reflects Equation~\ref{freqLinear}.
+The result of using Equations~\ref{vbrF} for FM is described in~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
+In fact, the simple and attractive FM behavior is usually observed with linear oscillations, such as in Equation~\ref{eq:fmEsp}, which yield less strident noisy sounds.
 
@@
 to note pitch, or a tremolo depth inversely proportional to
-pitch. Therefore, with equations \ref{vbrGamma}, \ref{vbrF} and \ref{trA}, it is possible to set:
+pitch. Therefore, with Equations~\ref{vbrGamma},~\ref{vbrF} and~\ref{trA}, it is possible to set:
 
@@
 
-\noindent with $f^{vbr}$ and $f^{tr}$ as $f'$ in the referenced equations. They can also be associated with vibrato and tremolo oscillation frequency of equation~\ref{vbrGamma}. 
+\noindent with $f^{vbr}$ and $f^{tr}$ as $f'$ in the referenced equations. They can also be associated with vibrato and tremolo oscillation frequency of Equation~\ref{vbrGamma}. 
  $\nu$ and $V_{dB}$ are the respective depth values of vibrato and
@@
 A musical pulse - such as specified by a BPM
-tempo - can be implied by an impulse at the start of each beat, with the purpose of establishing metric and rhythms: the convolution with an impulse shifts the sound to impulse position, as stated in section~\ref{subsec:conv}.
+tempo - can be implied by an impulse at the start of each beat, with the purpose of establishing metric and rhythms: the convolution with an impulse shifts the sound to impulse position, as stated in Section~\ref{subsec:conv}.
 For example, two impulses equally spaced build a binary division in the
@@
 \subsubsection{Moving source and receptor, Doppler effect}
-According to the description in Subsection~\ref{subsec:spac}, when an audio source or receptor is moving, its characteristics are ideally updated at each sample of the digital signal (if fast computational rendering is not at stake).
+According to the description in Section~\ref{subsec:spac}, when an audio source or receptor is moving, its characteristics are ideally updated at each sample of the digital signal (if fast computational rendering is not at stake).
 Velocity components should be found for each ear.
@@
 where $y_i=y_0+(v_s-v_r)\frac{i}{f_s}$ with $v_s$ and $v_r$ both
-horizontal, having null non-$y$ components. Amplitude changes with the distance and with the potency factor mentioned above (see subsection~\ref{subsec:volume} for potency to amplitude conversion).
+horizontal, having null non-$y$ components. Amplitude changes with the distance and with the potency factor mentioned above (see Section~\ref{subsec:volume} for potency to amplitude conversion).
 
@@
 
-\subsubsection{Filters and noises (subsections~\ref{subsec:ruidos} and~\ref{subsec:filtros})}
+\subsubsection{Filters and noises (Sections~\ref{subsec:ruidos} and~\ref{subsec:filtros})}
 With the use of filters, the possibilities are even wider.
@@
 Based on what happens in a choir of singers, the sound is synthesized using small and potentially arbitrary modifications of parameters like center frequency, presence (or absence) of vibrato or tremolo and its characteristics, equalization, volume, etc.
-As a final result, those versions of the original sound are mixed together (see equation~\ref{eq:mixagem}).
+As a final result, those versions of the original sound are mixed together (see Equation~\ref{eq:mixagem}).
 The music piece \emph{Child chorus} implements a chorus in many ways with different sounds.
@@
 \subsubsection{Reverberation}\label{subsubsec:reverb}
-Using the same terms of subsection~\ref{subsec:spac}, the
+Using the same terms of Section~\ref{subsec:spac}, the
 late reverberation can be achieved by a convolution with a section of pink, brown or black noise, with exponential decay of amplitude along time. Delay lines can be added as a prefix to the noise
@@
 $i$-th sample. The amplitude decreases exponentially. Following
-subsection~\ref{subsec:spac}, reverberation $R_i^1$ of the first period can be described as:
+Section~\ref{subsec:spac}, reverberation $R_i^1$ of the first period can be described as:
 
@@
 
-A sound with simulated reverberation can be achieved by a simple convolution of $R_i$ (called reverberation impulse response) with the sound sequence $T_i$, as described in subsection~\ref{subsec:filtros}.
+A sound with simulated reverberation can be achieved by a simple convolution of $R_i$ (called reverberation impulse response) with the sound sequence $T_i$, as described in Section~\ref{subsec:filtros}.
 Reverberation is well known for causing great interest in listeners and to provide sonorities that are more enjoyable.
@@
 
-\subsubsection{ADSR Envelopes}
+\subsubsection{ADSR envelopes}
 The variation of volume along the duration of sound is crucial to our timbre perception.
@@
 The elements of $S_j$ can be overlapped by mixing them together, as in
-equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, for building intervals and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of events
-in $S_j$, as in equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao}, builds melodic sequences and rhythms, which are associated with the `horizontal thought'.
+Equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, for building intervals and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of events
+in $S_j$, as in Equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao}, builds melodic sequences and rhythms, which are associated with the `horizontal thought'.
 The fundamental frequency $f$ and the starting moment
@@
 		Each of these intervals are related to one step less the their names suggest: a third is an interval with two steps.
-		As can be noticed in Table~\ref{eq:interfvalos}, one step is not one semitone.
+		As can be noticed in Table~\ref{eq:intervalos}, one step is not one semitone.
 		A step, in this sense, is yield by two consecutive notes in a musical scale.
@@
 	\item Tritone is a dissonance in Western music because
-		it characterizes the ``dominant'' chord in the tonal system (see subsection~\ref{subsec:harmonia}) and represents (or yields) instability.
+		it characterizes the ``dominant'' chord in the tonal system (see Section~\ref{subsec:harmonia}) and represents (or yields) instability.
 Some cultures considers the interval a consonance, using it in melodies and chants as a stable interval.
@@
         \item It is possible to build chords with notes different from triads, particularly if they obey a recurrent logic or musical concatenation that justifies these different notes.
-        \item Chords built by successive intervals different from thirds – like fourths and seconds -- are recurrent in compositions of advanced tonalism or experimental character.
+        \item Chords built by successive intervals different from thirds -- such as fourths and seconds -- are recurrent in compositions of advanced tonalism or experimental character.
         \item The repetition of chord successions (or of characteristics they hold) fixes a trajectory and makes it possible to 
@@
 Omission of basic tonal structures is the key to achieving modal and atonal harmonies. In the absence of minimal tonal organization,
-harmony is (usually) considered modal if the notes match with some diatonic scale (see equations~\ref{eq:escalas}) or if there is only a small number of notes.
+harmony is (usually) considered modal if the notes match with some diatonic scale (see Equations~\ref{eq:escalas}) or if there is only a small number of notes.
 If basic tonal progressions are absent and notes do not match any diatonic scale and are enough diverse and dissonant (between themselves) to avoid reduction of the notes by polarization\footnote{By polarization we mean having some notes that are way more important than others and to which the other notes are ornaments or subordinates.}, the harmony is atonal.
@@
      \item Use durations and pauses with liberty, respecting notes order.
-     \item Vary note sequences by temporal magnification and translation or pitch transposition and by pitch sequence inversion, retrograde and retrograde inversion.
-See subsections~\ref{subsec:motivos}
+     \item Vary note sequences by temporal magnification and translation or pitch transposition and by pitch sequence inversion, retrograde and retrograde inversion. See Sections~\ref{subsec:motivos}
      and~\ref{subsec:usosmusicais3} for what these terms mean.
@@
 In tonal or modal music, chords -- like the ones listed in
-equations~\ref{triades}-- built on the root note of each degree of a scale  -- such as listed in equations~\ref{eq:escalas} --  form the pillars of harmony.
+Equations~\ref{triades}-- built on the root note of each degree of a scale  -- such as listed in Equations~\ref{eq:escalas} --  form the pillars of harmony.
 Tonal (and modal) harmony aims at understanding the incidence of chord progressions and chaining rules.
@@
 In the traditional tonal music, a scale has its tonic (first
-degree) on any note, and can be major (with the same notes of Ionian mode) or minor (same notes as Eolian mode, 'natural minor', which has both harmonic and melodic versions, as in equations~\ref{eq:escalasMenores}).
+degree) on any note, and can be major (with the same notes of Ionian mode) or minor (same notes as Eolian mode, 'natural minor', which has both harmonic and melodic versions, as in Equations~\ref{eq:escalasMenores}).
 The
@@
 a fact that leads to the circle of fifths~\cite{Harmonia,Salzer,Koellheuteur,Harmony}.
-Other inventive ways to modulate are possible, to point but one common example, the minor thirds tetrad ($E_i^{tm}$ in equations~\ref{escSim}) can be sustained to bridge to other tonalities, with the ease of its both tritones.
+Other inventive ways to modulate are possible, to point but one common example, the minor thirds tetrad ($E_i^{tm}$ in Equations~\ref{escSim}) can be sustained to bridge to other tonalities, with the ease of its both tritones.
 The music piece \emph{Acorde cedo} explores these chord relations, and is implemented online as part of \massa\ toolbox~\cite{MASSA}.
@@
 Counterpoint is the conduction of simultaneous melodic lines, or ``voices''.
-The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues~\cite{count1, count2}.
+The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues~\cite{Fux,SchoenbergContra}.
 It is possible to
@@
 It is possible to summarize the durations heard as rhythm or pitch, and its transition,
-as in table~\ref{tab:duracoes}~\cite{Alfaix, microsound}.
+as in Table~\ref{tab:duracoes}~\cite{Alfaix, microsound}.
 
@@
 
-Each level $j$ has some sections $i$. When $i$ has three different
+Each level $j$ has some parts $i$. When $i$ has three different
 values (or multiple of three) there is a perfect (i.e. ternary) relation. When $i$ has only
@@
 of $H$ musical events $s_j$, each event with its $\Lambda_j$ samples $t_i^j$
-(refer to the beginning of this section~\ref{notasMusica} if needed). Bellow is a list of basic techniques
+(refer to the beginning of this Section~\ref{notasMusica} if needed). Bellow is a list of basic techniques
 for variation.
@@
         $\Xi$ is not a measure in semitones: $\Xi=1$ is not a semitone, it is a note with an audible frequency as rhythm, with
-		less than 9 occurrences each second (see table~\ref{tab:duracoes}).}
-
-        \item Interval inversion is either: 1) the inversion of note pitch order, within octave equivalence, such as described in~\ref{subsec:intervalos}, and $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{\varepsilon.
+		less than 9 occurrences each second (see Table~\ref{tab:duracoes}).}
+
+        \item Interval inversion is either: 1) the inversion of note pitch order, within octave equivalence, such as described in Section~\ref{subsec:intervalos}, and $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{\varepsilon.
 f_0}\right\}$ with selective $s_j=2,\;1/2,\;\text{or}\;1$; or 2) the inversion of interval orientation.
@@
         $\tilde{n}<0$ (elements delay) with the anti-clockwise rotation.
-        Additional information about rotations is given in subsection~\ref{estCic}.
+        Additional information about rotations is given in Section~\ref{estCic}.
 
@@
 In the auralization of a permutation, it is not necessary to permute elements of $S_i$,
-but only some characteristic. Thus, if $p$ is a permutation and $S_i$ is a sequence of basic notes as in the end of section~\ref{notaBasica}, the
+but only some characteristic. Thus, if $p$ is a permutation and $S_i$ is a sequence of basic notes as in the end of Section~\ref{notaBasica}, the
 sequence $S_i'=p^f(S_i)=\left\{s_i^{p(f)}\right\}$ consists of the same
@@
 relations established between elements along a music piece. In these matters,
-dichotomies are prominent, as explained in subsection~\ref{subsec:motivos}:
+dichotomies are prominent, as explained in Section~\ref{subsec:motivos}:
 repetition and variation, relaxation and tension, stability and instability, consonance and dissonance, etc.
@@
 \subsection{Musical usages}\label{subsec:usosmusicais3}
-First, the basic note was defined and characterized in quantitative terms in section~\ref{sec:notaDisc}.
+First, the basic note was defined and characterized in quantitative terms in Section~\ref{sec:notaDisc}.
 Next, the internal note
 composition was addressed within both internal transitions and elementary sonic treatment
-(section~\ref{varInternas}). Finally, this section aims at organizing these notes in music. The numerous resources and consequent infinitude
+(Section~\ref{sec:varInternas}). Finally, this section aims at organizing these notes in music. The numerous resources and consequent infinitude
 of praxis possibilities is typical and highly relevant for artistic contexts~\cite{Harmonia,Webern}.
@@
 
-As in subsection~\ref{subsec:mus2}, relations between
+As in Section~\ref{subsec:mus2}, relations between
 parameters are powerful ways to achieve musical pieces.
@@
 notes can vary during the music, revealing relationship with piece
-duration. Harmonies can be made from triads (equations~\ref{triades}) with duplicated
+duration. Harmonies can be made from triads (Equations~\ref{triades}) with duplicated
 notes at each octave and more numerous duplication when the depth and frequency of
-vibratos are lower (equations ~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}).
+vibratos are lower (Equations~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}).
 Incontestably, the possibilities are very wide, which is evidenced in the wide range of musical pieces and styles.
 
-The symmetries at octave divisions (equation~\ref{escSim}) and the
-symmetries presented as permutations (table~\ref{tab:change} and
-equations~\ref{eq:groups}) can be used together. In the music piece \emph{3 trios}
+The symmetries at octave divisions (Equation~\ref{escSim}) and the
+symmetries presented as permutations (Table~\ref{tab:change} and
+Equations~\ref{eq:groups}) can be used together. In the music piece \emph{3 trios}
 this association is done in a systematic way in order to achieve a specific style. This is an instrumental piece, not included as a source
@@
 We aimed at achieving a concise presentation of the subject because it involves many knowledge fields, and therefore can very easily blast into thousands of pages.
-Some readers might benefit from the text alone, but the \emph{Scripts} in the \massa toolbox, where all the relations presented here are directly and simply implemented as software (in Python), are very helpful for one to achieve elaborated implementations and deeper understandings.
+Some readers might benefit from the text alone, but the \emph{Scripts} in the \massa\ toolbox, where all the relations presented here are directly and simply implemented as software (in Python), are very helpful for one to achieve elaborated implementations and deeper understandings.
 The scripts include routines that render musical pieces that illustrate the concepts.
@@
 for many scientific, artistic and didactic purposes.
-A fact evidenced by a number of uses this framework had in courses, in software implementations and in making music~\ref{videos,vivace,massaRepo,dissertacao}.
+A fact evidenced by a number of uses this framework had in courses, in software implementations and in making music~\cite{vimeoLM,vivace,repoDissertacao,dissertacao}.
 
 This work systematically investigates parameterization issues (like for the tremolo, ADSR, etc.) in a sample-by-sample synthesis, which entails high/extreme fidelity and if of significant artistic convenience.
-Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, as far as the authors know, such as testified in the literature review in~\cite{dissertacao}, where the books, articles and open software that deal with the issue are related to this framework.
+Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, as far as the authors know, such as testified in the literature review~\cite{dissertacao}, where the books, articles and open software are related to this framework.
 

### 2017-07-30 article/acm/listings.tex
iniciado documento para as listagens dos elementos no artigo

\documentclass{scrreprt}
% \documentclass[12pt]{article}
\usepackage{graphicx}

\usepackage{xr}
\externaldocument{sample-acmsmall}

\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
        colorlinks,
        linkcolor={red!50!black},
        citecolor={blue!50!black},
        urlcolor={blue!80!black}
}
%
\usepackage{graphicx} %allows for including images
\usepackage{biblatex}

\newcommand{\massa}{{\large \textsc{mass}}}

\makeatletter
\newcommand{\usetocfromothersource}[1]{%
  \begingroup
  \IfFileExists{#1}{%
    \@input{#1}% 
    \@nobreakfalse
  }{}%
  \endgroup
}
\makeatother

\def\thesection{SI-\arabic{section}}

\begin{document}

\title{Listings of equations, figures, tables and sections of the
article '{\it \bf Musical elements in the discrete-time representation of
sound}' and of the scripts in the MASS toolbox}

\author{\\\\\\ Renato Fabbri (USP),\\
Vilson Vieira da Silva Junior (The Grid),\\
Ant\^onio Carlos Silvano Pessotti (Unimep),\\
D\'ebora Cristina Corr\^ea,\\
Osvaldo N. Oliveira Jr. (USP)
}

\maketitle

\begin{abstract}
The article is the main document of the \massa\ toolbox.
Being it of considerable length and complexity,
this document contains listings of its elements to facilitate
its navigation, apprehension and general usage.
\end{abstract}

\tableofcontents

\clearpage
\section{Table of Contents of the article}
\makeatletter
\let\Hy@linktoc\Hy@linktoc@none
\makeatother
\usetocfromothersource{sample-acmsmall.toc}

\clearpage
\section{Equations}

\clearpage
\section{Figures}

\clearpage
\section{Tables}

\clearpage
\section{Scripts}
\subsection{For all equations in each chapter}
\clearpage
\subsection{To render musical pieces}
\clearpage
\subsection{To render the figures used in the article}

\clearpage
\section{Other documents}
Latex files. PDF files.
MSc Dissertation.

\section{Final considerations}

\end{document}

### 2017-07-30 article/acm/sample-acmsmall.tex
finalizada revisao da intro
iniciado documento para as listagens dos elementos no artigo

@@
 % Metadata Information
-\acmJournal{TWEB}
+\acmJournal{CSUR}
 \acmVolume{0}
@@
 \acmMonth{0}
-\copyrightyear{2009}
+\copyrightyear{2017}
 %\acmArticleSeq{9}
@@
 %\setcopyright{acmcopyright}
-\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
+% \setcopyright{acmlicensed}
+\setcopyright{rightsretained}
 %\setcopyright{usgov}
@@
 %\setcopyright{cagovmixed}
+\acmPrice{0.00}
 
@@
 
-\newcommand{\massa}{{\large \textsc{massa}}}
+\newcommand{\massa}{{\large \textsc{mass}}}
 \newcommand{\mass}{{\large \textsc{mass}}}
@@
 % Title portion. Note the short title for running heads 
-\title[Psychophysics of musical elements in digital audio]{Psychophysics of musical elements in the discrete-time representation of sound}  
+\title[Music in digital audio]{Musical elements in the discrete-time representation of sound}  
 \author{Renato Fabbri}
@@
 and each of these relations are implemented in scripts within a software toolbox,
-referred to as \massa\ (Music and Audio in Sequences and Samples).
+referred to as \massa\ (Music and Audio in Sample Sequences).
 The toolbox also includes this article and every script necessary to render the images
@@
 It is possible to synthesize whole albums through collage of the scripts and parameterization specified by the user.
-With the paradigm of open source implementation, \massa\ toolbox can be promptly scrutinized, expanded in co-authorship processes and used freely by musicians, engineers and other interested parties.
+With the paradigm of open source implementation, the toolbox can be promptly scrutinized, expanded in co-authorship processes and used freely by musicians, engineers and other interested parties.
 In fact, \massa\ has already been employed by external users for diverse purposes which include music production,
@@
 
-\thanks{This work is supported by FAPESP (grant number etc ???).}
+\thanks{This work is supported by FAPESP and CNPq.}
 

### 2017-08-01 article/acm/body.tex
finished listing of equations

@@
 
-\section{Characterization of the discrete-time musical note} \label{sec:discNote}\label{sec:notaDisc}
+\section{Characterization of the musical note in dicrete-time audio} \label{sec:discNote}\label{sec:notaDisc}
 In diverse artistic and theoretical contexts, music is conceived as comprising fundamental units referred to as notes, ``atoms'' that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
@@
 
-\noindent where $\zeta$ is the distance between ears $\zeta$, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in ITD:
+\noindent where $\zeta$ is the distance between ears $\zeta$, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in Interaural Time Difference:
 
@@
 
-\section{Variation in the basic musical note}\label{sec:internalVar}\label{sec:varInternas}
+\section{Variation in the basic note}\label{sec:internalVar}\label{sec:varInternas}
 The basic digital music note defined in Section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (volume) and timbre.
@@
 
-  \item The brown noise received this name after Robert Brown, who described the Brownian movement\footnote{Although its origin is disparate with its color association, this noise became established with this specific name, in musical contexts.
+  \item The brown noise received this name after Robert Brown, who described the Brownian movement\footnote{Although its origin is disparate with its color association, this noise became established with this specific name in musical contexts.
 Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more vivid colors~\cite{Cook,guillaume}.} What characterizes brown noise is the decrease of $6dB$ per octave, with $\alpha_i$ in Equations~\ref{eq:rosa} being:
@@
 
-\newcommand{\OneColEqu}[1]{%
-\end{multicols}%
-\begin{twocolequfloat}%
-\begin{equation}
+\begin{equation}\label{eq:specAM}
+\begin{split}
 \{t_i'\}_0^{\Lambda-1} =\{(1+a_i) .
@@
                         =  \left\{P.\sin \left( f.2\pi\frac{i}{f_s -1}  \right ) +  \frac{P.M}{2} \left [ \sin \left( (f-f').2\pi\frac{i}{f_s -1}  \right )  + \sin \left( (f+f').2\pi\frac{i}{f_s -1}  \right ) \right ] \right \}_0^{\Lambda-1}
-\end{equation}
-\end{twocolequfloat}%
-\begin{multicols}{2}%
-}
+\end{split}
+\end{equation}
+
+% \newcommand{\OneColEqu}[1]{%
+% \end{multicols}%
+% \begin{twocolequfloat}%
+% \begin{equation}
+% \{t_i'\}_0^{\Lambda-1} =\{(1+a_i) .
+t_i\}_0^{\Lambda-1} = \left \{ \left [ 1+M.\sin \left ( f'.2\pi\frac{i}{f_s -1} \right ) \right] .P .\sin \left ( f.2\pi\frac{i}{f_s -1} \right ) \right \}_0^{\Lambda-1} = \\
+%                         =  \left\{P.\sin \left( f.2\pi\frac{i}{f_s -1}  \right ) +  \frac{P.M}{2} \left [ \sin \left( (f-f').2\pi\frac{i}{f_s -1}  \right )  + \sin \left( (f+f').2\pi\frac{i}{f_s -1}  \right ) \right ] \right \}_0^{\Lambda-1}
+% \end{equation}
+% \end{twocolequfloat}%
+% \begin{multicols}{2}%
+% }
 
@@
 
-\begin{equation}
+\begin{equation}\label{eq:rhythmicUnit}
 P^{ \{ j_k \} }_{ \{ i_{k} \}}

### 2017-08-01 article/acm/listings.tex
finished listing of equations

@@
 
+\begin{table*}[htp!]
+\centering
+\caption{Equation numbers and their descriptions.
+All these equations are implemented in file \texttt{src/sections/2.py}.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \ref{eq:dur} & relation between number of samples and duration \\
+ \ref{eq:potencia} & power of the wave \\
+ \ref{decibels} & decibels by difference by means of the power of each wave \\
+ \ref{eq:ampVol} & double amplitude implies $\approx 6dB$ \\
+ \ref{eq:potVol} & double power implies $\approx 3dB$ \\
+ \ref{eq:dobraVol} & double volume ($10dB$) implies a factor of $\approx 3.16$ for amplitude \\
+ \ref{ampDec} & direct relation between variations in amplitude and decibels \\
+ \ref{periodicidade} & equivalences in a periodic sound with respect to wavelength, frequency and sample rate \\
+ \ref{sinusoid} & sample amplitudes in a sinusoid \\
+ \ref{sawTooth} & sample amplitudes in a sawtooth wave \\
+ \ref{triangular} & sample amplitudes in a triangular wave \\
+ \ref{square} & sample amplitude in a square wave \\
+ \ref{sampleandoFormaDeOnda} & samples in a sound derived from a sampled wave period \\
+ \ref{recomposicaoFourier} & reconstruction of samples from the Fourier components \\
+ \ref{moduloEfase} & reconstruction of real samples (e.g. for audio) from Fourier components \\
+ \ref{coefsPareados} & number of pairs of Fourier coefficients which are related to the same frequency \\
+ \ref{equivalenciasFreqs} & indexes of equivalent frequencies and coefficients for real signals \\
+ \ref{equivalenciasModulos} & equal modules between samples of real signals \\
+ \ref{equivalenciasFases} & equivalence in phases between samples of real signals \\
+ \ref{eq:reconsCompleta} & complete reconstruction of the signal using Fourier components and previous equations \\
+ \ref{eq:notaBasica} & sample sequence related to a basic note \\
+ \ref{periodoUnico} & sample sequence of a period of an arbitrary waveform \\
+ \ref{eq:notaBasicaTimbre} & note samples derived from a sampled waveform \\
+ \ref{eq:distOuvidos} & distance of a (sound) source to each ear given the distance between the ears and an (x,y) position of the source \\
+ \ref{eq:dti} & Interaural Time Difference (ITD), the time difference of a sound reaching each ear \\
+ \ref{eq:dii} & Interaural Intensity Difference (IID), the intensity difference (in decibels) of a sound reaching each ear \\
+ \ref{eq:locImpl} & ITD and IID in terms of sample delays and their amplitudes \\
+ \ref{eq:angulo} & azimuthal angle of a (x, y) source \\
+ \ref{eq:mixagem} & samples that result from mixing sounds \\
+ \ref{eq:concatenacao} & samples that result from concatenating sounds \\
+\end{tabular}
+\end{table*}
+
+\begin{table*}[htp!]
+\centering
+\caption{Equation numbers and their descriptions.
+All these equations are implemented in file \texttt{src/sections/3.py}.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \ref{eq:lut} & sample sequence generated by means of a lookup table (LUT) \\
+ \ref{freqLinear} & frequency at each sample in a linear transition of frequency \\
+ \ref{indiceLinear} & indices of a LUT in a linear transition of frequency \\
+ \ref{serieAmostralLin} & sample sequence obtained through a LUT in a linear transition of frequency \\
+ \ref{freqExponencial} & frequency at each sample in an exponential transition of frequency (linear pitch) \\
+ \ref{indiceExponencial} & indices of a LUT in an exponential transition of frequency (linear pitch) \\
+ \ref{serieAmostralLog} & sample sequence obtained through a LUT in an exponential transition of frequency \\
+ \ref{seqAmp} & amplitude factors at each sample in an exponential transition of amplitude ($\approx$ linear volume) \\
+ \ref{transAmp} & sample sequence with an exponential transition of amplitude ($\approx$ linear volume) \\
+ \ref{seqAmpLin} & amplitude factors at each sample in a linear transition of amplitude \\
+ \ref{seqAmpDB} & sample sequence in an exponential transition of amplitude ($\approx$ linear volume) with difference given in decibels \\
+ \ref{eq:conv} & sample sequence obtained through the convolution of two other sequences (e.g. for applying FIR filters) \\
+ \ref{eq:diferencas} & difference equation (e.g. for applying IIR filters) \\
+ \ref{eq:passa-baixas} & IIR coefficients for a simple, useful and well-behaved low-pass filter \\
+ \ref{eq:passa-altas} & IIR coefficients for a simple, useful and well-behaved high-pass filter \\
+ \ref{eq:varAux} & auxiliary variables for the following band-pass and band-reject filters \\
+ \ref{eq:passa-banda} & IIR coefficients for a simple, useful and well-behaved band-pass filter \\
+ \ref{eq:rejeita-banda} & IIR coefficients for a simple, useful and well-behaved band-reject filter \\
+ \ref{eq:branco} & Fourier coefficients of a white noise \\
+ \ref{eq:rosa} & Fourier coefficients of a pink noise \\
+ \ref{eq:marrom} & Fourier coefficients of a brown noise \\
+ \ref{eq:azul} & Fourier coefficients of a blue noise \\
+ \ref{eq:violeta} & Fourier coefficients of a violet noise \\
+ \ref{eq:preto} & Fourier coefficients of a black noise \\
+ \ref{vbrGamma} & indices for a vibrato given its frequency and using a LUT \\
+ \ref{vbrAux} & samples for applying a vibrato \\
+ \ref{vbrF} & frequency at each sample of a sound with vibrato \\
+ \ref{vbrGamma2} & indices for LUT in a sound with vibrato \\
+ \ref{vbrT} & sample sequence of a sound with vibrato \\
+ \ref{trA} & amplitude at each sample for a tremolo \\
+ \ref{trT} & sample sequence of a sound with tremolo \\
+ \ref{eq:fmEsp} & components in FM synthesis when both modulator and carrier are sines \\
+ \ref{eq:Bessel} & the Bessel function \\
+ \ref{eq:specAM} & components in AM synthesis when both modulator and carrier are sines \\
+ \ref{fmGammaAux} & indices for LUT in modulator of an FM synthesis \\
+ \ref{fmAux} & sample sequence of a modulator in an FM synthesis \\
+ \ref{fmF} & frequeny at each sample of a sound derived from FM synthesis \\
+ \ref{fmGamma} & indices for the final signal in FM synthesis using LUT \\
+ \ref{fmT} & sample sequence of a sound generated through FM and using LUT \\
+ \ref{amA} & amplitude at each sample in a sound generated though AM \\
+ \ref{amT} & sample sequence of a sound generated through AM and using LUT \\
+\end{tabular}
+\end{table*}
+
+\begin{table*}[htp!]
+\centering
+\caption{Equation numbers and their descriptions.
+All these equations are implemented in file \texttt{src/sections/3.py}.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \ref{eq:vinculos} & an example of bonds between musical characteristics \\
+ \ref{eq:fDoppler} & relation between frequencies and speed in the Doppler effect \\
+ \ref{eq:aDoppler} & relation between position, speed and amplitude in the Doppler effect \\
+ \ref{eq:ffDoppler} & relation between position, speed and amplitude in the Doppler effect \\
+ \ref{eq:p1rev} & samples of a FIR filter for the first period of a reverberation \\
+ \ref{eq:p2rev} & samples of a FIR filter for the second period of a reverberation \\
+ \ref{eq:rev} & samples of the FIR filter for a reverberation (considering both first and second periods) \\
+ \ref{eq:adsr} & amplitude factors for each sample in an ADSR envelope \\
+ \ref{eq:adsrApl} & sample sequence of a sound with an ADSR envelope \\
+\end{tabular}
+\end{table*}
+
+\begin{table*}[htp!]
+\centering
+\caption{Equation numbers and their descriptions.
+All these equations are implemented in file \texttt{src/sections/4.py}.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \ref{escSim} & perfectly symmetric scales in each octave with the twelve semitones \\
+ \ref{eq:escalas} & diatonic scales \\
+ \ref{eq:relacaoDia} & the succession of tones and semitones of a diatonic scale \\
+ \ref{eq:escalasMenores} & sequences of semitones for the three types minor scales \\
+ \ref{eq:serieHarmonica} & harmonic series in terms of semitones \\
+ \ref{triades} & triads (chords constituted by thirds) \\
+ \ref{eq:rhythmicUnit} & a convention to specify a unit of rhythmic division or agglomeration \\
+ \ref{eq:groups} & definition of algebraic groups \\
+\end{tabular}
+\end{table*}
+
+\subsection{In Section 4 - Notes in music}
+
+\subsection{In Section 3 - Variations of the basic note}
+\subsection{In Section 4 - Notes in music}
+
 \clearpage

### 2017-08-05 article/acm/listings.tex
finished the listings

@@
 
-\subsection{In Section 4 - Notes in music}
+\clearpage
+\section{Figures}\label{sec:lfigures}
 
-\subsection{In Section 3 - Variations of the basic note}
-\subsection{In Section 4 - Notes in music}
-
-\clearpage
-\section{Figures}
+\begin{table*}[htp!]
+\centering
+\caption{Figure numbers and their descriptions.
+All these equations are implemented in files \texttt{src/aux/*}.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \ref{fig:PCM} & PCM PCM audio (discrete and digital) samples \\
+ \ref{fig:formasDeOnda} & synthetic and sampled waveforms \\
+ \ref{fig:espectroDeOndas} & spectrum of basic waveforms \\
+ \ref{fig:espectroOboe} & spectrum of a real note and of one derived from one sampled period \\
+ \ref{fig:amostras2} & sinusoid represented by two samples \\
+ \ref{fig:amostras3} & sinusoid represented by three samples \\
+ \ref{fig:amostras4} & sinusoids in 4 samples \\
+ \ref{fig:formas4} & basic waveforms within 4 samples \\
+ \ref{fig:amostras6} & sinusoids in 6 samples \\
+ \ref{fig:formas6} & basic waveforms withing 6 samples \\
+ \ref{fig:spac} & ITD and IID (spatial localization cues)\\
+ \ref{fig:mixagem} & mixing PCM audio \\
+ \ref{fig:concatenacao} & concatenation/juxtaposition of PCM audio \\
+ \ref{fig:lut} & Lookup table \\
+ \ref{fig:transicao} & transitions of intensity \\
+ \ref{fig:conv} & convolution \\
+ \ref{fig:delays} & convolution yielding time shifting, multiple time delays, sound amalgam \\
+ \ref{fig:iir} & frequency response of useful IIR filters \\
+ \ref{fig:ruidos} & spectrum and waveform of noises \\
+ \ref{fig:vibrato} & spectrogram of a vibrato \\
+ \ref{fig:tremolo} & waveform of a tremolo \\
+ \ref{fig:adsr} & ADSR envelope \\
+ \ref{fig:movContraponto} & counterpoint movements \\
+ \ref{fig:pulsoSubAgl} & musical metric in terms of divisions of temporal units \\
+ \ref{fig:climax} & distinctions of musical climax by localization \\
+\end{tabular}
+\end{table*}
 
@@
 
+\begin{table*}[htp!]
+\centering
+\caption{Table numbers and their descriptions.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \ref{eq:intervalos} & musical intervals, their notations and qualities \\
+ \ref{tab:harmonia} & tonal harmonic functions in the major scale \\
+ \ref{tab:duracoes} & duration scales yielding perception of pitch and rhythm \\
+ \ref{tab:change} & music by permutation of units (change ringing) \\
+\end{tabular}
+\end{table*}
+
 \clearpage
 \section{Scripts}
-\subsection{For all equations in each chapter}
+\subsection{For all equations and relations in each section.}
+
+\begin{table*}[htp!]
+\centering
+\caption{Script files and their descriptions.}
+\begin{tabular}{ c | p{12cm} }
+   Filename & Description \\\hline
+  \texttt{src/sections/2.py} & implementation of all the equations for the basic note in PCM audio in Section~\ref{sec:discNote} \\
+  \texttt{src/sections/3.py} & implementation of all the equations for variations within a note described in Section~\ref{sec:varInternas} \\
+  \texttt{src/sections/4.py} & implementation of all the techniques for assembling notes into music described in Section~\ref{sec:notesMusic} \\
+\end{tabular}
+\end{table*}
+
 \clearpage
 \subsection{To render musical pieces}
+
+\begin{table*}[htp!]
+\centering
+\caption{Piece names, script files and the concepts they exemplify from Section~\ref{sec:discNote}. All files are in the directory \texttt{src/pieces2/}.}
+\begin{tabular}{ p{5cm} | p{3.5cm} | p{6.2cm} }
+   Name & Filename & Description \\\hline
+ Reduced-fi & \texttt{reduced-fi.py} & concatenation of simple notes \\
+ Sonic pictures & \texttt{quadrosSonoros.py} & mixing of simple notes  \\
+\end{tabular}
+\end{table*}
+
+\begin{table*}[htp!]
+\centering
+\caption{Piece names, script files and the concepts they exemplify from Section~\ref{sec:internalVar}. All files are in the directory \texttt{src/pieces3/}.}
+\begin{tabular}{ p{5cm} | p{3.5cm} | p{6.2cm} }
+   Name & Filename & Description \\\hline
+ ADa and SaRah & \texttt{ADAandSaRah.py} & ADSR envelope \\
+ Tremolos, Vibratos and Frequency & \texttt{bonds.py} & bonds between tremolos, vibratos and frequency  \\
+ Bella Rugosi & \texttt{bellaRugosi.py} & rugosity achieved through frequencies between 15 adn 30 Hz \\
+ Children Choir & \texttt{childChoir.py} & achieving choir sonorities by small variations of the same basic note \\
+ Noisy Band & \texttt{noisyBanda.py} & using various noises \\
+ ParaMeter Transitions & \texttt{paraMeter.py} & gradual changes of parameters within one note \\
+ Little Train of Impulsive Hillbilies~\footnote{This names translates into ``Trenzinho de Caipiras Impulsivos'' which is a pun with the Villa-Lobos' ``Trenzinho do Capira''.} & \texttt{littleTrain.py} & the use of convolution with impulses to achieve rhythm \\
+ Shakes and Wiggles & \texttt{shakesWiggles.py} &  tremolos and vibratos \\
+\end{tabular}
+\end{table*}
+
+\begin{table*}[htp!]
+\centering
+\caption{Piece names, script files and the concepts they exemplify from Section~\ref{sec:notesMusic}. All files are in the directory \texttt{src/pieces4/}.}
+\begin{tabular}{ p{5cm} | p{3.5cm} | p{6.2cm} }
+   Name & Filename & Description \\\hline
+ Acorde Cedo & \texttt{acordeCedo.py} & chord successions and modulation \\
+ Conta Ponto & \texttt{contaPonto.py} & melodic lines conducted within the rules of counterpoint \\
+ Crystals & \texttt{crystals.py} & symmetric musical scales \\
+ Dirracional & \texttt{dirracional.py} & directional arcs \\
+ Intervals & \texttt{intervals.py} & musical intervals \\
+ MicroTone & \texttt{microTone.py} & microtonality (the use of intervals smaller than the semitone) \\
+ Poly-Hit-My & \texttt{polyHitMy.py} & polyrhythm (multiple metrics at once) \\
+\end{tabular}
+\end{table*}
+
 \clearpage
 \subsection{To render the figures used in the article}
+The files \texttt{src/figures/*} render each of the figures (listed above in Section~\ref{sec:lfigures}.
 
@@
 \section{Other documents}
-Latex files. PDF files.
-MSc Dissertation.
+The files in \texttt{latex/*} render the article PDF and this Supporting Information file.
 
-\section{Final considerations}
+The script \texttt{src/aux/iso226.py} can generate the Equal Loudness Contour as described in its latest revision~\cite{iso226}.
+
+The article is largely based on the MSc dissertation~\cite{dissertacao}.
 

### 2017-08-05 article/acm/body.tex
finished the listings
added a section on synonymy and polysemy and other minor improvements

@@
 Section~\ref{cap:conclusao} holds final considerations and further work.
-The Supporting Information document~\cite{SI} holds an informative listing of all the equations, figures, tables and sections in this document
+
+\subsection{Additional material}
+The Supporting Information document~\cite{SI} holds commented listings of all the equations, figures, tables and sections in this document
 and the scripts in the \massa\ toolbox\footnote{The toolbox contains a collection of Python scripts which:
@@
 Additionally, the documentation of the toolbox consists of this article and the scripts themselves.}.
+The musical pieces are not very traditional which facilitates the understanding of specific techniques
+and the extrapolation of the note concept.
+
+\subsection{Synonymy and polysemy (disclaimer)}
+Given that the main topic of this article (the expression of musical elements
+in PCM audio) is multidisciplinary and involves art,
+the reader should be aware that many of the vocabulary admit different choices of terms and definitions.
+More specifically, it is often often the case where many words can express the same concept
+and where one word can carry different meanings.
+This is a very deep issue which might receive a dedicated manuscript.
+The reader might need to read the rest of this document to understand this
+small selection of synonymy and polysemy in the literature,
+but it is important to illustrate the point before the more dense sections:
+\begin{itemize}
+	\item A ``note'' can mean a pitch or an abstract construct with pitch and duration or a sound emitted from a musical instrument or a specific note in a score or a music.
+	\item The sampling rate (discussed above) is also called the sampling frequency.
+	\item An harmonic in a sound is most often a sinusoidal component which is is the harmonic series of the fundamental frequency.
+Many times, however, the terms harmonic and component are not distinguished.
+A harmonic can also be a note performed in a instrument by preventing certain overtones (components).
+	\item Harmony can refer to chords or to note sets related to chords or even to ``harmony'' in a more general sense, as a kind of balance and consistency.
+	\item A ``tremolo'' can mean different things, e.g.
+in a piano score, a tremolo is a fast alternation of two notes (pitches) while in computer music theory it is (most often) an oscillation of intensity.
+\end{itemize}
+
+We strived to avoid nomenclature clashes and the use of more terms and needed.
+Also, some of these issues are exposed when the context is convenient, such as in the discussion of timbre.
 
@@
         \includegraphics[width=.5\columnwidth]{figures/mixagem__}}
-    \caption{Mixing of three sound sequences. The amplitudes are directly summed.}
+    \caption{Mixing of three sound sequences. The amplitudes are directly summed sample-by-sample.}
         \label{fig:mixagem}
@@
 Velocity components should be found for each ear.
-Hence, given the audio source velocity $v_s$, with positive values if the source moves away from receptor, and receptor velocity $v_r$, positive when it gets closer to audio source, the frequency is given by the well-known shift in the Doppler effect:
+Hence, given the audio source speed $s_s$, with positive values if the source moves away from receptor, and receptor speed $s_r$, positive when it gets closer to audio source, the frequency is given by the well-known shift in the Doppler effect:
 
 \begin{equation}\label{eq:fDoppler}
-    f=\left(\frac{v_{sound}+v_r}{v_{sound}+v_s}\right)f_0
+    f=\left(\frac{s_{sound}+s_r}{s_{sound}+s_s}\right)f_0
 \end{equation}

### 2017-08-05 article/acm/body.tex
equations reviwed to the end of section 3

@@
 
-\begin{equation}\label{eq:ampVol}
-\begin{split}
-t_i^{'}=2 . t_i \Rightarrow pot(T^{'}_i)=4 . pot(T_i) \Rightarrow \\ \Rightarrow V^{'}_{dB}=10log_{10} 4 \approx 6 dB
-\end{split}
-\end{equation}
-
-\begin{equation}\label{eq:potVol}
-\begin{split}
-pot(T^{'}_i)=2 pot(T_i) \Rightarrow \\ \Rightarrow V^{'}_{dB}=10log_{10} 2 \approx 3 dB
-\end{split}
-\end{equation}
+\begin{align}
+t_i^{'}=2 t_i \Rightarrow pot(T^{'}_i)=4 pot(T_i) \Rightarrow V^{'}_{dB}=10log_{10} 4 \approx 6 dB \label{eq:ampVol}\\
+t_i^{'}=\sqrt{2} t_i \Rightarrow pot(T^{'}_i)=2 pot(T_i) \Rightarrow V^{'}_{dB}=10log_{10} 2 \approx 3 dB\label{eq:potVol}
+\end{align}
 
@@
 
-\begin{multline}\label{eq:reconsCompleta}
-t_i = \frac{a_0}{\Lambda} + \frac{2}{\Lambda}\sum_{k=1}^{\tau}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]+ \\ \frac{ a_{\Lambda/2}}{\Lambda}.(1-\Lambda\% 2)
-\end{multline}
+\begin{equation}\label{eq:reconsCompleta}
+\begin{split}
+	t_i = \frac{a_0}{\Lambda} + \frac{ a_{\Lambda/2}}{\Lambda}(1-\Lambda\% 2) + \frac{2}{\Lambda}\sum_{k=1}^{\tau}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]
+\end{split}
+\end{equation}
 
@@
 T_i^{f,\,\Delta}=\left\{t_i^f\right\}_0^{\lfloor \, f_a .
-\Delta \, \rfloor -1} = & \left\{ \, \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} }\, \right\}_{0}^{\Lambda-1}\; , \quad \\ \text{where} \;\; \gamma_i = & \left \lfloor i .
-f \frac{ \widetilde{\Lambda}}{f_a} \right \rfloor
+\Delta \, \rfloor -1} = \left\{ \, \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} }\, \right\}_{0}^{\Lambda-1}\; , \quad \text{where} \;\; \gamma_i = \left \lfloor i f \frac{ \widetilde{\Lambda}}{f_a} \right \rfloor
 \end{split}
@@
 \begin{split}
- \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow \quad \gamma_i= \left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor \\
-\gamma_i=  \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{j}{\Lambda-1} \right ] \right \rfloor 
+	\Delta_{\gamma_i}=\frac{\widetilde{\Lambda}}{f_a}f_i \quad \Rightarrow \quad \gamma_i & = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f_j \right \rfloor \\
+	\gamma_i & =  \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{j}{\Lambda-1} \right ] \right \rfloor 
 \end{split}
@@
 \begin{split}
- \Delta_{\gamma_i}= f_i\frac{\widetilde{\Lambda}}{f_a} \quad \Rightarrow  \quad \gamma_i= & \left \lfloor \sum_{j=0}^{i} f_j\frac{\widetilde{\Lambda}}{f_a} \right \rfloor \\ \gamma_i = & \left \lfloor \sum_{j=0}^{i} f_0 \frac{\widetilde{\Lambda}}{f_a} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{j}{\Lambda -1}} \right \rfloor
+ \Delta_{\gamma_i}= \frac{\widetilde{\Lambda}}{f_a}f_i \quad \Rightarrow  \quad \gamma_i= & \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f_j \right \rfloor \\
+	\gamma_i = & \left \lfloor \sum_{j=0}^{i} f_0 \frac{\widetilde{\Lambda}}{f_a} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{j}{\Lambda -1}} \right \rfloor
 \end{split}
@@
 \begin{split}
- \{a_i\}_0^{\Lambda-1}= & \left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}= \\ = & \left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1} \text{ (with } a_0=1 \text{)}
-\end{split}
-\end{equation}
-and
+ \{a_i\}_0^{\Lambda-1}= \left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}= \left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
+\end{split}
+\end{equation}
+where $a_0=1$. And, therefore:
+
 \begin{equation}\label{transAmp}
@@
  \item The pink noise is characterized by a decrease of $3dB$ per octave. This noise is useful for testing electronic devices, being prominent in nature~\cite{Roederer}. 
-\begin{multline}\label{eq:rosa}
  %\begin{split}
- \qquad \qquad \qquad f_{\text{min}}  \approx 15 Hz \\
- f_i  = i \frac{f_a}{\Lambda} \;, \;\; \quad i \;\leq\; \frac{\Lambda}{2},\;\; i\;\in\;\mathbb{N}  \\
- \alpha_i  = \left(10^{-\frac{3}{20}}\right)^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )}  \\
- c_i  =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \\
- c_i  =e^{j.x} . \alpha_i\;, \; x \; \text{random} \; \in \; [0,2\pi]\;,\;\; \forall \; i \; : f_{\text{min}} \le f_i < f_{\lceil \Lambda/2-1 \rceil}  \\
- c_{\Lambda/2}  = \alpha_{\Lambda/2}\;, \; \text{if $\Lambda$ even} \\ 
- c_i  = c_{\Lambda - i}^*\;,\;\; \text{for}\;  i \; > \;  \Lambda/2 \qquad \qquad
+\begin{equation}\label{eq:rosa}
+\begin{split}
+	f_{\text{min}} & \approx 15 Hz \\
+	f_i  &= i \frac{f_a}{\Lambda} \;, \;\; \quad i \;\leq\; \frac{\Lambda}{2},\;\; i\;\in\;\mathbb{N}  \\
+	\alpha_i & = \left(10^{-\frac{3}{20}}\right)^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )}  \\
+	c_i & =0\;,\;\; \forall \; i \; : f_i<f_{\text{min}} \\
+	c_i & =e^{j.x} \alpha_i\;, \; x \; \text{random} \; \in \; [0,2\pi]\;,\;\; \forall \; i \; : f_{\text{min}} \le f_i < f_{\lceil \Lambda/2-1 \rceil}  \\
+	c_{\Lambda/2} & = \alpha_{\Lambda/2}\;, \; \text{if $\Lambda$ even} \\ 
+	c_i & = c_{\Lambda - i}^*\;,\;\; \text{for}\;  i \; > \;  \Lambda/2 \qquad \qquad
+\end{split}
+\end{equation}
  %\end{split}
-\end{multline}
 The minimum frequency $f_{\text{min}}$ is chosen with regard to the human hearing, since a sound component with frequency below $\approx\; 20Hz$ is virtually inaudible.
@@
 The definitions given here are common in contexts regarding music theory and electronic music.
-In addition, they are based on a broader literature than the one used for a specific instrument, practice or musical tradition~\cite{Lacerda,Harmonia}.}
-For the general case, vibrato is described as:
+In addition, they are based on a broader literature than the one used for a specific instrument, practice or musical tradition~\cite{Lacerda,Harmonia}.}.
+A vibrato can be achieved by:
 
@@
 
-\begin{multline}\label{vbrGamma2}
- \Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_s} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_s} \right \rfloor = \\ = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}f \left ( \frac{f + \mu }{f} \right )^{t_j'}  \right \rfloor= \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}f .
-2^{t_j'\frac{\nu}{12}}  \right \rfloor
-\end{multline}
+\begin{equation}\label{vbrGamma2}
+\begin{split}
+	\Delta_{\gamma_i}=\frac{\widetilde{\Lambda}}{f_s}f_i \quad \Rightarrow \quad \gamma_i & = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}f_j \right \rfloor \\ 
+	& = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}f \left ( \frac{f + \mu }{f} \right )^{t_j'}  \right \rfloor \\
+	& = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}f . 2^{t_j'\frac{\nu}{12}}  \right \rfloor
+\end{split}
+\end{equation}
 
@@
          \includegraphics[width=\columnwidth]{figures/vibrato_}
-     \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave, with $f_s=44.1kHz$.}
+     \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave, with $f_s=44.1kHz$. The color bar is uses decibels.}
          \label{fig:vibrato}
@@
 \end{equation}
-\noindent and finally: 
-\begin{multline}\label{trT}
+\noindent and, finally:
+
+\begin{equation}\label{trT}
+\begin{split}
  T_i^{tr(f')}=\left \{ t_i^{tr(f')} \right \}_0^{\Lambda-1}=\{ t_i .
-a_i \}_0^{\Lambda-1}= \\ =\left \{t_i .10^{t_i' \frac{V_{dB}}{20}}    \right \}_0^{\Lambda-1}=\left\{t_i .
+a_i \}_0^{\Lambda-1}= \left \{t_i .10^{t_i' \frac{V_{dB}}{20}}    \right \}_0^{\Lambda-1}=\left\{t_i .
 a_{\text{max}}^{t_i'} \right\}_0^{\Lambda-1}
-\end{multline}
+\end{split}
+\end{equation}
 
@@
 
-\begin{multline}\label{eq:Bessel}
-J_k(\mu) = \\ = \frac{2}{\pi} \int_0^{\frac{\pi}{2}}\left [ cos \left (\overline{k}\;\frac{\pi}{2} + \mu .
+\begin{equation}\label{eq:Bessel}
+J_k(\mu) = \frac{2}{\pi} \int_0^{\frac{\pi}{2}}\left [ cos \left (\overline{k}\;\frac{\pi}{2} + \mu .
 \sin w \right ) .
 cos \left ( \overline{k}\;\frac{\pi}{2} + k .
-w \right ) \right ] dw \\  \overline{k} = k \% 2 \;\;,\;\; k \in \mathbb{N}
-\end{multline}
+w \right ) \right ] dw \;,\; \overline{k} = k \% 2 \;,\; k \in \mathbb{N}
+\end{equation}
 
@@
 \begin{split}
-    R_i^1=\left\{r_i^1\right\}_0^{\Lambda_1-1}\;:\quad\quad\quad\quad\quad\quad\quad\quad \\ 
-\;r_i^1=\left\{
+	R_i^1  =\left\{r_i^1\right\}_0^{\Lambda_1-1}\;,\; \text{where}\;\;
+	r_i^1  =\left\{
         \begin{array}{l l}
@@
 \begin{equation}\label{eq:rev}
-    R_i=\left\{r_i\right\}_0^{\Lambda_R-1}\;:\;r_i=\left\{
+	R_i=\left\{r_i\right\}_0^{\Lambda_R-1}\;,\; \text{where} \; r_i=\left\{
         \begin{array}{l l}

### 2017-08-05 article/scripts/vibrato.py

@@ -39,6 +39,6 @@ p.xticks((0,10000,20000,30000,44000),(r"0",10000,20000,30000,44100),
         fontsize=16)
 p.yticks(fontsize=16)
 p.ylabel(u"frequency "+r"$ \in \; [0,\,\frac{f_a=44100}{2}=22050] \, \rightarrow $", fontsize=19)
-p.xlabel(r"samples $\quad \rightarrow$",fontsize=26)
+p.xlabel(r"samples $\quad \rightarrow$",fontsize=21)
 p.savefig("../figures/vibrato_.png")
 p.show()

### 2017-08-05 article/acm/body.tex
enhanced disclaimer, equations and notation

@@
 representing the sound with samples equally spaced
-by a duration $\lambda_s$, and specifying the amplitude of each sample by a fixed number of bits.
+by a duration $\delta_s$, and specifying the amplitude of each sample by a fixed number of bits.
 This is the Pulse Code Modulation (PCM) representation of sound.
 A PCM audio format has two essential attributes:
-a sampling frequency $f_s=\frac{1}{\lambda_s}$ (also called e.g. sampling rate or sample rate),
+a sampling frequency $f_s=\frac{1}{\delta_s}$ (also called e.g. sampling rate or sample rate),
 which is the number of samples used for representing a second of sound;
@@
 The objectives of this paper are:
-
 \begin{enumerate}
@@
 
-\subsection{Synonymy and polysemy (disclaimer)}
+\subsection{Synonymy, polysemy and theoretical frames (disclaimer)}
 Given that the main topic of this article (the expression of musical elements
@@
 
-We strived to avoid nomenclature clashes and the use of more terms and needed.
-Also, some of these issues are exposed when the context is convenient, such as in the discussion of timbre.
+We strived to avoid nomenclature clashes and the use of more terms than needed.
+Also, there are many theoretical standpoints for the understanding of musical phenomena,
+which is an evidence that most often there is not a single way to express or characterize musical structures.
+Therefore, in this article, adjectives such as "often" "commonly" "frequently" are abundant and they would probably be even more numerous if we wanted to be pedantically precise.
+Some of these issues are exposed when the context is convenient, such as in the first considerations of timbre.
 
@@
 That is, the integer part of the multiplication is considered, and an error of at most $\delta_s$ missing seconds is admitted, which is usually fine for musical purposes.
-As an example, if $f_s=44.1kHz \;\;\Rightarrow\;\;\lambda_s=\frac{1}{44100}\approx 23$ microseconds.
+As an example, if $f_s=44.1kHz \;\;\Rightarrow\;\;\delta_s=\frac{1}{44100}\approx 23$ microseconds.
 Thus:
@@
 
-\noindent where $\lambda_f=\frac{f_s}{f}=\frac{\delta_f}{\lambda_s}$ is the number of samples in the period.
+\noindent where $\lambda_f=\frac{f_s}{f}=\frac{\delta_f}{\delta_s}$ is the number of samples in the period.
 
@@
 \begin{equation}\label{escSim}
-\begin{split}
-\text{chromatic}    & = E_i^c = \{e_i^c\}_0^{11} = \\
-                    & =  \{0,1,2,3,4,5,6,7,8,9,10,11\} = \{i\}_0^{11}\\
-	\text{whole tones}  & = E_i^{wt} = \{e_i^{wt}\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
-\text{minor thirds} & = E_i^{mt} = \{e_i^{mt}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
-\text{major thirds} & = E_i^{Mt} = \{e_i^{Mt}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
-\text{tritones}     & = E_i^{tt} = \{e_i^{tt}\}_0^{1} = \{ 0, 6 \} = \{6.i\}_0^1
-\end{split}
+\begin{aligned}
+	\text{chromatic}    & = E_i^c    & = \{e_i^c\}_0^{11}   & =  \{0,1,2,3,4,5,6,7,8,9,10,11\} & = \{i\}_0^{11}\\
+	\text{whole tones}  & = E_i^{wt} & = \{e_i^{wt}\}_0^{5} & = \{0,2,4,6,8,10\}               & = \{2.i\}_0^{5} \\
+	\text{minor thirds} & = E_i^{mt} & = \{e_i^{mt}\}_0^{3} & = \{0,3,6,9\}                    & = \{3.i\}_0^3 \\
+	\text{major thirds} & = E_i^{Mt} & = \{e_i^{Mt}\}_0^{2} & = \{0,4,8\}                      & = \{4.i\}_0^2\\
+	\text{tritones}     & = E_i^{tt} & = \{e_i^{tt}\}_0^{1} & = \{ 0, 6 \}                     & = \{6.i\}_0^1
+\end{aligned}
 \end{equation}
@@
 \begin{equation}\label{eq:escalas}
-\begin{split}
-\text{aeolian}    & = \text{natural minor scale} = \\
-                  & = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
-\text{locrian}    & = E_i^{lo} = \{e_i^{lo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
-\text{ionian}     & = \text{major scale} =  \\
-                       & = E_i^M = \{e_i^M\}_0^6 = \{0,2,4,5,7,9,11\} \\
-\text{dorian}     & = E_i^{d} = \{e_i^{d}\}_0^6 = \{0,2,3,5,7,9,10\} \\
-\text{phrygian}   & = E_i^{p} = \{e_i^{p}\}_0^6 = \{0,1,3,5,7,8,10\} \\
-\text{lydian}     & = E_i^{l}=\{e_i^{l}\}_0^6 = \{0,2,4,6,7,9,11\} \\
-\text{mixolydian} & = E_i^{mi} = \{e_i^{mi}\}_0^6 = \{0,2,4,5,7,9,10\}
-\end{split}
+\begin{aligned}
+	\text{aeolian}    & = \text{natural minor scale}  & = & \\
+			  & = E_i^m = \{e_i^m\}_0^6       & = & \{0,2,3,5,7,8,10\} \\
+	\text{locrian}    & = E_i^{lo} = \{e_i^{lo}\}_0^6 & = & \{0,1,3,5,6,8,10\} \\ 
+	\text{ionian}     & = \text{major scale}          & = &  \\
+			  & = E_i^M = \{e_i^M\}_0^6       & = & \{0,2,4,5,7,9,11\} \\
+	\text{dorian}     & = E_i^{d} = \{e_i^{d}\}_0^6   & = & \{0,2,3,5,7,9,10\} \\
+	\text{phrygian}   & = E_i^{p} = \{e_i^{p}\}_0^6   & = & \{0,1,3,5,7,8,10\} \\
+	\text{lydian}     & = E_i^{l}=\{e_i^{l}\}_0^6     & = & \{0,2,4,6,7,9,11\} \\
+	\text{mixolydian} & = E_i^{mi} = \{e_i^{mi}\}_0^6 & = & \{0,2,4,5,7,9,10\}
+\end{aligned}
 \end{equation}
@@
 
-Although it is not a traditional scale, the harmonic series is often used as such::
+Although it is not a traditional scale, the harmonic series is often used as such:
 \begin{equation}\label{eq:serieHarmonica}
@@
 H_i = & \{h_i\}_0^{19}= \\
-      & \{ 0,12,19+0.02,  24,28-0.14, 31+0.2, 34-0.31, \\
+    =  & \{ 0,12,19+0.02,  24,28-0.14, 31+0.2, 34-0.31, \\
                      & 36, 38+0.04,40-0.14, 42-0.49, 43+0.02, \\
@@
      & \text{(associativity property)} \\
-\exists \;\; e \in G :                  \quad    & p \bullet e  = e \bullet p \;\;\;\; \forall p \in G  \\ 
-     &  \text{(identity element existence)} \\
+\exists \;\; e \in G :                  \quad    & p \bullet e  = e \bullet p \;,\;\;\; \forall\; p \in G  \\ 
+     &  \text{(existence of the identity element)} \\
 \forall \;\; p \in G, \;\exists\; p^{-1} :\quad  &  p\bullet p^{-1}i =p^{-1}\bullet p = e \\
-     &  \text{(inverse element existence)}
+     &  \text{(existence of the inverse element)}
 \end{split}
@@
 The possibilities provided by this exposition pour from both the organization of knowledge and the ability to achieve sounds which are extremely true to the models.
-For example, one can produce noises with arbitrary spectrum resolution and a musical note can be synthesized with the parameters (e.g. of a vibrato) updated sample-by-sample.
-These abilities can be crucial e.g. for psychoacoustic experiments or for the development of
+For example, one can produce noises with arbitrary resolution of the spectrum and a musical note can be synthesized with the parameters (e.g. of a vibrato) updated sample-by-sample.
+Furthermore, software for synthesis and processing of sounds for musical purposes
+by standard restrict the bit depth to 16 or 24.
+This is achievable in this presented framework but by standard Python uses more bits per floating point number.
+These ``higher fidelity'' characteristics can be crucial
+e.g. for psychoacoustic experiments or for the development of
 interfaces to generate high quality musical sounds or pieces.
 Simply put, it is compelling
-for many scientific, artistic and didactic purposes.
-A fact evidenced by a number of uses this framework had in courses, in software implementations and in making music~\cite{vimeoLM,vivace,repoDissertacao,dissertacao}.
+for many scientific and artistic purposes.
+The didactic potential of the framework is evident when noticed that:
+\begin{itemize}
+	\item the integrals and derivatives, ubiquitous in continuous signal processing,
+are all replaced by summations in discrete signals,
+which are more intuitive and does not require fluency in calculus.
+	\item The equations and concepts are implemented as software which can be easily assembled and inspected.
+\end{itemize}
+\noindent In fact, this framework was used in a number of contexts, including courses, software implementations and making music~\cite{vimeoLM,vivace,repoDissertacao,dissertacao}.
 

### 2017-08-07 article/acm/body.tex
arrumacoes finais antes da leitura final

@@
 More specifically, what follows is a summary of academic music composition theory and praxis.
+This section does not benefit from equations that dictate the amplitude of each sample as deeply as the previous sections.
+Even so, we understand that this content is the missing piece for synthesizing music given all the content above.
+The concepts are given algorithmic implementations in \mass~\cite{SI}.
 
@@
 
-\subsubsection{Atonal harmony}
+\subsubsection{Atonal harmony}\label{sec:atonal}
 In fact, atonal music techniques avoid that a direct relation of the notes with modes and tonality be established.
@@
 
-In practice, and historically, there is special incidence and use of the golden ratio. The Lucas sequence allows the generalization of
-Fibonacci sequence, making it easy to follow. Given any two numbers $x_0$
-and $x_1$, the Lucas sequence can be obtained as: $x_n=x_{n-1}+x_{n-2}$. The greater $n$ is, the more $\frac{x_{n}}{x_{n_1}}$ approaches the golden ratio:
-$1.61803398875...$. The sequence converges fast even with high discrepant
-initial values. If $x_0=1$ and $x_1=100$, and $y_n=\frac{x_n}{x_{n+1}}$, the error for the first values with
+In practice, and historically, there is special incidence and use of the golden ratio.
+The Fibonacci sequence might be generalized as follows in order for any two numbers be used
+to approach the golden ratio.
+Given any two numbers $x_0$
+and $x_1$, define the elements of the sequence $\{x_n\}$ as: $x_n=x_{n-1}+x_{n-2}$.
+The greater $n$ is, the more $\frac{x_{n}}{x_{n_1}}$ approaches the golden ratio
+($1.61803398875...$). The sequence converges fast even with discrepant
+initial values.
+E.g. let $x_0=1$, $x_1=100$ and $y_n=\frac{x_n}{x_{n+1}}$, the error for the first values with
 respect to the golden ratio is, approximately, $\{ e_n \}
 =\left\{100\frac{y_n}{1.61803398875}-100 \right\}_1^{10}=\{6080.33, -37.57, 23,
--7.14, 2.937, -1.09, 0.42, -0.1601, \\ 0.06125, -0.02338\}$. The Fibonacci sequence
+-7.14, 2.937, -1.09, 0.42, -0.1601, 0.06125, -0.02338\}$. The Fibonacci sequence
 presents the same error progression, but starts at the second step of a more discrepant initial setting
 ($\frac{1}{1}\approx\frac{100+1 = 101}{100}$).
+One might benefit from the On-Line Encyclopedia of Integer Sequences (OEIS~\cite{oeis})
+for exploring various sequences.
 
@@
 
-These arcs can be established by using a set of different permutations. As a historic
+These arcs can be established by using a set of permutations. As a historic
 example, the \emph{change ringing} tradition conceives music through
 bells played one after another and then played again, but in a different
-order. This process repeats until it reaches the initial ordering. The sequence of
+order. This process is repeated until it reaches the initial ordering. The sequence of
 different orderings is a \emph{peal}. Table~\ref{tab:change}
-presents a traditional \emph{peal}, named ``Hunt'', for 3 bells (\textcolor{red}{1}, \textcolor{blue}{2} and \textcolor{green}{3}), which explores
+presents a traditional \emph{peal}, named ``Plain Change''~\cite{change}, for 3 bells (\textcolor{red}{1}, \textcolor{blue}{2} and \textcolor{green}{3}), which explores
 all possible orderings. Each line indicates one bell ordering to be
@@
 The \massa\ toolbox presents a computational implementation that isolates
-permutation aplication to sonic characteristics, in order to deliver
+permutation application to sonic characteristics, in order to deliver
 musical structures~\cite{MASSA,figgusOriginal,figgusEspacializacao}.
+
+\subsection{Serialism and post-serial techniques}
+Expanding the concepts in Section~\ref{sec:atonal},
+sequences of characteristics can be predefined and
+used throughout a musical piece.
+These sequences can be of intensities, timbre, durations,
+density of events, etc.
+The sequences can be used very strictly or loose,
+such as by skipping some elements.
+The sequences can be of different sizes, yielding arcs
+until the initial condition is reached again (i.e. cycles).
+One paradigmatic case is the ``total serialism'' where all
+the musical characteristics are serialized.
+Although the use of sequences is inherent to music (e.g. scales, metric pulses),
+their use with greater emphasis than tonal (or modal) elements
+in western music, as an artistic trend, took place only in the
+first half of the twentieth century and is called ``serialism''.
+Post-serial techniques are numerous, but here is a description of
+important concepts:
+\begin{itemize}
+	\item Spectralism:
+	\item Spectromorphology:
+	\item Stochastic music:
+	\item Textures:
+\end{itemize}
 

### 2017-08-08 article/acm/body.tex
finished all notes before last read

@@
 This section does not benefit from equations that dictate the amplitude of each sample as deeply as the previous sections.
-Even so, we understand that this content is the missing piece for synthesizing music given all the content above.
-The concepts are given algorithmic implementations in \mass~\cite{SI}.
+Even so, we understand that this content is very useful for synthesizing music given all the content above.
+The concepts are given algorithmic implementations in \mass~\cite{SI} and can be further formalized~\cite{topos},
+although at a cost of intelligibility which we chose to avoid in this exposition.
 
@@
 
-% scramble 
 From these processes, many others are derived, such as the inverted retrograde, the temporal contraction with an external suffix, etc.
@@
 musical therapy, known by its utility in cases of depression and neurological injury.
-Also, tt is known that regions of the human brain responsible for sonic processing are also used for other activities, such as for performing verbal discourse and mathematics~\cite{Sacks,Roederer}.
+Also, it is known that regions of the human brain responsible for sonic processing are also used for other activities, such as for performing verbal discourse and mathematics~\cite{Sacks,Roederer}.
 
@@
 directional and cyclic arcs.
+
+For a scholastic discussion on the composition of motives, phrases, melodies, themes and music form (such as rondo, ternary, theme and variations) see~\cite{Schoenberg}.
+% melody, answer and reply, motives, etc
 
@@
 	\item Spectralism:
+		consists on use the (Fourier) spectrum of a sound or the harmonic series for music composition,
+		such as
+		to derive harmonies, sequences of pitches or a temporal evolution of the overall spectrum~\cite{grisey}.
+		In other words, the most prominent frequencies can be used as pitches,
+		real notes can be used to mimic an original spectrum
+		(e.g. use piano notes to mimic the spectrum of a spoken sentence) or portions of the spectrum made to vary~\cite{spectral}.
 	\item Spectromorphology:
+		can be considered a spectral music (spectralism) theoretical framework~\cite{smalley,schaeffer}
+		that examines the relation between sound spectra and their temporal evolution.
+		The theory poses e.g. different onsets, continuations and terminations; characteristics of (sonic) ``motion'';
+		and spectral density.
 	\item Stochastic music:
+		the use of random variables to describe musical elements are largely considered in stochastic music~\cite{formalized}.
+		In summary, one can use probability distributions for the synthesis of sounds and larger scale musical elements.
+		Changes in these distributions or in other characteristics yield the discourse.
 	\item Textures:
+		sounds can be assembled in terms of a ``sonic texture''.
+		Here, a texture is often thought about very abstractly
+		as a sonic counterpart of visual texture.
+		Parameters that can be used for such are: range between highest and lowest note,
+		density of notes, durations of notes, motives, number of voices, etc.
+		If the sounds are small enough (typically < $100ms$) the process can be though of in terms of \emph{granular synthesis}~\cite{microsound}.
 \end{itemize}

### 2017-08-08 article/acm/body.tex
finished final review of abstract and introduction

@@
 Since the speed of sound is $\approx 343.2 m/s$,
-such frequency limits imply wavelengths of $\frac{343.2}{20} \approx 17.16\,m$ and
+such frequency limits corresponds to  wavelengths of $\frac{343.2}{20} \approx 17.16\,m$ and
 $\frac{343.2}{20000} \approx 17.16\,mm$.
@@
 and higher level musical structures (such as tunings, scales and chords, in Section~\ref{sec:notesMusic}).
+
+\begin{figure*}[!h]
+    \centering
+        \includegraphics[width=.7\textwidth]{figures/pcm_}
+	\caption{Example of PCM audio:
+	a sound wave is represented by 25 samples equally spaced in time where each sample has an amplitude specified with 4 bits.}
+        \label{fig:PCM}
+\end{figure*}
 
@@
 which is the number of samples used for representing a second of sound;
-and a bit depth, which is the number of bits for each sample.
+and a bit depth, which is the number of bits for the amplitude of each sample.
 Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with a bit depth of $4$,
@@
 and a total of $4 \times 25= 100$ bits for representing the whole sound.
-
-\begin{figure*}
-    \centering
-        \includegraphics[width=.7\textwidth]{figures/pcm_}
-	\caption{Example of PCM audio:
-	a sound wave is represented by 25 samples equally spaced in time where each sample has an amplitude specified with 4 bits.}
-        \label{fig:PCM}
-\end{figure*}
 
@@
 that is, $f_s \geq 2\times 20kHz = 40kHz$.
-This is the fundamental reason for the adoption of sampling frequencies
+This is the basic reason for the adoption of sampling frequencies
 such as $44.1kHz$ and $48kHz$,
@@
 (such as atoms in matter or cells in macroscopic organisms) and,
-in practice, it can unfold into sounds that uphold other approaches.
+in practice, it can unfold into sounds that uphold other approaches to music.
 This is of capital importance because science and scholastic artists
@@
 (commercial) music had also incorporated
-sounds amalgams and abstract discursive arcs~\footnote{There
-are well known incidence of such musical techniques in ethnic music,
-such as in Pygmy music, but western music theory assimilated them
+sound amalgams and abstract discursive arcs~\footnote{There
+are well known incidence of such characteristics in ethnic music,
+such as in Pygmy music, but western theory assimilated them
 only in the last century~\cite{Wisnick}.}.
@@
 This article aims at representing musical structures and artifices
-by their discrete-time sonic characteristics.
+in discrete-time audio.
 The results include mathematical relations,
@@
 and their implementations as software routines both as
-very raw and direct algorithms and in the context of rendering musical pieces.
+very raw algorithms and in the context of rendering musical pieces.
 Despite the general interests involved,
@@
 \begin{enumerate}
-	\item Present a concise set of mathematical relations between musical basic elements and sequences of PCM audio samples.
+	\item Present a concise set of mathematical and algorithmic relations between musical basic elements and sequences of PCM audio samples.
 	\item Introduce a framework of sound and musical synthesis with control at sample level which entails potential uses in psychoacoustic experiments and synthesis with extreme precision (recap in Section~\ref{cap:conclusao}).
@@
 		the analytic relations presented in this article are implemented as public domain scripts,
-		i.e. small computer programs using accessible technologies for distribution and validation.
-		This constitutes the \massa\ toolbox, available in an open Git repository~\cite{gitBook}.
+		i.e. small computer programs for prompt distribution and validation.
+		This constitutes the \massa\ toolbox, available in an open source Git repository~\cite{gitBook}.
 		These scripts are written in Python and make use of Numpy,
 		which performs numerical routines efficiently (e.g. through LAPACK).
-		Part of the scripts have been transcribed to JavaScript and native Python,
-		which favors their use in Web browsers such as Firefox and
-		Chromium~\cite{numpy, audiolab, tutpython, python}.
+		Part of the scripts have been transcribed to JavaScript (which favors their use in Web browsers such as Firefox and Chromium) and native Python~\cite{numpy, audiolab, tutpython, python}.
 		These are all open technologies, published using licenses that allow for copy,
@@
 	\item Provide a didactic presentation of the content, which is highly multidisciplinary,
-		involving signal processing, music, psycho-acoustics and programming.
+		involving signal processing, music, psychoacoustics and programming.
 \end{enumerate}
@@
 and presents methods, results and discussions altogether.
-Section~\ref{cap:conclusao} holds final considerations and further work.
+Section~\ref{cap:conclusao} is dedicated to final considerations and further work.
 
@@
 \end{itemize}
-Additionally, the documentation of the toolbox consists of this article and the scripts themselves.}.
+The documentation of the toolbox consists of this article, the Supporting Information and the scripts themselves.}.
 The musical pieces are not very traditional which facilitates the understanding of specific techniques
@@
 in PCM audio) is multidisciplinary and involves art,
-the reader should be aware that many of the vocabulary admit different choices of terms and definitions.
-More specifically, it is often often the case where many words can express the same concept
+the reader should be aware that much of the vocabulary admit different choices of terms and definitions.
+More specifically, it is often the case where many words can express the same concept
 and where one word can carry different meanings.
@@
 	\item A ``note'' can mean a pitch or an abstract construct with pitch and duration or a sound emitted from a musical instrument or a specific note in a score or a music.
-	\item The sampling rate (discussed above) is also called the sampling frequency.
-	\item An harmonic in a sound is most often a sinusoidal component which is is the harmonic series of the fundamental frequency.
+	\item The sampling rate (discussed above) is also called the sampling frequency or sample rate.
+	\item An harmonic in a sound is most often a sinusoidal component which is in the harmonic series of the fundamental frequency.
 Many times, however, the terms harmonic and component are not distinguished.
-A harmonic can also be a note performed in a instrument by preventing certain overtones (components).
+A harmonic can also be a note performed in an instrument by preventing certain overtones (components).
 	\item Harmony can refer to chords or to note sets related to chords or even to ``harmony'' in a more general sense, as a kind of balance and consistency.
@@
 which is an evidence that most often there is not a single way to express or characterize musical structures.
-Therefore, in this article, adjectives such as "often" "commonly" "frequently" are abundant and they would probably be even more numerous if we wanted to be pedantically precise.
+Therefore, in this article, adjectives such as "often", "commonly" and "frequently" are abundant and they would probably be even more numerous if we wanted to be pedantically precise.
 Some of these issues are exposed when the context is convenient, such as in the first considerations of timbre.
@@
 \label{cap:conclusao}
-We believe that this article is effective in relating musical elements to digital audio.
+In our understanding,
+this article is effective in relating musical elements to digital audio.
 We aimed at achieving a concise presentation of the subject because it involves many knowledge fields, and therefore can very easily blast into thousands of pages.
-Some readers might benefit from the text alone, but the \emph{Scripts} in the \massa\ toolbox, where all the relations presented here are directly and simply implemented as software (in Python), are very helpful for one to achieve elaborated implementations and deeper understandings.
-The scripts include routines that render musical pieces that illustrate the concepts.
-This is valuable since art (e.g. music) can involve many non-trivial processes and is often deeply glamurized, which results in a nearly unmanagable terrain for a newcommer.
+Some readers might benefit from the text alone, but the \emph{Scripts} in the \massa\ toolbox, where all the equations and concepts are directly and simply implemented as software (in Python),
+are very helpful for one to achieve elaborated implementations and deeper understandings.
+The scripts include routines that render musical pieces that illustrate the concepts in practical contexts.
+This is valuable since art (music) can involve many non-trivial processes and is often deeply glamorized, which results in a nearly unmanageable terrain for a newcomer.
 Moreover, this didactic report and the supplied open source scripts should facilitate the use of the proposed framework.
+The Supporting Information document holds listings of sections, equations, figures, tables, scripts and other documents.
 
@@
 by standard restrict the bit depth to 16 or 24.
-This is achievable in this presented framework but by standard Python uses more bits per floating point number.
+This is achievable in this framework but by standard Python uses more bits per floating point number.
 These ``higher fidelity'' characteristics can be crucial
-e.g. for psychoacoustic experiments or for the development of
-interfaces to generate high quality musical sounds or pieces.
+e.g. for psychoacoustic experiments or to generate high quality musical sounds or pieces.
 Simply put, it is compelling
@@
 	\item the integrals and derivatives, ubiquitous in continuous signal processing,
-are all replaced by summations in discrete signals,
+are all replaced, in discrete signals, by summations,
 which are more intuitive and does not require fluency in calculus.
@@
 \noindent In fact, this framework was used in a number of contexts, including courses, software implementations and making music~\cite{vimeoLM,vivace,repoDissertacao,dissertacao}.
-
-This work systematically investigates parameterization issues (like for the tremolo, ADSR, etc.) in a sample-by-sample synthesis, which entails high/extreme fidelity and if of significant artistic convenience.
-Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, as far as the authors know, such as testified in the literature review~\cite{dissertacao}, where the books, articles and open software are related to this framework.
+Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, as far as the authors know, such as testified in the literature review (Appendix G of~\cite{dissertacao}), where books, articles and open software are related to this framework.
 
 The free software license and online availability of the content,
-with the respective codes and sonic examples, facilitate future collaborations and the generation of sub-products in a co-authorship fashion.
-As consequence, the expansion of \massa\ is favoured and straightforward, easing new implementations and development of musical pieces.
+with the respective codes and sonic examples, facilitate collaborations and the generation of sub-products in a co-authorship fashion,
+easing new implementations and development of musical pieces.
 The scripts can be divided in three groups: implementation of all the equations and topics of music theory; routines for rendering musical pieces that illustrate the concepts; scripts that render the figures of this article and the article itself.
 
-In addition, this framework permitted the formation of interest groups in topics such as musical creativity and computer music.
+This framework favored the formation of interest groups in topics such as musical creativity and computer music.
 In particular, the project \url{labMacambira.sourceforge.net} groups Brazilian and foreign co-workers
 in diverse areas that range from digital direct democracy and georeferencing to art and education.
-This was only possible because of the audiovisual utility in many contexts, in particular because of the knowledge and mastery condensed in the \massa\ framework.\footnote{There are more than 700 videos, written documentats,
- original software applications and contributions in well-known external softwares such as Firefox, Scilab, LibreOffice, GEM/Puredata, to name a few~\cite{siteLM,wikiLM,vimeoLM}.
+This was only possible because of the usefulness of audiovisual abilities in many contexts, in particular because of the knowledge and mastery condensed in the \massa\ framework.\footnote{There are more than 700 videos, written documents,
+ original software applications and contributions in well-known external software (such as Firefox, Scilab, LibreOffice, GEM/Puredata, to name just a few)~\cite{siteLM,wikiLM,vimeoLM}.
 Some of these efforts are available online~\cite{dissertacao}.
-It is obvious that all these contributions are a consequence of more that just \massa, but it is evident to the authors that \massa\ had a primary role in converging interests and attracting collaborators.}
-
-Future work should include application of these results in artificial intelligence
-for the generation of appealing artistic materials. Some psychoacoustical effects were detected,
-which needs validation and should be reported, specially with~\cite{quadrosSonoros}.\footnote{The sonic pictures where sent to a public mailing list and the fifth sonic picture was reported by some individuals to induce a state in which noises from the own tongue and teeth of the indidual echoed for some seconds~\cite{quadrosMetarec}.} Other foreseen advances
+It is evident that all these contributions are a consequence of more that just \massa, but it is also evident to the authors that \massa\ had a primary role in converging interests and attracting collaborators.}
+
+Future work might include application of these results in artificial intelligence
+for the generation of appealing artistic materials. Some psychoacoustic effects were detected,
+which needs validation and should be reported, specially with~\cite{quadrosSonoros}.\footnote{The sonic pictures where sent to a public mailing list and the fifth sonic picture was reported by some individuals to induce a state in which noises from the own tongue, teeth and jaw of the individual echoed for some seconds~\cite{quadrosMetarec}.} Other foreseen advances
 are: a JavaScript version of the toolbox, better hypermedia deliverables of this framework, user guides

### 2017-08-08 article/acm/sample-acmsmall.tex
finished all notes before last read
finished final review of abstract and introduction

@@
 \usepackage{multirow}
+\usepackage{amsmath}
+\usepackage{amssymb}
 
@@
 \begin{abstract}
-The representation of the basic elements of music - such as notes,
-ornaments and intervalar structures - in terms of discrete audio
+The representation of basic elements of music - such as notes,
+ornaments and intervals - in terms of discrete audio
 signals is often used in software for music creation and design.
-Nevertheless, there is no unified approach that relates these elements to the sound discrete samples.
-In this article, each musical element is related by equations to the discrete-time samples of sounds,
+Nevertheless, there is no unified approach that relates these elements to the discrete samples of digitized sound.
+In this article, each musical element is related by equations and algorithms to the discrete-time samples of sounds,
 and each of these relations are implemented in scripts within a software toolbox,
@@
 The toolbox also includes this article and every script necessary to render the images
-and musical examples, pieces and album, which adds to the educational and artistic uses.
+and musical examples, pieces and albums, which adds to the educational and artistic uses.
 The fundamental element, the musical note with duration, volume, pitch and timbre,
-is related quantitatively to the characteristics of the digital signal.
-Internal variations, such as tremolos, vibratos and spectral fluctuations,
+is related quantitatively to characteristics of the digital signal.
+Internal variations of a note, such as tremolos, vibratos and spectral fluctuations,
 are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities.
-With this representation of notes, resources are provided for the generation of further musical structures,
+With this representation of notes, resources are provided for the generation of higher scale musical structures,
 such as rhythmic meter, pitch intervals and cycles.
-Such a framework enables precise and trustful scientific experiments and is useful for education and art.
+This framework enables precise and trustful scientific experiments and is useful for education and art.
 The efficacy of \massa\ is confirmed by the synthesis of small musical pieces using basic notes,
-incremented notes and notes in music, which reflects the organization of the toolbox and of this article.
-It is possible to synthesize whole albums through collage of the scripts and parameterization specified by the user.
-With the paradigm of open source implementation, the toolbox can be promptly scrutinized, expanded in co-authorship processes and used freely by musicians, engineers and other interested parties.
+elaborated notes and notes in music, which reflects the organization of the toolbox and of this article.
+It is possible to synthesize whole albums through collage of the scripts and settings specified by the user.
+With the open source paradigm, the toolbox can be promptly scrutinized,
+expanded in co-authorship processes and used with freedom by musicians, engineers and other interested parties.
 In fact, \massa\ has already been employed by external users for diverse purposes which include music production,

### 2017-08-08 article/acm/body.tex
finished final review until the beginning of spatialization

@@
 
-\section{Characterization of the musical note in dicrete-time audio} \label{sec:discNote}\label{sec:notaDisc}
-In diverse artistic and theoretical contexts, music is conceived as comprising fundamental units referred to as notes, ``atoms'' that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
-These notes are understood as central elements of certain musical paradigms.
-In a cognitive perspective, the notes are seen as discretized elements that facilitate and enrich the flow of information through music~\cite{Roederer, Lacerda}.
-Canonically, the basic characteristics of a musical note are duration, loudness, pitch and timbre~\cite{Lacerda}, which can be quantified using the evenly time spaced sonic samples.
-
-All the relations described in this section are implemented in the file \emph{eqs2.1.py} of the \massa\ toolbox.
+\section{Characterization of the musical note in discrete-time audio} \label{sec:discNote}\label{sec:notaDisc}
+In diverse artistic and theoretical contexts, music is conceived as constituted by fundamental units referred to as notes, ``atoms'' that constitute music itself~\cite{Wisnick, Lovelock, Webern}.
+In a cognitive perspective, notes are seen as discernible elements that facilitate and enrich the transmission of information through music~\cite{Roederer, Lacerda}.
+Canonically, the basic characteristics of a musical note are duration, loudness, pitch and timbre~\cite{Lacerda}.
+All relations described in this section are implemented in the file \texttt{src/sections/eqs2.1.py}.
 The musical pieces \emph{5 sonic portraits} and \emph{reduced-fi} are also available online to corroborate the concepts.
@@
 \subsection{Duration}
-The sample frequency $f_s$ is defined as the number of samples in each second of the discrete-time signal. Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
-A musical note of duration $\Delta$ seconds can be expressed as a sequence $T_i^{\Delta}$ of $\lfloor \Delta . f_s \rfloor$ samples.
-That is, the integer part of the multiplication is considered, and an error of at most $\delta_s$ missing seconds is admitted, which is usually fine for musical purposes.
-As an example, if $f_s=44.1kHz \;\;\Rightarrow\;\;\delta_s=\frac{1}{44100}\approx 23$ microseconds.
-Thus:
+The sample frequency $f_s$ is defined as the number of samples in each second of the discrete-time signal.
+Let $T=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds ($f_s=44.1kHz \Rightarrow \delta_s=1/44100\approx 23ms$).
+A musical note of duration $\Delta$ seconds can be expressed as a sequence $T^{\Delta}$ with $\Lambda = \lfloor \Delta . f_s \rfloor$ samples.
+That is, the integer part of the multiplication is considered, and an error of at most $\delta_s$ missing seconds is admitted, which is usually fine for musical purposes. Thus:
 
 \begin{equation}\label{eq:dur}
-T_{i}^{\Delta}={\{t_i\}}_{i=0}^{\lfloor \Delta . f_s \rfloor -1}
-\end{equation}
-
-With $\Lambda=\lfloor \Delta . f_s \rfloor$, the number of samples in the sequence, a more condensed notation is $T_i^\Delta=\{t_i\}_0^{\Lambda-1}$.
+T^{\Delta}={\{t_i\}}_{i=0}^{\lfloor \Delta . f_s \rfloor -1} = \{t_i\}_0^{\Lambda-1}
+\end{equation}
 
 \subsection{Loudness}\label{subsec:volume}
-The sensation of loudness depends on reverberation and spectrum, among other characteristics described in Section~\ref{sec:varInternas}.
+Loudness\footnote{Loudness and ``volume'' are often used indistinctly.
+In technical contexts, loudness is used for the subjective sensation of sound intensity
+while volume might be used for some measurement of loudness or to a change
+in the signal by equipment.
+Accordingly, one can perceive a sound as loud or soft and change the volume
+by turning a knob.
+We will use the term loudness and avoid the more ambiguous term volume.} is a sensation that depends on reverberation,
+spectrum and other characteristics described in Section~\ref{sec:varInternas}.
 One can achieve loudness variations through the power of the wave~\cite{Chowning}:
@@
 \begin{equation}\label{eq:potencia}
-pow(T_i)=\frac{\sum_{i=0}^{\Lambda -1} t_i^2}{\Lambda}
+pow(T)=\frac{\sum_{i=0}^{\Lambda -1} t_i^2}{\Lambda}
 \end{equation}
@@
 Thus, what matters is the relative power of a note in relation to the others around it, or the power of a music section in relation to the rest.
-Differences in loudness are the result of complex psychophysical phenomena but can often be reasoned about in terms of decibels, calculated directly from the amplitudes through energy or potency:
+Differences in loudness are the result of complex psychophysical phenomena but can often be reasoned about in terms of decibels, calculated directly from the amplitudes through energy or power:
 
 \begin{equation}\label{decibels}
-V_{dB}=10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)}
+V_{dB}=10log_{10}\frac{pow(T^{'})}{pow(T)}
 \end{equation}
@@
 The quantity $V_{dB}$ has the decibel unit ($dB$). 
-For each $10dB$ a ``doubled volume'' is associated.
+By standard, a ``doubled loudness'' is associated to a gain of $10dB$ (10 violins yield the double of the loudness of a violin).
 A handy reference is $10dB$ for each step in the musical intensity scale: \emph{pianissimo}, \emph{piano}, \emph{mezzoforte}, \emph{forte} and \emph{fortissimo}.
-Other useful references are $dB$ values related to double amplitude or potency:
+Other useful references are $dB$ values related to double amplitude or power:
 
 \begin{align}
-t_i^{'}=2 t_i \Rightarrow pot(T^{'}_i)=4 pot(T_i) \Rightarrow V^{'}_{dB}=10log_{10} 4 \approx 6 dB \label{eq:ampVol}\\
-t_i^{'}=\sqrt{2} t_i \Rightarrow pot(T^{'}_i)=2 pot(T_i) \Rightarrow V^{'}_{dB}=10log_{10} 2 \approx 3 dB\label{eq:potVol}
+t_i^{'}=2 t_i \Rightarrow pow(T^{'})=4 pow(T) \Rightarrow V^{'}_{dB}=10log_{10} 4 \approx 6 dB \label{eq:ampVol}\\
+t_i^{'}=\sqrt{2} t_i \Rightarrow pow(T^{'})=2 pow(T) \Rightarrow V^{'}_{dB}=10log_{10} 2 \approx 3 dB\label{eq:potVol}
 \end{align}
 
-\noindent and the amplitude gain for a sequence whose volume has been doubled ($10dB$):
+\noindent and the amplitude gain for a sequence whose loudness has been doubled ($10dB$):
 
@@
 \begin{split}
-10log_{10}\frac{pot(T^{'}_i)}{pot(T_i)} = 10 \quad \Rightarrow \\ \Rightarrow \quad \sum_{i=0}^{\lfloor \Delta.f_a \rfloor -1}t^{'2}_i=10\sum_{i=0}^{\Lambda-1}t_i^2=\sum_{i=0}^{\Lambda-1}(\sqrt{10}.t_i)^2 \\
+10log_{10}\frac{pot(T^{'})}{pot(T)} = 10 \quad \Rightarrow \\ \Rightarrow \quad \sum_{i=0}^{\lfloor \Delta.f_s \rfloor -1}t^{'2}_i=10\sum_{i=0}^{\Lambda-1}t_i^2=\sum_{i=0}^{\Lambda-1}(\sqrt{10}.t_i)^2 \\
 \therefore \quad t^{'}_i=\sqrt{10}t_i \quad \Rightarrow \quad t^{'}_i \approx 3.16t_i
@@
 Thus, an amplitude increase by a factor slightly above 3 is required for achieving a doubled loudness.
-These values are guides for increasing or decreasing the absolute values in the sample sequences with musical purposes.
+These values are guides for increasing or decreasing the absolute values in sample sequences.
 The conversion from decibels to amplitude gain (or attenuation) is straightforward:
@@
 \subsection{Pitch}
-To a note corresponds a sequence $T_i$ in which duration and loudness are directly related to the size of the sequence and the amplitude of its samples, respectively.
-The pitch is specified by the fundamental frequency $f_0$ whose cycle has duration $\delta_{f_0}=1/f_0$.
-This duration, multiplied by the sampling frequency $f_s$, yields in the number of samples per cycle: $\lambda_{f_0}=f_s .
-\delta_{f_0} =f_a/f_0$.
-
-For didactic reasons, let $f_0$ be such that it divides $f_s$ and $\lambda_{f_0}$ results integer. If $T_i^f$ is a sonic sequence with fundamental frequency $f$, then:
+The pitch is specified by the fundamental frequency $f$ whose cycle has duration $\delta=1/f$.
+This duration, multiplied by the sampling frequency $f_s$, yields in the number of samples per cycle: $\lambda=f_s . \delta =f_s/f$.
+For didactic reasons, let $f$ divide $f_s$ and result $\lambda$ integer. If $T^f$ is a sonic sequence with fundamental frequency $f$, then:
 
 \begin{equation}\label{periodicidade}
-     T^f_i=\left\{ t_i^f \right\}=\left\{ t^f_{i+\lambda_{f}}  \right\}= \left\{ t^f_{i+\frac{f_a}{f}} \right\}
+     T^f=\left\{ t_i^f \right\}=\left\{ t^f_{i+\lambda}  \right\}= \left\{ t^f_{i+\frac{f_s}{f}} \right\}
 \end{equation}
@@
 \subsection{Timbre}
-In a sound with harmonic spectrum, there is a (wave) period which corresponds to the (wave) cycle duration and is given by the inverse of the fundamental frequency.
-The trajectory of the wave inside the period - called the waveform - defines a harmonic spectrum and, thus, entails a timbre\footnote{The timbre of a sound is a subjective and complex characteristic.
-Physically, the timbre is multidimensional and given by the temporal dynamics of energy in the spectral components that are harmonic or noisy.
+A spectrum is said harmonic if all the frequencies it contains are (whole number) multiples of the fundamental frequency.
+A sound with harmonic spectrum has a wave period which corresponds to the wave cycle duration and is given by the inverse of the fundamental frequency.
+The trajectory of the wave inside the period is the \emph{waveform} and defines the harmonic spectrum.
+The waveform thus entails a timbre\footnote{The timbre of a sound is a subjective and complex characteristic.
+The timbre can be considered by a temporal evolution of energy in the spectral components that are harmonic or noisy (and by deviations of the harmonics from the ideal harmonic spectrum).
 In addition, the word timbre is used to designate different things: one same note can have (be produced with) different timbres, an instrument has different timbres, two instruments of the same family have, at the same time, the same timbre that blends them in the same family, and different timbres as they are different instruments.
-It is worth mentioning that timbre is not always manifested in spectral traces, since cultural or circumstantial aspects alter our perception of timbre}.
-Sonic spectra with minimum differences can result in timbres with crucial differences and, consequently, distinct timbres can be produced using different spectra\cite{Roederer}.
-
-The simplest case is the spectrum with only one (fundamental) frequency $f$, which is a sinusoid, often regarded as a ``pure'' oscillation in terms of the \emph{simple harmonic motion}.
-Let $S_i^f$ be a sequence whose samples $s_i^f$ describe a sinusoid with frequency $f$:
+Timbre is not only about spectrum: culture and context alter our perception of timbre.~\cite{Roederer}}.
+Sonic spectra with minimum differences can result in timbres with crucial differences and, consequently, distinct timbres can be produced using different spectra~\cite{Roederer}.
+
+The simplest case is the spectrum with only one frequency, which is a sinusoid, often regarded as a ``pure'' oscillation (e.g. in terms of the \emph{simple harmonic motion}).
+Let $S^f$ be a sequence whose samples $s_i^f$ describe a sinusoid with frequency $f$:
 
 \begin{equation}\label{sinusoid}
-     S^f_i=\{ s^f_i \}=\Bigl\{ \sin\bigl(2\pi \frac{i}{\lambda_f} \bigr)  \Bigr\} = \Bigl\{ \sin\bigl(2\pi f \frac{i}{f_s}\bigr)  \Bigr\} 
+     S^f=\{ s^f_i \}=\Bigl\{ \sin\bigl(2\pi \frac{i}{\lambda_f} \bigr)  \Bigr\} = \Bigl\{ \sin\bigl(2\pi f \frac{i}{f_s}\bigr)  \Bigr\} 
 \end{equation}
@@
 In a similar fashion, other waveforms are used in music for their spectral qualities and simplicity.
-While the sinusoid is an isolated node in the spectrum, any other wave present a succession of harmonic components.
+While the sinusoid is an isolated node in the spectrum, any other waveform present a succession of harmonic components.
 The most basic waveforms are specified in Equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square}, and are illustrated in Figure~\ref{fig:formasDeOnda}.
-These artificial waveforms are traditionally used in music for synthesis and oscillatory control of variables. They are also useful outside musical contexts\cite{Openheim}.
-
-The sawtooth presents all components of the harmonic series with decreasing energy of $-6dB/octave$\footnote{In 
-musical jargon, an ``octave'' $\{f,2f\}$ means a frequency $f$ and twice such frequency $2f$.}. The sequence of temporal samples can be described as:
-
-\begin{equation}\label{sawTooth}
-     D^f_i=\left\{ d^f_i \right\}=\left\{ 2\frac{i\,\%\lambda_f}{\lambda_f} -1 \right\}
-\end{equation}
-
-The triangular waveform has only odd harmonics falling with $-12dB/octave$:
-\begin{equation}\label{triangular}
-     T^f_i=\left\{ t^f_i \right\}=\left\{1- \left| 2 - 4\frac{i\,\%\lambda_f}{\lambda_f} \right| \right\}
-\end{equation}
-
-The square wave preserves only odd harmonics falling at $-6dB/octave$:
-\begin{equation}\label{square}
-     Q^f_i=\left\{ q^f_i \right\}= \left\{
-         \begin{array}{l l}
-              1 & \quad \text{for } \; \; (i\,\%\lambda_f)   <  \lambda_f /2  \\
-              -1 & \quad \text{otherwise}\\
-         \end{array} \right.
-\end{equation}
-
-The square wave can be used in a subtractive synthesis with the purposes of mimicking a clarinet.
-This instrument has only the odd harmonic components and the square wave is convenient with its abundant energy at high frequencies.
-The sawtooth is a common starting point for a subtractive synthesis, because it has both odd and even harmonics with high energy.
-In general, these waveforms are appreciated as excessively rich in sharp harmonics, and attenuator filtering on treble and middle parts of the spectrum is especially useful for reaching a more natural and pleasant sound.
-The relatively attenuated harmonics of the triangle wave makes it the more functional - among the listed ones - to be used in the synthesis of musical notes without any further processing.
-The sinusoid is often a nice choice, but a problematic one.
-While pleasant if not loud in a very high pitch (above $\approx 500Hz$ it requires careful dosage), the pitch of a pure sinusoid is not accurately detected by the human auditory system, particularly at low frequencies.
-Also, it requires a large amplitude to give a sinusoid a level of loudness if compared to other waveforms.
-Both particularities are understood in the scientific literature as a consequence of the nonexistence of pure sinusoidal sounds in nature~\cite{Roederer}.
+These artificial waveforms are traditionally used in music for synthesis and oscillatory control of variables. They are also useful outside musical contexts~\cite{Openheim}.
 
@@
 
+The sawtooth presents all components of the harmonic series with decreasing energy of $-6dB/octave$\footnote{In 
+musical jargon, an ``octave'' means a frequency $f$ and twice such frequency $2f$ or the bandwidth $\{f,2f\}$.}. The sequence of temporal samples can be described as:
+
+\begin{equation}\label{sawTooth}
+     D^f=\left\{ d^f_i \right\}=\left\{ 2\frac{i\,\%\lambda_f}{\lambda_f} -1 \right\}
+\end{equation}
+
+The triangular waveform has only odd harmonics falling with $-12dB/octave$:
+\begin{equation}\label{triangular}
+     T^f=\left\{ t^f_i \right\}=\left\{1- \left| 2 - 4\frac{i\,\%\lambda_f}{\lambda_f} \right| \right\}
+\end{equation}
+
+The square wave also has only odd harmonics but falling at $-6dB/octave$:
+\begin{equation}\label{square}
+     Q^f=\left\{ q^f_i \right\}= \left\{
+         \begin{array}{l l}
+              1 & \quad \text{for } \; \; (i\,\%\lambda_f)   <  \lambda_f /2  \\
+              -1 & \quad \text{otherwise}\\
+         \end{array} \right.
+\end{equation}
+
+The square wave can be used in a subtractive synthesis with the purpose of mimicking a clarinet.
+This instrument has only the odd harmonics and the square wave is convenient with its abundant energy at high frequencies.
+The sawtooth is a common starting point for subtractive synthesis, because it has both odd and even harmonics with high energy.
+In general, these waveforms are appreciated as excessively rich in sharp harmonics, and attenuation by filtering on treble and middle parts of the spectrum is especially useful for reaching a more natural and pleasant sound.
+The relatively attenuated harmonics of the triangle wave makes it the more functional - among the listed ones - to be used in the synthesis of musical notes without any further processing.
+The sinusoid is often a nice choice, but a problematic one.
+While pleasant if not loud in a very high pitch (above $\approx 500Hz$ it requires careful dosage), the pitch of a pure sinusoid is not accurately detected by the human auditory system, particularly at low frequencies.
+Also, it requires a great amplitude gain for an increase in loudness of a sinusoid if compared to other waveforms.
+Both particularities are understood in the scientific literature as a consequence of the nonexistence of pure sinusoidal sounds in nature~\cite{Roederer}.
+
 Figure~\ref{fig:formasDeOnda} presents the waveforms described in Equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} for $\lambda_f=100$ (i.e.
 period of $100$ samples).
-If $f_s=44.1kHz$, the PCM standard for Compact Disks, the wave has fundamental frequency $f=\frac{f_a}{\lambda_f}=\frac{44100}{100} = 441 \; Hz$, around A4, just above the central "C", whatever the waveform is.
+If $f_s=44.1kHz$, the PCM standard for Compact Disks, the wave has fundamental frequency $f=\frac{f_s}{\lambda_f}=\frac{44100}{100} = 441 \; Hz$, around A4, just above the central "C", whatever the waveform is.
 
@@
         \includegraphics[width=.7\textwidth]{figures/waveSpectrum_}
-    \caption{Spectrum of basic artificial musical waveforms.}
+    \caption{Spectra of basic artificial musical waveforms.}
         \label{fig:espectroDeOndas}
@@
 Furthermore, the notion of absolute pitch relies on the similarity of the spectrum to the harmonic series~\cite{Roederer}.
-For a fixed length period and waveform, the spectrum is perfectly harmonic and static, each waveform entails different energy in the (same) harmonic components.
+For a fixed length period and waveform, the spectrum is perfectly harmonic and static.
+Each waveform has a different distribution of energy in the (same) harmonic components.
 High curvatures are a sign of energy in the high harmonics.
-Figure~\ref{fig:formasDeOnda} depicts a wave, labeled as ``sampled real sound'', with a period of $\Lambda_f=114$ samples, extracted from a relatively well behaved recorded sound.
+Figure~\ref{fig:formasDeOnda} depicts a wave, labeled as ``sampled real sound'', with a period of $\Lambda_f=114$ samples.
 The oboe waveform was also sampled at $44.1kHz$.
@@
 
-The sequence $R_i=\{ r_i \}_0^{\lambda_f-1}$ of samples in the real sound of Figure~\ref{fig:formasDeOnda} can be taken as basis for a sound $T_i^f$ in the following way:
+The sequence $R=\{ r_i \}_0^{\lambda_f-1}$ of samples in the real sound of Figure~\ref{fig:formasDeOnda} can be taken as basis for a sound $T^f$ in the following way:
 
 \begin{equation}\label{sampleandoFormaDeOnda}
-     T^f_i=\{ t_i^f \}=\Bigl\{ r_{(i\,\%\lambda_{f})} \Bigr\}
+     T^f=\{ t_i^f \}=\Bigl\{ r_{(i\,\%\lambda_{f})} \Bigr\}
 \end{equation}
@@
 The resulting sound has the spectrum of the original waveform.
-As a consequence of its repetition in an identical form, the spectrum is perfectly harmonic, without noise or variations typical of natural phenomena.
-This can be observed in Figure~\ref{fig:espectroOboe}, which shows the spectrum of the original oboe note and a note with the same duration, whose samples consists of the repetition of cycle of Figure~\ref{fig:formasDeOnda}.
-Summing up, the natural spectrum exhibits variations in the frequencies of the harmonics, in their intensities and some noise, while the note made from the sampled period has a perfectly harmonic and static spectrum.
+As a consequence of the identical repetitions, the spectrum is perfectly harmonic, without noise or variations typical of natural phenomena.
+This can be observed in Figure~\ref{fig:espectroOboe}, which shows the spectrum of the original oboe note and a note with the same duration, whose samples consist of the repetition of the cycle on Figure~\ref{fig:formasDeOnda}.
+Summing up, the spectrum of a naturally produced note (e.g.
+by a musical instrument) exhibits variations in the frequencies of the harmonics, in their intensities and noise, while a note made from a sampled period has a perfectly harmonic and static spectrum.
 
@@
         \includegraphics[width=.7\textwidth]{figures/oboeNaturalSampledSpectrum_}
-	\caption{Spectrum of the sonic waves of a natural oboe note and obtained through a sampled period.
+	\caption{Spectra of the sonic waves of a natural oboe note and obtained through a sampled period.
 The natural sound has fluctuations in the harmonics and in its noise, while the sampled period note has a perfectly harmonic (and static) spectrum.}
@@
 These sinusoidal components in the discretized sound have some particularities.
-Considering a signal $T_i$ and its corresponding Fourier decomposition $\mathcal{F}\langle T_i\rangle=C_i=\{c_i\}_0^{\Lambda-1}$, the recomposition is the sum of the frequency components as time samples\footnote{The factor $\frac{1}{\Lambda}$ can be distributed among the Fourier transform and its reconstruction, as preferred.}:
+Considering a signal $T$ and its corresponding Fourier decomposition $\mathcal{F}\langle T\rangle=C=\{c_i\}_0^{\Lambda-1}=\left\{ \sum_{k=0}^{\Lambda-1}t_ke^{-j2\pi ki/\Lambda} \right\}_0^{\Lambda-1}$, the recomposition is the sum of the frequency components to yield the temporal samples\footnote{The factor $\frac{1}{\Lambda}$ can be distributed among the Fourier transform and its reconstruction, as preferred.
+Note that $j$ here is the imaginary unit $j^2=-1$.}:
 
@@
 \noindent where $c_k = a_k + j .
-b_k$ defines the amplitude and phase of each frequency: $w_k=\frac{2\pi}{\Lambda}k$ in radians or $f_k=w_k\frac{f_a}{2\pi}=\frac{f_a}{\Lambda}k$ in Hertz, taking into account the respective limits in $\pi$ and in $\frac{f_a}{2}$ given by the Nyquist Theorem.
-$j$ is the complex number, with $j^2=-1$.
+b_k$ defines the amplitude and phase of each frequency: $w_k=\frac{2\pi}{\Lambda}k$ in radians or $f_k=w_k\frac{f_s}{2\pi}=\frac{f_s}{\Lambda}k$ in Hertz, taking into account the respective limits of $\pi$ and $\frac{f_s}{2}$ given by the Nyquist Theorem.
 
@@
 t_i& = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}\left[a_k cos(w_k i) -b_k sen(w_k i)\right] \\
-   & = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]
-\end{split}
-\end{equation}
-
-Equation~\ref{moduloEfase} shows how the imaginary term of $c_k$ adds a phase to the real sinusoid: the terms $b_k$ enable the phase sweep $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right]$ given by $tg^{-1}\left(\frac{b_k}{a_k}\right)$ which has this image.
-The terms $a_k$ specify the right or left side of the trigonometric circle, which completes the phase domain: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.
+   & = \frac{1}{\Lambda}\sum_{k=0}^{\Lambda-1}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - arctan\left(\frac{b_k}{a_k}\right)\right]
+\end{split}
+\end{equation}
+
+Equation~\ref{moduloEfase} shows how the imaginary term of $c_k$ adds a phase to the real sinusoid: the terms $b_k$ enable the phase sweep $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right]$ given by $arctan\left(\frac{b_k}{a_k}\right)$ which has this image.
+The sign of $a_k$ specify the right or left side of the trigonometric circle, which completes the phase domain: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.
 
@@
          \includegraphics[width=.7\textwidth]{figures/amostras2c___}
-     \caption{Oscillation of 2 samples (maximum frequency for any $f_a$).
+     \caption{Oscillation of 2 samples (maximum frequency for any $f_s$).
 The first coefficient reflects a constant detachment (called \emph{offset} or \emph{bias}) and the second coefficient specifies the oscillation amplitude.}
@@
 
-Figure~\ref{fig:amostras2} shows two samples and their spectral components, where the Fourier decomposition has one unique pair of coefficients $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relative to frequencies $\{f_k\}_0^1=\left\{w_k\frac{f_a}{2\pi}\right\}_0^1=\left\{k\frac{f_a}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_a}{2}=f_{\text{max}}\right\}$
+Figure~\ref{fig:amostras2} shows two samples and their spectral component.
+When there is only two samples, the Fourier decomposition has only one pair of coefficients $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relative to frequencies $\{f_k\}_0^1=\left\{w_k\frac{f_s}{2\pi}\right\}_0^1=\left\{k\frac{f_s}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_s}{2}=f_{\text{max}}\right\}$
 with energies $e_k=\frac{(c_k)^2}{\Lambda=2}$.
-The role of amplitudes $a_k$ is clearly observed with $\frac{a_0}{2}$, the fixed offset (also called \emph{bias}), and $\frac{a_1}{2}$, oscillation with frequency given by $f_k=k \frac{f_a}{\Lambda=2}$.
+The role of amplitudes $a_k$ is clearly observed with $\frac{a_0}{2}$, the fixed offset (also called \emph{bias}), and $\frac{a_1}{2}$ for the oscillation with frequency $f_1=\frac{f_s}{\Lambda=2}$.
 This case has special relevance.
-At least 2 samples are necessary to represent an oscillation and it yields the Nyquist frequency $f_{\text{max}}=\frac{f_a}{2}$, which is the maximum frequency in a sound sampled with $f_a$ samples per second.
-In fact, any sampled signal has this property, not only the digitalized sound.
-
-All fixed sequences $T_i$ of only $3$ samples also have just $1$ frequency, since the first harmonic would have $1.5$ samples and exceeds the bottom limit of 2 samples, i.e.\ the frequency of the harmonic would exceed the Nyquist frequency:  $\; \frac{2.
-f_a}{3} > \frac{f_a}{2}$.
+At least 2 samples are necessary to represent an oscillation and it yields the Nyquist frequency $f_{\text{max}}=\frac{f_s}{2}$, which is the maximum frequency in a sound sampled with $f_s$ samples per second.
+In fact, any discrete-time signal has this property, not only digitized sound.
+
+All fixed sequences $T$ of only $3$ samples also have just $1$ frequency, since the first harmonic would have $1.5$ samples and exceeds the bottom limit of 2 samples, i.e.\ the frequency of the harmonic would exceed the Nyquist frequency:  $\; \frac{2.
+f_s}{3} > \frac{f_s}{2}$.
 The coefficients $\{c_k\}_0^{\Lambda-1=2}$ are present in 3 frequency components.
-One is relative to frequency zero ($c_0$), and the other two ($c_1$ and $c_2$) have the same role for reconstructing a sinusoid with $f=f_a/3$.
+One is relative to frequency zero ($c_0$), and the other two ($c_1$ and $c_2$) have the same role for reconstructing a sinusoid with $f=f_s/3$.
 
@@
 They are complex conjugates: $a_{k1}=a_{k2}$ and $b_{k1}=-b_{k2}$ and, as a consequence, the modules are equal and phases have opposite signs.
-Recalling that $f_k = k\frac{f_a}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $.
-When $k > \frac{\Lambda}{2}$, the frequency $f_k$ is mirrored by $\frac{f_a}{2}$ in this way: $f_k=\frac{f_a}{2} - (f_k-\frac{f_a}{2})=f_a-f_k=f_a - k\frac{f_a}{\Lambda}=(\Lambda-k)\frac{f_a}{\Lambda} \;\;\;\; \Rightarrow \;\;\;\; f_k\equiv f_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
-
-The same applies to $w_k=f_k.\frac{2\pi}{f_a}$ and the periodicity $2\pi$: it follows that $w_k=-w_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
+Recalling that $f_k = k\frac{f_s}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $.
+When $k > \frac{\Lambda}{2}$, the frequency $f_k$ is mirrored through $\frac{f_s}{2}$ in this way: $f_k=\frac{f_s}{2} - (f_k-\frac{f_s}{2})=f_s-f_k=f_s - k\frac{f_s}{\Lambda}=(\Lambda-k)\frac{f_s}{\Lambda} \;\;\;\; \Rightarrow \;\;\;\; f_k\equiv f_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
+
+The same applies to $w_k=f_k.\frac{2\pi}{f_s}$ and the periodicity $2\pi$: it follows that $w_k=-w_{\Lambda-k} \; ,\;\; \forall \;\; k<\Lambda$.
 Given the cosine (an even function) and the inverse tangent (an odd function), the components in $w_k$ and $w_{\Lambda-k}$ contribute with coefficients $c_k$ and $c_{\Lambda-k}$ in the reconstruction of the real samples.
 In other words, in a decomposition of $\Lambda$ samples, the $\Lambda$ frequency components $\{c_i\}_0^{\Lambda-1}$ are equivalent in pairs,
-except for $f_0$, and, when $\Lambda$ is even, for $f_{\Lambda/2}=f_{\text{max}}=\frac{f_a}{2}$.
+except for $f_0$, and, when $\Lambda$ is even, for $f_{\Lambda/2}=f_{\text{max}}=\frac{f_s}{2}$.
 Both components are isolated, i.e.\ there is one and only one component at frequency $f_0$ or $f_{\Lambda/2}$ (if $\Lambda$ is even).
-This assertion can be verified with $k=0$ and $k=\Lambda/2$ like this: $f_{\Lambda/2}=f_{(\Lambda-\Lambda/2) = \Lambda/2}$ and $f_0=f_{(\Lambda-0)=\Lambda}=f_0$.
-Furthermore, these two frequencies (zero and Nyquist frequency) do not have phase variation, their coefficients being strictly real. Therefore, the number $\tau$ of equivalent coefficient pairs is:
+In fact when $k=0$ or $k=\Lambda/2$ the mirror of the frequencies are themselves: $f_{\Lambda/2}=f_{(\Lambda-\Lambda/2) = \Lambda/2}$ and $f_0=f_{(\Lambda-0)=\Lambda}=f_0$.
+Furthermore, these two frequencies (zero and Nyquist frequency) do not have phase offset: their coefficients are strictly real. Therefore, the number $\tau$ of equivalent coefficient pairs is:
 
@@
 
-$T_i \; \Rightarrow \; a_k = a_{\Lambda -k}\;\;$ and $\;\;b_k = - b_{\Lambda -k}$, and thus:
+Real $T \; \Rightarrow \; a_k = a_{\Lambda -k}\;\;$ and $\;\;b_k = - b_{\Lambda -k}$, and thus:
 
@@
 
-To discuss the general case for components combination in each sample $t_i$, one can gather relations in Equation~\ref{moduloEfase} for the real signal reconstruction, relations of modules (Equation~\ref{equivalenciasModulos}) and phase equivalences (Equation~\ref{equivalenciasFases}), the number of paired coefficients (Equation~\ref{coefsPareados}), and equivalence of paired frequencies (Equation~\ref{equivalenciasFreqs}):
+To discuss the general case for components combination in each sample $t_i$, one can gather relations in Equation~\ref{moduloEfase} for the real signal reconstruction, equivalences of modules (Equation~\ref{equivalenciasModulos}) and phases(Equation~\ref{equivalenciasFases}), the number of paired coefficients (Equation~\ref{coefsPareados}), and the equivalence of paired frequencies (Equation~\ref{equivalenciasFreqs}):
 
@@
 \begin{split}
-	t_i = \frac{a_0}{\Lambda} + \frac{ a_{\Lambda/2}}{\Lambda}(1-\Lambda\% 2) + \frac{2}{\Lambda}\sum_{k=1}^{\tau}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - tg^{-1}\left(\frac{b_k}{a_k}\right)\right]
+	t_i = \frac{a_0}{\Lambda} + \frac{ a_{\Lambda/2}}{\Lambda}(1-\Lambda\% 2) + \frac{2}{\Lambda}\sum_{k=1}^{\tau}\sqrt{a_k^2 + b_k^2} \; cos\left[w_k i - arctan\left(\frac{b_k}{a_k}\right)\right]
 \end{split}
@@
 
-With 4 samples it is possible to represent 1 or 2 frequencies in any proportions (i.e.
-with independence).
-Figure~\ref{fig:amostras4} depicts the basic waveforms with 4 samples and their two (possible) components.
-The individual components sum to the original waveform and a brief inspection reveals the major curvatures resulting from the higher frequency, while the fixed offset is captured in the zero frequency component.
-
-\begin{figure*}
+\begin{figure*}[!h!]
     \centering
@@
 
-Figure~\ref{fig:formas4} shows the harmonics for the basic waveforms of Equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} for the case of 4 samples.
-There is only 1 sinusoid for each waveform, with the exception of the sawtooth, which has even harmonics.
-
-Figure~\ref{fig:amostras6} shows the sinusoidal decomposition for 6 samples, while Figure~\ref{fig:formas6} presents the decomposition of the basic wave forms.
-In this case, the waveforms have spectra with fundamental differences: square and triangular have the same components but with different proportions, while the sawtooth has an extra component.
-
-\begin{figure*}
+With 4 samples it is possible to represent 1 or 2 frequencies in any proportions (i.e.
+with independence).
+Figure~\ref{fig:amostras4} depicts the basic waveforms with 4 samples and their two (possible) components.
+The individual components sum to the original waveform and a brief inspection reveals the major curvatures resulting from the higher frequency, while the fixed offset is captured in the component with frequency $f_0=0$.
+
+\begin{figure*}[!h!]
     \centering
@@
 
+Figure~\ref{fig:formas4} shows the harmonics for the basic waveforms of Equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} for the case of 4 samples.
+There is only 1 sinusoid for each waveform, with the exception of the sawtooth, which has even harmonics.
+
+Figure~\ref{fig:amostras6} shows the sinusoidal decomposition for 6 samples, while Figure~\ref{fig:formas6} presents the decomposition of the basic wave forms.
+In this case, the waveforms have spectra with fundamental differences: square and triangular have the same components but with different proportions, while the sawtooth has an extra component.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{The basic note}\label{notaBasica}
-In a nutshell, let $f$ be such that it divides $f_a$\footnote{As mentioned before, this limitation simplifies the explanation for now and will be overcome in the next section.}.
-A sequence $T_i$ of sonic samples separated by $\delta_a=1/f_a$ expresses a musical note with a frequency of $f$ Hertz and $\Delta$ seconds of duration if, and only if, it has the periodicity $\lambda_f=f_a/f$ and size $\Lambda=\lfloor f_a .
+In a nutshell, let $f$ be such that it divides $f_s$\footnote{As mentioned before, this limitation simplifies the explanation for now and will be overcome in the next section.}.
+A sequence $T$ of sonic samples separated by $\delta_a=1/f_s$ expresses a musical note with a frequency of $f$ Hertz and $\Delta$ seconds of duration if, and only if, it has the periodicity $\lambda_f=f_s/f$ and size $\Lambda=\lfloor f_s .
 \Delta \rfloor$:
@@
 \begin{equation}\label{eq:notaBasica}
-T_i^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{\Lambda-1}= \left \{t^f_{i \; \% \left( \frac{f_a}{f} \right) } \right \}_0^{\Lambda-1}
+T^{f,\; \Delta}=\{t_{i \, \% \lambda_f} \}_0^{\Lambda-1}= \left \{t^f_{i \; \% \left( \frac{f_s}{f} \right) } \right \}_0^{\Lambda-1}
 \end{equation}
@@
 Nevertheless, it is necessary to choose a waveform for the samples $t_i$ to have a value.
-A unique period from the basic waveforms can be used to specify the note, where $\lambda_f=\frac{f_a}{f}$ is the number of samples at the period.
-Here, $L_i^{f,\, \delta_f} $ is the sequence that describes a period of the waveform $L_i^f \in \{S_i^f,Q_i^f,T_i^f,D_i^f,R_i^f \}$ with duration $\delta_f=1/f$ (as given by Equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square}) and $R_i^f$ is a sampled real waveform:
+A unique period from one of the basic waveforms can be used to specify the note, where $\lambda_f=\frac{f_s}{f}$ is the number of samples at the period.
+Let $L^{f,\, \delta_f} $ be the sequence that describes a period of the waveform $L^f \in \{S^f,Q^f,T^f,D^f,R^f \}$ with duration $\delta_f=1/f$ (as given by Equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square} and let $R_i^f$ be a sampled real waveform):
 
 \begin{equation}\label{periodoUnico}
-L_i^{f , \delta_f } = \left\{ l_i^f \right\}_0^{\delta_f . f_a -1}=\left\{ l_i^f \right\}_0^{\lambda_f-1}
+L^{f , \delta_f } = \left\{ l_i^f \right\}_0^{\delta_f . f_s -1}=\left\{ l_i^f \right\}_0^{\lambda_f-1}
 \end{equation}
@@
 \begin{equation}\label{eq:notaBasicaTimbre}
-T_i^{f,\; \Delta}=\left\{t_i^f\right\}_0^{\lfloor f_a . \Delta \rfloor -1}=\left \{ l^f_{i\,\%\left(\frac{f_a}{f}\right)} \right \}_0^{\Lambda-1}
-\end{equation}
-
-\subsection{Spatial localization and spatialization}\label{subsec:spac}
-A musical note always has a spatial localization, even though this is not one of its four basic properties in canonical musical theory.
-The note has a source which has a position at the ordinary physical three dimensional space.
-This position is the spatial localization of the sound and might not be (modeled as) a single point.
-The reverberation in the environment in which sound occurs is one of the main topics of \emph{spatialization}, i.e.
-the consideration of the influence the environment has on the sound that reaches a listener.
-Both concepts, spatialization and spatial localization, are widely valued by audiophiles and the music industry~\cite{floEsp}.
+T^{f,\; \Delta}=\left\{t_i^f\right\}_0^{\lfloor f_s . \Delta \rfloor -1}=\left \{ l^f_{i\,\%\left(\frac{f_s}{f}\right)} \right \}_0^{\Lambda-1}
+\end{equation}
+
+\subsection{Spatialization: localization and reverberation}\label{subsec:spac}
+A musical note is always produced within the three dimensional physical space.
+The consideration of this issue is the subject of the spatialization knowledge field and practice\footnote{By
+spatialization one might both find: 1) the consideration of cues in the sound that derive from the environment,
+including the localization of the listener and the sound source;
+2) techniques to diffuse audio through various sources, such as loudspeakers, singers and traditional musical instruments, for musical purposes.
+We focus in the first item although issues of the second item are also tackled.}.
+A note is always spatialized
+even though it is not one of its four basic properties in canonical musical theory
+A note has a source which has a position at the ordinary three dimensional physical space.
+This position is the spatial localization of the sound.
+It is often (modeled as) a single point but can be a surface or a volume.
+The reverberation in the environment in which a sound occurs is another one of the main topics of spatialization.
+Both concepts, spatial localization and reverberation,
+are widely valued by composers, audiophiles and the music industry~\cite{floEsp}.
 
@@
 It is a crucial clue to localization perception of bass sounds and percussive sonorities~\cite{Heeger}.
-With $\Lambda_{ITD}=\lfloor ITD . f_a \rfloor$:
+With $\Lambda_{ITD}=\lfloor ITD . f_s \rfloor$:
 
@@
 \begin{split}
-\Lambda_{ITD} & = \left \lfloor \frac{d'-d}{343,2}  f_a \right \rfloor \\
+\Lambda_{ITD} & = \left \lfloor \frac{d'-d}{343,2}  f_s \right \rfloor \\
 IID_a & = \frac{d}{d'} \\
@@
 Smith III states that usual concert rooms have total reverberation time of $\approx 1.9$ seconds, and that the period of first reflections is around $0.1s$.
-With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08m$ ($83.79k$ samples in $f_a=44.1kHz$) before reaching the ear.
-In addition, sound reflections made after propagation for $34.32$ m ($4.41k$ samples in $f_a=44.1kHz$) have incidences less distinct by hearing.
+With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08m$ ($83.79k$ samples in $f_s=44.1kHz$) before reaching the ear.
+In addition, sound reflections made after propagation for $34.32$ m ($4.41k$ samples in $f_s=44.1kHz$) have incidences less distinct by hearing.
 These first reflections are particularly important to spatial sensation.
 The first incidence is the direct sound, described by ITD and IID of Equations~\ref{eq:dti} and~\ref{eq:dii}.
-Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_a=44.1kHz$).
+Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_s=44.1kHz$).
 It is experimentally verifiable that the number of reflections increases with the square of $\approx k.n^2$.
@@
 Figure~\ref{fig:mixagem} illustrates this overlapping process of discretized sound waves, each with 100 samples.
-If $f_a=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_a}{100/2}=882Hz$, $\frac{f_a}{100/4}=1764Hz$ and $\frac{f_a}{100/5}=2205Hz$.
-The samples duration is very short $\frac{f_a=44.1kHz}{100} \approx 2ms$.
+If $f_s=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_s}{100/2}=882Hz$, $\frac{f_s}{100/4}=1764Hz$ and $\frac{f_s}{100/5}=2205Hz$.
+The samples duration is very short $\frac{f_s=44.1kHz}{100} \approx 2ms$.
 One can complete the sequence with zeroes to sum (mix) sequences with different sizes.
@@
 Although the sequences are short for the usual sample rates, it is easy to visually observe the concatenation of sound sequences.
-In addition, each note has  a duration larger than $100ms$ if $f_a<1kHz$.
+In addition, each note has  a duration larger than $100ms$ if $f_s<1kHz$.
 
@@
 The musical piece \emph{reduced-fi} explores the temporal juxtaposition of notes, resulting in a homophonic piece.
-The vertical principle is demonstrated at the \emph{sounic pictures}, static sounds with peculiar spectrum.
+The vertical principle is demonstrated at the \emph{sonic portraits}, static sounds with peculiar spectrum.
 Both pieces were written in Python and are available as part of the \massa\ \emph{toolbox}.\cite{MASSA}
 
-With the basic musical digital note carefully described, the next section develops the temporal evolution of its contents as in \emph{glissandi} and volume envelopes.
+With the basic musical digital note carefully described, the next section develops the temporal evolution of its contents as in \emph{glissandi} and intensity envelopes.
 Filtering of spectral components and noise generation complements the musical note as a self-contained unity.
@@
 \section{Variation in the basic note}\label{sec:internalVar}\label{sec:varInternas}
-The basic digital music note defined in Section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (volume) and timbre.
+The basic digital music note defined in Section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (loudness) and timbre.
 This is a useful and paradigmatic model, but it does not exhaust all the aspects of a musical note.
@@
     \caption{Search in the \emph{lookup table} to synthesize sounds at different frequencies using a unique waveform with high resolution.
-	Each \emph{i-th} sample $t_i^f$ of a sound with frequency $f$ is related to the samples in the table $\widetilde{l}$\, by the relation $t_i^{f}=\widetilde{l}_{\left\lfloor i.f\frac{\widetilde{\Lambda}}{f_a} \right\rfloor \%\,\widetilde{\Lambda}}$ where $\widetilde{\Lambda}$ is the number of sample in $\widetilde{l}$ and $f_a$ is the sampling rate.}
+	Each \emph{i-th} sample $t_i^f$ of a sound with frequency $f$ is related to the samples in the table $\widetilde{l}$\, by the relation $t_i^{f}=\widetilde{l}_{\left\lfloor i.f\frac{\widetilde{\Lambda}}{f_s} \right\rfloor \%\,\widetilde{\Lambda}}$ where $\widetilde{\Lambda}$ is the number of sample in $\widetilde{l}$ and $f_s$ is the sampling rate.}
         \label{fig:lut}
@@
 \begin{split}
-T_i^{f,\,\Delta}=\left\{t_i^f\right\}_0^{\lfloor \, f_a .
-\Delta \, \rfloor -1} = \left\{ \, \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} }\, \right\}_{0}^{\Lambda-1}\; , \quad \text{where} \;\; \gamma_i = \left \lfloor i f \frac{ \widetilde{\Lambda}}{f_a} \right \rfloor
+T_i^{f,\,\Delta}=\left\{t_i^f\right\}_0^{\lfloor \, f_s .
+\Delta \, \rfloor -1} = \left\{ \, \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} }\, \right\}_{0}^{\Lambda-1}\; , \quad \text{where} \;\; \gamma_i = \left \lfloor i f \frac{ \widetilde{\Lambda}}{f_s} \right \rfloor
 \end{split}
@@
 The expression defining the variable $\gamma_i$ can be understood as $f_s$ being added to $i$ at each second.
-If $i$ is divided by the sample frequency, $\frac{i}{f_a}$
+If $i$ is divided by the sample frequency, $\frac{i}{f_s}$
 is incremented by $1$ at each second.
-Multiplied by the period, it results in $i \frac{\widetilde{\Lambda}}{f_a}$, which covers the period in one second.
+Multiplied by the period, it results in $i \frac{\widetilde{\Lambda}}{f_s}$, which covers the period in one second.
 Finally, with frequency $f$ it results in $i .
-f \frac{\widetilde{\Lambda}}{f_a}$ which completes $f$ periods $\widetilde{\Lambda}$ in $1$ second, i.e.
+f \frac{\widetilde{\Lambda}}{f_s}$ which completes $f$ periods $\widetilde{\Lambda}$ in $1$ second, i.e.
 the resulting sequence presents the fundamental frequency $f$.
@@
 There are important considerations here: it is possible to use practically any frequency $f$.
-Limits exist only at low frequencies when the size of table $\widetilde{\Lambda}$ is not sufficient for the sample rate $f_a$.
+Limits exist only at low frequencies when the size of table $\widetilde{\Lambda}$ is not sufficient for the sample rate $f_s$.
 The lookup procedure is virtually costless and replaces calculations by simple indexed searches (what is generally understood as an optimization process).
@@
 
-Consider a note with duration $\Delta = \frac{\Lambda}{f_a}$, in which the frequency $f=f_i$ varies linearly from $f_0$ to $f_{\Lambda -1}$. Thus:
+Consider a note with duration $\Delta = \frac{\Lambda}{f_s}$, in which the frequency $f=f_i$ varies linearly from $f_0$ to $f_{\Lambda -1}$. Thus:
 
@@
 \begin{split}
-	\Delta_{\gamma_i}=\frac{\widetilde{\Lambda}}{f_a}f_i \quad \Rightarrow \quad \gamma_i & = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f_j \right \rfloor \\
-	\gamma_i & =  \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{j}{\Lambda-1} \right ] \right \rfloor 
+	\Delta_{\gamma_i}=\frac{\widetilde{\Lambda}}{f_s}f_i \quad \Rightarrow \quad \gamma_i & = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}f_j \right \rfloor \\
+	\gamma_i & =  \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s} \left [f_0 + (f_{\Lambda-1}-f_0)\frac{j}{\Lambda-1} \right ] \right \rfloor 
 \end{split}
@@
 
-\noindent where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_a}$ is the LUT increment between two samples given the sound frequency of the first sample.
+\noindent where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_s}$ is the LUT increment between two samples given the sound frequency of the first sample.
 Therefore, it is handy to calculate the elements $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ by means of the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$.
@@
 \begin{split}
- \Delta_{\gamma_i}= \frac{\widetilde{\Lambda}}{f_a}f_i \quad \Rightarrow  \quad \gamma_i= & \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_a}f_j \right \rfloor \\
-	\gamma_i = & \left \lfloor \sum_{j=0}^{i} f_0 \frac{\widetilde{\Lambda}}{f_a} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{j}{\Lambda -1}} \right \rfloor
+ \Delta_{\gamma_i}= \frac{\widetilde{\Lambda}}{f_s}f_i \quad \Rightarrow  \quad \gamma_i= & \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}f_j \right \rfloor \\
+	\gamma_i = & \left \lfloor \sum_{j=0}^{i} f_0 \frac{\widetilde{\Lambda}}{f_s} \left ( \frac{f_{\Lambda-1}}{f_0} \right ) ^{\frac{j}{\Lambda -1}} \right \rfloor
 \end{split}
@@
 The term $\frac{i}{\Lambda-1}$ covers the interval $[0,1]$ and it is possible to raise it to a power in such a way that the beginning of the transition will be smoother or steeper.
-This procedure is useful for energy variations with the purpose of changing the volume\footnote{The volume (psychophysical quality) is a consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is the wave total energy.
-Wave energy is easy to modify (see Equation~\ref{eq:potencia}) and can be varied in many ways.
+This procedure is useful for energy variations with the purpose of changing the loudness\footnote{The
+loudness (psychophysical quality) is a consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is the wave total energy.
+Wave energy is easy to modify (see Section~\ref{subsec:volume}) and can be varied in many ways.
 The simplest way consists of modifying the amplitude by multiplying the whole sequence by a real number.
-The increase of energy without amplitude variation is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}.}.
+The increase of energy without changing amplitude ambit is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}.}.
 It is sufficient to multiply the original sequence by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is the given coefficient and $a_{\Lambda-1}$ is a fraction of the original amplitude, which is the value to be reached at the end of the transition.
@@
 It is often convenient to have $a_0=1$ to start a new sequence with the original amplitude and then progressively change it.
-If $\alpha=1$, the amplitude variation follows the geometric progression that implies the linear variation of volume.
+If $\alpha=1$, the amplitude variation follows the geometric progression that implies the linear variation of loudness.
 Figure~\ref{fig:transicao} depicts transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by Equation~\ref{eq:ampVol}.
@@
 Both cases make the procedure useless, once a ratio of any number in relation to zero is not well defined for our purposes.
-It is possible to solve this dilemma choosing a number that is small enough like $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum volume for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$).
+It is possible to solve this dilemma choosing a number that is small enough like $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum loudness for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$).
 A linear fade can be used then to reach zero amplitude, if needed.
@@
 The greater the value of $\alpha$, the smoother the sound introduction and more intense its end.
-$\alpha>1$ results in volume transitions commonly called \emph{slow fade}, while $\alpha<1$ results in \emph{fast fade}\cite{guillaume}.
+$\alpha>1$ results in loudness transitions commonly called \emph{slow fade}, while $\alpha<1$ results in \emph{fast fade}\cite{guillaume}.
 
@@
 	f_{\text{min}} & \approx 15 Hz \\
-	f_i  &= i \frac{f_a}{\Lambda} \;, \;\; \quad i \;\leq\; \frac{\Lambda}{2},\;\; i\;\in\;\mathbb{N}  \\
+	f_i  &= i \frac{f_s}{\Lambda} \;, \;\; \quad i \;\leq\; \frac{\Lambda}{2},\;\; i\;\in\;\mathbb{N}  \\
 	\alpha_i & = \left(10^{-\frac{3}{20}}\right)^{\log _2 \left ( \frac{f_i}{f_{\text{min}}} \right )}  \\
@@
 \subsection{Tremolo and vibrato, AM and FM}\label{subsec:tvaf}
-Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in volume (intensity)\footnote{The jargon may be different in other contexts.
+Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in loudness (intensity)\footnote{The jargon may be different in other contexts.
 For example, in piano, tremolo is a vibrato in the classification used here.
@@
 The curvature is due to the logarithmic progression of the intensity.
-The tremolo frequency is $1.5Hz$ if $f_a=44.1kHz$ because $\text{duration} = \frac{i_{\text{max}}=82000}{f_a}= 2s \; \Rightarrow \; \frac{3\text{oscillations}}{2s}=1.5$ oscillations per second.
+The tremolo frequency is $1.5Hz$ if $f_s=44.1kHz$ because $\text{duration} = \frac{i_{\text{max}}=82000}{f_s}= 2s \; \Rightarrow \; \frac{3\text{oscillations}}{2s}=1.5$ oscillations per second.
 
@@
 At this point the musical possibilities are very wide.
-All characteristics, like pitch (given by frequency), timbre (achieved by waveforms, filters and noise), volume (manipulated by intensity) and duration (given by the number of samples), can be considered in an absolute form or varied during the duration of the sound (clearly, with the exception of duration itself).
+All characteristics, like pitch (given by frequency), timbre (achieved by waveforms, filters and noise), loudness (manipulated by intensity) and duration (given by the number of samples), can be considered in an absolute form or varied during the duration of the sound (clearly, with the exception of duration itself).
 The following musical usages comprehend a collection of possibilities with the purpose of exemplifying types of sound manipulations that result in musical material.
@@
 A sound can be altered through different filtering and then mixed to create an effect known as \emph{chorus}.
-Based on what happens in a choir of singers, the sound is synthesized using small and potentially arbitrary modifications of parameters like center frequency, presence (or absence) of vibrato or tremolo and its characteristics, equalization, volume, etc.
+Based on what happens in a choir of singers, the sound is synthesized using small and potentially arbitrary modifications of parameters like center frequency, presence (or absence) of vibrato or tremolo and its characteristics, equalization, loudness, etc.
 As a final result, those versions of the original sound are mixed together (see Equation~\ref{eq:mixagem}).
@@
 \subsubsection{ADSR envelopes}
-The variation of volume along the duration of sound is crucial to our timbre perception.
-The volume envelope known as ADSR (\emph{Attack-Decay-Sustain-Release}) has many implementations in both hardware and software synthesizers.
+The variation of loudness along the duration of sound is crucial to our timbre perception.
+The intensity envelope known as ADSR (\emph{Attack-Decay-Sustain-Release}) has many implementations in both hardware and software synthesizers.
 A pioneering implementation can be found in the Hammond Novachord synthesizer of 1938 and some variants are mentioned below~\cite{ADSR}.
-The canonical ADSR envelope is characterized by 4 parameters: attack duration (time in which sound reaches its maximum amplitude), decay duration (follows the attack immediately), level of sustained volume (in which the volume remains stable after the decay) and release duration (after sustained section, this is the duration needed for amplitude to reach zero or final value).
+The canonical ADSR envelope is characterized by 4 parameters: attack duration (time in which sound reaches its maximum amplitude), decay duration (follows the attack immediately), level of sustained intensity (in which the intensity remains stable after the decay) and release duration (after sustained section, this is the duration needed for amplitude to reach zero or final value).
 Note that the sustain duration is not specified because it is the difference between the duration itself and the sum of the attack, decay and sustain durations.
@@
 for the generation of appealing artistic materials. Some psychoacoustic effects were detected,
-which needs validation and should be reported, specially with~\cite{quadrosSonoros}.\footnote{The sonic pictures where sent to a public mailing list and the fifth sonic picture was reported by some individuals to induce a state in which noises from the own tongue, teeth and jaw of the individual echoed for some seconds~\cite{quadrosMetarec}.} Other foreseen advances
+which needs validation and should be reported, specially with~\cite{quadrosSonoros}.\footnote{The sonic portraits where sent to a public mailing list and the fifth piece was reported by some individuals to induce a state in which noises from the own tongue, teeth and jaw of the individual echoed for some seconds~\cite{quadrosMetarec}.} Other foreseen advances
 are: a JavaScript version of the toolbox, better hypermedia deliverables of this framework, user guides

### 2017-08-09 article/acm/body.tex
finished final review until the end of spatialization

@@
 A musical note is always produced within the three dimensional physical space.
-The consideration of this issue is the subject of the spatialization knowledge field and practice\footnote{By
-spatialization one might both find: 1) the consideration of cues in the sound that derive from the environment,
+The consideration of this fact is the subject of the spatialization knowledge field and practice\footnote{By
+spatialization one might find both:
+1) the consideration of cues in sound that derive from the environment,
 including the localization of the listener and the sound source;
-2) techniques to diffuse audio through various sources, such as loudspeakers, singers and traditional musical instruments, for musical purposes.
+2) techniques to produce sound through various sources, such as loudspeakers, singers and traditional musical instruments, for musical purposes.
 We focus in the first item although issues of the second item are also tackled.}.
 A note is always spatialized
-even though it is not one of its four basic properties in canonical musical theory
+even though it is not one of its four basic properties in canonical musical theory (duration, loudness, pitch and timbre).
 A note has a source which has a position at the ordinary three dimensional physical space.
@@
 \subsubsection{Spatial localization}
-It is understood that the perception of sound localization occurs in our nervous system by three pieces of information: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, specially its chest, head and ears~\cite{Roederer, hrtf, Heeger}.
+It is understood that the perception of sound localization occurs in our nervous system mainly by three cues: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, specially its chest, head and ears~\cite{Roederer, hrtf, Heeger}.
 
@@
 
-\noindent where $\zeta$ is the distance between ears $\zeta$, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in Interaural Time Difference:
+\noindent where $\zeta$ is the distance between the ears, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in Interaural Time Difference:
 
@@
 
-\noindent which, converted to amplitude, yields $IID_a=\frac{d}{d'}$.
-$IID_a$ can be used as a multiplicative constant to the right channel of a stereo sound signal: $\{t_i'\}_0^{\Lambda -1}=\{IID_a .
+$IID_a=\frac{d}{d'}$ can be used as a multiplicative constant to the right channel of a stereo sound signal: $\{t_i'\}_0^{\Lambda -1}=\{IID_a .
 t_i\}_0^{\Lambda -1}$, where $\{t_i'\}$ are samples of the wave incident in the left ear.
-It is possible to use IID together with ITD as a time advance for the right channel.
-It is a crucial clue to localization perception of bass sounds and percussive sonorities~\cite{Heeger}.
+It is possible to use IID together with ITD as a temporal advance for the right channel which is a crucial cue to localization perception of bass sounds and percussive sonorities~\cite{Heeger}.
 With $\Lambda_{ITD}=\lfloor ITD . f_s \rfloor$:
@@
 
-\noindent with $t_i$ as the right channel and $t_i'$ the left channel.
-If $\Lambda_{ITD} < 0 $, it is only needed to change $t_i$ by $t_i'$ and to use $\Lambda_{ITD}'= | \Lambda_{ITD} | $ and $IID_a'=1 / IID_a$.
+\noindent where $t_i$ is the right channel and $t_i'$ is the left channel.
+If $\Lambda_{ITD} < 0$, it is necessary to change $t_i$ by $t_i'$
+and to use $\Lambda_{ITD}'= | \Lambda_{ITD} | $ and $IID_a'=1 / IID_a$.
 
@@
 \noindent with $x,y$ as presented in Figure~\ref{fig:spac}.
-Henceforth, there are problems when $\theta$ falls within the so-called "cone of confusion": the same pair of ITD and IID results in a large number of points inside the cone.
-On those points the inference of the azimuthal angle depends especially on the attenuation filtering for high frequencies, since the head interferes much more in the treble than in bass waves~\cite{Heeger,hrtf}.
-Also relevant to the hearing of lateral sources is that low frequencies diffract and the wave arrives to the opposite ear with a delay of $\approx 0.7ms$.\cite{floEsp}
-
+Furthermore, there are problems when $\theta$ falls within the so-called "cone of confusion": the same pair of ITD and IID are related to a large number of points inside the cone~\footnote{The cone itself is not shown in Figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not found in the literature.
+Given the filtering and diffraction dependent on the sound spectrum, it is hard, if not impossible, to correctly draw the cone.
+Even so, the cone of confusion can be pictured as a cone with its apex placed in the middle of the head and growing out in the direction of the ear~\cite{hrtf}.}.
+On those points the inference of the azimuthal angle depends especially on the filtering of the frequencies by the head and torso~\cite{Heeger,hrtf}.
 Figure~\ref{fig:spac} depicts the acoustic shadow of the cranium, an important phenomenon to perception of source azimuthal angle in the cone of confusion.
-The cone itself is not shown in Figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not encountered in the literature.
-Given the filtering and diffraction dependent on the sound spectrum, it is hard, if not impossible, to correctly draw the confusion cone.
-Even so, the cone of confusion can be understood as a cone with its apex placed in the middle of the head and growing out in the direction of each ear~\cite{hrtf}.
-
-On the other hand, the complete localization, including height and distance of sound source, is given by the Head Related Transfer Function (HRTF)~\cite{hrtf}.
+Also relevant to the hearing of lateral sources is that low frequencies diffract and the wave arrives to the opposite ear with a delay of $\approx 0.7ms$~\cite{floEsp}.
+The complete localization, including height and distance of sound source, is given by the Head Related Transfer Function (HRTF)~\cite{hrtf}.
 There are well known open databases of HRTF, such as CIPIC, and it is possible to apply those transfer functions in a sonic signal by convolution (see Equation~\ref{eq:conv})~\cite{CIPIC}.
@@
 
-\subsubsection{Spatialization}
-Spatialization results from sound reflections and absorption by room/environment surface where the note is played.
+\subsubsection{Reverberation}
+The reverberation results from sound reflections and absorption by the environment (e.g. a room) surface where a sound occurs.
 The sound propagates through the air with a speed of $\approx 343.2m/s$ and can be emitted from a source with any directionality pattern.
-When a sound pulse encounters a surface there are: 1) inversion of the propagation speed component normal to the surface;  2) energy absorption, especially in trebles.
+When a sound pulse encounters a surface there are: 1) inversion of the propagation speed component normal to the surface;  2) energy absorption, especially in high frequencies.
 The sonic waves propagate until they reach inaudible levels (and further but then can often be neglected).
@@
 It is possible to simulate reverberations that are impossible in real systems.
-For example, it is possible to use asymmetric reflections with relation to the axis perpendicular to the surface, or to increase specific frequency bands (known as 'resonances'); neither of these characteristics are found in real systems.
-
-There are reverberation models less related to each independent reflection, exploring valuable information to the auditory system.
+For example, it is possible to use asymmetric reflections with relation to the axis perpendicular to the surface, or to increase the intensity of the last reflections.
+
+There are reverberation models less related to each independent reflection and that explores valuable cues to the auditory system.
 In fact, reverberation can be modeled with a set of two temporal and two spectral regions~\cite{JOSPhy}:
@@
    \item First band: the bass has some resonance bandwidths relatively spaced.
-   \item Second band: mid and treble have progressive decay and smooth statistical fluctuations.
+   \item Second band: mid and treble have a progressive decay and smooth statistical fluctuations.
 \end{itemize}
@@
 Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_s=44.1kHz$).
-It is experimentally verifiable that the number of reflections increases with the square of $\approx k.n^2$.
+Also, it is experimentally verifiable that the number of reflections increases with the square of time.
 A discussion about the use of convolutions and filtering to favor implementation of these phenomena is provided in Section~\ref{subsec:mus2}, particularly in the paragraphs about reverberation.

### 2017-08-10 article/acm/body.tex
revisado ateh o final do LUT

@@
 \subsection{Spatialization: localization and reverberation}\label{subsec:spac}
-A musical note is always produced within the three dimensional physical space.
-The consideration of this fact is the subject of the spatialization knowledge field and practice\footnote{By
+A sound is always produced within the three dimensional physical space.
+The consideration of this fact is the subject of spatialization\footnote{By
 spatialization one might find both:
@@
 2) techniques to produce sound through various sources, such as loudspeakers, singers and traditional musical instruments, for musical purposes.
-We focus in the first item although issues of the second item are also tackled.}.
+We focus in the first item although issues of the item are also tackled.}.
 A note is always spatialized
@@
 
-\noindent where $\zeta$ is the distance between the ears, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in Interaural Time Difference:
+\noindent where $\zeta$ is the distance between the ears, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in the Interaural Time Difference:
 
@@
 
-\noindent where $t_i$ is the right channel and $t_i'$ is the left channel.
+\noindent where $t_i$ is a right channel sample and $t_i'$ is a left channel sample.
 If $\Lambda_{ITD} < 0$, it is necessary to change $t_i$ by $t_i'$
@@
 
-Spatial localization depends considerably on other cues. By using only ITD and IID it is possible to specify solely the horizontal angle (azimuthal) $\theta$ given by:
+Spatial localization depends considerably on other cues. By using only ITD and IID it is possible to specify solely the horizontal (azimuthal) angle $\theta$ given by:
 
@@
 The sound propagates through the air with a speed of $\approx 343.2m/s$ and can be emitted from a source with any directionality pattern.
-When a sound pulse encounters a surface there are: 1) inversion of the propagation speed component normal to the surface;  2) energy absorption, especially in high frequencies.
+When a sound wave encounters a surface there are: 1) inversion of the propagation speed component normal to the surface;  2) energy absorption, especially in high frequencies.
 The sonic waves propagate until they reach inaudible levels (and further but then can often be neglected).
-As a sonic front reaches the human ear, it can be described as the original sound, with the last reflection point as the source, and the absorption filters of each surface it has reached.
+As a sonic front reaches the human ear, it can be described as an ordinary sound, with the last reflection point as the source, and the absorption filters of each surface it has reached.
 It is possible to simulate reverberations that are impossible in real systems.
-For example, it is possible to use asymmetric reflections with relation to the axis perpendicular to the surface, or to increase the intensity of the last reflections.
-
-There are reverberation models less related to each independent reflection and that explores valuable cues to the auditory system.
-In fact, reverberation can be modeled with a set of two temporal and two spectral regions~\cite{JOSPhy}:
+For example, one can use asymmetric reflections with relation to the axis perpendicular to the surface, or increase the intensity of the last reflections.
+
+There are reverberation models less based in each independent reflection and that explores valuable cues to the auditory system.
+In fact, a reverberation can be modeled with a set of two temporal and two spectral regions~\cite{JOSPhy}:
 \begin{itemize}
@@
 Smith III states that usual concert rooms have total reverberation time of $\approx 1.9$ seconds, and that the period of first reflections is around $0.1s$.
-With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08m$ ($83.79k$ samples in $f_s=44.1kHz$) before reaching the ear.
-In addition, sound reflections made after propagation for $34.32$ m ($4.41k$ samples in $f_s=44.1kHz$) have incidences less distinct by hearing.
-These first reflections are particularly important to spatial sensation.
+Under these conditions, there are perceived wave pulses which propagate for $652.08m$ ($83.79k$ samples in $f_s=44.1kHz$) before reaching the ear.
+In addition, reflections that reach the ear after propagating  for $34.32m$ ($4.41k$ samples in $f_s=44.1kHz$) or more have incidences less distinct by hearing.
+The first reflections are particularly important to spatial sensation.
 The first incidence is the direct sound, described by ITD and IID of Equations~\ref{eq:dti} and~\ref{eq:dii}.
-Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_s=44.1kHz$).
+Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between each reflection is $8-90ms$ ($\approx 350-4000$ samples in $f_s=44.1kHz$).
 Also, it is experimentally verifiable that the number of reflections increases with the square of time.
@@
 \subsection{Musical usages}\label{subsec:basMus}
-Once the basic note is defined, it is didactically convenient to build musical structures with sequences based on these particles.
-The sum of elements with same index of $N$ sequences $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ with same size $\Lambda$ results in the overlapped spectral contents of each sequence, in a process called mixing:
+The sum of the respective $ith$ samples of $N$ sequences with the same size $\Lambda$
+results in the overlapped spectral contents of each sequence, in a process called mixing:
 
@@
 
-Figure~\ref{fig:mixagem} illustrates this overlapping process of discretized sound waves, each with 100 samples.
+Figure~\ref{fig:mixagem} illustrates this overlapping of discretized sound waves, each with 100 samples.
+The duration is very short $\frac{f_s=44.1kHz}{100} \approx 2ms$.
 If $f_s=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_s}{100/2}=882Hz$, $\frac{f_s}{100/4}=1764Hz$ and $\frac{f_s}{100/5}=2205Hz$.
-The samples duration is very short $\frac{f_s=44.1kHz}{100} \approx 2ms$.
-One can complete the sequence with zeroes to sum (mix) sequences with different sizes.
-
+One can complete the sequence with zeroes to sum (mix) sequences of different sizes.
 The mixed notes are generally separated by the ear according to the physical laws of resonance and by the nervous system~\cite{Roederer}.
-This process of mixing musical notes results in musical harmony whose intervals between frequencies and chords of simultaneous notes guide subjective and abstract aspects of music and its appreciation~\cite{Harmonia}, which is addressed in Section~\ref{sec:notesMusic}.
-
-Sequences can be concatenated in time.
-If the sequences $\{t_{k,i}\}_0^{\Lambda_k-1}$ of size $\Lambda_k$ represent $k$ musical notes, their concatenation in a unique sequence $T_i$ is a simple melodic sequence, or melody of its own:
-
-\begin{equation}\label{eq:concatenacao}
-\begin{split}
-\{t_i\}_0^{\sum\Delta_k-1}= & \{t_{l,i}\}_0^{\sum\Delta_k-1}, \;\; \\ l\text{ smallest integer } : & \quad \Lambda_l > i -\sum_{j=0}^{l-1}\Lambda_j
-\end{split}
-\end{equation}
-
-This mechanism is demonstrated in Figure~\ref{fig:concatenacao} with the same sequences of Figure~\ref{fig:mixagem}.
-Although the sequences are short for the usual sample rates, it is easy to visually observe the concatenation of sound sequences.
-In addition, each note has  a duration larger than $100ms$ if $f_s<1kHz$.
+This process of mixing musical notes results in musical harmony and counterpoint which guide
+subjective aspects of music appreciation~\cite{Harmonia} and are addressed in Section~\ref{sec:notesMusic}.
 
@@
 
+Sequences can be concatenated in time.
+If the sequences $\{t_{k,i}\}_0^{\Lambda_k-1}$ of size $\Lambda_k$ represent $k$ musical notes, their concatenation in a unique sequence $T$ results in a simple melodic sequence:
+
+\begin{equation}\label{eq:concatenacao}
+\begin{split}
+\{t_i\}_0^{\sum\Delta_k-1}= \{t_{l,i}\}_0^{\sum\Delta_k-1}, \;\; l\text{ smallest integer}: \;\; \sum_{j=0}^{l}\Lambda_j > i
+\end{split}
+\end{equation}
+
+This mechanism is demonstrated in Figure~\ref{fig:concatenacao} with the same sequences of Figure~\ref{fig:mixagem}.
+Although the sequences are short for the usual sample rates, it is easy to visually observe the concatenation of sound sequences.
+Moreover, each note has  a duration larger than $100ms$ if $f_s<1kHz$.
+
 The musical piece \emph{reduced-fi} explores the temporal juxtaposition of notes, resulting in a homophonic piece.
-The vertical principle is demonstrated at the \emph{sonic portraits}, static sounds with peculiar spectrum.
-Both pieces were written in Python and are available as part of the \massa\ \emph{toolbox}.\cite{MASSA}
-
-With the basic musical digital note carefully described, the next section develops the temporal evolution of its contents as in \emph{glissandi} and intensity envelopes.
-Filtering of spectral components and noise generation complements the musical note as a self-contained unity.
-Section~\ref{notasMusica} is dedicated to the organization of these notes as music by using metrics and trajectories, with regards to traditional music theory.
+The vertical principle is demonstrated in the \emph{sonic portraits} which are static sounds with peculiar spectrum.\cite{MASSA}
 
@@
 \section{Variation in the basic note}\label{sec:internalVar}\label{sec:varInternas}
-The basic digital music note defined in Section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (loudness) and timbre.
-This is a useful and paradigmatic model, but it does not exhaust all the aspects of a musical note.
-First of all, characteristics of the note change along the note itself~\cite{Chowning}.
-For example, a $3s$ piano note has intensity with abrupt rise at the beginning and progressive decay, has spectrum variations with harmonics decaying and some others emerging along time.
-These variations are not mandatory, but they are used in sound synthesis for music because they reflect how sounds appear in nature.
-This is considered so important that there is a rule of thumb: to make a sound that incites interest by itself, arrange internal variations on it~\cite{Roederer}.
-To explore all the ways in which variations occur within a note is out of the scope of any work, given the sensibility of the human ear and the complexity of human sound cognition.
-In the following, primary resources are presented to produce variations in the basic note.
-It is worthwhile to recall that all the relations in this and other sections are implemented in Python and published in the public domain \massa\ toolbox.
-The musical pieces \emph{Transit para meter}, \emph{Vibrates and shakes}, \emph{Tremolos, vibratos and the frequency}, \emph{Little train of impulsive hillbillies}, \emph{Noisy band}, \emph{Bela rugosi}, \emph{Children's chorus}, \emph{ADa and SaRa} were made to validate and illustrate concepts of this section.
-The code that synthesizes these pieces is also part of the toolbox\cite{MASSA}.
+The basic digital music note defined in Section~\ref{sec:notaDisc} has the following characteristics: duration, pitch, loudness and timbre.
+We also stated that a note is always spatialized and considered both localization and reverberation.
+This is a useful and paradigmatic model, but it does not exhaust the aspects of a musical note.
+First of all, a real note is not static~\cite{Chowning}.
+For example, a $3s$ piano note has an abrupt rise of intensity at the beginning and a progressive decline,
+has harmonic components decaying and emerging.
+These variations are not mandatory, but they are used in sound synthesis for music because 
+they reflect how sounds appear in nature and because they yield interesting sounds~\cite{Roederer}.
+To explore all the ways in which variations occur within a note is out of the scope of any work given
+the complexity and sensibility of the human cognition of sounds.
+Resources essential to produce variations in the basic note are presented in this section.
+The musical pieces \emph{Transit para meter}, \emph{Vibrates and shakes}, \emph{Tremolos, vibratos and the frequency}, \emph{Little train of impulsive hillbillies}, \emph{Noisy band}, \emph{Bela rugosi}, \emph{Children's chorus}, \emph{ADa and SaRa} were made to validate and illustrate concepts of this section.~\cite{MASSA}
 
 \subsection{Lookup table}\label{subsec:lookup}
-The \emph{Lookup Table} (or simply LUT) is an array for indexed operations which substitutes continuous and repetitive calculation.
+The \emph{Lookup Table} (LUT) is an array for indexed operations which substitutes continuous and repetitive calculation.
 It is used to reduce computational complexity and for employing functions without calculating them directly, e.g.
@@
     \caption{Search in the \emph{lookup table} to synthesize sounds at different frequencies using a unique waveform with high resolution.
-	Each \emph{i-th} sample $t_i^f$ of a sound with frequency $f$ is related to the samples in the table $\widetilde{l}$\, by the relation $t_i^{f}=\widetilde{l}_{\left\lfloor i.f\frac{\widetilde{\Lambda}}{f_s} \right\rfloor \%\,\widetilde{\Lambda}}$ where $\widetilde{\Lambda}$ is the number of sample in $\widetilde{l}$ and $f_s$ is the sampling rate.}
+	Each \emph{i-th} sample $t_i^f$ of a sound with frequency $f$ is related to the samples in the table $\widetilde{l}$\, by the relation $t_i^{f}=\widetilde{l}_{\left\lfloor i.f\frac{\widetilde{\Lambda}}{f_s} \right\rfloor \%\,\widetilde{\Lambda}}$ where $\widetilde{\Lambda}$ is the number of sample in $\widetilde{L}$ and $f_s$ is the sampling rate.}
         \label{fig:lut}
@@
 
-Let $\widetilde{\Lambda}$ be the wave period in samples and $\widetilde{L_i} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ the samples $\widetilde{l_i}$ (refer to Equation~\ref{periodoUnico}), a sequence $T_i^{f,\,\Delta}$ with samples of a sound with frequency $f$ and duration $\Delta$ can be obtained by means of $\widetilde{L_i}$:
+Let $\widetilde{\Lambda}$ be the wave period in samples and $\widetilde{L} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ the sequence of samples $\widetilde{l_i}$.
+A sequence $T^{f,\,\Delta}$ with samples of a sound with frequency $f$ and duration $\Delta$ can be obtained by means of $\widetilde{L}$:
 
@@
 \begin{split}
-T_i^{f,\,\Delta}=\left\{t_i^f\right\}_0^{\lfloor \, f_s .
+T^{f,\,\Delta}=\left\{t_i^f\right\}_0^{\lfloor \, f_s .
 \Delta \, \rfloor -1} = \left\{ \, \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} }\, \right\}_{0}^{\Lambda-1}\; , \quad \text{where} \;\; \gamma_i = \left \lfloor i f \frac{ \widetilde{\Lambda}}{f_s} \right \rfloor
@@
 
-In other words, with the right LUT indexes ($\gamma_i\%\widetilde{\Lambda}$) it is possible to synthesize sounds at any frequency.
+In other words, with the right LUT indexes ($\gamma_i\%\widetilde{\Lambda}$) it is possible to synthesize sounds at an arbitrary frequency.
 Figure~\ref{fig:lut} illustrates the calculation of $\{t_i\}$ sample from $\left\{\,\widetilde{l}_i\,\right\}$ for $f=200Hz$, $\widetilde{\Lambda}=128$ and adopting the sample rate of $f_s=44.1kHz$.
@@
 
-The expression defining the variable $\gamma_i$ can be understood as $f_s$ being added to $i$ at each second.
-If $i$ is divided by the sample frequency, $\frac{i}{f_s}$
+The calculation of $\gamma_i$ can be understood as follows:
+$\frac{i}{f_s}$
 is incremented by $1$ at each second.
-Multiplied by the period, it results in $i \frac{\widetilde{\Lambda}}{f_s}$, which covers the period in one second.
-Finally, with frequency $f$ it results in $i .
-f \frac{\widetilde{\Lambda}}{f_s}$ which completes $f$ periods $\widetilde{\Lambda}$ in $1$ second, i.e.
-the resulting sequence presents the fundamental frequency $f$.
+Multiplied by the period, $i \frac{\widetilde{\Lambda}}{f_s}$ covers $\widetilde{\Lambda}$ in one second.
+Finally, with frequency $f$ it results in $i f \frac{\widetilde{\Lambda}}{f_s}$ which completes $f$ cycles on $\widetilde{L}$ in $1$ second, i.e.
+the resulting sequence has the fundamental frequency $f$.
 
@@
 Limits exist only at low frequencies when the size of table $\widetilde{\Lambda}$ is not sufficient for the sample rate $f_s$.
-The lookup procedure is virtually costless and replaces calculations by simple indexed searches (what is generally understood as an optimization process).
+The lookup procedure replaces calculations by simple indexed searches (what is generally understood as an optimization process).
 Unless otherwise stated, this procedure will be used along all the text for every applicable case.
 LUTs are broadly used in computational implementations for music.
-A classical usage of LUTs is known as \emph{Wavetable Synthesis}, which consists of many LUTs used together to generate a quasi-periodic music note~\cite{Cook,Wavetable}.
+A classical usage of LUTs is known as \emph{Wavetable Synthesis}, which consists of many LUTs used together to generate a quasi-periodic musical note~\cite{Cook,Wavetable}.
 

### 2017-08-10 article/acm/body.tex
revisado ateh o final dos ruidos

@@
 
-\subsection{Spectrum of sampled sound}
+\subsection{Spectrum of sampled sound}\label{sec:spectrum}
 These sinusoidal components in the discretized sound have some particularities.
@@
 
-Consider a note with duration $\Delta = \frac{\Lambda}{f_s}$, in which the frequency $f=f_i$ varies linearly from $f_0$ to $f_{\Lambda -1}$. Thus:
+Consider a note with duration $\Delta = \frac{\Lambda}{f_s}$, in which the frequency $f_i$ varies linearly from $f_0$ to $f_{\Lambda -1}$:
 
 \begin{equation}\label{freqLinear}
- F_i=\{f_i\}_0^{\Lambda-1}=\left\{f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right\}_0^{\Lambda-1}
+ F=\{f_i\}_0^{\Lambda-1}=\left\{f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right\}_0^{\Lambda-1}
 \end{equation}
@@
 
-\noindent where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_s}$ is the LUT increment between two samples given the sound frequency of the first sample.
-Therefore, it is handy to calculate the elements $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ by means of the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$.
+\noindent where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_s}$ is the LUT index increment between two samples given the frequency of the first sample.
+Therefore, it is handy to calculate the amplitudes $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ by means of the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$.
 Equations~\ref{freqLinear},~\ref{indiceLinear} and~\ref{serieAmostralLin} are related with the linear progression of the frequency.
@@
 
+The term $\frac{i}{\Lambda-1}$ covers the interval $[0,1]$ and it is possible to raise it to a power in such a way that the beginning of the transition will be smoother or steeper.
+This procedure is useful for energy variations with the purpose of changing the loudness\footnote{The
+loudness (a psychophysic quality) is a consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is the wave total energy.
+Wave energy can be varied in many ways.
+The simplest way consists of modifying the amplitude by multiplying the whole sequence by a real number (see Section~\ref{subsec:volume}).
+The increase of energy without changing amplitude ambit is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}.}.
+It is sufficient to multiply the original sequence element-wise by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is an arbitrary coefficient and $t_{\Lambda-1}.a_{\Lambda-1}$ is the value to be reached at the end of the transition.
+
+Thus, for amplitude variations:
+\begin{equation}\label{seqAmp}
+\begin{split}
+ \{a_i\}_0^{\Lambda-1}= \left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}= \left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
+\end{split}
+\end{equation}
+where $a_0=1$. And, therefore:
+
+\begin{equation}\label{transAmp}
+\begin{split}
+ T^{'}=T \odot A = & \{t_i a_i\}_0^{\Lambda-1} \\ = & \left \{ t_i (a_{\Lambda-1} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
+\end{split}
+\end{equation}
+
 \begin{figure}[h!]
@@
 
-The term $\frac{i}{\Lambda-1}$ covers the interval $[0,1]$ and it is possible to raise it to a power in such a way that the beginning of the transition will be smoother or steeper.
-This procedure is useful for energy variations with the purpose of changing the loudness\footnote{The
-loudness (psychophysical quality) is a consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is the wave total energy.
-Wave energy is easy to modify (see Section~\ref{subsec:volume}) and can be varied in many ways.
-The simplest way consists of modifying the amplitude by multiplying the whole sequence by a real number.
-The increase of energy without changing amplitude ambit is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}.}.
-It is sufficient to multiply the original sequence by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is the given coefficient and $a_{\Lambda-1}$ is a fraction of the original amplitude, which is the value to be reached at the end of the transition.
-
-Thus, for amplitude variations:
-\begin{equation}\label{seqAmp}
-\begin{split}
- \{a_i\}_0^{\Lambda-1}= \left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}= \left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
-\end{split}
-\end{equation}
-where $a_0=1$. And, therefore:
-
-\begin{equation}\label{transAmp}
-\begin{split}
- T_i^{'}=T_i \odot A_i = & \{t_i . a_i\}_0^{\Lambda-1} \\ = & \left \{ t_i . (a_{\Lambda-1} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
-\end{split}
-\end{equation}
-
 It is often convenient to have $a_0=1$ to start a new sequence with the original amplitude and then progressively change it.
 If $\alpha=1$, the amplitude variation follows the geometric progression that implies the linear variation of loudness.
-Figure~\ref{fig:transicao} depicts transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by Equation~\ref{eq:ampVol}.
+Figure~\ref{fig:transicao} depicts transitions for different values of $\alpha$, a gain of $\approx 6dB$ as exposed by Equation~\ref{eq:ampVol}.
 
@@
 Both cases make the procedure useless, once a ratio of any number in relation to zero is not well defined for our purposes.
-It is possible to solve this dilemma choosing a number that is small enough like $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum loudness for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$).
+It is possible to solve this dilemma by choosing a number that is small enough such as $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum amplitude for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$).
 A linear fade can be used then to reach zero amplitude, if needed.
@@
 
-Here the conversion between decibels and amplitude is convenient, with Equations~\ref{ampDec} and~\ref{transAmp} specifying a transition of $V_{dB}$ decibels:
+Here, the conversion between decibels and amplitude is convenient. Given Equations~\ref{ampDec} and~\ref{transAmp}, a transition of $V_{dB}$ decibels is:
 
 \begin{equation}\label{seqAmpDB}
-T_i^{'}=\left\{ t_i 10^{\frac{V_{dB}}{20}\left( \frac{i}{\Lambda-1} \right)^\alpha} \right\}_0^{\Lambda-1}
+T^{'}=\left\{ t_i 10^{\frac{V_{dB}}{20}\left( \frac{i}{\Lambda-1} \right)^\alpha} \right\}_0^{\Lambda-1}
 \end{equation}
@@
 The greater the value of $\alpha$, the smoother the sound introduction and more intense its end.
-$\alpha>1$ results in loudness transitions commonly called \emph{slow fade}, while $\alpha<1$ results in \emph{fast fade}\cite{guillaume}.
+$\alpha>1$ results in loudness transitions commonly called \emph{slow fade}, while $\alpha<1$ results in \emph{fast fade}~\cite{guillaume}.
 
 The linear transitions will be used for AM and FM synthesis, while logarithmic transitions are proper tremolos and vibratos, as developed in Section~\ref{subsec:tvaf}.
-A non-oscillatory exploration of these variations is in the music piece \emph{Transit para meter}, whose code is online as part of the \massa\ toolbox\cite{MASSA}.
+A non-oscillatory exploration of these variations is in the music piece \emph{Transit para meter}~\cite{MASSA}.
 
 \subsection{Application of digital filters}\label{subsec:filtros}
-This subsection is limited to a description of sequences processing, by convolution and difference equations, and immediate applications, as a thorough discussion of filtering is beyond the scope of this study\footnote{The implementation of filters encompasses an area of recognized complexity, with dedicated literature and software implementations~\cite{Openheim,smith}.}.
-Filter applications can be part of the synthesis process or made subsequently as part of processes commonly referred to as ``acoustic/sound treatment''.
+This subsection is limited to a description of processing by convolution and difference equations, and immediate applications, as a thorough discussion of filtering is beyond the scope of this study\footnote{The
+design and implementation of filters is field of recognized complexity,
+with dedicated literature and software~\cite{Openheim,smith}.}.
+Filtering can be part of the sound synthesis or made subsequently as part of processes commonly referred to as ``acoustic/sound treatment''.
 
@@
 This sample representation is called `impulse response' $\{h_i\}$.
-FIR filters are applied in the time domain of digital sound by means of convolution with the respective impulse response of the filter\footnote{It is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both the sound and the impulse response, and then performing the inverse Fourier transform in the resulting spectrum.\cite{Openheim}}.
-For the purposes of this work, convolution is defined as:
+FIR filters are applied in the time domain of digital sound by means of convolution with the respective impulse response of the filter\footnote{It
+is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both the sound and the impulse response, and then performing the inverse Fourier transform of the resulting spectrum.
+This is usually computationally cheaper than performing the convolution in the time domain
+but might imply greater complexity especially if the sequence of samples is not fully available (e.g.
+in real-time applications)~\cite{Openheim}.}.
+For the purposes of this work, convolution $T*H$ of $T$ with $H$ is defined as:
 
@@
  \begin{split}
- \left\{t_i'\right\}_0^{\Lambda_t+\Lambda_h-2\; = \;\Lambda_{t\, '}-1} = & \{(T_j*H_j)_i\}_0^{\Lambda_{t \, '}-1} \\ = & \left \{ \sum_{j=0}^{min(\Lambda_h-1,i)}h_{j} t_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
+ \left\{t_i'\right\}_0^{\Lambda_t+\Lambda_h-2\; = \;\Lambda_{t\, '}-1} = & \{(T*H)_i\}_0^{\Lambda_{t \, '}-1} \\ = & \left \{ \sum_{j=0}^{min(\Lambda_h-1,i)}h_{j} t_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
      \\ = & \left \{ \sum_{j=max(i+1-\Lambda_h,0)}^{i}t_j h_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
@@
 In other words, the sound $\{t_i'\}$, resulting from the convolution of $\{t_i\}$, with the impulse response $\{h_i\}$, has each $i$-th sample $t_i$ overwritten by the sum of its last $\Lambda_h$ samples $\{t_{i-j}\}_{j=0}^{\Lambda_h-1}$ multiplied one-by-one by samples of the impulse response $\{h_i\}_0^{\Lambda_h-1}$.
-This procedure is illustrated in Figure~\ref{fig:conv}, where the impulse response $\{h_i\}$ is in its retrograde form, and $t_{12}'$ and $t_{32}'$ are two calculated samples using the convolution given by $(T_j*H_j)_i=t_i'$.
-The final signal always has the length of $\Lambda_t+\Lambda_h -1=\Lambda_{t'}$.
-
-With this procedure it is possible to achieve reverberators, equalizers, \emph{delays}, to name a few of a variety of other filters for sound processing used to obtain musical/artistic effects.
+This procedure is illustrated in Figure~\ref{fig:conv}, where the impulse response $\{h_i\}$ is in its retrograde form, and $t_{12}'$ and $t_{32}'$ are two samples from the convolution.
+The final signal always has the length $\Lambda_{t'}=\Lambda_t+\Lambda_h -1$.
+With this procedure it is possible to achieve reverberators, equalizers, \emph{delays}, to name a few of a variety of filters for sound processing used to obtain musical/artistic effects.
 
 The impulse response can be provided by physical measures or by pure synthesis.
-An impulse response for a reverberation application, for example, can be obtained by sound recording of the environment when someone triggers a click which resembles an impulse, or obtained by a sinusoidal sweep whose Fourier transform approximates its frequency response.
-Both are impulse responses which, properly convoluted with the sound sequence, result in the own sound with a reverberation that resembles the original environment where the measure was made~\cite{Cook}.
-The inverse Fourier transform of an even and real envelope is an impulse response of a FIR filter.
-Convoluted with a sound, it performs the frequency filtering specified by the envelope.
-The greater the number of samples, the higher the envelope resolution and the computational complexity, which should often be weighted, for convolution is expensive.
+An impulse response for a reverberation application, for example, can be obtained by recording the sound in a room when someone triggers a click which resembles an impulse, or be obtained by a sinusoidal sweep whose Fourier transform approximates its frequency response.
+Both are impulse responses which, when convoluted with a sequence of sound samples, result in the own sound with a reverberation that resembles the original environment where the measure was made~\cite{Cook}.
+The inverse Fourier transform of a sequence of coefficients might be used as an impulse response of a FIR filter.
+Convoluted with a sound, it performs the frequency filtering specified by coefficients\footnote{Given
+that the sound is described by a sequence of real numbers,
+the coefficients should follow the rule $c_k=c^*_{\Lambda-k}$ as discussed in Section~\ref{sec:spectrum}.}.
+The greater the number of samples, the higher the spectral resolution and the computational complexity, which should often be weighted, for convolution is expensive.
 
 An important property is the time shift caused by convolution with a shifted impulse.
-Despite being computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse for each reincidence of the sound.
+Despite being computationally expensive, it is possible to create \emph{delay lines} by means of convolution with an impulse response that has an impulse for each reincidence of the sound.
 Figure~\ref{fig:delays} shows the shift caused by convolution with an impulse.
-Depending on the density of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as pitch (from about 20 impulses per second and higher frequencies).
+Depending on the density of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as a sound amalgam (more than $\approx 20$ impulses per second).
 In the latter case, the process resembles granular synthesis, reverberation and equalization.
@@
 
-There are many methods and tools to obtain IIR filters.
-The text below lists a selection for didactic purposes and as a reference.
-They are well behaved filters whose aspects are described in Figure~\ref{fig:iir}.
-For filters of simple order, the cutoff frequency $f_c$ is where the filter performs an attenuation of $-3dB \approx 0.707 $ of the original amplitude.
+There are many methods and tools to obtain IIR filters. The text below lists a selection for didactic and reference purposes.
+They are well behaved filters whose frequency response are illustrated in Figure~\ref{fig:iir}.
+For the low-pass and high-pass filters, the cutoff frequency $f_c$ is where the filter performs an attenuation of $-3dB \approx 0.707 $ of the original amplitude.
 For band-pass and band-reject (or 'notch') filters, this attenuation has two specifications: $f_c$ (in this case, the `center frequency') and bandwidth $bw$.
@@
 It is possible to apply these filters successively in order to obtain filters with other frequency responses.
-Another possibility is to use a biquad 'filter recipe'\footnote{Short for 'biquadratic': its transfer function has two poles and two zeros, i.e.
+Other possibilities include the use of a biquad 'filter recipe'\footnote{Short 
+for 'biquadratic': its transfer function has two poles and two zeros, i.e.
 its first direct form consists of two quadratic polynomials in the fraction: $\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1- b_1.z^{-1} -b_2 .
-z^{-2}}$.} or the calculation of Chebichev filter coefficients\footnote{Butterworth and Elliptical filters can be considered as special cases of Chebichev filters~\cite{Openheim,smith}.}.
+z^{-2}}$.}
+or a Chebichev filter\footnote{Butterworth and Elliptical filters can be considered as special cases of Chebichev filters~\cite{Openheim,smith}.}.
 Both alternatives are explored by~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University~\cite{music-dsp,Openheim}.
@@
         \includegraphics[width=\textwidth]{figures/iir__}
-    \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of Equations~\ref{eq:passa-baixas},~\ref{eq:passa-altas},~\ref{eq:passa-banda} and~\ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
+    \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of Equations~\ref{eq:passa-baixas},~\ref{eq:passa-altas},~\ref{eq:passa-banda} and~\ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and bandwidths.}
         \label{fig:iir}
@@
 \begin{enumerate}
-  \item Low-pass with a simple pole, module of the frequency response in the upper left corner of Figure~\ref{fig:iir}.
-The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_s$ in which an attenuation of $3dB$ occurs.
-The coefficients $a_0$ and $b_1$ of the IIR filter are given by $x \in [e^{-\pi},1]$:
+  \item Low-pass with a simple pole and module of the frequency response in the upper left corner of Figure~\ref{fig:iir}.
+The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, which is the fraction of the sample frequency $f_s$ in which an attenuation of $3dB$ occurs.
+The coefficients $a_0$ and $b_1$ of the IIR filter are given by:
 \begin{equation}\label{eq:passa-baixas}
  \begin{split}
- x & =e^{-2\pi f_c} \\
- a_0 & =  1-x \\
- b_1 & =  x
- \end{split}
-\end{equation}
-  \item High-pass filter with a simple pole, module of its frequency responses at the upper right corner of Figure~\ref{fig:iir}.
-The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of $x \in [e^{-\pi},1]$:
+ a_0 & =  1-b_1 \\
+ b_1 & = e^{-2\pi f_c}
+ \end{split}
+\end{equation}
+  \item High-pass filter with a simple pole and module of the frequency responses at the upper right corner of Figure~\ref{fig:iir}.
+	  The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ has coefficients:
 \begin{equation}\label{eq:passa-altas}
  \begin{split}
- x & =e^{-2\pi f_c} \\
- a_0 & =  \frac{x+1}{2} \\
- a_1 & =  -\frac{x+1}{2} \\
- b_1 & =  x
+ a_0 & =  \frac{b_1+1}{2} \\
+ a_1 & =  -\frac{b_1+1}{2} \\
+ b_1 & = e^{-2\pi f_c}
  \end{split}
@@
 \item Notch filter.
-This filter is parametrized by a center frequency $f_c$ and bandwidth $bw$, both given as fraction of $f_s$, therefore $f,\; bw \in (0,0.5)$.
+This filter is parametrized by a center frequency $f_c$ and bandwidth $bw$, both given as a fraction of $f_s$, therefore $f,\; bw \in (0,0.5)$.
 Both frequencies $f_c \pm bw$ have $\approx 0.707$ of the amplitude, i.e.
-attenuation of $3dB$.
+an attenuation of $3dB$.
 The auxiliary variables $K$ and $R$ are defined as:
@@
 
-Some noises with static spectrum are listed below. They are called \emph{colored noise} since they are associated with colors for many reasons.
-Figure~\ref{fig:ruidos} shows the spectrum profile and the corresponding sonic sequence side-by-side.
+Some noises with static spectrum are listed below. They are called \emph{colored noises} since they are associated with colors for many reasons.
+Figure~\ref{fig:ruidos} shows the spectrum profile and the corresponding sonic samples sequence side-by-side.
 All five noises were generated with the same phase for each component, making it straightforward to observe the contributions of different parts of the spectrum.
@@
 
-The exponential $e^{j.x}$ is a way to obtain unitary module and random phase for the value of $c_i$. In addition, $c_{\Lambda/2}$ is always real (as discussed in the previous section).
-
- \item The pink noise is characterized by a decrease of $3dB$ per octave. This noise is useful for testing electronic devices, being prominent in nature~\cite{Roederer}. 
+The exponential $e^{j.x}$ is a way to obtain unitary module and random phase for the value of $c_i$. In addition, $c_{\Lambda/2}$ is always real (as discussed in Section~\ref{sec:spectrum}).
+
+ \item The pink noise is characterized by a decrease of $3dB$ per octave.
+	 This noise is useful for testing electronic devices and are prominent in nature~\cite{Roederer}. 
  %\begin{split}
@@
 
-  \item The brown noise received this name after Robert Brown, who described the Brownian movement\footnote{Although its origin is disparate with its color association, this noise became established with this specific name in musical contexts.
-Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more vivid colors~\cite{Cook,guillaume}.} What characterizes brown noise is the decrease of $6dB$ per octave, with $\alpha_i$ in Equations~\ref{eq:rosa} being:
+  \item The brown noise received this name after Robert Brown, who described the Brownian movement\footnote{Although
+	  its origin is disparate with its color association, this noise became established with this specific name in musical contexts.
+Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more vivid colors~\cite{Cook,guillaume}.}.
+What characterizes a brown noise is the decrease of $6dB$ per octave, with $\alpha_i$ in Equations~\ref{eq:rosa} being:
 
@@
  \item In the blue noise there is a gain of $3dB$ per octave in a band limited by the minimum frequency $f_{\text{min}}$ and the maximum frequency $f_{\text{max}}$.
-Therefore, the corresponding equation is also based on the Equations~\ref{eq:rosa}:
+	 Therefore, using Equations~\ref{eq:rosa} again:
 
@@
 
- \item The black noise has higher losses than $6dB$ for octave:
+ \item The black noise has higher losses than $6dB$ per octave:
 
@@
 Those curves are obtained by experiments and are imperative to obtain $\alpha_i$.
-An implementation of ISO 226, which is the last revision of these curves, is in the \massa\ toolbox~\cite{MASSA}.
+An implementation of ISO 226, which is the last revision of these curves, is in the file \texttt{src/aux/iso226.py}~\cite{MASSA}.
 \end{itemize}
@@
 This subsection discussed only noise with static spectrum.
-There are also characterizations for noise with dynamic spectrum during the time, and noises which are fundamentally transient, like clicks and chirps.
+There are also characterizations for noise with dynamic spectrum, and noises which are fundamentally transient, like clicks and chirps.
 The former are easily modeled by an impulse relatively isolated, while chirps are not in fact a noise, but a fast scan of some given frequency band~\cite{Cook}.

### 2017-08-10 article/acm/body.tex
finished review until the end of section 3

@@
 
-\subsection{Synonymy, polysemy and theoretical frames (disclaimer)}
+\subsection{Synonymy, polysemy and theoretical frames (disclaimer)}\label{sec:discl}
 Given that the main topic of this article (the expression of musical elements
@@
 Canonically, the basic characteristics of a musical note are duration, loudness, pitch and timbre~\cite{Lacerda}.
-All relations described in this section are implemented in the file \texttt{src/sections/eqs2.1.py}.
+All relations described in this section are implemented in the file \texttt{src/sections/1.py}.
 The musical pieces \emph{5 sonic portraits} and \emph{reduced-fi} are also available online to corroborate the concepts.
@@
 \subsection{Tremolo and vibrato, AM and FM}\label{subsec:tvaf}
-Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in loudness (intensity)\footnote{The jargon may be different in other contexts.
-For example, in piano, tremolo is a vibrato in the classification used here.
-The definitions given here are common in contexts regarding music theory and electronic music.
-In addition, they are based on a broader literature than the one used for a specific instrument, practice or musical tradition~\cite{Lacerda,Harmonia}.}.
-A vibrato can be achieved by:
+Vibrato is a periodic variation in pitch (frequency) and tremolo is a periodic variation in loudness (intensity)\footnote{The
+jargon may be different in other contexts as stated in Section~\ref{sec:discl}~\cite{Lacerda,Harmonia}.}.
+A vibrato with frequency $f'$ can be achieved by:
 
@@
 \begin{equation}\label{vbrT}
- T_i^{f, vbr(f',\,\nu)}=\left\{ t_i^{f,vbr(f',\,\nu)} \right\}_0^{\Lambda-1}=\left\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \right\}_0^{\Lambda-1}
+ T^{f, vbr(f',\,\nu)}=\left\{ t_i^{f,vbr(f',\,\nu)} \right\}_0^{\Lambda-1}=\left\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \right\}_0^{\Lambda-1}
 \end{equation}
@@
          \includegraphics[width=\columnwidth]{figures/vibrato_}
-     \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave, with $f_s=44.1kHz$. The color bar is uses decibels.}
+     \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave, with $f_s=44.1kHz$. The color bar is in decibels.}
          \label{fig:vibrato}
@@
 
-For the proper realization of the vibrato, it is important to pay attention to both tables and sequences.
-Table $\widetilde{M}_i$ with length $\widetilde{\Lambda}_M$ and the sequence with indices $\gamma_i'$ make the sequence $t_i'$ which is the oscillation pattern in the frequency while table $\widetilde{L}_i$ with length $\widetilde{\Lambda}$ and the sequence with indices $\gamma_i$ make $t_i$ which is the sound itself.
-Variables $\mu$ and $\nu$ quantify the vibrato intensity:
+It is important correctly distinguish both tables 
+($\widetilde{M}=\{\widetilde{m}\}_0^{\widetilde{\Lambda}_M-1}$ and $\widetilde{L}=\{\widetilde{l}\}_0^{\widetilde{\Lambda}-1}$)
+and sequences ($T'=\{t_i'\}_0^{\Lambda-1}$ and $T=\{t_i\}_0^{\Lambda-1}$).
+Table $\widetilde{M}$ and the indices $\gamma_i'$ yield the sequence $\{t_i'\}$
+which is the oscillation pattern in the frequency while table $\widetilde{L}$ and the indices $\gamma_i$
+yield $\{t_i\}$ which is the sound itself.
+Variables are useful to quantify the vibrato amplitude (or depth):
 \begin{itemize}
-    \item $\mu$ is a direct measure of how many Hertz are involved in the upper limit of the oscillation, while
-    \item $\nu$ is the direct measure of how many semitones (or half steps) are involved in the oscillation ($2\nu$ is the number of semitones between the upper and lower peaks of frequency oscillations of the sound $\{t_i\}$ in the vibrato).
+	\item let $\mu$ be frequency difference (in $Hz$) between the main frequency and the maximum frequency, and
+    \item let $\nu$ be the the number of semitones (or half steps) added and subtracted by the oscillation ($2\nu$ is the number of semitones between the upper and lower peaks of frequency in the sound).
 \end{itemize}
@@
 
-Figure~\ref{fig:vibrato} is the spectrogram of an artificial vibrato for a note with $1000Hz$ (between a \emph{B} and a \emph{C}), in which pitch deviation reaches one octave above and one below.
+Figure~\ref{fig:vibrato} is the spectrogram of an artificial vibrato for a note with $1000Hz$ (between a \emph{B} and a \emph{C}),
+in which the pitch deviation reaches one octave above and one below.
 Practically any waveform can be used to generate a sound and the vibrato oscillation pattern, with virtually any oscillation frequency and pitch deviation\footnote{The pitch deviation is called 'vibrato depth' and is generally given in semitones or cents (one cent $= \frac{1}{100}$ of a semitone).}.
-Those oscillations with precise waveforms and arbitrary amplitudes are not possible in traditional music instruments, and thus it introduces novelty in the artistic possibilities.
-
-Tremolo is similar: $f'$, $\gamma_i'$ and $t_i'$ remains the same.
+Oscillations with precise waveforms and arbitrary amplitudes are not possible in traditional music instruments, and thus it introduces novelty in the artistic possibilities.
+
+The tremolo is similar: $f'$, $\gamma_i'$ and $t_i'$ remains the same.
 The amplitude sequence to be multiplied by the original sequence $t_i$ is:
@@
 \begin{split}
- T_i^{tr(f')}=\left \{ t_i^{tr(f')} \right \}_0^{\Lambda-1}=\{ t_i .
+ T^{tr(f')}=\left \{ t_i^{tr(f')} \right \}_0^{\Lambda-1}=\{ t_i .
 a_i \}_0^{\Lambda-1}= \left \{t_i .10^{t_i' \frac{V_{dB}}{20}}    \right \}_0^{\Lambda-1}=\left\{t_i .
@@
 
-\noindent where $V_{dB}$ is the oscillation depth in decibels of tremolo and $a_{\text{max}}=10^{\frac{V_{dB}}{20}}$ is the maximum amplitude gain.
-The measurement in decibels is suitable because the maximum increase in amplitude is not equivalent to the related maximum decrease, while the difference in decibels remains.
+\noindent where $V_{dB}$ is the oscillation depth in decibels of the tremolo and $a_{\text{max}}=10^{\frac{V_{dB}}{20}}$ is the maximum amplitude gain.
+The measurement in decibels is suitable because the maximum increase in amplitude is not equivalent to the related maximum decrease, while the difference in decibels is invariant.
 
@@
 
-The music piece \emph{Vibrates and shakes} explores these possibilities given by tremolos and vibratos, both used in conjunction and independently, with frequencies $f'$, different depths ($\nu$ and $V_{dB}$), and progressive parameters variations (tremolos and vibratos occur many times together in a traditional music instrument and voices).
-Aiming at a qualitative appreciation, the piece also develops a comparison between vibratos and tremolos in logarithmic and linear scales.
-Its source code is available online as part of the \massa\ toolbox.
+The music piece \emph{Vibrates and shakes} explores these possibilities given by tremolos and vibratos
+used simultaneously and independently, with different frequencies $f'$, depths ($\nu$ and $V_{dB}$),
+waveforms, and incremental modifications of parameters
+(tremolos and vibratos occur often together in traditional music instruments and voices).
+Aiming at a qualitative appreciation, the piece also develops a comparison between vibratos and tremolos in logarithmic and linear scales.~\cite{MASSA}
 
@@
          \includegraphics[width=\textwidth]{figures/tremolo_}
-     \caption{Tremolo with a depth of $V_{dB}=12dB$, with a sawtooth waveform as its oscillatory pattern, with $f'=1.5Hz$ in a sine of $f=40Hz$ (sample frequency $f_s=44.1kHz$).}
+     \caption{Tremolo with a depth of $V_{dB}=12dB$, with a sawtooth waveform as its oscillatory pattern, with $f'=1.5Hz$ in a sine of $f=40Hz$ (if sample frequency is $f_s=44.1kHz$).}
          \label{fig:tremolo}
@@
 This roughness is largely appreciated both in traditional classical music and current electronic music, especially in the \emph{Dubstep} genre.
-Roughness is also generated by spectral content that produces beating~\cite{Porres,porres2009}.
-The sequence \emph{Bela Rugosi} explores this roughness threshold with concomitant tremolos and vibratos at the same voice, with different intensities and waveforms.
-The corresponding code is available online in the \massa\ toolbox.
+Roughness is also generated by spectral content that produces beats\footnote{A beat is an interference pattern between two sounds which is perceived as a periodic variation of loudness.
+If the sounds are sinusoids, the variation is also sinusoidal with a frequency that is the difference between the frequencies of the sinusoids~\cite{Porres,porres2009}.}.
+The sequence \emph{Bela Rugosi} explores this roughness threshold with concomitant tremolos and vibratos at the same voice, with different intensities and waveforms.~\cite{MASSA}
 
@@
 In this case, the oscillations are audible as pitch.
-Then, $f'$, $\mu$ and the waveform together change the audible spectrum of original sound $T_i$ in different ways for both tremolos and vibratos.
+Then, $f'$, $\mu$ and the waveform together change the audible spectrum of original sound $T$ in different ways for both tremolos and vibratos.
 They are called AM (\emph{Amplitude Modulation}) and FM (\emph{Frequency Modulation}) synthesis,
@@
 
-For musical goals, it is possible to understand FM based on the
-case of sines and, when other waveforms are employed, to decompose the signals into their respective Fourier components (i.e.\ sines as well).
-The FM synthesis performed with a sinusoidal vibrato of frequency $f'$ and depth $\mu$ in a sinusoidal sound $T_i$ with frequency $f$ generates bands centered around $f$ and far from each other by $f'$:
+For musical goals, it is convenient to understand FM when only pure
+sines are used and, when other waveforms are employed,
+to decompose the signals into their respective Fourier components (i.e.\ sines as well).
+The FM synthesis performed with a sinusoidal vibrato of frequency $f'$ and depth $\mu$
+in a sinusoidal sound $T$ with frequency $f$ generates bands centered around $f$ and far from each other by $f'$:
 
@@
 In these equations, the frequency variation introduced by $\{t_i'\}$ does not follow the geometric progression that yields linear pitch variation, but reflects Equation~\ref{freqLinear}.
-The result of using Equations~\ref{vbrF} for FM is described in~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
-In fact, the simple and attractive FM behavior is usually observed with linear oscillations, such as in Equation~\ref{eq:fmEsp}, which yield less strident noisy sounds.
+The result of using Equations~\ref{vbrF} for FM is described in the Appendix D of~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
+In fact, the simple and attractive FM behavior is usually observed within linear oscillations,
+such as in Equation~\ref{eq:fmEsp}, which yield less strident and less noisy sounds.
 
@@
 Again, this is obtained by variations in the linear scale of the amplitude.
-A discussion of the
-spectrum of an AM performed with oscillations in the logarithmic amplitude scale is available in~\cite{dissertacao}. The sequence $T_i$, with frequency $f$, called `carrier', is modulated by
-$f'$, called 'modulator'. In FM and AM jargon, $\mu$ and
-$\alpha=10^{\frac{V_{dB}}{20}}$ are `modulation indexes'. The following equations are defined for the vibration pattern of the modulator sequence $\{t_i'\}$:
+A description of the spectrum of an AM performed with oscillations in the logarithmic amplitude scale is also available in the Appendix D of~\cite{dissertacao}.
+In FM and AM jargon,
+the sequence $T$ is called `carrier' and is modulated by
+$T'$, called 'modulator'; $\mu$ and
+$\alpha=10^{\frac{V_{dB}}{20}}$ are the `modulation indexes' and $f'$ is the modulation frequency.
+The following equations are useful for the achievement of the oscillatory pattern of the modulator sequence $\{t_i'\}$:
 
@@
 
-The modulator $\{t_i'\}$ is applied to the carrier $\{t_i\}$:
+The modulator $\{t_i'\}$ is applied to the carrier $\{t_i\}$ in a FM by:
 
@@
 \begin{equation}\label{fmGamma}
-\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_s} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_s} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}(f+\mu .
+\Delta_{\gamma_i}=\frac{\widetilde{\Lambda}}{f_s}f_i \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s} f_j \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}(f+\mu .
 t_j') \right\rfloor
@@
 \begin{equation}\label{fmT}
-T_i^{f,\, FM(f',\,\mu)}=\left\{ t_i^{f,\,FM(f',\,\mu)} \right\}_0^{\Lambda-1}=\left\{\,\widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \,\right\}_0^{\Lambda-1}
-\end{equation}
-
-\noindent where $\widetilde{l}$ is the waveform period with a length of $\widetilde{\Lambda}$ samples, used for the carrier signal.
-
-To perform AM, the signal $\{t_i\}$ needs to be modulated with $\{t_i'\}$ using the following equations:
+T^{f,\, FM(f',\,\mu)}=\left\{ t_i^{f,\,FM(f',\,\mu)} \right\}_0^{\Lambda-1}=\left\{\,\widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \,\right\}_0^{\Lambda-1}
+\end{equation}
+
+\noindent where $\widetilde{l}$ is the waveform period with a length of $\widetilde{\Lambda}$ samples, used for the carrier signal,
+	as exposed above.
+
+To perform AM, the signal $\{t_i\}$ needs to be modulated by $\{t_i'\}$ using the following equations:
 
@@
 
-\begin{multline}\label{amT}
-T_i^{f,\,AM(f',\,\alpha)}=\left\{ t_i^{f,\,AM(f',\,\alpha)} \right\}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}= \\ \{t_i . (1 + \alpha . t_i')    \}_0^{\Lambda-1}
-\end{multline}
+\begin{equation}\label{amT}
+T_i^{f,\,AM(f',\,\alpha)}=\left\{ t_i^{f,\,AM(f',\,\alpha)} \right\}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}= \{t_i . (1 + \alpha . t_i')    \}_0^{\Lambda-1}
+\end{equation}
 
@@
 At this point the musical possibilities are very wide.
-All characteristics, like pitch (given by frequency), timbre (achieved by waveforms, filters and noise), loudness (manipulated by intensity) and duration (given by the number of samples), can be considered in an absolute form or varied during the duration of the sound (clearly, with the exception of duration itself).
 The following musical usages comprehend a collection of possibilities with the purpose of exemplifying types of sound manipulations that result in musical material.
@@
 
-\noindent with $f^{vbr}$ and $f^{tr}$ as $f'$ in the referenced equations. They can also be associated with vibrato and tremolo oscillation frequency of Equation~\ref{vbrGamma}. 
+\noindent with both $f^{vbr}$ and $f^{tr}$ as $f'$ in the equations of Section~\ref{subsec:tvaf}.
  $\nu$ and $V_{dB}$ are the respective depth values of vibrato and
@@
 frequency} explores such bonds and exhibits variations in the oscillation waveform with the purpose of building a \emph{musical
-language} (details in the next section). The corresponding code is also available online as part of the \massa\ toolbox.
+language} (details in the next section).~\cite{MASSA}
 
 \subsubsection{Convolution for rhythm and meter}
+With the purpose of establishing metric and rhythms,
+the convolution with an impulse shifts the sound to the position of the impulse, as stated in Section~\ref{subsec:conv}.
 A musical pulse - such as specified by a BPM
-tempo - can be implied by an impulse at the start of each beat, with the purpose of establishing metric and rhythms: the convolution with an impulse shifts the sound to impulse position, as stated in Section~\ref{subsec:conv}.
+tempo - can be implied by an impulse at the start of each beat.
 For example, two impulses equally spaced build a binary division in the
@@
 proportions among the amplitudes of the sonic reincidences. The use of convolution with impulses in this context is explored in the music piece \emph{Little train of impulsive hillbillies}.
-These features encompass
-the creation of `sound amalgams' based on granular synthesis; see Figure~\ref{fig:pulsoSubAgl}.
-This piece is a link to next section contents because of the explicit organization of notes in higher level structures.
-The source code of the music piece is online, as part of the \massa\ toolbox\cite{MASSA}.
+The piece also encompass
+the creation of `sound amalgams' e.g. by granular synthesis (see Figure~\ref{fig:pulsoSubAgl}).
+This piece is a link to the contents of next section because of the explicit organization of notes in higher level structures.~\cite{MASSA}.
 
 \subsubsection{Moving source and receptor, Doppler effect}
-According to the description in Section~\ref{subsec:spac}, when an audio source or receptor is moving, its characteristics are ideally updated at each sample of the digital signal (if fast computational rendering is not at stake).
-Velocity components should be found for each ear.
-Hence, given the audio source speed $s_s$, with positive values if the source moves away from receptor, and receptor speed $s_r$, positive when it gets closer to audio source, the frequency is given by the well-known shift in the Doppler effect:
+Given the audio source speed $s_s$, with positive values if the source moves away from receptor,
+and receptor speed $s_r$, positive when it gets closer to audio source, the frequency is given by the well-known shift in the Doppler effect:
 
@@
 
-With both frequencies $f$ and $f_0$, and the IID from the new source
-position, it is possible to simulate the Doppler effect.
+With both frequencies $f$ and $f_0$, and the IID and ITD from the new source
+position (see Section~\ref{subsec:spac}), it is possible to simulate the Doppler effect.
 There is an addendum to improve the fidelity of the physical phenomena: to increase the received power, which may be understood as being proportional to the wave shrinking or expansion: $\Delta P=P_0\left(\frac{v_r-v_s}{343.2}\right)$,
@@
 If the audio source is in front of the receptor with $y_0$ meters of horizontal distance and $z_0$ meters of height, the distance is given by
-$D_i=\left\{ d_i=\sqrt{ y_{i}^{2}+z_{0}^{2} } \right\}_0^{\Lambda-1}$,
+$D=\left\{ d_i=\sqrt{ y_{i}^{2}+z_{0}^{2} } \right\}_0^{\Lambda-1}$,
 where $y_i=y_0+(v_s-v_r)\frac{i}{f_s}$ with $v_s$ and $v_r$ both
-horizontal, having null non-$y$ components. Amplitude changes with the distance and with the potency factor mentioned above (see Section~\ref{subsec:volume} for potency to amplitude conversion).
+horizontal, having null non-$y$ components. Amplitude changes with the distance and with the potency factor mentioned above (see Section~\ref{subsec:volume} for potency to amplitude conversion). Thus:
 
 \begin{equation}\label{eq:aDoppler}
-    A_i=\left\{ \frac{z_0}{d_i}A_{\Delta P}\right\}_0^{\Lambda-1} = \left\{ \frac{z_0}{\sqrt{y_i^2+z_0^2}} \sqrt{\frac{v_r-v_s}{343,2}+1}  \,\right\}_0^{\Lambda-1}
+    A=\left\{ \frac{z_0}{d_i}A_{\Delta P}\right\}_0^{\Lambda-1} = \left\{ \frac{z_0}{\sqrt{y_i^2+z_0^2}} \sqrt{\frac{v_r-v_s}{343,2}+1}  \,\right\}_0^{\Lambda-1}
 \end{equation}
@@
 \begin{equation}\label{eq:ffDoppler}
-    F_i=\{f_i\}_0^{\Lambda-1}=\left\{\frac{v_{sound} + v_r\frac{y_i}{\sqrt{z_0^2+y_i^2}}}{v_{sound}+v_s\frac{y_i}{\sqrt{z_0^2+y_i^2}}}f_0\right\}_0^{\Lambda-1}
-\end{equation}
-
-There is a Python implementation of the Doppler effect, also considering the intersection between audio source and receptor, in \massa\ toolbox~\cite{MASSA}.
+    F=\{f_i\}_0^{\Lambda-1}=\left\{\frac{v_{sound} + v_r\frac{y_i}{\sqrt{z_0^2+y_i^2}}}{v_{sound}+v_s\frac{y_i}{\sqrt{z_0^2+y_i^2}}}f_0\right\}_0^{\Lambda-1}
+\end{equation}
+
+There is an implementation of the Doppler effect in \texttt{src/sections/3.py}~\cite{MASSA}.
+We considered this a musical usage given that it is not a basic method to introduce variation into a note
+and because it is a valued effect in audio plugins and electro-acoustic music~\cite{floEsp}.
 
@@
 With the use of filters, the possibilities are even wider.
-Convolve a signal to have a reverberated version of it, to remove its noise, to distort or to handle
+Convolve a signal to have a reverberated version of it, to remove noise, to distort or to handle
 the audio aesthetically in other ways.
-For example, sounds originated from an old television or telephone can be simulated with a band-pass filter, allowing only frequencies between $1kHz$ and $3kHz$.
-By rejecting the frequency of an electric oscillation (usually $50Hz$ or $60Hz$) and the harmonics, one can remove noises caused by audio devices connected to the power supply.
-A more musical application is to perform filtering in specific bands and to use those bands as an additional parameter to the notes.
+For example, sounds originated from an old television or telephone can be simulated with a band-pass filter, allowing only frequencies between $300Hz$ and $2kHz$.
+By rejecting the frequency of an electric oscillation (usually $50Hz$ or $60Hz$) and the harmonics, one can remove noises caused by audio devices connected to the power supply (called hiss noise).
+Another musical application is to perform filtering in specific bands and to use those bands as an additional parameter to the notes.
 
@@
 Chains of these filters can perform complex and more accurate filtering routines.
-The music piece \emph{Noisy band} explores filters and many kinds and noise synthesis.
-The source code is available online as part of \massa\ toolbox.
+The music piece \emph{Noisy band} explores filters and many kinds and noise synthesis.~\cite{MASSA}
 
 A sound can be altered through different filtering and then mixed to create an effect known as \emph{chorus}.
-Based on what happens in a choir of singers, the sound is synthesized using small and potentially arbitrary modifications of parameters like center frequency, presence (or absence) of vibrato or tremolo and its characteristics, equalization, loudness, etc.
-As a final result, those versions of the original sound are mixed together (see Equation~\ref{eq:mixagem}).
-The music piece \emph{Child chorus} implements a chorus in many ways with different sounds.
-Its source code is available in \massa\ toolbox.
+Inspired by what happens in a choir of singers, the same note is synthesized many times using small and potentially arbitrary modifications of parameters like center frequency, presence (or absence) of vibrato or tremolo and its characteristics, equalization, loudness, duration, etc.
+The versions of the sound are then mixed together (see Equation~\ref{eq:mixagem}).
+The music piece \emph{Child chorus} implements a chorus in many ways with different sounds.~\cite{MASSA}
 
@@
 late reverberation can be achieved by a convolution with a section of pink, brown or black noise, with exponential decay of amplitude along time. Delay lines can be added as a prefix to the noise
-with the decay, and this accounts for both time parts of the reverberation: the early reflections and the late reverberation.
+with the decay, and this accounts for both temporal parts of the reverberation:
+the early reflections and the late reverberation.
 Quality can be improved by varying the geometric trajectory and filtering by each surface where wavefront reflected before reaching the ear in the first $100-200ms$ (mainly with a LP).
-The colored noise can be gradually introduced with a \emph{fade-in}: the initial moment given by direct
+The colored noise can be gradually introduced with a \emph{fade-in}: the initial moment given by the direct
 incidence of sound (i.e.\ without any reflection and given by ITD and IID), reaching its maximum at the beginning of the 'late
-reverberation', when the geometric incidences lose their relevance to the statistical properties of noise decay.
-As an example, consider $\Delta_1$ as the duration of the first section and $\Delta_R$ as the duration of total reverberation ($\Lambda_1=\Delta_1 f_s$, $\Lambda_R=\Delta_R
-f_s$). Let $p_i$ be the probability of a sound to repeat in the
+reverberation', when the geometric incidences lose their relevance to the statistical properties of a decaying noise.
+As an example, consider $\Delta_1$ as the duration of the first section and $\Delta_R$ as the duration of total reverberation 
+(let also $\Lambda_1=\Delta_1 f_s$, $\Lambda_R=\Delta_R
+f_s$).
+Let $p_i$ be the probability of a sound being repeated in the
 $i$-th sample. The amplitude decreases exponentially. Following
-Section~\ref{subsec:spac}, reverberation $R_i^1$ of the first period can be described as:
+Section~\ref{subsec:spac}, the samples $R^1$ of the impulse response of the first period can be described as:
 
@@
 \begin{split}
-	R_i^1  =\left\{r_i^1\right\}_0^{\Lambda_1-1}\;,\; \text{where}\;\;
+	R^1  =\left\{r_i^1\right\}_0^{\Lambda_1-1}\;,\; \text{where}\;\;
 	r_i^1  =\left\{
@@
 \noindent where $V_{dB}$ is the total decay in decibels, typically $-80dB$ or $-120dB$.
-Reverberation $R_i^2$ of the second period can be emulated by a brown noise $R_i^m$ (or by a pink noise $R_i^r$) with exponential amplitude decay of the waveform:
+The samples $R^2$ of the impulse response of the second period can be taken from a brown noise (or by a pink noise) with an exponential amplitude decay of the waveform:
 
 \begin{equation}\label{eq:p2rev}
-    R_i^2=\left\{r_i^2\right\}_{\Lambda_1}^{\Lambda_R-1}=\left\{10^{\frac{V_{dB}}{20}\frac{i}{\Lambda_R-1}}\,.\,r_i^m\right\}_{\Lambda_1}^{\Lambda_R-1}
-\end{equation}
-
-\noindent with
+    R^2=\left\{r_i^2\right\}_{\Lambda_1}^{\Lambda_R-1}=\left\{10^{\frac{V_{dB}}{20}\frac{i}{\Lambda_R-1}}\,.\,r_i^m\right\}_{\Lambda_1}^{\Lambda_R-1}
+\end{equation}
+
+The reverberation impulse response is then:
 
 \begin{equation}\label{eq:rev}
-	R_i=\left\{r_i\right\}_0^{\Lambda_R-1}\;,\; \text{where} \; r_i=\left\{
+	R=\left\{r_i\right\}_0^{\Lambda_R-1}\;,\; \text{where} \; r_i=\left\{
         \begin{array}{l l}
@@
 
-A sound with simulated reverberation can be achieved by a simple convolution of $R_i$ (called reverberation impulse response) with the sound sequence $T_i$, as described in Section~\ref{subsec:filtros}.
-Reverberation is well known for causing great interest in listeners and to provide sonorities that are more enjoyable.
+A sound with simulated reverberation can be achieved by a simple convolution of $R$ with the sound sequence $T$, as described in Section~\ref{subsec:filtros}.
+Reverberation is well known for causing great interest in listeners and to provide sonorities that are more enjoyable than the sound without any reverberation.
 Furthermore, modifications in the reverberation space consist in a common resource (almost a \textit{clich\'{e}}) to surprise and attract the listener.
-An implementation of the reverb recipe described here is in \massa\ toolbox~\cite{MASSA}
+An implementation of the reverb recipe described here is in the file \texttt{src/sections/3.py}.~\cite{MASSA}.
 
@@
 A pioneering implementation can be found in the Hammond Novachord synthesizer of 1938 and some variants are mentioned below~\cite{ADSR}.
-The canonical ADSR envelope is characterized by 4 parameters: attack duration (time in which sound reaches its maximum amplitude), decay duration (follows the attack immediately), level of sustained intensity (in which the intensity remains stable after the decay) and release duration (after sustained section, this is the duration needed for amplitude to reach zero or final value).
+The canonical ADSR envelope is characterized by 4 parameters: attack duration (time in which sound reaches its maximum amplitude), decay duration (follows the attack immediately), level of sustained intensity (in which the intensity remains stable after the decay) and release duration (after the sustained section, this is the duration needed for amplitude to reach zero or final value).
 Note that the sustain duration is not specified because it is the difference between the duration itself and the sum of the attack, decay and sustain durations.
@@
 \noindent with $\Lambda_X=\lfloor \Delta_X . f_s \rfloor\;\;\forall\;\; X \; \in
-(A,D,R\;)$ with $\xi$ being a small value that provides a satisfactory \emph{fade in} and \emph{fade out}, e.g.\ $\xi=10^{\frac{-80}{20}}=10^{-4}\;$ or $\;\xi=10^{\frac{-40}{20}}=10^{-2}$.
+(A,D,R\;)$ and $\xi$ being a small value that provides a satisfactory \emph{fade in} and \emph{fade out}, e.g.\ $\xi=10^{\frac{-80}{20}}=10^{-4}\;$.
 The lower the $\xi$, the slower the \emph{fade}, like the $\alpha$ illustrated in Figure~\ref{fig:transicao}.
@@
 Another common example is the use of more elaborated outlines of attack or decay.
-The music piece \emph{ADa and SaRa}, available
-in \massa, explores many configurations of the ADSL envelope.
-
+To aply the ADSR envelope, simply multiply the amplitudes of Equation~\ref{eq:adsr} by the samples of the note or musical except:
 \begin{equation}\label{eq:adsrApl}
@@
 \end{equation}
+
+The music piece \emph{ADa and SaRa} explores many configurations of the ADSR envelope.~\cite{MASSA}
 

### 2017-08-10 article/acm/body.tex
finished review until the end of section 3
finished review until the end chords

@@
 
+\begin{figure}[htp!]
+    \centering
+        \includegraphics[width=\columnwidth]{figures/adsr_}
+    \caption{An ADSR envelope (\emph{Attack, Decay, Sustain, Release}) applied to an arbitrary sound sequence.
+The linear variation of the amplitude is above, in blue.
+Below the amplitude variation is exponential.}
+        \label{fig:adsr}
+\end{figure}
+
 The ADSR envelope with durations $\Delta_A$, $\Delta_D$ and $\Delta_R$, with total duration $\Delta$ and sustain level $a_S$, given by the fraction of the maximum amplitude, to be applied to any sound sequence $T_i=\{t_i\}$, can be expressed as:
@@
 
-\begin{figure}[htp!]
-    \centering
-        \includegraphics[width=\columnwidth]{figures/adsr_}
-    \caption{An ADSR envelope (\emph{Attack, Decay, Sustain, Release}) applied to an arbitrary sound sequence.
-The linear variation of the amplitude is above, in blue.
-Below the amplitude variation is exponential.}
-        \label{fig:adsr}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\afterpage{\blankpage}
-
-%\clearpage
-
 \section{Organization of notes in music}\label{notasMusica} \label{sec:notesMusic}
-Consider $S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1}$, the sequence $S_j$ of $H$ musical events $s_j$.
-Let $S_j$ be a `musical structure', composed by events $s_j$ which are musical structures themselves, e.g.\ notes.
-This section is dedicated to techniques that make $S_j$ interesting and enjoyable for audition.
+Let $S=\left\{  s_j=T^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1}$ be a sequence of $H$ musical events $s_j$.
+Consider $S$ as a `musical structure', composed by events which are also musical structures themselves, e.g.\ notes.
+This section is dedicated to techniques that make $S$ aesthetically appealing for audition.
 More specifically, what follows is a summary of academic music composition theory and praxis.
@@
 Even so, we understand that this content is very useful for synthesizing music given all the content above.
-The concepts are given algorithmic implementations in \mass~\cite{SI} and can be further formalized~\cite{topos},
+The concepts are given algorithmic implementations~\cite{MASSA} and can be further formalized~\cite{topos},
 although at a cost of intelligibility which we chose to avoid in this exposition.
 
-The elements of $S_j$ can be overlapped by mixing them together, as in
-Equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, for building intervals and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of events
-in $S_j$, as in Equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao}, builds melodic sequences and rhythms, which are associated with the `horizontal thought'.
+The elements of $S$ may be overlapped by mixing them together, as in
+Equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, for building intervals and chords.
+This reflects the `vertical thought' in music. On the other hand, the concatenation of events
+in $S$, as in Equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao}, builds melodic sequences and rhythms, which are associated with the `horizontal thought'.
 The fundamental frequency $f$ and the starting moment
-(attack) are generally the most important characteristics of the elements $s_j$ in $S_j$.
-This makes it possible to create music made by pitches (both harmony and melody) and by temporal metrics and rhythms.
+(attack or onset) are generally the most important characteristics of the elements $s_j$ in $S$.
+This makes it possible to create music that relies heavily on pitches (both for harmony and melody) and on rhythms.
 We will start by considering these aspects of musical organization as they are more traditional in music theory and usually easier to understand.
+One will verify in the scripts that illustrate these concepts that it is considerably straightforward to
+reach sonic materials of some artistic significance by
+extrapolating musical paradigms established on traditional notions of melody, harmony and rhythm.
 
@@
 \subsubsection{Tuning}
-Doubling the frequency is equivalent to ascending one octave ($f=2f_0$). The octave division in twelve steps is the canon of classical western music. Its usage has also been observed
-outside western tradition, e.g.
-in ceremonial/religious and ethnic contexts~\cite{Wisnick}.
+Doubling the frequency is equivalent to ascending one octave\footnote{An pitch
+is both considered equivalent and different from any of its octaves.
+It is different because one or more octaves higher or lower corresponds to
+a higher or lower pitch.
+It is equivalent in the sense that a pitch is very consonant to its octaves
+and that any other pitch holds a similar relation to a pitch and its octaves.} ($f=2f_0$).
+The octave division in twelve steps is the canon of classical western music.
+Its usage has also been observed
+outside western tradition, e.g. in ceremonial/religious and ethnic contexts~\cite{Wisnick}.
 The semitones need not to be equivalent, as will become clear in the next paragraphs, but, roughly, the factor given by $\varepsilon=2^{\frac{1}{12}}$ defines a semitone.
@@
 Twelve successive semitones (or half steps), equidistants to the human ear, yield an
-octave, equivalent to the human ear. In other words, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
+octave. In other words, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
 $f_0$ and $f$.
@@
 This absolute accuracy is usual in computational implementations.
-Real music instruments, however, often present semitones that are not exactly $2^{\frac{1}{12}}$ because the (real) harmonics of the notes are not perfect multiples of the fundamental frequency.
+Real music instruments, however, often present semitones that are not exactly $2^{\frac{1}{12}}$
+for two main reasons: 1) the harmonic series does not match multiples of $2^{\frac{1}{12}}$,
+and 2) because the harmonics of real sounds are not perfect multiples of the fundamental frequency.
 The fixed interval $\varepsilon=2^{\frac{1}{12}}$ characterizes an equally tempered tuning but there are other tunings.
-The first formalizations of tunings (that the scientific tradition has repported) date from around two thousand years before the advent of the equal temperament~\cite{Roederer}.
-). Two emblematic tunings are:
+The first formalizations of tunings (in the scientific literature) date from around two thousand years before the advent of the equal temperament~\cite{Roederer}.
+Two emblematic tunings are:
 \begin{itemize}
@@
 
-In order to account for micro-tonality\footnote{Micro-tonality is the use of intervals smaller
+In order to account for microtonality\footnote{Microtonality is the use of intervals smaller
 than one semitone and has ornamental and structuring functionalities in music.
@@
 is proposed with the equal division of the octave in $53$ notes:
-$\varepsilon=2^{\frac{1}{53}}$.~\cite{microtonalidade}
-Note that if $S_i$ is a pitch sequence related by means of $\varepsilon_1$, the sequence with the same notes, but related by $\varepsilon_2$, is 
-$S_i'=\left\{s_i'\right\}=\left\{
+$\varepsilon=2^{\frac{1}{53}}$~\cite{microtonalidade}.
+Notice that if $S$ is a pitch sequence related by means of $\varepsilon_1$,
+the sequence with the same notes, but related by $\varepsilon_2$, is 
+$\left\{s_i'\right\}=\left\{
 s_i \frac{\varepsilon_1}{\varepsilon_2}\right\}$. The music piece \emph{Micro
-tone} uses microtonal features and its code
-is part of \massa\ toolbox.
+tone} uses microtonal features.~\cite{MASSA}
 
 \subsubsection{Intervals}\label{subsec:intervalos}
+The frequency gap between two pitches is called an "interval" in musical jargon.
 Using the ratio $\varepsilon=2^{\frac{1}{12}}$ between note frequencies (i.e.\ one semitone) the intervals in the twelve note system can be represented by integers.
@@
 Major (M) and minor (m) thirds and sixths are the imperfect consonances.
-Minor seconds and major sevenths are the harsh dissonances.
-Major seconds and minor sevenths are the mild dissonances.
+	Minor seconds and major sevenths are the strong dissonances.
+	Major seconds and minor sevenths are the weak dissonances.
 	Perfect fourth is a special case, as it is a perfect consonance when considered as an inversion of the perfect fifth and a dissonance or an imperfect consonance otherwise.
@@
 	This interval is consonant in some cultures.
-	For tonal music, the tritone indicates dominant (chord or harmonic field, see Section~\ref{subsec:harmonia}) and seeks urgent resolution into a third or sixth.
+	For tonal music, the tritone indicates the dominant (chord or harmonic field, see Section~\ref{subsec:harmonia}) and seeks urgent resolution into a third or sixth.
 Due to this instability it is considered a dissonant interval.}
@@
 \begin{itemize}
-	\item Intervals are inspected first by the number of steps between notes. The simple intervals are: first (unison), second, third, fourth, fifth, sixth, seventh and eighth (octave).
-		Each of these intervals are related to one step less the their names suggest: a third is an interval with two steps.
+	\item Intervals are inspected first by the number of steps between notes in some scale.
 		As can be noticed in Table~\ref{eq:intervalos}, one step is not one semitone.
@@
 		A scale for now can be regarded as any arbitrary monotonic sequence of pitches and will be discussed in the next section.
+		The simple intervals are: first (unison), second, third, fourth, fifth, sixth, seventh and eighth (octave).
+		Each of these intervals are related to one step less then their names suggest: a third is an interval with two steps.
 	\item The intervals are represented by numeric digits, e.g. 1, 3, 5 are a unison, a third and a fifth, respectively.
@@
                 i.e.\ unison, fourth, fifth and octave -- are 'perfect'.
-The imperfect consonances -- i.e.\ thirds and sixths -- and dissonances -- i.e.\ seconds and sevenths -- can be major and minor.
+		The imperfect consonances -- i.e.\ thirds and sixths -- and dissonances -- i.e.\ seconds and sevenths -- can be 'major' or 'minor'.
 		The tritone is an exception to this rule because it is a dissonant interval and cannot be major or minor.
 	\item The perfect fourth can be a perfect consonance or a dissonance according to the context and theoretical background.
-As a general rule, it can be considered a consonance except when it is followed by a third or a fifth by the movement one of the notes.
-	\item Tritone is a dissonance in Western music because
+		As a general rule, it is considered a consonance except when it is followed by a third or a fifth
+		by the movement of one of the notes.
+	\item The tritone is a dissonance in Western music because
 		it characterizes the ``dominant'' chord in the tonal system (see Section~\ref{subsec:harmonia}) and represents (or yields) instability.
@@
 aug3 has five semitones). The augmented forth
-is also called tritone (aug4 ~ tri ~ TT).
+is also called tritone (aug4, tri or TT).
 	\item A perfect interval or a minor interval (m2, m3, m6 or m7), decreased by one semitone results in a diminished interval. The
-diminished fifth is also called tritone (dim5 ~ tri ~ TT).
+diminished fifth is also called tritone (dim5, tri or TT).
 
 	\item An augmented interval increased by one semitone results in a `doubly augmented' interval and a diminished interval decreased by one semitone results in a `doubly diminished' interval.
-	\item Notes played simultaneously yields a harmonic interval.
-	\item Notes played as a sequence in time yields a
-                melodic interval.  When the lowest note comes first there is an ascending interval, while a descending interval is observed when the highest comes note first. 
+	\item Notes played simultaneously yield a harmonic interval.
+	\item Notes played as a sequence in time yield a
+                melodic interval.  When the lowest note comes first there is an ascending interval,
+		while in a descending interval the highest note comes first.
 	\item A simple interval is inverted if the lower pitch is raised one octave, or if
@@
                 inverted augmented interval results in a diminished interval
-                and vice-versa (inverting a doublyi-augmented results in a
+                and vice-versa (inverting a doubly-augmented results in a
                 doubly-diminished and vice-versa, etc).
                 An inverted perfect interval is a perfect interval as well.
-	\item An interval wider than an octave (e.g.
-ninth, tenth, eleventh) is called a 'compound interval' and is classified in terms of the interval between the same notes but in the same octave.
-Their notation can be achieved by adding a
-                7 to the interval: P11 is an octave plus a forth ($7 + P4 = P11$), M9 is an octave plus a major second ($7 + M2 = M9$).
+	\item An interval wider than an octave (e.g. ninth, tenth, eleventh) is called a 'compound interval' and is classified in terms of the interval between the same notes but in the same octave.
+		Their notation can be achieved by adding
+                7 to the corresponding simple interval, e.g. P11 is an octave plus a forth ($7 + P4 = P11$), M9 is an octave plus a major second ($7 + M2 = M9$).
 \end{itemize}
@@
 Henceforth, in a \textit{C flat} major scale, the tonic -- first degree -- is \textit{C flat}, not \textit{B}, and the leading tone -- seventh degree -- is \textit{B flat}, not \textit{A sharp} or \textit{C double flat}.
-To grasp what this entails for intervals, let the second degree (second note) of a scale to be one semitone from the first degree.
-Let also the leading tone (i.e.
-the seventh degree at one ascending semitone from the first degree).
-There is a diminished third between the seventh and second scale degrees~\cite{Lacerda}.
-Notice that the dim3 is only two semitones wide, as is the major second (or e.g. an doubly-augmented unisson!).
+To grasp what this entails for intervals, let the second degree (second note) of a scale be one semitone from the first degree.
+Consider also the leading tone (i.e. the seventh degree at one ascending semitone from the first degree). There is a diminished third between the seventh and second scale degrees~\cite{Lacerda}.
+Notice that the dim3 is only two semitones wide, as is the major second (or e.g. a doubly-augmented unison!).
 
 This description summarizes the traditional theory of musical intervals~\cite{Lacerda}.
-The music piece \emph{Intervalos entre alturas} explores these intervals in both independent and related manners.
-The source code is available online in the \massa\ toolbox~\cite{MASSA}.
+The music piece \emph{Intervals between pitches} explores these intervals in both independent and related ways.~\cite{MASSA}
 
 \subsubsection{Scales}\label{subsec:escalas}
-A scale is an ordered set of pitches.
-Strictly speaking, any (ordered) set of pitches can be considered a scale.
-The complexity about musical scales lean on the tradition, i.e.
+A scale is an ordered set of pitches. Strictly speaking, any (ordered) set of pitches can be considered a scale.
+The complexity about musical scales lean mostly on the tradition, i.e.
 on the scales and their uses which result from practice throughout history.
@@
 There are 5 perfectly symmetric divisions of the octave within the chromatic scale.
-These divisions are often regarded as scales themselves owing to the easy and peculiar uses they entail.
+These divisions are often regarded as scales themselves owing to the easy and peculiar uses they admit.
 
@@
 \begin{aligned}
-	\text{chromatic}    & = E_i^c    & = \{e_i^c\}_0^{11}   & =  \{0,1,2,3,4,5,6,7,8,9,10,11\} & = \{i\}_0^{11}\\
-	\text{whole tones}  & = E_i^{wt} & = \{e_i^{wt}\}_0^{5} & = \{0,2,4,6,8,10\}               & = \{2.i\}_0^{5} \\
-	\text{minor thirds} & = E_i^{mt} & = \{e_i^{mt}\}_0^{3} & = \{0,3,6,9\}                    & = \{3.i\}_0^3 \\
-	\text{major thirds} & = E_i^{Mt} & = \{e_i^{Mt}\}_0^{2} & = \{0,4,8\}                      & = \{4.i\}_0^2\\
-	\text{tritones}     & = E_i^{tt} & = \{e_i^{tt}\}_0^{1} & = \{ 0, 6 \}                     & = \{6.i\}_0^1
+	\text{chromatic}    & = E^c    & = & \{e_i^c\}_0^{11}   & = & \{0,1,2,3,4,5,6,7,8,9,10,11\} & = & \{i\}_0^{11}\\
+	\text{whole tones}  & = E^{wt} & = & \{e_i^{wt}\}_0^{5} & = & \{0,2,4,6,8,10\}               & = & \{2i\}_0^{5} \\
+	\text{minor thirds} & = E^{mt} & = & \{e_i^{mt}\}_0^{3} & = & \{0,3,6,9\}                    & = & \{3i\}_0^3 \\
+	\text{major thirds} & = E^{Mt} & = & \{e_i^{Mt}\}_0^{2} & = & \{0,4,8\}                      & = & \{4i\}_0^2\\
+	\text{tritones}     & = E^{tt} & = & \{e_i^{tt}\}_0^{1} & = & \{ 0, 6 \}                     & = & \{6i\}_0^1
 \end{aligned}
@@
 
-For example, the third note of the whole tone scale with $f_0=200Hz$ is $f_3=\varepsilon^{e_2^{wt}}. f_0 = 2^{\frac{4}{12}} . 200 \approxeq 251.98
+For example, the third note of the whole tone scale with $f_0=200Hz$ is $f_3=\varepsilon^{e_2^{wt}}. f_0 = 2^{\frac{4}{12}} . 200 \approx 251.98
 Hz$.
@@
 Section~\ref{estCic} discusses symmetries.
-The music piece \emph{Cristais} uses each one of these scales, in both melodic and harmonic counterpart and their source code is part of \massa\ toolbox.
+The music piece \emph{Cristais} uses each one of these scales both melodically and harmonically.~\cite{MASSA}
 
@@
 \begin{aligned}
-	\text{aeolian}    & = \text{natural minor scale}  & = & \\
-			  & = E_i^m = \{e_i^m\}_0^6       & = & \{0,2,3,5,7,8,10\} \\
-	\text{locrian}    & = E_i^{lo} = \{e_i^{lo}\}_0^6 & = & \{0,1,3,5,6,8,10\} \\ 
-	\text{ionian}     & = \text{major scale}          & = &  \\
-			  & = E_i^M = \{e_i^M\}_0^6       & = & \{0,2,4,5,7,9,11\} \\
-	\text{dorian}     & = E_i^{d} = \{e_i^{d}\}_0^6   & = & \{0,2,3,5,7,9,10\} \\
-	\text{phrygian}   & = E_i^{p} = \{e_i^{p}\}_0^6   & = & \{0,1,3,5,7,8,10\} \\
-	\text{lydian}     & = E_i^{l}=\{e_i^{l}\}_0^6     & = & \{0,2,4,6,7,9,11\} \\
-	\text{mixolydian} & = E_i^{mi} = \{e_i^{mi}\}_0^6 & = & \{0,2,4,5,7,9,10\}
+	\text{Aeolian}    & = \text{natural minor scale}  & = & \\
+			  & = E^m = \{e_i^m\}_0^6       & = & \{0,2,3,5,7,8,10\} \\
+	\text{Locrian}    & = E^{lo} = \{e_i^{lo}\}_0^6 & = & \{0,1,3,5,6,8,10\} \\ 
+	\text{Ionian}     & = \text{major scale}          & = &  \\
+			  & = E^M = \{e_i^M\}_0^6       & = & \{0,2,4,5,7,9,11\} \\
+	\text{Dorian}     & = E^{d} = \{e_i^{d}\}_0^6   & = & \{0,2,3,5,7,9,10\} \\
+	\text{Phrygian}   & = E^{p} = \{e_i^{p}\}_0^6   & = & \{0,1,3,5,7,8,10\} \\
+	\text{Lydian}     & = E^{l}=\{e_i^{l}\}_0^6     & = & \{0,2,4,6,7,9,11\} \\
+	\text{Mixolydian} & = E^{mi} = \{e_i^{mi}\}_0^6 & = & \{0,2,4,5,7,9,10\}
 \end{aligned}
@@
 % This is the main reason why the locrian is sometimes not considered a diatonic scale.
-
 Diatonic scales follow a circular pattern of successive intervals \textit{tone, tone, semitone, tone, tone, tone, semitone}. Thus, it is possible to write:
@@
 For example, a brief inspection reveals that
-$e_i^{l}=d_{(i+2)\%7}+e_{i-1}^{l}$. Then, $\kappa=2$ for the lydian scale.
+$e_i^{l}=d_{(i+2)\%7}+e_{i-1}^{l}$. Then, $\kappa=2$ for the Lydian scale.
 
@@
 \begin{split}
-\text{natural minor}&  = E_i^m = \{e_i^m\}_0^6 = \\
+\text{natural minor}&  = E^m = \{e_i^m\}_0^6 = \\
                                            &  = \{0,2,3,5,7,8,10\} \\
-\text{harmonic minor}                      &  = E_i^{mh} = \{e_i^{mh}\}_0^6 = \\
+\text{harmonic minor}                      &  = E^{mh} = \{e_i^{mh}\}_0^6 = \\
                                            &  = \{0,2,3,5,7,8,11\} \\
-\text{melodic minor}                       &  = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \\
+\text{melodic minor}                       &  = E^{mm} = \{e_i^{mm}\}_0^{14} = \\
                                            &  = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
@@
 
-The different ascending and descending contour of the melodic minor scale is required in tonal harmony contexts.
-The minor scale has one whole tone between the seventh and eigth (or first) degrees but
+The different ascending and descending contours of the melodic minor scale is required in tonal harmony contexts.
+The minor scale has one whole tone between the seventh and eighth (or first) degrees but
 the separation by one semitone is critical to the polarization of the first degree.
@@
 \begin{split}
-H_i = & \{h_i\}_0^{19}= \\
+H = & \{h_i\}_0^{19}= \\
     =  & \{ 0,12,19+0.02,  24,28-0.14, 31+0.2, 34-0.31, \\
@@
 Many other scales can be expressed using the framework exposed in this section, e.g. the pentatonic scales and the modes of limited transposition of Messiaen~\cite{Messiaen}.
-
 One last observation: the words \emph{scale} and \emph{mode} are often used as synonyms both in the literature and in colloquial discussions.
@@
 \begin{split}
-\text{major triad} = A_i^M= \{a_i^M\}_0^2=\{0,4,7\} \\ 
-\text{minor triad} = A_i^m = \{a_i^m\}_0^2=\{0,3,7\} \\
-\text{diminished triad} = A_i^d = \{a_i^d\}_0^2=\{0,3,6\} \\
-\text{augmented triad} = A_i^a = \{a_i^a\}_0^2=\{0,4,8\}
+\text{major triad} = A^M= \{a_i^M\}_0^2=\{0,4,7\} \\ 
+\text{minor triad} = A^m = \{a_i^m\}_0^2=\{0,3,7\} \\
+\text{diminished triad} = A^d = \{a_i^d\}_0^2=\{0,3,6\} \\
+\text{augmented triad} =  A^a = \{a_i^a\}_0^2=\{0,4,8\}
 \end{split}
@@
 To have another third superimposed to the fifth, it is sufficient to add $10$ as the highest note in order to form a tetrad with minor seventh, or add $11$ in order to form a tetrad with major
-seventh. Inversions and open positions can be obtained with the 
-addition of $\pm 12$ to the selected component. Incomplete triadic chords, with extra notes ('dirty' chords), and non-triadic
+seventh. Inversions and open positions can be obtained by the 
+addition of $\pm 12$ to the selected pitch. Incomplete triadic chords, with extra notes ('dirty' chords), and non-triadic
 are also common.
@@
 		There are theoretical discussion about why this happens, and the most usual arguments are that the fifth is the first (non-octave) harmonic of a note and that the harmonics of the fifth are in the harmonics of the fundamental.
-		Important here is to grasp the fact that musical theory and practice assures that the fifth establishes the fundamental as the root of a chord.
+		Important here is to grasp the fact that musical theory and practice assures that the fifth enhances the sensation that
+		the root of a chord is the root.
         \item Major or minor thirds from the root entails major or minor chord qualities.

### 2017-08-12 article/acm/body.tex
finished review until the end cyclic structures

@@
 
-\subsection{Atonal and tonal harmonies, harmonic expansion and modulation}\label{subsec:harmonia}
-Omission of basic tonal structures is the key to achieving modal and atonal harmonies. In the absence of minimal tonal organization,
+\subsection{Tonal, modal and atonal music, harmonies, harmonic expansion and modulation}\label{subsec:harmonia}
+The omission of basic tonal structures is the key to achieving modal and atonal harmonies\footnote{To avoid very complicated discussions,
+the term 'harmony' is used in this article in a very loose sense:
+considerations, grouped by music theory as harmony, about how sounds are analyzed by hearing.
+These considerations stem substantially from chords and scales and the absence of them.}. In the absence of minimal tonal organization,
 harmony is (usually) considered modal if the notes match with some diatonic scale (see Equations~\ref{eq:escalas}) or if there is only a small number of notes.
-If basic tonal progressions are absent and notes do not match any diatonic scale and are enough diverse and dissonant (between themselves) to avoid reduction of the notes by polarization\footnote{By polarization we mean having some notes that are way more important than others and to which the other notes are ornaments or subordinates.}, the harmony is atonal.
+If basic tonal progressions are absent and notes do not match any diatonic scale and are diverse enough and dissonant (between themselves) to avoid reduction of the notes by polarization\footnote{By polarization we mean the presence of notes that are way more important than others and to which the other notes are ornaments or subordinates.}, the harmony is atonal.
 In this classification, the modal harmony is not tonal or atonal and is reduced to the incidence of notes within a (most often diatonic) scale and to the absence of tonal structures.
-Following this concept, one observes that atonal harmony is hard to be realized and, indeed, no matter how dissonant and diverse
+Atonal harmony is hard to be fulfilled and, indeed, no matter how dissonant and diverse
 a set of notes is, tonal harmonies arise very easily if not avoided~\cite{harmEXT}.
@@
 \subsubsection{Atonal harmony}\label{sec:atonal}
-In fact, atonal music techniques avoid that a direct relation of the notes with modes and tonality be established.
+In fact, atonal music techniques aim at avoiding that a direct relation of the notes with modes and tonality be established by the listener.
 Manifesting such atonal structures is of such difficulty that the dodecafonism emerged.
-The purpose of dodecafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
+Dodecafonism consists in using a sequence of pitches (ideally the 12 pitches of the chromatic scale),
+and performing each pitch, one by one, in the same
 order.
@@
 Nevertheless, the western listener automatically searches for tonal elements in music and obstinately finds them by unexpected and tortuous paths.
-The use of dissonant intervals (especially tritones) without resolution reinforces the absence of tonality.
-In this context, while creating a musical
+The use of dissonant intervals (especially tritones) without resolution reinforces the absence of tonality. In this context, while creating a musical
 piece, it is possible to:
 \begin{itemize}
-     \item Repeat notes. By considering immediate repetition as an extension of the previous incidence, the use of the same note in sequence does not add relevant information.
-     \item To play adjacent notes (e.g. of a dodecafonic progression) at the same time, making harmonic intervals and chords.
-     \item Use durations and pauses with liberty, respecting notes order.
+     \item Repeat pitches. By considering an immediate repetition as an extension of the previous incidence, the use of the same pitch in sequence does not add relevant information.
+     \item To play adjacent pitches (of the dodecafonic sequence) at the same time, making harmonic intervals and chords.
+     \item Use durations and pauses with freedom while respecting the order of incidence of the pitches.
      \item Vary note sequences by temporal magnification and translation or pitch transposition and by pitch sequence inversion, retrograde and retrograde inversion. See Sections~\ref{subsec:motivos}
-     and~\ref{subsec:usosmusicais3} for what these terms mean.
-     \item Make variations in orchestration, articulation, spatialization, among other possibilities in presenting the same notes.
+     and~\ref{subsec:usosmusicais3} for the meaning of these terms.
+     \item Make variations in timbre (e.g. orchestration), articulation, spatialization, among other possibilities in the presentation of the same pitches.
 \end{itemize}
 
-The atonal harmony can be observed, paradigmatically, within these presented conditions (which is a simple dodecaphonic model). Most of what was written by great dodecafonic composers,
+The atonal harmony can be observed, paradigmatically, within these presented conditions (which is a simple dodecaphonic model).
+Part of what was written by great dodecafonic composers,
 e.g. Alban Berg and even Schoenberg, had the purpose of mixing tonal and atonal techniques.
-Most frequently, atonal music is not strictly dodecafonic, but "serial", i.e. they use the same kind of techniques based in a sequence of notes (called the series or row).
+Most frequently, atonal music is not strictly dodecafonic, but "serial", i.e.
+they use the same kind of techniques as stated above, based in a sequence of pitches (called the series or row), but not within a 12 tone sequence and using sequences of other parameters (e.g.
+duration, timbre).
+Serialism and pos-serialism is briefly considered in Secion~\ref{sec:ser}.
 
@@
 In the XX century, music with emphasis on sonorities/timbres and rhythm, extended the concepts of tonality and harmony.
-Even so, tonal harmony is very often in artistic movements and commercial venues.
+Even so, tonal harmony is very common in artistic movements and commercial venues.
 In addition, dodecafonism itself is sometimes considered of tonal nature because it was conceived to deny tonal characteristics of polarization.
@@
 Equations~\ref{triades}-- built on the root note of each degree of a scale  -- such as listed in Equations~\ref{eq:escalas} --  form the pillars of harmony.
-Tonal (and modal) harmony aims at understanding the incidence of chord progressions and chaining rules.
-Even a monophonic melody entails harmonic fields, making it possible to perceive the chord progression even in unaccompanied melodies.
+Tonal (and modal) harmony theory aims at understanding the incidence of chord progressions and chaining rules.
+Even a monophonic melody entails harmonic fields, making it possible to perceive a chord progression even in unaccompanied melodies.
 
@@
 The remaining triads are associated to these three most important chords. In the major scale, the associated relative (relative tonic \textit{Tr}, relative
-subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built a third below, and the associated counter-relative (counter-relative tonic \textit{Ta}, counter-relative subdominant \textit{Sa} and the counter-relative dominant \textit{Da}) is the triad built in a third above.
-In the minor scale the same happens, but the triad a third below is called counter-relative (tA, sA) and the triad a third above is called relative (tR,
+subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built a third below, and the associated counter-relative (counter-relative tonic \textit{Tc}, counter-relative subdominant \textit{Sc} and the counter-relative dominant \textit{Dc}) is the triad built in a third above.
+In the minor scale the same happens, but the triad a third below is called counter-relative (tC, sC) and the triad a third above is called relative (tR,
 sR). The precise functions and musical effects of these chords are
@@
 \centering
-\caption{Summary of tonal harmonic functions on the major scale. Tonic is the musical center, the dominant goes to the tonic and the subdominant moves the music away from
+\caption{Summary of tonal harmonic functions on the major scale.
+	The tonic is the musical center, the dominant goes to the tonic and the subdominant moves the music away from
 the tonic. The three chords can, in principle, be freely replaced by their
-respective relative or counter-relatives.}
+respective relatives or counter-relatives.}
 \begin{tabular}{l | c | r}
@@
 on a fifth above or a fifth below, respectively. These individual dominants and subdominants,
-in the same way, have also subdominants and individual dominants of their own. Given a tonality, any chord can occur, no matter
+in the same way, also have subdominants and dominants of their own. Given a tonality, any chord can occur, no matter
 how distant it is from the most basic chords and from the notes of the scale.
@@
 There are two mediants, or 'chromatic mediants', for each chord: the upper
-mediant, formed with the root at the third of original chord; and the lower mediant, formed by the fifth at the third of original chord.
-Both chords also are formed by a third, but with a chromatic alteration regarding the original chord.
-If two chromatic alterations exist, i.e.\ two notes altered by one semitone each regarding the original chord, it is a 'doubly chromatic mediant'.
-Again, there are two forms for each chord: the upper form, with a third in the fifth of the original triad; and the lower form, with a third in the root
+mediant, with the root at the third of the original chord;
+and the lower mediant, with the fifth at the third of original chord.
+Both chords also are separated by a third from the original chord, like the relatives and counter-relatives,
+but with a chromatic alteration regarding the original chord.
+If two chromatic alterations exist, i.e.\ two notes altered by one semitone each regarding the original chord,
+it is a 'doubly chromatic mediant'.
+Again, there are two forms for each chord: the upper form, with the third on the fifth of the original triad;
+and the lower form, with the third on the root
 of the original triad.
@@
 (Recall that relatives and counter-relatives have opposite major-minor quality.)
-This relation between chords is considered advanced tonalism, sometimes even considered as an expansion and dissolution of tonalism,
+Mediants are considered of advanced tonalism,
+sometimes even considered as an expansion and dissolution of tonalism,
 with strong and impressive effects although they are simple, consonant triads.
-Chromatic mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss, among others~\cite{Harmonia,Salzer}.
+Chromatic mediants are used since the late 19th century by Wagner, Lizt, Richard Strauss, among others~\cite{Harmonia,Salzer}.
 
@@
 Keys are always (thought of as) related by fifths and their relatives and counter-relatives. Some ways to perform modulation include:
-
 \begin{itemize}
-    \item Transposing the discourse to a new key, without any preparation. It is a common Baroque procedure although incident in other periods as well. Sometimes it is 
+    \item Transposing the discourse to a new key, without any preparation.
+	    It is a common Baroque procedure although incident in other periods. Sometimes it is 
     called phrasal modulation or unprepared modulation.
     \item Careful use of an individual dominant, and perhaps also the individual
-    subdominant, to confirm change in key and harmonic field.
+    subdominant, to confirm a change in key.
     \item Use of chromatic alterations to reach a chord in the new key by starting from a chord in the previous key. Called chromatic modulation.
@@
     \item Changing the function, without changing the notes, of a chord to
-    contemplate a new key. This procedure is called enharmony.
+    reach a new key. This procedure is called enharmony.
     \item Maintaining the tonal center and changing the key quality from major to minor
@@
 The dominant has great importance and is a natural pivot in modulations,
-a fact that leads to the circle of fifths~\cite{Harmonia,Salzer,Koellheuteur,Harmony}.
-Other inventive ways to modulate are possible, to point but one common example, the minor thirds tetrad ($E_i^{tm}$ in Equations~\ref{escSim}) can be sustained to bridge to other tonalities, with the ease of its both tritones.
-The music piece \emph{Acorde cedo} explores these chord relations, and is implemented online as part of \massa\ toolbox~\cite{MASSA}.
+a fact that emphasizes the relevance of the 'circle of fifths'~\cite{Harmonia,Salzer,Koellheuteur,Harmony}.
+Other inventive ways to modulate are possible, to point but one common example, the minor thirds tetrad ($E_i^{tm}$ in Equations~\ref{escSim}) can be sustained as a bridge between tonalities, with the possible use of its both tritones.
+The music piece \emph{Acorde cedo} explores these chord relations.~\cite{MASSA}.
 
 \subsection{Counterpoint}\label{subsec:contraponto}
-Counterpoint is the conduction of simultaneous melodic lines, or ``voices''.
-The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues~\cite{Fux,SchoenbergContra}.
-It is possible to
+Counterpoint is a set of techniques for the conduction of simultaneous melodic lines, or ``voices''.
+The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues~\cite{Fux,SchoenbergContra}. It is possible to
 summarize counterpoint rules, and it is known that Beethoven --
@@
 neighbor note in the scale sound coherent and are prioritized. When having 3 or more voices, the melodic
-importance lies mainly in the highest and then in the lowest of the voices~\cite{Fux,Tragtenberg,SchoenbergContra}.
-
-These rules were used in the music piece \emph{Conta ponto}, whose source code is
-available online in \massa\ toolbox.
+relevance lies mainly in the highest and then in the lowest of the voices~\cite{Fux,Tragtenberg,SchoenbergContra}.
+These rules were used in the music piece \emph{Conta ponto}.~\cite{MASSA}
 
 \subsection{Rhythm}\label{subsec:ritmo}
-Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}, which can be heard individually if spaced by at least $50-63ms$.
-For the temporal separation between them to be perceived as duration, the period should even a bit larger, around $100ms$~\cite{microsound}.
+Rhythm is dependent on events separated by durations~\cite{Lacerda}.
+For the temporal separation between them to be perceived as duration,
+the period should be at least around $75ms$~\cite{microsound}.
 It is possible to summarize the durations heard as rhythm or pitch, and its transition,
@@
 
-The transition span in Table~\ref{tab:duracoes} is minimized because the limits
-are not well defined. In fact, the duration where someone begins to perceive a
-fundamental frequency or a separation between occurrences, depends on the
-listener and sonic characteristics~\cite{microsound,Roederer}. The rhythmic metric is commonly based on a basic duration called pulse, whose typical durations range between $0.25 and 1.5s$ ($240$
-and $40 BPM$, respectively~\footnote{BPM stands for Beats Per Minute and is just a frequency measure like Herz, but is the number of incidences per minute intead of second.
-BPM is often used as a measure of musical tempo and of heart rate.}).
-In music education and cognitive studies, it is common to associate this range of frequencies with the durations of the heart beat, movements of respiration and steps of a walking or running person~\cite{Lacerda,Roederer}.
-
-The pulse is subdivided into equal parts and is also repeated in sequence. These relations (division and concatenation) usually follow relations of small
-integers. By far, the most often musical pulse divisions (and their sequential groupings), in written and ethnic
-music, are: 2, 4 and 8; 3, 6 (two groups of 3 or 3 groups of 2), 9 and 12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, completing
-1-9 and 12.
-Other metrics are less common, like division or grouping in 13, 17, etc, and are mainly used in experimental music or classical music of the XX and XXI centuries.
-No matter how complex they seem, metrics are almost always compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
-A schematic illustration is shown in Figure~\ref{fig:pulsoSubAgl}.
-
 \begin{figure*}
@@
         pulse, is presented on the left. The time signature yielded by
-        groupings of the music pulse is presented on the right.}
+	groupings of the music pulse is presented on the right~\protect\footnotemark.}
         \label{fig:pulsoSubAgl}
 \end{figure*}
+
+\footnotetext{The
+	definition of the terms 'quarter note' and 'time signature' are beyond the scope of this work and might be found in elementary texts about musical theory~\cite{Lacerda}.
+	Anyway, what is important for the reader to grasp is that: 1) successive temporal elements of equal duration
+	are both grouped and subdivided into other elements of equal duration;
+	2) these groupings and subdivisions are usually made in terms of small integers and are discerned into
+	simple, compound and complex.}
+
+The transition span in Table~\ref{tab:duracoes} is minimized because the limits
+are not well defined,
+i.e. the exact duration that is perceived as a
+frequency or a separation between occurrences depends on the
+listener and the sonic characteristics~\cite{microsound,Roederer}.
+The rhythm is commonly based on a basic duration called pulse, whose typical duration range between $0.25$ and $1.5s$ ($240$
+and $40 BPM$, respectively~\footnote{BPM stands for Beats Per Minute and is just a frequency measure like Herz, but is the number of incidences per minute instead of second.
+BPM is often used as a measure of musical tempo and of heart rate.}).
+In music education and cognitive studies, this range of frequencies is often associated with the duration of the heart beat, movements of respiration and steps of a walking or running person~\cite{Lacerda,Roederer}.
+
+The pulse is subdivided into equal parts and is also repeated in sequence. These relations (division and concatenation) usually follow relations of small
+integers. By far, the most often musical pulse divisions (and their sequential groupings), in written, popular and ethnic
+music, are: 2, 4 and 8; 3, 6 (2 groups of 3 or 3 groups of 2), 9 and 12 (3 and 4 groups of 3). At last, the prime numbers 5 and 7, completing
+1-9 and 12. Other metrics are less common, like division or grouping in 13, 17, etc, and are mainly used in experimental music or classical music of the XX and XXI centuries.
+A schematic illustration is shown in Figure~\ref{fig:pulsoSubAgl}.~\cite{Gramani,Roederer}
 
@@
 weak, weak) and weak (also subdivided in strong, weak, weak). Another way to perform
-the division in 6 is ternary division units that subdivides as binary,
+the division in 6 is through a ternary division that subdivides as binary,
 resulting in: a strong unit (subdivided in strong and weak) and two weak units
@@
 
-An accent in the weak beat is the `backbeat', whereas a note starting in a weak beat persisting across the strong beat is a `syncope'.
+An accent in the weak beat is the `backbeat', whereas a note starting in a weak beat and persisting across the strong beat is a `syncope'.
 These are often found in ethnic and popular music and was used with parsimony in classical music before the XX century.
@@
 Let the pulse be the grouping level $j=0$, the first pulse subdivision be level $j=-1$,
-the first pulse agglomeration be level $j=1$ and so on.
-Accordingly, let $P_i^j$ be the $i$-th unit at grouping level $j$: $P^0_{10}$ is the tenth pulse, $P^{1}_3$ is the third grouped unit (possibly the third measure),
+the first pulse agglomeration be level $j=1$ and so on. Accordingly, let $P_i^j$ be the $i$-th unit at grouping level $j$: $P^0_{10}$ is the tenth pulse, $P^{1}_3$ is the third grouped unit,
 $P^{-1}_2$ is the second part of pulse subdivision.
@@
 As an example, consider $P^{-1,0,1}_{3,2,2}$ as the third subdivision $P^{-1}_3$ of the
-second pulse $P^0_2$ and of the second pulse group $P^1_2$ (possibly second measure). Each unit $P_i^j$ can be associated with a sequence of temporal samples $T_i$ that constitutes a
-note. The music piece \emph{Poli Hit Mia} uses different metrics (available as part of \massa\ toolbox).
+second pulse $P^0_2$ and of the second pulse group $P^1_2$.
+Each unit $P_i^j$ can be associated with a sequence of temporal samples $T$ that constitutes a
+note. The music piece \emph{Poli Hit Mia} uses different rhythmic metrics.~\cite{MASSA}
 
 \subsection{Repetition and variation: motifs and larger units}\label{subsec:motivos}
-Given the basic musical structures, both frequential (chords and scales) and rhythmic (simple, compound and complex beat divisions and agglomerations), it is
+Given the basic musical structures of the previous sections, it is
 natural to present these structures in a coherent and meaningful way~\cite{Boulez}.
-The concept of arcs is essential in this context: by departing from a place and returning, an arc is made.
+The concept of arc is essential in this context: by departing from a place and returning, an arc is made.
 One important case is the arc from and to the absence of a unit: from the beginning to the end.
@@
 Accordingly, the note can be considered the smallest (relevant) arc, and each motif and melody as an arc as well.
-Each beat and subdivision, each measure and music
-section, constitutes an arc. Music in which the arcs do not present consistency with one another can be understood as music with no coherence. Coherence impression
+Each beat and subdivision, each music and
+section, constitutes an arc.
+Music in which the arcs do not present consistency with one another can be understood as music with little or no coherence. Coherence impression
 comes, mostly, from the skilled handling of arcs in a music piece.
@@
 prone to variations. Let
-$S_j=\left\{s_j=T_i^j=\{t_i^{j}\}_0^{\Lambda_j-1}\right\}_0^{H-1}$ be a sequence
+$S=\left\{s_j=T^j=\{t_i^{j}\}_0^{\Lambda_j-1}\right\}_0^{H-1}$ be a sequence
 of $H$ musical events $s_j$, each event with its $\Lambda_j$ samples $t_i^j$
@@
     where $\Gamma$ is the duration between the beginning of the piece (or considered
-    section) and the first event $s_0$ of original structure $S_j$, and
-    $\delta$ is the time offset of the displacement.
+    section) and the first event $s_0$ of the original structure $S$, and
+    $\delta$ is the temporal offset of the displacement.
 
@@
     \item Temporal reversion consists on generating a sequence with elements
-    in the reverse order of the original sequence $S_j$, thus: $S_j'=\left\{s_j'\right\}_0^{H-1}=\left\{s_{(H-j-1)}\right\}_0^{H-1}$.
+    in the reverse order of the original sequence $S$, thus:
+		$S'=\left\{s_j'\right\}_0^{H-1}=\left\{s_{(H-j-1)}\right\}_0^{H-1}$.
 
     \item Pitch translation, or transposition, is a displacement $\tau$ of the material to a
-        different pitch $\Xi=\Xi_0 + \tau$. It is a variation that changes pitch
+        different pitch $\Xi'=\Xi + \tau$. It is a variation that changes pitch
         localization of material:
-        $\left\{s_j'\right\}=\left\{s_j^{\Xi'}\right\}=\left\{s_j^{\Xi+\tau}\right\}$
-        where $\Xi_0$ is a reference value, such as the pitch of a section $S_j$ or of the first event $s_0$.
-        If $\tau$ is given in semitones, the displacement in
-        frequency is $\tau_f=f_0.2^{\frac{\tau}{12}}$ where $f_0$ is the
-        reference frequency value: $f_0=\Xi_{f_0}Hz\;\sim \Xi$ absolute value of pitch. The frequency of any pitch value is $f=f_0.2^{\frac{\tau}{12}}$s; 
-        and the pitch of any frequency is: $\Xi=\Xi_0 +12
-		\log_2\left(\frac{f}{f_0}\right)$.\footnote{In the MIDI protocol, $\Xi_{f_0}=55Hz$ when pitch $\Xi_0=33$
-	(an \textit{A1} note). Another good MIDI reference is $\Xi_{f_0}=440Hz$ and
-		$\Xi_0=69$ (\textit{A4}). The difference ($\Xi_1 - \Xi_2$) is in semitones.
-        $\Xi$ is not a measure in semitones: $\Xi=1$ is not a semitone, it is a note with an audible frequency as rhythm, with
+        $\left\{s_j'\right\}=\left\{s_j^{\Xi'}\right\}=\left\{s_j^{\Xi+\tau}\right\}$.
+        If $\tau$ is given in semitones, new frequencies are given by $f'=f2^{\frac{\tau}{12}}$; 
+        and the new pitches: $\Xi'=\Xi +12 \log_2\left(\frac{f'}{f}\right)=\Xi+\tau$.\footnote{In
+		the MIDI protocol, $\Xi_{f}=55Hz$ when pitch $\Xi=33$
+	(an \textit{A1} note). Another good MIDI reference is $\Xi_{f}=440Hz$ and
+		$\Xi=69$ (\textit{A4}). The difference ($\tau=\Xi_1 - \Xi_2$) is in semitones.
+        Notice that $\Xi=1$ is not a semitone, it is a note with an audible frequency as rhythm, with
 		less than 9 occurrences each second (see Table~\ref{tab:duracoes}).}
 
-        \item Interval inversion is either: 1) the inversion of note pitch order, within octave equivalence, such as described in Section~\ref{subsec:intervalos}, and $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{\varepsilon.
-f_0}\right\}$ with selective $s_j=2,\;1/2,\;\text{or}\;1$; or 2) the inversion of interval orientation.
+        \item Interval inversion is either: 1) the inversion of note pitch order, within octave equivalence, such as described in Section~\ref{subsec:intervalos}; or 2) the inversion of interval orientation.
 In the former case, the number of semitones
         is preserved in the ``strict invertion'':
-        $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{-\varepsilon_j . f_0}\right\}$, where
+        $S'=\{s_j'\}_0^{H-1}=\left\{s_j^{-\varepsilon_j . f_0}\right\}$, where
         $\varepsilon_j$ is the factor between the frequency of event $s_j$ and the
@@
         considered in terms of the diaonic scale $E_k$:
-        $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{2^{\left(\frac{12-e_{\left(7-j_e\right)}}{12}\right)}
+        $S'=\{s_j'\}_0^{H-1}=\left\{s_j^{2^{\left(\frac{12-e_{\left(7-j_e\right)}}{12}\right)}
         . f_0}\right\}_0^{H-1}$ where $j_e$ is the index in
@@
         with events which are out of the slots. Thus, a
-        rotation of $\tilde{n}$ positions is $S'_n=S_{(n+\tilde{n})\%H}$. If
-        $\tilde{n}<0$, it is sufficient to use $\tilde{n}'=H-\tilde{n}$. It is
-        usual to associate $\tilde{n}<0$ (events advance) with the clockwise rotation and
-        $\tilde{n}<0$ (elements delay) with the anti-clockwise rotation.
+        rotation of $\tilde{n}$ positions is $s'_n=s_{(n+\tilde{n})\%H}$. If
+        $\tilde{n}<0$, it is sufficient to use $\tilde{n}'=H-\tilde{n}$.
+		One might associate $\tilde{n}>0$ (events advance) with the anti-clockwise rotation and
+        $\tilde{n}<0$ (elements delay) with the clockwise rotation.
         Additional information about rotations is given in Section~\ref{estCic}.
 
-        \item The insertion and removal of material in $S_j$ can be
-    ornamental or structural: $S_j'=\{s_j'\}=\{s_j \text{ if condition A,
+        \item The insertion and removal of material in $S$ can be
+    ornamental or structural: $S'=\{s_j'\}=\{s_j \text{ if condition A,
     otherwise } r_j\}$, for any music material $r_j$, including silence.
-    Elements of $R_j$ can be inserted at the beginning, like a prefix
-    in $S_j$; at the end, as a suffix; or in the middle, dividing $S_j$ and making
-    it the prefix and suffix. Both materials can be mixed in a variety of ways.
+    Elements of $R$ can be inserted at the beginning, as a prefix;
+		at the end, as a suffix; or in the middle, dividing $S$ and making
+    it the prefix and suffix of $R$. Both materials can be mixed in a variety of ways.
 
@@
 
-    \item Accompaniment. Both orchestration and melodic lines present when $S_j$ occurs can suffer modifications and be considered as a variation of $S_j$ itself.
+    \item Accompaniment.
+	    E.g. orchestration and melodic lines present when $S$ occurs can 
+	    be modified and be considered a variation of $S$ itself.
 \end{itemize}
 
-From these processes, many others are derived, such as the inverted retrograde, the temporal contraction with an external suffix, etc.
+From these processes many others are derived, such as the inverted retrograde, the temporal contraction with an external suffix, etc.
 Variations are often thought about in the terms above but are also often very loose, such as an arbitrary shuffle of the notes in a melody which the composer finds interesting.
@@
 musical therapy, known by its utility in cases of depression and neurological injury.
-Also, it is known that regions of the human brain responsible for sonic processing are also used for other activities, such as for performing verbal discourse and mathematics~\cite{Sacks,Roederer}.
+Also, it is known that regions of the human brain responsible for sonic processing are also used for other activities, such as for performing verbal discourse and mathematics.~\cite{Sacks,Roederer}
 
@@
 equivalence/difference, arrival/departure, near/far, stationary/moving,
-etc. All these dipoles are often thought of as parallel or even as equivalent. Ternary relations tend to relate to the circle and to unification. The
+etc. All these dipoles are often thought of as parallel or even as equivalent.
+Ternary constructions tend to relate to the circle and to unification. The
 lucid ternary communion, `modus perfectus', opposes to the passionate
@@
 directional and cyclic arcs.
-
 For a scholastic discussion on the composition of motives, phrases, melodies, themes and music form (such as rondo, ternary, theme and variations) see~\cite{Schoenberg}.
@@
 shown in Figure~\ref{fig:climax}. The varying parameter can be non-existent, a case in which
-the arc consists only of a reference structure~\cite{Schoenberg}\footnote{A case which resembles a note without the fundamental frequency.
-Brass instruments can emit such notes and the listener identifies the fundamental frequency even though it is absent.}.
+the arc consists only of a reference structure~\cite{Schoenberg}\footnote{A
+case which resembles a note without the fundamental frequency. In some cases, the listener is able to identify
+the fundamental frequency even though it is absent.}.
 
@@
         other arcs. The possibilities considered are: climax at the beginning, at the first half, in the middle, in the second half and 
-        at the end. The x and y-axis are not properly specified since the parameters can be non-existent, as in a reference structure.}
+        at the end.}
         \label{fig:climax}
@@
 
-Consider the $S_i=\{s_i\}_0^{H-1}$ increasing sequence. The sequence
-$R_i=\{r_i\}_0^{2H -2}=\left\{s_{(H-1-|H-1-i|)}\right\}_0^{2H-2}$
+Consider the $S=\{s_i\}_0^{H-1}$ increasing sequence. The sequence
+$R=\{r_i\}_0^{2H -2}=\left\{s_{(H-1-|H-1-i|)}\right\}_0^{2H-2}$
 presents perfect specular symmetry, i.e.\ the second half is the
@@
 
-In practice, and historically, there is special incidence and use of the golden ratio.
-The Fibonacci sequence might be generalized as follows in order for any two numbers be used
+In practice, and historically, there is special use of the golden ratio.
+The Fibonacci sequence might be generalized as follows in order to use any two numbers
 to approach the golden ratio.
@@
 for exploring various sequences.
-
-The music piece \emph{Dirracional} uses arcs into directional structures. Its source code is available online as part
-of \massa\ toolbox~\cite{MASSA}.
+The musical piece \emph{Dirracional} explores directional arcs,~\cite{MASSA}
 
@@
      &  \text{(existence of the identity element)} \\
-\forall \;\; p \in G, \;\exists\; p^{-1} :\quad  &  p\bullet p^{-1}i =p^{-1}\bullet p = e \\
+\forall \;\; p \in G, \;\exists\; p^{-1} :\quad  &  p\bullet p^{-1} =p^{-1}\bullet p = e \\
      &  \text{(existence of the inverse element)}
@@
 From the first property follows that two permutations act as one permutation. In fact, it is possible to apply a
-permutation $p_1$ and another permutation $p_2$, and, comparing both initial and final orderings, observe another permutation $p_3$.
+permutation $p_1$ and then another permutation $p_2$, and, comparing both initial and final orderings, observe another permutation $p_3$.
 Every element $p$ operated with itself a sufficient number of times $n$ reaches the identity element $p^n=e$ (otherwise the group generated by $p$ would be infinite).
@@
 
-These arcs can be established by using a set of permutations. As a historic
+As a historic
 example, the \emph{change ringing} tradition conceives music through
@@
 \centering
-\caption{Change Ringing: Traditional \emph{peal} for 3 bells. Permutations
-occur between each line. Each line is a bell ordering and each ordering is played at a time.} 
+\caption{A traditional change ringing \emph{peal} for 3 bells. Permutations
+occur between each line. Each line is a bell ordering to be played before next line.} 
 \begin{tabular}{l c r}
@@
 The use of permutations in music can be summarized in the following way:
-with $S_i=\{s_i\}$ a sequence of musical events $s_i$ (e.g.\ notes), and a
-permutation $p$. $S_i'=p(S_i)$ comprises the same elements of $S_i$ but in a
+with $S=\{s_i\}$ a sequence of musical events $s_i$ (e.g.\ notes), and a
+permutation $p$. $S'=p(S)$ comprises the same elements of $S$ but in a
 different order. Permutations have two notations: cyclic and
@@
 produces (ex. $[1\;3\;7\;0\;...]$). In the cyclic notation, a permutation consists
-of swaps by elements and its successors, and the last element by the first one.
+of swaps between elements and its successors, and between the last element and the first.
 E.g. $(1,2,5)(3,4)$ in cyclic notation is equivalent to $[0,2,5,4,3,1]$ in natural notation.
 
-In the auralization of a permutation, it is not necessary to permute elements of $S_i$,
-but only some characteristic. Thus, if $p$ is a permutation and $S_i$ is a sequence of basic notes as in the end of Section~\ref{notaBasica}, the
-sequence $S_i'=p^f(S_i)=\left\{s_i^{p(f)}\right\}$ consists of the same
+In the auralization of a permutation, it is not necessary to permute elements of $S$,
+but only some characteristic. Thus, if $p$ is a permutation and $S$ is a sequence of basic notes as in the end of Section~\ref{notaBasica}, the
+sequence $S'=p^f(S)=\left\{s_i^{p(f)}\right\}$ consists of the same
 music notes, following the same order and maintaining the same characteristics, but with the
@@
 Two subtleties of this procedure should be commented upon. 
-First, a permutation $p$ is not restricted to involve all elements of $S_i$, i.e.\ it can operate in a subset of $S_i$.
-Second, not all elements $s_i$ need to be executed at each access to $S_i$. To exemplify, let $S_i$ be a sequence of music notes $s_i$. 
+First, a permutation $p$ is not restricted to involve all elements of $S$, i.e.\ it can operate in a subset of $S$.
+Second, not all elements $s_i$ need to be executed at each access to $S$. To exemplify, let $S$ be a sequence of music notes $s_i$. 
 If $i$ goes from $0$ to $n$, and
@@
 notes.
-The other notes of $S_i$ can occur in other measures where permutations 
-allocate those notes to the first four events $s'_i$ of $S'_i$.
+The other notes of $S$ can occur in other measures where permutations 
+allocate those notes to the first four events $s'_i$ of $S'$.
 
 To each permutation $p_i$ described above, we have to determine:
-1) note dimensions where it operates (frequency, duration, \emph{fades},
-intensity, etc); and
+1) note characteristics where it operates (frequency, duration, \emph{fades},
+intensity, timbre, etc); and
 2) the period of incidence (how much consults before a permutation is
-applied). During realization of notes in $S_i$, an easy and coherent form is to
-execute the first $n$ notes, the execution of disjoint sets of $S_i$
+applied).
+For a coherent realization of the notes in $S$ is to
+execute the first $n$ notes given that the execution of disjoint sets of $S$
 is the same as modifying the permutation and executing the first $n$ notes.
-
-The \massa\ toolbox presents a computational implementation that isolates
-permutation application to sonic characteristics, in order to deliver
-musical structures~\cite{MASSA,figgusOriginal,figgusEspacializacao}.
-
-\subsection{Serialism and post-serial techniques}
+The PPEPPS is a set of musical pieces that uses permutations.~\cite{MASSA,figgusOriginal,figgusEspacializacao}.
+
+\subsection{Serialism and post-serial techniques}\label{sec:ser}
 Expanding the concepts in Section~\ref{sec:atonal},

### 2017-08-12 article/acm/body.tex
finished full review of the article

@@
 \subsection{Serialism and post-serial techniques}\label{sec:ser}
-Expanding the concepts in Section~\ref{sec:atonal},
+Expanding the concepts exposed in Section~\ref{sec:atonal},
 sequences of characteristics can be predefined and
@@
 density of events, etc.
-The sequences can be used very strictly or loose,
-such as by skipping some elements.
+The sequences can be used very strictly or loosely,
+such as by arbitrarily skipping some elements.
 The sequences can be of different sizes, yielding arcs
@@
 		such as
-		to derive harmonies, sequences of pitches or a temporal evolution of the overall spectrum~\cite{grisey}.
-		In other words, the most prominent frequencies can be used as pitches,
+		to derive harmonies, sequences of pitches or a temporal evolution of the overall spectrum.
+		Examples: the most prominent frequencies of a sound can be used as pitches,
 		real notes can be used to mimic an original spectrum
-		(e.g. use piano notes to mimic the spectrum of a spoken sentence) or portions of the spectrum made to vary~\cite{spectral}.
+		(e.g. use piano notes to mimic the spectrum of a spoken sentence) or portions of the spectrum made to vary.~\cite{grisey}
 	\item Spectromorphology:
-		can be considered a spectral music (spectralism) theoretical framework~\cite{smalley,schaeffer}
+		can be considered a spectral music (spectralism) theoretical framework
 		that examines the relation between sound spectra and their temporal evolution.
 		The theory poses e.g. different onsets, continuations and terminations; characteristics of (sonic) ``motion'';
-		and spectral density.
+		and spectral density.~\cite{smalley,schaeffer}
 	\item Stochastic music:
-		the use of random variables to describe musical elements are largely considered in stochastic music~\cite{formalized}.
+		the use of random variables to describe musical elements are largely considered in stochastic music.
 		In summary, one can use probability distributions for the synthesis of sounds and larger scale musical elements.
-		Changes in these distributions or in other characteristics yield the discourse.
+		Changes in these distributions or in other characteristics yield the discourse.~\cite{formalized}
 	\item Textures:
@@
 		density of notes, durations of notes, motives, number of voices, etc.
-		If the sounds are small enough (typically < $100ms$) the process can be though of in terms of \emph{granular synthesis}~\cite{microsound}.
+		If the sounds are small enough (typically $< 100ms$)
+		the process can be though of in terms of \emph{granular synthesis}~\cite{microsound}.
 \end{itemize}
@@
 repetition and variation, relaxation and tension, stability and instability, consonance and dissonance, etc.
-A thorough discussion of what can be considered a musical language is out of the scope of this article, but this brief consideration of the subject is useful as a onvergence of all the previous content.
+A thorough discussion of what can be considered a musical language is out of the scope of this article, but this brief consideration of the subject is useful as a convergence of all the previous content.
 
 \subsection{Musical usages}\label{subsec:usosmusicais3}
-First, the basic note was defined and characterized in quantitative terms in Section~\ref{sec:notaDisc}.
-Next, the internal note
-composition was addressed within both internal transitions and elementary sonic treatment
-(Section~\ref{sec:varInternas}). Finally, this section aims at organizing these notes in music. The numerous resources and consequent infinitude
+We strived to supply the reader with a very useful collection of
+fundamental concepts to elaborate musical structures
+with the content of Sections~\ref{sec:discNote} and~\ref{sec:internalVar}.
+The numerous resources and consequent infinitude
 of praxis possibilities is typical and highly relevant for artistic contexts~\cite{Harmonia,Webern}.
-
-There are studies and further developments for each of the presented resources.
 For example, it is possible to obtain `dirty' triadic harmonies (with notes out of the triad) by superposition of perfect fourths.
 Another interesting example is the superimposition of rhythms in different metrics, constituting what is
-called \emph{polyrhythm}. The music piece \emph{Poli-hit mia} in \massa\ toolbox~\cite{MASSA} explores these simultaneous metrics by impulse trains convolved with notes.
-
+called \emph{polyrhythm}. The music piece \emph{Poli-hit mia} explores these simultaneous metrics by impulse trains convolved with notes.~\cite{MASSA} 
 Microtonal scales are important for 20th
-century music~\cite{microtonalidade} and yielded diverse remarkable results throughout history, e.g.
+century music~\cite{microtonalidade} and yielded diverse remarkable results throughout history.
+E.g.
 fourths of a tone ($\epsilon=2^{\frac{1}{24}}$) is often used in some genres of Indian and Arabic music.
-The musical sequence \emph{MicroTom} in \massa\ toolbox~\cite{MASSA} explores these resources, including microtonal melodies and harmonies
-with many notes in a very reduced note scope.
-
-As in Section~\ref{subsec:mus2}, relations between
+The musical sequence \emph{MicroTom} explores microtonal melodies and harmonies.~\cite{MASSA}
+As stated in Section~\ref{subsec:mus2}, relations between
 parameters are powerful ways to achieve musical pieces.
@@
 duration. Harmonies can be made from triads (Equations~\ref{triades}) with duplicated
-notes at each octave and more numerous duplication when the depth and frequency of
-vibratos are lower (Equations~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}).
+notes at each octave and more numerous duplications entail lower depth and frequency of
+vibratos (Equations~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}).
 Incontestably, the possibilities are very wide, which is evidenced in the wide range of musical pieces and styles.
@@
 Equations~\ref{eq:groups}) can be used together. In the music piece \emph{3 trios}
-this association is done in a systematic way in order to achieve a specific style. This is an instrumental piece, not included as a source
+this association is done in a systematic way in order to achieve a specific style (or idiom).
+This is an instrumental piece, not included as a source
 code but available online~\cite{3Trios}.
@@
 \emph{PPEPPS} (Pure Python EP: Project Solvent) is an EP (Extended Play) synthesized using
-resources presented in this study. With minimal parametrization, the scripts
+resources presented in this article. With minimal parametrization, the scripts
 generate complete musical pieces, allowing easy composition of sets of
-music. A simple script of a few lines specifies music delivered as 16 bit
+music. A simple script of a few lines specifies music delivered as 16 bit,
 44.1kHz PCM files (WAVE). This facility and technological
-arrangement creates aesthetic possibilities for both sharing and education.
+arrangement creates aesthetic possibilities with potential to both
+sharing and education.
 
@@
 We aimed at achieving a concise presentation of the subject because it involves many knowledge fields, and therefore can very easily blast into thousands of pages.
-Some readers might benefit from the text alone, but the \emph{Scripts} in the \massa\ toolbox, where all the equations and concepts are directly and simply implemented as software (in Python),
-are very helpful for one to achieve elaborated implementations and deeper understandings.
-The scripts include routines that render musical pieces that illustrate the concepts in practical contexts.
+Some readers might benefit from the text alone, but the \emph{scripts} in the \massa\ toolbox, where all the equations and concepts are directly and simply implemented as software (in Python),
+are very helpful for one to achieve elaborated implementations and deeper insights.
+The scripts include routines that render musical pieces that illustrate the concepts in practical contexts
+even when we did not use equations in the descriptions (mainly in Section~\ref{sec:notesMusic}).
 This is valuable since art (music) can involve many non-trivial processes and is often deeply glamorized, which results in a nearly unmanageable terrain for a newcomer.
@@
 are all replaced, in discrete signals, by summations,
-which are more intuitive and does not require fluency in calculus.
+which are generally more intuitive and do not require fluency in calculus.
 	\item The equations and concepts are implemented as software which can be easily assembled and inspected.
@@
 with the respective codes and sonic examples, facilitate collaborations and the generation of sub-products in a co-authorship fashion,
-easing new implementations and development of musical pieces.
+favoring new implementations and musical pieces.
 The scripts can be divided in three groups: implementation of all the equations and topics of music theory; routines for rendering musical pieces that illustrate the concepts; scripts that render the figures of this article and the article itself.
-
-This framework favored the formation of interest groups in topics such as musical creativity and computer music.
-In particular, the project \url{labMacambira.sourceforge.net} groups Brazilian and foreign co-workers
-in diverse areas that range from digital direct democracy and georeferencing to art and education.
-This was only possible because of the usefulness of audiovisual abilities in many contexts, in particular because of the knowledge and mastery condensed in the \massa\ framework.\footnote{There are more than 700 videos, written documents,
- original software applications and contributions in well-known external software (such as Firefox, Scilab, LibreOffice, GEM/Puredata, to name just a few)~\cite{siteLM,wikiLM,vimeoLM}.
-Some of these efforts are available online~\cite{dissertacao}.
-It is evident that all these contributions are a consequence of more that just \massa, but it is also evident to the authors that \massa\ had a primary role in converging interests and attracting collaborators.}
+There are also a few auxiliary scripts which did not fit into these categories.~\cite{SI}
+
+% This framework favored the formation of interest groups in topics such as musical creativity and computer music.
+% In particular, the project \url{labMacambira.sourceforge.net} groups Brazilian and foreign co-workers
+% in diverse areas that range from digital direct democracy and georeferencing to art and education.
+% This was only possible because of the usefulness of audiovisual abilities in many contexts, in particular because of the knowledge and mastery condensed in the \massa\ framework.\footnote{There are more than 700 videos, written documents,
+%  original software applications and contributions in well-known external software (such as Firefox, Scilab, LibreOffice, GEM/Puredata, to name just a few)~\cite{siteLM,wikiLM,vimeoLM}.
+% Some of these efforts are available online~\cite{dissertacao}.
+% It is evident that all these contributions are a consequence of more that just \massa, but it is also evident to the authors that \massa\ had a primary role in converging interests and attracting collaborators.}
 
@@
 for the generation of appealing artistic materials. Some psychoacoustic effects were detected,
-which needs validation and should be reported, specially with~\cite{quadrosSonoros}.\footnote{The sonic portraits where sent to a public mailing list and the fifth piece was reported by some individuals to induce a state in which noises from the own tongue, teeth and jaw of the individual echoed for some seconds~\cite{quadrosMetarec}.} Other foreseen advances
+which needs validation and should be reported, especially with~\cite{quadrosSonoros}.\footnote{The sonic portraits where sent to a public mailing list and the fifth piece was reported by some individuals to induce a state in which noises from the own tongue, teeth and jaw of the individual echoed for some seconds~\cite{quadrosMetarec}.} Other foreseen advances
 are: a JavaScript version of the toolbox, better hypermedia deliverables of this framework, user guides

### 2017-08-12 article/acm/sample-acmsmall.tex
finished full review of the article

@@
 \begin{abstract}
-The representation of basic elements of music - such as notes,
-ornaments and intervals - in terms of discrete audio
+The representation of basic elements of music - such as notes - in terms of discrete audio
 signals is often used in software for music creation and design.
@@
 In this article, each musical element is related by equations and algorithms to the discrete-time samples of sounds,
-and each of these relations are implemented in scripts within a software toolbox,
-referred to as \massa\ (Music and Audio in Sample Sequences).
+and each of these relations are implemented in scripts within the
+    \massa\ software toolbox
+ (Music and Audio in Sample Sequences).
 The toolbox also includes this article and every script necessary to render the images
-and musical examples, pieces and albums, which adds to the educational and artistic uses.
+and musical examples, pieces and albums.
 The fundamental element, the musical note with duration, volume, pitch and timbre,
@@
 Internal variations of a note, such as tremolos, vibratos and spectral fluctuations,
-are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities.
-With this representation of notes, resources are provided for the generation of higher scale musical structures,
+are also described, which enables the synthesis of notes inspired by real instruments and new sonorities.
+With this representation of notes, resources are provided for the
+    generation of higher level musical structures,
 such as rhythmic meter, pitch intervals and cycles.
@@
 expanded in co-authorship processes and used with freedom by musicians, engineers and other interested parties.
-In fact, \massa\ has already been employed by external users for diverse purposes which include music production,
-artistic presentations, psychoacoustic experiments and computer language diffusion where the appeal of audiovisual artifacts is exploited for education.
 \end{abstract}

### 2017-08-16 article/acm/body.tex
incorporated suggestions from Chu

@@
-% \tableofcontents % enable to generate .toc file used in listings.tex
+% mi\tableofcontents % enable to generate .toc file used in listings.tex
 \section{Introduction}\label{sec:level1}
@@
 Since the speed of sound is $\approx 343.2 m/s$,
-such frequency limits corresponds to  wavelengths of $\frac{343.2}{20} \approx 17.16\,m$ and
+such frequency limits correspond to  wavelengths of $\frac{343.2}{20} \approx 17.16\,m$ and
 $\frac{343.2}{20000} \approx 17.16\,mm$.
@@
 e.g. as captured sounds are processed to generate new sonorities.
-Digital audio protocols often implies quality loss (to yield smaller files, ease storage and transfer)
+Digital audio protocols often imply quality loss (to yield smaller files, ease storage and transfer)
 and are called \emph{lossy}~\cite{loss}.
@@
 Non-lossy representations of digital audio, called \emph{lossless} protocols or formats,
-on the other hand, assures perfect reconstruction of the analog wave within any convenient precision.
+on the other hand, assure perfect reconstruction of the analog wave within any convenient precision.
 The standard paradigm of lossless audio consists of
@@
 sound amalgams and abstract discursive arcs~\footnote{There
-are well known incidence of such characteristics in ethnic music,
+are well known incidences of such characteristics in ethnic music,
 such as in Pygmy music, but western theory assimilated them
@@
 		which performs numerical routines efficiently (e.g. through LAPACK).
-		Part of the scripts have been transcribed to JavaScript (which favors their use in Web browsers such as Firefox and Chromium) and native Python~\cite{numpy, audiolab, tutpython, python}.
+		Part of the scripts has been transcribed to JavaScript (which favors their use in Web browsers such as Firefox and Chromium) and native Python~\cite{numpy, audiolab, tutpython, python}.
 		These are all open technologies, published using licenses that allow for copy,
@@
 	\item The sampling rate (discussed above) is also called the sampling frequency or sample rate.
-	\item An harmonic in a sound is most often a sinusoidal component which is in the harmonic series of the fundamental frequency.
+	\item A harmonic in a sound is most often a sinusoidal component which is in the harmonic series of the fundamental frequency.
 Many times, however, the terms harmonic and component are not distinguished.
@@
 The pitch is specified by the fundamental frequency $f$ whose cycle has duration $\delta=1/f$.
-This duration, multiplied by the sampling frequency $f_s$, yields in the number of samples per cycle: $\lambda=f_s . \delta =f_s/f$.
+This duration, multiplied by the sampling frequency $f_s$, yields the number of samples per cycle: $\lambda=f_s . \delta =f_s/f$.
 For didactic reasons, let $f$ divide $f_s$ and result $\lambda$ integer. If $T^f$ is a sonic sequence with fundamental frequency $f$, then:
@@
 In a similar fashion, other waveforms are used in music for their spectral qualities and simplicity.
-While the sinusoid is an isolated node in the spectrum, any other waveform present a succession of harmonic components.
+While the sinusoid is an isolated node in the spectrum, any other waveform presents a succession of harmonic components.
 The most basic waveforms are specified in Equations~\ref{sinusoid},~\ref{sawTooth},~\ref{triangular} and~\ref{square}, and are illustrated in Figure~\ref{fig:formasDeOnda}.
@@
 In general, these waveforms are appreciated as excessively rich in sharp harmonics, and attenuation by filtering on treble and middle parts of the spectrum is especially useful for reaching a more natural and pleasant sound.
-The relatively attenuated harmonics of the triangle wave makes it the more functional - among the listed ones - to be used in the synthesis of musical notes without any further processing.
+The relatively attenuated harmonics of the triangle wave make it the more functional - among the listed ones - to be used in the synthesis of musical notes without any further processing.
 The sinusoid is often a nice choice, but a problematic one.
@@
 Equation~\ref{moduloEfase} shows how the imaginary term of $c_k$ adds a phase to the real sinusoid: the terms $b_k$ enable the phase sweep $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right]$ given by $arctan\left(\frac{b_k}{a_k}\right)$ which has this image.
-The sign of $a_k$ specify the right or left side of the trigonometric circle, which completes the phase domain: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.
+The sign of $a_k$ specifies the right or left side of the trigonometric circle, which completes the phase domain: $\left[-\frac{\pi}{2},+\frac{\pi}{2}\right] \cup \left[\frac{\pi}{2},\frac{3\pi}{2}\right]\equiv [2\pi]$.
 
@@
 Figure~\ref{fig:amostras2} shows two samples and their spectral component.
-When there is only two samples, the Fourier decomposition has only one pair of coefficients $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relative to frequencies $\{f_k\}_0^1=\left\{w_k\frac{f_s}{2\pi}\right\}_0^1=\left\{k\frac{f_s}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_s}{2}=f_{\text{max}}\right\}$
+When there are only two samples, the Fourier decomposition has only one pair of coefficients $\{c_k=a_k-j.b_k\}_0^{\Lambda-1=1}$ relative to frequencies $\{f_k\}_0^1=\left\{w_k\frac{f_s}{2\pi}\right\}_0^1=\left\{k\frac{f_s}{\Lambda=2}\right\}_0^1=\left\{0,\frac{f_s}{2}=f_{\text{max}}\right\}$
 with energies $e_k=\frac{(c_k)^2}{\Lambda=2}$.
@@
 The coefficients $c_k$ are equivalent two by two, corresponding to the same frequencies and with the same contribution to its reconstruction.
-They are complex conjugates: $a_{k1}=a_{k2}$ and $b_{k1}=-b_{k2}$ and, as a consequence, the modules are equal and phases have opposite signs.
+They are complex conjugates: $a_{k1}=a_{k2}$ and $b_{k1}=-b_{k2}$ and, as a consequence, the moduli are equal and phases have opposite signs.
 Recalling that $f_k = k\frac{f_s}{\Lambda}, \; k \in \left\{0, ..., \left \lfloor \frac{\Lambda}{2} \right \rfloor \right\} $.
@@
 
-To discuss the general case for components combination in each sample $t_i$, one can gather relations in Equation~\ref{moduloEfase} for the real signal reconstruction, equivalences of modules (Equation~\ref{equivalenciasModulos}) and phases(Equation~\ref{equivalenciasFases}), the number of paired coefficients (Equation~\ref{coefsPareados}), and the equivalence of paired frequencies (Equation~\ref{equivalenciasFreqs}):
+To discuss the general case for components combination in each sample $t_i$, one can gather relations in Equation~\ref{moduloEfase} for the real signal reconstruction, equivalences of moduli (Equation~\ref{equivalenciasModulos}) and phases (Equation~\ref{equivalenciasFases}), the number of paired coefficients (Equation~\ref{coefsPareados}), and the equivalence of paired frequencies (Equation~\ref{equivalenciasFreqs}):
 
@@
 2) techniques to produce sound through various sources, such as loudspeakers, singers and traditional musical instruments, for musical purposes.
-We focus in the first item although issues of the item are also tackled.}.
+We focus in the first item although issues of the second item are also tackled.}.
 A note is always spatialized
@@
 \subsubsection{Spatial localization}
-It is understood that the perception of sound localization occurs in our nervous system mainly by three cues: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, specially its chest, head and ears~\cite{Roederer, hrtf, Heeger}.
+It is understood that the perception of sound localization occurs in our nervous system mainly by three cues: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, wspecially the chest, head and ears~\cite{Roederer, hrtf, Heeger}.
 
@@
 
-There are reverberation models less based in each independent reflection and that explores valuable cues to the auditory system.
+There are reverberation models less based on each independent reflection and that explore valuable cues to the auditory system.
 In fact, a reverberation can be modeled with a set of two temporal and two spectral regions~\cite{JOSPhy}:
@@
 from sampled data or hand picked values.
-In music its usage transcends these applications: it simplifies many operations and enables the use a single wave period to synthesize sounds in the whole audible spectrum, with any waveform.
+In music its usage transcends these applications: it simplifies many operations and enables the use of a single wave period to synthesize sounds in the whole audible spectrum, with any waveform.
 
@@
 This subsection is limited to a description of processing by convolution and difference equations, and immediate applications, as a thorough discussion of filtering is beyond the scope of this study\footnote{The
-design and implementation of filters is field of recognized complexity,
+design and implementation of filters is a field of recognized complexity,
 with dedicated literature and software~\cite{Openheim,smith}.}.
@@
 There are many methods and tools to obtain IIR filters. The text below lists a selection for didactic and reference purposes.
-They are well behaved filters whose frequency response are illustrated in Figure~\ref{fig:iir}.
+They are well behaved filters whose frequency responses are illustrated in Figure~\ref{fig:iir}.
 For the low-pass and high-pass filters, the cutoff frequency $f_c$ is where the filter performs an attenuation of $-3dB \approx 0.707 $ of the original amplitude.
@@
         \includegraphics[width=\textwidth]{figures/iir__}
-    \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of Equations~\ref{eq:passa-baixas},~\ref{eq:passa-altas},~\ref{eq:passa-banda} and~\ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and bandwidths.}
+    \caption{Moduli for the frequency response (a), (b), (c) and (d) for IIR filters of Equations~\ref{eq:passa-baixas},~\ref{eq:passa-altas},~\ref{eq:passa-banda} and~\ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and bandwidths.}
         \label{fig:iir}
@@
 \begin{enumerate}
-  \item Low-pass with a simple pole and module of the frequency response in the upper left corner of Figure~\ref{fig:iir}.
+  \item Low-pass with a simple pole and modulus of the frequency response in the upper left corner of Figure~\ref{fig:iir}.
 The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, which is the fraction of the sample frequency $f_s$ in which an attenuation of $3dB$ occurs.
@@
 \end{equation}
-  \item High-pass filter with a simple pole and module of the frequency responses at the upper right corner of Figure~\ref{fig:iir}.
+  \item High-pass filter with a simple pole and modulus of the frequency responses at the upper right corner of Figure~\ref{fig:iir}.
 	  The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ has coefficients:
@@
 emitted by a violin or a piano.
-Furthermore, many percussion instruments does not exhibit an unequivocal pitch and their sounds are generally regarded as noise~\cite{Roederer}.
+Furthermore, many percussion instruments do not exhibit an unequivocal pitch and their sounds are generally regarded as noise~\cite{Roederer}.
 In electronic music, including electro-acoustic and dance genres, noise has diverse uses and frequently characterizes the music style~\cite{Cook}.
@@
 
-The exponential $e^{j.x}$ is a way to obtain unitary module and random phase for the value of $c_i$. In addition, $c_{\Lambda/2}$ is always real (as discussed in Section~\ref{sec:spectrum}).
+The exponential $e^{j.x}$ is a way to obtain unitary modulus and random phase for the value of $c_i$. In addition, $c_{\Lambda/2}$ is always real (as discussed in Section~\ref{sec:spectrum}).
 
@@
 
-It is important correctly distinguish both tables 
+It is important to correctly distinguish both tables 
 ($\widetilde{M}=\{\widetilde{m}\}_0^{\widetilde{\Lambda}_M-1}$ and $\widetilde{L}=\{\widetilde{l}\}_0^{\widetilde{\Lambda}-1}$)
@@
 For example, two impulses equally spaced build a binary division in the
-pulse. Two signals, one with 2 impulses and the other with 3 impulses, both equally spaced in the pulse duration, yields a pulse
+pulse. Two signals, one with 2 impulses and the other with 3 impulses, both equally spaced in the pulse duration, yield a pulse
 maintenance with a rhythm which eases both binary or ternary
@@
 proportions among the amplitudes of the sonic reincidences. The use of convolution with impulses in this context is explored in the music piece \emph{Little train of impulsive hillbillies}.
-The piece also encompass
+The piece also encompasses
 the creation of `sound amalgams' e.g. by granular synthesis (see Figure~\ref{fig:pulsoSubAgl}).
@@
 \subsubsection{Tuning}
-Doubling the frequency is equivalent to ascending one octave\footnote{An pitch
+Doubling the frequency is equivalent to ascending one octave\footnote{A pitch
 is both considered equivalent and different from any of its octaves.
@@
 		As can be noticed in Table~\ref{eq:intervalos}, one step is not one semitone.
-		A step, in this sense, is yield by two consecutive notes in a musical scale.
+		A step, in this sense, is yielded by two consecutive notes in a musical scale.
 		A scale for now can be regarded as any arbitrary monotonic sequence of pitches and will be discussed in the next section.
@@
 		it characterizes the ``dominant'' chord in the tonal system (see Section~\ref{subsec:harmonia}) and represents (or yields) instability.
-Some cultures considers the interval a consonance, using it in melodies and chants as a stable interval.
+Some cultures consider the interval a consonance, using it in melodies and chants as a stable interval.
 	\item A major interval decreased by one semitone results in a minor interval. A minor interval increased by one semitone results in a major interval.
@@
 
-The minor scale have two additional forms, named harmonic and melodic:
+The minor scale has two additional forms, named harmonic and melodic:
 
@@
         \item A fifth confirms the root (fundamental).
-		There are theoretical discussion about why this happens, and the most usual arguments are that the fifth is the first (non-octave) harmonic of a note and that the harmonics of the fifth are in the harmonics of the fundamental.
+		There are theoretical discussions about why this happens, and the most usual arguments are that the fifth is the first (non-octave) harmonic of a note and that the harmonics of the fifth are in the harmonics of the fundamental.
 		Important here is to grasp the fact that musical theory and practice assures that the fifth enhances the sensation that
 		the root of a chord is the root.
-        \item Major or minor thirds from the root entails major or minor chord qualities.
+        \item Major or minor thirds from the root entail major or minor chord qualities.
         \item Every tritone, especially if built between a major third and a minor seventh, tends to resolve into a third or a sixth.
@@
 Most frequently, atonal music is not strictly dodecafonic, but "serial", i.e.
-they use the same kind of techniques as stated above, based in a sequence of pitches (called the series or row), but not within a 12 tone sequence and using sequences of other parameters (e.g.
+they use the same kind of techniques as stated above, based on a sequence of pitches (called the series or row), but not within a 12 tone sequence and using sequences of other parameters (e.g.
 duration, timbre).
@@
 The `functional harmony' ascribes functions to the three main chords and tries to understand their use by means of these functions.
-The chord built on top of the first degree is the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its function (role) consists on maintaining a center, usually referred to as a ``ground'' for the music.
+The chord built on top of the first degree is the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its function (role) consists in maintaining a center, usually referred to as a ``ground'' for the music.
 The chord built on the fifth degree is the \textbf{dominant} (\textit{D}, the dominant is always major) and its function is to lean for the tonic
@@
 listener and the sonic characteristics~\cite{microsound,Roederer}.
-The rhythm is commonly based on a basic duration called pulse, whose typical duration range between $0.25$ and $1.5s$ ($240$
+The rhythm is commonly based on a basic duration called pulse, whose typical duration ranges between $0.25$ and $1.5s$ ($240$
 and $40 BPM$, respectively~\footnote{BPM stands for Beats Per Minute and is just a frequency measure like Herz, but is the number of incidences per minute instead of second.
@@
 of $H$ musical events $s_j$, each event with its $\Lambda_j$ samples $t_i^j$
-(refer to the beginning of this Section~\ref{notasMusica} if needed). Bellow is a list of basic techniques
+(refer to the beginning of this Section~\ref{notasMusica} if needed). Below is a list of basic techniques
 for variation.
@@
 intensity, timbre, etc); and
-2) the period of incidence (how much consults before a permutation is
+2) the period of incidence (how many consults before a permutation is
 applied).
@@
 is the same as modifying the permutation and executing the first $n$ notes.
-The PPEPPS is a set of musical pieces that uses permutations.~\cite{MASSA,figgusOriginal,figgusEspacializacao}.
+PPEPPS is a set of musical pieces that uses permutations.~\cite{MASSA,figgusOriginal,figgusEspacializacao}.
 
@@
 	\item Spectralism:
-		consists on use the (Fourier) spectrum of a sound or the harmonic series for music composition,
+		consists in using the (Fourier) spectrum of a sound or the harmonic series for music composition,
 		such as
@@
 The numerous resources and consequent infinitude
-of praxis possibilities is typical and highly relevant for artistic contexts~\cite{Harmonia,Webern}.
+of praxis possibilities are typical and highly relevant for artistic contexts~\cite{Harmonia,Webern}.
 For example, it is possible to obtain `dirty' triadic harmonies (with notes out of the triad) by superposition of perfect fourths.
@@
 century music~\cite{microtonalidade} and yielded diverse remarkable results throughout history.
-E.g.
-fourths of a tone ($\epsilon=2^{\frac{1}{24}}$) is often used in some genres of Indian and Arabic music.
+For instance, 
+fourths of a tone ($\epsilon=2^{\frac{1}{24}}$) are often used in some genres of Indian and Arabic music.
 The musical sequence \emph{MicroTom} explores microtonal melodies and harmonies.~\cite{MASSA}
@@
 The number of permuted
-notes can vary during the music, revealing relationship with piece
+notes can vary during the music piece, revealing relationship with piece
 duration. Harmonies can be made from triads (Equations~\ref{triades}) with duplicated
@@
 vibratos (Equations~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}).
-Incontestably, the possibilities are very wide, which is evidenced in the wide range of musical pieces and styles.
+For sure, the possibilities are very wide, which is evidenced in the wide range of musical pieces and styles.
 
@@
 \label{cap:conclusao}
-In our understanding,
-this article is effective in relating musical elements to digital audio.
-We aimed at achieving a concise presentation of the subject because it involves many knowledge fields, and therefore can very easily blast into thousands of pages.
+In this article, we hope to have effectively related musical elements to digital audio.
+We aimed at achieving a concise presentation of the subject that involves many knowledge fields.
 Some readers might benefit from the text alone, but the \emph{scripts} in the \massa\ toolbox, where all the equations and concepts are directly and simply implemented as software (in Python),
 are very helpful for one to achieve elaborated implementations and deeper insights.
-The scripts include routines that render musical pieces that illustrate the concepts in practical contexts
+The scripts include routines that render musical pieces to illustrate the concepts in practical contexts
 even when we did not use equations in the descriptions (mainly in Section~\ref{sec:notesMusic}).
 This is valuable since art (music) can involve many non-trivial processes and is often deeply glamorized, which results in a nearly unmanageable terrain for a newcomer.
-Moreover, this didactic report and the supplied open source scripts should facilitate the use of the proposed framework.
+Moreover, this article and the open source scripts supplied should facilitate the use of the proposed framework.
 The Supporting Information document holds listings of sections, equations, figures, tables, scripts and other documents.
 
-The possibilities provided by this exposition pour from both the organization of knowledge and the ability to achieve sounds which are extremely true to the models.
+The possibilities provided stem from both the organization of knowledge and the ability to achieve sounds which are true to the models.
 For example, one can produce noises with arbitrary resolution of the spectrum and a musical note can be synthesized with the parameters (e.g. of a vibrato) updated sample-by-sample.
-Furthermore, software for synthesis and processing of sounds for musical purposes
-by standard restrict the bit depth to 16 or 24.
+Furthermore, standard software for synthesis and processing of sounds for musical purposes
+by restrict the bit depth to 16 or 24.
 This is achievable in this framework but by standard Python uses more bits per floating point number.
@@
 for many scientific and artistic purposes.
-The didactic potential of the framework is evident when noticed that:
+The didactic potential of the framework is evident by noticing that:
 \begin{itemize}
@@
 \end{itemize}
-\noindent In fact, this framework was used in a number of contexts, including courses, software implementations and making music~\cite{vimeoLM,vivace,repoDissertacao,dissertacao}.
-Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, as far as the authors know, such as testified in the literature review (Appendix G of~\cite{dissertacao}), where books, articles and open software are related to this framework.
+\noindent Significantly, this framework was used in a number of contexts, including courses, software implementations and making music~\cite{vimeoLM,vivace,repoDissertacao,dissertacao}.
+Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, as far as the authors know (see the literature review in the Appendix G of~\cite{dissertacao}), where books, articles and open software are related to this framework.
 
 The free software license and online availability of the content,
-with the respective codes and sonic examples, facilitate collaborations and the generation of sub-products in a co-authorship fashion,
+with the respective codes and sonic examples, facilitate collaborations and generation of sub-products in a co-authorship fashion,
 favoring new implementations and musical pieces.
@@
 for the generation of appealing artistic materials. Some psychoacoustic effects were detected,
-which needs validation and should be reported, especially with~\cite{quadrosSonoros}.\footnote{The sonic portraits where sent to a public mailing list and the fifth piece was reported by some individuals to induce a state in which noises from the own tongue, teeth and jaw of the individual echoed for some seconds~\cite{quadrosMetarec}.} Other foreseen advances
+which need validation and should be reported, especially with~\cite{quadrosSonoros}.\footnote{The sonic portraits were sent to a public mailing list and the fifth piece was reported by some individuals to induce a state in which noises from the own tongue, teeth and jaw of the individual echoed for some seconds~\cite{quadrosMetarec}.} Other foreseen advances
 are: a JavaScript version of the toolbox, better hypermedia deliverables of this framework, user guides

### 2017-09-22 article/acm/body.tex
temporal brench to move changes to other lap

--- 2017-08-12 article/acm/body.tex
@@
 Articles on the topic appear to be lacking, to the best of our knowledge.
-Although current music software use the analytical descriptions presented here,
+Although current music software uses the analytical descriptions presented here,
 there is no concise mathematical description of them, and it is far from trivial
@@
 
-\subsection{Synonymy, polysemy and theoretical frames (disclaimer)}\label{sec:discl}
+\subsection{Synonymy, polysemy and theoretical frames (disclaimer)}
 Given that the main topic of this article (the expression of musical elements
@@
 Canonically, the basic characteristics of a musical note are duration, loudness, pitch and timbre~\cite{Lacerda}.
-All relations described in this section are implemented in the file \texttt{src/sections/1.py}.
+All relations described in this section are implemented in the file \texttt{src/sections/eqs2.1.py}.
 The musical pieces \emph{5 sonic portraits} and \emph{reduced-fi} are also available online to corroborate the concepts.
@@
 
-\subsection{Spectrum of sampled sound}\label{sec:spectrum}
+\subsection{Spectrum of sampled sound}
 These sinusoidal components in the discretized sound have some particularities.
@@
 \subsection{Spatialization: localization and reverberation}\label{subsec:spac}
-A sound is always produced within the three dimensional physical space.
-The consideration of this fact is the subject of spatialization\footnote{By
+A musical note is always produced within the three dimensional physical space.
+The consideration of this fact is the subject of the spatialization knowledge field and practice\footnote{By
 spatialization one might find both:
@@
 
-\noindent where $\zeta$ is the distance between the ears, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in the Interaural Time Difference:
+\noindent where $\zeta$ is the distance between the ears, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in Interaural Time Difference:
 
@@
 
-\noindent where $t_i$ is a right channel sample and $t_i'$ is a left channel sample.
+\noindent where $t_i$ is the right channel and $t_i'$ is the left channel.
 If $\Lambda_{ITD} < 0$, it is necessary to change $t_i$ by $t_i'$
@@
 
-Spatial localization depends considerably on other cues. By using only ITD and IID it is possible to specify solely the horizontal (azimuthal) angle $\theta$ given by:
+Spatial localization depends considerably on other cues. By using only ITD and IID it is possible to specify solely the horizontal angle (azimuthal) $\theta$ given by:
 
@@
 The sound propagates through the air with a speed of $\approx 343.2m/s$ and can be emitted from a source with any directionality pattern.
-When a sound wave encounters a surface there are: 1) inversion of the propagation speed component normal to the surface;  2) energy absorption, especially in high frequencies.
+When a sound pulse encounters a surface there are: 1) inversion of the propagation speed component normal to the surface;  2) energy absorption, especially in high frequencies.
 The sonic waves propagate until they reach inaudible levels (and further but then can often be neglected).
-As a sonic front reaches the human ear, it can be described as an ordinary sound, with the last reflection point as the source, and the absorption filters of each surface it has reached.
+As a sonic front reaches the human ear, it can be described as the original sound, with the last reflection point as the source, and the absorption filters of each surface it has reached.
 It is possible to simulate reverberations that are impossible in real systems.
-For example, one can use asymmetric reflections with relation to the axis perpendicular to the surface, or increase the intensity of the last reflections.
-
-There are reverberation models less based in each independent reflection and that explores valuable cues to the auditory system.
-In fact, a reverberation can be modeled with a set of two temporal and two spectral regions~\cite{JOSPhy}:
+For example, it is possible to use asymmetric reflections with relation to the axis perpendicular to the surface, or to increase the intensity of the last reflections.
+
+There are reverberation models less related to each independent reflection and that explores valuable cues to the auditory system.
+In fact, reverberation can be modeled with a set of two temporal and two spectral regions~\cite{JOSPhy}:
 \begin{itemize}
@@
 Smith III states that usual concert rooms have total reverberation time of $\approx 1.9$ seconds, and that the period of first reflections is around $0.1s$.
-Under these conditions, there are perceived wave pulses which propagate for $652.08m$ ($83.79k$ samples in $f_s=44.1kHz$) before reaching the ear.
-In addition, reflections that reach the ear after propagating  for $34.32m$ ($4.41k$ samples in $f_s=44.1kHz$) or more have incidences less distinct by hearing.
-The first reflections are particularly important to spatial sensation.
+With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08m$ ($83.79k$ samples in $f_s=44.1kHz$) before reaching the ear.
+In addition, sound reflections made after propagation for $34.32$ m ($4.41k$ samples in $f_s=44.1kHz$) have incidences less distinct by hearing.
+These first reflections are particularly important to spatial sensation.
 The first incidence is the direct sound, described by ITD and IID of Equations~\ref{eq:dti} and~\ref{eq:dii}.
-Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between each reflection is $8-90ms$ ($\approx 350-4000$ samples in $f_s=44.1kHz$).
+Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_s=44.1kHz$).
 Also, it is experimentally verifiable that the number of reflections increases with the square of time.
@@
 \subsection{Musical usages}\label{subsec:basMus}
-The sum of the respective $ith$ samples of $N$ sequences with the same size $\Lambda$
-results in the overlapped spectral contents of each sequence, in a process called mixing:
+Once the basic note is defined, it is didactically convenient to build musical structures with sequences based on these particles.
+The sum of elements with same index of $N$ sequences $T_{k,i}=\{t_{k,i}\}_{k=0}^{N-1}$ with same size $\Lambda$ results in the overlapped spectral contents of each sequence, in a process called mixing:
 
@@
 
-Figure~\ref{fig:mixagem} illustrates this overlapping of discretized sound waves, each with 100 samples.
-The duration is very short $\frac{f_s=44.1kHz}{100} \approx 2ms$.
+Figure~\ref{fig:mixagem} illustrates this overlapping process of discretized sound waves, each with 100 samples.
 If $f_s=44.1kHz$, the frequencies of the sawtooth, square and sine wave are, respectively: $\frac{f_s}{100/2}=882Hz$, $\frac{f_s}{100/4}=1764Hz$ and $\frac{f_s}{100/5}=2205Hz$.
-One can complete the sequence with zeroes to sum (mix) sequences of different sizes.
+The samples duration is very short $\frac{f_s=44.1kHz}{100} \approx 2ms$.
+One can complete the sequence with zeroes to sum (mix) sequences with different sizes.
+
 The mixed notes are generally separated by the ear according to the physical laws of resonance and by the nervous system~\cite{Roederer}.
-This process of mixing musical notes results in musical harmony and counterpoint which guide
-subjective aspects of music appreciation~\cite{Harmonia} and are addressed in Section~\ref{sec:notesMusic}.
+This process of mixing musical notes results in musical harmony whose intervals between frequencies and chords of simultaneous notes guide subjective and abstract aspects of music and its appreciation~\cite{Harmonia}, which is addressed in Section~\ref{sec:notesMusic}.
+
+Sequences can be concatenated in time.
+If the sequences $\{t_{k,i}\}_0^{\Lambda_k-1}$ of size $\Lambda_k$ represent $k$ musical notes, their concatenation in a unique sequence $T_i$ is a simple melodic sequence, or melody of its own:
+
+\begin{equation}\label{eq:concatenacao}
+\begin{split}
+\{t_i\}_0^{\sum\Delta_k-1}= & \{t_{l,i}\}_0^{\sum\Delta_k-1}, \;\; \\ l\text{ smallest integer } : & \quad \Lambda_l > i -\sum_{j=0}^{l-1}\Lambda_j
+\end{split}
+\end{equation}
+
+This mechanism is demonstrated in Figure~\ref{fig:concatenacao} with the same sequences of Figure~\ref{fig:mixagem}.
+Although the sequences are short for the usual sample rates, it is easy to visually observe the concatenation of sound sequences.
+In addition, each note has  a duration larger than $100ms$ if $f_s<1kHz$.
 
@@
 
-Sequences can be concatenated in time.
-If the sequences $\{t_{k,i}\}_0^{\Lambda_k-1}$ of size $\Lambda_k$ represent $k$ musical notes, their concatenation in a unique sequence $T$ results in a simple melodic sequence:
-
-\begin{equation}\label{eq:concatenacao}
-\begin{split}
-\{t_i\}_0^{\sum\Delta_k-1}= \{t_{l,i}\}_0^{\sum\Delta_k-1}, \;\; l\text{ smallest integer}: \;\; \sum_{j=0}^{l}\Lambda_j > i
-\end{split}
-\end{equation}
-
-This mechanism is demonstrated in Figure~\ref{fig:concatenacao} with the same sequences of Figure~\ref{fig:mixagem}.
-Although the sequences are short for the usual sample rates, it is easy to visually observe the concatenation of sound sequences.
-Moreover, each note has  a duration larger than $100ms$ if $f_s<1kHz$.
-
 The musical piece \emph{reduced-fi} explores the temporal juxtaposition of notes, resulting in a homophonic piece.
-The vertical principle is demonstrated in the \emph{sonic portraits} which are static sounds with peculiar spectrum.\cite{MASSA}
+The vertical principle is demonstrated at the \emph{sonic portraits}, static sounds with peculiar spectrum.
+Both pieces were written in Python and are available as part of the \massa\ \emph{toolbox}.\cite{MASSA}
+
+With the basic musical digital note carefully described, the next section develops the temporal evolution of its contents as in \emph{glissandi} and intensity envelopes.
+Filtering of spectral components and noise generation complements the musical note as a self-contained unity.
+Section~\ref{notasMusica} is dedicated to the organization of these notes as music by using metrics and trajectories, with regards to traditional music theory.
 
@@
 \section{Variation in the basic note}\label{sec:internalVar}\label{sec:varInternas}
-The basic digital music note defined in Section~\ref{sec:notaDisc} has the following characteristics: duration, pitch, loudness and timbre.
-We also stated that a note is always spatialized and considered both localization and reverberation.
-This is a useful and paradigmatic model, but it does not exhaust the aspects of a musical note.
-First of all, a real note is not static~\cite{Chowning}.
-For example, a $3s$ piano note has an abrupt rise of intensity at the beginning and a progressive decline,
-has harmonic components decaying and emerging.
-These variations are not mandatory, but they are used in sound synthesis for music because 
-they reflect how sounds appear in nature and because they yield interesting sounds~\cite{Roederer}.
-To explore all the ways in which variations occur within a note is out of the scope of any work given
-the complexity and sensibility of the human cognition of sounds.
-Resources essential to produce variations in the basic note are presented in this section.
-The musical pieces \emph{Transit para meter}, \emph{Vibrates and shakes}, \emph{Tremolos, vibratos and the frequency}, \emph{Little train of impulsive hillbillies}, \emph{Noisy band}, \emph{Bela rugosi}, \emph{Children's chorus}, \emph{ADa and SaRa} were made to validate and illustrate concepts of this section.~\cite{MASSA}
+The basic digital music note defined in Section~\ref{sec:notaDisc} has the following parameters: duration, pitch, intensity (loudness) and timbre.
+This is a useful and paradigmatic model, but it does not exhaust all the aspects of a musical note.
+First of all, characteristics of the note change along the note itself~\cite{Chowning}.
+For example, a $3s$ piano note has intensity with abrupt rise at the beginning and progressive decay, has spectrum variations with harmonics decaying and some others emerging along time.
+These variations are not mandatory, but they are used in sound synthesis for music because they reflect how sounds appear in nature.
+This is considered so important that there is a rule of thumb: to make a sound that incites interest by itself, arrange internal variations on it~\cite{Roederer}.
+To explore all the ways in which variations occur within a note is out of the scope of any work, given the sensibility of the human ear and the complexity of human sound cognition.
+In the following, primary resources are presented to produce variations in the basic note.
+It is worthwhile to recall that all the relations in this and other sections are implemented in Python and published in the public domain \massa\ toolbox.
+The musical pieces \emph{Transit para meter}, \emph{Vibrates and shakes}, \emph{Tremolos, vibratos and the frequency}, \emph{Little train of impulsive hillbillies}, \emph{Noisy band}, \emph{Bela rugosi}, \emph{Children's chorus}, \emph{ADa and SaRa} were made to validate and illustrate concepts of this section.
+The code that synthesizes these pieces is also part of the toolbox\cite{MASSA}.
 
 \subsection{Lookup table}\label{subsec:lookup}
-The \emph{Lookup Table} (LUT) is an array for indexed operations which substitutes continuous and repetitive calculation.
+The \emph{Lookup Table} (or simply LUT) is an array for indexed operations which substitutes continuous and repetitive calculation.
 It is used to reduce computational complexity and for employing functions without calculating them directly, e.g.
@@
     \caption{Search in the \emph{lookup table} to synthesize sounds at different frequencies using a unique waveform with high resolution.
-	Each \emph{i-th} sample $t_i^f$ of a sound with frequency $f$ is related to the samples in the table $\widetilde{l}$\, by the relation $t_i^{f}=\widetilde{l}_{\left\lfloor i.f\frac{\widetilde{\Lambda}}{f_s} \right\rfloor \%\,\widetilde{\Lambda}}$ where $\widetilde{\Lambda}$ is the number of sample in $\widetilde{L}$ and $f_s$ is the sampling rate.}
+	Each \emph{i-th} sample $t_i^f$ of a sound with frequency $f$ is related to the samples in the table $\widetilde{l}$\, by the relation $t_i^{f}=\widetilde{l}_{\left\lfloor i.f\frac{\widetilde{\Lambda}}{f_s} \right\rfloor \%\,\widetilde{\Lambda}}$ where $\widetilde{\Lambda}$ is the number of sample in $\widetilde{l}$ and $f_s$ is the sampling rate.}
         \label{fig:lut}
@@
 
-Let $\widetilde{\Lambda}$ be the wave period in samples and $\widetilde{L} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ the sequence of samples $\widetilde{l_i}$.
-A sequence $T^{f,\,\Delta}$ with samples of a sound with frequency $f$ and duration $\Delta$ can be obtained by means of $\widetilde{L}$:
+Let $\widetilde{\Lambda}$ be the wave period in samples and $\widetilde{L_i} = \left\{\, \widetilde{l}_i \,\right\}_0^{\widetilde{\Lambda} -1}$ the samples $\widetilde{l_i}$ (refer to Equation~\ref{periodoUnico}), a sequence $T_i^{f,\,\Delta}$ with samples of a sound with frequency $f$ and duration $\Delta$ can be obtained by means of $\widetilde{L_i}$:
 
@@
 \begin{split}
-T^{f,\,\Delta}=\left\{t_i^f\right\}_0^{\lfloor \, f_s .
+T_i^{f,\,\Delta}=\left\{t_i^f\right\}_0^{\lfloor \, f_s .
 \Delta \, \rfloor -1} = \left\{ \, \widetilde{l}_{\gamma_i \% \widetilde{\Lambda} }\, \right\}_{0}^{\Lambda-1}\; , \quad \text{where} \;\; \gamma_i = \left \lfloor i f \frac{ \widetilde{\Lambda}}{f_s} \right \rfloor
@@
 
-In other words, with the right LUT indexes ($\gamma_i\%\widetilde{\Lambda}$) it is possible to synthesize sounds at an arbitrary frequency.
+In other words, with the right LUT indexes ($\gamma_i\%\widetilde{\Lambda}$) it is possible to synthesize sounds at any frequency.
 Figure~\ref{fig:lut} illustrates the calculation of $\{t_i\}$ sample from $\left\{\,\widetilde{l}_i\,\right\}$ for $f=200Hz$, $\widetilde{\Lambda}=128$ and adopting the sample rate of $f_s=44.1kHz$.
@@
 
-The calculation of $\gamma_i$ can be understood as follows:
-$\frac{i}{f_s}$
+The expression defining the variable $\gamma_i$ can be understood as $f_s$ being added to $i$ at each second.
+If $i$ is divided by the sample frequency, $\frac{i}{f_s}$
 is incremented by $1$ at each second.
-Multiplied by the period, $i \frac{\widetilde{\Lambda}}{f_s}$ covers $\widetilde{\Lambda}$ in one second.
-Finally, with frequency $f$ it results in $i f \frac{\widetilde{\Lambda}}{f_s}$ which completes $f$ cycles on $\widetilde{L}$ in $1$ second, i.e.
-the resulting sequence has the fundamental frequency $f$.
+Multiplied by the period, it results in $i \frac{\widetilde{\Lambda}}{f_s}$, which covers the period in one second.
+Finally, with frequency $f$ it results in $i .
+f \frac{\widetilde{\Lambda}}{f_s}$ which completes $f$ periods $\widetilde{\Lambda}$ in $1$ second, i.e.
+the resulting sequence presents the fundamental frequency $f$.
 
@@
 Limits exist only at low frequencies when the size of table $\widetilde{\Lambda}$ is not sufficient for the sample rate $f_s$.
-The lookup procedure replaces calculations by simple indexed searches (what is generally understood as an optimization process).
+The lookup procedure is virtually costless and replaces calculations by simple indexed searches (what is generally understood as an optimization process).
 Unless otherwise stated, this procedure will be used along all the text for every applicable case.
 LUTs are broadly used in computational implementations for music.
-A classical usage of LUTs is known as \emph{Wavetable Synthesis}, which consists of many LUTs used together to generate a quasi-periodic musical note~\cite{Cook,Wavetable}.
+A classical usage of LUTs is known as \emph{Wavetable Synthesis}, which consists of many LUTs used together to generate a quasi-periodic music note~\cite{Cook,Wavetable}.
 
@@
 
-Consider a note with duration $\Delta = \frac{\Lambda}{f_s}$, in which the frequency $f_i$ varies linearly from $f_0$ to $f_{\Lambda -1}$:
+Consider a note with duration $\Delta = \frac{\Lambda}{f_s}$, in which the frequency $f=f_i$ varies linearly from $f_0$ to $f_{\Lambda -1}$. Thus:
 
 \begin{equation}\label{freqLinear}
- F=\{f_i\}_0^{\Lambda-1}=\left\{f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right\}_0^{\Lambda-1}
+ F_i=\{f_i\}_0^{\Lambda-1}=\left\{f_0 + (f_{\Lambda-1}-f_0)\frac{i}{\Lambda-1} \right\}_0^{\Lambda-1}
 \end{equation}
@@
 
-\noindent where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_s}$ is the LUT index increment between two samples given the frequency of the first sample.
-Therefore, it is handy to calculate the amplitudes $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ by means of the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$.
+\noindent where $\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_s}$ is the LUT increment between two samples given the sound frequency of the first sample.
+Therefore, it is handy to calculate the elements $t_i^{\;\overline{f_0,f_{\Lambda-1}}}$ by means of the period $\left\{\widetilde{l}_i\right\}_0^{\Lambda-1}$.
 Equations~\ref{freqLinear},~\ref{indiceLinear} and~\ref{serieAmostralLin} are related with the linear progression of the frequency.
@@
 
-The term $\frac{i}{\Lambda-1}$ covers the interval $[0,1]$ and it is possible to raise it to a power in such a way that the beginning of the transition will be smoother or steeper.
-This procedure is useful for energy variations with the purpose of changing the loudness\footnote{The
-loudness (a psychophysic quality) is a consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is the wave total energy.
-Wave energy can be varied in many ways.
-The simplest way consists of modifying the amplitude by multiplying the whole sequence by a real number (see Section~\ref{subsec:volume}).
-The increase of energy without changing amplitude ambit is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}.}.
-It is sufficient to multiply the original sequence element-wise by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is an arbitrary coefficient and $t_{\Lambda-1}.a_{\Lambda-1}$ is the value to be reached at the end of the transition.
-
-Thus, for amplitude variations:
-\begin{equation}\label{seqAmp}
-\begin{split}
- \{a_i\}_0^{\Lambda-1}= \left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}= \left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
-\end{split}
-\end{equation}
-where $a_0=1$. And, therefore:
-
-\begin{equation}\label{transAmp}
-\begin{split}
- T^{'}=T \odot A = & \{t_i a_i\}_0^{\Lambda-1} \\ = & \left \{ t_i (a_{\Lambda-1} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
-\end{split}
-\end{equation}
-
 \begin{figure}[h!]
@@
 
+The term $\frac{i}{\Lambda-1}$ covers the interval $[0,1]$ and it is possible to raise it to a power in such a way that the beginning of the transition will be smoother or steeper.
+This procedure is useful for energy variations with the purpose of changing the loudness\footnote{The
+loudness (psychophysical quality) is a consequence of different sound characteristics, like reverberation and concentration of high harmonics, among which is the wave total energy.
+Wave energy is easy to modify (see Section~\ref{subsec:volume}) and can be varied in many ways.
+The simplest way consists of modifying the amplitude by multiplying the whole sequence by a real number.
+The increase of energy without changing amplitude ambit is the \emph{sound compression}, quite popular nowadays for music production~\cite{guillaume}.}.
+It is sufficient to multiply the original sequence by the sequence $a_{\Lambda-1}^{\left( \frac{i}{\Lambda-1} \right )^\alpha}$, where $\alpha$ is the given coefficient and $a_{\Lambda-1}$ is a fraction of the original amplitude, which is the value to be reached at the end of the transition.
+
+Thus, for amplitude variations:
+\begin{equation}\label{seqAmp}
+\begin{split}
+ \{a_i\}_0^{\Lambda-1}= \left \{ a_0 \left ( \frac{a_{\Lambda-1}}{a_0} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}= \left \{ \left ( {a_{\Lambda-1}} \right )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
+\end{split}
+\end{equation}
+where $a_0=1$. And, therefore:
+
+\begin{equation}\label{transAmp}
+\begin{split}
+ T_i^{'}=T_i \odot A_i = & \{t_i . a_i\}_0^{\Lambda-1} \\ = & \left \{ t_i . (a_{\Lambda-1} )^{\left ( \frac{i}{\Lambda-1} \right )^\alpha} \right \}_0^{\Lambda-1}
+\end{split}
+\end{equation}
+
 It is often convenient to have $a_0=1$ to start a new sequence with the original amplitude and then progressively change it.
 If $\alpha=1$, the amplitude variation follows the geometric progression that implies the linear variation of loudness.
-Figure~\ref{fig:transicao} depicts transitions for different values of $\alpha$, a gain of $\approx 6dB$ as exposed by Equation~\ref{eq:ampVol}.
+Figure~\ref{fig:transicao} depicts transitions between values 1 and 2 and for different values of $\alpha$, a gain of $\approx 6dB$ as given by Equation~\ref{eq:ampVol}.
 
@@
 Both cases make the procedure useless, once a ratio of any number in relation to zero is not well defined for our purposes.
-It is possible to solve this dilemma by choosing a number that is small enough such as $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum amplitude for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$).
+It is possible to solve this dilemma choosing a number that is small enough like $-80dB\;\Rightarrow a=10^{\frac{-80}{20}}=10^{-4}$ as the minimum loudness for a \emph{fade in} ($a_0=10^{-4}$) or for a \emph{fade out} ($a_{\Lambda-1}=10^{-4}$).
 A linear fade can be used then to reach zero amplitude, if needed.
@@
 
-Here, the conversion between decibels and amplitude is convenient. Given Equations~\ref{ampDec} and~\ref{transAmp}, a transition of $V_{dB}$ decibels is:
+Here the conversion between decibels and amplitude is convenient, with Equations~\ref{ampDec} and~\ref{transAmp} specifying a transition of $V_{dB}$ decibels:
 
 \begin{equation}\label{seqAmpDB}
-T^{'}=\left\{ t_i 10^{\frac{V_{dB}}{20}\left( \frac{i}{\Lambda-1} \right)^\alpha} \right\}_0^{\Lambda-1}
+T_i^{'}=\left\{ t_i 10^{\frac{V_{dB}}{20}\left( \frac{i}{\Lambda-1} \right)^\alpha} \right\}_0^{\Lambda-1}
 \end{equation}
@@
 The greater the value of $\alpha$, the smoother the sound introduction and more intense its end.
-$\alpha>1$ results in loudness transitions commonly called \emph{slow fade}, while $\alpha<1$ results in \emph{fast fade}~\cite{guillaume}.
+$\alpha>1$ results in loudness transitions commonly called \emph{slow fade}, while $\alpha<1$ results in \emph{fast fade}\cite{guillaume}.
 
 The linear transitions will be used for AM and FM synthesis, while logarithmic transitions are proper tremolos and vibratos, as developed in Section~\ref{subsec:tvaf}.
-A non-oscillatory exploration of these variations is in the music piece \emph{Transit para meter}~\cite{MASSA}.
+A non-oscillatory exploration of these variations is in the music piece \emph{Transit para meter}, whose code is online as part of the \massa\ toolbox\cite{MASSA}.
 
 \subsection{Application of digital filters}\label{subsec:filtros}
-This subsection is limited to a description of processing by convolution and difference equations, and immediate applications, as a thorough discussion of filtering is beyond the scope of this study\footnote{The
-design and implementation of filters is field of recognized complexity,
-with dedicated literature and software~\cite{Openheim,smith}.}.
-Filtering can be part of the sound synthesis or made subsequently as part of processes commonly referred to as ``acoustic/sound treatment''.
+This subsection is limited to a description of sequences processing, by convolution and difference equations, and immediate applications, as a thorough discussion of filtering is beyond the scope of this study\footnote{The implementation of filters encompasses an area of recognized complexity, with dedicated literature and software implementations~\cite{Openheim,smith}.}.
+Filter applications can be part of the synthesis process or made subsequently as part of processes commonly referred to as ``acoustic/sound treatment''.
 
@@
 This sample representation is called `impulse response' $\{h_i\}$.
-FIR filters are applied in the time domain of digital sound by means of convolution with the respective impulse response of the filter\footnote{It
-is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both the sound and the impulse response, and then performing the inverse Fourier transform of the resulting spectrum.
-This is usually computationally cheaper than performing the convolution in the time domain
-but might imply greater complexity especially if the sequence of samples is not fully available (e.g.
-in real-time applications)~\cite{Openheim}.}.
-For the purposes of this work, convolution $T*H$ of $T$ with $H$ is defined as:
+FIR filters are applied in the time domain of digital sound by means of convolution with the respective impulse response of the filter\footnote{It is possible to apply the filter in the frequency domain multiplying the Fourier coefficients of both the sound and the impulse response, and then performing the inverse Fourier transform in the resulting spectrum.\cite{Openheim}}.
+For the purposes of this work, convolution is defined as:
 
@@
  \begin{split}
- \left\{t_i'\right\}_0^{\Lambda_t+\Lambda_h-2\; = \;\Lambda_{t\, '}-1} = & \{(T*H)_i\}_0^{\Lambda_{t \, '}-1} \\ = & \left \{ \sum_{j=0}^{min(\Lambda_h-1,i)}h_{j} t_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
+ \left\{t_i'\right\}_0^{\Lambda_t+\Lambda_h-2\; = \;\Lambda_{t\, '}-1} = & \{(T_j*H_j)_i\}_0^{\Lambda_{t \, '}-1} \\ = & \left \{ \sum_{j=0}^{min(\Lambda_h-1,i)}h_{j} t_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
      \\ = & \left \{ \sum_{j=max(i+1-\Lambda_h,0)}^{i}t_j h_{i-j} \right \}_0^{\Lambda_{t\, '}-1}
@@
 In other words, the sound $\{t_i'\}$, resulting from the convolution of $\{t_i\}$, with the impulse response $\{h_i\}$, has each $i$-th sample $t_i$ overwritten by the sum of its last $\Lambda_h$ samples $\{t_{i-j}\}_{j=0}^{\Lambda_h-1}$ multiplied one-by-one by samples of the impulse response $\{h_i\}_0^{\Lambda_h-1}$.
-This procedure is illustrated in Figure~\ref{fig:conv}, where the impulse response $\{h_i\}$ is in its retrograde form, and $t_{12}'$ and $t_{32}'$ are two samples from the convolution.
-The final signal always has the length $\Lambda_{t'}=\Lambda_t+\Lambda_h -1$.
-With this procedure it is possible to achieve reverberators, equalizers, \emph{delays}, to name a few of a variety of filters for sound processing used to obtain musical/artistic effects.
+This procedure is illustrated in Figure~\ref{fig:conv}, where the impulse response $\{h_i\}$ is in its retrograde form, and $t_{12}'$ and $t_{32}'$ are two calculated samples using the convolution given by $(T_j*H_j)_i=t_i'$.
+The final signal always has the length of $\Lambda_t+\Lambda_h -1=\Lambda_{t'}$.
+
+With this procedure it is possible to achieve reverberators, equalizers, \emph{delays}, to name a few of a variety of other filters for sound processing used to obtain musical/artistic effects.
 
 The impulse response can be provided by physical measures or by pure synthesis.
-An impulse response for a reverberation application, for example, can be obtained by recording the sound in a room when someone triggers a click which resembles an impulse, or be obtained by a sinusoidal sweep whose Fourier transform approximates its frequency response.
-Both are impulse responses which, when convoluted with a sequence of sound samples, result in the own sound with a reverberation that resembles the original environment where the measure was made~\cite{Cook}.
-The inverse Fourier transform of a sequence of coefficients might be used as an impulse response of a FIR filter.
-Convoluted with a sound, it performs the frequency filtering specified by coefficients\footnote{Given
-that the sound is described by a sequence of real numbers,
-the coefficients should follow the rule $c_k=c^*_{\Lambda-k}$ as discussed in Section~\ref{sec:spectrum}.}.
-The greater the number of samples, the higher the spectral resolution and the computational complexity, which should often be weighted, for convolution is expensive.
+An impulse response for a reverberation application, for example, can be obtained by sound recording of the environment when someone triggers a click which resembles an impulse, or obtained by a sinusoidal sweep whose Fourier transform approximates its frequency response.
+Both are impulse responses which, properly convoluted with the sound sequence, result in the own sound with a reverberation that resembles the original environment where the measure was made~\cite{Cook}.
+The inverse Fourier transform of an even and real envelope is an impulse response of a FIR filter.
+Convoluted with a sound, it performs the frequency filtering specified by the envelope.
+The greater the number of samples, the higher the envelope resolution and the computational complexity, which should often be weighted, for convolution is expensive.
 
 An important property is the time shift caused by convolution with a shifted impulse.
-Despite being computationally expensive, it is possible to create \emph{delay lines} by means of convolution with an impulse response that has an impulse for each reincidence of the sound.
+Despite being computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse for each reincidence of the sound.
 Figure~\ref{fig:delays} shows the shift caused by convolution with an impulse.
-Depending on the density of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as a sound amalgam (more than $\approx 20$ impulses per second).
+Depending on the density of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as pitch (from about 20 impulses per second and higher frequencies).
 In the latter case, the process resembles granular synthesis, reverberation and equalization.
@@
 
-There are many methods and tools to obtain IIR filters. The text below lists a selection for didactic and reference purposes.
-They are well behaved filters whose frequency response are illustrated in Figure~\ref{fig:iir}.
-For the low-pass and high-pass filters, the cutoff frequency $f_c$ is where the filter performs an attenuation of $-3dB \approx 0.707 $ of the original amplitude.
+There are many methods and tools to obtain IIR filters.
+The text below lists a selection for didactic purposes and as a reference.
+They are well behaved filters whose aspects are described in Figure~\ref{fig:iir}.
+For filters of simple order, the cutoff frequency $f_c$ is where the filter performs an attenuation of $-3dB \approx 0.707 $ of the original amplitude.
 For band-pass and band-reject (or 'notch') filters, this attenuation has two specifications: $f_c$ (in this case, the `center frequency') and bandwidth $bw$.
@@
 It is possible to apply these filters successively in order to obtain filters with other frequency responses.
-Other possibilities include the use of a biquad 'filter recipe'\footnote{Short 
-for 'biquadratic': its transfer function has two poles and two zeros, i.e.
+Another possibility is to use a biquad 'filter recipe'\footnote{Short for 'biquadratic': its transfer function has two poles and two zeros, i.e.
 its first direct form consists of two quadratic polynomials in the fraction: $\mathbb{H}(z)=\frac{a_0+a_1.z^{-1}+a_2.x^{-2}}{1- b_1.z^{-1} -b_2 .
-z^{-2}}$.}
-or a Chebichev filter\footnote{Butterworth and Elliptical filters can be considered as special cases of Chebichev filters~\cite{Openheim,smith}.}.
+z^{-2}}$.} or the calculation of Chebichev filter coefficients\footnote{Butterworth and Elliptical filters can be considered as special cases of Chebichev filters~\cite{Openheim,smith}.}.
 Both alternatives are explored by~\cite{JOSFM,smith}, and by the collection of filters maintained by the \emph{Music-DSP} community of the Columbia University~\cite{music-dsp,Openheim}.
@@
         \includegraphics[width=\textwidth]{figures/iir__}
-    \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of Equations~\ref{eq:passa-baixas},~\ref{eq:passa-altas},~\ref{eq:passa-banda} and~\ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and bandwidths.}
+    \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of Equations~\ref{eq:passa-baixas},~\ref{eq:passa-altas},~\ref{eq:passa-banda} and~\ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
         \label{fig:iir}
@@
 \begin{enumerate}
-  \item Low-pass with a simple pole and module of the frequency response in the upper left corner of Figure~\ref{fig:iir}.
-The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, which is the fraction of the sample frequency $f_s$ in which an attenuation of $3dB$ occurs.
-The coefficients $a_0$ and $b_1$ of the IIR filter are given by:
+  \item Low-pass with a simple pole, module of the frequency response in the upper left corner of Figure~\ref{fig:iir}.
+The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_s$ in which an attenuation of $3dB$ occurs.
+The coefficients $a_0$ and $b_1$ of the IIR filter are given by $x \in [e^{-\pi},1]$:
 \begin{equation}\label{eq:passa-baixas}
  \begin{split}
- a_0 & =  1-b_1 \\
- b_1 & = e^{-2\pi f_c}
- \end{split}
-\end{equation}
-  \item High-pass filter with a simple pole and module of the frequency responses at the upper right corner of Figure~\ref{fig:iir}.
-	  The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ has coefficients:
+ x & =e^{-2\pi f_c} \\
+ a_0 & =  1-x \\
+ b_1 & =  x
+ \end{split}
+\end{equation}
+  \item High-pass filter with a simple pole, module of its frequency responses at the upper right corner of Figure~\ref{fig:iir}.
+The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of $x \in [e^{-\pi},1]$:
 \begin{equation}\label{eq:passa-altas}
  \begin{split}
- a_0 & =  \frac{b_1+1}{2} \\
- a_1 & =  -\frac{b_1+1}{2} \\
- b_1 & = e^{-2\pi f_c}
+ x & =e^{-2\pi f_c} \\
+ a_0 & =  \frac{x+1}{2} \\
+ a_1 & =  -\frac{x+1}{2} \\
+ b_1 & =  x
  \end{split}
@@
 \item Notch filter.
-This filter is parametrized by a center frequency $f_c$ and bandwidth $bw$, both given as a fraction of $f_s$, therefore $f,\; bw \in (0,0.5)$.
+This filter is parametrized by a center frequency $f_c$ and bandwidth $bw$, both given as fraction of $f_s$, therefore $f,\; bw \in (0,0.5)$.
 Both frequencies $f_c \pm bw$ have $\approx 0.707$ of the amplitude, i.e.
-an attenuation of $3dB$.
+attenuation of $3dB$.
 The auxiliary variables $K$ and $R$ are defined as:
@@
 
-Some noises with static spectrum are listed below. They are called \emph{colored noises} since they are associated with colors for many reasons.
-Figure~\ref{fig:ruidos} shows the spectrum profile and the corresponding sonic samples sequence side-by-side.
+Some noises with static spectrum are listed below. They are called \emph{colored noise} since they are associated with colors for many reasons.
+Figure~\ref{fig:ruidos} shows the spectrum profile and the corresponding sonic sequence side-by-side.
 All five noises were generated with the same phase for each component, making it straightforward to observe the contributions of different parts of the spectrum.
@@
 
-The exponential $e^{j.x}$ is a way to obtain unitary module and random phase for the value of $c_i$. In addition, $c_{\Lambda/2}$ is always real (as discussed in Section~\ref{sec:spectrum}).
-
- \item The pink noise is characterized by a decrease of $3dB$ per octave.
-	 This noise is useful for testing electronic devices and are prominent in nature~\cite{Roederer}. 
+The exponential $e^{j.x}$ is a way to obtain unitary module and random phase for the value of $c_i$. In addition, $c_{\Lambda/2}$ is always real (as discussed in the previous section).
+
+ \item The pink noise is characterized by a decrease of $3dB$ per octave. This noise is useful for testing electronic devices, being prominent in nature~\cite{Roederer}. 
  %\begin{split}
@@
 
-  \item The brown noise received this name after Robert Brown, who described the Brownian movement\footnote{Although
-	  its origin is disparate with its color association, this noise became established with this specific name in musical contexts.
-Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more vivid colors~\cite{Cook,guillaume}.}.
-What characterizes a brown noise is the decrease of $6dB$ per octave, with $\alpha_i$ in Equations~\ref{eq:rosa} being:
+  \item The brown noise received this name after Robert Brown, who described the Brownian movement\footnote{Although its origin is disparate with its color association, this noise became established with this specific name in musical contexts.
+Anyway, this association can be considered satisfactory once violet, blue, white and pink noises are more strident and associated with more vivid colors~\cite{Cook,guillaume}.} What characterizes brown noise is the decrease of $6dB$ per octave, with $\alpha_i$ in Equations~\ref{eq:rosa} being:
 
@@
  \item In the blue noise there is a gain of $3dB$ per octave in a band limited by the minimum frequency $f_{\text{min}}$ and the maximum frequency $f_{\text{max}}$.
-	 Therefore, using Equations~\ref{eq:rosa} again:
+Therefore, the corresponding equation is also based on the Equations~\ref{eq:rosa}:
 
@@
 
- \item The black noise has higher losses than $6dB$ per octave:
+ \item The black noise has higher losses than $6dB$ for octave:
 
@@
 Those curves are obtained by experiments and are imperative to obtain $\alpha_i$.
-An implementation of ISO 226, which is the last revision of these curves, is in the file \texttt{src/aux/iso226.py}~\cite{MASSA}.
+An implementation of ISO 226, which is the last revision of these curves, is in the \massa\ toolbox~\cite{MASSA}.
 \end{itemize}
@@
 This subsection discussed only noise with static spectrum.
-There are also characterizations for noise with dynamic spectrum, and noises which are fundamentally transient, like clicks and chirps.
+There are also characterizations for noise with dynamic spectrum during the time, and noises which are fundamentally transient, like clicks and chirps.
 The former are easily modeled by an impulse relatively isolated, while chirps are not in fact a noise, but a fast scan of some given frequency band~\cite{Cook}.
@@
 \subsection{Tremolo and vibrato, AM and FM}\label{subsec:tvaf}
-Vibrato is a periodic variation in pitch (frequency) and tremolo is a periodic variation in loudness (intensity)\footnote{The
-jargon may be different in other contexts as stated in Section~\ref{sec:discl}~\cite{Lacerda,Harmonia}.}.
-A vibrato with frequency $f'$ can be achieved by:
+Vibrato is a periodic variation in pitch (frequency) and tremolo is a period variation in loudness (intensity)\footnote{The jargon may be different in other contexts.
+For example, in piano, tremolo is a vibrato in the classification used here.
+The definitions given here are common in contexts regarding music theory and electronic music.
+In addition, they are based on a broader literature than the one used for a specific instrument, practice or musical tradition~\cite{Lacerda,Harmonia}.}.
+A vibrato can be achieved by:
 
@@
 \begin{equation}\label{vbrT}
- T^{f, vbr(f',\,\nu)}=\left\{ t_i^{f,vbr(f',\,\nu)} \right\}_0^{\Lambda-1}=\left\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \right\}_0^{\Lambda-1}
+ T_i^{f, vbr(f',\,\nu)}=\left\{ t_i^{f,vbr(f',\,\nu)} \right\}_0^{\Lambda-1}=\left\{ \widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \right\}_0^{\Lambda-1}
 \end{equation}
@@
          \includegraphics[width=\columnwidth]{figures/vibrato_}
-     \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave, with $f_s=44.1kHz$. The color bar is in decibels.}
+     \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave, with $f_s=44.1kHz$. The color bar is uses decibels.}
          \label{fig:vibrato}
@@
 
-It is important correctly distinguish both tables 
-($\widetilde{M}=\{\widetilde{m}\}_0^{\widetilde{\Lambda}_M-1}$ and $\widetilde{L}=\{\widetilde{l}\}_0^{\widetilde{\Lambda}-1}$)
-and sequences ($T'=\{t_i'\}_0^{\Lambda-1}$ and $T=\{t_i\}_0^{\Lambda-1}$).
-Table $\widetilde{M}$ and the indices $\gamma_i'$ yield the sequence $\{t_i'\}$
-which is the oscillation pattern in the frequency while table $\widetilde{L}$ and the indices $\gamma_i$
-yield $\{t_i\}$ which is the sound itself.
-Variables are useful to quantify the vibrato amplitude (or depth):
+For the proper realization of the vibrato, it is important to pay attention to both tables and sequences.
+Table $\widetilde{M}_i$ with length $\widetilde{\Lambda}_M$ and the sequence with indices $\gamma_i'$ make the sequence $t_i'$ which is the oscillation pattern in the frequency while table $\widetilde{L}_i$ with length $\widetilde{\Lambda}$ and the sequence with indices $\gamma_i$ make $t_i$ which is the sound itself.
+Variables $\mu$ and $\nu$ quantify the vibrato intensity:
 \begin{itemize}
-	\item let $\mu$ be frequency difference (in $Hz$) between the main frequency and the maximum frequency, and
-    \item let $\nu$ be the the number of semitones (or half steps) added and subtracted by the oscillation ($2\nu$ is the number of semitones between the upper and lower peaks of frequency in the sound).
+    \item $\mu$ is a direct measure of how many Hertz are involved in the upper limit of the oscillation, while
+    \item $\nu$ is the direct measure of how many semitones (or half steps) are involved in the oscillation ($2\nu$ is the number of semitones between the upper and lower peaks of frequency oscillations of the sound $\{t_i\}$ in the vibrato).
 \end{itemize}
@@
 
-Figure~\ref{fig:vibrato} is the spectrogram of an artificial vibrato for a note with $1000Hz$ (between a \emph{B} and a \emph{C}),
-in which the pitch deviation reaches one octave above and one below.
+Figure~\ref{fig:vibrato} is the spectrogram of an artificial vibrato for a note with $1000Hz$ (between a \emph{B} and a \emph{C}), in which pitch deviation reaches one octave above and one below.
 Practically any waveform can be used to generate a sound and the vibrato oscillation pattern, with virtually any oscillation frequency and pitch deviation\footnote{The pitch deviation is called 'vibrato depth' and is generally given in semitones or cents (one cent $= \frac{1}{100}$ of a semitone).}.
-Oscillations with precise waveforms and arbitrary amplitudes are not possible in traditional music instruments, and thus it introduces novelty in the artistic possibilities.
-
-The tremolo is similar: $f'$, $\gamma_i'$ and $t_i'$ remains the same.
+Those oscillations with precise waveforms and arbitrary amplitudes are not possible in traditional music instruments, and thus it introduces novelty in the artistic possibilities.
+
+Tremolo is similar: $f'$, $\gamma_i'$ and $t_i'$ remains the same.
 The amplitude sequence to be multiplied by the original sequence $t_i$ is:
@@
 \begin{split}
- T^{tr(f')}=\left \{ t_i^{tr(f')} \right \}_0^{\Lambda-1}=\{ t_i .
+ T_i^{tr(f')}=\left \{ t_i^{tr(f')} \right \}_0^{\Lambda-1}=\{ t_i .
 a_i \}_0^{\Lambda-1}= \left \{t_i .10^{t_i' \frac{V_{dB}}{20}}    \right \}_0^{\Lambda-1}=\left\{t_i .
@@
 
-\noindent where $V_{dB}$ is the oscillation depth in decibels of the tremolo and $a_{\text{max}}=10^{\frac{V_{dB}}{20}}$ is the maximum amplitude gain.
-The measurement in decibels is suitable because the maximum increase in amplitude is not equivalent to the related maximum decrease, while the difference in decibels is invariant.
+\noindent where $V_{dB}$ is the oscillation depth in decibels of tremolo and $a_{\text{max}}=10^{\frac{V_{dB}}{20}}$ is the maximum amplitude gain.
+The measurement in decibels is suitable because the maximum increase in amplitude is not equivalent to the related maximum decrease, while the difference in decibels remains.
 
@@
 
-The music piece \emph{Vibrates and shakes} explores these possibilities given by tremolos and vibratos
-used simultaneously and independently, with different frequencies $f'$, depths ($\nu$ and $V_{dB}$),
-waveforms, and incremental modifications of parameters
-(tremolos and vibratos occur often together in traditional music instruments and voices).
-Aiming at a qualitative appreciation, the piece also develops a comparison between vibratos and tremolos in logarithmic and linear scales.~\cite{MASSA}
+The music piece \emph{Vibrates and shakes} explores these possibilities given by tremolos and vibratos, both used in conjunction and independently, with frequencies $f'$, different depths ($\nu$ and $V_{dB}$), and progressive parameters variations (tremolos and vibratos occur many times together in a traditional music instrument and voices).
+Aiming at a qualitative appreciation, the piece also develops a comparison between vibratos and tremolos in logarithmic and linear scales.
+Its source code is available online as part of the \massa\ toolbox.
 
@@
          \includegraphics[width=\textwidth]{figures/tremolo_}
-     \caption{Tremolo with a depth of $V_{dB}=12dB$, with a sawtooth waveform as its oscillatory pattern, with $f'=1.5Hz$ in a sine of $f=40Hz$ (if sample frequency is $f_s=44.1kHz$).}
+     \caption{Tremolo with a depth of $V_{dB}=12dB$, with a sawtooth waveform as its oscillatory pattern, with $f'=1.5Hz$ in a sine of $f=40Hz$ (sample frequency $f_s=44.1kHz$).}
          \label{fig:tremolo}
@@
 This roughness is largely appreciated both in traditional classical music and current electronic music, especially in the \emph{Dubstep} genre.
-Roughness is also generated by spectral content that produces beats\footnote{A beat is an interference pattern between two sounds which is perceived as a periodic variation of loudness.
-If the sounds are sinusoids, the variation is also sinusoidal with a frequency that is the difference between the frequencies of the sinusoids~\cite{Porres,porres2009}.}.
-The sequence \emph{Bela Rugosi} explores this roughness threshold with concomitant tremolos and vibratos at the same voice, with different intensities and waveforms.~\cite{MASSA}
+Roughness is also generated by spectral content that produces beating~\cite{Porres,porres2009}.
+The sequence \emph{Bela Rugosi} explores this roughness threshold with concomitant tremolos and vibratos at the same voice, with different intensities and waveforms.
+The corresponding code is available online in the \massa\ toolbox.
 
@@
 In this case, the oscillations are audible as pitch.
-Then, $f'$, $\mu$ and the waveform together change the audible spectrum of original sound $T$ in different ways for both tremolos and vibratos.
+Then, $f'$, $\mu$ and the waveform together change the audible spectrum of original sound $T_i$ in different ways for both tremolos and vibratos.
 They are called AM (\emph{Amplitude Modulation}) and FM (\emph{Frequency Modulation}) synthesis,
@@
 
-For musical goals, it is convenient to understand FM when only pure
-sines are used and, when other waveforms are employed,
-to decompose the signals into their respective Fourier components (i.e.\ sines as well).
-The FM synthesis performed with a sinusoidal vibrato of frequency $f'$ and depth $\mu$
-in a sinusoidal sound $T$ with frequency $f$ generates bands centered around $f$ and far from each other by $f'$:
+For musical goals, it is possible to understand FM based on the
+case of sines and, when other waveforms are employed, to decompose the signals into their respective Fourier components (i.e.\ sines as well).
+The FM synthesis performed with a sinusoidal vibrato of frequency $f'$ and depth $\mu$ in a sinusoidal sound $T_i$ with frequency $f$ generates bands centered around $f$ and far from each other by $f'$:
 
@@
 In these equations, the frequency variation introduced by $\{t_i'\}$ does not follow the geometric progression that yields linear pitch variation, but reflects Equation~\ref{freqLinear}.
-The result of using Equations~\ref{vbrF} for FM is described in the Appendix D of~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
-In fact, the simple and attractive FM behavior is usually observed within linear oscillations,
-such as in Equation~\ref{eq:fmEsp}, which yield less strident and less noisy sounds.
+The result of using Equations~\ref{vbrF} for FM is described in~\cite{dissertacao}, where the spectral content of the FM synthesis is calculated for oscillations in the logarithmic scale.
+In fact, the simple and attractive FM behavior is usually observed with linear oscillations, such as in Equation~\ref{eq:fmEsp}, which yield less strident noisy sounds.
 
@@
 Again, this is obtained by variations in the linear scale of the amplitude.
-A description of the spectrum of an AM performed with oscillations in the logarithmic amplitude scale is also available in the Appendix D of~\cite{dissertacao}.
-In FM and AM jargon,
-the sequence $T$ is called `carrier' and is modulated by
-$T'$, called 'modulator'; $\mu$ and
-$\alpha=10^{\frac{V_{dB}}{20}}$ are the `modulation indexes' and $f'$ is the modulation frequency.
-The following equations are useful for the achievement of the oscillatory pattern of the modulator sequence $\{t_i'\}$:
+A discussion of the
+spectrum of an AM performed with oscillations in the logarithmic amplitude scale is available in~\cite{dissertacao}. The sequence $T_i$, with frequency $f$, called `carrier', is modulated by
+$f'$, called 'modulator'. In FM and AM jargon, $\mu$ and
+$\alpha=10^{\frac{V_{dB}}{20}}$ are `modulation indexes'. The following equations are defined for the vibration pattern of the modulator sequence $\{t_i'\}$:
 
@@
 
-The modulator $\{t_i'\}$ is applied to the carrier $\{t_i\}$ in a FM by:
+The modulator $\{t_i'\}$ is applied to the carrier $\{t_i\}$:
 
@@
 \begin{equation}\label{fmGamma}
-\Delta_{\gamma_i}=\frac{\widetilde{\Lambda}}{f_s}f_i \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s} f_j \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}(f+\mu .
+\Delta_{\gamma_i}=f_i\frac{\widetilde{\Lambda}}{f_s} \quad \Rightarrow \quad \gamma_i = \left \lfloor \sum_{j=0}^{i} f_j \frac{\widetilde{\Lambda}}{f_s} \right \rfloor = \left \lfloor \sum_{j=0}^{i} \frac{\widetilde{\Lambda}}{f_s}(f+\mu .
 t_j') \right\rfloor
@@
 \begin{equation}\label{fmT}
-T^{f,\, FM(f',\,\mu)}=\left\{ t_i^{f,\,FM(f',\,\mu)} \right\}_0^{\Lambda-1}=\left\{\,\widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \,\right\}_0^{\Lambda-1}
-\end{equation}
-
-\noindent where $\widetilde{l}$ is the waveform period with a length of $\widetilde{\Lambda}$ samples, used for the carrier signal,
-	as exposed above.
-
-To perform AM, the signal $\{t_i\}$ needs to be modulated by $\{t_i'\}$ using the following equations:
+T_i^{f,\, FM(f',\,\mu)}=\left\{ t_i^{f,\,FM(f',\,\mu)} \right\}_0^{\Lambda-1}=\left\{\,\widetilde{l}_{\gamma_i \%\; \widetilde{\Lambda} } \,\right\}_0^{\Lambda-1}
+\end{equation}
+
+\noindent where $\widetilde{l}$ is the waveform period with a length of $\widetilde{\Lambda}$ samples, used for the carrier signal.
+
+To perform AM, the signal $\{t_i\}$ needs to be modulated with $\{t_i'\}$ using the following equations:
 
@@
 
-\begin{equation}\label{amT}
-T_i^{f,\,AM(f',\,\alpha)}=\left\{ t_i^{f,\,AM(f',\,\alpha)} \right\}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}= \{t_i . (1 + \alpha . t_i')    \}_0^{\Lambda-1}
-\end{equation}
+\begin{multline}\label{amT}
+T_i^{f,\,AM(f',\,\alpha)}=\left\{ t_i^{f,\,AM(f',\,\alpha)} \right\}_0^{\Lambda-1}=\{ t_i . a_i \}_0^{\Lambda-1}= \\ \{t_i . (1 + \alpha . t_i')    \}_0^{\Lambda-1}
+\end{multline}
 
@@
 At this point the musical possibilities are very wide.
+All characteristics, like pitch (given by frequency), timbre (achieved by waveforms, filters and noise), loudness (manipulated by intensity) and duration (given by the number of samples), can be considered in an absolute form or varied during the duration of the sound (clearly, with the exception of duration itself).
 The following musical usages comprehend a collection of possibilities with the purpose of exemplifying types of sound manipulations that result in musical material.
@@
 
-\noindent with both $f^{vbr}$ and $f^{tr}$ as $f'$ in the equations of Section~\ref{subsec:tvaf}.
+\noindent with $f^{vbr}$ and $f^{tr}$ as $f'$ in the referenced equations. They can also be associated with vibrato and tremolo oscillation frequency of Equation~\ref{vbrGamma}. 
  $\nu$ and $V_{dB}$ are the respective depth values of vibrato and
@@
 frequency} explores such bonds and exhibits variations in the oscillation waveform with the purpose of building a \emph{musical
-language} (details in the next section).~\cite{MASSA}
+language} (details in the next section). The corresponding code is also available online as part of the \massa\ toolbox.
 
 \subsubsection{Convolution for rhythm and meter}
-With the purpose of establishing metric and rhythms,
-the convolution with an impulse shifts the sound to the position of the impulse, as stated in Section~\ref{subsec:conv}.
 A musical pulse - such as specified by a BPM
-tempo - can be implied by an impulse at the start of each beat.
+tempo - can be implied by an impulse at the start of each beat, with the purpose of establishing metric and rhythms: the convolution with an impulse shifts the sound to impulse position, as stated in Section~\ref{subsec:conv}.
 For example, two impulses equally spaced build a binary division in the
@@
 proportions among the amplitudes of the sonic reincidences. The use of convolution with impulses in this context is explored in the music piece \emph{Little train of impulsive hillbillies}.
-The piece also encompass
-the creation of `sound amalgams' e.g. by granular synthesis (see Figure~\ref{fig:pulsoSubAgl}).
-This piece is a link to the contents of next section because of the explicit organization of notes in higher level structures.~\cite{MASSA}.
+These features encompass
+the creation of `sound amalgams' based on granular synthesis; see Figure~\ref{fig:pulsoSubAgl}.
+This piece is a link to next section contents because of the explicit organization of notes in higher level structures.
+The source code of the music piece is online, as part of the \massa\ toolbox\cite{MASSA}.
 
 \subsubsection{Moving source and receptor, Doppler effect}
-Given the audio source speed $s_s$, with positive values if the source moves away from receptor,
-and receptor speed $s_r$, positive when it gets closer to audio source, the frequency is given by the well-known shift in the Doppler effect:
+According to the description in Section~\ref{subsec:spac}, when an audio source or receptor is moving, its characteristics are ideally updated at each sample of the digital signal (if fast computational rendering is not at stake).
+Velocity components should be found for each ear.
+Hence, given the audio source speed $s_s$, with positive values if the source moves away from receptor, and receptor speed $s_r$, positive when it gets closer to audio source, the frequency is given by the well-known shift in the Doppler effect:
 
@@
 
-With both frequencies $f$ and $f_0$, and the IID and ITD from the new source
-position (see Section~\ref{subsec:spac}), it is possible to simulate the Doppler effect.
+With both frequencies $f$ and $f_0$, and the IID from the new source
+position, it is possible to simulate the Doppler effect.
 There is an addendum to improve the fidelity of the physical phenomena: to increase the received power, which may be understood as being proportional to the wave shrinking or expansion: $\Delta P=P_0\left(\frac{v_r-v_s}{343.2}\right)$,
@@
 If the audio source is in front of the receptor with $y_0$ meters of horizontal distance and $z_0$ meters of height, the distance is given by
-$D=\left\{ d_i=\sqrt{ y_{i}^{2}+z_{0}^{2} } \right\}_0^{\Lambda-1}$,
+$D_i=\left\{ d_i=\sqrt{ y_{i}^{2}+z_{0}^{2} } \right\}_0^{\Lambda-1}$,
 where $y_i=y_0+(v_s-v_r)\frac{i}{f_s}$ with $v_s$ and $v_r$ both
-horizontal, having null non-$y$ components. Amplitude changes with the distance and with the potency factor mentioned above (see Section~\ref{subsec:volume} for potency to amplitude conversion). Thus:
+horizontal, having null non-$y$ components. Amplitude changes with the distance and with the potency factor mentioned above (see Section~\ref{subsec:volume} for potency to amplitude conversion).
 
 \begin{equation}\label{eq:aDoppler}
-    A=\left\{ \frac{z_0}{d_i}A_{\Delta P}\right\}_0^{\Lambda-1} = \left\{ \frac{z_0}{\sqrt{y_i^2+z_0^2}} \sqrt{\frac{v_r-v_s}{343,2}+1}  \,\right\}_0^{\Lambda-1}
+    A_i=\left\{ \frac{z_0}{d_i}A_{\Delta P}\right\}_0^{\Lambda-1} = \left\{ \frac{z_0}{\sqrt{y_i^2+z_0^2}} \sqrt{\frac{v_r-v_s}{343,2}+1}  \,\right\}_0^{\Lambda-1}
 \end{equation}
@@
 \begin{equation}\label{eq:ffDoppler}
-    F=\{f_i\}_0^{\Lambda-1}=\left\{\frac{v_{sound} + v_r\frac{y_i}{\sqrt{z_0^2+y_i^2}}}{v_{sound}+v_s\frac{y_i}{\sqrt{z_0^2+y_i^2}}}f_0\right\}_0^{\Lambda-1}
-\end{equation}
-
-There is an implementation of the Doppler effect in \texttt{src/sections/3.py}~\cite{MASSA}.
-We considered this a musical usage given that it is not a basic method to introduce variation into a note
-and because it is a valued effect in audio plugins and electro-acoustic music~\cite{floEsp}.
+    F_i=\{f_i\}_0^{\Lambda-1}=\left\{\frac{v_{sound} + v_r\frac{y_i}{\sqrt{z_0^2+y_i^2}}}{v_{sound}+v_s\frac{y_i}{\sqrt{z_0^2+y_i^2}}}f_0\right\}_0^{\Lambda-1}
+\end{equation}
+
+There is a Python implementation of the Doppler effect, also considering the intersection between audio source and receptor, in \massa\ toolbox~\cite{MASSA}.
 
@@
 With the use of filters, the possibilities are even wider.
-Convolve a signal to have a reverberated version of it, to remove noise, to distort or to handle
+Convolve a signal to have a reverberated version of it, to remove its noise, to distort or to handle
 the audio aesthetically in other ways.
-For example, sounds originated from an old television or telephone can be simulated with a band-pass filter, allowing only frequencies between $300Hz$ and $2kHz$.
-By rejecting the frequency of an electric oscillation (usually $50Hz$ or $60Hz$) and the harmonics, one can remove noises caused by audio devices connected to the power supply (called hiss noise).
-Another musical application is to perform filtering in specific bands and to use those bands as an additional parameter to the notes.
+For example, sounds originated from an old television or telephone can be simulated with a band-pass filter, allowing only frequencies between $1kHz$ and $3kHz$.
+By rejecting the frequency of an electric oscillation (usually $50Hz$ or $60Hz$) and the harmonics, one can remove noises caused by audio devices connected to the power supply.
+A more musical application is to perform filtering in specific bands and to use those bands as an additional parameter to the notes.
 
@@
 Chains of these filters can perform complex and more accurate filtering routines.
-The music piece \emph{Noisy band} explores filters and many kinds and noise synthesis.~\cite{MASSA}
+The music piece \emph{Noisy band} explores filters and many kinds and noise synthesis.
+The source code is available online as part of \massa\ toolbox.
 
 A sound can be altered through different filtering and then mixed to create an effect known as \emph{chorus}.
-Inspired by what happens in a choir of singers, the same note is synthesized many times using small and potentially arbitrary modifications of parameters like center frequency, presence (or absence) of vibrato or tremolo and its characteristics, equalization, loudness, duration, etc.
-The versions of the sound are then mixed together (see Equation~\ref{eq:mixagem}).
-The music piece \emph{Child chorus} implements a chorus in many ways with different sounds.~\cite{MASSA}
+Based on what happens in a choir of singers, the sound is synthesized using small and potentially arbitrary modifications of parameters like center frequency, presence (or absence) of vibrato or tremolo and its characteristics, equalization, loudness, etc.
+As a final result, those versions of the original sound are mixed together (see Equation~\ref{eq:mixagem}).
+The music piece \emph{Child chorus} implements a chorus in many ways with different sounds.
+Its source code is available in \massa\ toolbox.
 
@@
 late reverberation can be achieved by a convolution with a section of pink, brown or black noise, with exponential decay of amplitude along time. Delay lines can be added as a prefix to the noise
-with the decay, and this accounts for both temporal parts of the reverberation:
-the early reflections and the late reverberation.
+with the decay, and this accounts for both time parts of the reverberation: the early reflections and the late reverberation.
 Quality can be improved by varying the geometric trajectory and filtering by each surface where wavefront reflected before reaching the ear in the first $100-200ms$ (mainly with a LP).
-The colored noise can be gradually introduced with a \emph{fade-in}: the initial moment given by the direct
+The colored noise can be gradually introduced with a \emph{fade-in}: the initial moment given by direct
 incidence of sound (i.e.\ without any reflection and given by ITD and IID), reaching its maximum at the beginning of the 'late
-reverberation', when the geometric incidences lose their relevance to the statistical properties of a decaying noise.
-As an example, consider $\Delta_1$ as the duration of the first section and $\Delta_R$ as the duration of total reverberation 
-(let also $\Lambda_1=\Delta_1 f_s$, $\Lambda_R=\Delta_R
-f_s$).
-Let $p_i$ be the probability of a sound being repeated in the
+reverberation', when the geometric incidences lose their relevance to the statistical properties of noise decay.
+As an example, consider $\Delta_1$ as the duration of the first section and $\Delta_R$ as the duration of total reverberation ($\Lambda_1=\Delta_1 f_s$, $\Lambda_R=\Delta_R
+f_s$). Let $p_i$ be the probability of a sound to repeat in the
 $i$-th sample. The amplitude decreases exponentially. Following
-Section~\ref{subsec:spac}, the samples $R^1$ of the impulse response of the first period can be described as:
+Section~\ref{subsec:spac}, reverberation $R_i^1$ of the first period can be described as:
 
@@
 \begin{split}
-	R^1  =\left\{r_i^1\right\}_0^{\Lambda_1-1}\;,\; \text{where}\;\;
+	R_i^1  =\left\{r_i^1\right\}_0^{\Lambda_1-1}\;,\; \text{where}\;\;
 	r_i^1  =\left\{
@@
 \noindent where $V_{dB}$ is the total decay in decibels, typically $-80dB$ or $-120dB$.
-The samples $R^2$ of the impulse response of the second period can be taken from a brown noise (or by a pink noise) with an exponential amplitude decay of the waveform:
+Reverberation $R_i^2$ of the second period can be emulated by a brown noise $R_i^m$ (or by a pink noise $R_i^r$) with exponential amplitude decay of the waveform:
 
 \begin{equation}\label{eq:p2rev}
-    R^2=\left\{r_i^2\right\}_{\Lambda_1}^{\Lambda_R-1}=\left\{10^{\frac{V_{dB}}{20}\frac{i}{\Lambda_R-1}}\,.\,r_i^m\right\}_{\Lambda_1}^{\Lambda_R-1}
-\end{equation}
-
-The reverberation impulse response is then:
+    R_i^2=\left\{r_i^2\right\}_{\Lambda_1}^{\Lambda_R-1}=\left\{10^{\frac{V_{dB}}{20}\frac{i}{\Lambda_R-1}}\,.\,r_i^m\right\}_{\Lambda_1}^{\Lambda_R-1}
+\end{equation}
+
+\noindent with
 
 \begin{equation}\label{eq:rev}
-	R=\left\{r_i\right\}_0^{\Lambda_R-1}\;,\; \text{where} \; r_i=\left\{
+	R_i=\left\{r_i\right\}_0^{\Lambda_R-1}\;,\; \text{where} \; r_i=\left\{
         \begin{array}{l l}
@@
 
-A sound with simulated reverberation can be achieved by a simple convolution of $R$ with the sound sequence $T$, as described in Section~\ref{subsec:filtros}.
-Reverberation is well known for causing great interest in listeners and to provide sonorities that are more enjoyable than the sound without any reverberation.
+A sound with simulated reverberation can be achieved by a simple convolution of $R_i$ (called reverberation impulse response) with the sound sequence $T_i$, as described in Section~\ref{subsec:filtros}.
+Reverberation is well known for causing great interest in listeners and to provide sonorities that are more enjoyable.
 Furthermore, modifications in the reverberation space consist in a common resource (almost a \textit{clich\'{e}}) to surprise and attract the listener.
-An implementation of the reverb recipe described here is in the file \texttt{src/sections/3.py}.~\cite{MASSA}.
+An implementation of the reverb recipe described here is in \massa\ toolbox~\cite{MASSA}
 
@@
 A pioneering implementation can be found in the Hammond Novachord synthesizer of 1938 and some variants are mentioned below~\cite{ADSR}.
-The canonical ADSR envelope is characterized by 4 parameters: attack duration (time in which sound reaches its maximum amplitude), decay duration (follows the attack immediately), level of sustained intensity (in which the intensity remains stable after the decay) and release duration (after the sustained section, this is the duration needed for amplitude to reach zero or final value).
+The canonical ADSR envelope is characterized by 4 parameters: attack duration (time in which sound reaches its maximum amplitude), decay duration (follows the attack immediately), level of sustained intensity (in which the intensity remains stable after the decay) and release duration (after sustained section, this is the duration needed for amplitude to reach zero or final value).
 Note that the sustain duration is not specified because it is the difference between the duration itself and the sum of the attack, decay and sustain durations.
+
+The ADSR envelope with durations $\Delta_A$, $\Delta_D$ and $\Delta_R$, with total duration $\Delta$ and sustain level $a_S$, given by the fraction of the maximum amplitude, to be applied to any sound sequence $T_i=\{t_i\}$, can be expressed as:
+
+\begin{equation}\label{eq:adsr}
+\begin{split}
+\{a_i\}_0^{\Lambda_A-1}  = & \left\{\xi\left(\frac{1}{\xi}\right)^{\frac{i}{\Lambda_A-1}}\right\}_0^{\Lambda_A-1} \quad \text{ or }\\ = & \left\{\frac{i}{\Lambda_A-1}\right\}_0^{\Lambda_A}\\
+\{a_i\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} = & \left\{a_S^{\frac{i-\Lambda_A}{\Lambda_D-1}}  \right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} \quad \text{ or } \\ = &  \left\{1-(1-a_S)\frac{i-\Lambda_A}{\Lambda_D-1}\right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1}\\
+\{ a_i \}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} = & \left\{ a_S \right\}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} \\
+\{ a_i \}_{\Lambda-\Lambda_R}^{\Lambda-1}  = & \left\{ a_S\left(\frac{\xi}{a_S} \right)^{\frac{i-(\Lambda-\Lambda_R)}{\Lambda_R-1}} \right\}_{\Lambda-\Lambda_R}^{\Lambda-1} \quad \text{ or } \\ = &  \left\{ a_S - a_S\frac{i+\Lambda_R-\Lambda}{\Lambda_R-1}\right\}_{\Lambda-\Lambda_R}^{\Lambda-1}
+\end{split}
+\end{equation}
+
+\noindent with $\Lambda_X=\lfloor \Delta_X . f_s \rfloor\;\;\forall\;\; X \; \in
+(A,D,R\;)$ with $\xi$ being a small value that provides a satisfactory \emph{fade in} and \emph{fade out}, e.g.\ $\xi=10^{\frac{-80}{20}}=10^{-4}\;$ or $\;\xi=10^{\frac{-40}{20}}=10^{-2}$.
+The lower the $\xi$, the slower the \emph{fade}, like the $\alpha$ illustrated in Figure~\ref{fig:transicao}.
+The right side of Equations~\ref{eq:adsr} can cater for both introduction and ending of sound from zero intensity because they are linear.
+Schematically, Figure~\ref{fig:adsr} shows the ADSR envelope in a classical implementation that supports many variations.
+For example, between attack and decay it is possible to add an extra partition where the maximum amplitude remains for more than a peak.
+Another common example is the use of more elaborated outlines of attack or decay.
+The music piece \emph{ADa and SaRa}, available
+in \massa, explores many configurations of the ADSL envelope.
+
+\begin{equation}\label{eq:adsrApl}
+\left\{t_i^{ADSR}\right\}_0^{\Lambda-1} =\{t_i . a_i\}_0^{\Lambda-1}
+\end{equation}
 
@@
 
-The ADSR envelope with durations $\Delta_A$, $\Delta_D$ and $\Delta_R$, with total duration $\Delta$ and sustain level $a_S$, given by the fraction of the maximum amplitude, to be applied to any sound sequence $T_i=\{t_i\}$, can be expressed as:
-
-\begin{equation}\label{eq:adsr}
-\begin{split}
-\{a_i\}_0^{\Lambda_A-1}  = & \left\{\xi\left(\frac{1}{\xi}\right)^{\frac{i}{\Lambda_A-1}}\right\}_0^{\Lambda_A-1} \quad \text{ or }\\ = & \left\{\frac{i}{\Lambda_A-1}\right\}_0^{\Lambda_A}\\
-\{a_i\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} = & \left\{a_S^{\frac{i-\Lambda_A}{\Lambda_D-1}}  \right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1} \quad \text{ or } \\ = &  \left\{1-(1-a_S)\frac{i-\Lambda_A}{\Lambda_D-1}\right\}_{\Lambda_A}^{\Lambda_A+\Lambda_D-1}\\
-\{ a_i \}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} = & \left\{ a_S \right\}_{\Lambda_A+\Lambda_D}^{\Lambda-\Lambda_R-1} \\
-\{ a_i \}_{\Lambda-\Lambda_R}^{\Lambda-1}  = & \left\{ a_S\left(\frac{\xi}{a_S} \right)^{\frac{i-(\Lambda-\Lambda_R)}{\Lambda_R-1}} \right\}_{\Lambda-\Lambda_R}^{\Lambda-1} \quad \text{ or } \\ = &  \left\{ a_S - a_S\frac{i+\Lambda_R-\Lambda}{\Lambda_R-1}\right\}_{\Lambda-\Lambda_R}^{\Lambda-1}
-\end{split}
-\end{equation}
-
-\noindent with $\Lambda_X=\lfloor \Delta_X . f_s \rfloor\;\;\forall\;\; X \; \in
-(A,D,R\;)$ and $\xi$ being a small value that provides a satisfactory \emph{fade in} and \emph{fade out}, e.g.\ $\xi=10^{\frac{-80}{20}}=10^{-4}\;$.
-The lower the $\xi$, the slower the \emph{fade}, like the $\alpha$ illustrated in Figure~\ref{fig:transicao}.
-The right side of Equations~\ref{eq:adsr} can cater for both introduction and ending of sound from zero intensity because they are linear.
-Schematically, Figure~\ref{fig:adsr} shows the ADSR envelope in a classical implementation that supports many variations.
-For example, between attack and decay it is possible to add an extra partition where the maximum amplitude remains for more than a peak.
-Another common example is the use of more elaborated outlines of attack or decay.
-To aply the ADSR envelope, simply multiply the amplitudes of Equation~\ref{eq:adsr} by the samples of the note or musical except:
-\begin{equation}\label{eq:adsrApl}
-\left\{t_i^{ADSR}\right\}_0^{\Lambda-1} =\{t_i . a_i\}_0^{\Lambda-1}
-\end{equation}
-
-The music piece \emph{ADa and SaRa} explores many configurations of the ADSR envelope.~\cite{MASSA}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\afterpage{\blankpage}
+
+%\clearpage
 
 \section{Organization of notes in music}\label{notasMusica} \label{sec:notesMusic}
-Let $S=\left\{  s_j=T^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1}$ be a sequence of $H$ musical events $s_j$.
-Consider $S$ as a `musical structure', composed by events which are also musical structures themselves, e.g.\ notes.
-This section is dedicated to techniques that make $S$ aesthetically appealing for audition.
+Consider $S_j=\left\{  s_j=T_i^j=\{t_i^j\}_{i=0}^{\Lambda_j-1} \right\}_{j=0}^{H-1}$, the sequence $S_j$ of $H$ musical events $s_j$.
+Let $S_j$ be a `musical structure', composed by events $s_j$ which are musical structures themselves, e.g.\ notes.
+This section is dedicated to techniques that make $S_j$ interesting and enjoyable for audition.
 More specifically, what follows is a summary of academic music composition theory and praxis.
@@
 Even so, we understand that this content is very useful for synthesizing music given all the content above.
-The concepts are given algorithmic implementations~\cite{MASSA} and can be further formalized~\cite{topos},
+The concepts are given algorithmic implementations in \mass~\cite{SI} and can be further formalized~\cite{topos},
 although at a cost of intelligibility which we chose to avoid in this exposition.
 
-The elements of $S$ may be overlapped by mixing them together, as in
-Equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, for building intervals and chords.
-This reflects the `vertical thought' in music. On the other hand, the concatenation of events
-in $S$, as in Equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao}, builds melodic sequences and rhythms, which are associated with the `horizontal thought'.
+The elements of $S_j$ can be overlapped by mixing them together, as in
+Equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, for building intervals and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of events
+in $S_j$, as in Equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao}, builds melodic sequences and rhythms, which are associated with the `horizontal thought'.
 The fundamental frequency $f$ and the starting moment
-(attack or onset) are generally the most important characteristics of the elements $s_j$ in $S$.
-This makes it possible to create music that relies heavily on pitches (both for harmony and melody) and on rhythms.
+(attack) are generally the most important characteristics of the elements $s_j$ in $S_j$.
+This makes it possible to create music made by pitches (both harmony and melody) and by temporal metrics and rhythms.
 We will start by considering these aspects of musical organization as they are more traditional in music theory and usually easier to understand.
-One will verify in the scripts that illustrate these concepts that it is considerably straightforward to
-reach sonic materials of some artistic significance by
-extrapolating musical paradigms established on traditional notions of melody, harmony and rhythm.
 
@@
 \subsubsection{Tuning}
-Doubling the frequency is equivalent to ascending one octave\footnote{An pitch
-is both considered equivalent and different from any of its octaves.
-It is different because one or more octaves higher or lower corresponds to
-a higher or lower pitch.
-It is equivalent in the sense that a pitch is very consonant to its octaves
-and that any other pitch holds a similar relation to a pitch and its octaves.} ($f=2f_0$).
-The octave division in twelve steps is the canon of classical western music.
-Its usage has also been observed
-outside western tradition, e.g. in ceremonial/religious and ethnic contexts~\cite{Wisnick}.
+Doubling the frequency is equivalent to ascending one octave ($f=2f_0$). The octave division in twelve steps is the canon of classical western music. Its usage has also been observed
+outside western tradition, e.g.
+in ceremonial/religious and ethnic contexts~\cite{Wisnick}.
 The semitones need not to be equivalent, as will become clear in the next paragraphs, but, roughly, the factor given by $\varepsilon=2^{\frac{1}{12}}$ defines a semitone.
@@
 Twelve successive semitones (or half steps), equidistants to the human ear, yield an
-octave. In other words, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
+octave, equivalent to the human ear. In other words, if $f=2^{\frac{1}{12}}f_0$, there is a semitone between
 $f_0$ and $f$.
@@
 This absolute accuracy is usual in computational implementations.
-Real music instruments, however, often present semitones that are not exactly $2^{\frac{1}{12}}$
-for two main reasons: 1) the harmonic series does not match multiples of $2^{\frac{1}{12}}$,
-and 2) because the harmonics of real sounds are not perfect multiples of the fundamental frequency.
+Real music instruments, however, often present semitones that are not exactly $2^{\frac{1}{12}}$ because the (real) harmonics of the notes are not perfect multiples of the fundamental frequency.
 The fixed interval $\varepsilon=2^{\frac{1}{12}}$ characterizes an equally tempered tuning but there are other tunings.
-The first formalizations of tunings (in the scientific literature) date from around two thousand years before the advent of the equal temperament~\cite{Roederer}.
-Two emblematic tunings are:
+The first formalizations of tunings (that the scientific tradition has repported) date from around two thousand years before the advent of the equal temperament~\cite{Roederer}.
+). Two emblematic tunings are:
 \begin{itemize}
@@
 
-In order to account for microtonality\footnote{Microtonality is the use of intervals smaller
+In order to account for micro-tonality\footnote{Micro-tonality is the use of intervals smaller
 than one semitone and has ornamental and structuring functionalities in music.
@@
 is proposed with the equal division of the octave in $53$ notes:
-$\varepsilon=2^{\frac{1}{53}}$~\cite{microtonalidade}.
-Notice that if $S$ is a pitch sequence related by means of $\varepsilon_1$,
-the sequence with the same notes, but related by $\varepsilon_2$, is 
-$\left\{s_i'\right\}=\left\{
+$\varepsilon=2^{\frac{1}{53}}$.~\cite{microtonalidade}
+Note that if $S_i$ is a pitch sequence related by means of $\varepsilon_1$, the sequence with the same notes, but related by $\varepsilon_2$, is 
+$S_i'=\left\{s_i'\right\}=\left\{
 s_i \frac{\varepsilon_1}{\varepsilon_2}\right\}$. The music piece \emph{Micro
-tone} uses microtonal features.~\cite{MASSA}
+tone} uses microtonal features and its code
+is part of \massa\ toolbox.
 
 \subsubsection{Intervals}\label{subsec:intervalos}
-The frequency gap between two pitches is called an "interval" in musical jargon.
 Using the ratio $\varepsilon=2^{\frac{1}{12}}$ between note frequencies (i.e.\ one semitone) the intervals in the twelve note system can be represented by integers.
@@
 Major (M) and minor (m) thirds and sixths are the imperfect consonances.
-	Minor seconds and major sevenths are the strong dissonances.
-	Major seconds and minor sevenths are the weak dissonances.
+Minor seconds and major sevenths are the harsh dissonances.
+Major seconds and minor sevenths are the mild dissonances.
 Perfect fourth is a special case, as it is a perfect consonance when considered as an inversion of the perfect fifth and a dissonance or an imperfect consonance otherwise.
@@
 This interval is consonant in some cultures.
-	For tonal music, the tritone indicates the dominant (chord or harmonic field, see Section~\ref{subsec:harmonia}) and seeks urgent resolution into a third or sixth.
+	For tonal music, the tritone indicates dominant (chord or harmonic field, see Section~\ref{subsec:harmonia}) and seeks urgent resolution into a third or sixth.
 Due to this instability it is considered a dissonant interval.}
@@
 \begin{itemize}
-	\item Intervals are inspected first by the number of steps between notes in some scale.
+	\item Intervals are inspected first by the number of steps between notes. The simple intervals are: first (unison), second, third, fourth, fifth, sixth, seventh and eighth (octave).
+		Each of these intervals are related to one step less the their names suggest: a third is an interval with two steps.
 		As can be noticed in Table~\ref{eq:intervalos}, one step is not one semitone.
@@
 		A scale for now can be regarded as any arbitrary monotonic sequence of pitches and will be discussed in the next section.
-		The simple intervals are: first (unison), second, third, fourth, fifth, sixth, seventh and eighth (octave).
-		Each of these intervals are related to one step less then their names suggest: a third is an interval with two steps.
 	\item The intervals are represented by numeric digits, e.g. 1, 3, 5 are a unison, a third and a fifth, respectively.
@@
                 i.e.\ unison, fourth, fifth and octave -- are 'perfect'.
-		The imperfect consonances -- i.e.\ thirds and sixths -- and dissonances -- i.e.\ seconds and sevenths -- can be 'major' or 'minor'.
+The imperfect consonances -- i.e.\ thirds and sixths -- and dissonances -- i.e.\ seconds and sevenths -- can be major and minor.
 The tritone is an exception to this rule because it is a dissonant interval and cannot be major or minor.
 	\item The perfect fourth can be a perfect consonance or a dissonance according to the context and theoretical background.
-		As a general rule, it is considered a consonance except when it is followed by a third or a fifth
-		by the movement of one of the notes.
-	\item The tritone is a dissonance in Western music because
+As a general rule, it can be considered a consonance except when it is followed by a third or a fifth by the movement one of the notes.
+	\item Tritone is a dissonance in Western music because
 		it characterizes the ``dominant'' chord in the tonal system (see Section~\ref{subsec:harmonia}) and represents (or yields) instability.
@@
 aug3 has five semitones). The augmented forth
-is also called tritone (aug4, tri or TT).
+is also called tritone (aug4 ~ tri ~ TT).
 	\item A perfect interval or a minor interval (m2, m3, m6 or m7), decreased by one semitone results in a diminished interval. The
-diminished fifth is also called tritone (dim5, tri or TT).
+diminished fifth is also called tritone (dim5 ~ tri ~ TT).
 
 	\item An augmented interval increased by one semitone results in a `doubly augmented' interval and a diminished interval decreased by one semitone results in a `doubly diminished' interval.
-	\item Notes played simultaneously yield a harmonic interval.
-	\item Notes played as a sequence in time yield a
-                melodic interval.  When the lowest note comes first there is an ascending interval,
-		while in a descending interval the highest note comes first.
+	\item Notes played simultaneously yields a harmonic interval.
+	\item Notes played as a sequence in time yields a
+                melodic interval.  When the lowest note comes first there is an ascending interval, while a descending interval is observed when the highest comes note first. 
 	\item A simple interval is inverted if the lower pitch is raised one octave, or if
@@
                 inverted augmented interval results in a diminished interval
-                and vice-versa (inverting a doubly-augmented results in a
+                and vice-versa (inverting a doublyi-augmented results in a
                 doubly-diminished and vice-versa, etc).
                 An inverted perfect interval is a perfect interval as well.
-	\item An interval wider than an octave (e.g. ninth, tenth, eleventh) is called a 'compound interval' and is classified in terms of the interval between the same notes but in the same octave.
-		Their notation can be achieved by adding
-                7 to the corresponding simple interval, e.g. P11 is an octave plus a forth ($7 + P4 = P11$), M9 is an octave plus a major second ($7 + M2 = M9$).
+	\item An interval wider than an octave (e.g.
+ninth, tenth, eleventh) is called a 'compound interval' and is classified in terms of the interval between the same notes but in the same octave.
+Their notation can be achieved by adding a
+                7 to the interval: P11 is an octave plus a forth ($7 + P4 = P11$), M9 is an octave plus a major second ($7 + M2 = M9$).
 \end{itemize}
@@
 Henceforth, in a \textit{C flat} major scale, the tonic -- first degree -- is \textit{C flat}, not \textit{B}, and the leading tone -- seventh degree -- is \textit{B flat}, not \textit{A sharp} or \textit{C double flat}.
-To grasp what this entails for intervals, let the second degree (second note) of a scale be one semitone from the first degree.
-Consider also the leading tone (i.e. the seventh degree at one ascending semitone from the first degree). There is a diminished third between the seventh and second scale degrees~\cite{Lacerda}.
-Notice that the dim3 is only two semitones wide, as is the major second (or e.g. a doubly-augmented unison!).
+To grasp what this entails for intervals, let the second degree (second note) of a scale to be one semitone from the first degree.
+Let also the leading tone (i.e.
+the seventh degree at one ascending semitone from the first degree).
+There is a diminished third between the seventh and second scale degrees~\cite{Lacerda}.
+Notice that the dim3 is only two semitones wide, as is the major second (or e.g. an doubly-augmented unisson!).
 
 This description summarizes the traditional theory of musical intervals~\cite{Lacerda}.
-The music piece \emph{Intervals between pitches} explores these intervals in both independent and related ways.~\cite{MASSA}
+The music piece \emph{Intervalos entre alturas} explores these intervals in both independent and related manners.
+The source code is available online in the \massa\ toolbox~\cite{MASSA}.
 
 \subsubsection{Scales}\label{subsec:escalas}
-A scale is an ordered set of pitches. Strictly speaking, any (ordered) set of pitches can be considered a scale.
-The complexity about musical scales lean mostly on the tradition, i.e.
+A scale is an ordered set of pitches.
+Strictly speaking, any (ordered) set of pitches can be considered a scale.
+The complexity about musical scales lean on the tradition, i.e.
 on the scales and their uses which result from practice throughout history.
@@
 There are 5 perfectly symmetric divisions of the octave within the chromatic scale.
-These divisions are often regarded as scales themselves owing to the easy and peculiar uses they admit.
+These divisions are often regarded as scales themselves owing to the easy and peculiar uses they entail.
 
@@
 \begin{aligned}
-	\text{chromatic}    & = E^c    & = & \{e_i^c\}_0^{11}   & = & \{0,1,2,3,4,5,6,7,8,9,10,11\} & = & \{i\}_0^{11}\\
-	\text{whole tones}  & = E^{wt} & = & \{e_i^{wt}\}_0^{5} & = & \{0,2,4,6,8,10\}               & = & \{2i\}_0^{5} \\
-	\text{minor thirds} & = E^{mt} & = & \{e_i^{mt}\}_0^{3} & = & \{0,3,6,9\}                    & = & \{3i\}_0^3 \\
-	\text{major thirds} & = E^{Mt} & = & \{e_i^{Mt}\}_0^{2} & = & \{0,4,8\}                      & = & \{4i\}_0^2\\
-	\text{tritones}     & = E^{tt} & = & \{e_i^{tt}\}_0^{1} & = & \{ 0, 6 \}                     & = & \{6i\}_0^1
+	\text{chromatic}    & = E_i^c    & = \{e_i^c\}_0^{11}   & =  \{0,1,2,3,4,5,6,7,8,9,10,11\} & = \{i\}_0^{11}\\
+	\text{whole tones}  & = E_i^{wt} & = \{e_i^{wt}\}_0^{5} & = \{0,2,4,6,8,10\}               & = \{2.i\}_0^{5} \\
+	\text{minor thirds} & = E_i^{mt} & = \{e_i^{mt}\}_0^{3} & = \{0,3,6,9\}                    & = \{3.i\}_0^3 \\
+	\text{major thirds} & = E_i^{Mt} & = \{e_i^{Mt}\}_0^{2} & = \{0,4,8\}                      & = \{4.i\}_0^2\\
+	\text{tritones}     & = E_i^{tt} & = \{e_i^{tt}\}_0^{1} & = \{ 0, 6 \}                     & = \{6.i\}_0^1
 \end{aligned}
@@
 
-For example, the third note of the whole tone scale with $f_0=200Hz$ is $f_3=\varepsilon^{e_2^{wt}}. f_0 = 2^{\frac{4}{12}} . 200 \approx 251.98
+For example, the third note of the whole tone scale with $f_0=200Hz$ is $f_3=\varepsilon^{e_2^{wt}}. f_0 = 2^{\frac{4}{12}} . 200 \approxeq 251.98
 Hz$.
@@
 Section~\ref{estCic} discusses symmetries.
-The music piece \emph{Cristais} uses each one of these scales both melodically and harmonically.~\cite{MASSA}
+The music piece \emph{Cristais} uses each one of these scales, in both melodic and harmonic counterpart and their source code is part of \massa\ toolbox.
 
@@
 \begin{aligned}
-	\text{Aeolian}    & = \text{natural minor scale}  & = & \\
-			  & = E^m = \{e_i^m\}_0^6       & = & \{0,2,3,5,7,8,10\} \\
-	\text{Locrian}    & = E^{lo} = \{e_i^{lo}\}_0^6 & = & \{0,1,3,5,6,8,10\} \\ 
-	\text{Ionian}     & = \text{major scale}          & = &  \\
-			  & = E^M = \{e_i^M\}_0^6       & = & \{0,2,4,5,7,9,11\} \\
-	\text{Dorian}     & = E^{d} = \{e_i^{d}\}_0^6   & = & \{0,2,3,5,7,9,10\} \\
-	\text{Phrygian}   & = E^{p} = \{e_i^{p}\}_0^6   & = & \{0,1,3,5,7,8,10\} \\
-	\text{Lydian}     & = E^{l}=\{e_i^{l}\}_0^6     & = & \{0,2,4,6,7,9,11\} \\
-	\text{Mixolydian} & = E^{mi} = \{e_i^{mi}\}_0^6 & = & \{0,2,4,5,7,9,10\}
+	\text{aeolian}    & = \text{natural minor scale}  & = & \\
+			  & = E_i^m = \{e_i^m\}_0^6       & = & \{0,2,3,5,7,8,10\} \\
+	\text{locrian}    & = E_i^{lo} = \{e_i^{lo}\}_0^6 & = & \{0,1,3,5,6,8,10\} \\ 
+	\text{ionian}     & = \text{major scale}          & = &  \\
+			  & = E_i^M = \{e_i^M\}_0^6       & = & \{0,2,4,5,7,9,11\} \\
+	\text{dorian}     & = E_i^{d} = \{e_i^{d}\}_0^6   & = & \{0,2,3,5,7,9,10\} \\
+	\text{phrygian}   & = E_i^{p} = \{e_i^{p}\}_0^6   & = & \{0,1,3,5,7,8,10\} \\
+	\text{lydian}     & = E_i^{l}=\{e_i^{l}\}_0^6     & = & \{0,2,4,6,7,9,11\} \\
+	\text{mixolydian} & = E_i^{mi} = \{e_i^{mi}\}_0^6 & = & \{0,2,4,5,7,9,10\}
 \end{aligned}
@@
 % This is the main reason why the locrian is sometimes not considered a diatonic scale.
+
 Diatonic scales follow a circular pattern of successive intervals \textit{tone, tone, semitone, tone, tone, tone, semitone}. Thus, it is possible to write:
@@
 For example, a brief inspection reveals that
-$e_i^{l}=d_{(i+2)\%7}+e_{i-1}^{l}$. Then, $\kappa=2$ for the Lydian scale.
+$e_i^{l}=d_{(i+2)\%7}+e_{i-1}^{l}$. Then, $\kappa=2$ for the lydian scale.
 
@@
 \begin{split}
-\text{natural minor}&  = E^m = \{e_i^m\}_0^6 = \\
+\text{natural minor}&  = E_i^m = \{e_i^m\}_0^6 = \\
                                            &  = \{0,2,3,5,7,8,10\} \\
-\text{harmonic minor}                      &  = E^{mh} = \{e_i^{mh}\}_0^6 = \\
+\text{harmonic minor}                      &  = E_i^{mh} = \{e_i^{mh}\}_0^6 = \\
                                            &  = \{0,2,3,5,7,8,11\} \\
-\text{melodic minor}                       &  = E^{mm} = \{e_i^{mm}\}_0^{14} = \\
+\text{melodic minor}                       &  = E_i^{mm} = \{e_i^{mm}\}_0^{14} = \\
                                            &  = \{0,2,3,5,7,9,11,12,10,8,7,5,3,2,0\} \\
@@
 
-The different ascending and descending contours of the melodic minor scale is required in tonal harmony contexts.
-The minor scale has one whole tone between the seventh and eighth (or first) degrees but
+The different ascending and descending contour of the melodic minor scale is required in tonal harmony contexts.
+The minor scale has one whole tone between the seventh and eigth (or first) degrees but
 the separation by one semitone is critical to the polarization of the first degree.
@@
 \begin{split}
-H = & \{h_i\}_0^{19}= \\
+H_i = & \{h_i\}_0^{19}= \\
     =  & \{ 0,12,19+0.02,  24,28-0.14, 31+0.2, 34-0.31, \\
@@
 Many other scales can be expressed using the framework exposed in this section, e.g. the pentatonic scales and the modes of limited transposition of Messiaen~\cite{Messiaen}.
+
 One last observation: the words \emph{scale} and \emph{mode} are often used as synonyms both in the literature and in colloquial discussions.
@@
 \begin{split}
-\text{major triad} = A^M= \{a_i^M\}_0^2=\{0,4,7\} \\ 
-\text{minor triad} = A^m = \{a_i^m\}_0^2=\{0,3,7\} \\
-\text{diminished triad} = A^d = \{a_i^d\}_0^2=\{0,3,6\} \\
-\text{augmented triad} =  A^a = \{a_i^a\}_0^2=\{0,4,8\}
+\text{major triad} = A_i^M= \{a_i^M\}_0^2=\{0,4,7\} \\ 
+\text{minor triad} = A_i^m = \{a_i^m\}_0^2=\{0,3,7\} \\
+\text{diminished triad} = A_i^d = \{a_i^d\}_0^2=\{0,3,6\} \\
+\text{augmented triad} = A_i^a = \{a_i^a\}_0^2=\{0,4,8\}
 \end{split}
@@
 To have another third superimposed to the fifth, it is sufficient to add $10$ as the highest note in order to form a tetrad with minor seventh, or add $11$ in order to form a tetrad with major
-seventh. Inversions and open positions can be obtained by the 
-addition of $\pm 12$ to the selected pitch. Incomplete triadic chords, with extra notes ('dirty' chords), and non-triadic
+seventh. Inversions and open positions can be obtained with the 
+addition of $\pm 12$ to the selected component. Incomplete triadic chords, with extra notes ('dirty' chords), and non-triadic
 are also common.
@@
 		There are theoretical discussion about why this happens, and the most usual arguments are that the fifth is the first (non-octave) harmonic of a note and that the harmonics of the fifth are in the harmonics of the fundamental.
-		Important here is to grasp the fact that musical theory and practice assures that the fifth enhances the sensation that
-		the root of a chord is the root.
+		Important here is to grasp the fact that musical theory and practice assures that the fifth establishes the fundamental as the root of a chord.
         \item Major or minor thirds from the root entails major or minor chord qualities.
@@
 
-\subsection{Tonal, modal and atonal music, harmonies, harmonic expansion and modulation}\label{subsec:harmonia}
-The omission of basic tonal structures is the key to achieving modal and atonal harmonies\footnote{To avoid very complicated discussions,
-the term 'harmony' is used in this article in a very loose sense:
-considerations, grouped by music theory as harmony, about how sounds are analyzed by hearing.
-These considerations stem substantially from chords and scales and the absence of them.}. In the absence of minimal tonal organization,
+\subsection{Atonal and tonal harmonies, harmonic expansion and modulation}\label{subsec:harmonia}
+Omission of basic tonal structures is the key to achieving modal and atonal harmonies. In the absence of minimal tonal organization,
 harmony is (usually) considered modal if the notes match with some diatonic scale (see Equations~\ref{eq:escalas}) or if there is only a small number of notes.
-If basic tonal progressions are absent and notes do not match any diatonic scale and are diverse enough and dissonant (between themselves) to avoid reduction of the notes by polarization\footnote{By polarization we mean the presence of notes that are way more important than others and to which the other notes are ornaments or subordinates.}, the harmony is atonal.
+If basic tonal progressions are absent and notes do not match any diatonic scale and are enough diverse and dissonant (between themselves) to avoid reduction of the notes by polarization\footnote{By polarization we mean having some notes that are way more important than others and to which the other notes are ornaments or subordinates.}, the harmony is atonal.
 In this classification, the modal harmony is not tonal or atonal and is reduced to the incidence of notes within a (most often diatonic) scale and to the absence of tonal structures.
-Atonal harmony is hard to be fulfilled and, indeed, no matter how dissonant and diverse
+Following this concept, one observes that atonal harmony is hard to be realized and, indeed, no matter how dissonant and diverse
 a set of notes is, tonal harmonies arise very easily if not avoided~\cite{harmEXT}.
@@
 \subsubsection{Atonal harmony}\label{sec:atonal}
-In fact, atonal music techniques aim at avoiding that a direct relation of the notes with modes and tonality be established by the listener.
+In fact, atonal music techniques avoid that a direct relation of the notes with modes and tonality be established.
 Manifesting such atonal structures is of such difficulty that the dodecafonism emerged.
-Dodecafonism consists in using a sequence of pitches (ideally the 12 pitches of the chromatic scale),
-and performing each pitch, one by one, in the same
+The purpose of dodecafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
 order.
@@
 Nevertheless, the western listener automatically searches for tonal elements in music and obstinately finds them by unexpected and tortuous paths.
-The use of dissonant intervals (especially tritones) without resolution reinforces the absence of tonality. In this context, while creating a musical
+The use of dissonant intervals (especially tritones) without resolution reinforces the absence of tonality.
+In this context, while creating a musical
 piece, it is possible to:
 \begin{itemize}
-     \item Repeat pitches. By considering an immediate repetition as an extension of the previous incidence, the use of the same pitch in sequence does not add relevant information.
-     \item To play adjacent pitches (of the dodecafonic sequence) at the same time, making harmonic intervals and chords.
-     \item Use durations and pauses with freedom while respecting the order of incidence of the pitches.
+     \item Repeat notes. By considering immediate repetition as an extension of the previous incidence, the use of the same note in sequence does not add relevant information.
+     \item To play adjacent notes (e.g. of a dodecafonic progression) at the same time, making harmonic intervals and chords.
+     \item Use durations and pauses with liberty, respecting notes order.
      \item Vary note sequences by temporal magnification and translation or pitch transposition and by pitch sequence inversion, retrograde and retrograde inversion. See Sections~\ref{subsec:motivos}
-     and~\ref{subsec:usosmusicais3} for the meaning of these terms.
-     \item Make variations in timbre (e.g. orchestration), articulation, spatialization, among other possibilities in the presentation of the same pitches.
+     and~\ref{subsec:usosmusicais3} for what these terms mean.
+     \item Make variations in orchestration, articulation, spatialization, among other possibilities in presenting the same notes.
 \end{itemize}
 
-The atonal harmony can be observed, paradigmatically, within these presented conditions (which is a simple dodecaphonic model).
-Part of what was written by great dodecafonic composers,
+The atonal harmony can be observed, paradigmatically, within these presented conditions (which is a simple dodecaphonic model). Most of what was written by great dodecafonic composers,
 e.g. Alban Berg and even Schoenberg, had the purpose of mixing tonal and atonal techniques.
-Most frequently, atonal music is not strictly dodecafonic, but "serial", i.e.
-they use the same kind of techniques as stated above, based in a sequence of pitches (called the series or row), but not within a 12 tone sequence and using sequences of other parameters (e.g.
-duration, timbre).
-Serialism and pos-serialism is briefly considered in Secion~\ref{sec:ser}.
+Most frequently, atonal music is not strictly dodecafonic, but "serial", i.e. they use the same kind of techniques based in a sequence of notes (called the series or row).
 
@@
 In the XX century, music with emphasis on sonorities/timbres and rhythm, extended the concepts of tonality and harmony.
-Even so, tonal harmony is very common in artistic movements and commercial venues.
+Even so, tonal harmony is very often in artistic movements and commercial venues.
 In addition, dodecafonism itself is sometimes considered of tonal nature because it was conceived to deny tonal characteristics of polarization.
@@
 Equations~\ref{triades}-- built on the root note of each degree of a scale  -- such as listed in Equations~\ref{eq:escalas} --  form the pillars of harmony.
-Tonal (and modal) harmony theory aims at understanding the incidence of chord progressions and chaining rules.
-Even a monophonic melody entails harmonic fields, making it possible to perceive a chord progression even in unaccompanied melodies.
+Tonal (and modal) harmony aims at understanding the incidence of chord progressions and chaining rules.
+Even a monophonic melody entails harmonic fields, making it possible to perceive the chord progression even in unaccompanied melodies.
 
@@
 The remaining triads are associated to these three most important chords. In the major scale, the associated relative (relative tonic \textit{Tr}, relative
-subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built a third below, and the associated counter-relative (counter-relative tonic \textit{Tc}, counter-relative subdominant \textit{Sc} and the counter-relative dominant \textit{Dc}) is the triad built in a third above.
-In the minor scale the same happens, but the triad a third below is called counter-relative (tC, sC) and the triad a third above is called relative (tR,
+subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built a third below, and the associated counter-relative (counter-relative tonic \textit{Ta}, counter-relative subdominant \textit{Sa} and the counter-relative dominant \textit{Da}) is the triad built in a third above.
+In the minor scale the same happens, but the triad a third below is called counter-relative (tA, sA) and the triad a third above is called relative (tR,
 sR). The precise functions and musical effects of these chords are
@@
 \centering
-\caption{Summary of tonal harmonic functions on the major scale.
-	The tonic is the musical center, the dominant goes to the tonic and the subdominant moves the music away from
+\caption{Summary of tonal harmonic functions on the major scale. Tonic is the musical center, the dominant goes to the tonic and the subdominant moves the music away from
 the tonic. The three chords can, in principle, be freely replaced by their
-respective relatives or counter-relatives.}
+respective relative or counter-relatives.}
 \begin{tabular}{l | c | r}
@@
 on a fifth above or a fifth below, respectively. These individual dominants and subdominants,
-in the same way, also have subdominants and dominants of their own. Given a tonality, any chord can occur, no matter
+in the same way, have also subdominants and individual dominants of their own. Given a tonality, any chord can occur, no matter
 how distant it is from the most basic chords and from the notes of the scale.
@@
 There are two mediants, or 'chromatic mediants', for each chord: the upper
-mediant, with the root at the third of the original chord;
-and the lower mediant, with the fifth at the third of original chord.
-Both chords also are separated by a third from the original chord, like the relatives and counter-relatives,
-but with a chromatic alteration regarding the original chord.
-If two chromatic alterations exist, i.e.\ two notes altered by one semitone each regarding the original chord,
-it is a 'doubly chromatic mediant'.
-Again, there are two forms for each chord: the upper form, with the third on the fifth of the original triad;
-and the lower form, with the third on the root
+mediant, formed with the root at the third of original chord; and the lower mediant, formed by the fifth at the third of original chord.
+Both chords also are formed by a third, but with a chromatic alteration regarding the original chord.
+If two chromatic alterations exist, i.e.\ two notes altered by one semitone each regarding the original chord, it is a 'doubly chromatic mediant'.
+Again, there are two forms for each chord: the upper form, with a third in the fifth of the original triad; and the lower form, with a third in the root
 of the original triad.
@@
 (Recall that relatives and counter-relatives have opposite major-minor quality.)
-Mediants are considered of advanced tonalism,
-sometimes even considered as an expansion and dissolution of tonalism,
+This relation between chords is considered advanced tonalism, sometimes even considered as an expansion and dissolution of tonalism,
 with strong and impressive effects although they are simple, consonant triads.
-Chromatic mediants are used since the late 19th century by Wagner, Lizt, Richard Strauss, among others~\cite{Harmonia,Salzer}.
+Chromatic mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss, among others~\cite{Harmonia,Salzer}.
 
@@
 Keys are always (thought of as) related by fifths and their relatives and counter-relatives. Some ways to perform modulation include:
+
 \begin{itemize}
-    \item Transposing the discourse to a new key, without any preparation.
-	    It is a common Baroque procedure although incident in other periods. Sometimes it is 
+    \item Transposing the discourse to a new key, without any preparation. It is a common Baroque procedure although incident in other periods as well. Sometimes it is 
     called phrasal modulation or unprepared modulation.
     \item Careful use of an individual dominant, and perhaps also the individual
-    subdominant, to confirm a change in key.
+    subdominant, to confirm change in key and harmonic field.
     \item Use of chromatic alterations to reach a chord in the new key by starting from a chord in the previous key. Called chromatic modulation.
@@
     \item Changing the function, without changing the notes, of a chord to
-    reach a new key. This procedure is called enharmony.
+    contemplate a new key. This procedure is called enharmony.
     \item Maintaining the tonal center and changing the key quality from major to minor
@@
 The dominant has great importance and is a natural pivot in modulations,
-a fact that emphasizes the relevance of the 'circle of fifths'~\cite{Harmonia,Salzer,Koellheuteur,Harmony}.
-Other inventive ways to modulate are possible, to point but one common example, the minor thirds tetrad ($E_i^{tm}$ in Equations~\ref{escSim}) can be sustained as a bridge between tonalities, with the possible use of its both tritones.
-The music piece \emph{Acorde cedo} explores these chord relations.~\cite{MASSA}.
+a fact that leads to the circle of fifths~\cite{Harmonia,Salzer,Koellheuteur,Harmony}.
+Other inventive ways to modulate are possible, to point but one common example, the minor thirds tetrad ($E_i^{tm}$ in Equations~\ref{escSim}) can be sustained to bridge to other tonalities, with the ease of its both tritones.
+The music piece \emph{Acorde cedo} explores these chord relations, and is implemented online as part of \massa\ toolbox~\cite{MASSA}.
 
 \subsection{Counterpoint}\label{subsec:contraponto}
-Counterpoint is a set of techniques for the conduction of simultaneous melodic lines, or ``voices''.
-The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues~\cite{Fux,SchoenbergContra}. It is possible to
+Counterpoint is the conduction of simultaneous melodic lines, or ``voices''.
+The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues~\cite{Fux,SchoenbergContra}.
+It is possible to
 summarize counterpoint rules, and it is known that Beethoven --
@@
 neighbor note in the scale sound coherent and are prioritized. When having 3 or more voices, the melodic
-relevance lies mainly in the highest and then in the lowest of the voices~\cite{Fux,Tragtenberg,SchoenbergContra}.
-These rules were used in the music piece \emph{Conta ponto}.~\cite{MASSA}
+importance lies mainly in the highest and then in the lowest of the voices~\cite{Fux,Tragtenberg,SchoenbergContra}.
+
+These rules were used in the music piece \emph{Conta ponto}, whose source code is
+available online in \massa\ toolbox.
 
 \subsection{Rhythm}\label{subsec:ritmo}
-Rhythm is dependent on events separated by durations~\cite{Lacerda}.
-For the temporal separation between them to be perceived as duration,
-the period should be at least around $75ms$~\cite{microsound}.
+Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}, which can be heard individually if spaced by at least $50-63ms$.
+For the temporal separation between them to be perceived as duration, the period should even a bit larger, around $100ms$~\cite{microsound}.
 It is possible to summarize the durations heard as rhythm or pitch, and its transition,
@@
 
+The transition span in Table~\ref{tab:duracoes} is minimized because the limits
+are not well defined. In fact, the duration where someone begins to perceive a
+fundamental frequency or a separation between occurrences, depends on the
+listener and sonic characteristics~\cite{microsound,Roederer}. The rhythmic metric is commonly based on a basic duration called pulse, whose typical durations range between $0.25 and 1.5s$ ($240$
+and $40 BPM$, respectively~\footnote{BPM stands for Beats Per Minute and is just a frequency measure like Herz, but is the number of incidences per minute intead of second.
+BPM is often used as a measure of musical tempo and of heart rate.}).
+In music education and cognitive studies, it is common to associate this range of frequencies with the durations of the heart beat, movements of respiration and steps of a walking or running person~\cite{Lacerda,Roederer}.
+
+The pulse is subdivided into equal parts and is also repeated in sequence. These relations (division and concatenation) usually follow relations of small
+integers. By far, the most often musical pulse divisions (and their sequential groupings), in written and ethnic
+music, are: 2, 4 and 8; 3, 6 (two groups of 3 or 3 groups of 2), 9 and 12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, completing
+1-9 and 12.
+Other metrics are less common, like division or grouping in 13, 17, etc, and are mainly used in experimental music or classical music of the XX and XXI centuries.
+No matter how complex they seem, metrics are almost always compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
+A schematic illustration is shown in Figure~\ref{fig:pulsoSubAgl}.
+
 \begin{figure*}
@@
         pulse, is presented on the left. The time signature yielded by
-	groupings of the music pulse is presented on the right~\protect\footnotemark.}
+        groupings of the music pulse is presented on the right.}
         \label{fig:pulsoSubAgl}
 \end{figure*}
-
-\footnotetext{The
-	definition of the terms 'quarter note' and 'time signature' are beyond the scope of this work and might be found in elementary texts about musical theory~\cite{Lacerda}.
-	Anyway, what is important for the reader to grasp is that: 1) successive temporal elements of equal duration
-	are both grouped and subdivided into other elements of equal duration;
-	2) these groupings and subdivisions are usually made in terms of small integers and are discerned into
-	simple, compound and complex.}
-
-The transition span in Table~\ref{tab:duracoes} is minimized because the limits
-are not well defined,
-i.e. the exact duration that is perceived as a
-frequency or a separation between occurrences depends on the
-listener and the sonic characteristics~\cite{microsound,Roederer}.
-The rhythm is commonly based on a basic duration called pulse, whose typical duration range between $0.25$ and $1.5s$ ($240$
-and $40 BPM$, respectively~\footnote{BPM stands for Beats Per Minute and is just a frequency measure like Herz, but is the number of incidences per minute instead of second.
-BPM is often used as a measure of musical tempo and of heart rate.}).
-In music education and cognitive studies, this range of frequencies is often associated with the duration of the heart beat, movements of respiration and steps of a walking or running person~\cite{Lacerda,Roederer}.
-
-The pulse is subdivided into equal parts and is also repeated in sequence. These relations (division and concatenation) usually follow relations of small
-integers. By far, the most often musical pulse divisions (and their sequential groupings), in written, popular and ethnic
-music, are: 2, 4 and 8; 3, 6 (2 groups of 3 or 3 groups of 2), 9 and 12 (3 and 4 groups of 3). At last, the prime numbers 5 and 7, completing
-1-9 and 12. Other metrics are less common, like division or grouping in 13, 17, etc, and are mainly used in experimental music or classical music of the XX and XXI centuries.
-A schematic illustration is shown in Figure~\ref{fig:pulsoSubAgl}.~\cite{Gramani,Roederer}
 
@@
 weak, weak) and weak (also subdivided in strong, weak, weak). Another way to perform
-the division in 6 is through a ternary division that subdivides as binary,
+the division in 6 is ternary division units that subdivides as binary,
 resulting in: a strong unit (subdivided in strong and weak) and two weak units
@@
 
-An accent in the weak beat is the `backbeat', whereas a note starting in a weak beat and persisting across the strong beat is a `syncope'.
+An accent in the weak beat is the `backbeat', whereas a note starting in a weak beat persisting across the strong beat is a `syncope'.
 These are often found in ethnic and popular music and was used with parsimony in classical music before the XX century.
@@
 Let the pulse be the grouping level $j=0$, the first pulse subdivision be level $j=-1$,
-the first pulse agglomeration be level $j=1$ and so on. Accordingly, let $P_i^j$ be the $i$-th unit at grouping level $j$: $P^0_{10}$ is the tenth pulse, $P^{1}_3$ is the third grouped unit,
+the first pulse agglomeration be level $j=1$ and so on.
+Accordingly, let $P_i^j$ be the $i$-th unit at grouping level $j$: $P^0_{10}$ is the tenth pulse, $P^{1}_3$ is the third grouped unit (possibly the third measure),
 $P^{-1}_2$ is the second part of pulse subdivision.
@@
 As an example, consider $P^{-1,0,1}_{3,2,2}$ as the third subdivision $P^{-1}_3$ of the
-second pulse $P^0_2$ and of the second pulse group $P^1_2$.
-Each unit $P_i^j$ can be associated with a sequence of temporal samples $T$ that constitutes a
-note. The music piece \emph{Poli Hit Mia} uses different rhythmic metrics.~\cite{MASSA}
+second pulse $P^0_2$ and of the second pulse group $P^1_2$ (possibly second measure). Each unit $P_i^j$ can be associated with a sequence of temporal samples $T_i$ that constitutes a
+note. The music piece \emph{Poli Hit Mia} uses different metrics (available as part of \massa\ toolbox).
 
 \subsection{Repetition and variation: motifs and larger units}\label{subsec:motivos}
-Given the basic musical structures of the previous sections, it is
+Given the basic musical structures, both frequential (chords and scales) and rhythmic (simple, compound and complex beat divisions and agglomerations), it is
 natural to present these structures in a coherent and meaningful way~\cite{Boulez}.
-The concept of arc is essential in this context: by departing from a place and returning, an arc is made.
+The concept of arcs is essential in this context: by departing from a place and returning, an arc is made.
 One important case is the arc from and to the absence of a unit: from the beginning to the end.
@@
 Accordingly, the note can be considered the smallest (relevant) arc, and each motif and melody as an arc as well.
-Each beat and subdivision, each music and
-section, constitutes an arc.
-Music in which the arcs do not present consistency with one another can be understood as music with little or no coherence. Coherence impression
+Each beat and subdivision, each measure and music
+section, constitutes an arc. Music in which the arcs do not present consistency with one another can be understood as music with no coherence. Coherence impression
 comes, mostly, from the skilled handling of arcs in a music piece.
@@
 prone to variations. Let
-$S=\left\{s_j=T^j=\{t_i^{j}\}_0^{\Lambda_j-1}\right\}_0^{H-1}$ be a sequence
+$S_j=\left\{s_j=T_i^j=\{t_i^{j}\}_0^{\Lambda_j-1}\right\}_0^{H-1}$ be a sequence
 of $H$ musical events $s_j$, each event with its $\Lambda_j$ samples $t_i^j$
@@
     where $\Gamma$ is the duration between the beginning of the piece (or considered
-    section) and the first event $s_0$ of the original structure $S$, and
-    $\delta$ is the temporal offset of the displacement.
+    section) and the first event $s_0$ of original structure $S_j$, and
+    $\delta$ is the time offset of the displacement.
 
@@
     \item Temporal reversion consists on generating a sequence with elements
-    in the reverse order of the original sequence $S$, thus:
-		$S'=\left\{s_j'\right\}_0^{H-1}=\left\{s_{(H-j-1)}\right\}_0^{H-1}$.
+    in the reverse order of the original sequence $S_j$, thus: $S_j'=\left\{s_j'\right\}_0^{H-1}=\left\{s_{(H-j-1)}\right\}_0^{H-1}$.
 
     \item Pitch translation, or transposition, is a displacement $\tau$ of the material to a
-        different pitch $\Xi'=\Xi + \tau$. It is a variation that changes pitch
+        different pitch $\Xi=\Xi_0 + \tau$. It is a variation that changes pitch
         localization of material:
-        $\left\{s_j'\right\}=\left\{s_j^{\Xi'}\right\}=\left\{s_j^{\Xi+\tau}\right\}$.
-        If $\tau$ is given in semitones, new frequencies are given by $f'=f2^{\frac{\tau}{12}}$; 
-        and the new pitches: $\Xi'=\Xi +12 \log_2\left(\frac{f'}{f}\right)=\Xi+\tau$.\footnote{In
-		the MIDI protocol, $\Xi_{f}=55Hz$ when pitch $\Xi=33$
-	(an \textit{A1} note). Another good MIDI reference is $\Xi_{f}=440Hz$ and
-		$\Xi=69$ (\textit{A4}). The difference ($\tau=\Xi_1 - \Xi_2$) is in semitones.
-        Notice that $\Xi=1$ is not a semitone, it is a note with an audible frequency as rhythm, with
+        $\left\{s_j'\right\}=\left\{s_j^{\Xi'}\right\}=\left\{s_j^{\Xi+\tau}\right\}$
+        where $\Xi_0$ is a reference value, such as the pitch of a section $S_j$ or of the first event $s_0$.
+        If $\tau$ is given in semitones, the displacement in
+        frequency is $\tau_f=f_0.2^{\frac{\tau}{12}}$ where $f_0$ is the
+        reference frequency value: $f_0=\Xi_{f_0}Hz\;\sim \Xi$ absolute value of pitch. The frequency of any pitch value is $f=f_0.2^{\frac{\tau}{12}}$s; 
+        and the pitch of any frequency is: $\Xi=\Xi_0 +12
+		\log_2\left(\frac{f}{f_0}\right)$.\footnote{In the MIDI protocol, $\Xi_{f_0}=55Hz$ when pitch $\Xi_0=33$
+	(an \textit{A1} note). Another good MIDI reference is $\Xi_{f_0}=440Hz$ and
+		$\Xi_0=69$ (\textit{A4}). The difference ($\Xi_1 - \Xi_2$) is in semitones.
+        $\Xi$ is not a measure in semitones: $\Xi=1$ is not a semitone, it is a note with an audible frequency as rhythm, with
 		less than 9 occurrences each second (see Table~\ref{tab:duracoes}).}
 
-        \item Interval inversion is either: 1) the inversion of note pitch order, within octave equivalence, such as described in Section~\ref{subsec:intervalos}; or 2) the inversion of interval orientation.
+        \item Interval inversion is either: 1) the inversion of note pitch order, within octave equivalence, such as described in Section~\ref{subsec:intervalos}, and $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{\varepsilon.
+f_0}\right\}$ with selective $s_j=2,\;1/2,\;\text{or}\;1$; or 2) the inversion of interval orientation.
 In the former case, the number of semitones
         is preserved in the ``strict invertion'':
-        $S'=\{s_j'\}_0^{H-1}=\left\{s_j^{-\varepsilon_j . f_0}\right\}$, where
+        $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{-\varepsilon_j . f_0}\right\}$, where
         $\varepsilon_j$ is the factor between the frequency of event $s_j$ and the
@@
         considered in terms of the diaonic scale $E_k$:
-        $S'=\{s_j'\}_0^{H-1}=\left\{s_j^{2^{\left(\frac{12-e_{\left(7-j_e\right)}}{12}\right)}
+        $S_j'=\{s_j'\}_0^{H-1}=\left\{s_j^{2^{\left(\frac{12-e_{\left(7-j_e\right)}}{12}\right)}
         . f_0}\right\}_0^{H-1}$ where $j_e$ is the index in
@@
         with events which are out of the slots. Thus, a
-        rotation of $\tilde{n}$ positions is $s'_n=s_{(n+\tilde{n})\%H}$. If
-        $\tilde{n}<0$, it is sufficient to use $\tilde{n}'=H-\tilde{n}$.
-		One might associate $\tilde{n}>0$ (events advance) with the anti-clockwise rotation and
-        $\tilde{n}<0$ (elements delay) with the clockwise rotation.
+        rotation of $\tilde{n}$ positions is $S'_n=S_{(n+\tilde{n})\%H}$. If
+        $\tilde{n}<0$, it is sufficient to use $\tilde{n}'=H-\tilde{n}$. It is
+        usual to associate $\tilde{n}<0$ (events advance) with the clockwise rotation and
+        $\tilde{n}<0$ (elements delay) with the anti-clockwise rotation.
         Additional information about rotations is given in Section~\ref{estCic}.
 
-        \item The insertion and removal of material in $S$ can be
-    ornamental or structural: $S'=\{s_j'\}=\{s_j \text{ if condition A,
+        \item The insertion and removal of material in $S_j$ can be
+    ornamental or structural: $S_j'=\{s_j'\}=\{s_j \text{ if condition A,
     otherwise } r_j\}$, for any music material $r_j$, including silence.
-    Elements of $R$ can be inserted at the beginning, as a prefix;
-		at the end, as a suffix; or in the middle, dividing $S$ and making
-    it the prefix and suffix of $R$. Both materials can be mixed in a variety of ways.
+    Elements of $R_j$ can be inserted at the beginning, like a prefix
+    in $S_j$; at the end, as a suffix; or in the middle, dividing $S_j$ and making
+    it the prefix and suffix. Both materials can be mixed in a variety of ways.
 
@@
 
-    \item Accompaniment.
-	    E.g. orchestration and melodic lines present when $S$ occurs can 
-	    be modified and be considered a variation of $S$ itself.
+    \item Accompaniment. Both orchestration and melodic lines present when $S_j$ occurs can suffer modifications and be considered as a variation of $S_j$ itself.
 \end{itemize}
 
-From these processes many others are derived, such as the inverted retrograde, the temporal contraction with an external suffix, etc.
+From these processes, many others are derived, such as the inverted retrograde, the temporal contraction with an external suffix, etc.
 Variations are often thought about in the terms above but are also often very loose, such as an arbitrary shuffle of the notes in a melody which the composer finds interesting.
@@
 musical therapy, known by its utility in cases of depression and neurological injury.
-Also, it is known that regions of the human brain responsible for sonic processing are also used for other activities, such as for performing verbal discourse and mathematics.~\cite{Sacks,Roederer}
+Also, it is known that regions of the human brain responsible for sonic processing are also used for other activities, such as for performing verbal discourse and mathematics~\cite{Sacks,Roederer}.
 
@@
 equivalence/difference, arrival/departure, near/far, stationary/moving,
-etc. All these dipoles are often thought of as parallel or even as equivalent.
-Ternary constructions tend to relate to the circle and to unification. The
+etc. All these dipoles are often thought of as parallel or even as equivalent. Ternary relations tend to relate to the circle and to unification. The
 lucid ternary communion, `modus perfectus', opposes to the passionate
@@
 directional and cyclic arcs.
+
 For a scholastic discussion on the composition of motives, phrases, melodies, themes and music form (such as rondo, ternary, theme and variations) see~\cite{Schoenberg}.
@@
 shown in Figure~\ref{fig:climax}. The varying parameter can be non-existent, a case in which
-the arc consists only of a reference structure~\cite{Schoenberg}\footnote{A
-case which resembles a note without the fundamental frequency. In some cases, the listener is able to identify
-the fundamental frequency even though it is absent.}.
+the arc consists only of a reference structure~\cite{Schoenberg}\footnote{A case which resembles a note without the fundamental frequency.
+Brass instruments can emit such notes and the listener identifies the fundamental frequency even though it is absent.}.
 
@@
         other arcs. The possibilities considered are: climax at the beginning, at the first half, in the middle, in the second half and 
-        at the end.}
+        at the end. The x and y-axis are not properly specified since the parameters can be non-existent, as in a reference structure.}
         \label{fig:climax}
@@
 
-Consider the $S=\{s_i\}_0^{H-1}$ increasing sequence. The sequence
-$R=\{r_i\}_0^{2H -2}=\left\{s_{(H-1-|H-1-i|)}\right\}_0^{2H-2}$
+Consider the $S_i=\{s_i\}_0^{H-1}$ increasing sequence. The sequence
+$R_i=\{r_i\}_0^{2H -2}=\left\{s_{(H-1-|H-1-i|)}\right\}_0^{2H-2}$
 presents perfect specular symmetry, i.e.\ the second half is the
@@
 
-In practice, and historically, there is special use of the golden ratio.
-The Fibonacci sequence might be generalized as follows in order to use any two numbers
+In practice, and historically, there is special incidence and use of the golden ratio.
+The Fibonacci sequence might be generalized as follows in order for any two numbers be used
 to approach the golden ratio.
@@
 for exploring various sequences.
-The musical piece \emph{Dirracional} explores directional arcs,~\cite{MASSA}
+
+The music piece \emph{Dirracional} uses arcs into directional structures. Its source code is available online as part
+of \massa\ toolbox~\cite{MASSA}.
 
@@
      &  \text{(existence of the identity element)} \\
-\forall \;\; p \in G, \;\exists\; p^{-1} :\quad  &  p\bullet p^{-1} =p^{-1}\bullet p = e \\
+\forall \;\; p \in G, \;\exists\; p^{-1} :\quad  &  p\bullet p^{-1}i =p^{-1}\bullet p = e \\
      &  \text{(existence of the inverse element)}
@@
 From the first property follows that two permutations act as one permutation. In fact, it is possible to apply a
-permutation $p_1$ and then another permutation $p_2$, and, comparing both initial and final orderings, observe another permutation $p_3$.
+permutation $p_1$ and another permutation $p_2$, and, comparing both initial and final orderings, observe another permutation $p_3$.
 Every element $p$ operated with itself a sufficient number of times $n$ reaches the identity element $p^n=e$ (otherwise the group generated by $p$ would be infinite).
@@
 
-As a historic
+These arcs can be established by using a set of permutations. As a historic
 example, the \emph{change ringing} tradition conceives music through
@@
 \centering
-\caption{A traditional change ringing \emph{peal} for 3 bells. Permutations
-occur between each line. Each line is a bell ordering to be played before next line.} 
+\caption{Change Ringing: Traditional \emph{peal} for 3 bells. Permutations
+occur between each line. Each line is a bell ordering and each ordering is played at a time.} 
 \begin{tabular}{l c r}
@@
 The use of permutations in music can be summarized in the following way:
-with $S=\{s_i\}$ a sequence of musical events $s_i$ (e.g.\ notes), and a
-permutation $p$. $S'=p(S)$ comprises the same elements of $S$ but in a
+with $S_i=\{s_i\}$ a sequence of musical events $s_i$ (e.g.\ notes), and a
+permutation $p$. $S_i'=p(S_i)$ comprises the same elements of $S_i$ but in a
 different order. Permutations have two notations: cyclic and
@@
 produces (ex. $[1\;3\;7\;0\;...]$). In the cyclic notation, a permutation consists
-of swaps between elements and its successors, and between the last element and the first.
+of swaps by elements and its successors, and the last element by the first one.
 E.g. $(1,2,5)(3,4)$ in cyclic notation is equivalent to $[0,2,5,4,3,1]$ in natural notation.
 
-In the auralization of a permutation, it is not necessary to permute elements of $S$,
-but only some characteristic. Thus, if $p$ is a permutation and $S$ is a sequence of basic notes as in the end of Section~\ref{notaBasica}, the
-sequence $S'=p^f(S)=\left\{s_i^{p(f)}\right\}$ consists of the same
+In the auralization of a permutation, it is not necessary to permute elements of $S_i$,
+but only some characteristic. Thus, if $p$ is a permutation and $S_i$ is a sequence of basic notes as in the end of Section~\ref{notaBasica}, the
+sequence $S_i'=p^f(S_i)=\left\{s_i^{p(f)}\right\}$ consists of the same
 music notes, following the same order and maintaining the same characteristics, but with the
@@
 Two subtleties of this procedure should be commented upon. 
-First, a permutation $p$ is not restricted to involve all elements of $S$, i.e.\ it can operate in a subset of $S$.
-Second, not all elements $s_i$ need to be executed at each access to $S$. To exemplify, let $S$ be a sequence of music notes $s_i$. 
+First, a permutation $p$ is not restricted to involve all elements of $S_i$, i.e.\ it can operate in a subset of $S_i$.
+Second, not all elements $s_i$ need to be executed at each access to $S_i$. To exemplify, let $S_i$ be a sequence of music notes $s_i$. 
 If $i$ goes from $0$ to $n$, and
@@
 notes.
-The other notes of $S$ can occur in other measures where permutations 
-allocate those notes to the first four events $s'_i$ of $S'$.
+The other notes of $S_i$ can occur in other measures where permutations 
+allocate those notes to the first four events $s'_i$ of $S'_i$.
 
 To each permutation $p_i$ described above, we have to determine:
-1) note characteristics where it operates (frequency, duration, \emph{fades},
-intensity, timbre, etc); and
+1) note dimensions where it operates (frequency, duration, \emph{fades},
+intensity, etc); and
 2) the period of incidence (how much consults before a permutation is
-applied).
-For a coherent realization of the notes in $S$ is to
-execute the first $n$ notes given that the execution of disjoint sets of $S$
+applied). During realization of notes in $S_i$, an easy and coherent form is to
+execute the first $n$ notes, the execution of disjoint sets of $S_i$
 is the same as modifying the permutation and executing the first $n$ notes.
-The PPEPPS is a set of musical pieces that uses permutations.~\cite{MASSA,figgusOriginal,figgusEspacializacao}.
-
-\subsection{Serialism and post-serial techniques}\label{sec:ser}
+
+The \massa\ toolbox presents a computational implementation that isolates
+permutation application to sonic characteristics, in order to deliver
+musical structures~\cite{MASSA,figgusOriginal,figgusEspacializacao}.
+
+\subsection{Serialism and post-serial techniques}
 Expanding the concepts in Section~\ref{sec:atonal},
@@
 Furthermore, software for synthesis and processing of sounds for musical purposes
-by standard restrict the bit depth to 16 or 24.
+by standard restricts the bit depth to 16 or 24.
 This is achievable in this framework but by standard Python uses more bits per floating point number.

### 2017-09-22 article/acm/listings.tex
temporal brench to move changes to other lap

--- 2017-07-30 article/acm/listings.tex
@@
 
+\newcommand*{\reff}[1]{%
+	{\NoHyper\ref{#1}\endNoHyper}%
+	  }
+
 \def\thesection{SI-\arabic{section}}
@@
 \author{\\\\\\ Renato Fabbri (USP),\\
-Vilson Vieira da Silva Junior (The Grid),\\
+Vilson Vieira da Silva Junior (Cod.ai),\\
 Ant\^onio Carlos Silvano Pessotti (Unimep),\\
@@
 
+\begin{table*}[htp!]
+\centering
+\caption{Equation numbers and their descriptions.
+All these equations are implemented in file \texttt{src/sections/2.py}.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \reff{eq:dur} & relation between number of samples and duration \\
+ \reff{eq:potencia} & power of the wave \\
+ \reff{decibels} & decibels by difference by means of the power of each wave \\
+ \reff{eq:ampVol} & double amplitude implies $\approx 6dB$ \\
+ \reff{eq:potVol} & double power implies $\approx 3dB$ \\
+ \reff{eq:dobraVol} & double volume ($10dB$) implies a factor of $\approx 3.16$ for amplitude \\
+ \reff{ampDec} & direct relation between variations in amplitude and decibels \\
+ \reff{periodicidade} & equivalences in a periodic sound with respect to wavelength, frequency and sample rate \\
+ \reff{sinusoid} & sample amplitudes in a sinusoid \\
+ \reff{sawTooth} & sample amplitudes in a sawtooth wave \\
+ \reff{triangular} & sample amplitudes in a triangular wave \\
+ \reff{square} & sample amplitude in a square wave \\
+ \reff{sampleandoFormaDeOnda} & samples in a sound derived from a sampled wave period \\
+ \reff{recomposicaoFourier} & reconstruction of samples from the Fourier components \\
+ \reff{moduloEfase} & reconstruction of real samples (e.g. for audio) from Fourier components \\
+ \reff{coefsPareados} & number of pairs of Fourier coefficients which are related to the same frequency \\
+ \reff{equivalenciasFreqs} & indexes of equivalent frequencies and coefficients for real signals \\
+ \reff{equivalenciasModulos} & equal modules between samples of real signals \\
+ \reff{equivalenciasFases} & equivalence in phases between samples of real signals \\
+ \reff{eq:reconsCompleta} & complete reconstruction of the signal using Fourier components and previous equations \\
+ \reff{eq:notaBasica} & sample sequence related to a basic note \\
+ \reff{periodoUnico} & sample sequence of a period of an arbitrary waveform \\
+ \reff{eq:notaBasicaTimbre} & note samples derived from a sampled waveform \\
+ \reff{eq:distOuvidos} & distance of a (sound) source to each ear given the distance between the ears and an (x,y) position of the source \\
+ \reff{eq:dti} & Interaural Time Difference (ITD), the time difference of a sound reaching each ear \\
+ \reff{eq:dii} & Interaural Intensity Difference (IID), the intensity difference (in decibels) of a sound reaching each ear \\
+ \reff{eq:locImpl} & ITD and IID in terms of sample delays and their amplitudes \\
+ \reff{eq:angulo} & azimuthal angle of a (x, y) source \\
+ \reff{eq:mixagem} & samples that result from mixing sounds \\
+ \reff{eq:concatenacao} & samples that result from concatenating sounds \\
+\end{tabular}
+\end{table*}
+
+\begin{table*}[htp!]
+\centering
+\caption{Equation numbers and their descriptions.
+All these equations are implemented in file \texttt{src/sections/3.py}.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \reff{eq:lut} & sample sequence generated by means of a lookup table (LUT) \\
+ \reff{freqLinear} & frequency at each sample in a linear transition of frequency \\
+ \reff{indiceLinear} & indices of a LUT in a linear transition of frequency \\
+ \reff{serieAmostralLin} & sample sequence obtained through a LUT in a linear transition of frequency \\
+ \reff{freqExponencial} & frequency at each sample in an exponential transition of frequency (linear pitch) \\
+ \reff{indiceExponencial} & indices of a LUT in an exponential transition of frequency (linear pitch) \\
+ \reff{serieAmostralLog} & sample sequence obtained through a LUT in an exponential transition of frequency \\
+ \reff{seqAmp} & amplitude factors at each sample in an exponential transition of amplitude ($\approx$ linear volume) \\
+ \reff{transAmp} & sample sequence with an exponential transition of amplitude ($\approx$ linear volume) \\
+ \reff{seqAmpLin} & amplitude factors at each sample in a linear transition of amplitude \\
+ \reff{seqAmpDB} & sample sequence in an exponential transition of amplitude ($\approx$ linear volume) with difference given in decibels \\
+ \reff{eq:conv} & sample sequence obtained through the convolution of two other sequences (e.g. for applying FIR filters) \\
+ \reff{eq:diferencas} & difference equation (e.g. for applying IIR filters) \\
+ \reff{eq:passa-baixas} & IIR coefficients for a simple, useful and well-behaved low-pass filter \\
+ \reff{eq:passa-altas} & IIR coefficients for a simple, useful and well-behaved high-pass filter \\
+ \reff{eq:varAux} & auxiliary variables for the following band-pass and band-reject filters \\
+ \reff{eq:passa-banda} & IIR coefficients for a simple, useful and well-behaved band-pass filter \\
+ \reff{eq:rejeita-banda} & IIR coefficients for a simple, useful and well-behaved band-reject filter \\
+ \reff{eq:branco} & Fourier coefficients of a white noise \\
+ \reff{eq:rosa} & Fourier coefficients of a pink noise \\
+ \reff{eq:marrom} & Fourier coefficients of a brown noise \\
+ \reff{eq:azul} & Fourier coefficients of a blue noise \\
+ \reff{eq:violeta} & Fourier coefficients of a violet noise \\
+ \reff{eq:preto} & Fourier coefficients of a black noise \\
+ \reff{vbrGamma} & indices for a vibrato given its frequency and using a LUT \\
+ \reff{vbrAux} & samples for applying a vibrato \\
+ \reff{vbrF} & frequency at each sample of a sound with vibrato \\
+ \reff{vbrGamma2} & indices for LUT in a sound with vibrato \\
+ \reff{vbrT} & sample sequence of a sound with vibrato \\
+ \reff{trA} & amplitude at each sample for a tremolo \\
+ \reff{trT} & sample sequence of a sound with tremolo \\
+ \reff{eq:fmEsp} & components in FM synthesis when both modulator and carrier are sines \\
+ \reff{eq:Bessel} & the Bessel function \\
+ \reff{eq:specAM} & components in AM synthesis when both modulator and carrier are sines \\
+ \reff{fmGammaAux} & indices for LUT in modulator of an FM synthesis \\
+ \reff{fmAux} & sample sequence of a modulator in an FM synthesis \\
+ \reff{fmF} & frequeny at each sample of a sound derived from FM synthesis \\
+ \reff{fmGamma} & indices for the final signal in FM synthesis using LUT \\
+ \reff{fmT} & sample sequence of a sound generated through FM and using LUT \\
+ \reff{amA} & amplitude at each sample in a sound generated though AM \\
+ \reff{amT} & sample sequence of a sound generated through AM and using LUT \\
+\end{tabular}
+\end{table*}
+
+\begin{table*}[htp!]
+\centering
+\caption{Equation numbers and their descriptions.
+All these equations are implemented in file \texttt{src/sections/3.py}.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \reff{eq:vinculos} & an example of bonds between musical characteristics \\
+ \reff{eq:fDoppler} & relation between frequencies and speed in the Doppler effect \\
+ \reff{eq:aDoppler} & relation between position, speed and amplitude in the Doppler effect \\
+ \reff{eq:ffDoppler} & relation between position, speed and amplitude in the Doppler effect \\
+ \reff{eq:p1rev} & samples of a FIR filter for the first period of a reverberation \\
+ \reff{eq:p2rev} & samples of a FIR filter for the second period of a reverberation \\
+ \reff{eq:rev} & samples of the FIR filter for a reverberation (considering both first and second periods) \\
+ \reff{eq:adsr} & amplitude factors for each sample in an ADSR envelope \\
+ \reff{eq:adsrApl} & sample sequence of a sound with an ADSR envelope \\
+\end{tabular}
+\end{table*}
+
+\begin{table*}[htp!]
+\centering
+\caption{Equation numbers and their descriptions.
+All these equations are implemented in file \texttt{src/sections/4.py}.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \reff{escSim} & perfectly symmetric scales in each octave with the twelve semitones \\
+ \reff{eq:escalas} & diatonic scales \\
+ \reff{eq:relacaoDia} & the succession of tones and semitones of a diatonic scale \\
+ \reff{eq:escalasMenores} & sequences of semitones for the three types minor scales \\
+ \reff{eq:serieHarmonica} & harmonic series in terms of semitones \\
+ \reff{triades} & triads (chords constituted by thirds) \\
+ \reff{eq:rhythmicUnit} & a convention to specify a unit of rhythmic division or agglomeration \\
+ \reff{eq:groups} & definition of algebraic groups \\
+\end{tabular}
+\end{table*}
+
 \clearpage
-\section{Figures}
+\section{Figures}\label{sec:lfigures}
+
+\begin{table*}[htp!]
+\centering
+\caption{Figure numbers and their descriptions.
+All these equations are implemented in files \texttt{src/aux/*}.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \reff{fig:PCM} & PCM PCM audio (discrete and digital) samples \\
+ \reff{fig:formasDeOnda} & synthetic and sampled waveforms \\
+ \reff{fig:espectroDeOndas} & spectrum of basic waveforms \\
+ \reff{fig:espectroOboe} & spectrum of a real note and of one derived from one sampled period \\
+ \reff{fig:amostras2} & sinusoid represented by two samples \\
+ \reff{fig:amostras3} & sinusoid represented by three samples \\
+ \reff{fig:amostras4} & sinusoids in 4 samples \\
+ \reff{fig:formas4} & basic waveforms within 4 samples \\
+ \reff{fig:amostras6} & sinusoids in 6 samples \\
+ \reff{fig:formas6} & basic waveforms withing 6 samples \\
+ \reff{fig:spac} & ITD and IID (spatial localization cues)\\
+ \reff{fig:mixagem} & mixing PCM audio \\
+ \reff{fig:concatenacao} & concatenation/juxtaposition of PCM audio \\
+ \reff{fig:lut} & Lookup table \\
+ \reff{fig:transicao} & transitions of intensity \\
+ \reff{fig:conv} & convolution \\
+ \reff{fig:delays} & convolution yielding time shifting, multiple time delays, sound amalgam \\
+ \reff{fig:iir} & frequency response of useful IIR filters \\
+ \reff{fig:ruidos} & spectrum and waveform of noises \\
+ \reff{fig:vibrato} & spectrogram of a vibrato \\
+ \reff{fig:tremolo} & waveform of a tremolo \\
+ \reff{fig:adsr} & ADSR envelope \\
+ \reff{fig:movContraponto} & counterpoint movements \\
+ \reff{fig:pulsoSubAgl} & musical metric in terms of divisions of temporal units \\
+ \reff{fig:climax} & distinctions of musical climax by localization \\
+\end{tabular}
+\end{table*}
 
@@
 
+\begin{table*}[htp!]
+\centering
+\caption{Table numbers and their descriptions.}
+\begin{tabular}{ c | p{12cm} }
+   Number & Description \\\hline
+ \reff{eq:intervalos} & musical intervals, their notations and qualities \\
+ \reff{tab:harmonia} & tonal harmonic functions in the major scale \\
+ \reff{tab:duracoes} & duration scales yielding perception of pitch and rhythm \\
+ \reff{tab:change} & music by permutation of units (change ringing) \\
+\end{tabular}
+\end{table*}
+
 \clearpage
 \section{Scripts}
-\subsection{For all equations in each chapter}
+\subsection{For all equations and relations in each section.}
+
+\begin{table*}[htp!]
+\centering
+\caption{Script files and their descriptions.}
+\begin{tabular}{ c | p{12cm} }
+   Filename & Description \\\hline
+  \texttt{src/sections/2.py} & implementation of all the equations for the basic note in PCM audio in Section~\ref{sec:discNote} \\
+  \texttt{src/sections/3.py} & implementation of all the equations for variations within a note described in Section~\ref{sec:varInternas} \\
+  \texttt{src/sections/4.py} & implementation of all the techniques for assembling notes into music described in Section~\ref{sec:notesMusic} \\
+\end{tabular}
+\end{table*}
+
 \clearpage
 \subsection{To render musical pieces}
+
+\begin{table*}[htp!]
+\centering
+\caption{Piece names, script files and the concepts they exemplify from Section~\ref{sec:discNote}. All files are in the directory \texttt{src/pieces2/}.}
+\begin{tabular}{ p{5cm} | p{3.5cm} | p{6.2cm} }
+   Name & Filename & Description \\\hline
+ Reduced-fi & \texttt{reduced-fi.py} & concatenation of simple notes \\
+ Sonic pictures & \texttt{quadrosSonoros.py} & mixing of simple notes  \\
+\end{tabular}
+\end{table*}
+
+\begin{table*}[htp!]
+\centering
+\caption{Piece names, script files and the concepts they exemplify from Section~\ref{sec:internalVar}. All files are in the directory \texttt{src/pieces3/}.}
+\begin{tabular}{ p{5cm} | p{3.5cm} | p{6.2cm} }
+   Name & Filename & Description \\\hline
+ ADa and SaRah & \texttt{ADAandSaRah.py} & ADSR envelope \\
+ Tremolos, Vibratos and Frequency & \texttt{bonds.py} & bonds between tremolos, vibratos and frequency  \\
+ Bella Rugosi & \texttt{bellaRugosi.py} & rugosity achieved through frequencies between 15 adn 30 Hz \\
+ Children Choir & \texttt{childChoir.py} & achieving choir sonorities by small variations of the same basic note \\
+ Noisy Band & \texttt{noisyBanda.py} & using various noises \\
+ ParaMeter Transitions & \texttt{paraMeter.py} & gradual changes of parameters within one note \\
+ Little Train of Impulsive Hillbilies~\footnote{This names translates into ``Trenzinho de Caipiras Impulsivos'' which is a pun with the Villa-Lobos' ``Trenzinho do Capira''.} & \texttt{littleTrain.py} & the use of convolution with impulses to achieve rhythm \\
+ Shakes and Wiggles & \texttt{shakesWiggles.py} &  tremolos and vibratos \\
+\end{tabular}
+\end{table*}
+
+\begin{table*}[htp!]
+\centering
+\caption{Piece names, script files and the concepts they exemplify from Section~\ref{sec:notesMusic}. All files are in the directory \texttt{src/pieces4/}.}
+\begin{tabular}{ p{5cm} | p{3.5cm} | p{6.2cm} }
+   Name & Filename & Description \\\hline
+ Acorde Cedo & \texttt{acordeCedo.py} & chord successions and modulation \\
+ Conta Ponto & \texttt{contaPonto.py} & melodic lines conducted within the rules of counterpoint \\
+ Crystals & \texttt{crystals.py} & symmetric musical scales \\
+ Dirracional & \texttt{dirracional.py} & directional arcs \\
+ Intervals & \texttt{intervals.py} & musical intervals \\
+ MicroTone & \texttt{microTone.py} & microtonality (the use of intervals smaller than the semitone) \\
+ Poly-Hit-My & \texttt{polyHitMy.py} & polyrhythm (multiple metrics at once) \\
+\end{tabular}
+\end{table*}
+
 \clearpage
 \subsection{To render the figures used in the article}
+The files \texttt{src/figures/*} render each of the figures (listed above in Section~\ref{sec:lfigures}.
 
